### 2013-06-04 main.tex
Initial commit (working on chapter 2)

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[
		twoside,openright,titlepage,numbers=noenddot,headinclude,%1headlines,
                footinclude=true,cleardoublepage=empty,
                BCOR=10mm,paper=a4,fontsize=10pt, % Binding correction, paper type and font size
                ngerman,american, % Languages
                ]{scrreprt}

% Includes the file which contains all the document configurations and packages - make sure to edit this file
\input{classicthesis-config}
\usepackage{semantic}
\usepackage{paralist}
\usepackage{etoolbox}
\usepackage{stmaryrd}
\usepackage{amssymb}

\newtheorem{definition}{Definition}

% source code & types
\newcommand{\sep}[0]{\;\; | \;\;}
\newcommand{\kvd}[1]{\textnormal{\bfseries\sffamily #1}} %
\newcommand{\str}[1]{\textnormal{\ttfamily #1}} %
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\ctyp}[2]{C^{#1}#2}
\newcommand{\mtyp}[2]{M^{#1}#2}
\newcommand{\subst}[3]{#1[#2 \leftarrow #3]}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\catc}{\mathcal{C}}
\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }

\newcommand{\codesnip}[0]{\renewcommand{\baselinestretch}{0.9}\normalsize}
\newcommand{\codesnipout}[0]{\renewcommand{\baselinestretch}{1.0}\normalsize}

% Half line spacing around align
\newcommand{\zerodisplayskips}{%
  \setlength{\abovedisplayskip}{0.5em}
  \setlength{\belowdisplayskip}{0.5em}
  \setlength{\abovedisplayshortskip}{0.5em}
  \setlength{\belowdisplayshortskip}{0.5em}}
\appto{\normalsize}{\zerodisplayskips}
\appto{\small}{\zerodisplayskips}
\appto{\footnotesize}{\zerodisplayskips}

% half line spacing around compactitem
\setlength{\pltopsep}{0.5em}
\setlength{\plitemsep}{0.3em}

\begin{document}

\frenchspacing % Reduces space after periods to make text more compact

\raggedbottom % Makes all pages the height of the text on that page

\selectlanguage{american} % Select your default language - e.g. american or ngerman

%\renewcommand*{\bibname}{new name} % Uncomment to change the name of the bibliography
%\setbibpreamble{} % Uncomment to include a preamble to the bibliography - some text before the reference list starts

\pagenumbering{roman} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)

\pagestyle{plain} % Suppress headers for the pre-content pages

%----------------------------------------------------------------------------------------
%	PRE-CONTENT THESIS PAGES
%----------------------------------------------------------------------------------------

% \include{FrontBackMatter/Titlepage} % Main title page

% \include{FrontBackMatter/Titleback} % Back of the title page

% \cleardoublepage\include{FrontBackMatter/Dedication} % Dedication page

%\cleardoublepage\include{FrontBackMatter/Foreword} % Uncomment and create a Foreword.tex to include a foreword

% \cleardoublepage\include{FrontBackMatter/Abstract} % Abstract page

% \cleardoublepage\include{FrontBackMatter/Publication} % Publications from the thesis page

% \cleardoublepage\include{FrontBackMatter/Acknowledgments} % Acknowledgements page

\pagestyle{scrheadings} % Show chapter titles as headings

\cleardoublepage\include{text/Contents} % Contents, list of figures/tables/listings and acronyms

\pagenumbering{arabic} % Arabic page numbering for thesis content (1, 2, 3, etc)
%\setcounter{page}{90} % Uncomment to manually start the page counter at an arbitrary value (for example if you wish to count the pre-content pages in the page count)

\cleardoublepage % Avoids problems with pdfbookmark

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - CHAPTERS
%----------------------------------------------------------------------------------------

% \ctparttext{You can put some informational part preamble text here. Illo principalmente su nos. Non message \emph{occidental} angloromanic da. Debitas effortio simplificate sia se, auxiliar summarios da que, se avantiate publicationes via. Pan in terra summarios, capital interlingua se que. Al via multo esser specimen, campo responder que da. Le usate medical addresses pro, europa origine sanctificate nos se.} % Text on the Part 1 page describing  the content in Part 1

% \part{Some Kind of Manual} % First part of the thesis

% \include{Chapters/Chapter01} % Chapter 1

% \cleardoublepage % Empty page before the start of the next part

%------------------------------------------------

% \ctparttext{You can put some informational part preamble text here. Illo principalmente su nos. Non message \emph{occidental} angloromanic da. Debitas effortio simplificate sia se, auxiliar summarios da que, se avantiate publicationes via. Pan in terra summarios, capital interlingua se que. Al via multo esser specimen, campo responder que da. Le usate medical addresses pro, europa origine sanctificate nos se.} % Text on the Part 2 page describing the content in Part 2

% \part{The Showcase} % Second part of the thesis

\include{text/Chapter01} % Chapter 1
\include{text/Chapter02} % Chapter 2

% \include{Chapters/Chapter03} % Chapter 3
%\include{Chapters/Chapter04} % Chapter 4 - empty template

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - APPENDICES
%----------------------------------------------------------------------------------------

% \appendix

% \part{Appendix} % New part of the thesis for the appendix

% \include{Chapters/Chapter0A} % Appendix A
%\include{Chapters/Chapter0B} % Appendix B - empty template

%----------------------------------------------------------------------------------------
%	POST-CONTENT THESIS PAGES
%----------------------------------------------------------------------------------------

\cleardoublepage\include{Bibliography} % Bibliography
% \cleardoublepage\include{FrontBackMatter/Colophon} % Colophon
% \cleardoublepage\include{FrontBackMatter/Declaration} % Declaration

%----------------------------------------------------------------------------------------

\end{document}

### 2013-06-04 text/Chapter01.tex
Initial commit (working on chapter 2)

\chapter{Introduction} 
\label{ch:introduction}

%----------------------------------------------------------------------------------------

Some intro

\section{How lambda works}
The key difference - how lambda works

Effect systems, introduced by Gifford and Lucassen
\cite{effects-gifford}, track \emph{effects} of computations, such as
memory access or message-based
communication~\cite{effects-messagepassing}. Their approach augments
typing judgments with effect information: $\Gamma \vdash e :
\tau, F$.  Wadler and Thiemann explain how this shapes
effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the body.}
\end{quote}
%
In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
does not match this pattern. In the systems we consider, lambda abstraction places 
requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
(immediate requirements), resulting in different syntactic properties. 
We informally discuss three examples first that demonstrate how contextual
requirements propagate. Section ? then unifies these 
in a single calculus.

Also \cite{effects-producer-semantics}

\begin{quote}
\emph{We will define an effect as a producer effect if all computations with that effect 
can be thunked as "pure" computations for a domain-specific notion of purity.}
\end{quote}

Demonstrate using distributed computations

Demonstrate using dataflow/liveness

Perhaps mention LINQ as a motivation for cross-compilation..
\cite{app-linq}

\section{Associating with context}

-> Type providers make this important

%---------------------------------------------------------------------------------------------------

\section{Flat coeffect system}
~

resources?

%---------------------------------------------------------------------------------------------------

\section{Structural coeffect system}
\label{sec:structural}

The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
we need to track context-dependence in a more fine-grained way. To track neededness or security, 
we need to associate information with individual \emph{variables} of the context.

At the same time, we want to avoid developing multiple variants of coeffect systems -- indeed, 
our motivation is to develop a \emph{single unified mechanism} for tracking context-dependence.
In this section, we present a more powerful \emph{structural coeffect calculus}, which is a 
generalization of the flat calculus.

%We show that \clflt~can be obtained as a special case of \clstr. The
%structural version of our system is also where our work differs more significantly from well-known
%effect systems.

% --------------------------------------------------------------------------------------------------

\subsection{Motivation: Tracking array accesses}
\label{sec:structural-motivation}
\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
\newcommand{\dnat}{\ident{D}_\ident{nat}}
\newcommand{\cprd}{\times}
\newcommand{\cvop}{~\ident{max}~}

Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
respectively, but we use richer tag structure.

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\cprd$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
a single variable:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
Before looking at the formal definition, consider the typing of let bindings:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
\pastval{a}{15}+\pastval{c}{10}
\end{array}
\end{equation*}
%
The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.

The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
of $\cvop$ and it denotes empty coeffect. The typing context of the
body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.

To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
with the remaining tags from the body yielding the overall context:
$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
normalization mechanism (described later), this can be further reduced to
$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
\ident{b}.

%===================================================================================================

\section{Contributions}

### 2013-06-04 text/Chapter02.tex
Initial commit (working on chapter 2)

\chapter{Pathways to coeffects} 
\label{ch:related-work}

%---------------------------------------------------------------------------------------------------

There are many different directions from which the concept of \emph{coeffects} can be approached 
and, indeed, discovered. In the previous chapter, we motivated it by practical applications, but 
coeffects also naturally arise as an extension to a number of programming language theories.
Thanks to the Curry-Howard-Lambek correspondence, we can approach coeffects from the perspective of 
type theory, logic and also category theory. This chapter gives an overview of the most 
important directions.

We start by revisiting practical applications and existing language features that are related to 
coeffects (Section~\ref{sec:path-apps}), then we look at coeffects as the dual of effect systems
(Section~\ref{sec:path-eff}) and extend the duality to category theory, looking at the categorical 
dual of monads known as \emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically 
inspired type systems that are closely related to our structural coeffects 
(Section~\ref{sec:path-logic}).

This chapter serves two purposes. Firstly, it provides a high-level overview of the  related work, 
although technical details are often postponed until later. Secondly it recasts existing ideas in 
a way that naturally leads to the coeffect systems developed later in the thesis. For this reason, 
we are not always faithful to the referenced work -- sometimes we focus on aspects that the 
authors consider unimportant or present the work differently than originally intended. The reason 
is to fulfil the second goal of the chapter. When we do so, this is explicitly said in the text.

%===================================================================================================

\section{Through applications}
\label{sec:path-apps}

The general theme of this thesis is improving programming languages to better support writing
\emph{context-dependent} (or \emph{context-aware}) computations. With current trends in the 
computing industry such as mobile and ubiquitous computing, this is becoming an important topic.
In software engineering and programming community, a number of authors have addressed this problem
from different perspectives. Hirschfeld et al. propose \emph{Context-Oriented Programming} (COP)
as a methodology \cite{app-cop-method}, and the subject has also been addressed in mobile
computations \cite{app-cop-mobile,app-cop-mobile2}. In programming languages, Costanza 
\cite{app-cop-contextl} develops a domain-specific LISP-like language ContextL and Bardram 
\cite{app-cop-javafwk} proposes a Java framework for COP.

We approach the problem from a different perspective, building on the tradition of 
statically-typed functional programming languages and their theories. However, even in this field,
there is a number of calculi or language features that can be viewed as context-dependent.

%---------------------------------------------------------------------------------------------------

\subsection{Motivation for flat coeffects}

In a number of systems, the execution environment provides some additional data, resources or 
information about the execution context, but are independent of the variables used by the 
program. We look at implicit parameters and rebindable resources (that both provide additional
identifiers that can be accessed similarly to variables, but follow different scoping rules),
distributed programming, cross-compilation and data-flow.

%---------------------------------------------------------------------------------------------------

\paragraph{Implicit parameters} In Haskell, implicit parameters \cite{app-implicit-parameters} are 
a special kind of variables that may behave as dynamically scoped. This means, if a function uses 
parameter $\ident{?p}$, then the caller of the function must define $\ident{?p}$ and set its value.
Implicit parameters can be used to parameterise a computation (involving a chain of function calls)
without passing parameters explicitly as additional arguments of all involved functions. A simple 
language with implicit parameters has an expression $\ident{?p}$ to read a parameter value and an 
expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a different keyword to 
avoid confusion.} $\kvd{letdyn}~?p = e_1~\kvd{in}~e_2$ that sets a parameter $\ident{?p}$ to the value of $e_1$ 
and evaluates $e_2$ in a context containing $\ident{?p}$

An interesting question arises when we use implicit parameters in a nested function. The following 
function does some pre-processing and then returns a function that builds a formatted string based 
on two implicit parameters $\ident{?width}$ and $\ident{?size}$:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\
\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\
\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})\\
\end{array}
\end{equation*}
%
The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context 
$\{ \ident{?width} : \ident{int} \}$. The nested function (returned as a result) uses the parameter 
$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters of the nested 
function come from?

In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
However, in Haskell, implicit parameters behave as a combination of lexical and dynamic scoping. This means
that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$
In Haskell, this corresponds to the following type:
%
\begin{equation*}
(\ident{?width} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow 
  ((\ident{?size} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow \ident{string})
\end{equation*}
%
As a result, the function can be called as follows:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{formatHello} = \\
\quad(~\kvd{letdyn}~\ident{?width}=5~\kvd{in}\\
\quad~~\ident{format}~\texttt{"Hello"})~~\kvd{in}\\
\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}
\end{array}
\end{equation*}
%
This way of assigning type to \ident{format} and calling it is not the only possible, though. 
We could also say that the outer function requires both of the implicit parameters and the result
is a (pure) function with no context requirements. This interaction between implicit parameters 
and lambda abstraction demonstrates one of the key aspects of coeffects and will be discussed 
later. Implicit parameters will also sever as one of our examples in Chapter~Y.

%---------------------------------------------------------------------------------------------------

\paragraph{Type classes}
Implicit parameters are closely related to \emph{type classes} \cite{app-type-classes}. In Haskell,
type classes provide a principled form of ad-hoc polymorphism (overloading). When a code uses 
an overloaded operation (e.g.~comparison or numeric operators) a constraint is placed on the 
context in which the operation is used. For example:
%
\begin{equation*}
\begin{array}{l}
\ident{twoTimes}~::~\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha \\
\ident{twoTimes}~x=x+x
\end{array}
\end{equation*}
%
The constraint $\ident{Num}~a$ on the function type arises from the use of the $+$ operator. 
From the implementation perspective, the type class constraint means that the function takes 
a hidden parameter -- a dictionary that provides the operation $+ :: \alpha \rightarrow \alpha \rightarrow \alpha$.
Thus, the type $\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha$ can be viewed as
$(\ident{Num}_\alpha \times \alpha) \rightarrow \alpha$. Implicit parameters work in exactly
the same way -- they are passed around as hidden parameters.

The implementation of type classes and implicit parameters shows two important points about 
context-dependent properties. First, they are associated with some \emph{scope}, such as the body
of a function. Second, they are associated with the input. To call a function that takes an 
implicit parameter or has a type-class constraint, the caller needs to pass a (hidden) parameter
together with the function inputs.

%---------------------------------------------------------------------------------------------------

\paragraph{Rebindable resources}
The need for parameters that do not strictly follow static scoping rules also arises in distributed
computing. This problem has been addressed, for example, by Bierman et al. and Sewell et al. 
\cite{app-distributed-rebinding,app-distributed-acute}. To quote the first work: \emph{``Dynamic 
binding is required in various guises, for example when a marshalled value is received from the 
network, containing identifiers that must be rebound to local resources.''}

This situation arises when marshalling and transferring function values. A function may depend 
on a local resource (e.g.~a database available only on the server) and also resources that are 
available on the target node (e.g.~current time). In the following example, the construct
$\kvd{access}~\ident{Res}$ represents access to a re-bindable resource named $\ident{Res}$:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\
\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\
\quad\ident{query}~\ident{db}~\str{"SELECT * WHERE Date > \%1"}~(\kvd{access}~\ident{Clock})
\end{array}
\end{equation*}
%
When \ident{recentEvents} is created on a server and sent to a client, a remote reference to the 
database (available only on the server) must be captured. If the client device supports a clock, 
then \ident{Clock} can be locally \emph{rebound}, e.g., to accommodate time-zone changes. 
Otherwise, the date and time needs to be obtained from the server too.

The use of re-bindable resources creates a context requirement similar to the one arising from
the use of implicit parameters. For function values, such context-requirements can be satisfied
in different ways -- resources must be available either at the declaration site (i.e.~when a 
function is created) or at the call site (i.e.~when a function is called).

%---------------------------------------------------------------------------------------------------

\paragraph{Distributed computing and multi-targetting}

An increasing number of programming languages is capable of running across multiple different 
platforms or execution environments. Functional programming languages that can be compiled to
JavaScript (to target web and mobile clients) include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.

Links \cite{app-distributed-links}, F\# libraries \cite{app-fsharp-webapps,app-fsharp-webtools},
ML5 and QWeSST \cite{app-distributed-ml5, app-distributed-qwesst} and Hop \cite{app-hop-lang} go 
further and allow a single source program to be compiled to multiple target runtimes. This posses 
additional challenges -- it is necessary to track where each part of computation runs and statically 
guarantee that it will be possible to compile code to the required target platform 
(safe \emph{multi-targetting}).

We demonstrate the problem by looking at input validation. In distributed applications 
that communicate over unsecured HTTP channel, user input needs to be validated interactively
on the client-side (to provide immediate response) and then again on the server-side (to 
guarantee safety). For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\
\quad\ident{name} \neq \str{""} ~~\&\&~~ \ident{forall~isLetter~name}
\\[0.5em]
\kvd{let}~\ident{displayProduct} = \lambda \ident{name} \rightarrow\\
\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{displayProductPage~name}\\
\quad\kvd{else}~\ident{displayErrorPage}~() 
\end{array}
\end{equation*}
%
The function \ident{validateInput} can be compiled to both JavaScript (for client-side) and
native code (for server-side). However, \ident{displayProduct} uses other functionality
(generating web pages) that is only available on the server-side, so it can only be compiled to
native code.

In Links \cite{app-distributed-links}, functions can be annotated as client-side, server-side
and database-side. F\# WebTools \cite{app-fsharp-webtools} adds functions that support multiple
targets (mixed-side). However, these are single-purpose language features and they are not 
extensible. For example, in modern mobile development it is also important to track minimal 
supported version of runtime\footnote{Android Developer guide \cite{app-android-multitarget} 
demonstrates how difficult it is to solve the problem without language support.}.

Requirements on the execution environment can be viewed as contextual properties, but could be
also presented as effects (use of some API required only in certain environment is a computational
effect). We discuss the difference in Section~X. Furthermore, the theoretical foundations of
distributed languages like ML5 \cite{app-distributed-ml5} suggest that a contextual treatment
is more appropriate. We return to ML5 when discussing semantics in Section~\ref{sec:path-sem-contextdep}.

%---------------------------------------------------------------------------------------------------

\paragraph{Data-flow languages}

The examples discussed so far are all -- to some extent -- similar. They attach additional 
information (implicit parameters, dictionaries) or restrictions (on execution environment) to the
context where code evaluates. By \emph{context}, we mean, most importantly, the values of variables
and declarations that are in scope. The examples so far add more information to the context, but
do not operate on the variable values.

Data-flow languages provide a different example. Lucid \cite{app-lucid} is a declarative data-flow 
language designed by Wadge and Ashcroft. In Lucid, variables represent streams and programs
are written as transformations over streams. A function application $\mathit{square}(a)$ represents
a stream of squares calculated from the stream of values $a$.

The data-flow approach has been successfully used in domains such as development of real-time embedded 
application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the data-flow
paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
and implements program to count the number of edges on a Boolean stream:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{edge} = \ident{false}~\kvd{fby}~(\ident{input}~\&\&~\ident{not}~(\kvd{prev}~\ident{input}))
\\[0.5em]
\kvd{let}~\ident{edgeCount} = \\
\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~\kvd{prev}~\ident{edgeCount}\\
\quad\quad\quad~~~\, \kvd{else}~\kvd{prev}~\ident{edgeCount} ~)
\end{array}
\end{equation*}
%
The construct $\kvd{prev}~x$ returns a stream consisting of previous values of the stream 
$x$. The second value of $\kvd{prev}~x$ is first value of $x$ (and the first
value is undefined). The construct $y~\kvd{fby}~x$ returns a stream whose first element is the 
first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
such as \ident{false} and $0$ are constant streams. Formally, the construct are defined as follows
(writing $x_n$ for $n$-th element of a stream $x$):
%
\[ 
(\kvd{prev}~x)_n = \left\{ 
  \begin{array}{ll}
    nil     & \; \text{if $n=0$}\\
    x_{n-1} & \; \text{if $n>0$}
  \end{array} \right.
\quad
(y~\kvd{fby}~x)_n = \left\{ 
  \begin{array}{ll}
    y_0     & \; \text{if $n=0$}\\
    x_n     & \; \text{if $n>0$}
  \end{array} \right.
\]  
%
When reading data-flow programs, we do not need to think about variables in terms of streams --
we can see them as simple values. However, the operations \kvd{fby} and \kvd{prev} cannot operate
on plain values -- they require additional \emph{context} which provides past values of variables
(for \kvd{prev}) and information about the current location in the stream (for \kvd{fby}).

In this case, the context is not simply an additional (hidden) parameter. It completely changes
how variables must be represented. We may want to capture various \emph{contextual properties}
of Lucid programs. For example, how many past elements need to be cached when we evaluate the 
stream.

To understand the nature of the context, we later look at the semantics of Lucid. This can be
captured using a number of mathematical structures. Wadge \cite{app-lucid-monads} originally 
proposed to use monads, while Uustalu and Vene later used comonads \cite{app-dataflow-essence}.

%---------------------------------------------------------------------------------------------------

\subsection{Motivation for structural coeffects}

We now turn our attention to system where additional contextual information are associated not
with the context as a whole (or program scope), but with individual variables. We start by looking
simple static analysis -- variable \emph{liveness}. Then we revisit data-flow computations and
look at applications in security and software updating.

%---------------------------------------------------------------------------------------------------

\paragraph{Liveness analysis}

\emph{Live variable analysis} (LVA) \cite{app-modern-compiler} is a standard technique in compiler theory. 
It detects whether a free variable of an expression may be used by a program later (it is
\emph{live}) or whether it is definitely not needed (it is \emph{dead}). As an optimization, 
compiler can remove bindings to dead variables as the result is never accessed. Wadler 
\cite{app-strictness-absecnce} describes the property of a variable that is dead as the 
\emph{absence} of a variable.

In this thesis, we first use a restricted (and not practically useful) form of liveness analysis
to introduce the theory of indexed comonads (Section~X) and then use liveness analysis as one of the
motivations for structural coeffects. Consider the following two simple functions:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
\end{array}
\end{equation*}
%
In liveness analysis, we annotate the context with a value specifying whether the variables in
scope are \emph{live} or \emph{dead}. If we associate just a single value with the entire 
context, then the liveness analysis is very limited -- it can say that the context of the 
expression $42$ in the first function is dead, because no variables are accessed.

A useful liveness analysis needs to consider individual variables. For example, in the body of
the second function (\ident{value}), two variables are in scope. The variable \ident{value} is
accessed and thus is \emph{live}, but the variable \ident{x} is dead.

Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
guarantee). Liveness is a \emph{backward} analysis -- this means that the requirements propagates
from variables to their declaration sites. The distinction between \emph{must} and \emph{may} is 
apparent when we look at an example with conditionals:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
\end{array}
\end{equation*}
%
The liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
only if it was used in both of the branches (this is sometimes called \emph{neededness}).

The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
interaction between contextual properties and certain language constructs such as conditionals.

% --------------------------------------------------------------------------------------------------

\paragraph{Data-flow languages (revisited)}
When discussing data-flow languages in the previous section, we said that the context provides 
past values of variables. This can be viewed as a flat contextual property (the context needs
to keep all past values), but we can also view it as a structural property. Consider the following
example:
%
\begin{equation*}
\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
\end{equation*}
%
The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}.

As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
single number for each function. However, we can design a more precise \emph{structural} analysis
and track the number of required elements for individual variables.

% --------------------------------------------------------------------------------------------------

\paragraph{Tainting and provenance}
Tainting is a mechanism where variables coming from potentially untrusted sources are marked
(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
values are annotated with more detailed information about their source.

Statically typed systems that based on tainting have been use to prevent cross-site scripting
attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
\ident{msg}~+~\ident{name}
\end{array}
\end{equation*}
%
In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
(i.e.~to construct an SQL query).

In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
flag (for tainting) or more complex data (for provenance). In static checking, the information
need to be associated with the variables in the variable context. We use tainting as a motivating
example for \emph{structural} coeffects in Section~X.

% --------------------------------------------------------------------------------------------------

\paragraph{Security and core dependency calculus}

The checking of tainting is a special case of checking of the \emph{non-interference} property 
in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
more recent work. The checking of information flows has been also integrated (as a single-purpose
extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
using a monadic library.

Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\ident{L}, \ident{H}\}$ 
represents low and high secrecy, respectively with $\ident{L} \leq \ident{H}$ meaning that low security
values can be treated as high security (but not the other way round).

An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
example which may assign a new value to $z$:
%
\begin{equation*}
\kvd{if}~x>0~\kvd{then}~z := y
\end{equation*}
%
If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
assignment is performed or not performed leaks information in its own (this is an 
\emph{implicit} flow).

Abadi et al. realized that there is a number of analyses similar to secure information flow
and proposed to unify them using a single model called Dependency Core Calculus (DCC) \cite{app-dcc}.
It captures other cases where some information about expression relies on properties of variables
in the context where it executes.  The DCC captures, for example, \emph{binding time analysis}
\cite{app-binding-time-analysis}, which detects which parts of programs can be partially evaluated
(do not depend on user input) and \emph{program slicing} \cite{app-slicing-survey} that identifies
parts of programs that contribute to the output of an expression.

% --------------------------------------------------------------------------------------------------

\subsection{Beyond contextual information}

In the systems discussed

\paragraph{Calculus of capabilities}

a

% --------------------------------------------------------------------------------------------------

\paragraph{Software updating}
Dynamic software updating (DSU) is the ability to update programs at runtime without stopping them.

Stoyle et al. \cite{app-dsu-mutatis}.

% --------------------------------------------------------------------------------------------------

\paragraph{Safe locking}

\cite{app-safe-locking}

\paragraph{Thesis perspective}
?

\newpage
%===================================================================================================

\section{Through type and effect systems}
\label{sec:path-eff}

Introduced by Gifford and Lucassen \cite{effects-gifford,effects-polymorphic}, type and effect 
systems have been designed to track effectful operations performed by computations. Examples 
include tracking of reading and writing from and to memory locations \cite{effects-talpin-et-al}, 
communication in message-passing systems \cite{effects-messagepassing} and atomicity in concurrent 
applications \cite{effects-atomicity}.

Type and effect systems are usually specified judgements of the form $\Gamma \vdash e : \alpha, \sigma$, 
meaning that the expression $e$ has a type $\alpha$ in (free-variable) context $\Gamma$ and 
additionally may have effects described by $\sigma$. Effect systems are typically added to a 
language that already supports effectful operations as a way of increasing the safety -- the type
and effect system provides stronger guarantees than a plain type system. Filinsky 
\cite{effects-comprehensive} refers to this approach as \emph{descriptive}\footnote{In contrast
to \emph{prescriptive} effect systems that implement computational effects in a pure language 
-- such as monads in Haskell}.

%---------------------------------------------------------------------------------------------------

\begin{figure}[t]
\begin{equation*}
\inference[(var)]
  {x:\alpha \in \Gamma }
  {\Gamma \vdash x : \alpha, \emptyset }
\quad
\inference[(write)]
  {\Gamma \vdash e : \alpha, \sigma & l:\ident{ref}_\rho~\alpha\in \Gamma}
  {\Gamma \vdash l \leftarrow e : \ident{unit}, \sigma \cup \{\ident{write}(\rho)\} }
\end{equation*}
\begin{equation*}
\inference[(fun)]
  {\Gamma, x:\alpha_1 \vdash e : \beta, \sigma }
  {\Gamma \vdash \lambda x.e : \alpha \xrightarrow{\sigma} \beta, \emptyset }
\quad  
\inference[(app)]
  {\Gamma \vdash e_1 : \alpha \xrightarrow{\sigma_1} \beta, \sigma_2 \\
   \Gamma \vdash e_2 : \alpha, \sigma_3 }
  {\Gamma \vdash e_1~e_2 : \beta, \sigma_1 \cup \sigma_2 \cup \sigma_3 }
\end{equation*}

\caption{Simple effect system}
\label{fig:path-eff}
\end{figure}

%---------------------------------------------------------------------------------------------------

\paragraph{Simple effect system}
The structure of a simple effect system is demonstrated in Figure~\ref{fig:path-eff}. The example
shows typing rules for a simply typed lambda calculus with an additional (effectful) operation
$l \leftarrow e$ that writes the value of $e$ to a mutable location $l$. The type of locations
($\ident{ref}_\rho~\alpha$) is annotated with a \emph{memory region} $\rho$ of the location $l$.
The effects tracked by the type and effect system over-approximate the actual effects and memory
regions provide a convenient way to build such over-approximation. The effects are 
represented as a set of effectful actions that an expression may perform and the effectful action
(\emph{write}) adds a primitive effect $\ident{write}(\rho)$.

The remaining rules are shared by a majority of effect systems. Variable access (\emph{var}) 
has no effects, application (\emph{app}) combines the effects of both expressions, together with 
the latent effects of the function to be applied. Finally, lambda abstraction (\emph{fun}) is a
pure computation that turns the \emph{actual} effects of the body into \emph{latent} effects of 
the created function.

%---------------------------------------------------------------------------------------------------

\paragraph{Simple coeffect system}
When writing the judgements of coeffect systems, we want to emphasize the fact that coeffect 
systems talk about \emph{context} rather than \emph{results}. For this reason, we write the 
judgements in the form $\Gamma @ \sigma \vdash e : \alpha$, associating the additional information
with the context (left-hand side) of the judgement rather than with the result (right-hand side)
as in $\Gamma \vdash e : \alpha, \sigma$. This change alone would not be very interesting -- we 
simply used different syntax to write a predicate with four arguments. As already mentioned, the 
key difference follows from the lambda abstraction rule.

The language in Figure~\ref{fig:path-coeff} extends simple lambda calculus with resources and
with a construct $\kvd{access}~e$ that obtains the resource specified by the expression $e$.
Most of the typing rules correspond to those of effect systems. Variable access (\emph{var}) 
has no context requirements, application (\emph{app}) combines context requirements of the two
sub-expressions and latent context-requirements of the function.

The (\emph{fun}) rule is different -- the resources requirements of the body $\sigma_1 \cup \sigma_2$
are split between the \emph{immediate context-requirements} associated with the current context 
$\Gamma @ \sigma_1$ and the \emph{latent context-requirements} of the function.

As demonstrated by examples in the Chapter~\ref{ch:introduction}, this means that the resource
can be captured when a function is declared (e.g.~when it is constructed on the server-side
where database access is available), or when a function is called (e.g.~when a function created
on server-side requires access to current time-zone, it can use the resource available on the
client-side).

%---------------------------------------------------------------------------------------------------

\begin{figure}[t]
\begin{equation*}
\inference[(var)]
  {x:\alpha \in \Gamma }
  {\Gamma @ \emptyset \vdash x : \alpha }
\quad
\inference[(access)]
  { \Gamma @ \sigma \vdash e : \ident{res}_\rho~\alpha }
  {\Gamma @ \sigma_1 \cup \{\ident{access}(\rho)\} \vdash \kvd{access}~e : \alpha }
\end{equation*}
\begin{equation*}
\inference[(fun)]
  {\Gamma, x:\alpha @ \sigma_1 \cup \sigma_2 \vdash e : \beta }
  {\Gamma @ \sigma_1 \vdash \lambda x.e : \alpha \xrightarrow{\sigma_2} \beta}
\quad  
\inference[(app)]
  {\Gamma \vdash e_1 : \alpha \xrightarrow{\sigma_1} \beta, \sigma_2 \\
   \Gamma \vdash e_2 : \alpha, \sigma_3 }
  {\Gamma \vdash e_1~e_2 : \beta, \sigma_1 \cup \sigma_2 \cup \sigma_3 }
\end{equation*}

\caption{Simple effect system}
\label{fig:path-coeff}
\end{figure}

%===================================================================================================

\section{Through language semantics}
\label{sec:path-sem}

Another pathway to coeffects leads through the semantics of effectful and context-dependent 
computations. In a pioneering work, Moggi \cite{monad-notions} showed that effects (including
partiality, exceptions, non-determinism and I/O) can be modelled uisng the category theoretic
notion of \emph{monad}.

When using monads, we distinguish effect-free values $\alpha$ from programs, or 
computations $\mtyp{}{\alpha}$. The \emph{monad} $\mtyp{}{}$ abstracts the \emph{notion of 
computation} and provides a way of constructing and composing effectful computations:

\begin{definition}
A \emph{monad} over a category $\catc$ is a triple $(M, \ident{unit}, \ident{bind})$ where:
\begin{compactitem}
\item $M$ is a mapping on objects (types) $M : \catc \rightarrow \catc$
\item $\ident{unit}$ is a mapping $\alpha \rightarrow \mtyp{}{\alpha}$ 
\item $\ident{bind}$ is a mapping $(\alpha \rightarrow \mtyp{}{\beta}) 
  \rightarrow (\mtyp{}{\alpha} \rightarrow \mtyp{}{\beta})$
\end{compactitem}
such that, for all $f:\alpha \rightarrow \mtyp{}{\beta}, g:\beta \rightarrow \mtyp{}{\gamma}$:
\begin{align}
\tag{\emph{left identity}}
  \ident{bind}~\ident{unit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{bind}~f \circ \ident{unit} &= f
  \\
\tag{\emph{associativity}}
  \ident{bind}~(\ident{bind}~g \circ f) &= (\ident{bind}~f) \circ (\ident{bind}~g)
\end{align}
\end{definition}

\noindent
Without providing much details, we note that well known examples of monads include the partiality
monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in 
Haskell, list monad ($\mtyp{}{\alpha} = \mu \gamma.1 + (\alpha \times \gamma)$) and other.
In programming language semantics, monads can be used in two distinct ways.

%---------------------------------------------------------------------------------------------------

\subsection{Effectful languages and meta-languages}

Moggi uses monads to define two formal systems. In the first formal system, a monad is used to model 
the \emph{language} itself. This means that the semantics of a language is given in terms of a 
one specific monad and the semantics can be used to reason about programs in that language. To quote 
\emph{``When reasoning about programs one has only one monad, because the programming language is 
fixed, and the main aim is to prove properties of programs''} \cite[p. 5]{monad-notions}.

In the second formal system, monads are added to the programming language as type constructors, 
together with additional constructs corresponding to monadic \ident{bind} and \ident{unit}.
A single program can use multiple monads, but the key benefit is the ability to reason
about multiple languages. To quote \emph{``When reasoning about programming languages one has different 
monads, one for each programming language, and the main aim is to study how they relate to each 
other''} \cite[p. 5]{monad-notions}.

In this thesis, we generally follow the first approach -- this means that we work with an existing
programming language (without needing to add additional constructs corresponding to the primitives
of our semantics). To explain the difference in greater detail, the following two sections show a
minimal example of both formal systems. We follow Moggi and start with language where judgements have
the form $x:\alpha \vdash e : \beta$ with exactly one variable\footnote{This simplifies the examples
as we do not need \emph{strong} monad, but that is an orthogonal issue to the distinction between
language semantics and meta-language.}.

%---------------------------------------------------------------------------------------------------

\paragraph{Language semantics} When using monads to provide semantics of a language, we do not
need to extend the language in any way -- we assume that the language already contains the 
effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
of the form $x:\alpha \vdash e : \beta$ is interpreted as a morphism $\alpha \rightarrow \mtyp{}{\beta}$,
meaning that any expression is interpreted as an effectful computation. The semantics of variable
access ($x$) and the application of a primitive function $f$ is interpreted as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x:\alpha \vdash x : \alpha} &=& \ident{unit}_\mtyp{}{}\\
\sem{x:\alpha \vdash f~e : \gamma} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
\end{array}
\end{equation*}
%
Variable access is an effect-free computation, that returns the value of the variable, wrapped
using $\ident{unit}_\mtyp{}{}$. In the second rule, we assume that $e$ is an expression using
the variable $x$ and producing a value of type $\beta$ and that $f$ is a (primitive) function
$\beta \rightarrow \mtyp{}{\gamma}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
to a function $\mtyp{}{\beta} \rightarrow \mtyp{}{\gamma}$ which is compatible with the 
interpretation of the expression $e$.

%---------------------------------------------------------------------------------------------------

\paragraph{Meta-language interpretation} When designing meta-language based on monads, we need to
extend the lambda calculus with additional type(s) and expressions that correspond to monadic
primitives:
%
\begin{align*}
\alpha, \beta, \gamma &:= \tau \sep \alpha \rightarrow \beta \sep \mtyp{}{\alpha} \\
e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
\end{align*}
%
The types consist of primitive type ($\tau$), function type and a type constructor that 
represents monadic computations. This means that the expressions in the language can create both
effect-free values, such as $\alpha$ and computations $\mtyp{}{\alpha}$. The additional expression
$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no actual effects) from a
value and $\kvd{let}_\mtyp{}{}$ is used to sequence effectful computations. In the semantics, 
monads are not needed to interpret variable access and application, they are only used in the 
semantics of additional (monadic) constructs:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x:\alpha \vdash x : \alpha} &=& \idf{}\\
\sem{x:\alpha \vdash f~e : \beta} &=& f \circ \sem{e}\\
\sem{x:\alpha \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\beta}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
\sem{x:\alpha \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\beta}} &=& 
  \ident{bind}_\mtyp{}{}~\sem{e_2} \circ \sem{e_1}
\end{array}
\end{equation*}

\noindent
In this system, the interpretation of variable access becomes a simple identity function and
application is just composition. Monadic computations are constructed explicitly using 
$\kvd{return}_\mtyp{}{}$ (interpreted as $\ident{unit}_\mtyp{}{}$) and they are also sequenced
explicitly using the $\kvd{let}_\mtyp{}{}$ construct. As noted by Moggi, the first formal system
can be easily translated to the latter by inserting appropriate monadic constructs.

Moggi regards the meta-language system as more fundamental, because \emph{``its models are more 
general''}. Indeed, this is a valid and reasonable perspective. Yet, we follow the first style,
precisely because it is \emph{less general} -- our aim is to develop concrete context-aware 
programming languages (together with their type theory and semantics) rather than to build a 
general framework for reasoning about languages with context-dependent properties.

%---------------------------------------------------------------------------------------------------

\subsection{Marriage of effects and monads}

The work on effect systems and monads both tackle the same problem -- representing and tracking of 
computational effects. The two lines of research have been joined by Wadler and Thiemann
\cite{monads-effects-marriage}. This requires extending the categorical structure. A monadic
computation $\alpha \rightarrow \mtyp{}{\beta}$ means that the computation has \emph{some} 
effects while the judgement $\Gamma \vdash e : \alpha, \sigma$ specifies \emph{what} effects
the computation has.

To solve this mismatch, Wadler and Thiemann use a \emph{family} of monads $\mtyp{\sigma}{\alpha}$
with an annotation that specifies the effects that may be performed by the computation. In their
system, an effectful function $\alpha \xrightarrow{\sigma} \beta$ is modelled as a pure 
function returning monadic computation $\alpha \rightarrow \mtyp{\sigma}{\beta}$. Similarly, the
semantics of a judgement $x:\alpha \vdash e : \beta, \sigma$ can be given as a function 
$\alpha \rightarrow \mtyp{\sigma}{\beta}$. 
The precise nature of the family of monads has been later called \emph{indexed monads} (e.g.~by Tate
\cite{effects-producer-semantics}) and further developed by Atkey \cite{monads-parameterised-notions} 
in his work on \emph{parameterized monads}.

\paragraph{Thesis perspective}
The key takeaway for this thesis from the outlined line of research is that, if we want to develop a 
language with type system that captures context-dependent properties of programs more precisely,
the semantics of the language also needs to be a more fine-grained structure (akin to indexed 
monads). While monads have been used to model effects, an existing research links context-dependence
with \emph{comonads} -- the categorical dual of monads.

%---------------------------------------------------------------------------------------------------

\subsection{Context-dependent languages and meta-languages}
\label{sec:path-sem-contextdep}

The theoretical parts of this thesis extend the work of Uustalu and Vene who use comonads
to give the semantics of data-flow computations \cite{comonads-dataflow} and more generally, 
notions of \emph{context-dependent computations} \cite{comonads-notions}. The computations discussed 
in the latter work include streams, arrays and containers -- this is a more diverse set of examples, 
but they all mostly represent forms of collections. Ahman et al. \cite{comonads-containers} discuss
the relation between comonads and \emph{containers} in more details.

The utility of comonads has been explored by a number of authors before. Brookes and Geva
\cite{comonads-computational} use \emph{computational} comonads for intensional semantics\footnote{The
structure of computational comonad has been also used by the author of this thesis to abstract
evaluation order of monadic computations \cite{comonads-malias}.}. In functional programming,
Kieburtz \cite{comonads-and-codata} proposed to use comonads for stream programming, but also 
handling of I/O and interoperability.

Biermann and de Paiva used comonads to model the necessity modality $\square$ in intuitionistic
modal S4 \cite{logic-intuitionistic-modal}, linking programming languages derived from modal
logics to comonads. One such language has been reconstructed by Pfenning and Davies
\cite{logic-modal-reconstruction}. Nanevski et al. extend this work to Contextual Modal Type 
Theory (CMTT) \cite{logic-cmtt}, which again shows the importance of comonads for 
\emph{context-dependent} computations.

While Uustalu and Vene use comonads to define the \emph{language semantics} (the first style
of Moggi), Nanevski, Pfenning and Davies use comonads as part of meta-language, in the form 
of $\square$ modality, to reason about context-dependent computations (the second style of 
Moggi). Before looking at the details, we use the following definition of comonad:

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\alpha \rightarrow \mtyp{}{\beta}, g:\beta \rightarrow \mtyp{}{\gamma}$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(\ident{cobind}~g \circ f) &= (\ident{cobind}~f) \circ (\ident{cobind}~g)
\end{align}
\end{definition}

\noindent
The definition is similar to monad with ``reversed arrows''. Intuitively, the $\ident{counit}$ 
operation extracts a value $\alpha$ from a value that carries additional context $\ctyp{}{\alpha}$.
The $\ident{cobind}$ operation turns a context-dependent function 
$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
the context-dependent function to value(s) in the context and then propagates the context. The 
next section makes this intuitive definition more concrete. More detailed discussion about
comonads can be found in Orchard's PhD thesis \cite{comonads-dom-thesis}.

%---------------------------------------------------------------------------------------------------

\paragraph{Language semantics}
To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
$\ctyp{}{\alpha} = \mu \gamma.\alpha + (\alpha \times \gamma)$. A value of the type is either
the last element $\alpha$ or an element followed by another non-empty list $\alpha \times \gamma$.
Note that the list must be non-empty -- otherwise \ident{counit} would not be a complete 
function (it would be undefined on empty list). In the following, we write $(l_1, \ldots, l_n)$
for a list of $n$ elements:
%
\begin{equation*}
\begin{array}{rcl}
\ident{counit}~(l_1, \ldots, l_n) &=& l_1\\
\ident{cobind}~f~(l_1, \ldots, l_n) &=& (f (l_1, \ldots, l_n), f (l_2, \ldots, l_n), \ldots, f (l_n))
\end{array}
\end{equation*}
%
The \ident{counit} operation returns the current (first) element of the (non-empty) list.
The \ident{cobind} operation creates a new list by applying the context-dependent function $f$
to the entire list, to the suffix of the list, to the suffix of the suffix and so on.

In causal data-flow, we can interpret the list as a list consisting of past values, with the 
current value in the head. Then, the $\ident{cobind}$ operation calculates the current value
of the output based on the current and all past values of the input; the second element is
calculated based on all past values and the last element is calculated based just on the initial
input $(l_n)$. In addition to the operations of comonad, the model also uses some operations that
are specific to causal data-flow:
%
\begin{equation*}
\begin{array}{rcl}
\ident{prev}~(l_1, \ldots, l_n) &=& (l_2, \ldots, l_n)\\
\end{array}
\end{equation*}
%
The operation drops the first element from the list. In the data-flow interpretation, this means
that it returns the previous state of a value.

Now, consider a simple data-flow language with single-variable contexts, variables, 
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$ 
has a type $\alpha$, the expression $\kvd{prev}~e$ has also type $\alpha$. The fact that
the language models data-flow and values are lists (of past values) is a matter of semantics,
which is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x:\alpha \vdash x : \alpha} &=& \ident{counit}_\ctyp{}{}\\
\sem{x:\alpha \vdash f~e : \gamma} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x:\alpha \vdash \kvd{prev}~e : \gamma} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (obtain the value and ignore additional available context); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).

For example, the judgement $x:\alpha \vdash \kvd{prev}~(\kvd{prev}~x) : \alpha$ represents an 
expression that expects context with variable $x$ and returns a stream of values before the 
previous one. The semantics of the term expresses this behaviour: 
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed 
earlier.

In the outline presented here, we ignored lambda abstraction. Similarly to monadic semantics,
where lambda abstraction requires \emph{strong} monad, the comonadic semantics also requires
additional structure called \emph{symmetric (semi)monoidal} comonads. This structure is 
responsible for the splitting of context-requirements in lambda abstraction. We return to this
topic when discussing flat coeffect system later in the thesis.

%---------------------------------------------------------------------------------------------------

\begin{figure}
\begin{equation*}
\inference[(eval)]
  {\Gamma \vdash e : \ctyp{\emptyset}{\alpha}}
  {\Gamma \vdash !e : \alpha}
\quad
\inference[(letbox)]
  {\Gamma \vdash e_1 : \ctyp{\Phi, \Psi}{\alpha} & \Gamma, x : \ctyp{\Phi}{\alpha} \vdash e_2 : \beta }
  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{\Psi}{\beta}}
\end{equation*}

\caption{Typing for a comonadic language with contextual staged computations}
\label{fig:modal-meta}
\end{figure}

%---------------------------------------------------------------------------------------------------

\paragraph{Meta-language interpretation} To briefly demonstrate the approach that employs comonads
as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
Nanevski et al. We do not attempt to provide precise overview of their work. The main purpose 
of our discussion is to provide a different intuition behind comonads, and to give an example
of a language that includes comonad as a type constructor, together with language primitives
corresponding to comonadic operations\footnote{In fact, Pfenning and Davies \cite{logic-modal-reconstruction,logic-cmtt}
never mention comonads explicitly. This is done in later work by Gabbay et al. \cite{logic-cmtt-semantics}, 
but the connection between the language and comonads is not as direct as in case of monadic or
comonadic semantics covered in the last few pages.}.

In languages inspired by modal logics, types can have the form $\square \alpha$. In the work of
Pfenning and Davies, this means a term that is provable with no assumptions. In distributed 
programming language ML5, Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus} use the 
$\square \alpha$ type to mean \emph{mobile code}, that is code that can be evaluated at any node of a 
distributed system (the evaluation corresponds to the axiom $\square \alpha \rightarrow \alpha$). 
Finally, Davies and Pfenning \cite{logic-modal-staged} consider staged computations and interpret 
$\square \alpha$ as a type of (unevaluated) expressions of type $\alpha$.

In Contextual Modal Type Theory, the modality $\square$ is further annotated. To keep the syntax
consistent with earlier examples, we use $\ctyp{\Psi}{\alpha}$ for a type $\square \alpha$ with an
annotation $\Psi$. The type is a comonadic counterpart to the \emph{indexed monads} used by Wadler
and Thiemann when linking monads and effect systems and, indeed, it gives rise to a language that
tracks context-dependence of computations in a type system.

In staged computation, the type $\ctyp{\Psi}{\alpha}$ represents an expression 
that requires the context $\Psi$ (i.e.~the expression is an open term that requires variables $\Psi$).
The Figure~\ref{fig:modal-meta} shows two typing rules for such language. The rules directly
correspond to the two operations of a comonad and can be interpreted as follows:

\begin{itemize}
\item (\emph{eval}) corresponds to $\ident{counit} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha$. It means
  that we can evaluate a closed (unevaluated) term and obtain a value. Note that the rule requires
  a specific context annotation. It is not possible to evaluate an open term.

\item (\emph{letbox}) corresponds to $\ident{cobind} : (\ctyp{\Psi}{\alpha} \rightarrow \beta) 
  \rightarrow \ctyp{\Psi, \Phi}{\alpha} \rightarrow \ctyp{\Phi}{\beta}$. It means that given
  a term which requires variable context $\Psi, \Phi$ (expression $e_1$) and a function that turns 
  a term needing $\Psi$ into an evaluated value (expression $e_2$), we can construct a term 
  that requires just $\Phi$.
\end{itemize}

\noindent
The fact that the (\emph{eval}) rule requires a specific context is an interesting relaxation
from ordinary comonads where \ident{counit} needs to be defined for all values. Here, the indexed
\ident{counit} operation needs to be defined only on values annotated with $\emptyset$.

The annotated \ident{cobind} operation that corresponds to (\emph{letbox}) is in details 
introduced in Chapter~X. An interesting aspect is that it propagates the context-requirements
``backwards''. The input expression (second parameter) requires a combination of contexts that
are required by the two components -- those required by the input of the function (first
argument) and those required by the resulting expression (result). This is another key aspect
that distinguishes coeffects from effect systems.

\paragraph{Thesis perspective}
As mentioned earlier, we are interested in designing context-dependent languages and so we
use comonads as \emph{language semantics}. Uustalu and Vene present a semantics of 
context-dependent computations in terms of comonads. We provide the rest of the story known 
from the marriage of monads and effects. We develop coeffect calculus with a type system that 
tracks the context requirements more precisely (by annotating the types) and we add indexing 
to comonads and link the two by giving a formal semantics.

The \emph{meta-language} approach of Pfenning, Davies and Nanevski et al. is closely related to
our work. Most importantly, Contextual Modal Type Theory (CMTT) uses indexed $\square$ modality
which seems to correspond to indexed comonads (in a similar way in which effect systems 
correspond to indexed monads). The relation between CMTT and comonads has been suggested by
Gabbay et al. \cite{logic-cmtt-semantics}, but the meta-language employed by CMTT does not 
directly correspond to comonadic operations. For example, our \ident{let box} typing rule from
Figure~\ref{fig:modal-meta} is not a primitive of CMTT and would correspond to 
$\ident{box}(\Psi, \ident{letbox}(e_1, x, e_2))$. Nevertheless, the indexing in CMTT provides a useful
hint for adding indexing to the work of Uustalu and Vene.

%===================================================================================================

\section{Through sub-structural and bunched logics}
\label{sec:path-logic}

In the coeffect system for tracking resource usage outlined earlier, we associated additional
contextual information (set of available resources) with the variable context of the typing 
judgement: $\Gamma @ \sigma \vdash e : \alpha$. In other words, our work focuses on ``what is
happening on the left hand side of $\vdash$''.

In the case of resources, the additional information about the context are simply added to the
variable context (as a products), but we will later look at contextual properties that affect 
how variables are represented. More importantly, \emph{structural coeffects} link additional
information to individual variables in the context, rather than the context as a whole.

In this section, we look at type systems that reconsider $\Gamma$ in a number of ways. 
First of all, sub-structural type systems \cite{substruct-attpl-intro} restrict the use of variables
in the language. Most famously linear type systems introduced by Wadler \cite{substruct-linear-change} 
can guarantee that variable is used exactly once. This has interesting implications for memory
management and I/O.

In bunched typing developed by O'Hearn \cite{substruct-bunched}, the variable context is a tree 
formed by multiple different constructors (e.g.~one that allows sharing and one that does not). 
Most importantly, bunched typing has contributed to the development of separation logic
\cite{substruct-separation-logic} (starting a fruitful line of research in software verification), 
but it is also interesting on its own.

%---------------------------------------------------------------------------------------------------

\paragraph{Sub-structural type systems}

Traditionally, $\Gamma$ is viewed as a set of assumptions and typing rules admit (or explicitly
include) three operations that manipulate the variable contexts which are shown in 
Figure~\ref{fig:substructural-rules}. The (\emph{exchange}) rule allows us to reorder variables
(which is implicit, when assumptions are treated as set); (\emph{weakening}) makes it possible
to discard an assumption -- this has the implication that a variable may be declared but never
used. Finally, (\emph{contraction}) makes it possible to use a single variable multiple times
(by joining multiple variables into a single one using substitution).

\begin{figure}
\begin{equation*}
\inference[(exchange)]
  {\Gamma, x:\alpha, y:\beta \vdash e : \gamma}
  {\Gamma, y:\beta, x:\alpha \vdash e : \gamma}
\quad
\inference[(weakening)]
  {\Gamma, \Delta \vdash e : \gamma}
  {\Gamma, x:\alpha, \Delta \vdash e : \gamma}
\end{equation*}
\begin{equation*}
\inference[(contraction)]
  {\Gamma, x:\alpha, y:\alpha, \Delta \vdash e : \gamma}
  {\Gamma, x:\alpha, \Delta \vdash \subst{e}{y}{x} : \gamma}
\end{equation*}

\caption{Exchange, weakening and contraction typing rules}
\label{fig:substructural-rules}
\end{figure}

In sub-structural type systems, the assumptions are typically treated as a list. As a result,
they have to be manipulated explicitly. Different systems allow different subset of the rules.
For example, \emph{affine} systems allows exchange and weakening, leading to a system where 
variable may be used at most once; in \emph{linear} systems, only exchange is permitted and so 
every variable has to be used exactly once.

When tracking context-dependent properties associated with individual variables, we need to 
be more explicit in how variables are used. Sub-structural type systems provide a way to do this.
Even when we allow all three operations, we can track which variables are used and how
(and use that to track additional contextual information about variables).

%---------------------------------------------------------------------------------------------------

\paragraph{Bunched type systems}
Bunched typing makes one more refinement to how $\Gamma$ is treated. Rather than having a list
of assumptions, the context becomes a tree that contains variable typings (or special identity
values) in the leaves and has multiple different types of nodes. The context can be defined,
for example, as follows:
%
\begin{equation*}
\Gamma, \Delta, \Sigma := x:\alpha \sep I \sep \Gamma, \Gamma \sep 1 \sep \Gamma; \Gamma
\end{equation*}
%
The values $I$ and $1$ represent two kinds of ``empty'' contexts. More interestingly, non-empty
variable contexts may be constructed using two distinct constructors -- $\Gamma, \Gamma$ and 
$\Gamma; \Gamma$ -- that have different properties. In particular, weakening and contraction is
only allowed for the $;$ constructor, while exchange is allowed for both.

The structural rules for bunched typing are shown in Figure~\ref{fig:substructural-bunched}.
The syntax $\Gamma(\Delta)$ is used to mean an assumption tree that contains $\Delta$ as a 
sub-tree and so, for example, (\emph{exchange1}) can switch the order of contexts anywhere in the
tree. The remaining rules are similar to the rules of linear logic.

One important note about bunched typing is that it requires a different interpretation. The omission
of weakening and contraction in linear logic means that variable can be used exactly once. 
In bunched typing, variables may still be duplicated, but only using the ``;'' separator.
The type system can be interpreted as specifying whether a variable may be shared between the 
body of a function and the context where a function is declared. The system introduces two 
distinct function types $\alpha \rightarrow \beta$ and $\alpha~ \textendash\!\!\!\ast \beta$
(corresponding to ``;'' and ``,'' respectively). The key property is that only the first kind
of functions can share variables with the context where a function is declared, while the second
restricts such sharing. We do not attempt to give a detailed description here as it is not 
immediately to coeffects -- for more information, refer to O'Hearn's introduction 
\cite{substruct-bunched}.

\begin{figure}
\begin{equation*}
\inference[(exchange1)]
  {\Gamma(\Delta, \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma, \Delta) \vdash e : \alpha}
\quad
\inference[(weakening)]
  {\Gamma(\Delta) \vdash e : \alpha}
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
\end{equation*}
\begin{equation*}
\inference[(exchange2)]
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
\quad
\inference[(contraction)]
  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
  {\Gamma(\Delta) \vdash \subst{e}{\Sigma}{\Delta} : \alpha}
\end{equation*}
\caption{Exchange, weakening and contraction rules for bunched typing}
\label{fig:substructural-bunched}
\end{figure}

%---------------------------------------------------------------------------------------------------

\paragraph{Thesis perspective}

Our work can be viewed as annotating bunches. Such annotations then specify additional information
about the context -- or, more specifically, about the sub-tree of the context. Although this is not
the exact definition used in Chapter~X, we could define contexts as follows:
%
\begin{equation*}
\Gamma, \Delta, \Sigma := x:\alpha \sep 1 \sep \Gamma, \Gamma \sep \Gamma @ \sigma
\end{equation*}
%
Now we can not only annotate an entire context with some information (as in the simple coeffect
system for tracking resources that used judgements of a form $\Gamma @ \sigma \vdash e : \alpha$).
We can also annotate individual components. For example, a context containing variables $x,y,z$
where only $x$ is used could be written as $(x:\alpha \,@\, \ident{used}), ((y:\alpha, z:\alpha) 
\,@\, \ident{unused})$.

For the purpose of this introduction, we ignore important aspects such as how are nested annotations
interpreted. The main goal is to show that coeffects can be easily viewed as an extension to the 
work on bunched logic. Aside from this principal connection, \emph{structural coeffects} also 
use some of the proof techniques from the work on bunched logics, because they also use tree-like
structure of variable contexts.

%===================================================================================================

\section{Summary}
Oops!

% \section{Missing}
% ~
% 
% indexed/layered/etc. monads, productors or whatever

### 2013-06-04 text/Chapter02.tex
Adding more stuff in coeffect pathways

@@
 
-\subsection{Beyond contextual information}
-
-In the systems discussed
+\subsection{Beyond passive contexts}
+
+In the systems discussed so far, the context provides additional data (resources, implicit 
+parameters, historical values) or meta-data (security, provenance). However, it is impossible to
+write a function that modifies the context. We use the term \emph{passive} context for such 
+applications.
+
+However, there is a number of systems where the context may be changed -- not just be evaluating
+certain code block in a different scope (e.g. by wrapping it in $\ident{prev}$ in data-flow), but
+also by calling a function that, for example, acquires new capabilities. While this thesis focuses
+on systems with passive context, we quickly look at the most important examples of the 
+\emph{active} variant.
+
+% --------------------------------------------------------------------------------------------------
 
 \paragraph{Calculus of capabilities}
-
-a
+Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
+a sound system with region-based memory management for low-level code that can be easily 
+compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
+who developed an \emph{effect system} (discussed in Section~\ref{sec:path-sem-effects}) that uses
+lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.
+
+In the work of Tofte and Talpin, the context is \emph{passive}. They extend a simple functional language
+with the \kvd{letrgn} construct that defines a new memory region, evaluates an expression (possibly)
+using memory in that region and then deallocates the memory of the region:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
+\quad\kvd{letrgn}~\rho~\kvd{in}\\
+\quad\kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~!\ident{x}
+\end{array}
+\end{equation*}
+%
+The memory region $\rho$ is a part of the context, but only in the scope of the body of 
+\kvd{letrgn}. It is only available to the last line which allocates a memory cell in the region
+and reads it (before the region is deallocated). There is no way to allocate a region inside a 
+function and pass it back to the caller.
+
+Calculus of capabilities differs in two ways. First, it allows explicit allocation and deallocation
+of memory regions (and so region lifetimes do not follow strict LIFO ordering). Second, it
+uses continuation-passing style. We ignore the latter aspect and so the following example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
+\quad \kvd{letrgn}~\rho~\kvd{in}\\
+\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~\ident{x}
+\end{array}
+\end{equation*}
+%
+The example is almost identical to the previous one, except that it does not return the value
+of reference \ident{x}. Instead, it returns the reference, which is located in a newly allocated
+region. Together with the value, the function returns a \emph{capability} to access the region
+$\rho$.
+
+This is where systems with active context differ. To type check such programs, we do not only need
+to know what context is required to call \ident{calculate}. We also need to know what effects it
+has on the context when it evaluates and the current context meeds to be appropriately adjusted
+after a function call. We briefly consider this problem in Section~X. % future work
+
+% --------------------------------------------------------------------------------------------------
+
+% \paragraph{Safe locking}
+% Flanagan and Abadi \cite{app-safe-locking}
+%
+% => This has coeffect style judgments, but it has effect-style lambda
+%
 
@@
 % --------------------------------------------------------------------------------------------------
-
-\paragraph{Safe locking}
-
-\cite{app-safe-locking}
 
@@
 \subsection{Marriage of effects and monads}
+\label{sec:path-sem-effects}
 

### 2013-06-05 text/Chapter02.tex
Finished pathways, copy & past chapters 3 and 4.

@@
 guarantee safety). For example:
-%
+
 \begin{equation*}
@@
 is more appropriate. We return to ML5 when discussing semantics in Section~\ref{sec:path-sem-contextdep}.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Safe locking}
+In the previous examples, the context provides additional values or functions that may be accessed
+at runtime. However, it may also track \emph{permissions} to perform some operation. This is done
+in the type system for safe locking of Flanagan and Abadi \cite{app-safe-locking}.
+
+The system prevents race conditions (by only allowing access to mutable state under a lock)
+and avoids deadlocks (by imposing strict partial order on locks). The following 
+program uses a mutable state under a lock:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{newlock}~l:\rho~\kvd{in}\\
+\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\
+\kvd{sync}~l~(!\ident{state})
+\end{array}
+\end{equation*}
+%
+The declaration \kvd{newlock} creates a lock $l$ protecting memory region $\rho$. We can than
+allocate mutable variables in that memory region (second line). An access to mutable variable
+is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
+which locks a lock and evaluates an expression in a context that contains permission to access
+memory region of the lock ($\rho$ in the above example).
+
+The type system for safe locking associates the list of permission with the variable context.
+It uses judgements of a form $\Gamma, m \vdash e : \alpha$ specifying that an expression has a
+type in context $\Gamma$, given permissions (a list of locked regions) $m$. However, the treatment
+of lambda abstraction differs from the one for implicit parameters or rebindable resources.
+In the system for locking, code inside lambda function cannot use permissions from the scope
+where the function is declared. This is a necessary requirement -- a lambda function created 
+under a lock cannot access protected memory, because it will be executed later. We discuss how
+this restriction fits into our general coeffect framework in Section~X.Y.
+
+%
+% => This has coeffect style judgments, but it has effect-style lambda
+%
 
@@
 
-% \paragraph{Safe locking}
-% Flanagan and Abadi \cite{app-safe-locking}
-%
-% => This has coeffect style judgments, but it has effect-style lambda
-%
+\paragraph{Software updating}
+Dynamic software updating (DSU) \cite{app-dsu-programs,app-dsu} is the ability to update programs at
+runtime without stopping them. The Proteus system developed by Stoyle et al. \cite{app-dsu-mutatis} 
+investigates what language support is needed to enable safe dynamic software updating in C-like 
+languages. The system is based on the idea of capabilities.
+
+The system distinguishes between \emph{concrete} uses and \emph{abstract} uses of a value. When
+a value is used concretely, the program examines its representation (and so it is not safe to
+change the representation during an update). An abstract use of a value does not need to examine
+the representation and so updating the value does not break the program.
+
+The Proteus system uses capabilities to restrict what types may be used concretely after any point
+in the program. All other types, not listed in the capability, can be dynamically updated as this
+will not change concrete representation of types accessed later in the evaluation.
+
+Similarly to Capability Calculus, capabilities in DSU can be changed by a function call. For 
+example, calling a function that may update certain types makes it impossible to use those types
+concretely following the function call. This means that DSU uses the context \emph{actively}
+and not just \emph{passively}.
 
@@
 
-\paragraph{Software updating}
-Dynamic software updating (DSU) is the ability to update programs at runtime without stopping them.
-
-Stoyle et al. \cite{app-dsu-mutatis}.
-
-% --------------------------------------------------------------------------------------------------
-
 \paragraph{Thesis perspective}
-?
-
-\newpage
+
+As demonstrated in this section, there is a huge number of systems and applications that exhibit
+a form of context-dependence. The range includes different static analyses (liveness, provenance), 
+well-known programming language features (implicit parameters and type classes) as well as features
+not widely available (e.g. for distributed programming).
+
+It is impossible to cover all of these topics in a single coherent thesis and so we focus on 
+two key aspects:
+
+\begin{compactitem}
+\item \textbf{Flat vs. structural.} We look at both flat coeffects (single value for entire context) and 
+  structural coeffects (single value per variable). We use liveness, implicit parameters and 
+  data-flow to introduce flat coeffects (Section~X) and liveness, refined data-flow and tainting
+  to talk about structural coeffects (Section~Y).
+
+\item \textbf{Analysis vs. restriction.} Some of the discussed examples can be viewed as static
+  analyses that obtain some information about programs (i.e.~the number of required past values
+  in data-flow). Other examples provide type system that rules out certain invalid programs 
+  (e.g.~safe locking). We cover this topic when discussing \emph{partial coeffects} in Section~Z.
+
+\item \textbf{May vs. must analysis.} When discussing liveness, we observed that we can obtain 
+  two different analyses depending on how conditionals are treated. We discuss this topic in 
+  Section~X. % TBD - could be a chapter
+\end{compactitem}
+
+Although we also looked at examples of \emph{active} contextual computations (where developers can
+write functions that modify the context), we do not consider these applications, to keep the 
+material presented in this thesis focused. We briefly discuss them as future work in Section~X.
+
 %===================================================================================================
@@
 \section{Summary}
-Oops!
+
+This chapter presented four different pathways leading to the idea of coeffects. We also 
+introduced the most important related work, although presenting related work was not the
+main goal of the chapter. The main goal was to show the idea of coeffects as a logical follow up
+to a number of research directions. For this reason, we highlighted only certain aspects of 
+related work -- the remaining aspects as well as important technical details are covered 
+in later chapters.
+
+The first pathway looks at applications and systems that involve notion of \emph{context}.
+The two coeffect calculi we present aim to unify some of these systems. The second pathway
+follows as a dualization of well-known effect systems. However, this is not simply a syntactic
+transformation, because coeffect systems treat lambda abstraction differently. The third 
+pathway follows by extending comonadic semantics of context-dependent computations with 
+indexing and building a type system analogous to effect system from the ``marriage of effects
+and monads''. Finally, the fourth pathway starts with sub-structural type systems. Coeffect
+systems naturally arise by annotating bunches in bunched logics with additional information.
 

### 2013-06-05 text/Chapter03.tex
Finished pathways, copy & past chapters 3 and 4.

\chapter{Flat coeffects} 
\label{ch:flat-coeffects}

%---------------------------------------------------------------------------------------------------

% comonadic semantics
\newcommand{\iunit}{\mathsf{e}}
\newcommand{\imult}{\oplus}
\newcommand{\iftor}[1]{C^{#1}}
\newcommand{\iftorhat}[1]{\hat{C}^{#1}}

\newcommand{\cosem}[1]{\llbracket #1 \rrbracket}
\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
\newcommand{\ccat}[0]{\mathcal{C}}
\newcommand{\obj}[1]{\textnormal{obj}(#1)}

\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }
% identity function

% structural coeffect calculus - operation names

%---------------------------------------------------------------------------------------------------

TODO: Change this

Understanding how programs affect their environment is a well studied area: \emph{effect 
systems}~\cite{effects-talpin-et-al} provide a static analysis of effects and 
\emph{monads}~\cite{monad-notions} provide a unified semantics to different notions of effect.  
Wadler and Thiemann unify the two approaches~\cite{monads-effects-marriage}, \emph{indexing}
a monad with effect information, and showing that the propagation of effects in an effect system 
matches the semantic propagation of effects in the monadic approach.

No such unified mechanism exists for tracking the context requirements. We use the term 
\emph{coeffect} for such contextual program properties. 
Notions of context have been previously captured using comonads
\cite{comonads-notions} (the dual of monads) 
and by languages derived from modal logic
\cite{logic-modal-reconstruction,logic-cmtt}, but these approaches
do not capture many useful examples which motivate our work. 
We build mainly on the former comonadic direction (\S\ref{sec:comonads}) and discuss the
modal logic approach later (\S\ref{sec:related}).

We extend a simply typed lambda calculus with a coeffect system based on comonads,
replicating the successful approach of effect systems and monads.

\paragraph{Examples of coeffects.}
We present three examples that do not fit the traditional approach of
\emph{effect systems} and have not been considered using the \emph{modal logic}
perspective, but can be captured as \emph{coeffect systems} (\S\ref{sec:motivation}) --
the tracking of implicit dynamically-scoped parameters (or resources),
analysis of variable liveness, and tracking the
number of required past values in dataflow computations.

\paragraph{Coeffect calculus.}
Informed by the examples, we identify a general algebraic structure for coeffects. 
From this, we define a general \emph{coeffect calculus} that unifies the motivating 
examples (\S\ref{sec:calculus}) and discuss its syntactic properties (\S\ref{sec:syntactic}).

\paragraph{Indexed comonads.}
Our categorical semantics (\S\ref{sec:comonads}) extends the work of Uustalu and Vene 
\cite{comonads-notions}. By adding annotations, we generalize comonads to \emph{indexed comonads}, 
which capture notions of computation not captured by ordinary comonads.

% ==================================================================================================

\section{Motivation}
\label{sec:motivation}

Effect systems, introduced by Gifford and Lucassen
\cite{effects-gifford}, track \emph{effects} of computations, such as
memory access or message-based
communication~\cite{effects-messagepassing}. Their approach augments
typing judgments with effect information: $\Gamma \vdash e :
\tau, F$.  Wadler and Thiemann explain how this shapes
effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the body.}
\end{quote}
%
In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
does not match this pattern. In the systems we consider, lambda abstraction places 
requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
(immediate requirements), resulting in different syntactic properties. 
We informally discuss three examples first that demonstrate how contextual
requirements propagate. Section (\S\ref{sec:calculus}) then unifies these 
in a single calculus.

We write coeffect judgements $\ctyp{s}{\Gamma} \vdash e : \tau$ where the 
coeffect annotation $s$ associates context requirements with the 
free-variable context $\Gamma$. Function types have the form $\ctyp{s}{\tau_1} \rightarrow \tau_2$ associating
\emph{latent} coeffects $s$ with the parameter.
The $\ctyp{s}{\Gamma}$ syntax and $\ctyp{s}{\tau}$ types are a result of the
indexed comonadic semantics (\S\ref{sec:comonads}).

% --------------------------------------------------------------------------------------------------
\vspace{-1em}
\subsection{Implicit parameters and resources.}
\label{sec:coeffects-res}

Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
They can be used to parameterize a computation without propagating arguments explicitly through a 
chain of calls and are part of the context in which expressions evaluate. As correctly expected
\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
computations follow a similar pattern, but we discuss implicit parameters for simplicity.

The following function prints a number using implicit parameters
\ident{?culture} (determining the decimal mark) and \ident{?format}
(the number of decimal places):
%
\begin{equation*}
\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
\end{equation*}
%
Figure~\ref{fig:example-resources} shows a type-and-coeffect system
tracking the set of an expression's implicit parameters. 
For simplicity here, all implicit parameters have type $\rho$.

Context requirements are created in (\emph{access}), while (\emph{var}) requires no
implicit parameters; (\emph{app}) combines requirements of
both sub-expressions as well as the latent requirements of the
function. The (\emph{abs}) rule is where the example differs from
effect systems.  Function bodies can access the union of the
parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
judgements for the above example are:
%
\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
\begin{equation*}
\begin{array}{rcl}
\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
\end{array}
\end{equation*}
%
The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
Different judgments are desirable depending on how a function is used. 
In the first case, both parameters have to be provided by the caller.
In the second, both are available at declaration-site, but \ident{?format} may be
rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).

Implicit parameters can be captured by the
\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
the next example cannot be structured by \emph{any} monad.

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\vspace{-1.4em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
%
\quad\quad
%
\inference[(\emph{app})\;]
  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\vspace{-0.7em}
\begin{equation*}
\inference[(\emph{access})\;]
  {}
  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
%
\quad\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\vspace{-0.9em}
\caption{Selected coeffect rules for implicit parameters}
\label{fig:example-resources}
\vspace{-1em}
\end{figure}

% --------------------------------------------------------------------------------------------------

\begin{figure}[!b]
\vspace{-2em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
\quad
\inference[(\emph{app})\;]
  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\vspace{-1.5em}
\caption{Selected coeffect rules for liveness analysis}
\label{fig:example-need}
\vspace{-1.2em}
\end{figure}

% --------------------------------------------------------------------------------------------------

\vspace{-1em}
\subsection{Liveness analysis.}

Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
whether it is definitely not needed (\emph{dead}). A compiler can remove
bindings to dead variables as the result is never used.

We start with a restricted analysis and briefly mention how to make it practical
later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.

A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
A dead context may be marked as live by
letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).

The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
$g$ ignores its input (regardless of the evaluation order of the underlying language).
We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
Secondly, a \emph{pointwise composition} passes the same
argument to $g$ and $h$. The parameter is live if either the parameter
of $g$ or $h$ is live ($r \sqcup s$). Application combines the
two operations, so the context $\Gamma$ is live if it is needed by
$e_1$ \emph{or} by the function value \emph{and} by $e_2$.

An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
combines the context annotations using $\sqcap$. Thus, if the body uses some 
variables, both the function argument and the context of the declaration-site are marked as live.

Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
indexed comonads (\S\ref{sec:comonads}), we
model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
value when $r=\ident{D}$).

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\vspace{-1em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{0}{\Gamma} \vdash x : \tau }
%
\quad\quad
% 
\inference[(\emph{app})\;]
  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\inference[(\emph{prev})\;]
  {\ctyp{n}{\Gamma} \vdash e : \tau}
  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
%
\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\vspace{-1em}
\caption{Selected coeffect rules for causal data flow}
\label{fig:example-dataflow}
\vspace{-1em}
\end{figure}

% --------------------------------------------------------------------------------------------------

\vspace{-1em}
\subsection{Efficient dataflow.}
Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
computations over streams. In \emph{causal} data flow, program may
access past values -- in this setting, a function $\tau_1 \rightarrow
\tau_2$ becomes a function from a list of historical values $[\tau_1]
\rightarrow \tau_2$. A coeffect system here tracks how many past values
to cache.

Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.

The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
for the second function); passing the context to two subcomputations requires the maximum
number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
call-site must each provide at least the number of past values required by the function body
(the body may use variables coming from the declaration-site as well as the argument).

The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
(inefficient) unbounded lists of past elements. The above static analysis provides an
approximation of the number of required past elements and so we use just fixed-length lists.

% ==================================================================================================

\section{Generalized coeffect calculus}
\label{sec:calculus}

The previous three examples exhibit a number of commonalities. We capture these in 
the \emph{coeffect calculus}. We do not overly restrict the calculus to make
it open for notions of context-dependent computations not discussed above.

The syntax of our calculus is that of the simply-typed lambda calculus
(where $v$ ranges over variables, \ident{T} over base types, and $r$ over coeffect annotations):
%
\begin{equation*}
e ::= v \sep \lambda v.e \sep e_1~e_2
\quad\quad\quad\quad
\tau ::= \ident{T} \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{r}{\tau}
\end{equation*}

The type $\ctyp{r}{\tau}$ captures values of type $\tau$ in a context
specified by the annotation  $r$. This type appears only on the
left-hand side of a function arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$. 
In the semantics, $\ctyp{r}$ corresponds to some data type
(\eg{}~\ident{list} or \ident{Maybe}). Extensions
such as explicit \emph{let}-binding are discussed later
(\S\ref{sec:syntactic}).

The coeffect tags $r$, that were demonstrated in the previous section, can be generalized to a 
structure with three binary operators and a particular element.

\begin{definition}
A \emph{coeffect algebra} $(S, \imult, \vee, \wedge, \iunit)$ is a set $S$ with
an element $\iunit \in S$, a semi-lattice $(S, \vee)$, a monoid $(S, \imult, \iunit)$,
and a binary $\wedge$. That is, $\forall r,s,t\in S$:
\begin{equation}
\tag{monoid}
   r \imult (s \imult t) = (r \imult s) \imult t 
\quad\quad\quad
   \iunit \imult r = r = r \imult \iunit  \quad\quad
\end{equation}
\vspace{-1.5em}
\begin{equation}
\tag{semi-lattice}
   r \vee s = s \vee r 
\quad\quad
   r \vee (s \vee t) = (r \vee s) \vee t
\quad\quad
   r \vee r = r
\end{equation}
\end{definition}
%
The generalized coeffect calculus captures the three motivating examples (\S\ref{sec:motivation}),
where some operators of the coeffect algebra may coincide.

The $\imult$ operator represents \emph{sequential} composition; guided by the categorical
model (\S\ref{sec:comonads}), we require it to form a monoid with $\iunit$.
The operator $\vee$ corresponds to merging of context-requirements in \emph{pointwise composition} 
and the semi-lattice $(S, \vee)$ defines a partial order: $r \leq s$ when 
$r \vee s = s$. This ordering implies a sub-coeffecting rule.  
The coeffect $\iunit$ is often the top or bottom of the lattice.

The $\wedge$ operator corresponds to splitting requirements of a function body. 
The operator is unrestricted in the general system.
A number of additional laws holds for \emph{some} coeffects systems, \eg{}~semi-lattice structure of
$\wedge$ and a form of distributivity. Quite possibly, they should hold for all coeffect systems.  
We start with as few laws as possible so as not to limit possible uses of the calculus.  
We consider constrained variants that provide useful syntactic properties later (\S\ref{sec:syntactic}).

\vspace{-0.4em}
\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
union $\cup$ for all three operators. Variable access is annotated with $\iunit = \emptyset$
and $\leq$ is subset ordering.

\vspace{-0.4em}
\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
and $\imult$ are both $\sqcap$ (meet).

\vspace{-0.4em}
\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
$\vee, \wedge$ and $\imult$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
Variable access is annotated with $\iunit = 0$ and the order $\leq$ is the standard ordering
of natural numbers.

% --------------------------------------------------------------------------------------------------

\vspace{-1em}
\subsection{Coeffect typing rules.}
Figure~\ref{fig:coeffect-typing} shows the
rules of the coeffect calculus, given some coeffect algebra $(S, \imult, \vee, \wedge, \iunit)$.
The context required by a variable access (\emph{var}) is annotated with $\iunit$.
The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
to be generalized.

The (\emph{abs}) rule checks the body of the function in a context $r \wedge s$, which is a 
combination of the coeffects available in the context $r$ where the function is 
defined and in a context provided by the caller of the function. 
Note that none of the judgements create a \emph{value} of type $\ctyp{r}{\tau}$. This
type appears only immediately to the left of  an arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$.

In function application (\emph{app}), context requirements of both expressions and the 
function are combined as previously: the pointwise composition $\vee$ is used to combine
the coeffects of the expression representing a function $r$ and the coeffects of
the argument, sequentially composed with the coeffects of the function $s \imult t$.

For space reasons, we omit recursion. We note that this would require adding
effect variables and extending the coeffect algebra with a fixed point operation.

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\vspace{-1em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\iunit}{\Gamma} \vdash x : \tau }
%
\quad\quad
%
\inference[(\emph{app})\;]
  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
   \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\inference[(\emph{sub})\;]
  {\ctyp{s}{\Gamma} \vdash e : \tau }
  {\ctyp{r}{\Gamma} \vdash e : \tau }
\;(s\leq r)  
%
\quad\quad\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\caption{Type and coeffect system for the coeffect calculus}
\label{fig:coeffect-typing}
\vspace{-1em}
\end{figure}

% ==================================================================================================

\section{Coeffect semantics using indexed comonads}
\label{sec:comonads}

The approach of \emph{categorical semantics} interprets terms as
morphisms in some category.  For typed calculi, typing judgments $x_1 : \tau_1 \ldots x_n : \tau_n
\vdash e: \tau$ are usually mapped to morphisms $\interp{\tau_1} \times \ldots \times
\interp{\tau_n} \rightarrow \interp{\tau}$. Moggi showed the semantics of various effectful 
computations can be captured generally using the (\emph{strong}) \emph{monad}
structure~\cite{monad-notions}. Dually, Uustalu and Vene showed that
(\emph{monoidal}) \emph{comonads} capture various kinds of context-dependent 
computation~\cite{comonads-notions}.

We extend Uustalu and Vene's approach to give a semantics for
the coeffect calculus by generalising comonads to
\emph{indexed comonads}. We emphasise semantic intuition and
abbreviate the categorical foundations for space reasons.

\vspace{-1em}
\paragraph{Indexed comonads.}

Uustalu and Vene's approach interprets well-typed terms 
as morphisms $C (\tau_1 \times \ldots
\times \tau_n) \rightarrow \tau$, where $C$ encodes contexts
and has a comonad structure~\cite{comonads-notions}.
Indexed comonads comprise a \emph{family} of object mappings 
$C^r$ indexed by a coeffect $r$ describing the contextual requirements satisfied by
the encoded context. We interpret judgments $C^r (x_1 : \tau_1, \ldots, x_n : \tau_n) \vdash e : \tau$ as
morphisms $C^r (\interp{\tau_1} \times \ldots \times \interp{\tau_n}) \rightarrow \interp{\tau}$.

The indexed comonad structure provides a notion of composition
for computations with different contextual requirements.
%%
\begin{definition}
Given a monoid $(S, \imult, \iunit)$ with binary operator $\imult$ and
unit $\iunit$, an \emph{indexed comonad} over a category $\ccat$
comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
$\iftor r A \in \obj{\ccat}$ and:
%%
\begin{itemize}
\item a natural transformation $\counit_A : \iftor{\iunit} A \rightarrow A$, called the \emph{counit};
\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
  r A \rightarrow B$ to \\ morphisms $\iftor{r \imult{} s} A
  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
  \emph{coextend};
\end{itemize}
%
\noindent
such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
the following equations hold:
%
\renewcommand{\arraycolsep}{1.5em}
\begin{align*}
\begin{array}{ccc}
\counit \circ \cobind{f}{r, \iunit} = f & 
\cobind{{(\counit)}}{\iunit, r} = \ident{id} & 
\cobind{(g \circ \cobind{f}{r, s})}{(r \imult s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \imult t)} \\
\end{array}
\end{align*}
\end{definition}
%
The \emph{coextend} operation gives rise to an associative composition operation for
computations with contextual requirements (with \emph{counit} as the identity):
%
\[
\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
  \rightarrow (\iftor{r \imult s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
\]
%
The composition $\hat{\circ}$ best expresses the intention
of indexed comonads: contextual requirements of the composed
functions are combined. The properties of
the composition follow from the indexed comonad laws and the 
monoid $(S, \oplus, \iunit)$.

\paragraph{Example}
Indexed comonad are analogous to comonads (in coKleisli form), but
with the additional monoidal structure on indices.  Indeed, comonads
are a special case of indexed comonads with a trivial singleton
monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\iftor 1$
is the underlying functor of the comonad and $\counit$ and
$\cobind{(-)}{1, 1}$ are the usual comonad operations.
However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.

\paragraph{Example}
The \emph{indexed partiality comonad} encodes free-variable contexts
of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
\sqcap, \ident{L})$, where $\iftor{\ident L} A = A$ encodes live
contexts and $\iftor{\ident D} A = 1$ encodes dead contexts,
where $1$ is the unit type inhabited by a single value $()$.
The \emph{counit} operation $\counit{} : \iftor{\ident{L}} A \rightarrow A$
is defined $\counit\ x = x$ and \emph{coextend},
for all $f : \iftor r A \rightarrow B$, and thus
$\cobind{f}{r, s} : \; \iftor{r \sqcap s} A \rightarrow \iftor s B$,
is defined:
%
\begin{align*}
\begin{array}{llll}
\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{L}} x = f~x 
\end{array}
\end{align*}
%
The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
data type $\mathsf{Maybe}\ A = A + 1$.  This
type does not permit a comonad structure since
$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
For the indexed comonad, $\counit$ need only be defined
 for $\iftor{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
 contextual notions of computation than comonads.

 Moreover, indexed comonads are not restricted by the \emph{shape
   preservation} property of comonads~\cite{orchard12codo}: that
 a coextended function cannot change the \emph{shape} of the
 context. For example, in the second case above $f^\dagger_{\ident{D},
   \ident{L}} : \iftor{\ident{D}} A \rightarrow \iftor{\ident{L}} B$
 where the shape changes from $1$ (empty context) to $B$ (available
 context).

\vspace{-1em}
\subsection{Monoidal indexed comonads.}
Indexed comonads provide a semantics to sequential composition, but
additional structure is needed for the semantics of the full coeffect calculus.
Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
structure, which provides a monoidal operation $\mathsf{m} : C A
\times C B \rightarrow C (A \times B)$ for merging contexts (used in
the semantics of abstraction).

The semantics of the coeffect calculus requires an indexed
lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
%
\begin{itemize}
\item $\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \wedge s)}(A \times B)$ natural
in $A, B$;
\item $\csplit_{r, s} : \iftor{(r \vee s)}(A \times B) \rightarrow \iftor r A \times \iftor s B$ natural
in $A, B$;
\end{itemize}
%
\noindent
The $\cmerge_{r, s}$ operation merges contextual computations with
tags combined by $\wedge$ (greatest lower-bound), elucidating 
the behaviour of $\cmerge_{r, s}$: that merging may result in 
the loss of some parts of the contexts $r$ and $s$.

The $\csplit_{r, s}$ operation splits context-dependent computations
and thus the contextual requirements. To obtain coeffects $r$ and
$s$, the input needs to provide \emph{at least} $r$ and $s$, so the
tags are combined using the $\vee$ (least upper-bound).

For the sake of brevity, we elide the indexed versions of the laws required by Uustalu and Vene
(\eg~most importantly, merging two contexts and then adding the third is
equivalent to merging the last two and then adding the first; similar rule holds is required
for splitting).

\paragraph{Example}
For the indexed partiality comonad, given the liveness coeffect
algebra $(\{\ident{D}, \ident{L}\}, \sqcap, \sqcup, \sqcap, \ident{L})$,
the additional lax/colax monoidal operations are:
%%
\begin{align*}
\begin{array}{lll}
\cmerge_{\ident{L}, \ident{L}} (x, y) = (x, y) 
  \quad\quad& \csplit_{\ident{D}, \ident{D}} \;\; () \;\;\; = ((), ())
  \quad\quad& \csplit_{\ident{D}, \ident{L}} (x, y) = ((), y) \\
\cmerge_{\ident{r}, \ident{s}} \;\,(x, y) = () 
  & \csplit_{\ident{L}, \ident{D}} (x, y) = (x, ())
  & \csplit_{\ident{L}, \ident{L}} (x, y) = (x, y)
\end{array}
\end{align*}

\paragraph{Example}
Uustalu and Vene model causal dataflow computations using the
non-empty list comonad $\ident{NEList}~A = A \times (1 +
\ident{NEList}~A)$~\cite{comonads-notions}.  Whilst this comonad
implies a trivial indexed comonad, we define an indexed comonad with
integer indices for the number of past values demanded of the context.

We define $\ctyp{n}{A} = A \times (A \times \ldots \times A)$ where the first $A$ is the current
(always available) value, followed by a finite product of $n$ past values. The definition of the
operations is a straightforward extension of the work of Uustalu and Vene.

%\begin{example}
%The semantics of the coeffect calculus for
% dynamically-scoped implicit parameters, with coeffect
%algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
%is provided by the indexed comonad $\iftor rA = A \times (r \rightarrow \tau)$
%where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
%do we need to store the types too? Maybe should just be a product.}
%%
%\vspace{-1.0em}
%\begin{equation*}
%\begin{array}{l}
%\cobind{f}{r, s} \; : \; \iftor{r \cup s} A \rightarrow \iftor s B\\
%\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
%\\[1em]
%\counit{} (a, g) = a\\
%%\iota_{r, s} \; : \: \iftor r A \rightarrow \iftor s A\\
%\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
%\end{array}
%\quad\begin{array}{l}
%\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \cup s)}(A \times B)\\
%\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
%\\[1em]
%\csplit_{r, s} : \iftor{(r \cup s)}(A \times B) \rightarrow \iftor r A \times \iftor s B \\
%\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
%\end{array}
%\end{equation*}
%\end{example}

%The function created using \emph{coextend} expects a function $g$ that is defined for
%resource names $r \cup s$. We use function restriction $g|_r$ and $g|_s$ to get functions that are 
%defined on resources required by $f$ and the result of coextend, respectively.

%The operation $\cmerge_{r, s}$ pairs the values and combines the resources available in both contexts. 
%It combines a function $g_1|_{r\setminus s}$ (defined for resources in $r$ excluding those in $s$) and 
%function $g_2$ (defined for resources $s$). This definition is not symmetric as it prefers resources 
%defined in the second context. As a result, the caller of a function can rebind resources defined
%at the declaration site.

%Resources show that indexed comonads describe the structure of context more precisely
%than comonads. A comonad would have to use $TA = A \times (\Omega \rightharpoonup R)$ 
%with a \emph{partial} function on all resource names. Using indexed comonads, we can ensure that the function
%is defined on all resources tracked statically in the tag.\dnote{Needs tweaking}

\vspace{-1.5em}
\subsection{Categorical Semantics. }
\label{sec:comonads-semantics}

\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}

\begin{figure*}[t]
\vspace{-1em}
\newcommand{\uu}[1]{\overline{#1}}
% \begin{equation*} 
% \cosem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
%   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
% \end{equation*}
% \vspace{-0.4em}
\begin{equation*}
\begin{array}{rcll}
\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
    \emph{curry} \; (\cosem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
\\[0.1em]
\cosem{\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
    (\emph{uncurry} \;  \cosem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
 & & \quad\; (id \times \cobind{\cosem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
    \circ \csplit_{r, s \imult t} \circ \iftor{r \vee (s \imult t)}\Delta
\\[0.1em]
\cosem{\ctyp{\iunit}{\Gamma} \vdash x_i : \tau_i} &=&
    \pi_i \circ \counit                                                                  % &$(3)$
%\\[0.1em]
%\cosem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
%    \cosem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
%\\[0.1em]
%\cosem{\ctyp{r\vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
%    \cosem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
% ((\varepsilon \times \cosem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
%\iftorhat{r \vee (r \imult t)}\Delta
\end{array}
\end{equation*}
\vspace{-1.5em}
\caption{Categorical semantics for the coeffect calculus}
\label{fig:semantics}
\vspace{-1.0em}
\end{figure*}

Figure~\ref{fig:semantics} shows the
categorical semantics of the coeffect calculus
using additional operations 
$\pi_i$ for projection of the \emph{i}$^{th}$ element of a 
product, usual
\emph{curry} and \emph{uncurry} operations, and
 $\Delta : A \rightarrow A \times A$ duplicating a
value. While $\iftor{r}$ is a family of object mappings,
it is promoted to a family of functors with the derived morphism mapping 
$\iftor{r}(f) = \cobind{ (f \circ \counit) }{\iunit, r}$.

The semantics of variable access and abstraction are the same as in
Uustalu and Vene's semantics, modulo coeffects. Abstraction uses $\cmerge_{r, s}$ to merge the outer
context with the argument context for the context of the function body.
The indices of $\iunit$ for
$\counit$ and $r,s$ for $\cmerge_{r, s}$ match the coeffects
of the terms.
The semantics of application is more complex. It first duplicates the
free-variable values inside the context and then splits this context
using $\csplit_{r, {s \oplus t}}$. The two contexts (with
different coeffects) are passed to the two sub-expressions,
where the argument subexpression, passed a context $(s \oplus t)$,
is coextended to produce a context $s$ which is passed into
the parameter of the function subexpression (\emph{cf.} given $f : A \rightarrow (B \rightarrow C)$,
 $g : A \rightarrow B$, then $\emph{uncurry} \, f \, \circ \, (id \times g) \circ \Delta : A \rightarrow C$).

A semantics for sub-coeffecting is omitted, but may be provided
by an operation 
$\iota_{r, s} : \iftor r A \rightarrow \iftor s A$ natural in $A$, for all $r, s \in S$ where $s \leq r$,
which transforms a value $C^r A$ to $C^s A$ by
ignoring some of the encoded context.

% ==========================================================================================

\section{Syntax-based equational theory}
\label{sec:syntactic}

Operational semantics of every context-dependent language differs as the notion of context 
is always different. However, for coeffect calculi satisfying certain conditions we can 
define a universal equational theory. This suggests a pathway to an operational semantics for 
two out of our three examples (the notion of context for data-flow is more complex).

In a pure $\lambda$-calculus, $\beta$ and $\eta$ equality for
functions (also called \emph{local soundness} and \emph{completeness}
respectively~\cite{logic-modal-reconstruction})
describe how pairs of abstraction and application can be eliminated:
$(\lambda x . e_2) e_1 \equiv_\beta \subst{e_1}{x}{e_2}$ and $(\lambda x . e \, x) 
\equiv_\eta e$. The $\beta$ equality rule, using the usual Barendregt convention of 
syntactic substitution, implies a \emph{reduction},
giving part of an operational semantics for the calculus.

The call-by-name evaluation strategy modelled by $\beta$-reduction is
not suitable for impure calculi therefore a restricted $\beta$ rule,
corresponding to call-by-value, is used, \ie~$(\lambda x . e_2) v
\equiv \subst{e_2}{x}{v}$. Such reduction can be encoded by a
\emph{let}-binding term, $\kvd{let}~x=e_1~\kvd{in}~e_2$, which
corresponds to sequential composition of two computations, where the
resulting pure value of $e_1$ is substituted into
$e_2$~\cite{monads-inaction,monad-notions}.

%
% We consider here both a notion of \emph{let}-binding for the coeffect
% calculus, useful for a CBV evaluation, and a notion of substitution 
% for a CBV evaluation.
%

For an equational theory of coeffects, consider first a notion 
of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
has the following type and coeffect rule:
%
\begin{equation}
\inference
  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
  {\ctyp{r_1 \vee (r_2 \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
: \tau_2 }
\label{eq:let1}
\end{equation}

\noindent
For our examples, $\wedge$ is idempotent (\ie{}, $r \wedge r = r$)
implying a simpler rule:

% For the three examples we consider, a simpler rule gives a more
% precise coeffect. Because $r \wedge r = r$ for all our examples, we
% can also a coeffect $r \vee (r \imult s)$. Moreover,
%for our examples (but not necessarily for \emph{all} coeffect
%systems), $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$ meaning
%that the following gives more precise coeffects:
%
\begin{equation}
\inference
  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
   \ctyp{r}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
  {\ctyp{r \vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2 : \tau_2 }
\label{eq:let2}
\end{equation}
%%
For our examples (but not necessarily \emph{all} coeffect
systems), this defines a more ``precise'' coeffect with respect to $\leq$
where $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$.

This rule removes the non-principality of the first rule
(\ie~multiple possible typings).  However, using idempotency 
to split coeffects in abstraction would remove additional
flexibility needed by the implicit parameters example.

The coeffect $r \vee (r \imult s)$ can
also be simplified for all our examples, leading to more intuitive
rules -- for implicit parameters $r \cup (r \cup s) = r \cup s$; for
liveness we get that $r \sqcup (r \sqcap s) = r$ and for dataflow we
obtain $\textit{max}(r, r+s) = r + s$.

Our calculus can be extended with \emph{let}-binding and \eqref{eq:let2}.
However, we also consider the cases when a
syntactic substitution $e_2[x \leftarrow e_1]$ has the coeffects
specified by the above rule \eqref{eq:let2} 
and prove \emph{subject reduction} theorem
for certain coeffect calculi.  We consider two common special cases
when the coeffect of variables $\iunit$ is the greatest ($\top$) or
least ($\bot$) element of the semi-lattice $(S, \vee)$ and derive
additional conditions that have to hold about the coeffect algebra:

\begin{lemma}[Substitution]
\label{thm:subst}
Given $C^r (\Gamma, x : \tau_2) \vdash e_1 : \tau_1$ and $C^s \Gamma \vdash e_2 : \tau_2$
then $C^{r \vee (r \oplus s)} \Gamma \vdash \subst{e_2}{x}{e_1} : \tau_1$ if 
the coeffect algebra satisfies the conditions that 
$\iunit$ is either the greatest or least element of the semi-lattice,
$\oplus = \wedge$, and $\oplus$ distributes over $\vee$,
\ie{}, $X \oplus (Y \vee Z) = (X \oplus Y) \vee (X \oplus Z)$.
\end{lemma}

\begin{proof}
By induction over $\vdash$, using the laws (\S\ref{sec:calculus}) and additional assumptions.
\end{proof}

Assuming $\rightarrow_\beta$ is the usual call-by-name reduction, the
following theorem models the evaluation of coeffect calculi with
coeffect algebra that satisfies the above requirements. We do not
consider \emph{call-by-value}, because our calculus does not have a
notion of \emph{value}, unless explicitly provided by
\emph{let}-binding (even a function ``value'' $\lambda x.e$ may have
immediate contextual requirements).

\begin{theorem}[Subject reduction]
\label{thm:reduction}
For a coeffect calculus, satisfying the conditions of Lemma~\ref{thm:subst}, if
$\ctyp{r}{\Gamma} \vdash e : \tau$ and $e \rightarrow_\beta e'$ then 
$\ctyp{r}{\Gamma} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:subst}. 
\end{proof}

The above theorem holds for both the liveness and resources examples,
but not for dataflow.  In the case of liveness, $\iunit$ is the
greatest element ($r \vee \iunit = \iunit$); in the case of
resources, $\iunit$ is the \emph{least} element ($r \vee \iunit =
r$) and the proof relies on the fact that additional
context-requirements can be placed at the context $\ctyp{r}{\Gamma}$
(without affecting the type of function when substituted under
$\lambda$ abstraction).

However, the coeffect calculus also captures context-dependence in
languages with more complex evaluation strategies than
\emph{call-by-name} reduction based on syntactic substitution.  In
particular, syntactic substitution does not provide a suitable evaluation
for dataflow (because a substituted expression needs to capture the
context of the original scope).

Nevertheless, the above results show that -- unlike effects --
context-dependent properties can be integrated with
\emph{call-by-name} languages. Our work also provides a model of
existing work, namely Haskell implicit parameters
\cite{app-implicit-parameters}.

% ==================================================================================================

\section{Related and further work}
\label{sec:related}

This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
may split contextual requirements between the declaration-site and call-site.

Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).

% --------------------------------------------------------------------------------------------------

\vspace{-1em}
\paragraph{Comonads and modal logics.}

Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
$\square$ modality of an intuitionistic S4 modal logic using monoidal
comonads, which links our calculus to modal logics.  This link can be
materialized in two ways.

Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
\cite{monad-notions}) that includes $\square$ as a type
constructor. For example, in \cite{logic-modal-reconstruction}, the
modal type $\Box \tau$ represents closed terms.
In contrast, the \emph{semantic} approach uses monads or comonads
\emph{only} as a semantics.  This has been employed by Uustalu and
Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
follow the semantic approach.

Nanevski et al. extend an S4 term language to a \emph{contextual}
modal type theory (CMTT)~\cite{logic-cmtt}.
The \emph{context} is a set of variables required by a computation, which
makes CMTT useful for meta-programming and staged computations. Our contextual types are
indexed by a coeffect algebra, which is more general and can capture
variable contexts, but also integers, two-point lattices, \emph{etc.}.

The work on CMTT suggests two extensions to coeffects. The first is
developing the logical foundations. We briefly considered special cases
of our system that permits local soundness in \S\ref{sec:syntactic} and
local completeness can be treated similarly. The second problem is 
developing the coeffects \emph{metalanguage}. The use of coeffect algebras
would provide an additional flexibility over CMTT, allowing a wider range 
of applications.

% --------------------------------------------------------------------------------------------------

\vspace{-1em}
\paragraph{Relating effects and coeffects.} 
The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
semantic model (monads vs. comonads) is very different, we can consider extending the two to 
obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
the reader monad needs an operation that eagerly performs some effects of a function: 
$(\tau_1 \rightarrow \mtyp{r \oplus s}{\tau_2}) \rightarrow \mtyp{r}{(\tau_1 \rightarrow \mtyp{s}{\tau_2})}$.
To obtain a pure lambda abstraction
for coeffects, we need to restrict the $\cmerge_{r, s}$ 
operation of indexed comonads, so that the first parameter is annotated with $\iunit$ (meaning
no effects): $\iftor \iunit A \times \iftor r B \rightarrow \iftor{r}(A \times B)$.

\vspace{-1em}
\paragraph{Structural coeffects.} To make the liveness analysis practical, we need to associate
information with individual variables (rather than the entire context). We can generalize the 
calculus from this paper by adding a product operation $\times$ to the coeffect algebra.
A variable context $x:\tau_1, y:\tau_2, z:\tau_3$ is then annotated with
$r\times s \times t$ where each component of the tag corresponds to a single variable. The system
then needs to be extended with structural rules such as:
%
\begin{equation*}
\inference[(\emph{abs})]
  {\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\quad
\inference[(\emph{contr})]
  {\ctyp{r \cprd s}{(x:\tau_1, y:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r \cvop s }{(z:\tau_1)} \vdash \subst{\subst{e}{x}{z}}{y}{z} : \tau_2 }
\end{equation*}
%
The context-requirements associated with function are exactly those linked to the specific
variable of the lambda abstraction. Rules such as contraction manipulate variables and perform
a corresponding operation on the indices.

The structural coeffect system is related to bunched typing \cite{types-bunched} (but generalizes
it by adding indices). We are currently investigating how to use structural coeffects to capture 
fine-grained context-dependence properties such as secure information flow \cite{app-secure-flow}
or, more generally, those captured by dependency core calculus \cite{types-dcc}.

% ==================================================================================================

\section{Conclusions}

We examined three simple calculi with associated
static analyses (liveness analysis, implicit parameters, and dataflow
analysis). These were unified in the \emph{coeffect calculus},
providing a general coeffect system parameterised by an
algebraic structure describing the propagation of context
requirements throughout a program.

We model the semantics of coeffect calculus using \emph{indexed comonad} -- a novel structure, which
is more powerful than (monoidal) comonads. Indices of the indexed comonad operations manifest the 
semantic propagation of context such that the propagation of information in the general coeffect
type system corresponds exactly to the semantic propagation of context in our categorical model.

We consider the analysis of context to be essential, not least for the examples here but 
also given increasingly rich and diverse distributed systems.

### 2013-06-05 text/Chapter04.tex
Finished pathways, copy & past chapters 3 and 4.

\chapter{Structural coeffects} 
\label{ch:structural}

\newcommand{\ctimes}{ \hat{\times} }
\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }

% alculus names and rule names
\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}

\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}

%---------------------------------------------------------------------------------------------------

Introduction

%===================================================================================================

\section{Introduction}
\label{sec:structural-intro}

We already talked about tainting in pathways to coeffects... here is more stuff!

~

~

Our second motivating example is the use of \emph{tainting} for the prevention of SQL injection 
attacks \cite{app-secure-tainting}. The idea is to (statically) check that SQL commands cannot be 
directly constructed from, potentially dangerous, inputs provided by the user. Consider the type
checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
\ident{msg}~+~\ident{name}
\end{array}
\end{equation*}
%
We say that a variable is \emph{tainted} if it may come from untrusted input. In the above example,
\ident{id} must not be tainted. Otherwise, the attacker could specify values such as 
\str{"1; DROP TABLE Users"}. The variable \ident{msg} may or may not be tainted, because it is not
used in protected context (i.e.~to construct an SQL query). Using the coeffect calculus, security
information can be naturally attached to the context. We use tag \ident T for tainted and 
\ident N for not tainted:
%
\begin{equation*}
\ctyp{\ident T \times \ident N}{(\ident{id}:\ident{string}, \ident{msg}:\ident{string})} 
  \vdash (\ldots) : \ident{string}
\end{equation*}
%
The tags \ident T and \ident N correspond to variables \ident{id} and \ident{msg}, respectively.
The checking of tainting is a special case of checking of \emph{non-interference} in 
\emph{secure information flow} \cite{app-secure-information-flow}. Languages such as FlowCaml 
\cite{app-security-flowcaml} can check such properties statically, but it is also possible to 
embed such checking in a monadic library \cite{monad-secure-flow}.

Assuming the monadic type $\mtyp{s}{\tau}$ represents a value of type $\tau$ 
in a security context $s$ (either \ident T or \ident N): 
%
\begin{equation*}
\ident{id} : \mtyp{\ident T}{\ident{string}}, \ident{msg} : \mtyp{\ident N}{\ident{string}} 
  \vdash (\ldots) : \mtyp{\ident T}{\ident{string}}
\end{equation*}
%
In this example, the result is tagged with \ident T to denote that it may be tainted.
If the result was tagged with \ident N, the first parameter would also have to be \ident N
(and we would get fewer information about the function).

The monadic encoding has an unfortunate property from the semantic perspective. 
The function has monadic structure over the \emph{domain} as well as over the \emph{codomain}
and thus it does not match the desirable Kleisli model of expressions $\tau_1 \times \ldots \tau_n \rightarrow \mtyp{}{\tau}$. 
This makes the it difficult to give standard composable semantics of the library.

In coeffect calculus, expressions are modelled using the coKleisli form 
$\ctyp{r_1 \times \ldots \times r_n}{(\tau_1 \times \ldots \tau_n)} \rightarrow \tau$ with a tags 
$r_n$ that specify taintedness of inputs. As discussed in detail in Section~\ref{sec:structural-tainting}, 
this model gives a simple comonadic semantics of language-based security using  tainting.

%---------------------------------------------------------------------------------------------------
%===================================================================================================

% ==================================================================================================
\newcommand{\ctxtrans}[2]{#1 \Rightarrow_c #2}
\begin{figure*}[t]
% Constants, if we wanted them, would look like this
%
% \inference[]
%   {c : \tau \in \Phi}
%   {\ctyp{1}{()} \vdash c : \tau}&
% \textnormal{\rulest{const}}
% \\[1.6em]

%\paragraph{Typing rules}
\begin{equation*}
\boxed{\ctyp{r}{\Gamma} \vdash e : \tau }
\end{equation*}

\begin{equation*}
\begin{array}{cl}

\inference[]
  {}
  {\ctyp{0}{(v:\tau)} \vdash v : \tau }&
\textnormal{\rulest{var}}
\\[1.8em]
\inference
  {\ctyp{r \cprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
  {\ctyp{r}{\Gamma} \vdash \lambda v.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }&
\textnormal{\rulest{fun}}
\\[1.8em]
\inference
  {\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2 &
   \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1 }
  {\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
\textnormal{\rulest{app}}
\\[1.8em]
\inference
  {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
  \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1}
  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
\textnormal{\rulest{let}}
\\[1.8em]
\inference
  {\ctyp{r}{\Gamma} \vdash e : \tau &
   \ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}
  {\ctyp{r'}{\Gamma'} \vdash e : \tau}&
\textnormal{\rulest{ctx}}
\end{array}
\end{equation*}

%
% Empty variable context corresponds to unit, but things get messy
%

%\paragraph{Context rules}
\begin{equation*}
\boxed{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}} }
\end{equation*}

\begin{equation*}
\begin{array}{rcll}
\multicolumn{3}{c}{
  \inference
    {  \ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}} }
    {  \ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}} }
}&
\textnormal{\rulest{nest}}
\\[2.0em]
  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}&
\textnormal{\rulest{exch}}
\\[0.8em]
  \ctxtrans{ \ctyp{r \cprd 1}{(\Gamma, ())} &}{&
    \ctyp{r}{\Gamma}}&
\textnormal{\rulest{empty}}
\\[0.8em]
  \ctxtrans{\ctyp{r\cprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
\textnormal{\rulest{weak}}
\\[0.8em]
  \ctxtrans{\ctyp{r \cvop s}{(x:\tau)} &}{&
    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}&
\textnormal{\rulest{contr}}
\\[0.8em]
  \ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
\textnormal{\rulest{assoc}}
\\[0.8em]
\multicolumn{3}{c}{
  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \cvop r = s)
}&
\textnormal{\rulest{sub}}
\end{array}
\end{equation*}

% We do not need this rule (can be expressed in terms of the other one)
%
% \inference
%  {  \ctxtrans{\ctyp{s}{\Gamma_2}}{\ctyp{s'}{\Gamma_2'}} }
%  {  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\cprd s'}{(\Gamma_1, \Gamma_2')}} }&
% \textnormal{\rulest{right}}

\caption{Type system for \clstr}
\label{fig:structural-coeffect-typing}
\end{figure*}

\newpage

% --------------------------------------------------------------------------------------------------

\section{Structural coeffect system}
\label{sec:structural}

The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
we need to track context-dependence in a more fine-grained way. To track neededness or security, 
we need to associate information with individual \emph{variables} of the context.

At the same time, we want to avoid developing multiple variants of coeffect systems -- indeed, 
our motivation is to develop a \emph{single unified mechanism} for tracking context-dependence.
In this section, we present a more powerful \emph{structural coeffect calculus} \clstr, which is a 
generalization of \clflt. We show that \clflt~can be obtained as a special case of \clstr. The
structural version of our system is also where our work differs more significantly from well-known
effect systems.

% --------------------------------------------------------------------------------------------------

\subsection{Motivation: Tracking array accesses}
\label{sec:structural-motivation}
%\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
%\newcommand{\dnat}{\ident{D}_\ident{nat}}

Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
respectively, but we use richer tag structure.

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\cprd$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
a single variable:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
Before looking at the formal definition, consider the typing of let bindings:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
\pastval{a}{15}+\pastval{c}{10}
\end{array}
\end{equation*}
%
The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.

The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
of $\cvop$ and it denotes empty coeffect. The typing context of the
body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.

To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
with the remaining tags from the body yielding the overall context:
$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
normalization mechanism (described later), this can be further reduced to
$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
\ident{b}.

% --------------------------------------------------------------------------------------------------

\subsection{Structural coeffect tags}
In the previous section, the $\cvop$ operation behaves similarly to the flat $\vee$ operation.
However, the type system does not require some of the semilattice properties, because some uses
are replaced with the $\cprd$ operation. We do not require any properties about the $\cprd$ operation.
For example, in the previous example cannot be commutative (since a tag $15 \cprd 10$ has different 
meaning than $10 \cprd 15$). However, we relate the operations using distributivity laws to allow 
normalization that was hinted above.

\begin{definition}
A \emph{structural coeffect tag structure} $(S, \cprd, \cvop, 0, 1)$ is a tuple where 
$(S, \cvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
by a binary operation $\cprd$ and element $1\in S$ such that
for all $r,s,t\in S$, the following equalities hold: 
%
\begin{subequations}
\begin{align}
\tag{associativity}
r \cvop (s \cvop t) &= (r \cvop s) \cvop t\\
\tag{commutativity}
r \cvop s &= s \cvop r\\
\tag{lower bound}
r \cvop 0 &= r\\
\tag{distributivity}
r \cvop (s \cprd t) &= (r \cvop s) \cprd (r \cvop t)\\
\tag{upper bound}
1 \cvop r &= 1
\end{align}
\end{subequations}
%
\end{definition}
%
% 'upper bound' (not sure what to call it) says that empty context cannot carry information
% looks like this is the upper bound of the bounded lattice ???
%
% commutativity is not strictly required by any law or type preservation, but
% it simplifies the presentation and looks reasonable
%
The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
always accessed using standard variable access). The tag $1$ is used to annotate empty 
variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
to carry an annotation. We explain exactly how this works when we introduce the type system.
The fact that $1$ is the upper bound means that combining it with other coeffect annotations
does not affect it and so empty contexts cannot carry any information.

The structure generalizes the \emph{flat coeffect tag structure} introduced in 
Section~\ref{sec:flat-tag-structure}, but it additionally requires a special element $1$ 
representing the upper bound. Given a flat coeffect structure, we can construct a structural
coeffect structure (but not the other way round). For certain structures, the element $1$
may be already present, but in general, it can be added as a new element.
This construction will be important in Section~\ref{sec:generalization},
where we show that \clstr~calculus generalizes~\clflt.

\begin{lemma}
\label{thm:structural-generalizaton}
A flat coeffect tag structure $(S, \vee, 0)$ implies a structural coeffect tag structure.
\end{lemma}
\begin{proof}
Take $1 \notin S$, then $(S \cup \{1\}, \vee, \vee, 0, 1)$ is a structural coeffect tag 
structure. From the properties of flat coeffect tag structure, we get that the $\vee$ operation 
is associative, commutative and $0$ is the unit with respect to $\vee$.

To prove the distributivity, we need to show $r \vee (s \vee t) = (r \vee s) \vee (r \vee t)$.
This easily follows from commutativity, associativity and idempotence of the flat coeffect tag 
structure.
\end{proof}

% --------------------------------------------------------------------------------------------------

\subsection{Structural coeffect type system}
\label{sec:structural-types}

The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
a \emph{structural coeffect tag structure} $(S,\cprd, \cvop, 0)$.

The system differs from \clflt~in a significant aspect. It contains explicit 
structural rules that manipulate with the context. Such rules allow reordering, duplicating
and other manipulations with variable context. Such rules are known from affine or linear
type systems where they are removed to obtain more restrictive system. In our system, the
rules are present, but they manipulate the variable structure $\Gamma$ as well as the 
associated tag structure $r$.

As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
simple set. Instead, we use the following tree-like structure (which is more similar to 
bunched types than to linear or affine systems):
%
\begin{equation*}
\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
\end{equation*}
%
The syntax $()$ represents an empty context, so the structure defines a binary trees where
leaves are either variables or empty. Contexts such as $\ctyp{1\cprd(r \cprd 1)}{((), (x, ()))}$ 
contain unnecessary number of empty contexts $()$. However, we need to construct them
temporarily, because certain rules require splitting a context and, by our 
definition, the context $(x:\tau)$ is not splitable.

The typing rules of the system are shown in Figure~\ref{fig:structural-coeffect-typing}.
Many of the structural rules are expressed in terms of a helper judgement
$\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}}$.

\paragraph{Standard rules.}
Variable access (\rulest{var}) annotates the corresponding variable with an empty coeffect
$0$. The \rulest{fun} rule assumes that the context of the body can be split into the
variable of the function and other (potentially empty) context and it attaches the coeffect
associated with the function variable to the resulting function type $\ctyp{s}{\tau_1} \rightarrow \tau_2$.

The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
expression, argument and the function type respectively. 
The result of evaluating the argument in the context $t$ is passed to the function that
requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
the combination of coeffects $r \cvop t$. The variable context required to evaluate the 
function value is independent and so it is annotated just with coeffects $s$.
Finally, the \rulest{let} rule is derived from let binding and application, but we show
it separately to aid the understanding.

\paragraph{Structural rules.}
The remaining rules are not syntax-directed. They are embedded using the \rulest{ctx}
rule and expressed using a helper judgement $\ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}}$
that says that the context on the left-hand side can be transformed to the context on the
right-hand side. The transformation can be applied to any part of the context, which 
is captured using the \rulest{nest} rule (it is sufficient to apply the transformation 
on the left part of the context; the right part can be transformed using \rulest{exch}).

The \rulest{empty} rule allows attaching empty context to any existing context. The rule 
is needed to type-check lambda abstractions that do not capture any outer variable.
The \rulest{weak} rule is similar, but it represents \emph{weakening} where an unused variable is 
added. The associated coeffect tag does not associate context information with the variable.
This is needed to type-check lambda abstraction that does not use the argument.

The \rulest{contr} rule is a limited form of contraction. When a variable appears repeatedly,
it can be reduced to a single occurrence. The rule is needed to satisfy the side condition
of \rulest{fun} when the body uses the argument repeatedly. The two associativity rules
together with \rulest{exch} provide ways to rearrange variables. Finally, \rulest{sub} represents
sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.

% --------------------------------------------------------------------------------------------------

\subsection{Properties of reductions}
\label{sec:structural-metatheory}
Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
operational meaning, because that will differ for every concrete application. For example,
when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
access to a specified element.

Just like previously, we can state general properties of the reductions. As the syntax of expressions
is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
are also the same and can be found in Figure~\ref{fig:flat-substred}.

The structural coeffect calculus \clstr~associates information with individual variables. This means
that when an expression requires certain context, we know from what scope it comes -- the
context must be provided by a scope that defines the associated variable, which is either a
lambda abstraction or global scope. This distinguishes the structural system from the flat system
where context could have been provided by any scope and the lambda rule allowed arbitrary 
splitting of context requirements between the two scopes (or declaration and caller site).

\paragraph{Internalized substitution.}
Before looking at properties of the evaluation, we consider let binding, which can be 
viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
application and abstraction as follows.

\begin{lemma}[Definition of let binding]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
\end{lemma}
\begin{proof}
The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
correspond with the typing rule \rulest{let}:
\begin{equation*}
\inference[]
  {\inference
    {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
\end{equation*}
\end{proof}

The term $e_2$ which is substituted in $e_1$ is checked in a different variable
and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
the variable that we are substituting for. The let binding captures substitution for a 
specific variable (the context is of a form $\ctyp{r \cprd s}{\Gamma, v:\tau}$).
For a general substitution, we need to define the notion of context with a hole.

\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
\paragraph{Substitution and holes.}
In \clstr, the structure of the variable context is not a set, but a tree. When substituting
for a variable, we need to replace the variable in the context with the context of the
substituted expression. In general, this can occur anywhere in the tree. To formulate the
statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
in the free variable context and in a corresponding part of the coeffect tag:

\begin{equation*}
\begin{array}{rcl}
\Delta[-] &::=& \ctyp{1}{()} \\
       & \sep & \ctyp{r}{(x:\tau)}\\
       & \sep & \ctyp{-}{(-)} \\
       & \sep & \ctyp{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
\end{array}
\end{equation*}
%
Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
general substitution lemma for \clstr.

\begin{figure}

\begin{equation*}
\begin{array}{rcl}
                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
\chole{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \cprd r_2'}{(\Gamma_1', \Gamma_2')}\\
&\multicolumn{2}{l} {
  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
}
\end{array}
\end{equation*}
\caption{The definition of hole filling operation for $\Delta[-]$}
\label{fig:structural-hole-fill}
\vspace{-2em}
\end{figure}

\begin{lemma}[Substitution Lemma]
\label{thm:structural-subst}
If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
then $\chole{r}{\Gamma}{R \cvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
\end{lemma} 
\begin{proof}
Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
$(S, \cvop, 0, \cprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
\end{proof}

\begin{theorem}[Subject reduction]
\label{thm:structural-reduction}
If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
\end{theorem}
\begin{proof}
Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
\end{proof}

\paragraph{Local soundness and completeness.}
As with the previous calculus, we want to guarantee that the introduction and
elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
This can be done by showing \emph{local soundness} and \emph{local completeness},
which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
case of subject reduction and the latter is proved by a simple derivation:
%
\begin{theorem}[Local soundness]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
\end{theorem}
\begin{proof}
Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
\end{proof}
%
\begin{theorem}[Local completeness]
If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
\end{theorem}
\begin{proof} The property is proved by the following typing derivation:
%
\begin{equation*}
\inference
  {\inference
     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
     {\ctyp{r \cprd (s \cvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\qedhere
\end{equation*}
\end{proof}
%
In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
guarantees that $s\cvop0 = s$. Recall that in \clflt, the typing derivation for
$\lambda x.f x$ required for local completeness was not the only possible derivation.
In the last step, it was possible to split the coeffect tag arbitrarily between the 
context and the function type.

In the \clstr~calculus, this is not, in general, the case. The $\cprd$ operator is not required 
to be associative and to have units and so a unique splitting may exist.  For example, if we
define $\cprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
$t$, there are unique $r$ and $s$ such that $t = r \cprd s$. However, if the $\cprd$ operation 
has additional properties, then there may be other possible derivation.

% ==================================================================================================

\begin{figure*}[t]
\newcommand{\uu}[1]{\overline{#1}}
\begin{equation*}
\cosem{\ctyp{r_1 \cprd \ldots \cprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
  \ctyp{r_1 \cprd \ldots \cprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
\end{equation*}
\begin{equation*}
\begin{array}{rcll}
\cosem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
    \epsilon_0                                                                   &\quad$\ruless{var}$
\\[0.32em]
\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
    \Lambda (\cosem{\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
\\[0.32em]
\cosem{\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
    (\lambda (\gamma_1, \gamma_2) \rightarrow
    \cosem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
    (\cobind{\cosem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
\\[0.32em]
\cosem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
    \cosem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \cosem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{rclll}
\cosem{\ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)}}}
&=& \cmerge_{r, s} \circ (\cosem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
\textnormal{\rulessc{nest}}
\\[0.8em]
\cosem{\ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}}
&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
\textnormal{\rulessc{exch}}
\\[0.8em]
\cosem{\ctxtrans{\ctyp{r \cprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
&=&\ident{fst} \circ \csplit_{r, 1}&&
\textnormal{\rulessc{empty}}
\\[0.8em]
\cosem{\ctxtrans{ \ctyp{r\cprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
&=&\ident{fst} \circ \csplit_{r, 0}&&
\textnormal{\rulessc{weak}}
\\[0.8em]
\cosem{\ctxtrans{\ctyp{r \cvop s}{(x:\tau)}}{
    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}}
&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
\textnormal{\rulessc{contr}}
\\[0.8em]
\cosem{\ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
&=& \cmerge_{r \cprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \cprd t}&&
\textnormal{\rulessc{assoc}}
\\[0.8em]
\cosem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
&=&
\iota_{r, s}
&&
\textnormal{\rulessc{sub}}
\end{array}
\end{equation*}

\caption{Categorical semantics for~\clstr}
\label{fig:structural-semantics}
\vspace{-0.5em}
\end{figure*}

% --------------------------------------------------------------------------------------------------

\section{Semantics of structural coeffects}
\label{sec:structural-semantics}

The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
of the variable context. Thus an expression $\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
is modelled as a function $\ctyp{r \cprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.

As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
system is not a simple finite product, but instead a binary tree. To model this, we do not use
ordinary products in the domain of the semantic function, but instead use a special 
constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
the tag structure.

\subsection{Structural tagged comonads}
\label{sec:structural-tagged-comonads}

To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
$T^{r \cvop s} \tau_1 \rightarrow \tau_3$ and we use the $\cvop$ operation to combine the 
context-requirements.

However, functions that do not exist in context have only a single input variable (with a 
single corresponding tag). To model complex variable contexts, we need two additional
operations that allow manipulation with the variable context. Similarly to the model of \clflt,
we also require operations that model duplication and sub-coeffecting:

\begin{definition}[Structural tagged comonad]
Given a structural coeffect tag structure $(S, \cprd, \cvop, 0, 1)$ a \emph{structural tagged comonad} is a 
tagged comonad over $(S, \cvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
%
\begin{equation*}
\begin{array}{rcl}
\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \cprd s)}(A \ctimes B)\\
\csplit_{r, s} &:& T^{(r \cprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
\end{array}
\end{equation*}
%
% Delta is still defineable
% ?? \cdupl_{r, s}  &:& T^{(r \cvop s)}A \rightarrow T^r A \times T^s A\\
%
And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
%
%  TODO: laws?
%
\end{definition}

The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
and it can still be used to define a family of mappings that represents \emph{duplicating}
of variables while splitting the additional coeffect tags:
%
\begin{equation*}
\begin{array}{l}
\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
\end{array}
\end{equation*}
%
The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
but with two differences. Firstly, it combines tags using $\cprd$ instead of $\vee$, 
which corresponds to the fact that the variable context now consists of two parts (a tree node).
Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
(instead of $A \times B$ as previously).

To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
variables. The operation was not needed for \clflt, because there \emph{splitting} could
be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
with $r \cprd s$ that carries $A \ctimes B$.

Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
Before looking at them, we finish our discussion of categorical semantics.

\paragraph{Categorical notes.}
The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
variables in the context. Given a context modelled as $T^{r \cprd s}(A \ctimes B)$,
we can lift a morphism $f$ to $\hat{T}^{r \cprd s}(f)$, but we cannot manipulate the
variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.

This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
could apply $\cdupl_{r, s}$ on the input (if the tag $r \cprd s$ coincided
with tag $r \cvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.

This restriction is intentional -- at the semantic level, it prevents manipulations
with the context that would break the correspondence between tag structure and the
product structure.

\subsection{Categorical semantics}
The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
It uses the \emph{structural tagged comonad} structure introduced in the previous section,
together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
operations:
\begin{equation*}
\begin{array}{rcl}
\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
\end{array}
\end{equation*}

When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
$\cprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
type signature in $\cmerge_{r, s}$).

The rule \ruless{app} now uses the operation $\csplit_{s, (r \cvop t)}$ instead of $\cdupl_{s, (r \cvop t)}$,
which means that it splits the context instead of duplicating it. This makes the system more 
structural -- the expressions use disjunctive parts of the context -- and also explains why the
composed coeffect tag is $s \cprd (r \cvop t)$.

The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
of $\cosem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
takes a context, converts it to a new context which is compatible with the original expression $e$.
The context manipulation rules work as follows:

\begin{itemize}
\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
  Finally, they re-construct a single context using $\cmerge_{r, s}$.

\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
  and discard one part (containing either an unused variable or an empty context).

\item If we interpreted \rulessc{contr} by applying functor $T^{r\cvop s}$ to a function that duplicates
  a variable, the resulting context would be $\ctyp{r \cvop s}(x:\tau, x:\tau)$, which would break the
  correspondence between coeffect tag and context variable structure. However, that interpretation would be
  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
  a tag $r \cvop s$ into $r \cprd s$.

\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
  manipulates it (using \ident{assoc}) and then re-constructs it.

\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
  variable using the primitive natural transformation $\iota_{r, s}$.
\end{itemize}

\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
attaching the context separately to individual variables. This an expression
$\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
approach largely complicates the definition of application (where tag of all 
variables in a context is affected). Moreover, it makes it impossible to express
\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.

\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).

This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
the following natural transformation has to be provided:
%
\begin{equation*}
\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
\end{equation*}
%
This variant of the system could be used to define a system that ensures that all provided 
context is used and is not over-approximated. This difference is similar to the difference between
affine type systems (where a variable can be used at most once) and linear type systems 
(where a variable has to be used exactly once).

% --------------------------------------------------------------------------------------------------

\section{Examples of structural coeffects}

\subsection{Example: Neededness analysis}
\label{sec:structural-neededness}

\subsection{Example: Tracking tainting}
\label{sec:structural-tainting}

\section{Summary}
Oops!

### 2013-06-11 text/Abstract.tex
Refactor rules

% Abstract

\pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Abstract} % Abstract name

Static type systems are the most common form of program verification. In their most basic form, type
systems guarantee that a computation always returns a value of some type. The work on monads
and effect systems extends type systems to guarantee properties about \emph{effectful 
computations} that perform side-effects like non-termination, I/O operations or exceptions.

In this thesis we extend type systems to guarantee properties about \emph{context-dependent
computations}. Such computations are increasingly common in modern software. For example, applications
accessing data need to guarantee that the right data is available and distributed systems need to 
guarantee that resources are available on specific nodes.

We develop three \emph{coeffect} type systems that are capable of capturing different context-dependent
properties. All three systems are based on the unifying semantic structure called \emph{indexed
comonads}, but they use it differently.

Our \emph{flat coeffect calculus} tracks one annotation about the entire context. It can be used to
keep a set of required resources or a set of supported platforms. Our \emph{structural coeffect 
calculus} tracks one annotation for each variable. It can be used to track liveness of variables
or number of past values needed in data-flow computations. Finally, our \emph{coeffect meta-language}
embeds contextual values directly into a language. It can be used to guarantee security properties
of computations by using tainting.

\endgroup

\vfill

### 2013-06-11 text/Chapter04.tex
Refactor rules

@@
-\chapter{Structural coeffects} 
+\chapter{Structural coeffect language} 
 \label{ch:structural}
@@
 \newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
+
+\newcommand{\csprd}[0]{ \times }
+\newcommand{\csvop}[0]{ \vee }
 
@@
 % ==================================================================================================
-\newcommand{\ctxtrans}[2]{#1 \Rightarrow_c #2}
+
 \begin{figure*}[t]
@@
 \inference
-  {\ctyp{r \cprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
+  {\ctyp{r \csprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
   {\ctyp{r}{\Gamma} \vdash \lambda v.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }&
@@
    \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1 }
-  {\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
+  {\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
 \textnormal{\rulest{app}}
@@
 \inference
-  {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
+  {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
   \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1}
-  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
+  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
 \textnormal{\rulest{let}}
@@
     {  \ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}} }
-    {  \ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}} }
+    {  \ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)}} }
 }&
@@
 \\[2.0em]
-  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}&
+  \ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}&
 \textnormal{\rulest{exch}}
 \\[0.8em]
-  \ctxtrans{ \ctyp{r \cprd 1}{(\Gamma, ())} &}{&
+  \ctxtrans{ \ctyp{r \csprd 1}{(\Gamma, ())} &}{&
     \ctyp{r}{\Gamma}}&
@@
 \\[0.8em]
-  \ctxtrans{\ctyp{r\cprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
+  \ctxtrans{\ctyp{r\csprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
 \textnormal{\rulest{weak}}
 \\[0.8em]
-  \ctxtrans{\ctyp{r \cvop s}{(x:\tau)} &}{&
-    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}&
+  \ctxtrans{\ctyp{r \csvop s}{(x:\tau)} &}{&
+    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}&
 \textnormal{\rulest{contr}}
 \\[0.8em]
-  \ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
-    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
+  \ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
+    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
 \textnormal{\rulest{assoc}}
@@
 \multicolumn{3}{c}{
-  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \cvop r = s)
+  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \csvop r = s)
 }&
@@
 %  {  \ctxtrans{\ctyp{s}{\Gamma_2}}{\ctyp{s'}{\Gamma_2'}} }
-%  {  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\cprd s'}{(\Gamma_1, \Gamma_2')}} }&
+%  {  \ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\csprd s'}{(\Gamma_1, \Gamma_2')}} }&
 % \textnormal{\rulest{right}}
@@
 some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
-To track information about individual variables, we use a product-like operation $\cprd$ on tags 
+To track information about individual variables, we use a product-like operation $\csprd$ on tags 
 to mirrors the product structure of variables. For example:
@@
 \begin{array}{l}
-\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
+\ctyp{5 \csprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
   \vdash
@@
 %
-The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
+The coeffect tag $5 \csprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
 that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
@@
 \begin{array}{l}
-\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
+\ctyp{5 \csvop 10}{(\ident{c}:\dnat)}
   \vdash
@@
 %
-In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
+In this example, the operation $\csvop$ would be the \emph{max} function and so $5 \csvop 10 = 10$.
 Before looking at the formal definition, consider the typing of let bindings:
@@
 The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
-context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
-of $\cvop$ and it denotes empty coeffect. The typing context of the
-body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.
+context is $\ctyp{0 \csprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
+of $\csvop$ and it denotes empty coeffect. The typing context of the
+body is $\ctyp{15 \csprd 10}{(\ident{a}, \ident{c})}$.
 
 To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
-of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
+of the context in which \ident{c} was defined using the $\csvop$ operation. This is then combined 
 with the remaining tags from the body yielding the overall context:
-$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
+$\ctyp{15 \csprd (10 \csvop (0 \csprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
 normalization mechanism (described later), this can be further reduced to
-$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
+$\ctyp{(15 \csvop 10) \csprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
 we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
@@
 \subsection{Structural coeffect tags}
-In the previous section, the $\cvop$ operation behaves similarly to the flat $\vee$ operation.
+In the previous section, the $\csvop$ operation behaves similarly to the flat $\vee$ operation.
 However, the type system does not require some of the semilattice properties, because some uses
-are replaced with the $\cprd$ operation. We do not require any properties about the $\cprd$ operation.
-For example, in the previous example cannot be commutative (since a tag $15 \cprd 10$ has different 
-meaning than $10 \cprd 15$). However, we relate the operations using distributivity laws to allow 
+are replaced with the $\csprd$ operation. We do not require any properties about the $\csprd$ operation.
+For example, in the previous example cannot be commutative (since a tag $15 \csprd 10$ has different 
+meaning than $10 \csprd 15$). However, we relate the operations using distributivity laws to allow 
 normalization that was hinted above.
@@
 \begin{definition}
-A \emph{structural coeffect tag structure} $(S, \cprd, \cvop, 0, 1)$ is a tuple where 
-$(S, \cvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
-by a binary operation $\cprd$ and element $1\in S$ such that
+A \emph{structural coeffect tag structure} $(S, \csprd, \csvop, 0, 1)$ is a tuple where 
+$(S, \csvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
+by a binary operation $\csprd$ and element $1\in S$ such that
 for all $r,s,t\in S$, the following equalities hold: 
@@
 \tag{associativity}
-r \cvop (s \cvop t) &= (r \cvop s) \cvop t\\
+r \csvop (s \csvop t) &= (r \csvop s) \csvop t\\
 \tag{commutativity}
-r \cvop s &= s \cvop r\\
+r \csvop s &= s \csvop r\\
 \tag{lower bound}
-r \cvop 0 &= r\\
+r \csvop 0 &= r\\
 \tag{distributivity}
-r \cvop (s \cprd t) &= (r \cvop s) \cprd (r \cvop t)\\
+r \csvop (s \csprd t) &= (r \csvop s) \csprd (r \csvop t)\\
 \tag{upper bound}
-1 \cvop r &= 1
+1 \csvop r &= 1
 \end{align}
@@
 The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
-a \emph{structural coeffect tag structure} $(S,\cprd, \cvop, 0)$.
+a \emph{structural coeffect tag structure} $(S,\csprd, \csvop, 0)$.
 
@@
 The syntax $()$ represents an empty context, so the structure defines a binary trees where
-leaves are either variables or empty. Contexts such as $\ctyp{1\cprd(r \cprd 1)}{((), (x, ()))}$ 
+leaves are either variables or empty. Contexts such as $\ctyp{1\csprd(r \csprd 1)}{((), (x, ()))}$ 
 contain unnecessary number of empty contexts $()$. However, we need to construct them
@@
 requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
-the combination of coeffects $r \cvop t$. The variable context required to evaluate the 
+the combination of coeffects $r \csvop t$. The variable context required to evaluate the 
 function value is independent and so it is annotated just with coeffects $s$.
@@
   {\inference
-    {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
+    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
     {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
    \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
-  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
+  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
 \end{equation*}
@@
 the variable that we are substituting for. The let binding captures substitution for a 
-specific variable (the context is of a form $\ctyp{r \cprd s}{\Gamma, v:\tau}$).
+specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
 For a general substitution, we need to define the notion of context with a hole.
@@
        & \sep & \ctyp{-}{(-)} \\
-       & \sep & \ctyp{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
+       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
 \end{array}
@@
                              \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
-\chole{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \cprd r_2'}{(\Gamma_1', \Gamma_2')}\\
+\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
 &\multicolumn{2}{l} {
@@
 If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
-then $\chole{r}{\Gamma}{R \cvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
+then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
 \end{lemma} 
@@
 Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
-$(S, \cvop, 0, \cprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
+$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
 \end{proof}
@@
       \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
-     {\ctyp{r \cprd (s \cvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
+     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
   {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
@@
 In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
-guarantees that $s\cvop0 = s$. Recall that in \clflt, the typing derivation for
+guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
 $\lambda x.f x$ required for local completeness was not the only possible derivation.
@@
 
-In the \clstr~calculus, this is not, in general, the case. The $\cprd$ operator is not required 
+In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
 to be associative and to have units and so a unique splitting may exist.  For example, if we
-define $\cprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
-$t$, there are unique $r$ and $s$ such that $t = r \cprd s$. However, if the $\cprd$ operation 
+define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
+$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
 has additional properties, then there may be other possible derivation.
@@
 \begin{equation*}
-\cosem{\ctyp{r_1 \cprd \ldots \cprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-  \ctyp{r_1 \cprd \ldots \cprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
+\cosem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
+  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
 \end{equation*}
@@
 \cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \Lambda (\cosem{\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
+    \Lambda (\cosem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
 \\[0.32em]
-\cosem{\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
+\cosem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
     (\lambda (\gamma_1, \gamma_2) \rightarrow
@@
 \begin{array}{rclll}
-\cosem{\ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)}}}
+\cosem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
 &=& \cmerge_{r, s} \circ (\cosem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
@@
 \\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}}
+\cosem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
 &=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
@@
 \\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \cprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
+\cosem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
 &=&\ident{fst} \circ \csplit_{r, 1}&&
@@
 \\[0.8em]
-\cosem{\ctxtrans{ \ctyp{r\cprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
+\cosem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
 &=&\ident{fst} \circ \csplit_{r, 0}&&
@@
 \\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \cvop s}{(x:\tau)}}{
-    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}}
+\cosem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
+    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
 &=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
@@
 \\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
-    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
-&=& \cmerge_{r \cprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
-        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \cprd t}&&
+\cosem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
+    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
+&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
+        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
 \textnormal{\rulessc{assoc}}
@@
 \clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
-of the variable context. Thus an expression $\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
-is modelled as a function $\ctyp{r \cprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
+of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
+is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
 
@@
 $T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
-$T^{r \cvop s} \tau_1 \rightarrow \tau_3$ and we use the $\cvop$ operation to combine the 
+$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
 context-requirements.
@@
 \begin{definition}[Structural tagged comonad]
-Given a structural coeffect tag structure $(S, \cprd, \cvop, 0, 1)$ a \emph{structural tagged comonad} is a 
-tagged comonad over $(S, \cvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
+Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
+tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
 $\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
@@
 \begin{array}{rcl}
-\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \cprd s)}(A \ctimes B)\\
-\csplit_{r, s} &:& T^{(r \cprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
+\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
+\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
 \end{array}
@@
 % Delta is still defineable
-% ?? \cdupl_{r, s}  &:& T^{(r \cvop s)}A \rightarrow T^r A \times T^s A\\
+% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
 %
@@
 The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
-but with two differences. Firstly, it combines tags using $\cprd$ instead of $\vee$, 
+but with two differences. Firstly, it combines tags using $\csprd$ instead of $\vee$, 
 which corresponds to the fact that the variable context now consists of two parts (a tree node).
@@
 the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
-with $r \cprd s$ that carries $A \ctimes B$.
+with $r \csprd s$ that carries $A \ctimes B$.
 
@@
 Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
-variables in the context. Given a context modelled as $T^{r \cprd s}(A \ctimes B)$,
-we can lift a morphism $f$ to $\hat{T}^{r \cprd s}(f)$, but we cannot manipulate the
+variables in the context. Given a context modelled as $T^{r \csprd s}(A \ctimes B)$,
+we can lift a morphism $f$ to $\hat{T}^{r \csprd s}(f)$, but we cannot manipulate the
 variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
@@
 This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
-could apply $\cdupl_{r, s}$ on the input (if the tag $r \cprd s$ coincided
-with tag $r \cvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
+could apply $\cdupl_{r, s}$ on the input (if the tag $r \csprd s$ coincided
+with tag $r \csvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
 
@@
 variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
-$\cprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
+$\csprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
 type signature in $\cmerge_{r, s}$).
 
-The rule \ruless{app} now uses the operation $\csplit_{s, (r \cvop t)}$ instead of $\cdupl_{s, (r \cvop t)}$,
+The rule \ruless{app} now uses the operation $\csplit_{s, (r \csvop t)}$ instead of $\cdupl_{s, (r \csvop t)}$,
 which means that it splits the context instead of duplicating it. This makes the system more 
 structural -- the expressions use disjunctive parts of the context -- and also explains why the
-composed coeffect tag is $s \cprd (r \cvop t)$.
+composed coeffect tag is $s \csprd (r \csvop t)$.
 
@@
 
-\item If we interpreted \rulessc{contr} by applying functor $T^{r\cvop s}$ to a function that duplicates
-  a variable, the resulting context would be $\ctyp{r \cvop s}(x:\tau, x:\tau)$, which would break the
+\item If we interpreted \rulessc{contr} by applying functor $T^{r\csvop s}$ to a function that duplicates
+  a variable, the resulting context would be $\ctyp{r \csvop s}(x:\tau, x:\tau)$, which would break the
   correspondence between coeffect tag and context variable structure. However, that interpretation would be
@@
   rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
-  a tag $r \cvop s$ into $r \cprd s$.
+  a tag $r \csvop s$ into $r \csprd s$.
 
@@
 attaching the context separately to individual variables. This an expression
-$\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
+$\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
 $\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 

### 2013-06-11 text/Chapter05.tex
Refactor rules

--- 2013-06-05 text/Chapter04.tex
@@
-\chapter{Structural coeffects} 
-\label{ch:structural}
-
-\newcommand{\ctimes}{ \hat{\times} }
-\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
-
-% alculus names and rule names
-\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
-\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
-
-\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
-\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
-\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
-\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
-\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}
+\chapter{Coeffect meta-language} 
+\label{ch:metalanguage}
 
@@
 \section{Introduction}
-\label{sec:structural-intro}
+\label{sec:metalanguage-intro}
 
-We already talked about tainting in pathways to coeffects... here is more stuff!
-
-~
-
-~
-
-Our second motivating example is the use of \emph{tainting} for the prevention of SQL injection 
-attacks \cite{app-secure-tainting}. The idea is to (statically) check that SQL commands cannot be 
-directly constructed from, potentially dangerous, inputs provided by the user. Consider the type
-checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
-\ident{msg}~+~\ident{name}
-\end{array}
-\end{equation*}
-%
-We say that a variable is \emph{tainted} if it may come from untrusted input. In the above example,
-\ident{id} must not be tainted. Otherwise, the attacker could specify values such as 
-\str{"1; DROP TABLE Users"}. The variable \ident{msg} may or may not be tainted, because it is not
-used in protected context (i.e.~to construct an SQL query). Using the coeffect calculus, security
-information can be naturally attached to the context. We use tag \ident T for tainted and 
-\ident N for not tainted:
-%
-\begin{equation*}
-\ctyp{\ident T \times \ident N}{(\ident{id}:\ident{string}, \ident{msg}:\ident{string})} 
-  \vdash (\ldots) : \ident{string}
-\end{equation*}
-%
-The tags \ident T and \ident N correspond to variables \ident{id} and \ident{msg}, respectively.
-The checking of tainting is a special case of checking of \emph{non-interference} in 
-\emph{secure information flow} \cite{app-secure-information-flow}. Languages such as FlowCaml 
-\cite{app-security-flowcaml} can check such properties statically, but it is also possible to 
-embed such checking in a monadic library \cite{monad-secure-flow}.
-
-Assuming the monadic type $\mtyp{s}{\tau}$ represents a value of type $\tau$ 
-in a security context $s$ (either \ident T or \ident N): 
-%
-\begin{equation*}
-\ident{id} : \mtyp{\ident T}{\ident{string}}, \ident{msg} : \mtyp{\ident N}{\ident{string}} 
-  \vdash (\ldots) : \mtyp{\ident T}{\ident{string}}
-\end{equation*}
-%
-In this example, the result is tagged with \ident T to denote that it may be tainted.
-If the result was tagged with \ident N, the first parameter would also have to be \ident N
-(and we would get fewer information about the function).
-
-The monadic encoding has an unfortunate property from the semantic perspective. 
-The function has monadic structure over the \emph{domain} as well as over the \emph{codomain}
-and thus it does not match the desirable Kleisli model of expressions $\tau_1 \times \ldots \tau_n \rightarrow \mtyp{}{\tau}$. 
-This makes the it difficult to give standard composable semantics of the library.
-
-In coeffect calculus, expressions are modelled using the coKleisli form 
-$\ctyp{r_1 \times \ldots \times r_n}{(\tau_1 \times \ldots \tau_n)} \rightarrow \tau$ with a tags 
-$r_n$ that specify taintedness of inputs. As discussed in detail in Section~\ref{sec:structural-tainting}, 
-this model gives a simple comonadic semantics of language-based security using  tainting.
-
-%---------------------------------------------------------------------------------------------------
-%===================================================================================================
-
-% ==================================================================================================
-\newcommand{\ctxtrans}[2]{#1 \Rightarrow_c #2}
-\begin{figure*}[t]
-% Constants, if we wanted them, would look like this
-%
-% \inference[]
-%   {c : \tau \in \Phi}
-%   {\ctyp{1}{()} \vdash c : \tau}&
-% \textnormal{\rulest{const}}
-% \\[1.6em]
-
-%\paragraph{Typing rules}
-\begin{equation*}
-\boxed{\ctyp{r}{\Gamma} \vdash e : \tau }
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{cl}
-
-\inference[]
-  {}
-  {\ctyp{0}{(v:\tau)} \vdash v : \tau }&
-\textnormal{\rulest{var}}
-\\[1.8em]
-\inference
-  {\ctyp{r \cprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
-  {\ctyp{r}{\Gamma} \vdash \lambda v.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }&
-\textnormal{\rulest{fun}}
-\\[1.8em]
-\inference
-  {\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1 }
-  {\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
-\textnormal{\rulest{app}}
-\\[1.8em]
-\inference
-  {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
-  \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1}
-  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
-\textnormal{\rulest{let}}
-\\[1.8em]
-\inference
-  {\ctyp{r}{\Gamma} \vdash e : \tau &
-   \ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}
-  {\ctyp{r'}{\Gamma'} \vdash e : \tau}&
-\textnormal{\rulest{ctx}}
-\end{array}
-\end{equation*}
-
-%
-% Empty variable context corresponds to unit, but things get messy
-%
-
-%\paragraph{Context rules}
-\begin{equation*}
-\boxed{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}} }
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rcll}
-\multicolumn{3}{c}{
-  \inference
-    {  \ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}} }
-    {  \ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}} }
-}&
-\textnormal{\rulest{nest}}
-\\[2.0em]
-  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}&
-\textnormal{\rulest{exch}}
-\\[0.8em]
-  \ctxtrans{ \ctyp{r \cprd 1}{(\Gamma, ())} &}{&
-    \ctyp{r}{\Gamma}}&
-\textnormal{\rulest{empty}}
-\\[0.8em]
-  \ctxtrans{\ctyp{r\cprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
-\textnormal{\rulest{weak}}
-\\[0.8em]
-  \ctxtrans{\ctyp{r \cvop s}{(x:\tau)} &}{&
-    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}&
-\textnormal{\rulest{contr}}
-\\[0.8em]
-  \ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
-    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
-\textnormal{\rulest{assoc}}
-\\[0.8em]
-\multicolumn{3}{c}{
-  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \cvop r = s)
-}&
-\textnormal{\rulest{sub}}
-\end{array}
-\end{equation*}
-
-% We do not need this rule (can be expressed in terms of the other one)
-%
-% \inference
-%  {  \ctxtrans{\ctyp{s}{\Gamma_2}}{\ctyp{s'}{\Gamma_2'}} }
-%  {  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\cprd s'}{(\Gamma_1, \Gamma_2')}} }&
-% \textnormal{\rulest{right}}
-
-\caption{Type system for \clstr}
-\label{fig:structural-coeffect-typing}
-\end{figure*}
-
-\newpage
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Structural coeffect system}
-\label{sec:structural}
-
-The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
-we need to track context-dependence in a more fine-grained way. To track neededness or security, 
-we need to associate information with individual \emph{variables} of the context.
-
-At the same time, we want to avoid developing multiple variants of coeffect systems -- indeed, 
-our motivation is to develop a \emph{single unified mechanism} for tracking context-dependence.
-In this section, we present a more powerful \emph{structural coeffect calculus} \clstr, which is a 
-generalization of \clflt. We show that \clflt~can be obtained as a special case of \clstr. The
-structural version of our system is also where our work differs more significantly from well-known
-effect systems.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Motivation: Tracking array accesses}
-\label{sec:structural-motivation}
-%\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
-%\newcommand{\dnat}{\ident{D}_\ident{nat}}
-
-Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
-functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
-respectively, but we use richer tag structure.
-
-As an example, consider a language that allows us to get a value of a variable (representing
-some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
-To track information about individual variables, we use a product-like operation $\cprd$ on tags 
-to mirrors the product structure of variables. For example:
-%
-\begin{equation*}
-\begin{array}{l}
-\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
-  \vdash
-    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
-\end{array}
-\end{equation*}
-%
-The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
-that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
-for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
-a single variable:
-%
-\begin{equation*}
-\begin{array}{l}
-\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
-  \vdash
-    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
-\end{array}
-\end{equation*}
-%
-In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
-Before looking at the formal definition, consider the typing of let bindings:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
-\pastval{a}{15}+\pastval{c}{10}
-\end{array}
-\end{equation*}
-%
-The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
-is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
-The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.
-
-The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
-context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
-of $\cvop$ and it denotes empty coeffect. The typing context of the
-body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.
-
-To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
-of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
-with the remaining tags from the body yielding the overall context:
-$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
-normalization mechanism (described later), this can be further reduced to
-$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
-we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
-\ident{b}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffect tags}
-In the previous section, the $\cvop$ operation behaves similarly to the flat $\vee$ operation.
-However, the type system does not require some of the semilattice properties, because some uses
-are replaced with the $\cprd$ operation. We do not require any properties about the $\cprd$ operation.
-For example, in the previous example cannot be commutative (since a tag $15 \cprd 10$ has different 
-meaning than $10 \cprd 15$). However, we relate the operations using distributivity laws to allow 
-normalization that was hinted above.
-
-\begin{definition}
-A \emph{structural coeffect tag structure} $(S, \cprd, \cvop, 0, 1)$ is a tuple where 
-$(S, \cvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
-by a binary operation $\cprd$ and element $1\in S$ such that
-for all $r,s,t\in S$, the following equalities hold: 
-%
-\begin{subequations}
-\begin{align}
-\tag{associativity}
-r \cvop (s \cvop t) &= (r \cvop s) \cvop t\\
-\tag{commutativity}
-r \cvop s &= s \cvop r\\
-\tag{lower bound}
-r \cvop 0 &= r\\
-\tag{distributivity}
-r \cvop (s \cprd t) &= (r \cvop s) \cprd (r \cvop t)\\
-\tag{upper bound}
-1 \cvop r &= 1
-\end{align}
-\end{subequations}
-%
-\end{definition}
-%
-% 'upper bound' (not sure what to call it) says that empty context cannot carry information
-% looks like this is the upper bound of the bounded lattice ???
-%
-% commutativity is not strictly required by any law or type preservation, but
-% it simplifies the presentation and looks reasonable
-%
-The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
-always accessed using standard variable access). The tag $1$ is used to annotate empty 
-variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
-to carry an annotation. We explain exactly how this works when we introduce the type system.
-The fact that $1$ is the upper bound means that combining it with other coeffect annotations
-does not affect it and so empty contexts cannot carry any information.
-
-The structure generalizes the \emph{flat coeffect tag structure} introduced in 
-Section~\ref{sec:flat-tag-structure}, but it additionally requires a special element $1$ 
-representing the upper bound. Given a flat coeffect structure, we can construct a structural
-coeffect structure (but not the other way round). For certain structures, the element $1$
-may be already present, but in general, it can be added as a new element.
-This construction will be important in Section~\ref{sec:generalization},
-where we show that \clstr~calculus generalizes~\clflt.
-
-\begin{lemma}
-\label{thm:structural-generalizaton}
-A flat coeffect tag structure $(S, \vee, 0)$ implies a structural coeffect tag structure.
-\end{lemma}
-\begin{proof}
-Take $1 \notin S$, then $(S \cup \{1\}, \vee, \vee, 0, 1)$ is a structural coeffect tag 
-structure. From the properties of flat coeffect tag structure, we get that the $\vee$ operation 
-is associative, commutative and $0$ is the unit with respect to $\vee$.
-
-To prove the distributivity, we need to show $r \vee (s \vee t) = (r \vee s) \vee (r \vee t)$.
-This easily follows from commutativity, associativity and idempotence of the flat coeffect tag 
-structure.
-\end{proof}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffect type system}
-\label{sec:structural-types}
-
-The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
-The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
-a \emph{structural coeffect tag structure} $(S,\cprd, \cvop, 0)$.
-
-The system differs from \clflt~in a significant aspect. It contains explicit 
-structural rules that manipulate with the context. Such rules allow reordering, duplicating
-and other manipulations with variable context. Such rules are known from affine or linear
-type systems where they are removed to obtain more restrictive system. In our system, the
-rules are present, but they manipulate the variable structure $\Gamma$ as well as the 
-associated tag structure $r$.
-
-As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
-simple set. Instead, we use the following tree-like structure (which is more similar to 
-bunched types than to linear or affine systems):
-%
-\begin{equation*}
-\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
-\end{equation*}
-%
-The syntax $()$ represents an empty context, so the structure defines a binary trees where
-leaves are either variables or empty. Contexts such as $\ctyp{1\cprd(r \cprd 1)}{((), (x, ()))}$ 
-contain unnecessary number of empty contexts $()$. However, we need to construct them
-temporarily, because certain rules require splitting a context and, by our 
-definition, the context $(x:\tau)$ is not splitable.
-
-The typing rules of the system are shown in Figure~\ref{fig:structural-coeffect-typing}.
-Many of the structural rules are expressed in terms of a helper judgement
-$\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}}$.
-
-\paragraph{Standard rules.}
-Variable access (\rulest{var}) annotates the corresponding variable with an empty coeffect
-$0$. The \rulest{fun} rule assumes that the context of the body can be split into the
-variable of the function and other (potentially empty) context and it attaches the coeffect
-associated with the function variable to the resulting function type $\ctyp{s}{\tau_1} \rightarrow \tau_2$.
-
-The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
-expression, argument and the function type respectively. 
-The result of evaluating the argument in the context $t$ is passed to the function that
-requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
-the combination of coeffects $r \cvop t$. The variable context required to evaluate the 
-function value is independent and so it is annotated just with coeffects $s$.
-Finally, the \rulest{let} rule is derived from let binding and application, but we show
-it separately to aid the understanding.
-
-\paragraph{Structural rules.}
-The remaining rules are not syntax-directed. They are embedded using the \rulest{ctx}
-rule and expressed using a helper judgement $\ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}}$
-that says that the context on the left-hand side can be transformed to the context on the
-right-hand side. The transformation can be applied to any part of the context, which 
-is captured using the \rulest{nest} rule (it is sufficient to apply the transformation 
-on the left part of the context; the right part can be transformed using \rulest{exch}).
-
-The \rulest{empty} rule allows attaching empty context to any existing context. The rule 
-is needed to type-check lambda abstractions that do not capture any outer variable.
-The \rulest{weak} rule is similar, but it represents \emph{weakening} where an unused variable is 
-added. The associated coeffect tag does not associate context information with the variable.
-This is needed to type-check lambda abstraction that does not use the argument.
-
-The \rulest{contr} rule is a limited form of contraction. When a variable appears repeatedly,
-it can be reduced to a single occurrence. The rule is needed to satisfy the side condition
-of \rulest{fun} when the body uses the argument repeatedly. The two associativity rules
-together with \rulest{exch} provide ways to rearrange variables. Finally, \rulest{sub} represents
-sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Properties of reductions}
-\label{sec:structural-metatheory}
-Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
-operational meaning, because that will differ for every concrete application. For example,
-when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
-access to a specified element.
-
-Just like previously, we can state general properties of the reductions. As the syntax of expressions
-is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
-are also the same and can be found in Figure~\ref{fig:flat-substred}.
-
-The structural coeffect calculus \clstr~associates information with individual variables. This means
-that when an expression requires certain context, we know from what scope it comes -- the
-context must be provided by a scope that defines the associated variable, which is either a
-lambda abstraction or global scope. This distinguishes the structural system from the flat system
-where context could have been provided by any scope and the lambda rule allowed arbitrary 
-splitting of context requirements between the two scopes (or declaration and caller site).
-
-\paragraph{Internalized substitution.}
-Before looking at properties of the evaluation, we consider let binding, which can be 
-viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
-application and abstraction as follows.
-
-\begin{lemma}[Definition of let binding]
-If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
-then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
-\end{lemma}
-\begin{proof}
-The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
-correspond with the typing rule \rulest{let}:
-\begin{equation*}
-\inference[]
-  {\inference
-    {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
-    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
-   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
-  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
-\end{equation*}
-\end{proof}
-
-The term $e_2$ which is substituted in $e_1$ is checked in a different variable
-and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
-a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
-the variable that we are substituting for. The let binding captures substitution for a 
-specific variable (the context is of a form $\ctyp{r \cprd s}{\Gamma, v:\tau}$).
-For a general substitution, we need to define the notion of context with a hole.
-
-\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
-\paragraph{Substitution and holes.}
-In \clstr, the structure of the variable context is not a set, but a tree. When substituting
-for a variable, we need to replace the variable in the context with the context of the
-substituted expression. In general, this can occur anywhere in the tree. To formulate the
-statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
-in the free variable context and in a corresponding part of the coeffect tag:
-
-\begin{equation*}
-\begin{array}{rcl}
-\Delta[-] &::=& \ctyp{1}{()} \\
-       & \sep & \ctyp{r}{(x:\tau)}\\
-       & \sep & \ctyp{-}{(-)} \\
-       & \sep & \ctyp{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
-\end{array}
-\end{equation*}
-%
-Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
-operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
-$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
-Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
-general substitution lemma for \clstr.
-
-\begin{figure}
-
-\begin{equation*}
-\begin{array}{rcl}
-                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
-                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
-                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
-\chole{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \cprd r_2'}{(\Gamma_1', \Gamma_2')}\\
-&\multicolumn{2}{l} {
-  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
-}
-\end{array}
-\end{equation*}
-\caption{The definition of hole filling operation for $\Delta[-]$}
-\label{fig:structural-hole-fill}
-\vspace{-2em}
-\end{figure}
-
-\begin{lemma}[Substitution Lemma]
-\label{thm:structural-subst}
-If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
-then $\chole{r}{\Gamma}{R \cvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
-\end{lemma} 
-\begin{proof}
-Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
-$(S, \cvop, 0, \cprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
-\end{proof}
-
-\begin{theorem}[Subject reduction]
-\label{thm:structural-reduction}
-If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
-$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
-\end{theorem}
-\begin{proof}
-Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
-\end{proof}
-
-\paragraph{Local soundness and completeness.}
-As with the previous calculus, we want to guarantee that the introduction and
-elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
-This can be done by showing \emph{local soundness} and \emph{local completeness},
-which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
-case of subject reduction and the latter is proved by a simple derivation:
-%
-\begin{theorem}[Local soundness]
-If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
-$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
-\end{theorem}
-\begin{proof}
-Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
-\end{proof}
-%
-\begin{theorem}[Local completeness]
-If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
-$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
-\end{theorem}
-\begin{proof} The property is proved by the following typing derivation:
-%
-\begin{equation*}
-\inference
-  {\inference
-     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
-      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
-     {\ctyp{r \cprd (s \cvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\qedhere
-\end{equation*}
-\end{proof}
-%
-In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
-guarantees that $s\cvop0 = s$. Recall that in \clflt, the typing derivation for
-$\lambda x.f x$ required for local completeness was not the only possible derivation.
-In the last step, it was possible to split the coeffect tag arbitrarily between the 
-context and the function type.
-
-In the \clstr~calculus, this is not, in general, the case. The $\cprd$ operator is not required 
-to be associative and to have units and so a unique splitting may exist.  For example, if we
-define $\cprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
-$t$, there are unique $r$ and $s$ such that $t = r \cprd s$. However, if the $\cprd$ operation 
-has additional properties, then there may be other possible derivation.
-
-% ==================================================================================================
-
-\begin{figure*}[t]
-\newcommand{\uu}[1]{\overline{#1}}
-\begin{equation*}
-\cosem{\ctyp{r_1 \cprd \ldots \cprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-  \ctyp{r_1 \cprd \ldots \cprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
-\end{equation*}
-\begin{equation*}
-\begin{array}{rcll}
-\cosem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
-    \epsilon_0                                                                   &\quad$\ruless{var}$
-\\[0.32em]
-\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \Lambda (\cosem{\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
-\\[0.32em]
-\cosem{\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
-    (\lambda (\gamma_1, \gamma_2) \rightarrow
-    \cosem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
-    (\cobind{\cosem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
-\\[0.32em]
-\cosem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
-    \cosem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \cosem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
-\end{array}
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rclll}
-\cosem{\ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)}}}
-&=& \cmerge_{r, s} \circ (\cosem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
-\textnormal{\rulessc{nest}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}}
-&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
-\textnormal{\rulessc{exch}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \cprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 1}&&
-\textnormal{\rulessc{empty}}
-\\[0.8em]
-\cosem{\ctxtrans{ \ctyp{r\cprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 0}&&
-\textnormal{\rulessc{weak}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \cvop s}{(x:\tau)}}{
-    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}}
-&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
-\textnormal{\rulessc{contr}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
-    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
-&=& \cmerge_{r \cprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
-        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \cprd t}&&
-\textnormal{\rulessc{assoc}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
-&=&
-\iota_{r, s}
-&&
-\textnormal{\rulessc{sub}}
-\end{array}
-\end{equation*}
-
-\caption{Categorical semantics for~\clstr}
-\label{fig:structural-semantics}
-\vspace{-0.5em}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Semantics of structural coeffects}
-\label{sec:structural-semantics}
-
-The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
-\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
-of the variable context. Thus an expression $\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
-is modelled as a function $\ctyp{r \cprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
-
-As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
-system is not a simple finite product, but instead a binary tree. To model this, we do not use
-ordinary products in the domain of the semantic function, but instead use a special 
-constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
-the tag structure.
-
-\subsection{Structural tagged comonads}
-\label{sec:structural-tagged-comonads}
-
-To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
-Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
-$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
-$T^{r \cvop s} \tau_1 \rightarrow \tau_3$ and we use the $\cvop$ operation to combine the 
-context-requirements.
-
-However, functions that do not exist in context have only a single input variable (with a 
-single corresponding tag). To model complex variable contexts, we need two additional
-operations that allow manipulation with the variable context. Similarly to the model of \clflt,
-we also require operations that model duplication and sub-coeffecting:
-
-\begin{definition}[Structural tagged comonad]
-Given a structural coeffect tag structure $(S, \cprd, \cvop, 0, 1)$ a \emph{structural tagged comonad} is a 
-tagged comonad over $(S, \cvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
-$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
-$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \cprd s)}(A \ctimes B)\\
-\csplit_{r, s} &:& T^{(r \cprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
-\end{array}
-\end{equation*}
-%
-% Delta is still defineable
-% ?? \cdupl_{r, s}  &:& T^{(r \cvop s)}A \rightarrow T^r A \times T^s A\\
-%
-And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
-%
-%  TODO: laws?
-%
-\end{definition}
-
-The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
-and it can still be used to define a family of mappings that represents \emph{duplicating}
-of variables while splitting the additional coeffect tags:
-%
-\begin{equation*}
-\begin{array}{l}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
-\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
-\end{array}
-\end{equation*}
-%
-The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
-but with two differences. Firstly, it combines tags using $\cprd$ instead of $\vee$, 
-which corresponds to the fact that the variable context now consists of two parts (a tree node).
-Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
-(instead of $A \times B$ as previously).
-
-To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
-of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
-variables. The operation was not needed for \clflt, because there \emph{splitting} could
-be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
-the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
-with $r \cprd s$ that carries $A \ctimes B$.
-
-Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
-Before looking at them, we finish our discussion of categorical semantics.
-
-\paragraph{Categorical notes.}
-The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
-Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
-variables in the context. Given a context modelled as $T^{r \cprd s}(A \ctimes B)$,
-we can lift a morphism $f$ to $\hat{T}^{r \cprd s}(f)$, but we cannot manipulate the
-variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
-
-This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
-could apply $\cdupl_{r, s}$ on the input (if the tag $r \cprd s$ coincided
-with tag $r \cvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
-
-This restriction is intentional -- at the semantic level, it prevents manipulations
-with the context that would break the correspondence between tag structure and the
-product structure.
-
-\subsection{Categorical semantics}
-The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
-It uses the \emph{structural tagged comonad} structure introduced in the previous section,
-together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
-operations:
-\begin{equation*}
-\begin{array}{rcl}
-\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
-\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
-f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
-\end{array}
-\end{equation*}
-
-When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
-notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
-the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
-variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
-$\cprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
-type signature in $\cmerge_{r, s}$).
-
-The rule \ruless{app} now uses the operation $\csplit_{s, (r \cvop t)}$ instead of $\cdupl_{s, (r \cvop t)}$,
-which means that it splits the context instead of duplicating it. This makes the system more 
-structural -- the expressions use disjunctive parts of the context -- and also explains why the
-composed coeffect tag is $s \cprd (r \cvop t)$.
-
-The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
-of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
-of $\cosem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
-$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
-takes a context, converts it to a new context which is compatible with the original expression $e$.
-The context manipulation rules work as follows:
-
-\begin{itemize}
-\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
-  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
-  Finally, they re-construct a single context using $\cmerge_{r, s}$.
-
-\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
-  and discard one part (containing either an unused variable or an empty context).
-
-\item If we interpreted \rulessc{contr} by applying functor $T^{r\cvop s}$ to a function that duplicates
-  a variable, the resulting context would be $\ctyp{r \cvop s}(x:\tau, x:\tau)$, which would break the
-  correspondence between coeffect tag and context variable structure. However, that interpretation would be
-  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
-  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
-  a tag $r \cvop s$ into $r \cprd s$.
-
-\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
-  manipulates it (using \ident{assoc}) and then re-constructs it.
-
-\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
-  variable using the primitive natural transformation $\iota_{r, s}$.
-\end{itemize}
-
-\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
-attaching the context separately to individual variables. This an expression
-$\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
-$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
-approach largely complicates the definition of application (where tag of all 
-variables in a context is affected). Moreover, it makes it impossible to express
-\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.
-
-\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
-transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
-operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
-sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).
-
-This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
-typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
-the following natural transformation has to be provided:
-%
-\begin{equation*}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
-\end{equation*}
-%
-This variant of the system could be used to define a system that ensures that all provided 
-context is used and is not over-approximated. This difference is similar to the difference between
-affine type systems (where a variable can be used at most once) and linear type systems 
-(where a variable has to be used exactly once).
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Examples of structural coeffects}
-
-\subsection{Example: Neededness analysis}
-\label{sec:structural-neededness}
-
-\subsection{Example: Tracking tainting}
-\label{sec:structural-tainting}
+?
+\ExecuteMetaData[rules/rules.tex]{cml-types}
 

### 2013-06-19 text/Chapter01.tex
Merge changes

@@
 
-Some intro
+Modern software applications behave differently depending on the environment or context 
+in which they execute. They often run in increasingly rich environments that provide 
+resources (e.g. database or GPS sensor) and are gradually more diverse (e.g. multiple 
+versions of different mobile platforms). Web applications are split between client, server 
+and mobile components; mobile applications must be aware of the context and of the platform 
+while the ``internet of things'' makes the environments even more heterogeneous; 
+applications that access rich data sources need to propagate security policies and provenance 
+information about the data.
 
-\section{How lambda works}
-The key difference - how lambda works
+Writing such context-aware applications is a fundamental problem of modern software 
+engineering. The state of the art relies on ad-hoc approaches -- using hand-written conditions 
+or pre-processors for conditional compilation. Common problems that developers face include:
 
-Effect systems, introduced by Gifford and Lucassen
-\cite{effects-gifford}, track \emph{effects} of computations, such as
-memory access or message-based
-communication~\cite{effects-messagepassing}. Their approach augments
-typing judgments with effect information: $\Gamma \vdash e :
-\tau, F$.  Wadler and Thiemann explain how this shapes
-effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
+\begin{itemize}
+\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple 
+  targets (e.g. SQL, CUDA or JavaScript) the compilation often occurs at runtime and developers 
+  have no guarantee that it will succeed until the program is executed.
+
+\item \textbf{Platform versioning.} When developing application for multiple versions of a 
+  system (e.g. Android), developers rely on lazy loading at runtime or use conditional compilation 
+  using \ident{\#if}. The former delays errors to runtime, while the latter requires building all 
+  possible configurations to discover simple compile-time errors.
+
+\item \textbf{Resources \& data availability.} When programming applications that access 
+  resources or data provided by the environment (e.g. specific database table, GPS sensor), 
+  the program typically performs dynamic check for the resource availability. However, this 
+  is not checked in any way -- we have no easy way to tell what happens when the resource is 
+  not available (e.g. is there a fallback strategy or not?)
+
+\item \textbf{Meta-programming.} In distributed client-server programming, we often need to 
+  manipulate code using meta-programming techniques on the server and then pass a closed 
+  expression to client, translated to another language (JavaScript). We need to be able to 
+  guarantee that such expressions are well-formed before they are translated.
+\end{itemize}
+
+The issues discussed above pose an important problem to current and future state of computing.
+In this thesis, we develop a unifying mathematical model that allows us to capture the properties
+discussed above. We use the model in three different ways to capture different notions of 
+context-dependence.
+
+In the rest of the chapter, we first look at the distinguishing factor of two of the presented
+coeffect systems (how they differ from other existing work). Then we look at a number of concrete 
+problems that are discussed in greater details in later chapters. We use the examples to demonstrate 
+the real-world problems that motivate our work. (Theoretical and other motivations are discussed 
+later in Chapter~\ref{ch:related-work}).
+
+%===================================================================================================
+
+\section{Context and lambda abstraction}
+
+Our work on context-aware programming languages connects two directions in existing research on the 
+theory of programming languages. On one side, effect systems \cite{effects-gifford} and monadic 
+computations \cite{monad-notions,monads-effects-marriage} provide a detailed and established method 
+for tracking what effects programs have -- that is, how they affect the environment where they execute. 
+On the other side, the work on comonadic notions of computations \cite{comonads-notions} shows how 
+to use the mathematical dual of monads -- comonads -- to give categorical semantics of 
+context-dependent computations.
+
+Effect systems introduced track actions such as memory operations or communication. They are described 
+by typing judgments of a form $\Gamma \vdash e : \alpha, \sigma$ where $\Gamma$ is the context of a 
+program (typically available variables), $e$ is the expression (program) itself, $\alpha$ is the type of 
+values returned by the program (e.g. integer or boolean) and $\sigma$ is a set of possible effects. The 
+judgment states that, given the context $\Gamma$, an expression has a type $\alpha$ and can only perform 
+effects specified by the set $\sigma$. Wadler and Thiemann \cite{monads-effects-marriage} explain how 
+this shapes effect analysis of a lambda abstraction -- that is, how effect systems analyze the effects 
+associated with a definition of a function:
 %
@@
 %
-In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
-does not match this pattern. In the systems we consider, lambda abstraction places 
-requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
-(immediate requirements), resulting in different syntactic properties. 
-We informally discuss three examples first that demonstrate how contextual
-requirements propagate. Section ? then unifies these 
-in a single calculus.
-
-Also \cite{effects-producer-semantics}
-
+This means that, when a programmer defines a function, the system records that executing the 
+function will perform the effects of the body of the function. However, simply defining a function
+is an effect-free computation. Tate \cite{effects-producer-semantics} calls such effect systems
+\emph{producer} effect systems and generalizes the idea of function to more general ``thunking'':
+%
 \begin{quote}
@@
 \end{quote}
+%
+In contrast to the static analysis of (producer) effect systems, the analysis of 
+\emph{context-dependence} does not match this pattern. In the systems we consider, lambda 
+abstraction places requirements on both the \emph{call-site} (latent requirements) and the 
+\emph{declaration-site} (immediate requirements), resulting in different syntactic properties. 
+We informally discuss three examples first that demonstrate how contextual requirements propagate.
 
-Demonstrate using distributed computations
+%===================================================================================================
 
-Demonstrate using dataflow/liveness
+\section{Why coeffects matter}
 
-Perhaps mention LINQ as a motivation for cross-compilation..
-\cite{app-linq}
-
-\section{Associating with context}
-
--> Type providers make this important
+This section gives three examples of context-dependent computations whose properties can be captured
+by the tree calculi presented in this thesis. We look at an example of the \emph{flat coeffect calculus} 
+(Chapter~\ref{ch:flat-coeffects}), \emph{structural coeffect calculus} (Chapter~\ref{ch:structural-coeffects})
+and \emph{coeffect meta-language} (Chapter~\ref{ch:coeffect-metalanguage}).
 
@@
 
-\section{Flat coeffect system}
-~
+\subsection{Flat coeffect calculus}
 
-resources?
+The flat coeffect calculus associates a single piece of information with the context. As an example,
+we look at a simple distributed programming language that includes the concept of \emph{resources}.
+A resource may be accessed using a special construct $\kvd{access}~\ident{Res}$. The following 
+example shows a function taht lists recent events -- it accesses the resource \ident{News} 
+(representing a database) and a resource \ident{Clock} (with the current time):
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\
+\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\
+\quad\ident{query}~\ident{db}~\str{"SELECT * WHERE Date > \%1"}~(\kvd{access}~\ident{Clock})
+\end{array}
+\end{equation*}
+%
+Consider a scenario where the function is \emph{declared} on the server-side and then
+transferred and \emph{executed} on the client-side. The resource \ident{News} represents a 
+database that is only available on the server-side and so the function needs to keep a remote
+reference to the server. However, the \ident{Clock} resource may (or may not) be available on the
+client-side. If the resource is available on the client, then it may be re-bound and the function
+will use the current client's data -- for example, to accommodate for time-zone changes.
+
+This example demonstrates how lambda abstraction behaves for context-dependent computations. The 
+context requirement of the function body is a set of resources $\{\ident{Clock}, \ident{News}\}$.
+The context requirements are split between the \emph{declaration-site} and \emph{call-site}.
+However, there are multiple possible splittings. The splitting $\{\ident{News}\} \cup \{\ident{Clock}\}$
+models the case when the database is accessed from the server, but time is taken from the client,
+while the splitting $\{\ident{Clock}, \ident{News}\} \cup \{\}$ corresponds to the case when both
+resources are accessed from the server.
 
@@
 
-\section{Structural coeffect system}
-\label{sec:structural}
+\subsection{Structural coeffect calculus}
 
-The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
-we need to track context-dependence in a more fine-grained way. To track neededness or security, 
-we need to associate information with individual \emph{variables} of the context.
+\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
+\newcommand{\dnat}{\ident{stream}}
 
-At the same time, we want to avoid developing multiple variants of coeffect systems -- indeed, 
-our motivation is to develop a \emph{single unified mechanism} for tracking context-dependence.
-In this section, we present a more powerful \emph{structural coeffect calculus}, which is a 
-generalization of the flat calculus.
-
-%We show that \clflt~can be obtained as a special case of \clstr. The
-%structural version of our system is also where our work differs more significantly from well-known
-%effect systems.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Motivation: Tracking array accesses}
-\label{sec:structural-motivation}
-\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
-\newcommand{\dnat}{\ident{D}_\ident{nat}}
-\newcommand{\cprd}{\times}
-\newcommand{\cvop}{~\ident{max}~}
-
-Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
-functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
-respectively, but we use richer tag structure.
+The calculus used in the previous section annotates the entire context with a single value --  such
+as the set of required resources. However, sometimes it is desirable to annotate not the entire 
+context, but individual variables of the context.
 
@@
 some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
-To track information about individual variables, we use a product-like operation $\cprd$ on tags 
+To track information about individual variables, we use a product-like operation $\times$ on tags 
 to mirrors the product structure of variables. For example:
@@
 \begin{array}{l}
-\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
+(\ident{a}:\dnat, \ident{b}:\dnat) ~@~ 5 \times 10
   \vdash
@@
 %
-The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
-that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
-for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
-a single variable:
+The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
+that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
+\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
+maximum of the requirements of the individual variables:
 %
@@
 \begin{array}{l}
-\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
+(\ident{c}:\dnat) ~@~ \ident{max}(5, 10)
   \vdash
@@
 %
-In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
-Before looking at the formal definition, consider the typing of let bindings:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
-\pastval{a}{15}+\pastval{c}{10}
-\end{array}
-\end{equation*}
-%
-The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
-is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
-The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.
+This version of the calculus removes the non-determinism of the lambda abstraction from the previous
+version. As we associate information with individual variables, lambda abstraction creates a function
+that requires the context requirements associated with the variable that is being abstracted over.
 
-The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
-context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
-of $\cvop$ and it denotes empty coeffect. The typing context of the
-body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.
+For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
+\ident{b} from the declaration-site) then the function would have context requirements $5$ -- that
+is the number associated with the variable \ident{a}.
 
-To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
-of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
-with the remaining tags from the body yielding the overall context:
-$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
-normalization mechanism (described later), this can be further reduced to
-$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
-we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
-\ident{b}.
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+\todo{Give some example that uses the coeffects metalanguage style. This will probably be 
+meta-programming with open expressions (similarly to what Pfenning and Nanevski do with 
+contextual modal type theory).}
+
+%---------------------------------------------------------------------------------------------------
+
+\section{Why context matters}
+
+We claimed earlier that context-dependent comptuations are becoming increasingly common and our
+work is focused on annotating the (free-variable) context with additional information.
+The importance of context can be also demonstrated by looking at a technology that focuses 
+solely on making the context richer -- F\# type providers.
+
+\todo{Say more about type providers -- they extend the context $\Gamma$ so that it can be 
+lazily loaded and it can be huge. Potentially, it could be also annotated with additional
+meta-data...}
 
@@
 \section{Contributions}
+
+\todo{We present three calculi that model common notions of context-dependence and can be used
+as basis for developing context-aware programming langauges with static type systems.}

### 2013-06-19 text/Chapter03.tex
Refactor rules
Merge changes

@@
-\chapter{Flat coeffects} 
+\chapter{Flat coeffect language} 
 \label{ch:flat-coeffects}
@@
 
-TODO: Change this
+\todo{Update the introduction \& edit the whole chapter -- currently, this is just a direct copy
+from the accepted ICALP paper on flat coeffects.}
+
+\todo{Other extensions -- add more examples of indexed comonoads, consider recursion and 
+fixed-point operator, possibly also consider partially defined operations, which lets us
+express type systems that rule out some computations (rather than just static analyses). }
 
@@
 
+\newcommand{\cprd}{\times}
+\newcommand{\cvop}{\oplus}
+
 \vspace{-1em}

### 2013-06-19 text/Chapter04.tex
Merge changes

@@
 \chapter{Structural coeffect language} 
-\label{ch:structural}
+\label{ch:structural-coeffects}
 
@@
 
-Introduction
+The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
+we need to track context-dependence in a more fine-grained way. To track neededness or security, 
+we need to associate information with individual \emph{variables} of the context.
 
@@
 \section{Introduction}
-\label{sec:structural-intro}
-
-We already talked about tainting in pathways to coeffects... here is more stuff!
-
-~
-
-~
-
-Our second motivating example is the use of \emph{tainting} for the prevention of SQL injection 
-attacks \cite{app-secure-tainting}. The idea is to (statically) check that SQL commands cannot be 
-directly constructed from, potentially dangerous, inputs provided by the user. Consider the type
-checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
-\ident{msg}~+~\ident{name}
-\end{array}
-\end{equation*}
-%
-We say that a variable is \emph{tainted} if it may come from untrusted input. In the above example,
-\ident{id} must not be tainted. Otherwise, the attacker could specify values such as 
-\str{"1; DROP TABLE Users"}. The variable \ident{msg} may or may not be tainted, because it is not
-used in protected context (i.e.~to construct an SQL query). Using the coeffect calculus, security
-information can be naturally attached to the context. We use tag \ident T for tainted and 
-\ident N for not tainted:
-%
-\begin{equation*}
-\ctyp{\ident T \times \ident N}{(\ident{id}:\ident{string}, \ident{msg}:\ident{string})} 
-  \vdash (\ldots) : \ident{string}
-\end{equation*}
-%
-The tags \ident T and \ident N correspond to variables \ident{id} and \ident{msg}, respectively.
-The checking of tainting is a special case of checking of \emph{non-interference} in 
-\emph{secure information flow} \cite{app-secure-information-flow}. Languages such as FlowCaml 
-\cite{app-security-flowcaml} can check such properties statically, but it is also possible to 
-embed such checking in a monadic library \cite{monad-secure-flow}.
-
-Assuming the monadic type $\mtyp{s}{\tau}$ represents a value of type $\tau$ 
-in a security context $s$ (either \ident T or \ident N): 
-%
-\begin{equation*}
-\ident{id} : \mtyp{\ident T}{\ident{string}}, \ident{msg} : \mtyp{\ident N}{\ident{string}} 
-  \vdash (\ldots) : \mtyp{\ident T}{\ident{string}}
-\end{equation*}
-%
-In this example, the result is tagged with \ident T to denote that it may be tainted.
-If the result was tagged with \ident N, the first parameter would also have to be \ident N
-(and we would get fewer information about the function).
-
-The monadic encoding has an unfortunate property from the semantic perspective. 
-The function has monadic structure over the \emph{domain} as well as over the \emph{codomain}
-and thus it does not match the desirable Kleisli model of expressions $\tau_1 \times \ldots \tau_n \rightarrow \mtyp{}{\tau}$. 
-This makes the it difficult to give standard composable semantics of the library.
-
-In coeffect calculus, expressions are modelled using the coKleisli form 
-$\ctyp{r_1 \times \ldots \times r_n}{(\tau_1 \times \ldots \tau_n)} \rightarrow \tau$ with a tags 
-$r_n$ that specify taintedness of inputs. As discussed in detail in Section~\ref{sec:structural-tainting}, 
-this model gives a simple comonadic semantics of language-based security using  tainting.
-
-%---------------------------------------------------------------------------------------------------
-%===================================================================================================
-
-% ==================================================================================================
-
-\begin{figure*}[t]
-% Constants, if we wanted them, would look like this
-%
-% \inference[]
-%   {c : \tau \in \Phi}
-%   {\ctyp{1}{()} \vdash c : \tau}&
-% \textnormal{\rulest{const}}
-% \\[1.6em]
-
-%\paragraph{Typing rules}
-\begin{equation*}
-\boxed{\ctyp{r}{\Gamma} \vdash e : \tau }
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{cl}
-
-\inference[]
-  {}
-  {\ctyp{0}{(v:\tau)} \vdash v : \tau }&
-\textnormal{\rulest{var}}
-\\[1.8em]
-\inference
-  {\ctyp{r \csprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
-  {\ctyp{r}{\Gamma} \vdash \lambda v.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }&
-\textnormal{\rulest{fun}}
-\\[1.8em]
-\inference
-  {\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1 }
-  {\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
-\textnormal{\rulest{app}}
-\\[1.8em]
-\inference
-  {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
-  \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1}
-  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
-\textnormal{\rulest{let}}
-\\[1.8em]
-\inference
-  {\ctyp{r}{\Gamma} \vdash e : \tau &
-   \ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}
-  {\ctyp{r'}{\Gamma'} \vdash e : \tau}&
-\textnormal{\rulest{ctx}}
-\end{array}
-\end{equation*}
-
-%
-% Empty variable context corresponds to unit, but things get messy
-%
-
-%\paragraph{Context rules}
-\begin{equation*}
-\boxed{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}} }
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rcll}
-\multicolumn{3}{c}{
-  \inference
-    {  \ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}} }
-    {  \ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)}} }
-}&
-\textnormal{\rulest{nest}}
-\\[2.0em]
-  \ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}&
-\textnormal{\rulest{exch}}
-\\[0.8em]
-  \ctxtrans{ \ctyp{r \csprd 1}{(\Gamma, ())} &}{&
-    \ctyp{r}{\Gamma}}&
-\textnormal{\rulest{empty}}
-\\[0.8em]
-  \ctxtrans{\ctyp{r\csprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
-\textnormal{\rulest{weak}}
-\\[0.8em]
-  \ctxtrans{\ctyp{r \csvop s}{(x:\tau)} &}{&
-    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}&
-\textnormal{\rulest{contr}}
-\\[0.8em]
-  \ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
-    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
-\textnormal{\rulest{assoc}}
-\\[0.8em]
-\multicolumn{3}{c}{
-  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \csvop r = s)
-}&
-\textnormal{\rulest{sub}}
-\end{array}
-\end{equation*}
-
-% We do not need this rule (can be expressed in terms of the other one)
-%
-% \inference
-%  {  \ctxtrans{\ctyp{s}{\Gamma_2}}{\ctyp{s'}{\Gamma_2'}} }
-%  {  \ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\csprd s'}{(\Gamma_1, \Gamma_2')}} }&
-% \textnormal{\rulest{right}}
-
-\caption{Type system for \clstr}
-\label{fig:structural-coeffect-typing}
-\end{figure*}
-
-\newpage
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Structural coeffect system}
-\label{sec:structural}
-
-The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
-we need to track context-dependence in a more fine-grained way. To track neededness or security, 
-we need to associate information with individual \emph{variables} of the context.
-
-At the same time, we want to avoid developing multiple variants of coeffect systems -- indeed, 
-our motivation is to develop a \emph{single unified mechanism} for tracking context-dependence.
-In this section, we present a more powerful \emph{structural coeffect calculus} \clstr, which is a 
-generalization of \clflt. We show that \clflt~can be obtained as a special case of \clstr. The
-structural version of our system is also where our work differs more significantly from well-known
-effect systems.
 
@@
 sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.
+
+\ExecuteMetaData[rules/rules.tex]{scl-types}
 
@@
 
-\subsection{Example: Neededness analysis}
+\subsection{Example: Liveness analysis}
 \label{sec:structural-neededness}
 
-\subsection{Example: Tracking tainting}
+\subsection{Example: Data-flow (revisited) }
 \label{sec:structural-tainting}
 
-\section{Summary}
-Oops!
+\todo{Also, consider additional language features that we consider for flat coeffects
+  (mainly recursion and possibly conditionals)}
+
+\section{Conclusions}
+
+\todo{ (...) }

### 2013-06-19 text/Chapter05.tex
Merge changes

@@
 \chapter{Coeffect meta-language} 
-\label{ch:metalanguage}
+\label{ch:coeffect-metalanguage}
 
@@
 
-Introduction
+Both flat coeffect calculus and structural coeffect calculus (presented in the past two chapters)
+use indexed comonads to define the semantics of the langauge. In this section, we follow the 
+meta-language style and embed indexed comonads into the language -- the type constructor
+$\ctyp{r}{\alpha}$ becomes a first-class value and we add language constructs corresponding to
+primitive operations of the indexed comonad.
 
@@
 
-?
+\section{Type system}
 \ExecuteMetaData[rules/rules.tex]{cml-types}
 
+\section{Operational properties}
+
+\section{Categorical semantics}
+
+\section{Applications}
+
+\subsection{Meta-programming}
+
+\subsection{Mobile computations}
+
+\section{Related work}
+This chapter is closely related to Contextual Modal Type Theory (CMTT) of Nanevski et al. However
+they develop their language using model logic as a basis, while we use categorical foundations 
+as the basis - leading to a different system.
+
 \section{Summary}
-Oops!

### 2014-02-10 text/Abstract.tex
Merge changes
Adding popular introduction, title and new abstract

@@
 \chapter*{Abstract} % Abstract name
+The development of programming languages needs to reflect important changes in the industry.
+In recent years, this included \eg~the development of parallel programming models (in reaction 
+to the multi-core revolution). This thesis is a response to another such revolution -- the 
+diversification of devices and systems where programs run.
 
-Static type systems are the most common form of program verification. In their most basic form, type
-systems guarantee that a computation always returns a value of some type. The work on monads
-and effect systems extends type systems to guarantee properties about \emph{effectful 
-computations} that perform side-effects like non-termination, I/O operations or exceptions.
+We develop the foundations for statically typed functional languages that understand 
+the \emph{context} or environment in which programs execute. Such context includes different 
+platforms (and their versions) in cross-platform applications, resources available in different 
+execution environments (\eg \; GPS sensor on a phone and database on the server), but also more 
+traditional notions such as variable context (tracking variable usage in static analyses) or 
+past values in stream-based data-flow programming.
 
-In this thesis we extend type systems to guarantee properties about \emph{context-dependent
-computations}. Such computations are increasingly common in modern software. For example, applications
-accessing data need to guarantee that the right data is available and distributed systems need to 
-guarantee that resources are available on specific nodes.
+The thesis presents three \emph{coeffect} calculi that capture different notions of 
+context-awareness: \emph{flat} calculus capturing contextual properties of the execution 
+environment, \emph{structural} calculus capturing contextual properties related to 
+variable usage and \emph{meta-language} calculus which allows reasoning about multiple notions
+of context-dependence in a single language and provides pathway to embedding the other two
+calculi in existing languages.
 
-We develop three \emph{coeffect} type systems that are capable of capturing different context-dependent
-properties. All three systems are based on the unifying semantic structure called \emph{indexed
-comonads}, but they use it differently.
-
-Our \emph{flat coeffect calculus} tracks one annotation about the entire context. It can be used to
-keep a set of required resources or a set of supported platforms. Our \emph{structural coeffect 
-calculus} tracks one annotation for each variable. It can be used to track liveness of variables
-or number of past values needed in data-flow computations. Finally, our \emph{coeffect meta-language}
-embeds contextual values directly into a language. It can be used to guarantee security properties
-of computations by using tainting.
+Although the focus of this thesis is on the syntactical properties of the presented 
+systems, we also discuss their category-theoretical motivation. We introduce the notion of
+an \emph{indexed} comonad (the dual of monads) and show how they provide semantics of the
+presented three calculi.
 

### 2014-02-10 text/Chapter00.tex
Adding popular introduction, title and new abstract

\chapter{Why context-aware programming matters} 
\label{ch:introduction1}

\todo{This chapter is currently copy \& paste of a \texttt{}popular'' introduction to coeffects
that I wrote online (\url{http://tomasp.net/blog/2014/why-coeffects-matter}). It needs to be
rewritten in academic style - but hopefully it has the right structure for an intro chapter? }

% ==================================================================================================

Many advances in programming language design are driven by some practical motivations.
Sometimes, the practical motivations are easy to see - for example, when they come from
some external change such as the rise of multi-core processors. Sometimes, discovering 
the practical motivations is a tricky task - perhaps because everyone is used to a 
certain way of doing things that we do not even \emph{see} how poor our current solution is.

The following two examples are related to the work done in F\# (because this is what
I'm the most familiar with), but you can surely find similar examples in other languages:

\begin{itemize}
\item \textbf{Multi-core} is an easy to see challenge caused by an external development. 
   It led to the popularity of \emph{immutable} data structures (and functional programming,
   in general) and it was also partly motivation for asynchronous workflows \cite{xx}.

\item \textbf{Data access} is a more subtle challenge. Technologies like LINQ \cite{xx} make it
   significantly easier, but it was not easy to see that inline SQL was a poor solution.
   This is even more the case for F\# \emph{type providers}. You will not realize how poor the
   established data access story is until you \emph{see} something like 
   the WorldBank and R provider \cite{xx} or CSV type provider \cite{xx}.
\end{itemize}

I believe that the next important practical challenge for programming language designers
is of the kind that is not easy to see -- because we are so used to doing things in 
certain ways that we cannot see how poor they are. The problem is designing languages
that are better at working with (and understanding) the \emph{context} in which programs are
executed.

% ==================================================================================================

\section{Context-aware programming matters}

The phrase \emph{context in which programs are executed} sounds quite abstract and generic.
What are some concrete examples of such \emph{context}? For example:

\begin{itemize}
\item When writing a cross-platform application, different platforms (and even different 
   versions of the same platform) provide different contexts - the API functions that are 
   available.

\item When creating a mobile app, the different capabilities that you may (or may not) have
   access to are context (GPS sensor, accelerometer, battery status).

\item When working with data (be it sensitive database or social network data from Facebook),
   you have permissions to access only some of the data (depending on your identity) and
   you may want to track \emph{provenance} information. This is another example of a context.
\end{itemize}

These are all fairly standard problems that developers deal with today. As the number of 
devices where programs need to run increases, dealing with diverse contexts will be becoming
more and more important (and I'm not even talking about \emph{ubiquitous computing} where you need
to compile your code to a coffee machine).

We do not preceive the above things as problems (at best, annoyances that we just have to 
deal with), because we do not realize that there should be a better way. Let me digg into
four examples in a bit more detail.

% --------------------------------------------------------------------------------------------------

\subsection{Context awareness \#1: Platform versioning}

The number of different platform versions that you need to target is increasing, no matter what 
platform are you using. For Android, there is a number called API Level 
\footnote{\url{http://developer.android.com/guide/topics/manifest/uses-sdk-element.html}}
with mostly additive changes (\emph{mostly} sounds very reassuring). In the .NET world, there are multiple
versions, mobile editions, multiple versions of Silverlight etc. So, code that targets multiple
frameworks easily ends up looking as the following sample from the Http module in F\# Data\footnote{
\url{https://github.com/fsharp/FSharp.Data/blob/master/src/Net/Http.fs}}:

\begin{verbatim}
for header, value in headers do
  match header with
  | StringEquals "accept" -> req.Accept <- value
#if FX_NO_WEBREQUEST_USERAGENT
  | StringEquals "user-agent" -> 
      req.Headers.[HttpRequestHeader.UserAgent] <- value
#else
  | StringEquals "user-agent" -> req.UserAgent <- value
#endif
#if FX_NO_WEBREQUEST_REFERER
  | StringEquals "referer" -> 
      req.Headers.[HttpRequestHeader.Referer] <- value
#else
  | StringEquals "referer" -> req.Referer <- value
#endif
  | _ -> req.Headers.[header] <- value
\end{verbatim}

This is difficult to write (you won't know if the code even compiles until you try building all 
combinations) and maintaining such code is not particularly nice (try adding another platform version).
Maybe we could refactor the code in F\# Data to improve it, but that's not really my point - supporting
multiple platforms should be a lot easier.

On Android, you can access API from higher level platform dynamically using techniques like 
reflection and writing wrappers 
\footnote{\url{http://android-developers.blogspot.cz/2009/04/backward-compatibility-for-android.html}}
Again, this sounds very error prone - to quote the linked article 
\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}.
Testing is no doubt important. But at least in statically typed langauges, we
should not need to worry about calling a method that does not exist when writing
multi-platform applications!

% --------------------------------------------------------------------------------------------------

\subsubsection{Context awareness \#2: System capabilities}

Another example related to the previous one is when you're using something like LINQ or 
FunScript (\url{http://funscript.info/}) to translate code written in a sub-set of C\# or F\#
to some other language - such as SQL or JavaScript. This is another important area, because
you can use this technique to target JavaScript, GPUs, SQL, Erlang runtime or other components 
that use a particular programming language.

For example, take a look at the following LINQ query that selects product names where the
first upper case letter is "C":

\begin{verbatim}
var db = new NorthwindDataContext();

from p in db.Products
where p.ProductName.First(c => Char.IsUpper(c)) == "C"
select p.ProductName;
\end{verbatim}

This looks like a perfectly fine piece of code and it compiles fine. But when you try running
it, you get the following error:

\begin{quote}
Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators 
not supported for type \texttt{System.String}.
\end{quote}

The problem is that LINQ can only translate a \emph{subset} of normal C\# code. Here, we are using
the \texttt{First} method to iterate over characters of a string and that's not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach. If we target a 
limited language, we simply cannot support the full source language. Is this an important 
problem? If you search on Google for ``has no supported translation to SQL'' 
(which is a part of a similar error message that you get in another case), you get some 26900
links. So yes - this is an issue that people are hitting all the time.

% --------------------------------------------------------------------------------------------------

\subsubsection{Context awareness \#3: Confidentiality and provenance}

The previous two examples were mainly related to the (non-)existence of some API functions
or to their behaviour. However, this is not the only case of context-awarness that is important
in every day programming.

Most readers of this blog will immediately see what is wrong with the following code, but that
does not change the fact that it can be compiled and deployed (and that there is a large number
of systems that contain something similar):

\begin{verbatim}
let query = sprintf "SELECT * FROM Products WHERE ProductName = '%s'" name
use cmd = new SqlCommand(query)
use reader = cmd.ExecuteReader()
\end{verbatim}

The problem is obviously SQL Injenction \footnote{\url{http://en.wikipedia.org/wiki/SQL_injection}}. 
Concatenating strings to build an SQL command is a bad practice, but it is an example of a more general problem.

Sometimes, we have special kinds of variables that should have some contextual meta-data associated
with them. Such meta-data can dictate how the variable can be used. Here, \texttt{name} comes from the user
input and this \emph{provenance} information should propagate to \texttt{query} and we should get an error when
we try passing this - potentially unsafe - input to \texttt{SqlCommand}. Similarly, if you have \texttt{password} or
\texttt{creditCard}, it should be annotated with meta-data saying that this is a sensitive piece of data and
should not, for example, be sent over an unsecured network connection.
As a side note - this idea is related to a technique called taint checking \cite{xx}.

In another context, if you are working with data (e.g. in some data jouranlism application), it would be
great if your sources were annotated with meta-data about the quality and source of the data (e.g.
can it be trusted? how up-to-date is it?) The meta-data could propagate to the result and tell you how
accurate and trustworthy your result is.

% --------------------------------------------------------------------------------------------------

\subsubsection{Context-awareness \#4: Resource \& data availability}

A vast majority of applications accesses some data sources (like database) or resources
(like GPS sensor on a phone). This is more tricky for client/server applications where a part
of program runs on the server-side and another part runs on the client-side. I believe that 
these two parts should be written as a single program that is cross-compiled to two parts 
(and I tried to make that possible with F\# Web Tools \cite{xx})
a long time ago; more recently [WebSharper](http://websharper.com/) implemented a similar idea).

So, say we have a function \texttt{validateInput}, \texttt{readData} and \texttt{displayMessage} in my prorgam.
I want to look at their types and see what resources (or \emph{context}) they require. For example,
\texttt{readData} requires \emph{database} (or perhaps a database with a specific name), \texttt{displayMessage} 
requires access to \emph{user interface} and \texttt{validateInput} has no special requirements.

This means that I can call \texttt{validateInput} from both server-side and client-side code - it is
safe to share this piece of code, because it has no special requirements. However, when I write
a code that calls all three functions without any remote calls, it will only run on a thick
client that has access to a database as well as user interface.

I'll demonstrate this idea with a sample (pseudo-)code in a later section, so do not worry if
it sounds a bit abstract at first.

% ==================================================================================================

\section{Coeffects: Towards context-aware languages}

The above examples cover a couple of different scenarios, but they share a common theme - 
they all talk about some \emph{context} in which an expression is evaluated. The context has 
essentially two aspects:

\begin{itemize}
\item \textbf{Flat context} represents additional data, resources and meta-data that are 
   available in the execution environment (regardless of where in the program you 
   access them). Examples include resources like GPS sensors or databases, battery status,
   framework version and similar.

\item \textbf{Structural context} contains additional meta-data related to variables. This can include
   provenance (source of the variable value), usage information (how often is the value
   accessed) or security information (does it contain sensitive data). 
\end{itemize}

As a proponent of statically typed functional languages I believe that a context-aware 
programming language should capture such context information in the type system and make
sure that basic errors (like the ones demonstrated in the four examples above) are ruled
out at compile time.

This is essentially the idea behind \emph{coeffects}. Let's look at an example showing the
idea in (a very simplified) practice and then I'll say a few words about the theory
(which is the main topic of this thesis).

% --------------------------------------------------------------------------------------------------

\subsection{Case study: Coeffects in action}

So, how should a context-aware language look? Surely, there is a wide range of options, but I hope
I convinced you that it needs to be \emph{context-aware} in some way! I'll write my pseudo-example in a
language that looks like F\#, is fully statically typed and uses type inference.

I think that type inference is particularly important here - we want to check quite a few properties
that should not that difficult to infer (If I call a function that needs GPS, I need to have GPS access!)
Writing all such information by hand would be very cumbersome.

So, let's say that we want to write a client/server news reader where the news are stored in a 
database on a server. When a client (iPhone or Windows phone) runs, we get GPS location from the
phone and query the server that needs to connect to the "News" database using a password defined
somewhere earlier (perhaps loaded from a server-side config file):

\begin{verbatim}
let lookupNews(location) =
  let db = query("News", password)
  selectNews(db, location)

let readNews() =
  let loc = gpsLocation()       
  remote { 
    lookupNews(loc) 
  }

let iPhoneMain() =
  createCocoaWidget(readNews)

let windowsMain() =    
  createMetroWidget(readNews)
\end{verbatim}

The idea is that \texttt{lookupNews} is a server-side function that queries the "News" database based on 
the specified \texttt{location}. This is called from the client-side by \texttt{readNews} which get the current GPS
position and uses a \texttt{remote \{ .. \}} block to invoke the \texttt{lookupNews} function remotely (how exactly would
this be written is a separate question - but imagine a simple REST request here).

Then, we have two main functions, \texttt{iPhoneMain} and \texttt{windowsMain} that will serve as two entry points
for iPhone and Windows build of the client-side application. They are both using a corresponding
platform-specific function to build the user interface, which takes a function for reading news as an
argument.

If you wanted to write and compile something like this today, you could use F\# in Xamarin 
Studio to target iPhone and Window phone, but you'd either need two separate end-application projects 
or a large number of unmaintainable \texttt{\#if} constructs. Why not just use a single project, if 
the application is fairly simple?

I imagine that a context-aware statically typed language would let you write the above code and 
if you inspected the types of the functions, you would see something like this:

\begin{verbatim}
password      :   string { sensitive }
lookupNews    :   Location -{ database }-> list<News>

gpsLocation   :   unit -{ gps }-> Location 
readNews      :   unit -{ rpc, gps }-> Async<list<News>>

iPhoneMain    :   unit -{ cocoa, gps, rpc }-> unit 
windowsMain   :   unit -{ metro, gps, rpc }-> unit 
\end{verbatim}

The syntax is something that I just made up for the purpose of this article - it could
look different. Some information could even be mapped to other visual representations
(e.g. blueish background for the function body in your editor). The key thing is that
we can learn quite a lot about the context usage:

 - \texttt{password} is available in the context, but is sensitive and so we cannot return it
   as a result from a function that is called via an RPC call.
 - \texttt{lookupNews} requires database access and so it can only run on the server-side
   or on a thick client with local copy of the database.
 - \texttt{gpsLocation} accesses GPS and since we call it in \texttt{readNews}, this function
   also requires GPS (the requirement is propagated automatically).
 - We can compile the program for two client-side platforms - the entry points require
   GPS, the ability to make RPC calls and Cocoa or Metro UI platform, respectively.

When writing the application, I want to be always able to see this information (perhaps
similarly to how you can see type information in the various F\# editors). I want to be
able to reference multiple versions of base libraries - one for iPhone and another for
Windows and see all the API functions at the same time, with appropriate annotations.
When a function is available on both platforms, I want to be able to reuse the code that
calls it. When some function is available on only one platform, I want to solve this by
designing my own abstraction, rather than resorting to ugly \texttt{\#if} pragmas.

Then, I want to take this single program (again, strutured using whatever abstractions
I find appropriate) and compile it. As a result, I want to get a component (containing
\texttt{lookupNews}) that I can deploy to the server-side and two packages for iPhone and 
Windows respectively, that reference only one or the other platform.

% --------------------------------------------------------------------------------------------------

\subsection{Coeffects: Theory of context dependence}

If you're expecting a "Download!" button or (even better) a "Buy now!" button at the end of this article,
then I'll disappoint you. I have no implementation that would let you do all of this.
My work in this area has been (so far) on the theoretical side. This is a great way to 
understand what is \emph{actually} going on and what does the \emph{context} mean. And if you made
it this far, then it probably worked, because I understood the problem well enough to be 
write a readable article about it!

\subsubsection{Brief introduction to type systems}

I won't try to reproduce the entire content of the thesis in this introduction -- but I will
try to give you some background in case you are interested (that should make it easier to
look at the papers above). We'll start from the basics, so readers familiar with theory of 
programming languages can skip to the next section.

Type systems are usually described in the form of \emph{typing judgement} that have the following form:

\begin{equation}
\Gamma \vdash e : \tau
\end{equation}

The judgement means that, given some variables described by $\Gamma$, the expression or program $e$
has a type $\tau$. What does this mean? For example, what is a type of the expression \texttt{x + y}?
Well, this depends - in F\# it could be some numeric type or even a string, depending on the types
of \texttt{x} and \texttt{y}. That's why we need the variable context $\Gamma$ which specifies the types of variables.
So, for example, we can have:

\begin{equation}
x:int, y:int \vdash x+y : int
\end{equation}

Here, we assume that the types of \texttt{x} and \texttt{y} (on the left hand side) are both \texttt{int} and as a result,
we derive that the type of \texttt{x + y} is also an \texttt{int}. This is a valid typing, for the expression, but
not the only one possible - if \texttt{x} and \texttt{y} were of type \texttt{string}, then the result would also be 
 \texttt{string}.

\subsubsection{Checking what program does with effect systems}

Type systems can be extended in various interesting ways. Essentially, they give us an approximation
of the possible values that we can get as a result. For example refinement types \cite{xx}
can estimate numerical values more precisely (e.g. less than 100). However, it is also possible to 
track what a program does - how it \emph{affects} the world. For example, let's look at the following
expression that prints a number:

\begin{equation}
x:int \vdash print~x : unit
\end{equation}

This is a reasonable typing in F\# (and ML languages), but it ingores the important fact that the 
expression has a \emph{side-effect} and prints the number to the console. In Haskell, this would not be
a valid typing, because \texttt{print} would return an \texttt{IO} computation rather than just plain \texttt{unit} (for
more information see IO in Haskell \footnote{\url{http://www.haskell.org/haskellwiki/IO_inside}}.

However, monads are not the only way to be more precise about side-effects. Another option is
to use effect system \cite{xx} which essentially annotates the 
result of the typing judgement with more information about the \emph{effects} that occur as part of
evaluation of the expression:

\begin{equation}
x:int \vdash print~x : unit \& \{ IO \}
\end{equation}

The effect annotation is now part of the type - so, the expression has a type \texttt{unit \& \{ io \}} meaning
that it does not return anything useful, but it performs some I/O operation. Note that we do not track
what \emph{exactly} it does - just some useful over-approximation. How do we infer the information?
The compiler needs to know about certain language primitives (or basic library functions). Here, 
\texttt{print} is a function that performs I/O operation.

The main role of the type system is dealing with composition - so, if we have a function \texttt{read} that
reads from the console (I/O operation) and a function \texttt{send} that sends data over network, the type
system will tell us that the type and effects of \texttt{send (read ())} are \texttt{unit \& {io, network}}.

Effect systems are a fairly established idea - and they are a nice way to add better purity checking
to ML-like languages like F\#. However, they are not that widely adopted (interestingly, checked
exceptions in Java are probably the most major use of effect system). However, effect systems are
also a good example of general approach that we can use for tracking contextual information...

\subsubsection{Checking what program requires with coeffect systems}

How could we use the same idea of \emph{annotating} the types to capture information about the context?
Let's look at a part of the program from the case study that I described earlier:

\begin{equation}
pass : string \vdash query("News", pass) : NewsDb
\end{equation}

The expression queries a database and gets back a value of the \texttt{NewsDb} type (for now, let's say
that \texttt{"News"} is a constant string and \texttt{query} behaves like the SQL type provider in F\# \footnote{\url{http://www.pinksquirrellabs.com/post/2013/12/09/The-Erasing-SQL-type-provider.aspx}}
and generates the \texttt{NewsDb} type automatically).

What information do we want to capture? First of all, we want to add an annotation saying that
the expression requires \emph{database access}. Secondly, we want to mark the \texttt{pass} variable as 
\emph{secure} to guarantee that it will not be sent over an unsecured network connection etc.
The \emph{coeffect typing judgement} representing this information looks like this:

\begin{equation}
pass : string^{\{secure\}} @ \{database\} \vdash query("News", pass) : NewsDb
\end{equation}

Rather than attaching the annotations to the \emph{resulting type}, they are attached to the 
variable \emph{context}. In other words, the equation is saying -- given a variable \texttt{pass} that is
marked as secure and additional environment providing database access, the expression
\texttt{query("News", pass)} is well typed and returns a \texttt{NewsDb} value.

As a side-note, it is well known that \emph{effects} correspond to \emph{monads} (and Haskell uses
monads as a way of implementing limited effect checking). Quite interestingly, \emph{coeffects}
correspond to the dual concept called \emph{comonads} and, with a syntactic extension akin to 
the \texttt{do} notation or \emph{computation expressions} \cite{xx}, you could capture contextual 
propreties by adding comonads to a language.

% ==================================================================================================

\section{Summary}

I started by explaining the
motivation for my work - different problems that arise when we are writing programs that are
aware of the context in which they run. The context includes things such as execution environment
(databases, resources, available devices), platform and framework (different versions, different
platforms) and meta-data about data we access (sensitivity, security, provenance).

This may not be preceived as a major problem - we are all used to write code that deals with
such things. However, I believe that the area of \emph{context-aware} programming is a source of
many problems and pains - and programming languages can help!

In the second half of the article, I gave a brief introduction to \emph{coeffects} -- a programming
language theory that can simplify dealing with context.
The key idea is that we can use types to track and check additional information about
the \emph{context}. By propagating such information throughout the program (using type system
that is aware of the annotations), we can make sure that none of the errors that I used
as a motivation for this article happen.

### 2014-03-14 text/Chapter00.tex
Editing introduction chapter

@@
 
-\todo{This chapter is currently copy \& paste of a \texttt{}popular'' introduction to coeffects
-that I wrote online (\url{http://tomasp.net/blog/2014/why-coeffects-matter}). It needs to be
-rewritten in academic style - but hopefully it has the right structure for an intro chapter? }
+Many advances in programming language design are driven by some practical motivations. Sometimes, 
+the practical motivations are easy to see -- for example, when they come from an external change 
+such as the rise of multi-core processors. Sometimes, discovering the practical motivations is a 
+difficult task -- perhaps because we are so used to a certain way of doing things that we do not 
+even \emph{see} the flaws of our approach.
 
@@
 
-Many advances in programming language design are driven by some practical motivations.
-Sometimes, the practical motivations are easy to see - for example, when they come from
-some external change such as the rise of multi-core processors. Sometimes, discovering 
-the practical motivations is a tricky task - perhaps because everyone is used to a 
-certain way of doing things that we do not even \emph{see} how poor our current solution is.
+\section{Programming language and innovation}
 
-The following two examples are related to the work done in F\# (because this is what
-I'm the most familiar with), but you can surely find similar examples in other languages:
+We briefly consider two practical concerns that led to the development of new programming languages. 
+This helps to explain why context-aware programming is of importance, which we cover in the next section. 
+The examples are by no means representative, but they illustrate the motivations well.
 
-\begin{itemize}
-\item \textbf{Multi-core} is an easy to see challenge caused by an external development. 
-   It led to the popularity of \emph{immutable} data structures (and functional programming,
-   in general) and it was also partly motivation for asynchronous workflows \cite{xx}.
+\paragraph{Parallel programming.}
 
-\item \textbf{Data access} is a more subtle challenge. Technologies like LINQ \cite{xx} make it
-   significantly easier, but it was not easy to see that inline SQL was a poor solution.
-   This is even more the case for F\# \emph{type providers}. You will not realize how poor the
-   established data access story is until you \emph{see} something like 
-   the WorldBank and R provider \cite{xx} or CSV type provider \cite{xx}.
-\end{itemize}
+The rise of multi-core CPUs is a clear example of an external development influencing
+programming language research. As multi-core and multi-processor systems became de-facto standard, 
+languages had to provide better abstractions for parallel programming. This led to the industrial 
+popularity of \emph{immutable} data structures (and functional programming in general), software 
+transactional memory \cite{app-haskell-stm}, data-parallelism and also asynchronous 
+computing \cite{app-fsharp-async}.
 
-I believe that the next important practical challenge for programming language designers
-is of the kind that is not easy to see -- because we are so used to doing things in 
-certain ways that we cannot see how poor they are. The problem is designing languages
-that are better at working with (and understanding) the \emph{context} in which programs are
-executed.
+In this case, the motivation is easy to see -- writing multi-core programs using earlier 
+abstractions, such as threads and locks, is difficult and error-prone. At the same time, 
+multi-core CPUs become the standard very quickly and so the lack of good language abstractions
+was apparent.
+
+\paragraph{Data access.}
+
+Accessing data is an example of a more subtle challenge. Initiatives like open government data
+certainly make more data available. However, to access the data, one has to parse CSV and Excel files, 
+issue SQL or SPARQL queries (to query database and the semantic web, respectively).
+
+Technologies like LINQ \cite{app-linq} make querying data significantly easier. But perhaps 
+because accessing data became important more gradually, it was not easy to see that inline
+SQL is a poor solution.
+
+This is even more the case for \emph{type providers} -- a recent feature in F\# that integrates
+external data sources directly into the type system of the language and thus makes data explorable
+directly from the source code editor (through features such as auto-completion on object members).
+It is not easy to see the limitations of standard techniques (using HTTP requests to query REST 
+services or parsing CSV files and using string-based lookup) until one sees how type providers 
+change the data-scientist's workflow\footnote{This is difficult to explain in writing and so the 
+reader is encouraged to watch a video showing type providers for the WorldBank and CSV data 
+sources \cite{app-fsharp-world}.}.
+
+\paragraph{Context-aware programming.}
+
+In this chapter, we argue that the next important practical challenge for programming language 
+designers is designing languages that are better at working with (and understanding) the 
+\emph{context in which programs are executed}.
+
+This challenge is of the kind that is not easy to see -- perhaps because we are so used to 
+doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
+uncover such flaws -- we look at a number of basic programs that rely on contextual 
+information, we explain why they are inappropriate and then we briefly outline how this
+thesis remedies the situation.
+
+Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
+the goal of programming language research is generally to design languages that provide 
+more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and 
+more \emph{unified}. 
+These are exactly the aims that we follow in this thesis. In this chapter, we explain what
+the common problems are. In Chapter~\ref{ch:flat-coeffects} and Chapter~\ref{ch:structural-coeffects},
+we develop two simple calculi to understand and capture the structure of the problems and,
+finally, Chapter~\ref{ch:unified-coeffects} unifies the two abstractions.
 
@@
 
-\section{Context-aware programming matters}
+\section{Why context-aware programming matters}
 
-The phrase \emph{context in which programs are executed} sounds quite abstract and generic.
-What are some concrete examples of such \emph{context}? For example:
+The phrase \emph{context in which programs are executed} sounds rather abstract and generic.
+What notions of \emph{context} can be identified in modern software systems?
+Different environments provide different resources (\eg~database
+or GPS sensor), environments are increasingly diverse (\eg~multiple versions of different
+mobile platforms). Web applications are split between client, server and mobile components; mobile
+applications must be aware of the physical environment while the ``internet
+of things`` makes the environments even more heterogeneous. At the same time, applications access
+rich data sources and need to be aware of security policies and provenance information from the
+environment.
+
+Writing such context-aware (or environment-aware) applications is a fundamental problem of modern 
+software engineering. The state of the art relies on ad-hoc approaches -- using hand-written conditions 
+or pre-processors for conditional compilation. Common problems that developers face include:
 
 \begin{itemize}
-\item When writing a cross-platform application, different platforms (and even different 
-   versions of the same platform) provide different contexts - the API functions that are 
-   available.
+\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple 
+  targets (e.g. SQL \cite{app-linq}, OpenCL or JavaScript \cite{app-hop-lang}) a part of the compilation 
+  (\eg~SQL generation) often occurs at runtime and developers have no guarantee that it will succeed 
+  until the program is executed.
 
-\item When creating a mobile app, the different capabilities that you may (or may not) have
-   access to are context (GPS sensor, accelerometer, battery status).
+\item \textbf{Platform versions.} When developing cross-platform applications, different platforms
+  (and different versions of the same platform) provide different API functions. Writing a 
+  cross-platform code usually relies on (fragile) conditional compilation or (equally fragile) 
+  dynamic loading.
 
-\item When working with data (be it sensitive database or social network data from Facebook),
-   you have permissions to access only some of the data (depending on your identity) and
-   you may want to track \emph{provenance} information. This is another example of a context.
+\item \textbf{Security and provenance.} When working with data (be it sensitive database or social 
+   network data), we have permissions to access only some of the data and we may want to track 
+   \emph{provenance} information. However, this is not checked -- if a program attempts to access 
+   unavailable data, the access will be refused at run-time.
+
+\item \textbf{Resources \& data availability.} When creating a mobile application, the program may
+  (or may not) be granted access to device capabilities such as GPS sensor, social updates or battery
+  status. We would like to know which of the capabilities are required and which are optional
+  (\ie~enhance the user experience, but there is a fallback strategy). Equally, on the server-side,
+  we might have access to different database tables and other information sources.
 \end{itemize}
 
-These are all fairly standard problems that developers deal with today. As the number of 
-devices where programs need to run increases, dealing with diverse contexts will be becoming
-more and more important (and I'm not even talking about \emph{ubiquitous computing} where you need
-to compile your code to a coffee machine).
+Most developers do not perceive the above as programming language flaws -- they are simply common
+programming problems (at most somewhat annoying and tedious) that had to be solved. However, this 
+is because we do not realize that a suitable language extension could make the above problems 
+significantly easier to solve. As the number of distinct contexts and their diversity increases,
+these problems will become even more commonplace.
 
-We do not preceive the above things as problems (at best, annoyances that we just have to 
-deal with), because we do not realize that there should be a better way. Let me digg into
-four examples in a bit more detail.
+The following sub-sections explore four examples in more details. The examples are chosen to 
+demonstrate two distinct forms of contexts that are studied in this thesis.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\kvd{for}~\ident{header},\ident{value}~\kvd{in}~\ident{header}~\kvd{do}\\[-0.25em]
+\hspace{2em}\kvd{match}~\ident{header}~\kvd{with}\\[-0.25em]
+\hspace{2em}|~\str{accept} \rightarrow~ \ident{req.Accept} \leftarrow \ident{value}\\[-0.25em]
+%
+\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_USERAGENT}\\[-0.25em]
+\hspace{2em}|~\str{user-agent} \rightarrow~ \ident{req.UserAgent} \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#else}\\[-0.25em]
+\hspace{2em}|~\str{user-agent} \rightarrow~
+  \ident{req.Headers}.[\,\ident{HttpHeader.UserAgent}\,] \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#endif}\\[-0.25em]
+%
+\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_REFERER}\\[-0.25em]
+\hspace{2em}|~\str{referer} \rightarrow~ \ident{req.Referer} \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#else}\\[-0.25em]
+\hspace{2em}|~\str{user-agent} \rightarrow~
+  \ident{req.Headers}.[\,\ident{HttpHeader.Referer}\,] \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#endif}\\[-0.25em]
+%
+\hspace{2em}|~\ident{other} \rightarrow~ \ident{req.Headers}.[\,\ident{other}\,] \leftarrow \ident{value}\\[-0.25em]
+\end{array}
+\end{equation*}
+
+\vspace{1em}
+\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
+\vspace{-1.5em}
+\caption{Conditional compilation in the HTTP module of the F\# Data library}
+\label{fig:introduction-context-http}
+\end{figure}
 
@@
 
-The number of different platform versions that you need to target is increasing, no matter what 
-platform are you using. For Android, there is a number called API Level 
-\footnote{\url{http://developer.android.com/guide/topics/manifest/uses-sdk-element.html}}
-with mostly additive changes (\emph{mostly} sounds very reassuring). In the .NET world, there are multiple
-versions, mobile editions, multiple versions of Silverlight etc. So, code that targets multiple
-frameworks easily ends up looking as the following sample from the Http module in F\# Data\footnote{
-\url{https://github.com/fsharp/FSharp.Data/blob/master/src/Net/Http.fs}}:
+The diversity across devices means that developers need to target an increasing number of platforms
+and possibly also multiple versions of each platform. For Android, there is a number called 
+API level \cite{app-android-apilevel} which ``uniquely identifies the framework API revision offered 
+by a version of the Android platform''. Most changes in the libraries (but not all) are additive.
 
-\begin{verbatim}
-for header, value in headers do
-  match header with
-  | StringEquals "accept" -> req.Accept <- value
-#if FX_NO_WEBREQUEST_USERAGENT
-  | StringEquals "user-agent" -> 
-      req.Headers.[HttpRequestHeader.UserAgent] <- value
-#else
-  | StringEquals "user-agent" -> req.UserAgent <- value
-#endif
-#if FX_NO_WEBREQUEST_REFERER
-  | StringEquals "referer" -> 
-      req.Headers.[HttpRequestHeader.Referer] <- value
-#else
-  | StringEquals "referer" -> req.Referer <- value
-#endif
-  | _ -> req.Headers.[header] <- value
-\end{verbatim}
+Equally, in the .NET ecosystem, there are multiple versions of the .NET runtime, mobile and 
+portable versions of the framework etc. The differences may be subtle -- for example, some members
+are omitted to make the mobile version of the library smaller, some functionality is not available
+at all, but naming can also vary between versions.
 
-This is difficult to write (you won't know if the code even compiles until you try building all 
-combinations) and maintaining such code is not particularly nice (try adding another platform version).
-Maybe we could refactor the code in F\# Data to improve it, but that's not really my point - supporting
-multiple platforms should be a lot easier.
+For example, the Figure~\ref{fig:introduction-context-http} shows an excerpt from the \ident{Http} module 
+in the F\# Data library\footnote{The file version shown here is available at:
+\url{https://github.com/fsharp/FSharp.Data/blob/b4c58f4015a63bb9f8bb4449ab93853b90f93790/src/Net/Http.fs}}.
+The example uses conditional compilation to target multiple versions of the .NET framework. Such code
+is difficult to write -- to see whether a change is correct, it had to be recompiled for all combinations
+of pre-processor flags -- and maintaining the code is equally hard. The above example could be refactored
+and the .NET API could be cleaner, but the fundamental issue remains. If the language does not understand
+the context (here, the different platforms and platform versions), it cannot provide any static
+guarantees about the code.
 
-On Android, you can access API from higher level platform dynamically using techniques like 
-reflection and writing wrappers 
-\footnote{\url{http://android-developers.blogspot.cz/2009/04/backward-compatibility-for-android.html}}
-Again, this sounds very error prone - to quote the linked article 
-\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}.
-Testing is no doubt important. But at least in statically typed langauges, we
-should not need to worry about calling a method that does not exist when writing
-multi-platform applications!
+As an alternative to conditional compilation, developers can use dynamic loading. For example, 
+on Android, programs can access API from higher level platform dynamically using techniques like 
+reflection and writing wrappers. This is even more error prone. As noted in an article introducing
+the technique\footnote{Retrieved from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}:
+\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}. Again, it would be 
+reasonable to expect that statically-typed languages could provide a better solution.
 
@@
 
-\subsubsection{Context awareness \#2: System capabilities}
+\subsection{Context awareness \#2: System capabilities}
 
-Another example related to the previous one is when you're using something like LINQ or 
-FunScript (\url{http://funscript.info/}) to translate code written in a sub-set of C\# or F\#
-to some other language - such as SQL or JavaScript. This is another important area, because
-you can use this technique to target JavaScript, GPUs, SQL, Erlang runtime or other components 
-that use a particular programming language.
+Another example related to the previous one is when libraries use meta-programming techniques
+(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
+written in a subset of a host language to some other target language, such as SQL, OpenCL or 
+JavaScript. This is an important technique, because it lets developers targets multiple 
+heterogeneous runtimes that have limited execution capabilities.
 
-For example, take a look at the following LINQ query that selects product names where the
-first upper case letter is "C":
-
-\begin{verbatim}
-var db = new NorthwindDataContext();
-
-from p in db.Products
-where p.ProductName.First(c => Char.IsUpper(c)) == "C"
-select p.ProductName;
-\end{verbatim}
-
-This looks like a perfectly fine piece of code and it compiles fine. But when you try running
-it, you get the following error:
-
+For example, consider the following LINQ query written in C\# that queries a database and
+selects product names where the first upper case letter is "C":
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
+\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
+\kvd{where}~\ident{p.ProductName.First}(\ident{c} \Rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
+\kvd{select}~\ident{p.ProductName};
+\end{array}
+\end{equation*}
+%
+This appears as a perfectly valid code ant the C\# compiler accepts it. However, when the program
+is executed, it fails with the following error:
+%
 \begin{quote}
@@
 \end{quote}
+%
+The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
+uses \ident{First} method to iterate over characters of a string, which is not supported. This is
+not a technical limitation of LINQ, but a fundamental problem of the approach.
 
-The problem is that LINQ can only translate a \emph{subset} of normal C\# code. Here, we are using
-the \texttt{First} method to iterate over characters of a string and that's not supported. This is
-not a technical limitation of LINQ, but a fundamental problem of the approach. If we target a 
-limited language, we simply cannot support the full source language. Is this an important 
-problem? If you search on Google for ``has no supported translation to SQL'' 
-(which is a part of a similar error message that you get in another case), you get some 26900
-links. So yes - this is an issue that people are hitting all the time.
+When cross-compiling to a limited environment, we cannot always support the full source language.
+The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
+Google search returns 11800 results for the message above and even more (44100 results) for a
+LINQ error message \emph{``Method X has no supported translation to SQL``} caused by a similar 
+limitation.
 
@@
 
-\subsubsection{Context awareness \#3: Confidentiality and provenance}
+\subsection{Context awareness \#3: Confidentiality and provenance}
 
-The previous two examples were mainly related to the (non-)existence of some API functions
-or to their behaviour. However, this is not the only case of context-awarness that is important
-in every day programming.
+The previous two examples were related to the non-existence of some library functions in another
+environment. Another common factor was that they were related to the execution context of the whole 
+program or a scope. However, contextual properties can be also related to specific variables.
 
-Most readers of this blog will immediately see what is wrong with the following code, but that
-does not change the fact that it can be compiled and deployed (and that there is a large number
-of systems that contain something similar):
+For example, consider the following code sample that accesses database by building a SQL query
+using string concatenation:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
+\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
+\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
+\end{array}
+\end{equation*}
+%
+The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
+(an attacker could enter \str{'; DROP TABLE Products -{}-} as their \ident{name} and delete the 
+database table with products). For this reason, most libraries discourage building SQL commands by
+string concatenation, but there are still many systems that do so.
 
-\begin{verbatim}
-let query = sprintf "SELECT * FROM Products WHERE ProductName = '%s'" name
-use cmd = new SqlCommand(query)
-use reader = cmd.ExecuteReader()
-\end{verbatim}
+The example demonstrates a more general property. Sometimes, it is desirable to track additional 
+meta-data about variables that are in some ways special. Such meta-data can determine how the variables
+can be used. Here, \ident{name} comes from the user input. This \emph{provenance} information should 
+be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
+not directly contain user input (in an unchecked form). Such marking of values (but at run-time)
+is also called tainting \cite{app-tainting-sql}.
 
-The problem is obviously SQL Injenction \footnote{\url{http://en.wikipedia.org/wiki/SQL_injection}}. 
-Concatenating strings to build an SQL command is a bad practice, but it is an example of a more general problem.
+Similarly, if we had \ident{password} or \ident{creditCard} variables in a client/server web application,
+these should be annotated as sensitive and it should not be possible to send their values over an
+unsecured network connection.
 
-Sometimes, we have special kinds of variables that should have some contextual meta-data associated
-with them. Such meta-data can dictate how the variable can be used. Here, \texttt{name} comes from the user
-input and this \emph{provenance} information should propagate to \texttt{query} and we should get an error when
-we try passing this - potentially unsafe - input to \texttt{SqlCommand}. Similarly, if you have \texttt{password} or
-\texttt{creditCard}, it should be annotated with meta-data saying that this is a sensitive piece of data and
-should not, for example, be sent over an unsecured network connection.
-As a side note - this idea is related to a technique called taint checking \cite{xx}.
-
-In another context, if you are working with data (e.g. in some data jouranlism application), it would be
-great if your sources were annotated with meta-data about the quality and source of the data (e.g.
-can it be trusted? how up-to-date is it?) The meta-data could propagate to the result and tell you how
-accurate and trustworthy your result is.
+In another context, when working with data (\eg~in data journalism), it would be desirable to 
+track meta-data about the quality and the source of the data. For example, is the source trustworthy?
+Is the data up-to-date? Such meta-data could propagate to the result and tell us important 
+information about the calculated results.
 
@@
 
-\subsubsection{Context-awareness \#4: Resource \& data availability}
+\subsection{Context-awareness \#4: Access patterns in data-flow languages}
+
+As an example, consider a language that allows us to get a value of a variable (representing
+some changing data-source) \ident{x} versions back using the syntax $\ident{x}_{[a]}$. 
+To track information about individual variables, we use a product-like operation $\times$ on tags 
+to mirrors the product structure of variables. For example:
+%
+\begin{equation*}
+?
+\end{equation*}
+%
+The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
+that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
+\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
+maximum of the requirements of the individual variables:
+%
+\begin{equation*}
+\begin{array}{l}
+?
+\end{array}
+\end{equation*}
+%
+This version of the calculus removes the non-determinism of the lambda abstraction from the previous
+version. As we associate information with individual variables, lambda abstraction creates a function
+that requires the context requirements associated with the variable that is being abstracted over.
+
+For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
+\ident{b} from the declaration-site) then the function would have context requirements $5$ -- that
+is the number associated with the variable \ident{a}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Context-awareness \#4: Resource \& data availability}
 

### 2014-03-14 text/Chapter01.tex
Editing introduction chapter

@@
 %----------------------------------------------------------------------------------------
-
-Modern software applications behave differently depending on the environment or context 
-in which they execute. They often run in increasingly rich environments that provide 
-resources (e.g. database or GPS sensor) and are gradually more diverse (e.g. multiple 
-versions of different mobile platforms). Web applications are split between client, server 
-and mobile components; mobile applications must be aware of the context and of the platform 
-while the ``internet of things'' makes the environments even more heterogeneous; 
-applications that access rich data sources need to propagate security policies and provenance 
-information about the data.
-
-Writing such context-aware applications is a fundamental problem of modern software 
-engineering. The state of the art relies on ad-hoc approaches -- using hand-written conditions 
-or pre-processors for conditional compilation. Common problems that developers face include:
-
-\begin{itemize}
-\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple 
-  targets (e.g. SQL, CUDA or JavaScript) the compilation often occurs at runtime and developers 
-  have no guarantee that it will succeed until the program is executed.
-
-\item \textbf{Platform versioning.} When developing application for multiple versions of a 
-  system (e.g. Android), developers rely on lazy loading at runtime or use conditional compilation 
-  using \ident{\#if}. The former delays errors to runtime, while the latter requires building all 
-  possible configurations to discover simple compile-time errors.
-
-\item \textbf{Resources \& data availability.} When programming applications that access 
-  resources or data provided by the environment (e.g. specific database table, GPS sensor), 
-  the program typically performs dynamic check for the resource availability. However, this 
-  is not checked in any way -- we have no easy way to tell what happens when the resource is 
-  not available (e.g. is there a fallback strategy or not?)
-
-\item \textbf{Meta-programming.} In distributed client-server programming, we often need to 
-  manipulate code using meta-programming techniques on the server and then pass a closed 
-  expression to client, translated to another language (JavaScript). We need to be able to 
-  guarantee that such expressions are well-formed before they are translated.
-\end{itemize}
-
-The issues discussed above pose an important problem to current and future state of computing.
-In this thesis, we develop a unifying mathematical model that allows us to capture the properties
-discussed above. We use the model in three different ways to capture different notions of 
-context-dependence.
 

### 2014-03-14 text/Chapter05.tex
Editing introduction chapter

--- 2013-06-11 text/Chapter05.tex
@@
-\chapter{Coeffect meta-language} 
-\label{ch:metalanguage}
+\chapter{Unified coeffect language} 
+\label{ch:unified-coeffects}
 
-%---------------------------------------------------------------------------------------------------
+% ==================================================================================================
+% Macros - for ICFP 2014
+% ==================================================================================================
 
-Introduction
+% comonadic semantics
+\newcommand{\F}{\mathsf{F}}
+\newcommand{\D}{\mathsf{D}}
+\newcommand{\Ds}[3]{\D^{\sclrd{#1}}_{#2} #3}
+\newcommand{\Icat}{\cclrd{\mathbb{I}}}
+\newcommand{\id}{\mathit{id}}
+\newcommand{\B}[1]{\bar{#1}}
+\newcommand{\bbullet}{\mathop{\bar{\bullet}}}
+\renewcommand{\idf}[0]{ \ident{id} }
+\newcommand{\m}{\mathsf{m}}
+\newcommand{\n}{\mathsf{n}}
+
+\newcommand{\shortar}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar_{#4}"#1";"#2"}
+\newcommand{\shortaro}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar^{#4}"#1";"#2"}
+\newcommand{\shortarn}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar_{#4}"#1";"#2"}
+\newcommand{\shortaron}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar^{#4}"#1";"#2"}
+\newcommand{\diagar}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}^{#4}"#1";"#2" }
+\newcommand{\diagaro}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}_{#4}"#1";"#2" }
+
+% Typing rule and typing statement (single line)
+\newcommand{\tyrule}[3]{ \inference[(\footnotesize{#1})~]{#2}{#3} }
+\newcommand{\tyruleCompact}[3]{ \text{(\footnotesize{#1})}~\dfrac{#2}{#3} }
+\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
+% Context transformation and context transformation with new line in the middle
+\renewcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
+\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
+% Ordinary context where colouring is done by hand 
+\newcommand{\coctx}[2]{ #1 \text{\scriptsize @} {\footnotesize{\text{$#2$}}}  }
+
+% Simple coeffect systems in the introduction
+\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}
+
+% Coeffect algebra
+\definecolor{cclr}{rgb}{0.0,0.4,0.0}
+\newcommand{\cclrd}[1]{ \textcolor{cclr}{#1} }
+
+\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
+\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
+\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
+\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }} % \bullet} } % \textcolor{cclr}{\bullet}}  
+\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
+\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
+
+% \newcommand{\single}[{\langle{#1}\rangle}
+
+\newcommand{\uncurry}{\Lambda^{\hspace{-0.2em}-1}}
+
+% Coeffect shapes 
+\definecolor{sclr}{rgb}{0.5,0,0.0}
+\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }
+
+\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
+\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
+\renewcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
+\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
+
+\newcommand{\sflat}{\star}
+
+% Coeffect algebra
+\definecolor{aclr}{rgb}{0.0,0.0,0.7}
+\newcommand{\aclrd}[1]{ \textcolor{aclr}{#1} }
+
+\newcommand{\mksh}[1]{{\sclrd{[}#1\sclrd{]}}}
+
+\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
+\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
+\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq 
+% Tomas: I prefer \circledast here because it is a circled (lifted) version of \cseq 
+% { \textcolor{aclr}{\bullet}}
+\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}} 
+\newcommand{\azero}{ \textcolor{aclr}{\bot} }
+\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
+\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }
+
+% Free monoid used in the *structural* coeffect algebra
+\newcommand{\fcons}{ \times } 
+\newcommand{\fempty}{\epsilon}
 
@@
 \section{Introduction}
-\label{sec:metalanguage-intro}
 
-?
-\ExecuteMetaData[rules/rules.tex]{cml-types}
+\emph{Context} is important for defining meaning-- not just in natural languages, but also in logics and 
+programming languages.
+The standard notion of context in programming is an environment 
+providing values for free variables. An open term with free variables is context dependent -- its 
+meaning depends on the free-variable context. The simply-typed $\lambda$-calculus famously 
+analyses such context usage. 
+Other systems go further. %capturing how 
+%the context is used. 
+For example, bounded linear logic tracks the number of times a 
+variable is used~\cite{girard1992bounded}.
 
-\section{Summary}
-Oops!
+In software engineering, ``context'' provides more than just free-variable values. For example, in a
+distributed system, the context provides different resources that may be available on different 
+devices (\eg{}, a database on a server or a GPS sensor on a phone).
+
+In this paper we develop a calculus for capturing various notions of 
+context in programming. A key feature and contribution
+of the calculus is its \emph{coeffect system} which provides a static analysis for contextual properties (coeffects). The system
+follows the style of type and effect systems, but captures a different class of properties. % (Section~\ref{sec:intro-pathways}). 
+Another key contribution of the calculus is its semantics which can be smoothly instantiated for specific notions of context.
+%The semantics of the calculus is general and can be instantiated for particular notions of context.
+
+Coeffect systems were previously introduced as a generic analysis of
+context dependence which can be instantiated for various notions of
+context~\cite{petricek2013coeffects}. However, the formalization was restricted to
+tracking a class of \emph{whole-context} properties where a term has just
+one coeffect. This limited the applications and precision of any
+analysis. For example, a whole-context liveness analysis marks the 
+free-variable context as live (some variable may be used) or dead 
+(no variable is used), but it cannot record liveness \emph{per-variable}.
+
+We develop a more general system which captures both
+\emph{per-variable} coeffects, which we call \emph{structural}, and
+\emph{whole-context} coeffects, which we call \emph{flat}, and more. Our
+key contributions are:
+
+\begin{itemize}
+\item We present a general coeffect type system (Section~\ref{sec:coeffects}) and demonstrate two 
+  concrete uses -- \emph{flat} coeffect systems that tracks whole-context information and 
+  \emph{structural} coeffect systems which tracks fine-grained per-variable information.
+
+\item We show practical examples, instantiating the calculus for structural systems capturing 
+  variable usage based on bounded linear logic, dataflow caching, and precise liveness analysis.
+  We also instantiate the calculus to flat systems, building on and extending previous examples~\cite{petricek2013coeffects}.
+% variants of the coeffect system capture examples described   in our earlier work
+
+\item We discuss the syntactic properties of the \emph{structural} coeffect system 
+  (Section~\ref{sec:syntax}). It satisfies type preservation under both $\beta$-reduction 
+  and $\eta$-expansion, allowing its use with both call-by-name and call-by-value languages.
+  This important property distinguishes it from both effect systems and flat coeffects.
+
+\item We provide a denotational semantics, resisting and extending the notion of \emph{indexed comonads} to the 
+structural setting (Section~\ref{sec:semantics}). We prove soundness by showing 
+  the correspondence between syntactic and semantic properties of coeffect systems.
+\end{itemize}
+
+\noindent
+Coeffects can be approached from multiple directions (Section~\ref{sec:intro-pathways}) including 
+syntactic (effect systems), semantic, and proof-theoretic. We 
+emphasize the syntactic view, though we also outline a categorical semantics
+and note the interesting technical details.
+
+% ==================================================================================================
+
+\section{Why coeffects matter}
+\label{sec:intro-coeffects}
+
+Coeffects are way to describe notions of context in programming that keep turning up. 
+To illustrate this, we overview three systems tracking contextual properties that 
+ motivate our general coeffect system. Two systems track per-variable properties (bounded 
+linear logic and dataflow) and one tracks whole-context properties (implicit parameters).
+ We start with some background and finish with a brief overview of the literature leading to coeffects.
+
+\subsection{Background, scalars and vectors}
+
+The $\lambda$-calculus is asymmetric-- it maps a context with \emph{multiple} variables to a 
+\emph{single} result. An expression with $n$ free variables of types $\sigma_i$ can be modelled
+by a function
+$\sigma_1 \times \ldots \times \sigma_n \rightarrow \tau$ with a product on the left, but a single value
+on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
+we attach coeffects to the context $\sigma_1 \times \ldots \times \sigma_n$ and we often (but not always)
+have one coeffect per each variable. We call the overall coeffect a \emph{vector} consisting of
+\emph{scalar} coeffects. This asymmetry explains why coeffect systems are not trivially dual to 
+effect systems.
+
+It is useful to clarify how vectors are used in this paper. Suppose we have a set $\C$ of
+\emph{scalars} such as $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ over $\C$
+is a tuple $\langle \cclrd{r_1}, \ldots, \cclrd{r_n}\rangle$ of scalars. 
+We use letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for 
+vectors and $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars.\footnote{For better readability, the paper 
+distinguishes different structures using colours. However ignoring the colour does not introduce 
+any ambiguity.} We also say that a \emph{shape} of a vector $\mksh{\aclrd{R}}$ (or more generally any container) 
+is the set of \emph{positions} in a vector. So, a vector of length
+$n$ has shape $\{ 1, 2, \ldots, n \}$.
+
+Just as in scalar-vector multiplication, we lift any binary operation on scalars into a scalar-vector one:
+$\cclrd{s} \bullet \aclrd{R} = \langle \cclrd{s}\bullet\cclrd{r_1}, \ldots, \cclrd{s}\bullet\cclrd{r_n}\rangle$.
+Given two vectors $\aclrd{R},\aclrd{S}$ of the same shape, containing partially ordered scalars, we 
+write $\aclrd{R} \leq \aclrd{S}$ for the pointwise extension of $\leq$ on scalars. Finally,
+the associative operation $\times$ concatenates vectors.
+
+We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
+but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
+be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.
+
+\subsection{Bounded reuse}
+\label{sec:intro-coeffects-bll}
+
+Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
+This is done by limiting the number of times a value (proposition) can be used. An assumption
+$!_k A$ means that a variable can be used at most $k$ times. 
+We attach annotations to the whole context rather than individual assumptions and so a context
+$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
+This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
+%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.
+
+Bounded linear logic includes explicit weakening and contraction rules
+that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
+\[
+\inference
+  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
+  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
+\quad
+\inference
+  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
+  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
+\]
+The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
+$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
+In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
+occurrences of a proposition are joined by adding the number of uses.
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\begin{align*}
+\begin{array}{c}
+%
+\hspace{-0.6em}
+\begin{array}{c}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
+\quad 
+\tyruleCompact{weak}
+  {\coctx{\Gamma}{R} \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
+\\[1.8em]
+\tyruleCompact{sub}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
+\quad 
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
+\\[1.2em]
+\end{array} \\
+%
+% ------------------------ 
+%
+\begin{array}{c}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
+    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
+\\[1.3em]
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
+\\[1.3em]
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
+\end{array} 
+% ------------------------ 
+%
+% \hspace{-1em}
+%\begin{array}{c}
+%
+% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
+% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
+% to keep it simple.
+%
+% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
+% (I do not understand why but it makes the line somehow thicker...)
+%
+% \tyruleCompact{let}
+%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
+%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
+%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
+\end{array}
+\end{align*}
+\vspace{-0.5em}
+\caption{type and coeffect system for bounded reuse}
+\label{fig:bounded-coeff}
+\end{figure}
+
+\paragraph{Bounded linear coeffects.}
+The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
+Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
+$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
+the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
+exchange (\emph{exch}) rules manipulate variables in the context and
+modify the annotations accordingly -- adding the number of uses in
+contraction and switching vector elements in exchange.
+
+For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
+and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
+\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
+annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.
+
+Application (\emph{app}) describes call-by-name evaluation.  Applying
+a function that uses its parameter $\cclrd{t}$-times to an argument
+that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
+total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
+Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
+vector. Meanwhile, the variables in $\Gamma_1$ are used just
+$\aclrd{R}$-times when reducing the expression $e_1$ to a function
+value.
+
+Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
+the number of uses using the pointwise
+$\leq$ relation. We can view any variable as being used a greater
+number of times than it actually is.
+
+\paragraph{Example.} To demonstrate, consider a term
+$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
+times -- once directly inside the function and twice via the variable $v$ after 
+substitution. Similarly, $y$ is used 
+twice. Assuming a judgment for the function body, abstraction yields:
+%
+\begin{equation*}
+\tyrule{abs}
+ { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
+ { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
+\end{equation*}
+%
+To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
+Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
+a judgment that is simplified as follows:
+\[
+\hspace{-0.7em}\inference
+  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
+          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
+{\inference
+  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
+  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
+\]
+%
+The first step performs scalar multiplication, producing the vector
+$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
+$x$ and $x'$ from the function and argument terms respectively.
+
+It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
+the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
+show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Dataflow and data access}
+
+Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
+An expression is re-evaluated when new inputs are available (push) or when more output
+is demanded (pull).  In causal dataflow, programs can access past
+values of a stream. We consider a language where $\kvd{prev}~e$
+returns the previous value of $e$, where 
+$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.
+
+An implementation of causal dataflow may cache past values of
+variables as an optimisation. The question is, how many past values
+should be cached?  This can be approximated by a coeffect system.
+
+\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
+similar to the one for bounded reuse in that it tracks a vector
+of numbers $\aclrd{R}$ as part of the context
+$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
+past values (\emph{causality depth}) required for a variable.
+
+Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
+coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
+accessed variables (\emph{var}) are annotated with
+$0$ meaning that no past value is required (only the
+current one). The (\emph{prev}) rule crates caching requirements --
+it increments the number of required values for all variables
+used in $e$ using scalar-vector addition.
+
+Application and contraction have the same structure as before, 
+but use different operators. If two variables 
+are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
+values, then overall we need at most
+$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
+two caches are combined with the maximum of the two requirements, 
+which satisfy the smaller requirements.
+
+In (\emph{app}), the function requires $\cclrd{t}$ past values of
+its parameter.  This means $\cclrd{t}$ past values of
+$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
+variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
+call (\eg{}, we need $1+\aclrd{S}$ values to
+get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
+values to get $2$ past values of $\sigma$, \emph{etc.}).
+
+\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
+$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
+and one bound:
+\[
+\dfrac
+  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
+  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
+\]
+The expression always requires the previous value of $y$ and adds it to
+a previous value of the parameter $x$. Evaluating the value of the
+argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
+and so the overall requirement is $3$ past values:
+\[
+\inference
+  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
+{\inference
+  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
+  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
+\]
+The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
+(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
+the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
+Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
+number of enclosing $\kvd{prev}$s.% expressions.
+
+We previously used dataflow as an example of
+coeffects~\cite{petricek2013coeffects}, but tracked caching
+requirements on the whole context. The system outlined here is more
+powerful and practically useful, with finer-grained coeffects tracking
+caching requirements per-variable.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\[
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
+\]
+\[
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
+    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
+\]
+\[
+\tyrule{var}{}
+  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
+\quad\;
+\tyrule{prev}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
+  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
+\]
+\caption{type and coeffect system for dataflow caching}
+\label{fig:dataflow-coeff}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Implicit parameters}
+\label{sec:intro-coeffects-impl}
+
+As our third example, we revisit Haskell implicit parameters \cite{lewis2000implicit} 
+used in our earlier coeffect work \cite{petricek2013coeffects}. Implicit parameters are variables
+that mix aspects of dynamic and lexical scoping. Implicit parameters are a distinct syntactic 
+category to variables and we write them as \ident{?p}. For simplicity, we omit \emph{let}-binding for
+implicit parameters and focus just on tracking requirements.
+
+\paragraph{Implicit parameters coeffects.} Implicit parameters are a whole-context coeffect 
+not linked to ordinary variables. We keep track of sets of implicit parameters that are required
+by an expression (and their types). For example $\coctx{\Gamma}{\{ \ident{?p}_1 : \tau_1, \ldots, \ident{?p}_n : \tau_n \}}$
+means that a context provides ordinary variables $\Gamma$ and values for implicit parameters $\ident{?p}_i$.
+Unlike in the previous examples, we no longer need to distinguish between coeffects attached
+to variables (scalars) and coeffects attached to contexts (vectors), so we write $\cclrd{r},\cclrd{s},\cclrd{t}$
+for both.
+
+Despite the differences, the type system in Figure~\ref{fig:implpar-coeff} follows the same 
+structure as the earlier two examples. Context requirements are created when accessing an implicit
+parameter (\emph{param}) (a system-specific rule). Structural rules (exchange, weaken, contract) 
+do not affect the coeffects. For example parameters are reordered in (\emph{exch}),
+but this has no effect as set union $\cup$ is commutative.
+
+In abstraction and application, the structural $\cons$ operator (previously vector 
+concatenation) becomes $\cup$. Sets of implicit parameters are not associated to
+individual variables and so they are unioned. The (\emph{app}) rule uses $\cup$ to
+combine the implicit parameters required by the function with the requirements of the argument too.
+
+We call this a \emph{flat} coeffect system since coeffects have only
+one shape (there is no scalar/vector distinction).  Other flat
+coeffect systems may use a richer structure~\cite{petricek2013coeffects}. In
+particular, the operations used in abstraction and application may differ
+(to accommodate over-approximation). We return to this in
+Section~\ref{sec:coeffects-structural}.
+
+\paragraph{Example.} Unlike structural coeffect systems, flat systems
+do not necessarily have principal coeffects. 
+This arises from the (\emph{abs}) rule which can freely split
+requirements between the function type and the declaring
+context. Consider a function %filters news based on a location:
+$\lambda ().\ident{?p}_1 + \ident{?p}_2$.  There are nine possible 
+ type and coeffect derivations, two of which are:
+%% 1 p1p2  p1p2
+%% 2 p1p2  p1
+%% 3 p1p2  p2
+%% 4 p1p2  
+%% 5 p1    p1p2
+%% 6 p1    p2
+%% 7 p2    p1p2
+%% 8 p2    p1
+%% 9       p1p2
+\[
+\begin{array}{rl}
+\coctx{\emptyset}{ \{  \} } &\hspace{-0.7em}\vdash (\ldots) : 
+  \ident{unit} \xrightarrow{ \{ \ident{?p}_1:\mathbb{Z}, \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
+\\
+\coctx{\emptyset}{ \{ \ident{?p}_1:\mathbb{Z} \} } &\hspace{-0.7em} \vdash (\ldots) : 
+  \ident{unit} \xrightarrow{ \{ \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
+\end{array}
+\]
+In the first case, both parameters are dynamically scoped and have to be provided by the caller.
+In the second case, the parameter $\ident{?p}_1$ is available in the declaring scope and so it is
+(lexically) captured.
+
+Although structural coeffects have more desirable syntactic properties, we aim to capture this
+non-principality too as it is practically useful -- Haskell's implicit parameters use it and
+it can be used to model resource rebinding in distributed systems such as~\cite{app-distributed-acute}).
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\[
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau,y:\sigma,\Gamma_2}{r \cup \cclrd{s} \cup \cclrd{t} \cup \cclrd{q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\sigma,x:\tau,\Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s} \cup \cclrd{q}} \vdash e : \tau}
+\]
+\[
+\tyrule{app}
+  { \coctx{\Gamma_1}{\cclrd{r}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
+    \coctx{\Gamma_2}{\cclrd{s}} \vdash e_2 : \sigma }
+  { \coctx{\Gamma_1, \Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s}} \vdash e_1 \, e_2 : \tau} 
+\]
+\[
+\tyrule{param}{}
+  {\coctx{()}{ \{ \ident{?p}:\tau \} } \vdash \ident{?p} : \tau}
+\quad\;
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \sigma}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
+\]
+\vspace{-0.5em}
+\caption{Type and coeffect system for implicit parameters}
+\label{fig:implpar-coeff}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Pathways to coeffects}
+\label{sec:intro-pathways}
+
+This paper mainly follows work on effect systems and their link to categorical semantics. We briefly 
+review this and other directions leading to coeffects. An eager reader can return to this section later.
+
+\paragraph{Effect systems.}
+
+Effect systems \cite{gifford1986effects} track
+effectful operations of computations such as memory access or lock
+usage \cite{flanagan2003type}. They are written as judgments $\Gamma
+\vdash e : \tau ~\&~\cclrd{\rho}$ associating effects
+$\cclrd{\rho}$ with the result. Effect systems capture \emph{output
+  effects} where, as Tate puts it, \emph{``all computations with [an]
+  effect can be thunked as pure computations for a domain-specific
+  notion of purity.''}~\cite{effects-producer-semantics}.  This
+thunking is typically a $\lambda$-abstraction. Given an effectful
+expression $e$, the function $\lambda x . e$ is an effect-free value
+that delays all effects:
+%
+\begin{equation*}
+\inference[({abs})\;]
+  {\Gamma, x\!:\!\tau_1 \vdash e : \tau_2 ~\&~\cclrd{\rho}} 
+  {\Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{\rho}} \tau_2~\&~\cclrd{\emptyset}}
+\end{equation*}
+%
+Coeffects do not follow this pattern.  In contrast to effect systems,
+context requirements cannot be easily ``thunked'' as
+pure values. Lambda abstraction can split context requirements
+between \emph{immediate} and \emph{latent} requirements. This is akin
+to how lambda abstraction splits a free-variable context into 
+the bound parameter (call site) and the remaining free variables (declaration site).
+
+\paragraph{Categorical semantics.}
+Moggi models effectful computations as functions of type 
+$\tau_1 \rightarrow \mtyp{}{\tau_2}$ where $M$ is monad providing composition of effectful 
+computations~\cite{moggi1991monads}. Wadler and Thiemann \cite{wadler2003marriage} link effect systems with monads using 
+annotated monads $\tau_1 \rightarrow \mtyp{\cclrd{\rho}}{\tau_2}$ whose semantics has been 
+provided by Katsumata~\cite{katsumata2014parametric}.
+
+Context-dependent computations require a different model. Uustalu and Vene \cite{uustalu2008notions} 
+use functions $C{}{\tau_1} \rightarrow \tau_2$ where $C$ is a \emph{comonad}. Our earlier work 
+\cite{petricek2013coeffects} used indexed comonads with denotations $C^{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ adding 
+annotations akin to Wadler and Thiemann. In Section~\ref{sec:semantics} we extend
+indexed comonads to capture the general coeffect systems of this paper,  
+in the style of Katsumata. 
+% 
+% TP: Mentioned Katsumata here. Thought it is nice to say we build on new work here (as opposed to ICALP)
+%
+
+\paragraph{Language and meta-language.}
+Moggi uses monads in two systems~\cite{moggi1991monads}. In the first system, a monad is used to model
+an effectful language itself -- the semantics of a language uses a specific monad. In the second system, 
+monads are added as type constructors, together with syntax corresponding to \emph{unit} and \emph{bind}
+operations.
+
+Looking at context, Uustalu and Vene \cite{uustalu2008notions} follow the first approach 
+(using a concrete comonad to model dataflow). Contextual-Modal Type Theory (CMTT) of Nanevski et al.
+\cite{logic-cmtt} follows the latter approach, adding a comonad via the $\Box$ modality of modal S4
+to the language. We focus on concrete languages using the first approach. 
+A ``coeffect meta-language'' is an interesting future work.
+
+\paragraph{Sub-structural and bunched types.}
+% \cite{substruct-linear-change} 
+Sub-structural type systems restrict how a context is used. This is achieved by removing some
+of the structural typing rules (weakening, contraction, exchange). As the bounded linear logic 
+example (Section~\ref{sec:intro-coeffects-bll}) shows, our system can be viewed as a generalization.
+
+% ==================================================================================================
+
+\section{The coeffect calculus}
+\label{sec:coeffects}
+
+The three calculi shown in the previous section track two kinds of
+contextual properties: bounded reuse and dataflow are structural
+(per-variable) systems, and implicit parameters and our earlier coeffect
+systems~\cite{petricek2013coeffects} are flat (whole-context) systems. 
+This section presents our primary contribution: the general coeffect
+calculus.
+
+The calculus is parameterised by an algebraic structure of coeffects.
+To capture both structural and flat systems, coeffect annotations are
+indexed by a \emph{shape}. In flat systems, the shape is a singleton set
+and so annotations are \emph{scalar} values. Structural systems use shapes
+matching the number of variables in a free-variable context and so annotations
+are \emph{vectors}. However, the coeffect calculus could also use shapes describing
+trees and other structures.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Understanding coeffects: syntax and semantics}
+
+The coeffect calculus provides both an analysis of
+context dependence (its coeffect system) and a semantics for
+context (see Section~\ref{sec:semantics}). These two features of the calculus provide different
+perspectives on coeffect annotations $\aclrd{R}$ in a judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$:
+%%
+\begin{itemize}
+\item Syntactically, coeffects model \emph{contextual requirements} and 
+  may be overapproximated, such that more capabilities are required 
+  than necessary at runtime.
+
+\item Semantically, coeffects model \emph{contextual capabilities} and behave 
+  like containers of capabilities, such that the semantics may throw away 
+  capabilities that will not be needed.
+\end{itemize}
+%
+Thus there are two dual ways to understand coeffect annotations. Each
+perspective implies an alternate reading of the typing rules.
+%
+\begin{itemize}
+\item
+As \emph{contextual requirements}, the rules should be read top-down.
+The requirements of multiple sub-terms are \emph{merged} and the
+requirements of function body are \emph{split} between immediate (declaration-site)
+and latent (call-site) coeffects.
+
+\item
+As \emph{contextual capabilities}, the rules should be read bottom-up.
+The capabilities provided to a larger term are \emph{split} between 
+sub-terms; for functions, the capabilities of declaration-site and call-site
+are \emph{merged} and passed to the body.
+\end{itemize}
+%
+The reason for this asymmetry follows from the fact that context appears in a \emph{negative
+position} in the model. In Section~\ref{sec:semantics}, the denotation of a judgment
+is a function of the form $\D_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau}$ where $\D_{\aclrd{R}} \interp{\Gamma}$ 
+encodes the contextual capabilities used to evaluate a term. Similarly
+functions have models of the form $\D_{\cclrd{s}} \interp{\sigma} \rightarrow \interp{\tau}$
+with additional contextual capabilities attached to the input.
+
+\subsection{Structure of coeffects}
+
+We describe the algebraic structure of coeffects in three
+steps. First, we define a \emph{\cclrd{coeffect scalar}} structure 
+which defines the basic building blocks of coeffect information;
+then we define \emph{\sclrd{coeffect shapes}} which determines how
+coeffect scalar values are related to the free-variable context. 
+Finally, we define the \emph{\aclrd{coeffect
+    algebra}} which consists of shape-indexed coeffect scalar values.
+
+For example, in bounded reuse the coeffect scalar structure
+comprised natural numbers $\cclrd{\mathbb{N}}$ with $\cclrd{+}$ and
+$\cclrd{\ast}$ operators. The shape for bounded reuse is the length of the
+free-variable context and so the coeffect annotation is a vector of matching length.
+Finally, the coeffect algebra specifies how vectors are
+concatenated and split in abstraction and application.
+
+In the coeffect system of the calculus, contexts are annotated with shape-indexed coeffects (\eg{}, vectors)  as in
+$\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$. However, functions take just a single 
+input parameter and so are annotated with scalar coeffect values as in $\sigma \xrightarrow{\cclrd{r}} \tau$.
+
+\paragraph{Coeffect scalar.}
+Coeffect scalar structures are equipped with two operations. In bounded reuse, those 
+were $\cclrd{\ast}$ for sequencing (in function application) and $\cclrd{+}$ for context 
+sharing (in contraction). Additional structure is needed for variable access and sub-coeffecting.
+
+\begin{definition}
+\emph{\cclrd{Coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set $\C$
+together with elements $\cunit,\czero \in \C$, relation $\cleq$ and binary operations 
+$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids
+and $(\C, \cleq)$ is a pre-order. The following distributivity axioms are required:
+\begin{align*}
+(\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+\end{definition}
+
+\noindent
+The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
+category in the semantics (Section~\ref{sec:semantics}). 
+It models sequential composition with variable access
+($\cunit$) as the identity. The other element ($\czero$) is used for
+variables that are not accessed. The operation $\cpar$ combines coeffects
+for contexts used in multiple places (contraction). 
+The notation is inspired by the bounded reuse example, which uses coeffect scalar structure
+$(\cclrd{\mathbb{N}}, \ast, +, 1,0, \cclrd{\leq})$, but be aware that $\cseq$ and $\cpar$ do
+not always mean $\ast$ and $+$.
+
+The context annotations can be viewed as \emph{containers} of scalar coeffects.
+For structural coeffects, the container is a vector, while for flat coeffects, it is
+a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{abbott2005containers}
+which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.
+
+\paragraph{Coeffect shapes.}
+The coeffect system is parameterised by a set of shapes $\SH$. A coeffect annotation 
+is indexed by a shape $\sclrd{s} \in \SH$ calculated from the shape of the free-variable vector. The correspondence is not necessarily 
+bijective. For example, flat coeffect systems have just a single shape $\SH = \{ \ast \}$.
+
+In the coeffect judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$, the coeffect annotation
+$\aclrd{R}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
+We write $\aclrd{s} = \mksh{\Gamma}$ for the shape corresponding to $\Gamma$. We define shapes by a 
+\emph{set} of positions and so we can write $\aclrd{R} \in \sclrd{s} \rightarrow \C$ as a
+mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
+the exponent $\aclrd{R} \in \C^{\sclrd{s}}$.
+
+The set of shapes is equipped with an operation that combines shapes (when we combine 
+variable contexts), an operation that computes shape from the free-variable contexts, 
+and two special shapes in $\SH$ representing empty context and singleton context.
+
+\begin{definition}
+  A \emph{\sclrd{coeffect shape}} structure $(\SH, \mksh{\textnormal{--}}, \stimes,
+  \sempty, \sunit)$ comprises a set $\SH$ with a binary operation
+  $\stimes$ on $\SH$ for shape composition, a mapping from contexts to
+  shapes $\mksh{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
+  \SH$ such that $(\SH, \stimes, \sempty)$ is a monoid. The 
+  elements $\sempty$ and $\sunit$ represent the shapes of empty and
+  singleton free-variable contexts respectively.
+\end{definition}
+
+\noindent
+As said earlier, we use two kinds of shape structures that describe the shape
+of vectors and the shape of trivial singleton container:
+
+\begin{itemize}
+\item
+Structural coeffect shape is defined as $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$.
+We treat numbers as sets $0 = \{\}, 1=\{ \emptyset \}, 2=\{ \emptyset, 1\}, 3=\{\emptyset, 1, 2 \} \ldots$ 
+(so that a number is a set of positions). The shape mapping ${|}\Gamma{|}$ returns the number
+of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, respectively,
+and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = {|}\Gamma_1{|} + {|}\Gamma_2{|}$.
+Therefore, a coeffect annotation is a \emph{vector} $\aclrd{R} \in
+\C^{\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{R}(i) \in \C$ for each
+variable $x_i$ in the context.
+
+\item
+Flat coeffect shape is defined as $\sclrd{(\{\sflat \}, \mathsf{const}\,\,\sflat, \diamond, \sflat, \sflat)}$ where
+$\sflat\,\diamond\,\sflat = \sflat$ and $\star=\{\emptyset\}$. That is, there is a single shape
+$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
+Therefore, a coeffect annotation is drawn from $\C^{\sflat}$ which is iso\-morphic to $\C$
+and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
+\end{itemize}
+
+\noindent
+Using a shape with \emph{no} positions reduces our system to the simply-typed $\lambda$-calculus with no context annotations.
+Trees could be used to build a system akin to bunched typing \cite{substruct-bunched}.
+
+\paragraph{Coeffect algebra.}
+The coeffect calculus annotates judgments with 
+shape-indexed (or, \emph{shaped}) coeffects.
+The \emph{\aclrd{coeffect
+    algebra}} structure combines a coeffect scalar and coeffect shape structures
+to define shaped coeffects and operations for combining these.
+In Section~\ref{sec:intro-coeffects}, shaped coeffects were combined
+by the tensor $\times$ in structural examples and
+$\cup$ in the implicit parameters example. To capture the examples so far and
+those described previously~\cite{petricek2013coeffects}, we distinguish two
+operators for combining shaped coeffects.
+
+\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
+
+\begin{definition}
+  Given a \cclrd{coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero,
+  \cleq)$ and a \sclrd{coeffect shape} $(\SH, \mksh{\textnormal{--}}, \stimes,
+  \sempty, \sunit)$ a \emph{\aclrd{coeffect algebra}} extends the two
+  structures with $(\amerge, \asplit, \azero)$ where $\azero \in
+  \C^{\sempty}$ is a coeffect annotation for the empty context and
+  $\amerge, \asplit$ are families of operations that combine coeffect
+  annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m}
+  \in \SH$:
+\[
+\asplit_{\sclrd{m},\sclrd{n}}, \amerge_{\sclrd{m},\sclrd{n}} ~:~ \C^{\sclrd{m}} \times \C^{\sclrd{n}} \rightarrow \C^{\sclrd{m}\,\stimes\,\sclrd{n}}
+\]
+\end{definition}
+
+\noindent
+A coeffect algebra induces the following two additional operations:
+\[
+\begin{array}{rclrcl}
+\alift{\textnormal{--}} &\narrow{:}& \C \rightarrow \C^{\sunit}          \quad\quad
+   & \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{m}}  
+\\ 
+\alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}  
+   & \cclrd{r}\ \aseq\ \aclrd{S} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\ \cseq\ (\aclrd{S}(\sclrd{s}))
+\end{array}
+\]
+$\alift{\textnormal{--}}$ lifts a scalar coeffect 
+to a shaped coeffect indexed by the singleton context shape. 
+The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a vector by scalar.
+As we always use lower-case for scalars and upper-case for vectors, using the same
+symbol is not ambiguous. We also tend to omit the subscript
+$\sclrd{m}$ and write just $\aseq$.
+
+The operators $\asplit$ and $\amerge$ combine shaped
+coeffects associated with two contexts. For example,
+assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects 
+$\aclrd{R} \in \C^{\sclrd{m}}$ and $\aclrd{S} \in \C^{\sclrd{n}}$. In
+the structural system, the context shapes $\sclrd{m}, \sclrd{n}$
+denote the number of variables in the two contexts. The combined
+context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and
+the combined coeffects $\aclrd{R} \,\amerge\, \aclrd{S}, \aclrd{R} \,\asplit\, \aclrd{S} \in \C^{\sclrd{m} \stimes \sclrd{n}}$
+are indexed by that shape.
+
+For structural coeffect systems such as bounded reuse, both $\asplit$ and $\amerge$ are
+just the tensor product $\times$ of vectors. However, we need to distinguish them for flat coeffect 
+systems discussed later.
+
+The difference is explained by the semantics (Section~\ref{sec:semantics}), where
+$\aclrd{R}\,\amerge\,\aclrd{S}$ is an annotation of the codomain of 
+a morphism that merges the capabilities provided by two contexts (in the syntactic 
+reading, splits the context requirements); $\aclrd{R}\,\asplit\,\aclrd{S}$ is an annotation of
+the domain of a morphism that splits the capabilities of a single context into two parts (in the
+syntactic reading, merges their context requi\-rements). Syntactically, this means that we always use
+$\amerge$ in the rule \emph{assumptions} and $\asplit$ in \emph{conclusions}. For now, it suffices to use the 
+bounded reuse intuition and read the operations as tensor products.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+$\boxed{
+\coctx{\Gamma}{R} \vdash e : \tau
+}$
+\[
+\begin{array}{c}
+\tyrule{const}{\;}
+  { \coctx{ () }{ \azero } \vdash c : \iota }
+\quad\quad
+\tyrule{var}{\;}
+  { \coctx{ (x:\tau) }{ \aunit } \vdash x : \tau }
+\\[1.8em]
+\tyrule{abs}
+  { \coctx{\Gamma, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{s}}} \vdash e : \tau }
+  { \coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau }
+\\[1.8em]
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \tau \xrightarrow{\cclrd{t}} \sigma &
+    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \tau }
+  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash e_1~e_2 : \sigma } 
+\\[1.8em]
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{S}} \vdash e_1 : \sigma & 
+    \coctx{\Gamma_2, x:\sigma}{\aclrd{R}\; \amerge\; \alift{\cclrd{t}}} \vdash e_2 : \tau  }
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R}\; \asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau }
+\\[1.8em]
+\tyrule{ctx}
+  { \coctx{\Gamma}{\aclrd{R}} \vdash e : \tau & \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma'}{\aclrd{R'}} }{\theta}}
+  { \coctx{\Gamma'}{\aclrd{R'}} \vdash \theta e : \tau }
+\end{array}
+\]
+
+$\boxed{
+\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}
+}$
+
+\[
+\tystmt{weak}
+  { \quad\quad 
+    \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x:\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}
+    \quad\quad\quad }
+\]
+\[ 
+\tystmt{exch}
+  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau, y:\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
+               { \quad \coctx{\Gamma_1, y:\sigma, x:\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }{\emptyset} }
+\]
+\[
+\tystmt{contr}
+  { \ctxtransnl{ \coctx{\Gamma_1, y:\tau, z:\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
+               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]}}
+\]
+\[
+\tystmt{sub}
+  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau,\Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \aclrd{T} } }
+               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s'}} \,\asplit\, \aclrd{T}} }{\emptyset} }
+\quad(\cclrd{s} \;\cleq\; \cclrd{s'} )                 
+\]
+\caption{The general coeffect calculus}
+\vspace{-1em}
+\label{fig:coeffects-general}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{General coeffect type system}
+
+In the previous section, we developed an algebraic structure capable of
+capturing different concrete context-dependent properties discussed in
+Section~\ref{sec:intro-coeffects}. Now, we use the structure to
+define the general coeffect calculus in
+Figure~\ref{fig:coeffects-general}.
+
+Coeffect annotations on free-variable contexts
+are shape-indexed coeffects $\aclrd{R}, \aclrd{S}, \aclrd{T} \in \C^\SH$ 
+and function types are annotated with
+coeffects scalars $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.  The
+rules manipulate coeffect annotations using the operations provided by
+coeffect algebra $(\amerge, \asplit, \azero)$ and the derived
+constructs $\alift{\textnormal{--}}$ and $\aseq$. Free-variable contexts $\Gamma$
+are treated as vectors modulo duplicate use of variables 
+-- the associativity is built-in. The order of
+variables matters, but can be changed using the structural rules. To
+make the system easier to follow, structural rules are expressed
+using a separate judgment.
+
+\paragraph{Typing rules.} 
+Constants (\emph{const}) and variable access (\emph{var}) annotate the context with special
+values. Empty unused context is annotated with $\azero
+\in \C^\sempty$, while singleton context 
+is annotated with $\aunit \in \C^\sunit$.  Note that the shapes
+$\sempty, \sunit$ match the shape of the variable contexts.
+
+Lambda abstraction \emph{splits} the context requirements using $\amerge$ into a coeffect 
+$\aclrd{R}$ and a coeffect $\alift{\cclrd{s}}$ of a shape $\sunit$ (semantically, it \emph{merges}
+capabilities provided by the declaration-site and call-site contexts). In structural systems such
+as bounded reuse, $\asplit$ is not symmetric and so this gives us a coeffect associated with the bound variable.
+
+The (\emph{app}) rule follows the patterns seen
+earlier -- it uses the scalar-vector multiplication ($\cclrd{t}\,\aseq\,\aclrd{S}$) on 
+coeffects associated with $\Gamma_2$.
+Using the syntactic reading, it then \emph{merges} context requirements for $\Gamma_1$ and $\Gamma_2$.
+In the dual semantic reading,
+it \emph{splits} the provided context into two parts passed to the sub-expressions.
+
+The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
+Syntactically, the context requirements are first split using $\amerge$ and then re-combined
+using $\asplit$.
+
+\paragraph{Structural rules.}
+The coeffect-annotated context can be transformed using structural rules that are
+not syntax-directed. These are captured by (\emph{ctx}), which uses a helper judgment representing
+context transformations $\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}$.
+The rule models that a context used in the rule conclusion $\coctx{\Gamma'}{R'}$ can be transformed
+to a context required by the assumptions $\coctx{\Gamma}{R}$ (using the semantic bottom-up reading).
+In the rule, $\theta$ is a variable substitution generated by the transformation, which is 
+used in the (\emph{contr}) rule.
+
+Exchange and contraction decompose and reconstruct coeffect annotations 
+using $\amerge_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\asplit_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
+The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \mksh{\Gamma_1}$
+and $\sclrd{n} = \mksh{\Gamma_2}$.
+
+The (\emph{weak}) rule drops an ignored variable annotated with $\aweak$ (compare with (\emph{var})
+annotated using $\aunit$). The (\emph{exch}) rule switches the values 
+while (\emph{contr}) combines them using $\cpar$ to represent sharing of the context.
+Finally, (\emph{sub}) represents sub-coeffecting that can be applied (point-wise) to any
+individual coeffect.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffects}
+\label{sec:coeffects-structural}
+
+The coeffect system uses a general notion of context shape, but
+it has been designed with structural and flat systems in
+mind. The structural system is new in this paper and so we
+look at it first.
+
+Recall the coeffect shapes that characterise structural systems: the shape is formed by natural
+numbers (with addition) modelling the number of variables in the
+context. The coeffect algebra is therefore formed by the free monoid
+(vectors) over a coeffect scalar. This means that the system keeps a vector
+of basic coeffect annotations -- one for each variable. An empty
+context (\eg{}, in the (\emph{const}) rule) is annotated with an empty
+vector.
+
+\begin{definition}
+Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural
+coeffect system} has: 
+\begin{itemize}{}
+\item[--] Coeffect shape $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$ formed by natural numbers
+\item[--] Coeffect algebra $\aclrd{(\fcons, \fcons, \fempty)}$ where $\fcons$ and $\fempty$ are
+  shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
+  That is $\fcons : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n + m}}$ appends vectors
+  (lists) and $\fempty : \C^{\sclrd{0}}$ represents empty vectors (lists)
+\end{itemize}
+\end{definition}
+
+\noindent
+The definition is valid since the shape operations form a monoid
+$\sclrd{(\mathbb{N}, +, 0)}$ and $\mksh{\textnormal{--}}$ (calculating the length of
+a list) is a monoid homomorphism from the free monoid to the monoid of
+shapes.
+
+\paragraph{Examples.} 
+Defining a concrete structural coeffect system is easy, we just provide the coeffect scalar structure and the rest is free.
+
+\begin{itemize}
+\item
+To recreate the system for bounded reuse, we use coeffect scalars formed by
+$(\mathbb{N}, \ast, +, 1, 0, \leq)$. As in the system of Figure~\ref{fig:bounded-coeff},
+\emph{used} variables are annotated with $1$ and \emph{unused} with $0$. Contraction adds the number
+of uses via $+$ and application (sequencing) multiplies the uses.
+
+\item
+\emph{Dataflow} uses natural numbers (of past values), but differently: $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$.
+Variables are initially annotated with $0$ (and can be incremented using the \kvd{prev} keyword).
+Annotations of a shared variable are combined by taking maximum (of past values needed) and
+sequencing uses $+$.
+
+\item
+Another use of the system is to track \emph{variable liveness}. The annotations are formed by
+$\C=\{ \mathsf{D}, \mathsf{L} \}$ where $\mathsf{L}$ represents a \emph{live} (used) variable
+and $\mathsf{D}$ represents a \emph{dead} (unused) variable. The scalars are given as
+$(\C, \sqcap, \sqcup, \mathsf{L}, \mathsf{D}, \sqsubseteq)$.
+
+In sequential composition ($\sqcap$), a variable is live only if it is required by both of the 
+computations ($\mathsf{L} \sqcap \mathsf{L} = \mathsf{L}$), otherwise it is marked as 
+dead ($\mathsf{D}$). A computation is not evaluated if its result is not needed.
+A shared variable ($\sqcup$) is live if either of the uses is live 
+($\mathsf{D} \sqcup \mathsf{D} = \mathsf{D}$, otherwise $\mathsf{L}$).
+\end{itemize}
+
+\noindent
+Structural liveness is a practically useful, precise version of an example from our earlier
+work, which was a flat system overapproximating liveness to the entire
+context~\cite{petricek2013coeffects}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Flat coeffects}
+\label{sec:coeffects-flat}
+
+The same general coeffect system can be used to define systems that track whole-context coeffects
+as in the implicit parameters example 
+(Section~\ref{sec:intro-coeffects-impl}). Flat coeffect systems are
+characterised by a singleton set of shapes, such as ${ \{ \sflat \} }$. In
+this setting, the context annotations $\C^{ \sflat }$ are
+equivalent to coeffect scalars $\C$.
+
+In addition to the coeffect scalar structure, we also need to define $\asplit$ and $\amerge$.
+Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\asplit$ (merging of shaped 
+coeffect annotations). However, the $\amerge$ operation needs to be provided explicitly.
+
+\begin{definition}
+Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary 
+$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, we define:
+\begin{itemize}{}
+\item[--] Flat coeffect shape $\sclrd{(\{ \sflat \}, \emph{const} \, \sflat, \diamond, \sflat, \sflat)}$ where 
+  $\sflat \diamond \sflat = \sflat$
+\item[--] Flat coeffect algebra $(\cclrd{\wedge}, \cpar, \czero)$, \ie{}, the $\asplit = \cpar$ and $\azero = 
+\czero$ with the additional binary operation $\amerge = \wedge$. 
+\end{itemize}
+\end{definition}
+
+\noindent
+The requirement $(\cclrd{r \wedge s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$ guarantees 
+exchange and contraction preserve the coeffect of the assumption in the conclusion. Thus, flat coeffect
+calculi do not require substructural-style rules.
+
+\paragraph{Examples.}
+Implicit parameters are the prime example of a flat coeffect system, but other examples
+include rebindable resources \cite{app-distributed-acute} and Haskell type classes \cite{orchard2013thesis}.
+
+In the \emph{implicit parameters} system (Section~\ref{sec:intro-coeffects-impl}), 
+the co\-effect scalars are sets of names with types 
+$\C = \mathcal{P}(\ident{Name} \times \ident{Types}) $. Variables
+are always annotated with $\emptyset$ and coeffects are combined or split using set union $\cup$. Thus
+the system is given by coeffect scalar structure $(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \cup, \emptyset, \emptyset, \subseteq)$
+with $\cclrd{\wedge} = \cup$.
+
+\begin{remark}
+We previously described flat systems for \emph{liveness} and \emph{dataflow}~\cite{petricek2013coeffects}.
+Turning a structural system to flat requires finding $\wedge$ that 
+underapproximates the capabilities of combined contexts. For dataflow, this is given by the \textsf{min} function as
+$\mathsf{min}(\cclrd{r},\cclrd{s}) \,\cleq\, \mathsf{max}(\cclrd{r},\cclrd{s})$.
+
+In flat \emph{dataflow}, we annotate the
+entire context with the maximal number of past elements required
+overall. We use the same coeffect scalars $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$
+as in the structural version, but with
+$\cclrd{\wedge} = \ident{min}$. Abstraction (which is the only rule using $\wedge$) becomes:
+%%
+\[
+\tyrule{abs}
+  { \coctx{\Gamma, x\!:\!\sigma}{\ident{min}(\cclrd{r}, \cclrd{s})} \vdash e : \tau }
+  { \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau }
+\]
+%%
+Both the declaration-site and the call-site need to provide at least
+the number of past values required by the body. The overapproximation means that both $\cclrd{r}$
+and $\cclrd{s}$ can be greater than actually required.
+For dataflow, we could annotate
+both contexts with the same coeffect,
+but that would require treating $\amerge$ as a partial function.
+\end{remark}
+
+% ==================================================================================================
+
+\section{Equational theory}
+\label{sec:syntax}
+\newcommand{\reduce}{\leadsto}
+
+Each of the concrete coeffect systems discussed in this paper has a different notion of 
+context-dependence, much like various effectful languages have different notions of effects
+(such as state or exceptions). However, there are equational properties that hold for all (or some) of
+the systems we consider.
+
+The equational theory in this section illuminates 
+the axioms of coeffect algebra and the semantics of the calculus.
+We discuss syntactic substitution as it can form the basis for
+reduction in a concrete operational semantics. 
+We consider structural and flat systems
+separately. This provides better insight into how the two systems work and
+differ. In particular, call-by-name evaluation is \emph{coeffect
+preserving} for all structural, but only some flat systems.
+
+The properties and proofs in this section are syntactic.
+In Section~\ref{sec:semantics-soundness} we show that
+our denotational model of the coeffect calculus is sound with respect
+to the equational theory here.
+
+We use standard syntactic substitution written as $\subst{e_1}{x}{e_2}$, 
+$\beta\textnormal{-reduction}$ and $\eta$-expansion, written as $\reduce_\beta$ and $\reduce_\eta$.
+Equality of terms $e_1$ and $e_2$, defined by a relation
+$\equiv$, requires the equality of their contexts, types, and 
+coeffects, written $\coctx{\Gamma}{\aclrd{R}} \vdash e_1 \equiv e_2 : \tau$.
+
+%\begin{definition}[Subject reduction]
+%Given reduction $e \reduce e'$ and $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$
+%then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$, \ie{}, $\reduce$ is 
+%\emph{type-and-coeffect preserving}. 
+%\label{def:pres}
+%\end{definition}
+
+%Definition~\eqref{def:pres} is also known as the \emph{subject reduction} property.
+
+\subsection{Structural coeffect systems}
+\label{prop:structural}
+
+For structural coeffect systems, recall that coeffects are vectors
+with $\asplit = \amerge = \cons$ (vector concatenation) and $\azero =
+\langle{}\rangle$ (the empty vector), thus coeffect annotations
+comprise the \emph{free monoid} \ie, lists over the coeffect scalars
+(although we continue using the vector terminology). We first show
+substitution:
+
+\begin{lemma}[Substitution lemma] 
+\label{thm:structural-subst}
+In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
+\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
+\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\vspace{-1.5em}
+\label{lemm:structural-subst}
+\end{lemma}
+\begin{proof}
+By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
+and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
+\end{proof}
+
+\noindent
+Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
+$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
+$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
+the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
+Using the substitution lemma, we can demonstrate $\beta$-equality:
+%
+\begin{equation*}
+\inference{
+  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
+             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
+           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
+  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
+\equiv
+\subst{e_1}{x}{e_2} : \tau}
+\end{equation*}
+%%
+\noindent
+As a result, $\beta$-reduction preserves the type and coeffects of a term.
+This gives the following subject reduction property:
+
+\begin{theorem}[Subject reduction]
+In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
+$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
+\end{proof}
+
+\noindent
+Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
+\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
+\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
+and application does not eliminate too much.
+%
+\begin{equation*}
+\inference{
+  \inference 
+    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
+     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
+    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
+: \sigma \xrightarrow{\cclrd{s}} \tau}
+\end{equation*}
+%
+The last step uses the fact that 
+$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
+arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
+
+This highlights another difference between coeffects and effects, as $\eta$-equality does
+not hold for many notions of effect. For example, in a language with output
+effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
+its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
+hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
+allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
+$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
+the exact immediate requirements of $\lambda x . e x$ to match those of $e$.
+
+\subsection{Flat coeffect systems}
+
+The equational theory for flat coeffect systems is somewhat similar to effect systems
+where (co)effects are not linked to individual variables. In effectful languages,
+substituting an effectful computation for $y$ in $\lambda x.y$ changes the latent effect 
+associated with the function.
+
+Similarly, for some of the flat coeffect systems, substituting a context-dependent computation
+for $y$ in $\lambda x.y$ adds latent context requirements to the function type. However, this is not
+the case for \emph{all} flat coeffect systems -- for example, call-by-name reduction preserves
+types and coeffects for the implicit parameters system (which makes it a suitable model for Haskell).
+For other systems, we first briefly consider call-by-value reduction.
+
+\paragraph{What is a value?}
+The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
+As discussed earlier, a function $(\lambda x.e)$ is not necessarily a value in coeffect calculi, 
+because it may not delay all context requirements of $e$. Thus we say that $e$ is a value if it
+has no immediate context requirements.
+
+\begin{definition} 
+An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
+where $\aclrd{\mathsf{VAL}} : \C^{\mksh{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
+always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
+\end{definition}
+
+\paragraph{Call-by-value.}
+In call-by-value, a pure \emph{value} is substituted for a variable. 
+The right-hand side of an application is evaluated to a value before the
+$\beta$-reduction, but the discharging of coeffects prior to substitution is
+different for each concrete coeffect system.
+
+Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
+as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
+merging of context-requirements, respectively.
+
+\begin{lemma}[Call-by-value substitution]
+\label{thm:flat-subst-cbv}
+In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+and the $\cclrd{\wedge}$ operator:
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\end{lemma}
+\begin{proof}
+By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
+with $\cunit$.
+\end{proof}
+
+\noindent
+Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
+the operational semantics must provide a way of partially evaluating a term 
+with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
+Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:
+
+\begin{theorem}[Call-by-value reduction]
+In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
+$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
+\end{proof}
+
+\paragraph{Call-by-name.}
+The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
+may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
+$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.
+
+% TODO
+%
+% The context requirements of the 
+% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
+% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
+% lemma should be $r \fmax (s \fseq r)$.
+
+We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
+$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
+Liveness analysis is an example of top-pointed coeffects
+as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.
+
+\begin{lemma}[Top-pointed substitution]
+\label{thm:cbn-substitution-top}
+In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+and the $\cclrd{\wedge}$ operator:
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\vspace{-1.5em}
+\end{lemma}
+\begin{proof}
+Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
+\end{proof}
+%
+\noindent
+As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
+use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).
+
+In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
+requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
+the context requirements arising from the substitution can be associated with the context
+$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
+requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
+(all three operators are set union) and allows the following substitution lemma:
+
+\begin{lemma}[Bottom-pointed substitution]
+\label{thm:cbn-substitution-bot}
+In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\vspace{-1.5em}
+\end{lemma}
+\begin{proof}
+By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
+the free-variable context.
+\end{proof}
+\noindent
+The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
+variant described here is flexible enough to let us always re-associate new context requirements with the 
+free-variable context.
+
+The two substitution lemmas show that the call-by-name evaluation strategy can be used for
+certain coeffect calculi, including liveness and implicit parameters.
+Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
+reduction, the following theorem holds:
+
+\begin{theorem}[Call-by-name reduction]
+\raggedright
+In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
+Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
+\end{proof}
+
+%
+% TODO: We could have a section about let binding covering the alternative CBV and the let binding
+% we included in the system, but we'd have to clean this up!
+%
+%\subsection{Syntactic let binding}
+%\begin{remark}
+%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
+%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
+%calculus of Section~\ref{}.
+%%%
+%\begin{align*}
+%{\footnotesize{\trule{let}}}
+%\dfrac
+%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
+%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
+%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
+%\end{align*}
+%%%
+%Therefore, for example, we see that \emph{let} has the coeffects
+%as substitution, for example:
+%%%
+%\begin{align*}
+%& {\footnotesize{\trule{let}}}
+%\dfrac
+%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
+%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
+%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
+%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
+%\end{align*}
+%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
+%the judgment:
+%%%
+%$$
+%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
+%$$
+%%%
+%\end{remark}
+%
+% 
+%\paragraph{Syntactical let binding}
+%Has the same typing derivation as abstraction/application
+%
+%\[
+%\inference
+% {\inference
+%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
+%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
+%  &
+%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
+%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
+%\]
+%
+%We can extend the CBV behaviour to \emph{let}, where the bound term
+%is evaluated to a pure value first, discharging any contextual
+%requirements before substitution into the body.
+%
+%\begin{lemma}[CBV \emph{let}]
+%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
+%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
+%: \tau$.
+%\end{lemma}
+
+% ==================================================================================================
+
+\section{Semantics}
+\label{sec:semantics}
+
+\newcommand{\Di}{\F}
+
+Coeffects provide a unified notion of context-dependence. In the 
+previous sections, we used this to define a unified coeffect calculus.
+We now define a unified (categorical) semantics for the 
+coeffect calculus. The semantics can be instantiated for different 
+notions of context dependence and thus can model a wide range of
+context-aware languages (both for flat and structural systems).
+
+We relate the semantics to the equational theory and show that it is
+sound with respect to term equality. For a variant
+of the the flat system, a similar result has already been shown in the 
+second author's dissertation~\cite{orchard2013thesis}.
+The semantics is introduced in pieces:
+
+\begin{itemize}
+\item
+Section~\ref{sem:interp}
+describes the range and domain (signature) of the interpretation $\interp{-}$, 
+gives the interpretations for types and free-variable contexts (in
+flat and structural systems), and defines the signature of
+functors $\D$ which encode contexts.
+
+\item
+The first part of the semantics (Section~\ref{sem:seq}) defines \emph{sequential
+composition} of context-dependent computations via \emph{indexed
+  comonads} (introduced briefly in our previous
+work~\cite{petricek2013coeffects}) and the \emph{indexed structural
+  comonad} structure (new here).
+
+\item
+More structure is needed for application and abstraction. 
+Section~\ref{sem:monoidal} defines
+indexed monoidal operations for splitting and merging contexts.
+Concrete structures are given throughout for the semantics of the
+structural bounded reuse and flat implicit parameter systems.
+
+\item
+Section~\ref{sem:together} puts the pieces together, defining the
+semantics of the coeffect calculus along with structural rules.  
+The semantics is illustrated by executing an 
+ example bounded-reuse program in 
+the semantics (Example~\ref{exm:run}).% and the concrete structures for
+%the semantics of structural dataflow (Example~\ref{exm:dataflow-full}).
+
+\item
+Section~\ref{sec:semantics-soundness} shows our
+semantics sound with respect to the syntactic equational theory of
+Section~\ref{sec:syntax}. This uses the derivation of 
+categorical structures for the semantics as \emph{lax homomorphisms} between
+structure on the coeffect category $\Icat$ and the base $\mathbb{C}$.
+
+\end{itemize}
+
+%
+% TP: Removed \paragraph just because I think this looks nicer with the list
+%
+\noindent 
+In this section, 
+$\mathbb{C}, \mathbb{D}, \mathbb{I}$ range over categories.  The
+objects of a category $\mathbb{C}$ are written
+$\emph{obj}(\mathbb{C})$.  The category of functors between
+$\mathbb{C}$ and $\mathbb{D}$ is written $[\mathbb{C}, \mathbb{D}]$.
+Exponential objects, representing function types in our model, 
+are written in two ways, either $B^A$ or $A \Rightarrow B$.
+
+\subsection{Interpreting contexts and judgments} 
+\label{sem:interp}
+
+The semantics is parameterised by a coeffect algebra, with scalar coeffects $(\C,
+\cseq, \cpar, \cunit, \czero, \cleq)$, coeffect shape $(\SH, \mksh{-},
+\stimes, \sempty, \sunit)$, and $(\amerge, \asplit, \azero)$. An
+interpretation $\interp{-}$ is given to types, free-variable contexts,
+and type and coeffect judgments, with a base Cartesian-closed category
+$\mathbb{C}$ for denotations and a category $\Icat$ of scalar coeffects, where $\emph{obj}(\Icat)
+= \C$. Since $\mathbb{C}$ is Cartesian-closed, we use the
+$\lambda$-calculus as the syntax for giving concrete definitions.
+
+The interpretation $\interp{-}$ is parameterised by categorical 
+structures which model a particular notion of context. %We show  
+%instances of these structures for bounded reuse and dataflow later. 
+The interpretation of free-variable contexts depends on shape, for
+which we give concrete definitions for flat and structural shapes.
+
+\paragraph{Interpreting judgments.}
+type and coeffect judgments are interpreted (given denotations) 
+as morphisms in $\mathbb{C}$, of the form:
+%
+\begin{align*}
+\interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \D^{\mksh{\Gamma}}_{\aclrd{R}}
+\interp{\Gamma} \rightarrow \interp{\tau}
+\end{align*}
+%%
+The interpretation is a morphism from an interpretation of the context $\Gamma$
+to the interpretation of the result. The functor $\D^{\mksh{\Gamma}}_{\aclrd{R}}$ over the
+context encodes the semantic notion of context and is indexed by the free-variable 
+context shape $\mksh{\Gamma}$ and coeffect annotation $\aclrd{R}$.
+
+The structure $\D$ can be thought of as a coproduct of functors $\D^{\sclrd{n}}$ for every
+possible shape $\sclrd{n} \in \SH$ of free-variable context:
+%%
+\begin{equation*}
+\D : \Sigma_{\sclrd{n} : \SH}. \D^{\sclrd{n}}
+\quad \text{where} \quad
+\D^{\sclrd{n}} : \Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}]
+\end{equation*}
+% %
+For a fixed context shape $\sclrd{n}$ the functor $\D^{\sclrd{n}} :
+\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}]$ 
+maps an $\sclrd{n}$-indexed coeffect (think positions) to a functor from a
+context $\mathbb{C}^n$ to an object in $\mathbb{C}$. That is, given a coeffect
+annotation (matching the shape of the context), we get a 
+functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$. 
+%
+
+From a programming 
+perspective, this functor defines a data structure that models the additional
+context provided to the program. The shape of this data structure depends on the coeffect
+annotation $\Icat^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
+the number of values needed for each variable and the functor will be 
+formed by lists of length matching the required number.
+
+%
+% TP: I think this is easier to understand in the later paragraph
+%
+% Thus free-variable contexts are interpreted
+% $\interp{\Gamma} \in \mathbb{C}^{\mksh{\Gamma}}$ and
+
+%
+% TP: I did not get the following - but I think I tried to say the same thing above...?
+%
+% $\D$ couples the
+% shape of the coeffect with the shape of the free-variable context.
+%Below, the interpretation of free-variable contexts is defined for
+%classes of flat and structural system.
+
+\paragraph{Types.} 
+Types are interpreted as objects of $\mathbb{C}$, that is 
+$\interp{\tau} : \emph{obj}(\mathbb{C})$
+ where function types have the interpretation as exponents: 
+%
+\begin{align*}
+  \interp{\sigma \xrightarrow{\cclrd{r}} \tau} = \D^{\sunit}_{\cclrd{r}} \interp{\sigma}
+  \Rightarrow \interp{\tau}
+\end{align*}
+%%
+The parameter of a function is wrapped by 
+a functor $\D^{\sunit}_{\cclrd{r}}$ that defines a context with singleton shape
+$\sunit$, matching the single value that it contains.
+This interpretation is shared by all coeffect calculi.
+
+%
+% TP [Deleted] "although it may be extended to other type constructors."
+% (I did not understand what you mean here)
+%
+
+\paragraph{Free-variable contexts.} 
+As described above, free-variable contexts $\Gamma$ are given an
+interpretation as objects in $\mathbb{C}^{\mksh{\Gamma}}$. Thus,
+the interpretation of contexts is shape dependent.
+
+%\tpnote{This is tricky but clever! Tried to add explanation}
+We define $\interp{-}$ on free-variable context for structural and flat
+systems. For flat systems, there is only a single shape, so the interpretation
+is a product type inside the Cartesian closed category $\mathbb{C}$. For 
+structural systems, the shape matches the number of variables and so the model 
+is a value in the product category $\mathbb{C} \times \ldots \times \mathbb{C}$.
+
+\paragraph{\textnormal{\emph{Flat coeffects.}}} 
+Recall that $\SH = \{\sflat\}$ and $\mksh{\Gamma} = \sflat$.
+Since the set of positions $\sflat$ is a singleton, then $\mathbb{C}^{\sflat}$
+is isomorphic to $\mathbb{C}$.
+%
+% TP: I see why the following bit is important, but I think we can assume people
+% will just believe us here and make this a bit simpler
+%
+% as categories formed from a singleton set are terminal objects in \textbf{Cat}. 
+%
+Therefore $\interp{\Gamma} : \emph{obj}(\mathbb{C})$, which is defined as:
+%%
+% \begin{align*}
+% \interp{x : \tau, \Gamma'} = \interp{\tau} \times \interp{\Gamma'}
+% \qquad\qquad
+% \interp{\cdot} = 1 
+% \end{align*}
+%
+% TP: Is there anything wrong with the following? I find it a bit simpler...
+%
+\begin{align*}
+\interp{x_1 : \tau_1, \ldots, x_n : \tau_n} = \interp{\tau_1} \times \ldots \times \interp{\tau_n}
+\end{align*}
+%
+Denotations of typing judgments in a flat coeffect system are thus of the form 
+(where $\aclrd{R} \in \Icat$): 
+%%
+$$\interp{\coctx{x_1 : \tau_1, .., 
+x_n : \tau_n}{\aclrd{R}} \vdash e : \tau} : \D^{\sflat}_{\aclrd{R}}(\interp{\tau_1} \times ... \times \interp{\tau_n})
+\rightarrow \interp{\tau}
+$$
+
+\paragraph{\textnormal{\emph{Structural coeffects.}}}
+Recall that $\SH = \mathbb{N}$ and $\mksh{\Gamma} = |\Gamma|$
+(number of free variables), thus 
+$\interp{\Gamma} : \emph{obj}(\mathbb{C}^{|\Gamma|})$. This is defined similarly 
+to the above, but instead of using products in $\mathbb{C}$, %and the terminal object
+%$1 : \mathbb{C}$, 
+we use the product of categories. % (\ie{}, $\times$ 
+% in \textbf{Cat}) with the singleton category $1 \in \mathbf{Cat}$. 
+%
+% TP: If you think that we should say how empty products are treated
+% then I would just say "Empty contexts are modelled using the singleton category"
+% (those who would ask the question will understand this and it does not add
+% too much fancy notation to the explanation)
+%
+Thus, denotations have the form: 
+$$
+\interp{\coctx{x_1 : \tau_1, \ldots, x_n : \tau_n}{\aclrd{R}} \vdash e : \tau} : \D^n_{\aclrd{R}}
+(\interp{\tau_1}, \ldots, \interp{\tau_n}) \rightarrow \interp{\tau}
+$$
+where $|R| = n$ and we use commas (instead of $\times$) to denote the product
+of categories. %in $\mathbf{Cat}$. 
+This means that $\D^n : \Icat^n \rightarrow [\mathbb{C}^n,
+\mathbb{C}]$ is a functor between an $n$-length vector of coeffects
+indices and an \emph{n}-ary endofunctor.
+As mentioned, the key
+difference between the flat and structural interpretations of
+free-variable contexts is that flat uses products of objects in
+$\mathbb{C}$ and the structural uses products of $\mathbb{C}$ in 
+the category of categories. %\textbf{Cat}.
+
+\newcommand{\BLL}{\mathsf{B}}
+
+\begin{example}[Bounded reuse] Recall
+bounded reuse has coeffect scalars $\C = \mathbb{N}$ and shapes $\SH = \mathbb{N}$.
+We model contexts by replicating the value of each variable so 
+there is a value for each use. This matches the model used by 
+Girard \emph{et al.}~\cite{girard1992bounded}. Contexts are described by 
+ $\BLL : \Sigma_{\sclrd{n} : \mathbb{N}}. (\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$, 
+where for $\aclrd{R} = \langle{\cclrd{r_1}, \ldots, \cclrd{r_n}}\rangle$: 
+%%
+%\tpnote{\\I do not understand what the $a_i$ functions mean (in the morphism mapping part)?}
+\begin{align*}
+\BLL^n_{\aclrd{R}} (A_1, \ldots, A_n) & = A_1^{\cclrd{r_1}} \times \ldots \times A_n^{\cclrd{r_n}} \\
+\BLL^n_{\aclrd{R}} (f_1, \;\ldots\;, f_n) & = \lambda \langle{a_1, \ldots, a_n}\rangle . \langle{(f_1 \circ a_1), \ldots, (f_n \circ a_n)}\rangle
+\end{align*}
+%
+Thus each object in the free-variable context $A_i$ is indexed by its associated coeffect $\cclrd{r_i}$.
+For the morphism mapping, $f_i : A_i \rightarrow B_i$ and $a_i : {A_i}^{\cclrd{r_i}}$ thus $(f_i \circ a_i) : {B_i}^{\cclrd{r_i}}$. 
+The exponent ${A_i}^{\cclrd{r_i}}$ can be read as a product of $\cclrd{r_i}$ copies of $A_i$, \eg:
+$$\BLL^3_{1,0,2} (A, B, C) = A^1 \times B^0 \times C^2 = (A) \times 1 \times (C \times C)$$
+\label{exm:bll1}
+\end{example}
+
+\vspace{-2em}
+\newcommand{\Imp}{\mathsf{I}}
+\begin{example}[Implicit paramters]
+Recall the implicit parameter calculus with scalar coeffects as 
+sets of names paired with types $\C = \mathcal{P}(\ident{Name} \times \ident{Types})$ 
+and flat shape with singleton $\SH = {\{\sflat\}}$.
+
+%\tpnote{Could you say, let $\mathsf{Impl}=\textbf{Set}$ be a category of pairs of vars and values and use $\mathsf{Impl}$ throughout?\\ I find Set non-intuitive }
+Its contexts are defined by 
+$\Imp^\star :  \Sigma_{\sclrd{n} : \{\star\}}, (\Icat^{\sclrd{n}} \rightarrow [\mathbf{Set}^{\sclrd{n}}, \mathbf{Set}])$,
+which is equivalent to $\Icat \rightarrow [\mathbf{Set}, \mathbf{Set}]$) and defined as follows:
+%%
+\begin{align*}
+\Imp^{\sflat}_{\aclrd{R}} A = A \times \interp{\aclrd{R}} \qquad\qquad
+\Imp^{\sflat}_{\aclrd{R}} f = \lambda (a, r) . (f\ a, r)
+\end{align*}
+%%
+The interpretation $\interp{\aclrd{R}}$ maps a set of variable-type
+pairs to an object representing a set of variables-values pairs in $\mathbf{Set}$.
+\label{exm:impl1}
+\end{example}
+
+\subsection{Sequential composition}
+\label{sem:seq}
+
+Following the usual categorical semantics approach, we require a
+notion of sequential composition for our denotations.  We show first a
+special case for $\D^\sunit$, where $\Icat^\sunit = \Icat$
+and $\mathbb{C}^\sunit = \mathbb{C}$ in both flat and structural
+systems\footnote{since $\sunit = 1$ in structural and $\sunit \cong
+  1$ in flat, \ie{}, $\star$ is isomorphic to 1} and thus $\D^\sunit :
+\Icat \rightarrow [\mathbb{C}, \mathbb{C}]$. Composition of  
+morphisms $f : \D^\sunit_S A \rightarrow B$ and $g : \D^\sunit_R B
+\rightarrow C$ is defined by an \emph{indexed comonad}  
+(which we previously introduced briefly~\cite{petricek2013coeffects}).
+
+\begin{definition}
+An \emph{indexed comonad} comprises a strict monoidal category
+$(\mathbb{I}, \bullet, I)$ and a functor $\Di : \mathbb{I}
+\rightarrow [\mathbb{C}, \mathbb{C}]$ with two natural transformations
+(where we write $(\Di\,R)\,A$ as $\Di_R A$):
+%\tpnote{Added brackets in $(\Di\,R)\,A$ -- otherwise I was confused}
+%%
+\begin{align*}
+(\delta_{X,Y})_A : \Di_{(X \bullet Y)} A \rightarrow \Di_X (\Di_Y A) \qquad
+(\varepsilon_I)_A : \Di_I A \rightarrow A
+%\delta_{X,Y} : \Di (X \bullet Y) \xrightarrow{.} \Di X \circ \Di Y \qquad
+%\varepsilon_I : \Di I \xrightarrow{.} 1_{\mathbb{C}}
+\end{align*}
+%%
+where $\delta$ is called \emph{comultiplication} and $\varepsilon$ is called \emph{counit}.
+We require indexed 
+analogues of the usual comonad axioms (\emph{cf.}~\cite{uustalu2008notions}): 
+%$$
+%\Di_R \varepsilon_I \circ \delta_{R,I} = {\varepsilon_I}_{\Di_R} \circ \delta_{I, R} = id \quad 
+%\delta_{R,S}\Di_T \delta_{R \bullet{} S, T} = \Di_R \delta_{S,T} \circ \delta_{R, S \bullet{} T}
+%$$ 
+%
+%\tpnote{I'd consider omitting the axioms (too much theory?) if we can refer to your thesis?}
+%\tpnote{[Read 2] OK - I commented them out - but feel free to revert!}
+\begin{equation*}
+\hspace{-1em}
+\xymatrix@R=1em{
+\Di_R \ar[r]^-{\delta_{R, I}} \ar@{=}[dr] \ar[d]_{\delta_{I,R}} \ar@{}[dr]^{\hspace{0.3em}\text{[C2]}} & \Di_R \Di_I \ar[d]^{\Di_R \varepsilon_I} \\
+\ar@{}[ur]_{\hspace{-1.8em}\text{[C1]}} \Di_I \Di_R \ar[r]_{\varepsilon_I \Di_R} & \Di_R
+}
+\hspace{0.6em}
+\xymatrix@R=1em{
+\ar@{}[dr]|{\text{[C3]}}
+\Di_{R \bullet{} S \bullet{} T} \ar[d]_{\delta_{R, S \bullet{} T}} \ar[r]^{\delta_{R \bullet{} S, T}} & \Di_{R \bullet{} S}  \Di_T \ar[d]^{\delta_{R,S}\Di_T} \\
+\Di_R \Di_{S \bullet{} T} \ar[r]_{\Di_R \delta_{S,T}} & \Di_R \Di_S \Di_T
+}
+\label{eq:indexed-comonads-axioms}
+\end{equation*}
+\end{definition}
+
+\noindent
+An indexed comonad $\Di : \mathbb{I} \rightarrow
+[\mathbb{C}, \mathbb{C}]$ induces a notion of composition
+for all $f : \Di_S A \rightarrow B, g : \Di_R B \rightarrow C$: 
+%
+\begin{equation*}
+ g \hat{\circ} f = g \circ \Di_R f \circ \delta_{R,S} \, : \, \Di_{R \bullet S} A \rightarrow B
+\end{equation*}
+%
+with the identity $\hat{\id}_A = (\varepsilon_I)_A : \Di_I A \rightarrow A$ for all $A$. Thus
+in\-dexed comonads induce a category which has the same objects as $\mathbb{C}$ and morphisms 
+  $\mathbb{C}_\Di(A, B) = \bigcup_{R \in \mathbb{I}} \mathbb{C}(\Di_R A, B)$.
+Note that an indexed comonad is not a family of (ordinary) comonads, because the identity
+is only defined for the functor $\Di_I$. %and the functor in the result of composition (and comultiplication) 
+%may be any $\Di_R$.
+
+Therefore, if $\D^\sunit$ is an indexed comonad, there is a notion of composition for
+denotations with a single coeffect index. % in our semantics.
+
+\begin{example}[Bounded reuse]
+$\BLL^\sunit_R$ (Example~\ref{exm:bll1}) has an indexed comonad structure, where
+the monoid $(\cclrd{\mathbb{N}}, \cclrd{\ast}, \cclrd{1})$ from the coeffect scalar for bounded reuse
+induces a monoidal category structure on $\Icat$ (with $1 : \Icat$ and the bifunctor
+$\ast : \Icat \times \Icat \rightarrow \Icat$), with operations:
+%
+%\begin{align*}
+%\begin{array}{l}
+%\delta^\sunit_{R,S} = \lambda \langle{a_1 ..., a_{R S}}\rangle . \\
+%\hspace{3.5em} \langle{~\langle{a_1 ..., a_S}\rangle, \langle{a_{S+1},..., a_{S+S}}\rangle}, \ldots \\
+%\hspace{4.5em} \;\;~\langle{a_{(R - 1)S + 1},..., a_{R S}}\rangle~~\rangle 
+%\end{array}
+%\quad\quad
+%\begin{array}{l}
+%\varepsilon^\sunit_{1} = \lambda \langle{a_1}\rangle . a_1  \\[0.5em] 
+%\\
+%\end{array}
+%\end{align*}
+\begin{align*}
+\begin{array}{rl}
+\varepsilon^\sunit_{1} & \hspace{-1em} = \lambda \langle{a_1}\rangle . a_1 \\
+\hspace{-0.5em}
+\delta^\sunit_{R,S} & \hspace{-1em} = \lambda \langle{a_1 ..., a_{R S}}\rangle . \\
+& \hspace{0.4em} \langle{\langle{a_1 ..., a_S}\rangle, \langle{a_{S+1},..., a_{S+S}}\rangle}, ..., \langle{a_{(R - 1)S + 1},..., a_{R S}}\rangle\rangle
+\end{array}
+\end{align*}
+%\tpnote{Does this explanation for couint / comult. make sense?}
+Indexed comonads essentially model single-variable contexts.
+%Recall that we only consider single-variable context.
+The counit requires a single copy of the value from the context. 
+The comultiplication splits $R$ times $S$ copies of a value into
+$R$ copies of a context where each context contains just $S$ copies of the value.
+\label{exm:bll2}
+\end{example}
+
+\begin{remark}
+A semantics for dataflow coeffects is similar to bounded reuse with
+$\D^n_R (A_1, \ldots, A_n) = (A_1 \times A_1^{R_1}) \times ... \times (A_n \times {A_n}^{R_n})$, \ie{},
+each free-variable has an extra value representing the ``current'' value. A dataflow indexed
+comonad is similar to the above but with additive rather than multiplicative behaviour. 
+\end{remark}
+
+\begin{example}[Implicit parameters]
+For the coeffect scalar monoid $\cclrd{(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \emptyset{})}$ 
+of implicit parameters, 
+$\Imp^\ast$ (Example~\ref{exm:impl1}) there is an indexed comonad structure, with operations: 
+%where induces a monoidal category structure on $\Icat$, with operations:
+%For a set $X$, with the monoidal category of $X$-subsets 
+%$(\mathcal{P}(X), \cup, \emptyset)$, 
+%then $\D : \mathcal{P}(X) \rightarrow [\mathbf{Set}, \mathbf{Set}]$ where
+% $\D_X A = A \times X$ and $\D_X f = f \times \id_X$ 
+%is an indexed comonad with:
+%
+\begin{align*}
+\varepsilon_{\emptyset} \, (a, \emptyset{}) \mapsto a 
+\qquad\qquad\quad~~~
+\delta_{R,S} (a, \gamma) \mapsto ((a, {\gamma|}_S ), {\gamma|}_R) %\\
+%\qquad \textbf{where}~~~{\gamma|}_R = \{(x, v) \mid (x, v) \in \gamma, (x, t) \in R \}
+%\delta_{R,S} (a, x) \mapsto ((a,  x-(R-S)), x-(S-R))
+\end{align*}
+%%`
+where ${\gamma|}_R = \{(x, v) \mid (x, v) \in \gamma, (x, t) \in R \}$ filters incoming implicit parameters to those
+variable-value pairs where the variable is in the coeffect $R$.
+%%
+%where $x : R \cup S$ and thus the $R$-only subset of $x$ is $x - (S - R)$
+%and the $S$-only subset of $x$ is $x - (R - S)$ (where ($-$) is set difference).
+\label{exm:impl2}
+\end{example}
+
+\indent
+These two examples (which are new here) provide composition for context-dependent
+computations indexed by coeffects in a flat calculus. For structural coeffects,
+we need to compose morphisms which have more than a single coeffect annotation. For this, we
+introduced the notion of \emph{structural indexed comonads}.
+
+\begin{definition}
+\newcommand{\sunitc}{\sunit} %{{\sclrd{1}}}
+%
+A \emph{structural indexed comonad} comprises a functor 
+$\D : \Sigma_{n : \SH}. (\mathbb{I}^n \rightarrow [\mathbb{C}^n, \mathbb{C}])$
+where $(\mathbb{I}, \bullet, I)$ is a strict monoidal category,  
+$\sunit \in \SH$ and 
+$\D^{\sunit} : \mathbb{I}^\sunit \rightarrow [\mathbb{C}^\sunit, \mathbb{C}]$
+ is an indexed comonad (with 
+$(\delta^\sunit_{X,Y})_A : \D^\sunit_{X \bullet Y} A \rightarrow \D^\sunit_{X} \D^\sunit_{Y} A$ 
+and 
+$(\varepsilon_I)_A : \D^\sunit_I A \rightarrow A$) and 
+a \emph{structural comultiplication} natural transformation:
+$$
+(\delta^{\sclrd{n}}_{r, S})_{A^n} : \D^{\sclrd{n}}_{r \bar{\bullet} S} A^n \rightarrow \D^\sunit_r \D^{\sclrd{n}}_S A^n
+$$ 
+%$$
+%(\delta^{\sclrd{n}}_{R, S})_{A_\sunitc, \ldots, A_n} : \D^{\sclrd{n}}_{R \bar{\bullet} S} (A_1, \ldots, A_n) \rightarrow %\D^\sunitc_R \D^{\sclrd{n}}_S (A_1, \ldots, A_n)
+%$$ 
+where $A^n \in \mathbb{C}^n$, $r \in \Icat$, $S \in \Icat^{\sclrd{n}}$ and
+ $\bbullet : \mathbb{I} \times \mathbb{I}^{\sclrd{n}} \rightarrow \mathbb{I}^{\sclrd{n}}$ is 
+ the monoid left action for $\bullet$ lifting scalar coeffects to shaped coeffects (\eg{}, 
+the scalar-vector version of $\bullet$).
+Analogous laws to monoid left actions for unitality and associativity hold for structural comultiplication:
+$$
+\varepsilon_I \circ \delta^{\sclrd{n}}_{I, r} = id 
+\,\qquad
+\D^\sunitc_r \delta^{\sclrd{n}}_{S,T} \circ \delta^{\sclrd{n}}_{r, S \bar{\bullet{}} T}
+= 
+\delta^\sunitc_{r,S}\D^{\sclrd{n}}_T \circ \delta^{\sclrd{n}}_{r \bar{\bullet{}} S, T}
+$$
+%Structural comultiplication satisfies 
+%$\varepsilon_I \circ \delta^{\sclrd{n}}_{I, R} = id$ 
+%, but
+%with appropriate indices.
+%
+%\tpnote{Totally \emph{not} sure if this is right -- but I \emph{believe} we could keep law details for MFPS}
+%
+%
+%%
+%\begin{align}
+%\hspace{-1em}
+%\xymatrix@R=1.4em{
+%\D^{\sclrd{n}}_{I \bar{\bullet} R} \ar[r]^-{\delta^{\sclrd{n}}_{I,R}} \ar@{=}[dr] \ar@{}[dr]^{\hspace{0.3em}\text{[SC1]}} & \D^\sunitc_I \D^{\sclrd{n}}_R \ar[d]^{\varepsilon_I} \\
+% & \D_R
+%}
+%\;
+%\xymatrix@R=1em{
+%\ar@{}[dr]|{\text{[SC2]}}
+%\D^{\sclrd{n}}_{R \bar{\bullet{}} S \bar{\bullet{}} T} \ar[d]_{\delta^{\sclrd{n}}_{R, S \bar{\bullet{}} T}} \ar[r]^{\delta^{\sclrd{n}}_{R \bar{\bullet{}} S, T}} %& \D^\sunitc_{R \bullet{} S}  \D^{\sclrd{n}}_T \ar[d]^{\delta^\sunitc_{R,S}\D^{\sclrd{n}}_T} \\
+%\D^\sunitc_R \D^{\sclrd{n}}_{S \bar{\bullet{}} T} \ar[r]_{\D^\sunitc_R \delta^{\sclrd{n}}_{S,T}} & \D^\sunitc_R \D^\sunitc_S \D^{\sclrd{n}}_T
+%}
+%\vspace{-1em}
+%\label{eq:structural-indexed-comonad-axioms}
+%\end{align}
+%for unit and associativity respectively 
+Note the indexed comonad comultiplication $\delta^\sunitc$ for associativity. 
+\label{def:structural-indexed-comonad}
+\end{definition}
+
+\noindent
+Structural indexed comonads provide composition for morphisms
+%$f : \D^{\sclrd{n}}_{S} (A_1 \times \ldots \times A_n) \rightarrow B$ and $g : \D^\sunit_R B \rightarrow C$:
+$f : \D^{\sclrd{n}}_{S} A^n \rightarrow B$ and singleton-shaped $g : \D^\sunit_r B \rightarrow C$:
+%%
+$$
+g \hat{\circ} f = g \circ \D^{\sunit}_r f \circ \delta^{\sclrd{n}}_{r,S} 
+: \D^{\sclrd{n}}_{r \bbullet S} A^n \rightarrow C
+$$
+%$$
+%g \hat{\circ} f = g \circ \D^{\sunit}_R f \circ \delta^{\sclrd{n}}_{R,S} 
+%: \D^{\sclrd{n}}_{R \bbullet S} (A_1 \times \ldots \times A_n)  \rightarrow C
+%$$
+%%
+Note that this composition is asymmetric: the left morphism and right morphisms 
+have different shapes. To compose morphisms which both have
+non-trivial context shapes requires additional structure for manipulating contexts
+(shown in the next section).
+
+\begin{example}[Bounded reuse]
+$\BLL :\Sigma_{\sclrd{n} : \mathbb{N}}, (\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$ has a structural indexed comonad structure with the indexed comonad $\BLL^\sunit$ (Example~\ref{exm:bll2}) and the following structural comultiplication:
+%%
+%\tpnote{Is there a mismatch between $,$ and $\times$?}
+\arraycolsep=1pt\def\arraystretch{0.4}
+\begin{align*}
+%\arraycolsep=3pt\def\arraystretch{0.4}
+& \delta^n_{r,S} = \lambda (\langle{a^1_1, \ldots, a^1_{r \ast S_1}}\rangle, \ldots, \langle{a^n_1, \ldots a^n_{r \ast S_n}}\rangle) . \\[-0.3em]
+& \begin{array}{lllll}
+    ( &  ( &   \langle{a^1_1, \ldots, a^1_{S_1}}\rangle, & ..., & \langle{a^n_1,\ldots, a^n_{S_n}}\rangle ), \\[0.3em]
+            &  ( &   \langle{a^1_{(S_1 + 1)}, ..., a^1_{(S_1+1) + S_1}}\rangle, & ..., & \langle{a^n_{(S_n + 1)}, \ldots, a^n_{(S_n+1) + S_n}}\rangle ), \\[0.6em]
+            &  & \ldots  \\[0.2em]
+            &  ( &   \langle{a^1_{(r - 1) \ast s_1 + 1}, .., a^1_{r \ast S_1}}\rangle, & ..., & \langle{a^n_{(r - 1) \ast S_n + 1}, ..., a^n_{r \ast S_n}}\rangle ) ) 
+   \end{array}
+\end{align*}
+%
+The input is an $n$-variable context 
+containing $r$ times $S_i$ copies of $a^i$ for each variable. The output
+has $r$ copies of a single $n$-variable context containing $S_i$ copies of $a^i$ for each variable. Thus,
+$\delta^n_{r,S}$ partitions the incoming context into $r$-sized contexts. 
+%into a single-variable context, containing $R$ copies of the input vector. In each copy, there are $S$ copies of each
+%variable.
+\label{exm:bll3}
+%\tpnote{Please check - does this make sense?}
+\end{example}
+
+\noindent
+Note that in the case of the flat system, a structural indexed co\-mo\-nad collapses
+to a regular indexed comonad on $\D^{\sunit}$.
+
+\subsection{Splitting and merging contexts}
+\label{sem:monoidal}
+
+Indexed comonads and structural indexed comonads give a semantics for
+sequential composition of contextual computations. However, this does
+not provide enough structure for a full semantics of the coeffect
+calculus. Core to the semantics of abstraction and application is the
+merging and splitting of contexts. Recall the free-variable contexts 
+and coeffects in the (abs) and (app) rules:
+%%
+\begin{align*}
+{\text{\scriptsize{(app)}}}
+\dfrac
+      {\coctx{\Gamma_1}{R} \vdash e_1 ... \;
+       \coctx{\Gamma_2}{S} \vdash e_2 ...}
+      {\coctx{\Gamma_1, \Gamma_2}{ R \asplit (T \aseq S)} 
+\vdash e_1 \, e_2 ...}
+\;
+& 
+\;
+{\text{\scriptsize{(abs)}}}
+\dfrac
+  {\coctx{\Gamma, v : \sigma}{R \amerge \alift{S}} \vdash e ...}
+  {\coctx{\Gamma}{R} \vdash \lambda v . e ... : \sigma \xrightarrow{S}\ldots}
+\end{align*}
+%
+Reading (app) bottom-up, the context of the application
+is split into two contexts for each subterm $e_1$ and $e_2$. 
+Reading (abs) bottom-up, the context of the abstraction
+is merged with the singleton context of the parameter. 
+Capturing these notions in the denotational semantics requires
+some additional structure.
+
+% \tpnote{I removed some details about $\sqcup$ and $\sqcap$ to hide old notation}
+A (non-indexed) comonadic semantics for the $\lambda$-calculus
+requires a \emph{monoidal comonad} with operation $\m_{A,B} : \F A
+\times \F B \rightarrow \F (A \times B)$~\cite{uustalu2008notions}.
+Previously, we defined a similar operation for the semantics of a flat
+coeffect system, with an indexed monoidal operation $\m^{R,S}_{A,B}$ for
+merging contexts. Dually, contexts were split with 
+$\n^{R,S}_{A,B}$~\cite{petricek2013coeffects}. We used two operations for combining and
+splitting the coeffect annotations, respectively. Here we generalize these 
+as $\amerge$ and $\asplit$ using a shape-indexed version.
+
+\begin{definition}
+$\D : \Sigma_{\sclrd{n} :
+    \SH}, (\aclrd{\mathbb{I}}^{\sclrd{n}} \rightarrow
+  [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$ is a 
+   \emph{indexed lax (semi)monoidal functor} and/or \emph{colax (semi)monoidal functor} if it has the following
+ natural transformations respectively: 
+%%
+\begin{align*}
+\mathsf{m}^{n,m}_{R, S} & : \D^n_{R} A \times \D^m_{S} B \rightarrow \D^{n \stimes m}_{R \amerge S} (A \times B) 
+\\
+\mathsf{n}^{n,m}_{R, S} & : \D^{n \stimes m}_{R \asplit S} (A \times B)  
+\rightarrow \D^n_{R} A \times \D^m_{S} B 
+\end{align*}
+%
+In both, shape descriptions are combined by $\stimes$. The first operation models 
+context merging and combines coeffects using $\amerge$. The second models
+context splitting, with $\asplit$ for the pre-split coeffect. 
+\end{definition}
+
+\begin{example}[Bounded reuse]
+For bounded reuse, $\BLL$ is an indexed lax and colax semimonoidal functor
+with the following operations:
+\begin{align*}
+\m^{n,m}_{R,S} & = \lambda (\langle{a_1, ..., a_n}\rangle \times \langle{b_1, ..., b_m}\rangle) . (\langle{a_1, ..., a_n}\rangle, \langle{b_1, ..., b_m}\rangle)
+ \\
+\n^{n,m}_{R,S} & = \lambda (\langle{a_1, ..., a_n}\rangle, \langle{b_1, ..., b_m}\rangle) . (\langle{a_1, ..., a_n}\rangle \times \langle{b_1, ..., b_m}\rangle)
+\end{align*}
+%%%
+\ie{}, %\tpnote{You are not using the tuple notation $(x,y)$ -- but you say so?}
+ $\m^{m,m}_{R,S}$ takes a pair of contexts and merges them simply by replacing the 
+product in $\mathbb{C}$ which pairs the two arguments 
+(written using $\times$) with products inside of $\BLL$ (written using tuple notation $(x, y)$). 
+The operation $\n^{n,m}_{R,S}$ is the inverse.
+\end{example}
+
+\begin{example}[Implicit parameters]
+For implicit parameters, $\Imp^\star$ is an indexed lax and colax semimonoidal
+functor with operations:
+%%
+\begin{align*}
+\m^{\star,\star}_{R,S} & = \lambda ((a, \gamma_R), (b, \gamma_S)) . ((a, b), \gamma_R \cup \gamma_S) \\
+\n^{\star,\star}_{R,S} & = \lambda ((a, b), \gamma) . ((a, {\gamma|}_R), (b, {\gamma|}_S))
+\end{align*}
+%% 
+As in Example~\ref{exm:impl2}, ${\gamma|}_R$ and ${\gamma|}_S$ restrict the set of implicit parameters $\gamma$ to 
+the variable-values pairs for variables in $R$ and $S$. 
+\label{exm:impl3}
+\end{example}
+
+\subsection{Putting it together}
+\label{sem:together}
+The semantics of the structural coeffect calculus $\interp{-}$ is defined in Figure~\ref{fig:semantics},
+using the structures described in the previous sections.
+
+\paragraph{Core rules.}
+The denotation in (\emph{var}) maps a context of the singleton shape $\sunit$
+containing just a single variable $\tau$ (with coeffect $I$) to a $\tau$ value 
+using the counit operation.
+
+The premise of (\emph{abs}) takes a context of shape $\sclrd{n} \,\stimes\, \sunit$ with
+coeffects $\aclrd{R} \amerge \alift{\cclrd{s}}$ and a free-variables context consisting of $\Gamma$ and
+an additional variable $x$. The denotation $g : \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \amerge 
+  \alift{\cclrd{s}}} \interp{\Gamma, v : \sigma} \rightarrow \interp{\tau}$ is
+pre-composed with $\m$, such that its context
+is obtained by merging the declaration-site context ($\Gamma$) and call-site context ($\sigma$):
+%%
+$$
+g \circ \m^{\sclrd{n},\sunit}_{\aclrd{R}, \alift{\cclrd{s}}} : 
+(\D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \times \D^\sunit_{\alift{\cclrd{s}}} \interp{\sigma}) \rightarrow \interp{\tau}
+$$
+%%
+This is uncurried to give a denotation from a context to an exponential object
+representing the abstraction, where the singleton-shaped context becomes the source of the exponential.
+
+% --------------------------------------------------------------------------------------------------
+% Figure - semantics
+% --------------------------------------------------------------------------------------------------
+
+\newcommand{\exchgl}{R \amerge \langle{t}\rangle \amerge \langle{s}\rangle \amerge Q}
+\newcommand{\exchgr}{R \amerge \langle{s}\rangle \amerge \langle{t}\rangle \amerge Q}
+\newcommand{\contrl}{R \amerge \langle{s}\rangle \amerge \langle{t}\rangle \amerge Q}
+\newcommand{\contrr}{R \amerge \langle{s \cpar t}\rangle \amerge Q}
+\begin{figure*}[t]
+\vspace{-0.5em}
+\begin{align*}
+\begin{array}{c}
+%
+% VAR semantics 
+%
+\tyrule{var}
+  {\quad}
+  {\interp{\coctx{x : \tau}{ \alift{\cunit} } \vdash x : \tau} = \varepsilon_I : \D^\sunit_I \interp{\tau} \rightarrow \interp{\tau}}
+\qquad
+%
+% ABS semantics
+%
+\tyrule{abs}
+  {\interp{\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \amerge \alift{\cclrd{s}}} \vdash e : \tau} =
+              g : \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \amerge {\alift{\cclrd{s}}}} \interp{\Gamma, x \!:\! \sigma} \rightarrow \interp{\tau}}
+  {\interp{\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
+                = \Lambda (g \circ \m^{\sclrd{n},\sunit}_{\aclrd{R}, \alift{\cclrd{s}}})
+: \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \rightarrow (\D^\sunit_{\alift{\cclrd{s}}} \interp{\sigma} \Rightarrow \interp{\tau})}
+\\[1.8em]
+%
+% APP semantics
+%
+\tyrule{app}
+   { \interp{\coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau} 
+        = g_1 : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \rightarrow (\D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma} \Rightarrow \interp{\tau})
+     \quad\qquad   
+     \interp{\coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \tau} 
+        = g_2 : \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} \rightarrow \interp{\sigma} }
+   { \interp{\coctx{\Gamma_1, \Gamma_2}{\aclrd{R}\ \asplit\ (\cclrd{t}\ \aseq\ \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
+        = \Lambda^{-1} g_1 \circ (\mathsf{id} \times (\D^{\sunit}_{\alift{\cclrd{t}}} g_2 \circ \delta^{\sclrd{m}}_{\alift{\cclrd{t}}, \aclrd{S}})) \circ \n^{\sclrd{n},\sclrd{m}}_{\aclrd{R}, \cclrd{t} \aseq \aclrd{S}}
+: \D^{\sclrd{n} \stimes \sclrd{m}}_{\aclrd{R} \asplit (\cclrd{t} \aseq \aclrd{S})} \interp{\Gamma_1, \Gamma_2} \rightarrow \interp{\tau} }
+\\[2em]
+\tyrule{ctx}
+  { \interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} = f : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau} 
+    \qquad \interp{\ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma'}{\aclrd{R'}} }{\emptyset}} = c : \D^{\sclrd{m}}_{\aclrd{R'}} \interp{\Gamma'}
+\rightarrow \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma}
+  }
+  { \interp{\coctx{\Gamma'}{\aclrd{R'}} \vdash e : \tau} = f \circ c : \D^{\sclrd{m}}_{\aclrd{R'}} \interp{\Gamma'} \rightarrow \interp{\tau}}
+\end{array}
+\end{align*}
+%
+%
+% -------------- WEAK --------------
+%
+\begin{equation*}
+{\text{\footnotesize{(weak)}}}\quad
+\; \interp{\ctxtrans{\coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x\!:\!\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}}
+%
+% Denotation
+%
+= \pi_1 \circ \n^{\sclrd{n},\sunit}_{\aclrd{R},\aweak}
+: \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \asplit \aweak} (\interp{\Gamma} \times \interp{\sigma}) \rightarrow \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma}
+\end{equation*}
+%
+%
+% -------------- CONTR --------------
+%
+\vspace{-1em}
+\begin{equation*}
+{\text{\footnotesize{(contr)}}} \quad
+\interp{
+  \ctxtrans
+    {\coctx{\Gamma_1, y\!:\!\tau, z\!:\!\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
+    {\coctx{\Gamma_1, x\!:\!\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]} }
+%
+% Denotation
+%
+ = \m^{\sclrd{n},\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s} \cpar \cclrd{t}},\aclrd{Q}} \circ (\mathsf{id} \times \Delta_{\cclrd{s},\cclrd{t}} \times \mathsf{id}) \circ
+\n^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s}},\alift{\cclrd{t}},\aclrd{Q}} \\[1.1em]
+\end{equation*}
+% 
+%
+% -------------- EXCH --------------
+%
+\vspace{-2em}
+\begin{equation*}
+{\text{\footnotesize{(exch)}}} \quad
+\interp{
+  \ctxtrans
+     {\coctx{\Gamma_1, x\!:\!\tau, y\!:\!\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
+     {\coctx{\Gamma_1, y\!:\!\sigma, x\!:\!\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }
+       {\emptyset} }  
+%
+% Denotation
+%
+ = \m^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s}},\alift{\cclrd{t}},\aclrd{Q}} \circ (\mathsf{id} \times \mathsf{swap} \times \mathsf{id}) \circ \n^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{t}},\alift{\cclrd{s}},\aclrd{Q}} \\[0.7em]
+\end{equation*}
+%
+% ------------------ Definitions ---------------------
+%
+\vspace{-1.5em}
+\begin{equation*}
+\textbf{where}~~ \mathsf{swap} : A \times B \rightarrow B \times A~~~\textbf{and}~~
+  \Lambda, \Lambda^{-1} ~~\textnormal{denote currying and uncurrying respectively}
+\end{equation*}
+\vspace{-0.5em}
+\caption{Denotational semantics for the coeffect calculus}
+\label{fig:semantics}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+The application rule (\emph{app})  has two sub-expressions for the function and argument, with denotations
+requiring two distinct contexts:
+%%
+\begin{align*}
+g_1 : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \rightarrow (\D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma} \Rightarrow \interp{\tau})
+\qquad
+g_2 : \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} \rightarrow \interp{\sigma}
+\end{align*}
+%%
+The target of $g_1$ is an exponential object with singleton shape for the parameter of type $\sigma$. 
+To evaluate $g_1$ and $g_2$, the semantics of (\emph{app}) 
+splits the incoming context over $\Gamma_1, \Gamma_2$ using $\n$:
+%% 
+\begin{align*}
+\D^{\sclrd{n} \stimes \sclrd{m}}_{\aclrd{R}\,\asplit\,(\alift{\cclrd{t}}\,\bbullet\,\aclrd{S})}
+(\interp{\Gamma_1} \times \interp{\Gamma_2})
+\xrightarrow{\n^{\sclrd{n},\sclrd{m}}_{\aclrd{R}, \alift{\cclrd{t}}\bbullet\aclrd{S}}} 
+\D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \times \D^{\sclrd{m}}_{\alift{\cclrd{t}} \bbullet \aclrd{S}} \interp{\Gamma_2}
+\end{align*}
+%%
+Since $e_2$ computes the parameter for function $e_1$, the denotation
+$g_2$ must be sequentially composed with the parameter part of
+$g_1$. Thus, the structural indexed comonad is used with $g_2$
+to compute the correct context for the parameter of $g_1$:
+%
+\begin{align*}
+\D^{\sclrd{m}}_{\alift{\cclrd{t}} \bbullet \aclrd{S}} \interp{\Gamma_2} 
+    \xrightarrow{\delta^{\sclrd{m}}_{\cclrd{t},\aclrd{S}}} 
+ \D^\sunit_{\alift{\cclrd{t}}} \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} 
+    \xrightarrow{\D^\sunit_{\alift{\cclrd{t}}} g_2} \D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma}
+%\label{eq:app-morph2}
+\end{align*}
+%
+This is composed with the previous equation by lifting
+to the right-component of the product:
+%
+\begin{align*}
+\D^{\sclrd{n}}_R \interp{\Gamma_1} \times \D^{\sclrd{m}}_S \interp{\Gamma_2}
+\xrightarrow{id \times (\D^\sunit_{\alift{\cclrd{t}}} g_2 \, \circ \, \delta^{\sclrd{m}}_{t,S})}
+\D^{\sclrd{n}}_R \interp{\Gamma_1} \times \D^{\sunit}_{\alift{\cclrd{t}}}  \interp{\sigma}
+\end{align*}
+%%
+This equation computes the calling context and
+parameter context for the function $e_1$, which is then composed with
+the uncurried $g_1$ denotation as shown in the (\emph{app}) rule in Figure~\ref{fig:semantics}.
+
+\paragraph{Structural rules.}
+
+%The second part of Figure~\ref{fig:semantics} shows the semantics of the
+%structural rules for the coeffect calculus. 
+In Figure~\ref{fig:semantics}, (\emph{ctx}) composes the
+denotation of an expression with a transformation $c$ providing the semantic 
+ structural rules. 
+The semantics of structural rules are defined by using
+$\n^{n,m}_{R,S}$ to split contexts, transforming the components, and merging the transformed
+contexts using $\m^{n,m}_{R,S}$. The (\emph{contr}) rule uses an additional operation
+which duplicates a variable inside a context:
+%%
+\begin{align*}
+\Delta_{\cclrd{r},\cclrd{s}} & : \D^{\sunit}_{\cclrd{r} \cpar \cclrd{s}} A
+    \rightarrow \D^{\sunit \stimes \sunit}_{\alift{\cclrd{r}} \amerge \alift{\cclrd{s}}}(A \times A)
+\end{align*}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{example}
+We demonstrate the semantics with a concrete example 
+for the bounded reuse calculus.
+Consider the following:
+%%
+$$
+\coctx{f:\mathbb{Z}\xrightarrow{2}\mathbb{Z}, x:\mathbb{Z}}{\langle 2, 4 \rangle } \vdash (\lambda z.z+z)~(f\ x)
+$$
+We elide the function body's denotation prior to contraction,
+ $g = \interp{x : \mathbb{Z}, y : \mathbb{Z} \vdash (+x) y : \mathbb{Z}}$\footnote{
+the full semantics has $\interp{+} : \D^0_{\langle{}\rangle} 1 \rightarrow (\D^1_1
+\mathbb{Z} \Rightarrow (\D^1_1 \mathbb{Z} \Rightarrow \mathbb{Z}))$}. %  as primitive and uses double application $(+\ e_1) e_2$
+The term therefore has denotation: 
+%%
+\begin{align}
+\label{eq:exm-denot5}
+& \interp{\coctx{}{\langle  \rangle} \vdash
+\lambda z . (+ z) z : \mathbb{Z} \xrightarrow{2} \mathbb{Z}} = \Lambda (g \circ \Delta_{\langle{1}\rangle,\langle{1}\rangle} \circ \m^{0,1}_{\langle{}\rangle,\langle{1}\rangle}) \\
+\notag & \interp{\coctx{f : \mathbb{Z} \xrightarrow{2} \mathbb{Z}, x : \mathbb{Z}}{\langle{1, 2}\rangle} \vdash 
+f x : \mathbb{Z}} \\
+%\label{eq:exm-denot6} 
+\notag
+& \hspace{1.5em} = \uncurry \varepsilon_1 \circ (id \times (\D \varepsilon_1 \circ \delta^1_{2, \langle{1}\rangle})) \circ \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle} \\ 
+\label{eq:exm-denot6} 
+& \hspace{1.5em} = \uncurry \varepsilon_1 \circ \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle} \\ 
+\notag 
+& \interp{\coctx{f : \mathbb{Z} \xrightarrow{2} \mathbb{Z},
+x : \mathbb{Z}}{\langle{2, 4}\rangle}
+\vdash 
+(\lambda z . (+ z) z)\ (f x) : \mathbb{Z}} 
+\\
+\notag
+& \hspace{1.5em} = \uncurry \eqref{eq:exm-denot5} \circ (id \times \D \eqref{eq:exm-denot6} 
+\circ \delta^2_{2,\langle{2,1}\rangle}) 
+\circ \n^{0,2}_{\langle{}\rangle,{\langle{4,2}\rangle}}
+\\
+\label{eq:exm-denot-final}
+& \hspace{1.5em} = g \circ \Delta_{r,s} \circ \m^{0,1}_{\langle{}\rangle,\langle{1}\rangle} \circ (id \times \D \eqref{eq:exm-denot6} 
+\circ \delta^2_{2,\langle{2,1}\rangle}) 
+\circ \n^{0,2}_{\langle{}\rangle,{\langle{4,2}\rangle}}
+\end{align}
+%%
+where \eqref{eq:exm-denot6} and \eqref{eq:exm-denot-final} are simplified. 
+We ``run'' this semantics on some input, evaluating each step of the denotation as a function. 
+%(\ie{}, we take $\mathbb{C} = \mathbf{Set}$). 
+We write context objects, \eg{}, $\D^2_{\langle{R, S}\rangle} (A, B)$ as $\langle(a_1, ..., a_R), (b_1, ..., b_R)\rangle$ and products of
+ contexts in $\mathbb{C}$, \eg{},  $\D^n_R A \times \D^m_S B$, as $(a \times b)$.
+
+\vspace{0.2em}
+\newcommand{\arw}[1]{\makebox[4.3em]{\scriptsize{${#1}$}}}
+\hspace{-2.4em}
+{\scalebox{0.9}{
+\begin{minipage}{1.04\linewidth}
+\begin{align*}
+\begin{array}{c l l}
+%%%% START
+\hspace{-0.5em} & \langle(f_1, f_2),(x_1, x_2, x_3, x_4)\rangle  &\hspace{-5em} : \D^{2}_{\langle{2,4}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z}) \\[-0.5em]
+%%%% STAGE 1
+\xrightarrow{\arw{{n}^{0,2}_{\langle{}\rangle,\langle{2,4}\rangle}}} \hspace{-0.5em} &
+ \langle\rangle \times \langle(f_1, f_2), (x_1, x_2, x_3, x_4)\rangle &  \hspace{-3em}: \D^0_{\langle{}\rangle} 1 \times \D^{2}_{\langle{2,4}\rangle} (\textit{as above}) \\[-0.3em]
+%%
+%\hspace{-0.5em} & 
+%\multicolumn{2}{l}{
+%\hspace{5.91em} : \D^0_{\langle{}\rangle} 1 \times \D^{2}_{\langle{2,4}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z}) 
+%}
+%\\
+%%%% STAGE 2
+\xrightarrow{\arw{\id \times \delta^2_{2, \langle{1, 2}\rangle}}} \hspace{-0.5em} & 
+ \langle\rangle \times \langle\langle f_1, (x_1, x_2)\rangle, \langle f_2, (x_3, x_4)\rangle\rangle \\[0.33em]
+%%
+\hspace{-0.5em} & 
+\multicolumn{2}{l}{
+\hspace{5.91em}
+: \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} \D^{2}_{\langle{2,1}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z})} \\[-0.3em]
+%%%% STAGE 3
+\xrightarrow{\arw{\id \times \D \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle}}} \hspace{-0.5em} &
+ \langle\rangle \times \langle\langle f_1\rangle \times \langle x_1, x_2\rangle, \langle f_2\rangle \times \langle x_3, x_4\rangle\rangle \\[0.34em]
+%%
+\hspace{-0.5em} & 
+\multicolumn{2}{l}{
+\hspace{5.91em} 
+: \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} (\D^{1}_{\langle{1}\rangle} (\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \D^{1}_{\langle{2}\rangle} \mathbb{Z})} \\[-0.1em]
+%%%% STAGE 4
+\xrightarrow{\arw{\id \times \D (\uncurry \varepsilon_1)}}  
+\hspace{-0.5em} & \langle\rangle \times \langle f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em} : \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} \mathbb{Z} \\[-0.2em]
+%%%% STAGE 5
+\xrightarrow{\arw{\m^{0,1}_{\langle{}\rangle,\langle{1}\rangle}}} 
+\hspace{-0.5em} & \langle f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em}  : \D^1_{\langle{2}\rangle} \mathbb{Z} \\
+%%%% STAGE 6
+\xrightarrow{\arw{\Delta^2_{\langle{1}\rangle,\langle{1}\rangle}}}
+\hspace{-0.5em} & \langle(f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle)\rangle & \hspace{-3.5em} : \D^1_{\langle{1,1}\rangle} (\mathbb{Z} \times \mathbb{Z}) \\[0.54em]
+%%%%% STAGE 7
+\xrightarrow{\arw{g}} 
+ \hspace{-0.5em} & \interp{+}\ \langle f_1 \langle x_1, x_2\rangle\rangle\ \langle f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em} : \mathbb{Z} 
+\end{array}
+\end{align*}
+\end{minipage}
+}}
+\label{exm:run}
+%This demonstrates the successive splitting, merging, and composing of contexts to the final result. 
+\end{example}
+
+\subsection{Soundness, with respect to equational theory}
+\label{sec:semantics-soundness}
+
+\noindent
+Our denotational semantics for the coeffect
+calculus is sound with respect the equational theory of Section~\ref{sec:syntax}. That is:
+%%
+\begin{theorem}[Soundness]
+$$\coctx{\Gamma}{R} \vdash e \equiv e' : \tau \Rightarrow \interp{\coctx{\Gamma}{R} \vdash e : \tau} 
+\equiv \interp{\coctx{\Gamma}{R} \vdash e' : \tau}$$
+\end{theorem}
+%%
+\noindent
+Proof of this follows from an interesting result which we 
+first unpack: determining whether $\interp{\coctx{\Gamma}{R} \vdash
+  e_1 : \tau} {\equiv} \interp{\coctx{\Gamma}{S} \vdash e_2 : \tau}$ follows
+from a proof on coeffect annotations that $R = S$. 
+%
+\begin{lemma}
+Every coeffect algebra axiom corresponds to an axiom 
+of one of the categorical structures introduced here (indexed (structural) comonad or 
+  indexed (co)lax monoidal functor). 
+\end{lemma}
+%
+\noindent
+For example, the monoid axiom $X \cseq \cunit =
+X$ for scalar coeffects corresponds to indexed comonad axiom $\D^{\sunit}_X \varepsilon_{\cunit}
+\circ \delta^{\sunit}_{X,{\cunit}} = id_{\D^{\sunit}_X}$ (which requires the monoid axiom to hold). 
+This lemma follows from our derivation of 
+the indexed categorical structures here. They are not derived \emph{ad hoc} but systematically 
+as (lax) \emph{homomorphisms} (structure preserving maps) between the structure of 
+coeffect annotations in $\Icat$ and the structure of denotations in $\mathbb{C}$.
+
+\begin{proposition}
+An indexed comonad on $\D$ witnesses that
+$\D$ is a \emph{colax monoid homomorphism} between the (strict)
+monoidal categories $(\mathbb{I}, \cseq, \cunit)$ and 
+$([\mathbb{C}, \mathbb{C}], 
+\circ, 1_{\mathbb{C}})$ (endofunctor composition). 
+\end{proposition}
+%%
+Unpacking this, a \emph{monoid homomorphism} maps between
+the underlying sets of two monoids, preserving the monoid structure of one into the other,
+\ie{}, given monoids $(X, \bullet, I)$ and $(Y, \otimes, E)$ then
+a monoid homomorphism is a mapping $F : X \rightarrow Y$ such that:
+%%
+\begin{align*}
+F X \otimes F Y \equiv F (X \bullet Y) \quad\qquad E \equiv F I
+\end{align*}
+%%
+The axioms of each monoid are preserved trivially by these equalities, 
+\eg{}, $F X \equiv F (X \bullet I) \equiv F X \otimes F I \equiv F X \otimes E \equiv F X$. 
+In a categorical setting, monoids are taken on categories with binary operations as 
+bifunctors. A homomorphism is \emph{lax} if the above equalities are instead morphisms 
+(which we say \emph{witness} the homomorphism) and \emph{colax} if these morphisms go in opposite direction.
+Thus, a colax monoid homomorphism is witnessed by: 
+%%
+\begin{align*}
+\delta : F X \otimes F Y \leftarrow F (X \bullet Y) \quad\qquad \varepsilon : E \leftarrow F I
+\end{align*}
+%%
+Note our choice of morphism names.  
+$F$ no longer preserves the monoid axioms \emph{up to equality} but has axioms on $\delta$ and $\varepsilon$, %% F X {=\joinrel=} 
+\eg{}, $F (X \bullet I) \xrightarrow{\delta} F X \otimes F I \xrightarrow{id \otimes \varepsilon} F X \otimes E$ equals $F X \xrightarrow{id} F X$. 
+%%  {=\joinrel=} F X
+
+Our indexed comonad definition is equivalent to $\D$ being a colax homomorphism between a monoidal category $(X, \bullet, I)$ 
+and the monoidal category of $\mathbb{C}$ endofunctors $(Y, \otimes, E) = ([\mathbb{C}, \mathbb{C}], \circ, 1_{\mathbb{C}})$, with endofunctor composition 
+$\circ$ and the trivial endofunctor $1_{\mathbb{C}}$. The indexed comonads axioms are the axioms of the colax homomorphism.
+Equivalently, $\D$ is a \emph{colax monoidal functor}.
+
+A similar approach is taken to deriving the remaining structures.%, though we give less detail for brevity.
+
+\begin{proposition}
+  A structural indexed comonad provides $\delta^n_{R,S} :
+  \D^\sunit_r \D^n_S A^n \leftarrow \D^n_{r \aseq S} A^n$ which witnesses that $\D$ is a 
+  colax homomorphism between the following \emph{monoid left-actions} for
+  $(\mathbb{I}, \cseq, \czero)$ and $([\mathbb{C}, \mathbb{C}], \circ,
+  1_{\mathbb{C}})$: 
+\begin{align*}
+\begin{array}{rclcl}
+(r : \mathbb{I}) & \narrow{\aseq} & (\langle{s_1, ..., s_n}\rangle : \mathbb{I}^n) &\narrow{=}& \langle{r \cseq s_1, ..., r \cseq s_n}\rangle : \mathbb{I}^n  \\
+(\D^\sunit_r : [\mathbb{C}, \mathbb{C}]) & \narrow{\circ} & (\D^n_S : [\mathbb{C}^n, \mathbb{C}]) &\narrow{=}& \D^\sunit_r \D^n_S : [\mathbb{C}^n, \mathbb{C}]
+\end{array}
+%(x : \mathbb{I}) \aseq (\langle{y_1, \ldots, y_n}\rangle : \mathbb{I}^n) & = \langle{x \cseq y_1, \ldots, x \cseq y_n}\rangle  \\
+%(\D^\sunit_r : \mathbb{C} \rightarrow \mathbb{C}) \circ (\D^n_S : \mathbb{C}^n \rightarrow \mathbb{C}) & = \D^\sunit_R \D^n_S : \mathbb{C}^n \rightarrow \mathbb{C}
+\end{align*}
+The axioms are the lax versions of the monoid left-action laws. 
+%witnesses that $\D$
+%is a colax homomorphism between the induced \emph{monoid left-action} for $(\mathbb{I}, \cseq, \czero)$
+%of $n$-length coeffect vectors $(\mathbb{I}^n, \aseq_n : \mathbb{I} \times \mathbb{I}^n 
+%\rightarrow \mathbb{I}^n)$ for monoid, and a trivial monoid left-action for $([\mathbb{C}, \mathbb{C}], \circ, 1_{\mathbb{C}})$ 
+%defined by composition $([\mathbb{C}^n, \mathbb{C}], \circ : [\mathbb{C}, \mathbb{C}] \times [\mathbb{C}^n, \mathbb{C}] \rightarrow [\mathbb{C}^n, 
+%\mathbb{C}])$, 
+%\ie{},
+\end{proposition}
+%%
+The lax and colax indexed monoidal operations $\m^{n,m}_{R,S}$ and $\n^{n,m}_{R,S}$ follow a similar derivation
+but as lax and colax monoid homomorphism between composite monoids on coeffect annotations and shapes
+and $\times$ in $\mathbb{C}$. The details are elided here.
+
+Returning to soundness, our semantics is therefore defined in terms of structures whose axioms correspond 
+to axioms of the syntactic equational theory. Consequently, semantic proofs correspond to 
+syntactic proofs, modulo naturality laws and product/exponent laws in $\mathbb{C}$. 
+This result holds in the general coeffect calculus and semantics since every semantic structure
+has a unique corresponding structure on coeffect annotations (\ie{}, $(\C, \cseq, \cunit)$ for
+sequential composition of unary denotations, $(\C, \asplit)$ for splitting contexts, 
+$(\C, \amerge)$ for joining contexts).
+
+\begin{example}
+Section~\ref{prop:structural} showed $\eta$-equality for structural systems,
+which uses the properties (1) $\asplit = \amerge = \times$ for structural systems and (2) 
+$s \aseq \langle{\cunit}\rangle
+= \langle{s \cseq \cunit}\rangle = \langle{s}\rangle$. The semantics here is sound
+with respect to $\eta$-equality, the proof of which uses the corresponding axioms
+(1) $\n^{n,m}_{R,S} \circ \m^{n,m}_{R,S} = id$ and (2) $\D^{\sunit}_s \varepsilon_{\cunit} \circ \delta^m_{s,\cunit}$
+(structural indexed comonad unit law, Definition~\ref{def:structural-indexed-comonad}). 
+% $\D^{n}_R A \times \D^{m}_S B \xrightarrow{\m^{n,m}_{R,S}} \D^{n,m}_{R \amerge S} (A \times B) \equiv
+% \D^{n,m}_{R \asplit S} (A \times B) \xrightarrow{\n^{n,m}_{R,S}} \D^n_R A \times \D^m_S$ (\ie{}, $\n \circ \m = id$) and 
+\end{example}
+%
+\noindent
+The accompanying technical report shows the full semantic proofs for $\beta\eta$-equality whose structure corresponds
+exactly to the syntactic proofs with the corresponding coherence conditions~\cite{appendix}.
+
+%\begin{prosition}
+%A lax indexed monoidal functor on $\D$ witnesses that $\D$
+%is a lax monoid homomorphism between the $(\mathbb{I}, \amerge, \azero)$ and 
+%a composite monoid of $
+%\end{proposition}
+
+%The indexed comonad structure is derived from a
+%\emph{colax} \emph{homomorphism} between the monoidal structure
+%$(\mathbb{I}, \bullet, I)$ on indices and monoidal structure of
+%endofunctor composition. This approach is used to derive the rest
+%of our semantics, though we keep the details of the derivation light
+%here and present instead the resulting structures.
+
+%\begin{remark}
+%The category of $\mathbb{C}$-endofunctors $[\mathbb{C}, \mathbb{C}]$
+%is a \emph{strict} monoidal category $([\mathbb{C}, \mathbb{C}], \circ,
+%1_{\mathbb{C}})$ with composition $\times$ as the binary operation and the identity
+%functor $1_{\mathbb{C}}$ as the unit. 
+%\end{remark}
+
+% ==================================================================================================
+
+\section{Related work}
+\label{sec:related-work}
+
+We expand briefly on the overview of related work in 
+Section~\ref{sec:intro-pathways}.
+
+%\paragraph{Bounded reuse.}
+The (\emph{storage}) rule for bounded linear logic explains the contextual requirements 
+induced by proposition reuse~\cite{girard1992bounded}:
+$$
+\inference[\text{(storage)}]{!_{\overline{Y}} \Gamma \vdash A}{!_{X\overline{Y}} \Gamma \vdash !_X A} 
+$$
+where $X\overline{Y} = \langle{XY_1, .., XY_n}\rangle$ is the scalar multiple of a vector. 
+This rule is akin to the $\delta^n$ operation of structural indexed comonads,
+indeed, we can model it exactly using $\delta^{n}_{X,\bar{Y}}$ and the lifting $\D^n_X$.
+
+In BLL, the modality $!_X$ is a constructor and may 
+  appear both on the left- and right-hand sides of $\vdash$.
+  In this paper, reuse bounds 
+  annotate typing rules, thus there is no constructor
+  corresponding to bounded reuse in the language; reuse bounds are
+  meta-level. Our choice to work at the meta-level means that the 
+  coeffect calculus provides a unified analysis and semantics to
+  different notions of context and its term language is that of standard $\lambda$-calculus.
+
+%\paragraph{Semantics.}
+Previously we briefly introduced indexed comonads~\cite{petricek2013coeffects}
+without derivation.  Here we derived indexed comonads as colax
+homomorphisms. This is dual to the \emph{parametric effect monad}
+structure defined as a lax
+homomorphism~\cite{katsumata2014parametric}. Our semantics 
+requires additional structure not needed for effects due to the
+asymmetry inherent in the $\lambda$-calculus.
+
+The necessity modality $\Box$ in S4 logic corresponds to a comonad
+with lax monoidal functor structure $\m : \Box A \times \Box B \rightarrow \Box (A \times B)$. 
+Bierman and de Paiva \cite{bierman2000intuitionistic} defined a term language corresponding to a natural deduction S4, where contexts contain sequences of $\Box$-wrapped assumptions
+$x_1 \!:\! \Box A_1, \ldots x_n \!:\! \Box A_n$. Modelling these judgments does not require a context splitting operation $\n$ 
+unlike in our approach. Our approach can be thought of as having a single $\Box$ modality over the context 
+which allows both flat whole-context dependence and structural per-variable dependence.
+
+%\noindent
+%\todo{Discuss related work not covered in the background section. More details on the difference
+%between our work and modal logic approaches. etc!}
+
+\section{Conclusions}
+
+In this paper, we looked at two forms of context-dependence
+analysis -- \emph{flat} coeffect systems that track whole-context
+requirements (such as implicit parameters, resources, or platform version) and
+\emph{structural} coeffects that track per-variable requirements (such
+as usage or data access patterns). 
+The newly introduced structural system makes applications such as liveness,
+bounded reuse, and dataflow analysis (from our earlier work)
+practically useful. With the move towards cross-platform systems running in 
+diverse environments, analysing context dependence is vital for reasoning and compilation. The coeffect calculus
+provides a foundation for further study, similar to the type-and-effect discipline.
+
+We presented the system together with its syntactic equational theory and categorical semantics.
+The equational theory is presented in order to explain how the systems work, but it also
+provides a basis for an operational semantics for concrete systems. Exploring these, 
+and their connection to the denotational semantics, is further work.

### 2014-03-14 text/Chapter00.tex
Add dataflow example to introduction

@@
 
-\subsection{Context-awareness \#4: Access patterns in data-flow languages}
+\subsection{Context-awareness \#4: Checking array access patterns}
 
-As an example, consider a language that allows us to get a value of a variable (representing
-some changing data-source) \ident{x} versions back using the syntax $\ident{x}_{[a]}$. 
-To track information about individual variables, we use a product-like operation $\times$ on tags 
-to mirrors the product structure of variables. For example:
-%
-\begin{equation*}
-?
-\end{equation*}
-%
-The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
-that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
-\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
-maximum of the requirements of the individual variables:
+The last example leaves the topic of cross-platform and distributed computing. We focus on 
+checking how arrays are accessed. This is a simpler version of the data-flow programming
+examples used later in the thesis.
+
+Consider a simple programming language with arrays where $n^{\textnormal{th}}$ element of an
+array \ident{arr} is accessed using $\ident{arr}[n]$. Furthermore, we focus on performing local
+transformations and we assume that the keyword $\kvd{cursor}$ returns the \emph{current} location
+in the array.
+
+The following example implements a simple one-dimensional cellular automata, reading from the
+\ident{input} array and writing to \ident{output}:
 %
@@
 \begin{array}{l}
-?
+\kvd{let}~\ident{sum}=\ident{input}[\kvd{cursor} - 1] ~+~ \ident{input}[\kvd{cursor}] ~+~ \ident{input}[\kvd{cursor} + 1]\\[-0.25em]
+\kvd{if}~\ident{sum}=2 ~||~ (\ident{sum}=1 ~\&\&~ \ident{input}[\kvd{cursor} - 1] = 0)\\[-0.25em] 
+\kvd{then}~\ident{output}[\kvd{cursor}] \leftarrow 1 ~\kvd{else}~\ident{output}[\kvd{cursor}] \leftarrow 0
+
 \end{array}
@@
 %
-This version of the calculus removes the non-determinism of the lambda abstraction from the previous
-version. As we associate information with individual variables, lambda abstraction creates a function
-that requires the context requirements associated with the variable that is being abstracted over.
+In this example, we use the term \emph{context} to refer to the values in the array around the 
+current location provided by \kvd{cursor}. The interesting question is, how much of the context
+(\ie~how far in the array) does the program access.
 
-For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
-\ident{b} from the declaration-site) then the function would have context requirements $5$ -- that
-is the number associated with the variable \ident{a}.
+This is contextual information attached to individual (array) variables. 
+In the above example, we want to track that \ident{input} is accessed in the range $\langle-1, 1 \rangle$
+while \ident{output} is accessed in the range $\langle 0, 0 \rangle$. When calculating the ranges,
+we need to be able to compose ranges $\langle-1,-1$, $\langle 0,0 \rangle$ and $\langle 1,1 \rangle$
+(based on the accesses on the first line).
+
+The information about access patterns can be used to efficiently compile the computation
+(as we know which sub-range of the array might be accessed) and it also allows better handling
+of boundaries. For example, wrap-around behaviour we could pad the \ident{input} with a known
+number of elements from the other side of the array.
 

### 2014-03-14 text/Chapter00.tex
Remove resources example from introduction

@@
 
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Context-awareness \#4: Resource \& data availability}
-
-A vast majority of applications accesses some data sources (like database) or resources
-(like GPS sensor on a phone). This is more tricky for client/server applications where a part
-of program runs on the server-side and another part runs on the client-side. I believe that 
-these two parts should be written as a single program that is cross-compiled to two parts 
-(and I tried to make that possible with F\# Web Tools \cite{xx})
-a long time ago; more recently [WebSharper](http://websharper.com/) implemented a similar idea).
-
-So, say we have a function \texttt{validateInput}, \texttt{readData} and \texttt{displayMessage} in my prorgam.
-I want to look at their types and see what resources (or \emph{context}) they require. For example,
-\texttt{readData} requires \emph{database} (or perhaps a database with a specific name), \texttt{displayMessage} 
-requires access to \emph{user interface} and \texttt{validateInput} has no special requirements.
-
-This means that I can call \texttt{validateInput} from both server-side and client-side code - it is
-safe to share this piece of code, because it has no special requirements. However, when I write
-a code that calls all three functions without any remote calls, it will only run on a thick
-client that has access to a database as well as user interface.
-
-I'll demonstrate this idea with a sample (pseudo-)code in a later section, so do not worry if
-it sounds a bit abstract at first.
-
 % ==================================================================================================

### 2014-03-14 text/Chapter00.tex
Edits in the middle of the intro chapter

@@
 
-\section{Coeffects: Towards context-aware languages}
+\section{Towards context-aware languages}
 
-The above examples cover a couple of different scenarios, but they share a common theme - 
-they all talk about some \emph{context} in which an expression is evaluated. The context has 
-essentially two aspects:
+The four examples presented in the previous section cover different notions of \emph{context}.
+The context can be viewed as execution environment, capabilities provided by the environment 
+or input and meta-data about the input.
+
+The different notions of context can be broadly classified into two categories -- those that
+speak about the environment and those that speak about individual inputs (variables). In this
+thesis, we refer to them as \emph{flat coeffects} and \emph{structural coeffects}, respectively:
 
 \begin{itemize}
-\item \textbf{Flat context} represents additional data, resources and meta-data that are 
-   available in the execution environment (regardless of where in the program you 
-   access them). Examples include resources like GPS sensors or databases, battery status,
-   framework version and similar.
+\item \textbf{Flat coeffects} represent additional data, resources and meta-data that are 
+   available in the execution environment (regardless of how they are accessed in a program).
+   Examples include resources such as GPS sensors and battery status (on a phone), databases 
+   (on the server), or framework version.
 
-\item \textbf{Structural context} contains additional meta-data related to variables. This can include
-   provenance (source of the variable value), usage information (how often is the value
-   accessed) or security information (does it contain sensitive data). 
+\item \textbf{Structural coeffects} capture additional meta-data related to inputs. This can include
+   provenance (source of the input value), usage information (how often is the value
+   accessed and in what ways) or security information (whether it contain sensitive data or not). 
 \end{itemize}
+%
+This thesis follows the tradition of statically typed programming languages. As such, we
+attempt to capture such contextual information in the type system of context-aware programming
+languages. The type system should provide both safety guarantees (as in the first three examples)
+and also static analysis useful for optimization (as in the last example).
 
-As a proponent of statically typed functional languages I believe that a context-aware 
-programming language should capture such context information in the type system and make
-sure that basic errors (like the ones demonstrated in the four examples above) are ruled
-out at compile time.
-
-This is essentially the idea behind \emph{coeffects}. Let's look at an example showing the
-idea in (a very simplified) practice and then I'll say a few words about the theory
-(which is the main topic of this thesis).
+Although the main focus of this thesis is on the underlying theory of \emph{coeffects} and on
+their structure, the following section briefly demonstrates the features that a practical
+context-aware language, based on the theory of coeffects, can provide.
 
@@
 
-\subsection{Case study: Coeffects in action}
+\section{Context-aware languages in action}
 
@@
 
-\subsection{Coeffects: Theory of context dependence}
+\section{Coeffects: Theory of context dependence}
 
@@
 
-\subsubsection{Brief introduction to type systems}
+\subsection{Brief introduction to type systems}
 
@@
 
-\subsubsection{Checking what program does with effect systems}
+\subsection{Checking what program does with effect systems}
 
@@
 
-\subsubsection{Checking what program requires with coeffect systems}
+\subsection{Checking what program requires with coeffect systems}
 

### 2014-03-27 text/Chapter01.tex
Finish introduction chapter and restructure the rest

--- 2013-06-04 text/Chapter01.tex
@@
-\chapter{Introduction} 
-\label{ch:introduction}
+% ==================================================================================================
 
-%----------------------------------------------------------------------------------------
+\chapter{Why context-aware programming matters} 
+\label{ch:intro}
 
-Some intro
+Many advances in programming language design are driven by some practical motivations. Sometimes, 
+the practical motivations are easy to see -- for example, when they come from an external change 
+such as the rise of multi-core processors. Sometimes, discovering the practical motivations is a 
+difficult task -- perhaps because we are so used to a certain way of doing things that we do not 
+even \emph{see} the flaws of our approach.
 
-\section{How lambda works}
-The key difference - how lambda works
+Before exploring the motivations leading to this thesis, we briefly consider two recent practical 
+concerns that led to the development of new programming languages. This helps to explain why 
+context-aware programming is of importance. The examples are by no means representative, but they 
+illustrate different kinds of motivations well.
 
-Effect systems, introduced by Gifford and Lucassen
-\cite{effects-gifford}, track \emph{effects} of computations, such as
-memory access or message-based
-communication~\cite{effects-messagepassing}. Their approach augments
-typing judgments with effect information: $\Gamma \vdash e :
-\tau, F$.  Wadler and Thiemann explain how this shapes
-effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
-%
-\begin{quote}
-\emph{In the rule for abstraction, the effect is empty because evaluation immediately
-returns the function, with no side effects. The effect on the function arrow
-is the same as the effect for the function body, because applying the function will
-have the same side effects as evaluating the body.}
-\end{quote}
-%
-In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
-does not match this pattern. In the systems we consider, lambda abstraction places 
-requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
-(immediate requirements), resulting in different syntactic properties. 
-We informally discuss three examples first that demonstrate how contextual
-requirements propagate. Section ? then unifies these 
-in a single calculus.
+\paragraph{Parallel programming.}
 
-Also \cite{effects-producer-semantics}
+The rise of multi-core CPUs is a clear example of an external development influencing
+programming language research. As multi-core and multi-processor systems became de-facto standard, 
+languages had to provide better abstractions for parallel programming. This led to the industrial 
+popularity of \emph{immutable} data structures (and functional programming in general), software 
+transactional memory \cite{app-haskell-stm}, data-parallelism and also asynchronous 
+computing \cite{app-fsharp-async}.
 
-\begin{quote}
-\emph{We will define an effect as a producer effect if all computations with that effect 
-can be thunked as "pure" computations for a domain-specific notion of purity.}
-\end{quote}
+In this case, the motivation is easy to see -- writing multi-core programs using earlier 
+abstractions, such as threads and locks, is difficult and error-prone. At the same time, 
+multi-core CPUs become the standard very quickly and so the lack of good language abstractions
+was apparent.
 
-Demonstrate using distributed computations
+\paragraph{Data access.}
 
-Demonstrate using dataflow/liveness
+Accessing data is an example of a more subtle challenge. Initiatives like open government data\footnote{
+In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
+certainly make more data available. However, to access the data, one has to parse CSV and Excel files, 
+issue SQL or SPARQL queries (to query database and the semantic web, respectively).
 
-Perhaps mention LINQ as a motivation for cross-compilation..
-\cite{app-linq}
+Technologies like LINQ \cite{app-linq} make querying data significantly easier. But perhaps 
+because accessing data became important more gradually, it was not easy to see that inline
+SQL is a poor solution \emph{before} better approaches were developed.
 
-\section{Associating with context}
+This is even more the case for \emph{type providers} -- a recent feature in F\# that integrates
+external data sources directly into the type system of the language and thus makes data explorable
+directly from the source code editor (through features such as auto-completion on object members).
+It is not easy to see the limitations of standard techniques (using HTTP requests to query REST 
+services or parsing CSV files and using string-based lookup) until one sees how type providers 
+change the data-scientist's workflow\footnote{This is difficult to explain in writing and so the 
+reader is encouraged to watch a video showing type providers for the WorldBank and CSV data 
+sources \cite{app-fsharp-world}.}.
 
--> Type providers make this important
+\paragraph{Context-aware programming.}
 
-%---------------------------------------------------------------------------------------------------
+In this chapter, we argue that the next important practical challenge for programming language 
+designers is designing languages that are better at working with (and understanding) the 
+\emph{context in which programs are executed}.
 
-\section{Flat coeffect system}
-~
+This challenge is of the kind that is not easy to see -- perhaps because we are so used to 
+doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
+uncover such flaws -- we look at a number of basic programs that rely on contextual 
+information, we explain why they are inappropriate and then we briefly outline how this
+thesis remedies the situation.
 
-resources?
+Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
+the goal of programming language research is generally to design languages that provide 
+more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and 
+more \emph{unified}. 
+These are exactly the aims that we follow in this thesis. In this chapter, we explain what
+the common problems are. In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural},
+we develop two simple calculi to understand and capture the structure of the problems and,
+finally, Chapter~\ref{ch:unified} unifies the two abstractions.
 
-%---------------------------------------------------------------------------------------------------
+% ==================================================================================================
 
-\section{Structural coeffect system}
-\label{sec:structural}
+\section{Why context-aware programming matters}
 
-The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
-we need to track context-dependence in a more fine-grained way. To track neededness or security, 
-we need to associate information with individual \emph{variables} of the context.
+The phrase \emph{context in which programs are executed} sounds rather abstract and generic.
+What notions of \emph{context} can be identified in modern software systems?
+Different environments provide different resources (\eg~database
+or GPS sensor), environments are increasingly diverse (\eg~multiple versions of different
+mobile platforms). Web applications are split between client, server and mobile components; mobile
+applications must be aware of the physical environment while the ``internet
+of things`` makes the environments even more heterogeneous. At the same time, applications access
+rich data sources and need to be aware of security policies and provenance information from the
+environment.
 
-At the same time, we want to avoid developing multiple variants of coeffect systems -- indeed, 
-our motivation is to develop a \emph{single unified mechanism} for tracking context-dependence.
-In this section, we present a more powerful \emph{structural coeffect calculus}, which is a 
-generalization of the flat calculus.
+Writing such context-aware (or environment-aware) applications is a fundamental problem of modern 
+software engineering. The state of the art relies on ad-hoc approaches -- using hand-written conditions 
+or pre-processors for conditional compilation. Common problems that developers face include:
 
-%We show that \clflt~can be obtained as a special case of \clstr. The
-%structural version of our system is also where our work differs more significantly from well-known
-%effect systems.
+\begin{itemize}
+\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple 
+  targets (e.g. SQL \cite{app-linq}, OpenCL or JavaScript \cite{app-hop-lang}) a part of the compilation 
+  (generating the SQL query) often occurs at runtime and developers have no guarantee that it will succeed 
+  until the program is executed.
+
+\item \textbf{Platform versions.} When developing cross-platform applications, different platforms
+  (and different versions of the same platform) provide different API functions. Writing a 
+  cross-platform code usually relies on (fragile) conditional compilation or (equally fragile) 
+  dynamic loading.
+
+\item \textbf{Security and provenance.} When working with data (be it sensitive database or social 
+   network data), we have permissions to access only some of the data and we may want to track 
+   \emph{provenance} information. However, this is not checked -- if a program attempts to access 
+   unavailable data, the access will be refused at run-time.
+
+\item \textbf{Resources \& data availability.} When creating a mobile application, the program may
+  (or may not) be granted access to device capabilities such as GPS sensor, social updates or battery
+  status. We would like to know which of the capabilities are required and which are optional
+  (\ie~enhance the user experience, but there is a fallback strategy). Equally, on the server-side,
+  we might have access to different database tables and other information sources.
+\end{itemize}
+
+Most developers do not perceive the above as programming language flaws -- they are simply common
+programming problems (at most somewhat annoying and tedious) that had to be solved. However, this 
+is because we do not realize that a suitable language extension could make the above problems 
+significantly easier to solve. As the number of distinct contexts and their diversity increases,
+these problems will become even more commonplace.
+
+The following sub-sections explore four examples in more details. The examples are chosen to 
+demonstrate two distinct forms of contexts that are studied in this thesis.
 
@@
 
-\subsection{Motivation: Tracking array accesses}
-\label{sec:structural-motivation}
-\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
-\newcommand{\dnat}{\ident{D}_\ident{nat}}
-\newcommand{\cprd}{\times}
-\newcommand{\cvop}{~\ident{max}~}
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\kvd{for}~\ident{header},\ident{value}~\kvd{in}~\ident{header}~\kvd{do}\\[-0.25em]
+\hspace{2em}\kvd{match}~\ident{header}~\kvd{with}\\[-0.25em]
+\hspace{2em}|~\str{accept} \rightarrow~ \ident{req.Accept} \leftarrow \ident{value}\\[-0.25em]
+%
+\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_USERAGENT}\\[-0.25em]
+\hspace{2em}|~\str{user-agent} \rightarrow~ \ident{req.UserAgent} \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#else}\\[-0.25em]
+\hspace{2em}|~\str{user-agent} \rightarrow~
+  \ident{req.Headers}.[\,\ident{HttpHeader.UserAgent}\,] \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#endif}\\[-0.25em]
+%
+\prepk{\#if}~\prepi{FX\_NO\_WEBREQUEST\_REFERER}\\[-0.25em]
+\hspace{2em}|~\str{referer} \rightarrow~ \ident{req.Referer} \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#else}\\[-0.25em]
+\hspace{2em}|~\str{user-agent} \rightarrow~
+  \ident{req.Headers}.[\,\ident{HttpHeader.Referer}\,] \leftarrow \ident{value}\\[-0.25em]
+\prepk{\#endif}\\[-0.25em]
+%
+\hspace{2em}|~\ident{other} \rightarrow~ \ident{req.Headers}.[\,\ident{other}\,] \leftarrow \ident{value}\\[-0.25em]
+\end{array}
+\end{equation*}
 
-Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
-functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
-respectively, but we use richer tag structure.
+\vspace{1em}
+\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
+\vspace{-1.5em}
+\caption{Conditional compilation in the HTTP module of the F\# Data library}
+\label{fig:introduction-context-http}
+\end{figure}
 
-As an example, consider a language that allows us to get a value of a variable (representing
-some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
-To track information about individual variables, we use a product-like operation $\cprd$ on tags 
-to mirrors the product structure of variables. For example:
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Context awareness \#1: Platform versioning}
+
+The diversity across devices means that developers need to target an increasing number of platforms
+and possibly also multiple versions of each platform. For Android, there is a number called 
+API level \cite{app-android-apilevel} which ``uniquely identifies the framework API revision offered 
+by a version of the Android platform''. Most changes in the libraries (but not all) are additive.
+
+Equally, in the .NET ecosystem, there are multiple versions of the .NET runtime, mobile and 
+portable versions of the framework etc. The differences may be subtle -- for example, some members
+are omitted to make the mobile version of the library smaller, some functionality is not available
+at all, but naming can also vary between versions.
+
+For example, the Figure~\ref{fig:introduction-context-http} shows an excerpt from the \ident{Http} module 
+in the F\# Data library\footnote{The file version shown here is available at:
+\url{https://github.com/fsharp/FSharp.Data/blob/b4c58f4015a63bb9f8bb4449ab93853b90f93790/src/Net/Http.fs}}.
+The example uses conditional compilation to target multiple versions of the .NET framework. Such code
+is difficult to write -- to see whether a change is correct, it had to be recompiled for all combinations
+of pre-processor flags -- and maintaining the code is equally hard. The above example could be refactored
+and the .NET API could be cleaner, but the fundamental issue remains. If the language does not understand
+the context (here, the different platforms and platform versions), it cannot provide any static
+guarantees about the code.
+
+As an alternative to conditional compilation, developers can use dynamic loading. For example, 
+on Android, programs can access API from higher level platform dynamically using techniques like 
+reflection and writing wrappers. This is even more error prone. As noted in an article introducing
+the technique\footnote{Retrieved from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}:
+\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}. Again, it would be 
+reasonable to expect that statically-typed languages could provide a better solution.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Context awareness \#2: System capabilities}
+
+Another example related to the previous one is when libraries use meta-programming techniques
+(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
+written in a subset of a host language to some other target language, such as SQL, OpenCL or 
+JavaScript. This is an important technique, because it lets developers targets multiple 
+heterogeneous runtimes that have limited execution capabilities.
+
+For example, consider the following LINQ query written in C\# that queries a database and
+selects product names where the first upper case letter is "C":
 %
@@
 \begin{array}{l}
-\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
-  \vdash
-    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
+\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
+\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
+\kvd{where}~\ident{p.ProductName.First}(\ident{c} \Rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
+\kvd{select}~\ident{p.ProductName};
 \end{array}
@@
 %
-The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
-that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
-for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
-a single variable:
+This appears as a perfectly valid code ant the C\# compiler accepts it. However, when the program
+is executed, it fails with the following error:
+%
+\begin{quote}
+Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators 
+not supported for type \texttt{System.String}.
+\end{quote}
+%
+The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
+uses \ident{First} method to iterate over characters of a string, which is not supported. This is
+not a technical limitation of LINQ, but a fundamental problem of the approach.
+
+When cross-compiling to a limited environment, we cannot always support the full source language.
+The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
+Google search returns 11800 results for the message above and even more (44100 results) for a
+LINQ error message \emph{``Method X has no supported translation to SQL``} caused by a similar 
+limitation.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Context awareness \#3: Confidentiality and provenance}
+
+The previous two examples were related to the non-existence of some library functions in another
+environment. Another common factor was that they were related to the execution context of the whole 
+program or a scope. However, contextual properties can be also related to specific variables.
+
+For example, consider the following code sample that accesses database by building a SQL query
+using string concatenation:
 %
@@
 \begin{array}{l}
-\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
-  \vdash
-    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
+\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
+\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
+\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
 \end{array}
@@
 %
-In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
-Before looking at the formal definition, consider the typing of let bindings:
+The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
+(an attacker could enter \str{'; DROP TABLE Products -{}-} as their \ident{name} and delete the 
+database table with products). For this reason, most libraries discourage building SQL commands by
+string concatenation, but there are still many systems that do so.
+
+The example demonstrates a more general property. Sometimes, it is desirable to track additional 
+meta-data about variables that are in some ways special. Such meta-data can determine how the variables
+can be used. Here, \ident{name} comes from the user input. This \emph{provenance} information should 
+be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
+not directly contain user input (in an unchecked form). Such marking of values (but at run-time)
+is also called tainting \cite{app-tainting-sql}.
+
+Similarly, if we had \ident{password} or \ident{creditCard} variables in a client/server web application,
+these should be annotated as sensitive and it should not be possible to send their values over an
+unsecured network connection.
+
+In another context, when working with data (\eg~in data journalism), it would be desirable to 
+track meta-data about the quality and the source of the data. For example, is the source trustworthy?
+Is the data up-to-date? Such meta-data could propagate to the result and tell us important 
+information about the calculated results.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Context-awareness \#4: Checking array access patterns}
+
+The last example leaves the topic of cross-platform and distributed computing. We focus on 
+checking how arrays are accessed. This is a simpler version of the data-flow programming
+examples used later in the thesis.
+
+Consider a simple programming language with arrays where $n^{\textnormal{th}}$ element of an
+array \ident{arr} is accessed using $\ident{arr}[n]$. Furthermore, we focus on performing local
+transformations and we assume that the keyword $\kvd{cursor}$ returns the \emph{current} location
+in the array.
+
+The following example implements a simple one-dimensional cellular automata, reading from the
+\ident{input} array and writing to \ident{output}:
 %
@@
 \begin{array}{l}
-\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
-\pastval{a}{15}+\pastval{c}{10}
+\kvd{let}~\ident{sum}=\ident{input}[\kvd{cursor} - 1] ~+~ \ident{input}[\kvd{cursor}] ~+~ \ident{input}[\kvd{cursor} + 1]\\[-0.25em]
+\kvd{if}~\ident{sum}=2 ~||~ (\ident{sum}=1 ~\&\&~ \ident{input}[\kvd{cursor} - 1] = 0)\\[-0.25em] 
+\kvd{then}~\ident{output}[\kvd{cursor}] \leftarrow 1 ~\kvd{else}~\ident{output}[\kvd{cursor}] \leftarrow 0
 \end{array}
@@
 %
-The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
-is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
-The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.
+In this example, we use the term \emph{context} to refer to the values in the array around the 
+current location provided by \kvd{cursor}. The interesting question is, how much of the context
+(\ie~how far in the array) does the program access.
 
-The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
-context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
-of $\cvop$ and it denotes empty coeffect. The typing context of the
-body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.
+This is contextual information attached to individual (array) variables. 
+In the above example, we want to track that \ident{input} is accessed in the range $\langle-1, 1 \rangle$
+while \ident{output} is accessed in the range $\langle 0, 0 \rangle$. When calculating the ranges,
+we need to be able to compose ranges $\langle-1,-1$, $\langle 0,0 \rangle$ and $\langle 1,1 \rangle$
+(based on the accesses on the first line).
 
-To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
-of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
-with the remaining tags from the body yielding the overall context:
-$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
-normalization mechanism (described later), this can be further reduced to
-$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
-we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
-\ident{b}.
+The information about access patterns can be used to efficiently compile the computation
+(as we know which sub-range of the array might be accessed) and it also allows better handling
+of boundaries. For example, wrap-around behaviour we could pad the \ident{input} with a known
+number of elements from the other side of the array.
 
-%===================================================================================================
+% ==================================================================================================
 
-\section{Contributions}
+\section{Towards context-aware languages}
+
+The four examples presented in the previous section cover different notions of \emph{context}.
+The context can be viewed as execution environment, capabilities provided by the environment 
+or input and meta-data about the input.
+
+The different notions of context can be broadly classified into two categories -- those that
+speak about the environment and those that speak about individual inputs (variables). In this
+thesis, we refer to them as \emph{flat coeffects} and \emph{structural coeffects}, respectively:
+
+\begin{itemize}
+\item \textbf{Flat coeffects} represent additional data, resources and meta-data that are 
+   available in the execution environment (regardless of how they are accessed in a program).
+   Examples include resources such as GPS sensors and battery status (on a phone), databases 
+   (on the server), or framework version.
+
+\item \textbf{Structural coeffects} capture additional meta-data related to inputs. This can include
+   provenance (source of the input value), usage information (how often is the value
+   accessed and in what ways) or security information (whether it contain sensitive data or not). 
+\end{itemize}
+%
+This thesis follows the tradition of statically typed programming languages. As such, we
+attempt to capture such contextual information in the type system of context-aware programming
+languages. The type system should provide both safety guarantees (as in the first three examples)
+and also static analysis useful for optimization (as in the last example).
+
+Although the main focus of this thesis is on the underlying theory of \emph{coeffects} and on
+their structure, the following section briefly demonstrates the features that a practical
+context-aware language, based on the theory of coeffects, can provide.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Context-aware language in action}
+
+As an example, consider a news reader consisting of server-side (which stores the news in a 
+database) and a number of clients applications for popular platforms (Android, Windows Phone, etc.). 
+A simplified code excerpt that might appear somewhere in the implementation is shown in
+Figure~\ref{fig:introduction-sample}.
+
+We assume that the language supports cross-compilation and splits the single program into three
+components: one for the server-side and two for the client-side, for iPhone and Windows platforms, 
+respectively. The cross-compilation could be done in a way similar to Links \cite{app-distributed-links},
+but we do not require explicit annotations specifying the target platform.
+
+If we were writing the code using current main-stream technologies, we would have to create three 
+completely separate components. The server-side would include the \ident{fetchNews} function, which
+queries the database. The iPhone version would include \ident{fetchLocalNews}, which gets the current
+GPS location and performs a call to the remote server and \ident{iPhoneMain}, which constructs the
+user-interface. For Windows, we would also need \ident{fetchLocalNews}, but this time with 
+\ident{windowsMain}. When using a language that can be compiled for all of the platforms, we would need
+a number of \prepk{\#if} blocks to delimit the platform-specific parts.
+
+To support cross-compilation, the language needs to be context-aware. Each of the function has a 
+number of context requirements. The \ident{fetchNews} function needs to have access to a database;
+\ident{fetchLocalNews} needs access to a GPS sensor and to a network (to perform the remote call).
+However, it does not need a specific platform -- it can work on both iPhone and Windows. The last
+two platform-specific functions inherit the requirements of \ident{fetchLocalNews} and additionally
+also require a specific platform.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{fetchNews}(\ident{loc}) =\\[-0.25em]
+\hspace{1em} \kvd{let}~\ident{cmd} = \ident{sprintf}~\str{SELECT * FROM News WHERE Location='\%s'}~\ident{loc}\\[-0.25em]
+\hspace{1em} \ident{query}(\ident{cmd}, \ident{password})\\[-0.25em]
+\\[-1.0em]
+\kvd{let}~\ident{fetchLocalNews}() =\\[-0.25em]
+\hspace{1em} \kvd{let}~\ident{loc} = \ident{gpsLocation}()\\[-0.25em]
+\hspace{1em} \kvd{remote}~\ident{fetchNews}(\ident{loc})\\[-0.25em]
+\\[-1.0em]
+\kvd{let}~\ident{iPhoneMain}() = \\[-0.25em]
+\hspace{1em} \ident{createCocoaListing}(\ident{fetchLocalNews})\\[-0.25em]
+\\[-1.0em]
+\kvd{let}~\ident{windowsMain}() = \\[-0.25em]
+\hspace{1em} \ident{createMetroListing}(\ident{fetchLocalNews})
+\end{array}
+\end{equation*}
+
+\vspace{1em}
+\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
+\vspace{-1.5em}
+\caption{News reader implemented in a context-aware language}
+\label{fig:introduction-sample}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Understanding context with types}
+
+The approach advocated in this thesis is to track information about context requirements using
+the type system. To make this practical, the system needs to provide at least partial support for 
+automatic type inference, as the information about context requirements makes the types more complex.
+An inspiring example might be the F\# support for units of measure \cite{app-units} -- the user has
+to explicitly annotate constants, but the rest of the information is inferred automatically.
+
+Furthermore, integrating contextual information into the type system can provide information for
+modern developer tools. For example, many editors for F\# display inferred types when placing mouse
+pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
+%
+\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
+\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
+\newcommand{\introfunc}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \{ #2 \}} $\rightarrow$ \ident{#3}  }
+\newcommand{\introvar}[2]{ \ident{#1 \hspace{-1em}\footnotesize{@} #2 } }
+\vspace{0.5em}
+\begin{equation*}
+\hspace{5em}
+\begin{array}{l}
+\ident{\textbf{fetchLocalNews}}\\
+\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
+	\introfunc{unit}{gps, rpc}{(news~list)~async}
+}
+\end{array}
+\end{equation*}
+\vspace{0.5em}
+
+Here, we use the notation $\tau_1 \,\ident{\hspace{-1em}\footnotesize{@}}\, c \rightarrow \tau_2$ to denote 
+a function that takes an input of type $\tau_1$, produces a result of type $\tau_2$ and has additional 
+context requirements specified by $c$. In the above example, the annotation $c$ is simply a set of 
+required resources or capabilities. However, a more complex structure could be used as well, for example,
+including the Android API level.
+
+The following summary shows the types of the functions from the code sample in 
+Figure~\ref{fig:introduction-sample}. These guide the code generation by specifying which 
+function should be compiled for which of the platforms, but they also provide documentation
+for the developers:
+
+\begin{equation*}
+\begin{array}{lcl}
+\ident{password} &:& $\introvar{string}{sensitive}$\\[-0.25em]
+\ident{fetchNews} &:& $\introfunc{location}{database}{news~list}$\\[0.5em]
+\ident{gpsLocation} &:& $\introfunc{unit}{gps}{location}$\\[-0.25em]
+\ident{fetchLocalNews} &:& $\introfunc{location}{gps, rpc}{news~list}$\\[0.5em]
+\ident{iPhoneMain} &:& $\introfunc{unit}{ cocoa, gps, rpc}{unit}$\\[-0.25em]
+\ident{windowsMain} &:& $\introfunc{unit}{ windows, gps, rpc}{unit}$
+\end{array}
+\end{equation*}
+
+As mentioned earlier, the concrete syntax used here is just for illustration. Furthermore,
+some information could even be mapped to other visual representations -- for example, differently
+coloured backgrounds for platform-specific functions. The key point is that the type provides a
+number of useful information:
+
+\begin{itemize}
+\item The \ident{password} variable is available in the context (we assume it has been declared 
+  earlier), but is marked as sensitive, which restricts how it can be used. In particular, we 
+  cannot return it as a result of a function that is called via a remote call (\eg~\ident{fetchNews})
+  as that would leak sensitive data over an unsecured connection.
+
+\item The \ident{fetchNews} function requires database access and so it can only run on the server-side
+  (or on a thick client with local copy of the database, such as a desktop computer with an offline mode).
+
+\item The \ident{gpsLocation} function accesses the GPS sensor and since we call it in 
+  from \texttt{fetchLocalNews}, this function also requires GPS (the requirement is propagated 
+  automatically).
+
+\item We can compile the program for two client-side platforms - the entry points are \ident{iPhoneMain}
+  and \ident{windowsMain} and require Cocoa and Windows user-interface libraries, together with
+  GPS and the ability to perform remote calls over the network.
+\end{itemize}
+
+The details of how the cross-compilation would work are out of the scope of this thesis. However,
+one can imagine that the compiler would take multiple sets of references (representing the different 
+platforms), expose the \emph{union} of the functions, but annotate each with the required platform.
+Then, it would produce multiple different binaries -- here, one for the server-side (containing
+\ident{fetchNews}), one for iPhone and one for Windows.
+
+In this scenario, the main benefit of using an integrated context-aware language would be the 
+ability to design appropriate abstractions using standard mechanisms of the language. For 
+cross-compilation, we can structure code using functions, rather than relying on \prepk{\#if}
+directives. Similarly, the splitting between client-side, server-side and shared code can be
+done using ordinary functions and modules -- rather than having to split the application into 
+separate independent libraries or projects.
+
+The purpose of this section was to show that many modern programs rely on the context in which 
+they execute in non-trivial ways. Thus designing context-aware languages is an important 
+practical problem for language designers. The sample serves more as a motivation than as a 
+technical background for this thesis. We explore more concrete examples of properties that 
+can be tracked using the systems developed in this thesis in Chapter~\ref{ch:applications}.
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Theory of context dependence}
+
+The previous section introduced the idea of context-aware languages from the practical perspective.
+As already discussed, we approach the problem from the perspective of statically typed programming
+languages. This section outlines how can contextual information be integrated into the 
+standard framework of static typing. This section is intended only as an informal overview and the
+related work is discussed in Chapter~\ref{ch:pathways}.
+
+\paragraph{Type systems.}
+A type system is a form of static analysis that is usually specified by \emph{typing judgements}
+such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
+the context $\Gamma$, the expression $e$ has a type $\tau$. 
+The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an 
+expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int}
+or \ident{float}, depending on the types of the variables. For example, the following is a valid
+typing judgement in F\#:
+%
+\begin{equation*}
+x\-:\-\ident{int},~y\-:\-\ident{int} \vdash x+y : \ident{int}
+\end{equation*}
+%
+This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
+be \ident{int}. The expression would also be typeable in a context $x\-:\-\ident{int},~y\-:\-\ident{int}$,
+but not, for example, in a context where $x$ has a type \ident{unit}.
+
+\paragraph{Tracking evaluation effects.}
+Type systems can be extended in numerous ways. The types can be more precise, for example, by specifying
+the range of an integer. However, it is also possible to track what program \emph{does} when executed.
+In ML-like languages, the following is a valid judgement:
+%
+\begin{equation*}
+x\-:\-\ident{int} \vdash \ident{print}~x : \ident{unit}
+\end{equation*}
+%
+The judgement states that the expression $\ident{print}~x$ has a type \ident{unit}. This is correct, 
+but it ignores the important fact that the expression has a \emph{side-effect} and prints a number
+to the console. In purely functional languages, this would not be possible. For example, in Haskell,
+the type would be $\ident{IO unit}$ meaning that the result is a \emph{computation} that performs I/O
+effects and then returns \ident{unit} value.
+
+Another option for tracking effects is to extend the judgement with additional information 
+about the effects. The judgement in a language with effect system would look as follows:
+%
+\begin{equation*}
+x\-:\-\ident{int} \vdash \ident{print}~x : \ident{unit~\&~\{ console \}}
+\end{equation*}
+%
+Effect systems add \emph{effect annotation} as another component of the typing judgement. In the above
+example, the return type is \ident{unit}, but the effect annotation informs us that the expression
+also accesses \ident{console} as part of the evaluation. To track such information, the compiler needs
+to understand the effects of primitive built-in functions -- such as \ident{print}.
+
+The crucial part of type systems is dealing with different forms of composition. For example, assume
+we have a function \ident{read} that reads from the console and a function \ident{send} that sends
+data over the network. In that case, the type system should correctly infer that the effects of an 
+expression $\ident{send}(\ident{read}())$ are $\{ \ident{console}, \ident{network} \}$.
+
+Effect systems are an established idea, but they are suitable only for tracking properties of a
+certain kind. They can be used for properties that describe how programs \emph{affect} the environment.
+For context-aware languages, we instead need to track what programs \emph{require} from the environment.
+
+\paragraph{Tracking context requirements.}
+
+The systems for tracking of context requirements developed in this thesis are inspired by the idea
+of effect systems. To demonstrate our approach, consider the following call from the sample program
+shown earlier -- first using standard ML-like type system:
+%
+\begin{equation*}
+\ident{password}\-:\-\ident{string},~\ident{cmd}\-:\-\ident{string} 
+  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
+\end{equation*}
+%
+The expression queries a database and gets back a list of news values as the result. Recall from the
+earlier discussion that there are two contextual information that are desirable to track for this 
+expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
+the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
+an unsecure network connection.
+The \emph{coeffect systems} developed in this thesis capture this information in the following way: 
+%
+\begin{equation*}
+\begin{array}{l}
+ ( \ident{password}\-:\-\ident{string \hspace{-1em}{\footnotesize @ sensitive}},
+   \ident{cmd}\-:\-\ident{string})
+ \hspace{-0.5em}
+ \ident{\footnotesize @ \{ database \}} \vdash \\[-0.1em]
+ \hspace{3em} \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
+\end{array}
+\end{equation*}
+%
+Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
+context $\Gamma$. In other words, coeffect systems track more detailed information about the context,
+not just the available variables. In the above example, it tracks meta-data about the variables
+and annotates \ident{password} as sensitive. Furthermore, it tracks requirements about the execution
+environment -- for example, that the execution requires an access to database.
+
+The example demonstrates the two kinds of coeffect systems outlined earlier. The tracking of 
+\emph{whole-context} information (such as environment requirements) is captured by the \emph{flat 
+coeffect calculus} developed in Chapter~\ref{ch:flat}, while the tracking of \emph{per-variable} 
+information is captured by the \emph{structural coeffect calculus} developed in Chapter~\ref{ch:structural}.
+
+As mentioned earlier, it is well-known fact that \emph{effects} correspond to \emph{monads} and
+languages such sa Haskell use monads to provide a limited form of effect system. An interesting 
+observation made in this thesis is that \emph{coeffects}, or systems for tracking contextual
+information, correspond to the category theoretical dual of monads called \emph{comonads}. The
+details are explained when discussing the semantics of coeffects throughout the thesis.
+
+% ==================================================================================================
+
+\section{Thesis outline}
+
+The key claim of this thesis is that programming languages need to provide better ways of capturing
+how programs rely on the context in which they execute. This chapter shows why this is an important
+problem. We looked at a number of properties related to context that are currently handled in 
+ad-hoc and error-prone ways. Next, we considered the properties in a simplified, but realistic 
+example of a client/server application for displaying local news.
+
+Tracking of contextual properties may not be initially perceived as a major problem -- perhaps 
+because we are so used to write code in certain ways that prevent us from seeing the flaws. The 
+purpose of this chapter was to uncover the flaws and convince the reader that there should be a 
+better solution. Finding the foundations of such better solution is the goal of this thesis:
+
+\begin{itemize}
+\item In Chapter~\ref{ch:pathways} we give an overview of related work. Most importantly, we show
+  that the idea of context-aware computations can be naturally approached from a number of 
+  directions developed recently in theories of programming languages. Chapter~\ref{ch:applications}
+  follows by showing practical motivation for coeffects -- we look at a number of systems that 
+  can be captured using the systems developed later.
+
+\item In Chapter~\ref{ch:pathways} and Chapter~\ref{ch:applications}, we present the key novel
+  contributions of this thesis. We develop the \emph{flat} and \emph{structural} calculi, show 
+  how they capture important contextual properties and develop their categorical semantics using
+  a notion based on comonads. Chapter~\ref{ch:unified} links the two systems into a single 
+  formalism that is capable of capturing both flat and structural properties.
+
+\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis, but 
+  one important direction deserves further exploration. In Chapter~\ref{ch:meta}, we look at 
+  a different approach to tracking contextual information that arises from modal logics. 
+  Finally, Chapter~\ref{ch:conclusions} discusses approaches for implementing the presented theory
+  in main-stream programming languages and concludes.
+\end{itemize}

### 2014-03-27 text/Chapter02.tex
Finish introduction chapter and restructure the rest

--- 2013-06-04 text/Chapter02.tex
@@
 \chapter{Pathways to coeffects} 
-\label{ch:related-work}
+\label{ch:pathways}
 
@@
 
-%===================================================================================================
-
-\section{Through applications}
-\label{sec:path-apps}
-
-The general theme of this thesis is improving programming languages to better support writing
-\emph{context-dependent} (or \emph{context-aware}) computations. With current trends in the 
-computing industry such as mobile and ubiquitous computing, this is becoming an important topic.
-In software engineering and programming community, a number of authors have addressed this problem
-from different perspectives. Hirschfeld et al. propose \emph{Context-Oriented Programming} (COP)
-as a methodology \cite{app-cop-method}, and the subject has also been addressed in mobile
-computations \cite{app-cop-mobile,app-cop-mobile2}. In programming languages, Costanza 
-\cite{app-cop-contextl} develops a domain-specific LISP-like language ContextL and Bardram 
-\cite{app-cop-javafwk} proposes a Java framework for COP.
-
-We approach the problem from a different perspective, building on the tradition of 
-statically-typed functional programming languages and their theories. However, even in this field,
-there is a number of calculi or language features that can be viewed as context-dependent.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Motivation for flat coeffects}
-
-In a number of systems, the execution environment provides some additional data, resources or 
-information about the execution context, but are independent of the variables used by the 
-program. We look at implicit parameters and rebindable resources (that both provide additional
-identifiers that can be accessed similarly to variables, but follow different scoping rules),
-distributed programming, cross-compilation and data-flow.
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Implicit parameters} In Haskell, implicit parameters \cite{app-implicit-parameters} are 
-a special kind of variables that may behave as dynamically scoped. This means, if a function uses 
-parameter $\ident{?p}$, then the caller of the function must define $\ident{?p}$ and set its value.
-Implicit parameters can be used to parameterise a computation (involving a chain of function calls)
-without passing parameters explicitly as additional arguments of all involved functions. A simple 
-language with implicit parameters has an expression $\ident{?p}$ to read a parameter value and an 
-expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a different keyword to 
-avoid confusion.} $\kvd{letdyn}~?p = e_1~\kvd{in}~e_2$ that sets a parameter $\ident{?p}$ to the value of $e_1$ 
-and evaluates $e_2$ in a context containing $\ident{?p}$
-
-An interesting question arises when we use implicit parameters in a nested function. The following 
-function does some pre-processing and then returns a function that builds a formatted string based 
-on two implicit parameters $\ident{?width}$ and $\ident{?size}$:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\
-\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\
-\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
-         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})\\
-\end{array}
-\end{equation*}
-%
-The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context 
-$\{ \ident{?width} : \ident{int} \}$. The nested function (returned as a result) uses the parameter 
-$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters of the nested 
-function come from?
-
-In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
-However, in Haskell, implicit parameters behave as a combination of lexical and dynamic scoping. This means
-that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$
-In Haskell, this corresponds to the following type:
-%
-\begin{equation*}
-(\ident{?width} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow 
-  ((\ident{?size} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow \ident{string})
-\end{equation*}
-%
-As a result, the function can be called as follows:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{formatHello} = \\
-\quad(~\kvd{letdyn}~\ident{?width}=5~\kvd{in}\\
-\quad~~\ident{format}~\texttt{"Hello"})~~\kvd{in}\\
-\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}
-\end{array}
-\end{equation*}
-%
-This way of assigning type to \ident{format} and calling it is not the only possible, though. 
-We could also say that the outer function requires both of the implicit parameters and the result
-is a (pure) function with no context requirements. This interaction between implicit parameters 
-and lambda abstraction demonstrates one of the key aspects of coeffects and will be discussed 
-later. Implicit parameters will also sever as one of our examples in Chapter~Y.
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Type classes}
-Implicit parameters are closely related to \emph{type classes} \cite{app-type-classes}. In Haskell,
-type classes provide a principled form of ad-hoc polymorphism (overloading). When a code uses 
-an overloaded operation (e.g.~comparison or numeric operators) a constraint is placed on the 
-context in which the operation is used. For example:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{twoTimes}~::~\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha \\
-\ident{twoTimes}~x=x+x
-\end{array}
-\end{equation*}
-%
-The constraint $\ident{Num}~a$ on the function type arises from the use of the $+$ operator. 
-From the implementation perspective, the type class constraint means that the function takes 
-a hidden parameter -- a dictionary that provides the operation $+ :: \alpha \rightarrow \alpha \rightarrow \alpha$.
-Thus, the type $\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha$ can be viewed as
-$(\ident{Num}_\alpha \times \alpha) \rightarrow \alpha$. Implicit parameters work in exactly
-the same way -- they are passed around as hidden parameters.
-
-The implementation of type classes and implicit parameters shows two important points about 
-context-dependent properties. First, they are associated with some \emph{scope}, such as the body
-of a function. Second, they are associated with the input. To call a function that takes an 
-implicit parameter or has a type-class constraint, the caller needs to pass a (hidden) parameter
-together with the function inputs.
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Rebindable resources}
-The need for parameters that do not strictly follow static scoping rules also arises in distributed
-computing. This problem has been addressed, for example, by Bierman et al. and Sewell et al. 
-\cite{app-distributed-rebinding,app-distributed-acute}. To quote the first work: \emph{``Dynamic 
-binding is required in various guises, for example when a marshalled value is received from the 
-network, containing identifiers that must be rebound to local resources.''}
-
-This situation arises when marshalling and transferring function values. A function may depend 
-on a local resource (e.g.~a database available only on the server) and also resources that are 
-available on the target node (e.g.~current time). In the following example, the construct
-$\kvd{access}~\ident{Res}$ represents access to a re-bindable resource named $\ident{Res}$:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\
-\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\
-\quad\ident{query}~\ident{db}~\str{"SELECT * WHERE Date > \%1"}~(\kvd{access}~\ident{Clock})
-\end{array}
-\end{equation*}
-%
-When \ident{recentEvents} is created on a server and sent to a client, a remote reference to the 
-database (available only on the server) must be captured. If the client device supports a clock, 
-then \ident{Clock} can be locally \emph{rebound}, e.g., to accommodate time-zone changes. 
-Otherwise, the date and time needs to be obtained from the server too.
-
-The use of re-bindable resources creates a context requirement similar to the one arising from
-the use of implicit parameters. For function values, such context-requirements can be satisfied
-in different ways -- resources must be available either at the declaration site (i.e.~when a 
-function is created) or at the call site (i.e.~when a function is called).
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Distributed computing and multi-targetting}
-
-An increasing number of programming languages is capable of running across multiple different 
-platforms or execution environments. Functional programming languages that can be compiled to
-JavaScript (to target web and mobile clients) include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.
-
-Links \cite{app-distributed-links}, F\# libraries \cite{app-fsharp-webapps,app-fsharp-webtools},
-ML5 and QWeSST \cite{app-distributed-ml5, app-distributed-qwesst} and Hop \cite{app-hop-lang} go 
-further and allow a single source program to be compiled to multiple target runtimes. This posses 
-additional challenges -- it is necessary to track where each part of computation runs and statically 
-guarantee that it will be possible to compile code to the required target platform 
-(safe \emph{multi-targetting}).
-
-We demonstrate the problem by looking at input validation. In distributed applications 
-that communicate over unsecured HTTP channel, user input needs to be validated interactively
-on the client-side (to provide immediate response) and then again on the server-side (to 
-guarantee safety). For example:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\
-\quad\ident{name} \neq \str{""} ~~\&\&~~ \ident{forall~isLetter~name}
-\\[0.5em]
-\kvd{let}~\ident{displayProduct} = \lambda \ident{name} \rightarrow\\
-\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{displayProductPage~name}\\
-\quad\kvd{else}~\ident{displayErrorPage}~() 
-\end{array}
-\end{equation*}
-%
-The function \ident{validateInput} can be compiled to both JavaScript (for client-side) and
-native code (for server-side). However, \ident{displayProduct} uses other functionality
-(generating web pages) that is only available on the server-side, so it can only be compiled to
-native code.
-
-In Links \cite{app-distributed-links}, functions can be annotated as client-side, server-side
-and database-side. F\# WebTools \cite{app-fsharp-webtools} adds functions that support multiple
-targets (mixed-side). However, these are single-purpose language features and they are not 
-extensible. For example, in modern mobile development it is also important to track minimal 
-supported version of runtime\footnote{Android Developer guide \cite{app-android-multitarget} 
-demonstrates how difficult it is to solve the problem without language support.}.
-
-Requirements on the execution environment can be viewed as contextual properties, but could be
-also presented as effects (use of some API required only in certain environment is a computational
-effect). We discuss the difference in Section~X. Furthermore, the theoretical foundations of
-distributed languages like ML5 \cite{app-distributed-ml5} suggest that a contextual treatment
-is more appropriate. We return to ML5 when discussing semantics in Section~\ref{sec:path-sem-contextdep}.
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Data-flow languages}
-
-The examples discussed so far are all -- to some extent -- similar. They attach additional 
-information (implicit parameters, dictionaries) or restrictions (on execution environment) to the
-context where code evaluates. By \emph{context}, we mean, most importantly, the values of variables
-and declarations that are in scope. The examples so far add more information to the context, but
-do not operate on the variable values.
-
-Data-flow languages provide a different example. Lucid \cite{app-lucid} is a declarative data-flow 
-language designed by Wadge and Ashcroft. In Lucid, variables represent streams and programs
-are written as transformations over streams. A function application $\mathit{square}(a)$ represents
-a stream of squares calculated from the stream of values $a$.
-
-The data-flow approach has been successfully used in domains such as development of real-time embedded 
-application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the data-flow
-paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
-and implements program to count the number of edges on a Boolean stream:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{edge} = \ident{false}~\kvd{fby}~(\ident{input}~\&\&~\ident{not}~(\kvd{prev}~\ident{input}))
-\\[0.5em]
-\kvd{let}~\ident{edgeCount} = \\
-\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~\kvd{prev}~\ident{edgeCount}\\
-\quad\quad\quad~~~\, \kvd{else}~\kvd{prev}~\ident{edgeCount} ~)
-\end{array}
-\end{equation*}
-%
-The construct $\kvd{prev}~x$ returns a stream consisting of previous values of the stream 
-$x$. The second value of $\kvd{prev}~x$ is first value of $x$ (and the first
-value is undefined). The construct $y~\kvd{fby}~x$ returns a stream whose first element is the 
-first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
-such as \ident{false} and $0$ are constant streams. Formally, the construct are defined as follows
-(writing $x_n$ for $n$-th element of a stream $x$):
-%
-\[ 
-(\kvd{prev}~x)_n = \left\{ 
-  \begin{array}{ll}
-    nil     & \; \text{if $n=0$}\\
-    x_{n-1} & \; \text{if $n>0$}
-  \end{array} \right.
-\quad
-(y~\kvd{fby}~x)_n = \left\{ 
-  \begin{array}{ll}
-    y_0     & \; \text{if $n=0$}\\
-    x_n     & \; \text{if $n>0$}
-  \end{array} \right.
-\]  
-%
-When reading data-flow programs, we do not need to think about variables in terms of streams --
-we can see them as simple values. However, the operations \kvd{fby} and \kvd{prev} cannot operate
-on plain values -- they require additional \emph{context} which provides past values of variables
-(for \kvd{prev}) and information about the current location in the stream (for \kvd{fby}).
-
-In this case, the context is not simply an additional (hidden) parameter. It completely changes
-how variables must be represented. We may want to capture various \emph{contextual properties}
-of Lucid programs. For example, how many past elements need to be cached when we evaluate the 
-stream.
-
-To understand the nature of the context, we later look at the semantics of Lucid. This can be
-captured using a number of mathematical structures. Wadge \cite{app-lucid-monads} originally 
-proposed to use monads, while Uustalu and Vene later used comonads \cite{app-dataflow-essence}.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Motivation for structural coeffects}
-
-We now turn our attention to system where additional contextual information are associated not
-with the context as a whole (or program scope), but with individual variables. We start by looking
-simple static analysis -- variable \emph{liveness}. Then we revisit data-flow computations and
-look at applications in security and software updating.
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Liveness analysis}
-
-\emph{Live variable analysis} (LVA) \cite{app-modern-compiler} is a standard technique in compiler theory. 
-It detects whether a free variable of an expression may be used by a program later (it is
-\emph{live}) or whether it is definitely not needed (it is \emph{dead}). As an optimization, 
-compiler can remove bindings to dead variables as the result is never accessed. Wadler 
-\cite{app-strictness-absecnce} describes the property of a variable that is dead as the 
-\emph{absence} of a variable.
-
-In this thesis, we first use a restricted (and not practically useful) form of liveness analysis
-to introduce the theory of indexed comonads (Section~X) and then use liveness analysis as one of the
-motivations for structural coeffects. Consider the following two simple functions:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
-\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
-\end{array}
-\end{equation*}
-%
-In liveness analysis, we annotate the context with a value specifying whether the variables in
-scope are \emph{live} or \emph{dead}. If we associate just a single value with the entire 
-context, then the liveness analysis is very limited -- it can say that the context of the 
-expression $42$ in the first function is dead, because no variables are accessed.
-
-A useful liveness analysis needs to consider individual variables. For example, in the body of
-the second function (\ident{value}), two variables are in scope. The variable \ident{value} is
-accessed and thus is \emph{live}, but the variable \ident{x} is dead.
-
-Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
-propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
-guarantee). Liveness is a \emph{backward} analysis -- this means that the requirements propagates
-from variables to their declaration sites. The distinction between \emph{must} and \emph{may} is 
-apparent when we look at an example with conditionals:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
-\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
-\end{array}
-\end{equation*}
-%
-The liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
-\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
-\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
-only if it was used in both of the branches (this is sometimes called \emph{neededness}).
-
-The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
-interaction between contextual properties and certain language constructs such as conditionals.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Data-flow languages (revisited)}
-When discussing data-flow languages in the previous section, we said that the context provides 
-past values of variables. This can be viewed as a flat contextual property (the context needs
-to keep all past values), but we can also view it as a structural property. Consider the following
-example:
-%
-\begin{equation*}
-\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
-\end{equation*}
-%
-The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
-To evaluate a current value of the stream, we need the current value of \ident{left} and one past
-value of \ident{right}.
-
-As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
-values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
-single number for each function. However, we can design a more precise \emph{structural} analysis
-and track the number of required elements for individual variables.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Tainting and provenance}
-Tainting is a mechanism where variables coming from potentially untrusted sources are marked
-(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
-can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
-(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
-case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
-values are annotated with more detailed information about their source.
-
-Statically typed systems that based on tainting have been use to prevent cross-site scripting
-attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
-\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
-cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
-type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
-\ident{msg}~+~\ident{name}
-\end{array}
-\end{equation*}
-%
-In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
-untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
-The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
-(i.e.~to construct an SQL query).
-
-In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
-flag (for tainting) or more complex data (for provenance). In static checking, the information
-need to be associated with the variables in the variable context. We use tainting as a motivating
-example for \emph{structural} coeffects in Section~X.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Security and core dependency calculus}
-
-The checking of tainting is a special case of checking of the \emph{non-interference} property 
-in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
-as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
-network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
-system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
-more recent work. The checking of information flows has been also integrated (as a single-purpose
-extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
-Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
-using a monadic library.
-
-Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
-where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\ident{L}, \ident{H}\}$ 
-represents low and high secrecy, respectively with $\ident{L} \leq \ident{H}$ meaning that low security
-values can be treated as high security (but not the other way round).
-
-An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
-example which may assign a new value to $z$:
-%
-\begin{equation*}
-\kvd{if}~x>0~\kvd{then}~z := y
-\end{equation*}
-%
-If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
-(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
-$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
-assignment is performed or not performed leaks information in its own (this is an 
-\emph{implicit} flow).
-
-Abadi et al. realized that there is a number of analyses similar to secure information flow
-and proposed to unify them using a single model called Dependency Core Calculus (DCC) \cite{app-dcc}.
-It captures other cases where some information about expression relies on properties of variables
-in the context where it executes.  The DCC captures, for example, \emph{binding time analysis}
-\cite{app-binding-time-analysis}, which detects which parts of programs can be partially evaluated
-(do not depend on user input) and \emph{program slicing} \cite{app-slicing-survey} that identifies
-parts of programs that contribute to the output of an expression.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Beyond contextual information}
-
-In the systems discussed
-
-\paragraph{Calculus of capabilities}
-
-a
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Software updating}
-Dynamic software updating (DSU) is the ability to update programs at runtime without stopping them.
-
-Stoyle et al. \cite{app-dsu-mutatis}.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Safe locking}
-
-\cite{app-safe-locking}
-
-\paragraph{Thesis perspective}
-?
-
-\newpage
 %===================================================================================================

### 2014-03-27 text/Chapter03.tex
Finish introduction chapter and restructure the rest

--- 2013-06-05 text/Chapter03.tex
@@
-\chapter{Flat coeffects} 
-\label{ch:flat-coeffects}
+\chapter{Context-aware applications} 
+\label{ch:applications}
 
@@
 
-% comonadic semantics
-\newcommand{\iunit}{\mathsf{e}}
-\newcommand{\imult}{\oplus}
-\newcommand{\iftor}[1]{C^{#1}}
-\newcommand{\iftorhat}[1]{\hat{C}^{#1}}
+TODO
 
-\newcommand{\cosem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
-\newcommand{\ccat}[0]{\mathcal{C}}
-\newcommand{\obj}[1]{\textnormal{obj}(#1)}
+%===================================================================================================
 
-\newcommand{\cobind}[2]{#1^\dagger_{#2}}
-\newcommand{\cmerge}[0]{ \ident{m} }
-\newcommand{\csplit}[0]{ \ident{n} }
-\newcommand{\counit}[0]{ \varepsilon }
-% identity function
+\section{Through applications}
+\label{sec:path-apps}
 
-% structural coeffect calculus - operation names
+The general theme of this thesis is improving programming languages to better support writing
+\emph{context-dependent} (or \emph{context-aware}) computations. With current trends in the 
+computing industry such as mobile and ubiquitous computing, this is becoming an important topic.
+In software engineering and programming community, a number of authors have addressed this problem
+from different perspectives. Hirschfeld et al. propose \emph{Context-Oriented Programming} (COP)
+as a methodology \cite{app-cop-method}, and the subject has also been addressed in mobile
+computations \cite{app-cop-mobile,app-cop-mobile2}. In programming languages, Costanza 
+\cite{app-cop-contextl} develops a domain-specific LISP-like language ContextL and Bardram 
+\cite{app-cop-javafwk} proposes a Java framework for COP.
+
+We approach the problem from a different perspective, building on the tradition of 
+statically-typed functional programming languages and their theories. However, even in this field,
+there is a number of calculi or language features that can be viewed as context-dependent.
 
@@
 
-TODO: Change this
+\subsection{Motivation for flat coeffects}
 
-Understanding how programs affect their environment is a well studied area: \emph{effect 
-systems}~\cite{effects-talpin-et-al} provide a static analysis of effects and 
-\emph{monads}~\cite{monad-notions} provide a unified semantics to different notions of effect.  
-Wadler and Thiemann unify the two approaches~\cite{monads-effects-marriage}, \emph{indexing}
-a monad with effect information, and showing that the propagation of effects in an effect system 
-matches the semantic propagation of effects in the monadic approach.
+In a number of systems, the execution environment provides some additional data, resources or 
+information about the execution context, but are independent of the variables used by the 
+program. We look at implicit parameters and rebindable resources (that both provide additional
+identifiers that can be accessed similarly to variables, but follow different scoping rules),
+distributed programming, cross-compilation and data-flow.
 
-No such unified mechanism exists for tracking the context requirements. We use the term 
-\emph{coeffect} for such contextual program properties. 
-Notions of context have been previously captured using comonads
-\cite{comonads-notions} (the dual of monads) 
-and by languages derived from modal logic
-\cite{logic-modal-reconstruction,logic-cmtt}, but these approaches
-do not capture many useful examples which motivate our work. 
-We build mainly on the former comonadic direction (\S\ref{sec:comonads}) and discuss the
-modal logic approach later (\S\ref{sec:related}).
+%---------------------------------------------------------------------------------------------------
 
-We extend a simply typed lambda calculus with a coeffect system based on comonads,
-replicating the successful approach of effect systems and monads.
+\paragraph{Implicit parameters} In Haskell, implicit parameters \cite{app-implicit-parameters} are 
+a special kind of variables that may behave as dynamically scoped. This means, if a function uses 
+parameter $\ident{?p}$, then the caller of the function must define $\ident{?p}$ and set its value.
+Implicit parameters can be used to parameterise a computation (involving a chain of function calls)
+without passing parameters explicitly as additional arguments of all involved functions. A simple 
+language with implicit parameters has an expression $\ident{?p}$ to read a parameter value and an 
+expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a different keyword to 
+avoid confusion.} $\kvd{letdyn}~?p = e_1~\kvd{in}~e_2$ that sets a parameter $\ident{?p}$ to the value of $e_1$ 
+and evaluates $e_2$ in a context containing $\ident{?p}$
 
-\paragraph{Examples of coeffects.}
-We present three examples that do not fit the traditional approach of
-\emph{effect systems} and have not been considered using the \emph{modal logic}
-perspective, but can be captured as \emph{coeffect systems} (\S\ref{sec:motivation}) --
-the tracking of implicit dynamically-scoped parameters (or resources),
-analysis of variable liveness, and tracking the
-number of required past values in dataflow computations.
-
-\paragraph{Coeffect calculus.}
-Informed by the examples, we identify a general algebraic structure for coeffects. 
-From this, we define a general \emph{coeffect calculus} that unifies the motivating 
-examples (\S\ref{sec:calculus}) and discuss its syntactic properties (\S\ref{sec:syntactic}).
-
-\paragraph{Indexed comonads.}
-Our categorical semantics (\S\ref{sec:comonads}) extends the work of Uustalu and Vene 
-\cite{comonads-notions}. By adding annotations, we generalize comonads to \emph{indexed comonads}, 
-which capture notions of computation not captured by ordinary comonads.
-
-% ==================================================================================================
-
-\section{Motivation}
-\label{sec:motivation}
-
-Effect systems, introduced by Gifford and Lucassen
-\cite{effects-gifford}, track \emph{effects} of computations, such as
-memory access or message-based
-communication~\cite{effects-messagepassing}. Their approach augments
-typing judgments with effect information: $\Gamma \vdash e :
-\tau, F$.  Wadler and Thiemann explain how this shapes
-effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
-%
-\begin{quote}
-\emph{In the rule for abstraction, the effect is empty because evaluation immediately
-returns the function, with no side effects. The effect on the function arrow
-is the same as the effect for the function body, because applying the function will
-have the same side effects as evaluating the body.}
-\end{quote}
-%
-In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
-does not match this pattern. In the systems we consider, lambda abstraction places 
-requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
-(immediate requirements), resulting in different syntactic properties. 
-We informally discuss three examples first that demonstrate how contextual
-requirements propagate. Section (\S\ref{sec:calculus}) then unifies these 
-in a single calculus.
-
-We write coeffect judgements $\ctyp{s}{\Gamma} \vdash e : \tau$ where the 
-coeffect annotation $s$ associates context requirements with the 
-free-variable context $\Gamma$. Function types have the form $\ctyp{s}{\tau_1} \rightarrow \tau_2$ associating
-\emph{latent} coeffects $s$ with the parameter.
-The $\ctyp{s}{\Gamma}$ syntax and $\ctyp{s}{\tau}$ types are a result of the
-indexed comonadic semantics (\S\ref{sec:comonads}).
-
-% --------------------------------------------------------------------------------------------------
-\vspace{-1em}
-\subsection{Implicit parameters and resources.}
-\label{sec:coeffects-res}
-
-Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
-They can be used to parameterize a computation without propagating arguments explicitly through a 
-chain of calls and are part of the context in which expressions evaluate. As correctly expected
-\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
-computations follow a similar pattern, but we discuss implicit parameters for simplicity.
-
-The following function prints a number using implicit parameters
-\ident{?culture} (determining the decimal mark) and \ident{?format}
-(the number of decimal places):
+An interesting question arises when we use implicit parameters in a nested function. The following 
+function does some pre-processing and then returns a function that builds a formatted string based 
+on two implicit parameters $\ident{?width}$ and $\ident{?size}$:
 %
 \begin{equation*}
-\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
-\end{equation*}
-%
-Figure~\ref{fig:example-resources} shows a type-and-coeffect system
-tracking the set of an expression's implicit parameters. 
-For simplicity here, all implicit parameters have type $\rho$.
-
-Context requirements are created in (\emph{access}), while (\emph{var}) requires no
-implicit parameters; (\emph{app}) combines requirements of
-both sub-expressions as well as the latent requirements of the
-function. The (\emph{abs}) rule is where the example differs from
-effect systems.  Function bodies can access the union of the
-parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
-at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
-judgements for the above example are:
-%
-\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
-\begin{equation*}
-\begin{array}{rcl}
-\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
-\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
+\begin{array}{l}
+\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\
+\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\
+\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
+         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})\\
 \end{array}
@@
 %
-The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
-Different judgments are desirable depending on how a function is used. 
-In the first case, both parameters have to be provided by the caller.
-In the second, both are available at declaration-site, but \ident{?format} may be
-rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).
+The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context 
+$\{ \ident{?width} : \ident{int} \}$. The nested function (returned as a result) uses the parameter 
+$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters of the nested 
+function come from?
 
-Implicit parameters can be captured by the
-\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
-  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
-modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
-the next example cannot be structured by \emph{any} monad.
+In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
+However, in Haskell, implicit parameters behave as a combination of lexical and dynamic scoping. This means
+that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$
+In Haskell, this corresponds to the following type:
+%
+\begin{equation*}
+(\ident{?width} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow 
+  ((\ident{?size} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow \ident{string})
+\end{equation*}
+%
+As a result, the function can be called as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{formatHello} = \\
+\quad(~\kvd{letdyn}~\ident{?width}=5~\kvd{in}\\
+\quad~~\ident{format}~\texttt{"Hello"})~~\kvd{in}\\
+\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}
+\end{array}
+\end{equation*}
+%
+This way of assigning type to \ident{format} and calling it is not the only possible, though. 
+We could also say that the outer function requires both of the implicit parameters and the result
+is a (pure) function with no context requirements. This interaction between implicit parameters 
+and lambda abstraction demonstrates one of the key aspects of coeffects and will be discussed 
+later. Implicit parameters will also sever as one of our examples in Chapter~Y.
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Type classes}
+Implicit parameters are closely related to \emph{type classes} \cite{app-type-classes}. In Haskell,
+type classes provide a principled form of ad-hoc polymorphism (overloading). When a code uses 
+an overloaded operation (e.g.~comparison or numeric operators) a constraint is placed on the 
+context in which the operation is used. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{twoTimes}~::~\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha \\
+\ident{twoTimes}~x=x+x
+\end{array}
+\end{equation*}
+%
+The constraint $\ident{Num}~a$ on the function type arises from the use of the $+$ operator. 
+From the implementation perspective, the type class constraint means that the function takes 
+a hidden parameter -- a dictionary that provides the operation $+ :: \alpha \rightarrow \alpha \rightarrow \alpha$.
+Thus, the type $\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha$ can be viewed as
+$(\ident{Num}_\alpha \times \alpha) \rightarrow \alpha$. Implicit parameters work in exactly
+the same way -- they are passed around as hidden parameters.
+
+The implementation of type classes and implicit parameters shows two important points about 
+context-dependent properties. First, they are associated with some \emph{scope}, such as the body
+of a function. Second, they are associated with the input. To call a function that takes an 
+implicit parameter or has a type-class constraint, the caller needs to pass a (hidden) parameter
+together with the function inputs.
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Rebindable resources}
+The need for parameters that do not strictly follow static scoping rules also arises in distributed
+computing. This problem has been addressed, for example, by Bierman et al. and Sewell et al. 
+\cite{app-distributed-rebinding,app-distributed-acute}. To quote the first work: \emph{``Dynamic 
+binding is required in various guises, for example when a marshalled value is received from the 
+network, containing identifiers that must be rebound to local resources.''}
+
+This situation arises when marshalling and transferring function values. A function may depend 
+on a local resource (e.g.~a database available only on the server) and also resources that are 
+available on the target node (e.g.~current time). In the following example, the construct
+$\kvd{access}~\ident{Res}$ represents access to a re-bindable resource named $\ident{Res}$:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\
+\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\
+\quad\ident{query}~\ident{db}~\str{"SELECT * WHERE Date > \%1"}~(\kvd{access}~\ident{Clock})
+\end{array}
+\end{equation*}
+%
+When \ident{recentEvents} is created on a server and sent to a client, a remote reference to the 
+database (available only on the server) must be captured. If the client device supports a clock, 
+then \ident{Clock} can be locally \emph{rebound}, e.g., to accommodate time-zone changes. 
+Otherwise, the date and time needs to be obtained from the server too.
+
+The use of re-bindable resources creates a context requirement similar to the one arising from
+the use of implicit parameters. For function values, such context-requirements can be satisfied
+in different ways -- resources must be available either at the declaration site (i.e.~when a 
+function is created) or at the call site (i.e.~when a function is called).
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Distributed computing and multi-targetting}
+
+An increasing number of programming languages is capable of running across multiple different 
+platforms or execution environments. Functional programming languages that can be compiled to
+JavaScript (to target web and mobile clients) include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.
+
+Links \cite{app-distributed-links}, F\# libraries \cite{app-fsharp-webapps,app-fsharp-webtools},
+ML5 and QWeSST \cite{app-distributed-ml5, app-distributed-qwesst} and Hop \cite{app-hop-lang} go 
+further and allow a single source program to be compiled to multiple target runtimes. This posses 
+additional challenges -- it is necessary to track where each part of computation runs and statically 
+guarantee that it will be possible to compile code to the required target platform 
+(safe \emph{multi-targetting}).
+
+We demonstrate the problem by looking at input validation. In distributed applications 
+that communicate over unsecured HTTP channel, user input needs to be validated interactively
+on the client-side (to provide immediate response) and then again on the server-side (to 
+guarantee safety). For example:
+
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\
+\quad\ident{name} \neq \str{""} ~~\&\&~~ \ident{forall~isLetter~name}
+\\[0.5em]
+\kvd{let}~\ident{displayProduct} = \lambda \ident{name} \rightarrow\\
+\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{displayProductPage~name}\\
+\quad\kvd{else}~\ident{displayErrorPage}~() 
+\end{array}
+\end{equation*}
+%
+The function \ident{validateInput} can be compiled to both JavaScript (for client-side) and
+native code (for server-side). However, \ident{displayProduct} uses other functionality
+(generating web pages) that is only available on the server-side, so it can only be compiled to
+native code.
+
+In Links \cite{app-distributed-links}, functions can be annotated as client-side, server-side
+and database-side. F\# WebTools \cite{app-fsharp-webtools} adds functions that support multiple
+targets (mixed-side). However, these are single-purpose language features and they are not 
+extensible. For example, in modern mobile development it is also important to track minimal 
+supported version of runtime\footnote{Android Developer guide \cite{app-android-multitarget} 
+demonstrates how difficult it is to solve the problem without language support.}.
+
+Requirements on the execution environment can be viewed as contextual properties, but could be
+also presented as effects (use of some API required only in certain environment is a computational
+effect). We discuss the difference in Section~X. Furthermore, the theoretical foundations of
+distributed languages like ML5 \cite{app-distributed-ml5} suggest that a contextual treatment
+is more appropriate. We return to ML5 when discussing semantics in Section~\ref{sec:path-sem-contextdep}.
 
@@
 
-\begin{figure}[t]
-\vspace{-1.4em}
+\paragraph{Safe locking}
+In the previous examples, the context provides additional values or functions that may be accessed
+at runtime. However, it may also track \emph{permissions} to perform some operation. This is done
+in the type system for safe locking of Flanagan and Abadi \cite{app-safe-locking}.
+
+The system prevents race conditions (by only allowing access to mutable state under a lock)
+and avoids deadlocks (by imposing strict partial order on locks). The following 
+program uses a mutable state under a lock:
+%
 \begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
+\begin{array}{l}
+\kvd{newlock}~l:\rho~\kvd{in}\\
+\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\
+\kvd{sync}~l~(!\ident{state})
+\end{array}
+\end{equation*}
 %
-\quad\quad
+The declaration \kvd{newlock} creates a lock $l$ protecting memory region $\rho$. We can than
+allocate mutable variables in that memory region (second line). An access to mutable variable
+is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
+which locks a lock and evaluates an expression in a context that contains permission to access
+memory region of the lock ($\rho$ in the above example).
+
+The type system for safe locking associates the list of permission with the variable context.
+It uses judgements of a form $\Gamma, m \vdash e : \alpha$ specifying that an expression has a
+type in context $\Gamma$, given permissions (a list of locked regions) $m$. However, the treatment
+of lambda abstraction differs from the one for implicit parameters or rebindable resources.
+In the system for locking, code inside lambda function cannot use permissions from the scope
+where the function is declared. This is a necessary requirement -- a lambda function created 
+under a lock cannot access protected memory, because it will be executed later. We discuss how
+this restriction fits into our general coeffect framework in Section~X.Y.
+
 %
-\inference[(\emph{app})\;]
-  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
+% => This has coeffect style judgments, but it has effect-style lambda
+%
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Data-flow languages}
+
+The examples discussed so far are all -- to some extent -- similar. They attach additional 
+information (implicit parameters, dictionaries) or restrictions (on execution environment) to the
+context where code evaluates. By \emph{context}, we mean, most importantly, the values of variables
+and declarations that are in scope. The examples so far add more information to the context, but
+do not operate on the variable values.
+
+Data-flow languages provide a different example. Lucid \cite{app-lucid} is a declarative data-flow 
+language designed by Wadge and Ashcroft. In Lucid, variables represent streams and programs
+are written as transformations over streams. A function application $\mathit{square}(a)$ represents
+a stream of squares calculated from the stream of values $a$.
+
+The data-flow approach has been successfully used in domains such as development of real-time embedded 
+application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the data-flow
+paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
+and implements program to count the number of edges on a Boolean stream:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{edge} = \ident{false}~\kvd{fby}~(\ident{input}~\&\&~\ident{not}~(\kvd{prev}~\ident{input}))
+\\[0.5em]
+\kvd{let}~\ident{edgeCount} = \\
+\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~\kvd{prev}~\ident{edgeCount}\\
+\quad\quad\quad~~~\, \kvd{else}~\kvd{prev}~\ident{edgeCount} ~)
+\end{array}
 \end{equation*}
-\vspace{-0.7em}
+%
+The construct $\kvd{prev}~x$ returns a stream consisting of previous values of the stream 
+$x$. The second value of $\kvd{prev}~x$ is first value of $x$ (and the first
+value is undefined). The construct $y~\kvd{fby}~x$ returns a stream whose first element is the 
+first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
+such as \ident{false} and $0$ are constant streams. Formally, the construct are defined as follows
+(writing $x_n$ for $n$-th element of a stream $x$):
+%
+\[ 
+(\kvd{prev}~x)_n = \left\{ 
+  \begin{array}{ll}
+    nil     & \; \text{if $n=0$}\\
+    x_{n-1} & \; \text{if $n>0$}
+  \end{array} \right.
+\quad
+(y~\kvd{fby}~x)_n = \left\{ 
+  \begin{array}{ll}
+    y_0     & \; \text{if $n=0$}\\
+    x_n     & \; \text{if $n>0$}
+  \end{array} \right.
+\]  
+%
+When reading data-flow programs, we do not need to think about variables in terms of streams --
+we can see them as simple values. However, the operations \kvd{fby} and \kvd{prev} cannot operate
+on plain values -- they require additional \emph{context} which provides past values of variables
+(for \kvd{prev}) and information about the current location in the stream (for \kvd{fby}).
+
+In this case, the context is not simply an additional (hidden) parameter. It completely changes
+how variables must be represented. We may want to capture various \emph{contextual properties}
+of Lucid programs. For example, how many past elements need to be cached when we evaluate the 
+stream.
+
+To understand the nature of the context, we later look at the semantics of Lucid. This can be
+captured using a number of mathematical structures. Wadge \cite{app-lucid-monads} originally 
+proposed to use monads, while Uustalu and Vene later used comonads \cite{app-dataflow-essence}.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Motivation for structural coeffects}
+
+We now turn our attention to system where additional contextual information are associated not
+with the context as a whole (or program scope), but with individual variables. We start by looking
+simple static analysis -- variable \emph{liveness}. Then we revisit data-flow computations and
+look at applications in security and software updating.
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Liveness analysis}
+
+\emph{Live variable analysis} (LVA) \cite{app-modern-compiler} is a standard technique in compiler theory. 
+It detects whether a free variable of an expression may be used by a program later (it is
+\emph{live}) or whether it is definitely not needed (it is \emph{dead}). As an optimization, 
+compiler can remove bindings to dead variables as the result is never accessed. Wadler 
+\cite{app-strictness-absecnce} describes the property of a variable that is dead as the 
+\emph{absence} of a variable.
+
+In this thesis, we first use a restricted (and not practically useful) form of liveness analysis
+to introduce the theory of indexed comonads (Section~X) and then use liveness analysis as one of the
+motivations for structural coeffects. Consider the following two simple functions:
+%
 \begin{equation*}
-\inference[(\emph{access})\;]
-  {}
-  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
+\begin{array}{l}
+\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
+\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
+\end{array}
+\end{equation*}
 %
-\quad\quad\quad
+In liveness analysis, we annotate the context with a value specifying whether the variables in
+scope are \emph{live} or \emph{dead}. If we associate just a single value with the entire 
+context, then the liveness analysis is very limited -- it can say that the context of the 
+expression $42$ in the first function is dead, because no variables are accessed.
+
+A useful liveness analysis needs to consider individual variables. For example, in the body of
+the second function (\ident{value}), two variables are in scope. The variable \ident{value} is
+accessed and thus is \emph{live}, but the variable \ident{x} is dead.
+
+Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
+propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
+guarantee). Liveness is a \emph{backward} analysis -- this means that the requirements propagates
+from variables to their declaration sites. The distinction between \emph{must} and \emph{may} is 
+apparent when we look at an example with conditionals:
 %
-\inference[(\emph{abs})\;]
-  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
+\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
+\end{array}
 \end{equation*}
-\vspace{-0.9em}
-\caption{Selected coeffect rules for implicit parameters}
-\label{fig:example-resources}
-\vspace{-1em}
-\end{figure}
+%
+The liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
+\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
+\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
+only if it was used in both of the branches (this is sometimes called \emph{neededness}).
+
+The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
+interaction between contextual properties and certain language constructs such as conditionals.
 
@@
 
-\begin{figure}[!b]
-\vspace{-2em}
+\paragraph{Data-flow languages (revisited)}
+When discussing data-flow languages in the previous section, we said that the context provides 
+past values of variables. This can be viewed as a flat contextual property (the context needs
+to keep all past values), but we can also view it as a structural property. Consider the following
+example:
+%
 \begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
-\quad
-\inference[(\emph{app})\;]
-  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
-    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
-  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
 \end{equation*}
-\vspace{-1.5em}
-\caption{Selected coeffect rules for liveness analysis}
-\label{fig:example-need}
-\vspace{-1.2em}
-\end{figure}
+%
+The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
+To evaluate a current value of the stream, we need the current value of \ident{left} and one past
+value of \ident{right}.
+
+As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
+values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
+single number for each function. However, we can design a more precise \emph{structural} analysis
+and track the number of required elements for individual variables.
 
@@
 
-\vspace{-1em}
-\subsection{Liveness analysis.}
+\paragraph{Tainting and provenance}
+Tainting is a mechanism where variables coming from potentially untrusted sources are marked
+(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
+can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
+(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
+case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
+values are annotated with more detailed information about their source.
 
-Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
-whether it is definitely not needed (\emph{dead}). A compiler can remove
-bindings to dead variables as the result is never used.
+Statically typed systems that based on tainting have been use to prevent cross-site scripting
+attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
+\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
+cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
+type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
+\ident{msg}~+~\ident{name}
+\end{array}
+\end{equation*}
+%
+In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
+untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
+The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
+(i.e.~to construct an SQL query).
 
-We start with a restricted analysis and briefly mention how to make it practical
-later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
-to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
-
-A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
-($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
-Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
-if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
-A dead context may be marked as live by
-letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
-
-The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
-\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
-The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
-The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
-$g$ ignores its input (regardless of the evaluation order of the underlying language).
-We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
-Secondly, a \emph{pointwise composition} passes the same
-argument to $g$ and $h$. The parameter is live if either the parameter
-of $g$ or $h$ is live ($r \sqcup s$). Application combines the
-two operations, so the context $\Gamma$ is live if it is needed by
-$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
-
-An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
-combines the context annotations using $\sqcap$. Thus, if the body uses some 
-variables, both the function argument and the context of the declaration-site are marked as live.
-
-Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
-In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
-indexed comonads (\S\ref{sec:comonads}), we
-model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
-$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
-(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
-value when $r=\ident{D}$).
+In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
+flag (for tainting) or more complex data (for provenance). In static checking, the information
+need to be associated with the variables in the variable context. We use tainting as a motivating
+example for \emph{structural} coeffects in Section~X.
 
@@
 
-\begin{figure}[t]
-\vspace{-1em}
+\paragraph{Security and core dependency calculus}
+
+The checking of tainting is a special case of checking of the \emph{non-interference} property 
+in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
+as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
+network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
+system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
+more recent work. The checking of information flows has been also integrated (as a single-purpose
+extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
+Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
+using a monadic library.
+
+Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
+where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\ident{L}, \ident{H}\}$ 
+represents low and high secrecy, respectively with $\ident{L} \leq \ident{H}$ meaning that low security
+values can be treated as high security (but not the other way round).
+
+An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
+example which may assign a new value to $z$:
+%
 \begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{0}{\Gamma} \vdash x : \tau }
+\kvd{if}~x>0~\kvd{then}~z := y
+\end{equation*}
 %
-\quad\quad
-% 
-\inference[(\emph{app})\;]
-  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\inference[(\emph{prev})\;]
-  {\ctyp{n}{\Gamma} \vdash e : \tau}
-  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
-%
-\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\vspace{-1em}
-\caption{Selected coeffect rules for causal data flow}
-\label{fig:example-dataflow}
-\vspace{-1em}
-\end{figure}
+If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
+(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
+$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
+assignment is performed or not performed leaks information in its own (this is an 
+\emph{implicit} flow).
+
+Abadi et al. realized that there is a number of analyses similar to secure information flow
+and proposed to unify them using a single model called Dependency Core Calculus (DCC) \cite{app-dcc}.
+It captures other cases where some information about expression relies on properties of variables
+in the context where it executes.  The DCC captures, for example, \emph{binding time analysis}
+\cite{app-binding-time-analysis}, which detects which parts of programs can be partially evaluated
+(do not depend on user input) and \emph{program slicing} \cite{app-slicing-survey} that identifies
+parts of programs that contribute to the output of an expression.
 
@@
 
-\vspace{-1em}
-\subsection{Efficient dataflow.}
-Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
-computations over streams. In \emph{causal} data flow, program may
-access past values -- in this setting, a function $\tau_1 \rightarrow
-\tau_2$ becomes a function from a list of historical values $[\tau_1]
-\rightarrow \tau_2$. A coeffect system here tracks how many past values
-to cache.
+\subsection{Beyond passive contexts}
 
-Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
-number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
-The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
-additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
+In the systems discussed so far, the context provides additional data (resources, implicit 
+parameters, historical values) or meta-data (security, provenance). However, it is impossible to
+write a function that modifies the context. We use the term \emph{passive} context for such 
+applications.
 
-The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
-adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
-for the second function); passing the context to two subcomputations requires the maximum
-number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
-needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
-call-site must each provide at least the number of past values required by the function body
-(the body may use variables coming from the declaration-site as well as the argument).
-
-The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
-\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
-$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
-(inefficient) unbounded lists of past elements. The above static analysis provides an
-approximation of the number of required past elements and so we use just fixed-length lists.
-
-% ==================================================================================================
-
-\section{Generalized coeffect calculus}
-\label{sec:calculus}
-
-The previous three examples exhibit a number of commonalities. We capture these in 
-the \emph{coeffect calculus}. We do not overly restrict the calculus to make
-it open for notions of context-dependent computations not discussed above.
-
-The syntax of our calculus is that of the simply-typed lambda calculus
-(where $v$ ranges over variables, \ident{T} over base types, and $r$ over coeffect annotations):
-%
-\begin{equation*}
-e ::= v \sep \lambda v.e \sep e_1~e_2
-\quad\quad\quad\quad
-\tau ::= \ident{T} \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{r}{\tau}
-\end{equation*}
-
-The type $\ctyp{r}{\tau}$ captures values of type $\tau$ in a context
-specified by the annotation  $r$. This type appears only on the
-left-hand side of a function arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$. 
-In the semantics, $\ctyp{r}$ corresponds to some data type
-(\eg{}~\ident{list} or \ident{Maybe}). Extensions
-such as explicit \emph{let}-binding are discussed later
-(\S\ref{sec:syntactic}).
-
-The coeffect tags $r$, that were demonstrated in the previous section, can be generalized to a 
-structure with three binary operators and a particular element.
-
-\begin{definition}
-A \emph{coeffect algebra} $(S, \imult, \vee, \wedge, \iunit)$ is a set $S$ with
-an element $\iunit \in S$, a semi-lattice $(S, \vee)$, a monoid $(S, \imult, \iunit)$,
-and a binary $\wedge$. That is, $\forall r,s,t\in S$:
-\begin{equation}
-\tag{monoid}
-   r \imult (s \imult t) = (r \imult s) \imult t 
-\quad\quad\quad
-   \iunit \imult r = r = r \imult \iunit  \quad\quad
-\end{equation}
-\vspace{-1.5em}
-\begin{equation}
-\tag{semi-lattice}
-   r \vee s = s \vee r 
-\quad\quad
-   r \vee (s \vee t) = (r \vee s) \vee t
-\quad\quad
-   r \vee r = r
-\end{equation}
-\end{definition}
-%
-The generalized coeffect calculus captures the three motivating examples (\S\ref{sec:motivation}),
-where some operators of the coeffect algebra may coincide.
-
-The $\imult$ operator represents \emph{sequential} composition; guided by the categorical
-model (\S\ref{sec:comonads}), we require it to form a monoid with $\iunit$.
-The operator $\vee$ corresponds to merging of context-requirements in \emph{pointwise composition} 
-and the semi-lattice $(S, \vee)$ defines a partial order: $r \leq s$ when 
-$r \vee s = s$. This ordering implies a sub-coeffecting rule.  
-The coeffect $\iunit$ is often the top or bottom of the lattice.
-
-The $\wedge$ operator corresponds to splitting requirements of a function body. 
-The operator is unrestricted in the general system.
-A number of additional laws holds for \emph{some} coeffects systems, \eg{}~semi-lattice structure of
-$\wedge$ and a form of distributivity. Quite possibly, they should hold for all coeffect systems.  
-We start with as few laws as possible so as not to limit possible uses of the calculus.  
-We consider constrained variants that provide useful syntactic properties later (\S\ref{sec:syntactic}).
-
-\vspace{-0.4em}
-\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
-union $\cup$ for all three operators. Variable access is annotated with $\iunit = \emptyset$
-and $\leq$ is subset ordering.
-
-\vspace{-0.4em}
-\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
-$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
-and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
-and $\imult$ are both $\sqcap$ (meet).
-
-\vspace{-0.4em}
-\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
-$\vee, \wedge$ and $\imult$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
-Variable access is annotated with $\iunit = 0$ and the order $\leq$ is the standard ordering
-of natural numbers.
+However, there is a number of systems where the context may be changed -- not just be evaluating
+certain code block in a different scope (e.g. by wrapping it in $\ident{prev}$ in data-flow), but
+also by calling a function that, for example, acquires new capabilities. While this thesis focuses
+on systems with passive context, we quickly look at the most important examples of the 
+\emph{active} variant.
 
@@
 
-\vspace{-1em}
-\subsection{Coeffect typing rules.}
-Figure~\ref{fig:coeffect-typing} shows the
-rules of the coeffect calculus, given some coeffect algebra $(S, \imult, \vee, \wedge, \iunit)$.
-The context required by a variable access (\emph{var}) is annotated with $\iunit$.
-The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
-to be generalized.
+\paragraph{Calculus of capabilities}
+Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
+a sound system with region-based memory management for low-level code that can be easily 
+compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
+who developed an \emph{effect system} (discussed in Section~\ref{sec:path-sem-effects}) that uses
+lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.
 
-The (\emph{abs}) rule checks the body of the function in a context $r \wedge s$, which is a 
-combination of the coeffects available in the context $r$ where the function is 
-defined and in a context provided by the caller of the function. 
-Note that none of the judgements create a \emph{value} of type $\ctyp{r}{\tau}$. This
-type appears only immediately to the left of  an arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$.
+In the work of Tofte and Talpin, the context is \emph{passive}. They extend a simple functional language
+with the \kvd{letrgn} construct that defines a new memory region, evaluates an expression (possibly)
+using memory in that region and then deallocates the memory of the region:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
+\quad\kvd{letrgn}~\rho~\kvd{in}\\
+\quad\kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~!\ident{x}
+\end{array}
+\end{equation*}
+%
+The memory region $\rho$ is a part of the context, but only in the scope of the body of 
+\kvd{letrgn}. It is only available to the last line which allocates a memory cell in the region
+and reads it (before the region is deallocated). There is no way to allocate a region inside a 
+function and pass it back to the caller.
 
-In function application (\emph{app}), context requirements of both expressions and the 
-function are combined as previously: the pointwise composition $\vee$ is used to combine
-the coeffects of the expression representing a function $r$ and the coeffects of
-the argument, sequentially composed with the coeffects of the function $s \imult t$.
+Calculus of capabilities differs in two ways. First, it allows explicit allocation and deallocation
+of memory regions (and so region lifetimes do not follow strict LIFO ordering). Second, it
+uses continuation-passing style. We ignore the latter aspect and so the following example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
+\quad \kvd{letrgn}~\rho~\kvd{in}\\
+\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~\ident{x}
+\end{array}
+\end{equation*}
+%
+The example is almost identical to the previous one, except that it does not return the value
+of reference \ident{x}. Instead, it returns the reference, which is located in a newly allocated
+region. Together with the value, the function returns a \emph{capability} to access the region
+$\rho$.
 
-For space reasons, we omit recursion. We note that this would require adding
-effect variables and extending the coeffect algebra with a fixed point operation.
+This is where systems with active context differ. To type check such programs, we do not only need
+to know what context is required to call \ident{calculate}. We also need to know what effects it
+has on the context when it evaluates and the current context meeds to be appropriately adjusted
+after a function call. We briefly consider this problem in Section~X. % future work
 
@@
 
-\begin{figure}[t]
-\vspace{-1em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\iunit}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-%
-\inference[(\emph{app})\;]
-  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\inference[(\emph{sub})\;]
-  {\ctyp{s}{\Gamma} \vdash e : \tau }
-  {\ctyp{r}{\Gamma} \vdash e : \tau }
-\;(s\leq r)  
-%
-\quad\quad\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\caption{Type and coeffect system for the coeffect calculus}
-\label{fig:coeffect-typing}
-\vspace{-1em}
-\end{figure}
+\paragraph{Software updating}
+Dynamic software updating (DSU) \cite{app-dsu-programs,app-dsu} is the ability to update programs at
+runtime without stopping them. The Proteus system developed by Stoyle et al. \cite{app-dsu-mutatis} 
+investigates what language support is needed to enable safe dynamic software updating in C-like 
+languages. The system is based on the idea of capabilities.
 
-% ==================================================================================================
+The system distinguishes between \emph{concrete} uses and \emph{abstract} uses of a value. When
+a value is used concretely, the program examines its representation (and so it is not safe to
+change the representation during an update). An abstract use of a value does not need to examine
+the representation and so updating the value does not break the program.
 
-\section{Coeffect semantics using indexed comonads}
-\label{sec:comonads}
+The Proteus system uses capabilities to restrict what types may be used concretely after any point
+in the program. All other types, not listed in the capability, can be dynamically updated as this
+will not change concrete representation of types accessed later in the evaluation.
 
-The approach of \emph{categorical semantics} interprets terms as
-morphisms in some category.  For typed calculi, typing judgments $x_1 : \tau_1 \ldots x_n : \tau_n
-\vdash e: \tau$ are usually mapped to morphisms $\interp{\tau_1} \times \ldots \times
-\interp{\tau_n} \rightarrow \interp{\tau}$. Moggi showed the semantics of various effectful 
-computations can be captured generally using the (\emph{strong}) \emph{monad}
-structure~\cite{monad-notions}. Dually, Uustalu and Vene showed that
-(\emph{monoidal}) \emph{comonads} capture various kinds of context-dependent 
-computation~\cite{comonads-notions}.
-
-We extend Uustalu and Vene's approach to give a semantics for
-the coeffect calculus by generalising comonads to
-\emph{indexed comonads}. We emphasise semantic intuition and
-abbreviate the categorical foundations for space reasons.
-
-\vspace{-1em}
-\paragraph{Indexed comonads.}
-
-Uustalu and Vene's approach interprets well-typed terms 
-as morphisms $C (\tau_1 \times \ldots
-\times \tau_n) \rightarrow \tau$, where $C$ encodes contexts
-and has a comonad structure~\cite{comonads-notions}.
-Indexed comonads comprise a \emph{family} of object mappings 
-$C^r$ indexed by a coeffect $r$ describing the contextual requirements satisfied by
-the encoded context. We interpret judgments $C^r (x_1 : \tau_1, \ldots, x_n : \tau_n) \vdash e : \tau$ as
-morphisms $C^r (\interp{\tau_1} \times \ldots \times \interp{\tau_n}) \rightarrow \interp{\tau}$.
-
-The indexed comonad structure provides a notion of composition
-for computations with different contextual requirements.
-%%
-\begin{definition}
-Given a monoid $(S, \imult, \iunit)$ with binary operator $\imult$ and
-unit $\iunit$, an \emph{indexed comonad} over a category $\ccat$
-comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
-$\iftor r A \in \obj{\ccat}$ and:
-%%
-\begin{itemize}
-\item a natural transformation $\counit_A : \iftor{\iunit} A \rightarrow A$, called the \emph{counit};
-\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
-  r A \rightarrow B$ to \\ morphisms $\iftor{r \imult{} s} A
-  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
-  \emph{coextend};
-\end{itemize}
-%
-\noindent
-such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
-the following equations hold:
-%
-\renewcommand{\arraycolsep}{1.5em}
-\begin{align*}
-\begin{array}{ccc}
-\counit \circ \cobind{f}{r, \iunit} = f & 
-\cobind{{(\counit)}}{\iunit, r} = \ident{id} & 
-\cobind{(g \circ \cobind{f}{r, s})}{(r \imult s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \imult t)} \\
-\end{array}
-\end{align*}
-\end{definition}
-%
-The \emph{coextend} operation gives rise to an associative composition operation for
-computations with contextual requirements (with \emph{counit} as the identity):
-%
-\[
-\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
-  \rightarrow (\iftor{r \imult s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
-g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
-\]
-%
-The composition $\hat{\circ}$ best expresses the intention
-of indexed comonads: contextual requirements of the composed
-functions are combined. The properties of
-the composition follow from the indexed comonad laws and the 
-monoid $(S, \oplus, \iunit)$.
-
-\paragraph{Example}
-Indexed comonad are analogous to comonads (in coKleisli form), but
-with the additional monoidal structure on indices.  Indeed, comonads
-are a special case of indexed comonads with a trivial singleton
-monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\iftor 1$
-is the underlying functor of the comonad and $\counit$ and
-$\cobind{(-)}{1, 1}$ are the usual comonad operations.
-However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.
-
-\paragraph{Example}
-The \emph{indexed partiality comonad} encodes free-variable contexts
-of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
-\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
-\sqcap, \ident{L})$, where $\iftor{\ident L} A = A$ encodes live
-contexts and $\iftor{\ident D} A = 1$ encodes dead contexts,
-where $1$ is the unit type inhabited by a single value $()$.
-The \emph{counit} operation $\counit{} : \iftor{\ident{L}} A \rightarrow A$
-is defined $\counit\ x = x$ and \emph{coextend},
-for all $f : \iftor r A \rightarrow B$, and thus
-$\cobind{f}{r, s} : \; \iftor{r \sqcap s} A \rightarrow \iftor s B$,
-is defined:
-%
-\begin{align*}
-\begin{array}{llll}
-\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
-\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
-\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
-\cobind{f}{\ident{L}, \ident{L}} x = f~x 
-\end{array}
-\end{align*}
-%
-The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
-data type $\mathsf{Maybe}\ A = A + 1$.  This
-type does not permit a comonad structure since
-$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
-For the indexed comonad, $\counit$ need only be defined
- for $\iftor{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
- contextual notions of computation than comonads.
-
- Moreover, indexed comonads are not restricted by the \emph{shape
-   preservation} property of comonads~\cite{orchard12codo}: that
- a coextended function cannot change the \emph{shape} of the
- context. For example, in the second case above $f^\dagger_{\ident{D},
-   \ident{L}} : \iftor{\ident{D}} A \rightarrow \iftor{\ident{L}} B$
- where the shape changes from $1$ (empty context) to $B$ (available
- context).
-
-\vspace{-1em}
-\subsection{Monoidal indexed comonads.}
-Indexed comonads provide a semantics to sequential composition, but
-additional structure is needed for the semantics of the full coeffect calculus.
-Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
-structure, which provides a monoidal operation $\mathsf{m} : C A
-\times C B \rightarrow C (A \times B)$ for merging contexts (used in
-the semantics of abstraction).
-
-The semantics of the coeffect calculus requires an indexed
-lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
-\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
-by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
-%
-\begin{itemize}
-\item $\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \wedge s)}(A \times B)$ natural
-in $A, B$;
-\item $\csplit_{r, s} : \iftor{(r \vee s)}(A \times B) \rightarrow \iftor r A \times \iftor s B$ natural
-in $A, B$;
-\end{itemize}
-%
-\noindent
-The $\cmerge_{r, s}$ operation merges contextual computations with
-tags combined by $\wedge$ (greatest lower-bound), elucidating 
-the behaviour of $\cmerge_{r, s}$: that merging may result in 
-the loss of some parts of the contexts $r$ and $s$.
-
-The $\csplit_{r, s}$ operation splits context-dependent computations
-and thus the contextual requirements. To obtain coeffects $r$ and
-$s$, the input needs to provide \emph{at least} $r$ and $s$, so the
-tags are combined using the $\vee$ (least upper-bound).
-
-For the sake of brevity, we elide the indexed versions of the laws required by Uustalu and Vene
-(\eg~most importantly, merging two contexts and then adding the third is
-equivalent to merging the last two and then adding the first; similar rule holds is required
-for splitting).
-
-\paragraph{Example}
-For the indexed partiality comonad, given the liveness coeffect
-algebra $(\{\ident{D}, \ident{L}\}, \sqcap, \sqcup, \sqcap, \ident{L})$,
-the additional lax/colax monoidal operations are:
-%%
-\begin{align*}
-\begin{array}{lll}
-\cmerge_{\ident{L}, \ident{L}} (x, y) = (x, y) 
-  \quad\quad& \csplit_{\ident{D}, \ident{D}} \;\; () \;\;\; = ((), ())
-  \quad\quad& \csplit_{\ident{D}, \ident{L}} (x, y) = ((), y) \\
-\cmerge_{\ident{r}, \ident{s}} \;\,(x, y) = () 
-  & \csplit_{\ident{L}, \ident{D}} (x, y) = (x, ())
-  & \csplit_{\ident{L}, \ident{L}} (x, y) = (x, y)
-\end{array}
-\end{align*}
-
-\paragraph{Example}
-Uustalu and Vene model causal dataflow computations using the
-non-empty list comonad $\ident{NEList}~A = A \times (1 +
-\ident{NEList}~A)$~\cite{comonads-notions}.  Whilst this comonad
-implies a trivial indexed comonad, we define an indexed comonad with
-integer indices for the number of past values demanded of the context.
-
-We define $\ctyp{n}{A} = A \times (A \times \ldots \times A)$ where the first $A$ is the current
-(always available) value, followed by a finite product of $n$ past values. The definition of the
-operations is a straightforward extension of the work of Uustalu and Vene.
-
-%\begin{example}
-%The semantics of the coeffect calculus for
-% dynamically-scoped implicit parameters, with coeffect
-%algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
-%is provided by the indexed comonad $\iftor rA = A \times (r \rightarrow \tau)$
-%where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
-%do we need to store the types too? Maybe should just be a product.}
-%%
-%\vspace{-1.0em}
-%\begin{equation*}
-%\begin{array}{l}
-%\cobind{f}{r, s} \; : \; \iftor{r \cup s} A \rightarrow \iftor s B\\
-%\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
-%\\[1em]
-%\counit{} (a, g) = a\\
-%%\iota_{r, s} \; : \: \iftor r A \rightarrow \iftor s A\\
-%\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
-%\end{array}
-%\quad\begin{array}{l}
-%\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \cup s)}(A \times B)\\
-%\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
-%\\[1em]
-%\csplit_{r, s} : \iftor{(r \cup s)}(A \times B) \rightarrow \iftor r A \times \iftor s B \\
-%\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
-%\end{array}
-%\end{equation*}
-%\end{example}
-
-%The function created using \emph{coextend} expects a function $g$ that is defined for
-%resource names $r \cup s$. We use function restriction $g|_r$ and $g|_s$ to get functions that are 
-%defined on resources required by $f$ and the result of coextend, respectively.
-
-%The operation $\cmerge_{r, s}$ pairs the values and combines the resources available in both contexts. 
-%It combines a function $g_1|_{r\setminus s}$ (defined for resources in $r$ excluding those in $s$) and 
-%function $g_2$ (defined for resources $s$). This definition is not symmetric as it prefers resources 
-%defined in the second context. As a result, the caller of a function can rebind resources defined
-%at the declaration site.
-
-%Resources show that indexed comonads describe the structure of context more precisely
-%than comonads. A comonad would have to use $TA = A \times (\Omega \rightharpoonup R)$ 
-%with a \emph{partial} function on all resource names. Using indexed comonads, we can ensure that the function
-%is defined on all resources tracked statically in the tag.\dnote{Needs tweaking}
-
-\vspace{-1.5em}
-\subsection{Categorical Semantics. }
-\label{sec:comonads-semantics}
-
-\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
-\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}
-
-\begin{figure*}[t]
-\vspace{-1em}
-\newcommand{\uu}[1]{\overline{#1}}
-% \begin{equation*} 
-% \cosem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-%   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
-% \end{equation*}
-% \vspace{-0.4em}
-\begin{equation*}
-\begin{array}{rcll}
-\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \emph{curry} \; (\cosem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
-\\[0.1em]
-\cosem{\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
-    (\emph{uncurry} \;  \cosem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
- & & \quad\; (id \times \cobind{\cosem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
-    \circ \csplit_{r, s \imult t} \circ \iftor{r \vee (s \imult t)}\Delta
-\\[0.1em]
-\cosem{\ctyp{\iunit}{\Gamma} \vdash x_i : \tau_i} &=&
-    \pi_i \circ \counit                                                                  % &$(3)$
-%\\[0.1em]
-%\cosem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
-%    \cosem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
-%\\[0.1em]
-%\cosem{\ctyp{r\vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
-%    \cosem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
-% ((\varepsilon \times \cosem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
-%\iftorhat{r \vee (r \imult t)}\Delta
-\end{array}
-\end{equation*}
-\vspace{-1.5em}
-\caption{Categorical semantics for the coeffect calculus}
-\label{fig:semantics}
-\vspace{-1.0em}
-\end{figure*}
-
-Figure~\ref{fig:semantics} shows the
-categorical semantics of the coeffect calculus
-using additional operations 
-$\pi_i$ for projection of the \emph{i}$^{th}$ element of a 
-product, usual
-\emph{curry} and \emph{uncurry} operations, and
- $\Delta : A \rightarrow A \times A$ duplicating a
-value. While $\iftor{r}$ is a family of object mappings,
-it is promoted to a family of functors with the derived morphism mapping 
-$\iftor{r}(f) = \cobind{ (f \circ \counit) }{\iunit, r}$.
-
-The semantics of variable access and abstraction are the same as in
-Uustalu and Vene's semantics, modulo coeffects. Abstraction uses $\cmerge_{r, s}$ to merge the outer
-context with the argument context for the context of the function body.
-The indices of $\iunit$ for
-$\counit$ and $r,s$ for $\cmerge_{r, s}$ match the coeffects
-of the terms.
-The semantics of application is more complex. It first duplicates the
-free-variable values inside the context and then splits this context
-using $\csplit_{r, {s \oplus t}}$. The two contexts (with
-different coeffects) are passed to the two sub-expressions,
-where the argument subexpression, passed a context $(s \oplus t)$,
-is coextended to produce a context $s$ which is passed into
-the parameter of the function subexpression (\emph{cf.} given $f : A \rightarrow (B \rightarrow C)$,
- $g : A \rightarrow B$, then $\emph{uncurry} \, f \, \circ \, (id \times g) \circ \Delta : A \rightarrow C$).
-
-A semantics for sub-coeffecting is omitted, but may be provided
-by an operation 
-$\iota_{r, s} : \iftor r A \rightarrow \iftor s A$ natural in $A$, for all $r, s \in S$ where $s \leq r$,
-which transforms a value $C^r A$ to $C^s A$ by
-ignoring some of the encoded context.
-
-% ==========================================================================================
-
-\section{Syntax-based equational theory}
-\label{sec:syntactic}
-
-Operational semantics of every context-dependent language differs as the notion of context 
-is always different. However, for coeffect calculi satisfying certain conditions we can 
-define a universal equational theory. This suggests a pathway to an operational semantics for 
-two out of our three examples (the notion of context for data-flow is more complex).
-
-In a pure $\lambda$-calculus, $\beta$ and $\eta$ equality for
-functions (also called \emph{local soundness} and \emph{completeness}
-respectively~\cite{logic-modal-reconstruction})
-describe how pairs of abstraction and application can be eliminated:
-$(\lambda x . e_2) e_1 \equiv_\beta \subst{e_1}{x}{e_2}$ and $(\lambda x . e \, x) 
-\equiv_\eta e$. The $\beta$ equality rule, using the usual Barendregt convention of 
-syntactic substitution, implies a \emph{reduction},
-giving part of an operational semantics for the calculus.
-
-The call-by-name evaluation strategy modelled by $\beta$-reduction is
-not suitable for impure calculi therefore a restricted $\beta$ rule,
-corresponding to call-by-value, is used, \ie~$(\lambda x . e_2) v
-\equiv \subst{e_2}{x}{v}$. Such reduction can be encoded by a
-\emph{let}-binding term, $\kvd{let}~x=e_1~\kvd{in}~e_2$, which
-corresponds to sequential composition of two computations, where the
-resulting pure value of $e_1$ is substituted into
-$e_2$~\cite{monads-inaction,monad-notions}.
-
-%
-% We consider here both a notion of \emph{let}-binding for the coeffect
-% calculus, useful for a CBV evaluation, and a notion of substitution 
-% for a CBV evaluation.
-%
-
-For an equational theory of coeffects, consider first a notion 
-of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
-has the following type and coeffect rule:
-%
-\begin{equation}
-\inference
-  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
-   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
-  {\ctyp{r_1 \vee (r_2 \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
-: \tau_2 }
-\label{eq:let1}
-\end{equation}
-
-\noindent
-For our examples, $\wedge$ is idempotent (\ie{}, $r \wedge r = r$)
-implying a simpler rule:
-
-% For the three examples we consider, a simpler rule gives a more
-% precise coeffect. Because $r \wedge r = r$ for all our examples, we
-% can also a coeffect $r \vee (r \imult s)$. Moreover,
-%for our examples (but not necessarily for \emph{all} coeffect
-%systems), $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$ meaning
-%that the following gives more precise coeffects:
-%
-\begin{equation}
-\inference
-  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
-   \ctyp{r}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
-  {\ctyp{r \vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2 : \tau_2 }
-\label{eq:let2}
-\end{equation}
-%%
-For our examples (but not necessarily \emph{all} coeffect
-systems), this defines a more ``precise'' coeffect with respect to $\leq$
-where $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$.
-
-This rule removes the non-principality of the first rule
-(\ie~multiple possible typings).  However, using idempotency 
-to split coeffects in abstraction would remove additional
-flexibility needed by the implicit parameters example.
-
-The coeffect $r \vee (r \imult s)$ can
-also be simplified for all our examples, leading to more intuitive
-rules -- for implicit parameters $r \cup (r \cup s) = r \cup s$; for
-liveness we get that $r \sqcup (r \sqcap s) = r$ and for dataflow we
-obtain $\textit{max}(r, r+s) = r + s$.
-
-Our calculus can be extended with \emph{let}-binding and \eqref{eq:let2}.
-However, we also consider the cases when a
-syntactic substitution $e_2[x \leftarrow e_1]$ has the coeffects
-specified by the above rule \eqref{eq:let2} 
-and prove \emph{subject reduction} theorem
-for certain coeffect calculi.  We consider two common special cases
-when the coeffect of variables $\iunit$ is the greatest ($\top$) or
-least ($\bot$) element of the semi-lattice $(S, \vee)$ and derive
-additional conditions that have to hold about the coeffect algebra:
-
-\begin{lemma}[Substitution]
-\label{thm:subst}
-Given $C^r (\Gamma, x : \tau_2) \vdash e_1 : \tau_1$ and $C^s \Gamma \vdash e_2 : \tau_2$
-then $C^{r \vee (r \oplus s)} \Gamma \vdash \subst{e_2}{x}{e_1} : \tau_1$ if 
-the coeffect algebra satisfies the conditions that 
-$\iunit$ is either the greatest or least element of the semi-lattice,
-$\oplus = \wedge$, and $\oplus$ distributes over $\vee$,
-\ie{}, $X \oplus (Y \vee Z) = (X \oplus Y) \vee (X \oplus Z)$.
-\end{lemma}
-
-\begin{proof}
-By induction over $\vdash$, using the laws (\S\ref{sec:calculus}) and additional assumptions.
-\end{proof}
-
-Assuming $\rightarrow_\beta$ is the usual call-by-name reduction, the
-following theorem models the evaluation of coeffect calculi with
-coeffect algebra that satisfies the above requirements. We do not
-consider \emph{call-by-value}, because our calculus does not have a
-notion of \emph{value}, unless explicitly provided by
-\emph{let}-binding (even a function ``value'' $\lambda x.e$ may have
-immediate contextual requirements).
-
-\begin{theorem}[Subject reduction]
-\label{thm:reduction}
-For a coeffect calculus, satisfying the conditions of Lemma~\ref{thm:subst}, if
-$\ctyp{r}{\Gamma} \vdash e : \tau$ and $e \rightarrow_\beta e'$ then 
-$\ctyp{r}{\Gamma} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-A direct consequence of Lemma~\ref{thm:subst}. 
-\end{proof}
-
-The above theorem holds for both the liveness and resources examples,
-but not for dataflow.  In the case of liveness, $\iunit$ is the
-greatest element ($r \vee \iunit = \iunit$); in the case of
-resources, $\iunit$ is the \emph{least} element ($r \vee \iunit =
-r$) and the proof relies on the fact that additional
-context-requirements can be placed at the context $\ctyp{r}{\Gamma}$
-(without affecting the type of function when substituted under
-$\lambda$ abstraction).
-
-However, the coeffect calculus also captures context-dependence in
-languages with more complex evaluation strategies than
-\emph{call-by-name} reduction based on syntactic substitution.  In
-particular, syntactic substitution does not provide a suitable evaluation
-for dataflow (because a substituted expression needs to capture the
-context of the original scope).
-
-Nevertheless, the above results show that -- unlike effects --
-context-dependent properties can be integrated with
-\emph{call-by-name} languages. Our work also provides a model of
-existing work, namely Haskell implicit parameters
-\cite{app-implicit-parameters}.
-
-% ==================================================================================================
-
-\section{Related and further work}
-\label{sec:related}
-
-This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
-and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
-\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
-may split contextual requirements between the declaration-site and call-site.
-
-Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
-\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
-and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
-\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
-general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).
+Similarly to Capability Calculus, capabilities in DSU can be changed by a function call. For 
+example, calling a function that may update certain types makes it impossible to use those types
+concretely following the function call. This means that DSU uses the context \emph{actively}
+and not just \emph{passively}.
 
@@
 
-\vspace{-1em}
-\paragraph{Comonads and modal logics.}
+\paragraph{Thesis perspective}
 
-Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
-$\square$ modality of an intuitionistic S4 modal logic using monoidal
-comonads, which links our calculus to modal logics.  This link can be
-materialized in two ways.
+As demonstrated in this section, there is a huge number of systems and applications that exhibit
+a form of context-dependence. The range includes different static analyses (liveness, provenance), 
+well-known programming language features (implicit parameters and type classes) as well as features
+not widely available (e.g. for distributed programming).
 
-Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
-correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
-building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
-\cite{monad-notions}) that includes $\square$ as a type
-constructor. For example, in \cite{logic-modal-reconstruction}, the
-modal type $\Box \tau$ represents closed terms.
-In contrast, the \emph{semantic} approach uses monads or comonads
-\emph{only} as a semantics.  This has been employed by Uustalu and
-Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
-follow the semantic approach.
+It is impossible to cover all of these topics in a single coherent thesis and so we focus on 
+two key aspects:
 
-Nanevski et al. extend an S4 term language to a \emph{contextual}
-modal type theory (CMTT)~\cite{logic-cmtt}.
-The \emph{context} is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. Our contextual types are
-indexed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}.
+\begin{compactitem}
+\item \textbf{Flat vs. structural.} We look at both flat coeffects (single value for entire context) and 
+  structural coeffects (single value per variable). We use liveness, implicit parameters and 
+  data-flow to introduce flat coeffects (Section~X) and liveness, refined data-flow and tainting
+  to talk about structural coeffects (Section~Y).
 
-The work on CMTT suggests two extensions to coeffects. The first is
-developing the logical foundations. We briefly considered special cases
-of our system that permits local soundness in \S\ref{sec:syntactic} and
-local completeness can be treated similarly. The second problem is 
-developing the coeffects \emph{metalanguage}. The use of coeffect algebras
-would provide an additional flexibility over CMTT, allowing a wider range 
-of applications.
+\item \textbf{Analysis vs. restriction.} Some of the discussed examples can be viewed as static
+  analyses that obtain some information about programs (i.e.~the number of required past values
+  in data-flow). Other examples provide type system that rules out certain invalid programs 
+  (e.g.~safe locking). We cover this topic when discussing \emph{partial coeffects} in Section~Z.
 
-% --------------------------------------------------------------------------------------------------
+\item \textbf{May vs. must analysis.} When discussing liveness, we observed that we can obtain 
+  two different analyses depending on how conditionals are treated. We discuss this topic in 
+  Section~X. % TBD - could be a chapter
+\end{compactitem}
 
-\vspace{-1em}
-\paragraph{Relating effects and coeffects.} 
-The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
-semantic model (monads vs. comonads) is very different, we can consider extending the two to 
-obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
-the reader monad needs an operation that eagerly performs some effects of a function: 
-$(\tau_1 \rightarrow \mtyp{r \oplus s}{\tau_2}) \rightarrow \mtyp{r}{(\tau_1 \rightarrow \mtyp{s}{\tau_2})}$.
-To obtain a pure lambda abstraction
-for coeffects, we need to restrict the $\cmerge_{r, s}$ 
-operation of indexed comonads, so that the first parameter is annotated with $\iunit$ (meaning
-no effects): $\iftor \iunit A \times \iftor r B \rightarrow \iftor{r}(A \times B)$.
+Although we also looked at examples of \emph{active} contextual computations (where developers can
+write functions that modify the context), we do not consider these applications, to keep the 
+material presented in this thesis focused. We briefly discuss them as future work in Section~X.
 
-\vspace{-1em}
-\paragraph{Structural coeffects.} To make the liveness analysis practical, we need to associate
-information with individual variables (rather than the entire context). We can generalize the 
-calculus from this paper by adding a product operation $\times$ to the coeffect algebra.
-A variable context $x:\tau_1, y:\tau_2, z:\tau_3$ is then annotated with
-$r\times s \times t$ where each component of the tag corresponds to a single variable. The system
-then needs to be extended with structural rules such as:
+%===================================================================================================
+
+\section{Summary}
+
+TODO
+
+% \section{Missing}
+% ~
 % 
-\begin{equation*}
-\inference[(\emph{abs})]
-  {\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\quad
-\inference[(\emph{contr})]
-  {\ctyp{r \cprd s}{(x:\tau_1, y:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r \cvop s }{(z:\tau_1)} \vdash \subst{\subst{e}{x}{z}}{y}{z} : \tau_2 }
-\end{equation*}
-%
-The context-requirements associated with function are exactly those linked to the specific
-variable of the lambda abstraction. Rules such as contraction manipulate variables and perform
-a corresponding operation on the indices.
-
-The structural coeffect system is related to bunched typing \cite{types-bunched} (but generalizes
-it by adding indices). We are currently investigating how to use structural coeffects to capture 
-fine-grained context-dependence properties such as secure information flow \cite{app-secure-flow}
-or, more generally, those captured by dependency core calculus \cite{types-dcc}.
-
-% ==================================================================================================
-
-\section{Conclusions}
-
-We examined three simple calculi with associated
-static analyses (liveness analysis, implicit parameters, and dataflow
-analysis). These were unified in the \emph{coeffect calculus},
-providing a general coeffect system parameterised by an
-algebraic structure describing the propagation of context
-requirements throughout a program.
-
-We model the semantics of coeffect calculus using \emph{indexed comonad} -- a novel structure, which
-is more powerful than (monoidal) comonads. Indices of the indexed comonad operations manifest the 
-semantic propagation of context such that the propagation of information in the general coeffect
-type system corresponds exactly to the semantic propagation of context in our categorical model.
-
-We consider the analysis of context to be essential, not least for the examples here but 
-also given increasingly rich and diverse distributed systems.
+% indexed/layered/etc. monads, productors or whatever

### 2014-03-27 text/Chapter04.tex
Finish introduction chapter and restructure the rest

@@
-\chapter{Structural coeffect language} 
-\label{ch:structural-coeffects}
-
-\newcommand{\ctimes}{ \hat{\times} }
-\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
-
-\newcommand{\csprd}[0]{ \times }
-\newcommand{\csvop}[0]{ \vee }
-
-% alculus names and rule names
-\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
-\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
-
-\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
-\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
-\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
-\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
-\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}
+\chapter{Flat coeffect language} 
+\label{ch:flat}
 
@@
 
-The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
-we need to track context-dependence in a more fine-grained way. To track neededness or security, 
-we need to associate information with individual \emph{variables} of the context.
+% comonadic semantics
+\newcommand{\iunit}{\mathsf{e}}
+\newcommand{\imult}{\oplus}
+\newcommand{\iftor}[1]{C^{#1}}
+\newcommand{\iftorhat}[1]{\hat{C}^{#1}}
 
-%===================================================================================================
+\newcommand{\cosem}[1]{\llbracket #1 \rrbracket}
+\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
+\newcommand{\ccat}[0]{\mathcal{C}}
+\newcommand{\obj}[1]{\textnormal{obj}(#1)}
 
-\section{Introduction}
+\newcommand{\cobind}[2]{#1^\dagger_{#2}}
+\newcommand{\cmerge}[0]{ \ident{m} }
+\newcommand{\csplit}[0]{ \ident{n} }
+\newcommand{\counit}[0]{ \varepsilon }
+% identity function
+
+% structural coeffect calculus - operation names
+
+%---------------------------------------------------------------------------------------------------
+
+\todo{Update the introduction \& edit the whole chapter -- currently, this is just a direct copy
+from the accepted ICALP paper on flat coeffects.}
+
+\todo{Other extensions -- add more examples of indexed comonoads, consider recursion and 
+fixed-point operator, possibly also consider partially defined operations, which lets us
+express type systems that rule out some computations (rather than just static analyses). }
+
+Understanding how programs affect their environment is a well studied area: \emph{effect 
+systems}~\cite{effects-talpin-et-al} provide a static analysis of effects and 
+\emph{monads}~\cite{monad-notions} provide a unified semantics to different notions of effect.  
+Wadler and Thiemann unify the two approaches~\cite{monads-effects-marriage}, \emph{indexing}
+a monad with effect information, and showing that the propagation of effects in an effect system 
+matches the semantic propagation of effects in the monadic approach.
+
+No such unified mechanism exists for tracking the context requirements. We use the term 
+\emph{coeffect} for such contextual program properties. 
+Notions of context have been previously captured using comonads
+\cite{comonads-notions} (the dual of monads) 
+and by languages derived from modal logic
+\cite{logic-modal-reconstruction,logic-cmtt}, but these approaches
+do not capture many useful examples which motivate our work. 
+We build mainly on the former comonadic direction (\S\ref{sec:comonads}) and discuss the
+modal logic approach later (\S\ref{sec:related}).
+
+We extend a simply typed lambda calculus with a coeffect system based on comonads,
+replicating the successful approach of effect systems and monads.
+
+\paragraph{Examples of coeffects.}
+We present three examples that do not fit the traditional approach of
+\emph{effect systems} and have not been considered using the \emph{modal logic}
+perspective, but can be captured as \emph{coeffect systems} (\S\ref{sec:motivation}) --
+the tracking of implicit dynamically-scoped parameters (or resources),
+analysis of variable liveness, and tracking the
+number of required past values in dataflow computations.
+
+\paragraph{Coeffect calculus.}
+Informed by the examples, we identify a general algebraic structure for coeffects. 
+From this, we define a general \emph{coeffect calculus} that unifies the motivating 
+examples (\S\ref{sec:calculus}) and discuss its syntactic properties (\S\ref{sec:syntactic}).
+
+\paragraph{Indexed comonads.}
+Our categorical semantics (\S\ref{sec:comonads}) extends the work of Uustalu and Vene 
+\cite{comonads-notions}. By adding annotations, we generalize comonads to \emph{indexed comonads}, 
+which capture notions of computation not captured by ordinary comonads.
+
+% ==================================================================================================
+
+\section{Motivation}
+\label{sec:motivation}
+
+Effect systems, introduced by Gifford and Lucassen
+\cite{effects-gifford}, track \emph{effects} of computations, such as
+memory access or message-based
+communication~\cite{effects-messagepassing}. Their approach augments
+typing judgments with effect information: $\Gamma \vdash e :
+\tau, F$.  Wadler and Thiemann explain how this shapes
+effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
+%
+\begin{quote}
+\emph{In the rule for abstraction, the effect is empty because evaluation immediately
+returns the function, with no side effects. The effect on the function arrow
+is the same as the effect for the function body, because applying the function will
+have the same side effects as evaluating the body.}
+\end{quote}
+%
+In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
+does not match this pattern. In the systems we consider, lambda abstraction places 
+requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
+(immediate requirements), resulting in different syntactic properties. 
+We informally discuss three examples first that demonstrate how contextual
+requirements propagate. Section (\S\ref{sec:calculus}) then unifies these 
+in a single calculus.
+
+We write coeffect judgements $\ctyp{s}{\Gamma} \vdash e : \tau$ where the 
+coeffect annotation $s$ associates context requirements with the 
+free-variable context $\Gamma$. Function types have the form $\ctyp{s}{\tau_1} \rightarrow \tau_2$ associating
+\emph{latent} coeffects $s$ with the parameter.
+The $\ctyp{s}{\Gamma}$ syntax and $\ctyp{s}{\tau}$ types are a result of the
+indexed comonadic semantics (\S\ref{sec:comonads}).
+
+% --------------------------------------------------------------------------------------------------
+\vspace{-1em}
+\subsection{Implicit parameters and resources.}
+\label{sec:coeffects-res}
+
+Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
+They can be used to parameterize a computation without propagating arguments explicitly through a 
+chain of calls and are part of the context in which expressions evaluate. As correctly expected
+\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
+computations follow a similar pattern, but we discuss implicit parameters for simplicity.
+
+The following function prints a number using implicit parameters
+\ident{?culture} (determining the decimal mark) and \ident{?format}
+(the number of decimal places):
+%
+\begin{equation*}
+\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
+\end{equation*}
+%
+Figure~\ref{fig:example-resources} shows a type-and-coeffect system
+tracking the set of an expression's implicit parameters. 
+For simplicity here, all implicit parameters have type $\rho$.
+
+Context requirements are created in (\emph{access}), while (\emph{var}) requires no
+implicit parameters; (\emph{app}) combines requirements of
+both sub-expressions as well as the latent requirements of the
+function. The (\emph{abs}) rule is where the example differs from
+effect systems.  Function bodies can access the union of the
+parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
+at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
+judgements for the above example are:
+%
+\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
+\begin{equation*}
+\begin{array}{rcl}
+\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
+\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
+\end{array}
+\end{equation*}
+%
+The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
+Different judgments are desirable depending on how a function is used. 
+In the first case, both parameters have to be provided by the caller.
+In the second, both are available at declaration-site, but \ident{?format} may be
+rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).
+
+Implicit parameters can be captured by the
+\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
+  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
+modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
+the next example cannot be structured by \emph{any} monad.
 
@@
 
-\subsection{Motivation: Tracking array accesses}
-\label{sec:structural-motivation}
-%\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
-%\newcommand{\dnat}{\ident{D}_\ident{nat}}
-
-Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
-functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
-respectively, but we use richer tag structure.
-
-As an example, consider a language that allows us to get a value of a variable (representing
-some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
-To track information about individual variables, we use a product-like operation $\csprd$ on tags 
-to mirrors the product structure of variables. For example:
+\begin{figure}[t]
+\vspace{-1.4em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
 %
+\quad\quad
+%
+\inference[(\emph{app})\;]
+  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
+   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
+  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\vspace{-0.7em}
 \begin{equation*}
-\begin{array}{l}
-\ctyp{5 \csprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
-  \vdash
-    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
-\end{array}
+\inference[(\emph{access})\;]
+  {}
+  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
+%
+\quad\quad\quad
+%
+\inference[(\emph{abs})\;]
+  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
 \end{equation*}
-%
-The coeffect tag $5 \csprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
-that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
-for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
-a single variable:
-%
-\begin{equation*}
-\begin{array}{l}
-\ctyp{5 \csvop 10}{(\ident{c}:\dnat)}
-  \vdash
-    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
-\end{array}
-\end{equation*}
-%
-In this example, the operation $\csvop$ would be the \emph{max} function and so $5 \csvop 10 = 10$.
-Before looking at the formal definition, consider the typing of let bindings:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
-\pastval{a}{15}+\pastval{c}{10}
-\end{array}
-\end{equation*}
-%
-The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
-is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
-The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.
-
-The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
-context is $\ctyp{0 \csprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
-of $\csvop$ and it denotes empty coeffect. The typing context of the
-body is $\ctyp{15 \csprd 10}{(\ident{a}, \ident{c})}$.
-
-To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
-of the context in which \ident{c} was defined using the $\csvop$ operation. This is then combined 
-with the remaining tags from the body yielding the overall context:
-$\ctyp{15 \csprd (10 \csvop (0 \csprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
-normalization mechanism (described later), this can be further reduced to
-$\ctyp{(15 \csvop 10) \csprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
-we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
-\ident{b}.
+\vspace{-0.9em}
+\caption{Selected coeffect rules for implicit parameters}
+\label{fig:example-resources}
+\vspace{-1em}
+\end{figure}
 
@@
 
-\subsection{Structural coeffect tags}
-In the previous section, the $\csvop$ operation behaves similarly to the flat $\vee$ operation.
-However, the type system does not require some of the semilattice properties, because some uses
-are replaced with the $\csprd$ operation. We do not require any properties about the $\csprd$ operation.
-For example, in the previous example cannot be commutative (since a tag $15 \csprd 10$ has different 
-meaning than $10 \csprd 15$). However, we relate the operations using distributivity laws to allow 
-normalization that was hinted above.
-
-\begin{definition}
-A \emph{structural coeffect tag structure} $(S, \csprd, \csvop, 0, 1)$ is a tuple where 
-$(S, \csvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
-by a binary operation $\csprd$ and element $1\in S$ such that
-for all $r,s,t\in S$, the following equalities hold: 
-%
-\begin{subequations}
-\begin{align}
-\tag{associativity}
-r \csvop (s \csvop t) &= (r \csvop s) \csvop t\\
-\tag{commutativity}
-r \csvop s &= s \csvop r\\
-\tag{lower bound}
-r \csvop 0 &= r\\
-\tag{distributivity}
-r \csvop (s \csprd t) &= (r \csvop s) \csprd (r \csvop t)\\
-\tag{upper bound}
-1 \csvop r &= 1
-\end{align}
-\end{subequations}
-%
-\end{definition}
-%
-% 'upper bound' (not sure what to call it) says that empty context cannot carry information
-% looks like this is the upper bound of the bounded lattice ???
-%
-% commutativity is not strictly required by any law or type preservation, but
-% it simplifies the presentation and looks reasonable
-%
-The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
-always accessed using standard variable access). The tag $1$ is used to annotate empty 
-variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
-to carry an annotation. We explain exactly how this works when we introduce the type system.
-The fact that $1$ is the upper bound means that combining it with other coeffect annotations
-does not affect it and so empty contexts cannot carry any information.
-
-The structure generalizes the \emph{flat coeffect tag structure} introduced in 
-Section~\ref{sec:flat-tag-structure}, but it additionally requires a special element $1$ 
-representing the upper bound. Given a flat coeffect structure, we can construct a structural
-coeffect structure (but not the other way round). For certain structures, the element $1$
-may be already present, but in general, it can be added as a new element.
-This construction will be important in Section~\ref{sec:generalization},
-where we show that \clstr~calculus generalizes~\clflt.
-
-\begin{lemma}
-\label{thm:structural-generalizaton}
-A flat coeffect tag structure $(S, \vee, 0)$ implies a structural coeffect tag structure.
-\end{lemma}
-\begin{proof}
-Take $1 \notin S$, then $(S \cup \{1\}, \vee, \vee, 0, 1)$ is a structural coeffect tag 
-structure. From the properties of flat coeffect tag structure, we get that the $\vee$ operation 
-is associative, commutative and $0$ is the unit with respect to $\vee$.
-
-To prove the distributivity, we need to show $r \vee (s \vee t) = (r \vee s) \vee (r \vee t)$.
-This easily follows from commutativity, associativity and idempotence of the flat coeffect tag 
-structure.
-\end{proof}
+\begin{figure}[!b]
+\vspace{-2em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
+\quad
+\inference[(\emph{app})\;]
+  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
+    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
+  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\vspace{-1.5em}
+\caption{Selected coeffect rules for liveness analysis}
+\label{fig:example-need}
+\vspace{-1.2em}
+\end{figure}
 
@@
 
-\subsection{Structural coeffect type system}
-\label{sec:structural-types}
+\vspace{-1em}
+\subsection{Liveness analysis.}
 
-The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
-The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
-a \emph{structural coeffect tag structure} $(S,\csprd, \csvop, 0)$.
+Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
+whether it is definitely not needed (\emph{dead}). A compiler can remove
+bindings to dead variables as the result is never used.
 
-The system differs from \clflt~in a significant aspect. It contains explicit 
-structural rules that manipulate with the context. Such rules allow reordering, duplicating
-and other manipulations with variable context. Such rules are known from affine or linear
-type systems where they are removed to obtain more restrictive system. In our system, the
-rules are present, but they manipulate the variable structure $\Gamma$ as well as the 
-associated tag structure $r$.
+We start with a restricted analysis and briefly mention how to make it practical
+later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
+to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
 
-As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
-simple set. Instead, we use the following tree-like structure (which is more similar to 
-bunched types than to linear or affine systems):
-%
-\begin{equation*}
-\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
-\end{equation*}
-%
-The syntax $()$ represents an empty context, so the structure defines a binary trees where
-leaves are either variables or empty. Contexts such as $\ctyp{1\csprd(r \csprd 1)}{((), (x, ()))}$ 
-contain unnecessary number of empty contexts $()$. However, we need to construct them
-temporarily, because certain rules require splitting a context and, by our 
-definition, the context $(x:\tau)$ is not splitable.
+A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
+($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
+Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
+if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
+A dead context may be marked as live by
+letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
 
-The typing rules of the system are shown in Figure~\ref{fig:structural-coeffect-typing}.
-Many of the structural rules are expressed in terms of a helper judgement
-$\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}}$.
+The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
+\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
+The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
+The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
+$g$ ignores its input (regardless of the evaluation order of the underlying language).
+We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
+Secondly, a \emph{pointwise composition} passes the same
+argument to $g$ and $h$. The parameter is live if either the parameter
+of $g$ or $h$ is live ($r \sqcup s$). Application combines the
+two operations, so the context $\Gamma$ is live if it is needed by
+$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
 
-\paragraph{Standard rules.}
-Variable access (\rulest{var}) annotates the corresponding variable with an empty coeffect
-$0$. The \rulest{fun} rule assumes that the context of the body can be split into the
-variable of the function and other (potentially empty) context and it attaches the coeffect
-associated with the function variable to the resulting function type $\ctyp{s}{\tau_1} \rightarrow \tau_2$.
+An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
+combines the context annotations using $\sqcap$. Thus, if the body uses some 
+variables, both the function argument and the context of the declaration-site are marked as live.
 
-The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
-expression, argument and the function type respectively. 
-The result of evaluating the argument in the context $t$ is passed to the function that
-requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
-the combination of coeffects $r \csvop t$. The variable context required to evaluate the 
-function value is independent and so it is annotated just with coeffects $s$.
-Finally, the \rulest{let} rule is derived from let binding and application, but we show
-it separately to aid the understanding.
-
-\paragraph{Structural rules.}
-The remaining rules are not syntax-directed. They are embedded using the \rulest{ctx}
-rule and expressed using a helper judgement $\ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}}$
-that says that the context on the left-hand side can be transformed to the context on the
-right-hand side. The transformation can be applied to any part of the context, which 
-is captured using the \rulest{nest} rule (it is sufficient to apply the transformation 
-on the left part of the context; the right part can be transformed using \rulest{exch}).
-
-The \rulest{empty} rule allows attaching empty context to any existing context. The rule 
-is needed to type-check lambda abstractions that do not capture any outer variable.
-The \rulest{weak} rule is similar, but it represents \emph{weakening} where an unused variable is 
-added. The associated coeffect tag does not associate context information with the variable.
-This is needed to type-check lambda abstraction that does not use the argument.
-
-The \rulest{contr} rule is a limited form of contraction. When a variable appears repeatedly,
-it can be reduced to a single occurrence. The rule is needed to satisfy the side condition
-of \rulest{fun} when the body uses the argument repeatedly. The two associativity rules
-together with \rulest{exch} provide ways to rearrange variables. Finally, \rulest{sub} represents
-sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.
-
-\ExecuteMetaData[rules/rules.tex]{scl-types}
+Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
+In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
+indexed comonads (\S\ref{sec:comonads}), we
+model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
+$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
+(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
+value when $r=\ident{D}$).
 
@@
 
-\subsection{Properties of reductions}
-\label{sec:structural-metatheory}
-Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
-operational meaning, because that will differ for every concrete application. For example,
-when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
-access to a specified element.
-
-Just like previously, we can state general properties of the reductions. As the syntax of expressions
-is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
-are also the same and can be found in Figure~\ref{fig:flat-substred}.
-
-The structural coeffect calculus \clstr~associates information with individual variables. This means
-that when an expression requires certain context, we know from what scope it comes -- the
-context must be provided by a scope that defines the associated variable, which is either a
-lambda abstraction or global scope. This distinguishes the structural system from the flat system
-where context could have been provided by any scope and the lambda rule allowed arbitrary 
-splitting of context requirements between the two scopes (or declaration and caller site).
-
-\paragraph{Internalized substitution.}
-Before looking at properties of the evaluation, we consider let binding, which can be 
-viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
-application and abstraction as follows.
-
-\begin{lemma}[Definition of let binding]
-If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
-then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
-\end{lemma}
-\begin{proof}
-The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
-correspond with the typing rule \rulest{let}:
+\begin{figure}[t]
+\vspace{-1em}
 \begin{equation*}
-\inference[]
-  {\inference
-    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
-    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
-   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
-  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{0}{\Gamma} \vdash x : \tau }
+%
+\quad\quad
+% 
+\inference[(\emph{app})\;]
+  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
+   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
+  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
-\end{proof}
-
-The term $e_2$ which is substituted in $e_1$ is checked in a different variable
-and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
-a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
-the variable that we are substituting for. The let binding captures substitution for a 
-specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
-For a general substitution, we need to define the notion of context with a hole.
-
-\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
-\paragraph{Substitution and holes.}
-In \clstr, the structure of the variable context is not a set, but a tree. When substituting
-for a variable, we need to replace the variable in the context with the context of the
-substituted expression. In general, this can occur anywhere in the tree. To formulate the
-statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
-in the free variable context and in a corresponding part of the coeffect tag:
-
 \begin{equation*}
-\begin{array}{rcl}
-\Delta[-] &::=& \ctyp{1}{()} \\
-       & \sep & \ctyp{r}{(x:\tau)}\\
-       & \sep & \ctyp{-}{(-)} \\
-       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
-\end{array}
+\inference[(\emph{prev})\;]
+  {\ctyp{n}{\Gamma} \vdash e : \tau}
+  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
+%
+\quad\quad
+%
+\inference[(\emph{abs})\;]
+  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
 \end{equation*}
-%
-Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
-operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
-$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
-Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
-general substitution lemma for \clstr.
-
-\begin{figure}
-
-\begin{equation*}
-\begin{array}{rcl}
-                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
-                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
-                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
-\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
-&\multicolumn{2}{l} {
-  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
-}
-\end{array}
-\end{equation*}
-\caption{The definition of hole filling operation for $\Delta[-]$}
-\label{fig:structural-hole-fill}
-\vspace{-2em}
+\vspace{-1em}
+\caption{Selected coeffect rules for causal data flow}
+\label{fig:example-dataflow}
+\vspace{-1em}
 \end{figure}
 
-\begin{lemma}[Substitution Lemma]
-\label{thm:structural-subst}
-If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
-then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
-\end{lemma} 
-\begin{proof}
-Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
-$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
-\end{proof}
+% --------------------------------------------------------------------------------------------------
 
-\begin{theorem}[Subject reduction]
-\label{thm:structural-reduction}
-If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
-$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
-\end{theorem}
-\begin{proof}
-Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
-\end{proof}
+\vspace{-1em}
+\subsection{Efficient dataflow.}
+Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
+computations over streams. In \emph{causal} data flow, program may
+access past values -- in this setting, a function $\tau_1 \rightarrow
+\tau_2$ becomes a function from a list of historical values $[\tau_1]
+\rightarrow \tau_2$. A coeffect system here tracks how many past values
+to cache.
 
-\paragraph{Local soundness and completeness.}
-As with the previous calculus, we want to guarantee that the introduction and
-elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
-This can be done by showing \emph{local soundness} and \emph{local completeness},
-which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
-case of subject reduction and the latter is proved by a simple derivation:
-%
-\begin{theorem}[Local soundness]
-If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
-$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
-\end{theorem}
-\begin{proof}
-Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
-\end{proof}
-%
-\begin{theorem}[Local completeness]
-If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
-$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
-\end{theorem}
-\begin{proof} The property is proved by the following typing derivation:
-%
-\begin{equation*}
-\inference
-  {\inference
-     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
-      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
-     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\qedhere
-\end{equation*}
-\end{proof}
-%
-In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
-guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
-$\lambda x.f x$ required for local completeness was not the only possible derivation.
-In the last step, it was possible to split the coeffect tag arbitrarily between the 
-context and the function type.
+Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
+number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
+The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
+additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
 
-In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
-to be associative and to have units and so a unique splitting may exist.  For example, if we
-define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
-$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
-has additional properties, then there may be other possible derivation.
+The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
+adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
+for the second function); passing the context to two subcomputations requires the maximum
+number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
+needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
+call-site must each provide at least the number of past values required by the function body
+(the body may use variables coming from the declaration-site as well as the argument).
+
+The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
+\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
+$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
+(inefficient) unbounded lists of past elements. The above static analysis provides an
+approximation of the number of required past elements and so we use just fixed-length lists.
 
@@
 
-\begin{figure*}[t]
-\newcommand{\uu}[1]{\overline{#1}}
+\section{Generalized coeffect calculus}
+\label{sec:calculus}
+
+The previous three examples exhibit a number of commonalities. We capture these in 
+the \emph{coeffect calculus}. We do not overly restrict the calculus to make
+it open for notions of context-dependent computations not discussed above.
+
+The syntax of our calculus is that of the simply-typed lambda calculus
+(where $v$ ranges over variables, \ident{T} over base types, and $r$ over coeffect annotations):
+%
 \begin{equation*}
-\cosem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
-\end{equation*}
-\begin{equation*}
-\begin{array}{rcll}
-\cosem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
-    \epsilon_0                                                                   &\quad$\ruless{var}$
-\\[0.32em]
-\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \Lambda (\cosem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
-\\[0.32em]
-\cosem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
-    (\lambda (\gamma_1, \gamma_2) \rightarrow
-    \cosem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
-    (\cobind{\cosem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
-\\[0.32em]
-\cosem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
-    \cosem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \cosem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
-\end{array}
+e ::= v \sep \lambda v.e \sep e_1~e_2
+\quad\quad\quad\quad
+\tau ::= \ident{T} \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{r}{\tau}
 \end{equation*}
 
-\begin{equation*}
-\begin{array}{rclll}
-\cosem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
-&=& \cmerge_{r, s} \circ (\cosem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
-\textnormal{\rulessc{nest}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
-&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
-\textnormal{\rulessc{exch}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 1}&&
-\textnormal{\rulessc{empty}}
-\\[0.8em]
-\cosem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 0}&&
-\textnormal{\rulessc{weak}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
-    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
-&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
-\textnormal{\rulessc{contr}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
-    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
-&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
-        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
-\textnormal{\rulessc{assoc}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
-&=&
-\iota_{r, s}
-&&
-\textnormal{\rulessc{sub}}
-\end{array}
-\end{equation*}
+The type $\ctyp{r}{\tau}$ captures values of type $\tau$ in a context
+specified by the annotation  $r$. This type appears only on the
+left-hand side of a function arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$. 
+In the semantics, $\ctyp{r}$ corresponds to some data type
+(\eg{}~\ident{list} or \ident{Maybe}). Extensions
+such as explicit \emph{let}-binding are discussed later
+(\S\ref{sec:syntactic}).
 
-\caption{Categorical semantics for~\clstr}
-\label{fig:structural-semantics}
-\vspace{-0.5em}
-\end{figure*}
+The coeffect tags $r$, that were demonstrated in the previous section, can be generalized to a 
+structure with three binary operators and a particular element.
+
+\begin{definition}
+A \emph{coeffect algebra} $(S, \imult, \vee, \wedge, \iunit)$ is a set $S$ with
+an element $\iunit \in S$, a semi-lattice $(S, \vee)$, a monoid $(S, \imult, \iunit)$,
+and a binary $\wedge$. That is, $\forall r,s,t\in S$:
+\begin{equation}
+\tag{monoid}
+   r \imult (s \imult t) = (r \imult s) \imult t 
+\quad\quad\quad
+   \iunit \imult r = r = r \imult \iunit  \quad\quad
+\end{equation}
+\vspace{-1.5em}
+\begin{equation}
+\tag{semi-lattice}
+   r \vee s = s \vee r 
+\quad\quad
+   r \vee (s \vee t) = (r \vee s) \vee t
+\quad\quad
+   r \vee r = r
+\end{equation}
+\end{definition}
+%
+The generalized coeffect calculus captures the three motivating examples (\S\ref{sec:motivation}),
+where some operators of the coeffect algebra may coincide.
+
+The $\imult$ operator represents \emph{sequential} composition; guided by the categorical
+model (\S\ref{sec:comonads}), we require it to form a monoid with $\iunit$.
+The operator $\vee$ corresponds to merging of context-requirements in \emph{pointwise composition} 
+and the semi-lattice $(S, \vee)$ defines a partial order: $r \leq s$ when 
+$r \vee s = s$. This ordering implies a sub-coeffecting rule.  
+The coeffect $\iunit$ is often the top or bottom of the lattice.
+
+The $\wedge$ operator corresponds to splitting requirements of a function body. 
+The operator is unrestricted in the general system.
+A number of additional laws holds for \emph{some} coeffects systems, \eg{}~semi-lattice structure of
+$\wedge$ and a form of distributivity. Quite possibly, they should hold for all coeffect systems.  
+We start with as few laws as possible so as not to limit possible uses of the calculus.  
+We consider constrained variants that provide useful syntactic properties later (\S\ref{sec:syntactic}).
+
+\vspace{-0.4em}
+\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
+union $\cup$ for all three operators. Variable access is annotated with $\iunit = \emptyset$
+and $\leq$ is subset ordering.
+
+\vspace{-0.4em}
+\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
+$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
+and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
+and $\imult$ are both $\sqcap$ (meet).
+
+\vspace{-0.4em}
+\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
+$\vee, \wedge$ and $\imult$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
+Variable access is annotated with $\iunit = 0$ and the order $\leq$ is the standard ordering
+of natural numbers.
 
@@
 
-\section{Semantics of structural coeffects}
-\label{sec:structural-semantics}
+\vspace{-1em}
+\subsection{Coeffect typing rules.}
+Figure~\ref{fig:coeffect-typing} shows the
+rules of the coeffect calculus, given some coeffect algebra $(S, \imult, \vee, \wedge, \iunit)$.
+The context required by a variable access (\emph{var}) is annotated with $\iunit$.
+The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
+to be generalized.
 
-The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
-\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
-of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
-is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
+The (\emph{abs}) rule checks the body of the function in a context $r \wedge s$, which is a 
+combination of the coeffects available in the context $r$ where the function is 
+defined and in a context provided by the caller of the function. 
+Note that none of the judgements create a \emph{value} of type $\ctyp{r}{\tau}$. This
+type appears only immediately to the left of  an arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$.
 
-As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
-system is not a simple finite product, but instead a binary tree. To model this, we do not use
-ordinary products in the domain of the semantic function, but instead use a special 
-constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
-the tag structure.
+In function application (\emph{app}), context requirements of both expressions and the 
+function are combined as previously: the pointwise composition $\vee$ is used to combine
+the coeffects of the expression representing a function $r$ and the coeffects of
+the argument, sequentially composed with the coeffects of the function $s \imult t$.
 
-\subsection{Structural tagged comonads}
-\label{sec:structural-tagged-comonads}
-
-To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
-Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
-$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
-$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
-context-requirements.
-
-However, functions that do not exist in context have only a single input variable (with a 
-single corresponding tag). To model complex variable contexts, we need two additional
-operations that allow manipulation with the variable context. Similarly to the model of \clflt,
-we also require operations that model duplication and sub-coeffecting:
-
-\begin{definition}[Structural tagged comonad]
-Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
-tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
-$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
-$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
-\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
-\end{array}
-\end{equation*}
-%
-% Delta is still defineable
-% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
-%
-And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
-%
-%  TODO: laws?
-%
-\end{definition}
-
-The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
-and it can still be used to define a family of mappings that represents \emph{duplicating}
-of variables while splitting the additional coeffect tags:
-%
-\begin{equation*}
-\begin{array}{l}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
-\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
-\end{array}
-\end{equation*}
-%
-The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
-but with two differences. Firstly, it combines tags using $\csprd$ instead of $\vee$, 
-which corresponds to the fact that the variable context now consists of two parts (a tree node).
-Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
-(instead of $A \times B$ as previously).
-
-To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
-of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
-variables. The operation was not needed for \clflt, because there \emph{splitting} could
-be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
-the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
-with $r \csprd s$ that carries $A \ctimes B$.
-
-Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
-Before looking at them, we finish our discussion of categorical semantics.
-
-\paragraph{Categorical notes.}
-The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
-Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
-variables in the context. Given a context modelled as $T^{r \csprd s}(A \ctimes B)$,
-we can lift a morphism $f$ to $\hat{T}^{r \csprd s}(f)$, but we cannot manipulate the
-variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
-
-This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
-could apply $\cdupl_{r, s}$ on the input (if the tag $r \csprd s$ coincided
-with tag $r \csvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
-
-This restriction is intentional -- at the semantic level, it prevents manipulations
-with the context that would break the correspondence between tag structure and the
-product structure.
-
-\subsection{Categorical semantics}
-The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
-It uses the \emph{structural tagged comonad} structure introduced in the previous section,
-together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
-operations:
-\begin{equation*}
-\begin{array}{rcl}
-\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
-\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
-f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
-\end{array}
-\end{equation*}
-
-When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
-notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
-the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
-variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
-$\csprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
-type signature in $\cmerge_{r, s}$).
-
-The rule \ruless{app} now uses the operation $\csplit_{s, (r \csvop t)}$ instead of $\cdupl_{s, (r \csvop t)}$,
-which means that it splits the context instead of duplicating it. This makes the system more 
-structural -- the expressions use disjunctive parts of the context -- and also explains why the
-composed coeffect tag is $s \csprd (r \csvop t)$.
-
-The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
-of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
-of $\cosem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
-$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
-takes a context, converts it to a new context which is compatible with the original expression $e$.
-The context manipulation rules work as follows:
-
-\begin{itemize}
-\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
-  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
-  Finally, they re-construct a single context using $\cmerge_{r, s}$.
-
-\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
-  and discard one part (containing either an unused variable or an empty context).
-
-\item If we interpreted \rulessc{contr} by applying functor $T^{r\csvop s}$ to a function that duplicates
-  a variable, the resulting context would be $\ctyp{r \csvop s}(x:\tau, x:\tau)$, which would break the
-  correspondence between coeffect tag and context variable structure. However, that interpretation would be
-  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
-  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
-  a tag $r \csvop s$ into $r \csprd s$.
-
-\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
-  manipulates it (using \ident{assoc}) and then re-constructs it.
-
-\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
-  variable using the primitive natural transformation $\iota_{r, s}$.
-\end{itemize}
-
-\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
-attaching the context separately to individual variables. This an expression
-$\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
-$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
-approach largely complicates the definition of application (where tag of all 
-variables in a context is affected). Moreover, it makes it impossible to express
-\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.
-
-\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
-transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
-operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
-sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).
-
-This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
-typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
-the following natural transformation has to be provided:
-%
-\begin{equation*}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
-\end{equation*}
-%
-This variant of the system could be used to define a system that ensures that all provided 
-context is used and is not over-approximated. This difference is similar to the difference between
-affine type systems (where a variable can be used at most once) and linear type systems 
-(where a variable has to be used exactly once).
+For space reasons, we omit recursion. We note that this would require adding
+effect variables and extending the coeffect algebra with a fixed point operation.
 
@@
 
-\section{Examples of structural coeffects}
+\begin{figure}[t]
+\vspace{-1em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{\iunit}{\Gamma} \vdash x : \tau }
+%
+\quad\quad
+%
+\inference[(\emph{app})\;]
+  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
+   \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
+  {\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\inference[(\emph{sub})\;]
+  {\ctyp{s}{\Gamma} \vdash e : \tau }
+  {\ctyp{r}{\Gamma} \vdash e : \tau }
+\;(s\leq r)  
+%
+\quad\quad\quad\quad
+%
+\inference[(\emph{abs})\;]
+  {\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
+\end{equation*}
+\caption{Type and coeffect system for the coeffect calculus}
+\label{fig:coeffect-typing}
+\vspace{-1em}
+\end{figure}
 
-\subsection{Example: Liveness analysis}
-\label{sec:structural-neededness}
+% ==================================================================================================
 
-\subsection{Example: Data-flow (revisited) }
-\label{sec:structural-tainting}
+\section{Coeffect semantics using indexed comonads}
+\label{sec:comonads}
 
-\todo{Also, consider additional language features that we consider for flat coeffects
-  (mainly recursion and possibly conditionals)}
+The approach of \emph{categorical semantics} interprets terms as
+morphisms in some category.  For typed calculi, typing judgments $x_1 : \tau_1 \ldots x_n : \tau_n
+\vdash e: \tau$ are usually mapped to morphisms $\interp{\tau_1} \times \ldots \times
+\interp{\tau_n} \rightarrow \interp{\tau}$. Moggi showed the semantics of various effectful 
+computations can be captured generally using the (\emph{strong}) \emph{monad}
+structure~\cite{monad-notions}. Dually, Uustalu and Vene showed that
+(\emph{monoidal}) \emph{comonads} capture various kinds of context-dependent 
+computation~\cite{comonads-notions}.
+
+We extend Uustalu and Vene's approach to give a semantics for
+the coeffect calculus by generalising comonads to
+\emph{indexed comonads}. We emphasise semantic intuition and
+abbreviate the categorical foundations for space reasons.
+
+\vspace{-1em}
+\paragraph{Indexed comonads.}
+
+Uustalu and Vene's approach interprets well-typed terms 
+as morphisms $C (\tau_1 \times \ldots
+\times \tau_n) \rightarrow \tau$, where $C$ encodes contexts
+and has a comonad structure~\cite{comonads-notions}.
+Indexed comonads comprise a \emph{family} of object mappings 
+$C^r$ indexed by a coeffect $r$ describing the contextual requirements satisfied by
+the encoded context. We interpret judgments $C^r (x_1 : \tau_1, \ldots, x_n : \tau_n) \vdash e : \tau$ as
+morphisms $C^r (\interp{\tau_1} \times \ldots \times \interp{\tau_n}) \rightarrow \interp{\tau}$.
+
+The indexed comonad structure provides a notion of composition
+for computations with different contextual requirements.
+%%
+\begin{definition}
+Given a monoid $(S, \imult, \iunit)$ with binary operator $\imult$ and
+unit $\iunit$, an \emph{indexed comonad} over a category $\ccat$
+comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
+$\iftor r A \in \obj{\ccat}$ and:
+%%
+\begin{itemize}
+\item a natural transformation $\counit_A : \iftor{\iunit} A \rightarrow A$, called the \emph{counit};
+\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
+  r A \rightarrow B$ to \\ morphisms $\iftor{r \imult{} s} A
+  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
+  \emph{coextend};
+\end{itemize}
+%
+\noindent
+such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
+the following equations hold:
+%
+\renewcommand{\arraycolsep}{1.5em}
+\begin{align*}
+\begin{array}{ccc}
+\counit \circ \cobind{f}{r, \iunit} = f & 
+\cobind{{(\counit)}}{\iunit, r} = \ident{id} & 
+\cobind{(g \circ \cobind{f}{r, s})}{(r \imult s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \imult t)} \\
+\end{array}
+\end{align*}
+\end{definition}
+%
+The \emph{coextend} operation gives rise to an associative composition operation for
+computations with contextual requirements (with \emph{counit} as the identity):
+%
+\[
+\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
+  \rightarrow (\iftor{r \imult s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
+g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
+\]
+%
+The composition $\hat{\circ}$ best expresses the intention
+of indexed comonads: contextual requirements of the composed
+functions are combined. The properties of
+the composition follow from the indexed comonad laws and the 
+monoid $(S, \oplus, \iunit)$.
+
+\paragraph{Example}
+Indexed comonad are analogous to comonads (in coKleisli form), but
+with the additional monoidal structure on indices.  Indeed, comonads
+are a special case of indexed comonads with a trivial singleton
+monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\iftor 1$
+is the underlying functor of the comonad and $\counit$ and
+$\cobind{(-)}{1, 1}$ are the usual comonad operations.
+However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.
+
+\paragraph{Example}
+The \emph{indexed partiality comonad} encodes free-variable contexts
+of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
+\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
+\sqcap, \ident{L})$, where $\iftor{\ident L} A = A$ encodes live
+contexts and $\iftor{\ident D} A = 1$ encodes dead contexts,
+where $1$ is the unit type inhabited by a single value $()$.
+The \emph{counit} operation $\counit{} : \iftor{\ident{L}} A \rightarrow A$
+is defined $\counit\ x = x$ and \emph{coextend},
+for all $f : \iftor r A \rightarrow B$, and thus
+$\cobind{f}{r, s} : \; \iftor{r \sqcap s} A \rightarrow \iftor s B$,
+is defined:
+%
+\begin{align*}
+\begin{array}{llll}
+\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
+\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
+\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
+\cobind{f}{\ident{L}, \ident{L}} x = f~x 
+\end{array}
+\end{align*}
+%
+The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
+data type $\mathsf{Maybe}\ A = A + 1$.  This
+type does not permit a comonad structure since
+$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
+For the indexed comonad, $\counit$ need only be defined
+ for $\iftor{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
+ contextual notions of computation than comonads.
+
+ Moreover, indexed comonads are not restricted by the \emph{shape
+   preservation} property of comonads~\cite{orchard12codo}: that
+ a coextended function cannot change the \emph{shape} of the
+ context. For example, in the second case above $f^\dagger_{\ident{D},
+   \ident{L}} : \iftor{\ident{D}} A \rightarrow \iftor{\ident{L}} B$
+ where the shape changes from $1$ (empty context) to $B$ (available
+ context).
+
+\vspace{-1em}
+\subsection{Monoidal indexed comonads.}
+Indexed comonads provide a semantics to sequential composition, but
+additional structure is needed for the semantics of the full coeffect calculus.
+Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
+structure, which provides a monoidal operation $\mathsf{m} : C A
+\times C B \rightarrow C (A \times B)$ for merging contexts (used in
+the semantics of abstraction).
+
+The semantics of the coeffect calculus requires an indexed
+lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
+\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
+by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
+%
+\begin{itemize}
+\item $\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \wedge s)}(A \times B)$ natural
+in $A, B$;
+\item $\csplit_{r, s} : \iftor{(r \vee s)}(A \times B) \rightarrow \iftor r A \times \iftor s B$ natural
+in $A, B$;
+\end{itemize}
+%
+\noindent
+The $\cmerge_{r, s}$ operation merges contextual computations with
+tags combined by $\wedge$ (greatest lower-bound), elucidating 
+the behaviour of $\cmerge_{r, s}$: that merging may result in 
+the loss of some parts of the contexts $r$ and $s$.
+
+The $\csplit_{r, s}$ operation splits context-dependent computations
+and thus the contextual requirements. To obtain coeffects $r$ and
+$s$, the input needs to provide \emph{at least} $r$ and $s$, so the
+tags are combined using the $\vee$ (least upper-bound).
+
+For the sake of brevity, we elide the indexed versions of the laws required by Uustalu and Vene
+(\eg~most importantly, merging two contexts and then adding the third is
+equivalent to merging the last two and then adding the first; similar rule holds is required
+for splitting).
+
+\paragraph{Example}
+For the indexed partiality comonad, given the liveness coeffect
+algebra $(\{\ident{D}, \ident{L}\}, \sqcap, \sqcup, \sqcap, \ident{L})$,
+the additional lax/colax monoidal operations are:
+%%
+\begin{align*}
+\begin{array}{lll}
+\cmerge_{\ident{L}, \ident{L}} (x, y) = (x, y) 
+  \quad\quad& \csplit_{\ident{D}, \ident{D}} \;\; () \;\;\; = ((), ())
+  \quad\quad& \csplit_{\ident{D}, \ident{L}} (x, y) = ((), y) \\
+\cmerge_{\ident{r}, \ident{s}} \;\,(x, y) = () 
+  & \csplit_{\ident{L}, \ident{D}} (x, y) = (x, ())
+  & \csplit_{\ident{L}, \ident{L}} (x, y) = (x, y)
+\end{array}
+\end{align*}
+
+\paragraph{Example}
+Uustalu and Vene model causal dataflow computations using the
+non-empty list comonad $\ident{NEList}~A = A \times (1 +
+\ident{NEList}~A)$~\cite{comonads-notions}.  Whilst this comonad
+implies a trivial indexed comonad, we define an indexed comonad with
+integer indices for the number of past values demanded of the context.
+
+We define $\ctyp{n}{A} = A \times (A \times \ldots \times A)$ where the first $A$ is the current
+(always available) value, followed by a finite product of $n$ past values. The definition of the
+operations is a straightforward extension of the work of Uustalu and Vene.
+
+%\begin{example}
+%The semantics of the coeffect calculus for
+% dynamically-scoped implicit parameters, with coeffect
+%algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
+%is provided by the indexed comonad $\iftor rA = A \times (r \rightarrow \tau)$
+%where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
+%do we need to store the types too? Maybe should just be a product.}
+%%
+%\vspace{-1.0em}
+%\begin{equation*}
+%\begin{array}{l}
+%\cobind{f}{r, s} \; : \; \iftor{r \cup s} A \rightarrow \iftor s B\\
+%\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
+%\\[1em]
+%\counit{} (a, g) = a\\
+%%\iota_{r, s} \; : \: \iftor r A \rightarrow \iftor s A\\
+%\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
+%\end{array}
+%\quad\begin{array}{l}
+%\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \cup s)}(A \times B)\\
+%\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
+%\\[1em]
+%\csplit_{r, s} : \iftor{(r \cup s)}(A \times B) \rightarrow \iftor r A \times \iftor s B \\
+%\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
+%\end{array}
+%\end{equation*}
+%\end{example}
+
+%The function created using \emph{coextend} expects a function $g$ that is defined for
+%resource names $r \cup s$. We use function restriction $g|_r$ and $g|_s$ to get functions that are 
+%defined on resources required by $f$ and the result of coextend, respectively.
+
+%The operation $\cmerge_{r, s}$ pairs the values and combines the resources available in both contexts. 
+%It combines a function $g_1|_{r\setminus s}$ (defined for resources in $r$ excluding those in $s$) and 
+%function $g_2$ (defined for resources $s$). This definition is not symmetric as it prefers resources 
+%defined in the second context. As a result, the caller of a function can rebind resources defined
+%at the declaration site.
+
+%Resources show that indexed comonads describe the structure of context more precisely
+%than comonads. A comonad would have to use $TA = A \times (\Omega \rightharpoonup R)$ 
+%with a \emph{partial} function on all resource names. Using indexed comonads, we can ensure that the function
+%is defined on all resources tracked statically in the tag.\dnote{Needs tweaking}
+
+\vspace{-1.5em}
+\subsection{Categorical Semantics. }
+\label{sec:comonads-semantics}
+
+\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
+\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}
+
+\begin{figure*}[t]
+\vspace{-1em}
+\newcommand{\uu}[1]{\overline{#1}}
+% \begin{equation*} 
+% \cosem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
+%   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
+% \end{equation*}
+% \vspace{-0.4em}
+\begin{equation*}
+\begin{array}{rcll}
+\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
+    \emph{curry} \; (\cosem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
+\\[0.1em]
+\cosem{\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
+    (\emph{uncurry} \;  \cosem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
+ & & \quad\; (id \times \cobind{\cosem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
+    \circ \csplit_{r, s \imult t} \circ \iftor{r \vee (s \imult t)}\Delta
+\\[0.1em]
+\cosem{\ctyp{\iunit}{\Gamma} \vdash x_i : \tau_i} &=&
+    \pi_i \circ \counit                                                                  % &$(3)$
+%\\[0.1em]
+%\cosem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
+%    \cosem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
+%\\[0.1em]
+%\cosem{\ctyp{r\vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
+%    \cosem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
+% ((\varepsilon \times \cosem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
+%\iftorhat{r \vee (r \imult t)}\Delta
+\end{array}
+\end{equation*}
+\vspace{-1.5em}
+\caption{Categorical semantics for the coeffect calculus}
+\label{fig:semantics}
+\vspace{-1.0em}
+\end{figure*}
+
+Figure~\ref{fig:semantics} shows the
+categorical semantics of the coeffect calculus
+using additional operations 
+$\pi_i$ for projection of the \emph{i}$^{th}$ element of a 
+product, usual
+\emph{curry} and \emph{uncurry} operations, and
+ $\Delta : A \rightarrow A \times A$ duplicating a
+value. While $\iftor{r}$ is a family of object mappings,
+it is promoted to a family of functors with the derived morphism mapping 
+$\iftor{r}(f) = \cobind{ (f \circ \counit) }{\iunit, r}$.
+
+The semantics of variable access and abstraction are the same as in
+Uustalu and Vene's semantics, modulo coeffects. Abstraction uses $\cmerge_{r, s}$ to merge the outer
+context with the argument context for the context of the function body.
+The indices of $\iunit$ for
+$\counit$ and $r,s$ for $\cmerge_{r, s}$ match the coeffects
+of the terms.
+The semantics of application is more complex. It first duplicates the
+free-variable values inside the context and then splits this context
+using $\csplit_{r, {s \oplus t}}$. The two contexts (with
+different coeffects) are passed to the two sub-expressions,
+where the argument subexpression, passed a context $(s \oplus t)$,
+is coextended to produce a context $s$ which is passed into
+the parameter of the function subexpression (\emph{cf.} given $f : A \rightarrow (B \rightarrow C)$,
+ $g : A \rightarrow B$, then $\emph{uncurry} \, f \, \circ \, (id \times g) \circ \Delta : A \rightarrow C$).
+
+A semantics for sub-coeffecting is omitted, but may be provided
+by an operation 
+$\iota_{r, s} : \iftor r A \rightarrow \iftor s A$ natural in $A$, for all $r, s \in S$ where $s \leq r$,
+which transforms a value $C^r A$ to $C^s A$ by
+ignoring some of the encoded context.
+
+% ==========================================================================================
+
+\section{Syntax-based equational theory}
+\label{sec:syntactic}
+
+Operational semantics of every context-dependent language differs as the notion of context 
+is always different. However, for coeffect calculi satisfying certain conditions we can 
+define a universal equational theory. This suggests a pathway to an operational semantics for 
+two out of our three examples (the notion of context for data-flow is more complex).
+
+In a pure $\lambda$-calculus, $\beta$ and $\eta$ equality for
+functions (also called \emph{local soundness} and \emph{completeness}
+respectively~\cite{logic-modal-reconstruction})
+describe how pairs of abstraction and application can be eliminated:
+$(\lambda x . e_2) e_1 \equiv_\beta \subst{e_1}{x}{e_2}$ and $(\lambda x . e \, x) 
+\equiv_\eta e$. The $\beta$ equality rule, using the usual Barendregt convention of 
+syntactic substitution, implies a \emph{reduction},
+giving part of an operational semantics for the calculus.
+
+The call-by-name evaluation strategy modelled by $\beta$-reduction is
+not suitable for impure calculi therefore a restricted $\beta$ rule,
+corresponding to call-by-value, is used, \ie~$(\lambda x . e_2) v
+\equiv \subst{e_2}{x}{v}$. Such reduction can be encoded by a
+\emph{let}-binding term, $\kvd{let}~x=e_1~\kvd{in}~e_2$, which
+corresponds to sequential composition of two computations, where the
+resulting pure value of $e_1$ is substituted into
+$e_2$~\cite{monads-inaction,monad-notions}.
+
+%
+% We consider here both a notion of \emph{let}-binding for the coeffect
+% calculus, useful for a CBV evaluation, and a notion of substitution 
+% for a CBV evaluation.
+%
+
+For an equational theory of coeffects, consider first a notion 
+of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
+has the following type and coeffect rule:
+%
+\begin{equation}
+\inference
+  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
+   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
+  {\ctyp{r_1 \vee (r_2 \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
+: \tau_2 }
+\label{eq:let1}
+\end{equation}
+
+\noindent
+For our examples, $\wedge$ is idempotent (\ie{}, $r \wedge r = r$)
+implying a simpler rule:
+
+% For the three examples we consider, a simpler rule gives a more
+% precise coeffect. Because $r \wedge r = r$ for all our examples, we
+% can also a coeffect $r \vee (r \imult s)$. Moreover,
+%for our examples (but not necessarily for \emph{all} coeffect
+%systems), $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$ meaning
+%that the following gives more precise coeffects:
+%
+\begin{equation}
+\inference
+  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
+   \ctyp{r}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
+  {\ctyp{r \vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2 : \tau_2 }
+\label{eq:let2}
+\end{equation}
+%%
+For our examples (but not necessarily \emph{all} coeffect
+systems), this defines a more ``precise'' coeffect with respect to $\leq$
+where $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$.
+
+This rule removes the non-principality of the first rule
+(\ie~multiple possible typings).  However, using idempotency 
+to split coeffects in abstraction would remove additional
+flexibility needed by the implicit parameters example.
+
+The coeffect $r \vee (r \imult s)$ can
+also be simplified for all our examples, leading to more intuitive
+rules -- for implicit parameters $r \cup (r \cup s) = r \cup s$; for
+liveness we get that $r \sqcup (r \sqcap s) = r$ and for dataflow we
+obtain $\textit{max}(r, r+s) = r + s$.
+
+Our calculus can be extended with \emph{let}-binding and \eqref{eq:let2}.
+However, we also consider the cases when a
+syntactic substitution $e_2[x \leftarrow e_1]$ has the coeffects
+specified by the above rule \eqref{eq:let2} 
+and prove \emph{subject reduction} theorem
+for certain coeffect calculi.  We consider two common special cases
+when the coeffect of variables $\iunit$ is the greatest ($\top$) or
+least ($\bot$) element of the semi-lattice $(S, \vee)$ and derive
+additional conditions that have to hold about the coeffect algebra:
+
+\begin{lemma}[Substitution]
+\label{thm:subst}
+Given $C^r (\Gamma, x : \tau_2) \vdash e_1 : \tau_1$ and $C^s \Gamma \vdash e_2 : \tau_2$
+then $C^{r \vee (r \oplus s)} \Gamma \vdash \subst{e_2}{x}{e_1} : \tau_1$ if 
+the coeffect algebra satisfies the conditions that 
+$\iunit$ is either the greatest or least element of the semi-lattice,
+$\oplus = \wedge$, and $\oplus$ distributes over $\vee$,
+\ie{}, $X \oplus (Y \vee Z) = (X \oplus Y) \vee (X \oplus Z)$.
+\end{lemma}
+
+\begin{proof}
+By induction over $\vdash$, using the laws (\S\ref{sec:calculus}) and additional assumptions.
+\end{proof}
+
+Assuming $\rightarrow_\beta$ is the usual call-by-name reduction, the
+following theorem models the evaluation of coeffect calculi with
+coeffect algebra that satisfies the above requirements. We do not
+consider \emph{call-by-value}, because our calculus does not have a
+notion of \emph{value}, unless explicitly provided by
+\emph{let}-binding (even a function ``value'' $\lambda x.e$ may have
+immediate contextual requirements).
+
+\begin{theorem}[Subject reduction]
+\label{thm:reduction}
+For a coeffect calculus, satisfying the conditions of Lemma~\ref{thm:subst}, if
+$\ctyp{r}{\Gamma} \vdash e : \tau$ and $e \rightarrow_\beta e'$ then 
+$\ctyp{r}{\Gamma} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+A direct consequence of Lemma~\ref{thm:subst}. 
+\end{proof}
+
+The above theorem holds for both the liveness and resources examples,
+but not for dataflow.  In the case of liveness, $\iunit$ is the
+greatest element ($r \vee \iunit = \iunit$); in the case of
+resources, $\iunit$ is the \emph{least} element ($r \vee \iunit =
+r$) and the proof relies on the fact that additional
+context-requirements can be placed at the context $\ctyp{r}{\Gamma}$
+(without affecting the type of function when substituted under
+$\lambda$ abstraction).
+
+However, the coeffect calculus also captures context-dependence in
+languages with more complex evaluation strategies than
+\emph{call-by-name} reduction based on syntactic substitution.  In
+particular, syntactic substitution does not provide a suitable evaluation
+for dataflow (because a substituted expression needs to capture the
+context of the original scope).
+
+Nevertheless, the above results show that -- unlike effects --
+context-dependent properties can be integrated with
+\emph{call-by-name} languages. Our work also provides a model of
+existing work, namely Haskell implicit parameters
+\cite{app-implicit-parameters}.
+
+% ==================================================================================================
+
+\section{Related and further work}
+\label{sec:related}
+
+This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
+and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
+\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
+may split contextual requirements between the declaration-site and call-site.
+
+Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
+\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
+and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
+\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
+general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).
+
+% --------------------------------------------------------------------------------------------------
+
+\vspace{-1em}
+\paragraph{Comonads and modal logics.}
+
+Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
+$\square$ modality of an intuitionistic S4 modal logic using monoidal
+comonads, which links our calculus to modal logics.  This link can be
+materialized in two ways.
+
+Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
+correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
+building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
+\cite{monad-notions}) that includes $\square$ as a type
+constructor. For example, in \cite{logic-modal-reconstruction}, the
+modal type $\Box \tau$ represents closed terms.
+In contrast, the \emph{semantic} approach uses monads or comonads
+\emph{only} as a semantics.  This has been employed by Uustalu and
+Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
+follow the semantic approach.
+
+Nanevski et al. extend an S4 term language to a \emph{contextual}
+modal type theory (CMTT)~\cite{logic-cmtt}.
+The \emph{context} is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. Our contextual types are
+indexed by a coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, \emph{etc.}.
+
+The work on CMTT suggests two extensions to coeffects. The first is
+developing the logical foundations. We briefly considered special cases
+of our system that permits local soundness in \S\ref{sec:syntactic} and
+local completeness can be treated similarly. The second problem is 
+developing the coeffects \emph{metalanguage}. The use of coeffect algebras
+would provide an additional flexibility over CMTT, allowing a wider range 
+of applications.
+
+% --------------------------------------------------------------------------------------------------
+
+\vspace{-1em}
+\paragraph{Relating effects and coeffects.} 
+The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
+semantic model (monads vs. comonads) is very different, we can consider extending the two to 
+obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
+the reader monad needs an operation that eagerly performs some effects of a function: 
+$(\tau_1 \rightarrow \mtyp{r \oplus s}{\tau_2}) \rightarrow \mtyp{r}{(\tau_1 \rightarrow \mtyp{s}{\tau_2})}$.
+To obtain a pure lambda abstraction
+for coeffects, we need to restrict the $\cmerge_{r, s}$ 
+operation of indexed comonads, so that the first parameter is annotated with $\iunit$ (meaning
+no effects): $\iftor \iunit A \times \iftor r B \rightarrow \iftor{r}(A \times B)$.
+
+\newcommand{\cprd}{\times}
+\newcommand{\cvop}{\oplus}
+
+\vspace{-1em}
+\paragraph{Structural coeffects.} To make the liveness analysis practical, we need to associate
+information with individual variables (rather than the entire context). We can generalize the 
+calculus from this paper by adding a product operation $\times$ to the coeffect algebra.
+A variable context $x:\tau_1, y:\tau_2, z:\tau_3$ is then annotated with
+$r\times s \times t$ where each component of the tag corresponds to a single variable. The system
+then needs to be extended with structural rules such as:
+%
+\begin{equation*}
+\inference[(\emph{abs})]
+  {\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
+\quad
+\inference[(\emph{contr})]
+  {\ctyp{r \cprd s}{(x:\tau_1, y:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r \cvop s }{(z:\tau_1)} \vdash \subst{\subst{e}{x}{z}}{y}{z} : \tau_2 }
+\end{equation*}
+%
+The context-requirements associated with function are exactly those linked to the specific
+variable of the lambda abstraction. Rules such as contraction manipulate variables and perform
+a corresponding operation on the indices.
+
+The structural coeffect system is related to bunched typing \cite{types-bunched} (but generalizes
+it by adding indices). We are currently investigating how to use structural coeffects to capture 
+fine-grained context-dependence properties such as secure information flow \cite{app-secure-flow}
+or, more generally, those captured by dependency core calculus \cite{types-dcc}.
+
+% ==================================================================================================
 
@@
 
-\todo{ (...) }
+We examined three simple calculi with associated
+static analyses (liveness analysis, implicit parameters, and dataflow
+analysis). These were unified in the \emph{coeffect calculus},
+providing a general coeffect system parameterised by an
+algebraic structure describing the propagation of context
+requirements throughout a program.
+
+We model the semantics of coeffect calculus using \emph{indexed comonad} -- a novel structure, which
+is more powerful than (monoidal) comonads. Indices of the indexed comonad operations manifest the 
+semantic propagation of context such that the propagation of information in the general coeffect
+type system corresponds exactly to the semantic propagation of context in our categorical model.
+
+We consider the analysis of context to be essential, not least for the examples here but 
+also given increasingly rich and diverse distributed systems.

### 2014-03-27 text/Chapter05.tex
Finish introduction chapter and restructure the rest

--- 2013-06-11 text/Chapter05.tex
@@
-\chapter{Coeffect meta-language} 
-\label{ch:metalanguage}
+\chapter{Structural coeffect language} 
+\label{ch:structural}
+
+\newcommand{\ctimes}{ \hat{\times} }
+\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
+
+\newcommand{\csprd}[0]{ \times }
+\newcommand{\csvop}[0]{ \vee }
+
+% alculus names and rule names
+\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
+\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
+
+\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
+\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
+\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
+\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
+\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}
 
@@
 \section{Introduction}
-\label{sec:metalanguage-intro}
 
-?
-\ExecuteMetaData[rules/rules.tex]{cml-types}
+% --------------------------------------------------------------------------------------------------
 
-\section{Summary}
-Oops!
+\subsection{Motivation: Tracking array accesses}
+\label{sec:structural-motivation}
+%\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
+%\newcommand{\dnat}{\ident{D}_\ident{nat}}
+
+Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
+functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
+respectively, but we use richer tag structure.
+
+As an example, consider a language that allows us to get a value of a variable (representing
+some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
+To track information about individual variables, we use a product-like operation $\csprd$ on tags 
+to mirrors the product structure of variables. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+\ctyp{5 \csprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
+  \vdash
+    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
+\end{array}
+\end{equation*}
+%
+The coeffect tag $5 \csprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
+that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
+for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
+a single variable:
+%
+\begin{equation*}
+\begin{array}{l}
+\ctyp{5 \csvop 10}{(\ident{c}:\dnat)}
+  \vdash
+    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
+\end{array}
+\end{equation*}
+%
+In this example, the operation $\csvop$ would be the \emph{max} function and so $5 \csvop 10 = 10$.
+Before looking at the formal definition, consider the typing of let bindings:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
+\pastval{a}{15}+\pastval{c}{10}
+\end{array}
+\end{equation*}
+%
+The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
+is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
+The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.
+
+The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
+context is $\ctyp{0 \csprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
+of $\csvop$ and it denotes empty coeffect. The typing context of the
+body is $\ctyp{15 \csprd 10}{(\ident{a}, \ident{c})}$.
+
+To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
+of the context in which \ident{c} was defined using the $\csvop$ operation. This is then combined 
+with the remaining tags from the body yielding the overall context:
+$\ctyp{15 \csprd (10 \csvop (0 \csprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
+normalization mechanism (described later), this can be further reduced to
+$\ctyp{(15 \csvop 10) \csprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
+we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
+\ident{b}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffect tags}
+In the previous section, the $\csvop$ operation behaves similarly to the flat $\vee$ operation.
+However, the type system does not require some of the semilattice properties, because some uses
+are replaced with the $\csprd$ operation. We do not require any properties about the $\csprd$ operation.
+For example, in the previous example cannot be commutative (since a tag $15 \csprd 10$ has different 
+meaning than $10 \csprd 15$). However, we relate the operations using distributivity laws to allow 
+normalization that was hinted above.
+
+\begin{definition}
+A \emph{structural coeffect tag structure} $(S, \csprd, \csvop, 0, 1)$ is a tuple where 
+$(S, \csvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
+by a binary operation $\csprd$ and element $1\in S$ such that
+for all $r,s,t\in S$, the following equalities hold: 
+%
+\begin{subequations}
+\begin{align}
+\tag{associativity}
+r \csvop (s \csvop t) &= (r \csvop s) \csvop t\\
+\tag{commutativity}
+r \csvop s &= s \csvop r\\
+\tag{lower bound}
+r \csvop 0 &= r\\
+\tag{distributivity}
+r \csvop (s \csprd t) &= (r \csvop s) \csprd (r \csvop t)\\
+\tag{upper bound}
+1 \csvop r &= 1
+\end{align}
+\end{subequations}
+%
+\end{definition}
+%
+% 'upper bound' (not sure what to call it) says that empty context cannot carry information
+% looks like this is the upper bound of the bounded lattice ???
+%
+% commutativity is not strictly required by any law or type preservation, but
+% it simplifies the presentation and looks reasonable
+%
+The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
+always accessed using standard variable access). The tag $1$ is used to annotate empty 
+variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
+to carry an annotation. We explain exactly how this works when we introduce the type system.
+The fact that $1$ is the upper bound means that combining it with other coeffect annotations
+does not affect it and so empty contexts cannot carry any information.
+
+The structure generalizes the \emph{flat coeffect tag structure} introduced in 
+Section~\ref{sec:flat-tag-structure}, but it additionally requires a special element $1$ 
+representing the upper bound. Given a flat coeffect structure, we can construct a structural
+coeffect structure (but not the other way round). For certain structures, the element $1$
+may be already present, but in general, it can be added as a new element.
+This construction will be important in Section~\ref{sec:generalization},
+where we show that \clstr~calculus generalizes~\clflt.
+
+\begin{lemma}
+\label{thm:structural-generalizaton}
+A flat coeffect tag structure $(S, \vee, 0)$ implies a structural coeffect tag structure.
+\end{lemma}
+\begin{proof}
+Take $1 \notin S$, then $(S \cup \{1\}, \vee, \vee, 0, 1)$ is a structural coeffect tag 
+structure. From the properties of flat coeffect tag structure, we get that the $\vee$ operation 
+is associative, commutative and $0$ is the unit with respect to $\vee$.
+
+To prove the distributivity, we need to show $r \vee (s \vee t) = (r \vee s) \vee (r \vee t)$.
+This easily follows from commutativity, associativity and idempotence of the flat coeffect tag 
+structure.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffect type system}
+\label{sec:structural-types}
+
+The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
+The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
+a \emph{structural coeffect tag structure} $(S,\csprd, \csvop, 0)$.
+
+The system differs from \clflt~in a significant aspect. It contains explicit 
+structural rules that manipulate with the context. Such rules allow reordering, duplicating
+and other manipulations with variable context. Such rules are known from affine or linear
+type systems where they are removed to obtain more restrictive system. In our system, the
+rules are present, but they manipulate the variable structure $\Gamma$ as well as the 
+associated tag structure $r$.
+
+As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
+simple set. Instead, we use the following tree-like structure (which is more similar to 
+bunched types than to linear or affine systems):
+%
+\begin{equation*}
+\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
+\end{equation*}
+%
+The syntax $()$ represents an empty context, so the structure defines a binary trees where
+leaves are either variables or empty. Contexts such as $\ctyp{1\csprd(r \csprd 1)}{((), (x, ()))}$ 
+contain unnecessary number of empty contexts $()$. However, we need to construct them
+temporarily, because certain rules require splitting a context and, by our 
+definition, the context $(x:\tau)$ is not splitable.
+
+The typing rules of the system are shown in Figure~\ref{fig:structural-coeffect-typing}.
+Many of the structural rules are expressed in terms of a helper judgement
+$\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}}$.
+
+\paragraph{Standard rules.}
+Variable access (\rulest{var}) annotates the corresponding variable with an empty coeffect
+$0$. The \rulest{fun} rule assumes that the context of the body can be split into the
+variable of the function and other (potentially empty) context and it attaches the coeffect
+associated with the function variable to the resulting function type $\ctyp{s}{\tau_1} \rightarrow \tau_2$.
+
+The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
+expression, argument and the function type respectively. 
+The result of evaluating the argument in the context $t$ is passed to the function that
+requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
+the combination of coeffects $r \csvop t$. The variable context required to evaluate the 
+function value is independent and so it is annotated just with coeffects $s$.
+Finally, the \rulest{let} rule is derived from let binding and application, but we show
+it separately to aid the understanding.
+
+\paragraph{Structural rules.}
+The remaining rules are not syntax-directed. They are embedded using the \rulest{ctx}
+rule and expressed using a helper judgement $\ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}}$
+that says that the context on the left-hand side can be transformed to the context on the
+right-hand side. The transformation can be applied to any part of the context, which 
+is captured using the \rulest{nest} rule (it is sufficient to apply the transformation 
+on the left part of the context; the right part can be transformed using \rulest{exch}).
+
+The \rulest{empty} rule allows attaching empty context to any existing context. The rule 
+is needed to type-check lambda abstractions that do not capture any outer variable.
+The \rulest{weak} rule is similar, but it represents \emph{weakening} where an unused variable is 
+added. The associated coeffect tag does not associate context information with the variable.
+This is needed to type-check lambda abstraction that does not use the argument.
+
+The \rulest{contr} rule is a limited form of contraction. When a variable appears repeatedly,
+it can be reduced to a single occurrence. The rule is needed to satisfy the side condition
+of \rulest{fun} when the body uses the argument repeatedly. The two associativity rules
+together with \rulest{exch} provide ways to rearrange variables. Finally, \rulest{sub} represents
+sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.
+
+\ExecuteMetaData[rules/rules.tex]{scl-types}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Properties of reductions}
+\label{sec:structural-metatheory}
+Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
+operational meaning, because that will differ for every concrete application. For example,
+when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
+access to a specified element.
+
+Just like previously, we can state general properties of the reductions. As the syntax of expressions
+is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
+are also the same and can be found in Figure~\ref{fig:flat-substred}.
+
+The structural coeffect calculus \clstr~associates information with individual variables. This means
+that when an expression requires certain context, we know from what scope it comes -- the
+context must be provided by a scope that defines the associated variable, which is either a
+lambda abstraction or global scope. This distinguishes the structural system from the flat system
+where context could have been provided by any scope and the lambda rule allowed arbitrary 
+splitting of context requirements between the two scopes (or declaration and caller site).
+
+\paragraph{Internalized substitution.}
+Before looking at properties of the evaluation, we consider let binding, which can be 
+viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
+application and abstraction as follows.
+
+\begin{lemma}[Definition of let binding]
+If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
+then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
+\end{lemma}
+\begin{proof}
+The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
+correspond with the typing rule \rulest{let}:
+\begin{equation*}
+\inference[]
+  {\inference
+    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
+    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
+   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
+  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
+\end{equation*}
+\end{proof}
+
+The term $e_2$ which is substituted in $e_1$ is checked in a different variable
+and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
+a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
+the variable that we are substituting for. The let binding captures substitution for a 
+specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
+For a general substitution, we need to define the notion of context with a hole.
+
+\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
+\paragraph{Substitution and holes.}
+In \clstr, the structure of the variable context is not a set, but a tree. When substituting
+for a variable, we need to replace the variable in the context with the context of the
+substituted expression. In general, this can occur anywhere in the tree. To formulate the
+statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
+in the free variable context and in a corresponding part of the coeffect tag:
+
+\begin{equation*}
+\begin{array}{rcl}
+\Delta[-] &::=& \ctyp{1}{()} \\
+       & \sep & \ctyp{r}{(x:\tau)}\\
+       & \sep & \ctyp{-}{(-)} \\
+       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
+\end{array}
+\end{equation*}
+%
+Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
+operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
+$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
+Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
+general substitution lemma for \clstr.
+
+\begin{figure}
+
+\begin{equation*}
+\begin{array}{rcl}
+                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
+                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
+                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
+\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
+&\multicolumn{2}{l} {
+  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
+}
+\end{array}
+\end{equation*}
+\caption{The definition of hole filling operation for $\Delta[-]$}
+\label{fig:structural-hole-fill}
+\vspace{-2em}
+\end{figure}
+
+\begin{lemma}[Substitution Lemma]
+\label{thm:structural-subst}
+If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
+then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
+\end{lemma} 
+\begin{proof}
+Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
+$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
+\end{proof}
+
+\begin{theorem}[Subject reduction]
+\label{thm:structural-reduction}
+If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
+$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
+\end{theorem}
+\begin{proof}
+Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
+\end{proof}
+
+\paragraph{Local soundness and completeness.}
+As with the previous calculus, we want to guarantee that the introduction and
+elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
+This can be done by showing \emph{local soundness} and \emph{local completeness},
+which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
+case of subject reduction and the latter is proved by a simple derivation:
+%
+\begin{theorem}[Local soundness]
+If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
+$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
+\end{theorem}
+\begin{proof}
+Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
+\end{proof}
+%
+\begin{theorem}[Local completeness]
+If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
+$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
+\end{theorem}
+\begin{proof} The property is proved by the following typing derivation:
+%
+\begin{equation*}
+\inference
+  {\inference
+     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
+      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
+     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
+  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
+\qedhere
+\end{equation*}
+\end{proof}
+%
+In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
+guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
+$\lambda x.f x$ required for local completeness was not the only possible derivation.
+In the last step, it was possible to split the coeffect tag arbitrarily between the 
+context and the function type.
+
+In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
+to be associative and to have units and so a unique splitting may exist.  For example, if we
+define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
+$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
+has additional properties, then there may be other possible derivation.
+
+% ==================================================================================================
+
+\begin{figure*}[t]
+\newcommand{\uu}[1]{\overline{#1}}
+\begin{equation*}
+\cosem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
+  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
+\end{equation*}
+\begin{equation*}
+\begin{array}{rcll}
+\cosem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
+    \epsilon_0                                                                   &\quad$\ruless{var}$
+\\[0.32em]
+\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
+    \Lambda (\cosem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
+\\[0.32em]
+\cosem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
+    (\lambda (\gamma_1, \gamma_2) \rightarrow
+    \cosem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
+    (\cobind{\cosem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
+\\[0.32em]
+\cosem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
+    \cosem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \cosem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
+\end{array}
+\end{equation*}
+
+\begin{equation*}
+\begin{array}{rclll}
+\cosem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
+&=& \cmerge_{r, s} \circ (\cosem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
+\textnormal{\rulessc{nest}}
+\\[0.8em]
+\cosem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
+&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
+\textnormal{\rulessc{exch}}
+\\[0.8em]
+\cosem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
+&=&\ident{fst} \circ \csplit_{r, 1}&&
+\textnormal{\rulessc{empty}}
+\\[0.8em]
+\cosem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
+&=&\ident{fst} \circ \csplit_{r, 0}&&
+\textnormal{\rulessc{weak}}
+\\[0.8em]
+\cosem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
+    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
+&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
+\textnormal{\rulessc{contr}}
+\\[0.8em]
+\cosem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
+    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
+&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
+        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
+\textnormal{\rulessc{assoc}}
+\\[0.8em]
+\cosem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
+&=&
+\iota_{r, s}
+&&
+\textnormal{\rulessc{sub}}
+\end{array}
+\end{equation*}
+
+\caption{Categorical semantics for~\clstr}
+\label{fig:structural-semantics}
+\vspace{-0.5em}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Semantics of structural coeffects}
+\label{sec:structural-semantics}
+
+The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
+\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
+of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
+is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
+
+As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
+system is not a simple finite product, but instead a binary tree. To model this, we do not use
+ordinary products in the domain of the semantic function, but instead use a special 
+constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
+the tag structure.
+
+\subsection{Structural tagged comonads}
+\label{sec:structural-tagged-comonads}
+
+To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
+Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
+$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
+$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
+context-requirements.
+
+However, functions that do not exist in context have only a single input variable (with a 
+single corresponding tag). To model complex variable contexts, we need two additional
+operations that allow manipulation with the variable context. Similarly to the model of \clflt,
+we also require operations that model duplication and sub-coeffecting:
+
+\begin{definition}[Structural tagged comonad]
+Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
+tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
+$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
+$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
+\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
+\end{array}
+\end{equation*}
+%
+% Delta is still defineable
+% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
+%
+And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
+%
+%  TODO: laws?
+%
+\end{definition}
+
+The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
+and it can still be used to define a family of mappings that represents \emph{duplicating}
+of variables while splitting the additional coeffect tags:
+%
+\begin{equation*}
+\begin{array}{l}
+\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
+\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
+\end{array}
+\end{equation*}
+%
+The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
+but with two differences. Firstly, it combines tags using $\csprd$ instead of $\vee$, 
+which corresponds to the fact that the variable context now consists of two parts (a tree node).
+Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
+(instead of $A \times B$ as previously).
+
+To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
+of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
+variables. The operation was not needed for \clflt, because there \emph{splitting} could
+be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
+the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
+with $r \csprd s$ that carries $A \ctimes B$.
+
+Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
+Before looking at them, we finish our discussion of categorical semantics.
+
+\paragraph{Categorical notes.}
+The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
+Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
+variables in the context. Given a context modelled as $T^{r \csprd s}(A \ctimes B)$,
+we can lift a morphism $f$ to $\hat{T}^{r \csprd s}(f)$, but we cannot manipulate the
+variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
+
+This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
+could apply $\cdupl_{r, s}$ on the input (if the tag $r \csprd s$ coincided
+with tag $r \csvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
+
+This restriction is intentional -- at the semantic level, it prevents manipulations
+with the context that would break the correspondence between tag structure and the
+product structure.
+
+\subsection{Categorical semantics}
+The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
+It uses the \emph{structural tagged comonad} structure introduced in the previous section,
+together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
+operations:
+\begin{equation*}
+\begin{array}{rcl}
+\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
+\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
+f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
+\end{array}
+\end{equation*}
+
+When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
+notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
+the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
+variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
+$\csprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
+type signature in $\cmerge_{r, s}$).
+
+The rule \ruless{app} now uses the operation $\csplit_{s, (r \csvop t)}$ instead of $\cdupl_{s, (r \csvop t)}$,
+which means that it splits the context instead of duplicating it. This makes the system more 
+structural -- the expressions use disjunctive parts of the context -- and also explains why the
+composed coeffect tag is $s \csprd (r \csvop t)$.
+
+The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
+of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
+of $\cosem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
+$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
+takes a context, converts it to a new context which is compatible with the original expression $e$.
+The context manipulation rules work as follows:
+
+\begin{itemize}
+\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
+  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
+  Finally, they re-construct a single context using $\cmerge_{r, s}$.
+
+\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
+  and discard one part (containing either an unused variable or an empty context).
+
+\item If we interpreted \rulessc{contr} by applying functor $T^{r\csvop s}$ to a function that duplicates
+  a variable, the resulting context would be $\ctyp{r \csvop s}(x:\tau, x:\tau)$, which would break the
+  correspondence between coeffect tag and context variable structure. However, that interpretation would be
+  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
+  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
+  a tag $r \csvop s$ into $r \csprd s$.
+
+\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
+  manipulates it (using \ident{assoc}) and then re-constructs it.
+
+\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
+  variable using the primitive natural transformation $\iota_{r, s}$.
+\end{itemize}
+
+\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
+attaching the context separately to individual variables. This an expression
+$\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
+$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
+approach largely complicates the definition of application (where tag of all 
+variables in a context is affected). Moreover, it makes it impossible to express
+\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.
+
+\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
+transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
+operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
+sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).
+
+This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
+typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
+the following natural transformation has to be provided:
+%
+\begin{equation*}
+\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
+\end{equation*}
+%
+This variant of the system could be used to define a system that ensures that all provided 
+context is used and is not over-approximated. This difference is similar to the difference between
+affine type systems (where a variable can be used at most once) and linear type systems 
+(where a variable has to be used exactly once).
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Examples of structural coeffects}
+
+\subsection{Example: Liveness analysis}
+\label{sec:structural-neededness}
+
+\subsection{Example: Data-flow (revisited) }
+\label{sec:structural-tainting}
+
+\todo{Also, consider additional language features that we consider for flat coeffects
+  (mainly recursion and possibly conditionals)}
+
+\section{Conclusions}
+
+\todo{ (...) }

### 2014-03-27 text/Chapter06.tex
Finish introduction chapter and restructure the rest

\chapter{Unified coeffect language} 
\label{ch:unified}

% ==================================================================================================
% Macros - for ICFP 2014
% ==================================================================================================

% comonadic semantics
\newcommand{\F}{\mathsf{F}}
\newcommand{\D}{\mathsf{D}}
\newcommand{\Ds}[3]{\D^{\sclrd{#1}}_{#2} #3}
\newcommand{\Icat}{\cclrd{\mathbb{I}}}
\newcommand{\id}{\mathit{id}}
\newcommand{\B}[1]{\bar{#1}}
\newcommand{\bbullet}{\mathop{\bar{\bullet}}}
\renewcommand{\idf}[0]{ \ident{id} }
\newcommand{\m}{\mathsf{m}}
\newcommand{\n}{\mathsf{n}}

\newcommand{\shortar}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar_{#4}"#1";"#2"}
\newcommand{\shortaro}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar^{#4}"#1";"#2"}
\newcommand{\shortarn}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar_{#4}"#1";"#2"}
\newcommand{\shortaron}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar^{#4}"#1";"#2"}
\newcommand{\diagar}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}^{#4}"#1";"#2" }
\newcommand{\diagaro}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}_{#4}"#1";"#2" }

% Typing rule and typing statement (single line)
\newcommand{\tyrule}[3]{ \inference[(\footnotesize{#1})~]{#2}{#3} }
\newcommand{\tyruleCompact}[3]{ \text{(\footnotesize{#1})}~\dfrac{#2}{#3} }
\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
% Context transformation and context transformation with new line in the middle
\renewcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
% Ordinary context where colouring is done by hand 
\newcommand{\coctx}[2]{ #1 \text{\scriptsize @} {\footnotesize{\text{$#2$}}}  }

% Simple coeffect systems in the introduction
\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}

% Coeffect algebra
\definecolor{cclr}{rgb}{0.0,0.4,0.0}
\newcommand{\cclrd}[1]{ \textcolor{cclr}{#1} }

\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }} % \bullet} } % \textcolor{cclr}{\bullet}}  
\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
\newcommand{\cleq}{ \textcolor{cclr}{\leq} }

% \newcommand{\single}[{\langle{#1}\rangle}

\newcommand{\uncurry}{\Lambda^{\hspace{-0.2em}-1}}

% Coeffect shapes 
\definecolor{sclr}{rgb}{0.5,0,0.0}
\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }

\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\renewcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }

\newcommand{\sflat}{\star}

% Coeffect algebra
\definecolor{aclr}{rgb}{0.0,0.0,0.7}
\newcommand{\aclrd}[1]{ \textcolor{aclr}{#1} }

\newcommand{\mksh}[1]{{\sclrd{[}#1\sclrd{]}}}

\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq 
% Tomas: I prefer \circledast here because it is a circled (lifted) version of \cseq 
% { \textcolor{aclr}{\bullet}}
\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}} 
\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }

% Free monoid used in the *structural* coeffect algebra
\newcommand{\fcons}{ \times } 
\newcommand{\fempty}{\epsilon}

%===================================================================================================

\section{Introduction}

\emph{Context} is important for defining meaning-- not just in natural languages, but also in logics and 
programming languages.
The standard notion of context in programming is an environment 
providing values for free variables. An open term with free variables is context dependent -- its 
meaning depends on the free-variable context. The simply-typed $\lambda$-calculus famously 
analyses such context usage. 
Other systems go further. %capturing how 
%the context is used. 
For example, bounded linear logic tracks the number of times a 
variable is used~\cite{girard1992bounded}.

In software engineering, ``context'' provides more than just free-variable values. For example, in a
distributed system, the context provides different resources that may be available on different 
devices (\eg{}, a database on a server or a GPS sensor on a phone).

In this paper we develop a calculus for capturing various notions of 
context in programming. A key feature and contribution
of the calculus is its \emph{coeffect system} which provides a static analysis for contextual properties (coeffects). The system
follows the style of type and effect systems, but captures a different class of properties. % (Section~\ref{sec:intro-pathways}). 
Another key contribution of the calculus is its semantics which can be smoothly instantiated for specific notions of context.
%The semantics of the calculus is general and can be instantiated for particular notions of context.

Coeffect systems were previously introduced as a generic analysis of
context dependence which can be instantiated for various notions of
context~\cite{petricek2013coeffects}. However, the formalization was restricted to
tracking a class of \emph{whole-context} properties where a term has just
one coeffect. This limited the applications and precision of any
analysis. For example, a whole-context liveness analysis marks the 
free-variable context as live (some variable may be used) or dead 
(no variable is used), but it cannot record liveness \emph{per-variable}.

We develop a more general system which captures both
\emph{per-variable} coeffects, which we call \emph{structural}, and
\emph{whole-context} coeffects, which we call \emph{flat}, and more. Our
key contributions are:

\begin{itemize}
\item We present a general coeffect type system (Section~\ref{sec:coeffects}) and demonstrate two 
  concrete uses -- \emph{flat} coeffect systems that tracks whole-context information and 
  \emph{structural} coeffect systems which tracks fine-grained per-variable information.

\item We show practical examples, instantiating the calculus for structural systems capturing 
  variable usage based on bounded linear logic, dataflow caching, and precise liveness analysis.
  We also instantiate the calculus to flat systems, building on and extending previous examples~\cite{petricek2013coeffects}.
% variants of the coeffect system capture examples described   in our earlier work

\item We discuss the syntactic properties of the \emph{structural} coeffect system 
  (Section~\ref{sec:syntax}). It satisfies type preservation under both $\beta$-reduction 
  and $\eta$-expansion, allowing its use with both call-by-name and call-by-value languages.
  This important property distinguishes it from both effect systems and flat coeffects.

\item We provide a denotational semantics, resisting and extending the notion of \emph{indexed comonads} to the 
structural setting (Section~\ref{sec:semantics}). We prove soundness by showing 
  the correspondence between syntactic and semantic properties of coeffect systems.
\end{itemize}

\noindent
Coeffects can be approached from multiple directions (Section~\ref{sec:intro-pathways}) including 
syntactic (effect systems), semantic, and proof-theoretic. We 
emphasize the syntactic view, though we also outline a categorical semantics
and note the interesting technical details.

% ==================================================================================================

\section{Why coeffects matter}
\label{sec:intro-coeffects}

Coeffects are way to describe notions of context in programming that keep turning up. 
To illustrate this, we overview three systems tracking contextual properties that 
 motivate our general coeffect system. Two systems track per-variable properties (bounded 
linear logic and dataflow) and one tracks whole-context properties (implicit parameters).
 We start with some background and finish with a brief overview of the literature leading to coeffects.

\subsection{Background, scalars and vectors}

The $\lambda$-calculus is asymmetric-- it maps a context with \emph{multiple} variables to a 
\emph{single} result. An expression with $n$ free variables of types $\sigma_i$ can be modelled
by a function
$\sigma_1 \times \ldots \times \sigma_n \rightarrow \tau$ with a product on the left, but a single value
on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
we attach coeffects to the context $\sigma_1 \times \ldots \times \sigma_n$ and we often (but not always)
have one coeffect per each variable. We call the overall coeffect a \emph{vector} consisting of
\emph{scalar} coeffects. This asymmetry explains why coeffect systems are not trivially dual to 
effect systems.

It is useful to clarify how vectors are used in this paper. Suppose we have a set $\C$ of
\emph{scalars} such as $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ over $\C$
is a tuple $\langle \cclrd{r_1}, \ldots, \cclrd{r_n}\rangle$ of scalars. 
We use letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for 
vectors and $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars.\footnote{For better readability, the paper 
distinguishes different structures using colours. However ignoring the colour does not introduce 
any ambiguity.} We also say that a \emph{shape} of a vector $\mksh{\aclrd{R}}$ (or more generally any container) 
is the set of \emph{positions} in a vector. So, a vector of length
$n$ has shape $\{ 1, 2, \ldots, n \}$.

Just as in scalar-vector multiplication, we lift any binary operation on scalars into a scalar-vector one:
$\cclrd{s} \bullet \aclrd{R} = \langle \cclrd{s}\bullet\cclrd{r_1}, \ldots, \cclrd{s}\bullet\cclrd{r_n}\rangle$.
Given two vectors $\aclrd{R},\aclrd{S}$ of the same shape, containing partially ordered scalars, we 
write $\aclrd{R} \leq \aclrd{S}$ for the pointwise extension of $\leq$ on scalars. Finally,
the associative operation $\times$ concatenates vectors.

We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.

\subsection{Bounded reuse}
\label{sec:intro-coeffects-bll}

Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
This is done by limiting the number of times a value (proposition) can be used. An assumption
$!_k A$ means that a variable can be used at most $k$ times. 
We attach annotations to the whole context rather than individual assumptions and so a context
$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.

Bounded linear logic includes explicit weakening and contraction rules
that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
\[
\inference
  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
\quad
\inference
  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
\]
The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
occurrences of a proposition are joined by adding the number of uses.

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\begin{array}{c}
%
\hspace{-0.6em}
\begin{array}{c}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
\quad 
\tyruleCompact{weak}
  {\coctx{\Gamma}{R} \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
\\[1.8em]
\tyruleCompact{sub}
  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
\quad 
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
\\[1.2em]
\end{array} \\
%
% ------------------------ 
%
\begin{array}{c}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
\\[1.3em]
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
\\[1.3em]
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
\end{array} 
% ------------------------ 
%
% \hspace{-1em}
%\begin{array}{c}
%
% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
% to keep it simple.
%
% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
% (I do not understand why but it makes the line somehow thicker...)
%
% \tyruleCompact{let}
%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
\end{array}
\end{align*}
\vspace{-0.5em}
\caption{type and coeffect system for bounded reuse}
\label{fig:bounded-coeff}
\end{figure}

\paragraph{Bounded linear coeffects.}
The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
exchange (\emph{exch}) rules manipulate variables in the context and
modify the annotations accordingly -- adding the number of uses in
contraction and switching vector elements in exchange.

For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.

Application (\emph{app}) describes call-by-name evaluation.  Applying
a function that uses its parameter $\cclrd{t}$-times to an argument
that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
vector. Meanwhile, the variables in $\Gamma_1$ are used just
$\aclrd{R}$-times when reducing the expression $e_1$ to a function
value.

Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
the number of uses using the pointwise
$\leq$ relation. We can view any variable as being used a greater
number of times than it actually is.

\paragraph{Example.} To demonstrate, consider a term
$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
times -- once directly inside the function and twice via the variable $v$ after 
substitution. Similarly, $y$ is used 
twice. Assuming a judgment for the function body, abstraction yields:
%
\begin{equation*}
\tyrule{abs}
 { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
 { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
\end{equation*}
%
To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
a judgment that is simplified as follows:
\[
\hspace{-0.7em}\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
{\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
\]
%
The first step performs scalar multiplication, producing the vector
$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
$x$ and $x'$ from the function and argument terms respectively.

It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.

% --------------------------------------------------------------------------------------------------

\subsection{Dataflow and data access}

Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
An expression is re-evaluated when new inputs are available (push) or when more output
is demanded (pull).  In causal dataflow, programs can access past
values of a stream. We consider a language where $\kvd{prev}~e$
returns the previous value of $e$, where 
$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.

An implementation of causal dataflow may cache past values of
variables as an optimisation. The question is, how many past values
should be cached?  This can be approximated by a coeffect system.

\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
similar to the one for bounded reuse in that it tracks a vector
of numbers $\aclrd{R}$ as part of the context
$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
past values (\emph{causality depth}) required for a variable.

Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
accessed variables (\emph{var}) are annotated with
$0$ meaning that no past value is required (only the
current one). The (\emph{prev}) rule crates caching requirements --
it increments the number of required values for all variables
used in $e$ using scalar-vector addition.

Application and contraction have the same structure as before, 
but use different operators. If two variables 
are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
values, then overall we need at most
$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
two caches are combined with the maximum of the two requirements, 
which satisfy the smaller requirements.

In (\emph{app}), the function requires $\cclrd{t}$ past values of
its parameter.  This means $\cclrd{t}$ past values of
$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
call (\eg{}, we need $1+\aclrd{S}$ values to
get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
values to get $2$ past values of $\sigma$, \emph{etc.}).

\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
and one bound:
\[
\dfrac
  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
\]
The expression always requires the previous value of $y$ and adds it to
a previous value of the parameter $x$. Evaluating the value of the
argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
and so the overall requirement is $3$ past values:
\[
\inference
  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
{\inference
  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
\]
The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
number of enclosing $\kvd{prev}$s.% expressions.

We previously used dataflow as an example of
coeffects~\cite{petricek2013coeffects}, but tracked caching
requirements on the whole context. The system outlined here is more
powerful and practically useful, with finer-grained coeffects tracking
caching requirements per-variable.

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\[
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
\]
\[
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
\]
\[
\tyrule{var}{}
  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
\quad\;
\tyrule{prev}
  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
\]
\caption{type and coeffect system for dataflow caching}
\label{fig:dataflow-coeff}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Implicit parameters}
\label{sec:intro-coeffects-impl}

As our third example, we revisit Haskell implicit parameters \cite{lewis2000implicit} 
used in our earlier coeffect work \cite{petricek2013coeffects}. Implicit parameters are variables
that mix aspects of dynamic and lexical scoping. Implicit parameters are a distinct syntactic 
category to variables and we write them as \ident{?p}. For simplicity, we omit \emph{let}-binding for
implicit parameters and focus just on tracking requirements.

\paragraph{Implicit parameters coeffects.} Implicit parameters are a whole-context coeffect 
not linked to ordinary variables. We keep track of sets of implicit parameters that are required
by an expression (and their types). For example $\coctx{\Gamma}{\{ \ident{?p}_1 : \tau_1, \ldots, \ident{?p}_n : \tau_n \}}$
means that a context provides ordinary variables $\Gamma$ and values for implicit parameters $\ident{?p}_i$.
Unlike in the previous examples, we no longer need to distinguish between coeffects attached
to variables (scalars) and coeffects attached to contexts (vectors), so we write $\cclrd{r},\cclrd{s},\cclrd{t}$
for both.

Despite the differences, the type system in Figure~\ref{fig:implpar-coeff} follows the same 
structure as the earlier two examples. Context requirements are created when accessing an implicit
parameter (\emph{param}) (a system-specific rule). Structural rules (exchange, weaken, contract) 
do not affect the coeffects. For example parameters are reordered in (\emph{exch}),
but this has no effect as set union $\cup$ is commutative.

In abstraction and application, the structural $\cons$ operator (previously vector 
concatenation) becomes $\cup$. Sets of implicit parameters are not associated to
individual variables and so they are unioned. The (\emph{app}) rule uses $\cup$ to
combine the implicit parameters required by the function with the requirements of the argument too.

We call this a \emph{flat} coeffect system since coeffects have only
one shape (there is no scalar/vector distinction).  Other flat
coeffect systems may use a richer structure~\cite{petricek2013coeffects}. In
particular, the operations used in abstraction and application may differ
(to accommodate over-approximation). We return to this in
Section~\ref{sec:coeffects-structural}.

\paragraph{Example.} Unlike structural coeffect systems, flat systems
do not necessarily have principal coeffects. 
This arises from the (\emph{abs}) rule which can freely split
requirements between the function type and the declaring
context. Consider a function %filters news based on a location:
$\lambda ().\ident{?p}_1 + \ident{?p}_2$.  There are nine possible 
 type and coeffect derivations, two of which are:
%% 1 p1p2  p1p2
%% 2 p1p2  p1
%% 3 p1p2  p2
%% 4 p1p2  
%% 5 p1    p1p2
%% 6 p1    p2
%% 7 p2    p1p2
%% 8 p2    p1
%% 9       p1p2
\[
\begin{array}{rl}
\coctx{\emptyset}{ \{  \} } &\hspace{-0.7em}\vdash (\ldots) : 
  \ident{unit} \xrightarrow{ \{ \ident{?p}_1:\mathbb{Z}, \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
\\
\coctx{\emptyset}{ \{ \ident{?p}_1:\mathbb{Z} \} } &\hspace{-0.7em} \vdash (\ldots) : 
  \ident{unit} \xrightarrow{ \{ \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
\end{array}
\]
In the first case, both parameters are dynamically scoped and have to be provided by the caller.
In the second case, the parameter $\ident{?p}_1$ is available in the declaring scope and so it is
(lexically) captured.

Although structural coeffects have more desirable syntactic properties, we aim to capture this
non-principality too as it is practically useful -- Haskell's implicit parameters use it and
it can be used to model resource rebinding in distributed systems such as~\cite{app-distributed-acute}).

% --------------------------------------------------------------------------------------------------

\begin{figure}
\[
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau,y:\sigma,\Gamma_2}{r \cup \cclrd{s} \cup \cclrd{t} \cup \cclrd{q}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x:\tau,\Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s} \cup \cclrd{q}} \vdash e : \tau}
\]
\[
\tyrule{app}
  { \coctx{\Gamma_1}{\cclrd{r}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
    \coctx{\Gamma_2}{\cclrd{s}} \vdash e_2 : \sigma }
  { \coctx{\Gamma_1, \Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s}} \vdash e_1 \, e_2 : \tau} 
\]
\[
\tyrule{param}{}
  {\coctx{()}{ \{ \ident{?p}:\tau \} } \vdash \ident{?p} : \tau}
\quad\;
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
\]
\vspace{-0.5em}
\caption{Type and coeffect system for implicit parameters}
\label{fig:implpar-coeff}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Pathways to coeffects}
\label{sec:intro-pathways}

This paper mainly follows work on effect systems and their link to categorical semantics. We briefly 
review this and other directions leading to coeffects. An eager reader can return to this section later.

\paragraph{Effect systems.}

Effect systems \cite{gifford1986effects} track
effectful operations of computations such as memory access or lock
usage \cite{flanagan2003type}. They are written as judgments $\Gamma
\vdash e : \tau ~\&~\cclrd{\rho}$ associating effects
$\cclrd{\rho}$ with the result. Effect systems capture \emph{output
  effects} where, as Tate puts it, \emph{``all computations with [an]
  effect can be thunked as pure computations for a domain-specific
  notion of purity.''}~\cite{effects-producer-semantics}.  This
thunking is typically a $\lambda$-abstraction. Given an effectful
expression $e$, the function $\lambda x . e$ is an effect-free value
that delays all effects:
%
\begin{equation*}
\inference[({abs})\;]
  {\Gamma, x\!:\!\tau_1 \vdash e : \tau_2 ~\&~\cclrd{\rho}} 
  {\Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{\rho}} \tau_2~\&~\cclrd{\emptyset}}
\end{equation*}
%
Coeffects do not follow this pattern.  In contrast to effect systems,
context requirements cannot be easily ``thunked'' as
pure values. Lambda abstraction can split context requirements
between \emph{immediate} and \emph{latent} requirements. This is akin
to how lambda abstraction splits a free-variable context into 
the bound parameter (call site) and the remaining free variables (declaration site).

\paragraph{Categorical semantics.}
Moggi models effectful computations as functions of type 
$\tau_1 \rightarrow \mtyp{}{\tau_2}$ where $M$ is monad providing composition of effectful 
computations~\cite{moggi1991monads}. Wadler and Thiemann \cite{wadler2003marriage} link effect systems with monads using 
annotated monads $\tau_1 \rightarrow \mtyp{\cclrd{\rho}}{\tau_2}$ whose semantics has been 
provided by Katsumata~\cite{katsumata2014parametric}.

Context-dependent computations require a different model. Uustalu and Vene \cite{uustalu2008notions} 
use functions $C{}{\tau_1} \rightarrow \tau_2$ where $C$ is a \emph{comonad}. Our earlier work 
\cite{petricek2013coeffects} used indexed comonads with denotations $C^{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ adding 
annotations akin to Wadler and Thiemann. In Section~\ref{sec:semantics} we extend
indexed comonads to capture the general coeffect systems of this paper,  
in the style of Katsumata. 
% 
% TP: Mentioned Katsumata here. Thought it is nice to say we build on new work here (as opposed to ICALP)
%

\paragraph{Language and meta-language.}
Moggi uses monads in two systems~\cite{moggi1991monads}. In the first system, a monad is used to model
an effectful language itself -- the semantics of a language uses a specific monad. In the second system, 
monads are added as type constructors, together with syntax corresponding to \emph{unit} and \emph{bind}
operations.

Looking at context, Uustalu and Vene \cite{uustalu2008notions} follow the first approach 
(using a concrete comonad to model dataflow). Contextual-Modal Type Theory (CMTT) of Nanevski et al.
\cite{logic-cmtt} follows the latter approach, adding a comonad via the $\Box$ modality of modal S4
to the language. We focus on concrete languages using the first approach. 
A ``coeffect meta-language'' is an interesting future work.

\paragraph{Sub-structural and bunched types.}
% \cite{substruct-linear-change} 
Sub-structural type systems restrict how a context is used. This is achieved by removing some
of the structural typing rules (weakening, contraction, exchange). As the bounded linear logic 
example (Section~\ref{sec:intro-coeffects-bll}) shows, our system can be viewed as a generalization.

% ==================================================================================================

\section{The coeffect calculus}
\label{sec:coeffects}

The three calculi shown in the previous section track two kinds of
contextual properties: bounded reuse and dataflow are structural
(per-variable) systems, and implicit parameters and our earlier coeffect
systems~\cite{petricek2013coeffects} are flat (whole-context) systems. 
This section presents our primary contribution: the general coeffect
calculus.

The calculus is parameterised by an algebraic structure of coeffects.
To capture both structural and flat systems, coeffect annotations are
indexed by a \emph{shape}. In flat systems, the shape is a singleton set
and so annotations are \emph{scalar} values. Structural systems use shapes
matching the number of variables in a free-variable context and so annotations
are \emph{vectors}. However, the coeffect calculus could also use shapes describing
trees and other structures.

% --------------------------------------------------------------------------------------------------

\subsection{Understanding coeffects: syntax and semantics}

The coeffect calculus provides both an analysis of
context dependence (its coeffect system) and a semantics for
context (see Section~\ref{sec:semantics}). These two features of the calculus provide different
perspectives on coeffect annotations $\aclrd{R}$ in a judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$:
%%
\begin{itemize}
\item Syntactically, coeffects model \emph{contextual requirements} and 
  may be overapproximated, such that more capabilities are required 
  than necessary at runtime.

\item Semantically, coeffects model \emph{contextual capabilities} and behave 
  like containers of capabilities, such that the semantics may throw away 
  capabilities that will not be needed.
\end{itemize}
%
Thus there are two dual ways to understand coeffect annotations. Each
perspective implies an alternate reading of the typing rules.
%
\begin{itemize}
\item
As \emph{contextual requirements}, the rules should be read top-down.
The requirements of multiple sub-terms are \emph{merged} and the
requirements of function body are \emph{split} between immediate (declaration-site)
and latent (call-site) coeffects.

\item
As \emph{contextual capabilities}, the rules should be read bottom-up.
The capabilities provided to a larger term are \emph{split} between 
sub-terms; for functions, the capabilities of declaration-site and call-site
are \emph{merged} and passed to the body.
\end{itemize}
%
The reason for this asymmetry follows from the fact that context appears in a \emph{negative
position} in the model. In Section~\ref{sec:semantics}, the denotation of a judgment
is a function of the form $\D_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau}$ where $\D_{\aclrd{R}} \interp{\Gamma}$ 
encodes the contextual capabilities used to evaluate a term. Similarly
functions have models of the form $\D_{\cclrd{s}} \interp{\sigma} \rightarrow \interp{\tau}$
with additional contextual capabilities attached to the input.

\subsection{Structure of coeffects}

We describe the algebraic structure of coeffects in three
steps. First, we define a \emph{\cclrd{coeffect scalar}} structure 
which defines the basic building blocks of coeffect information;
then we define \emph{\sclrd{coeffect shapes}} which determines how
coeffect scalar values are related to the free-variable context. 
Finally, we define the \emph{\aclrd{coeffect
    algebra}} which consists of shape-indexed coeffect scalar values.

For example, in bounded reuse the coeffect scalar structure
comprised natural numbers $\cclrd{\mathbb{N}}$ with $\cclrd{+}$ and
$\cclrd{\ast}$ operators. The shape for bounded reuse is the length of the
free-variable context and so the coeffect annotation is a vector of matching length.
Finally, the coeffect algebra specifies how vectors are
concatenated and split in abstraction and application.

In the coeffect system of the calculus, contexts are annotated with shape-indexed coeffects (\eg{}, vectors)  as in
$\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$. However, functions take just a single 
input parameter and so are annotated with scalar coeffect values as in $\sigma \xrightarrow{\cclrd{r}} \tau$.

\paragraph{Coeffect scalar.}
Coeffect scalar structures are equipped with two operations. In bounded reuse, those 
were $\cclrd{\ast}$ for sequencing (in function application) and $\cclrd{+}$ for context 
sharing (in contraction). Additional structure is needed for variable access and sub-coeffecting.

\begin{definition}
\emph{\cclrd{Coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set $\C$
together with elements $\cunit,\czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids
and $(\C, \cleq)$ is a pre-order. The following distributivity axioms are required:
\begin{align*}
(\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

\noindent
The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
category in the semantics (Section~\ref{sec:semantics}). 
It models sequential composition with variable access
($\cunit$) as the identity. The other element ($\czero$) is used for
variables that are not accessed. The operation $\cpar$ combines coeffects
for contexts used in multiple places (contraction). 
The notation is inspired by the bounded reuse example, which uses coeffect scalar structure
$(\cclrd{\mathbb{N}}, \ast, +, 1,0, \cclrd{\leq})$, but be aware that $\cseq$ and $\cpar$ do
not always mean $\ast$ and $+$.

The context annotations can be viewed as \emph{containers} of scalar coeffects.
For structural coeffects, the container is a vector, while for flat coeffects, it is
a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{abbott2005containers}
which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.

\paragraph{Coeffect shapes.}
The coeffect system is parameterised by a set of shapes $\SH$. A coeffect annotation 
is indexed by a shape $\sclrd{s} \in \SH$ calculated from the shape of the free-variable vector. The correspondence is not necessarily 
bijective. For example, flat coeffect systems have just a single shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$, the coeffect annotation
$\aclrd{R}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
We write $\aclrd{s} = \mksh{\Gamma}$ for the shape corresponding to $\Gamma$. We define shapes by a 
\emph{set} of positions and so we can write $\aclrd{R} \in \sclrd{s} \rightarrow \C$ as a
mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
the exponent $\aclrd{R} \in \C^{\sclrd{s}}$.

The set of shapes is equipped with an operation that combines shapes (when we combine 
variable contexts), an operation that computes shape from the free-variable contexts, 
and two special shapes in $\SH$ representing empty context and singleton context.

\begin{definition}
  A \emph{\sclrd{coeffect shape}} structure $(\SH, \mksh{\textnormal{--}}, \stimes,
  \sempty, \sunit)$ comprises a set $\SH$ with a binary operation
  $\stimes$ on $\SH$ for shape composition, a mapping from contexts to
  shapes $\mksh{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
  \SH$ such that $(\SH, \stimes, \sempty)$ is a monoid. The 
  elements $\sempty$ and $\sunit$ represent the shapes of empty and
  singleton free-variable contexts respectively.
\end{definition}

\noindent
As said earlier, we use two kinds of shape structures that describe the shape
of vectors and the shape of trivial singleton container:

\begin{itemize}
\item
Structural coeffect shape is defined as $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$.
We treat numbers as sets $0 = \{\}, 1=\{ \emptyset \}, 2=\{ \emptyset, 1\}, 3=\{\emptyset, 1, 2 \} \ldots$ 
(so that a number is a set of positions). The shape mapping ${|}\Gamma{|}$ returns the number
of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, respectively,
and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = {|}\Gamma_1{|} + {|}\Gamma_2{|}$.
Therefore, a coeffect annotation is a \emph{vector} $\aclrd{R} \in
\C^{\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{R}(i) \in \C$ for each
variable $x_i$ in the context.

\item
Flat coeffect shape is defined as $\sclrd{(\{\sflat \}, \mathsf{const}\,\,\sflat, \diamond, \sflat, \sflat)}$ where
$\sflat\,\diamond\,\sflat = \sflat$ and $\star=\{\emptyset\}$. That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^{\sflat}$ which is iso\-morphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{itemize}

\noindent
Using a shape with \emph{no} positions reduces our system to the simply-typed $\lambda$-calculus with no context annotations.
Trees could be used to build a system akin to bunched typing \cite{substruct-bunched}.

\paragraph{Coeffect algebra.}
The coeffect calculus annotates judgments with 
shape-indexed (or, \emph{shaped}) coeffects.
The \emph{\aclrd{coeffect
    algebra}} structure combines a coeffect scalar and coeffect shape structures
to define shaped coeffects and operations for combining these.
In Section~\ref{sec:intro-coeffects}, shaped coeffects were combined
by the tensor $\times$ in structural examples and
$\cup$ in the implicit parameters example. To capture the examples so far and
those described previously~\cite{petricek2013coeffects}, we distinguish two
operators for combining shaped coeffects.

\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}

\begin{definition}
  Given a \cclrd{coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero,
  \cleq)$ and a \sclrd{coeffect shape} $(\SH, \mksh{\textnormal{--}}, \stimes,
  \sempty, \sunit)$ a \emph{\aclrd{coeffect algebra}} extends the two
  structures with $(\amerge, \asplit, \azero)$ where $\azero \in
  \C^{\sempty}$ is a coeffect annotation for the empty context and
  $\amerge, \asplit$ are families of operations that combine coeffect
  annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m}
  \in \SH$:
\[
\asplit_{\sclrd{m},\sclrd{n}}, \amerge_{\sclrd{m},\sclrd{n}} ~:~ \C^{\sclrd{m}} \times \C^{\sclrd{n}} \rightarrow \C^{\sclrd{m}\,\stimes\,\sclrd{n}}
\]
\end{definition}

\noindent
A coeffect algebra induces the following two additional operations:
\[
\begin{array}{rclrcl}
\alift{\textnormal{--}} &\narrow{:}& \C \rightarrow \C^{\sunit}          \quad\quad
   & \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{m}}  
\\ 
\alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}  
   & \cclrd{r}\ \aseq\ \aclrd{S} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\ \cseq\ (\aclrd{S}(\sclrd{s}))
\end{array}
\]
$\alift{\textnormal{--}}$ lifts a scalar coeffect 
to a shaped coeffect indexed by the singleton context shape. 
The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a vector by scalar.
As we always use lower-case for scalars and upper-case for vectors, using the same
symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write just $\aseq$.

The operators $\asplit$ and $\amerge$ combine shaped
coeffects associated with two contexts. For example,
assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects 
$\aclrd{R} \in \C^{\sclrd{m}}$ and $\aclrd{S} \in \C^{\sclrd{n}}$. In
the structural system, the context shapes $\sclrd{m}, \sclrd{n}$
denote the number of variables in the two contexts. The combined
context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and
the combined coeffects $\aclrd{R} \,\amerge\, \aclrd{S}, \aclrd{R} \,\asplit\, \aclrd{S} \in \C^{\sclrd{m} \stimes \sclrd{n}}$
are indexed by that shape.

For structural coeffect systems such as bounded reuse, both $\asplit$ and $\amerge$ are
just the tensor product $\times$ of vectors. However, we need to distinguish them for flat coeffect 
systems discussed later.

The difference is explained by the semantics (Section~\ref{sec:semantics}), where
$\aclrd{R}\,\amerge\,\aclrd{S}$ is an annotation of the codomain of 
a morphism that merges the capabilities provided by two contexts (in the syntactic 
reading, splits the context requirements); $\aclrd{R}\,\asplit\,\aclrd{S}$ is an annotation of
the domain of a morphism that splits the capabilities of a single context into two parts (in the
syntactic reading, merges their context requi\-rements). Syntactically, this means that we always use
$\amerge$ in the rule \emph{assumptions} and $\asplit$ in \emph{conclusions}. For now, it suffices to use the 
bounded reuse intuition and read the operations as tensor products.

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
$\boxed{
\coctx{\Gamma}{R} \vdash e : \tau
}$
\[
\begin{array}{c}
\tyrule{const}{\;}
  { \coctx{ () }{ \azero } \vdash c : \iota }
\quad\quad
\tyrule{var}{\;}
  { \coctx{ (x:\tau) }{ \aunit } \vdash x : \tau }
\\[1.8em]
\tyrule{abs}
  { \coctx{\Gamma, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{s}}} \vdash e : \tau }
  { \coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau }
\\[1.8em]
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \tau \xrightarrow{\cclrd{t}} \sigma &
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \tau }
  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash e_1~e_2 : \sigma } 
\\[1.8em]
\tyrule{let}
  { \coctx{\Gamma_1}{\aclrd{S}} \vdash e_1 : \sigma & 
    \coctx{\Gamma_2, x:\sigma}{\aclrd{R}\; \amerge\; \alift{\cclrd{t}}} \vdash e_2 : \tau  }
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R}\; \asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau }
\\[1.8em]
\tyrule{ctx}
  { \coctx{\Gamma}{\aclrd{R}} \vdash e : \tau & \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma'}{\aclrd{R'}} }{\theta}}
  { \coctx{\Gamma'}{\aclrd{R'}} \vdash \theta e : \tau }
\end{array}
\]

$\boxed{
\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}
}$

\[
\tystmt{weak}
  { \quad\quad 
    \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x:\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}
    \quad\quad\quad }
\]
\[ 
\tystmt{exch}
  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau, y:\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
               { \quad \coctx{\Gamma_1, y:\sigma, x:\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }{\emptyset} }
\]
\[
\tystmt{contr}
  { \ctxtransnl{ \coctx{\Gamma_1, y:\tau, z:\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]}}
\]
\[
\tystmt{sub}
  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau,\Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \aclrd{T} } }
               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s'}} \,\asplit\, \aclrd{T}} }{\emptyset} }
\quad(\cclrd{s} \;\cleq\; \cclrd{s'} )                 
\]
\caption{The general coeffect calculus}
\vspace{-1em}
\label{fig:coeffects-general}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{General coeffect type system}

In the previous section, we developed an algebraic structure capable of
capturing different concrete context-dependent properties discussed in
Section~\ref{sec:intro-coeffects}. Now, we use the structure to
define the general coeffect calculus in
Figure~\ref{fig:coeffects-general}.

Coeffect annotations on free-variable contexts
are shape-indexed coeffects $\aclrd{R}, \aclrd{S}, \aclrd{T} \in \C^\SH$ 
and function types are annotated with
coeffects scalars $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.  The
rules manipulate coeffect annotations using the operations provided by
coeffect algebra $(\amerge, \asplit, \azero)$ and the derived
constructs $\alift{\textnormal{--}}$ and $\aseq$. Free-variable contexts $\Gamma$
are treated as vectors modulo duplicate use of variables 
-- the associativity is built-in. The order of
variables matters, but can be changed using the structural rules. To
make the system easier to follow, structural rules are expressed
using a separate judgment.

\paragraph{Typing rules.} 
Constants (\emph{const}) and variable access (\emph{var}) annotate the context with special
values. Empty unused context is annotated with $\azero
\in \C^\sempty$, while singleton context 
is annotated with $\aunit \in \C^\sunit$.  Note that the shapes
$\sempty, \sunit$ match the shape of the variable contexts.

Lambda abstraction \emph{splits} the context requirements using $\amerge$ into a coeffect 
$\aclrd{R}$ and a coeffect $\alift{\cclrd{s}}$ of a shape $\sunit$ (semantically, it \emph{merges}
capabilities provided by the declaration-site and call-site contexts). In structural systems such
as bounded reuse, $\asplit$ is not symmetric and so this gives us a coeffect associated with the bound variable.

The (\emph{app}) rule follows the patterns seen
earlier -- it uses the scalar-vector multiplication ($\cclrd{t}\,\aseq\,\aclrd{S}$) on 
coeffects associated with $\Gamma_2$.
Using the syntactic reading, it then \emph{merges} context requirements for $\Gamma_1$ and $\Gamma_2$.
In the dual semantic reading,
it \emph{splits} the provided context into two parts passed to the sub-expressions.

The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
Syntactically, the context requirements are first split using $\amerge$ and then re-combined
using $\asplit$.

\paragraph{Structural rules.}
The coeffect-annotated context can be transformed using structural rules that are
not syntax-directed. These are captured by (\emph{ctx}), which uses a helper judgment representing
context transformations $\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}$.
The rule models that a context used in the rule conclusion $\coctx{\Gamma'}{R'}$ can be transformed
to a context required by the assumptions $\coctx{\Gamma}{R}$ (using the semantic bottom-up reading).
In the rule, $\theta$ is a variable substitution generated by the transformation, which is 
used in the (\emph{contr}) rule.

Exchange and contraction decompose and reconstruct coeffect annotations 
using $\amerge_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\asplit_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \mksh{\Gamma_1}$
and $\sclrd{n} = \mksh{\Gamma_2}$.

The (\emph{weak}) rule drops an ignored variable annotated with $\aweak$ (compare with (\emph{var})
annotated using $\aunit$). The (\emph{exch}) rule switches the values 
while (\emph{contr}) combines them using $\cpar$ to represent sharing of the context.
Finally, (\emph{sub}) represents sub-coeffecting that can be applied (point-wise) to any
individual coeffect.

% --------------------------------------------------------------------------------------------------

\subsection{Structural coeffects}
\label{sec:coeffects-structural}

The coeffect system uses a general notion of context shape, but
it has been designed with structural and flat systems in
mind. The structural system is new in this paper and so we
look at it first.

Recall the coeffect shapes that characterise structural systems: the shape is formed by natural
numbers (with addition) modelling the number of variables in the
context. The coeffect algebra is therefore formed by the free monoid
(vectors) over a coeffect scalar. This means that the system keeps a vector
of basic coeffect annotations -- one for each variable. An empty
context (\eg{}, in the (\emph{const}) rule) is annotated with an empty
vector.

\begin{definition}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural
coeffect system} has: 
\begin{itemize}{}
\item[--] Coeffect shape $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$ formed by natural numbers
\item[--] Coeffect algebra $\aclrd{(\fcons, \fcons, \fempty)}$ where $\fcons$ and $\fempty$ are
  shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
  That is $\fcons : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n + m}}$ appends vectors
  (lists) and $\fempty : \C^{\sclrd{0}}$ represents empty vectors (lists)
\end{itemize}
\end{definition}

\noindent
The definition is valid since the shape operations form a monoid
$\sclrd{(\mathbb{N}, +, 0)}$ and $\mksh{\textnormal{--}}$ (calculating the length of
a list) is a monoid homomorphism from the free monoid to the monoid of
shapes.

\paragraph{Examples.} 
Defining a concrete structural coeffect system is easy, we just provide the coeffect scalar structure and the rest is free.

\begin{itemize}
\item
To recreate the system for bounded reuse, we use coeffect scalars formed by
$(\mathbb{N}, \ast, +, 1, 0, \leq)$. As in the system of Figure~\ref{fig:bounded-coeff},
\emph{used} variables are annotated with $1$ and \emph{unused} with $0$. Contraction adds the number
of uses via $+$ and application (sequencing) multiplies the uses.

\item
\emph{Dataflow} uses natural numbers (of past values), but differently: $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$.
Variables are initially annotated with $0$ (and can be incremented using the \kvd{prev} keyword).
Annotations of a shared variable are combined by taking maximum (of past values needed) and
sequencing uses $+$.

\item
Another use of the system is to track \emph{variable liveness}. The annotations are formed by
$\C=\{ \mathsf{D}, \mathsf{L} \}$ where $\mathsf{L}$ represents a \emph{live} (used) variable
and $\mathsf{D}$ represents a \emph{dead} (unused) variable. The scalars are given as
$(\C, \sqcap, \sqcup, \mathsf{L}, \mathsf{D}, \sqsubseteq)$.

In sequential composition ($\sqcap$), a variable is live only if it is required by both of the 
computations ($\mathsf{L} \sqcap \mathsf{L} = \mathsf{L}$), otherwise it is marked as 
dead ($\mathsf{D}$). A computation is not evaluated if its result is not needed.
A shared variable ($\sqcup$) is live if either of the uses is live 
($\mathsf{D} \sqcup \mathsf{D} = \mathsf{D}$, otherwise $\mathsf{L}$).
\end{itemize}

\noindent
Structural liveness is a practically useful, precise version of an example from our earlier
work, which was a flat system overapproximating liveness to the entire
context~\cite{petricek2013coeffects}.

% --------------------------------------------------------------------------------------------------

\subsection{Flat coeffects}
\label{sec:coeffects-flat}

The same general coeffect system can be used to define systems that track whole-context coeffects
as in the implicit parameters example 
(Section~\ref{sec:intro-coeffects-impl}). Flat coeffect systems are
characterised by a singleton set of shapes, such as ${ \{ \sflat \} }$. In
this setting, the context annotations $\C^{ \sflat }$ are
equivalent to coeffect scalars $\C$.

In addition to the coeffect scalar structure, we also need to define $\asplit$ and $\amerge$.
Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\asplit$ (merging of shaped 
coeffect annotations). However, the $\amerge$ operation needs to be provided explicitly.

\begin{definition}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary 
$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, we define:
\begin{itemize}{}
\item[--] Flat coeffect shape $\sclrd{(\{ \sflat \}, \emph{const} \, \sflat, \diamond, \sflat, \sflat)}$ where 
  $\sflat \diamond \sflat = \sflat$
\item[--] Flat coeffect algebra $(\cclrd{\wedge}, \cpar, \czero)$, \ie{}, the $\asplit = \cpar$ and $\azero = 
\czero$ with the additional binary operation $\amerge = \wedge$. 
\end{itemize}
\end{definition}

\noindent
The requirement $(\cclrd{r \wedge s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$ guarantees 
exchange and contraction preserve the coeffect of the assumption in the conclusion. Thus, flat coeffect
calculi do not require substructural-style rules.

\paragraph{Examples.}
Implicit parameters are the prime example of a flat coeffect system, but other examples
include rebindable resources \cite{app-distributed-acute} and Haskell type classes \cite{orchard2013thesis}.

In the \emph{implicit parameters} system (Section~\ref{sec:intro-coeffects-impl}), 
the co\-effect scalars are sets of names with types 
$\C = \mathcal{P}(\ident{Name} \times \ident{Types}) $. Variables
are always annotated with $\emptyset$ and coeffects are combined or split using set union $\cup$. Thus
the system is given by coeffect scalar structure $(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \cup, \emptyset, \emptyset, \subseteq)$
with $\cclrd{\wedge} = \cup$.

\begin{remark}
We previously described flat systems for \emph{liveness} and \emph{dataflow}~\cite{petricek2013coeffects}.
Turning a structural system to flat requires finding $\wedge$ that 
underapproximates the capabilities of combined contexts. For dataflow, this is given by the \textsf{min} function as
$\mathsf{min}(\cclrd{r},\cclrd{s}) \,\cleq\, \mathsf{max}(\cclrd{r},\cclrd{s})$.

In flat \emph{dataflow}, we annotate the
entire context with the maximal number of past elements required
overall. We use the same coeffect scalars $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$
as in the structural version, but with
$\cclrd{\wedge} = \ident{min}$. Abstraction (which is the only rule using $\wedge$) becomes:
%%
\[
\tyrule{abs}
  { \coctx{\Gamma, x\!:\!\sigma}{\ident{min}(\cclrd{r}, \cclrd{s})} \vdash e : \tau }
  { \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau }
\]
%%
Both the declaration-site and the call-site need to provide at least
the number of past values required by the body. The overapproximation means that both $\cclrd{r}$
and $\cclrd{s}$ can be greater than actually required.
For dataflow, we could annotate
both contexts with the same coeffect,
but that would require treating $\amerge$ as a partial function.
\end{remark}

% ==================================================================================================

\section{Equational theory}
\label{sec:syntax}
\newcommand{\reduce}{\leadsto}

Each of the concrete coeffect systems discussed in this paper has a different notion of 
context-dependence, much like various effectful languages have different notions of effects
(such as state or exceptions). However, there are equational properties that hold for all (or some) of
the systems we consider.

The equational theory in this section illuminates 
the axioms of coeffect algebra and the semantics of the calculus.
We discuss syntactic substitution as it can form the basis for
reduction in a concrete operational semantics. 
We consider structural and flat systems
separately. This provides better insight into how the two systems work and
differ. In particular, call-by-name evaluation is \emph{coeffect
preserving} for all structural, but only some flat systems.

The properties and proofs in this section are syntactic.
In Section~\ref{sec:semantics-soundness} we show that
our denotational model of the coeffect calculus is sound with respect
to the equational theory here.

We use standard syntactic substitution written as $\subst{e_1}{x}{e_2}$, 
$\beta\textnormal{-reduction}$ and $\eta$-expansion, written as $\reduce_\beta$ and $\reduce_\eta$.
Equality of terms $e_1$ and $e_2$, defined by a relation
$\equiv$, requires the equality of their contexts, types, and 
coeffects, written $\coctx{\Gamma}{\aclrd{R}} \vdash e_1 \equiv e_2 : \tau$.

%\begin{definition}[Subject reduction]
%Given reduction $e \reduce e'$ and $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$
%then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$, \ie{}, $\reduce$ is 
%\emph{type-and-coeffect preserving}. 
%\label{def:pres}
%\end{definition}

%Definition~\eqref{def:pres} is also known as the \emph{subject reduction} property.

\subsection{Structural coeffect systems}
\label{prop:structural}

For structural coeffect systems, recall that coeffects are vectors
with $\asplit = \amerge = \cons$ (vector concatenation) and $\azero =
\langle{}\rangle$ (the empty vector), thus coeffect annotations
comprise the \emph{free monoid} \ie, lists over the coeffect scalars
(although we continue using the vector terminology). We first show
substitution:

\begin{lemma}[Substitution lemma] 
\label{thm:structural-subst}
In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\label{lemm:structural-subst}
\end{lemma}
\begin{proof}
By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
\end{proof}

\noindent
Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
Using the substitution lemma, we can demonstrate $\beta$-equality:
%
\begin{equation*}
\inference{
  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
\equiv
\subst{e_1}{x}{e_2} : \tau}
\end{equation*}
%%
\noindent
As a result, $\beta$-reduction preserves the type and coeffects of a term.
This gives the following subject reduction property:

\begin{theorem}[Subject reduction]
In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
\end{proof}

\noindent
Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
and application does not eliminate too much.
%
\begin{equation*}
\inference{
  \inference 
    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
: \sigma \xrightarrow{\cclrd{s}} \tau}
\end{equation*}
%
The last step uses the fact that 
$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.

This highlights another difference between coeffects and effects, as $\eta$-equality does
not hold for many notions of effect. For example, in a language with output
effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
the exact immediate requirements of $\lambda x . e x$ to match those of $e$.

\subsection{Flat coeffect systems}

The equational theory for flat coeffect systems is somewhat similar to effect systems
where (co)effects are not linked to individual variables. In effectful languages,
substituting an effectful computation for $y$ in $\lambda x.y$ changes the latent effect 
associated with the function.

Similarly, for some of the flat coeffect systems, substituting a context-dependent computation
for $y$ in $\lambda x.y$ adds latent context requirements to the function type. However, this is not
the case for \emph{all} flat coeffect systems -- for example, call-by-name reduction preserves
types and coeffects for the implicit parameters system (which makes it a suitable model for Haskell).
For other systems, we first briefly consider call-by-value reduction.

\paragraph{What is a value?}
The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
As discussed earlier, a function $(\lambda x.e)$ is not necessarily a value in coeffect calculi, 
because it may not delay all context requirements of $e$. Thus we say that $e$ is a value if it
has no immediate context requirements.

\begin{definition} 
An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
where $\aclrd{\mathsf{VAL}} : \C^{\mksh{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
\end{definition}

\paragraph{Call-by-value.}
In call-by-value, a pure \emph{value} is substituted for a variable. 
The right-hand side of an application is evaluated to a value before the
$\beta$-reduction, but the discharging of coeffects prior to substitution is
different for each concrete coeffect system.

Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
merging of context-requirements, respectively.

\begin{lemma}[Call-by-value substitution]
\label{thm:flat-subst-cbv}
In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator:
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\end{lemma}
\begin{proof}
By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
with $\cunit$.
\end{proof}

\noindent
Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
the operational semantics must provide a way of partially evaluating a term 
with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:

\begin{theorem}[Call-by-value reduction]
In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
\end{proof}

\paragraph{Call-by-name.}
The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.

% TODO
%
% The context requirements of the 
% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
% lemma should be $r \fmax (s \fseq r)$.

We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
Liveness analysis is an example of top-pointed coeffects
as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.

\begin{lemma}[Top-pointed substitution]
\label{thm:cbn-substitution-top}
In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator:
\[
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\end{lemma}
\begin{proof}
Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
\end{proof}
%
\noindent
As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).

In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
the context requirements arising from the substitution can be associated with the context
$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
(all three operators are set union) and allows the following substitution lemma:

\begin{lemma}[Bottom-pointed substitution]
\label{thm:cbn-substitution-bot}
In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
\[
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\end{lemma}
\begin{proof}
By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
the free-variable context.
\end{proof}
\noindent
The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
variant described here is flexible enough to let us always re-associate new context requirements with the 
free-variable context.

The two substitution lemmas show that the call-by-name evaluation strategy can be used for
certain coeffect calculi, including liveness and implicit parameters.
Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
reduction, the following theorem holds:

\begin{theorem}[Call-by-name reduction]
\raggedright
In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
\end{proof}

%
% TODO: We could have a section about let binding covering the alternative CBV and the let binding
% we included in the system, but we'd have to clean this up!
%
%\subsection{Syntactic let binding}
%\begin{remark}
%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
%calculus of Section~\ref{}.
%%%
%\begin{align*}
%{\footnotesize{\trule{let}}}
%\dfrac
%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
%\end{align*}
%%%
%Therefore, for example, we see that \emph{let} has the coeffects
%as substitution, for example:
%%%
%\begin{align*}
%& {\footnotesize{\trule{let}}}
%\dfrac
%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
%\end{align*}
%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
%the judgment:
%%%
%$$
%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
%$$
%%%
%\end{remark}
%
% 
%\paragraph{Syntactical let binding}
%Has the same typing derivation as abstraction/application
%
%\[
%\inference
% {\inference
%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
%  &
%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
%\]
%
%We can extend the CBV behaviour to \emph{let}, where the bound term
%is evaluated to a pure value first, discharging any contextual
%requirements before substitution into the body.
%
%\begin{lemma}[CBV \emph{let}]
%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
%: \tau$.
%\end{lemma}

% ==================================================================================================

\section{Semantics}
\label{sec:semantics}

\newcommand{\Di}{\F}

Coeffects provide a unified notion of context-dependence. In the 
previous sections, we used this to define a unified coeffect calculus.
We now define a unified (categorical) semantics for the 
coeffect calculus. The semantics can be instantiated for different 
notions of context dependence and thus can model a wide range of
context-aware languages (both for flat and structural systems).

We relate the semantics to the equational theory and show that it is
sound with respect to term equality. For a variant
of the the flat system, a similar result has already been shown in the 
second author's dissertation~\cite{orchard2013thesis}.
The semantics is introduced in pieces:

\begin{itemize}
\item
Section~\ref{sem:interp}
describes the range and domain (signature) of the interpretation $\interp{-}$, 
gives the interpretations for types and free-variable contexts (in
flat and structural systems), and defines the signature of
functors $\D$ which encode contexts.

\item
The first part of the semantics (Section~\ref{sem:seq}) defines \emph{sequential
composition} of context-dependent computations via \emph{indexed
  comonads} (introduced briefly in our previous
work~\cite{petricek2013coeffects}) and the \emph{indexed structural
  comonad} structure (new here).

\item
More structure is needed for application and abstraction. 
Section~\ref{sem:monoidal} defines
indexed monoidal operations for splitting and merging contexts.
Concrete structures are given throughout for the semantics of the
structural bounded reuse and flat implicit parameter systems.

\item
Section~\ref{sem:together} puts the pieces together, defining the
semantics of the coeffect calculus along with structural rules.  
The semantics is illustrated by executing an 
 example bounded-reuse program in 
the semantics (Example~\ref{exm:run}).% and the concrete structures for
%the semantics of structural dataflow (Example~\ref{exm:dataflow-full}).

\item
Section~\ref{sec:semantics-soundness} shows our
semantics sound with respect to the syntactic equational theory of
Section~\ref{sec:syntax}. This uses the derivation of 
categorical structures for the semantics as \emph{lax homomorphisms} between
structure on the coeffect category $\Icat$ and the base $\mathbb{C}$.

\end{itemize}

%
% TP: Removed \paragraph just because I think this looks nicer with the list
%
\noindent 
In this section, 
$\mathbb{C}, \mathbb{D}, \mathbb{I}$ range over categories.  The
objects of a category $\mathbb{C}$ are written
$\emph{obj}(\mathbb{C})$.  The category of functors between
$\mathbb{C}$ and $\mathbb{D}$ is written $[\mathbb{C}, \mathbb{D}]$.
Exponential objects, representing function types in our model, 
are written in two ways, either $B^A$ or $A \Rightarrow B$.

\subsection{Interpreting contexts and judgments} 
\label{sem:interp}

The semantics is parameterised by a coeffect algebra, with scalar coeffects $(\C,
\cseq, \cpar, \cunit, \czero, \cleq)$, coeffect shape $(\SH, \mksh{-},
\stimes, \sempty, \sunit)$, and $(\amerge, \asplit, \azero)$. An
interpretation $\interp{-}$ is given to types, free-variable contexts,
and type and coeffect judgments, with a base Cartesian-closed category
$\mathbb{C}$ for denotations and a category $\Icat$ of scalar coeffects, where $\emph{obj}(\Icat)
= \C$. Since $\mathbb{C}$ is Cartesian-closed, we use the
$\lambda$-calculus as the syntax for giving concrete definitions.

The interpretation $\interp{-}$ is parameterised by categorical 
structures which model a particular notion of context. %We show  
%instances of these structures for bounded reuse and dataflow later. 
The interpretation of free-variable contexts depends on shape, for
which we give concrete definitions for flat and structural shapes.

\paragraph{Interpreting judgments.}
type and coeffect judgments are interpreted (given denotations) 
as morphisms in $\mathbb{C}$, of the form:
%
\begin{align*}
\interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \D^{\mksh{\Gamma}}_{\aclrd{R}}
\interp{\Gamma} \rightarrow \interp{\tau}
\end{align*}
%%
The interpretation is a morphism from an interpretation of the context $\Gamma$
to the interpretation of the result. The functor $\D^{\mksh{\Gamma}}_{\aclrd{R}}$ over the
context encodes the semantic notion of context and is indexed by the free-variable 
context shape $\mksh{\Gamma}$ and coeffect annotation $\aclrd{R}$.

The structure $\D$ can be thought of as a coproduct of functors $\D^{\sclrd{n}}$ for every
possible shape $\sclrd{n} \in \SH$ of free-variable context:
%%
\begin{equation*}
\D : \Sigma_{\sclrd{n} : \SH}. \D^{\sclrd{n}}
\quad \text{where} \quad
\D^{\sclrd{n}} : \Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}]
\end{equation*}
% %
For a fixed context shape $\sclrd{n}$ the functor $\D^{\sclrd{n}} :
\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}]$ 
maps an $\sclrd{n}$-indexed coeffect (think positions) to a functor from a
context $\mathbb{C}^n$ to an object in $\mathbb{C}$. That is, given a coeffect
annotation (matching the shape of the context), we get a 
functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$. 
%

From a programming 
perspective, this functor defines a data structure that models the additional
context provided to the program. The shape of this data structure depends on the coeffect
annotation $\Icat^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
the number of values needed for each variable and the functor will be 
formed by lists of length matching the required number.

%
% TP: I think this is easier to understand in the later paragraph
%
% Thus free-variable contexts are interpreted
% $\interp{\Gamma} \in \mathbb{C}^{\mksh{\Gamma}}$ and

%
% TP: I did not get the following - but I think I tried to say the same thing above...?
%
% $\D$ couples the
% shape of the coeffect with the shape of the free-variable context.
%Below, the interpretation of free-variable contexts is defined for
%classes of flat and structural system.

\paragraph{Types.} 
Types are interpreted as objects of $\mathbb{C}$, that is 
$\interp{\tau} : \emph{obj}(\mathbb{C})$
 where function types have the interpretation as exponents: 
%
\begin{align*}
  \interp{\sigma \xrightarrow{\cclrd{r}} \tau} = \D^{\sunit}_{\cclrd{r}} \interp{\sigma}
  \Rightarrow \interp{\tau}
\end{align*}
%%
The parameter of a function is wrapped by 
a functor $\D^{\sunit}_{\cclrd{r}}$ that defines a context with singleton shape
$\sunit$, matching the single value that it contains.
This interpretation is shared by all coeffect calculi.

%
% TP [Deleted] "although it may be extended to other type constructors."
% (I did not understand what you mean here)
%

\paragraph{Free-variable contexts.} 
As described above, free-variable contexts $\Gamma$ are given an
interpretation as objects in $\mathbb{C}^{\mksh{\Gamma}}$. Thus,
the interpretation of contexts is shape dependent.

%\tpnote{This is tricky but clever! Tried to add explanation}
We define $\interp{-}$ on free-variable context for structural and flat
systems. For flat systems, there is only a single shape, so the interpretation
is a product type inside the Cartesian closed category $\mathbb{C}$. For 
structural systems, the shape matches the number of variables and so the model 
is a value in the product category $\mathbb{C} \times \ldots \times \mathbb{C}$.

\paragraph{\textnormal{\emph{Flat coeffects.}}} 
Recall that $\SH = \{\sflat\}$ and $\mksh{\Gamma} = \sflat$.
Since the set of positions $\sflat$ is a singleton, then $\mathbb{C}^{\sflat}$
is isomorphic to $\mathbb{C}$.
%
% TP: I see why the following bit is important, but I think we can assume people
% will just believe us here and make this a bit simpler
%
% as categories formed from a singleton set are terminal objects in \textbf{Cat}. 
%
Therefore $\interp{\Gamma} : \emph{obj}(\mathbb{C})$, which is defined as:
%%
% \begin{align*}
% \interp{x : \tau, \Gamma'} = \interp{\tau} \times \interp{\Gamma'}
% \qquad\qquad
% \interp{\cdot} = 1 
% \end{align*}
%
% TP: Is there anything wrong with the following? I find it a bit simpler...
%
\begin{align*}
\interp{x_1 : \tau_1, \ldots, x_n : \tau_n} = \interp{\tau_1} \times \ldots \times \interp{\tau_n}
\end{align*}
%
Denotations of typing judgments in a flat coeffect system are thus of the form 
(where $\aclrd{R} \in \Icat$): 
%%
$$\interp{\coctx{x_1 : \tau_1, .., 
x_n : \tau_n}{\aclrd{R}} \vdash e : \tau} : \D^{\sflat}_{\aclrd{R}}(\interp{\tau_1} \times ... \times \interp{\tau_n})
\rightarrow \interp{\tau}
$$

\paragraph{\textnormal{\emph{Structural coeffects.}}}
Recall that $\SH = \mathbb{N}$ and $\mksh{\Gamma} = |\Gamma|$
(number of free variables), thus 
$\interp{\Gamma} : \emph{obj}(\mathbb{C}^{|\Gamma|})$. This is defined similarly 
to the above, but instead of using products in $\mathbb{C}$, %and the terminal object
%$1 : \mathbb{C}$, 
we use the product of categories. % (\ie{}, $\times$ 
% in \textbf{Cat}) with the singleton category $1 \in \mathbf{Cat}$. 
%
% TP: If you think that we should say how empty products are treated
% then I would just say "Empty contexts are modelled using the singleton category"
% (those who would ask the question will understand this and it does not add
% too much fancy notation to the explanation)
%
Thus, denotations have the form: 
$$
\interp{\coctx{x_1 : \tau_1, \ldots, x_n : \tau_n}{\aclrd{R}} \vdash e : \tau} : \D^n_{\aclrd{R}}
(\interp{\tau_1}, \ldots, \interp{\tau_n}) \rightarrow \interp{\tau}
$$
where $|R| = n$ and we use commas (instead of $\times$) to denote the product
of categories. %in $\mathbf{Cat}$. 
This means that $\D^n : \Icat^n \rightarrow [\mathbb{C}^n,
\mathbb{C}]$ is a functor between an $n$-length vector of coeffects
indices and an \emph{n}-ary endofunctor.
As mentioned, the key
difference between the flat and structural interpretations of
free-variable contexts is that flat uses products of objects in
$\mathbb{C}$ and the structural uses products of $\mathbb{C}$ in 
the category of categories. %\textbf{Cat}.

\newcommand{\BLL}{\mathsf{B}}

\begin{example}[Bounded reuse] Recall
bounded reuse has coeffect scalars $\C = \mathbb{N}$ and shapes $\SH = \mathbb{N}$.
We model contexts by replicating the value of each variable so 
there is a value for each use. This matches the model used by 
Girard \emph{et al.}~\cite{girard1992bounded}. Contexts are described by 
 $\BLL : \Sigma_{\sclrd{n} : \mathbb{N}}. (\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$, 
where for $\aclrd{R} = \langle{\cclrd{r_1}, \ldots, \cclrd{r_n}}\rangle$: 
%%
%\tpnote{\\I do not understand what the $a_i$ functions mean (in the morphism mapping part)?}
\begin{align*}
\BLL^n_{\aclrd{R}} (A_1, \ldots, A_n) & = A_1^{\cclrd{r_1}} \times \ldots \times A_n^{\cclrd{r_n}} \\
\BLL^n_{\aclrd{R}} (f_1, \;\ldots\;, f_n) & = \lambda \langle{a_1, \ldots, a_n}\rangle . \langle{(f_1 \circ a_1), \ldots, (f_n \circ a_n)}\rangle
\end{align*}
%
Thus each object in the free-variable context $A_i$ is indexed by its associated coeffect $\cclrd{r_i}$.
For the morphism mapping, $f_i : A_i \rightarrow B_i$ and $a_i : {A_i}^{\cclrd{r_i}}$ thus $(f_i \circ a_i) : {B_i}^{\cclrd{r_i}}$. 
The exponent ${A_i}^{\cclrd{r_i}}$ can be read as a product of $\cclrd{r_i}$ copies of $A_i$, \eg:
$$\BLL^3_{1,0,2} (A, B, C) = A^1 \times B^0 \times C^2 = (A) \times 1 \times (C \times C)$$
\label{exm:bll1}
\end{example}

\vspace{-2em}
\newcommand{\Imp}{\mathsf{I}}
\begin{example}[Implicit paramters]
Recall the implicit parameter calculus with scalar coeffects as 
sets of names paired with types $\C = \mathcal{P}(\ident{Name} \times \ident{Types})$ 
and flat shape with singleton $\SH = {\{\sflat\}}$.

%\tpnote{Could you say, let $\mathsf{Impl}=\textbf{Set}$ be a category of pairs of vars and values and use $\mathsf{Impl}$ throughout?\\ I find Set non-intuitive }
Its contexts are defined by 
$\Imp^\star :  \Sigma_{\sclrd{n} : \{\star\}}, (\Icat^{\sclrd{n}} \rightarrow [\mathbf{Set}^{\sclrd{n}}, \mathbf{Set}])$,
which is equivalent to $\Icat \rightarrow [\mathbf{Set}, \mathbf{Set}]$) and defined as follows:
%%
\begin{align*}
\Imp^{\sflat}_{\aclrd{R}} A = A \times \interp{\aclrd{R}} \qquad\qquad
\Imp^{\sflat}_{\aclrd{R}} f = \lambda (a, r) . (f\ a, r)
\end{align*}
%%
The interpretation $\interp{\aclrd{R}}$ maps a set of variable-type
pairs to an object representing a set of variables-values pairs in $\mathbf{Set}$.
\label{exm:impl1}
\end{example}

\subsection{Sequential composition}
\label{sem:seq}

Following the usual categorical semantics approach, we require a
notion of sequential composition for our denotations.  We show first a
special case for $\D^\sunit$, where $\Icat^\sunit = \Icat$
and $\mathbb{C}^\sunit = \mathbb{C}$ in both flat and structural
systems\footnote{since $\sunit = 1$ in structural and $\sunit \cong
  1$ in flat, \ie{}, $\star$ is isomorphic to 1} and thus $\D^\sunit :
\Icat \rightarrow [\mathbb{C}, \mathbb{C}]$. Composition of  
morphisms $f : \D^\sunit_S A \rightarrow B$ and $g : \D^\sunit_R B
\rightarrow C$ is defined by an \emph{indexed comonad}  
(which we previously introduced briefly~\cite{petricek2013coeffects}).

\begin{definition}
An \emph{indexed comonad} comprises a strict monoidal category
$(\mathbb{I}, \bullet, I)$ and a functor $\Di : \mathbb{I}
\rightarrow [\mathbb{C}, \mathbb{C}]$ with two natural transformations
(where we write $(\Di\,R)\,A$ as $\Di_R A$):
%\tpnote{Added brackets in $(\Di\,R)\,A$ -- otherwise I was confused}
%%
\begin{align*}
(\delta_{X,Y})_A : \Di_{(X \bullet Y)} A \rightarrow \Di_X (\Di_Y A) \qquad
(\varepsilon_I)_A : \Di_I A \rightarrow A
%\delta_{X,Y} : \Di (X \bullet Y) \xrightarrow{.} \Di X \circ \Di Y \qquad
%\varepsilon_I : \Di I \xrightarrow{.} 1_{\mathbb{C}}
\end{align*}
%%
where $\delta$ is called \emph{comultiplication} and $\varepsilon$ is called \emph{counit}.
We require indexed 
analogues of the usual comonad axioms (\emph{cf.}~\cite{uustalu2008notions}): 
%$$
%\Di_R \varepsilon_I \circ \delta_{R,I} = {\varepsilon_I}_{\Di_R} \circ \delta_{I, R} = id \quad 
%\delta_{R,S}\Di_T \delta_{R \bullet{} S, T} = \Di_R \delta_{S,T} \circ \delta_{R, S \bullet{} T}
%$$ 
%
%\tpnote{I'd consider omitting the axioms (too much theory?) if we can refer to your thesis?}
%\tpnote{[Read 2] OK - I commented them out - but feel free to revert!}
\begin{equation*}
\hspace{-1em}
\xymatrix@R=1em{
\Di_R \ar[r]^-{\delta_{R, I}} \ar@{=}[dr] \ar[d]_{\delta_{I,R}} \ar@{}[dr]^{\hspace{0.3em}\text{[C2]}} & \Di_R \Di_I \ar[d]^{\Di_R \varepsilon_I} \\
\ar@{}[ur]_{\hspace{-1.8em}\text{[C1]}} \Di_I \Di_R \ar[r]_{\varepsilon_I \Di_R} & \Di_R
}
\hspace{0.6em}
\xymatrix@R=1em{
\ar@{}[dr]|{\text{[C3]}}
\Di_{R \bullet{} S \bullet{} T} \ar[d]_{\delta_{R, S \bullet{} T}} \ar[r]^{\delta_{R \bullet{} S, T}} & \Di_{R \bullet{} S}  \Di_T \ar[d]^{\delta_{R,S}\Di_T} \\
\Di_R \Di_{S \bullet{} T} \ar[r]_{\Di_R \delta_{S,T}} & \Di_R \Di_S \Di_T
}
\label{eq:indexed-comonads-axioms}
\end{equation*}
\end{definition}

\noindent
An indexed comonad $\Di : \mathbb{I} \rightarrow
[\mathbb{C}, \mathbb{C}]$ induces a notion of composition
for all $f : \Di_S A \rightarrow B, g : \Di_R B \rightarrow C$: 
%
\begin{equation*}
 g \hat{\circ} f = g \circ \Di_R f \circ \delta_{R,S} \, : \, \Di_{R \bullet S} A \rightarrow B
\end{equation*}
%
with the identity $\hat{\id}_A = (\varepsilon_I)_A : \Di_I A \rightarrow A$ for all $A$. Thus
in\-dexed comonads induce a category which has the same objects as $\mathbb{C}$ and morphisms 
  $\mathbb{C}_\Di(A, B) = \bigcup_{R \in \mathbb{I}} \mathbb{C}(\Di_R A, B)$.
Note that an indexed comonad is not a family of (ordinary) comonads, because the identity
is only defined for the functor $\Di_I$. %and the functor in the result of composition (and comultiplication) 
%may be any $\Di_R$.

Therefore, if $\D^\sunit$ is an indexed comonad, there is a notion of composition for
denotations with a single coeffect index. % in our semantics.

\begin{example}[Bounded reuse]
$\BLL^\sunit_R$ (Example~\ref{exm:bll1}) has an indexed comonad structure, where
the monoid $(\cclrd{\mathbb{N}}, \cclrd{\ast}, \cclrd{1})$ from the coeffect scalar for bounded reuse
induces a monoidal category structure on $\Icat$ (with $1 : \Icat$ and the bifunctor
$\ast : \Icat \times \Icat \rightarrow \Icat$), with operations:
%
%\begin{align*}
%\begin{array}{l}
%\delta^\sunit_{R,S} = \lambda \langle{a_1 ..., a_{R S}}\rangle . \\
%\hspace{3.5em} \langle{~\langle{a_1 ..., a_S}\rangle, \langle{a_{S+1},..., a_{S+S}}\rangle}, \ldots \\
%\hspace{4.5em} \;\;~\langle{a_{(R - 1)S + 1},..., a_{R S}}\rangle~~\rangle 
%\end{array}
%\quad\quad
%\begin{array}{l}
%\varepsilon^\sunit_{1} = \lambda \langle{a_1}\rangle . a_1  \\[0.5em] 
%\\
%\end{array}
%\end{align*}
\begin{align*}
\begin{array}{rl}
\varepsilon^\sunit_{1} & \hspace{-1em} = \lambda \langle{a_1}\rangle . a_1 \\
\hspace{-0.5em}
\delta^\sunit_{R,S} & \hspace{-1em} = \lambda \langle{a_1 ..., a_{R S}}\rangle . \\
& \hspace{0.4em} \langle{\langle{a_1 ..., a_S}\rangle, \langle{a_{S+1},..., a_{S+S}}\rangle}, ..., \langle{a_{(R - 1)S + 1},..., a_{R S}}\rangle\rangle
\end{array}
\end{align*}
%\tpnote{Does this explanation for couint / comult. make sense?}
Indexed comonads essentially model single-variable contexts.
%Recall that we only consider single-variable context.
The counit requires a single copy of the value from the context. 
The comultiplication splits $R$ times $S$ copies of a value into
$R$ copies of a context where each context contains just $S$ copies of the value.
\label{exm:bll2}
\end{example}

\begin{remark}
A semantics for dataflow coeffects is similar to bounded reuse with
$\D^n_R (A_1, \ldots, A_n) = (A_1 \times A_1^{R_1}) \times ... \times (A_n \times {A_n}^{R_n})$, \ie{},
each free-variable has an extra value representing the ``current'' value. A dataflow indexed
comonad is similar to the above but with additive rather than multiplicative behaviour. 
\end{remark}

\begin{example}[Implicit parameters]
For the coeffect scalar monoid $\cclrd{(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \emptyset{})}$ 
of implicit parameters, 
$\Imp^\ast$ (Example~\ref{exm:impl1}) there is an indexed comonad structure, with operations: 
%where induces a monoidal category structure on $\Icat$, with operations:
%For a set $X$, with the monoidal category of $X$-subsets 
%$(\mathcal{P}(X), \cup, \emptyset)$, 
%then $\D : \mathcal{P}(X) \rightarrow [\mathbf{Set}, \mathbf{Set}]$ where
% $\D_X A = A \times X$ and $\D_X f = f \times \id_X$ 
%is an indexed comonad with:
%
\begin{align*}
\varepsilon_{\emptyset} \, (a, \emptyset{}) \mapsto a 
\qquad\qquad\quad~~~
\delta_{R,S} (a, \gamma) \mapsto ((a, {\gamma|}_S ), {\gamma|}_R) %\\
%\qquad \textbf{where}~~~{\gamma|}_R = \{(x, v) \mid (x, v) \in \gamma, (x, t) \in R \}
%\delta_{R,S} (a, x) \mapsto ((a,  x-(R-S)), x-(S-R))
\end{align*}
%%`
where ${\gamma|}_R = \{(x, v) \mid (x, v) \in \gamma, (x, t) \in R \}$ filters incoming implicit parameters to those
variable-value pairs where the variable is in the coeffect $R$.
%%
%where $x : R \cup S$ and thus the $R$-only subset of $x$ is $x - (S - R)$
%and the $S$-only subset of $x$ is $x - (R - S)$ (where ($-$) is set difference).
\label{exm:impl2}
\end{example}

\indent
These two examples (which are new here) provide composition for context-dependent
computations indexed by coeffects in a flat calculus. For structural coeffects,
we need to compose morphisms which have more than a single coeffect annotation. For this, we
introduced the notion of \emph{structural indexed comonads}.

\begin{definition}
\newcommand{\sunitc}{\sunit} %{{\sclrd{1}}}
%
A \emph{structural indexed comonad} comprises a functor 
$\D : \Sigma_{n : \SH}. (\mathbb{I}^n \rightarrow [\mathbb{C}^n, \mathbb{C}])$
where $(\mathbb{I}, \bullet, I)$ is a strict monoidal category,  
$\sunit \in \SH$ and 
$\D^{\sunit} : \mathbb{I}^\sunit \rightarrow [\mathbb{C}^\sunit, \mathbb{C}]$
 is an indexed comonad (with 
$(\delta^\sunit_{X,Y})_A : \D^\sunit_{X \bullet Y} A \rightarrow \D^\sunit_{X} \D^\sunit_{Y} A$ 
and 
$(\varepsilon_I)_A : \D^\sunit_I A \rightarrow A$) and 
a \emph{structural comultiplication} natural transformation:
$$
(\delta^{\sclrd{n}}_{r, S})_{A^n} : \D^{\sclrd{n}}_{r \bar{\bullet} S} A^n \rightarrow \D^\sunit_r \D^{\sclrd{n}}_S A^n
$$ 
%$$
%(\delta^{\sclrd{n}}_{R, S})_{A_\sunitc, \ldots, A_n} : \D^{\sclrd{n}}_{R \bar{\bullet} S} (A_1, \ldots, A_n) \rightarrow %\D^\sunitc_R \D^{\sclrd{n}}_S (A_1, \ldots, A_n)
%$$ 
where $A^n \in \mathbb{C}^n$, $r \in \Icat$, $S \in \Icat^{\sclrd{n}}$ and
 $\bbullet : \mathbb{I} \times \mathbb{I}^{\sclrd{n}} \rightarrow \mathbb{I}^{\sclrd{n}}$ is 
 the monoid left action for $\bullet$ lifting scalar coeffects to shaped coeffects (\eg{}, 
the scalar-vector version of $\bullet$).
Analogous laws to monoid left actions for unitality and associativity hold for structural comultiplication:
$$
\varepsilon_I \circ \delta^{\sclrd{n}}_{I, r} = id 
\,\qquad
\D^\sunitc_r \delta^{\sclrd{n}}_{S,T} \circ \delta^{\sclrd{n}}_{r, S \bar{\bullet{}} T}
= 
\delta^\sunitc_{r,S}\D^{\sclrd{n}}_T \circ \delta^{\sclrd{n}}_{r \bar{\bullet{}} S, T}
$$
%Structural comultiplication satisfies 
%$\varepsilon_I \circ \delta^{\sclrd{n}}_{I, R} = id$ 
%, but
%with appropriate indices.
%
%\tpnote{Totally \emph{not} sure if this is right -- but I \emph{believe} we could keep law details for MFPS}
%
%
%%
%\begin{align}
%\hspace{-1em}
%\xymatrix@R=1.4em{
%\D^{\sclrd{n}}_{I \bar{\bullet} R} \ar[r]^-{\delta^{\sclrd{n}}_{I,R}} \ar@{=}[dr] \ar@{}[dr]^{\hspace{0.3em}\text{[SC1]}} & \D^\sunitc_I \D^{\sclrd{n}}_R \ar[d]^{\varepsilon_I} \\
% & \D_R
%}
%\;
%\xymatrix@R=1em{
%\ar@{}[dr]|{\text{[SC2]}}
%\D^{\sclrd{n}}_{R \bar{\bullet{}} S \bar{\bullet{}} T} \ar[d]_{\delta^{\sclrd{n}}_{R, S \bar{\bullet{}} T}} \ar[r]^{\delta^{\sclrd{n}}_{R \bar{\bullet{}} S, T}} %& \D^\sunitc_{R \bullet{} S}  \D^{\sclrd{n}}_T \ar[d]^{\delta^\sunitc_{R,S}\D^{\sclrd{n}}_T} \\
%\D^\sunitc_R \D^{\sclrd{n}}_{S \bar{\bullet{}} T} \ar[r]_{\D^\sunitc_R \delta^{\sclrd{n}}_{S,T}} & \D^\sunitc_R \D^\sunitc_S \D^{\sclrd{n}}_T
%}
%\vspace{-1em}
%\label{eq:structural-indexed-comonad-axioms}
%\end{align}
%for unit and associativity respectively 
Note the indexed comonad comultiplication $\delta^\sunitc$ for associativity. 
\label{def:structural-indexed-comonad}
\end{definition}

\noindent
Structural indexed comonads provide composition for morphisms
%$f : \D^{\sclrd{n}}_{S} (A_1 \times \ldots \times A_n) \rightarrow B$ and $g : \D^\sunit_R B \rightarrow C$:
$f : \D^{\sclrd{n}}_{S} A^n \rightarrow B$ and singleton-shaped $g : \D^\sunit_r B \rightarrow C$:
%%
$$
g \hat{\circ} f = g \circ \D^{\sunit}_r f \circ \delta^{\sclrd{n}}_{r,S} 
: \D^{\sclrd{n}}_{r \bbullet S} A^n \rightarrow C
$$
%$$
%g \hat{\circ} f = g \circ \D^{\sunit}_R f \circ \delta^{\sclrd{n}}_{R,S} 
%: \D^{\sclrd{n}}_{R \bbullet S} (A_1 \times \ldots \times A_n)  \rightarrow C
%$$
%%
Note that this composition is asymmetric: the left morphism and right morphisms 
have different shapes. To compose morphisms which both have
non-trivial context shapes requires additional structure for manipulating contexts
(shown in the next section).

\begin{example}[Bounded reuse]
$\BLL :\Sigma_{\sclrd{n} : \mathbb{N}}, (\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$ has a structural indexed comonad structure with the indexed comonad $\BLL^\sunit$ (Example~\ref{exm:bll2}) and the following structural comultiplication: 
%%
%\tpnote{Is there a mismatch between $,$ and $\times$?}
\arraycolsep=1pt\def\arraystretch{0.4}
\begin{align*}
%\arraycolsep=3pt\def\arraystretch{0.4}
& \delta^n_{r,S} = \lambda (\langle{a^1_1, \ldots, a^1_{r \ast S_1}}\rangle, \ldots, \langle{a^n_1, \ldots a^n_{r \ast S_n}}\rangle) . \\[-0.3em]
& \begin{array}{lllll}
    ( &  ( &   \langle{a^1_1, \ldots, a^1_{S_1}}\rangle, & ..., & \langle{a^n_1,\ldots, a^n_{S_n}}\rangle ), \\[0.3em]
            &  ( &   \langle{a^1_{(S_1 + 1)}, ..., a^1_{(S_1+1) + S_1}}\rangle, & ..., & \langle{a^n_{(S_n + 1)}, \ldots, a^n_{(S_n+1) + S_n}}\rangle ), \\[0.6em]
            &  & \ldots  \\[0.2em]
            &  ( &   \langle{a^1_{(r - 1) \ast s_1 + 1}, .., a^1_{r \ast S_1}}\rangle, & ..., & \langle{a^n_{(r - 1) \ast S_n + 1}, ..., a^n_{r \ast S_n}}\rangle ) ) 
   \end{array}
\end{align*}
%
The input is an $n$-variable context 
containing $r$ times $S_i$ copies of $a^i$ for each variable. The output
has $r$ copies of a single $n$-variable context containing $S_i$ copies of $a^i$ for each variable. Thus,
$\delta^n_{r,S}$ partitions the incoming context into $r$-sized contexts. 
%into a single-variable context, containing $R$ copies of the input vector. In each copy, there are $S$ copies of each
%variable.
\label{exm:bll3}
%\tpnote{Please check - does this make sense?}
\end{example}

\noindent
Note that in the case of the flat system, a structural indexed co\-mo\-nad collapses
to a regular indexed comonad on $\D^{\sunit}$.

\subsection{Splitting and merging contexts}
\label{sem:monoidal}

Indexed comonads and structural indexed comonads give a semantics for
sequential composition of contextual computations. However, this does
not provide enough structure for a full semantics of the coeffect
calculus. Core to the semantics of abstraction and application is the
merging and splitting of contexts. Recall the free-variable contexts 
and coeffects in the (abs) and (app) rules:
%%
\begin{align*}
{\text{\scriptsize{(app)}}}
\dfrac
      {\coctx{\Gamma_1}{R} \vdash e_1 ... \;
       \coctx{\Gamma_2}{S} \vdash e_2 ...}
      {\coctx{\Gamma_1, \Gamma_2}{ R \asplit (T \aseq S)} 
\vdash e_1 \, e_2 ...}
\;
& 
\;
{\text{\scriptsize{(abs)}}}
\dfrac
  {\coctx{\Gamma, v : \sigma}{R \amerge \alift{S}} \vdash e ...}
  {\coctx{\Gamma}{R} \vdash \lambda v . e ... : \sigma \xrightarrow{S}\ldots}
\end{align*}
%
Reading (app) bottom-up, the context of the application
is split into two contexts for each subterm $e_1$ and $e_2$. 
Reading (abs) bottom-up, the context of the abstraction
is merged with the singleton context of the parameter. 
Capturing these notions in the denotational semantics requires
some additional structure.

% \tpnote{I removed some details about $\sqcup$ and $\sqcap$ to hide old notation}
A (non-indexed) comonadic semantics for the $\lambda$-calculus
requires a \emph{monoidal comonad} with operation $\m_{A,B} : \F A
\times \F B \rightarrow \F (A \times B)$~\cite{uustalu2008notions}.
Previously, we defined a similar operation for the semantics of a flat
coeffect system, with an indexed monoidal operation $\m^{R,S}_{A,B}$ for
merging contexts. Dually, contexts were split with 
$\n^{R,S}_{A,B}$~\cite{petricek2013coeffects}. We used two operations for combining and
splitting the coeffect annotations, respectively. Here we generalize these 
as $\amerge$ and $\asplit$ using a shape-indexed version.

\begin{definition}
$\D : \Sigma_{\sclrd{n} :
    \SH}, (\aclrd{\mathbb{I}}^{\sclrd{n}} \rightarrow
  [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$ is a 
   \emph{indexed lax (semi)monoidal functor} and/or \emph{colax (semi)monoidal functor} if it has the following
 natural transformations respectively: 
%%
\begin{align*}
\mathsf{m}^{n,m}_{R, S} & : \D^n_{R} A \times \D^m_{S} B \rightarrow \D^{n \stimes m}_{R \amerge S} (A \times B) 
\\
\mathsf{n}^{n,m}_{R, S} & : \D^{n \stimes m}_{R \asplit S} (A \times B)  
\rightarrow \D^n_{R} A \times \D^m_{S} B 
\end{align*}
%
In both, shape descriptions are combined by $\stimes$. The first operation models 
context merging and combines coeffects using $\amerge$. The second models
context splitting, with $\asplit$ for the pre-split coeffect. 
\end{definition}

\begin{example}[Bounded reuse]
For bounded reuse, $\BLL$ is an indexed lax and colax semimonoidal functor
with the following operations:
\begin{align*}
\m^{n,m}_{R,S} & = \lambda (\langle{a_1, ..., a_n}\rangle \times \langle{b_1, ..., b_m}\rangle) . (\langle{a_1, ..., a_n}\rangle, \langle{b_1, ..., b_m}\rangle)
 \\
\n^{n,m}_{R,S} & = \lambda (\langle{a_1, ..., a_n}\rangle, \langle{b_1, ..., b_m}\rangle) . (\langle{a_1, ..., a_n}\rangle \times \langle{b_1, ..., b_m}\rangle)
\end{align*}
%%%
\ie{}, %\tpnote{You are not using the tuple notation $(x,y)$ -- but you say so?}
 $\m^{m,m}_{R,S}$ takes a pair of contexts and merges them simply by replacing the 
product in $\mathbb{C}$ which pairs the two arguments 
(written using $\times$) with products inside of $\BLL$ (written using tuple notation $(x, y)$). 
The operation $\n^{n,m}_{R,S}$ is the inverse.
\end{example}

\begin{example}[Implicit parameters]
For implicit parameters, $\Imp^\star$ is an indexed lax and colax semimonoidal
functor with operations:
%%
\begin{align*}
\m^{\star,\star}_{R,S} & = \lambda ((a, \gamma_R), (b, \gamma_S)) . ((a, b), \gamma_R \cup \gamma_S) \\
\n^{\star,\star}_{R,S} & = \lambda ((a, b), \gamma) . ((a, {\gamma|}_R), (b, {\gamma|}_S))
\end{align*}
%% 
As in Example~\ref{exm:impl2}, ${\gamma|}_R$ and ${\gamma|}_S$ restrict the set of implicit parameters $\gamma$ to 
the variable-values pairs for variables in $R$ and $S$. 
\label{exm:impl3}
\end{example}

\subsection{Putting it together}
\label{sem:together}
The semantics of the structural coeffect calculus $\interp{-}$ is defined in Figure~\ref{fig:semantics},
using the structures described in the previous sections.

\paragraph{Core rules.}
The denotation in (\emph{var}) maps a context of the singleton shape $\sunit$
containing just a single variable $\tau$ (with coeffect $I$) to a $\tau$ value 
using the counit operation.

The premise of (\emph{abs}) takes a context of shape $\sclrd{n} \,\stimes\, \sunit$ with
coeffects $\aclrd{R} \amerge \alift{\cclrd{s}}$ and a free-variables context consisting of $\Gamma$ and
an additional variable $x$. The denotation $g : \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \amerge 
  \alift{\cclrd{s}}} \interp{\Gamma, v : \sigma} \rightarrow \interp{\tau}$ is
pre-composed with $\m$, such that its context
is obtained by merging the declaration-site context ($\Gamma$) and call-site context ($\sigma$):
%%
$$
g \circ \m^{\sclrd{n},\sunit}_{\aclrd{R}, \alift{\cclrd{s}}} : 
(\D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \times \D^\sunit_{\alift{\cclrd{s}}} \interp{\sigma}) \rightarrow \interp{\tau}
$$
%%
This is uncurried to give a denotation from a context to an exponential object
representing the abstraction, where the singleton-shaped context becomes the source of the exponential.

% --------------------------------------------------------------------------------------------------
% Figure - semantics
% --------------------------------------------------------------------------------------------------

\newcommand{\exchgl}{R \amerge \langle{t}\rangle \amerge \langle{s}\rangle \amerge Q}
\newcommand{\exchgr}{R \amerge \langle{s}\rangle \amerge \langle{t}\rangle \amerge Q}
\newcommand{\contrl}{R \amerge \langle{s}\rangle \amerge \langle{t}\rangle \amerge Q}
\newcommand{\contrr}{R \amerge \langle{s \cpar t}\rangle \amerge Q}
\begin{figure*}[t]
\vspace{-0.5em}
\begin{align*}
\begin{array}{c}
%
% VAR semantics 
%
\tyrule{var}
  {\quad}
  {\interp{\coctx{x : \tau}{ \alift{\cunit} } \vdash x : \tau} = \varepsilon_I : \D^\sunit_I \interp{\tau} \rightarrow \interp{\tau}}
\qquad
%
% ABS semantics
%
\tyrule{abs}
  {\interp{\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \amerge \alift{\cclrd{s}}} \vdash e : \tau} =
              g : \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \amerge {\alift{\cclrd{s}}}} \interp{\Gamma, x \!:\! \sigma} \rightarrow \interp{\tau}}
  {\interp{\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
                = \Lambda (g \circ \m^{\sclrd{n},\sunit}_{\aclrd{R}, \alift{\cclrd{s}}})
: \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \rightarrow (\D^\sunit_{\alift{\cclrd{s}}} \interp{\sigma} \Rightarrow \interp{\tau})}
\\[1.8em]
%
% APP semantics
%
\tyrule{app}
   { \interp{\coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau} 
        = g_1 : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \rightarrow (\D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma} \Rightarrow \interp{\tau})
     \quad\qquad   
     \interp{\coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \tau} 
        = g_2 : \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} \rightarrow \interp{\sigma} }
   { \interp{\coctx{\Gamma_1, \Gamma_2}{\aclrd{R}\ \asplit\ (\cclrd{t}\ \aseq\ \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
        = \Lambda^{-1} g_1 \circ (\mathsf{id} \times (\D^{\sunit}_{\alift{\cclrd{t}}} g_2 \circ \delta^{\sclrd{m}}_{\alift{\cclrd{t}}, \aclrd{S}})) \circ \n^{\sclrd{n},\sclrd{m}}_{\aclrd{R}, \cclrd{t} \aseq \aclrd{S}}  
: \D^{\sclrd{n} \stimes \sclrd{m}}_{\aclrd{R} \asplit (\cclrd{t} \aseq \aclrd{S})} \interp{\Gamma_1, \Gamma_2} \rightarrow \interp{\tau} }
\\[2em]
\tyrule{ctx}
  { \interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} = f : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau} 
    \qquad \interp{\ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma'}{\aclrd{R'}} }{\emptyset}} = c : \D^{\sclrd{m}}_{\aclrd{R'}} \interp{\Gamma'}
\rightarrow \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma}
  }
  { \interp{\coctx{\Gamma'}{\aclrd{R'}} \vdash e : \tau} = f \circ c : \D^{\sclrd{m}}_{\aclrd{R'}} \interp{\Gamma'} \rightarrow \interp{\tau}}
\end{array}
\end{align*}
%
%
% -------------- WEAK --------------
%
\begin{equation*}
{\text{\footnotesize{(weak)}}}\quad
\; \interp{\ctxtrans{\coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x\!:\!\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}}
%
% Denotation
%
= \pi_1 \circ \n^{\sclrd{n},\sunit}_{\aclrd{R},\aweak}
: \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \asplit \aweak} (\interp{\Gamma} \times \interp{\sigma}) \rightarrow \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma}
\end{equation*}
%
%
% -------------- CONTR --------------
%
\vspace{-1em}
\begin{equation*}
{\text{\footnotesize{(contr)}}} \quad
\interp{
  \ctxtrans
    {\coctx{\Gamma_1, y\!:\!\tau, z\!:\!\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
    {\coctx{\Gamma_1, x\!:\!\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]} }
%
% Denotation
%
 = \m^{\sclrd{n},\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s} \cpar \cclrd{t}},\aclrd{Q}} \circ (\mathsf{id} \times \Delta_{\cclrd{s},\cclrd{t}} \times \mathsf{id}) \circ
\n^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s}},\alift{\cclrd{t}},\aclrd{Q}} \\[1.1em]
\end{equation*}
% 
%
% -------------- EXCH --------------
%
\vspace{-2em}
\begin{equation*}
{\text{\footnotesize{(exch)}}} \quad
\interp{
  \ctxtrans
     {\coctx{\Gamma_1, x\!:\!\tau, y\!:\!\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
     {\coctx{\Gamma_1, y\!:\!\sigma, x\!:\!\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }
       {\emptyset} }  
%
% Denotation
%
 = \m^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s}},\alift{\cclrd{t}},\aclrd{Q}} \circ (\mathsf{id} \times \mathsf{swap} \times \mathsf{id}) \circ \n^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{t}},\alift{\cclrd{s}},\aclrd{Q}} \\[0.7em]
\end{equation*}
%
% ------------------ Definitions ---------------------
%
\vspace{-1.5em}
\begin{equation*}
\textbf{where}~~ \mathsf{swap} : A \times B \rightarrow B \times A~~~\textbf{and}~~
  \Lambda, \Lambda^{-1} ~~\textnormal{denote currying and uncurrying respectively}
\end{equation*}
\vspace{-0.5em}
\caption{Denotational semantics for the coeffect calculus}
\label{fig:semantics}
\end{figure*}

% --------------------------------------------------------------------------------------------------

The application rule (\emph{app})  has two sub-expressions for the function and argument, with denotations
requiring two distinct contexts:
%%
\begin{align*}
g_1 : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \rightarrow (\D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma} \Rightarrow \interp{\tau})
\qquad
g_2 : \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} \rightarrow \interp{\sigma}
\end{align*}
%%
The target of $g_1$ is an exponential object with singleton shape for the parameter of type $\sigma$. 
To evaluate $g_1$ and $g_2$, the semantics of (\emph{app}) 
splits the incoming context over $\Gamma_1, \Gamma_2$ using $\n$:
%% 
\begin{align*}
\D^{\sclrd{n} \stimes \sclrd{m}}_{\aclrd{R}\,\asplit\,(\alift{\cclrd{t}}\,\bbullet\,\aclrd{S})}
(\interp{\Gamma_1} \times \interp{\Gamma_2})
\xrightarrow{\n^{\sclrd{n},\sclrd{m}}_{\aclrd{R}, \alift{\cclrd{t}}\bbullet\aclrd{S}}} 
\D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \times \D^{\sclrd{m}}_{\alift{\cclrd{t}} \bbullet \aclrd{S}} \interp{\Gamma_2}
\end{align*}
%%
Since $e_2$ computes the parameter for function $e_1$, the denotation
$g_2$ must be sequentially composed with the parameter part of
$g_1$. Thus, the structural indexed comonad is used with $g_2$
to compute the correct context for the parameter of $g_1$:
%
\begin{align*}
\D^{\sclrd{m}}_{\alift{\cclrd{t}} \bbullet \aclrd{S}} \interp{\Gamma_2} 
    \xrightarrow{\delta^{\sclrd{m}}_{\cclrd{t},\aclrd{S}}} 
 \D^\sunit_{\alift{\cclrd{t}}} \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} 
    \xrightarrow{\D^\sunit_{\alift{\cclrd{t}}} g_2} \D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma}
%\label{eq:app-morph2}
\end{align*}
%
This is composed with the previous equation by lifting
to the right-component of the product:
%
\begin{align*}
\D^{\sclrd{n}}_R \interp{\Gamma_1} \times \D^{\sclrd{m}}_S \interp{\Gamma_2}
\xrightarrow{id \times (\D^\sunit_{\alift{\cclrd{t}}} g_2 \, \circ \, \delta^{\sclrd{m}}_{t,S})}
\D^{\sclrd{n}}_R \interp{\Gamma_1} \times \D^{\sunit}_{\alift{\cclrd{t}}}  \interp{\sigma}
\end{align*}
%%
This equation computes the calling context and
parameter context for the function $e_1$, which is then composed with
the uncurried $g_1$ denotation as shown in the (\emph{app}) rule in Figure~\ref{fig:semantics}.

\paragraph{Structural rules.}

%The second part of Figure~\ref{fig:semantics} shows the semantics of the
%structural rules for the coeffect calculus. 
In Figure~\ref{fig:semantics}, (\emph{ctx}) composes the
denotation of an expression with a transformation $c$ providing the semantic 
 structural rules. 
The semantics of structural rules are defined by using
$\n^{n,m}_{R,S}$ to split contexts, transforming the components, and merging the transformed
contexts using $\m^{n,m}_{R,S}$. The (\emph{contr}) rule uses an additional operation
which duplicates a variable inside a context:
%%
\begin{align*}
\Delta_{\cclrd{r},\cclrd{s}} & : \D^{\sunit}_{\cclrd{r} \cpar \cclrd{s}} A
    \rightarrow \D^{\sunit \stimes \sunit}_{\alift{\cclrd{r}} \amerge \alift{\cclrd{s}}}(A \times A)
\end{align*}

% --------------------------------------------------------------------------------------------------

\begin{example}
We demonstrate the semantics with a concrete example 
for the bounded reuse calculus.
Consider the following:
%%
$$
\coctx{f:\mathbb{Z}\xrightarrow{2}\mathbb{Z}, x:\mathbb{Z}}{\langle 2, 4 \rangle } \vdash (\lambda z.z+z)~(f\ x)
$$
We elide the function body's denotation prior to contraction,
 $g = \interp{x : \mathbb{Z}, y : \mathbb{Z} \vdash (+x) y : \mathbb{Z}}$\footnote{
the full semantics has $\interp{+} : \D^0_{\langle{}\rangle} 1 \rightarrow (\D^1_1
\mathbb{Z} \Rightarrow (\D^1_1 \mathbb{Z} \Rightarrow \mathbb{Z}))$}. %  as primitive and uses double application $(+\ e_1) e_2$
The term therefore has denotation: 
%%
\begin{align}
\label{eq:exm-denot5}
& \interp{\coctx{}{\langle  \rangle} \vdash
\lambda z . (+ z) z : \mathbb{Z} \xrightarrow{2} \mathbb{Z}} = \Lambda (g \circ \Delta_{\langle{1}\rangle,\langle{1}\rangle} \circ \m^{0,1}_{\langle{}\rangle,\langle{1}\rangle}) \\
\notag & \interp{\coctx{f : \mathbb{Z} \xrightarrow{2} \mathbb{Z}, x : \mathbb{Z}}{\langle{1, 2}\rangle} \vdash 
f x : \mathbb{Z}} \\
%\label{eq:exm-denot6} 
\notag
& \hspace{1.5em} = \uncurry \varepsilon_1 \circ (id \times (\D \varepsilon_1 \circ \delta^1_{2, \langle{1}\rangle})) \circ \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle} \\ 
\label{eq:exm-denot6} 
& \hspace{1.5em} = \uncurry \varepsilon_1 \circ \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle} \\ 
\notag 
& \interp{\coctx{f : \mathbb{Z} \xrightarrow{2} \mathbb{Z},
x : \mathbb{Z}}{\langle{2, 4}\rangle}
\vdash 
(\lambda z . (+ z) z)\ (f x) : \mathbb{Z}} 
\\
\notag
& \hspace{1.5em} = \uncurry \eqref{eq:exm-denot5} \circ (id \times \D \eqref{eq:exm-denot6} 
\circ \delta^2_{2,\langle{2,1}\rangle}) 
\circ \n^{0,2}_{\langle{}\rangle,{\langle{4,2}\rangle}}
\\
\label{eq:exm-denot-final}
& \hspace{1.5em} = g \circ \Delta_{r,s} \circ \m^{0,1}_{\langle{}\rangle,\langle{1}\rangle} \circ (id \times \D \eqref{eq:exm-denot6} 
\circ \delta^2_{2,\langle{2,1}\rangle}) 
\circ \n^{0,2}_{\langle{}\rangle,{\langle{4,2}\rangle}}
\end{align}
%%
where \eqref{eq:exm-denot6} and \eqref{eq:exm-denot-final} are simplified. 
We ``run'' this semantics on some input, evaluating each step of the denotation as a function. 
%(\ie{}, we take $\mathbb{C} = \mathbf{Set}$). 
We write context objects, \eg{}, $\D^2_{\langle{R, S}\rangle} (A, B)$ as $\langle(a_1, ..., a_R), (b_1, ..., b_R)\rangle$ and products of
 contexts in $\mathbb{C}$, \eg{},  $\D^n_R A \times \D^m_S B$, as $(a \times b)$.

\vspace{0.2em}
\newcommand{\arw}[1]{\makebox[4.3em]{\scriptsize{${#1}$}}}
\hspace{-2.4em}
{\scalebox{0.9}{
\begin{minipage}{1.04\linewidth}
\begin{align*}
\begin{array}{c l l}
%%%% START
\hspace{-0.5em} & \langle(f_1, f_2),(x_1, x_2, x_3, x_4)\rangle  &\hspace{-5em} : \D^{2}_{\langle{2,4}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z}) \\[-0.5em]
%%%% STAGE 1
\xrightarrow{\arw{{n}^{0,2}_{\langle{}\rangle,\langle{2,4}\rangle}}} \hspace{-0.5em} &
 \langle\rangle \times \langle(f_1, f_2), (x_1, x_2, x_3, x_4)\rangle &  \hspace{-3em}: \D^0_{\langle{}\rangle} 1 \times \D^{2}_{\langle{2,4}\rangle} (\textit{as above}) \\[-0.3em]
%%
%\hspace{-0.5em} & 
%\multicolumn{2}{l}{
%\hspace{5.91em} : \D^0_{\langle{}\rangle} 1 \times \D^{2}_{\langle{2,4}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z}) 
%}
%\\
%%%% STAGE 2
\xrightarrow{\arw{\id \times \delta^2_{2, \langle{1, 2}\rangle}}} \hspace{-0.5em} & 
 \langle\rangle \times \langle\langle f_1, (x_1, x_2)\rangle, \langle f_2, (x_3, x_4)\rangle\rangle \\[0.33em]
%%
\hspace{-0.5em} & 
\multicolumn{2}{l}{
\hspace{5.91em}
: \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} \D^{2}_{\langle{2,1}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z})} \\[-0.3em]
%%%% STAGE 3
\xrightarrow{\arw{\id \times \D \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle}}} \hspace{-0.5em} &
 \langle\rangle \times \langle\langle f_1\rangle \times \langle x_1, x_2\rangle, \langle f_2\rangle \times \langle x_3, x_4\rangle\rangle \\[0.34em]
%%
\hspace{-0.5em} & 
\multicolumn{2}{l}{
\hspace{5.91em} 
: \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} (\D^{1}_{\langle{1}\rangle} (\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \D^{1}_{\langle{2}\rangle} \mathbb{Z})} \\[-0.1em]
%%%% STAGE 4
\xrightarrow{\arw{\id \times \D (\uncurry \varepsilon_1)}}  
\hspace{-0.5em} & \langle\rangle \times \langle f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em} : \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} \mathbb{Z} \\[-0.2em]
%%%% STAGE 5
\xrightarrow{\arw{\m^{0,1}_{\langle{}\rangle,\langle{1}\rangle}}} 
\hspace{-0.5em} & \langle f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em}  : \D^1_{\langle{2}\rangle} \mathbb{Z} \\
%%%% STAGE 6
\xrightarrow{\arw{\Delta^2_{\langle{1}\rangle,\langle{1}\rangle}}}
\hspace{-0.5em} & \langle(f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle)\rangle & \hspace{-3.5em} : \D^1_{\langle{1,1}\rangle} (\mathbb{Z} \times \mathbb{Z}) \\[0.54em]
%%%%% STAGE 7
\xrightarrow{\arw{g}} 
 \hspace{-0.5em} & \interp{+}\ \langle f_1 \langle x_1, x_2\rangle\rangle\ \langle f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em} : \mathbb{Z} 
\end{array}
\end{align*}
\end{minipage}
}}
\label{exm:run}
%This demonstrates the successive splitting, merging, and composing of contexts to the final result. 
\end{example}

\subsection{Soundness, with respect to equational theory}
\label{sec:semantics-soundness}

\noindent
Our denotational semantics for the coeffect
calculus is sound with respect the equational theory of Section~\ref{sec:syntax}. That is:
%%
\begin{theorem}[Soundness]
$$\coctx{\Gamma}{R} \vdash e \equiv e' : \tau \Rightarrow \interp{\coctx{\Gamma}{R} \vdash e : \tau} 
\equiv \interp{\coctx{\Gamma}{R} \vdash e' : \tau}$$
\end{theorem}
%%
\noindent
Proof of this follows from an interesting result which we 
first unpack: determining whether $\interp{\coctx{\Gamma}{R} \vdash
  e_1 : \tau} {\equiv} \interp{\coctx{\Gamma}{S} \vdash e_2 : \tau}$ follows
from a proof on coeffect annotations that $R = S$. 
%
\begin{lemma}
Every coeffect algebra axiom corresponds to an axiom 
of one of the categorical structures introduced here (indexed (structural) comonad or 
  indexed (co)lax monoidal functor). 
\end{lemma}
%
\noindent
For example, the monoid axiom $X \cseq \cunit =
X$ for scalar coeffects corresponds to indexed comonad axiom $\D^{\sunit}_X \varepsilon_{\cunit}
\circ \delta^{\sunit}_{X,{\cunit}} = id_{\D^{\sunit}_X}$ (which requires the monoid axiom to hold). 
This lemma follows from our derivation of 
the indexed categorical structures here. They are not derived \emph{ad hoc} but systematically 
as (lax) \emph{homomorphisms} (structure preserving maps) between the structure of 
coeffect annotations in $\Icat$ and the structure of denotations in $\mathbb{C}$.

\begin{proposition}
An indexed comonad on $\D$ witnesses that
$\D$ is a \emph{colax monoid homomorphism} between the (strict)
monoidal categories $(\mathbb{I}, \cseq, \cunit)$ and 
$([\mathbb{C}, \mathbb{C}], 
\circ, 1_{\mathbb{C}})$ (endofunctor composition). 
\end{proposition}
%%
Unpacking this, a \emph{monoid homomorphism} maps between
the underlying sets of two monoids, preserving the monoid structure of one into the other,
\ie{}, given monoids $(X, \bullet, I)$ and $(Y, \otimes, E)$ then
a monoid homomorphism is a mapping $F : X \rightarrow Y$ such that:
%%
\begin{align*}
F X \otimes F Y \equiv F (X \bullet Y) \quad\qquad E \equiv F I
\end{align*}
%%
The axioms of each monoid are preserved trivially by these equalities, 
\eg{}, $F X \equiv F (X \bullet I) \equiv F X \otimes F I \equiv F X \otimes E \equiv F X$. 
In a categorical setting, monoids are taken on categories with binary operations as 
bifunctors. A homomorphism is \emph{lax} if the above equalities are instead morphisms 
(which we say \emph{witness} the homomorphism) and \emph{colax} if these morphisms go in opposite direction.
Thus, a colax monoid homomorphism is witnessed by: 
%%
\begin{align*}
\delta : F X \otimes F Y \leftarrow F (X \bullet Y) \quad\qquad \varepsilon : E \leftarrow F I
\end{align*}
%%
Note our choice of morphism names.  
$F$ no longer preserves the monoid axioms \emph{up to equality} but has axioms on $\delta$ and $\varepsilon$, %% F X {=\joinrel=} 
\eg{}, $F (X \bullet I) \xrightarrow{\delta} F X \otimes F I \xrightarrow{id \otimes \varepsilon} F X \otimes E$ equals $F X \xrightarrow{id} F X$. 
%%  {=\joinrel=} F X

Our indexed comonad definition is equivalent to $\D$ being a colax homomorphism between a monoidal category $(X, \bullet, I)$ 
and the monoidal category of $\mathbb{C}$ endofunctors $(Y, \otimes, E) = ([\mathbb{C}, \mathbb{C}], \circ, 1_{\mathbb{C}})$, with endofunctor composition 
$\circ$ and the trivial endofunctor $1_{\mathbb{C}}$. The indexed comonads axioms are the axioms of the colax homomorphism.
Equivalently, $\D$ is a \emph{colax monoidal functor}.

A similar approach is taken to deriving the remaining structures.%, though we give less detail for brevity.

\begin{proposition}
  A structural indexed comonad provides $\delta^n_{R,S} :
  \D^\sunit_r \D^n_S A^n \leftarrow \D^n_{r \aseq S} A^n$ which witnesses that $\D$ is a 
  colax homomorphism between the following \emph{monoid left-actions} for
  $(\mathbb{I}, \cseq, \czero)$ and $([\mathbb{C}, \mathbb{C}], \circ,
  1_{\mathbb{C}})$: 
\begin{align*}
\begin{array}{rclcl}
(r : \mathbb{I}) & \narrow{\aseq} & (\langle{s_1, ..., s_n}\rangle : \mathbb{I}^n) &\narrow{=}& \langle{r \cseq s_1, ..., r \cseq s_n}\rangle : \mathbb{I}^n  \\
(\D^\sunit_r : [\mathbb{C}, \mathbb{C}]) & \narrow{\circ} & (\D^n_S : [\mathbb{C}^n, \mathbb{C}]) &\narrow{=}& \D^\sunit_r \D^n_S : [\mathbb{C}^n, \mathbb{C}]
\end{array}
%(x : \mathbb{I}) \aseq (\langle{y_1, \ldots, y_n}\rangle : \mathbb{I}^n) & = \langle{x \cseq y_1, \ldots, x \cseq y_n}\rangle  \\
%(\D^\sunit_r : \mathbb{C} \rightarrow \mathbb{C}) \circ (\D^n_S : \mathbb{C}^n \rightarrow \mathbb{C}) & = \D^\sunit_R \D^n_S : \mathbb{C}^n \rightarrow \mathbb{C}
\end{align*}
The axioms are the lax versions of the monoid left-action laws. 
%witnesses that $\D$
%is a colax homomorphism between the induced \emph{monoid left-action} for $(\mathbb{I}, \cseq, \czero)$
%of $n$-length coeffect vectors $(\mathbb{I}^n, \aseq_n : \mathbb{I} \times \mathbb{I}^n 
%\rightarrow \mathbb{I}^n)$ for monoid, and a trivial monoid left-action for $([\mathbb{C}, \mathbb{C}], \circ, 1_{\mathbb{C}})$ 
%defined by composition $([\mathbb{C}^n, \mathbb{C}], \circ : [\mathbb{C}, \mathbb{C}] \times [\mathbb{C}^n, \mathbb{C}] \rightarrow [\mathbb{C}^n, 
%\mathbb{C}])$, 
%\ie{},
\end{proposition}
%%
The lax and colax indexed monoidal operations $\m^{n,m}_{R,S}$ and $\n^{n,m}_{R,S}$ follow a similar derivation
but as lax and colax monoid homomorphism between composite monoids on coeffect annotations and shapes
and $\times$ in $\mathbb{C}$. The details are elided here.

Returning to soundness, our semantics is therefore defined in terms of structures whose axioms correspond 
to axioms of the syntactic equational theory. Consequently, semantic proofs correspond to 
syntactic proofs, modulo naturality laws and product/exponent laws in $\mathbb{C}$. 
This result holds in the general coeffect calculus and semantics since every semantic structure
has a unique corresponding structure on coeffect annotations (\ie{}, $(\C, \cseq, \cunit)$ for
sequential composition of unary denotations, $(\C, \asplit)$ for splitting contexts, 
$(\C, \amerge)$ for joining contexts).

\begin{example}
Section~\ref{prop:structural} showed $\eta$-equality for structural systems,
which uses the properties (1) $\asplit = \amerge = \times$ for structural systems and (2) 
$s \aseq \langle{\cunit}\rangle
= \langle{s \cseq \cunit}\rangle = \langle{s}\rangle$. The semantics here is sound
with respect to $\eta$-equality, the proof of which uses the corresponding axioms
(1) $\n^{n,m}_{R,S} \circ \m^{n,m}_{R,S} = id$ and (2) $\D^{\sunit}_s \varepsilon_{\cunit} \circ \delta^m_{s,\cunit}$
(structural indexed comonad unit law, Definition~\ref{def:structural-indexed-comonad}). 
% $\D^{n}_R A \times \D^{m}_S B \xrightarrow{\m^{n,m}_{R,S}} \D^{n,m}_{R \amerge S} (A \times B) \equiv
% \D^{n,m}_{R \asplit S} (A \times B) \xrightarrow{\n^{n,m}_{R,S}} \D^n_R A \times \D^m_S$ (\ie{}, $\n \circ \m = id$) and 
\end{example}
%
\noindent
The accompanying technical report shows the full semantic proofs for $\beta\eta$-equality whose structure corresponds
exactly to the syntactic proofs with the corresponding coherence conditions~\cite{appendix}.

%\begin{prosition}
%A lax indexed monoidal functor on $\D$ witnesses that $\D$
%is a lax monoid homomorphism between the $(\mathbb{I}, \amerge, \azero)$ and 
%a composite monoid of $
%\end{proposition}

%The indexed comonad structure is derived from a
%\emph{colax} \emph{homomorphism} between the monoidal structure
%$(\mathbb{I}, \bullet, I)$ on indices and monoidal structure of
%endofunctor composition. This approach is used to derive the rest
%of our semantics, though we keep the details of the derivation light
%here and present instead the resulting structures.

%\begin{remark}
%The category of $\mathbb{C}$-endofunctors $[\mathbb{C}, \mathbb{C}]$
%is a \emph{strict} monoidal category $([\mathbb{C}, \mathbb{C}], \circ,
%1_{\mathbb{C}})$ with composition $\times$ as the binary operation and the identity
%functor $1_{\mathbb{C}}$ as the unit. 
%\end{remark}

% ==================================================================================================

\section{Related work}
\label{sec:related-work}

We expand briefly on the overview of related work in 
Section~\ref{sec:intro-pathways}.

%\paragraph{Bounded reuse.}
The (\emph{storage}) rule for bounded linear logic explains the contextual requirements 
induced by proposition reuse~\cite{girard1992bounded}:
$$
\inference[\text{(storage)}]{!_{\overline{Y}} \Gamma \vdash A}{!_{X\overline{Y}} \Gamma \vdash !_X A} 
$$
where $X\overline{Y} = \langle{XY_1, .., XY_n}\rangle$ is the scalar multiple of a vector. 
This rule is akin to the $\delta^n$ operation of structural indexed comonads,
indeed, we can model it exactly using $\delta^{n}_{X,\bar{Y}}$ and the lifting $\D^n_X$.

In BLL, the modality $!_X$ is a constructor and may 
  appear both on the left- and right-hand sides of $\vdash$.
  In this paper, reuse bounds 
  annotate typing rules, thus there is no constructor
  corresponding to bounded reuse in the language; reuse bounds are
  meta-level. Our choice to work at the meta-level means that the 
  coeffect calculus provides a unified analysis and semantics to
  different notions of context and its term language is that of standard $\lambda$-calculus.

%\paragraph{Semantics.}
Previously we briefly introduced indexed comonads~\cite{petricek2013coeffects}
without derivation.  Here we derived indexed comonads as colax
homomorphisms. This is dual to the \emph{parametric effect monad}
structure defined as a lax
homomorphism~\cite{katsumata2014parametric}. Our semantics 
requires additional structure not needed for effects due to the
asymmetry inherent in the $\lambda$-calculus.

The necessity modality $\Box$ in S4 logic corresponds to a comonad
with lax monoidal functor structure $\m : \Box A \times \Box B \rightarrow \Box (A \times B)$. 
Bierman and de Paiva \cite{bierman2000intuitionistic} defined a term language corresponding to a natural deduction S4, where contexts contain sequences of $\Box$-wrapped assumptions
$x_1 \!:\! \Box A_1, \ldots x_n \!:\! \Box A_n$. Modelling these judgments does not require a context splitting operation $\n$ 
unlike in our approach. Our approach can be thought of as having a single $\Box$ modality over the context 
which allows both flat whole-context dependence and structural per-variable dependence.

%\noindent
%\todo{Discuss related work not covered in the background section. More details on the difference
%between our work and modal logic approaches. etc!}

\section{Conclusions}

In this paper, we looked at two forms of context-dependence
analysis -- \emph{flat} coeffect systems that track whole-context
requirements (such as implicit parameters, resources, or platform version) and
\emph{structural} coeffects that track per-variable requirements (such
as usage or data access patterns). 
The newly introduced structural system makes applications such as liveness,
bounded reuse, and dataflow analysis (from our earlier work)
practically useful. With the move towards cross-platform systems running in 
diverse environments, analysing context dependence is vital for reasoning and compilation. The coeffect calculus
provides a foundation for further study, similar to the type-and-effect discipline.

We presented the system together with its syntactic equational theory and categorical semantics.
The equational theory is presented in order to explain how the systems work, but it also
provides a basis for an operational semantics for concrete systems. Exploring these, 
and their connection to the denotational semantics, is further work.

### 2014-03-27 text/Chapter07.tex
Finish introduction chapter and restructure the rest

\chapter{Coeffect meta-language} 
\label{ch:meta}

%---------------------------------------------------------------------------------------------------

Both flat coeffect calculus and structural coeffect calculus (presented in the past two chapters)
use indexed comonads to define the semantics of the langauge. In this section, we follow the 
meta-language style and embed indexed comonads into the language -- the type constructor
$\ctyp{r}{\alpha}$ becomes a first-class value and we add language constructs corresponding to
primitive operations of the indexed comonad.

%===================================================================================================

\section{Introduction}
\label{sec:metalanguage-intro}

\section{Type system}
\ExecuteMetaData[rules/rules.tex]{cml-types}

\section{Operational properties}

\section{Categorical semantics}

\section{Applications}

\subsection{Meta-programming}

\subsection{Mobile computations}

\section{Related work}
This chapter is closely related to Contextual Modal Type Theory (CMTT) of Nanevski et al. However
they develop their language using model logic as a basis, while we use categorical foundations 
as the basis - leading to a different system.

\section{Summary}

### 2014-03-27 text/Chapter08.tex
Finish introduction chapter and restructure the rest

--- 2014-03-27 text/Chapter07.tex
@@
-\chapter{Coeffect meta-language} 
-\label{ch:meta}
+\chapter{Conclusions} 
+\label{ch:conclusions}
 
@@
 
-Both flat coeffect calculus and structural coeffect calculus (presented in the past two chapters)
-use indexed comonads to define the semantics of the langauge. In this section, we follow the 
-meta-language style and embed indexed comonads into the language -- the type constructor
-$\ctyp{r}{\alpha}$ becomes a first-class value and we add language constructs corresponding to
-primitive operations of the indexed comonad.
+How to make this practical?
 
-%===================================================================================================
+codo notation / computation expressions
 
-\section{Introduction}
-\label{sec:metalanguage-intro}
-
-\section{Type system}
-\ExecuteMetaData[rules/rules.tex]{cml-types}
-
-\section{Operational properties}
-
-\section{Categorical semantics}
-
-\section{Applications}
-
-\subsection{Meta-programming}
-
-\subsection{Mobile computations}
-
-\section{Related work}
-This chapter is closely related to Contextual Modal Type Theory (CMTT) of Nanevski et al. However
-they develop their language using model logic as a basis, while we use categorical foundations 
-as the basis - leading to a different system.
-
-\section{Summary}
+embedding algebraic things

### 2014-03-27 text/backup/Chapter01.tex
Finish introduction chapter and restructure the rest

\chapter{Introduction} 
\label{ch:introduction}

%----------------------------------------------------------------------------------------

In the rest of the chapter, we first look at the distinguishing factor of two of the presented
coeffect systems (how they differ from other existing work). Then we look at a number of concrete 
problems that are discussed in greater details in later chapters. We use the examples to demonstrate 
the real-world problems that motivate our work. (Theoretical and other motivations are discussed 
later in Chapter~\ref{ch:related-work}).

%===================================================================================================

\section{Context and lambda abstraction}

Our work on context-aware programming languages connects two directions in existing research on the 
theory of programming languages. On one side, effect systems \cite{effects-gifford} and monadic 
computations \cite{monad-notions,monads-effects-marriage} provide a detailed and established method 
for tracking what effects programs have -- that is, how they affect the environment where they execute. 
On the other side, the work on comonadic notions of computations \cite{comonads-notions} shows how 
to use the mathematical dual of monads -- comonads -- to give categorical semantics of 
context-dependent computations.

Effect systems introduced track actions such as memory operations or communication. They are described 
by typing judgments of a form $\Gamma \vdash e : \alpha, \sigma$ where $\Gamma$ is the context of a 
program (typically available variables), $e$ is the expression (program) itself, $\alpha$ is the type of 
values returned by the program (e.g. integer or boolean) and $\sigma$ is a set of possible effects. The 
judgment states that, given the context $\Gamma$, an expression has a type $\alpha$ and can only perform 
effects specified by the set $\sigma$. Wadler and Thiemann \cite{monads-effects-marriage} explain how 
this shapes effect analysis of a lambda abstraction -- that is, how effect systems analyze the effects 
associated with a definition of a function:
%
\begin{quote} 
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the body.}
\end{quote}
%
This means that, when a programmer defines a function, the system records that executing the 
function will perform the effects of the body of the function. However, simply defining a function
is an effect-free computation. Tate \cite{effects-producer-semantics} calls such effect systems
\emph{producer} effect systems and generalizes the idea of function to more general ``thunking'':
%
\begin{quote}
\emph{We will define an effect as a producer effect if all computations with that effect 
can be thunked as "pure" computations for a domain-specific notion of purity.}
\end{quote}
%
In contrast to the static analysis of (producer) effect systems, the analysis of 
\emph{context-dependence} does not match this pattern. In the systems we consider, lambda 
abstraction places requirements on both the \emph{call-site} (latent requirements) and the 
\emph{declaration-site} (immediate requirements), resulting in different syntactic properties. 
We informally discuss three examples first that demonstrate how contextual requirements propagate.

%===================================================================================================

\section{Why coeffects matter}

This section gives three examples of context-dependent computations whose properties can be captured
by the tree calculi presented in this thesis. We look at an example of the \emph{flat coeffect calculus} 
(Chapter~\ref{ch:flat-coeffects}), \emph{structural coeffect calculus} (Chapter~\ref{ch:structural-coeffects})
and \emph{coeffect meta-language} (Chapter~\ref{ch:coeffect-metalanguage}).

%---------------------------------------------------------------------------------------------------

\subsection{Flat coeffect calculus}

The flat coeffect calculus associates a single piece of information with the context. As an example,
we look at a simple distributed programming language that includes the concept of \emph{resources}.
A resource may be accessed using a special construct $\kvd{access}~\ident{Res}$. The following 
example shows a function taht lists recent events -- it accesses the resource \ident{News} 
(representing a database) and a resource \ident{Clock} (with the current time):
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\
\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\
\quad\ident{query}~\ident{db}~\str{"SELECT * WHERE Date > \%1"}~(\kvd{access}~\ident{Clock})
\end{array}
\end{equation*}
%
Consider a scenario where the function is \emph{declared} on the server-side and then
transferred and \emph{executed} on the client-side. The resource \ident{News} represents a 
database that is only available on the server-side and so the function needs to keep a remote
reference to the server. However, the \ident{Clock} resource may (or may not) be available on the
client-side. If the resource is available on the client, then it may be re-bound and the function
will use the current client's data -- for example, to accommodate for time-zone changes.

This example demonstrates how lambda abstraction behaves for context-dependent computations. The 
context requirement of the function body is a set of resources $\{\ident{Clock}, \ident{News}\}$.
The context requirements are split between the \emph{declaration-site} and \emph{call-site}.
However, there are multiple possible splittings. The splitting $\{\ident{News}\} \cup \{\ident{Clock}\}$
models the case when the database is accessed from the server, but time is taken from the client,
while the splitting $\{\ident{Clock}, \ident{News}\} \cup \{\}$ corresponds to the case when both
resources are accessed from the server.

%---------------------------------------------------------------------------------------------------

\subsection{Structural coeffect calculus}

The calculus used in the previous section annotates the entire context with a single value --  such
as the set of required resources. However, sometimes it is desirable to annotate not the entire 
context, but individual variables of the context.

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\times$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
(\ident{a}:\dnat, \ident{b}:\dnat) ~@~ 5 \times 10
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The annotations $5 \times 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}, respectively. If we substitute 
\ident{c} for both \ident{a} and \ident{b}, we need to combine the context-requirements and take the
maximum of the requirements of the individual variables:
%
\begin{equation*}
\begin{array}{l}
(\ident{c}:\dnat) ~@~ \ident{max}(5, 10)
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
This version of the calculus removes the non-determinism of the lambda abstraction from the previous
version. As we associate information with individual variables, lambda abstraction creates a function
that requires the context requirements associated with the variable that is being abstracted over.

For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
\ident{b} from the declaration-site) then the function would have context requirements $5$ -- that
is the number associated with the variable \ident{a}.

%---------------------------------------------------------------------------------------------------

\subsection{Coeffect meta-language}

\todo{Give some example that uses the coeffects metalanguage style. This will probably be 
meta-programming with open expressions (similarly to what Pfenning and Nanevski do with 
contextual modal type theory).}

%---------------------------------------------------------------------------------------------------

\section{Why context matters}

We claimed earlier that context-dependent comptuations are becoming increasingly common and our
work is focused on annotating the (free-variable) context with additional information.
The importance of context can be also demonstrated by looking at a technology that focuses 
solely on making the context richer -- F\# type providers.

\todo{Say more about type providers -- they extend the context $\Gamma$ so that it can be 
lazily loaded and it can be huge. Potentially, it could be also annotated with additional
meta-data...}

%===================================================================================================

\section{Contributions}

\todo{We present three calculi that model common notions of context-dependence and can be used
as basis for developing context-aware programming langauges with static type systems.}

### 2014-04-02 text/Chapter03.tex
Early work on chapter 3

@@
+% ==================================================================================================
+
 \chapter{Context-aware applications} 
@@
 
-%---------------------------------------------------------------------------------------------------
+Software developers as well as programming language researchers choose abstractions based not 
+just on how appropriate they are. Other factors may include social aspects -- how well is the 
+abstraction known, how well is it documented and whether it is a standard tool of the 
+\emph{research programme}\footnote{ Research programme, as introduced by Lakatos \cite{philosophy-lakatos},
+is a network of scientists sharing the same basic assumptions and techniques.}.
 
-TODO
+In Chapter~\ref{ch:intro}, we argued that context-awareness had, so far, only limited influence 
+on the design of programming languages, possibly because it is a challenge that is not easy to
+see. However, many of the properties that we treat uniformly as \emph{coeffects} have been
+previously tracked by other means. This includes special-purpose type systems, systematic 
+approaches arising from modal logic S4, as well as techniques based on abstractions designed 
+for other purpose, most frequently monads.
+
+In this chapter, we describe a number of simple calculi for tracking a wide range of contextual
+properties. The systems are adapted from existing work, but the uniform presentation in this 
+chapter is a novel contribution. The fact that we find a common structure in all systems presented 
+here lets us develop unified coeffect calculi in the upcoming three chapters.
 
@@
 
-\section{Through applications}
-\label{sec:path-apps}
+\section{Structure of coeffect systems}
 
-The general theme of this thesis is improving programming languages to better support writing
-\emph{context-dependent} (or \emph{context-aware}) computations. With current trends in the 
-computing industry such as mobile and ubiquitous computing, this is becoming an important topic.
-In software engineering and programming community, a number of authors have addressed this problem
-from different perspectives. Hirschfeld et al. propose \emph{Context-Oriented Programming} (COP)
-as a methodology \cite{app-cop-method}, and the subject has also been addressed in mobile
-computations \cite{app-cop-mobile,app-cop-mobile2}. In programming languages, Costanza 
-\cite{app-cop-contextl} develops a domain-specific LISP-like language ContextL and Bardram 
-\cite{app-cop-javafwk} proposes a Java framework for COP.
+When introducing coeffect systems in Section~\ref{sec:intro-theory}, we related coeffect systems
+with effect systems. Effect systems track how program affects the environment, or, in other words 
+capture some \emph{output impurity}. In contrast, coeffect systems track what program requires from the 
+environment, or \emph{input impurity}.
 
-We approach the problem from a different perspective, building on the tradition of 
-statically-typed functional programming languages and their theories. However, even in this field,
-there is a number of calculi or language features that can be viewed as context-dependent.
+Effect systems generally use judgements of the form $\Gamma \vdash e : \tau \;\&\; \aclrd{\sigma}$, 
+associating effects $\sigma$ with the output type. In contrast, we choose to write coeffect 
+systems using judgements of the form $\coctx{\Gamma}{\aclrd{\sigma}} \vdash e : \tau$, associating
+the context requirements with $\Gamma$. Thus, we extend the traditional notion of free-variable
+context $\Gamma$ with richer notions of context. Besides the notation, there are more important 
+differences between effects and coeffects.
 
-%---------------------------------------------------------------------------------------------------
+% --------------------------------------------------------------------------------------------------
 
-\subsection{Motivation for flat coeffects}
+\subsection{Lambda abstraction}
+
+The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
+The typical lambda abstraction rule for effect systems looks as (\emph{abs-eff}) in 
+Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how 
+the effect analysis works `as follows:
+%
+\begin{quote}
+\emph{In the rule for abstraction, the effect is empty because evaluation immediately
+returns the function, with no side effects. The effect on the function arrow
+is the same as the effect for the function body, because applying the function will
+have the same side effects as evaluating the body.}
+\end{quote}
+%
+This is the key property of \emph{output impurity}. The effects are only produced when the
+function is evaluated and so the effects of the body are attached to the function. A recent
+work by Tate~\cite{effects-producer-semantics} uses the term \emph{producer} effect systems
+for such standard systems and characterises them as follows:
+%
+\begin{quote}
+\emph{Indeed, we will define an effect as a producer effect if all computations with that 
+effect can be thunked as ``pure'' computations for a domain-specific notion of purity.}
+\end{quote} 
+%
+The thunking is typically a lambda abstraction -- given an effectful expression $e$, the
+function $\lambda x.e$ is effect free value (thunk) that delays all effects.
+
+Coeffects do not follow this pattern. Indeed, we can see this even when we consider the classical
+notion of context. Standard lambda abstraction (\emph{app-pure}) in Figure~\ref{fig:applications-abs}
+splits the free-variable context of an expression between variables defined at the \emph{declaration 
+site} (the lexical scope) and those provided by the \emph{call site} (bound variable, or the 
+parameter). This is also the case for other notions of context that we consider in this thesis.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\[
+\tyrule{abs-pure}
+  { \Gamma, x\-:\-\tau_1 \vdash e : \tau_2}
+  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 1}
+\;\;
+\tyrule{abs-eff}
+  { \Gamma, x\-:\-\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
+  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{\sigma}} \tau_2 \,\&\, \aclrd{\emptyset} }
+\]
+\caption{Lambda abstraction for pure and effectful computations}
+\label{fig:applications-abs}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Notions of context}
+
+Aside from the standard free-variable context, we consider two notions of context -- 
+\emph{flat} refers to overall properties of the environment and \emph{structural} refer to 
+properties attached to individual variables. Similarly to free-variable context, both of these
+notions split the context between declaration-site (lexical scope) and call-site (dynamic scope).
+
+\paragraph{Flat coeffects.}
+
+In distributed 
+systems, additional resources available at certain nodes (\eg~a database or GPS sensor) can be 
+also viewed as a part of typing assumptions.
+
+Similarly, resources in a distributed system can be satisfied by both the declaration site (by 
+capturing a remote reference) and the call site (by resource rebinding)~\cite{app-distributed-acute}. 
+In this case, the resource requirements are split freely.
+
+\paragraph{Structural coeffects.}
+
+The previously described notion of a \emph{coeffect system} \cite{petricek2013coeffects} tracks
+contextual properties of entire programs, which is practically useful for implicit parameters 
+or resources. However, for per-variable coeffects, the previous work provides only an approximation 
+with limited practical usefulness (\eg~marking the whole context as live when just one 
+variable in the context is live).
+
+Furthermore, contextual requirements may relate to 
+specific variables. For example, liveness or access patterns, such as the number of accesses in 
+bounded linear logic or past values in causal dataflow.
+
+For contextual properties associated with 
+variables (\eg~liveness or dataflow), the splitting tracks free variables -- requirements associated 
+with a \emph{bound} variable should be satisfied by the call site, while those associated with 
+\emph{free} variables should be provided by the declaration site.
+
+lambda abstraction places 
+requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
+(immediate requirements),
+
+ We start with some background and finish with a brief overview of the literature leading to coeffects.
+
+In the rest of the chapter, we look at instances of both. But before, we need to look at vectors.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Scalars and vectors}
+
+The $\lambda$-calculus is asymmetric-- it maps a context with \emph{multiple} variables to a 
+\emph{single} result. An expression with $n$ free variables of types $\sigma_i$ can be modelled
+by a function
+$\sigma_1 \times \ldots \times \sigma_n \rightarrow \tau$ with a product on the left, but a single value
+on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
+we attach coeffects to the context $\sigma_1 \times \ldots \times \sigma_n$ and we often (but not always)
+have one coeffect per each variable. We call the overall coeffect a \emph{vector} consisting of
+\emph{scalar} coeffects. This asymmetry explains why coeffect systems are not trivially dual to 
+effect systems.
+
+It is useful to clarify how vectors are used in this paper. Suppose we have a set $\C$ of
+\emph{scalars} such as $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ over $\C$
+is a tuple $\langle \cclrd{r_1}, \ldots, \cclrd{r_n}\rangle$ of scalars. 
+We use letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for 
+vectors and $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars.\footnote{For better readability, the paper 
+distinguishes different structures using colours. However ignoring the colour does not introduce 
+any ambiguity.} We also say that a \emph{shape} of a vector $\slift{\aclrd{R}}$ (or more generally any container) 
+is the set of \emph{positions} in a vector. So, a vector of length
+$n$ has shape $\{ 1, 2, \ldots, n \}$.
+
+Just as in scalar-vector multiplication, we lift any binary operation on scalars into a scalar-vector one:
+$\cclrd{s} \bullet \aclrd{R} = \langle \cclrd{s}\bullet\cclrd{r_1}, \ldots, \cclrd{s}\bullet\cclrd{r_n}\rangle$.
+Given two vectors $\aclrd{R},\aclrd{S}$ of the same shape, containing partially ordered scalars, we 
+write $\aclrd{R} \leq \aclrd{S}$ for the pointwise extension of $\leq$ on scalars. Finally,
+the associative operation $\times$ concatenates vectors.
+
+We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
+but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
+be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.
+
+%===================================================================================================
+
+\section{Flat coeffects (1.)}
 
@@
 
-%---------------------------------------------------------------------------------------------------
+% ==================================================================================================
 
-\subsection{Motivation for structural coeffects}
+\section{Flat coeffects (2.)}
+
+% --------------------------------------------------------------------------------------------------
+\subsection{Implicit parameters and resources.}
+\label{sec:coeffects-res}
+
+Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
+They can be used to parameterize a computation without propagating arguments explicitly through a 
+chain of calls and are part of the context in which expressions evaluate. As correctly expected
+\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
+computations follow a similar pattern, but we discuss implicit parameters for simplicity.
+
+The following function prints a number using implicit parameters
+\ident{?culture} (determining the decimal mark) and \ident{?format}
+(the number of decimal places):
+%
+\begin{equation*}
+\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
+\end{equation*}
+%
+Figure~\ref{fig:example-resources} shows a type-and-coeffect system
+tracking the set of an expression's implicit parameters. 
+For simplicity here, all implicit parameters have type $\rho$.
+
+Context requirements are created in (\emph{access}), while (\emph{var}) requires no
+implicit parameters; (\emph{app}) combines requirements of
+both sub-expressions as well as the latent requirements of the
+function. The (\emph{abs}) rule is where the example differs from
+effect systems.  Function bodies can access the union of the
+parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
+at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
+judgements for the above example are:
+%
+\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
+\begin{equation*}
+\begin{array}{rcl}
+\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
+\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
+\end{array}
+\end{equation*}
+%
+The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
+Different judgments are desirable depending on how a function is used. 
+In the first case, both parameters have to be provided by the caller.
+In the second, both are available at declaration-site, but \ident{?format} may be
+rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).
+
+Implicit parameters can be captured by the
+\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
+  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
+modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
+the next example cannot be structured by \emph{any} monad.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-1.4em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
+%
+\quad\quad
+%
+\inference[(\emph{app})\;]
+  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
+   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
+  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\vspace{-0.7em}
+\begin{equation*}
+\inference[(\emph{access})\;]
+  {}
+  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
+%
+\quad\quad\quad
+%
+\inference[(\emph{abs})\;]
+  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
+\end{equation*}
+\vspace{-0.9em}
+\caption{Selected coeffect rules for implicit parameters}
+\label{fig:example-resources}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[!b]
+\vspace{-2em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
+\quad
+\inference[(\emph{app})\;]
+  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
+    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
+  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\vspace{-1.5em}
+\caption{Selected coeffect rules for liveness analysis}
+\label{fig:example-need}
+\vspace{-1.2em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\vspace{-1em}
+\subsection{Liveness analysis.}
+
+Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
+whether it is definitely not needed (\emph{dead}). A compiler can remove
+bindings to dead variables as the result is never used.
+
+We start with a restricted analysis and briefly mention how to make it practical
+later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
+to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
+
+A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
+($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
+Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
+if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
+A dead context may be marked as live by
+letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
+
+The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
+\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
+The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
+The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
+$g$ ignores its input (regardless of the evaluation order of the underlying language).
+We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
+Secondly, a \emph{pointwise composition} passes the same
+argument to $g$ and $h$. The parameter is live if either the parameter
+of $g$ or $h$ is live ($r \sqcup s$). Application combines the
+two operations, so the context $\Gamma$ is live if it is needed by
+$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
+
+An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
+combines the context annotations using $\sqcap$. Thus, if the body uses some 
+variables, both the function argument and the context of the declaration-site are marked as live.
+
+Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
+In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
+indexed comonads (\S\ref{sec:comonads}), we
+model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
+$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
+(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
+value when $r=\ident{D}$).
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-1em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{0}{\Gamma} \vdash x : \tau }
+%
+\quad\quad
+% 
+\inference[(\emph{app})\;]
+  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
+   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
+  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\inference[(\emph{prev})\;]
+  {\ctyp{n}{\Gamma} \vdash e : \tau}
+  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
+%
+\quad\quad
+%
+\inference[(\emph{abs})\;]
+  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
+\end{equation*}
+\vspace{-1em}
+\caption{Selected coeffect rules for causal data flow}
+\label{fig:example-dataflow}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\vspace{-1em}
+\subsection{Efficient dataflow.}
+Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
+computations over streams. In \emph{causal} data flow, program may
+access past values -- in this setting, a function $\tau_1 \rightarrow
+\tau_2$ becomes a function from a list of historical values $[\tau_1]
+\rightarrow \tau_2$. A coeffect system here tracks how many past values
+to cache.
+
+Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
+number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
+The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
+additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
+
+The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
+adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
+for the second function); passing the context to two subcomputations requires the maximum
+number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
+needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
+call-site must each provide at least the number of past values required by the function body
+(the body may use variables coming from the declaration-site as well as the argument).
+
+The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
+\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
+$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
+(inefficient) unbounded lists of past elements. The above static analysis provides an
+approximation of the number of required past elements and so we use just fixed-length lists.
+
+% ==================================================================================================
+
+\section{Structural coeffects (1.)}
 
@@
 
+%===================================================================================================
+
+\section{Structural coeffects}
+\label{sec:intro-coeffects}
+
+Coeffects are way to describe notions of context in programming that keep turning up. 
+To illustrate this, we overview three systems tracking contextual properties that 
+ motivate our general coeffect system. Two systems track per-variable properties (bounded 
+linear logic and dataflow) and one tracks whole-context properties (implicit parameters).
+
+\subsection{Bounded reuse}
+\label{sec:intro-coeffects-bll}
+
+Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
+This is done by limiting the number of times a value (proposition) can be used. An assumption
+$!_k A$ means that a variable can be used at most $k$ times. 
+We attach annotations to the whole context rather than individual assumptions and so a context
+$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
+This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
+%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.
+
+Bounded linear logic includes explicit weakening and contraction rules
+that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
+\[
+\inference
+  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
+  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
+\quad
+\inference
+  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
+  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
+\]
+The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
+$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
+In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
+occurrences of a proposition are joined by adding the number of uses.
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\begin{align*}
+\begin{array}{c}
+%
+\hspace{-0.6em}
+\begin{array}{c}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
+\quad 
+\tyruleCompact{weak}
+  {\coctx{\Gamma}{R} \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
+\\[1.8em]
+\tyruleCompact{sub}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
+\quad 
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
+\\[1.2em]
+\end{array} \\
+%
+% ------------------------ 
+%
+\begin{array}{c}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
+    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
+\\[1.3em]
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
+\\[1.3em]
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
+\end{array} 
+% ------------------------ 
+%
+% \hspace{-1em}
+%\begin{array}{c}
+%
+% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
+% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
+% to keep it simple.
+%
+% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
+% (I do not understand why but it makes the line somehow thicker...)
+%
+% \tyruleCompact{let}
+%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
+%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
+%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
+\end{array}
+\end{align*}
+\vspace{-0.5em}
+\caption{type and coeffect system for bounded reuse}
+\label{fig:bounded-coeff}
+\end{figure}
+
+\paragraph{Bounded linear coeffects.}
+The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
+Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
+$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
+the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
+exchange (\emph{exch}) rules manipulate variables in the context and
+modify the annotations accordingly -- adding the number of uses in
+contraction and switching vector elements in exchange.
+
+For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
+and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
+\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
+annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.
+
+Application (\emph{app}) describes call-by-name evaluation.  Applying
+a function that uses its parameter $\cclrd{t}$-times to an argument
+that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
+total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
+Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
+vector. Meanwhile, the variables in $\Gamma_1$ are used just
+$\aclrd{R}$-times when reducing the expression $e_1$ to a function
+value.
+
+Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
+the number of uses using the pointwise
+$\leq$ relation. We can view any variable as being used a greater
+number of times than it actually is.
+
+\paragraph{Example.} To demonstrate, consider a term
+$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
+times -- once directly inside the function and twice via the variable $v$ after 
+substitution. Similarly, $y$ is used 
+twice. Assuming a judgment for the function body, abstraction yields:
+%
+\begin{equation*}
+\tyrule{abs}
+ { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
+ { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
+\end{equation*}
+%
+To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
+Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
+a judgment that is simplified as follows:
+\[
+\hspace{-0.7em}\inference
+  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
+          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
+{\inference
+  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
+  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
+\]
+%
+The first step performs scalar multiplication, producing the vector
+$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
+$x$ and $x'$ from the function and argument terms respectively.
+
+It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
+the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
+show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.
+
 % --------------------------------------------------------------------------------------------------
 
-\subsection{Beyond passive contexts}
+\subsection{Dataflow and data access}
+
+Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
+An expression is re-evaluated when new inputs are available (push) or when more output
+is demanded (pull).  In causal dataflow, programs can access past
+values of a stream. We consider a language where $\kvd{prev}~e$
+returns the previous value of $e$, where 
+$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.
+
+An implementation of causal dataflow may cache past values of
+variables as an optimisation. The question is, how many past values
+should be cached?  This can be approximated by a coeffect system.
+
+\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
+similar to the one for bounded reuse in that it tracks a vector
+of numbers $\aclrd{R}$ as part of the context
+$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
+past values (\emph{causality depth}) required for a variable.
+
+Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
+coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
+accessed variables (\emph{var}) are annotated with
+$0$ meaning that no past value is required (only the
+current one). The (\emph{prev}) rule crates caching requirements --
+it increments the number of required values for all variables
+used in $e$ using scalar-vector addition.
+
+Application and contraction have the same structure as before, 
+but use different operators. If two variables 
+are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
+values, then overall we need at most
+$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
+two caches are combined with the maximum of the two requirements, 
+which satisfy the smaller requirements.
+
+In (\emph{app}), the function requires $\cclrd{t}$ past values of
+its parameter.  This means $\cclrd{t}$ past values of
+$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
+variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
+call (\eg{}, we need $1+\aclrd{S}$ values to
+get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
+values to get $2$ past values of $\sigma$, \emph{etc.}).
+
+\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
+$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
+and one bound:
+\[
+\dfrac
+  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
+  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
+\]
+The expression always requires the previous value of $y$ and adds it to
+a previous value of the parameter $x$. Evaluating the value of the
+argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
+and so the overall requirement is $3$ past values:
+\[
+\inference
+  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
+{\inference
+  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
+  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
+\]
+The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
+(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
+the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
+Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
+number of enclosing $\kvd{prev}$s.% expressions.
+
+We previously used dataflow as an example of
+coeffects~\cite{petricek2013coeffects}, but tracked caching
+requirements on the whole context. The system outlined here is more
+powerful and practically useful, with finer-grained coeffects tracking
+caching requirements per-variable.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\[
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
+\]
+\[
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
+    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
+\]
+\[
+\tyrule{var}{}
+  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
+\quad\;
+\tyrule{prev}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
+  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
+\]
+\caption{type and coeffect system for dataflow caching}
+\label{fig:dataflow-coeff}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Implicit parameters}
+\label{sec:intro-coeffects-impl}
+
+As our third example, we revisit Haskell implicit parameters \cite{lewis2000implicit} 
+used in our earlier coeffect work \cite{petricek2013coeffects}. Implicit parameters are variables
+that mix aspects of dynamic and lexical scoping. Implicit parameters are a distinct syntactic 
+category to variables and we write them as \ident{?p}. For simplicity, we omit \emph{let}-binding for
+implicit parameters and focus just on tracking requirements.
+
+\paragraph{Implicit parameters coeffects.} Implicit parameters are a whole-context coeffect 
+not linked to ordinary variables. We keep track of sets of implicit parameters that are required
+by an expression (and their types). For example $\coctx{\Gamma}{\{ \ident{?p}_1 : \tau_1, \ldots, \ident{?p}_n : \tau_n \}}$
+means that a context provides ordinary variables $\Gamma$ and values for implicit parameters $\ident{?p}_i$.
+Unlike in the previous examples, we no longer need to distinguish between coeffects attached
+to variables (scalars) and coeffects attached to contexts (vectors), so we write $\cclrd{r},\cclrd{s},\cclrd{t}$
+for both.
+
+Despite the differences, the type system in Figure~\ref{fig:implpar-coeff} follows the same 
+structure as the earlier two examples. Context requirements are created when accessing an implicit
+parameter (\emph{param}) (a system-specific rule). Structural rules (exchange, weaken, contract) 
+do not affect the coeffects. For example parameters are reordered in (\emph{exch}),
+but this has no effect as set union $\cup$ is commutative.
+
+In abstraction and application, the structural $\cons$ operator (previously vector 
+concatenation) becomes $\cup$. Sets of implicit parameters are not associated to
+individual variables and so they are unioned. The (\emph{app}) rule uses $\cup$ to
+combine the implicit parameters required by the function with the requirements of the argument too.
+
+We call this a \emph{flat} coeffect system since coeffects have only
+one shape (there is no scalar/vector distinction).  Other flat
+coeffect systems may use a richer structure~\cite{petricek2013coeffects}. In
+particular, the operations used in abstraction and application may differ
+(to accommodate over-approximation). We return to this in
+Section~\ref{sec:coeffects-structural}.
+
+\paragraph{Example.} Unlike structural coeffect systems, flat systems
+do not necessarily have principal coeffects. 
+This arises from the (\emph{abs}) rule which can freely split
+requirements between the function type and the declaring
+context. Consider a function %filters news based on a location:
+$\lambda ().\ident{?p}_1 + \ident{?p}_2$.  There are nine possible 
+ type and coeffect derivations, two of which are:
+%% 1 p1p2  p1p2
+%% 2 p1p2  p1
+%% 3 p1p2  p2
+%% 4 p1p2  
+%% 5 p1    p1p2
+%% 6 p1    p2
+%% 7 p2    p1p2
+%% 8 p2    p1
+%% 9       p1p2
+\[
+\begin{array}{rl}
+\coctx{\emptyset}{ \{  \} } &\hspace{-0.7em}\vdash (\ldots) : 
+  \ident{unit} \xrightarrow{ \{ \ident{?p}_1:\mathbb{Z}, \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
+\\
+\coctx{\emptyset}{ \{ \ident{?p}_1:\mathbb{Z} \} } &\hspace{-0.7em} \vdash (\ldots) : 
+  \ident{unit} \xrightarrow{ \{ \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
+\end{array}
+\]
+In the first case, both parameters are dynamically scoped and have to be provided by the caller.
+In the second case, the parameter $\ident{?p}_1$ is available in the declaring scope and so it is
+(lexically) captured.
+
+Although structural coeffects have more desirable syntactic properties, we aim to capture this
+non-principality too as it is practically useful -- Haskell's implicit parameters use it and
+it can be used to model resource rebinding in distributed systems such as~\cite{app-distributed-acute}).
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\[
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau,y:\sigma,\Gamma_2}{r \cup \cclrd{s} \cup \cclrd{t} \cup \cclrd{q}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\sigma,x:\tau,\Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s} \cup \cclrd{q}} \vdash e : \tau}
+\]
+\[
+\tyrule{app}
+  { \coctx{\Gamma_1}{\cclrd{r}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
+    \coctx{\Gamma_2}{\cclrd{s}} \vdash e_2 : \sigma }
+  { \coctx{\Gamma_1, \Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s}} \vdash e_1 \, e_2 : \tau} 
+\]
+\[
+\tyrule{param}{}
+  {\coctx{()}{ \{ \ident{?p}:\tau \} } \vdash \ident{?p} : \tau}
+\quad\;
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \sigma}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
+\]
+\vspace{-0.5em}
+\caption{Type and coeffect system for implicit parameters}
+\label{fig:implpar-coeff}
+\end{figure}
+
+% ==================================================================================================
+
+\section{Beyond passive contexts}
 
@@
 
+\section{Context oriented programming}
+
+The importance of context-aware computations is perhaps most obvious when considering mobile
+application, client/server web applications or even the internet of things. A pioneering work
+in the area using functional languages has been done by Serrano \cite{app-hop-diffuse,app-hop-lang}
+(which also inspired the example presented in Chapter~\ref{ch:intro}). His HOP language supports 
+cross-compilation and programs execute in different contexts. However, HOP is not statically 
+type checked.
+
+In the software engineering community, a number of authors have addressed the
+problem of context-aware computations. Hirschfeld et al. propose \emph{Context-Oriented Programming} 
+(COP) as a methodology \cite{app-cop-method}. The COP paradigm has been later implemented by
+programming language features. Costanza \cite{app-cop-contextl} develops a domain-specific LISP-like 
+language ContextL and Bardram \cite{app-cop-javafwk} proposes a Java framework for COP.
+
+Finally, the subject of context-awareness has also been addressed in work focusing on the development 
+of mobile applications \cite{app-cop-mobile,app-cop-mobile2}. Here, the \emph{context} focuses more 
+on concrete physical context (obtained from the device sensors) than context as an abstract 
+language feature.
+
+We approach the problem from a different perspective, building on the tradition of 
+statically-typed functional programming languages and their theories.
+
+%===================================================================================================
+
 \section{Summary}

### 2014-04-02 text/Chapter04.tex
Early work on chapter 3

@@
 which capture notions of computation not captured by ordinary comonads.
-
-% ==================================================================================================
-
-\section{Motivation}
-\label{sec:motivation}
-
-Effect systems, introduced by Gifford and Lucassen
-\cite{effects-gifford}, track \emph{effects} of computations, such as
-memory access or message-based
-communication~\cite{effects-messagepassing}. Their approach augments
-typing judgments with effect information: $\Gamma \vdash e :
-\tau, F$.  Wadler and Thiemann explain how this shapes
-effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
-%
-\begin{quote}
-\emph{In the rule for abstraction, the effect is empty because evaluation immediately
-returns the function, with no side effects. The effect on the function arrow
-is the same as the effect for the function body, because applying the function will
-have the same side effects as evaluating the body.}
-\end{quote}
-%
-In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
-does not match this pattern. In the systems we consider, lambda abstraction places 
-requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
-(immediate requirements), resulting in different syntactic properties. 
-We informally discuss three examples first that demonstrate how contextual
-requirements propagate. Section (\S\ref{sec:calculus}) then unifies these 
-in a single calculus.
-
-We write coeffect judgements $\ctyp{s}{\Gamma} \vdash e : \tau$ where the 
-coeffect annotation $s$ associates context requirements with the 
-free-variable context $\Gamma$. Function types have the form $\ctyp{s}{\tau_1} \rightarrow \tau_2$ associating
-\emph{latent} coeffects $s$ with the parameter.
-The $\ctyp{s}{\Gamma}$ syntax and $\ctyp{s}{\tau}$ types are a result of the
-indexed comonadic semantics (\S\ref{sec:comonads}).
-
-% --------------------------------------------------------------------------------------------------
-\vspace{-1em}
-\subsection{Implicit parameters and resources.}
-\label{sec:coeffects-res}
-
-Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
-They can be used to parameterize a computation without propagating arguments explicitly through a 
-chain of calls and are part of the context in which expressions evaluate. As correctly expected
-\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
-computations follow a similar pattern, but we discuss implicit parameters for simplicity.
-
-The following function prints a number using implicit parameters
-\ident{?culture} (determining the decimal mark) and \ident{?format}
-(the number of decimal places):
-%
-\begin{equation*}
-\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
-\end{equation*}
-%
-Figure~\ref{fig:example-resources} shows a type-and-coeffect system
-tracking the set of an expression's implicit parameters. 
-For simplicity here, all implicit parameters have type $\rho$.
-
-Context requirements are created in (\emph{access}), while (\emph{var}) requires no
-implicit parameters; (\emph{app}) combines requirements of
-both sub-expressions as well as the latent requirements of the
-function. The (\emph{abs}) rule is where the example differs from
-effect systems.  Function bodies can access the union of the
-parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
-at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
-judgements for the above example are:
-%
-\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
-\begin{equation*}
-\begin{array}{rcl}
-\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
-\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
-\end{array}
-\end{equation*}
-%
-The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
-Different judgments are desirable depending on how a function is used. 
-In the first case, both parameters have to be provided by the caller.
-In the second, both are available at declaration-site, but \ident{?format} may be
-rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).
-
-Implicit parameters can be captured by the
-\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
-  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
-modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
-the next example cannot be structured by \emph{any} monad.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-1.4em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-%
-\inference[(\emph{app})\;]
-  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\vspace{-0.7em}
-\begin{equation*}
-\inference[(\emph{access})\;]
-  {}
-  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
-%
-\quad\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\vspace{-0.9em}
-\caption{Selected coeffect rules for implicit parameters}
-\label{fig:example-resources}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[!b]
-\vspace{-2em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
-\quad
-\inference[(\emph{app})\;]
-  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
-    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
-  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\vspace{-1.5em}
-\caption{Selected coeffect rules for liveness analysis}
-\label{fig:example-need}
-\vspace{-1.2em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\subsection{Liveness analysis.}
-
-Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
-whether it is definitely not needed (\emph{dead}). A compiler can remove
-bindings to dead variables as the result is never used.
-
-We start with a restricted analysis and briefly mention how to make it practical
-later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
-to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
-
-A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
-($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
-Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
-if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
-A dead context may be marked as live by
-letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
-
-The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
-\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
-The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
-The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
-$g$ ignores its input (regardless of the evaluation order of the underlying language).
-We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
-Secondly, a \emph{pointwise composition} passes the same
-argument to $g$ and $h$. The parameter is live if either the parameter
-of $g$ or $h$ is live ($r \sqcup s$). Application combines the
-two operations, so the context $\Gamma$ is live if it is needed by
-$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
-
-An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
-combines the context annotations using $\sqcap$. Thus, if the body uses some 
-variables, both the function argument and the context of the declaration-site are marked as live.
-
-Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
-In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
-indexed comonads (\S\ref{sec:comonads}), we
-model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
-$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
-(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
-value when $r=\ident{D}$).
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-1em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{0}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-% 
-\inference[(\emph{app})\;]
-  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\inference[(\emph{prev})\;]
-  {\ctyp{n}{\Gamma} \vdash e : \tau}
-  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
-%
-\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\vspace{-1em}
-\caption{Selected coeffect rules for causal data flow}
-\label{fig:example-dataflow}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\subsection{Efficient dataflow.}
-Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
-computations over streams. In \emph{causal} data flow, program may
-access past values -- in this setting, a function $\tau_1 \rightarrow
-\tau_2$ becomes a function from a list of historical values $[\tau_1]
-\rightarrow \tau_2$. A coeffect system here tracks how many past values
-to cache.
-
-Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
-number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
-The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
-additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
-
-The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
-adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
-for the second function); passing the context to two subcomputations requires the maximum
-number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
-needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
-call-site must each provide at least the number of past values required by the function body
-(the body may use variables coming from the declaration-site as well as the argument).
-
-The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
-\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
-$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
-(inefficient) unbounded lists of past elements. The above static analysis provides an
-approximation of the number of required past elements and so we use just fixed-length lists.
 

### 2014-04-02 text/Chapter06.tex
Early work on chapter 3

@@
 % Typing rule and typing statement (single line)
-\newcommand{\tyrule}[3]{ \inference[(\footnotesize{#1})~]{#2}{#3} }
-\newcommand{\tyruleCompact}[3]{ \text{(\footnotesize{#1})}~\dfrac{#2}{#3} }
 \newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
@@
 \newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
-% Ordinary context where colouring is done by hand 
-\newcommand{\coctx}[2]{ #1 \text{\scriptsize @} {\footnotesize{\text{$#2$}}}  }
-
-% Simple coeffect systems in the introduction
-\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}
-
-% Coeffect algebra
-\definecolor{cclr}{rgb}{0.0,0.4,0.0}
-\newcommand{\cclrd}[1]{ \textcolor{cclr}{#1} }
-
-\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
+
 \newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
@@
 % Coeffect shapes
-\definecolor{sclr}{rgb}{0.5,0,0.0}
-\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }
 
@@
 \newcommand{\sflat}{\star}
-
-% Coeffect algebra
-\definecolor{aclr}{rgb}{0.0,0.0,0.7}
-\newcommand{\aclrd}[1]{ \textcolor{aclr}{#1} }
-
-\newcommand{\mksh}[1]{{\sclrd{[}#1\sclrd{]}}}
 
@@
 % { \textcolor{aclr}{\bullet}}
-\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}} 
 \newcommand{\azero}{ \textcolor{aclr}{\bot} }
@@
 
-\section{Why coeffects matter}
-\label{sec:intro-coeffects}
-
-Coeffects are way to describe notions of context in programming that keep turning up. 
-To illustrate this, we overview three systems tracking contextual properties that 
- motivate our general coeffect system. Two systems track per-variable properties (bounded 
-linear logic and dataflow) and one tracks whole-context properties (implicit parameters).
- We start with some background and finish with a brief overview of the literature leading to coeffects.
-
-\subsection{Background, scalars and vectors}
-
-The $\lambda$-calculus is asymmetric-- it maps a context with \emph{multiple} variables to a 
-\emph{single} result. An expression with $n$ free variables of types $\sigma_i$ can be modelled
-by a function
-$\sigma_1 \times \ldots \times \sigma_n \rightarrow \tau$ with a product on the left, but a single value
-on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
-we attach coeffects to the context $\sigma_1 \times \ldots \times \sigma_n$ and we often (but not always)
-have one coeffect per each variable. We call the overall coeffect a \emph{vector} consisting of
-\emph{scalar} coeffects. This asymmetry explains why coeffect systems are not trivially dual to 
-effect systems.
-
-It is useful to clarify how vectors are used in this paper. Suppose we have a set $\C$ of
-\emph{scalars} such as $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ over $\C$
-is a tuple $\langle \cclrd{r_1}, \ldots, \cclrd{r_n}\rangle$ of scalars. 
-We use letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for 
-vectors and $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars.\footnote{For better readability, the paper 
-distinguishes different structures using colours. However ignoring the colour does not introduce 
-any ambiguity.} We also say that a \emph{shape} of a vector $\mksh{\aclrd{R}}$ (or more generally any container) 
-is the set of \emph{positions} in a vector. So, a vector of length
-$n$ has shape $\{ 1, 2, \ldots, n \}$.
-
-Just as in scalar-vector multiplication, we lift any binary operation on scalars into a scalar-vector one:
-$\cclrd{s} \bullet \aclrd{R} = \langle \cclrd{s}\bullet\cclrd{r_1}, \ldots, \cclrd{s}\bullet\cclrd{r_n}\rangle$.
-Given two vectors $\aclrd{R},\aclrd{S}$ of the same shape, containing partially ordered scalars, we 
-write $\aclrd{R} \leq \aclrd{S}$ for the pointwise extension of $\leq$ on scalars. Finally,
-the associative operation $\times$ concatenates vectors.
-
-We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
-but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
-be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.
-
-\subsection{Bounded reuse}
-\label{sec:intro-coeffects-bll}
-
-Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
-This is done by limiting the number of times a value (proposition) can be used. An assumption
-$!_k A$ means that a variable can be used at most $k$ times. 
-We attach annotations to the whole context rather than individual assumptions and so a context
-$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
-This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
-%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.
-
-Bounded linear logic includes explicit weakening and contraction rules
-that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
-\[
-\inference
-  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
-  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
-\quad
-\inference
-  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
-  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
-\]
-The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
-$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
-In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
-occurrences of a proposition are joined by adding the number of uses.
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\begin{array}{c}
-%
-\hspace{-0.6em}
-\begin{array}{c}
-\tyrule{var}{}
-  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
-\quad 
-\tyruleCompact{weak}
-  {\coctx{\Gamma}{R} \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
-\\[1.8em]
-\tyruleCompact{sub}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
-\quad 
-\tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
-\\[1.2em]
-\end{array} \\
-%
-% ------------------------ 
-%
-\begin{array}{c}
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
-    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
-\\[1.3em]
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
-\\[1.3em]
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
-\end{array} 
-% ------------------------ 
-%
-% \hspace{-1em}
-%\begin{array}{c}
-%
-% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
-% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
-% to keep it simple.
-%
-% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
-% (I do not understand why but it makes the line somehow thicker...)
-%
-% \tyruleCompact{let}
-%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
-%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
-%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
-\end{array}
-\end{align*}
-\vspace{-0.5em}
-\caption{type and coeffect system for bounded reuse}
-\label{fig:bounded-coeff}
-\end{figure}
-
-\paragraph{Bounded linear coeffects.}
-The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
-Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
-$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
-the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
-exchange (\emph{exch}) rules manipulate variables in the context and
-modify the annotations accordingly -- adding the number of uses in
-contraction and switching vector elements in exchange.
-
-For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
-and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
-\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
-annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.
-
-Application (\emph{app}) describes call-by-name evaluation.  Applying
-a function that uses its parameter $\cclrd{t}$-times to an argument
-that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
-total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
-Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
-vector. Meanwhile, the variables in $\Gamma_1$ are used just
-$\aclrd{R}$-times when reducing the expression $e_1$ to a function
-value.
-
-Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
-the number of uses using the pointwise
-$\leq$ relation. We can view any variable as being used a greater
-number of times than it actually is.
-
-\paragraph{Example.} To demonstrate, consider a term
-$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
-times -- once directly inside the function and twice via the variable $v$ after 
-substitution. Similarly, $y$ is used 
-twice. Assuming a judgment for the function body, abstraction yields:
-%
-\begin{equation*}
-\tyrule{abs}
- { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
- { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
-\end{equation*}
-%
-To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
-Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
-a judgment that is simplified as follows:
-\[
-\hspace{-0.7em}\inference
-  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
-          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
-{\inference
-  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
-  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
-\]
-%
-The first step performs scalar multiplication, producing the vector
-$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
-$x$ and $x'$ from the function and argument terms respectively.
-
-It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
-the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
-show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Dataflow and data access}
-
-Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
-An expression is re-evaluated when new inputs are available (push) or when more output
-is demanded (pull).  In causal dataflow, programs can access past
-values of a stream. We consider a language where $\kvd{prev}~e$
-returns the previous value of $e$, where 
-$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.
-
-An implementation of causal dataflow may cache past values of
-variables as an optimisation. The question is, how many past values
-should be cached?  This can be approximated by a coeffect system.
-
-\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
-similar to the one for bounded reuse in that it tracks a vector
-of numbers $\aclrd{R}$ as part of the context
-$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
-past values (\emph{causality depth}) required for a variable.
-
-Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
-coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
-accessed variables (\emph{var}) are annotated with
-$0$ meaning that no past value is required (only the
-current one). The (\emph{prev}) rule crates caching requirements --
-it increments the number of required values for all variables
-used in $e$ using scalar-vector addition.
-
-Application and contraction have the same structure as before, 
-but use different operators. If two variables 
-are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
-values, then overall we need at most
-$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
-two caches are combined with the maximum of the two requirements, 
-which satisfy the smaller requirements.
-
-In (\emph{app}), the function requires $\cclrd{t}$ past values of
-its parameter.  This means $\cclrd{t}$ past values of
-$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
-variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
-call (\eg{}, we need $1+\aclrd{S}$ values to
-get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
-values to get $2$ past values of $\sigma$, \emph{etc.}).
-
-\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
-$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
-and one bound:
-\[
-\dfrac
-  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
-  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
-\]
-The expression always requires the previous value of $y$ and adds it to
-a previous value of the parameter $x$. Evaluating the value of the
-argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
-and so the overall requirement is $3$ past values:
-\[
-\inference
-  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
-{\inference
-  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
-  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
-\]
-The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
-(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
-the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
-Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
-number of enclosing $\kvd{prev}$s.% expressions.
-
-We previously used dataflow as an example of
-coeffects~\cite{petricek2013coeffects}, but tracked caching
-requirements on the whole context. The system outlined here is more
-powerful and practically useful, with finer-grained coeffects tracking
-caching requirements per-variable.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\[
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
-\]
-\[
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
-    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
-\]
-\[
-\tyrule{var}{}
-  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
-\quad\;
-\tyrule{prev}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
-  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
-\]
-\caption{type and coeffect system for dataflow caching}
-\label{fig:dataflow-coeff}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Implicit parameters}
-\label{sec:intro-coeffects-impl}
-
-As our third example, we revisit Haskell implicit parameters \cite{lewis2000implicit} 
-used in our earlier coeffect work \cite{petricek2013coeffects}. Implicit parameters are variables
-that mix aspects of dynamic and lexical scoping. Implicit parameters are a distinct syntactic 
-category to variables and we write them as \ident{?p}. For simplicity, we omit \emph{let}-binding for
-implicit parameters and focus just on tracking requirements.
-
-\paragraph{Implicit parameters coeffects.} Implicit parameters are a whole-context coeffect 
-not linked to ordinary variables. We keep track of sets of implicit parameters that are required
-by an expression (and their types). For example $\coctx{\Gamma}{\{ \ident{?p}_1 : \tau_1, \ldots, \ident{?p}_n : \tau_n \}}$
-means that a context provides ordinary variables $\Gamma$ and values for implicit parameters $\ident{?p}_i$.
-Unlike in the previous examples, we no longer need to distinguish between coeffects attached
-to variables (scalars) and coeffects attached to contexts (vectors), so we write $\cclrd{r},\cclrd{s},\cclrd{t}$
-for both.
-
-Despite the differences, the type system in Figure~\ref{fig:implpar-coeff} follows the same 
-structure as the earlier two examples. Context requirements are created when accessing an implicit
-parameter (\emph{param}) (a system-specific rule). Structural rules (exchange, weaken, contract) 
-do not affect the coeffects. For example parameters are reordered in (\emph{exch}),
-but this has no effect as set union $\cup$ is commutative.
-
-In abstraction and application, the structural $\cons$ operator (previously vector 
-concatenation) becomes $\cup$. Sets of implicit parameters are not associated to
-individual variables and so they are unioned. The (\emph{app}) rule uses $\cup$ to
-combine the implicit parameters required by the function with the requirements of the argument too.
-
-We call this a \emph{flat} coeffect system since coeffects have only
-one shape (there is no scalar/vector distinction).  Other flat
-coeffect systems may use a richer structure~\cite{petricek2013coeffects}. In
-particular, the operations used in abstraction and application may differ
-(to accommodate over-approximation). We return to this in
-Section~\ref{sec:coeffects-structural}.
-
-\paragraph{Example.} Unlike structural coeffect systems, flat systems
-do not necessarily have principal coeffects. 
-This arises from the (\emph{abs}) rule which can freely split
-requirements between the function type and the declaring
-context. Consider a function %filters news based on a location:
-$\lambda ().\ident{?p}_1 + \ident{?p}_2$.  There are nine possible 
- type and coeffect derivations, two of which are:
-%% 1 p1p2  p1p2
-%% 2 p1p2  p1
-%% 3 p1p2  p2
-%% 4 p1p2  
-%% 5 p1    p1p2
-%% 6 p1    p2
-%% 7 p2    p1p2
-%% 8 p2    p1
-%% 9       p1p2
-\[
-\begin{array}{rl}
-\coctx{\emptyset}{ \{  \} } &\hspace{-0.7em}\vdash (\ldots) : 
-  \ident{unit} \xrightarrow{ \{ \ident{?p}_1:\mathbb{Z}, \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
-\\
-\coctx{\emptyset}{ \{ \ident{?p}_1:\mathbb{Z} \} } &\hspace{-0.7em} \vdash (\ldots) : 
-  \ident{unit} \xrightarrow{ \{ \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
-\end{array}
-\]
-In the first case, both parameters are dynamically scoped and have to be provided by the caller.
-In the second case, the parameter $\ident{?p}_1$ is available in the declaring scope and so it is
-(lexically) captured.
-
-Although structural coeffects have more desirable syntactic properties, we aim to capture this
-non-principality too as it is practically useful -- Haskell's implicit parameters use it and
-it can be used to model resource rebinding in distributed systems such as~\cite{app-distributed-acute}).
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}
-\[
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau,y:\sigma,\Gamma_2}{r \cup \cclrd{s} \cup \cclrd{t} \cup \cclrd{q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\sigma,x:\tau,\Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s} \cup \cclrd{q}} \vdash e : \tau}
-\]
-\[
-\tyrule{app}
-  { \coctx{\Gamma_1}{\cclrd{r}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
-    \coctx{\Gamma_2}{\cclrd{s}} \vdash e_2 : \sigma }
-  { \coctx{\Gamma_1, \Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s}} \vdash e_1 \, e_2 : \tau} 
-\]
-\[
-\tyrule{param}{}
-  {\coctx{()}{ \{ \ident{?p}:\tau \} } \vdash \ident{?p} : \tau}
-\quad\;
-\tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \sigma}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
-\]
-\vspace{-0.5em}
-\caption{Type and coeffect system for implicit parameters}
-\label{fig:implpar-coeff}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Pathways to coeffects}
-\label{sec:intro-pathways}
-
-This paper mainly follows work on effect systems and their link to categorical semantics. We briefly 
-review this and other directions leading to coeffects. An eager reader can return to this section later.
-
-\paragraph{Effect systems.}
-
-Effect systems \cite{gifford1986effects} track
-effectful operations of computations such as memory access or lock
-usage \cite{flanagan2003type}. They are written as judgments $\Gamma
-\vdash e : \tau ~\&~\cclrd{\rho}$ associating effects
-$\cclrd{\rho}$ with the result. Effect systems capture \emph{output
-  effects} where, as Tate puts it, \emph{``all computations with [an]
-  effect can be thunked as pure computations for a domain-specific
-  notion of purity.''}~\cite{effects-producer-semantics}.  This
-thunking is typically a $\lambda$-abstraction. Given an effectful
-expression $e$, the function $\lambda x . e$ is an effect-free value
-that delays all effects:
-%
-\begin{equation*}
-\inference[({abs})\;]
-  {\Gamma, x\!:\!\tau_1 \vdash e : \tau_2 ~\&~\cclrd{\rho}} 
-  {\Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{\rho}} \tau_2~\&~\cclrd{\emptyset}}
-\end{equation*}
-%
-Coeffects do not follow this pattern.  In contrast to effect systems,
-context requirements cannot be easily ``thunked'' as
-pure values. Lambda abstraction can split context requirements
-between \emph{immediate} and \emph{latent} requirements. This is akin
-to how lambda abstraction splits a free-variable context into 
-the bound parameter (call site) and the remaining free variables (declaration site).
-
-\paragraph{Categorical semantics.}
-Moggi models effectful computations as functions of type 
-$\tau_1 \rightarrow \mtyp{}{\tau_2}$ where $M$ is monad providing composition of effectful 
-computations~\cite{moggi1991monads}. Wadler and Thiemann \cite{wadler2003marriage} link effect systems with monads using 
-annotated monads $\tau_1 \rightarrow \mtyp{\cclrd{\rho}}{\tau_2}$ whose semantics has been 
-provided by Katsumata~\cite{katsumata2014parametric}.
-
-Context-dependent computations require a different model. Uustalu and Vene \cite{uustalu2008notions} 
-use functions $C{}{\tau_1} \rightarrow \tau_2$ where $C$ is a \emph{comonad}. Our earlier work 
-\cite{petricek2013coeffects} used indexed comonads with denotations $C^{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ adding 
-annotations akin to Wadler and Thiemann. In Section~\ref{sec:semantics} we extend
-indexed comonads to capture the general coeffect systems of this paper,  
-in the style of Katsumata. 
-% 
-% TP: Mentioned Katsumata here. Thought it is nice to say we build on new work here (as opposed to ICALP)
-%
-
-\paragraph{Language and meta-language.}
-Moggi uses monads in two systems~\cite{moggi1991monads}. In the first system, a monad is used to model
-an effectful language itself -- the semantics of a language uses a specific monad. In the second system, 
-monads are added as type constructors, together with syntax corresponding to \emph{unit} and \emph{bind}
-operations.
-
-Looking at context, Uustalu and Vene \cite{uustalu2008notions} follow the first approach 
-(using a concrete comonad to model dataflow). Contextual-Modal Type Theory (CMTT) of Nanevski et al.
-\cite{logic-cmtt} follows the latter approach, adding a comonad via the $\Box$ modality of modal S4
-to the language. We focus on concrete languages using the first approach. 
-A ``coeffect meta-language'' is an interesting future work.
-
-\paragraph{Sub-structural and bunched types.}
-% \cite{substruct-linear-change} 
-Sub-structural type systems restrict how a context is used. This is achieved by removing some
-of the structural typing rules (weakening, contraction, exchange). As the bounded linear logic 
-example (Section~\ref{sec:intro-coeffects-bll}) shows, our system can be viewed as a generalization.
-
-% ==================================================================================================
-
 \section{The coeffect calculus}
@@
 $\aclrd{R}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
-We write $\aclrd{s} = \mksh{\Gamma}$ for the shape corresponding to $\Gamma$. We define shapes by a 
+We write $\aclrd{s} = \slift{\Gamma}$ for the shape corresponding to $\Gamma$. We define shapes by a 
 \emph{set} of positions and so we can write $\aclrd{R} \in \sclrd{s} \rightarrow \C$ as a
@@
 \begin{definition}
-  A \emph{\sclrd{coeffect shape}} structure $(\SH, \mksh{\textnormal{--}}, \stimes,
+  A \emph{\sclrd{coeffect shape}} structure $(\SH, \slift{\textnormal{--}}, \stimes,
   \sempty, \sunit)$ comprises a set $\SH$ with a binary operation
   $\stimes$ on $\SH$ for shape composition, a mapping from contexts to
-  shapes $\mksh{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
+  shapes $\slift{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
   \SH$ such that $(\SH, \stimes, \sempty)$ is a monoid. The 
@@
   Given a \cclrd{coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero,
-  \cleq)$ and a \sclrd{coeffect shape} $(\SH, \mksh{\textnormal{--}}, \stimes,
+  \cleq)$ and a \sclrd{coeffect shape} $(\SH, \slift{\textnormal{--}}, \stimes,
   \sempty, \sunit)$ a \emph{\aclrd{coeffect algebra}} extends the two
@@
 using $\amerge_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\asplit_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
-The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \mksh{\Gamma_1}$
-and $\sclrd{n} = \mksh{\Gamma_2}$.
+The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slift{\Gamma_1}$
+and $\sclrd{n} = \slift{\Gamma_2}$.
 
@@
 The definition is valid since the shape operations form a monoid
-$\sclrd{(\mathbb{N}, +, 0)}$ and $\mksh{\textnormal{--}}$ (calculating the length of
+$\sclrd{(\mathbb{N}, +, 0)}$ and $\slift{\textnormal{--}}$ (calculating the length of
 a list) is a monoid homomorphism from the free monoid to the monoid of
@@
 An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
-where $\aclrd{\mathsf{VAL}} : \C^{\mksh{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
+where $\aclrd{\mathsf{VAL}} : \C^{\slift{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
 always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
@@
 The semantics is parameterised by a coeffect algebra, with scalar coeffects $(\C,
-\cseq, \cpar, \cunit, \czero, \cleq)$, coeffect shape $(\SH, \mksh{-},
+\cseq, \cpar, \cunit, \czero, \cleq)$, coeffect shape $(\SH, \slift{-},
 \stimes, \sempty, \sunit)$, and $(\amerge, \asplit, \azero)$. An
@@
 \begin{align*}
-\interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \D^{\mksh{\Gamma}}_{\aclrd{R}}
+\interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \D^{\slift{\Gamma}}_{\aclrd{R}}
 \interp{\Gamma} \rightarrow \interp{\tau}
@@
 The interpretation is a morphism from an interpretation of the context $\Gamma$
-to the interpretation of the result. The functor $\D^{\mksh{\Gamma}}_{\aclrd{R}}$ over the
+to the interpretation of the result. The functor $\D^{\slift{\Gamma}}_{\aclrd{R}}$ over the
 context encodes the semantic notion of context and is indexed by the free-variable 
-context shape $\mksh{\Gamma}$ and coeffect annotation $\aclrd{R}$.
+context shape $\slift{\Gamma}$ and coeffect annotation $\aclrd{R}$.
 
@@
 % Thus free-variable contexts are interpreted
-% $\interp{\Gamma} \in \mathbb{C}^{\mksh{\Gamma}}$ and
+% $\interp{\Gamma} \in \mathbb{C}^{\slift{\Gamma}}$ and
 
@@
 As described above, free-variable contexts $\Gamma$ are given an
-interpretation as objects in $\mathbb{C}^{\mksh{\Gamma}}$. Thus,
+interpretation as objects in $\mathbb{C}^{\slift{\Gamma}}$. Thus,
 the interpretation of contexts is shape dependent.
@@
 \paragraph{\textnormal{\emph{Flat coeffects.}}} 
-Recall that $\SH = \{\sflat\}$ and $\mksh{\Gamma} = \sflat$.
+Recall that $\SH = \{\sflat\}$ and $\slift{\Gamma} = \sflat$.
 Since the set of positions $\sflat$ is a singleton, then $\mathbb{C}^{\sflat}$
@@
 \paragraph{\textnormal{\emph{Structural coeffects.}}}
-Recall that $\SH = \mathbb{N}$ and $\mksh{\Gamma} = |\Gamma|$
+Recall that $\SH = \mathbb{N}$ and $\slift{\Gamma} = |\Gamma|$
 (number of free variables), thus 

### 2014-04-07 text/Chapter03.tex
Put all coeffect examples in chapter 3

@@
 \emph{research programme}\footnote{ Research programme, as introduced by Lakatos \cite{philosophy-lakatos},
-is a network of scientists sharing the same basic assumptions and techniques.}.
+is a network of scientists sharing the same basic assumptions and techniques.}. This may
+partly be why no unified context tracking mechanism has been developed so far.
 
 In Chapter~\ref{ch:intro}, we argued that context-awareness had, so far, only limited influence 
-on the design of programming languages, possibly because it is a challenge that is not easy to
-see. However, many of the properties that we treat uniformly as \emph{coeffects} have been
+on the design of programming languages because it is a challenge that is not easy to
+see. However, many of the properties that this thesis treats uniformly as \emph{coeffects} have been
 previously tracked by other means. This includes special-purpose type systems, systematic 
@@
 Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how 
-the effect analysis works `as follows:
+the effect analysis works as follows:
 %
@@
 is the same as the effect for the function body, because applying the function will
-have the same side effects as evaluating the body.}
+have the same side effects as evaluating the~body.}
 \end{quote}
@@
 %
-The thunking is typically a lambda abstraction -- given an effectful expression $e$, the
-function $\lambda x.e$ is effect free value (thunk) that delays all effects.
-
-Coeffects do not follow this pattern. Indeed, we can see this even when we consider the classical
-notion of context. Standard lambda abstraction (\emph{app-pure}) in Figure~\ref{fig:applications-abs}
-splits the free-variable context of an expression between variables defined at the \emph{declaration 
-site} (the lexical scope) and those provided by the \emph{call site} (bound variable, or the 
-parameter). This is also the case for other notions of context that we consider in this thesis.
+The thunking is typically performed by a lambda abstraction -- given an effectful expression 
+$e$, the function $\lambda x.e$ is an effect free value (thunk) that delays all effects.
+As shown in the next section, contextual properties do not follow this pattern.
 
@@
 
-Aside from the standard free-variable context, we consider two notions of context -- 
-\emph{flat} refers to overall properties of the environment and \emph{structural} refer to 
-properties attached to individual variables. Similarly to free-variable context, both of these
-notions split the context between declaration-site (lexical scope) and call-site (dynamic scope).
+We look at three notions of context. The first is the standard free-variable context in 
+$\lambda$-calculus. This is well understood and we use it to demonstrate how contextual 
+properties behave. Then we consider two notions of context introduced in this thesis --
+\emph{flat coeffects} refer to overall properties of the environment and \emph{structural coeffects} 
+refer to properties attached to individual variables.
+
+\paragraph{Variable coeffects.}
+
+In standard $\lambda$-calculus, variable access can be seen as a primitive operation that 
+accesses the context. The expression $x$ introduces a context requirement -- the expression 
+is typeable only in a context that contains $x:\tau$ for some type $\tau$.
+
+The standard lambda abstraction (\emph{abs-pure}), shown in Figure~\ref{fig:applications-abs},
+splits the free-variable context of an expression into two parts. The value of the parameter
+has to be provided by the \emph{call site} (dynamic scope) and the remaining values are provided 
+by the \emph{declaration site} (lexical scope). Here, the splitting is determined syntactically --
+the notation $\lambda x.e$ names the variable whose value comes from the call site.
+
+The flat and structural coeffects behave in the same way. They also split context-requirements
+between the declaration site and call site, but they do it in two different ways.
 
 \paragraph{Flat coeffects.}
+
+In Chapter~\ref{ch:intro}, we used \emph{resources} in a distributed system as an example of flat
+coeffects. These could be, for example, a database, GPS sensor or access to current time.
+We also outlined that such context requirements can be tracked as part of the typing assumption,
+for example, say we have an expression $e$ that requires GPS and current time. The context of
+such expression will be $\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.
+
+The interesting case is when we construct a lambda function $\lambda x.e$, marshall it and
+send it to another node. In that case, the context requirements can be satisfied in a number of
+ways. When the same resource is available at the target machine (\eg.~current time), we can
+send the function with a context requirement and \emph{rebind} the resource. However, if the
+resource is not available (\eg.~GPS on the server), we need to capture a \emph{remote reference}.
+
+In the example discussed here, $\lambda x.e$ would require a context 
+$\coctx{\Gamma}{ \aclrd{\ident{\{ gps \}}} }$
+
+Context requirements in distributed systems can be satisfied in a number of ways. When the 
+resource is available locally,
 
@@
 In this case, the resource requirements are split freely.
+
+We discuss this example in detail in Section~X.
 

### 2014-04-07 text/Chapter01.tex
Early work on chapter 3
Finish section 3.1 on coeffect structures

@@
 \subsection{Context-awareness \#4: Checking array access patterns}
+\label{sec:intro-why-array}
 
@@
 \subsection{Context-aware language in action}
+\label{sec:intro-context-example}
 
@@
 \section{Theory of context dependence}
+\label{sec:intro-theory}
 
@@
 \begin{equation*}
-x\-:\-\ident{int},~y\-:\-\ident{int} \vdash x+y : \ident{int}
+x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
 \end{equation*}
@@
 This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
-be \ident{int}. The expression would also be typeable in a context $x\-:\-\ident{int},~y\-:\-\ident{int}$,
+be \ident{int}. The expression would also be typeable in a context $x\!:\!\ident{int},~y\!:\!\ident{int}$,
 but not, for example, in a context where $x$ has a type \ident{unit}.
@@
 \begin{equation*}
-x\-:\-\ident{int} \vdash \ident{print}~x : \ident{unit}
+x\!:\!\ident{int} \vdash \ident{print}~x : \ident{unit}
 \end{equation*}
@@
 \begin{equation*}
-x\-:\-\ident{int} \vdash \ident{print}~x : \ident{unit~\&~\{ console \}}
+x\!:\!\ident{int} \vdash \ident{print}~x : \ident{unit~\&\aclrd{\{ console \}}}
 \end{equation*}
@@
 \begin{equation*}
-\ident{password}\-:\-\ident{string},~\ident{cmd}\-:\-\ident{string} 
+\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string} 
   \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
@@
 \begin{array}{l}
- ( \ident{password}\-:\-\ident{string \hspace{-1em}{\footnotesize @ sensitive}},
-   \ident{cmd}\-:\-\ident{string})
+ ( \ident{password}\!:\!\ident{string \hspace{-1em}{\footnotesize @ \aclrd{sensitive}}},
+   \ident{cmd}\!:\!\ident{string})
  \hspace{-0.5em}
- \ident{\footnotesize @ \{ database \}} \vdash \\[-0.1em]
+ \ident{\footnotesize @ \aclrd{\{ database \}}} \vdash \\[-0.1em]
  \hspace{3em} \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}

### 2014-04-07 text/Chapter03.tex
Finish section 3.1 on coeffect structures

@@
 \tyrule{abs-pure}
-  { \Gamma, x\-:\-\tau_1 \vdash e : \tau_2}
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2}
   { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 1}
@@
 \tyrule{abs-eff}
-  { \Gamma, x\-:\-\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
   { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{\sigma}} \tau_2 \,\&\, \aclrd{\emptyset} }
@@
 
-In Chapter~\ref{ch:intro}, we used \emph{resources} in a distributed system as an example of flat
-coeffects. These could be, for example, a database, GPS sensor or access to current time.
-We also outlined that such context requirements can be tracked as part of the typing assumption,
-for example, say we have an expression $e$ that requires GPS and current time. The context of
-such expression will be $\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.
+In Section~\ref{sec:intro-context-example}, we used \emph{resources} in a distributed system as an 
+example of flat coeffects. These could be, for example, a database, GPS sensor or access to the current 
+time. We also outlined that such context requirements can be tracked as part of the typing assumption,
+for example, say we have an expression $e$ that requires GPS coordinates and the current time. 
+The context of such expression will be $\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.
 
@@
 
-In the example discussed here, $\lambda x.e$ would require a context 
-$\coctx{\Gamma}{ \aclrd{\ident{\{ gps \}}} }$
-
-Context requirements in distributed systems can be satisfied in a number of ways. When the 
-resource is available locally,
-
-In distributed 
-systems, additional resources available at certain nodes (\eg~a database or GPS sensor) can be 
-also viewed as a part of typing assumptions.
-
-Similarly, resources in a distributed system can be satisfied by both the declaration site (by 
-capturing a remote reference) and the call site (by resource rebinding)~\cite{app-distributed-acute}. 
-In this case, the resource requirements are split freely.
-
-We discuss this example in detail in Section~X.
+In the example discussed here, $\lambda x.e$ would require GPS sensor from the declaration
+site (lexical scope) where the function is declared, which is attached to the current context
+as $\coctx{\Gamma}{ \aclrd{\ident{\{ gps \}}} }$. The current time is required from the caller
+of the function. So, the context requirement on the call site (dynamic scope) will be
+$\aclrd{r}=\aclrd{\footnotesize\ident{\{ time \}}}$. In coeffect systems, we attach this information
+to the function writing $\tau_1 \xrightarrow{ \aclrd{r} } \tau_2$.
+
+We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
+The important point here is that in flat coeffect systems, contextual requirements are 
+\emph{split} between the call-site and declaration-site. Furthermore, in case of distributed
+programming, the resources can be freely distributed between the two sites.
 
@@
 
-The previously described notion of a \emph{coeffect system} \cite{petricek2013coeffects} tracks
-contextual properties of entire programs, which is practically useful for implicit parameters 
-or resources. However, for per-variable coeffects, the previous work provides only an approximation 
-with limited practical usefulness (\eg~marking the whole context as live when just one 
-variable in the context is live).
-
-Furthermore, contextual requirements may relate to 
-specific variables. For example, liveness or access patterns, such as the number of accesses in 
-bounded linear logic or past values in causal dataflow.
-
-For contextual properties associated with 
-variables (\eg~liveness or dataflow), the splitting tracks free variables -- requirements associated 
-with a \emph{bound} variable should be satisfied by the call site, while those associated with 
-\emph{free} variables should be provided by the declaration site.
-
-lambda abstraction places 
-requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
-(immediate requirements),
-
- We start with some background and finish with a brief overview of the literature leading to coeffects.
-
-In the rest of the chapter, we look at instances of both. But before, we need to look at vectors.
+On the one hand, variable context provides a \emph{fine-grained tracking} mechanism of how context
+(variables) are used. On the other hand, flat coeffects let us track \emph{additional information} about 
+the context. The purpose of \emph{structural coeffects} is to reconcile the two and to provide a way
+for fine-grained tracking of additional information related to variables in programs.
+
+In Section~\ref{sec:intro-why-array}, we used an example of tracking array access patterns. For every
+variable, the additional coeffect annotation keeps a range of indices that may be accessed relatively
+to the current cursor. For example, consider an expression 
+$x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$.
+
+Here, the variable context $\Gamma$ contains two variables, both of type \ident{Arr}. This means
+$\Gamma = x\!:\!\ident{Arr},\, y\-!\-!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a 
+language primitive. The coeffect annotations will be $(0,0)$ for $x$ and $(-1,1)$ for $y$, 
+denoting that we access only the current value in $x$, but we need access to both left and right 
+neighbours in the $y$ array. In order to unify the flat and structural notions, we attach this information 
+as a \emph{vector} of annotations associated with a \emph{vector} of variable and write:
+$\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
+The unification is discussed in Chapter~\ref{ch:unified}.
+
+Unlike in flat coeffects, in the structural systems, splitting of context determined by the syntax. 
+For example, consider a function that takes $y$ and contains the above body:
+$\lambda y.x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$. Here, the declaration site
+contains $x$ and needs to provide access at least within a range $(0,0)$. The call site provides
+a value for $y$, which needs to be accessible at least within $(-1, 1)$. In this way, structural
+coeffects remove the non-determinism of flat coeffect systems.
+
+Before looking at concrete flat and structural systems in mode details, we briefly overview 
+some notation used in this thesis. As structural coeffects keep annotations as \emph{vectors},
+we use a number of operations related to scalars and vectors.
 
@@
 
-The $\lambda$-calculus is asymmetric-- it maps a context with \emph{multiple} variables to a 
-\emph{single} result. An expression with $n$ free variables of types $\sigma_i$ can be modelled
-by a function
-$\sigma_1 \times \ldots \times \sigma_n \rightarrow \tau$ with a product on the left, but a single value
+The $\lambda$-calculus is asymmetric -- it maps a context with \emph{multiple} variables to a 
+\emph{single} result. An expression with free variables of types $\tau_i$ can be modelled by a function 
+$\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
 on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
-we attach coeffects to the context $\sigma_1 \times \ldots \times \sigma_n$ and we often (but not always)
-have one coeffect per each variable. We call the overall coeffect a \emph{vector} consisting of
-\emph{scalar} coeffects. This asymmetry explains why coeffect systems are not trivially dual to 
+we attach a coeffect annotation to the context $\tau_1 \times \ldots \times \tau_n$.
+
+Structural coeffects have one coeffect annotation per each variable. Thus, the annotation consists
+of multiple values -- one belonging to each variable. To distinguish between the overall annotation
+and individual (per-variable) annotations, we call the overall coeffect a \emph{vector} consisting of 
+\emph{scalar} coeffects. This asymmetry also explains why coeffect systems are not trivially dual to 
 effect systems.
 
-It is useful to clarify how vectors are used in this paper. Suppose we have a set $\C$ of
-\emph{scalars} such as $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ over $\C$
-is a tuple $\langle \cclrd{r_1}, \ldots, \cclrd{r_n}\rangle$ of scalars. 
-We use letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for 
-vectors and $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars.\footnote{For better readability, the paper 
-distinguishes different structures using colours. However ignoring the colour does not introduce 
-any ambiguity.} We also say that a \emph{shape} of a vector $\slift{\aclrd{R}}$ (or more generally any container) 
-is the set of \emph{positions} in a vector. So, a vector of length
-$n$ has shape $\{ 1, 2, \ldots, n \}$.
-
-Just as in scalar-vector multiplication, we lift any binary operation on scalars into a scalar-vector one:
-$\cclrd{s} \bullet \aclrd{R} = \langle \cclrd{s}\bullet\cclrd{r_1}, \ldots, \cclrd{s}\bullet\cclrd{r_n}\rangle$.
-Given two vectors $\aclrd{R},\aclrd{S}$ of the same shape, containing partially ordered scalars, we 
-write $\aclrd{R} \leq \aclrd{S}$ for the pointwise extension of $\leq$ on scalars. Finally,
-the associative operation $\times$ concatenates vectors.
+It is useful to clarify how vectors are used in this thesis. Suppose we have a set $\C$ of
+\emph{scalars} such that $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ 
+over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars. 
+We use upper-case letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for vectors and lower-case
+letters $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
+also distinguishes different structures using colours. However ignoring the colour does not introduce 
+any ambiguity.}. We also say that a \emph{shape} of a vector $\slift{\aclrd{R}}$ (or more generally any container) 
+is the set of \emph{positions} in a vector. So, a vector of length $n$ has shape $\{ 1, 2, \ldots, n \}$. 
+We discuss containers and shapes further in Chapter~\ref{ch:unified} and also discuss how our use
+relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.
+
+Just as in the usual multiplication of a vector by scalar, we lift any binary operation on scalars into a 
+scalar-vector one. For any binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
+ $\cclrd{s} \circ \aclrd{R} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
+Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{R}, \aclrd{S}$ of the
+same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define 
+$\aclrd{R} \varpropto \aclrd{S} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
+Finally, we often concatenate vectors -- for example, when joining two variable contexts.
+Given vectors $\aclrd{R}, \aclrd{S}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and 
+$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as 
+$\aclrd{R}\times\aclrd{S} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.
 
@@
 
-\paragraph{Implicit parameters} In Haskell, implicit parameters \cite{app-implicit-parameters} are 
+\subsection{Implicit parameters} In Haskell, implicit parameters \cite{app-implicit-parameters} are 
 a special kind of variables that may behave as dynamically scoped. This means, if a function uses 
@@
 
-\paragraph{Type classes}
+\subsection{Type classes}
 Implicit parameters are closely related to \emph{type classes} \cite{app-type-classes}. In Haskell,
@@
 
-\paragraph{Rebindable resources}
+\subsection{Rebindable resources}
+\label{sec:applications-flat-distr}
+
 The need for parameters that do not strictly follow static scoping rules also arises in distributed

### 2014-04-10 text/Chapter03.tex
Chapter 3: Implicit parameters and distributed

@@
 
-\chapter{Context-aware applications} 
+\chapter{Context-aware systems} 
 \label{ch:applications}
@@
 \subsection{Lambda abstraction}
+\label{sec:applications-structure-lam}
 
@@
 
-\section{Flat coeffects (1.)}
-
-In a number of systems, the execution environment provides some additional data, resources or 
-information about the execution context, but are independent of the variables used by the 
-program. We look at implicit parameters and rebindable resources (that both provide additional
-identifiers that can be accessed similarly to variables, but follow different scoping rules),
-distributed programming, cross-compilation and data-flow.
+\section{Flat coeffect systems}
+
+In flat coeffect systems, the additional contextual information are independent of the variables.
+As such, flat coeffects capture properties where the execution environment provides some 
+additional data, resources or information about the execution context.
+
+In this section, we look at a number of examples ranging from Haskell's type constraints
+and implicit parameters to distributed computing. For three of our examples --
+implicit parameters, liveness analysis and data-flow -- we show an ad-hoc type system that
+captures their properties. This serves as a basis for Chapter~\ref{ch:flat}, which develops
+a unified flat coeffect calculus.
 
@@
 
-\subsection{Implicit parameters} In Haskell, implicit parameters \cite{app-implicit-parameters} are 
-a special kind of variables that may behave as dynamically scoped. This means, if a function uses 
-parameter $\ident{?p}$, then the caller of the function must define $\ident{?p}$ and set its value.
-Implicit parameters can be used to parameterise a computation (involving a chain of function calls)
-without passing parameters explicitly as additional arguments of all involved functions. A simple 
-language with implicit parameters has an expression $\ident{?p}$ to read a parameter value and an 
-expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a different keyword to 
-avoid confusion.} $\kvd{letdyn}~?p = e_1~\kvd{in}~e_2$ that sets a parameter $\ident{?p}$ to the value of $e_1$ 
-and evaluates $e_2$ in a context containing $\ident{?p}$
-
-An interesting question arises when we use implicit parameters in a nested function. The following 
-function does some pre-processing and then returns a function that builds a formatted string based 
-on two implicit parameters $\ident{?width}$ and $\ident{?size}$:
+\subsection{Implicit parameters and type classes}
+
+Haskell provides two examples of flat coeffects -- type class and implicit parameter constraints
+\cite{app-type-classes,app-implicit-parameters}. Both of the features introduce additional 
+\emph{constraints} on the context requiring that the environment provides certain operations for
+a type (type classes) or that it provides values for named implicit parameters.
+In the Haskell type system, constraints $C$ are attached to the types of top-level declarations,
+such as let-bound functions. The Haskell notation $\Gamma \vdash e : C \Rightarrow \tau$ 
+corresponds to our notation $\coctx{\Gamma}{C} \vdash e : \tau$.
+
+In this section, we present a type system for implicit parameters in terms of the coeffect typing
+judgement. We briefly consider type classes, but do no give a full type system.
+
+\paragraph{Implicit parameters.}
+Implicit parameters are a special kind of variables that support dynamic scoping.
+They can be used to parameterise a computation (involving a long chain of function calls)
+without passing parameters explicitly as additional arguments of all involved functions.
+
+The dynamic scoping means that if a function uses a parameter $\ident{?param}$ then the caller of the 
+function must set a value of $\ident{?param}$ before calling the function. However, implicit 
+parameters also support lexical scoping. If the parameter $\ident{?param}$ is available in the 
+lexical scope where a function (which uses it) is defined, then the function will not require a
+value from the caller.
+
+A simple language with implicit parameters has an expression $\ident{?param}$ to read a parameter 
+and an expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a 
+different keyword to avoid confusion.} $\kvd{letdyn}~\ident{?param} = e_1~\kvd{in}~e_2$ that sets a 
+parameter $\ident{?param}$ to the value of $e_1$ and evaluates $e_2$ in a context containing 
+$\ident{?param}$.
+
+The fact that implicit parameters support both lexical and dynamic scoping becomes interesting
+when we consider nested functions. The following function does some pre-processing and then returns a 
+function that builds a formatted string based on two implicit parameters $\ident{?width}$ and 
+$\ident{?size}$:
 %
@@
 \begin{array}{l}
-\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\
-\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\
+\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\[-0.25em]
+\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\[-0.25em]
 \quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
-         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})\\
+         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})
 \end{array}
@@
 $\{ \ident{?width} : \ident{int} \}$. The nested function (returned as a result) uses the parameter 
-$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters of the nested 
-function come from?
-
+$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters used by the 
+nested function come from?
+
+To keep examples in this chapter uniform, we do not use the Haskell notation and instead
+write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified $r$.
 In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
-However, in Haskell, implicit parameters behave as a combination of lexical and dynamic scoping. This means
-that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$
-In Haskell, this corresponds to the following type:
-%
-\begin{equation*}
-(\ident{?width} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow 
-  ((\ident{?size} :: \ident{Int}) \Rightarrow \ident{String} \rightarrow \ident{string})
-\end{equation*}
-%
-As a result, the function can be called as follows:
+However, implicit parameters behave as a combination of lexical and dynamic scoping. This means
+that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$.
+The following shows the two options:
+%
+\begin{equation}
+\tag{dynamic}
+\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
+  (\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int}, \ident{?size} : \ident{int} \} } \ident{string})
+\end{equation}
+\vspace{-1em}
+\begin{equation}
+\tag{mixed}
+\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
+  (\ident{string} \xrightarrow{ \{ \ident{?size} : \ident{int} \} } \ident{string})
+\end{equation}
+%
+This is not a complete list of possible typings, but it demonstrates the options. The \emph{dynamic}
+case requires the parameter \ident{?width} twice (this may be confusing when the caller provides
+two different values). In the \emph{mixed} case, the nested function captures the \ident{?width} 
+parameter available from the declaration site. As a result, the latter function can be called as follows:\
 %
@@
 \begin{array}{l}
-\kvd{let}~\ident{formatHello} = \\
-\quad(~\kvd{letdyn}~\ident{?width}=5~\kvd{in}\\
-\quad~~\ident{format}~\texttt{"Hello"})~~\kvd{in}\\
-\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}
-\end{array}
-\end{equation*}
-%
-This way of assigning type to \ident{format} and calling it is not the only possible, though. 
-We could also say that the outer function requires both of the implicit parameters and the result
-is a (pure) function with no context requirements. This interaction between implicit parameters 
-and lambda abstraction demonstrates one of the key aspects of coeffects and will be discussed 
-later. Implicit parameters will also sever as one of our examples in Chapter~Y.
+\kvd{let}~\ident{formatHello} = \\[-0.25em]
+\quad(~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
+\kvd{in}\,(~\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}~)
+\end{array}
+\end{equation*}
+%
+For different typings of \ident{format}, different ways of calling it are valid. This illustrates
+the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect systems may 
+introduce certain non-determinism in the typing. The following section shows how this looks in the
+type system for implicit parameters.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma}{\aclrd{\emptyset}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{param}
+  {}
+  {\coctx{\Gamma}{ \{\ident{?param}:\tau \} } \vdash \ident{?param} : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \subseteq \aclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\aclrd{r} \cup \aclrd{s} \cup \aclrd{t}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\aclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\aclrd{r \cup s}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
+\end{equation*}
+\vspace{-0.9em}
+\caption{Selected coeffect rules for implicit parameters}
+\label{fig:applications-flat-impl}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Type system.}
+
+Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's 
+implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau$
+meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set 
+of implicit parameters specified by $\aclrd{r}$. The annotations $\aclrd{r},\aclrd{s},\aclrd{t}$ are sets of 
+pairs consisting of implicit parameter names and types, \ie~$\aclrd{r},\aclrd{s},\aclrd{t} \subseteq 
+  \ident{Names} \times \ident{Types}$. The expressions include \ident{?param} to read implicit
+parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
+annotated with the set of implicit parameters that must be available on the call-site, \ie~
+$\tau_1 \xrightarrow{\aclrd{s}} \tau_2$.
+
+Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that 
+introduces primitive context requirements is (\emph{param}) -- accessing a parameter \ident{?param} 
+of type $\tau$ requires it to be available in the context. The context may provide more (unused) 
+implicit parameters thanks to the (\emph{sub}) rule.
+
+When we read the rules from the top to the bottom, application (\emph{app}) and let binding 
+(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
+(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
+the body $\aclrd{r} \cup \aclrd{s}$ can be freely split between the declaration-site ($\coctx{\Gamma}{\aclrd{r}}$)
+and the call-site ($\tau_1 \xrightarrow{\aclrd{s}} \tau_2$).
+
+The union operation $\cup$ is not a disjoint union, which means that the values for implicit 
+parameters can also be provided by both sites. For example, consider a function with a body
+$\ident{?a} + \ident{?b}$. Assuming that the function takes and returns $\ident{int}$, the following
+list shows 4 out of 9 possible valid typing:
+%
+\begin{equation*}
+\begin{array}{rcllllr}
+\coctx{\Gamma}{\aclrd{ \{ \ident{?a}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{ \{ \ident{?b}:\ident{int} \} } \ident{int} &\qquad\qquad&(1) \\
+\coctx{\Gamma}{\aclrd{ \{ \ident{?b}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{ \{ \ident{?a}:\ident{int} \} } \ident{int} &&(2)\\
+\coctx{\Gamma}{\aclrd{ \{\ident{?a}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} } \ident{int} &&(3)\\
+\coctx{\Gamma}{\aclrd{ \emptyset }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} } \ident{int} &&(4)
+\end{array}
+\end{equation*}
+%
+The first two examples demonstrate why the system does not have the principal typing property. 
+Both ($1$) and ($2$) are valid typings and they may both be desirable in certain contexts where
+the function is used.
+
+In ($3$), the parameter \ident{?a} has to be provided from both the declaration-site and call-site.
+We describe system that supports dynamic rebinding, meaning that when the caller provides a 
+value, it hides the value that may be available from the declaration-site. This means that 
+$4$ is a more precise typing modelling the same situation.
 
@@
 
-\subsection{Type classes}
-Implicit parameters are closely related to \emph{type classes} \cite{app-type-classes}. In Haskell,
-type classes provide a principled form of ad-hoc polymorphism (overloading). When a code uses 
-an overloaded operation (e.g.~comparison or numeric operators) a constraint is placed on the 
-context in which the operation is used. For example:
-%
-\begin{equation*}
+\begin{figure}
+
+{\small The semantics is defined inductively over the typing derivation:}
+\begin{equation}
+\tag{\emph{var}}
+\sem{\coctx{\Gamma}{\aclrd{r}} \vdash x_i : \tau_i} = \lambda ((x_1, \ldots, x_n), \_) \rightarrow x_i
+\end{equation}
+\vspace{-1.5em}
+\begin{equation}
+\tag{\emph{param}}
+\sem{\coctx{\Gamma}{\aclrd{r}} \vdash ?p : \sigma} = \lambda (\_, f) \rightarrow f~?p
+\end{equation}
+\vspace{-1.5em}
+\begin{equation}
+\tag{\emph{sub}}
+\sem{\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau} = \lambda (x, f) \rightarrow 
+  \sem{\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau}~(x, \restr{f}{\aclrd{r'}})
+\end{equation}
+\vspace{-1.5em}
+\begin{equation}
+\tag{\emph{abs}}
+\hspace{-0.5em}
 \begin{array}{l}
-\ident{twoTimes}~::~\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha \\
+  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\
+  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{r}\cup\aclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), f \uplus g)   	 
+\end{array}
+\end{equation}
+\vspace{-1.0em}
+\begin{equation*}
+\tag{\emph{app}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{r}\cup\aclrd{s}\cup\aclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 }~(x, \restr{f}{\aclrd{r}})\\
+  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }~(x, \restr{f}{\aclrd{s}} ), \restr{f}{\aclrd{t}})
+\end{array}
+\end{equation*}
+
+\vspace{1em}
+{\small Monadic semantics using the reader monads differs as follows:}
+\begin{equation}
+\tag{\emph{rdabs}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\
+  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{r}\cup\aclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)   	 
+\end{array}
+\end{equation}
+
+\vspace{1em}
+{\small Where $\uplus$ and $\restr{f}{\aclrd{r}}$ are auxiliary definitions:}
+\begin{equation*}
+\begin{array}{rcl}
+\restr{f}{\aclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \aclrd{r} \}\\
+f \uplus g &=& \restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g \\
+\end{array}
+\end{equation*}
+
+\caption{Semantics of a language with implicit parameters}
+\label{fig:applications-flat-implsem}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Semantics.}
+Implicit parameters can be implemented by passing around a hidden dictionary that provides values
+to the implicit parameters. Accessing a parameter then becomes a lookup in the dictionary and
+the \kvd{letdyn} construct extends the dictionary. To elucidate how such hidden dictionaries 
+are propagated through the program when using lambda abstractions and applications, we present a 
+simple semantics for implicit parameters. The goal here is not to prove properties of the language,
+but simply to provide a better explanation. A detailed semantics in terms of indexed comonads is
+shown in Chapter~\ref{ch:flat}.
+
+For simplicity, we assume that all implicit parameters have the same type $\sigma$. In that setting, 
+coeffect annotations $\aclrd{r}$ are just sets of names, \ie~$\aclrd{r}, \aclrd{s}, \aclrd{t} \subseteq \ident{Names}$.
+Given an expression $e$ of type $\tau$ that requires free variables $\Gamma$ and implicit parameters 
+$\aclrd{r}$, our interpretation is a function that takes a product of variables from $\Gamma$ together 
+with a hidden dictionary of implicit parameters and returns $\tau$:
+%
+\begin{equation*}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{r} } \vdash e : \tau} 
+  ~:~ (\tau_1 \times \ldots \times \tau_n) \times (\aclrd{r} \rightarrow \sigma) \rightarrow \tau
+\end{equation*}
+%
+The hidden dictionary is represented as a function from $\aclrd{r}$ to $\sigma$. This means that it 
+provides a $\sigma$ value for all implicit parameters that are required according to the typing.
+Note that the domain of the function is not the set of all possible implicit parameter names, but
+only a set of those that are specified by the type system.
+
+A hidden dictionary also needs to be attached to all functions. A function $\tau_1 \xrightarrow{\aclrd{s}} \tau_2$
+is interpreted by a function that takes $\tau_1$ together with a dictionary that defines values for
+implicit parameters in $\aclrd{s}$:
+%
+\begin{equation*}
+\begin{array}{l}
+\sem{\tau_1 \xrightarrow{\aclrd{s}} \tau_2} = \tau_1 \times (\aclrd{s} \rightarrow \sigma) \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+The definition of the semantics is shown in Figure~\ref{fig:applications-flat-implsem}. Let binding
+can be viewed as a syntactic sugar for $(\lambda x.e_2)~e_1$ and so it is omitted for brevity. The 
+(\emph{var}) and (\emph{param}) rules are simple -- they project the appropriate variable and 
+implicit parameter, respectively.
+
+When an expression requires implicit parameters $\aclrd{r}$, the semantics always provides a 
+dictionary defined \emph{exactly} on $\aclrd{r}$. To achieve this, the (\emph{sub}) rule restricts
+the function to $\aclrd{r'}$ (which is valid because $\aclrd{r'} \subseteq \aclrd{r}$).
+
+The most interesting rules are (\emph{abs}) and (\emph{app}). In abstraction, we get two dictionaries
+$f$ and $g$ (from the declaration-site and call-site, respectively), which are combined and passed 
+to the body of the function. The semantics prefers values from the call-site, which is captured by
+the $\uplus$ operation. In application, we first evaluate the expression $e_1$, then $e_2$ and finally
+call the returned function. The three calls use (possibly overlapping) restrictions of the dictionary
+as required by the static types.
+
+Without providing a proof here, we note that the semantics is sounds with respect to the type 
+system -- when evaluating an expression, it provides it with a dictionary that is guaranteed to
+contain values for all implicit parameters that may be accessed. This can be easily checked by
+examining the semantic rules (and noting that the restriction and union always provide the
+expected set of parameters).
+
+\paragraph{Monadic semantics.}
+Implicit parameters are related to the \emph{reader mo\-nad}. The type 
+$\tau_1 \times (\aclrd{r} \rightarrow \sigma) \rightarrow \tau_2$ is equivalent to
+$\tau_1 \rightarrow ((\aclrd{r} \rightarrow \sigma) \rightarrow \tau_2)$ through currying. Thus, we can
+express the function as $\tau_1 \rightarrow M\tau_2$ for $M\tau = (\aclrd{r} \rightarrow \sigma) \rightarrow \tau$.
+Indeed, the reader monad can be used to model dynamic scoping. However, there is an important distinction
+from implicit parameters. The usual monadic semantics models fully dynamic scoping, while implicit
+parameters combine lexical and dynamic scoping.
+
+The (\emph{rdabs}) rule in Figure~\ref{fig:applications-flat-implsem} shows a semantics that
+matches the usual monadic semantics using the reader monad. Note that the declaration-site dictionary
+is ignored and the body is called with only the dictionary provided by the call-site. This is
+a consequence of the fact that monadic functions are always pure values created using \emph{unit}.
+
+As we discuss later in Section~3.?, the reader monad can be extended to model rebinding.
+However, later examples in this chapter, such as liveness in Section~\ref{sec:applications-flat-live}
+show that other context-aware computations cannot be captured by \emph{any} monad.
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Type classes.}
+Another type of constraints in Haskell that is closely related to implicit parameters are 
+\emph{type class} constraints \cite{app-type-classes}. They provide a principled form of ad-hoc
+polymorphism (overloading). When a code uses an overloaded operation (e.g.~comparison or numeric 
+operators) a constraint is placed on the context in which the operation is used. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{twoTimes}~::~\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha \\[-0.25em]
 \ident{twoTimes}~x=x+x
@@
 %
-The constraint $\ident{Num}~a$ on the function type arises from the use of the $+$ operator. 
-From the implementation perspective, the type class constraint means that the function takes 
-a hidden parameter -- a dictionary that provides the operation $+ :: \alpha \rightarrow \alpha \rightarrow \alpha$.
-Thus, the type $\ident{Num}~\alpha \Rightarrow \alpha \rightarrow \alpha$ can be viewed as
-$(\ident{Num}_\alpha \times \alpha) \rightarrow \alpha$. Implicit parameters work in exactly
-the same way -- they are passed around as hidden parameters.
-
-The implementation of type classes and implicit parameters shows two important points about 
-context-dependent properties. First, they are associated with some \emph{scope}, such as the body
+The constraint $\ident{Num}~\alpha$ on the function type arises from the use of the $+$ operator. 
+Similarly to implicit parameters, type classes can be implemented using a hidden dictionary. In 
+the above case, the function \ident{twoTimes} takes a hidden dictionary that provides an operation
+$+$ of type $\alpha \times \alpha \rightarrow \alpha$.
+
+Type classes could be modelled as a coeffect system. The type system would annotate the context
+with a set of required type classes. The typing of the body of \ident{twoTimes} would look as 
+follows:
+%
+\begin{equation*}
+\coctx{x\!:\!\alpha}{ \{ \ident{Num}_\alpha \} } \vdash x + x : \alpha
+\end{equation*}
+%
+Similarly, the semantics of a language with type class constraints can be defined in a way
+similar to implicit parameters. The interpretation of the body is a function that takes $\alpha$
+together with a hidden dictionary of operations: $\alpha \times \ident{Num}_\alpha \rightarrow \alpha$.
+
+Type classes and implicit parameters show two important points about flat coeffect systems.
+First, the context requirements are associated with some \emph{scope}, such as the body
 of a function. Second, they are associated with the input. To call a function that takes an 
@@
 
-\subsection{Rebindable resources}
+\subsection{Distributed computing}
 \label{sec:applications-flat-distr}
 
-The need for parameters that do not strictly follow static scoping rules also arises in distributed
-computing. This problem has been addressed, for example, by Bierman et al. and Sewell et al. 
-\cite{app-distributed-rebinding,app-distributed-acute}. To quote the first work: \emph{``Dynamic 
+Distributed programming was used as one of the motivating examples for coeffects in 
+Chapter~\ref{ch:intro}. This section explores the use case. We look at rebindable resources and
+cross-compilation. The structure of both is very similar to implicit parameters and type
+class constraints, but they demonstrate that there is a broader use for coeffect systems.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Rebindable resources.}
+
+The need for parameters that support dynamic scoping also arises in distributed computing.
+To quote an example discussed by Bierman et al. \cite{app-distributed-rebinding}: \emph{``Dynamic 
 binding is required in various guises, for example when a marshalled value is received from the 
@@
 
-This situation arises when marshalling and transferring function values. A function may depend 
-on a local resource (e.g.~a database available only on the server) and also resources that are 
-available on the target node (e.g.~current time). In the following example, the construct
-$\kvd{access}~\ident{Res}$ represents access to a re-bindable resource named $\ident{Res}$:
+Rebindable parameters are identifiers that refer to some specific resource. When a function value
+is marshalled and sent to another machine, rebindable resources can be handled in two ways. 
+First, if the resource is available on the target machine, the parameter is \emph{rebound} to
+the resource on the new machine. This is captured by dynamic scoping rules. Second, if the 
+resource is not available on the target machine, the resource is either marshalled or a \emph{remote 
+reference} is created. This is captured by lexical scoping rules.
+
+A practical language that supports rebindable resources is for example Acute \cite{app-distributed-acute}.
+In the following example, we use the construct $\kvd{access}~\ident{Res}$ to represent
+access to a rebindable resource named $\ident{Res}$. The following simple function accessed
+a database and a current date and filters values based on the date:
 %
@@
 \begin{array}{l}
-\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\
-\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\
-\quad\ident{query}~\ident{db}~\str{"SELECT * WHERE Date > \%1"}~(\kvd{access}~\ident{Clock})
-\end{array}
-\end{equation*}
-%
-When \ident{recentEvents} is created on a server and sent to a client, a remote reference to the 
-database (available only on the server) must be captured. If the client device supports a clock, 
-then \ident{Clock} can be locally \emph{rebound}, e.g., to accommodate time-zone changes. 
+\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\[-0.25em]
+\quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\[-0.25em]
+\quad\ident{query}~\ident{db}~\str{SELECT * WHERE Date > \%1}~(\kvd{access}~\ident{Clock})
+\end{array}
+\end{equation*}
+%
+When \ident{recentEvents} is created on the server and sent to the client, a remote reference to 
+the database (available only on the server) must be captured. If the client device supports a 
+clock, then \ident{Clock} can be locally \emph{rebound}, e.g., to accommodate time-zone changes. 
 Otherwise, the date and time needs to be obtained from the server too.
 
-The use of re-bindable resources creates a context requirement similar to the one arising from
-the use of implicit parameters. For function values, such context-requirements can be satisfied
-in different ways -- resources must be available either at the declaration site (i.e.~when a 
-function is created) or at the call site (i.e.~when a function is called).
+The type system and semantics for rebindable resources are essentially the same as those for
+implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword. 
+Lambda abstraction splits the resources non-deterministically between declaration-site 
+(capturing remote reference) and call-site (representing rebinding). For this reason, we do not
+discuss the system in details and instead look at other uses.
 
@@
 
-\paragraph{Distributed computing and multi-targetting}
-
-An increasing number of programming languages is capable of running across multiple different 
-platforms or execution environments. Functional programming languages that can be compiled to
-JavaScript (to target web and mobile clients) include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.
-
-Links \cite{app-distributed-links}, F\# libraries \cite{app-fsharp-webapps,app-fsharp-webtools},
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\cmt{// Checks that input is valid; can run on both server and client}\\[-0.25em]
+\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\[-0.25em]
+\quad\ident{name} \neq \str{} ~~\&\&~~ \ident{forall~isLetter~name}
+\\[0.5em]
+\cmt{// Searches database for a product; can run on the server-side}\\[-0.25em]
+\kvd{let}~\ident{retrieveProduct} = \lambda \ident{name} \rightarrow\\[-0.25em]
+\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{Some}(\ident{queryProductDb~name})\\[-0.25em]
+\quad\kvd{else}~\ident{None}
+\\[0.5em]
+\cmt{// Client-side function to show price or error (for invalid inputs)}\\[-0.25em]
+\kvd{let}~\ident{showPrice} = \lambda \ident{name} \rightarrow\\[-0.25em]
+\quad\kvd{if}~\ident{validateInput name}~\kvd{then}\\[-0.25em]
+\quad\quad\kvd{match}~(\kvd{remote}~\ident{retrieveProduct}())~\kvd{with}\\[-0.25em]
+\quad\quad|~\ident{Some}~\ident{p} \rightarrow \ident{showPrice}~(\ident{getPrice}~\ident{p})\\[-0.25em]
+\quad\quad|~\ident{None}~\rightarrow \ident{showError}~\str{Invalid input on the server}\\[-0.25em]
+\quad\kvd{else}~\ident{showError}~\str{Invalid input on the client}
+\end{array}
+\end{equation*}
+
+\vspace{1em}
+\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
+\vspace{-1.5em}
+\caption{Sample client-server application with input validation}
+\label{fig:applications-flat-distr}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
+
+\paragraph{Cross-compilation.}
+A related issue with distributed programming is the need to target increasing number of diverse
+platforms. Modern applications often need to run on multiple platforms (iOS, Android, Windows or
+as JavaScript) or multiple versions of the same platform. Many programming languages are capable
+of targeting multiple different platforms. For example, functional languages that can be compiled 
+to native code and JavaScript include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.
+
+Links \cite{app-distributed-links}, F\# WebTools and WebSharper \cite{app-fsharp-webapps,app-fsharp-webtools},
 ML5 and QWeSST \cite{app-distributed-ml5, app-distributed-qwesst} and Hop \cite{app-hop-lang} go 
-further and allow a single source program to be compiled to multiple target runtimes. This posses 
-additional challenges -- it is necessary to track where each part of computation runs and statically 
-guarantee that it will be possible to compile code to the required target platform 
-(safe \emph{multi-targetting}).
-
-We demonstrate the problem by looking at input validation. In distributed applications 
-that communicate over unsecured HTTP channel, user input needs to be validated interactively
-on the client-side (to provide immediate response) and then again on the server-side (to 
-guarantee safety). For example:
-
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\
-\quad\ident{name} \neq \str{""} ~~\&\&~~ \ident{forall~isLetter~name}
-\\[0.5em]
-\kvd{let}~\ident{displayProduct} = \lambda \ident{name} \rightarrow\\
-\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{displayProductPage~name}\\
-\quad\kvd{else}~\ident{displayErrorPage}~() 
-\end{array}
-\end{equation*}
-%
-The function \ident{validateInput} can be compiled to both JavaScript (for client-side) and
-native code (for server-side). However, \ident{displayProduct} uses other functionality
-(generating web pages) that is only available on the server-side, so it can only be compiled to
-native code.
+further and allow including code for multiple distinct platforms in a single source file. 
+A single program is then automatically split and compiled to multiple target runtimes. This
+posses additional challenges -- it is necessary to check where each part of the program can run
+and statically guarantee that it will be possible to compile code to the required target 
+platform (safe \emph{multi-targetting}).
+
+We demonstrate the problem by looking at input validation. In applications that communicate over 
+unsecured HTTP channel, user input needs to be validated interactively on the client-side (to 
+provide immediate response) and then again on the server-side (to guarantee safety).
+
+Consider the client-server example in Figure~\ref{fig:applications-flat-distr}. The 
+\ident{retrieveProduct} function represents the server-side, while \ident{showPrice} is called 
+on the client-side and performs a remote call to the server-side function (how this is implemented 
+is not our concern here). To ensure that the input is valid \emph{both} functions call 
+\ident{validateInput} -- however, this is fine, because \ident{validateInput} uses only basic
+functions and language features that can be cross-compiled to both client-side and server-side.
 
 In Links \cite{app-distributed-links}, functions can be annotated as client-side, server-side
-and database-side. F\# WebTools \cite{app-fsharp-webtools} adds functions that support multiple
-targets (mixed-side). However, these are single-purpose language features and they are not 
-extensible. For example, in modern mobile development it is also important to track minimal 
-supported version of runtime\footnote{Android Developer guide \cite{app-android-multitarget} 
-demonstrates how difficult it is to solve the problem without language support.}.
-
-Requirements on the execution environment can be viewed as contextual properties, but could be
-also presented as effects (use of some API required only in certain environment is a computational
-effect). We discuss the difference in Section~X. Furthermore, the theoretical foundations of
-distributed languages like ML5 \cite{app-distributed-ml5} suggest that a contextual treatment
-is more appropriate. We return to ML5 when discussing semantics in Section~\ref{sec:path-sem-contextdep}.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Safe locking}
+and database-side. F\# WebTools \cite{app-fsharp-webtools} supports cross-compiled (mixed-side)
+functions similar to \ident{validateInput}. However, these are single-purpose language features 
+and they are not extensible. A practical implementation needs to be able to capture multiple
+different patterns -- sets of environments (client, server, mobile) for distributed computing,
+but also Android API level \cite{app-android-multitarget} to cross-compile for multiple versions 
+of the same platform.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Set based type system for cross-compilation, inspired by Links \cite{app-distributed-links}}
+
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \supseteq \aclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\aclrd{r} \cap \aclrd{s} \cap \aclrd{t}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
+\end{equation*}
+\vspace{0.5em}
+
+{\small b.) Version number based type system, inspired by Android API level \cite{app-android-multitarget}}
+
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \leq \aclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\ident{max} \{\aclrd{r}, \aclrd{s}, \aclrd{t} \}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 }
+\end{equation*}
+
+\caption{Two variants of coeffect typing rules for cross-compilation}
+\label{fig:applications-flat-cross}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Type systems.}
+
+Cross-compilation may seem similar to the tracking of resources (and thus to the tracking of implicit 
+parameters), but it actually demonstrates a couple of new ideas that are important for flat coeffect 
+systems. Unlike with implicit parameters, we will not present a specific existing system in this 
+section -- instead we briefly look at two examples that let us explore the range of possibilities.
+
+In the first system, shown in Figure~\ref{fig:applications-flat-cross} (a), the coeffect annotations
+are sets of execution environments, \ie~$\aclrd{r}, \aclrd{s}, \aclrd{t} \subseteq \{ \ident{client},
+\ident{server}, \ident{database} \}$. Sub-coeffecting (\emph{sub}) lets us ignore some of the supported 
+execution environments; application (\emph{app}) can be only executed in the \emph{intersection} of the
+environments required by the two expressions and the function value.
+
+Sub-coeffecting and application are trivially dual to the rules for implicit parameters. We just track
+supported environments using intersection as opposed to tracking of required parameters using union.
+However, this symmetry does not hold for lambda abstraction (\emph{abs}), which still uses \emph{union}.
+This models the case where the function can be executed in two different ways:
+%
+\begin{itemize}
+\item[--] The function is represented as executable code for an environment available at the call-site
+  and is executed there, possibly after it is marshalled and transferred to another machine.
+\vspace{-0.5em}
+\item[--] The function body is compiled for the environment available at the declaration-site; the value
+  that is returned is a remote reference to the code and function calls are performed as remote invocations.
+\end{itemize}
+%
+This example ignores important considerations -- for example, it is likely desirable to make this
+difference explicit and the implementation (and semantics) needs to be clarified. However, the
+example shows that the algebraic structure of coeffect annotations may be more complex. Here, using
+$\cap$ for application and $\cup$ for abstraction.
+
+The second system, shown in Figure~\ref{fig:applications-flat-cross} (b) is inspired by the API
+level requirements in Android. Coeffect annotations are simply numbers representing the level
+($\aclrd{r}, \aclrd{s}, \aclrd{t} \in \mathbb{N}$). Levels are ordered increasingly, so we can
+always require higher level (\emph{sub}). The requirement on function application (\emph{app}) is
+the highest level of the levels required by the sub-expressions and the function. The system uses 
+yet another variant of lambda abstraction (\emph{abs}) -- the requirements of the body are duplicated
+and placed on \emph{both} the declaration-site and the call-site.
+
+In addition to the work discussed already, ML5 \cite{app-distributed-ml5} is another important work 
+that looks at tracking of execution environments. It uses modalities of modal S4 to represent the 
+environment -- this approach is similar to coeffects, both from the practical perspective, but also 
+through deeper theoretical links. We return to this topic in Chapter~\ref{ch:meta}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[!b]
+\vspace{-2em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
+\quad
+\inference[(\emph{app})\;]
+  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
+    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
+  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\vspace{-1.5em}
+\caption{Selected coeffect rules for liveness analysis}
+\label{fig:example-need}
+\vspace{-1.2em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Liveness analysis.}
+\label{sec:applications-flat-live}
+
+Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
+whether it is definitely not needed (\emph{dead}). A compiler can remove
+bindings to dead variables as the result is never used.
+
+We start with a restricted analysis and briefly mention how to make it practical
+later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
+to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
+
+A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
+($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
+Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
+if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
+A dead context may be marked as live by
+letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
+
+The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
+\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
+The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
+The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
+$g$ ignores its input (regardless of the evaluation order of the underlying language).
+We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
+Secondly, a \emph{pointwise composition} passes the same
+argument to $g$ and $h$. The parameter is live if either the parameter
+of $g$ or $h$ is live ($r \sqcup s$). Application combines the
+two operations, so the context $\Gamma$ is live if it is needed by
+$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
+
+An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
+combines the context annotations using $\sqcap$. Thus, if the body uses some 
+variables, both the function argument and the context of the declaration-site are marked as live.
+
+Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
+In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
+indexed comonads (\S\ref{sec:comonads}), we
+model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
+$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
+(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
+value when $r=\ident{D}$).
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-1em}
+\begin{equation*}
+\inference[(\emph{var})\;]
+  {x : \tau \in \Gamma}
+  {\ctyp{0}{\Gamma} \vdash x : \tau }
+%
+\quad\quad
+% 
+\inference[(\emph{app})\;]
+  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
+   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
+  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\inference[(\emph{prev})\;]
+  {\ctyp{n}{\Gamma} \vdash e : \tau}
+  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
+%
+\quad\quad
+%
+\inference[(\emph{abs})\;]
+  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
+\end{equation*}
+\vspace{-1em}
+\caption{Selected coeffect rules for causal data flow}
+\label{fig:example-dataflow}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\vspace{-1em}
+\subsection{Efficient dataflow.}
+Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
+computations over streams. In \emph{causal} data flow, program may
+access past values -- in this setting, a function $\tau_1 \rightarrow
+\tau_2$ becomes a function from a list of historical values $[\tau_1]
+\rightarrow \tau_2$. A coeffect system here tracks how many past values
+to cache.
+
+Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
+number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
+The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
+additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
+
+The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
+adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
+for the second function); passing the context to two subcomputations requires the maximum
+number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
+needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
+call-site must each provide at least the number of past values required by the function body
+(the body may use variables coming from the declaration-site as well as the argument).
+
+The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
+\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
+$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
+(inefficient) unbounded lists of past elements. The above static analysis provides an
+approximation of the number of required past elements and so we use just fixed-length lists.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Safe locking}
 In the previous examples, the context provides additional values or functions that may be accessed
@@
 proposed to use monads, while Uustalu and Vene later used comonads \cite{app-dataflow-essence}.
-
-% ==================================================================================================
-
-\section{Flat coeffects (2.)}
-
-% --------------------------------------------------------------------------------------------------
-\subsection{Implicit parameters and resources.}
-\label{sec:coeffects-res}
-
-Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
-They can be used to parameterize a computation without propagating arguments explicitly through a 
-chain of calls and are part of the context in which expressions evaluate. As correctly expected
-\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
-computations follow a similar pattern, but we discuss implicit parameters for simplicity.
-
-The following function prints a number using implicit parameters
-\ident{?culture} (determining the decimal mark) and \ident{?format}
-(the number of decimal places):
-%
-\begin{equation*}
-\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
-\end{equation*}
-%
-Figure~\ref{fig:example-resources} shows a type-and-coeffect system
-tracking the set of an expression's implicit parameters. 
-For simplicity here, all implicit parameters have type $\rho$.
-
-Context requirements are created in (\emph{access}), while (\emph{var}) requires no
-implicit parameters; (\emph{app}) combines requirements of
-both sub-expressions as well as the latent requirements of the
-function. The (\emph{abs}) rule is where the example differs from
-effect systems.  Function bodies can access the union of the
-parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
-at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
-judgements for the above example are:
-%
-\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
-\begin{equation*}
-\begin{array}{rcl}
-\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
-\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
-\end{array}
-\end{equation*}
-%
-The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
-Different judgments are desirable depending on how a function is used. 
-In the first case, both parameters have to be provided by the caller.
-In the second, both are available at declaration-site, but \ident{?format} may be
-rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).
-
-Implicit parameters can be captured by the
-\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
-  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
-modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
-the next example cannot be structured by \emph{any} monad.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-1.4em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-%
-\inference[(\emph{app})\;]
-  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\vspace{-0.7em}
-\begin{equation*}
-\inference[(\emph{access})\;]
-  {}
-  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
-%
-\quad\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\vspace{-0.9em}
-\caption{Selected coeffect rules for implicit parameters}
-\label{fig:example-resources}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[!b]
-\vspace{-2em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
-\quad
-\inference[(\emph{app})\;]
-  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
-    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
-  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\vspace{-1.5em}
-\caption{Selected coeffect rules for liveness analysis}
-\label{fig:example-need}
-\vspace{-1.2em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\subsection{Liveness analysis.}
-
-Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
-whether it is definitely not needed (\emph{dead}). A compiler can remove
-bindings to dead variables as the result is never used.
-
-We start with a restricted analysis and briefly mention how to make it practical
-later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
-to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
-
-A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
-($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
-Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
-if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
-A dead context may be marked as live by
-letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
-
-The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
-\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
-The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
-The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
-$g$ ignores its input (regardless of the evaluation order of the underlying language).
-We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
-Secondly, a \emph{pointwise composition} passes the same
-argument to $g$ and $h$. The parameter is live if either the parameter
-of $g$ or $h$ is live ($r \sqcup s$). Application combines the
-two operations, so the context $\Gamma$ is live if it is needed by
-$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
-
-An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
-combines the context annotations using $\sqcap$. Thus, if the body uses some 
-variables, both the function argument and the context of the declaration-site are marked as live.
-
-Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
-In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
-indexed comonads (\S\ref{sec:comonads}), we
-model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
-$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
-(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
-value when $r=\ident{D}$).
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-1em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{0}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-% 
-\inference[(\emph{app})\;]
-  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\inference[(\emph{prev})\;]
-  {\ctyp{n}{\Gamma} \vdash e : \tau}
-  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
-%
-\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\vspace{-1em}
-\caption{Selected coeffect rules for causal data flow}
-\label{fig:example-dataflow}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\subsection{Efficient dataflow.}
-Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
-computations over streams. In \emph{causal} data flow, program may
-access past values -- in this setting, a function $\tau_1 \rightarrow
-\tau_2$ becomes a function from a list of historical values $[\tau_1]
-\rightarrow \tau_2$. A coeffect system here tracks how many past values
-to cache.
-
-Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
-number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
-The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
-additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
-
-The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
-adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
-for the second function); passing the context to two subcomputations requires the maximum
-number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
-needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
-call-site must each provide at least the number of past values required by the function body
-(the body may use variables coming from the declaration-site as well as the argument).
-
-The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
-\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
-$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
-(inefficient) unbounded lists of past elements. The above static analysis provides an
-approximation of the number of required past elements and so we use just fixed-length lists.
 

### 2014-04-16 text/Chapter03.tex
Flat coeffect systems in Chapter 03

@@
 \vspace{-0.9em}
-\caption{Selected coeffect rules for implicit parameters}
+\caption{Coeffect rules for tracking implicit parameters}
 \label{fig:applications-flat-impl}
@@
 $\ident{?a} + \ident{?b}$. Assuming that the function takes and returns $\ident{int}$, the following
-list shows 4 out of 9 possible valid typing:
+list shows 4 out of 9 possible valid typing. Full typing derivations can be found in Appendix~?:
 %
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\
+  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\[-0.25em]
   \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{r}\cup\aclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), f \uplus g)   	 
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}\cup\aclrd{s}\cup\aclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 }~(x, \restr{f}{\aclrd{r}})\\
+  \sem{\coctx{\Gamma}{\aclrd{r}\cup\aclrd{s}\cup\aclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 }~(x, \restr{f}{\aclrd{r}})\\[-0.25em]
   \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }~(x, \restr{f}{\aclrd{s}} ), \restr{f}{\aclrd{t}})
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\
+  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
   \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{r}\cup\aclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)   	 
@@
 \begin{array}{rcl}
-\restr{f}{\aclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \aclrd{r} \}\\
-f \uplus g &=& \restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g \\
+\restr{f}{\aclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \aclrd{r} \}\\[-0.25em]
+f \uplus g &=& \restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g 
 \end{array}
@@
 
-\begin{figure}[!b]
-\vspace{-2em}
-\begin{equation*}
-\inference[(\emph{var})\;]
+\subsection{Liveness analysis}
+\label{sec:applications-flat-live}
+
+\emph{Live variable analysis} (LVA) \cite{app-modern-compiler} is a standard technique in compiler 
+theory. It detects whether a free variable of an expression may be used by a program during its
+evaluation (it is \emph{live}) or whether it is definitely not needed (it is \emph{dead}). As an 
+optimization, compiler can remove bindings to dead variables as they are never accessed. Wadler 
+\cite{app-strictness-absecnce} describes the property of a variable that is dead as the 
+\emph{absence} of a variable.
+
+\paragraph{Flat liveness analysis.}
+In this section, we discuss a restricted form of liveness analysis. We do not track liveness of 
+\emph{individual} variables, but of the \emph{entire} variable context. This is not practically
+useful, but it provides interesting insight into how flat coeffects work. A per-variable liveness
+analysis can be captured using structural coeffects and is discussed in Section~\ref{sec:applications-struct-live}.
+Consider the following two examples:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
+\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
+\end{array}
+\end{equation*}
+%
+The body of the first function is just a constant $42$ and so the context of the body is marked
+as \emph{dead}. The parameter (call-site) of the function is not used and can also be marked as dead. 
+Similarly, no variables from the declaration-site are used and so they are also marked as dead.
+
+In contrast, the body of the second function accesses a variable \ident{value} and so the body 
+of the function is marked as \emph{live}. In the flat system, we do not track \emph{which} 
+variable was used and so we have to mark both the call-site and declaration-site as live (this will
+be refined in structural liveness system).
+
+\paragraph{Forward vs. backward \& may vs. must.}
+Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
+propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
+guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variables 
+to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look 
+at an example with conditionals:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
+\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
+\end{array}
+\end{equation*}
+%
+Liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
+\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
+\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
+only if it was used in both of the branches (this is sometimes called \emph{neededness}).
+
+The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
+interaction between contextual properties and certain language constructs such as conditionals.
+We discuss this in Section~\ref{sec:flat-conditionals}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+{\small a.) The operations of a two-point lattice $\mathcal{L} = \{\ident{L}, \ident{D}\}$ 
+where $\ident{D} \sqsubseteq \ident{L}$ are defined as:}
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{L} \sqcup \ident{L} &=& \ident{L}\\
+\ident{D} \sqcup \ident{L} &=& \ident{D}\\
+\end{array}
+\qquad
+\begin{array}{rcl}
+\ident{L} \sqcup \ident{D} &=& \ident{D}\\
+\ident{D} \sqcup \ident{D} &=& \ident{D}
+\end{array}
+\qquad
+\begin{array}{rcl}
+\ident{L} \sqcap \ident{L} &=& \ident{L}\\
+\ident{D} \sqcap \ident{L} &=& \ident{L}\\
+\end{array}
+\qquad
+\begin{array}{rcl}
+\ident{L} \sqcap \ident{D} &=& \ident{L}\\
+\ident{D} \sqcap \ident{D} &=& \ident{D}
+\end{array}
+\end{equation*}
+
+{\small b.) Sequential composition of (semantic) functions composes annotations using $\sqcup$:}
+\begin{equation*}
+\begin{array}{llll}
+f : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 \qquad\qquad&
+g : \tau_2 \xrightarrow{\aclrd{s}} \tau_3 \qquad\qquad&
+g \circ f : \tau_1 \xrightarrow{\aclrd{r} \sqcup \aclrd{s}} \tau_3 \qquad\qquad& ~
+\\[0.75em]
+f : \tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2 &
+g : \tau_2 \xrightarrow{\aclrd{\ident{L}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_3 & (1)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2 &
+g : \tau_2 \xrightarrow{\aclrd{\ident{L}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 & (2)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2 &
+g : \tau_2 \xrightarrow{\aclrd{\ident{D}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 & (3)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2 &
+g : \tau_2 \xrightarrow{\aclrd{\ident{D}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 & (4)
+\end{array}
+\end{equation*}
+
+{\small c.) Pointwise composition of (semantic) functions composes annotations using $\sqcap$:}
+\begin{equation*}
+\begin{array}{llll}
+f : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 \qquad\qquad&
+h : \tau_1 \xrightarrow{\aclrd{s}} \tau_3 \qquad\qquad&
+\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{r} \sqcap \aclrd{s}} \tau_2 \times \tau_3 \qquad&~
+\\[0.75em]
+f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 &
+h : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 \times \tau_3 & (1)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 &
+h : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 \times \tau_3 & (2)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 &
+h : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 \times \tau_3 & (3)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 &
+h : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 \times \tau_3 & (4)
+\end{array}
+\end{equation*}
+
+\caption{Liveness annotations with sequential and pointwise composition}
+\label{fig:applications-flat-livealg}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Type system.}
+A type system that captures whole-context liveness annotates the context with value of a 
+two-point lattice $\mathcal{L} = \{ \ident{L}, \ident{D} \}$. The annotation \ident{L} marks
+the context as \emph{live} and \ident{D} stands for a \emph{dead} context. 
+Figure~\ref{fig:applications-flat-livealg} (a) defines the ordering $\sqsubseteq$, meet $\sqcup$ and join 
+operations $\sqcap$ of the lattice.
+
+The typing rules for tracking whole-context liveness are shown in Figure~\ref{fig:applications-flat-liveness}.
+The language now includes constants $c:\tau \in \Delta$. Accessing a constant (\emph{const}) annotates
+the context as dead using \ident{D}. This contrasts with variable access (\emph{var}), which marks the
+context as live using \ident{L}. A dead context (definitely not needed) can be treated as live context
+(which may be used) using the (\emph{sub}) rule. This captures the \emph{may} nature of the analysis.
+
+The (\emph{app}) rule is best understood by discussing its semantics. The semantics uses 
+\emph{sequential composition} to compose the semantics of $e_2$ with the function obtained
+as the result of $e_1$. However, we need more than just sequential composition. The same input
+context is passed to the expression $e_1$ (in order to get the function value) and to the
+sequentially composed function (to evaluate $e_2$ followed by the function call). This is captured 
+by \emph{pointwise composition}.
+
+Consider first \emph{sequential composition} of (semantic) functions $f, g$ annotated with 
+$\aclrd{r}, \aclrd{s}$. The composed function $g \circ f$ is annotated with $\aclrd{r} \sqcup \aclrd{s}$
+as shown in Figure~\ref{fig:applications-flat-livealg} (b).
+The argument of the function $g \circ f$ is live only when the arguments of both $f$ and $g$ are 
+live ($1$). When the argument of $f$ is dead, but $g$ requires $\tau_2$ ($2$), we can evaluate 
+$f$ without any input and obtain $\tau_2$, which is then passed to $g$. When $g$ does not require
+its argument ($3, 4$), we can just evaluate $g$, without evaluating $f$. Here, the semantics
+\emph{implements} the dead code elimination optimization.
+
+Secondly, a \emph{pointwise composition} passes the same argument to $f$ and $h$. The parameter 
+is live if either the parameter of $f$ or $h$ is live. The pointwise composition is written as
+$\langle f, h \rangle$ and it combines annotations using $\sqcap$ as shown in Figure~\ref{fig:applications-flat-livealg} (c).
+Here, the argument is not needed only when both $f$ and $h$ do not need it ($1$). In all other cases,
+the parameter is needed and is then used either once ($2,3$) or twice ($4$). The rule for function
+application (\emph{app}) combines the two operations. The context $\Gamma$ is live if it is needed by
+$e_1$ (which always needs to be evaluated) \emph{or} when it is needed by the function value \emph{and}
+by $e_2$.
+
+The (\emph{abs}) rule duplicates the annotation of the body, similarly to the cross-compilation 
+example in Figure~\ref{fig:applications-flat-cross}. When the body accesses any variables, it 
+requires both the argument and the variables from declaration-site. When it does not use any variables,
+it marks both as dead. Finally, the (\emph{let}) rule annotates the composed expression with the
+liveness of the expression $e_2$ -- if the context of $e_2$ is live, then it also requires variables
+from $\Gamma$; if it is dead, then it does not require $\Gamma$ or $x$. 
+As further discussed later in Section~?, the (\emph{let}) rule is again just a syntactic sugar for 
+$(\lambda x.e_2)~e_1$. Briefly, this follows from the simple observation that 
+$\aclrd{r} \sqcup (\aclrd{s} \sqcap \aclrd{r}) = \aclrd{r}$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
   {x : \tau \in \Gamma}
-  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
-\quad
-\inference[(\emph{app})\;]
-  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
-    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
-  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\vspace{-1.5em}
-\caption{Selected coeffect rules for liveness analysis}
-\label{fig:example-need}
+  {\coctx{\Gamma}{\aclrd{\ident{L} }} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c : \tau \in \Delta}
+  {\coctx{\Gamma}{ \aclrd{\ident{D}} } \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \sqsubseteq \aclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\aclrd{r} \sqcup (\aclrd{s} \sqcap \aclrd{t})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\aclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\aclrd{s}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 }
+\end{equation*}
+\vspace{-0.9em}
+
+\caption{Coeffect rules for tracking whole-context liveness}
+\label{fig:applications-flat-liveness}
 \vspace{-1.2em}
@@
 
-\subsection{Liveness analysis.}
-\label{sec:applications-flat-live}
-
-Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
-whether it is definitely not needed (\emph{dead}). A compiler can remove
-bindings to dead variables as the result is never used.
-
-We start with a restricted analysis and briefly mention how to make it practical
-later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
-to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.
-
-A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
-($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
-Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
-if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
-A dead context may be marked as live by
-letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).
-
-The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
-\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
-The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
-The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
-$g$ ignores its input (regardless of the evaluation order of the underlying language).
-We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
-Secondly, a \emph{pointwise composition} passes the same
-argument to $g$ and $h$. The parameter is live if either the parameter
-of $g$ or $h$ is live ($r \sqcup s$). Application combines the
-two operations, so the context $\Gamma$ is live if it is needed by
-$e_1$ \emph{or} by the function value \emph{and} by $e_2$.
-
-An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
-combines the context annotations using $\sqcap$. Thus, if the body uses some 
-variables, both the function argument and the context of the declaration-site are marked as live.
-
-Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
-In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
-indexed comonads (\S\ref{sec:comonads}), we
-model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
-$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
-(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
-value when $r=\ident{D}$).
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-1em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{0}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-% 
-\inference[(\emph{app})\;]
-  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\inference[(\emph{prev})\;]
-  {\ctyp{n}{\Gamma} \vdash e : \tau}
-  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
-%
-\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\vspace{-1em}
-\caption{Selected coeffect rules for causal data flow}
-\label{fig:example-dataflow}
-\vspace{-1em}
+\paragraph{Examples.}
+Before looking at the semantics, we consider a number of simple examples to demonstrate the
+key aspects of the system. Full typing derivations are shown in Appendix~?:
+%
+\begin{equation*}
+\begin{array}{lll}
+(\lambda x \rightarrow 42)~y &~\hspace{1em}~&(1)\\
+\ident{twoTimes}~42          &&(2)\\
+(\lambda x \rightarrow x)~42 &&(3)\\
+\end{array}
+\end{equation*}
+%
+In the first case, the context is dead. In ($1$), the function's parameter is dead and so the
+overall context is dead, even though the argument uses a variable $y$ -- the semantics evaluates 
+the function without passing it an actual argument. In the second case ($2$), the function is
+a variable that needs to be obtained and so the context is live. In the last case ($3$), the 
+function accesses a variable and so its declaration-site is marked as requiring the context 
+(\emph{abs}). This is where structural coeffect analysis would be more precise -- the system shown
+here cannot capture the fact that $x$ is a bound variable.
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation}
+\tag{\emph{var}}
+\sem{\coctx{\Gamma}{\aclrd{ \ident{L} }} \vdash x_i : \tau_i} = \lambda (x_1, \ldots, x_n) \rightarrow x_i
+\end{equation}
+\vspace{-1.75em}
+\begin{equation}
+\tag{\emph{const}}
+\sem{\coctx{\Gamma}{\aclrd{ \ident{D} }} \vdash c_i : \tau_i} = \lambda () \rightarrow \delta(c_i)
+\end{equation}
+%
+%
+\vspace{-0.5em}
+%
+%
+\begin{equation}
+\tag{\emph{sub-1}}
+\sem{\coctx{\Gamma}{\aclrd{\ident{L} }} \vdash e : \tau} = \lambda x \rightarrow 
+  \sem{\coctx{\Gamma}{\aclrd{\ident{D} }} \vdash e : \tau}~()
+\end{equation}
+\vspace{-1.75em}
+\begin{equation}
+\tag{\emph{sub-2}}
+\sem{\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau} = \lambda x \rightarrow 
+  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau}~x
+\end{equation}
+%
+%
+\vspace{-0.5em}
+%
+%
+\begin{equation}
+\tag{\emph{abs-1}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{\ident{L}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2 } = \lambda (x_1, \ldots, x_n) \rightarrow\\[-0.25em]
+  \qquad\lambda y \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{\ident{L}}} \vdash e : \tau_2 }~(x_1, \ldots, x_n, y)   	 
+\end{array}
+\end{equation}
+\vspace{-1.0em}
+\begin{equation}
+\tag{\emph{abs-2}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{\ident{D}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2 } = \lambda () \rightarrow\\[-0.25em]
+  \qquad\lambda () \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{\ident{D}}} \vdash e : \tau_2 }~()
+\end{array}
+\end{equation}
+%
+%
+\vspace{0.0em}
+%
+%
+%
+\begin{equation*}
+\tag{\emph{app-1}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 }~x
+    ~\kvd{in}~g~()
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\tag{\emph{app-2}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{\ident{L}}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{ \ident{L} }} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 }~x
+    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{ \ident{D} }} \vdash e_2 : \tau_1 }~())
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\tag{\emph{app-3}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{r} \sqcup (\aclrd{s} \sqcap \aclrd{t}) } \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 }~x
+    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }~x)
+\end{array}
+\end{equation*}
+
+\caption{Semantics of a language with liveness analysis}
+\label{fig:applications-flat-livsem}
 \end{figure}
@@
 
-\vspace{-1em}
-\subsection{Efficient dataflow.}
-Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
-computations over streams. In \emph{causal} data flow, program may
-access past values -- in this setting, a function $\tau_1 \rightarrow
-\tau_2$ becomes a function from a list of historical values $[\tau_1]
-\rightarrow \tau_2$. A coeffect system here tracks how many past values
-to cache.
-
-Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
-number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
-The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
-additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.
-
-The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
-adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
-for the second function); passing the context to two subcomputations requires the maximum
-number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
-needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
-call-site must each provide at least the number of past values required by the function body
-(the body may use variables coming from the declaration-site as well as the argument).
-
-The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
-\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
-$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
-(inefficient) unbounded lists of past elements. The above static analysis provides an
-approximation of the number of required past elements and so we use just fixed-length lists.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Safe locking}
-In the previous examples, the context provides additional values or functions that may be accessed
-at runtime. However, it may also track \emph{permissions} to perform some operation. This is done
-in the type system for safe locking of Flanagan and Abadi \cite{app-safe-locking}.
-
-The system prevents race conditions (by only allowing access to mutable state under a lock)
-and avoids deadlocks (by imposing strict partial order on locks). The following 
-program uses a mutable state under a lock:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{newlock}~l:\rho~\kvd{in}\\
-\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\
-\kvd{sync}~l~(!\ident{state})
-\end{array}
-\end{equation*}
-%
-The declaration \kvd{newlock} creates a lock $l$ protecting memory region $\rho$. We can than
-allocate mutable variables in that memory region (second line). An access to mutable variable
-is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
-which locks a lock and evaluates an expression in a context that contains permission to access
-memory region of the lock ($\rho$ in the above example).
-
-The type system for safe locking associates the list of permission with the variable context.
-It uses judgements of a form $\Gamma, m \vdash e : \alpha$ specifying that an expression has a
-type in context $\Gamma$, given permissions (a list of locked regions) $m$. However, the treatment
-of lambda abstraction differs from the one for implicit parameters or rebindable resources.
-In the system for locking, code inside lambda function cannot use permissions from the scope
-where the function is declared. This is a necessary requirement -- a lambda function created 
-under a lock cannot access protected memory, because it will be executed later. We discuss how
-this restriction fits into our general coeffect framework in Section~X.Y.
-
-%
-% => This has coeffect style judgments, but it has effect-style lambda
-%
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Data-flow languages}
-
-The examples discussed so far are all -- to some extent -- similar. They attach additional 
-information (implicit parameters, dictionaries) or restrictions (on execution environment) to the
-context where code evaluates. By \emph{context}, we mean, most importantly, the values of variables
-and declarations that are in scope. The examples so far add more information to the context, but
-do not operate on the variable values.
-
-Data-flow languages provide a different example. Lucid \cite{app-lucid} is a declarative data-flow 
-language designed by Wadge and Ashcroft. In Lucid, variables represent streams and programs
-are written as transformations over streams. A function application $\mathit{square}(a)$ represents
-a stream of squares calculated from the stream of values $a$.
+\paragraph{Semantics.} 
+The type system presented above requires the semantics to \emph{implement} dead code elimination.
+This means that when a function does not require an input (it is marked as dead), the semantics
+does not evaluate the argument and passes an empty value as the input instead.
+
+We can represent such empty values using the option type (known as \ident{Maybe} in Haskell).
+We use the notation $\tau + 1$ to denote option types. Given a context with variables $x_i$ of
+type $\tau_i$, the semantics is a function taking $(\tau_1 \times \ldots \times \tau_n) + 1$.
+When the context is live, it will be called with the left value (product of variable assignments);
+when the context is dead, it will be called with the right value (containing no information).
+
+However, ordinary option type is not sufficient. We need to capture the fact that the 
+representation depends on the annotation -- in other words, the type is \emph{indexed} by 
+the coeffect annotation. The indexing is discussed in details in Section~X. For now, it suffices
+to define the semantics using two separate rules:
+%
+\begin{equation*}
+\begin{array}{rlrcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{\ident{L}} } \vdash e : \tau} 
+  &:& (\tau_1 \times \ldots \times \tau_n) &\narrow{\rightarrow}& \tau\\
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{\ident{D}} } \vdash e : \tau} 
+  &:& 1 &\narrow{\rightarrow}& \tau
+\end{array}
+\end{equation*}
+%
+The semantics of functions is defined similarly. When the argument of a function is live, the function 
+takes the input value; when the argument is dead, the semantic function takes a unit as its argument:
+%
+\begin{equation*}
+\begin{array}{l}
+\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2\\
+\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+Unlike with implicit parameters, the coeffect system for liveness tracking cannot be modelled 
+using monads. Any monadic semantics would express functions as $\tau_1 \rightarrow M\, \tau_2$.
+Unless laziness is already built-in in the semantics, there is no way to call such function without
+first obtaining a value $\tau_1$. The above semantics makes this possible by taking a unit $1$ when
+the argument is not live.
+
+In Figure~\ref{fig:applications-flat-livsem}, we define the semantics directly. We write $()$ for
+the only value of type $1$. This appears, for example, in (\emph{const}) which takes $()$ as the
+input and returns constant using a global dictionary $\delta$. In (\emph{var}), the context is live
+and so the semantics performs a projection. Sub-coeffecting is captured by two rules. A dead context 
+can be treated as live using (\emph{abs-1}); in other cases, the annotation is not changed (\emph{abs-2}).
+
+Lambda abstraction can be annotated in just two ways. When the body requires context (\emph{abs-1}),
+the value of a bound variable $y$ is added to the context $\Gamma$ before passing it to the body.
+When the body does not require context (\emph{abs-2}), it is called with $()$ as the input.
+
+For application, there are 8 possible combinations of annotations. The semantics of some of them
+is the same, so we only need to show 3 cases. The rules should be read as ML-style pattern matching,
+where the last rule handles all cases not covered by the first two. In (\emph{app-1}), we handle the
+case when the function $g$ does not require its argument -- $e_2$ is not used and instead, the function
+is called with $()$ as the argument. The case (\emph{app-2}) covers the case when the expression
+$e_1$ does not require a context, but $e_1$ does. Finally, in (\emph{app-3}), the same input
+(which may be either tuple of variables or unit) is propagated uniformly to both $e_1$ and $e_2$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Data-flow languages}
+The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an 
+operation that accesses a context. In case of arrays, the context is neighbourhood of a current
+location in the array specified by a cursor. In this section, we make the example more concrete,
+using a simpler and better studied programming model, data-flow languages.
+
+Lucid \cite{app-lucid} is a declarative data-flow language designed by Wadge and Ashcroft. In Lucid, 
+variables represent streams and programs are written as transformations over streams. A function 
+application $\ident{square}(x)$ represents a stream of squares calculated from the stream of values $x$.
 
@@
 \\[0.5em]
-\kvd{let}~\ident{edgeCount} = \\
-\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~\kvd{prev}~\ident{edgeCount}\\
+\kvd{let}~\ident{edgeCount} = \\[-0.25em]
+\quad 0~\kvd{fby}~ (~\kvd{if}~\ident{edge}~\kvd{then}~1 + (\kvd{prev}~\ident{edgeCount})\\[-0.25em]
 \quad\quad\quad~~~\, \kvd{else}~\kvd{prev}~\ident{edgeCount} ~)
@@
 first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
-such as \ident{false} and $0$ are constant streams. Formally, the construct are defined as follows
+such as \ident{false} and $0$ are constant streams. Formally, the constructs are defined as follows
 (writing $x_n$ for $n$-th element of a stream $x$):
@@
 When reading data-flow programs, we do not need to think about variables in terms of streams --
-we can see them as simple values. However, the operations \kvd{fby} and \kvd{prev} cannot operate
-on plain values -- they require additional \emph{context} which provides past values of variables
+we can see them as simple values. Most of the operations perform calculation just on the 
+\emph{current} value of the stream. However, the operation \kvd{fby} and \kvd{prev} are different.
+They require additional \emph{context} which provides past values of variables
 (for \kvd{prev}) and information about the current location in the stream (for \kvd{fby}).
 
-In this case, the context is not simply an additional (hidden) parameter. It completely changes
-how variables must be represented. We may want to capture various \emph{contextual properties}
-of Lucid programs. For example, how many past elements need to be cached when we evaluate the 
-stream.
-
-To understand the nature of the context, we later look at the semantics of Lucid. This can be
-captured using a number of mathematical structures. Wadge \cite{app-lucid-monads} originally 
-proposed to use monads, while Uustalu and Vene later used comonads \cite{app-dataflow-essence}.
+The semantics of Lucid-like languages can be captured using a number of mathematical 
+structures. Wadge \cite{app-lucid-monads} originally proposed to use monads, while Uustalu and 
+Vene later used comonads \cite{app-dataflow-essence}. In Chapter~\ref{ch:flat}, we extend
+the latter approach. However, the present chapter presents a sketch of a data-flow semantics
+defined directly on streams.
+
+In the introductory example with array access patterns, we used coeffects to track the range
+of values accessed. In this section, we look at a simpler example -- we only consider the
+\kvd{prev} operation and track the maximal number of \emph{past values} needed. This is an 
+important information for efficient implementation of data-flow languages. When we can guarantee
+that at most $x$ past values are accessed, the values can be stored in a pre-allocated buffer
+rather than using \eg~on-demand computed lazy streams.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{prev}
+  {\coctx{\Gamma}{n} \vdash e : \tau}
+  {\coctx{\Gamma}{n+1} \vdash \kvd{prev}~e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
+   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
+\end{equation*}
+
+\caption{Coeffect rules for tracking context-usage in data-flow language}
+\label{fig:applications-flat-dataflow}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Type system.}
+A type system that tracks the maximal number of accessed past values annotates the context with
+a single integer. The current value is always present, so $0$ means that no past values are 
+needed, but the current value is still available. The typing rules of the system are shown
+in Figure~\ref{fig:applications-flat-dataflow}.
+
+Variable access (\emph{var}) annotates the context with $0$; sub-coeffecting (\emph{sub}) allows
+us to require more values than is actually needed. Primitive context-requirements are introduced
+in (\emph{prev}), which increments the number of past values by one. Thus, for example, 
+$\kvd{prev}~(\kvd{prev}~x)$ requires 2 past values.
+
+The (\emph{app}) rule follows the same intuition as for liveness. It combines \emph{sequential} 
+and \emph{pointwise} composition of semantic functions. In case of data-flow, the operations
+combine annotations using $+$ and \emph{max} operations:
+%
+\begin{equation*}
+\begin{array}{lll}
+f : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 \qquad\qquad&
+g : \tau_2 \xrightarrow{\aclrd{n}} \tau_3 \qquad\qquad&
+g \circ f : \tau_1 \xrightarrow{\aclrd{m} + \aclrd{n}} \tau_3 \qquad\qquad
+\\
+f : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 \qquad\qquad&
+h : \tau_1 \xrightarrow{\aclrd{n}} \tau_3 \qquad\qquad&
+\langle f, h \rangle : \tau_1 \xrightarrow{\textnormal{max}(\aclrd{m}, \aclrd{s})} \tau_2 \times \tau_3 \qquad
+\end{array}
+\end{equation*}
+%
+Sequential composition adds the annotations. The function $f$ needs $\aclrd{m}$ past values to 
+produce a single $\tau_2$ value. To produce two $\tau_2$ values, we thus need $\aclrd{m}+1$ past
+values of $\tau_1$; to produce three $\tau_2$ values, we need $\aclrd{m}+2$ past values of $\tau_1$,
+and so on. To produce $\ident{n}$ past values that are required as the input of $g$, we need
+$\aclrd{m}+\aclrd{n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
+the same stream to evaluate functions requiring $\aclrd{m}$ and $\aclrd{n}$ past values, and so it
+needs maximum of the two at most.
+
+In summary, function application (\emph{app}) requires maximum of the values needed to evaluate 
+$e_1$ and the number of values needed to evaluate the argument $e_2$, sequentially composed with
+the function.
+
+In function abstraction (\emph{abs}), the requirements of the body are duplicated on the declaration-site
+and the call-site as in liveness analysis. If the body requires $\aclrd{n}$ past values, it may access
+$\aclrd{n}$ values of any variables -- including those available in $\Gamma$, as well as the parameter
+$x$. Finally, the (\emph{let}) rule simply adds the two requirements. This corresponds to the sequential
+composition operation, but it is also a rule that we obtain by treating let-binding as a syntactic 
+sugar for $(\lambda x.e_2)~e_1$.
+
+\paragraph{Example.} 
+In the following example, \ident{counter} generates positive integers 
+and \ident{tick} flips between $0$ and $1$:
+%
+\begin{equation*}
+\begin{array}{l}
+(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0~\kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\
+\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\quad(\kvd{prev}~\ident{counter})
+\end{array}
+\end{equation*}
+%
+The function on the left-hand side returns the previous value of the argument every other 
+step. Applied to a stream $\langle{\ldots,4,3,2,1}\rangle$ (where $1$ is the ``current''
+value) it yields the stream $\langle{\ldots,4,4,2,2}\rangle$. To obtain the function, 
+we need one past value from the context (for $\kvd{prev}~\ident{tick}$). The returned function 
+needs either none or one past value (thus a subtyping rule is required to type it
+as requiring one past value).
+
+The function is called with $\kvd{prev}~\ident{counter}$ as an argument, meaning that the result
+is either the first or second past element. Given 
+$\ident{counter}\hspace{-0.1em}=\hspace{-0.1em}\langle{\ldots,5,4,3,2,1}\rangle$, the argument 
+is $\langle{\ldots,5,4,3,2}\rangle$ and so the result is a stream $\langle{\ldots,5,5,3,3}\rangle$.
+
+How many past values does the whole expression need? The argument requires one past value and
+the function one additional ($\kvd{prev}~(\kvd{prev}~\ident{counter}))$, adding to $2$. To obtain
+the function, one past value is needed. Overall we need the maximum of what is needed to evaluate the 
+function and what is needed by passing the argument to the function: $max(1, 2)=2$. This is 
+captured by (\emph{app}) in Figure~\ref{figure:intro-dataflow}.
+
+The (\emph{fun}) rule again does not allow delaying of context requirements. If the body 
+requires $m$ past values, these have to be available both from the outer context and from the
+argument provided by the caller. Later, we make the rule more flexible to make it compatible
+with (\emph{fun}) for implicit parameters that \emph{splits} requirements.
+
+\paragraph{Semantics.}
+In this setting, a function $\tau_1 \rightarrow
+\tau_2$ becomes a function from a list of historical values $[\tau_1]
+\rightarrow \tau_2$. A coeffect system here tracks how many past values
+to cache.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Safe locking}
+In the previous examples, the context provides additional values or functions that may be accessed
+at runtime. However, it may also track \emph{permissions} to perform some operation. This is done
+in the type system for safe locking of Flanagan and Abadi \cite{app-safe-locking}.
+
+The system prevents race conditions (by only allowing access to mutable state under a lock)
+and avoids deadlocks (by imposing strict partial order on locks). The following 
+program uses a mutable state under a lock:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{newlock}~l:\rho~\kvd{in}\\
+\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\
+\kvd{sync}~l~(!\ident{state})
+\end{array}
+\end{equation*}
+%
+The declaration \kvd{newlock} creates a lock $l$ protecting memory region $\rho$. We can than
+allocate mutable variables in that memory region (second line). An access to mutable variable
+is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
+which locks a lock and evaluates an expression in a context that contains permission to access
+memory region of the lock ($\rho$ in the above example).
+
+The type system for safe locking associates the list of permission with the variable context.
+It uses judgements of a form $\Gamma, m \vdash e : \alpha$ specifying that an expression has a
+type in context $\Gamma$, given permissions (a list of locked regions) $m$. However, the treatment
+of lambda abstraction differs from the one for implicit parameters or rebindable resources.
+In the system for locking, code inside lambda function cannot use permissions from the scope
+where the function is declared. This is a necessary requirement -- a lambda function created 
+under a lock cannot access protected memory, because it will be executed later. We discuss how
+this restriction fits into our general coeffect framework in Section~X.Y.
+
+%
+% => This has coeffect style judgments, but it has effect-style lambda
+%
 
@@
 look at applications in security and software updating.
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Liveness analysis}
-
-\emph{Live variable analysis} (LVA) \cite{app-modern-compiler} is a standard technique in compiler theory. 
-It detects whether a free variable of an expression may be used by a program later (it is
-\emph{live}) or whether it is definitely not needed (it is \emph{dead}). As an optimization, 
-compiler can remove bindings to dead variables as the result is never accessed. Wadler 
-\cite{app-strictness-absecnce} describes the property of a variable that is dead as the 
-\emph{absence} of a variable.
-
-In this thesis, we first use a restricted (and not practically useful) form of liveness analysis
-to introduce the theory of indexed comonads (Section~X) and then use liveness analysis as one of the
-motivations for structural coeffects. Consider the following two simple functions:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{constant42} = \lambda \ident{x} \rightarrow 42\\
-\kvd{let}~\ident{constant} = \lambda \ident{value} \rightarrow \lambda \ident{x} \rightarrow \ident{value}
-\end{array}
-\end{equation*}
-%
-In liveness analysis, we annotate the context with a value specifying whether the variables in
-scope are \emph{live} or \emph{dead}. If we associate just a single value with the entire 
-context, then the liveness analysis is very limited -- it can say that the context of the 
-expression $42$ in the first function is dead, because no variables are accessed.
-
-A useful liveness analysis needs to consider individual variables. For example, in the body of
-the second function (\ident{value}), two variables are in scope. The variable \ident{value} is
-accessed and thus is \emph{live}, but the variable \ident{x} is dead.
-
-Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they 
-propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
-guarantee). Liveness is a \emph{backward} analysis -- this means that the requirements propagates
-from variables to their declaration sites. The distinction between \emph{must} and \emph{may} is 
-apparent when we look at an example with conditionals:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
-\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
-\end{array}
-\end{equation*}
-%
-The liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
-\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
-\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
-only if it was used in both of the branches (this is sometimes called \emph{neededness}).
-
-The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
-interaction between contextual properties and certain language constructs such as conditionals.
 
@@
 % --------------------------------------------------------------------------------------------------
+
+\subsection{Liveness}
+\label{sec:applications-struct-live}
+
+todo
 
@@
 
-\section{Context oriented programming}
-
-The importance of context-aware computations is perhaps most obvious when considering mobile
-application, client/server web applications or even the internet of things. A pioneering work
-in the area using functional languages has been done by Serrano \cite{app-hop-diffuse,app-hop-lang}
-(which also inspired the example presented in Chapter~\ref{ch:intro}). His HOP language supports 
-cross-compilation and programs execute in different contexts. However, HOP is not statically 
-type checked.
-
-In the software engineering community, a number of authors have addressed the
-problem of context-aware computations. Hirschfeld et al. propose \emph{Context-Oriented Programming} 
-(COP) as a methodology \cite{app-cop-method}. The COP paradigm has been later implemented by
-programming language features. Costanza \cite{app-cop-contextl} develops a domain-specific LISP-like 
-language ContextL and Bardram \cite{app-cop-javafwk} proposes a Java framework for COP.
-
-Finally, the subject of context-awareness has also been addressed in work focusing on the development 
-of mobile applications \cite{app-cop-mobile,app-cop-mobile2}. Here, the \emph{context} focuses more 
-on concrete physical context (obtained from the device sensors) than context as an abstract 
-language feature.
-
-We approach the problem from a different perspective, building on the tradition of 
-statically-typed functional programming languages and their theories.
-
-%===================================================================================================
-
 \section{Summary}

### 2014-04-16 text/Chapter03.tex
Finish flat dataflow in chapter 3

@@
 \tyrule{prev}
-  {\coctx{\Gamma}{n} \vdash e : \tau}
-  {\coctx{\Gamma}{n+1} \vdash \kvd{prev}~e : \tau}
+  {\coctx{\Gamma}{ \aclrd{n} } \vdash e : \tau}
+  {\coctx{\Gamma}{ \aclrd{n}+1} \vdash \kvd{prev}~e : \tau}
 \end{equation*}
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \paragraph{Example.} 
-In the following example, \ident{counter} generates positive integers 
-and \ident{tick} flips between $0$ and $1$:
-%
-\begin{equation*}
-\begin{array}{l}
-(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0~\kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\
-\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\quad(\kvd{prev}~\ident{counter})
-\end{array}
-\end{equation*}
-%
-The function on the left-hand side returns the previous value of the argument every other 
-step. Applied to a stream $\langle{\ldots,4,3,2,1}\rangle$ (where $1$ is the ``current''
-value) it yields the stream $\langle{\ldots,4,4,2,2}\rangle$. To obtain the function, 
-we need one past value from the context (for $\kvd{prev}~\ident{tick}$). The returned function 
-needs either none or one past value (thus a subtyping rule is required to type it
-as requiring one past value).
-
+As with the liveness example, the application rule might require more explanation. The following 
+example is somewhat arbitrary, but it demonstrates the rule well. We assume that \ident{counter} 
+is a stream of positive integers (starting from zero) and \ident{tick} flips between $0$ and $1$.
+The full typing derivation is shown in Appendix~?:
+%
+\begin{equation*}
+\begin{array}{l}
+(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
+\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
+\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
+\end{array}
+\end{equation*}
+%
+The left-hand side of the application returns a function depending on the \emph{previous}
+value of \ident{tick}. The resulting stream of functions flips between a function returning
+a current value and a function returning the previous value. If the current \ident{tick} is 0, and
+the function is applied to a stream $\langle{\ldots,4,3,2,1}\rangle$ 
+(where $1$ is the current value), it yields the stream $\langle{\ldots,4,4,2,2}\rangle$.
+
+To obtain the function, we need one past value from the context (for $\kvd{prev}~\ident{tick}$). The 
+returned function needs either none or one past value (thus a subtyping rule is required to type 
+it as requiring one past value). So, the annotations for (\emph{app}) are $\aclrd{m}=1, \aclrd{p}=1$.
 The function is called with $\kvd{prev}~\ident{counter}$ as an argument, meaning that the result
@@
 $\ident{counter}\hspace{-0.1em}=\hspace{-0.1em}\langle{\ldots,5,4,3,2,1}\rangle$, the argument 
-is $\langle{\ldots,5,4,3,2}\rangle$ and so the result is a stream $\langle{\ldots,5,5,3,3}\rangle$.
-
-How many past values does the whole expression need? The argument requires one past value and
-the function one additional ($\kvd{prev}~(\kvd{prev}~\ident{counter}))$, adding to $2$. To obtain
-the function, one past value is needed. Overall we need the maximum of what is needed to evaluate the 
-function and what is needed by passing the argument to the function: $max(1, 2)=2$. This is 
-captured by (\emph{app}) in Figure~\ref{figure:intro-dataflow}.
-
-The (\emph{fun}) rule again does not allow delaying of context requirements. If the body 
-requires $m$ past values, these have to be available both from the outer context and from the
-argument provided by the caller. Later, we make the rule more flexible to make it compatible
-with (\emph{fun}) for implicit parameters that \emph{splits} requirements.
+is $\langle{\ldots,5,4,3,2}\rangle$ and so the overall result is a stream $\langle{\ldots,5,5,3,3}\rangle$.
+From the argument, we get the requirement $\aclrd{n}=1$.
+
+Using the (\emph{app}) rule, we get that the overall number of past elements needed is
+$\mathit{max}(1, 1+1) = 2$. This should match the intuition about the code -- when the first function
+is applied to the argument, the computation will first access $\kvd{prev}~\ident{tick}$ (using one
+past value) and then $\kvd{prev}~(\kvd{prev}~\ident{counter}))$ (using two past values).
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation}
+\tag{\emph{var}}
+\hspace{-1em}
+\sem{\coctx{\Gamma}{\aclrd{0}} \vdash x_i : \tau_i} = \lambda \langle(x_0, \ldots, x_n)\rangle \rightarrow x_i
+\end{equation}
+\vspace{-1.0em}
+\begin{equation}
+\tag{\emph{prev}}
+\hspace{-1.5em}
+\begin{array}{l}
+\sem{\coctx{\Gamma}{\aclrd{n}+1} \vdash \kvd{prev}~e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}+1} \rangle \rightarrow\\[-0.1em]
+  \qquad\sem{ \coctx{\Gamma}{ \aclrd{n} } \vdash e : \tau }~\langle \mathbf{v}_1, \ldots, \mathbf{v}_{ \aclrd{n}+1 }\rangle
+\end{array}
+\end{equation}
+\vspace{-0.5em}
+\begin{equation}
+\tag{\emph{sub}}
+\hspace{-1.5em}
+\begin{array}{l}
+\sem{\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}} \rangle \rightarrow\\[-0.1em]
+  \qquad\sem{ \coctx{\Gamma}{ \aclrd{n'} } \vdash e : \tau }~\langle \mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{n'} }\rangle
+\end{array}
+\end{equation}
+\vspace{-0.5em}
+\begin{equation}
+\tag{\emph{abs}}
+\hspace{-1.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\aclrd{n}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 } = 
+    \lambda \langle \mathbf{v}_0, \ldots \mathbf{v}_{\aclrd{n}} \rangle \rightarrow\\[-0.1em]
+  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{n}} \vdash e : \tau_2 
+    }~\langle (\mathbf{v}_0, y_0), \ldots, (\mathbf{v}_{ \aclrd{n}  }, y_{ \aclrd{n} } ) \rangle
+\end{array}
+\end{equation}
+\vspace{-0.5em}
+\begin{equation*}
+\tag{\emph{app}}
+\hspace{-1.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n}+\aclrd{p})} \vdash e_1~e_2 : \tau_2 } = 
+    \lambda (\mathbf{v}_0, \ldots, \mathbf{v}_{ \textnormal{max}(\aclrd{m}, \aclrd{n}+\aclrd{p}) } ) \rightarrow \\[-0.1em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 
+    }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{m} })\\[-0.1em]
+  \qquad\kvd{in}~g~(~\sem{ \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{n} }),\ldots, \\[-0.1em]
+  \qquad\qquad\;~\sem{ \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_{ \aclrd{p} }, \ldots, \mathbf{v}_{ \aclrd{n}+\aclrd{p} })~)
+\end{array}
+\end{equation*}
+
+\caption{Semantics of a language with implicit parameters}
+\label{fig:applications-flat-dfsem}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
 
 \paragraph{Semantics.}
-In this setting, a function $\tau_1 \rightarrow
-\tau_2$ becomes a function from a list of historical values $[\tau_1]
-\rightarrow \tau_2$. A coeffect system here tracks how many past values
-to cache.
+The sample language discussed in this section is a \emph{causal} data-flow language. This means that
+a computation can access \emph{past} values of the stream (but not future values). In the semantics, 
+we again need richer structure over the input.
+
+Uustalu and Vene \cite{comonads-notions} model causal data-flow computations using a non-empty list
+$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$ over the input. A function $\tau_1 \rightarrow \tau_2$
+is thus modelled as $\ident{NeList}~\tau_1 \rightarrow \tau_2$. This model is difficult to implement
+efficiently, as it creates unbounded lists of past elements.
+
+The coeffect system tracks maximal number of past values and so we can define the semantics using
+a list of fixed length. As with liveness, this is a data structure \emph{indexed} by the coeffect
+annotation. We write $\tau^{\aclrd{n}}$ for a list containing $\aclrd{n}$ elements (which can be 
+also viewed as an $\aclrd{n}$-element product $\tau \times \ldots \times \tau$).
+
+As with the previous examples, our semantics interprets a judgement using a (semantic) function;
+functions in the language are modelled as functions taking a list of inputs:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{n} } \vdash e : \tau} 
+  &:& (\tau_1 \times \ldots \times \tau_n)^{\aclrd{n}+1} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\aclrd{n}} \tau_2} &:& \tau_1^{\aclrd{n}+1} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+Note that the semantics requires one more value than is the number of past values. This is because 
+the first value is the current value and has to be always available, even when the annotation is
+zero as in (\emph{var}).
+
+The rules defining the semantics are shown in Figure~\ref{fig:applications-flat-dfsem}. The
+semantics of the context is a \emph{list of pairs}. To make the rules easier to follow, we write
+$\langle \mathbf{v}_1, \ldots, \mathbf{v}_n \rangle$ for an $n$-element list containing pairs.
+Pairs that model the entire context such as $\mathbf{v}_1$ are written in bold. When we access
+individual variables, we write $\mathbf{v} = (x_1, \ldots, x_m)$ where $x_i$ denote individual
+variables of the context.
+
+In (\emph{var}), the context is a singleton-list containing a product of variables, from which
+we project the right one. In (\emph{prev}) and (\emph{sub}), we drop some of the elements from 
+the history (from the front and end, respectively) and then evaluate the original expression.
+
+Lambda abstractions (\emph{abs}) receives two lists of the same size -- one containing values of
+the variables from the declaration-site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}} \rangle$ and one
+containing the argument provided by the call-site $\langle y_0, \ldots, v_{\aclrd{n}} \rangle$.
+The semantics applies the well-known \emph{zip} operation on the lists and passes the result to the
+body.
+
+Finally, application (\emph{abs}) uses the input context in two ways, which gives rise to the
+two requirements combined using \emph{max}. First, it evaluates the expression $e_1$ which is 
+called with the past $\aclrd{m}$ values. The resulting function $g$ is then sequentially composed
+with the semantics of $e_2$. To call the function, we need to evaluate $e_2$ repeatedly -- namely,
+$\aclrd{p}+1$ times, which results in the overall requirement for $\aclrd{n}+\aclrd{p}$ past values.
+
+\paragraph{Summary}
+
+\todo{Key points...}
 

### 2014-04-17 text/Chapter03.tex
Finish flat coeffects in Chapter 3, start structural

@@
 
+\paragraph{Summary.}
+Implicit parameters are the simplest example of a system where function abstraction does not 
+delay all impurities of the body. As discussed in Section~\ref{sec:applications-structure-lam},
+this is the defining feature of \emph{coeffect} systems.
+
+In this section, we have seen how this affects both the type system and the semantics of the 
+language. In the type system, the (\emph{abs}) rule places context-requirements on both the 
+declaration-site and the call-site. For implicit parameters, this rule introduces non-determinism,
+because the parameters can be split arbitrarily. However, as we show in the next section, this is 
+not always the case. Semantically, lambda abstraction \emph{merges} two parts of context (implicit 
+parameter dictionaries) that are provided by the call-site and declaration-site.
+
 %---------------------------------------------------------------------------------------------------
@@
 
+\paragraph{Summary.}
+Unlike with implicit parameters, the lambda abstraction for liveness analysis does not introduce 
+non-determinism. It simply duplicates the context requirements. However, this still matches the
+property of coeffects that impurities cannot be thunked.
+
+The semantics of liveness reveals a number of interesting properties too. Firstly, the semantics
+cannot be captured by any monad. Secondly, the system would not work without the coeffect annotations. 
+The shape of the semantic function depends on the annotation (the input is either $1$ or $\tau$) and 
+is \emph{indexed} by the annotation. Finally, we discussed at length how the semantics of application
+arises from \emph{sequential} and \emph{pointwise} composition. This is another important aspect of
+coeffect systems -- categorical semantics typically builds on \emph{sequential} composition, but to
+model full $\lambda$ calculus it needs more. For coeffect systems, we need \emph{pointwise} composition
+where the same context is shared by multiple sub-expressions.
+
 % --------------------------------------------------------------------------------------------------
@@
 
-\caption{Semantics of a language with implicit parameters}
+\caption{Semantics of a simple data-flow language}
 \label{fig:applications-flat-dfsem}
@@
 
-\paragraph{Summary}
-
-\todo{Key points...}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Safe locking}
-In the previous examples, the context provides additional values or functions that may be accessed
-at runtime. However, it may also track \emph{permissions} to perform some operation. This is done
-in the type system for safe locking of Flanagan and Abadi \cite{app-safe-locking}.
-
-The system prevents race conditions (by only allowing access to mutable state under a lock)
-and avoids deadlocks (by imposing strict partial order on locks). The following 
+\paragraph{Summary.}
+The most interesting point about the data-flow example is that it is remarkably similar to our
+earlier liveness example. In the type system, abstraction (\emph{abs}) duplicates the context
+requirements and application (\emph{abs}) arises from sequential and pointwise composition.
+We capture this striking similarity in Chapter~\ref{ch:flat}. Before doing that, we look at one
+more example and then explore the \emph{structural} class of systems.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Permissions and safe locking}
+In the implicit parameters and data-flow examples, the context provides additional resources or 
+values that may be accessed at runtime. However, it may also track \emph{permissions} or 
+\emph{capabilities} to perform some operation. Liveness can be seen as a trivial example -- when
+the context is live, it contains a permission to access variables. In this section, we briefly
+consider a system for safe locking of Flanagan and Abadi \cite{app-safe-locking} as one, more
+advanced example. Calculus of capabilities of Cray et al. \cite{app-capabilities} is discussed 
+later in Section~\ref{sec:applications-active-ccc}.
+
+\paragraph{Safe locking.}
+The system for safe locking prevents race conditions (by only allowing access to mutable state 
+under a lock) and avoids deadlocks (by imposing strict partial order on locks). The following 
 program uses a mutable state under a lock:
@@
 \begin{array}{l}
-\kvd{newlock}~l:\rho~\kvd{in}\\
-\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\
+\kvd{newlock}~l:\rho~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\[-0.25em]
 \kvd{sync}~l~(!\ident{state})
@@
 
-The type system for safe locking associates the list of permission with the variable context.
-It uses judgements of a form $\Gamma, m \vdash e : \alpha$ specifying that an expression has a
-type in context $\Gamma$, given permissions (a list of locked regions) $m$. However, the treatment
-of lambda abstraction differs from the one for implicit parameters or rebindable resources.
-In the system for locking, code inside lambda function cannot use permissions from the scope
-where the function is declared. This is a necessary requirement -- a lambda function created 
-under a lock cannot access protected memory, because it will be executed later. We discuss how
-this restriction fits into our general coeffect framework in Section~X.Y.
-
-%
-% => This has coeffect style judgments, but it has effect-style lambda
-%
+The type system for safe locking associates a list of acquired locks with the context. 
+Interestingly, the original presentation of the system uses a coeffect-style judgements
+of a form $\Gamma; p \vdash e : \tau$ where $p$ is a list of accessible regions (protected by
+an acquired lock). Using our notation, the rule for \kvd{sync} looks as follows:
+%
+\begin{equation*}
+\tyrule{sync}
+  { \coctx{\Gamma}{\aclrd{p}} \vdash e_1 : m& 
+    \coctx{\Gamma}{\aclrd{p} \cup \{m\} } \vdash e_2 : \tau }
+  { \coctx{\Gamma}{\aclrd{p}} \vdash \kvd{sync}~e_1~e_2 : \tau }
+\end{equation*}
+%
+The rule requires that $e_1$ yields a value of a singleton type $m$. The type is added as an
+indicator of the locked region to the context $\aclrd{p}\cup \{m\}$ which is then used to evaluate
+the expression $e_2$.
+
+\paragraph{Summary.}
+Despite attaching annotations to the variable context, the system for safe locking uses 
+effect-style lambda abstraction. Lambda abstraction associates all requirements with the 
+call-site -- a lambda function created under a lock cannot access protected memory available at 
+the time of creation. It will be executed later and can only access the memory available then.
+This suggests that safe locking is perhaps better seen as an effect system.
+
+Another interesting aspect is the extension to avoid deadlocks. In that case, the type system
+needs to reject programs that acquire locks in an invalid order. One way to model this is to 
+replace $\aclrd{p} \cup \{m\}$ with a \emph{partial} operation $\aclrd{p} \uplus \{m\}$ which
+is only defined when the lock $m$ can be added to the set $\aclrd{p}$. Supporting partial 
+operations on coeffect annotations is an interesting extension which we discuss in Section~?.
+The extension also lets us capture systems with effect-style lambda abstraction such as safe
+locking.
 
@@
 
-\section{Structural coeffects (1.)}
+\section{Structural coeffect systems}
 
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Data-flow languages (revisited)}
-When discussing data-flow languages in the previous section, we said that the context provides 
-past values of variables. This can be viewed as a flat contextual property (the context needs
-to keep all past values), but we can also view it as a structural property. Consider the following
-example:
-%
-\begin{equation*}
-\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
-\end{equation*}
-%
-The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
-To evaluate a current value of the stream, we need the current value of \ident{left} and one past
-value of \ident{right}.
-
-As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
-values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
-single number for each function. However, we can design a more precise \emph{structural} analysis
-and track the number of required elements for individual variables.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Tainting and provenance}
-Tainting is a mechanism where variables coming from potentially untrusted sources are marked
-(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
-can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
-(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
-case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
-values are annotated with more detailed information about their source.
-
-Statically typed systems that based on tainting have been use to prevent cross-site scripting
-attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
-\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
-cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
-type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
-\ident{msg}~+~\ident{name}
-\end{array}
-\end{equation*}
-%
-In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
-untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
-The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
-(i.e.~to construct an SQL query).
-
-In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
-flag (for tainting) or more complex data (for provenance). In static checking, the information
-need to be associated with the variables in the variable context. We use tainting as a motivating
-example for \emph{structural} coeffects in Section~X.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Security and core dependency calculus}
-
-The checking of tainting is a special case of checking of the \emph{non-interference} property 
-in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
-as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
-network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
-system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
-more recent work. The checking of information flows has been also integrated (as a single-purpose
-extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
-Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
-using a monadic library.
-
-Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
-where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\ident{L}, \ident{H}\}$ 
-represents low and high secrecy, respectively with $\ident{L} \leq \ident{H}$ meaning that low security
-values can be treated as high security (but not the other way round).
-
-An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
-example which may assign a new value to $z$:
-%
-\begin{equation*}
-\kvd{if}~x>0~\kvd{then}~z := y
-\end{equation*}
-%
-If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
-(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
-$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
-assignment is performed or not performed leaks information in its own (this is an 
-\emph{implicit} flow).
-
-Abadi et al. realized that there is a number of analyses similar to secure information flow
-and proposed to unify them using a single model called Dependency Core Calculus (DCC) \cite{app-dcc}.
-It captures other cases where some information about expression relies on properties of variables
-in the context where it executes.  The DCC captures, for example, \emph{binding time analysis}
-\cite{app-binding-time-analysis}, which detects which parts of programs can be partially evaluated
-(do not depend on user input) and \emph{program slicing} \cite{app-slicing-survey} that identifies
-parts of programs that contribute to the output of an expression.
-
-%===================================================================================================
-
-\section{Structural coeffects}
-\label{sec:intro-coeffects}
+% % % % % % % %
 
@@
 
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Liveness}
+\label{sec:applications-struct-live}
+
+todo
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Bounded reuse}
-\label{sec:intro-coeffects-bll}
 
@@
 
-\subsection{Liveness}
-\label{sec:applications-struct-live}
-
-todo
-
-\subsection{Dataflow and data access}
+\subsection{Data-flow languages (revisited)}
+When discussing data-flow languages in the previous section, we said that the context provides 
+past values of variables. This can be viewed as a flat contextual property (the context needs
+to keep all past values), but we can also view it as a structural property. Consider the following
+example:
+%
+\begin{equation*}
+\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
+\end{equation*}
+%
+The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
+To evaluate a current value of the stream, we need the current value of \ident{left} and one past
+value of \ident{right}.
+
+As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
+values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
+single number for each function. However, we can design a more precise \emph{structural} analysis
+and track the number of required elements for individual variables.
+
+% % % % % %
+
+\paragraph{Dataflow and data access}
 
@@
 
-\subsection{Implicit parameters}
-\label{sec:intro-coeffects-impl}
-
-As our third example, we revisit Haskell implicit parameters \cite{lewis2000implicit} 
-used in our earlier coeffect work \cite{petricek2013coeffects}. Implicit parameters are variables
-that mix aspects of dynamic and lexical scoping. Implicit parameters are a distinct syntactic 
-category to variables and we write them as \ident{?p}. For simplicity, we omit \emph{let}-binding for
-implicit parameters and focus just on tracking requirements.
-
-\paragraph{Implicit parameters coeffects.} Implicit parameters are a whole-context coeffect 
-not linked to ordinary variables. We keep track of sets of implicit parameters that are required
-by an expression (and their types). For example $\coctx{\Gamma}{\{ \ident{?p}_1 : \tau_1, \ldots, \ident{?p}_n : \tau_n \}}$
-means that a context provides ordinary variables $\Gamma$ and values for implicit parameters $\ident{?p}_i$.
-Unlike in the previous examples, we no longer need to distinguish between coeffects attached
-to variables (scalars) and coeffects attached to contexts (vectors), so we write $\cclrd{r},\cclrd{s},\cclrd{t}$
-for both.
-
-Despite the differences, the type system in Figure~\ref{fig:implpar-coeff} follows the same 
-structure as the earlier two examples. Context requirements are created when accessing an implicit
-parameter (\emph{param}) (a system-specific rule). Structural rules (exchange, weaken, contract) 
-do not affect the coeffects. For example parameters are reordered in (\emph{exch}),
-but this has no effect as set union $\cup$ is commutative.
-
-In abstraction and application, the structural $\cons$ operator (previously vector 
-concatenation) becomes $\cup$. Sets of implicit parameters are not associated to
-individual variables and so they are unioned. The (\emph{app}) rule uses $\cup$ to
-combine the implicit parameters required by the function with the requirements of the argument too.
-
-We call this a \emph{flat} coeffect system since coeffects have only
-one shape (there is no scalar/vector distinction).  Other flat
-coeffect systems may use a richer structure~\cite{petricek2013coeffects}. In
-particular, the operations used in abstraction and application may differ
-(to accommodate over-approximation). We return to this in
-Section~\ref{sec:coeffects-structural}.
-
-\paragraph{Example.} Unlike structural coeffect systems, flat systems
-do not necessarily have principal coeffects. 
-This arises from the (\emph{abs}) rule which can freely split
-requirements between the function type and the declaring
-context. Consider a function %filters news based on a location:
-$\lambda ().\ident{?p}_1 + \ident{?p}_2$.  There are nine possible 
- type and coeffect derivations, two of which are:
-%% 1 p1p2  p1p2
-%% 2 p1p2  p1
-%% 3 p1p2  p2
-%% 4 p1p2  
-%% 5 p1    p1p2
-%% 6 p1    p2
-%% 7 p2    p1p2
-%% 8 p2    p1
-%% 9       p1p2
-\[
-\begin{array}{rl}
-\coctx{\emptyset}{ \{  \} } &\hspace{-0.7em}\vdash (\ldots) : 
-  \ident{unit} \xrightarrow{ \{ \ident{?p}_1:\mathbb{Z}, \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
-\\
-\coctx{\emptyset}{ \{ \ident{?p}_1:\mathbb{Z} \} } &\hspace{-0.7em} \vdash (\ldots) : 
-  \ident{unit} \xrightarrow{ \{ \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
-\end{array}
-\]
-In the first case, both parameters are dynamically scoped and have to be provided by the caller.
-In the second case, the parameter $\ident{?p}_1$ is available in the declaring scope and so it is
-(lexically) captured.
-
-Although structural coeffects have more desirable syntactic properties, we aim to capture this
-non-principality too as it is practically useful -- Haskell's implicit parameters use it and
-it can be used to model resource rebinding in distributed systems such as~\cite{app-distributed-acute}).
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}
-\[
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau,y:\sigma,\Gamma_2}{r \cup \cclrd{s} \cup \cclrd{t} \cup \cclrd{q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\sigma,x:\tau,\Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s} \cup \cclrd{q}} \vdash e : \tau}
-\]
-\[
-\tyrule{app}
-  { \coctx{\Gamma_1}{\cclrd{r}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
-    \coctx{\Gamma_2}{\cclrd{s}} \vdash e_2 : \sigma }
-  { \coctx{\Gamma_1, \Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s}} \vdash e_1 \, e_2 : \tau} 
-\]
-\[
-\tyrule{param}{}
-  {\coctx{()}{ \{ \ident{?p}:\tau \} } \vdash \ident{?p} : \tau}
-\quad\;
-\tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \sigma}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
-\]
-\vspace{-0.5em}
-\caption{Type and coeffect system for implicit parameters}
-\label{fig:implpar-coeff}
-\end{figure}
+\subsection{Tainting and provenance}
+Tainting is a mechanism where variables coming from potentially untrusted sources are marked
+(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
+can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
+(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
+case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
+values are annotated with more detailed information about their source.
+
+Statically typed systems that based on tainting have been use to prevent cross-site scripting
+attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
+\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
+cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
+type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
+\ident{msg}~+~\ident{name}
+\end{array}
+\end{equation*}
+%
+In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
+untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
+The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
+(i.e.~to construct an SQL query).
+
+In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
+flag (for tainting) or more complex data (for provenance). In static checking, the information
+need to be associated with the variables in the variable context. We use tainting as a motivating
+example for \emph{structural} coeffects in Section~X.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Security and core dependency calculus}
+
+The checking of tainting is a special case of checking of the \emph{non-interference} property 
+in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
+as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
+network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
+system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
+more recent work. The checking of information flows has been also integrated (as a single-purpose
+extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
+Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
+using a monadic library.
+
+Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
+where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\ident{L}, \ident{H}\}$ 
+represents low and high secrecy, respectively with $\ident{L} \leq \ident{H}$ meaning that low security
+values can be treated as high security (but not the other way round).
+
+An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
+example which may assign a new value to $z$:
+%
+\begin{equation*}
+\kvd{if}~x>0~\kvd{then}~z := y
+\end{equation*}
+%
+If the value of $y$ is high-secure, then $z$ becomes high-secure after the assignment
+(this is an \emph{explicit} flow). However, if $x$ is high-secure, then the value of
+$z$ becomes high-secure, regardless of the security level of $y$, because the fact whether an 
+assignment is performed or not performed leaks information in its own (this is an 
+\emph{implicit} flow).
+
+Abadi et al. realized that there is a number of analyses similar to secure information flow
+and proposed to unify them using a single model called Dependency Core Calculus (DCC) \cite{app-dcc}.
+It captures other cases where some information about expression relies on properties of variables
+in the context where it executes.  The DCC captures, for example, \emph{binding time analysis}
+\cite{app-binding-time-analysis}, which detects which parts of programs can be partially evaluated
+(do not depend on user input) and \emph{program slicing} \cite{app-slicing-survey} that identifies
+parts of programs that contribute to the output of an expression.
 
@@
 \paragraph{Calculus of capabilities}
+\label{sec:applications-active-ccc}
+
 Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 

### 2014-04-18 text/Chapter03.tex
Working on structural liveness (Chapter 3)

@@
 \subsection{Scalars and vectors}
+\label{sec:applications-strucutre-vec}
 
@@
 over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars. 
-We use upper-case letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for vectors and lower-case
+We use bold-face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for vectors and lower-case
 letters $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
 also distinguishes different structures using colours. However ignoring the colour does not introduce 
-any ambiguity.}. We also say that a \emph{shape} of a vector $\slift{\aclrd{R}}$ (or more generally any container) 
+any ambiguity.}. We also say that a \emph{shape} of a vector $\slift{\aclrd{\textbf{r}}}$ (or more generally any container) 
 is the set of \emph{positions} in a vector. So, a vector of length $n$ has shape $\{ 1, 2, \ldots, n \}$. 
@@
 scalar-vector one. For any binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
- $\cclrd{s} \circ \aclrd{R} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
-Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{R}, \aclrd{S}$ of the
+ $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
+Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
 same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define 
-$\aclrd{R} \varpropto \aclrd{S} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
+$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
 Finally, we often concatenate vectors -- for example, when joining two variable contexts.
-Given vectors $\aclrd{R}, \aclrd{S}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and 
+Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and 
 $\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as 
-$\aclrd{R}\times\aclrd{S} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.
+$\aclrd{\textbf{r}}\times\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.
 
@@
 
-We now turn our attention to system where additional contextual information are associated not
-with the context as a whole (or program scope), but with individual variables. We start by looking
-simple static analysis -- variable \emph{liveness}. Then we revisit data-flow computations and
-look at applications in security and software updating.
-
-% % % % % % % %
-
-Coeffects are way to describe notions of context in programming that keep turning up. 
-To illustrate this, we overview three systems tracking contextual properties that 
- motivate our general coeffect system. Two systems track per-variable properties (bounded 
-linear logic and dataflow) and one tracks whole-context properties (implicit parameters).
+In structural coeffect systems, the additional information are associated with individual variables.
+This is very often information about how the variables are used, or, in which contexts they are 
+used.
+
+In Chapter~\ref{ch:intro}, we introduced the idea using an example that tracks array access patterns.  
+Each variable is annotated with a range specifying which elements of the corresponding array 
+may be accessed. In this section, we look at a number of examples. We first consider an example
+inspired by linear logic. Then we revisit liveness and data-flow, for which the structural system 
+provides a more precise analysis. Finally, we look at a number of other practical uses including 
+security, tainting and provenance tracking.
 
@@
 
-\subsection{Liveness}
+\subsection{Liveness analysis revisited}
 \label{sec:applications-struct-live}
 
-todo
-
-% --------------------------------------------------------------------------------------------------
+The flat system for liveness analysis presented in Section~\ref{sec:applications-flat-live} is 
+interesting from a theoretical perspective, but it is not practically useful. In this section, we
+revisit the problem and define a structural system that tracks liveness per-variable.
+
+\paragraph{Structural liveness.} 
+Recall two examples discussed earlier where the flat liveness analysis marked the whole context
+as (syntactically) live, despite the fact part of it was (semantically) dead:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{constant} = \lambda y \rightarrow \lambda x \rightarrow y\\
+\kvd{let}~\ident{answer} = (\lambda x \rightarrow x)~42
+\end{array}
+\end{equation*}
+%
+In the first case, the variable $x$ is dead, but was marked as live. In the second example, the
+declaration-site of the \ident{answer} value is dead, but was marked as live. This is because in 
+both of the expressions, \emph{some} variable is accessed. However, the (\emph{abs}) rule of flat
+liveness has no way of determining \emph{which} variables are used by the body -- and, in particular,
+whether the accessed variable is the \emph{bound} variable or some of the \emph{free} variables.
+
+As discussed earlier, we can resolve this by attaching a \emph{vector} of liveness annotations to
+a \emph{vector} of variables. In the first example, the available variables are $y$ and $x$, so
+the variable context $\Gamma$ is a vector $\langle y\!:\!\tau, x\!:\!\tau \rangle$. Only the variable $y$
+is used and so the annotated context is: $\coctx{y\!:\!\tau, x\!:\!\tau}{ \alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}} }$.
+When writing the contexts, we omit angle brackets around variables, but it should still be viewed
+as a vector. There are two important points:
+
+\begin{itemize}
+\item The fact that variables are now a vector means that we cannot freely re-order them. This
+  guarantees that $\coctx{x\!:\!\tau, y\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$
+  can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$.
+  We need to define the type system in a way that is similar to sub-structural systems 
+  (discussed in Section~\ref{sec:path-logic}) which provide explicit rules for manipulating
+  the context.
+
+\item We choose to attach a vector of annotations to a vector of variables, rather than attaching
+  individual annotations to individual variables. This lets us unify and combine flat and 
+  structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
+  explored in Chapter~\ref{ch:meta}.
+\end{itemize}
+
+\paragraph{Type system.}
+The structural system for liveness uses the same two-point lattice of annotations
+$\mathcal{L}=\{ \ident{L}, \ident{D} \}$ that was used by the flat system. We also use the
+$\sqcup, \sqcap$ and $\sqsubseteq$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.
+
+The rules of the system are split into two groups. Figure~\ref{fig:applications-struct-live} (a) shows 
+the standard syntax-driven rules plus sub-coeffecting. In (\emph{var}), the context contains just the 
+single accessed variable, which is annotated as live. Other variables can be introduced using weakening. 
+A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
+sub-coeffecting rule (\emph{sub}) uses a point-wise extension of the $\sqsubseteq$ relation over two
+vectors as defined in Section~\ref{sec:applications-strucutre-vec}.
+
+In the (\emph{abs}) rule, the variable context of the body $\Gamma, x\!:\!\tau_1$ is annotated with
+a vector $\mathbf{\aclrd{r}}\cons\alift{\cclrd{s}}$, where the vector $\mathbf{\aclrd{r}}$ corresponds
+to $\Gamma$ and the singleton annotation $\cclrd{s}$ corresponds to the variable $x$. Thus, the
+function is annotated with $\cclrd{s}$. Note that the free-variable context is annotated with vectors,
+but functions take only a single input and so are annotated with primitive annotations.
+
+The (\emph{app}) rule is similar to function applications in flat systems, but there is an important
+difference. In structural systems, the two sub-expressions have separate variable contexts
+$\Gamma_1$ and $\Gamma_2$. Therefore, the composed expression just concatenates the variables
+and their corresponding annotations. (We can still use the same variable in both sub-expressions
+thanks to the structural contraction rule.)
+
+The context $\Gamma_1$ is used to evaluate $e_1$ and is thus annotated with $\aclrd{\mathbf{r}}$. 
+The annotation for $\Gamma_2$ is more interesting. It is a result of sequential composition of two
+semantic functions -- the first one takes the (multi-variable) context $\Gamma_2$ and evaluates 
+$e_2$; the second takes the result of type $\tau_1$ and passes it to the function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$.
+The composition is defined as follows:
+%
+\begin{equation*}
+g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\aclrd{\mathbf{s}}} \sigma
+\qquad
+f : \sigma \xrightarrow{\cclrd{t}} \tau
+\qquad
+f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \sqcup \aclrd{\mathbf{s}}} \tau
+\end{equation*}
+%
+This definition is only for illustration and is revised in Chapter~\ref{ch:structural}. The function
+$g$ takes a product of multiple variables (and is annotated with a vector). The function $f$ takes
+just a single value and is annotated with the scalar. As in the flat system, sequential composition
+is modelled using $\sqcup$ -- but here, we use a scalar-vector extension of the operation. Finally,
+the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$.
+
+\paragraph{Structural typing rules.}
+
+The rules are split into two groups. We first discuss the standard rules.
+
+is good
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
+\begin{equation*}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c:\tau \in \Delta}
+  {\coctx{()}{\alift{}} \vdash c : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1, x : \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+\end{equation*}
+\vspace{0.5em}
+
+{\small b.) Structural rules for context manipulation}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+\end{equation*}
+
+\caption{Structural coeffect liveness analysis}
+\label{fig:applications-struct-live}
+\vspace{-1em}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}[t]
-\vspace{-0.5em}
-\begin{align*}
-\begin{array}{c}
-%
-\hspace{-0.6em}
-\begin{array}{c}
+{\small Ordinary rules}
+
+\begin{equation*}
 \tyrule{var}{}
   {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
-\quad 
-\tyruleCompact{weak}
-  {\coctx{\Gamma}{R} \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
-\\[1.8em]
-\tyruleCompact{sub}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
-\quad 
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
+\end{equation*}
+\begin{equation*}
 \tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
-\\[1.2em]
-\end{array} \\
-%
-% ------------------------ 
-%
-\begin{array}{c}
+  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
+\end{equation*}
+\begin{equation*}
 \tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
-    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
-\\[1.3em]
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1, v : \sigma}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau}
+\end{equation*}
+
+{\small Structural rules}
+
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
+\end{equation*}
+\begin{equation*}
 \tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
-\\[1.3em]
+  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+\end{equation*}
+\begin{equation*}
 \tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
-\end{array} 
-% ------------------------ 
-%
-% \hspace{-1em}
-%\begin{array}{c}
-%
-% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
-% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
-% to keep it simple.
-%
-% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
-% (I do not understand why but it makes the line somehow thicker...)
-%
-% \tyruleCompact{let}
-%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
-%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
-%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
-\end{array}
-\end{align*}
-\vspace{-0.5em}
-\caption{type and coeffect system for bounded reuse}
-\label{fig:bounded-coeff}
+  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+\end{equation*}
+
+\caption{Coeffect rules for tracking implicit parameters}
+\label{fig:applications-struct-bll}
+\vspace{-1em}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
 

### 2014-04-19 text/Chapter03.tex
Semantics of structural liveness (Chapter 3)

@@
 
-\paragraph{Structural typing rules.}
-
-The rules are split into two groups. We first discuss the standard rules.
-
-is good
-
 % --------------------------------------------------------------------------------------------------
@@
 \tyrule{let}
-  { \coctx{\Gamma_1, x : \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
+  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
     \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
@@
 \tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} 
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+\quad
+\begin{array}{l}
+\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
+\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\end{array}
 \end{equation*}
@@
 \tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}\sqcap\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+~
+\begin{array}{l}
+\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
+\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\end{array}
 \end{equation*}
@@
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Structural typing rules.}
+The structural typing rules are shown in Figure~\ref{fig:applications-struct-live} (b). They mirror
+the rules know from sub-structural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
+extends the context with a single unused variable $x$ and adds the $\ident{D}$ annotation to the
+vector of coeffects.
+
+The variable is always added to the end as in the (\emph{abs}) rule. However, the exchange rule 
+(\emph{exch}) lets us arbitrarily reorder variables. It flips the variables $x$ and $x'$ and their
+corresponding coeffect annotations in the vector. This is done by requiring that the lengths of the
+remaining, unchanged, parts of the vectors match.
+
+Finally, contraction (\emph{contr}) makes it possible to use a single variable multiple times.
+Given a judgement that contains variables $y$ and $z$, we can derive a judgement for an expression
+where both $z$ and $y$ are replaced by a single variable $x$. Their annotations $\cclrd{s}, \cclrd{t}$
+are combined into $\cclrd{s} \sqcap \cclrd{t}$, which means that $x$ is live if either $z$ or $y$
+were live in the original expression.
+
+\paragraph{Example.} To demonstrate how the system works, we consider the expression
+$(\lambda x \rightarrow v)~y$. This is similar to an example where flat liveness mistakenly marks
+the entire context as live. Despite the fact that the variable $y$ is accessed (syntactically), it
+is not live -- because the function that takes it as an argument always returns $v$.
+
+The typing derivation for the body uses (\emph{var}) and (\emph{abs}). However, we also need (\emph{weak})
+to add the unused variable $x$ to the context:
+%
+\begin{equation*}
+\tyrule{weak}
+{ \tyruler{var}
+    {}
+    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash v : \tau } }
+{ \tyruler{abs}
+    { \coctx{v\!:\!\tau, x\!:\!\tau}{\alift{ \cclrd{\ident{L}},\cclrd{\ident{D}} }} \vdash v : \tau }
+    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash (\lambda x \rightarrow v) : \tau \xrightarrow{\cclrd{\ident{D}}} \tau }}
+\end{equation*}
+%
+The interesting part is the use of the (\emph{app}) rule in the next step. Although the variable $y$ is live in the expression $y$,
+it is marked as dead in the overall expression, because the function is annotated with $\ident{\cclrd{D}}$:
+%
+\begin{equation*}
+\hspace{-2em}
+\tyrule{app}
+  {
+    \begin{array}{l}
+    \vspace{-1.2em}
+    \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash (\lambda x \rightarrow v) : \tau \xrightarrow{\cclrd{\ident{D}}} \tau 
+    \end{array} &
+    \tyruler{var}{}{\coctx{y\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash y : \tau}
+  }
+  { 
+  \inference
+  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\ident{L}} } \cons\, ({ \cclrd{\ident{D}} \sqcup \alift{\cclrd{\ident{L}}} })} 
+  	     \vdash (\lambda x \rightarrow v)~y : \tau }
+  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\ident{L}}, \cclrd{\ident{D}} }} \vdash (\lambda x \rightarrow v)~y : \tau }  	
+  }
+\end{equation*}
+%
+The application is written in two steps -- the first one directly applies the (\emph{app}) rule
+and the second one simplifies the coeffect annotation. The key part is the use of the scalar-vector 
+operator $\cclrd{\ident{D}} \sqcup \alift{\cclrd{\ident{L}}}$. Using the definition of the scalar-vector
+extension, this equals $\alift{\cclrd{\ident{D}} \sqcup \cclrd{\ident{L}}}$ which is
+$\alift{\cclrd{\ident{D}}}$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Semantics of ordinary expressions}
+\begin{equation*}
+\tag{\emph{var}}
+\sem{\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} = \lambda(x) \rightarrow x
+\end{equation*}
+\vspace{-1.5em}
+\begin{equation*}
+\tag{\emph{const}}
+\sem{\coctx{()}{\alift{}} \vdash c : \tau} = \lambda () \rightarrow \delta(c)
+\end{equation*}
+\vspace{-1.5em}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{abs}}
+\begin{array}{l}
+\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda y . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \lambda \textbf{v} \rightarrow\\[-0.25em]
+\qquad\lambda y \rightarrow \sem{\coctx{\Gamma, y \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}~(\textbf{v}, y)
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{app-1}}
+\begin{array}{l}
+\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{L}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
+  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
+  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{app-2}}
+\begin{array}{l}
+\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{D}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
+  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 }~\mathbf{v_1}~
+  \kvd{in}~g~()
+\end{array}
+\end{equation*}
+\vspace{0.5em}
+
+{\small b.) Semantics of structural context manipulation}
+\begin{equation*}
+\tag{\emph{weak}}
+\sem{\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} = 
+  \lambda(\textbf{v}, ()) \rightarrow
+    \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}~\textbf{v}
+\end{equation*}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{exch}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} = 
+  \lambda(\mathbf{v_1}, y, x, \mathbf{v_2}) \rightarrow\\[-0.25em]
+\qquad
+  \sem{\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~
+  (\mathbf{v_1}, x, y, \mathbf{v_2})
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{contr-1}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, (), \mathbf{v_2}) \rightarrow \\[-0.25em]
+\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} ~(\mathbf{v_1}, (), (), \mathbf{v_2})
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{contr-2}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, x, \mathbf{v_2}) \rightarrow \\[-0.25em]
+\quad\begin{cases}
+~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, x, \mathbf{v_2})\\[-0.25em]
+~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, (), x, \mathbf{v_2})\\[-0.25em]
+~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, (), \mathbf{v_2})
+\end{cases}
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\caption{Semantics of structural liveness}
+\label{fig:applications-struct-livesem}
+\end{figure}
+\vspace{-0.5em}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Semantics.}
+When defining the semantics of flat liveness calculus, we used an indexed form of the option type
+$1 + \tau$ (which is $1$ for dead contexts and $\tau$ for live contexts). In the semantics of 
+expressions, the type wrapped the entire context, \ie~$1+(\tau_1 \times \ldots \times \tau_n)$.
+In the structural version, the semantics wraps individual elements of the free-variable context
+pair: $(1+\tau_1) \times \ldots \times (1+ \tau_n)$. For each variable, the type is indexed by 
+the corresponding annotation. More formally:
+%
+\begin{equation*}
+\begin{array}{r}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
+  ~:~ (\tau'_1 \times \ldots \times \tau'_n) \rightarrow \tau\\[0.5em]
+\textnormal{where}~\tau'_i = \begin{cases}
+\tau_i & (\cclrd{r_i} = \cclrd{\ident{L}})\\
+1      & (\cclrd{r_i} = \cclrd{\ident{D}})
+\end{cases}
+\end{array}
+\end{equation*}
+%
+Note that the product of the free variables is not an ordinary tuple of our language, but a special
+construction. This follows from the asymmetry of $\lambda$-calculus, as discussed in 
+Section~\ref{sec:applications-strucutre-vec}. Functions take just a single input and so they are
+interpreted in the same way as in flat calculus:
+%
+\begin{equation*}
+\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2 \qquad\qquad
+\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
+\end{equation*}
+%
+The rules that define the semantics are shown in Figure~\ref{fig:applications-struct-livesem}.
+To make the definition simpler, we are somewhat vague when working with products. We write 
+variables of product type such as $\mathbf{v}$ in bold-face and individual values like $x$ in 
+normal face. We freely re-associate products and so $(\mathbf{v}, x)$ should not be seen as a
+nested product, but simply a product with a number of variables represented as another product
+$\mathbf{v}$ with one more variable $x$ at the end. We shall be more precise in Chapter~\ref{ch:structural}.
+
+In (\emph{var}), the context contains just a single variable and so we do not even need to apply
+projection; (\emph{cosnt}) receives no variables and uses global constant lookup function $\delta$.
+In (\emph{abs}), we obtain two parts of the context and combine them into $(\mathbf{v}, x)$. This
+works the same way regardless of whether the variables are live or dead. For simplicity, we omit
+sub-coeffecting, which just turns some of the available values $v_i$ to unit values $()$.
+
+In application, we again need to ``implement'' dead code elimination. When the input parameter
+of the function $g$ is live (\emph{app-1}), we first evaluate $e_2$ and then pass the result to
+$g$. When the parameter is dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values
+in $\mathbf{v_2}$ can be dead, \ie~$()$.
+
+In the structural rules, (\emph{weak}) receives context containing a dead variable as the last one.
+It drops the $()$ value and evaluates the expression in a context $\mathbf{v}$. Exchange (\emph{exch})
+simply swaps two variables. In contraction, we either duplicate a dead value (\emph{contr-1}),
+or a live value (\emph{contr-2}). In the latter, one of the duplicates may be dead and 
+so we need to consider three separate cases.
+
+\paragraph{Summary.}
+The structural coeffect system for tracking liveness analysis
+
+semantics is ``the other way round''
 

### 2014-04-20 text/Chapter03.tex
Tweaks and additions in bounded reuse (Chapter 3)

@@
 \paragraph{Summary.}
-The structural coeffect system for tracking liveness analysis
-
-semantics is ``the other way round''
+The structural liveness calculus is a typical example of a system that tracks per-variable 
+annotations. In a number of ways, the system is simpler than the flat coeffect calculi. In
+lambda abstraction, we simply annotate function with the annotation of a matching variable 
+(this rule is the same for all upcoming systems). In application, the \emph{pointwise} composition
+is no longer needed, because the sub-expressions use separate contexts. On the other hand, 
+we had to add weakening, contraction and exchange rules to let us manipulate the contexts.
+
+The semantics of weakening demonstrates an important point about coeffects that may be quite
+confusing. When we read the \emph{typing rule} from top to bottom, weakening adds a variable
+to the context. When we read the \emph{semantic rule}, weakening drops a variable value from the 
+context! This duality is caused by the fact that coeffects talk about context -- they describe
+how to build the context required by the sub-expressions and so the semantics implements 
+transformation from the context in the (typing) conclusion to the (typing) assumption.
+
+The structural systems discussed in the upcoming sections are remarkably similar to the one
+shown here. We discuss two more examples in details to explore the design space, but we shall
+omit details that are the shared with the system in this section.
 
@@
 
-\subsection{Bounded reuse}
-
-Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
+\subsection{Bounded variable use}
+
+Liveness analysis checks whether a variable is used or unused. With structural coeffects, we can go
+further and track how many times is the variable accessed. Girard et al. \cite{logic-bounded} coined
+this idea as \emph{bounded linear logic} and use it to restrict well-typed programs to 
+polynomial-time algorithms. We first introduce the system in our, coeffect, style and then 
+relate it with the original formulation.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
+\begin{equation*}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+\end{equation*}
+\vspace{0.5em}
+
+{\small b.) Structural rules for context manipulation}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma} 
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+\quad
+\begin{array}{l}
+\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
+\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+~
+\begin{array}{l}
+\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
+\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\caption{Structural coeffect liveness analysis}
+\label{fig:applications-struct-bll}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Bounded variable use.} 
+The system discussed in this section tracks the number of times a variable is accessed in the 
+call-by-name evaluation. Although we look at an example that tracks \emph{variable usage}, the same
+system could be used to track access to resources that are always passed as a reference (and behave
+effectively as call-by-name) and so the system is relevant for call-by-value languages too.
+To demonstrate the idea, consider the following term:
+%
+\begin{equation*}
+(\lambda v.x + v + v)~(x+y)
+\end{equation*}
+%
+When evaluated, the body of the function directly accesses $x$ once and then twice indirectly, via
+the function argument. Similarly, $y$ is accessed twice indirectly. Thus, the overall expression uses
+$x$ three times and $y$ twice.
+
+As discussed in Chapter~\ref{ch:structural}, the system preserves type and coeffect annotations under
+the $\beta$-reduction. Reducing the expression in this case gives $x + (x+y) + (x+y)$. This has the 
+same bounds as the original expression -- $x$ is used three times and $y$ twice.
+
+\paragraph{Type system.}
+The type system in Figure~\ref{fig:applications-struct-bll} annotates contexts with vectors of integers.
+The rules have the same structure as those of the system for liveness analysis. The only difference is
+how annotations are combined -- here, we use integer multiplication ($\ast$) and addition ($+$).
+
+Variable access (\emph{var}) annotates a variable with $1$, meaning that it has been used once. An
+unused variable (\emph{weak}) is annotated with $0$. Multiple occurrences of the same variable are 
+introduced by contraction (\emph{contr}), which adds the numbers of the two contracted variables.
+
+As previously, application (\emph{app}) and let binding (\emph{let}) combine two separate contexts.
+The second part applies a function that uses its parameter $\cclrd{t}$-times to an argument that uses
+variables in $\Gamma_2$ at most $\aclrd{\mathbf{s}}$-times (here, $\aclrd{\mathbf{s}}$ is a vector of
+integers with an annotations for each variable in $\Gamma_2$). The sequential composition (modelling
+call-by-name) multiplies the uses, meaning that the total number of uses is $(\cclrd{t} \ast \aclrd{\mathbf{s}})$
+(where $\ast$ is a multiplication of a vector by a scalar). This models the fact that for each use of
+the function parameter, we replicate the variable uses in $e_2$.
+
+Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
+the number of uses using the pointwise
+$\leq$ relation. We can view any variable as being used a greater
+number of times than it actually is.
+
+\paragraph{Example.} To type check the expression $(\lambda v.x+v+v)~(x+y)$ discussed earlier, we need
+to use abstraction, application, but also the contraction rule. Assuming the type judgement for the body,
+abstractions yields:
+%
+\begin{equation*}
+\tyrule{abs}
+ { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
+ { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
+\end{equation*}
+%
+To type-check the application, the contexts of $e_1$ and $e_2$ need to contain disjoint variables.
+For this reason, we $\alpha$-rename $x$ to $x'$ in the argument $(x+y)$ and later join $x$ and $x'$ using
+the contraction rule. Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the 
+application rule yields a judgement that is simplified as follows:
+%
+\begin{equation*}
+\inference
+  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
+          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
+{\tyrule{contr}
+  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
+  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
+\end{equation*}
+%
+The first step performs scalar multiplication, producing the vector
+$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
+$x$ and $x'$ from the function and argument terms respectively.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Semantics.}
+In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
+$\lambda$-calculus language to encode the model. More formally, this could be, for example,
+a Cartesian closed category. In that model, we can reuse variables arbitrarily and so it is not
+a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
+linear logic in an (ordinary) linear logic where variables can be used at most once.
+
+Following the same approach, we could model a variable $\tau$, annotated with $\cclrd{r}$ as 
+a product containing $\cclrd{r}$ copies of $\tau$, that is $\tau^{\cclrd{r}}$:
+%
+\begin{equation*}
+\begin{array}{r}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
+  ~:~ (\tau^{\cclrd{r_1}}_1 \times \ldots \times \tau^{\cclrd{r_n}}_n) \rightarrow \tau\\[0.5em]
+\textnormal{where}~\tau^{\cclrd{r_i}}_i = \underbrace{\tau_i \times \ldots \times \tau_i}_{\cclrd{r_i}-\text{times}}
+\end{array}
+\end{equation*}
+%
+The functions are interpreted similarly. A function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$ is modelled
+as a function taking $\cclrd{t}$-element product of $\tau_1$ values: $\tau_1^{\cclrd{t}} \rightarrow \tau_2$.
+
+The rules that define the semantics of bounded calculus are mostly the same as (or easy to adapt 
+from) the semantic rules of liveness in Figure~\ref{fig:applications-struct-livesem}. The ones that
+differ are those that use sequential composition (application and let binding) and the contraction
+rule, which represents pointwise composition. The bounded reuse calculus rules look as follows:
+
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{app}}
+\begin{array}{l}
+\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
+  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
+  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
+\end{array}
+\end{equation*}
+
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{contr}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em] 
+\qquad  \lambda(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}+\cclrd{t}}), \mathbf{v_2}) \rightarrow \\[-0.25em]
+\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
+\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{t}}), (x_{\cclrd{t}+1}, \ldots, x_{\cclrd{t}+\cclrd{s}} ), \mathbf{v_2})
+\end{array}
+\end{equation*}
+
+TODO
+
+\paragraph{Bounded linear logic.}
+
+Bounded linear logic \cite{logic-bounded} restricts well-typed terms to polynomial-time algorithms.  
 This is done by limiting the number of times a value (proposition) can be used. An assumption
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-{\small Ordinary rules}
-
-\begin{equation*}
-\tyrule{var}{}
-  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
-    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1, v : \sigma}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau \quad 
-    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \sigma}
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau}
-\end{equation*}
-
-{\small Structural rules}
-
-\begin{equation*}
-\tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{\textbf{r}} \cons \alift{0}} \vdash e : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \tau}
-\end{equation*}
-
-\caption{Coeffect rules for tracking implicit parameters}
-\label{fig:applications-struct-bll}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Bounded linear coeffects.}
-The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
-Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
-$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
-the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
-exchange (\emph{exch}) rules manipulate variables in the context and
-modify the annotations accordingly -- adding the number of uses in
-contraction and switching vector elements in exchange.
-
-For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
-and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
-\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
-annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.
-
-Application (\emph{app}) describes call-by-name evaluation.  Applying
-a function that uses its parameter $\cclrd{t}$-times to an argument
-that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
-total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
-Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
-vector. Meanwhile, the variables in $\Gamma_1$ are used just
-$\aclrd{R}$-times when reducing the expression $e_1$ to a function
-value.
-
-Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
-the number of uses using the pointwise
-$\leq$ relation. We can view any variable as being used a greater
-number of times than it actually is.
-
-\paragraph{Example.} To demonstrate, consider a term
-$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
-times -- once directly inside the function and twice via the variable $v$ after 
-substitution. Similarly, $y$ is used 
-twice. Assuming a judgment for the function body, abstraction yields:
-%
-\begin{equation*}
-\tyrule{abs}
- { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
- { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
-\end{equation*}
-%
-To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
-Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
-a judgment that is simplified as follows:
-\[
-\hspace{-0.7em}\inference
-  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
-          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
-{\inference
-  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
-  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
-\]
-%
-The first step performs scalar multiplication, producing the vector
-$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
-$x$ and $x'$ from the function and argument terms respectively.
-
-It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
-the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
-show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.
+\paragraph{Summary.}
+
+We really need just: $\ast$, $+$, $1$, $0$, $\leq$
+
+composing flat and structural - meta vs. language semantics
 

### 2014-04-28 text/Chapter03.tex
Finish BLL, start working on structural dataflow (Chapter 3)

@@
 
-\caption{Structural coeffect liveness analysis}
+\caption{Structural coeffect bounded reuse analysis}
 \label{fig:applications-struct-bll}
@@
 In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
-$\lambda$-calculus language to encode the model. More formally, this could be, for example,
-a Cartesian closed category. In that model, we can reuse variables arbitrarily and so it is not
+$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian closed 
+category. In that model, we can reuse variables arbitrarily and so it is not
 a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
@@
 differ are those that use sequential composition (application and let binding) and the contraction
-rule, which represents pointwise composition. The bounded reuse calculus rules look as follows:
-
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{app}}
-\begin{array}{l}
-\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
-  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
-  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
-\end{array}
-\end{equation*}
-
+rule, which represents pointwise composition.
+
+In the following, we use variable names $\mathbf{v}_i$ for context containing multiple variables (where 
+each variable may be available multiple times), \ie~have a type $\tau_1^{\cclrd{r_1}}\times\ldots\times\tau_m^{\cclrd{r_m}}$;
+We do not explicitly write the sizes of these vectors (number of variables in a context; number of 
+instances of a variable) as these are clear from the coeffect annotations. We assume that $\Gamma_2$ 
+contains $n$ variables and that $\aclrd{s}=\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}$:
+%
 \begin{equation*}
@@
 \qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
-\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{t}}), (x_{\cclrd{t}+1}, \ldots, x_{\cclrd{t}+\cclrd{s}} ), \mathbf{v_2})
-\end{array}
-\end{equation*}
-
-TODO
+\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}}), (x_{\cclrd{s}+1}, \ldots, x_{\cclrd{s}+\cclrd{t}} ), \mathbf{v_2})
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{app}}
+\begin{array}{l}
+\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = \\[-0.25em]
+  \qquad\lambda (\mathbf{v_1}, ((x_{1,1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{n,\cclrd{t}\ast\cclrd{s}_n}) ) \rightarrow \\[-0.25em]
+  \qquad\qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
+  \qquad\qquad\kvd{let}~\mathbf{y}_1 = ((x_{1,1}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{1,\cclrd{s}_n})  )\\[-0.25em]
+  \qquad\qquad\kvd{let}~\ldots \\[-0.25em]
+  \qquad\qquad\kvd{let}~\mathbf{y}_{ \cclrd{t} } = ((x_{1,(\cclrd{t}-1)\ast\cclrd{s}_1 + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, \\[-0.25em]
+  \hspace{8.2em}                                         (x_{n,(\cclrd{t}-1)\ast\cclrd{s}_n + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_n})  )\\[-0.25em]
+  \qquad\qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_1, \ldots, 
+    \sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_{\cclrd{t}})
+\end{array}
+\end{equation*}
+%
+In the (\emph{contr}) rule, the semantic function is called with $\cclrd{s}+\cclrd{t}$ copies of
+a value for the $x$ variable. The values are split between $\cclrd{s}$ and $\cclrd{t}$ separate
+copies of variables $y$ and $z$, respectively.
+
+The (\emph{app}) rule is similar in that it needs to split the input variable context. However,
+it needs to split values of multiple variables -- in $x_{i,j}$, the index $i$ stands for an
+index of the variable while $j$ is an index of one of multiple copies of the value. In the
+semantic function, the second
+part of the context consists of $n$ variables where the multiplicity of each value is specified 
+by the annotation $\cclrd{s}_i$ multiplied by $\cclrd{t}$. The rule needs to evaluate the argument 
+$e_2$ $\cclrd{t}$-times and each call requires $\cclrd{s}_i$ copies of the $i^\textnormal{th}$ 
+variable. To do this, we create contexts $\mathbf{y}_1$ to $\mathbf{y}_{\cclrd{t}}$, each containing
+$\cclrd{s}_i$ copies of the variable (and so we require $\cclrd{s}_i\ast\cclrd{t}$ copies of each 
+variable). Note that the contexts are created such that each value is used exactly once.
+
+It is worth noting that the (\emph{var}) rule requires exactly one copy of a variable and so 
+the system tracks precisely the number of uses. However, the (\emph{sub}) rule lets us 
+ignore additional copies of a value. Thus, permitting (\emph{sub}) rule is only possible if the
+underlying model is \emph{affine} rather than \emph{linear}.
 
 \paragraph{Bounded linear logic.}
-
-Bounded linear logic \cite{logic-bounded} restricts well-typed terms to polynomial-time algorithms.  
-This is done by limiting the number of times a value (proposition) can be used. An assumption
-$!_k A$ means that a variable can be used at most $k$ times. 
-We attach annotations to the whole context rather than individual assumptions and so a context
-$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
-This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
-%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.
-
-Bounded linear logic includes explicit weakening and contraction rules
-that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
+The system presented in this section differs from bounded linear logic (BLL) \cite{logic-bounded}.
+Using the terminology from Section~\ref{sec:path-sem-contextdep}, our system is written in the
+\emph{language semantics} style, while BLL is written in the \emph{meta-language} style.
+
+This means that the terms and types of our system are the terms and types of ordinary 
+$\lambda$-calculus, with the only difference that functions carry coeffect annotations.
+In BLL, the language of types is extended with a type constructor $!_k A$ (where $A$ is
+a proposition, corresponding to a type $\tau$ in our system). The type denotes a value
+$A$ that can be used at most $k$ times.
+
+As a result, BLL does not need to attach additional annotation to the variable context 
+as a whole. The requirements are attached to individual variables and so our context
+$\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$ corresponds
+to a BLL assumption $!_{k_1} A_1, ..., !_{k_n} A_n$.
+
+Using the formulation of bounded logic (and omitting the terms), the weakening and
+contraction rules are written as follows:
+%
 \[
-\inference
-  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
-  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
+\tyrule{weak}
+  {\Gamma \vdash B}
+  {\Gamma, !_0 A \vdash B}
 \quad
-\inference
-  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
-  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
+\tyrule{contr}
+  {\Gamma, !_n A, !_m A \vdash B}
+  {\Gamma, !_{n+m} A \vdash B}
 \]
-The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
-$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
-In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
-occurrences of a proposition are joined by adding the number of uses.
+%
+The system captures the same idea as the structural coeffect system presented above.
+Variable access in bounded linear logic is simply an operation that produces a value
+$!_n A$ and so the system further introduces \emph{dereliction} rule which lets us 
+treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
+semantics} and \emph{meta-language} and also revisit the BLL example in 
+Chapter~\ref{ch:meta}.
 
 \paragraph{Summary.}
-
-We really need just: $\ast$, $+$, $1$, $0$, $\leq$
-
-composing flat and structural - meta vs. language semantics
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Data-flow languages (revisited)}
+Comparing the structural coeffect calculus for tracking liveness and for bounded variable reuse
+reveals which parts of the systems differ and which parts are shared. In particular, both systems
+use the same vector operations ($\times$, $\alift{\cclrd{\textnormal{--}}}$) and also share the
+lambda abstraction rule (\emph{abs}). They differ in the primitive values used to annotate used
+and unused variables (\cclrd{L}, \cclrd{D} and $1$, $0$, respectively) and in the operators used
+for sequential composition and contraction ($\sqcup$, $\sqcap$ and $\ast$, $+$, respectively).
+The algebraic structure capturing these operators is developed in Chapter~\ref{ch:structural}.
+
+The brief overview of bounded linear logic shows an alternative approach to tracking properties
+related to individual variables -- we could attach annotations to the variables themselves 
+rather than attaching a \emph{vector} of annotations to the entire context. Our approach has
+two benefits -- it lets you unify flat and structural systems (Chapter~\ref{ch:unified}) and 
+it also makes it possible to build composed systems that mix both flat and structural properties.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Data-flow languages revisited}
+
 When discussing data-flow languages in the previous section, we said that the context provides 
@@
 and track the number of required elements for individual variables.
-
-% % % % % %
-
-\paragraph{Dataflow and data access}
-
-Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
-An expression is re-evaluated when new inputs are available (push) or when more output
-is demanded (pull).  In causal dataflow, programs can access past
-values of a stream. We consider a language where $\kvd{prev}~e$
-returns the previous value of $e$, where 
-$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.
-
-An implementation of causal dataflow may cache past values of
-variables as an optimisation. The question is, how many past values
-should be cached?  This can be approximated by a coeffect system.
 
@@
 % --------------------------------------------------------------------------------------------------
+% variable annotated with 0 (not 1)
+% sub and abs omitted
+% app uses +, not \ast
+% let is default
 
 \begin{figure}[t]
-\[
+\begin{equation*}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{\cclrd{0}}} \vdash x : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{prev}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{1 + \aclrd{\textbf{r}}} \vdash \kvd{prev}~e : \tau}  
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} + \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma} 
+\end{equation*}
+\begin{equation*}
 \tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
-\]
-\[
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
-    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
-\]
-\[
-\tyrule{var}{}
-  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
-\quad\;
-\tyrule{prev}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
-  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
-\]
-\caption{type and coeffect system for dataflow caching}
-\label{fig:dataflow-coeff}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\textnormal{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+\end{equation*}
+
+\caption{Structural coeffect bounded reuse analysis}
+\label{fig:applications-struct-df}
+\vspace{-1em}
 \end{figure}

### 2014-04-29 text/Chapter03.tex
Structural data-flow (Chapter 3)

@@
 \subsection{Data-flow languages}
+\label{sec:applications-flat-dataflow}
+
 The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an 
@@
 When discussing data-flow languages in the previous section, we said that the context provides 
-past values of variables. This can be viewed as a flat contextual property (the context needs
-to keep all past values), but we can also view it as a structural property. Consider the following
-example:
-%
-\begin{equation*}
-\kvd{let}~\ident{offsetZip} = 0~\kvd{fby}~(\ident{left} + \kvd{prev}~\ident{right})
+past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as 
+a \emph{flat} property, which gives us a system that keeps the same number of past values for
+all variables. However, data-flow can also be adapted to a structural system which keeps the number 
+of required past values individually for each variable. Consider the 
+following example:
+%
+\begin{equation*}
+\kvd{let}~\ident{offsetZip} = \ident{left} + \kvd{prev}~\ident{right}
 \end{equation*}
@@
 To evaluate a current value of the stream, we need the current value of \ident{left} and one past
-value of \ident{right}.
-
-As mentioned earlier, a static analysis for data-flow computations could calculate how many past 
-values must be cached. This can be done as a \emph{flat} coeffect analysis that produces just a 
-single number for each function. However, we can design a more precise \emph{structural} analysis
-and track the number of required elements for individual variables.
-
-\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
-similar to the one for bounded reuse in that it tracks a vector
-of numbers $\aclrd{R}$ as part of the context
-$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
-past values (\emph{causality depth}) required for a variable.
-
-Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
-coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
-accessed variables (\emph{var}) are annotated with
-$0$ meaning that no past value is required (only the
-current one). The (\emph{prev}) rule crates caching requirements --
-it increments the number of required values for all variables
-used in $e$ using scalar-vector addition.
-
-Application and contraction have the same structure as before, 
-but use different operators. If two variables 
-are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
-values, then overall we need at most
-$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
-two caches are combined with the maximum of the two requirements, 
-which satisfy the smaller requirements.
-
-In (\emph{app}), the function requires $\cclrd{t}$ past values of
-its parameter.  This means $\cclrd{t}$ past values of
-$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
-variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
-call (\eg{}, we need $1+\aclrd{S}$ values to
-get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
-values to get $2$ past values of $\sigma$, \emph{etc.}).
-
-\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
-$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
-and one bound:
-\[
-\dfrac
-  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
-  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
-\]
-The expression always requires the previous value of $y$ and adds it to
-a previous value of the parameter $x$. Evaluating the value of the
-argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
-and so the overall requirement is $3$ past values:
-\[
-\inference
-  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
-{\inference
-  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
-  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
-\]
-The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
-(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
-the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
-Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
-number of enclosing $\kvd{prev}$s.% expressions.
-
-We previously used dataflow as an example of
-coeffects~\cite{petricek2013coeffects}, but tracked caching
-requirements on the whole context. The system outlined here is more
-powerful and practically useful, with finer-grained coeffects tracking
-caching requirements per-variable.
-
-% --------------------------------------------------------------------------------------------------
-% variable annotated with 0 (not 1)
-% sub and abs omitted
-% app uses +, not \ast
-% let is default
+value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
+requires $1$ past values of both streams in the variable context.
+
+Turning a flat data-flow system to a structural data-flow system is a change similar to the one
+between flat ans structural liveness. In case of liveness analysis, we included the flat system
+only as an illustration (it is not practically useful).
+
+For data-flow, the flat system is less precise, but still practically useful (simplicity may outweigh 
+precision). As discussed in Section~X, the structural system is necessary when we allow arbitrary 
+recursion and cannot (easily) determine the number of required values statically.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Type system.} 
+The type system in Figure~\ref{fig:applications-struct-df} annotates the variable context with a
+vector of integers. This is similar as in the bounded reuse system, but the integers \emph{mean} a
+different thing. Consequently, they are also calculated differently. We omit rules that are the
+same for all structural coeffect systems (exchange, lambda abstraction).
+
+In data-flow, we annotate both used variables (\emph{var}) and unused variables (\emph{weak}) with
+$0$, meaning that no past values are required. This is the same as in flat data-flow, but different
+from bounded reuse and liveness (where difference between using and not using a variable matters).
+Primitive requirements are introduced by the (\emph{prev}) rule, which increments the annotation of
+all variables in the context.
+
+In flat data-flow, we identified sequential composition and point-wise composition as two primitive
+operations that were used in the (flat) application. In the structural system, these are used in 
+(\emph{app}) and (\emph{contr}), respectively. Thus application combines coeffect annotations using
+$+$ and contraction using \emph{max}. This contrasts with bounded reuse, which uses $\ast$ and $+$,
+respectively.
+
+\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
+$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
+and one bound. The (\emph{abs}) rule splits the annotations between the declaration-site and call-site
+of the function:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
+  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
+\end{equation*}
+%
+The expression always requires the previous value of $y$ and adds it to a previous value of the 
+parameter $x$. Evaluating the value of the argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past 
+values of $y$ and so the overall requirement for the (free) variable $y$ is $3$ past values. In 
+order to use the contraction rule, we rename $y$ to $y'$ in the argument:
+%
+\begin{equation*}
+\inference
+  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} &
+    \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~y') : \mathbb{Z} }
+{\inference
+  { \coctx{y\!:\!\mathbb{Z}, y'\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y')) : \mathbb{Z} }
+  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
+\end{equation*}
+%
+The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then (\emph{contr}) to take 
+the maximum, showing three past values are sufficient.
+
+Note that we get the same requirements when we perform $\beta$ reduction of the expression. 
+Substituting the argument for $x$ yields the expression $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
+Semantically, this performs stream lookups $y[1]$ and $y[3]$ where the indices are the 
+number of enclosing $\kvd{prev}$ constructs.
+
+\paragraph{Semantics.}
+To define the semantics of our structural data-flow language, we can use the same approach as when
+adapting flat liveness to structural liveness. Rather than wrapping the whole context in some wrapper
+(list or option type), we now wrap individual components of the product representing the variables
+in the context.
+
+The result is similar as the structure used for bounded reuse. The only difference is that, given
+a variable annotated with $\cclrd{r}$, we need $1+\cclrd{r}$ values. That is, we need the current
+value, followed by $\cclrd{r}$ past values:
+%
+\begin{equation*}
+\begin{array}{l}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
+  ~:~ (\tau^{(\cclrd{r_1}+1)}_1 \times \ldots \times \tau^{(\cclrd{r_n}+1)}_n) \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{s}} \tau_2} ~=~ \tau_1^{(\cclrd{s}+1)} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+Despite the similarity with the semantics for bounded reuse, the values here \emph{represent}
+different things. Rather than providing multiple copies of a value (out of which each can be
+used just once), the pair provides past values (that can be reused and freely accessed).
+To illustrate the behaviour we consider the semantics of the \kvd{prev} construct and of the
+structural contraction rule:
+%
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{prev}}
+\begin{array}{l}
+\sem{\coctx{\Gamma}{\alift{(\cclrd{s}_1 + 1), \ldots, (\cclrd{s}_n + 1)}} \vdash \kvd{prev}~e : \tau} = \\[-0.25em]
+  \qquad\lambda ((x_{1,0}, \ldots, x_{1,{\cclrd{s}_1 + 1}}), \ldots, (x_{n,0}, \ldots, x_{n,{\cclrd{s}_n + 1}})) \rightarrow \\[-0.25em]
+\qquad\qquad\sem{\coctx{\Gamma}{\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}} \vdash e : \tau}\\[-0.25em]
+\qquad\qquad\qquad ((x_{1,0}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,0}, \ldots, x_{n,\cclrd{s}_n}))
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{contr}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\textnormal{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em] 
+\qquad  \lambda(\mathbf{v_1}, (x_0, x_1, \ldots, x_{ \textnormal{max}(\cclrd{s},\cclrd{t}) }), \mathbf{v_2}) \rightarrow \\[-0.25em]
+\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
+\qquad\qquad\qquad (\mathbf{v_1}, (x_0, \ldots, x_{\cclrd{s}}), (x_0, \ldots, x_{\cclrd{t}} ), \mathbf{v_2})
+\end{array}
+\end{equation*}
+%
+In (\emph{prev}), the semantic function is called with an argument that stores values of $n$
+variables, such that a variable $x_i$ has values ranging from $x_{i,0}$ to $x_{i,\cclrd{s}_i + 1}$.
+Thus, there is one current value, followed by $\cclrd{s_i} + 1$ past values. The expression $e$
+nested under $\kvd{prev}$ requires only $\cclrd{s_i}$ past values and so the semantics simply
+drops the last value.
+
+In the (\emph{contr}) rule, the semantic function receives $\emph{max}(\cclrd{s}, \cclrd{t})$ values
+of a specific variable $x$. It needs to produce values for two separate variables, $y$ and $z$ that
+require $\cclrd{s}$ and $\cclrd{t}$ past values. Both of these numbers are certainly smaller than 
+(or equal to) the number of values available. Thus we simply take the first values. Unlike in the
+contraction for BLL, the values are duplicated and the same values are used for both variables.
+
+\paragraph{Summary.}
+Two of the structural examples shown so far (liveness and data-flow) extend an earlier flat 
+version of a similar system. We discuss this relation in general later. However, a flat system 
+can generally be turned into a structural one -- although this only gives a useful system when
+the flat version captures properties related to variables.
+
+The data-flow example demonstrates that the a flat system can also be turned into structural 
+system. In general, this only works for systems where lambda abstraction duplicates context
+requirements (as in Figure~\ref{fig:applications-flat-liveness}).
 

### 2014-04-29 text/Chapter03.tex
Tweak style and finish Chapter 3

@@
 
-\subsection{Tainting and provenance}
+\subsection{Security, tainting and provenance}
 Tainting is a mechanism where variables coming from potentially untrusted sources are marked
@@
 \begin{array}{l}
-\kvd{let}~\ident{name} = \ident{query}~(\str{"SELECT Name WHERE Id = "}~+~\ident{id})~\kvd{in}\\
+\kvd{let}~\ident{name} = \ident{query}(\str{SELECT Name WHERE Id = \%1}, \ident{id})\\[-0.25em]
 \ident{msg}~+~\ident{name}
@@
 In this example, \ident{id} must not come directly from a user input, because \ident{query} requires 
-untainted string. Otherwise, the attacker could specify values such as \str{"1; DROP TABLE Users"}. 
+untainted string. Otherwise, the attacker could specify values such as \str{1; DROP TABLE Users}. 
 The variable \ident{msg} may or may not be tainted, because it is not used in protected context 
@@
 flag (for tainting) or more complex data (for provenance). In static checking, the information
-need to be associated with the variables in the variable context. We use tainting as a motivating
-example for \emph{structural} coeffects in Section~X.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Security and core dependency calculus}
-
+need to be associated with the variables in the variable context.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Core dependency calculus.}
 The checking of tainting is a special case of checking of the \emph{non-interference} property 
-in \emph{secure information flow}. Here, the aim is to guarantee that sensitive information (such
+in \emph{secure information flow}. There, the aim is to guarantee that sensitive information (such
 as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
@@
 
+\paragraph{Implicit flows.}
 An important aspect of secure information flow is called \emph{implicit flows}. Consider the following
-example which may assign a new value to $z$:
-%
-\begin{equation*}
-\kvd{if}~x>0~\kvd{then}~z := y
+example which returns either $y$ or zero, depending on the value of $x$:
+%
+\begin{equation*}
+\kvd{let}~z = \kvd{if}~x>0~\kvd{then}~y~\kvd{else}~0
 \end{equation*}
@@
 
+\paragraph{Coeffect systems.}
+The work outlined in this section is another area where coeffect systems could be applied.
+We do not develop coeffect systems for tracking of tainting, security and provenance in details,
+but briefly mention some examples in the upcoming chapters.
+
+The systems work in the same way as the examples discussed already. For example, consider the
+tainting example with the \ident{query} function calling an SQL database. To capture such 
+tainting, we annotate variables with $\cclrd{\ident{T}}$ for \emph{tainted} and with 
+$\cclrd{\ident{U}}$ for \emph{untainted}. Simply accessing variable does not introduce taint, 
+but using a variable in certain contexts -- such as in arguments of \ident{query} -- does introduce
+a taint. This is captured using the standard application rule (\emph{app}):
+%
+\begin{equation*}
+\tyrule{app}
+  { \coctx{\Gamma}{\aclrd{r}} \vdash \ident{query} : \ident{string} \xrightarrow{\cclrd{\ident{T}}} \ident{Table} \qquad
+    \coctx{\ident{id}:\ident{string}}{ \alift{\cclrd{\ident{U}}}} \vdash \ident{id} : \ident{string} }
+  { \coctx{\Gamma, \ident{id}:\ident{string}}{ \aclrd{r} \cons \alift {\cclrd{\ident{T}}} \vdash \ident{query}(\str{...}, \ident{id}) : \ident{Table} } }
+\end{equation*}
+%
+The derivation assumes that \ident{query} is a standard function that requires the parameters
+to be tainted (it does not have to be a built-in language construct). The argument is a 
+variable and so it is not tainted in the assumptions.
+
+In the conclusion, we need to derive an annotation for the variable \ident{id}. To do this, we
+combine \cclrd{\ident{T}} (from the function) and \cclrd{\ident{U}} (from the argument). In case
+of tainting, the variable is tainted whenever it is already tainted \emph{or} the function marks
+it as tainted. For different kinds of annotations, the composition would work differently -- for
+example, for provenance, we could union the \emph{set} of possible data sources, or even combine 
+\emph{probability distributions} modelling the influence of different sources on the value.
+However, expanding such ideas is beyond the scope of this thesis.
+
 % ==================================================================================================
@@
 
-In the systems discussed so far, the context provides additional data (resources, implicit 
-parameters, historical values) or meta-data (security, provenance). However, it is impossible to
-write a function that modifies the context. We use the term \emph{passive} context for such 
-applications.
-
-However, there is a number of systems where the context may be changed -- not just be evaluating
-certain code block in a different scope (e.g. by wrapping it in $\ident{prev}$ in data-flow), but
-also by calling a function that, for example, acquires new capabilities. While this thesis focuses
-on systems with passive context, we quickly look at the most important examples of the 
-\emph{active} variant.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Calculus of capabilities}
-\label{sec:applications-active-ccc}
-
+In both flat and structural systems discussed so far, the context provides additional data (resources, 
+implicit parameters, historical values) or meta-data (security, provenance). However, \emph{within} 
+the language, it is impossible to write a function that modifies the context. We use the term 
+\emph{passive} context for such applications.
+
+There is a number of systems that also capture contextual properties, but that make it possible to 
+\emph{change} the context -- not just be evaluating certain code block in a different scope (e.g. by 
+wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example, acquires 
+new capabilities and returns those to the caller. While this thesis focuses on systems with passive 
+contexts, we briefly consider the most important examples of the \emph{active} variant.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Calculus of capabilities.}
 Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 
@@
 compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
-who developed an \emph{effect system} (discussed in Section~\ref{sec:path-sem-effects}) that uses
+who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
 lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.
@@
 \begin{array}{l}
-\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
-\quad\kvd{letrgn}~\rho~\kvd{in}\\
-\quad\kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~!\ident{x}
+\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\[-0.25em]
+\quad \kvd{letrgn}~\rho~\kvd{in}\\[-0.25em]
+\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho ~\ident{input}~\kvd{in}\\[-0.25em]
+\quad \ident{x} :=\, !\ident{x} + 1;~ !\ident{x}
 \end{array}
@@
 The memory region $\rho$ is a part of the context, but only in the scope of the body of 
-\kvd{letrgn}. It is only available to the last line which allocates a memory cell in the region
-and reads it (before the region is deallocated). There is no way to allocate a region inside a 
-function and pass it back to the caller.
+\kvd{letrgn}. It is only available to the last two lines which allocate a memory cell in the region,
+increment a value in the region and then read it. The region is de-allocated when the execution 
+leaves its lexical scope -- there is no way to allocate a region inside a function and pass it back 
+to the caller.
 
@@
 of memory regions (and so region lifetimes do not follow strict LIFO ordering). Second, it
-uses continuation-passing style. We ignore the latter aspect and so the following example:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\
-\quad \kvd{letrgn}~\rho~\kvd{in}\\
-\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho \ident{input}~\kvd{in}~\ident{x}
-\end{array}
-\end{equation*}
-%
-The example is almost identical to the previous one, except that it does not return the value
-of reference \ident{x}. Instead, it returns the reference, which is located in a newly allocated
-region. Together with the value, the function returns a \emph{capability} to access the region
-$\rho$.
+uses continuation-passing style. We ignore the latter aspect. The following example is almost 
+identical as the previous one:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{calculate} = \lambda \ident{input} \rightarrow\\[-0.25em]
+\quad \kvd{letrgn}~\rho~\kvd{in}\\[-0.25em]
+\quad \kvd{let}~\ident{x} = \kvd{ref}_\rho ~\ident{input}~\kvd{in}\\[-0.25em]
+\quad \ident{x} :=\, !\ident{x} + 1;~ \ident{x}
+\end{array}
+\end{equation*}
+%
+The difference is that the example does not return the \emph{value} of a reference using 
+$!\ident{x}$, but returns the reference $\ident{x}$ itself. The reference is allocated in a newly 
+created region $\rho$. Together with the value, the function returns a \emph{capability} to access 
+the region $\rho$.
 
@@
 to know what context is required to call \ident{calculate}. We also need to know what effects it
-has on the context when it evaluates and the current context meeds to be appropriately adjusted
-after a function call. We briefly consider this problem in Section~X. % future work
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Software updating}
-Dynamic software updating (DSU) \cite{app-dsu-programs,app-dsu} is the ability to update programs at
-runtime without stopping them. The Proteus system developed by Stoyle et al. \cite{app-dsu-mutatis} 
+has on the context when it evaluates and the current context meeds to be updated after a function 
+call.
+
+\paragraph{Active contexts.}
+In a systems with passive contexts, we only need an annotation that specifies the required 
+context. In semantics, this is reflected by having some structure (data type) $\C$ over the
+\emph{input} of the function. Without giving any details, the semantics generally has the 
+following structure:
+%
+\begin{equation*}
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} = \C^{\cclrd{r}} \tau_1 \rightarrow \tau_2
+\end{equation*}
+%
+Systems with active contexts require two annotations -- one that specifies the context required
+before the call is performed and one that specifies how the context changes after the call (this
+could be either a \emph{new} context or \emph{update} to the original context). Thus the structure
+of the semantics would look as follows:
+%
+\begin{equation*}
+\sem{\tau_1 \xrightarrow{\cclrd{r}, \cclrd{s}} \tau_2} = \C^{\cclrd{r}} \tau_1 \rightarrow \M^{\cclrd{s}} \tau_2
+\end{equation*}
+%
+In case of Calculus of Capabilities, both of the structures could be the same and they could
+carry a set of available memory regions. In this thesis, we focus only on passive contexts. 
+However, we briefly consider the problem of active contexts in the Section~X of the future work 
+chapter.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Software updating.}
+Another example of a system that uses contextual information actively is dynamic software updating 
+(DSU) \cite{app-dsu-programs,app-dsu}. The DSU systems have the ability to update programs at
+runtime without stopping them. For example, Proteus developed by Stoyle et al. \cite{app-dsu-mutatis} 
 investigates what language support is needed to enable safe dynamic software updating in C-like 
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Thesis perspective}
-
-As demonstrated in this section, there is a huge number of systems and applications that exhibit
-a form of context-dependence. The range includes different static analyses (liveness, provenance), 
-well-known programming language features (implicit parameters and type classes) as well as features
-not widely available (e.g. for distributed programming).
-
-It is impossible to cover all of these topics in a single coherent thesis and so we focus on 
-two key aspects:
-
-\begin{compactitem}
-\item \textbf{Flat vs. structural.} We look at both flat coeffects (single value for entire context) and 
-  structural coeffects (single value per variable). We use liveness, implicit parameters and 
-  data-flow to introduce flat coeffects (Section~X) and liveness, refined data-flow and tainting
-  to talk about structural coeffects (Section~Y).
-
-\item \textbf{Analysis vs. restriction.} Some of the discussed examples can be viewed as static
-  analyses that obtain some information about programs (i.e.~the number of required past values
-  in data-flow). Other examples provide type system that rules out certain invalid programs 
-  (e.g.~safe locking). We cover this topic when discussing \emph{partial coeffects} in Section~Z.
-
-\item \textbf{May vs. must analysis.} When discussing liveness, we observed that we can obtain 
-  two different analyses depending on how conditionals are treated. We discuss this topic in 
-  Section~X. % TBD - could be a chapter
-\end{compactitem}
-
-Although we also looked at examples of \emph{active} contextual computations (where developers can
-write functions that modify the context), we do not consider these applications, to keep the 
-material presented in this thesis focused. We briefly discuss them as future work in Section~X.
-
 %===================================================================================================
@@
 
-TODO
-
-% \section{Missing}
-% ~
-% 
-% indexed/layered/etc. monads, productors or whatever
+This chapter served two purposes. The first aim was to present existing work on programming 
+languages and systems that include some notion of \emph{context}. Because there was no well-known
+abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
+from principled approaches based on comonads, ad-hoc type system extensions and static analyses as
+well as approaches based on monads. We looked at a number of applications inclding Haskell's implicit 
+parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of 
+security properties.
+
+The second aim of this chapter was to re-formulate the existing work in a more uniform style and thus
+reveal that all \emph{context-dependent} languages share a common structure. In the upcoming three 
+chapters, we identify the common structure more precisely and develop three calculi to capture it. We 
+will then be able to re-create many of the examples discussed in this chapter by instantiating our 
+unified calculi.
+
+This chapter was divided into two major sections. First, we looked at \emph{flat} systems, which
+track whole-context properties. Next, we look a \emph{structural} systems, which track per-variable
+properties. Both of the variants are useful and important -- for example, implicit parameters can
+only be expressed as \emph{flat} system, but liveness analysis is only useful as \emph{structural}.
+For this reason, we explore both of these variants in this thesis (Chapter~\ref{ch:flat} and 
+Chapter~\ref{ch:structural}, respectively). We can, however, unify the two variants into a single
+system discussed in Chapter~\ref{ch:unified}.

### 2014-05-10 text/Chapter04.tex
Start working on Chapter 4

@@
-\chapter{Flat coeffect language} 
+%\begin{flushright}{\slshape    
+%Even the most puritanical rationalist will be forced to stop arguing and \\
+%use propaganda [...] because the psychological conditions have disappeared \\
+%that allowed effective argument and therefore influence over the others.} \\ \medskip
+%--- Paul Feyerabend, \emph{Against Method} \cite{philosophy-feyerabend}
+%\end{flushright}
+%\vspace{2em}
+
+% ==================================================================================================
+
+\chapter{Flat coeffect calculus} 
 \label{ch:flat}
+
+Successful programming language abstractions need to generalize a wide range of recurring
+problems while capturing the key commonalities. These two aims are typically in opposition -- 
+more general abstractions provide smaller value, while less general abstractions cannot be
+used as often.
+
+In the previous chapter, we outlined a number of systems that capture how computations
+access the environment in which they are executed. We identified two kinds of systems --
+\emph{flat} capturing whole-context properties and \emph{structural} capturing per-variable
+properties. As we show in Chapter~\ref{ch:unified}, the systems can be unified using a single 
+abstraction. This is useful when implementing and composing the systems, but such abstraction 
+provides \emph{smaller value} -- \ie~its generality hides useful properties that we can see 
+when we consider the systems separately. For this reason, this and the next chapter discusses 
+\emph{flat} and \emph{structural} systems separately.
+
+% ==================================================================================================
+
+\section{Introduction}
+\label{sec:flat-intro}
+
+Liveness is more similar to data-flow.
+
+Implicit parameters (or resources) are a bit different, but important.
+
+The closest work is Uustalu and Vene
+
+% ==================================================================================================
+
+\section{Flat coeffect calculus}
+\label{sec:flat-calculus}
+
+This section revisits the flat coeffect calculus, which provides a 
+basis for the structural coeffect calculus that we introduce in Section~... We 
+make a number of improvements -- we cover more standard language features and require additional laws
+which allows a more useful equational theory.
+The formulation is general, so as not to restrict the possible uses unnecessarily, but is
+specialised enough to provide useful properties.
+
+The flat coeffect calculus maintains a single annotation of the contextual requirements associated
+with the variable context. It can be instantiated to capture implicit parameters, the system for 
+dataflow and simple liveness analysis.
+
+The expressions are those of the $\lambda$-calculus with \emph{let} binding; assuming $T$ ranges
+over base types, types are extended as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\end{array}
+\end{equation*}
+%
+where $T$ ranges over type constants. 
+We discuss pairs and recursion in Sections~\ref{sec:flat-synprop} and \ref{sec:structural-properties}.
+The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
+represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
+It can be viewed as a pure function that takes $\tau_1$ \emph{with} context $r$.
+
+%Note that
+%$\ctxtyp{r}{\tau}$ does not exist as syntactic value (although it has a semantic meaning
+%discussed in Section~\ref{sec:flat-semantics}). The annotations $r$ are formed by 
+%an algebraic structure discussed next.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Flat coeffect algebra}
+
+\begin{definition}
+A \emph{\cclrd{flat coeffect algebra}} $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is a set 
+$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
+$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids,
+$(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a semigroup. That is, for all $r,s,t\in \C$:
+%
+\begin{equation*}
+\begin{array}{ccr}
+r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
+\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
+\textnormal{(monoid)}   
+\\
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
+\czero \;\cpar\; r = r = r \;\cpar\; \czero &
+\textnormal{(monoid)}   
+\\
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
+t\; \cleq\; t &
+\textnormal{(pre-order)}   
+\\
+r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
+\textnormal{(assoc)}   
+\end{array}
+\end{equation*}
+%
+In addition, the following distributivity axioms hold:
+\begin{align*}
+\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+\end{definition}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Flat coeffect types}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Examples of flat coeffects}
+
+% ==================================================================================================
+
+\section{Categorical motivation}
+\label{sec:flat-semantics}
+
+\subsection{Comonads}
+\subsection{Indexed comonads}
+\subsection{Semantics of flat calculus}
+\subsection{Examples}
+
+% ==================================================================================================
+
+\section{Equational theory}
+\label{sec:flat-syntax}
+
+\subsection{Call-by-value evaluation}
+\subsection{Call-by-name evaluation}
+
+% ==================================================================================================
+
+\section{Syntactic extensions}
+\label{sec:flat-exts}
+
+\subsection{Lambda abstraction}
+\subsection{Constants and pairs}
+\subsection{Recursion}
+
+% ==================================================================================================
+
+\section{Type inference}
+\label{sec:flat-inference}
+
+\subsection{Semi-lattice formulation}
+\subsection{Type inference algorithm}
+
+% ==================================================================================================
+
+\section{Summary}
+
+\section{----------------JUNK!}
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
+   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
+\end{equation*}
+
+\caption{Flat coeffect calculus}
+\label{fig:flat-types}
+\end{figure}
+
+% ------------------
+
+\section{Type inference}
+
+% ==================================================================================================
+
+~
+\newpage
+~
 

### 2014-05-12 text/Chapter04.tex
Introduction in Chapter 4

@@
 problems while capturing the key commonalities. These two aims are typically in opposition -- 
-more general abstractions provide smaller value, while less general abstractions cannot be
+more general abstractions are less powerful, while less general abstractions cannot be
 used as often.
@@
 abstraction. This is useful when implementing and composing the systems, but such abstraction 
-provides \emph{smaller value} -- \ie~its generality hides useful properties that we can see 
+is  \emph{less powerful} -- \ie~its generality hides useful properties that we can see 
 when we consider the systems separately. For this reason, this and the next chapter discusses 
@@
 
-Liveness is more similar to data-flow.
+In the previous chapter, we looked at three important examples of systems that track whole-context 
+properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live}) 
+and whole-context data-flow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
+structure -- their lambda abstraction duplicates the requirements and their application arises
+from the combination of \emph{sequential} and \emph{point-wise} composition.
 
-Implicit parameters (or resources) are a bit different, but important.
+The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl}), and
+similar systems for rebindable resources, differ in two ways. In lambda abstraction, they split
+the context requirements between the declaration-site and the call-site and they use only a single
+operator on the indices, typically $\cup$.
 
-The closest work is Uustalu and Vene
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Contributions}
+
+All of the examples are practically useful and important and so we want to be able to capture all
+of them. Despite the differences, the systems can fit the same framework. The contributions of this
+chapter are as follows:
+
+\begin{itemize}
+\item We present a \emph{flat coeffect calculus} with a type system that is parameterized by a 
+  \emph{flat coeffect algebra} and can be instantiated to obtain all of the three examples
+  discussed (Section~\ref{sec:flat-calculus}).
+
+\item We give the equational theory of the calculus and discuss type-preservation for call-by-name
+  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
+  with pairs and recursion (Section~\ref{sec:flat-exts}).
+
+\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which is a
+  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
+  The semantics provides deeper insight into how (and why) the calculus works.
+
+\item We develop an alternative presentation of the system in terms of a simple structure 
+  (semi-lattice) and use it to develop a type inference algorithm for flat coeffect calculus.
+\end{itemize}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Related work}
+
+The development in this chapter can be seen as a counterpart to the well-known development of 
+\emph{effect systems} \cite{effects-gifford} and the use of \emph{monads} \cite{monad-notions}
+in programming languages. The syntax and type system of the flat coeffect calculus follows 
+similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
+in the structure, as explained in the previous chapter, most importantly in lambda abstraction.
+
+Wadler and Thiemann famously show a correspondence between effect systems to monads 
+\cite{monads-effects-marriage}, relating effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ 
+to monadic computations $M^\sigma \tau_1 \rightarrow \tau_2$. In this chapter, we show a similar
+correspondence between \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry 
+of $\lambda$-calculus, this is not a simple mechanical dualization.
+
+The main purpose of the comonadic semantics presented in this chapter is to provide a semantic
+motivation for the flat coeffect calculus. The semantics is inspired by the work of Uustalu and
+Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
+data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
+comonads} annotate the structure with information about the required context, \ie~$C^\sigma \tau_1 \rightarrow \tau_2$.
+This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.
 
@@
 
-This section revisits the flat coeffect calculus, which provides a 
-basis for the structural coeffect calculus that we introduce in Section~... We 
-make a number of improvements -- we cover more standard language features and require additional laws
-which allows a more useful equational theory.
-The formulation is general, so as not to restrict the possible uses unnecessarily, but is
-specialised enough to provide useful properties.
-
-The flat coeffect calculus maintains a single annotation of the contextual requirements associated
-with the variable context. It can be instantiated to capture implicit parameters, the system for 
-dataflow and simple liveness analysis.
-
-The expressions are those of the $\lambda$-calculus with \emph{let} binding; assuming $T$ ranges
-over base types, types are extended as follows:
+The flat coeffect calculus is defined in terms of \emph{flat coeffect algebra}, which defines
+the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
+sets of implicit parameters, integers or other values. The expressions of the calculus are those
+of the $\lambda$-calculus with \emph{let} binding; assuming $T$ ranges over base types, the 
+types of the calculus are defined as follows:
 %
@@
 %
-where $T$ ranges over type constants. 
-We discuss pairs and recursion in Sections~\ref{sec:flat-synprop} and \ref{sec:structural-properties}.
-The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
+We discuss pairs and recursion in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
 represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
-It can be viewed as a pure function that takes $\tau_1$ \emph{with} context $r$.
+It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
+context $r$.
 
-%Note that
-%$\ctxtyp{r}{\tau}$ does not exist as syntactic value (although it has a semantic meaning
-%discussed in Section~\ref{sec:flat-semantics}). The annotations $r$ are formed by 
-%an algebraic structure discussed next.
+In the categorical semantics, the function $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is modelled
+by a morphism $C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the object $C^{\cclrd{r}}$
+does not exist as a syntactical value. This is because we use comonads to define the 
+\emph{semantics} rather than \emph{embedding} them into the language as in the meta-language
+approaches (the distinction between the two approaches has been discussed in detail in 
+Section~\ref{sec:path-sem-langs}). The annotations $\cclrd{r}$ are formed by an algebraic
+structure discussed next.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Reconciling lambda abstraction}
+
+Recall the lambda abstraction rules for the implicit parameters system (annotating the context
+with sets of required parameters) and the data-flow system (annotating the context with the
+number of past required values):
+%
+\begin{equation*}
+\tyrule{abs-imp}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
+\;
+\tyrule{abs-df}
+  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
+\end{equation*}
+
+~
+
+In order to capture both systems using a single calculus, we need a way of unifying the two
+systems. For the data-flow system, this can be achieved by over-approximating the number of 
+required past elements:
+%
+\begin{equation*}
+\tyrule{abs-min}
+  {\coctx{\Gamma, x:\tau_1}{\textnormal{min}(\aclrd{n}, \aclrd{m})} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 }
+\end{equation*}
+%
+The rule (\emph{abs-df}) is admissible in a system that includes the (\emph{abs-min}) rule. 
+If we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on 
+annotations of contexts), then the reverse is also true -- because 
+$\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{m}$ and $\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{n}$.
 
@@
 \subsection{Flat coeffect algebra}
+To make the flat coeffect system general enough, the algebra consists of three operations.
+Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{point-wise} composition, 
+respectively and the third one, $\czip$ represents context \emph{merging}. The term merging should be 
+understood semantically -- the operation models what happens when the semantics of lambda abstraction 
+combines context available at the declaration-site and the call-site.
+
+In addition to the three operations, we also require two special values used to annotate
+variable access and constant access and a relation that defines the ordering.
 
@@
 
+\noindent
+In two of the three systems, some of the operators of the flat coeffect algebra coincide,
+but the data-flow system requires all three. Similarly, the two special elements also 
+coincide in some, but not all systems. The required laws are motivated by the aim to capture
+common properties of the three examples, without unnecessarily restricting the system:
+
+\begin{itemize}
+\item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
+functions. The laws of a monoid are required in order to form a category structure in the 
+categorical model (Section~\ref{sec:flat-semantics}).
+
+\item The monoid $(\C, \cpar, \czero)$ represents \emph{point-wise} composition, \ie~the case when the
+same context is passed to multiple (independent) computations. The monoid laws guarantee 
+that usual syntactic transformations on tuples and the unit value (Section~\ref{sec:flat-exts})
+preserve the coeffect.
+
+\item For the $\czip$ operation, we only require associativity in general. However, the operator can,
+in some cases, form a monoid with the unit being the greatest element of the set. This 
+alternative is discussed when we consider recursion (Section~\ref{sec:flat-exts}).
+\end{itemize}
+
+It is worth noting that the operators $\cpar$ and $\czip$ are dual in some of the systems. For 
+example, in data-flow computations, they are \emph{max} and \emph{min} respectively. However, this
+duality does not hold for implicit parameters. Using the syntactic reading, they represent 
+\emph{merging} and \emph{splitting} of context requirements.
+
+\paragraph{Ordering.}
+
+The flat coeffect algebra requires a pre-order relation $\cleq$, which is used to define 
+sub-coeffecting rule of the type system. When the monoid $(\C, \cpar, \czero)$ is idempotent
+and commutative monoid (semi-lattice), the $\cleq$ relation can be defined in terms of $\cpar$ as:
+%
+\begin{equation*}
+r \;\cleq\; s \;\Longleftrightarrow\; r \;\cpar\; s \;=\; s
+\end{equation*}
+%
+This definition is consistent with all three examples that motivate flat coeffect calculus.
+However, it cannot be used with the structural coeffects (where it fails for the bounded reuse 
+calculus) and so we also do not use it for flat coeffects.
+
+Furthermore, the $\cunit$ coeffect is often the top (greatest) or the bottom (smallest) 
+element of the semi-lattice, but not in general. However, when this is the case, we are able
+to prove certain properties of the calculus (Section~\ref{sec:syntax}).
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma}{\cunit} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c : \tau \in \Delta}
+  {\coctx{\Gamma}{\czero} \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+
+\caption{Type system for the flat coeffect calculus}
+\label{fig:flat-types}
+\end{figure}
+
 %---------------------------------------------------------------------------------------------------
@@
 \subsection{Examples of flat coeffects}
+
+\vspace{-0.4em}
+\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
+union $\cup$ for all three operators. Variable access is annotated with $\cunit = \emptyset$
+and $\leq$ is subset ordering.
+
+\vspace{-0.4em}
+\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
+$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
+and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
+and $\cseq$ are both $\sqcap$ (meet).
+
+\vspace{-0.4em}
+\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
+$\vee, \wedge$ and $\cseq$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
+Variable access is annotated with $\cunit = 0$ and the order $\leq$ is the standard ordering
+of natural numbers.
 
@@
 
+\section{Related work}
+
+\subsection{What can be monad}
+\label{sec:flat-related-monads}
+
+% ==================================================================================================
+
 \section{Summary}
@@
 \section{----------------JUNK!}
-
-\begin{figure}[t]
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
-   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
-\end{equation*}
-
-\caption{Flat coeffect calculus}
-\label{fig:flat-types}
-\end{figure}
-
-% ------------------
-
-\section{Type inference}
 
@@
 % comonadic semantics
-\newcommand{\iunit}{\mathsf{e}}
-\newcommand{\imult}{\oplus}
 \newcommand{\iftor}[1]{C^{#1}}
@@
 \newcommand{\counit}[0]{ \varepsilon }
-% identity function
-
-% structural coeffect calculus - operation names
-
-%---------------------------------------------------------------------------------------------------
-
-\todo{Update the introduction \& edit the whole chapter -- currently, this is just a direct copy
-from the accepted ICALP paper on flat coeffects.}
-
-\todo{Other extensions -- add more examples of indexed comonoads, consider recursion and 
-fixed-point operator, possibly also consider partially defined operations, which lets us
-express type systems that rule out some computations (rather than just static analyses). }
-
-Understanding how programs affect their environment is a well studied area: \emph{effect 
-systems}~\cite{effects-talpin-et-al} provide a static analysis of effects and 
-\emph{monads}~\cite{monad-notions} provide a unified semantics to different notions of effect.  
-Wadler and Thiemann unify the two approaches~\cite{monads-effects-marriage}, \emph{indexing}
-a monad with effect information, and showing that the propagation of effects in an effect system 
-matches the semantic propagation of effects in the monadic approach.
-
-No such unified mechanism exists for tracking the context requirements. We use the term 
-\emph{coeffect} for such contextual program properties. 
-Notions of context have been previously captured using comonads
-\cite{comonads-notions} (the dual of monads) 
-and by languages derived from modal logic
-\cite{logic-modal-reconstruction,logic-cmtt}, but these approaches
-do not capture many useful examples which motivate our work. 
-We build mainly on the former comonadic direction (\S\ref{sec:comonads}) and discuss the
-modal logic approach later (\S\ref{sec:related}).
-
-We extend a simply typed lambda calculus with a coeffect system based on comonads,
-replicating the successful approach of effect systems and monads.
-
-\paragraph{Examples of coeffects.}
-We present three examples that do not fit the traditional approach of
-\emph{effect systems} and have not been considered using the \emph{modal logic}
-perspective, but can be captured as \emph{coeffect systems} (\S\ref{sec:motivation}) --
-the tracking of implicit dynamically-scoped parameters (or resources),
-analysis of variable liveness, and tracking the
-number of required past values in dataflow computations.
-
-\paragraph{Coeffect calculus.}
-Informed by the examples, we identify a general algebraic structure for coeffects. 
-From this, we define a general \emph{coeffect calculus} that unifies the motivating 
-examples (\S\ref{sec:calculus}) and discuss its syntactic properties (\S\ref{sec:syntactic}).
-
-\paragraph{Indexed comonads.}
-Our categorical semantics (\S\ref{sec:comonads}) extends the work of Uustalu and Vene 
-\cite{comonads-notions}. By adding annotations, we generalize comonads to \emph{indexed comonads}, 
-which capture notions of computation not captured by ordinary comonads.
-
-% ==================================================================================================
-
-\section{Generalized coeffect calculus}
-\label{sec:calculus}
-
-The previous three examples exhibit a number of commonalities. We capture these in 
-the \emph{coeffect calculus}. We do not overly restrict the calculus to make
-it open for notions of context-dependent computations not discussed above.
-
-The syntax of our calculus is that of the simply-typed lambda calculus
-(where $v$ ranges over variables, \ident{T} over base types, and $r$ over coeffect annotations):
-%
-\begin{equation*}
-e ::= v \sep \lambda v.e \sep e_1~e_2
-\quad\quad\quad\quad
-\tau ::= \ident{T} \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{r}{\tau}
-\end{equation*}
-
-The type $\ctyp{r}{\tau}$ captures values of type $\tau$ in a context
-specified by the annotation  $r$. This type appears only on the
-left-hand side of a function arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$. 
-In the semantics, $\ctyp{r}$ corresponds to some data type
-(\eg{}~\ident{list} or \ident{Maybe}). Extensions
-such as explicit \emph{let}-binding are discussed later
-(\S\ref{sec:syntactic}).
-
-The coeffect tags $r$, that were demonstrated in the previous section, can be generalized to a 
-structure with three binary operators and a particular element.
-
-\begin{definition}
-A \emph{coeffect algebra} $(S, \imult, \vee, \wedge, \iunit)$ is a set $S$ with
-an element $\iunit \in S$, a semi-lattice $(S, \vee)$, a monoid $(S, \imult, \iunit)$,
-and a binary $\wedge$. That is, $\forall r,s,t\in S$:
-\begin{equation}
-\tag{monoid}
-   r \imult (s \imult t) = (r \imult s) \imult t 
-\quad\quad\quad
-   \iunit \imult r = r = r \imult \iunit  \quad\quad
-\end{equation}
-\vspace{-1.5em}
-\begin{equation}
-\tag{semi-lattice}
-   r \vee s = s \vee r 
-\quad\quad
-   r \vee (s \vee t) = (r \vee s) \vee t
-\quad\quad
-   r \vee r = r
-\end{equation}
-\end{definition}
-%
-The generalized coeffect calculus captures the three motivating examples (\S\ref{sec:motivation}),
-where some operators of the coeffect algebra may coincide.
-
-The $\imult$ operator represents \emph{sequential} composition; guided by the categorical
-model (\S\ref{sec:comonads}), we require it to form a monoid with $\iunit$.
-The operator $\vee$ corresponds to merging of context-requirements in \emph{pointwise composition} 
-and the semi-lattice $(S, \vee)$ defines a partial order: $r \leq s$ when 
-$r \vee s = s$. This ordering implies a sub-coeffecting rule.  
-The coeffect $\iunit$ is often the top or bottom of the lattice.
-
-The $\wedge$ operator corresponds to splitting requirements of a function body. 
-The operator is unrestricted in the general system.
-A number of additional laws holds for \emph{some} coeffects systems, \eg{}~semi-lattice structure of
-$\wedge$ and a form of distributivity. Quite possibly, they should hold for all coeffect systems.  
-We start with as few laws as possible so as not to limit possible uses of the calculus.  
-We consider constrained variants that provide useful syntactic properties later (\S\ref{sec:syntactic}).
-
-\vspace{-0.4em}
-\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
-union $\cup$ for all three operators. Variable access is annotated with $\iunit = \emptyset$
-and $\leq$ is subset ordering.
-
-\vspace{-0.4em}
-\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
-$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
-and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
-and $\imult$ are both $\sqcap$ (meet).
-
-\vspace{-0.4em}
-\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
-$\vee, \wedge$ and $\imult$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
-Variable access is annotated with $\iunit = 0$ and the order $\leq$ is the standard ordering
-of natural numbers.
 
@@
 Figure~\ref{fig:coeffect-typing} shows the
-rules of the coeffect calculus, given some coeffect algebra $(S, \imult, \vee, \wedge, \iunit)$.
-The context required by a variable access (\emph{var}) is annotated with $\iunit$.
+rules of the coeffect calculus, given some coeffect algebra $(S, \cseq, \vee, \wedge, \cunit)$.
+The context required by a variable access (\emph{var}) is annotated with $\cunit$.
 The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
@@
 the coeffects of the expression representing a function $r$ and the coeffects of
-the argument, sequentially composed with the coeffects of the function $s \imult t$.
+the argument, sequentially composed with the coeffects of the function $s \cseq t$.
 
@@
   {x : \tau \in \Gamma}
-  {\ctyp{\iunit}{\Gamma} \vdash x : \tau }
+  {\ctyp{\cunit}{\Gamma} \vdash x : \tau }
 %
@@
    \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
+  {\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 \begin{definition}
-Given a monoid $(S, \imult, \iunit)$ with binary operator $\imult$ and
-unit $\iunit$, an \emph{indexed comonad} over a category $\ccat$
+Given a monoid $(S, \cseq, \cunit)$ with binary operator $\cseq$ and
+unit $\cunit$, an \emph{indexed comonad} over a category $\ccat$
 comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
@@
 \begin{itemize}
-\item a natural transformation $\counit_A : \iftor{\iunit} A \rightarrow A$, called the \emph{counit};
+\item a natural transformation $\counit_A : \iftor{\cunit} A \rightarrow A$, called the \emph{counit};
 \item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
-  r A \rightarrow B$ to \\ morphisms $\iftor{r \imult{} s} A
+  r A \rightarrow B$ to \\ morphisms $\iftor{r \cseq{} s} A
   \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
@@
 \begin{array}{ccc}
-\counit \circ \cobind{f}{r, \iunit} = f & 
-\cobind{{(\counit)}}{\iunit, r} = \ident{id} & 
-\cobind{(g \circ \cobind{f}{r, s})}{(r \imult s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \imult t)} \\
+\counit \circ \cobind{f}{r, \cunit} = f & 
+\cobind{{(\counit)}}{\cunit, r} = \ident{id} & 
+\cobind{(g \circ \cobind{f}{r, s})}{(r \cseq s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \cseq t)} \\
 \end{array}
@@
 \hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
-  \rightarrow (\iftor{r \imult s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
+  \rightarrow (\iftor{r \cseq s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
 g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
@@
 the composition follow from the indexed comonad laws and the 
-monoid $(S, \oplus, \iunit)$.
+monoid $(S, \oplus, \cunit)$.
 
@@
 \\[0.1em]
-\cosem{\ctyp{r \vee (s \imult t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
+\cosem{\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
     (\emph{uncurry} \;  \cosem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
  & & \quad\; (id \times \cobind{\cosem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
-    \circ \csplit_{r, s \imult t} \circ \iftor{r \vee (s \imult t)}\Delta
+    \circ \csplit_{r, s \cseq t} \circ \iftor{r \vee (s \cseq t)}\Delta
 \\[0.1em]
-\cosem{\ctyp{\iunit}{\Gamma} \vdash x_i : \tau_i} &=&
+\cosem{\ctyp{\cunit}{\Gamma} \vdash x_i : \tau_i} &=&
     \pi_i \circ \counit                                                                  % &$(3)$
@@
 %\\[0.1em]
-%\cosem{\ctyp{r\vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
+%\cosem{\ctyp{r\vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
 %    \cosem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
 % ((\varepsilon \times \cosem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
-%\iftorhat{r \vee (r \imult t)}\Delta
+%\iftorhat{r \vee (r \cseq t)}\Delta
 \end{array}
@@
 it is promoted to a family of functors with the derived morphism mapping 
-$\iftor{r}(f) = \cobind{ (f \circ \counit) }{\iunit, r}$.
+$\iftor{r}(f) = \cobind{ (f \circ \counit) }{\cunit, r}$.
 
@@
 context with the argument context for the context of the function body.
-The indices of $\iunit$ for
+The indices of $\cunit$ for
 $\counit$ and $r,s$ for $\cmerge_{r, s}$ match the coeffects
@@
    \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
-  {\ctyp{r_1 \vee (r_2 \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
+  {\ctyp{r_1 \vee (r_2 \cseq s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
 : \tau_2 }
@@
 % precise coeffect. Because $r \wedge r = r$ for all our examples, we
-% can also a coeffect $r \vee (r \imult s)$. Moreover,
+% can also a coeffect $r \vee (r \cseq s)$. Moreover,
 %for our examples (but not necessarily for \emph{all} coeffect
-%systems), $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$ meaning
+%systems), $r \vee (r \cseq s) \leq r_1 \vee (r_2 \cseq s)$ meaning
 %that the following gives more precise coeffects:
@@
    \ctyp{r}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
-  {\ctyp{r \vee (r \imult s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2 : \tau_2 }
+  {\ctyp{r \vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2 : \tau_2 }
 \label{eq:let2}
@@
 systems), this defines a more ``precise'' coeffect with respect to $\leq$
-where $r \vee (r \imult s) \leq r_1 \vee (r_2 \imult s)$.
+where $r \vee (r \cseq s) \leq r_1 \vee (r_2 \cseq s)$.
 
@@
 
-The coeffect $r \vee (r \imult s)$ can
+The coeffect $r \vee (r \cseq s)$ can
 also be simplified for all our examples, leading to more intuitive
@@
 for certain coeffect calculi.  We consider two common special cases
-when the coeffect of variables $\iunit$ is the greatest ($\top$) or
+when the coeffect of variables $\cunit$ is the greatest ($\top$) or
 least ($\bot$) element of the semi-lattice $(S, \vee)$ and derive
@@
 the coeffect algebra satisfies the conditions that 
-$\iunit$ is either the greatest or least element of the semi-lattice,
+$\cunit$ is either the greatest or least element of the semi-lattice,
 $\oplus = \wedge$, and $\oplus$ distributes over $\vee$,
@@
 The above theorem holds for both the liveness and resources examples,
-but not for dataflow.  In the case of liveness, $\iunit$ is the
-greatest element ($r \vee \iunit = \iunit$); in the case of
-resources, $\iunit$ is the \emph{least} element ($r \vee \iunit =
+but not for dataflow.  In the case of liveness, $\cunit$ is the
+greatest element ($r \vee \cunit = \cunit$); in the case of
+resources, $\cunit$ is the \emph{least} element ($r \vee \cunit =
 r$) and the proof relies on the fact that additional
@@
 for coeffects, we need to restrict the $\cmerge_{r, s}$ 
-operation of indexed comonads, so that the first parameter is annotated with $\iunit$ (meaning
-no effects): $\iftor \iunit A \times \iftor r B \rightarrow \iftor{r}(A \times B)$.
+operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
+no effects): $\iftor \cunit A \times \iftor r B \rightarrow \iftor{r}(A \times B)$.
 

### 2014-05-17 text/Chapter04.tex
Flat type system (Chapter 4)

@@
 \subsection{Reconciling lambda abstraction}
+\label{sec:flat-calculus-lambda}
 
@@
 $\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids,
-$(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a semigroup. That is, for all $r,s,t\in \C$:
+$(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup). That is, 
+for all $r,s,t\in \C$:
 %
@@
 \\
+r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &
+r\; \czip\; r = r &
+\textnormal{(band)}   
+\\
 \textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
@@
 \textnormal{(pre-order)}   
-\\
-r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
-\textnormal{(assoc)}   
 \end{array}
@@
 
-\item For the $\czip$ operation, we only require associativity in general. However, the operator can,
-in some cases, form a monoid with the unit being the greatest element of the set. This 
-alternative is discussed when we consider recursion (Section~\ref{sec:flat-exts}).
+\item For the $\czip$ operation, we require associativity and idempotence. The idempotence
+requirement makes it possible to duplicate the coeffects and place the same requirement on both
+call-site and declaration-site, \ie~it makes the (\emph{abs-df}) rule admissible. In some cases, 
+the operator forms a monoid with the unit being the greatest element of the set. This alternative is 
+discussed when we consider recursion (Section~\ref{sec:flat-exts}).
 \end{itemize}
@@
 duality does not hold for implicit parameters. Using the syntactic reading, they represent 
-\emph{merging} and \emph{splitting} of context requirements.
+\emph{merging} and \emph{splitting} of context requirements -- in the (\emph{abs}) rule, 
+$\czip$ appears in the assumption and the combined context requirements of the body are split 
+between two positions in the conclusions; in the (\emph{app}) rule, $\cpar$ appears in the 
+conclusion and combines two context requirements from the assumptions.
 
@@
 %
-This definition is consistent with all three examples that motivate flat coeffect calculus.
-However, it cannot be used with the structural coeffects (where it fails for the bounded reuse 
-calculus) and so we also do not use it for flat coeffects.
+This definition is consistent with all three examples that motivate flat coeffect calculus, but
+it cannot be used with the structural coeffects (where it fails for the bounded reuse 
+calculus) and so we choose not to use it.
 
 Furthermore, the $\cunit$ coeffect is often the top (greatest) or the bottom (smallest) 
-element of the semi-lattice, but not in general. However, when this is the case, we are able
+element of the semi-lattice, but not in general. When this is the case, we are able
 to prove certain properties of the calculus (Section~\ref{sec:syntax}).
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Understanding flat coeffects}
+
+Before looking at the type system in Figure~\ref{fig:flat-types}, let us clarify how the rules
+should be understood. The coeffect calculus provides both analysis of context dependence (type 
+system) and semantics for context (how it is propagated). These two aspects provide different
+ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
+used to define it.
+
+\begin{itemize}
+\item \textbf{Analysis of context dependence.}
+Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context requirements}. This means
+we can over-approximate them and require more than is actually needed at runtime.
+
+Syntactically, the typing rules should be read top-down. In (\emph{app}), the context requirements 
+of multiple assumptions are \emph{merged}; in (\emph{abs}), they are split between the declaration-site
+and the call-site.
+
+\item \textbf{Semantics of context passing.}
+Semantically, coeffect annotations $\cclrd{r}$ mo\-del \emph{contextual capabilities}. This means
+that we can throw away capabilities, if a sub-expression requires fewer than we 
+currently have.
+
+Semantically, the typing rules should be read bottom-up. In application, the capabilities 
+provided to the term $e_1~e_2$ are \emph{split} between the two sub-expressions; in abstraction,
+the capabilities provided by the call-site and declaration-site are \emph{merged} and passed
+to the body.
+\end{itemize}
+
+The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
+position} in the semantic model (Section~\ref{sec:flat-semantics}). It means that we need to be
+careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
+opposite things. To disambiguate, we always use the term \emph{context requirements} when using
+the syntactic view and \emph{context capabilities} or just \emph{available context} when using 
+the semantic view.
 
@@
 \begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
 \tyrule{let}
@@
 \end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\end{equation*}
 
@@
 \subsection{Flat coeffect types}
+The type system for flat coeffect calculus is shown in Figure~\ref{fig:flat-types}. Variables 
+(\emph{var}) and constants (\emph{const}) are annotated with special values provided by the 
+coeffect algebra. Following the top-down syntactic reading, the (\emph{sub}) rule allows us to 
+treat an expression with fewer context requirements as an expression with more context requirements.
+
+The (\emph{abs}) rule is defined as discussed in Section~\ref{sec:flat-calculus-lambda}. The
+body is annotated with context requirements $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
+between the context-requirements on the declaration-site $\cclrd{r}$ and context-requirements on
+the call-site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.
+
+In function application (\emph{app}), context requirements of both expressions and the 
+function are combined as discussed in Chapter~\ref{ch:applications}. The pointwise composition
+$\cpar$ is used to combine the context requirements of the expression representing a function 
+$\cclrd{r}$ and the context requirements of the argument, sequentially composed with the 
+context-requirements of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.
+
+The type system also includes a rule for let-binding. The rule is \emph{not} equivalent to the
+derivation for $(\lambda x.e_2)~e_1$, but it represents one admissible typing derivation. We
+return to let-binding after looking a number of examples. Additional constructs such 
+as recursion and tuples are covered in Section~\ref{sec:flat-exts}.
 
@@
 
-\vspace{-0.4em}
-\paragraph{Implicit parameters} use sets of names $S = \mathcal{P}(\ident{Id})$ as tags with
-union $\cup$ for all three operators. Variable access is annotated with $\cunit = \emptyset$
-and $\leq$ is subset ordering.
-
-\vspace{-0.4em}
-\paragraph{Liveness} uses a two point lattice $S=\{ \ident{D}, \ident{L} \}$ where
-$\ident{D} \sqsubseteq \ident{L}$. Variables are annotated with the top element $e = \ident{L}$ 
-and constants with bottom $\ident{D}$. The $\vee$ operation is $\sqcup$ (join) and $\wedge$
-and $\cseq$ are both $\sqcap$ (meet).
-
-\vspace{-0.4em}
-\paragraph{Dataflow} tags are natural numbers $S = \mathbb{N}$ and operations
-$\vee, \wedge$ and $\cseq$ correspond to \textit{max}, \textit{min} and $+$, respectively. 
-Variable access is annotated with $\cunit = 0$ and the order $\leq$ is the standard ordering
-of natural numbers.
+The flat coeffect calculus generalizes the flat systems discussed in 
+Section~\ref{sec:applications-flat} of the previous chapter. We can instantiate it to a specific
+use just by providing a flat coeffect algebra. The following summary defines the systems for implicit 
+parameters, liveness and data-flow. For the latter two, we obtain more general (but compatible) rule 
+for lambda abstraction.
+
+\begin{example}[Implicit parameters]
+Assuming \ident{Id} is a set of implicit parameter names, the flat coeffect algebra 
+is formed by $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset, \emptyset, \subseteq)$.
+\end{example}
+
+\noindent
+For simplicity, we assume that all parameters have the same type $\rho$ and so the annotations only
+track sets of names. The definition uses set union for all three operations. Both variables and
+constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The 
+definition satisfies the flat coeffect algebra laws because $(S, \cup, \emptyset)$ is an idempotent, 
+commutative monoid. The system has a single additional typing rule for accessing the value of a
+parameter:
+%
+\begin{equation*}
+\tyrule{param}
+  { \ident{?p} \in \cclrd{c} }
+  { \coctx{\Gamma}{\cclrd{c}} \vdash \ident{?p} : \rho }
+\end{equation*}
+%
+The rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of required
+parameters $\cclrd{c}$. As discussed earlier, we use the same type $\rho$ for all parameters, but
+it is easy to define an extension tracking set of parameters with type annotations.
+
+\begin{example}[Liveness]
+Let $L=\{ \ident{L}, \ident{D} \}$ be a two-point lattice such that $\ident{D} \sqsubseteq \ident{L}$
+with a join $\sqcup$ and meet $\sqcap$. The flat coeffect algebra for liveness is then formed by
+$(L, \sqcup, \sqcap, \sqcap, \ident{L}, \ident{D}, \sqsubseteq)$.
+\end{example}
+
+\noindent
+As in Section~\ref{sec:applications-flat-live}, sequential composition $\cseq$ is modelled by 
+the join operation $\sqcup$ and point-wise composition $\cpar$ is modelled by meet $\sqcap$. 
+Two-point lattice is a commutative, idempotent monoid. The distributivity 
+$(r \sqcap s) \sqcup t = (r \sqcup t) \sqcap (s \sqcup t)$ does not hold for \emph{every} 
+lattice, but it trivially holds for a two-point lattice used here.
+
+The definition uses meet $\sqcap$ for the $\czip$ operator that is used by lambda abstraction.
+This means that, when the body is live $\ident{L}$, both declaration-site and call-site are 
+marked as live $\ident{L}$. When the body is dead $\ident{D}$, the declaration-site and call-site
+can be marked as dead $\ident{D}$, or as live $\ident{L}$, which is less precise, but permissible
+over-approximation, which could otherwise be achieved via sub-typing.
+
+\begin{example}[Data-flow]
+In data-flow, context is annotated with natural numbers and the flat coeffect algebra is formed 
+by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$.
+\end{example}
+
+\noindent
+As discussed earlier, sequential composition $\cseq$ is represented by $+$ and point-wise 
+composition $\cpar$ uses $\emph{max}$. For data-flow, we need a third separate operator for
+lambda abstraction. Annotating the body with $\emph{min}(\cclrd{r}, \cclrd{s})$ ensures that
+both call-site and declaration-site annotations are equal or greater than the annotation 
+of the body. As with liveness, this allows over-approximation.
+
+As required by the laws, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
+and $(\mathbb{N}, \mathit{min})$ forms a band. Note that data-flow is our first example where 
+$+$ is not idempotent. The distributivity laws require the following to be the case:
+$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see. Finally, a simple 
+data-flow language includes an additional rule for $\kvd{prev}$:
+%
+\begin{equation*}
+\tyrule{prev}
+  { \coctx{\Gamma}{\cclrd{c}} \vdash e : \tau }
+  { \coctx{\Gamma}{\cclrd{c}+1} \vdash \kvd{prev}~e : \tau }
+\end{equation*}
+%
+As a further example that was not covered earlier, it is also possible to combine liveness analysis
+and data-flow. In the above calculus, $0$ denotes that we require current value, but no previous
+values. However, for constants, we do not even need the current value.
+
+\begin{example}[Optimized data-flow]
+In optimized data-flow, context is annotated with natural numbers extended with the $\bot$ element,
+that is $\mathbb{N}_{\bot} = \mathbb{N} \cup \{\bot \}$ such that $\forall n \in \mathbb{N}. \bot \leq n$.
+The flat coeffect algebra is $(\mathbb{N}_{\bot}, +, \mathit{max}, \mathit{min}, 0, \bot, \leq)$
+where $m + n$ is $\bot$ whenever $m=\bot$ or $n=\bot$ and \emph{min}, \emph{max} treat $\bot$ as the
+least element.
+\end{example}
+
+\noindent
+Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$,
+$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid and $(\mathbb{N}, \emph{min})$ is a band.
+The required distributivity laws also holds for this algebra.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Typing of let binding}
+
+Recall the (\emph{let}) rule in Figure~\ref{fig:flat-types}. It annotates the expression 
+$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context requirements $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
+This is a special case of typing of an expression $(\lambda x.e_2)~e_1$, using the idempotence
+of $\czip$ as follows:
+%
+\begin{equation*}
+\tyrule{app}
+  {\begin{array}{l}
+   \vspace{-1.5em}
+   \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1
+   \end{array} &
+   \tyruler{abs}
+       { \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2 }
+       { \coctx{\Gamma}{\cclrd{s}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+  { \coctx{\Gamma}{\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }    
+\end{equation*}
+%
+This design decision is similar to ML value restriction, but it works the other way round. Our
+\emph{let} binding is more restrictive rather than more general. The choice is motivated by the 
+fact that the typing obtained using the special rule for let-binding is more precise (with respect 
+to sub-coeffecting) for all the examples considered in this chapter. Table~\ref{tab:flat-simplelet}
+shows how the coeffect annotations are simplified for our examples.
+
+\begin{table}[!h]
+\begin{center}
+\begin{tabular}{ | l | c | c |}
+\hline
+& \textbf{\footnotesize Definition\hspace{1em}} & \textbf{\footnotesize Simplified\hspace{1em}} \\ \hline
+\hspace{-1em}{\footnotesize Implicit parameters} & $\cclrd{s} \cup (\cclrd{s} \cup \cclrd{r})$ & $\cclrd{s} \cup \cclrd{r}$ \\ \hline
+\hspace{-1em}{\footnotesize Liveness} & $\cclrd{s} \sqcap (\cclrd{s} \sqcup \cclrd{r})$ & $\cclrd{s}$ \\ \hline
+\hspace{-1em}{\footnotesize Data-flow} & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
+\end{tabular}
+\end{center}
+
+\vspace{-0.5em}
+\caption{Simplified annotation for let binding in sample flat calculi}
+\label{tab:flat-simplelet}
+\end{table}
+
+\noindent
+The simplified annotations directly follow from the definitions of particular flat coeffect 
+algebras. It is perhaps somewhat unexpected that the annotation can be simplified in different
+ways for different examples.
+
+To see that the simplified annotations are \emph{better}, assume that we used arbitrary 
+splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$ rather than idempotence. The
+``Definition'' column would use $\cclrd{s_1}$ and $\cclrd{s_2}$ for the first and second 
+$\cclrd{s}$, respectively. The corresponding simplified annotation (using idempotence) would
+have $\cclrd{s_1}\,\czip\,\cclrd{s_2}$ instead of $\cclrd{s}$. For all our systems, the 
+simplified annotation (on the right) is more precise than the original (on the left):
+%
+\begin{equation*}
+\begin{array}{rclll}
+\cclrd{s_1} \cup (\cclrd{s_2} \cup \cclrd{r}) &\supseteq& (\cclrd{s_1} \cup \cclrd{s_2}) \cup \cclrd{r} 
+  && \textnormal{(implicit parameters)}\\
+\cclrd{s_1} \sqcap (\cclrd{s_2} \sqcup \cclrd{r}) &\sqsupseteq&  (\cclrd{s_1} \sqcap \cclrd{s_2}) 
+  && \textnormal{(liveness)} \\
+\mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r}) &\geq& \mathit{min}(\cclrd{s_1}, \cclrd{s_2}) + \cclrd{r} 
+  &\quad& \textnormal{(data-flow)} \\
+\end{array}
+\end{equation*}
+%
+The inequality cannot be proved from other properties of the flat coeffect algebra. To make
+the flat coeffect system as general as possible, we do not \emph{in general} require it as
+an additional axiom, although the above examples provide reasonable basis for requiring 
+that the specialized annotation for let binding is the least possible annotation for the 
+expression $(\lambda x.e_2)~e_1$.
 
@@
 \label{sec:flat-semantics}
+
+a
 
@@
 \newcommand{\counit}[0]{ \varepsilon }
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\subsection{Coeffect typing rules.}
-Figure~\ref{fig:coeffect-typing} shows the
-rules of the coeffect calculus, given some coeffect algebra $(S, \cseq, \vee, \wedge, \cunit)$.
-The context required by a variable access (\emph{var}) is annotated with $\cunit$.
-The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
-to be generalized.
-
-The (\emph{abs}) rule checks the body of the function in a context $r \wedge s$, which is a 
-combination of the coeffects available in the context $r$ where the function is 
-defined and in a context provided by the caller of the function. 
-Note that none of the judgements create a \emph{value} of type $\ctyp{r}{\tau}$. This
-type appears only immediately to the left of  an arrow $\ctyp{r}{\tau_1} \rightarrow \tau_2$.
-
-In function application (\emph{app}), context requirements of both expressions and the 
-function are combined as previously: the pointwise composition $\vee$ is used to combine
-the coeffects of the expression representing a function $r$ and the coeffects of
-the argument, sequentially composed with the coeffects of the function $s \cseq t$.
-
-For space reasons, we omit recursion. We note that this would require adding
-effect variables and extending the coeffect algebra with a fixed point operation.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-1em}
-\begin{equation*}
-\inference[(\emph{var})\;]
-  {x : \tau \in \Gamma}
-  {\ctyp{\cunit}{\Gamma} \vdash x : \tau }
-%
-\quad\quad
-%
-\inference[(\emph{app})\;]
-  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
-   \ctyp{t}{\Gamma} \vdash e_2 : \tau_1 }
-  {\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\inference[(\emph{sub})\;]
-  {\ctyp{s}{\Gamma} \vdash e : \tau }
-  {\ctyp{r}{\Gamma} \vdash e : \tau }
-\;(s\leq r)  
-%
-\quad\quad\quad\quad
-%
-\inference[(\emph{abs})\;]
-  {\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\end{equation*}
-\caption{Type and coeffect system for the coeffect calculus}
-\label{fig:coeffect-typing}
-\vspace{-1em}
-\end{figure}
 

### 2014-05-19 text/Chapter04.tex
Categorical semantics of flat calculus (Chapter 4)

@@
 
-a
-
-\subsection{Comonads}
-\subsection{Indexed comonads}
-\subsection{Semantics of flat calculus}
+The type system of flat coeffect calculus arises as a generalization of the examples discussed in 
+Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical semantics
+of context-dependent computations. This is a direction that we explore in this section. Although
+the development presented here is interesting in its own, our main focus is \emph{using} categorical
+semantics to motivate and explain the design of flat coeffect calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Categorical semantics}
+
+As discussed in Section~\ref{sec:path-sem}, categorical semantics interprets terms as morphisms
+in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets typing 
+judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
+$\sem{\tau_1} \times \ldots \times \sem{\tau_n} \rightarrow \sem{\tau}$.
+
+As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful 
+computations can be captured uniformly using the (\emph{strong}) \emph{monad} structure. In that 
+approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
+for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models 
+partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
+monad) and $\mtyp{}{\alpha} = (\alpha \times S)^S$ models side-effects (state monad). Here, the 
+structure of a strong monad provides necessary ``plumbing'' for composing monadic computations.
+
+Following similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that 
+(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent 
+computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
+$\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$ are modelled using
+the non-empty list comonad.
+
+The monadic and comonadic model outlined here represents at most a binary analysis of effects or 
+context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no 
+context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
+$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context. In the next section, we introduce
+\emph{indexed comonads}, which provide a more precise analysis and let us model computations with
+context requirements $\cclrd{r}$ as functions $\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using
+an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Introducing comonads}
+
+In category theory, \emph{comonad} is a dual of \emph{monad}. Informally, we get a comonad by 
+taking a monad and ``reversing the arrows''. More formally, one of the equivalent definitions of
+comonad looks as follows:
+
+\begin{definition}
+A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
+\begin{compactitem}
+\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
+\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
+\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
+  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
+\end{compactitem}
+such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
+\begin{align}
+\tag{\emph{left identity}}
+  \ident{cobind}~\ident{counit} &= \idf{}
+  \\
+\tag{\emph{right identity}}
+  \ident{counit} \circ \ident{cobind}~f &= f
+  \\
+\tag{\emph{associativity}}
+  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
+\end{align}
+\end{definition}
+
+\noindent
+From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
+\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries 
+additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function 
+$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
+the context-dependent function to value(s) in the context and then propagates the context.
+
+As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
+computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.
+
+\begin{example}[Non-empty list]
+A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
+We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The 
+type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
+%
+\begin{equation*}
+\begin{array}{rclll}
+\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
+\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
+\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
+\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{counit} operation returns the head of the non-empty list. Note that it is crucial that
+the list is \emph{non-empty}, because we always need to be able to obtain a value. The \ident{cobind}
+defined here returns a list of the same length as the original where, for each element, the 
+function $f$ is applied on a \emph{suffix} list starting from the element. Using a simplified
+notation for list, the result of applying \ident{cobind} to a function that sums elements of a
+list gives the following behaviour:
+%
+\begin{equation*}
+\ident{cobind}~\ident{sum}~(7,6,5,4,3,2,1,0) = (28,21,15,10,6,3,1,0)
+\end{equation*}
+%
+The fact that the function $f$ is applied to a \emph{suffix} is important in order to satisfy the
+\emph{left identity} law, which requires that $\ident{cobind}~\ident{counit}~l = l$.
+
+It is also interesting to examine some data types that do \emph{not} form a comonad. As already
+mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
+because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. Similarly,
+the \ident{Maybe} data type defined as $1 + \alpha$ is not a comonad for the same reason.
+However, if we consider flat coeffect calculus for liveness, it appears natural to model computations
+as function $\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such model, we first need to 
+generalise comonads to \emph{indexed comonads}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Generalising to indexed comonads}
+The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
+sequential composition, where the annotation $\cunit$ represents a variable access. An indexed 
+comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the annotation 
+$\cclrd{r}$ determines what context is required.
+
+\begin{definition}
+Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
+\emph{indexed comonad} over a category $\catc$ is a triple 
+$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where:
+
+\begin{compactitem}
+\item $\ctyp{\cclrd{r}}{}$ for all $\cclrd{r} \in \C$ is a family of object mappings 
+\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\cunit}{\alpha} \rightarrow \alpha$ 
+\item $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ is a mapping $(\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
+  \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})$
+\end{compactitem}
+such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$
+and the identity $\idf{\cclrd{s}} : \ctyp{\cclrd{s}}\alpha \rightarrow \ctyp{\cclrd{s}}\alpha$:
+\begin{align}
+\tag{\emph{left identity}}
+  \ident{cobind}_{\cunit, \cclrd{s}}~\ident{counit}_{\cunit} &= \idf{}
+  \\
+\tag{\emph{right identity}}
+  \ident{counit}_{\cunit} \circ \ident{cobind}_{\cclrd{r}, \cunit}~f &= f
+  \\
+\tag{\emph{associativity}}
+\hspace{-10em}
+  \ident{cobind}_{\cclrd{r}\cseq\cclrd{s},\cclrd{t}}~(g \circ \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f) &= 
+    (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g) \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}~f)
+\end{align}
+\end{definition}
+
+\noindent
+Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings 
+$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the operation $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
+operation is now also formed by a \emph{family} of mappings for different pairs of indices 
+$\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations 
+and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
+For the purpose of this thesis, it is sufficient to treat $\ident{cobind}$ as a family of
+mappings or, when it does not  introduce ambiguity, view it as a single mapping.
+
+The $\ident{counit}$ operation is not defined for all $\cclrd{r} \in \C$, but only for 
+the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
+but this is merely for symmetry. Crucially, this means that the operation is defined only
+for some special contexts.
+
+If we look at the indices in the laws, we can see that the left and right identity 
+require $\cunit$ to be the unit of $\cseq$. Similarly, the associativity law implies the 
+associativity of the $\cseq$ operator.
+
+The category that models sequential composition is formed by the unit arrow $\ident{counit}$ 
+together with the (associative) composition operation that composes computations with 
+contextual requirements as follows:
+%
+\begin{equation*}
+\begin{array}{ccl}
+\textnormal{--}\, \hat{\circ} \,\textnormal{--}&\narrow{:}& (\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2) 
+  \rightarrow (\ctyp{\cclrd{s}}{\tau_2} \rightarrow \tau_3) 
+  \rightarrow (\ctyp{\cclrd{r} \cseq \cclrd{s}}{\tau_1} \rightarrow \tau_3) \\
+g \, \hat{\circ} \, f &\narrow{=}& g \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}} f)
+\end{array}
+\end{equation*}
+%
+The composition $\hat{\circ}$ best expresses the intention of indexed comonads. Given two functions
+with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
+requires $\cclrd{r}\,\cseq,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
+are attached to the input of the composed function.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Examples}
 
-% ==================================================================================================
-
-\section{Equational theory}
-\label{sec:flat-syntax}
-
-\subsection{Call-by-value evaluation}
-\subsection{Call-by-name evaluation}
-
-% ==================================================================================================
-
-\section{Syntactic extensions}
-\label{sec:flat-exts}
-
-\subsection{Lambda abstraction}
-\subsection{Constants and pairs}
-\subsection{Recursion}
-
-% ==================================================================================================
-
-\section{Type inference}
-\label{sec:flat-inference}
-
-\subsection{Semi-lattice formulation}
-\subsection{Type inference algorithm}
-
-% ==================================================================================================
-
-\section{Related work}
-
-\subsection{What can be monad}
-\label{sec:flat-related-monads}
-
-% ==================================================================================================
-
-\section{Summary}
-
-\section{----------------JUNK!}
-
-% ==================================================================================================
-
-~
-\newpage
-~
-
-%---------------------------------------------------------------------------------------------------
-
-% comonadic semantics
-\newcommand{\iftor}[1]{C^{#1}}
-\newcommand{\iftorhat}[1]{\hat{C}^{#1}}
-
-\newcommand{\cosem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
+Any comonad can be turned into an indexed comonad using a trivial monoid. However, indexed comonads
+are more general and can be used with other data types, including indexed \ident{Maybe}.
+
+\begin{example}[Comonads]
+Any comonad $\ctyp{}{}$ is an indexed comonad with an index provided by a trivial monoid $(\{1\},\ast,1)$
+where $1\ast 1 = 1$ and $\ctyp{1}{}$ is the underlying mapping $\ctyp{}{}$ of the original comonad. The
+operations $\ident{counit}_1$ and $\ident{cobind}_{1,1}$ are defined by the operations $\ident{counit}$
+and $\ident{cobind}$ of the comonad.
+\end{example}
+
+\begin{example}[Indexed option]
+The indexed option comonad is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
+where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
+Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
+\begin{equation*}
+\begin{array}{l}
+\ctyp{\ident{L}}{\alpha} = \alpha\\
+\ctyp{\ident{D}}{\alpha} = 1\\
+\\
+\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
+\ident{counit}_{\ident{L}}~v = v\\
+\end{array}
+\qquad
+\begin{array}{lcl}
+\multicolumn{3}{l}{
+  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
+    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
+\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
+\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
+\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
+\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The indexed option comonad models the semantics of the liveness coeffect system discussed in 
+\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
+and $\ctyp{\ident{D}}{\alpha}=1$ models a dead context which does not contain a value. The \ident{counit}
+operation extracts a value from a live context; \ident{cobind} can be seen as an implementation of 
+dead code elimination. The definition only evaluates $f$ when the result is marked as live and is thus
+required, and it only accesses $x$ if the function $f$ requires its input.
+
+The indexed family $\ctyp{\cclrd{r}}{}$ in the above example is analogous to the \ident{Maybe}
+(or option) data type $\ident{Maybe}\,\alpha = 1 + \alpha$. As mentioned earlier, this type does not 
+permit (non-indexed) comonad structure, because $\ident{counit}~()$ is not defined. This is not a 
+problem with indexed comonads, because \ident{counit} only needs to be defined on live context.
+
+\begin{example}[Indexed product]
+The semantics of implicit parameters
+
+\end{example}
+
+\paragraph{Related work}
+
+Not a family of comonads
+
+Katsumata
+
 \newcommand{\ccat}[0]{\mathcal{C}}
-\newcommand{\obj}[1]{\textnormal{obj}(#1)}
-
 \newcommand{\cobind}[2]{#1^\dagger_{#2}}
@@
 
-% ==================================================================================================
-
-\section{Coeffect semantics using indexed comonads}
-\label{sec:comonads}
-
-The approach of \emph{categorical semantics} interprets terms as
-morphisms in some category.  For typed calculi, typing judgments $x_1 : \tau_1 \ldots x_n : \tau_n
-\vdash e: \tau$ are usually mapped to morphisms $\interp{\tau_1} \times \ldots \times
-\interp{\tau_n} \rightarrow \interp{\tau}$. Moggi showed the semantics of various effectful 
-computations can be captured generally using the (\emph{strong}) \emph{monad}
-structure~\cite{monad-notions}. Dually, Uustalu and Vene showed that
-(\emph{monoidal}) \emph{comonads} capture various kinds of context-dependent 
-computation~\cite{comonads-notions}.
-
-We extend Uustalu and Vene's approach to give a semantics for
-the coeffect calculus by generalising comonads to
-\emph{indexed comonads}. We emphasise semantic intuition and
-abbreviate the categorical foundations for space reasons.
-
-\vspace{-1em}
-\paragraph{Indexed comonads.}
-
-Uustalu and Vene's approach interprets well-typed terms 
-as morphisms $C (\tau_1 \times \ldots
-\times \tau_n) \rightarrow \tau$, where $C$ encodes contexts
-and has a comonad structure~\cite{comonads-notions}.
-Indexed comonads comprise a \emph{family} of object mappings 
-$C^r$ indexed by a coeffect $r$ describing the contextual requirements satisfied by
-the encoded context. We interpret judgments $C^r (x_1 : \tau_1, \ldots, x_n : \tau_n) \vdash e : \tau$ as
-morphisms $C^r (\interp{\tau_1} \times \ldots \times \interp{\tau_n}) \rightarrow \interp{\tau}$.
-
-The indexed comonad structure provides a notion of composition
-for computations with different contextual requirements.
-%%
-\begin{definition}
-Given a monoid $(S, \cseq, \cunit)$ with binary operator $\cseq$ and
-unit $\cunit$, an \emph{indexed comonad} over a category $\ccat$
-comprises a family of object mappings $\iftor r$ where for all $r \in S$ and $A \in \obj{\ccat}$ then 
-$\iftor r A \in \obj{\ccat}$ and:
-%%
-\begin{itemize}
-\item a natural transformation $\counit_A : \iftor{\cunit} A \rightarrow A$, called the \emph{counit};
-\item a family of mappings $\cobind{(-)}{r, s}$ from morphisms $\iftor
-  r A \rightarrow B$ to \\ morphisms $\iftor{r \cseq{} s} A
-  \rightarrow \iftor s B$ in $\ccat$, natural in $A,B$, called
-  \emph{coextend};
-\end{itemize}
-%
-\noindent
-such that for all $f : \iftor r \tau_1 \rightarrow \tau_2$ and $g : \iftor s \tau_2 \rightarrow \tau_3$ 
-the following equations hold:
-%
-\renewcommand{\arraycolsep}{1.5em}
-\begin{align*}
-\begin{array}{ccc}
-\counit \circ \cobind{f}{r, \cunit} = f & 
-\cobind{{(\counit)}}{\cunit, r} = \ident{id} & 
-\cobind{(g \circ \cobind{f}{r, s})}{(r \cseq s), t} = \cobind{g}{s, t} \circ \cobind{f}{r, (s \cseq t)} \\
-\end{array}
-\end{align*}
-\end{definition}
-%
-The \emph{coextend} operation gives rise to an associative composition operation for
-computations with contextual requirements (with \emph{counit} as the identity):
-%
-\[
-\hat{\circ} : (\iftor r \tau_1 \rightarrow \tau_2) \rightarrow (\iftor s \tau_2 \rightarrow \tau_3) 
-  \rightarrow (\iftor{r \cseq s} \tau_1 \rightarrow \tau_3) \quad\quad\quad
-g \, \hat{\circ} \, f = g \circ f^\dagger_{r,s}\quad
-\]
-%
-The composition $\hat{\circ}$ best expresses the intention
-of indexed comonads: contextual requirements of the composed
-functions are combined. The properties of
-the composition follow from the indexed comonad laws and the 
-monoid $(S, \oplus, \cunit)$.
-
 \paragraph{Example}
@@
 are a special case of indexed comonads with a trivial singleton
-monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\iftor 1$
+monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\ccat^1$
 is the underlying functor of the comonad and $\counit$ and
@@
 \emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
-\sqcap, \ident{L})$, where $\iftor{\ident L} A = A$ encodes live
-contexts and $\iftor{\ident D} A = 1$ encodes dead contexts,
+\sqcap, \ident{L})$, where $\ccat^{\ident L} A = A$ encodes live
+contexts and $\ccat^{\ident D} A = 1$ encodes dead contexts,
 where $1$ is the unit type inhabited by a single value $()$.
-The \emph{counit} operation $\counit{} : \iftor{\ident{L}} A \rightarrow A$
+The \emph{counit} operation $\counit{} : \ccat^{\ident{L}} A \rightarrow A$
 is defined $\counit\ x = x$ and \emph{coextend},
-for all $f : \iftor r A \rightarrow B$, and thus
-$\cobind{f}{r, s} : \; \iftor{r \sqcap s} A \rightarrow \iftor s B$,
+for all $f : \ccat^r A \rightarrow B$, and thus
+$\cobind{f}{r, s} : \; \ccat^{r \sqcap s} A \rightarrow \ccat^s B$,
 is defined:
@@
 For the indexed comonad, $\counit$ need only be defined
- for $\iftor{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
+ for $\ccat^{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
  contextual notions of computation than comonads.
@@
  context. For example, in the second case above $f^\dagger_{\ident{D},
-   \ident{L}} : \iftor{\ident{D}} A \rightarrow \iftor{\ident{L}} B$
+   \ident{L}} : \ccat^{\ident{D}} A \rightarrow \ccat^{\ident{L}} B$
  where the shape changes from $1$ (empty context) to $B$ (available
  context). 
+a
+
+\subsection{Indexed monoidal comonads}
+
+!!!
+
+not a family of comonads
+
+\subsection{Semantics of flat calculus}
+
+% ==================================================================================================
+
+\section{Equational theory}
+\label{sec:flat-syntax}
+
+\subsection{Call-by-value evaluation}
+\subsection{Call-by-name evaluation}
+
+% ==================================================================================================
+
+\section{Syntactic extensions}
+\label{sec:flat-exts}
+
+\subsection{Lambda abstraction}
+\subsection{Constants and pairs}
+\subsection{Recursion}
+
+% ==================================================================================================
+
+\section{Type inference}
+\label{sec:flat-inference}
+
+\subsection{Semi-lattice formulation}
+\subsection{Type inference algorithm}
+
+% ==================================================================================================
+
+\section{Related work}
+
+\subsection{What can be monad}
+\label{sec:flat-related-monads}
+
+% ==================================================================================================
+
+\section{Summary}
+
+\section{----------------JUNK!}
+
+% ==================================================================================================
+
+~
+\newpage
+~
+
+% ==================================================================================================
+
+\section{Coeffect semantics using indexed comonads}
+\label{sec:comonads}
 
@@
 \begin{itemize}
-\item $\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \wedge s)}(A \times B)$ natural
+\item $\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \wedge s)}(A \times B)$ natural
 in $A, B$;
-\item $\csplit_{r, s} : \iftor{(r \vee s)}(A \times B) \rightarrow \iftor r A \times \iftor s B$ natural
+\item $\csplit_{r, s} : \ccat^{(r \vee s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B$ natural
 in $A, B$;
@@
 %algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
-%is provided by the indexed comonad $\iftor rA = A \times (r \rightarrow \tau)$
+%is provided by the indexed comonad $\ccat^rA = A \times (r \rightarrow \tau)$
 %where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
@@
 %\begin{array}{l}
-%\cobind{f}{r, s} \; : \; \iftor{r \cup s} A \rightarrow \iftor s B\\
+%\cobind{f}{r, s} \; : \; \ccat^{r \cup s} A \rightarrow \ccat^s B\\
 %\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
@@
 %\counit{} (a, g) = a\\
-%%\iota_{r, s} \; : \: \iftor r A \rightarrow \iftor s A\\
+%%\iota_{r, s} \; : \: \ccat^r A \rightarrow \ccat^s A\\
 %\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
@@
 %\quad\begin{array}{l}
-%\cmerge_{r, s} : \iftor r A \times \iftor s B \rightarrow \iftor{(r \cup s)}(A \times B)\\
+%\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \cup s)}(A \times B)\\
 %\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
 %\\[1em]
-%\csplit_{r, s} : \iftor{(r \cup s)}(A \times B) \rightarrow \iftor r A \times \iftor s B \\
+%\csplit_{r, s} : \ccat^{(r \cup s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B \\
 %\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
@@
 % \begin{equation*} 
-% \cosem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
+% \sem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
 %   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
@@
 \begin{array}{rcll}
-\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \emph{curry} \; (\cosem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
+\sem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
+    \emph{curry} \; (\sem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
 \\[0.1em]
-\cosem{\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
-    (\emph{uncurry} \;  \cosem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
- & & \quad\; (id \times \cobind{\cosem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
-    \circ \csplit_{r, s \cseq t} \circ \iftor{r \vee (s \cseq t)}\Delta
+\sem{\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
+    (\emph{uncurry} \;  \sem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
+ & & \quad\; (id \times \cobind{\sem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
+    \circ \csplit_{r, s \cseq t} \circ \ccat^{r \vee (s \cseq t)}\Delta
 \\[0.1em]
-\cosem{\ctyp{\cunit}{\Gamma} \vdash x_i : \tau_i} &=&
+\sem{\ctyp{\cunit}{\Gamma} \vdash x_i : \tau_i} &=&
     \pi_i \circ \counit                                                                  % &$(3)$
 %\\[0.1em]
-%\cosem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
-%    \cosem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
+%\sem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
+%    \sem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
 %\\[0.1em]
-%\cosem{\ctyp{r\vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
-%    \cosem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
-% ((\varepsilon \times \cosem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
+%\sem{\ctyp{r\vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
+%    \sem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
+% ((\varepsilon \times \sem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
 %\iftorhat{r \vee (r \cseq t)}\Delta
@@
  $\Delta : A \rightarrow A \times A$ duplicating a
-value. While $\iftor{r}$ is a family of object mappings,
+value. While $\ccat^{r}$ is a family of object mappings,
 it is promoted to a family of functors with the derived morphism mapping 
-$\iftor{r}(f) = \cobind{ (f \circ \counit) }{\cunit, r}$.
+$\ccat^{r}(f) = \cobind{ (f \circ \counit) }{\cunit, r}$.
 
@@
 by an operation 
-$\iota_{r, s} : \iftor r A \rightarrow \iftor s A$ natural in $A$, for all $r, s \in S$ where $s \leq r$,
+$\iota_{r, s} : \ccat^r A \rightarrow \ccat^s A$ natural in $A$, for all $r, s \in S$ where $s \leq r$,
 which transforms a value $C^r A$ to $C^s A$ by
@@
 operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
-no effects): $\iftor \cunit A \times \iftor r B \rightarrow \iftor{r}(A \times B)$.
+no effects): $\ccat^{\cunit} A \times \ccat^r B \rightarrow \ccat^{r}(A \times B)$.
 

### 2014-05-20 text/Chapter04.tex
Adding product and related notions (Chapter 4)

@@
 
-\subsection{Examples}
+\subsection{Examples of indexed comonads}
 
@@
 Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
+%
 \begin{equation*}
@@
 \begin{example}[Indexed product]
-The semantics of implicit parameters
-
+The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
+$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
+As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
+= \alpha \times (\cclrd{r} \rightarrow R)$ represents a value $\alpha$ together with a function that
+associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
+The cobind and counit operations are defined as:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
+\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
+\end{array}
+\quad
+\begin{array}{l}
+\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
+    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
+\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
+\end{array}
+\end{equation*}
 \end{example}
 
-\paragraph{Related work}
-
-Not a family of comonads
-
-Katsumata
+\noindent
+The definition of \ident{counit} simply ignores the function and returns the value in the context.
+The \ident{cobind} operation uses the restriction operation $\restr{f}{\cclrd{r}}$, which we 
+already defined when discussing semantics of implicit parameters in Section~\ref{sec:applications-flat-impl}
+(indeed, \ident{cobind} here captures an essential part of the semantics).
+
+The function $g$ in \ident{cobind} is defined on the union of the implicit parameters, 
+\ie~$\cclrd{r}\cup\cclrd{s}\rightarrow\rho$. When passing it to $f$, we restrict it to
+just $\cclrd{r}$ and when returning it as a result, we restrict it to $\cclrd{s}$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Properties and related notions}
+
+We discuss additional examples in Section~\ref{sec:flat-semantics-monoidal}, after we look at the
+remaining structure that is needed to define the semantics of flat coeffect calculus. Before doing 
+so, we discuss additional properties and categorical structures that have been proposed mainly in 
+the context of monads and effects and are related to indexed comonads.
+
+\paragraph{Shape preservation.}
+Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
+this means that the shape of the additional context does not change during the computation. For
+example, in the \ident{NEList} comonad, the length of the list stays the same after applying 
+\ident{cobind}.
+
+Indexed comonads are not restricted by this property of comonads. For example, given the indexed
+product monad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$ above, the shape of
+the context changes from containing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to containing
+just implicit parameters $\cclrd{s}$.
+
+\paragraph{Families of monads.}
+When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
+\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
+comonads}, is defined as follows.
+
+\begin{definition}
+A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
+  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
+$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
+$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
+certain coherence conditions.
+\end{definition}
+
+Family of comonads is more restrictive than indexed comonad, because each of the data types needs
+to form a comonad separately. For example, our indexed option does not form a family of comonads
+(again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given a 
+family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
+an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
+we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
+$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
+
+\paragraph{Parameteric effect monads.}
+Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
+indexed comonads) are closely related to our definition.  Although presented in a more general 
+categorical framework (and using monads), the model defines \ident{unit} operation only on the 
+unit of a monoid and \ident{bind} operation composes effect annotations using the provided monoidal 
+structure.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Indexed monoidal comonads}
+\label{sec:flat-semantics-monoidal}
 
@@
 
-\paragraph{Example}
-Indexed comonad are analogous to comonads (in coKleisli form), but
-with the additional monoidal structure on indices.  Indeed, comonads
-are a special case of indexed comonads with a trivial singleton
-monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\ccat^1$
-is the underlying functor of the comonad and $\counit$ and
-$\cobind{(-)}{1, 1}$ are the usual comonad operations.
-However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.
-
-\paragraph{Example}
-The \emph{indexed partiality comonad} encodes free-variable contexts
-of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
-\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
-\sqcap, \ident{L})$, where $\ccat^{\ident L} A = A$ encodes live
-contexts and $\ccat^{\ident D} A = 1$ encodes dead contexts,
-where $1$ is the unit type inhabited by a single value $()$.
-The \emph{counit} operation $\counit{} : \ccat^{\ident{L}} A \rightarrow A$
-is defined $\counit\ x = x$ and \emph{coextend},
-for all $f : \ccat^r A \rightarrow B$, and thus
-$\cobind{f}{r, s} : \; \ccat^{r \sqcap s} A \rightarrow \ccat^s B$,
-is defined:
-%
-\begin{align*}
-\begin{array}{llll}
-\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
-\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
-\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
-\cobind{f}{\ident{L}, \ident{L}} x = f~x 
-\end{array}
-\end{align*}
-%
-The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
-data type $\mathsf{Maybe}\ A = A + 1$.  This
-type does not permit a comonad structure since
-$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
-For the indexed comonad, $\counit$ need only be defined
- for $\ccat^{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
- contextual notions of computation than comonads.
-
- Moreover, indexed comonads are not restricted by the \emph{shape
-   preservation} property of comonads~\cite{orchard12codo}: that
- a coextended function cannot change the \emph{shape} of the
- context. For example, in the second case above $f^\dagger_{\ident{D},
-   \ident{L}} : \ccat^{\ident{D}} A \rightarrow \ccat^{\ident{L}} B$
- where the shape changes from $1$ (empty context) to $B$ (available
- context). 
-a
-
-\subsection{Indexed monoidal comonads}
-
-!!!
-
-not a family of comonads
-
-\subsection{Semantics of flat calculus}
-
-% ==================================================================================================
-
-\section{Equational theory}
-\label{sec:flat-syntax}
-
-\subsection{Call-by-value evaluation}
-\subsection{Call-by-name evaluation}
-
-% ==================================================================================================
-
-\section{Syntactic extensions}
-\label{sec:flat-exts}
-
-\subsection{Lambda abstraction}
-\subsection{Constants and pairs}
-\subsection{Recursion}
-
-% ==================================================================================================
-
-\section{Type inference}
-\label{sec:flat-inference}
-
-\subsection{Semi-lattice formulation}
-\subsection{Type inference algorithm}
-
-% ==================================================================================================
-
-\section{Related work}
-
-\subsection{What can be monad}
-\label{sec:flat-related-monads}
-
-% ==================================================================================================
-
-\section{Summary}
-
-\section{----------------JUNK!}
-
-% ==================================================================================================
-
-~
-\newpage
-~
-
-% ==================================================================================================
-
-\section{Coeffect semantics using indexed comonads}
-\label{sec:comonads}
-
-\vspace{-1em}
-\subsection{Monoidal indexed comonads.}
 Indexed comonads provide a semantics to sequential composition, but
@@
 for splitting).
+
+\subsection{Semantics of flat calculus}
+
+% ==================================================================================================
+
+\section{Equational theory}
+\label{sec:flat-syntax}
+
+\subsection{Call-by-value evaluation}
+\subsection{Call-by-name evaluation}
+
+% ==================================================================================================
+
+\section{Syntactic extensions}
+\label{sec:flat-exts}
+
+\subsection{Lambda abstraction}
+\subsection{Constants and pairs}
+\subsection{Recursion}
+
+% ==================================================================================================
+
+\section{Type inference}
+\label{sec:flat-inference}
+
+\subsection{Semi-lattice formulation}
+\subsection{Type inference algorithm}
+
+% ==================================================================================================
+
+\section{Related work}
+
+\subsection{What can be monad}
+\label{sec:flat-related-monads}
+
+% ==================================================================================================
+
+\section{Summary}
+
+\section{----------------JUNK!}
+
+% ==================================================================================================
+
+~
+\newpage
+~
+
+% ==================================================================================================
 

### 2014-05-21 text/Chapter04.tex
Minor additions to categorical semantics (Chapter 4)

@@
 
-\subsection{Examples of indexed comonads}
+\paragraph{Examples.}
 
@@
 \begin{definition}
+\label{def:flat-family}
 A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
@@
 
+Indexed comonads model the semantics of sequential composition, but additional structure is needed
+to model the semantics of the flat coeffect calculus. This is where the duality between monads and 
+comonads can no longer help us, because context is propagated differently than effects in lambda 
+abstraction and application.
+
+Whereas Moggi~\cite{monad-notions} requires \emph{strong} monad to model effectful $\lambda$-calculus, 
+Uustalu and Vene~\cite{comonads-notions} require \emph{lax semi-monoidal} comonad to model 
+$\lambda$-calculus with contextual properties. The structure requires a monoidal operation:
+%
+\begin{equation*}
+\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
+\end{equation*}
+%
+The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
+contexts and is used to merge the context of the declaration-site (containing free variables)
+and the call-site (containing bound variable). For example, for implicit parameters, this combines
+the additional parameters defined in the two contexts.
+
+The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
+\emph{splitting} of contexts. These are provided by \emph{lax} and \emph{oplax} monoidal 
+structures. In addition, we also need a lifting operation (similar to $\iota$ from 
+Definition~\ref{def:flat-family}) to model sub-coeffecting.
+
+\begin{definition}
+Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+an \emph{indexed monoidal comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
+equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
+and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
+%
+\begin{compactitem}
+\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
+  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
+\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
+  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
+\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
+  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\end{compactitem}
+%
+% We should require lift to be comonad morphism
+%
+% \tag{\emph{comonad morphism}}
+%   \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f \circ \ident{lift}_{\cclrd{r'}\cseq\cclrd{s'}, \cclrd{r}\cseq\cclrd{s}} &=
+%   \ident{lift}_{\cclrd{s'}, \cclrd{s}} \circ \ident{cobind}_{\cclrd{r'}, \cclrd{s'}}~(f \circ \ident{lift}_{\cclrd{r'}, \cclrd{r}} )
+%
+% ... but adapting the coherence conditions for counit/cobind is ugly!
+%
+\end{definition}
+
+\noindent
+The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
+values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combineds them into a 
+single value with additional context $\cclrd{r}\czip\cclrd{s}$. The $\czip$ operation often represents
+\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
+bounds for the liveness and data-flow examples, but not for implicit parameters. However, they are useful 
+as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
+the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.
+
+The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
+into two separate values. Note that this does not simply duplicate the value, because the additional
+context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide 
+\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often 
+the \emph{least upper-bound}\footnotemark[1].
+
+Finally, $\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ is a family of operations that ``forget'' some part of
+a context. This models the sub-coeffecting operation and lets us, for example, forget some of the
+available implicit parameters, or turn a live context (containing a value) into a dead context (empty).
+
+\paragraph{Simpler.}
+
+\paragraph{Examples.}
+
+\subsection{Semantics of flat calculus}
+
+% ==================================================================================================
+
 \newcommand{\ccat}[0]{\mathcal{C}}
@@
 \newcommand{\counit}[0]{ \varepsilon }
-
-Indexed comonads provide a semantics to sequential composition, but
-additional structure is needed for the semantics of the full coeffect calculus.
-Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
-structure, which provides a monoidal operation $\mathsf{m} : C A
-\times C B \rightarrow C (A \times B)$ for merging contexts (used in
-the semantics of abstraction).
-
-The semantics of the coeffect calculus requires an indexed
-lax semi-monoidal structure for combining contexts \emph{as well as} an indexed
-\emph{colax} monoidal structure for \emph{splitting} contexts. These are provided
-by two families of morphisms (given a coeffect algebra with $\vee$ and $\wedge$):
-%
-\begin{itemize}
-\item $\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \wedge s)}(A \times B)$ natural
-in $A, B$;
-\item $\csplit_{r, s} : \ccat^{(r \vee s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B$ natural
-in $A, B$;
-\end{itemize}
-%
-\noindent
-The $\cmerge_{r, s}$ operation merges contextual computations with
-tags combined by $\wedge$ (greatest lower-bound), elucidating 
-the behaviour of $\cmerge_{r, s}$: that merging may result in 
-the loss of some parts of the contexts $r$ and $s$.
-
-The $\csplit_{r, s}$ operation splits context-dependent computations
-and thus the contextual requirements. To obtain coeffects $r$ and
-$s$, the input needs to provide \emph{at least} $r$ and $s$, so the
-tags are combined using the $\vee$ (least upper-bound).
-
-For the sake of brevity, we elide the indexed versions of the laws required by Uustalu and Vene
-(\eg~most importantly, merging two contexts and then adding the third is
-equivalent to merging the last two and then adding the first; similar rule holds is required
-for splitting).
-
-\subsection{Semantics of flat calculus}
-
-% ==================================================================================================
 

### 2014-05-26 text/Chapter04.tex
Finish categorical semantics (Chapter 4)

@@
 \subsection{Flat coeffect types}
+\label{sec:flat-calculus-types}
+
 The type system for flat coeffect calculus is shown in Figure~\ref{fig:flat-types}. Variables 
@@
 with a join $\sqcup$ and meet $\sqcap$. The flat coeffect algebra for liveness is then formed by
-$(L, \sqcup, \sqcap, \sqcap, \ident{L}, \ident{D}, \sqsubseteq)$.
+$(L, \sqcap, \sqcup, \sqcap, \ident{L}, \ident{D}, \sqsubseteq)$.
 \end{example}
@@
 As in Section~\ref{sec:applications-flat-live}, sequential composition $\cseq$ is modelled by 
-the join operation $\sqcup$ and point-wise composition $\cpar$ is modelled by meet $\sqcap$. 
+the meet operation $\sqcap$ and point-wise composition $\cpar$ is modelled by join $\sqcup$. 
 Two-point lattice is a commutative, idempotent monoid. The distributivity 
-$(r \sqcap s) \sqcup t = (r \sqcup t) \sqcap (s \sqcup t)$ does not hold for \emph{every} 
+$(r \sqcup s) \sqcap t = (r \sqcap t) \sqcup (s \sqcap t)$ does not hold for \emph{every} 
 lattice, but it trivially holds for a two-point lattice used here.
 
-The definition uses meet $\sqcap$ for the $\czip$ operator that is used by lambda abstraction.
+The definition uses join $\sqcup$ for the $\czip$ operator that is used by lambda abstraction.
 This means that, when the body is live $\ident{L}$, both declaration-site and call-site are 
@@
 judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
-$\sem{\tau_1} \times \ldots \times \sem{\tau_n} \rightarrow \sem{\tau}$.
+$\sem{\tau_1 \times \ldots \times \tau_n} \rightarrow \sem{\tau}$.
 
@@
 \subsection{Generalising to indexed comonads}
+\label{sec:flat-semantics-idx}
+
 The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
@@
 
-\subsection{Indexed monoidal comonads}
+\subsection{Flat indexed comonads}
 \label{sec:flat-semantics-monoidal}
@@
 Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-an \emph{indexed monoidal comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
+an \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
 equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
@@
 
-\paragraph{Simpler.}
+\paragraph{Alternative definition.}
+Although we do not require this as a general law, in all our systems, it is the case that
+$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$.
+This allows a simpler definition of \emph{indexed flat comonad} by expressing the \ident{split} operation
+in terms of the lifting (sub-coeffecting) as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
+\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
+  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c), 
+    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
+\end{array}
+\end{equation*}
+%
+The $\ident{map}_{\cclrd{r}}$ operation is the mapping on functions that corresponds to the object 
+mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map} 
+for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
+are first and second projections from a two-element pair. To define the 
+$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we duplicate the argument $c$, then use 
+lifting to throw away additional parts of the context and then transform the values in the 
+context.
+
+This alternative is valid for our examples, but we do not use it for two reasons. Firstly, it 
+requires duplication of the value $c$, which is not required elsewhere in our model. So, using 
+explicit \ident{split}, our model could be embedded in a linear or affine model. Secondly, it
+is similar to the definition that is needed for structural coeffects in Chapter~\ref{ch:structural}
+and so it makes the connection between the two system easier to see.
 
 \paragraph{Examples.}
+All examples of \emph{indexed comonads} discussed in Section~\ref{sec:flat-semantics-idx} can
+be extended into \emph{flat indexed comonads}.
+
+\begin{example}[Monoidal comonads]
+Just like indexed comonads generalise co\-monads, the additional structure of
+flat indexed comonads generalises symmetric semimonoidal comonads of Uustalu 
+and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
+where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the 
+monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation 
+is defined by duplication and $\ident{lift}_{1,1}$ is the identity function.
+\end{example}
+
+\begin{example}[Indexed option]
+Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively 
+and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined 
+as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations 
+of a flat indexed comonad are defined as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
+\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
+\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
+\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
+\end{array}
+\quad
+\begin{array}{rcl}
+\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
+\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
+\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
+\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& (a, b)\\
+\end{array}
+\quad
+\begin{array}{rcl}
+\ident{lift}_{\ident{L}, \ident{D}}~v &\narrow{=}& ()\\
+\ident{lift}_{\ident{L}, \ident{L}}~v &\narrow{=}& v\\
+\ident{lift}_{\ident{D}, \ident{D}}~() &\narrow{=}& ()\\
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+Without the indexing, the \ident{merge} operations implements \emph{zip} on option values,
+returning an option only when both values are present. The behaviour of the \ident{split} 
+operation is partly determined by the indices. When the input is \emph{dead}, both values have 
+to be dead (this is also the only solution of $\ident{D}=\cclrd{r}\sqcap\cclrd{D}$), but when
+the input is \emph{live}, the operation can perform implicit sub-coeffecting and drop one of
+the values.
+
+Explicit sub-coeffecting using the (\emph{sub}) rule is modelled by the \ident{lift} operation.
+This can turn a \emph{live} value $v$ into a dead value $()$, or it can behave as identity.
+The behaviour is, again, determined by the index.
+
+\begin{example}[Indexed product]
+For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
+$\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
+is $\alpha \times (\cclrd{r} \rightarrow R)$ where $R$ is some representation of a parameter value.
+The additional operations are defined as:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{split}_{\cclrd{r}, \cclrd{s}}~((a,b), g) &\narrow{=}& ((a, \restr{g}{\cclrd{r}}), (b, \restr{g}{\cclrd{s}}))\\
+\ident{merge}_{\cclrd{r}, \cclrd{s}}~((a, f), (b, g)) &\narrow{=}& ((a, b), f \uplus g)\\
+\ident{lift}_{\cclrd{r'}, \cclrd{r}}~(a, g) &\narrow{=}& (a, \restr{g}{\cclrd{r}})\\
+\end{array}
+\quad
+\begin{array}{l}
+\textnormal{where}~f \uplus g = \\[-0.25em]
+\quad\restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g 
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{split} operation splits the tuple and restricts the function (representing available
+implicit parameters) to the required sub-sets. This corresponds to the definition in terms
+of \ident{lift}, which performs just the restriction. The \ident{merge} operation is more 
+interesting. It uses $\uplus$ operation that we defined when introducing implicit parameters
+in Section~\ref{sec:applications-flat-impl}. It merges the values, preferring the definitions from
+the right-hand side (call-site) over left-hand side (declaration-site). Thus the operation is not
+symmetric.
+
+\begin{example}[Indexed list]
+Our last example provides the semantics of data-flow computations. The flat coeffect algebra 
+is formed by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$. In a 
+non-indexed version, the semantics is provided by a non-empty list. In the indexed semantics,
+the index represents the length of the storing past values. The data type is then a pair of 
+the current value, followed by $n$ past values. The mappings that form the flat indexed comonad 
+are defined as follows:
+
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{0}\langle a_0 \rangle = a_0
+\\[0.45em]
+\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
+\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
+\\[0.45em]
+\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
+\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
+\\[0.45em]
+\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
+\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
+\\[0.45em]
+\ident{lift}_{n', n} \langle a_0, \ldots, a_n' \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
+\quad \langle a_0, \ldots, a_n \rangle
+\end{array}
+\begin{array}{l}
+\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[11em]
+~\\
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The reader is invited to check that the number of required past elements in each of the mappings
+matches the number specified by the indices. The index specifies the number of \emph{past} elements
+and so the list always contains at least one value. Thus \ident{counit} returns the element of a
+singleton list.
+
+The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results 
+of the $f$ function, which itself requires $m$ past values. When combining two lists, 
+$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter 
+argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths. 
+Finally, the lifting operation just drops some number of elements from a list.
+
+% --------------------------------------------------------------------------------------------------
 
 \subsection{Semantics of flat calculus}
+
+In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
+computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed 
+comonad} structure, we can now define a single uniform semantics that is capable capturing of all 
+our examples, as well as other computations that can be modelled by the structure.
+
+\paragraph{Contexts and functions.}
+The modelling of contexts and functions generalizes the earlier concrete examples. We use the 
+family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed) data-type that wraps the product of 
+free variables of the context and the arguments of functions:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau} 
+  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+
+% --------------------------------------------------------------------------------------------------
+
+\newcommand{\ctx}{\textit{ctx}}
+\begin{figure*}[t]
+
+\begin{equation*}
+\begin{array}{ll}
+\sem{\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }~\ctx =
+  \pi_i~(\ident{counit}_{\cunit}~\ctx) & (\emph{var})
+\\[0.5em]
+\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
+  \delta~(c_i) & (\emph{const})
+\\[0.5em]
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }~\ctx = & (\emph{sub})\\[-0.25em]
+  \hspace{3em}\sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx)~
+    \hspace{7.5em}(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
+\\[0.5em]
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
+  \hspace{3em}\sem{\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v))
+\\[0.5em]
+\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
+  \hspace{2.5em}
+  \begin{array}{l}  
+  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}}~
+    (\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~(\lambda x.(x,x))~\ctx) \\[-0.25em]
+  \kvd{in}~\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
+      (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~\sem{\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~\ctx_2)
+  \end{array}    
+\end{array}
+\end{equation*}
+
+\caption{Categorical semantics of the flat coeffect calculus}
+\label{fig:flat-semantics}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Expressions.}
+The definition of the semantics is shown in Figure~\ref{fig:flat-semantics}. For readability, we 
+write the definitions in a simple programming language notation as opposed to the point-free 
+categorical style. However, it can be equally written using just the operations of flat indexed 
+comonad together with $i^{th}$ projection from a tuple represented by $\pi_i$, \emph{curry} and 
+\emph{uncurry}, function composition, value duplication ($\Delta : A \rightarrow A \times A$) and 
+function pairing (given $f:A\rightarrow B$ and $g:C\rightarrow D$ then $f\times g : A\times C \rightarrow B \times D$). 
+These operations can be provided by \eg~a Cartesian Closed Category.
+
+The semantics of variable access and abstraction are the same as in the semantics of Uustalu and 
+Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses 
+$\ident{counit}_{\cunit}$ to extract product of free-variables from the context and then projection
+$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) takes the context $\ctx$ and function argument 
+$v$ and merges their additional contexts using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. Assuming
+the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives us a 
+value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
+Assuming that $n$-element tuples are associated to the left, the wrapped context is equivalent to
+$\sigma_1 \times \ldots \times \sigma_n \times \tau_1$, which can then be passed to the body of the
+function.
+
+The semantics of application is more complex. It first duplicates the free-variable product inside the 
+context (using $\ident{map}_{\cclrd{r}}$ and duplication). Then it splits this context using 
+$\ident{split}_{\cclrd{r}, \cclrd{s} \cpar \cclrd{t}}$. The two contexts contain the same variables
+(as required by sub-expressions $e_1$ and $e_2$), but different coeffect annotations. The first
+context (with index $\cclrd{r}$) is used to evaluate $e_1$, resulting in a function 
+$\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. To obtain the result, we compose this with a function
+created by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$ on the semantics of sub-expression
+$e_2$, which is of type $\ctyp{\cclrd{s}\cseq\cclrd{t}}{\sigma_1 \times \ldots \times \sigma_n}
+\rightarrow \ctyp{\cclrd{t}}{\tau_1}$.
+
+Finally, constants (\emph{const}) are modelled by a global dictionary $\delta$ and sub-coeffecting
+is interpreted by dropping additional context from the provided context $\ctx$ using 
+$\ident{lift}_{\cclrd{r}, \cclrd{r'}}$ and providing it to the semantics of the assumption.
+
+\paragraph{Properties.}
+
+The categorical semantics can be used to embed context-dependent computations in functional 
+programming languages, similarly to how monads provide a way of embedding effectful computations.
+More importantly, it also provides validation for the design of the type system developed in 
+Section~\ref{sec:flat-calculus-types}. As stated in the following theorem, the annotations in 
+the type system match those of the semantic functions.
+
+\vspace{2em}
+
+\begin{remark}[Correspondence]
+In all of the typing rules of the flat coeffect system, the context annotations $\cclrd{r}$ of typing 
+judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ 
+correspond to the indices of mappings $\ctyp{\cclrd{r}}{}$ in the corresponding semantic function defined 
+by $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}$.
+\end{remark}
+\begin{proof}
+By analysis of the semantic rules in Figure~\ref{fig:flat-semantics}.
+\end{proof}
+
+\noindent
+Thanks to the indexing, the statement of the remark is significantly stronger than for a 
+non-indexed system, because it provides the justification for our choice of indices in the typing
+rules. In particular, we can see that the annotations follow from the annotations on primitive 
+functions that define the semantics. Also, each function defining the semantics uses a distinct 
+operation of the coeffect algebra and so the type system is the most general possible definition
+(within the comonadic framework we use).
 
@@
 \newcommand{\counit}[0]{ \varepsilon }
+\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
+\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}
 
@@
 % ==================================================================================================
-
-\paragraph{Example}
-For the indexed partiality comonad, given the liveness coeffect
-algebra $(\{\ident{D}, \ident{L}\}, \sqcap, \sqcup, \sqcap, \ident{L})$,
-the additional lax/colax monoidal operations are:
-%%
-\begin{align*}
-\begin{array}{lll}
-\cmerge_{\ident{L}, \ident{L}} (x, y) = (x, y) 
-  \quad\quad& \csplit_{\ident{D}, \ident{D}} \;\; () \;\;\; = ((), ())
-  \quad\quad& \csplit_{\ident{D}, \ident{L}} (x, y) = ((), y) \\
-\cmerge_{\ident{r}, \ident{s}} \;\,(x, y) = () 
-  & \csplit_{\ident{L}, \ident{D}} (x, y) = (x, ())
-  & \csplit_{\ident{L}, \ident{L}} (x, y) = (x, y)
-\end{array}
-\end{align*}
-
-\paragraph{Example}
-Uustalu and Vene model causal dataflow computations using the
-non-empty list comonad $\ident{NEList}~A = A \times (1 +
-\ident{NEList}~A)$~\cite{comonads-notions}.  Whilst this comonad
-implies a trivial indexed comonad, we define an indexed comonad with
-integer indices for the number of past values demanded of the context.
-
-We define $\ctyp{n}{A} = A \times (A \times \ldots \times A)$ where the first $A$ is the current
-(always available) value, followed by a finite product of $n$ past values. The definition of the
-operations is a straightforward extension of the work of Uustalu and Vene.
-
-%\begin{example}
-%The semantics of the coeffect calculus for
-% dynamically-scoped implicit parameters, with coeffect
-%algebra $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset)$, 
-%is provided by the indexed comonad $\ccat^rA = A \times (r \rightarrow \tau)$
-%where $r \in \mathcal{P}(\ident{Id})$ with operations:\dnote{The type here is wrong,
-%do we need to store the types too? Maybe should just be a product.}
-%%
-%\vspace{-1.0em}
-%\begin{equation*}
-%\begin{array}{l}
-%\cobind{f}{r, s} \; : \; \ccat^{r \cup s} A \rightarrow \ccat^s B\\
-%\cobind{f}{r, s} (a, g) = (f(a, g|_r), g|_s)
-%\\[1em]
-%\counit{} (a, g) = a\\
-%%\iota_{r, s} \; : \: \ccat^r A \rightarrow \ccat^s A\\
-%\iota_{r, s}(a, g) = (a, g|_s)\quad(\textnormal{if}~\leq r)
-%\end{array}
-%\quad\begin{array}{l}
-%\cmerge_{r, s} : \ccat^r A \times \ccat^s B \rightarrow \ccat^{(r \cup s)}(A \times B)\\
-%\cmerge_{r, s} ((a_1, g_1), (a_2, g_2)) = ((a_1, a_2), g_1|_{r \setminus s} \cup g_2 )
-%\\[1em]
-%\csplit_{r, s} : \ccat^{(r \cup s)}(A \times B) \rightarrow \ccat^r A \times \ccat^s B \\
-%\csplit_{r, s} ((a_1, a_2), g) = ((a_1, g|_r), (a_1, g|_s))
-%\end{array}
-%\end{equation*}
-%\end{example}
-
-%The function created using \emph{coextend} expects a function $g$ that is defined for
-%resource names $r \cup s$. We use function restriction $g|_r$ and $g|_s$ to get functions that are 
-%defined on resources required by $f$ and the result of coextend, respectively.
-
-%The operation $\cmerge_{r, s}$ pairs the values and combines the resources available in both contexts. 
-%It combines a function $g_1|_{r\setminus s}$ (defined for resources in $r$ excluding those in $s$) and 
-%function $g_2$ (defined for resources $s$). This definition is not symmetric as it prefers resources 
-%defined in the second context. As a result, the caller of a function can rebind resources defined
-%at the declaration site.
-
-%Resources show that indexed comonads describe the structure of context more precisely
-%than comonads. A comonad would have to use $TA = A \times (\Omega \rightharpoonup R)$ 
-%with a \emph{partial} function on all resource names. Using indexed comonads, we can ensure that the function
-%is defined on all resources tracked statically in the tag.\dnote{Needs tweaking}
-
-\vspace{-1.5em}
-\subsection{Categorical Semantics. }
-\label{sec:comonads-semantics}
-
-\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
-\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}
-
-\begin{figure*}[t]
-\vspace{-1em}
-\newcommand{\uu}[1]{\overline{#1}}
-% \begin{equation*} 
-% \sem{\ctyp{r}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-%   \ctyp{r}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
-% \end{equation*}
-% \vspace{-0.4em}
-\begin{equation*}
-\begin{array}{rcll}
-\sem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \emph{curry} \; (\sem{\ctyp{r \wedge s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    %&$(1)$
-\\[0.1em]
-\sem{\ctyp{r \vee (s \cseq t)}{\Gamma} \vdash e_1~e_2 : \tau} &=& 
-    (\emph{uncurry} \;  \sem{\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{s}{\tau_1} \rightarrow \tau_2}) \, \circ \\
- & & \quad\; (id \times \cobind{\sem{\ctyp{t}{\Gamma}\vdash e_2 : \tau_1}}{t,s})
-    \circ \csplit_{r, s \cseq t} \circ \ccat^{r \vee (s \cseq t)}\Delta
-\\[0.1em]
-\sem{\ctyp{\cunit}{\Gamma} \vdash x_i : \tau_i} &=&
-    \pi_i \circ \counit                                                                  % &$(3)$
-%\\[0.1em]
-%\sem{\ctyp{r}{\Gamma} \vdash e : \tau} &=&
-%    \sem{\ctyp{s}{\Gamma} \vdash e : \tau} \circ \iota_{r, s}                              %   &%$(4)$
-%\\[0.1em]
-%\sem{\ctyp{r\vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2} &=&
-%    \sem{\ctyp{r}{(\Gamma, x:\tau_1)} \vdash e_2 : \tau_2} \circ
-% ((\varepsilon \times \sem{ \ctyp{s}{\Gamma} \vdash e_1 : \tau_1 }) \circ 
-%\iftorhat{r \vee (r \cseq t)}\Delta
-\end{array}
-\end{equation*}
-\vspace{-1.5em}
-\caption{Categorical semantics for the coeffect calculus}
-\label{fig:semantics}
-\vspace{-1.0em}
-\end{figure*}
-
-Figure~\ref{fig:semantics} shows the
-categorical semantics of the coeffect calculus
-using additional operations 
-$\pi_i$ for projection of the \emph{i}$^{th}$ element of a 
-product, usual
-\emph{curry} and \emph{uncurry} operations, and
- $\Delta : A \rightarrow A \times A$ duplicating a
-value. While $\ccat^{r}$ is a family of object mappings,
-it is promoted to a family of functors with the derived morphism mapping 
-$\ccat^{r}(f) = \cobind{ (f \circ \counit) }{\cunit, r}$.
-
-The semantics of variable access and abstraction are the same as in
-Uustalu and Vene's semantics, modulo coeffects. Abstraction uses $\cmerge_{r, s}$ to merge the outer
-context with the argument context for the context of the function body.
-The indices of $\cunit$ for
-$\counit$ and $r,s$ for $\cmerge_{r, s}$ match the coeffects
-of the terms.
-The semantics of application is more complex. It first duplicates the
-free-variable values inside the context and then splits this context
-using $\csplit_{r, {s \oplus t}}$. The two contexts (with
-different coeffects) are passed to the two sub-expressions,
-where the argument subexpression, passed a context $(s \oplus t)$,
-is coextended to produce a context $s$ which is passed into
-the parameter of the function subexpression (\emph{cf.} given $f : A \rightarrow (B \rightarrow C)$,
- $g : A \rightarrow B$, then $\emph{uncurry} \, f \, \circ \, (id \times g) \circ \Delta : A \rightarrow C$).
-
-A semantics for sub-coeffecting is omitted, but may be provided
-by an operation 
-$\iota_{r, s} : \ccat^r A \rightarrow \ccat^s A$ natural in $A$, for all $r, s \in S$ where $s \leq r$,
-which transforms a value $C^r A$ to $C^s A$ by
-ignoring some of the encoded context.
-
-% ==========================================================================================
 

### 2014-06-02 text/AppendixA.tex
Syntactic properties (Chapter 4)

\chapter{Appendix A} 
\label{ch:appendix}

%---------------------------------------------------------------------------------------------------

\section{Internalized substitution}

\subsection{First transformation}

\begin{equation*}
(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 \leadsto \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
\end{equation*}

\begin{equation*}
\tyrule{app}
  { \tyrule{glet}
      { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
        \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 }
      { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 } &
    \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3  }
  { \coctx{\Gamma}{(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) } \vdash (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 : \tau_2 }
\end{equation*}

to

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
    \tyrule{app}
      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 & 
        \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3 }
      { \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})} \vdash e_2~e_3 : \tau_2 } }
  { \coctx{\Gamma}{ (\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})) \;\cpar\; (\cclrd{s}\,\cseq\,(\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})))  }
      \vdash \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3) : \tau_2 }
\end{equation*}

meaning
\begin{equation*}
\begin{array}{l}
(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) =
\end{array}
\end{equation*}

\subsection{Second transformation}

Second transformation

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
  { \coctx{\Gamma}{
      \cclrd{t}
      \;\cpar\;
      (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
         \,\cseq\, 
         \cclrd{t} )
    } 
    \vdash \kvd{glet}~x_r=(\kvd{glet}~x_s=e_s~\kvd{in}~e_r)~\kvd{in}~e_t : \tau_t }
\end{equation*}

or

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
  { \coctx{\Gamma}{
      (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
      \;\cpar\;
      (  \cclrd{s}
         \,\cseq\, 
         (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) )
    } 
    \vdash \kvd{glet}~x_s=e_s~\kvd{in}~(\kvd{glet}~x_r=e_r~\kvd{in}~e_t) : \tau_t }
\end{equation*}

\begin{equation*}
\begin{array}{l}
  \cclrd{t}
  \;\cpar\;
  (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
     \,\cseq\, 
     \cclrd{t} ) = 
\\
  \cclrd{t}
  \;\cpar\;
  (\cclrd{r}\,\cseq\, \cclrd{t})
  \;\cpar\;
  (\cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t}) = 
\\
  \cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t} 
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
  (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
  \;\cpar\;
  (  \cclrd{s}
     \,\cseq\, 
     (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) ) = 
\\     
  \cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})
  \;\cpar\;
  (\cclrd{s} \,\cseq\, \cclrd{t})
  \;\cpar\;
  (\cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t})  = 
\\     
  \cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t}
\end{array}
\end{equation*}

require

\begin{equation*}
\cclrd{r}\;\cpar\;(\cclrd{r} \,\cseq\, \cclrd{s}) = \cclrd{r} \,\cseq\, \cclrd{s}
\end{equation*}

### 2014-06-02 text/Chapter04.tex
Syntactic properties (Chapter 4)

@@
 
+\section{Equational theory}
+\label{sec:flat-syntax}
+
+Each of the concrete coeffect calculi discussed in this chapter has a different notion of context,
+much like various effectful languages have different notions of effects (such as exceptions or
+mutable state). However, in all of the calculi, the context has a number of common properties that 
+are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
+that hold for all of the systems we consider. Further properties hold for systems where the context
+satisfies additional properties.
+
+In this section, we look at such shared syntactic properties. This accompanies the previous section, 
+which provided a \emph{semantic} justification for the axioms of coeffect algebra with a 
+\emph{syntactic} justification. Operationally, this section can also be viewed as providing a 
+pathway to an operational semantics for two of our systems (implicit parameters and liveness),
+which can be based on syntactic substitution. As we discuss later, the notion of context for
+data-flow is more complex.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Syntactic properties}
+
+Before discussing the syntactic properties of general coeffect calculus formally, it should be 
+clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
+that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
+Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
+equations define four syntactic reductions on the terms:
+%
+\begin{equation*}
+\begin{array}{rllcl}
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
+(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized substitution})\\
+e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
+\end{array}
+\end{equation*}
+%
+The rules capture syntactic reductions that can be performed in a general calculus, without any 
+knowledge of the specific notion of context. The \kvd{glet} notation models explicit sequencing
+of context-dependent computations an is inspired by Filinski \cite{monads-inaction}. In the rest
+of the section, we briefly outline the interpretation of the four rules and then we focus on
+call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name (Section~\ref{sec:flat-syntax-cbn})
+in more details.
+
+The focus of this work is on the general coeffect system and so we do not discuss the operational
+semantics of the specific notions of context. However, some work in that area has been done
+by Brunel et al. \cite{coeffects-quantitative}.
+
+\paragraph{Call-by-name.} 
+In call-by-name, the argument is syntactically substituted for all occurrences of a variable. It
+can be used as the basis for operational semantics of purely functional languages. However, using
+the rule in effectful languages breaks the \emph{type preservation} property. For example, consider
+a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
+A function $\lambda x.y$ is a effect-free:
+%
+\begin{equation*}
+y\!:\!\tau_1 \vdash \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 \;\&\; \emptyset
+\end{equation*}
+%
+Substituting an expression $e$ with effects $\{ \ident{write} \}$ for $y$ changes the type of 
+the function by adding latent effects (without changing the immediate effects):
+%
+\begin{equation*}
+\vdash \lambda x.e : \tau_1 \xrightarrow{ \{ \ident{write} \} } \tau_2 \;\&\; \emptyset
+\end{equation*}
+%
+Similarly to effect systems, substituting a context-dependent computation $e$ for a variable $y$ can
+add latent coeffects to the function type. However, this is not the case for \emph{all} flat coeffect
+calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
+parameters system. This makes the model suitable for languages such as Haskell.
+
+\paragraph{Call-by-value.}
+
+The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
+first reduced to a \emph{value} before performing the substitution. In effectful languages, 
+value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
+values are identifiers $x$ or functions $(\lambda x.e)$.
+
+The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
+A function $(\lambda x.e)$ does not delay all context requirements of the body $e$ and may have
+immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
+sense \emph{and} has not immediate context requirements. We define this formally in 
+Section~\ref{sec:flat-syntax-cbv}.
+
+The call-by-value evaluation strategy holds for a wide range of flat coeffect calculi, including all
+our three examples. However, it is rather weak -- in order to use it, the concrete semantics needs
+to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ to
+a term $\coctx{\Gamma}{\cunit} e' : \tau$ with no context requirements.
+
+\paragraph{Internalized substitution.}
+
+The (\emph{internalized substitution}) rule models an operational semantics where the language
+provides a construct that models \emph{sequential composition} and expressions can be reduced to 
+a normal form, consisting of a sequenced context-dependent operations. We choose to write the 
+sequencing operation as \kvd{glet} to emphasize that this is a separate primitive and not an 
+ordinary syntactic \kvd{let}. The normal form looks as follows:
+%
+\begin{equation*}
+\kvd{glet}~x_1 = e_1~\kvd{in}~\ldots~\kvd{glet}~x_n = e_n~\kvd{in}~e
+\end{equation*}
+%
+Here, the expressions $e_1, \ldots, e_n, e$ do not contain further nested \kvd{glet} constructs.
+This evaluation strategy provides context-dependent counterpart to operational view of monads
+developed by Filinsky \cite{monads-inaction}. We discuss how expressions reduce to the normal form
+in Section~\ref{sec:flat-syntax-norm}
+
+The (\emph{internalized substitution}) rule is useful when defining a concrete semantics for 
+a language that provides constructs for explicitly providing the context. For example, consider
+a language with implicit parameters where a parameter is defined by $e_1~\kvd{with}~\ident{?p}=e_2$.
+Semantics for such language would provide a reduction rule:
+%
+\begin{equation*}
+(\kvd{glet}~x_1 = \ident{?p}~\kvd{in}~e)~\kvd{with}~\ident{?p}=e_1 ~~\leadsto~~ \kvd{let}~x_1 = e_1~\kvd{in}~e
+\end{equation*}
+%
+Here, the \kvd{glet} construct provides a way of sequentialising the context-requi\-rements so
+that they can be discharged by matching constructs providing the required contexts. As discussed
+earlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
+this form of evaluation (rather than looking at semantics for concrete systems).
+
+\paragraph{Local soundness and completeness.}
+Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
+are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
+introduce a function arrow (lambda abstraction) and eliminate it (application) are not too strong
+and sufficiently strong.
+
+The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
+already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
+coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Call-by-value evaluation}
+\label{sec:flat-syntax-cbv}
+
+\begin{definition} 
+An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
+where $\aclrd{\mathsf{VAL}} : \C^{mksh{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
+always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
+\end{definition}
+
+Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
+as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
+merging of context-requirements, respectively.
+
+\begin{lemma}[Call-by-value substitution]
+\label{thm:flat-subst-cbv}
+In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+and the $\cclrd{\wedge}$ operator:
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\end{lemma}
+\begin{proof}
+By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
+with $\cunit$.
+\end{proof}
+
+\noindent
+Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
+the operational semantics must provide a way of partially evaluating a term 
+with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
+Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:
+
+\begin{theorem}[Call-by-value reduction]
+In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
+$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
+\end{proof}
+
+\subsection{Call-by-name evaluation}
+\label{sec:flat-syntax-cbn}
+
+The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
+may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
+$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.
+
+% TODO
+%
+% The context requirements of the 
+% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
+% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
+% lemma should be $r \fmax (s \fseq r)$.
+
+We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
+$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
+Liveness analysis is an example of top-pointed coeffects
+as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.
+
+\begin{lemma}[Top-pointed substitution]
+\label{thm:cbn-substitution-top}
+In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+and the $\cclrd{\wedge}$ operator:
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\vspace{-1.5em}
+\end{lemma}
+\begin{proof}
+Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
+\end{proof}
+%
+\noindent
+As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
+use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).
+
+In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
+requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
+the context requirements arising from the substitution can be associated with the context
+$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
+requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
+(all three operators are set union) and allows the following substitution lemma:
+
+\begin{lemma}[Bottom-pointed substitution]
+\label{thm:cbn-substitution-bot}
+In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
+and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\vspace{-1.5em}
+\end{lemma}
+\begin{proof}
+By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
+the free-variable context.
+\end{proof}
+\noindent
+The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
+variant described here is flexible enough to let us always re-associate new context requirements with the 
+free-variable context.
+
+The two substitution lemmas show that the call-by-name evaluation strategy can be used for
+certain coeffect calculi, including liveness and implicit parameters.
+Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
+reduction, the following theorem holds:
+
+\begin{theorem}[Call-by-name reduction]
+\raggedright
+In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
+Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
+\end{proof}
+
+ALSO: Something on eta expansion
+
+\subsection{Normal form}
+\label{sec:flat-syntax-norm}
+
+\begin{equation*}
+\tyrule{glet}
+  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_2 }
+  { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+
+First transformation
+\begin{equation*}
+(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
+\end{equation*}
+
+Second transformation
+
+\begin{equation*}
+\kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~
+\kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
+\end{equation*}
+
+or
+
+%
+% TODO: We could have a section about let binding covering the alternative CBV and the let binding
+% we included in the system, but we'd have to clean this up!
+%
+%\subsection{Syntactic let binding}
+%\begin{remark}
+%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
+%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
+%calculus of Section~\ref{}.
+%%%
+%\begin{align*}
+%{\footnotesize{\trule{let}}}
+%\dfrac
+%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
+%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
+%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
+%\end{align*}
+%%%
+%Therefore, for example, we see that \emph{let} has the coeffects
+%as substitution, for example:
+%%%
+%\begin{align*}
+%& {\footnotesize{\trule{let}}}
+%\dfrac
+%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
+%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
+%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
+%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
+%\end{align*}
+%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
+%the judgment:
+%%%
+%$$
+%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
+%$$
+%%%
+%\end{remark}
+%
+% 
+%\paragraph{Syntactical let binding}
+%Has the same typing derivation as abstraction/application
+%
+%\[
+%\inference
+% {\inference
+%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
+%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
+%  &
+%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
+%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
+%\]
+%
+%We can extend the CBV behaviour to \emph{let}, where the bound term
+%is evaluated to a pure value first, discharging any contextual
+%requirements before substitution into the body.
+%
+%\begin{lemma}[CBV \emph{let}]
+%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
+%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
+%: \tau$.
+%\end{lemma}
+
+% ==================================================================================================
+
+\section{Syntactic extensions}
+\label{sec:flat-exts}
+
+\subsection{Lambda abstraction}
+\subsection{Constants and pairs}
+\subsection{Recursion}
+
+% ==================================================================================================
+
+\section{Type inference}
+\label{sec:flat-inference}
+
+\subsection{Semi-lattice formulation}
+\subsection{Type inference algorithm}
+
+% ==================================================================================================
+
+\section{Related work}
+
+\subsection{What can be monad}
+\label{sec:flat-related-monads}
+
+% ==================================================================================================
+
+\section{Summary}
+
+\section{----------------JUNK!}
+
+% ==================================================================================================
+
+~
+\newpage
+~
+
 \newcommand{\ccat}[0]{\mathcal{C}}
@@
 
-\section{Equational theory}
-\label{sec:flat-syntax}
-
-\subsection{Call-by-value evaluation}
-\subsection{Call-by-name evaluation}
-
-% ==================================================================================================
-
-\section{Syntactic extensions}
-\label{sec:flat-exts}
-
-\subsection{Lambda abstraction}
-\subsection{Constants and pairs}
-\subsection{Recursion}
-
-% ==================================================================================================
-
-\section{Type inference}
-\label{sec:flat-inference}
-
-\subsection{Semi-lattice formulation}
-\subsection{Type inference algorithm}
-
-% ==================================================================================================
-
-\section{Related work}
-
-\subsection{What can be monad}
-\label{sec:flat-related-monads}
-
-% ==================================================================================================
-
-\section{Summary}
-
-\section{----------------JUNK!}
-
-% ==================================================================================================
-
-~
-\newpage
-~
-
 % ==================================================================================================
@@
 \label{sec:syntactic}
-
-Operational semantics of every context-dependent language differs as the notion of context 
-is always different. However, for coeffect calculi satisfying certain conditions we can 
-define a universal equational theory. This suggests a pathway to an operational semantics for 
-two out of our three examples (the notion of context for data-flow is more complex).
-
-In a pure $\lambda$-calculus, $\beta$ and $\eta$ equality for
-functions (also called \emph{local soundness} and \emph{completeness}
-respectively~\cite{logic-modal-reconstruction})
-describe how pairs of abstraction and application can be eliminated:
-$(\lambda x . e_2) e_1 \equiv_\beta \subst{e_1}{x}{e_2}$ and $(\lambda x . e \, x) 
-\equiv_\eta e$. The $\beta$ equality rule, using the usual Barendregt convention of 
-syntactic substitution, implies a \emph{reduction},
-giving part of an operational semantics for the calculus.
-
-The call-by-name evaluation strategy modelled by $\beta$-reduction is
-not suitable for impure calculi therefore a restricted $\beta$ rule,
-corresponding to call-by-value, is used, \ie~$(\lambda x . e_2) v
-\equiv \subst{e_2}{x}{v}$. Such reduction can be encoded by a
-\emph{let}-binding term, $\kvd{let}~x=e_1~\kvd{in}~e_2$, which
-corresponds to sequential composition of two computations, where the
-resulting pure value of $e_1$ is substituted into
-$e_2$~\cite{monads-inaction,monad-notions}.
-
-%
-% We consider here both a notion of \emph{let}-binding for the coeffect
-% calculus, useful for a CBV evaluation, and a notion of substitution 
-% for a CBV evaluation.
-%
 

### 2014-06-03 text/Chapter04.tex
Call-by-value and more intros (Chapter 4)

@@
 
+As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
+calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
+term to a \emph{value} has to be provided for each system. Syntactic category for values is 
+defined as:
+%
+\begin{equation*}
+\begin{array}{llll}
+v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
+n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
+e\in\mathit{Expr}     & e &::=& v \sep n 
+\end{array}
+\end{equation*}
+%
+The category \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
+coeffect calculus cannot be defined purely syntactically.
+
 \begin{definition} 
-An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
-where $\aclrd{\mathsf{VAL}} : \C^{mksh{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
-always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
+An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if it is a syntactic value,
+\ie~$e \in \mathit{SynVal}$ and it has no context-dependencies, \ie~$\coctx{\Gamma}{ \cunit } \vdash e : \tau$.
 \end{definition}
 
-Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
-as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
-merging of context-requirements, respectively.
+\noindent
+The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
+variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
+context-requirements of the term:
 
@@
 \label{thm:flat-subst-cbv}
-In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-and the $\cclrd{\wedge}$ operator:
-\[
-\begin{array}{l}
-\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
+In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
+$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
+change the type and context requirements: $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
 \end{lemma}
 \begin{proof}
-By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
-with $\cunit$.
+By induction over the type derivation, using the fact that $x$ and $v$ are annotated
+with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
 \end{proof}
@@
 \noindent
-Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
-the operational semantics must provide a way of partially evaluating a term 
-with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
-Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:
+The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
+proving that call-by-value reduction preserves typing requires an additional constraint on the
+flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations:
+%
+\begin{equation*}
+\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
+\hspace{10em}\textnormal{(approximation)}
+\end{equation*}
+%
+Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
+under-approximates the actual context capabilities while the $\cpar$ operation (combining of
+context requirements) over-approximates the actual context requirements.
+
+The property holds for all three systems we consider -- for implicit parameters, this is 
+an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
+is smaller than the least upper bound.
+
+Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
+$(\lambda x.e)\,v$ to a term $\substack{e}{x}{v}$, the type preservation theorem is
+stated as follows:
 
 \begin{theorem}[Call-by-value reduction]
-In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+In a flat coeffect system with the (\emph{approxi\-mation}) property, 
+if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
 and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
@@
 \begin{proof}
-A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
-$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
+Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
+\begin{equation*}
+\inference
+  { \inference
+  { \inference
+      {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
+      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
+   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
+\end{equation*}
+The last step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
+From Lemma~\ref{thm:flat-subst-cbv} $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
+As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
+%
+\begin{equation*}
+\tyrule{sub}
+ { \coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
+ { \coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
+\end{equation*}
+%
+Thus, the reduction preserves type and coeffect annotation (although this may not be the
+\emph{only} typing of the original term).
 \end{proof}
@@
 
+This does not work for dataflow, because
+
+(e1 + e2) max (e3 + f)
+e1 + (e2 max (e3 + f))
+
+But it works for the other ones...
+
 Second transformation
-
 \begin{equation*}

### 2014-07-09 text/Chapter04.tex
Finish cbn section, notes on normal forms (Chapter 4)

@@
+%!TEX root = ../main.tex
+
 %\begin{flushright}{\slshape    
@@
 \subsection{Syntactic properties}
+\label{sec:flat-syntax-props}
 
@@
 
-The (\emph{internalized substitution}) rule models an operational semantics where the language
-provides a construct that models \emph{sequential composition} and expressions can be reduced to 
+The (\emph{internalized substitution}) rule captures an operational semantics where the language
+provides a construct representing \emph{sequential composition} and expressions can be reduced to 
 a normal form, consisting of a sequenced context-dependent operations. We choose to write the 
@@
 that they can be discharged by matching constructs providing the required contexts. As discussed
-earlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
+tfearlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
 this form of evaluation (rather than looking at semantics for concrete systems).
@@
 \begin{theorem}[Call-by-value reduction]
+\label{thm:cbv-reduction}
 In a flat coeffect system with the (\emph{approxi\-mation}) property, 
@@
 \tyrule{sub}
- { \coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
- { \coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
+ { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
+ { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
 \end{equation*}
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Call-by-name evaluation}
@@
 
-The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
-may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
-$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.
-
-% TODO
-%
-% The context requirements of the 
-% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
-% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
-% lemma should be $r \fmax (s \fseq r)$.
-
+In the call-by-name reduction of $(\lambda x.e_1)~e_2$, the expression $e_2$ is substituted for all 
+occurrences of the variable $v$ in an expression $e_1$. As discussed in 
+Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in general} preserve 
+the type of a program, but it does preserve the typing in some interesting cases. The typing is
+preserved for different reasons in two of our systems, so we briefly review the concrete examples.
+
+\paragraph{Data-flow.} 
+The type preservation property does not hold for data-flow. This case is similar to the example
+shown earlier with effectful computations. As a minimal example, consider the substitution of 
+$\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1,z\!:\!\tau_1}{0} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{0} \tau_2 & & (\textnormal{before}) \\
+ \coctx{z\!:\!\tau_1}{1} &\narrow{\vdash}& \lambda x.\kvd{prev}~z : \tau_1 \xrightarrow{1} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+After the substitution, the coeffect of the body is $1$. The rule for lambda abstraction requires
+that $1=\mathit{min}(r,s)$ and so the least solution is to set both $r,s$ to $1$. The substitution
+this affects the coeffects attached both to the function type and the overall context.
+
+Semantically, the coeffect over-approximates the actual requirements -- the code does not access
+previous value of the argument $x$. This cannot be caputred by a flat coeffect system, but can
+be captured using the structural system discussed in Chapter~\ref{ch:structural}.
+
+\paragraph{Implicit parameters.} In data-flow, there is no typing for the resulting expression that
+preserves the type of the function. However, this is not the case for all systems. Consider substituting
+an implicit parameter access $\ident{?p}$ for a variable $y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1}{\emptyset} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{before}) \\
+ \coctx{\emptyset}{ \{\ident{?p}\} } &\narrow{\vdash}& \lambda x.\ident{?p} : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+The above shows one possible typing of the body -- one that does not change the coeffects of the 
+function type and attaches all additional coeffects (implicit parameters) to the context. In case
+of implicit parameters (and, more generally, systems with set-like annotations) this is always 
+possible.
+
+\paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
+substituting any expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1}{\ident{L}} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{before}) \\
+ \coctx{\emptyset}{\ident{L}} &\narrow{\vdash}& \lambda x.e : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+In the original expression, both the overall context and the function type are annotated with \ident{L}, 
+because the body contains a variable access. An expression $e$ can always be treated as being annotated
+with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so substitution does not
+change any coeffects.
+
+\paragraph{Reduction theorem.}
+The above examples (implicit parameters and liveness) demonstrate two particular kinds of coeffect 
+algebra for which typing preservation holds. Proving the type preservation separately provides 
+more insight into how the systems work and so we choose to consider separately.
+
+\begin{definition}
 We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
-$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
-Liveness analysis is an example of top-pointed coeffects
-as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.
+($\forall r \in \C \;.\; r \,\cleq\, \cunit $) and \emph{bottom-pointed} if it is the smallest (bottom) 
+element ($\forall r \in \C \;.\; r \,\cgeq\, \cunit $).
+\end{definition}
+
+\noindent
+Liveness is an example of top-pointed coeffects as variables are annotated with 
+$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and data-flow are examples
+of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
+additional requirements:
 
@@
 \label{thm:cbn-substitution-top}
-In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-and the $\cclrd{\wedge}$ operator:
-\[
+In a top-pointed flat coeffect calculus with an aglebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
+not change the coeffects of $e_r$:
+%
+\begin{equation*}
 \begin{array}{l}
-\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\vspace{-1.5em}
+ \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
+   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
+ \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
+\end{array}
+\end{equation*}
 \end{lemma}
+
 \begin{proof}
-Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
+Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
 \end{proof}
@@
 \noindent
-As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
-use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).
+As variables are annotated with the top element $\cunit$, we can substitute the term $e_s$ 
+for any variable and use sub-coeffecting to get the original typing (because
+$\cclrd{s} \,\cleq\, \cunit$).
 
 In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
-requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
-the context requirements arising from the substitution can be associated with the context
-$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
-requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
-(all three operators are set union) and allows the following substitution lemma:
+requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
+then the context requirements arising from the substitution can be associated with the context
+$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
+does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
+our implicit parameters example (all three operators are set union) and for other set-like 
+coeffects. It allows the following substitution lemma:
 
@@
 \label{thm:cbn-substitution-bot}
-In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
-\[
+In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
+where $\czip = \cseq = \cpar$ and the operation is also idempotent and commutative:
+%
+\begin{equation*}
 \begin{array}{l}
-\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\vspace{-1.5em}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
+\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
+\end{array}
+\end{equation*}
+
 \end{lemma}
@@
 By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
-the free-variable context.
+the free-variable context. See Appendix~?.
 \end{proof}
-\noindent
-The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
-variant described here is flexible enough to let us always re-associate new context requirements with the 
-free-variable context.
+
+\noindent
+The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
+requirements (arising from the substitution) with the free-variable context. In contrast, the 
+structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
+coeffects associated with individual variables -- thus preserving typing in a complementary way.
 
 The two substitution lemmas show that the call-by-name evaluation strategy can be used for
-certain coeffect calculi, including liveness and implicit parameters.
-Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
-reduction, the following theorem holds:
+certain coeffect calculi, including liveness and implicit parameters. Assuming
+$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following subject reduction
+theorem holds:
 
 \begin{theorem}[Call-by-name reduction]
-\raggedright
 In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
@@
 \begin{proof}
-A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
+
+For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
+to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \cleq \cunit$
+and $r \,\czip\, t \;=\; r \,\cpar\, t$.
+For bottom-pointed coeffect algebra, consider the typing derivation for the term
+$(\lambda x.e)\,v$ before reduction:
+%
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r}
+      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{r}} \tau_r } &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash (\lambda x.e_r)\,e_s : \tau_r }
+\end{equation*}
+%
+The derivation uses the idempotence of $\czip$ in the first step, followed by the 
+(\emph{app}) rule. The type of the term after substitution, using Lemma~\ref{thm:cbn-substitution-bot} is:
+%
+\begin{equation*}
+\inference
+  { \coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r & 
+    \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
+  { \coctx{\Gamma, x:\tau_r}{\cclrd{r}\;\cseq\;\cclrd{s}} \vdash \subst{e_r}{x}{e_s} : \tau_s }
+\end{equation*}
+%
+From the assumptions of Lemma~\ref{thm:cbn-substitution-bot}, we know that $\cseq\,=\,\cpar$
+and the operation is idempotent, so trivially:
+$\cclrd{r}\;\cseq\;\cclrd{s} = \cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})$
 \end{proof}
 
-ALSO: Something on eta expansion
-
-\subsection{Normal form}
+\paragraph{Expansion theorem.}
+The $\eta$-expansion (local completeness) is similar to $\beta$-reduction (local soundness) in that
+it holds for some flat coeffect systems, but not for all. Out of the examples we discuss, it 
+holds for implicit parameters, but does not hold for liveness and data-flow.
+
+Recall that $\eta$-expansion turns $e$ into $\lambxa x.e~x$. In the case of liveness, the 
+expression $e$ may require no variables (both immediate and latent coeffects are marked as
+$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
+marking the context and function argument as live. In case of data-flow, the coeffects are
+made larger by the lambda abstraction. We remedy this limitation in the next chapter.
+
+However, the $\eta$-expansion preserves the type for implicit parameters and, more generally,
+for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the 
+standard $\eta$-reduction:
+
+\begin{theorem}[$\eta$-expansion]
+\label{thm:cbn-substitution-bot}
+In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
+where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
+and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
+\end{theorem}
+\begin{proof}
+The following derivation shows that $\lambda x.e~x$ has the same type as $e$:
+\begin{equation*}
+\inference
+  { \inference
+    { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
+      \coctx{x:\tau_1}{\cunit} \vdash x:\tau_1 }
+    { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash e~x : \tau_2 } }
+  { \inference
+    { \inference
+      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash e~x : \tau_2 }
+      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash e~x : \tau_2 } }
+    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.e~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+\end{equation*}
+\end{proof}
+
+\noindent
+The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
+application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
+original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
+is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
+this is sufficient for showing type preservation.
+
+In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
+there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
+these include liveness and implicit parameters. Moreover, for implicit parameters (and more
+generally, any set-like flat coeffect algebra), the $\eta$-expansion holds as well, giving us
+both local soundness and local completeness as coined by Pfenning and Davies \cite{logic-modal-reconstruction}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Normal form?}
 \label{sec:flat-syntax-norm}
@@
 
+%
+% (e2 + (e1 * e2)) + (e3 * f)
+% ----(glet)------
+% ----------(app)------------
+%
+% = e2 + e1*e2 + e3*f
+
+% 
+% (e2 + (e3 * f)) + ((e2 + (e3 * f)) * e1)
+% ----(app)------    ----(app)-----
+% ---------------(glet)-------------------
+%
+% = e2 + e3*f + e2*f + e1*e3*f
+
+% Data-flow: (e1 + e2) max (e3 + f) = e1 + (e2 max (e3 + f))    XXXXX
+% Liveness:  e2 ^ (e3 v f) = e2 ^ (e3 v f)
+% Resources: OK
+
 This does not work for dataflow, because
@@
 
-or
+% e3 + ((e2 + (e1 * e2)) * e3)
+%       ------(glet)---
+% ---------(glet)-------------
+%
+% = e3 + e2*e3 + e1*e2*e3
+
+% (e2 + (e3 * e2)) + (e1 * (e2 + (e3 * e2)))
+% -----(glet)-----         -----(glet)-----
+% ------------------(glet)------------------
+%
+% = e2 + e3*e2 + e1*e2 + e1*e2*e3
+
+% Data-flow: e3 + (e2 + e1) = (e2 + e3) + e1
+% Liveness:  e3 = e3
+% Resources: OK
 

### 2014-07-15 text/Chapter01.tex
Finish equational theory, subtyping and lambda (Chapter 4)

@@
 mobile platforms). Web applications are split between client, server and mobile components; mobile
-applications must be aware of the physical environment while the ``internet
-of things`` makes the environments even more heterogeneous. At the same time, applications access
+applications must be aware of the physical environment while the ``internet of things'' makes
+the environments even more heterogeneous. At the same time, applications access
 rich data sources and need to be aware of security policies and provenance information from the

### 2014-07-15 text/Chapter04.tex
Finish equational theory, subtyping and lambda (Chapter 4)

@@
 \subsection{Typing of let binding}
+\label{sec:flat-calculus-let}
 
@@
 (\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
-(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized substitution})\\
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
 e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
@@
 
-\paragraph{Internalized substitution.}
-
-The (\emph{internalized substitution}) rule captures an operational semantics where the language
+\paragraph{Internalized sequencing.}
+
+The (\emph{internalized sequencing}) rule captures an operational semantics where the language
 provides a construct representing \emph{sequential composition} and expressions can be reduced to 
@@
 
-The (\emph{internalized substitution}) rule is useful when defining a concrete semantics for 
+The (\emph{internalized sequencing}) rule is useful when defining a concrete semantics for 
 a language that provides constructs for explicitly providing the context. For example, consider
@@
 
-Recall that $\eta$-expansion turns $e$ into $\lambxa x.e~x$. In the case of liveness, the 
+Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
 expression $e$ may require no variables (both immediate and latent coeffects are marked as
@@
 
-\subsection{Normal form?}
+\subsection{Internalized sequencing}
 \label{sec:flat-syntax-norm}
 
+The call-by-value and call-by-name evaluation strategies discussed in the previous section are
+the key techniques for defining equational theory of flat coeffects. In this section, we briefly 
+discuss another approach that follows the style of generic operational semantics designed by Filinski
+\cite{monads-inaction} for effectful computations.
+
+The idea is to embed sequencing of context-dependent computations as an explicit construct into
+the language and define a \emph{normal form} that consists of sequenced expressions. The reduction
+to the normal form is generic for all coeffect systems (satisfying certain conditions), while the
+reduction of the normal form is provided by each concrete coeffect system. The extended language
+with the \kvd{glet} construct and its typing is defined as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2 \sep \kvd{glet}~x = e_1~\kvd{in}~e_2\\
+n &::=& e \sep \kvd{glet}~x=e~\kvd{in}~n\\
+\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\end{array}
+\end{equation*}
 \begin{equation*}
@@
 \end{equation*}
-
-First transformation
-\begin{equation*}
-(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
-\end{equation*}
-
-%
-% (e2 + (e1 * e2)) + (e3 * f)
-% ----(glet)------
-% ----------(app)------------
-%
-% = e2 + e1*e2 + e3*f
-
-% 
-% (e2 + (e3 * f)) + ((e2 + (e3 * f)) * e1)
-% ----(app)------    ----(app)-----
-% ---------------(glet)-------------------
-%
-% = e2 + e3*f + e2*f + e1*e3*f
-
-% Data-flow: (e1 + e2) max (e3 + f) = e1 + (e2 max (e3 + f))    XXXXX
-% Liveness:  e2 ^ (e3 v f) = e2 ^ (e3 v f)
-% Resources: OK
-
-This does not work for dataflow, because
-
-(e1 + e2) max (e3 + f)
-e1 + (e2 max (e3 + f))
-
-But it works for the other ones...
-
-Second transformation
-\begin{equation*}
-\kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~
-\kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
-\end{equation*}
-
-% e3 + ((e2 + (e1 * e2)) * e3)
-%       ------(glet)---
-% ---------(glet)-------------
-%
-% = e3 + e2*e3 + e1*e2*e3
-
-% (e2 + (e3 * e2)) + (e1 * (e2 + (e3 * e2)))
-% -----(glet)-----         -----(glet)-----
-% ------------------(glet)------------------
-%
-% = e2 + e3*e2 + e1*e2 + e1*e2*e3
-
-% Data-flow: e3 + (e2 + e1) = (e2 + e3) + e1
-% Liveness:  e3 = e3
-% Resources: OK
-
-%
-% TODO: We could have a section about let binding covering the alternative CBV and the let binding
-% we included in the system, but we'd have to clean this up!
-%
-%\subsection{Syntactic let binding}
-%\begin{remark}
-%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
-%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
-%calculus of Section~\ref{}.
-%%%
-%\begin{align*}
-%{\footnotesize{\trule{let}}}
-%\dfrac
-%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
-%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
-%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
-%\end{align*}
-%%%
-%Therefore, for example, we see that \emph{let} has the coeffects
-%as substitution, for example:
-%%%
-%\begin{align*}
-%& {\footnotesize{\trule{let}}}
-%\dfrac
-%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
-%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
-%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
-%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
-%\end{align*}
-%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
-%the judgment:
-%%%
-%$$
-%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
-%$$
-%%%
-%\end{remark}
-%
-% 
-%\paragraph{Syntactical let binding}
-%Has the same typing derivation as abstraction/application
-%
-%\[
-%\inference
-% {\inference
-%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
-%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
-%  &
-%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
-%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
-%\]
-%
-%We can extend the CBV behaviour to \emph{let}, where the bound term
-%is evaluated to a pure value first, discharging any contextual
-%requirements before substitution into the body.
-%
-%\begin{lemma}[CBV \emph{let}]
-%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
-%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
-%: \tau$.
-%\end{lemma}
+%
+The newly introduced syntactic form $n$ represents a normal form. The construct $\kvd{glet}~x=e~\kvd{in}~n$
+models an explicit sequencing of an expression $e$, followed by an expression $n$. Note that \kvd{glet} can 
+only contain further \kvd{glet} constructs in the body, but not in the argument. The typing of the \kvd{glet}
+is the same as the typing of ordinary \kvd{let} construct -- as discussed in Section~\ref{sec:flat-calculus-types},
+for the examples we consider, this gives a \emph{more precise} typing than the typing of the
+term $(\lambda x.n)~e$.
+
+The reduction rules that produce a normal form are shown in Figure~\ref{fig:flat-normal-form}. The
+(\emph{eval}) rule introduces \kvd{glet} by reducing a redex $(\lambda x.e_2)~e_1$ to an expression
+representing explicit sequencing of $e_1$ and $e_2$. The remaining two rules specify how \kvd{glet}
+distributes with other constructs (\kvd{glet} and application). In (\emph{glet-app}), the \kvd{glet}
+construct appearing inside an application is lifted to the top-level; similarly (\emph{glet-glet}) 
+lifts a \kvd{glet} construct nested in the argument of another \kvd{glet}.
+
+As with \emph{call-by-value} and \emph{call-by-name} strategies, the \emph{internalized sequencing}
+strategy can only be used with coeffect systems satisfying certain conditions. The general form of
+the conditions is summarized in Appendix~?. Here, we briefly consider our three examples.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+\begin{equation*}
+\tag{\emph{eval}}
+(\lambda x.e_2)~e_1 ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~e_2
+\end{equation*}
+
+\begin{equation*}
+\tag{\emph{glet-app}}
+\begin{array}{l}
+  (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~\\ \qquad \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
+\end{array}
+\qquad x \notin \textit{fv}(e_3)\qquad
+\end{equation*}
+
+\begin{equation*}
+\tag{\emph{glet-glet}}
+\begin{array}{l}
+  \kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~\\
+  \qquad \kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
+\end{array}
+\qquad x_1 \notin \textit{fv}(e_3)\qquad
+\end{equation*}
+
+\caption{Reduction to normal form.}
+\label{fig:flat-normal-form}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Conditions.}
+The (\emph{eval}) reduction can be safely applied for any flat coeffect calculus which satisfies the
+condition that the typing of the \kvd{glet} expression (shown above) is equivalent or more precise
+than typing of the expression $(\lambda x.e_2)~e_1$. More formally:
+
+\begin{definition}
+A flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is \emph{oriented} if for all
+$\cclrd{s}, \cclrd{s_1}, \cclrd{s_2}, \cclrd{r} \in \C$ such that $\cclrd{s}=\cclrd{s_1}\,\czip\,\cclrd{s_2}$
+it is the case that $\cclrd{s} \;\cseq\; (\cclrd{s} \,\cpar\, \cclrd{r}) \;\cleq\; \cclrd{s_1} \;\cseq\; (\cclrd{s_2} \,\cpar\, \cclrd{r})$.
+\end{definition}
+
+\noindent
+As discussed in Section~\ref{sec:flat-calculus-let}, this condition is satisfied for all our
+three examples (strictly for liveness and data-flow; and by equality for implicit parameters).
+For the (\emph{glet-app}) and (\emph{glet-glet}) rules, additional conditions arise from the
+typing of the original and reduced expression (showed in Appendix~?). The results are summarized
+in the following two tables.
+
+Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
+$\coctx{\Gamma}{\cclrd{r_2}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{s}} \tau_2$ and
+$\coctx{\Gamma}{\cclrd{r_3}} \vdash e_3 : \tau_3$, the typings of the original and
+reduced expressions in (\emph{glet-app}) rule are:
+
+\begin{center}
+\begin{tabular}{ | l | c | c | c |}
+\hline
+  & \footnotesize Before\hspace{1em} & \footnotesize After\hspace{1em}  & \footnotesize Satisfied\hspace{1em} \\ \hline
+\hspace{-1em}{\footnotesize Parameters} 
+  & $\mathit{max}(\cclrd{r_1} + \cclrd{r_2}, \cclrd{r_3}+\cclrd{s})$
+  & $\cclrd{r_1}+\mathit{max}(\cclrd{r_2}, \cclrd{r_3} + \cclrd{s})$ & $\times$ \\ \hline
+\hspace{-1em}{\footnotesize Liveness} 
+  & $\cclrd{r_2} \sqcap (\cclrd{r_3} \sqcup \cclrd{s})$ 
+  & $\cclrd{r_2} \sqcap (\cclrd{r_3} \sqcup \cclrd{s})$ & \checkmark \\ \hline
+\hspace{-1em}{\footnotesize Data-flow} 
+  & $(\cclrd{r_2} \cup \cclrd{r_1}) \cup (\cclrd{r_3} \cup \cclrd{s}) $ 
+  & $(\cclrd{r_2} \cup (\cclrd{r_3} \cup \cclrd{s})) \cup \cclrd{r_1} $ & \checkmark \\ \hline
+\end{tabular}
+\end{center}
+
+Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
+$\coctx{\Gamma}{\cclrd{r_2}} \vdash e_2 : \tau_2$ and
+$\coctx{\Gamma}{\cclrd{r_3}} \vdash e_3 : \tau_3$, the typings of the original and
+reduced expressions in (\emph{glet-glet}) rule are:
+
+\begin{center}
+\begin{tabular}{ | l | c | c | c |}
+\hline
+  & \footnotesize Before\hspace{1em} & \footnotesize After\hspace{1em}  & \footnotesize Satisfied\hspace{1em} \\ \hline
+\hspace{-1em}{\footnotesize Parameters} 
+  & $\cclrd{r_3} + (\cclrd{r_2} + \cclrd{r_1}) $ 
+  & $(\cclrd{r_3} + \cclrd{r_2}) + \cclrd{r_1}$ & \checkmark \\ \hline
+\hspace{-1em}{\footnotesize Liveness} 
+  & $\cclrd{r_3}$ 
+  & $\cclrd{r_3}$ & \checkmark \\ \hline
+\hspace{-1em}{\footnotesize Data-flow} 
+  & $\cclrd{r_3} \cup (\cclrd{r_2} \cup \cclrd{r_1}) $ 
+  & $(\cclrd{r_3} \cup \cclrd{r_2}) \cup \cclrd{r_1}$ & \checkmark \\ \hline
+\end{tabular}
+\end{center}
+
+This means that \emph{internalized sequencing} provides a basis for operational semantics of
+liveness and implicit parameters, but not for data-flow languages. For other coeffect systems,
+the conditions in Appendix~? have to be re-examined.
 
@@
 
-\section{Syntactic extensions}
+\section{Syntactic properties and extensions}
 \label{sec:flat-exts}
 
-\subsection{Lambda abstraction}
+The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of laws.
+The laws are required for three distinct reasons -- to be able to define the categorical structure
+in Section~\ref{sec:flat-semantics}, to prove equational properties in Section~\ref{sec:flat-syntax}
+and finally, to guarantee intuitive syntactic properties for constructs such as 
+$\lambda$-abstraction and pairs in context-aware calculi.
+
+In this section, we look at the last point. We discuss what syntactic equivalences are permitted
+by the properties of $\czip$, we extend the calculus with pairs and units and discuss their 
+syntactic properties and, finall, we consider properties arising from recursion. In the following
+section, we further develop subtyping relation for the calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Subtyping for coeffects}
+
+The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
+which makes it possible to treat an expression with smaller context requirements as an expression
+with greater context requirements. In the corresponding categorical semantics, this means that
+we can \emph{drop} some of the provided context.
+
+Figure~\ref{fig:flat-types-sub} adds sub-typing on function types, making it possible to treat
+a function with smaller context requirements as a function with greater context requirements. 
+The definition uses the standard reflexive and transitive $<:$ operator. As the (\emph{sub-fun})
+shows, the function type is contra-variant in the input and co-variant in the output. The 
+(\emph{typ}) rule allows using sub-typing on an expression type in the coeffect calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+\begin{equation*}
+\boxed{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
+\end{equation*}
+
+\begin{equation*}
+\tyrule{typ}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau & \tau <: \tau' }
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' }
+\quad
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau & \cclrd{r'} \cleq \cclrd{r} }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
+\end{equation*}
+
+\begin{equation*}
+\boxed{\tau <: \tau'}
+\end{equation*}
+
+\begin{equation*}
+\tyrule{sub-trans}
+  { \tau_1 <: \tau_2 & \tau_2 <: \tau_3 }
+  { \tau_1 <: \tau_3  }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub-fun}
+  { \tau_1' <: \tau_1 & \tau_2 <: \tau_2' & \cclrd{r'} \cgeq \cclrd{r} }
+  { \tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2' }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub-refl}
+  { }
+  { \tau <: \tau }
+\end{equation*}
+
+\caption{Subtyping rules for flat coeffect calculus}
+\label{fig:flat-types-sub}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Semantics.} We follow the same approach as with the rest of the calculus and use
+a categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
+of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
+As shown in Figure~\ref{fig:flat-semantics-sub}, the seamntics of the sub-typing rule (\emph{typ}) 
+then just composes the semantics of the original expression with the conversion produced by the 
+semantics of the sub-typing judgement.
+
+The rest of the Figure~\ref{fig:flat-semantics-sub} shows the rules that define the semantics
+of $<:$. The reflexivity and transitivity are just the identity function and function composition,
+respectively. The (\emph{sub-fun}) case is interesting -- recall that the semantics of a function
+$\tau_1' \xrightarrow{\cclrd{r'}} \tau_2'$ is $C^{\cclrd{r'}} \tau_1' \rightarrow \tau_2'$. To 
+build the required function, we first drop unnecessary context using $\ident{lift}_{\cclrd{r'}, \cclrd{r}}
+ : C^{\cclrd{r'}} \tau_1' \rightarrow C^{\cclrd{r}} \tau_1'$ and use the $\ident{map}_{\cclrd{r}}$
+function to transform the nested $\tau_1'$ to $\tau_1$. Then we evaluate the original function $f$
+and turn the resulting $\tau_2$ into the required result of type $\tau_2'$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+\begin{equation*}
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' } = 
+  \sem{\tau <: \tau'} \circ \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } \qquad\quad (\emph{typ})
+\end{equation*}
+
+\begin{equation*}
+\begin{array}{ll}
+\sem{\tau <: \tau} = \ident{id} & (\emph{sub-refl})
+\\[0.5em]
+\sem{\tau_1 <: \tau_3} = \sem{\tau_2 <: \tau_3} \circ \sem{\tau_1 <: \tau_2} & (\emph{sub-trans})
+\\[0.5em]
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} = \lambda f. & (\emph{sub-fun}) \\
+\qquad\sem{\tau_2 <: \tau_2'} \circ f \circ \ident{map}_{\cclrd{r}}~\sem{\tau_1' <: \tau_1} 
+  \circ \ident{lift}_{\cclrd{r'}, \cclrd{r}}
+\end{array}
+\end{equation*}
+
+\caption{Semantics of subtyping for flat coeffects}
+\label{fig:flat-semantics-sub}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Alternative lambda abstraction}
+
+In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
+lambda abstraction (for implicit parameters, the lambda function splits context requirements
+using $\cclrd{r} \cup \cclrd{s}$; for data-flow it suffices to duplicate the requirement 
+$\cclrd{r}$). We introduced the $\czip$ operation as a way of providing the additional abstraction.
+Here, we identify coeffect calculi for which the simpler (duplicating) rule is sufficient.
+
+\paragraph{Idempotence.}
+Recall that $(\C, \czip)$ is a band, meaning that $\czip$  is idempotent and associative. The 
+idempotence means that the context requirements of the body can be required from both the 
+declaration site and the call site. Thus, the following (\emph{idabs}) typing is valid 
+(for reference, it is shown side-by-side with the ordinary lambda abstraction rule):
+%
+\begin{equation*}
+\tyrule{idabs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\,\czip\,\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
+\end{equation*}
+% 
+To derive (\emph{idabs}), we use idempotence on the body annotation $\cclrd{r}\;=\;\cclrd{r}\,\czip\,\cclrd{r}$
+and then use the standard (\emph{abs}) rule. So, (\emph{idabs}) follows from (\emph{abs}), 
+but the other direction is not necessarily the case. The condition below identifies 
+coeffect calculi where (\emph{abs}) follows from (\emph{idabs}).
+
+\begin{definition}
+A flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is \emph{strictly oriented} if for all
+$\cclrd{s}, \cclrd{r} \in \C$ it is the case that $\cclrd{r} \,\czip\, \cclrd{s} \;\cleq\; \cclrd{r}$.
+\end{definition}
+
+\begin{remark}
+\label{thm:flat-alt-abs}
+For a flat coeffect calculus with a strictly oriented algebra, the standard (abs) rule can 
+be derived from the (idfun) rule.
+\end{remark}
+\begin{proof}
+The following derives the conclusion of (\emph{abs}) using (\emph{abs}), sub-coeffecting, 
+sub-typing and the fact that the algebra is \emph{stricly oriented}:
+
+\begin{equation*}
+\tyrule{typ}
+  {\hspace{-5em} \tyrule{sub}
+     {\hspace{-4em} \tyrule{idabs}  
+        {\coctx{\Gamma, x:\tau_1}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash e : \tau_2}
+        {\coctx{\Gamma}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r} \,\czip\, \cclrd{s}} \tau_2 } }
+     {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r} \,\czip\, \cclrd{s}} \tau_2} \;
+           \textnormal{\footnotesize{($\cclrd{r} \cleq \cclrd{r} \,\czip\, \cclrd{s}$)}} \hspace{-5em} }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} \; 
+           \textnormal{\footnotesize{($\cclrd{r} \cleq \cclrd{r} \,\czip\, \cclrd{s}$)}}
+\end{equation*}
+\end{proof}
+
+\noindent
+The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
+oriented coeffect calculi (such as our liveness and data-flow computations), we can use
+the (\emph{idabs}) rule and get an equivalent type system. This alternative formulation
+removes the non-determinism that arises from the splitting of context requirements in 
+the original (\emph{abs}) rule.
+
+\paragraph{Symmetry.}
+The $\czip$ operation is idempotent and associative. In all of the three examples considered in 
+this chapter, the operation is also \emph{symmetric}. To make our definitions more general, we
+do not require this to be the case for \emph{all} flat coeffect systems. However, systems with
+symmetric $\czip$ have the following property.
+
+\begin{remark}
+For a flat coeffect calculus such that $\cclrd{r}\,\czip\,\cclrd{s} = \cclrd{s}\,\czip\,\cclrd{r}$,
+assuming that $\cclrd{r'}, \cclrd{s'}, \cclrd{t'}$ is a permutation of $\cclrd{r},\cclrd{s},\cclrd{t}$:
+%
+\begin{equation*}
+\inference
+  {\coctx{\Gamma, x:\tau_1, y:\tau_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t}} \vdash e : \tau_3}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash \lambda x.\lambda y.e : 
+      \tau_1 \xrightarrow{\cclrd{s'}} (\tau_2 \xrightarrow{\cclrd{t'}} \tau_3) }
+\end{equation*} 
+\end{remark}
+
+\noindent
+Intuitively, this means that the context requirements of a function with multiple arguments can be 
+split arbitrarily between the declaration site and (multiple) call sites. In other words, it does
+not matter how the context requirements are satisfied.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Constants and pairs}
+
+The flat coeffect calculus annotates constants with $\czero$ (which is needed, e.g.~for liveness).
+To support pairs, we add the following rule for the constructor $(e_1, e_2)$:
+%
+\begin{equation*}
+\rulefcl{pair}
+  {\coctx{r}{\Gamma} \vdash e_1 : \tau_1 & \coctx{s}{\Gamma} \vdash e_2 : \tau_2}
+  {\coctx{r \fmax s}{\Gamma} \vdash (e_1, e_2) : \tau_1 \times \tau_2 }
+\end{equation*}
+%
+The operator $\fmax$ combines context requirements of the two sub-expressions. It models
+the case when the available context is split and passed to two independent expressions. The rule
+for projection is uninteresting, because $\pi_i$ can be viewed as a pure function.
+
+In a language with a constant $() : \ident{unit}$ and pairs, we would expect the following 
+isomorphisms to hold:
+%
+\begin{equation*}
+(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)
+\quad
+(e_1, e_2) \simeq (e_2, e_1)
+\quad
+((), e) \simeq e
+\end{equation*}
+%
+The coeffect is not a part of the type of pairs, so the types are isomorphic. The isomorphism
+of the context requirements is guaranteed by the fact that $(S, \cpar, \czero)$ is a bounded
+semilattice using associativity, symmetry and boundedness, respectively.
+
+% --------------------------------------------------------------------------------------------------
+
+$(\C, \cseq, \cunit)$ is monoid
+
+$(\C, \cpar, \czero)$ is monoid
+
+$(\C, \czip)$ is a band (associative, idempotent)
+
+$(r \cpar s) \cseq t = (r \cseq t) \cpar (s \cseq t)$
+
+$t \cseq (r \cpar s) = (t \cseq r) \cpar (t \cseq s)$
+
 \subsection{Recursion}
@@
 \section{----------------JUNK!}
-
-% ==================================================================================================
-
-~
-\newpage
-~
-
-\newcommand{\ccat}[0]{\mathcal{C}}
-\newcommand{\cobind}[2]{#1^\dagger_{#2}}
-\newcommand{\cmerge}[0]{ \ident{m} }
-\newcommand{\csplit}[0]{ \ident{n} }
-\newcommand{\counit}[0]{ \varepsilon }
-\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
-\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}
-
-% ==================================================================================================
-
-\section{Syntax-based equational theory}
-\label{sec:syntactic}
-
-For an equational theory of coeffects, consider first a notion 
-of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
-has the following type and coeffect rule:
-%
-\begin{equation}
-\inference
-  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
-   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
-  {\ctyp{r_1 \vee (r_2 \cseq s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
-: \tau_2 }
-\label{eq:let1}
-\end{equation}
-
-\noindent
-For our examples, $\wedge$ is idempotent (\ie{}, $r \wedge r = r$)
-implying a simpler rule:
-
-% For the three examples we consider, a simpler rule gives a more
-% precise coeffect. Because $r \wedge r = r$ for all our examples, we
-% can also a coeffect $r \vee (r \cseq s)$. Moreover,
-%for our examples (but not necessarily for \emph{all} coeffect
-%systems), $r \vee (r \cseq s) \leq r_1 \vee (r_2 \cseq s)$ meaning
-%that the following gives more precise coeffects:
-%
-\begin{equation}
-\inference
-  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
-   \ctyp{r}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
-  {\ctyp{r \vee (r \cseq s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2 : \tau_2 }
-\label{eq:let2}
-\end{equation}
-%%
-For our examples (but not necessarily \emph{all} coeffect
-systems), this defines a more ``precise'' coeffect with respect to $\leq$
-where $r \vee (r \cseq s) \leq r_1 \vee (r_2 \cseq s)$.
-
-This rule removes the non-principality of the first rule
-(\ie~multiple possible typings).  However, using idempotency 
-to split coeffects in abstraction would remove additional
-flexibility needed by the implicit parameters example.
-
-The coeffect $r \vee (r \cseq s)$ can
-also be simplified for all our examples, leading to more intuitive
-rules -- for implicit parameters $r \cup (r \cup s) = r \cup s$; for
-liveness we get that $r \sqcup (r \sqcap s) = r$ and for dataflow we
-obtain $\textit{max}(r, r+s) = r + s$.
-
-Our calculus can be extended with \emph{let}-binding and \eqref{eq:let2}.
-However, we also consider the cases when a
-syntactic substitution $e_2[x \leftarrow e_1]$ has the coeffects
-specified by the above rule \eqref{eq:let2} 
-and prove \emph{subject reduction} theorem
-for certain coeffect calculi.  We consider two common special cases
-when the coeffect of variables $\cunit$ is the greatest ($\top$) or
-least ($\bot$) element of the semi-lattice $(S, \vee)$ and derive
-additional conditions that have to hold about the coeffect algebra:
-
-\begin{lemma}[Substitution]
-\label{thm:subst}
-Given $C^r (\Gamma, x : \tau_2) \vdash e_1 : \tau_1$ and $C^s \Gamma \vdash e_2 : \tau_2$
-then $C^{r \vee (r \oplus s)} \Gamma \vdash \subst{e_2}{x}{e_1} : \tau_1$ if 
-the coeffect algebra satisfies the conditions that 
-$\cunit$ is either the greatest or least element of the semi-lattice,
-$\oplus = \wedge$, and $\oplus$ distributes over $\vee$,
-\ie{}, $X \oplus (Y \vee Z) = (X \oplus Y) \vee (X \oplus Z)$.
-\end{lemma}
-
-\begin{proof}
-By induction over $\vdash$, using the laws (\S\ref{sec:calculus}) and additional assumptions.
-\end{proof}
-
-Assuming $\rightarrow_\beta$ is the usual call-by-name reduction, the
-following theorem models the evaluation of coeffect calculi with
-coeffect algebra that satisfies the above requirements. We do not
-consider \emph{call-by-value}, because our calculus does not have a
-notion of \emph{value}, unless explicitly provided by
-\emph{let}-binding (even a function ``value'' $\lambda x.e$ may have
-immediate contextual requirements).
-
-\begin{theorem}[Subject reduction]
-\label{thm:reduction}
-For a coeffect calculus, satisfying the conditions of Lemma~\ref{thm:subst}, if
-$\ctyp{r}{\Gamma} \vdash e : \tau$ and $e \rightarrow_\beta e'$ then 
-$\ctyp{r}{\Gamma} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-A direct consequence of Lemma~\ref{thm:subst}. 
-\end{proof}
-
-The above theorem holds for both the liveness and resources examples,
-but not for dataflow.  In the case of liveness, $\cunit$ is the
-greatest element ($r \vee \cunit = \cunit$); in the case of
-resources, $\cunit$ is the \emph{least} element ($r \vee \cunit =
-r$) and the proof relies on the fact that additional
-context-requirements can be placed at the context $\ctyp{r}{\Gamma}$
-(without affecting the type of function when substituted under
-$\lambda$ abstraction).
-
-However, the coeffect calculus also captures context-dependence in
-languages with more complex evaluation strategies than
-\emph{call-by-name} reduction based on syntactic substitution.  In
-particular, syntactic substitution does not provide a suitable evaluation
-for dataflow (because a substituted expression needs to capture the
-context of the original scope).
-
-Nevertheless, the above results show that -- unlike effects --
-context-dependent properties can be integrated with
-\emph{call-by-name} languages. Our work also provides a model of
-existing work, namely Haskell implicit parameters
-\cite{app-implicit-parameters}.
-
-% ==================================================================================================
-
-\section{Related and further work}
-\label{sec:related}
 
@@
 To obtain a pure lambda abstraction
-for coeffects, we need to restrict the $\cmerge_{r, s}$ 
+for coeffects, we need to restrict the $\cpar_{r, s}$ 
 operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
-no effects): $\ccat^{\cunit} A \times \ccat^r B \rightarrow \ccat^{r}(A \times B)$.
+no effects): $\C^{\cunit} A \times \C^r B \rightarrow \C^{r}(A \times B)$.
 

### 2014-08-02 text/Chapter04.tex
Finish pairs and units (Chapter 4)

@@
   and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
-  with pairs and recursion (Section~\ref{sec:flat-exts}).
+  with subtyping and pairs (Section~\ref{sec:flat-exts}).
 
@@
   The semantics provides deeper insight into how (and why) the calculus works.
-
-\item We develop an alternative presentation of the system in terms of a simple structure 
-  (semi-lattice) and use it to develop a type inference algorithm for flat coeffect calculus.
 \end{itemize}
@@
 %
-We discuss pairs and recursion in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
+We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
 represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
@@
 call-site and declaration-site, \ie~it makes the (\emph{abs-df}) rule admissible. In some cases, 
-the operator forms a monoid with the unit being the greatest element of the set. This alternative is 
-discussed when we consider recursion (Section~\ref{sec:flat-exts}).
+the operator forms a monoid with the unit being the greatest element of the set. 
 \end{itemize}
@@
 return to let-binding after looking a number of examples. Additional constructs such 
-as recursion and tuples are covered in Section~\ref{sec:flat-exts}.
+as pairs are covered in Section~\ref{sec:flat-exts}.
 
@@
 In this section, we look at the last point. We discuss what syntactic equivalences are permitted
-by the properties of $\czip$, we extend the calculus with pairs and units and discuss their 
-syntactic properties and, finall, we consider properties arising from recursion. In the following
-section, we further develop subtyping relation for the calculus.
+by the properties of $\czip$ and we extend the calculus with pairs and units and discuss their 
+syntactic properties. In the following section, we further develop subtyping relation for the 
+calculus.
 
@@
 
-\subsection{Constants and pairs}
-
-The flat coeffect calculus annotates constants with $\czero$ (which is needed, e.g.~for liveness).
-To support pairs, we add the following rule for the constructor $(e_1, e_2)$:
-%
-\begin{equation*}
-\rulefcl{pair}
-  {\coctx{r}{\Gamma} \vdash e_1 : \tau_1 & \coctx{s}{\Gamma} \vdash e_2 : \tau_2}
-  {\coctx{r \fmax s}{\Gamma} \vdash (e_1, e_2) : \tau_1 \times \tau_2 }
-\end{equation*}
-%
-The operator $\fmax$ combines context requirements of the two sub-expressions. It models
-the case when the available context is split and passed to two independent expressions. The rule
-for projection is uninteresting, because $\pi_i$ can be viewed as a pure function.
-
-In a language with a constant $() : \ident{unit}$ and pairs, we would expect the following 
-isomorphisms to hold:
-%
-\begin{equation*}
-(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)
-\quad
-(e_1, e_2) \simeq (e_2, e_1)
-\quad
-((), e) \simeq e
-\end{equation*}
-%
-The coeffect is not a part of the type of pairs, so the types are isomorphic. The isomorphism
-of the context requirements is guaranteed by the fact that $(S, \cpar, \czero)$ is a bounded
-semilattice using associativity, symmetry and boundedness, respectively.
-
-% --------------------------------------------------------------------------------------------------
-
-$(\C, \cseq, \cunit)$ is monoid
-
-$(\C, \cpar, \czero)$ is monoid
-
-$(\C, \czip)$ is a band (associative, idempotent)
-
-$(r \cpar s) \cseq t = (r \cseq t) \cpar (s \cseq t)$
-
-$t \cseq (r \cpar s) = (t \cseq r) \cpar (t \cseq s)$
-
-\subsection{Recursion}
-
-% ==================================================================================================
-
-\section{Type inference}
-\label{sec:flat-inference}
-
-\subsection{Semi-lattice formulation}
-\subsection{Type inference algorithm}
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{pair}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 & \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\cclrd{r}\,\cpar\,\cclrd{s}}{\Gamma} \vdash (e_1, e_2) : \tau_1 \times \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{proj}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \times \tau_2 }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \pi_i~e : \tau_{i} }
+\end{equation*}
+\begin{equation*}
+\tyrule{unit}
+  {}
+  {\coctx{\Gamma}{\czero} \vdash () : \ident{unit} }
+\end{equation*}
+
+\caption{Typing rules for pairs and units}
+\label{fig:flat-ext-types}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Language with pairs and unit}
+
+The calculus introduced in Section~\ref{sec:flat-calculus} consisted only of variables, abstraction, 
+application and let binding to show the key aspects of flat coeffect systems. Here, we extend it 
+with pairs and the unit value to sketch how it can be turned into a full programming language.
+The syntax of the language is extended as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& \ldots \sep () \sep e_1, e_2 \\
+\tau &::=& \ldots \sep \ident{unit} \sep \tau \times \tau
+\end{array}
+\end{equation*}
+%
+The typing rules for pairs and the unit value are shown in Figure~\ref{fig:flat-ext-types}.
+The unit value (\emph{unit}) is annotated with the $\czero$ coeffect (the same as other constants).
+Pairs, created using the $(e_1, e_2)$ expression, are annotated with a coeffect that combines
+the coeffects of the two sub-expressions using the \emph{point-wise} operator $\cpar$. The operator
+models the case when the (same) available context is split and passed to two independent 
+sub-expressions. This matches the semantics of pairs discussed shortly. Finally, the (\emph{proj})
+rule is uninteresting, because $\pi_i$ can be viewed as a pure function.
+
+\paragraph{Properties.}
+Pairs and the unit value in a lambda calculus should form a monoid -- associativity means that 
+the expression $(e_1, (e_2, e_3))$ should be isomorphic to $((e_1, e_2), e_3)$ and that
+$((), e) \simeq e \simeq (e, ())$, where the isomorphism appropriately transforms the values,
+without affecting other properties (here coeffects) of the expression.
+
+In the following, we assume that \ident{assoc} is a pure function transforming a pair $(x_1, (x_2, x_3))$ 
+to a pair $((x_1, x_2), x_3)$. We write $e \equiv e'$ when for all $\Gamma, \tau$ and $\cclrd{r}$,
+it is the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ if and only if
+$\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+
+\begin{theorem}
+For a flat coeffect calculus with pairs and units, the following holds:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \ident{assoc}~(e_1, (e_2, e_3)) &\equiv& ((e_1, e_2), e_3) &\qquad\qquad&(\emph{associativity}) \\
+ \pi_1~(e, ()) &\equiv& e &&(\emph{right unit})\\
+ \pi_2~((), e) &\equiv& e &&(\emph{left unit})\\
+\end{array} 
+\end{equation*}
+\end{theorem}
+\begin{proof}
+Follows from the fact that $(\C, \cpar, \czero)$ is a monoid and \ident{assoc}, $\pi_1$ and
+$\pi_2$ are pure functions (treated as contstants in the langauge).
+\end{proof}
+
+The above properties follow from the laws of the flat coeffect algebra. In addition, if the 
+$\cpar$ operation is symmetric (which is the case for all our examples in this chapter), it 
+also holds that $\ident{swap}~(e_1, e_2) \equiv (e_2, e_1)$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Language with recursion}
+
+\begin{equation*}
+\tyrule{loop}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \ident{bool} &
+    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \ident{unit} }
+  { \coctx{\Gamma}{\cclrd{r} \;\cseq\; (\cclrd{s} \,\cseq\, \cclrd{r})^{\cclrd{*}} } \vdash \kvd{while}~e_1~\kvd{do}~e_2 : \ident{unit} }
+\end{equation*}
+
+\begin{equation*}
+\tyrule{rec}
+  { \coctx{\Gamma, x:\tau_1 \xrightarrow{\cclrd{\phi}} \tau_2}{\cclrd{s}} \vdash e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
+  { \coctx{\Gamma}{\cclrd{s^{\ast}}} \vdash \kvd{fix}~x. e : \tau_1 \xrightarrow{\ident{fix}~\cclrd{\phi}.\cclrd{r}} \tau_2 }
+\end{equation*}
 
@@
 
-% ==================================================================================================
-
-\section{Summary}
-
-\section{----------------JUNK!}
-
-This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
-and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
-\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
-may split contextual requirements between the declaration-site and call-site.
-
-Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
-\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
-and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
-\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
-general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\paragraph{Comonads and modal logics.}
-
-Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
-$\square$ modality of an intuitionistic S4 modal logic using monoidal
-comonads, which links our calculus to modal logics.  This link can be
-materialized in two ways.
-
-Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
-correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
-building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
-\cite{monad-notions}) that includes $\square$ as a type
-constructor. For example, in \cite{logic-modal-reconstruction}, the
-modal type $\Box \tau$ represents closed terms.
-In contrast, the \emph{semantic} approach uses monads or comonads
-\emph{only} as a semantics.  This has been employed by Uustalu and
-Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
-follow the semantic approach.
-
-Nanevski et al. extend an S4 term language to a \emph{contextual}
-modal type theory (CMTT)~\cite{logic-cmtt}.
-The \emph{context} is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. Our contextual types are
-indexed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}.
-
-The work on CMTT suggests two extensions to coeffects. The first is
-developing the logical foundations. We briefly considered special cases
-of our system that permits local soundness in \S\ref{sec:syntactic} and
-local completeness can be treated similarly. The second problem is 
-developing the coeffects \emph{metalanguage}. The use of coeffect algebras
-would provide an additional flexibility over CMTT, allowing a wider range 
-of applications.
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\paragraph{Relating effects and coeffects.} 
 The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
@@
 
-\newcommand{\cprd}{\times}
-\newcommand{\cvop}{\oplus}
-
-\vspace{-1em}
-\paragraph{Structural coeffects.} To make the liveness analysis practical, we need to associate
-information with individual variables (rather than the entire context). We can generalize the 
-calculus from this paper by adding a product operation $\times$ to the coeffect algebra.
-A variable context $x:\tau_1, y:\tau_2, z:\tau_3$ is then annotated with
-$r\times s \times t$ where each component of the tag corresponds to a single variable. The system
-then needs to be extended with structural rules such as:
-%
-\begin{equation*}
-\inference[(\emph{abs})]
-  {\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\quad
-\inference[(\emph{contr})]
-  {\ctyp{r \cprd s}{(x:\tau_1, y:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r \cvop s }{(z:\tau_1)} \vdash \subst{\subst{e}{x}{z}}{y}{z} : \tau_2 }
-\end{equation*}
-%
-The context-requirements associated with function are exactly those linked to the specific
-variable of the lambda abstraction. Rules such as contraction manipulate variables and perform
-a corresponding operation on the indices.
-
-The structural coeffect system is related to bunched typing \cite{types-bunched} (but generalizes
-it by adding indices). We are currently investigating how to use structural coeffects to capture 
-fine-grained context-dependence properties such as secure information flow \cite{app-secure-flow}
-or, more generally, those captured by dependency core calculus \cite{types-dcc}.
-
 % ==================================================================================================

### 2014-08-02 text/Chapter07.tex
Finish pairs and units (Chapter 4)

@@
 \label{sec:metalanguage-intro}
+
+% ==================================================================================================
+
+\section{Type inference}
+\label{sec:flat-inference}
+
+\subsection{Semi-lattice formulation}
+\subsection{Type inference algorithm}
+
+% ==================================================================================================
 
@@
 
+\section{Also related work}
 \section{Summary}
+
+This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
+and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
+\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
+may split contextual requirements between the declaration-site and call-site.
+
+Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
+\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
+and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
+\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
+general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).
+
+% --------------------------------------------------------------------------------------------------
+
+\vspace{-1em}
+\paragraph{Comonads and modal logics.}
+
+Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
+$\square$ modality of an intuitionistic S4 modal logic using monoidal
+comonads, which links our calculus to modal logics.  This link can be
+materialized in two ways.
+
+Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
+correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
+building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
+\cite{monad-notions}) that includes $\square$ as a type
+constructor. For example, in \cite{logic-modal-reconstruction}, the
+modal type $\Box \tau$ represents closed terms.
+In contrast, the \emph{semantic} approach uses monads or comonads
+\emph{only} as a semantics.  This has been employed by Uustalu and
+Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
+follow the semantic approach.
+
+Nanevski et al. extend an S4 term language to a \emph{contextual}
+modal type theory (CMTT)~\cite{logic-cmtt}.
+The \emph{context} is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. Our contextual types are
+indexed by a coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, \emph{etc.}.
+
+The work on CMTT suggests two extensions to coeffects. The first is
+developing the logical foundations. We briefly considered special cases
+of our system that permits local soundness in \S\ref{sec:syntactic} and
+local completeness can be treated similarly. The second problem is 
+developing the coeffects \emph{metalanguage}. The use of coeffect algebras
+would provide an additional flexibility over CMTT, allowing a wider range 
+of applications.
+
+% --------------------------------------------------------------------------------------------------
+
+\newcommand{\cprd}{\times}
+\newcommand{\cvop}{\oplus}
+
+\vspace{-1em}
+\paragraph{Structural coeffects.} To make the liveness analysis practical, we need to associate
+information with individual variables (rather than the entire context). We can generalize the 
+calculus from this paper by adding a product operation $\times$ to the coeffect algebra.
+A variable context $x:\tau_1, y:\tau_2, z:\tau_3$ is then annotated with
+$r\times s \times t$ where each component of the tag corresponds to a single variable. The system
+then needs to be extended with structural rules such as:
+%
+\begin{equation*}
+\inference[(\emph{abs})]
+  {\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
+\quad
+\inference[(\emph{contr})]
+  {\ctyp{r \cprd s}{(x:\tau_1, y:\tau_1)} \vdash e : \tau_2}
+  {\ctyp{r \cvop s }{(z:\tau_1)} \vdash \subst{\subst{e}{x}{z}}{y}{z} : \tau_2 }
+\end{equation*}
+%
+The context-requirements associated with function are exactly those linked to the specific
+variable of the lambda abstraction. Rules such as contraction manipulate variables and perform
+a corresponding operation on the indices.
+
+The structural coeffect system is related to bunched typing \cite{types-bunched} (but generalizes
+it by adding indices). We are currently investigating how to use structural coeffects to capture 
+fine-grained context-dependence properties such as secure information flow \cite{app-secure-flow}
+or, more generally, those captured by dependency core calculus \cite{types-dcc}.
+
+\section{Summary}

### 2014-08-22 text/Chapter04.tex
Drop recursion section (Chapter 4)
Calculus and semantics work (Chapter 5)

@@
 % ==================================================================================================
+%                                                                                      
+%     ###           #                       #                 #       #                 
+%      #            #                       #                 #                         
+%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
+%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
+%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
+%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
+%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
+%                                                                                     
+% ==================================================================================================
 
@@
 \cite{monads-effects-marriage}, relating effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ 
-to monadic computations $M^\sigma \tau_1 \rightarrow \tau_2$. In this chapter, we show a similar
+to monadic computations $\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar
 correspondence between \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry 
@@
 
+% ==================================================================================================
+%                                                            
+%       ###           ##                   ##                 
+%      #   #           #                    #                 
+%      #       ###     #     ###   #   #    #    #   #   ###  
+%      #          #    #    #   #  #   #    #    #   #  #     
+%      #       ####    #    #      #   #    #    #   #   ###  
+%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
+%       ###    ####   ###    ###    ## #   ###    ## #  ####  
+%                                                           
 % ==================================================================================================
@@
 \subsection{Examples of flat coeffects}
+\label{sec:flat-calculus-examples}
 
@@
 
+% ==================================================================================================
+%                                              
+%      #####  #                                 
+%        #    #                                 
+%        #    # ##    ###    ###   # ##   #   # 
+%        #    ##  #  #   #  #   #  ##  #  #   # 
+%        #    #   #  #####  #   #  #      #  ## 
+%        #    #   #  #      #   #  #       ## # 
+%        #    #   #   ###    ###   #          # 
+%                                         #   # 
+%                                          ###                                                                         
 % ==================================================================================================
@@
 % ==================================================================================================
+%                                                                   
+%       ###                  #                    #       #          
+%      #   #                 #                    #                  
+%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
+%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
+%          #  #  ##  #   #   #      ####  #       #       #    #     
+%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
+%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
+%             #   #                                                  
+%              ###                                                   
+%
+% ==================================================================================================
 
@@
 % ==================================================================================================
+%
+%      #####          #                            #                        
+%      #              #                                                     
+%      #      #   #  ####    ###   # ##    ###    ##     ###   # ##    ###  
+%      ####    # #    #     #   #  ##  #  #        #    #   #  ##  #  #     
+%      #        #     #     #####  #   #   ###     #    #   #  #   #   ###  
+%      #       # #    #  #  #      #   #      #    #    #   #  #   #      # 
+%      #####  #   #    ##    ###   #   #  ####    ###    ###   #   #  ####  
+%
+% ==================================================================================================
 
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Language with recursion}
-
-\begin{equation*}
-\tyrule{loop}
-  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \ident{bool} &
-    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \ident{unit} }
-  { \coctx{\Gamma}{\cclrd{r} \;\cseq\; (\cclrd{s} \,\cseq\, \cclrd{r})^{\cclrd{*}} } \vdash \kvd{while}~e_1~\kvd{do}~e_2 : \ident{unit} }
-\end{equation*}
-
-\begin{equation*}
-\tyrule{rec}
-  { \coctx{\Gamma, x:\tau_1 \xrightarrow{\cclrd{\phi}} \tau_2}{\cclrd{s}} \vdash e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
-  { \coctx{\Gamma}{\cclrd{s^{\ast}}} \vdash \kvd{fix}~x. e : \tau_1 \xrightarrow{\ident{fix}~\cclrd{\phi}.\cclrd{r}} \tau_2 }
-\end{equation*}
-
 % ==================================================================================================
+%                                                 
+%      ####           ##            #                # 
+%      #   #           #            #                # 
+%      #   #   ###     #     ###   ####    ###    ## # 
+%      ####   #   #    #        #   #     #   #  #  ## 
+%      # #    #####    #     ####   #     #####  #   # 
+%      #  #   #        #    #   #   #  #  #      #  ## 
+%      #   #   ###    ###    ####    ##    ###    ## # 
+%
+% ==================================================================================================
 
 \section{Related work}
-
-\subsection{What can be monad}
 \label{sec:flat-related-monads}
+
+Most of the related work has been already discussed in Chapter~\ref{ch:pathways} or throughout
+the chapter. In this section, we look in more details at the relationship between coeffects,
+comonads and monads.
+
+\subsection{When coeffect is not a monad}
+Coeffect systems differ from effect systems in three important ways:
+
+\begin{itemize}
+\item Semantically, coeffects capture very different notions of computation. As demonstrated in 
+  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
+  computation, many of which cannot be captured by a monad (\eg~liveness or data-flow).
+
+\item Syntactically, coeffect calculi use a richer algebraic structure with point-wise composition,
+  sequential composition and context merging $(\cpar, \cseq, \czip)$ while most effect systems
+  only use a single operation for sequential composition (monadic bind).
+
+\item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In 
+  coeffect systems, the context requirements of the body can be split between declaration-site
+  and call-site, while monadic effect systems delay all effects.
+\end{itemize}
+
+\noindent
+Despite the differences, our implicit parameters example can be also represented by a monad.
+Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
+we use the $\cup$ operation for all three operations of the coeffect algebra. However, the last
+point requires us to extend monadic lambda abstraction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{When coeffect is a monad}
+As discussed in \ref{sec:applications-flat-impl} ...
+
+Product comonad
+
+$C_{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma)$
+
+Reader monad
+
+$M_{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau$
+
+\begin{equation*}
+\begin{array}{l}
+ \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 = \\
+ (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
+ \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) \\
+ \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \\
+\end{array}
+\end{equation*}
 
@@
 the reader monad needs an operation that eagerly performs some effects of a function:
-$(\tau_1 \rightarrow \mtyp{r \oplus s}{\tau_2}) \rightarrow \mtyp{r}{(\tau_1 \rightarrow \mtyp{s}{\tau_2})}$.
+
+\begin{equation*}
+(\tau_1 \rightarrow \mtyp{\cclrd{r} \cseq \cclrd{s}}{\tau_2}) \rightarrow 
+  \mtyp{\cclrd{r}}{(\tau_1 \rightarrow \mtyp{\cclrd{s}}{\tau_2})}
+\end{equation*}
+
 To obtain a pure lambda abstraction
@@
 
+\begin{equation*}
+\begin{array}{rcl}
+ \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
+ \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
+\end{array}
+\end{equation*}
 % ==================================================================================================

### 2014-08-22 text/Chapter05.tex
Calculus and semantics work (Chapter 5)

@@
+%!TEX root = ../main.tex
+
 \chapter{Structural coeffect language} 
 \label{ch:structural}
+
+As already discussed, the aim of this thesis is to identify abstractions for context-aware
+programming languages. We attempt to find abstractions that are general enough to capture a wide
+range of useful programming language features, but specific enough to let us identify
+interesting properties of the languages.
+
+In Chapter~\ref{ch:applications}, we identified two notions of context. We generalized the 
+class of flat calculi that capture whole-context properties in Chapter~\ref{ch:flat}. 
+In this chapter, we turn our attention to \emph{structural} coeffect calculi that capture
+per-variable properties.
+
+The flat coeffect system captures interesting use-cases (implicit parameters, liveness and
+data-flow), but provides relatively weak properties. We can define its categorical semantics,
+but the equational theory proofs had numerous additional requirements. For this reason, it
+is worthwhile to consider structural systems in a separate chapter. We will see that 
+structural coeffects have a number of desirable properties that hold for all instances of the
+calculus.
+
+% ==================================================================================================
+%                                                                                      
+%     ###           #                       #                 #       #                 
+%      #            #                       #                 #                         
+%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
+%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
+%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
+%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
+%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
+%                                                                                     
+% ==================================================================================================
+
+\section{Introduction}
+\label{sec:flat-intro}
+
+Two examples of flat systems from the previous chapter were liveness and data-flow. As discussed
+in \ref{sec:applications-structural}, these are interesting for theoretical reasons. However, tracking
+liveness of the whole context is not practically useful. Structural versions of liveness and 
+data-flow let us track more fine-grained properties. Moreover, the equational theory of flat 
+coeffect calculus did not reveal many useful properties for flat liveness and data-flow. 
+As we show in this chapter, this is not the case with structural versions.
+
+In this chapter, we focus on three example applications. We look at structural liveness and
+data-flow and we also consider calculus for bounded reuse, which checks how many times a variable
+is accessed and generalizes linear logics (that restrict variables to be used exactly once).
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Contributions}
+
+Compared to the previous chapter, the structural coeffect calculi we consider are more 
+homogeneous and so finding the common pattern is in some ways easier. However, the systems
+are somewhat more complicated as they need to keep annotations attached to individual variables. 
+The contributions of this chapter are as follows:
+
+\begin{itemize}
+\item We present a \emph{structural coeffect calculus} with a type system that is parameterized by a 
+  \emph{structural coeffect algebra} and can be instantiated to obtain all of the three examples
+  discussed (Section~\ref{sec:struct-calculus}).
+
+\item We give the equational theory of the calculus. We prove the type-preservation property 
+  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}).
+
+\item We show how to extend indexed comonads introduced in the previous section to 
+  \emph{structural indexed comonads} and use them to give the semantics of structural 
+  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
+  categorical semantics provides a motivation for the design of the calculus.
+\end{itemize}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Related work}
+
+In the previous chapter, we discussed the correspondence between coeffects and effects (and 
+between comonads and monads). As noted earlier, the $\lambda$-calculus is assymetric in that
+an expression has multiple inputs (variables in the context), but just a single result 
+(the resulting value) and so monads and effects have no notion directly corresponding to 
+structural coeffect systems.
+
+The work in this chapter is more closely related to sub-structural type systems \cite{substruct-attpl-intro}.
+While sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
+\emph{exchange} rules, our systems keep them, but use them to manipulate both the context
+and its annotations.
+
+Our work follows the language semantics style in that we provide a structural semantics to
+the terms of ordinary $\lambda$-calculus. The most closely related work has been done in the
+meta-langauge style, which extends the terms and types with constructs working with the context
+explicitly. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
+may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
+$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
+annotations as stand-alone types.
+
+Structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
+For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
+respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
+This means that the typing judgements have the same structure as those of the flat coeffect
+calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify the two 
+systems and compose tracking of flat and structural properties.
+
+% =================================================================================================
+%                                                            
+%       ###           ##                   ##                 
+%      #   #           #                    #                 
+%      #       ###     #     ###   #   #    #    #   #   ###  
+%      #          #    #    #   #  #   #    #    #   #  #     
+%      #       ####    #    #      #   #    #    #   #   ###  
+%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
+%       ###    ####   ###    ###    ## #   ###    ## #  ####  
+%                                                           
+% =================================================================================================
+
+\section{Structural coeffect calculus}
+\label{sec:struct-calculus}
+
+In the structural coeffect calculus, a vector of variables in the free-variable context is 
+annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
+for different coeffect calculi and their properties are captured by the \emph{structural 
+coeffect scalar} definition below. The scalar annotations can be integers (how many past values we 
+need) or annotations specifying whether a variable is live or not.
+
+Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in 
+the previous chapter). Functions always have exactly one input variable and so they are annotated
+with a coeffect scalar. Thus the expressions and types of structural coeffect calculi are the same
+as in the previous chapter (except that annotation on function type is now a structural coeffect 
+scalar):
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\end{array}
+\end{equation*}
+%
+In the previous chapter, the free variable context $\Gamma$ has been treated as a set. In the 
+structural coeffect calculus, the order of variables matters. Thus we treat free variable context
+as a vector with a uniqueness condition. We also write $\xlift{\textnormal{--}}$ for the length
+of the vector:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \Gamma &\narrow{=}& \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
+ \xlift{\Gamma} &\narrow{=}& n
+\end{array}
+\end{equation*}
+%
+For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
+typing judgements, but the free variable context should be understood as a vector.
+Furthermore, the usual notation $\Gamma_1, \Gamma_2$ stands for the tensor product.
+Given $\Gamma_1 = \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
+$\Gamma_2 = \langle x_{n+1}\!:\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
+$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
+
+The free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
+we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
+over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
+and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
+is written as $\slift{\aclrd{\textbf{r}}}$. The structure for working with vectors
+of coeffects is provided by the \emph{structural coeffect algebra} definition below.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffect algebra}
+
+The structural coeffect scalar structure is similar to \emph{flat coeffect algebra} with the 
+exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
+modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
+point-wise composition, as well as a relation $\cleq$ that defines sub-coeffecting.
+
+\begin{definition}
+A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
+$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
+$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
+$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
+%
+\begin{equation*}
+\begin{array}{ccr}
+r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
+\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
+\textnormal{(monoid)}   
+\\
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
+\czero \;\cpar\; r = r = r \;\cpar\; \czero &
+\textnormal{(monoid)}   
+\\
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
+t\; \cleq\; t &
+\textnormal{(pre-order)}   
+\end{array}
+\end{equation*}
+%
+In addition, the following distributivity axioms hold:
+\begin{align*}
+\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+\end{definition}
+
+\noindent
+In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
+available from the declaration-site and the call-site or, in the syntactic reading, to split the 
+context requirements.
+
+In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
+coeffects becomes just vector a concatenation or splitting, respectively, which is provided
+by the tensor product. The operations on vectors are indexed by integers representing the lengths
+of the vectors. The additional structure required by the type system for structural coeffect 
+calculi is given by the following definition.
+
+\begin{definition}
+A \emph{\aclrd{structural coeffect algebra}} is formed by a structural coeffect scalar 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:
+
+\begin{itemize}
+\item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
+  structural coeffect scalars indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.
+
+\item An operation that constructs a vector from scalars indexed by the vector length
+  $\alift{\textnormal{--}}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
+  and an operation that returns the vector length such that 
+  $\slift{\aclrd{\textbf{r}}} = \sclrd{n}$ for $\aclrd{\textbf{r}} : \C^{\sclrd{n}}$
+
+\item A point-wise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
+  such that\\ $\cclrd{t} \,\aseq\, \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} = 
+    \alift{\cclrd{t} \,\cseq\,\cclrd{r_1}, \ldots, \cclrd{t} \,\cseq\,\cclrd{r_n}}$.
+
+\item An indexed tensor product $\atimes_{\sclrd{n},\sclrd{m}} : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n}+\sclrd{m}}$
+  that is used in both directions -- for vector concatenation and for splitting -- 
+  which is defined as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} 
+     \atimes_{\sclrd{n}, \sclrd{m}} \alift{\cclrd{s_1}, \ldots, \cclrd{s_m}}
+   = \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}, \cclrd{s_1}, \ldots, \cclrd{s_m}}$
+\end{itemize}
+\end{definition}
+
+\noindent
+The fact that the tensor product $\atimes_{\sclrd{n},\sclrd{m}}$ is indexed by the lengths
+of the two vectors means that we can use it unambiguously for both concatenation of vectors
+and for splitting of vectors, provided that the lengths of the resulting vectors are known.
+In the following text, we usually omit the indices and write just 
+$\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}}$, because the lengths of the coeffect vectors 
+can be determined from the lengths of the matching free variable context vectors.
+
+More generally, we could see the the coeffect annotations as a \emph{container} 
+\cite{types-containers} that supports certain operations. This approach is used in 
+Chapter~\ref{ch:unified} as a way of unifying the flat and structural systems.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffect types}
+\label{sec:struct-calculus-types}
+
+The type system for structural coeffect calculus is similar to sub-structural type systems
+in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
+weakening, exchange or contraction -- this is done by checking the types of sub-expressions
+in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
+allows weakening, exchange and contraction, but using explicit \emph{structural} rules that
+perform corresponding transformation on the coeffect annoation.
+
+\paragraph{Syntax-driven rules.}
+The variable access rule (\emph{var}) annotates the corresponding variable as being used
+using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
+\emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
+Constants (\emph{const}) are type checked in an empty variable context, which is annotated with
+an empty vector of coeffect annotations.
+
+The (\emph{abs}) rule assumes that the free variable context of the body can be split into
+a potentially empty \emph{declaration site} and a singleton context containing the bound
+variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
+the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
+non-determinism present in flat coeffect systems.
+
+In (\emph{app}), the sub-expressions $e_1$ and $e_2$ use free variable contexts $\Gamma_1, \Gamma_2$
+with coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$, respectively. The function
+value is annotated with a coeffect scalar $\cclrd{t}$.
+The coeffect annotation of the composed expression is obtained by combining the annotations
+associated with variables in $\Gamma_1$ and $\Gamma_2$. Variables in $\Gamma_1$ are only
+used to obtain the function value, resulting in coeffects $\aclrd{\textbf{r}}$. The 
+variables in $\Gamma_2$ are used to obtain the argument value, which is then sequentially
+composed with the function, resulting in $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$.
+
+\paragraph{Structural rules.}
+The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
+They allow different transformation of the free variable context. We include sub-coeffecting
+(\emph{sub}) as one of the rules, allowing sub-coeffecting on coeffect scalars belonging
+to individual variables. The remaining rules capture \emph{weakening}, \emph{exchange} and
+\emph{contraction} known from sub-structural systems.
+
+The (\emph{weak}) allows adding a variable to the context, extending the coeffect
+vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
+in the context, performing the same reordering on the coeffect vector. Finally recall that 
+variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
+allows re-using a variable as we can type check sub-expressions using two separate varaibles and 
+then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
+is the only place in the structural coeffect system where context requiremens are combined,
+or semantically, where the same context is shared.
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Syntax-driven typing rules:}
+\begin{equation*}
+\tyrule{var}
+  { }
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c \!:\! \tau \in \Delta}
+  {\coctx{()}{\alift{}} \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Structural rules for context manipulation:}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+(\cclrd{s'} \cleq \cclrd{s})
+\end{equation*}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{ \czero }} \vdash e : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+\begin{array}{l}
+ \xlift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{2.5em} 
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~
+\begin{array}{l}
+ \xlift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\caption{Type system for the structural coeffect calculus}
+\label{fig:struct-types}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Understanding structural coeffects}
+
+The type system for structural coeffects appears more complicated when compared to the flat
+version, but it is in many ways simpler -- it removes the ambiguity arising from the use
+of $\czip$ in lambda abstraction and, as discussed in Section~\ref{sec:struct-syntax}, has
+a cleaner equational theory.
+
+\paragraph{Flat and structural context.} In flat systems, lambda abstraction splits context
+requirements using $\czip$ and application combines them using $\cpar$. In the structural
+version, both of these are replaced with $\atimes$. The $\czip$ operation is not needed, but
+$\cpar$ is still used in the (\emph{contr}) rule.
+
+This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
+as over- and under-approximations of $\atimes$. This is demonstrated by the (\emph{approximation})
+requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
+$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
+combines available context, potentially discarding parts of it (under-approximation), while flat
+application splits available context, potentially duplicating parts of it (over-approxi\-mation)\footnote{
+Because of this duality, earlier version of coeffects in \cite{coeffects-icalp13}
+used $\wedge$ and $\vee$.}.
+
+Second, the operator $\cpar$ is used when the semantics passes the same context to multiple
+sub-expressions. In flat systems, this happens in (\emph{app}) and (\emph{pair}), because the 
+sub-expressions may share variables. In structural systems, this is separated into an 
+explicit contraction rule.
+
+\paragraph{Let binding.}
+The other aspect that makes structural systems simpler is that they remove the need for 
+separate let binding. As discussed in Section~\ref{sec:flat-calculus-let}, flat calculi include
+let binding that gives a \emph{more precise} typing than combination of abstraction and 
+application. This is not the case for structural coeffects.
+
+\begin{remark}[Let binding]
+In a structural coeffect calculus, the typing of $(\lambda x.e_2)~e_1$ is equivalent
+to the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$.
+\end{remark}
+\begin{proof}
+Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
+in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
+%
+\begin{equation*}
+\inference
+  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
+    \inference
+      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
+      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
+  { \inference
+      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
+      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
+\end{equation*}
+
+\noindent
+The assumptions and conclusions match those of the (\emph{let}) rule. 
+\end{proof}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Examples of structural coeffects}
+
+The structural coeffect calculus can be instantiated to obtain the structural coeffect calculi
+presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
+structural liveness provide a more precise tracking of properties that can be tracked using 
+flat systems. Formally, a flat coeffect algebra can be turned into a structural coeffect algebra
+(by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
+example, it is not clear why one would associate implicit parameters with individual variables.
+
+On the other hand, some of the structural systems do not have a flat equivalent, typically 
+because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
+algebra. This is the case, for example, for the bounded variable use.
+
+\begin{example}[Structural liveness]
+The structural coeffect algebra for liveness is formed by
+$(L, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $L=\{ \ident{L}, \ident{D} \}$ is 
+the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
+with a join $\sqcup$ and meet $\sqcap$. 
+\end{example}
+
+\begin{example}[Structural data-flow]
+In data-flow, context is annotated with natural numbers and the flat coeffect algebra is formed 
+by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
+\end{example}
+
+\noindent
+For the two examples that have both flat and structural version, obtaining the strucutral coeffect
+algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
+required by a structural coeffect algebra are the same as those required by the flat version
+and so the above definitions are both valid. Similar construction can be used for the
+\emph{optimized data-flow} example from Section~\ref{sec:flat-calculus-examples}.
+
+It is important to note that this gives us a systems with \emph{different} properties. The 
+information are now tracked per-variable rather than for the enitre context. For data-flow,
+we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
+for a point-wise extension of the $+$ operator, such that
+$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} + k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
+%
+\begin{equation*}
+\tyrule{prev}
+  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
+  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
+\end{equation*}
+%
+The rule appears similar to the flat one, but there is an important difference. Because of the
+structural nature of the type system, it only increments the required number of values for 
+variables that are used in the expression $e$. Annotations of other variables can be left 
+unchanged.
+
+Before looking at the semantics and equational properties of structural coeffect systems, 
+we consider bounded variable use, which is an example of structural system that does not
+have a flat counterpart.
+
+\begin{example}[Bounded variable reuse]
+The structural coeffect algebra for tracking bounded variable use is given by
+$(\mathbb{N}, \ast, +, 1, 0, \leq)$
+\end{example}
+
+\noindent
+Similarly to the structural calculus for data-flow, the calculus for bounded variable reuse
+annotates each variable with an integer. However, the integer denotes how many times is the variable 
+\emph{accessed} rather than how many \emph{past values} are needed. The resulting type system is 
+the one shown in Figure~\ref{fig:applications-struct-live} in Chapter~\ref{ch:applications}.
+
+% ==================================================================================================
+%                                              
+%      #####  #                                 
+%        #    #                                 
+%        #    # ##    ###    ###   # ##   #   # 
+%        #    ##  #  #   #  #   #  ##  #  #   # 
+%        #    #   #  #####  #   #  #      #  ## 
+%        #    #   #  #      #   #  #       ## # 
+%        #    #   #   ###    ###   #          # 
+%                                         #   # 
+%                                          ###                                                                         
+% ==================================================================================================
+
+\section{Categorical motivation}
+\label{sec:struct-semantics}
+
+When introducing structural coeffect systems in Section~\ref{sec:applications-structural}, we 
+included a concrete semantics of structural liveness and bounded variable reuse. In this section,
+we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
+extension of \emph{indexed comonad} structure. As in the previous chapter, the main aim of this 
+section is to motivate and explain the design of the structural coeffect calculus shown in 
+Section~\ref{sec:struct-calculus}. The semantics highlights the similarities and differences 
+between the two systems.
+
+Most of the differences between flat and structural systems arise from the fact that contexts
+in structural coeffect systems are treated as \emph{vectors} rather than sets modelled using 
+categorical products, so we start by discussing our treatment of vectors.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of vectors}
+
+In the flat coeffect calculus, the context is interpreted as a product and so a typing judgement
+$\coctx{x_1\!:\tau_1, \ldots, x_n:\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a morphism 
+$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, we can 
+freely transform the value contained in the context modelled using an indexed comonad
+$\ctyp{\cclrd{r}}$. For example, the function $\ident{map}_{\cclrd{r}}~\pi_i$ transforms
+a context $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a value 
+$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
+
+The ability to freely transform the variable structure is not desirable in the model of 
+structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
+the coeffect annotations matches the structure of variables. To achieve this, we model vectors
+using a structure distinct from ordinary products which we denote $-\xtimes-$. 
+For example, the judgement $\coctx{x_1\!:\tau_1, \ldots, x_n:\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
+is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
+
+The operator is a bifunctor, but it is \emph{not} a product in the categorical sense. In particular, 
+there is no way to turn $\tau_1 \xtimes \ldots \xtimes \tau_n$ into $\tau_i$ (the structure does 
+not have projections) and so there is also no way of turning
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$ into
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{\tau_i}$, which would break the correspondence 
+between coeffect annotations and variable structure.
+
+The structure created using $-\xtimes-$ can be manipulated only using operations provided by
+the \emph{strucutral indexed comonad}, which operate over variable contexts contained in an 
+indexed comonad $\ctyp{\aclrd{\textbf{r}}}$.
+
+vectors
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural indexed comonads}
+
+Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
+\emph{indexed comonad} over a category $\catc$ is a triple 
+$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where....
+
+\begin{definition}
+Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+an \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
+equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
+and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
+%
+\begin{compactitem}
+\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
+  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
+\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
+  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
+\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
+  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\end{compactitem}
+\end{definition}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Examples}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of structural caluculus}
+
+\newcommand{\ctx}{\textit{ctx}}
+
+\begin{figure*}[t]
+
+\begin{equation*}
+\begin{array}{ll}
+ \sem{\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }~\ctx =
+   \ident{counit}_{\cunit}~\ctx & (\emph{var})
+\\[0.5em]
+\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
+  \delta~(c_i) & (\emph{const})
+\\[0.5em]
+\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
+  \hspace{2.5em}
+  \begin{array}{l}  
+  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}~\ctx\\[-0.25em]
+  \kvd{in}~\sem{\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
+      (\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~\sem{\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\ctx_2)
+  \end{array}
+\\[1.5em]
+\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
+\hspace{2em}\sem{\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  ~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v))
+\\[0.5em]
+\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
+  \hspace{2em}
+  \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx~~\kvd{in}~~
+  \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}~\ctx_1
+\\[0.5em]
+\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{sub})\\
+  \hspace{2em}
+  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
+  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
+\\[0.5em]
+\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
+        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\
+  \hspace{2em}
+  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
+  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
+\\[0.5em]
+\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
+        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\
+  \hspace{2em}
+  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
+  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)
+\end{array}
+\end{equation*}
+\\[0.5em]
+{\small Assuming the following auxiliary definitions:}
+\begin{equation*}
+\hspace{-2em}
+\begin{array}{l}
+ \ident{swap}_{\cclrd{t},\cclrd{s}} ~:~ 
+  \ctyp{\alift{\cclrd{t}, \cclrd{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\alift{\cclrd{s}, \cclrd{t}}}{(\beta \xtimes \alpha)} \\[-0.25em]
+\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx = \\
+  \hspace{2em} \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}}~\ctx \\[-0.25em]
+  \hspace{2em} \kvd{in}~\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)
+ \\[0.5em]
+\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} ~:~
+  (\ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{s'}}}{\beta'}) \rightarrow
+    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta \xtimes \gamma)} \rightarrow
+    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s'}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta' \xtimes \gamma)}\\[-0.25em]
+\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f~\ctx =\\
+  \hspace{2em} \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx \\[-0.25em]
+  \hspace{2em} \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}~\ctx' \\[-0.25em] 
+  \hspace{2em} \kvd{in}~\ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~
+        (\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
+\end{array}
+\end{equation*}
+
+\caption{Categorical semantics of the structural coeffect calculus}
+\label{fig:struct-semantics}
+\end{figure*}
+
+~
+\newpage
+
+- variable access gets single-variable context and so it just projects the value
+- application does not duplicate before splitting - it splits directly (to get two sub-contexts)
+- abs is pretty much the same
+- weakening just drops a value with ``ign'' annotation - we write this using pseuod-pattern matching, but it could be special operation in the category
+- sub, exchange and contraction are quite similar - they extract and then merge
+~ 
+\newpage
 
@@
 \newcommand{\csvop}[0]{ \vee }
+\newcommand{\cmerge}[0]{ \cpar }
+\newcommand{\csplit}[0]{ \czip }
 
@@
 
-%---------------------------------------------------------------------------------------------------
+\newcommand{\cobind}[1]{{#1}^\dagger}
+\newcommand{\obj}[0]{obj}
+\newcommand{\ccat}[0]{C}
 
-The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
-we need to track context-dependence in a more fine-grained way. To track neededness or security, 
-we need to associate information with individual \emph{variables} of the context.
+\begin{figure*}[t]
+\newcommand{\uu}[1]{\overline{#1}}
+\begin{equation*}
+\sem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
+  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
+\end{equation*}
+\begin{equation*}
+\begin{array}{rcll}
+\sem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
+    \epsilon_0                                                                   &\quad$\ruless{var}$
+\\[0.32em]
+\sem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
+    \Lambda (\sem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
+\\[0.32em]
+\sem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
+    (\lambda (\gamma_1, \gamma_2) \rightarrow
+    \sem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
+    (\cobind{\sem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
+\\[0.32em]
+\sem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
+    \sem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \sem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
+\end{array}
+\end{equation*}
 
-%===================================================================================================
+\begin{equation*}
+\begin{array}{rclll}
+\sem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
+&=& \cmerge_{r, s} \circ (\sem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
+\textnormal{\rulessc{nest}}
+\\[0.8em]
+\sem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
+&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
+\textnormal{\rulessc{exch}}
+\\[0.8em]
+\sem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
+&=&\ident{fst} \circ \csplit_{r, 1}&&
+\textnormal{\rulessc{empty}}
+\\[0.8em]
+\sem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
+&=&\ident{fst} \circ \csplit_{r, 0}&&
+\textnormal{\rulessc{weak}}
+\\[0.8em]
+\sem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
+    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
+&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
+\textnormal{\rulessc{contr}}
+\\[0.8em]
+\sem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
+    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
+&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
+        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
+\textnormal{\rulessc{assoc}}
+\\[0.8em]
+\sem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
+&=&
+\iota_{r, s}
+&&
+\textnormal{\rulessc{sub}}
+\end{array}
+\end{equation*}
 
-\section{Introduction}
+\caption{Categorical semantics for~\clstr}
+\label{fig:structural-semantics}
+\vspace{-0.5em}
+\end{figure*}
 
@@
 
-\subsection{Motivation: Tracking array accesses}
-\label{sec:structural-motivation}
-%\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
-%\newcommand{\dnat}{\ident{D}_\ident{nat}}
+\section{Semantics of structural coeffects}
+\label{sec:structural-semantics}
 
-Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
-functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
-respectively, but we use richer tag structure.
+The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
+\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
+of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
+is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
 
-As an example, consider a language that allows us to get a value of a variable (representing
-some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
-To track information about individual variables, we use a product-like operation $\csprd$ on tags 
-to mirrors the product structure of variables. For example:
+As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
+system is not a simple finite product, but instead a binary tree. To model this, we do not use
+ordinary products in the domain of the semantic function, but instead use a special 
+constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
+the tag structure.
+
+\subsection{Structural tagged comonads}
+\label{sec:structural-tagged-comonads}
+
+To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
+Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
+$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
+$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
+context-requirements.
+
+However, functions that do not exist in context have only a single input variable (with a 
+single corresponding tag). To model complex variable contexts, we need two additional
+operations that allow manipulation with the variable context. Similarly to the model of \clflt,
+we also require operations that model duplication and sub-coeffecting:
+
+\begin{definition}[Structural tagged comonad]
+Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
+tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
+$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
+$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
+\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
+\end{array}
+\end{equation*}
+%
+% Delta is still defineable
+% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
+%
+And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
+%
+%  TODO: laws?
+%
+\end{definition}
+
+The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
+and it can still be used to define a family of mappings that represents \emph{duplicating}
+of variables while splitting the additional coeffect tags:
 %
@@
 \begin{array}{l}
-\ctyp{5 \csprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
-  \vdash
-    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
+\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
+\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
 \end{array}
@@
 %
-The coeffect tag $5 \csprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
-that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
-for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
-a single variable:
+The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
+but with two differences. Firstly, it combines tags using $\csprd$ instead of $\vee$, 
+which corresponds to the fact that the variable context now consists of two parts (a tree node).
+Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
+(instead of $A \times B$ as previously).
+
+To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
+of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
+variables. The operation was not needed for \clflt, because there \emph{splitting} could
+be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
+the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
+with $r \csprd s$ that carries $A \ctimes B$.
+
+Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
+Before looking at them, we finish our discussion of categorical semantics.
+
+\paragraph{Categorical notes.}
+The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
+Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
+variables in the context. Given a context modelled as $T^{r \csprd s}(A \ctimes B)$,
+we can lift a morphism $f$ to $\hat{T}^{r \csprd s}(f)$, but we cannot manipulate the
+variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
+
+This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
+could apply $\cdupl_{r, s}$ on the input (if the tag $r \csprd s$ coincided
+with tag $r \csvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
+
+This restriction is intentional -- at the semantic level, it prevents manipulations
+with the context that would break the correspondence between tag structure and the
+product structure.
+
+\subsection{Categorical semantics}
+The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
+It uses the \emph{structural tagged comonad} structure introduced in the previous section,
+together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
+operations:
+\begin{equation*}
+\begin{array}{rcl}
+\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
+\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
+f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
+\end{array}
+\end{equation*}
+
+When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
+notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
+the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
+variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
+$\csprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
+type signature in $\cmerge_{r, s}$).
+
+The rule \ruless{app} now uses the operation $\csplit_{s, (r \csvop t)}$ instead of $\cdupl_{s, (r \csvop t)}$,
+which means that it splits the context instead of duplicating it. This makes the system more 
+structural -- the expressions use disjunctive parts of the context -- and also explains why the
+composed coeffect tag is $s \csprd (r \csvop t)$.
+
+The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
+of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
+of $\sem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
+$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
+takes a context, converts it to a new context which is compatible with the original expression $e$.
+The context manipulation rules work as follows:
+
+\begin{itemize}
+\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
+  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
+  Finally, they re-construct a single context using $\cmerge_{r, s}$.
+
+\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
+  and discard one part (containing either an unused variable or an empty context).
+
+\item If we interpreted \rulessc{contr} by applying functor $T^{r\csvop s}$ to a function that duplicates
+  a variable, the resulting context would be $\ctyp{r \csvop s}(x:\tau, x:\tau)$, which would break the
+  correspondence between coeffect tag and context variable structure. However, that interpretation would be
+  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
+  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
+  a tag $r \csvop s$ into $r \csprd s$.
+
+\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
+  manipulates it (using \ident{assoc}) and then re-constructs it.
+
+\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
+  variable using the primitive natural transformation $\iota_{r, s}$.
+\end{itemize}
+
+\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
+attaching the context separately to individual variables. This an expression
+$\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
+$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
+approach largely complicates the definition of application (where tag of all 
+variables in a context is affected). Moreover, it makes it impossible to express
+\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.
+
+\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
+transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
+operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
+sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).
+
+This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
+typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
+the following natural transformation has to be provided:
 %
 \begin{equation*}
-\begin{array}{l}
-\ctyp{5 \csvop 10}{(\ident{c}:\dnat)}
-  \vdash
-    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
-\end{array}
+\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
 \end{equation*}
 %
-In this example, the operation $\csvop$ would be the \emph{max} function and so $5 \csvop 10 = 10$.
-Before looking at the formal definition, consider the typing of let bindings:
+This variant of the system could be used to define a system that ensures that all provided 
+context is used and is not over-approximated. This difference is similar to the difference between
+affine type systems (where a variable can be used at most once) and linear type systems 
+(where a variable has to be used exactly once).
+
+\newpage
+~
+\newpage
+~
+
+% ==================================================================================================
 %
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
-\pastval{a}{15}+\pastval{c}{10}
-\end{array}
-\end{equation*}
+%                               ##      
+%      ####                     ##      
+%        ##                     ##      
+%        ##   ##   ##  ## ###   ##  ##  
+%        ##   ##   ##  ###  ##  ## ##   
+%    ##  ##   ##   ##  ##   ##  ####    
+%    ##  ##   ##  ###  ##   ##  ## ##   
+%     ####     ### ##  ##   ##  ##  ##  
 %                                      
-The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
-is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
-The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.
-
-The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
-context is $\ctyp{0 \csprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
-of $\csvop$ and it denotes empty coeffect. The typing context of the
-body is $\ctyp{15 \csprd 10}{(\ident{a}, \ident{c})}$.
-
-To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
-of the context in which \ident{c} was defined using the $\csvop$ operation. This is then combined 
-with the remaining tags from the body yielding the overall context:
-$\ctyp{15 \csprd (10 \csvop (0 \csprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
-normalization mechanism (described later), this can be further reduced to
-$\ctyp{(15 \csvop 10) \csprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
-we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
-\ident{b}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffect tags}
-In the previous section, the $\csvop$ operation behaves similarly to the flat $\vee$ operation.
-However, the type system does not require some of the semilattice properties, because some uses
-are replaced with the $\csprd$ operation. We do not require any properties about the $\csprd$ operation.
-For example, in the previous example cannot be commutative (since a tag $15 \csprd 10$ has different 
-meaning than $10 \csprd 15$). However, we relate the operations using distributivity laws to allow 
-normalization that was hinted above.
-
-\begin{definition}
-A \emph{structural coeffect tag structure} $(S, \csprd, \csvop, 0, 1)$ is a tuple where 
-$(S, \csvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
-by a binary operation $\csprd$ and element $1\in S$ such that
-for all $r,s,t\in S$, the following equalities hold: 
-%
-\begin{subequations}
-\begin{align}
-\tag{associativity}
-r \csvop (s \csvop t) &= (r \csvop s) \csvop t\\
-\tag{commutativity}
-r \csvop s &= s \csvop r\\
-\tag{lower bound}
-r \csvop 0 &= r\\
-\tag{distributivity}
-r \csvop (s \csprd t) &= (r \csvop s) \csprd (r \csvop t)\\
-\tag{upper bound}
-1 \csvop r &= 1
-\end{align}
-\end{subequations}
-%
-\end{definition}
-%
-% 'upper bound' (not sure what to call it) says that empty context cannot carry information
-% looks like this is the upper bound of the bounded lattice ???
-%
-% commutativity is not strictly required by any law or type preservation, but
-% it simplifies the presentation and looks reasonable
-%
-The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
-always accessed using standard variable access). The tag $1$ is used to annotate empty 
-variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
-to carry an annotation. We explain exactly how this works when we introduce the type system.
-The fact that $1$ is the upper bound means that combining it with other coeffect annotations
-does not affect it and so empty contexts cannot carry any information.
-
-The structure generalizes the \emph{flat coeffect tag structure} introduced in 
-Section~\ref{sec:flat-tag-structure}, but it additionally requires a special element $1$ 
-representing the upper bound. Given a flat coeffect structure, we can construct a structural
-coeffect structure (but not the other way round). For certain structures, the element $1$
-may be already present, but in general, it can be added as a new element.
-This construction will be important in Section~\ref{sec:generalization},
-where we show that \clstr~calculus generalizes~\clflt.
-
-\begin{lemma}
-\label{thm:structural-generalizaton}
-A flat coeffect tag structure $(S, \vee, 0)$ implies a structural coeffect tag structure.
-\end{lemma}
-\begin{proof}
-Take $1 \notin S$, then $(S \cup \{1\}, \vee, \vee, 0, 1)$ is a structural coeffect tag 
-structure. From the properties of flat coeffect tag structure, we get that the $\vee$ operation 
-is associative, commutative and $0$ is the unit with respect to $\vee$.
-
-To prove the distributivity, we need to show $r \vee (s \vee t) = (r \vee s) \vee (r \vee t)$.
-This easily follows from commutativity, associativity and idempotence of the flat coeffect tag 
-structure.
-\end{proof}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffect type system}
-\label{sec:structural-types}
-
-The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
-The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
-a \emph{structural coeffect tag structure} $(S,\csprd, \csvop, 0)$.
-
-The system differs from \clflt~in a significant aspect. It contains explicit 
-structural rules that manipulate with the context. Such rules allow reordering, duplicating
-and other manipulations with variable context. Such rules are known from affine or linear
-type systems where they are removed to obtain more restrictive system. In our system, the
-rules are present, but they manipulate the variable structure $\Gamma$ as well as the 
-associated tag structure $r$.
-
-As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
-simple set. Instead, we use the following tree-like structure (which is more similar to 
-bunched types than to linear or affine systems):
-%
-\begin{equation*}
-\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
-\end{equation*}
-%
-The syntax $()$ represents an empty context, so the structure defines a binary trees where
-leaves are either variables or empty. Contexts such as $\ctyp{1\csprd(r \csprd 1)}{((), (x, ()))}$ 
-contain unnecessary number of empty contexts $()$. However, we need to construct them
-temporarily, because certain rules require splitting a context and, by our 
-definition, the context $(x:\tau)$ is not splitable.
-
-The typing rules of the system are shown in Figure~\ref{fig:structural-coeffect-typing}.
-Many of the structural rules are expressed in terms of a helper judgement
-$\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}}$.
-
-\paragraph{Standard rules.}
-Variable access (\rulest{var}) annotates the corresponding variable with an empty coeffect
-$0$. The \rulest{fun} rule assumes that the context of the body can be split into the
-variable of the function and other (potentially empty) context and it attaches the coeffect
-associated with the function variable to the resulting function type $\ctyp{s}{\tau_1} \rightarrow \tau_2$.
-
-The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
-expression, argument and the function type respectively. 
-The result of evaluating the argument in the context $t$ is passed to the function that
-requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
-the combination of coeffects $r \csvop t$. The variable context required to evaluate the 
-function value is independent and so it is annotated just with coeffects $s$.
-Finally, the \rulest{let} rule is derived from let binding and application, but we show
-it separately to aid the understanding.
-
-\paragraph{Structural rules.}
-The remaining rules are not syntax-directed. They are embedded using the \rulest{ctx}
-rule and expressed using a helper judgement $\ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}}$
-that says that the context on the left-hand side can be transformed to the context on the
-right-hand side. The transformation can be applied to any part of the context, which 
-is captured using the \rulest{nest} rule (it is sufficient to apply the transformation 
-on the left part of the context; the right part can be transformed using \rulest{exch}).
-
-The \rulest{empty} rule allows attaching empty context to any existing context. The rule 
-is needed to type-check lambda abstractions that do not capture any outer variable.
-The \rulest{weak} rule is similar, but it represents \emph{weakening} where an unused variable is 
-added. The associated coeffect tag does not associate context information with the variable.
-This is needed to type-check lambda abstraction that does not use the argument.
-
-The \rulest{contr} rule is a limited form of contraction. When a variable appears repeatedly,
-it can be reduced to a single occurrence. The rule is needed to satisfy the side condition
-of \rulest{fun} when the body uses the argument repeatedly. The two associativity rules
-together with \rulest{exch} provide ways to rearrange variables. Finally, \rulest{sub} represents
-sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.
-
-\ExecuteMetaData[rules/rules.tex]{scl-types}
+% ==================================================================================================
 
@@
 
-\begin{figure*}[t]
-\newcommand{\uu}[1]{\overline{#1}}
-\begin{equation*}
-\cosem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
-\end{equation*}
-\begin{equation*}
-\begin{array}{rcll}
-\cosem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
-    \epsilon_0                                                                   &\quad$\ruless{var}$
-\\[0.32em]
-\cosem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \Lambda (\cosem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
-\\[0.32em]
-\cosem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
-    (\lambda (\gamma_1, \gamma_2) \rightarrow
-    \cosem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
-    (\cobind{\cosem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
-\\[0.32em]
-\cosem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
-    \cosem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \cosem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
-\end{array}
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rclll}
-\cosem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
-&=& \cmerge_{r, s} \circ (\cosem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
-\textnormal{\rulessc{nest}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
-&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
-\textnormal{\rulessc{exch}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 1}&&
-\textnormal{\rulessc{empty}}
-\\[0.8em]
-\cosem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 0}&&
-\textnormal{\rulessc{weak}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
-    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
-&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
-\textnormal{\rulessc{contr}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
-    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
-&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
-        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
-\textnormal{\rulessc{assoc}}
-\\[0.8em]
-\cosem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
-&=&
-\iota_{r, s}
-&&
-\textnormal{\rulessc{sub}}
-\end{array}
-\end{equation*}
-
-\caption{Categorical semantics for~\clstr}
-\label{fig:structural-semantics}
-\vspace{-0.5em}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Semantics of structural coeffects}
-\label{sec:structural-semantics}
-
-The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
-\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
-of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
-is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
-
-As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
-system is not a simple finite product, but instead a binary tree. To model this, we do not use
-ordinary products in the domain of the semantic function, but instead use a special 
-constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
-the tag structure.
-
-\subsection{Structural tagged comonads}
-\label{sec:structural-tagged-comonads}
-
-To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
-Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
-$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
-$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
-context-requirements.
-
-However, functions that do not exist in context have only a single input variable (with a 
-single corresponding tag). To model complex variable contexts, we need two additional
-operations that allow manipulation with the variable context. Similarly to the model of \clflt,
-we also require operations that model duplication and sub-coeffecting:
-
-\begin{definition}[Structural tagged comonad]
-Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
-tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
-$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
-$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
-\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
-\end{array}
-\end{equation*}
-%
-% Delta is still defineable
-% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
-%
-And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
-%
-%  TODO: laws?
-%
-\end{definition}
-
-The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
-and it can still be used to define a family of mappings that represents \emph{duplicating}
-of variables while splitting the additional coeffect tags:
-%
-\begin{equation*}
-\begin{array}{l}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
-\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
-\end{array}
-\end{equation*}
-%
-The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
-but with two differences. Firstly, it combines tags using $\csprd$ instead of $\vee$, 
-which corresponds to the fact that the variable context now consists of two parts (a tree node).
-Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
-(instead of $A \times B$ as previously).
-
-To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
-of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
-variables. The operation was not needed for \clflt, because there \emph{splitting} could
-be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
-the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
-with $r \csprd s$ that carries $A \ctimes B$.
-
-Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
-Before looking at them, we finish our discussion of categorical semantics.
-
-\paragraph{Categorical notes.}
-The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
-Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
-variables in the context. Given a context modelled as $T^{r \csprd s}(A \ctimes B)$,
-we can lift a morphism $f$ to $\hat{T}^{r \csprd s}(f)$, but we cannot manipulate the
-variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
-
-This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
-could apply $\cdupl_{r, s}$ on the input (if the tag $r \csprd s$ coincided
-with tag $r \csvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
-
-This restriction is intentional -- at the semantic level, it prevents manipulations
-with the context that would break the correspondence between tag structure and the
-product structure.
-
-\subsection{Categorical semantics}
-The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
-It uses the \emph{structural tagged comonad} structure introduced in the previous section,
-together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
-operations:
-\begin{equation*}
-\begin{array}{rcl}
-\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
-\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
-f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
-\end{array}
-\end{equation*}
-
-When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
-notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
-the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
-variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
-$\csprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
-type signature in $\cmerge_{r, s}$).
-
-The rule \ruless{app} now uses the operation $\csplit_{s, (r \csvop t)}$ instead of $\cdupl_{s, (r \csvop t)}$,
-which means that it splits the context instead of duplicating it. This makes the system more 
-structural -- the expressions use disjunctive parts of the context -- and also explains why the
-composed coeffect tag is $s \csprd (r \csvop t)$.
-
-The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
-of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
-of $\cosem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
-$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
-takes a context, converts it to a new context which is compatible with the original expression $e$.
-The context manipulation rules work as follows:
-
-\begin{itemize}
-\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
-  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
-  Finally, they re-construct a single context using $\cmerge_{r, s}$.
-
-\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
-  and discard one part (containing either an unused variable or an empty context).
-
-\item If we interpreted \rulessc{contr} by applying functor $T^{r\csvop s}$ to a function that duplicates
-  a variable, the resulting context would be $\ctyp{r \csvop s}(x:\tau, x:\tau)$, which would break the
-  correspondence between coeffect tag and context variable structure. However, that interpretation would be
-  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
-  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
-  a tag $r \csvop s$ into $r \csprd s$.
-
-\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
-  manipulates it (using \ident{assoc}) and then re-constructs it.
-
-\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
-  variable using the primitive natural transformation $\iota_{r, s}$.
-\end{itemize}
-
-\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
-attaching the context separately to individual variables. This an expression
-$\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
-$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
-approach largely complicates the definition of application (where tag of all 
-variables in a context is affected). Moreover, it makes it impossible to express
-\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.
-
-\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
-transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
-operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
-sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).
-
-This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
-typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
-the following natural transformation has to be provided:
-%
-\begin{equation*}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
-\end{equation*}
-%
-This variant of the system could be used to define a system that ensures that all provided 
-context is used and is not over-approximated. This difference is similar to the difference between
-affine type systems (where a variable can be used at most once) and linear type systems 
-(where a variable has to be used exactly once).
-
-% --------------------------------------------------------------------------------------------------
-
 \section{Examples of structural coeffects}

### 2014-08-22 text/Chapter05.tex
Structural indexed comonads (Chapter 05)

@@
 structural coeffect calculus, the order of variables matters. Thus we treat free variable context
-as a vector with a uniqueness condition. We also write $\xlift{\textnormal{--}}$ for the length
+as a vector with a uniqueness condition. We also write $\xlift{-}$ for the length
 of the vector:
@@
 \item An operation that constructs a vector from scalars indexed by the vector length
-  $\alift{\textnormal{--}}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
+  $\alift{-}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
   and an operation that returns the vector length such that 
@@
 
-vectors
+In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
+We assume that the use of the operator can be freely re-associated. If an operation requires an 
+input in the form $(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
+we call it with $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ as an argument and assume that the 
+appropriate transformation is inserted.
 
@@
 
-\subsection{Structural indexed comonads}
-
-Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
-\emph{indexed comonad} over a category $\catc$ is a triple 
-$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where....
+\subsection{Indexed comonads, revisited}
+
+The semantics of structural coeffect calculus reuses the definition of \emph{indexed comonad} almost
+without a change. The additional structure that is required for context manipulation (merging and 
+splitting) is different and is provided by the \emph{structural indexed comonad} structure that we
+introduce in this section.
+
+Recall the definition from Section~\ref{sec:flat-semantics-idx}, which defines an indexed comonad 
+over a monoid $(\C, \cseq, \cunit)$ as a triple $(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, 
+\ident{cobind}_{\cclrd{r}, \cclrd{s}})$. The triple consists of a family of object mappings $\ctyp{\cclrd{r}{}}$,
+and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau \rightarrow \tau'$.
+
+In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau \rightarrow \tau'$
+representing function values (appearing in the language), but also of the form
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$, 
+modelling expressions in a context. To capture this, we need to generalize some of the indices from
+\emph{coeffect scalars} $\cclrd{r}, \cclrd{s}, \cclrd{t}$ to \emph{coeffect vectors} $\aclrd{r}, \aclrd{s}, \aclrd{t}$.
 
 \begin{definition}
-Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-an \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
-equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
-and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
-%
+Given a monoid $(\C, \cseq, \cunit)$ with a point-wise extension of the $\cseq$ operator to a vector 
+(written as $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$) and an operation lifting scalars to vectors 
+$\alift{-}$, an \emph{indexed comonad} over a category $\catc$ is a triple 
+$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}})$:
+
 \begin{compactitem}
-\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
-  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
-\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
-  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
-\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
-  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\item $\ctyp{\aclrd{\textbf{r}}}{}$ for all $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$ is a family of object mappings 
+\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\alift{\cunit}}{\alpha} \rightarrow \alpha$ 
+\item $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$ is a mapping $(\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta) 
+  \rightarrow (\ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{s}}}{\beta})$
 \end{compactitem}
@@
 
+\noindent
+The object mapping $\ctyp{\aclrd{\textbf{r}}}$ is now indexed by a vector rather than by a scalar 
+$\ctyp{\cclrd{r}}$ as in the previous chapter. This new definition supersedes the old one, because a 
+flat coeffect annotation can be seen as singleton vectors.
+
+The operation $\ident{counit}_{\cunit}$ operates on a singleton-vector. This means that it will always
+return a single variable value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
+operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This assymmetry
+is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
+takes a vector of variables, but always produces just a single value. Thus the resulting function 
+also takes a vector of variables, but always returns a context with singleton variable vector.
+In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
+no way of constructing values containing $\xtimes$.
+
 % --------------------------------------------------------------------------------------------------
 
+\subsection{Structural indexed comonads}
+
+The flat indexed comonad structure extends indexed comonads with operations 
+$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
+that combine or split the additional (flat) context and are annotated with the flat 
+coeffect operations $\czip$ and $\cpar$, respectively. In the structural version, we use
+corresponding operations that operate on variable vectors represented using $\xtimes$ and
+are annotated with a tensor $\atimes$ which mirrors the variable structure.
+
+The following definition also includes $\ident{lift}_{\cclrd{r'},\cclrd{r}}$, which is similar as
+before and models sub-coeffecting and also $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ which models
+duplication of a variable in a context needed for the semantics of contraction:
+
+\begin{definition}
+Given a structural coeffect algebra formed by $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$
+with operations $\alift{-}$ and $\aseq$, a \emph{structural indexed comonad} is an indexed comonad over 
+the monoid $(\C, \cseq, \cunit)$ equipped with families of operations $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$, 
+$\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$,
+$\ident{dup}_{\cclrd{r}, \cclrd{s}}$ and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
+%
+\begin{compactitem}
+\item $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
+  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}$
+\item $\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
+  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
+\item $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ is a family of mappings
+  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \xtimes \alpha)}$
+\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
+  $\ctyp{\alift{\cclrd{r'}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\end{compactitem}
+%
+Such that the following equalities hold:
+\begin{equation*}
+\begin{array}{l}
+  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \equiv \ident{id} \\
+  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \equiv \ident{id} 
+\end{array}
+\end{equation*}
+
+\end{definition}
+
+\noindent
+The operations differ from those of the flat indexed comonad in that the \ident{merge} and 
+\ident{split} operations are required to be inverse functions and to preserve the additional 
+information about the context. This was not required for the flat system where the operations
+could under- or over-approximate. Note that the operations use $\xtimes$ to combine or
+split the contained values. This means that they operate on free-variable vectors rather than 
+on ordinary products.
+
+The \ident{dup} mapping is a new operation that was not required for a flat calculus. It 
+takes a variable context with a single variable annotated with $\cclrd{r}\cpar\cclrd{s}$,
+duplicates the value of the variable $\alpha$ and splits the additional context between the two
+new variables. In flat calculus, this operation has been expressed using ordinary tuple 
+construction, which is not possible here -- the returned context needs to contain a 
+two-element vector $\alpha \xtimes \alpha$.
+
+Finally, the \ident{lift} mapping is almost the same as in the flat version. It
+operates on a singleton vector, which is equivalent to operating on a scalar as before.
+The operation could easily be extended to a vector in a point-wise way, but we keep it simple
+and perform sub-coeffecting separately on individual variables.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Examples}
+~
+\newpage
+~
 

### 2014-08-23 text/Chapter05.tex
Examples of structural comonads (Chapter 5)

@@
 The structural coeffect algebra for liveness is formed by
-$(L, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $L=\{ \ident{L}, \ident{D} \}$ is 
+$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
 the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
@@
 
-\subsection{Examples}
-~
-\newpage
-~
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of structural caluculus}
-
-\newcommand{\ctx}{\textit{ctx}}
-
-\begin{figure*}[t]
-
+\begin{figure*}[!t]
 \begin{equation*}
@@
   \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
-  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
+  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \alift{\cclrd{s'}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
 \\[0.5em]
 \sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
-        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\
-  \hspace{2em}
-  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
-  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
+        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\[-0.25em]
+  \hspace{2em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
+     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
+  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, 
+     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
 \\[0.5em]
 \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
-        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\
-  \hspace{2em}
-  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
-  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)
+        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\[-0.25em]
+  \hspace{2em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
+  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, 
+      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)
 \end{array}
@@
 \end{equation*}
-
 \caption{Categorical semantics of the structural coeffect calculus}
@@
 
-~
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of structural caluculus}
+
+The concrete semantics for liveness and bounded variable use shown in 
+Sections~\ref{sec:applications-struct-live} and \ref{sec:applications-struct-bll}
+suggests that semantics of structural coeffect calculi tend to be more complex 
+than semantics of flat coeffect calculi. The complexity comes from the fact
+that we need a more expressive representation of the variable context -- \eg~a vector
+of optional values -- and that the structural system needs to pass separate variable contexts
+to the sub-expressions.
+
+The latter aspect is fully captured by the semantics shown in this section. The
+earlier point is left to the concrete notion of structural coeffect. Our model still gives
+us the flexibility of defining the concrete representation of variable vectors. We explore
+a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
+at the unified categorical semantics defined in terms of \emph{structural indexed comonads}.
+
+\paragraph{Contexts and functions.}
+In the structural coeffect calculus, expressions in context are interpreted as functions taking
+a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
+such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
+structure is annotated with a scalar, such as $\ctyp{\cclrd{r}}$, which we treat as being equivalent
+to a singleton vector annotation $\ctyp{\alift{\cclrd{r}}}$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
+  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+Note that the instances of flat indexed comonad ignored the fact that the variable 
+context wrapped in the data structure is a product. This is not generally the case for the 
+structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
+are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
+rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need for examining the structure
+of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
+
+\paragraph{Expressions.}
+The semantics of strucutural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
+the previous chapter, the semantics is written in a programming language style using constructs
+such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
+can be written using standard primitives (currying, uncurrying, function pairing etc.).
+
+The following summarizes how the standard syntax-driven rules work, highlighting the differences
+from the flat version:
+
+\begin{itemize}
+\item When accessing a variable (\emph{var}), the context now contains \emph{only} the accessed
+  variable and so the semantics is just $\ident{counit}_{\cunit}$ without a projection.
+  Constants (\emph{const}) are interpreted using a global dictionary $\delta$ as earlier.
+
+\item The semantics of flat function application first duplicated the context so that the same 
+  variables can be passed to both sub-expressions. This is no longer needed -- the 
+  (\emph{app}) rule splits the variables \emph{including} the additional context into two parts.
+  Passing the first context to the semantics of $e_1$ gives us a function
+  $\ctyp{\alift{\cclrd{t}}}{\tau_1} \rightarrow \tau_2$.
+
+  The argument for the function is obtained by applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ 
+  to the semantics of $e_2$. The resulting function 
+  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
+  is then called with the latter part of the context to obtain argument for the first function.
+
+\item The semantic of function abstraction (\emph{abs}) is syntactically the same as in the 
+  flat version -- the only difference is that we now merge a free-variable context with a 
+  singleton vector, both at the level of variable assignments and at the level of coeffect 
+  annotations.
+\end{itemize}
+
+The semantics for the non-syntax-driven rules performs transformations on the 
+free-variable context. Weakening (\emph{weak}) splits the context and ignores the part corresponding
+to the removed variable. If we were modelling the semantics in a language with a linear 
+type system, this would require an additional operation for ignoring a context annotated with 
+$\aclrd{\czero}$.
+
+The remaining rules perform a transformation anywhere inside the free-variable vector. 
+To simplify writing the semantics, we define a helper operation 
+$\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
+that splits the variable vector into three parts, transforms the middle part and then 
+merges them, using the newly transformed middle part.
+
+The transformations on the middle part are quite simple. The (\emph{sub}) rule 
+uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
+the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
+uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a varaible while 
+splitting its additional context.
+
+\paragraph{Properties.}
+
+As in the flat calculus, the main reason for defining the categorical semantics in this chapter
+is to provide validation for the design of the calculus. As we show in the next section, the 
+discussed examples (liveness, data-flow, bounded variable reuse) form \emph{structural indexed
+comonads} and so the calculus captures them correctly if the coeffect annotations in the typing
+rules match the indices in the semantics. More formally:
+
+\begin{remark}[Correspondence]
+In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$ 
+and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types 
+$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$ 
+and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined 
+by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
+\end{remark}
+\begin{proof}
+By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
+\end{proof}
+
+\noindent
+As in the flat calculus, the primitive operations of the structural indexed comonad are all 
+annotated with different operations provided by the coeffect annotations. This means that the
+semantics uniquely determines the structure of the typing rules of the strucutral coeffect
+calculus. Thanks to the correspondence between the product structure $\atimes$ of the annotations
+and the variable context $\xtimes$, the correspondence property also guarantees that variable
+values are split correctly, as required by the structural nature of the type system.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Examples of structural indexed comonads}
+\label{sec:struct-semantics-examples}
+
+The categorical semantics for structural coeffect calculus can be easily instantiated to give
+a semantics of a concrete calculus. In this section, we look at the three examples discussed
+throughout this chapter -- structural liveness and data-flow and bounded variable reuse. 
+Some aspects of the earlier two examples will be similar to flat versions discussed in
+Section~\ref{sec:flat-semantics} -- they are based on the same data structures (option and a 
+list, respectively), but the data structures are composed differently. Generally speaking -- 
+rather than having a data structure over a product of variables, we now have 
+a vector of variables over a specific data structure.
+
+The abstract semantics does not specify how vectors of variables should be represented, so this can
+vary in concrete instantiations. In all our examples, we represent a vector of variables as a 
+product written using $\times$. To distinguish between products representing vectors and ordinary
+products (\eg~a product of contexts returned by \ident{split}), we write vectors using 
+$\langle a, \ldots, b \rangle$ rather than using parentheses.
+
+\paragraph{Data-flow.}
+It is interesting to note that the semantics of data-flow and bounded variable both keep
+a product of multiple values for each variable, so they are both built around an \emph{indexed list}
+data structure. However, their \ident{cobind} and \ident{dup} operations work differently. We start
+by looking at the structure modelling data-flow computations (variables written in bold face such
+as $\mathbf{a}_1$ range over vectors while $a_1$ ranges over individual values).
+
+\begin{example}[Indexed list for data-flow]
+The indexed list model of data-flow computations is defined over a structural coeffect
+algebra $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
+is indexed by required number of past variables for each individual variable. It is defined
+over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
+containing a current value followed by $n_i$ past values:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
+\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
+  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{(n_k+1)-\textnormal{times}}
+\end{equation*}
+%
+The mappings that define the structural indexed comonad include the \ident{split} and \ident{merge}
+operations that are shared by the other two examples (discussed below):
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{merge}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
+  (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle) = \\[-0.25em]
+\quad \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle
+\\[0.45em]
+\ident{split}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
+   \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle = \\[-0.25em]
+\quad (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle)
+\end{array}
+\end{equation*}
+%
+The remaining mappings that are required by structural indexed comonad and capture the
+essence of data-flow computations are defined as:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{0}~\langle \langle a_0 \rangle \rangle = a_0
+\\[0.45em]
+\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
+  \langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle = \\[-0.25em]
+\quad \langle\langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle, ~\ldots~, \\[-0.25em]
+\quad ~~~\; f \langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle~\; \rangle\rangle
+\\[0.45em]
+\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{\mathit{max}(m,n)} \rangle \rangle =
+   \langle \langle a_1, \ldots, a_m \rangle, \langle a_1, \ldots, a_n \rangle \rangle
+\\[0.45em]
+\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
+  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
+\end{array}
+\end{equation*}
+\end{example}
+
+The definition of the indexed list data structure relies on the fact that the number of annotations
+corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
+containing $n_i+1$ values for $i$-\textit{th} variable (the annotation represents the number of 
+required \emph{past} values so one more value is required).
+
+The \ident{split} and \ident{merge} operations are defined separately, because they are not specific
+to the example. They operate on the top-level vectors of variables (without looking at the 
+representation of the variable). This means that we can re-use the same definitions for the following
+two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will represent options
+rather than lists).
+
+The mappings that explain how data-flow computations work are \ident{cobind} (representing sequential
+composition) and \ident{dup} (representing context sharing or parallel composition). In \ident{cobind},
+we get $k$ vectors corresponding to $k$ variables, each with $m+n_i$ values. The operation calls 
+$f$ $m$-times to obtain $m$ past values required as the result of type $\ctyp{\langle m \rangle} \beta$.
+
+The \ident{dup} operation needs to produce a two-varaible context containing $m$ and $n$ values,
+respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
+is simply a matter of restriction. Finally, \ident{counit} extracts the (only) value of the (only) variable
+and \ident{lift} drops additional past values that are not required.
+
+\paragraph{Bounded reuse.}
+As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list 
+structure. Rather than representing possibly different past values that can be shared (\emph{c.f.}~\ident{dup}), 
+the list now represents multiple copies of the same value that cannot be shared.
+
+\begin{example}[Indexed list for bounded reuse]
+The indexed list model of bounded variable reuse is defined over a structural coeffect
+algebra $(\mathbb{N}, \ast, +, 1, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$ 
+is a vector containing $n_i$ values of $i$-\textit{th} variable:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
+\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
+  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{n_k-\textnormal{times}}
+\end{equation*}
+%
+The \ident{merge} and \ident{split} operations are defined as in \emph{indexed list for data-flow}.
+The operations that capture the behaviour of bounded reuse are defined as:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{1}~\langle \langle a_0 \rangle \rangle = a_0
+\\[0.45em]
+\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
+  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
+\\[0.45em]
+\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{m+n} \rangle \rangle = 
+\quad \langle \langle a_1, \ldots, a_m \rangle, \langle a_{m+1}, \ldots, a_{m+n} \rangle \rangle
+\\[0.45em]
+\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
+  \langle \langle a_{1,0}, \ldots a_{1,m\ast{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m\ast{n_k}} \rangle \rangle = \\[-0.25em]
+\quad \langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1 - 1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k - 1}} \rangle \rangle, ~\ldots~, \\[-0.25em]
+\quad ~~ f \langle \langle a_{1,(m-1)\ast n_1}, \ldots a_{1,(m-1) \ast n_1} \rangle, \ldots, 
+                   \langle a_{k,m\ast n_k - 1}, \ldots a_{k,m\ast n_k - 1} \rangle \rangle~\; \rangle
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{counit} and \ident{lift} operations are defined as previously -- variable access 
+extracts the only value of the only variable and sub-coeffecting allows discarding multiple
+copies of a value that are not needed.
+
+In the bounded variable reuse system, variable sharing is annotated with $+$ (in contrast with
+\textit{max} used in data-flow). The \indent{dup} operation thus splits the $m+n$ available
+values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The 
+\ident{cobind} operation works similarly -- it splits $m\ast n_i$ available values of each
+variable into $m$ vectors containing $n_i$ copies and then calls the $f$ function $m$-times to 
+obtain $m$ resulting values without sharing any input value.
+
+\paragraph{Liveness.}
+In both data-flow and bounded reuse, the data type is defined as a vector of values obtained
+by applying some parameterized data type (indexed list) to types of individual variables. We can
+generalize this pattern and define $\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of 
+$D^{l}$ where $D^{l}$ is a simpler indexed data type. For liveness, the definition lets us 
+reuse the mapping used when defining the semantics of flat liveness. However, we cannot fully
+define the semantics of the structural version in terms of the flat version -- the 
+\ident{cobind} operation is different and we need to provide the \ident{dup} operation.
+
+\begin{example}[Structural indexed option]
+Given a structural coeffect algebra formed by $(\{ \ident{L}, \ident{D} \}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$
+and the indexed option data type $D^l$, such that $D^{\ident{D}}\alpha = 1$ and $D^{\ident{L}}\alpha=\alpha$,
+the data type for structural indexed option comonad is:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
+  D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
+\end{equation*}
+%
+The \ident{merge} and \ident{split} operations are defined as earlier. The remaining operations
+model variable liveness as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{cobind}_{\ident{L}, \langle l_1, \ldots, l_n \rangle}~f~\langle a_1, \ldots, a_n \rangle = \langle f~\langle a_1, \ldots, a_n \rangle \rangle \\[-0.25em]
+\ident{cobind}_{\ident{D}, \langle \ident{D}, \ldots, \ident{D} \rangle}~f~\langle (), \ldots, () \rangle = \langle \ident{D} \rangle
+\\[0.5em]
+\hspace{-0.5em}
+\begin{array}{lll}
+~\\[-0.85em]
+\ident{dup}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle (), () \rangle \\[-0.25em]
+\ident{dup}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle a, () \rangle \\[-0.25em]
+\ident{dup}_{\ident{D}, \ident{L}} \langle a \rangle &\narrow{=}& \langle (), a \rangle \\[-0.25em]
+\ident{dup}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a, a \rangle 
+\end{array}
+\qquad\quad
+\begin{array}{lll}
+\ident{counit}_{\ident{L}}~\langle a \rangle &\narrow{=}& a
+\\[0.45em]
+\ident{lift}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a \rangle \\[-0.25em]
+\ident{lift}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
+\ident{lift}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
+\end{array}
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+When the expected result of the \ident{cobind} operation is dead (second case), the operation can
+ignore all inputs and directly return the unit value $()$. Otherwise, it passes the vector of 
+input variables to $f$ as-is -- no matter whether the individual values are live or dead. The
+\ident{L} annotation is a unit with respect to $\cap$ and so the annotations expected by $f$ are
+the same as those required by the result of \ident{cobind}.
+
+The \ident{dup} operation resembles with the flat version of \ident{split} -- this is expected as
+duplication in the flat calculus is performed by first duplicating the variable context (using 
+\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair which may or
+may not contain value, depending on the annotations.
+
+Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
+$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models 
+sub-coeffecting which may drop an available value (second case) or behaves as identity.
+
+% ==================================================================================================
+%                                                                   
+%       ###                  #                    #       #          
+%      #   #                 #                    #                  
+%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
+%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
+%          #  #  ##  #   #   #      ####  #       #       #    #     
+%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
+%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
+%             #   #                                                  
+%              ###                                                   
+%
+% ==================================================================================================
+
+\section{Equational theory}
+\label{sec:struct-syntax}
+
+??
+
 \newpage
-
-- variable access gets single-variable context and so it just projects the value
-- application does not duplicate before splitting - it splits directly (to get two sub-contexts)
-- abs is pretty much the same
-- weakening just drops a value with ``ign'' annotation - we write this using pseuod-pattern matching, but it could be special operation in the category
-- sub, exchange and contraction are quite similar - they extract and then merge
-~ 
-\newpage
-
-\newcommand{\ctimes}{ \hat{\times} }
-\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
-
-\newcommand{\csprd}[0]{ \times }
-\newcommand{\csvop}[0]{ \vee }
-\newcommand{\cmerge}[0]{ \cpar }
-\newcommand{\csplit}[0]{ \czip }
-
-% alculus names and rule names
-\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
-\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
-
-\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
-\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
-\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
-\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
-\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}
-
-\newcommand{\cobind}[1]{{#1}^\dagger}
-\newcommand{\obj}[0]{obj}
-\newcommand{\ccat}[0]{C}
-
-\begin{figure*}[t]
-\newcommand{\uu}[1]{\overline{#1}}
-\begin{equation*}
-\sem{\ctyp{r_1 \csprd \ldots \csprd r_n}{(x_1 : \tau_1 \times \ldots \times x_n : \tau_n)} \vdash e : \tau} : 
-  \ctyp{r_1 \csprd \ldots \csprd r_n}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau
-\end{equation*}
-\begin{equation*}
-\begin{array}{rcll}
-\sem{\ctyp{0}{\Gamma} \vdash x_i : \tau_i} &=&
-    \epsilon_0                                                                   &\quad$\ruless{var}$
-\\[0.32em]
-\sem{\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2} &=& 
-    \Lambda (\sem{\ctyp{r \csprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2} \circ \cmerge_{r, s})    &\quad$\ruless{fun}$
-\\[0.32em]
-\sem{\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau} &=& 
-    (\lambda (\gamma_1, \gamma_2) \rightarrow
-    \sem{\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2}~\gamma_1~
-    (\cobind{\sem{\ctyp{t}{\Gamma_2}\vdash e_2 : \tau_1}}{t,r}~\gamma_2)) \circ \csplit_{s,(r \vee t)}  &\quad$\ruless{app}$
-\\[0.32em]
-\sem{\ctyp{r'}{\Gamma'} \vdash e : \tau} &=&
-    \sem{\ctyp{r}{\Gamma} \vdash e : \tau} \circ \sem{\ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}       &\quad$\ruless{ctx}$
-\end{array}
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rclll}
-\sem{\ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)}}}
-&=& \cmerge_{r, s} \circ (\sem{\ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}}} \times \ident{id})~\circ \csplit_{r', s}\quad&\quad\quad\quad& % spaces to align
-\textnormal{\rulessc{nest}}
-\\[0.8em]
-\sem{\ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} }{ \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}}
-&=& \cmerge_{s, r} \circ \ident{swap} \circ \csplit_{r, s}&&
-\textnormal{\rulessc{exch}}
-\\[0.8em]
-\sem{\ctxtrans{\ctyp{r \csprd 1}{(\Gamma, ())}}{\ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 1}&&
-\textnormal{\rulessc{empty}}
-\\[0.8em]
-\sem{\ctxtrans{ \ctyp{r\csprd 0}{(\Gamma, x:\tau)} }{ \ctyp{r}{\Gamma}}}
-&=&\ident{fst} \circ \csplit_{r, 0}&&
-\textnormal{\rulessc{weak}}
-\\[0.8em]
-\sem{\ctxtrans{\ctyp{r \csvop s}{(x:\tau)}}{
-    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}}
-&=&\cmerge_{r, s} \circ \cdupl_{r, s}&&
-\textnormal{\rulessc{contr}}
-\\[0.8em]
-\sem{\ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))}}{
-    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}}\;
-&=& \cmerge_{r \csprd s, t} \circ (\cmerge_{r, s} \times \ident{id}) \circ \ident{assoc}_1 \circ
-        (\ident{id} \times \csplit_{s, t}) \circ \csplit_{r, s \csprd t}&&
-\textnormal{\rulessc{assoc}}
-\\[0.8em]
-\sem{\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}}
-&=&
-\iota_{r, s}
-&&
-\textnormal{\rulessc{sub}}
-\end{array}
-\end{equation*}
-
-\caption{Categorical semantics for~\clstr}
-\label{fig:structural-semantics}
-\vspace{-0.5em}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Semantics of structural coeffects}
-\label{sec:structural-semantics}
-
-The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
-\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
-of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
-is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
-
-As discussed in \ref{sec:structural-types}, the variable context $\Gamma$ in structural coeffect
-system is not a simple finite product, but instead a binary tree. To model this, we do not use
-ordinary products in the domain of the semantic function, but instead use a special 
-constructor $\ctimes$. This way, we can guarantee that the variable structure corresponds to 
-the tag structure.
-
-\subsection{Structural tagged comonads}
-\label{sec:structural-tagged-comonads}
-
-To model composition of functions, we reuse the definition of \emph{tagged comonads} from 
-Section~\ref{sec:tagged-comonads} without any change. This means that composing morphisms
-$T^r \tau_1 \rightarrow \tau_2$ with $T^s \tau_2 \rightarrow \tau_3$ still gives us a morphism
-$T^{r \csvop s} \tau_1 \rightarrow \tau_3$ and we use the $\csvop$ operation to combine the 
-context-requirements.
-
-However, functions that do not exist in context have only a single input variable (with a 
-single corresponding tag). To model complex variable contexts, we need two additional
-operations that allow manipulation with the variable context. Similarly to the model of \clflt,
-we also require operations that model duplication and sub-coeffecting:
-
-\begin{definition}[Structural tagged comonad]
-Given a structural coeffect tag structure $(S, \csprd, \csvop, 0, 1)$ a \emph{structural tagged comonad} is a 
-tagged comonad over $(S, \csvop, 0)$ comprising of $T^r$, $\epsilon_0$ and
-$\cobind{(-)}{r, s}$ together with a mapping $- \ctimes -$ from a pair of objects 
-$\obj{\ccat} \times \obj{\ccat}$ to an object $\obj{\ccat}$ and families of mappings:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\cmerge_{r, s} &:& T^r A \times T^s B \rightarrow T^{(r \csprd s)}(A \ctimes B)\\
-\csplit_{r, s} &:& T^{(r \csprd s)}(A \ctimes B) \rightarrow T^r A \times T^s N\\
-\end{array}
-\end{equation*}
-%
-% Delta is still defineable
-% ?? \cdupl_{r, s}  &:& T^{(r \csvop s)}A \rightarrow T^r A \times T^s A\\
-%
-And with a family of mappings $\iota_{r, s} : T^r A \rightarrow T^s A$ for all $r, s \in S$ such that $r \vee s = r$.
-%
-%  TODO: laws?
-%
-\end{definition}
-
-The family of mappings $\iota_{r, s}$ is the same as for \emph{flat} coeffects
-and it can still be used to define a family of mappings that represents \emph{duplicating}
-of variables while splitting the additional coeffect tags:
-%
-\begin{equation*}
-\begin{array}{l}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A\\
-\cdupl_{r, s}(\gamma) = (\iota_{(r\vee s), r}~\gamma, \iota_{(r\vee s), s}~\gamma)
-\end{array}
-\end{equation*}
-%
-The type of the $\cmerge_{r,s}$ operation looks similar to the one used for \emph{flat} coeffects,
-but with two differences. Firstly, it combines tags using $\csprd$ instead of $\vee$, 
-which corresponds to the fact that the variable context now consists of two parts (a tree node).
-Secondly, to model the tree node, the resulting context is modelled as $A \ctimes B$
-(instead of $A \times B$ as previously).
-
-To model structural coeffects, we also need $\csplit_{r, s}$, which serves as the dual
-of $\cmerge_{r, s}$. It represents \emph{splitting} of context containing multiple 
-variables. The operation was not needed for \clflt, because there \emph{splitting} could
-be defined in terms of \emph{duplication} provided by $\cdupl_{r, s}$. For \clstr,
-the situation is different. The $\csplit_{r, s}$ operation takes a context annotated
-with $r \csprd s$ that carries $A \ctimes B$.
-
-Examples of \emph{structural tagged comonads} are shown in Section~\ref{sec:structural-tainting}.
-Before looking at them, we finish our discussion of categorical semantics.
-
-\paragraph{Categorical notes.}
-The mapping $T^r$ can be extended to an endo-functor $\hat{T}^r$ in the same way as in
-Section~\ref{sec:tagged-comonads}. However, we still cannot freely manipulate the 
-variables in the context. Given a context modelled as $T^{r \csprd s}(A \ctimes B)$,
-we can lift a morphism $f$ to $\hat{T}^{r \csprd s}(f)$, but we cannot manipulate the
-variables, because $A \ctimes B$ is not a product and does not have projections $\pi_i$.
-
-This also explains why $\csplit$ cannot be defined in terms of $\cdupl$. Even if we 
-could apply $\cdupl_{r, s}$ on the input (if the tag $r \csprd s$ coincided
-with tag $r \csvop s$) we would still not be able to obtain $T^r A$ from $T^r (A \ctimes B)$.
-
-This restriction is intentional -- at the semantic level, it prevents manipulations
-with the context that would break the correspondence between tag structure and the
-product structure.
-
-\subsection{Categorical semantics}
-The categorical semantics of \clstr is shown in Figure~\ref{fig:structural-semantics}.
-It uses the \emph{structural tagged comonad} structure introduced in the previous section,
-together with the helper operation $\cdupl_{r, s}$ and the following simple helper 
-operations:
-\begin{equation*}
-\begin{array}{rcl}
-\ident{assoc} &=& \lambda (\delta_r, (\delta_s, \delta_t)) \rightarrow ((\delta_r, \delta_s), \delta_t)\\
-\ident{swap}  &=& \lambda (\gamma_1, \gamma_2) \rightarrow (\gamma_2, \gamma_1)\\
-f \times g &=& \lambda (x, y) \rightarrow (f~x, g~y)
-\end{array}
-\end{equation*}
-
-When compared with the semantics of \clflt (Figure~\ref{fig:flat-semantics}), there is a number of 
-notable differences. Firstly, the rule \ruless{var} is now interpreted as $\epsilon_0$ without 
-the need for projection $\pi_i$. When accessing a variable, the context contains only the accessed
-variable. The \ruless{fun} rule has the same structure -- the only difference is that we use the
-$\csprd$ operator for combining context tags instead of $\vee$ (which is a result of the change of
-type signature in $\cmerge_{r, s}$).
-
-The rule \ruless{app} now uses the operation $\csplit_{s, (r \csvop t)}$ instead of $\cdupl_{s, (r \csvop t)}$,
-which means that it splits the context instead of duplicating it. This makes the system more 
-structural -- the expressions use disjunctive parts of the context -- and also explains why the
-composed coeffect tag is $s \csprd (r \csvop t)$.
-
-The only rule from \clflt~that was not syntax-directed (\rulefs{sub}) is now generalized to a number
-of non-syntax-directed rules \rulessc{} that perform various manipulations with the context. The semantics
-of $\sem{ \ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}} }$ is a function that, when given a context
-$\ctyp{r_1}{\Gamma_1}$ produces a new context $\ctyp{r_2}{\Gamma_2}$. The semantics in \ruless{ctx} then 
-takes a context, converts it to a new context which is compatible with the original expression $e$.
-The context manipulation rules work as follows:
-
-\begin{itemize}
-\item The \rulessc{nest} and \rulessc{exch} rules use $\csplit_{r, s}$ to split the context into a product
-  of contexts, then perform some operation with the contexts -- transform one and swap them, respectively.
-  Finally, they re-construct a single context using $\cmerge_{r, s}$.
-
-\item The \rulessc{empty} and \rulessc{weak} rules have the same semantics. They both split the context
-  and discard one part (containing either an unused variable or an empty context).
-
-\item If we interpreted \rulessc{contr} by applying functor $T^{r\csvop s}$ to a function that duplicates
-  a variable, the resulting context would be $\ctyp{r \csvop s}(x:\tau, x:\tau)$, which would break the
-  correspondence between coeffect tag and context variable structure. However, that interpretation would be
-  incorrect, because we use $\ctimes$ instead of normal product for variable contexts. As a result, the
-  rule has to be interpreted as a composition of $\cdupl_{r, s}$ and $\cmerge_{r, s}$, which also turns
-  a tag $r \csvop s$ into $r \csprd s$.
-
-\item The \rulessc{assoc} rule is similar to \rulessc{exch} in the sense that it de-constructs the context,
-  manipulates it (using \ident{assoc}) and then re-constructs it.
-
-\item Finally, the \rulessc{sub} rule interprets sub-coeffecting on the context associated with a single
-  variable using the primitive natural transformation $\iota_{r, s}$.
-\end{itemize}
-
-\paragraph{Alternative: Separate variables.} As an alternative, we could model an expression by
-attaching the context separately to individual variables. This an expression
-$\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ would be modelled as 
-$\ctyp{r}{\Gamma_1} \times \ctyp{s}{\Gamma_2} \rightarrow \tau$. However, this 
-approach largely complicates the definition of application (where tag of all 
-variables in a context is affected). Moreover, it makes it impossible to express
-\clflt~in terms of \clstr~as discussed in Section~\ref{sec:generalization}.
-
-\paragraph{Alternative: Without sub-coeffecting.} The semantics presented above uses the natural
-transformation $\iota_{r, s}$, which represents sub-coeffecting, to define the duplication
-operation $\cdupl_{r, s}$. However, structural coeffect calculus \clstr does not require 
-sub-coeffecting in the same way as flat \clflt (where it is required for subject reduction).
-
-This means that it is possible to define a variant of the system that does not have the \rulest{sub} 
-typing rule. Then the semantics does not need the $\iota_{r, s}$ transformation, but instead,
-the following natural transformation has to be provided:
-%
-\begin{equation*}
-\cdupl_{r, s} : T^{(r \vee s)}A \rightarrow T^r A \times T^s A
-\end{equation*}
-%
-This variant of the system could be used to define a system that ensures that all provided 
-context is used and is not over-approximated. This difference is similar to the difference between
-affine type systems (where a variable can be used at most once) and linear type systems 
-(where a variable has to be used exactly once).
-
-\newpage
-~
-\newpage
-~
 
@@
 
-% --------------------------------------------------------------------------------------------------
+\newcommand{\ctimes}{ \hat{\times} }
+\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
+
+\newcommand{\csprd}[0]{ \times }
+\newcommand{\csvop}[0]{ \vee }
+\newcommand{\cmerge}[0]{ \cpar }
+\newcommand{\csplit}[0]{ \czip }
+
+\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
+\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
+\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
 

### 2014-09-05 text/Abstract.tex
Equational theory (Chapter 5) and Abstract

@@
-% Abstract
-
+%!TEX root = ../main.tex
 \pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer
@@
 \chapter*{Abstract} % Abstract name
-The development of programming languages needs to reflect important changes in the industry.
-In recent years, this included \eg~the development of parallel programming models (in reaction 
-to the multi-core revolution). This thesis is a response to another such revolution -- the 
-diversification of devices and systems where programs run.
+The development of programming languages needs to reflect important changes in the way
+programs execute. In recent years, this included \eg~the development of parallel programming 
+models (in reaction to the multi-core revolution) or improvements in data access technologies. 
+This thesis is a response to another such revolution -- the diversification of devices and 
+systems where programs run.
 
-We develop the foundations for statically typed functional languages that understand 
-the \emph{context} or environment in which programs execute. Such context includes different 
-platforms (and their versions) in cross-platform applications, resources available in different 
-execution environments (\eg \; GPS sensor on a phone and database on the server), but also more 
-traditional notions such as variable context (tracking variable usage in static analyses) or 
-past values in stream-based data-flow programming.
+The key point made by this thesis is the realization that execution environment or
+\emph{context} is fundamental for writing modern applications and that programming 
+languages should provide abstractions for programming with context and verifying how 
+it is accessed.
 
-The thesis presents three \emph{coeffect} calculi that capture different notions of 
-context-awareness: \emph{flat} calculus capturing contextual properties of the execution 
-environment, \emph{structural} calculus capturing contextual properties related to 
-variable usage and \emph{meta-language} calculus which allows reasoning about multiple notions
-of context-dependence in a single language and provides pathway to embedding the other two
-calculi in existing languages.
+We identify a number of program properties that were not connected before, but model some notion 
+of context. Our examples include tracking different execution platforms (and their versions) 
+in cross-platform development, resources available in different execution environments (\eg~GPS
+sensor on a phone and database on the server), but also more traditional notions such as 
+variable usage (\eg~in liveness analaysis) or past values in stream-based data-flow programming.
+
+Our first contribution is the discovery of the connection between the above examples and
+their novel presentation -- in the form of calculi with type systems and formal semantics --
+that highlights this relationship. Our second and third contribution are two \emph{coeffect}
+calculi that capture commonalities in the presented examples. In particular, our 
+\emph{flat} coeffect calculus models languages with contextual properties of the execution 
+environment and our \emph{structural} coeffect calculus models languages where the contextual
+properties are attached to the variable usage.
 
@@
 an \emph{indexed} comonad (the dual of monads) and show how they provide semantics of the
-presented three calculi.
+presented two calculi.
 

### 2014-09-05 text/Chapter05.tex
Equational theory (Chapter 5) and Abstract

@@
 
-??
+Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
+different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
+certain equational properties hold for all instances of the calculus.
+
+\subsection{From flat to structural coeffects}
+
+The discussion of equational theory for flat calculus in Section~\ref{sec:flat-syntax} considered
+different reductions that can be used as a basis for operational semantics for calculi satisfying
+different additional properties, as none of the techniques works universally. The structural 
+coeffect calculus has more desirable equational properties -- in particular, both $\beta$-reduction
+and $\eta$-expansion can be proved using just the properties of strucutral coeffect algebra. Thus
+this section focuses on these reductions. 
+Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
+the structural coeffect calculus satisfies both are \emph{local soundness} and \emph{local completeness}.
+
+% The structural coeffect calculus clstr~associates information with individual variables. This means
+% that when an expression requires certain context, we know from what scope it comes -- the
+% context must be provided by a scope that defines the associated variable, which is either a
+% lambda abstraction or global scope. This distinguishes the structural system from the flat system
+% where context could have been provided by any scope and the lambda rule allowed arbitrary 
+% splitting of context requirements between the two scopes (or declaration and caller site).
+
+\subsection{Holes substitution lemma}
+
+if
+
+\begin{equation*}
+\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s
+\end{equation*}
+
+and
+
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_s}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+  \coctx{x_k\!:\!\tau_s}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
+\end{equation*}
+
+then
+
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{r}}} \,|\, \ldots \,|\, 
+  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{r}}}\,] \vdash \subst{\subst{e_r}{x_1}{e_s}\ldots}{x_k}{e_s} : \tau_r
+\end{equation*}
+
+\subsection{Structural coeffect systems}
+\label{prop:structural}
+
+For structural coeffect systems, recall that coeffects are vectors
+with $\czip = \cpar = \cons$ (vector concatenation) and $\czero =
+\langle{}\rangle$ (the empty vector), thus coeffect annotations
+comprise the \emph{free monoid} \ie, lists over the coeffect scalars
+(although we continue using the vector terminology). We first show
+substitution:
+
+\begin{lemma}[Substitution lemma] 
+\label{thm:structural-subst}
+In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
+\[
+\begin{array}{l}
+\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
+\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
+\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
+\end{array}
+\]
+\vspace{-1.5em}
+\label{lemm:structural-subst}
+\end{lemma}
+\begin{proof}
+By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
+and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
+\end{proof}
+
+\noindent
+Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
+$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
+$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
+the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
+Using the substitution lemma, we can demonstrate $\beta$-equality:
+%
+\begin{equation*}
+\inference{
+  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
+             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
+           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
+  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
+\equiv
+\subst{e_1}{x}{e_2} : \tau}
+\end{equation*}
+%%
+\noindent
+As a result, $\beta$-reduction preserves the type and coeffects of a term.
+This gives the following subject reduction property:
+
+\begin{theorem}[Subject reduction]
+In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
+$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
+\end{proof}
+
+\noindent
+Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
+\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
+\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
+and application does not eliminate too much.
+%
+\begin{equation*}
+\inference{
+  \inference 
+    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
+     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
+    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
+  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
+: \sigma \xrightarrow{\cclrd{s}} \tau}
+\end{equation*}
+%
+The last step uses the fact that 
+$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
+arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
+
+This highlights another difference between coeffects and effects, as $\eta$-equality does
+not hold for many notions of effect. For example, in a language with output
+effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
+its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
+hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
+allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
+$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
+the exact immediate requirements of $\lambda x . e x$ to match those of $e$.
 
@@
 \newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
-
-\subsection{Properties of reductions}
-\label{sec:structural-metatheory}
-Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
-operational meaning, because that will differ for every concrete application. For example,
-when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
-access to a specified element.
-
-Just like previously, we can state general properties of the reductions. As the syntax of expressions
-is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
-are also the same and can be found in Figure~\ref{fig:flat-substred}.
-
-The structural coeffect calculus \clstr~associates information with individual variables. This means
-that when an expression requires certain context, we know from what scope it comes -- the
-context must be provided by a scope that defines the associated variable, which is either a
-lambda abstraction or global scope. This distinguishes the structural system from the flat system
-where context could have been provided by any scope and the lambda rule allowed arbitrary 
-splitting of context requirements between the two scopes (or declaration and caller site).
 

### 2014-09-05 text/Chapter05.tex
Substitution for structural coeffects (Chapter 5)

@@
 Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
-different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
-certain equational properties hold for all instances of the calculus.
-
-\subsection{From flat to structural coeffects}
-
-The discussion of equational theory for flat calculus in Section~\ref{sec:flat-syntax} considered
-different reductions that can be used as a basis for operational semantics for calculi satisfying
-different additional properties, as none of the techniques works universally. The structural 
-coeffect calculus has more desirable equational properties -- in particular, both $\beta$-reduction
-and $\eta$-expansion can be proved using just the properties of strucutral coeffect algebra. Thus
-this section focuses on these reductions. 
+different notion of context and thus a different operational interpretation. As before, the 
+properties of the flat coeffect algebra guarantee that certain equational properties hold for 
+all instances of the calculus.
+
+We start the discussion by briefly considering the key aspects
+that make the equational theory of flat and structural coeffects different.
+
+\subsection{From flat coeffects to structural coeffects}
+\label{sec:struct-syntax-intro}
+
+When discussing equational theory for the flat calculus in Section~\ref{sec:flat-syntax}, we 
+noted that no single technique works universally for all flat coeffect calculi. We considered
+multiple different reductions that can be used as the basis for operational semantics for
+calculi satisfying different additional properties.
+
+The structural coeffect calculus has more desirable equational properties. In particular, 
+we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of strucutral 
+coeffect algebra. For this reason, we focus on these two reductions in this section.
 Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
-the structural coeffect calculus satisfies both are \emph{local soundness} and \emph{local completeness}.
-
-% The structural coeffect calculus clstr~associates information with individual variables. This means
-% that when an expression requires certain context, we know from what scope it comes -- the
-% context must be provided by a scope that defines the associated variable, which is either a
-% lambda abstraction or global scope. This distinguishes the structural system from the flat system
-% where context could have been provided by any scope and the lambda rule allowed arbitrary 
-% splitting of context requirements between the two scopes (or declaration and caller site).
-
-\subsection{Holes substitution lemma}
-
-if
-
-\begin{equation*}
-\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s
-\end{equation*}
-
-and
-
-\begin{equation*}
-\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_s}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
-  \coctx{x_k\!:\!\tau_s}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
-\end{equation*}
-
-then
-
-\begin{equation*}
-\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{r}}} \,|\, \ldots \,|\, 
-  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{r}}}\,] \vdash \subst{\subst{e_r}{x_1}{e_s}\ldots}{x_k}{e_s} : \tau_r
-\end{equation*}
-
-\subsection{Structural coeffect systems}
-\label{prop:structural}
-
-For structural coeffect systems, recall that coeffects are vectors
-with $\czip = \cpar = \cons$ (vector concatenation) and $\czero =
-\langle{}\rangle$ (the empty vector), thus coeffect annotations
-comprise the \emph{free monoid} \ie, lists over the coeffect scalars
-(although we continue using the vector terminology). We first show
-substitution:
-
-\begin{lemma}[Substitution lemma] 
-\label{thm:structural-subst}
-In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
-\[
-\begin{array}{l}
-\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
-\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
-\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\vspace{-1.5em}
-\label{lemm:structural-subst}
+the structural coeffect calculus satisfies both the \emph{local soundness} and 
+the \emph{local completeness} properties.
+
+\paragraph{Substitution for flat coeffects.} 
+The more interesting variant of the substitution lemma for flat coeffects 
+(Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
+to coincide. This enables the substitution to preserve the type of expressions, because all
+additional requirements arising as the result of the substitution can be associated with the
+declaration context. For example, consider the following example where implicit parameter 
+$\ident{?offset}$ is substituted for the variable $y$:
+%
+\begin{equation*}
+\begin{array}{rclcll}
+ \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{before}) \\
+\coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
+  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{after})
+\end{array}  
+\end{equation*}
+%
+The typing judgement obtained in (\emph{after}) preserves the type of the expression
+(function value) from the original typing (\emph{before}). This is possible thanks to the
+non-determinism involved in lambda abstraction -- as all operators of the flat coeffect
+algebra used here are $\cup$, we can place the additional requirement on the outer context.
+Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.
+
+Here, the flat coeffect calculus gives us typing with limited \emph{precision}, but 
+enough \emph{flexibility} to prove the substitution lemma.
+
+\paragraph{Substitution for structural coeffects.}
+In contrast, the substitution lemma for structural coeffects can be proven ecause 
+structural coeffect systems provide enough \emph{precision} to identify exactly with 
+which variable should a context requirement be associated.
+
+The following example shows a situation similar to the previous one -- here, we use
+structural data-flow calculus (with the \kvd{prev} construct to obtain previous
+value of an expression) and we substitute $w+z$ for $y$:
+%
+\begin{equation*}
+\begin{array}{rclcll}
+ \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} } 
+  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{before}) \\
+\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
+  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{after}) \\
+\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
+  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{final})
+\end{array}  
+\end{equation*}
+%
+The type of the function does not change, because the structural type system associates
+the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not 
+affect how the variable $x$ is used.
+
+The other aspect demonstrated in the example is how the coeffect of the substituted 
+variable affects the free-variable context of the substituted expression. Here, the
+original variable $y$ is annotated with $\cclrd{2}$ and we substitute it for an
+expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
+The substitution applies the operation $\aseq$ (modelling sequential composition) to 
+the annotation of the new context -- in the above example
+$\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}=\alift{\cclrd{2}, \cclrd{2}}$.
+
+\subsection{Holes and substitution lemma}
+
+As demonstrated in the previous section, reduction (and substitution) in the structural coeffect 
+calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
+Furthermore, we may also need to substitute for multiple variables in the variable context.
+
+For example, the expression $\lambda x.x+x$ is type-checked by type-checking $x_1 + x_2$,
+contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction 
+$(\lambda x.x+x)~(y+z)$ we thus need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
+This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
+for example in the bunched typing system \cite{substruct-bunched}.
+To express the substitution lemma, we define the notion of a \emph{context with holes}
+
+\begin{definition}[Context with holes]
+A \emph{context with holes} is a context such as $\coctx{x_1\!:\!\tau_1, \ldots, x_k\!:\!\tau_k}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_k}}}$,
+where some of the variable typings $x_i\!:\!\tau_i$ and corresponding coeffects $\cclrd{r_i}$ 
+are replaced by \emph{holes} written as $-$.
+
+\begin{equation*}
+~\Delta[\,\coctx{-}{-}\,]_n = \Delta[\,\underbrace{\coctx{-}{-} \,|\, \ldots \,|\, \coctx{-}{-}}_{ n-\textit{times} } \,]
+\end{equation*}
+\begin{equation*}
+\begin{array}{lcl}
+ \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
+   \coctx{-, \Gamma}{\alift{-} \atimes \aclrd{\textbf{s}} }
+   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_{n-1} \\
+ \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
+   \coctx{x\!:\!\tau, \Gamma}{\alift{\cclrd{r}} \atimes \aclrd{\textbf{s}} }
+   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_n \\
+ \Delta[\,\coctx{-}{-}\,]_0 &\narrow{:=}&
+   \coctx{()}{\alift{}}
+\end{array} 
+\end{equation*} 
+\end{definition}
+
+\noindent
+A context with $n$ holes may either start with a hole, followed by a context with $n-1$ holes, or it may
+start with a variable followed by a context with $n$ holes. Note that the definition ensures that
+the location of variable holes correspond to the locations of coeffect annotation holes. Given a context
+with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
+obtain an ordinary coeffect-annotated context.
+
+\begin{definition}[Hole filling] Given a context with $n$ holes $\coctx{\Delta}{\textbf{s}} \in 
+\Delta[\,\coctx{-}{-}\,]_n$, the hole filling operation written as 
+$\coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\mathbf{r_1}}} \,|\, \ldots \,|\, \coctx{\Gamma_n}{\aclrd{\mathbf{r_n}}} \,]$,
+which replaces the holes by the specified variables and corresponding coeffect annotations, is defined as:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \coctx{-, \Delta}{\alift{-} \atimes \textbf{s} }
+    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
+      \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}_1} \atimes \aclrd{\textbf{r}_2} } \\
+ && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
+     \coctx{\Delta}{\textbf{s}}[\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
+\\[0.5em]
+ \coctx{x_1\!:\!\tau, \Delta}{\alift{\cclrd{r_1}} \atimes \textbf{s} }
+    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
+      \coctx{x_1\!:\!\tau, \Gamma_2}{\alift{\cclrd{r_1}} \atimes \aclrd{\textbf{r}_2} } \\
+ && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
+     \coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
+\\[0.5em]
+ \coctx{()}{\alift{}}~[~] &\narrow{=}& \coctx{()}{\alift{}}
+\end{array} 
+\end{equation*} 
+\end{definition}
+
+\noindent
+When we substitute an expression with coeffects $\aclrd{\textbf{t}}$ (associated with variables $\Gamma$)
+for a variable that has coeffects $\cclrd{s}$, the resulting coeffects of $\Gamma$ need to 
+combine $\aclrd{\textbf{t}}$ and $\cclrd{s}$. Unlike in the flat coeffect systems, the structural 
+substitution does not require all coeffect algebra operations to coincide and so the combination
+is more interesting than in the bottom-pointed substitution for flat coeffects, where it used
+the only available operator (Lemma~\ref{thm:cbn-substitution-bot}).
+
+Substitution can be seen as sequential composition. Informally -- we first need to obtain the 
+value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
+requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
+are annotated with $\cclrd{s} \aseq \aclrd{\textbf{t}}$, using the (scalar-vector extension)
+of the sequential composition operator $\cseq$.
+
+\begin{lemma}[Multi-nary substitution]
+Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
+%
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_1}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+  \coctx{x_k\!:\!\tau_k}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
+\end{equation*}
+%
+and a expressions $e_i$ with free-variable contexts $\Gamma_i$ annotated with $\aclrd{\mathbf{t_i}}$:
+%
+\begin{equation*}
+\coctx{\Gamma_1}{\aclrd{\textbf{t}_1}} \vdash e_1 : \tau_1 
+\quad \ldots \quad
+\coctx{\Gamma_k}{\aclrd{\textbf{t}_k}} \vdash e_k : \tau_k
+\end{equation*}
+%
+then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
+where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
+%
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
+  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
+\end{equation*}
 \end{lemma}
 \begin{proof}
-By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
-and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
+By induction over $\vdash$, using the multi-nary aspect of the substitution 
+in the proof of the contraction case. See Appendix~?.
 \end{proof}
 
-\noindent
+\subsection{Reduction and expansion}
+
+\todo{Exterminate!}
+
+~
+
 Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
@@
 
-% ==================================================================================================
-%
-%                               ##      
-%      ####                     ##      
-%        ##                     ##      
-%        ##   ##   ##  ## ###   ##  ##  
-%        ##   ##   ##  ###  ##  ## ##   
-%    ##  ##   ##   ##  ##   ##  ####    
-%    ##  ##   ##  ###  ##   ##  ## ##   
-%     ####     ### ##  ##   ##  ##  ##  
-%                                      
-% ==================================================================================================
-
-\newcommand{\ctimes}{ \hat{\times} }
-\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }
-
-\newcommand{\csprd}[0]{ \times }
-\newcommand{\csvop}[0]{ \vee }
-\newcommand{\cmerge}[0]{ \cpar }
-\newcommand{\csplit}[0]{ \czip }
-
-\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
-\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}
-\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
-
-\paragraph{Internalized substitution.}
-Before looking at properties of the evaluation, we consider let binding, which can be 
-viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
-application and abstraction as follows.
-
-\begin{lemma}[Definition of let binding]
-If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
-then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
-\end{lemma}
-\begin{proof}
-The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
-correspond with the typing rule \rulest{let}:
-\begin{equation*}
-\inference[]
-  {\inference
-    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
-    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
-   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
-  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
-\end{equation*}
-\end{proof}
-
-The term $e_2$ which is substituted in $e_1$ is checked in a different variable
-and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
-a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
-the variable that we are substituting for. The let binding captures substitution for a 
-specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
-For a general substitution, we need to define the notion of context with a hole.
-
-\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
-\paragraph{Substitution and holes.}
-In \clstr, the structure of the variable context is not a set, but a tree. When substituting
-for a variable, we need to replace the variable in the context with the context of the
-substituted expression. In general, this can occur anywhere in the tree. To formulate the
-statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
-in the free variable context and in a corresponding part of the coeffect tag:
-
-\begin{equation*}
-\begin{array}{rcl}
-\Delta[-] &::=& \ctyp{1}{()} \\
-       & \sep & \ctyp{r}{(x:\tau)}\\
-       & \sep & \ctyp{-}{(-)} \\
-       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
-\end{array}
-\end{equation*}
-%
-Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
-operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
-$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
-Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
-general substitution lemma for \clstr.
-
-\begin{figure}
-
-\begin{equation*}
-\begin{array}{rcl}
-                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
-                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
-                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
-\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
-&\multicolumn{2}{l} {
-  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
-}
-\end{array}
-\end{equation*}
-\caption{The definition of hole filling operation for $\Delta[-]$}
-\label{fig:structural-hole-fill}
-\vspace{-2em}
-\end{figure}
-
-\begin{lemma}[Substitution Lemma]
-\label{thm:structural-subst}
-If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
-then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
-\end{lemma} 
-\begin{proof}
-Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
-$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
-\end{proof}
-
-\begin{theorem}[Subject reduction]
-\label{thm:structural-reduction}
-If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
-$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
-\end{theorem}
-\begin{proof}
-Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
-\end{proof}
-
-\paragraph{Local soundness and completeness.}
-As with the previous calculus, we want to guarantee that the introduction and
-elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
-This can be done by showing \emph{local soundness} and \emph{local completeness},
-which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
-case of subject reduction and the latter is proved by a simple derivation:
-%
-\begin{theorem}[Local soundness]
-If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
-$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
-\end{theorem}
-\begin{proof}
-Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
-\end{proof}
-%
-\begin{theorem}[Local completeness]
-If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
-$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
-\end{theorem}
-\begin{proof} The property is proved by the following typing derivation:
-%
-\begin{equation*}
-\inference
-  {\inference
-     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
-      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
-     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\qedhere
-\end{equation*}
-\end{proof}
-%
-In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
-guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
-$\lambda x.f x$ required for local completeness was not the only possible derivation.
-In the last step, it was possible to split the coeffect tag arbitrarily between the 
-context and the function type.
-
-In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
-to be associative and to have units and so a unique splitting may exist.  For example, if we
-define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
-$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
-has additional properties, then there may be other possible derivation.
-
-% ==================================================================================================
-
-\section{Examples of structural coeffects}
-
-\subsection{Example: Liveness analysis}
-\label{sec:structural-neededness}
-
-\subsection{Example: Data-flow (revisited) }
-\label{sec:structural-tainting}
-
-\todo{Also, consider additional language features that we consider for flat coeffects
-  (mainly recursion and possibly conditionals)}
-
 \section{Conclusions}

### 2014-09-11 text/Abstract.tex
Tweaks at the beginning and lofty closing paragraph (Chapter 1)

@@
 The development of programming languages needs to reflect important changes in the way
-programs execute. In recent years, this included \eg~the development of parallel programming 
+programs execute. In recent years, this has included \eg~the development of parallel programming 
 models (in reaction to the multi-core revolution) or improvements in data access technologies. 
@@
 sensor on a phone and database on the server), but also more traditional notions such as 
-variable usage (\eg~in liveness analaysis) or past values in stream-based data-flow programming.
+variable usage (\eg~in liveness analaysis and linear logics) or past values in 
+stream-based data-flow programming.
 
 Our first contribution is the discovery of the connection between the above examples and
-their novel presentation -- in the form of calculi with type systems and formal semantics --
-that highlights this relationship. Our second and third contribution are two \emph{coeffect}
-calculi that capture commonalities in the presented examples. In particular, our 
-\emph{flat} coeffect calculus models languages with contextual properties of the execution 
-environment and our \emph{structural} coeffect calculus models languages where the contextual
-properties are attached to the variable usage.
+their novel presentation in the form of calculi (\emph{coeffect systems}). The presented type 
+systems and formal semantics highlight the relationship between different notions of context. 
+Our second and third contributions are two unified coeffect calculi that capture commonalities 
+in the presented examples. In particular, our \emph{flat coeffect calculus} models languages 
+with contextual properties of the execution environment and our \emph{structural coeffect 
+calculus} models languages where the contextual properties are attached to the variable usage.
 
-Although the focus of this thesis is on the syntactical properties of the presented 
+Although the focus of this thesis is on the syntactic properties of the presented 
 systems, we also discuss their category-theoretical motivation. We introduce the notion of
-an \emph{indexed} comonad (the dual of monads) and show how they provide semantics of the
-presented two calculi.
+an \emph{indexed} comonad (based on dualisation of the well-known monad structure) and show 
+how they provide semantics of the two coeffect calculi.
 

### 2014-09-11 text/Chapter01.tex
Tweaks at the beginning and lofty closing paragraph (Chapter 1)

@@
+%!TEX root = ../main.tex
+
 % ==================================================================================================
@@
 
-Many advances in programming language design are driven by some practical motivations. Sometimes,
-the practical motivations are easy to see -- for example, when they come from an external change
+Many advances in programming language design are driven by practical motivations. Sometimes,
+these practical motivations are easy to see -- for example, when they come from an external change
 such as the rise of multi-core processors. Sometimes, discovering the practical motivations is a
@@
 
-Before exploring the motivations leading to this thesis, we briefly consider two recent practical
-concerns that led to the development of new programming languages. This helps to explain why
-context-aware programming is of importance. The examples are by no means representative, but they
-illustrate different kinds of motivations well.
+Before exploring the motivations for to this thesis, we briefly consider two recent practical
+concerns that have led to the development of new programming languages. This helps to explain why
+context-aware programming is important. The examples are by no means representative, but they
+illustrate various kinds of motivations well.
 
@@
 The rise of multi-core CPUs is a clear example of an external development influencing
-programming language research. As multi-core and multi-processor systems became de-facto standard,
+programming language research. As multi-core and multi-processor systems became ubiquitous,
 languages had to provide better abstractions for parallel programming. This led to the industrial
@@
 abstractions, such as threads and locks, is difficult and error-prone. At the same time,
-multi-core CPUs become the standard very quickly and so the lack of good language abstractions
+multi-core CPUs became a standard very quickly and so the lack of good language abstractions
 was apparent.
@@
 
-Accessing data is an example of a more subtle challenge. Initiatives like open government data\footnote{
-In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
+Accessing ``big data'' sources is an example of a more subtle challenge. Initiatives like open 
+government data\footnote{In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
 certainly make more data available. However, to access the data, one has to parse CSV and Excel files,
@@
 Technologies like LINQ \cite{app-linq} make querying data significantly easier. But perhaps
-because accessing data became important more gradually, it was not easy to see that inline
-SQL is a poor solution \emph{before} better approaches were developed.
+because accessing data became important more gradually, it was not easy to see that SQL queries,
+embedded as parameterized strings\footnote{The dominant approach is demonstrated, for example, 
+by a review of SQL injection prevention techniques by Clarke \cite{app-sql-injection}}, are a 
+poor solution \emph{before} better approaches were developed.
 
@@
 It is not easy to see the limitations of standard techniques (using HTTP requests to query REST
-services or parsing CSV files and using string-based lookup) until one sees how type providers
-change the data-scientist's workflow\footnote{This is difficult to explain in writing and so the
-reader is encouraged to watch a video showing type providers for the WorldBank and CSV data
-sources \cite{app-fsharp-world}.}.
+services or parsing CSV files and using string-based lookup) until one sees just how much type 
+providers change the data-scientist's workflow\footnote{This is difficult to explain in writing 
+and so the reader is encouraged to watch a video showing type providers for the WorldBank and CSV 
+data sources \cite{app-fsharp-world}.}.
+
+~\\
 
@@
 
-In this chapter, we argue that the next important practical challenge for programming language
+In this thesis, we argue that the next important practical challenge for programming language
 designers is designing languages that are better at working with (and understanding) the
@@
 
-This challenge is of the kind that is not easy to see -- perhaps because we are so used to
+This challenge is of the kind that is not easy to see, perhaps because we are so used to
 doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
-uncover such flaws -- we look at a number of basic programs that rely on contextual
-information, we explain why they are inappropriate and then we briefly outline how this
-thesis remedies the situation.
+expose such flaws. We look at a number of basic programs that rely on contextual
+information, we explain why the currently dominant solutions are inappropriate and then 
+briefly outline how this thesis solves the problems.
 
@@
 more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and
-more \emph{unified}.
-These are exactly the aims that we follow in this thesis. In this chapter, we explain what
-the common problems are. In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural},
-we develop two simple calculi to understand and capture the structure of the problems and,
-finally, Chapter~\ref{ch:unified} unifies the two abstractions.
+more \emph{unified}. These are exactly the aims that we follow in this thesis. In this chapter, 
+we explain what the common problems in context-dependent programming are. In 
+Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we develop two simple calculi to 
+understand and capture the structure of those problems and, finally, Chapter~\ref{ch:unified} 
+unifies the two abstractions.
 
 % ==================================================================================================
+%                                                                                             
+%     ###                  #                    #                                              
+%    #   #                 #                    #                                              
+%    #       ###   # ##   ####    ###   #   #  ####           ###   #   #   ###   # ##    ###  
+%    #      #   #  ##  #   #     #   #   # #    #     #####      #  #   #      #  ##  #  #   # 
+%    #      #   #  #   #   #     #####    #     #             ####  # # #   ####  #      ##### 
+%    #   #  #   #  #   #   #  #  #       # #    #  #         #   #  # # #  #   #  #      #     
+%     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###  
+%                                                                                                                                                                       % ==================================================================================================
 
@@
 What notions of \emph{context} can be identified in modern software systems?
-Different environments provide different resources (\eg~database
-or GPS sensor), environments are increasingly diverse (\eg~multiple versions of different
-mobile platforms). Web applications are split between client, server and mobile components; mobile
+Different environments provide different resources (\eg~a database or GPS sensors), environments 
+are increasingly diverse (\eg~different mobile platforms with multiple partially incompatible versions). 
+Web applications are split between client, server and mobile components; mobile
 applications must be aware of the physical environment while the ``internet of things'' makes
-the environments even more heterogeneous. At the same time, applications access
-rich data sources and need to be aware of security policies and provenance information from the
-environment.
+the environment even more heterogeneous. At the same time, applications access
+rich data sources and need to be aware of provenance information and respect the security policies 
+from the environment.
 
@@
 \begin{itemize}
-\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple
-  targets (e.g. SQL \cite{app-linq}, OpenCL or JavaScript \cite{app-hop-lang}) a part of the compilation
-  (generating the SQL query) often occurs at runtime and developers have no guarantee that it will succeed
-  until the program is executed.
+\item \textbf{System capabilities.} Libraries such as LINQ \cite{app-linq} let developers write code
+  in a host language like C\# and then cross-compile it to multiple targets (including SQL, OpenCL 
+  or JavaScript \cite{app-hop-lang}). Part of the compilation (\eg~generating the SQL query) occurs 
+  at runtime and developers have no guarantee that it will succeed until the program is executed,
+  because only subset of the host language is supported.
 
@@
 \item \textbf{Security and provenance.} When working with data (be it sensitive database or social
-   network data), we have permissions to access only some of the data and we may want to track
+   network data), we may have permission to access only some of the data and we may want to track
    \emph{provenance} information. However, this is not checked -- if a program attempts to access
@@
 
+\noindent
 Most developers do not perceive the above as programming language flaws -- they are simply common
-programming problems (at most somewhat annoying and tedious) that had to be solved. However, this
-is because we do not realize that a suitable language extension could make the above problems
+programming problems (at most somewhat annoying and tedious) that have to be solved. However, this
+is because it is not apparent that a suitable language extension could make the above problems
 significantly easier to solve. As the number of distinct contexts and their diversity increases,
@@
 
-The following sub-sections explore four examples in more details. The examples are chosen to
-demonstrate two distinct forms of contexts that are studied in this thesis.
+The following sub-sections explore 4 examples in more detail. The examples are chosen to
+demonstrate two distinct forms of contexts that are studied in this thesis -- first two are
+related to the program environment and the latter two are associated with individual variables of 
+the program.
 
@@
 Equally, in the .NET ecosystem, there are multiple versions of the .NET runtime, mobile and
-portable versions of the framework etc. The differences may be subtle -- for example, some members
-are omitted to make the mobile version of the library smaller, some functionality is not available
-at all, but naming can also vary between versions.
+portable versions of the framework etc. The differences may be subtle -- for example, some instance
+methods and properties are omitted to make the mobile version of the library smaller, some functionality 
+is not available at all, but naming can also vary between versions.
 
@@
 on Android, programs can access API from higher level platform dynamically using techniques like
-reflection and writing wrappers. This is even more error prone. As noted in an article introducing
-the technique\footnote{Retrieved from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}:
-\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}. Again, it would be
-reasonable to expect that statically-typed languages could provide a better solution.
+reflection and writing wrappers. This is even more error prone. As noted in an article\footnote{Retrieved 
+from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}
+introducing the technique ``Remember the mantra: if you haven't tried it, it doesn't work.'' 
+Again, it would be reasonable to expect that statically-typed languages could provide a better solution.
 
@@
 written in a subset of a host language to some other target language, such as SQL, OpenCL or
-JavaScript. This is an important technique, because it lets developers targets multiple
-heterogeneous runtimes that have limited execution capabilities.
+JavaScript. For database access, this is a recently developed technique replacing embedded SQL
+discussed in the introduction, but it is a more generally important technique for programming
+in heterogeneous environments. It lets developers targets multiple runtimes that have limited 
+execution capabilities.
 
-For example, consider the following LINQ query written in C\# that queries a database and
-selects product names where the first upper case letter is "C":
+For example, the following LINQ query written in C\# queries a database and
+selects those product names where the first upper case letter is ``C'':
 %
@@
 \kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
-\kvd{where}~\ident{p.ProductName.First}(\ident{c} \Rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
+\kvd{where}~\ident{p.ProductName.First}(\lambda \ident{c} \rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
 \kvd{select}~\ident{p.ProductName};
@@
 %
-This appears as a perfectly valid code ant the C\# compiler accepts it. However, when the program
+This appears as a perfectly valid code and the C\# compiler accepts it. However, when the program
 is executed, it fails with the following error:
@@
 The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
-uses \ident{First} method to iterate over characters of a string, which is not supported. This is
+uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
 not a technical limitation of LINQ, but a fundamental problem of the approach.
@@
 The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
-Google search returns 11800 results for the message above and even more (44100 results) for a
-LINQ error message \emph{``Method X has no supported translation to SQL``} caused by a similar
+Google search returns 11800 results for the message above and even more results (44100) for a
+LINQ error message \emph{``Method X has no supported translation to SQL''} caused by a similar
 limitation.
@@
 
-The previous two examples were related to the non-existence of some library functions in another
-environment. Another common factor was that they were related to the execution context of the whole
-program or a scope. However, contextual properties can be also related to specific variables.
+The previous two examples were related to the non-existence of some library functions in a different
+execution environment. Another common factor was that they were related to the execution context of 
+the whole program or a function scope. However, contextual properties can also be associated with
+a specific variables.
 
-For example, consider the following code sample that accesses database by building a SQL query
-using string concatenation:
+For example, consider the following code sample that accesses a database by building a SQL query
+using string concatenation. For the purpose of the demonstration, this example does not use LINQ, 
+but an older approach with a parameterized SQL query written as a string:
 %
@@
 The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
-(an attacker could enter \str{'; DROP TABLE Products -{}-} as their \ident{name} and delete the
-database table with products). For this reason, most libraries discourage building SQL commands by
-string concatenation, but there are still many systems that do so.
+\cite{app-sql-injection} (an attacker could enter \str{'; DROP TABLE Products -{}-} as their 
+\ident{name} and delete the database table ``Products''). For this reason, most libraries discourage 
+building SQL commands by string concatenation, but there are still many systems that do so.
 
@@
 meta-data about variables that are in some ways special. Such meta-data can determine how the variables
-can be used. Here, \ident{name} comes from the user input. This \emph{provenance} information should
+can be used. Here, \ident{name} comes from the user input. This information about the value should
 be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
-not directly contain user input (in an unchecked form). Such marking of values (but at run-time)
-is also called tainting \cite{app-tainting-sql}.
+not directly contain user input (in an unchecked form).
 
@@
 
-In another context, when working with data (\eg~in data journalism), it would be desirable to
-track meta-data about the quality and the source of the data. For example, is the source trustworthy?
-Is the data up-to-date? Such meta-data could propagate to the result and tell us important
-information about the calculated results.
+In the security context, such marking of values (but at run-time) is called \emph{tainting} 
+\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
+tracking. This can be useful when working with data in other contexts. For example, 
+data jounralsts might want to propagate meta-data about the quality and the information source -- 
+is the source trustworthy? Is the data up-to-date? Such meta-data could propagate to 
+the result and tell us important information about the calculated results.
 
@@
 
-The last example leaves the topic of cross-platform and distributed computing. We focus on
+The final example leaves the topic of cross-platform and distributed computing. We focus on
 checking how arrays are accessed. This is a simpler version of the data-flow programming
@@
 
-Consider a simple programming language with arrays where $n^{\textnormal{th}}$ element of an
-array \ident{arr} is accessed using $\ident{arr}[n]$. Furthermore, we focus on performing local
-transformations and we assume that the keyword $\kvd{cursor}$ returns the \emph{current} location
-in the array.
+Consider a simple programming language with arrays where $n^{\mathit{th}}$ element of an
+array \ident{arr} is accessed using $\ident{arr}[n]$. We focus on writing stencil computations
+(such as image blurring, Conway's game of life or convolution) where all arrays are of
+the same size and the system provides a \emph{cursor} pointing to a current location in the 
+stencil. We assume that the keyword \kvd{cursor} returns the current location in the stencil.
 
-The following example implements a simple one-dimensional cellular automata, reading from the
+The following example implements a simple one-dimensional cellular automaton, reading from the
 \ident{input} array and writing to \ident{output}:
@@
 while \ident{output} is accessed in the range $\langle 0, 0 \rangle$. When calculating the ranges,
-we need to be able to compose ranges $\langle-1,-1$, $\langle 0,0 \rangle$ and $\langle 1,1 \rangle$
-(based on the accesses on the first line).
+we need to be able to compose ranges $\langle-1,-1 \rangle$, $\langle 0,0 \rangle$ and $\langle 1,1 \rangle$
+(based on the three accesses on the first line).
 
 The information about access patterns can be used to efficiently compile the computation
-(as we know which sub-range of the array might be accessed) and it also allows better handling
-of boundaries. For example, wrap-around behaviour we could pad the \ident{input} with a known
-number of elements from the other side of the array.
+by preallocating the necessary space (as we know which sub-range of the array might be accessed).
+It also allows better handling of boundaries. For example, to simplify wrap-around behaviour we 
+could pad the \ident{input} with a known number of elements from the other side of the array.
 
+% ==================================================================================================
+%
+%    #                                                             
+%    #                                                             
+%    #       ###   # ##    ## #  #   #   ###    ## #   ###    ###  
+%    #          #  ##  #  #  #   #   #      #  #  #   #   #  #     
+%    #       ####  #   #   ##    #   #   ####   ##    #####   ###  
+%    #      #   #  #   #  #      #  ##  #   #  #      #          # 
+%    #####   ####  #   #   ###    ## #   ####   ###    ###   ####  
+%                         #   #                #   #               
+%                          ###                  ###                
+%
 % ==================================================================================================
@@
 
-The four examples presented in the previous section cover different notions of \emph{context}.
-The context can be viewed as execution environment, capabilities provided by the environment
-or input and meta-data about the input.
+The four examples presented in the previous section cover different kinds of \emph{context}.
+The context includes notions such as execution environment, capabilities provided by the 
+environment or input and meta-data about the input.
 
-The different notions of context can be broadly classified into two categories -- those that
+The different applications can be broadly classified into two categories -- those that
 speak about the environment and those that speak about individual inputs (variables). In this
@@
    Examples include resources such as GPS sensors and battery status (on a phone), databases
-   (on the server), or framework version.
+   (on the server), or software framework (or library) version.
 
@@
 
-\subsection{Context-aware language in action}
+\subsection{Context-aware languages in action}
 \label{sec:intro-context-example}
 
-As an example, consider a news reader consisting of server-side (which stores the news in a
-database) and a number of clients applications for popular platforms (Android, Windows Phone, etc.).
-A simplified code excerpt that might appear somewhere in the implementation is shown in
+As an example, consider a news reader app consisting of a server-side component (which stores the 
+news in an SQL database) and a number of clients applications for popular platforms (Android, Windows 
+Phone, etc.). A simplified code excerpt that might appear somewhere in the implementation is shown in
 Figure~\ref{fig:introduction-sample}.
@@
 
-If we were writing the code using current main-stream technologies, we would have to create three
+If we were writing the code using current mainstream technologies, we would have to create three
 completely separate components. The server-side would include the \ident{fetchNews} function, which
@@
 \kvd{let}~\ident{iPhoneMain}() = \\[-0.25em]
-\hspace{1em} \ident{createCocoaListing}(\ident{fetchLocalNews})\\[-0.25em]
+\hspace{1em} \ident{createiPhoneListing}(\ident{fetchLocalNews})\\[-0.25em]
 \\[-1.0em]
 \kvd{let}~\ident{windowsMain}() = \\[-0.25em]
-\hspace{1em} \ident{createMetroListing}(\ident{fetchLocalNews})
+\hspace{1em} \ident{createWindowsListing}(\ident{fetchLocalNews})
 \end{array}
@@
 
+\noindent
 Here, we use the notation $\tau_1 \,\ident{\hspace{-1em}\footnotesize{@}}\, c \rightarrow \tau_2$ to denote
@@
 required resources or capabilities. However, a more complex structure could be used as well, for example,
-including the Android API level.
+including the Android API level as an integer.
 
 The following summary shows the types of the functions from the code sample in
-Figure~\ref{fig:introduction-sample}. These guide the code generation by specifying which
+Figure~\ref{fig:introduction-sample}. These guide code generation by specifying which
 function should be compiled for which of the platforms, but they also provide documentation
-for the developers:
-
+for the developers. In addition to function annotations, we also show the annotation attached
+to the \ident{password} variable:
+%
 \begin{equation*}
@@
  \ident{fetchLocalNews} &:& $\introfunc{location}{gps, rpc}{news~list}$\\[0.5em]
-\ident{iPhoneMain} &:& $\introfunc{unit}{ cocoa, gps, rpc}{unit}$\\[-0.25em]
+ \ident{iPhoneMain} &:& $\introfunc{unit}{ ios, gps, rpc}{unit}$\\[-0.25em]
  \ident{windowsMain} &:& $\introfunc{unit}{ windows, gps, rpc}{unit}$
@@
 \end{equation*}
+%
+The example combines two separate notions of context. The variable \ident{password} is annotated 
+with a single (per-variable) annotation specifying tainting while
+functions are annotated with a set of resource requirements.
 
-As mentioned earlier, the concrete syntax used here is just for illustration. Furthermore,
+The concrete syntax used here is just for illustration. Furthermore,
 some information could even be mapped to other visual representations -- for example, differently
@@
 \item We can compile the program for two client-side platforms - the entry points are \ident{iPhoneMain}
-  and \ident{windowsMain} and require Cocoa and Windows user-interface libraries, together with
+  and \ident{windowsMain} and require iOS and Windows user-interface libraries, together with
   GPS and the ability to perform remote calls over the network.
@@
 languages. This section outlines how can contextual information be integrated into the
-standard framework of static typing. This section is intended only as an informal overview and the
-related work is discussed in Chapter~\ref{ch:pathways}.
+standard framework of static typing. This section is intended only as an informal overview and 
+complete review of related work is available in Chapter~\ref{ch:pathways}.
 
@@
 The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
-expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int}
-or \ident{float}, depending on the types of the variables. For example, the following is a valid
-typing judgement in F\#:
+expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
+\ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
+the following is a valid typing judgement in F\#:
 %
@@
 This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
-be \ident{int}. The expression would also be typeable in a context $x\!:\!\ident{int},~y\!:\!\ident{int}$,
-but not, for example, in a context where $x$ has a type \ident{unit}.
+be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
+but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.
 
@@
 the type would be $\ident{IO unit}$ meaning that the result is a \emph{computation} that performs I/O
-effects and then returns \ident{unit} value.
-
-Another option for tracking effects is to extend the judgement with additional information
-about the effects. The judgement in a language with effect system would look as follows:
+effects and then returns \ident{unit} value. Here, we look at another option for tracking effects, 
+which is to extend the judgement with additional information about the effects. The judgement in a 
+language with effect system would look as follows:
 %
 \begin{equation*}
-x\!:\!\ident{int} \vdash \ident{print}~x : \ident{unit~\&\aclrd{\{ console \}}}
+x\!:\!\ident{int} \vdash \ident{print}~x : \ident{unit}~\&~\aclrd{\{ \ident{console} \}}
 \end{equation*}
@@
 
-The crucial part of type systems is dealing with different forms of composition. For example, assume
+The crucial part of type systems is dealing with different forms of composition. Assume
 we have a function \ident{read} that reads from the console and a function \ident{send} that sends
-data over the network. In that case, the type system should correctly infer that the effects of an
+data over the network. The type system should correctly infer that the effects of an
 expression $\ident{send}(\ident{read}())$ are $\{ \ident{console}, \ident{network} \}$.
@@
 \begin{array}{l}
- ( \ident{password}\!:\!\ident{string \hspace{-1em}{\footnotesize @ \aclrd{sensitive}}},
-   \ident{cmd}\!:\!\ident{string})
- \hspace{-0.5em}
- \ident{\footnotesize @ \aclrd{\{ database \}}} \vdash \\[-0.1em]
- \hspace{3em} \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
+ \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
+   \ident{cmd}\!:\!\ident{string})}{ \cclrd{\{ \ident{database} \}}}  \\[-0.1em]
+ \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
 \end{array}
@@
 Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
-context $\Gamma$. In other words, coeffect systems track more detailed information about the context,
-not just the available variables. In the above example, it tracks meta-data about the variables
-and annotates \ident{password} as sensitive. Furthermore, it tracks requirements about the execution
-environment -- for example, that the execution requires an access to database.
+context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
+in the context -- they also capture detailed information about the execution environment.
+In the above example, the system tracks meta-data about the variables and annotates \ident{password} 
+as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
+that the execution requires an access to database.
 
@@
 
-As mentioned earlier, it is well-known fact that \emph{effects} correspond to \emph{monads} and
-languages such sa Haskell use monads to provide a limited form of effect system. An interesting
-observation made in this thesis is that \emph{coeffects}, or systems for tracking contextual
-information, correspond to the category theoretical dual of monads called \emph{comonads}. The
-details are explained when discussing the semantics of coeffects throughout the thesis.
+It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell 
+use monads to provide a limited form of effect system. An interesting observation made in this thesis 
+is that \emph{coeffects}, or systems for tracking contextual information, correspond to the category 
+theoretical dual of monads called \emph{comonads}. The details are explained when discussing the 
+semantics of coeffects throughout the thesis.
 
+% ==================================================================================================
+%
+%     ###           #      ##      #                 
+%    #   #          #       #                        
+%    #   #  #   #  ####     #     ##    # ##    ###  
+%    #   #  #   #   #       #      #    ##  #  #   # 
+%    #   #  #   #   #       #      #    #   #  ##### 
+%    #   #  #  ##   #  #    #      #    #   #  #     
+%     ###    ## #    ##    ###    ###   #   #   ###  
+%
 % ==================================================================================================
@@
 The key claim of this thesis is that programming languages need to provide better ways of capturing
-how programs rely on the context in which they execute. This chapter shows why this is an important
-problem. We looked at a number of properties related to context that are currently handled in
-ad-hoc and error-prone ways. Next, we considered the properties in a simplified, but realistic
-example of a client/server application for displaying local news.
+how programs rely on the context, or execution environment, in which they execute. This chapter 
+shows why this is an important problem. We looked at a number of properties related to context that 
+are currently handled in ad-hoc and error-prone ways. Next, we considered the properties in a 
+simplified, but realistic example of a client/server application for displaying local news.
 
@@
 because we are so used to write code in certain ways that prevent us from seeing the flaws. The
-purpose of this chapter was to uncover the flaws and convince the reader that there should be a
+purpose of this chapter was to expose the flaws and convince the reader that there should be a
 better solution. Finding the foundations of such better solution is the goal of this thesis:
@@
   that the idea of context-aware computations can be naturally approached from a number of
-  directions developed recently in theories of programming languages. Chapter~\ref{ch:applications}
-  follows by showing practical motivation for coeffects -- we look at a number of systems that
-  can be captured using the systems developed later.
+  directions developed recently in theories of programming languages (including type and effect
+  systems, categorical semantics and sub-structural logics).
 
-\item In Chapter~\ref{ch:pathways} and Chapter~\ref{ch:applications}, we present the key novel
+\item Chapter~\ref{ch:applications} presents the first contribution of the thesis -- the discovery
+  of the connection between a number of existing programming language features that are related
+  to context. The chapter presents type systems and semantics for a number of systems and analyses 
+  (including data-flow, liveness analysis, distributed programming and Haskell's type classes).
+  Our novel presentation reveals their similarity.
+
+\item In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we present the key
   contributions of this thesis. We develop the \emph{flat} and \emph{structural} calculi, show
   how they capture important contextual properties and develop their categorical semantics using
-  a notion based on comonads. Chapter~\ref{ch:unified} links the two systems into a single
-  formalism that is capable of capturing both flat and structural properties.
+  a notion based on comonads. Chapter~\ref{ch:unified}  links the two systems using a single 
+  formalism that is capable of capturing both flat and structural properties and also discusses
+  an alternative presentation that is more suitable for automatic type inference of coeffects.
 
-\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis, but
-  one important direction deserves further exploration. In Chapter~\ref{ch:meta}, we look at
-  a different approach to tracking contextual information that arises from modal logics.
-  Finally, Chapter~\ref{ch:conclusions} discusses approaches for implementing the presented theory
-  in main-stream programming languages and concludes.
+\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis. 
+  One important direction deserves further exploration, and so Chapter~\ref{ch:conclusions}
+  starts with a brief discussion of a different approach to tracking contextual information that 
+  arises from modal logics. Finally, the rest of Chapter~\ref{ch:conclusions} discusses approaches 
+  for implementing the presented theory in main-stream programming languages and concludes.
 \end{itemize}
+
+\noindent
+If there is a one thing that the reader should remember from this thesis, it is the fact that
+there is a unified notion of \emph{context}, capturing many common scenarios in programming, 
+and that programming language designers need to provide ways for working with this context (using 
+\emph{coeffects} or not). This greatly reduces the number of distinct concepts that software 
+developers need to keep in mind of when building applications for the rich and diverse 
+execution environments of the future.

### 2014-09-11 text/Chapter04.tex
Examples of structural comonads (Chapter 5)
When is coeffect a monad (Chapter 4)

@@
 \ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
-\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& (a, b)\\
+\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
 \end{array}
@@
 \ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
-\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
+\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
 \\[0.45em]
@@
 \\[0.45em]
-\ident{lift}_{n', n} \langle a_0, \ldots, a_n' \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
+\ident{lift}_{n', n} \langle a_0, \ldots, a_{n'} \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
 \quad \langle a_0, \ldots, a_n \rangle
@@
 
-\newcommand{\ctx}{\textit{ctx}}
 \begin{figure*}[t]
@@
 \section{Related work}
-\label{sec:flat-related-monads}
-
-Most of the related work has been already discussed in Chapter~\ref{ch:pathways} or throughout
-the chapter. In this section, we look in more details at the relationship between coeffects,
-comonads and monads.
-
-\subsection{When coeffect is not a monad}
+\label{sec:flat-related}
+
+Most of the related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
+and we covered work related to individual concepts throughout the chapter. In this section, we
+do not repeat the discussion present elsewhere -- we discuss one specific question that often 
+arises when discussing coeffects and that is \emph{when is coeffect (not) an effect?}
+
+We start with a quick overview of the ways in which effects and coeffects differ and then
+we briefly look at one (but illustrative) example where the two concepts overlap. We focus 
+mainly on the equivalence between the \emph{categorical semantics}, which reveals the nature
+of the computations -- rather than considering just the syntactic aspects of the type system.
+
+\subsection{When is coeffect not a monad}
 Coeffect systems differ from effect systems in three important ways:
@@
 
-\subsection{When coeffect is a monad}
-As discussed in \ref{sec:applications-flat-impl} ...
-
-Product comonad
-
-$C_{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma)$
-
-Reader monad
-
-$M_{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau$
-
+\subsection{When is coeffect a monad}
+\label{sec:flat-related-monads}
+
+As discussed in Section~\ref{sec:applications-flat-impl}, one of our examples, implicit 
+parameters, can be also captured by a monad. However, \emph{just} a monad is not enough
+because lambda abstraction in effect systems does not provide a way of splitting the context
+requirements between declaration-site and call-site (or combining the implicit parameters
+available in the scope where the function is deined and those specified by the caller).
+
+\paragraph{Categorical relationship.}
+Before looking at the necessary extensions, consider the two ways of modelling implicit 
+parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
+for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
+are:
+%
+\begin{equation*}
+\begin{array}{lll}
+ \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
+ \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
+\end{array} 
+\end{equation*}
+%
+The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
+the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
+As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the 
+two representations are equivalent:
+%
+\begin{remark}
+Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
+are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using
+the currying/uncurrying isomoprhism.
+\end{remark}
+\begin{proof}
+The isomorphism is demonstrated by the following equation:
 \begin{equation*}
@@
  \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) \\
- \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \\
-\end{array}
-\end{equation*}
-
-The difference between effects and coeffects is mainly in the (\emph{abs}) rule. While the 
-semantic model (monads vs. comonads) is very different, we can consider extending the two to 
-obtain equivalent syntactic rules. To allow splitting of implicit parameters in lambda abstraction, 
-the reader monad needs an operation that eagerly performs some effects of a function:
-
-\begin{equation*}
-(\tau_1 \rightarrow \mtyp{\cclrd{r} \cseq \cclrd{s}}{\tau_2}) \rightarrow 
-  \mtyp{\cclrd{r}}{(\tau_1 \rightarrow \mtyp{\cclrd{s}}{\tau_2})}
-\end{equation*}
-
-To obtain a pure lambda abstraction
-for coeffects, we need to restrict the $\cpar_{r, s}$ 
-operation of indexed comonads, so that the first parameter is annotated with $\cunit$ (meaning
-no effects): $\C^{\cunit} A \times \C^r B \rightarrow \C^{r}(A \times B)$.
-
-\begin{equation*}
-\begin{array}{rcl}
- \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
- \sem{\lambda x.e} &=& \ident{unit}~(\lambda x. \sem{e}) \\
-\end{array}
-\end{equation*}
+ \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
+\end{array}
+\end{equation*}
+\end{proof}
+
+\noindent
+The equivalence holds for monads and comonads (as well as \emph{indexed} monads
+and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
+the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging.
+
+\paragraph{Delaying effects in monads.} 
+In the syntax of the language, the above difference is manifested in the (\emph{abs}) rules for 
+monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
+side-by-side, using the coeffect notation for both of them:
+%
+\begin{equation*}
+\tyrule{cabs}
+  { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \vdash e : \tau_2 }
+  { \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\quad
+\tyrule{mabs}
+  { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \vdash e : \tau_2 }
+  { \coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 }
+\qquad
+\end{equation*}
+%
+In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
+the monadic rule (\emph{mabs}) places all requirements on the call-site. This follows from the
+fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
+%
+\begin{equation*}
+\sem{\lambda x.e} \;=\; \ident{unit}~(\lambda x. \sem{e})
+\end{equation*}
+%
+The type of $\ident{unit}$ is $\alpha\rightarrow\mtyp{\alpha}{\cclrd{\emptyset}}$, but in this specific
+case, the $\alpha$ is instantiated to be $\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}$ and so this
+use of \ident{unit} has a type:
+%
+\begin{equation*}
+\ident{unit}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{\emptyset}}(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2})
+\end{equation*}
+%
+In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
+the declaration-site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
+call-site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
+(which we call \ident{delay}) with the following signature:
+%
+\begin{equation*}
+\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{r}}(\tau_1\rightarrow\mtyp{\cclrd{s}}{\tau_2})
+\end{equation*}
+%
+The operation reveals the difference between effects and coeffects -- intuitively, given a function
+with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
+function, \emph{before} the function actually perforsm the effectful operation with the effects.
+The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
+from the effect annotation of the body.
+
+Another important aspect of the signature is that the function needs to be indexed by the coeffect
+annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context requirements
+$\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function.
+
+The operation cannot be implemented in a useful way for most standard monads, but the 
+reader monad is, indeed, an exception. It is not difficult to see how it can be implemented
+when we expand the definitions of $\mtyp{\cclrd{r}}\tau$:
+%
+\begin{equation*}
+\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~
+(\tau_1\rightarrow(\cclrd{r \cup s} \rightarrow \sigma) \rightarrow \tau_2) \rightarrow 
+((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_1\rightarrow(\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2)
+\end{equation*}
+
+\paragraph{Restricting coeffects in comonads.} 
+As just demonstrated, we can extend monads so that the reader monad is capable of capturing
+the semantics of implicit parameters, including the splitting of implicit parameter requirements
+in lambda abstraction. Can we also go the other way round and \emph{restrict} the comonadic
+semantics so that all requirements are delayed as in the (\emph{mabs}) rule, thus modelling 
+fully dynamically scoped parameters?
+
+This is, indeed, also possible. Recall that the semantics of lambda abstraction in the flat
+coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
+two contexts (wrapped in a comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried values and
+additional contextual information (implicit parameters). To obtain the (\emph{mabs}) rule, we 
+can restrict the first parameter, which corresponds to the declaration-site context:
+%
+\begin{equation*}
+\begin{array}{rcll}
+ \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{r}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{r \cup s}}(\alpha \times \beta)
+ &\hspace{4em}(\textit{normal}) \\
+ \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{\emptyset}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{s}}(\alpha \times \beta)
+ &\hspace{4em}(\textit{restricted})
+\end{array} 
+\end{equation*}
+%
+In the (\emph{restricted}) version of the operation, the declaration-site context requires
+no implicit parameters and so all implicit parameters have to be satisfied by the call-site.
+The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.
+
+The idea of restricting the operations of the coeffect calculus semantics could be used more
+generally. We could allow any of the coeffect algebra operations $\cseq, \czip, \cpar$ to be
+\emph{partial} and thus the restricted (fully dynamically-scoped) version of implicit parameters
+could be obtained just by changing the definition of $\czip$. Similarly, we could obtain \eg~a
+fully lexically-scoped version of the system. Similar idea has been used for the
+semantics of effectful computations by Tate~\cite{effects-producer-semantics}.
+
 % ==================================================================================================
-
+%                                                                             
+%        ###                         ##                    #                        
+%       #   #                         #                                             
+%       #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###  
+%       #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #     
+%       #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###  
+%       #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      # 
+%        ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####  
+%                                                                             
+% ==================================================================================================                                                                             
 \section{Conclusions}
 
-We examined three simple calculi with associated
-static analyses (liveness analysis, implicit parameters, and dataflow
-analysis). These were unified in the \emph{coeffect calculus},
-providing a general coeffect system parameterised by an
-algebraic structure describing the propagation of context
-requirements throughout a program.
-
-We model the semantics of coeffect calculus using \emph{indexed comonad} -- a novel structure, which
-is more powerful than (monoidal) comonads. Indices of the indexed comonad operations manifest the 
-semantic propagation of context such that the propagation of information in the general coeffect
-type system corresponds exactly to the semantic propagation of context in our categorical model.
-
-We consider the analysis of context to be essential, not least for the examples here but 
-also given increasingly rich and diverse distributed systems.
+This chapter presented the \emph{flat coeffect calculus} -- a unified system for tracking 
+contextual properties that are \emph{whole-context}, meaning that they are related to the 
+execution environment or the enitre context in which computations are executed.
+This is the first of the two \emph{coeffect calculi} developed in this thesis.
+
+The flat coeffect calculus is parameterized by a \emph{flat coeffect algebra} that captures
+the structure of the information tracked by the type system. We demonstrated how to 
+instantiate the system to capture three specific systems discussed earlier in 
+Chapter~\ref{ch:applications}, namely liveness, data-flow and implicit parameters.
+
+Next, we introduced the notion of \emph{flat indexed comonad}, which is a generalization of
+comonad, equipped with additional operations needed to provide categorical semantics of the 
+flat coeffect calculus. The indices of the flat indexed comonad operations correspond to the
+coeffect annotations in the type system and provide a foundation for the design of the 
+calculus.
+
+Finally, we discussed the equational theory for flat coeffect calculus. Although each 
+concrete instance of flat coeffect calculus models different notion of context, there are
+syntactic properties that hold for all flat coeffect systems satisfying certain additional
+conditions. In particular, two \emph{subject reduction} theorems prove that the operational
+semantics for two classes of flat coeffect calculi (including liveness and implicit parameters) 
+can be based on call-by-name reduction.
+
+In the upcoming chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
+properties to \emph{structural} coeffect calculi, tracking per-variable information, thus
+covering systems from the second half of Chapter~\ref{ch:applications}.

### 2014-09-19 text/Chapter04.tex
Finish conclusions and tweaks (Chapter 5)

@@
 For bottom-pointed coeffect algebra, consider the typing derivation for the term
-$(\lambda x.e)\,v$ before reduction:
+$(\lambda x.e_r)\,e_s$ before reduction:
 %
@@
 \begin{theorem}[$\eta$-expansion]
-\label{thm:cbn-substitution-bot}
 In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
@@
 \begin{proof}
-The following derivation shows that $\lambda x.e~x$ has the same type as $e$:
+The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
 \begin{equation*}
@@
   { \inference
-    { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
+    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
       \coctx{x:\tau_1}{\cunit} \vdash x:\tau_1 }
-    { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash e~x : \tau_2 } }
+    { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
   { \inference
     { \inference
-      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash e~x : \tau_2 }
-      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash e~x : \tau_2 } }
-    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.e~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
+      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
+    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
 \end{equation*}

### 2014-09-19 text/Chapter05.tex
Finish conclusions and tweaks (Chapter 5)

@@
 
-\chapter{Structural coeffect language} 
+\chapter{Structural coeffect calculus} 
 \label{ch:structural}
@@
 \begin{remark}[Let binding]
+\label{thm:structural-letbinding}
 In a structural coeffect calculus, the typing of $(\lambda x.e_2)~e_1$ is equivalent
@@
 properties of the flat coeffect algebra guarantee that certain equational properties hold for 
-all instances of the calculus.
+all instances of the calculus. In this section, we look at these common properties that we get
+``for free'' in all structural coeffect calculi.
 
@@
 When discussing equational theory for the flat calculus in Section~\ref{sec:flat-syntax}, we 
-noted that no single technique works universally for all flat coeffect calculi. We considered
-multiple different reductions that can be used as the basis for operational semantics for
-calculi satisfying different additional properties.
+noted that no single technique of specifying syntactic reduction works universally for all 
+flat coeffect calculi. We considered multiple different options (call-by-name, call-by-value, 
+internalized substitution) that can be used as the basis for operational semantics for
+different calculi that satisfy different additional properties.
 
@@
 \paragraph{Substitution for flat coeffects.} 
-The more interesting variant of the substitution lemma for flat coeffects 
+The less obvious (\emph{top-pointed}) variant of the substitution lemma for flat coeffects 
 (Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
@@
 \paragraph{Substitution for structural coeffects.}
-In contrast, the substitution lemma for structural coeffects can be proven ecause 
+In contrast, the substitution lemma for structural coeffects can be proven because 
 structural coeffect systems provide enough \emph{precision} to identify exactly with 
@@
 The following example shows a situation similar to the previous one -- here, we use
-structural data-flow calculus (with the \kvd{prev} construct to obtain previous
-value of an expression) and we substitute $w+z$ for $y$:
+structural data-flow calculus (writing $\kvd{prev}~e$ to obtain previous value of the 
+expression $e$) and we substitute $w+z$ for $y$:
 %
@@
 calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
-Furthermore, we may also need to substitute for multiple variables in the variable context.
-
-For example, the expression $\lambda x.x+x$ is type-checked by type-checking $x_1 + x_2$,
-contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction 
-$(\lambda x.x+x)~(y+z)$ we thus need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
+More importantly, because the system uses explicit contraction, we may also need to substitute 
+for multiple variables in the variable context at the same time.
+
+Consider the expression $\lambda x.x+x$. It is type-checked by type-checking $x_1 + x_2$,
+contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction of 
+$(\lambda x.x+x)~(y+z)$ we need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
 This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
-for example in the bunched typing system \cite{substruct-bunched}.
-To express the substitution lemma, we define the notion of a \emph{context with holes}
+such as the bunched typing system \cite{substruct-bunched}. To express the substitution 
+lemma, we define the notion of a \emph{context with holes}:
 
@@
 start with a variable followed by a context with $n$ holes. Note that the definition ensures that
-the location of variable holes correspond to the locations of coeffect annotation holes. Given a context
+the locations of variable holes correspond to the locations of coeffect annotation holes. Given a context
 with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
@@
 requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
-are annotated with $\cclrd{s} \aseq \aclrd{\textbf{t}}$, using the (scalar-vector extension)
-of the sequential composition operator $\cseq$.
+are annotated with $\cclrd{s} \,\aseq\, \aclrd{\textbf{t}}$, using the (scalar-vector extension)
+of the sequential composition operator $\cseq$. This suggests that, to evaluate each of the variables
+annotated with coeffects in the vector $\aclrd{\textbf{t}}$, we first need to evaluate the substituted
+expression with coeffects $\cclrd{s}$, followed by the rest of the expression with coeffects specified
+by $\aclrd{\textbf{t}}$.
 
 \begin{lemma}[Multi-nary substitution]
+\label{thm:structural-substitution}
 Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
@@
 
-\todo{Exterminate!}
-
-~
-
-Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
-$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
-$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
-the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
-Using the substitution lemma, we can demonstrate $\beta$-equality:
-%
-\begin{equation*}
-\inference{
-  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
-             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
-           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
-  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
-\equiv
-\subst{e_1}{x}{e_2} : \tau}
-\end{equation*}
-%%
-\noindent
-As a result, $\beta$-reduction preserves the type and coeffects of a term.
-This gives the following subject reduction property:
-
-\begin{theorem}[Subject reduction]
-In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
-$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
+In the previous chapter, we discussed call-by-value separately from call-by-name, because the
+proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
+$\beta$-reduction (local soundness), which encompasses both types of evaluation and 
+$\eta$-expansion (local completeness). Both of the properties hold for a system with any
+structural coeffect algebra.
+
+\paragraph{Reduction theorem}
+In a full $\beta$-reduction,
+written as $\rightarrow_\beta$, we replace the redex $(\lambda x.e_2)~e_1$ by the
+expression $\subst{e_r}{x}{e_s}$ anywhere inside a term. The subject reduction theorem
+guarantees that this does not change the type of the term.
+
+\begin{theorem}[$\beta$-reduction]
+In a structural coeffect system with a structural coeffect algebra formed by 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
+if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
+and $e \rightarrow_{\beta} e'$ using the full $\beta$-reduction then
+$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
 \end{theorem}
 \begin{proof}
-Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
+Consider the typing derivation for the redex $(\lambda x.e_r)\,e_s$ before the reduction
+(note that this is similar to the typing of let binding discussed in Remark~\ref{thm:structural-letbinding},
+but we do not swap the two parts of the free-variable context):
+%
+\begin{equation*}
+\inference
+  { \begin{array}{l}\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s\vspace{-2em}\end{array} &
+    \inference
+      { \coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r }
+      { \coctx{\Gamma_r}{\aclrd{\textbf{r}}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{t}} \tau_r} \vspace{0.2em} }
+  { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})} \vdash (\lambda x.e_r)~e_s : \tau_r }
+\end{equation*}
+%
+For the substitution lemma, we first rewrite the typing judgement for $e_r$, \ie~
+$\coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r$
+as a context with a single hole filled by the $x$ variable:
+$\coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes\,-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r$.
+Now we can perform the substitution using Lemma~\ref{thm:structural-substitution}:
+%
+\begin{equation*}
+\inference
+  { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r & 
+    \coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s }
+  { \inference
+      { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{\Gamma_s}{\cclrd{t}\,\aseq\,\aclrd{\textbf{s}} }\,] \vdash \subst{e_r}{x}{e_s} : \tau_r }
+      { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}}\,\atimes\,(\cclrd{t}\,\aseq\,\aclrd{\textbf{s}})} \vdash \subst{e_r}{x}{e_s} : \tau_r } }
+\end{equation*}
+%
+The last step applies the hole filling operation, showing that substitution preserves 
+the type of the term.
 \end{proof}
@@
 \noindent
+Because of the vector (free monoid) structure, coeffect annotations $\aclrd{\textbf{r}}$, 
+$\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$ are uniquely associated with $\Gamma_r$, 
+$\Gamma_s$, and $x$ respectively. Therefore, substituting $e_s$ (which has coeffects 
+$\aclrd{\textbf{s}}$) for $x$ introduces the context dependencies specified by $\aclrd{\textbf{s}}$
+which are composed with the existing requirements $\cclrd{t}$ on $x$.
+
+\paragraph{Expansion theorem}
 Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
-\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
-\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
-and application does not eliminate too much.
-%
-\begin{equation*}
-\inference{
+\emph{local soundness} and \emph{local completeness}. Informally, this means that 
+abstraction does not introduce too much, and application does not eliminate too much.
+
+\begin{theorem}[$\eta$-expansion]
+In a structural coeffect system with a structural coeffect algebra formed by 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
+if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
+and $e \rightarrow_{\eta} e'$ using the full $\eta$-reduction then
+$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
+\begin{equation*}
 \inference
-    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
-     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
-    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
-: \sigma \xrightarrow{\cclrd{s}} \tau}
-\end{equation*}
-%
-The last step uses the fact that 
+  { \inference
+    { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
+      \coctx{x:\tau_1}{\alift{\cunit}} \vdash x:\tau_1 }
+    { \coctx{\Gamma, x:\tau_1}{\aclrd{\textbf{r}} \;\atimes\;(\cclrd{s}\,\aseq\,\alift{\cunit}) } \vdash f~x : \tau_2 } }
+  { \inference
+    { \coctx{\Gamma, x:\tau_1}{\aclrd{\textbf{r}} \;\atimes\,\alift{\cclrd{s}} } \vdash f~x : \tau_2 }
+    { \coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+\end{equation*}
+%
+The second step uses the fact that 
 $\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
 arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
-
-This highlights another difference between coeffects and effects, as $\eta$-equality does
-not hold for many notions of effect. For example, in a language with output
-effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
-its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
-hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
-allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
-$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
-the exact immediate requirements of $\lambda x . e x$ to match those of $e$.
-
-\newpage
+\end{proof}
+
+\noindent
+The $\eta$-expansion property discussed in this section highlights another difference 
+between coeffects and effects. The $\eta$-equality property does not hold for many notions 
+of effect. For example, in a language with output effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ 
+has different effects to its $\eta$-converted form $\lambda x . e x$ because the immediate 
+effects of $e$ are hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, 
+the (\emph{abs}) rule allows immediate contextual requirements of $e$ to ``float outside'' 
+of the enclosing $\lambda$. Furthermore, the free monoid nature of $\atimes$ in structural 
+coeffect systems allows the exact immediate requirements of $\lambda x . e x$ to match 
+those of $e$.
+
+% ==================================================================================================
+%
+%     ###                         ##                    #                 
+%    #   #                         #                                      
+%    #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##  
+%    #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  # 
+%    #      #   #  #   #  #        #    #   #   ###     #    #   #  #   # 
+%    #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   # 
+%     ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   # 
+%
+% ==================================================================================================
 
@@
 
-\todo{ (...) }
+This chapter completes the key development of this thesis -- the presentation of unified calculi
+for context-aware computations that capture the motivating examples introduced in 
+Chapter~\ref{ch:applications}. In the previous chapter, we focused on the first category of 
+contet-aware computations and we developed \emph{flat coeffect calculus} that captures
+\emph{whole-context} properties. This chapter develops \emph{structural coeffect calculus},
+capturing \emph{per-variable} contextual properties. The system provides a precise 
+analysis of liveness and data-flow and allows other interesting uses such as tracking of variable
+accesses based on bounded linear logic.
+
+Following the structure of the previous chapter, the structural coeffect calculus is parameterized
+by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations 
+$\cseq$ and $\cpar$ that model sequential and point-wise composition, respectively. For flat 
+coeffects, we further require $\czip$ to model context merging. For structural coeffects, we 
+instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
+as $\czip$.
+
+In order to keep track of separate annotations for each variable, we use a system with 
+explicit structural rules (contraction, weakening and exchange) that manipulate the structure
+of variables and the structure of annotations at the same time.
+
+The chapter presents two technical results. The first is the semantics of structural coeffect
+calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
+it unifies our three example calculi (bounded reuse, data-flow and liveness) and it demonstrates
+suitability of the type system. The second technical result is equational theory for structural
+coeffect calculi. In particular, we show that $\beta$-reduction and $\eta$-expansion preserve
+the typing for any structural coeffect calculi. These two strong properties are desirable for
+programming languages, but are often not satisfied (\eg~by effectful langauges).
+
+Two questions remain opened for the upcoming chapter. Firstly, is there a way to present 
+flat and structural coeffect calculi in a uniform way as a single system? Secondly, can we
+simplify the definitions to make type inference easier and add other practical language
+aspects such as recursion?

### 2014-09-19 text/Chapter02.tex
Introduction in Chapter 4
Categorical semantics of flat calculus (Chapter 4)
Cleanup files, fix some invalid references

@@
+% "Language is a labyrinth of paths. You approach from one side and know your way about; you 
+% approach the same place from another side and no longer know your way about.”
+
 \chapter{Pathways to coeffects} 
@@
 
-We start by revisiting practical applications and existing language features that are related to 
-coeffects (Section~\ref{sec:path-apps}), then we look at coeffects as the dual of effect systems
-(Section~\ref{sec:path-eff}) and extend the duality to category theory, looking at the categorical 
-dual of monads known as \emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically 
-inspired type systems that are closely related to our structural coeffects 
-(Section~\ref{sec:path-logic}).
+We start by looking at coeffects as the dual of effect systems (Section~\ref{sec:path-eff}). Next,
+we extend the duality to category theory, looking at the categorical dual of monads known as 
+\emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically inspired type systems 
+that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).
 
@@
 
-As demonstrated by examples in the Chapter~\ref{ch:introduction}, this means that the resource
+As demonstrated by examples in the Chapter~\ref{ch:intro}, this means that the resource
 can be captured when a function is declared (e.g.~when it is constructed on the server-side
@@
 \subsection{Effectful languages and meta-languages}
+\label{sec:path-sem-langs}
 
@@
 
+\todo{Maybe remove this from here, we repeate it in section 4}
+
 \begin{definition}
@@
 \end{compactitem}
-such that, for all $f:\alpha \rightarrow \mtyp{}{\beta}, g:\beta \rightarrow \mtyp{}{\gamma}$:
+such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
 \begin{align}
@@
 \tag{\emph{associativity}}
-  \ident{cobind}~(\ident{cobind}~g \circ f) &= (\ident{cobind}~f) \circ (\ident{cobind}~g)
+  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
 \end{align}
 \end{definition}
+
+\todo{The following text is duplicated in chapter 4}
 

### 2014-09-19 text/Chapter03.tex
Introduction in Chapter 4
Flat type system (Chapter 4)
Calculus and semantics work (Chapter 5)
Examples of structural comonads (Chapter 5)
When is coeffect a monad (Chapter 4)
Cleanup files, fix some invalid references

@@
 
-%===================================================================================================
+% ===================================================================================================
+%	                                                               
+% 	  ###    #                           #                         
+% 	 #   #   #                           #                         
+% 	 #      ####   # ##   #   #   ###   ####   #   #  # ##    ###  
+% 	  ###    #     ##  #  #   #  #   #   #     #   #  ##  #  #   # 
+% 	     #   #     #      #   #  #       #     #   #  #      ##### 
+% 	 #   #   #  #  #      #  ##  #   #   #  #  #  ##  #      #     
+% 	  ###     ##   #       ## #   ###     ##    ## #  #       ###  
+%
+% ===================================================================================================
 
@@
   { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2}
-  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 1}
+  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2}
 \;\;
@@
 
-%===================================================================================================
+% ===================================================================================================
+%	                            
+% 	 #####   ##            #    
+% 	 #        #            #    
+% 	 #        #     ###   ####  
+% 	 ####     #        #   #    
+% 	 #        #     ####   #    
+% 	 #        #    #   #   #  # 
+% 	 #       ###    ####    ##  
+%	                            
+% ===================================================================================================
 
 \section{Flat coeffect systems}
+\label{sec:applications-flat}
 
@@
 \subsection{Implicit parameters and type classes} 
+\label{sec:applications-flat-impl}
 
@@
 
-As we discuss later in Section~3.?, the reader monad can be extended to model rebinding.
-However, later examples in this chapter, such as liveness in Section~\ref{sec:applications-flat-live}
-show that other context-aware computations cannot be captured by \emph{any} monad.
+As we discuss later in Section~\ref{sec:flat-related-monads}, the reader monad can be extended 
+to model rebinding. However, later examples in this chapter, such as liveness in 
+Section~\ref{sec:applications-flat-live} show that other context-aware computations 
+cannot be captured by \emph{any} monad.
 
@@
 environment -- this approach is similar to coeffects, both from the practical perspective, but also 
-through deeper theoretical links. We return to this topic in Chapter~\ref{ch:meta}.
+through deeper theoretical links. We return to this topic in Section~\ref{sec:conclusions-meta} of
+the final chapter.
 
@@
 only if it was used in both of the branches (this is sometimes called \emph{neededness}).
-
 The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
 interaction between contextual properties and certain language constructs such as conditionals.
-We discuss this in Section~\ref{sec:flat-conditionals}
 
@@
 % ==================================================================================================
+%	                                                                      
+% 	  ###    #                           #                           ##   
+% 	 #   #   #                           #                            #   
+% 	 #      ####   # ##   #   #   ###   ####   #   #  # ##    ###     #   
+% 	  ###    #     ##  #  #   #  #   #   #     #   #  ##  #      #    #   
+% 	     #   #     #      #   #  #       #     #   #  #       ####    #   
+% 	 #   #   #  #  #      #  ##  #   #   #  #  #  ##  #      #   #    #   
+% 	  ###     ##   #       ## #   ###     ##    ## #  #       ####   ###  
+%	                                                                      
+% ==================================================================================================
 
 \section{Structural coeffect systems}
+\label{sec:applications-structural}
 
@@
   structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
-  explored in Chapter~\ref{ch:meta}.
+  explored in Section~\ref{sec:conclusions-meta}.
 \end{itemize}
@@
 \subsection{Bounded variable use}
+\label{sec:applications-struct-bll}
 
@@
 treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
-semantics} and \emph{meta-language} and also revisit the BLL example in 
-Chapter~\ref{ch:meta}.
+semantics} and \emph{meta-language} in Section~\ref{sec:conclusions-meta}.
 
@@
 \paragraph{Calculus of capabilities.}
+\label{sec:applications-active-ccc}
+
 Crary et al. \cite{app-capabilities} introduced the Calculus of Capabilities to provide 

### 2014-09-19 text/Chapter06.tex
Start working on Chapter 4
Calculus and semantics work (Chapter 5)
Cleanup files, fix some invalid references

@@
+%!TEX root = ../main.tex
+
 \chapter{Unified coeffect language} 
@@
 
+%===================================================================================================
+
+\section{Introduction}
+
+%===================================================================================================
+
+\section{Semi-lattice formulation}
+\label{sec:unified-semilattice}
+
+\subsection{Semi-lattice formulation}
+\subsection{Type inference algorithm}
+
 % ==================================================================================================
-% Macros - for ICFP 2014
-% ==================================================================================================
-
-% comonadic semantics
-\newcommand{\F}{\mathsf{F}}
-\newcommand{\D}{\mathsf{D}}
-\newcommand{\Ds}[3]{\D^{\sclrd{#1}}_{#2} #3}
-\newcommand{\Icat}{\cclrd{\mathbb{I}}}
-\newcommand{\id}{\mathit{id}}
-\newcommand{\B}[1]{\bar{#1}}
-\newcommand{\bbullet}{\mathop{\bar{\bullet}}}
-\renewcommand{\idf}[0]{ \ident{id} }
-\newcommand{\m}{\mathsf{m}}
-\newcommand{\n}{\mathsf{n}}
-
-\newcommand{\shortar}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar_{#4}"#1";"#2"}
-\newcommand{\shortaro}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar^{#4}"#1";"#2"}
-\newcommand{\shortarn}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar_{#4}"#1";"#2"}
-\newcommand{\shortaron}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar^{#4}"#1";"#2"}
-\newcommand{\diagar}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}^{#4}"#1";"#2" }
-\newcommand{\diagaro}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}_{#4}"#1";"#2" }
-
-% Typing rule and typing statement (single line)
+
 \newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
-% Context transformation and context transformation with new line in the middle
-\renewcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
+\newcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
 \newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
-
-\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
-\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
-\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }} % \bullet} } % \textcolor{cclr}{\bullet}}  
-\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
-\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
-
-% \newcommand{\single}[{\langle{#1}\rangle}
-
-\newcommand{\uncurry}{\Lambda^{\hspace{-0.2em}-1}}
-
-% Coeffect shapes
 
@@
 \newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
-\renewcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
+\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
 \newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
-
 \newcommand{\sflat}{\star}
@@
 \newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
-\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq 
-% Tomas: I prefer \circledast here because it is a circled (lifted) version of \cseq 
-% { \textcolor{aclr}{\bullet}}
 \newcommand{\azero}{ \textcolor{aclr}{\bot} }
@@
 
-% Free monoid used in the *structural* coeffect algebra
 \newcommand{\fcons}{ \times } 
@@
 
-%===================================================================================================
-
-\section{Introduction}
-
-\emph{Context} is important for defining meaning-- not just in natural languages, but also in logics and 
-programming languages.
-The standard notion of context in programming is an environment 
-providing values for free variables. An open term with free variables is context dependent -- its 
-meaning depends on the free-variable context. The simply-typed $\lambda$-calculus famously 
-analyses such context usage. 
-Other systems go further. %capturing how 
-%the context is used. 
-For example, bounded linear logic tracks the number of times a 
-variable is used~\cite{girard1992bounded}.
-
-In software engineering, ``context'' provides more than just free-variable values. For example, in a
-distributed system, the context provides different resources that may be available on different 
-devices (\eg{}, a database on a server or a GPS sensor on a phone).
-
-In this paper we develop a calculus for capturing various notions of 
-context in programming. A key feature and contribution
-of the calculus is its \emph{coeffect system} which provides a static analysis for contextual properties (coeffects). The system
-follows the style of type and effect systems, but captures a different class of properties. % (Section~\ref{sec:intro-pathways}). 
-Another key contribution of the calculus is its semantics which can be smoothly instantiated for specific notions of context.
-%The semantics of the calculus is general and can be instantiated for particular notions of context.
-
-Coeffect systems were previously introduced as a generic analysis of
-context dependence which can be instantiated for various notions of
-context~\cite{petricek2013coeffects}. However, the formalization was restricted to
-tracking a class of \emph{whole-context} properties where a term has just
-one coeffect. This limited the applications and precision of any
-analysis. For example, a whole-context liveness analysis marks the 
-free-variable context as live (some variable may be used) or dead 
-(no variable is used), but it cannot record liveness \emph{per-variable}.
-
-We develop a more general system which captures both
-\emph{per-variable} coeffects, which we call \emph{structural}, and
-\emph{whole-context} coeffects, which we call \emph{flat}, and more. Our
-key contributions are:
-
-\begin{itemize}
-\item We present a general coeffect type system (Section~\ref{sec:coeffects}) and demonstrate two 
-  concrete uses -- \emph{flat} coeffect systems that tracks whole-context information and 
-  \emph{structural} coeffect systems which tracks fine-grained per-variable information.
-
-\item We show practical examples, instantiating the calculus for structural systems capturing 
-  variable usage based on bounded linear logic, dataflow caching, and precise liveness analysis.
-  We also instantiate the calculus to flat systems, building on and extending previous examples~\cite{petricek2013coeffects}.
-% variants of the coeffect system capture examples described   in our earlier work
-
-\item We discuss the syntactic properties of the \emph{structural} coeffect system 
-  (Section~\ref{sec:syntax}). It satisfies type preservation under both $\beta$-reduction 
-  and $\eta$-expansion, allowing its use with both call-by-name and call-by-value languages.
-  This important property distinguishes it from both effect systems and flat coeffects.
-
-\item We provide a denotational semantics, resisting and extending the notion of \emph{indexed comonads} to the 
-structural setting (Section~\ref{sec:semantics}). We prove soundness by showing 
-  the correspondence between syntactic and semantic properties of coeffect systems.
-\end{itemize}
-
-\noindent
-Coeffects can be approached from multiple directions (Section~\ref{sec:intro-pathways}) including 
-syntactic (effect systems), semantic, and proof-theoretic. We 
-emphasize the syntactic view, though we also outline a categorical semantics
-and note the interesting technical details.
-
-% ==================================================================================================
-
-\section{The coeffect calculus}
-\label{sec:coeffects}
-
-The three calculi shown in the previous section track two kinds of
-contextual properties: bounded reuse and dataflow are structural
-(per-variable) systems, and implicit parameters and our earlier coeffect
-systems~\cite{petricek2013coeffects} are flat (whole-context) systems. 
-This section presents our primary contribution: the general coeffect
-calculus.
-
-The calculus is parameterised by an algebraic structure of coeffects.
-To capture both structural and flat systems, coeffect annotations are
-indexed by a \emph{shape}. In flat systems, the shape is a singleton set
-and so annotations are \emph{scalar} values. Structural systems use shapes
-matching the number of variables in a free-variable context and so annotations
-are \emph{vectors}. However, the coeffect calculus could also use shapes describing
-trees and other structures.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Understanding coeffects: syntax and semantics}
-
-The coeffect calculus provides both an analysis of
-context dependence (its coeffect system) and a semantics for
-context (see Section~\ref{sec:semantics}). These two features of the calculus provide different
-perspectives on coeffect annotations $\aclrd{R}$ in a judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$:
-%%
-\begin{itemize}
-\item Syntactically, coeffects model \emph{contextual requirements} and 
-  may be overapproximated, such that more capabilities are required 
-  than necessary at runtime.
-
-\item Semantically, coeffects model \emph{contextual capabilities} and behave 
-  like containers of capabilities, such that the semantics may throw away 
-  capabilities that will not be needed.
-\end{itemize}
-%
-Thus there are two dual ways to understand coeffect annotations. Each
-perspective implies an alternate reading of the typing rules.
-%
-\begin{itemize}
-\item
-As \emph{contextual requirements}, the rules should be read top-down.
-The requirements of multiple sub-terms are \emph{merged} and the
-requirements of function body are \emph{split} between immediate (declaration-site)
-and latent (call-site) coeffects.
-
-\item
-As \emph{contextual capabilities}, the rules should be read bottom-up.
-The capabilities provided to a larger term are \emph{split} between 
-sub-terms; for functions, the capabilities of declaration-site and call-site
-are \emph{merged} and passed to the body.
-\end{itemize}
-%
-The reason for this asymmetry follows from the fact that context appears in a \emph{negative
-position} in the model. In Section~\ref{sec:semantics}, the denotation of a judgment
-is a function of the form $\D_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau}$ where $\D_{\aclrd{R}} \interp{\Gamma}$ 
-encodes the contextual capabilities used to evaluate a term. Similarly
-functions have models of the form $\D_{\cclrd{s}} \interp{\sigma} \rightarrow \interp{\tau}$
-with additional contextual capabilities attached to the input.
+\section{The unified coeffect calculus}
 
@@
 The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
-category in the semantics (Section~\ref{sec:semantics}). 
+category in the semantics. 
 It models sequential composition with variable access
@@
 For structural coeffects, the container is a vector, while for flat coeffects, it is
-a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{abbott2005containers}
+a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{types-containers}
 which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.
@@
 operators for combining shaped coeffects.
-
-\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
 
@@
 \vspace{-1em}
-\label{fig:coeffects-general}
+\label{fig:unified-coeffects}
 \end{figure}
@@
 \subsection{Structural coeffects}
-\label{sec:coeffects-structural}
+\label{sec:unified-structural}
 
@@
 \subsection{Flat coeffects}
-\label{sec:coeffects-flat}
+\label{sec:unified-flat}
 
@@
 
-\section{Equational theory}
-\label{sec:syntax}
-\newcommand{\reduce}{\leadsto}
-
-Each of the concrete coeffect systems discussed in this paper has a different notion of 
-context-dependence, much like various effectful languages have different notions of effects
-(such as state or exceptions). However, there are equational properties that hold for all (or some) of
-the systems we consider.
-
-The equational theory in this section illuminates 
-the axioms of coeffect algebra and the semantics of the calculus.
-We discuss syntactic substitution as it can form the basis for
-reduction in a concrete operational semantics. 
-We consider structural and flat systems
-separately. This provides better insight into how the two systems work and
-differ. In particular, call-by-name evaluation is \emph{coeffect
-preserving} for all structural, but only some flat systems.
-
-The properties and proofs in this section are syntactic.
-In Section~\ref{sec:semantics-soundness} we show that
-our denotational model of the coeffect calculus is sound with respect
-to the equational theory here.
-
-We use standard syntactic substitution written as $\subst{e_1}{x}{e_2}$, 
-$\beta\textnormal{-reduction}$ and $\eta$-expansion, written as $\reduce_\beta$ and $\reduce_\eta$.
-Equality of terms $e_1$ and $e_2$, defined by a relation
-$\equiv$, requires the equality of their contexts, types, and 
-coeffects, written $\coctx{\Gamma}{\aclrd{R}} \vdash e_1 \equiv e_2 : \tau$.
-
-%\begin{definition}[Subject reduction]
-%Given reduction $e \reduce e'$ and $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$
-%then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$, \ie{}, $\reduce$ is 
-%\emph{type-and-coeffect preserving}. 
-%\label{def:pres}
-%\end{definition}
-
-%Definition~\eqref{def:pres} is also known as the \emph{subject reduction} property.
-
-\subsection{Structural coeffect systems}
-\label{prop:structural}
-
-For structural coeffect systems, recall that coeffects are vectors
-with $\asplit = \amerge = \cons$ (vector concatenation) and $\azero =
-\langle{}\rangle$ (the empty vector), thus coeffect annotations
-comprise the \emph{free monoid} \ie, lists over the coeffect scalars
-(although we continue using the vector terminology). We first show
-substitution:
-
-\begin{lemma}[Substitution lemma] 
-\label{thm:structural-subst}
-In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
-\[
-\begin{array}{l}
-\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
-\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
-\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\vspace{-1.5em}
-\label{lemm:structural-subst}
-\end{lemma}
-\begin{proof}
-By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
-and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
-\end{proof}
-
-\noindent
-Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
-$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
-$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
-the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
-Using the substitution lemma, we can demonstrate $\beta$-equality:
-%
-\begin{equation*}
-\inference{
-  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
-             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
-           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
-  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
-\equiv
-\subst{e_1}{x}{e_2} : \tau}
-\end{equation*}
-%%
-\noindent
-As a result, $\beta$-reduction preserves the type and coeffects of a term.
-This gives the following subject reduction property:
-
-\begin{theorem}[Subject reduction]
-In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
-$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
-\end{proof}
-
-\noindent
-Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
-\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
-\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
-and application does not eliminate too much.
-%
-\begin{equation*}
-\inference{
-  \inference 
-    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
-     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
-    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
-  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
-: \sigma \xrightarrow{\cclrd{s}} \tau}
-\end{equation*}
-%
-The last step uses the fact that 
-$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
-arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
-
-This highlights another difference between coeffects and effects, as $\eta$-equality does
-not hold for many notions of effect. For example, in a language with output
-effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
-its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
-hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
-allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
-$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
-the exact immediate requirements of $\lambda x . e x$ to match those of $e$.
-
-\subsection{Flat coeffect systems}
-
-The equational theory for flat coeffect systems is somewhat similar to effect systems
-where (co)effects are not linked to individual variables. In effectful languages,
-substituting an effectful computation for $y$ in $\lambda x.y$ changes the latent effect 
-associated with the function.
-
-Similarly, for some of the flat coeffect systems, substituting a context-dependent computation
-for $y$ in $\lambda x.y$ adds latent context requirements to the function type. However, this is not
-the case for \emph{all} flat coeffect systems -- for example, call-by-name reduction preserves
-types and coeffects for the implicit parameters system (which makes it a suitable model for Haskell).
-For other systems, we first briefly consider call-by-value reduction.
-
-\paragraph{What is a value?}
-The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
-As discussed earlier, a function $(\lambda x.e)$ is not necessarily a value in coeffect calculi, 
-because it may not delay all context requirements of $e$. Thus we say that $e$ is a value if it
-has no immediate context requirements.
-
-\begin{definition} 
-An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
-where $\aclrd{\mathsf{VAL}} : \C^{\slift{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
-always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
-\end{definition}
-
-\paragraph{Call-by-value.}
-In call-by-value, a pure \emph{value} is substituted for a variable. 
-The right-hand side of an application is evaluated to a value before the
-$\beta$-reduction, but the discharging of coeffects prior to substitution is
-different for each concrete coeffect system.
-
-Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
-as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
-merging of context-requirements, respectively.
-
-\begin{lemma}[Call-by-value substitution]
-\label{thm:flat-subst-cbv}
-In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-and the $\cclrd{\wedge}$ operator:
-\[
-\begin{array}{l}
-\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\end{lemma}
-\begin{proof}
-By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
-with $\cunit$.
-\end{proof}
-
-\noindent
-Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
-the operational semantics must provide a way of partially evaluating a term 
-with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
-Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:
-
-\begin{theorem}[Call-by-value reduction]
-In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
-$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
-\end{proof}
-
-\paragraph{Call-by-name.}
-The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
-may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
-$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.
-
-% TODO
-%
-% The context requirements of the 
-% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
-% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
-% lemma should be $r \fmax (s \fseq r)$.
-
-We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
-$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
-Liveness analysis is an example of top-pointed coeffects
-as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.
-
-\begin{lemma}[Top-pointed substitution]
-\label{thm:cbn-substitution-top}
-In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-and the $\cclrd{\wedge}$ operator:
-\[
-\begin{array}{l}
-\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\vspace{-1.5em}
-\end{lemma}
-\begin{proof}
-Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
-\end{proof}
-%
-\noindent
-As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
-use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).
-
-In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
-requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
-the context requirements arising from the substitution can be associated with the context
-$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
-requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
-(all three operators are set union) and allows the following substitution lemma:
-
-\begin{lemma}[Bottom-pointed substitution]
-\label{thm:cbn-substitution-bot}
-In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
-and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
-\[
-\begin{array}{l}
-\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
-\end{array}
-\]
-\vspace{-1.5em}
-\end{lemma}
-\begin{proof}
-By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
-the free-variable context.
-\end{proof}
-\noindent
-The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
-variant described here is flexible enough to let us always re-associate new context requirements with the 
-free-variable context.
-
-The two substitution lemmas show that the call-by-name evaluation strategy can be used for
-certain coeffect calculi, including liveness and implicit parameters.
-Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
-reduction, the following theorem holds:
-
-\begin{theorem}[Call-by-name reduction]
-\raggedright
-In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
-Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
-\end{proof}
-
-%
-% TODO: We could have a section about let binding covering the alternative CBV and the let binding
-% we included in the system, but we'd have to clean this up!
-%
-%\subsection{Syntactic let binding}
-%\begin{remark}
-%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
-%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
-%calculus of Section~\ref{}.
-%%%
-%\begin{align*}
-%{\footnotesize{\trule{let}}}
-%\dfrac
-%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
-%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
-%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
-%\end{align*}
-%%%
-%Therefore, for example, we see that \emph{let} has the coeffects
-%as substitution, for example:
-%%%
-%\begin{align*}
-%& {\footnotesize{\trule{let}}}
-%\dfrac
-%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
-%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
-%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
-%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
-%\end{align*}
-%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
-%the judgment:
-%%%
-%$$
-%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
-%$$
-%%%
-%\end{remark}
-%
-% 
-%\paragraph{Syntactical let binding}
-%Has the same typing derivation as abstraction/application
-%
-%\[
-%\inference
-% {\inference
-%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
-%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
-%  &
-%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
-%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
-%\]
-%
-%We can extend the CBV behaviour to \emph{let}, where the bound term
-%is evaluated to a pure value first, discharging any contextual
-%requirements before substitution into the body.
-%
-%\begin{lemma}[CBV \emph{let}]
-%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
-%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
-%: \tau$.
-%\end{lemma}
-
-% ==================================================================================================
-
-\section{Semantics}
-\label{sec:semantics}
-
-\newcommand{\Di}{\F}
-
-Coeffects provide a unified notion of context-dependence. In the 
-previous sections, we used this to define a unified coeffect calculus.
-We now define a unified (categorical) semantics for the 
-coeffect calculus. The semantics can be instantiated for different 
-notions of context dependence and thus can model a wide range of
-context-aware languages (both for flat and structural systems).
-
-We relate the semantics to the equational theory and show that it is
-sound with respect to term equality. For a variant
-of the the flat system, a similar result has already been shown in the 
-second author's dissertation~\cite{orchard2013thesis}.
-The semantics is introduced in pieces:
-
-\begin{itemize}
-\item
-Section~\ref{sem:interp}
-describes the range and domain (signature) of the interpretation $\interp{-}$, 
-gives the interpretations for types and free-variable contexts (in
-flat and structural systems), and defines the signature of
-functors $\D$ which encode contexts.
-
-\item
-The first part of the semantics (Section~\ref{sem:seq}) defines \emph{sequential
-composition} of context-dependent computations via \emph{indexed
-  comonads} (introduced briefly in our previous
-work~\cite{petricek2013coeffects}) and the \emph{indexed structural
-  comonad} structure (new here).
-
-\item
-More structure is needed for application and abstraction. 
-Section~\ref{sem:monoidal} defines
-indexed monoidal operations for splitting and merging contexts.
-Concrete structures are given throughout for the semantics of the
-structural bounded reuse and flat implicit parameter systems.
-
-\item
-Section~\ref{sem:together} puts the pieces together, defining the
-semantics of the coeffect calculus along with structural rules.  
-The semantics is illustrated by executing an 
- example bounded-reuse program in 
-the semantics (Example~\ref{exm:run}).% and the concrete structures for
-%the semantics of structural dataflow (Example~\ref{exm:dataflow-full}).
-
-\item
-Section~\ref{sec:semantics-soundness} shows our
-semantics sound with respect to the syntactic equational theory of
-Section~\ref{sec:syntax}. This uses the derivation of 
-categorical structures for the semantics as \emph{lax homomorphisms} between
-structure on the coeffect category $\Icat$ and the base $\mathbb{C}$.
-
-\end{itemize}
-
-%
-% TP: Removed \paragraph just because I think this looks nicer with the list
-%
-\noindent 
-In this section, 
-$\mathbb{C}, \mathbb{D}, \mathbb{I}$ range over categories.  The
-objects of a category $\mathbb{C}$ are written
-$\emph{obj}(\mathbb{C})$.  The category of functors between
-$\mathbb{C}$ and $\mathbb{D}$ is written $[\mathbb{C}, \mathbb{D}]$.
-Exponential objects, representing function types in our model, 
-are written in two ways, either $B^A$ or $A \Rightarrow B$.
-
-\subsection{Interpreting contexts and judgments} 
-\label{sem:interp}
-
-The semantics is parameterised by a coeffect algebra, with scalar coeffects $(\C,
-\cseq, \cpar, \cunit, \czero, \cleq)$, coeffect shape $(\SH, \slift{-},
-\stimes, \sempty, \sunit)$, and $(\amerge, \asplit, \azero)$. An
-interpretation $\interp{-}$ is given to types, free-variable contexts,
-and type and coeffect judgments, with a base Cartesian-closed category
-$\mathbb{C}$ for denotations and a category $\Icat$ of scalar coeffects, where $\emph{obj}(\Icat)
-= \C$. Since $\mathbb{C}$ is Cartesian-closed, we use the
-$\lambda$-calculus as the syntax for giving concrete definitions.
-
-The interpretation $\interp{-}$ is parameterised by categorical 
-structures which model a particular notion of context. %We show  
-%instances of these structures for bounded reuse and dataflow later. 
-The interpretation of free-variable contexts depends on shape, for
-which we give concrete definitions for flat and structural shapes.
-
-\paragraph{Interpreting judgments.}
-type and coeffect judgments are interpreted (given denotations) 
-as morphisms in $\mathbb{C}$, of the form:
-%
-\begin{align*}
-\interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \D^{\slift{\Gamma}}_{\aclrd{R}}
-\interp{\Gamma} \rightarrow \interp{\tau}
-\end{align*}
-%%
-The interpretation is a morphism from an interpretation of the context $\Gamma$
-to the interpretation of the result. The functor $\D^{\slift{\Gamma}}_{\aclrd{R}}$ over the
-context encodes the semantic notion of context and is indexed by the free-variable 
-context shape $\slift{\Gamma}$ and coeffect annotation $\aclrd{R}$.
-
-The structure $\D$ can be thought of as a coproduct of functors $\D^{\sclrd{n}}$ for every
-possible shape $\sclrd{n} \in \SH$ of free-variable context:
-%%
-\begin{equation*}
-\D : \Sigma_{\sclrd{n} : \SH}. \D^{\sclrd{n}}
-\quad \text{where} \quad
-\D^{\sclrd{n}} : \Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}]
-\end{equation*}
-% %
-For a fixed context shape $\sclrd{n}$ the functor $\D^{\sclrd{n}} :
-\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}]$ 
-maps an $\sclrd{n}$-indexed coeffect (think positions) to a functor from a
-context $\mathbb{C}^n$ to an object in $\mathbb{C}$. That is, given a coeffect
-annotation (matching the shape of the context), we get a 
-functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$. 
-%
-
-From a programming 
-perspective, this functor defines a data structure that models the additional
-context provided to the program. The shape of this data structure depends on the coeffect
-annotation $\Icat^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
-the number of values needed for each variable and the functor will be 
-formed by lists of length matching the required number.
-
-%
-% TP: I think this is easier to understand in the later paragraph
-%
-% Thus free-variable contexts are interpreted
-% $\interp{\Gamma} \in \mathbb{C}^{\slift{\Gamma}}$ and
-
-%
-% TP: I did not get the following - but I think I tried to say the same thing above...?
-%
-% $\D$ couples the
-% shape of the coeffect with the shape of the free-variable context.
-%Below, the interpretation of free-variable contexts is defined for
-%classes of flat and structural system.
-
-\paragraph{Types.} 
-Types are interpreted as objects of $\mathbb{C}$, that is 
-$\interp{\tau} : \emph{obj}(\mathbb{C})$
- where function types have the interpretation as exponents: 
-%
-\begin{align*}
-  \interp{\sigma \xrightarrow{\cclrd{r}} \tau} = \D^{\sunit}_{\cclrd{r}} \interp{\sigma}
-  \Rightarrow \interp{\tau}
-\end{align*}
-%%
-The parameter of a function is wrapped by 
-a functor $\D^{\sunit}_{\cclrd{r}}$ that defines a context with singleton shape
-$\sunit$, matching the single value that it contains.
-This interpretation is shared by all coeffect calculi.
-
-%
-% TP [Deleted] "although it may be extended to other type constructors."
-% (I did not understand what you mean here)
-%
-
-\paragraph{Free-variable contexts.} 
-As described above, free-variable contexts $\Gamma$ are given an
-interpretation as objects in $\mathbb{C}^{\slift{\Gamma}}$. Thus,
-the interpretation of contexts is shape dependent.
-
-%\tpnote{This is tricky but clever! Tried to add explanation}
-We define $\interp{-}$ on free-variable context for structural and flat
-systems. For flat systems, there is only a single shape, so the interpretation
-is a product type inside the Cartesian closed category $\mathbb{C}$. For 
-structural systems, the shape matches the number of variables and so the model 
-is a value in the product category $\mathbb{C} \times \ldots \times \mathbb{C}$.
-
-\paragraph{\textnormal{\emph{Flat coeffects.}}} 
-Recall that $\SH = \{\sflat\}$ and $\slift{\Gamma} = \sflat$.
-Since the set of positions $\sflat$ is a singleton, then $\mathbb{C}^{\sflat}$
-is isomorphic to $\mathbb{C}$.
-%
-% TP: I see why the following bit is important, but I think we can assume people
-% will just believe us here and make this a bit simpler
-%
-% as categories formed from a singleton set are terminal objects in \textbf{Cat}. 
-%
-Therefore $\interp{\Gamma} : \emph{obj}(\mathbb{C})$, which is defined as:
-%%
-% \begin{align*}
-% \interp{x : \tau, \Gamma'} = \interp{\tau} \times \interp{\Gamma'}
-% \qquad\qquad
-% \interp{\cdot} = 1 
-% \end{align*}
-%
-% TP: Is there anything wrong with the following? I find it a bit simpler...
-%
-\begin{align*}
-\interp{x_1 : \tau_1, \ldots, x_n : \tau_n} = \interp{\tau_1} \times \ldots \times \interp{\tau_n}
-\end{align*}
-%
-Denotations of typing judgments in a flat coeffect system are thus of the form 
-(where $\aclrd{R} \in \Icat$): 
-%%
-$$\interp{\coctx{x_1 : \tau_1, .., 
-x_n : \tau_n}{\aclrd{R}} \vdash e : \tau} : \D^{\sflat}_{\aclrd{R}}(\interp{\tau_1} \times ... \times \interp{\tau_n})
-\rightarrow \interp{\tau}
-$$
-
-\paragraph{\textnormal{\emph{Structural coeffects.}}}
-Recall that $\SH = \mathbb{N}$ and $\slift{\Gamma} = |\Gamma|$
-(number of free variables), thus 
-$\interp{\Gamma} : \emph{obj}(\mathbb{C}^{|\Gamma|})$. This is defined similarly 
-to the above, but instead of using products in $\mathbb{C}$, %and the terminal object
-%$1 : \mathbb{C}$, 
-we use the product of categories. % (\ie{}, $\times$ 
-% in \textbf{Cat}) with the singleton category $1 \in \mathbf{Cat}$. 
-%
-% TP: If you think that we should say how empty products are treated
-% then I would just say "Empty contexts are modelled using the singleton category"
-% (those who would ask the question will understand this and it does not add
-% too much fancy notation to the explanation)
-%
-Thus, denotations have the form: 
-$$
-\interp{\coctx{x_1 : \tau_1, \ldots, x_n : \tau_n}{\aclrd{R}} \vdash e : \tau} : \D^n_{\aclrd{R}}
-(\interp{\tau_1}, \ldots, \interp{\tau_n}) \rightarrow \interp{\tau}
-$$
-where $|R| = n$ and we use commas (instead of $\times$) to denote the product
-of categories. %in $\mathbf{Cat}$. 
-This means that $\D^n : \Icat^n \rightarrow [\mathbb{C}^n,
-\mathbb{C}]$ is a functor between an $n$-length vector of coeffects
-indices and an \emph{n}-ary endofunctor.
-As mentioned, the key
-difference between the flat and structural interpretations of
-free-variable contexts is that flat uses products of objects in
-$\mathbb{C}$ and the structural uses products of $\mathbb{C}$ in 
-the category of categories. %\textbf{Cat}.
-
-\newcommand{\BLL}{\mathsf{B}}
-
-\begin{example}[Bounded reuse] Recall
-bounded reuse has coeffect scalars $\C = \mathbb{N}$ and shapes $\SH = \mathbb{N}$.
-We model contexts by replicating the value of each variable so 
-there is a value for each use. This matches the model used by 
-Girard \emph{et al.}~\cite{girard1992bounded}. Contexts are described by 
- $\BLL : \Sigma_{\sclrd{n} : \mathbb{N}}. (\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$, 
-where for $\aclrd{R} = \langle{\cclrd{r_1}, \ldots, \cclrd{r_n}}\rangle$: 
-%%
-%\tpnote{\\I do not understand what the $a_i$ functions mean (in the morphism mapping part)?}
-\begin{align*}
-\BLL^n_{\aclrd{R}} (A_1, \ldots, A_n) & = A_1^{\cclrd{r_1}} \times \ldots \times A_n^{\cclrd{r_n}} \\
-\BLL^n_{\aclrd{R}} (f_1, \;\ldots\;, f_n) & = \lambda \langle{a_1, \ldots, a_n}\rangle . \langle{(f_1 \circ a_1), \ldots, (f_n \circ a_n)}\rangle
-\end{align*}
-%
-Thus each object in the free-variable context $A_i$ is indexed by its associated coeffect $\cclrd{r_i}$.
-For the morphism mapping, $f_i : A_i \rightarrow B_i$ and $a_i : {A_i}^{\cclrd{r_i}}$ thus $(f_i \circ a_i) : {B_i}^{\cclrd{r_i}}$. 
-The exponent ${A_i}^{\cclrd{r_i}}$ can be read as a product of $\cclrd{r_i}$ copies of $A_i$, \eg:
-$$\BLL^3_{1,0,2} (A, B, C) = A^1 \times B^0 \times C^2 = (A) \times 1 \times (C \times C)$$
-\label{exm:bll1}
-\end{example}
-
-\vspace{-2em}
-\newcommand{\Imp}{\mathsf{I}}
-\begin{example}[Implicit paramters]
-Recall the implicit parameter calculus with scalar coeffects as 
-sets of names paired with types $\C = \mathcal{P}(\ident{Name} \times \ident{Types})$ 
-and flat shape with singleton $\SH = {\{\sflat\}}$.
-
-%\tpnote{Could you say, let $\mathsf{Impl}=\textbf{Set}$ be a category of pairs of vars and values and use $\mathsf{Impl}$ throughout?\\ I find Set non-intuitive }
-Its contexts are defined by 
-$\Imp^\star :  \Sigma_{\sclrd{n} : \{\star\}}, (\Icat^{\sclrd{n}} \rightarrow [\mathbf{Set}^{\sclrd{n}}, \mathbf{Set}])$,
-which is equivalent to $\Icat \rightarrow [\mathbf{Set}, \mathbf{Set}]$) and defined as follows:
-%%
-\begin{align*}
-\Imp^{\sflat}_{\aclrd{R}} A = A \times \interp{\aclrd{R}} \qquad\qquad
-\Imp^{\sflat}_{\aclrd{R}} f = \lambda (a, r) . (f\ a, r)
-\end{align*}
-%%
-The interpretation $\interp{\aclrd{R}}$ maps a set of variable-type
-pairs to an object representing a set of variables-values pairs in $\mathbf{Set}$.
-\label{exm:impl1}
-\end{example}
-
-\subsection{Sequential composition}
-\label{sem:seq}
-
-Following the usual categorical semantics approach, we require a
-notion of sequential composition for our denotations.  We show first a
-special case for $\D^\sunit$, where $\Icat^\sunit = \Icat$
-and $\mathbb{C}^\sunit = \mathbb{C}$ in both flat and structural
-systems\footnote{since $\sunit = 1$ in structural and $\sunit \cong
-  1$ in flat, \ie{}, $\star$ is isomorphic to 1} and thus $\D^\sunit :
-\Icat \rightarrow [\mathbb{C}, \mathbb{C}]$. Composition of  
-morphisms $f : \D^\sunit_S A \rightarrow B$ and $g : \D^\sunit_R B
-\rightarrow C$ is defined by an \emph{indexed comonad}  
-(which we previously introduced briefly~\cite{petricek2013coeffects}).
-
-\begin{definition}
-An \emph{indexed comonad} comprises a strict monoidal category
-$(\mathbb{I}, \bullet, I)$ and a functor $\Di : \mathbb{I}
-\rightarrow [\mathbb{C}, \mathbb{C}]$ with two natural transformations
-(where we write $(\Di\,R)\,A$ as $\Di_R A$):
-%\tpnote{Added brackets in $(\Di\,R)\,A$ -- otherwise I was confused}
-%%
-\begin{align*}
-(\delta_{X,Y})_A : \Di_{(X \bullet Y)} A \rightarrow \Di_X (\Di_Y A) \qquad
-(\varepsilon_I)_A : \Di_I A \rightarrow A
-%\delta_{X,Y} : \Di (X \bullet Y) \xrightarrow{.} \Di X \circ \Di Y \qquad
-%\varepsilon_I : \Di I \xrightarrow{.} 1_{\mathbb{C}}
-\end{align*}
-%%
-where $\delta$ is called \emph{comultiplication} and $\varepsilon$ is called \emph{counit}.
-We require indexed 
-analogues of the usual comonad axioms (\emph{cf.}~\cite{uustalu2008notions}): 
-%$$
-%\Di_R \varepsilon_I \circ \delta_{R,I} = {\varepsilon_I}_{\Di_R} \circ \delta_{I, R} = id \quad 
-%\delta_{R,S}\Di_T \delta_{R \bullet{} S, T} = \Di_R \delta_{S,T} \circ \delta_{R, S \bullet{} T}
-%$$ 
-%
-%\tpnote{I'd consider omitting the axioms (too much theory?) if we can refer to your thesis?}
-%\tpnote{[Read 2] OK - I commented them out - but feel free to revert!}
-\begin{equation*}
-\hspace{-1em}
-\xymatrix@R=1em{
-\Di_R \ar[r]^-{\delta_{R, I}} \ar@{=}[dr] \ar[d]_{\delta_{I,R}} \ar@{}[dr]^{\hspace{0.3em}\text{[C2]}} & \Di_R \Di_I \ar[d]^{\Di_R \varepsilon_I} \\
-\ar@{}[ur]_{\hspace{-1.8em}\text{[C1]}} \Di_I \Di_R \ar[r]_{\varepsilon_I \Di_R} & \Di_R
-}
-\hspace{0.6em}
-\xymatrix@R=1em{
-\ar@{}[dr]|{\text{[C3]}}
-\Di_{R \bullet{} S \bullet{} T} \ar[d]_{\delta_{R, S \bullet{} T}} \ar[r]^{\delta_{R \bullet{} S, T}} & \Di_{R \bullet{} S}  \Di_T \ar[d]^{\delta_{R,S}\Di_T} \\
-\Di_R \Di_{S \bullet{} T} \ar[r]_{\Di_R \delta_{S,T}} & \Di_R \Di_S \Di_T
-}
-\label{eq:indexed-comonads-axioms}
-\end{equation*}
-\end{definition}
-
-\noindent
-An indexed comonad $\Di : \mathbb{I} \rightarrow
-[\mathbb{C}, \mathbb{C}]$ induces a notion of composition
-for all $f : \Di_S A \rightarrow B, g : \Di_R B \rightarrow C$: 
-%
-\begin{equation*}
- g \hat{\circ} f = g \circ \Di_R f \circ \delta_{R,S} \, : \, \Di_{R \bullet S} A \rightarrow B
-\end{equation*}
-%
-with the identity $\hat{\id}_A = (\varepsilon_I)_A : \Di_I A \rightarrow A$ for all $A$. Thus
-in\-dexed comonads induce a category which has the same objects as $\mathbb{C}$ and morphisms 
-  $\mathbb{C}_\Di(A, B) = \bigcup_{R \in \mathbb{I}} \mathbb{C}(\Di_R A, B)$.
-Note that an indexed comonad is not a family of (ordinary) comonads, because the identity
-is only defined for the functor $\Di_I$. %and the functor in the result of composition (and comultiplication) 
-%may be any $\Di_R$.
-
-Therefore, if $\D^\sunit$ is an indexed comonad, there is a notion of composition for
-denotations with a single coeffect index. % in our semantics.
-
-\begin{example}[Bounded reuse]
-$\BLL^\sunit_R$ (Example~\ref{exm:bll1}) has an indexed comonad structure, where
-the monoid $(\cclrd{\mathbb{N}}, \cclrd{\ast}, \cclrd{1})$ from the coeffect scalar for bounded reuse
-induces a monoidal category structure on $\Icat$ (with $1 : \Icat$ and the bifunctor
-$\ast : \Icat \times \Icat \rightarrow \Icat$), with operations:
-%
-%\begin{align*}
-%\begin{array}{l}
-%\delta^\sunit_{R,S} = \lambda \langle{a_1 ..., a_{R S}}\rangle . \\
-%\hspace{3.5em} \langle{~\langle{a_1 ..., a_S}\rangle, \langle{a_{S+1},..., a_{S+S}}\rangle}, \ldots \\
-%\hspace{4.5em} \;\;~\langle{a_{(R - 1)S + 1},..., a_{R S}}\rangle~~\rangle 
-%\end{array}
-%\quad\quad
-%\begin{array}{l}
-%\varepsilon^\sunit_{1} = \lambda \langle{a_1}\rangle . a_1  \\[0.5em] 
-%\\
-%\end{array}
-%\end{align*}
-\begin{align*}
-\begin{array}{rl}
-\varepsilon^\sunit_{1} & \hspace{-1em} = \lambda \langle{a_1}\rangle . a_1 \\
-\hspace{-0.5em}
-\delta^\sunit_{R,S} & \hspace{-1em} = \lambda \langle{a_1 ..., a_{R S}}\rangle . \\
-& \hspace{0.4em} \langle{\langle{a_1 ..., a_S}\rangle, \langle{a_{S+1},..., a_{S+S}}\rangle}, ..., \langle{a_{(R - 1)S + 1},..., a_{R S}}\rangle\rangle
-\end{array}
-\end{align*}
-%\tpnote{Does this explanation for couint / comult. make sense?}
-Indexed comonads essentially model single-variable contexts.
-%Recall that we only consider single-variable context.
-The counit requires a single copy of the value from the context. 
-The comultiplication splits $R$ times $S$ copies of a value into
-$R$ copies of a context where each context contains just $S$ copies of the value.
-\label{exm:bll2}
-\end{example}
-
-\begin{remark}
-A semantics for dataflow coeffects is similar to bounded reuse with
-$\D^n_R (A_1, \ldots, A_n) = (A_1 \times A_1^{R_1}) \times ... \times (A_n \times {A_n}^{R_n})$, \ie{},
-each free-variable has an extra value representing the ``current'' value. A dataflow indexed
-comonad is similar to the above but with additive rather than multiplicative behaviour. 
-\end{remark}
-
-\begin{example}[Implicit parameters]
-For the coeffect scalar monoid $\cclrd{(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \emptyset{})}$ 
-of implicit parameters, 
-$\Imp^\ast$ (Example~\ref{exm:impl1}) there is an indexed comonad structure, with operations: 
-%where induces a monoidal category structure on $\Icat$, with operations:
-%For a set $X$, with the monoidal category of $X$-subsets 
-%$(\mathcal{P}(X), \cup, \emptyset)$, 
-%then $\D : \mathcal{P}(X) \rightarrow [\mathbf{Set}, \mathbf{Set}]$ where
-% $\D_X A = A \times X$ and $\D_X f = f \times \id_X$ 
-%is an indexed comonad with:
-%
-\begin{align*}
-\varepsilon_{\emptyset} \, (a, \emptyset{}) \mapsto a 
-\qquad\qquad\quad~~~
-\delta_{R,S} (a, \gamma) \mapsto ((a, {\gamma|}_S ), {\gamma|}_R) %\\
-%\qquad \textbf{where}~~~{\gamma|}_R = \{(x, v) \mid (x, v) \in \gamma, (x, t) \in R \}
-%\delta_{R,S} (a, x) \mapsto ((a,  x-(R-S)), x-(S-R))
-\end{align*}
-%%`
-where ${\gamma|}_R = \{(x, v) \mid (x, v) \in \gamma, (x, t) \in R \}$ filters incoming implicit parameters to those
-variable-value pairs where the variable is in the coeffect $R$.
-%%
-%where $x : R \cup S$ and thus the $R$-only subset of $x$ is $x - (S - R)$
-%and the $S$-only subset of $x$ is $x - (R - S)$ (where ($-$) is set difference).
-\label{exm:impl2}
-\end{example}
-
-\indent
-These two examples (which are new here) provide composition for context-dependent
-computations indexed by coeffects in a flat calculus. For structural coeffects,
-we need to compose morphisms which have more than a single coeffect annotation. For this, we
-introduced the notion of \emph{structural indexed comonads}.
-
-\begin{definition}
-\newcommand{\sunitc}{\sunit} %{{\sclrd{1}}}
-%
-A \emph{structural indexed comonad} comprises a functor 
-$\D : \Sigma_{n : \SH}. (\mathbb{I}^n \rightarrow [\mathbb{C}^n, \mathbb{C}])$
-where $(\mathbb{I}, \bullet, I)$ is a strict monoidal category,  
-$\sunit \in \SH$ and 
-$\D^{\sunit} : \mathbb{I}^\sunit \rightarrow [\mathbb{C}^\sunit, \mathbb{C}]$
- is an indexed comonad (with 
-$(\delta^\sunit_{X,Y})_A : \D^\sunit_{X \bullet Y} A \rightarrow \D^\sunit_{X} \D^\sunit_{Y} A$ 
-and 
-$(\varepsilon_I)_A : \D^\sunit_I A \rightarrow A$) and 
-a \emph{structural comultiplication} natural transformation:
-$$
-(\delta^{\sclrd{n}}_{r, S})_{A^n} : \D^{\sclrd{n}}_{r \bar{\bullet} S} A^n \rightarrow \D^\sunit_r \D^{\sclrd{n}}_S A^n
-$$ 
-%$$
-%(\delta^{\sclrd{n}}_{R, S})_{A_\sunitc, \ldots, A_n} : \D^{\sclrd{n}}_{R \bar{\bullet} S} (A_1, \ldots, A_n) \rightarrow %\D^\sunitc_R \D^{\sclrd{n}}_S (A_1, \ldots, A_n)
-%$$ 
-where $A^n \in \mathbb{C}^n$, $r \in \Icat$, $S \in \Icat^{\sclrd{n}}$ and
- $\bbullet : \mathbb{I} \times \mathbb{I}^{\sclrd{n}} \rightarrow \mathbb{I}^{\sclrd{n}}$ is 
- the monoid left action for $\bullet$ lifting scalar coeffects to shaped coeffects (\eg{}, 
-the scalar-vector version of $\bullet$).
-Analogous laws to monoid left actions for unitality and associativity hold for structural comultiplication:
-$$
-\varepsilon_I \circ \delta^{\sclrd{n}}_{I, r} = id 
-\,\qquad
-\D^\sunitc_r \delta^{\sclrd{n}}_{S,T} \circ \delta^{\sclrd{n}}_{r, S \bar{\bullet{}} T}
-= 
-\delta^\sunitc_{r,S}\D^{\sclrd{n}}_T \circ \delta^{\sclrd{n}}_{r \bar{\bullet{}} S, T}
-$$
-%Structural comultiplication satisfies 
-%$\varepsilon_I \circ \delta^{\sclrd{n}}_{I, R} = id$ 
-%, but
-%with appropriate indices.
-%
-%\tpnote{Totally \emph{not} sure if this is right -- but I \emph{believe} we could keep law details for MFPS}
-%
-%
-%%
-%\begin{align}
-%\hspace{-1em}
-%\xymatrix@R=1.4em{
-%\D^{\sclrd{n}}_{I \bar{\bullet} R} \ar[r]^-{\delta^{\sclrd{n}}_{I,R}} \ar@{=}[dr] \ar@{}[dr]^{\hspace{0.3em}\text{[SC1]}} & \D^\sunitc_I \D^{\sclrd{n}}_R \ar[d]^{\varepsilon_I} \\
-% & \D_R
-%}
-%\;
-%\xymatrix@R=1em{
-%\ar@{}[dr]|{\text{[SC2]}}
-%\D^{\sclrd{n}}_{R \bar{\bullet{}} S \bar{\bullet{}} T} \ar[d]_{\delta^{\sclrd{n}}_{R, S \bar{\bullet{}} T}} \ar[r]^{\delta^{\sclrd{n}}_{R \bar{\bullet{}} S, T}} %& \D^\sunitc_{R \bullet{} S}  \D^{\sclrd{n}}_T \ar[d]^{\delta^\sunitc_{R,S}\D^{\sclrd{n}}_T} \\
-%\D^\sunitc_R \D^{\sclrd{n}}_{S \bar{\bullet{}} T} \ar[r]_{\D^\sunitc_R \delta^{\sclrd{n}}_{S,T}} & \D^\sunitc_R \D^\sunitc_S \D^{\sclrd{n}}_T
-%}
-%\vspace{-1em}
-%\label{eq:structural-indexed-comonad-axioms}
-%\end{align}
-%for unit and associativity respectively 
-Note the indexed comonad comultiplication $\delta^\sunitc$ for associativity. 
-\label{def:structural-indexed-comonad}
-\end{definition}
-
-\noindent
-Structural indexed comonads provide composition for morphisms
-%$f : \D^{\sclrd{n}}_{S} (A_1 \times \ldots \times A_n) \rightarrow B$ and $g : \D^\sunit_R B \rightarrow C$:
-$f : \D^{\sclrd{n}}_{S} A^n \rightarrow B$ and singleton-shaped $g : \D^\sunit_r B \rightarrow C$:
-%%
-$$
-g \hat{\circ} f = g \circ \D^{\sunit}_r f \circ \delta^{\sclrd{n}}_{r,S} 
-: \D^{\sclrd{n}}_{r \bbullet S} A^n \rightarrow C
-$$
-%$$
-%g \hat{\circ} f = g \circ \D^{\sunit}_R f \circ \delta^{\sclrd{n}}_{R,S} 
-%: \D^{\sclrd{n}}_{R \bbullet S} (A_1 \times \ldots \times A_n)  \rightarrow C
-%$$
-%%
-Note that this composition is asymmetric: the left morphism and right morphisms 
-have different shapes. To compose morphisms which both have
-non-trivial context shapes requires additional structure for manipulating contexts
-(shown in the next section).
-
-\begin{example}[Bounded reuse]
-$\BLL :\Sigma_{\sclrd{n} : \mathbb{N}}, (\Icat^{\sclrd{n}} \rightarrow [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$ has a structural indexed comonad structure with the indexed comonad $\BLL^\sunit$ (Example~\ref{exm:bll2}) and the following structural comultiplication:
-%%
-%\tpnote{Is there a mismatch between $,$ and $\times$?}
-\arraycolsep=1pt\def\arraystretch{0.4}
-\begin{align*}
-%\arraycolsep=3pt\def\arraystretch{0.4}
-& \delta^n_{r,S} = \lambda (\langle{a^1_1, \ldots, a^1_{r \ast S_1}}\rangle, \ldots, \langle{a^n_1, \ldots a^n_{r \ast S_n}}\rangle) . \\[-0.3em]
-& \begin{array}{lllll}
-    ( &  ( &   \langle{a^1_1, \ldots, a^1_{S_1}}\rangle, & ..., & \langle{a^n_1,\ldots, a^n_{S_n}}\rangle ), \\[0.3em]
-            &  ( &   \langle{a^1_{(S_1 + 1)}, ..., a^1_{(S_1+1) + S_1}}\rangle, & ..., & \langle{a^n_{(S_n + 1)}, \ldots, a^n_{(S_n+1) + S_n}}\rangle ), \\[0.6em]
-            &  & \ldots  \\[0.2em]
-            &  ( &   \langle{a^1_{(r - 1) \ast s_1 + 1}, .., a^1_{r \ast S_1}}\rangle, & ..., & \langle{a^n_{(r - 1) \ast S_n + 1}, ..., a^n_{r \ast S_n}}\rangle ) ) 
-   \end{array}
-\end{align*}
-%
-The input is an $n$-variable context 
-containing $r$ times $S_i$ copies of $a^i$ for each variable. The output
-has $r$ copies of a single $n$-variable context containing $S_i$ copies of $a^i$ for each variable. Thus,
-$\delta^n_{r,S}$ partitions the incoming context into $r$-sized contexts. 
-%into a single-variable context, containing $R$ copies of the input vector. In each copy, there are $S$ copies of each
-%variable.
-\label{exm:bll3}
-%\tpnote{Please check - does this make sense?}
-\end{example}
-
-\noindent
-Note that in the case of the flat system, a structural indexed co\-mo\-nad collapses
-to a regular indexed comonad on $\D^{\sunit}$.
-
-\subsection{Splitting and merging contexts}
-\label{sem:monoidal}
-
-Indexed comonads and structural indexed comonads give a semantics for
-sequential composition of contextual computations. However, this does
-not provide enough structure for a full semantics of the coeffect
-calculus. Core to the semantics of abstraction and application is the
-merging and splitting of contexts. Recall the free-variable contexts 
-and coeffects in the (abs) and (app) rules:
-%%
-\begin{align*}
-{\text{\scriptsize{(app)}}}
-\dfrac
-      {\coctx{\Gamma_1}{R} \vdash e_1 ... \;
-       \coctx{\Gamma_2}{S} \vdash e_2 ...}
-      {\coctx{\Gamma_1, \Gamma_2}{ R \asplit (T \aseq S)} 
-\vdash e_1 \, e_2 ...}
-\;
-& 
-\;
-{\text{\scriptsize{(abs)}}}
-\dfrac
-  {\coctx{\Gamma, v : \sigma}{R \amerge \alift{S}} \vdash e ...}
-  {\coctx{\Gamma}{R} \vdash \lambda v . e ... : \sigma \xrightarrow{S}\ldots}
-\end{align*}
-%
-Reading (app) bottom-up, the context of the application
-is split into two contexts for each subterm $e_1$ and $e_2$. 
-Reading (abs) bottom-up, the context of the abstraction
-is merged with the singleton context of the parameter. 
-Capturing these notions in the denotational semantics requires
-some additional structure.
-
-% \tpnote{I removed some details about $\sqcup$ and $\sqcap$ to hide old notation}
-A (non-indexed) comonadic semantics for the $\lambda$-calculus
-requires a \emph{monoidal comonad} with operation $\m_{A,B} : \F A
-\times \F B \rightarrow \F (A \times B)$~\cite{uustalu2008notions}.
-Previously, we defined a similar operation for the semantics of a flat
-coeffect system, with an indexed monoidal operation $\m^{R,S}_{A,B}$ for
-merging contexts. Dually, contexts were split with 
-$\n^{R,S}_{A,B}$~\cite{petricek2013coeffects}. We used two operations for combining and
-splitting the coeffect annotations, respectively. Here we generalize these 
-as $\amerge$ and $\asplit$ using a shape-indexed version.
-
-\begin{definition}
-$\D : \Sigma_{\sclrd{n} :
-    \SH}, (\aclrd{\mathbb{I}}^{\sclrd{n}} \rightarrow
-  [\mathbb{C}^{\sclrd{n}}, \mathbb{C}])$ is a 
-   \emph{indexed lax (semi)monoidal functor} and/or \emph{colax (semi)monoidal functor} if it has the following
- natural transformations respectively: 
-%%
-\begin{align*}
-\mathsf{m}^{n,m}_{R, S} & : \D^n_{R} A \times \D^m_{S} B \rightarrow \D^{n \stimes m}_{R \amerge S} (A \times B) 
-\\
-\mathsf{n}^{n,m}_{R, S} & : \D^{n \stimes m}_{R \asplit S} (A \times B)  
-\rightarrow \D^n_{R} A \times \D^m_{S} B 
-\end{align*}
-%
-In both, shape descriptions are combined by $\stimes$. The first operation models 
-context merging and combines coeffects using $\amerge$. The second models
-context splitting, with $\asplit$ for the pre-split coeffect. 
-\end{definition}
-
-\begin{example}[Bounded reuse]
-For bounded reuse, $\BLL$ is an indexed lax and colax semimonoidal functor
-with the following operations:
-\begin{align*}
-\m^{n,m}_{R,S} & = \lambda (\langle{a_1, ..., a_n}\rangle \times \langle{b_1, ..., b_m}\rangle) . (\langle{a_1, ..., a_n}\rangle, \langle{b_1, ..., b_m}\rangle)
- \\
-\n^{n,m}_{R,S} & = \lambda (\langle{a_1, ..., a_n}\rangle, \langle{b_1, ..., b_m}\rangle) . (\langle{a_1, ..., a_n}\rangle \times \langle{b_1, ..., b_m}\rangle)
-\end{align*}
-%%%
-\ie{}, %\tpnote{You are not using the tuple notation $(x,y)$ -- but you say so?}
- $\m^{m,m}_{R,S}$ takes a pair of contexts and merges them simply by replacing the 
-product in $\mathbb{C}$ which pairs the two arguments 
-(written using $\times$) with products inside of $\BLL$ (written using tuple notation $(x, y)$). 
-The operation $\n^{n,m}_{R,S}$ is the inverse.
-\end{example}
-
-\begin{example}[Implicit parameters]
-For implicit parameters, $\Imp^\star$ is an indexed lax and colax semimonoidal
-functor with operations:
-%%
-\begin{align*}
-\m^{\star,\star}_{R,S} & = \lambda ((a, \gamma_R), (b, \gamma_S)) . ((a, b), \gamma_R \cup \gamma_S) \\
-\n^{\star,\star}_{R,S} & = \lambda ((a, b), \gamma) . ((a, {\gamma|}_R), (b, {\gamma|}_S))
-\end{align*}
-%% 
-As in Example~\ref{exm:impl2}, ${\gamma|}_R$ and ${\gamma|}_S$ restrict the set of implicit parameters $\gamma$ to 
-the variable-values pairs for variables in $R$ and $S$. 
-\label{exm:impl3}
-\end{example}
-
-\subsection{Putting it together}
-\label{sem:together}
-The semantics of the structural coeffect calculus $\interp{-}$ is defined in Figure~\ref{fig:semantics},
-using the structures described in the previous sections.
-
-\paragraph{Core rules.}
-The denotation in (\emph{var}) maps a context of the singleton shape $\sunit$
-containing just a single variable $\tau$ (with coeffect $I$) to a $\tau$ value 
-using the counit operation.
-
-The premise of (\emph{abs}) takes a context of shape $\sclrd{n} \,\stimes\, \sunit$ with
-coeffects $\aclrd{R} \amerge \alift{\cclrd{s}}$ and a free-variables context consisting of $\Gamma$ and
-an additional variable $x$. The denotation $g : \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \amerge 
-  \alift{\cclrd{s}}} \interp{\Gamma, v : \sigma} \rightarrow \interp{\tau}$ is
-pre-composed with $\m$, such that its context
-is obtained by merging the declaration-site context ($\Gamma$) and call-site context ($\sigma$):
-%%
-$$
-g \circ \m^{\sclrd{n},\sunit}_{\aclrd{R}, \alift{\cclrd{s}}} : 
-(\D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \times \D^\sunit_{\alift{\cclrd{s}}} \interp{\sigma}) \rightarrow \interp{\tau}
-$$
-%%
-This is uncurried to give a denotation from a context to an exponential object
-representing the abstraction, where the singleton-shaped context becomes the source of the exponential.
-
-% --------------------------------------------------------------------------------------------------
-% Figure - semantics
-% --------------------------------------------------------------------------------------------------
-
-\newcommand{\exchgl}{R \amerge \langle{t}\rangle \amerge \langle{s}\rangle \amerge Q}
-\newcommand{\exchgr}{R \amerge \langle{s}\rangle \amerge \langle{t}\rangle \amerge Q}
-\newcommand{\contrl}{R \amerge \langle{s}\rangle \amerge \langle{t}\rangle \amerge Q}
-\newcommand{\contrr}{R \amerge \langle{s \cpar t}\rangle \amerge Q}
-\begin{figure*}[t]
-\vspace{-0.5em}
-\begin{align*}
-\begin{array}{c}
-%
-% VAR semantics 
-%
-\tyrule{var}
-  {\quad}
-  {\interp{\coctx{x : \tau}{ \alift{\cunit} } \vdash x : \tau} = \varepsilon_I : \D^\sunit_I \interp{\tau} \rightarrow \interp{\tau}}
-\qquad
-%
-% ABS semantics
-%
-\tyrule{abs}
-  {\interp{\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \amerge \alift{\cclrd{s}}} \vdash e : \tau} =
-              g : \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \amerge {\alift{\cclrd{s}}}} \interp{\Gamma, x \!:\! \sigma} \rightarrow \interp{\tau}}
-  {\interp{\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
-                = \Lambda (g \circ \m^{\sclrd{n},\sunit}_{\aclrd{R}, \alift{\cclrd{s}}})
-: \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \rightarrow (\D^\sunit_{\alift{\cclrd{s}}} \interp{\sigma} \Rightarrow \interp{\tau})}
-\\[1.8em]
-%
-% APP semantics
-%
-\tyrule{app}
-   { \interp{\coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau} 
-        = g_1 : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \rightarrow (\D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma} \Rightarrow \interp{\tau})
-     \quad\qquad   
-     \interp{\coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \tau} 
-        = g_2 : \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} \rightarrow \interp{\sigma} }
-   { \interp{\coctx{\Gamma_1, \Gamma_2}{\aclrd{R}\ \asplit\ (\cclrd{t}\ \aseq\ \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
-        = \Lambda^{-1} g_1 \circ (\mathsf{id} \times (\D^{\sunit}_{\alift{\cclrd{t}}} g_2 \circ \delta^{\sclrd{m}}_{\alift{\cclrd{t}}, \aclrd{S}})) \circ \n^{\sclrd{n},\sclrd{m}}_{\aclrd{R}, \cclrd{t} \aseq \aclrd{S}}
-: \D^{\sclrd{n} \stimes \sclrd{m}}_{\aclrd{R} \asplit (\cclrd{t} \aseq \aclrd{S})} \interp{\Gamma_1, \Gamma_2} \rightarrow \interp{\tau} }
-\\[2em]
-\tyrule{ctx}
-  { \interp{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} = f : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau} 
-    \qquad \interp{\ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma'}{\aclrd{R'}} }{\emptyset}} = c : \D^{\sclrd{m}}_{\aclrd{R'}} \interp{\Gamma'}
-\rightarrow \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma}
-  }
-  { \interp{\coctx{\Gamma'}{\aclrd{R'}} \vdash e : \tau} = f \circ c : \D^{\sclrd{m}}_{\aclrd{R'}} \interp{\Gamma'} \rightarrow \interp{\tau}}
-\end{array}
-\end{align*}
-%
-%
-% -------------- WEAK --------------
-%
-\begin{equation*}
-{\text{\footnotesize{(weak)}}}\quad
-\; \interp{\ctxtrans{\coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x\!:\!\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}}
-%
-% Denotation
-%
-= \pi_1 \circ \n^{\sclrd{n},\sunit}_{\aclrd{R},\aweak}
-: \D^{\sclrd{n} \stimes \sunit}_{\aclrd{R} \asplit \aweak} (\interp{\Gamma} \times \interp{\sigma}) \rightarrow \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma}
-\end{equation*}
-%
-%
-% -------------- CONTR --------------
-%
-\vspace{-1em}
-\begin{equation*}
-{\text{\footnotesize{(contr)}}} \quad
-\interp{
-  \ctxtrans
-    {\coctx{\Gamma_1, y\!:\!\tau, z\!:\!\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
-    {\coctx{\Gamma_1, x\!:\!\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]} }
-%
-% Denotation
-%
- = \m^{\sclrd{n},\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s} \cpar \cclrd{t}},\aclrd{Q}} \circ (\mathsf{id} \times \Delta_{\cclrd{s},\cclrd{t}} \times \mathsf{id}) \circ
-\n^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s}},\alift{\cclrd{t}},\aclrd{Q}} \\[1.1em]
-\end{equation*}
-% 
-%
-% -------------- EXCH --------------
-%
-\vspace{-2em}
-\begin{equation*}
-{\text{\footnotesize{(exch)}}} \quad
-\interp{
-  \ctxtrans
-     {\coctx{\Gamma_1, x\!:\!\tau, y\!:\!\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
-     {\coctx{\Gamma_1, y\!:\!\sigma, x\!:\!\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }
-       {\emptyset} }  
-%
-% Denotation
-%
- = \m^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{s}},\alift{\cclrd{t}},\aclrd{Q}} \circ (\mathsf{id} \times \mathsf{swap} \times \mathsf{id}) \circ \n^{\sclrd{n},\sunit,\sunit,\sclrd{m}}_{\aclrd{R},\alift{\cclrd{t}},\alift{\cclrd{s}},\aclrd{Q}} \\[0.7em]
-\end{equation*}
-%
-% ------------------ Definitions ---------------------
-%
-\vspace{-1.5em}
-\begin{equation*}
-\textbf{where}~~ \mathsf{swap} : A \times B \rightarrow B \times A~~~\textbf{and}~~
-  \Lambda, \Lambda^{-1} ~~\textnormal{denote currying and uncurrying respectively}
-\end{equation*}
-\vspace{-0.5em}
-\caption{Denotational semantics for the coeffect calculus}
-\label{fig:semantics}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-The application rule (\emph{app})  has two sub-expressions for the function and argument, with denotations
-requiring two distinct contexts:
-%%
-\begin{align*}
-g_1 : \D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \rightarrow (\D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma} \Rightarrow \interp{\tau})
-\qquad
-g_2 : \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} \rightarrow \interp{\sigma}
-\end{align*}
-%%
-The target of $g_1$ is an exponential object with singleton shape for the parameter of type $\sigma$. 
-To evaluate $g_1$ and $g_2$, the semantics of (\emph{app}) 
-splits the incoming context over $\Gamma_1, \Gamma_2$ using $\n$:
-%% 
-\begin{align*}
-\D^{\sclrd{n} \stimes \sclrd{m}}_{\aclrd{R}\,\asplit\,(\alift{\cclrd{t}}\,\bbullet\,\aclrd{S})}
-(\interp{\Gamma_1} \times \interp{\Gamma_2})
-\xrightarrow{\n^{\sclrd{n},\sclrd{m}}_{\aclrd{R}, \alift{\cclrd{t}}\bbullet\aclrd{S}}} 
-\D^{\sclrd{n}}_{\aclrd{R}} \interp{\Gamma_1} \times \D^{\sclrd{m}}_{\alift{\cclrd{t}} \bbullet \aclrd{S}} \interp{\Gamma_2}
-\end{align*}
-%%
-Since $e_2$ computes the parameter for function $e_1$, the denotation
-$g_2$ must be sequentially composed with the parameter part of
-$g_1$. Thus, the structural indexed comonad is used with $g_2$
-to compute the correct context for the parameter of $g_1$:
-%
-\begin{align*}
-\D^{\sclrd{m}}_{\alift{\cclrd{t}} \bbullet \aclrd{S}} \interp{\Gamma_2} 
-    \xrightarrow{\delta^{\sclrd{m}}_{\cclrd{t},\aclrd{S}}} 
- \D^\sunit_{\alift{\cclrd{t}}} \D^{\sclrd{m}}_{\aclrd{S}} \interp{\Gamma_2} 
-    \xrightarrow{\D^\sunit_{\alift{\cclrd{t}}} g_2} \D^\sunit_{\alift{\cclrd{t}}} \interp{\sigma}
-%\label{eq:app-morph2}
-\end{align*}
-%
-This is composed with the previous equation by lifting
-to the right-component of the product:
-%
-\begin{align*}
-\D^{\sclrd{n}}_R \interp{\Gamma_1} \times \D^{\sclrd{m}}_S \interp{\Gamma_2}
-\xrightarrow{id \times (\D^\sunit_{\alift{\cclrd{t}}} g_2 \, \circ \, \delta^{\sclrd{m}}_{t,S})}
-\D^{\sclrd{n}}_R \interp{\Gamma_1} \times \D^{\sunit}_{\alift{\cclrd{t}}}  \interp{\sigma}
-\end{align*}
-%%
-This equation computes the calling context and
-parameter context for the function $e_1$, which is then composed with
-the uncurried $g_1$ denotation as shown in the (\emph{app}) rule in Figure~\ref{fig:semantics}.
-
-\paragraph{Structural rules.}
-
-%The second part of Figure~\ref{fig:semantics} shows the semantics of the
-%structural rules for the coeffect calculus. 
-In Figure~\ref{fig:semantics}, (\emph{ctx}) composes the
-denotation of an expression with a transformation $c$ providing the semantic 
- structural rules. 
-The semantics of structural rules are defined by using
-$\n^{n,m}_{R,S}$ to split contexts, transforming the components, and merging the transformed
-contexts using $\m^{n,m}_{R,S}$. The (\emph{contr}) rule uses an additional operation
-which duplicates a variable inside a context:
-%%
-\begin{align*}
-\Delta_{\cclrd{r},\cclrd{s}} & : \D^{\sunit}_{\cclrd{r} \cpar \cclrd{s}} A
-    \rightarrow \D^{\sunit \stimes \sunit}_{\alift{\cclrd{r}} \amerge \alift{\cclrd{s}}}(A \times A)
-\end{align*}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{example}
-We demonstrate the semantics with a concrete example 
-for the bounded reuse calculus.
-Consider the following:
-%%
-$$
-\coctx{f:\mathbb{Z}\xrightarrow{2}\mathbb{Z}, x:\mathbb{Z}}{\langle 2, 4 \rangle } \vdash (\lambda z.z+z)~(f\ x)
-$$
-We elide the function body's denotation prior to contraction,
- $g = \interp{x : \mathbb{Z}, y : \mathbb{Z} \vdash (+x) y : \mathbb{Z}}$\footnote{
-the full semantics has $\interp{+} : \D^0_{\langle{}\rangle} 1 \rightarrow (\D^1_1
-\mathbb{Z} \Rightarrow (\D^1_1 \mathbb{Z} \Rightarrow \mathbb{Z}))$}. %  as primitive and uses double application $(+\ e_1) e_2$
-The term therefore has denotation: 
-%%
-\begin{align}
-\label{eq:exm-denot5}
-& \interp{\coctx{}{\langle  \rangle} \vdash
-\lambda z . (+ z) z : \mathbb{Z} \xrightarrow{2} \mathbb{Z}} = \Lambda (g \circ \Delta_{\langle{1}\rangle,\langle{1}\rangle} \circ \m^{0,1}_{\langle{}\rangle,\langle{1}\rangle}) \\
-\notag & \interp{\coctx{f : \mathbb{Z} \xrightarrow{2} \mathbb{Z}, x : \mathbb{Z}}{\langle{1, 2}\rangle} \vdash 
-f x : \mathbb{Z}} \\
-%\label{eq:exm-denot6} 
-\notag
-& \hspace{1.5em} = \uncurry \varepsilon_1 \circ (id \times (\D \varepsilon_1 \circ \delta^1_{2, \langle{1}\rangle})) \circ \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle} \\ 
-\label{eq:exm-denot6} 
-& \hspace{1.5em} = \uncurry \varepsilon_1 \circ \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle} \\ 
-\notag 
-& \interp{\coctx{f : \mathbb{Z} \xrightarrow{2} \mathbb{Z},
-x : \mathbb{Z}}{\langle{2, 4}\rangle}
-\vdash 
-(\lambda z . (+ z) z)\ (f x) : \mathbb{Z}} 
-\\
-\notag
-& \hspace{1.5em} = \uncurry \eqref{eq:exm-denot5} \circ (id \times \D \eqref{eq:exm-denot6} 
-\circ \delta^2_{2,\langle{2,1}\rangle}) 
-\circ \n^{0,2}_{\langle{}\rangle,{\langle{4,2}\rangle}}
-\\
-\label{eq:exm-denot-final}
-& \hspace{1.5em} = g \circ \Delta_{r,s} \circ \m^{0,1}_{\langle{}\rangle,\langle{1}\rangle} \circ (id \times \D \eqref{eq:exm-denot6} 
-\circ \delta^2_{2,\langle{2,1}\rangle}) 
-\circ \n^{0,2}_{\langle{}\rangle,{\langle{4,2}\rangle}}
-\end{align}
-%%
-where \eqref{eq:exm-denot6} and \eqref{eq:exm-denot-final} are simplified. 
-We ``run'' this semantics on some input, evaluating each step of the denotation as a function. 
-%(\ie{}, we take $\mathbb{C} = \mathbf{Set}$). 
-We write context objects, \eg{}, $\D^2_{\langle{R, S}\rangle} (A, B)$ as $\langle(a_1, ..., a_R), (b_1, ..., b_R)\rangle$ and products of
- contexts in $\mathbb{C}$, \eg{},  $\D^n_R A \times \D^m_S B$, as $(a \times b)$.
-
-\vspace{0.2em}
-\newcommand{\arw}[1]{\makebox[4.3em]{\scriptsize{${#1}$}}}
-\hspace{-2.4em}
-{\scalebox{0.9}{
-\begin{minipage}{1.04\linewidth}
-\begin{align*}
-\begin{array}{c l l}
-%%%% START
-\hspace{-0.5em} & \langle(f_1, f_2),(x_1, x_2, x_3, x_4)\rangle  &\hspace{-5em} : \D^{2}_{\langle{2,4}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z}) \\[-0.5em]
-%%%% STAGE 1
-\xrightarrow{\arw{{n}^{0,2}_{\langle{}\rangle,\langle{2,4}\rangle}}} \hspace{-0.5em} &
- \langle\rangle \times \langle(f_1, f_2), (x_1, x_2, x_3, x_4)\rangle &  \hspace{-3em}: \D^0_{\langle{}\rangle} 1 \times \D^{2}_{\langle{2,4}\rangle} (\textit{as above}) \\[-0.3em]
-%%
-%\hspace{-0.5em} & 
-%\multicolumn{2}{l}{
-%\hspace{5.91em} : \D^0_{\langle{}\rangle} 1 \times \D^{2}_{\langle{2,4}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z}) 
-%}
-%\\
-%%%% STAGE 2
-\xrightarrow{\arw{\id \times \delta^2_{2, \langle{1, 2}\rangle}}} \hspace{-0.5em} & 
- \langle\rangle \times \langle\langle f_1, (x_1, x_2)\rangle, \langle f_2, (x_3, x_4)\rangle\rangle \\[0.33em]
-%%
-\hspace{-0.5em} & 
-\multicolumn{2}{l}{
-\hspace{5.91em}
-: \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} \D^{2}_{\langle{2,1}\rangle} ((\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \mathbb{Z})} \\[-0.3em]
-%%%% STAGE 3
-\xrightarrow{\arw{\id \times \D \n^{1,1}_{\langle{1}\rangle,\langle{2}\rangle}}} \hspace{-0.5em} &
- \langle\rangle \times \langle\langle f_1\rangle \times \langle x_1, x_2\rangle, \langle f_2\rangle \times \langle x_3, x_4\rangle\rangle \\[0.34em]
-%%
-\hspace{-0.5em} & 
-\multicolumn{2}{l}{
-\hspace{5.91em} 
-: \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} (\D^{1}_{\langle{1}\rangle} (\D^1_{\langle{2}\rangle} \mathbb{Z} \Rightarrow \mathbb{Z}) \times \D^{1}_{\langle{2}\rangle} \mathbb{Z})} \\[-0.1em]
-%%%% STAGE 4
-\xrightarrow{\arw{\id \times \D (\uncurry \varepsilon_1)}}  
-\hspace{-0.5em} & \langle\rangle \times \langle f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em} : \D^0_{\langle{}\rangle} 1 \times \D^{1}_{\langle{2}\rangle} \mathbb{Z} \\[-0.2em]
-%%%% STAGE 5
-\xrightarrow{\arw{\m^{0,1}_{\langle{}\rangle,\langle{1}\rangle}}} 
-\hspace{-0.5em} & \langle f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em}  : \D^1_{\langle{2}\rangle} \mathbb{Z} \\
-%%%% STAGE 6
-\xrightarrow{\arw{\Delta^2_{\langle{1}\rangle,\langle{1}\rangle}}}
-\hspace{-0.5em} & \langle(f_1 \langle x_1, x_2\rangle, f_2 \langle x_3, x_4\rangle)\rangle & \hspace{-3.5em} : \D^1_{\langle{1,1}\rangle} (\mathbb{Z} \times \mathbb{Z}) \\[0.54em]
-%%%%% STAGE 7
-\xrightarrow{\arw{g}} 
- \hspace{-0.5em} & \interp{+}\ \langle f_1 \langle x_1, x_2\rangle\rangle\ \langle f_2 \langle x_3, x_4\rangle\rangle & \hspace{-3.5em} : \mathbb{Z} 
-\end{array}
-\end{align*}
-\end{minipage}
-}}
-\label{exm:run}
-%This demonstrates the successive splitting, merging, and composing of contexts to the final result. 
-\end{example}
-
-\subsection{Soundness, with respect to equational theory}
-\label{sec:semantics-soundness}
-
-\noindent
-Our denotational semantics for the coeffect
-calculus is sound with respect the equational theory of Section~\ref{sec:syntax}. That is:
-%%
-\begin{theorem}[Soundness]
-$$\coctx{\Gamma}{R} \vdash e \equiv e' : \tau \Rightarrow \interp{\coctx{\Gamma}{R} \vdash e : \tau} 
-\equiv \interp{\coctx{\Gamma}{R} \vdash e' : \tau}$$
-\end{theorem}
-%%
-\noindent
-Proof of this follows from an interesting result which we 
-first unpack: determining whether $\interp{\coctx{\Gamma}{R} \vdash
-  e_1 : \tau} {\equiv} \interp{\coctx{\Gamma}{S} \vdash e_2 : \tau}$ follows
-from a proof on coeffect annotations that $R = S$. 
-%
-\begin{lemma}
-Every coeffect algebra axiom corresponds to an axiom 
-of one of the categorical structures introduced here (indexed (structural) comonad or 
-  indexed (co)lax monoidal functor). 
-\end{lemma}
-%
-\noindent
-For example, the monoid axiom $X \cseq \cunit =
-X$ for scalar coeffects corresponds to indexed comonad axiom $\D^{\sunit}_X \varepsilon_{\cunit}
-\circ \delta^{\sunit}_{X,{\cunit}} = id_{\D^{\sunit}_X}$ (which requires the monoid axiom to hold). 
-This lemma follows from our derivation of 
-the indexed categorical structures here. They are not derived \emph{ad hoc} but systematically 
-as (lax) \emph{homomorphisms} (structure preserving maps) between the structure of 
-coeffect annotations in $\Icat$ and the structure of denotations in $\mathbb{C}$.
-
-\begin{proposition}
-An indexed comonad on $\D$ witnesses that
-$\D$ is a \emph{colax monoid homomorphism} between the (strict)
-monoidal categories $(\mathbb{I}, \cseq, \cunit)$ and 
-$([\mathbb{C}, \mathbb{C}], 
-\circ, 1_{\mathbb{C}})$ (endofunctor composition). 
-\end{proposition}
-%%
-Unpacking this, a \emph{monoid homomorphism} maps between
-the underlying sets of two monoids, preserving the monoid structure of one into the other,
-\ie{}, given monoids $(X, \bullet, I)$ and $(Y, \otimes, E)$ then
-a monoid homomorphism is a mapping $F : X \rightarrow Y$ such that:
-%%
-\begin{align*}
-F X \otimes F Y \equiv F (X \bullet Y) \quad\qquad E \equiv F I
-\end{align*}
-%%
-The axioms of each monoid are preserved trivially by these equalities, 
-\eg{}, $F X \equiv F (X \bullet I) \equiv F X \otimes F I \equiv F X \otimes E \equiv F X$. 
-In a categorical setting, monoids are taken on categories with binary operations as 
-bifunctors. A homomorphism is \emph{lax} if the above equalities are instead morphisms 
-(which we say \emph{witness} the homomorphism) and \emph{colax} if these morphisms go in opposite direction.
-Thus, a colax monoid homomorphism is witnessed by: 
-%%
-\begin{align*}
-\delta : F X \otimes F Y \leftarrow F (X \bullet Y) \quad\qquad \varepsilon : E \leftarrow F I
-\end{align*}
-%%
-Note our choice of morphism names.  
-$F$ no longer preserves the monoid axioms \emph{up to equality} but has axioms on $\delta$ and $\varepsilon$, %% F X {=\joinrel=} 
-\eg{}, $F (X \bullet I) \xrightarrow{\delta} F X \otimes F I \xrightarrow{id \otimes \varepsilon} F X \otimes E$ equals $F X \xrightarrow{id} F X$. 
-%%  {=\joinrel=} F X
-
-Our indexed comonad definition is equivalent to $\D$ being a colax homomorphism between a monoidal category $(X, \bullet, I)$ 
-and the monoidal category of $\mathbb{C}$ endofunctors $(Y, \otimes, E) = ([\mathbb{C}, \mathbb{C}], \circ, 1_{\mathbb{C}})$, with endofunctor composition 
-$\circ$ and the trivial endofunctor $1_{\mathbb{C}}$. The indexed comonads axioms are the axioms of the colax homomorphism.
-Equivalently, $\D$ is a \emph{colax monoidal functor}.
-
-A similar approach is taken to deriving the remaining structures.%, though we give less detail for brevity.
-
-\begin{proposition}
-  A structural indexed comonad provides $\delta^n_{R,S} :
-  \D^\sunit_r \D^n_S A^n \leftarrow \D^n_{r \aseq S} A^n$ which witnesses that $\D$ is a 
-  colax homomorphism between the following \emph{monoid left-actions} for
-  $(\mathbb{I}, \cseq, \czero)$ and $([\mathbb{C}, \mathbb{C}], \circ,
-  1_{\mathbb{C}})$: 
-\begin{align*}
-\begin{array}{rclcl}
-(r : \mathbb{I}) & \narrow{\aseq} & (\langle{s_1, ..., s_n}\rangle : \mathbb{I}^n) &\narrow{=}& \langle{r \cseq s_1, ..., r \cseq s_n}\rangle : \mathbb{I}^n  \\
-(\D^\sunit_r : [\mathbb{C}, \mathbb{C}]) & \narrow{\circ} & (\D^n_S : [\mathbb{C}^n, \mathbb{C}]) &\narrow{=}& \D^\sunit_r \D^n_S : [\mathbb{C}^n, \mathbb{C}]
-\end{array}
-%(x : \mathbb{I}) \aseq (\langle{y_1, \ldots, y_n}\rangle : \mathbb{I}^n) & = \langle{x \cseq y_1, \ldots, x \cseq y_n}\rangle  \\
-%(\D^\sunit_r : \mathbb{C} \rightarrow \mathbb{C}) \circ (\D^n_S : \mathbb{C}^n \rightarrow \mathbb{C}) & = \D^\sunit_R \D^n_S : \mathbb{C}^n \rightarrow \mathbb{C}
-\end{align*}
-The axioms are the lax versions of the monoid left-action laws. 
-%witnesses that $\D$
-%is a colax homomorphism between the induced \emph{monoid left-action} for $(\mathbb{I}, \cseq, \czero)$
-%of $n$-length coeffect vectors $(\mathbb{I}^n, \aseq_n : \mathbb{I} \times \mathbb{I}^n 
-%\rightarrow \mathbb{I}^n)$ for monoid, and a trivial monoid left-action for $([\mathbb{C}, \mathbb{C}], \circ, 1_{\mathbb{C}})$ 
-%defined by composition $([\mathbb{C}^n, \mathbb{C}], \circ : [\mathbb{C}, \mathbb{C}] \times [\mathbb{C}^n, \mathbb{C}] \rightarrow [\mathbb{C}^n, 
-%\mathbb{C}])$, 
-%\ie{},
-\end{proposition}
-%%
-The lax and colax indexed monoidal operations $\m^{n,m}_{R,S}$ and $\n^{n,m}_{R,S}$ follow a similar derivation
-but as lax and colax monoid homomorphism between composite monoids on coeffect annotations and shapes
-and $\times$ in $\mathbb{C}$. The details are elided here.
-
-Returning to soundness, our semantics is therefore defined in terms of structures whose axioms correspond 
-to axioms of the syntactic equational theory. Consequently, semantic proofs correspond to 
-syntactic proofs, modulo naturality laws and product/exponent laws in $\mathbb{C}$. 
-This result holds in the general coeffect calculus and semantics since every semantic structure
-has a unique corresponding structure on coeffect annotations (\ie{}, $(\C, \cseq, \cunit)$ for
-sequential composition of unary denotations, $(\C, \asplit)$ for splitting contexts, 
-$(\C, \amerge)$ for joining contexts).
-
-\begin{example}
-Section~\ref{prop:structural} showed $\eta$-equality for structural systems,
-which uses the properties (1) $\asplit = \amerge = \times$ for structural systems and (2) 
-$s \aseq \langle{\cunit}\rangle
-= \langle{s \cseq \cunit}\rangle = \langle{s}\rangle$. The semantics here is sound
-with respect to $\eta$-equality, the proof of which uses the corresponding axioms
-(1) $\n^{n,m}_{R,S} \circ \m^{n,m}_{R,S} = id$ and (2) $\D^{\sunit}_s \varepsilon_{\cunit} \circ \delta^m_{s,\cunit}$
-(structural indexed comonad unit law, Definition~\ref{def:structural-indexed-comonad}). 
-% $\D^{n}_R A \times \D^{m}_S B \xrightarrow{\m^{n,m}_{R,S}} \D^{n,m}_{R \amerge S} (A \times B) \equiv
-% \D^{n,m}_{R \asplit S} (A \times B) \xrightarrow{\n^{n,m}_{R,S}} \D^n_R A \times \D^m_S$ (\ie{}, $\n \circ \m = id$) and 
-\end{example}
-%
-\noindent
-The accompanying technical report shows the full semantic proofs for $\beta\eta$-equality whose structure corresponds
-exactly to the syntactic proofs with the corresponding coherence conditions~\cite{appendix}.
-
-%\begin{prosition}
-%A lax indexed monoidal functor on $\D$ witnesses that $\D$
-%is a lax monoid homomorphism between the $(\mathbb{I}, \amerge, \azero)$ and 
-%a composite monoid of $
-%\end{proposition}
-
-%The indexed comonad structure is derived from a
-%\emph{colax} \emph{homomorphism} between the monoidal structure
-%$(\mathbb{I}, \bullet, I)$ on indices and monoidal structure of
-%endofunctor composition. This approach is used to derive the rest
-%of our semantics, though we keep the details of the derivation light
-%here and present instead the resulting structures.
-
-%\begin{remark}
-%The category of $\mathbb{C}$-endofunctors $[\mathbb{C}, \mathbb{C}]$
-%is a \emph{strict} monoidal category $([\mathbb{C}, \mathbb{C}], \circ,
-%1_{\mathbb{C}})$ with composition $\times$ as the binary operation and the identity
-%functor $1_{\mathbb{C}}$ as the unit. 
-%\end{remark}
-
-% ==================================================================================================
-
-\section{Related work}
-\label{sec:related-work}
-
-We expand briefly on the overview of related work in 
-Section~\ref{sec:intro-pathways}.
-
-%\paragraph{Bounded reuse.}
-The (\emph{storage}) rule for bounded linear logic explains the contextual requirements 
-induced by proposition reuse~\cite{girard1992bounded}:
-$$
-\inference[\text{(storage)}]{!_{\overline{Y}} \Gamma \vdash A}{!_{X\overline{Y}} \Gamma \vdash !_X A} 
-$$
-where $X\overline{Y} = \langle{XY_1, .., XY_n}\rangle$ is the scalar multiple of a vector. 
-This rule is akin to the $\delta^n$ operation of structural indexed comonads,
-indeed, we can model it exactly using $\delta^{n}_{X,\bar{Y}}$ and the lifting $\D^n_X$.
-
-In BLL, the modality $!_X$ is a constructor and may 
-  appear both on the left- and right-hand sides of $\vdash$.
-  In this paper, reuse bounds 
-  annotate typing rules, thus there is no constructor
-  corresponding to bounded reuse in the language; reuse bounds are
-  meta-level. Our choice to work at the meta-level means that the 
-  coeffect calculus provides a unified analysis and semantics to
-  different notions of context and its term language is that of standard $\lambda$-calculus.
-
-%\paragraph{Semantics.}
-Previously we briefly introduced indexed comonads~\cite{petricek2013coeffects}
-without derivation.  Here we derived indexed comonads as colax
-homomorphisms. This is dual to the \emph{parametric effect monad}
-structure defined as a lax
-homomorphism~\cite{katsumata2014parametric}. Our semantics 
-requires additional structure not needed for effects due to the
-asymmetry inherent in the $\lambda$-calculus.
-
-The necessity modality $\Box$ in S4 logic corresponds to a comonad
-with lax monoidal functor structure $\m : \Box A \times \Box B \rightarrow \Box (A \times B)$. 
-Bierman and de Paiva \cite{bierman2000intuitionistic} defined a term language corresponding to a natural deduction S4, where contexts contain sequences of $\Box$-wrapped assumptions
-$x_1 \!:\! \Box A_1, \ldots x_n \!:\! \Box A_n$. Modelling these judgments does not require a context splitting operation $\n$ 
-unlike in our approach. Our approach can be thought of as having a single $\Box$ modality over the context 
-which allows both flat whole-context dependence and structural per-variable dependence.
-
-%\noindent
-%\todo{Discuss related work not covered in the background section. More details on the difference
-%between our work and modal logic approaches. etc!}
-
 \section{Conclusions}
-
-In this paper, we looked at two forms of context-dependence
-analysis -- \emph{flat} coeffect systems that track whole-context
-requirements (such as implicit parameters, resources, or platform version) and
-\emph{structural} coeffects that track per-variable requirements (such
-as usage or data access patterns). 
-The newly introduced structural system makes applications such as liveness,
-bounded reuse, and dataflow analysis (from our earlier work)
-practically useful. With the move towards cross-platform systems running in 
-diverse environments, analysing context dependence is vital for reasoning and compilation. The coeffect calculus
-provides a foundation for further study, similar to the type-and-effect discipline.
-
-We presented the system together with its syntactic equational theory and categorical semantics.
-The equational theory is presented in order to explain how the systems work, but it also
-provides a basis for an operational semantics for concrete systems. Exploring these, 
-and their connection to the denotational semantics, is further work.

### 2014-09-19 text/Chapter07.tex
Cleanup files, fix some invalid references

@@
-\chapter{Coeffect meta-language} 
-\label{ch:meta}
+%!TEX root = ../main.tex
+
+\chapter{Conclusions} 
+\label{ch:conclusions}
+
+%---------------------------------------------------------------------------------------------------
+
+How to make this practical?
+
+codo notation / computation expressions
+
+embedding algebraic things
 
@@
 
+\newcommand{\munit}{\mathsf{e}}
+\newcommand{\mseq}[0]{ \oplus }
+
 \section{Introduction}
-\label{sec:metalanguage-intro}
+
+\section{Meta-language}
+\label{sec:conclusions-meta}
+
+\begin{figure}
+\begin{equation*}
+\tyrule{var}
+  {x : \alpha \in \Gamma}
+  {\Gamma \vdash x : \alpha }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \alpha \rightarrow \beta &
+   \Gamma \vdash e_2 : \alpha }
+  {\Gamma \vdash e_1~e_2 : \beta}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x:\alpha \vdash e : \beta}
+  {\Gamma \vdash \lambda x.e : \alpha \rightarrow \beta }
+\end{equation*}
+\begin{equation*}
+\tyrule{letbox}
+  {\Gamma \vdash e_1 : \ctyp{r \mseq s}{\alpha} & \Gamma, x : \ctyp{r}{\alpha} \vdash e_2 : \beta }
+  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{s}{\beta}}
+\end{equation*}
+\begin{equation*}
+\tyrule{eval}
+  {\Gamma \vdash e : \ctyp{\munit}{\alpha}}
+  {\Gamma \vdash !e : \alpha}
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\Gamma \vdash e : \ctyp{s}{\alpha} }
+  {\Gamma \vdash e : \ctyp{r}{\alpha} }
+\;(s\leq r)  
+\end{equation*}
+
+\caption{Type system for the coeffect meta-language cml}
+\label{fig:cml-types}
+\end{figure}
 
 % ==================================================================================================
-
-\section{Type inference}
-\label{sec:flat-inference}
-
-\subsection{Semi-lattice formulation}
-\subsection{Type inference algorithm}
-
-% ==================================================================================================
-
-\section{Type system}
-\ExecuteMetaData[rules/rules.tex]{cml-types}
-
-\section{Operational properties}
-
-\section{Categorical semantics}
-
-\section{Applications}
-
-\subsection{Meta-programming}
-
-\subsection{Mobile computations}
 
@@
 \section{Also related work}
-\section{Summary}
 
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\newcommand{\cprd}{\times}
-\newcommand{\cvop}{\oplus}
-
-\vspace{-1em}
-\paragraph{Structural coeffects.} To make the liveness analysis practical, we need to associate
-information with individual variables (rather than the entire context). We can generalize the 
-calculus from this paper by adding a product operation $\times$ to the coeffect algebra.
-A variable context $x:\tau_1, y:\tau_2, z:\tau_3$ is then annotated with
-$r\times s \times t$ where each component of the tag corresponds to a single variable. The system
-then needs to be extended with structural rules such as:
-%
-\begin{equation*}
-\inference[(\emph{abs})]
-  {\ctyp{r \cprd s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
-\quad
-\inference[(\emph{contr})]
-  {\ctyp{r \cprd s}{(x:\tau_1, y:\tau_1)} \vdash e : \tau_2}
-  {\ctyp{r \cvop s }{(z:\tau_1)} \vdash \subst{\subst{e}{x}{z}}{y}{z} : \tau_2 }
-\end{equation*}
-%
-The context-requirements associated with function are exactly those linked to the specific
-variable of the lambda abstraction. Rules such as contraction manipulate variables and perform
-a corresponding operation on the indices.
-
-The structural coeffect system is related to bunched typing \cite{types-bunched} (but generalizes
-it by adding indices). We are currently investigating how to use structural coeffects to capture 
-fine-grained context-dependence properties such as secure information flow \cite{app-secure-flow}
-or, more generally, those captured by dependency core calculus \cite{types-dcc}.
-
 \section{Summary}

### 2014-09-22 text/Chapter07.tex
Starting the final chapter (Chapter 7)

@@
 
-\chapter{Conclusions} 
+\chapter{Further work and conclusions} 
 \label{ch:conclusions}
 
-%---------------------------------------------------------------------------------------------------
+The main goal of this thesis is to demonstrate that there is a unified notion of \emph{context}
+that captures many common scenarios in computer programming. The examples we explore include
+many notions of context important in modern software development (including distributed and
+cross-platform development) as well as context studied in fundamental programming language 
+research (including data-flow, linear logics and variable liveness).
 
-How to make this practical?
+This thesis demonstrates that all these notions of context have something in common. In 
+particular, they can be modelled by our flat or structural coeffect calculi. Although the 
+results presented in this thesis are mainly of a theoretical nature, we, indeed, believe that 
+coeffects should be integrated in main-stream programming languages.
 
-codo notation / computation expressions
+In this chapter, we present further work in two directions. Firstly, we outline one possible
+approach for practical implementations of coeffects (Section~\ref{sec:conclusions-impl}). 
+Secondly, we discuss an alternative approach to defining coeffect systems which highlights
+the relationship between our work and related work arising from modal logics 
+(Section~\ref{sec:conclusions-meta}). Finally, the Section~\ref{sec:conclusions-summary} summarizes
+and concludes the thesis.
 
-embedding algebraic things
+% =================================================================================================
+%                                                                   
+%      ####                         #       #                   ##   
+%      #   #                        #                            #   
+%      #   #  # ##    ###    ###   ####    ##     ###    ###     #   
+%      ####   ##  #      #  #   #   #       #    #   #      #    #   
+%      #      #       ####  #       #       #    #       ####    #   
+%      #      #      #   #  #   #   #  #    #    #   #  #   #    #   
+%      #      #       ####   ###     ##    ###    ###    ####   ###  
+%                                                                
+% =================================================================================================
 
-%---------------------------------------------------------------------------------------------------
+\section{Towards practical coeffects}
+\label{sec:conclusions-impl}
+
+As discussed earlier, the main focus of this thesis is the development of the much needed 
+\emph{theory of context-aware computations} and so discussing the details of a practical 
+implementation of coeffect tracking is beyond the scope of the thesis. However, this section
+briefly outlines one possible pathway towards this goal.
+
+Many of the examples of contextual computation that we discussed earlier (\eg~implicit
+parameters \cite{app-implicit-parameters} or distributed computations 
+\cite{app-distributed-ml5,app-distributed-links}) have been implemented as a single-purpose
+programming language feature. However, the main contribution of this thesis is that it captures
+\emph{multiple} different notions of context-aware computations using just a \emph{single}
+common abstraction. For this reason, we advocate that future practical implementations of
+coeffects should not be single-purpose language features, but instead reusable abstractions that
+can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
+
+In order to do this, programming languages need to provide two features; one that allows 
+embedding of context-aware computations themselves in programs (akin to the ``do'' notation
+in Haskell) and one that allows tracking of the contextual information in the type system.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding contextual computations}
+
+The embedding of \emph{contextual} computations in programming languages can learn from better 
+explored emedding of \emph{effectful} computations. In purely functional programming languages
+such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
+necessary (monadic) plumbing. This is made easier by the ``do'' notation that inserts the monadic
+operations automatically.
+
+The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
+context-aware computations based on comonads. The notation is close to the semantics of our 
+flat coeffect calculus (Chapter~\ref{ch:flat}). Extending it to support calculi based on the
+structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
+this requires explicitly manipulating individual context variables and application of structural
+rules (which is not needed in flat coeffects).
+
+In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
+but they can still use a special notation for explicitly marking effectful (coeffectful) 
+blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
+that differ from the ``do'' notation in two ways. First, they support wider range of 
+constructs, making it possible to wrap existing F\# code in a block without other changes.
+Second, they support other abstractions including monads, applicative functors and monad
+transformers. It would be interesting to see if computation expressions can be extended to
+handle computations based on flat/structural indexed comonads.
+
+Finally, more lightweight syntax for effectful computation can be obtained using
+techniques that automatically insert the necessary monadic plumbing (wihtout using any
+special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
+and similar approach would be worth exploring for coeffects.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect annotations as types}
+
+The other aspect of practical implementation of coeffects is tracking the context requirements
+(coeffect annotations) in the type system. To achieve this, the type system needs to be able to 
+capture the different examples of coeffect algebras. The structures used in this thesis include sets 
+(with union or interesection), natural numbers (with addition, maximum, minimum and multiplication),
+two-point lattice (for liveness) and free monoids (vectors of annotations).
+
+A recent work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
+recent additions to the Haskell type system are powerful enough to implement structures such
+as sets at the type level.
+
+\cite{types-dependent-ml}
+\cite{types-units-of-measure}
+\cite{app-inforich}
+
+% =================================================================================================
+%                                                                                             
+%    #   #          #                    ##                                                    
+%    #   #          #                     #                                                    
+%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
+%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
+%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
+%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
+%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
+%                                                            #   #                #   #        
+%                                                             ###                  ###         
+% =================================================================================================
+
+\section{Meta-language}
+\label{sec:conclusions-meta}
+
+Related and further work has been discussed throughout, but one important thing remains.
 
@@
 
-%===================================================================================================
-
 \newcommand{\munit}{\mathsf{e}}
@@
 
-\section{Introduction}
-
-\section{Meta-language}
-\label{sec:conclusions-meta}
-
-\begin{figure}
+%\begin{figure}
 \begin{equation*}
@@
 
-\caption{Type system for the coeffect meta-language cml}
-\label{fig:cml-types}
-\end{figure}
+%\caption{Type system for the coeffect meta-language cml}
+%\label{fig:cml-types}
+%\end{figure}
 
@@
 \section{Summary}
+\label{sec:conclusions-summary}

### 2014-09-23 text/Chapter07.tex
Finish further work and conclusions (Chapter 7)

@@
 
-The main goal of this thesis is to demonstrate that there is a unified notion of \emph{context}
-that captures many common scenarios in computer programming. The examples we explore include
-many notions of context important in modern software development (including distributed and
-cross-platform development) as well as context studied in fundamental programming language 
-research (including data-flow, linear logics and variable liveness).
+The main goal of this thesis is to demonstrate that different programming language features
+and libraries involving \emph{context} can be seen as instances of the same unified notion.
+The examples we explore include many properties of context important in modern software 
+development (including distributed and cross-platform development) as well as context studied 
+in fundamental programming language research (including data-flow, linear logics and variable 
+liveness).
 
@@
 particular, they can be modelled by our flat or structural coeffect calculi. Although the 
-results presented in this thesis are mainly of a theoretical nature, we, indeed, believe that 
+results presented in this thesis are mainly of a theoretical nature, we indeed believe that 
 coeffects should be integrated in main-stream programming languages.
@@
 
-Many of the examples of contextual computation that we discussed earlier (\eg~implicit
-parameters \cite{app-implicit-parameters} or distributed computations 
-\cite{app-distributed-ml5,app-distributed-links}) have been implemented as a single-purpose
-programming language feature. However, the main contribution of this thesis is that it captures
+Many of the examples of contextual computation that we discussed earlier have been implemented as 
+a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters} 
+or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However, 
+the main contribution of this thesis is that it captures
 \emph{multiple} different notions of context-aware computations using just a \emph{single}
@@
 such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
-necessary (monadic) plumbing. This is made easier by the ``do'' notation that inserts the monadic
-operations automatically.
+necessary (monadic) plumbing. This is made easier by the ``do'' notation \cite{other-haskell98} 
+that inserts the monadic operations automatically.
 
@@
 context-aware computations based on comonads. The notation is close to the semantics of our 
-flat coeffect calculus (Chapter~\ref{ch:flat}). Extending it to support calculi based on the
-structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
+flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based 
+on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
 this requires explicitly manipulating individual context variables and application of structural
-rules (which is not needed in flat coeffects).
+rules, which is not needed in flat coeffects.
 
@@
 
-Finally, more lightweight syntax for effectful computation can be obtained using
-techniques that automatically insert the necessary monadic plumbing (wihtout using any
+More lightweight syntax for effectful computation can be obtained using
+techniques that automatically insert the necessary monadic plumbing (wihtout a
 special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
@@
 The other aspect of practical implementation of coeffects is tracking the context requirements
-(coeffect annotations) in the type system. To achieve this, the type system needs to be able to 
-capture the different examples of coeffect algebras. The structures used in this thesis include sets 
-(with union or interesection), natural numbers (with addition, maximum, minimum and multiplication),
-two-point lattice (for liveness) and free monoids (vectors of annotations).
+(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
+language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
+The structures used in this thesis include sets (with union or interesection), natural numbers (with 
+addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
+(vectors of annotations).
 
-A recent work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
-recent additions to the Haskell type system are powerful enough to implement structures such
-as sets at the type level.
+The work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
+recent additions to the Haskell type system provide enough poer to implement structures such
+as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful 
+future direction for applied coeffects.
 
-\cite{types-dependent-ml}
-\cite{types-units-of-measure}
-\cite{app-inforich}
+In dependently-typed programming languages such as Agda \cite{other-agda} or Idris \cite{other-idris},
+the embedding of coeffects can be implemented more directly. However, we believe that coeffect
+tracking does not require full dependent types and can be made accessible in more main-stream
+languages. Dependent ML \cite{types-dependent-ml} provides an interesting example of a language
+with some dependent typing which is still close to its non-dependently-typed predecessor ML.
+
+Another approach for embedding computations into the type system has been pioneered by F\#
+\emph{type providers} \cite{app-inforich}. Technically, type providers are compiler extensions 
+that generate types based on external data sources or other information in order to provide easy 
+access to data or services. A similar approach could be used for embedding \emph{algberas} such
+as coeffect algebras into the type system. A \emph{algebra provider} would be a library that 
+specifies the objects of the algebra, equational laws and generalization rules for type inference.
+This could provide an easy to use way of embedding coeffect tracking in pragmatic languages
+such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
+\cite{types-units-of-measure}, which could be provided via one such \emph{algebra provider}.
 
@@
 
-\section{Meta-language}
+\section{Coeffect meta-language}
 \label{sec:conclusions-meta}
 
-Related and further work has been discussed throughout, but one important thing remains.
+In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
+language semantics introduced by Moggi \cite{monad-notions}. The first approach is to use
+monads in the \emph{semantics} of an effectful language. The second approach is to extend the
+language with (additional) monadic constructs that can then be used for writing effectful
+monads explicitly.
 
-Both flat coeffect calculus and structural coeffect calculus (presented in the past two chapters)
-use indexed comonads to define the semantics of the langauge. In this section, we follow the 
-meta-language style and embed indexed comonads into the language -- the type constructor
-$\ctyp{r}{\alpha}$ becomes a first-class value and we add language constructs corresponding to
-primitive operations of the indexed comonad.
+In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
+the term language is that of simply-typed $\lambda$-calculus, and we used (flat or structural)
+indexed comonads to give the semantics for the language and to derive type system for it.
 
-\newcommand{\munit}{\mathsf{e}}
-\newcommand{\mseq}[0]{ \oplus }
+In this section, we briefly discuss the alternative approach. That is, we embed indexed
+comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
+context (semantically, this corresponds to an indexed comonad) and we add language constructs
+that correspond to the operations of indexed comonads.
 
-%\begin{figure}
+This section provides a brief sketch of coeffect meta-language to highlight the relationship
+between coeffects and important related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
+Developing the system further is an interesting future research direction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffects and contextual modal type theory}
+
+As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
+to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
+meta-languages that include $\square$ as a type constructor
+\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
+Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
+From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
+coeffect calculus.
+
+\paragraph{Context in CMTT and coeffects.}
+Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
+comonads (the $\square$ modality) into the meta-language, there are two important differences.
+
+Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
+requirements are formed by a coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, \emph{etc.}
+
+Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
+the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+constructor refers to a value of type $\tau$ \emph{together} with additional context specified
+by $\cclrd{r}$ (\eg~list of past values or additional resources). In contrast in 
+CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
+original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
+context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
+two operations of a comonad:
+\begin{equation*}
+\begin{array}{rcl}
+ \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
+ \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
+    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
+\end{array}
+\end{equation*}
+Both readings are possible, but they give quite different meanings to the operations:
+
+\begin{itemize}
+\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
+  and does not require any additional context; the \ident{cobind} operation requires
+  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
+  evaluate the function, ending with a value $\beta$ together with remaining context
+  $\cclrd{s}$.
+\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
+  that requires no additional context to obtain a $\alpha$ value; given a function that
+  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
+  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
+  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
+  context requirements is eliminated by the function).
+\end{itemize}
+
+\noindent
+Although the different reading does not affect formal properties of the systems, it is important
+to understand the difference when discussing the two systems.
+
+The sketch of a coeffect meta-language in the following section attempts to bridge the gap
+between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
+it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
+sets of variables. Future work on the coeffect meta-language would thus be an interesting 
+development for both coeffect systems and CMTT.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
+For simplicity, this section only discusses the flat variant. The syntax of types and terms
+of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four expression forms:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
+ e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
+  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
+  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
+  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
+\end{array}
+\end{equation*}
+%
+The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
+operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
+\kvd{split} and \kvd{merge} that embed the corresponding operations.
+
+\paragraph{Types for coeffect meta-language.}
+The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
+shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
+rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
+
+The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
+a value from a comonadic context and corresponds to variable access in coeffect calculi. The
+\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
+and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
+The result is a computation that produces a $\tau_2$ value with the remaining context specified
+by $\cclrd{s}$. Note that the expression $e_2$ and $e_1$ corresponds to the first and second
+arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
+CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
+although it differs because of our generalization of comonads where \ident{bind} composes
+coeffect annotations rather than requiring the same annotation everywhere.}.
+
+The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
+transformation on one or two values in a context and then add the new value as a fresh variable
+to the variable context. For simplicity, we omit sub-coeffecting, but it could be easily added 
+following the pattern used elsewhere.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Typing rules for the simply typed $\lambda$-calculus}
+
 \begin{equation*}
 \tyrule{var}
-  {x : \alpha \in \Gamma}
-  {\Gamma \vdash x : \alpha }
+  {x : \tau \in \Gamma}
+  {\Gamma \vdash x : \tau }
 \end{equation*}
@@
 \tyrule{app}
-  {\Gamma \vdash e_1 : \alpha \rightarrow \beta &
-   \Gamma \vdash e_2 : \alpha }
-  {\Gamma \vdash e_1~e_2 : \beta}
+  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
+   \Gamma \vdash e_2 : \tau_2 }
+  {\Gamma \vdash e_1~e_2 : \tau_1 }
 \end{equation*}
@@
 \tyrule{abs}
-  {\Gamma, x:\alpha \vdash e : \beta}
-  {\Gamma \vdash \lambda x.e : \alpha \rightarrow \beta }
+  {\Gamma, x:\tau_1 \vdash e : \tau_2}
+  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
+
+\begin{equation*}
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
 \end{equation*}
 \begin{equation*}
-\tyrule{letbox}
-  {\Gamma \vdash e_1 : \ctyp{r \mseq s}{\alpha} & \Gamma, x : \ctyp{r}{\alpha} \vdash e_2 : \beta }
-  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{s}{\beta}}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash !e : \tau}
 \end{equation*}
 \begin{equation*}
-\tyrule{eval}
-  {\Gamma \vdash e : \ctyp{\munit}{\alpha}}
-  {\Gamma \vdash !e : \alpha}
+\tyrule{split}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
+   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
 \end{equation*}
 \begin{equation*}
-\tyrule{sub}
-  {\Gamma \vdash e : \ctyp{s}{\alpha} }
-  {\Gamma \vdash e : \ctyp{r}{\alpha} }
-\;(s\leq r)  
+\tyrule{merge}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
+   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
+   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
+  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
 \end{equation*}
 
-%\caption{Type system for the coeffect meta-language cml}
-%\label{fig:cml-types}
-%\end{figure}
+\caption{Type system for the (flat) coeffect meta-language}
+\label{fig:conclusions-cml}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding flat coeffect calculus}
+
+The \emph{meta-language} approach of embedding comonads in a language is more general than the
+\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
+a type system for context-aware programming languages.
+
+However, it is worth noting that 
+(flat) coeffect calculus can be embedded in the meta-language described above. This may be 
+desirable, \eg~when using the meta-language for reasoning about context-aware computations. 
+We briefly consider the embedding as it illuminates the relationship between coeffects and
+CMTT (although it is not possible to embed coeffect calculi in CMTT because of the more general
+annotations structure).
+
+Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in a flat coeffect calculus,
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
+meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
+to represent the entire variable context of the source language. The translation is defined in
+Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
+Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+the operations of a monad.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{l}
+ \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
+   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
+   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
+      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
+   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
+\end{array}
+\end{equation*}
+\caption{Embedding flat coeffect calculus in coeffect meta-language }
+\label{fig:conclusions-embed}
+\end{figure}
 
 % ==================================================================================================
-
-\section{Related work}
-This chapter is closely related to Contextual Modal Type Theory (CMTT) of Nanevski et al. However
-they develop their language using model logic as a basis, while we use categorical foundations 
-as the basis - leading to a different system.
-
-\section{Also related work}
-
-This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
-and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
-\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
-may split contextual requirements between the declaration-site and call-site.
-
-Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
-\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
-and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
-\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
-general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).
-
-% --------------------------------------------------------------------------------------------------
-
-\vspace{-1em}
-\paragraph{Comonads and modal logics.}
-
-Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
-$\square$ modality of an intuitionistic S4 modal logic using monoidal
-comonads, which links our calculus to modal logics.  This link can be
-materialized in two ways.
-
-Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
-correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
-building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
-\cite{monad-notions}) that includes $\square$ as a type
-constructor. For example, in \cite{logic-modal-reconstruction}, the
-modal type $\Box \tau$ represents closed terms.
-In contrast, the \emph{semantic} approach uses monads or comonads
-\emph{only} as a semantics.  This has been employed by Uustalu and
-Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
-follow the semantic approach.
-
-Nanevski et al. extend an S4 term language to a \emph{contextual}
-modal type theory (CMTT)~\cite{logic-cmtt}.
-The \emph{context} is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. Our contextual types are
-indexed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}.
-
-The work on CMTT suggests two extensions to coeffects. The first is
-developing the logical foundations. We briefly considered special cases
-of our system that permits local soundness in \S\ref{sec:syntactic} and
-local completeness can be treated similarly. The second problem is 
-developing the coeffects \emph{metalanguage}. The use of coeffect algebras
-would provide an additional flexibility over CMTT, allowing a wider range 
-of applications.
+%
+%       ###                                            
+%      #   #                                           
+%      #      #   #  ## #   ## #    ###   # ##   #   # 
+%       ###   #   #  # # #  # # #      #  ##  #  #   # 
+%          #  #   #  # # #  # # #   ####  #      #  ## 
+%      #   #  #  ##  # # #  # # #  #   #  #       ## # 
+%       ###    ## #  #   #  #   #   ####  #          # 
+%                                                #   # 
+%                                                 ###  
+% =================================================================================================
 
@@
 \label{sec:conclusions-summary}
+
+Some of the most fundamental academic work is not the one solving hard research problems, but the 
+one that changes how we understand the world. Some philosophers argue that \emph{language} is the 
+key for understanding how we think, while in science the dominant thinking is determined by 
+\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
+In a way, programming languages play a similar role for computer science and software development.
+
+This thesis aims to change how developers and programming language designers think about 
+\emph{context} or \emph{execution environments} of programs. Such context or execution environment
+has numerous forms -- mobile applications access network, GPS location or user's personal data; 
+journalists obtain information published on the web or through open government data initiatives. 
+This thesis aims to change our understanding of \emph{context} so that the above examples are 
+viewed uniformly through a single programming language abstraction, which we call \emph{coeffects}, 
+rather than as disjoint cases. This is done through the following three contributions.
+
+\paragraph{Unified presentation.}
+In $\lambda$-calculus, the term \emph{context} is usually used for the free-variable context.
+However, there are many other programming language features that are related to context or program's
+execution environment. In Chapter~\ref{ch:applications}, we revisit many of such features --
+including resource tracking in distributed computing, cross-platform development, data-flow 
+programming and liveness analysis, but also Haskell's type classes and implicit parameters.
+
+The main contribution of the chapter is that it presents the disjoint language features in a 
+unified way. We show type systems and semantics for many of the languages, illuminating
+the fact that they are closely related.
+
+Considering applications is one way of approaching the theory of coeffects introduced in this thesis.
+Other pathways to coeffects are discussed in Chapter~\ref{ch:pathways}, which looks at theoretical 
+developments leading to coeffects, including the work on effect systems, comonadic semantics and 
+linear logics.
+
+\paragraph{Flat coeffect calculus.}
+The applications discussed in Chapter~\ref{ch:applications} fall into two categories. In the first
+category (Section~\ref{sec:applications-flat}), the additional contextual information are 
+\emph{whole-context} properties. They either capture properties of the execution environment or
+affect the whole free-variable context.
+
+In Chapter~\ref{ch:flat}, we develop a \emph{flat coeffect calculus} which gives us an unified way
+of tracking \emph{whole-context} properties. The calculus is parameterized by a \emph{flat coeffect
+algebra} that captures the algebraic properties that contextual information satisfy. Concrete 
+instances of flat coeffects include Haskell's implicit parameters, whole-context liveness and
+whole-context data-flow.
+
+Our focus is on the syntactic properties of the calculus -- we give a type system 
+(Section~\ref{sec:flat-calculus}) and discuss equational theory of the calculus (Section~\ref{sec:flat-syntax}).
+In the flat coeffect calculus, $\beta$-reduction and $\eta$-expansion do not generally preserve
+type of expressions, but we identify two conditions when this is the case -- this gives us
+a basis for operational semantics of our calculi for liveness and implicit parameters. 
+The design of the type system is further validated by categorical semantics
+(Section~\ref{sec:flat-semantics}) which models flat coeffect calculus in terms of 
+\emph{flat indexed comonad} -- a structure based on categorical dual of monads.
+
+\paragraph{Structural coeffect calculus.}
+In the second category of context-aware systems discussed in Section~\ref{sec:applications-structural},
+contextual properties are \emph{per-variable}. The systems discussed here resemble sub-structural
+logics, but rather than \emph{restricting} variable use, they \emph{track} additional information
+about how variables are used.
+
+We unify systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
+which describes our \emph{structural coeffect calculus} (Section~\ref{sec:struct-calculus}). 
+Similarly to the flat variant, the calculus is parameterized by a \emph{structural coeffect algebra}. 
+Concrete instances of the calculus track bounded variable use (\ie~how many times is variable 
+accessed), data-flow properties (how many past values are needed) and liveness (\ie~can variable 
+be accessed).
+
+The structural coeffect calculus has desirable equational properties (Section~\ref{sec:struct-syntax}).
+In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all structural
+coeffect calculi. This follows from the fact that structural coeffect associates contextual requirements
+with individual variables and preserves the connection by including explicit structural rules
+(weakening, exchange and contraction).
+
+The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
+thesis. In Chapter~\ref{ch:unified}, we develop the two calculi further by introducing a unified
+presentation of the two (by parameterizing the calculus by the \emph{shape} of context) and we present
+an alternative presentation that makes type inference easier.

### 2014-09-26 text/Chapter04.tex
Cleanup files, fix some invalid references
Add proof for bottom pointed substitution (Appendix A)

@@
 Furthermore, the $\cunit$ coeffect is often the top (greatest) or the bottom (smallest) 
-element of the semi-lattice, but not in general. When this is the case, we are able
-to prove certain properties of the calculus (Section~\ref{sec:syntax}).
+element of the semi-lattice, but not in general. As discussed in Section~\ref{sec:flat-syntax},
+when this is the case, we are able to prove certain properties of the calculus.
 
@@
 to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ to
-a term $\coctx{\Gamma}{\cunit} e' : \tau$ with no context requirements.
+a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
 
@@
 In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
-where $\czip = \cseq = \cpar$ and the operation is also idempotent and commutative:
+where $\czip = \cseq = \cpar$ and the operation is also idempotent and commutative and
+$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
 %
@@
 By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
-the free-variable context. See Appendix~?.
+the free-variable context. See Appendix~\ref{sec:appendix-flat-cbn}.
 \end{proof}

### 2014-11-17 text/Chapter05.tex
Cleanup files, fix some invalid references
Add proof for bottom pointed substitution (Appendix A)
Unified calculus, move future work (Chapter 6)

@@
 \section{Introduction}
-\label{sec:flat-intro}
+\label{sec:structural-intro}
 
@@
 structural coeffect calculus, the order of variables matters. Thus we treat free variable context
-as a vector with a uniqueness condition. We also write $\xlift{-}$ for the length
+as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
 of the vector:
@@
  \Gamma &\narrow{=}& \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
- \xlift{\Gamma} &\narrow{=}& n
+ \xlen{\Gamma} &\narrow{=}& n
 \end{array}
@@
 and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
-is written as $\slift{\aclrd{\textbf{r}}}$. The structure for working with vectors
+is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
 of coeffects is provided by the \emph{structural coeffect algebra} definition below.
@@
 \begin{definition}
+\label{def:structural-scalar}
 A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
@@
   and an operation that returns the vector length such that 
-  $\slift{\aclrd{\textbf{r}}} = \sclrd{n}$ for $\aclrd{\textbf{r}} : \C^{\sclrd{n}}$
+  $\alen{\aclrd{\textbf{r}}} = \sclrd{n}$ for $\aclrd{\textbf{r}} : \C^{\sclrd{n}}$
 
@@
   {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{ \czero }} \vdash e : \tau} 
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau} 
 \end{equation*}
@@
 \begin{array}{l}
- \xlift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
- \xlift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+ \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
 \end{array}
@@
 \\[0.5em]
-\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
+\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
   \hspace{2em}
@@
 By induction over $\vdash$, using the multi-nary aspect of the substitution 
-in the proof of the contraction case. See Appendix~?.
+in the proof of the contraction case.
 \end{proof}

### 2014-11-17 text/Chapter06.tex
Unified calculus, move future work (Chapter 6)

@@
 
-%===================================================================================================
+The main goal of this thesis is to provide a \emph{unified} calculus for tracking context 
+dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
+two kinds of contextual properties that we further covered separately -- flat coeffects in 
+Chapter~\ref{ch:flat} track whole-context properties and structural coeffects, covered in 
+Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify
+the two notions. We introduce a \emph{unified coeffect} calculus that generalizes the two systems
+and can be instantiated to track both flat and structural properties (Section~\ref{sec:unified-unified}).
 
-\section{Semi-lattice formulation}
-\label{sec:unified-semilattice}
-
-\subsection{Semi-lattice formulation}
-\subsection{Type inference algorithm}
+Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
+believe that coeffects should be integrated in main-stream programming languages. In the second
+part of this chapter, we outline one possible approach for practical implementations of coeffects 
+(Section~\ref{sec:unified-impl}). Finally, we discuss an alternative approach to defining coeffect 
+systems which highlights the relationship between our work and related work arising from modal logics 
+(Section~\ref{sec:unified-meta}).
 
 % ==================================================================================================
+
+\section{The unified coeffect calculus}
+\label{sec:unified-unified}
+
+The flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural coeffect calculus 
+differ in a number of ways (Figure~\ref{fig:struct-types}). Understanding the differences is the 
+key to reconciling the two systems:
+
+\begin{itemize}
+\item Structural coeffect calculus contains explicit rules for context manipulation
+  (weakening, contraction, exchange). In flat coeffect calculus, these rules are not defined
+  explicitly, but are admissible.
+
+\item In structural coeffect calculus, the variable context is treated as a vector
+  and is annotated with a vector of (scalar) coeffects. In flat coeffect calculus,
+  the variable context is a set and is annotated with a single (scalar) coeffect.
+
+\item In flat coeffect calculus, we distinguish between splitting of the context requirements
+  and merging of context requirements ($\cpar$ and $\czip$, respectively). In structural
+  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
+  and so the structural coeffect algebra requires just $\atimes$.
+\end{itemize}
+
+\noindent
+In the unified calculus presented in this section, we address the three differences as follows.
+We use calculus with explicit rules for context manipulation. In systems that arise from flat
+calculi, the rules can be applied freely without changing the coeffects. We generalize the 
+structure of coeffect annotations using the notion of a ``container'' which can be specialized 
+to obtain a single annotation or a vector of annotations. Finally, we distinguish between 
+splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
+respectively). For structural coeffect calculi, the two operators coincide, but for flat
+coeffect calculi, they provide the needed flexibility.
+
+\subsection{Shapes and containers}
+
+Our notion of a \emph{coeffect container} is based on the idea of a container introduced by 
+Abbott et al. \cite{types-containers}. Interestingly, the work on containers has later been linked
+to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
+types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of lists).
+For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
+specified length). More formally:
+
+\begin{definition}
+A \emph{container} $S \triangleleft P$ is given by a set $S$ of shapes and a shape-indexed family
+$P : S \rightarrow \ident{Set}$ of positions.
+\end{definition}
+
+\noindent
+Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams and 
+singleton data type (which contains exactly one element). Two containers relevant to our work
+are lists and singleton data types:
+
+\begin{itemize}
+\item The container representing lists is given by $S \triangleleft P$ where 
+  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
+  given length $n$ is a set of indices $P n = \{ 1 \ldots n \}$.
+
+\item The contianer representing singleton data type is given by $S \triangleleft P$ where
+  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
+  shape $\ast$ contains exactly one position $P \ast = \{ 0 \}$.
+\end{itemize}
+
+\noindent
+In the unified coeffect calculus, the structure of coeffect annotations is defined by a 
+container with additional operations (discussed later) that links it with the free-variable 
+context $\Gamma$.
+
+\subsection{Structure of coeffects}
+In structural coeffect calculus, the structure annotation was formed by a vector of coeffect
+sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
+\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
+a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
+(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
+\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
+define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
+are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
+algebra}} which consists of shape-indexed coeffect scalar values.
+
+As in the structural coeffect calculus, the contexts in the unified calculus are annotated with 
+shape-indexed coeffects, written as $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$; 
+functions take just a single input parameter and so are annotated with scalar coeffect values 
+$\sigma \xrightarrow{\cclrd{r}} \tau$.
+
+\paragraph{Coeffect scalar.}
+The following definition of the coeffect scalar structure repeats the Definition~\ref{def:structural-scalar}
+from the previous chapter.
+
+\begin{definition}
+A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
+$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
+$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
+$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
+%
+\begin{equation*}
+\begin{array}{ccr}
+r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
+\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
+\textnormal{(monoid)}   
+\\
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
+\czero \;\cpar\; r = r = r \;\cpar\; \czero &
+\textnormal{(monoid)}   
+\\
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
+t\; \cleq\; t &
+\textnormal{(pre-order)}   
+\end{array}
+\end{equation*}
+\end{definition}
+
+\noindent
+As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
+an underlying category structre; $\cunit$ and $\czero$ represent an accessed and unused variable, 
+respectively.
+
+The $\cpar$ operation models combining of context requirements arising from multiple parts of a 
+program. The meaning depends on the coeffect container. The operation can either combine requirements
+of individual variables (structural) or requirements attached to the whole context of multiple
+sub-exp\-ressions (flat).
+
+\paragraph{Coeffect containers.}
+The coeffect container is a container that determines how are scalar coeffect annotations attached 
+to free-variable contexts. In addition to a container $\SHP$ with shapes and shape-indexed positions, the
+coeffect container also provides a mapping that returns the shape of a free-variable context. The
+mapping between the shape of the variable context and the shape of the coeffect annotation
+is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
+shape $\SH = \{ \ast \}$.
+
+In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
+$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
+We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation $\SP\sclrd{s}$
+returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in \SP\sclrd{s} \rightarrow \C$ 
+as a mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
+the exponent $\aclrd{\textbf{r}} \in \C^{\SP\sclrd{s}}$.
+
+The coeffect container is also equipped with an operation that appends shapes (when we concatenate
+variable contexts) and two special shapes in $\SH$ representing empty context and singleton context.
+
+\begin{definition}
+A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$ 
+comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a 
+mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
+\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.
+
+The elements $\sempty$ and $\sunit$ represent the shapes of empty and singleton free-variable contexts 
+respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts. Given
+$\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
+we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
+\end{definition}
+
+\noindent
+As said earlier, we use two kinds of coeffect containers that describe the structure of vectors 
+(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):
+
+\begin{example}
+Structural coeffect shape is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
+where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
+number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
+respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
+{|}\Gamma_1{|} + {|}\Gamma_2{|}$.
+
+Therefore, a coeffect annotation is a \emph{vector} 
+$\aclrd{\textbf{r}} \in \C^{\SP\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
+for each position (corresponding to a variable $x_i$ in the context).
+\end{example}
+
+\begin{example}
+Flat coeffect shape is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
+The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
+with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.
+
+That is, there is a single shape
+$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
+Therefore, a coeffect annotation is drawn from $\C^\{\sflat\}$ which is isomorphic to $\C$
+and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
+\end{example}
+
+\newpage
+\begin{example}
+Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
+\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP \sflat = \emptyset$,
+${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.
+
+This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
+$\SP\sflat=\emptyset$ and so coeffect annotations would be from the set $\C^\emptyset$.
+\end{example}
+
+\paragraph{Unified coeffect algebra.}
+The coeffect calculus annotates judgments with shape-indexed coeffect annotations.
+The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and coeffect container
+to define shape-indexed coeffects and operations for manipulating these.
+
+The definition here reconciles the third point discussed in Section~\ref{sec:unified-unified} --
+the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
+while structural coeffects use tensor $\atimes$. In the unified calculus, we use two operators
+that can, however, be coincide.
+
+\begin{definition}
+Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
+\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
+coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero$ is a 
+coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
+combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
+%
+\begin{equation*}
+\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
+\end{equation*}
+\end{definition}
+
+\noindent
+A coeffect algebra induces the following three additional operations:
+%
+\begin{equation*}
+\begin{array}{rclcrcl}
+ \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
+ \alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}   \\
+\\[-0.5em]
+ \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
+ \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
+\\[-0.5em]
+ \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
+ \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
+\end{array}
+\end{equation*}
+%
+The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
+the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
+vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
+as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
+$\sclrd{m}$ and write $\aseq$.
+
+Finally, we define $\alen{}$ as an operation that returns the shape of a given shape-indexed 
+coeffect. The only purpose is to simplify notation, as we often need to specify that shapes of 
+variable context and coeffect match, \eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.
+
+\paragraph{Splitting and merging coeffects.}
+The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts. 
+For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
+\C^{\SP\sclrd{m}}$ and $\aclrd{\textbf{s}} \in \C^{\SP\sclrd{n}}$. In the structural system, the 
+context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
+combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined 
+coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}} 
+\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.
+
+For structural coeffect systems such as bounded reuse, both $\apar$ and $\azip$ are
+just the tensor product $\times$ of vectors. For flat coeffect systems, the operations
+can be defined independently, letting $\azip=\czip$ and $\apar=\cpar$.
+
+The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:unified-semantics}), 
+where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of 
+a morphism that merges the capabilities provided by two contexts (in the syntactic 
+reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$ 
+is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single 
+context into two parts (in the syntactic reading, merges their context requirements). 
+Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and 
+$\apar$ in \emph{conclusions}.
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
-\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
-\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
-\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
-\newcommand{\sflat}{\star}
+\begin{figure}[t]
 
-\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
-\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
-\newcommand{\azero}{ \textcolor{aclr}{\bot} }
-\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
-\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }
+{\small a.) Syntax-driven typing rules:}
+\begin{equation*}
+\tyrule{var}
+  { }
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c \!:\! \tau \in \Delta}
+  {\coctx{()}{\azero} \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\apar\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\azip\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\azip\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\apar\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
 
-\newcommand{\fcons}{ \times } 
-\newcommand{\fempty}{\epsilon}
+\vspace{1em}
+{\small b.) Structural rules for context manipulation:}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+(\cclrd{s'} \cleq \cclrd{s})
+\end{equation*}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+\begin{array}{l}
+ \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{2.5em} 
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~
+\begin{array}{l}
+ \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
 
-\section{The unified coeffect calculus}
-
-\subsection{Structure of coeffects}
-
-We describe the algebraic structure of coeffects in three
-steps. First, we define a \emph{\cclrd{coeffect scalar}} structure 
-which defines the basic building blocks of coeffect information;
-then we define \emph{\sclrd{coeffect shapes}} which determines how
-coeffect scalar values are related to the free-variable context. 
-Finally, we define the \emph{\aclrd{coeffect
-    algebra}} which consists of shape-indexed coeffect scalar values.
-
-For example, in bounded reuse the coeffect scalar structure
-comprised natural numbers $\cclrd{\mathbb{N}}$ with $\cclrd{+}$ and
-$\cclrd{\ast}$ operators. The shape for bounded reuse is the length of the
-free-variable context and so the coeffect annotation is a vector of matching length.
-Finally, the coeffect algebra specifies how vectors are
-concatenated and split in abstraction and application.
-
-In the coeffect system of the calculus, contexts are annotated with shape-indexed coeffects (\eg{}, vectors)  as in
-$\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$. However, functions take just a single 
-input parameter and so are annotated with scalar coeffect values as in $\sigma \xrightarrow{\cclrd{r}} \tau$.
-
-\paragraph{Coeffect scalar.}
-Coeffect scalar structures are equipped with two operations. In bounded reuse, those 
-were $\cclrd{\ast}$ for sequencing (in function application) and $\cclrd{+}$ for context 
-sharing (in contraction). Additional structure is needed for variable access and sub-coeffecting.
-
-\begin{definition}
-\emph{\cclrd{Coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set $\C$
-together with elements $\cunit,\czero \in \C$, relation $\cleq$ and binary operations 
-$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids
-and $(\C, \cleq)$ is a pre-order. The following distributivity axioms are required:
-\begin{align*}
-(\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
-\cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
-\end{align*}
-\end{definition}
-
-\noindent
-The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
-category in the semantics. 
-It models sequential composition with variable access
-($\cunit$) as the identity. The other element ($\czero$) is used for
-variables that are not accessed. The operation $\cpar$ combines coeffects
-for contexts used in multiple places (contraction). 
-The notation is inspired by the bounded reuse example, which uses coeffect scalar structure
-$(\cclrd{\mathbb{N}}, \ast, +, 1,0, \cclrd{\leq})$, but be aware that $\cseq$ and $\cpar$ do
-not always mean $\ast$ and $+$.
-
-The context annotations can be viewed as \emph{containers} of scalar coeffects.
-For structural coeffects, the container is a vector, while for flat coeffects, it is
-a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{types-containers}
-which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.
-
-\paragraph{Coeffect shapes.}
-The coeffect system is parameterised by a set of shapes $\SH$. A coeffect annotation 
-is indexed by a shape $\sclrd{s} \in \SH$ calculated from the shape of the free-variable vector. The correspondence is not necessarily 
-bijective. For example, flat coeffect systems have just a single shape $\SH = \{ \ast \}$.
-
-In the coeffect judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$, the coeffect annotation
-$\aclrd{R}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
-We write $\aclrd{s} = \slift{\Gamma}$ for the shape corresponding to $\Gamma$. We define shapes by a 
-\emph{set} of positions and so we can write $\aclrd{R} \in \sclrd{s} \rightarrow \C$ as a
-mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
-the exponent $\aclrd{R} \in \C^{\sclrd{s}}$.
-
-The set of shapes is equipped with an operation that combines shapes (when we combine 
-variable contexts), an operation that computes shape from the free-variable contexts, 
-and two special shapes in $\SH$ representing empty context and singleton context.
-
-\begin{definition}
-  A \emph{\sclrd{coeffect shape}} structure $(\SH, \slift{\textnormal{--}}, \stimes,
-  \sempty, \sunit)$ comprises a set $\SH$ with a binary operation
-  $\stimes$ on $\SH$ for shape composition, a mapping from contexts to
-  shapes $\slift{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
-  \SH$ such that $(\SH, \stimes, \sempty)$ is a monoid. The 
-  elements $\sempty$ and $\sunit$ represent the shapes of empty and
-  singleton free-variable contexts respectively.
-\end{definition}
-
-\noindent
-As said earlier, we use two kinds of shape structures that describe the shape
-of vectors and the shape of trivial singleton container:
-
-\begin{itemize}
-\item
-Structural coeffect shape is defined as $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$.
-We treat numbers as sets $0 = \{\}, 1=\{ \emptyset \}, 2=\{ \emptyset, 1\}, 3=\{\emptyset, 1, 2 \} \ldots$ 
-(so that a number is a set of positions). The shape mapping ${|}\Gamma{|}$ returns the number
-of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, respectively,
-and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = {|}\Gamma_1{|} + {|}\Gamma_2{|}$.
-Therefore, a coeffect annotation is a \emph{vector} $\aclrd{R} \in
-\C^{\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{R}(i) \in \C$ for each
-variable $x_i$ in the context.
-
-\item
-Flat coeffect shape is defined as $\sclrd{(\{\sflat \}, \mathsf{const}\,\,\sflat, \diamond, \sflat, \sflat)}$ where
-$\sflat\,\diamond\,\sflat = \sflat$ and $\star=\{\emptyset\}$. That is, there is a single shape
-$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
-Therefore, a coeffect annotation is drawn from $\C^{\sflat}$ which is iso\-morphic to $\C$
-and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
-\end{itemize}
-
-\noindent
-Using a shape with \emph{no} positions reduces our system to the simply-typed $\lambda$-calculus with no context annotations.
-Trees could be used to build a system akin to bunched typing \cite{substruct-bunched}.
-
-\paragraph{Coeffect algebra.}
-The coeffect calculus annotates judgments with 
-shape-indexed (or, \emph{shaped}) coeffects.
-The \emph{\aclrd{coeffect
-    algebra}} structure combines a coeffect scalar and coeffect shape structures
-to define shaped coeffects and operations for combining these.
-In Section~\ref{sec:intro-coeffects}, shaped coeffects were combined
-by the tensor $\times$ in structural examples and
-$\cup$ in the implicit parameters example. To capture the examples so far and
-those described previously~\cite{petricek2013coeffects}, we distinguish two
-operators for combining shaped coeffects.
-
-\begin{definition}
-  Given a \cclrd{coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero,
-  \cleq)$ and a \sclrd{coeffect shape} $(\SH, \slift{\textnormal{--}}, \stimes,
-  \sempty, \sunit)$ a \emph{\aclrd{coeffect algebra}} extends the two
-  structures with $(\amerge, \asplit, \azero)$ where $\azero \in
-  \C^{\sempty}$ is a coeffect annotation for the empty context and
-  $\amerge, \asplit$ are families of operations that combine coeffect
-  annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m}
-  \in \SH$:
-\[
-\asplit_{\sclrd{m},\sclrd{n}}, \amerge_{\sclrd{m},\sclrd{n}} ~:~ \C^{\sclrd{m}} \times \C^{\sclrd{n}} \rightarrow \C^{\sclrd{m}\,\stimes\,\sclrd{n}}
-\]
-\end{definition}
-
-\noindent
-A coeffect algebra induces the following two additional operations:
-\[
-\begin{array}{rclrcl}
-\alift{\textnormal{--}} &\narrow{:}& \C \rightarrow \C^{\sunit}          \quad\quad
-   & \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{m}}  
-\\ 
-\alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}  
-   & \cclrd{r}\ \aseq\ \aclrd{S} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\ \cseq\ (\aclrd{S}(\sclrd{s}))
-\end{array}
-\]
-$\alift{\textnormal{--}}$ lifts a scalar coeffect 
-to a shaped coeffect indexed by the singleton context shape. 
-The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a vector by scalar.
-As we always use lower-case for scalars and upper-case for vectors, using the same
-symbol is not ambiguous. We also tend to omit the subscript
-$\sclrd{m}$ and write just $\aseq$.
-
-The operators $\asplit$ and $\amerge$ combine shaped
-coeffects associated with two contexts. For example,
-assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects 
-$\aclrd{R} \in \C^{\sclrd{m}}$ and $\aclrd{S} \in \C^{\sclrd{n}}$. In
-the structural system, the context shapes $\sclrd{m}, \sclrd{n}$
-denote the number of variables in the two contexts. The combined
-context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and
-the combined coeffects $\aclrd{R} \,\amerge\, \aclrd{S}, \aclrd{R} \,\asplit\, \aclrd{S} \in \C^{\sclrd{m} \stimes \sclrd{n}}$
-are indexed by that shape.
-
-For structural coeffect systems such as bounded reuse, both $\asplit$ and $\amerge$ are
-just the tensor product $\times$ of vectors. However, we need to distinguish them for flat coeffect 
-systems discussed later.
-
-The difference is explained by the semantics (Section~\ref{sec:semantics}), where
-$\aclrd{R}\,\amerge\,\aclrd{S}$ is an annotation of the codomain of 
-a morphism that merges the capabilities provided by two contexts (in the syntactic 
-reading, splits the context requirements); $\aclrd{R}\,\asplit\,\aclrd{S}$ is an annotation of
-the domain of a morphism that splits the capabilities of a single context into two parts (in the
-syntactic reading, merges their context requi\-rements). Syntactically, this means that we always use
-$\amerge$ in the rule \emph{assumptions} and $\asplit$ in \emph{conclusions}. For now, it suffices to use the 
-bounded reuse intuition and read the operations as tensor products.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-$\boxed{
-\coctx{\Gamma}{R} \vdash e : \tau
-}$
-\[
-\begin{array}{c}
-\tyrule{const}{\;}
-  { \coctx{ () }{ \azero } \vdash c : \iota }
-\quad\quad
-\tyrule{var}{\;}
-  { \coctx{ (x:\tau) }{ \aunit } \vdash x : \tau }
-\\[1.8em]
-\tyrule{abs}
-  { \coctx{\Gamma, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{s}}} \vdash e : \tau }
-  { \coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau }
-\\[1.8em]
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \tau \xrightarrow{\cclrd{t}} \sigma &
-    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \tau }
-  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash e_1~e_2 : \sigma } 
-\\[1.8em]
-\tyrule{let}
-  { \coctx{\Gamma_1}{\aclrd{S}} \vdash e_1 : \sigma & 
-    \coctx{\Gamma_2, x:\sigma}{\aclrd{R}\; \amerge\; \alift{\cclrd{t}}} \vdash e_2 : \tau  }
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R}\; \asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau }
-\\[1.8em]
-\tyrule{ctx}
-  { \coctx{\Gamma}{\aclrd{R}} \vdash e : \tau & \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma'}{\aclrd{R'}} }{\theta}}
-  { \coctx{\Gamma'}{\aclrd{R'}} \vdash \theta e : \tau }
-\end{array}
-\]
-
-$\boxed{
-\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}
-}$
-
-\[
-\tystmt{weak}
-  { \quad\quad 
-    \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x:\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}
-    \quad\quad\quad }
-\]
-\[ 
-\tystmt{exch}
-  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau, y:\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
-               { \quad \coctx{\Gamma_1, y:\sigma, x:\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }{\emptyset} }
-\]
-\[
-\tystmt{contr}
-  { \ctxtransnl{ \coctx{\Gamma_1, y:\tau, z:\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
-               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]}}
-\]
-\[
-\tystmt{sub}
-  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau,\Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \aclrd{T} } }
-               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s'}} \,\asplit\, \aclrd{T}} }{\emptyset} }
-\quad(\cclrd{s} \;\cleq\; \cclrd{s'} )                 
-\]
-\caption{The general coeffect calculus}
-\vspace{-1em}
-\label{fig:unified-coeffects}
+\caption{Type system for the unified coeffect calculus}
+\label{fig:unif-types}
 \end{figure}
@@
 
-\subsection{General coeffect type system}
+\subsection{Unified coeffect type system}
 
-In the previous section, we developed an algebraic structure capable of
-capturing different concrete context-dependent properties discussed in
-Section~\ref{sec:intro-coeffects}. Now, we use the structure to
-define the general coeffect calculus in
-Figure~\ref{fig:coeffects-general}.
+The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
+shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
+the key differences between the two.
 
-Coeffect annotations on free-variable contexts
-are shape-indexed coeffects $\aclrd{R}, \aclrd{S}, \aclrd{T} \in \C^\SH$ 
-and function types are annotated with
-coeffects scalars $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.  The
-rules manipulate coeffect annotations using the operations provided by
-coeffect algebra $(\amerge, \asplit, \azero)$ and the derived
-constructs $\alift{\textnormal{--}}$ and $\aseq$. Free-variable contexts $\Gamma$
-are treated as vectors modulo duplicate use of variables 
--- the associativity is built-in. The order of
-variables matters, but can be changed using the structural rules. To
-make the system easier to follow, structural rules are expressed
-using a separate judgment.
+The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
+and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
 
-\paragraph{Typing rules.} 
+As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
+use of variables -- the associativity is built-int. The order of variables matters, but can be changed 
+using the structural rules. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}} \in \C^\SH$ 
+are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
+annotated with coeffects scalars. % $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.
+
+\paragraph{Differences from structural system.}
+
+% => \azero for constants 
+% => \azip in assumptions (let, abs), \apar in conclusions (let, app)
+% => the same also happened in structural rules
+%      --> how does that make contraction etc. permissible?
+% => <s,r> -> <s> x <r>
+
 Constants (\emph{const}) and variable access (\emph{var}) annotate the context with special
@@
 
-Lambda abstraction \emph{splits} the context requirements using $\amerge$ into a coeffect 
+Lambda abstraction \emph{splits} the context requirements using $\azip$ into a coeffect 
 $\aclrd{R}$ and a coeffect $\alift{\cclrd{s}}$ of a shape $\sunit$ (semantically, it \emph{merges}
 capabilities provided by the declaration-site and call-site contexts). In structural systems such
-as bounded reuse, $\asplit$ is not symmetric and so this gives us a coeffect associated with the bound variable.
+as bounded reuse, $\apar$ is not symmetric and so this gives us a coeffect associated with the bound variable.
 
@@
 The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
-Syntactically, the context requirements are first split using $\amerge$ and then re-combined
-using $\asplit$.
+Syntactically, the context requirements are first split using $\azip$ and then re-combined
+using $\apar$.
 
@@
 Exchange and contraction decompose and reconstruct coeffect annotations 
-using $\amerge_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\asplit_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
-The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slift{\Gamma_1}$
-and $\sclrd{n} = \slift{\Gamma_2}$.
+using $\azip_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\apar_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
+The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slen{\Gamma_1}$
+and $\sclrd{n} = \slen{\Gamma_2}$.
 
@@
 \item[--] Coeffect shape $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$ formed by natural numbers
-\item[--] Coeffect algebra $\aclrd{(\fcons, \fcons, \fempty)}$ where $\fcons$ and $\fempty$ are
+\item[--] Coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
   shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
-  That is $\fcons : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n + m}}$ appends vectors
-  (lists) and $\fempty : \C^{\sclrd{0}}$ represents empty vectors (lists)
+  That is $\times : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n + m}}$ appends vectors
+  (lists) and $\epsilon : \C^{\sclrd{0}}$ represents empty vectors (lists)
 \end{itemize}
@@
 The definition is valid since the shape operations form a monoid
-$\sclrd{(\mathbb{N}, +, 0)}$ and $\slift{\textnormal{--}}$ (calculating the length of
+$\sclrd{(\mathbb{N}, +, 0)}$ and $\slen{\textnormal{--}}$ (calculating the length of
 a list) is a monoid homomorphism from the free monoid to the monoid of
@@
 
-In addition to the coeffect scalar structure, we also need to define $\asplit$ and $\amerge$.
-Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\asplit$ (merging of shaped 
-coeffect annotations). However, the $\amerge$ operation needs to be provided explicitly.
+In addition to the coeffect scalar structure, we also need to define $\apar$ and $\azip$.
+Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ (merging of shaped 
+coeffect annotations). However, the $\azip$ operation needs to be provided explicitly.
 
@@
   $\sflat \diamond \sflat = \sflat$
-\item[--] Flat coeffect algebra $(\cclrd{\wedge}, \cpar, \czero)$, \ie{}, the $\asplit = \cpar$ and $\azero = 
-\czero$ with the additional binary operation $\amerge = \wedge$. 
+\item[--] Flat coeffect algebra $(\cclrd{\wedge}, \cpar, \czero)$, \ie{}, the $\apar = \cpar$ and $\azero = 
+\czero$ with the additional binary operation $\azip = \wedge$. 
 \end{itemize}
@@
 both contexts with the same coeffect,
-but that would require treating $\amerge$ as a partial function.
+but that would require treating $\azip$ as a partial function.
 \end{remark}
+
+\subsection{Semanitcs}
+\label{sec:unified-semantics}
+
+Not really - but sketch.
+
+%===================================================================================================
+
+\section{Semi-lattice formulation}
+\label{sec:unified-semilattice}
+
+\subsection{Semi-lattice formulation}
+
+\begin{equation*}
+\begin{array}{lcll}
+ \ident{counit}_{\cunit} &\narrow{:}& 
+    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
+ \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
+    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
+\ident{merge}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
+    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)} \\
+\ident{split}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
+    \ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \\
+\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
+    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} & 
+    (\cclrd{s}\,\cleq\,\cclrd{r})
+\end{array}
+\end{equation*}
+
+Change to
+
+\begin{equation*}
+\begin{array}{lcll}
+ \ident{counit}_{\cunit} &\narrow{:}& 
+    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
+ \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
+    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
+\ident{merge}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
+    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{t}}{(\alpha \times \beta)} &
+    (\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s}) \\
+\ident{split}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
+    \ctyp{\cclrd{t}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} &
+    (\cclrd{r}\,\cleq\,\cclrd{t}, \cclrd{s}\,\cleq\,\cclrd{t}) \\
+\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
+    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} & 
+    (\cclrd{s}\,\cleq\,\cclrd{r})
+\end{array}
+\end{equation*}
+
+Gives us..
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma}{\cunit} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c : \tau \in \Delta}
+  {\coctx{\Gamma}{\czero} \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+
+\caption{Type system for the flat coeffect calculus}
+\label{fig:flat-types2}
+\end{figure}
+
+q
+\newpage
+a
+\newpage
+a
+\newpage
+
+\subsection{Type inference algorithm}
 
@@
 
+% =================================================================================================
+%                                                                   
+%      ####                         #       #                   ##   
+%      #   #                        #                            #   
+%      #   #  # ##    ###    ###   ####    ##     ###    ###     #   
+%      ####   ##  #      #  #   #   #       #    #   #      #    #   
+%      #      #       ####  #       #       #    #       ####    #   
+%      #      #      #   #  #   #   #  #    #    #   #  #   #    #   
+%      #      #       ####   ###     ##    ###    ###    ####   ###  
+%                                                                
+% =================================================================================================
+
+\section{Towards practical coeffects}
+\label{sec:unified-impl}
+
+As discussed earlier, the main focus of this thesis is the development of the much needed 
+\emph{theory of context-aware computations} and so discussing the details of a practical 
+implementation of coeffect tracking is beyond the scope of the thesis. However, this section
+briefly outlines one possible pathway towards this goal.
+
+Many of the examples of contextual computation that we discussed earlier have been implemented as 
+a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters} 
+or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However, 
+the main contribution of this thesis is that it captures
+\emph{multiple} different notions of context-aware computations using just a \emph{single}
+common abstraction. For this reason, we advocate that future practical implementations of
+coeffects should not be single-purpose language features, but instead reusable abstractions that
+can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
+
+In order to do this, programming languages need to provide two features; one that allows 
+embedding of context-aware computations themselves in programs (akin to the ``do'' notation
+in Haskell) and one that allows tracking of the contextual information in the type system.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding contextual computations}
+
+The embedding of \emph{contextual} computations in programming languages can learn from better 
+explored emedding of \emph{effectful} computations. In purely functional programming languages
+such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
+necessary (monadic) plumbing. This is made easier by the ``do'' notation \cite{other-haskell98} 
+that inserts the monadic operations automatically.
+
+The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
+context-aware computations based on comonads. The notation is close to the semantics of our 
+flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based 
+on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
+this requires explicitly manipulating individual context variables and application of structural
+rules, which is not needed in flat coeffects.
+
+In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
+but they can still use a special notation for explicitly marking effectful (coeffectful) 
+blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
+that differ from the ``do'' notation in two ways. First, they support wider range of 
+constructs, making it possible to wrap existing F\# code in a block without other changes.
+Second, they support other abstractions including monads, applicative functors and monad
+transformers. It would be interesting to see if computation expressions can be extended to
+handle computations based on flat/structural indexed comonads.
+
+More lightweight syntax for effectful computation can be obtained using
+techniques that automatically insert the necessary monadic plumbing (wihtout a
+special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
+and similar approach would be worth exploring for coeffects.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect annotations as types}
+
+The other aspect of practical implementation of coeffects is tracking the context requirements
+(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
+language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
+The structures used in this thesis include sets (with union or interesection), natural numbers (with 
+addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
+(vectors of annotations).
+
+The work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
+recent additions to the Haskell type system provide enough poer to implement structures such
+as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful 
+future direction for applied coeffects.
+
+In dependently-typed programming languages such as Agda \cite{other-agda} or Idris \cite{other-idris},
+the embedding of coeffects can be implemented more directly. However, we believe that coeffect
+tracking does not require full dependent types and can be made accessible in more main-stream
+languages. Dependent ML \cite{types-dependent-ml} provides an interesting example of a language
+with some dependent typing which is still close to its non-dependently-typed predecessor ML.
+
+Another approach for embedding computations into the type system has been pioneered by F\#
+\emph{type providers} \cite{app-inforich}. Technically, type providers are compiler extensions 
+that generate types based on external data sources or other information in order to provide easy 
+access to data or services. A similar approach could be used for embedding \emph{algberas} such
+as coeffect algebras into the type system. A \emph{algebra provider} would be a library that 
+specifies the objects of the algebra, equational laws and generalization rules for type inference.
+This could provide an easy to use way of embedding coeffect tracking in pragmatic languages
+such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
+\cite{types-units-of-measure}, which could be provided via one such \emph{algebra provider}.
+
+% =================================================================================================
+%                                                                                             
+%    #   #          #                    ##                                                    
+%    #   #          #                     #                                                    
+%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
+%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
+%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
+%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
+%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
+%                                                            #   #                #   #        
+%                                                             ###                  ###         
+% =================================================================================================
+
+\section{Coeffect meta-language}
+\label{sec:unified-meta}
+
+In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
+language semantics introduced by Moggi \cite{monad-notions}. The first approach is to use
+monads in the \emph{semantics} of an effectful language. The second approach is to extend the
+language with (additional) monadic constructs that can then be used for writing effectful
+monads explicitly.
+
+In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
+the term language is that of simply-typed $\lambda$-calculus, and we used (flat or structural)
+indexed comonads to give the semantics for the language and to derive type system for it.
+
+In this section, we briefly discuss the alternative approach. That is, we embed indexed
+comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
+context (semantically, this corresponds to an indexed comonad) and we add language constructs
+that correspond to the operations of indexed comonads.
+
+This section provides a brief sketch of coeffect meta-language to highlight the relationship
+between coeffects and important related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
+Developing the system further is an interesting future research direction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffects and contextual modal type theory}
+
+As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
+to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
+meta-languages that include $\square$ as a type constructor
+\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
+Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
+From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
+coeffect calculus.
+
+\paragraph{Context in CMTT and coeffects.}
+Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
+comonads (the $\square$ modality) into the meta-language, there are two important differences.
+
+Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
+requirements are formed by a coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, \emph{etc.}
+
+Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
+the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+constructor refers to a value of type $\tau$ \emph{together} with additional context specified
+by $\cclrd{r}$ (\eg~list of past values or additional resources). In contrast in 
+CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
+original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
+context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
+two operations of a comonad:
+\begin{equation*}
+\begin{array}{rcl}
+ \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
+ \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
+    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
+\end{array}
+\end{equation*}
+Both readings are possible, but they give quite different meanings to the operations:
+
+\begin{itemize}
+\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
+  and does not require any additional context; the \ident{cobind} operation requires
+  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
+  evaluate the function, ending with a value $\beta$ together with remaining context
+  $\cclrd{s}$.
+\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
+  that requires no additional context to obtain a $\alpha$ value; given a function that
+  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
+  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
+  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
+  context requirements is eliminated by the function).
+\end{itemize}
+
+\noindent
+Although the different reading does not affect formal properties of the systems, it is important
+to understand the difference when discussing the two systems.
+
+The sketch of a coeffect meta-language in the following section attempts to bridge the gap
+between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
+it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
+sets of variables. Future work on the coeffect meta-language would thus be an interesting 
+development for both coeffect systems and CMTT.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
+For simplicity, this section only discusses the flat variant. The syntax of types and terms
+of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four expression forms:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
+ e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
+  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
+  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
+  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
+\end{array}
+\end{equation*}
+%
+The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
+operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
+\kvd{split} and \kvd{merge} that embed the corresponding operations.
+
+\paragraph{Types for coeffect meta-language.}
+The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
+shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
+rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
+
+The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
+a value from a comonadic context and corresponds to variable access in coeffect calculi. The
+\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
+and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
+The result is a computation that produces a $\tau_2$ value with the remaining context specified
+by $\cclrd{s}$. Note that the expression $e_2$ and $e_1$ corresponds to the first and second
+arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
+CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
+although it differs because of our generalization of comonads where \ident{bind} composes
+coeffect annotations rather than requiring the same annotation everywhere.}.
+
+The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
+transformation on one or two values in a context and then add the new value as a fresh variable
+to the variable context. For simplicity, we omit sub-coeffecting, but it could be easily added 
+following the pattern used elsewhere.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Typing rules for the simply typed $\lambda$-calculus}
+
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\Gamma \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
+   \Gamma \vdash e_2 : \tau_2 }
+  {\Gamma \vdash e_1~e_2 : \tau_1 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x:\tau_1 \vdash e : \tau_2}
+  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
+
+\begin{equation*}
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
+\end{equation*}
+\begin{equation*}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash !e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{split}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
+   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{merge}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
+   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
+   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
+  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
+\end{equation*}
+
+\caption{Type system for the (flat) coeffect meta-language}
+\label{fig:conclusions-cml}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding flat coeffect calculus}
+
+The \emph{meta-language} approach of embedding comonads in a language is more general than the
+\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
+a type system for context-aware programming languages.
+
+However, it is worth noting that 
+(flat) coeffect calculus can be embedded in the meta-language described above. This may be 
+desirable, \eg~when using the meta-language for reasoning about context-aware computations. 
+We briefly consider the embedding as it illuminates the relationship between coeffects and
+CMTT (although it is not possible to embed coeffect calculi in CMTT because of the more general
+annotations structure).
+
+Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in a flat coeffect calculus,
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
+meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
+to represent the entire variable context of the source language. The translation is defined in
+Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
+Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+the operations of a monad.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{l}
+ \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
+   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
+   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
+      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
+   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
+\end{array}
+\end{equation*}
+\caption{Embedding flat coeffect calculus in coeffect meta-language }
+\label{fig:conclusions-embed}
+\end{figure}
+
 \section{Conclusions}

### 2014-11-17 text/Chapter07.tex
Unified calculus, move future work (Chapter 6)

--- 2014-09-19 text/Chapter07.tex
@@
 
-%---------------------------------------------------------------------------------------------------
+% ==================================================================================================
+%
+%       ###                                            
+%      #   #                                           
+%      #      #   #  ## #   ## #    ###   # ##   #   # 
+%       ###   #   #  # # #  # # #      #  ##  #  #   # 
+%          #  #   #  # # #  # # #   ####  #      #  ## 
+%      #   #  #  ##  # # #  # # #  #   #  #       ## # 
+%       ###    ## #  #   #  #   #   ####  #          # 
+%                                                #   # 
+%                                                 ###  
+% =================================================================================================
 
-How to make this practical?
+Some of the most fundamental academic work is not the one solving hard research problems, but the 
+one that changes how we understand the world. Some philosophers argue that \emph{language} is the 
+key for understanding how we think, while in science the dominant thinking is determined by 
+\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
+In a way, programming languages play a similar role for computer science and software development.
 
-codo notation / computation expressions
-
-embedding algebraic things
-
-%---------------------------------------------------------------------------------------------------
-
-Both flat coeffect calculus and structural coeffect calculus (presented in the past two chapters)
-use indexed comonads to define the semantics of the langauge. In this section, we follow the 
-meta-language style and embed indexed comonads into the language -- the type constructor
-$\ctyp{r}{\alpha}$ becomes a first-class value and we add language constructs corresponding to
-primitive operations of the indexed comonad.
-
-%===================================================================================================
-
-\newcommand{\munit}{\mathsf{e}}
-\newcommand{\mseq}[0]{ \oplus }
-
-\section{Introduction}
-
-\section{Meta-language}
-\label{sec:conclusions-meta}
-
-\begin{figure}
-\begin{equation*}
-\tyrule{var}
-  {x : \alpha \in \Gamma}
-  {\Gamma \vdash x : \alpha }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\Gamma \vdash e_1 : \alpha \rightarrow \beta &
-   \Gamma \vdash e_2 : \alpha }
-  {\Gamma \vdash e_1~e_2 : \beta}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\Gamma, x:\alpha \vdash e : \beta}
-  {\Gamma \vdash \lambda x.e : \alpha \rightarrow \beta }
-\end{equation*}
-\begin{equation*}
-\tyrule{letbox}
-  {\Gamma \vdash e_1 : \ctyp{r \mseq s}{\alpha} & \Gamma, x : \ctyp{r}{\alpha} \vdash e_2 : \beta }
-  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{s}{\beta}}
-\end{equation*}
-\begin{equation*}
-\tyrule{eval}
-  {\Gamma \vdash e : \ctyp{\munit}{\alpha}}
-  {\Gamma \vdash !e : \alpha}
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\Gamma \vdash e : \ctyp{s}{\alpha} }
-  {\Gamma \vdash e : \ctyp{r}{\alpha} }
-\;(s\leq r)  
-\end{equation*}
-
-\caption{Type system for the coeffect meta-language cml}
-\label{fig:cml-types}
-\end{figure}
-
-% ==================================================================================================
-
-\section{Related work}
-This chapter is closely related to Contextual Modal Type Theory (CMTT) of Nanevski et al. However
-they develop their language using model logic as a basis, while we use categorical foundations 
-as the basis - leading to a different system.
-
-\section{Also related work}
-
-This paper follows the approaches of effect systems \cite{effects-gifford,effects-talpin-et-al,monads-effects-marriage}
-and categorical semantics based on monads and comonads \cite{monad-notions,comonads-notions}. Syntactically,
-\emph{coeffects} differ from \emph{effects} in that they model systems where $\lambda$-abstraction 
-may split contextual requirements between the declaration-site and call-site.
-
-Our \emph{indexed (monoidal) comonads} (\S\ref{sec:comonads}) fill the gap between (non-indexed)
-\emph{(monoidal) comonads} of Uustalu and Vene \cite{comonads-notions}
-and indexed monads of Atkey~\cite{monads-parameterised-notions}, Wadler and Thiemann
-\cite{monads-effects-marriage}. Interestingly, \emph{indexed} comonads are \emph{more
-general} than comonads, capturing more notions of context-dependence (\S\ref{sec:motivation}).
+This thesis aims to change how developers and programming language designers think about 
+\emph{context} or \emph{execution environments} of programs. Such context or execution environment
+has numerous forms -- mobile applications access network, GPS location or user's personal data; 
+journalists obtain information published on the web or through open government data initiatives. 
+This thesis aims to change our understanding of \emph{context} so that the above examples are 
+viewed uniformly through a single programming language abstraction, which we call \emph{coeffects}, 
+rather than as disjoint cases. This is done through the following three contributions.
 
@@
 
-\vspace{-1em}
-\paragraph{Comonads and modal logics.}
+\section{Unified presentation.}
 
-Bierman and de Paiva \cite{logic-intuitionistic-modal} model the
-$\square$ modality of an intuitionistic S4 modal logic using monoidal
-comonads, which links our calculus to modal logics.  This link can be
-materialized in two ways.
+In $\lambda$-calculus, the term \emph{context} is usually used for the free-variable context.
+However, there are many other programming language features that are related to context or program's
+execution environment. In Chapter~\ref{ch:applications}, we revisit many of such features --
+including resource tracking in distributed computing, cross-platform development, data-flow 
+programming and liveness analysis, but also Haskell's type classes and implicit parameters.
 
-Pfenning et al. and Nanevski et al.  derive term languages using the Curry-Howard
-correspondence~\cite{logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt},
-building a \emph{metalanguage} (akin to Moggi's monadic metalanguage
-\cite{monad-notions}) that includes $\square$ as a type
-constructor. For example, in \cite{logic-modal-reconstruction}, the
-modal type $\Box \tau$ represents closed terms.
-In contrast, the \emph{semantic} approach uses monads or comonads
-\emph{only} as a semantics.  This has been employed by Uustalu and
-Vene and (again) Moggi \cite{monad-notions,comonads-notions}.  We
-follow the semantic approach.
+The main contribution of the chapter is that it presents the disjoint language features in a 
+unified way. We show type systems and semantics for many of the languages, illuminating
+the fact that they are closely related.
 
-Nanevski et al. extend an S4 term language to a \emph{contextual}
-modal type theory (CMTT)~\cite{logic-cmtt}.
-The \emph{context} is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. Our contextual types are
-indexed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}.
+Considering applications is one way of approaching the theory of coeffects introduced in this thesis.
+Other pathways to coeffects are discussed in Chapter~\ref{ch:pathways}, which looks at theoretical 
+developments leading to coeffects, including the work on effect systems, comonadic semantics and 
+linear logics.
 
-The work on CMTT suggests two extensions to coeffects. The first is
-developing the logical foundations. We briefly considered special cases
-of our system that permits local soundness in \S\ref{sec:syntactic} and
-local completeness can be treated similarly. The second problem is 
-developing the coeffects \emph{metalanguage}. The use of coeffect algebras
-would provide an additional flexibility over CMTT, allowing a wider range 
-of applications.
+% --------------------------------------------------------------------------------------------------
+
+\section{Flat coeffect calculus.}
+
+The applications discussed in Chapter~\ref{ch:applications} fall into two categories. In the first
+category (Section~\ref{sec:applications-flat}), the additional contextual information are 
+\emph{whole-context} properties. They either capture properties of the execution environment or
+affect the whole free-variable context.
+
+In Chapter~\ref{ch:flat}, we develop a \emph{flat coeffect calculus} which gives us an unified way
+of tracking \emph{whole-context} properties. The calculus is parameterized by a \emph{flat coeffect
+algebra} that captures the algebraic properties that contextual information satisfy. Concrete 
+instances of flat coeffects include Haskell's implicit parameters, whole-context liveness and
+whole-context data-flow.
+
+Our focus is on the syntactic properties of the calculus -- we give a type system 
+(Section~\ref{sec:flat-calculus}) and discuss equational theory of the calculus (Section~\ref{sec:flat-syntax}).
+In the flat coeffect calculus, $\beta$-reduction and $\eta$-expansion do not generally preserve
+type of expressions, but we identify two conditions when this is the case -- this gives us
+a basis for operational semantics of our calculi for liveness and implicit parameters. 
+The design of the type system is further validated by categorical semantics
+(Section~\ref{sec:flat-semantics}) which models flat coeffect calculus in terms of 
+\emph{flat indexed comonad} -- a structure based on categorical dual of monads.
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Structural coeffect calculus.}
+
+In the second category of context-aware systems discussed in Section~\ref{sec:applications-structural},
+contextual properties are \emph{per-variable}. The systems discussed here resemble sub-structural
+logics, but rather than \emph{restricting} variable use, they \emph{track} additional information
+about how variables are used.
+
+We unify systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
+which describes our \emph{structural coeffect calculus} (Section~\ref{sec:struct-calculus}). 
+Similarly to the flat variant, the calculus is parameterized by a \emph{structural coeffect algebra}. 
+Concrete instances of the calculus track bounded variable use (\ie~how many times is variable 
+accessed), data-flow properties (how many past values are needed) and liveness (\ie~can variable 
+be accessed).
+
+The structural coeffect calculus has desirable equational properties (Section~\ref{sec:struct-syntax}).
+In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all structural
+coeffect calculi. This follows from the fact that structural coeffect associates contextual requirements
+with individual variables and preserves the connection by including explicit structural rules
+(weakening, exchange and contraction).
+
+The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
+thesis. In Chapter~\ref{ch:unified}, we develop the two calculi further by introducing a unified
+presentation of the two (by parameterizing the calculus by the \emph{shape} of context) and we present
+an alternative presentation that makes type inference easier.
+
+% --------------------------------------------------------------------------------------------------
 

### 2014-11-17 text/Preface.tex
Unified calculus, move future work (Chapter 6)

%!TEX root = ../main.tex
\pdfbookmark[1]{Preface}{Preface} % Bookmark name visible in a PDF viewer

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Declaration} % Abstract name

This dissertation is my own work and includes nothing which is the outcome of work done in 
collaboration with others except where specifically indicated.
The work presented in Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural} has been previously published
in peer-reviewed papers \cite{coeffects-icalp13} and \cite{coeffects-icfp14}.

The published papers present the key ideas of the dissertation from Chapters~\ref{ch:flat} and 
Chapter~\ref{ch:structural} as well as the motivation discussed in Chapter~\ref{ch:intro}.
They present the flat and structural coeffect calculi and their equational theory as 
developed in this dissertation.

The categorical semantics presented in the published papers has been developed in collaboration and
differs from the one presented in this dissertation. The semantics 
presented here (Section~\ref{sec:flat-semantics} and Section~\ref{sec:struct-semantics}) 
has been developed by the author.

\endgroup

\vfill

### 2014-11-30 text/Chapter04.tex
Finish Ch6 and tweaks everywhere (Chapter 6)

@@
 \subsection{Flat coeffect algebra}
+\label{sec:flat-calculus-algebra}
+
 To make the flat coeffect system general enough, the algebra consists of three operations.
@@
 $\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
-$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids,
-$(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup). That is, 
-for all $r,s,t\in \C$:
-%
-\begin{equation*}
-\begin{array}{ccr}
-r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
-\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
-\textnormal{(monoid)}   
+$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
+idempotent monoid, $(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup). 
+That is, for all $r,s,t\in \C$: 
+%
+\begin{equation*}
+\begin{array}{ccc}
+r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t &&
+\cunit \;\cseq\; r = r = r \;\cseq\; \cunit 
 \\
-r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
-\czero \;\cpar\; r = r = r \;\cpar\; \czero &
-\textnormal{(monoid)}   
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &&
+\czero \;\cpar\; r = r = r \;\cpar\; \czero 
 \\
-r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &
-r\; \czip\; r = r &
-\textnormal{(band)}   
+r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
+r\; \czip\; r = r \qquad\quad
+r\; \cpar\; r = r 
 \\
-\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
-t\; \cleq\; t &
-\textnormal{(pre-order)}   
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &~~~~&
+t\; \cleq\; t 
 \end{array}
@@
 
-\item The monoid $(\C, \cpar, \czero)$ represents \emph{point-wise} composition, \ie~the case when the
-same context is passed to multiple (independent) computations. The monoid laws guarantee 
-that usual syntactic transformations on tuples and the unit value (Section~\ref{sec:flat-exts})
-preserve the coeffect.
+\item The idempotent monoid $(\C, \cpar, \czero)$ represents \emph{point-wise} composition, 
+\ie~the case when the same context is passed to multiple (independent) computations. The monoid 
+laws guarantee that usual syntactic transformations on tuples and the unit value 
+(Section~\ref{sec:flat-exts}) preserve the coeffect. Idempotence holds for all our examples
+and allows us to unify the flat and structural systems in Chapter~\ref{ch:unified}.
 
@@
 
+\noindent
 It is worth noting that the operators $\cpar$ and $\czip$ are dual in some of the systems. For 
@@
 \begin{equation*}
-r \;\cleq\; s \;\Longleftrightarrow\; r \;\cpar\; s \;=\; s
+\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
 \end{equation*}
@@
 \subsection{Understanding flat coeffects}
+\label{sec:flat-calculus-undestanding}
 
@@
 \subsection{Alternative lambda abstraction}
+\label{sec:flat-exts-lambda}
 

### 2014-11-30 text/Chapter06.tex
Finish Ch6 and tweaks everywhere (Chapter 6)

@@
 \item Structural coeffect calculus contains explicit rules for context manipulation
-  (weakening, contraction, exchange). In flat coeffect calculus, these rules are not defined
+  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
   explicitly, but are admissible.
 
-\item In structural coeffect calculus, the variable context is treated as a vector
-  and is annotated with a vector of (scalar) coeffects. In flat coeffect calculus,
+\item In the structural coeffect calculus, the variable context is treated as a vector
+  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
   the variable context is a set and is annotated with a single (scalar) coeffect.
 
-\item In flat coeffect calculus, we distinguish between splitting of the context requirements
-  and merging of context requirements ($\cpar$ and $\czip$, respectively). In structural
+\item In the flat coeffect calculus, we distinguish between splitting of the context requirements
+  and merging of context requirements ($\cpar$ and $\czip$, respectively). In the structural
   coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
@@
 \subsection{Structure of coeffects}
-In structural coeffect calculus, the structure annotation was formed by a vector of coeffect
+In the structural coeffect calculus, the structure annotation was formed by a vector of coeffect
 sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
@@
 \begin{example}
-Structural coeffect shape is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
+\label{ex:unified-struct-shape}
+Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
 where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
@@
 \begin{example}
-Flat coeffect shape is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
+\label{ex:unified-flat-shape}
+Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
 The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
@@
 \sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
-coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero$ is a 
-coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
+coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP \sempty}$
+is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
 combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
@@
  \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
- \alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}   \\
+ \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
 \\[-0.5em]
@@
 
-\vspace{1em}
+\vspace{1.5em}
 {\small b.) Structural rules for context manipulation:}
@@
 
+\vspace{0.5em}
 \caption{Type system for the unified coeffect calculus}
@@
 \subsection{Unified coeffect type system}
-
+\label{sec:unified-types}
 The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
@@
 and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
-
 As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
@@
 
-\paragraph{Differences from structural system.}
-
-% => \azero for constants 
-% => \azip in assumptions (let, abs), \apar in conclusions (let, app)
-% => the same also happened in structural rules
-%      --> how does that make contraction etc. permissible?
-% => <s,r> -> <s> x <r>
-
-Constants (\emph{const}) and variable access (\emph{var}) annotate the context with special
-values. Empty unused context is annotated with $\azero
-\in \C^\sempty$, while singleton context 
-is annotated with $\aunit \in \C^\sunit$.  Note that the shapes
-$\sempty, \sunit$ match the shape of the variable contexts.
-
-Lambda abstraction \emph{splits} the context requirements using $\azip$ into a coeffect 
-$\aclrd{R}$ and a coeffect $\alift{\cclrd{s}}$ of a shape $\sunit$ (semantically, it \emph{merges}
-capabilities provided by the declaration-site and call-site contexts). In structural systems such
-as bounded reuse, $\apar$ is not symmetric and so this gives us a coeffect associated with the bound variable.
-
-The (\emph{app}) rule follows the patterns seen
-earlier -- it uses the scalar-vector multiplication ($\cclrd{t}\,\aseq\,\aclrd{S}$) on 
-coeffects associated with $\Gamma_2$.
-Using the syntactic reading, it then \emph{merges} context requirements for $\Gamma_1$ and $\Gamma_2$.
-In the dual semantic reading,
-it \emph{splits} the provided context into two parts passed to the sub-expressions.
-
-The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
-Syntactically, the context requirements are first split using $\azip$ and then re-combined
-using $\apar$.
+\paragraph{Syntax-driven rules.} 
+The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
+differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
+coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$. 
+The (\emph{const}) rule annotates empty context with a special annotation of the unit shape.
+
+In a structural system, the two annotations correspond to a singleton and empty vector, respectively.
+However, for a singleton shape with one position, the annotations are equivalent to annotating
+variables with a scalar $\cunit$ and constants with a scalar $\czero$.
+
+In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
+system is that vector concatentation $\atimes$ is now replaced with context splitting/merging 
+of the unified coeffect algebra. In particular, we use splitting of context requirements $\azip$
+in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.
+
+Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As 
+discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read 
+the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual 
+information and conclusions split available contextual information.
 
 \paragraph{Structural rules.}
-The coeffect-annotated context can be transformed using structural rules that are
-not syntax-directed. These are captured by (\emph{ctx}), which uses a helper judgment representing
-context transformations $\ctxtrans{ \coctx{\Gamma}{R} }{ \coctx{\Gamma'}{R'} }{\theta}$.
-The rule models that a context used in the rule conclusion $\coctx{\Gamma'}{R'}$ can be transformed
-to a context required by the assumptions $\coctx{\Gamma}{R}$ (using the semantic bottom-up reading).
-In the rule, $\theta$ is a variable substitution generated by the transformation, which is 
-used in the (\emph{contr}) rule.
-
-Exchange and contraction decompose and reconstruct coeffect annotations 
-using $\azip_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\apar_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
-The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slen{\Gamma_1}$
-and $\sclrd{n} = \slen{\Gamma_2}$.
-
-The (\emph{weak}) rule drops an ignored variable annotated with $\aweak$ (compare with (\emph{var})
-annotated using $\aunit$). The (\emph{exch}) rule switches the values 
-while (\emph{contr}) combines them using $\cpar$ to represent sharing of the context.
-Finally, (\emph{sub}) represents sub-coeffecting that can be applied (point-wise) to any
-individual coeffect.
+The merging/splitting operations in the structural rules are changed in the same way as in the
+rules above. It is also worth noting that structural coeffect system used vectors of multiple 
+elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch}) 
+rule denoted a two-element vector. In the unified system, this is replaced with merging/splitting
+of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.
+
+In structural systems, this means the same thing -- we are simply concatenating or splitting two 
+singleton vectors. However, the generalization allows us to capture flat coeffect systems as well.
+The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
+$\apar$ correspond to operations on coeffect scalars. And, as discussed in Section~\ref{sec:unified-flat},
+thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not 
+affect the coeffect annotation are admissible for all flat systems.
 
@@
 
-The coeffect system uses a general notion of context shape, but
-it has been designed with structural and flat systems in
-mind. The structural system is new in this paper and so we
-look at it first.
-
-Recall the coeffect shapes that characterise structural systems: the shape is formed by natural
-numbers (with addition) modelling the number of variables in the
-context. The coeffect algebra is therefore formed by the free monoid
-(vectors) over a coeffect scalar. This means that the system keeps a vector
-of basic coeffect annotations -- one for each variable. An empty
-context (\eg{}, in the (\emph{const}) rule) is annotated with an empty
-vector.
+The unified coeffect system uses a general notion of context shape, but it has been designed with 
+structural and flat systems in mind. In this and the next section, we show how it captures 
+the two kinds of systems.
+
+The unified calculus is modelled after the structural system and so using it to model structural
+systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector} 
+of annotations (Example~\ref{ex:unified-struct-shape}) and define a coeffect algebra formed by
+a vector (free monoid) of scalars.
 
 \begin{definition}
-Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural
-coeffect system} has: 
+\label{def:unified-struct}
+Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system} 
+is defined by:
+
 \begin{itemize}{}
-\item[--] Coeffect shape $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$ formed by natural numbers
+\item[--] Coeffect container $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$ where $\SH=\mathbb{N}$ and
+  $\SP \sclrd{n}=\{ 1 \ldots n \}$ and  ${|}x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n{|}=n$.
+
 \item[--] Coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
   shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
-  That is $\times : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n + m}}$ appends vectors
-  (lists) and $\epsilon : \C^{\sclrd{0}}$ represents empty vectors (lists)
+  That is $\times : \C^{\SP\sclrd{n}} \times \C^{\SP\sclrd{m}} \rightarrow \C^{\SP(\sclrd{n + m})}$ 
+  appends vectors and $\epsilon : \C^{\SP\sclrd{0}}$ represents empty vectors.
 \end{itemize}
@@
 
-\paragraph{Examples.} 
-Defining a concrete structural coeffect system is easy, we just provide the coeffect scalar structure and the rest is free.
-
-\begin{itemize}
-\item
-To recreate the system for bounded reuse, we use coeffect scalars formed by
-$(\mathbb{N}, \ast, +, 1, 0, \leq)$. As in the system of Figure~\ref{fig:bounded-coeff},
-\emph{used} variables are annotated with $1$ and \emph{unused} with $0$. Contraction adds the number
-of uses via $+$ and application (sequencing) multiplies the uses.
-
-\item
-\emph{Dataflow} uses natural numbers (of past values), but differently: $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$.
-Variables are initially annotated with $0$ (and can be incremented using the \kvd{prev} keyword).
-Annotations of a shared variable are combined by taking maximum (of past values needed) and
-sequencing uses $+$.
-
-\item
-Another use of the system is to track \emph{variable liveness}. The annotations are formed by
-$\C=\{ \mathsf{D}, \mathsf{L} \}$ where $\mathsf{L}$ represents a \emph{live} (used) variable
-and $\mathsf{D}$ represents a \emph{dead} (unused) variable. The scalars are given as
-$(\C, \sqcap, \sqcup, \mathsf{L}, \mathsf{D}, \sqsubseteq)$.
-
-In sequential composition ($\sqcap$), a variable is live only if it is required by both of the 
-computations ($\mathsf{L} \sqcap \mathsf{L} = \mathsf{L}$), otherwise it is marked as 
-dead ($\mathsf{D}$). A computation is not evaluated if its result is not needed.
-A shared variable ($\sqcup$) is live if either of the uses is live 
-($\mathsf{D} \sqcup \mathsf{D} = \mathsf{D}$, otherwise $\mathsf{L}$).
-\end{itemize}
-
-\noindent
-Structural liveness is a practically useful, precise version of an example from our earlier
-work, which was a flat system overapproximating liveness to the entire
-context~\cite{petricek2013coeffects}.
+\paragraph{Properties.}
+An important property of the unified system is that, when used in a structural way as discussed
+above, it gives calculi with the same properties as the structural system described in 
+Chapter~\ref{ch:structural}. This can be easily seen by comparing the Figure~\ref{fig:unif-types}
+with the Figure~\ref{fig:struct-types} and using the free monoid interpretation of the unified
+coeffect algebra.
+
+\begin{remark}
+The system described in Definition~\ref{def:unified-struct} is equivalent to the structural coeffect
+system described in Figure~\ref{fig:struct-types}. That is, a typing derivation using a structural
+coeffect embedded in the unified system is valid if and only if the corresponding derivation is
+valid in the structural system.
+\end{remark}
+
+\noindent
+Using the above definition, our unified coeffect system can capture per-variable coeffect properties
+discussed in Chapter~\ref{ch:applications}, Section~\ref{sec:applications-structural}. This includes
+the system for bounded reuse (which is only used in the structural form) and precise tracking 
+of per-variable dataflow and liveness.
 
@@
 
-The same general coeffect system can be used to define systems that track whole-context coeffects
-as in the implicit parameters example 
-(Section~\ref{sec:intro-coeffects-impl}). Flat coeffect systems are
-characterised by a singleton set of shapes, such as ${ \{ \sflat \} }$. In
-this setting, the context annotations $\C^{ \sflat }$ are
-equivalent to coeffect scalars $\C$.
-
-In addition to the coeffect scalar structure, we also need to define $\apar$ and $\azip$.
-Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ (merging of shaped 
-coeffect annotations). However, the $\azip$ operation needs to be provided explicitly.
+The same unified coeffect system can be used to capture systems that track whole-context (flat) 
+coeffects such as implicit parameters. This is achieved using a singleton-shaped container for
+coeffect annotations. The resulting system is has explicit structural rules (and is syntactically
+different from the standard flat coeffect system), but we show that they are equivalent.
+
+Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:unified-flat-shape})
+In this setting, the context annotations $\C^{ \SP \sflat } = \{0\} \rightarrow \C$ are equivalent 
+to coeffect scalars $\C$. In addition to the coeffect scalar structure, we also need to define $\apar$ 
+and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ 
+(merging of shaped coeffect annotations). The $\azip$ operation (corresponding to $\czip$
+in flat coeffect calculus) needs to be provided explicitly.
 
 \begin{definition}
-Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary 
-$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, we define:
+\label{def:unified-flat}
+Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary operation 
+$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, 
+the unified coeffect algebra modelling flat coeffect systems consists of:
+%
 \begin{itemize}{}
-\item[--] Flat coeffect shape $\sclrd{(\{ \sflat \}, \emph{const} \, \sflat, \diamond, \sflat, \sflat)}$ where 
-  $\sflat \diamond \sflat = \sflat$
-\item[--] Flat coeffect algebra $(\cclrd{\wedge}, \cpar, \czero)$, \ie{}, the $\apar = \cpar$ and $\azero = 
-\czero$ with the additional binary operation $\azip = \wedge$. 
+\item[--] Flat coeffect container $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ as defined in 
+  Example~\ref{ex:unified-flat-shape}.
+
+\item[--] Flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$ 
+  and $\azero = \czero$ with an additional binary operation $\azip = \czip$. 
 \end{itemize}
@@
 \noindent
-The requirement $(\cclrd{r \wedge s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$ guarantees 
-exchange and contraction preserve the coeffect of the assumption in the conclusion. Thus, flat coeffect
-calculi do not require substructural-style rules.
-
-\paragraph{Examples.}
-Implicit parameters are the prime example of a flat coeffect system, but other examples
-include rebindable resources \cite{app-distributed-acute} and Haskell type classes \cite{orchard2013thesis}.
-
-In the \emph{implicit parameters} system (Section~\ref{sec:intro-coeffects-impl}), 
-the co\-effect scalars are sets of names with types 
-$\C = \mathcal{P}(\ident{Name} \times \ident{Types}) $. Variables
-are always annotated with $\emptyset$ and coeffects are combined or split using set union $\cup$. Thus
-the system is given by coeffect scalar structure $(\mathcal{P}(\ident{Name} \times \ident{Types}), \cup, \cup, \emptyset, \emptyset, \subseteq)$
-with $\cclrd{\wedge} = \cup$.
-
-\begin{remark}
-We previously described flat systems for \emph{liveness} and \emph{dataflow}~\cite{petricek2013coeffects}.
-Turning a structural system to flat requires finding $\wedge$ that 
-underapproximates the capabilities of combined contexts. For dataflow, this is given by the \textsf{min} function as
-$\mathsf{min}(\cclrd{r},\cclrd{s}) \,\cleq\, \mathsf{max}(\cclrd{r},\cclrd{s})$.
-
-In flat \emph{dataflow}, we annotate the
-entire context with the maximal number of past elements required
-overall. We use the same coeffect scalars $(\mathbb{N}, +, \mathsf{max}, 0, 0, \leq)$
-as in the structural version, but with
-$\cclrd{\wedge} = \ident{min}$. Abstraction (which is the only rule using $\wedge$) becomes:
-%%
-\[
-\tyrule{abs}
-  { \coctx{\Gamma, x\!:\!\sigma}{\ident{min}(\cclrd{r}, \cclrd{s})} \vdash e : \tau }
-  { \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau }
-\]
-%%
-Both the declaration-site and the call-site need to provide at least
-the number of past values required by the body. The overapproximation means that both $\cclrd{r}$
-and $\cclrd{s}$ can be greater than actually required.
-For dataflow, we could annotate
-both contexts with the same coeffect,
-but that would require treating $\azip$ as a partial function.
-\end{remark}
-
-\subsection{Semanitcs}
+Intuitively, the requirement $(\cclrd{r}\,\czip\,\cclrd{s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$,
+which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
+denotes that splitting context requirements and then re-combining them preserves all the requirements
+from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
+it cannot lose requirements. This is fundamental for showing that exchange and contraction are admissible
+in the unified system.
+
+\paragraph{Properties.}
+To show that the typing of flat properties in the unified system is equivalent to the typing in the
+flat system, we show that a valid typing judgement in the first system is a valid typing judgement in 
+the second system and vice versa.
+
+In one direction, we show that the unified system (capturing flat properties) permits weakening, 
+contraction and exchange rules that do not change the coeffect annotations. This guarantees that
+a valid judgement in flat system is also valid in the unified system.
+
+\begin{lemma}
+\label{thm:unified-weak}
+A unified coeffect calculus capturing flat properties admits weakening that does not change the
+coeffect annotation.
+\end{lemma}
+\begin{proof}
+The rule is admissible using the following derivation:
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma}{ \cclrd{r} } \vdash e : \tau}
+      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \;\apar\; \alift{ \czero }} \vdash e : \tau} }
+  { \inference
+      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \,\cpar\, \czero} \vdash e : \tau}      
+      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r}} \vdash e : \tau} }
+\end{equation*}
+
+\noindent
+We write $\cclrd{r}$ rather than $\aclrd{\textbf{r}}$, because we are tracking flat properties.
+The first step is an application of the (\emph{weak}) rule. Next, we use the fact that
+$\apar=\cpar$ and $\alift{\czero}=\czero$, which is the unit of the monoid $(\C, \cpar, \czero)$.
+\end{proof}
+
+\begin{lemma}
+A unified coeffect calculus capturing flat properties admits exchange that does not change the
+coeffect annotation.
+\end{lemma}
+\begin{proof}
+We use the idempotence of $\czip$ and $\cpar$ together with the (\emph{exch}) rule:
+\begin{equation*}
+\inference
+  { \inference
+      { \inference
+          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r}} \vdash e : \tau}
+          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r}} \vdash e : \tau} }
+      { \coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau } }
+  { \inference 
+      {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r}} \apar \alift{\cclrd{r}} \apar \cclrd{r}} \vdash e : \tau}
+      {\inference
+         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}} \vdash e : \tau}
+         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}} \vdash e : \tau} } }
+\end{equation*}
+
+\noindent
+Using idempotence, we first duplicate the annotation $\cclrd{r}$ to get a coeffect in the form
+$\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}$ as required by the
+assumption of the (\emph{exch}) rule. Note that $\alift{-}$ can be added freely as
+$\alift{\cclrd{r}}$ is equivalent to $\cclrd{r}$. After applying (\emph{exch}), we use 
+idempotence of $\apar$.
+\end{proof}
+
+\begin{lemma}
+A unified coeffect calculus capturing flat properties admits contraction that does not change the
+coeffect annotation.
+\end{lemma}
+\begin{proof}
+Similarly to exchange, the proof uses idempotence of $\czip$ and $\cpar$:
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \czip\, \cclrd{r} \czip\, \cclrd{r} \czip\, \cclrd{r}} \vdash e : \tau}
+      {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau} }
+  { \inference
+      {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r} \,\cpar\, \cclrd{r}} \apar \cclrd{r}} \vdash \subst{e}{z,y}{x} : \tau}
+      {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{r} \,\cpar\, \cclrd{r} \,\cpar\, \cclrd{r}} \vdash \subst{e}{z,y}{x} : \tau} }
+\begin{array}{r} \\[2.5em] \qedhere \end{array}
+\end{equation*}
+\end{proof}
+
+\noindent
+In the last two cases, we need to turn the coeffect into a form that is required by the exchange
+and contraction rules. Aside from idempotence, we could use the unit property and obtain 
+\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this appraoch does not 
+work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
+the existence of unit of $\czip$).
+
+In the other direction, we need to show that any valid judgement in the unified system (tracking
+flat properties) is also valid in the flat system. The syntax-directed rules are the same in both
+systems, but we need to show that any use of (explicit) weakening, contraction and exchange can be
+derived in the flat system.
+
+\begin{lemma}
+Weakening, as defined in a unified coeffect calculus capturing flat properties, is admissible in 
+the flat coeffect calculus.
+\end{lemma}
+\begin{proof}
+Similar to the proof in Lemma~\ref{thm:unified-weak}. The property follows from the fact that 
+$\bot=\alift{\czero}$ is the unit of $\cpar$.
+\end{proof}
+
+\begin{lemma}
+\label{thm:unified-rev-contr}
+Contraction, as defined in a unified coeffect calculus capturing flat properties, is admissible in 
+the flat coeffect calculus.
+\end{lemma}
+\begin{proof}
+The application of (\emph{contr}) rule has the following form:
+\begin{equation*}
+\inference
+ { \inference
+    {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}} \vdash e : \tau}
+    {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\azip\, \alift{\cclrd{s}} \,\azip\, \alift{\cclrd{t}} \,\azip\, \cclrd{q}} \vdash e : \tau} }
+ { \inference
+    {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\apar\, \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \,\apar\, \cclrd{q}} \vdash \subst{e}{z,y}{x} : \tau}
+    {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{q}} \vdash \subst{e}{z,y}{x} : \tau} } 
+\end{equation*}
+%
+From Definition~\ref{def:unified-flat}, we know that $(\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}) 
+\;\cleq\; (\cclrd{r} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{q})$. Thus, the judgement
+ can be derived using the (\emph{sub}) rule of flat coeffect calculus.
+\end{proof}
+
+\begin{lemma}
+\label{thm:unified-rev-exch}
+Exchange, as defined in a unified coeffect calculus capturing flat properties, is admissible in 
+the flat coeffect calculus.
+\end{lemma}
+\begin{proof}
+The application of (\emph{exch}) rule has the following form:
+\begin{equation*}
+\inference
+ {\inference
+    {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}} \vdash e : \tau}
+    {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\azip\, \alift{\cclrd{s}} \,\azip\, \alift{\cclrd{t}} \,\azip\, \cclrd{q}} \vdash e : \tau} }
+ {\inference
+    {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\apar\, \alift{\cclrd{t}} \,\apar\, \alift{\cclrd{s}} \,\apar\, \cclrd{q}} \vdash e : \tau}
+    {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{q}} \vdash e : \tau} }
+\end{equation*}
+%
+From idempotence of $\czip$, we get that $\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q} =
+\cclrd{r} \,\czip\, \cclrd{t} \,\czip\, \cclrd{s} \,\czip\, \cclrd{q}$. Thus the judgement can be derived using
+(\emph{sub}) as in contraction.
+\end{proof}
+
+\noindent
+A consequence of the equivalence discussed above is that the unified coeffect system can capture 
+all properties that can be captured by the flat coeffect system -- including implicit parameters,
+rebindable resources, Haskell type classes (discussed by Orchard~\cite{comonads-dom-thesis}),
+data-flow and variable liveness.
+
+%--------------------------------------------------------------------------------------------------
+
+\subsection{Semanitcs of unified calculus}
 \label{sec:unified-semantics}
 
-Not really - but sketch.
-
-%===================================================================================================
-
-\section{Semi-lattice formulation}
-\label{sec:unified-semilattice}
-
-\subsection{Semi-lattice formulation}
-
-\begin{equation*}
-\begin{array}{lcll}
- \ident{counit}_{\cunit} &\narrow{:}& 
-    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
- \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
-    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
-\ident{merge}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
-    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)} \\
-\ident{split}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
-    \ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \\
-\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
-    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} & 
-    (\cclrd{s}\,\cleq\,\cclrd{r})
+The semantics of unified calculus can be obtained by generalizing the semantics of the structural
+calculus in the same way as we generalized the type system. However, it is worth exploring the type 
+signatures of the operations of the underlying (comonad-based) structure.
+
+The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
+structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The 
+index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of 
+the structure $\C^{\SP \sclrd{n}}$ for a specified container $\SH \triangleleft \SP$ and a shape 
+$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in 
+Section~\ref{sec:unified-types}, we need to modify the domains and codomains of the \ident{merge}, 
+\ident{split} and \ident{dup} operations as follows.
+%
+\begin{equation*}
+\begin{array}{lcl}
+ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
+  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
+ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
+  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
+ \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
+  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
 \end{array}
 \end{equation*}
-
-Change to
-
+%
+The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
+\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
+(in \ident{merge}, but also in \ident{dup} which models contraction).
+This means that the splitting and merging the context can change the coeffect annotation:
+%
+\begin{equation*}
+\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
+  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
+  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} 
+\end{equation*}
+%
+For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
+splitting and then merging a context may lose some of the avilable contextual capabilities
+(and increase context requirements). This corresponds to the fact that our proofs of 
+Lemma~\ref{thm:unified-rev-contr} and Lemma~\ref{thm:unified-rev-exch} relied on sub-coeffecting
+in the flat system.
+
+\paragraph{Related and future work.}
+The semantics presented here serves mainly to inform the design of the type system. For this 
+reason, we keep the semantics simple and close to the concrete notions used in the type system
+such as flat, structural and unified coeffect algebras. In particular, the indexed comonad 
+structure consists of object mapping $\ctyp{r}{}$ that is indexed by a corresponding algebra:
+
+\begin{compactitem}
+\item[-] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
+\item[-] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is 
+  indexed by vectors of scalars $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$
+\item[-] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
+  the mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is indexed by shape-indexed scalars 
+  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP\sclrd{m}}$
+\end{compactitem}
+
+\noindent
+In this treatment, the indices are formed by objects that are modelled outside of category theory.
+A fully categorical semantics can be found in a joint paper with Orchard \cite{coeffects-icfp14}. 
+There, the index (coeffect scalars and shape-indexed scalars) are modelled as categories too.
+The definition requires the following notation:
+
+\begin{compactitem}
+\item[-] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
+\item[-] $B^A$ is an exponential object $A \Rightarrow B$ in cartesian closed category
+\end{compactitem}
+
+\noindent
+In the categorical semantics of coeffect calculi, coeffect scalars are modelled as a
+category of scalars $\cclrd{\mathbb{I}}$ such that $\textit{obj}(\cclrd{\mathbb{I}}) = \C$.
+The structure of the context (indexed comonad) over the input then becomes a functor
+indexed by the free-variable context shape $\slen{\Gamma}$ and coeffect annotations
+$\aclrd{\textbf{r}}$, \ie~$\ctyp{\slen{\Gamma}}_{\aclrd{\textbf{r}}}$:
+%
+\begin{equation*}
+\sem{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \ctyp{\slen{\Gamma}}_{\aclrd{R}}
+\sem{\Gamma} \rightarrow \sem{\tau}
+\end{equation*}
+%
+The structure $\ctyp{}$ can be thought of as a dependent product of functors $\ctyp{\sclrd{n}}$ 
+over possible shapes $\sclrd{n} \in \SH$ where:
+%
+\begin{equation*}
+\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP \sclrd{n}} \rightarrow [\mathbb{C}^{\SP \sclrd{n}}, \mathbb{C}]
+\end{equation*}
+%
+For a fixed context shape $\sclrd{n}$ the functor 
+$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP \sclrd{n}} \rightarrow [\mathbb{C}^{\SP \sclrd{n}}, \mathbb{C}]$ 
+maps a coeffect indexed by positions $\SP \sclrd{n}$ to a functor from a context 
+$\mathbb{C}^{\sclrd{n}}$ to an object in $\mathbb{C}$. That is, given a coeffect annotation 
+(matching the shape of the context), we get a functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$. 
+From a programming perspective, this functor defines a data structure that models the additional
+context provided to the program. The shape of this data structure depends on the coeffect
+annotation $\cclrd{\mathbb{I}}^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
+the number of values needed for each variable and the functor will be 
+formed by lists of length matching the required number.
+
+In the model above, shapes and positions are still treated as ordinary sets. Indeed, the paper
+\cite{coeffects-icfp14} treats shapes as just sets of sets. In this chapter, we instead define
+\emph{coeffect container} as an extension of containers, which have a well-defined 
+categorical structure \cite{semantics-containers}. A future work is adapting the semantics of
+the unified coeffect calculus so that it is based on the semantics of containers. However, note 
+that context is \emph{not} a (simple) container, because containers are used as indices (coeffect
+annotations) of the object that wrap the free-variable context.
+
+% =================================================================================================
+%                                                                   
+%      ####                         #       #                   ##   
+%      #   #                        #                            #   
+%      #   #  # ##    ###    ###   ####    ##     ###    ###     #   
+%      ####   ##  #      #  #   #   #       #    #   #      #    #   
+%      #      #       ####  #       #       #    #       ####    #   
+%      #      #      #   #  #   #   #  #    #    #   #  #   #    #   
+%      #      #       ####   ###     ##    ###    ###    ####   ###  
+%                                                                
+% =================================================================================================
+
+\section{Towards practical coeffects}
+\label{sec:unified-impl}
+
+As discussed earlier, the main focus of this thesis is the development of the much needed 
+\emph{theory of context-aware computations} and so discussing the details of a practical 
+implementation of coeffect tracking is beyond the scope of the thesis. However, this section
+briefly outlines one possible pathway towards this goal.
+
+Many of the examples of contextual computation that we discussed earlier have been implemented as 
+a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters} 
+or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However, 
+the main contribution of this thesis is that it captures
+\emph{multiple} different notions of context-aware computations using just a \emph{single}
+common abstraction. For this reason, we advocate that future practical implementations of
+coeffects should not be single-purpose language features, but instead reusable abstractions that
+can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
+
+In order to do this, programming languages need to provide two features; one that allows 
+embedding of context-aware computations themselves in programs akin to the ``do'' notation
+in Haskell (Section~\ref{sec:unified-impl-embed}) and one that allows tracking of the 
+contextual information in the type system (Section~\ref{sec:unified-impl-types}). To make
+constraint resolution in type inference easier, we also discuss how the coeffect algebra 
+can be simplified (Section~\ref{unified-impl-semilattice})
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding contextual computations}
+\label{sec:unified-impl-embed}
+
+The embedding of \emph{contextual} computations in programming languages can learn from better 
+explored emedding of \emph{effectful} computations. In purely functional programming languages
+such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
+necessary (monadic) plumbing. This is made easier by the ``do'' notation \cite{other-haskell98} 
+that inserts the monadic operations automatically.
+
+The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
+context-aware computations based on comonads. The notation is close to the semantics of our 
+flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based 
+on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
+this requires explicitly manipulating individual context variables and application of structural
+rules, which is not needed in flat coeffects.
+
+In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
+but they can still use a special notation for explicitly marking effectful (coeffectful) 
+blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
+that differ from the ``do'' notation in two ways. First, they support wider range of 
+constructs, making it possible to wrap existing F\# code in a block without other changes.
+Second, they support other abstractions including monads, applicative functors and monad
+transformers. It would be interesting to see if computation expressions can be extended to
+handle computations based on flat/structural indexed comonads.
+
+More lightweight syntax for effectful computation can be obtained using
+techniques that automatically insert the necessary monadic plumbing (wihtout a
+special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
+and similar approach would be worth exploring for coeffects.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect annotations as types}
+\label{sec:unified-impl-types}
+
+The other aspect of practical implementation of coeffects is tracking the context requirements
+(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
+language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
+The structures used in this thesis include sets (with union or interesection), natural numbers (with 
+addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
+(vectors of annotations).
+
+The work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
+recent additions to the Haskell type system provide enough poer to implement structures such
+as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful 
+future direction for applied coeffects.
+
+In dependently-typed programming languages such as Agda \cite{other-agda} or Idris \cite{other-idris},
+the embedding of coeffects can be implemented more directly. However, we believe that coeffect
+tracking does not require full dependent types and can be made accessible in more main-stream
+languages. Dependent ML \cite{types-dependent-ml} provides an interesting example of a language
+with some dependent typing which is still close to its non-dependently-typed predecessor ML.
+
+Another approach for embedding computations into the type system has been pioneered by F\#
+\emph{type providers} \cite{app-inforich}. Technically, type providers are compiler extensions 
+that generate types based on external data sources or other information in order to provide easy 
+access to data or services. A similar approach could be used for embedding \emph{algberas} such
+as coeffect algebras into the type system. A \emph{algebra provider} would be a library that 
+specifies the objects of the algebra, equational laws and generalization rules for type inference.
+This could provide an easy to use way of embedding coeffect tracking in pragmatic languages
+such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
+\cite{types-units-of-measure}, which could be provided via one such \emph{algebra provider}.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Alternative formulation using coeffect lattice}
+\label{sec:unified-impl-semilattice}
+
+The fact that coeffect scalar structure consists of four operations ($\cseq,\czip,\cpar$ 
+and a relation $\cleq$) makes working with coeffect annotations difficult -- it compilcates tasks
+such as type generalization and constraint solving in type inference. Here, we look at one possible 
+simplification of the structure.
+
+\begin{itemize}
+\item As mentioned in Section~\ref{sec:flat-calculus-algebra}, when the monoid $(\C,\cpar,\czero)$ is a
+semi-lattice, the $\cleq$ relation can often be expressed in terms of the $\cpar$ operation 
+($\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}$).
+This does not give a definition equivalent to our, but it is a simplification that is consistent 
+with all examples in this thesis.
+
+\item Another approach that works for most (but not all) systems we discussed is to require the structure
+$(\C,\cpar,\czip)$ to form a lattice. This is consistent with all our structural coeffect systems as
+well as flat systems for liveness and data-flow, but not with the system for tracking of implicit
+parameters (and other sets).
+\end{itemize}
+
+\noindent
+The system for tracking implicit parameters does not fit the lattice-based model, because it uses the 
+same operation $\cup$ for both $\cpar$ and $\czip$. This also means that the system for implicit parameters 
+cannot be used with a system where lambda abstraction duplicates the context requirements 
+(Section~\ref{sec:flat-exts-lambda}).
+
+\paragraph{Lattice-based coeffects.} Putting implicit parameters aside, there are many other coeffect
+systems for which we could use the following definition of coeffect scalars (for simplicity, we only 
+consider flat systems):
+
+\begin{definition}
+A \emph{\cclrd{flat coeffect lattice}} $(\C, \cseq, \cleq, \cunit, \czero)$ is a set 
+$\C$ together with elements $\cunit, \czero \in \C$ and an ordering $\cleq$ such 
+that $(\C, \cseq, \cunit)$ is a monoid and $(\C, \cleq)$ is a lattice with the least 
+element $\czero$.
+
+Assuming $\cpar$ and $\czip$ are the \emph{least upper bound} and \emph{greatest lower bound}
+of the lattice, we require the following distributivity axioms:
+%
+\begin{align*}
+\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+\end{definition}
+
+\noindent
+The definition provides the same structure as flat coeffect algebra, but it defines a more rigid 
+structure. In particular, $\czip$ and $\cpar$ are defined in terms of $\cleq$. This simplifies the 
+system as the operations of the underlying indexed comonad structure can be expressed using just
+$\cseq$ and $\cleq$:
+%
 \begin{equation*}
@@
 \end{equation*}
-
-Gives us..
+%
+Note that we still need two \emph{distinct} operators in the system. The one used for
+sequential composition $\cseq$ is not necessarilly related in any way to the ordering on
+the lattice $\cleq$.
+
+However, the operations $\ident{merge}$, $\ident{split}$ and $\ident{lift}$ can now all
+use ordering. For example, the \ident{merge} operation previously returned
+a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
+element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit sub-coeffecting
+to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and 
+$\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
+This means that we can compose the \ident{merge} operation which returns greatest lower
+bound with $\ident{lift}$ to get the (less precise) operation required here.
+
+% -------------------------------------------------------------------------------------------------
 
 \begin{figure}[t]
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\cunit} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{const}
-  {c : \tau \in \Delta}
-  {\coctx{\Gamma}{\czero} \vdash c : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
+{\small a.) Littice-based flat systems (selected rules):}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{t}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+~~(\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s})
 \end{equation*}
@@
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-
-\caption{Type system for the flat coeffect calculus}
-\label{fig:flat-types2}
+  {\coctx{\Gamma}{\cclrd{u}} \vdash e_1~e_2 : \tau_2}
+~~(\cclrd{r}\,\cleq\,\cclrd{u}, (\cclrd{s} \,\cseq\, \cclrd{t})\,\cleq\,\cclrd{u})
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Littice-based structural systems (selected rules):}
+\begin{equation*}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{u}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~~(\cclrd{s}\,\cleq\,\cclrd{u}, \cclrd{t}\,\cleq\,\cclrd{u})
+\end{equation*}
+
+\caption{Flat and structural coeffects using lattice-based formulation}
+\label{fig:unified-lattice-types}
 \end{figure}
 
-q
-\newpage
-a
-\newpage
-a
-\newpage
-
-\subsection{Type inference algorithm}
-
-% ==================================================================================================
-
-% =================================================================================================
-%                                                                   
-%      ####                         #       #                   ##   
-%      #   #                        #                            #   
-%      #   #  # ##    ###    ###   ####    ##     ###    ###     #   
-%      ####   ##  #      #  #   #   #       #    #   #      #    #   
-%      #      #       ####  #       #       #    #       ####    #   
-%      #      #      #   #  #   #   #  #    #    #   #  #   #    #   
-%      #      #       ####   ###     ##    ###    ###    ####   ###  
-%                                                                
-% =================================================================================================
-
-\section{Towards practical coeffects}
-\label{sec:unified-impl}
-
-As discussed earlier, the main focus of this thesis is the development of the much needed 
-\emph{theory of context-aware computations} and so discussing the details of a practical 
-implementation of coeffect tracking is beyond the scope of the thesis. However, this section
-briefly outlines one possible pathway towards this goal.
-
-Many of the examples of contextual computation that we discussed earlier have been implemented as 
-a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters} 
-or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However, 
-the main contribution of this thesis is that it captures
-\emph{multiple} different notions of context-aware computations using just a \emph{single}
-common abstraction. For this reason, we advocate that future practical implementations of
-coeffects should not be single-purpose language features, but instead reusable abstractions that
-can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
-
-In order to do this, programming languages need to provide two features; one that allows 
-embedding of context-aware computations themselves in programs (akin to the ``do'' notation
-in Haskell) and one that allows tracking of the contextual information in the type system.
-
 % -------------------------------------------------------------------------------------------------
 
-\subsection{Embedding contextual computations}
-
-The embedding of \emph{contextual} computations in programming languages can learn from better 
-explored emedding of \emph{effectful} computations. In purely functional programming languages
-such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
-necessary (monadic) plumbing. This is made easier by the ``do'' notation \cite{other-haskell98} 
-that inserts the monadic operations automatically.
-
-The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
-context-aware computations based on comonads. The notation is close to the semantics of our 
-flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based 
-on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
-this requires explicitly manipulating individual context variables and application of structural
-rules, which is not needed in flat coeffects.
-
-In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
-but they can still use a special notation for explicitly marking effectful (coeffectful) 
-blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
-that differ from the ``do'' notation in two ways. First, they support wider range of 
-constructs, making it possible to wrap existing F\# code in a block without other changes.
-Second, they support other abstractions including monads, applicative functors and monad
-transformers. It would be interesting to see if computation expressions can be extended to
-handle computations based on flat/structural indexed comonads.
-
-More lightweight syntax for effectful computation can be obtained using
-techniques that automatically insert the necessary monadic plumbing (wihtout a
-special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
-and similar approach would be worth exploring for coeffects.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect annotations as types}
-
-The other aspect of practical implementation of coeffects is tracking the context requirements
-(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
-language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
-The structures used in this thesis include sets (with union or interesection), natural numbers (with 
-addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
-(vectors of annotations).
-
-The work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
-recent additions to the Haskell type system provide enough poer to implement structures such
-as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful 
-future direction for applied coeffects.
-
-In dependently-typed programming languages such as Agda \cite{other-agda} or Idris \cite{other-idris},
-the embedding of coeffects can be implemented more directly. However, we believe that coeffect
-tracking does not require full dependent types and can be made accessible in more main-stream
-languages. Dependent ML \cite{types-dependent-ml} provides an interesting example of a language
-with some dependent typing which is still close to its non-dependently-typed predecessor ML.
-
-Another approach for embedding computations into the type system has been pioneered by F\#
-\emph{type providers} \cite{app-inforich}. Technically, type providers are compiler extensions 
-that generate types based on external data sources or other information in order to provide easy 
-access to data or services. A similar approach could be used for embedding \emph{algberas} such
-as coeffect algebras into the type system. A \emph{algebra provider} would be a library that 
-specifies the objects of the algebra, equational laws and generalization rules for type inference.
-This could provide an easy to use way of embedding coeffect tracking in pragmatic languages
-such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
-\cite{types-units-of-measure}, which could be provided via one such \emph{algebra provider}.
+\paragraph{Lattice-based type system.}
+
+Using lattice-based coeffect structure leads to the typing rules shown in Figure~\ref{fig:unified-lattice-types}.
+In the flat coeffect system, this affects both application and abstraction -- in abstraction,
+the context-requirements of the body should be smaller than the context available on the 
+call-site and declaration-site. In application, we require more than what is required by
+the sub-expressions.
+
+Structural coeffect systems only require least upper bound $\cpar$ in the (\emph{contr}) rule.
+This can be rewritten similarly to flat application --  when contracting two variables, the
+context requirements of the resulting variable are greater than the requirements associated with
+each of the original variables.
+
+In smmary, the lattice-based coeffect system has some advantages over the design used in this
+thesis. In particular, it uses better-known structures and is simpler. However, we do not use
+it as it rules out one of the important motivating examples (implicit parameters) and, even with
+the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.
 
@@
 
+% ==================================================================================================
+%                                                                             
+%        ###                         ##                    #                        
+%       #   #                         #                                             
+%       #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###  
+%       #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #     
+%       #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###  
+%       #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      # 
+%        ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####  
+%                                                                             
+% ==================================================================================================
+
 \section{Conclusions}
+
+In this chapter, we completed our goal of presenting a \emph{unified} system for tracking contextual
+properties and then we discussed two important topics of future and related work that require 
+in-depth discussion.
+
+The key technical contribution of this chapter is the \emph{unified coeffect system} 
+(Section~\ref{sec:unified-unified}), which unifies the flat and structural system presented
+in the previous two chapters. To achieve this, we introduced \emph{coeffect container}, which
+determines how are coeffect annotations attached to variable contexts. We then discussed two
+instances of the structure that capture flat and structural properties.
+
+In the rest of the chapter, we discussed two topis of future and related work. First
+(Section~\ref{sec:unified-impl}), we considered pathways to practical implementations of
+coeffect systems, including a discussion of a novel lattice-based coeffect system, which is simpler,
+but cannot capture all our motivating examples. Finally, we discussed how our work relates to 
+meta-language based on comonads (Section~\ref{sec:unified-meta}). We present a \emph{coeffect 
+meta-language} that follows similar style to CMTT, but is based on indexed comonads. Next, we 
+discussed how to embed flat coeffect systems in this coeffect meta-language, which elucidates the 
+relationship between our work and CMTT.

### 2014-12-01 text/Abstract.tex
Proofreading chapters 1 and 2, reorder chapter 6

@@
 
-The key point made by this thesis is the realization that execution environment or
-\emph{context} is fundamental for writing modern applications and that programming 
+The key point made by this thesis is the realization that an execution environment or
+a \emph{context} is fundamental for writing modern applications and that programming 
 languages should provide abstractions for programming with context and verifying how 
@@
 calculus} models languages where the contextual properties are attached to the variable usage.
+We close the thesis with a \emph{unified coeffect calculus} that captures the two notions
+using a single parameterized system.
 

### 2014-12-01 text/Chapter01.tex
Proofreading chapters 1 and 2, reorder chapter 6

@@
   (\ie~enhance the user experience, but there is a fallback strategy). Equally, on the server-side,
-  we might have access to different database tables and other information sources.
+  we might have access to different database tables, depending on the role of the user.
 \end{itemize}
@@
 \noindent
-Most developers do not perceive the above as programming language flaws -- they are simply common
+Most developers do not perceive the above as programming language flaws. They are simply common
 programming problems (at most somewhat annoying and tedious) that have to be solved. However, this
@@
 \prepk{\#else}\\[-0.25em]
-\hspace{2em}|~\str{user-agent} \rightarrow~
+\hspace{2em}|~\str{referer} \rightarrow~
   \ident{req.Headers}.[\,\ident{HttpHeader.Referer}\,] \leftarrow \ident{value}\\[-0.25em]
@@
 
-\vspace{1em}
-\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
-\vspace{-1.5em}
 \caption{Conditional compilation in the HTTP module of the F\# Data library}
@@
 introducing the technique ``Remember the mantra: if you haven't tried it, it doesn't work.'' 
-Again, it would be reasonable to expect that statically-typed languages could provide a better solution.
+Again, it would be reasonable to expect that statically-typed languages can provide a better solution.
 
@@
 
-Another example related to the previous one is when libraries use meta-programming techniques
-(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
-written in a subset of a host language to some other target language, such as SQL, OpenCL or
-JavaScript. For database access, this is a recently developed technique replacing embedded SQL
-discussed in the introduction, but it is a more generally important technique for programming
+Another example related to the previous one is when libraries use meta-programming techniques,
+such as LINQ \cite{app-linq,app-linq-essence} or F\# quotations \cite{app-fsharp-metaprog}, to 
+translate code written in a subset of a host language to some other target language, such as SQL, 
+OpenCL or JavaScript. For database access, this is a recently developed technique replacing embedded 
+SQL discussed in the introduction, but it is a more broadly applicable technique for programming
 in heterogeneous environments. It lets developers targets multiple runtimes that have limited 
@@
 The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
-Google search returns 11800 results for the message above and even more results (44100) for a
+Google search returns 11,800 results for the message above and even more results (44,100) for a
 LINQ error message \emph{``Method X has no supported translation to SQL''} caused by a similar
@@
 the whole program or a function scope. However, contextual properties can also be associated with
-a specific variables.
+specific variables.
 
@@
 The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
-\cite{app-sql-injection} (an attacker could enter \str{'; DROP TABLE Products -{}-} as their 
-\ident{name} and delete the database table ``Products''). For this reason, most libraries discourage 
+\cite{app-sql-injection}. An attacker could enter \str{'; DROP TABLE Products -{}-} as their 
+\ident{name} and delete the database table ``Products''! For this reason, most libraries discourage 
 building SQL commands by string concatenation, but there are still many systems that do so.
 
-The example demonstrates a more general property. Sometimes, it is desirable to track additional
+Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
 meta-data about variables that are in some ways special. Such meta-data can determine how the variables
@@
 \cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
-tracking. This can be useful when working with data in other contexts. For example, 
-data jounralsts might want to propagate meta-data about the quality and the information source -- 
-is the source trustworthy? Is the data up-to-date? Such meta-data could propagate to 
-the result and tell us important information about the calculated results.
+tracking \cite{app-provenance-future}. This can be useful when working with data in other contexts. 
+For example, data jounralsts might want to propagate meta-data about the quality and the 
+information source -- is the source trustworthy? Is the data up-to-date? Such meta-data could 
+propagate to the result and tell us important information about the calculated results.
 
@@
 by preallocating the necessary space (as we know which sub-range of the array might be accessed).
-It also allows better handling of boundaries. For example, to simplify wrap-around behaviour we 
-could pad the \ident{input} with a known number of elements from the other side of the array.
+It also allows better handling of boundaries \cite{app-stencil-dsl}. For example, to simplify 
+wrap-around behaviour we could pad the \ident{input} with a known number of elements from the 
+other side of the array.
 
@@
 The context includes notions such as execution environment, capabilities provided by the 
-environment or input and meta-data about the input.
+environment or input and meta-data about the input and variables through which it is accessed.
 
@@
 
-\vspace{1em}
-\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
-\vspace{-1.5em}
-\caption{News reader implemented in a context-aware language}
+\caption{Client/server news reader app implemented in a context-aware language}
 \label{fig:introduction-sample}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\newcommand{\introfuncblack}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \{ #2 \}} $\rightarrow$ \ident{#3}  }
+\newcommand{\introfunc}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \aclrd{\{ #2 \}}} $\rightarrow$ \ident{#3}  }
+\newcommand{\introvar}[2]{ \ident{#1 \hspace{-1em}\footnotesize{@} \cclrd{#2} } }
 
@@
 The approach advocated in this thesis is to track information about context requirements using
-the type system. To make this practical, the system needs to provide at least partial support for
+the type system. To make this practical, the system should also provide at least a partial support for
 automatic type inference, as the information about context requirements makes the types more complex.
@@
 \definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
-\newcommand{\introfunc}[3]{ \ident{#1 \hspace{-1em}\footnotesize{@} \{ #2 \}} $\rightarrow$ \ident{#3}  }
-\newcommand{\introvar}[2]{ \ident{#1 \hspace{-1em}\footnotesize{@} #2 } }
 \vspace{0.5em}
@@
 \hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
-	\introfunc{unit}{gps, rpc}{(news~list)~async}
+	\introfuncblack{unit}{gps, rpc}{(news~list)~async}
 }
@@
 \item The \ident{fetchNews} function requires database access and so it can only run on the server-side
-  (or on a thick client with local copy of the database, such as a desktop computer with an offline mode).
+  (or on a thick client with a local copy of the database, such as a desktop computer with an offline mode).
 
@@
 directives. Similarly, the splitting between client-side, server-side and shared code can be
-done using ordinary functions and modules -- rather than having to split the application into
-separate independent libraries or projects.
+done using ordinary functions and modules (with shared functions reused) -- rather than having 
+to split the application into separate independent libraries or projects.
 
@@
 \ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
-the following is a valid typing judgement in F\#:
+the following is a valid typing judgement in F\# \cite{app-fsharp}:
 %
@@
 the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
-an unsecure network connection.
-The \emph{coeffect systems} developed in this thesis capture this information in the following way:
+an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this 
+information in the following way (we slightly refine the notation later):
 %
@@
  \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
-   \ident{cmd}\!:\!\ident{string})}{ \cclrd{\{ \ident{database} \}}}  \\[-0.1em]
+   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
  \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
@@
 is that \emph{coeffects}, or systems for tracking contextual information, correspond to the category 
-theoretical dual of monads called \emph{comonads}. The details are explained when discussing the 
-semantics of coeffects throughout the thesis.
+theoretical dual of monads called \emph{comonads}. The details are explained throughout the thesis.
 
@@
   that the idea of context-aware computations can be naturally approached from a number of
-  directions developed recently in theories of programming languages (including type and effect
+  directions developed recently in the theory of programming languages (including type and effect
   systems, categorical semantics and sub-structural logics).
@@
 \item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis. 
-  One important direction deserves further exploration, and so Chapter~\ref{ch:conclusions}
-  starts with a brief discussion of a different approach to tracking contextual information that 
-  arises from modal logics. Finally, the rest of Chapter~\ref{ch:conclusions} discusses approaches 
-  for implementing the presented theory in main-stream programming languages and concludes.
+  One important direction deserves further exploration, and so Chapter~\ref{ch:unified}
+  also includes a brief discussion of a different approach to tracking contextual information that 
+  arises from modal logics. This is followed by an outline of the future work that is needed
+  in order to implement the presented theory in main-stream programming languages.
 \end{itemize}

### 2014-12-01 text/Chapter02.tex
Proofreading chapters 1 and 2, reorder chapter 6

@@
-% "Language is a labyrinth of paths. You approach from one side and know your way about; you 
-% approach the same place from another side and no longer know your way about.”
+%!TEX root = ../main.tex
+
+% ==================================================================================================
 
@@
 
-Type and effect systems are usually specified judgements of the form $\Gamma \vdash e : \alpha, \sigma$, 
-meaning that the expression $e$ has a type $\alpha$ in (free-variable) context $\Gamma$ and 
-additionally may have effects described by $\sigma$. Effect systems are typically added to a 
+Type and effect systems are usually specified as judgements of the form $\Gamma \vdash e : \tau, \cclrd{r}$, 
+meaning that the expression $e$ has a type $\tau$ in a (free-variable) context $\Gamma$ and 
+additionally may have effects described by $\cclrd{r}$. Effect systems are typically added to a 
 language that already supports effectful operations as a way of increasing the safety -- the type 
@@
 to \emph{prescriptive} effect systems that implement computational effects in a pure language 
--- such as monads in Haskell}.
+-- such as monads in Haskell.}.
 
@@
 \begin{equation*}
-\inference[(var)]
-  {x:\alpha \in \Gamma }
-  {\Gamma \vdash x : \alpha, \emptyset }
-\quad
-\inference[(write)]
-  {\Gamma \vdash e : \alpha, \sigma & l:\ident{ref}_\rho~\alpha\in \Gamma}
-  {\Gamma \vdash l \leftarrow e : \ident{unit}, \sigma \cup \{\ident{write}(\rho)\} }
+\tyrule{var}
+  {x\!:\!\tau \in \Gamma }
+  {\Gamma \vdash x : \tau, \cclrd{\emptyset} }
 \end{equation*}
 \begin{equation*}
-\inference[(fun)]
-  {\Gamma, x:\alpha_1 \vdash e : \beta, \sigma }
-  {\Gamma \vdash \lambda x.e : \alpha \xrightarrow{\sigma} \beta, \emptyset }
-\quad  
-\inference[(app)]
-  {\Gamma \vdash e_1 : \alpha \xrightarrow{\sigma_1} \beta, \sigma_2 \\
-   \Gamma \vdash e_2 : \alpha, \sigma_3 }
-  {\Gamma \vdash e_1~e_2 : \beta, \sigma_1 \cup \sigma_2 \cup \sigma_3 }
+\tyrule{write}
+  {\Gamma \vdash e : \tau, \cclrd{r} & l:\ident{ref}_\rho~\tau\in \Gamma}
+  {\Gamma \vdash l \leftarrow e : \ident{unit}, \cclrd{r} \cup \cclrd{\{\ident{write}(\rho)\}} }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x\!:\!\tau_1 \vdash e : \tau_2, \cclrd{r} }
+  {\Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2, \cclrd{\emptyset} }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_1 \xrightarrow{\cclrd{r}} \tau_2, \cclrd{s} &
+   \Gamma \vdash e_2 : \tau_1, \cclrd{t} }
+  {\Gamma \vdash e_1~e_2 : \tau_2, \cclrd{r} \cup \cclrd{s} \cup \cclrd{t} }
 \end{equation*}
@@
 
-\paragraph{Simple effect system}
-The structure of a simple effect system is demonstrated in Figure~\ref{fig:path-eff}. The example
+\subsection{Simple effect system.}
+The structure of a simple effect system\footnote{Most work on effect systems uses $\sigma$ for 
+effect annotations. We use letters $\cclrd{r}, \cclrd{s}, \cclrd{t}$ and also distinguish effect 
+or coeffect annotations by colour.} is demonstrated in Figure~\ref{fig:path-eff}. The example
 shows typing rules for a simply typed lambda calculus with an additional (effectful) operation
 $l \leftarrow e$ that writes the value of $e$ to a mutable location $l$. The type of locations
-($\ident{ref}_\rho~\alpha$) is annotated with a \emph{memory region} $\rho$ of the location $l$.
+($\ident{ref}_\rho~\tau$) is annotated with a \emph{memory region} $\rho$ of the location $l$.
 The effects tracked by the type and effect system over-approximate the actual effects and memory
@@
 has no effects, application (\emph{app}) combines the effects of both expressions, together with 
-the latent effects of the function to be applied. Finally, lambda abstraction (\emph{fun}) is a
+the latent effects of the function to be applied. Finally, lambda abstraction (\emph{abs}) is a
 pure computation that turns the \emph{actual} effects of the body into \emph{latent} effects of 
@@
 
-\paragraph{Simple coeffect system}
+\subsection{Simple coeffect system.}
 When writing the judgements of coeffect systems, we want to emphasize the fact that coeffect 
 systems talk about \emph{context} rather than \emph{results}. For this reason, we write the 
-judgements in the form $\Gamma @ \sigma \vdash e : \alpha$, associating the additional information
-with the context (left-hand side) of the judgement rather than with the result (right-hand side)
-as in $\Gamma \vdash e : \alpha, \sigma$. This change alone would not be very interesting -- we 
-simply used different syntax to write a predicate with four arguments. As already mentioned, the 
-key difference follows from the lambda abstraction rule.
+judgements in the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$, associating the additional 
+information with the context (left-hand side) of the judgement rather than with the result 
+(right-hand side) as in $\Gamma \vdash e : \tau, \cclrd{r}$. This change alone would not be 
+very interesting -- we simply used different syntax to write a predicate with four arguments. 
+The more interesting difference is how the lambda abstraction rule looks.
 
@@
 
-The (\emph{fun}) rule is different -- the resources requirements of the body $\sigma_1 \cup \sigma_2$
-are split between the \emph{immediate context-requirements} associated with the current context 
-$\Gamma @ \sigma_1$ and the \emph{latent context-requirements} of the function.
+However, the (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
+resource requirements of the body $\cclrd{r} \cup \cclrd{s}$ are split between the \emph{immediate 
+context-requirements} associated with the current context $\coctx{\Gamma}{\cclrd{r}}$ and the 
+\emph{latent context-requirements} of the function.
 
-As demonstrated by examples in the Chapter~\ref{ch:intro}, this means that the resource
-can be captured when a function is declared (e.g.~when it is constructed on the server-side
-where database access is available), or when a function is called (e.g.~when a function created
-on server-side requires access to current time-zone, it can use the resource available on the
-client-side).
+As will be discussed in more detail in Section~\ref{sec:applications-flat-impl}, this means that a 
+resource can be captured when a function is declared (e.g.~when it is constructed on the 
+server-side where database access is available), or when a function is called (\eg~when a function 
+created on server-side requires access to current time-zone, it can use the resource available on
+ the client-side). In other words, resources in this example support both static (lexical) and
+ dynamic scoping.
 
@@
 \begin{equation*}
-\inference[(var)]
-  {x:\alpha \in \Gamma }
-  {\Gamma @ \emptyset \vdash x : \alpha }
-\quad
-\inference[(access)]
-  { \Gamma @ \sigma \vdash e : \ident{res}_\rho~\alpha }
-  {\Gamma @ \sigma_1 \cup \{\ident{access}(\rho)\} \vdash \kvd{access}~e : \alpha }
+\tyrule{var}
+  {x\!:\!\tau \in \Gamma }
+  {\coctx{\Gamma}{\cclrd{\emptyset}} \vdash x : \tau }
 \end{equation*} 
 \begin{equation*}
-\inference[(fun)]
-  {\Gamma, x:\alpha @ \sigma_1 \cup \sigma_2 \vdash e : \beta }
-  {\Gamma @ \sigma_1 \vdash \lambda x.e : \alpha \xrightarrow{\sigma_2} \beta}
-\quad  
-\inference[(app)]
-  {\Gamma \vdash e_1 : \alpha \xrightarrow{\sigma_1} \beta, \sigma_2 \\
-   \Gamma \vdash e_2 : \alpha, \sigma_3 }
-  {\Gamma \vdash e_1~e_2 : \beta, \sigma_1 \cup \sigma_2 \cup \sigma_3 }
+\tyrule{access}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \ident{res}_\rho~\tau }
+  {\coctx{\Gamma}{\cclrd{r} \cup \cclrd{ \{\ident{access}(\rho)\} }} \vdash \kvd{access}~e : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{(\Gamma, x\!:\!\tau_1)}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2 }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_1 \xrightarrow{\cclrd{r}} \tau_2, \cclrd{s} \\
+   \Gamma \vdash e_2 : \tau_1, \cclrd{t} }
+  {\Gamma \vdash e_1~e_2 : \tau_2, \cclrd{r}\cup\cclrd{s}\cup\cclrd{t}  }
 \end{equation*}
 
-\caption{Simple effect system}
+\caption{Simple coeffect system}
 \label{fig:path-coeff}
@@
 
-When using monads, we distinguish effect-free values $\alpha$ from programs, or 
-computations $\mtyp{}{\alpha}$. The \emph{monad} $\mtyp{}{}$ abstracts the \emph{notion of 
+When using monads, we distinguish effect-free values $\tau$ from programs, or 
+computations $\mtyp{}{\tau}$. The \emph{monad} $\mtyp{}{}$ abstracts the \emph{notion of 
 computation} and provides a way of constructing and composing effectful computations:
@@
 \end{compactitem}
-such that, for all $f:\alpha \rightarrow \mtyp{}{\beta}, g:\beta \rightarrow \mtyp{}{\gamma}$:
+such that, for all $f:\alpha \rightarrow \mtyp{}{\beta}$ and $g:\beta \rightarrow \mtyp{}{\gamma}$:
 \begin{align}
@@
 In this thesis, we generally follow the first approach -- this means that we work with an existing
-programming language (without needing to add additional constructs corresponding to the primitives
-of our semantics). To explain the difference in greater detail, the following two sections show a
-minimal example of both formal systems. We follow Moggi and start with language where judgements have
-the form $x:\alpha \vdash e : \beta$ with exactly one variable\footnote{This simplifies the examples
-as we do not need \emph{strong} monad, but that is an orthogonal issue to the distinction between
-language semantics and meta-language.}.
+programming language without needing to add additional constructs corresponding to the primitives
+of our semantics (the alternative is discussed in Section~\ref{sec:unified-meta}). To clarify the 
+difference, the following two sections show a minimal example of both formal systems. We follow 
+Moggi and start with language where judgements have the form $x\!:\!\alpha \vdash e : \beta$ with 
+exactly one variable\footnote{This simplifies the examples as we do not need \emph{strong} monad, 
+but that is an orthogonal issue to the distinction between language semantics and meta-language.}.
 
@@
 
-\paragraph{Language semantics} When using monads to provide semantics of a language, we do not
+\paragraph{Language semantics.} When using monads to provide semantics of a language, we do not
 need to extend the language in any way -- we assume that the language already contains the 
 effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
-of the form $x:\alpha \vdash e : \beta$ is interpreted as a morphism $\alpha \rightarrow \mtyp{}{\beta}$,
+of the form $x\!:\!\tau_1 \vdash e : \tau_2$ is interpreted as a morphism $\tau_1 \rightarrow \mtyp{}{\tau_2}$,
 meaning that any expression is interpreted as an effectful computation. The semantics of variable
-access ($x$) and the application of a primitive function $f$ is interpreted as follows:
+access and the application of a primitive function $f$ is interpreted as follows:
 %
@@
 \begin{array}{rcl}
-\sem{x:\alpha \vdash x : \alpha} &=& \ident{unit}_\mtyp{}{}\\
-\sem{x:\alpha \vdash f~e : \gamma} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
+ \sem{x\!:\!\tau_1 \vdash x : \tau_1} &=& \ident{unit}_\mtyp{}{}\\
+ \sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
 \end{array}
@@
 using $\ident{unit}_\mtyp{}{}$. In the second rule, we assume that $e$ is an expression using
-the variable $x$ and producing a value of type $\beta$ and that $f$ is a (primitive) function
-$\beta \rightarrow \mtyp{}{\gamma}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
-to a function $\mtyp{}{\beta} \rightarrow \mtyp{}{\gamma}$ which is compatible with the 
+the variable $x$ and producing a value of type $\tau_2$ and that $f$ is a (primitive) function
+$\tau_2 \rightarrow \mtyp{}{\tau_3}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
+to a function $\mtyp{}{\tau_2} \rightarrow \mtyp{}{\tau_3}$ which is compatible with the 
 interpretation of the expression $e$.
@@
 
-\paragraph{Meta-language interpretation} When designing meta-language based on monads, we need to
+\paragraph{Meta-language interpretation.} ~When designing meta-language based on monads, we need to
 extend the lambda calculus with additional type(s) and expressions that correspond to monadic
-primitives:
+primitives. Assuming $\alpha$ is a primitive type:
 %
 \begin{align*}
-\alpha, \beta, \gamma &:= \tau \sep \alpha \rightarrow \beta \sep \mtyp{}{\alpha} \\
+\tau &:= \alpha \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
    e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
@@
 %
-The types consist of primitive type ($\tau$), function type and a type constructor that 
+The types consist of the primitive type, function type and a type constructor that 
 represents monadic computations. This means that the expressions in the language can create both
-effect-free values, such as $\alpha$ and computations $\mtyp{}{\alpha}$. The additional expression
+effect-free values, such as $\tau$ and computations $\mtyp{}{\tau}$. The additional expression
 $\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no actual effects) from a
@@
 \begin{array}{rcl}
-\sem{x:\alpha \vdash x : \alpha} &=& \idf{}\\
-\sem{x:\alpha \vdash f~e : \beta} &=& f \circ \sem{e}\\
-\sem{x:\alpha \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\beta}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
-\sem{x:\alpha \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\beta}} &=& 
+\sem{x\!:\!\tau \vdash x : \tau} &=& \idf{}\\
+\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ \sem{e}\\
+\sem{x\!:\!\tau_1 \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\tau_2}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
+\sem{x\!:\!\tau_1 \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\tau_3}} &=& 
   \ident{bind}_\mtyp{}{}~\sem{e_2} \circ \sem{e_1}
@@
 general''}. Indeed, this is a valid and reasonable perspective. Yet, we follow the first style,
-precisely because it is \emph{less general} -- our aim is to develop concrete context-aware 
-programming languages (together with their type theory and semantics) rather than to build a 
-general framework for reasoning about languages with context-dependent properties.
+precisely because it is \emph{less general}. As discussed in Chapter~\ref{ch:intro}, our aim is 
+to develop concrete context-aware programming languages (together with their type theory and 
+semantics) rather than to build a general framework for reasoning about languages with 
+context-dependent properties.
 
@@
 \cite{monads-effects-marriage}. This requires extending the categorical structure. A monadic
-computation $\alpha \rightarrow \mtyp{}{\beta}$ means that the computation has \emph{some} 
-effects while the judgement $\Gamma \vdash e : \alpha, \sigma$ specifies \emph{what} effects
+computation $\tau_1 \rightarrow \mtyp{}{\tau_2}$ means that the computation has \emph{some} 
+effects while the judgement $x\!:\!\tau_1 \vdash e : \tau_2, \cclrd{r}$ specifies \emph{what} effects
 the computation has.
 
-To solve this mismatch, Wadler and Thiemann use a \emph{family} of monads $\mtyp{\sigma}{\alpha}$
+To solve this mismatch, Wadler and Thiemann use a \emph{family} of monads $\mtyp{\cclrd{r}}{\tau}$
 with an annotation that specifies the effects that may be performed by the computation. In their
-system, an effectful function $\alpha \xrightarrow{\sigma} \beta$ is modelled as a pure 
-function returning monadic computation $\alpha \rightarrow \mtyp{\sigma}{\beta}$. Similarly, the
-semantics of a judgement $x:\alpha \vdash e : \beta, \sigma$ can be given as a function 
-$\alpha \rightarrow \mtyp{\sigma}{\beta}$. 
-The precise nature of the family of monads has been later called \emph{indexed monads} (e.g.~by Tate
-\cite{effects-producer-semantics}) and further developed by Atkey \cite{monads-parameterised-notions} 
-in his work on \emph{parameterized monads}.
+system, an effectful function $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is modelled as a pure 
+function returning monadic computation $\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$. Similarly, the
+semantics of a judgement $x\!:\!\tau_1 \vdash e : \tau_2, \cclrd{r}$ can be given as a function 
+$\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$. 
+The precise nature of the family of monads has been later called \emph{indexed monads} by Tate
+\cite{effects-producer-semantics} and further developed by Atkey \cite{monads-parameterised-notions} 
+in his work on \emph{parameterized monads} and Katsumata \cite{monads-parametric}.
 
-\paragraph{Thesis perspective}
+\paragraph{Thesis perspective.}
 The key takeaway for this thesis from the outlined line of research is that, if we want to develop a 
@@
 notions of \emph{context-dependent computations} \cite{comonads-notions}. The computations discussed 
-in the latter work include streams, arrays and containers -- this is a more diverse set of examples, 
+in the latter work include streams, arrays and containers. This is a more diverse set of examples, 
 but they all mostly represent forms of collections. Ahman et al. \cite{comonads-containers} discuss
-the relation between comonads and \emph{containers} in more details.
+the relation between comonads and \emph{containers} \cite{types-containers} in more details.
 
@@
 \cite{comonads-computational} use \emph{computational} comonads for intensional semantics\footnote{The
-structure of computational comonad has been also used by the author of this thesis to abstract
+structure of a computational comonad has been also used by the author of this thesis to abstract
 evaluation order of monadic computations \cite{comonads-malias}.}. In functional programming,
@@
 Moggi). Before looking at the details, we use the following definition of comonad:
-
-\todo{Maybe remove this from here, we repeate it in section 4}
 
@@
 
-\todo{The following text is duplicated in chapter 4}
-
 \noindent
-The definition is similar to monad with ``reversed arrows''. Intuitively, the $\ident{counit}$ 
+The definition is similar to a monad with ``reversed arrows''. Intuitively, the $\ident{counit}$ 
 operation extracts a value $\alpha$ from a value that carries additional context $\ctyp{}{\alpha}$.
@@
 
-\paragraph{Language semantics}
+\paragraph{Language semantics.}
 To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
 $\ctyp{}{\alpha} = \mu \gamma.\alpha + (\alpha \times \gamma)$. A value of the type is either
-the last element $\alpha$ or an element followed by another non-empty list $\alpha \times \gamma$.
-Note that the list must be non-empty -- otherwise \ident{counit} would not be a complete 
-function (it would be undefined on empty list). In the following, we write $(l_1, \ldots, l_n)$
-for a list of $n$ elements:
+the last element $\alpha$ or an element followed by another non-empty list $\alpha \times \gamma$
+(consisting of the head $\alpha$ and the tail $\gamma$). Note that the list must be non-empty, 
+otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In 
+the following, we write $(l_1, \ldots, l_n)$ for a list of $n$ elements:
 %
@@
 The \ident{cobind} operation creates a new list by applying the context-dependent function $f$
-to the entire list, to the suffix of the list, to the suffix of the suffix and so on.
+to the entire list, to the suffix of the list, to the suffix of the suffix and so on. Interestingly,
+it preserves the \emph{shape} of the list as it turns a list of $n$ elements into another list
+of $n$ elements.
 
@@
 value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$ 
-has a type $\alpha$, the expression $\kvd{prev}~e$ has also type $\alpha$. The fact that
+has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
 the language models data-flow and values are lists (of past values) is a matter of semantics,
@@
 \begin{array}{rcl}
-\sem{x:\alpha \vdash x : \alpha} &=& \ident{counit}_\ctyp{}{}\\
-\sem{x:\alpha \vdash f~e : \gamma} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
-\sem{x:\alpha \vdash \kvd{prev}~e : \gamma} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
+\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
+\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
+\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
 \end{array}
@@
 
-For example, the judgement $x:\alpha \vdash \kvd{prev}~(\kvd{prev}~x) : \alpha$ represents an 
+For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an 
 expression that expects context with variable $x$ and returns a stream of values before the 
@@
 additional structure called \emph{symmetric (semi)monoidal} comonads. This structure is 
-responsible for the splitting of context-requirements in lambda abstraction. We return to this
-topic when discussing flat coeffect system later in the thesis.
+responsible for the splitting of context-requirements in lambda abstraction. Note that this is 
+what happens in the unusual (\emph{abs}) rule in Figure~\ref{fig:path-coeff}, which distinguishes
+coeffect systems from effect systems.
+
+We return to this topic when discussing lambda abstraction in Section~\ref{sec:applications-structure-lam} 
+and semantics of flat coeffect systems in Section~\ref{sec:flat-semantics}.
 
@@
 
-\begin{figure}
-\begin{equation*}
-\inference[(eval)]
-  {\Gamma \vdash e : \ctyp{\emptyset}{\alpha}}
-  {\Gamma \vdash !e : \alpha}
-\quad
-\inference[(letbox)]
-  {\Gamma \vdash e_1 : \ctyp{\Phi, \Psi}{\alpha} & \Gamma, x : \ctyp{\Phi}{\alpha} \vdash e_2 : \beta }
-  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{\Psi}{\beta}}
-\end{equation*}
-
-\caption{Typing for a comonadic language with contextual staged computations}
-\label{fig:modal-meta}
-\end{figure}
-
-%---------------------------------------------------------------------------------------------------
-
-\paragraph{Meta-language interpretation} To briefly demonstrate the approach that employs comonads
+\paragraph{Meta-language interpretation.} ~To briefly demonstrate the app\-roach that employs comonads
 as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
@@
 
-In languages inspired by modal logics, types can have the form $\square \alpha$. In the work of
+In languages inspired by modal logics, types can have the form $\square \tau$. In the work of
 Pfenning and Davies, this means a term that is provable with no assumptions. In distributed 
 programming language ML5, Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus} use the 
-$\square \alpha$ type to mean \emph{mobile code}, that is code that can be evaluated at any node of a 
-distributed system (the evaluation corresponds to the axiom $\square \alpha \rightarrow \alpha$). 
+$\square \tau$ type to mean \emph{mobile code}, that is code that can be evaluated at any node of a 
+distributed system (the evaluation corresponds to the axiom $\square \tau \rightarrow \tau$). 
 Finally, Davies and Pfenning \cite{logic-modal-staged} consider staged computations and interpret 
-$\square \alpha$ as a type of (unevaluated) expressions of type $\alpha$.
+$\square \tau$ as a type of unevaluated expressions of type $\tau$ (with no free variables).
 
-In Contextual Modal Type Theory, the modality $\square$ is further annotated. To keep the syntax
-consistent with earlier examples, we use $\ctyp{\Psi}{\alpha}$ for a type $\square \alpha$ with an
-annotation $\Psi$. The type is a comonadic counterpart to the \emph{indexed monads} used by Wadler
-and Thiemann when linking monads and effect systems and, indeed, it gives rise to a language that
-tracks context-dependence of computations in a type system.
+In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variales
+of the (unevaluated) expression. To keep the syntax consistent with earlier examples, we use 
+$\ctyp{\cclrd{\Psi}}{\tau}$ for a type $\square \tau$ with an annotation $\cclrd{\Psi}$. The type 
+is a comonadic counterpart to the \emph{indexed monads} used by Wadler and Thiemann when linking 
+monads and effect systems and, indeed, it gives rise to a language that tracks context-dependence 
+of computations in a type system.
 
-In staged computation, the type $\ctyp{\Psi}{\alpha}$ represents an expression 
+In staged computation, the type $\ctyp{\cclrd{\Psi}}{\alpha}$ represents an expression 
 that requires the context $\Psi$ (i.e.~the expression is an open term that requires variables $\Psi$).
@@
 \begin{itemize}
-\item (\emph{eval}) corresponds to $\ident{counit} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha$. It means
-  that we can evaluate a closed (unevaluated) term and obtain a value. Note that the rule requires
-  a specific context annotation. It is not possible to evaluate an open term.
+\item (\emph{eval}) corresponds to $\ident{counit} : \ctyp{\cclrd{\emptyset}}{\alpha} \rightarrow \alpha$. 
+  It means that we can evaluate a closed (unevaluated) term and obtain a value. Note that the 
+  rule requires a specific context annotation (empty set of free variables). 
+  It is not possible to evaluate an open term.
 
-\item (\emph{letbox}) corresponds to $\ident{cobind} : (\ctyp{\Psi}{\alpha} \rightarrow \beta) 
-  \rightarrow \ctyp{\Psi, \Phi}{\alpha} \rightarrow \ctyp{\Phi}{\beta}$. It means that given
-  a term which requires variable context $\Psi, \Phi$ (expression $e_1$) and a function that turns 
-  a term needing $\Psi$ into an evaluated value (expression $e_2$), we can construct a term 
-  that requires just $\Phi$.
+\item (\emph{letbox}) corresponds to $\ident{cobind} : (\ctyp{\cclrd{\Psi}}{\alpha} \rightarrow \beta) 
+  \rightarrow \ctyp{\cclrd{\Psi}, \clrd{\Phi}}{\alpha} \rightarrow \ctyp{\cclrd{\Phi}}{\beta}$. 
+  It means that given a term which requires variable context $\cclrd{\Psi}, \cclrd{\Phi}$ 
+  (expression $e_1$) and a function that turns a term needing $\cclrd{\Psi}$ into an evaluated 
+  value (expression $e_2$), we can construct a term that requires just $\cclrd{\Phi}$.
 \end{itemize}
@@
 from ordinary comonads where \ident{counit} needs to be defined for all values. Here, the indexed
-\ident{counit} operation needs to be defined only on values annotated with $\emptyset$.
+\ident{counit} operation needs to be defined \emph{only} on values annotated with $\emptyset$.
 
-The annotated \ident{cobind} operation that corresponds to (\emph{letbox}) is in details 
-introduced in Chapter~X. An interesting aspect is that it propagates the context-requirements
-``backwards''. The input expression (second parameter) requires a combination of contexts that
-are required by the two components -- those required by the input of the function (first
-argument) and those required by the resulting expression (result). This is another key aspect
-that distinguishes coeffects from effect systems.
+The annotated \ident{cobind} operation that corresponds to (\emph{letbox}). We discuss this 
+in Section~\ref{sec:unified-meta}. An interesting aspect is that it propagates the 
+context-requirements ``backwards''. The input expression (second parameter) requires a 
+combination of contexts that are required by the two components -- those required by the 
+input of the function (first argument) and those required by the resulting expression (result). 
+This is another key aspect that distinguishes coeffects from effect systems.
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\inference[(eval)]
+  {\Gamma \vdash e : \ctyp{\cclrd{\emptyset}}{\tau}}
+  {\Gamma \vdash !e : \tau}
+\end{equation*}
+\begin{equation*}
+\inference[(letbox)]
+  { \Gamma \vdash e_1 : \ctyp{\cclrd{\Phi}, \cclrd{\Psi}}{\tau_1} & 
+    \Gamma, x\!:\!\ctyp{\cclrd{\Phi}}{\tau_1} \vdash e_2 : \tau_2 }
+  { \Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{\Psi}}{\tau_2}}
+\end{equation*}
+
+\caption{Typing for a comonadic language with contextual staged computations}
+\label{fig:modal-meta}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
 
@@
 our work. Most importantly, Contextual Modal Type Theory (CMTT) uses indexed $\square$ modality
-which seems to correspond to indexed comonads (in a similar way in which effect systems 
-correspond to indexed monads). The relation between CMTT and comonads has been suggested by
+which corresponds to indexed comonads (in a similar way in which effect systems correspond to 
+indexed monads). The relation between CMTT and comonads has been suggested by
 Gabbay et al. \cite{logic-cmtt-semantics}, but the meta-language employed by CMTT does not 
-directly correspond to comonadic operations. For example, our \ident{let box} typing rule from
+directly correspond to comonadic operations. For example, our (\emph{letbox}) typing rule from
 Figure~\ref{fig:modal-meta} is not a primitive of CMTT and would correspond to 
-$\ident{box}(\Psi, \ident{letbox}(e_1, x, e_2))$. Nevertheless, the indexing in CMTT provides a useful
-hint for adding indexing to the work of Uustalu and Vene.
+$\ident{box}(\cclrd{\Psi}, \ident{letbox}(e_1, x, e_2))$. Nevertheless, the indexing in 
+CMTT provides a useful hint for adding indexing to the work of Uustalu and Vene.
 
@@
 \inference[(exchange)]
-  {\Gamma, x:\alpha, y:\beta \vdash e : \gamma}
-  {\Gamma, y:\beta, x:\alpha \vdash e : \gamma}
+  {\Gamma, x\!:\!\alpha, y\!:\!\tau_2 \vdash e : \gamma}
+  {\Gamma, y\!:\!\tau_2, x\!:\!\alpha \vdash e : \gamma}
 \quad
@@
   {\Gamma, \Delta \vdash e : \gamma}
-  {\Gamma, x:\alpha, \Delta \vdash e : \gamma}
+  {\Gamma, x\!:\!\alpha, \Delta \vdash e : \gamma}
 \end{equation*}
@@
 \inference[(contraction)]
-  {\Gamma, x:\alpha, y:\alpha, \Delta \vdash e : \gamma}
-  {\Gamma, x:\alpha, \Delta \vdash \subst{e}{y}{x} : \gamma}
+  {\Gamma, x\!:\!\alpha, y\!:\!\alpha, \Delta \vdash e : \gamma}
+  {\Gamma, x\!:\!\alpha, \Delta \vdash \subst{e}{y}{x} : \gamma}
 \end{equation*}
@@
 \begin{equation*}
-\Gamma, \Delta, \Sigma := x:\alpha \sep I \sep \Gamma, \Gamma \sep 1 \sep \Gamma; \Gamma
+\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep I \sep \Gamma, \Gamma \sep 1 \sep \Gamma; \Gamma
 \end{equation*}
@@
 body of a function and the context where a function is declared. The system introduces two 
-distinct function types $\alpha \rightarrow \beta$ and $\alpha~ \textendash\!\!\!\ast \beta$
+distinct function types $\alpha \rightarrow \tau_2$ and $\alpha~ \textendash\!\!\!\ast \tau_2$
 (corresponding to ``;'' and ``,'' respectively). The key property is that only the first kind
@@
 \begin{equation*}
-\Gamma, \Delta, \Sigma := x:\alpha \sep 1 \sep \Gamma, \Gamma \sep \Gamma @ \sigma
+\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep 1 \sep \Gamma, \Gamma \sep \Gamma @ \sigma
 \end{equation*}
@@
 We can also annotate individual components. For example, a context containing variables $x,y,z$
-where only $x$ is used could be written as $(x:\alpha \,@\, \ident{used}), ((y:\alpha, z:\alpha) 
+where only $x$ is used could be written as $(x\!:\!\alpha \,@\, \ident{used}), ((y\!:\!\alpha, z\!:\!\alpha) 
 \,@\, \ident{unused})$.

### 2014-12-01 text/Chapter06.tex
Proofreading chapters 1 and 2, reorder chapter 6

@@
 believe that coeffects should be integrated in main-stream programming languages. In the second
-part of this chapter, we outline one possible approach for practical implementations of coeffects 
-(Section~\ref{sec:unified-impl}). Finally, we discuss an alternative approach to defining coeffect 
+part of this chapter, we discuss an alternative approach to defining coeffect 
 systems which highlights the relationship between our work and related work arising from modal logics 
-(Section~\ref{sec:unified-meta}).
+(Section~\ref{sec:unified-meta}). Finally, we outline one possible approach for practical
+implementations of coeffects (Section~\ref{sec:unified-impl}).
 
@@
 annotations) of the object that wrap the free-variable context.
+
+% =================================================================================================
+%                                                                                             
+%    #   #          #                    ##                                                    
+%    #   #          #                     #                                                    
+%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
+%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
+%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
+%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
+%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
+%                                                            #   #                #   #        
+%                                                             ###                  ###         
+% =================================================================================================
+
+\section{Coeffect meta-language}
+\label{sec:unified-meta}
+
+In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
+language semantics introduced by Moggi \cite{monad-notions}. The first approach is to use
+monads in the \emph{semantics} of an effectful language. The second approach is to extend the
+language with (additional) monadic constructs that can then be used for writing effectful
+monads explicitly.
+
+In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
+the term language is that of simply-typed $\lambda$-calculus, and we used (flat or structural)
+indexed comonads to give the semantics for the language and to derive type system for it.
+
+In this section, we briefly discuss the alternative approach. That is, we embed indexed
+comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
+context (semantically, this corresponds to an indexed comonad) and we add language constructs
+that correspond to the operations of indexed comonads.
+
+This section provides a brief sketch of coeffect meta-language to highlight the relationship
+between coeffects and important related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
+Developing the system further is an interesting future research direction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffects and contextual modal type theory}
+
+As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
+to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
+meta-languages that include $\square$ as a type constructor
+\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
+Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
+From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
+coeffect calculus.
+
+\paragraph{Context in CMTT and coeffects.}
+Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
+comonads (the $\square$ modality) into the meta-language, there are two important differences.
+
+Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
+requirements are formed by a coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, \emph{etc.}
+
+Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
+the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+constructor refers to a value of type $\tau$ \emph{together} with additional context specified
+by $\cclrd{r}$ (\eg~list of past values or additional resources). In contrast in 
+CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
+original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
+context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
+two operations of a comonad:
+\begin{equation*}
+\begin{array}{rcl}
+ \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
+ \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
+    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
+\end{array}
+\end{equation*}
+Both readings are possible, but they give quite different meanings to the operations:
+
+\begin{itemize}
+\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
+  and does not require any additional context; the \ident{cobind} operation requires
+  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
+  evaluate the function, ending with a value $\beta$ together with remaining context
+  $\cclrd{s}$.
+\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
+  that requires no additional context to obtain a $\alpha$ value; given a function that
+  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
+  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
+  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
+  context requirements is eliminated by the function).
+\end{itemize}
+
+\noindent
+Although the different reading does not affect formal properties of the systems, it is important
+to understand the difference when discussing the two systems.
+
+The sketch of a coeffect meta-language in the following section attempts to bridge the gap
+between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
+it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
+sets of variables. Future work on the coeffect meta-language would thus be an interesting 
+development for both coeffect systems and CMTT.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
+For simplicity, this section only discusses the flat variant. The syntax of types and terms
+of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four expression forms:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
+ e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
+  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
+  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
+  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
+\end{array}
+\end{equation*}
+%
+The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
+operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
+\kvd{split} and \kvd{merge} that embed the corresponding operations.
+
+\paragraph{Types for coeffect meta-language.}
+The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
+shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
+rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
+
+The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
+a value from a comonadic context and corresponds to variable access in coeffect calculi. The
+\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
+and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
+The result is a computation that produces a $\tau_2$ value with the remaining context specified
+by $\cclrd{s}$. Note that the expression $e_2$ and $e_1$ corresponds to the first and second
+arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
+CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
+although it differs because of our generalization of comonads where \ident{bind} composes
+coeffect annotations rather than requiring the same annotation everywhere.}.
+
+The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
+transformation on one or two values in a context and then add the new value as a fresh variable
+to the variable context. For simplicity, we omit sub-coeffecting, but it could be easily added 
+following the pattern used elsewhere.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Typing rules for the simply typed $\lambda$-calculus}
+
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\Gamma \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
+   \Gamma \vdash e_2 : \tau_2 }
+  {\Gamma \vdash e_1~e_2 : \tau_1 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x:\tau_1 \vdash e : \tau_2}
+  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
+
+\begin{equation*}
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
+\end{equation*}
+\begin{equation*}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash !e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{split}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
+   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{merge}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
+   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
+   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
+  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
+\end{equation*}
+
+\caption{Type system for the (flat) coeffect meta-language}
+\label{fig:conclusions-cml}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding flat coeffect calculus}
+
+The \emph{meta-language} approach of embedding comonads in a language is more general than the
+\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
+a type system for context-aware programming languages.
+
+However, it is worth noting that 
+(flat) coeffect calculus can be embedded in the meta-language described above. This may be 
+desirable, \eg~when using the meta-language for reasoning about context-aware computations. 
+We briefly consider the embedding as it illuminates the relationship between coeffects and
+CMTT (although it is not possible to embed coeffect calculi in CMTT because of the more general
+annotations structure).
+
+Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in a flat coeffect calculus,
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
+meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
+to represent the entire variable context of the source language. The translation is defined in
+Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
+Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+the operations of a monad.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{l}
+ \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
+   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
+   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
+      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
+   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
+\end{array}
+\end{equation*}
+\caption{Embedding flat coeffect calculus in coeffect meta-language }
+\label{fig:conclusions-embed}
+\end{figure}
 
@@
 
-% =================================================================================================
-%                                                                                             
-%    #   #          #                    ##                                                    
-%    #   #          #                     #                                                    
-%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
-%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
-%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
-%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
-%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
-%                                                            #   #                #   #        
-%                                                             ###                  ###         
-% =================================================================================================
-
-\section{Coeffect meta-language}
-\label{sec:unified-meta}
-
-In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
-language semantics introduced by Moggi \cite{monad-notions}. The first approach is to use
-monads in the \emph{semantics} of an effectful language. The second approach is to extend the
-language with (additional) monadic constructs that can then be used for writing effectful
-monads explicitly.
-
-In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
-the term language is that of simply-typed $\lambda$-calculus, and we used (flat or structural)
-indexed comonads to give the semantics for the language and to derive type system for it.
-
-In this section, we briefly discuss the alternative approach. That is, we embed indexed
-comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
-type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
-context (semantically, this corresponds to an indexed comonad) and we add language constructs
-that correspond to the operations of indexed comonads.
-
-This section provides a brief sketch of coeffect meta-language to highlight the relationship
-between coeffects and important related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
-Developing the system further is an interesting future research direction.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffects and contextual modal type theory}
-
-As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
-to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
-meta-languages that include $\square$ as a type constructor
-\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
-Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
-From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
-coeffect calculus.
-
-\paragraph{Context in CMTT and coeffects.}
-Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
-comonads (the $\square$ modality) into the meta-language, there are two important differences.
-
-Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
-requirements are formed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}
-
-Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
-the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
-constructor refers to a value of type $\tau$ \emph{together} with additional context specified
-by $\cclrd{r}$ (\eg~list of past values or additional resources). In contrast in 
-CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
-original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
-context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
-two operations of a comonad:
-\begin{equation*}
-\begin{array}{rcl}
- \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
- \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
-    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
-\end{array}
-\end{equation*}
-Both readings are possible, but they give quite different meanings to the operations:
-
-\begin{itemize}
-\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
-  and does not require any additional context; the \ident{cobind} operation requires
-  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
-  evaluate the function, ending with a value $\beta$ together with remaining context
-  $\cclrd{s}$.
-\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
-  that requires no additional context to obtain a $\alpha$ value; given a function that
-  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
-  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
-  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
-  context requirements is eliminated by the function).
-\end{itemize}
-
-\noindent
-Although the different reading does not affect formal properties of the systems, it is important
-to understand the difference when discussing the two systems.
-
-The sketch of a coeffect meta-language in the following section attempts to bridge the gap
-between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
-it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
-sets of variables. Future work on the coeffect meta-language would thus be an interesting 
-development for both coeffect systems and CMTT.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect meta-language}
-
-The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
-For simplicity, this section only discusses the flat variant. The syntax of types and terms
-of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four expression forms:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
- e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
-  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
-  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
-  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
-\end{array}
-\end{equation*}
-%
-The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
-operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
-\kvd{split} and \kvd{merge} that embed the corresponding operations.
-
-\paragraph{Types for coeffect meta-language.}
-The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
-shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
-rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
-
-The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
-a value from a comonadic context and corresponds to variable access in coeffect calculi. The
-\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
-and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
-The result is a computation that produces a $\tau_2$ value with the remaining context specified
-by $\cclrd{s}$. Note that the expression $e_2$ and $e_1$ corresponds to the first and second
-arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
-CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
-although it differs because of our generalization of comonads where \ident{bind} composes
-coeffect annotations rather than requiring the same annotation everywhere.}.
-
-The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
-transformation on one or two values in a context and then add the new value as a fresh variable
-to the variable context. For simplicity, we omit sub-coeffecting, but it could be easily added 
-following the pattern used elsewhere.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-
-{\small a.) Typing rules for the simply typed $\lambda$-calculus}
-
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\Gamma \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
-   \Gamma \vdash e_2 : \tau_2 }
-  {\Gamma \vdash e_1~e_2 : \tau_1 }
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\Gamma, x:\tau_1 \vdash e : \tau_2}
-  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
-\end{equation*}
-
-\vspace{1em}
-{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
-
-\begin{equation*}
-\tyrule{cobind}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
-\end{equation*}
-\begin{equation*}
-\tyrule{counit}
-  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
-  {\Gamma \vdash !e : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{split}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
-   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{merge}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
-   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
-   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
-  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
-\end{equation*}
-
-\caption{Type system for the (flat) coeffect meta-language}
-\label{fig:conclusions-cml}
-\end{figure}
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Embedding flat coeffect calculus}
-
-The \emph{meta-language} approach of embedding comonads in a language is more general than the
-\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
-a type system for context-aware programming languages.
-
-However, it is worth noting that 
-(flat) coeffect calculus can be embedded in the meta-language described above. This may be 
-desirable, \eg~when using the meta-language for reasoning about context-aware computations. 
-We briefly consider the embedding as it illuminates the relationship between coeffects and
-CMTT (although it is not possible to embed coeffect calculi in CMTT because of the more general
-annotations structure).
-
-Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in a flat coeffect calculus,
-we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
-meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
-to represent the entire variable context of the source language. The translation is defined in
-Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
-Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
-the operations of a monad.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\begin{equation*}
-\begin{array}{l}
- \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
-   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
-   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
-      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
-   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
-\end{array}
-\end{equation*}
-\caption{Embedding flat coeffect calculus in coeffect meta-language }
-\label{fig:conclusions-embed}
-\end{figure}
-
 % ==================================================================================================

### 2014-12-04 text/Chapter02.tex
Proofreading (Chapter 2)

@@
 monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in 
-Haskell, list monad ($\mtyp{}{\alpha} = \mu \gamma.1 + (\alpha \times \gamma)$) and other.
+Haskell, list monad ($\mtyp{}{\tau} = \mu \gamma.1 + (\tau \times \gamma)$) and other.
 In programming language semantics, monads can be used in two distinct ways.
@@
 difference, the following two sections show a minimal example of both formal systems. We follow 
-Moggi and start with language where judgements have the form $x\!:\!\alpha \vdash e : \beta$ with 
+Moggi and start with language where judgements have the form $x\!:\!\tau_1 \vdash e : \tau_2$ with 
 exactly one variable\footnote{This simplifies the examples as we do not need \emph{strong} monad, 
@@
 To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
-$\ctyp{}{\alpha} = \mu \gamma.\alpha + (\alpha \times \gamma)$. A value of the type is either
-the last element $\alpha$ or an element followed by another non-empty list $\alpha \times \gamma$
-(consisting of the head $\alpha$ and the tail $\gamma$). Note that the list must be non-empty, 
+$\ctyp{}{\tau} = \mu \gamma.\tau + (\tau \times \gamma)$. A value of the type is either
+the last element $\tau$ or an element followed by another non-empty list $\tau \times \gamma$
+(consisting of the head $\tau$ and the tail $\gamma$). Note that the list must be non-empty, 
 otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In 
@@
 as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
-Nanevski et al. We do not attempt to provide precise overview of their work. The main purpose 
-of our discussion is to provide a different intuition behind comonads, and to give an example
-of a language that includes comonad as a type constructor, together with language primitives
-corresponding to comonadic operations\footnote{In fact, Pfenning and Davies \cite{logic-modal-reconstruction,logic-cmtt}
-never mention comonads explicitly. This is done in later work by Gabbay et al. \cite{logic-cmtt-semantics}, 
-but the connection between the language and comonads is not as direct as in case of monadic or
-comonadic semantics covered in the last few pages.}.
+Nanevski. We do not attempt to provide a precise overview of their work. The main purpose 
+of the following discussion is to provide a different intuition behind comonads, and to present an 
+example of a language that includes comonad as a type constructor, together with language primitives
+corresponding to comonadic operations\footnote{In fact, Pfenning, Davies and Nanevski 
+\cite{logic-modal-reconstruction,logic-cmtt} never mention comonads explicitly. This is done in later 
+work by Gabbay et al. \cite{logic-cmtt-semantics},  but the connection between the language and comonads
+is not as direct as in case of monadic or comonadic semantics covered in the previous section.}.
 
 In languages inspired by modal logics, types can have the form $\square \tau$. In the work of
-Pfenning and Davies, this means a term that is provable with no assumptions. In distributed 
-programming language ML5, Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus} use the 
-$\square \tau$ type to mean \emph{mobile code}, that is code that can be evaluated at any node of a 
+Pfenning and Davies, this is the type of a term that is provable with no assumptions. In distributed 
+programming language ML5 by Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus}, the 
+$\square \tau$ type means \emph{mobile code}, that is code that can be evaluated at any node of a 
 distributed system (the evaluation corresponds to the axiom $\square \tau \rightarrow \tau$). 
@@
 In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variales
-of the (unevaluated) expression. To keep the syntax consistent with earlier examples, we use 
-$\ctyp{\cclrd{\Psi}}{\tau}$ for a type $\square \tau$ with an annotation $\cclrd{\Psi}$. The type 
-is a comonadic counterpart to the \emph{indexed monads} used by Wadler and Thiemann when linking 
-monads and effect systems and, indeed, it gives rise to a language that tracks context-dependence 
-of computations in a type system.
+of the (unevaluated) expression. We write $\square^{\cclrd{\Psi}} \tau$ for a type of expressions
+that requires a context $\Psi$. The type is a comonadic counterpart to \emph{indexed monads} used by 
+Wadler and Thiemann when linking monads and effect systems and, indeed, it gives rise to a language 
+that tracks context-dependence of computations in a type system.
 
-In staged computation, the type $\ctyp{\cclrd{\Psi}}{\alpha}$ represents an expression 
+In staged computation, the type $\ctyp{\cclrd{\Psi}}{\tau}$ represents an expression 
 that requires the context $\Psi$ (i.e.~the expression is an open term that requires variables $\Psi$).
@@
 \item (\emph{eval}) corresponds to $\ident{counit} : \ctyp{\cclrd{\emptyset}}{\alpha} \rightarrow \alpha$. 
-  It means that we can evaluate a closed (unevaluated) term and obtain a value. Note that the 
+  It indicates that we can evaluate a closed (unevaluated) term and obtain a value. Interestingly, the 
   rule requires a specific context annotation (empty set of free variables). 
@@
 \item (\emph{letbox}) corresponds to $\ident{cobind} : (\ctyp{\cclrd{\Psi}}{\alpha} \rightarrow \beta) 
-  \rightarrow \ctyp{\cclrd{\Psi}, \clrd{\Phi}}{\alpha} \rightarrow \ctyp{\cclrd{\Phi}}{\beta}$. 
-  It means that given a term which requires variable context $\cclrd{\Psi}, \cclrd{\Phi}$ 
+  \rightarrow \ctyp{\cclrd{\Psi}, \cclrd{\Phi}}{\alpha} \rightarrow \ctyp{\cclrd{\Phi}}{\beta}$. 
+  Given a term which requires variable context $\cclrd{\Psi}, \cclrd{\Phi}$ 
   (expression $e_1$) and a function that turns a term needing $\cclrd{\Psi}$ into an evaluated 
@@
 
-The annotated \ident{cobind} operation that corresponds to (\emph{letbox}). We discuss this 
-in Section~\ref{sec:unified-meta}. An interesting aspect is that it propagates the 
-context-requirements ``backwards''. The input expression (second parameter) requires a 
-combination of contexts that are required by the two components -- those required by the 
+The annotated \ident{cobind} operation that corresponds to (\emph{letbox}). An interesting aspect 
+is that it propagates the context-requirements ``backwards''. The input expression (second parameter) 
+requires a combination of contexts that are required by the two components -- those required by the 
 input of the function (first argument) and those required by the resulting expression (result). 
-This is another key aspect that distinguishes coeffects from effect systems.
+This is another key aspect that distinguishes coeffects from effect systems. We return back to 
+the meta-language approach of embedding comonads in Section~\ref{sec:unified-meta}.
 
@@
 \inference[(eval)]
-  {\Gamma \vdash e : \ctyp{\cclrd{\emptyset}}{\tau}}
+  {\Gamma \vdash e : \square^{\cclrd{\emptyset}}{\tau}}
   {\Gamma \vdash !e : \tau}
@@
 \inference[(letbox)]
-  { \Gamma \vdash e_1 : \ctyp{\cclrd{\Phi}, \cclrd{\Psi}}{\tau_1} & 
-    \Gamma, x\!:\!\ctyp{\cclrd{\Phi}}{\tau_1} \vdash e_2 : \tau_2 }
-  { \Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{\Psi}}{\tau_2}}
+  { \Gamma \vdash e_1 : \square^{\cclrd{\Phi}, \cclrd{\Psi}}{\tau_1} & 
+    \Gamma, x\!:\!\square^{\cclrd{\Phi}}{\tau_1} \vdash e_2 : \tau_2 }
+  { \Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \square^{\cclrd{\Psi}}{\tau_2}}
 \end{equation*}
@@
 
-\paragraph{Thesis perspective}
+\paragraph{Thesis perspective.}
 As mentioned earlier, we are interested in designing context-dependent languages and so we
-use comonads as \emph{language semantics}. Uustalu and Vene present a semantics of 
+use comonads for \emph{language semantics}. Uustalu and Vene present a semantics of 
 context-dependent computations in terms of comonads. We provide the rest of the story known 
-from the marriage of monads and effects. We develop coeffect calculus with a type system that 
+from the marriage of monads and effects. We develop coeffect calculus with an type system that 
 tracks the context requirements more precisely (by annotating the types) and we add indexing 
-to comonads and link the two by giving a formal semantics.
+to comonads and link the two by giving a formal semantics. The indexing allows us to capture 
+applications that do not fit into the model provided by plain comonads.
 
-The \emph{meta-language} approach of Pfenning, Davies and Nanevski et al. is closely related to
+The \emph{meta-language} approach of Pfenning, Davies and Nanevski is closely related to
 our work. Most importantly, Contextual Modal Type Theory (CMTT) uses indexed $\square$ modality
@@
 
-\section{Through sub-structural and bunched logics}
+\section{Through substructural and bunched logics}
 \label{sec:path-logic}
@@
 contextual information (set of available resources) with the variable context of the typing 
-judgement: $\Gamma @ \sigma \vdash e : \alpha$. In other words, our work focuses on ``what is
-happening on the left hand side of $\vdash$''.
+judgement: $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$. In other words, our work focuses on 
+``what is happening on the left hand side of $\vdash$''.
 
@@
 In this section, we look at type systems that reconsider $\Gamma$ in a number of ways. 
-First of all, sub-structural type systems \cite{substruct-attpl-intro} restrict the use of variables
+First of all, substructural type systems \cite{substruct-attpl-intro} restrict the use of variables
 in the language. Most famously linear type systems introduced by Wadler \cite{substruct-linear-change} 
-can guarantee that variable is used exactly once. This has interesting implications for memory
+can guarantee that a variable is used exactly once. This has interesting implications for memory
 management and I/O.
@@
 formed by multiple different constructors (e.g.~one that allows sharing and one that does not). 
-Most importantly, bunched typing has contributed to the development of separation logic
+Most famously, bunched typing has contributed to the development of separation logic
 \cite{substruct-separation-logic} (starting a fruitful line of research in software verification), 
@@
 
-\paragraph{Sub-structural type systems}
+\subsection{Substructural type systems.}
 
 Traditionally, $\Gamma$ is viewed as a set of assumptions and typing rules admit (or explicitly
-include) three operations that manipulate the variable contexts which are shown in 
-Figure~\ref{fig:substructural-rules}. The (\emph{exchange}) rule allows us to reorder variables
-(which is implicit, when assumptions are treated as set); (\emph{weakening}) makes it possible
+include) three transformations that manipulate the variable contexts which are shown in 
+Figure~\ref{fig:substructural-rules}. The (\emph{exchange}) rule allows reordering of variables
+(which is implicit when assumptions are treated as set); (\emph{weakening}) makes it possible
 to discard an assumption -- this has the implication that a variable may be declared but never
 used. Finally, (\emph{contraction}) makes it possible to use a single variable multiple times
-(by joining multiple variables into a single one using substitution).
+(in the rule, this is done explicitly by joining multiple variables into a single one using 
+substitution).
+
+In substructural type systems, the assumptions are typically treated as a list. As a result,
+they have to be manipulated explicitly. Different systems allow different subsets of the rules.
+For example, \emph{affine} systems allows exchange and weakening, leading to a system where 
+variable may be used at most once; in \emph{linear} systems, only exchange is permitted and so 
+every variable has to be used exactly once.
+
+When tracking context-dependent properties associated with individual variables, we need to 
+be more explicit in how variables are used. Substructural type systems provide a way to do this.
+Even if we allow all three operations, we can use a variation on the three rules (exchange, 
+weakening and contraction) to track which variables are used and how (and to track additional 
+contextual information about variables).
 
@@
 \begin{equation*}
-\inference[(exchange)]
-  {\Gamma, x\!:\!\alpha, y\!:\!\tau_2 \vdash e : \gamma}
-  {\Gamma, y\!:\!\tau_2, x\!:\!\alpha \vdash e : \gamma}
-\quad
-\inference[(weakening)]
-  {\Gamma, \Delta \vdash e : \gamma}
-  {\Gamma, x\!:\!\alpha, \Delta \vdash e : \gamma}
+\tyrule{exchange}
+  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_2 \vdash e : \gamma}
+  {\Gamma, y\!:\!\tau_2, x\!:\!\tau_1 \vdash e : \gamma}
 \end{equation*}
 \begin{equation*}
-\inference[(contraction)]
-  {\Gamma, x\!:\!\alpha, y\!:\!\alpha, \Delta \vdash e : \gamma}
-  {\Gamma, x\!:\!\alpha, \Delta \vdash \subst{e}{y}{x} : \gamma}
+\tyrule{weakening}
+  {\Gamma, \Delta \vdash e : \gamma}
+  {\Gamma, x\!:\!\tau, \Delta \vdash e : \gamma}
+\end{equation*}
+\begin{equation*}
+\tyrule{contraction}
+  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_1, \Delta \vdash e : \tau_2}
+  {\Gamma, x\!:\!\tau_1, \Delta \vdash \subst{e}{y}{x} : \tau_2}
 \end{equation*}
@@
 
-In sub-structural type systems, the assumptions are typically treated as a list. As a result,
-they have to be manipulated explicitly. Different systems allow different subset of the rules.
-For example, \emph{affine} systems allows exchange and weakening, leading to a system where 
-variable may be used at most once; in \emph{linear} systems, only exchange is permitted and so 
-every variable has to be used exactly once.
-
-When tracking context-dependent properties associated with individual variables, we need to 
-be more explicit in how variables are used. Sub-structural type systems provide a way to do this.
-Even when we allow all three operations, we can track which variables are used and how
-(and use that to track additional contextual information about variables).
-
 %---------------------------------------------------------------------------------------------------
 
-\paragraph{Bunched type systems}
+\subsection{Bunched type systems.}
 Bunched typing makes one more refinement to how $\Gamma$ is treated. Rather than having a list
@@
 The type system can be interpreted as specifying whether a variable may be shared between the 
-body of a function and the context where a function is declared. The system introduces two 
-distinct function types $\alpha \rightarrow \tau_2$ and $\alpha~ \textendash\!\!\!\ast \tau_2$
+body of a function and the context where a function is declared.
+
+The system introduces two 
+distinct function types $\tau_1 \rightarrow \tau_2$ and $\tau_1~ \textendash\!\!\!\ast \tau_2$
 (corresponding to ``;'' and ``,'' respectively). The key property is that only the first kind
@@
 restricts such sharing. We do not attempt to give a detailed description here as it is not 
-immediately to coeffects -- for more information, refer to O'Hearn's introduction 
+immediately related to coeffects -- for more information, refer to O'Hearn's introduction 
 \cite{substruct-bunched}.
@@
 \begin{equation*}
-\inference[(exchange1)]
+\tyrule{exchange1}
   {\Gamma(\Delta, \Sigma) \vdash e : \alpha}
   {\Gamma(\Sigma, \Delta) \vdash e : \alpha}
-\quad
-\inference[(weakening)]
+\end{equation*}
+\begin{equation*}
+\tyrule{exchange2}
+  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
+  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
+\end{equation*}
+\begin{equation*}
+\tyrule{weakening}
   {\Gamma(\Delta) \vdash e : \alpha}
@@
 \begin{equation*}
-\inference[(exchange2)]
-  {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
-  {\Gamma(\Sigma; \Delta) \vdash e : \alpha}
-\quad
-\inference[(contraction)]
+\tyrule{contraction}
   {\Gamma(\Delta; \Sigma) \vdash e : \alpha}
@@
 
-\paragraph{Thesis perspective}
+\paragraph{Thesis perspective.}
 
-Our work can be viewed as annotating bunches. Such annotations then specify additional information
-about the context -- or, more specifically, about the sub-tree of the context. Although this is not
-the exact definition used in Chapter~X, we could define contexts as follows:
+From the perspective of substructural and bunched types, our work can be viewed as annotating 
+bunches. Such annotations then specify additional information about the context -- or, more
+specifically, about the sub-tree of the context. Although this is not the exact definition used in 
+Chapter~\ref{ch:unified}, we could define contexts as follows:
 %
 \begin{equation*}
-\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep 1 \sep \Gamma, \Gamma \sep \Gamma @ \sigma
+\Gamma, \Delta, \Sigma := x\!:\!\alpha \sep 1 \sep \Gamma, \Gamma \sep \coctx{\Gamma}{\cclrd{r}}
 \end{equation*}
@@
 Now we can not only annotate an entire context with some information (as in the simple coeffect
-system for tracking resources that used judgements of a form $\Gamma @ \sigma \vdash e : \alpha$).
+system for tracking resources that used judgements of a form $\cclrd{\Gamma}{\cclrd{r}} \vdash e : \tau$).
 We can also annotate individual components. For example, a context containing variables $x,y,z$
-where only $x$ is used could be written as $(x\!:\!\alpha \,@\, \ident{used}), ((y\!:\!\alpha, z\!:\!\alpha) 
-\,@\, \ident{unused})$.
+where only $x$ is used could be written as $\coctx{(x\!:\!\tau_1}{\ident{\cclrd{used}}}, \coctx{(y\!:\!\tau_2, z\!:\!\tau_3)}{\ident{\cclrd{unused}}}$.
 
@@
 work on bunched logic. Aside from this principal connection, \emph{structural coeffects} also 
-use some of the proof techniques from the work on bunched logics, because they also use tree-like
-structure of variable contexts.
+use some of the proof techniques from the work on bunched logics.
 
@@
 in the area using functional languages has been done by Serrano \cite{app-hop-diffuse,app-hop-lang}
-(which also inspired the example presented in Chapter~\ref{ch:intro}). His HOP language supports 
+(which also inspired the motivating example presented in Chapter~\ref{ch:intro}). His HOP language supports 
 cross-compilation and programs execute in different contexts. However, HOP is not statically 
@@
 We approach the problem from a different perspective, building on the tradition of 
-statically-typed functional programming languages and their theories.
+statically-typed functional programming languages, focusing on type systems as the primary way
+of capturing contextual properties.
 
@@
 
-This chapter presented four different pathways leading to the idea of coeffects. We also 
+This chapter presented three different pathways leading to the idea of coeffects. We also 
 introduced the most important related work, although presenting related work was not the
-main goal of the chapter. The main goal was to show the idea of coeffects as a logical follow up
-to a number of research directions. For this reason, we highlighted only certain aspects of 
-related work -- the remaining aspects as well as important technical details are covered 
-in later chapters.
+primary goal of the chapter. The primary goal was to present the idea of coeffects as a logical 
+follow up to a number of research directions. For this reason, we highlighted only certain aspects 
+of the discussed related work -- the remaining aspects as well as important technical details are
+covered throughout the thesis.
 
-The first pathway looks at applications and systems that involve notion of \emph{context}.
-The two coeffect calculi we present aim to unify some of these systems. The second pathway
-follows as a dualization of well-known effect systems. However, this is not simply a syntactic
-transformation, because coeffect systems treat lambda abstraction differently. The third 
-pathway follows by extending comonadic semantics of context-dependent computations with 
-indexing and building a type system analogous to effect system from the ``marriage of effects
-and monads''. Finally, the fourth pathway starts with sub-structural type systems. Coeffect
-systems naturally arise by annotating bunches in bunched logics with additional information.
-
-% \section{Missing}
-% ~
-% 
-% indexed/layered/etc. monads, productors or whatever
+The first pathway follows as a dualization of well-known effect systems. However, this is not simply 
+a syntactic transformation. As we further discuss in the next chapter, coeffect systems treat lambda 
+abstraction differently. The second pathway follows by extending comonadic semantics of context-dependent 
+computations with indexing and building a type system analogous to effect system from the ``marriage of 
+effects and monads''. Finally, the third pathway starts with substructural type systems. Coeffect
+systems naturally arise by annotating bunches in bunched logics with additional information. In this
+thesis, we mostly follow the first two approaches.

### 2014-12-06 text/AppendixB.tex
Proofreading (Chapter 3)

%!TEX root = ../main.tex

\chapter{Appendix A} 
\label{ch:appendix}

This appendinx provides additional details for some of the proofs for equational theory
of flat coeffect calculus from Chapter~\ref{ch:flat} and structural coeffect calculus
from Chapter~\ref{ch:structural}.

\section{Substitution for flat coeffects}
\label{sec:appendix-flat-cbn}
In Section~\ref{sec:flat-syntax-cbn}, we stated that, for a bottom-pointed flat coeffect
algebra (\ie~$\forall r \in \C \;.\; r \,\cgeq\, \cunit $), the call-by-name substitution 
preserves type if all operators of the flat coeffect algebra coincide (Lemma~\ref{thm:cbn-substitution-bot}).
This section provides the corresponding proof.

\begin{lemma*}[Bottom-pointed substitution]
In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
where $\czip = \cseq = \cpar$ and the operation is also idempotent and commutative and
$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
%
\begin{equation*}
\begin{array}{l}
 \coctx{\Gamma}{\cclrd{S}} \vdash e_s : \tau_s \;\; \wedge \;\; 
 \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{S} } \vdash \subst{e_r}{x}{e_s} : \tau_r
\end{array}
\end{equation*}

\end{lemma*}
\begin{proof}
Assume that $\coctx{\Gamma}{\cclrd{S}} \vdash e_s : \tau_s$ and we are substituting a term $e_s$ for 
a variable $x$. Note that we use upper-case $\cclrd{S}$ to distinguish the coeffect of the expression
that is being substituted into an expression. Using structural induction over $\vdash$:

\paragraph{(var)} Given the following derivation using (\emph{var}):
\[
\inference
  { }
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit} \vdash y : \tau }
\]
There are two cases depending on whether $y$ is the variable $x$ or not:
\begin{itemize}
\item[--] If $y=x$ then also $\tau = \tau_s$ and thus $\subst{y}{x}{e_s} = e_s$. Using the assumption,
implicit weakening and the fact that $\cunit$ is a unit of $\cseq$:
\[
\inference
 {\inference 
  {\coctx{\Gamma}{\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau_s }
  {\coctx{\Gamma_1, \Gamma, \Gamma_2}{\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau } }
 {\coctx{\Gamma_1, \Gamma, \Gamma_2}{\cunit\,\cseq\,\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau } 
\]
\item[--] If $y\neq x$ then $\subst{y}{x}{e_s} = y$. Using the fact that $\cunit$ is the bottom element
and sub-coeffecting:
\[
\inference
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit} \vdash y : \tau }
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit\,\cseq\,\cclrd{S}} \vdash y : \tau }
\]
\end{itemize}

\paragraph{(const)} Similar to the (\emph{var}) case when the variable is not substituted.

\paragraph{(sub)} Given the following typing derivation using (\emph{sub}):
\[
\inference
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{r'}} \vdash e : \tau }
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
\]
From the induction hypothesis, we have that
$\coctx{\Gamma_1,\Gamma,\Gamma_2}{\cclrd{r'}\,\cseq\,\cclrd{S}} \vdash \subst{e}{x}{e_s} : \tau$.
The condition on $\cleq$ means that $\cclrd{r'}\,\cseq\,\cclrd{S}\;\cleq\;\cclrd{r}\,\cseq\,\cclrd{S}$
and so we can apply the (\emph{sub}) rule to obtain
$\coctx{\Gamma_1,\Gamma,\Gamma_2}{\cclrd{r}\,\cseq\,\cclrd{S}} \vdash \subst{e}{x}{e_s} : \tau$.

\paragraph{(abs)} Given the following typing derivation using (\emph{abs}):
\[
\inference
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2, y:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{r}} \vdash \lambda y.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\]
Assume w.l.o.g. that $x\neq y$. From the induction hypothesis, we have that:
\[
\coctx{\Gamma_1,\Gamma,\Gamma_2, y:\tau_1}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e}{x}{e_s} : \tau_2
\]
Now using the fact that $\czip\,=\,\cseq$, associativity and commutativity and (\emph{abs}):
\[
\inference
 {\inference
  {\coctx{\Gamma_1,\Gamma,\Gamma_2, y:\tau_1}{ (\cclrd{r} \,\czip\, \cclrd{s})\,\cseq\,\cclrd{S} } \vdash \subst{e}{x}{e_s} : \tau_2}
  {\coctx{\Gamma_1,\Gamma,\Gamma_2, y:\tau_1}{ (\cclrd{r} \,\cseq\, \cclrd{S})\,\czip\,\cclrd{s} } \vdash \subst{e}{x}{e_s} : \tau_2}}
 {\inference
  {\coctx{\Gamma_1,\Gamma,\Gamma_2}{\cclrd{r} \,\cseq\, \cclrd{S}} \vdash \lambda y.(\subst{e}{x}{e_s}) : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
  {\coctx{\Gamma_1,\Gamma,\Gamma_2}{\cclrd{r} \,\cseq\, \cclrd{S}} \vdash \subst{(\lambda y.e)}{x}{e_s} : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} }
\]

\paragraph{(app)} Given the following typing derivation using (\emph{app}):
\[
\inference
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
   \coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
\]
From the induction hypothesis, we have that:
\[
\begin{array}{l}
 \coctx{\Gamma_1, \Gamma, \Gamma_2}{\cclrd{r}\,\cseq\,\cclrd{S}} \vdash \subst{e_1}{x}{e_s} : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \\
 \coctx{\Gamma_1, \Gamma, \Gamma_2}{\cclrd{s}\,\cseq\,\cclrd{S}} \vdash \subst{e_2}{x}{e_s} : \tau_1
\end{array}\quad(*)
\]
Now using (\emph{app}) rule and the fact that $\cpar\,=\,\cseq$, associativity, commutativity and idempotence
(note that all three properties are needed):
\[
\inference
 { (*) }
 {\inference
   { \coctx{\Gamma_1,\Gamma,\Gamma_2}{(\cclrd{r}\,\cseq\,\cclrd{S}) \;\cpar\; ((\cclrd{s}\,\cseq\,\cclrd{S}) \,\cseq\, \cclrd{t})} 
       \vdash \subst{e_1}{x}{e_s}~\subst{e_2}{x}{e_s} : \tau_2}
   { \coctx{\Gamma_1,\Gamma,\Gamma_2}{(\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t}))\;\cseq\;\cclrd{S}} 
       \vdash \subst{(e_1~e_2)}{x}{e_s} : \tau_2} }
\]

\paragraph{(let)} Given the following typing derivation using (\emph{let}):
\[
\inference
  { \coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma_1,x:\tau_s,\Gamma_2, y:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma_1,x:\tau_s,\Gamma_2}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~y=e_1~\kvd{in}~e_2 : \tau_2 }
\]
From the induction hypothesis, we have that:
\[
\begin{array}{l}
 \coctx{\Gamma_1, \Gamma, \Gamma_2}{\cclrd{r}\,\cseq\,\cclrd{S}} \vdash \subst{e_1}{x}{e_s} : \tau_1 \\
 \coctx{\Gamma_1, \Gamma, \Gamma_2, y:\tau_1}{\cclrd{s}\,\cseq\,\cclrd{S}} \vdash \subst{e_2}{x}{e_s} : \tau_2
\end{array}\quad(\dagger)
\]
Now using (\emph{let}) rule and similarly to the (\emph{app}) case:
\[
\inference
 { (\dagger) }
 { \inference
   { \coctx{\Gamma_1, \Gamma, \Gamma_2}{(\cclrd{s}\,\cseq\,\cclrd{S}) \;\cpar\; ((\cclrd{s}\,\cseq\,\cclrd{S}) \,\cseq\, (\cclrd{r}\,\cseq\,\cclrd{S}))} 
         \vdash \kvd{let}~y=\subst{e_1}{x}{e_s}~\kvd{in}~\subst{e_2}{x}{e_s} : \tau_2 } 
   { \coctx{\Gamma_1, \Gamma, \Gamma_2}{(\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})) \,\cseq\, \cclrd{S}} 
         \vdash \subst{(\kvd{let}~y=e_1~\kvd{in}~e_2)}{x}{e_s} : \tau_2 } }
\]
\end{proof}

%
% (e2 + (e1 * e2)) + (e3 * f)
% ----(glet)------
% ----------(app)------------
%
% = e2 + e1*e2 + e3*f
%  
% ~~~>
%
% (e2 + (e3 * f)) + ((e2 + (e3 * f)) * e1)
% ----(app)------    ----(app)-----
% ---------------(glet)-------------------
%
% = e2 + e3*f + e2*f + e1*e3*f

% e3 + ((e2 + (e1 * e2)) * e3)
%       ------(glet)---
% ---------(glet)-------------
%
% = e3 + e2*e3 + e1*e2*e3
%  
% ~~~>
%
% (e2 + (e3 * e2)) + (e1 * (e2 + (e3 * e2)))
% -----(glet)-----         -----(glet)-----
% ------------------(glet)------------------
%
% = e2 + e3*e2 + e1*e2 + e1*e2*e3

% Data-flow: e3 + (e2 + e1) = (e2 + e3) + e1
% Liveness:  e3 = e3
% Resources: OK

%---------------------------------------------------------------------------------------------------

% \section{Internalized substitution}

% \subsection{First transformation}

% \begin{equation*}
% (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 \leadsto \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
% \end{equation*}

% \begin{equation*}
% \tyrule{app}
%   { \tyrule{glet}
%       { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
%         \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 }
%       { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 } &
%     \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3  }
%   { \coctx{\Gamma}{(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) } \vdash (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 : \tau_2 }
% \end{equation*}

% to

% \begin{equation*}
% \tyrule{glet}
%   { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
%     \tyrule{app}
%       { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 & 
%         \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3 }
%       { \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})} \vdash e_2~e_3 : \tau_2 } }
%   { \coctx{\Gamma}{ (\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})) \;\cpar\; (\cclrd{s}\,\cseq\,(\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})))  }
%       \vdash \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3) : \tau_2 }
% \end{equation*}

% meaning
% \begin{equation*}
% \begin{array}{l}
% (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) =
% \end{array}
% \end{equation*}

% \subsection{Second transformation}

% Second transformation

% \begin{equation*}
% \tyrule{glet}
%   { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
%   { \coctx{\Gamma}{
%       \cclrd{t}
%       \;\cpar\;
%       (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
%          \,\cseq\, 
%          \cclrd{t} )
%     } 
%     \vdash \kvd{glet}~x_r=(\kvd{glet}~x_s=e_s~\kvd{in}~e_r)~\kvd{in}~e_t : \tau_t }
% \end{equation*}

% or

% \begin{equation*}
% \tyrule{glet}
%   { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
%   { \coctx{\Gamma}{
%       (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
%       \;\cpar\;
%       (  \cclrd{s}
%          \,\cseq\, 
%          (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) )
%     } 
%     \vdash \kvd{glet}~x_s=e_s~\kvd{in}~(\kvd{glet}~x_r=e_r~\kvd{in}~e_t) : \tau_t }
% \end{equation*}

% \begin{equation*}
% \begin{array}{l}
%   \cclrd{t}
%   \;\cpar\;
%   (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
%      \,\cseq\, 
%      \cclrd{t} ) = 
% \\
%   \cclrd{t}
%   \;\cpar\;
%   (\cclrd{r}\,\cseq\, \cclrd{t})
%   \;\cpar\;
%   (\cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t}) = 
% \\
%   \cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t} 
% \end{array}
% \end{equation*}

% \begin{equation*}
% \begin{array}{l}
%   (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
%   \;\cpar\;
%   (  \cclrd{s}
%      \,\cseq\, 
%      (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) ) = 
% \\     
%   \cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})
%   \;\cpar\;
%   (\cclrd{s} \,\cseq\, \cclrd{t})
%   \;\cpar\;
%   (\cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t})  = 
% \\     
%   \cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t}
% \end{array}
% \end{equation*}

% require

% \begin{equation*}
% \cclrd{r}\;\cpar\;(\cclrd{r} \,\cseq\, \cclrd{s}) = \cclrd{r} \,\cseq\, \cclrd{s}
% \end{equation*}

### 2014-12-06 text/Chapter01.tex
Proofreading (Chapter 3)

@@
 \begin{itemize}
-\item \textbf{System capabilities.} Libraries such as LINQ \cite{app-linq} let developers write code
+\item \textsc{System capabilities.} Libraries such as LINQ \cite{app-linq} let developers write code
   in a host language like C\# and then cross-compile it to multiple targets (including SQL, OpenCL 
@@
 
-\item \textbf{Platform versions.} When developing cross-platform applications, different platforms
+\item \textsc{Platform versions.} When developing cross-platform applications, different platforms
   (and different versions of the same platform) provide different API functions. Writing a
@@
 
-\item \textbf{Security and provenance.} When working with data (be it sensitive database or social
+\item \textsc{Security and provenance.} When working with data (be it sensitive database or social
    network data), we may have permission to access only some of the data and we may want to track
@@
 
-\item \textbf{Resources \& data availability.} When creating a mobile application, the program may
+\item \textsc{Resources \& data availability.} When creating a mobile application, the program may
   (or may not) be granted access to device capabilities such as GPS sensor, social updates or battery
@@
 \begin{itemize}
-\item \textbf{Flat coeffects} represent additional data, resources and meta-data that are
+\item \textsc{Flat coeffects} -- represent additional data, resources and meta-data that are
    available in the execution environment (regardless of how they are accessed in a program).
@@
 
-\item \textbf{Structural coeffects} capture additional meta-data related to inputs. This can include
+\item \textsc{Structural coeffects} -- capture additional meta-data related to inputs. This can include
    provenance (source of the input value), usage information (how often is the value
@@
 \begin{itemize}
-\item In Chapter~\ref{ch:pathways} we give an overview of related work. Most importantly, we show
+\item In Chapter~\ref{ch:pathways}, we give an overview of related work. Most importantly, we show
   that the idea of context-aware computations can be naturally approached from a number of
@@
 
-\item Chapter~\ref{ch:applications} presents the first contribution of the thesis -- the discovery
+\item In Chapter~\ref{ch:applications}, we present the first contribution of the thesis -- the discovery
   of the connection between a number of existing programming language features that are related

### 2014-12-06 text/Chapter02.tex
Proofreading (Chapter 3)

@@
 \subsection{Simple coeffect system.}
+\label{sec:path-effects-coeff}
 When writing the judgements of coeffect systems, we want to emphasize the fact that coeffect 

### 2014-12-06 text/Chapter03.tex
Finish Ch6 and tweaks everywhere (Chapter 6)
Proofreading (Chapter 3)

@@
 Software developers as well as programming language researchers choose abstractions based not 
-just on how appropriate they are. Other factors may include social aspects -- how well is the 
+just on how appropriate they are. Other factors include social aspects -- how well is the 
 abstraction known, how well is it documented and whether it is a standard tool of the
-\emph{research programme}\footnote{ Research programme, as introduced by Lakatos \cite{philosophy-lakatos},
-is a network of scientists sharing the same basic assumptions and techniques.}. This may
-partly be why no unified context tracking mechanism has been developed so far.
-
-In Chapter~\ref{ch:intro}, we argued that context-awareness had, so far, only limited influence 
-on the design of programming languages because it is a challenge that is not easy to
-see. However, many of the properties that this thesis treats uniformly as \emph{coeffects} have been
-previously tracked by other means. This includes special-purpose type systems, systematic 
-approaches arising from modal logic S4, as well as techniques based on abstractions designed 
-for other purpose, most frequently monads.
-
-In this chapter, we describe a number of simple calculi for tracking a wide range of contextual
-properties. The systems are adapted from existing work, but the uniform presentation in this 
-chapter is a novel contribution. The fact that we find a common structure in all systems presented 
-here lets us develop unified coeffect calculi in the upcoming three chapters.
+\emph{research programme}\footnote{A research programme, as introduced by Lakatos \cite{philosophy-lakatos},
+is a network of scientists sharing the same basic assumptions and techniques.} that the researcher
+unconsciously subscribes to.
+
+For tracking of computational effects, such \emph{standard tools} are well known. When 
+faced with an effectful computation, programming language designers immediately pick monads. For 
+context-aware computations, there are no standard tools. This means that contextual properties may, 
+at first, appear as a set of disconnected examples. Consequently, existing systems that capture 
+contextual properties use a wide range of methods including special-purpose type systems,  
+approaches arising from modal logic S4, as well as techniques based on abstractions designed for other 
+purpose, most frequently monads and effect systems.
+
+\paragraph{Chapter structure and contributions}
+\begin{itemize}
+\item[--] We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
+  explains what is a \emph{coeffect} and constrasts it with a better known notion of 
+  \emph{effect}. It explains what is the nature of properties that can be tracked using 
+  coeffect systems presented in this theis.
+
+\item[--] We describe a number of simple calculi for tracking a wide range of contextual properties. 
+  The systems are adapted from diverse sources (type systems, static analyses, logics) and apply to
+  various domains (cross-compilation, liveness, distributed computing, data-flow, security), but
+  share a common structure.
+
+\item[--] The uniform presentation of the systems is the key contribution of this chapter. We distinguish
+  between \emph{flat coeffect} systems (Section~\ref{sec:applications-flat}) and \emph{structural coeffect}
+  systems (Section~\ref{sec:applications-structural}). The fact that we find a common structure in
+  all systems presented here lets us develop unified coeffect calculi in the upcoming three chapters.
+\end{itemize}
+
+\noindent
+As mentioned earlier, this chapter may initially appear as a collection of disconnected 
+examples\footnote{The different properties captured by monads may appear similarly 
+disconnected at first!}. But at the end, we will see that they share a common pattern.
 
@@
 \section{Structure of coeffect systems}
-
-When introducing coeffect systems in Section~\ref{sec:intro-theory}, we related coeffect systems
-with effect systems. Effect systems track how program affects the environment, or, in other words 
-capture some \emph{output impurity}. In contrast, coeffect systems track what program requires from the 
-environment, or \emph{input impurity}.
+\label{sec:applications-structure}
+
+When introducing coeffect systems in Section~\ref{sec:path-effects-coeff}, we related coeffect systems
+with effect systems. Effect systems track how a program affects the environment, or, in other words 
+capture some \emph{output impurity}. In contrast, coeffect systems track what a program requires from 
+the execution environment, or \emph{input impurity}.
 
 Effect systems generally use judgements of the form $\Gamma \vdash e : \tau \;\&\; \aclrd{\sigma}$, 
-associating effects $\sigma$ with the output type. In contrast, we choose to write coeffect 
+associating effects $\sigma$ with the output type. We write coeffect 
 systems using judgements of the form $\coctx{\Gamma}{\aclrd{\sigma}} \vdash e : \tau$, associating
 the context requirements with $\Gamma$. Thus, we extend the traditional notion of free-variable
-context $\Gamma$ with richer notions of context. Besides the notation, there are more important 
-differences between effects and coeffects.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Lambda abstraction}
+context $\Gamma$ with richer notions of context. This notation emphasizes the right intuition, 
+but there are more important differences between effects and coeffects.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Effectful lambda abstraction}
 \label{sec:applications-structure-lam}
@@
 The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
-The typical lambda abstraction rule for effect systems looks as (\emph{abs-eff}) in 
+The typical lambda abstraction rule for effect systems looks as (\emph{effect}) in 
 Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how 
@@
 \begin{figure}
-\[
-\tyrule{abs-pure}
+\begin{equation*}
+\tyrule{pure}
   { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2}
   { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2}
-\;\;
-\tyrule{abs-eff}
+\end{equation*}
+\begin{equation*}
+\tyrule{effect}
   { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
   { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{\sigma}} \tau_2 \,\&\, \aclrd{\emptyset} }
-\]
+\end{equation*}
 \caption{Lambda abstraction for pure and effectful computations}
@@
 \emph{flat coeffects} refer to overall properties of the environment and \emph{structural coeffects} 
-refer to properties attached to individual variables.
+refer to properties attached to individual variables. We could track properties associated
+with values in data structures (\eg~fields of a tuple), but this is left as future work.
 
@@
 In standard $\lambda$-calculus, variable access can be seen as a primitive operation that 
-accesses the context. The expression $x$ introduces a context requirement -- the expression 
-is typeable only in a context that contains $x:\tau$ for some type $\tau$.
-
-The standard lambda abstraction (\emph{abs-pure}), shown in Figure~\ref{fig:applications-abs},
-splits the free-variable context of an expression into two parts. The value of the parameter
+accesses the context. The variable access expression introduces a context requirement -- the 
+expression $x$ is typeable only in a context that contains $x:\tau$ for some type $\tau$.
+
+The standard lambda abstraction (\emph{pure}), shown in Figure~\ref{fig:applications-abs},
+splits the free-variable context of an expression into two parts. At runtime, the value of the parameter
 has to be provided by the \emph{call site} (dynamic scope) and the remaining values are provided 
-by the \emph{declaration site} (lexical scope). Here, the splitting is determined syntactically --
-the notation $\lambda x.e$ names the variable whose value comes from the call site.
-
-The flat and structural coeffects behave in the same way. They also split context-requirements
-between the declaration site and call site, but they do it in two different ways.
+by the \emph{declaration site} (lexical scope). In the type checking, the splitting is determined 
+syntactically -- the notation $\lambda x.e$ names the variable whose value comes from the call site.
+
+Flat and structural coeffects also split context-requirements between the declaration site and 
+the call site. The flat and structural coeffects capture two different ways of doing this.
 
@@
 for example, say we have an expression $e$ that requires GPS coordinates and the current time. 
-The context of such expression will be $\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.
+The variable context of such expression will be annotated with a set of required resources, 
+\ie~$\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.
 
 The interesting case is when we construct a lambda function $\lambda x.e$, marshall it and
-send it to another node. In that case, the context requirements can be satisfied in a number of
-ways. When the same resource is available at the target machine (\eg.~current time), we can
-send the function with a context requirement and \emph{rebind} the resource. However, if the
-resource is not available (\eg.~GPS on the server), we need to capture a \emph{remote reference}.
+send it to another node. In systems such as Acute \cite{app-distributed-acute}, the context 
+requirements can be satisfied in a number of ways. When the same resource is available at the target
+machine (\eg~current time), we can transfer the function with a context requirement and \emph{rebind} 
+the resource. However, if the resource is not available (\eg.~GPS on the server), we need to a 
+capture \emph{remote reference}.
 
@@
 $\aclrd{r}=\aclrd{\footnotesize\ident{\{ time \}}}$. In coeffect systems, we attach this information
-to the function writing $\tau_1 \xrightarrow{ \aclrd{r} } \tau_2$.
+to the function, writing $\tau_1 \xrightarrow{ \aclrd{r} } \tau_2$.
 
@@
 The important point here is that in flat coeffect systems, contextual requirements are 
-\emph{split} between the call-site and declaration-site. Furthermore, in case of distributed
-programming, the resources can be freely distributed between the two sites.
+\emph{split} between the call-site and declaration-site. Furthermore, there is no syntactic
+structure that determines how the requirements are split. In the case of distributed
+programming, the resources can be freely associated with either of the two sites.
 
 \paragraph{Structural coeffects.}
-
 On the one hand, variable context provides a \emph{fine-grained tracking} mechanism of how context
@@
 the context. The purpose of \emph{structural coeffects} is to reconcile the two and to provide a way
-for fine-grained tracking of additional information related to variables in programs.
+for fine-grained tracking of additional information linked to variables in programs. Structural 
+coeffects follow the lexical scoping structure determined by the typing rules.
 
@@
 Here, the variable context $\Gamma$ contains two variables, both of type \ident{Arr}. This means
-$\Gamma = x\!:\!\ident{Arr},\, y\-!\-!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a 
+$\Gamma = x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a 
 language primitive. The coeffect annotations will be $(0,0)$ for $x$ and $(-1,1)$ for $y$, 
@@
 
-Unlike in flat coeffects, in the structural systems, splitting of context determined by the syntax. 
+In structural systems, the splitting of context is determined by the name (variable) binding. 
 For example, consider a function that takes $y$ and contains the above body:
@@
 a value for $y$, which needs to be accessible at least within $(-1, 1)$. In this way, structural
-coeffects remove the non-determinism of flat coeffect systems.
-
-Before looking at concrete flat and structural systems in mode details, we briefly overview 
-some notation used in this thesis. As structural coeffects keep annotations as \emph{vectors},
-we use a number of operations related to scalars and vectors.
+coeffects remove the non-determinism arising from the splitting of requirements in flat coeffect systems.
+
+Before looking at concrete flat and structural systems, we briefly overview 
+some notation used in this thesis. Structural coeffects keep annotations as \emph{vectors} and
+use a number of operations related to scalars and vectors.
 
@@
 
-The $\lambda$-calculus is asymmetric -- it maps a context with \emph{multiple} variables to a 
-\emph{single} result. An expression with free variables of types $\tau_i$ can be modelled by a function 
+The $\lambda$-calculus is asymmetric. It maps a context with \emph{multiple} variables to a 
+\emph{single} result. An expression with $n$ free variables of types $\tau_i$ can be modelled by a function 
 $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
-on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
-we attach a coeffect annotation to the context $\tau_1 \times \ldots \times \tau_n$.
+on the right. In both effect systems and coeffect systems, we \emph{write} the annotation as part of
+the function arrow. However, in the underlyng categorical model, effects are attached to the result
+$\tau$, while coeffects are attached to the context $\tau_1 \times \ldots \times \tau_n$.
 
@@
 It is useful to clarify how vectors are used in this thesis. Suppose we have a set $\C$ of
-\emph{scalars} such that $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ 
+\emph{scalars} ranged over by $\cclrd{r},\cclrd{s},\cclrd{t}$. A vector $\aclrd{R}$ 
 over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars. 
-We use bold-face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for vectors and lower-case
-letters $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
+We use bold face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for vectors and 
+normal face $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
 also distinguishes different structures using colours. However ignoring the colour does not introduce 
-any ambiguity.}. We also say that a \emph{shape} of a vector $\slift{\aclrd{\textbf{r}}}$ (or more generally any container) 
-is the set of \emph{positions} in a vector. So, a vector of length $n$ has shape $\{ 1, 2, \ldots, n \}$. 
-We discuss containers and shapes further in Chapter~\ref{ch:unified} and also discuss how our use
-relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.
-
-Just as in the usual multiplication of a vector by scalar, we lift any binary operation on scalars into a 
-scalar-vector one. For any binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
+any ambiguity.}. We also say that a \emph{shape} of a vector $\slen{\aclrd{\textbf{r}}}$ (or more generally any container) 
+determines the set of \emph{positions} in a vector. So, a vector of a shape (length) $n$ has positions 
+$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Chapter~\ref{ch:unified} and also 
+discuss how our use relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.
+
+Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a 
+scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
  $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
@@
 
-In flat coeffect systems, the additional contextual information are independent of the variables.
-As such, flat coeffects capture properties where the execution environment provides some 
+In flat coeffect systems, the additional contextual information are independent of lexically scoped
+variables. As such, flat coeffects capture properties where the execution environment provides some 
 additional data, resources or information about the execution context.
 
-In this section, we look at a number of examples ranging from Haskell's type constraints
-and implicit parameters to distributed computing. For three of our examples --
-implicit parameters, liveness analysis and data-flow -- we show an ad-hoc type system that
-captures their properties. This serves as a basis for Chapter~\ref{ch:flat}, which develops
-a unified flat coeffect calculus.
+As mentioned in the introduction, coeffect systems in this chapter may appear as a disconnected
+set of examples at first. Indeed, this section covers a diverse set of calculi including
+Haskell's implicit parameters (Section~\ref{sec:applications-flat-impl}), distributed computing and 
+cross-compilation (Section~\ref{sec:applications-flat-distr}), liveness analysis 
+(Section~\ref{sec:applications-flat-live}) and data-flow (Section~\ref{sec:applications-flat-dataflow}).
+
+For three of the examples, we present a type system and a simple semantics. Although the examples 
+are not new, our novel presentation of the systems (and the fact that they appear side-by-side)
+makes it possible to see that they share a common structure. The structure is captured by
+a unified \emph{flat coeffect calculus} in Chapter~\ref{ch:flat}.
 
@@
 
-Haskell provides two examples of flat coeffects -- type class and implicit parameter constraints
-\cite{app-type-classes,app-implicit-parameters}. Both of the features introduce additional 
+Haskell provides two examples of flat coeffects -- type class constraints and implicit parameter 
+constraints \cite{app-type-classes,app-implicit-parameters}. Both of the features introduce additional 
 \emph{constraints} on the context requiring that the environment provides certain operations for
@@
 In this section, we present a type system for implicit parameters in terms of the coeffect typing
-judgement. We briefly consider type classes, but do no give a full type system.
+judgement. We briefly consider type classes, but do not give a full type system.
 
@@
 Implicit parameters are a special kind of variables that support dynamic scoping.
-They can be used to parameterise a computation (involving a long chain of function calls)
+They make it possible to parameterise a computation (involving a long chain of function calls)
 without passing parameters explicitly as additional arguments of all involved functions.
@@
 parameters also support lexical scoping. If the parameter $\ident{?param}$ is available in the 
-lexical scope where a function (which uses it) is defined, then the function will not require a
-value from the caller.
+lexical scope where a function is defined, then the function will not require a value from the caller.
 
@@
 To keep examples in this chapter uniform, we do not use the Haskell notation and instead
-write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified $r$.
+write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
 In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
@@
 \begin{equation}
-\tag{dynamic}
+\tag{\emph{dynamic}}
 \ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
@@
 \begin{equation}
-\tag{mixed}
+\tag{\emph{mixed}}
 \ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
@@
 %
-This is not a complete list of possible typings, but it demonstrates the options. The \emph{dynamic}
+This is not a complete list of possible typings, but it demonstrates the options. The (\emph{dynamic})
 case requires the parameter \ident{?width} twice (this may be confusing when the caller provides
-two different values). In the \emph{mixed} case, the nested function captures the \ident{?width} 
-parameter available from the declaration site. As a result, the latter function can be called as follows:\
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{formatHello} = \\[-0.25em]
-\quad(~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
+two different values). In the (\emph{mixed}) case, the nested function captures the \ident{?width} 
+parameter available from the declaration site. Using the latter typing, the function can be called as follows:\
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{formatHello} = (~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
 \kvd{in}\,(~\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}~)
@@
 meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set 
-of implicit parameters specified by $\aclrd{r}$. The annotations $\aclrd{r},\aclrd{s},\aclrd{t}$ are sets of 
-pairs consisting of implicit parameter names and types, \ie~$\aclrd{r},\aclrd{s},\aclrd{t} \subseteq 
-  \ident{Names} \times \ident{Types}$. The expressions include \ident{?param} to read implicit
+of implicit parameters specified by $\aclrd{r}$. The annotations $\aclrd{r},\aclrd{s},\aclrd{t}$ are 
+finite partial functions mapping implicit parameter names to types, \ie~$\aclrd{r},\aclrd{s},\aclrd{t} \subseteq 
+\ident{Names} \mapsto \ident{Types}$. The expressions include \ident{?param} to read implicit
 parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
@@
 Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that 
-introduces primitive context requirements is (\emph{param}) -- accessing a parameter \ident{?param} 
+introduces primitive context requirements is (\emph{param}). Accessing a parameter \ident{?param} 
 of type $\tau$ requires it to be available in the context. The context may provide more (unused) 
@@
 A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
-sub-coeffecting rule (\emph{sub}) uses a point-wise extension of the $\sqsubseteq$ relation over two
+sub-coeffecting rule (\emph{sub}) uses a pointwise extension of the $\sqsubseteq$ relation over two
 vectors as defined in Section~\ref{sec:applications-strucutre-vec}.
@@
 \begin{array}{l}
-\slift{\Gamma_1} = \slift{\aclrd{\textbf{r}}}\\[-0.25em]
-\slift{\Gamma_2} = \slift{\aclrd{\textbf{s}}}
+\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
+\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
 \end{array}
@@
 
-In flat data-flow, we identified sequential composition and point-wise composition as two primitive
+In flat data-flow, we identified sequential composition and pointwise composition as two primitive
 operations that were used in the (flat) application. In the structural system, these are used in 

### 2014-12-06 text/Chapter04.tex
Proofreading (Chapter 3)

@@
 structure -- their lambda abstraction duplicates the requirements and their application arises
-from the combination of \emph{sequential} and \emph{point-wise} composition.
+from the combination of \emph{sequential} and \emph{pointwise} composition.
 
@@
 To make the flat coeffect system general enough, the algebra consists of three operations.
-Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{point-wise} composition, 
+Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{pointwise} composition, 
 respectively and the third one, $\czip$ represents context \emph{merging}. The term merging should be 
@@
 
-\item The idempotent monoid $(\C, \cpar, \czero)$ represents \emph{point-wise} composition, 
+\item The idempotent monoid $(\C, \cpar, \czero)$ represents \emph{pointwise} composition, 
 \ie~the case when the same context is passed to multiple (independent) computations. The monoid 
@@
 \begin{itemize}
-\item \textbf{Analysis of context dependence.}
+\item \textsc{Analysis of context dependence.}
 Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context requirements}. This means
@@
 
-\item \textbf{Semantics of context passing.}
+\item \textsc{Semantics of context passing.}
 Semantically, coeffect annotations $\cclrd{r}$ mo\-del \emph{contextual capabilities}. This means
@@
 As in Section~\ref{sec:applications-flat-live}, sequential composition $\cseq$ is modelled by 
-the meet operation $\sqcap$ and point-wise composition $\cpar$ is modelled by join $\sqcup$. 
+the meet operation $\sqcap$ and pointwise composition $\cpar$ is modelled by join $\sqcup$. 
 Two-point lattice is a commutative, idempotent monoid. The distributivity 
@@
 \noindent
-As discussed earlier, sequential composition $\cseq$ is represented by $+$ and point-wise 
+As discussed earlier, sequential composition $\cseq$ is represented by $+$ and pointwise 
 composition $\cpar$ uses $\emph{max}$. For data-flow, we need a third separate operator for
@@
 Pairs, created using the $(e_1, e_2)$ expression, are annotated with a coeffect that combines
-the coeffects of the two sub-expressions using the \emph{point-wise} operator $\cpar$. The operator
+the coeffects of the two sub-expressions using the \emph{pointwise} operator $\cpar$. The operator
 models the case when the (same) available context is split and passed to two independent 
@@
 
-\item Syntactically, coeffect calculi use a richer algebraic structure with point-wise composition,
+\item Syntactically, coeffect calculi use a richer algebraic structure with pointwise composition,
   sequential composition and context merging $(\cpar, \cseq, \czip)$ while most effect systems

### 2014-12-06 text/Chapter05.tex
Proofreading (Chapter 3)

@@
 modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
-point-wise composition, as well as a relation $\cleq$ that defines sub-coeffecting.
+pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.
 
@@
 
-\item A point-wise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
+\item A pointwise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
   such that\\ $\cclrd{t} \,\aseq\, \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} = 
@@
 we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
-for a point-wise extension of the $+$ operator, such that
+for a pointwise extension of the $+$ operator, such that
 $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} + k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
@@
 \begin{definition}
-Given a monoid $(\C, \cseq, \cunit)$ with a point-wise extension of the $\cseq$ operator to a vector 
+Given a monoid $(\C, \cseq, \cunit)$ with a pointwise extension of the $\cseq$ operator to a vector 
 (written as $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$) and an operation lifting scalars to vectors 
@@
 operates on a singleton vector, which is equivalent to operating on a scalar as before.
-The operation could easily be extended to a vector in a point-wise way, but we keep it simple
+The operation could easily be extended to a vector in a pointwise way, but we keep it simple
 and perform sub-coeffecting separately on individual variables.
@@
 by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations 
-$\cseq$ and $\cpar$ that model sequential and point-wise composition, respectively. For flat 
+$\cseq$ and $\cpar$ that model sequential and pointwise composition, respectively. For flat 
 coeffects, we further require $\czip$ to model context merging. For structural coeffects, we 

### 2014-12-07 text/AppendixA.tex
Finish cbn section, notes on normal forms (Chapter 4)
Finish equational theory, subtyping and lambda (Chapter 4)
Finish tweaks in flat coeffects (Chapter 3)

@@
+%!TEX root = ../main.tex
+
 \chapter{Appendix A} 
@@
 
-%---------------------------------------------------------------------------------------------------
+This appendix gives full typing derivations for the examplespresented in Chapter~\ref{ch:applications},
+which demonstrate simple coeffect calculi for implicit parameters, liveness and data-flow.
 
-\section{Internalized substitution}
+% --------------------------------------------------------------------------------------------------
 
-\subsection{First transformation}
+\section{Coeffect typing for implicit parameters}
+\label{sec:appendixa-implicit}
+
+The example on page~\pageref{pg:applications-flat-paramsex} considers the typing of a function that
+adds the values of two implicit parameters: $\lambda x.\ident{?a} + \ident{?b}$. Note that addition
+is a function (and so we write $(+)~\ident{?a}~\ident{?b}$) that requires no implicit parameters.
+Thus our typing derivation starts with: 
+%
+\begin{equation*}
+\Gamma_0 = (+):\ident{int}\xrightarrow{\cclrd{\emptyset}}\ident{int}\xrightarrow{\cclrd{\emptyset}}\ident{int}
+\end{equation*}
+%
+The typing for the sub-expression $(+)~\ident{?a}$ is shown in Figure~\ref{fig:appendixa-impl} (a).
+Note that the resulting function does not require any implicit parameters, so there is no non-determinism so far.
+It is worth noting that this would change if we used $\eta$-expansion and wrote $(\lambda y.(+)~\ident{?a})$.
+We refer to the above as $\Delta$. The rest of the expresion is shown in Figure~\ref{fig:appendixa-impl} (b).
+
+Here, the last rule applies (\emph{app}) and so we non-deterministically split the set of required resources.
+This means that we need $\cclrd{r_1}, \cclrd{r_2}$ such that $\cclrd{r_1}\,\cup\,\cclrd{r_2}=\{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \}$.
+The Figure~\ref{fig:appendixa-impl} (c) summarizes the 9 options.
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Coeffect typing for liveness}
+\label{sec:appendixa-liveness}
+
+\newcommand{\vrl}{\cclrd{\ident{L}}}
+\newcommand{\vrd}{\cclrd{\ident{D}}}
+
+The Section~\ref{sec:applications-flat-live} discusses a coeffect system where the coeffect 
+annotations capture whether a variable may be accessed (marked as live using $\vrl$) or 
+whether it is definitely not used (marked as dead using $\vrd$). The following three examples
+are considered:
+%
+\begin{equation*}
+\begin{array}{lll}
+(\lambda x . 42)~y &~\hspace{1em}~&(1)\\
+\ident{twoTimes}~42          &&(2)\\
+(\lambda x . x)~42 &&(3)\\
+\end{array}
+\end{equation*}
+%
+The typing derivation for the expression ($1$) is shown in Figure~\ref{fig:appendixa-liv} (a).
+The most interesting aspect about the previous example is the use of the (\emph{app}) rule, which
+marks the resulting context as dead, even though a variable is accessed in the second part of the
+expression (this part never needs to be evaluated).
+
+Assuming $\Gamma_0 = \ident{twoTimes} : \ident{int} \xrightarrow{\vrl} \ident{int}$,
+the typing derivation for ($2$) is shown in Figure~\ref{fig:appendixa-liv} (b).
+Here, the variable context is marked as live. This is not because the argument of \ident{twoTimes} 
+is marked as live, but because the function itself is a variable that (always) needs to be obtained
+from the variable context. Finally, the derivation for ($3$) is shown in Figure~\ref{fig:appendixa-liv} (c).
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Coeffect typing for data-flow}
+\label{sec:appendixa-dataflow}
+
+The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
+maximal number of past values required by a data-flow computation. The discussion includes the 
+following example:
+%
+\begin{equation*}
+\begin{array}{l}
+(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
+\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
+\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
+\end{array}
+\end{equation*}
+%
+In order to give typing for the above example, we first need to extend the language with 
+conditionals. The typing rule for the $\kvd{if}$ construct is:
+%
+\begin{equation*}
+\tyrule{if}
+  { \coctx{\Gamma}{\cclrd{m}} \vdash e : \ident{bool} &
+    \coctx{\Gamma}{\cclrd{n}} \vdash e_1 : \tau &
+    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau }
+  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m},\cclrd{n})} \vdash \kvd{if}~e~\kvd{then}~e_1~\kvd{else}~e_2 : \tau_1 }
+\end{equation*}
+%
+Given a condition that requires $\cclrd{m}$ past and two (alternative) bodies that each require
+$\cclrd{n}$ past values, the composed expression requires at most the maximum of the two, \ie~the
+same context is passed to both the condition and the body using point-wise composition. As we will
+see, the fact that both branches require the same number of past values means that we need to use
+the sub-coeffecting rule.
+
+Assuming $\Gamma_0 = \ident{tick}:\ident{int}, \ident{counter}:\ident{int}$, the 
+Figure~\ref{fig:appendixa-dfl} shows the typing derivation for the example. In the first part,
+we derive the types for the sub-expressions of the application and the conditional. Note that 
+to obtain compatible function types, we use sub-coeffecting \emph{before} abstraction in the
+typing of $(\lambda x.x)$ -- the type $\ident{int} \xrightarrow{\cclrd{1}} \ident{int}$ is  a
+valid overapproximation. Finally, the typing of the application yields the requirement of two 
+past values, calculated as $\textnormal{max}(\cclrd{1}, \cclrd{1}+\cclrd{1})$.
+
+\newpage
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+{\small a.) Typing for the sub-expression $(+)~\ident{?a}$}
 
 \begin{equation*}
-(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 \leadsto \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
+\begin{array}{l}
+ \tyrule{app}
+  { \hspace{-15em}\tyrule{var}{}
+    { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \emptyset }} \vdash
+      (+) : \ident{int} \xrightarrow{\cclrd{\emptyset}} \ident{int} \xrightarrow{\cclrd{\emptyset}} \ident{int}} \\ \hspace{10em}
+    \tyruler{param}{}
+    { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{ \ident{?a}:\ident{int} \} }} \vdash
+      \ident{?a} : \ident{int} } }
+  { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{ \ident{?a}:\ident{int} \} }} \vdash
+    (+)~\ident{?a} : \ident{int} \xrightarrow{\cclrd{\emptyset}} \ident{int} }
+\end{array}
 \end{equation*}
+
+\vspace{1em}
+{\small b.) Typing for the expression $(+)~\ident{?a}~\ident{?b}$}
 
@@
 \tyrule{app}
-  { \tyrule{glet}
-      { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
-        \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 }
-      { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 } &
-    \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3  }
-  { \coctx{\Gamma}{(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) } \vdash (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 : \tau_2 }
+{ \Delta\qquad\qquad &
+  \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{ \ident{?b}:\ident{int} \} }} \vdash
+  \ident{?b} : \ident{int} }
+{ \tyrule{abs}
+  { \coctx{\Gamma_0, x:\ident{int}}{\cclrd{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} }} \vdash 
+    (+)~\ident{?a}~\ident{?b} : \ident{int} }
+  { \coctx{\Gamma_0}{\cclrd{ r_1 }} \vdash 
+    \lambda x.(+)~\ident{?a}~\ident{?b} : 
+    \ident{int} \xrightarrow{ \cclrd{r_2} } \ident{int} } }
 \end{equation*}
 
-to
+\vspace{1em}
+{\small c.) Possible splittings of the implicit parameters }
 
 \begin{equation*}
-\tyrule{glet}
-  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
-    \tyrule{app}
-      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 & 
-        \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3 }
-      { \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})} \vdash e_2~e_3 : \tau_2 } }
-  { \coctx{\Gamma}{ (\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})) \;\cpar\; (\cclrd{s}\,\cseq\,(\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})))  }
-      \vdash \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3) : \tau_2 }
-\end{equation*}
-
-meaning
-\begin{equation*}
-\begin{array}{l}
-(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) =
+\begin{array}{ll}
+ \cclrd{r_1} = \{ \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
+ \cclrd{r_1} = \{ \ident{?a}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
+ \cclrd{r_1} = \{ \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
+ \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} \\
+ \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \}\\
+ \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int} \}\\
+ \cclrd{r_1} = \{ \ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?b}:\ident{int} \}\\
+ \cclrd{r_1} = \{ \ident{?a}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?b}:\ident{int} \} \\
+ \cclrd{r_1} = \{ \ident{?b}:\ident{int} \} & \cclrd{r_2} = \{ \ident{?a}:\ident{int} \} \\
 \end{array}
@@
 
-\subsection{Second transformation}
+\figcaption{Coeffect typing for implicit parameters}
+\label{fig:appendixa-impl}
+\end{figure}
 
-Second transformation
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Typing for the expression $(\lambda x . 42)~y$}
 
 \begin{equation*}
-\tyrule{glet}
-  { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
-    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
-    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
-  { \coctx{\Gamma}{
-      \cclrd{t}
-      \;\cpar\;
-      (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
-         \,\cseq\, 
-         \cclrd{t} )
-    } 
-    \vdash \kvd{glet}~x_r=(\kvd{glet}~x_s=e_s~\kvd{in}~e_r)~\kvd{in}~e_t : \tau_t }
+\tyrule{app}
+  { \hspace{-5em}
+    \tyrule{abs}
+      { \hspace{-5em}
+        \tyrule{const}{}{ \coctx{y\!:\!\tau, x\!:\!\tau}{\vrd} \vdash 42 : \ident{int} } }
+      { \coctx{y\!:\!\tau}{\vrd} \vdash \lambda x . 42 : \tau \xrightarrow{\vrd} \ident{int} }  &
+    \tyrule{var}{}{\coctx{y\!:\!\tau}{\vrl} \vdash y : \tau } }
+  { \inference 
+    { \coctx{y\!:\!\tau}{\vrd \sqcup (\vrl \sqcap \vrd)} \vdash (\lambda x . 42)~y : \ident{int} }
+    { \coctx{y\!:\!\tau}{\vrd} \vdash (\lambda x . 42)~y : \ident{int} } }
 \end{equation*}
 
-or
+\vspace{1em}
+{\small b.) Typing for the expression $\ident{twoTimes}~42$}
 
 \begin{equation*}
-\tyrule{glet}
-  { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
-    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
-    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
-  { \coctx{\Gamma}{
-      (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
-      \;\cpar\;
-      (  \cclrd{s}
-         \,\cseq\, 
-         (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) )
-    } 
-    \vdash \kvd{glet}~x_s=e_s~\kvd{in}~(\kvd{glet}~x_r=e_r~\kvd{in}~e_t) : \tau_t }
+\tyrule{app}
+  { \hspace{-5em}
+    \tyrule{var}{}{ \coctx{\Gamma_0}{\vrl} \vdash \ident{twoTimes} : \ident{int} \xrightarrow{\vrl} \ident{int}  }&
+    \tyrule{var}{}{ \coctx{\Gamma_0}{\vrd} \vdash 42 : \ident{int} } }
+  { \inference 
+      { \coctx{\Gamma_0}{\vrl \sqcup (\vrd \sqcap \vrl)} \vdash \ident{twoTimes}~42 : \ident{int}}
+      { \coctx{\Gamma_0}{\vrl} \vdash \ident{twoTimes}~42 : \ident{int}} }
 \end{equation*}
 
+\vspace{1em}
+{\small c.) Typing for the expression $(\lambda x . x)~42$}
+
 \begin{equation*}
-\begin{array}{l}
-  \cclrd{t}
-  \;\cpar\;
-  (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
-     \,\cseq\, 
-     \cclrd{t} ) = 
-\\
-  \cclrd{t}
-  \;\cpar\;
-  (\cclrd{r}\,\cseq\, \cclrd{t})
-  \;\cpar\;
-  (\cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t}) = 
-\\
-  \cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t} 
-\end{array}
+\tyrule{app}
+  { \hspace{-5em}
+    \tyrule{abs}
+      { \hspace{-5em}
+        \tyrule{var}{}{ \coctx{x\!:\!\ident{int}}{\vrl} \vdash x : \ident{int} } }
+      { \coctx{()}{\vrl} \vdash \lambda x . x : \ident{int} \xrightarrow{\vrl} \ident{int} }  &
+    \tyrule{const}{}{\coctx{()}{\vrd} \vdash 42 : \ident{int} } }
+  { \inference 
+    { \coctx{()}{\vrl \sqcup (\vrl \sqcap \vrd)} \vdash (\lambda x . x)~42 : \ident{int} }
+    { \coctx{()}{\vrl} \vdash (\lambda x . x)~42 : \ident{int} } }
 \end{equation*}
 
+\figcaption{Coeffect typing for liveness}
+\label{fig:appendixa-liv}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Typing for sub-expressions of the conditional and for the argument}
+
 \begin{equation*}
-\begin{array}{l}
-  (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
-  \;\cpar\;
-  (  \cclrd{s}
-     \,\cseq\, 
-     (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) ) = 
-\\     
-  \cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})
-  \;\cpar\;
-  (\cclrd{s} \,\cseq\, \cclrd{t})
-  \;\cpar\;
-  (\cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t})  = 
-\\     
-  \cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t}
-\end{array}
+\tyrule{prev}
+ { \coctx{\Gamma_0}{\cclrd{0}} \vdash \ident{tick} : \ident{int} }
+ { \inference
+   { \coctx{\Gamma_0}{\cclrd{1}} \vdash \kvd{prev}~\ident{tick} : \ident{int} }
+   { \coctx{\Gamma_0}{\cclrd{1}} \vdash (\kvd{prev}~\ident{tick}) = 0 : \ident{bool} } }
+\end{equation*}
+\vspace{1em}
+\begin{equation*}
+\tyrule{prev}
+ { \tyrule{var}
+    {} 
+    { \coctx{\Gamma_0,x\!:\!\ident{int}}{\cclrd{0}} \vdash x : \ident{int} } }
+ { \tyrule{abs}
+   { \coctx{\Gamma_0,x\!:\!\ident{int}}{\cclrd{1}} \vdash \kvd{prev}~x : \ident{int} }
+   { \coctx{\Gamma_0}{\cclrd{1}} \vdash \lambda x . \kvd{prev}~x : \ident{int} \xrightarrow{\cclrd{1}} \ident{int} } }
+\end{equation*}
+\vspace{1em}
+\begin{equation*}
+\tyrule{abs}
+  { \tyrule{sub}
+      { \tyrule{var}{}{ \coctx{\Gamma_0, x\!:\!\ident{int}}{\cclrd{0}} \vdash x : \ident{int} } }
+      { \coctx{\Gamma_0, x\!:\!\ident{int}}{\cclrd{1}} \vdash x : \ident{int} } } 
+  { \coctx{\Gamma_0}{\cclrd{0}} \vdash \lambda x . x : \ident{int} \xrightarrow{\cclrd{1}} \ident{int} }
+\end{equation*}
+\vspace{1em}
+\begin{equation*}
+\tyrule{prev}
+  { \tyrule{var}{}{ \coctx{\Gamma_0}{\cclrd{0}} \vdash \ident{counter} : \ident{int} } }
+  { \coctx{\Gamma_0}{\cclrd{1}} \vdash \kvd{prev}~\ident{counter} : \ident{int} }
 \end{equation*}
 
-require
+\vspace{1em}
+{\small b.) Typing for the composed expression}
 
 \begin{equation*}
-\cclrd{r}\;\cpar\;(\cclrd{r} \,\cseq\, \cclrd{s}) = \cclrd{r} \,\cseq\, \cclrd{s}
+\tyrule{app}
+  { \hspace{-5em}
+    \tyrule{if}
+      { (\ldots) } 
+      { \coctx{\Gamma_0}{\cclrd{1}} \vdash (\kvd{if}\ldots\kvd{then}\ldots\kvd{else}\ldots) : \ident{int} \xrightarrow{\cclrd{1}} \ident{int} } & 
+        \coctx{\Gamma_0}{\cclrd{1}} \vdash (\ldots) : \ident{int} }
+  { \hspace{-1em}
+    \inference
+     { \coctx{\Gamma_0}{\textnormal{max}(\cclrd{1}, \cclrd{1}+\cclrd{1})} \vdash (\kvd{if}\ldots\kvd{then}\ldots\kvd{else}\ldots)~(\ldots) : \ident{int} }
+     { \coctx{\Gamma_0}{\cclrd{2}} \vdash (\kvd{if}\ldots\kvd{then}\ldots\kvd{else}\ldots)~(\ldots) : \ident{int} } }
 \end{equation*}
+
+\figcaption{Coeffect typing for data-flow}
+\label{fig:appendixa-dfl}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------

### 2014-12-07 text/Chapter01.tex
Finish tweaks in flat coeffects (Chapter 3)

@@
 
-\caption{Conditional compilation in the HTTP module of the F\# Data library}
+\figcaption{Conditional compilation in the HTTP module of the F\# Data library}
 \label{fig:introduction-context-http}
@@
 
-\caption{Client/server news reader app implemented in a context-aware language}
+\figcaption{Client/server news reader app implemented in a context-aware language}
 \label{fig:introduction-sample}

### 2014-12-07 text/Chapter02.tex
Finish tweaks in flat coeffects (Chapter 3)

@@
 
-\caption{Simple effect system}
+\figcaption{Simple effect system}
 \label{fig:path-eff}
@@
 
-\caption{Simple coeffect system}
+\figcaption{Simple coeffect system}
 \label{fig:path-coeff}
@@
 
-\caption{Typing for a comonadic language with contextual staged computations}
+\figcaption{Typing for a comonadic language with contextual staged computations}
 \label{fig:modal-meta}
@@
 
-\caption{Exchange, weakening and contraction typing rules}
+\figcaption{Exchange, weakening and contraction typing rules}
 \label{fig:substructural-rules}
@@
 \end{equation*}
-\caption{Exchange, weakening and contraction rules for bunched typing}
+\figcaption{Exchange, weakening and contraction rules for bunched typing}
 \label{fig:substructural-bunched}

### 2014-12-07 text/Chapter03.tex
Finish tweaks in flat coeffects (Chapter 3)

@@
 
-For tracking of computational effects, such \emph{standard tools} are well known. When 
-faced with an effectful computation, programming language designers immediately pick monads. For 
-context-aware computations, there are no standard tools. This means that contextual properties may, 
-at first, appear as a set of disconnected examples. Consequently, existing systems that capture 
-contextual properties use a wide range of methods including special-purpose type systems,  
-approaches arising from modal logic S4, as well as techniques based on abstractions designed for other 
-purpose, most frequently monads and effect systems.
+For tracking of effects, such \emph{standard tools} are well known. When faced with an effectful 
+computation, programming language designers immediately pick monads. For context-aware computations, 
+there are no standard tools. Thus contextual properties may, at first, appear as a set of disconnected 
+examples. Existing systems that capture contextual properties use a wide range of methods including 
+special-purpose type systems, approaches arising from modal logic S4, as well as techniques based 
+on abstractions designed for other purpose, most frequently monads.
 
@@
   all systems presented here lets us develop unified coeffect calculi in the upcoming three chapters.
+
+\item[--] In addition, the coeffect systems for tracking the number of accessed past values in 
+  data-flow languagges (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
+  presents novel results and can be used to optimize data-flow programs.
 \end{itemize}
@@
 \noindent
-As mentioned earlier, this chapter may initially appear as a collection of disconnected 
+As mentioned, this chapter may appear as a collection of disconnected 
 examples\footnote{The different properties captured by monads may appear similarly 
@@
 
-Effect systems generally use judgements of the form $\Gamma \vdash e : \tau \;\&\; \aclrd{\sigma}$, 
+Effect systems generally use judgements of the form $\Gamma \vdash e : \tau \;\&\; \cclrd{\sigma}$, 
 associating effects $\sigma$ with the output type. We write coeffect 
-systems using judgements of the form $\coctx{\Gamma}{\aclrd{\sigma}} \vdash e : \tau$, associating
+systems using judgements of the form $\coctx{\Gamma}{\cclrd{\sigma}} \vdash e : \tau$, associating
 the context requirements with $\Gamma$. Thus, we extend the traditional notion of free-variable
@@
 \tyrule{effect}
-  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \,\&\, \aclrd{\sigma} }
-  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{\sigma}} \tau_2 \,\&\, \aclrd{\emptyset} }
-\end{equation*}
-\caption{Lambda abstraction for pure and effectful computations}
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \,\&\, \cclrd{\sigma} }
+  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{\sigma}} \tau_2 \,\&\, \cclrd{\emptyset} }
+\end{equation*}
+
+\figcaption{Lambda abstraction for pure and effectful computations}
 \label{fig:applications-abs}
@@
 The variable context of such expression will be annotated with a set of required resources, 
-\ie~$\coctx{\Gamma}{ \aclrd{\ident{\{ gps, time \}}} }$.
+\ie~$\coctx{\Gamma}{ \cclrd{\ident{\{ gps, time \}}} }$.
 
@@
 site (lexical scope) where the function is declared, which is attached to the current context
-as $\coctx{\Gamma}{ \aclrd{\ident{\{ gps \}}} }$. The current time is required from the caller
+as $\coctx{\Gamma}{ \cclrd{\ident{\{ gps \}}} }$. The current time is required from the caller
 of the function. So, the context requirement on the call site (dynamic scope) will be
-$\aclrd{r}=\aclrd{\footnotesize\ident{\{ time \}}}$. In coeffect systems, we attach this information
-to the function, writing $\tau_1 \xrightarrow{ \aclrd{r} } \tau_2$.
+$\cclrd{r}=\cclrd{\footnotesize\ident{\{ time \}}}$. In coeffect systems, we attach this information
+to the function, writing $\tau_1 \xrightarrow{ \cclrd{r} } \tau_2$.
 
@@
   {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\aclrd{\emptyset}} \vdash x : \tau }
+  {\coctx{\Gamma}{\cclrd{\emptyset}} \vdash x : \tau }
 \end{equation*}
@@
   {}
-  {\coctx{\Gamma}{ \{\ident{?param}:\tau \} } \vdash \ident{?param} : \tau }
+  {\coctx{\Gamma}{\cclrd{ \{\ident{?param}:\tau \} }} \vdash \ident{?param} : \tau }
 \end{equation*}
@@
 \tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \subseteq \aclrd{r})
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \subseteq \cclrd{r})
 \end{equation*}
@@
 \tyrule{app}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
-   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\aclrd{r} \cup \aclrd{s} \cup \aclrd{t}} \vdash e_1~e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \cup \cclrd{s} \cup \cclrd{t}} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 \tyrule{let}
-  { \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x:\tau_1}{\aclrd{s}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma}{\aclrd{r \cup s}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r \cup s}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
 \end{equation*}
@@
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
-\end{equation*}
-\vspace{-0.9em}
-\caption{Coeffect rules for tracking implicit parameters}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{letdyn}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r \cup (s \setminus \{ \ident{?p}:\tau_1 \})}} \vdash \kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+
+\figcaption{Coeffect rules for tracking implicit parameters}
 \label{fig:applications-flat-impl}
@@
 Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's 
-implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau$
+implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
 meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set 
-of implicit parameters specified by $\aclrd{r}$. The annotations $\aclrd{r},\aclrd{s},\aclrd{t}$ are 
-finite partial functions mapping implicit parameter names to types, \ie~$\aclrd{r},\aclrd{s},\aclrd{t} \subseteq 
+of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are 
+finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq 
 \ident{Names} \mapsto \ident{Types}$. The expressions include \ident{?param} to read implicit
@@
 annotated with the set of implicit parameters that must be available on the call-site, \ie~
-$\tau_1 \xrightarrow{\aclrd{s}} \tau_2$.
+$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.
 
@@
 (\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
-the body $\aclrd{r} \cup \aclrd{s}$ can be freely split between the declaration-site ($\coctx{\Gamma}{\aclrd{r}}$)
-and the call-site ($\tau_1 \xrightarrow{\aclrd{s}} \tau_2$).
+the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration-site ($\coctx{\Gamma}{\cclrd{r}}$)
+and the call-site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
+parameter and removes it from the set of requirements.
 
@@
 $\ident{?a} + \ident{?b}$. Assuming that the function takes and returns $\ident{int}$, the following
-list shows 4 out of 9 possible valid typing. Full typing derivations can be found in Appendix~?:
-%
+list shows 4 out of 9 possible valid typing. Full typing derivations can be found in Appendix~\ref{sec:appendixa-implicit}:
+%
+\label{pg:applications-flat-paramsex}
 \begin{equation*}
 \begin{array}{rcllllr}
-\coctx{\Gamma}{\aclrd{ \{ \ident{?a}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
-  \ident{int} \xrightarrow{ \{ \ident{?b}:\ident{int} \} } \ident{int} &\qquad\qquad&(1) \\
-\coctx{\Gamma}{\aclrd{ \{ \ident{?b}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
-  \ident{int} \xrightarrow{ \{ \ident{?a}:\ident{int} \} } \ident{int} &&(2)\\
-\coctx{\Gamma}{\aclrd{ \{\ident{?a}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
-  \ident{int} \xrightarrow{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} } \ident{int} &&(3)\\
-\coctx{\Gamma}{\aclrd{ \emptyset }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
-  \ident{int} \xrightarrow{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} } \ident{int} &&(4)
-\end{array}
-\end{equation*}
-%
-The first two examples demonstrate why the system does not have the principal typing property. 
+\coctx{\Gamma}{\cclrd{ \{ \ident{?a}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?b}:\ident{int} \} }} \ident{int} &\qquad\qquad&(1) \\
+\coctx{\Gamma}{\cclrd{ \{ \ident{?b}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?a}:\ident{int} \} }} \ident{int} &&(2)\\
+\coctx{\Gamma}{\cclrd{ \{\ident{?a}:\ident{int} \} }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{\cclrd{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} }} \ident{int} &&(3)\\
+\coctx{\Gamma}{\cclrd{ \emptyset }} &\vdash& \lambda x.\ident{?a} + \ident{?b} &:& 
+  \ident{int} \xrightarrow{\cclrd{ \{\ident{?a}:\ident{int}, \ident{?b}:\ident{int} \} }} \ident{int} &&(4)
+\end{array}
+\end{equation*}
+%
+The first two examples demonstrate that the system does not have the principal typing property. 
 Both ($1$) and ($2$) are valid typings and they may both be desirable in certain contexts where
@@
 
-In ($3$), the parameter \ident{?a} has to be provided from both the declaration-site and call-site.
-We describe system that supports dynamic rebinding, meaning that when the caller provides a 
-value, it hides the value that may be available from the declaration-site. This means that 
-$4$ is a more precise typing modelling the same situation.
+The next typing derivation ($3$) requires the parameter \ident{?a} from both the declaration-site and 
+the call-site. This means that, at runtime, two values will be available. Our semantics for the 
+system describes \emph{dynamic rebinding}, meaning that when the caller provides a value for a 
+parameter that is already specified by the declaration-site, the new value hides the old one. This
+means that only the value from the call-site is actually used. This ($4$) gives a more precise
+typing for this situation.
 
@@
 \tag{\emph{var}}
-\sem{\coctx{\Gamma}{\aclrd{r}} \vdash x_i : \tau_i} = \lambda ((x_1, \ldots, x_n), \_) \rightarrow x_i
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash x_i : \tau_i} = \lambda ((x_1, \ldots, x_n), \_) \rightarrow x_i
 \end{equation}
@@
 \tag{\emph{param}}
-\sem{\coctx{\Gamma}{\aclrd{r}} \vdash ?p : \sigma} = \lambda (\_, f) \rightarrow f~?p
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash ?p : \sigma} = \lambda (\_, f) \rightarrow f~?p
 \end{equation}
@@
 \tag{\emph{sub}}
-\sem{\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau} = \lambda (x, f) \rightarrow 
-  \sem{\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau}~(x, \restr{f}{\aclrd{r'}})
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau} = \lambda (x, f) \rightarrow 
+  \sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}~(x, \restr{f}{\cclrd{r'}})
 \end{equation}
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\[-0.25em]
-  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{r}\cup\aclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), f \uplus g)   	 
+  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\[-0.25em]
+  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{r}\cup\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), f \uplus g)   	 
 \end{array}
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}\cup\aclrd{s}\cup\aclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 }~(x, \restr{f}{\aclrd{r}})\\[-0.25em]
-  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }~(x, \restr{f}{\aclrd{s}} ), \restr{f}{\aclrd{t}})
-\end{array}
-\end{equation*}
+  \sem{\coctx{\Gamma}{\cclrd{r}\cup\cclrd{s}\cup\cclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~(x, \restr{f}{\cclrd{r}})\\[-0.25em]
+  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~(x, \restr{f}{\cclrd{s}} ), \restr{f}{\cclrd{t}})
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation}
+\tag{\emph{letdyn}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{r \cup (s \setminus \{ \ident{?p}:\tau_1 \})}} \vdash \kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 } =
+     \lambda (x, f) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~v=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 }~(x, \restr{f}{\cclrd{r}})\\[-0.25em]
+  \qquad\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2 }~
+     (x, \restr{f}{\cclrd{s} \setminus \cclrd{\{ \ident{?p}:\tau_1 \}}} \uplus \{ \ident{?p} \mapsto v \} )
+\end{array}
+\end{equation}
 
 \vspace{1em}
-{\small Monadic semantics using the reader monads differs as follows:}
-\begin{equation}
-\tag{\emph{rdabs}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
-  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{r}\cup\aclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)   	 
-\end{array}
-\end{equation}
-
-\vspace{1em}
-{\small Where $\uplus$ and $\restr{f}{\aclrd{r}}$ are auxiliary definitions:}
+{\small Here $\uplus$ and $\restr{f}{\cclrd{r}}$ are auxiliary definitions:}
 \begin{equation*}
 \begin{array}{rcl}
- \restr{f}{\aclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \aclrd{r} \}\\[-0.25em]
+ \restr{f}{\cclrd{r}} &=& \{ (p,v) \;|\; (p,v) \in f, \;p \in \cclrd{r} \}\\[-0.25em]
  f \uplus g &=& \restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g 
@@
 
-\caption{Semantics of a language with implicit parameters}
+\figcaption{Semantics of a language with implicit parameters}
 \label{fig:applications-flat-implsem}
@@
 to the implicit parameters. Accessing a parameter then becomes a lookup in the dictionary and
-the \kvd{letdyn} construct extends the dictionary. To elucidate how such hidden dictionaries 
+the new \kvd{letdyn} construct extends the dictionary. To elucidate how such hidden dictionaries 
 are propagated through the program when using lambda abstractions and applications, we present a 
@@
 
-For simplicity, we assume that all implicit parameters have the same type $\sigma$. In that setting, 
-coeffect annotations $\aclrd{r}$ are just sets of names, \ie~$\aclrd{r}, \aclrd{s}, \aclrd{t} \subseteq \ident{Names}$.
+For simplicity, we assume that all implicit parameters have a type $\sigma$. In that setting, 
+coeffect annotations $\cclrd{r}$ are just sets of names, \ie~$\cclrd{r}, \cclrd{s}, \cclrd{t} \subseteq \ident{Names}$.
 Given an expression $e$ of type $\tau$ that requires free variables $\Gamma$ and implicit parameters 
-$\aclrd{r}$, our interpretation is a function that takes a product of variables from $\Gamma$ together 
-with a hidden dictionary of implicit parameters and returns $\tau$:
-%
-\begin{equation*}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{r} } \vdash e : \tau} 
-  ~:~ (\tau_1 \times \ldots \times \tau_n) \times (\aclrd{r} \rightarrow \sigma) \rightarrow \tau
-\end{equation*}
-%
-The hidden dictionary is represented as a function from $\aclrd{r}$ to $\sigma$. This means that it 
+$\cclrd{r}$, the semantics is a function that takes a product of variables from $\Gamma$ together 
+with a dictionary of implicit parameters and returns $\tau$:
+
+\begin{equation*}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau} 
+  ~:~ (\tau_1 \times \ldots \times \tau_n) \times (\cclrd{r} \rightarrow \sigma) \rightarrow \tau
+\end{equation*}
+%
+The dictionary is represented as a function from $\cclrd{r}$ to $\sigma$. This means that it 
 provides a $\sigma$ value for all implicit parameters that are required according to the typing.
 Note that the domain of the function is not the set of all possible implicit parameter names, but
-only a set of those that are specified by the type system.
-
-A hidden dictionary also needs to be attached to all functions. A function $\tau_1 \xrightarrow{\aclrd{s}} \tau_2$
+only the finite subset of names that are required according to the typing.
+
+The dictionary is also attached to the inputs of all functions. That is, a function $\tau_1 \xrightarrow{\cclrd{s}} \tau_2$
 is interpreted by a function that takes $\tau_1$ together with a dictionary that defines values for
-implicit parameters in $\aclrd{s}$:
-%
-\begin{equation*}
-\begin{array}{l}
-\sem{\tau_1 \xrightarrow{\aclrd{s}} \tau_2} = \tau_1 \times (\aclrd{s} \rightarrow \sigma) \rightarrow \tau_2
-\end{array}
-\end{equation*}
-%
-The definition of the semantics is shown in Figure~\ref{fig:applications-flat-implsem}. Let binding
-can be viewed as a syntactic sugar for $(\lambda x.e_2)~e_1$ and so it is omitted for brevity. The 
+implicit parameters in $\cclrd{s}$:
+%
+\begin{equation*}
+\begin{array}{l}
+ \sem{\tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \tau_1 \times (\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+The definition of the semantics is shown in Figure~\ref{fig:applications-flat-implsem}. The 
 (\emph{var}) and (\emph{param}) rules are simple -- they project the appropriate variable and 
@@
 
-When an expression requires implicit parameters $\aclrd{r}$, the semantics always provides a 
-dictionary defined \emph{exactly} on $\aclrd{r}$. To achieve this, the (\emph{sub}) rule restricts
-the function to $\aclrd{r'}$ (which is valid because $\aclrd{r'} \subseteq \aclrd{r}$).
+When an expression requires implicit parameters $\cclrd{r}$, the semantics always provides a 
+dictionary defined \emph{exactly} on $\cclrd{r}$. To achieve this, the (\emph{sub}) rule restricts
+the function to $\cclrd{r'}$ (which is valid because $\cclrd{r'} \subseteq \cclrd{r}$).
 
@@
 
+Finally, the (\emph{letdyn}) rule specifies the semantics of the \kvd{letdyn} construct, which 
+assigns a value to an implicit parameter. This is similar to (\emph{app}), because it needs to 
+evaluate the sub-expression first. After evaluating $e_1$, the result is added to the dictionary
+using $\uplus$. The semantics of ordinary let binding is omitted, because let binding can be 
+treated as a syntactic sugar for $(\lambda x.e_2)~e_1$.
+
 Without providing a proof here, we note that the semantics is sounds with respect to the type 
@@
 Implicit parameters are related to the \emph{reader mo\-nad}. The type 
-$\tau_1 \times (\aclrd{r} \rightarrow \sigma) \rightarrow \tau_2$ is equivalent to
-$\tau_1 \rightarrow ((\aclrd{r} \rightarrow \sigma) \rightarrow \tau_2)$ through currying. Thus, we can
-express the function as $\tau_1 \rightarrow M\tau_2$ for $M\tau = (\aclrd{r} \rightarrow \sigma) \rightarrow \tau$.
+$\tau_1 \times (\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2$ is equivalent to
+$\tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2)$ through currying. Thus, we can
+express the function as $\tau_1 \rightarrow M\tau_2$ for $M\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau$.
 Indeed, the reader monad can be used to model dynamic scoping. However, there is an important distinction
@@
 
-The (\emph{rdabs}) rule in Figure~\ref{fig:applications-flat-implsem} shows a semantics that
-matches the usual monadic semantics using the reader monad. Note that the declaration-site dictionary
+When using the usualmonadic semantics based on the reader monad, the semantics of the (\emph{abs})
+rule would be modified as follows:
+%
+\begin{equation*}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = 
+      \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
+  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)      
+\end{array}
+\end{equation*}
+%
+Note that the declaration-site dictionary
 is ignored and the body is called with only the dictionary provided by the call-site. This is
-a consequence of the fact that monadic functions are always pure values created using \emph{unit}.
+a consequence of the fact that monadic functions are always pure values created using monadic
+\emph{unit}, which turns a function $\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}$ into a monadic
+computation with no side-effects $\mtyp{\cclrd{\emptyset}}{\tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2}}$.
 
@@
 \emph{type class} constraints \cite{app-type-classes}. They provide a principled form of ad-hoc
-polymorphism (overloading). When a code uses an overloaded operation (e.g.~comparison or numeric 
+polymorphism (overloading). When code uses an overloaded operation (\eg~comparison or numeric 
 operators) a constraint is placed on the context in which the operation is used. For example:
@@
 Similarly to implicit parameters, type classes can be implemented using a hidden dictionary. In 
-the above case, the function \ident{twoTimes} takes a hidden dictionary that provides an operation
+the above case, the function \ident{twoTimes} takes an additional dictionary that provides an operation
 $+$ of type $\alpha \times \alpha \rightarrow \alpha$.
@@
 Implicit parameters are the simplest example of a system where function abstraction does not 
-delay all impurities of the body. As discussed in Section~\ref{sec:applications-structure-lam},
+``delay'' all impurities of the body. Here, the term ``delay'' refers to the fact that some 
+implicit parameters may be captured (from the declaration-site) at the time when the function is 
+defined, but before it is executed. As discussed in Section~\ref{sec:applications-structure-lam},
 this is the defining feature of \emph{coeffect} systems.
@@
 language. In the type system, the (\emph{abs}) rule places context-requirements on both the 
-declaration-site and the call-site. For implicit parameters, this rule introduces non-determinism,
-because the parameters can be split arbitrarily. However, as we show in the next section, this is 
-not always the case. Semantically, lambda abstraction \emph{merges} two parts of context (implicit 
-parameter dictionaries) that are provided by the call-site and declaration-site.
+declaration-site and the call-site. For implicit parameters, this rule introduces non-determinism
+in the type-inference, because the parameters can be split arbitrarily. However, as we show in the 
+next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two 
+parts of context (implicit parameter dictionaries) that are provided by the call-site and 
+declaration-site.
 
@@
 is marshalled and sent to another machine, rebindable resources can be handled in two ways. 
-First, if the resource is available on the target machine, the parameter is \emph{rebound} to
-the resource on the new machine. This is captured by dynamic scoping rules. Second, if the 
+First, if the resource is available on the target machine, the parameter may be \emph{rebound} to
+the resource on the new machine. This is captured by the dynamic scoping rule. Second, if the 
 resource is not available on the target machine, the resource is either marshalled or a \emph{remote 
-reference} is created. This is captured by lexical scoping rules.
+reference} is created. This is captured by the lexical scoping rule.
 
@@
 In the following example, we use the construct $\kvd{access}~\ident{Res}$ to represent
-access to a rebindable resource named $\ident{Res}$. The following simple function accessed
-a database and a current date and filters values based on the date:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{recentEvents} = \lambda () \rightarrow\\[-0.25em]
+access to a rebindable resource named $\ident{Res}$. The following simple function accesses
+a database together with a current date; then it filters from the database based on the date:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{recentNews} = \lambda () \rightarrow\\[-0.25em]
  \quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\[-0.25em]
@@
 %
-When \ident{recentEvents} is created on the server and sent to the client, a remote reference to 
+When \ident{recentNews} is created on the server and sent to the client, a remote reference to 
 the database (available only on the server) must be captured. If the client device supports a 
-clock, then \ident{Clock} can be locally \emph{rebound}, e.g., to accommodate time-zone changes. 
+clock, then \ident{Clock} can be locally \emph{rebound}, \eg, to accommodate time-zone changes. 
 Otherwise, the date and time needs to be obtained from the server too.
@@
 implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword. 
-Lambda abstraction splits the resources non-deterministically between declaration-site 
-(capturing remote reference) and call-site (representing rebinding). For this reason, we do not
-discuss the system in details and instead look at other uses.
+Lambda abstraction splits the requirements between declaration-site (capturing remote reference) 
+and call-site (representing rebinding). For this reason, we do not discuss the system in details
+and instead look at other uses.
 
@@
 \\[0.5em]
-\cmt{// Searches database for a product; can run on the server-side}\\[-0.25em]
+\cmt{// Searches database for a product; must run on the server-side}\\[-0.25em]
 \kvd{let}~\ident{retrieveProduct} = \lambda \ident{name} \rightarrow\\[-0.25em]
@@
 
-\vspace{1em}
-\noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
-\vspace{-1.5em}
-\caption{Sample client-server application with input validation}
+\figcaption{Sample client-server application with input validation}
 \label{fig:applications-flat-distr}
@@
 We demonstrate the problem by looking at input validation. In applications that communicate over 
-unsecured HTTP channel, user input needs to be validated interactively on the client-side (to 
+an unsecured HTTP channel, user input needs to be validated interactively on the client-side (to 
 provide immediate response) and then again on the server-side (to guarantee safety).
@@
 
-{\small a.) Set based type system for cross-compilation, inspired by Links \cite{app-distributed-links}}
+{\small a.) Set-based type system for cross-compilation, inspired by Links \cite{app-distributed-links}}
 
@@
 \tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \supseteq \aclrd{r})
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \supseteq \cclrd{r})
 \end{equation*}
@@
 \tyrule{app}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
-   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\aclrd{r} \cap \aclrd{s} \cap \aclrd{t}} \vdash e_1~e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \cap \cclrd{s} \cap \cclrd{t}} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
 \end{equation*}
@@
 
-{\small b.) Version number based type system, inspired by Android API level \cite{app-android-multitarget}}
+{\small b.) Version-based type system, inspired by Android API level \cite{app-android-multitarget}}
 
@@
 \tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \leq \aclrd{r})
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \leq \cclrd{r})
 \end{equation*}
@@
 \tyrule{app}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
-   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\ident{max} \{\aclrd{r}, \aclrd{s}, \aclrd{t} \}} \vdash e_1~e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\ident{max} \{\cclrd{r}, \cclrd{s}, \cclrd{t} \}} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{r}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 }
-\end{equation*}
-
-\caption{Two variants of coeffect typing rules for cross-compilation}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
+\end{equation*}
+
+\figcaption{Two variants of coeffect typing rules for cross-compilation}
 \label{fig:applications-flat-cross}
@@
 
-Cross-compilation may seem similar to the tracking of resources (and thus to the tracking of implicit 
+Cross-compilation may seem similar to resource tracking (and thus to the tracking of implicit 
 parameters), but it actually demonstrates a couple of new ideas that are important for flat coeffect 
@@
 In the first system, shown in Figure~\ref{fig:applications-flat-cross} (a), the coeffect annotations
-are sets of execution environments, \ie~$\aclrd{r}, \aclrd{s}, \aclrd{t} \subseteq \{ \ident{client},
+are sets of execution environments, \ie~$\cclrd{r}, \cclrd{s}, \cclrd{t} \subseteq \{ \ident{client},
 \ident{server}, \ident{database} \}$. Sub-coeffecting (\emph{sub}) lets us ignore some of the supported 
@@
 Sub-coeffecting and application are trivially dual to the rules for implicit parameters. We just track
-supported environments using intersection as opposed to tracking of required parameters using union.
+supported environments using intersection as opposed to tracking required parameters using union.
 However, this symmetry does not hold for lambda abstraction (\emph{abs}), which still uses \emph{union}.
-This models the case where the function can be executed in two different ways:
+This models the case when there are two ways of executing the function:
 %
 \begin{itemize}
-\item[--] The function is represented as executable code for an environment available at the call-site
+\item[--] The function is represented as executable code for an call-site environment
   and is executed there, possibly after it is marshalled and transferred to another machine.
 \vspace{-0.5em}
-\item[--] The function body is compiled for the environment available at the declaration-site; the value
+\item[--] The function body is compiled for the declaration-site environment; the value
   that is returned is a remote reference to the code and function calls are performed as remote invocations.
@@
 This example ignores important considerations -- for example, it is likely desirable to make this
-difference explicit and the implementation (and semantics) needs to be clarified. However, the
-example shows that the algebraic structure of coeffect annotations may be more complex. Here, using
-$\cap$ for application and $\cup$ for abstraction.
+difference explicit (\eg~using explicit wrapping of unevaluated expressions) and the implementation
+also needs to be clarified. For a system that does this, see \eg~ML5 \cite{app-distributed-ml5}).
+The key point of our brief example is that the algebraic structure of coeffect annotations may be more 
+complex and use, for example, $\cap$ for application and $\cup$ for abstraction.
 
@@
 level requirements in Android. Coeffect annotations are simply numbers representing the level
-($\aclrd{r}, \aclrd{s}, \aclrd{t} \in \mathbb{N}$). Levels are ordered increasingly, so we can
+($\cclrd{r}, \cclrd{s}, \cclrd{t} \in \mathbb{N}$). Levels are ordered increasingly, so we can
 always require higher level (\emph{sub}). The requirement on function application (\emph{app}) is
@@
 
-In addition to the work discussed already, ML5 \cite{app-distributed-ml5} is another important work 
-that looks at tracking of execution environments. It uses modalities of modal S4 to represent the 
-environment -- this approach is similar to coeffects, both from the practical perspective, but also 
-through deeper theoretical links. We return to this topic in Section~\ref{sec:conclusions-meta} of
-the final chapter.
+The ML5 language \cite{app-distributed-ml5} mentioned above served as an inspiration for our example.
+It tracks execution environments using modalities of modal S4 to represent the environment -- this 
+approach is similar to coeffects, both from the practical perspective, but also through deeper 
+theoretical links. However, it is based on the \emph{meta-language} style of embedding modalities
+rather than on the \emph{language-semantics} style (see~Section~\ref{sec:path-sem-langs}). We 
+return to this topic in Section~\ref{sec:unified-meta}.
 
@@
 \label{sec:applications-flat-live}
+Our next example shows the idea of coeffects from a different perspective. Rather than
+keeping additional information independent of the variable context, we track properties about how
+variables are used. Nevertheless, we still look at the left-hand side of $\vdash$ and the structure
+of the typing rules and semantics will be very similar.
 
@@
 \emph{individual} variables, but of the \emph{entire} variable context. This is not practically
-useful, but it provides interesting insight into how flat coeffects work. A per-variable liveness
+useful, but it provides an interesting insight into how flat coeffects work. A per-variable liveness
 analysis can be captured using structural coeffects and is discussed in Section~\ref{sec:applications-struct-live}.
@@
 of the function is marked as \emph{live}. In the flat system, we do not track \emph{which} 
-variable was used and so we have to mark both the call-site and declaration-site as live (this will
-be refined in structural liveness system).
+variable was used and so we have to mark both the call-site and the declaration-site as live (this will
+be refined in a structural version of the system).
 
@@
 \ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
-only if it was used in both of the branches (this is sometimes called \emph{neededness}).
+only if it was used in both of the branches (this is sometimes called \emph{neededness} or
+\emph{very busy} variable/expression).
+
 The distinction between \emph{may} and \emph{must} analyses demonstrates the importance of 
@@
 \begin{array}{llll}
-f : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 \qquad\qquad&
-g : \tau_2 \xrightarrow{\aclrd{s}} \tau_3 \qquad\qquad&
-g \circ f : \tau_1 \xrightarrow{\aclrd{r} \sqcup \aclrd{s}} \tau_3 \qquad\qquad& ~
+f : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 \qquad\qquad&
+g : \tau_2 \xrightarrow{\cclrd{s}} \tau_3 \qquad\qquad&
+g \circ f : \tau_1 \xrightarrow{\cclrd{r} \sqcup \cclrd{s}} \tau_3 \qquad\qquad& ~
 \\[0.75em]
-f : \tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2 &
-g : \tau_2 \xrightarrow{\aclrd{\ident{L}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_3 & (1)
+f : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\ident{L}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_3 & (1)
 \\[-0.10em]
-f : \tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2 &
-g : \tau_2 \xrightarrow{\aclrd{\ident{L}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 & (2)
+f : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\ident{L}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 & (2)
 \\[-0.10em]
-f : \tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2 &
-g : \tau_2 \xrightarrow{\aclrd{\ident{D}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 & (3)
+f : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\ident{D}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 & (3)
 \\[-0.10em]
-f : \tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2 &
-g : \tau_2 \xrightarrow{\aclrd{\ident{D}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 & (4)
+f : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\ident{D}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 & (4)
 \end{array}
@@
 \begin{array}{llll}
-f : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 \qquad\qquad&
-h : \tau_1 \xrightarrow{\aclrd{s}} \tau_3 \qquad\qquad&
-\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{r} \sqcap \aclrd{s}} \tau_2 \times \tau_3 \qquad&~
+f : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 \qquad\qquad&
+h : \tau_1 \xrightarrow{\cclrd{s}} \tau_3 \qquad\qquad&
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{r} \sqcap \cclrd{s}} \tau_2 \times \tau_3 \qquad&~
 \\[0.75em]
-f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 &
-h : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 \times \tau_3 & (1)
+f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 \times \tau_3 & (1)
 \\[-0.10em]
-f : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 &
-h : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 \times \tau_3 & (2)
+f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 \times \tau_3 & (2)
 \\[-0.10em]
-f : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 &
-h : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 \times \tau_3 & (3)
+f : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 \times \tau_3 & (3)
 \\[-0.10em]
-f : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 &
-h : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 \times \tau_3 & (4)
-\end{array}
-\end{equation*}
-
-\caption{Liveness annotations with sequential and pointwise composition}
+f : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 \times \tau_3 & (4)
+\end{array}
+\end{equation*}
+
+\figcaption{Liveness annotations with sequential and pointwise composition}
 \label{fig:applications-flat-livealg}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma}{\cclrd{\ident{L} }} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c : \tau \in \Delta}
+  {\coctx{\Gamma}{ \cclrd{\ident{D}} } \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \sqsubseteq \cclrd{r})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{t})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
+\end{equation*}
+\vspace{-0.9em}
+
+\figcaption{Coeffect rules for tracking whole-context liveness}
+\label{fig:applications-flat-liveness}
+\vspace{-1.2em}
 \end{figure}
@@
 context as live using \ident{L}. A dead context (definitely not needed) can be treated as live context
-(which may be used) using the (\emph{sub}) rule. This captures the \emph{may} nature of the analysis.
+using the (\emph{sub}) rule. This captures the \emph{may} nature of the analysis.
 
@@
 Consider first \emph{sequential composition} of (semantic) functions $f, g$ annotated with 
-$\aclrd{r}, \aclrd{s}$. The composed function $g \circ f$ is annotated with $\aclrd{r} \sqcup \aclrd{s}$
+$\cclrd{r}, \cclrd{s}$. The composed function $g \circ f$ is annotated with $\cclrd{r} \sqcup \cclrd{s}$
 as shown in Figure~\ref{fig:applications-flat-livealg} (b).
@@
 As further discussed later in Section~?, the (\emph{let}) rule is again just a syntactic sugar for 
-$(\lambda x.e_2)~e_1$. Briefly, this follows from the simple observation that 
-$\aclrd{r} \sqcup (\aclrd{s} \sqcap \aclrd{r}) = \aclrd{r}$.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\aclrd{\ident{L} }} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{const}
-  {c : \tau \in \Delta}
-  {\coctx{\Gamma}{ \aclrd{\ident{D}} } \vdash c : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau }\quad\quad(\aclrd{r'} \sqsubseteq \aclrd{r})
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 &
-   \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\aclrd{r} \sqcup (\aclrd{s} \sqcap \aclrd{t})} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x:\tau_1}{\aclrd{s}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma}{\aclrd{s}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{r}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{r}} \tau_2 }
-\end{equation*}
-\vspace{-0.9em}
-
-\caption{Coeffect rules for tracking whole-context liveness}
-\label{fig:applications-flat-liveness}
-\vspace{-1.2em}
-\end{figure}
+$(\lambda x.e_2)~e_1$. This follows from the simple observation that 
+$\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{r}) = \cclrd{r}$.
 
@@
 Before looking at the semantics, we consider a number of simple examples to demonstrate the
-key aspects of the system. Full typing derivations are shown in Appendix~?:
+key aspects of the system. Full typing derivations are shown in Appendix~\ref{sec:appendixa-liveness}:
 %
@@
 \begin{array}{lll}
-(\lambda x \rightarrow 42)~y &~\hspace{1em}~&(1)\\
+(\lambda x . 42)~y &~\hspace{1em}~&(1)\\
 \ident{twoTimes}~42          &&(2)\\
-(\lambda x \rightarrow x)~42 &&(3)\\
+(\lambda x . x)~42 &&(3)\\
 \end{array}
@@
 
-\begin{figure}
-\begin{equation}
-\tag{\emph{var}}
-\sem{\coctx{\Gamma}{\aclrd{ \ident{L} }} \vdash x_i : \tau_i} = \lambda (x_1, \ldots, x_n) \rightarrow x_i
-\end{equation}
-\vspace{-1.75em}
-\begin{equation}
-\tag{\emph{const}}
-\sem{\coctx{\Gamma}{\aclrd{ \ident{D} }} \vdash c_i : \tau_i} = \lambda () \rightarrow \delta(c_i)
-\end{equation}
-%
-%
-\vspace{-0.5em}
-%
-%
-\begin{equation}
-\tag{\emph{sub-1}}
-\sem{\coctx{\Gamma}{\aclrd{\ident{L} }} \vdash e : \tau} = \lambda x \rightarrow 
-  \sem{\coctx{\Gamma}{\aclrd{\ident{D} }} \vdash e : \tau}~()
-\end{equation}
-\vspace{-1.75em}
-\begin{equation}
-\tag{\emph{sub-2}}
-\sem{\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau} = \lambda x \rightarrow 
-  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash e : \tau}~x
-\end{equation}
-%
-%
-\vspace{-0.5em}
-%
-%
-\begin{equation}
-\tag{\emph{abs-1}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{\ident{L}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2 } = \lambda (x_1, \ldots, x_n) \rightarrow\\[-0.25em]
-  \qquad\lambda y \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{\ident{L}}} \vdash e : \tau_2 }~(x_1, \ldots, x_n, y)   	 
-\end{array}
-\end{equation}
-\vspace{-1.0em}
-\begin{equation}
-\tag{\emph{abs-2}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{\ident{D}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2 } = \lambda () \rightarrow\\[-0.25em]
-  \qquad\lambda () \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{\ident{D}}} \vdash e : \tau_2 }~()
-\end{array}
-\end{equation}
-%
-%
-\vspace{0.0em}
-%
-%
-%
-\begin{equation*}
-\tag{\emph{app-1}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{ \ident{D} }} \tau_2 }~x
-    ~\kvd{in}~g~()
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\tag{\emph{app-2}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{\ident{L}}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{ \ident{L} }} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{ \ident{L} }} \tau_2 }~x
-    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{ \ident{D} }} \vdash e_2 : \tau_1 }~())
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\tag{\emph{app-3}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{r} \sqcup (\aclrd{s} \sqcap \aclrd{t}) } \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{t}} \tau_2 }~x
-    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\aclrd{s}} \vdash e_2 : \tau_1 }~x)
-\end{array}
-\end{equation*}
-
-\caption{Semantics of a language with liveness analysis}
-\label{fig:applications-flat-livsem}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \paragraph{Semantics.} 
-The type system presented above requires the semantics to \emph{implement} dead code elimination.
-This means that when a function does not require an input (it is marked as dead), the semantics
-does not evaluate the argument and passes an empty value as the input instead.
+As showed in the examples, the type system for the liveness coeffect calculus marks the context 
+of an expression $(\lambda x.42)~y$ as dead. This means that the semantics of the above expression
+must not evaluate the argument $y$. In other words, the type system is only sound if the semantics
+includes dead code elimination.
+
+To capture dead code elimination in the semantics, we add a special empty value and pass it as an
+argument to a function whose argument is not needed, so $(\lambda x.42)$ will be called with 
+an empty value as argument (because it does not need its argument).
 
@@
 representation depends on the annotation -- in other words, the type is \emph{indexed} by 
-the coeffect annotation. The indexing is discussed in details in Section~X. For now, it suffices
-to define the semantics using two separate rules:
+the coeffect annotation. The indexing is discussed in details in Section~\ref{sec:flat-semantics-idx}. 
+For now, it suffices to define the semantics using two separate rules:
 %
@@
 \begin{array}{rlrcl}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{\ident{L}} } \vdash e : \tau} 
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{\ident{L}} } \vdash e : \tau} 
   &:& (\tau_1 \times \ldots \times \tau_n) &\narrow{\rightarrow}& \tau\\
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{\ident{D}} } \vdash e : \tau} 
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{\ident{D}} } \vdash e : \tau} 
   &:& 1 &\narrow{\rightarrow}& \tau
@@
 \begin{array}{l}
-\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2\\
-\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
+\sem{\tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2\\
+\sem{\tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
 \end{array}
@@
 using monads. Any monadic semantics would express functions as $\tau_1 \rightarrow M\, \tau_2$.
-Unless laziness is already built-in in the semantics, there is no way to call such function without
+Unless laziness is already built-in, there is no way to call such function without
 first obtaining a value $\tau_1$. The above semantics makes this possible by taking a unit $1$ when
@@
 the only value of type $1$. This appears, for example, in (\emph{const}) which takes $()$ as the
-input and returns constant using a global dictionary $\delta$. In (\emph{var}), the context is live
+input and returns a constant using a global dictionary $\delta$. In (\emph{var}), the context is live
 and so the semantics performs a projection. Sub-coeffecting is captured by two rules. A dead context 
@@
 
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation}
+\tag{\emph{var}}
+\sem{\coctx{\Gamma}{\cclrd{ \ident{L} }} \vdash x_i : \tau_i} = \lambda (x_1, \ldots, x_n) \rightarrow x_i
+\end{equation}
+\vspace{-1.75em}
+\begin{equation}
+\tag{\emph{const}}
+\sem{\coctx{\Gamma}{\cclrd{ \ident{D} }} \vdash c_i : \tau_i} = \lambda () \rightarrow \delta(c_i)
+\end{equation}
+%
+%
+\vspace{-0.5em}
+%
+%
+\begin{equation}
+\tag{\emph{sub-1}}
+\sem{\coctx{\Gamma}{\cclrd{\ident{L} }} \vdash e : \tau} = \lambda x \rightarrow 
+  \sem{\coctx{\Gamma}{\cclrd{\ident{D} }} \vdash e : \tau}~()
+\end{equation}
+\vspace{-1.75em}
+\begin{equation}
+\tag{\emph{sub-2}}
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau} = \lambda x \rightarrow 
+  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}~x
+\end{equation}
+%
+%
+\vspace{-0.5em}
+%
+%
+\begin{equation}
+\tag{\emph{abs-1}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 } = \lambda (x_1, \ldots, x_n) \rightarrow\\[-0.25em]
+  \qquad\lambda y \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{\ident{L}}} \vdash e : \tau_2 }~(x_1, \ldots, x_n, y)     
+\end{array}
+\end{equation}
+\vspace{-1.0em}
+\begin{equation}
+\tag{\emph{abs-2}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{\ident{D}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 } = \lambda () \rightarrow\\[-0.25em]
+  \qquad\lambda () \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{\ident{D}}} \vdash e : \tau_2 }~()
+\end{array}
+\end{equation}
+%
+%
+\vspace{0.0em}
+%
+%
+%
+\begin{equation*}
+\tag{\emph{app-1}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 }~x
+    ~\kvd{in}~g~()
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\tag{\emph{app-2}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{ \ident{L} }} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 }~x
+    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{ \ident{D} }} \vdash e_2 : \tau_1 }~())
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\tag{\emph{app-3}}
+\hspace{-0.5em}
+\begin{array}{l}
+  \sem{\coctx{\Gamma}{\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{t}) } \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~x
+    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~x)
+\end{array}
+\end{equation*}
+
+\figcaption{Semantics that implements dead code elimination for $\lambda$-calculus}
+\label{fig:applications-flat-livsem}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
 \paragraph{Summary.}
-Unlike with implicit parameters, the lambda abstraction for liveness analysis does not introduce 
+Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce 
 non-determinism. It simply duplicates the context requirements. However, this still matches the
-property of coeffects that impurities cannot be thunked.
-
-The semantics of liveness reveals a number of interesting properties too. Firstly, the semantics
-cannot be captured by any monad. Secondly, the system would not work without the coeffect annotations. 
+property of coeffects that impurities cannot be delayed or thunked and attached just to the 
+function arrow -- we place requirements on both call-site and declaration-site.
+
+The semantics of liveness reveals three interesting properties. Firstly, the coeffect calculus for
+liveness cannot be modelled as a monadic computation of the form $\tau_1 \rightarrow \mtyp{}{\tau_2}$.
+Secondly, the system would not work without the coeffect annotations. 
 The shape of the semantic function depends on the annotation (the input is either $1$ or $\tau$) and 
-is \emph{indexed} by the annotation. Finally, we discussed at length how the semantics of application
-arises from \emph{sequential} and \emph{pointwise} composition. This is another important aspect of
-coeffect systems -- categorical semantics typically builds on \emph{sequential} composition, but to
-model full $\lambda$ calculus it needs more. For coeffect systems, we need \emph{pointwise} composition
-where the same context is shared by multiple sub-expressions.
+is \emph{indexed} by the annotation.
+
+Finally, we discussed how the semantics of application arises from \emph{sequential} and 
+\emph{pointwise} composition. This is an important aspect of coeffect systems -- categorical 
+semantics typically builds on \emph{sequential} composition, but to model full $\lambda$ calculus 
+it needs more. For coeffects, we need \emph{pointwise} composition where the same context 
+is shared by multiple sub-expressions.
 
@@
 \label{sec:applications-flat-dataflow}
+
+We used implicit parameters as our first example, because they show the simplest form of coeffects.
+Liveness requires a richer coeffect annotation structure, but the flat version is not practical.
+In this section, we look at a system with a structure similar to liveness that is not a toy example.
 
@@
 paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
-and implements program to count the number of edges on a Boolean stream:
+and implements a program to count the number of edges on a Boolean stream:
 %
@@
 first element of $y$ and the remaining elements are values of $x$. Note that in Lucid, the constants
-such as \ident{false} and $0$ are constant streams. Formally, the constructs are defined as follows
-(writing $x_n$ for $n$-th element of a stream $x$):
+such as \ident{false} and $0$ are constant streams.
+
+Formally, the constructs are defined as follows (writing $x_n$ for $n$-th element of a stream $x$):
 %
@@
 The semantics of Lucid-like languages can be captured using a number of mathematical 
-structures. Wadge \cite{app-lucid-monads} originally proposed to use monads, while Uustalu and 
-Vene later used comonads \cite{app-dataflow-essence}. In Chapter~\ref{ch:flat}, we extend
-the latter approach. However, the present chapter presents a sketch of a data-flow semantics
+structures. Wadge \cite{app-lucid-monads} originally defined a monadic semantics, while Uustalu 
+and Vene later used comonads \cite{app-dataflow-essence}. In Chapter~\ref{ch:flat}, we extend
+the latter approach. The present chapter presents a sketch of a concrete data-flow semantics
 defined directly on streams.
@@
   {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\aclrd{ 0 }} \vdash x : \tau }
+  {\coctx{\Gamma}{\cclrd{ 0 }} \vdash x : \tau }
 \end{equation*}
@@
 \tyrule{prev}
-  {\coctx{\Gamma}{ \aclrd{n} } \vdash e : \tau}
-  {\coctx{\Gamma}{ \aclrd{n}+1} \vdash \kvd{prev}~e : \tau}
+  {\coctx{\Gamma}{ \cclrd{n} } \vdash e : \tau}
+  {\coctx{\Gamma}{ \cclrd{n}+1} \vdash \kvd{prev}~e : \tau}
 \end{equation*}
@@
 \tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{n'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau }\quad\quad(\aclrd{n'} \leq \aclrd{n})
+  {\coctx{\Gamma}{\cclrd{n'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }\quad\quad(\cclrd{n'} \leq \cclrd{n})
 \end{equation*}
@@
 \tyrule{app}
-  {\coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 &
-   \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n} + \aclrd{p})} \vdash e_1~e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{p}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m}, \cclrd{n} + \cclrd{p})} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 \tyrule{let}
-  { \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma}{\aclrd{n}+\aclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+  { \coctx{\Gamma}{\cclrd{m}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x:\tau_1}{\cclrd{n}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{n}+\cclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
 \end{equation*}
@@
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
-\end{equation*}
-
-\caption{Coeffect rules for tracking context-usage in data-flow language}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{n}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 }
+\end{equation*}
+
+\figcaption{Coeffect rules for tracking context-usage in data-flow language}
 \label{fig:applications-flat-dataflow}
@@
 \paragraph{Type system.}
-A type system that tracks the maximal number of accessed past values annotates the context with
-a single integer. The current value is always present, so $0$ means that no past values are 
-needed, but the current value is still available. The typing rules of the system are shown
-in Figure~\ref{fig:applications-flat-dataflow}.
+We can use a coeffect type system to track the maximal number of accessed past values. Here,
+the context is annotated with a single integer. The current value is always present, so $0$ means 
+that no past values are needed, but the current value is still available. The typing rules of 
+the system are shown in Figure~\ref{fig:applications-flat-dataflow}.
 
@@
 \begin{array}{lll}
-f : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 \qquad\qquad&
-g : \tau_2 \xrightarrow{\aclrd{n}} \tau_3 \qquad\qquad&
-g \circ f : \tau_1 \xrightarrow{\aclrd{m} + \aclrd{n}} \tau_3 \qquad\qquad
+f : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 \qquad\qquad&
+g : \tau_2 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
+g \circ f : \tau_1 \xrightarrow{\cclrd{m} + \cclrd{n}} \tau_3 \qquad\qquad
 \\
-f : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 \qquad\qquad&
-h : \tau_1 \xrightarrow{\aclrd{n}} \tau_3 \qquad\qquad&
-\langle f, h \rangle : \tau_1 \xrightarrow{\textnormal{max}(\aclrd{m}, \aclrd{s})} \tau_2 \times \tau_3 \qquad
-\end{array}
-\end{equation*}
-%
-Sequential composition adds the annotations. The function $f$ needs $\aclrd{m}$ past values to 
-produce a single $\tau_2$ value. To produce two $\tau_2$ values, we thus need $\aclrd{m}+1$ past
-values of $\tau_1$; to produce three $\tau_2$ values, we need $\aclrd{m}+2$ past values of $\tau_1$,
+f : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 \qquad\qquad&
+h : \tau_1 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
+\langle f, h \rangle : \tau_1 \xrightarrow{\textnormal{max}(\cclrd{m}, \cclrd{s})} \tau_2 \times \tau_3 \qquad
+\end{array}
+\end{equation*}
+%
+Sequential composition adds the annotations. The function $f$ needs $\cclrd{m}$ past values to 
+produce a single $\tau_2$ value. To produce two $\tau_2$ values, we thus need $\cclrd{m}+1$ past
+values of $\tau_1$; to produce three $\tau_2$ values, we need $\cclrd{m}+2$ past values of $\tau_1$,
 and so on. To produce $\ident{n}$ past values that are required as the input of $g$, we need
-$\aclrd{m}+\aclrd{n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
-the same stream to evaluate functions requiring $\aclrd{m}$ and $\aclrd{n}$ past values, and so it
+$\cclrd{m}+\cclrd{n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
+the same stream to evaluate functions requiring $\cclrd{m}$ and $\cclrd{n}$ past values, and so it
 needs maximum of the two at most.
@@
 In function abstraction (\emph{abs}), the requirements of the body are duplicated on the declaration-site
-and the call-site as in liveness analysis. If the body requires $\aclrd{n}$ past values, it may access
-$\aclrd{n}$ values of any variables -- including those available in $\Gamma$, as well as the parameter
+and the call-site as in liveness analysis. If the body requires $\cclrd{n}$ past values, it may access
+$\cclrd{n}$ values of any variables -- including those available in $\Gamma$, as well as the parameter
 $x$. Finally, the (\emph{let}) rule simply adds the two requirements. This corresponds to the sequential
@@
 is a stream of positive integers (starting from zero) and \ident{tick} flips between $0$ and $1$.
-The full typing derivation is shown in Appendix~?:
+The full typing derivation is shown in Appendix~\ref{sec:appendixa-dataflow}:
 %
@@
 returned function needs either none or one past value (thus a subtyping rule is required to type 
-it as requiring one past value). So, the annotations for (\emph{app}) are $\aclrd{m}=1, \aclrd{p}=1$.
+it as requiring one past value). So, the annotations for (\emph{app}) are $\cclrd{m}=1, \cclrd{p}=1$.
 The function is called with $\kvd{prev}~\ident{counter}$ as an argument, meaning that the result
@@
 is $\langle{\ldots,5,4,3,2}\rangle$ and so the overall result is a stream $\langle{\ldots,5,5,3,3}\rangle$.
-From the argument, we get the requirement $\aclrd{n}=1$.
+From the argument, we get the requirement $\cclrd{n}=1$.
 
@@
 \hspace{-1em}
-\sem{\coctx{\Gamma}{\aclrd{0}} \vdash x_i : \tau_i} = \lambda \langle(x_0, \ldots, x_n)\rangle \rightarrow x_i
+\sem{\coctx{\Gamma}{\cclrd{0}} \vdash x_i : \tau_i} = \lambda \langle(x_0, \ldots, x_n)\rangle \rightarrow x_i
 \end{equation}
@@
 \begin{array}{l}
-\sem{\coctx{\Gamma}{\aclrd{n}+1} \vdash \kvd{prev}~e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}+1} \rangle \rightarrow\\[-0.1em]
-  \qquad\sem{ \coctx{\Gamma}{ \aclrd{n} } \vdash e : \tau }~\langle \mathbf{v}_1, \ldots, \mathbf{v}_{ \aclrd{n}+1 }\rangle
+\sem{\coctx{\Gamma}{\cclrd{n}+1} \vdash \kvd{prev}~e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}+1} \rangle \rightarrow\\[-0.1em]
+  \qquad\sem{ \coctx{\Gamma}{ \cclrd{n} } \vdash e : \tau }~\langle \mathbf{v}_1, \ldots, \mathbf{v}_{ \cclrd{n}+1 }\rangle
 \end{array}
@@
 \begin{array}{l}
-\sem{\coctx{\Gamma}{\aclrd{n}} \vdash e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}} \rangle \rightarrow\\[-0.1em]
-  \qquad\sem{ \coctx{\Gamma}{ \aclrd{n'} } \vdash e : \tau }~\langle \mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{n'} }\rangle
+\sem{\coctx{\Gamma}{\cclrd{n}} \vdash e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle \rightarrow\\[-0.1em]
+  \qquad\sem{ \coctx{\Gamma}{ \cclrd{n'} } \vdash e : \tau }~\langle \mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{n'} }\rangle
 \end{array}
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\aclrd{n}} \vdash \lambda y. e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 } = 
-    \lambda \langle \mathbf{v}_0, \ldots \mathbf{v}_{\aclrd{n}} \rangle \rightarrow\\[-0.1em]
-  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\aclrd{n}} \vdash e : \tau_2 
-    }~\langle (\mathbf{v}_0, y_0), \ldots, (\mathbf{v}_{ \aclrd{n}  }, y_{ \aclrd{n} } ) \rangle
+  \sem{\coctx{\Gamma}{\cclrd{n}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 } = 
+    \lambda \langle \mathbf{v}_0, \ldots \mathbf{v}_{\cclrd{n}} \rangle \rightarrow\\[-0.1em]
+  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{n}} \vdash e : \tau_2 
+    }~\langle (\mathbf{v}_0, y_0), \ldots, (\mathbf{v}_{ \cclrd{n}  }, y_{ \cclrd{n} } ) \rangle
 \end{array}
@@
 \begin{array}{l}
-  \sem{\coctx{\Gamma}{\textnormal{max}(\aclrd{m}, \aclrd{n}+\aclrd{p})} \vdash e_1~e_2 : \tau_2 } = 
-    \lambda (\mathbf{v}_0, \ldots, \mathbf{v}_{ \textnormal{max}(\aclrd{m}, \aclrd{n}+\aclrd{p}) } ) \rightarrow \\[-0.1em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\aclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\aclrd{p}} \tau_2 
-    }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{m} })\\[-0.1em]
-  \qquad\kvd{in}~g~(~\sem{ \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \aclrd{n} }),\ldots, \\[-0.1em]
-  \qquad\qquad\;~\sem{ \coctx{\Gamma}{\aclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_{ \aclrd{p} }, \ldots, \mathbf{v}_{ \aclrd{n}+\aclrd{p} })~)
-\end{array}
-\end{equation*}
-
-\caption{Semantics of a simple data-flow language}
+  \sem{\coctx{\Gamma}{\textnormal{max}(\cclrd{m}, \cclrd{n}+\cclrd{p})} \vdash e_1~e_2 : \tau_2 } = 
+    \lambda (\mathbf{v}_0, \ldots, \mathbf{v}_{ \textnormal{max}(\cclrd{m}, \cclrd{n}+\cclrd{p}) } ) \rightarrow \\[-0.1em]
+  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{p}} \tau_2 
+    }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{m} })\\[-0.1em]
+  \qquad\kvd{in}~g~(~\sem{ \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{n} }),\ldots, \\[-0.1em]
+  \qquad\qquad\;~\sem{ \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_{ \cclrd{p} }, \ldots, \mathbf{v}_{ \cclrd{n}+\cclrd{p} })~)
+\end{array}
+\end{equation*}
+
+\figcaption{Semantics showing how past values are accessed in a data-flow language}
 \label{fig:applications-flat-dfsem}
@@
 a list of fixed length. As with liveness, this is a data structure \emph{indexed} by the coeffect
-annotation. We write $\tau^{\aclrd{n}}$ for a list containing $\aclrd{n}$ elements (which can be 
-also viewed as an $\aclrd{n}$-element product $\tau \times \ldots \times \tau$).
+annotation. We write $\tau^{\cclrd{n}}$ for a list containing $\cclrd{n}$ elements (which can be 
+also viewed as an $\cclrd{n}$-element product $\tau \times \ldots \times \tau$).
 
@@
 \begin{array}{rcl}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \aclrd{n} } \vdash e : \tau} 
-  &:& (\tau_1 \times \ldots \times \tau_n)^{\aclrd{n}+1} \rightarrow \tau\\
-\sem{\tau_1 \xrightarrow{\aclrd{n}} \tau_2} &:& \tau_1^{\aclrd{n}+1} \rightarrow \tau_2
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{n} } \vdash e : \tau} 
+  &:& (\tau_1 \times \ldots \times \tau_n)^{\cclrd{n}+1} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{n}} \tau_2} &:& \tau_1^{\cclrd{n}+1} \rightarrow \tau_2
 \end{array}
@@
 The rules defining the semantics are shown in Figure~\ref{fig:applications-flat-dfsem}. The
-semantics of the context is a \emph{list of pairs}. To make the rules easier to follow, we write
-$\langle \mathbf{v}_1, \ldots, \mathbf{v}_n \rangle$ for an $n$-element list containing pairs.
-Pairs that model the entire context such as $\mathbf{v}_1$ are written in bold. When we access
+semantics of the context is a \emph{list of products}. To make the rules easier to follow, we write
+$\langle \mathbf{v}_1, \ldots, \mathbf{v}_n \rangle$ for an $n$-element list containing products.
+Products that model the entire context such as $\mathbf{v}_1$ are written in bold. When we access
 individual variables, we write $\mathbf{v} = (x_1, \ldots, x_m)$ where $x_i$ denote individual
@@
 Lambda abstractions (\emph{abs}) receives two lists of the same size -- one containing values of
-the variables from the declaration-site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\aclrd{n}} \rangle$ and one
-containing the argument provided by the call-site $\langle y_0, \ldots, v_{\aclrd{n}} \rangle$.
+the variables (list of products) from the declaration-site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle$ 
+and one containing the argument (list of values) provided by the call-site $\langle y_0, \ldots, y_{\cclrd{n}} \rangle$.
 The semantics applies the well-known \emph{zip} operation on the lists and passes the result to the
@@
 
-Finally, application (\emph{abs}) uses the input context in two ways, which gives rise to the
+Finally, application (\emph{app}) uses the input context in two ways, which gives rise to the
 two requirements combined using \emph{max}. First, it evaluates the expression $e_1$ which is 
-called with the past $\aclrd{m}$ values. The resulting function $g$ is then sequentially composed
+called with the past $\cclrd{m}$ values. The resulting function $g$ is then sequentially composed
 with the semantics of $e_2$. To call the function, we need to evaluate $e_2$ repeatedly -- namely,
-$\aclrd{p}+1$ times, which results in the overall requirement for $\aclrd{n}+\aclrd{p}$ past values.
+$\cclrd{p}+1$ times, which results in the overall requirement for $\cclrd{n}+\cclrd{p}$ past values.
 
 \paragraph{Summary.}
-The most interesting point about the data-flow example is that it is remarkably similar to our
+Type systems have been used in the context of data-flow languages, for example to check
+initialization properties \cite{app-dataflow-init}, but to our knowledge, not for checking
+the maximal number of required past values. Thus this section serves not just as an example, 
+but also shows how coeffects can lead to novel results.
+
+The most interesting point about the data-flow system is that it is remarkably similar to our
 earlier liveness example. In the type system, abstraction (\emph{abs}) duplicates the context
@@
 In the implicit parameters and data-flow examples, the context provides additional resources or 
-values that may be accessed at runtime. However, it may also track \emph{permissions} or 
-\emph{capabilities} to perform some operation. Liveness can be seen as a trivial example -- when
-the context is live, it contains a permission to access variables. In this section, we briefly
-consider a system for safe locking of Flanagan and Abadi \cite{app-safe-locking} as one, more
-advanced example. Calculus of capabilities of Cray et al. \cite{app-capabilities} is discussed 
-later in Section~\ref{sec:applications-active-ccc}.
+values that may be accessed at runtime. However, coeffects can also track \emph{permissions} or 
+\emph{capabilities} to perform some operation. We can invert the intuition behind liveness and 
+use it as a trivial example -- when the context is live, it contains a \emph{permission} to access 
+variables. In this section, we briefly consider a system for safe locking of Flanagan and 
+Abadi~\cite{app-safe-locking} as one, more advanced example. Calculus of capabilities of 
+Cray et al.~\cite{app-capabilities} is discussed later in Section~\ref{sec:applications-active-ccc}.
 
@@
 The declaration \kvd{newlock} creates a lock $l$ protecting memory region $\rho$. We can than
-allocate mutable variables in that memory region (second line). An access to mutable variable
-is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
+allocate mutable variables in that memory region (second line). An access to one or more mutable 
+variables is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
 which locks a lock and evaluates an expression in a context that contains permission to access
@@
 The type system for safe locking associates a list of acquired locks with the context. 
-Interestingly, the original presentation of the system uses a coeffect-style judgements
-of a form $\Gamma; p \vdash e : \tau$ where $p$ is a list of accessible regions (protected by
-an acquired lock). Using our notation, the rule for \kvd{sync} looks as follows:
-%
+Interestingly, the original presentation of the system by Flanagan and Abadi \cite{app-safe-locking}
+uses a coeffect-style judgements of a form $\Gamma; p \vdash e : \tau$ where $p$ is a list of 
+accessible regions (protected by an acquired lock). Using our notation, the rule for \kvd{sync} 
+looks as follows:
+
 \begin{equation*}
 \tyrule{sync}
-  { \coctx{\Gamma}{\aclrd{p}} \vdash e_1 : m& 
-    \coctx{\Gamma}{\aclrd{p} \cup \{m\} } \vdash e_2 : \tau }
-  { \coctx{\Gamma}{\aclrd{p}} \vdash \kvd{sync}~e_1~e_2 : \tau }
+  { \coctx{\Gamma}{\cclrd{p}} \vdash e_1 : m& 
+    \coctx{\Gamma}{\cclrd{p} \cup \{m\} } \vdash e_2 : \tau }
+  { \coctx{\Gamma}{\cclrd{p}} \vdash \kvd{sync}~e_1~e_2 : \tau }
 \end{equation*}
@@
 The rule requires that $e_1$ yields a value of a singleton type $m$. The type is added as an
-indicator of the locked region to the context $\aclrd{p}\cup \{m\}$ which is then used to evaluate
+indicator of the locked region to the context $\cclrd{p}\cup \{m\}$ which is then used to evaluate
 the expression $e_2$.
@@
 the time of creation. It will be executed later and can only access the memory available then.
-This suggests that safe locking is perhaps better seen as an effect system.
+This suggests that safe locking is better seen as an effect system.
 
@@
 needs to reject programs that acquire locks in an invalid order. One way to model this is to 
-replace $\aclrd{p} \cup \{m\}$ with a \emph{partial} operation $\aclrd{p} \uplus \{m\}$ which
-is only defined when the lock $m$ can be added to the set $\aclrd{p}$. Supporting partial 
-operations on coeffect annotations is an interesting extension which we discuss in Section~?.
-The extension also lets us capture systems with effect-style lambda abstraction such as safe
-locking.
+replace $\cclrd{p} \cup \{m\}$ with a \emph{partial} operation $\cclrd{p} \uplus \{m\}$ which
+is only defined when the lock $m$ can be added to the set $\cclrd{p}$. Supporting partial 
+operations on coeffect annotations is an interesting future extension for coeffect systems.
 
@@
 \begin{array}{l}
-\kvd{let}~\ident{constant} = \lambda y \rightarrow \lambda x \rightarrow y\\
+\kvd{let}~\ident{constant} = \lambda y \rightarrow \lambda x \rightarrow y\\[-0.25em]
 \kvd{let}~\ident{answer} = (\lambda x \rightarrow x)~42
@@
 
-\caption{Structural coeffect liveness analysis}
+\figcaption{Structural coeffect liveness analysis}
 \label{fig:applications-struct-live}
@@
 \vspace{-0.5em}
-\caption{Semantics of structural liveness}
+\figcaption{Semantics of structural liveness}
 \label{fig:applications-struct-livesem}
@@
 
-\caption{Structural coeffect bounded reuse analysis}
+\figcaption{Structural coeffect bounded reuse analysis}
 \label{fig:applications-struct-bll}
@@
 \subsection{Data-flow languages revisited}
+\label{sec:applications-structural-dataflow}
 
@@
 
-\caption{Structural coeffect bounded reuse analysis}
+\figcaption{Structural coeffect bounded reuse analysis}
 \label{fig:applications-struct-df}
@@
 can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
-(e.g.~in the Perl language) or statically using a type system. Tainting can be viewed as a special
+(\eg~in the Perl language) or statically using a type system. Tainting can be viewed as a special
 case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
@@
 in \emph{secure information flow}. There, the aim is to guarantee that sensitive information (such
-as credit card number) cannot be leaked to contexts with low secrecy (e.g.~sent via an unsecured
+as credit card number) cannot be leaked to contexts with low secrecy (\eg~sent via an unsecured
 network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
@@
 There is a number of systems that also capture contextual properties, but that make it possible to 
-\emph{change} the context -- not just be evaluating certain code block in a different scope (e.g. by 
+\emph{change} the context -- not just be evaluating certain code block in a different scope (\eg by 
 wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example, acquires 

### 2014-12-08 text/Chapter03.tex
Finish proofreading and tweaking (Chapter 3)

@@
 
-In structural coeffect systems, the additional information are associated with individual variables.
-This is very often information about how the variables are used, or, in which contexts they are 
-used.
-
+In structural coeffect systems, the additional information is associated with individual variables.
+This is very often information about how the variables are used, or, in which contexts they are used. 
 In Chapter~\ref{ch:intro}, we introduced the idea using an example that tracks array access patterns.  
 Each variable is annotated with a range specifying which elements of the corresponding array 
-may be accessed. In this section, we look at a number of examples. We first consider an example
-inspired by linear logic. Then we revisit liveness and data-flow, for which the structural system 
-provides a more precise analysis. Finally, we look at a number of other practical uses including 
-security, tainting and provenance tracking.
+may be accessed.
+
+In this section, we look at three examples in detail -- we revisit liveness and show a practically
+useful structural version of the system; we consider an example inspired by linear logic; finally,
+we revisit data-flow to get a more precise analysis. Although quite different, the common pattern
+among these three examples is somewhat easier to see, because they all track information about 
+variable usage. We finish the section with a brief outline of several other applications.
 
@@
 The flat system for liveness analysis presented in Section~\ref{sec:applications-flat-live} is 
-interesting from a theoretical perspective, but it is not practically useful. In this section, we
+interesting from a theoretical perspective, but it is not practically useful. Here, we
 revisit the problem and define a structural system that tracks liveness per-variable.
@@
   We need to define the type system in a way that is similar to sub-structural systems 
-  (discussed in Section~\ref{sec:path-logic}) which provide explicit rules for manipulating
+  (discussed in Section~\ref{sec:path-logic}) and add explicit rules for manipulating
   the context.
@@
   individual annotations to individual variables. This lets us unify and combine flat and 
-  structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
-  explored in Section~\ref{sec:conclusions-meta}.
+  structural systems as discussed in Section~\ref{sec:unified-unified}, but the alternative is briefly
+  explored in Section~\ref{sec:unified-meta}.
 \end{itemize}
@@
 The structural system for liveness uses the same two-point lattice of annotations
-$\mathcal{L}=\{ \ident{L}, \ident{D} \}$ that was used by the flat system. We also use the
+$\mathcal{L}=\{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$ that was used by the flat system. We also use the
 $\sqcup, \sqcap$ and $\sqsubseteq$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.
@@
 the standard syntax-driven rules plus sub-coeffecting. In (\emph{var}), the context contains just the 
-single accessed variable, which is annotated as live. Other variables can be introduced using weakening. 
+single accessed variable, which is annotated as live. Unused variables can be introduced using weakening. 
 A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
@@
 \qquad
-f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \sqcup \aclrd{\mathbf{s}}} \tau
+f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \,\sqcup\, \aclrd{\mathbf{s}}} \tau
 \end{equation*}
@@
 is modelled using $\sqcup$ -- but here, we use a scalar-vector extension of the operation. Finally,
-the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$.
+the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$).
 
@@
 \begin{figure}[t]
-{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
+{\small a.) Ordinary, syntax-driven rules along with sub-coeffecting}
 \begin{equation*}
@@
   {\coctx{()}{\alift{}} \vdash c : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
 \end{equation*}
@@
   {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
 \end{equation*}
@@
 
-\begin{figure}[t]
-{\small a.) Semantics of ordinary expressions}
-\begin{equation*}
-\tag{\emph{var}}
-\sem{\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} = \lambda(x) \rightarrow x
-\end{equation*}
-\vspace{-1.5em}
-\begin{equation*}
-\tag{\emph{const}}
-\sem{\coctx{()}{\alift{}} \vdash c : \tau} = \lambda () \rightarrow \delta(c)
-\end{equation*}
-\vspace{-1.5em}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{abs}}
-\begin{array}{l}
-\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda y . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \lambda \textbf{v} \rightarrow\\[-0.25em]
-\qquad\lambda y \rightarrow \sem{\coctx{\Gamma, y \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}~(\textbf{v}, y)
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{app-1}}
-\begin{array}{l}
-\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{L}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
-  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
-  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{app-2}}
-\begin{array}{l}
-\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{D}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
-  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 }~\mathbf{v_1}~
-  \kvd{in}~g~()
-\end{array}
-\end{equation*}
-\vspace{0.5em}
-
-{\small b.) Semantics of structural context manipulation}
-\begin{equation*}
-\tag{\emph{weak}}
-\sem{\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} = 
-  \lambda(\textbf{v}, ()) \rightarrow
-    \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}~\textbf{v}
-\end{equation*}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{exch}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} = 
-  \lambda(\mathbf{v_1}, y, x, \mathbf{v_2}) \rightarrow\\[-0.25em]
-\qquad
-  \sem{\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~
-  (\mathbf{v_1}, x, y, \mathbf{v_2})
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{contr-1}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, (), \mathbf{v_2}) \rightarrow \\[-0.25em]
-\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} ~(\mathbf{v_1}, (), (), \mathbf{v_2})
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{contr-2}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, x, \mathbf{v_2}) \rightarrow \\[-0.25em]
-\quad\begin{cases}
-~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, x, \mathbf{v_2})\\[-0.25em]
-~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, (), x, \mathbf{v_2})\\[-0.25em]
-~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, (), \mathbf{v_2})
-\end{cases}
-\end{array}
-\end{equation*}
-\vspace{-0.5em}
-\figcaption{Semantics of structural liveness}
-\label{fig:applications-struct-livesem}
-\end{figure}
-\vspace{-0.5em}
-
-% --------------------------------------------------------------------------------------------------
-
 \paragraph{Semantics.}
@@
 $1 + \tau$ (which is $1$ for dead contexts and $\tau$ for live contexts). In the semantics of 
-expressions, the type wrapped the entire context, \ie~$1+(\tau_1 \times \ldots \times \tau_n)$.
-In the structural version, the semantics wraps individual elements of the free-variable context
-pair: $(1+\tau_1) \times \ldots \times (1+ \tau_n)$. For each variable, the type is indexed by 
-the corresponding annotation. More formally:
+expressions, the type constructor was applied to the entire context, \ie~$1+(\tau_1 \times \ldots \times \tau_n)$.
+In the structural version, the semantics applies the option type constructor to individual elements 
+of the free-variable context pair: $(1+\tau_1) \times \ldots \times (1+ \tau_n)$. For each variable, 
+the type is indexed by the corresponding annotation:
 %
@@
 Note that the product of the free variables is not an ordinary tuple of our language, but a special
-construction. This follows from the asymmetry of $\lambda$-calculus, as discussed in 
-Section~\ref{sec:applications-strucutre-vec}. Functions take just a single input and so they are
-interpreted in the same way as in flat calculus:
+construction (we return to this topic in Section~\ref{sec:struct-semantics}). This follows from the 
+asymmetry of $\lambda$-calculus, as discussed in Section~\ref{sec:applications-strucutre-vec}. Functions 
+take just a single input and so they are interpreted in the same way as in flat calculus:
 %
@@
 normal face. We freely re-associate products and so $(\mathbf{v}, x)$ should not be seen as a
-nested product, but simply a product with a number of variables represented as another product
-$\mathbf{v}$ with one more variable $x$ at the end. We shall be more precise in Chapter~\ref{ch:structural}.
+nested product, but simply as a product containing all variables from the product $\mathbf{v}$ 
+together with one additional variable $x$ at the end. We shall be more precise in 
+Chapter~\ref{ch:structural}.
 
@@
 
-In application, we again need to ``implement'' dead code elimination. When the input parameter
-of the function $g$ is live (\emph{app-1}), we first evaluate $e_2$ and then pass the result to
-$g$. When the parameter is dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values
-in $\mathbf{v_2}$ can be dead, \ie~$()$.
+As dictated by the semantics, the application again needs to ``implement'' dead code elimination
+(otherwise the type system would be unsound). When the input parameter of the function $g$ is live 
+(\emph{app-1}), we first evaluate $e_2$ and then pass the result to $g$. When the parameter is 
+dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values in $\mathbf{v_2}$ can be 
+dead, \ie~$()$.
 
@@
 so we need to consider three separate cases.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Semantics of ordinary expressions}
+\begin{equation*}
+\tag{\emph{var}}
+\sem{\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} = \lambda(x) \rightarrow x
+\end{equation*}
+\vspace{-1.5em}
+\begin{equation*}
+\tag{\emph{const}}
+\sem{\coctx{()}{\alift{}} \vdash c : \tau} = \lambda () \rightarrow \delta(c)
+\end{equation*}
+\vspace{-1.5em}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{abs}}
+\begin{array}{l}
+\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda y . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \lambda \textbf{v} \rightarrow\\[-0.25em]
+\qquad\lambda y \rightarrow \sem{\coctx{\Gamma, y \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}~(\textbf{v}, y)
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{app-1}}
+\begin{array}{l}
+\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{L}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
+  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
+  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
+\end{array}
+\end{equation*}
+\vspace{-1.0em}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{app-2}}
+\begin{array}{l}
+\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{D}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = 
+  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
+  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 }~\mathbf{v_1}~
+  \kvd{in}~g~()
+\end{array}
+\end{equation*}
+\vspace{0.5em}
+
+{\small b.) Semantics of structural context manipulation}
+\begin{equation*}
+\tag{\emph{weak}}
+\sem{\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} = 
+  \lambda(\textbf{v}, ()) \rightarrow
+    \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}~\textbf{v}
+\end{equation*}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{exch}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} = 
+  \lambda(\mathbf{v_1}, y, x, \mathbf{v_2}) \rightarrow\\[-0.25em]
+\qquad
+  \sem{\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~
+  (\mathbf{v_1}, x, y, \mathbf{v_2})
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{contr-1}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, (), \mathbf{v_2}) \rightarrow \\[-0.25em]
+\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} ~(\mathbf{v_1}, (), (), \mathbf{v_2})
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{-0.5em}
+\tag{\emph{contr-2}}
+\begin{array}{l}
+\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, x, \mathbf{v_2}) \rightarrow \\[-0.25em]
+\quad\begin{cases}
+~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, x, \mathbf{v_2})\\[-0.25em]
+~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, (), x, \mathbf{v_2})\\[-0.25em]
+~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, (), \mathbf{v_2})
+\end{cases}
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\figcaption{Semantics of structural liveness}
+\label{fig:applications-struct-livesem}
+\end{figure}
+\vspace{-0.5em}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 is no longer needed, because the sub-expressions use separate contexts. On the other hand, 
-we had to add weakening, contraction and exchange rules to let us manipulate the contexts.
+we had to add weakening, contraction and exchange rules to let us manipulate contexts.
 
@@
 how to build the context required by the sub-expressions and so the semantics implements 
-transformation from the context in the (typing) conclusion to the (typing) assumption.
+transformation from the context in the (typing) conclusion to the (typing) assumption. How
+should coeffects be understood, in general, is discussed further in Section~\ref{sec:flat-calculus-undestanding}.
 
 The structural systems discussed in the upcoming sections are remarkably similar to the one
-shown here. We discuss two more examples in details to explore the design space, but we shall
+shown here. We discuss two more examples to explore the design space, but we shall
 omit details that are the shared with the system in this section.
@@
 relate it with the original formulation.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
-\begin{equation*}
-\tyrule{var}{}
-  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
-    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
-    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
-\end{equation*}
-\vspace{0.5em}
-
-{\small b.) Structural rules for context manipulation}
-\begin{equation*}
-\tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
-  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma} 
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-\quad
-\begin{array}{l}
-\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
-\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-\begin{equation*}
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
-~
-\begin{array}{l}
-\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
-\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-
-\figcaption{Structural coeffect bounded reuse analysis}
-\label{fig:applications-struct-bll}
-\vspace{-1em}
-\end{figure}
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \paragraph{Type system.}
@@
 The rules have the same structure as those of the system for liveness analysis. The only difference is
-how annotations are combined -- here, we use integer multiplication ($\ast$) and addition ($+$).
+how annotations are combined. Here, we use integer multiplication ($\ast$) for sequential composition
+and addition ($+$) for point-wise composition.
 
@@
 call-by-name) multiplies the uses, meaning that the total number of uses is $(\cclrd{t} \ast \aclrd{\mathbf{s}})$
-(where $\ast$ is a multiplication of a vector by a scalar). This models the fact that for each use of
-the function parameter, we replicate the variable uses in $e_2$.
-
-Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
-the number of uses using the pointwise
-$\leq$ relation. We can view any variable as being used a greater
+(where $\ast$ is a point-wise multiplication of a vector by a scalar). This models the fact that 
+for each use of the function parameter, we replicate the variable uses in $e_2$.
+
+Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates the number of accesses using 
+the pointwise $\leq$ relation. We can view any variable as being used a greater
 number of times than it actually is.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Ordinary, syntax-driven rules along with sub-coeffecting}
+\begin{equation*}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
+\end{equation*}
+\vspace{0.5em}
+
+{\small b.) Structural rules for context manipulation}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma} 
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+\quad
+\begin{array}{l}
+\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
+\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+~
+\begin{array}{l}
+\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
+\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\figcaption{Structural coeffect bounded reuse analysis}
+\label{fig:applications-struct-bll}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
-$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian closed 
-category. In that model, we can reuse variables arbitrarily and so it is not
+$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian Closed 
+Category. In that model, we can reuse variables arbitrarily and so it is not
 a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
@@
 
-In the following, we use variable names $\mathbf{v}_i$ for context containing multiple variables (where 
-each variable may be available multiple times), \ie~have a type $\tau_1^{\cclrd{r_1}}\times\ldots\times\tau_m^{\cclrd{r_m}}$;
-We do not explicitly write the sizes of these vectors (number of variables in a context; number of 
-instances of a variable) as these are clear from the coeffect annotations. We assume that $\Gamma_2$ 
+In the following, we use vector names $\mathbf{v}_i$ for contexts containing multiple variables
+\ie~have a type $\tau_1^{\cclrd{r_1}}\times\ldots\times\tau_m^{\cclrd{r_m}}$. Each vector contains 
+multiple copies of each variable, to model the fact that variables are used in an affine way (at most 
+once). We do not explicitly write the sizes of these vectors (number of variables in a context; number 
+of instances of a variable) as these are clear from the coeffect annotations. We assume that $\Gamma_2$ 
 contains $n$ variables and that $\aclrd{s}=\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}$:
@@
     \sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_{\cclrd{t}})
+\\[1em]
 \end{array}
@@
 \paragraph{Bounded linear logic.}
-The system presented in this section differs from bounded linear logic (BLL) \cite{logic-bounded}.
-Using the terminology from Section~\ref{sec:path-sem-contextdep}, our system is written in the
-\emph{language semantics} style, while BLL is written in the \emph{meta-language} style.
-
-This means that the terms and types of our system are the terms and types of ordinary 
-$\lambda$-calculus, with the only difference that functions carry coeffect annotations.
-In BLL, the language of types is extended with a type constructor $!_k A$ (where $A$ is
-a proposition, corresponding to a type $\tau$ in our system). The type denotes a value
-$A$ that can be used at most $k$ times.
+The system presented in this section is based on the idea of bounded linear logic (BLL)
+\cite{logic-bounded}, but it is adapted to follow the structure of other coeffect systems 
+discussed in this chapter. This elucidates the connection between BLL and coeffects.
+
+The big difference, using the terminology from Section~\ref{sec:path-sem-contextdep}, is 
+that our system is written in \emph{language semantics} style, while BLL is written 
+in \emph{meta-language} style. We briefly consider the original BLL formulation.
+
+The terms and types of our system are the terms and types of an ordinary $\lambda$-calculus, 
+with the only difference that functions carry coeffect annotations. In BLL, the language of 
+types is extended with a type constructor $!_k A$ (where $A$ is a proposition, corresponding 
+to a type $\tau$ in our system). The type denotes a value $A$ that can be used at most $k$ times.
 
@@
 $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$ corresponds
-to a BLL assumption $!_{k_1} A_1, ..., !_{k_n} A_n$.
-
-Using the formulation of bounded logic (and omitting the terms), the weakening and
-contraction rules are written as follows:
+to a BLL assumption $!_{k_1} A_1, ..., !_{k_n} A_n$. Using the formulation of bounded logic 
+(and omitting the terms), the weakening and contraction rules are written as follows:
 %
@@
 treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
-semantics} and \emph{meta-language} in Section~\ref{sec:conclusions-meta}.
+semantics} and \emph{meta-language} in Section~\ref{sec:unified-meta}.
 
@@
 related to individual variables -- we could attach annotations to the variables themselves 
-rather than attaching a \emph{vector} of annotations to the entire context. Our approach has
-two benefits -- it lets you unify flat and structural systems (Chapter~\ref{ch:unified}) and 
-it also makes it possible to build composed systems that mix both flat and structural properties.
+rather than attaching a \emph{vector} of annotations to the entire context. The main benefit
+of our approach is that it lets us unify flat and structural systems (Chapter~\ref{ch:unified}).
 
@@
 
-When discussing data-flow languages in the previous section, we said that the context provides 
+When discussing data-flow languages in an earlier section, we said that the context provides 
 past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as 
@@
 following example:
-%
-\begin{equation*}
-\kvd{let}~\ident{offsetZip} = \ident{left} + \kvd{prev}~\ident{right}
-\end{equation*}
-%
-The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
+
+\begin{equation*}
+\kvd{let}~\ident{offsetAdd} = \ident{left} + \kvd{prev}~\ident{right}
+\end{equation*}
+%
+The value \ident{offsetAdd} adds values of \ident{left} with previous values of \ident{right}.
 To evaluate a current value of the stream, we need the current value of \ident{left} and one past
@@
 between flat ans structural liveness. In case of liveness analysis, we included the flat system
-only as an illustration (it is not practically useful).
-
-For data-flow, the flat system is less precise, but still practically useful (simplicity may outweigh 
-precision). As discussed in Section~X, the structural system is necessary when we allow arbitrary 
-recursion and cannot (easily) determine the number of required values statically.
+only as an illustration (it is not practically useful). For data-flow, the flat system is less 
+precise, but still practically useful (simplicity may outweigh precision).
 
@@
 To define the semantics of our structural data-flow language, we can use the same approach as when
-adapting flat liveness to structural liveness. Rather than wrapping the whole context in some wrapper
-(list or option type), we now wrap individual components of the product representing the variables
-in the context.
+adapting flat liveness to structural liveness. Rather than wrapping the whole context in a type 
+constructor (list or option), we now wrap the individual components of the product representing 
+the variables in the context.
 
@@
 can generally be turned into a structural one -- although this only gives a useful system when
-the flat version captures properties related to variables.
+the flat version captures statically scoped properties, \ie~related to variables.
 
@@
 (\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
-can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
-(\eg~in the Perl language) or statically using a type system. Tainting can be viewed as a special
+can cause security issues or other problems. Tainting can be done dynamically using a runtime mark
+(\eg~in the Perl language) or using a static type system. Tainting can be viewed as a special
 case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
@@
 
-Statically typed systems that based on tainting have been use to prevent cross-site scripting
-attacks \cite{app-tainting-xss} and a well known attack known as SQL injection
-\cite{app-tainting-sql,app-tainting-wasp}. In the latter chase, we want to check that SQL commands 
-cannot be directly constructed from, potentially dangerous, inputs provided by the user. Consider the 
-type checking of the following expression in a context containing variables \ident{id} and \ident{msg}:
+Static typed systems based on tainting have been use to prevent cross-site scripting attacks 
+\cite{app-tainting-xss} and SQL injection attacks \cite{app-tainting-sql,app-tainting-wasp}. 
+In the latter case, we want to check that SQL commands cannot be directly constructed from,
+potentially dangerous, inputs provided by the user. Consider the type checking of the following 
+expression in a context containing variables \ident{id} and \ident{msg}:
 %
@@
 
-In runtime checking, all (string) values need to be wrapped in an object that stores Boolean 
+In runtime checking, all (string) values need to be wrapped in an object with a Boolean 
 flag (for tainting) or more complex data (for provenance). In static checking, the information
@@
 \paragraph{Core dependency calculus.}
-The checking of tainting is a special case of checking of the \emph{non-interference} property 
+Taint checking is a special case of checking of the \emph{non-interference} property 
 in \emph{secure information flow}. There, the aim is to guarantee that sensitive information (such
@@
 network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type 
-system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
-more recent work. The checking of information flows has been also integrated (as a single-purpose
+system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} surveys
+more recent work. Information flow checking has been also integrated (as a single-purpose
 extension) in the FlowCaml \cite{app-security-flowcaml} language. Finally, Russo et al. and 
-Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that the properties can be checked
+Swamy et al. \cite{monad-secure-flow,monads-lightweight-ml} show that such properties can be checked
 using a monadic library.
@@
 
-Abadi et al. realized that there is a number of analyses similar to secure information flow
-and proposed to unify them using a single model called Dependency Core Calculus (DCC) \cite{app-dcc}.
-It captures other cases where some information about expression relies on properties of variables
-in the context where it executes.  The DCC captures, for example, \emph{binding time analysis}
+Although we do not describe a coeffect calculus for information flow checking, it is worth noting 
+that Abadi et al. \cite{app-dcc} realized that there is a number of analyses similar to secure information 
+flow and unified them using a single model called Dependency Core Calculus (DCC). This would be
+a useful basis for coeffect-based information flow checking.
+
+The DCC captures other cases where some information about expression relies on properties of variables
+in the context where it executes.  This includes, for example, \emph{binding time analysis}
 \cite{app-binding-time-analysis}, which detects which parts of programs can be partially evaluated
@@
 The work outlined in this section is another area where coeffect systems could be applied.
-We do not develop coeffect systems for tracking of tainting, security and provenance in details,
+We do not develop coeffect systems for taint tracking, security and provenance in detail,
 but briefly mention some examples in the upcoming chapters.
@@
 tainting, we annotate variables with $\cclrd{\ident{T}}$ for \emph{tainted} and with 
-$\cclrd{\ident{U}}$ for \emph{untainted}. Simply accessing variable does not introduce taint, 
-but using a variable in certain contexts -- such as in arguments of \ident{query} -- does introduce
-a taint. This is captured using the standard application rule (\emph{app}):
+$\cclrd{\ident{U}}$ for \emph{untainted}. Accessing a variable marks it as untained,
+but using an expression that depends on some variable in certain dangerous contexts -- such 
+as in arguments of \ident{query} -- does introduce a taint on all the variables contributing to
+the expression. This is captured using the standard application rule (\emph{app}):
 %
@@
 There is a number of systems that also capture contextual properties, but that make it possible to 
-\emph{change} the context -- not just be evaluating certain code block in a different scope (\eg by 
-wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example, acquires 
-new capabilities and returns those to the caller. While this thesis focuses on systems with passive 
-contexts, we briefly consider the most important examples of the \emph{active} variant.
+\emph{change} the context -- not just by evaluating certain code block in a locally modified context
+(\eg by wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example, 
+acquires new capabilities and returns those to the caller. Such actions appear to be closer to 
+effects than to coeffects. While this thesis focuses on systems with passive contexts, we briefly 
+consider the most important examples of the \emph{active} variant.
 
@@
 
-Calculus of capabilities differs in two ways. First, it allows explicit allocation and deallocation
-of memory regions (and so region lifetimes do not follow strict LIFO ordering). Second, it
-uses continuation-passing style. We ignore the latter aspect. The following example is almost 
-identical as the previous one:
+The calculus of capabilities differs in two ways. First, it allows explicit allocation and deallocation
+of memory regions (and so region lifetimes do not necessarily follow strict LIFO ordering). Second, 
+it uses continuation-passing style. We ignore the latter aspect. The following example is almost 
+identical to the previous one:
 %
@@
 
-This is where systems with active context differ. To type check such programs, we do not only need
-to know what context is required to call \ident{calculate}. We also need to know what effects it
+This is where systems with active context differ from systems with passive context. To type check 
+such programs, we do not only need to know what context is required to call \ident{calculate}
+(\ie~context on the left-hand side of $\vdash$). We also need to know what effects an expression
 has on the context when it evaluates and the current context meeds to be updated after a function 
-call.
+call. This is an effectful property that would appear on the right-hand side of $\vdash$.
 
@@
 \emph{input} of the function. Without giving any details, the semantics generally has the 
-following structure:
+following structure (with comonad to model coeffects on the left):
 %
@@
 could be either a \emph{new} context or \emph{update} to the original context). Thus the structure
-of the semantics would look as follows:
+of the semantics would look as follows (with comonad to model coeffects on the left and monad to 
+model effects on the right):
 %
@@
 carry a set of available memory regions. In this thesis, we focus only on passive contexts. 
-However, we briefly consider the problem of active contexts in the Section~X of the future work 
-chapter.
+However, capturing active contexts is an interesting future work.
 
@@
 investigates what language support is needed to enable safe dynamic software updating in C-like 
-languages. The system is based on the idea of capabilities.
+languages. The work is based on capabilities and follows a structure similar to the Calculus
+of Capabilities \cite{app-capabilities}.
 
@@
 a value is used concretely, the program examines its representation (and so it is not safe to
-change the representation during an update). An abstract use of a value does not need to examine
+change the representation during an update). An abstract use of a value does not examine
 the representation and so updating the value does not break the program.
@@
 abstraction capturing contextual properties, the languages use a wide range of formalisms -- including
-from principled approaches based on comonads, ad-hoc type system extensions and static analyses as
-well as approaches based on monads. We looked at a number of applications inclding Haskell's implicit 
+principled approaches based on comonads and modal S4, ad-hoc type system extensions and static analyses 
+as well as approaches based on monads. We looked at a number of applications inclding Haskell's implicit 
 parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of 
@@
 chapters, we identify the common structure more precisely and develop three calculi to capture it. We 
-will then be able to re-create many of the examples discussed in this chapter by instantiating our 
+will then be able to re-create many of the examples discussed in this chapter just by instantiating our 
 unified calculi.

### 2014-12-12 text/Chapter03.tex
Proofreading and English conventions (Chapter 4)

@@
 The important point here is that in flat coeffect systems, contextual requirements are 
-\emph{split} between the call-site and declaration-site. Furthermore, there is no syntactic
+\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
 structure that determines how the requirements are split. In the case of distributed
@@
 parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
-annotated with the set of implicit parameters that must be available on the call-site, \ie~
+annotated with the set of implicit parameters that must be available on the call site, \ie~
 $\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.
@@
 (\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
-the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration-site ($\coctx{\Gamma}{\cclrd{r}}$)
-and the call-site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
+the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
+and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
 parameter and removes it from the set of requirements.
@@
 
-The next typing derivation ($3$) requires the parameter \ident{?a} from both the declaration-site and 
-the call-site. This means that, at runtime, two values will be available. Our semantics for the 
+The next typing derivation ($3$) requires the parameter \ident{?a} from both the declaration site and 
+the call site. This means that, at runtime, two values will be available. Our semantics for the 
 system describes \emph{dynamic rebinding}, meaning that when the caller provides a value for a 
-parameter that is already specified by the declaration-site, the new value hides the old one. This
-means that only the value from the call-site is actually used. This ($4$) gives a more precise
+parameter that is already specified by the declaration site, the new value hides the old one. This
+means that only the value from the call site is actually used. This ($4$) gives a more precise
 typing for this situation.
@@
 The most interesting rules are (\emph{abs}) and (\emph{app}). In abstraction, we get two dictionaries
-$f$ and $g$ (from the declaration-site and call-site, respectively), which are combined and passed 
-to the body of the function. The semantics prefers values from the call-site, which is captured by
+$f$ and $g$ (from the declaration site and call site, respectively), which are combined and passed 
+to the body of the function. The semantics prefers values from the call site, which is captured by
 the $\uplus$ operation. In application, we first evaluate the expression $e_1$, then $e_2$ and finally
@@
 %
-Note that the declaration-site dictionary
-is ignored and the body is called with only the dictionary provided by the call-site. This is
+Note that the declaration site dictionary
+is ignored and the body is called with only the dictionary provided by the call site. This is
 a consequence of the fact that monadic functions are always pure values created using monadic
@@
 ``delay'' all impurities of the body. Here, the term ``delay'' refers to the fact that some 
-implicit parameters may be captured (from the declaration-site) at the time when the function is 
+implicit parameters may be captured (from the declaration site) at the time when the function is 
 defined, but before it is executed. As discussed in Section~\ref{sec:applications-structure-lam},
@@
 language. In the type system, the (\emph{abs}) rule places context-requirements on both the 
-declaration-site and the call-site. For implicit parameters, this rule introduces non-determinism
+declaration site and the call site. For implicit parameters, this rule introduces non-determinism
 in the type-inference, because the parameters can be split arbitrarily. However, as we show in the 
 next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two 
-parts of context (implicit parameter dictionaries) that are provided by the call-site and 
-declaration-site.
+parts of context (implicit parameter dictionaries) that are provided by the call site and 
+declaration site.
 
@@
 implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword. 
-Lambda abstraction splits the requirements between declaration-site (capturing remote reference) 
-and call-site (representing rebinding). For this reason, we do not discuss the system in details
+Lambda abstraction splits the requirements between declaration site (capturing remote reference) 
+and call site (representing rebinding). For this reason, we do not discuss the system in details
 and instead look at other uses.
@@
 \begin{itemize}
-\item[--] The function is represented as executable code for an call-site environment
+\item[--] The function is represented as executable code for an call site environment
   and is executed there, possibly after it is marshalled and transferred to another machine.
 \vspace{-0.5em}
-\item[--] The function body is compiled for the declaration-site environment; the value
+\item[--] The function body is compiled for the declaration site environment; the value
   that is returned is a remote reference to the code and function calls are performed as remote invocations.
@@
 yet another variant of lambda abstraction (\emph{abs}) -- the requirements of the body are duplicated
-and placed on \emph{both} the declaration-site and the call-site.
+and placed on \emph{both} the declaration site and the call site.
 
@@
 The body of the first function is just a constant $42$ and so the context of the body is marked
-as \emph{dead}. The parameter (call-site) of the function is not used and can also be marked as dead. 
-Similarly, no variables from the declaration-site are used and so they are also marked as dead.
+as \emph{dead}. The parameter (call site) of the function is not used and can also be marked as dead. 
+Similarly, no variables from the declaration site are used and so they are also marked as dead.
 
@@
 of the function is marked as \emph{live}. In the flat system, we do not track \emph{which} 
-variable was used and so we have to mark both the call-site and the declaration-site as live (this will
+variable was used and so we have to mark both the call site and the declaration site as live (this will
 be refined in a structural version of the system).
@@
 example in Figure~\ref{fig:applications-flat-cross}. When the body accesses any variables, it 
-requires both the argument and the variables from declaration-site. When it does not use any variables,
+requires both the argument and the variables from declaration site. When it does not use any variables,
 it marks both as dead. Finally, the (\emph{let}) rule annotates the composed expression with the
@@
 a variable that needs to be obtained and so the context is live. In the last case ($3$), the 
-function accesses a variable and so its declaration-site is marked as requiring the context 
+function accesses a variable and so its declaration site is marked as requiring the context 
 (\emph{abs}). This is where structural coeffect analysis would be more precise -- the system shown
@@
 property of coeffects that impurities cannot be delayed or thunked and attached just to the 
-function arrow -- we place requirements on both call-site and declaration-site.
+function arrow -- we place requirements on both call site and declaration site.
 
@@
 
-In function abstraction (\emph{abs}), the requirements of the body are duplicated on the declaration-site
-and the call-site as in liveness analysis. If the body requires $\cclrd{n}$ past values, it may access
+In function abstraction (\emph{abs}), the requirements of the body are duplicated on the declaration site
+and the call site as in liveness analysis. If the body requires $\cclrd{n}$ past values, it may access
 $\cclrd{n}$ values of any variables -- including those available in $\Gamma$, as well as the parameter
@@
 Lambda abstractions (\emph{abs}) receives two lists of the same size -- one containing values of
-the variables (list of products) from the declaration-site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle$ 
-and one containing the argument (list of values) provided by the call-site $\langle y_0, \ldots, y_{\cclrd{n}} \rangle$.
+the variables (list of products) from the declaration site $\langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle$ 
+and one containing the argument (list of values) provided by the call site $\langle y_0, \ldots, y_{\cclrd{n}} \rangle$.
 The semantics applies the well-known \emph{zip} operation on the lists and passes the result to the
@@
 effect-style lambda abstraction. Lambda abstraction associates all requirements with the 
-call-site -- a lambda function created under a lock cannot access protected memory available at 
+call site -- a lambda function created under a lock cannot access protected memory available at 
 the time of creation. It will be executed later and can only access the memory available then.
@@
 In the first case, the variable $x$ is dead, but was marked as live. In the second example, the
-declaration-site of the \ident{answer} value is dead, but was marked as live. This is because in 
+declaration site of the \ident{answer} value is dead, but was marked as live. This is because in 
 both of the expressions, \emph{some} variable is accessed. However, the (\emph{abs}) rule of flat
@@
 In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
-$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian Closed 
-Category. In that model, we can reuse variables arbitrarily and so it is not
+$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian-closed 
+category. In that model, we can reuse variables arbitrarily and so it is not
 a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
@@
 $\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
-and one bound. The (\emph{abs}) rule splits the annotations between the declaration-site and call-site
+and one bound. The (\emph{abs}) rule splits the annotations between the declaration site and call site
 of the function:

### 2014-12-12 text/Chapter04.tex
Proofreading and English conventions (Chapter 4)

@@
 
-\chapter{Flat coeffect calculus} 
+\chapter{The flat coeffect calculus} 
 \label{ch:flat}
@@
 access the environment in which they are executed. We identified two kinds of systems --
-\emph{flat} capturing whole-context properties and \emph{structural} capturing per-variable
-properties. As we show in Chapter~\ref{ch:unified}, the systems can be unified using a single 
-abstraction. This is useful when implementing and composing the systems, but such abstraction 
-is  \emph{less powerful} -- \ie~its generality hides useful properties that we can see 
-when we consider the systems separately. For this reason, this and the next chapter discusses 
-\emph{flat} and \emph{structural} systems separately.
+\emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing 
+per-variable properties. As we show in Chapter~\ref{ch:unified}, the systems can be unified 
+using a single abstraction. This is useful when implementing and composing the systems, but such 
+abstraction is  \emph{less powerful} -- \ie~its generality hides useful properties that we can see 
+when we consider the systems separately. For this reason, we discuss \emph{flat coeffects} and
+\emph{structural coeffects} separately in this and the next chapter.
+
+\paragraph{Chapter structure and contributions}
+\begin{itemize}
+\item[--] We present a \emph{flat coeffect calculus} as a type system that is parameterized by a 
+  \emph{flat coeffect algebra} (Section~\ref{sec:flat-calculus}). We show that the system can be 
+  instantiated to obtain three of the systems discussed in previous section (implicit parameters, 
+  liveness and data-flow).
+
+\item[--] We give the equational theory of the calculus and discuss type-pre\-ser\-vation for call-by-name
+  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
+  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections motivate the laws
+  of the flat coeffect algebra.
+
+\item[--] We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
+  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
+  This provides a deeper insight into how (and why) the calculus works and shows an 
+  intriguing link with effects.
+\end{itemize}
 
@@
 and whole-context data-flow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
-structure -- their lambda abstraction duplicates the requirements and their application arises
-from the combination of \emph{sequential} and \emph{pointwise} composition.
-
-The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl}), and
-similar systems for rebindable resources, differ in two ways. In lambda abstraction, they split
-the context requirements between the declaration-site and the call-site and they use only a single
-operator on the indices, typically $\cup$.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Contributions}
-
-All of the examples are practically useful and important and so we want to be able to capture all
-of them. Despite the differences, the systems can fit the same framework. The contributions of this
-chapter are as follows:
-
-\begin{itemize}
-\item We present a \emph{flat coeffect calculus} with a type system that is parameterized by a 
-  \emph{flat coeffect algebra} and can be instantiated to obtain all of the three examples
-  discussed (Section~\ref{sec:flat-calculus}).
-
-\item We give the equational theory of the calculus and discuss type-preservation for call-by-name
-  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
-  with subtyping and pairs (Section~\ref{sec:flat-exts}).
-
-\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which is a
-  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
-  The semantics provides deeper insight into how (and why) the calculus works.
-\end{itemize}
+structure. First, their lambda abstraction duplicates the requirements. Given a body with context 
+requirements $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
+annotated with $\cclrd{r}$. Second, their application arises from the combination of \emph{sequential} 
+and \emph{pointwise} composition.
+
+The system for tracking of implicit parameters and similar (Section~\ref{sec:applications-flat-impl})
+differ in two ways. In lambda abstraction, they split the context requirements between the 
+declaration site and the call site and they use only a single operator on the indices, typically $\cup$.
+
+Despite the differences, the systems fit the same unified framework. This becomes apparent when
+we consider the categorical structure (Section~\ref{sec:flat-semantics}). However, rather than 
+starting from the semantics, we first explain how the systems can be unified syntactically
+(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as a justification.
 
@@
 in programming languages. The syntax and type system of the flat coeffect calculus follows 
-similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
-in the structure, as explained in the previous chapter, most importantly in lambda abstraction.
-
-Wadler and Thiemann famously show a correspondence between effect systems to monads 
-\cite{monads-effects-marriage}, relating effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ 
-to monadic computations $\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar
-correspondence between \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry 
-of $\lambda$-calculus, this is not a simple mechanical dualization.
-
-The main purpose of the comonadic semantics presented in this chapter is to provide a semantic
-motivation for the flat coeffect calculus. The semantics is inspired by the work of Uustalu and
+asimilar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
+in the structure, as explained in the previous chapter, most importantly in lambda abstraction
+(the relationship with monads is further discussed in Section~\ref{sec:flat-related}).
+
+Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
+systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
+and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
+relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
+$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
+\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
+defining the semantics in terms of comonadic computations is not a simple mechanical dualization
+of the work on effect systems and monads.
+
+The main purpose of the comonadic semantics presented in this chapter is to provide a motivation 
+for the typing of the flat coeffect calculus. Our approach is inspired by the work of Uustalu and
 Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
 data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
-comonads} annotate the structure with information about the required context, \ie~$C^\sigma \tau_1 \rightarrow \tau_2$.
+comonads} annotate the structure with information about the required context, \ie~$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
 This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.
@@
 the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
-sets of implicit parameters, integers or other values. The expressions of the calculus are those
-of the $\lambda$-calculus with \emph{let} binding; assuming $T$ ranges over base types, the 
-types of the calculus are defined as follows:
+sets of implicit parameters, versions represented as integers or other values. The expressions of 
+the calculus are those of the $\lambda$-calculus with \emph{let} binding; assuming $\alpha$ ranges 
+over base types, the types of the calculus are defined as:
 %
@@
 e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
-\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\tau &::=& \alpha \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
 \end{array}
@@
 by a morphism $C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the object $C^{\cclrd{r}}$
-does not exist as a syntactical value. This is because we use comonads to define the 
+does not itself exist as a syntactical value. This is because we use comonads to define the 
 \emph{semantics} rather than \emph{embedding} them into the language as in the meta-language
-approaches (the distinction between the two approaches has been discussed in detail in 
-Section~\ref{sec:path-sem-langs}). The annotations $\cclrd{r}$ are formed by an algebraic
-structure discussed next.
+approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}). The annotations
+$\cclrd{r}$ are formed by an algebraic structure discussed next.
 
@@
 \begin{equation*}
-\tyrule{abs-imp}
+\tyrule{param}
   {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
@@
 \;
-\tyrule{abs-df}
+\tyrule{df1}
   {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
@@
 \end{equation*}
-
-~
 
@@
 \begin{equation*}
-\tyrule{abs-min}
+\tyrule{df2}
   {\coctx{\Gamma, x:\tau_1}{\textnormal{min}(\aclrd{n}, \aclrd{m})} \vdash e : \tau_2}
@@
 %
-The rule (\emph{abs-df}) is admissible in a system that includes the (\emph{abs-min}) rule. 
-If we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on 
+The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule. 
+Furthermore, if we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on 
 annotations of contexts), then the reverse is also true -- because 
 $\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{m}$ and $\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{n}$.
+In other words (\emph{df1}) is more precise, but (\emph{df2}) gives a sound over-approximation
+with a structure that can be unified with (\emph{param}).
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Understanding flat coeffects}
+\label{sec:flat-calculus-undestanding}
+
+Before looking at the type system in Figure~\ref{fig:flat-types}, let us clarify how the rules
+should be understood. The coeffect calculus provides both analysis of context dependence (type 
+system) and semantics for context (how it is propagated). These two aspects provide different
+ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
+used to define it.
+
+\begin{itemize}
+\item \textsc{Analysis of context dependence.}
+Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context requirements}. This means
+we can over-approximate them and require more than is actually needed at runtime.
+
+Syntactically, the typing rules should be read top-down. In function application, the context 
+requirements of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in 
+lambda abstraction, the requirements of a single expression (the body) are split between
+the declaration site and the call site.
+
+\item \textsc{Semantics of context passing.}
+Semantically, coeffect annotations $\cclrd{r}$ mo\-del \emph{contextual capabilities}. This means
+that we can throw away capabilities, if a sub-expression requires fewer than we 
+currently have.
+
+Semantically, the typing rules should be read bottom-up. In application, the capabilities 
+provided to the term $e_1~e_2$ are \emph{split} between the two sub-expressions; in abstraction,
+the capabilities provided by the call site and declaration site are \emph{merged} and passed
+to the body.
+\end{itemize}
+
+The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
+position} in the semantic model (Section~\ref{sec:flat-semantics}). It means that we need to be
+careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
+exactly the opposite things. To disambiguate, we always use the term \emph{context requirements} 
+when using the syntactic view, especially in the rest of Section~\ref{sec:flat-calculus}, and 
+\emph{context capabilities} or just \emph{available context} when using the semantic view, 
+especially in Section~\ref{sec:flat-semantics}.
 
@@
 Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{pointwise} composition, 
-respectively and the third one, $\czip$ represents context \emph{merging}. The term merging should be 
-understood semantically -- the operation models what happens when the semantics of lambda abstraction 
-combines context available at the declaration-site and the call-site.
+which are mainly used in function application. The third operator, $\czip$ is used in lambda 
+abstraction and represents \emph{splitting} of context requirements (or, semantically, \emph{merging} 
+of available context capabilities).
 
@@
 \\
-r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &&
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
+r\; \cpar\; r = r~~~~&
 \czero \;\cpar\; r = r = r \;\cpar\; \czero 
@@
 r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
-r\; \czip\; r = r \qquad\quad
-r\; \cpar\; r = r 
+r\; \czip\; r = r \qquad
 \\
-\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &~~~~&
-t\; \cleq\; t 
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
+t\; \cleq\; t \\[0.5em]
 \end{array}
@@
 \noindent
-In two of the three systems, some of the operators of the flat coeffect algebra coincide,
-but the data-flow system requires all three. Similarly, the two special elements also 
+In two of the three systems, some of the operators of the flat coeffect algebra coincide, but
+in the data-flow system all three are distinct. Similarly, the two special elements  
 coincide in some, but not all systems. The required laws are motivated by the aim to capture
@@
 functions. The laws of a monoid are required in order to form a category structure in the 
-categorical model (Section~\ref{sec:flat-semantics}).
+semantics (Section~\ref{sec:flat-semantics}).
 
@@
 requirement makes it possible to duplicate the coeffects and place the same requirement on both
-call-site and declaration-site, \ie~it makes the (\emph{abs-df}) rule admissible. In some cases, 
-the operator forms a monoid with the unit being the greatest element of the set. 
+call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
+this guarantees that the rule (\emph{df1}) is not a special case, but can always be derived 
+from (\emph{df2}). In some cases, the operator forms a monoid with the unit being the greatest 
+element of the set $\C$. 
 \end{itemize}
@@
 \noindent
-It is worth noting that the operators $\cpar$ and $\czip$ are dual in some of the systems. For 
-example, in data-flow computations, they are \emph{max} and \emph{min} respectively. However, this
-duality does not hold for implicit parameters. Using the syntactic reading, they represent 
-\emph{merging} and \emph{splitting} of context requirements -- in the (\emph{abs}) rule, 
-$\czip$ appears in the assumption and the combined context requirements of the body are split 
-between two positions in the conclusions; in the (\emph{app}) rule, $\cpar$ appears in the 
-conclusion and combines two context requirements from the assumptions.
+It is worth noting that, in some of the systems, the operators $\cpar$ and $\czip$ are the least 
+upper bound and the greatest lower bounds of a lattice. For example, in data-flow computations, they 
+are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
+(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.
+
+Using the syntactic reading, the two operators represent \emph{merging} and \emph{splitting} of context 
+requirements -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context 
+requirements of the body are split between two positions in the conclusions; in the (\emph{app}) rule, 
+$\cpar$ appears in the conclusion and combines two context requirements from the assumptions.
 
@@
 The flat coeffect algebra requires a pre-order relation $\cleq$, which is used to define 
-sub-coeffecting rule of the type system. When the monoid $(\C, \cpar, \czero)$ is idempotent
-and commutative monoid (semi-lattice), the $\cleq$ relation can be defined in terms of $\cpar$ as:
+sub-coeffecting rule of the type system. When the idempotent monoid $(\C, \cpar, \czero)$ also 
+has the commutative property (\ie~forms a semi-lattice), the $\cleq$ relation can be defined as the
+ordering of the semi-lattice:
 %
@@
 
-Furthermore, the $\cunit$ coeffect is often the top (greatest) or the bottom (smallest) 
-element of the semi-lattice, but not in general. As discussed in Section~\ref{sec:flat-syntax},
-when this is the case, we are able to prove certain properties of the calculus.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Understanding flat coeffects}
-\label{sec:flat-calculus-undestanding}
-
-Before looking at the type system in Figure~\ref{fig:flat-types}, let us clarify how the rules
-should be understood. The coeffect calculus provides both analysis of context dependence (type 
-system) and semantics for context (how it is propagated). These two aspects provide different
-ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
-used to define it.
-
-\begin{itemize}
-\item \textsc{Analysis of context dependence.}
-Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context requirements}. This means
-we can over-approximate them and require more than is actually needed at runtime.
-
-Syntactically, the typing rules should be read top-down. In (\emph{app}), the context requirements 
-of multiple assumptions are \emph{merged}; in (\emph{abs}), they are split between the declaration-site
-and the call-site.
-
-\item \textsc{Semantics of context passing.}
-Semantically, coeffect annotations $\cclrd{r}$ mo\-del \emph{contextual capabilities}. This means
-that we can throw away capabilities, if a sub-expression requires fewer than we 
-currently have.
-
-Semantically, the typing rules should be read bottom-up. In application, the capabilities 
-provided to the term $e_1~e_2$ are \emph{split} between the two sub-expressions; in abstraction,
-the capabilities provided by the call-site and declaration-site are \emph{merged} and passed
-to the body.
-\end{itemize}
-
-The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
-position} in the semantic model (Section~\ref{sec:flat-semantics}). It means that we need to be
-careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
-opposite things. To disambiguate, we always use the term \emph{context requirements} when using
-the syntactic view and \emph{context capabilities} or just \emph{available context} when using 
-the semantic view.
+Furthermore, the $\cunit$ coeffect is often the top or the bottom element of the semi-lattice. 
+As discussed in Section~\ref{sec:flat-syntax}, when this is the case, we are able to prove certain 
+syntactic properties of the calculus.
 
@@
 
-\caption{Type system for the flat coeffect calculus}
+\figcaption{Type system for the flat coeffect calculus}
 \label{fig:flat-types}
@@
 body is annotated with context requirements $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
-between the context-requirements on the declaration-site $\cclrd{r}$ and context-requirements on
-the call-site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.
+between the context-requirements on the declaration site $\cclrd{r}$ and context-requirements on
+the call site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.
 
 In function application (\emph{app}), context requirements of both expressions and the 
-function are combined as discussed in Chapter~\ref{ch:applications}. The pointwise composition
+function are combined. As discussed in Chapter~\ref{ch:applications}, the pointwise composition
 $\cpar$ is used to combine the context requirements of the expression representing a function 
@@
 The type system also includes a rule for let-binding. The rule is \emph{not} equivalent to the
-derivation for $(\lambda x.e_2)~e_1$, but it represents one admissible typing derivation. We
-return to let-binding after looking a number of examples. Additional constructs such 
-as pairs are covered in Section~\ref{sec:flat-exts}.
+typing derivation for $(\lambda x.e_2)~e_1$, but it corresponds to \emph{one} possible typing 
+derivation. As we show in \ref{sec:flat-calculus-let}, the typing used in (\emph{let}) is more
+precise than the general rule that can be derived from $(\lambda x.e_2)~e_1$. Additional 
+constructs such as pairs are covered in Section~\ref{sec:flat-exts}.
 
@@
 use just by providing a flat coeffect algebra. The following summary defines the systems for implicit 
-parameters, liveness and data-flow. For the latter two, we obtain more general (but compatible) rule 
-for lambda abstraction.
+parameters, liveness and data-flow. For the latter two, the general calculus has a lambda abstraction
+that is compatible with those discussed in Chapter~\ref{ch:flat}, but includes implicit sub-coeffecting.
 
@@
 parameters $\cclrd{c}$. As discussed earlier, we use the same type $\rho$ for all parameters, but
-it is easy to define an extension tracking set of parameters with type annotations.
+it is easy to define an extension that uses mappings from names to types.
 
 \begin{example}[Liveness]
-Let $L=\{ \ident{L}, \ident{D} \}$ be a two-point lattice such that $\ident{D} \sqsubseteq \ident{L}$
+Let $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ be a two-point lattice such that $\ident{D} \sqsubseteq \ident{L}$
 with a join $\sqcup$ and meet $\sqcap$. The flat coeffect algebra for liveness is then formed by
-$(L, \sqcap, \sqcup, \sqcap, \ident{L}, \ident{D}, \sqsubseteq)$.
+$(\mathcal{L}, \sqcap, \sqcup, \sqcap, \ident{L}, \ident{D}, \sqsubseteq)$.
 \end{example}
@@
 the meet operation $\sqcap$ and pointwise composition $\cpar$ is modelled by join $\sqcup$. 
-Two-point lattice is a commutative, idempotent monoid. The distributivity 
+The two-point lattice is a commutative, idempotent monoid. Distributivity 
 $(r \sqcup s) \sqcap t = (r \sqcap t) \sqcup (s \sqcap t)$ does not hold for \emph{every} 
-lattice, but it trivially holds for a two-point lattice used here.
+lattice, but it trivially holds for the two-point lattice used here.
 
 The definition uses join $\sqcup$ for the $\czip$ operator that is used by lambda abstraction.
-This means that, when the body is live $\ident{L}$, both declaration-site and call-site are 
-marked as live $\ident{L}$. When the body is dead $\ident{D}$, the declaration-site and call-site
-can be marked as dead $\ident{D}$, or as live $\ident{L}$, which is less precise, but permissible
-over-approximation, which could otherwise be achieved via sub-typing.
+This means that, when the body is live $\ident{L}$, both declaration site and call site are 
+marked as live $\ident{L}$. When the body is dead $\ident{D}$, the declaration site and call site
+can be marked as dead $\ident{D}$, or as live $\ident{L}$. The latter is less precise, but 
+permissible over-approximation, which could otherwise be obtained via sub-typing.
 
@@
 lambda abstraction. Annotating the body with $\emph{min}(\cclrd{r}, \cclrd{s})$ ensures that
-both call-site and declaration-site annotations are equal or greater than the annotation 
+both call site and declaration site annotations are equal or greater than the annotation 
 of the body. As with liveness, this allows over-approximation.
@@
 and $(\mathbb{N}, \mathit{min})$ forms a band. Note that data-flow is our first example where 
-$+$ is not idempotent. The distributivity laws require the following to be the case:
+$\cseq$ is not idempotent. The distributivity laws require the following to be the case:
 $\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see. Finally, a simple 
@@
 As a further example that was not covered earlier, it is also possible to combine liveness analysis
-and data-flow. In the above calculus, $0$ denotes that we require current value, but no previous
+and data-flow. In the above calculus, $0$ denotes that we require the current value, but no previous
 values. However, for constants, we do not even need the current value.
@@
 In optimized data-flow, context is annotated with natural numbers extended with the $\bot$ element,
-that is $\mathbb{N}_{\bot} = \mathbb{N} \cup \{\bot \}$ such that $\forall n \in \mathbb{N}. \bot \leq n$.
+that is $\mathbb{N}_{\bot} = \{\bot, 0, 1, 2, 3, \ldots \}$ such that $\forall n \in \mathbb{N}. \bot \leq n$.
 The flat coeffect algebra is $(\mathbb{N}_{\bot}, +, \mathit{max}, \mathit{min}, 0, \bot, \leq)$
@@
 \noindent
-Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$,
-$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid and $(\mathbb{N}, \emph{min})$ is a band.
-The required distributivity laws also holds for this algebra.
+Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$; for the
+bottom element $0 + \bot = \bot$ and for natural numbers $0 + n = n$. The structure 
+$(\mathbb{N}, \emph{max}, \bot)$ is also a monoid, because $\bot$ is the least element and so
+$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band (the extended
+\emph{min} is still idempotent and associative) and the distributivity law also holds
+for $\mathbb{N}_{\bot}$.
 
@@
 $\kvd{let}~x=e_1~\kvd{in}~e_2$ with context requirements $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
-This is a special case of typing of an expression $(\lambda x.e_2)~e_1$, using the idempotence
+This is a special case of typing an expression $(\lambda x.e_2)~e_1$, using the idempotence
 of $\czip$ as follows:
@@
 This design decision is similar to ML value restriction, but it works the other way round. Our
-\emph{let} binding is more restrictive rather than more general. The choice is motivated by the 
-fact that the typing obtained using the special rule for let-binding is more precise (with respect 
-to sub-coeffecting) for all the examples considered in this chapter. Table~\ref{tab:flat-simplelet}
-shows how the coeffect annotations are simplified for our examples.
+\emph{let} binding is more restrictive than the typing of abstraction-application, rather than
+being more general. The choice is motivated by the fact that the typing obtained using the special 
+rule for let-binding is more precise (with respect to sub-coeffecting) for all the examples consider 
+in this chapter. Table~\ref{tab:flat-simplelet} shows how the coeffect annotations are simplified 
+for our examples.
 
@@
 \vspace{-0.5em}
-\caption{Simplified annotation for let binding in sample flat calculi}
+\figcaption{Simplified annotation for let binding in sample flat calculi instances}
 \label{tab:flat-simplelet}
@@
 
-To see that the simplified annotations are \emph{better}, assume that we used arbitrary 
+To see that the simplified annotations are more precise, assume that we used arbitrary 
 splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$ rather than idempotence. The
 ``Definition'' column would use $\cclrd{s_1}$ and $\cclrd{s_2}$ for the first and second 
-$\cclrd{s}$, respectively. The corresponding simplified annotation (using idempotence) would
-have $\cclrd{s_1}\,\czip\,\cclrd{s_2}$ instead of $\cclrd{s}$. For all our systems, the 
+$\cclrd{s}$, respectively. The corresponding simplified annotation would have 
+$\cclrd{s_1}\,\czip\,\cclrd{s_2}$ instead of $\cclrd{s}$. For all our systems, the 
 simplified annotation (on the right) is more precise than the original (on the left):
@@
 %
-The inequality cannot be proved from other properties of the flat coeffect algebra. To make
+In other words, the inequality states that using idempotence, we get a more precise typing.
+This can be expressed using the abstract operators of the flat coeffect algebra as:
+%
+\begin{equation*}
+\cclrd{s_1} \;\cpar\; (\cclrd{s_2} \,\cseq\, \cclrd{r}) \;\cgeq\; 
+  (\cclrd{s_1} \,\czip\, \cclrd{s_2}) \;\cpar\; ((\cclrd{s_1} \,\czip\, \cclrd{s_2}) \,\cseq\, \cclrd{r})
+\end{equation*}
+%
+This property cannot be proved from other properties of the flat coeffect algebra. To make
 the flat coeffect system as general as possible, we do not \emph{in general} require it as
@@
 
-The type system of flat coeffect calculus arises as a generalization of the examples discussed in 
-Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical semantics
-of context-dependent computations. This is a direction that we explore in this section. Although
-the development presented here is interesting in its own, our main focus is \emph{using} categorical
-semantics to motivate and explain the design of flat coeffect calculus.
+The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
+discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical 
+semantics of context-dependent computations. This is a direction that we explore in this section. 
+Although the development presented here is interesting in its own, our main focus is \emph{using} 
+categorical semantics to motivate and explain the design of flat coeffect calculus.
 
@@
 As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful 
-computations can be captured uniformly using the (\emph{strong}) \emph{monad} structure. In that 
+computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that 
 approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
@@
 partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
-monad) and $\mtyp{}{\alpha} = (\alpha \times S)^S$ models side-effects (state monad). Here, the 
-structure of a strong monad provides necessary ``plumbing'' for composing monadic computations.
+monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$ 
+(state monad). Here, the structure of a strong monad provides necessary ``plumbing'' for composing 
+monadic computations -- sequential composition and strength for lifting free variables into the
+body of computation under a lambda abstraction.
 
@@
 computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
-$\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$ are modelled using
-the non-empty list comonad.
+are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
 
@@
 context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
-$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context. In the next section, we introduce
-\emph{indexed comonads}, which provide a more precise analysis and let us model computations with
-context requirements $\cclrd{r}$ as functions $\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using
-an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.
+$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an 
+over-simplification as we can use \eg~stacks of monad transformers and model functions with
+two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers 
+require defining complex system of lifting to be composable. Consequently, they are usually used
+for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
+capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
+stateful computation).}.
+
+In the next section, we introduce \emph{indexed comonads}, which provide a more precise analysis 
+and let us model computations with context requirements $\cclrd{r}$ as functions 
+$\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.
 
@@
 
-In category theory, \emph{comonad} is a dual of \emph{monad}. Informally, we get a comonad by 
-taking a monad and ``reversing the arrows''. More formally, one of the equivalent definitions of
-comonad looks as follows:
+In category theory, \emph{comonad} is a dual of \emph{monad}. As already outlined in 
+Section~\ref{sec:path-sem}, we get a definition of a comonad by taking a definition of a monad 
+and ``reversing the arrows''. More formally, one of the equivalent definitions of comonad 
+looks as follows:
 
@@
 the list is \emph{non-empty}, because we always need to be able to obtain a value. The \ident{cobind}
-defined here returns a list of the same length as the original where, for each element, the 
+operation defined here returns a list of the same length as the original where, for each element, the 
 function $f$ is applied on a \emph{suffix} list starting from the element. Using a simplified
@@
 mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
-because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. Similarly,
-the \ident{Maybe} data type defined as $1 + \alpha$ is not a comonad for the same reason.
-However, if we consider flat coeffect calculus for liveness, it appears natural to model computations
-as function $\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such model, we first need to 
-generalise comonads to \emph{indexed comonads}.
+because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. The \ident{Maybe} 
+data type defined as $1 + \alpha$ is not a comonad for the same reason. However, if we consider 
+flat coeffect calculus for liveness, it appears natural to model computations as functions 
+$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalise comonads 
+to \emph{indexed comonads}.
 
@@
 \end{compactitem}
-such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$
-and the identity $\idf{\cclrd{s}} : \ctyp{\cclrd{s}}\alpha \rightarrow \ctyp{\cclrd{s}}\alpha$:
+such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$:f
 \begin{align}
@@
 Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings 
-$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the operation $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
-operation is now also formed by a \emph{family} of mappings for different pairs of indices 
+$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
+operation is now formed by a \emph{family} of mappings for different pairs of indices 
 $\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations 
 and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
-For the purpose of this thesis, it is sufficient to treat $\ident{cobind}$ as a family of
-mappings or, when it does not  introduce ambiguity, view it as a single mapping.
+For the purpose of this thesis, it is sufficient to omit the superscripts and treat $\ident{cobind}$ just 
+as a family of mappings (rather than natural transformations). When this does not introduce ambiguity, 
+we also occasionally omit the subscripts.
 
@@
 the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
-but this is merely for symmetry. Crucially, this means that the operation is defined only
-for some special contexts.
+but this is merely for symmetry and as a useful reminder to the reader. Crucially, this means that 
+the operation is defined only for special contexts.
 
@@
 
-The category that models sequential composition is formed by the unit arrow $\ident{counit}$ 
-together with the (associative) composition operation that composes computations with 
-contextual requirements as follows:
+\paragraph{Composition.}
+The co-Kleisli category that models sequential composition is formed by the unit arrow (provided
+by $\ident{counit}$) together with the (associative) composition operation that composes computations 
+with contextual requirements as follows:
 %
@@
 with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
-requires $\cclrd{r}\,\cseq,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
+requires $\cclrd{r}\,\cseq\,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
 are attached to the input of the composed function.
@@
 Any comonad $\ctyp{}{}$ is an indexed comonad with an index provided by a trivial monoid $(\{1\},\ast,1)$
-where $1\ast 1 = 1$ and $\ctyp{1}{}$ is the underlying mapping $\ctyp{}{}$ of the original comonad. The
+where $1\ast 1 = 1$. The mapping $\ctyp{1}{}$ is the mapping $\ctyp{}{}$ of the underlying comonad. The
 operations $\ident{counit}_1$ and $\ident{cobind}_{1,1}$ are defined by the operations $\ident{counit}$
@@
 \begin{example}[Indexed option]
-The indexed option comonad is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
+The \emph{indexed option comonad} is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
 where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
@@
 \noindent
-The indexed option comonad models the semantics of the liveness coeffect system discussed in 
-\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
+The \emph{indexed option comonad} models the semantics of the liveness coeffect system discussed in 
+Section~\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
 and $\ctyp{\ident{D}}{\alpha}=1$ models a dead context which does not contain a value. The \ident{counit}
-operation extracts a value from a live context; \ident{cobind} can be seen as an implementation of 
-dead code elimination. The definition only evaluates $f$ when the result is marked as live and is thus
-required, and it only accesses $x$ if the function $f$ requires its input.
+operation extracts a value from a live context. As in the direct model discussed in Chapter~\ref{ch:appendix},
+the \ident{cobind} operation can be seen as an implementation of dead code elimination. The definition 
+only evaluates $f$ when the result is marked as live and is thus required, and it only accesses $x$ if 
+the function $f$ requires its input.
 
@@
 permit (non-indexed) comonad structure, because $\ident{counit}~()$ is not defined. This is not a 
-problem with indexed comonads, because \ident{counit} only needs to be defined on live context.
+problem with indexed comonads, because live contexts are distinguished by the (type-level) coeffect 
+annotation and \ident{counit} only needs to be defined on live contexts.
 
@@
 As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
-= \alpha \times (\cclrd{r} \rightarrow R)$ represents a value $\alpha$ together with a function that
+= \alpha \times (\cclrd{r} \rightarrow \rho)$ represents a value $\alpha$ together with a function that
 associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
@@
 \noindent
-The definition of \ident{counit} simply ignores the function and returns the value in the context.
-The \ident{cobind} operation uses the restriction operation $\restr{f}{\cclrd{r}}$, which we 
-already defined when discussing semantics of implicit parameters in Section~\ref{sec:applications-flat-impl}
-(indeed, \ident{cobind} here captures an essential part of the semantics).
-
-The function $g$ in \ident{cobind} is defined on the union of the implicit parameters, 
-\ie~$\cclrd{r}\cup\cclrd{s}\rightarrow\rho$. When passing it to $f$, we restrict it to
-just $\cclrd{r}$ and when returning it as a result, we restrict it to $\cclrd{s}$.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Properties and related notions}
-
-We discuss additional examples in Section~\ref{sec:flat-semantics-monoidal}, after we look at the
-remaining structure that is needed to define the semantics of flat coeffect calculus. Before doing 
-so, we discuss additional properties and categorical structures that have been proposed mainly in 
-the context of monads and effects and are related to indexed comonads.
-
-\paragraph{Shape preservation.}
-Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
-this means that the shape of the additional context does not change during the computation. For
-example, in the \ident{NEList} comonad, the length of the list stays the same after applying 
-\ident{cobind}.
-
-Indexed comonads are not restricted by this property of comonads. For example, given the indexed
-product monad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$ above, the shape of
-the context changes from containing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to containing
-just implicit parameters $\cclrd{s}$.
-
-\paragraph{Families of monads.}
-When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
-\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
-comonads}, is defined as follows.
-
-\begin{definition}
-\label{def:flat-family}
-A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
-  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
-$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
-$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
-certain coherence conditions.
-\end{definition}
-
-Family of comonads is more restrictive than indexed comonad, because each of the data types needs
-to form a comonad separately. For example, our indexed option does not form a family of comonads
-(again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given a 
-family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
-an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
-we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
-$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
-
-\paragraph{Parameteric effect monads.}
-Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
-indexed comonads) are closely related to our definition.  Although presented in a more general 
-categorical framework (and using monads), the model defines \ident{unit} operation only on the 
-unit of a monoid and \ident{bind} operation composes effect annotations using the provided monoidal 
-structure.
+In the definition, we use the notation $(a, g)$ for a pair containing the value of type $\alpha$
+together with $g$, which is a function $\cclrd{r} \rightarrow \rho$. The \ident{counit} operation
+takes a value and a function (with empty set as a domain), ignores the function and extracts the
+value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to 
+restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in oder to get
+implicit parameters required by the argument of $f$ and by the resulting computation, respectively
+(\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed 
+to \ident{cobind} is initially defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is 
+valid in both cases.
+
+The structure of \emph{indexed comonads} is sufficient to model sequential composition of 
+computations that use a single variable (as discussed in Section~\ref{sec:path-sem}). To model 
+full $\lambda$-calculus with lambda abstraction and multiple-variable contexts, we need 
+additional operations introduced in the next section.
 
@@
 
-Indexed comonads model the semantics of sequential composition, but additional structure is needed
-to model the semantics of the flat coeffect calculus. This is where the duality between monads and 
-comonads can no longer help us, because context is propagated differently than effects in lambda 
-abstraction and application.
-
-Whereas Moggi~\cite{monad-notions} requires \emph{strong} monad to model effectful $\lambda$-calculus, 
-Uustalu and Vene~\cite{comonads-notions} require \emph{lax semi-monoidal} comonad to model 
-$\lambda$-calculus with contextual properties. The structure requires a monoidal operation:
+Because of the assymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
+the duality between monads and comonads can no longer help us with defining the additional structure
+required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
+to the context. In application and lambda abstraction, the context is propagated differently than 
+in effectful computations.
+
+To model the effectful $\lambda$-calculus, Moggi~\cite{monad-notions} requires a \emph{strong} monad which
+has an additional operation $\ident{strength} : \alpha\times\mtyp{}{\beta} \rightarrow \mtyp{}{(\alpha\times\beta)}$.
+This allows lifting of free variables into an effectful computation. In Haskell, strength can be expressed
+in the host language and so it is implicit.
+
+To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions} 
+require \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
 %
@@
 The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
-contexts and is used to merge the context of the declaration-site (containing free variables)
-and the call-site (containing bound variable). For example, for implicit parameters, this combines
+contexts and is used to merge the context of the declaration site (containing free variables)
+and the call site (containing bound variable). For example, for implicit parameters, this combines
 the additional parameters defined in the two contexts.
@@
 The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
-\emph{splitting} of contexts. These are provided by \emph{lax} and \emph{oplax} monoidal 
-structures. In addition, we also need a lifting operation (similar to $\iota$ from 
-Definition~\ref{def:flat-family}) to model sub-coeffecting.
+\emph{splitting} of contexts. In addition, we also need a lifting operation (similar to $\iota$ 
+from Definition~\ref{def:flat-family}) to model sub-coeffecting.
 
@@
 Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-an \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
+a \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
 equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
@@
 \end{compactitem}
-%
-% We should require lift to be comonad morphism
-%
-% \tag{\emph{comonad morphism}}
-%   \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f \circ \ident{lift}_{\cclrd{r'}\cseq\cclrd{s'}, \cclrd{r}\cseq\cclrd{s}} &=
-%   \ident{lift}_{\cclrd{s'}, \cclrd{s}} \circ \ident{cobind}_{\cclrd{r'}, \cclrd{s'}}~(f \circ \ident{lift}_{\cclrd{r'}, \cclrd{r}} )
-%
-% ... but adapting the coherence conditions for counit/cobind is ugly!
-%
 \end{definition}
@@
 values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combineds them into a 
-single value with additional context $\cclrd{r}\czip\cclrd{s}$. The $\czip$ operation often represents
+single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
 \emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
-bounds for the liveness and data-flow examples, but not for implicit parameters. However, they are useful 
+bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful 
 as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
@@
 \paragraph{Alternative definition.}
-Although we do not require this as a general law, in all our systems, it is the case that
+Although we do not demand this as a general law, in all our systems, it is the case that
 $\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$.
-This allows a simpler definition of \emph{indexed flat comonad} by expressing the \ident{split} operation
-in terms of the lifting (sub-coeffecting) as follows:
+This special case allows a simpler definition of \emph{indexed flat comonad} by expressing the 
+\ident{split} operation in terms of lifting (sub-coeffecting) as follows:
 %
@@
 %
-The $\ident{map}_{\cclrd{r}}$ operation is the mapping on functions that corresponds to the object 
+The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object 
 mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map} 
@@
 are first and second projections from a two-element pair. To define the 
-$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we duplicate the argument $c$, then use 
-lifting to throw away additional parts of the context and then transform the values in the 
+$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
+to throw away additional parts of the context and then transform the values in the 
 context.
 
-This alternative is valid for our examples, but we do not use it for two reasons. Firstly, it 
-requires duplication of the value $c$, which is not required elsewhere in our model. So, using 
-explicit \ident{split}, our model could be embedded in a linear or affine model. Secondly, it
-is similar to the definition that is needed for structural coeffects in Chapter~\ref{ch:structural}
-and so it makes the connection between the two system easier to see.
+This alternative is valid for our examples, but we do not use it for two reasons. Firstly, this
+would be the only place where our semantics uses a variable \emph{twice} (in this case $c$). 
+Thus using an explicit \ident{split} means that the structure required by our semantics does not
+need to provide variable duplication and our model could be embedded in linear or affine category.
+Secondly, explicit \ident{split} is similar to the definition that is needed for structural 
+coeffects in Chapter~\ref{ch:structural} and so it makes the connection between the two system 
+easier to see.
 
 \paragraph{Examples.}
-All examples of \emph{indexed comonads} discussed in Section~\ref{sec:flat-semantics-idx} can
-be extended into \emph{flat indexed comonads}.
+All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:flat-semantics-idx} can
+be extended into \emph{flat indexed comonads}. Note that this is not a \emph{general} statement, 
+because each example requires us to define additional operations, specific for the example.
 
@@
 operation is partly determined by the indices. When the input is \emph{dead}, both values have 
-to be dead (this is also the only solution of $\ident{D}=\cclrd{r}\sqcap\cclrd{D}$), but when
+to be dead (this is also the only solution of $\cclrd{\ident{D}} =\cclrd{r}\sqcap\cclrd{s}$), but when
 the input is \emph{live}, the operation can perform implicit sub-coeffecting and drop one of
@@
 $\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
-is $\alpha \times (\cclrd{r} \rightarrow R)$ where $R$ is some representation of a parameter value.
+is $\alpha \times (\cclrd{r} \rightarrow \rho)$ where $\rho$ is the type of implicit parameter values.
 The additional operations are defined as:
@@
 of \ident{lift}, which performs just the restriction. The \ident{merge} operation is more 
-interesting. It uses $\uplus$ operation that we defined when introducing implicit parameters
+interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
 in Section~\ref{sec:applications-flat-impl}. It merges the values, preferring the definitions from
-the right-hand side (call-site) over left-hand side (declaration-site). Thus the operation is not
+the right-hand side (call site) over left-hand side (declaration site). Thus the operation is not
 symmetric.
@@
 non-indexed version, the semantics is provided by a non-empty list. In the indexed semantics,
-the index represents the length of the storing past values. The data type is then a pair of 
+the index represents the number of available past values. The data type is then a pair of 
 the current value, followed by $n$ past values. The mappings that form the flat indexed comonad 
@@
 
+\subsection{Properties and related notions}
+
+The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
+calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
+additional properties and other categorical structures that have been proposed mainly in the 
+context of monads and effects and we look how they relate to indexed comonads.
+
+\paragraph{Shape preservation.}
+Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
+this means that the core comonad structure does not provide a way of modeling computations where
+the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
+the length of the list stays the same after applying \ident{cobind}.
+
+~
+
+Indexed comonads are not restricted by this property of comonads. For example, given the indexed
+product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
+the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
+just implicit parameters $\cclrd{s}$.
+
+\paragraph{Families of monads.}
+When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
+\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
+comonads}, is defined as follows.
+
+\begin{definition}
+\label{def:flat-family}
+A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
+  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
+$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
+$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
+certain coherence conditions.
+\end{definition}
+
+\noindent
+A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data 
+types needs to form a comonad separately. For example, our indexed option does not form a family of 
+comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given 
+a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
+an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
+we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
+$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
+
+\paragraph{Parameteric effect monads.}
+Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
+indexed comonads) are closely related to our definition.  Although presented in a more general 
+categorical framework (and using monads), the model defines the \ident{unit} operation only on the 
+unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal 
+structure.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Semantics of flat calculus}
+\label{sec:flat-semantics-calculus}
 
@@
 computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed 
-comonad} structure, we can now define a single uniform semantics that is capable capturing of all 
+comonad} structure, we can now define a single uniform semantics that is capable of capturing all 
 our examples, as well as other computations that can be modelled by the structure.
@@
   \delta~(c_i) & (\emph{const})
-\\[0.5em]
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }~\ctx = & (\emph{sub})\\[-0.25em]
-  \hspace{3em}\sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx)~
-    \hspace{7.5em}(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
 \\[0.5em]
@@
   \end{array}    
-\end{array}
-\end{equation*}
-
-\caption{Categorical semantics of the flat coeffect calculus}
+\\[1.25em]
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }~\ctx = & (\emph{sub})\\[-0.25em]
+  \hspace{3em}\sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx)~
+    \hspace{7.5em}(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
+\end{array}
+\end{equation*}
+
+\figcaption{Categorical semantics of the flat coeffect calculus}
 \label{fig:flat-semantics}
@@
 function pairing (given $f:A\rightarrow B$ and $g:C\rightarrow D$ then $f\times g : A\times C \rightarrow B \times D$). 
-These operations can be provided by \eg~a Cartesian Closed Category.
+These operations can be provided by \eg~a Cartesian-closed category.
 
@@
 
-\begin{remark}[Correspondence]
+\begin{theorem}[Correspondence]
+\label{thm:flat-correspondence}
 In all of the typing rules of the flat coeffect system, the context annotations $\cclrd{r}$ of typing 
@@
 by $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}$.
-\end{remark}
+\end{theorem}
 \begin{proof}
@@
 \noindent
-Thanks to the indexing, the statement of the remark is significantly stronger than for a 
+Thanks to indexing, the statement of the theorem is significantly stronger than for a 
 non-indexed system, because it provides the justification for our choice of indices in the typing
@@
 (within the comonadic framework we use).
+
+Although the notion of indexed comonads presented in this section is novel and interesting 
+in its own, the main reason for introducing it is to motivate the flat coeffect type system.
+This is captured by the Theorem~\ref{thm:flat-correspondence}. In the next section, we shift our
+focus from (categorical) semantics to syntactic properties of the calculus.
 
@@
 
-\section{Equational theory}
+\section{Syntactic equational theory}
 \label{sec:flat-syntax}
@@
 are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
-that hold for all of the systems we consider. Further properties hold for systems where the context
+that hold for all of the coeffect systems. Further properties hold for systems where the context
 satisfies additional properties.
@@
 The rules capture syntactic reductions that can be performed in a general calculus, without any 
-knowledge of the specific notion of context. The \kvd{glet} notation models explicit sequencing
-of context-dependent computations an is inspired by Filinski \cite{monads-inaction}. In the rest
-of the section, we briefly outline the interpretation of the four rules and then we focus on
-call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name (Section~\ref{sec:flat-syntax-cbn})
-in more details.
-
-The focus of this work is on the general coeffect system and so we do not discuss the operational
+knowledge of the specific notion of context. If the reductions preserve the type of the expression
+(type preservation), then operational semantics can be defined as a repeated application of the 
+rules, until a specified normal form (\ie~a value) is reached.
+
+The first two equations define well-known call-by-name and call-by-value reductions. The \kvd{glet} 
+reduction (inspired by Filinski \cite{monads-inaction}) models a system where program is reduced to 
+a sequence of primitive operations, sequentially composed using the special \kvd{glet} construct. 
+In the rest of the section, we briefly outline the interpretation of the four 
+rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name 
+(Section~\ref{sec:flat-syntax-cbn}) in more details.
+
+The focus of our work is on the general coeffect system and so we do not discuss the operational
 semantics of the specific notions of context. However, some work in that area has been done
@@
 a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
-A function $\lambda x.y$ is a effect-free:
+A function $\lambda x.y$ is effect-free:
 %
@@
 calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
-parameters system. This makes the model suitable for languages such as Haskell.
+parameters system. This means that certain coeffect systems support call-by-name evaluation strategy
+and could be embedded in purely functional language (such as Haskell).
 
@@
 The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
-A function $(\lambda x.e)$ does not delay all context requirements of the body $e$ and may have
+A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
 immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
@@
 
-The call-by-value evaluation strategy holds for a wide range of flat coeffect calculi, including all
-our three examples. However, it is rather weak -- in order to use it, the concrete semantics needs
-to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ to
-a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
+The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
+including all our three examples. However, it is rather weak -- in order to use it, the concrete 
+semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
 
@@
 provides a construct representing \emph{sequential composition} and expressions can be reduced to 
-a normal form, consisting of a sequenced context-dependent operations. We choose to write the 
+a normal form, consisting of a sequence of context-dependent operations. We choose to write the 
 sequencing operation as \kvd{glet} to emphasize that this is a separate primitive and not an 
@@
 a language that provides constructs for explicitly providing the context. For example, consider
-a language with implicit parameters where a parameter is defined by $e_1~\kvd{with}~\ident{?p}=e_2$.
+a language with implicit parameters where a parameter is defined by $\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2$.
 Semantics for such language would provide a reduction rule:
@@
 \begin{equation*}
-(\kvd{glet}~x_1 = \ident{?p}~\kvd{in}~e)~\kvd{with}~\ident{?p}=e_1 ~~\leadsto~~ \kvd{let}~x_1 = e_1~\kvd{in}~e
-\end{equation*}
-%
-Here, the \kvd{glet} construct provides a way of sequentialising the context-requi\-rements so
-that they can be discharged by matching constructs providing the required contexts. As discussed
+\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~(\kvd{glet}~x = \ident{?p}~\kvd{in}~e)
+ ~~\leadsto~~ \kvd{let}~x = e_1~\kvd{in}~e
+\end{equation*}
+%
+The \kvd{glet} construct provides a way of sequentialising the context-require\-ments so
+that they can be discharged by matching constructs providing the required context. As discussed
 tfearlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
@@
 are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
-introduce a function arrow (lambda abstraction) and eliminate it (application) are not too strong
-and sufficiently strong.
+introduce a function arrow (lambda abstraction) and eliminate it (application) are sufficiently 
+strong, but not too strong.
 
@@
 calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
-term to a \emph{value} has to be provided for each system. Syntactic category for values is 
-defined as:
+term to a \emph{value} has to be provided for each system. Syntactic values are defined
+in the usual way:
 %
@@
 %
-The category \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
-coeffect calculus cannot be defined purely syntactically.
+The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
+coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
+not automatically defer all context requirements.
 
@@
 variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
-context-requirements of the term:
+context requirements of the term:
 
@@
 $\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
-change the type and context requirements: $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
+change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
 \end{lemma}
@@
 proving that call-by-value reduction preserves typing requires an additional constraint on the
-flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations:
-%
+flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
+by the (\emph{approximation}) property:
+
 \begin{equation*}
 \cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
-\hspace{10em}\textnormal{(approximation)}
+\hspace{10em}(\textit{approximation})
 \end{equation*}
@@
 
-The property holds for all three systems we consider -- for implicit parameters, this is 
+The property holds for the three systems we consider -- for implicit parameters, this is 
 an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
@@
 Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
-$(\lambda x.e)\,v$ to a term $\substack{e}{x}{v}$, the type preservation theorem is
+$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the soundness theorem is
 stated as follows:
 
-\begin{theorem}[Call-by-value reduction]
+\begin{theorem}[Call-by-value soundness]
 \label{thm:cbv-reduction}
@@
 \end{equation*}
-The last step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
-From Lemma~\ref{thm:flat-subst-cbv} $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
+The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
+From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
 As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
@@
 %
-Thus, the reduction preserves type and coeffect annotation (although this may not be the
-\emph{only} typing of the original term).
+Comparing the final conclusions of the above two typing derivations shows that 
+the reduction preserves type and coeffect annotation.
 \end{proof}
@@
 
-In the call-by-name reduction of $(\lambda x.e_1)~e_2$, the expression $e_2$ is substituted for all 
-occurrences of the variable $v$ in an expression $e_1$. As discussed in 
-Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in general} preserve 
-the type of a program, but it does preserve the typing in some interesting cases. The typing is
-preserved for different reasons in two of our systems, so we briefly review the concrete examples.
+When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the 
+sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$. 
+As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in 
+general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two 
+interesting cases.
+
+The typing is preserved for different reasons in two of our systems, so we briefly review the 
+concrete examples. Then, we prove the substitution lemma for two special cases of flat coeffects
+(Lemma~\ref{thm:cbn-substitution-top} and Lemma~\ref{thm:cbn-substitution-bot}) and finally, we
+state the conditions under which typing preservation hold for flat coeffect calculi 
+(Theorem~\ref{thm:cbn-flat}).
 
@@
 shown earlier with effectful computations. As a minimal example, consider the substitution of 
-$\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
+a context-dependent expression $\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
 %
@@
 
-Semantically, the coeffect over-approximates the actual requirements -- the code does not access
-previous value of the argument $x$. This cannot be caputred by a flat coeffect system, but can
-be captured using the structural system discussed in Chapter~\ref{ch:structural}.
+Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
+actually access a previous value of the argument $x$. This cannot be caputred by a flat coeffect 
+system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
 
@@
 preserves the type of the function. However, this is not the case for all systems. Consider substituting
-an implicit parameter access $\ident{?p}$ for a variable $y$:
+an implicit parameter access $\ident{?p}$ for a free variable $y$ under a lambda:
 %
@@
 \paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
-substituting any expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
+substituting an arbitrary expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
 %
@@
 because the body contains a variable access. An expression $e$ can always be treated as being annotated
-with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so substitution does not
-change any coeffects.
+with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so we can also treat $e$
+as being annotated with coeffects $\ident{L}$. As a result, substitution does not change the coeffect.
 
@@
 algebra for which typing preservation holds. Proving the type preservation separately provides 
-more insight into how the systems work and so we choose to consider separately.
+more insight into how the systems work. We consider the two cases separately, but find a more general
+formulation for both of them.
 
@@
 does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
-our implicit parameters example (all three operators are set union) and for other set-like 
+our implicit parameters example (all three operators are a set union) and for other set-like 
 coeffects. It allows the following substitution lemma:
@@
 In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
-where $\czip = \cseq = \cpar$ and the operation is also idempotent and commutative and
+where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
 $\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
@@
 The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
-requirements (arising from the substitution) with the free-variable context. In contrast, the 
+requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
 structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
-coeffects associated with individual variables -- thus preserving typing in a complementary way.
-
-The two substitution lemmas show that the call-by-name evaluation strategy can be used for
-certain coeffect calculi, including liveness and implicit parameters. Assuming
-$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following subject reduction
+coeffects associated with individual variables, thus preserving typing in a complementary way.
+
+The two substitution lemmas discussed above show that the call-by-name evaluation strategy can be 
+used for certain coeffect calculi, including liveness and implicit parameters. Assuming
+$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following soundness
 theorem holds:
 
-\begin{theorem}[Call-by-name reduction]
+\begin{theorem}[Call-by-name soundness]
+\label{thm:cbn-flat}
 In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
 Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+and $e \rightarrow_{\ident{cbn}} e'$ then it is also the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
 \end{theorem}
@@
 For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
-to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \cleq \cunit$
+to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
 and $r \,\czip\, t \;=\; r \,\cpar\, t$.
@@
 $\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
-marking the context and function argument as live. In case of data-flow, the coeffects are
-made larger by the lambda abstraction. We remedy this limitation in the next chapter.
-
-However, the $\eta$-expansion preserves the type for implicit parameters and, more generally,
+marking the context and function argument as live. In case of data-flow, the immediate coeffects 
+are made larger by the lambda abstraction -- the context requirements of the function value are 
+imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
+the next chapter.
+
+However, $\eta$-expansion preserves the type for implicit parameters and, more generally,
 for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the 
@@
 
-\begin{theorem}[$\eta$-expansion]
+\begin{theorem}[Soundness of $\eta$-expansion]
 In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
@@
 \end{equation*}
-\end{proof}
-
-\noindent
+%
 The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
@@
 this is sufficient for showing type preservation.
-
+\end{proof}
+
+\noindent
 In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
@@
 The call-by-value and call-by-name evaluation strategies discussed in the previous section are
-the key techniques for defining equational theory of flat coeffects. In this section, we briefly 
-discuss another approach that follows the style of generic operational semantics designed by Filinski
-\cite{monads-inaction} for effectful computations.
-
-The idea is to embed sequencing of context-dependent computations as an explicit construct into
-the language and define a \emph{normal form} that consists of sequenced expressions. The reduction
-to the normal form is generic for all coeffect systems (satisfying certain conditions), while the
-reduction of the normal form is provided by each concrete coeffect system. The extended language
-with the \kvd{glet} construct and its typing is defined as follows:
+the most important techniques for defining equational theory of flat coeffects. In this section, 
+we briefly discuss another approach that follows the style of generic operational semantics 
+designed by Filinski \cite{monads-inaction} for effectful computations.
+
+The idea is to embed sequencing of context-dependent (effectful) computations as an explicit 
+construct into the language and define a \emph{normal form} that consists of sequentially composed 
+context-dependent expressions. The reduction to the normal form is generic for all coeffect systems 
+(satisfying certain conditions), while the reduction of the normal form is provided by each 
+concrete coeffect system. The extended language with the \kvd{glet} construct and its typing is 
+defined as follows:
 %
@@
 \begin{array}{rcl}
-e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2 \sep \kvd{glet}~x = e_1~\kvd{in}~e_2\\
+e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2 \\
 n &::=& e \sep \kvd{glet}~x=e~\kvd{in}~n\\
@@
 \end{equation*}
-%
+
 The newly introduced syntactic form $n$ represents a normal form. The construct $\kvd{glet}~x=e~\kvd{in}~n$
@@
 representing explicit sequencing of $e_1$ and $e_2$. The remaining two rules specify how \kvd{glet}
-distributes with other constructs (\kvd{glet} and application). In (\emph{glet-app}), the \kvd{glet}
+distributes across other constructs (\kvd{glet} and application). In (\emph{glet-app}), the \kvd{glet}
 construct appearing inside an application is lifted to the top-level; similarly (\emph{glet-glet}) 
-lifts a \kvd{glet} construct nested in the argument of another \kvd{glet}.
+lifts a \kvd{glet} construct nested in the argument of another \kvd{glet}. The rules provide a way
+for turning an expression into a normal form.
 
@@
 
-\caption{Reduction to normal form.}
+\figcaption{Reduction to normal form.}
 \label{fig:flat-normal-form}
@@
 As discussed in Section~\ref{sec:flat-calculus-let}, this condition is satisfied for all our
-three examples (strictly for liveness and data-flow; and by equality for implicit parameters).
+three examples, strictly for liveness and data-flow and by equality for implicit parameters.
 For the (\emph{glet-app}) and (\emph{glet-glet}) rules, additional conditions arise from the
-typing of the original and reduced expression (showed in Appendix~?). The results are summarized
-in the following two tables.
+typing of the original and reduced expression. The results are summarized in the following two tables.
 
@@
 \hspace{-1em}{\footnotesize Parameters} 
-  & $\mathit{max}(\cclrd{r_1} + \cclrd{r_2}, \cclrd{r_3}+\cclrd{s})$
-  & $\cclrd{r_1}+\mathit{max}(\cclrd{r_2}, \cclrd{r_3} + \cclrd{s})$ & $\times$ \\ \hline
+  & $(\cclrd{r_2} \cup \cclrd{r_1}) \cup (\cclrd{r_3} \cup \cclrd{s}) $ 
+  & $(\cclrd{r_2} \cup (\cclrd{r_3} \cup \cclrd{s})) \cup \cclrd{r_1} $ & \checkmark \\ \hline
 \hspace{-1em}{\footnotesize Liveness} 
@@
 \hspace{-1em}{\footnotesize Data-flow} 
-  & $(\cclrd{r_2} \cup \cclrd{r_1}) \cup (\cclrd{r_3} \cup \cclrd{s}) $ 
-  & $(\cclrd{r_2} \cup (\cclrd{r_3} \cup \cclrd{s})) \cup \cclrd{r_1} $ & \checkmark \\ \hline
+  & $\mathit{max}(\cclrd{r_1} + \cclrd{r_2}, \cclrd{r_3}+\cclrd{s})$
+  & $\cclrd{r_1}+\mathit{max}(\cclrd{r_2}, \cclrd{r_3} + \cclrd{s})$ & $\times$ \\ \hline
 \end{tabular}
@@
 
+\noindent
 Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
@@
 
+\noindent
 This means that \emph{internalized sequencing} provides a basis for operational semantics of
-liveness and implicit parameters, but not for data-flow languages. For other coeffect systems,
-the conditions in Appendix~? have to be re-examined.
+liveness and implicit parameters, but (again) not for data-flow languages. For other coeffect 
+systems, the conditions arising from the syntactic reductions have to be re-examined.
 
@@
 The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of laws.
-The laws are required for three distinct reasons -- to be able to define the categorical structure
+The laws are required for three reasons -- to be able to define the categorical structure
 in Section~\ref{sec:flat-semantics}, to prove equational properties in Section~\ref{sec:flat-syntax}
@@
 In this section, we look at the last point. We discuss what syntactic equivalences are permitted
-by the properties of $\czip$ and we extend the calculus with pairs and units and discuss their 
-syntactic properties. In the following section, we further develop subtyping relation for the 
-calculus.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Subtyping for coeffects}
-
-The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
-which makes it possible to treat an expression with smaller context requirements as an expression
-with greater context requirements. In the corresponding categorical semantics, this means that
-we can \emph{drop} some of the provided context.
-
-Figure~\ref{fig:flat-types-sub} adds sub-typing on function types, making it possible to treat
-a function with smaller context requirements as a function with greater context requirements. 
-The definition uses the standard reflexive and transitive $<:$ operator. As the (\emph{sub-fun})
-shows, the function type is contra-variant in the input and co-variant in the output. The 
-(\emph{typ}) rule allows using sub-typing on an expression type in the coeffect calculus.
+by the properties of $\czip$ (Section~\ref{sec:flat-exts-lambda}) and we extend the calculus with 
+pairs and units and discuss their syntactic properties (Section~\ref{sec:flat-exts-tup}). Before
+doing that, the following section further develops subtyping relation for the calculus.
 
@@
 
-\caption{Subtyping rules for flat coeffect calculus}
+\figcaption{Subtyping rules for flat coeffect calculus}
 \label{fig:flat-types-sub}
@@
 % --------------------------------------------------------------------------------------------------
-
-\paragraph{Semantics.} We follow the same approach as with the rest of the calculus and use
-a categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
+~
+
+~
+
+\subsection{Subtyping for coeffects}
+\label{sec:flat-exts-sub}
+
+The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
+which makes it possible to treat an expression with smaller context requirements as an expression
+with greater context requirements. In the corresponding categorical semantics, this means that
+we can \emph{drop} some of the provided context. However, sub-coeffecting only affects the 
+immediate coeffects attached to the free-variable context.
+
+In Figure~\ref{fig:flat-types-sub}, we add sub-typing on function types, making it possible to treat
+a function with smaller context requirements as a function with greater context requirements. 
+The definition uses the standard reflexive and transitive $<:$ operator. As the (\emph{sub-fun})
+shows, the function type is contra-variant in the input and co-variant in the output. The 
+(\emph{typ}) rule allows using sub-typing on expressions in the coeffect calculus.
+
+\paragraph{Semantics.} We follow the same approach as in Section~\ref{sec:flat-semantics} and use
+categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
 of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
@@
 
-The rest of the Figure~\ref{fig:flat-semantics-sub} shows the rules that define the semantics
-of $<:$. The reflexivity and transitivity are just the identity function and function composition,
-respectively. The (\emph{sub-fun}) case is interesting -- recall that the semantics of a function
-$\tau_1' \xrightarrow{\cclrd{r'}} \tau_2'$ is $C^{\cclrd{r'}} \tau_1' \rightarrow \tau_2'$. To 
-build the required function, we first drop unnecessary context using $\ident{lift}_{\cclrd{r'}, \cclrd{r}}
- : C^{\cclrd{r'}} \tau_1' \rightarrow C^{\cclrd{r}} \tau_1'$ and use the $\ident{map}_{\cclrd{r}}$
-function to transform the nested $\tau_1'$ to $\tau_1$. Then we evaluate the original function $f$
-and turn the resulting $\tau_2$ into the required result of type $\tau_2'$.
+The rest of the Figure~\ref{fig:flat-semantics-sub} defines the semantics of the $<:$ operation. 
+The reflexivity and transitivity are just the identity function and function composition,
+respectively. The (\emph{sub-fun}) case is interesting -- recall that the semantics of functions:
+%
+\begin{equation*}
+\sem{\tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} ~=~ C^{\cclrd{r'}} \tau_1' \rightarrow \tau_2'
+\end{equation*}
+%
+We build the transformation using an explicit lambda abstraction that takes the original function
+$f$ as an argument. To build the required function, we first drop unnecessary context using 
+$\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ of type $C^{\cclrd{r'}} \tau_1' \rightarrow C^{\cclrd{r}} \tau_1'$,
+then we use the $\ident{map}_{\cclrd{r}}$ function to transform the nested $\tau_1'$ to $\tau_1$. 
+Finally we evaluate the original function $f$ and turn the resulting value of type $\tau_2$ into 
+the required result of type $\tau_2'$.
 
@@
 \begin{equation*}
+\begin{array}{lll}
 \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' } = 
-  \sem{\tau <: \tau'} \circ \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } \qquad\quad (\emph{typ})
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{ll}
-\sem{\tau <: \tau} = \ident{id} & (\emph{sub-refl})
+  \sem{\tau <: \tau'} \circ \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } &\qquad& (\emph{typ})
+\\[1.75em]  
+\sem{\tau <: \tau} = \ident{id} && (\emph{sub-refl})
 \\[0.5em]
-\sem{\tau_1 <: \tau_3} = \sem{\tau_2 <: \tau_3} \circ \sem{\tau_1 <: \tau_2} & (\emph{sub-trans})
+\sem{\tau_1 <: \tau_3} = \sem{\tau_2 <: \tau_3} \circ \sem{\tau_1 <: \tau_2} && (\emph{sub-trans})
 \\[0.5em]
-\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} = \lambda f. & (\emph{sub-fun}) \\
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} = \lambda f. && (\emph{sub-fun}) \\
 \qquad\sem{\tau_2 <: \tau_2'} \circ f \circ \ident{map}_{\cclrd{r}}~\sem{\tau_1' <: \tau_1} 
@@
 
-\caption{Semantics of subtyping for flat coeffects}
+\figcaption{Semantics of subtyping for flat coeffects}
 \label{fig:flat-semantics-sub}
@@
 
-\subsection{Alternative lambda abstraction}
+\subsection{Properties of lambda abstraction}
 \label{sec:flat-exts-lambda}
@@
 In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
-lambda abstraction (for implicit parameters, the lambda function splits context requirements
-using $\cclrd{r} \cup \cclrd{s}$; for data-flow it suffices to duplicate the requirement 
-$\cclrd{r}$). We introduced the $\czip$ operation as a way of providing the additional abstraction.
-Here, we identify coeffect calculi for which the simpler (duplicating) rule is sufficient.
-
-\paragraph{Idempotence.}
-Recall that $(\C, \czip)$ is a band, meaning that $\czip$  is idempotent and associative. The 
+lambda abstraction -- for implicit parameters, the lambda function needs to split context 
+requirements using $\cclrd{r} \cup \cclrd{s}$, but for data-flow and liveness it suffices to 
+duplicate the requirement $\cclrd{r}$ of the body. We introduced the $\czip$ operation as 
+a way of providing the additional abstraction.
+
+In this section, we first identify coeffect calculi for which the simpler (duplicating) 
+rule is sufficient. Then we look at syntactic transformations corresponding to other common 
+properties of the $\czip$ operation.
+
+\paragraph{Simplified abstraction.}
+Recall that $(\C, \czip)$ is a band, that is, $\czip$  is idempotent and associative. The 
 idempotence means that the context requirements of the body can be required from both the 
@@
 % 
-To derive (\emph{idabs}), we use idempotence on the body annotation $\cclrd{r}\;=\;\cclrd{r}\,\czip\,\cclrd{r}$
+To derive (\emph{idabs}), we use idempotence on the body annotation $\cclrd{r}\;=\;\cclrd{r}\;\czip\;\cclrd{r}$
 and then use the standard (\emph{abs}) rule. So, (\emph{idabs}) follows from (\emph{abs}), 
-but the other direction is not necessarily the case. The condition below identifies 
-coeffect calculi where (\emph{abs}) follows from (\emph{idabs}).
+but the other direction is not necessarily the case. The following condition identifies 
+coeffect calculi where (\emph{abs}) can be derived from (\emph{idabs}).
 
@@
 \begin{proof}
-The following derives the conclusion of (\emph{abs}) using (\emph{abs}), sub-coeffecting, 
+The following derives the conclusion of (\emph{abs}) using (\emph{idabs}), sub-coeffecting, 
 sub-typing and the fact that the algebra is \emph{stricly oriented}:
@@
      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r} \,\czip\, \cclrd{s}} \tau_2} \;
-           \textnormal{\footnotesize{($\cclrd{r} \cleq \cclrd{r} \,\czip\, \cclrd{s}$)}} \hspace{-5em} }
+           \textnormal{\footnotesize{($\cclrd{r} \;\cleq\; \cclrd{r} \,\czip\, \cclrd{s}$)}} \hspace{-5em} }
   {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} \; 
-           \textnormal{\footnotesize{($\cclrd{r} \cleq \cclrd{r} \,\czip\, \cclrd{s}$)}}
+           \textnormal{\footnotesize{($\cclrd{r} \;\cleq\; \cclrd{r} \,\czip\, \cclrd{s}$)}}
 \end{equation*}
@@
 the (\emph{idabs}) rule and get an equivalent type system. This alternative formulation
-removes the non-determinism that arises from the splitting of context requirements in 
-the original (\emph{abs}) rule.
+removes the non-determinism of type checking that arises from the splitting of context 
+requirements in the original (\emph{abs}) rule. Furthermore, for data-flow and liveness
+the (\emph{idabs}) rule is more precise than (\emph{abs}).
 
@@
 
-\caption{Typing rules for pairs and units}
+\figcaption{Typing rules for pairs and units}
 \label{fig:flat-ext-types}
@@
 \subsection{Language with pairs and unit}
-
-The calculus introduced in Section~\ref{sec:flat-calculus} consisted only of variables, abstraction, 
-application and let binding to show the key aspects of flat coeffect systems. Here, we extend it 
-with pairs and the unit value to sketch how it can be turned into a full programming language.
-The syntax of the language is extended as follows:
+\label{sec:flat-exts-tup}
+
+To show the key aspects of flat coeffect systems, the calculus introduced in Section~\ref{sec:flat-calculus} 
+consists only of variables, abstraction, application and let binding. Here, we extend it 
+with pairs and the unit value to sketch how it can be turned into a more complete programming 
+language and to motivate the laws required about $\cpar$. The syntax of the language is extended as follows:
 %
@@
 models the case when the (same) available context is split and passed to two independent 
-sub-expressions. This matches the semantics of pairs discussed shortly. Finally, the (\emph{proj})
-rule is uninteresting, because $\pi_i$ can be viewed as a pure function.
+sub-expressions. Finally, the (\emph{proj}) rule is uninteresting, because $\pi_i$ can be 
+viewed as a pure function.
 
 \paragraph{Properties.}
-Pairs and the unit value in a lambda calculus should form a monoid -- associativity means that 
-the expression $(e_1, (e_2, e_3))$ should be isomorphic to $((e_1, e_2), e_3)$ and that
-$((), e) \simeq e \simeq (e, ())$, where the isomorphism appropriately transforms the values,
-without affecting other properties (here coeffects) of the expression.
-
+Pairs and the unit value in a lambda calculus typically form a monoid. Assuming $\simeq$ is an
+isomorphism that performs appropriate transformation on values, without affecting other 
+properties (here, coeffects) of the expressions. The monoid laws then correspond to 
+the requirement that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
+requirement that $((), e) \simeq e \simeq (e, ())$ (unit).
+
+Thanks to the properties of $\cpar$, the flat coeffect calculus obeys the monoid laws for pairs. 
 In the following, we assume that \ident{assoc} is a pure function transforming a pair $(x_1, (x_2, x_3))$ 
@@
 \begin{theorem}
+\label{thm:flat-tup-eq}
 For a flat coeffect calculus with pairs and units, the following holds:
@@
 
-The above properties follow from the laws of the flat coeffect algebra. In addition, if the 
-$\cpar$ operation is symmetric (which is the case for all our examples in this chapter), it 
-also holds that $\ident{swap}~(e_1, e_2) \equiv (e_2, e_1)$.
+\noindent
+The Theorem~\ref{thm:flat-tup-eq} motivates the requirement of the monoid structure 
+$(\C, \cpar, \czero)$ of the flat coeffect algebra. We require only unit and associativity
+laws. In our three examples, the $\cpar$ operator is also symmetric, which additionally
+gives us the property that $(e_1, e_2) \simeq (e_2, e_1)$.
 
@@
 and we covered work related to individual concepts throughout the chapter. In this section, we
-do not repeat the discussion present elsewhere -- we discuss one specific question that often 
-arises when discussing coeffects and that is \emph{when is coeffect (not) an effect?}
+do not repeat the discussion present elsewhere. Instead, we discuss one specific question that often 
+arises when discussing coeffects and that is \emph{when is a coeffect (not) an effect?}
 
@@
 \begin{itemize}
-\item Semantically, coeffects capture very different notions of computation. As demonstrated in 
+\item Semantically, coeffects capture different notions of computation. As demonstrated in 
   Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
@@
 \item Syntactically, coeffect calculi use a richer algebraic structure with pointwise composition,
-  sequential composition and context merging $(\cpar, \cseq, \czip)$ while most effect systems
-  only use a single operation for sequential composition (monadic bind).
+  sequential composition and context merging ($\cpar, \cseq$, and $\czip$) while most effect systems
+  only use a single operation for sequential composition (used by monadic bind).
 
 \item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In 
-  coeffect systems, the context requirements of the body can be split between declaration-site
-  and call-site, while monadic effect systems delay all effects.
+  coeffect systems, the context requirements of the body can be split between (or duplicated
+  at) declaration site and call site, while monadic effect systems always defer all effects.
 \end{itemize}
@@
 Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
-we use the $\cup$ operation for all three operations of the coeffect algebra. However, the last
-point requires us to extend monadic lambda abstraction.
+we use the $\cup$ operation for all three operations of the coeffect algebra. However, to enable
+splitting of implicit parameter requirements using the reader monad, we need to extend the 
+monad structure and change the translation of monadic lambda abstraction.
 
@@
 
-As discussed in Section~\ref{sec:applications-flat-impl}, one of our examples, implicit 
-parameters, can be also captured by a monad. However, \emph{just} a monad is not enough
-because lambda abstraction in effect systems does not provide a way of splitting the context
-requirements between declaration-site and call-site (or combining the implicit parameters
-available in the scope where the function is deined and those specified by the caller).
+Implicit parameters can be captured by a monad, but \emph{just} a monad is not enough.
+Lambda abstraction in effect systems does not provide a way of splitting the context
+requirements between declaration site and call site (or, semantically, combining the implicit 
+parameters available in the scope where the function is deined and those specified by the caller).
 
@@
 Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
-are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using
-the currying/uncurrying isomoprhism.
+are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
+reader monad via currying/uncurrying isomoprhism.
 \end{remark}
@@
 \begin{equation*}
-\begin{array}{l}
- \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 = \\
+\begin{array}{lll}
+ \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
  (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
- \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) \\
+ &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
  \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
@@
 \noindent
-The equivalence holds for monads and comonads (as well as \emph{indexed} monads
+This equivalence holds for monads and comonads (as well as \emph{indexed} monads
 and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
@@
 \paragraph{Delaying effects in monads.} 
-In the syntax of the language, the above difference is manifested in the (\emph{abs}) rules for 
+In the syntax of the language, the above difference is manifested by the (\emph{abs}) rules for 
 monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
-side-by-side, using the coeffect notation for both of them:
+side-by-side, using the effect system notation for both of them:
 %
@@
 \tyrule{cabs}
-  { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \vdash e : \tau_2 }
-  { \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s} }
+  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 \;\&\; \cclrd{r}}
 \quad
 \tyrule{mabs}
-  { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \vdash e : \tau_2 }
-  { \coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 }
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s}}
+  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 \;\&\; \cclrd{\emptyset} }
 \qquad
@@
 In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
-the monadic rule (\emph{mabs}) places all requirements on the call-site. This follows from the
+the monadic rule (\emph{mabs}) places all requirements on the call site. This follows from the
 fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
@@
 In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
-the declaration-site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
-call-site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
+the declaration site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
+call site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
 (which we call \ident{delay}) with the following signature:
@@
 annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context requirements
-$\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function.
+$\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function at run-time.
 
@@
 
-This is, indeed, also possible. Recall that the semantics of lambda abstraction in the flat
+This is, indeed, possible. Recall that the semantics of lambda abstraction in the flat
 coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
-two contexts (wrapped in a comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried values and
-additional contextual information (implicit parameters). To obtain the (\emph{mabs}) rule, we 
-can restrict the first parameter, which corresponds to the declaration-site context:
+two contexts (wrapped in an indexed comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried 
+values and additional contextual information (implicit parameters). To obtain the (\emph{mabs}) 
+rule, we can restrict the first parameter, which corresponds to the declaration site context:
 %
@@
 %
-In the (\emph{restricted}) version of the operation, the declaration-site context requires
-no implicit parameters and so all implicit parameters have to be satisfied by the call-site.
+In the (\emph{restricted}) version of the operation, the declaration site context requires
+no implicit parameters and so all implicit parameters have to be satisfied by the call site.
 The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.
@@
 could be obtained just by changing the definition of $\czip$. Similarly, we could obtain \eg~a
-fully lexically-scoped version of the system. Similar idea has been used for the
-semantics of effectful computations by Tate~\cite{effects-producer-semantics}.
+fully lexically-scoped version of the system. The ability to restrict operations to partial 
+functions has been used in the semantics of effectful computations by Tate~\cite{effects-producer-semantics}.
 
@@
 This chapter presented the \emph{flat coeffect calculus} -- a unified system for tracking 
-contextual properties that are \emph{whole-context}, meaning that they are related to the 
-execution environment or the enitre context in which computations are executed.
-This is the first of the two \emph{coeffect calculi} developed in this thesis.
+\emph{whole-context} properties of computations, that is properties related to the 
+execution environment or the enitre context in which programs are executed.
+This is the first of the three \emph{coeffect calculi} developed in this thesis.
 
 The flat coeffect calculus is parameterized by a \emph{flat coeffect algebra} that captures
-the structure of the information tracked by the type system. We demonstrated how to 
-instantiate the system to capture three specific systems discussed earlier in 
-Chapter~\ref{ch:applications}, namely liveness, data-flow and implicit parameters.
-
-Next, we introduced the notion of \emph{flat indexed comonad}, which is a generalization of
-comonad, equipped with additional operations needed to provide categorical semantics of the 
-flat coeffect calculus. The indices of the flat indexed comonad operations correspond to the
+the structure of the information tracked by the type system. We instantiated the system to 
+capture three specific systems, namely liveness, data-flow and implicit parameters. However,
+the system is more general and an capture numerous other applications outlined in Section~\ref{sec:applications-flat}.
+
+Next, we introduced the notion of \emph{flat indexed comonad}, which generalizes of comonads
+and adds additional operations needed to provide categorical semantics of the flat coeffect 
+calculus. The indices of the flat indexed comonad operations correspond to the
 coeffect annotations in the type system and provide a foundation for the design of the 
@@
 syntactic properties that hold for all flat coeffect systems satisfying certain additional
-conditions. In particular, two \emph{subject reduction} theorems prove that the operational
+conditions. In particular, two \emph{typing preservation} theorems prove that the operational
 semantics for two classes of flat coeffect calculi (including liveness and implicit parameters) 
-can be based on call-by-name reduction.
+can be based on the standard call-by-name reduction.
 

### 2014-12-12 text/Chapter05.tex
Proofreading and English conventions (Chapter 4)

@@
 
-\chapter{Structural coeffect calculus} 
+\chapter{The structural coeffect calculus} 
 \label{ch:structural}
@@
 In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
-available from the declaration-site and the call-site or, in the syntactic reading, to split the 
+available from the declaration site and the call site or, in the syntactic reading, to split the 
 context requirements.

### 2014-12-12 text/Chapter06.tex
Finish proofreading and tweaking (Chapter 3)
Proofreading and English conventions (Chapter 4)

@@
 
-\chapter{Unified coeffect language} 
+\chapter{Unified coeffect language ???} 
 \label{ch:unified}
@@
 \item[-] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
-\item[-] $B^A$ is an exponential object $A \Rightarrow B$ in cartesian closed category
+\item[-] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
 \end{compactitem}
@@
 the context-requirements of the body should be smaller than the context available on the 
-call-site and declaration-site. In application, we require more than what is required by
+call site and declaration site. In application, we require more than what is required by
 the sub-expressions.

### 2014-12-12 text/backup/Chapter01.tex
Proofreading and English conventions (Chapter 4)

@@
 \emph{context-dependence} does not match this pattern. In the systems we consider, lambda 
-abstraction places requirements on both the \emph{call-site} (latent requirements) and the 
-\emph{declaration-site} (immediate requirements), resulting in different syntactic properties. 
+abstraction places requirements on both the \emph{call site} (latent requirements) and the 
+\emph{declaration site} (immediate requirements), resulting in different syntactic properties. 
 We informally discuss three examples first that demonstrate how contextual requirements propagate.
@@
 context requirement of the function body is a set of resources $\{\ident{Clock}, \ident{News}\}$.
-The context requirements are split between the \emph{declaration-site} and \emph{call-site}.
+The context requirements are split between the \emph{declaration site} and \emph{call site}.
 However, there are multiple possible splittings. The splitting $\{\ident{News}\} \cup \{\ident{Clock}\}$
@@
 For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
-\ident{b} from the declaration-site) then the function would have context requirements $5$ -- that
+\ident{b} from the declaration site) then the function would have context requirements $5$ -- that
 is the number associated with the variable \ident{a}.

### 2014-12-13 text/Chapter04.tex
Proofreading and editing (Chapter 5)

@@
   \emph{flat coeffect algebra} (Section~\ref{sec:flat-calculus}). We show that the system can be 
-  instantiated to obtain three of the systems discussed in previous section (implicit parameters, 
-  liveness and data-flow).
+  instantiated to obtain three of the systems discussed in Section~\ref{sec:applications-flat},
+  namely implicit parameters, liveness and data-flow.
 
@@
 Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
-$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the soundness theorem is
+$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the type preservation theorem is
 stated as follows:
 
-\begin{theorem}[Call-by-value soundness]
+\begin{theorem}[Type preservation for call-by-value]
 \label{thm:cbv-reduction}
@@
 used for certain coeffect calculi, including liveness and implicit parameters. Assuming
-$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following soundness
-theorem holds:
-
-\begin{theorem}[Call-by-name soundness]
+$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following theorem holds:
+
+\begin{theorem}[Type preservation for call-by-name]
 \label{thm:cbn-flat}
@@
 
-\begin{theorem}[Soundness of $\eta$-expansion]
+\begin{theorem}[Type preservation of $\eta$-expansion]
 In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 

### 2014-12-13 text/Chapter05.tex
Proofreading and editing (Chapter 5)

@@
 
-As already discussed, the aim of this thesis is to identify abstractions for context-aware
-programming languages. We attempt to find abstractions that are general enough to capture a wide
-range of useful programming language features, but specific enough to let us identify
-interesting properties of the languages.
-
-In Chapter~\ref{ch:applications}, we identified two notions of context. We generalized the 
-class of flat calculi that capture whole-context properties in Chapter~\ref{ch:flat}. 
-In this chapter, we turn our attention to \emph{structural} coeffect calculi that capture
-per-variable properties.
-
-The flat coeffect system captures interesting use-cases (implicit parameters, liveness and
-data-flow), but provides relatively weak properties. We can define its categorical semantics,
-but the equational theory proofs had numerous additional requirements. For this reason, it
-is worthwhile to consider structural systems in a separate chapter. We will see that 
-structural coeffects have a number of desirable properties that hold for all instances of the
-calculus.
+In this chapter, we take the next step towards a unified theory of context-aware programming
+languages. As already discussed, we attempt to find abstractions that are \emph{general enough} 
+to capture a wide range of useful programming language features, but \emph{specific enough} to 
+let us identify and capture interesting properties of the context.
+
+In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware prorgamming
+languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
+calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
+and we introduce the \emph{structural coeffect calculus}.
+
+The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness 
+and data-flow), but provides relatively imprecise approximation and weak syntactical properties. 
+Some of those examples can be also seen as per-variable properties. For those, structural coeffect
+systems give a more precise information about the context. However, we also look at other applications 
+that arise from the work on sub-structural logics, a pathway to coeffects discussed in 
+Section~\ref{sec:path-logic}.
+
+As we will see, the categorical semantics of the structural coeffect calculus is based on the
+same underlying structure as the semantics of the flat calculus. However, the syntactical properties
+of the calculus are quite different and so it is worth considering it in a separate chapter.
+
+\paragraph{Chapter structure and contributions}
+\begin{itemize}
+\item[--] We present a \emph{structural coeffect calculus} as a type system that is parameterized 
+  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the 
+  system captures fine-grained liveness and data-flow information, as well as a calculus for bounded 
+  reuse (checking how many times is a variable accessed).
+
+\item[--] We extend indexed comonads introduced in the previous chapter to 
+  \emph{structural indexed comonads} and use them to provide the semantics of structural 
+  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
+  semantics provides a motivation for the design of the calculus.
+
+\item[--] We discuss the equational theory of the calculus. For structural coeffects, the substitution 
+  lemma holds for all instances of the calculus. Thus, we prove the type-preservation property 
+  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}). 
+\end{itemize}
 
@@
 
-Two examples of flat systems from the previous chapter were liveness and data-flow. As discussed
-in \ref{sec:applications-structural}, these are interesting for theoretical reasons. However, tracking
-liveness of the whole context is not practically useful. Structural versions of liveness and 
-data-flow let us track more fine-grained properties. Moreover, the equational theory of flat 
-coeffect calculus did not reveal many useful properties for flat liveness and data-flow. 
-As we show in this chapter, this is not the case with structural versions.
-
-In this chapter, we focus on three example applications. We look at structural liveness and
-data-flow and we also consider calculus for bounded reuse, which checks how many times a variable
-is accessed and generalizes linear logics (that restrict variables to be used exactly once).
+Compared to the previous chapter, structural coeffect calculi we consider are more 
+homogeneous and so finding the common pattern is in some ways easier. However, the systems
+are more complicated as they need to keep annotations attached to individual variables
+and thus require explicit structural rules. Before looking at the system, we briefly consider
+the most important related work.
 
@@
 
-\subsection{Contributions}
-
-Compared to the previous chapter, the structural coeffect calculi we consider are more 
-homogeneous and so finding the common pattern is in some ways easier. However, the systems
-are somewhat more complicated as they need to keep annotations attached to individual variables. 
-The contributions of this chapter are as follows:
-
-\begin{itemize}
-\item We present a \emph{structural coeffect calculus} with a type system that is parameterized by a 
-  \emph{structural coeffect algebra} and can be instantiated to obtain all of the three examples
-  discussed (Section~\ref{sec:struct-calculus}).
-
-\item We give the equational theory of the calculus. We prove the type-preservation property 
-  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}).
-
-\item We show how to extend indexed comonads introduced in the previous section to 
-  \emph{structural indexed comonads} and use them to give the semantics of structural 
-  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
-  categorical semantics provides a motivation for the design of the calculus.
-\end{itemize}
-
-%---------------------------------------------------------------------------------------------------
-
 \subsection{Related work}
@@
 In the previous chapter, we discussed the correspondence between coeffects and effects (and 
-between comonads and monads). As noted earlier, the $\lambda$-calculus is assymetric in that
+between comonads and monads). As noted earlier, the $\lambda$-calculus is asymmetric in that
 an expression has multiple inputs (variables in the context), but just a single result 
@@
 
-Our work follows the language semantics style in that we provide a structural semantics to
-the terms of ordinary $\lambda$-calculus. The most closely related work has been done in the
-meta-langauge style, which extends the terms and types with constructs working with the context
-explicitly. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
+Our work follows the ``language semantics'' style in that we provide a structural semantics to
+the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
+the meta-langauge style, which extends the terms and types of a language with constructs for explicitly
+manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
 may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
 $A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
-annotations as stand-alone types.
-
-Structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
+annotations as stand-alone types (indexed comonads only appear in the semantics).
+
+Our structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
 For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
 respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
-This means that the typing judgements have the same structure as those of the flat coeffect
-calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify the two 
-systems and compose tracking of flat and structural properties.
+A benefit of this approach is that the typing judgements have the same structure as those of the 
+flat coeffect calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify 
+the two systems.
 
@@
 
-In the structural coeffect calculus, a vector of variables in the free-variable context is 
+In the structural coeffect calculus, a vector of variables forming the free-variable context is 
 annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
-for different coeffect calculi and their properties are captured by the \emph{structural 
-coeffect scalar} definition below. The scalar annotations can be integers (how many past values we 
-need) or annotations specifying whether a variable is live or not.
+for different coeffect calculi and their properties are captured by the definition of \emph{structural 
+coeffect scalar} below. The scalar annotations can be \eg~integers (how many past values we 
+need) or values of a two-point lattice specifying whether a variable is live or not.
 
@@
 the previous chapter). Functions always have exactly one input variable and so they are annotated
-with a coeffect scalar. Thus the expressions and types of structural coeffect calculi are the same
-as in the previous chapter (except that annotation on function type is now a structural coeffect 
-scalar):
+with a single coeffect scalar. Thus the expressions and types of structural coeffect calculi are the 
+same as in the previous chapter. The only difference is that $\cclrd{r}, \cclrd{s}, \cclrd{t}$ now
+range over values of a \emph{structural coeffect scalar} rather than over values of a \emph{flat 
+coeffect algebra} (the definitions are related, but not the same).
 %
@@
 %
-In the previous chapter, the free variable context $\Gamma$ has been treated as a set. In the 
-structural coeffect calculus, the order of variables matters. Thus we treat free variable context
-as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
+In the previous chapter, the free variable context $\Gamma$ was treated as a set. In the 
+structural coeffect calculus, the order of variables matters. Thus we treat a free variable 
+context as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
 of the vector:
@@
 \begin{array}{rcl}
- \Gamma &\narrow{=}& \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
+ \Gamma &\narrow{=}& \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
  \xlen{\Gamma} &\narrow{=}& n
@@
 typing judgements, but the free variable context should be understood as a vector.
-Furthermore, the usual notation $\Gamma_1, \Gamma_2$ stands for the tensor product.
-Given $\Gamma_1 = \langle x_1\!:\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
-$\Gamma_2 = \langle x_{n+1}\!:\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
-$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
-
-The free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
+Furthermore, the notation $\Gamma_1, \Gamma_2$ is here seen as the tensor product.
+Given $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
+$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
+$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
+
+Free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
 we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
@@
 is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
-of coeffects is provided by the \emph{structural coeffect algebra} definition below.
+of coeffects is provided by the definition of \emph{structural coeffect algebra} below.
 
@@
 
-The structural coeffect scalar structure is similar to \emph{flat coeffect algebra} with the 
+The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the 
 exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
@@
 available from the declaration site and the call site or, in the syntactic reading, to split the 
-context requirements.
+context requirements. The other difference is that the $\cpar$ operation of structural coeffect
+scalar is not required to be idempotent. The bounded reuse example uses $\cpar$ which is not 
+idempotent.
 
 In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
-coeffects becomes just vector a concatenation or splitting, respectively, which is provided
+coeffects becomes just vector concatenation or splitting, respectively; these are provided
 by the tensor product. The operations on vectors are indexed by integers representing the lengths
@@
 
-The type system for structural coeffect calculus is similar to sub-structural type systems
+The type system for the structural coeffect calculus is similar to sub-structural type systems
 in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
@@
 in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
-allows weakening, exchange and contraction, but using explicit \emph{structural} rules that
+does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
 perform corresponding transformation on the coeffect annoation.
@@
 \paragraph{Syntax-driven rules.}
-The variable access rule (\emph{var}) annotates the corresponding variable as being used
+The syntax-driven rules of the type system are shown in Figure~\ref{fig:struct-types} (a).
+The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
 using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
@@
 the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
-non-determinism present in flat coeffect systems.
+non-determinism of the type inference present in flat coeffect systems.
 
@@
 The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
-They allow different transformation of the free variable context. We include sub-coeffecting
-(\emph{sub}) as one of the rules, allowing sub-coeffecting on coeffect scalars belonging
-to individual variables. The remaining rules capture \emph{weakening}, \emph{exchange} and
-\emph{contraction} known from sub-structural systems.
-
-The (\emph{weak}) allows adding a variable to the context, extending the coeffect
+They allow different transformation of the free variable context. We include (\emph{sub}) 
+as one of the rules, allowing sub-coeffecting on coeffect scalars belonging to individual 
+variables. The remaining rules capture \emph{weakening}, \emph{exchange} and \emph{contraction} 
+known from sub-structural systems.
+
+Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
 vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
@@
 variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
-allows re-using a variable as we can type check sub-expressions using two separate varaibles and 
+allows re-using a variable. We can type check sub-expressions using two separate varaibles and 
 then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
-is the only place in the structural coeffect system where context requiremens are combined,
-or semantically, where the same context is shared.
+is the only place in the structural coeffect system where context requiremens are combined
+(semantically, this is where the available context is shared).
 
@@
 
-\caption{Type system for the structural coeffect calculus}
+\figcaption{Type system for the structural coeffect calculus}
 \label{fig:struct-types}
@@
 This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
-as over- and under-approximations of $\atimes$. This is demonstrated by the (\emph{approximation})
-requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
+as over-approximations and under-approximations of $\atimes$. This is demonstrated by the 
+(\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
 $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
-combines available context, potentially discarding parts of it (under-approximation), while flat
-application splits available context, potentially duplicating parts of it (over-approxi\-mation)\footnote{
-Because of this duality, earlier version of coeffects in \cite{coeffects-icalp13}
-used $\wedge$ and $\vee$.}.
-
-Second, the operator $\cpar$ is used when the semantics passes the same context to multiple
-sub-expressions. In flat systems, this happens in (\emph{app}) and (\emph{pair}), because the 
-sub-expressions may share variables. In structural systems, this is separated into an 
+combines two values representing the available context, potentially discarding parts of it 
+(under-approximation), while flat application splits the available context (a single value), 
+potentially duplicating parts of it (over-approxi\-mation)\footnote{Because of this duality, earlier 
+version of coeffect systems \cite{coeffects-icalp13} used $\wedge$ and $\vee$.}.
+
+Secondly, the operator $\cpar$ is used when the semantics passes a given context to multiple
+sub-expressions. In flat systems, the context is shared in (\emph{app}) and (\emph{pair}), because 
+the sub-expressions may share variables. In structural systems, the sharing is isolated into an 
 explicit contraction rule.
@@
 let binding that gives a \emph{more precise} typing than combination of abstraction and 
-application. This is not the case for structural coeffects.
+application. Structural coeffect systems avoid this issue.
 
@@
 \label{thm:structural-letbinding}
-In a structural coeffect calculus, the typing of $(\lambda x.e_2)~e_1$ is equivalent
-to the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$.
+In a structural coeffect calculus, the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$ can be
+seen as a derived rule, \ie~its typing is equivalent to the typing of the expression
+$(\lambda x.e_2)~e_1$.
+
 \end{remark}
@@
 
-The structural coeffect calculus can be instantiated to obtain the structural coeffect calculi
+The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
 presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
 structural liveness provide a more precise tracking of properties that can be tracked using 
-flat systems. Formally, a flat coeffect algebra can be turned into a structural coeffect algebra
+flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
 (by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
-example, it is not clear why one would associate implicit parameters with individual variables.
+example, implicit parameters could be attached to individual variables, but, to our knowledge, this
+has not been used in any practical system.
 
@@
 because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
-algebra. This is the case, for example, for the bounded variable use.
+scalar. This is the case, for example, for the bounded variable use.
 
 \begin{example}[Structural liveness]
-The structural coeffect algebra for liveness is formed by
+The structural coeffect scalar for liveness is formed by
 $(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
 the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
-with a join $\sqcup$ and meet $\sqcap$. 
+with a join $\sqcup$ and a meet $\sqcap$. 
 \end{example}
@@
 \begin{example}[Structural data-flow]
-In data-flow, context is annotated with natural numbers and the flat coeffect algebra is formed 
+In data-flow, context is annotated with natural numbers and the structural coeffect scalar is formed 
 by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
@@
 \noindent
-For the two examples that have both flat and structural version, obtaining the strucutral coeffect
+For the two examples that have both flat and structural version, obtaining the structural coeffect
 algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
@@
 It is important to note that this gives us a systems with \emph{different} properties. The 
-information are now tracked per-variable rather than for the enitre context. For data-flow,
+information are now tracked per-variable rather than for enitre contexts. For data-flow,
 we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
 for a pointwise extension of the $+$ operator, such that
-$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} + k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
+$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
 %
@@
 
-When introducing structural coeffect systems in Section~\ref{sec:applications-structural}, we 
+When we introduced structural coeffect systems in Section~\ref{sec:applications-structural}, we 
 included a concrete semantics of structural liveness and bounded variable reuse. In this section,
 we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
-extension of \emph{indexed comonad} structure. As in the previous chapter, the main aim of this 
-section is to motivate and explain the design of the structural coeffect calculus shown in 
-Section~\ref{sec:struct-calculus}. The semantics highlights the similarities and differences 
-between the two systems.
-
-Most of the differences between flat and structural systems arise from the fact that contexts
-in structural coeffect systems are treated as \emph{vectors} rather than sets modelled using 
-categorical products, so we start by discussing our treatment of vectors.
+extension of the \emph{indexed comonad} structure. Similarly to \emph{flat indexed comonad}, the
+structural variant adds operations that are needed to embed full $\lambda$-calculus, this time 
+with per-variable contexts.
+
+As in the previous chapter, the main aim of this section is to motivate and explain the design of 
+the structural coeffect calculus shown in Section~\ref{sec:struct-calculus}. However, the semantics 
+also highlights the similarities and differences between the flat and structural system.
+Most of the differences arise from the fact that contexts in structural coeffect systems are treated 
+as \emph{vectors} rather than sets.
 
@@
 
-In the flat coeffect calculus, the context is interpreted as a product and so a typing judgement
-$\coctx{x_1\!:\tau_1, \ldots, x_n:\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a morphism 
-$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, we can 
-freely transform the value contained in the context modelled using an indexed comonad
-$\ctyp{\cclrd{r}}$. For example, the function $\ident{map}_{\cclrd{r}}~\pi_i$ transforms
-a context $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a value 
+Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing 
+judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a 
+morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, 
+we can freely transform the value contained in the context modelled using an indexed comonad
+$\ctyp{\cclrd{r}}$.
+
+Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
+which transforms the value inside the context without affecting the coeffect annotation. Thus,
+we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
+$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value 
 $\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
@@
 using a structure distinct from ordinary products which we denote $-\xtimes-$. 
-For example, the judgement $\coctx{x_1\!:\tau_1, \ldots, x_n:\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
+For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
 is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
+We assume that the operator is equipped with necessary associativity transformations allowing us
+to use it freely on more than two values.
 
@@
 The structure created using $-\xtimes-$ can be manipulated only using operations provided by
-the \emph{strucutral indexed comonad}, which operate over variable contexts contained in an 
-indexed comonad $\ctyp{\aclrd{\textbf{r}}}$.
+the \emph{structural indexed comonad}, which operate over variable contexts contained in an 
+indexed comonad $\ctyp{\aclrd{\textbf{r}}}$ and are designed to preserve the correspondence
+between vectors and annotations.
 
 In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
-We assume that the use of the operator can be freely re-associated. If an operation requires an 
-input in the form $(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
-we call it with $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ as an argument and assume that the 
+As mentioned, we assume that the use of the operator can be freely re-associated. For example, when
+calling an operation that requires input of the form 
+$(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
+we use an argument $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ and assume that the 
 appropriate transformation is inserted.
@@
 The semantics of structural coeffect calculus reuses the definition of \emph{indexed comonad} almost
-without a change. The additional structure that is required for context manipulation (merging and 
-splitting) is different and is provided by the \emph{structural indexed comonad} structure that we
-introduce in this section.
-
-Recall the definition from Section~\ref{sec:flat-semantics-idx}, which defines an indexed comonad 
+without change. The additional structure that is required for context manipulation (merging and 
+splitting) is different and is here provided by the \emph{structural indexed comonad} structure that 
+we introduce in this section.
+
+Recall the definition from Section~\ref{sec:flat-semantics-idx}, which defines an indexed como\-nad 
 over a monoid $(\C, \cseq, \cunit)$ as a triple $(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, 
 \ident{cobind}_{\cclrd{r}, \cclrd{s}})$. The triple consists of a family of object mappings $\ctyp{\cclrd{r}{}}$,
-and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau \rightarrow \tau'$.
-
-In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau \rightarrow \tau'$
+and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
+
+In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$
 representing function values (appearing in the language), but also of the form
 $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$, 
-modelling expressions in a context. To capture this, we need to generalize some of the indices from
-\emph{coeffect scalars} $\cclrd{r}, \cclrd{s}, \cclrd{t}$ to \emph{coeffect vectors} $\aclrd{r}, \aclrd{s}, \aclrd{t}$.
+modelling expressions in a context. To capture this, we need to revisit the definition and use 
+\emph{coeffect vectors} in some of the operations.
 
@@
 
-The operation $\ident{counit}_{\cunit}$ operates on a singleton-vector. This means that it will always
-return a single variable value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
+The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
+return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
 operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This assymmetry
@@
 takes a vector of variables, but always produces just a single value. Thus the resulting function 
-also takes a vector of variables, but always returns a context with singleton variable vector.
+also takes a vector of variables, but always returns a context with a vector containing just one value.
 In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
@@
 The flat indexed comonad structure extends indexed comonads with operations 
-$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
-that combine or split the additional (flat) context and are annotated with the flat 
-coeffect operations $\czip$ and $\cpar$, respectively. In the structural version, we use
-corresponding operations that operate on variable vectors represented using $\xtimes$ and
-are annotated with a tensor $\atimes$ which mirrors the variable structure.
-
-The following definition also includes $\ident{lift}_{\cclrd{r'},\cclrd{r}}$, which is similar as
+$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ that combine or 
+split the additional (flat) context and are annotated with the flat coeffect operations $\czip$ 
+and $\cpar$, respectively.
+
+In the structural version, the corresponding operations operate convert between a (wrapped) vector 
+of values represented using $\xtimes$ and ordinary pairs of contexts containing parts of the vector. 
+The vectors of coeffect annotations are split or merged using $\atimes$ of the structural coeffect 
+algebra, in a way that mirrors the wrapped vectors (variable structure).
+
+The following definition includes $\ident{lift}_{\cclrd{r'},\cclrd{r}}$, which is similar as
 before and models sub-coeffecting and also $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ which models
@@
 %
-Such that the following equalities hold:
-\begin{equation*}
-\begin{array}{l}
-  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \equiv \ident{id} \\
-  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \equiv \ident{id} 
-\end{array}
+Here, the following equalities must hold:
+\begin{equation*}
+\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~\equiv~ \ident{id} \qquad\qquad
+\ident{id} ~\equiv~ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} 
 \end{equation*}
@@
 \noindent
-The operations differ from those of the flat indexed comonad in that the \ident{merge} and 
+These operations differ from those of the flat indexed comonad in that the \ident{merge} and 
 \ident{split} operations are required to be inverse functions and to preserve the additional 
 information about the context. This was not required for the flat system where the operations
-could under- or over-approximate. Note that the operations use $\xtimes$ to combine or
-split the contained values. This means that they operate on free-variable vectors rather than 
+could under-approximate or over-approximate. Note that the operations use $\xtimes$ to combine 
+or split the contained values. This means that they operate on free-variable vectors rather than 
 on ordinary products.
@@
 The \ident{dup} mapping is a new operation that was not required for a flat calculus. It 
-takes a variable context with a single variable annotated with $\cclrd{r}\cpar\cclrd{s}$,
+takes a variable context with a single variable annotated with $\cclrd{r}\,\cpar\,\cclrd{s}$,
 duplicates the value of the variable $\alpha$ and splits the additional context between the two
-new variables. In flat calculus, this operation has been expressed using ordinary tuple 
+new variables. In a flat calculus, this operation was expressed using ordinary tuple 
 construction, which is not possible here -- the returned context needs to contain a 
@@
 operates on a singleton vector, which is equivalent to operating on a scalar as before.
-The operation could easily be extended to a vector in a pointwise way, but we keep it simple
-and perform sub-coeffecting separately on individual variables.
+The operation could be extended to a vector in a pointwise way, but we keep it simple
+and perform sub-coeffecting on individual variables.
 
@@
    \ident{counit}_{\cunit}~\ctx & (\emph{var})
-\\[0.5em]
+\\[0.4em]
 \sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
   \delta~(c_i) & (\emph{const})
-\\[0.5em]
+\\[0.4em]
 \sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
-  \hspace{2.5em}
+  \hspace{1.5em}
   \begin{array}{l}  
@@
   \end{array}
-\\[1.5em]
+\\[1.4em]
 \sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
-\hspace{2em}\sem{\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+\hspace{1.75em}\sem{\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
   ~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v))
-\\[0.5em]
+\\[0.4em]
 \sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
-  \hspace{2em}
+  \hspace{1.75em}
   \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx~~\kvd{in}~~
   \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}~\ctx_1
-\\[0.5em]
+\\[0.4em]
 \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{sub})\\
-  \hspace{2em}
+  \hspace{1.75em}
   \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
   (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \alift{\cclrd{s'}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
-\\[0.5em]
+\\[0.4em]
 \sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
         \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\[-0.25em]
-  \hspace{2em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
+  \hspace{1.75em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
@@
      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
-\\[0.5em]
+\\[0.4em]
 \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
         \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\[-0.25em]
-  \hspace{2em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
+  \hspace{1.75em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
       {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
@@
 \end{equation*}
-\\[0.5em]
-{\small Assuming the following auxiliary definitions:}
+\\[0.0em]
+{\small The definition uses the following auxiliary definitions:}
 \begin{equation*}
@@
 \ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx = \\
-  \hspace{2em} \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}}~\ctx \\[-0.25em]
-  \hspace{2em} \kvd{in}~\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)
- \\[0.5em]
+  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}}~\ctx \\[-0.25em]
+  \hspace{1.75em} \kvd{in}~\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)
+ \\[0.4em]
 \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} ~:~
@@
 \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f~\ctx =\\
-  \hspace{2em} \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx \\[-0.25em]
-  \hspace{2em} \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}~\ctx' \\[-0.25em] 
-  \hspace{2em} \kvd{in}~\ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~
+  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx \\[-0.25em]
+  \hspace{1.75em} \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}~\ctx' \\[-0.25em] 
+  \hspace{1.75em} \kvd{in}~\ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~
         (\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
@@
 \end{equation*}
-\caption{Categorical semantics of the structural coeffect calculus}
+\figcaption{Categorical semantics of the structural coeffect calculus}
 \label{fig:struct-semantics}
+\vspace{-0.5em}
 \end{figure*}
-
 % --------------------------------------------------------------------------------------------------
@@
 that we need a more expressive representation of the variable context -- \eg~a vector
-of optional values -- and that the structural system needs to pass separate variable contexts
-to the sub-expressions.
+of optional values (\ie~wrapped in the \ident{Maybe} data type). Additionally,
+the structural system needs to pass separate variable contexts to the sub-expressions.
 
@@
 a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
-at the unified categorical semantics defined in terms of \emph{structural indexed comonads}.
+at a unified categorical semantics defined in terms of \emph{structural indexed comonads}.
 
@@
 are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
-rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need for examining the structure
+rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need to examine the structure
 of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
@@
 \paragraph{Expressions.}
-The semantics of strucutural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
+A semantics of strucutural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
 the previous chapter, the semantics is written in a programming language style using constructs
 such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
-can be written using standard primitives (currying, uncurrying, function pairing etc.).
+can equally be written using primitives such as currying, uncurrying, function pairing etc.
 
@@
 
-  The argument for the function is obtained by applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ 
-  to the semantics of $e_2$. The resulting function 
+  A value $\ctyp{\alift{\cclrd{t}}}{\tau_1}$ required in order to call the function is obtained by 
+  applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ to the semantics of $e_2$. The result 
   $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
-  is then called with the latter part of the context to obtain argument for the first function.
-
-\item The semantic of function abstraction (\emph{abs}) is syntactically the same as in the 
+  is then called with the latter part of the split input context.
+
+\item The semantics of function abstraction (\emph{abs}) is syntactically the same as in the 
   flat version -- the only difference is that we now merge a free-variable context with a 
@@
 
-The semantics for the non-syntax-driven rules performs transformations on the 
-free-variable context. Weakening (\emph{weak}) splits the context and ignores the part corresponding
-to the removed variable. If we were modelling the semantics in a language with a linear 
-type system, this would require an additional operation for ignoring a context annotated with 
-$\aclrd{\czero}$.
+\noindent
+The semantics for the non-syntax-driven rules (weakening, exchange, contraction) performs 
+transformations on the free-variable context. Weakening (\emph{weak}) splits the context and 
+ignores the part corresponding to the removed variable. If we were modelling the semantics in 
+a language with a linear type system, this would require an additional operation for ignoring an
+unused context annotated with $\aclrd{\czero}$.
 
 The remaining rules perform a transformation anywhere inside the free-variable vector. 
-To simplify writing the semantics, we define a helper operation 
+To simplify writing the semantics, we define a helper 
 $\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
@@
 As in the flat calculus, the main reason for defining the categorical semantics in this chapter
-is to provide validation for the design of the calculus. As we show in the next section, the 
-discussed examples (liveness, data-flow, bounded variable reuse) form \emph{structural indexed
-comonads} and so the calculus captures them correctly if the coeffect annotations in the typing
-rules match the indices in the semantics. More formally:
-
-\begin{remark}[Correspondence]
+is to provide validation for the design of the calculus. The following correspondence theorem
+states that the annotations in the typing rules of the structural coeffect calculus correspond 
+to the indices of the semantics. Thus, the calculus captures a context-dependent property if
+it can be modelled by a \emph{structural indexed comonad}. As we show in the next section, 
+this is the case for all three discussed examples (liveness, data-flow, bounded variable reuse).
+
+\begin{theorem}[Correspondence]
 In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$ 
@@
 by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
-\end{remark}
+\end{theorem}
 \begin{proof}
@@
 \noindent
-As in the flat calculus, the primitive operations of the structural indexed comonad are all 
-annotated with different operations provided by the coeffect annotations. This means that the
-semantics uniquely determines the structure of the typing rules of the strucutral coeffect
-calculus. Thanks to the correspondence between the product structure $\atimes$ of the annotations
-and the variable context $\xtimes$, the correspondence property also guarantees that variable
-values are split correctly, as required by the structural nature of the type system.
+As in the flat calculus, the primitive operations of the structural indexed comonad are each
+annotated with a unique operations provided by the coeffect annotations. This means that the
+semantics fully determines the structure of the typing rules of the strucutral coeffect
+calculus.
 
@@
 
-The categorical semantics for structural coeffect calculus can be easily instantiated to give
-a semantics of a concrete calculus. In this section, we look at the three examples discussed
-throughout this chapter -- structural liveness and data-flow and bounded variable reuse. 
-Some aspects of the earlier two examples will be similar to flat versions discussed in
+The categorical semantics for structural coeffect calculus is easily instantiated to give
+semantics for a concrete calculus. In this section, we revisit the three examples discussed
+throughout this chapter -- structural liveness, data-flow and bounded variable reuse. 
+Some aspects of the first two examples will be similar to flat versions discussed in
 Section~\ref{sec:flat-semantics} -- they are based on the same data structures (option and a 
-list, respectively), but the data structures are composed differently. Generally speaking -- 
+list, respectively), but the data structures are composed differently. Generally speaking,
 rather than having a data structure over a product of variables, we now have 
@@
 products (\eg~a product of contexts returned by \ident{split}), we write vectors using 
-$\langle a, \ldots, b \rangle$ rather than using parentheses.
+$\langle a, \ldots, b \rangle$ rather than the parentheses, used for ordinary tuples.
 
 \paragraph{Data-flow.}
-It is interesting to note that the semantics of data-flow and bounded variable both keep
-a product of multiple values for each variable, so they are both built around an \emph{indexed list}
-data structure. However, their \ident{cobind} and \ident{dup} operations work differently. We start
-by looking at the structure modelling data-flow computations (variables written in bold face such
-as $\mathbf{a}_1$ range over vectors while $a_1$ ranges over individual values).
+It is interesting to note that the semantics of data-flow and bounded variable reuse (discussed
+next) both keep a product of multiple values for each variable, so they are both built around an 
+\emph{indexed list} data structure. However, their \ident{cobind} and \ident{dup} operations work 
+differently. We start by looking at the structure modelling data-flow computations. For readability,
+variables in bold face (such as $\mathbf{a_1}$) range over vectors while ordinary notation (such
+as $a_1$) is used for individual values.
+
+~
+
+~
 
 \begin{example}[Indexed list for data-flow]
+\label{ex:struct-dfl}
 The indexed list model of data-flow computations is defined over a structural coeffect
@@
 
+\noindent
 The definition of the indexed list data structure relies on the fact that the number of annotations
 corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
-containing $n_i+1$ values for $i$-\textit{th} variable (the annotation represents the number of 
+containing $n_i+1$ values for the $i$-\textit{th} variable (the annotation represents the number of 
 required \emph{past} values so one more value is required).
@@
 representation of the variable). This means that we can re-use the same definitions for the following
-two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will represent options
+two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will there represent options
 rather than lists).
@@
 
-The \ident{dup} operation needs to produce a two-varaible context containing $m$ and $n$ values,
+The $\ident{dup}_{m,n}$ operation needs to produce a two-varaible context containing $m$ and $n$ values,
 respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
-is simply a matter of restriction. Finally, \ident{counit} extracts the (only) value of the (only) variable
+is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable
 and \ident{lift} drops additional past values that are not required.
@@
 As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list 
-structure. Rather than representing possibly different past values that can be shared (\emph{c.f.}~\ident{dup}), 
-the list now represents multiple copies of the same value that cannot be shared.
+structure. Rather than representing possibly different past values that can be shared (see the 
+definition of \ident{dup}), the list now represents multiple copies of the same value as each value
+can only be accessed once. This semantics follows that of Girard \cite{logic-bounded}.
 
@@
 %
-The \ident{merge} and \ident{split} operations are defined as in \emph{indexed list for data-flow}.
-The operations that capture the behaviour of bounded reuse are defined as:
+The \ident{merge} and \ident{split} operations are defined as in Example~\ref{ex:struct-dfl}.
+The operations that capture the behaviour of bounded reuse are:
 %
@@
 
-In the bounded variable reuse system, variable sharing is annotated with $+$ (in contrast with
-\textit{max} used in data-flow). The \indent{dup} operation thus splits the $m+n$ available
+In the bounded variable reuse system, variable sharing is annotated with the $+$ operator (in contrast 
+with \textit{max} used in data-flow). The $\ident{dup}_{m,n}$ operation thus splits the $m+n$ available
 values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The 
@@
 In both data-flow and bounded reuse, the data type is defined as a vector of values obtained
-by applying some parameterized data type (indexed list) to types of individual variables. We can
-generalize this pattern and define $\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of 
-$D^{l}$ where $D^{l}$ is a simpler indexed data type. For liveness, the definition lets us 
-reuse the mapping used when defining the semantics of flat liveness. However, we cannot fully
-define the semantics of the structural version in terms of the flat version -- the 
-\ident{cobind} operation is different and we need to provide the \ident{dup} operation.
+by applying the indexed list type constructor to types of individual variables. We can
+generalize this pattern. Given a parameterized (indexed) type constructor $D^l \alpha$, we define 
+$\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of a vector of $D^{l_i}$ types.
+For liveness, the definition lets us reuse some of the mappings used when defining the semantics of
+flat liveness. However, we cannot fully define the semantics of the structural version in terms of 
+the flat version -- the \ident{cobind} operation is different and we need an appropriate 
+\ident{dup} operation.
 
@@
 duplication in the flat calculus is performed by first duplicating the variable context (using 
-\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair which may or
-may not contain value, depending on the annotations.
+\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair. Depending on the
+required coeffect annotations, this may copy (duplicate) the value, or it may produce an empty context.
 
@@
 $\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models 
-sub-coeffecting which may drop an available value (second case) or behaves as identity.
+sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
+it behaves as identity.
 
@@
 
-\section{Equational theory}
+\section{Syntactic equational theory}
 \label{sec:struct-syntax}
@@
 Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
-different notion of context and thus a different operational interpretation. As before, the 
-properties of the flat coeffect algebra guarantee that certain equational properties hold for 
-all instances of the calculus. In this section, we look at these common properties that we get
-``for free'' in all structural coeffect calculi.
-
-We start the discussion by briefly considering the key aspects
+different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
+certain equational properties hold in all instances of the structural coeffect calculus.
+
+In this section, we look at these common properties that we get ``for free'' from the structural 
+coeffect calculus. We start the discussion by briefly considering the key aspects
 that make the equational theory of flat and structural coeffects different.
@@
 
-When discussing equational theory for the flat calculus in Section~\ref{sec:flat-syntax}, we 
-noted that no single technique of specifying syntactic reduction works universally for all 
-flat coeffect calculi. We considered multiple different options (call-by-name, call-by-value, 
+When discussing syntactic reductions for the flat calculus in Section~\ref{sec:flat-syntax}, 
+we noted that no single technique of specifying syntactic reduction works universally for all 
+flat coeffect calculi. We considered multiple alternatives (call-by-name, call-by-value, 
 internalized substitution) that can be used as the basis for operational semantics for
-different calculi that satisfy different additional properties.
-
-The structural coeffect calculus has more desirable equational properties. In particular, 
-we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of strucutral 
+different calculi that satisfy certain additional properties.
+
+The structural coeffect calculus has more preferable equational properties. In particular, 
+we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of structural 
 coeffect algebra. For this reason, we focus on these two reductions in this section.
@@
 the structural coeffect calculus satisfies both the \emph{local soundness} and 
-the \emph{local completeness} properties.
+the \emph{local completeness} properties, witnessed by $\beta$-reduction and $\eta$-expansion,
+respectively.
 
@@
 (Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
-to coincide. This enables the substitution to preserve the type of expressions, because all
+to coincide. This enables substitution to preserve the type of expressions, because all
 additional requirements arising as the result of the substitution can be associated with the
@@
  \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{before}) \\
+   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textit{before}) \\
 \coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
-  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{after})
-\end{array}  
-\end{equation*}
-%
-The typing judgement obtained in (\emph{after}) preserves the type of the expression
-(function value) from the original typing (\emph{before}). This is possible thanks to the
-non-determinism involved in lambda abstraction -- as all operators of the flat coeffect
+  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textit{after})
+\end{array}  
+\end{equation*}
+%
+The typing judgement obtained in (\emph{after}) preserves the type of the expression (function 
+value) from the original typing (\emph{before}). This is possible thanks to the non-determinism 
+involved in the typing rule for lambda abstraction -- as all operators of the flat coeffect 
 algebra used here are $\cup$, we can place the additional requirement on the outer context. 
@@
 \paragraph{Substitution for structural coeffects.}
-In contrast, the substitution lemma for structural coeffects can be proven because 
-structural coeffect systems provide enough \emph{precision} to identify exactly with 
-which variable should a context requirement be associated.
-
-The following example shows a situation similar to the previous one -- here, we use
+In contrast, the substitution lemma (Lemma~\ref{thm:structural-substitution}, shown later on 
+page~\pageref{thm:structural-substitution}) for structural coeffects can be proven because 
+structural coeffect systems provide enough \emph{precision} to identify exactly with which 
+variable should a context requirement be associated.
+
+The following example shows a situation similar to the previous one. Here, we use
 structural data-flow calculus (writing $\kvd{prev}~e$ to obtain previous value of the 
@@
   &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{before}) \\
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{before}) \\
 \coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
   &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{after}) \\
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{after}) \\
 \coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
   &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{final})
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{final})
 \end{array}  
@@
 such as the bunched typing system \cite{substruct-bunched}. To express the substitution 
-lemma, we define the notion of a \emph{context with holes}:
+lemma later in this section, we first define the notion of a \emph{context with holes}:
 
@@
 
-Substitution can be seen as sequential composition. Informally -- we first need to obtain the 
+Substitution can be seen as a form of sequencing, \ie~the term $\subst{e_1}{x}{e_2}$ should have
+the same type as the term $\kvd{let}~x=e_2~\kvd{in}~e_1$. In other words, we first need to obtain the 
 value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
@@
 are annotated with $\cclrd{s} \,\aseq\, \aclrd{\textbf{t}}$, using the (scalar-vector extension)
-of the sequential composition operator $\cseq$. This suggests that, to evaluate each of the variables
+of the sequential composition operator $\cseq$.
+
+An operational reading of the resulting coeffect annotation is that, to evaluate each of the variables
 annotated with coeffects in the vector $\aclrd{\textbf{t}}$, we first need to evaluate the substituted
@@
 By induction over $\vdash$, using the multi-nary aspect of the substitution 
-in the proof of the contraction case.
+in the proof of the contraction case (see Appendix~\ref{sec:appendix-struct-cbn}).
 \end{proof}
@@
 
-In the previous chapter, we discussed call-by-value separately from call-by-name, because the
+In the Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
 proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
-$\beta$-reduction (local soundness), which encompasses both types of evaluation and 
-$\eta$-expansion (local completeness). Both of the properties hold for a system with any
+$\beta$-reduction, which encompasses both call-by-value and call-by-name. We also show that 
+$\eta$-expansion preserves the types. Both of the properties hold for a system with any
 structural coeffect algebra.
 
-\paragraph{Reduction theorem}
-In a full $\beta$-reduction,
-written as $\rightarrow_\beta$, we replace the redex $(\lambda x.e_2)~e_1$ by the
+\paragraph{Reduction theorem.} In a full $\beta$-reduction,
+written as $\rightarrow_\beta$, we can replace the redex $(\lambda x.e_2)~e_1$ by the
 expression $\subst{e_r}{x}{e_s}$ anywhere inside a term. The subject reduction theorem
@@
 
-\begin{theorem}[$\beta$-reduction]
+\begin{theorem}[Type preservation]
 In a structural coeffect system with a structural coeffect algebra formed by 
@@
 \noindent
-Because of the vector (free monoid) structure, coeffect annotations $\aclrd{\textbf{r}}$, 
-$\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$ are uniquely associated with $\Gamma_r$, 
+Because of the vector (free monoid) structure of coeffect annotations $\aclrd{\textbf{r}}$, 
+$\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$, these are uniquely associated with $\Gamma_r$, 
 $\Gamma_s$, and $x$ respectively. Therefore, substituting $e_s$ (which has coeffects 
-$\aclrd{\textbf{s}}$) for $x$ introduces the context dependencies specified by $\aclrd{\textbf{s}}$
-which are composed with the existing requirements $\cclrd{t}$ on $x$.
-
-\paragraph{Expansion theorem}
+$\aclrd{\textbf{s}}$) for $x$ introduces the context requirements specified by $\aclrd{\textbf{s}}$
+which are composed with the requirements $\cclrd{t}$ associated with $x$, \ie~the variable being 
+substituted.
+
+\paragraph{Expansion theorem.}
 Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
-\emph{local soundness} and \emph{local completeness}. Informally, this means that 
-abstraction does not introduce too much, and application does not eliminate too much.
+\emph{local soundness} and \emph{local completeness} as required by Pfenning and Davies
+\cite{logic-modal-reconstruction}. Informally, this means that abstraction does not introduce 
+too much, and application does not eliminate too much.
 
@@
 \begin{proof}
-The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
+The following derivation shows that $\lambda x.f~x$ has the same type and coeffects as 
+the original expression $f$:
 \begin{equation*}
@@
     { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
-      \coctx{x:\tau_1}{\alift{\cunit}} \vdash x:\tau_1 }
-    { \coctx{\Gamma, x:\tau_1}{\aclrd{\textbf{r}} \;\atimes\;(\cclrd{s}\,\aseq\,\alift{\cunit}) } \vdash f~x : \tau_2 } }
+      \coctx{x\!:\!\tau_1}{\alift{\cunit}} \vdash x:\tau_1 }
+    { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}} \;\atimes\;(\cclrd{s}\,\aseq\,\alift{\cunit}) } \vdash f~x : \tau_2 } }
   { \inference
-    { \coctx{\Gamma, x:\tau_1}{\aclrd{\textbf{r}} \;\atimes\,\alift{\cclrd{s}} } \vdash f~x : \tau_2 }
+    { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}} \;\atimes\,\alift{\cclrd{s}} } \vdash f~x : \tau_2 }
     { \coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
@@
 of effect. For example, in a language with output effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ 
-has different effects to its $\eta$-converted form $\lambda x . e x$ because the immediate 
+has different effects to its $\eta$-converted form $\lambda x . e\,x$ because the immediate 
 effects of $e$ are hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, 
@@
 
-This chapter completes the key development of this thesis -- the presentation of unified calculi
-for context-aware computations that capture the motivating examples introduced in 
-Chapter~\ref{ch:applications}. In the previous chapter, we focused on the first category of 
-contet-aware computations and we developed \emph{flat coeffect calculus} that captures
-\emph{whole-context} properties. This chapter develops \emph{structural coeffect calculus},
-capturing \emph{per-variable} contextual properties. The system provides a precise 
-analysis of liveness and data-flow and allows other interesting uses such as tracking of variable
-accesses based on bounded linear logic.
+This chapter completes the key development of this thesis -- the presentation of the coeffect 
+framework, consisting of two calculi capturing properties of context-aware computations introduced
+in Chapter~\ref{ch:applications}. In the previous chapter, we focused on the whole-context 
+properties of computations and we developed \emph{flat coeffect calculus} to capture them.
+This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
+contextual properties. The system provides a precise analysis of liveness and data-flow and allows 
+other interesting uses such as tracking of variable accesses based on bounded linear logic.
 
@@
 $\cseq$ and $\cpar$ that model sequential and pointwise composition, respectively. For flat 
-coeffects, we further require $\czip$ to model context merging. For structural coeffects, we 
+coeffects, we required $\czip$ to model context merging. For structural coeffects, we 
 instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
@@
 calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
-it unifies our three example calculi (bounded reuse, data-flow and liveness) and it demonstrates
-suitability of the type system. The second technical result is equational theory for structural
-coeffect calculi. In particular, we show that $\beta$-reduction and $\eta$-expansion preserve
-the typing for any structural coeffect calculi. These two strong properties are desirable for
-programming languages, but are often not satisfied (\eg~by effectful langauges).
-
-Two questions remain opened for the upcoming chapter. Firstly, is there a way to present 
-flat and structural coeffect calculi in a uniform way as a single system? Secondly, can we
-simplify the definitions to make type inference easier and add other practical language
-aspects such as recursion?
+it unifies our three examples (bounded reuse, data-flow and liveness) and it demonstrates
+suitability of the type system. The second technical result is equational theory for the structural
+coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
+typing preservation property for any instance of the structural coeffect calculus. These two 
+strong properties are desirable for programming languages, but are often not satisfied 
+(\eg~by langauges with effects).
+
+Two questions remain open for the upcoming chapter. Firstly, is there a way to present 
+flat and structural coeffect calculi in a uniform way as a single system? Secondly, what
+is the best pathway towards a practical implementation of programming languages based on 
+coeffect systems?

### 2014-12-14 text/AppendixB.tex
Finish tweaks in flat coeffects (Chapter 3)
Proofreading and English conventions (Chapter 4)
Proofreading and editing (Chapter 5)
Proving things (Chapter 5)

@@
 
-\chapter{Appendix A} 
+\chapter{Appendix B} 
 \label{ch:appendix}
@@
 from Chapter~\ref{ch:structural}.
+
+% =================================================================================================
+%
+%      #####   ##            #    
+%      #        #            #    
+%      #        #     ###   ####  
+%      ####     #        #   #    
+%      #        #     ####   #    
+%      #        #    #   #   #  # 
+%      #       ###    ####    ##  
+%                               
+% =================================================================================================
 
@@
 \begin{itemize}
-\item[--] If $y=x$ then also $\tau = \tau_s$ and thus $\subst{y}{x}{e_s} = e_s$. Using the assumption,
+\item If $y=x$ then also $\tau = \tau_s$ and thus $\subst{y}{x}{e_s} = e_s$. Using the assumption,
 implicit weakening and the fact that $\cunit$ is a unit of $\cseq$:
@@
 \]
-\item[--] If $y\neq x$ then $\subst{y}{x}{e_s} = y$. Using the fact that $\cunit$ is the bottom element
+\item If $y\neq x$ then $\subst{y}{x}{e_s} = y$. Using the fact that $\cunit$ is the bottom element
 and sub-coeffecting:
@@
 \]
-Now using (\emph{app}) rule and the fact that $\cpar\,=\,\cseq$, associativity, commutativity and idempotence
+Now using the (\emph{app}) rule and the fact that $\cpar\,=\,\cseq$, associativity, commutativity and idempotence
 (note that all three properties are needed):
@@
 \]
-Now using (\emph{let}) rule and similarly to the (\emph{app}) case:
+Now using the (\emph{let}) rule and similarly to the (\emph{app}) case:
 \[
+\hspace{-2em}
 \inference
@@
 \end{proof}
+\newpage
 
+% =================================================================================================
 %                                                                      
-% (e2 + (e1 * e2)) + (e3 * f)
-% ----(glet)------
-% ----------(app)------------
+%     ###    #                           #                           ##   
+%    #   #   #                           #                            #   
+%    #      ####   # ##   #   #   ###   ####   #   #  # ##    ###     #   
+%     ###    #     ##  #  #   #  #   #   #     #   #  ##  #      #    #   
+%        #   #     #      #   #  #       #     #   #  #       ####    #   
+%    #   #   #  #  #      #  ##  #   #   #  #  #  ##  #      #   #    #   
+%     ###     ##   #       ## #   ###     ##    ## #  #       ####   ###  
 %
-% = e2 + e1*e2 + e3*f
+% =================================================================================================
+
+\section{Substitution for structural coeffects}
+\label{sec:appendix-struct-cbn}
+
+In order to prove that $\beta$-reduction and $\eta$-expansion preserve the type of an expression
+in Section~\ref{sec:struct-syntactic-subst}, we required a multi-nary form of the substitution 
+lemma (Lemma~\ref{thm:structural-substitution}). This section provides the corresponding proof.
+
+\begin{lemma*}[Multi-nary substitution]
+Given an expression with multiple holes filled by variables $x_{Ti}\!:\!\tau_{Ti}$ with coeffects $\cclrd{s_k}$:
 %
-% ~~~>
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+  \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
+\end{equation*}
 %
-% (e2 + (e3 * f)) + ((e2 + (e3 * f)) * e1)
-% ----(app)------    ----(app)-----
-% ---------------(glet)-------------------
+and a expressions $e_{Ti}$ with free-variable contexts $\Gamma_{Ti}$ annotated with $\aclrd{\mathbf{T_i}}$:
 %
-% = e2 + e3*f + e2*f + e1*e3*f
+\begin{equation*}
+\coctx{\Gamma_1}{\aclrd{\mathbf{T_1}}} \vdash e_{T1} : \tau_{T1}
+\quad \ldots \quad
+\coctx{\Gamma_k}{\aclrd{\mathbf{T_k}}} \vdash e_{Tk} : \tau_{Tk}
+\end{equation*}
+%
+then substituting the expressions $e_{Ti}$ for variables $x_{Ti}$ results in an expression with a context
+where the original holes are filled by contexts $\Gamma_{Ti}$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{T_i}}$:
+%
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+  \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,] \vdash \subst{\subst{e_r}{x_{T1}}{e_{T1}}\ldots}{x_{Tk}}{e_{Tk}} : \tau_r
+\end{equation*}
+\end{lemma*}
+\begin{proof}
+Assume that $\coctx{\Gamma}{\aclrd{\mathbf{T_i}}} \vdash e_{Ti} : \tau_{Ti}$ and we are substituting terms $e_{Ti}$ for 
+variables $x_{Ti}$. Furthermore, we assume that are variables that are being substituted for actually appear in the original 
+term (which means that $k$ is at most the number of variables). Note that we use upper-case $\aclrd{\textbf{T}}$ to distinguish 
+the coeffect of the expression that is being substituted into an expression. Using structural induction over $\vdash$:
 
-% e3 + ((e2 + (e1 * e2)) * e3)
-%       ------(glet)---
-% ---------(glet)-------------
-%
-% = e3 + e2*e3 + e1*e2*e3
-%  
-% ~~~>
-%
-% (e2 + (e3 * e2)) + (e1 * (e2 + (e3 * e2)))
-% -----(glet)-----         -----(glet)-----
-% ------------------(glet)------------------
-%
-% = e2 + e3*e2 + e1*e2 + e1*e2*e3
+% -------------------------------------------------------------------------------------------------
 
-% Data-flow: e3 + (e2 + e1) = (e2 + e3) + e1
-% Liveness:  e3 = e3
-% Resources: OK
+\paragraph{Syntax-driven typing rules}
+\paragraph{(var)} Given the following derivation using (\emph{var}):
+\[
+\inference
+  { }
+  {\coctx{y:\tau}{\alift{\cunit}} \vdash y : \tau }
+\]
+Here, the context contains exactly one variable and so $k=1$. There are two cases depending on whether $y$ is 
+the (only substituted) variables $x_{T1}$ or not:
+\begin{itemize}
+\item If $y=x_{T1}$ then also $\tau = \tau_{T1}$ and thus $\subst{y}{x_{T1}}{e_{T_1}} = e_{T1}$. 
+In this case, the context contains only a single hole $\coctx{\Gamma}{\aclrd{\mathbf{r}}} = \coctx{-}{-}$. Using the
+assumption and the fact that $\cunit$ is a unit of $\cseq$:
+\[
+\inference
+ { \inference 
+   {\coctx{\Gamma_{T1}}{\aclrd{\mathbf{T_i}}} \vdash \subst{y}{x_{T1}}{e_{T1}} : \tau_r }
+   {\coctx{\Gamma}{\aclrd{\mathbf{r}}}~[\,\coctx{\Gamma_{T1}}{\aclrd{\mathbf{T_i}}}\,] \vdash \subst{y}{x_{T1}}{e_{T1}} : \tau_r } }
+ { \coctx{\Gamma}{\aclrd{\mathbf{r}}}~[\,\coctx{\Gamma_{T1}}{\cunit\,\aseq\,\aclrd{\mathbf{T_i}}}\,] \vdash \subst{y}{x_{T1}}{e_{T1}} : \tau_r }   
+\]
+\item If $y\neq x_{T1}$ then there is no substitution that could be performed ($y$ does not 
+appear in the context) and so (trivially):
+\[
+\coctx{\Gamma}{\aclrd{\mathbf{r}}}~[\,] \vdash y : \tau_r
+\]
+\end{itemize}
 
-%---------------------------------------------------------------------------------------------------
+\paragraph{(const)} Similar to the (\emph{var}) case when the variable is not substituted.
 
-% \section{Internalized substitution}
+\paragraph{(app)} In the (\emph{app}) rule, the context $\Gamma$ is obtained as a tensor product $\Gamma_1, \Gamma_2$.
+  Given $\Gamma$ of length $k$, we assume that $\Gamma_1$ has length $l$ and $\Gamma_2$ has length $k-l$. 
+  Now, given the following typing derivation using (\emph{app}):
+\[
+\inference
+  {\coctx{\Gamma_1}{\aclrd{\mathbf{s_1}}} 
+   [\,\coctx{x_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tl}}{\alift{\cclrd{s_l}}}\,] 
+      \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \\
+   \coctx{\Gamma_2}{\aclrd{\mathbf{s_2}}} 
+   [\,\coctx{x_{Tl+1}}{\alift{\cclrd{s_{l+1}}}} \,|\, \ldots \,|\, \coctx{x_{Tk}}{\alift{\cclrd{s_k}}}\,]
+      \vdash e_2 : \tau_1 }
+  { \begin{array}{l}
+    \coctx{\Gamma_1, \Gamma_2}{\aclrd{\mathbf{s_1}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\mathbf{s_2}})}}~
+     [\,\coctx{x_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tl}}{\alift{\cclrd{s_l}}}, \\[-0.35em] \hspace{9.3em}
+       \coctx{x_{Tl+1}}{\alift{\cclrd{t} \,\cseq\, \cclrd{s_{l+1}}}} \,|\, \ldots \,|\, 
+          \coctx{x_{Tk}}{\alift{\cclrd{t} \,\cseq\, \cclrd{s_k}}}\,] \vdash e_1~e_2 : \tau_2 
+    \end{array} }
+\]
+Here, we use $\cclrd{t}\;\aseq\;\aclrd{\mathbf{s_2}}$ as a pointwise extension of $\czip$ that is additionally
+applied to holes such that $\cclrd{t}\;\czip\; - = -$. That is, a hole in the context remains a hole and so
+the second part of the context is filled with $\cclrd{t} \,\cseq\, \cclrd{s_i}$ for all $i \in \{ l+1 \ldots k \}$.
+Next, from the induction hypothesis, we have that:
+\[
+\begin{array}{l}
+ \coctx{\Gamma_1}{\aclrd{\mathbf{s_1}}} 
+   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\,\aseq\,\aclrd{\mathbf{T_1}}} \,|\, \ldots \,|\, \coctx{\Gamma_{Tl}}{\cclrd{s_l}\aseq\,\aclrd{\mathbf{T_l}}}\,] 
+   \vdash e_1~[\,\ldots\,] : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \\
+ \coctx{\Gamma_2}{\aclrd{\mathbf{s_{2}}}} 
+   [\,\coctx{\Gamma_{Tl+1}}{\cclrd{s_{l+1}}\aseq\,\aclrd{\mathbf{T}_{l+1}}} \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\mathbf{T_k}}}\,]
+   \vdash e_2~[\,\ldots\,] : \tau_1
+\end{array}\quad(*)
+\]
+Now using the (\emph{app}) rule in the first step and associativity of $\cseq$ on the second part of the context in the second step:
+\[
+\hspace{-2em}
+\inference 
+ { \inference
+   { (*) }
+   { \begin{array}{l}
+      \coctx{\Gamma_1, \Gamma_2}{\aclrd{\mathbf{s_1}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\mathbf{s_2}})}} ~~
+       [\,\coctx{\Gamma_{T1}}{ \cclrd{s_1}\,\aseq\,\aclrd{\mathbf{T_1}} } \,|\, \ldots \,|\, 
+          \coctx{\Gamma_{Tl}}{ \cclrd{s_l}\,\aseq\,\aclrd{\mathbf{T_l}} }, 
+         \\[-0.35em] \qquad
+         \coctx{\Gamma_{Tl+1}}{\cclrd{t} \,\aseq\, (\cclrd{s_{l+1}}\,\aseq\,\aclrd{\mathbf{T_{l+1}}}) } \,|\, \ldots \,|\,
+         \coctx{\Gamma_{Tk}}{\cclrd{t} \,\aseq\, (\cclrd{s_k}\,\aseq\,\aclrd{\mathbf{T_k}})}\,] \vdash (e_1~e_2)~[\,\ldots\,] : \tau_2 
+      \end{array} } }
+ { \begin{array}{l}
+    \coctx{\Gamma_1, \Gamma_2}{\aclrd{\mathbf{s_1}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\mathbf{s_2}})}} ~~
+     [\,\coctx{\Gamma_{T1}}{ \cclrd{s_1}\,\aseq\,\aclrd{\mathbf{T_1}} } \,|\, \ldots \,|\, 
+        \coctx{\Gamma_{Tl}}{ \cclrd{s_l}\,\aseq\,\aclrd{\mathbf{T_l}} }, 
+       \\[-0.35em] \qquad
+       \coctx{\Gamma_{Tl+1}}{(\cclrd{t} \,\cseq\, \cclrd{s_{l+1}})\,\aseq\,\aclrd{\mathbf{T_{l+1}}} } \,|\, \ldots \,|\,
+       \coctx{\Gamma_{Tk}}{(\cclrd{t} \,\cseq\, \cclrd{s_k})\,\aseq\,\aclrd{\mathbf{T_k}}}\,] \vdash (e_1~e_2)~[\,\ldots\,] : \tau_2 
+    \end{array} }
+\]
 
-% \subsection{First transformation}
+\paragraph{(abs)} Without the loss of generality, we can assume that the bound variable is 
+not one of the variables being substituted. Thus, the last variable (and the corresponding coeffect)
+are not holes. The typing derivation using (\emph{abs}) then looks as follows:
+\[
+\inference
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} 
+   ~[\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+    \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} 
+   ~[\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+    \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\]
+From the induction hypothesis, we have that:
+\[
+\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}}~
+ [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+    \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,] \vdash e~[\,\ldots\,] : \tau_2
+\]
+Because the last position in the vector of variables is an actual variable rather than a hole, we
+just need to apply the (\emph{abs}) rule:
+\[
+\inference 
+ { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}}~
+   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+      \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,] \vdash e~[\,\ldots\,] : \tau_2 }
+ { \coctx{\Gamma}{\aclrd{\textbf{r}}}~
+   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+      \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,] \vdash \lambda x.e~[\,\ldots\,] : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\]
 
-% \begin{equation*}
-% (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 \leadsto \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
-% \end{equation*}
+\paragraph{(let)} In the structural coeffect calculus let-binding can be viewed as a syntactic
+sugar for abstraction/application and so the case follows from (\emph{abs}) and (\emph{app}).
 
-% \begin{equation*}
-% \tyrule{app}
-%   { \tyrule{glet}
-%       { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
-%         \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 }
-%       { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 } &
-%     \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3  }
-%   { \coctx{\Gamma}{(\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) } \vdash (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3 : \tau_2 }
-% \end{equation*}
+\vspace{0.75em}
+\noindent
+Compared to the similar proof for the flat coeffect calculus, the proof for the structural system
+requires fewer properties of the coeffect algebra. In particular, we only needed associativity of 
+$\cseq$ in the (\emph{app}) rule the fact that $\cunit$ is a unit of $\cseq$ in (\emph{var}).
 
-% to
+% -------------------------------------------------------------------------------------------------
 
-% \begin{equation*}
-% \tyrule{glet}
-%   { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
-%     \tyrule{app}
-%       { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{t}} \tau_2 & 
-%         \coctx{\Gamma}{\cclrd{u}} \vdash e_3 : \tau_3 }
-%       { \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})} \vdash e_2~e_3 : \tau_2 } }
-%   { \coctx{\Gamma}{ (\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})) \;\cpar\; (\cclrd{s}\,\cseq\,(\cclrd{r}\;\cpar\;(\cclrd{u}\,\cseq\,\cclrd{t})))  }
-%       \vdash \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3) : \tau_2 }
-% \end{equation*}
+\paragraph{Structural rules}
 
-% meaning
-% \begin{equation*}
-% \begin{array}{l}
-% (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})) \;\cpar\; (\cclrd{u}\,\cseq\,\cclrd{t}) =
-% \end{array}
-% \end{equation*}
+\paragraph{(contr)} In case of contraction, we can assume that the two variables to be contracted
+(in the assumption) are not the variables that are being substituted. However, the resulting
+variable  (in the conclusion) can be one of the variables being substituted for.
 
-% \subsection{Second transformation}
+\begin{itemize}
+\item Assuming that $x \neq x_{Ti}$ for all $i$, the original derivation is:
+\[
+\inference
+  {{ \begin{array}{l}
+     \coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} 
+     \\[-0.35em] \hspace{4em}
+     [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+    \vdash e : \tau
+    \end{array} }}
+  { \begin{array}{l}
+    \coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} 
+    \\[-0.35em] \quad
+    [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+    \vdash \subst{e}{z,y}{x} : \tau
+    \end{array} }
+\]
+Applying (\emph{contr}) to the induction hypothesis gives the required result:
+\[
+\inference
+  {{ \begin{array}{l}
+     \coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} 
+     \\[-0.35em] \hspace{4em}
+     [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
+    \vdash e~[\,\ldots\,] : \tau
+    \end{array} }}
+  { \begin{array}{l}
+    \coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} 
+    \\[-0.35em] \quad
+    [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
+    \vdash \subst{e}{z,y}{x}[\,\ldots\,] : \tau
+    \end{array} }
+\]
 
-% Second transformation
+\item In the other case, $x=x_{Ti}$ for some $i$. The original typing is:
+\[
+\inference
+  {{ \begin{array}{l}
+     \coctx{\Gamma_1,-,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-,-} \atimes \aclrd{\textbf{q}}}~
+         [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+     \\[-0.35em] \hspace{4em}
+     \coctx{y\!:\!\tau_1}{\alift{\cclrd{s}}} \,|\,
+     \coctx{z\!:\!\tau_1}{\alift{\cclrd{t}}}
+     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+    \vdash e : \tau
+    \end{array} }}
+  { \begin{array}{l}
+    \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}}~
+        [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+     \\[-0.35em] \quad
+     \coctx{x\!:\!\tau_1}{ \alift{\cclrd{s} \,\cpar\, \cclrd{t}} }
+     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+    \vdash \subst{e}{z,y}{x} : \tau
+    \end{array} }
+\]
+Applying (\emph{contr}) to the induction hypothesis gives the following result:
+\[
+\inference
+  {{ \begin{array}{l}
+     \coctx{\Gamma_1,-,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-,-} \atimes \aclrd{\textbf{q}}}~
+         [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\,
+     \\[-0.35em] \quad
+     \coctx{\Gamma_{Ti}}{\cclrd{s}\,\aseq\,\aclrd{\mathbf{T_k}} } \,|\,
+     \coctx{\Gamma_{Ti}}{\cclrd{t}\,\aseq\,\aclrd{\mathbf{T_k}} }
+     \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
+    \vdash e~[\,\ldots\,] : \tau
+    \end{array} }}
+  {{ \begin{array}{l}
+     \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}}~
+         [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\,
+     \\[-0.35em] \quad
+     \coctx{\Gamma_{Ti}}{ (\cclrd{s}\,\aseq\,\aclrd{\mathbf{T_k}}) \aparstr (\cclrd{t}\,\aseq\,\aclrd{\mathbf{T_k}})}
+     \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
+    \vdash e~[\,\ldots\,] : \tau
+    \end{array} }}
+\]
+Here the $\aparstr$ operation represents a pointwise extension of $\cpar$. Thus for $i^{\textit{th}}$ 
+substituted coeffect, we have $(\cclrd{s}\,\cseq\,\cclrd{T_i})\;\cpar\;(\cclrd{t}\,\cseq\,\cclrd{T_i})$.
+Using the distributivity law of structural coeffect algebra, we obtain the required structure:
+$(\cclrd{s}\,\cpar\,\cclrd{t})\;\cseq\;\cclrd{T_i}$.
 
-% \begin{equation*}
-% \tyrule{glet}
-%   { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
-%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
-%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
-%   { \coctx{\Gamma}{
-%       \cclrd{t}
-%       \;\cpar\;
-%       (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
-%          \,\cseq\, 
-%          \cclrd{t} )
-%     } 
-%     \vdash \kvd{glet}~x_r=(\kvd{glet}~x_s=e_s~\kvd{in}~e_r)~\kvd{in}~e_t : \tau_t }
-% \end{equation*}
+\paragraph{(sub)} As in the (\emph{contr}) case, in the (\emph{sub}) case we distinguish two situations.
+If the sub-coeffecting is applied to variable that is \emph{not} being substituted for, then the case is
+easy (sub-coeffecting does not interact with substitution), so we only consider the case when $x$ is one
+of the variables being substituted for:
+\[
+\inference
+  {{\begin{array}{l}
+   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
+   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+     \\[-0.35em] \qquad
+     \coctx{x\!:\!\tau_1}{\alift{\cclrd{s'}}}
+     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+   \vdash e : \tau
+   \end{array}}}
+  {{\begin{array}{l}
+   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
+   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+     \\[-0.35em] \qquad
+     \coctx{x\!:\!\tau_1}{\alift{\cclrd{s}}}
+     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+   \vdash e : \tau
+   \end{array}}}
+\quad(\cclrd{s'}\;\cleq\;\cclrd{s})   
+\]
+From the induction hypothesis, we have the following:
+\[
+  {{\begin{array}{l}
+   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
+   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\,
+     \\[-0.35em] \qquad
+     \coctx{\Gamma_{Ti}}{\cclrd{s'}\aseq\,\aclrd{\mathbf{T_k}} }
+     \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
+   \vdash e : \tau
+   \end{array}}}
+\]
+To complete the case, we need to apply (\emph{sub}) repeatedly on each of the variables in 
+$\Gamma_{Ti}$. For $i^{\textit{th}}$ variable $x_i$, the coeffect annotation is $\cclrd{s'}\cseq\,\cclrd{T_i}$.
+Using the fact that combining coeffects with $\cseq$ preserves the ordering, we get that 
+$(\cclrd{s'}\cseq\,\cclrd{T_i}) \cleq (\cclrd{s}\,\cseq\,\cclrd{T_i})$ and so the conditions
+of (\emph{sub}) are satisfied.
 
-% or
+\paragraph{(weak)} We again need to consider whether the removed variable is 
+one of the variables that are being substituted for. If this is not the case,
+the proof is easy, so we only look at the other case:
+\[
+\inference
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} }~
+   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
+   \vdash e : \tau}
+  {\coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
+   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,|\,\coctx{x\!:\!\tau_1}{\czero}\,]
+   \vdash e : \tau} 
+\]
+Now, we use the induction hypothesis, apply the (\emph{weak}) rule and use properties of the
+structural coeffect algebra:
+\[
+\inference
+  {{ \begin{array}{l}
+    \coctx{\Gamma}{ \aclrd{\textbf{r}} }~
+    [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+     \\[-0.35em] \qquad
+     \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}\,]
+   \vdash e~[\,\ldots\,] : \tau 
+   \end{array} }}
+{\tyrule{sub}
+ {{\begin{array}{l}
+   \coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
+   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+    \\[-0.35em] \qquad
+    \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}
+      \,|\,\coctx{x\!:\!\tau_1}{\czero}\,]
+   \vdash e~[\,\ldots\,] : \tau
+   \end{array} }} 
+ {{\begin{array}{l}
+   \coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
+   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
+    \\[-0.35em] \qquad
+    \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}
+      \,|\,\coctx{x\!:\!\tau_1}{\czero \aseq\,\aclrd{\textbf{T}_k} }\,]
+   \vdash e~[\,\ldots\,] : \tau
+   \end{array} }}  }
+\]
+The derivation first applies the standard (\emph{weak}) rule and then uses sub-coeffecting
+rule and the property $\czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})$ to obtain conclusion of the required form.
 
-% \begin{equation*}
-% \tyrule{glet}
-%   { \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s &
-%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_r : \tau_r &
-%     \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e_t : \tau_t }
-%   { \coctx{\Gamma}{
-%       (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
-%       \;\cpar\;
-%       (  \cclrd{s}
-%          \,\cseq\, 
-%          (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) )
-%     } 
-%     \vdash \kvd{glet}~x_s=e_s~\kvd{in}~(\kvd{glet}~x_r=e_r~\kvd{in}~e_t) : \tau_t }
-% \end{equation*}
+\paragraph{(exch)} In the (\emph{exch}) case, the property follows directly from the 
+induction hypothesis. The required conclusion is obtained by applying (\emph{exch}) 
+repeatedly (as we now need to exchange not just two individual variables, but two 
+contexts, possibly containing multiple variables).
 
-% \begin{equation*}
-% \begin{array}{l}
-%   \cclrd{t}
-%   \;\cpar\;
-%   (  (\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r}))
-%      \,\cseq\, 
-%      \cclrd{t} ) = 
-% \\
-%   \cclrd{t}
-%   \;\cpar\;
-%   (\cclrd{r}\,\cseq\, \cclrd{t})
-%   \;\cpar\;
-%   (\cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t}) = 
-% \\
-%   \cclrd{s}\,\cseq\,\cclrd{r}\,\cseq\, \cclrd{t} 
-% \end{array}
-% \end{equation*}
-
-% \begin{equation*}
-% \begin{array}{l}
-%   (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) 
-%   \;\cpar\;
-%   (  \cclrd{s}
-%      \,\cseq\, 
-%      (\cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})) ) = 
-% \\     
-%   \cclrd{t}\;\cpar\;(\cclrd{r}\,\cseq\,\cclrd{t})
-%   \;\cpar\;
-%   (\cclrd{s} \,\cseq\, \cclrd{t})
-%   \;\cpar\;
-%   (\cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t})  = 
-% \\     
-%   \cclrd{s} \,\cseq\, \cclrd{r}\,\cseq\,\cclrd{t}
-% \end{array}
-% \end{equation*}
-
-% require
-
-% \begin{equation*}
-% \cclrd{r}\;\cpar\;(\cclrd{r} \,\cseq\, \cclrd{s}) = \cclrd{r} \,\cseq\, \cclrd{s}
-% \end{equation*}
+\end{itemize}
+\end{proof}

### 2014-12-14 text/Chapter05.tex
Proving things (Chapter 5)

@@
 \textnormal{(pre-order)}   
-\end{array}
-\end{equation*}
-%
-In addition, the following distributivity axioms hold:
+\\
+\end{array}
+\end{equation*}
+%
+The following distributivity axioms hold:
 \begin{align*}
@@
 \end{align*}
+%
+In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
+\begin{equation*}
+\begin{array}{l}
+ \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
+ \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
+\end{array}
+\end{equation*}
 \end{definition}
@@
 idempotent.
+
+Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
+via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma 
+(Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
+for all our examples; it is needed for the structural version of the subtitution lemma 
+(Lemma~\ref{thm:structural-substitution}).
 
@@
 
-More generally, we could see the the coeffect annotations as a \emph{container} 
-\cite{types-containers} that supports certain operations. This approach is used in 
-Chapter~\ref{ch:unified} as a way of unifying the flat and structural systems.
+More generally, we could see the the coeffect annotations as \emph{containers} 
+\cite{types-containers}. This approach is used in Chapter~\ref{ch:unified} to unify flat 
+and structural systems.
 
@@
 
-~
-
-~
-
 \begin{example}[Indexed list for data-flow]
@@
 and \ident{lift} drops additional past values that are not required.
+
+~
 
@@
 \subsection{Holes and substitution lemma}
+\label{sec:struct-syntactic-subst}
 
@@
 
+~
+
+~
+
+~
+
 \begin{lemma}[Multi-nary substitution]
@@
 \begin{equation*}
-\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
-  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_1}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
+  \coctx{\Gamma_k}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
 \end{equation*}

### 2014-12-14 text/Chapter03.tex
Finishing proofreading (Chapter 6)

@@
 \begin{itemize}
-\item[--] We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
+\item We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
   explains what is a \emph{coeffect} and constrasts it with a better known notion of 
@@
 
-\item[--] We describe a number of simple calculi for tracking a wide range of contextual properties. 
+\item We describe a number of simple calculi for tracking a wide range of contextual properties. 
   The systems are adapted from diverse sources (type systems, static analyses, logics) and apply to
@@
 
-\item[--] The uniform presentation of the systems is the key contribution of this chapter. We distinguish
+\item The uniform presentation of the systems is the key contribution of this chapter. We distinguish
   between \emph{flat coeffect} systems (Section~\ref{sec:applications-flat}) and \emph{structural coeffect}
@@
 
-\item[--] In addition, the coeffect systems for tracking the number of accessed past values in 
+\item In addition, the coeffect systems for tracking the number of accessed past values in 
   data-flow languagges (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
@@
 \begin{itemize}
-\item[--] The function is represented as executable code for an call site environment
+\item The function is represented as executable code for an call site environment
   and is executed there, possibly after it is marshalled and transferred to another machine.
 \vspace{-0.5em}
-\item[--] The function body is compiled for the declaration site environment; the value
+\item The function body is compiled for the declaration site environment; the value
   that is returned is a remote reference to the code and function calls are performed as remote invocations.

### 2014-12-14 text/Chapter04.tex
Finishing proofreading (Chapter 6)

@@
 \begin{itemize}
-\item[--] We present a \emph{flat coeffect calculus} as a type system that is parameterized by a 
+\item We present a \emph{flat coeffect calculus} as a type system that is parameterized by a 
   \emph{flat coeffect algebra} (Section~\ref{sec:flat-calculus}). We show that the system can be 
@@
 
-\item[--] We give the equational theory of the calculus and discuss type-pre\-ser\-vation for call-by-name
+\item We give the equational theory of the calculus and discuss type-pre\-ser\-vation for call-by-name
   and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
@@
 
-\item[--] We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
+\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
   generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).

### 2014-12-14 text/Chapter05.tex
Finishing proofreading (Chapter 6)

@@
 \begin{itemize}
-\item[--] We present a \emph{structural coeffect calculus} as a type system that is parameterized 
+\item We present a \emph{structural coeffect calculus} as a type system that is parameterized 
   by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the 
@@
 
-\item[--] We extend indexed comonads introduced in the previous chapter to 
+\item We extend indexed comonads introduced in the previous chapter to 
   \emph{structural indexed comonads} and use them to provide the semantics of structural 
@@
 
-\item[--] We discuss the equational theory of the calculus. For structural coeffects, the substitution 
+\item We discuss the equational theory of the calculus. For structural coeffects, the substitution 
   lemma holds for all instances of the calculus. Thus, we prove the type-preservation property 

### 2014-12-14 text/Chapter06.tex
Finishing proofreading (Chapter 6)

@@
 
-\chapter{Unified coeffect language ???} 
+\chapter{Unified coeffect languages} 
 \label{ch:unified}
@@
 %===================================================================================================
-
-\section{Introduction}
 
@@
 dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
-two kinds of contextual properties that we further covered separately -- flat coeffects in 
-Chapter~\ref{ch:flat} track whole-context properties and structural coeffects, covered in 
-Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify
-the two notions. We introduce a \emph{unified coeffect} calculus that generalizes the two systems
-and can be instantiated to track both flat and structural properties (Section~\ref{sec:unified-unified}).
+two kinds of contextual properties that we further covered separately. Flat coeffects in 
+Chapter~\ref{ch:flat} track whole-context properties, while structural coeffects, covered in 
+Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify the two 
+notions.
 
 Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
-believe that coeffects should be integrated in main-stream programming languages. In the second
-part of this chapter, we discuss an alternative approach to defining coeffect 
-systems which highlights the relationship between our work and related work arising from modal logics 
-(Section~\ref{sec:unified-meta}). Finally, we outline one possible approach for practical
-implementations of coeffects (Section~\ref{sec:unified-impl}).
+believe that coeffects should be integrated in main-stream programming languages. This is addressed
+in the second part of the chapter, where we outline possible approach for a practical implementation
+of coeffect systems. Finally, we conclude the chapter with a brief discussion of an alternative
+approach to coeffect system, based on the meta-language style.
+
+\paragraph{Chapter structure and contributions}
+\begin{itemize}
+\item We start by unifying the two kinds of coeffects discussed so far. In 
+  Section~\ref{sec:unified-unified}, we introduce the \emph{unified coeffect calculus}, which 
+  generalizes flat and structural systems and can be instantiated to track both flat and structural 
+  properties.
+
+\item We outline one possible approach for practical implementations of coeffects 
+  (Section~\ref{sec:unified-impl}). The technique is based on embedding comonadic computations
+  via a lightweight syntactic extension and using rich type systems for capturing the structure of 
+  coeffect algebra.
+
+\item Finally, we discuss an alternative approach to defining coeffect systems based on the
+  meta-language style, which highlights the relationship between our work and related work arising 
+  from modal logics such as CMTT \cite{logic-cmtt} (Section~\ref{sec:unified-meta}).
+\end{itemize}
+
+\noindent
+The unified coeffect calculus presented in the first part of the chapter is a novel result that
+completes the theory developed in this thesis. The next two sections present important future and
+related work, respectively.
 
@@
 
-The flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural coeffect calculus 
-differ in a number of ways (Figure~\ref{fig:struct-types}). Understanding the differences is the 
-key to reconciling the two systems:
+The type systems of the flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural 
+coeffect calculus (Figure~\ref{fig:struct-types}) differ in a number of ways. Understanding the 
+differences is the key to reconciling the two systems:
 
 \begin{itemize}
-\item Structural coeffect calculus contains explicit rules for context manipulation
+\item The structural coeffect calculus contains explicit rules for context manipulation
   (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
@@
   and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
-  the variable context is a set and is annotated with a single (scalar) coeffect.
-
-\item In the flat coeffect calculus, we distinguish between splitting of the context requirements
-  and merging of context requirements ($\cpar$ and $\czip$, respectively). In the structural
-  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
-  and so the structural coeffect algebra requires just $\atimes$.
+  the variable context is a finite partial function (from names to types) and is annotated with 
+  a single (scalar) coeffect.
+
+\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
+  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
+  coeffect calculus, the operations (which model appending and splitting vectors) are invertible 
+  and so the structural coeffect algebra uses the same tensor product $\atimes$.
 \end{itemize}
@@
 In the unified calculus presented in this section, we address the three differences as follows.
-We use calculus with explicit rules for context manipulation. In systems that arise from flat
-calculi, the rules can be applied freely without changing the coeffects. We generalize the 
-structure of coeffect annotations using the notion of a ``container'' which can be specialized 
-to obtain a single annotation or a vector of annotations. Finally, we distinguish between 
-splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
+We include explicit rules for context manipulation in the calculus; in systems that arise from flat
+calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize 
+the structure of coeffect annotations using the notion of a ``container'' which can be specialized 
+to obtain a single annotation or a vector of annotations. This could potentially be used to capture
+other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish 
+between splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
 respectively). For structural coeffect calculi, the two operators coincide, but for flat
@@
 \subsection{Shapes and containers}
-
-Our notion of a \emph{coeffect container} is based on the idea of a container introduced by 
-Abbott et al. \cite{types-containers}. Interestingly, the work on containers has later been linked
+Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of 
+Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
 to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
-types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of lists).
+types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of 
+lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise 
+as some containers may have multiple shapes of the same length, \eg~differently ballanced trees.}.
 For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
@@
 \begin{definition}
-A \emph{container} $S \triangleleft P$ is given by a set $S$ of shapes and a shape-indexed family
-$P : S \rightarrow \ident{Set}$ of positions.
+A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$. 
+Here, $S$ is a set of shapes and $P$ is a shape-indexed family $P : S \rightarrow \ident{Set}$ 
+of positions.
 \end{definition}
@@
 \noindent
-Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams and 
-singleton data type (which contains exactly one element). Two containers relevant to our work
-are lists and singleton data types:
+Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams,
+trees and the singleton data type (which contains exactly one element). Containers relevant 
+to our work are lists and singleton data types:
 
@@
   shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
-  given length $n$ is a set of indices $P n = \{ 1 \ldots n \}$.
-
-\item The contianer representing singleton data type is given by $S \triangleleft P$ where
+  given length $n$ is the set of indices $P(n)= \{ 1 \ldots n \}$.
+
+\item The contianer representing the singleton data type is given by $S \triangleleft P$ where
   shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
-  shape $\ast$ contains exactly one position $P \ast = \{ 0 \}$.
+  shape $\ast$ contains exactly one position. To follow the intuition based on offset or 
+  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
 \end{itemize}
@@
 \subsection{Structure of coeffects}
-In the structural coeffect calculus, the structure annotation was formed by a vector of coeffect
-sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
-\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
-a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
-(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
-\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
-define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
-are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
-algebra}} which consists of shape-indexed coeffect scalar values.
+In the structural coeffect calculus, the coeffect annotation was formed by a vector of coeffect
+scalars. The annotations in the unified coeffect calculus are similar, but a \emph{vector} is 
+replaced with a more general \emph{container}. The primitive coeffect annotations in the unified 
+calculus are formed of \emph{coeffect scalars}. The coeffect scalars are equipped with the same
+structure as in structural coeffect calculus (Definition~\ref{def:structural-scalar}). In this 
+section, we refer to it as \emph{\cclrd{unified coeffect scalar}} (and we repeat the definition 
+below). Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect 
+scalar values are attached to the free-variable context. Finally, we define the 
+\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.
 
@@
 
-\paragraph{Coeffect scalar.}
-The following definition of the coeffect scalar structure repeats the Definition~\ref{def:structural-scalar}
-from the previous chapter.
+\paragraph{Coeffect scalars.}
+The following definition of the coeffect scalar structure repeats the 
+Definition~\ref{def:structural-scalar} from the previous chapter.
 
@@
 \textnormal{(pre-order)}   
+\\
+\end{array}
+\end{equation*}
+%
+The following distributivity axioms hold:
+\begin{align*}
+\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+%
+In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
+\begin{equation*}
+\begin{array}{l}
+ \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
+ \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
 \end{array}
@@
 program. The meaning depends on the coeffect container. The operation can either combine requirements
-of individual variables (structural) or requirements attached to the whole context of multiple
-sub-exp\-ressions (flat).
+of individual variables (structural coeffects) or requirements attached to the whole context of 
+multiple sub-exp\-ressions (flat coeffects).
 
 \paragraph{Coeffect containers.}
-The coeffect container is a container that determines how are scalar coeffect annotations attached 
-to free-variable contexts. In addition to a container $\SHP$ with shapes and shape-indexed positions, the
-coeffect container also provides a mapping that returns the shape of a free-variable context. The
-mapping between the shape of the variable context and the shape of the coeffect annotation
+The coeffect container is a container that determines how are scalar coeffects attached to 
+free-variable contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed positions, 
+the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.
+
+The mapping between the shape of the variable context and the shape of the coeffect annotation
 is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
@@
 In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
-$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
-We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation $\SP\sclrd{s}$
-returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in \SP\sclrd{s} \rightarrow \C$ 
-as a mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
-the exponent $\aclrd{\textbf{r}} \in \C^{\SP\sclrd{s}}$.
-
+$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
+to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation 
+$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in 
+  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects. 
+We usually write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
 The coeffect container is also equipped with an operation that appends shapes (when we concatenate
-variable contexts) and two special shapes in $\SH$ representing empty context and singleton context.
+variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton 
+context.
 
@@
 
-The elements $\sempty$ and $\sunit$ represent the shapes of empty and singleton free-variable contexts 
-respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts. Given
-$\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
+The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable 
+context, respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts.
+Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
 we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
@@
 \noindent
-As said earlier, we use two kinds of coeffect containers that describe the structure of vectors 
+As discussed earlier, we use two kinds of coeffect containers that describe the structure of vectors 
 (for structural coeffects) and the shape of trivial singleton container (for flat coeffects):
@@
 Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
-where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
+where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
 number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
@@
 Therefore, a coeffect annotation is a \emph{vector} 
-$\aclrd{\textbf{r}} \in \C^{\SP\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
+$\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
 for each position (corresponding to a variable $x_i$ in the context).
@@
 Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
-The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
+The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
 with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.
@@
 $\sflat$ with a single position and all free-variable contexts have the same singleton shape.
-Therefore, a coeffect annotation is drawn from $\C^\{\sflat\}$ which is isomorphic to $\C$
+Therefore, a coeffect annotation is drawn from $\C^{\{\sflat\}}$ which is isomorphic to $\C$
 and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
@@
 
-\newpage
 \begin{example}
 Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
-\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP \sflat = \emptyset$,
+\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
 ${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$. 
-
 This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
-$\SP\sflat=\emptyset$ and so coeffect annotations would be from the set $\C^\emptyset$.
+$\SP(\sflat)=\emptyset$ and so coeffect annotations would be from the empty set $\C^\emptyset$.
 \end{example}
@@
 \paragraph{Unified coeffect algebra.}
-The coeffect calculus annotates judgments with shape-indexed coeffect annotations.
-The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and coeffect container
+The unified coeffect calculus annotates typing judgments with shape-indexed coeffect annotations.
+The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and a coeffect container
 to define shape-indexed coeffects and operations for manipulating these.
@@
 the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
-while structural coeffects use tensor $\atimes$. In the unified calculus, we use two operators
-that can, however, be coincide.
+while structural coeffects use the tensor $\atimes$. In the unified calculus, we use two operators
+that can, however, coincide.
 
@@
 \sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
-coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP \sempty}$
+coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
 is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
@@
 \begin{array}{rclcrcl}
- \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
+ \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP(\sunit)}  \\
  \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
 \\[-0.5em]
- \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
+ \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
  \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
 \\[-0.5em]
- \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
+ \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
  \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
@@
 The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
-the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
-vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
-as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
-$\sclrd{m}$ and write $\aseq$.
-
-Finally, we define $\alen{}$ as an operation that returns the shape of a given shape-indexed 
-coeffect. The only purpose is to simplify notation, as we often need to specify that shapes of 
-variable context and coeffect match, \eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.
+the shape of a singleton context. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
+vector by a scalar. As we always use bold face for vectors and ordinary face for scalars (as well 
+as a distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
+$\sclrd{m}$ and write~$\aseq$.
+
+Finally, we define $\alen{-}$ as an operation that returns the shape of a given shape-indexed 
+coeffect. The only purpose is to simplify notation, as we tend to avoid subscripts, but often need 
+to specify that shapes of variable context and coeffect match, 
+\eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.
 
@@
 For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
-\C^{\SP\sclrd{m}}$ and $\aclrd{\textbf{s}} \in \C^{\SP\sclrd{n}}$. In the structural system, the 
+\C^{\SP(\sclrd{m})}$ and $\aclrd{\textbf{s}} \in \C^{\SP(\sclrd{n})}$. In the structural system, the 
 context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
@@
 \vspace{0.5em}
-\caption{Type system for the unified coeffect calculus}
+\figcaption{Type system for the unified coeffect calculus}
 \label{fig:unif-types}
@@
 shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
-the key differences between the two.
+how the unified system differs from the structural system.
 
@@
 As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
-use of variables -- the associativity is built-int. The order of variables matters, but can be changed 
-using the structural rules. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}} \in \C^\SH$ 
+use of variables -- the associativity is built-in. The order of variables matters, but can be changed 
+using the exchane rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ 
 are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
-annotated with coeffects scalars. % $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.
+annotated with coeffects scalars.
 
@@
 coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$. 
-The (\emph{const}) rule annotates empty context with a special annotation of the unit shape.
+The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.
 
@@
 In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
-system is that vector concatentation $\atimes$ is now replaced with context splitting/merging 
-of the unified coeffect algebra. In particular, we use splitting of context requirements $\azip$
+system is that the vector concatentation $\atimes$ is now replaced with context splitting/merging 
+of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
 in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.
@@
 elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch}) 
-rule denoted a two-element vector. In the unified system, this is replaced with merging/splitting
+rule denoted two-element vectors. In the unified system, this is replaced with merging/splitting
 of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.
 
-In structural systems, this means the same thing -- we are simply concatenating or splitting two 
-singleton vectors. However, the generalization allows us to capture flat coeffect systems as well.
+In structural systems, the two notations mean the same thing -- we are simply concatenating or splitting two 
+singleton vectors. However, this generalization allows us to capture flat coeffect systems as well.
 The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
-$\apar$ correspond to operations on coeffect scalars. And, as discussed in Section~\ref{sec:unified-flat},
+$\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:unified-flat},
 thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not 
-affect the coeffect annotation are admissible for all flat systems.
+affect the coeffect annotation are admissible for all flat systems embedded in the unified 
+coeffect calculus.
 
@@
 structural and flat systems in mind. In this and the next section, we show how it captures 
-the two kinds of systems.
-
-The unified calculus is modelled after the structural system and so using it to model structural
+the two coeffect systems from Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}.
+
+The unified calculus closely resembles the structural system and so using it to model structural
 systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector} 
@@
 
-\begin{itemize}{}
-\item[--] Coeffect container $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$ where $\SH=\mathbb{N}$ and
-  $\SP \sclrd{n}=\{ 1 \ldots n \}$ and  ${|}x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n{|}=n$.
-
-\item[--] Coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
+\begin{itemize}
+\item A coeffect container $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$ where $\SH=\mathbb{N}$ and
+  $\SP(\sclrd{n})=\{ 1 \ldots n \}$ and  ${|}x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n{|}=n$.
+
+\item A coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
   shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
-  That is $\times : \C^{\SP\sclrd{n}} \times \C^{\SP\sclrd{m}} \rightarrow \C^{\SP(\sclrd{n + m})}$ 
-  appends vectors and $\epsilon : \C^{\SP\sclrd{0}}$ represents empty vectors.
+  That is $\epsilon : \C^{\SP(\sclrd{0})}$ is an empty vector and 
+  $\times : \C^{\SP(\sclrd{n})} \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{n + m})}$ 
+  appends vectors.
 \end{itemize}
@@
 Using the above definition, our unified coeffect system can capture per-variable coeffect properties
-discussed in Chapter~\ref{ch:applications}, Section~\ref{sec:applications-structural}. This includes
-the system for bounded reuse (which is only used in the structural form) and precise tracking 
-of per-variable dataflow and liveness.
+discussed in Section~\ref{sec:applications-structural}. This includes the system for bounded reuse 
+(which is only used in the structural form) and precise tracking of per-variable dataflow and liveness.
 
@@
 coeffects such as implicit parameters. This is achieved using a singleton-shaped container for
-coeffect annotations. The resulting system is has explicit structural rules (and is syntactically
+coeffect annotations. The resulting system has explicit structural rules (and is syntactically
 different from the standard flat coeffect system), but we show that they are equivalent.
 
-Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:unified-flat-shape})
-In this setting, the context annotations $\C^{ \SP \sflat } = \{0\} \rightarrow \C$ are equivalent 
-to coeffect scalars $\C$. In addition to the coeffect scalar structure, we also need to define $\apar$ 
+Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:unified-flat-shape}).
+In this setting, the context annotations $\C^{ \SP(\sflat) } = \{0\} \rightarrow \C$ are equivalent 
+to coeffect scalars $\C$. In addition to the coeffect scalar structure, we need to define $\apar$ 
 and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ 
@@
 $\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, 
-the unified coeffect algebra modelling flat coeffect systems consists of:
+the unified coeffect algebra modelling a flat coeffect systems consists of:
 %
 \begin{itemize}{}
-\item[--] Flat coeffect container $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ as defined in 
+\item A flat coeffect container $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ as defined in 
   Example~\ref{ex:unified-flat-shape}.
 
-\item[--] Flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$ 
-  and $\azero = \czero$ with an additional binary operation $\azip = \czip$. 
+\item A flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$ 
+  and $\azero = \czero$ with an\\ additional binary operation $\azip = \czip$. 
 \end{itemize}
@@
 work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
-the existence of unit of $\czip$).
+the existence of a unit for $\czip$).
 
@@
 
-The semantics of unified calculus can be obtained by generalizing the semantics of the structural
+The semantics of the unified calculus can be obtained by generalizing the semantics of the structural
 calculus in the same way as we generalized the type system. However, it is worth exploring the type 
@@
 index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of 
-the structure $\C^{\SP \sclrd{n}}$ for a specified container $\SH \triangleleft \SP$ and a shape 
+the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape 
 $\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in 
@@
 splitting and then merging a context may lose some of the avilable contextual capabilities
-(and increase context requirements). This corresponds to the fact that our proofs of 
-Lemma~\ref{thm:unified-rev-contr} and Lemma~\ref{thm:unified-rev-exch} relied on sub-coeffecting
-in the flat system.
+(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:unified-rev-contr} 
+and Lemma~\ref{thm:unified-rev-exch} relied on sub-coeffecting in the flat system.
 
 \paragraph{Related and future work.}
-The semantics presented here serves mainly to inform the design of the type system. For this 
+The semantics presented here serves main\-ly to inform the design of the type system. For this 
 reason, we keep the semantics simple and close to the concrete notions used in the type system
@@
 \begin{compactitem}
-\item[-] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
-\item[-] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is 
+\item[--] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
+\item[--] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is 
   indexed by vectors of scalars $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$
-\item[-] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
+\item[--] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
   the mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is indexed by shape-indexed scalars 
-  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP\sclrd{m}}$
+  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP(\sclrd{m})}$
 \end{compactitem}
@@
 There, the index (coeffect scalars and shape-indexed scalars) are modelled as categories too.
-The definition requires the following notation:
+The definition is written using the following notation:
 
 \begin{compactitem}
-\item[-] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
-\item[-] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
+\item[--] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
+\item[--] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
 \end{compactitem}
@@
 \begin{equation*}
-\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP \sclrd{n}} \rightarrow [\mathbb{C}^{\SP \sclrd{n}}, \mathbb{C}]
+\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]
 \end{equation*}
@@
 For a fixed context shape $\sclrd{n}$ the functor 
-$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP \sclrd{n}} \rightarrow [\mathbb{C}^{\SP \sclrd{n}}, \mathbb{C}]$ 
-maps a coeffect indexed by positions $\SP \sclrd{n}$ to a functor from a context 
+$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]$ 
+maps a coeffect indexed by positions $\SP(\sclrd{n})$ to a functor from a context 
 $\mathbb{C}^{\sclrd{n}}$ to an object in $\mathbb{C}$. That is, given a coeffect annotation 
@@
 
-In the model above, shapes and positions are still treated as ordinary sets. Indeed, the paper
-\cite{coeffects-icfp14} treats shapes as just sets of sets. In this chapter, we instead define
-\emph{coeffect container} as an extension of containers, which have a well-defined 
+In the model above, shapes and positions are still treated as ordinary sets. Indeed, in the 
+joint paper \cite{coeffects-icfp14}, we treat shapes as just sets of sets. In this chapter, we 
+instead define \emph{coeffect container} as an extension of containers, which have a well-defined 
 categorical structure \cite{semantics-containers}. A future work is adapting the semantics of
@@
 annotations) of the object that wrap the free-variable context.
-
-% =================================================================================================
-%                                                                                             
-%    #   #          #                    ##                                                    
-%    #   #          #                     #                                                    
-%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
-%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
-%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
-%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
-%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
-%                                                            #   #                #   #        
-%                                                             ###                  ###         
-% =================================================================================================
-
-\section{Coeffect meta-language}
-\label{sec:unified-meta}
-
-In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
-language semantics introduced by Moggi \cite{monad-notions}. The first approach is to use
-monads in the \emph{semantics} of an effectful language. The second approach is to extend the
-language with (additional) monadic constructs that can then be used for writing effectful
-monads explicitly.
-
-In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
-the term language is that of simply-typed $\lambda$-calculus, and we used (flat or structural)
-indexed comonads to give the semantics for the language and to derive type system for it.
-
-In this section, we briefly discuss the alternative approach. That is, we embed indexed
-comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
-type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
-context (semantically, this corresponds to an indexed comonad) and we add language constructs
-that correspond to the operations of indexed comonads.
-
-This section provides a brief sketch of coeffect meta-language to highlight the relationship
-between coeffects and important related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
-Developing the system further is an interesting future research direction.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffects and contextual modal type theory}
-
-As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
-to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
-meta-languages that include $\square$ as a type constructor
-\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
-Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
-From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
-coeffect calculus.
-
-\paragraph{Context in CMTT and coeffects.}
-Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
-comonads (the $\square$ modality) into the meta-language, there are two important differences.
-
-Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
-requirements are formed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}
-
-Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
-the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
-constructor refers to a value of type $\tau$ \emph{together} with additional context specified
-by $\cclrd{r}$ (\eg~list of past values or additional resources). In contrast in 
-CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
-original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
-context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
-two operations of a comonad:
-\begin{equation*}
-\begin{array}{rcl}
- \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
- \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
-    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
-\end{array}
-\end{equation*}
-Both readings are possible, but they give quite different meanings to the operations:
-
-\begin{itemize}
-\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
-  and does not require any additional context; the \ident{cobind} operation requires
-  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
-  evaluate the function, ending with a value $\beta$ together with remaining context
-  $\cclrd{s}$.
-\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
-  that requires no additional context to obtain a $\alpha$ value; given a function that
-  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
-  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
-  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
-  context requirements is eliminated by the function).
-\end{itemize}
-
-\noindent
-Although the different reading does not affect formal properties of the systems, it is important
-to understand the difference when discussing the two systems.
-
-The sketch of a coeffect meta-language in the following section attempts to bridge the gap
-between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
-it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
-sets of variables. Future work on the coeffect meta-language would thus be an interesting 
-development for both coeffect systems and CMTT.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect meta-language}
-
-The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
-For simplicity, this section only discusses the flat variant. The syntax of types and terms
-of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four expression forms:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
- e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
-  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
-  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
-  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
-\end{array}
-\end{equation*}
-%
-The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
-operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
-\kvd{split} and \kvd{merge} that embed the corresponding operations.
-
-\paragraph{Types for coeffect meta-language.}
-The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
-shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
-rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
-
-The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
-a value from a comonadic context and corresponds to variable access in coeffect calculi. The
-\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
-and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
-The result is a computation that produces a $\tau_2$ value with the remaining context specified
-by $\cclrd{s}$. Note that the expression $e_2$ and $e_1$ corresponds to the first and second
-arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
-CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
-although it differs because of our generalization of comonads where \ident{bind} composes
-coeffect annotations rather than requiring the same annotation everywhere.}.
-
-The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
-transformation on one or two values in a context and then add the new value as a fresh variable
-to the variable context. For simplicity, we omit sub-coeffecting, but it could be easily added 
-following the pattern used elsewhere.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-
-{\small a.) Typing rules for the simply typed $\lambda$-calculus}
-
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\Gamma \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
-   \Gamma \vdash e_2 : \tau_2 }
-  {\Gamma \vdash e_1~e_2 : \tau_1 }
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\Gamma, x:\tau_1 \vdash e : \tau_2}
-  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
-\end{equation*}
-
-\vspace{1em}
-{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
-
-\begin{equation*}
-\tyrule{cobind}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
-\end{equation*}
-\begin{equation*}
-\tyrule{counit}
-  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
-  {\Gamma \vdash !e : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{split}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
-   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{merge}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
-   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
-   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
-  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
-\end{equation*}
-
-\caption{Type system for the (flat) coeffect meta-language}
-\label{fig:conclusions-cml}
-\end{figure}
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Embedding flat coeffect calculus}
-
-The \emph{meta-language} approach of embedding comonads in a language is more general than the
-\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
-a type system for context-aware programming languages.
-
-However, it is worth noting that 
-(flat) coeffect calculus can be embedded in the meta-language described above. This may be 
-desirable, \eg~when using the meta-language for reasoning about context-aware computations. 
-We briefly consider the embedding as it illuminates the relationship between coeffects and
-CMTT (although it is not possible to embed coeffect calculi in CMTT because of the more general
-annotations structure).
-
-Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in a flat coeffect calculus,
-we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
-meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
-to represent the entire variable context of the source language. The translation is defined in
-Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
-Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
-the operations of a monad.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\begin{equation*}
-\begin{array}{l}
- \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
-   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
-   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
-      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
-   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
-\end{array}
-\end{equation*}
-\caption{Embedding flat coeffect calculus in coeffect meta-language }
-\label{fig:conclusions-embed}
-\end{figure}
 
@@
 As discussed earlier, the main focus of this thesis is the development of the much needed 
-\emph{theory of context-aware computations} and so discussing the details of a practical 
+\emph{theory of context-aware programming languages} and so discussing the details of a practical 
 implementation of coeffect tracking is beyond the scope of the thesis. However, this section
@@
 common abstraction. For this reason, we advocate that future practical implementations of
-coeffects should not be single-purpose language features, but instead reusable abstractions that
+coeffects should not be single-purpose language features, but rather reusable abstractions that
 can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
@@
 
-The embedding of \emph{contextual} computations in programming languages can learn from better 
-explored emedding of \emph{effectful} computations. In purely functional programming languages
-such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
-necessary (monadic) plumbing. This is made easier by the ``do'' notation \cite{other-haskell98} 
-that inserts the monadic operations automatically.
-
-The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
-context-aware computations based on comonads. The notation is close to the semantics of our 
+The embedding of \emph{contextual} computations into programming languages can follow the successful
+model of \emph{effectful} computations. In purely functional programming languages such as Haskell, 
+effectful computations are embedded by \emph{implementing} their model within the language 
+and inserting the necessary (monadic) plumbing. This is made easier by the ``do'' notation or
+monad comprehensions \cite{other-haskell98,monads-compre}, which insert the monadic operations 
+automatically.
+
+The recently proposed ``codo'' notation \cite{comonads-codo} provides similar automation for
+context-aware computations based on comonads. The notation follows the semantics of our 
 flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based 
-on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
-this requires explicitly manipulating individual context variables and application of structural
+on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting future work -- 
+this would require explicitly manipulating individual context variables and application of structural
 rules, which is not needed in flat coeffects.
@@
 In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
-but they can still use a special notation for explicitly marking effectful (coeffectful) 
+but they can still benefit from a special notation for explicitly marking effectful (coeffectful) 
 blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
 that differ from the ``do'' notation in two ways. First, they support wider range of 
-constructs, making it possible to wrap existing F\# code in a block without other changes.
+constructs, making it possible to wrap existing F\# code within a block without other changes.
 Second, they support other abstractions including monads, applicative functors and monad
 transformers. It would be interesting to see if computation expressions can be extended to
-handle computations based on flat/structural indexed comonads.
+support programming with computations based on flat/structural indexed comonads.
 
 More lightweight syntax for effectful computation can be obtained using
-techniques that automatically insert the necessary monadic plumbing (wihtout a
-special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
-and similar approach would be worth exploring for coeffects.
+techniques that automatically insert the necessary monadic plumbing (without
+special syntax). This has been done for effecful computations \cite{monads-lightweight-ml}
+and a similar approach would be worth exploring for coeffects.
 
@@
 
-The other aspect of practical implementation of coeffects is tracking the context requirements
+The other aspect of practical implementation of coeffects is that of tracking the context requirements
 (coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
 language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
-The structures used in this thesis include sets (with union or interesection), natural numbers (with 
+The structures required in this thesis include sets (with union or interesection), natural numbers (with 
 addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
@@
 
-The work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
-recent additions to the Haskell type system provide enough poer to implement structures such
+The work \cite{effects-embedding} on embedding effect systems in Haskell demonstrated that the
+recent additions to the Haskell type system provide enough power to implement structures such
 as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful 
@@
 
-In dependently-typed programming languages such as Agda \cite{other-agda} or Idris \cite{other-idris},
-the embedding of coeffects can be implemented more directly. However, we believe that coeffect
-tracking does not require full dependent types and can be made accessible in more main-stream
-languages. Dependent ML \cite{types-dependent-ml} provides an interesting example of a language
-with some dependent typing which is still close to its non-dependently-typed predecessor ML.
+In dependently-typed programming languages such as Agda or Idris \cite{other-agda,other-idris},
+the embedding of coeffects can be implemented more directly (as terms implementing sets or lattices
+can be lifted to the type level). However, we believe that coeffect tracking does not require full 
+dependent types and can be made accessible in more mainstream languages. Dependent ML 
+\cite{types-dependent-ml} provides an interesting example of a language with some dependent typing 
+(\eg~computations with integers) which is still close to its non-dependently-typed predecessor ML.
 
@@
 access to data or services. A similar approach could be used for embedding \emph{algberas} such
-as coeffect algebras into the type system. A \emph{algebra provider} would be a library that 
-specifies the objects of the algebra, equational laws and generalization rules for type inference.
-This could provide an easy to use way of embedding coeffect tracking in pragmatic languages
+as coeffect algebras into the type system. An \emph{algebra provider} would then be a library that 
+specifies the objects of the algebra, its equational laws and generalization rules for type inference.
+This could provide an easy-to-use way of embedding coeffect tracking in pragmatic languages
 such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
-\cite{types-units-of-measure}, which could be provided via one such \emph{algebra provider}.
+\cite{types-units-of-measure}; these could be alternatively provided via one such \emph{algebra
+provider}.
 
@@
 
-The fact that coeffect scalar structure consists of four operations ($\cseq,\czip,\cpar$ 
-and a relation $\cleq$) makes working with coeffect annotations difficult -- it compilcates tasks
+Compared to typical effect systems based on sets, coeffect systems have richer structure consisting
+of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting 
+properties, but it makes working with coeffect annotations difficult -- it compilcates tasks
 such as type generalization and constraint solving in type inference. Here, we look at one possible 
@@
 well as flat systems for liveness and data-flow, but not with the system for tracking of implicit
-parameters (and other sets).
+parameters (or sets of rebindable resources).
 \end{itemize}
@@
 The system for tracking implicit parameters does not fit the lattice-based model, because it uses the 
-same operation $\cup$ for both $\cpar$ and $\czip$. This also means that the system for implicit parameters 
-cannot be used with a system where lambda abstraction duplicates the context requirements 
-(Section~\ref{sec:flat-exts-lambda}).
+same operation $\cup$ for both $\cpar$ and $\czip$.
+
+This also means that the system for implicit parameters cannot be used with a system where lambda 
+abstraction duplicates the context requirements (Section~\ref{sec:flat-exts-lambda}).
 
@@
 \noindent
-The definition provides the same structure as flat coeffect algebra, but it defines a more rigid 
-structure. In particular, $\czip$ and $\cpar$ are defined in terms of $\cleq$. This simplifies the 
-system as the operations of the underlying indexed comonad structure can be expressed using just
+The definition provides all the operations of a flat coeffect algebra, but it defines a more rigid 
+structure. In particular, $\czip$ and $\cpar$ are both defined in terms of $\cleq$. This simplifies 
+the system as the operations of the underlying indexed comonad structure can be expressed using just
 $\cseq$ and $\cleq$:
@@
 %
-Note that we still need two \emph{distinct} operators in the system. The one used for
-sequential composition $\cseq$ is not necessarilly related in any way to the ordering on
-the lattice $\cleq$.
+Note that we still need \emph{two distinct} operators in the system. Analysis of our examples shows
+that the operator used for sequential composition $\cseq$ is not necessarilly related in any way to 
+the ordering on the lattice provided by $\cleq$.
 
@@
 $\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
-This means that we can compose the \ident{merge} operation which returns greatest lower
+This means that we can compose the original \ident{merge} operation which returns greatest lower
 bound with $\ident{lift}$ to get the (less precise) operation required here.
@@
 {\small a.) Littice-based flat systems (selected rules):}
+
 \begin{equation*}
@@
 
-\vspace{1em}
+\vspace{1.5em}
 {\small b.) Littice-based structural systems (selected rules):}
+
 \begin{equation*}
@@
 
-\caption{Flat and structural coeffects using lattice-based formulation}
+\figcaption{Flat and structural coeffects using lattice-based formulation}
 \label{fig:unified-lattice-types}
@@
 \paragraph{Lattice-based type system.}
-
+\hspace{-1em}
 Using lattice-based coeffect structure leads to the typing rules shown in Figure~\ref{fig:unified-lattice-types}.
@@
 Structural coeffect systems only require least upper bound $\cpar$ in the (\emph{contr}) rule.
-This can be rewritten similarly to flat application --  when contracting two variables, the
+This can be rewritten similarly to the flat (\emph{app}) rule --  when contracting two variables, the
 context requirements of the resulting variable are greater than the requirements associated with
@@
 the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.
+
+% =================================================================================================
+%                                                                                             
+%    #   #          #                    ##                                                    
+%    #   #          #                     #                                                    
+%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
+%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
+%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
+%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
+%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
+%                                                            #   #                #   #        
+%                                                             ###                  ###         
+% =================================================================================================
+
+\section{Coeffect meta-language}
+\label{sec:unified-meta}
+
+In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
+language semantics as introduced by Moggi \cite{monad-notions}. The first approach is to use
+monads in the \emph{semantics} of an effectful language. The second approach is to extend the
+language with (additional) monadic constructs that can then be used for writing effectful
+monads explicitly.
+
+In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
+the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
+indexed comonads to give the semantics for the language and to derive type system for it.
+
+In this section, we briefly discuss the other technique. That is, we embed indexed
+comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
+context (semantically, this corresponds to an indexed comonad) and we add language constructs
+that correspond to the operations of indexed comonads.
+
+This section provides a brief sketch of the coeffect meta-language in order to highlight the relationship
+between coeffects and closely related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
+Developing the system further is an interesting future research direction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffects and contextual modal type theory}
+
+As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
+to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
+meta-languages that include $\square$ as a type constructor
+\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
+Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
+From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
+coeffect calculus.
+
+\paragraph{Context in CMTT and coeffects.}
+Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
+comonads (the $\square$ modality) into the meta-language, there are two important differences.
+
+Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
+requirements are formed by an abstract coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, etc.
+
+Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
+the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+constructor refers to a value of type $\tau$ \emph{together} with additional context specified
+by the coeffect annotation $\cclrd{r}$ (\eg~a list of past values or additional resources).
+
+In contrast in CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
+original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
+context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
+two operations of a comonad:
+\begin{equation*}
+\begin{array}{rcl}
+ \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
+ \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
+    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
+\end{array}
+\end{equation*}
+Two readings of the signatures are possible, but they give quite different meanings to the operations:
+
+\begin{itemize}
+\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
+  and does not require any additional context; the \ident{cobind} operation requires
+  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
+  evaluate the function, ending with a value $\beta$ together with remaining context
+  $\cclrd{s}$.
+\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
+  that requires no additional context to obtain a $\alpha$ value; given a function that
+  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
+  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
+  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
+  context requirements is eliminated by the function).
+\end{itemize}
+
+\noindent
+Although the different reading does not affect formal properties of the systems, it is important
+to understand the difference when discussing the two systems as they provide a different 
+intuition.
+
+The sketch of a coeffect meta-language in the following section attempts to bridge the gap
+between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
+it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
+sets of variables. Future work on the coeffect meta-language would thus be an interesting 
+development for both coeffect systems and CMTT.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
+For simplicity, this section only discusses the flat variant. The syntax of types and terms
+of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
+language constructs:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
+ e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
+  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
+  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
+  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
+\end{array}
+\end{equation*}
+%
+The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
+operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
+\kvd{split} and \kvd{merge} that embed the corresponding operations.
+
+\paragraph{Types for coeffect meta-language.}
+The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
+shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
+rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
+
+The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
+a value from a comonadic context and corresponds to variable access in coeffect calculi. The
+\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
+and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
+The result is a computation that produces a $\tau_2$ value with the remaining context specified
+by $\cclrd{s}$. Note that the expressions $e_2$ and $e_1$ correspond to the first and second
+arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
+CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
+although it differs because of our generalization of comonads where \ident{bind} composes
+coeffect annotations rather than requiring the same annotation everywhere.}.
+
+The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
+transformation on one or two values in a context and then add the new value as a fresh variable
+to the variable context. We omit sub-coeffecting, but it could be easily added following 
+the method used elsewhere.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Typing rules for the simply typed $\lambda$-calculus}
+
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\Gamma \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
+   \Gamma \vdash e_2 : \tau_2 }
+  {\Gamma \vdash e_1~e_2 : \tau_1 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x:\tau_1 \vdash e : \tau_2}
+  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
+
+\begin{equation*}
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
+\end{equation*}
+\begin{equation*}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash !e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{split}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
+   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{merge}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
+   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
+   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
+  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
+\end{equation*}
+
+\figcaption{Type system for the (flat) coeffect meta-language}
+\label{fig:conclusions-cml}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding flat coeffect calculus}
+
+The \emph{meta-language} approach of embedding comonads in a language is more general than the
+\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
+a type system for context-aware programming languages.
+
+However, it is worth demonstrating that the (flat) coeffect calculus can be embedded in the 
+meta-language described in the previous section. This may be desirable, \eg~when using the 
+meta-language for reasoning about context-aware computations. We briefly consider the embedding 
+as it illuminates the relationship between coeffects and CMTT (unfortunately it is not possible 
+to embed coeffect calculi in CMTT because of the more general annotations structure).
+
+Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
+meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
+to represent the entire variable context of the source language. The translation is defined in
+Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
+Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+the operations of an indexed comonad.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{l}
+ \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
+   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
+   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
+      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
+   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
+\end{array}
+\end{equation*}
+\figcaption{Embedding flat coeffect calculus in coeffect meta-language }
+\label{fig:conclusions-embed}
+\end{figure}
 
@@
 
-In this chapter, we completed our goal of presenting a \emph{unified} system for tracking contextual
+In this chapter, we completed our quest of presenting a \emph{unified} system for tracking contextual
 properties and then we discussed two important topics of future and related work that require 

### 2014-12-15 text/Chapter07.tex
Further work and conclusions (Chapter 7)

@@
 
-\chapter{Conclusions} 
+\chapter{Conclusions and further work} 
 \label{ch:conclusions}
+
+Some of the most fundamental academic work is not the one solving hard research problems, but the 
+one that changes how we understand the world. Some philosophers argue that \emph{language} is the 
+key for understanding how we think, while in science the dominant thinking is determined by 
+\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
+In a way, programming languages play a similar role for computer science and software development.
+
+This thesis aims to change how developers and programming language designers think about 
+\emph{context} or \emph{execution environments} of programs. Such context or execution environment
+has numerous forms -- mobile applications access network, GPS location or user's personal data; 
+journalists obtain information published on the web or through open government data initiatives;
+in data-flow programming, we have access to past values of expressions.
+This thesis aims to change our understanding of \emph{context} so that the above examples are 
+viewed uniformly through a single programming language abstraction, which we call \emph{coeffects}, 
+rather than as disjoint cases.
+
+In this chapter, we give a brief summary of the technical development completed in this thesis
+(Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
+common patterns and captures those using three \emph{coeffect calculi}. Next, we give an overview
+of further work (Section~\ref{sec:conc-further}), often referring to longer discussion in earlier
+chapters. Finally, Section~\ref{sec:conc-conclusions} concludes the thesis.
 
@@
 
-Some of the most fundamental academic work is not the one solving hard research problems, but the 
-one that changes how we understand the world. Some philosophers argue that \emph{language} is the 
-key for understanding how we think, while in science the dominant thinking is determined by 
-\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
-In a way, programming languages play a similar role for computer science and software development.
+\section{Summary}
+\label{sec:conc-summary}
 
-This thesis aims to change how developers and programming language designers think about 
-\emph{context} or \emph{execution environments} of programs. Such context or execution environment
-has numerous forms -- mobile applications access network, GPS location or user's personal data; 
-journalists obtain information published on the web or through open government data initiatives. 
-This thesis aims to change our understanding of \emph{context} so that the above examples are 
-viewed uniformly through a single programming language abstraction, which we call \emph{coeffects}, 
-rather than as disjoint cases. This is done through the following three contributions.
+Modern computer programs run in rich and diverse environments or \emph{contexts}. The richness means 
+that environments provides additional resources and capabilities that may be accessed by the program. 
+The diversity means that programs often need to run in multiple different environments, such as mobile 
+phones, servers, web browsers or even on the GPU. In this thesis, we present the foundations for
+building programming languages that simplify writing software for such rich and diverse environments.
 
@@
 
-\section{Unified presentation.}
+\paragraph{Notions of context.}
 
-In $\lambda$-calculus, the term \emph{context} is usually used for the free-variable context.
-However, there are many other programming language features that are related to context or program's
+In $\lambda$-calculus, the term \emph{context} is usually refers to the free-variable context.
+However, other programming language features are also related to context or program's
 execution environment. In Chapter~\ref{ch:applications}, we revisit many of such features --
 including resource tracking in distributed computing, cross-platform development, data-flow 
-programming and liveness analysis, but also Haskell's type classes and implicit parameters.
+programming and liveness analysis, but also Haskell's implicit parameters.
 
@@
 
-\section{Flat coeffect calculus.}
+\paragraph{Flat coeffect calculus.}
 
@@
 of tracking \emph{whole-context} properties. The calculus is parameterized by a \emph{flat coeffect
-algebra} that captures the algebraic properties that contextual information satisfy. Concrete 
+algebra} that captures the algebraic properties of contextual information. Concrete 
 instances of flat coeffects include Haskell's implicit parameters, whole-context liveness and
@@
 In the flat coeffect calculus, $\beta$-reduction and $\eta$-expansion do not generally preserve
-type of expressions, but we identify two conditions when this is the case -- this gives us
+the type of an expression, but we identify two conditions when this is the case -- this gives us
 a basis for operational semantics of our calculi for liveness and implicit parameters.
+
 The design of the type system is further validated by categorical semantics
 (Section~\ref{sec:flat-semantics}) which models flat coeffect calculus in terms of 
-\emph{flat indexed comonad} -- a structure based on categorical dual of monads.
+\emph{flat indexed comonad} -- a structure based on a categorical dual of monads.
 
@@
 
-\section{Structural coeffect calculus.}
+\paragraph{Structural coeffect calculus.}
 
-In the second category of context-aware systems discussed in Section~\ref{sec:applications-structural},
-contextual properties are \emph{per-variable}. The systems discussed here resemble sub-structural
-logics, but rather than \emph{restricting} variable use, they \emph{track} additional information
-about how variables are used.
+~ The second category of context-aware systems discussed in Section~\ref{sec:applications-structural}
+captures \emph{per-variable} contextual properties. The systems discussed here resemble sub-structural
+logics, but rather than \emph{restricting} variable use, they \emph{track} how the variables are used.
 
@@
 Similarly to the flat variant, the calculus is parameterized by a \emph{structural coeffect algebra}. 
-Concrete instances of the calculus track bounded variable use (\ie~how many times is variable 
+Concrete instances of the calculus track bounded variable use (\ie~how many times is a variable 
 accessed), data-flow properties (how many past values are needed) and liveness (\ie~can variable 
@@
 The structural coeffect calculus has desirable equational properties (Section~\ref{sec:struct-syntax}).
-In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all structural
-coeffect calculi. This follows from the fact that structural coeffect associates contextual requirements
-with individual variables and preserves the connection by including explicit structural rules
-(weakening, exchange and contraction).
-
-The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
-thesis. In Chapter~\ref{ch:unified}, we develop the two calculi further by introducing a unified
-presentation of the two (by parameterizing the calculus by the \emph{shape} of context) and we present
-an alternative presentation that makes type inference easier.
+In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all instances of the
+structural coeffect calculus. This follows from the fact that structural coeffect 
+associates contextual requirements with individual variables and preserves the connection by 
+including explicit structural rules (weakening, exchange and contraction).
 
@@
 
+\paragraph{Unified calculus.}
+
+The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
+thesis. In Chapter~\ref{ch:unified}, we discuss the similarities and differences between the two. 
+Although the calculi are worth considering separately (as they have different equational properties),
+they can be presented in a uniform way. In Section~\ref{sec:unified-unified}, we present the 
+\emph{unified coeffect calculus}, which can be instantiate to track both per-variable and 
+whole-context properties. The key trick is to parameterize the calculus by the \emph{shape} of 
+a context, which can be a vector (corresponding to the vector of free variables) or a singleton
+container (attaching one annotation to the whole context).
+
+% =================================================================================================
+%                                                                                    
+% 	 #####                 #     #                                                #     
+% 	 #                     #     #                                                #     
+% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   # 
+% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #  
+% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###   
+% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #  
+% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   # 
+%                                                                                    
+% =================================================================================================
+
+\section{Further work}
+\label{sec:conc-further}
+
+Most of the further work has been discussed throughout the thesis, so this section
+serves mainly as an overview. We follow the same structure as when discussing pathways to 
+coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
+to coeffect systems, categorical semantics of coeffects, and connections with sub-structural type 
+systems and bunched logics.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Syntactic coeffect systems.}
+The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
+variables, lambda abstraction, application and let-binding. This lets us focus on fundamental
+properties of coeffect systems (and explain how coeffect system differ from better-known effect
+systems), but it is hardly sufficient for a realistic programming language.
+
+Further work is to extend the coeffect systems presented here to a full programming language. A
+useful reference is the work of Nielson and Nielson~\cite{effects-nielson} who consider a 
+similar development for effect systems, adding conditionals, recursion and polymorphism.
+
+Coeffect annotations in context-aware programming languages can contain significant amount
+of information. Thus, the programming langauge also needs a form of type inference that can propagate
+such information. This has been done for effect systems \cite{effects-polymorphic}. For 
+\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
+for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
+to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
+be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where 
+\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Language semantics.}
+
+As discussed in Section~\ref{sec:path-sem}, comonads can be used to define the semantics of a
+programming language in two ways. The first, ``language semantics'', approach is to use a 
+single comonad to describe the semantics of a specific context-aware property. The second,
+``meta-language'' approach is to extend the language with explicit constructs representing the 
+operations of the underlying comonad.
+
+The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
+guide for the development of the type systems. Further work includes more precise treatment of
+the categorical structure of indexed comonads. This has partly been done for flat coeffects by
+Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
+first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
+work is to develop a similar categorical model for the unified coeffect calculus, possibly using
+the categorical notion of containers \cite{types-containers}.
+
+We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:unified-meta}.
+Developing this technique further could unify coeffect systems with the work on Contextual Modal
+Type-Theory (CMTT) \cite{logic-cmtt} and allow other interesting applications of coeffects such
+as meta-programming \cite{logic-cmtt} and distributed computing with explicit modalities \cite{app-distributed-ml5}.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Substructural and bunched logics.}
+
+In the \emph{flat} and \emph{sturctural} coeffect calculi, we attach annotations to the \emph{entire
+context} and to \emph{individual variables}, respectively. We later unified the two systems 
+in Section~\ref{sec:unified-unified}.
+
+An intriguing question is whether coeffects can generalize \emph{bunched typing} \cite{substruct-bunched},
+which uses a tree-like structure of variable contexts (also discussed in Section~\ref{sec:path-logic}).
+The definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
+typing requires tree-like variable context, while we use a vector. Furthermore, bunched typing
+annotates internal nodes of the tree (sub-contexts are combined using ``;'' or using ``,'') rather 
+than just variables. Finding a simple coeffect system that is capable of capturing bunched typing
+is thus an interesting further work. Indeed, this could lead to numerous uses of coeffects as 
+bunched typing is the basis for widely-used separation logic \cite{substruct-separation-logic}.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Applications.}
+The last direction for further development is developing practical programming langauges or
+libraries based on the theory presented in this thesis. We believe that the most useful approach
+is to extend programming languages with a reusable coeffect system and allow library developers 
+define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
+For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
+``do'' notation.
+
+Another option is to use coeffects just as the underlying theory of a single-purpose programming
+language feature. When discussing why context-aware programming matters (Section~\ref{sec:intro-whymatters}),
+we covered a number of concrete problems that developers are facing today. Each of those would
+presents an important further work. The most important practical challenges are cross-compilation 
+(often solved using complex $\prepk{\#if}$ pre-processor rules) and tracking of provenance
+and security properties.
+
+% =================================================================================================                                                                     
+%
+% 	  ###                         ##                    #                        
+% 	 #   #                         #                                             
+% 	 #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###  
+% 	 #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #     
+% 	 #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###  
+% 	 #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      # 
+% 	  ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####  
+%
+% =================================================================================================
+
 \section{Summary}
+\label{sec:conc-conclusions}
+
+We believe that understanding what programs \emph{require} from the world is equally important as 
+how programs \emph{affect} the world.
+
+The latter has been uniformly captured by effect systems and monads. Those provide not just 
+technical tools for defining semantics and designing type systems, but they also \emph{shape}
+our thinking -- they let us view seemingly unrelated programming language features (exceptions,
+state, I/O) as instances of the same concept and thus reduce the number of distinct language
+features that developers need to understand.
+
+This thesis aims to provide a similar unifying theory and tools for capturing context-dependence
+in programming languages. We showed that programming language features (liveness, data-flow, 
+implicit parameters, etc.) that were previously treated separately can be captured by a common 
+framework developed in this thesis. The main technical contribution of this thesis is that it  
+provides the necessary tools for programming language designers -- including parameterized type 
+systems, categorical semantics based on indexed comonads and equational theory.
+
+If there is a one thing that the reader should remember from this thesis, it is the fact that
+there is a unified notion of \emph{context}, capturing many common scenarios in programming.

### 2014-12-16 text/Chapter02.tex
Spell checking

@@
 computations. In a pioneering work, Moggi \cite{monad-notions} showed that effects (including
-partiality, exceptions, non-determinism and I/O) can be modelled uisng the category theoretic
+partiality, exceptions, non-determinism and I/O) can be modelled using the category theoretic
 notion of \emph{monad}.
@@
 
-In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variales
+In Contextual Modal Type Theory, the modality $\square$ is further annotated with the free variables
 of the (unevaluated) expression. We write $\square^{\cclrd{\Psi}} \tau$ for a type of expressions
@@
 
-The first pathway follows as a dualization of well-known effect systems. However, this is not simply 
+The first pathway follows as a dualisation of well-known effect systems. However, this is not simply 
 a syntactic transformation. As we further discuss in the next chapter, coeffect systems treat lambda 

### 2014-12-16 text/Chapter03.tex
Spell checking

@@
 \item We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
-  explains what is a \emph{coeffect} and constrasts it with a better known notion of 
+  explains what is a \emph{coeffect} and contrasts it with a better known notion of 
   \emph{effect}. It explains what is the nature of properties that can be tracked using 
-  coeffect systems presented in this theis.
+  coeffect systems presented in this thesis.
 
@@
 \item In addition, the coeffect systems for tracking the number of accessed past values in 
-  data-flow languagges (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
+  data-flow languages (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
   presents novel results and can be used to optimize data-flow programs.
@@
 capture some \emph{output impurity}. In contrast, coeffect systems track what a program requires from 
-the execution environment, or \emph{input impurity}.
+the execution envionment, or \emph{input impurity}.
 
@@
 on the right. In both effect systems and coeffect systems, we \emph{write} the annotation as part of
-the function arrow. However, in the underlyng categorical model, effects are attached to the result
+the function arrow. However, in the underlying categorical model, effects are attached to the result
 $\tau$, while coeffects are attached to the context $\tau_1 \times \ldots \times \tau_n$.
@@
 
-We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
+We note thatt an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
 but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
@@
 
-When using the usualmonadic semantics based on the reader monad, the semantics of the (\emph{abs})
+When using the usual monadic semantics based on the reader monad, the semantics of the (\emph{abs})
 rule would be modified as follows:
@@
 
-%---------------------------------------------------------------------------------------------------
+%----------------------------------------------------------------------------------------------------
 
@@
 tainting, we annotate variables with $\cclrd{\ident{T}}$ for \emph{tainted} and with 
-$\cclrd{\ident{U}}$ for \emph{untainted}. Accessing a variable marks it as untained,
+$\cclrd{\ident{U}}$ for \emph{untainted}. Accessing a variable marks it as untainted,
 but using an expression that depends on some variable in certain dangerous contexts -- such 
@@
 compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
-who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
+who developed an effect system (as discussed in Section~\ref{sec:paath-sem-effects}) that uses
 lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.
@@
 principled approaches based on comonads and modal S4, ad-hoc type system extensions and static analyses 
-as well as approaches based on monads. We looked at a number of applications inclding Haskell's implicit 
+as well as approaches based on monads. We looked at a number of applications including Haskell's implicit 
 parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of 

### 2014-12-16 text/Chapter04.tex
Spell checking

@@
 in programming languages. The syntax and type system of the flat coeffect calculus follows 
-asimilar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
+a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
 in the structure, as explained in the previous chapter, most importantly in lambda abstraction
@@
 \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
-defining the semantics in terms of comonadic computations is not a simple mechanical dualization
+defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
 of the work on effect systems and monads.
@@
 sets of implicit parameters, versions represented as integers or other values. The expressions of 
-the calculus are those of the $\lambda$-calculus with \emph{let} binding; assuming $\alpha$ ranges 
+the calculus are those of the $\lambda$-calculus wwith \emph{let} binding; assuming $\alpha$ ranges 
 over base types, the types of the calculus are defined as:
@@
 
-Because of the assymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
+Because of the asymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
 the duality between monads and comonads can no longer help us with defining the additional structure
@@
 The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
-values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combineds them into a 
+values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a 
 single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
@@
 that they can be discharged by matching constructs providing the required context. As discussed
-tfearlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
+earlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
 this form of evaluation (rather than looking at semantics for concrete systems).
@@
 In a flat coeffect system with the (\emph{approxi\-mation}) property, 
-if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+if $\ctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
 and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
@@
 Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
-actually access a previous value of the argument $x$. This cannot be caputred by a flat coeffect 
+actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
 system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
@@
 \label{thm:cbn-substitution-top}
-In a top-pointed flat coeffect calculus with an aglebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
 substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
@@
 of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
-As shown in Figure~\ref{fig:flat-semantics-sub}, the seamntics of the sub-typing rule (\emph{typ}) 
+As shown in Figure~\ref{fig:flat-semantics-sub}, the semantics of the sub-typing rule (\emph{typ}) 
 then just composes the semantics of the original expression with the conversion produced by the 
@@
 The following derives the conclusion of (\emph{abs}) using (\emph{idabs}), sub-coeffecting, 
-sub-typing and the fact that the algebra is \emph{stricly oriented}:
+sub-typing and the fact that the algebra is \emph{strictly oriented}:
 
@@
 Follows from the fact that $(\C, \cpar, \czero)$ is a monoid and \ident{assoc}, $\pi_1$ and
-$\pi_2$ are pure functions (treated as contstants in the langauge).
+$\pi_2$ are pure functions (treated as constants in the language).
 \end{proof}
@@
 requirements between declaration site and call site (or, semantically, combining the implicit 
-parameters available in the scope where the function is deined and those specified by the caller).
+parameters available in the scope where the function is defined and those specified by the caller).
 
@@
  \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
- \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
+ \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{eader monad})
 \end{array} 
@@
 are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
-reader monad via currying/uncurrying isomoprhism.
+reader monad via currying/uncurrying isomorphism.
 \end{remark}
@@
 with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
-function, \emph{before} the function actually perforsm the effectful operation with the effects.
+function, \emph{before} the function actually performs the effectful operation with the effects.
 The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
@@
 \emph{whole-context} properties of computations, that is properties related to the 
-execution environment or the enitre context in which programs are executed.
+execution environment or the entire context in which programs are executed.
 This is the first of the three \emph{coeffect calculi} developed in this thesis.

### 2014-12-16 text/Chapter05.tex
Spell checking

@@
 
-In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware prorgamming
+In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware programming
 languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
@@
 the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
-the meta-langauge style, which extends the terms and types of a language with constructs for explicitly
+the meta-language style, which extends the terms and types of a language with constructs for explicitly
 manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
@@
 (Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
-for all our examples; it is needed for the structural version of the subtitution lemma 
+for all our examples; it is needed for the structural version of the substitution lemma 
 (Lemma~\ref{thm:structural-substitution}).
@@
 does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
-perform corresponding transformation on the coeffect annoation.
+perform corresponding transformation on the coeffect annotation.
 
@@
 variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
-allows re-using a variable. We can type check sub-expressions using two separate varaibles and 
+allows re-using a variable. We can type check sub-expressions using two separate variables and 
 then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
-is the only place in the structural coeffect system where context requiremens are combined
+is the only place in the structural coeffect system where context requirements are combined
 (semantically, this is where the available context is shared).
@@
 It is important to note that this gives us a systems with \emph{different} properties. The 
-information are now tracked per-variable rather than for enitre contexts. For data-flow,
+information are now tracked per-variable rather than for entire contexts. For data-flow,
 we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
@@
 return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
-operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This assymmetry
+operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
 is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
@@
 
-\subsection{Semantics of structural caluculus}
+\subsection{Semantics of structural calculus}
 
@@
 \paragraph{Expressions.}
-A semantics of strucutural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
+A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
 the previous chapter, the semantics is written in a programming language style using constructs
@@
 the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
-uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a varaible while 
+uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a variable while 
 splitting its additional context.
@@
 annotated with a unique operations provided by the coeffect annotations. This means that the
-semantics fully determines the structure of the typing rules of the strucutral coeffect
+semantics fully determines the structure of the typing rules of the structural coeffect
 calculus.
@@
 
-The $\ident{dup}_{m,n}$ operation needs to produce a two-varaible context containing $m$ and $n$ values,
+The $\ident{dup}_{m,n}$ operation needs to produce a two-variable context containing $m$ and $n$ values,
 respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
@@
 strong properties are desirable for programming languages, but are often not satisfied 
-(\eg~by langauges with effects).
+(\eg~by languages with effects).
 

### 2014-12-16 text/Chapter06.tex
Spell checking

@@
 lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise 
-as some containers may have multiple shapes of the same length, \eg~differently ballanced trees.}.
+as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
 For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
@@
 
-\item The contianer representing the singleton data type is given by $S \triangleleft P$ where
+\item The container representing the singleton data type is given by $S \triangleleft P$ where
   shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
@@
 As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
-an underlying category structre; $\cunit$ and $\czero$ represent an accessed and unused variable, 
+an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable, 
 respectively.
@@
 The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable 
-context, respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts.
+context, respectively. The $\stimes$ operation corresponds to concatenation of free-variable contexts.
 Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
@@
 use of variables -- the associativity is built-in. The order of variables matters, but can be changed 
-using the exchane rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ 
+using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ 
 are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
@@
 In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
-system is that the vector concatentation $\atimes$ is now replaced with context splitting/merging 
+system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging 
 of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
@@
 and contraction rules. Aside from idempotence, we could use the unit property and obtain 
-\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this appraoch does not 
+\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this approach does not 
 work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
@@
 
-\subsection{Semanitcs of unified calculus}
+\subsection{Semantics of unified calculus}
 \label{sec:unified-semantics}
@@
 For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
-splitting and then merging a context may lose some of the avilable contextual capabilities
+splitting and then merging a context may lose some of the available contextual capabilities
 (and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:unified-rev-contr} 
@@
 techniques that automatically insert the necessary monadic plumbing (without
-special syntax). This has been done for effecful computations \cite{monads-lightweight-ml}
+special syntax). This has been done for effectful computations \cite{monads-lightweight-ml}
 and a similar approach would be worth exploring for coeffects.
@@
 language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
-The structures required in this thesis include sets (with union or interesection), natural numbers (with 
+The structures required in this thesis include sets (with union or intersection), natural numbers (with 
 addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
@@
 of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting 
-properties, but it makes working with coeffect annotations difficult -- it compilcates tasks
+properties, but it makes working with coeffect annotations difficult -- it complicates tasks
 such as type generalization and constraint solving in type inference. Here, we look at one possible 
@@
 Note that we still need \emph{two distinct} operators in the system. Analysis of our examples shows
-that the operator used for sequential composition $\cseq$ is not necessarilly related in any way to 
+that the operator used for sequential composition $\cseq$ is not necessarily related in any way to 
 the ordering on the lattice provided by $\cleq$.
@@
 \begin{figure}[t]
-{\small a.) Littice-based flat systems (selected rules):}
+{\small a.) Lattice-based flat systems (selected rules):}
 
@@
 \vspace{1.5em}
-{\small b.) Littice-based structural systems (selected rules):}
+{\small b.) Lattice-based structural systems (selected rules):}
 
@@
 
-In smmary, the lattice-based coeffect system has some advantages over the design used in this
+In summary, the lattice-based coeffect system has some advantages over the design used in this
 thesis. In particular, it uses better-known structures and is simpler. However, we do not use
@@
 
-The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
+The coeffect meta-language could be designed using both flat and structural indexed comonads.
 For simplicity, this section only discusses the flat variant. The syntax of types and terms
-of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
+of the language includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
 language constructs:
@@
 The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
-operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
+operation of the comonad. To define meta-language for flat indexed comonads, we also include 
 \kvd{split} and \kvd{merge} that embed the corresponding operations.
@@
 Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
-we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
 meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
@@
 
-In the rest of the chapter, we discussed two topis of future and related work. First
+In the rest of the chapter, we discussed two topics of future and related work. First
 (Section~\ref{sec:unified-impl}), we considered pathways to practical implementations of

### 2014-12-16 text/Chapter07.tex
Spell checking

@@
 Coeffect annotations in context-aware programming languages can contain significant amount
-of information. Thus, the programming langauge also needs a form of type inference that can propagate
+of information. Thus, the programming language also needs a form of type inference that can propagate
 such information. This has been done for effect systems \cite{effects-polymorphic}. For 
@@
 
-In the \emph{flat} and \emph{sturctural} coeffect calculi, we attach annotations to the \emph{entire
+In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
 context} and to \emph{individual variables}, respectively. We later unified the two systems 
@@
 \paragraph{Applications.}
-The last direction for further development is developing practical programming langauges or
+The last direction for further development is developing practical programming languages or
 libraries based on the theory presented in this thesis. We believe that the most useful approach

### 2014-12-16 text/Chapter03.tex
Undo evil spell checker breaking changes

@@
 
-We note thatt an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
+We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
 but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
@@
 compiled to assembly language. They build on the work of Tofte and Talpin \cite{app-region-memory}
-who developed an effect system (as discussed in Section~\ref{sec:paath-sem-effects}) that uses
+who developed an effect system (as discussed in Section~\ref{sec:path-sem-effects}) that uses
 lexically scoped \emph{memory regions} to provide an efficient and controlled memory management.

### 2014-12-16 text/Chapter04.tex
Undo evil spell checker breaking changes

@@
 sets of implicit parameters, versions represented as integers or other values. The expressions of 
-the calculus are those of the $\lambda$-calculus wwith \emph{let} binding; assuming $\alpha$ ranges 
+the calculus are those of the $\lambda$-calculus with \emph{let} binding; assuming $\alpha$ ranges 
 over base types, the types of the calculus are defined as:
@@
 In a flat coeffect system with the (\emph{approxi\-mation}) property, 
-if $\ctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
 and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
@@
  \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
- \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{eader monad})
+ \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
 \end{array} 

### 2014-12-16 text/AppendixA.tex

@@
 
-This appendix gives full typing derivations for the examplespresented in Chapter~\ref{ch:applications},
+This appendix gives full typing derivations for the examples presented in Chapter~\ref{ch:applications},
 which demonstrate simple coeffect calculi for implicit parameters, liveness and data-flow.

### 2014-12-16 main.tex
Finished pathways, copy & past chapters 3 and 4.
Refactor rules
Merge changes
Adding popular introduction, title and new abstract
Editing introduction chapter
Add dataflow example to introduction
Finish introduction chapter and restructure the rest
Early work on chapter 3
Chapter 3: Implicit parameters and distributed
Flat coeffect systems in Chapter 03
Working on structural liveness (Chapter 3)
Semantics of structural liveness (Chapter 3)
Tweak style and finish Chapter 3
Start working on Chapter 4
Categorical semantics of flat calculus (Chapter 4)
Syntactic properties (Chapter 4)
Call-by-valu

@@
-%----------------------------------------------------------------------------------------
+%%----------------------------------------------------------------------------------------
 %	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
@@
 \usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{catchfilebetweentags}
 
-\newtheorem{definition}{Definition}
+\newcounter{mythmcounter}
+\newcounter{mydefcounter}
+\newcounter{myexpcounter}
+\newtheorem{example}[myexpcounter]{Example}
+\newtheorem{remark}[mythmcounter]{Remark}
+\newtheorem{definition}[mydefcounter]{Definition}
+\newtheorem{theorem}[mythmcounter]{Theorem}
+\newtheorem{lemma}[mythmcounter]{Lemma}
+\newtheorem{proposition}[mythmcounter]{Proposition}
+\newtheorem*{lemma*}{Lemma}
 
-% source code & types
-\newcommand{\sep}[0]{\;\; | \;\;}
-\newcommand{\kvd}[1]{\textnormal{\bfseries\sffamily #1}} %
-\newcommand{\str}[1]{\textnormal{\ttfamily #1}} %
+% Formatting for source code & types
+\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
+\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
+\definecolor{strclr}{rgb}{0.5,0.1,0.0}
+\definecolor{prepclr}{rgb}{0.0,0.0,0.0}
+
+\newcommand{\sep}[0]{\; | \;}
+\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
+\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
+\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
+\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
 \newcommand{\ident}[1]{\textnormal{\sffamily #1}}
+\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
+
 \newcommand{\ctyp}[2]{C^{#1}#2}
@@
 \newcommand{\subst}[3]{#1[#2 \leftarrow #3]}
+
+% Typing rule and typing statement (single line)
+\newcommand{\tyrule}[3]{ \inference[\footnotesize{(\emph{#1})}~~]{#2}{#3} }
+\newcommand{\tyruler}[3]{ \inference{#2}{#3}~~\footnotesize{(\emph{#1})} }
+
+% Coeffect algebra
+\definecolor{aclr}{rgb}{0.6,0.3,0.0}
+\newcommand{\aclrd}[1]{\textcolor{aclr}{#1}}
+
+% Coeffect algebra
+\definecolor{cclr}{rgb}{0.0,0.5,0.0}
+\newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}
+
+\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
+\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
+\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }} 
+\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
+\newcommand{\czip}{ \textcolor{cclr}{\wedge} } % only used by flat coeffects
+\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
+\newcommand{\cgeq}{ \textcolor{cclr}{\geq} }
+
+\newcommand{\atimes}{ \textcolor{aclr}{\times} }
+\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq 
+\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq
+
+% Unified
+
+\newcommand{\SH}{ {\textcolor{sclr}{S}} }
+\newcommand{\SP}{ {\textcolor{sclr}{P}} }
+\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
+\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
+\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
+\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
+\newcommand{\sflat}{\star}
+
+\newcommand{\azero}{ \textcolor{aclr}{\bot} }
+\newcommand{\aunit}{ \alift{\cunit} }
+\newcommand{\aweak}{ \alift{\czero} }
+
+\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
+\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
+
+\newcommand{\xtimes}{\hat{\times}}
+
+\newcommand{\ctx}{\textit{ctx}}
+
+% Ordinary context where colouring is done by hand 
+\newcommand{\coctx}[2]{ #1 \,\text{\scriptsize @}\, {{\text{$#2$}}}  }
+
+% Simple coeffect systems in the introduction
+\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}
+
+\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}}
+
+\definecolor{sclr}{rgb}{0.2,0,0.8}
+\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }
+
+\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
+\newcommand{\M}{ {\textcolor{cclr}{\mathcal{M}}} }
+
+\newcommand{\alen}[1]{{\aclrd{\textit{len}(}#1\aclrd{)}}}
+\newcommand{\slen}[1]{{\sclrd{\textit{len}(}#1\sclrd{)}}}
+\newcommand{\xlen}[1]{{\textit{len}(#1)}}
+
+\newcommand{\reduce}{\longrightarrow}
+
+\newcommand{\restr}[2]{#1|_{#2}}
+\newcommand{\narrow}[1]{\hspace{-0.5em} #1 \hspace{-0.65em}~}
+
+\definecolor{todoclr}{rgb}{0.8,0,0.5}
+\newcommand{\todo}[1]{\textcolor{todoclr}{ \begin{quotation} \noindent\textbf{TODO:} #1 \end{quotation} }}
 
@@
 
-\newcommand{\codesnip}[0]{\renewcommand{\baselinestretch}{0.9}\normalsize}
-\newcommand{\codesnipout}[0]{\renewcommand{\baselinestretch}{1.0}\normalsize}
-
+\newcommand{\figcaption}[1]{
+  \vspace{0.5em}
+  \noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}} 
+  \vspace{-2em}
+  \caption{#1}
+}
 % Half line spacing around align
@@
 
-% \include{FrontBackMatter/Titlepage} % Main title page
+\include{text/Titlepage} % Main title page
 
@@
 
-% \cleardoublepage\include{FrontBackMatter/Abstract} % Abstract page
+\cleardoublepage\include{text/Declaration} % Publications from the thesis page
 
-% \cleardoublepage\include{FrontBackMatter/Publication} % Publications from the thesis page
+\cleardoublepage\include{text/Abstract} % Abstract page
 
@@
 
-\include{text/Chapter01} % Chapter 1
-\include{text/Chapter02} % Chapter 2
-
-% \include{Chapters/Chapter03} % Chapter 3
-%\include{Chapters/Chapter04} % Chapter 4 - empty template
+\include{text/Chapter01}
+\include{text/Chapter02}
+\include{text/Chapter03}
+\include{text/Chapter04}
+\include{text/Chapter05}
+\include{text/Chapter06}
+\include{text/Chapter07}
 
@@
 
-% \appendix
-
 % \part{Appendix} % New part of the thesis for the appendix
 
-% \include{Chapters/Chapter0A} % Appendix A
-%\include{Chapters/Chapter0B} % Appendix B - empty template
+% \include{Chapters/Chapter0B} % Appendix B - empty template
 
@@
 
+\appendix
+\include{text/AppendixA}
+\include{text/AppendixB}
+
 \end{document}

### 2015-06-16 response.tex
Add first response draft

% ==================================================================================================
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
% ==================================================================================================

\documentclass[
		twoside,openright,titlepage,numbers=noenddot,headinclude,%1headlines,
                footinclude=true,cleardoublepage=empty,
                BCOR=10mm,paper=a4,fontsize=10pt, % Binding correction, paper type and font size
                ngerman,american, % Languages
                ]{scrreprt}

\input{classicthesis-config}
\usepackage{semantic}
\usepackage{paralist}
\usepackage{etoolbox}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\newcounter{mythmcounter}
\newcounter{mydefcounter}
\newcounter{myexpcounter}
\newtheorem{example}[myexpcounter]{Example}
\newtheorem{remark}[mythmcounter]{Remark}
\newtheorem{definition}[mydefcounter]{Definition}
\newtheorem{theorem}[mythmcounter]{Theorem}
\newtheorem{lemma}[mythmcounter]{Lemma}
\newtheorem{proposition}[mythmcounter]{Proposition}
\newtheorem*{lemma*}{Lemma}

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\sep}[0]{\; | \;}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

\newcommand{\ctyp}[2]{C^{#1}#2}
\newcommand{\mtyp}[2]{M^{#1}#2}
\newcommand{\subst}[3]{#1[#2 \leftarrow #3]}

% Typing rule and typing statement (single line)
\newcommand{\tyrule}[3]{ \inference[\footnotesize{(\emph{#1})}~~]{#2}{#3} }
\newcommand{\tyruler}[3]{ \inference{#2}{#3}~~\footnotesize{(\emph{#1})} }

% Coeffect algebra
\definecolor{aclr}{rgb}{0.6,0.3,0.0}
\newcommand{\aclrd}[1]{\textcolor{aclr}{#1}}

% Coeffect algebra
\definecolor{cclr}{rgb}{0.0,0.5,0.0}
\newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}

\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }}
\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
\newcommand{\czip}{ \textcolor{cclr}{\wedge} } % only used by flat coeffects
\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
\newcommand{\cgeq}{ \textcolor{cclr}{\geq} }

\newcommand{\atimes}{ \textcolor{aclr}{\times} }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq

% Unified

\newcommand{\SH}{ {\textcolor{sclr}{S}} }
\newcommand{\SP}{ {\textcolor{sclr}{P}} }
\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
\newcommand{\sflat}{\star}

\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} }
\newcommand{\aweak}{ \alift{\czero} }

\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }

\newcommand{\xtimes}{\hat{\times}}

\newcommand{\ctx}{\textit{ctx}}

% Ordinary context where colouring is done by hand
\newcommand{\coctx}[2]{ #1 \,\text{\scriptsize @}\, {{\text{$#2$}}}  }

% Simple coeffect systems in the introduction
\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}

\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}}

\definecolor{sclr}{rgb}{0.2,0,0.8}
\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }

\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
\newcommand{\M}{ {\textcolor{cclr}{\mathcal{M}}} }

\newcommand{\alen}[1]{{\aclrd{\textit{len}(}#1\aclrd{)}}}
\newcommand{\slen}[1]{{\sclrd{\textit{len}(}#1\sclrd{)}}}
\newcommand{\xlen}[1]{{\textit{len}(#1)}}

\newcommand{\reduce}{\longrightarrow}

\newcommand{\restr}[2]{#1|_{#2}}
\newcommand{\narrow}[1]{\hspace{-0.5em} #1 \hspace{-0.65em}~}

\definecolor{todoclr}{rgb}{0.8,0,0.5}
\newcommand{\todo}[1]{\textcolor{todoclr}{ \begin{quotation} \noindent\textbf{TODO:} #1 \end{quotation} }}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\catc}{\mathcal{C}}
\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }

\newcommand{\figcaption}[1]{
  \vspace{0.5em}
  \noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
  \vspace{-2em}
  \caption{#1}
}
% Half line spacing around align
\newcommand{\zerodisplayskips}{%
  \setlength{\abovedisplayskip}{0.5em}
  \setlength{\belowdisplayskip}{0.5em}
  \setlength{\abovedisplayshortskip}{0.5em}
  \setlength{\belowdisplayshortskip}{0.5em}}
\appto{\normalsize}{\zerodisplayskips}
\appto{\small}{\zerodisplayskips}
\appto{\footnotesize}{\zerodisplayskips}

% half line spacing around compactitem
\setlength{\pltopsep}{0.5em}
\setlength{\plitemsep}{0.3em}

\patchcmd{\thebibliography}{\chapter*}{\section*}{}{}

\begin{document}

\frenchspacing % Reduces space after periods to make text more compact
\raggedbottom % Makes all pages the height of the text on that page
\selectlanguage{american} % Select your default language - e.g. american or ngerman
\pagenumbering{roman} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
\pagestyle{plain} % Suppress headers for the pre-content pages

% ==================================================================================================

\chapter*{Proposal for thesis corrections}
\renewcommand*\thesection{\arabic{section}}

This document addresses some of the concerns discussed in the joint report on a viva of the thesis
``Context-aware programming languages'' and to proposes an outline for corrections that would
resolve the issues discussed in the report.

The main concern presented in the report is related to the use of ``coeffect-directed semantics''
in the thesis. We start by discussing additional background that provides a better perspective on
this approach as well as why it was employed in the thesis (Section~\ref{sec:background}). Next,
we turn our attention to the issue of \emph{implicit parameters} (Section~\ref{sec:ambiguity}) --
this problem (one of the three problems used throughout the thesis) demonstrates the ambiguity
of type checking that arises in \emph{flat coeffect calculus} and relates it with prior work in
languages such as Haskell. Finally, the last section (Section~\ref{sec:proposal}) outlines four
concrete steps that address the reviewer's concerns.

\section{Background}
\label{sec:background}

Perhaps the most important objection raised in the report is that the semantics presented throughout
the thesis is given based on a concrete typing derivation. In other words, the meaning of a term
depends on the concrete typing assigned by a compiler\footnote{The thesis does not clearly and
explicitly state this, leading to potential confusion -- this can be clarified, both in the
text and typographically (by writing the semantics over a derivation rule, rather than over
a judgement) to avoid confusion.} In this section, we discuss additional programming language
background\footnote{Which will be added in the final version of the thesis.} that provides
additional grounds for specifying the meaning of programs in this way.

\subsection{Type-directed semantics in practice}
\label{sec:background-practice}

One of the aims of this thesis is to model how real-world functional-first programming languages
could be extended with the support for context-aware programming and so the work is done in
the context of languages such as F\# \cite{app-fsharp-spec} and Scala \cite{app-scala-spec}.

In theoretically-focused programming languages, it is possible to define the meaning of a term
without considering the typing, but there are many examples where this is not the case in languages
such as F\#:
%
\begin{equation*}
\begin{array}{l}
  \kvd{let}~\ident{nextTwo}~\ident{rnd}~=~(\ident{rnd.Next}(), \ident{rnd.Next}()) \\
  \kvd{let}~\ident{add}~\ident{a}~\ident{b}~=~\ident{a}+\ident{b}
\end{array}
\end{equation*}
%
In the first example, the meaning of $\ident{rnd.Next}()$ depends on the type of \ident{rnd}.
This can be any .NET class with the \ident{Next} method, but the F\# compiler can only compile
the code if it knows the concrete type. Thus, the meaning depends the typing derivation.

In the second example, the meaning of $+$ depends on the types of the operands (it can be an
addition or string concatenation). The F\# compiler can compile the function
without additional information, because the type inference has a defaulting mechanism (using
\ident{int}), but if the program contains e.g.~$\ident{add}~\str{hi}~\str{there}$, the compiler
will use \ident{string}.

As demonstrated in this section, it is not uncommon in practical programming languages that the
meaning of a term depends on its typing derivation. In F\#, this is addressed in two ways. In case
of \ident{nextTwo}, the compiler rejects the program if it does not know the type of \ident{rnd}
(an explicit type annotation would be required to disambiguate). In case of \ident{add}, the
compiler uses a defaulting mechanism to pick one of the avialable typing derivations. We return to
these two options when discussing the resolution of ambiguities in \emph{coeffect calculi} in
Section~\ref{sec:ambiguity}.

\subsection{Type-directed semantics in theory}

In a number of places in the thesis, we give the semantics of based on typing derivation (both for
the concrete languages in Chapter~3 and for the coeffect calculi in Chapters~4 and 5). The notation
used in the thesis is (using simple let-binding as an example):
%
\begin{equation*}
\begin{array}{l}
  \sem{x:\tau \vdash \kvd{let}~x_1=e_2~\kvd{in}~e_2 : \tau_2} = \\
  \quad \sem{x_1:\tau_1 \vdash e_2 : \tau_2} \circ \sem{x:\tau \vdash e_1 : \tau_1}
\end{array}
\end{equation*}
%
This notation is somewhat ambiguous as it suggests that we are defining a function from a typing
judgement for the expression on the left-hand side. However, this should be read as giving an
interpretation of the term based on the typing derivation.

It is worth noting that giving an interpretation of a term based on its typing derivation is not
an unusual construction used only in the thesis. In fact, Moggi in his seminal paper on monads uses
the same approach and writes \cite[p7]{monad-notions} (emphasis mine):

\begin{quotation}
\noindent
An interpretation $\sem{-}$ of the language in a category $C$ is parametric in an interpretation
of the symbols in the signature and is defined by \emph{induction on the derivation} of
well-formedness for (types,) terms and equations.
\end{quotation}

\noindent
The notation used by Moggi is unambiguous (and we intend to adopt the same clear notation in the
revised version of the thesis). Using the same example as above:
%
\begin{equation*}
  \begin{array}{r}
    x:\tau \vdash e_1 : \tau_1 \\
    x_1:\tau_1 \vdash e_2 : \tau_2 \\ \hline
    x:\tau \vdash \kvd{let}~x_1=e_2~\kvd{in}~e_2 : \tau_2
  \end{array}
  \begin{array}{l}
    = g_1 \\
    = g_2 \\
    = g_2 \circ g_1
  \end{array}
\end{equation*}
%
In systems that have unique typing derivation (such as the language of Moggi), the fact that the
semantics is defined by induction over a typing derivation means merely that we only define
semantics for well-typed prorgams (as there is no other typing derivation). However, there are two
possible directions that can be taken from this simple starting point.

We can extend the type system with non-syntax-driven rules (for example, with sub-effecting or
explicit structural rules like exchange, weakening and contraction). As a result, the language
loses uniqueness of typing, but the semantics of terms is equivalent for all typing derivations.
This is the \emph{coherence property} that has been proved for certain systems
\cite{semantics-bltres}.

However, when we consider programming languages where the meaning of a term \emph{depends} on a
typing derivation (as discussed in Section~\ref{sec:background-practice}), we can no longer
prove the coherence property -- because the semantics of a term is different for different
typing derivations. As further discussed in Section~\ref{sec:ambiguity}, this is exactly the case
of the flat coeffect calculus for implicit parameters presented in the thesis. Before discussing
implicit parameters, the following subsection provides a more general argument for why this is the
correct approach for context-aware computations.

\subsection{Why type-directed semantics for coeffects?}

One aim of the thesis is to make explicit some of the context-dependencies that are often implicit
in today's programming languages. For example, a language today might provide a construct for
accessing a resource provided by the environment such as $\kvd{access}~\ident{Date}$ (typically,
this is just a library call). Operationally, the meaning of the meaning of the term depends on
the context -- if \ident{Date} is available on the client-side, it is accessed from there; if
\ident{Date} is available on the server-side, it is accessed from there instead and if \ident{Date}
is not available, the program raises an exception.

We aim to make the programming language aware of these contextual dependencies through the use
of type-and-coeffect systems. Our calculi lift some of the operational properties into the type
system as coeffects. This makes the existing context-dependencies more visible -- and as a
consequence, the semantics can only be defined for a concrete typing derivation. An alternative
approach is to make the language \emph{more explicit}, but we do not attempt to do this and
instead follow an approach similar to F\#\footnote{This demarcation for the work
in the thesis is not clearly discussed in the text and will be clarified.} as outlined in
Section~\ref{sec:background-practice}.

\section{Resolving ambiguity}
\label{sec:ambiguity}

The ambiguity in typing derivation that is ``a problem'' in the thesis is best demonstrated
using implicit parameters. Here, we look at how Haskell resolves this (and how the same
problem appears elsewhere).

The presented thesis does not attempt to resolve the ambiguity and so it provides a more
interesting insight. We discuss why this is the case.

\subsection{Haskell implicit parameters}

If we try to understand implicit parameters in a more general way, then we can see that the meaning
indeed depends on the typing derivation. (Haskell resolves this by preferring one specific typing derivation.)

Haskell implicit parameters

\subsection{Exceptions}
Exceptions

\subsection{Counter example}

\subsection{Summary}

Haskell does one thing, exceptions do (strictly) another thing.
(Maybe for good reasons, but we should view this more generally)

\section{Proposal}
\label{sec:proposal}

Coeffect-directed meaning makes sense.

The thesis does not use category theory to give a semantics in the traditional sense with
traiditional properties (and this needs to be said clearly)

\subsection{Translation}

The "coeffect-directed semantics" should not be seen as a "semantics of a term", but it should
instead be seen as a translation that - given a typing derivation - produces a new term that
unambiguously defines the meaning. The chapter 3 does so using direct denotational style; the
chapters 4 and 5 show that we can abstract common structures used in these definitions.

Doing so also explains why our type system (about which we also prove syntactic properties) is
appropriately constructed (the structure of the typing follows from the structure of this
coeffect-directed semantics - and BTW this was the hard thing to figure out when working on the thesis!)

As a result, we can see the "coeffect-directed semantics" more as a translation from a type-and-coeffect
derivation to an unambiguous term language (which I'm happy to define more explicitly and precisely
in a syntactic style, as opposed to categorical style).

\subsection{Unique typing derivation}

>> 1. The problem occurs because the type derivation in all the proposed
>> systems is not unique, due to features such as sub-typing

After doing the changes outlined above, this still remains a valid point. In Haskell's implicit
parameters, this is resolved by choosing one canonical typing derivation (I'll give an example how
this works). The same approach can be used to correct this in the thesis - that is, I would like
to give an algorithmic formulation of the type system that produces a canonical typing derivation.
(There are three sources of the problem - subtyping, lambda abstraction and structural system; I
intend to make subtyping an explicit term, add externally provided function for resolving ambiguities
in lambda abstraction and have algorithmic typing rules for the structural systems).

\subsection{Lack of equational structure}

>> 3. The co-monadic structures proposed in the categorical semantics
>> lack an equational structure

This is correct - and it follows from the fact that the categorical structures are used to motivate
the design of the type system, rather than - which is more common - to provide a reasoning
framework about coeffectful programs.

\subsection{Generalization}

>> 4. It is unclear as to whether the concrete denotational semantics
>> are instances of the categorical framework.

This is a valid point and I fully accept it. I intend to add sections showing that this is, indeed, the case.

% ==================================================================================================

\section{Summary}

In practice, people use monads as common structure - not for proving program properties.
This thesis aims to find a similar common structure in context-aware programming.

This is done in two ways pursued in parallel - by finding a type system with desirable syntactic
properties and by finding a way to structure the translational semantics using a common structure.

Finding common structure.

yo

\bibliographystyle{abbrv}
\bibliography{Bibliography}

\end{document}

### 2015-07-06 response.tex
Finish response, add FAKE script

@@
 \selectlanguage{american} % Select your default language - e.g. american or ngerman
-\pagenumbering{roman} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
+\pagenumbering{arabic} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
 \pagestyle{plain} % Suppress headers for the pre-content pages
@@
 
-This document addresses some of the concerns discussed in the joint report on a viva of the thesis
-``Context-aware programming languages'' and to proposes an outline for corrections that would
-resolve the issues discussed in the report.
+The starting point for addressing the reviewers concerns is observing that in many practical
+programming languages, the value of an expression depends on the type derivation
+(Section~\ref{sec:background-practice}). Often, the choice is hidden behind a mechanism that
+selects one type as primary\footnote{An F\# example is discussed in Section~\ref{sec:background-practice}
+and Haskell example is discussed in Section~\ref{sec:ambiguity}.}. This resolves the ambiguity,
+but also narrows down the design space.
 
-The main concern presented in the report is related to the use of ``coeffect-directed semantics''
-in the thesis. We start by discussing additional background that provides a better perspective on
-this approach as well as why it was employed in the thesis (Section~\ref{sec:background}). Next,
-we turn our attention to the issue of \emph{implicit parameters} (Section~\ref{sec:ambiguity}) --
-this problem (one of the three problems used throughout the thesis) demonstrates the ambiguity
-of type checking that arises in \emph{flat coeffect calculus} and relates it with prior work in
-languages such as Haskell. Finally, the last section (Section~\ref{sec:proposal}) outlines four
-concrete steps that address the reviewer's concerns.
+The same is the case for the coeffect calculi presented in the thesis \emph{Context-aware programming
+languages}. However, our approach makes the dependence on the typing explicit and we resolve it by
+defining the semantics as a function of a typing derivation\footnote{The thesis does not explicitly
+state this and it will be clarified in the final version.}. As discussed in Section~\ref{sec:background-theory}
+this approach does not diverge from, but rather extends Moggi's seminal work \cite{monad-notions}.
 
-\section{Background}
-\label{sec:background}
+In the following two sections, we argue that specifying program semantics based on a type derivation
+is not unusual in practice (Section~\ref{sec:background-practice}) and theory
+(Section~\ref{sec:background-theory})\footnote{This background will be added in the final version of the thesis.}.
+Finally, Section~\ref{sec:proposal} concludes with a concrete correction proposal.
 
-Perhaps the most important objection raised in the report is that the semantics presented throughout
-the thesis is given based on a concrete typing derivation. In other words, the meaning of a term
-depends on the concrete typing assigned by a compiler\footnote{The thesis does not clearly and
-explicitly state this, leading to potential confusion -- this can be clarified, both in the
-text and typographically (by writing the semantics over a derivation rule, rather than over
-a judgement) to avoid confusion.} In this section, we discuss additional programming language
-background\footnote{Which will be added in the final version of the thesis.} that provides
-additional grounds for specifying the meaning of programs in this way.
-
-\subsection{Type-directed semantics in practice}
+\section{Type-directed semantics in practice}
 \label{sec:background-practice}
 
-One of the aims of this thesis is to model how real-world functional-first programming languages
-could be extended with the support for context-aware programming and so the work is done in
-the context of languages such as F\# \cite{app-fsharp-spec} and Scala \cite{app-scala-spec}.
-
-In theoretically-focused programming languages, it is possible to define the meaning of a term
-without considering the typing, but there are many examples where this is not the case in languages
-such as F\#:
+The aim of the thesis is to provide foundations for extending practical functional-first
+programming languages such as F\# \cite{app-fsharp-spec} and Scala \cite{app-scala-spec} with
+support for context-aware programming, so we consider two F\# examples. In both, the meaning of
+a term depends on the typing:
 %
@@
 %
-In the first example, the meaning of $\ident{rnd.Next}()$ depends on the type of \ident{rnd}.
-This can be any .NET class with the \ident{Next} method, but the F\# compiler can only compile
-the code if it knows the concrete type. Thus, the meaning depends the typing derivation.
+The two examples demonstrate two ways of resolving ambiguities in typing derivations and
+corresponding semantics:
 
-In the second example, the meaning of $+$ depends on the types of the operands (it can be an
-addition or string concatenation). The F\# compiler can compile the function
-without additional information, because the type inference has a defaulting mechanism (using
-\ident{int}), but if the program contains e.g.~$\ident{add}~\str{hi}~\str{there}$, the compiler
-will use \ident{string}.
+\begin{itemize}
+\item The first example illustrates an ordinary .NET method invocation. In this case,
+  the meaning of $\ident{rnd.Next}()$ depends on the type of \ident{rnd}. This could be any
+  class with the \ident{Next} method. The F\# compiler can emit .NET bytecode only if it knows the
+  concrete type and so the developer has to resolve the ambiguity through a \emph{type annotation}.
 
-As demonstrated in this section, it is not uncommon in practical programming languages that the
-meaning of a term depends on its typing derivation. In F\#, this is addressed in two ways. In case
-of \ident{nextTwo}, the compiler rejects the program if it does not know the type of \ident{rnd}
-(an explicit type annotation would be required to disambiguate). In case of \ident{add}, the
-compiler uses a defaulting mechanism to pick one of the avialable typing derivations. We return to
-these two options when discussing the resolution of ambiguities in \emph{coeffect calculi} in
-Section~\ref{sec:ambiguity}.
+\item Operators are treated differently. The meaning of $+$ depends on the types of the
+  operands\footnote{F\# differs from OCaml, which uses $+$ for integers, $+.$ for floats and
+  $\hat{}$ for string concatenation.}, but F\# can compile it without additional information thanks
+  to a \emph{defaulting mechanism}, which chooses \ident{int} if there are no other constraints.
+  If the program contains $\ident{add}~\str{hi}~\str{there}$, F\# will use string concatenation
+  instead of the default (integer addition).
+\end{itemize}
 
-\subsection{Type-directed semantics in theory}
+\noindent
+In both examples, the meaning of the term depends on a typing derivation, but the compiler includes
+a mechanism for finding a unique typing.
 
-In a number of places in the thesis, we give the semantics of based on typing derivation (both for
-the concrete languages in Chapter~3 and for the coeffect calculi in Chapters~4 and 5). The notation
-used in the thesis is (using simple let-binding as an example):
-%
+\section{Type-directed semantics in theory}
+\label{sec:background-theory}
+
+In the thesis, we give the semantics of terms based on a typing derivation (both for concrete
+languages and for coeffect calculi). Using the notation used in the thesis, the semantics of
+a (plain) let-binding is written as:
+
 \begin{equation*}
 \begin{array}{l}
-  \sem{x:\tau \vdash \kvd{let}~x_1=e_2~\kvd{in}~e_2 : \tau_2} = \\
-  \quad \sem{x_1:\tau_1 \vdash e_2 : \tau_2} \circ \sem{x:\tau \vdash e_1 : \tau_1}
+ \sem{x:\tau \vdash \kvd{let}~x_1=e_2~\kvd{in}~e_2 : \tau_2} = \\ \qquad
+ \sem{x_1:\tau_1 \vdash e_2 : \tau_2} \circ
+ \sem{x:\tau \vdash e_1 : \tau_1}
 \end{array}
@@
 %
-This notation is somewhat ambiguous as it suggests that we are defining a function from a typing
-judgement for the expression on the left-hand side. However, this should be read as giving an
-interpretation of the term based on the typing derivation.
-
-It is worth noting that giving an interpretation of a term based on its typing derivation is not
-an unusual construction used only in the thesis. In fact, Moggi in his seminal paper on monads uses
-the same approach and writes \cite[p7]{monad-notions} (emphasis mine):
+The reader might incorrectly see the equation as defining a function from typing \emph{judgements},
+as opposed to typing \emph{derivations}. The correct reading is the latter. In fact,
+the same approach has been used by Moggi in his seminal paper on monads, who writes
+\cite[p7]{monad-notions} (emphasis mine):
 
@@
 The notation used by Moggi is unambiguous (and we intend to adopt the same clear notation in the
-revised version of the thesis). Using the same example as above:
+revised version of the thesis):
 %
@@
   \begin{array}{r}
-    x:\tau \vdash e_1 : \tau_1 \\
-    x_1:\tau_1 \vdash e_2 : \tau_2 \\ \hline
-    x:\tau \vdash \kvd{let}~x_1=e_2~\kvd{in}~e_2 : \tau_2
+    \sem{x:\tau \vdash e_1 : \tau_1} \\
+    \sem{x_1:\tau_1 \vdash e_2 : \tau_2} \\ \hline
+    \sem{x:\tau \vdash \kvd{let}~x_1=e_2~\kvd{in}~e_2 : \tau_2}
   \end{array}
@@
 %
-In systems that have unique typing derivation (such as the language of Moggi), the fact that the
-semantics is defined by induction over a typing derivation means merely that we only define
-semantics for well-typed prorgams (as there is no other typing derivation). However, there are two
-possible directions that can be taken from this simple starting point.
+In systems that have a unique typing derivation (including the language of Moggi), the fact that
+the semantics is defined by induction \emph{over a typing derivation} means merely that we only
+define semantics for well-typed prorgams. More complex languages with non-syntax-driven rules
+(and no unique typing derivation) have two options:
 
-We can extend the type system with non-syntax-driven rules (for example, with sub-effecting or
-explicit structural rules like exchange, weakening and contraction). As a result, the language
-loses uniqueness of typing, but the semantics of terms is equivalent for all typing derivations.
-This is the \emph{coherence property} that has been proved for certain systems
-\cite{semantics-bltres}.
+\begin{itemize}
+\item \textsc{Coherent semantics.} For certain systems \cite{semantics-bltres}, the semantics of
+  a term is the same regardless of a typing derivation. In that case, we can define the semantics
+  over a typing derivation, but show that all derviations yield the same semantics.
 
-However, when we consider programming languages where the meaning of a term \emph{depends} on a
-typing derivation (as discussed in Section~\ref{sec:background-practice}), we can no longer
-prove the coherence property -- because the semantics of a term is different for different
-typing derivations. As further discussed in Section~\ref{sec:ambiguity}, this is exactly the case
-of the flat coeffect calculus for implicit parameters presented in the thesis. Before discussing
-implicit parameters, the following subsection provides a more general argument for why this is the
-correct approach for context-aware computations.
+\item \textsc{Type-directed semantics.} Systems where the meaning of a term \emph{does depend}
+  on a typing derivation (cf.~Section~\ref{sec:background-practice}) do not have the coherence
+  property and thus require \emph{type-directed} semantics. This is the case for example,
+  for the flat coeffect system for implicit parameters (Section~\ref{sec:ambiguity}).
+\end{itemize}
 
-\subsection{Why type-directed semantics for coeffects?}
-
-One aim of the thesis is to make explicit some of the context-dependencies that are often implicit
-in today's programming languages. For example, a language today might provide a construct for
-accessing a resource provided by the environment such as $\kvd{access}~\ident{Date}$ (typically,
-this is just a library call). Operationally, the meaning of the meaning of the term depends on
-the context -- if \ident{Date} is available on the client-side, it is accessed from there; if
-\ident{Date} is available on the server-side, it is accessed from there instead and if \ident{Date}
-is not available, the program raises an exception.
-
-We aim to make the programming language aware of these contextual dependencies through the use
-of type-and-coeffect systems. Our calculi lift some of the operational properties into the type
-system as coeffects. This makes the existing context-dependencies more visible -- and as a
-consequence, the semantics can only be defined for a concrete typing derivation. An alternative
-approach is to make the language \emph{more explicit}, but we do not attempt to do this and
-instead follow an approach similar to F\#\footnote{This demarcation for the work
-in the thesis is not clearly discussed in the text and will be clarified.} as outlined in
-Section~\ref{sec:background-practice}.
+\noindent
+Contextual dependencies can often be satisfied in multiple ways\footnote{For example, see the
+discussion about distributed programming in Section~3.2.2 of the thesis.}. By capturing the
+contextual properties in the type-and-coeffect system, we are merely making the existing
+run-time ambiguities \emph{more explicit}. Thus the proposal is to focus on systems with
+\emph{coeffect-directed semantics} rather than modifying the calculi that inspired the work
+in the thesis in order to force coherence.
 
@@
 
-The ambiguity in typing derivation that is ``a problem'' in the thesis is best demonstrated
-using implicit parameters. Here, we look at how Haskell resolves this (and how the same
-problem appears elsewhere).
+The case where the semantics of a context-aware program depends on the typing derivation is best
+demonstrated using the system for implicit parameters (Section~3.2.1 of the thesis). Consider the
+following example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{f} = (~\kvd{let}~\ident{?x} = 10~\kvd{in}~
+  \lambda \ident{v} \rightarrow \ident{?x})~\kvd{in}\\
+\kvd{let}~\ident{?x} = 15~\kvd{in}~\ident{f}~0
+\end{array}
+\end{equation*}
+%
+Here, a function value $\lambda \ident{v} \rightarrow \ident{?x}$ is created in a context where the
+implicit parameter $\ident{?x}$ has a value 10. The function is then called in a context where
+$\ident{?x}$ has a value 15. Is the result of the expression $10$ or $15$? The choice of the
+run-time semantics needs to be reflected in the type system.
 
-The presented thesis does not attempt to resolve the ambiguity and so it provides a more
-interesting insight. We discuss why this is the case.
+\paragraph{Resolution in Haskell.}
+In Haskell, the implicit parameters from the \emph{declaration-site} are preferred and so the result
+is 10. Correspondingly, the type system only permits one typing -- with the parameter \ident{?x}
+required on the declaration-side (allowing no rebinding at the call-site). However, if the function
+value was defined in a context not containing \ident{?x}, the value from the call-site would be used.
+Thus, Haskell uses a \emph{defaulting mechanism} for implicit parameters, similar to that of F\#
+operators discussed in Section~\ref{sec:background-practice}.
 
-\subsection{Haskell implicit parameters}
+\paragraph{Resolution for exceptions.}
+In case of implicit parameters, the choice is between \emph{lexical} and \emph{dynamic} scoping.
+The same choice appears for other language constructs, such as exception handling. Exception
+handlers in most modern languages are dynamically scoped, but Beta and original version of Smalltalk
+support lexical scoping \cite{app-exceptions}. A language with checked exception that supports
+both dynamic and lexical exception handlers would expose the same problem as implicit parameters
+in Haskell.
 
-If we try to understand implicit parameters in a more general way, then we can see that the meaning
-indeed depends on the typing derivation. (Haskell resolves this by preferring one specific typing derivation.)
+\paragraph{Resolution in coeffect systems.}
+Coeffect systems presented in the thesis explore the most general approach. For example, the
+calculi for implicit parameters permits both lexical and dynamic scoping without a built-in ad-hoc
+defaulting mechanism.
 
-Haskell implicit parameters
+Our design choice means that semantics of a term depends on the typing. In other words, the
+inherent choice in the semantics is reflected in the fact that a term can have multiple types.
+We argue that a \emph{coeffect-directed semantics} (as used in the thesis) provides more insight
+into the problem of context-aware programming as it lets us explore a range of design
+possibilities.
 
-\subsection{Exceptions}
-Exceptions
-
-\subsection{Counter example}
-
-\subsection{Summary}
-
-Haskell does one thing, exceptions do (strictly) another thing.
-(Maybe for good reasons, but we should view this more generally)
-
-\section{Proposal}
+\section{Correction proposal}
 \label{sec:proposal}
 
-Coeffect-directed meaning makes sense.
+In the preceding sections, we argued that \emph{coeffect-directed semantics} is the right approach
+for discussing context-aware programming languages. This clarifies the confusion arising from the
+fact that coeffect systems (like implicit parameters and others) do not satisfy categorical
+\emph{coherence}. This section outlines four specific improvements to address the reviewers concerns.
 
-The thesis does not use category theory to give a semantics in the traditional sense with
-traiditional properties (and this needs to be said clearly)
+\paragraph{4.1 Type erasure.}
+The coeffect-directed semantics in the thesis is presented in the style of a categorical semantics
+(as it was, indeed, inspired by categorical structures). However, it does not follow the full rigour
+of category theory and so it is better seen as \emph{translational semantics} or as \emph{type
+erasure}. Given a typing derivation we produce a new term that unambiguously defines the meaning of
+the original term (Chapter 3 uses direct denotational style; Chapters 4 and 5 use a common
+abstract structure).
 
-\subsection{Translation}
+The main value of the semantics is guiding the definition of the type system, because the structure
+of the coeffect annotations in the typing rules follows from the structure of this coeffect-directed
+translation.
 
-The "coeffect-directed semantics" should not be seen as a "semantics of a term", but it should
-instead be seen as a translation that - given a typing derivation - produces a new term that
-unambiguously defines the meaning. The chapter 3 does so using direct denotational style; the
-chapters 4 and 5 show that we can abstract common structures used in these definitions.
+In the revised thesis, we intend to make this more explicit and treat the formalism as a
+``coeffect-directed translation'', as well as explicitly define the target term language
+for the translation.
 
-Doing so also explains why our type system (about which we also prove syntactic properties) is
-appropriately constructed (the structure of the typing follows from the structure of this
-coeffect-directed semantics - and BTW this was the hard thing to figure out when working on the thesis!)
+\paragraph{4.2 Unique typing derivation.}
 
-As a result, we can see the "coeffect-directed semantics" more as a translation from a type-and-coeffect
-derivation to an unambiguous term language (which I'm happy to define more explicitly and precisely
-in a syntactic style, as opposed to categorical style).
+As argued above, the fact the that semantics depends on the typing derivation is an important
+aspect of our work. However, the fact that we do not define a unique typing for coeffect
+calculi makes them impractical when compared with e.g. Haskell's implicit parameters.
 
-\subsection{Unique typing derivation}
+In the final version of the thesis, we intend to follow the example of Haskell and include an
+algorithmic formulation of the type system that produces a single default typing
+derivation\footnote{There are three sources of typing ambiguity (and hence semantic ambiguity) --
+subtyping, lambda abstraction and structural rules. We intend to make subtyping explicit in the
+language, add externally provided function for resolving ambiguities in lambda abstraction and
+have algorithmic typing rules instead of contraction, weakening and exchange rules.} and
+resolves the ambiguity.
 
->> 1. The problem occurs because the type derivation in all the proposed
->> systems is not unique, due to features such as sub-typing
+\paragraph{4.3. Equational structure.}
 
-After doing the changes outlined above, this still remains a valid point. In Haskell's implicit
-parameters, this is resolved by choosing one canonical typing derivation (I'll give an example how
-this works). The same approach can be used to correct this in the thesis - that is, I would like
-to give an algorithmic formulation of the type system that produces a canonical typing derivation.
-(There are three sources of the problem - subtyping, lambda abstraction and structural system; I
-intend to make subtyping an explicit term, add externally provided function for resolving ambiguities
-in lambda abstraction and have algorithmic typing rules for the structural systems).
+One specific concern raised in the referees report is the lack of equational structure for the
+categorically-inspired structures. This arises from the fact that our usage of categorical
+structures (indexed comonads) is mainly to inspire the abstractions that can be used to structure
+context-aware computations (as opposed to focusing on using category theory to prove properties
+of context-aware programs).
 
-\subsection{Lack of equational structure}
+We intend to make this more explicit in the final version. As discussed in Section~4.1, the thesis
+presents a coeffect-directed translation to a target language, which contains operations inspired
+by indexed comonads. Explaining this link is important in order to relate coeffect calculi with
+previous work \cite{comonads-notions}, however our structure need not be seen as categorical
+semantics.
 
->> 3. The co-monadic structures proposed in the categorical semantics
->> lack an equational structure
+\paragraph{4.4. Generalization.}
 
-This is correct - and it follows from the fact that the categorical structures are used to motivate
-the design of the type system, rather than - which is more common - to provide a reasoning
-framework about coeffectful programs.
+One of the key claims of the thesis is that the comonad-inspired structure of the target
+language appropriately captures the structure of concrete examples of context-aware computations
+(presented in Chapter 3) and as a result, it provides an appropriate type system.
 
-\subsection{Generalization}
-
->> 4. It is unclear as to whether the concrete denotational semantics
->> are instances of the categorical framework.
-
-This is a valid point and I fully accept it. I intend to add sections showing that this is, indeed, the case.
+This claim is left implicit in the text and we intend to address this in the final version (most
+importantly, by proving that the three concrete examples spelled out in Chapter 3 do, indeed, fit
+the common patterns identified later in Chapters 4 and 5).
 
@@
 
-In practice, people use monads as common structure - not for proving program properties.
-This thesis aims to find a similar common structure in context-aware programming.
+An interesting aspect of many context-aware computations is that they require access to some
+context, but do not explicitly specify where the context comes from. In case of implicit parameters,
+the values can be provided either by the call-site or by the declaration-site.
 
-This is done in two ways pursued in parallel - by finding a type system with desirable syntactic
-properties and by finding a way to structure the translational semantics using a common structure.
+In some systems like Haskell's implicit parameters, this aspect is hidden through a mechanism that
+chooses a \emph{default typing} (and a \emph{default semantics}) and thus hides the ambiguity. Our approach is
+to make the ambiguity explicit in the type system. We follow F\# and accept the fact that this
+makes the semantics of a term dependent on the typing and we extend Moggi's \emph{type-directed
+semantics} into a \emph{coeffect-directed semantics}.
 
-Finding common structure.
+Our use of categorical structures (namely \emph{indexed comonads}) is similar to practical Haskell
+use of monads in that we use them to \emph{capture a common structure}. This is in contrast with
+use of category theory for \emph{proving program properties}. Correspondingly, our focus is on
+finding the right structure and showing that it captures our motivating examples and gives a type
+system with desirable properties.
 
-yo
+This document explains an additional background and motivation that is missing in the submitted
+version of the thesis (Section~\ref{sec:background-practice} and \ref{sec:background-theory}) as
+well as additional clarification showing how the semantics of implicit parameters depends on the
+typing (Section~\ref{sec:ambiguity}). Finally, we list four specific improvements that will be
+made in the final version of the thesis to better reflect our approach and the methodology
+clarified in this document (Section~\ref{sec:proposal}).
 

### 2016-01-27 responses/response.tex
Move responses

--- 2015-07-06 response.tex
@@
 selects one type as primary\footnote{An F\# example is discussed in Section~\ref{sec:background-practice}
-and Haskell example is discussed in Section~\ref{sec:ambiguity}.}. This resolves the ambiguity,
-but also narrows down the design space.
+and Haskell example is discussed in Section~\ref{sec:ambiguity}.}.
 
-The same is the case for the coeffect calculi presented in the thesis \emph{Context-aware programming
-languages}. However, our approach makes the dependence on the typing explicit and we resolve it by
+The same is the case for the coeffect calculi presented in the thesis, but 
+our approach makes the dependence on the typing explicit and we resolve it by
 defining the semantics as a function of a typing derivation\footnote{The thesis does not explicitly
-state this and it will be clarified in the final version.}. As discussed in Section~\ref{sec:background-theory}
-this approach does not diverge from, but rather extends Moggi's seminal work \cite{monad-notions}.
+state this and it will be clarified in the final version.}. 
+this approach does not diverge from, but rather extends Moggi's work \cite{monad-notions}
+(Section~\ref{sec:background-theory}).
 
@@
 \label{sec:background-practice}
+\vspace{-1em}
 
@@
 \begin{array}{l}
-  \kvd{let}~\ident{nextTwo}~\ident{rnd}~=~(\ident{rnd.Next}(), \ident{rnd.Next}()) \\
+  \kvd{let}~\ident{nextTwo}~\ident{rnd}~=~(\ident{rnd.Next}(), \ident{rnd.Next}()) \\[-0.2em]
   \kvd{let}~\ident{add}~\ident{a}~\ident{b}~=~\ident{a}+\ident{b}
@@
 \label{sec:background-theory}
+\vspace{-1em}
 
@@
 a (plain) let-binding is written as:
-
+%
 \begin{equation*}
@@
 
+\vspace{-0.5em}
 \begin{quotation}
@@
 \end{quotation}
+\vspace{-0.5em}
 
@@
 %
-In systems that have a unique typing derivation (including the language of Moggi), the fact that
+In systems that have a unique typing (including Moggi's), the fact that
 the semantics is defined by induction \emph{over a typing derivation} means merely that we only
-define semantics for well-typed prorgams. More complex languages with non-syntax-driven rules
+define semantics for well-typed programs. Languages with non-syntax-driven rules
 (and no unique typing derivation) have two options:
@@
 discussion about distributed programming in Section~3.2.2 of the thesis.}. By capturing the
-contextual properties in the type-and-coeffect system, we are merely making the existing
+contextual properties in the type-and-coeffect system, we are making the existing
 run-time ambiguities \emph{more explicit}. Thus the proposal is to focus on systems with
@@
 \label{sec:ambiguity}
+\vspace{-1em}
 
@@
 \kvd{let}~\ident{f} = (~\kvd{let}~\ident{?x} = 10~\kvd{in}~
-  \lambda \ident{v} \rightarrow \ident{?x})~\kvd{in}\\
+  \lambda \ident{v} \rightarrow \ident{?x})~\kvd{in}\\[-0.2em]
 \kvd{let}~\ident{?x} = 15~\kvd{in}~\ident{f}~0
@@
 handlers in most modern languages are dynamically scoped, but Beta and original version of Smalltalk
-support lexical scoping \cite{app-exceptions}. A language with checked exception that supports
-both dynamic and lexical exception handlers would expose the same problem as implicit parameters
-in Haskell.
+support lexical scoping \cite{app-exceptions}. A language that supports both dynamic and lexical
+exception handlers would expose the same problem.
 
@@
 \label{sec:proposal}
+\vspace{-1em}
 
@@
 
+\vspace{-0.5em}
 \paragraph{4.1 Type erasure.}
@@
 
+\vspace{-0.5em}
 \paragraph{4.2 Unique typing derivation.}
@@
 As argued above, the fact the that semantics depends on the typing derivation is an important
-aspect of our work. However, the fact that we do not define a unique typing for coeffect
-calculi makes them impractical when compared with e.g. Haskell's implicit parameters.
+aspect of our work. This lets us better explore the design space, but it does not guide
+concrete language implementations that need a unique typing (cf. Haskell's implicit parameters).
 
@@
 
+\vspace{-0.5em}
 \paragraph{4.3. Equational structure.}
@@
 
-We intend to make this more explicit in the final version. As discussed in Section~4.1, the thesis
-presents a coeffect-directed translation to a target language, which contains operations inspired
-by indexed comonads. Explaining this link is important in order to relate coeffect calculi with
+Our work should be seen as a coeffect-directed translation to a target language with structures
+inspired by indexed comonads. Preserving this link is important in order to relate coeffect calculi with
 previous work \cite{comonads-notions}, however our structure need not be seen as categorical
@@
 
+\vspace{-0.5em}
 \paragraph{4.4. Generalization.}
@@
 
-This claim is left implicit in the text and we intend to address this in the final version (most
-importantly, by proving that the three concrete examples spelled out in Chapter 3 do, indeed, fit
+This claim is left implicit and we intend to address this in the final version 
+(by proving that the three concrete examples spelled out in Chapter 3 do, indeed, fit
 the common patterns identified later in Chapters 4 and 5).
+
+\vspace{-0.5em}
+\paragraph{4.5. Universal type systems.}
+
+The thesis does not clearly explain the benefit of introducing the universal type systems
+based on the coeffect calculi, as we do not prove the usual type-safety property for the concrete systems.
+This becomes clearer when we consider the system as \emph{type erasure}.
+
+The type systems guarantee that \emph{coeffect indices} will match in the resulting 
+translation. The relationship from indices to values is injective in the general case, but 
+we can also show a number of important properties (which will be clarified in the revised
+version) for the concrete systems, \eg~an upper bound on the number of required
+past values in causal data-flow.
 

### 2016-01-29 translation.tex
Working on translation semantics

% ==================================================================================================
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
% ==================================================================================================

\documentclass[
		twoside,openright,titlepage,numbers=noenddot,headinclude,%1headlines,
                footinclude=true,cleardoublepage=empty,
                BCOR=10mm,paper=a4,fontsize=10pt, % Binding correction, paper type and font size
                ngerman,american, % Languages
                ]{scrreprt}

\input{classicthesis-config}
\usepackage{semantic}
\usepackage{paralist}
\usepackage{etoolbox}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\newcounter{mythmcounter}
\newcounter{mydefcounter}
\newcounter{myexpcounter}
\newtheorem{example}[myexpcounter]{Example}
\newtheorem{remark}[mythmcounter]{Remark}
\newtheorem{definition}[mydefcounter]{Definition}
\newtheorem{theorem}[mythmcounter]{Theorem}
\newtheorem{lemma}[mythmcounter]{Lemma}
\newtheorem{proposition}[mythmcounter]{Proposition}
\newtheorem*{lemma*}{Lemma}

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\sep}[0]{\; | \;}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

\newcommand{\ctyp}[2]{C^{#1}#2}
\newcommand{\mtyp}[2]{M^{#1}#2}
\newcommand{\subst}[3]{#1[#2 \leftarrow #3]}

% Typing rule and typing statement (single line)
\newcommand{\tyrule}[3]{ \inference[\footnotesize{(\emph{#1})}~~]{#2}{#3} }
\newcommand{\tyruler}[3]{ \inference{#2}{#3}~~\footnotesize{(\emph{#1})} }

% Coeffect algebra
\definecolor{aclr}{rgb}{0.6,0.3,0.0}
\newcommand{\aclrd}[1]{\textcolor{aclr}{#1}}

% Coeffect algebra
\definecolor{cclr}{rgb}{0.0,0.5,0.0}
\newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}

\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }}
\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
\newcommand{\czip}{ \textcolor{cclr}{\wedge} } % only used by flat coeffects
\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
\newcommand{\cgeq}{ \textcolor{cclr}{\geq} }

\newcommand{\atimes}{ \textcolor{aclr}{\times} }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq

% Unified

\newcommand{\SH}{ {\textcolor{sclr}{S}} }
\newcommand{\SP}{ {\textcolor{sclr}{P}} }
\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
\newcommand{\sflat}{\star}

\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} }
\newcommand{\aweak}{ \alift{\czero} }

\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }

\newcommand{\xtimes}{\hat{\times}}

\newcommand{\ctx}{\textit{ctx}}

% Ordinary context where colouring is done by hand
\newcommand{\coctx}[2]{ #1 \,\text{\scriptsize @}\, {{\text{$#2$}}}  }

% Simple coeffect systems in the introduction
\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}

\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}}

\definecolor{sclr}{rgb}{0.2,0,0.8}
\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }

\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
\newcommand{\M}{ {\textcolor{cclr}{\mathcal{M}}} }

\newcommand{\alen}[1]{{\aclrd{\textit{len}(}#1\aclrd{)}}}
\newcommand{\slen}[1]{{\sclrd{\textit{len}(}#1\sclrd{)}}}
\newcommand{\xlen}[1]{{\textit{len}(#1)}}

\newcommand{\reduce}{\longrightarrow}

\newcommand{\restr}[2]{#1|_{#2}}
\newcommand{\narrow}[1]{\hspace{-0.5em} #1 \hspace{-0.65em}~}

\definecolor{todoclr}{rgb}{0.8,0,0.5}
\newcommand{\todo}[1]{\textcolor{todoclr}{ \begin{quotation} \noindent\textbf{TODO:} #1 \end{quotation} }}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\catc}{\mathcal{C}}
\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }

\newcommand{\figcaption}[1]{
  \vspace{0.5em}
  \noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
  \vspace{-2em}
  \caption{#1}
}
% Half line spacing around align
\newcommand{\zerodisplayskips}{%
  \setlength{\abovedisplayskip}{0.5em}
  \setlength{\belowdisplayskip}{0.5em}
  \setlength{\abovedisplayshortskip}{0.5em}
  \setlength{\belowdisplayshortskip}{0.5em}}
\appto{\normalsize}{\zerodisplayskips}
\appto{\small}{\zerodisplayskips}
\appto{\footnotesize}{\zerodisplayskips}

% half line spacing around compactitem
\setlength{\pltopsep}{0.5em}
\setlength{\plitemsep}{0.3em}

\patchcmd{\thebibliography}{\chapter*}{\section*}{}{}

\begin{document}

\frenchspacing % Reduces space after periods to make text more compact
\raggedbottom % Makes all pages the height of the text on that page
\selectlanguage{american} % Select your default language - e.g. american or ngerman
\pagenumbering{arabic} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
\pagestyle{plain} % Suppress headers for the pre-content pages

% ==================================================================================================

\chapter*{Translation semantics}
\renewcommand*\thesection{\arabic{section}}

Coeffects let us easily construct context-aware domain-specific languages. The general coeffect
language framework provides a parameterized type system (flat or structural) and a translation
that turns a program written in a context-aware DSL into a program in a simple (non-context-aware)
functional programming langauge with additional comonadically-inspired coeffect-specific primitives.

In this section, we discuss the safety that is guaranteed by the coeffect language framework.
The general coeffect language framework itself does not automatically guarantee safety of a 
concrete context-aware DSL. This depends on the additional coeffect-specific primitives.

However, the framework makes it easier to prove such safety -- we only need to define the
meaning of the comonadically-inspired coeffect-specific data type and its associated operations and 
then show that those do not ``go wrong''. In doing so, we can use the coeffect annotations
provided by the type system. In this section, we:

\begin{itemize}
\item Define the common subset of the target functional programming language. This includes the
  syntax of the language, reduction rules and typing rules, but it does \emph{not} include 
  coeffect-specific definitions. Well-typed programs written using the common subset of the 
  target language do not get stuck (via progress and preservation), but they may reduce to 
  \ident{error}, e.g. when accessing the head of an empty list.

\item We then extend the language with coeffect-specific comonadically-inspired
  data types and primtiives for dataflow and for implicit parameters. We show that the extension
  preserves the progress and preservation properties.

\item Next, we consider only programs in the target language that were produced by a
  translation from the coeffect domain-specific language and we show that such programs not only
  do not get stuck, but they also do not reduce to the \ident{error} value. In other words,
  ``well-typed coeffect programs do not get hungry'' requiring more context than guaranteed
  by the coeffect type system.

\item We show how the approach extends to structural coeffect systems and we argue that our
  proof can be generalized -- rather than reconsidering progress and preservation of the 
  whole language, we can rely just on the correctness of the coeffect-specific comonadically-inspired
  primitives and abstraction mechanism provided by languages such as ML and Haskell.  
\end{itemize}

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\paragraph{Language syntax}  
\begin{equation*}
\begin{array}{rcl}
  v &  = & \ident{error} \sep s \sep n \sep \lambda x.e \sep (v_1, \ldots, v_n) \sep v_1::v_2 \sep []\\
\\[-0.5em]
  e &  = & x \sep \ident{error} \sep s \sep n \sep \pi_i~e \sep (e_1, \ldots, e_n)  \\
    &\sep& e_1~e_2 \sep \lambda x.e \sep \ident{head}~e \sep \ident{tail}~e \sep e_1 :: e_2 \sep [] \\
    &\sep& \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4\\
\\[-0.5em]
  \tau &  = & \ident{str} \sep \ident{num} \sep [\tau] \sep \tau_1 \times \ldots \times \tau_n \sep \tau_1 \rightarrow \tau_2 \\
\\[-0.5em]
  C  & = & (v_1, \ldots, v_{i-1}, \_, e_{i+1}, \ldots e_n) \\
    &\sep& v~\_ \sep \_~e \sep \pi_i~\_ \sep \ident{head}~\_ \sep \ident{tail}~\_ \sep \_::e \sep v::\_  \\
    &\sep& \kvd{if}~\_ = e~\kvd{then}~e_1~\kvd{else}~e_2 \sep \kvd{if}~v = \_~\kvd{then}~e_1~\kvd{else}~e_2

\end{array}
\end{equation*}

~
\paragraph{Reduction rules}
\begin{equation*}
\begin{array}{rll}
  \footnotesize{(\emph{fn})}  & (\lambda x.e)~v \rightarrow e[x\leftarrow v] \\[0.5em]
  \footnotesize{(\emph{prj})} & \pi_i (v_1, \ldots, v_n) \rightarrow v_i\\[0.5em]
  \footnotesize{(\emph{hd})}  & \ident{head}~(v_1::v_2) \rightarrow v_1\\[0.5em]
  \footnotesize{(\emph{tl})}  & \ident{tail}~(v_1::v_2) \rightarrow v_2\\[0.5em]
  \footnotesize{(\emph{hde})}  & \ident{head}~[] \rightarrow \ident{error}\\[0.5em]
  \footnotesize{(\emph{tle})}  & \ident{tail}~[] \rightarrow \ident{error}\\[0.5em]
  \footnotesize{(\emph{err})}  & C[\ident{error}] \rightarrow \ident{error}\\[0.5em]
  \footnotesize{(\emph{ifnt})}  & \kvd{if}~n_1 = n_2~\kvd{then}~e_1~\kvd{else}~e_2 \rightarrow e_1 & (\textnormal{when}~n_1 = n_2)\\[0.5em]
  \footnotesize{(\emph{ifnt})}  & \kvd{if}~n_1 = n_2~\kvd{then}~e_1~\kvd{else}~e_2 \rightarrow e_2 & (\textnormal{when}~n_1 \neq n_2)\\[0.5em]
  \footnotesize{(\emph{ifst})}  & \kvd{if}~s_1 = s_2~\kvd{then}~e_1~\kvd{else}~e_2 \rightarrow e_1 & (\textnormal{when}~s_1 = s_2)\\[0.5em]
  \footnotesize{(\emph{ifst})}  & \kvd{if}~s_1 = s_2~\kvd{then}~e_1~\kvd{else}~e_2 \rightarrow e_2 & (\textnormal{when}~s_1 \neq s_2)\\[0.5em]
  \footnotesize{(\emph{ctx})}  & C[e] \rightarrow C[e']  & (\textnormal{when}~e\rightarrow e')\\[0.5em]
\end{array}
\end{equation*}

\caption{Common syntax and reduction rules of the target langauge}
\label{fig:transl-target}
\end{figure}

% --------------------------------------------------------------------------------------------------

\section{Target langauge}

The target langauge for the translation is a simply typed lambda calculus with strings, numbers,
tuples and lists. For each concrete coeffect domain-specific language, we then add additional
primitives -- a comonadically-inspired data type and its associated primitive operations. In this
section, we define the common parts of the language.

The syntax of the target programming langauge is shown in Figure~\ref{fig:transl-target}. 
The values include primitive strings $s$ and numbers $n$, tuples and list values. The value
\ident{error} represents the result of a failed computation.

The expressions include variables $x$, values lambda abstraction and application, conditionals
(limited to equality check) and operations on tuples and lists. We do not need pattern matching
on lists and recursion (although a realistic programming language would include both). In what
follows, we also use the following syntactic sugar for lists and let binding:
%
\begin{equation*}
\begin{array}{rcl}
  \langle e_1, \ldots, e_n \rangle &=& e_1 :: \ldots :: e_n :: [] \\
  \kvd{let}~x=e_1~\kvd{in}~e_2 &=& (\lambda x.e_2) e_1 \\
\end{array}
\end{equation*}
%
Finally, $C[e]$ defines the context in which sub-expressions are evaluated. Together with the
evaluation rules shown in  Figure~\ref{fig:transl-target}, this captures the standard call-by-name
semantics of the common parts of the target language.

The (\emph{hde}) and (\emph{tle}) reduction rules describe the situation when we attempt to access
the head or the tail of an empty list. In that case, the expression reduces to \ident{error} value.
An \ident{error} value appearing in any context then propagates to the top-level as defined by 
(\emph{err}).

The typing rules for the common expressions of the target langauge are shown in 
Figure~\ref{fig:transl-targetty}. The rules are standard. As usual, the 
\ident{error} term is allowed to have any type (\emph{err}).

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\paragraph{Typing rules}
\begin{equation*}
\begin{array}{ll}
\tyrule{var}
  { x:\tau \in \Gamma }
  { \Gamma \vdash x:\tau }
&
\tyrule{err}
  { ~ }
  { \Gamma \vdash \ident{error}:\tau }
\\[1em]
\tyrule{str}
  { ~ }
  { \Gamma \vdash s:\ident{str} }
&
\tyrule{num}
  { ~ }
  { \Gamma \vdash n:\ident{num} }
\\[1em]
\tyrule{nil}
  { ~ }
  { \Gamma \vdash []:[\tau] }
&
\tyrule{head}
  { \Gamma \vdash e :[\tau] }
  { \Gamma \vdash \ident{head}~e:\tau }
\\[1em]
\tyrule{tail}
  { \Gamma \vdash e :[\tau] }
  { \Gamma \vdash \ident{tail}~e:[\tau] }
\qquad~&
\tyrule{abs}
  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 }
  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
\end{array}
\end{equation*}
\begin{equation*}
\begin{array}{l}
\tyrule{app}
  { \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2  & \Gamma \vdash e_2 : \tau_1 }
  { \Gamma \vdash e_1~e_2 : \tau_2 }
\\[1.25em]
\tyrule{cons}
  { \Gamma \vdash e_1 :\tau & \Gamma \vdash e_2 : [\tau] }
  { \Gamma \vdash e_1\!::\!e_2 : [\tau] }
\\[1.25em]
\tyrule{proj}
  { \Gamma \vdash e : \tau_1 \times \ldots \tau_i \times \ldots \times \tau_n }
  { \Gamma \vdash \pi_i~e:\tau_i }
\\[1.25em]
\tyrule{tup}
  { \forall i\in\{1 \ldots n\}.~ \Gamma \vdash e_i : \tau_i }
  { \Gamma \vdash (e_1, \ldots, e_n) : \tau_1 \times \ldots \times \tau_n }
\\[1.25em]
\tyrule{ifn}
  { \Gamma \vdash e_1 : \ident{num} & \Gamma \vdash e_2 : \ident{num} &
    \Gamma \vdash e_3 : \tau & \Gamma \vdash e_4 : \tau}
  { \Gamma \vdash \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4 : \tau}
\\[1.25em]
\tyrule{ifs}
  { \Gamma \vdash e_1 : \ident{str} & \Gamma \vdash e_2 : \ident{str} &
    \Gamma \vdash e_3 : \tau & \Gamma \vdash e_4 : \tau}
  { \Gamma \vdash \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4 : \tau}
\end{array}
\end{equation*}

\caption{Typing rules for the common syntax of the target language}
\label{fig:transl-targetty}
\end{figure}

% --------------------------------------------------------------------------------------------------

\section{Properties}

The subset of the langauge described so far models a simple ML-like functional programming 
langauge (or, Haskell-like language, if we choose call-by-name evaluation). We discuss the properties
of the language after introducing coeffect-specific primitives. As will be seen, ``well-typed
programs do not get stuck'' in our language, although they may produce the \ident{error} value.
For example, $\ident{head}~42$ is a stuck expression ruled out by the common rules of the type
system, however, $\ident{head}~[]$ is a well typed expression that produces \ident{error}.

% And also syntactic sugar \ident{map}

%    &\sep& \ident{cobind}_{\cclrd{s},\cclrd{r}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~e 
%       \sep \ident{split}_{\cclrd{r},\cclrd{s}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \\

%
% &\sep& \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e 
%   \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~e \\

\section{Dataflow}
Domain specific coeffect annotations are non-negative integers, type is a tuple and new kinds
of values are lists:
\begin{equation*}
\begin{array}{rcl}
  v_c &  = &  \langle v_1, \ldots, v_n \rangle
\end{array}
\end{equation*}
%
With the following typing:
%
\begin{equation*}  
\tyrule{val}
  { \forall v_i . \vdash v_i : \tau }
  { \vdash \langle v_1, \ldots, v_n \rangle : \ctyp{n}{\tau} }
\end{equation*}
%
And domain specific operations are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{0}\langle a_0 \rangle \rightarrow a_0
\\[0.45em]
\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle \rightarrow
\\[-0.25em]
\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
\\[0.45em]
\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) \rightarrow
\\[-0.25em]
\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
\\[0.45em]
\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle  \rightarrow
\\[-0.25em]
\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
\\[0.45em]
\ident{lift}_{n', n} \langle a_0, \ldots, a_{n'} \rangle \rightarrow
\qquad(\hspace{-0.75em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
\quad \langle a_0, \ldots, a_n \rangle
\end{array}
\end{equation*}

We could define those in terms of head/tail functions and then prove that they work (and translated
program does not contain head/tail), but perhaps that's unnecessary overkill. Or perhaps it would
be useful in order to show how we avoid errors... But doing it directly looks just simpler (we
also do not need a ``list'' type).

\begin{theorem}[Type preservation]
  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
\end{theorem}
\begin{proof}
  Rule induction over $\rightarrow$

Simple for normal parts of the language

We have to check this for dataflow reductions  
\end{proof}

\begin{theorem}[Progress]
  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
\end{theorem}
\begin{proof}
  By rule induction over $\vdash$.

Interesting case is $e_1 e_2$ where $e_1 = v$ is one of the primitives. Then we have to go through
them and make sure they can progress.  
\end{proof}

\end{document}

### 2016-02-02 text/Chapter05.tex
Move translation into chapters

@@
-%!TEX root = ../main.tex
-
-\chapter{The structural coeffect calculus} 
-\label{ch:structural}
-
-In this chapter, we take the next step towards a unified theory of context-aware programming
-languages. As already discussed, we attempt to find abstractions that are \emph{general enough} 
-to capture a wide range of useful programming language features, but \emph{specific enough} to 
-let us identify and capture interesting properties of the context.
-
-In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware programming
-languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
-calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
-and we introduce the \emph{structural coeffect calculus}.
-
-The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness 
-and data-flow), but provides relatively imprecise approximation and weak syntactical properties. 
-Some of those examples can be also seen as per-variable properties. For those, structural coeffect
-systems give a more precise information about the context. However, we also look at other applications 
-that arise from the work on sub-structural logics, a pathway to coeffects discussed in 
-Section~\ref{sec:path-logic}.
-
-As we will see, the categorical semantics of the structural coeffect calculus is based on the
-same underlying structure as the semantics of the flat calculus. However, the syntactical properties
-of the calculus are quite different and so it is worth considering it in a separate chapter.
+% ==================================================================================================
+%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
+% ==================================================================================================
+
+\documentclass[
+		twoside,openright,titlepage,numbers=noenddot,headinclude,%1headlines,
+                footinclude=true,cleardoublepage=empty,
+                BCOR=10mm,paper=a4,fontsize=10pt, % Binding correction, paper type and font size
+                ngerman,american, % Languages
+                ]{scrreprt}
+
+\input{classicthesis-config}
+\usepackage{semantic}
+\usepackage{paralist}
+\usepackage{etoolbox}
+\usepackage{stmaryrd}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsthm}
+
+\newcounter{mythmcounter}
+\newcounter{mydefcounter}
+\newcounter{myexpcounter}
+\newtheorem{example}[myexpcounter]{Example}
+\newtheorem{remark}[mythmcounter]{Remark}
+\newtheorem{definition}[mydefcounter]{Definition}
+\newtheorem{theorem}[mythmcounter]{Theorem}
+\newtheorem{lemma}[mythmcounter]{Lemma}
+\newtheorem{proposition}[mythmcounter]{Proposition}
+\newtheorem*{lemma*}{Lemma}
+
+% Formatting for source code & types
+\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
+\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
+\definecolor{strclr}{rgb}{0.25,0.1,0.0}
+\definecolor{prepclr}{rgb}{0.0,0.0,0.0}
+
+\newcommand{\sep}[0]{\; | \;}
+\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
+\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
+\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
+\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
+\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
+\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
+
+\newcommand{\ctyp}[2]{C^{#1}#2}
+\newcommand{\mtyp}[2]{M^{#1}#2}
+\newcommand{\subst}[3]{#1[#2 \leftarrow #3]}
+
+% Typing rule and typing statement (single line)
+\newcommand{\tyrule}[3]{ \inference[\footnotesize{(\emph{#1})}~~]{#2}{#3} }
+\newcommand{\tyruler}[3]{ \inference{#2}{#3}~~\footnotesize{(\emph{#1})} }
+
+% Coeffect algebra
+\definecolor{aclr}{rgb}{0.6,0.3,0.0}
+\newcommand{\aclrd}[1]{\textcolor{aclr}{#1}}
+
+% Coeffect algebra
+\definecolor{cclr}{rgb}{0.0,0.5,0.0}
+\newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}
+
+\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
+\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
+\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }}
+\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
+\newcommand{\czip}{ \textcolor{cclr}{\wedge} } % only used by flat coeffects
+\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
+\newcommand{\cgeq}{ \textcolor{cclr}{\geq} }
+
+\newcommand{\atimes}{ \textcolor{aclr}{\times} }
+\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
+\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq
+
+% Unified
+
+\newcommand{\SH}{ {\textcolor{sclr}{S}} }
+\newcommand{\SP}{ {\textcolor{sclr}{P}} }
+\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
+\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
+\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
+\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
+\newcommand{\sflat}{\star}
+
+\newcommand{\azero}{ \textcolor{aclr}{\bot} }
+\newcommand{\aunit}{ \alift{\cunit} }
+\newcommand{\aweak}{ \alift{\czero} }
+
+\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
+\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
+
+\newcommand{\xtimes}{\hat{\times}}
+
+\newcommand{\ctx}{\textit{ctx}}
+
+% Ordinary context where colouring is done by hand
+\newcommand{\coctx}[2]{ #1 \,\text{\scriptsize @}\, {{\text{$#2$}}}  }
+
+% Simple coeffect systems in the introduction
+\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}
+
+\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}}
+
+\definecolor{sclr}{rgb}{0.2,0,0.8}
+\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }
+
+\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
+\newcommand{\M}{ {\textcolor{cclr}{\mathcal{M}}} }
+
+\newcommand{\alen}[1]{{\aclrd{\textit{len}(}#1\aclrd{)}}}
+\newcommand{\slen}[1]{{\sclrd{\textit{len}(}#1\sclrd{)}}}
+\newcommand{\xlen}[1]{{\textit{len}(#1)}}
+
+\newcommand{\reduce}{\longrightarrow}
+
+\newcommand{\restr}[2]{#1|_{#2}}
+\newcommand{\narrow}[1]{\hspace{-0.5em} #1 \hspace{-0.65em}~}
+
+\definecolor{todoclr}{rgb}{0.8,0,0.5}
+\newcommand{\todo}[1]{\textcolor{todoclr}{ \begin{quotation} \noindent\textbf{TODO:} #1 \end{quotation} }}
+
+\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
+\newcommand{\catc}{\mathcal{C}}
+\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }
+
+\newcommand{\figcaption}[1]{
+  \vspace{0.5em}
+  \noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
+  \vspace{-2em}
+  \caption{#1}
+}
+% Half line spacing around align
+\newcommand{\zerodisplayskips}{%
+  \setlength{\abovedisplayskip}{0.5em}
+  \setlength{\belowdisplayskip}{0.5em}
+  \setlength{\abovedisplayshortskip}{0.5em}
+  \setlength{\belowdisplayshortskip}{0.5em}}
+\appto{\normalsize}{\zerodisplayskips}
+\appto{\small}{\zerodisplayskips}
+\appto{\footnotesize}{\zerodisplayskips}
+
+% half line spacing around compactitem
+\setlength{\pltopsep}{0.5em}
+\setlength{\plitemsep}{0.3em}
+
+\patchcmd{\thebibliography}{\chapter*}{\section*}{}{}
+
+\begin{document}
+
+\frenchspacing % Reduces space after periods to make text more compact
+\raggedbottom % Makes all pages the height of the text on that page
+\selectlanguage{american} % Select your default language - e.g. american or ngerman
+\pagenumbering{arabic} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
+\pagestyle{plain} % Suppress headers for the pre-content pages
+
+\newcommand{\semdeff}[4]{
+\begin{array}{rcl}
+  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
+  #2 & = & #4 \\
+\end{array}  
+}
+\newcommand{\semdefff}[6]{
+\begin{array}{rcl}
+  #1 & = & #4 \\
+  #2 & = & #5 \\ \cline{1-1}\cline{3-3}
+  #3 & = & #6 \\
+\end{array}  
+}
+\newcommand{\semdef}[2]{
+\begin{array}{rcl}
+  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
+  #1 & = & #2 \\
+\end{array}  
+}
+% ==================================================================================================
+
+\chapter{Translation semantics}
+\renewcommand*\thesection{\arabic{section}}
+
+The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
+of context-aware systems discussed earlier in Chapter~X. The coeffect calculus can be seen
+as a \emph{language framework} that simplifies the construction of concrete \emph{domain-specific}
+coeffect lnguages. First, it provides a parameterized type system that tracks the required context.
+Second, the comonadic semantics provides a way for implementing the langauge by translating
+it into a non-context-aware functional programming language with additional comonadically-inspired
+coeffect-specific primitives that implement the concrete notion of context-awareness.
+
+In this chapter, we give three concrete examples of how the coeffect \emph{language framework}
+gives rise to concrete \emph{context-aware domain-specific} languages and we discuss the safety
+guarantees provided by the coeffect type system.
+
+For each concrete context-aware langauge, the safety depends on the coeffect-specific 
+comonadically-inspired primitives, but the coeffect framework makes it easier to prove such safety.
+We only need to define the meaning of the comonadically-inspired coeffect-specific primitives
+and then show that those do not ``go wrong''. In doing so, we can use the coeffect annotations
+provided by the type system.
 
 \paragraph{Chapter structure and contributions}
+
 \begin{itemize}
-\item We present a \emph{structural coeffect calculus} as a type system that is parameterized 
-  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the 
-  system captures fine-grained liveness and data-flow information, as well as a calculus for bounded 
-  reuse (checking how many times is a variable accessed).
-
-\item We extend indexed comonads introduced in the previous chapter to 
-  \emph{structural indexed comonads} and use them to provide the semantics of structural 
-  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
-  semantics provides a motivation for the design of the calculus.
-
-\item We discuss the equational theory of the calculus. For structural coeffects, the substitution 
-  lemma holds for all instances of the calculus. Thus, we prove the type-preservation property 
-  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}). 
+\item 
+  We define the common subset of the target functional programming language. This includes the
+  syntax of the language, reduction rules and typing rules, but it does \emph{not} include 
+  coeffect-specific definitions. Well-typed programs written using the common subset of the 
+  target language do not get stuck (via progress and preservation), but they may reduce to 
+  \ident{error}, e.g. when accessing the head of an empty list.
+
+\item We then extend the language with coeffect-specific comonadically-inspired
+  data types and primtiives for dataflow and for implicit parameters. We show that the extension
+  preserves the progress and preservation properties.
+
+\item Next, we consider only programs in the target language that were produced by a
+  translation from the coeffect domain-specific language and we show that such programs not only
+  do not get stuck, but they also do not reduce to the \ident{error} value. In other words,
+  ``well-typed coeffect programs do not get hungry'' requiring more context than guaranteed
+  by the coeffect type system.
+
+\item We show how the approach extends to structural coeffect systems and we argue that our
+  proof can be generalized -- rather than reconsidering progress and preservation of the 
+  whole language, we can rely just on the correctness of the coeffect-specific comonadically-inspired
+  primitives and abstraction mechanism provided by languages such as ML and Haskell.  
 \end{itemize}
 
-% ==================================================================================================
-%                                                                                      
-%     ###           #                       #                 #       #                 
-%      #            #                       #                 #                         
-%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
-%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
-%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
-%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
-%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
-%                                                                                     
-% ==================================================================================================
+% --------------------------------------------------------------------------------------------------
 
 \section{Introduction}
-\label{sec:structural-intro}
-
-Compared to the previous chapter, structural coeffect calculi we consider are more 
-homogeneous and so finding the common pattern is in some ways easier. However, the systems
-are more complicated as they need to keep annotations attached to individual variables
-and thus require explicit structural rules. Before looking at the system, we briefly consider
-the most important related work.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Related work}
-
-In the previous chapter, we discussed the correspondence between coeffects and effects (and 
-between comonads and monads). As noted earlier, the $\lambda$-calculus is asymmetric in that
-an expression has multiple inputs (variables in the context), but just a single result 
-(the resulting value) and so monads and effects have no notion directly corresponding to 
-structural coeffect systems.
-
-The work in this chapter is more closely related to sub-structural type systems \cite{substruct-attpl-intro}.
-While sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
-\emph{exchange} rules, our systems keep them, but use them to manipulate both the context
-and its annotations.
-
-Our work follows the ``language semantics'' style in that we provide a structural semantics to
-the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
-the meta-language style, which extends the terms and types of a language with constructs for explicitly
-manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
-may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
-$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
-annotations as stand-alone types (indexed comonads only appear in the semantics).
-
-Our structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
-For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
-respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
-A benefit of this approach is that the typing judgements have the same structure as those of the 
-flat coeffect calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify 
-the two systems.
-
-% =================================================================================================
-%                                                            
-%       ###           ##                   ##                 
-%      #   #           #                    #                 
-%      #       ###     #     ###   #   #    #    #   #   ###  
-%      #          #    #    #   #  #   #    #    #   #  #     
-%      #       ####    #    #      #   #    #    #   #   ###  
-%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
-%       ###    ####   ###    ###    ## #   ###    ## #  ####  
-%                                                           
-% =================================================================================================
-
-\section{Structural coeffect calculus}
-\label{sec:struct-calculus}
-
-In the structural coeffect calculus, a vector of variables forming the free-variable context is 
-annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
-for different coeffect calculi and their properties are captured by the definition of \emph{structural 
-coeffect scalar} below. The scalar annotations can be \eg~integers (how many past values we 
-need) or values of a two-point lattice specifying whether a variable is live or not.
-
-Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in 
-the previous chapter). Functions always have exactly one input variable and so they are annotated
-with a single coeffect scalar. Thus the expressions and types of structural coeffect calculi are the 
-same as in the previous chapter. The only difference is that $\cclrd{r}, \cclrd{s}, \cclrd{t}$ now
-range over values of a \emph{structural coeffect scalar} rather than over values of a \emph{flat 
-coeffect algebra} (the definitions are related, but not the same).
-%
+
+The development in this chapter closely follows the example of effectful computations. 
+Effect systems provide a type system for tracking effects and monadic translation can be used as 
+a basis for implementing effectful domain-specific languages (e.g.~through the do-notation in 
+Haskell).
+
+This chapter also links together all the different technical developments presented in this thesis.
+The abstract comonadic semantics is used as a \emph{translation} that gives a concrete semantics
+to a number of concrete context-aware languages. The type system is used to guarantee that the
+resulting programs are correct. Finally, the development in this chapter is closely mirrored by 
+the implementation presented in Chapter~X.
+
+\subsection{Safety properties}
+The key claim of this thesis is that writing context-aware programs using coeffects is easier and
+safer. For example, consider the simple problem of writing a dataflow function that calculates the 
+difference between the current and the previous value.
+
+The following shows the code written in a coeffect dataflow language (left) and using lists to
+represent past values in a ML-like language (right):
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
+ \quad \ident{x}-\kvd{prev}~\ident{x}
+\end{array} 
+\qquad
+\begin{array}{l}
+\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
+ \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
+\end{array} 
+\end{equation*}
+%
+The function on the right has a type $\ident{num}~\ident{list}\rightarrow\ident{num}$. The function
+fails for input lists containing only zero or one elements. However, this is not reflected in the
+type and it is not enforced by the type checker.
+
+The function on the left has a type $\ctyp{\cclrd{1}}\ident{num}\rightarrow\ident{num}$ meaning 
+that it requires one past value (in addition to the current value). The information about required
+context is now reflected in the type. In this chapter, we show that well-typed (and well-coeffected!) 
+programs do not ``get stuck''. That is, the coeffect annotation in the type system is a safe
+over-approximation of the actual contextual information that will be accessed when a program runs.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Comonadic translation}
+
+The previous chapter gives the semantics of coeffect calculus in terms of indexed comonads. In this
+chapter, we follow the example of effects and monads and we use the semantics to define a 
+\emph{translation}.
+
+A context-aware \emph{source} program written using a concrete context-aware domain-specific 
+language (capturing dataflow, implicit parameters or other kinds of context awareness) with
+domain-specific language extensions (the \kvd{prev} keyword, or the \ident{?impl} syntax)
+is translated to a \emph{target} langauge that is not context-aware. The target language is
+a small functional language consisting of:
+%
+\begin{itemize}
+  \item Simple functional subset formed by lambda calculus with tuples and numbers.
+  \item Comonadically-inspired primitives corresponding to \emph{counit}, \emph{cobind} and
+    other operations of flat indexed comonads.
+  \item Additional primitives that model contextual operations of each concrete coeffect language
+    (\emph{prev} for the \kvd{prev} keyword, \emph{lookup} for the \ident{?impl} syntax etc.)
+\end{itemize}
+%
+The syntx, typing and reduction rules of the first part (simple functional language) are used by 
+all concrete coeffect domain-specific langauges. The syntax and typing rules of the second part 
+(comonadically-inspired) primitives are also shared by all coeffect DSLs, however the \emph{reduction
+rules} for the comonadically-inspired primitives differ -- they capture the concrete notion of
+context. Finally, the third part (domain-specific primitives) will differ for each coeffect DSL.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\paragraph{Language syntax}  
 \begin{equation*}
 \begin{array}{rcl}
-e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
-\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
-\end{array}
-\end{equation*}
-%
-In the previous chapter, the free variable context $\Gamma$ was treated as a set. In the 
-structural coeffect calculus, the order of variables matters. Thus we treat a free variable 
-context as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
-of the vector:
+  v &  = & n \sep \lambda x.e \sep (v_1, \ldots, v_n) \\
+  e &  = & x \sep n \sep \pi_i~e \sep (e_1, \ldots, e_n) \sep e_1~e_2 \sep \lambda x.e \\
+  \tau &  = & \ident{num} \sep \tau_1 \times \ldots \times \tau_n \sep \tau_1 \rightarrow \tau_2 \\
+  C  & = & (v_1, \ldots, v_{i-1}, \_, e_{i+1}, \ldots e_n) \sep v~\_ \sep \_~e \sep \pi_i~\_
+
+\end{array}
+\end{equation*}
+
+~
+\paragraph{Reduction rules}
+\begin{equation*}
+\begin{array}{rll}
+  \footnotesize{(\emph{fn})}  & (\lambda x.e)~v \rightarrow e[x\leftarrow v] \\[0.25em]
+  \footnotesize{(\emph{prj})} & \pi_i (v_1, \ldots, v_n) \rightarrow v_i\\[0.25em]
+  \footnotesize{(\emph{ctx})}  & C[e] \rightarrow C[e']  & (\textnormal{when}~e\rightarrow e')\\[0.25em]
+\end{array}
+\end{equation*}
+
+\caption{Common syntax and reduction rules of the target langauge}
+\label{fig:transl-target}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Target langauge}
+
+The target langauge for the translation is a simply typed lambda calculus with numbers and tuples.
+The translation uses tuples as it keeps a tuple with variable assignments (encoding those without 
+tuples would be possible, but cumbersome) and we add numbers as a basic concrete data type.
+In this section, we define the common parts of the language without the comonadically-inspired
+primitives.
+
+The syntax of the target programming langauge is shown in Figure~\ref{fig:transl-target}. 
+The values include numbers $n$, tuples and function values.
+The expressions include variables $x$, values, lambda abstraction and application and operations 
+on tuples. We do not need recursion (although a realistic programming language would include it). 
+In what follows, we also use the following syntactic sugar for let binding:
 %
@@
 \begin{array}{rcl}
- \Gamma &\narrow{=}& \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
- \xlen{\Gamma} &\narrow{=}& n
-\end{array}
-\end{equation*}
-%
-For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
-typing judgements, but the free variable context should be understood as a vector.
-Furthermore, the notation $\Gamma_1, \Gamma_2$ is here seen as the tensor product.
-Given $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
-$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
-$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
-
-Free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
-we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
-over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
-and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
-is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
-of coeffects is provided by the definition of \emph{structural coeffect algebra} below.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffect algebra}
-
-The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the 
-exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
-modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
-pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.
-
-\begin{definition}
-\label{def:structural-scalar}
-A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
-$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
-$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
-$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
-%
-\begin{equation*}
-\begin{array}{ccr}
-r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
-\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
-\textnormal{(monoid)}   
-\\
-r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
-\czero \;\cpar\; r = r = r \;\cpar\; \czero &
-\textnormal{(monoid)}   
-\\
-\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
-t\; \cleq\; t &
-\textnormal{(pre-order)}   
-\\
-\end{array}
-\end{equation*}
-%
-The following distributivity axioms hold:
-\begin{align*}
-\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
-\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
-\end{align*}
-%
-In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
+  \kvd{let}~x=e_1~\kvd{in}~e_2 &=& (\lambda x.e_2) e_1
+\end{array}
+\end{equation*}
+%
+Finally, $C[e]$ defines the context in which sub-expressions are evaluated. Together with the
+evaluation rules shown in  Figure~\ref{fig:transl-target}, this captures the standard call-by-name
+semantics of the common parts of the target language.
+The (standard) typing rules for the common expressions of the target langauge are shown in 
+Figure~\ref{fig:transl-targetty}. The rules are standard.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\paragraph{Typing rules}
+\begin{equation*}
+\begin{array}{ll}
+\tyrule{var}
+  { x:\tau \in \Gamma }
+  { \Gamma \vdash x:\tau }
+&
+\tyrule{err}
+  { ~ }
+  { \Gamma \vdash \ident{error}:\tau }
+\\[1em]
+\tyrule{str}
+  { ~ }
+  { \Gamma \vdash s:\ident{str} }
+&
+\tyrule{num}
+  { ~ }
+  { \Gamma \vdash n:\ident{num} }
+\\[1em]
+\tyrule{nil}
+  { ~ }
+  { \Gamma \vdash []:[\tau] }
+&
+\tyrule{head}
+  { \Gamma \vdash e :[\tau] }
+  { \Gamma \vdash \ident{head}~e:\tau }
+\\[1em]
+\tyrule{tail}
+  { \Gamma \vdash e :[\tau] }
+  { \Gamma \vdash \ident{tail}~e:[\tau] }
+\qquad~&
+\tyrule{abs}
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 }
+  { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{array}
+\end{equation*}
 \begin{equation*}
 \begin{array}{l}
- \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
- \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
-\end{array}
-\end{equation*}
-\end{definition}
-
-\noindent
-In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
-available from the declaration site and the call site or, in the syntactic reading, to split the 
-context requirements. The other difference is that the $\cpar$ operation of structural coeffect
-scalar is not required to be idempotent. The bounded reuse example uses $\cpar$ which is not 
-idempotent.
-
-Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
-via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma 
-(Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
-for all our examples; it is needed for the structural version of the substitution lemma 
-(Lemma~\ref{thm:structural-substitution}).
-
-In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
-coeffects becomes just vector concatenation or splitting, respectively; these are provided
-by the tensor product. The operations on vectors are indexed by integers representing the lengths
-of the vectors. The additional structure required by the type system for structural coeffect 
-calculi is given by the following definition.
-
-\begin{definition}
-A \emph{\aclrd{structural coeffect algebra}} is formed by a structural coeffect scalar 
-$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:
-
-\begin{itemize}
-\item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
-  structural coeffect scalars indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.
-
-\item An operation that constructs a vector from scalars indexed by the vector length
-  $\alift{-}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
-  and an operation that returns the vector length such that 
-  $\alen{\aclrd{\textbf{r}}} = \sclrd{n}$ for $\aclrd{\textbf{r}} : \C^{\sclrd{n}}$
-
-\item A pointwise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
-  such that\\ $\cclrd{t} \,\aseq\, \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} = 
-    \alift{\cclrd{t} \,\cseq\,\cclrd{r_1}, \ldots, \cclrd{t} \,\cseq\,\cclrd{r_n}}$.
-
-\item An indexed tensor product $\atimes_{\sclrd{n},\sclrd{m}} : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n}+\sclrd{m}}$
-  that is used in both directions -- for vector concatenation and for splitting -- 
-  which is defined as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} 
-     \atimes_{\sclrd{n}, \sclrd{m}} \alift{\cclrd{s_1}, \ldots, \cclrd{s_m}}
-   = \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}, \cclrd{s_1}, \ldots, \cclrd{s_m}}$
-\end{itemize}
-\end{definition}
-
-\noindent
-The fact that the tensor product $\atimes_{\sclrd{n},\sclrd{m}}$ is indexed by the lengths
-of the two vectors means that we can use it unambiguously for both concatenation of vectors
-and for splitting of vectors, provided that the lengths of the resulting vectors are known.
-In the following text, we usually omit the indices and write just 
-$\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}}$, because the lengths of the coeffect vectors 
-can be determined from the lengths of the matching free variable context vectors.
-
-More generally, we could see the the coeffect annotations as \emph{containers} 
-\cite{types-containers}. This approach is used in Chapter~\ref{ch:unified} to unify flat 
-and structural systems.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffect types}
-\label{sec:struct-calculus-types}
-
-The type system for the structural coeffect calculus is similar to sub-structural type systems
-in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
-weakening, exchange or contraction -- this is done by checking the types of sub-expressions
-in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
-does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
-perform corresponding transformation on the coeffect annotation.
-
-\paragraph{Syntax-driven rules.}
-The syntax-driven rules of the type system are shown in Figure~\ref{fig:struct-types} (a).
-The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
-using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
-\emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
-Constants (\emph{const}) are type checked in an empty variable context, which is annotated with
-an empty vector of coeffect annotations.
-
-The (\emph{abs}) rule assumes that the free variable context of the body can be split into
-a potentially empty \emph{declaration site} and a singleton context containing the bound
-variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
-the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
-non-determinism of the type inference present in flat coeffect systems.
-
-In (\emph{app}), the sub-expressions $e_1$ and $e_2$ use free variable contexts $\Gamma_1, \Gamma_2$
-with coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$, respectively. The function
-value is annotated with a coeffect scalar $\cclrd{t}$.
-The coeffect annotation of the composed expression is obtained by combining the annotations
-associated with variables in $\Gamma_1$ and $\Gamma_2$. Variables in $\Gamma_1$ are only
-used to obtain the function value, resulting in coeffects $\aclrd{\textbf{r}}$. The 
-variables in $\Gamma_2$ are used to obtain the argument value, which is then sequentially
-composed with the function, resulting in $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$.
-
-\paragraph{Structural rules.}
-The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
-They allow different transformation of the free variable context. We include (\emph{sub}) 
-as one of the rules, allowing sub-coeffecting on coeffect scalars belonging to individual 
-variables. The remaining rules capture \emph{weakening}, \emph{exchange} and \emph{contraction} 
-known from sub-structural systems.
-
-Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
-vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
-in the context, performing the same reordering on the coeffect vector. Finally recall that 
-variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
-allows re-using a variable. We can type check sub-expressions using two separate variables and 
-then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
-is the only place in the structural coeffect system where context requirements are combined
-(semantically, this is where the available context is shared).
-
-%---------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-{\small a.) Syntax-driven typing rules:}
-\begin{equation*}
-\tyrule{var}
-  { }
-  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{const}
-  {c \!:\! \tau \in \Delta}
-  {\coctx{()}{\alift{}} \vdash c : \tau }
-\end{equation*}
-\begin{equation*}
 \tyrule{app}
-  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
-   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-
-\vspace{1em}
-{\small b.) Structural rules for context manipulation:}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-(\cclrd{s'} \cleq \cclrd{s})
-\end{equation*}
-\begin{equation*}
-\tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-\begin{array}{l}
- \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
- \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{2.5em} 
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-~
-\begin{array}{l}
- \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
- \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-
-\figcaption{Type system for the structural coeffect calculus}
-\label{fig:struct-types}
+  { \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2  & \Gamma \vdash e_2 : \tau_1 }
+  { \Gamma \vdash e_1~e_2 : \tau_2 }
+\\[1.25em]
+\tyrule{cons}
+  { \Gamma \vdash e_1 :\tau & \Gamma \vdash e_2 : [\tau] }
+  { \Gamma \vdash e_1\!::\!e_2 : [\tau] }
+\\[1.25em]
+\tyrule{proj}
+  { \Gamma \vdash e : \tau_1 \times \ldots \tau_i \times \ldots \times \tau_n }
+  { \Gamma \vdash \pi_i~e:\tau_i }
+\\[1.25em]
+\tyrule{tup}
+  { \forall i\in\{1 \ldots n\}.~ \Gamma \vdash e_i : \tau_i }
+  { \Gamma \vdash (e_1, \ldots, e_n) : \tau_1 \times \ldots \times \tau_n }
+\\[1.25em]
+\tyrule{ifn}
+  { \Gamma \vdash e_1 : \ident{num} & \Gamma \vdash e_2 : \ident{num} &
+    \Gamma \vdash e_3 : \tau & \Gamma \vdash e_4 : \tau}
+  { \Gamma \vdash \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4 : \tau}
+\\[1.25em]
+\tyrule{ifs}
+  { \Gamma \vdash e_1 : \ident{str} & \Gamma \vdash e_2 : \ident{str} &
+    \Gamma \vdash e_3 : \tau & \Gamma \vdash e_4 : \tau}
+  { \Gamma \vdash \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4 : \tau}
+\end{array}
+\end{equation*}
+
+\caption{Typing rules for the common syntax of the target language}
+\label{fig:transl-targetty}
 \end{figure}
 
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Understanding structural coeffects}
-
-The type system for structural coeffects appears more complicated when compared to the flat
-version, but it is in many ways simpler -- it removes the ambiguity arising from the use
-of $\czip$ in lambda abstraction and, as discussed in Section~\ref{sec:struct-syntax}, has
-a cleaner equational theory.
-
-\paragraph{Flat and structural context.} In flat systems, lambda abstraction splits context
-requirements using $\czip$ and application combines them using $\cpar$. In the structural
-version, both of these are replaced with $\atimes$. The $\czip$ operation is not needed, but
-$\cpar$ is still used in the (\emph{contr}) rule.
-
-This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
-as over-approximations and under-approximations of $\atimes$. This is demonstrated by the 
-(\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
-$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
-combines two values representing the available context, potentially discarding parts of it 
-(under-approximation), while flat application splits the available context (a single value), 
-potentially duplicating parts of it (over-approxi\-mation)\footnote{Because of this duality, earlier 
-version of coeffect systems \cite{coeffects-icalp13} used $\wedge$ and $\vee$.}.
-
-Secondly, the operator $\cpar$ is used when the semantics passes a given context to multiple
-sub-expressions. In flat systems, the context is shared in (\emph{app}) and (\emph{pair}), because 
-the sub-expressions may share variables. In structural systems, the sharing is isolated into an 
-explicit contraction rule.
-
-\paragraph{Let binding.}
-The other aspect that makes structural systems simpler is that they remove the need for 
-separate let binding. As discussed in Section~\ref{sec:flat-calculus-let}, flat calculi include
-let binding that gives a \emph{more precise} typing than combination of abstraction and 
-application. Structural coeffect systems avoid this issue.
-
-\begin{remark}[Let binding]
-\label{thm:structural-letbinding}
-In a structural coeffect calculus, the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$ can be
-seen as a derived rule, \ie~its typing is equivalent to the typing of the expression
-$(\lambda x.e_2)~e_1$.
-
-\end{remark}
-\begin{proof}
-Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
-in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
-%
-\begin{equation*}
-\inference
-  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
-    \inference
-      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
-      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
-  { \inference
-      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
-      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
-\end{equation*}
-
-\noindent
-The assumptions and conclusions match those of the (\emph{let}) rule. 
-\end{proof}
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Examples of structural coeffects}
-
-The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
-presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
-structural liveness provide a more precise tracking of properties that can be tracked using 
-flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
-(by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
-example, implicit parameters could be attached to individual variables, but, to our knowledge, this
-has not been used in any practical system.
-
-On the other hand, some of the structural systems do not have a flat equivalent, typically 
-because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
-scalar. This is the case, for example, for the bounded variable use.
-
-\begin{example}[Structural liveness]
-The structural coeffect scalar for liveness is formed by
-$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
-the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
-with a join $\sqcup$ and a meet $\sqcap$. 
-\end{example}
-
-\begin{example}[Structural data-flow]
-In data-flow, context is annotated with natural numbers and the structural coeffect scalar is formed 
-by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
-\end{example}
-
-\noindent
-For the two examples that have both flat and structural version, obtaining the structural coeffect
-algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
-required by a structural coeffect algebra are the same as those required by the flat version
-and so the above definitions are both valid. Similar construction can be used for the
-\emph{optimized data-flow} example from Section~\ref{sec:flat-calculus-examples}.
-
-It is important to note that this gives us a systems with \emph{different} properties. The 
-information are now tracked per-variable rather than for entire contexts. For data-flow,
-we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
-for a pointwise extension of the $+$ operator, such that
-$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
-%
-\begin{equation*}
-\tyrule{prev}
-  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
-  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
-\end{equation*}
-%
-The rule appears similar to the flat one, but there is an important difference. Because of the
-structural nature of the type system, it only increments the required number of values for 
-variables that are used in the expression $e$. Annotations of other variables can be left 
-unchanged.
-
-Before looking at the semantics and equational properties of structural coeffect systems, 
-we consider bounded variable use, which is an example of structural system that does not
-have a flat counterpart.
-
-\begin{example}[Bounded variable reuse]
-The structural coeffect algebra for tracking bounded variable use is given by
-$(\mathbb{N}, \ast, +, 1, 0, \leq)$
-\end{example}
-
-\noindent
-Similarly to the structural calculus for data-flow, the calculus for bounded variable reuse
-annotates each variable with an integer. However, the integer denotes how many times is the variable 
-\emph{accessed} rather than how many \emph{past values} are needed. The resulting type system is 
-the one shown in Figure~\ref{fig:applications-struct-live} in Chapter~\ref{ch:applications}.
-
-% ==================================================================================================
-%                                              
-%      #####  #                                 
-%        #    #                                 
-%        #    # ##    ###    ###   # ##   #   # 
-%        #    ##  #  #   #  #   #  ##  #  #   # 
-%        #    #   #  #####  #   #  #      #  ## 
-%        #    #   #  #      #   #  #       ## # 
-%        #    #   #   ###    ###   #          # 
-%                                         #   # 
-%                                          ###                                                                         
-% ==================================================================================================
-
-\section{Categorical motivation}
-\label{sec:struct-semantics}
-
-When we introduced structural coeffect systems in Section~\ref{sec:applications-structural}, we 
-included a concrete semantics of structural liveness and bounded variable reuse. In this section,
-we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
-extension of the \emph{indexed comonad} structure. Similarly to \emph{flat indexed comonad}, the
-structural variant adds operations that are needed to embed full $\lambda$-calculus, this time 
-with per-variable contexts.
-
-As in the previous chapter, the main aim of this section is to motivate and explain the design of 
-the structural coeffect calculus shown in Section~\ref{sec:struct-calculus}. However, the semantics 
-also highlights the similarities and differences between the flat and structural system.
-Most of the differences arise from the fact that contexts in structural coeffect systems are treated 
-as \emph{vectors} rather than sets.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of vectors}
-
-Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing 
-judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a 
-morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, 
-we can freely transform the value contained in the context modelled using an indexed comonad
-$\ctyp{\cclrd{r}}$.
-
-Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
-which transforms the value inside the context without affecting the coeffect annotation. Thus,
-we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
-$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value 
-$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
-
-The ability to freely transform the variable structure is not desirable in the model of 
-structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
-the coeffect annotations matches the structure of variables. To achieve this, we model vectors
-using a structure distinct from ordinary products which we denote $-\xtimes-$. 
-For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
-is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
-We assume that the operator is equipped with necessary associativity transformations allowing us
-to use it freely on more than two values.
-
-The operator is a bifunctor, but it is \emph{not} a product in the categorical sense. In particular, 
-there is no way to turn $\tau_1 \xtimes \ldots \xtimes \tau_n$ into $\tau_i$ (the structure does 
-not have projections) and so there is also no way of turning
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$ into
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{\tau_i}$, which would break the correspondence 
-between coeffect annotations and variable structure.
-
-The structure created using $-\xtimes-$ can be manipulated only using operations provided by
-the \emph{structural indexed comonad}, which operate over variable contexts contained in an 
-indexed comonad $\ctyp{\aclrd{\textbf{r}}}$ and are designed to preserve the correspondence
-between vectors and annotations.
-
-In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
-As mentioned, we assume that the use of the operator can be freely re-associated. For example, when
-calling an operation that requires input of the form 
-$(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
-we use an argument $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ and assume that the 
-appropriate transformation is inserted.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Indexed comonads, revisited}
-
-The semantics of structural coeffect calculus reuses the definition of \emph{indexed comonad} almost
-without change. The additional structure that is required for context manipulation (merging and 
-splitting) is different and is here provided by the \emph{structural indexed comonad} structure that 
-we introduce in this section.
-
-Recall the definition from Section~\ref{sec:flat-semantics-idx}, which defines an indexed como\-nad 
-over a monoid $(\C, \cseq, \cunit)$ as a triple $(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, 
-\ident{cobind}_{\cclrd{r}, \cclrd{s}})$. The triple consists of a family of object mappings $\ctyp{\cclrd{r}{}}$,
-and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
-
-In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$
-representing function values (appearing in the language), but also of the form
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$, 
-modelling expressions in a context. To capture this, we need to revisit the definition and use 
-\emph{coeffect vectors} in some of the operations.
-
-\begin{definition}
-Given a monoid $(\C, \cseq, \cunit)$ with a pointwise extension of the $\cseq$ operator to a vector 
-(written as $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$) and an operation lifting scalars to vectors 
-$\alift{-}$, an \emph{indexed comonad} over a category $\catc$ is a triple 
-$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}})$:
-
-\begin{compactitem}
-\item $\ctyp{\aclrd{\textbf{r}}}{}$ for all $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$ is a family of object mappings 
-\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\alift{\cunit}}{\alpha} \rightarrow \alpha$ 
-\item $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$ is a mapping $(\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta) 
-  \rightarrow (\ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{s}}}{\beta})$
-\end{compactitem}
-\end{definition}
-
-\noindent
-The object mapping $\ctyp{\aclrd{\textbf{r}}}$ is now indexed by a vector rather than by a scalar 
-$\ctyp{\cclrd{r}}$ as in the previous chapter. This new definition supersedes the old one, because a 
-flat coeffect annotation can be seen as singleton vectors.
-
-The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
-return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
-operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
-is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
-takes a vector of variables, but always produces just a single value. Thus the resulting function 
-also takes a vector of variables, but always returns a context with a vector containing just one value.
-In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
-no way of constructing values containing $\xtimes$.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural indexed comonads}
-
-The flat indexed comonad structure extends indexed comonads with operations 
-$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ that combine or 
-split the additional (flat) context and are annotated with the flat coeffect operations $\czip$ 
-and $\cpar$, respectively.
-
-In the structural version, the corresponding operations operate convert between a (wrapped) vector 
-of values represented using $\xtimes$ and ordinary pairs of contexts containing parts of the vector. 
-The vectors of coeffect annotations are split or merged using $\atimes$ of the structural coeffect 
-algebra, in a way that mirrors the wrapped vectors (variable structure).
-
-The following definition includes $\ident{lift}_{\cclrd{r'},\cclrd{r}}$, which is similar as
-before and models sub-coeffecting and also $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ which models
-duplication of a variable in a context needed for the semantics of contraction:
-
-\begin{definition}
-Given a structural coeffect algebra formed by $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$
-with operations $\alift{-}$ and $\aseq$, a \emph{structural indexed comonad} is an indexed comonad over 
-the monoid $(\C, \cseq, \cunit)$ equipped with families of operations $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$, 
-$\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$,
-$\ident{dup}_{\cclrd{r}, \cclrd{s}}$ and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
-%
-\begin{compactitem}
-\item $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
-  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}$
-\item $\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
-  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
-\item $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ is a family of mappings
-  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \xtimes \alpha)}$
-\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
-  $\ctyp{\alift{\cclrd{r'}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
-\end{compactitem}
-%
-Here, the following equalities must hold:
-\begin{equation*}
-\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~\equiv~ \ident{id} \qquad\qquad
-\ident{id} ~\equiv~ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} 
-\end{equation*}
-
-\end{definition}
-
-\noindent
-These operations differ from those of the flat indexed comonad in that the \ident{merge} and 
-\ident{split} operations are required to be inverse functions and to preserve the additional 
-information about the context. This was not required for the flat system where the operations
-could under-approximate or over-approximate. Note that the operations use $\xtimes$ to combine 
-or split the contained values. This means that they operate on free-variable vectors rather than 
-on ordinary products.
-
-The \ident{dup} mapping is a new operation that was not required for a flat calculus. It 
-takes a variable context with a single variable annotated with $\cclrd{r}\,\cpar\,\cclrd{s}$,
-duplicates the value of the variable $\alpha$ and splits the additional context between the two
-new variables. In a flat calculus, this operation was expressed using ordinary tuple 
-construction, which is not possible here -- the returned context needs to contain a 
-two-element vector $\alpha \xtimes \alpha$.
-
-Finally, the \ident{lift} mapping is almost the same as in the flat version. It
-operates on a singleton vector, which is equivalent to operating on a scalar as before.
-The operation could be extended to a vector in a pointwise way, but we keep it simple
-and perform sub-coeffecting on individual variables.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[!t]
-\begin{equation*}
-\begin{array}{ll}
- \sem{\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }~\ctx =
-   \ident{counit}_{\cunit}~\ctx & (\emph{var})
-\\[0.4em]
-\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
-  \delta~(c_i) & (\emph{const})
-\\[0.4em]
-\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
-  \hspace{1.5em}
-  \begin{array}{l}  
-  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}~\ctx\\[-0.25em]
-  \kvd{in}~\sem{\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
-      (\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~\sem{\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\ctx_2)
-  \end{array}
-\\[1.4em]
-\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
-\hspace{1.75em}\sem{\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
-  ~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v))
-\\[0.4em]
-\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
-  \hspace{1.75em}
-  \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx~~\kvd{in}~~
-  \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}~\ctx_1
-\\[0.4em]
-\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{sub})\\
-  \hspace{1.75em}
-  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
-  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \alift{\cclrd{s'}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
-\\[0.4em]
-\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
-        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\[-0.25em]
-  \hspace{1.75em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
-     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
-  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, 
-     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
-\\[0.4em]
-\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
-        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\[-0.25em]
-  \hspace{1.75em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
-      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
-  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, 
-      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)
-\end{array}
-\end{equation*}
-\\[0.0em]
-{\small The definition uses the following auxiliary definitions:}
-\begin{equation*}
-\hspace{-2em}
-\begin{array}{l}
- \ident{swap}_{\cclrd{t},\cclrd{s}} ~:~ 
-  \ctyp{\alift{\cclrd{t}, \cclrd{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\alift{\cclrd{s}, \cclrd{t}}}{(\beta \xtimes \alpha)} \\[-0.25em]
-\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx = \\
-  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}}~\ctx \\[-0.25em]
-  \hspace{1.75em} \kvd{in}~\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)
- \\[0.4em]
-\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} ~:~
-  (\ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{s'}}}{\beta'}) \rightarrow
-    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta \xtimes \gamma)} \rightarrow
-    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s'}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta' \xtimes \gamma)}\\[-0.25em]
-\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f~\ctx =\\
-  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx \\[-0.25em]
-  \hspace{1.75em} \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}~\ctx' \\[-0.25em] 
-  \hspace{1.75em} \kvd{in}~\ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~
-        (\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
-\end{array}
-\end{equation*}
-\figcaption{Categorical semantics of the structural coeffect calculus}
-\label{fig:struct-semantics}
-\vspace{-0.5em}
-\end{figure*}
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of structural calculus}
-
-The concrete semantics for liveness and bounded variable use shown in 
-Sections~\ref{sec:applications-struct-live} and \ref{sec:applications-struct-bll}
-suggests that semantics of structural coeffect calculi tend to be more complex 
-than semantics of flat coeffect calculi. The complexity comes from the fact
-that we need a more expressive representation of the variable context -- \eg~a vector
-of optional values (\ie~wrapped in the \ident{Maybe} data type). Additionally,
-the structural system needs to pass separate variable contexts to the sub-expressions.
-
-The latter aspect is fully captured by the semantics shown in this section. The
-earlier point is left to the concrete notion of structural coeffect. Our model still gives
-us the flexibility of defining the concrete representation of variable vectors. We explore
-a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
-at a unified categorical semantics defined in terms of \emph{structural indexed comonads}.
-
-\paragraph{Contexts and functions.}
-In the structural coeffect calculus, expressions in context are interpreted as functions taking
-a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
-such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
-structure is annotated with a scalar, such as $\ctyp{\cclrd{r}}$, which we treat as being equivalent
-to a singleton vector annotation $\ctyp{\alift{\cclrd{r}}}$:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
-  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
-\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
-\end{array}
-\end{equation*}
-%
-Note that the instances of flat indexed comonad ignored the fact that the variable 
-context wrapped in the data structure is a product. This is not generally the case for the 
-structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
-are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
-rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need to examine the structure
-of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
-
-\paragraph{Expressions.}
-A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
-the previous chapter, the semantics is written in a programming language style using constructs
-such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
-can equally be written using primitives such as currying, uncurrying, function pairing etc.
-
-The following summarizes how the standard syntax-driven rules work, highlighting the differences
-from the flat version:
-
-\begin{itemize}
-\item When accessing a variable (\emph{var}), the context now contains \emph{only} the accessed
-  variable and so the semantics is just $\ident{counit}_{\cunit}$ without a projection.
-  Constants (\emph{const}) are interpreted using a global dictionary $\delta$ as earlier.
-
-\item The semantics of flat function application first duplicated the context so that the same 
-  variables can be passed to both sub-expressions. This is no longer needed -- the 
-  (\emph{app}) rule splits the variables \emph{including} the additional context into two parts.
-  Passing the first context to the semantics of $e_1$ gives us a function
-  $\ctyp{\alift{\cclrd{t}}}{\tau_1} \rightarrow \tau_2$.
-
-  A value $\ctyp{\alift{\cclrd{t}}}{\tau_1}$ required in order to call the function is obtained by 
-  applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ to the semantics of $e_2$. The result 
-  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
-  is then called with the latter part of the split input context.
-
-\item The semantics of function abstraction (\emph{abs}) is syntactically the same as in the 
-  flat version -- the only difference is that we now merge a free-variable context with a 
-  singleton vector, both at the level of variable assignments and at the level of coeffect 
-  annotations.
-\end{itemize}
-
-\noindent
-The semantics for the non-syntax-driven rules (weakening, exchange, contraction) performs 
-transformations on the free-variable context. Weakening (\emph{weak}) splits the context and 
-ignores the part corresponding to the removed variable. If we were modelling the semantics in 
-a language with a linear type system, this would require an additional operation for ignoring an
-unused context annotated with $\aclrd{\czero}$.
-
-The remaining rules perform a transformation anywhere inside the free-variable vector. 
-To simplify writing the semantics, we define a helper 
-$\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
-that splits the variable vector into three parts, transforms the middle part and then 
-merges them, using the newly transformed middle part.
-
-The transformations on the middle part are quite simple. The (\emph{sub}) rule 
-uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
-the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
-uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a variable while 
-splitting its additional context.
-
-\paragraph{Properties.}
-
-As in the flat calculus, the main reason for defining the categorical semantics in this chapter
-is to provide validation for the design of the calculus. The following correspondence theorem
-states that the annotations in the typing rules of the structural coeffect calculus correspond 
-to the indices of the semantics. Thus, the calculus captures a context-dependent property if
-it can be modelled by a \emph{structural indexed comonad}. As we show in the next section, 
-this is the case for all three discussed examples (liveness, data-flow, bounded variable reuse).
-
-\begin{theorem}[Correspondence]
-In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$ 
-and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types 
-$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$ 
-and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined 
-by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Properties}
+
+The subset of the langauge described so far models a simple ML-like functional programming 
+langauge (or, Haskell-like language, if we choose call-by-name evaluation). The subset of the
+langauge introduced so far satisfies the property that ``well-typed programs do not get stuck'',
+i.e.~both type preservation (reduction does not change the type of an expression) and the progress 
+property (an well-typed expression is either a value or can be reduced).
+We first show this for the subset of the language discussed so far and later extend the proof to 
+also cover the comonadically-inspired primitives that will be added in the next section.
+
+\begin{theorem}[Type preservation]
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
 \end{theorem}
 \begin{proof}
-By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
+  Rule induction over $\rightarrow$.
 \end{proof}
 
-\noindent
-As in the flat calculus, the primitive operations of the structural indexed comonad are each
-annotated with a unique operations provided by the coeffect annotations. This means that the
-semantics fully determines the structure of the typing rules of the structural coeffect
-calculus.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Examples of structural indexed comonads}
-\label{sec:struct-semantics-examples}
-
-The categorical semantics for structural coeffect calculus is easily instantiated to give
-semantics for a concrete calculus. In this section, we revisit the three examples discussed
-throughout this chapter -- structural liveness, data-flow and bounded variable reuse. 
-Some aspects of the first two examples will be similar to flat versions discussed in
-Section~\ref{sec:flat-semantics} -- they are based on the same data structures (option and a 
-list, respectively), but the data structures are composed differently. Generally speaking,
-rather than having a data structure over a product of variables, we now have 
-a vector of variables over a specific data structure.
-
-The abstract semantics does not specify how vectors of variables should be represented, so this can
-vary in concrete instantiations. In all our examples, we represent a vector of variables as a 
-product written using $\times$. To distinguish between products representing vectors and ordinary
-products (\eg~a product of contexts returned by \ident{split}), we write vectors using 
-$\langle a, \ldots, b \rangle$ rather than the parentheses, used for ordinary tuples.
-
-\paragraph{Data-flow.}
-It is interesting to note that the semantics of data-flow and bounded variable reuse (discussed
-next) both keep a product of multiple values for each variable, so they are both built around an 
-\emph{indexed list} data structure. However, their \ident{cobind} and \ident{dup} operations work 
-differently. We start by looking at the structure modelling data-flow computations. For readability,
-variables in bold face (such as $\mathbf{a_1}$) range over vectors while ordinary notation (such
-as $a_1$) is used for individual values.
-
-\begin{example}[Indexed list for data-flow]
-\label{ex:struct-dfl}
-The indexed list model of data-flow computations is defined over a structural coeffect
-algebra $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
-is indexed by required number of past variables for each individual variable. It is defined
-over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
-containing a current value followed by $n_i$ past values:
-%
-\begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
-\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
-  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{(n_k+1)-\textnormal{times}}
-\end{equation*}
-%
-The mappings that define the structural indexed comonad include the \ident{split} and \ident{merge}
-operations that are shared by the other two examples (discussed below):
-%
+\begin{theorem}[Progress]
+  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+\end{theorem}
+\begin{proof}
+  By rule induction over $\vdash$.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect-specific extnesions}
+
+Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ of a concrete 
+coeffect domain-specific langauge, we first extend the language syntax and typing rules with 
+the terms that correspond to the comonadically-inspired operations. This is done in the same way
+for all concrete coeffect DSLs and so we give the additional syntax, evaluation context and 
+typing rules just once in Figure~\ref{fig:transl-ext}.
+
+The figure defines the syntax and the typing rules, but it does not define the reduction rules.
+Those -- together with the values for a concrete notion of context -- will be defined separately
+for each individual coeffect DSL. We first consider the DSL for dataflow programming.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\paragraph{Language syntax}
+Given $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$, extend the syntax:
+
+\begin{equation*}
+\begin{array}{rcl}
+  e &  =  &\ldots \sep\ident{cobind}_{\cclrd{s},\cclrd{r}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~e 
+         \sep \ident{split}_{\cclrd{r},\cclrd{s}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \\
+  \tau &  = &\ldots \sep \ctyp{\cclrd{r}}{\tau} \\
+  C  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~\_~e \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
+    &\sep& \ident{merge}_{\cclrd{r},\cclrd{s}}~\_
+           \sep \ident{split}_{\cclrd{r},\cclrd{s}}~\_ \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~\_ \\
+
+\end{array}
+\end{equation*}
+
+~
+
+~
+
+\paragraph{Typing rules}
+Given $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$, add the typing rules:
+
 \begin{equation*}
 \begin{array}{l}
-\ident{merge}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
-  (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle) = \\[-0.25em]
-\quad \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle
-\\[0.45em]
-\ident{split}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
-   \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle = \\[-0.25em]
-\quad (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle)
-\end{array}
-\end{equation*}
-%
-The remaining mappings that are required by structural indexed comonad and capture the
-essence of data-flow computations are defined as:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{counit}_{0}~\langle \langle a_0 \rangle \rangle = a_0
-\\[0.45em]
-\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
-  \langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle = \\[-0.25em]
-\quad \langle\langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle, ~\ldots~, \\[-0.25em]
-\quad ~~~\; f \langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle~\; \rangle\rangle
-\\[0.45em]
-\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{\mathit{max}(m,n)} \rangle \rangle =
-   \langle \langle a_1, \ldots, a_m \rangle, \langle a_1, \ldots, a_n \rangle \rangle
-\\[0.45em]
-\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
-  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The definition of the indexed list data structure relies on the fact that the number of annotations
-corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
-containing $n_i+1$ values for the $i$-\textit{th} variable (the annotation represents the number of 
-required \emph{past} values so one more value is required).
-
-The \ident{split} and \ident{merge} operations are defined separately, because they are not specific
-to the example. They operate on the top-level vectors of variables (without looking at the 
-representation of the variable). This means that we can re-use the same definitions for the following
-two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will there represent options
-rather than lists).
-
-The mappings that explain how data-flow computations work are \ident{cobind} (representing sequential
-composition) and \ident{dup} (representing context sharing or parallel composition). In \ident{cobind},
-we get $k$ vectors corresponding to $k$ variables, each with $m+n_i$ values. The operation calls 
-$f$ $m$-times to obtain $m$ past values required as the result of type $\ctyp{\langle m \rangle} \beta$.
-
-The $\ident{dup}_{m,n}$ operation needs to produce a two-variable context containing $m$ and $n$ values,
-respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
-is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable
-and \ident{lift} drops additional past values that are not required.
-
-~
-
-\paragraph{Bounded reuse.}
-As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list 
-structure. Rather than representing possibly different past values that can be shared (see the 
-definition of \ident{dup}), the list now represents multiple copies of the same value as each value
-can only be accessed once. This semantics follows that of Girard \cite{logic-bounded}.
-
-\begin{example}[Indexed list for bounded reuse]
-The indexed list model of bounded variable reuse is defined over a structural coeffect
-algebra $(\mathbb{N}, \ast, +, 1, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$ 
-is a vector containing $n_i$ values of $i$-\textit{th} variable:
-%
-\begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
-\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
-  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{n_k-\textnormal{times}}
-\end{equation*}
-%
-The \ident{merge} and \ident{split} operations are defined as in Example~\ref{ex:struct-dfl}.
-The operations that capture the behaviour of bounded reuse are:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{counit}_{1}~\langle \langle a_0 \rangle \rangle = a_0
-\\[0.45em]
-\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
-  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
-\\[0.45em]
-\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{m+n} \rangle \rangle = 
-\quad \langle \langle a_1, \ldots, a_m \rangle, \langle a_{m+1}, \ldots, a_{m+n} \rangle \rangle
-\\[0.45em]
-\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
-  \langle \langle a_{1,0}, \ldots a_{1,m\ast{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m\ast{n_k}} \rangle \rangle = \\[-0.25em]
-\quad \langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1 - 1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k - 1}} \rangle \rangle, ~\ldots~, \\[-0.25em]
-\quad ~~ f \langle \langle a_{1,(m-1)\ast n_1}, \ldots a_{1,(m-1) \ast n_1} \rangle, \ldots, 
-                   \langle a_{k,m\ast n_k - 1}, \ldots a_{k,m\ast n_k - 1} \rangle \rangle~\; \rangle
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The \ident{counit} and \ident{lift} operations are defined as previously -- variable access 
-extracts the only value of the only variable and sub-coeffecting allows discarding multiple
-copies of a value that are not needed.
-
-In the bounded variable reuse system, variable sharing is annotated with the $+$ operator (in contrast 
-with \textit{max} used in data-flow). The $\ident{dup}_{m,n}$ operation thus splits the $m+n$ available
-values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The 
-\ident{cobind} operation works similarly -- it splits $m\ast n_i$ available values of each
-variable into $m$ vectors containing $n_i$ copies and then calls the $f$ function $m$-times to 
-obtain $m$ resulting values without sharing any input value.
-
-\paragraph{Liveness.}
-In both data-flow and bounded reuse, the data type is defined as a vector of values obtained
-by applying the indexed list type constructor to types of individual variables. We can
-generalize this pattern. Given a parameterized (indexed) type constructor $D^l \alpha$, we define 
-$\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of a vector of $D^{l_i}$ types.
-For liveness, the definition lets us reuse some of the mappings used when defining the semantics of
-flat liveness. However, we cannot fully define the semantics of the structural version in terms of 
-the flat version -- the \ident{cobind} operation is different and we need an appropriate 
-\ident{dup} operation.
-
-\begin{example}[Structural indexed option]
-Given a structural coeffect algebra formed by $(\{ \ident{L}, \ident{D} \}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$
-and the indexed option data type $D^l$, such that $D^{\ident{D}}\alpha = 1$ and $D^{\ident{L}}\alpha=\alpha$,
-the data type for structural indexed option comonad is:
-%
-\begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
-  D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
-\end{equation*}
-%
-The \ident{merge} and \ident{split} operations are defined as earlier. The remaining operations
-model variable liveness as follows:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{cobind}_{\ident{L}, \langle l_1, \ldots, l_n \rangle}~f~\langle a_1, \ldots, a_n \rangle = \langle f~\langle a_1, \ldots, a_n \rangle \rangle \\[-0.25em]
-\ident{cobind}_{\ident{D}, \langle \ident{D}, \ldots, \ident{D} \rangle}~f~\langle (), \ldots, () \rangle = \langle \ident{D} \rangle
-\\[0.5em]
-\hspace{-0.5em}
-\begin{array}{lll}
-~\\[-0.85em]
-\ident{dup}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle (), () \rangle \\[-0.25em]
-\ident{dup}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle a, () \rangle \\[-0.25em]
-\ident{dup}_{\ident{D}, \ident{L}} \langle a \rangle &\narrow{=}& \langle (), a \rangle \\[-0.25em]
-\ident{dup}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a, a \rangle 
-\end{array}
-\qquad\quad
-\begin{array}{lll}
-\ident{counit}_{\ident{L}}~\langle a \rangle &\narrow{=}& a
-\\[0.45em]
-\ident{lift}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a \rangle \\[-0.25em]
-\ident{lift}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
-\ident{lift}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
-\end{array}
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-When the expected result of the \ident{cobind} operation is dead (second case), the operation can
-ignore all inputs and directly return the unit value $()$. Otherwise, it passes the vector of 
-input variables to $f$ as-is -- no matter whether the individual values are live or dead. The
-\ident{L} annotation is a unit with respect to $\cap$ and so the annotations expected by $f$ are
-the same as those required by the result of \ident{cobind}.
-
-The \ident{dup} operation resembles with the flat version of \ident{split} -- this is expected as
-duplication in the flat calculus is performed by first duplicating the variable context (using 
-\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair. Depending on the
-required coeffect annotations, this may copy (duplicate) the value, or it may produce an empty context.
-
-Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
-$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models 
-sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
-it behaves as identity.
-
-% ==================================================================================================
-%                                                                   
-%       ###                  #                    #       #          
-%      #   #                 #                    #                  
-%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
-%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
-%          #  #  ##  #   #   #      ####  #       #       #    #     
-%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
-%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
-%             #   #                                                  
-%              ###                                                   
-%
-% ==================================================================================================
-
-\section{Syntactic equational theory}
-\label{sec:struct-syntax}
-
-Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
-different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
-certain equational properties hold in all instances of the structural coeffect calculus.
-
-In this section, we look at these common properties that we get ``for free'' from the structural 
-coeffect calculus. We start the discussion by briefly considering the key aspects
-that make the equational theory of flat and structural coeffects different.
-
-\subsection{From flat coeffects to structural coeffects}
-\label{sec:struct-syntax-intro}
-
-When discussing syntactic reductions for the flat calculus in Section~\ref{sec:flat-syntax}, 
-we noted that no single technique of specifying syntactic reduction works universally for all 
-flat coeffect calculi. We considered multiple alternatives (call-by-name, call-by-value, 
-internalized substitution) that can be used as the basis for operational semantics for
-different calculi that satisfy certain additional properties.
-
-The structural coeffect calculus has more preferable equational properties. In particular, 
-we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of structural 
-coeffect algebra. For this reason, we focus on these two reductions in this section.
-Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
-the structural coeffect calculus satisfies both the \emph{local soundness} and 
-the \emph{local completeness} properties, witnessed by $\beta$-reduction and $\eta$-expansion,
-respectively.
-
-\paragraph{Substitution for flat coeffects.} 
-The less obvious (\emph{top-pointed}) variant of the substitution lemma for flat coeffects 
-(Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
-to coincide. This enables substitution to preserve the type of expressions, because all
-additional requirements arising as the result of the substitution can be associated with the
-declaration context. For example, consider the following example where implicit parameter 
-$\ident{?offset}$ is substituted for the variable $y$:
-%
-\begin{equation*}
-\begin{array}{rclcll}
- \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textit{before}) \\
-\coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
-  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textit{after})
-\end{array}  
-\end{equation*}
-%
-The typing judgement obtained in (\emph{after}) preserves the type of the expression (function 
-value) from the original typing (\emph{before}). This is possible thanks to the non-determinism 
-involved in the typing rule for lambda abstraction -- as all operators of the flat coeffect 
-algebra used here are $\cup$, we can place the additional requirement on the outer context. 
-Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.
-
-Here, the flat coeffect calculus gives us typing with limited \emph{precision}, but 
-enough \emph{flexibility} to prove the substitution lemma.
-
-\paragraph{Substitution for structural coeffects.}
-In contrast, the substitution lemma (Lemma~\ref{thm:structural-substitution}, shown later on 
-page~\pageref{thm:structural-substitution}) for structural coeffects can be proven because 
-structural coeffect systems provide enough \emph{precision} to identify exactly with which 
-variable should a context requirement be associated.
-
-The following example shows a situation similar to the previous one. Here, we use
-structural data-flow calculus (writing $\kvd{prev}~e$ to obtain previous value of the 
-expression $e$) and we substitute $w+z$ for $y$:
-%
-\begin{equation*}
-\begin{array}{rclcll}
- \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} } 
-  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{before}) \\
-\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
-  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{after}) \\
-\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
-  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{final})
-\end{array}  
-\end{equation*}
-%
-The type of the function does not change, because the structural type system associates
-the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not 
-affect how the variable $x$ is used.
-
-The other aspect demonstrated in the example is how the coeffect of the substituted 
-variable affects the free-variable context of the substituted expression. Here, the
-original variable $y$ is annotated with $\cclrd{2}$ and we substitute it for an
-expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
-The substitution applies the operation $\aseq$ (modelling sequential composition) to 
-the annotation of the new context -- in the above example
-$\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}=\alift{\cclrd{2}, \cclrd{2}}$.
-
-\subsection{Holes and substitution lemma}
-\label{sec:struct-syntactic-subst}
-
-As demonstrated in the previous section, reduction (and substitution) in the structural coeffect 
-calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
-More importantly, because the system uses explicit contraction, we may also need to substitute 
-for multiple variables in the variable context at the same time.
-
-Consider the expression $\lambda x.x+x$. It is type-checked by type-checking $x_1 + x_2$,
-contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction of 
-$(\lambda x.x+x)~(y+z)$ we need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
-This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
-such as the bunched typing system \cite{substruct-bunched}. To express the substitution 
-lemma later in this section, we first define the notion of a \emph{context with holes}:
-
-\begin{definition}[Context with holes]
-A \emph{context with holes} is a context such as $\coctx{x_1\!:\!\tau_1, \ldots, x_k\!:\!\tau_k}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_k}}}$,
-where some of the variable typings $x_i\!:\!\tau_i$ and corresponding coeffects $\cclrd{r_i}$ 
-are replaced by \emph{holes} written as $-$.
-
-\begin{equation*}
-~\Delta[\,\coctx{-}{-}\,]_n = \Delta[\,\underbrace{\coctx{-}{-} \,|\, \ldots \,|\, \coctx{-}{-}}_{ n-\textit{times} } \,]
-\end{equation*}
-\begin{equation*}
-\begin{array}{lcl}
- \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
-   \coctx{-, \Gamma}{\alift{-} \atimes \aclrd{\textbf{s}} }
-   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_{n-1} \\
- \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
-   \coctx{x\!:\!\tau, \Gamma}{\alift{\cclrd{r}} \atimes \aclrd{\textbf{s}} }
-   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_n \\
- \Delta[\,\coctx{-}{-}\,]_0 &\narrow{:=}&
-   \coctx{()}{\alift{}}
-\end{array} 
-\end{equation*} 
-\end{definition}
-
-\noindent
-A context with $n$ holes may either start with a hole, followed by a context with $n-1$ holes, or it may
-start with a variable followed by a context with $n$ holes. Note that the definition ensures that
-the locations of variable holes correspond to the locations of coeffect annotation holes. Given a context
-with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
-obtain an ordinary coeffect-annotated context.
-
-\begin{definition}[Hole filling] Given a context with $n$ holes $\coctx{\Delta}{\textbf{s}} \in 
-\Delta[\,\coctx{-}{-}\,]_n$, the hole filling operation written as 
-$\coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\mathbf{r_1}}} \,|\, \ldots \,|\, \coctx{\Gamma_n}{\aclrd{\mathbf{r_n}}} \,]$,
-which replaces the holes by the specified variables and corresponding coeffect annotations, is defined as:
-%
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash \ident{counit}_{\cunit}~e : \tau} \\
+\\
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 & \Gamma \vdash e_2 : \ctyp{\cclrd{r}\cseq\cclrd{s}}{\tau_1} }
+  {\Gamma \vdash \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2 : \ctyp{\cclrd{s}}{\tau_2}} \\
+  \\
+\tyrule{merge}
+  {\Gamma \vdash  e : \ctyp{\cclrd{r}}{\tau_1} \times \ctyp{\cclrd{s}}{\tau_2} } 
+  {\Gamma \vdash  \ident{merge}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\cclrd{r}\czip\cclrd{s}}{(\tau_1 \times \tau_2)} }\\
+    \\
+\tyrule{split}
+  {\Gamma \vdash  e : \ctyp{\cclrd{r}\cpar\cclrd{s}}{(\tau_1 \times \tau_2)} }
+  {\Gamma \vdash  \ident{split}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\cclrd{r}}{\tau_1} \times \ctyp{\cclrd{s}}{\tau_2}} \\
+\\
+\tyrule{lift}
+  {\Gamma \vdash  e : \ctyp{\cclrd{r'}}{\tau} }
+  {\Gamma \vdash \ident{lift}_{\cclrd{r'},\cclrd{r}}~e : \ctyp{\cclrd{r}}{\tau} } \qquad\forall \cclrd{r'},\cclrd{r}, \cclrd{r}\,\cleq\,\cclrd{r'}
+\\
+\end{array}
+\end{equation*}
+
+\caption{Comonadically-inspired extensions for the target langauge}
+\label{fig:transl-ext}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Comonadically-inspired translation}
+
+In Chapter 4, we presented the semantics of the flat coeffect calculus in terms of indexed comonads.
+We treated the semantics as denotational -- interpreting the meaning of a given typing derivation
+of a program in terms of category theory.
+
+In this chapter, we use the same structure in a different way. Rather than treating the rules as
+\emph{denotation} in categorical sense, we treat them as \emph{translation} from a source 
+domain-specific coeffect language into a target language with comonadically-inspired primitives
+described in the previous section.
+
+Similarly, the interpretation of contexts and types in the category now becomes a translation from types and
+contexts in the source language into the types of the target language:
+
 \begin{equation*}
 \begin{array}{rcl}
- \coctx{-, \Delta}{\alift{-} \atimes \textbf{s} }
-    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
-      \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}_1} \atimes \aclrd{\textbf{r}_2} } \\
- && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
-     \coctx{\Delta}{\textbf{s}}[\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
-\\[0.5em]
- \coctx{x_1\!:\!\tau, \Delta}{\alift{\cclrd{r_1}} \atimes \textbf{s} }
-    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
-      \coctx{x_1\!:\!\tau, \Gamma_2}{\alift{\cclrd{r_1}} \atimes \aclrd{\textbf{r}_2} } \\
- && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
-     \coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
-\\[0.5em]
- \coctx{()}{\alift{}}~[~] &\narrow{=}& \coctx{()}{\alift{}}
-\end{array} 
-\end{equation*} 
-\end{definition}
-
-\noindent
-When we substitute an expression with coeffects $\aclrd{\textbf{t}}$ (associated with variables $\Gamma$)
-for a variable that has coeffects $\cclrd{s}$, the resulting coeffects of $\Gamma$ need to 
-combine $\aclrd{\textbf{t}}$ and $\cclrd{s}$. Unlike in the flat coeffect systems, the structural 
-substitution does not require all coeffect algebra operations to coincide and so the combination
-is more interesting than in the bottom-pointed substitution for flat coeffects, where it used
-the only available operator (Lemma~\ref{thm:cbn-substitution-bot}).
-
-Substitution can be seen as a form of sequencing, \ie~the term $\subst{e_1}{x}{e_2}$ should have
-the same type as the term $\kvd{let}~x=e_2~\kvd{in}~e_1$. In other words, we first need to obtain the 
-value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
-requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
-are annotated with $\cclrd{s} \,\aseq\, \aclrd{\textbf{t}}$, using the (scalar-vector extension)
-of the sequential composition operator $\cseq$.
-
-An operational reading of the resulting coeffect annotation is that, to evaluate each of the variables
-annotated with coeffects in the vector $\aclrd{\textbf{t}}$, we first need to evaluate the substituted
-expression with coeffects $\cclrd{s}$, followed by the rest of the expression with coeffects specified
-by $\aclrd{\textbf{t}}$.
-
-~
-
-~
-
-~
-
-\begin{lemma}[Multi-nary substitution]
-\label{thm:structural-substitution}
-Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
-%
-\begin{equation*}
-\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_1}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
-  \coctx{x_k\!:\!\tau_k}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
-\end{equation*}
-%
-and a expressions $e_i$ with free-variable contexts $\Gamma_i$ annotated with $\aclrd{\mathbf{t_i}}$:
-%
-\begin{equation*}
-\coctx{\Gamma_1}{\aclrd{\textbf{t}_1}} \vdash e_1 : \tau_1 
-\quad \ldots \quad
-\coctx{\Gamma_k}{\aclrd{\textbf{t}_k}} \vdash e_k : \tau_k
-\end{equation*}
-%
-then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
-where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
-%
-\begin{equation*}
-\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_1}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
-  \coctx{\Gamma_k}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
-\end{equation*}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}}} &=& \ctyp{\cclrd{r}}{(\sem{\tau_1} \times \ldots \times \sem{\tau_n})} \\[0.5em]
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} & = & \ctyp{\cclrd{r}}{\sem{\tau_1}} \rightarrow \sem{\tau_2} \\[-0.2em]
+\sem{\ident{num}} & = & \ident{num} \\
+\end{array}
+\end{equation*}
+
+Here, a context becomes a comonadically-inspired data type wrapping a tuple of variable values 
+and a coeffectful function is translated into an ordinary function in the target language
+with a comonadically-inspired data type wrapping the input type.
+
+The rules shown in Figure~\ref{fig:transl-translation} are very similar to those shown earlier
+Figure~X in Chapter~3. Thanks to the equivalence between $\lambda$-calculus and category theory,
+we now interpret them as denoting a translation. Furthermore, the program produced by the above
+rules is well-typed.
+
+\begin{lemma}[Well-typedness of the translation]
+Given a well-typed coeffect program $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ with a 
+typing derivation such that:
+%  
+\begin{equation*}  
+  \semdeff
+    {(\ldots)\qquad~}
+    {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
+    {(\ldots)}
+    { ~~f }
+\end{equation*}  
+%
+Then $e$ is a valid expression in the target langauge and $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
 \end{lemma}
 \begin{proof}
-By induction over $\vdash$, using the multi-nary aspect of the substitution 
-in the proof of the contraction case (see Appendix~\ref{sec:appendix-struct-cbn}).
+  By rule induction over the translation rules.
 \end{proof}
 
-\subsection{Reduction and expansion}
-
-In the Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
-proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
-$\beta$-reduction, which encompasses both call-by-value and call-by-name. We also show that 
-$\eta$-expansion preserves the types. Both of the properties hold for a system with any
-structural coeffect algebra.
-
-\paragraph{Reduction theorem.} In a full $\beta$-reduction,
-written as $\rightarrow_\beta$, we can replace the redex $(\lambda x.e_2)~e_1$ by the
-expression $\subst{e_r}{x}{e_s}$ anywhere inside a term. The subject reduction theorem
-guarantees that this does not change the type of the term.
+\begin{figure*}[t]
+
+\begin{equation*}
+\hspace{-3em}
+\begin{array}{ll}
+\hspace{4.8em}\semdef
+  {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
+  {\lambda\ctx.\pi_i~(\ident{counit}_{\cunit}~\ctx)} 
+& (\emph{var})
+\\[1em]
+\hspace{5.3em}\semdef
+  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
+  {\lambda\ctx.n} 
+& (\emph{const})
+\\[1.5em]
+\hspace{1.5em}\semdeff
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+  {f}
+  {\lambda \ctx.\lambda v.f~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v)) }
+& (\emph{abs})
+\\[2em]
+\semdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\begin{array}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\\[-0.3em]~\\[-0.3em]~\end{array}}
+  {f}
+  {g}
+  {\begin{array}{l}  
+  \lambda\ctx.~~\\[-0.3em]
+    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~(\lambda x.(x,x))~\ctx\\[-0.3em]
+    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}}~\ctx_0\\[-0.3em]
+    \quad f~\ctx_1~
+      (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g~\ctx_2)
+  \end{array} }   
+& (\emph{app})
+\\[4em]
+\hspace{6em}\semdeff
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } 
+  {f}
+  {\lambda\ctx f~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx}
+\quad(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
+& (\emph{sub})\\[-0.25em]
+\end{array}
+\end{equation*}
+
+\figcaption{Translation from a flat DSL to a comonadically-inspired target language}
+\label{fig:transl-translation}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Coeffect langauge for dataflow}
+
+The types of the comonadically-inspired operations are the same for each concrete coeffect DSL, but
+each DSL introduces its own \emph{values} of type $\ctyp{\cclrd{r}}\tau$ and also its own reduction
+rules that define how comonadically-inspired operations evaluate.
+
+We first consider the extensions needed for dataflow computations. As discussed earlier in the
+semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
+values.
+
+When evaluating translated programs, the comonadic values will not be directly manipulated by the
+user code and so we introduce a new class of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$. 
+Furthermore, the \ident{Df} values will only appear as fully evaluated inputs and they will be manipulated 
+using the comonadically-inspired operations. Thus, the typing rule (\emph{df}) only needs to 
+check values. We do not need a similar rule for expressions. Construct such as $\ident{Df}\langle(\kvd{fun}~x.x)~1\rangle$ 
+are not allowed in the target language. The (\emph{df}) rule also guarantees that the number of 
+elements in the list matches the number in the coeffect:
+
+\begin{equation*}
+\begin{array}{l}
+v = \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
+e = \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
+\end{array}  
+\begin{array}{l}
+\tyrule{df}
+  { \forall i\in\{ 0 \ldots n \}. ~ \vdash v : \tau}
+  { \Gamma \vdash \ident{Df}\langle v_0, \ldots, v_n \rangle : \ctyp{\cclrd{n}}{\tau} }
+\end{array}  
+\end{equation*}
+~
+
+\noindent
+The additional reduction rules mirror the semantics that we discussed earlier when talking about
+indexed dataflow comonad.
+
+\begin{equation*}
+\begin{array}{rl}
+{\footnotesize(\emph{counit})} & \begin{array}{l}
+\ident{counit}_{\cclrd{0}}(\ident{Df} \langle v_0 \rangle)   \rightarrow v_0
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{cobind})} & \begin{array}{l}
+\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f (\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
+\\[-0.25em]
+\quad (\ident{Df}\langle f (\ident{Df}\langle v_0, \ldots, v_m \rangle), \ldots, f (\ident{Df}\langle v_{n}, \ldots, v_{m+n} \rangle) \rangle)
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{merge})} & \begin{array}{l}
+\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle b_0, \ldots, b_n\rangle)) \rightarrow
+\\[-0.25em]
+\quad (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle)
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{split})} & \begin{array}{l}
+\ident{split}_{\cclrd{m}, \cclrd{n}} (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle)  \rightarrow
+\\[-0.25em]
+\quad \ident{Df}\langle v_0, \ldots, v_m \rangle, (\ident{Df}\langle b_0, \ldots, b_n\rangle
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{lift})} & \begin{array}{l}
+\ident{lift}_{\cclrd{n'}, \cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_{n'} \rangle) \rightarrow
+\qquad(\textnormal{when}~n\leq n') \\[-0.25em]
+\quad \ident{Df}\langle v_0, \ldots, v_n \rangle
+\end{array}
+\end{array}
+\end{equation*}
+
+~
+
+\noindent
+Now consider a target langauge consisting of the core (ML-subset) defined in Figure~\ref{fig:transl-target}
+with typing rules defined in Figure~\ref{fig:transl-targetty} and comonadically-inspired primtiives
+defined in Figure~\ref{fig:transl-ext} and also concrete notion of comonadically-inspired value
+and reduction rules for data flow as defined above.
 
 \begin{theorem}[Type preservation]
-In a structural coeffect system with a structural coeffect algebra formed by 
-$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
-if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
-and $e \rightarrow_{\beta} e'$ using the full $\beta$-reduction then
-$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
 \end{theorem}
 \begin{proof}
-Consider the typing derivation for the redex $(\lambda x.e_r)\,e_s$ before the reduction
-(note that this is similar to the typing of let binding discussed in Remark~\ref{thm:structural-letbinding},
-but we do not swap the two parts of the free-variable context):
-%
-\begin{equation*}
-\inference
-  { \begin{array}{l}\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s\vspace{-2em}\end{array} &
-    \inference
-      { \coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r }
-      { \coctx{\Gamma_r}{\aclrd{\textbf{r}}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{t}} \tau_r} \vspace{0.2em} }
-  { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})} \vdash (\lambda x.e_r)~e_s : \tau_r }
-\end{equation*}
-%
-For the substitution lemma, we first rewrite the typing judgement for $e_r$, \ie~
-$\coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r$
-as a context with a single hole filled by the $x$ variable:
-$\coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes\,-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r$.
-Now we can perform the substitution using Lemma~\ref{thm:structural-substitution}:
-%
-\begin{equation*}
-\inference
-  { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r & 
-    \coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s }
-  { \inference
-      { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{\Gamma_s}{\cclrd{t}\,\aseq\,\aclrd{\textbf{s}} }\,] \vdash \subst{e_r}{x}{e_s} : \tau_r }
-      { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}}\,\atimes\,(\cclrd{t}\,\aseq\,\aclrd{\textbf{s}})} \vdash \subst{e_r}{x}{e_s} : \tau_r } }
-\end{equation*}
-%
-The last step applies the hole filling operation, showing that substitution preserves 
-the type of the term.
+As before, using rule induction over $\rightarrow$.
+
+Interesting new cases are the reduction rules (\emph{counit}), (\emph{cobind}), (\emph{merge}),
+(\emph{split}) and (\emph{lift}). We need to show that, given a well typed input value, the resulting
+value is a \ident{Df} value with the right number of elements (for (\emph{cobind}), (\emph{lift}) and (\emph{merge}),
+a tuple of \ident{Df} values with the right number of elements (for (\emph{split})) and a value
+of correct type (for (\emph{unit})).
+
+From the fact that the reduction occurred, we know that the input was a \ident{Df} value with the
+required number of inputs and from the typing rule (\emph{df}) we know that the values in the input
+were all of correct types.  
 \end{proof}
 
-\noindent
-Because of the vector (free monoid) structure of coeffect annotations $\aclrd{\textbf{r}}$, 
-$\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$, these are uniquely associated with $\Gamma_r$, 
-$\Gamma_s$, and $x$ respectively. Therefore, substituting $e_s$ (which has coeffects 
-$\aclrd{\textbf{s}}$) for $x$ introduces the context requirements specified by $\aclrd{\textbf{s}}$
-which are composed with the requirements $\cclrd{t}$ associated with $x$, \ie~the variable being 
-substituted.
-
-\paragraph{Expansion theorem.}
-Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
-\emph{local soundness} and \emph{local completeness} as required by Pfenning and Davies
-\cite{logic-modal-reconstruction}. Informally, this means that abstraction does not introduce 
-too much, and application does not eliminate too much.
-
-\begin{theorem}[$\eta$-expansion]
-In a structural coeffect system with a structural coeffect algebra formed by 
-$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
-if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
-and $e \rightarrow_{\eta} e'$ using the full $\eta$-reduction then
-$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
+\begin{theorem}[Progress]
+  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
 \end{theorem}
 \begin{proof}
-The following derivation shows that $\lambda x.f~x$ has the same type and coeffects as 
-the original expression $f$:
-\begin{equation*}
-\inference
-  { \inference
-    { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
-      \coctx{x\!:\!\tau_1}{\alift{\cunit}} \vdash x:\tau_1 }
-    { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}} \;\atimes\;(\cclrd{s}\,\aseq\,\alift{\cunit}) } \vdash f~x : \tau_2 } }
-  { \inference
-    { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}} \;\atimes\,\alift{\cclrd{s}} } \vdash f~x : \tau_2 }
-    { \coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
-\end{equation*}
-%
-The second step uses the fact that 
-$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
-arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
+  As before, using rule induction over $\vdash$.
+
+Interesting new cases are those arising from the typing rules for the comonadically-inspired
+primitives in Figure~\ref{fig:transl-ext}, that is (\emph{counit}), (\emph{cobind}), (\emph{merge}),
+(\emph{split}) and (\emph{lift}).  
+In all of the cases, $e$ is not a value and so it needs to reduce and the only reduction rules
+are the ones for dataflow computations (above).
+
+If the argument is a value, one of the reduction rules (above) can be used (the typing guarantees
+that the list has the required number of elements). If the argument is an expressions, we use 
+the induction hypothesis.
 \end{proof}
 
-\noindent
-The $\eta$-expansion property discussed in this section highlights another difference 
-between coeffects and effects. The $\eta$-equality property does not hold for many notions 
-of effect. For example, in a language with output effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ 
-has different effects to its $\eta$-converted form $\lambda x . e\,x$ because the immediate 
-effects of $e$ are hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, 
-the (\emph{abs}) rule allows immediate contextual requirements of $e$ to ``float outside'' 
-of the enclosing $\lambda$. Furthermore, the free monoid nature of $\atimes$ in structural 
-coeffect systems allows the exact immediate requirements of $\lambda x . e x$ to match 
-those of $e$.
-
-% ==================================================================================================
-%
-%     ###                         ##                    #                 
-%    #   #                         #                                      
-%    #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##  
-%    #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  # 
-%    #      #   #  #   #  #        #    #   #   ###     #    #   #  #   # 
-%    #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   # 
-%     ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   # 
-%
-% ==================================================================================================
-
-\section{Conclusions}
-
-This chapter completes the key development of this thesis -- the presentation of the coeffect 
-framework, consisting of two calculi capturing properties of context-aware computations introduced
-in Chapter~\ref{ch:applications}. In the previous chapter, we focused on the whole-context 
-properties of computations and we developed \emph{flat coeffect calculus} to capture them.
-This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
-contextual properties. The system provides a precise analysis of liveness and data-flow and allows 
-other interesting uses such as tracking of variable accesses based on bounded linear logic.
-
-Following the structure of the previous chapter, the structural coeffect calculus is parameterized
-by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations 
-$\cseq$ and $\cpar$ that model sequential and pointwise composition, respectively. For flat 
-coeffects, we required $\czip$ to model context merging. For structural coeffects, we 
-instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
-as $\czip$.
-
-In order to keep track of separate annotations for each variable, we use a system with 
-explicit structural rules (contraction, weakening and exchange) that manipulate the structure
-of variables and the structure of annotations at the same time.
-
-The chapter presents two technical results. The first is the semantics of structural coeffect
-calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
-it unifies our three examples (bounded reuse, data-flow and liveness) and it demonstrates
-suitability of the type system. The second technical result is equational theory for the structural
-coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
-typing preservation property for any instance of the structural coeffect calculus. These two 
-strong properties are desirable for programming languages, but are often not satisfied 
-(\eg~by languages with effects).
-
-Two questions remain open for the upcoming chapter. Firstly, is there a way to present 
-flat and structural coeffect calculi in a uniform way as a single system? Secondly, what
-is the best pathway towards a practical implementation of programming languages based on 
-coeffect systems?
+\newpage
+\section{Coeffect language for implicit parameters}
+
+\begin{equation*}
+v = \ident{Impl}(v)
+\end{equation*}
+
+a
+
+\begin{equation*}
+\tyrule{impl}
+  { \Gamma \vdash v : \ident{string} \rightarrow \ident{num}}
+  { a }
+\end{equation*}
+
+b
+
+% Simple for normal parts of the language
+% We have to check this for dataflow reductions  
+%
+% Interesting case is $e_1 e_2$ where $e_1 = v$ is one of the primitives. Then we have to go through
+% them and make sure they can progress.
+
+% And also syntactic sugar \ident{map}
+
+%    &\sep& \ident{cobind}_{\cclrd{s},\cclrd{r}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~e 
+%       \sep \ident{split}_{\cclrd{r},\cclrd{s}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \\
+
+%
+% &\sep& \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e 
+%   \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~e \\
+
+\section{Dataflow}
+Domain specific coeffect annotations are non-negative integers, type is a tuple and new kinds
+of values are lists:
+\begin{equation*}
+\begin{array}{rcl}
+  v_c &  = &  \langle v_1, \ldots, v_n \rangle
+\end{array}
+\end{equation*}
+%
+With the following typing:
+%
+\begin{equation*}  
+\tyrule{val}
+  { \forall v_i . \vdash v_i : \tau }
+  { \vdash \langle v_1, \ldots, v_n \rangle : \ctyp{n}{\tau} }
+\end{equation*}
+%
+
+We could define those in terms of head/tail functions and then prove that they work (and translated
+program does not contain head/tail), but perhaps that's unnecessary overkill. Or perhaps it would
+be useful in order to show how we avoid errors... But doing it directly looks just simpler (we
+also do not need a ``list'' type).
+
+\begin{theorem}[Type preservation]
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
+\end{theorem}
+\begin{proof}
+  Rule induction over $\rightarrow$
+
+Simple for normal parts of the language
+
+We have to check this for dataflow reductions  
+\end{proof}
+
+\begin{theorem}[Progress]
+  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+\end{theorem}
+\begin{proof}
+  By rule induction over $\vdash$.
+
+Interesting case is $e_1 e_2$ where $e_1 = v$ is one of the primitives. Then we have to go through
+them and make sure they can progress.  
+\end{proof}
+
+\end{document}

### 2016-02-03 main.tex
Reorg chapter 4 and 5, adding unique typing to chapter 4

@@
 \usepackage{amsthm}
-\usepackage{catchfilebetweentags}
+%\usepackage{catchfilebetweentags}
 
@@
 \newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}
+
+% Red color
+\definecolor{rclr}{rgb}{0.75,0.0,0.0}
+\newcommand{\rclrd}[1]{\textcolor{rclr}{#1}}
 
@@
 }
+
+% For definining semantics of things over a typing derivation
+\newcommand{\semdeff}[4]{
+\begin{array}{rcl}
+  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
+  #2 & = & #4 \\
+\end{array}  
+}
+\newcommand{\semdefff}[6]{
+\begin{array}{rcl}
+  #1 & = & #4 \\
+  #2 & = & #5 \\ \cline{1-1}\cline{3-3}
+  #3 & = & #6 \\
+\end{array}  
+}
+\newcommand{\semdef}[2]{
+\begin{array}{rcl}
+  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
+  #1 & = & #2 \\
+\end{array}  
+}
+
 % Half line spacing around align
@@
 
-\include{text/Titlepage} % Main title page
+%\include{text/Titlepage} % Main title page
 
-% \include{FrontBackMatter/Titleback} % Back of the title page
+%\cleardoublepage\include{text/Declaration} % Publications from the thesis page
 
-% \cleardoublepage\include{FrontBackMatter/Dedication} % Dedication page
-
-%\cleardoublepage\include{FrontBackMatter/Foreword} % Uncomment and create a Foreword.tex to include a foreword
-
-\cleardoublepage\include{text/Declaration} % Publications from the thesis page
-
-\cleardoublepage\include{text/Abstract} % Abstract page
-
-% \cleardoublepage\include{FrontBackMatter/Acknowledgments} % Acknowledgements page
+%\cleardoublepage\include{text/Abstract} % Abstract page
 
@@
 
-\include{text/Chapter01}
-\include{text/Chapter02}
-\include{text/Chapter03}
+%\include{text/Chapter01}
+%\include{text/Chapter02}
+%\include{text/Chapter03}
+
+\ctparttext{This part presents the key novel contributions of the thesis. We develop \emph{flat}
+  and \emph{structural} coeffect calculi that capture a wide range of context-aware programming
+  languages. We discuss their type systems and semantics together with safety properties.}
+\setcounter{part}{1}
+\part{Coeffect calculi}
+
 \include{text/Chapter04}
 \include{text/Chapter05}
-\include{text/Chapter06}
-\include{text/Chapter07}
+%\include{text/Chapter06}
+%\include{text/Chapter07}
+%\include{text/Chapter08}
 
@@
 
-\appendix
-\include{text/AppendixA}
-\include{text/AppendixB}
+%\appendix
+%\include{text/AppendixA}
+%\include{text/AppendixB}
 

### 2016-02-03 text/Chapter04.tex
Reorg chapter 4 and 5, adding unique typing to chapter 4

@@
-%!TEX root = ../main.tex
-
-%\begin{flushright}{\slshape    
-%Even the most puritanical rationalist will be forced to stop arguing and \\
-%use propaganda [...] because the psychological conditions have disappeared \\
-%that allowed effective argument and therefore influence over the others.} \\ \medskip
-%--- Paul Feyerabend, \emph{Against Method} \cite{philosophy-feyerabend}
-%\end{flushright}
-%\vspace{2em}
-
 % ==================================================================================================
 
-\chapter{The flat coeffect calculus} 
+\setcounter{chapter}{3}
+\chapter{Types for flat coeffects} 
 \label{ch:flat}
@@
 \emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing 
-per-variable properties. As we show in Chapter~\ref{ch:unified}, the systems can be unified 
-using a single abstraction. This is useful when implementing and composing the systems, but such 
-abstraction is  \emph{less powerful} -- \ie~its generality hides useful properties that we can see 
-when we consider the systems separately. For this reason, we discuss \emph{flat coeffects} and
-\emph{structural coeffects} separately in this and the next chapter.
+per-variable properties. As we show in Section~X, the systems can be further unified using a 
+single abstraction, but such abstraction is  \emph{less powerful} -- \ie~its generality hides 
+useful properties that we can see when we consider the systems separately. For this reason, we 
+discuss \emph{flat coeffects} and \emph{structural coeffects} separately.
+
+In this and the next chapter, we discuss the type system and the semantics of flat coeffect 
+systems, respectively. In this chapter, we develop a parameterized type system for flat coeffect
+systems and we study its general syntactic properties. We also consider variations of the type system
+that resolve the ambiguity of coeffectful lambda abstraction for concrete instances of the system.
+In the next chapter, we give operational meaning of concrete coeffect languages using the unified
+system and we discuss their safety.
 
@@
   instantiated to obtain three of the systems discussed in Section~\ref{sec:applications-flat},
-  namely implicit parameters, liveness and data-flow.
-
-\item We give the equational theory of the calculus and discuss type-pre\-ser\-vation for call-by-name
+  namely implicit parameters, liveness and dataflow.
+
+\item The type system permits multiple typing derivation for certain programs due to the ambiguity
+  inherent in conextual lambda abstraction rule. In Section~\ref{sec:flat-unique}, we discuss
+  variations of the type system that resolve the ambiguity and give a unique typing derivation
+  for the three coeffect systems covered in this chapter.
+
+\item We discuss syntactic properties of the calculus, covering type-pre\-ser\-vation for call-by-name
   and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
@@
   of the flat coeffect algebra.
-
-\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
-  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
-  This provides a deeper insight into how (and why) the calculus works and shows an 
-  intriguing link with effects.
 \end{itemize}
@@
 %
-%     ###           #                       #                 #       #                 
-%      #            #                       #                 #                         
-%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
-%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
-%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
-%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
-%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
+%    ###
+%     #  #    # ##### #####   ####  #####  #    #  ####  ##### #  ####  #    #
+%     #  ##   #   #   #    # #    # #    # #    # #    #   #   # #    # ##   #
+%     #  # #  #   #   #    # #    # #    # #    # #        #   # #    # # #  #
+%     #  #  # #   #   #####  #    # #    # #    # #        #   # #    # #  # #
+%     #  #   ##   #   #   #  #    # #    # #    # #    #   #   # #    # #   ##
+%    ### #    #   #   #    #  ####  #####   ####   ####    #   #  ####  #    #
 %
@@
 
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Related work}
-
 The development in this chapter can be seen as a counterpart to the well-known development of 
-\emph{effect systems} \cite{effects-gifford} and the use of \emph{monads} \cite{monad-notions}
-in programming languages. The syntax and type system of the flat coeffect calculus follows 
+\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:transl} then links \emph{coeffects} 
+with \emph{comonads} in the same way in which effect systems can be linked with monads
+\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows 
 a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
@@
 
-Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
-systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
-and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
-relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
-$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
-\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
-defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
-of the work on effect systems and monads.
-
-The main purpose of the comonadic semantics presented in this chapter is to provide a motivation 
-for the typing of the flat coeffect calculus. Our approach is inspired by the work of Uustalu and
-Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
-data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
-comonads} annotate the structure with information about the required context, \ie~$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
-This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.
-
 % ==================================================================================================
 %                                                            
-%       ###           ##                   ##                 
-%      #   #           #                    #                 
-%      #       ###     #     ###   #   #    #    #   #   ###  
-%      #          #    #    #   #  #   #    #    #   #  #     
-%      #       ####    #    #      #   #    #    #   #   ###  
-%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
-%       ###    ####   ###    ###    ## #   ###    ## #  ####  
+%     #####
+%    #     #   ##   #       ####  #    # #      #    #  ####
+%    #        #  #  #      #    # #    # #      #    # #
+%    #       #    # #      #      #    # #      #    #  ####
+%    #       ###### #      #      #    # #      #    #      #
+%    #     # #    # #      #    # #    # #      #    # #    #
+%     #####  #    # ######  ####   ####  ######  ####   ####
 %                                                           
@@
 sets of implicit parameters, versions represented as integers or other values. The expressions of 
-the calculus are those of the $\lambda$-calculus with \emph{let} binding; assuming $\alpha$ ranges 
-over base types, the types of the calculus are defined as:
+the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type 
+\ident{num} as an example of a concrete base type with numerical constants written as $n$:
 %
@@
 \begin{array}{rcl}
-e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
-\tau &::=& \alpha \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
 \end{array}
@@
 %
+Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
+annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a 
+unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
+we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations} 
+rather than solving the problem of type reconstruction.
+
 We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
@@
 It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
-context $r$.
-
-In the categorical semantics, the function $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is modelled
-by a morphism $C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the object $C^{\cclrd{r}}$
-does not itself exist as a syntactical value. This is because we use comonads to define the 
-\emph{semantics} rather than \emph{embedding} them into the language as in the meta-language
-approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}). The annotations
-$\cclrd{r}$ are formed by an algebraic structure discussed next.
+context $\cclrd{r}$.
+
+In the categorically-inspired translation in the next chapter, the function 
+$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function 
+$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
+does not itself exist as a syntactical value in the coeffect calculus. This is because we use 
+comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in 
+the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
+The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.
 
@@
 \tyrule{param}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{r} \cup \aclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{s}} \tau_2 }
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
 \;
 \tyrule{df1}
-  {\coctx{\Gamma, x:\tau_1}{\aclrd{n}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{n}} \tau_2 }
-\end{equation*}
-
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{n}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 }
+\end{equation*}
+%
 In order to capture both systems using a single calculus, we need a way of unifying the two
@@
 \tyrule{df2}
-  {\coctx{\Gamma, x:\tau_1}{\textnormal{min}(\aclrd{n}, \aclrd{m})} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\aclrd{m}} \tau_2 }
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{\textnormal{min}(n, m)}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 }
 \end{equation*}
@@
 annotations of contexts), then the reverse is also true -- because 
-$\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{m}$ and $\textit{min}(\aclrd{n}, \aclrd{m}) \leq \aclrd{n}$.
+$\cclrd{\textit{min}(n, m)} \leq \cclrd{m}$ and $\cclrd{\textit{min}(n, m)} \leq \cclrd{n}$.
 In other words (\emph{df1}) is more precise, but (\emph{df2}) gives a sound over-approximation
 with a structure that can be unified with (\emph{param}).
+
+Using a rule such as (\emph{df2}) allows us to give a unified formulation of the flat coeffect 
+calculus in Section~\ref{sec:flat-calculus-types}, however the coeffect-specific handling of
+lambda abstraction remains important in practical implementation in order to obtain a unique
+typing derivation for each coeffect program as discussed in Section~\ref{sec:flat-unique}
+(and in the implementation in Chatper~\ref{ch:impl}).
 
@@
 exactly the opposite things. To disambiguate, we always use the term \emph{context requirements} 
-when using the syntactic view, especially in the rest of Section~\ref{sec:flat-calculus}, and 
+when using the syntactic view, especially in the rest of Chapter~\ref{chf:flat}, and 
 \emph{context capabilities} or just \emph{available context} when using the semantic view, 
-especially in Section~\ref{sec:flat-semantics}.
+especially in Chapter~\ref{ch:transl}.
 
@@
 $\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
-idempotent monoid, $(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup). 
-That is, for all $r,s,t\in \C$: 
+idempotent monoid and $(\C, \czip)$ is a band (idempotent semigroup). That is, for all $r,s,t\in \C$: 
 %
@@
 r\; \czip\; r = r \qquad
-\\
-\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
-t\; \cleq\; t \\[0.5em]
 \end{array}
@@
 are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
-(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.
+(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}).
 
@@
 
-\paragraph{Ordering.}
-
-The flat coeffect algebra requires a pre-order relation $\cleq$, which is used to define 
-sub-coeffecting rule of the type system. When the idempotent monoid $(\C, \cpar, \czero)$ also 
-has the commutative property (\ie~forms a semi-lattice), the $\cleq$ relation can be defined as the
-ordering of the semi-lattice:
-%
-\begin{equation*}
-\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
-\end{equation*}
-%
-This definition is consistent with all three examples that motivate flat coeffect calculus, but
-it cannot be used with the structural coeffects (where it fails for the bounded reuse 
-calculus) and so we choose not to use it.
-
-Furthermore, the $\cunit$ coeffect is often the top or the bottom element of the semi-lattice. 
-As discussed in Section~\ref{sec:flat-syntax}, when this is the case, we are able to prove certain 
-syntactic properties of the calculus.
-
 %---------------------------------------------------------------------------------------------------
@@
 \tyrule{const}
-  {c : \tau \in \Delta}
-  {\coctx{\Gamma}{\czero} \vdash c : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
+  {~}
+  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
 \end{equation*}
@@
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
 \end{equation*}
@@
   { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
+    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
   {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
@@
 (\emph{var}) and constants (\emph{const}) are annotated with special values provided by the 
-coeffect algebra. Following the top-down syntactic reading, the (\emph{sub}) rule allows us to 
-treat an expression with fewer context requirements as an expression with more context requirements.
+coeffect algebra.
 
@@
 typing derivation for $(\lambda x.e_2)~e_1$, but it corresponds to \emph{one} possible typing 
-derivation. As we show in \ref{sec:flat-calculus-let}, the typing used in (\emph{let}) is more
+derivation. As we show in \ref{sec:flat-exts-let}, the typing used in (\emph{let}) is more
 precise than the general rule that can be derived from $(\lambda x.e_2)~e_1$. Additional 
-constructs such as pairs are covered in Section~\ref{sec:flat-exts}.
+constructs such as pairs, sub-coeffecting and sub-typing are covered in Section~\ref{sec:flat-exts}.
 
@@
 \noindent
-For simplicity, we assume that all parameters have the same type $\rho$ and so the annotations only
-track sets of names. The definition uses set union for all three operations. Both variables and
+For simplicity, we assume that all parameters have the same type $\ident{num}$ and so the annotations only
+track sets of names. The definition uses a set union for all three operations. Both variables and
 constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The 
 definition satisfies the flat coeffect algebra laws because $(S, \cup, \emptyset)$ is an idempotent, 
-commutative monoid. The system has a single additional typing rule for accessing the value of a
-parameter:
-%
+commutative monoid. The language has additional syntax for defining an implicit parameter and for
+accessing it, together with associated typing rules:
+%
+\begin{equation*}
+\quad~ e ~::=~ \ldots \sep \ident{?p} \sep \kvd{let}~\ident{?p}~=~e_1~\kvd{in}~e_2
+\end{equation*}
 \begin{equation*}  
 \tyrule{param}
-  { \ident{?p} \in \cclrd{c} }
-  { \coctx{\Gamma}{\cclrd{c}} \vdash \ident{?p} : \rho }
-\end{equation*}
-%
-The rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of required
-parameters $\cclrd{c}$. As discussed earlier, we use the same type $\rho$ for all parameters, but
-it is easy to define an extension that uses mappings from names to types.
+  { ~ }
+  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
+\end{equation*}
+\begin{equation*}
+\tyrule{letpar}
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} } \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+%
+The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of required
+parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all parameters, but
+it is possible to define a coeffect calculus that uses mappings from names to types (care is needed to 
+avoid assigning multiple types to a parameter of the same type).
+
+The (\emph{letpar}) rule is the same as the one discussed in Section~\ref{sec:applications-flat-impl}.
+As both of the rules are specific to implicit parameters, we write the operations on coeffects directly
+using set operations -- coeffect-specific operations such as set subtraction are not a part of the
+unified coeffect algebra.
 
@@
 \noindent
+The liveness example is interesting because it does not require any additional syntactic extensions
+to the langauge. It annotates constants and variables with $\ident{D}$ and $\ident{L}$, respectively
+and it captures how those annotation propagate through the remaining language constructs.
+
 As in Section~\ref{sec:applications-flat-live}, sequential composition $\cseq$ is modelled by 
@@
 $\cseq$ is not idempotent. The distributivity laws require the following to be the case:
-$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see. Finally, a simple 
-data-flow language includes an additional rule for $\kvd{prev}$:
-%
+$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see.
+
+A simple dataflow langauge includes an additional construct $\kvd{prev}$ for accessing the 
+previous value in a stream with an additional typing rule that look as follows:
+%
+\begin{equation*}
+\quad~ e ~::=~ \ldots \sep \kvd{next}~e
+\end{equation*}
 \begin{equation*}
 \tyrule{prev}
-  { \coctx{\Gamma}{\cclrd{c}} \vdash e : \tau }
-  { \coctx{\Gamma}{\cclrd{c}+1} \vdash \kvd{prev}~e : \tau }
+  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
+  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
 \end{equation*}
@@
 
+% ==================================================================================================
+%
+%       #
+%      # #   #    # #####   ####  # #    # # ##### #   #
+%     #   #  ##  ## #    # #    # # #    # #   #    # #
+%    #     # # ## # #####  #      # #    # #   #     #
+%    ####### #    # #    # #  ### # #    # #   #     #
+%    #     # #    # #    # #    # # #    # #   #     #
+%    #     # #    # #####   ####  #  ####  #   #     #
+%
+% ==================================================================================================
+
+\section{Choosing unique typing}
+\label{sec:flat-unique}
+
+As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
+differs from the rule for effect systems in that it does not delay all context requirements. 
+In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the requirements
+can be satisfied either by the call-site or by the declaration-site. In case of dataflow and
+liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
+because it allows multiple valid typing derivations.
+
+Furthermore, the semantics of context-aware languages in Chapter~\ref{sec:applications} and
+also in Chapter~\ref{sec:transl} is defined over \emph{typing derivation} and so the same program
+could have a different meaning, depending on the typing derivation chosen. In this section, we
+specify how to choose \emph{unique} typing derivation in each of the coeffect systems we consider.
+
+The most interesting case is that of implicit parameters. For example, consider the following 
+program written using the coeffect calculus with implicit parameter extensions:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~f~= (\kvd{let}~\ident{?x}=1~\kvd{in}~(\lambda y.~\ident{?x}))~\kvd{in} \\[-0.2em]
+\kvd{let}~\ident{?x}=2~\kvd{in}~f~0 \\[-0.2em]
+\end{array}  
+\end{equation*}
+%
+There are two possible typings allowed by the typing rules discussed in Section~\ref{sec:flat-calculus-types}
+that lead to two possible meanings of the program -- evaluating to $1$ and $2$, respectively:
+%
+\begin{itemize}
+  \item $f : \ident{num} \xrightarrow{\cclrd{\emptyset}} \ident{num}$ -- in this case, the value
+    of \ident{?x} is captured from the declaration-site and the program produces $1$.
+  \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
+    parameter \ident{?x} is required from the call-site and the program produces $2$.
+\end{itemize}
+%
+The coeffect calculus intentionally allows both of the options, acknowledging the fact that the
+choice needs to be made for each individual concrete context-aware programming language. In this
+section, we discuss the choices for implicit parameters, dataflow and liveness.
+
+In this section, we use the fact that the coeffect calculus uses Church-style syntax for lambda
+abstraction and has a type annotation for the type of the variable. This does not affect the 
+handling of coeffects (those are not defined by the type annotation), but it lets us prove 
+uniqueness typing property of the specialized coeffect type systems. This shows that we define a 
+\emph{unique} way of assigning coeffects to otherwise well-typed programs.
+
 %---------------------------------------------------------------------------------------------------
 
-\subsection{Typing of let binding}
-\label{sec:flat-calculus-let}
-
-Recall the (\emph{let}) rule in Figure~\ref{fig:flat-types}. It annotates the expression 
-$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context requirements $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
-This is a special case of typing an expression $(\lambda x.e_2)~e_1$, using the idempotence
-of $\czip$ as follows:
-%
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cunit} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {~}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\czero} \vdash n : \ident{num} }
+\end{equation*}
 \begin{equation*}
 \tyrule{app}
-  {\begin{array}{l}
-   \vspace{-1.5em}
-   \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1
-   \end{array} &
-   \tyruler{abs}
-       { \coctx{\Gamma, x:\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2 }
-       { \coctx{\Gamma}{\cclrd{s}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
-  { \coctx{\Gamma}{\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }    
-\end{equation*}
-%
-This design decision is similar to ML value restriction, but it works the other way round. Our
-\emph{let} binding is more restrictive than the typing of abstraction-application, rather than
-being more general. The choice is motivated by the fact that the typing obtained using the special 
-rule for let-binding is more precise (with respect to sub-coeffecting) for all the examples consider 
-in this chapter. Table~\ref{tab:flat-simplelet} shows how the coeffect annotations are simplified 
-for our examples.
-
-\begin{table}[!h]
-\begin{center}
-\begin{tabular}{ | l | c | c |}
-\hline
-& \textbf{\footnotesize Definition\hspace{1em}} & \textbf{\footnotesize Simplified\hspace{1em}} \\ \hline
-\hspace{-1em}{\footnotesize Implicit parameters} & $\cclrd{s} \cup (\cclrd{s} \cup \cclrd{r})$ & $\cclrd{s} \cup \cclrd{r}$ \\ \hline
-\hspace{-1em}{\footnotesize Liveness} & $\cclrd{s} \sqcap (\cclrd{s} \sqcup \cclrd{r})$ & $\cclrd{s}$ \\ \hline
-\hspace{-1em}{\footnotesize Data-flow} & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
-\end{tabular}
-\end{center}
-
-\vspace{-0.5em}
-\figcaption{Simplified annotation for let binding in sample flat calculi instances}
-\label{tab:flat-simplelet}
-\end{table}
-
-\noindent
-The simplified annotations directly follow from the definitions of particular flat coeffect 
-algebras. It is perhaps somewhat unexpected that the annotation can be simplified in different
-ways for different examples.
-
-To see that the simplified annotations are more precise, assume that we used arbitrary 
-splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$ rather than idempotence. The
-``Definition'' column would use $\cclrd{s_1}$ and $\cclrd{s_2}$ for the first and second 
-$\cclrd{s}$, respectively. The corresponding simplified annotation would have 
-$\cclrd{s_1}\,\czip\,\cclrd{s_2}$ instead of $\cclrd{s}$. For all our systems, the 
-simplified annotation (on the right) is more precise than the original (on the left):
-%
-\begin{equation*}
-\begin{array}{rclll}
-\cclrd{s_1} \cup (\cclrd{s_2} \cup \cclrd{r}) &\supseteq& (\cclrd{s_1} \cup \cclrd{s_2}) \cup \cclrd{r} 
-  && \textnormal{(implicit parameters)}\\
-\cclrd{s_1} \sqcap (\cclrd{s_2} \sqcup \cclrd{r}) &\sqsupseteq&  (\cclrd{s_1} \sqcap \cclrd{s_2}) 
-  && \textnormal{(liveness)} \\
-\mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r}) &\geq& \mathit{min}(\cclrd{s_1}, \cclrd{s_2}) + \cclrd{r} 
-  &\quad& \textnormal{(data-flow)} \\
-\end{array}
-\end{equation*}
-%
-In other words, the inequality states that using idempotence, we get a more precise typing.
-This can be expressed using the abstract operators of the flat coeffect algebra as:
-%
-\begin{equation*}
-\cclrd{s_1} \;\cpar\; (\cclrd{s_2} \,\cseq\, \cclrd{r}) \;\cgeq\; 
-  (\cclrd{s_1} \,\czip\, \cclrd{s_2}) \;\cpar\; ((\cclrd{s_1} \,\czip\, \cclrd{s_2}) \,\cseq\, \cclrd{r})
-\end{equation*}
-%
-This property cannot be proved from other properties of the flat coeffect algebra. To make
-the flat coeffect system as general as possible, we do not \emph{in general} require it as
-an additional axiom, although the above examples provide reasonable basis for requiring 
-that the specialized annotation for let binding is the least possible annotation for the 
-expression $(\lambda x.e_2)~e_1$.
-
-% ==================================================================================================
-%                                              
-%      #####  #                                 
-%        #    #                                 
-%        #    # ##    ###    ###   # ##   #   # 
-%        #    ##  #  #   #  #   #  ##  #  #   # 
-%        #    #   #  #####  #   #  #      #  ## 
-%        #    #   #  #      #   #  #       ## # 
-%        #    #   #   ###    ###   #          # 
-%                                         #   # 
-%                                          ###                                                                         
-% ==================================================================================================
-
-\section{Categorical motivation}
-\label{sec:flat-semantics}
-
-The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
-discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical 
-semantics of context-dependent computations. This is a direction that we explore in this section. 
-Although the development presented here is interesting in its own, our main focus is \emph{using} 
-categorical semantics to motivate and explain the design of flat coeffect calculus.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Categorical semantics}
-
-As discussed in Section~\ref{sec:path-sem}, categorical semantics interprets terms as morphisms
-in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets typing 
-judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
-$\sem{\tau_1 \times \ldots \times \tau_n} \rightarrow \sem{\tau}$.
-
-As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful 
-computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that 
-approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
-for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models 
-partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
-monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$ 
-(state monad). Here, the structure of a strong monad provides necessary ``plumbing'' for composing 
-monadic computations -- sequential composition and strength for lifting free variables into the
-body of computation under a lambda abstraction.
-
-Following similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that 
-(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent 
-computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
-are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
-
-The monadic and comonadic model outlined here represents at most a binary analysis of effects or 
-context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no 
-context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
-$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an 
-over-simplification as we can use \eg~stacks of monad transformers and model functions with
-two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers 
-require defining complex system of lifting to be composable. Consequently, they are usually used
-for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
-capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
-stateful computation).}.
-
-In the next section, we introduce \emph{indexed comonads}, which provide a more precise analysis 
-and let us model computations with context requirements $\cclrd{r}$ as functions 
-$\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Introducing comonads}
-
-In category theory, \emph{comonad} is a dual of \emph{monad}. As already outlined in 
-Section~\ref{sec:path-sem}, we get a definition of a comonad by taking a definition of a monad 
-and ``reversing the arrows''. More formally, one of the equivalent definitions of comonad 
-looks as follows:
-
-\begin{definition}
-A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
-\begin{compactitem}
-\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
-\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
-\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
-  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
-\end{compactitem}
-such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
-\begin{align}
-\tag{\emph{left identity}}
-  \ident{cobind}~\ident{counit} &= \idf{}
-  \\
-\tag{\emph{right identity}}
-  \ident{counit} \circ \ident{cobind}~f &= f
-  \\
-\tag{\emph{associativity}}
-  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
-\end{align}
-\end{definition}
-
-\noindent
-From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
-\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries 
-additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function 
-$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
-the context-dependent function to value(s) in the context and then propagates the context.
-
-As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
-computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.
-
-\begin{example}[Non-empty list]
-A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
-We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The 
-type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
-%
-\begin{equation*}
-\begin{array}{rclll}
-\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
-\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
-\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
-\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The \ident{counit} operation returns the head of the non-empty list. Note that it is crucial that
-the list is \emph{non-empty}, because we always need to be able to obtain a value. The \ident{cobind}
-operation defined here returns a list of the same length as the original where, for each element, the 
-function $f$ is applied on a \emph{suffix} list starting from the element. Using a simplified
-notation for list, the result of applying \ident{cobind} to a function that sums elements of a
-list gives the following behaviour:
-%
-\begin{equation*}
-\ident{cobind}~\ident{sum}~(7,6,5,4,3,2,1,0) = (28,21,15,10,6,3,1,0)
-\end{equation*}
-%
-The fact that the function $f$ is applied to a \emph{suffix} is important in order to satisfy the
-\emph{left identity} law, which requires that $\ident{cobind}~\ident{counit}~l = l$.
-
-It is also interesting to examine some data types that do \emph{not} form a comonad. As already
-mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
-because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. The \ident{Maybe} 
-data type defined as $1 + \alpha$ is not a comonad for the same reason. However, if we consider 
-flat coeffect calculus for liveness, it appears natural to model computations as functions 
-$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalise comonads 
-to \emph{indexed comonads}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Generalising to indexed comonads}
-\label{sec:flat-semantics-idx}
-
-The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
-sequential composition, where the annotation $\cunit$ represents a variable access. An indexed 
-comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the annotation 
-$\cclrd{r}$ determines what context is required.
-
-\begin{definition}
-Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
-\emph{indexed comonad} over a category $\catc$ is a triple 
-$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where:
-
-\begin{compactitem}
-\item $\ctyp{\cclrd{r}}{}$ for all $\cclrd{r} \in \C$ is a family of object mappings 
-\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\cunit}{\alpha} \rightarrow \alpha$ 
-\item $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ is a mapping $(\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
-  \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})$
-\end{compactitem}
-such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$:f
-\begin{align}
-\tag{\emph{left identity}}
-  \ident{cobind}_{\cunit, \cclrd{s}}~\ident{counit}_{\cunit} &= \idf{}
-  \\
-\tag{\emph{right identity}}
-  \ident{counit}_{\cunit} \circ \ident{cobind}_{\cclrd{r}, \cunit}~f &= f
-  \\
-\tag{\emph{associativity}}
-\hspace{-10em}
-  \ident{cobind}_{\cclrd{r}\cseq\cclrd{s},\cclrd{t}}~(g \circ \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f) &= 
-    (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g) \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}~f)
-\end{align}
-\end{definition}
-
-\noindent
-Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings 
-$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
-operation is now formed by a \emph{family} of mappings for different pairs of indices 
-$\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations 
-and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
-For the purpose of this thesis, it is sufficient to omit the superscripts and treat $\ident{cobind}$ just 
-as a family of mappings (rather than natural transformations). When this does not introduce ambiguity, 
-we also occasionally omit the subscripts.
-
-The $\ident{counit}$ operation is not defined for all $\cclrd{r} \in \C$, but only for 
-the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
-but this is merely for symmetry and as a useful reminder to the reader. Crucially, this means that 
-the operation is defined only for special contexts.
-
-If we look at the indices in the laws, we can see that the left and right identity 
-require $\cunit$ to be the unit of $\cseq$. Similarly, the associativity law implies the 
-associativity of the $\cseq$ operator.
-
-\paragraph{Composition.}
-The co-Kleisli category that models sequential composition is formed by the unit arrow (provided
-by $\ident{counit}$) together with the (associative) composition operation that composes computations 
-with contextual requirements as follows:
-%
-\begin{equation*}
-\begin{array}{ccl}
-\textnormal{--}\, \hat{\circ} \,\textnormal{--}&\narrow{:}& (\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2) 
-  \rightarrow (\ctyp{\cclrd{s}}{\tau_2} \rightarrow \tau_3) 
-  \rightarrow (\ctyp{\cclrd{r} \cseq \cclrd{s}}{\tau_1} \rightarrow \tau_3) \\
-g \, \hat{\circ} \, f &\narrow{=}& g \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}} f)
-\end{array}
-\end{equation*}
-%
-The composition $\hat{\circ}$ best expresses the intention of indexed comonads. Given two functions
-with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
-requires $\cclrd{r}\,\cseq\,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
-are attached to the input of the composed function.
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Examples.}
-
-Any comonad can be turned into an indexed comonad using a trivial monoid. However, indexed comonads
-are more general and can be used with other data types, including indexed \ident{Maybe}.
-
-\begin{example}[Comonads]
-Any comonad $\ctyp{}{}$ is an indexed comonad with an index provided by a trivial monoid $(\{1\},\ast,1)$
-where $1\ast 1 = 1$. The mapping $\ctyp{1}{}$ is the mapping $\ctyp{}{}$ of the underlying comonad. The
-operations $\ident{counit}_1$ and $\ident{cobind}_{1,1}$ are defined by the operations $\ident{counit}$
-and $\ident{cobind}$ of the comonad.
-\end{example}
-
-\begin{example}[Indexed option]
-The \emph{indexed option comonad} is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
-where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
-Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
-%
-\begin{equation*}
-\begin{array}{l}
-\ctyp{\ident{L}}{\alpha} = \alpha\\
-\ctyp{\ident{D}}{\alpha} = 1\\
-\\
-\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
-\ident{counit}_{\ident{L}}~v = v\\
-\end{array}
-\qquad
-\begin{array}{lcl}
-\multicolumn{3}{l}{
-  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
-    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
-\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
-\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
-\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
-\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The \emph{indexed option comonad} models the semantics of the liveness coeffect system discussed in 
-Section~\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
-and $\ctyp{\ident{D}}{\alpha}=1$ models a dead context which does not contain a value. The \ident{counit}
-operation extracts a value from a live context. As in the direct model discussed in Chapter~\ref{ch:appendix},
-the \ident{cobind} operation can be seen as an implementation of dead code elimination. The definition 
-only evaluates $f$ when the result is marked as live and is thus required, and it only accesses $x$ if 
-the function $f$ requires its input.
-
-The indexed family $\ctyp{\cclrd{r}}{}$ in the above example is analogous to the \ident{Maybe}
-(or option) data type $\ident{Maybe}\,\alpha = 1 + \alpha$. As mentioned earlier, this type does not 
-permit (non-indexed) comonad structure, because $\ident{counit}~()$ is not defined. This is not a 
-problem with indexed comonads, because live contexts are distinguished by the (type-level) coeffect 
-annotation and \ident{counit} only needs to be defined on live contexts.
-
-\begin{example}[Indexed product]
-The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
-$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
-As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
-= \alpha \times (\cclrd{r} \rightarrow \rho)$ represents a value $\alpha$ together with a function that
-associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
-The cobind and counit operations are defined as:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
-\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
-\end{array}
-\quad
-\begin{array}{l}
-\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
-    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
-\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-In the definition, we use the notation $(a, g)$ for a pair containing the value of type $\alpha$
-together with $g$, which is a function $\cclrd{r} \rightarrow \rho$. The \ident{counit} operation
-takes a value and a function (with empty set as a domain), ignores the function and extracts the
-value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to 
-restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in oder to get
-implicit parameters required by the argument of $f$ and by the resulting computation, respectively
-(\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed 
-to \ident{cobind} is initially defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is 
-valid in both cases.
-
-The structure of \emph{indexed comonads} is sufficient to model sequential composition of 
-computations that use a single variable (as discussed in Section~\ref{sec:path-sem}). To model 
-full $\lambda$-calculus with lambda abstraction and multiple-variable contexts, we need 
-additional operations introduced in the next section.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Flat indexed comonads}
-\label{sec:flat-semantics-monoidal}
-
-Because of the asymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
-the duality between monads and comonads can no longer help us with defining the additional structure
-required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
-to the context. In application and lambda abstraction, the context is propagated differently than 
-in effectful computations.
-
-To model the effectful $\lambda$-calculus, Moggi~\cite{monad-notions} requires a \emph{strong} monad which
-has an additional operation $\ident{strength} : \alpha\times\mtyp{}{\beta} \rightarrow \mtyp{}{(\alpha\times\beta)}$.
-This allows lifting of free variables into an effectful computation. In Haskell, strength can be expressed
-in the host language and so it is implicit.
-
-To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions} 
-require \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
-%
-\begin{equation*}
-\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
-\end{equation*}
-%
-The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
-contexts and is used to merge the context of the declaration site (containing free variables)
-and the call site (containing bound variable). For example, for implicit parameters, this combines
-the additional parameters defined in the two contexts.
-
-The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
-\emph{splitting} of contexts. In addition, we also need a lifting operation (similar to $\iota$ 
-from Definition~\ref{def:flat-family}) to model sub-coeffecting.
-
-\begin{definition}
-Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-a \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
-equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
-and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
-%
-\begin{compactitem}
-\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
-  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
-\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
-  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
-\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
-  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
-\end{compactitem}
-\end{definition}
-
-\noindent
-The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
-values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a 
-single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
-\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
-bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful 
-as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
-the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.
-
-The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
-into two separate values. Note that this does not simply duplicate the value, because the additional
-context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide 
-\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often 
-the \emph{least upper-bound}\footnotemark[1].
-
-Finally, $\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ is a family of operations that ``forget'' some part of
-a context. This models the sub-coeffecting operation and lets us, for example, forget some of the
-available implicit parameters, or turn a live context (containing a value) into a dead context (empty).
-
-\paragraph{Alternative definition.}
-Although we do not demand this as a general law, in all our systems, it is the case that
-$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$.
-This special case allows a simpler definition of \emph{indexed flat comonad} by expressing the 
-\ident{split} operation in terms of lifting (sub-coeffecting) as follows:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
-\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
-  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c), 
-    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
-\end{array}
-\end{equation*}
-%
-The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object 
-mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map} 
-for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
-are first and second projections from a two-element pair. To define the 
-$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
-to throw away additional parts of the context and then transform the values in the 
-context.
-
-This alternative is valid for our examples, but we do not use it for two reasons. Firstly, this
-would be the only place where our semantics uses a variable \emph{twice} (in this case $c$). 
-Thus using an explicit \ident{split} means that the structure required by our semantics does not
-need to provide variable duplication and our model could be embedded in linear or affine category.
-Secondly, explicit \ident{split} is similar to the definition that is needed for structural 
-coeffects in Chapter~\ref{ch:structural} and so it makes the connection between the two system 
-easier to see.
-
-\paragraph{Examples.}
-All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:flat-semantics-idx} can
-be extended into \emph{flat indexed comonads}. Note that this is not a \emph{general} statement, 
-because each example requires us to define additional operations, specific for the example.
-
-\begin{example}[Monoidal comonads]
-Just like indexed comonads generalise co\-monads, the additional structure of
-flat indexed comonads generalises symmetric semimonoidal comonads of Uustalu 
-and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
-where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the 
-monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation 
-is defined by duplication and $\ident{lift}_{1,1}$ is the identity function.
-\end{example}
-
-\begin{example}[Indexed option]
-Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively 
-and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined 
-as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations 
-of a flat indexed comonad are defined as follows:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
-\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
-\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
-\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
-\end{array}
-\quad
-\begin{array}{rcl}
-\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
-\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
-\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
-\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
-\end{array}
-\quad
-\begin{array}{rcl}
-\ident{lift}_{\ident{L}, \ident{D}}~v &\narrow{=}& ()\\
-\ident{lift}_{\ident{L}, \ident{L}}~v &\narrow{=}& v\\
-\ident{lift}_{\ident{D}, \ident{D}}~() &\narrow{=}& ()\\
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-Without the indexing, the \ident{merge} operations implements \emph{zip} on option values,
-returning an option only when both values are present. The behaviour of the \ident{split} 
-operation is partly determined by the indices. When the input is \emph{dead}, both values have 
-to be dead (this is also the only solution of $\cclrd{\ident{D}} =\cclrd{r}\sqcap\cclrd{s}$), but when
-the input is \emph{live}, the operation can perform implicit sub-coeffecting and drop one of
-the values.
-
-Explicit sub-coeffecting using the (\emph{sub}) rule is modelled by the \ident{lift} operation.
-This can turn a \emph{live} value $v$ into a dead value $()$, or it can behave as identity.
-The behaviour is, again, determined by the index.
-
-\begin{example}[Indexed product]
-For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
-$\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
-is $\alpha \times (\cclrd{r} \rightarrow \rho)$ where $\rho$ is the type of implicit parameter values.
-The additional operations are defined as:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\ident{split}_{\cclrd{r}, \cclrd{s}}~((a,b), g) &\narrow{=}& ((a, \restr{g}{\cclrd{r}}), (b, \restr{g}{\cclrd{s}}))\\
-\ident{merge}_{\cclrd{r}, \cclrd{s}}~((a, f), (b, g)) &\narrow{=}& ((a, b), f \uplus g)\\
-\ident{lift}_{\cclrd{r'}, \cclrd{r}}~(a, g) &\narrow{=}& (a, \restr{g}{\cclrd{r}})\\
-\end{array}
-\quad
-\begin{array}{l}
-\textnormal{where}~f \uplus g = \\[-0.25em]
-\quad\restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g 
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The \ident{split} operation splits the tuple and restricts the function (representing available
-implicit parameters) to the required sub-sets. This corresponds to the definition in terms
-of \ident{lift}, which performs just the restriction. The \ident{merge} operation is more 
-interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
-in Section~\ref{sec:applications-flat-impl}. It merges the values, preferring the definitions from
-the right-hand side (call site) over left-hand side (declaration site). Thus the operation is not
-symmetric.
-
-\begin{example}[Indexed list]
-Our last example provides the semantics of data-flow computations. The flat coeffect algebra 
-is formed by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$. In a 
-non-indexed version, the semantics is provided by a non-empty list. In the indexed semantics,
-the index represents the number of available past values. The data type is then a pair of 
-the current value, followed by $n$ past values. The mappings that form the flat indexed comonad 
-are defined as follows:
-
-\begin{equation*}
-\begin{array}{l}
-\ident{counit}_{0}\langle a_0 \rangle = a_0
-\\[0.45em]
-\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
-\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
-\\[0.45em]
-\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
-\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
-\\[0.45em]
-\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
-\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
-\\[0.45em]
-\ident{lift}_{n', n} \langle a_0, \ldots, a_{n'} \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
-\quad \langle a_0, \ldots, a_n \rangle
-\end{array}
-\begin{array}{l}
-\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[11em]
-~\\
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The reader is invited to check that the number of required past elements in each of the mappings
-matches the number specified by the indices. The index specifies the number of \emph{past} elements
-and so the list always contains at least one value. Thus \ident{counit} returns the element of a
-singleton list.
-
-The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results 
-of the $f$ function, which itself requires $m$ past values. When combining two lists, 
-$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter 
-argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths. 
-Finally, the lifting operation just drops some number of elements from a list.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Properties and related notions}
-
-The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
-calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
-additional properties and other categorical structures that have been proposed mainly in the 
-context of monads and effects and we look how they relate to indexed comonads.
-
-\paragraph{Shape preservation.}
-Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
-this means that the core comonad structure does not provide a way of modeling computations where
-the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
-the length of the list stays the same after applying \ident{cobind}.
-
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma;\cclrd{\Delta}}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}  
+\tyrule{param}
+  { ~ }
+  { \coctx{\Gamma;\cclrd{\Delta}}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{\Delta}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{r}\setminus\cclrd{\Delta}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{letpar}
+  { \coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e_1 : \ident{num} &
+    \coctx{\Gamma;\cclrd{\Delta \cup \{ \ident{?p} \}}}{\cclrd{s}} \vdash e_2 : \tau}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} } \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau }
+\end{equation*}
+
+\figcaption{Choosing unique typing for implicit parameters}
+\label{fig:flat-resolve-impl}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Implicit parameters}
+
+For implicit parameters we follow the behaviour implemented by Haskell \cite{app-implicit-parameters}
+where function abstraction captures all parameters that are available at the declaration-site and 
+places all other requirements on the call-site. For the example in the introduction, this means
+that the body of $f$ captures the value of \ident{?p} available from the declaration-site
+and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The 
+program thus evaluates to $1$.
+
+To express this behaviour formally, we extend the coeffect type system to additionally track 
+implicit parameters that are currently in scope. The typing judgement becomes:
+%
+\begin{equation*}
+\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau
+\end{equation*}
+%
+Here, $\cclrd{\Delta}$ is a set of implicit parameters that are in scope at the declaration-site.
+The modified typing rules are shown in Figure~\ref{fig:flat-resolve-impl}. The rules (\emph{var}),
+(\emph{const}), (\emph{app}) and (\emph{let}) are modified to use the new typing judgement, but they
+simply propagate the information tracked by $\cclrd{\Delta}$ to all assumptions. The (\emph{param})
+rule also remains unchanged -- the implicit parameter access is still tracked by the coeffect 
+$\cclrd{r}$ meaning that we still allow a form of dynamic binding (the parameter does not have
+to be in static scope).
+
+The most interesting rule is (\emph{abs}). The body of a function requires implicit parameters
+tracked by $\cclrd{r}$ and the parameters currently in (static) scope are $\cclrd{\Delta}$.
+The coeffect on the declaration site becomes $\cclrd{\Delta}$ (capture all available parameters)
+and the latent coeffect attached to the function becomes $\cclrd{r\setminus\Delta}$ (require all
+remaining parameters from the call-site). Finally, in the (\emph{letpar}) rule, we add the newly
+bound implicit parameter \ident{?p} to the static scope in the sub-expression $e_2$.
+
+\paragraph{Properties.}
+If a written in a coeffect language with implicit parameters is well-typed according to the 
+type system presented in Figure~\ref{fig:flat-resolve-impl}, then the type system gives a unique
+derivation. We use this unique typing derivation to give the semantics of coeffect language
+with implicit parameters in Chapter~\ref{ch:transl} and we also implement this algorithm as
+discussed in Chatper~\ref{ch:impl}.
+
+The type system is more restrictive than the fully general one and it reject certain programs that
+could be typed using the more general system. This is expected -- we are restricting the fully
+general coeffect calculus to match the typing and semantics of implicit parameters as known from 
+Haskell.
+
+In order to prove the uniqueness of typing theorem (Theorem~\ref{thm:flat-impl-unique}), we first
+need the inversion lemma (Lemma~\ref{thm:flat-impl-invert}).
+
+\begin{lemma}[Inversion lemma for implicit parameters]
+\label{thm:flat-impl-invert} 
 ~
-
-Indexed comonads are not restricted by this property of comonads. For example, given the indexed
-product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
-the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
-just implicit parameters $\cclrd{s}$.
-
-\paragraph{Families of monads.}
-When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
-\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
-comonads}, is defined as follows.
-
-\begin{definition}
-\label{def:flat-family}
-A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
-  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
-$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
-$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
-certain coherence conditions.
-\end{definition}
-
-\noindent
-A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data 
-types needs to form a comonad separately. For example, our indexed option does not form a family of 
-comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given 
-a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
-an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
-we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
-$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
-
-\paragraph{Parameteric effect monads.}
-Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
-indexed comonads) are closely related to our definition.  Although presented in a more general 
-categorical framework (and using monads), the model defines the \ident{unit} operation only on the 
-unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal 
-structure.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of flat calculus}
-\label{sec:flat-semantics-calculus}
-
-In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
-computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed 
-comonad} structure, we can now define a single uniform semantics that is capable of capturing all 
-our examples, as well as other computations that can be modelled by the structure.
-
-\paragraph{Contexts and functions.}
-The modelling of contexts and functions generalizes the earlier concrete examples. We use the 
-family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed) data-type that wraps the product of 
-free variables of the context and the arguments of functions:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau} 
-  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
-\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
-\end{array}
-\end{equation*}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[t]
-
-\begin{equation*}
-\begin{array}{ll}
-\sem{\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }~\ctx =
-  \pi_i~(\ident{counit}_{\cunit}~\ctx) & (\emph{var})
-\\[0.5em]
-\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
-  \delta~(c_i) & (\emph{const})
-\\[0.5em]
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
-  \hspace{3em}\sem{\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v))
-\\[0.5em]
-\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
-  \hspace{2.5em}
-  \begin{array}{l}  
-  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}}~
-    (\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~(\lambda x.(x,x))~\ctx) \\[-0.25em]
-  \kvd{in}~\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
-      (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~\sem{\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~\ctx_2)
-  \end{array}    
-\\[1.25em]
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }~\ctx = & (\emph{sub})\\[-0.25em]
-  \hspace{3em}\sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx)~
-    \hspace{7.5em}(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
-\end{array}
-\end{equation*}
-
-\figcaption{Categorical semantics of the flat coeffect calculus}
-\label{fig:flat-semantics}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Expressions.}
-The definition of the semantics is shown in Figure~\ref{fig:flat-semantics}. For readability, we 
-write the definitions in a simple programming language notation as opposed to the point-free 
-categorical style. However, it can be equally written using just the operations of flat indexed 
-comonad together with $i^{th}$ projection from a tuple represented by $\pi_i$, \emph{curry} and 
-\emph{uncurry}, function composition, value duplication ($\Delta : A \rightarrow A \times A$) and 
-function pairing (given $f:A\rightarrow B$ and $g:C\rightarrow D$ then $f\times g : A\times C \rightarrow B \times D$). 
-These operations can be provided by \eg~a Cartesian-closed category.
-
-The semantics of variable access and abstraction are the same as in the semantics of Uustalu and 
-Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses 
-$\ident{counit}_{\cunit}$ to extract product of free-variables from the context and then projection
-$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) takes the context $\ctx$ and function argument 
-$v$ and merges their additional contexts using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. Assuming
-the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives us a 
-value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
-Assuming that $n$-element tuples are associated to the left, the wrapped context is equivalent to
-$\sigma_1 \times \ldots \times \sigma_n \times \tau_1$, which can then be passed to the body of the
-function.
-
-The semantics of application is more complex. It first duplicates the free-variable product inside the 
-context (using $\ident{map}_{\cclrd{r}}$ and duplication). Then it splits this context using 
-$\ident{split}_{\cclrd{r}, \cclrd{s} \cpar \cclrd{t}}$. The two contexts contain the same variables
-(as required by sub-expressions $e_1$ and $e_2$), but different coeffect annotations. The first
-context (with index $\cclrd{r}$) is used to evaluate $e_1$, resulting in a function 
-$\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. To obtain the result, we compose this with a function
-created by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$ on the semantics of sub-expression
-$e_2$, which is of type $\ctyp{\cclrd{s}\cseq\cclrd{t}}{\sigma_1 \times \ldots \times \sigma_n}
-\rightarrow \ctyp{\cclrd{t}}{\tau_1}$.
-
-Finally, constants (\emph{const}) are modelled by a global dictionary $\delta$ and sub-coeffecting
-is interpreted by dropping additional context from the provided context $\ctx$ using 
-$\ident{lift}_{\cclrd{r}, \cclrd{r'}}$ and providing it to the semantics of the assumption.
-
-\paragraph{Properties.}
-
-The categorical semantics can be used to embed context-dependent computations in functional 
-programming languages, similarly to how monads provide a way of embedding effectful computations.
-More importantly, it also provides validation for the design of the type system developed in 
-Section~\ref{sec:flat-calculus-types}. As stated in the following theorem, the annotations in 
-the type system match those of the semantic functions.
-
-\vspace{2em}
-
-\begin{theorem}[Correspondence]
-\label{thm:flat-correspondence}
-In all of the typing rules of the flat coeffect system, the context annotations $\cclrd{r}$ of typing 
-judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ 
-correspond to the indices of mappings $\ctyp{\cclrd{r}}{}$ in the corresponding semantic function defined 
-by $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}$.
+%
+\begin{enumerate}
+\raggedright
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash x : \tau$ then $x : \tau \in \Gamma$ and $\cclrd{c} = \cunit$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash n : \tau$ then $\tau = \ident{num}$ and $\cclrd{c} = \czero$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash e_1~e_2 : \tau_2$
+ then there is some $\tau_1,\cclrd{r},\cclrd{s}$ and $\cclrd{t}$ such that
+ $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2$ and 
+ $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{s}} \vdash e_2 : \tau_1$ and also
+ $\cclrd{c} = \cclrd{r \cup s \cup t}$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$
+ then there is some $\tau_1, \cclrd{s}$ and $\cclrd{r}$ such that
+ $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e_1 : \tau_1$ and
+ $\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{s}} \vdash e_2 : \tau_2}$ and also
+ $\cclrd{c} = \cclrd{s \cup r}$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash \ident{?p} : \ident{num}$ then 
+ $\ident{?p} \in \cclrd{c}$ and $\cclrd{c} = \cclrd{ \{\ident{?p}\} }$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash \lambda x\!:\!\tau_1.e : \tau$ then 
+  $\tau = \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ for some $\tau_2$ with 
+  ${\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2}$ and
+  $\cclrd{c} = \cclrd{\Delta}$ and also
+  $\cclrd{s} = \cclrd{r\setminus\Delta}$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash \kvd{let}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau }$ then 
+  there is some $\cclrd{r}, \cclrd{s}$ such that
+  $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e_1 : \ident{num}$ and
+  $\coctx{\Gamma;\cclrd{\Delta \cup \{ \ident{?p} \}}}{\cclrd{s}} \vdash e_2 : \tau$ and also
+  $\cclrd{c}=\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}}$.
+
+\end{enumerate}  
+\end{lemma}
+\begin{proof}
+Follows from the individual rules given in Figure~\ref{fig:flat-resolve-impl}.
+\end{proof}
+
+\begin{theorem}[Uniqueness of coeffect typing for implicit parameters]
+\label{thm:flat-impl-unique}
+In the type system for implicit parameters defined in Figure~\ref{fig:flat-resolve-impl}, when
+$\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau$ and 
+$\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r'}} \vdash e : \tau'$ then $\tau = \tau'$ and $\cclrd{r} = \cclrd{r'}$.
 \end{theorem}
 \begin{proof}
-By analysis of the semantic rules in Figure~\ref{fig:flat-semantics}.
+Suppose that (A) $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash e : \tau$ and 
+(B) $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c'}} \vdash e : \tau'$. We show by induction over the typing 
+derivation of $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash e : \tau$ that $\tau = \tau'$ and $\cclrd{c}=\cclrd{c'}$.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{abs}): $e = \lambda x\!:\!\tau_1.e_1$ and $\cclrd{c}=\cclrd{\Delta}$.
+  For some $\cclrd{r}$, 
+  $\tau = \tau_1 \xrightarrow{\cclrd{r}\setminus\cclrd{\Delta}} \tau_2$ and also
+  $\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2}$. 
+  By case (6) of Lemma~\ref{thm:flat-impl-invert}, the final rule of the derivation (B) must
+  have also been (\emph{abs}) and this derivation has a sub-derivation with a conclusion
+  ${\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2'}$.
+  By the induction hypothesis $\tau_2 = \tau_2'$ and so $\tau = \tau'$ and 
+  also $\cclrd{c}=\cclrd{c'}=\cclrd{\Delta}$.
+
+\vspace{0.5em}\noindent\hangindent=0.7cm 
+Case (\emph{param}): $e = \ident{?p}$, from Lemma~\ref{thm:flat-impl-invert}, $\tau = \tau' = \ident{int}$
+  and $\cclrd{c} = \cclrd{c'} = \cclrd{\{ \ident{?p} \}}$.
+
+\vspace{0.5em}\noindent\hangindent=0.7cm 
+Cases (\emph{var}), (\emph{const}) are direct consequence of Lemma~\ref{thm:flat-impl-invert}.
+
+\vspace{0.5em}\noindent\hangindent=0.7cm 
+Cases (\emph{var}), (\emph{const}), (\emph{app}), (\emph{let}), (\emph{param}) and (\emph{letpar})
+  similarly to (\emph{abs}).
 \end{proof}
 
-\noindent
-Thanks to indexing, the statement of the theorem is significantly stronger than for a 
-non-indexed system, because it provides the justification for our choice of indices in the typing
-rules. In particular, we can see that the annotations follow from the annotations on primitive 
-functions that define the semantics. Also, each function defining the semantics uses a distinct 
-operation of the coeffect algebra and so the type system is the most general possible definition
-(within the comonadic framework we use).
-
-Although the notion of indexed comonads presented in this section is novel and interesting 
-in its own, the main reason for introducing it is to motivate the flat coeffect type system.
-This is captured by the Theorem~\ref{thm:flat-correspondence}. In the next section, we shift our
-focus from (categorical) semantics to syntactic properties of the calculus.
+\paragraph{Implementation.}
+From the presentation in this section, it might appear that resolving the ambiguity related to
+lambda abstraction for implicit parameters requires a type system that is quite different from
+the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
+case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are 
+simpler.
+
+Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets 
+of implicit parameters to satisfy the constraints. We still need to track implicit parameters in 
+scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
+to the one from Figure~\ref{fig:flat-types}:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}} 
+\end{equation*}
+%
+Given a typing derivation for the body that produced constraints $C$, we generate an additional
+constraint that restricts $\cclrd{r}$ (declaration-site requirements) to those available in the
+current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds 
+the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Dataflow and liveness}
+
+\newpage
+\newpage
 
@@
 %
-%       ###                  #                    #       #          
-%      #   #                 #                    #                  
-%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
-%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
-%          #  #  ##  #   #   #      ####  #       #       #    #     
-%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
-%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
-%             #   #                                                  
-%              ###                                                   
-%
-% ==================================================================================================
-
-\section{Syntactic equational theory}
-\label{sec:flat-syntax}
-
-Each of the concrete coeffect calculi discussed in this chapter has a different notion of context,
-much like various effectful languages have different notions of effects (such as exceptions or
-mutable state). However, in all of the calculi, the context has a number of common properties that 
-are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
-that hold for all of the coeffect systems. Further properties hold for systems where the context
-satisfies additional properties.
-
-In this section, we look at such shared syntactic properties. This accompanies the previous section, 
-which provided a \emph{semantic} justification for the axioms of coeffect algebra with a 
-\emph{syntactic} justification. Operationally, this section can also be viewed as providing a 
-pathway to an operational semantics for two of our systems (implicit parameters and liveness),
-which can be based on syntactic substitution. As we discuss later, the notion of context for
-data-flow is more complex.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Syntactic properties}
-\label{sec:flat-syntax-props}
-
-Before discussing the syntactic properties of general coeffect calculus formally, it should be 
-clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
-that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
-Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
-equations define four syntactic reductions on the terms:
-%
-\begin{equation*}
-\begin{array}{rllcl}
-(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
-(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
-(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
-e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
-\end{array}
-\end{equation*}
-%
-The rules capture syntactic reductions that can be performed in a general calculus, without any 
-knowledge of the specific notion of context. If the reductions preserve the type of the expression
-(type preservation), then operational semantics can be defined as a repeated application of the 
-rules, until a specified normal form (\ie~a value) is reached.
-
-The first two equations define well-known call-by-name and call-by-value reductions. The \kvd{glet} 
-reduction (inspired by Filinski \cite{monads-inaction}) models a system where program is reduced to 
-a sequence of primitive operations, sequentially composed using the special \kvd{glet} construct. 
-In the rest of the section, we briefly outline the interpretation of the four 
-rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name 
-(Section~\ref{sec:flat-syntax-cbn}) in more details.
-
-The focus of our work is on the general coeffect system and so we do not discuss the operational
-semantics of the specific notions of context. However, some work in that area has been done
-by Brunel et al. \cite{coeffects-quantitative}.
-
-\paragraph{Call-by-name.} 
-In call-by-name, the argument is syntactically substituted for all occurrences of a variable. It
-can be used as the basis for operational semantics of purely functional languages. However, using
-the rule in effectful languages breaks the \emph{type preservation} property. For example, consider
-a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
-A function $\lambda x.y$ is effect-free:
-%
-\begin{equation*}
-y\!:\!\tau_1 \vdash \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 \;\&\; \emptyset
-\end{equation*}
-%
-Substituting an expression $e$ with effects $\{ \ident{write} \}$ for $y$ changes the type of 
-the function by adding latent effects (without changing the immediate effects):
-%
-\begin{equation*}
-\vdash \lambda x.e : \tau_1 \xrightarrow{ \{ \ident{write} \} } \tau_2 \;\&\; \emptyset
-\end{equation*}
-%
-Similarly to effect systems, substituting a context-dependent computation $e$ for a variable $y$ can
-add latent coeffects to the function type. However, this is not the case for \emph{all} flat coeffect
-calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
-parameters system. This means that certain coeffect systems support call-by-name evaluation strategy
-and could be embedded in purely functional language (such as Haskell).
-
-\paragraph{Call-by-value.}
-
-The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
-first reduced to a \emph{value} before performing the substitution. In effectful languages, 
-value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
-values are identifiers $x$ or functions $(\lambda x.e)$.
-
-The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
-A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
-immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
-sense \emph{and} has not immediate context requirements. We define this formally in 
-Section~\ref{sec:flat-syntax-cbv}.
-
-The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
-including all our three examples. However, it is rather weak -- in order to use it, the concrete 
-semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
-
-\paragraph{Internalized sequencing.}
-
-The (\emph{internalized sequencing}) rule captures an operational semantics where the language
-provides a construct representing \emph{sequential composition} and expressions can be reduced to 
-a normal form, consisting of a sequence of context-dependent operations. We choose to write the 
-sequencing operation as \kvd{glet} to emphasize that this is a separate primitive and not an 
-ordinary syntactic \kvd{let}. The normal form looks as follows:
-%
-\begin{equation*}
-\kvd{glet}~x_1 = e_1~\kvd{in}~\ldots~\kvd{glet}~x_n = e_n~\kvd{in}~e
-\end{equation*}
-%
-Here, the expressions $e_1, \ldots, e_n, e$ do not contain further nested \kvd{glet} constructs.
-This evaluation strategy provides context-dependent counterpart to operational view of monads
-developed by Filinsky \cite{monads-inaction}. We discuss how expressions reduce to the normal form
-in Section~\ref{sec:flat-syntax-norm}
-
-The (\emph{internalized sequencing}) rule is useful when defining a concrete semantics for 
-a language that provides constructs for explicitly providing the context. For example, consider
-a language with implicit parameters where a parameter is defined by $\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2$.
-Semantics for such language would provide a reduction rule:
-%
-\begin{equation*}
-\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~(\kvd{glet}~x = \ident{?p}~\kvd{in}~e)
- ~~\leadsto~~ \kvd{let}~x = e_1~\kvd{in}~e
-\end{equation*}
-%
-The \kvd{glet} construct provides a way of sequentialising the context-require\-ments so
-that they can be discharged by matching constructs providing the required context. As discussed
-earlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
-this form of evaluation (rather than looking at semantics for concrete systems).
-
-\paragraph{Local soundness and completeness.}
-Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
-are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
-introduce a function arrow (lambda abstraction) and eliminate it (application) are sufficiently 
-strong, but not too strong.
-
-The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
-already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
-coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Call-by-value evaluation}
-\label{sec:flat-syntax-cbv}
-
-As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
-calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
-term to a \emph{value} has to be provided for each system. Syntactic values are defined
-in the usual way:
-%
-\begin{equation*}
-\begin{array}{llll}
-v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
-n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
-e\in\mathit{Expr}     & e &::=& v \sep n 
-\end{array}
-\end{equation*}
-%
-The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
-coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
-not automatically defer all context requirements.
-
-\begin{definition} 
-An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if it is a syntactic value,
-\ie~$e \in \mathit{SynVal}$ and it has no context-dependencies, \ie~$\coctx{\Gamma}{ \cunit } \vdash e : \tau$.
-\end{definition}
-
-\noindent
-The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
-variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
-context requirements of the term:
-
-\begin{lemma}[Call-by-value substitution]
-\label{thm:flat-subst-cbv}
-In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
-$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
-change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
-\end{lemma}
-\begin{proof}
-By induction over the type derivation, using the fact that $x$ and $v$ are annotated
-with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
-\end{proof}
-
-\noindent
-The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
-proving that call-by-value reduction preserves typing requires an additional constraint on the
-flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
-by the (\emph{approximation}) property:
-
-\begin{equation*}
-\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
-\hspace{10em}(\textit{approximation})
-\end{equation*}
-%
-Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
-under-approximates the actual context capabilities while the $\cpar$ operation (combining of
-context requirements) over-approximates the actual context requirements.
-
-The property holds for the three systems we consider -- for implicit parameters, this is 
-an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
-is smaller than the least upper bound.
-
-Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
-$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the type preservation theorem is
-stated as follows:
-
-\begin{theorem}[Type preservation for call-by-value]
-\label{thm:cbv-reduction}
-In a flat coeffect system with the (\emph{approxi\-mation}) property, 
-if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
-\begin{equation*}
-\inference
-  { \inference
-  { \inference
-      {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
-      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
-   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
-\end{equation*}
-The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
-From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
-As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
-%
-\begin{equation*}
-\tyrule{sub}
- { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
- { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
-\end{equation*}
-%
-Comparing the final conclusions of the above two typing derivations shows that 
-the reduction preserves type and coeffect annotation.
-\end{proof}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Call-by-name evaluation}
-\label{sec:flat-syntax-cbn}
-
-When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the 
-sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$. 
-As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in 
-general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two 
-interesting cases.
-
-The typing is preserved for different reasons in two of our systems, so we briefly review the 
-concrete examples. Then, we prove the substitution lemma for two special cases of flat coeffects
-(Lemma~\ref{thm:cbn-substitution-top} and Lemma~\ref{thm:cbn-substitution-bot}) and finally, we
-state the conditions under which typing preservation hold for flat coeffect calculi 
-(Theorem~\ref{thm:cbn-flat}).
-
-\paragraph{Data-flow.} 
-The type preservation property does not hold for data-flow. This case is similar to the example
-shown earlier with effectful computations. As a minimal example, consider the substitution of 
-a context-dependent expression $\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
-%
-\begin{equation*}
-\begin{array}{rclcl}
- \coctx{y\!:\!\tau_1,z\!:\!\tau_1}{0} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{0} \tau_2 & & (\textnormal{before}) \\
- \coctx{z\!:\!\tau_1}{1} &\narrow{\vdash}& \lambda x.\kvd{prev}~z : \tau_1 \xrightarrow{1} \tau_2 & & (\textnormal{after}) \\
-\end{array}
-\end{equation*}
-%
-After the substitution, the coeffect of the body is $1$. The rule for lambda abstraction requires
-that $1=\mathit{min}(r,s)$ and so the least solution is to set both $r,s$ to $1$. The substitution
-this affects the coeffects attached both to the function type and the overall context.
-
-Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
-actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
-system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
-
-\paragraph{Implicit parameters.} In data-flow, there is no typing for the resulting expression that
-preserves the type of the function. However, this is not the case for all systems. Consider substituting
-an implicit parameter access $\ident{?p}$ for a free variable $y$ under a lambda:
-%
-\begin{equation*}
-\begin{array}{rclcl}
- \coctx{y\!:\!\tau_1}{\emptyset} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{before}) \\
- \coctx{\emptyset}{ \{\ident{?p}\} } &\narrow{\vdash}& \lambda x.\ident{?p} : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{after}) \\
-\end{array}
-\end{equation*}
-%
-The above shows one possible typing of the body -- one that does not change the coeffects of the 
-function type and attaches all additional coeffects (implicit parameters) to the context. In case
-of implicit parameters (and, more generally, systems with set-like annotations) this is always 
-possible.
-
-\paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
-substituting an arbitrary expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
-%
-\begin{equation*}
-\begin{array}{rclcl}
- \coctx{y\!:\!\tau_1}{\ident{L}} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{before}) \\
- \coctx{\emptyset}{\ident{L}} &\narrow{\vdash}& \lambda x.e : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{after}) \\
-\end{array}
-\end{equation*}
-%
-In the original expression, both the overall context and the function type are annotated with \ident{L}, 
-because the body contains a variable access. An expression $e$ can always be treated as being annotated
-with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so we can also treat $e$
-as being annotated with coeffects $\ident{L}$. As a result, substitution does not change the coeffect.
-
-\paragraph{Reduction theorem.}
-The above examples (implicit parameters and liveness) demonstrate two particular kinds of coeffect 
-algebra for which typing preservation holds. Proving the type preservation separately provides 
-more insight into how the systems work. We consider the two cases separately, but find a more general
-formulation for both of them.
-
-\begin{definition}
-We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
-($\forall r \in \C \;.\; r \,\cleq\, \cunit $) and \emph{bottom-pointed} if it is the smallest (bottom) 
-element ($\forall r \in \C \;.\; r \,\cgeq\, \cunit $).
-\end{definition}
-
-\noindent
-Liveness is an example of top-pointed coeffects as variables are annotated with 
-$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and data-flow are examples
-of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
-additional requirements:
-
-\begin{lemma}[Top-pointed substitution]
-\label{thm:cbn-substitution-top}
-In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
-not change the coeffects of $e_r$:
-%
-\begin{equation*}
-\begin{array}{l}
- \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
-   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
- \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
-\end{array}
-\end{equation*}
-\end{lemma}
-
-\begin{proof}
-Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
-\end{proof}
-%
-\noindent
-As variables are annotated with the top element $\cunit$, we can substitute the term $e_s$ 
-for any variable and use sub-coeffecting to get the original typing (because
-$\cclrd{s} \,\cleq\, \cunit$).
-
-In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
-requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
-then the context requirements arising from the substitution can be associated with the context
-$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
-does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
-our implicit parameters example (all three operators are a set union) and for other set-like 
-coeffects. It allows the following substitution lemma:
-
-\begin{lemma}[Bottom-pointed substitution]
-\label{thm:cbn-substitution-bot}
-In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
-where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
-$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
-%
-\begin{equation*}
-\begin{array}{l}
-\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
-\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
-\end{array}
-\end{equation*}
-
-\end{lemma}
-\begin{proof}
-By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
-the free-variable context. See Appendix~\ref{sec:appendix-flat-cbn}.
-\end{proof}
-
-\noindent
-The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
-requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
-structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
-coeffects associated with individual variables, thus preserving typing in a complementary way.
-
-The two substitution lemmas discussed above show that the call-by-name evaluation strategy can be 
-used for certain coeffect calculi, including liveness and implicit parameters. Assuming
-$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following theorem holds:
-
-\begin{theorem}[Type preservation for call-by-name]
-\label{thm:cbn-flat}
-In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
-Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \rightarrow_{\ident{cbn}} e'$ then it is also the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-
-For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
-to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
-and $r \,\czip\, t \;=\; r \,\cpar\, t$.
-For bottom-pointed coeffect algebra, consider the typing derivation for the term
-$(\lambda x.e_r)\,e_s$ before reduction:
-%
-\begin{equation*}
-\inference
-  { \inference
-      {\coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r}
-      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{r}} \tau_r } &
-   \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash (\lambda x.e_r)\,e_s : \tau_r }
-\end{equation*}
-%
-The derivation uses the idempotence of $\czip$ in the first step, followed by the 
-(\emph{app}) rule. The type of the term after substitution, using Lemma~\ref{thm:cbn-substitution-bot} is:
-%
-\begin{equation*}
-\inference
-  { \coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r & 
-    \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
-  { \coctx{\Gamma, x:\tau_r}{\cclrd{r}\;\cseq\;\cclrd{s}} \vdash \subst{e_r}{x}{e_s} : \tau_s }
-\end{equation*}
-%
-From the assumptions of Lemma~\ref{thm:cbn-substitution-bot}, we know that $\cseq\,=\,\cpar$
-and the operation is idempotent, so trivially:
-$\cclrd{r}\;\cseq\;\cclrd{s} = \cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})$
-\end{proof}
-
-\paragraph{Expansion theorem.}
-The $\eta$-expansion (local completeness) is similar to $\beta$-reduction (local soundness) in that
-it holds for some flat coeffect systems, but not for all. Out of the examples we discuss, it 
-holds for implicit parameters, but does not hold for liveness and data-flow.
-
-Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
-expression $e$ may require no variables (both immediate and latent coeffects are marked as
-$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
-marking the context and function argument as live. In case of data-flow, the immediate coeffects 
-are made larger by the lambda abstraction -- the context requirements of the function value are 
-imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
-the next chapter.
-
-However, $\eta$-expansion preserves the type for implicit parameters and, more generally,
-for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the 
-standard $\eta$-reduction:
-
-\begin{theorem}[Type preservation of $\eta$-expansion]
-In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
-where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
-and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
-\end{theorem}
-\begin{proof}
-The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
-\begin{equation*}
-\inference
-  { \inference
-    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
-      \coctx{x:\tau_1}{\cunit} \vdash x:\tau_1 }
-    { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
-  { \inference
-    { \inference
-      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
-      { \coctx{\Gamma, x:\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
-    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
-\end{equation*}
-%
-The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
-application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
-original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
-is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
-this is sufficient for showing type preservation.
-\end{proof}
-
-\noindent
-In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
-there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
-these include liveness and implicit parameters. Moreover, for implicit parameters (and more
-generally, any set-like flat coeffect algebra), the $\eta$-expansion holds as well, giving us
-both local soundness and local completeness as coined by Pfenning and Davies \cite{logic-modal-reconstruction}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Internalized sequencing}
-\label{sec:flat-syntax-norm}
-
-The call-by-value and call-by-name evaluation strategies discussed in the previous section are
-the most important techniques for defining equational theory of flat coeffects. In this section, 
-we briefly discuss another approach that follows the style of generic operational semantics 
-designed by Filinski \cite{monads-inaction} for effectful computations.
-
-The idea is to embed sequencing of context-dependent (effectful) computations as an explicit 
-construct into the language and define a \emph{normal form} that consists of sequentially composed 
-context-dependent expressions. The reduction to the normal form is generic for all coeffect systems 
-(satisfying certain conditions), while the reduction of the normal form is provided by each 
-concrete coeffect system. The extended language with the \kvd{glet} construct and its typing is 
-defined as follows:
-%
-\begin{equation*}
-\begin{array}{rcl}
-e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2 \\
-n &::=& e \sep \kvd{glet}~x=e~\kvd{in}~n\\
-\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
-\end{array}
-\end{equation*}
-\begin{equation*}
-\tyrule{glet}
-  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_2 }
-  { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-
-The newly introduced syntactic form $n$ represents a normal form. The construct $\kvd{glet}~x=e~\kvd{in}~n$
-models an explicit sequencing of an expression $e$, followed by an expression $n$. Note that \kvd{glet} can 
-only contain further \kvd{glet} constructs in the body, but not in the argument. The typing of the \kvd{glet}
-is the same as the typing of ordinary \kvd{let} construct -- as discussed in Section~\ref{sec:flat-calculus-types},
-for the examples we consider, this gives a \emph{more precise} typing than the typing of the
-term $(\lambda x.n)~e$.
-
-The reduction rules that produce a normal form are shown in Figure~\ref{fig:flat-normal-form}. The
-(\emph{eval}) rule introduces \kvd{glet} by reducing a redex $(\lambda x.e_2)~e_1$ to an expression
-representing explicit sequencing of $e_1$ and $e_2$. The remaining two rules specify how \kvd{glet}
-distributes across other constructs (\kvd{glet} and application). In (\emph{glet-app}), the \kvd{glet}
-construct appearing inside an application is lifted to the top-level; similarly (\emph{glet-glet}) 
-lifts a \kvd{glet} construct nested in the argument of another \kvd{glet}. The rules provide a way
-for turning an expression into a normal form.
-
-As with \emph{call-by-value} and \emph{call-by-name} strategies, the \emph{internalized sequencing}
-strategy can only be used with coeffect systems satisfying certain conditions. The general form of
-the conditions is summarized in Appendix~?. Here, we briefly consider our three examples.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-
-\begin{equation*}
-\tag{\emph{eval}}
-(\lambda x.e_2)~e_1 ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~e_2
-\end{equation*}
-
-\begin{equation*}
-\tag{\emph{glet-app}}
-\begin{array}{l}
-  (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~\\ \qquad \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
-\end{array}
-\qquad x \notin \textit{fv}(e_3)\qquad
-\end{equation*}
-
-\begin{equation*}
-\tag{\emph{glet-glet}}
-\begin{array}{l}
-  \kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~\\
-  \qquad \kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
-\end{array}
-\qquad x_1 \notin \textit{fv}(e_3)\qquad
-\end{equation*}
-
-\figcaption{Reduction to normal form.}
-\label{fig:flat-normal-form}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\paragraph{Conditions.}
-The (\emph{eval}) reduction can be safely applied for any flat coeffect calculus which satisfies the
-condition that the typing of the \kvd{glet} expression (shown above) is equivalent or more precise
-than typing of the expression $(\lambda x.e_2)~e_1$. More formally:
-
-\begin{definition}
-A flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is \emph{oriented} if for all
-$\cclrd{s}, \cclrd{s_1}, \cclrd{s_2}, \cclrd{r} \in \C$ such that $\cclrd{s}=\cclrd{s_1}\,\czip\,\cclrd{s_2}$
-it is the case that $\cclrd{s} \;\cseq\; (\cclrd{s} \,\cpar\, \cclrd{r}) \;\cleq\; \cclrd{s_1} \;\cseq\; (\cclrd{s_2} \,\cpar\, \cclrd{r})$.
-\end{definition}
-
-\noindent
-As discussed in Section~\ref{sec:flat-calculus-let}, this condition is satisfied for all our
-three examples, strictly for liveness and data-flow and by equality for implicit parameters.
-For the (\emph{glet-app}) and (\emph{glet-glet}) rules, additional conditions arise from the
-typing of the original and reduced expression. The results are summarized in the following two tables.
-
-Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
-$\coctx{\Gamma}{\cclrd{r_2}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{s}} \tau_2$ and
-$\coctx{\Gamma}{\cclrd{r_3}} \vdash e_3 : \tau_3$, the typings of the original and
-reduced expressions in (\emph{glet-app}) rule are:
-
-\begin{center}
-\begin{tabular}{ | l | c | c | c |}
-\hline
-  & \footnotesize Before\hspace{1em} & \footnotesize After\hspace{1em}  & \footnotesize Satisfied\hspace{1em} \\ \hline
-\hspace{-1em}{\footnotesize Parameters} 
-  & $(\cclrd{r_2} \cup \cclrd{r_1}) \cup (\cclrd{r_3} \cup \cclrd{s}) $ 
-  & $(\cclrd{r_2} \cup (\cclrd{r_3} \cup \cclrd{s})) \cup \cclrd{r_1} $ & \checkmark \\ \hline
-\hspace{-1em}{\footnotesize Liveness} 
-  & $\cclrd{r_2} \sqcap (\cclrd{r_3} \sqcup \cclrd{s})$ 
-  & $\cclrd{r_2} \sqcap (\cclrd{r_3} \sqcup \cclrd{s})$ & \checkmark \\ \hline
-\hspace{-1em}{\footnotesize Data-flow} 
-  & $\mathit{max}(\cclrd{r_1} + \cclrd{r_2}, \cclrd{r_3}+\cclrd{s})$
-  & $\cclrd{r_1}+\mathit{max}(\cclrd{r_2}, \cclrd{r_3} + \cclrd{s})$ & $\times$ \\ \hline
-\end{tabular}
-\end{center}
-
-\noindent
-Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
-$\coctx{\Gamma}{\cclrd{r_2}} \vdash e_2 : \tau_2$ and
-$\coctx{\Gamma}{\cclrd{r_3}} \vdash e_3 : \tau_3$, the typings of the original and
-reduced expressions in (\emph{glet-glet}) rule are:
-
-\begin{center}
-\begin{tabular}{ | l | c | c | c |}
-\hline
-  & \footnotesize Before\hspace{1em} & \footnotesize After\hspace{1em}  & \footnotesize Satisfied\hspace{1em} \\ \hline
-\hspace{-1em}{\footnotesize Parameters} 
-  & $\cclrd{r_3} + (\cclrd{r_2} + \cclrd{r_1}) $ 
-  & $(\cclrd{r_3} + \cclrd{r_2}) + \cclrd{r_1}$ & \checkmark \\ \hline
-\hspace{-1em}{\footnotesize Liveness} 
-  & $\cclrd{r_3}$ 
-  & $\cclrd{r_3}$ & \checkmark \\ \hline
-\hspace{-1em}{\footnotesize Data-flow} 
-  & $\cclrd{r_3} \cup (\cclrd{r_2} \cup \cclrd{r_1}) $ 
-  & $(\cclrd{r_3} \cup \cclrd{r_2}) \cup \cclrd{r_1}$ & \checkmark \\ \hline
-\end{tabular}
-\end{center}
-
-\noindent
-This means that \emph{internalized sequencing} provides a basis for operational semantics of
-liveness and implicit parameters, but (again) not for data-flow languages. For other coeffect 
-systems, the conditions arising from the syntactic reductions have to be re-examined.
-
-% ==================================================================================================
-%
-%      #####          #                            #                        
-%      #              #                                                     
-%      #      #   #  ####    ###   # ##    ###    ##     ###   # ##    ###  
-%      ####    # #    #     #   #  ##  #  #        #    #   #  ##  #  #     
-%      #        #     #     #####  #   #   ###     #    #   #  #   #   ###  
-%      #       # #    #  #  #      #   #      #    #    #   #  #   #      # 
-%      #####  #   #    ##    ###   #   #  ####    ###    ###   #   #  ####  
+%    #######
+%    #       #    # ##### ###### #    #  ####  #  ####  #    #  ####
+%    #        #  #    #   #      ##   # #      # #    # ##   # #
+%    #####     ##     #   #####  # #  #  ####  # #    # # #  #  ####
+%    #         ##     #   #      #  # #      # # #    # #  # #      #
+%    #        #  #    #   #      #   ## #    # # #    # #   ## #    #
+%    ####### #    #   #   ###### #    #  ####  #  ####  #    #  ####
 %
@@
 \tyrule{sub}
-  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau & \cclrd{r'} \cleq \cclrd{r} }
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
 \end{equation*}
@@
 
-\subsection{Subtyping for coeffects}
+\subsection{Subcoeffecting and subtyping}
 \label{sec:flat-exts-sub}
 
+\paragraph{Ordering.}
+Add $(\C, \cleq)$ is a pre-order that is:
+
+\begin{equation*}
+\begin{array}{ccc}
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
+t\; \cleq\; t \\[0.5em]
+\end{array}
+\end{equation*}
+
+The flat coeffect algebra requires a pre-order relation $\cleq$, which is used to define 
+sub-coeffecting rule of the type system. When the idempotent monoid $(\C, \cpar, \czero)$ also 
+has the commutative property (\ie~forms a semi-lattice), the $\cleq$ relation can be defined as the
+ordering of the semi-lattice:
+%
+\begin{equation*}
+\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
+\end{equation*}
+%
+This definition is consistent with all three examples that motivate flat coeffect calculus, but
+it cannot be used with the structural coeffects (where it fails for the bounded reuse 
+calculus) and so we choose not to use it.
+
+Furthermore, the $\cunit$ coeffect is often the top or the bottom element of the semi-lattice. 
+As discussed in Section~\ref{sec:flat-syntax}, when this is the case, we are able to prove certain 
+syntactic properties of the calculus.
+
+\paragraph{Something else}
 The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
@@
 
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Typing of let binding}
+\label{sec:flat-exts-let}
+
+Recall the (\emph{let}) rule in Figure~\ref{fig:flat-types}. It annotates the expression 
+$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context requirements $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
+This is a special case of typing an expression $(\lambda x.e_2)~e_1$, using the idempotence
+of $\czip$ as follows:
+%
+\begin{equation*}
+\tyrule{app}
+  {\begin{array}{l}
+   \vspace{-1.5em}
+   \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1
+   \end{array} &
+   \tyruler{abs}
+       { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2 }
+       { \coctx{\Gamma}{\cclrd{s}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+  { \coctx{\Gamma}{\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }    
+\end{equation*}
+%
+This design decision is similar to ML value restriction, but it works the other way round. Our
+\emph{let} binding is more restrictive than the typing of abstraction-application, rather than
+being more general. The choice is motivated by the fact that the typing obtained using the special 
+rule for let-binding is more precise for all the examples consider 
+in this chapter. Table~\ref{tab:flat-simplelet} shows how the coeffect annotations are simplified 
+for our examples.
+
+\begin{table}[!h]
+\begin{center}
+\begin{tabular}{ | l | c | c |}
+\hline
+& \textbf{\footnotesize Definition\hspace{1em}} & \textbf{\footnotesize Simplified\hspace{1em}} \\ \hline
+\hspace{-1em}{\footnotesize Implicit parameters} & $\cclrd{s} \cup (\cclrd{s} \cup \cclrd{r})$ & $\cclrd{s} \cup \cclrd{r}$ \\ \hline
+\hspace{-1em}{\footnotesize Liveness} & $\cclrd{s} \sqcap (\cclrd{s} \sqcup \cclrd{r})$ & $\cclrd{s}$ \\ \hline
+\hspace{-1em}{\footnotesize Data-flow} & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
+\end{tabular}
+\end{center}
+
+\vspace{-0.5em}
+\figcaption{Simplified annotation for let binding in sample flat calculi instances}
+\label{tab:flat-simplelet}
+\end{table}
+
+\noindent
+The simplified annotations directly follow from the definitions of particular flat coeffect 
+algebras. It is perhaps somewhat unexpected that the annotation can be simplified in different
+ways for different examples.
+
+To see that the simplified annotations are more precise, assume that we used arbitrary 
+splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$ rather than idempotence. The
+``Definition'' column would use $\cclrd{s_1}$ and $\cclrd{s_2}$ for the first and second 
+$\cclrd{s}$, respectively. The corresponding simplified annotation would have 
+$\cclrd{s_1}\,\czip\,\cclrd{s_2}$ instead of $\cclrd{s}$. For all our systems, the 
+simplified annotation (on the right) is more precise than the original (on the left):
+%
+\begin{equation*}
+\begin{array}{rclll}
+\cclrd{s_1} \cup (\cclrd{s_2} \cup \cclrd{r}) &\supseteq& (\cclrd{s_1} \cup \cclrd{s_2}) \cup \cclrd{r} 
+  && \textnormal{(implicit parameters)}\\
+\cclrd{s_1} \sqcap (\cclrd{s_2} \sqcup \cclrd{r}) &\sqsupseteq&  (\cclrd{s_1} \sqcap \cclrd{s_2}) 
+  && \textnormal{(liveness)} \\
+\mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r}) &\geq& \mathit{min}(\cclrd{s_1}, \cclrd{s_2}) + \cclrd{r} 
+  &\quad& \textnormal{(data-flow)} \\
+\end{array}
+\end{equation*}
+%
+In other words, the inequality states that using idempotence, we get a more precise typing.
+Using the $\cgeq$ operator this property can be expressed using the abstract operators of the flat coeffect algebra as:
+%
+\begin{equation*}
+\cclrd{s_1} \;\cpar\; (\cclrd{s_2} \,\cseq\, \cclrd{r}) \;\cgeq\; 
+  (\cclrd{s_1} \,\czip\, \cclrd{s_2}) \;\cpar\; ((\cclrd{s_1} \,\czip\, \cclrd{s_2}) \,\cseq\, \cclrd{r})
+\end{equation*}
+%
+This property cannot be proved from other properties of the flat coeffect algebra. To make
+the flat coeffect system as general as possible, we do not \emph{in general} require it as
+an additional axiom, although the above examples provide reasonable basis for requiring 
+that the specialized annotation for let binding is the least possible annotation for the 
+expression $(\lambda x.e_2)~e_1$.
+
 % --------------------------------------------------------------------------------------------------
@@
 \tyrule{idabs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e : \tau_2}
   {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\,\czip\,\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\,\czip\,\cclrd{r}} \vdash e : \tau_2}
   {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
@@
      {\hspace{-4em} \tyrule{idabs}  
-        {\coctx{\Gamma, x:\tau_1}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash e : \tau_2}
+        {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash e : \tau_2}
         {\coctx{\Gamma}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r} \,\czip\, \cclrd{s}} \tau_2 } }
@@
 \inference
-  {\coctx{\Gamma, x:\tau_1, y:\tau_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t}} \vdash e : \tau_3}
+  {\coctx{\Gamma, x\!:\!\tau_1, y:\tau_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t}} \vdash e : \tau_3}
   {\coctx{\Gamma}{\cclrd{r'}} \vdash \lambda x.\lambda y.e : 
@@
 %
-%      ####           ##            #                # 
-%      #   #           #            #                # 
-%      #   #   ###     #     ###   ####    ###    ## # 
-%      ####   #   #    #        #   #     #   #  #  ## 
-%      # #    #####    #     ####   #     #####  #   # 
-%      #  #   #        #    #   #   #  #  #      #  ## 
-%      #   #   ###    ###    ####    ##    ###    ## # 
+%    #####
+%    #     # #   # #    # #####   ##    ####  ##### #  ####
+%    #        # #  ##   #   #    #  #  #    #   #   # #    #
+%     #####    #   # #  #   #   #    # #        #   # #
+%          #   #   #  # #   #   ###### #        #   # #
+%    #     #   #   #   ##   #   #    # #    #   #   # #    #
+%     #####    #   #    #   #   #    #  ####    #   #  ####
 %
@@
 
-\section{Related work}
-\label{sec:flat-related}
-
-Most of the related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
-and we covered work related to individual concepts throughout the chapter. In this section, we
-do not repeat the discussion present elsewhere. Instead, we discuss one specific question that often 
-arises when discussing coeffects and that is \emph{when is a coeffect (not) an effect?}
-
-We start with a quick overview of the ways in which effects and coeffects differ and then
-we briefly look at one (but illustrative) example where the two concepts overlap. We focus 
-mainly on the equivalence between the \emph{categorical semantics}, which reveals the nature
-of the computations -- rather than considering just the syntactic aspects of the type system.
-
-\subsection{When is coeffect not a monad}
-Coeffect systems differ from effect systems in three important ways:
-
-\begin{itemize}
-\item Semantically, coeffects capture different notions of computation. As demonstrated in 
-  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
-  computation, many of which cannot be captured by a monad (\eg~liveness or data-flow).
-
-\item Syntactically, coeffect calculi use a richer algebraic structure with pointwise composition,
-  sequential composition and context merging ($\cpar, \cseq$, and $\czip$) while most effect systems
-  only use a single operation for sequential composition (used by monadic bind).
-
-\item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In 
-  coeffect systems, the context requirements of the body can be split between (or duplicated
-  at) declaration site and call site, while monadic effect systems always defer all effects.
-\end{itemize}
+\section{Syntactic equational theory}
+\label{sec:flat-syntax}
+
+Each of the concrete coeffect calculi discussed in this chapter has a different notion of context,
+much like various effectful languages have different notions of effects (such as exceptions or
+mutable state). However, in all of the calculi, the context has a number of common properties that 
+are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
+that hold for all of the coeffect systems. Further properties hold for systems where the context
+satisfies additional properties.
+
+In this section, we look at such shared syntactic properties. This accompanies the previous section, 
+which provided a \emph{semantic} justification for the axioms of coeffect algebra with a 
+\emph{syntactic} justification. Operationally, this section can also be viewed as providing a 
+pathway to an operational semantics for two of our systems (implicit parameters and liveness),
+which can be based on syntactic substitution. As we discuss later, the notion of context for
+data-flow is more complex.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Syntactic properties}
+\label{sec:flat-syntax-props}
+
+Before discussing the syntactic properties of general coeffect calculus formally, it should be 
+clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
+that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
+Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
+equations define four syntactic reductions on the terms:
+%
+\begin{equation*}
+\begin{array}{rllcl}
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
+(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
+e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
+\end{array}
+\end{equation*}
+%
+The rules capture syntactic reductions that can be performed in a general calculus, without any 
+knowledge of the specific notion of context. If the reductions preserve the type of the expression
+(type preservation), then operational semantics can be defined as a repeated application of the 
+rules, until a specified normal form (\ie~a value) is reached.
+
+The first two equations define well-known call-by-name and call-by-value reductions. The \kvd{glet} 
+reduction (inspired by Filinski \cite{monads-inaction}) models a system where program is reduced to 
+a sequence of primitive operations, sequentially composed using the special \kvd{glet} construct. 
+In the rest of the section, we briefly outline the interpretation of the four 
+rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name 
+(Section~\ref{sec:flat-syntax-cbn}) in more details.
+
+The focus of our work is on the general coeffect system and so we do not discuss the operational
+semantics of the specific notions of context. However, some work in that area has been done
+by Brunel et al. \cite{coeffects-quantitative}.
+
+\paragraph{Call-by-name.} 
+In call-by-name, the argument is syntactically substituted for all occurrences of a variable. It
+can be used as the basis for operational semantics of purely functional languages. However, using
+the rule in effectful languages breaks the \emph{type preservation} property. For example, consider
+a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
+A function $\lambda x.y$ is effect-free:
+%
+\begin{equation*}
+y\!:\!\tau_1 \vdash \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 \;\&\; \emptyset
+\end{equation*}
+%
+Substituting an expression $e$ with effects $\{ \ident{write} \}$ for $y$ changes the type of 
+the function by adding latent effects (without changing the immediate effects):
+%
+\begin{equation*}
+\vdash \lambda x.e : \tau_1 \xrightarrow{ \{ \ident{write} \} } \tau_2 \;\&\; \emptyset
+\end{equation*}
+%
+Similarly to effect systems, substituting a context-dependent computation $e$ for a variable $y$ can
+add latent coeffects to the function type. However, this is not the case for \emph{all} flat coeffect
+calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
+parameters system. This means that certain coeffect systems support call-by-name evaluation strategy
+and could be embedded in purely functional language (such as Haskell).
+
+\paragraph{Call-by-value.}
+
+The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
+first reduced to a \emph{value} before performing the substitution. In effectful languages, 
+value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
+values are identifiers $x$ or functions $(\lambda x.e)$.
+
+The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
+A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
+immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
+sense \emph{and} has not immediate context requirements. We define this formally in 
+Section~\ref{sec:flat-syntax-cbv}.
+
+The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
+including all our three examples. However, it is rather weak -- in order to use it, the concrete 
+semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
+
+\paragraph{Local soundness and completeness.}
+Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
+are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
+introduce a function arrow (lambda abstraction) and eliminate it (application) are sufficiently 
+strong, but not too strong.
+
+The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
+already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
+coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Call-by-value evaluation}
+\label{sec:flat-syntax-cbv}
+
+As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
+calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
+term to a \emph{value} has to be provided for each system. Syntactic values are defined
+in the usual way:
+%
+\begin{equation*}
+\begin{array}{llll}
+v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
+n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
+e\in\mathit{Expr}     & e &::=& v \sep n 
+\end{array}
+\end{equation*}
+%
+The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
+coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
+not automatically defer all context requirements.
+
+\begin{definition} 
+An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if it is a syntactic value,
+\ie~$e \in \mathit{SynVal}$ and it has no context-dependencies, \ie~$\coctx{\Gamma}{ \cunit } \vdash e : \tau$.
+\end{definition}
 
 \noindent
-Despite the differences, our implicit parameters example can be also represented by a monad.
-Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
-we use the $\cup$ operation for all three operations of the coeffect algebra. However, to enable
-splitting of implicit parameter requirements using the reader monad, we need to extend the 
-monad structure and change the translation of monadic lambda abstraction.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{When is coeffect a monad}
-\label{sec:flat-related-monads}
-
-Implicit parameters can be captured by a monad, but \emph{just} a monad is not enough.
-Lambda abstraction in effect systems does not provide a way of splitting the context
-requirements between declaration site and call site (or, semantically, combining the implicit 
-parameters available in the scope where the function is defined and those specified by the caller).
-
-\paragraph{Categorical relationship.}
-Before looking at the necessary extensions, consider the two ways of modelling implicit 
-parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
-for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
-are:
-%
-\begin{equation*}
-\begin{array}{lll}
- \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
- \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
+The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
+variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
+context requirements of the term:
+
+\begin{lemma}[Call-by-value substitution]
+\label{thm:flat-subst-cbv}
+In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
+$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
+change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
+\end{lemma}
+\begin{proof}
+By induction over the type derivation, using the fact that $x$ and $v$ are annotated
+with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
+\end{proof}
+
+\noindent
+The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
+proving that call-by-value reduction preserves typing requires an additional constraint on the
+flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
+by the (\emph{approximation}) property:
+
+\begin{equation*}
+\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
+\hspace{10em}(\textit{approximation})
+\end{equation*}
+%
+Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
+under-approximates the actual context capabilities while the $\cpar$ operation (combining of
+context requirements) over-approximates the actual context requirements.
+
+The property holds for the three systems we consider -- for implicit parameters, this is 
+an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
+is smaller than the least upper bound.
+
+Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
+$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the type preservation theorem is
+stated as follows:
+
+\begin{theorem}[Type preservation for call-by-value]
+\label{thm:cbv-reduction}
+In a flat coeffect system with the (\emph{approxi\-mation}) property, 
+if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
+\begin{equation*}
+\inference
+  { \inference
+  { \inference
+      {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
+      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
+   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
+\end{equation*}
+The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
+From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
+As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
+%
+\begin{equation*}
+\tyrule{sub}
+ { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
+ { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
+\end{equation*}
+%
+Comparing the final conclusions of the above two typing derivations shows that 
+the reduction preserves type and coeffect annotation.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Call-by-name evaluation}
+\label{sec:flat-syntax-cbn}
+
+When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the 
+sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$. 
+As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in 
+general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two 
+interesting cases.
+
+The typing is preserved for different reasons in two of our systems, so we briefly review the 
+concrete examples. Then, we prove the substitution lemma for two special cases of flat coeffects
+(Lemma~\ref{thm:cbn-substitution-top} and Lemma~\ref{thm:cbn-substitution-bot}) and finally, we
+state the conditions under which typing preservation hold for flat coeffect calculi 
+(Theorem~\ref{thm:cbn-flat}).
+
+\paragraph{Data-flow.} 
+The type preservation property does not hold for data-flow. This case is similar to the example
+shown earlier with effectful computations. As a minimal example, consider the substitution of 
+a context-dependent expression $\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1,z\!:\!\tau_1}{0} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{0} \tau_2 & & (\textnormal{before}) \\
+ \coctx{z\!:\!\tau_1}{1} &\narrow{\vdash}& \lambda x.\kvd{prev}~z : \tau_1 \xrightarrow{1} \tau_2 & & (\textnormal{after}) \\
 \end{array}
@@
 %
-The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
-the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
-As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the 
-two representations are equivalent:
-%
-\begin{remark}
-Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
-are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
-reader monad via currying/uncurrying isomorphism.
-\end{remark}
+After the substitution, the coeffect of the body is $1$. The rule for lambda abstraction requires
+that $1=\mathit{min}(r,s)$ and so the least solution is to set both $r,s$ to $1$. The substitution
+this affects the coeffects attached both to the function type and the overall context.
+
+Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
+actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
+system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
+
+\paragraph{Implicit parameters.} In data-flow, there is no typing for the resulting expression that
+preserves the type of the function. However, this is not the case for all systems. Consider substituting
+an implicit parameter access $\ident{?p}$ for a free variable $y$ under a lambda:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1}{\emptyset} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{before}) \\
+ \coctx{\emptyset}{ \{\ident{?p}\} } &\narrow{\vdash}& \lambda x.\ident{?p} : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+The above shows one possible typing of the body -- one that does not change the coeffects of the 
+function type and attaches all additional coeffects (implicit parameters) to the context. In case
+of implicit parameters (and, more generally, systems with set-like annotations) this is always 
+possible.
+
+\paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
+substituting an arbitrary expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1}{\ident{L}} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{before}) \\
+ \coctx{\emptyset}{\ident{L}} &\narrow{\vdash}& \lambda x.e : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+In the original expression, both the overall context and the function type are annotated with \ident{L}, 
+because the body contains a variable access. An expression $e$ can always be treated as being annotated
+with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so we can also treat $e$
+as being annotated with coeffects $\ident{L}$. As a result, substitution does not change the coeffect.
+
+\paragraph{Reduction theorem.}
+The above examples (implicit parameters and liveness) demonstrate two particular kinds of coeffect 
+algebra for which typing preservation holds. Proving the type preservation separately provides 
+more insight into how the systems work. We consider the two cases separately, but find a more general
+formulation for both of them.
+
+\begin{definition}
+We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
+($\forall r \in \C \;.\; r \,\cleq\, \cunit $) and \emph{bottom-pointed} if it is the smallest (bottom) 
+element ($\forall r \in \C \;.\; r \,\cgeq\, \cunit $).
+\end{definition}
+
+\noindent
+Liveness is an example of top-pointed coeffects as variables are annotated with 
+$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and data-flow are examples
+of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
+additional requirements:
+
+\begin{lemma}[Top-pointed substitution]
+\label{thm:cbn-substitution-top}
+In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
+not change the coeffects of $e_r$:
+%
+\begin{equation*}
+\begin{array}{l}
+ \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
+   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
+ \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
+\end{array}
+\end{equation*}
+\end{lemma}
+
 \begin{proof}
-The isomorphism is demonstrated by the following equation:
-\begin{equation*}
-\begin{array}{lll}
- \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
- (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
- &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
- \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
+Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
+\end{proof}
+%
+\noindent
+As variables are annotated with the top element $\cunit$, we can substitute the term $e_s$ 
+for any variable and use sub-coeffecting to get the original typing (because
+$\cclrd{s} \,\cleq\, \cunit$).
+
+In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
+requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
+then the context requirements arising from the substitution can be associated with the context
+$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
+does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
+our implicit parameters example (all three operators are a set union) and for other set-like 
+coeffects. It allows the following substitution lemma:
+
+\begin{lemma}[Bottom-pointed substitution]
+\label{thm:cbn-substitution-bot}
+In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
+where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
+$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
+%
+\begin{equation*}
+\begin{array}{l}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
+\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
 \end{array}
 \end{equation*}
+
+\end{lemma}
+\begin{proof}
+By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
+the free-variable context. See Appendix~\ref{sec:appendix-flat-cbn}.
 \end{proof}
@@
 \noindent
-This equivalence holds for monads and comonads (as well as \emph{indexed} monads
-and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
-the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging.
-
-\paragraph{Delaying effects in monads.} 
-In the syntax of the language, the above difference is manifested by the (\emph{abs}) rules for 
-monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
-side-by-side, using the effect system notation for both of them:
-%
-\begin{equation*}
-\tyrule{cabs}
-  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s} }
-  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 \;\&\; \cclrd{r}}
-\quad
-\tyrule{mabs}
-  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s}}
-  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 \;\&\; \cclrd{\emptyset} }
-\qquad
-\end{equation*}
-%
-In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
-the monadic rule (\emph{mabs}) places all requirements on the call site. This follows from the
-fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
-%
-\begin{equation*}
-\sem{\lambda x.e} \;=\; \ident{unit}~(\lambda x. \sem{e})
-\end{equation*}
-%
-The type of $\ident{unit}$ is $\alpha\rightarrow\mtyp{\alpha}{\cclrd{\emptyset}}$, but in this specific
-case, the $\alpha$ is instantiated to be $\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}$ and so this
-use of \ident{unit} has a type:
-%
-\begin{equation*}
-\ident{unit}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{\emptyset}}(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2})
-\end{equation*}
-%
-In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
-the declaration site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
-call site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
-(which we call \ident{delay}) with the following signature:
-%
-\begin{equation*}
-\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{r}}(\tau_1\rightarrow\mtyp{\cclrd{s}}{\tau_2})
-\end{equation*}
-%
-The operation reveals the difference between effects and coeffects -- intuitively, given a function
-with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
-function, \emph{before} the function actually performs the effectful operation with the effects.
-The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
-from the effect annotation of the body.
-
-Another important aspect of the signature is that the function needs to be indexed by the coeffect
-annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context requirements
-$\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function at run-time.
-
-The operation cannot be implemented in a useful way for most standard monads, but the 
-reader monad is, indeed, an exception. It is not difficult to see how it can be implemented
-when we expand the definitions of $\mtyp{\cclrd{r}}\tau$:
-%
-\begin{equation*}
-\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~
-(\tau_1\rightarrow(\cclrd{r \cup s} \rightarrow \sigma) \rightarrow \tau_2) \rightarrow 
-((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_1\rightarrow(\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2)
-\end{equation*}
-
-\paragraph{Restricting coeffects in comonads.} 
-As just demonstrated, we can extend monads so that the reader monad is capable of capturing
-the semantics of implicit parameters, including the splitting of implicit parameter requirements
-in lambda abstraction. Can we also go the other way round and \emph{restrict} the comonadic
-semantics so that all requirements are delayed as in the (\emph{mabs}) rule, thus modelling 
-fully dynamically scoped parameters?
-
-This is, indeed, possible. Recall that the semantics of lambda abstraction in the flat
-coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
-two contexts (wrapped in an indexed comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried 
-values and additional contextual information (implicit parameters). To obtain the (\emph{mabs}) 
-rule, we can restrict the first parameter, which corresponds to the declaration site context:
-%
-\begin{equation*}
-\begin{array}{rcll}
- \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{r}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{r \cup s}}(\alpha \times \beta)
- &\hspace{4em}(\textit{normal}) \\
- \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{\emptyset}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{s}}(\alpha \times \beta)
- &\hspace{4em}(\textit{restricted})
-\end{array} 
-\end{equation*}
-%
-In the (\emph{restricted}) version of the operation, the declaration site context requires
-no implicit parameters and so all implicit parameters have to be satisfied by the call site.
-The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.
-
-The idea of restricting the operations of the coeffect calculus semantics could be used more
-generally. We could allow any of the coeffect algebra operations $\cseq, \czip, \cpar$ to be
-\emph{partial} and thus the restricted (fully dynamically-scoped) version of implicit parameters
-could be obtained just by changing the definition of $\czip$. Similarly, we could obtain \eg~a
-fully lexically-scoped version of the system. The ability to restrict operations to partial 
-functions has been used in the semantics of effectful computations by Tate~\cite{effects-producer-semantics}.
+The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
+requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
+structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
+coeffects associated with individual variables, thus preserving typing in a complementary way.
+
+The two substitution lemmas discussed above show that the call-by-name evaluation strategy can be 
+used for certain coeffect calculi, including liveness and implicit parameters. Assuming
+$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following theorem holds:
+
+\begin{theorem}[Type preservation for call-by-name]
+\label{thm:cbn-flat}
+In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
+Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \rightarrow_{\ident{cbn}} e'$ then it is also the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+
+For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
+to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
+and $r \,\czip\, t \;=\; r \,\cpar\, t$.
+For bottom-pointed coeffect algebra, consider the typing derivation for the term
+$(\lambda x.e_r)\,e_s$ before reduction:
+%
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r}
+      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{r}} \tau_r } &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash (\lambda x.e_r)\,e_s : \tau_r }
+\end{equation*}
+%
+The derivation uses the idempotence of $\czip$ in the first step, followed by the 
+(\emph{app}) rule. The type of the term after substitution, using Lemma~\ref{thm:cbn-substitution-bot} is:
+%
+\begin{equation*}
+\inference
+  { \coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r & 
+    \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
+  { \coctx{\Gamma, x:\tau_r}{\cclrd{r}\;\cseq\;\cclrd{s}} \vdash \subst{e_r}{x}{e_s} : \tau_s }
+\end{equation*}
+%
+From the assumptions of Lemma~\ref{thm:cbn-substitution-bot}, we know that $\cseq\,=\,\cpar$
+and the operation is idempotent, so trivially:
+$\cclrd{r}\;\cseq\;\cclrd{s} = \cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})$
+\end{proof}
+
+\paragraph{Expansion theorem.}
+The $\eta$-expansion (local completeness) is similar to $\beta$-reduction (local soundness) in that
+it holds for some flat coeffect systems, but not for all. Out of the examples we discuss, it 
+holds for implicit parameters, but does not hold for liveness and data-flow.
+
+Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
+expression $e$ may require no variables (both immediate and latent coeffects are marked as
+$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
+marking the context and function argument as live. In case of data-flow, the immediate coeffects 
+are made larger by the lambda abstraction -- the context requirements of the function value are 
+imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
+the next chapter.
+
+However, $\eta$-expansion preserves the type for implicit parameters and, more generally,
+for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the 
+standard $\eta$-reduction:
+
+\begin{theorem}[Type preservation of $\eta$-expansion]
+In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
+where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
+and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
+\end{theorem}
+\begin{proof}
+The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
+\begin{equation*}
+\inference
+  { \inference
+    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
+      \coctx{x\!:\!\tau_1}{\cunit} \vdash x\!:\!\tau_1 }
+    { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
+  { \inference
+    { \inference
+      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
+      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
+    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+\end{equation*}
+%
+The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
+application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
+original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
+is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
+this is sufficient for showing type preservation.
+\end{proof}
+
+\noindent
+In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
+there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
+these include liveness and implicit parameters. Moreover, for implicit parameters (and more
+generally, any set-like flat coeffect algebra), the $\eta$-expansion holds as well, giving us
+both local soundness and local completeness as coined by Pfenning and Davies \cite{logic-modal-reconstruction}.
 

### 2016-02-03 text/Chapter05.tex
Reorg chapter 4 and 5, adding unique typing to chapter 4

@@
 % ==================================================================================================
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-% ==================================================================================================
-
-\documentclass[
-		twoside,openright,titlepage,numbers=noenddot,headinclude,%1headlines,
-                footinclude=true,cleardoublepage=empty,
-                BCOR=10mm,paper=a4,fontsize=10pt, % Binding correction, paper type and font size
-                ngerman,american, % Languages
-                ]{scrreprt}
-
-\input{classicthesis-config}
-\usepackage{semantic}
-\usepackage{paralist}
-\usepackage{etoolbox}
-\usepackage{stmaryrd}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsthm}
-
-\newcounter{mythmcounter}
-\newcounter{mydefcounter}
-\newcounter{myexpcounter}
-\newtheorem{example}[myexpcounter]{Example}
-\newtheorem{remark}[mythmcounter]{Remark}
-\newtheorem{definition}[mydefcounter]{Definition}
-\newtheorem{theorem}[mythmcounter]{Theorem}
-\newtheorem{lemma}[mythmcounter]{Lemma}
-\newtheorem{proposition}[mythmcounter]{Proposition}
-\newtheorem*{lemma*}{Lemma}
-
-% Formatting for source code & types
-\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
-\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
-\definecolor{strclr}{rgb}{0.25,0.1,0.0}
-\definecolor{prepclr}{rgb}{0.0,0.0,0.0}
-
-\newcommand{\sep}[0]{\; | \;}
-\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
-\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
-\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
-\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
-\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
-\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
-
-\newcommand{\ctyp}[2]{C^{#1}#2}
-\newcommand{\mtyp}[2]{M^{#1}#2}
-\newcommand{\subst}[3]{#1[#2 \leftarrow #3]}
-
-% Typing rule and typing statement (single line)
-\newcommand{\tyrule}[3]{ \inference[\footnotesize{(\emph{#1})}~~]{#2}{#3} }
-\newcommand{\tyruler}[3]{ \inference{#2}{#3}~~\footnotesize{(\emph{#1})} }
-
-% Coeffect algebra
-\definecolor{aclr}{rgb}{0.6,0.3,0.0}
-\newcommand{\aclrd}[1]{\textcolor{aclr}{#1}}
-
-% Coeffect algebra
-\definecolor{cclr}{rgb}{0.0,0.5,0.0}
-\newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}
-
-\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
-\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
-\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }}
-\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
-\newcommand{\czip}{ \textcolor{cclr}{\wedge} } % only used by flat coeffects
-\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
-\newcommand{\cgeq}{ \textcolor{cclr}{\geq} }
-
-\newcommand{\atimes}{ \textcolor{aclr}{\times} }
-\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
-\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq
-
-% Unified
-
-\newcommand{\SH}{ {\textcolor{sclr}{S}} }
-\newcommand{\SP}{ {\textcolor{sclr}{P}} }
-\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
-\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
-\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
-\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
-\newcommand{\sflat}{\star}
-
-\newcommand{\azero}{ \textcolor{aclr}{\bot} }
-\newcommand{\aunit}{ \alift{\cunit} }
-\newcommand{\aweak}{ \alift{\czero} }
-
-\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
-\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
-
-\newcommand{\xtimes}{\hat{\times}}
-
-\newcommand{\ctx}{\textit{ctx}}
-
-% Ordinary context where colouring is done by hand
-\newcommand{\coctx}[2]{ #1 \,\text{\scriptsize @}\, {{\text{$#2$}}}  }
-
-% Simple coeffect systems in the introduction
-\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}
-
-\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}}
-
-\definecolor{sclr}{rgb}{0.2,0,0.8}
-\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }
-
-\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
-\newcommand{\M}{ {\textcolor{cclr}{\mathcal{M}}} }
-
-\newcommand{\alen}[1]{{\aclrd{\textit{len}(}#1\aclrd{)}}}
-\newcommand{\slen}[1]{{\sclrd{\textit{len}(}#1\sclrd{)}}}
-\newcommand{\xlen}[1]{{\textit{len}(#1)}}
-
-\newcommand{\reduce}{\longrightarrow}
-
-\newcommand{\restr}[2]{#1|_{#2}}
-\newcommand{\narrow}[1]{\hspace{-0.5em} #1 \hspace{-0.65em}~}
-
-\definecolor{todoclr}{rgb}{0.8,0,0.5}
-\newcommand{\todo}[1]{\textcolor{todoclr}{ \begin{quotation} \noindent\textbf{TODO:} #1 \end{quotation} }}
-
-\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\catc}{\mathcal{C}}
-\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }
-
-\newcommand{\figcaption}[1]{
-  \vspace{0.5em}
-  \noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
-  \vspace{-2em}
-  \caption{#1}
-}
-% Half line spacing around align
-\newcommand{\zerodisplayskips}{%
-  \setlength{\abovedisplayskip}{0.5em}
-  \setlength{\belowdisplayskip}{0.5em}
-  \setlength{\abovedisplayshortskip}{0.5em}
-  \setlength{\belowdisplayshortskip}{0.5em}}
-\appto{\normalsize}{\zerodisplayskips}
-\appto{\small}{\zerodisplayskips}
-\appto{\footnotesize}{\zerodisplayskips}
-
-% half line spacing around compactitem
-\setlength{\pltopsep}{0.5em}
-\setlength{\plitemsep}{0.3em}
-
-\patchcmd{\thebibliography}{\chapter*}{\section*}{}{}
-
-\begin{document}
-
-\frenchspacing % Reduces space after periods to make text more compact
-\raggedbottom % Makes all pages the height of the text on that page
-\selectlanguage{american} % Select your default language - e.g. american or ngerman
-\pagenumbering{arabic} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
-\pagestyle{plain} % Suppress headers for the pre-content pages
-
-\newcommand{\semdeff}[4]{
-\begin{array}{rcl}
-  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
-  #2 & = & #4 \\
-\end{array}  
-}
-\newcommand{\semdefff}[6]{
-\begin{array}{rcl}
-  #1 & = & #4 \\
-  #2 & = & #5 \\ \cline{1-1}\cline{3-3}
-  #3 & = & #6 \\
-\end{array}  
-}
-\newcommand{\semdef}[2]{
-\begin{array}{rcl}
-  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
-  #1 & = & #2 \\
-\end{array}  
-}
-% ==================================================================================================
-
-\chapter{Translation semantics}
-\renewcommand*\thesection{\arabic{section}}
+
+\chapter{Semantics of flat coeffects}
+\label{ch:transl}
 
@@
 \begin{itemize}
+\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
+  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
+  This provides a deeper insight into how (and why) the calculus works and shows an 
+  intriguing link with effects.
+
 \item 
@@
 
-% --------------------------------------------------------------------------------------------------
+% ==================================================================================================
+%
+%    ###
+%     #  #    # ##### #####   ####  #####  #    #  ####  ##### #  ####  #    #
+%     #  ##   #   #   #    # #    # #    # #    # #    #   #   # #    # ##   #
+%     #  # #  #   #   #    # #    # #    # #    # #        #   # #    # # #  #
+%     #  #  # #   #   #####  #    # #    # #    # #        #   # #    # #  # #
+%     #  #   ##   #   #   #  #    # #    # #    # #    #   #   # #    # #   ##
+%    ### #    #   #   #    #  ####  #####   ####   ####    #   #  ####  #    #
+%
+% ==================================================================================================
 
 \section{Introduction}
+\label{sec:transl-intro}
 
@@
 over-approximation of the actual contextual information that will be accessed when a program runs.
+
+\subsection{Related work}
+
+Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
+systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
+and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
+relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
+$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
+\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
+defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
+of the work on effect systems and monads.
+
+The main purpose of the comonadic semantics presented in this chapter is to provide a motivation 
+for the typing of the flat coeffect calculus. Our approach is inspired by the work of Uustalu and
+Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
+data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
+comonads} annotate the structure with information about the required context, \ie~$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
+This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.
+
+% ==================================================================================================
+%                                              
+%      #####  #                                 
+%        #    #                                 
+%        #    # ##    ###    ###   # ##   #   # 
+%        #    ##  #  #   #  #   #  ##  #  #   # 
+%        #    #   #  #####  #   #  #      #  ## 
+%        #    #   #  #      #   #  #       ## # 
+%        #    #   #   ###    ###   #          # 
+%                                         #   # 
+%                                          ###                                                                         
+% ==================================================================================================
+
+\section{Categorical motivation}
+\label{sec:flat-semantics}
+
+The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
+discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical 
+semantics of context-dependent computations. This is a direction that we explore in this section. 
+Although the development presented here is interesting in its own, our main focus is \emph{using} 
+categorical semantics to motivate and explain the design of flat coeffect calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Categorical semantics}
+
+As discussed in Section~\ref{sec:path-sem}, categorical semantics interprets terms as morphisms
+in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets typing 
+judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
+$\sem{\tau_1 \times \ldots \times \tau_n} \rightarrow \sem{\tau}$.
+
+As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful 
+computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that 
+approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
+for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models 
+partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
+monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$ 
+(state monad). Here, the structure of a strong monad provides necessary ``plumbing'' for composing 
+monadic computations -- sequential composition and strength for lifting free variables into the
+body of computation under a lambda abstraction.
+
+Following similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that 
+(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent 
+computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
+are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
+
+The monadic and comonadic model outlined here represents at most a binary analysis of effects or 
+context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no 
+context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
+$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an 
+over-simplification as we can use \eg~stacks of monad transformers and model functions with
+two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers 
+require defining complex system of lifting to be composable. Consequently, they are usually used
+for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
+capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
+stateful computation).}.
+
+In the next section, we introduce \emph{indexed comonads}, which provide a more precise analysis 
+and let us model computations with context requirements $\cclrd{r}$ as functions 
+$\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Introducing comonads}
+
+In category theory, \emph{comonad} is a dual of \emph{monad}. As already outlined in 
+Section~\ref{sec:path-sem}, we get a definition of a comonad by taking a definition of a monad 
+and ``reversing the arrows''. More formally, one of the equivalent definitions of comonad 
+looks as follows:
+
+\begin{definition}
+A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
+\begin{compactitem}
+\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
+\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
+\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
+  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
+\end{compactitem}
+such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
+\begin{align}
+\tag{\emph{left identity}}
+  \ident{cobind}~\ident{counit} &= \idf{}
+  \\
+\tag{\emph{right identity}}
+  \ident{counit} \circ \ident{cobind}~f &= f
+  \\
+\tag{\emph{associativity}}
+  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
+\end{align}
+\end{definition}
+
+\noindent
+From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
+\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries 
+additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function 
+$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
+the context-dependent function to value(s) in the context and then propagates the context.
+
+As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
+computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.
+
+\begin{example}[Non-empty list]
+A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
+We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The 
+type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
+%
+\begin{equation*}
+\begin{array}{rclll}
+\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
+\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
+\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
+\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{counit} operation returns the head of the non-empty list. Note that it is crucial that
+the list is \emph{non-empty}, because we always need to be able to obtain a value. The \ident{cobind}
+operation defined here returns a list of the same length as the original where, for each element, the 
+function $f$ is applied on a \emph{suffix} list starting from the element. Using a simplified
+notation for list, the result of applying \ident{cobind} to a function that sums elements of a
+list gives the following behaviour:
+%
+\begin{equation*}
+\ident{cobind}~\ident{sum}~(7,6,5,4,3,2,1,0) = (28,21,15,10,6,3,1,0)
+\end{equation*}
+%
+The fact that the function $f$ is applied to a \emph{suffix} is important in order to satisfy the
+\emph{left identity} law, which requires that $\ident{cobind}~\ident{counit}~l = l$.
+
+It is also interesting to examine some data types that do \emph{not} form a comonad. As already
+mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
+because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. The \ident{Maybe} 
+data type defined as $1 + \alpha$ is not a comonad for the same reason. However, if we consider 
+flat coeffect calculus for liveness, it appears natural to model computations as functions 
+$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalise comonads 
+to \emph{indexed comonads}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Generalising to indexed comonads}
+\label{sec:flat-semantics-idx}
+
+The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
+sequential composition, where the annotation $\cunit$ represents a variable access. An indexed 
+comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the annotation 
+$\cclrd{r}$ determines what context is required.
+
+\begin{definition}
+Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
+\emph{indexed comonad} over a category $\catc$ is a triple 
+$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where:
+
+\begin{compactitem}
+\item $\ctyp{\cclrd{r}}{}$ for all $\cclrd{r} \in \C$ is a family of object mappings 
+\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\cunit}{\alpha} \rightarrow \alpha$ 
+\item $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ is a mapping $(\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
+  \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})$
+\end{compactitem}
+such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$:f
+\begin{align}
+\tag{\emph{left identity}}
+  \ident{cobind}_{\cunit, \cclrd{s}}~\ident{counit}_{\cunit} &= \idf{}
+  \\
+\tag{\emph{right identity}}
+  \ident{counit}_{\cunit} \circ \ident{cobind}_{\cclrd{r}, \cunit}~f &= f
+  \\
+\tag{\emph{associativity}}
+\hspace{-10em}
+  \ident{cobind}_{\cclrd{r}\cseq\cclrd{s},\cclrd{t}}~(g \circ \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f) &= 
+    (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g) \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}~f)
+\end{align}
+\end{definition}
+
+\noindent
+Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings 
+$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
+operation is now formed by a \emph{family} of mappings for different pairs of indices 
+$\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations 
+and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
+For the purpose of this thesis, it is sufficient to omit the superscripts and treat $\ident{cobind}$ just 
+as a family of mappings (rather than natural transformations). When this does not introduce ambiguity, 
+we also occasionally omit the subscripts.
+
+The $\ident{counit}$ operation is not defined for all $\cclrd{r} \in \C$, but only for 
+the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
+but this is merely for symmetry and as a useful reminder to the reader. Crucially, this means that 
+the operation is defined only for special contexts.
+
+If we look at the indices in the laws, we can see that the left and right identity 
+require $\cunit$ to be the unit of $\cseq$. Similarly, the associativity law implies the 
+associativity of the $\cseq$ operator.
+
+\paragraph{Composition.}
+The co-Kleisli category that models sequential composition is formed by the unit arrow (provided
+by $\ident{counit}$) together with the (associative) composition operation that composes computations 
+with contextual requirements as follows:
+%
+\begin{equation*}
+\begin{array}{ccl}
+\textnormal{--}\, \hat{\circ} \,\textnormal{--}&\narrow{:}& (\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2) 
+  \rightarrow (\ctyp{\cclrd{s}}{\tau_2} \rightarrow \tau_3) 
+  \rightarrow (\ctyp{\cclrd{r} \cseq \cclrd{s}}{\tau_1} \rightarrow \tau_3) \\
+g \, \hat{\circ} \, f &\narrow{=}& g \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}} f)
+\end{array}
+\end{equation*}
+%
+The composition $\hat{\circ}$ best expresses the intention of indexed comonads. Given two functions
+with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
+requires $\cclrd{r}\,\cseq\,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
+are attached to the input of the composed function.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Examples.}
+
+Any comonad can be turned into an indexed comonad using a trivial monoid. However, indexed comonads
+are more general and can be used with other data types, including indexed \ident{Maybe}.
+
+\begin{example}[Comonads]
+Any comonad $\ctyp{}{}$ is an indexed comonad with an index provided by a trivial monoid $(\{1\},\ast,1)$
+where $1\ast 1 = 1$. The mapping $\ctyp{1}{}$ is the mapping $\ctyp{}{}$ of the underlying comonad. The
+operations $\ident{counit}_1$ and $\ident{cobind}_{1,1}$ are defined by the operations $\ident{counit}$
+and $\ident{cobind}$ of the comonad.
+\end{example}
+
+\begin{example}[Indexed option]
+The \emph{indexed option comonad} is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
+where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
+Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\ctyp{\ident{L}}{\alpha} = \alpha\\
+\ctyp{\ident{D}}{\alpha} = 1\\
+\\
+\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
+\ident{counit}_{\ident{L}}~v = v\\
+\end{array}
+\qquad
+\begin{array}{lcl}
+\multicolumn{3}{l}{
+  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
+    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
+\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
+\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
+\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
+\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \emph{indexed option comonad} models the semantics of the liveness coeffect system discussed in 
+Section~\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
+and $\ctyp{\ident{D}}{\alpha}=1$ models a dead context which does not contain a value. The \ident{counit}
+operation extracts a value from a live context. As in the direct model discussed in Chapter~\ref{ch:appendix},
+the \ident{cobind} operation can be seen as an implementation of dead code elimination. The definition 
+only evaluates $f$ when the result is marked as live and is thus required, and it only accesses $x$ if 
+the function $f$ requires its input.
+
+The indexed family $\ctyp{\cclrd{r}}{}$ in the above example is analogous to the \ident{Maybe}
+(or option) data type $\ident{Maybe}\,\alpha = 1 + \alpha$. As mentioned earlier, this type does not 
+permit (non-indexed) comonad structure, because $\ident{counit}~()$ is not defined. This is not a 
+problem with indexed comonads, because live contexts are distinguished by the (type-level) coeffect 
+annotation and \ident{counit} only needs to be defined on live contexts.
+
+\begin{example}[Indexed product]
+The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
+$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
+As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
+= \alpha \times (\cclrd{r} \rightarrow \rho)$ represents a value $\alpha$ together with a function that
+associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
+The cobind and counit operations are defined as:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
+\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
+\end{array}
+\quad
+\begin{array}{l}
+\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
+    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
+\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+In the definition, we use the notation $(a, g)$ for a pair containing the value of type $\alpha$
+together with $g$, which is a function $\cclrd{r} \rightarrow \rho$. The \ident{counit} operation
+takes a value and a function (with empty set as a domain), ignores the function and extracts the
+value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to 
+restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in oder to get
+implicit parameters required by the argument of $f$ and by the resulting computation, respectively
+(\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed 
+to \ident{cobind} is initially defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is 
+valid in both cases.
+
+The structure of \emph{indexed comonads} is sufficient to model sequential composition of 
+computations that use a single variable (as discussed in Section~\ref{sec:path-sem}). To model 
+full $\lambda$-calculus with lambda abstraction and multiple-variable contexts, we need 
+additional operations introduced in the next section.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Flat indexed comonads}
+\label{sec:flat-semantics-monoidal}
+
+Because of the asymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
+the duality between monads and comonads can no longer help us with defining the additional structure
+required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
+to the context. In application and lambda abstraction, the context is propagated differently than 
+in effectful computations.
+
+To model the effectful $\lambda$-calculus, Moggi~\cite{monad-notions} requires a \emph{strong} monad which
+has an additional operation $\ident{strength} : \alpha\times\mtyp{}{\beta} \rightarrow \mtyp{}{(\alpha\times\beta)}$.
+This allows lifting of free variables into an effectful computation. In Haskell, strength can be expressed
+in the host language and so it is implicit.
+
+To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions} 
+require \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
+%
+\begin{equation*}
+\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
+\end{equation*}
+%
+The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
+contexts and is used to merge the context of the declaration site (containing free variables)
+and the call site (containing bound variable). For example, for implicit parameters, this combines
+the additional parameters defined in the two contexts.
+
+The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
+\emph{splitting} of contexts. In addition, we also need a lifting operation (similar to $\iota$ 
+from Definition~\ref{def:flat-family}) to model sub-coeffecting.
+
+\begin{definition}
+Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+a \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
+equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
+and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
+%
+\begin{compactitem}
+\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
+  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
+\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
+  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
+\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
+  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\end{compactitem}
+\end{definition}
+
+\noindent
+The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
+values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a 
+single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
+\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
+bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful 
+as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
+the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.
+
+The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
+into two separate values. Note that this does not simply duplicate the value, because the additional
+context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide 
+\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often 
+the \emph{least upper-bound}\footnotemark[1].
+
+Finally, $\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ is a family of operations that ``forget'' some part of
+a context. This models the sub-coeffecting operation and lets us, for example, forget some of the
+available implicit parameters, or turn a live context (containing a value) into a dead context (empty).
+
+\paragraph{Alternative definition.}
+Although we do not demand this as a general law, in all our systems, it is the case that
+$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$.
+This special case allows a simpler definition of \emph{indexed flat comonad} by expressing the 
+\ident{split} operation in terms of lifting (sub-coeffecting) as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
+\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
+  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c), 
+    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
+\end{array}
+\end{equation*}
+%
+The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object 
+mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map} 
+for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
+are first and second projections from a two-element pair. To define the 
+$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
+to throw away additional parts of the context and then transform the values in the 
+context.
+
+This alternative is valid for our examples, but we do not use it for two reasons. Firstly, this
+would be the only place where our semantics uses a variable \emph{twice} (in this case $c$). 
+Thus using an explicit \ident{split} means that the structure required by our semantics does not
+need to provide variable duplication and our model could be embedded in linear or affine category.
+Secondly, explicit \ident{split} is similar to the definition that is needed for structural 
+coeffects in Chapter~\ref{ch:structural} and so it makes the connection between the two system 
+easier to see.
+
+\paragraph{Examples.}
+All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:flat-semantics-idx} can
+be extended into \emph{flat indexed comonads}. Note that this is not a \emph{general} statement, 
+because each example requires us to define additional operations, specific for the example.
+
+\begin{example}[Monoidal comonads]
+Just like indexed comonads generalise co\-monads, the additional structure of
+flat indexed comonads generalises symmetric semimonoidal comonads of Uustalu 
+and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
+where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the 
+monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation 
+is defined by duplication and $\ident{lift}_{1,1}$ is the identity function.
+\end{example}
+
+\begin{example}[Indexed option]
+Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively 
+and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined 
+as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations 
+of a flat indexed comonad are defined as follows:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
+\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
+\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
+\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
+\end{array}
+\quad
+\begin{array}{rcl}
+\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
+\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
+\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
+\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
+\end{array}
+\quad
+\begin{array}{rcl}
+\ident{lift}_{\ident{L}, \ident{D}}~v &\narrow{=}& ()\\
+\ident{lift}_{\ident{L}, \ident{L}}~v &\narrow{=}& v\\
+\ident{lift}_{\ident{D}, \ident{D}}~() &\narrow{=}& ()\\
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+Without the indexing, the \ident{merge} operations implements \emph{zip} on option values,
+returning an option only when both values are present. The behaviour of the \ident{split} 
+operation is partly determined by the indices. When the input is \emph{dead}, both values have 
+to be dead (this is also the only solution of $\cclrd{\ident{D}} =\cclrd{r}\sqcap\cclrd{s}$), but when
+the input is \emph{live}, the operation can perform implicit sub-coeffecting and drop one of
+the values.
+
+Explicit sub-coeffecting using the (\emph{sub}) rule is modelled by the \ident{lift} operation.
+This can turn a \emph{live} value $v$ into a dead value $()$, or it can behave as identity.
+The behaviour is, again, determined by the index.
+
+\begin{example}[Indexed product]
+For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
+$\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
+is $\alpha \times (\cclrd{r} \rightarrow \rho)$ where $\rho$ is the type of implicit parameter values.
+The additional operations are defined as:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{split}_{\cclrd{r}, \cclrd{s}}~((a,b), g) &\narrow{=}& ((a, \restr{g}{\cclrd{r}}), (b, \restr{g}{\cclrd{s}}))\\
+\ident{merge}_{\cclrd{r}, \cclrd{s}}~((a, f), (b, g)) &\narrow{=}& ((a, b), f \uplus g)\\
+\ident{lift}_{\cclrd{r'}, \cclrd{r}}~(a, g) &\narrow{=}& (a, \restr{g}{\cclrd{r}})\\
+\end{array}
+\quad
+\begin{array}{l}
+\textnormal{where}~f \uplus g = \\[-0.25em]
+\quad\restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g 
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{split} operation splits the tuple and restricts the function (representing available
+implicit parameters) to the required sub-sets. This corresponds to the definition in terms
+of \ident{lift}, which performs just the restriction. The \ident{merge} operation is more 
+interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
+in Section~\ref{sec:applications-flat-impl}. It merges the values, preferring the definitions from
+the right-hand side (call site) over left-hand side (declaration site). Thus the operation is not
+symmetric.
+
+\begin{example}[Indexed list]
+Our last example provides the semantics of data-flow computations. The flat coeffect algebra 
+is formed by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$. In a 
+non-indexed version, the semantics is provided by a non-empty list. In the indexed semantics,
+the index represents the number of available past values. The data type is then a pair of 
+the current value, followed by $n$ past values. The mappings that form the flat indexed comonad 
+are defined as follows:
+
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{0}\langle a_0 \rangle = a_0
+\\[0.45em]
+\ident{cobind}_{m, n}~f \langle a_0, \ldots a_{m+n} \rangle = \\[-0.25em]
+\quad \langle f \langle a_0, \ldots, a_m \rangle, \ldots, f \langle a_{n}, \ldots, a_{m+n} \rangle \rangle
+\\[0.45em]
+\ident{merge}_{m, n} (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle) = \\[-0.25em]
+\quad \langle (a_0, b_0), \ldots, (a_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle
+\\[0.45em]
+\ident{split}_{m, n} \langle (a_0, b_0), \ldots, (a_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle = \\[-0.25em]
+\quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
+\\[0.45em]
+\ident{lift}_{n', n} \langle a_0, \ldots, a_{n'} \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
+\quad \langle a_0, \ldots, a_n \rangle
+\end{array}
+\begin{array}{l}
+\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[11em]
+~\\
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The reader is invited to check that the number of required past elements in each of the mappings
+matches the number specified by the indices. The index specifies the number of \emph{past} elements
+and so the list always contains at least one value. Thus \ident{counit} returns the element of a
+singleton list.
+
+The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results 
+of the $f$ function, which itself requires $m$ past values. When combining two lists, 
+$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter 
+argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths. 
+Finally, the lifting operation just drops some number of elements from a list.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Properties and related notions}
+
+The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
+calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
+additional properties and other categorical structures that have been proposed mainly in the 
+context of monads and effects and we look how they relate to indexed comonads.
+
+\paragraph{Shape preservation.}
+Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
+this means that the core comonad structure does not provide a way of modeling computations where
+the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
+the length of the list stays the same after applying \ident{cobind}.
+
+~
+
+Indexed comonads are not restricted by this property of comonads. For example, given the indexed
+product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
+the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
+just implicit parameters $\cclrd{s}$.
+
+\paragraph{Families of monads.}
+When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
+\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
+comonads}, is defined as follows.
+
+\begin{definition}
+\label{def:flat-family}
+A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
+  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
+$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
+$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
+certain coherence conditions.
+\end{definition}
+
+\noindent
+A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data 
+types needs to form a comonad separately. For example, our indexed option does not form a family of 
+comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given 
+a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
+an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
+we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
+$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
+
+\paragraph{Parameteric effect monads.}
+Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
+indexed comonads) are closely related to our definition.  Although presented in a more general 
+categorical framework (and using monads), the model defines the \ident{unit} operation only on the 
+unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal 
+structure.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of flat calculus}
+\label{sec:flat-semantics-calculus}
+
+In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
+computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed 
+comonad} structure, we can now define a single uniform semantics that is capable of capturing all 
+our examples, as well as other computations that can be modelled by the structure.
+
+\paragraph{Contexts and functions.}
+The modelling of contexts and functions generalizes the earlier concrete examples. We use the 
+family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed) data-type that wraps the product of 
+free variables of the context and the arguments of functions:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau} 
+  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[t]
+
+\begin{equation*}
+\begin{array}{ll}
+\sem{\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }~\ctx =
+  \pi_i~(\ident{counit}_{\cunit}~\ctx) & (\emph{var})
+\\[0.5em]
+\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
+  \delta~(c_i) & (\emph{const})
+\\[0.5em]
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
+  \hspace{3em}\sem{\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v))
+\\[0.5em]
+\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
+  \hspace{2.5em}
+  \begin{array}{l}  
+  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}}~
+    (\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~(\lambda x.(x,x))~\ctx) \\[-0.25em]
+  \kvd{in}~\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
+      (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~\sem{\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~\ctx_2)
+  \end{array}    
+\\[1.25em]
+\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }~\ctx = & (\emph{sub})\\[-0.25em]
+  \hspace{3em}\sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx)~
+    \hspace{7.5em}(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
+\end{array}
+\end{equation*}
+
+\figcaption{Categorical semantics of the flat coeffect calculus}
+\label{fig:flat-semantics}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Expressions.}
+The definition of the semantics is shown in Figure~\ref{fig:flat-semantics}. For readability, we 
+write the definitions in a simple programming language notation as opposed to the point-free 
+categorical style. However, it can be equally written using just the operations of flat indexed 
+comonad together with $i^{th}$ projection from a tuple represented by $\pi_i$, \emph{curry} and 
+\emph{uncurry}, function composition, value duplication ($\Delta : A \rightarrow A \times A$) and 
+function pairing (given $f:A\rightarrow B$ and $g:C\rightarrow D$ then $f\times g : A\times C \rightarrow B \times D$). 
+These operations can be provided by \eg~a Cartesian-closed category.
+
+The semantics of variable access and abstraction are the same as in the semantics of Uustalu and 
+Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses 
+$\ident{counit}_{\cunit}$ to extract product of free-variables from the context and then projection
+$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) takes the context $\ctx$ and function argument 
+$v$ and merges their additional contexts using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. Assuming
+the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives us a 
+value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
+Assuming that $n$-element tuples are associated to the left, the wrapped context is equivalent to
+$\sigma_1 \times \ldots \times \sigma_n \times \tau_1$, which can then be passed to the body of the
+function.
+
+The semantics of application is more complex. It first duplicates the free-variable product inside the 
+context (using $\ident{map}_{\cclrd{r}}$ and duplication). Then it splits this context using 
+$\ident{split}_{\cclrd{r}, \cclrd{s} \cpar \cclrd{t}}$. The two contexts contain the same variables
+(as required by sub-expressions $e_1$ and $e_2$), but different coeffect annotations. The first
+context (with index $\cclrd{r}$) is used to evaluate $e_1$, resulting in a function 
+$\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. To obtain the result, we compose this with a function
+created by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$ on the semantics of sub-expression
+$e_2$, which is of type $\ctyp{\cclrd{s}\cseq\cclrd{t}}{\sigma_1 \times \ldots \times \sigma_n}
+\rightarrow \ctyp{\cclrd{t}}{\tau_1}$.
+
+Finally, constants (\emph{const}) are modelled by a global dictionary $\delta$ and sub-coeffecting
+is interpreted by dropping additional context from the provided context $\ctx$ using 
+$\ident{lift}_{\cclrd{r}, \cclrd{r'}}$ and providing it to the semantics of the assumption.
+
+\paragraph{Properties.}
+
+The categorical semantics can be used to embed context-dependent computations in functional 
+programming languages, similarly to how monads provide a way of embedding effectful computations.
+More importantly, it also provides validation for the design of the type system developed in 
+Section~\ref{sec:flat-calculus-types}. As stated in the following theorem, the annotations in 
+the type system match those of the semantic functions.
+
+\vspace{2em}
+
+\begin{theorem}[Correspondence]
+\label{thm:flat-correspondence}
+In all of the typing rules of the flat coeffect system, the context annotations $\cclrd{r}$ of typing 
+judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ 
+correspond to the indices of mappings $\ctyp{\cclrd{r}}{}$ in the corresponding semantic function defined 
+by $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}$.
+\end{theorem}
+\begin{proof}
+By analysis of the semantic rules in Figure~\ref{fig:flat-semantics}.
+\end{proof}
+
+\noindent
+Thanks to indexing, the statement of the theorem is significantly stronger than for a 
+non-indexed system, because it provides the justification for our choice of indices in the typing
+rules. In particular, we can see that the annotations follow from the annotations on primitive 
+functions that define the semantics. Also, each function defining the semantics uses a distinct 
+operation of the coeffect algebra and so the type system is the most general possible definition
+(within the comonadic framework we use).
+
+Although the notion of indexed comonads presented in this section is novel and interesting 
+in its own, the main reason for introducing it is to motivate the flat coeffect type system.
+This is captured by the Theorem~\ref{thm:flat-correspondence}. In the next section, we shift our
+focus from (categorical) semantics to syntactic properties of the calculus.
 
@@
 
-\end{document}
+\newpage
+\newpage
+
+% ==================================================================================================
+%                                                 
+%      ####           ##            #                # 
+%      #   #           #            #                # 
+%      #   #   ###     #     ###   ####    ###    ## # 
+%      ####   #   #    #        #   #     #   #  #  ## 
+%      # #    #####    #     ####   #     #####  #   # 
+%      #  #   #        #    #   #   #  #  #      #  ## 
+%      #   #   ###    ###    ####    ##    ###    ## # 
+%
+% ==================================================================================================
+
+\section{Related work}
+\label{sec:flat-related}
+
+Most of the related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
+and we covered work related to individual concepts throughout the chapter. In this section, we
+do not repeat the discussion present elsewhere. Instead, we discuss one specific question that often 
+arises when discussing coeffects and that is \emph{when is a coeffect (not) an effect?}
+
+We start with a quick overview of the ways in which effects and coeffects differ and then
+we briefly look at one (but illustrative) example where the two concepts overlap. We focus 
+mainly on the equivalence between the \emph{categorical semantics}, which reveals the nature
+of the computations -- rather than considering just the syntactic aspects of the type system.
+
+\subsection{When is coeffect not a monad}
+Coeffect systems differ from effect systems in three important ways:
+
+\begin{itemize}
+\item Semantically, coeffects capture different notions of computation. As demonstrated in 
+  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
+  computation, many of which cannot be captured by a monad (\eg~liveness or data-flow).
+
+\item Syntactically, coeffect calculi use a richer algebraic structure with pointwise composition,
+  sequential composition and context merging ($\cpar, \cseq$, and $\czip$) while most effect systems
+  only use a single operation for sequential composition (used by monadic bind).
+
+\item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In 
+  coeffect systems, the context requirements of the body can be split between (or duplicated
+  at) declaration site and call site, while monadic effect systems always defer all effects.
+\end{itemize}
+
+\noindent
+Despite the differences, our implicit parameters example can be also represented by a monad.
+Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
+we use the $\cup$ operation for all three operations of the coeffect algebra. However, to enable
+splitting of implicit parameter requirements using the reader monad, we need to extend the 
+monad structure and change the translation of monadic lambda abstraction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{When is coeffect a monad}
+\label{sec:flat-related-monads}
+
+Implicit parameters can be captured by a monad, but \emph{just} a monad is not enough.
+Lambda abstraction in effect systems does not provide a way of splitting the context
+requirements between declaration site and call site (or, semantically, combining the implicit 
+parameters available in the scope where the function is defined and those specified by the caller).
+
+\paragraph{Categorical relationship.}
+Before looking at the necessary extensions, consider the two ways of modelling implicit 
+parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
+for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
+are:
+%
+\begin{equation*}
+\begin{array}{lll}
+ \ctyp{\cclrd{r}}\tau = \tau \times (\cclrd{r} \rightarrow \sigma) &~\hspace{10em}~& (\emph{product comonad}) \\
+ \mtyp{\cclrd{r}}\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau && (\emph{reader monad})
+\end{array} 
+\end{equation*}
+%
+The \emph{product comonad} simply pairs the value $\tau$ with the lookup function, while
+the \emph{reader monad} is a function that, given a lookup function, produces a $\tau$ value.
+As noted by Orchard \cite{comonads-vs-monads}, when used to model computation semantics, the 
+two representations are equivalent:
+%
+\begin{remark}
+Computations modelled as $\ctyp{\cclrd{r}}{\tau_1}\rightarrow\tau_2$ using the product comonad
+are isomorphic to computations modelled as $\tau_1\rightarrow\mtyp{\cclrd{r}}{\tau_2}$ using the
+reader monad via currying/uncurrying isomorphism.
+\end{remark}
+\begin{proof}
+The isomorphism is demonstrated by the following equation:
+\begin{equation*}
+\begin{array}{lll}
+ \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 &\narrow{=}&
+ (\tau_1 \times (\cclrd{r} \rightarrow \sigma)) \rightarrow \tau_2 \\
+ &\narrow{=}& \tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2) =
+ \tau_1 \rightarrow \mtyp{\cclrd{r}}{\tau_2} \hspace{23em}\qedhere \\
+\end{array}
+\end{equation*}
+\end{proof}
+
+\noindent
+This equivalence holds for monads and comonads (as well as \emph{indexed} monads
+and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
+the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging.
+
+\paragraph{Delaying effects in monads.} 
+In the syntax of the language, the above difference is manifested by the (\emph{abs}) rules for 
+monadic effect systems and comonadic coeffect systems. The following listing shows the two rules 
+side-by-side, using the effect system notation for both of them:
+%
+\begin{equation*}
+\tyrule{cabs}
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s} }
+  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 \;\&\; \cclrd{r}}
+\quad
+\tyrule{mabs}
+  { \Gamma, x\!:\!\tau_1 \vdash e : \tau_2 \;\&\; \cclrd{r}\,\cclrd{\cup}\,\cclrd{s}}
+  { \Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}\,\cclrd{\cup}\,\cclrd{s}} \tau_2 \;\&\; \cclrd{\emptyset} }
+\qquad
+\end{equation*}
+%
+In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
+the monadic rule (\emph{mabs}) places all requirements on the call site. This follows from the
+fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
+%
+\begin{equation*}
+\sem{\lambda x.e} \;=\; \ident{unit}~(\lambda x. \sem{e})
+\end{equation*}
+%
+The type of $\ident{unit}$ is $\alpha\rightarrow\mtyp{\alpha}{\cclrd{\emptyset}}$, but in this specific
+case, the $\alpha$ is instantiated to be $\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}$ and so this
+use of \ident{unit} has a type:
+%
+\begin{equation*}
+\ident{unit}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{\emptyset}}(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2})
+\end{equation*}
+%
+In order to split the implicit parameters of the body ($\cclrd{r \cup s}$ on the left-hand side) between 
+the declaration site ($\cclrd{\emptyset}$ on the outer $\mtyp{}$ on the right-hand side) and the 
+call site ($\cclrd{r \cup s}$ on the inner $\mtyp{}$ on the right-hand side), we need an operation
+(which we call \ident{delay}) with the following signature:
+%
+\begin{equation*}
+\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~(\tau_1\rightarrow\mtyp{\cclrd{r \cup s}}{\tau_2}) \rightarrow \mtyp{\cclrd{r}}(\tau_1\rightarrow\mtyp{\cclrd{s}}{\tau_2})
+\end{equation*}
+%
+The operation reveals the difference between effects and coeffects -- intuitively, given a function
+with effects $\cclrd{r \cup s}$, it should execute the effects $\cclrd{r}$ when wrapping the 
+function, \emph{before} the function actually performs the effectful operation with the effects.
+The remaining effects $\cclrd{s}$ are delayed as usual, while effects $\cclrd{r}$ are removed
+from the effect annotation of the body.
+
+Another important aspect of the signature is that the function needs to be indexed by the coeffect
+annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context requirements
+$\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function at run-time.
+
+The operation cannot be implemented in a useful way for most standard monads, but the 
+reader monad is, indeed, an exception. It is not difficult to see how it can be implemented
+when we expand the definitions of $\mtyp{\cclrd{r}}\tau$:
+%
+\begin{equation*}
+\ident{delay}_{\cclrd{r}, \cclrd{s}}~:~
+(\tau_1\rightarrow(\cclrd{r \cup s} \rightarrow \sigma) \rightarrow \tau_2) \rightarrow 
+((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_1\rightarrow(\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2)
+\end{equation*}
+
+\paragraph{Restricting coeffects in comonads.} 
+As just demonstrated, we can extend monads so that the reader monad is capable of capturing
+the semantics of implicit parameters, including the splitting of implicit parameter requirements
+in lambda abstraction. Can we also go the other way round and \emph{restrict} the comonadic
+semantics so that all requirements are delayed as in the (\emph{mabs}) rule, thus modelling 
+fully dynamically scoped parameters?
+
+This is, indeed, possible. Recall that the semantics of lambda abstraction in the flat
+coeffect calculus is modelled using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. The operation takes
+two contexts (wrapped in an indexed comonad $\ctyp{\cclrd{r}}\alpha$), combines their carried 
+values and additional contextual information (implicit parameters). To obtain the (\emph{mabs}) 
+rule, we can restrict the first parameter, which corresponds to the declaration site context:
+%
+\begin{equation*}
+\begin{array}{rcll}
+ \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{r}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{r \cup s}}(\alpha \times \beta)
+ &\hspace{4em}(\textit{normal}) \\
+ \ident{merge}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& \ctyp{\cclrd{\emptyset}}\alpha \times \ctyp{\cclrd{s}}\beta \rightarrow \ctyp{\cclrd{s}}(\alpha \times \beta)
+ &\hspace{4em}(\textit{restricted})
+\end{array} 
+\end{equation*}
+%
+In the (\emph{restricted}) version of the operation, the declaration site context requires
+no implicit parameters and so all implicit parameters have to be satisfied by the call site.
+The semantics using the restricted version corresponds to the (\emph{mabs}) rule shown above.
+
+The idea of restricting the operations of the coeffect calculus semantics could be used more
+generally. We could allow any of the coeffect algebra operations $\cseq, \czip, \cpar$ to be
+\emph{partial} and thus the restricted (fully dynamically-scoped) version of implicit parameters
+could be obtained just by changing the definition of $\czip$. Similarly, we could obtain \eg~a
+fully lexically-scoped version of the system. The ability to restrict operations to partial 
+functions has been used in the semantics of effectful computations by Tate~\cite{effects-producer-semantics}.
+
+~

### 2016-02-03 text/Chapter06.tex
Reorg chapter 4 and 5, adding unique typing to chapter 4

--- 2014-12-12 text/Chapter06.tex
@@
-%!TEX root = ../main.tex
-
-\chapter{Unified coeffect language ???} 
-\label{ch:unified}
-
-%===================================================================================================
+\chapter{The structural coeffect calculus} 
+\label{ch:structural}
+
+In this chapter, we take the next step towards a unified theory of context-aware programming
+languages. As already discussed, we attempt to find abstractions that are \emph{general enough} 
+to capture a wide range of useful programming language features, but \emph{specific enough} to 
+let us identify and capture interesting properties of the context.
+
+In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware programming
+languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
+calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
+and we introduce the \emph{structural coeffect calculus}.
+
+The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness 
+and data-flow), but provides relatively imprecise approximation and weak syntactical properties. 
+Some of those examples can be also seen as per-variable properties. For those, structural coeffect
+systems give a more precise information about the context. However, we also look at other applications 
+that arise from the work on sub-structural logics, a pathway to coeffects discussed in 
+Section~\ref{sec:path-logic}.
+
+As we will see, the categorical semantics of the structural coeffect calculus is based on the
+same underlying structure as the semantics of the flat calculus. However, the syntactical properties
+of the calculus are quite different and so it is worth considering it in a separate chapter.
+
+\paragraph{Chapter structure and contributions}
+\begin{itemize}
+\item We present a \emph{structural coeffect calculus} as a type system that is parameterized 
+  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the 
+  system captures fine-grained liveness and data-flow information, as well as a calculus for bounded 
+  reuse (checking how many times is a variable accessed).
+
+\item We extend indexed comonads introduced in the previous chapter to 
+  \emph{structural indexed comonads} and use them to provide the semantics of structural 
+  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
+  semantics provides a motivation for the design of the calculus.
+
+\item We discuss the equational theory of the calculus. For structural coeffects, the substitution 
+  lemma holds for all instances of the calculus. Thus, we prove the type-preservation property 
+  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}). 
+\end{itemize}
+
+% ==================================================================================================
+%                                                                                      
+%     ###           #                       #                 #       #                 
+%      #            #                       #                 #                         
+%      #    # ##   ####   # ##    ###    ## #  #   #   ###   ####    ##     ###   # ##  
+%      #    ##  #   #     ##  #  #   #  #  ##  #   #  #   #   #       #    #   #  ##  # 
+%      #    #   #   #     #      #   #  #   #  #   #  #       #       #    #   #  #   # 
+%      #    #   #   #  #  #      #   #  #  ##  #  ##  #   #   #  #    #    #   #  #   # 
+%     ###   #   #    ##   #       ###    ## #   ## #   ###     ##    ###    ###   #   # 
+%                                                                                     
+% ==================================================================================================
 
 \section{Introduction}
-
-The main goal of this thesis is to provide a \emph{unified} calculus for tracking context 
-dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
-two kinds of contextual properties that we further covered separately -- flat coeffects in 
-Chapter~\ref{ch:flat} track whole-context properties and structural coeffects, covered in 
-Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify
-the two notions. We introduce a \emph{unified coeffect} calculus that generalizes the two systems
-and can be instantiated to track both flat and structural properties (Section~\ref{sec:unified-unified}).
-
-Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
-believe that coeffects should be integrated in main-stream programming languages. In the second
-part of this chapter, we discuss an alternative approach to defining coeffect 
-systems which highlights the relationship between our work and related work arising from modal logics 
-(Section~\ref{sec:unified-meta}). Finally, we outline one possible approach for practical
-implementations of coeffects (Section~\ref{sec:unified-impl}).
-
-% ==================================================================================================
-
-\section{The unified coeffect calculus}
-\label{sec:unified-unified}
-
-The flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural coeffect calculus 
-differ in a number of ways (Figure~\ref{fig:struct-types}). Understanding the differences is the 
-key to reconciling the two systems:
-
-\begin{itemize}
-\item Structural coeffect calculus contains explicit rules for context manipulation
-  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
-  explicitly, but are admissible.
-
-\item In the structural coeffect calculus, the variable context is treated as a vector
-  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
-  the variable context is a set and is annotated with a single (scalar) coeffect.
-
-\item In the flat coeffect calculus, we distinguish between splitting of the context requirements
-  and merging of context requirements ($\cpar$ and $\czip$, respectively). In the structural
-  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
-  and so the structural coeffect algebra requires just $\atimes$.
-\end{itemize}
-
-\noindent
-In the unified calculus presented in this section, we address the three differences as follows.
-We use calculus with explicit rules for context manipulation. In systems that arise from flat
-calculi, the rules can be applied freely without changing the coeffects. We generalize the 
-structure of coeffect annotations using the notion of a ``container'' which can be specialized 
-to obtain a single annotation or a vector of annotations. Finally, we distinguish between 
-splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
-respectively). For structural coeffect calculi, the two operators coincide, but for flat
-coeffect calculi, they provide the needed flexibility.
-
-\subsection{Shapes and containers}
-
-Our notion of a \emph{coeffect container} is based on the idea of a container introduced by 
-Abbott et al. \cite{types-containers}. Interestingly, the work on containers has later been linked
-to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
-types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of lists).
-For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
-specified length). More formally:
+\label{sec:structural-intro}
+
+Compared to the previous chapter, structural coeffect calculi we consider are more 
+homogeneous and so finding the common pattern is in some ways easier. However, the systems
+are more complicated as they need to keep annotations attached to individual variables
+and thus require explicit structural rules. Before looking at the system, we briefly consider
+the most important related work.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Related work}
+
+In the previous chapter, we discussed the correspondence between coeffects and effects (and 
+between comonads and monads). As noted earlier, the $\lambda$-calculus is asymmetric in that
+an expression has multiple inputs (variables in the context), but just a single result 
+(the resulting value) and so monads and effects have no notion directly corresponding to 
+structural coeffect systems.
+
+The work in this chapter is more closely related to sub-structural type systems \cite{substruct-attpl-intro}.
+While sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
+\emph{exchange} rules, our systems keep them, but use them to manipulate both the context
+and its annotations.
+
+Our work follows the ``language semantics'' style in that we provide a structural semantics to
+the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in 
+the meta-language style, which extends the terms and types of a language with constructs for explicitly
+manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
+may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT, 
+$A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect 
+annotations as stand-alone types (indexed comonads only appear in the semantics).
+
+Our structural coeffect systems annotate the whole variable context with a \emph{vector} of annotations.
+For example, a context with variables $x$ and $y$ annotated with $\cclrd{s}$ and $\cclrd{t}$, 
+respectively is written as $\coctx{x\!:\!\tau_1,y\!:\!\tau_2}{\alift{\cclrd{s},\cclrd{t}}}$.
+A benefit of this approach is that the typing judgements have the same structure as those of the 
+flat coeffect calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify 
+the two systems.
+
+% =================================================================================================
+%                                                            
+%       ###           ##                   ##                 
+%      #   #           #                    #                 
+%      #       ###     #     ###   #   #    #    #   #   ###  
+%      #          #    #    #   #  #   #    #    #   #  #     
+%      #       ####    #    #      #   #    #    #   #   ###  
+%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
+%       ###    ####   ###    ###    ## #   ###    ## #  ####  
+%                                                           
+% =================================================================================================
+
+\section{Structural coeffect calculus}
+\label{sec:struct-calculus}
+
+In the structural coeffect calculus, a vector of variables forming the free-variable context is 
+annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
+for different coeffect calculi and their properties are captured by the definition of \emph{structural 
+coeffect scalar} below. The scalar annotations can be \eg~integers (how many past values we 
+need) or values of a two-point lattice specifying whether a variable is live or not.
+
+Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in 
+the previous chapter). Functions always have exactly one input variable and so they are annotated
+with a single coeffect scalar. Thus the expressions and types of structural coeffect calculi are the 
+same as in the previous chapter. The only difference is that $\cclrd{r}, \cclrd{s}, \cclrd{t}$ now
+range over values of a \emph{structural coeffect scalar} rather than over values of a \emph{flat 
+coeffect algebra} (the definitions are related, but not the same).
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\end{array}
+\end{equation*}
+%
+In the previous chapter, the free variable context $\Gamma$ was treated as a set. In the 
+structural coeffect calculus, the order of variables matters. Thus we treat a free variable 
+context as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
+of the vector:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \Gamma &\narrow{=}& \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
+ \xlen{\Gamma} &\narrow{=}& n
+\end{array}
+\end{equation*}
+%
+For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
+typing judgements, but the free variable context should be understood as a vector.
+Furthermore, the notation $\Gamma_1, \Gamma_2$ is here seen as the tensor product.
+Given $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
+$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
+$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
+
+Free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
+we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
+over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
+and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
+is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
+of coeffects is provided by the definition of \emph{structural coeffect algebra} below.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffect algebra}
+
+The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the 
+exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
+modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
+pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.
 
 \begin{definition}
-A \emph{container} $S \triangleleft P$ is given by a set $S$ of shapes and a shape-indexed family
-$P : S \rightarrow \ident{Set}$ of positions.
-\end{definition}
-
-\noindent
-Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams and 
-singleton data type (which contains exactly one element). Two containers relevant to our work
-are lists and singleton data types:
-
-\begin{itemize}
-\item The container representing lists is given by $S \triangleleft P$ where 
-  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
-  given length $n$ is a set of indices $P n = \{ 1 \ldots n \}$.
-
-\item The contianer representing singleton data type is given by $S \triangleleft P$ where
-  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
-  shape $\ast$ contains exactly one position $P \ast = \{ 0 \}$.
-\end{itemize}
-
-\noindent
-In the unified coeffect calculus, the structure of coeffect annotations is defined by a 
-container with additional operations (discussed later) that links it with the free-variable 
-context $\Gamma$.
-
-\subsection{Structure of coeffects}
-In the structural coeffect calculus, the structure annotation was formed by a vector of coeffect
-sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
-\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
-a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
-(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
-\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
-define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
-are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
-algebra}} which consists of shape-indexed coeffect scalar values.
-
-As in the structural coeffect calculus, the contexts in the unified calculus are annotated with 
-shape-indexed coeffects, written as $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$; 
-functions take just a single input parameter and so are annotated with scalar coeffect values 
-$\sigma \xrightarrow{\cclrd{r}} \tau$.
-
-\paragraph{Coeffect scalar.}
-The following definition of the coeffect scalar structure repeats the Definition~\ref{def:structural-scalar}
-from the previous chapter.
-
-\begin{definition}
-A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
+\label{def:structural-scalar}
+A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
 $\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
@@
 \textnormal{(pre-order)}   
-\end{array}
-\end{equation*}
-\end{definition}
-
-\noindent
-As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
-an underlying category structre; $\cunit$ and $\czero$ represent an accessed and unused variable, 
-respectively.
-
-The $\cpar$ operation models combining of context requirements arising from multiple parts of a 
-program. The meaning depends on the coeffect container. The operation can either combine requirements
-of individual variables (structural) or requirements attached to the whole context of multiple
-sub-exp\-ressions (flat).
-
-\paragraph{Coeffect containers.}
-The coeffect container is a container that determines how are scalar coeffect annotations attached 
-to free-variable contexts. In addition to a container $\SHP$ with shapes and shape-indexed positions, the
-coeffect container also provides a mapping that returns the shape of a free-variable context. The
-mapping between the shape of the variable context and the shape of the coeffect annotation
-is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
-shape $\SH = \{ \ast \}$.
-
-In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
-$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
-We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation $\SP\sclrd{s}$
-returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in \SP\sclrd{s} \rightarrow \C$ 
-as a mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
-the exponent $\aclrd{\textbf{r}} \in \C^{\SP\sclrd{s}}$.
-
-The coeffect container is also equipped with an operation that appends shapes (when we concatenate
-variable contexts) and two special shapes in $\SH$ representing empty context and singleton context.
-
-\begin{definition}
-A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$ 
-comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a 
-mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
-\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.
-
-The elements $\sempty$ and $\sunit$ represent the shapes of empty and singleton free-variable contexts 
-respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts. Given
-$\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
-we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
-\end{definition}
-
-\noindent
-As said earlier, we use two kinds of coeffect containers that describe the structure of vectors 
-(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):
-
-\begin{example}
-\label{ex:unified-struct-shape}
-Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
-where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
-number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
-respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
-{|}\Gamma_1{|} + {|}\Gamma_2{|}$.
-
-Therefore, a coeffect annotation is a \emph{vector} 
-$\aclrd{\textbf{r}} \in \C^{\SP\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
-for each position (corresponding to a variable $x_i$ in the context).
-\end{example}
-
-\begin{example}
-\label{ex:unified-flat-shape}
-Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
-The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
-with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.
-
-That is, there is a single shape
-$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
-Therefore, a coeffect annotation is drawn from $\C^\{\sflat\}$ which is isomorphic to $\C$
-and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
-\end{example}
-
-\newpage
-\begin{example}
-Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
-\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP \sflat = \emptyset$,
-${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.
-
-This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
-$\SP\sflat=\emptyset$ and so coeffect annotations would be from the set $\C^\emptyset$.
-\end{example}
-
-\paragraph{Unified coeffect algebra.}
-The coeffect calculus annotates judgments with shape-indexed coeffect annotations.
-The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and coeffect container
-to define shape-indexed coeffects and operations for manipulating these.
-
-The definition here reconciles the third point discussed in Section~\ref{sec:unified-unified} --
-the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
-while structural coeffects use tensor $\atimes$. In the unified calculus, we use two operators
-that can, however, be coincide.
-
-\begin{definition}
-Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
-\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
-coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP \sempty}$
-is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
-combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
-%
-\begin{equation*}
-\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
-\end{equation*}
-\end{definition}
-
-\noindent
-A coeffect algebra induces the following three additional operations:
-%
-\begin{equation*}
-\begin{array}{rclcrcl}
- \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
- \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
-\\[-0.5em]
- \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
- \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
-\\[-0.5em]
- \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
- \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
-\end{array}
-\end{equation*}
-%
-The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
-the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
-vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
-as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
-$\sclrd{m}$ and write $\aseq$.
-
-Finally, we define $\alen{}$ as an operation that returns the shape of a given shape-indexed 
-coeffect. The only purpose is to simplify notation, as we often need to specify that shapes of 
-variable context and coeffect match, \eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.
-
-\paragraph{Splitting and merging coeffects.}
-The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts. 
-For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
-\C^{\SP\sclrd{m}}$ and $\aclrd{\textbf{s}} \in \C^{\SP\sclrd{n}}$. In the structural system, the 
-context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
-combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined 
-coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}} 
-\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.
-
-For structural coeffect systems such as bounded reuse, both $\apar$ and $\azip$ are
-just the tensor product $\times$ of vectors. For flat coeffect systems, the operations
-can be defined independently, letting $\azip=\czip$ and $\apar=\cpar$.
-
-The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:unified-semantics}), 
-where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of 
-a morphism that merges the capabilities provided by two contexts (in the syntactic 
-reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$ 
-is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single 
-context into two parts (in the syntactic reading, merges their context requirements). 
-Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and 
-$\apar$ in \emph{conclusions}.
-
-% --------------------------------------------------------------------------------------------------
-
-\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
-\newcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
-\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
-
-\begin{figure}[t]
-
-{\small a.) Syntax-driven typing rules:}
-\begin{equation*}
-\tyrule{var}
-  { }
-  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{const}
-  {c \!:\! \tau \in \Delta}
-  {\coctx{()}{\azero} \vdash c : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
-   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\apar\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\azip\,\alift{\cclrd{s}}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\azip\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\apar\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-
-\vspace{1.5em}
-{\small b.) Structural rules for context manipulation:}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-(\cclrd{s'} \cleq \cclrd{s})
-\end{equation*}
-\begin{equation*}
-\tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-\begin{array}{l}
- \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
- \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{2.5em} 
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-~
-\begin{array}{l}
- \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
- \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-
-\vspace{0.5em}
-\caption{Type system for the unified coeffect calculus}
-\label{fig:unif-types}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Unified coeffect type system}
-\label{sec:unified-types}
-The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
-shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
-the key differences between the two.
-
-The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
-$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
-and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
-As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
-use of variables -- the associativity is built-int. The order of variables matters, but can be changed 
-using the structural rules. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}} \in \C^\SH$ 
-are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
-annotated with coeffects scalars. % $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.
-
-\paragraph{Syntax-driven rules.} 
-The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
-differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
-coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$. 
-The (\emph{const}) rule annotates empty context with a special annotation of the unit shape.
-
-In a structural system, the two annotations correspond to a singleton and empty vector, respectively.
-However, for a singleton shape with one position, the annotations are equivalent to annotating
-variables with a scalar $\cunit$ and constants with a scalar $\czero$.
-
-In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
-system is that vector concatentation $\atimes$ is now replaced with context splitting/merging 
-of the unified coeffect algebra. In particular, we use splitting of context requirements $\azip$
-in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.
-
-Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As 
-discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read 
-the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual 
-information and conclusions split available contextual information.
-
-\paragraph{Structural rules.}
-The merging/splitting operations in the structural rules are changed in the same way as in the
-rules above. It is also worth noting that structural coeffect system used vectors of multiple 
-elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch}) 
-rule denoted a two-element vector. In the unified system, this is replaced with merging/splitting
-of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.
-
-In structural systems, this means the same thing -- we are simply concatenating or splitting two 
-singleton vectors. However, the generalization allows us to capture flat coeffect systems as well.
-The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
-$\apar$ correspond to operations on coeffect scalars. And, as discussed in Section~\ref{sec:unified-flat},
-thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not 
-affect the coeffect annotation are admissible for all flat systems.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural coeffects}
-\label{sec:unified-structural}
-
-The unified coeffect system uses a general notion of context shape, but it has been designed with 
-structural and flat systems in mind. In this and the next section, we show how it captures 
-the two kinds of systems.
-
-The unified calculus is modelled after the structural system and so using it to model structural
-systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector} 
-of annotations (Example~\ref{ex:unified-struct-shape}) and define a coeffect algebra formed by
-a vector (free monoid) of scalars.
-
-\begin{definition}
-\label{def:unified-struct}
-Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system} 
-is defined by:
-
-\begin{itemize}{}
-\item[--] Coeffect container $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$ where $\SH=\mathbb{N}$ and
-  $\SP \sclrd{n}=\{ 1 \ldots n \}$ and  ${|}x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n{|}=n$.
-
-\item[--] Coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
-  shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
-  That is $\times : \C^{\SP\sclrd{n}} \times \C^{\SP\sclrd{m}} \rightarrow \C^{\SP(\sclrd{n + m})}$ 
-  appends vectors and $\epsilon : \C^{\SP\sclrd{0}}$ represents empty vectors.
-\end{itemize}
-\end{definition}
-
-\noindent
-The definition is valid since the shape operations form a monoid
-$\sclrd{(\mathbb{N}, +, 0)}$ and $\slen{\textnormal{--}}$ (calculating the length of
-a list) is a monoid homomorphism from the free monoid to the monoid of
-shapes.
-
-\paragraph{Properties.}
-An important property of the unified system is that, when used in a structural way as discussed
-above, it gives calculi with the same properties as the structural system described in 
-Chapter~\ref{ch:structural}. This can be easily seen by comparing the Figure~\ref{fig:unif-types}
-with the Figure~\ref{fig:struct-types} and using the free monoid interpretation of the unified
-coeffect algebra.
-
-\begin{remark}
-The system described in Definition~\ref{def:unified-struct} is equivalent to the structural coeffect
-system described in Figure~\ref{fig:struct-types}. That is, a typing derivation using a structural
-coeffect embedded in the unified system is valid if and only if the corresponding derivation is
-valid in the structural system.
-\end{remark}
-
-\noindent
-Using the above definition, our unified coeffect system can capture per-variable coeffect properties
-discussed in Chapter~\ref{ch:applications}, Section~\ref{sec:applications-structural}. This includes
-the system for bounded reuse (which is only used in the structural form) and precise tracking 
-of per-variable dataflow and liveness.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Flat coeffects}
-\label{sec:unified-flat}
-
-The same unified coeffect system can be used to capture systems that track whole-context (flat) 
-coeffects such as implicit parameters. This is achieved using a singleton-shaped container for
-coeffect annotations. The resulting system is has explicit structural rules (and is syntactically
-different from the standard flat coeffect system), but we show that they are equivalent.
-
-Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:unified-flat-shape})
-In this setting, the context annotations $\C^{ \SP \sflat } = \{0\} \rightarrow \C$ are equivalent 
-to coeffect scalars $\C$. In addition to the coeffect scalar structure, we also need to define $\apar$ 
-and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ 
-(merging of shaped coeffect annotations). The $\azip$ operation (corresponding to $\czip$
-in flat coeffect calculus) needs to be provided explicitly.
-
-\begin{definition}
-\label{def:unified-flat}
-Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary operation 
-$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$, 
-the unified coeffect algebra modelling flat coeffect systems consists of:
-%
-\begin{itemize}{}
-\item[--] Flat coeffect container $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ as defined in 
-  Example~\ref{ex:unified-flat-shape}.
-
-\item[--] Flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$ 
-  and $\azero = \czero$ with an additional binary operation $\azip = \czip$. 
-\end{itemize}
-\end{definition}
-
-\noindent
-Intuitively, the requirement $(\cclrd{r}\,\czip\,\cclrd{s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$,
-which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
-denotes that splitting context requirements and then re-combining them preserves all the requirements
-from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
-it cannot lose requirements. This is fundamental for showing that exchange and contraction are admissible
-in the unified system.
-
-\paragraph{Properties.}
-To show that the typing of flat properties in the unified system is equivalent to the typing in the
-flat system, we show that a valid typing judgement in the first system is a valid typing judgement in 
-the second system and vice versa.
-
-In one direction, we show that the unified system (capturing flat properties) permits weakening, 
-contraction and exchange rules that do not change the coeffect annotations. This guarantees that
-a valid judgement in flat system is also valid in the unified system.
-
-\begin{lemma}
-\label{thm:unified-weak}
-A unified coeffect calculus capturing flat properties admits weakening that does not change the
-coeffect annotation.
-\end{lemma}
-\begin{proof}
-The rule is admissible using the following derivation:
-\begin{equation*}
-\inference
-  { \inference
-      {\coctx{\Gamma}{ \cclrd{r} } \vdash e : \tau}
-      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \;\apar\; \alift{ \czero }} \vdash e : \tau} }
-  { \inference
-      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \,\cpar\, \czero} \vdash e : \tau}      
-      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r}} \vdash e : \tau} }
-\end{equation*}
-
-\noindent
-We write $\cclrd{r}$ rather than $\aclrd{\textbf{r}}$, because we are tracking flat properties.
-The first step is an application of the (\emph{weak}) rule. Next, we use the fact that
-$\apar=\cpar$ and $\alift{\czero}=\czero$, which is the unit of the monoid $(\C, \cpar, \czero)$.
-\end{proof}
-
-\begin{lemma}
-A unified coeffect calculus capturing flat properties admits exchange that does not change the
-coeffect annotation.
-\end{lemma}
-\begin{proof}
-We use the idempotence of $\czip$ and $\cpar$ together with the (\emph{exch}) rule:
-\begin{equation*}
-\inference
-  { \inference
-      { \inference
-          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r}} \vdash e : \tau}
-          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r}} \vdash e : \tau} }
-      { \coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau } }
-  { \inference 
-      {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r}} \apar \alift{\cclrd{r}} \apar \cclrd{r}} \vdash e : \tau}
-      {\inference
-         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}} \vdash e : \tau}
-         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}} \vdash e : \tau} } }
-\end{equation*}
-
-\noindent
-Using idempotence, we first duplicate the annotation $\cclrd{r}$ to get a coeffect in the form
-$\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}$ as required by the
-assumption of the (\emph{exch}) rule. Note that $\alift{-}$ can be added freely as
-$\alift{\cclrd{r}}$ is equivalent to $\cclrd{r}$. After applying (\emph{exch}), we use 
-idempotence of $\apar$.
-\end{proof}
-
-\begin{lemma}
-A unified coeffect calculus capturing flat properties admits contraction that does not change the
-coeffect annotation.
-\end{lemma}
-\begin{proof}
-Similarly to exchange, the proof uses idempotence of $\czip$ and $\cpar$:
-\begin{equation*}
-\inference
-  { \inference
-      {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \czip\, \cclrd{r} \czip\, \cclrd{r} \czip\, \cclrd{r}} \vdash e : \tau}
-      {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau} }
-  { \inference
-      {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r} \,\cpar\, \cclrd{r}} \apar \cclrd{r}} \vdash \subst{e}{z,y}{x} : \tau}
-      {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{r} \,\cpar\, \cclrd{r} \,\cpar\, \cclrd{r}} \vdash \subst{e}{z,y}{x} : \tau} }
-\begin{array}{r} \\[2.5em] \qedhere \end{array}
-\end{equation*}
-\end{proof}
-
-\noindent
-In the last two cases, we need to turn the coeffect into a form that is required by the exchange
-and contraction rules. Aside from idempotence, we could use the unit property and obtain 
-\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this appraoch does not 
-work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
-the existence of unit of $\czip$).
-
-In the other direction, we need to show that any valid judgement in the unified system (tracking
-flat properties) is also valid in the flat system. The syntax-directed rules are the same in both
-systems, but we need to show that any use of (explicit) weakening, contraction and exchange can be
-derived in the flat system.
-
-\begin{lemma}
-Weakening, as defined in a unified coeffect calculus capturing flat properties, is admissible in 
-the flat coeffect calculus.
-\end{lemma}
-\begin{proof}
-Similar to the proof in Lemma~\ref{thm:unified-weak}. The property follows from the fact that 
-$\bot=\alift{\czero}$ is the unit of $\cpar$.
-\end{proof}
-
-\begin{lemma}
-\label{thm:unified-rev-contr}
-Contraction, as defined in a unified coeffect calculus capturing flat properties, is admissible in 
-the flat coeffect calculus.
-\end{lemma}
-\begin{proof}
-The application of (\emph{contr}) rule has the following form:
-\begin{equation*}
-\inference
- { \inference
-    {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}} \vdash e : \tau}
-    {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\azip\, \alift{\cclrd{s}} \,\azip\, \alift{\cclrd{t}} \,\azip\, \cclrd{q}} \vdash e : \tau} }
- { \inference
-    {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\apar\, \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \,\apar\, \cclrd{q}} \vdash \subst{e}{z,y}{x} : \tau}
-    {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{q}} \vdash \subst{e}{z,y}{x} : \tau} } 
-\end{equation*}
-%
-From Definition~\ref{def:unified-flat}, we know that $(\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}) 
-\;\cleq\; (\cclrd{r} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{q})$. Thus, the judgement
- can be derived using the (\emph{sub}) rule of flat coeffect calculus.
-\end{proof}
-
-\begin{lemma}
-\label{thm:unified-rev-exch}
-Exchange, as defined in a unified coeffect calculus capturing flat properties, is admissible in 
-the flat coeffect calculus.
-\end{lemma}
-\begin{proof}
-The application of (\emph{exch}) rule has the following form:
-\begin{equation*}
-\inference
- {\inference
-    {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}} \vdash e : \tau}
-    {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\azip\, \alift{\cclrd{s}} \,\azip\, \alift{\cclrd{t}} \,\azip\, \cclrd{q}} \vdash e : \tau} }
- {\inference
-    {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\apar\, \alift{\cclrd{t}} \,\apar\, \alift{\cclrd{s}} \,\apar\, \cclrd{q}} \vdash e : \tau}
-    {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{q}} \vdash e : \tau} }
-\end{equation*}
-%
-From idempotence of $\czip$, we get that $\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q} =
-\cclrd{r} \,\czip\, \cclrd{t} \,\czip\, \cclrd{s} \,\czip\, \cclrd{q}$. Thus the judgement can be derived using
-(\emph{sub}) as in contraction.
-\end{proof}
-
-\noindent
-A consequence of the equivalence discussed above is that the unified coeffect system can capture 
-all properties that can be captured by the flat coeffect system -- including implicit parameters,
-rebindable resources, Haskell type classes (discussed by Orchard~\cite{comonads-dom-thesis}),
-data-flow and variable liveness.
-
-%--------------------------------------------------------------------------------------------------
-
-\subsection{Semanitcs of unified calculus}
-\label{sec:unified-semantics}
-
-The semantics of unified calculus can be obtained by generalizing the semantics of the structural
-calculus in the same way as we generalized the type system. However, it is worth exploring the type 
-signatures of the operations of the underlying (comonad-based) structure.
-
-The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
-structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The 
-index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of 
-the structure $\C^{\SP \sclrd{n}}$ for a specified container $\SH \triangleleft \SP$ and a shape 
-$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in 
-Section~\ref{sec:unified-types}, we need to modify the domains and codomains of the \ident{merge}, 
-\ident{split} and \ident{dup} operations as follows.
-%
-\begin{equation*}
-\begin{array}{lcl}
- \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
-  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
- \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}& 
-  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
- \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
-  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
-\end{array}
-\end{equation*}
-%
-The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
-\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
-(in \ident{merge}, but also in \ident{dup} which models contraction).
-This means that the splitting and merging the context can change the coeffect annotation:
-%
-\begin{equation*}
-\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
-  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
-  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} 
-\end{equation*}
-%
-For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
-splitting and then merging a context may lose some of the avilable contextual capabilities
-(and increase context requirements). This corresponds to the fact that our proofs of 
-Lemma~\ref{thm:unified-rev-contr} and Lemma~\ref{thm:unified-rev-exch} relied on sub-coeffecting
-in the flat system.
-
-\paragraph{Related and future work.}
-The semantics presented here serves mainly to inform the design of the type system. For this 
-reason, we keep the semantics simple and close to the concrete notions used in the type system
-such as flat, structural and unified coeffect algebras. In particular, the indexed comonad 
-structure consists of object mapping $\ctyp{r}{}$ that is indexed by a corresponding algebra:
-
-\begin{compactitem}
-\item[-] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
-\item[-] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is 
-  indexed by vectors of scalars $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$
-\item[-] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
-  the mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is indexed by shape-indexed scalars 
-  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP\sclrd{m}}$
-\end{compactitem}
-
-\noindent
-In this treatment, the indices are formed by objects that are modelled outside of category theory.
-A fully categorical semantics can be found in a joint paper with Orchard \cite{coeffects-icfp14}. 
-There, the index (coeffect scalars and shape-indexed scalars) are modelled as categories too.
-The definition requires the following notation:
-
-\begin{compactitem}
-\item[-] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
-\item[-] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
-\end{compactitem}
-
-\noindent
-In the categorical semantics of coeffect calculi, coeffect scalars are modelled as a
-category of scalars $\cclrd{\mathbb{I}}$ such that $\textit{obj}(\cclrd{\mathbb{I}}) = \C$.
-The structure of the context (indexed comonad) over the input then becomes a functor
-indexed by the free-variable context shape $\slen{\Gamma}$ and coeffect annotations
-$\aclrd{\textbf{r}}$, \ie~$\ctyp{\slen{\Gamma}}_{\aclrd{\textbf{r}}}$:
-%
-\begin{equation*}
-\sem{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \ctyp{\slen{\Gamma}}_{\aclrd{R}}
-\sem{\Gamma} \rightarrow \sem{\tau}
-\end{equation*}
-%
-The structure $\ctyp{}$ can be thought of as a dependent product of functors $\ctyp{\sclrd{n}}$ 
-over possible shapes $\sclrd{n} \in \SH$ where:
-%
-\begin{equation*}
-\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP \sclrd{n}} \rightarrow [\mathbb{C}^{\SP \sclrd{n}}, \mathbb{C}]
-\end{equation*}
-%
-For a fixed context shape $\sclrd{n}$ the functor 
-$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP \sclrd{n}} \rightarrow [\mathbb{C}^{\SP \sclrd{n}}, \mathbb{C}]$ 
-maps a coeffect indexed by positions $\SP \sclrd{n}$ to a functor from a context 
-$\mathbb{C}^{\sclrd{n}}$ to an object in $\mathbb{C}$. That is, given a coeffect annotation 
-(matching the shape of the context), we get a functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$. 
-From a programming perspective, this functor defines a data structure that models the additional
-context provided to the program. The shape of this data structure depends on the coeffect
-annotation $\cclrd{\mathbb{I}}^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
-the number of values needed for each variable and the functor will be 
-formed by lists of length matching the required number.
-
-In the model above, shapes and positions are still treated as ordinary sets. Indeed, the paper
-\cite{coeffects-icfp14} treats shapes as just sets of sets. In this chapter, we instead define
-\emph{coeffect container} as an extension of containers, which have a well-defined 
-categorical structure \cite{semantics-containers}. A future work is adapting the semantics of
-the unified coeffect calculus so that it is based on the semantics of containers. However, note 
-that context is \emph{not} a (simple) container, because containers are used as indices (coeffect
-annotations) of the object that wrap the free-variable context.
-
-% =================================================================================================
-%                                                                                             
-%    #   #          #                    ##                                                    
-%    #   #          #                     #                                                    
-%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
-%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
-%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
-%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
-%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
-%                                                            #   #                #   #        
-%                                                             ###                  ###         
-% =================================================================================================
-
-\section{Coeffect meta-language}
-\label{sec:unified-meta}
-
-In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
-language semantics introduced by Moggi \cite{monad-notions}. The first approach is to use
-monads in the \emph{semantics} of an effectful language. The second approach is to extend the
-language with (additional) monadic constructs that can then be used for writing effectful
-monads explicitly.
-
-In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
-the term language is that of simply-typed $\lambda$-calculus, and we used (flat or structural)
-indexed comonads to give the semantics for the language and to derive type system for it.
-
-In this section, we briefly discuss the alternative approach. That is, we embed indexed
-comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
-type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
-context (semantically, this corresponds to an indexed comonad) and we add language constructs
-that correspond to the operations of indexed comonads.
-
-This section provides a brief sketch of coeffect meta-language to highlight the relationship
-between coeffects and important related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
-Developing the system further is an interesting future research direction.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffects and contextual modal type theory}
-
-As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
-to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
-meta-languages that include $\square$ as a type constructor
-\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
-Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
-From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
-coeffect calculus.
-
-\paragraph{Context in CMTT and coeffects.}
-Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
-comonads (the $\square$ modality) into the meta-language, there are two important differences.
-
-Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
-requirements are formed by a coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, \emph{etc.}
-
-Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
-the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
-constructor refers to a value of type $\tau$ \emph{together} with additional context specified
-by $\cclrd{r}$ (\eg~list of past values or additional resources). In contrast in 
-CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the 
-original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the 
-context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the 
-two operations of a comonad:
-\begin{equation*}
-\begin{array}{rcl}
- \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
- \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow 
-    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
-\end{array}
-\end{equation*}
-Both readings are possible, but they give quite different meanings to the operations:
-
-\begin{itemize}
-\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value 
-  and does not require any additional context; the \ident{cobind} operation requires
-  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to 
-  evaluate the function, ending with a value $\beta$ together with remaining context
-  $\cclrd{s}$.
-\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation 
-  that requires no additional context to obtain a $\alpha$ value; given a function that
-  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
-  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$ 
-  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
-  context requirements is eliminated by the function).
-\end{itemize}
-
-\noindent
-Although the different reading does not affect formal properties of the systems, it is important
-to understand the difference when discussing the two systems.
-
-The sketch of a coeffect meta-language in the following section attempts to bridge the gap
-between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
-it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
-sets of variables. Future work on the coeffect meta-language would thus be an interesting 
-development for both coeffect systems and CMTT.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect meta-language}
-
-The coeffect meta-langauge could be designed using both flat and structural indexed comonads.
-For simplicity, this section only discusses the flat variant. The syntax of types and terms
-of the langauge includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four expression forms:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
- e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
-  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
-  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
-  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 
-\end{array}
-\end{equation*}
-%
-The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
-operation of the comonad. To define meta-langauge for flat indexed comonads, we also include 
-\kvd{split} and \kvd{merge} that embed the corresponding operations.
-
-\paragraph{Types for coeffect meta-language.}
-The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
-shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing 
-rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
-
-The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
-a value from a comonadic context and corresponds to variable access in coeffect calculi. The
-\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
-and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
-The result is a computation that produces a $\tau_2$ value with the remaining context specified
-by $\cclrd{s}$. Note that the expression $e_2$ and $e_1$ corresponds to the first and second
-arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following 
-CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
-although it differs because of our generalization of comonads where \ident{bind} composes
-coeffect annotations rather than requiring the same annotation everywhere.}.
-
-The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some 
-transformation on one or two values in a context and then add the new value as a fresh variable
-to the variable context. For simplicity, we omit sub-coeffecting, but it could be easily added 
-following the pattern used elsewhere.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-
-{\small a.) Typing rules for the simply typed $\lambda$-calculus}
-
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\Gamma \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
-   \Gamma \vdash e_2 : \tau_2 }
-  {\Gamma \vdash e_1~e_2 : \tau_1 }
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\Gamma, x:\tau_1 \vdash e : \tau_2}
-  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
-\end{equation*}
-
-\vspace{1em}
-{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
-
-\begin{equation*}
-\tyrule{cobind}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
-\end{equation*}
-\begin{equation*}
-\tyrule{counit}
-  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
-  {\Gamma \vdash !e : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{split}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
-   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{merge}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
-   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
-   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
-  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
-\end{equation*}
-
-\caption{Type system for the (flat) coeffect meta-language}
-\label{fig:conclusions-cml}
-\end{figure}
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Embedding flat coeffect calculus}
-
-The \emph{meta-language} approach of embedding comonads in a language is more general than the
-\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
-a type system for context-aware programming languages.
-
-However, it is worth noting that 
-(flat) coeffect calculus can be embedded in the meta-language described above. This may be 
-desirable, \eg~when using the meta-language for reasoning about context-aware computations. 
-We briefly consider the embedding as it illuminates the relationship between coeffects and
-CMTT (although it is not possible to embed coeffect calculi in CMTT because of the more general
-annotations structure).
-
-Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in a flat coeffect calculus,
-we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its emebdding in the coeffect
-meta-language. Note that the translation is indexed by $v$, which is a name of variable used 
-to represent the entire variable context of the source language. The translation is defined in
-Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in 
-Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
-the operations of a monad.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\begin{equation*}
-\begin{array}{l}
- \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
-   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
-   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
-      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
-   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
-\end{array}
-\end{equation*}
-\caption{Embedding flat coeffect calculus in coeffect meta-language }
-\label{fig:conclusions-embed}
-\end{figure}
-
-% =================================================================================================
-%                                                                   
-%      ####                         #       #                   ##   
-%      #   #                        #                            #   
-%      #   #  # ##    ###    ###   ####    ##     ###    ###     #   
-%      ####   ##  #      #  #   #   #       #    #   #      #    #   
-%      #      #       ####  #       #       #    #       ####    #   
-%      #      #      #   #  #   #   #  #    #    #   #  #   #    #   
-%      #      #       ####   ###     ##    ###    ###    ####   ###  
-%                                                                
-% =================================================================================================
-
-\section{Towards practical coeffects}
-\label{sec:unified-impl}
-
-As discussed earlier, the main focus of this thesis is the development of the much needed 
-\emph{theory of context-aware computations} and so discussing the details of a practical 
-implementation of coeffect tracking is beyond the scope of the thesis. However, this section
-briefly outlines one possible pathway towards this goal.
-
-Many of the examples of contextual computation that we discussed earlier have been implemented as 
-a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters} 
-or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However, 
-the main contribution of this thesis is that it captures
-\emph{multiple} different notions of context-aware computations using just a \emph{single}
-common abstraction. For this reason, we advocate that future practical implementations of
-coeffects should not be single-purpose language features, but instead reusable abstractions that
-can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
-
-In order to do this, programming languages need to provide two features; one that allows 
-embedding of context-aware computations themselves in programs akin to the ``do'' notation
-in Haskell (Section~\ref{sec:unified-impl-embed}) and one that allows tracking of the 
-contextual information in the type system (Section~\ref{sec:unified-impl-types}). To make
-constraint resolution in type inference easier, we also discuss how the coeffect algebra 
-can be simplified (Section~\ref{unified-impl-semilattice})
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Embedding contextual computations}
-\label{sec:unified-impl-embed}
-
-The embedding of \emph{contextual} computations in programming languages can learn from better 
-explored emedding of \emph{effectful} computations. In purely functional programming languages
-such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
-necessary (monadic) plumbing. This is made easier by the ``do'' notation \cite{other-haskell98} 
-that inserts the monadic operations automatically.
-
-The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
-context-aware computations based on comonads. The notation is close to the semantics of our 
-flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based 
-on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
-this requires explicitly manipulating individual context variables and application of structural
-rules, which is not needed in flat coeffects.
-
-In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
-but they can still use a special notation for explicitly marking effectful (coeffectful) 
-blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
-that differ from the ``do'' notation in two ways. First, they support wider range of 
-constructs, making it possible to wrap existing F\# code in a block without other changes.
-Second, they support other abstractions including monads, applicative functors and monad
-transformers. It would be interesting to see if computation expressions can be extended to
-handle computations based on flat/structural indexed comonads.
-
-More lightweight syntax for effectful computation can be obtained using
-techniques that automatically insert the necessary monadic plumbing (wihtout a
-special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
-and similar approach would be worth exploring for coeffects.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect annotations as types}
-\label{sec:unified-impl-types}
-
-The other aspect of practical implementation of coeffects is tracking the context requirements
-(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
-language feature) the type system needs to be able to capture various kinds of coeffect algebras. 
-The structures used in this thesis include sets (with union or interesection), natural numbers (with 
-addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids 
-(vectors of annotations).
-
-The work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
-recent additions to the Haskell type system provide enough poer to implement structures such
-as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful 
-future direction for applied coeffects.
-
-In dependently-typed programming languages such as Agda \cite{other-agda} or Idris \cite{other-idris},
-the embedding of coeffects can be implemented more directly. However, we believe that coeffect
-tracking does not require full dependent types and can be made accessible in more main-stream
-languages. Dependent ML \cite{types-dependent-ml} provides an interesting example of a language
-with some dependent typing which is still close to its non-dependently-typed predecessor ML.
-
-Another approach for embedding computations into the type system has been pioneered by F\#
-\emph{type providers} \cite{app-inforich}. Technically, type providers are compiler extensions 
-that generate types based on external data sources or other information in order to provide easy 
-access to data or services. A similar approach could be used for embedding \emph{algberas} such
-as coeffect algebras into the type system. A \emph{algebra provider} would be a library that 
-specifies the objects of the algebra, equational laws and generalization rules for type inference.
-This could provide an easy to use way of embedding coeffect tracking in pragmatic languages
-such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
-\cite{types-units-of-measure}, which could be provided via one such \emph{algebra provider}.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Alternative formulation using coeffect lattice}
-\label{sec:unified-impl-semilattice}
-
-The fact that coeffect scalar structure consists of four operations ($\cseq,\czip,\cpar$ 
-and a relation $\cleq$) makes working with coeffect annotations difficult -- it compilcates tasks
-such as type generalization and constraint solving in type inference. Here, we look at one possible 
-simplification of the structure.
-
-\begin{itemize}
-\item As mentioned in Section~\ref{sec:flat-calculus-algebra}, when the monoid $(\C,\cpar,\czero)$ is a
-semi-lattice, the $\cleq$ relation can often be expressed in terms of the $\cpar$ operation 
-($\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}$).
-This does not give a definition equivalent to our, but it is a simplification that is consistent 
-with all examples in this thesis.
-
-\item Another approach that works for most (but not all) systems we discussed is to require the structure
-$(\C,\cpar,\czip)$ to form a lattice. This is consistent with all our structural coeffect systems as
-well as flat systems for liveness and data-flow, but not with the system for tracking of implicit
-parameters (and other sets).
-\end{itemize}
-
-\noindent
-The system for tracking implicit parameters does not fit the lattice-based model, because it uses the 
-same operation $\cup$ for both $\cpar$ and $\czip$. This also means that the system for implicit parameters 
-cannot be used with a system where lambda abstraction duplicates the context requirements 
-(Section~\ref{sec:flat-exts-lambda}).
-
-\paragraph{Lattice-based coeffects.} Putting implicit parameters aside, there are many other coeffect
-systems for which we could use the following definition of coeffect scalars (for simplicity, we only 
-consider flat systems):
-
-\begin{definition}
-A \emph{\cclrd{flat coeffect lattice}} $(\C, \cseq, \cleq, \cunit, \czero)$ is a set 
-$\C$ together with elements $\cunit, \czero \in \C$ and an ordering $\cleq$ such 
-that $(\C, \cseq, \cunit)$ is a monoid and $(\C, \cleq)$ is a lattice with the least 
-element $\czero$.
-
-Assuming $\cpar$ and $\czip$ are the \emph{least upper bound} and \emph{greatest lower bound}
-of the lattice, we require the following distributivity axioms:
-%
+\\
+\end{array}
+\end{equation*}
+%
+The following distributivity axioms hold:
 \begin{align*}
@@
 \noindent
-The definition provides the same structure as flat coeffect algebra, but it defines a more rigid 
-structure. In particular, $\czip$ and $\cpar$ are defined in terms of $\cleq$. This simplifies the 
-system as the operations of the underlying indexed comonad structure can be expressed using just
-$\cseq$ and $\cleq$:
-%
-\begin{equation*}
-\begin{array}{lcll}
- \ident{counit}_{\cunit} &\narrow{:}& 
-    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
- \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}& 
-    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
-\ident{merge}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
-    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{t}}{(\alpha \times \beta)} &
-    (\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s}) \\
-\ident{split}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}& 
-    \ctyp{\cclrd{t}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} &
-    (\cclrd{r}\,\cleq\,\cclrd{t}, \cclrd{s}\,\cleq\,\cclrd{t}) \\
-\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}& 
-    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} & 
-    (\cclrd{s}\,\cleq\,\cclrd{r})
-\end{array}
-\end{equation*}
-%
-Note that we still need two \emph{distinct} operators in the system. The one used for
-sequential composition $\cseq$ is not necessarilly related in any way to the ordering on
-the lattice $\cleq$.
-
-However, the operations $\ident{merge}$, $\ident{split}$ and $\ident{lift}$ can now all
-use ordering. For example, the \ident{merge} operation previously returned
-a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
-element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit sub-coeffecting
-to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and 
-$\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
-This means that we can compose the \ident{merge} operation which returns greatest lower
-bound with $\ident{lift}$ to get the (less precise) operation required here.
+In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
+available from the declaration site and the call site or, in the syntactic reading, to split the 
+context requirements. The other difference is that the $\cpar$ operation of structural coeffect
+scalar is not required to be idempotent. The bounded reuse example uses $\cpar$ which is not 
+idempotent.
+
+Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
+via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma 
+(Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
+for all our examples; it is needed for the structural version of the substitution lemma 
+(Lemma~\ref{thm:structural-substitution}).
+
+In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
+coeffects becomes just vector concatenation or splitting, respectively; these are provided
+by the tensor product. The operations on vectors are indexed by integers representing the lengths
+of the vectors. The additional structure required by the type system for structural coeffect 
+calculi is given by the following definition.
+
+\begin{definition}
+A \emph{\aclrd{structural coeffect algebra}} is formed by a structural coeffect scalar 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:
+
+\begin{itemize}
+\item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
+  structural coeffect scalars indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.
+
+\item An operation that constructs a vector from scalars indexed by the vector length
+  $\alift{-}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
+  and an operation that returns the vector length such that 
+  $\alen{\aclrd{\textbf{r}}} = \sclrd{n}$ for $\aclrd{\textbf{r}} : \C^{\sclrd{n}}$
+
+\item A pointwise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
+  such that\\ $\cclrd{t} \,\aseq\, \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} = 
+    \alift{\cclrd{t} \,\cseq\,\cclrd{r_1}, \ldots, \cclrd{t} \,\cseq\,\cclrd{r_n}}$.
+
+\item An indexed tensor product $\atimes_{\sclrd{n},\sclrd{m}} : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n}+\sclrd{m}}$
+  that is used in both directions -- for vector concatenation and for splitting -- 
+  which is defined as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} 
+     \atimes_{\sclrd{n}, \sclrd{m}} \alift{\cclrd{s_1}, \ldots, \cclrd{s_m}}
+   = \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}, \cclrd{s_1}, \ldots, \cclrd{s_m}}$
+\end{itemize}
+\end{definition}
+
+\noindent
+The fact that the tensor product $\atimes_{\sclrd{n},\sclrd{m}}$ is indexed by the lengths
+of the two vectors means that we can use it unambiguously for both concatenation of vectors
+and for splitting of vectors, provided that the lengths of the resulting vectors are known.
+In the following text, we usually omit the indices and write just 
+$\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}}$, because the lengths of the coeffect vectors 
+can be determined from the lengths of the matching free variable context vectors.
+
+More generally, we could see the the coeffect annotations as \emph{containers} 
+\cite{types-containers}. This approach is used in Chapter~\ref{ch:unified} to unify flat 
+and structural systems.
 
@@
 
+\subsection{Structural coeffect types}
+\label{sec:struct-calculus-types}
+
+The type system for the structural coeffect calculus is similar to sub-structural type systems
+in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
+weakening, exchange or contraction -- this is done by checking the types of sub-expressions
+in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
+does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
+perform corresponding transformation on the coeffect annotation.
+
+\paragraph{Syntax-driven rules.}
+The syntax-driven rules of the type system are shown in Figure~\ref{fig:struct-types} (a).
+The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
+using $\cunit$. Note that, as in sub-structural systems, the free variable context contains 
+\emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
+Constants (\emph{const}) are type checked in an empty variable context, which is annotated with
+an empty vector of coeffect annotations.
+
+The (\emph{abs}) rule assumes that the free variable context of the body can be split into
+a potentially empty \emph{declaration site} and a singleton context containing the bound
+variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
+the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
+non-determinism of the type inference present in flat coeffect systems.
+
+In (\emph{app}), the sub-expressions $e_1$ and $e_2$ use free variable contexts $\Gamma_1, \Gamma_2$
+with coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$, respectively. The function
+value is annotated with a coeffect scalar $\cclrd{t}$.
+The coeffect annotation of the composed expression is obtained by combining the annotations
+associated with variables in $\Gamma_1$ and $\Gamma_2$. Variables in $\Gamma_1$ are only
+used to obtain the function value, resulting in coeffects $\aclrd{\textbf{r}}$. The 
+variables in $\Gamma_2$ are used to obtain the argument value, which is then sequentially
+composed with the function, resulting in $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$.
+
+\paragraph{Structural rules.}
+The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
+They allow different transformation of the free variable context. We include (\emph{sub}) 
+as one of the rules, allowing sub-coeffecting on coeffect scalars belonging to individual 
+variables. The remaining rules capture \emph{weakening}, \emph{exchange} and \emph{contraction} 
+known from sub-structural systems.
+
+Rule (\emph{weak}) allows adding a variable to the context, extending the coeffect
+vector with $\czero$ to mark it as unused, (\emph{exch}) provides a way to rearrange variables
+in the context, performing the same reordering on the coeffect vector. Finally recall that 
+variables in the free variable context are required to be \emph{unique}. The (\emph{contr}) rule 
+allows re-using a variable. We can type check sub-expressions using two separate variables and 
+then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
+is the only place in the structural coeffect system where context requirements are combined
+(semantically, this is where the available context is shared).
+
+%---------------------------------------------------------------------------------------------------
+
 \begin{figure}[t]
-{\small a.) Littice-based flat systems (selected rules):}
+{\small a.) Syntax-driven typing rules:}
+\begin{equation*}
+\tyrule{var}
+  { }
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c \!:\! \tau \in \Delta}
+  {\coctx{()}{\alift{}} \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
 \begin{equation*}
 \tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{t}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-~~(\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s})
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
-   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\cclrd{u}} \vdash e_1~e_2 : \tau_2}
-~~(\cclrd{r}\,\cleq\,\cclrd{u}, (\cclrd{s} \,\cseq\, \cclrd{t})\,\cleq\,\cclrd{u})
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
 \end{equation*}
@@
 \vspace{1em}
-{\small b.) Littice-based structural systems (selected rules):}
-\begin{equation*}
+{\small b.) Structural rules for context manipulation:}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+(\cclrd{s'} \cleq \cclrd{s})
+\end{equation*}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau} 
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+\begin{array}{l}
+ \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{2.5em} 
 \tyrule{contr}
   {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{u}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-~~(\cclrd{s}\,\cleq\,\cclrd{u}, \cclrd{t}\,\cleq\,\cclrd{u})
-\end{equation*}
-
-\caption{Flat and structural coeffects using lattice-based formulation}
-\label{fig:unified-lattice-types}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~
+\begin{array}{l}
+ \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\figcaption{Type system for the structural coeffect calculus}
+\label{fig:struct-types}
 \end{figure}
 
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Lattice-based type system.}
-
-Using lattice-based coeffect structure leads to the typing rules shown in Figure~\ref{fig:unified-lattice-types}.
-In the flat coeffect system, this affects both application and abstraction -- in abstraction,
-the context-requirements of the body should be smaller than the context available on the 
-call site and declaration site. In application, we require more than what is required by
-the sub-expressions.
-
-Structural coeffect systems only require least upper bound $\cpar$ in the (\emph{contr}) rule.
-This can be rewritten similarly to flat application --  when contracting two variables, the
-context requirements of the resulting variable are greater than the requirements associated with
-each of the original variables.
-
-In smmary, the lattice-based coeffect system has some advantages over the design used in this
-thesis. In particular, it uses better-known structures and is simpler. However, we do not use
-it as it rules out one of the important motivating examples (implicit parameters) and, even with
-the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Understanding structural coeffects}
+
+The type system for structural coeffects appears more complicated when compared to the flat
+version, but it is in many ways simpler -- it removes the ambiguity arising from the use
+of $\czip$ in lambda abstraction and, as discussed in Section~\ref{sec:struct-syntax}, has
+a cleaner equational theory.
+
+\paragraph{Flat and structural context.} In flat systems, lambda abstraction splits context
+requirements using $\czip$ and application combines them using $\cpar$. In the structural
+version, both of these are replaced with $\atimes$. The $\czip$ operation is not needed, but
+$\cpar$ is still used in the (\emph{contr}) rule.
+
+This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
+as over-approximations and under-approximations of $\atimes$. This is demonstrated by the 
+(\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
+$\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$. Semantically, flat abstraction
+combines two values representing the available context, potentially discarding parts of it 
+(under-approximation), while flat application splits the available context (a single value), 
+potentially duplicating parts of it (over-approxi\-mation)\footnote{Because of this duality, earlier 
+version of coeffect systems \cite{coeffects-icalp13} used $\wedge$ and $\vee$.}.
+
+Secondly, the operator $\cpar$ is used when the semantics passes a given context to multiple
+sub-expressions. In flat systems, the context is shared in (\emph{app}) and (\emph{pair}), because 
+the sub-expressions may share variables. In structural systems, the sharing is isolated into an 
+explicit contraction rule.
+
+\paragraph{Let binding.}
+The other aspect that makes structural systems simpler is that they remove the need for 
+separate let binding. As discussed in Section~\ref{sec:flat-calculus-let}, flat calculi include
+let binding that gives a \emph{more precise} typing than combination of abstraction and 
+application. Structural coeffect systems avoid this issue.
+
+\begin{remark}[Let binding]
+\label{thm:structural-letbinding}
+In a structural coeffect calculus, the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$ can be
+seen as a derived rule, \ie~its typing is equivalent to the typing of the expression
+$(\lambda x.e_2)~e_1$.
+
+\end{remark}
+\begin{proof}
+Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
+in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
+%
+\begin{equation*}
+\inference
+  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
+    \inference
+      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
+      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
+  { \inference
+      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
+      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
+\end{equation*}
+
+\noindent
+The assumptions and conclusions match those of the (\emph{let}) rule. 
+\end{proof}
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Examples of structural coeffects}
+
+The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
+presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
+structural liveness provide a more precise tracking of properties that can be tracked using 
+flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
+(by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
+example, implicit parameters could be attached to individual variables, but, to our knowledge, this
+has not been used in any practical system.
+
+On the other hand, some of the structural systems do not have a flat equivalent, typically 
+because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
+scalar. This is the case, for example, for the bounded variable use.
+
+\begin{example}[Structural liveness]
+The structural coeffect scalar for liveness is formed by
+$(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is 
+the same two-point lattice as in the flat version, that is $\ident{D} \sqsubseteq \ident{L}$
+with a join $\sqcup$ and a meet $\sqcap$. 
+\end{example}
+
+\begin{example}[Structural data-flow]
+In data-flow, context is annotated with natural numbers and the structural coeffect scalar is formed 
+by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
+\end{example}
+
+\noindent
+For the two examples that have both flat and structural version, obtaining the structural coeffect
+algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
+required by a structural coeffect algebra are the same as those required by the flat version
+and so the above definitions are both valid. Similar construction can be used for the
+\emph{optimized data-flow} example from Section~\ref{sec:flat-calculus-examples}.
+
+It is important to note that this gives us a systems with \emph{different} properties. The 
+information are now tracked per-variable rather than for entire contexts. For data-flow,
+we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$ 
+for a pointwise extension of the $+$ operator, such that
+$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} \aclrd{+} k = \alift{\cclrd{r_1} + k, \ldots, \cclrd{r_n} + k}$.
+%
+\begin{equation*}
+\tyrule{prev}
+  { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau }
+  { \coctx{\Gamma}{\aclrd{\textbf{r}} \,\aclrd{+}\, 1} \vdash \kvd{prev}~e : \tau }
+\end{equation*}
+%
+The rule appears similar to the flat one, but there is an important difference. Because of the
+structural nature of the type system, it only increments the required number of values for 
+variables that are used in the expression $e$. Annotations of other variables can be left 
+unchanged.
+
+Before looking at the semantics and equational properties of structural coeffect systems, 
+we consider bounded variable use, which is an example of structural system that does not
+have a flat counterpart.
+
+\begin{example}[Bounded variable reuse]
+The structural coeffect algebra for tracking bounded variable use is given by
+$(\mathbb{N}, \ast, +, 1, 0, \leq)$
+\end{example}
+
+\noindent
+Similarly to the structural calculus for data-flow, the calculus for bounded variable reuse
+annotates each variable with an integer. However, the integer denotes how many times is the variable 
+\emph{accessed} rather than how many \emph{past values} are needed. The resulting type system is 
+the one shown in Figure~\ref{fig:applications-struct-live} in Chapter~\ref{ch:applications}.
+
+% ==================================================================================================
+%                                              
+%      #####  #                                 
+%        #    #                                 
+%        #    # ##    ###    ###   # ##   #   # 
+%        #    ##  #  #   #  #   #  ##  #  #   # 
+%        #    #   #  #####  #   #  #      #  ## 
+%        #    #   #  #      #   #  #       ## # 
+%        #    #   #   ###    ###   #          # 
+%                                         #   # 
+%                                          ###                                                                         
+% ==================================================================================================
+
+\section{Categorical motivation}
+\label{sec:struct-semantics}
+
+When we introduced structural coeffect systems in Section~\ref{sec:applications-structural}, we 
+included a concrete semantics of structural liveness and bounded variable reuse. In this section,
+we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
+extension of the \emph{indexed comonad} structure. Similarly to \emph{flat indexed comonad}, the
+structural variant adds operations that are needed to embed full $\lambda$-calculus, this time 
+with per-variable contexts.
+
+As in the previous chapter, the main aim of this section is to motivate and explain the design of 
+the structural coeffect calculus shown in Section~\ref{sec:struct-calculus}. However, the semantics 
+also highlights the similarities and differences between the flat and structural system.
+Most of the differences arise from the fact that contexts in structural coeffect systems are treated 
+as \emph{vectors} rather than sets.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of vectors}
+
+Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing 
+judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a 
+morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, 
+we can freely transform the value contained in the context modelled using an indexed comonad
+$\ctyp{\cclrd{r}}$.
+
+Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
+which transforms the value inside the context without affecting the coeffect annotation. Thus,
+we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
+$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value 
+$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
+
+The ability to freely transform the variable structure is not desirable in the model of 
+structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
+the coeffect annotations matches the structure of variables. To achieve this, we model vectors
+using a structure distinct from ordinary products which we denote $-\xtimes-$. 
+For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
+is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
+We assume that the operator is equipped with necessary associativity transformations allowing us
+to use it freely on more than two values.
+
+The operator is a bifunctor, but it is \emph{not} a product in the categorical sense. In particular, 
+there is no way to turn $\tau_1 \xtimes \ldots \xtimes \tau_n$ into $\tau_i$ (the structure does 
+not have projections) and so there is also no way of turning
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$ into
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{\tau_i}$, which would break the correspondence 
+between coeffect annotations and variable structure.
+
+The structure created using $-\xtimes-$ can be manipulated only using operations provided by
+the \emph{structural indexed comonad}, which operate over variable contexts contained in an 
+indexed comonad $\ctyp{\aclrd{\textbf{r}}}$ and are designed to preserve the correspondence
+between vectors and annotations.
+
+In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
+As mentioned, we assume that the use of the operator can be freely re-associated. For example, when
+calling an operation that requires input of the form 
+$(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
+we use an argument $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ and assume that the 
+appropriate transformation is inserted.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Indexed comonads, revisited}
+
+The semantics of structural coeffect calculus reuses the definition of \emph{indexed comonad} almost
+without change. The additional structure that is required for context manipulation (merging and 
+splitting) is different and is here provided by the \emph{structural indexed comonad} structure that 
+we introduce in this section.
+
+Recall the definition from Section~\ref{sec:flat-semantics-idx}, which defines an indexed como\-nad 
+over a monoid $(\C, \cseq, \cunit)$ as a triple $(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, 
+\ident{cobind}_{\cclrd{r}, \cclrd{s}})$. The triple consists of a family of object mappings $\ctyp{\cclrd{r}{}}$,
+and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
+
+In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$
+representing function values (appearing in the language), but also of the form
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$, 
+modelling expressions in a context. To capture this, we need to revisit the definition and use 
+\emph{coeffect vectors} in some of the operations.
+
+\begin{definition}
+Given a monoid $(\C, \cseq, \cunit)$ with a pointwise extension of the $\cseq$ operator to a vector 
+(written as $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$) and an operation lifting scalars to vectors 
+$\alift{-}$, an \emph{indexed comonad} over a category $\catc$ is a triple 
+$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}})$:
+
+\begin{compactitem}
+\item $\ctyp{\aclrd{\textbf{r}}}{}$ for all $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$ is a family of object mappings 
+\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\alift{\cunit}}{\alpha} \rightarrow \alpha$ 
+\item $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$ is a mapping $(\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta) 
+  \rightarrow (\ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{s}}}{\beta})$
+\end{compactitem}
+\end{definition}
+
+\noindent
+The object mapping $\ctyp{\aclrd{\textbf{r}}}$ is now indexed by a vector rather than by a scalar 
+$\ctyp{\cclrd{r}}$ as in the previous chapter. This new definition supersedes the old one, because a 
+flat coeffect annotation can be seen as singleton vectors.
+
+The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
+return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
+operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
+is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
+takes a vector of variables, but always produces just a single value. Thus the resulting function 
+also takes a vector of variables, but always returns a context with a vector containing just one value.
+In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
+no way of constructing values containing $\xtimes$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural indexed comonads}
+
+The flat indexed comonad structure extends indexed comonads with operations 
+$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ that combine or 
+split the additional (flat) context and are annotated with the flat coeffect operations $\czip$ 
+and $\cpar$, respectively.
+
+In the structural version, the corresponding operations operate convert between a (wrapped) vector 
+of values represented using $\xtimes$ and ordinary pairs of contexts containing parts of the vector. 
+The vectors of coeffect annotations are split or merged using $\atimes$ of the structural coeffect 
+algebra, in a way that mirrors the wrapped vectors (variable structure).
+
+The following definition includes $\ident{lift}_{\cclrd{r'},\cclrd{r}}$, which is similar as
+before and models sub-coeffecting and also $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ which models
+duplication of a variable in a context needed for the semantics of contraction:
+
+\begin{definition}
+Given a structural coeffect algebra formed by $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$
+with operations $\alift{-}$ and $\aseq$, a \emph{structural indexed comonad} is an indexed comonad over 
+the monoid $(\C, \cseq, \cunit)$ equipped with families of operations $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$, 
+$\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$,
+$\ident{dup}_{\cclrd{r}, \cclrd{s}}$ and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
+%
+\begin{compactitem}
+\item $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
+  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}$
+\item $\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
+  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
+\item $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ is a family of mappings
+  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \xtimes \alpha)}$
+\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
+  $\ctyp{\alift{\cclrd{r'}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\end{compactitem}
+%
+Here, the following equalities must hold:
+\begin{equation*}
+\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~\equiv~ \ident{id} \qquad\qquad
+\ident{id} ~\equiv~ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} 
+\end{equation*}
+
+\end{definition}
+
+\noindent
+These operations differ from those of the flat indexed comonad in that the \ident{merge} and 
+\ident{split} operations are required to be inverse functions and to preserve the additional 
+information about the context. This was not required for the flat system where the operations
+could under-approximate or over-approximate. Note that the operations use $\xtimes$ to combine 
+or split the contained values. This means that they operate on free-variable vectors rather than 
+on ordinary products.
+
+The \ident{dup} mapping is a new operation that was not required for a flat calculus. It 
+takes a variable context with a single variable annotated with $\cclrd{r}\,\cpar\,\cclrd{s}$,
+duplicates the value of the variable $\alpha$ and splits the additional context between the two
+new variables. In a flat calculus, this operation was expressed using ordinary tuple 
+construction, which is not possible here -- the returned context needs to contain a 
+two-element vector $\alpha \xtimes \alpha$.
+
+Finally, the \ident{lift} mapping is almost the same as in the flat version. It
+operates on a singleton vector, which is equivalent to operating on a scalar as before.
+The operation could be extended to a vector in a pointwise way, but we keep it simple
+and perform sub-coeffecting on individual variables.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[!t]
+\begin{equation*}
+\begin{array}{ll}
+ \sem{\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }~\ctx =
+   \ident{counit}_{\cunit}~\ctx & (\emph{var})
+\\[0.4em]
+\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
+  \delta~(c_i) & (\emph{const})
+\\[0.4em]
+\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
+  \hspace{1.5em}
+  \begin{array}{l}  
+  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}~\ctx\\[-0.25em]
+  \kvd{in}~\sem{\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
+      (\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~\sem{\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\ctx_2)
+  \end{array}
+\\[1.4em]
+\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
+\hspace{1.75em}\sem{\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  ~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v))
+\\[0.4em]
+\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
+  \hspace{1.75em}
+  \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx~~\kvd{in}~~
+  \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}~\ctx_1
+\\[0.4em]
+\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{sub})\\
+  \hspace{1.75em}
+  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
+  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \alift{\cclrd{s'}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
+\\[0.4em]
+\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
+        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\[-0.25em]
+  \hspace{1.75em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
+     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
+  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, 
+     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
+\\[0.4em]
+\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
+        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\[-0.25em]
+  \hspace{1.75em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
+  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, 
+      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)
+\end{array}
+\end{equation*}
+\\[0.0em]
+{\small The definition uses the following auxiliary definitions:}
+\begin{equation*}
+\hspace{-2em}
+\begin{array}{l}
+ \ident{swap}_{\cclrd{t},\cclrd{s}} ~:~ 
+  \ctyp{\alift{\cclrd{t}, \cclrd{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\alift{\cclrd{s}, \cclrd{t}}}{(\beta \xtimes \alpha)} \\[-0.25em]
+\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx = \\
+  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}}~\ctx \\[-0.25em]
+  \hspace{1.75em} \kvd{in}~\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)
+ \\[0.4em]
+\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} ~:~
+  (\ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{s'}}}{\beta'}) \rightarrow
+    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta \xtimes \gamma)} \rightarrow
+    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s'}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta' \xtimes \gamma)}\\[-0.25em]
+\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f~\ctx =\\
+  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx \\[-0.25em]
+  \hspace{1.75em} \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}~\ctx' \\[-0.25em] 
+  \hspace{1.75em} \kvd{in}~\ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~
+        (\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
+\end{array}
+\end{equation*}
+\figcaption{Categorical semantics of the structural coeffect calculus}
+\label{fig:struct-semantics}
+\vspace{-0.5em}
+\end{figure*}
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of structural calculus}
+
+The concrete semantics for liveness and bounded variable use shown in 
+Sections~\ref{sec:applications-struct-live} and \ref{sec:applications-struct-bll}
+suggests that semantics of structural coeffect calculi tend to be more complex 
+than semantics of flat coeffect calculi. The complexity comes from the fact
+that we need a more expressive representation of the variable context -- \eg~a vector
+of optional values (\ie~wrapped in the \ident{Maybe} data type). Additionally,
+the structural system needs to pass separate variable contexts to the sub-expressions.
+
+The latter aspect is fully captured by the semantics shown in this section. The
+earlier point is left to the concrete notion of structural coeffect. Our model still gives
+us the flexibility of defining the concrete representation of variable vectors. We explore
+a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
+at a unified categorical semantics defined in terms of \emph{structural indexed comonads}.
+
+\paragraph{Contexts and functions.}
+In the structural coeffect calculus, expressions in context are interpreted as functions taking
+a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
+such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
+structure is annotated with a scalar, such as $\ctyp{\cclrd{r}}$, which we treat as being equivalent
+to a singleton vector annotation $\ctyp{\alift{\cclrd{r}}}$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
+  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+Note that the instances of flat indexed comonad ignored the fact that the variable 
+context wrapped in the data structure is a product. This is not generally the case for the 
+structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
+are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
+rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need to examine the structure
+of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
+
+\paragraph{Expressions.}
+A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
+the previous chapter, the semantics is written in a programming language style using constructs
+such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
+can equally be written using primitives such as currying, uncurrying, function pairing etc.
+
+The following summarizes how the standard syntax-driven rules work, highlighting the differences
+from the flat version:
+
+\begin{itemize}
+\item When accessing a variable (\emph{var}), the context now contains \emph{only} the accessed
+  variable and so the semantics is just $\ident{counit}_{\cunit}$ without a projection.
+  Constants (\emph{const}) are interpreted using a global dictionary $\delta$ as earlier.
+
+\item The semantics of flat function application first duplicated the context so that the same 
+  variables can be passed to both sub-expressions. This is no longer needed -- the 
+  (\emph{app}) rule splits the variables \emph{including} the additional context into two parts.
+  Passing the first context to the semantics of $e_1$ gives us a function
+  $\ctyp{\alift{\cclrd{t}}}{\tau_1} \rightarrow \tau_2$.
+
+  A value $\ctyp{\alift{\cclrd{t}}}{\tau_1}$ required in order to call the function is obtained by 
+  applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ to the semantics of $e_2$. The result 
+  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
+  is then called with the latter part of the split input context.
+
+\item The semantics of function abstraction (\emph{abs}) is syntactically the same as in the 
+  flat version -- the only difference is that we now merge a free-variable context with a 
+  singleton vector, both at the level of variable assignments and at the level of coeffect 
+  annotations.
+\end{itemize}
+
+\noindent
+The semantics for the non-syntax-driven rules (weakening, exchange, contraction) performs 
+transformations on the free-variable context. Weakening (\emph{weak}) splits the context and 
+ignores the part corresponding to the removed variable. If we were modelling the semantics in 
+a language with a linear type system, this would require an additional operation for ignoring an
+unused context annotated with $\aclrd{\czero}$.
+
+The remaining rules perform a transformation anywhere inside the free-variable vector. 
+To simplify writing the semantics, we define a helper 
+$\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
+that splits the variable vector into three parts, transforms the middle part and then 
+merges them, using the newly transformed middle part.
+
+The transformations on the middle part are quite simple. The (\emph{sub}) rule 
+uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
+the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
+uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a variable while 
+splitting its additional context.
+
+\paragraph{Properties.}
+
+As in the flat calculus, the main reason for defining the categorical semantics in this chapter
+is to provide validation for the design of the calculus. The following correspondence theorem
+states that the annotations in the typing rules of the structural coeffect calculus correspond 
+to the indices of the semantics. Thus, the calculus captures a context-dependent property if
+it can be modelled by a \emph{structural indexed comonad}. As we show in the next section, 
+this is the case for all three discussed examples (liveness, data-flow, bounded variable reuse).
+
+\begin{theorem}[Correspondence]
+In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$ 
+and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types 
+$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$ 
+and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined 
+by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
+\end{theorem}
+\begin{proof}
+By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
+\end{proof}
+
+\noindent
+As in the flat calculus, the primitive operations of the structural indexed comonad are each
+annotated with a unique operations provided by the coeffect annotations. This means that the
+semantics fully determines the structure of the typing rules of the structural coeffect
+calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Examples of structural indexed comonads}
+\label{sec:struct-semantics-examples}
+
+The categorical semantics for structural coeffect calculus is easily instantiated to give
+semantics for a concrete calculus. In this section, we revisit the three examples discussed
+throughout this chapter -- structural liveness, data-flow and bounded variable reuse. 
+Some aspects of the first two examples will be similar to flat versions discussed in
+Section~\ref{sec:flat-semantics} -- they are based on the same data structures (option and a 
+list, respectively), but the data structures are composed differently. Generally speaking,
+rather than having a data structure over a product of variables, we now have 
+a vector of variables over a specific data structure.
+
+The abstract semantics does not specify how vectors of variables should be represented, so this can
+vary in concrete instantiations. In all our examples, we represent a vector of variables as a 
+product written using $\times$. To distinguish between products representing vectors and ordinary
+products (\eg~a product of contexts returned by \ident{split}), we write vectors using 
+$\langle a, \ldots, b \rangle$ rather than the parentheses, used for ordinary tuples.
+
+\paragraph{Data-flow.}
+It is interesting to note that the semantics of data-flow and bounded variable reuse (discussed
+next) both keep a product of multiple values for each variable, so they are both built around an 
+\emph{indexed list} data structure. However, their \ident{cobind} and \ident{dup} operations work 
+differently. We start by looking at the structure modelling data-flow computations. For readability,
+variables in bold face (such as $\mathbf{a_1}$) range over vectors while ordinary notation (such
+as $a_1$) is used for individual values.
+
+\begin{example}[Indexed list for data-flow]
+\label{ex:struct-dfl}
+The indexed list model of data-flow computations is defined over a structural coeffect
+algebra $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
+is indexed by required number of past variables for each individual variable. It is defined
+over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
+containing a current value followed by $n_i$ past values:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
+\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
+  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{(n_k+1)-\textnormal{times}}
+\end{equation*}
+%
+The mappings that define the structural indexed comonad include the \ident{split} and \ident{merge}
+operations that are shared by the other two examples (discussed below):
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{merge}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
+  (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle) = \\[-0.25em]
+\quad \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle
+\\[0.45em]
+\ident{split}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
+   \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle = \\[-0.25em]
+\quad (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle)
+\end{array}
+\end{equation*}
+%
+The remaining mappings that are required by structural indexed comonad and capture the
+essence of data-flow computations are defined as:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{0}~\langle \langle a_0 \rangle \rangle = a_0
+\\[0.45em]
+\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
+  \langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle = \\[-0.25em]
+\quad \langle\langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle, ~\ldots~, \\[-0.25em]
+\quad ~~~\; f \langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle~\; \rangle\rangle
+\\[0.45em]
+\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{\mathit{max}(m,n)} \rangle \rangle =
+   \langle \langle a_1, \ldots, a_m \rangle, \langle a_1, \ldots, a_n \rangle \rangle
+\\[0.45em]
+\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
+  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The definition of the indexed list data structure relies on the fact that the number of annotations
+corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
+containing $n_i+1$ values for the $i$-\textit{th} variable (the annotation represents the number of 
+required \emph{past} values so one more value is required).
+
+The \ident{split} and \ident{merge} operations are defined separately, because they are not specific
+to the example. They operate on the top-level vectors of variables (without looking at the 
+representation of the variable). This means that we can re-use the same definitions for the following
+two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will there represent options
+rather than lists).
+
+The mappings that explain how data-flow computations work are \ident{cobind} (representing sequential
+composition) and \ident{dup} (representing context sharing or parallel composition). In \ident{cobind},
+we get $k$ vectors corresponding to $k$ variables, each with $m+n_i$ values. The operation calls 
+$f$ $m$-times to obtain $m$ past values required as the result of type $\ctyp{\langle m \rangle} \beta$.
+
+The $\ident{dup}_{m,n}$ operation needs to produce a two-variable context containing $m$ and $n$ values,
+respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
+is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable
+and \ident{lift} drops additional past values that are not required.
+
+~
+
+\paragraph{Bounded reuse.}
+As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list 
+structure. Rather than representing possibly different past values that can be shared (see the 
+definition of \ident{dup}), the list now represents multiple copies of the same value as each value
+can only be accessed once. This semantics follows that of Girard \cite{logic-bounded}.
+
+\begin{example}[Indexed list for bounded reuse]
+The indexed list model of bounded variable reuse is defined over a structural coeffect
+algebra $(\mathbb{N}, \ast, +, 1, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$ 
+is a vector containing $n_i$ values of $i$-\textit{th} variable:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
+\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
+  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{n_k-\textnormal{times}}
+\end{equation*}
+%
+The \ident{merge} and \ident{split} operations are defined as in Example~\ref{ex:struct-dfl}.
+The operations that capture the behaviour of bounded reuse are:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{1}~\langle \langle a_0 \rangle \rangle = a_0
+\\[0.45em]
+\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
+  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
+\\[0.45em]
+\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{m+n} \rangle \rangle = 
+\quad \langle \langle a_1, \ldots, a_m \rangle, \langle a_{m+1}, \ldots, a_{m+n} \rangle \rangle
+\\[0.45em]
+\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
+  \langle \langle a_{1,0}, \ldots a_{1,m\ast{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m\ast{n_k}} \rangle \rangle = \\[-0.25em]
+\quad \langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1 - 1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k - 1}} \rangle \rangle, ~\ldots~, \\[-0.25em]
+\quad ~~ f \langle \langle a_{1,(m-1)\ast n_1}, \ldots a_{1,(m-1) \ast n_1} \rangle, \ldots, 
+                   \langle a_{k,m\ast n_k - 1}, \ldots a_{k,m\ast n_k - 1} \rangle \rangle~\; \rangle
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{counit} and \ident{lift} operations are defined as previously -- variable access 
+extracts the only value of the only variable and sub-coeffecting allows discarding multiple
+copies of a value that are not needed.
+
+In the bounded variable reuse system, variable sharing is annotated with the $+$ operator (in contrast 
+with \textit{max} used in data-flow). The $\ident{dup}_{m,n}$ operation thus splits the $m+n$ available
+values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The 
+\ident{cobind} operation works similarly -- it splits $m\ast n_i$ available values of each
+variable into $m$ vectors containing $n_i$ copies and then calls the $f$ function $m$-times to 
+obtain $m$ resulting values without sharing any input value.
+
+\paragraph{Liveness.}
+In both data-flow and bounded reuse, the data type is defined as a vector of values obtained
+by applying the indexed list type constructor to types of individual variables. We can
+generalize this pattern. Given a parameterized (indexed) type constructor $D^l \alpha$, we define 
+$\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of a vector of $D^{l_i}$ types.
+For liveness, the definition lets us reuse some of the mappings used when defining the semantics of
+flat liveness. However, we cannot fully define the semantics of the structural version in terms of 
+the flat version -- the \ident{cobind} operation is different and we need an appropriate 
+\ident{dup} operation.
+
+\begin{example}[Structural indexed option]
+Given a structural coeffect algebra formed by $(\{ \ident{L}, \ident{D} \}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$
+and the indexed option data type $D^l$, such that $D^{\ident{D}}\alpha = 1$ and $D^{\ident{L}}\alpha=\alpha$,
+the data type for structural indexed option comonad is:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
+  D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
+\end{equation*}
+%
+The \ident{merge} and \ident{split} operations are defined as earlier. The remaining operations
+model variable liveness as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{cobind}_{\ident{L}, \langle l_1, \ldots, l_n \rangle}~f~\langle a_1, \ldots, a_n \rangle = \langle f~\langle a_1, \ldots, a_n \rangle \rangle \\[-0.25em]
+\ident{cobind}_{\ident{D}, \langle \ident{D}, \ldots, \ident{D} \rangle}~f~\langle (), \ldots, () \rangle = \langle \ident{D} \rangle
+\\[0.5em]
+\hspace{-0.5em}
+\begin{array}{lll}
+~\\[-0.85em]
+\ident{dup}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle (), () \rangle \\[-0.25em]
+\ident{dup}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle a, () \rangle \\[-0.25em]
+\ident{dup}_{\ident{D}, \ident{L}} \langle a \rangle &\narrow{=}& \langle (), a \rangle \\[-0.25em]
+\ident{dup}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a, a \rangle 
+\end{array}
+\qquad\quad
+\begin{array}{lll}
+\ident{counit}_{\ident{L}}~\langle a \rangle &\narrow{=}& a
+\\[0.45em]
+\ident{lift}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a \rangle \\[-0.25em]
+\ident{lift}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
+\ident{lift}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
+\end{array}
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+When the expected result of the \ident{cobind} operation is dead (second case), the operation can
+ignore all inputs and directly return the unit value $()$. Otherwise, it passes the vector of 
+input variables to $f$ as-is -- no matter whether the individual values are live or dead. The
+\ident{L} annotation is a unit with respect to $\cap$ and so the annotations expected by $f$ are
+the same as those required by the result of \ident{cobind}.
+
+The \ident{dup} operation resembles with the flat version of \ident{split} -- this is expected as
+duplication in the flat calculus is performed by first duplicating the variable context (using 
+\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair. Depending on the
+required coeffect annotations, this may copy (duplicate) the value, or it may produce an empty context.
+
+Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
+$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models 
+sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
+it behaves as identity.
+
+% ==================================================================================================
+%                                                                   
+%       ###                  #                    #       #          
+%      #   #                 #                    #                  
+%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
+%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
+%          #  #  ##  #   #   #      ####  #       #       #    #     
+%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
+%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
+%             #   #                                                  
+%              ###                                                   
+%
+% ==================================================================================================
+
+\section{Syntactic equational theory}
+\label{sec:struct-syntax}
+
+Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
+different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
+certain equational properties hold in all instances of the structural coeffect calculus.
+
+In this section, we look at these common properties that we get ``for free'' from the structural 
+coeffect calculus. We start the discussion by briefly considering the key aspects
+that make the equational theory of flat and structural coeffects different.
+
+\subsection{From flat coeffects to structural coeffects}
+\label{sec:struct-syntax-intro}
+
+When discussing syntactic reductions for the flat calculus in Section~\ref{sec:flat-syntax}, 
+we noted that no single technique of specifying syntactic reduction works universally for all 
+flat coeffect calculi. We considered multiple alternatives (call-by-name, call-by-value, 
+internalized substitution) that can be used as the basis for operational semantics for
+different calculi that satisfy certain additional properties.
+
+The structural coeffect calculus has more preferable equational properties. In particular, 
+we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of structural 
+coeffect algebra. For this reason, we focus on these two reductions in this section.
+Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
+the structural coeffect calculus satisfies both the \emph{local soundness} and 
+the \emph{local completeness} properties, witnessed by $\beta$-reduction and $\eta$-expansion,
+respectively.
+
+\paragraph{Substitution for flat coeffects.} 
+The less obvious (\emph{top-pointed}) variant of the substitution lemma for flat coeffects 
+(Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
+to coincide. This enables substitution to preserve the type of expressions, because all
+additional requirements arising as the result of the substitution can be associated with the
+declaration context. For example, consider the following example where implicit parameter 
+$\ident{?offset}$ is substituted for the variable $y$:
+%
+\begin{equation*}
+\begin{array}{rclcll}
+ \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textit{before}) \\
+\coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
+  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textit{after})
+\end{array}  
+\end{equation*}
+%
+The typing judgement obtained in (\emph{after}) preserves the type of the expression (function 
+value) from the original typing (\emph{before}). This is possible thanks to the non-determinism 
+involved in the typing rule for lambda abstraction -- as all operators of the flat coeffect 
+algebra used here are $\cup$, we can place the additional requirement on the outer context. 
+Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.
+
+Here, the flat coeffect calculus gives us typing with limited \emph{precision}, but 
+enough \emph{flexibility} to prove the substitution lemma.
+
+\paragraph{Substitution for structural coeffects.}
+In contrast, the substitution lemma (Lemma~\ref{thm:structural-substitution}, shown later on 
+page~\pageref{thm:structural-substitution}) for structural coeffects can be proven because 
+structural coeffect systems provide enough \emph{precision} to identify exactly with which 
+variable should a context requirement be associated.
+
+The following example shows a situation similar to the previous one. Here, we use
+structural data-flow calculus (writing $\kvd{prev}~e$ to obtain previous value of the 
+expression $e$) and we substitute $w+z$ for $y$:
+%
+\begin{equation*}
+\begin{array}{rclcll}
+ \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} } 
+  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{before}) \\
+\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
+  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{after}) \\
+\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
+  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{final})
+\end{array}  
+\end{equation*}
+%
+The type of the function does not change, because the structural type system associates
+the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not 
+affect how the variable $x$ is used.
+
+The other aspect demonstrated in the example is how the coeffect of the substituted 
+variable affects the free-variable context of the substituted expression. Here, the
+original variable $y$ is annotated with $\cclrd{2}$ and we substitute it for an
+expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
+The substitution applies the operation $\aseq$ (modelling sequential composition) to 
+the annotation of the new context -- in the above example
+$\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}=\alift{\cclrd{2}, \cclrd{2}}$.
+
+\subsection{Holes and substitution lemma}
+\label{sec:struct-syntactic-subst}
+
+As demonstrated in the previous section, reduction (and substitution) in the structural coeffect 
+calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
+More importantly, because the system uses explicit contraction, we may also need to substitute 
+for multiple variables in the variable context at the same time.
+
+Consider the expression $\lambda x.x+x$. It is type-checked by type-checking $x_1 + x_2$,
+contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction of 
+$(\lambda x.x+x)~(y+z)$ we need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
+This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
+such as the bunched typing system \cite{substruct-bunched}. To express the substitution 
+lemma later in this section, we first define the notion of a \emph{context with holes}:
+
+\begin{definition}[Context with holes]
+A \emph{context with holes} is a context such as $\coctx{x_1\!:\!\tau_1, \ldots, x_k\!:\!\tau_k}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_k}}}$,
+where some of the variable typings $x_i\!:\!\tau_i$ and corresponding coeffects $\cclrd{r_i}$ 
+are replaced by \emph{holes} written as $-$.
+
+\begin{equation*}
+~\Delta[\,\coctx{-}{-}\,]_n = \Delta[\,\underbrace{\coctx{-}{-} \,|\, \ldots \,|\, \coctx{-}{-}}_{ n-\textit{times} } \,]
+\end{equation*}
+\begin{equation*}
+\begin{array}{lcl}
+ \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
+   \coctx{-, \Gamma}{\alift{-} \atimes \aclrd{\textbf{s}} }
+   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_{n-1} \\
+ \Delta[\,\coctx{-}{-}\,]_n &\narrow{:=}&
+   \coctx{x\!:\!\tau, \Gamma}{\alift{\cclrd{r}} \atimes \aclrd{\textbf{s}} }
+   \qquad \textnormal{where}~\coctx{\Gamma}{\aclrd{\textbf{s}}} \in \Delta[\,\coctx{-}{-}\,]_n \\
+ \Delta[\,\coctx{-}{-}\,]_0 &\narrow{:=}&
+   \coctx{()}{\alift{}}
+\end{array} 
+\end{equation*} 
+\end{definition}
+
+\noindent
+A context with $n$ holes may either start with a hole, followed by a context with $n-1$ holes, or it may
+start with a variable followed by a context with $n$ holes. Note that the definition ensures that
+the locations of variable holes correspond to the locations of coeffect annotation holes. Given a context
+with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
+obtain an ordinary coeffect-annotated context.
+
+\begin{definition}[Hole filling] Given a context with $n$ holes $\coctx{\Delta}{\textbf{s}} \in 
+\Delta[\,\coctx{-}{-}\,]_n$, the hole filling operation written as 
+$\coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\mathbf{r_1}}} \,|\, \ldots \,|\, \coctx{\Gamma_n}{\aclrd{\mathbf{r_n}}} \,]$,
+which replaces the holes by the specified variables and corresponding coeffect annotations, is defined as:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \coctx{-, \Delta}{\alift{-} \atimes \textbf{s} }
+    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
+      \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}_1} \atimes \aclrd{\textbf{r}_2} } \\
+ && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
+     \coctx{\Delta}{\textbf{s}}[\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
+\\[0.5em]
+ \coctx{x_1\!:\!\tau, \Delta}{\alift{\cclrd{r_1}} \atimes \textbf{s} }
+    ~[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] &\narrow{=}&
+      \coctx{x_1\!:\!\tau, \Gamma_2}{\alift{\cclrd{r_1}} \atimes \aclrd{\textbf{r}_2} } \\
+ && \qquad \hspace{-9.3em} \textnormal{where}~\coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} =
+     \coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\textbf{r}_1}} \,|\, \coctx{\Gamma_2}{\aclrd{\textbf{r}_2}} \,|\, \ldots \,] 
+\\[0.5em]
+ \coctx{()}{\alift{}}~[~] &\narrow{=}& \coctx{()}{\alift{}}
+\end{array} 
+\end{equation*} 
+\end{definition}
+
+\noindent
+When we substitute an expression with coeffects $\aclrd{\textbf{t}}$ (associated with variables $\Gamma$)
+for a variable that has coeffects $\cclrd{s}$, the resulting coeffects of $\Gamma$ need to 
+combine $\aclrd{\textbf{t}}$ and $\cclrd{s}$. Unlike in the flat coeffect systems, the structural 
+substitution does not require all coeffect algebra operations to coincide and so the combination
+is more interesting than in the bottom-pointed substitution for flat coeffects, where it used
+the only available operator (Lemma~\ref{thm:cbn-substitution-bot}).
+
+Substitution can be seen as a form of sequencing, \ie~the term $\subst{e_1}{x}{e_2}$ should have
+the same type as the term $\kvd{let}~x=e_2~\kvd{in}~e_1$. In other words, we first need to obtain the 
+value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
+requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
+are annotated with $\cclrd{s} \,\aseq\, \aclrd{\textbf{t}}$, using the (scalar-vector extension)
+of the sequential composition operator $\cseq$.
+
+An operational reading of the resulting coeffect annotation is that, to evaluate each of the variables
+annotated with coeffects in the vector $\aclrd{\textbf{t}}$, we first need to evaluate the substituted
+expression with coeffects $\cclrd{s}$, followed by the rest of the expression with coeffects specified
+by $\aclrd{\textbf{t}}$.
+
+~
+
+~
+
+~
+
+\begin{lemma}[Multi-nary substitution]
+\label{thm:structural-substitution}
+Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
+%
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_1}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
+  \coctx{x_k\!:\!\tau_k}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
+\end{equation*}
+%
+and a expressions $e_i$ with free-variable contexts $\Gamma_i$ annotated with $\aclrd{\mathbf{t_i}}$:
+%
+\begin{equation*}
+\coctx{\Gamma_1}{\aclrd{\textbf{t}_1}} \vdash e_1 : \tau_1 
+\quad \ldots \quad
+\coctx{\Gamma_k}{\aclrd{\textbf{t}_k}} \vdash e_k : \tau_k
+\end{equation*}
+%
+then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
+where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
+%
+\begin{equation*}
+\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_1}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
+  \coctx{\Gamma_k}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
+\end{equation*}
+\end{lemma}
+\begin{proof}
+By induction over $\vdash$, using the multi-nary aspect of the substitution 
+in the proof of the contraction case (see Appendix~\ref{sec:appendix-struct-cbn}).
+\end{proof}
+
+\subsection{Reduction and expansion}
+
+In the Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
+proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
+$\beta$-reduction, which encompasses both call-by-value and call-by-name. We also show that 
+$\eta$-expansion preserves the types. Both of the properties hold for a system with any
+structural coeffect algebra.
+
+\paragraph{Reduction theorem.} In a full $\beta$-reduction,
+written as $\rightarrow_\beta$, we can replace the redex $(\lambda x.e_2)~e_1$ by the
+expression $\subst{e_r}{x}{e_s}$ anywhere inside a term. The subject reduction theorem
+guarantees that this does not change the type of the term.
+
+\begin{theorem}[Type preservation]
+In a structural coeffect system with a structural coeffect algebra formed by 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
+if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
+and $e \rightarrow_{\beta} e'$ using the full $\beta$-reduction then
+$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+Consider the typing derivation for the redex $(\lambda x.e_r)\,e_s$ before the reduction
+(note that this is similar to the typing of let binding discussed in Remark~\ref{thm:structural-letbinding},
+but we do not swap the two parts of the free-variable context):
+%
+\begin{equation*}
+\inference
+  { \begin{array}{l}\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s\vspace{-2em}\end{array} &
+    \inference
+      { \coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r }
+      { \coctx{\Gamma_r}{\aclrd{\textbf{r}}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{t}} \tau_r} \vspace{0.2em} }
+  { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})} \vdash (\lambda x.e_r)~e_s : \tau_r }
+\end{equation*}
+%
+For the substitution lemma, we first rewrite the typing judgement for $e_r$, \ie~
+$\coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r$
+as a context with a single hole filled by the $x$ variable:
+$\coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes\,-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r$.
+Now we can perform the substitution using Lemma~\ref{thm:structural-substitution}:
+%
+\begin{equation*}
+\inference
+  { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r & 
+    \coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s }
+  { \inference
+      { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{\Gamma_s}{\cclrd{t}\,\aseq\,\aclrd{\textbf{s}} }\,] \vdash \subst{e_r}{x}{e_s} : \tau_r }
+      { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}}\,\atimes\,(\cclrd{t}\,\aseq\,\aclrd{\textbf{s}})} \vdash \subst{e_r}{x}{e_s} : \tau_r } }
+\end{equation*}
+%
+The last step applies the hole filling operation, showing that substitution preserves 
+the type of the term.
+\end{proof}
+
+\noindent
+Because of the vector (free monoid) structure of coeffect annotations $\aclrd{\textbf{r}}$, 
+$\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$, these are uniquely associated with $\Gamma_r$, 
+$\Gamma_s$, and $x$ respectively. Therefore, substituting $e_s$ (which has coeffects 
+$\aclrd{\textbf{s}}$) for $x$ introduces the context requirements specified by $\aclrd{\textbf{s}}$
+which are composed with the requirements $\cclrd{t}$ associated with $x$, \ie~the variable being 
+substituted.
+
+\paragraph{Expansion theorem.}
+Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
+\emph{local soundness} and \emph{local completeness} as required by Pfenning and Davies
+\cite{logic-modal-reconstruction}. Informally, this means that abstraction does not introduce 
+too much, and application does not eliminate too much.
+
+\begin{theorem}[$\eta$-expansion]
+In a structural coeffect system with a structural coeffect algebra formed by 
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
+if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
+and $e \rightarrow_{\eta} e'$ using the full $\eta$-reduction then
+$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+The following derivation shows that $\lambda x.f~x$ has the same type and coeffects as 
+the original expression $f$:
+\begin{equation*}
+\inference
+  { \inference
+    { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
+      \coctx{x\!:\!\tau_1}{\alift{\cunit}} \vdash x:\tau_1 }
+    { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}} \;\atimes\;(\cclrd{s}\,\aseq\,\alift{\cunit}) } \vdash f~x : \tau_2 } }
+  { \inference
+    { \coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}} \;\atimes\,\alift{\cclrd{s}} } \vdash f~x : \tau_2 }
+    { \coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+\end{equation*}
+%
+The second step uses the fact that 
+$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
+arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
+\end{proof}
+
+\noindent
+The $\eta$-expansion property discussed in this section highlights another difference 
+between coeffects and effects. The $\eta$-equality property does not hold for many notions 
+of effect. For example, in a language with output effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ 
+has different effects to its $\eta$-converted form $\lambda x . e\,x$ because the immediate 
+effects of $e$ are hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, 
+the (\emph{abs}) rule allows immediate contextual requirements of $e$ to ``float outside'' 
+of the enclosing $\lambda$. Furthermore, the free monoid nature of $\atimes$ in structural 
+coeffect systems allows the exact immediate requirements of $\lambda x . e x$ to match 
+those of $e$.
 
@@
 %    #   #                         #                                      
-%       #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###  
-%       #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #     
-%       #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###  
-%       #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      # 
-%        ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####  
+%    #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##  
+%    #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  # 
+%    #      #   #  #   #  #        #    #   #   ###     #    #   #  #   # 
+%    #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   # 
+%     ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   # 
 %
@@
 
-In this chapter, we completed our goal of presenting a \emph{unified} system for tracking contextual
-properties and then we discussed two important topics of future and related work that require 
-in-depth discussion.
-
-The key technical contribution of this chapter is the \emph{unified coeffect system} 
-(Section~\ref{sec:unified-unified}), which unifies the flat and structural system presented
-in the previous two chapters. To achieve this, we introduced \emph{coeffect container}, which
-determines how are coeffect annotations attached to variable contexts. We then discussed two
-instances of the structure that capture flat and structural properties.
-
-In the rest of the chapter, we discussed two topis of future and related work. First
-(Section~\ref{sec:unified-impl}), we considered pathways to practical implementations of
-coeffect systems, including a discussion of a novel lattice-based coeffect system, which is simpler,
-but cannot capture all our motivating examples. Finally, we discussed how our work relates to 
-meta-language based on comonads (Section~\ref{sec:unified-meta}). We present a \emph{coeffect 
-meta-language} that follows similar style to CMTT, but is based on indexed comonads. Next, we 
-discussed how to embed flat coeffect systems in this coeffect meta-language, which elucidates the 
-relationship between our work and CMTT.
+This chapter completes the key development of this thesis -- the presentation of the coeffect 
+framework, consisting of two calculi capturing properties of context-aware computations introduced
+in Chapter~\ref{ch:applications}. In the previous chapter, we focused on the whole-context 
+properties of computations and we developed \emph{flat coeffect calculus} to capture them.
+This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
+contextual properties. The system provides a precise analysis of liveness and data-flow and allows 
+other interesting uses such as tracking of variable accesses based on bounded linear logic.
+
+Following the structure of the previous chapter, the structural coeffect calculus is parameterized
+by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations 
+$\cseq$ and $\cpar$ that model sequential and pointwise composition, respectively. For flat 
+coeffects, we required $\czip$ to model context merging. For structural coeffects, we 
+instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
+as $\czip$.
+
+In order to keep track of separate annotations for each variable, we use a system with 
+explicit structural rules (contraction, weakening and exchange) that manipulate the structure
+of variables and the structure of annotations at the same time.
+
+The chapter presents two technical results. The first is the semantics of structural coeffect
+calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
+it unifies our three examples (bounded reuse, data-flow and liveness) and it demonstrates
+suitability of the type system. The second technical result is equational theory for the structural
+coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
+typing preservation property for any instance of the structural coeffect calculus. These two 
+strong properties are desirable for programming languages, but are often not satisfied 
+(\eg~by languages with effects).
+
+Two questions remain open for the upcoming chapter. Firstly, is there a way to present 
+flat and structural coeffect calculi in a uniform way as a single system? Secondly, what
+is the best pathway towards a practical implementation of programming languages based on 
+coeffect systems?

### 2016-02-05 text/Chapter04.tex
Choosing unique typing

@@
 $\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
-idempotent monoid and $(\C, \czip)$ is a band (idempotent semigroup). That is, for all $r,s,t\in \C$: 
+idempotent monoid, $(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup). 
+That is, for all $r,s,t\in \C$: 
 %
@@
 r\; \czip\; r = r \qquad
+\\
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
+t\; \cleq\; t \\[0.5em]
 \end{array}
@@
 are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
-(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}).
+(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.
 
@@
 $\cpar$ appears in the conclusion and combines two context requirements from the assumptions.
+
+\paragraph{Ordering.}
+
+The flat coeffect algebra requires a pre-order relation $\cleq$. This will be used to introduce
+sub-coeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
+coeffect algebra, as it will be useful for characterization of different kinds of coeffect calculi.
+When the idempotent monoid $(\C, \cpar, \czero)$ also has the commutative property (\ie~forms a 
+semi-lattice), the $\cleq$ relation can be defined as the ordering of the semi-lattice:
+%
+\begin{equation*}
+\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
+\end{equation*}
+%
+This definition is consistent with all three examples that motivate flat coeffect calculus, but
+it cannot be used with the structural coeffects (where it fails for the bounded reuse 
+calculus) and so we choose not to use it.
+
+Furthermore, the $\cunit$ coeffect is often the top or the bottom element of the semi-lattice. 
+As discussed in Section~\ref{sec:flat-syntax}, when this is the case, we are able to prove certain 
+syntactic properties of the calculus.
 
@@
 \end{equation*}
-%
+
+\noindent
 There are two possible typings allowed by the typing rules discussed in Section~\ref{sec:flat-calculus-types}
@@
 \paragraph{Properties.}
-If a written in a coeffect language with implicit parameters is well-typed according to the 
+If a program written in a coeffect language with implicit parameters is well-typed according to the 
 type system presented in Figure~\ref{fig:flat-resolve-impl}, then the type system gives a unique
@@
 \label{thm:flat-impl-invert} 
-~
+For the type system defined in Figure~\ref{fig:flat-resolve-impl}:
 %
@@
 \raggedright
-\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash x : \tau$ then $x : \tau \in \Gamma$ and $\cclrd{c} = \cunit$.
-\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash n : \tau$ then $\tau = \ident{num}$ and $\cclrd{c} = \czero$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash x : \tau$ then $x : \tau \in \Gamma$ and $\cclrd{c} = \cclrd{\emptyset}$.
+\item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash n : \tau$ then $\tau = \ident{num}$ and $\cclrd{c} = \cclrd{\emptyset}$.
 \item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash e_1~e_2 : \tau_2$
@@
 \item If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{c}} \vdash \lambda x\!:\!\tau_1.e : \tau$ then 
-  $\tau = \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ for some $\tau_2$ with 
+  there is some $\tau_2$ such that $\tau = \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
   ${\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2}$ and
@@
 Case (\emph{abs}): $e = \lambda x\!:\!\tau_1.e_1$ and $\cclrd{c}=\cclrd{\Delta}$.
-  For some $\cclrd{r}$, 
-  $\tau = \tau_1 \xrightarrow{\cclrd{r}\setminus\cclrd{\Delta}} \tau_2$ and also
+  $\tau = \tau_1 \xrightarrow{\cclrd{r}\setminus\cclrd{\Delta}} \tau_2$ for some $\cclrd{r}, \tau_2$ and also
   $\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2}$. 
@@
   ${\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau_2'}$.
-  By the induction hypothesis $\tau_2 = \tau_2'$ and so $\tau = \tau'$ and 
-  also $\cclrd{c}=\cclrd{c'}=\cclrd{\Delta}$.
+  By the induction hypothesis $\tau_2 = \tau_2'$ and $\cclrd{c}=\cclrd{c'}$ and 
+  therefore $\tau = \tau'$.
 
@@
 
-\newpage
-\newpage
+Resolving the ambiguity for liveness and dataflow computations is easier than for implicit 
+parameters. It suffices to use a lambda abstraction rule that duplicates the coeffects of the 
+body:
+%
+\begin{equation*}
+\tyrule{idabs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
+\end{equation*}
+%
+This is the rule that we originally used for liveness and dataflow computations in 
+Chapter~\ref{ch:applications}. This rule cannot be used with implicit parameters and so 
+the additional flexibility provided by the $\czip$ operator is needed in the general flat
+coeffect calculus.
+
+For livenes and dataflow, the (\emph{idabs}) rule provides the most precise coeffect.
+Assume we have lambda abstraction with body that has coeffects $\cclrd{r}$. The 
+ordinary (\emph{abs}) rule requires us to find $\cclrd{s}, \cclrd{t}$ such that
+$\cclrd{r}=\cclrd{s}\,\czip\,\cclrd{t}$.
+
+\begin{itemize}
+  \item[--] For dataflow, this is $\cclrd{r}=\mathit{min}(\cclrd{s}, \cclrd{t})$. The smallest
+    $\cclrd{s}, \cclrd{t}$ such that the equality holds are $\cclrd{s}=\cclrd{t}=\cclrd{r}$.
+ \item[--] For livenes, this is $\cclrd{r}=\cclrd{s} \sqcup \cclrd{t}$. When $\cclrd{r}=\ident{L}$,
+   the only solution is $\cclrd{s}=\cclrd{t}=\ident{L}$; when $\cclrd{r}=\ident{D}$, the most
+   precise solution is $\cclrd{s}=\cclrd{t}=\ident{D}$ because $\ident{D}\sqsubseteq\ident{L}$.    
+\end{itemize}
+
+The notion of ``more precise'' solution can be defined in terms of sub-coeffecting and subtyping.
+We return to this topic in Section~\ref{sec:flat-exts-lambda} and we also precisely characterise
+for which coeffect system is the (\emph{idabs}) rule preferable over the (\emph{abs}) rule.
+
+\paragraph{Properties.}
+If a program written in a coeffect language for liveness or dataflow is well-typed according to
+the type system presented in Figure~\ref{fig:flat-types} with the (\emph{abs}) rule replaced by
+(\emph{idabs}), then the type system gives a unique derivation. As for implicit parameters, this
+defines the semantics of coeffect program (Chapter~\ref{ch:transl}) and it is used in the
+implementation (Chatper~\ref{ch:impl}).
+
+In order to prove the uniqueness of typing theorem (Theorem~\ref{thm:flat-ldf-unique}), we first
+need the inversion lemma (Lemma~\ref{thm:flat-ldf-invert}).
+
+\begin{lemma}[Inversion lemma for liveness and dataflow]
+\label{thm:flat-ldf-invert} 
+For the type system defined in Figure~\ref{fig:flat-types} with the (abs) rule replaced
+by (idabs):
+%
+\begin{enumerate}
+\raggedright
+\item If $\coctx{\Gamma}{\cclrd{c}} \vdash x : \tau$ then $x : \tau \in \Gamma$ and $\cclrd{c} = \cunit$.
+\item If $\coctx{\Gamma}{\cclrd{c}} \vdash n : \tau$ then $\tau = \ident{num}$ and $\cclrd{c} = \czero$.
+\item If $\coctx{\Gamma}{\cclrd{c}} \vdash e_1~e_2 : \tau_2$
+ then there is some $\tau_1,\cclrd{r},\cclrd{s}$ and $\cclrd{t}$ such that
+ $\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2$ and 
+ $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1$ and also
+ $\cclrd{c} = \cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})$.
+\item If $\coctx{\Gamma}{\cclrd{c}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$
+ then there is some $\tau_1, \cclrd{s}$ and $\cclrd{r}$ such that
+ $\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1$ and
+ $\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}$ and also
+ $\cclrd{c} = \cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})$.
+\item If $\coctx{\Gamma}{\cclrd{c}} \vdash \lambda x\!:\!\tau_1.e : \tau$ then there is some  
+  $\tau_2$ such that $\tau = \tau_1 \xrightarrow{\cclrd{c}} \tau_2$ and
+  ${\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{c}} \vdash e : \tau_2}$.
+\end{enumerate}  
+\end{lemma}
+\begin{proof}
+Follows from the individual rules given in Figure~\ref{fig:flat-resolve-impl}.
+\end{proof}
+
+\begin{theorem}[Uniqueness of coeffect typing for liveness and dataflow]
+\label{thm:flat-ldf-unique}
+In the type system for implicit parameters defined in Figure~\ref{fig:flat-types} with the (abs)
+rule replaced by (idabs), when $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and 
+$\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau'$ then $\tau = \tau'$ and $\cclrd{r} = \cclrd{r'}$.
+\end{theorem}
+\begin{proof}
+Suppose that (A) $\coctx{\Gamma}{\cclrd{c}} \vdash e : \tau$ and 
+(B) $\coctx{\Gamma}{\cclrd{c'}} \vdash e : \tau'$. We show by induction over the typing 
+derivation of $\coctx{\Gamma}{\cclrd{c}} \vdash e : \tau$ that $\tau = \tau'$ and $\cclrd{c}=\cclrd{c'}$.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{abs}): $e = \lambda x\!:\!\tau_1.e_1$. $\tau = \tau_1 \xrightarrow{\cclrd{c}} \tau_2$ 
+  for some $\tau_2$ and $\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{c}} \vdash e : \tau_2}$. 
+  By case (5) of Lemma~\ref{thm:flat-ldf-invert}, the final rule of the derivation (B) must
+  have also been (\emph{abs}) and this derivation has a sub-derivation with a conclusion
+  ${\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{c'}} \vdash e : \tau_2'}$.
+  By the induction hypothesis $\tau_2 = \tau_2'$ and $\cclrd{c}=\cclrd{c'}$ and therefore
+  also so $\tau = \tau'$.
+
+\vspace{0.5em}\noindent\hangindent=0.7cm 
+Cases (\emph{var}), (\emph{const}) are direct consequence of Lemma~\ref{thm:flat-impl-invert}.
+
+\vspace{0.5em}\noindent\hangindent=0.7cm 
+Cases (\emph{var}), (\emph{const}), (\emph{app}), (\emph{let}), (\emph{param}) and (\emph{letpar})
+  similarly to (\emph{abs}).
+\end{proof}
+
+\paragraph{Implementation.}
+As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
+require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
+by (\emph{idabs}) can be easily obtained by generating additional constraints:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}} 
+\end{equation*}
+%
+Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
+coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by 
+(\emph{idabs}).
+
+% ==================================================================================================
+%
+%    #####
+%    #     # #   # #    # #####   ##    ####  ##### #  ####
+%    #        # #  ##   #   #    #  #  #    #   #   # #    #
+%     #####    #   # #  #   #   #    # #        #   # #
+%          #   #   #  # #   #   ###### #        #   # #
+%    #     #   #   #   ##   #   #    # #    #   #   # #    #
+%     #####    #   #    #   #   #    #  ####    #   #  ####
+%
+% ==================================================================================================
+
+\section{Syntactic equational theory}
+\label{sec:flat-syntax}
+
+Each of the concrete coeffect calculi discussed in this chapter has a different notion of context,
+much like various effectful languages have different notions of effects (such as exceptions or
+mutable state). However, in all of the calculi, the context has a number of common properties that 
+are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
+that hold for all of the coeffect systems. Further properties hold for systems where the context
+satisfies additional properties.
+
+In this section, we look at such shared syntactic properties. This accompanies the previous section, 
+which provided a \emph{semantic} justification for the axioms of coeffect algebra with a 
+\emph{syntactic} justification. Operationally, this section can also be viewed as providing a 
+pathway to an operational semantics for two of our systems (implicit parameters and liveness),
+which can be based on syntactic substitution. As we discuss later, the notion of context for
+data-flow is more complex.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Syntactic properties}
+\label{sec:flat-syntax-props}
+
+Before discussing the syntactic properties of general coeffect calculus formally, it should be 
+clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
+that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
+Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
+equations define four syntactic reductions on the terms:
+%
+\begin{equation*}
+\begin{array}{rllcl}
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
+(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
+(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
+e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
+\end{array}
+\end{equation*}
+%
+The rules capture syntactic reductions that can be performed in a general calculus, without any 
+knowledge of the specific notion of context. If the reductions preserve the type of the expression
+(type preservation), then operational semantics can be defined as a repeated application of the 
+rules, until a specified normal form (\ie~a value) is reached.
+
+The first two equations define well-known call-by-name and call-by-value reductions. The \kvd{glet} 
+reduction (inspired by Filinski \cite{monads-inaction}) models a system where program is reduced to 
+a sequence of primitive operations, sequentially composed using the special \kvd{glet} construct. 
+In the rest of the section, we briefly outline the interpretation of the four 
+rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name 
+(Section~\ref{sec:flat-syntax-cbn}) in more details.
+
+The focus of our work is on the general coeffect system and so we do not discuss the operational
+semantics of the specific notions of context. However, some work in that area has been done
+by Brunel et al. \cite{coeffects-quantitative}.
+
+\paragraph{Call-by-name.} 
+In call-by-name, the argument is syntactically substituted for all occurrences of a variable. It
+can be used as the basis for operational semantics of purely functional languages. However, using
+the rule in effectful languages breaks the \emph{type preservation} property. For example, consider
+a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
+A function $\lambda x.y$ is effect-free:
+%
+\begin{equation*}
+y\!:\!\tau_1 \vdash \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 \;\&\; \emptyset
+\end{equation*}
+%
+Substituting an expression $e$ with effects $\{ \ident{write} \}$ for $y$ changes the type of 
+the function by adding latent effects (without changing the immediate effects):
+%
+\begin{equation*}
+\vdash \lambda x.e : \tau_1 \xrightarrow{ \{ \ident{write} \} } \tau_2 \;\&\; \emptyset
+\end{equation*}
+%
+Similarly to effect systems, substituting a context-dependent computation $e$ for a variable $y$ can
+add latent coeffects to the function type. However, this is not the case for \emph{all} flat coeffect
+calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
+parameters system. This means that certain coeffect systems support call-by-name evaluation strategy
+and could be embedded in purely functional language (such as Haskell).
+
+\paragraph{Call-by-value.}
+
+The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
+first reduced to a \emph{value} before performing the substitution. In effectful languages, 
+value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
+values are identifiers $x$ or functions $(\lambda x.e)$.
+
+The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
+A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
+immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
+sense \emph{and} has not immediate context requirements. We define this formally in 
+Section~\ref{sec:flat-syntax-cbv}.
+
+The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
+including all our three examples. However, it is rather weak -- in order to use it, the concrete 
+semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
+
+\paragraph{Local soundness and completeness.}
+Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
+are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
+introduce a function arrow (lambda abstraction) and eliminate it (application) are sufficiently 
+strong, but not too strong.
+
+The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
+already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
+coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Call-by-value evaluation}
+\label{sec:flat-syntax-cbv}
+
+As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
+calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
+term to a \emph{value} has to be provided for each system. Syntactic values are defined
+in the usual way:
+%
+\begin{equation*}
+\begin{array}{llll}
+v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
+n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
+e\in\mathit{Expr}     & e &::=& v \sep n 
+\end{array}
+\end{equation*}
+%
+The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
+coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
+not automatically defer all context requirements.
+
+\begin{definition} 
+An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if it is a syntactic value,
+\ie~$e \in \mathit{SynVal}$ and it has no context-dependencies, \ie~$\coctx{\Gamma}{ \cunit } \vdash e : \tau$.
+\end{definition}
+
+\noindent
+The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
+variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
+context requirements of the term:
+
+\begin{lemma}[Call-by-value substitution]
+\label{thm:flat-subst-cbv}
+In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
+$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
+change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
+\end{lemma}
+\begin{proof}
+By induction over the type derivation, using the fact that $x$ and $v$ are annotated
+with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
+\end{proof}
+
+\noindent
+The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
+proving that call-by-value reduction preserves typing requires an additional constraint on the
+flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
+by the (\emph{approximation}) property:
+
+\begin{equation*}
+\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
+\hspace{10em}(\textit{approximation})
+\end{equation*}
+%
+Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
+under-approximates the actual context capabilities while the $\cpar$ operation (combining of
+context requirements) over-approximates the actual context requirements.
+
+The property holds for the three systems we consider -- for implicit parameters, this is 
+an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
+is smaller than the least upper bound.
+
+Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
+$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the type preservation theorem is
+stated as follows:
+
+\begin{theorem}[Type preservation for call-by-value]
+\label{thm:cbv-reduction}
+In a flat coeffect system with the (\emph{approxi\-mation}) property, 
+if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
+\begin{equation*}
+\inference
+  { \inference
+  { \inference
+      {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
+      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
+   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
+\end{equation*}
+The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
+From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
+As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
+%
+\begin{equation*}
+\tyrule{sub}
+ { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
+ { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
+\end{equation*}
+%
+Comparing the final conclusions of the above two typing derivations shows that 
+the reduction preserves type and coeffect annotation.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Call-by-name evaluation}
+\label{sec:flat-syntax-cbn}
+
+When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the 
+sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$. 
+As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in 
+general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two 
+interesting cases.
+
+The typing is preserved for different reasons in two of our systems, so we briefly review the 
+concrete examples. Then, we prove the substitution lemma for two special cases of flat coeffects
+(Lemma~\ref{thm:cbn-substitution-top} and Lemma~\ref{thm:cbn-substitution-bot}) and finally, we
+state the conditions under which typing preservation hold for flat coeffect calculi 
+(Theorem~\ref{thm:cbn-flat}).
+
+\paragraph{Data-flow.} 
+The type preservation property does not hold for data-flow. This case is similar to the example
+shown earlier with effectful computations. As a minimal example, consider the substitution of 
+a context-dependent expression $\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1,z\!:\!\tau_1}{0} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{0} \tau_2 & & (\textnormal{before}) \\
+ \coctx{z\!:\!\tau_1}{1} &\narrow{\vdash}& \lambda x.\kvd{prev}~z : \tau_1 \xrightarrow{1} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+After the substitution, the coeffect of the body is $1$. The rule for lambda abstraction requires
+that $1=\mathit{min}(r,s)$ and so the least solution is to set both $r,s$ to $1$. The substitution
+this affects the coeffects attached both to the function type and the overall context.
+
+Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
+actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
+system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
+
+\paragraph{Implicit parameters.} In data-flow, there is no typing for the resulting expression that
+preserves the type of the function. However, this is not the case for all systems. Consider substituting
+an implicit parameter access $\ident{?p}$ for a free variable $y$ under a lambda:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1}{\emptyset} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{before}) \\
+ \coctx{\emptyset}{ \{\ident{?p}\} } &\narrow{\vdash}& \lambda x.\ident{?p} : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+The above shows one possible typing of the body -- one that does not change the coeffects of the 
+function type and attaches all additional coeffects (implicit parameters) to the context. In case
+of implicit parameters (and, more generally, systems with set-like annotations) this is always 
+possible.
+
+\paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
+substituting an arbitrary expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
+%
+\begin{equation*}
+\begin{array}{rclcl}
+ \coctx{y\!:\!\tau_1}{\ident{L}} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{before}) \\
+ \coctx{\emptyset}{\ident{L}} &\narrow{\vdash}& \lambda x.e : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{after}) \\
+\end{array}
+\end{equation*}
+%
+In the original expression, both the overall context and the function type are annotated with \ident{L}, 
+because the body contains a variable access. An expression $e$ can always be treated as being annotated
+with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so we can also treat $e$
+as being annotated with coeffects $\ident{L}$. As a result, substitution does not change the coeffect.
+
+\paragraph{Reduction theorem.}
+The above examples (implicit parameters and liveness) demonstrate two particular kinds of coeffect 
+algebra for which typing preservation holds. Proving the type preservation separately provides 
+more insight into how the systems work. We consider the two cases separately, but find a more general
+formulation for both of them.
+
+\begin{definition}
+We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
+($\forall r \in \C \;.\; r \,\cleq\, \cunit $) and \emph{bottom-pointed} if it is the smallest (bottom) 
+element ($\forall r \in \C \;.\; r \,\cgeq\, \cunit $).
+\end{definition}
+
+\noindent
+Liveness is an example of top-pointed coeffects as variables are annotated with 
+$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and data-flow are examples
+of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
+additional requirements:
+
+\begin{lemma}[Top-pointed substitution]
+\label{thm:cbn-substitution-top}
+In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
+not change the coeffects of $e_r$:
+%
+\begin{equation*}
+\begin{array}{l}
+ \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
+   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
+ \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
+\end{array}
+\end{equation*}
+\end{lemma}
+
+\begin{proof}
+Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
+\end{proof}
+%
+\noindent
+As variables are annotated with the top element $\cunit$, we can substitute the term $e_s$ 
+for any variable and use sub-coeffecting to get the original typing (because
+$\cclrd{s} \,\cleq\, \cunit$).
+
+In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
+requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
+then the context requirements arising from the substitution can be associated with the context
+$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
+does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
+our implicit parameters example (all three operators are a set union) and for other set-like 
+coeffects. It allows the following substitution lemma:
+
+\begin{lemma}[Bottom-pointed substitution]
+\label{thm:cbn-substitution-bot}
+In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
+where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
+$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
+%
+\begin{equation*}
+\begin{array}{l}
+\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
+\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
+\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
+\end{array}
+\end{equation*}
+
+\end{lemma}
+\begin{proof}
+By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
+the free-variable context. See Appendix~\ref{sec:appendix-flat-cbn}.
+\end{proof}
+
+\noindent
+The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
+requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
+structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
+coeffects associated with individual variables, thus preserving typing in a complementary way.
+
+The two substitution lemmas discussed above show that the call-by-name evaluation strategy can be 
+used for certain coeffect calculi, including liveness and implicit parameters. Assuming
+$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following theorem holds:
+
+\begin{theorem}[Type preservation for call-by-name]
+\label{thm:cbn-flat}
+In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
+Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
+and $e \rightarrow_{\ident{cbn}} e'$ then it is also the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
+\end{theorem}
+\begin{proof}
+
+For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
+to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
+and $r \,\czip\, t \;=\; r \,\cpar\, t$.
+For bottom-pointed coeffect algebra, consider the typing derivation for the term
+$(\lambda x.e_r)\,e_s$ before reduction:
+%
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r}
+      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{r}} \tau_r } &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
+  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash (\lambda x.e_r)\,e_s : \tau_r }
+\end{equation*}
+%
+The derivation uses the idempotence of $\czip$ in the first step, followed by the 
+(\emph{app}) rule. The type of the term after substitution, using Lemma~\ref{thm:cbn-substitution-bot} is:
+%
+\begin{equation*}
+\inference
+  { \coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r & 
+    \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
+  { \coctx{\Gamma, x:\tau_r}{\cclrd{r}\;\cseq\;\cclrd{s}} \vdash \subst{e_r}{x}{e_s} : \tau_s }
+\end{equation*}
+%
+From the assumptions of Lemma~\ref{thm:cbn-substitution-bot}, we know that $\cseq\,=\,\cpar$
+and the operation is idempotent, so trivially:
+$\cclrd{r}\;\cseq\;\cclrd{s} = \cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})$
+\end{proof}
+
+\paragraph{Expansion theorem.}
+The $\eta$-expansion (local completeness) is similar to $\beta$-reduction (local soundness) in that
+it holds for some flat coeffect systems, but not for all. Out of the examples we discuss, it 
+holds for implicit parameters, but does not hold for liveness and data-flow.
+
+Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
+expression $e$ may require no variables (both immediate and latent coeffects are marked as
+$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
+marking the context and function argument as live. In case of data-flow, the immediate coeffects 
+are made larger by the lambda abstraction -- the context requirements of the function value are 
+imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
+the next chapter.
+
+However, $\eta$-expansion preserves the type for implicit parameters and, more generally,
+for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the 
+standard $\eta$-reduction:
+
+\begin{theorem}[Type preservation of $\eta$-expansion]
+In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
+where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
+and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
+\end{theorem}
+\begin{proof}
+The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
+\begin{equation*}
+\inference
+  { \inference
+    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
+      \coctx{x\!:\!\tau_1}{\cunit} \vdash x\!:\!\tau_1 }
+    { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
+  { \inference
+    { \inference
+      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
+      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
+    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
+\end{equation*}
+%
+The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
+application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
+original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
+is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
+this is sufficient for showing type preservation.
+\end{proof}
+
+\noindent
+In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
+there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
+these include liveness and implicit parameters. Moreover, for implicit parameters (and more
+generally, any set-like flat coeffect algebra), the $\eta$-expansion holds as well, giving us
+both local soundness and local completeness as coined by Pfenning and Davies \cite{logic-modal-reconstruction}.
 
@@
 
-\paragraph{Ordering.}
-Add $(\C, \cleq)$ is a pre-order that is:
-
-\begin{equation*}
-\begin{array}{ccc}
-\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
-t\; \cleq\; t \\[0.5em]
-\end{array}
-\end{equation*}
-
-The flat coeffect algebra requires a pre-order relation $\cleq$, which is used to define 
-sub-coeffecting rule of the type system. When the idempotent monoid $(\C, \cpar, \czero)$ also 
-has the commutative property (\ie~forms a semi-lattice), the $\cleq$ relation can be defined as the
-ordering of the semi-lattice:
-%
-\begin{equation*}
-\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
-\end{equation*}
-%
-This definition is consistent with all three examples that motivate flat coeffect calculus, but
-it cannot be used with the structural coeffects (where it fails for the bounded reuse 
-calculus) and so we choose not to use it.
-
-Furthermore, the $\cunit$ coeffect is often the top or the bottom element of the semi-lattice. 
-As discussed in Section~\ref{sec:flat-syntax}, when this is the case, we are able to prove certain 
-syntactic properties of the calculus.
-
-\paragraph{Something else}
 The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
@@
 %     #####
-%    #     # #   # #    # #####   ##    ####  ##### #  ####
-%    #        # #  ##   #   #    #  #  #    #   #   # #    #
-%     #####    #   # #  #   #   #    # #        #   # #
-%          #   #   #  # #   #   ###### #        #   # #
-%    #     #   #   #   ##   #   #    # #    #   #   # #    #
-%     #####    #   #    #   #   #    #  ####    #   #  ####
-%
-% ==================================================================================================
-
-\section{Syntactic equational theory}
-\label{sec:flat-syntax}
-
-Each of the concrete coeffect calculi discussed in this chapter has a different notion of context,
-much like various effectful languages have different notions of effects (such as exceptions or
-mutable state). However, in all of the calculi, the context has a number of common properties that 
-are captured by the \emph{flat coeffect algebra}. This means that there are equational properties
-that hold for all of the coeffect systems. Further properties hold for systems where the context
-satisfies additional properties.
-
-In this section, we look at such shared syntactic properties. This accompanies the previous section, 
-which provided a \emph{semantic} justification for the axioms of coeffect algebra with a 
-\emph{syntactic} justification. Operationally, this section can also be viewed as providing a 
-pathway to an operational semantics for two of our systems (implicit parameters and liveness),
-which can be based on syntactic substitution. As we discuss later, the notion of context for
-data-flow is more complex.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Syntactic properties}
-\label{sec:flat-syntax-props}
-
-Before discussing the syntactic properties of general coeffect calculus formally, it should be 
-clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
-that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
-Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
-equations define four syntactic reductions on the terms:
-%
-\begin{equation*}
-\begin{array}{rllcl}
-(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
-(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
-(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
-e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
-\end{array}
-\end{equation*}
-%
-The rules capture syntactic reductions that can be performed in a general calculus, without any 
-knowledge of the specific notion of context. If the reductions preserve the type of the expression
-(type preservation), then operational semantics can be defined as a repeated application of the 
-rules, until a specified normal form (\ie~a value) is reached.
-
-The first two equations define well-known call-by-name and call-by-value reductions. The \kvd{glet} 
-reduction (inspired by Filinski \cite{monads-inaction}) models a system where program is reduced to 
-a sequence of primitive operations, sequentially composed using the special \kvd{glet} construct. 
-In the rest of the section, we briefly outline the interpretation of the four 
-rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name 
-(Section~\ref{sec:flat-syntax-cbn}) in more details.
-
-The focus of our work is on the general coeffect system and so we do not discuss the operational
-semantics of the specific notions of context. However, some work in that area has been done
-by Brunel et al. \cite{coeffects-quantitative}.
-
-\paragraph{Call-by-name.} 
-In call-by-name, the argument is syntactically substituted for all occurrences of a variable. It
-can be used as the basis for operational semantics of purely functional languages. However, using
-the rule in effectful languages breaks the \emph{type preservation} property. For example, consider
-a language with effect system where functions are annotated with sets of effects such as $\{ \ident{write} \}$.
-A function $\lambda x.y$ is effect-free:
-%
-\begin{equation*}
-y\!:\!\tau_1 \vdash \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 \;\&\; \emptyset
-\end{equation*}
-%
-Substituting an expression $e$ with effects $\{ \ident{write} \}$ for $y$ changes the type of 
-the function by adding latent effects (without changing the immediate effects):
-%
-\begin{equation*}
-\vdash \lambda x.e : \tau_1 \xrightarrow{ \{ \ident{write} \} } \tau_2 \;\&\; \emptyset
-\end{equation*}
-%
-Similarly to effect systems, substituting a context-dependent computation $e$ for a variable $y$ can
-add latent coeffects to the function type. However, this is not the case for \emph{all} flat coeffect
-calculi. For example, call-by-name reduction preserves types and coeffects for the implicit 
-parameters system. This means that certain coeffect systems support call-by-name evaluation strategy
-and could be embedded in purely functional language (such as Haskell).
-
-\paragraph{Call-by-value.}
-
-The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
-first reduced to a \emph{value} before performing the substitution. In effectful languages, 
-value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
-values are identifiers $x$ or functions $(\lambda x.e)$.
-
-The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
-A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
-immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
-sense \emph{and} has not immediate context requirements. We define this formally in 
-Section~\ref{sec:flat-syntax-cbv}.
-
-The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi, 
-including all our three examples. However, it is rather weak -- in order to use it, the concrete 
-semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
-
-\paragraph{Local soundness and completeness.}
-Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
-are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which 
-introduce a function arrow (lambda abstraction) and eliminate it (application) are sufficiently 
-strong, but not too strong.
-
-The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
-already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
-coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Call-by-value evaluation}
-\label{sec:flat-syntax-cbv}
-
-As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
-calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
-term to a \emph{value} has to be provided for each system. Syntactic values are defined
-in the usual way:
-%
-\begin{equation*}
-\begin{array}{llll}
-v\in\mathit{SynVal}   & v &::=& x \sep c \sep (\lambda x.e)\\
-n\in\mathit{NonVal}   & n &::=& e_1\;e_2 \sep \kvd{let}~x=e_1~\kvd{in}~e_2\\
-e\in\mathit{Expr}     & e &::=& v \sep n 
-\end{array}
-\end{equation*}
-%
-The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in 
-coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
-not automatically defer all context requirements.
-
-\begin{definition} 
-An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if it is a syntactic value,
-\ie~$e \in \mathit{SynVal}$ and it has no context-dependencies, \ie~$\coctx{\Gamma}{ \cunit } \vdash e : \tau$.
-\end{definition}
-
-\noindent
-The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
-variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
-context requirements of the term:
-
-\begin{lemma}[Call-by-value substitution]
-\label{thm:flat-subst-cbv}
-In a flat coeffect calculus with a coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-given a value $\coctx{\Gamma}{\cunit} \vdash v : \sigma$ and an expression
-$\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
-change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
-\end{lemma}
-\begin{proof}
-By induction over the type derivation, using the fact that $x$ and $v$ are annotated
-with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
-\end{proof}
-
-\noindent
-The substitution lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems. However, 
-proving that call-by-value reduction preserves typing requires an additional constraint on the
-flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
-by the (\emph{approximation}) property:
-
-\begin{equation*}
-\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}
-\hspace{10em}(\textit{approximation})
-\end{equation*}
-%
-Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
-under-approximates the actual context capabilities while the $\cpar$ operation (combining of
-context requirements) over-approximates the actual context requirements.
-
-The property holds for the three systems we consider -- for implicit parameters, this is 
-an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
-is smaller than the least upper bound.
-
-Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction that reduces the term 
-$(\lambda x.e)\,v$ to a term $\subst{e}{x}{v}$, the type preservation theorem is
-stated as follows:
-
-\begin{theorem}[Type preservation for call-by-value]
-\label{thm:cbv-reduction}
-In a flat coeffect system with the (\emph{approxi\-mation}) property, 
-if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-Consider the typing derivation for the term $(\lambda x.e)\,v$ before reduction:
-\begin{equation*}
-\inference
-  { \inference
-  { \inference
-      {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash e : \tau_2}
-      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 } &
-   \coctx{\Gamma}{\cunit} \vdash v : \tau_1 }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{t})} \vdash (\lambda x.e)\,v : \tau_2} }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; \cclrd{t}} \vdash (\lambda x.e)\,v : \tau_2}
-\end{equation*}
-The final step simplifies the coeffect annotation using the fact that $\cunit$ is a unit of $\cseq$.
-From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
-As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
-%
-\begin{equation*}
-\tyrule{sub}
- { \coctx{\Gamma}{\cclrd{r}\;\czip\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 }
- { \coctx{\Gamma}{\cclrd{r}\;\cpar\;\cclrd{t}} \vdash \subst{e}{x}{v} : \tau_2 } 
-\end{equation*}
-%
-Comparing the final conclusions of the above two typing derivations shows that 
-the reduction preserves type and coeffect annotation.
-\end{proof}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Call-by-name evaluation}
-\label{sec:flat-syntax-cbn}
-
-When reducing the expression $(\lambda x.e_1)~e_2$ using the call-by-name strategy, the 
-sub-expression $e_2$ is substituted for all occurrences of the variable $v$ in an expression $e_1$. 
-As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in 
-general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two 
-interesting cases.
-
-The typing is preserved for different reasons in two of our systems, so we briefly review the 
-concrete examples. Then, we prove the substitution lemma for two special cases of flat coeffects
-(Lemma~\ref{thm:cbn-substitution-top} and Lemma~\ref{thm:cbn-substitution-bot}) and finally, we
-state the conditions under which typing preservation hold for flat coeffect calculi 
-(Theorem~\ref{thm:cbn-flat}).
-
-\paragraph{Data-flow.} 
-The type preservation property does not hold for data-flow. This case is similar to the example
-shown earlier with effectful computations. As a minimal example, consider the substitution of 
-a context-dependent expression $\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
-%
-\begin{equation*}
-\begin{array}{rclcl}
- \coctx{y\!:\!\tau_1,z\!:\!\tau_1}{0} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{0} \tau_2 & & (\textnormal{before}) \\
- \coctx{z\!:\!\tau_1}{1} &\narrow{\vdash}& \lambda x.\kvd{prev}~z : \tau_1 \xrightarrow{1} \tau_2 & & (\textnormal{after}) \\
-\end{array}
-\end{equation*}
-%
-After the substitution, the coeffect of the body is $1$. The rule for lambda abstraction requires
-that $1=\mathit{min}(r,s)$ and so the least solution is to set both $r,s$ to $1$. The substitution
-this affects the coeffects attached both to the function type and the overall context.
-
-Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
-actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
-system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
-
-\paragraph{Implicit parameters.} In data-flow, there is no typing for the resulting expression that
-preserves the type of the function. However, this is not the case for all systems. Consider substituting
-an implicit parameter access $\ident{?p}$ for a free variable $y$ under a lambda:
-%
-\begin{equation*}
-\begin{array}{rclcl}
- \coctx{y\!:\!\tau_1}{\emptyset} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{before}) \\
- \coctx{\emptyset}{ \{\ident{?p}\} } &\narrow{\vdash}& \lambda x.\ident{?p} : \tau_1 \xrightarrow{\emptyset} \tau_2 & & (\textnormal{after}) \\
-\end{array}
-\end{equation*}
-%
-The above shows one possible typing of the body -- one that does not change the coeffects of the 
-function type and attaches all additional coeffects (implicit parameters) to the context. In case
-of implicit parameters (and, more generally, systems with set-like annotations) this is always 
-possible.
-
-\paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
-substituting an arbitrary expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
-%
-\begin{equation*}
-\begin{array}{rclcl}
- \coctx{y\!:\!\tau_1}{\ident{L}} &\narrow{\vdash}& \lambda x.y : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{before}) \\
- \coctx{\emptyset}{\ident{L}} &\narrow{\vdash}& \lambda x.e : \tau_1 \xrightarrow{\ident{L}} \tau_2 & & (\textnormal{after}) \\
-\end{array}
-\end{equation*}
-%
-In the original expression, both the overall context and the function type are annotated with \ident{L}, 
-because the body contains a variable access. An expression $e$ can always be treated as being annotated
-with $\ident{L}$ (because $\ident{L}$ is the top element of the lattice) and so we can also treat $e$
-as being annotated with coeffects $\ident{L}$. As a result, substitution does not change the coeffect.
-
-\paragraph{Reduction theorem.}
-The above examples (implicit parameters and liveness) demonstrate two particular kinds of coeffect 
-algebra for which typing preservation holds. Proving the type preservation separately provides 
-more insight into how the systems work. We consider the two cases separately, but find a more general
-formulation for both of them.
-
-\begin{definition}
-We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
-($\forall r \in \C \;.\; r \,\cleq\, \cunit $) and \emph{bottom-pointed} if it is the smallest (bottom) 
-element ($\forall r \in \C \;.\; r \,\cgeq\, \cunit $).
-\end{definition}
-
-\noindent
-Liveness is an example of top-pointed coeffects as variables are annotated with 
-$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and data-flow are examples
-of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
-additional requirements:
-
-\begin{lemma}[Top-pointed substitution]
-\label{thm:cbn-substitution-top}
-In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
-not change the coeffects of $e_r$:
-%
-\begin{equation*}
-\begin{array}{l}
- \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
-   \coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
- \quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau_r
-\end{array}
-\end{equation*}
-\end{lemma}
-
-\begin{proof}
-Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
-\end{proof}
-%
-\noindent
-As variables are annotated with the top element $\cunit$, we can substitute the term $e_s$ 
-for any variable and use sub-coeffecting to get the original typing (because
-$\cclrd{s} \,\cleq\, \cunit$).
-
-In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
-requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
-then the context requirements arising from the substitution can be associated with the context
-$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
-does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
-our implicit parameters example (all three operators are a set union) and for other set-like 
-coeffects. It allows the following substitution lemma:
-
-\begin{lemma}[Bottom-pointed substitution]
-\label{thm:cbn-substitution-bot}
-In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
-where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
-$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
-%
-\begin{equation*}
-\begin{array}{l}
-\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\; 
-\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
-\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
-\end{array}
-\end{equation*}
-
-\end{lemma}
-\begin{proof}
-By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
-the free-variable context. See Appendix~\ref{sec:appendix-flat-cbn}.
-\end{proof}
-
-\noindent
-The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
-requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
-structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
-coeffects associated with individual variables, thus preserving typing in a complementary way.
-
-The two substitution lemmas discussed above show that the call-by-name evaluation strategy can be 
-used for certain coeffect calculi, including liveness and implicit parameters. Assuming
-$\reduce_{\ident{cbn}}$ is the standard call-by-name reduction, the following theorem holds:
-
-\begin{theorem}[Type preservation for call-by-name]
-\label{thm:cbn-flat}
-In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
-Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-and $e \rightarrow_{\ident{cbn}} e'$ then it is also the case that $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
-\end{theorem}
-\begin{proof}
-
-For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
-to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
-and $r \,\czip\, t \;=\; r \,\cpar\, t$.
-For bottom-pointed coeffect algebra, consider the typing derivation for the term
-$(\lambda x.e_r)\,e_s$ before reduction:
-%
-\begin{equation*}
-\inference
-  { \inference
-      {\coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r}
-      {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{r}} \tau_r } &
-   \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
-  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash (\lambda x.e_r)\,e_s : \tau_r }
-\end{equation*}
-%
-The derivation uses the idempotence of $\czip$ in the first step, followed by the 
-(\emph{app}) rule. The type of the term after substitution, using Lemma~\ref{thm:cbn-substitution-bot} is:
-%
-\begin{equation*}
-\inference
-  { \coctx{\Gamma, x:\tau_s}{\cclrd{r}} \vdash e_r : \tau_r & 
-    \coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s }
-  { \coctx{\Gamma, x:\tau_r}{\cclrd{r}\;\cseq\;\cclrd{s}} \vdash \subst{e_r}{x}{e_s} : \tau_s }
-\end{equation*}
-%
-From the assumptions of Lemma~\ref{thm:cbn-substitution-bot}, we know that $\cseq\,=\,\cpar$
-and the operation is idempotent, so trivially:
-$\cclrd{r}\;\cseq\;\cclrd{s} = \cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})$
-\end{proof}
-
-\paragraph{Expansion theorem.}
-The $\eta$-expansion (local completeness) is similar to $\beta$-reduction (local soundness) in that
-it holds for some flat coeffect systems, but not for all. Out of the examples we discuss, it 
-holds for implicit parameters, but does not hold for liveness and data-flow.
-
-Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
-expression $e$ may require no variables (both immediate and latent coeffects are marked as
-$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
-marking the context and function argument as live. In case of data-flow, the immediate coeffects 
-are made larger by the lambda abstraction -- the context requirements of the function value are 
-imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
-the next chapter.
-
-However, $\eta$-expansion preserves the type for implicit parameters and, more generally,
-for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the 
-standard $\eta$-reduction:
-
-\begin{theorem}[Type preservation of $\eta$-expansion]
-In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ 
-where $\czip = \cpar$, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ 
-and $e \rightarrow_\eta e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 $.
-\end{theorem}
-\begin{proof}
-The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
-\begin{equation*}
-\inference
-  { \inference
-    { \coctx{\Gamma}{\cclrd{r}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
-      \coctx{x\!:\!\tau_1}{\cunit} \vdash x\!:\!\tau_1 }
-    { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; (\cunit \,\cseq\, \cclrd{s}) } \vdash f~x : \tau_2 } }
-  { \inference
-    { \inference
-      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\cpar\; \cclrd{s} } \vdash f~x : \tau_2 }
-      { \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \;\czip\; \cclrd{s} } \vdash f~x : \tau_2 } }
-    { \coctx{\Gamma}{ \cclrd{r} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
-\end{equation*}
-%
-The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
-application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
-original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
-is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
-this is sufficient for showing type preservation.
-\end{proof}
-
-\noindent
-In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
-there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
-these include liveness and implicit parameters. Moreover, for implicit parameters (and more
-generally, any set-like flat coeffect algebra), the $\eta$-expansion holds as well, giving us
-both local soundness and local completeness as coined by Pfenning and Davies \cite{logic-modal-reconstruction}.
-
-% ==================================================================================================
-%                                                                             
-%        ###                         ##                    #                        
-%       #   #                         #                                             
-%       #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###  
+%    #     #  ####  #    #  ####  #      #    #  ####  #  ####  #    #  ####
 %    #       #    # ##   # #    # #      #    # #      # #    # ##   # #
-%       #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###  
-%       #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      # 
-%        ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####  
+%    #       #    # # #  # #      #      #    #  ####  # #    # # #  #  ####
+%    #       #    # #  # # #      #      #    #      # # #    # #  # #      #
+%    #     # #    # #   ## #    # #      #    # #    # # #    # #   ## #    #
+%     #####   ####  #    #  ####  ######  ####   ####  #  ####  #    #  ####
 %                                                                             

### 2016-02-05 text/Chapter04.tex
Delete rest of glet section

@@
 (\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
-(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
 e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
@@
 
-The first two equations define well-known call-by-name and call-by-value reductions. The \kvd{glet} 
-reduction (inspired by Filinski \cite{monads-inaction}) models a system where program is reduced to 
-a sequence of primitive operations, sequentially composed using the special \kvd{glet} construct. 
-In the rest of the section, we briefly outline the interpretation of the four 
+In the rest of the section, we briefly outline the interpretation of the three
 rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name 
@@
 
-The focus of our work is on the general coeffect system and so we do not discuss the operational
+The focus of this chapter is on the general coeffect system and so we do not discuss the operational
 semantics of the specific notions of context. However, some work in that area has been done
-by Brunel et al. \cite{coeffects-quantitative}.
+by Brunel et al. \cite{coeffects-quantitative}. We discuss concrete semantics of implicit 
+parameters and dataflow in Chapter~\ref{ch:transl}.
 
@@
 by the (\emph{approximation}) property:
-
+%
 \begin{equation*}

### 2016-02-05 text/Chapter04.tex
Finish edits in chapter 4, move stuff to chapter 5

@@
 \subsection{Dataflow and liveness}
+\label{sec:flat-unique-ldf}
 
@@
 
-In this section, we look at the last point. We discuss what syntactic equivalences are permitted
+In this section, we look at the last point. We consider sub-coeffecting and subtyping 
+(Section~\ref{sec:flat-exts-sub}), discuss what syntactic equivalences are permitted
 by the properties of $\czip$ (Section~\ref{sec:flat-exts-lambda}) and we extend the calculus with 
-pairs and units and discuss their syntactic properties (Section~\ref{sec:flat-exts-tup}). Before
-doing that, the following section further develops subtyping relation for the calculus.
+pairs and units and discuss their syntactic properties (Section~\ref{sec:flat-exts-tup}).
 
@@
 \begin{figure}[t]
-
-\begin{equation*}
-\boxed{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
-\end{equation*}
-
+\begin{equation*}
+\tyrule{sub-trans}
+  { \tau_1 <: \tau_2 & \tau_2 <: \tau_3 }
+  { \tau_1 <: \tau_3  }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub-fun}
+  { \tau_1' <: \tau_1 & \tau_2 <: \tau_2' & \cclrd{r'} \cgeq \cclrd{r} }
+  { \tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2' }
+\end{equation*}
+\begin{equation*}
+\tyrule{sub-refl}
+  { }
+  { \tau <: \tau }
+\end{equation*}
+
+\figcaption{Subtyping rules for flat coeffect calculus}
+\label{fig:flat-types-sub}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Subcoeffecting and subtyping}
+\label{sec:flat-exts-sub}
+
+The \emph{flat coeffect algebra} includes the $\cleq$ relation which captures the ordering of
+coeffects and can be used to define sub-coeffecting. Syntactically, an expression with context
+requirements $\cclrd{r}$ can be treated as an expression with greater context requirements:
+%
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
+\end{equation*}
+%
+Semantically, this means that we can \emph{drop} some of the provided context. For example, 
+if an expression requires implicit parameters $\{ \ident{?p} \}$ it can be treated as requiring
+$\{ \ident{?p}, \ident{?q} \}$. The semantic function will then be provided with a dictionary
+containing both assignments and it can drop the value for the unused parameter \ident{?q}.
+
+Sub-coeffecting only affects the immediate coeffects attached to the free-variable context.
+In Figure~\ref{fig:flat-types-sub}, we add sub-typing on function types, making it possible to treat
+a function with smaller context requirements as a function with greater context requirements:
+%
 \begin{equation*}
@@
   { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' }
-\quad
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \cleq \cclrd{r})
-\end{equation*}
-
-\begin{equation*}
-\boxed{\tau <: \tau'}
-\end{equation*}
-
-\begin{equation*}
-\tyrule{sub-trans}
-  { \tau_1 <: \tau_2 & \tau_2 <: \tau_3 }
-  { \tau_1 <: \tau_3  }
-\end{equation*}
-\begin{equation*}
-\tyrule{sub-fun}
-  { \tau_1' <: \tau_1 & \tau_2 <: \tau_2' & \cclrd{r'} \cgeq \cclrd{r} }
-  { \tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2' }
-\end{equation*}
-\begin{equation*}
-\tyrule{sub-refl}
-  { }
-  { \tau <: \tau }
-\end{equation*}
-
-\figcaption{Subtyping rules for flat coeffect calculus}
-\label{fig:flat-types-sub}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-~
-
-~
-
-\subsection{Subcoeffecting and subtyping}
-\label{sec:flat-exts-sub}
-
-The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
-which makes it possible to treat an expression with smaller context requirements as an expression
-with greater context requirements. In the corresponding categorical semantics, this means that
-we can \emph{drop} some of the provided context. However, sub-coeffecting only affects the 
-immediate coeffects attached to the free-variable context.
-
-In Figure~\ref{fig:flat-types-sub}, we add sub-typing on function types, making it possible to treat
-a function with smaller context requirements as a function with greater context requirements. 
+\end{equation*}
+%
 The definition uses the standard reflexive and transitive $<:$ operator. As the (\emph{sub-fun})
@@
 (\emph{typ}) rule allows using sub-typing on expressions in the coeffect calculus.
-
-\paragraph{Semantics.} We follow the same approach as in Section~\ref{sec:flat-semantics} and use
-categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
-of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
-As shown in Figure~\ref{fig:flat-semantics-sub}, the semantics of the sub-typing rule (\emph{typ}) 
-then just composes the semantics of the original expression with the conversion produced by the 
-semantics of the sub-typing judgement.
-
-The rest of the Figure~\ref{fig:flat-semantics-sub} defines the semantics of the $<:$ operation. 
-The reflexivity and transitivity are just the identity function and function composition,
-respectively. The (\emph{sub-fun}) case is interesting -- recall that the semantics of functions:
-%
-\begin{equation*}
-\sem{\tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} ~=~ C^{\cclrd{r'}} \tau_1' \rightarrow \tau_2'
-\end{equation*}
-%
-We build the transformation using an explicit lambda abstraction that takes the original function
-$f$ as an argument. To build the required function, we first drop unnecessary context using 
-$\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ of type $C^{\cclrd{r'}} \tau_1' \rightarrow C^{\cclrd{r}} \tau_1'$,
-then we use the $\ident{map}_{\cclrd{r}}$ function to transform the nested $\tau_1'$ to $\tau_1$. 
-Finally we evaluate the original function $f$ and turn the resulting value of type $\tau_2$ into 
-the required result of type $\tau_2'$.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-
-\begin{equation*}
-\begin{array}{lll}
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' } = 
-  \sem{\tau <: \tau'} \circ \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } &\qquad& (\emph{typ})
-\\[1.75em]  
-\sem{\tau <: \tau} = \ident{id} && (\emph{sub-refl})
-\\[0.5em]
-\sem{\tau_1 <: \tau_3} = \sem{\tau_2 <: \tau_3} \circ \sem{\tau_1 <: \tau_2} && (\emph{sub-trans})
-\\[0.5em]
-\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} = \lambda f. && (\emph{sub-fun}) \\
-\qquad\sem{\tau_2 <: \tau_2'} \circ f \circ \ident{map}_{\cclrd{r}}~\sem{\tau_1' <: \tau_1} 
-  \circ \ident{lift}_{\cclrd{r'}, \cclrd{r}}
-\end{array}
-\end{equation*}
-
-\figcaption{Semantics of subtyping for flat coeffects}
-\label{fig:flat-semantics-sub}
-\end{figure}
 
@@
 
-\begin{table}[!h]
+\begin{table}
 \begin{center}
@@
 
-\noindent
 The simplified annotations directly follow from the definitions of particular flat coeffect 
@@
 
-In this section, we first identify coeffect calculi for which the simpler (duplicating) 
-rule is sufficient. Then we look at syntactic transformations corresponding to other common 
-properties of the $\czip$ operation.
+In this section, we first identify coeffect calculi for which the simpler (\emph{idabs}) rule
+introduced in Section~\ref{ec:flat-unique-ldf} is sufficient. Then we look at syntactic 
+transformations corresponding to other common properties of $\czip$.
 
@@
 idempotence means that the context requirements of the body can be required from both the 
-declaration site and the call site. Thus, the following (\emph{idabs}) typing is valid 
-(for reference, it is shown side-by-side with the ordinary lambda abstraction rule):
-%
+declaration site and the call site. Thus, the (\emph{idabs}) typing is valid. For reference,
+we repeat the earlier definitions here:
+
 \begin{equation*}
@@
 \label{thm:flat-alt-abs}
-For a flat coeffect calculus with a strictly oriented algebra, the standard (abs) rule can 
-be derived from the (idfun) rule.
+For a flat coeffect calculus with a strictly oriented algebra, equipped with sub-coeffecting and
+subtyping, the standard (abs) rule can be derived from the (idabs) rule.
 \end{remark}
@@
 removes the non-determinism of type checking that arises from the splitting of context 
-requirements in the original (\emph{abs}) rule. Furthermore, for data-flow and liveness
-the (\emph{idabs}) rule is more precise than (\emph{abs}).
+requirements in the original (\emph{abs}) rule.
 
@@
 language and to motivate the laws required about $\cpar$. The syntax of the language is extended as follows:
-%
+
+\noindent
 \begin{equation*}
@@
 
-\begin{theorem}
+\begin{remark}
 \label{thm:flat-tup-eq}
@@
 \end{equation*}
-\end{theorem}
+\end{remark}
 \begin{proof}
@@
 \noindent
-The Theorem~\ref{thm:flat-tup-eq} motivates the requirement of the monoid structure 
+The Remark~\ref{thm:flat-tup-eq} motivates the requirement of the monoid structure 
 $(\C, \cpar, \czero)$ of the flat coeffect algebra. We require only unit and associativity
@@
 execution environment or the entire context in which programs are executed.
-This is the first of the three \emph{coeffect calculi} developed in this thesis.
+This is the first of the two \emph{coeffect calculi} developed in this thesis.
 
@@
 capture three specific systems, namely liveness, data-flow and implicit parameters. However,
-the system is more general and an capture numerous other applications outlined in Section~\ref{sec:applications-flat}.
-
-Next, we introduced the notion of \emph{flat indexed comonad}, which generalizes of comonads
-and adds additional operations needed to provide categorical semantics of the flat coeffect 
-calculus. The indices of the flat indexed comonad operations correspond to the
-coeffect annotations in the type system and provide a foundation for the design of the 
-calculus.
-
-Finally, we discussed the equational theory for flat coeffect calculus. Although each 
+the system is more general and an capture numerous other applications outlined in 
+Section~\ref{sec:applications-flat}.
+
+An inherent property of flat coeffect systems is the ambiguity of the typing for lambda 
+abstraction. The body of a function requires certain context, but the context can be often
+provided by either the declaration-site or the call-site. Resolving this ambiguity has to be
+done differently for each concrete coeffect system, depending on its specific notion of context.
+We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec}.
+
+Finally, we introduced the equational theory for flat coeffect calculus. Although each 
 concrete instance of flat coeffect calculus models different notion of context, there are
@@
 
-In the upcoming chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
-properties to \emph{structural} coeffect calculi, tracking per-variable information, thus
-covering systems from the second half of Chapter~\ref{ch:applications}.
+In the next section, we move from abstract treatment of the flat coeffect calculus to a more
+concrete discussion. We explain its category-theoretical motivation, we use it to define 
+translational semantics (akin to Haskell's do notation) and we prove that well-typed programs
+in flat coeffect calcului for implicit parameters and dataflow do not get stuck.

### 2016-02-05 text/Chapter05.tex
Finish edits in chapter 4, move stuff to chapter 5

@@
 
-~
+\section{SUmmary}
+
+Next, we introduced the notion of \emph{flat indexed comonad}, which generalizes of comonads
+and adds additional operations needed to provide categorical semantics of the flat coeffect 
+calculus. The indices of the flat indexed comonad operations correspond to the
+coeffect annotations in the type system and provide a foundation for the design of the 
+calculus.

### 2016-02-08 text/Chapter04.tex
Chapter 5 intro and semantics

@@
 The development in this chapter can be seen as a counterpart to the well-known development of 
-\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:transl} then links \emph{coeffects} 
+\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:semantics} then links \emph{coeffects} 
 with \emph{comonads} in the same way in which effect systems can be linked with monads
@@
 \emph{context capabilities} or just \emph{available context} when using the semantic view, 
-especially in Chapter~\ref{ch:transl}.
+especially in Chapter~\ref{ch:semantics}.
 
@@
 Furthermore, the semantics of context-aware languages in Chapter~\ref{sec:applications} and
-also in Chapter~\ref{sec:transl} is defined over \emph{typing derivation} and so the same program
+also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivation} and so the same program
 could have a different meaning, depending on the typing derivation chosen. In this section, we
@@
 derivation. We use this unique typing derivation to give the semantics of coeffect language
-with implicit parameters in Chapter~\ref{ch:transl} and we also implement this algorithm as
+with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm as
 discussed in Chatper~\ref{ch:impl}.
@@
 (\emph{idabs}), then the type system gives a unique derivation. As for implicit parameters, this
-defines the semantics of coeffect program (Chapter~\ref{ch:transl}) and it is used in the
+defines the semantics of coeffect program (Chapter~\ref{ch:semantics}) and it is used in the
 implementation (Chatper~\ref{ch:impl}).
@@
 by Brunel et al. \cite{coeffects-quantitative}. We discuss concrete semantics of implicit 
-parameters and dataflow in Chapter~\ref{ch:transl}.
+parameters and dataflow in Chapter~\ref{ch:semantics}.
 

### 2016-02-08 text/Chapter05.tex
Chapter 5 intro and semantics

@@
 \chapter{Semantics of flat coeffects}
-\label{ch:transl}
+\label{ch:semantics}
 
 The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
-of context-aware systems discussed earlier in Chapter~X. The coeffect calculus can be seen
-as a \emph{language framework} that simplifies the construction of concrete \emph{domain-specific}
-coeffect lnguages. First, it provides a parameterized type system that tracks the required context.
-Second, the comonadic semantics provides a way for implementing the langauge by translating
-it into a non-context-aware functional programming language with additional comonadically-inspired
-coeffect-specific primitives that implement the concrete notion of context-awareness.
-
-In this chapter, we give three concrete examples of how the coeffect \emph{language framework}
-gives rise to concrete \emph{context-aware domain-specific} languages and we discuss the safety
-guarantees provided by the coeffect type system.
-
-For each concrete context-aware langauge, the safety depends on the coeffect-specific 
-comonadically-inspired primitives, but the coeffect framework makes it easier to prove such safety.
-We only need to define the meaning of the comonadically-inspired coeffect-specific primitives
-and then show that those do not ``go wrong''. In doing so, we can use the coeffect annotations
-provided by the type system.
+of context-aware systems discussed in Chapter~\ref{ch:applications}. The coeffect calculus 
+can be seen as a \emph{language framework} that simplifies the construction of concrete 
+\emph{domain-specific} coeffect lnguages. In the previous chapter, we discussed how it provides a  
+type system that tracks the required context. In this chapter, we show that the langauge framework 
+also provides a way for defining the semantics of concrete domain-specific coeffect languages,  
+guides their implementation and simplifies safety proofs.
+
+This is done using a \emph{comonadically-inspired translation}. We translate a program written 
+using the coeffect calculus into a simple functional language with additional coeffect-specific 
+comonadically-inspired primitives that implement the concrete notion of context-awareness.
+
+We use comonads in a syntactic way, following the example of Haskell's use of monads. The 
+translation is the same for all coeffect languages, but the safety depends on the concrete
+coeffect-specific comonadically-inspired primitives. We prove the soundness of two concrete
+coeffect calculi (dataflow and implicit parameters). We note that the proof crucially relies on a 
+relationship between coeffect annotations (provided by the type system) and the comonadically-inspired
+primitives (defining the semantics), which makes it easy to extend it to other concrete 
+context-aware langauges.
 
@@
 \begin{itemize}
-\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
-  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
-  This provides a deeper insight into how (and why) the calculus works and shows an 
-  intriguing link with effects.
-
-\item 
-  We define the common subset of the target functional programming language. This includes the
-  syntax of the language, reduction rules and typing rules, but it does \emph{not} include 
-  coeffect-specific definitions. Well-typed programs written using the common subset of the 
-  target language do not get stuck (via progress and preservation), but they may reduce to 
-  \ident{error}, e.g. when accessing the head of an empty list.
-
-\item We then extend the language with coeffect-specific comonadically-inspired
-  data types and primtiives for dataflow and for implicit parameters. We show that the extension
-  preserves the progress and preservation properties.
-
-\item Next, we consider only programs in the target language that were produced by a
-  translation from the coeffect domain-specific language and we show that such programs not only
-  do not get stuck, but they also do not reduce to the \ident{error} value. In other words,
-  ``well-typed coeffect programs do not get hungry'' requiring more context than guaranteed
-  by the coeffect type system.
-
-\item We show how the approach extends to structural coeffect systems and we argue that our
-  proof can be generalized -- rather than reconsidering progress and preservation of the 
-  whole language, we can rely just on the correctness of the coeffect-specific comonadically-inspired
-  primitives and abstraction mechanism provided by languages such as ML and Haskell.  
+\item We introduce \emph{indexed comonads}, a generalization of comonads, a category-theoretical 
+  dual of monads (Section~\ref{sec:semantics-theory}) and we discuss how they provide semantics
+  for coeffect calculus. This provides an insight into how (and why) the calculus works and shows 
+  an intriguing link with effects and monads.
+
+\item We use indexed comonads to guide our \emph{translational semantics} of coeffect calculus
+  (Section~\ref{sec:semantics-translation}). We define a simple functional programming language
+  (with type system and operational semantics) and prove that well-typed programs in the language
+  do not get stuck. We extend it with uninterpreted comonadically-inspired primitives and
+  define a translation that turns well-typed context-aware coeffect programs into programs of 
+  our functional language.
+
+\item For two of the coeffect calculi discussed earlier (dataflow and implicit parameters),
+  we give reduction rules for the comonadically-inspired primitives and we extend the progress
+  and preservation proofs, showing that well-typed programs produced by translation from two
+  coeffect languages do not get stuck (Section~\ref{sec:semantics-proofs})
+
+\item We leave the details to future work, but we note that the proof for concrete coeffect language 
+  (dataflow and implicit parameters) can be generalized -- rather than reconsidering progress and 
+  preservation of the whole target language, we rely just on the correctness of the coeffect-specific 
+  comonadically-inspired primitives and abstraction mechanism provided by languages 
+  such as ML and Haskell (Section~\ref{sec:semantics-related}).
 \end{itemize}
@@
 
-\section{Introduction}
-\label{sec:transl-intro}
+\section{Introduction and safety}
+\label{sec:semantics-intro}
+
+This chapter links together all the different technical developments presented in this thesis.
+We take the flat coeffect calculus introduced in Chatper~\ref{ch:flat}, define its \emph{abstract 
+comonadic semantics} and use it to define a translation that gives a \emph{concrete operational
+semantics} to a number of concrete context-aware languages. The type system is used to guarantee 
+that the resulting programs are correct. Finally, the development in this chapter is closely 
+mirrored by the implementation presented in Chapter~\ref{ch:impl}, which implements the 
+translation together with an interpreter for the target language.
+
+The key claim of this thesis is that writing context-aware programs using coeffects is easier and
+safer. In this chapter, we substantiate the claim by showing that programs written in the coeffect
+calculus and evaluated using the translation provided here do not ``go wrong''.
+
+To provide an intuition, consider two context-aware programs. The first calls a function that
+adds two implicit parameters in a context where one of them is defined. The second calculates
+the difference between the current and the previous value in a dataflow computation.
+For comparison, we show the code written in a coeffect dataflow language (on the left) and 
+using standard ML-like libraries (on the right):
+%
+\begin{equation*}
+\begin{array}{lcl}
+\begin{array}{l}
+\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
+ \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
+\kvd{let}~\ident{?one} = 10~\kvd{in}\\[-0.2em] 
+\ident{add}~0
+\end{array} 
+&\quad&
+\begin{array}{l}
+\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}~\ident{params}\rightarrow\\[-0.2em]
+ \quad \ident{lookup}~\str{one}~\ident{params}~+ \\[-0.2em]
+ \quad\quad \ident{lookup}~\str{two}~\ident{params}~\kvd{in}\\[-0.2em]
+\ident{add}~0~(\ident{add}~\str{one}~10~\ident{params})
+\end{array} 
+\\[3.5em]
+\begin{array}{l}
+\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
+ \quad \ident{x}-\kvd{prev}~\ident{x}
+\end{array} 
+&&
+\begin{array}{l}
+\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
+ \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
+\end{array} 
+\end{array} 
+\end{equation*}
+%
+The \ident{add} function (on the left) has a type $\ident{int}\xrightarrow{\cclrd{ \{ \ident{?one},\ident{?two} \} }}\ident{int}$.
+We call it in a context containing $\cclrd{\ident{?one}}$ and so the coeffect of the program is
+$\cclrd{ \{ \ident{?two} \} }$. The safety property for implicit parameters
+(Theorem~\ref{thm:semantics-sound-impl}) guarantees that, when executed in a context that
+provides a value for the implicit parameter $\cclrd{\ident{?one}}$, the program reduces to 
+a value of an correct type (or never terminates).
+
+If we wrote the code without coeffects (on the right), we could use a dynamic map to pass around
+a dictionary of parameters (the \ident{lookup} function obtains a value and \ident{add} adds a new
+assignment to the map). In that case, the type of \ident{add} is just $\ident{int}\rightarrow\ident{int}$
+and so the user does not know which implicit parameters it will need.
+
+Similarly, the \ident{diff} function can be implemented in terms of lists (on the right) as
+a function of type $\ident{num}~\ident{list}\rightarrow\ident{num}$. The function fails for input 
+lists containing only zero or one elements and this is not reflected in the type and it is not 
+enforced by the type checker.
+
+Using coeffects (on the left), the function has a type $\ident{num}\xrightarrow{\cclrd{1}}\ident{num}$ 
+meaning that it requires one past value (in addition to the current value). The safety property
+for dataflow (Theorem~\ref{thm:semantics-sound-df}) shows that, when called with a context that
+contains the required number of past values as captured by the coeffect type system, the function 
+does not get stuck.
+
+In summary, coeffect type system, captures certain runtime 
+requirements of context-aware programs and (as we show in this chapter), eliminates common errors 
+related to working with context.
+
+% ==================================================================================================
+%                
+%    #######
+%       #    #    # ######  ####  #####  #   #
+%       #    #    # #      #    # #    #  # #
+%       #    ###### #####  #    # #    #   #
+%       #    #    # #      #    # #####    #
+%       #    #    # #      #    # #   #    #
+%       #    #    # ######  ####  #    #   #                              
+%
+% ==================================================================================================
+
+\section{Categorical motivation}
+\label{sec:semantics-theory}
+
+The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
+discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical 
+semantics of context-dependent computations. This is a direction that we explore in this section. 
+Although the development presented here is interesting in its own, our main focus is \emph{using} 
+categorical semantics to motivate and explain the translation discussed in 
+Section~\ref{sec:semantics-translation}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Comonads are to coeffects what monads are to effects}
 
@@
 
-This chapter also links together all the different technical developments presented in this thesis.
-The abstract comonadic semantics is used as a \emph{translation} that gives a concrete semantics
-to a number of concrete context-aware languages. The type system is used to guarantee that the
-resulting programs are correct. Finally, the development in this chapter is closely mirrored by 
-the implementation presented in Chapter~X.
-
-\subsection{Safety properties}
-The key claim of this thesis is that writing context-aware programs using coeffects is easier and
-safer. For example, consider the simple problem of writing a dataflow function that calculates the 
-difference between the current and the previous value.
-
-The following shows the code written in a coeffect dataflow language (left) and using lists to
-represent past values in a ML-like language (right):
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
- \quad \ident{x}-\kvd{prev}~\ident{x}
-\end{array} 
-\qquad
-\begin{array}{l}
-\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
- \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
-\end{array} 
-\end{equation*}
-%
-The function on the right has a type $\ident{num}~\ident{list}\rightarrow\ident{num}$. The function
-fails for input lists containing only zero or one elements. However, this is not reflected in the
-type and it is not enforced by the type checker.
-
-The function on the left has a type $\ctyp{\cclrd{1}}\ident{num}\rightarrow\ident{num}$ meaning 
-that it requires one past value (in addition to the current value). The information about required
-context is now reflected in the type. In this chapter, we show that well-typed (and well-coeffected!) 
-programs do not ``get stuck''. That is, the coeffect annotation in the type system is a safe
-over-approximation of the actual contextual information that will be accessed when a program runs.
-
-\subsection{Related work}
-
-Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
-systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
-and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
-relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
-$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
+The correspondence between effect system and monads has been pointed out by Wadler and Thiemann 
+\cite{monads-effects-marriage} and further explored by Atkey \cite{monads-parameterised-notions}
+and Vazou and Leijen \cite{monads-effects-remarriage}). This line of work relates effectful 
+functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
+$\tau_1 \rightarrow \mtyp{\sigma} \tau_2$. In this chapter, we show a similar correspondence between 
 \emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
@@
 
-The main purpose of the comonadic semantics presented in this chapter is to provide a motivation 
-for the typing of the flat coeffect calculus. Our approach is inspired by the work of Uustalu and
-Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
-data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
-comonads} annotate the structure with information about the required context, \ie~$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
-This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.
-
-% ==================================================================================================
-%                                              
-%      #####  #                                 
-%        #    #                                 
-%        #    # ##    ###    ###   # ##   #   # 
-%        #    ##  #  #   #  #   #  ##  #  #   # 
-%        #    #   #  #####  #   #  #      #  ## 
-%        #    #   #  #      #   #  #       ## # 
-%        #    #   #   ###    ###   #          # 
-%                                         #   # 
-%                                          ###                                                                         
-% ==================================================================================================
-
-\section{Categorical motivation}
-\label{sec:flat-semantics}
-
-The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
-discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical 
-semantics of context-dependent computations. This is a direction that we explore in this section. 
-Although the development presented here is interesting in its own, our main focus is \emph{using} 
-categorical semantics to motivate and explain the design of flat coeffect calculus.
+Our approach is inspired by the work of Uustalu and Vene \cite{comonads-notions} who present the 
+semantics of contextual computations (mainly for dataflow) in terms of comonadic functions 
+$\ctyp{}\tau_1 \rightarrow \tau_2$. Our \emph{indexed comonads} annotate the structure with 
+information about the required context, \ie~$\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
+This is similar to the recent development on monads and effects by Katsumata \cite{monads-parametric}
+who parameterizes monads in a similar way to our parameterization of comonads.
 
@@
 In category theory, \emph{comonad} is a dual of \emph{monad}. As already outlined in
-Section~\ref{sec:path-sem}, we get a definition of a comonad by taking a definition of a monad 
+Chapter~\ref{ch:pathways}, we obtain a definition of a comonad by taking a definition of a monad 
 and ``reversing the arrows''. More formally, one of the equivalent definitions of comonad 
-looks as follows:
+looks as follows (repeated from Section~\ref{sec:path-sem}):
 
@@
 \subsection{Generalising to indexed comonads}
-\label{sec:flat-semantics-idx}
+\label{sec:semantics-flat-idx}
 
@@
 \subsection{Flat indexed comonads}
-\label{sec:flat-semantics-monoidal}
+\label{sec:semantics-flat-monoidal}
 
@@
 The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
-\emph{splitting} of contexts. In addition, we also need a lifting operation (similar to $\iota$ 
-from Definition~\ref{def:flat-family}) to model sub-coeffecting.
+\emph{splitting} of contexts.
 
@@
 a \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
-equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
-and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
+equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ where:
 %
@@
   $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
-\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
-  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
 \end{compactitem}
@@
 
-Finally, $\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ is a family of operations that ``forget'' some part of
-a context. This models the sub-coeffecting operation and lets us, for example, forget some of the
-available implicit parameters, or turn a live context (containing a value) into a dead context (empty).
-
-\paragraph{Alternative definition.}
+\paragraph{Semantics of sub-coeffecting.}
+Although we do not include sub-coeffecting in the core flat coeffect calculus, it is an interesting 
+extension to consider. Semantically, sub-coeffecting drops some of the available contextual 
+capabilities (drops some of the implicit parameters or some of the past values). This can be
+modelled by adding a (family of) lifting operation(s):
+%
+\begin{itemize}
+ \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
+   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
+\end{itemize}
+%
 Although we do not demand this as a general law, in all our systems, it is the case that
@@
 
-This alternative is valid for our examples, but we do not use it for two reasons. Firstly, this
-would be the only place where our semantics uses a variable \emph{twice} (in this case $c$). 
-Thus using an explicit \ident{split} means that the structure required by our semantics does not
-need to provide variable duplication and our model could be embedded in linear or affine category.
-Secondly, explicit \ident{split} is similar to the definition that is needed for structural 
-coeffects in Chapter~\ref{ch:structural} and so it makes the connection between the two system 
-easier to see.
+This alternative is valid for our examples, but we do not use it for three reasons. First, it
+requires making sub-coeffecting a part of the core definition. Second, this would be the only 
+place where our semantics uses a variable \emph{twice} (in this case $c$). Thus using an explicit 
+\ident{split} means that the structure required by our semantics does not need to provide variable 
+duplication and our model could be embedded in linear or affine category. Finally, explicit 
+\ident{split} is similar to the definition that is needed for structural coeffects in 
+Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.
 
 \paragraph{Examples.}
-All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:flat-semantics-idx} can
-be extended into \emph{flat indexed comonads}. Note that this is not a \emph{general} statement, 
+All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:semantics-flat-idx} can
+be extended into \emph{flat indexed comonads}. Note that this is not true \emph{in general}, 
 because each example requires us to define additional operations, specific for the example.
@@
 monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation 
-is defined by duplication and $\ident{lift}_{1,1}$ is the identity function.
+is defined by duplication.
 \end{example}
@@
 \end{array}
-\quad
+\qquad
 \begin{array}{rcl}
@@
 \ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
-\end{array}
-\quad
-\begin{array}{rcl}
-\ident{lift}_{\ident{L}, \ident{D}}~v &\narrow{=}& ()\\
-\ident{lift}_{\ident{L}, \ident{L}}~v &\narrow{=}& v\\
-\ident{lift}_{\ident{D}, \ident{D}}~() &\narrow{=}& ()\\
 \end{array}
@@
 
-Explicit sub-coeffecting using the (\emph{sub}) rule is modelled by the \ident{lift} operation.
-This can turn a \emph{live} value $v$ into a dead value $()$, or it can behave as identity.
-The behaviour is, again, determined by the index.
-
 \begin{example}[Indexed product]
@@
 \ident{merge}_{\cclrd{r}, \cclrd{s}}~((a, f), (b, g)) &\narrow{=}& ((a, b), f \uplus g)\\
-\ident{lift}_{\cclrd{r'}, \cclrd{r}}~(a, g) &\narrow{=}& (a, \restr{g}{\cclrd{r}})\\
 \end{array}
@@
 The \ident{split} operation splits the tuple and restricts the function (representing available
-implicit parameters) to the required sub-sets. This corresponds to the definition in terms
-of \ident{lift}, which performs just the restriction. The \ident{merge} operation is more 
+implicit parameters) to the required sub-sets. The \ident{merge} operation is more 
 interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
@@
 are defined as follows:
-
+%
 \begin{equation*}
@@
 \quad (\langle a_0, \ldots, a_m \rangle, \langle b_0, \ldots, b_n\rangle)
-\\[0.45em]
-\ident{lift}_{n', n} \langle a_0, \ldots, a_{n'} \rangle = \qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~n\leq n') \\[-0.25em]
-\quad \langle a_0, \ldots, a_n \rangle
 \end{array}
@@
 
+~
+
+~
+
+~
+
 % --------------------------------------------------------------------------------------------------
 
-\subsection{Properties and related notions}
-
-The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
-calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
-additional properties and other categorical structures that have been proposed mainly in the 
-context of monads and effects and we look how they relate to indexed comonads.
-
-\paragraph{Shape preservation.}
-Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
-this means that the core comonad structure does not provide a way of modeling computations where
-the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
-the length of the list stays the same after applying \ident{cobind}.
+\begin{figure*}[t]
+The semantics is defined over a typing derivation:
+%  
+\begin{equation*}
+\hspace{-3em}
+\begin{array}{ll}
+\hspace{4.8em}\semdef
+  {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
+  {\pi_i \circ \ident{counit}_{\cunit}} 
+& (\emph{var})
+\\[1em]
+\hspace{5.3em}\semdef
+  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
+  {\ident{const}~n} 
+& (\emph{const})
+\\[1.5em]
+\hspace{1.5em}\semdeff
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+  {f}
+  {f \circ \ident{curry}~\ident{merge}_{\cclrd{r}, \cclrd{s}} }
+& (\emph{abs})
+\\[2em]
+\semdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\begin{array}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\end{array}}
+  {f}
+  {g}
+  {\begin{array}{l}  
+  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g)~\circ~\ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}} \\[-0.3em]
+  ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
+  \end{array} }   
+& (\emph{app})
+\end{array}
+\end{equation*}
 
@@
 
-Indexed comonads are not restricted by this property of comonads. For example, given the indexed
-product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
-the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
-just implicit parameters $\cclrd{s}$.
-
-\paragraph{Families of monads.}
-When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
-\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
-comonads}, is defined as follows.
-
-\begin{definition}
-\label{def:flat-family}
-A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
-  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
-$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
-$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
-certain coherence conditions.
-\end{definition}
-
-\noindent
-A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data 
-types needs to form a comonad separately. For example, our indexed option does not form a family of 
-comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given 
-a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
-an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
-we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
-$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
-
-\paragraph{Parameteric effect monads.}
-Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
-indexed comonads) are closely related to our definition.  Although presented in a more general 
-categorical framework (and using monads), the model defines the \ident{unit} operation only on the 
-unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal 
-structure.
+Assuming the following auxiliary operations:
+%
+\begin{equation*}
+\begin{array}{rcl}
+  \ident{const}~v &=& \lambda x.v \\[-0.3em]
+  \ident{curry}~f~x~y &=& \lambda f.\lambda x.\lambda y.f~(x, y)\\[-0.3em]
+  \ident{dup}~x &=& (x, x)\\[-0.3em]
+  f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.3em]
+  \ident{app}~f~x &=& f~x
+\end{array}
+\end{equation*}
+
+\figcaption{Categorical semantics of the flat coeffect calculus}
+\label{fig:flat-semantics}
+\end{figure*}
 
@@
 \subsection{Semantics of flat calculus}
-\label{sec:flat-semantics-calculus}
+\label{sec:semantics-flat-calculus}
 
@@
 
+As discussed in Section~\ref{sec:flat-unique}, different typing derivations of coeffect programs 
+may have different meaning (\eg~when working with implicit parameters) and so the semantics is
+defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a 
+semantics to a term, we choose its unique coeffect-specific typing derivation as defined in 
+Section~\ref{sec:flat-unique}.
+
 \paragraph{Contexts and functions.}
@@
 \end{equation*}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[t]
-
-\begin{equation*}
-\begin{array}{ll}
-\sem{\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }~\ctx =
-  \pi_i~(\ident{counit}_{\cunit}~\ctx) & (\emph{var})
-\\[0.5em]
-\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
-  \delta~(c_i) & (\emph{const})
-\\[0.5em]
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
-  \hspace{3em}\sem{\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v))
-\\[0.5em]
-\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
-  \hspace{2.5em}
-  \begin{array}{l}  
-  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}}~
-    (\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~(\lambda x.(x,x))~\ctx) \\[-0.25em]
-  \kvd{in}~\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
-      (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~\sem{\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~\ctx_2)
-  \end{array}    
-\\[1.25em]
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }~\ctx = & (\emph{sub})\\[-0.25em]
-  \hspace{3em}\sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx)~
-    \hspace{7.5em}(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
-\end{array}
-\end{equation*}
-
-\figcaption{Categorical semantics of the flat coeffect calculus}
-\label{fig:flat-semantics}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
 
@@
 
+\begin{theorem}[Generalization]
+\label{thm:flat-generalization}
+The semantics is the same as the concrete one from Chapter~\ref{ch:applications}
+\end{theorem}
+\begin{proof}
+It is really boring.
+\end{proof}
+
 \noindent
@@
 \caption{Common syntax and reduction rules of the target langauge}
-\label{fig:transl-target}
+\label{fig:semantics-target}
 \end{figure}
@@
 
-\section{Target langauge}
+\section{Translation}
+\label{sec:semantics-translation}
+
+\subsection{Target langauge}
 
@@
 \caption{Typing rules for the common syntax of the target language}
-\label{fig:transl-targetty}
+\label{fig:semantics-targetty}
 \end{figure}
@@
 \caption{Comonadically-inspired extensions for the target langauge}
-\label{fig:transl-ext}
+\label{fig:semantics-ext}
 \end{figure}
@@
 \figcaption{Translation from a flat DSL to a comonadically-inspired target language}
-\label{fig:transl-translation}
+\label{fig:semantics-translation}
 \end{figure*}
@@
 
-\section{Coeffect langauge for dataflow}
+\section{Proofs}
+\label{sec:semantics-proofs}
+
+\subsection{Coeffect langauge for dataflow}
 
@@
 
+\begin{theorem}[Soundness]
+\label{thm:semantics-sound-df}  
+Yo
+\end{theorem}
+
 \newpage
@@
 \end{proof}
+
+\begin{theorem}[Soundness]
+\label{thm:semantics-sound-impl}  
+Yo
+\end{theorem}
 
@@
 
-\section{Related work}
-\label{sec:flat-related}
+\section{Related and futrue work}
+\label{sec:semantics-related}
 
@@
 of the computations -- rather than considering just the syntactic aspects of the type system.
+
+\subsection{Generalising}
+... works for any good comonad...
+
+\subsection{Properties and related notions}
+
+The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
+calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
+additional properties and other categorical structures that have been proposed mainly in the 
+context of monads and effects and we look how they relate to indexed comonads.
+
+\paragraph{Shape preservation.}
+Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
+this means that the core comonad structure does not provide a way of modeling computations where
+the additional context changes during the computation. For example, in the \ident{NEList} comonad, 
+the length of the list stays the same after applying \ident{cobind}.
+
+~
+
+Indexed comonads are not restricted by this property of comonads. For example, given the indexed
+product comonad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$, the shape of
+the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
+just implicit parameters $\cclrd{s}$.
+
+\paragraph{Families of monads.}
+When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
+\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
+comonads}, is defined as follows.
+
+\begin{definition}
+\label{def:flat-family}
+A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
+  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
+$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
+$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
+certain coherence conditions.
+\end{definition}
+
+\noindent
+A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data 
+types needs to form a comonad separately. For example, our indexed option does not form a family of 
+comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given 
+a family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
+an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
+we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
+$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.
+
+\paragraph{Parameteric effect monads.}
+Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
+indexed comonads) are closely related to our definition.  Although presented in a more general 
+categorical framework (and using monads), the model defines the \ident{unit} operation only on the 
+unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal 
+structure.
 
@@
 \subsection{When is coeffect a monad}
-\label{sec:flat-related-monads}
+\label{sec:semantics-related-monads}
 
@@
 calculus.
+
+In the upcoming chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
+properties to \emph{structural} coeffect calculi, tracking per-variable information, thus
+covering systems from the second half of Chapter~\ref{ch:applications}.

### 2016-02-12 text/Chapter05.tex
More syntactic theory in Chapter 5

@@
 \begin{example}[Indexed option]
+\label{thm:semantics-indexed-opt}
+
 Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively 
@@
 \begin{example}[Indexed product]
+\label{thm:semantics-indexed-prod}
+
 For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
@@
 \begin{example}[Indexed list]
+\label{thm:semantics-indexed-list}
+
 Our last example provides the semantics of data-flow computations. The flat coeffect algebra 
@@
 \begin{array}{l}
-\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[11em]
+\hspace{-2em}\ctyp{n}{\alpha} = \underbrace{\alpha \times \ldots \times \alpha}_{(n+1)-\textnormal{times}}\\[8em]
 ~\\
@@
 
-~
-
-~
-
-~
-
 % --------------------------------------------------------------------------------------------------
@@
 \begin{array}{ll}
-\hspace{4.8em}\semdef
+\hspace{5.2em}\semdef
   {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
@@
 \\[1em]
-\hspace{5.3em}\semdef
+\hspace{4.9em}\semdef
   {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
   {\ident{const}~n} 
-& (\emph{const})
+& (\emph{num})
 \\[1.5em]
-\hspace{1.5em}\semdeff
+\hspace{2.2em}\semdeff
   {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
@@
   {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\begin{array}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\end{array}}
+  {\begin{array}{l}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\end{array}}
   {f}
   {g}
-  {\begin{array}{l}  
+  {\hspace{-0.5em}\begin{array}{l} 
   \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g)~\circ~\ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}} \\[-0.3em]
   ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
-  \end{array} }   
+  \end{array}\hspace{-0.5em}}   
 & (\emph{app})
@@
 \begin{array}{rcl}
+  \ident{map}_\cclrd{r}}~f &=& \ident{cobind}_{\cunit, \cclrd{r}}~(f\circ\ident{counit}_\cunit) \\[-0.3em]
   \ident{const}~v &=& \lambda x.v \\[-0.3em]
@@
   f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.3em]
-  \ident{app}~f~x &=& f~x
+  \ident{app}~(f, x) &=& f~x
 \end{array}
@@
 \figcaption{Categorical semantics of the flat coeffect calculus}
-\label{fig:flat-semantics}
+\label{fig:semantics-flat}
 \end{figure*}
@@
 
-\paragraph{Contexts and functions.}
+\paragraph{Contexts and types.}
 The modelling of contexts and functions generalizes the earlier concrete examples. We use the 
@@
 \paragraph{Expressions.}
-The definition of the semantics is shown in Figure~\ref{fig:flat-semantics}. For readability, we 
-write the definitions in a simple programming language notation as opposed to the point-free 
-categorical style. However, it can be equally written using just the operations of flat indexed 
-comonad together with $i^{th}$ projection from a tuple represented by $\pi_i$, \emph{curry} and 
-\emph{uncurry}, function composition, value duplication ($\Delta : A \rightarrow A \times A$) and 
-function pairing (given $f:A\rightarrow B$ and $g:C\rightarrow D$ then $f\times g : A\times C \rightarrow B \times D$). 
-These operations can be provided by \eg~a Cartesian-closed category.
+The definition of the semantics is shown in Figure~\ref{fig:semantics-flat}. For consistency
+with earlier work \cite{comonads-notions,comonads-dom-thesis}, the definitions use a point-free 
+categorical notation. The semantics uses a number of auxiliary definitions that can be expressed
+in a Cartesian-closed category such as currying \ident{curry}, value duplication \ident{dup},
+function pairing (given $f:A\rightarrow B$ and $g:C\rightarrow D$ then $f\times g : A\times C \rightarrow B \times D$)
+and application \ident{app}. We will embed the definitions in a simple programming language 
+later (Section~\ref{sec:semantics-translation}).
 
@@
 Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses 
-$\ident{counit}_{\cunit}$ to extract product of free-variables from the context and then projection
-$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) takes the context $\ctx$ and function argument 
-$v$ and merges their additional contexts using $\ident{merge}_{\cclrd{r}, \cclrd{s}}$. Assuming
-the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives us a 
-value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
+$\ident{counit}_{\cunit}$ to extract a product of free variables and projection
+$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) is interpreted as a curried
+function that takes the declaration-site context and a function argument, merges them using
+$\ident{merge}_{\cclrd{r}, \cclrd{s}}$ and passes the result to the semantics of the body $f$. 
+Assuming the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives 
+us a value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
 Assuming that $n$-element tuples are associated to the left, the wrapped context is equivalent to
@@
 
-The semantics of application is more complex. It first duplicates the free-variable product inside the 
+The semantics of application (\emph{app}) first duplicates the free-variable product inside the 
 context (using $\ident{map}_{\cclrd{r}}$ and duplication). Then it splits this context using 
@@
 (as required by sub-expressions $e_1$ and $e_2$), but different coeffect annotations. The first
-context (with index $\cclrd{r}$) is used to evaluate $e_1$, resulting in a function 
-$\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. To obtain the result, we compose this with a function
-created by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$ on the semantics of sub-expression
-$e_2$, which is of type $\ctyp{\cclrd{s}\cseq\cclrd{t}}{\sigma_1 \times \ldots \times \sigma_n}
-\rightarrow \ctyp{\cclrd{t}}{\tau_1}$.
-
-Finally, constants (\emph{const}) are modelled by a global dictionary $\delta$ and sub-coeffecting
-is interpreted by dropping additional context from the provided context $\ctx$ using 
-$\ident{lift}_{\cclrd{r}, \cclrd{r'}}$ and providing it to the semantics of the assumption.
+context (with index $\cclrd{r}$) is used to evaluate $e_1$ using the semantic function $f$. The 
+result is a function $\ctyp{\cclrd{t}}{\tau_1} \rightarrow \tau_2$. The second context
+(with index $\cclrd{s}\cseq\cclrd{t}$) is used to evaluate $e_2$ and using the semantic function $g$
+and wrap it with context required by the function $e_1$ by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$.
+The \ident{app} operation than applies the function (first element) on the argument (second element).
+Finally, numbers (\emph{num}) become constant functions that ignore the context.
 
 \paragraph{Properties.}
-
-The categorical semantics can be used to embed context-dependent computations in functional 
-programming languages, similarly to how monads provide a way of embedding effectful computations.
-More importantly, it also provides validation for the design of the type system developed in 
-Section~\ref{sec:flat-calculus-types}. As stated in the following theorem, the annotations in 
-the type system match those of the semantic functions.
-
-\vspace{2em}
-
-\begin{theorem}[Correspondence]
+We the categorical semantics  in Section~\ref{sec:semantics-translation} to define a translation 
+that embeds context-dependent computations in a functional programming langauge, similarly to how 
+monads and the do notation provide a way of embedding effectful computations in Haskell.
+
+An important property of the translation is that it respects the coeffect annotations provided 
+by the type system. The annotations of the semantic functions match the annotations in the typing
+judgement and so the semantics is well-defined. 
+This provides a further validation for the design of the type system developed in 
+Section~\ref{sec:flat-calculus-types} -- if the coeffect annotations for (\emph{app}) and (\emph{abs}) 
+were different, we would not be able to provide a well-defined semantics using flat indexed comonads.
+
+Informally, the following states that if we see the semantics as a translation, the resulting code
+is well-typed. We revisit the property in Lemma~\ref{thm:semantics-welltyped} once we define
+the target langauge and its typing.
+
+\begin{lemma}[Correspondence]
 \label{thm:flat-correspondence}
-In all of the typing rules of the flat coeffect system, the context annotations $\cclrd{r}$ of typing 
-judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ 
-correspond to the indices of mappings $\ctyp{\cclrd{r}}{}$ in the corresponding semantic function defined 
-by $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}$.
+In the semantics defined in Figure~\ref{fig:semantics-flat}, the context annotations $\cclrd{r}$ of 
+typing judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types 
+$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$  on the left-hand side correspond to the indices of mappings 
+$\ctyp{\cclrd{r}}{}$ in the corresponding semantic function on the right-hand side.
+\end{lemma}
+\begin{proof}
+By analysis of the semantic rules in Figure~\ref{fig:semantics-flat}.
+\end{proof}
+
+\noindent
+Thanks to indexing, the correspondence provides more guarantees than for a non-indexed system.
+In the semantics, we not only know which values are comonadic, but we also know what contextual
+information they are required to provide. In Section~\ref{sec:semantics-generalising}, we note
+that this lets us generalize the proofs about concrete languages discussed in this chapter to
+a more general setting.
+
+The semantics is also a generalization of the concrete semantics given when introducing
+context-aware programming languages in Chapter~\ref{ch:applications}.
+
+\begin{theorem}[Generalization]
+\label{thm:flat-generalization}
+
+Consider a typing derivation obtained according to the rules for finding unique typing derivations
+as specified in Section~\ref{sec:flat-unique} for a coeffect language with liveness, dataflow or 
+implicit parameters.
+
+The semantics obtained by instantiating the rules in Figure~\ref{fig:semantics-flat} with the
+concrete operations defined in Example~\ref{thm:semantics-indexed-opt},
+Example~\ref{thm:semantics-indexed-prod} or Example~\ref{thm:semantics-indexed-list} is the
+same as the one defined in Figure~\ref{fig:applications-flat-livsem},
+Figure~\ref{fig:applications-flat-implsem} and Figure~\ref{fig:applications-flat-dfsem},
+respectively.
 \end{theorem}
 \begin{proof}
-By analysis of the semantic rules in Figure~\ref{fig:flat-semantics}.
+Simple expansion of the definitions for the unique typing derivation.
 \end{proof}
 
-\begin{theorem}[Generalization]
-\label{thm:flat-generalization}
-The semantics is the same as the concrete one from Chapter~\ref{ch:applications}
-\end{theorem}
-\begin{proof}
-It is really boring.
-\end{proof}
-
-\noindent
-Thanks to indexing, the statement of the theorem is significantly stronger than for a 
-non-indexed system, because it provides the justification for our choice of indices in the typing
-rules. In particular, we can see that the annotations follow from the annotations on primitive 
-functions that define the semantics. Also, each function defining the semantics uses a distinct 
-operation of the coeffect algebra and so the type system is the most general possible definition
-(within the comonadic framework we use).
-
-Although the notion of indexed comonads presented in this section is novel and interesting 
-in its own, the main reason for introducing it is to motivate the flat coeffect type system.
-This is captured by the Theorem~\ref{thm:flat-correspondence}. In the next section, we shift our
-focus from (categorical) semantics to syntactic properties of the calculus.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Comonadic translation}
-
-The previous chapter gives the semantics of coeffect calculus in terms of indexed comonads. In this
-chapter, we follow the example of effects and monads and we use the semantics to define a 
-\emph{translation}.
-
-A context-aware \emph{source} program written using a concrete context-aware domain-specific 
-language (capturing dataflow, implicit parameters or other kinds of context awareness) with
-domain-specific language extensions (the \kvd{prev} keyword, or the \ident{?impl} syntax)
-is translated to a \emph{target} langauge that is not context-aware. The target language is
-a small functional language consisting of:
-%
-\begin{itemize}
-  \item Simple functional subset formed by lambda calculus with tuples and numbers.
-  \item Comonadically-inspired primitives corresponding to \emph{counit}, \emph{cobind} and
-    other operations of flat indexed comonads.
-  \item Additional primitives that model contextual operations of each concrete coeffect language
-    (\emph{prev} for the \kvd{prev} keyword, \emph{lookup} for the \ident{?impl} syntax etc.)
-\end{itemize}
-%
-The syntx, typing and reduction rules of the first part (simple functional language) are used by 
-all concrete coeffect domain-specific langauges. The syntax and typing rules of the second part 
-(comonadically-inspired) primitives are also shared by all coeffect DSLs, however the \emph{reduction
-rules} for the comonadically-inspired primitives differ -- they capture the concrete notion of
-context. Finally, the third part (domain-specific primitives) will differ for each coeffect DSL.
-
-% --------------------------------------------------------------------------------------------------
+% ==================================================================================================
+%
+%      #######
+%      #       #    # #####  ###### #####  #####  # #    #  ####
+%      #       ##  ## #    # #      #    # #    # # ##   # #    #
+%      #####   # ## # #####  #####  #    # #    # # # #  # #
+%      #       #    # #    # #      #    # #    # # #  # # #  ###
+%      #       #    # #    # #      #    # #    # # #   ## #    #
+%      ####### #    # #####  ###### #####  #####  # #    #  ####
+%
+% ==================================================================================================
 
@@
   C  & = & (v_1, \ldots, v_{i-1}, \_, e_{i+1}, \ldots e_n) \sep v~\_ \sep \_~e \sep \pi_i~\_   
-
 \end{array}
@@
 
-\caption{Common syntax and reduction rules of the target langauge}
-\label{fig:semantics-target}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\section{Translation}
-\label{sec:semantics-translation}
-
-\subsection{Target langauge}
-
-The target langauge for the translation is a simply typed lambda calculus with numbers and tuples.
-The translation uses tuples as it keeps a tuple with variable assignments (encoding those without 
-tuples would be possible, but cumbersome) and we add numbers as a basic concrete data type.
-In this section, we define the common parts of the language without the comonadically-inspired
-primitives.
-
-The syntax of the target programming langauge is shown in Figure~\ref{fig:transl-target}. 
-The values include numbers $n$, tuples and function values.
-The expressions include variables $x$, values, lambda abstraction and application and operations 
-on tuples. We do not need recursion (although a realistic programming language would include it). 
-In what follows, we also use the following syntactic sugar for let binding:
-%
-\begin{equation*}
-\begin{array}{rcl}
-  \kvd{let}~x=e_1~\kvd{in}~e_2 &=& (\lambda x.e_2) e_1
-\end{array}
-\end{equation*}
-%
-Finally, $C[e]$ defines the context in which sub-expressions are evaluated. Together with the
-evaluation rules shown in  Figure~\ref{fig:transl-target}, this captures the standard call-by-name
-semantics of the common parts of the target language.
-The (standard) typing rules for the common expressions of the target langauge are shown in 
-Figure~\ref{fig:transl-targetty}. The rules are standard.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
+~
 \paragraph{Typing rules}
 \begin{equation*}
-\begin{array}{ll}
+\begin{array}{l}
 \tyrule{var}
@@
   { \Gamma \vdash x:\tau }
-&
-\tyrule{err}
-  { ~ }
-  { \Gamma \vdash \ident{error}:\tau }
-\\[1em]
-\tyrule{str}
-  { ~ }
-  { \Gamma \vdash s:\ident{str} }
-&
+\\[1.25em]
 \tyrule{num}
@@
   { \Gamma \vdash n:\ident{num} }
-\\[1em]
-\tyrule{nil}
-  { ~ }
-  { \Gamma \vdash []:[\tau] }
-&
-\tyrule{head}
-  { \Gamma \vdash e :[\tau] }
-  { \Gamma \vdash \ident{head}~e:\tau }
-\\[1em]
-\tyrule{tail}
-  { \Gamma \vdash e :[\tau] }
-  { \Gamma \vdash \ident{tail}~e:[\tau] }
-\qquad~&
+\\[1.25em]
 \tyrule{abs}
@@
   { \Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
-\end{array}
-\end{equation*}
-\begin{equation*}
-\begin{array}{l}
+\\[1.25em]
 \tyrule{app}
@@
   { \Gamma \vdash e_1~e_2 : \tau_2 }
-\\[1.25em]
-\tyrule{cons}
-  { \Gamma \vdash e_1 :\tau & \Gamma \vdash e_2 : [\tau] }
-  { \Gamma \vdash e_1\!::\!e_2 : [\tau] }
 \\[1.25em]
@@
   { \Gamma \vdash (e_1, \ldots, e_n) : \tau_1 \times \ldots \times \tau_n }
-\\[1.25em]
-\tyrule{ifn}
-  { \Gamma \vdash e_1 : \ident{num} & \Gamma \vdash e_2 : \ident{num} &
-    \Gamma \vdash e_3 : \tau & \Gamma \vdash e_4 : \tau}
-  { \Gamma \vdash \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4 : \tau}
-\\[1.25em]
-\tyrule{ifs}
-  { \Gamma \vdash e_1 : \ident{str} & \Gamma \vdash e_2 : \ident{str} &
-    \Gamma \vdash e_3 : \tau & \Gamma \vdash e_4 : \tau}
-  { \Gamma \vdash \kvd{if}~e_1 = e_2~\kvd{then}~e_3~\kvd{else}~e_4 : \tau}
-\end{array}
-\end{equation*}
-
-\caption{Typing rules for the common syntax of the target language}
-\label{fig:semantics-targetty}
+\end{array}
+\end{equation*}
+
+\caption{Common syntax and reduction rules of the target langauge}
+\label{fig:semantics-target}
 \end{figure}
@@
 
-\subsection{Properties}
-
-The subset of the langauge described so far models a simple ML-like functional programming 
-langauge (or, Haskell-like language, if we choose call-by-name evaluation). The subset of the
-langauge introduced so far satisfies the property that ``well-typed programs do not get stuck'',
-i.e.~both type preservation (reduction does not change the type of an expression) and the progress 
-property (an well-typed expression is either a value or can be reduced).
-We first show this for the subset of the language discussed so far and later extend the proof to 
-also cover the comonadically-inspired primitives that will be added in the next section.
+\section{Embedding coeffect languages}
+\label{sec:semantics-translation}
+
+Although the notion of indexed comonads presented in the previous section is novel and interesting 
+in its own, the main reason for introducing it is that we can view it as a translation that provides 
+embedding of context-aware domain-specific languages in a simple target functional language.
+In this section, we follow the example of effects and monads and we use the semantics to define a 
+translation akin to the do notation in Haskell.
+
+A context-aware \emph{source} program written using a concrete context-aware domain-specific 
+language (capturing dataflow, implicit parameters or other kinds of context awareness) with
+domain-specific language extensions (the \kvd{prev} keyword, or the \ident{?impl} syntax)
+is translated to a \emph{target} langauge that is not context-aware. The target language is
+a small functional language consisting of:
+%
+\begin{itemize}
+  \item Simple functional subset formed by lambda calculus with support for tuples and numbers.
+  \item Comonadically-inspired primitives corresponding to \emph{counit}, \emph{cobind} and
+    other operations of flat indexed comonads.
+  \item Additional primitives that model contextual operations of each concrete coeffect language
+    (\emph{prev} for the \kvd{prev} keyword, \emph{lookup} for the \ident{?p} syntax and
+    \emph{letimpl} for the $\kvd{let}~\ident{?p}=\ldots$ notation).
+\end{itemize}
+%
+The syntx, typing and reduction rules of the first part (simple functional language) are used by 
+all concrete coeffect domain-specific langauges. The syntax and typing rules of the second part 
+(comonadically-inspired) primitives are also shared by all coeffect DSLs, however the \emph{reduction
+rules} for the comonadically-inspired primitives differ -- they capture the concrete notion of
+context. Finally, the third part (domain-specific primitives) will differ for each coeffect DSL.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Functional target langauge}
+
+The target langauge for the translation is a simply typed lambda calculus with integers and tuples.
+We include integers as an example of a concrete type. Tuples are needed by the translation, which
+keeps a tuple of variable assignments. Encoding those without tuples would be possible, but 
+cumbersome. In this section, we define the common parts of the language without the 
+comonadically-inspired primitives.
+
+The syntax of the target programming langauge is shown in Figure~\ref{fig:semantics-target}. 
+The values include numbers $n$, tuples and function values. The expressions include variables $x$, 
+values, lambda abstraction and application and operations on tuples. We do not need recursion or 
+other data types (although a realistic programming language would include them). 
+In what follows, we also use the following syntactic sugar for let binding:
+%
+\begin{equation*}
+\begin{array}{rcl}
+  \kvd{let}~x=e_1~\kvd{in}~e_2 &=& (\lambda x.e_2)~e_1
+\end{array}
+\end{equation*}
+%
+Finally, $C[e]$ defines the context in which sub-expressions are evaluated. Together with the
+evaluation rules shown in  Figure~\ref{fig:semantics-target}, this captures the standard 
+call-by-name semantics of the common parts of the target language. The (standard) typing rules 
+for the common expressions of the target langauge are also shown in Figure~\ref{fig:semantics-target}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Safety of functional target language}
+
+The functional subset of the langauge described so far models a simple ML-like langauge. We 
+choose call-by-value over call-by-name for no particular reason and Haskell-like language would
+work equally well.
+
+The subset of the language introduced so far is type-safe in the standard sense that
+``well-typed programs do not get stuck''. Although standard, we outline the important parts
+of the proof for the functional subset here, before we extend it to concrete context-aware
+languages in Section~\ref{sec:semantics-proofs}.
+
+We use the standard syntactic approach to type safety introduced by Milner \cite{syntactic-sml}.
+Following Wright, Felleisen and Pierce \cite{syntactic-tapl,syntactic-approach}, we prove the
+type preservation property (reduction does not change the type of an expression) and the progress 
+property (a well-typed expression is either a value or can be further reduced).
+
+\begin{lemma}[Canonical forms]
+\label{thm:semantics-fp-canon}
+For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
+\begin{enumerate}
+  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
+  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
+  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
+\end{enumerate}  
+\end{lemma}
+\begin{proof}
+  For (1), the last typing rule must have been (\emph{num}); for (2), it must have been
+  (\emph{abs}) and for (3), the last typing rule must have been (\emph{tup})
+\end{proof}
+
+\begin{lemma}[Preservation under substitution]
+\label{thm:semantics-fp-pres-subst}
+For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$ 
+then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
+\end{lemma}
+\begin{proof}
+By induction over the derivation of $\Gamma, x:\tau \vdash e : \tau'$. 
+\end{proof}
 
 \begin{theorem}[Type preservation]
+\label{thm:semantics-fp-pres}
   If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
@@
   Rule induction over $\rightarrow$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{fn}): $e = (\lambda x.e_0)~v$, from Lemma~\ref{thm:semantics-fp-pres-subst}
+  it follows that $\Gamma \vdash e_0[x \leftarrow v] : \tau$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{prj}): $e = \pi_i(v_1, \ldots, v_n)$ and so the last applied typing rule must have been
+  (\emph{tup}) and $\Gamma \vdash (v_1, \ldots, v_n) : \tau_1 \times\ldots\times \tau_n$ and
+  $\tau = \tau_i$. After applyng (\emph{prj}) reduction, $e' = v_i$ and so $\Gamma \vdash e' : \tau_i$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{ctx}): By induction hypothesis, the type of the reduced sub-expression does not change
+  and the last used rule in the derivation of $\Gamma \vdash e : \tau$ also applies on $e'$
+  giving $\Gamma \vdash e' : \tau$.  
 \end{proof}
@@
 \begin{theorem}[Progress]
-  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+\label{thm:semantics-fp-prog}  
+  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
 \end{theorem}
@@
   By rule induction over $\vdash$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{num}): $e = n$ for some $n$ and so $e$ is a value.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{abs}): $e = \lambda x.e'$ for some $x, e'$, which is a value.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{var}): This case cannot occur, because $e$ is a closed expression.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{app}): $e = e_1~e_2$ which is not a value. By induction, $e_1$ is either
+a value or it can reduce. If it can reduce, apply (\emph{ctx}) reduction with context $\_~e$. Otherwise
+consider $e_2$. If it can reduce, apply (\emph{ctx}) with context $v~\_$. If both are values,
+Lemma~\ref{thm:semantics-fp-canon} guarantees that $e_1 = \lambda x.e_1'$ and so we can apply
+reduction (\emph{fn}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{proj}): $e = \pi_i e_0$ and $\tau=\tau_1\times\ldots\tau_n$. If $e_0$ can be reduced, 
+  apply (\emph{ctx}) with context $\pi_i~\_$. Otherwise from Lemma~\ref{thm:semantics-fp-canon}, 
+  we have that $e_0=(v_1, \ldots, v_n)$ and we can apply reduction (\emph{prj}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{tup}): $e = (e_1, \ldots, e_n)$. If all sub-expressions are values, then $e$ is also
+  a value. Otherwise, we can apply reduction using (\emph{ctx}) with a context $(v_1, \ldots, v_{i-1}, \_, e_{i+1}, \ldots, e_n)$.
 \end{proof}
 
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect-specific extnesions}
-
-Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ of a concrete 
-coeffect domain-specific langauge, we first extend the language syntax and typing rules with 
-the terms that correspond to the comonadically-inspired operations. This is done in the same way
-for all concrete coeffect DSLs and so we give the additional syntax, evaluation context and 
-typing rules just once in Figure~\ref{fig:transl-ext}.
-
-The figure defines the syntax and the typing rules, but it does not define the reduction rules.
-Those -- together with the values for a concrete notion of context -- will be defined separately
-for each individual coeffect DSL. We first consider the DSL for dataflow programming.
+\begin{theorem}[Safety]
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow^* e'$ then either $e'$ is a value or there
+  exists $e''$ such that $e' \rightarrow e''$
+\end{theorem}
+\begin{proof}
+  Rule induction over $\rightarrow^*$ using Theorem~\ref{thm:semantics-fp-pres} and Theorem~\ref{thm:semantics-fp-prog}.
+\end{proof}
 
@@
   C  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~\_~e \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
-    &\sep& \ident{merge}_{\cclrd{r},\cclrd{s}}~\_
+    && \hspace{1.4em} \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~\_
            \sep \ident{split}_{\cclrd{r},\cclrd{s}}~\_ \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~\_ \\  
-
 \end{array}
@@
   {\Gamma \vdash  \ident{split}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\cclrd{r}}{\tau_1} \times \ctyp{\cclrd{s}}{\tau_2}} \\
-\\
-\tyrule{lift}
-  {\Gamma \vdash  e : \ctyp{\cclrd{r'}}{\tau} }
-  {\Gamma \vdash \ident{lift}_{\cclrd{r'},\cclrd{r}}~e : \ctyp{\cclrd{r}}{\tau} } \qquad\forall \cclrd{r'},\cclrd{r}, \cclrd{r}\,\cleq\,\cclrd{r'}
-\\
 \end{array}
@@
 
-In Chapter 4, we presented the semantics of the flat coeffect calculus in terms of indexed comonads.
-We treated the semantics as denotational -- interpreting the meaning of a given typing derivation
-of a program in terms of category theory.
+In Section~\ref{sec:semantics-theory}, we presented the semantics of the flat coeffect calculus in 
+terms of indexed comonads. We treated the semantics as denotational -- interpreting the meaning of 
+a given typing derivation of a program in terms of category theory.
 
@@
 
-Similarly, the interpretation of contexts and types in the category now becomes a translation from types and
+\paragraph{Language extension.}
+Given a coeffect language with a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
+we first extend the language syntax and typing rules with terms that correspond to the 
+comonadically-inspired operations. This is done in the same way for all concrete coeffect 
+domain-specific languages and so we give the additional syntax, evaluation context and 
+typing rules in Figure~\ref{fig:semantics-ext} for all languages we consider later in 
+Section~\ref{sec:semantics-proofs}.
+
+The new type $\ctyp{\cclrd{r}}$ represents an indexed comonad, which is left abstract for now.
+The additional expressions such as $\ident{counit}_\cunit$ and $\ident{cobind}_{\cclrd{r,s}}$ 
+correspond to the operations of indexed comonads. Note that the we embed the coeffect annotations 
+into the target language -- these are known when translationg a typed term from a source language
+and they will be useful when proving that sufficient context (as specified by the coeffect
+annotations) is available.
+
+The figure defines the syntax and the typing rules, but it does not define the reduction rules.
+Those -- together with the values for a concrete notion of context -- will be defined separately
+for each individual coeffect language.
+
+\paragraph{Contexts and types.}
+The interpretation of contexts and types in the category now becomes a translation from types and
 contexts in the source language into the types of the target language:
-
+%
 \begin{equation*}
@@
 
+\noindent
 Here, a context becomes a comonadically-inspired data type wrapping a tuple of variable values 
@@
 
-The rules shown in Figure~\ref{fig:transl-translation} are very similar to those shown earlier
-Figure~X in Chapter~3. Thanks to the equivalence between $\lambda$-calculus and category theory,
-we now interpret them as denoting a translation. Furthermore, the program produced by the above
-rules is well-typed.
-
-\begin{lemma}[Well-typedness of the translation]
-Given a well-typed coeffect program $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ with a 
-typing derivation such that:
-%  
-\begin{equation*}  
-  \semdeff
-    {(\ldots)\qquad~}
-    {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
-    {(\ldots)}
-    { ~~f }
-\end{equation*}  
-%
-Then $e$ is a valid expression in the target langauge and $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
-\end{lemma}
-\begin{proof}
-  By rule induction over the translation rules.
-\end{proof}
+% --------------------------------------------------------------------------------------------------
 
@@
 
+The translation is defined over a typing derivation:
+%
 \begin{equation*}
@@
 \\[1em]
-\hspace{5.3em}\semdef
+\hspace{4.3em}\semdef
   {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
   {\lambda\ctx.n} 
-& (\emph{const})
+& (\emph{num})
 \\[1.5em]
 \hspace{1.5em}\semdeff
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+  {\coctx{\Gamma, x_i:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
+  {\hspace{-1em}\begin{array}{l}  
+    \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x_i.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2
+    \\[-0.3em]~\\[-0.3em]~ \\[-0.3em]~
+  \end{array} }
   {f}
-  {\lambda \ctx.\lambda v.f~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v)) }
+  {\hspace{-1em}\begin{array}{l}  
+    \lambda \ctx.\lambda v.\\[-0.3em]
+    \quad \kvd{let}~\ident{assoc} = \lambda x.\\[-0.3em]
+    \quad\quad (\pi_1(\pi_1~x), \ldots, \pi_{i-1}(\pi_1~x), \pi_2~x)\\[-0.3em]
+    \quad f~(\ident{map}_{\cclrd{r}\czip\cclrd{s}}~\ident{assoc}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v)))
+   \end{array} }
 & (\emph{abs})
-\\[2em]
+\\[4em]
 \semdefff
@@
   {g}
-  {\begin{array}{l}  
+  {\hspace{-1em}\begin{array}{l}  
   \lambda\ctx.~~\\[-0.3em]
-    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~(\lambda x.(x,x))~\ctx\\[-0.3em]
+    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}~\ctx\\[-0.3em]
     \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}}~\ctx_0\\[-0.3em]
@@
 & (\emph{app})
-\\[4em]
-\hspace{6em}\semdeff
-  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } 
-  {f}
-  {\lambda\ctx f~(\ident{lift}_{\cclrd{r}, \cclrd{r'}}~\ctx}
-\quad(\hspace{-0.7em}\textnormal{\footnotesize when}~\cclrd{r'} \leq \cclrd{r})
-& (\emph{sub})\\[-0.25em]
+\end{array}
+\end{equation*}
+
+~
+
+~
+
+Assuming the following auxiliary operations:
+%
+\begin{equation*}
+\begin{array}{rcl}
+  \ident{map}_\cclrd{r}}~f &=& \ident{cobind}_{\cunit, \cclrd{r}}~(\lambda x.f~({\ident{counit}_\cunit}~x)) \\[-0.3em]
+  \ident{dup} &=& \lambda x. (x, x)
 \end{array}
@@
 
-\section{Proofs}
+\paragraph{Expressions.}
+The rules shown in Figure~\ref{fig:semantics-translation} define how expressions of the source
+language are translated into the target language. The rules are very similar to those shown earlier
+in Figure~\ref{fig:semantics-flat}. The result is now written as source code in the
+target programming language rather than as composition of morphisms in a category. However,
+thanks to the equivalence between $\lambda$-calculus and category theory, both interpretations 
+are equivalent.
+
+One change from Figure~\ref{fig:semantics-flat} is that we are now more explicit about the tuple
+that contains variable assignments. Previously, we assumed that the tuple is appropriately 
+reassociated. Now, we perform the reassociation explicitly. We keep a flat tuple of variables, so
+given $\Gamma=x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n$, the tuple has a type $\tau_1\times\ldots\times\tau_n$.
+In (\emph{var}), we access a variable using $\pi$, but in (\emph{abs}), the \ident{merge} operation
+produces a tuple $(\tau_1\times\ldots\times\tau_{i-1})\times\tau_i$ that we turn into a flat
+tuple $\tau_1\times\ldots\times\tau_{i-1}\times\tau_i$ using the \ident{assoc} function.
+
+\paragraph{Properties.} 
+The most important property of the translation is that it produces well-typed programs in the
+target language. This is akin to the correspondence property of the semantics discussed earlier
+(Theorem~\ref{thm:flat-correspondence}), but now it has more obvious practical consequences.
+
+In Section~\ref{sec:semantics-proofs}, we will prove safety properties of well-typed programs in the 
+target language. Thanks to the fact that the translation produces a well-typed program (as complex
+as it might be) means that we are also proving safety of well-typed programs in the source 
+context aware languages.
+
+\begin{theorem}[Well-typedness of the translation]
+\label{thm:semantics-welltyped}
+Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$
+written in a context-aware programming langugae that is translated to the target langauge as:
+%  
+\begin{equation*}  
+  \semdeff
+    {(\ldots)\qquad~}
+    {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
+    {(\ldots)}
+    { ~~f }
+\end{equation*}  
+%
+Then $f$ is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
+\end{theorem}
+\begin{proof}
+By rule induction over the derivation of the translation. Given a judgement
+$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\cclrd{c}} \vdash e : \tau$, the translation
+constructs a function of type $\ctyp{\cclrd{c}}(\sem{\tau_1} \times\ldots\times \sem{\tau_n}) \rightarrow \sem{\tau}$.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{var}): $\cclrd{c}=\cunit$ and $\tau = \tau_i$ and so $\pi_i ({\ident{counit}_\cunit}~\ctx)$ is well-typed.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{num}): $\tau = \ident{num}$ and so the body $n$ is well-typed.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\cclrd{r}}(\ldots)$ and the type of $v$ is $\ctyp{\cclrd{s}}\tau_1$,
+  calling $\ident{merge}_{\cclrd{r},\cclrd{s}}$ and reassociating produces
+  $\ctyp{\cclrd{r}\czip\cclrd{s}}(\ldots)$ as expected by $f$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{num}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of 
+  $\ctx_1, \ctx_2$ are $\ctyp{\cclrd{r}}(\ldots)$ and $\ctyp{\cclrd{s}\cseq\cclrd{t}}(\ldots)$, respectively.
+  $g$ requires $\ctyp{\cclrd{s}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{s},\cclrd{t}}$ is
+  $\ctyp{\cclrd{t}}\tau_1$ as required by $f$.
+\end{proof}
+
+% ==================================================================================================
+%
+%       #####
+%      #     #   ##   ###### ###### ##### #   #
+%      #        #  #  #      #        #    # #
+%       #####  #    # #####  #####    #     #
+%            # ###### #      #        #     #
+%      #     # #    # #      #        #     #
+%       #####  #    # #      ######   #     #
+%
+% ==================================================================================================
+
+\section{Safety of context-aware languages}
 \label{sec:semantics-proofs}
 
+The previous two languages provide a general structrue that we now use to prove the safety of 
+concrete context-aware programming languages based on the coeffect language framework. We
+consider a langauge for dataflow computations (Section~\ref{sec:semantics-proofs-df}) and
+for implicit parameters (Section~\ref{sec:semantics-proofs-impl}). In both cases, we extend
+the progress and preservation theorems of the functional subset of the target language, but the
+approach can be generalised as discussed in Section~\ref{sec:semantics-generalising}.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Coeffect langauge for dataflow}
+\label{sec:semantics-proofs-df}
 
@@
 
-We first consider the extensions needed for dataflow computations. As discussed earlier in the
+We first consider dataflow computations. As discussed earlier in the
 semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
-values.
-
-When evaluating translated programs, the comonadic values will not be directly manipulated by the
-user code and so we introduce a new class of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$. 
-Furthermore, the \ident{Df} values will only appear as fully evaluated inputs and they will be manipulated 
-using the comonadically-inspired operations. Thus, the typing rule (\emph{df}) only needs to 
-check values. We do not need a similar rule for expressions. Construct such as $\ident{Df}\langle(\kvd{fun}~x.x)~1\rangle$ 
-are not allowed in the target language. The (\emph{df}) rule also guarantees that the number of 
-elements in the list matches the number in the coeffect:
-
-\begin{equation*}
-\begin{array}{l}
-v = \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
-e = \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
-\end{array}  
-\begin{array}{l}
-\tyrule{df}
-  { \forall i\in\{ 0 \ldots n \}. ~ \vdash v : \tau}
-  { \Gamma \vdash \ident{Df}\langle v_0, \ldots, v_n \rangle : \ctyp{\cclrd{n}}{\tau} }
-\end{array}  
-\end{equation*}
-~
-
-\noindent
+values. When reducing translated programs, the comonadic values will not be directly manipulated by the
+user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
+operations are the comonadically-inspired ones defined earlier.
+
+The Figure~\ref{fig:semantics-ext-df} shows extensions to the target language for modelling
+dataflow computations. We introduce a new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
+and a matching kind of expressions. We add a language primitive modelling the semantics of the
+\kvd{prev} keyword together with its typing rule (\emph{prev}) and we also add
+a typing rule (\emph{df}) that checks the types of the elements of the stream and also guarantees 
+that the number of elements in the stream matches the number in the coeffect.
+
 The additional reduction rules mirror the semantics that we discussed earlier when talking about
 indexed dataflow comonad.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\paragraph{Language syntax}
+
+\begin{equation*}
+\begin{array}{rcl}
+v &=& \ldots \sep \ident{Df}\langle v_0, \ldots, v_n\rangle \\
+e &=& \ldots \sep \ident{Df}\langle e_0, \ldots, e_n\rangle \sep \ident{prev}_{\cclrd{n}}~e\\
+C &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_
+\end{array}  
+\end{equation*}
+
+\vspace{1.25em}
+\paragraph{Typing rules}
+
+\begin{equation*}
+\tyrule{df}
+  { \forall i\in\{ 0 \ldots n \}. ~ \Gamma \vdash e_i : \tau}
+  { \Gamma \vdash \ident{Df}\langle e_0, \ldots, e_n \rangle : \ctyp{\cclrd{n}}{\tau} }
+\end{equation*}
+\begin{equation*}
+\tyrule{prev}
+  { \Gamma \vdash e : \ctyp{\cclrd{n+1}}{\tau} }
+  { \Gamma \vdash \ident{prev}_{\cclrd{n}}~e : \ctyp{\cclrd{n}}{\tau} }
+\end{equation*}
+
+\vspace{1.25em}
+\paragraph{Translation}
+
+\begin{equation*}
+\semdeff
+  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
+  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
+  {f}
+  {\lambda ctx.\ident{prev}_{\cclrd{n}}~\ctx }
+\end{equation*}
+
+\vspace{1.25em}
+\paragraph{Reduction rules}
 
@@
 {\footnotesize(\emph{cobind})} & \begin{array}{l}
-\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f (\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
+\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
 \\[-0.25em]
@@
 {\footnotesize(\emph{merge})} & \begin{array}{l}
-\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle b_0, \ldots, b_n\rangle)) \rightarrow
+\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle v'_0, \ldots, v'_n\rangle)) \rightarrow
 \\[-0.25em]
-\quad (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{min}(m,n)}, b_{\mathit{min}(m,n)}) \rangle)
+\quad (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{min}(m,n)}, v'_{\mathit{min}(m,n)}) \rangle)
 \end{array}
@@
 \\[0.5em]
-{\footnotesize(\emph{lift})} & \begin{array}{l}
-\ident{lift}_{\cclrd{n'}, \cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_{n'} \rangle) \rightarrow
-\qquad(\textnormal{when}~n\leq n') \\[-0.25em]
+{\footnotesize(\emph{prev})} & \begin{array}{l}
+\ident{prev}_{\cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_n, v_{n+1} \rangle) \rightarrow
+\\[-0.25em]
 \quad \ident{Df}\langle v_0, \ldots, v_n \rangle
@@
 
-~
-
-\noindent
-Now consider a target langauge consisting of the core (ML-subset) defined in Figure~\ref{fig:transl-target}
-with typing rules defined in Figure~\ref{fig:transl-targetty} and comonadically-inspired primtiives
-defined in Figure~\ref{fig:transl-ext} and also concrete notion of comonadically-inspired value
-and reduction rules for data flow as defined above.
+\caption{Additional constructs for modelling dataflow in the target langauge}
+\label{fig:semantics-ext-df}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Properties.}
+Now consider a target langauge consisting of the core (ML-subset) defined by the syntax,
+reduction rules and typing rules given in Figure~\ref{fig:semantics-target} and comonadically-inspired 
+primtives defined in Figure~\ref{fig:semantics-ext} and also concrete notion of comonadically-inspired 
+value and reduction rules for dataflow as defined in Figure~\ref{fig:semantics-ext-df}.
+
+In order to prove type safety, we first extend the canonical forms lemma
+(Lemma~\ref{thm:semantics-fp-canon}) and the preservation under substitution lemma
+(Lemma~\ref{thm:semantics-fp-pres-subst}). Those need to consider the new (\emph{df}) 
+and (\emph{prev}) typing rules and substitution under the newly introduced expression forms 
+$\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\cclrd{n}}$. Then 
+we extend the type preservation (Theorem~\ref{thm:semantics-fp-pres}) and progress
+(Theorem~\ref{thm:semantics-fp-prog}).
+
+\begin{lemma}[Canonical forms]
+\label{thm:semantics-df-canon}
+For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
+\begin{enumerate}
+  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
+  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
+  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
+  \item If $\tau=\ctyp{\cclrd{n}}\tau_1$ then $e = \ident{Df}\langle v_0, \ldots v_n\rangle$ for some $v_i$
+\end{enumerate}  
+\end{lemma}
+\begin{proof}
+  (1,2,3) as before; for (4) the last typing rule must have been (\emph{df}).
+\end{proof}
+
+\begin{lemma}[Preservation under substitution]
+\label{thm:semantics-df-pres-subst}
+For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$ 
+then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $\Gamma, x:\tau \vdash e : \tau'$ as before, with new
+  cases for $\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\cclrd{n}}$.
+\end{proof}
 
 \begin{theorem}[Type preservation]
+\label{thm:semantics-df-pres}
   If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
@@
 \begin{proof}
-As before, using rule induction over $\rightarrow$.
-
-Interesting new cases are the reduction rules (\emph{counit}), (\emph{cobind}), (\emph{merge}),
-(\emph{split}) and (\emph{lift}). We need to show that, given a well typed input value, the resulting
-value is a \ident{Df} value with the right number of elements (for (\emph{cobind}), (\emph{lift}) and (\emph{merge}),
-a tuple of \ident{Df} values with the right number of elements (for (\emph{split})) and a value
-of correct type (for (\emph{unit})).
-
-From the fact that the reduction occurred, we know that the input was a \ident{Df} value with the
-required number of inputs and from the typing rule (\emph{df}) we know that the values in the input
-were all of correct types.  
+  Rule induction over $\rightarrow$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{fn}, \emph{prj}, \emph{ctx}): As before, using Lemma~\ref{thm:semantics-df-pres-subst} for (\emph{fn}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{counit}): $e = \ident{counit}_{\cclrd{0}}(\ident{Df} \langle v_0 \rangle)$. The last rule in the type
+  derivation of $e$ must have been (\emph{counit}) with $\Gamma \vdash \ident{Df} \langle v_0 \rangle : \ctyp{\cclrd{0}}\tau$
+  and therefore $\Gamma \vdash v_0 : \tau$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle)$. The
+  last rule in the type derivation of $e$ must have been (\emph{cobind}) with a type $\tau = \ctyp{\cclrd{n}}\tau_2$ and assumptions
+  $\Gamma \vdash f : \ctyp{\cclrd{m}}{\tau_1} \rightarrow \tau_2$ and $\Gamma \vdash \ident{Df}\langle v_0, \ldots v_{m+n} \rangle : \ctyp{\cclrd{m+n}}{\tau}$.
+  The reduced expression has a type $\ctyp{\cclrd{n}}\tau_2$:
+\begin{equation*}
+\inference
+  { \inference 
+      { \Gamma \vdash f : \ctyp{\cclrd{m}}{\tau_1} \rightarrow \tau_2 & \forall i\in 0 \ldots n.~\Gamma\vdash \ident{Df}\langle v_i, \ldots, v_{i+m} \rangle :  \ctyp{\cclrd{m}}{\tau_1}}  
+      { \forall i\in 0 \ldots n.~\Gamma\vdash f (\ident{Df}\langle v_i, \ldots, v_{i+m} \rangle) : \tau_2} }
+  { \Gamma \vdash \ident{Df}\langle f (\ident{Df}\langle v_0, \ldots, v_m \rangle), \ldots, f (\ident{Df}\langle v_{n}, \ldots, v_{m+n} \rangle) \rangle : \ctyp{\cclrd{n}}\tau_2 }
+\end{equation*}
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{merge}, \emph{split}, \emph{next}): Similar. In all three cases, the last typing rule in the derivation of $e$
+  guarantees that the stream contains a sufficient number of elements of correct type.
 \end{proof}
@@
 \begin{theorem}[Progress]
-  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
-\end{theorem}
-\begin{proof}
-  As before, using rule induction over $\vdash$.
-
-Interesting new cases are those arising from the typing rules for the comonadically-inspired
-primitives in Figure~\ref{fig:transl-ext}, that is (\emph{counit}), (\emph{cobind}), (\emph{merge}),
-(\emph{split}) and (\emph{lift}).  
-In all of the cases, $e$ is not a value and so it needs to reduce and the only reduction rules
-are the ones for dataflow computations (above).
-
-If the argument is a value, one of the reduction rules (above) can be used (the typing guarantees
-that the list has the required number of elements). If the argument is an expressions, we use 
-the induction hypothesis.
-\end{proof}
-
-\begin{theorem}[Soundness]
-\label{thm:semantics-sound-df}  
-Yo
-\end{theorem}
-
-\newpage
-\section{Coeffect language for implicit parameters}
-
-\begin{equation*}
-v = \ident{Impl}(v)
-\end{equation*}
-
-a
-
-\begin{equation*}
-\tyrule{impl}
-  { \Gamma \vdash v : \ident{string} \rightarrow \ident{num}}
-  { a }
-\end{equation*}
-
-b
-
-% Simple for normal parts of the language
-% We have to check this for dataflow reductions  
-%
-% Interesting case is $e_1 e_2$ where $e_1 = v$ is one of the primitives. Then we have to go through
-% them and make sure they can progress.
-
-% And also syntactic sugar \ident{map}
-
-%    &\sep& \ident{cobind}_{\cclrd{s},\cclrd{r}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~e 
-%       \sep \ident{split}_{\cclrd{r},\cclrd{s}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \\
-
-%
-% &\sep& \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e 
-%   \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~e \\
-
-\section{Dataflow}
-Domain specific coeffect annotations are non-negative integers, type is a tuple and new kinds
-of values are lists:
-\begin{equation*}
-\begin{array}{rcl}
-  v_c &  = &  \langle v_1, \ldots, v_n \rangle
-\end{array}
-\end{equation*}
-%
-With the following typing:
-%
-\begin{equation*}  
-\tyrule{val}
-  { \forall v_i . \vdash v_i : \tau }
-  { \vdash \langle v_1, \ldots, v_n \rangle : \ctyp{n}{\tau} }
-\end{equation*}
-%
-
-We could define those in terms of head/tail functions and then prove that they work (and translated
-program does not contain head/tail), but perhaps that's unnecessary overkill. Or perhaps it would
-be useful in order to show how we avoid errors... But doing it directly looks just simpler (we
-also do not need a ``list'' type).
-
-\begin{theorem}[Type preservation]
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
-\end{theorem}
-\begin{proof}
-  Rule induction over $\rightarrow$
-
-Simple for normal parts of the language
-
-We have to check this for dataflow reductions  
-\end{proof}
-
-\begin{theorem}[Progress]
-  If $\Gamma \vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+\label{thm:semantics-df-prog}  
+  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
 \end{theorem}
@@
 
-Interesting case is $e_1 e_2$ where $e_1 = v$ is one of the primitives. Then we have to go through
-them and make sure they can progress.  
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{num,abs,var,app,proj,tup}): As before, using the adapted canonical forms lemma
+  (Lemma~\ref{thm:semantics-df-canon}) for (\emph{app}) and (\emph{proj}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{counit}): $e = {\ident{counit}_\cunit}~e_1$. If $e_1$ is not a value, it can be reduced 
+  using (\emph{ctx}) with context ${\ident{counit}_\cunit}~\_$, otherwise it is a value. From Lemma~\ref{thm:semantics-df-canon},
+  $e_1 = \ident{Df}\langle v\rangle$ and so we can apply (\emph{counit}) reduction rule.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{m}, \cclrd{n}}~e_1~e_2$. If $e_1$ is not a value,
+  reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{m},\cclrd{n}}}~\_~e$. If $e_2$ is
+  not a value reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{m},\cclrd{n}}}~v~\_$.
+  If both are values, then from Lemma~\ref{thm:semantics-df-canon}, we have that 
+  $e_2 = \ident{Df}\langle v_0, \ldots v_{m+n} \rangle$ and so we can apply the (\emph{cobind})
+  reduction.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{merge}): $e = \ident{merge}_{\cclrd{m}, \cclrd{n}} e_1$. If $e_1$ is not a value, 
+  reduce using (\emph{ctx}) with context $e = \ident{merge}_{\cclrd{m}, \cclrd{n}}~\_$. If $e_1$ is
+  a value, it must be a pair of streams $(\ident{Df}\langle v_0, \ldots, v_m \rangle, \ident{Df}\langle v'_0, \ldots, v'_n\rangle)$
+  using Lemma~\ref{thm:semantics-df-canon} and it can reduce using (\emph{merge}) reduction.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{split}, \emph{prev}): Similar. Either sub-expression is not a value, or the type 
+  guarantees that it is a stream with correct number of elements to enable the (\emph{split}) 
+  or (\emph{prev}) reduction, respectively.
 \end{proof}
 
-\begin{theorem}[Soundness]
-\label{thm:semantics-sound-impl}  
-Yo
+\begin{theorem}[Safety]
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow^* e'$ then either $e'$ is a value or there
+  exists $e''$ such that $e' \rightarrow e''$
 \end{theorem}
+\begin{proof}
+  Rule induction over $\rightarrow^*$ using Theorem~\ref{thm:semantics-df-pres} and Theorem~\ref{thm:semantics-df-prog}.
+\end{proof}
+
+\newpage
+
+~
+
+\newpage
+
+\section{Generalising}
+\label{sec:semantics-generalising}
+... works for any good comonad...
 
@@
 of the computations -- rather than considering just the syntactic aspects of the type system.
-
-\subsection{Generalising}
-... works for any good comonad...
 

### 2016-02-12 text/Chapter05.tex
Adding translation for implicits

@@
 \vspace{0.25em}\noindent\hangindent=0.6cm 
-Case (\emph{num}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of 
+Case (\emph{app}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of 
   $\ctx_1, \ctx_2$ are $\ctyp{\cclrd{r}}(\ldots)$ and $\ctyp{\cclrd{s}\cseq\cclrd{t}}(\ldots)$, respectively.
@@
 and (\emph{prev}) typing rules and substitution under the newly introduced expression forms 
-$\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\cclrd{n}}$. Then 
-we extend the type preservation (Theorem~\ref{thm:semantics-fp-pres}) and progress
-(Theorem~\ref{thm:semantics-fp-prog}).
+$\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\cclrd{n}}$. We show that the translation 
+rule for \kvd{prev} produces well-typed expressions. Finally, we extend the type preservation 
+(Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog}).
+
+\begin{theorem}[Well-typedness of the \kvd{prev} translation]
+\label{thm:semantics-welltyped}
+Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$,
+the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
+and Figure~\ref{fig:semantics-ext-df} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
+\end{theorem}
+\begin{proof}
+By rule induction over the derivation of the translation.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{var, num, abs, app}): As before.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{prev}): Type of $\ctx$ is $\ctyp{\cclrd{n+1}}\tau$ and so we can apply the (\emph{prev}) rule.
+\end{proof}
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect langauge for implicit parameters}
+\label{sec:semantics-proofs-impl}
+
+::::TODO::::
+
+The types of the comonadically-inspired operations are the same for each concrete coeffect DSL, but
+each DSL introduces its own \emph{values} of type $\ctyp{\cclrd{r}}\tau$ and also its own reduction
+rules that define how comonadically-inspired operations evaluate.
+
+We first consider dataflow computations. As discussed earlier in the
+semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
+values. When reducing translated programs, the comonadic values will not be directly manipulated by the
+user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
+operations are the comonadically-inspired ones defined earlier.
+
+The Figure~\ref{fig:semantics-ext-df} shows extensions to the target language for modelling
+dataflow computations. We introduce a new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
+and a matching kind of expressions. We add a language primitive modelling the semantics of the
+\kvd{prev} keyword together with its typing rule (\emph{prev}) and we also add
+a typing rule (\emph{df}) that checks the types of the elements of the stream and also guarantees 
+that the number of elements in the stream matches the number in the coeffect.
+
+The additional reduction rules mirror the semantics that we discussed earlier when talking about
+indexed dataflow comonad.
+
+\paragraph{Properties.}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[h!]
+\paragraph{Language syntax}
+
+\begin{equation*}
+\begin{array}{rcl}
+v &=& \ldots \sep \ident{Impl}(v, \langle \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n\rangle) \\
+e &=& \ldots \sep \ident{Impl}(e, \langle\ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n\rangle) \\
+  && \hspace{1.4em} \sep \ident{lookup}_{\cclrd{\ident{?p}}}~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~e_1~e_2\\
+C &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_
+\end{array}  
+\end{equation*}
+
+\vspace{1.25em}
+\paragraph{Typing rules}
+
+\begin{equation*}
+\tyrule{impl}
+  { \Gamma \vdash e:\tau & \forall i\in\{ 1 \ldots n \}. ~ \Gamma \vdash e_i : \ident{num}}
+  { \Gamma \vdash \ident{Impl}(e, \langle \ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n \rangle) : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \}}}{\tau} }
+\end{equation*}
+\begin{equation*}
+\tyrule{lookup}
+  { \Gamma \vdash e : \ctyp{\cclrd{ \{ \ldots, \ident{?p}, \ldots \} }}{\tau} }
+  { \Gamma \vdash \ident{lookup}_{\cclrd{\ident{?p}}}~e : \ident{num} }
+\end{equation*}
+\begin{equation*}
+\tyrule{letimpl}
+  { \Gamma \vdash e_1 : \ident{num} &
+    \Gamma \vdash e_2 : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}{\tau} }
+  { \Gamma \vdash \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}~e_1~e_2 : 
+      \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n, \ident{?p} \} }}{\tau} }
+\end{equation*}
+
+\vspace{1.25em}
+\paragraph{Translation}
+
+\begin{equation*}
+\semdef
+  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
+  {\lambda ctx.\ident{lookup}_{\cclrd{\ident{?p}}}~\ctx }
+\end{equation*}
+
+\begin{equation*}
+\semdefff
+  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1}
+  { \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\hspace{-1em}\begin{array}{l}  
+      \coctx{\Gamma}{\cclrd{r} \;\cclrd{\cup\; (\cclrd{s} \setminus \{ \ident{?p}) \}} }\\[-0.3em]
+        \quad \vdash \kvd{let}~\ident{?p}=e_1\\[-0.3em]
+        \quyad\qquad \kvd{in}~e_2 : \tau_2 \\[-0.3em]
+        ~ \\[-0.3em]
+   \end{array}}
+  {f}
+  {g}
+  {\hspace{-1em}\begin{array}{l}  
+  \lambda\ctx.~~\\[-0.3em]
+    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cclrd{\cup}\; \cclrd{(s\setminus{\ident{?p}})}}~\ident{dup}~\ctx\\[-0.3em]
+    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{(s\setminus{\ident{?p}})}}~\ctx_0\\[-0.3em]
+    \quad g~(\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{(s\setminus{\ident{?p}})}}~(f~\ctx_1)~\ctx_2)
+  \end{array} }   
+\end{equation*}
+
+\vspace{1.25em}
+\paragraph{Reduction rules}
+
+::: TODO ::: Everything except for counit
+
+\begin{equation*}
+\begin{array}{rl}
+{\footnotesize(\emph{counit})} & \begin{array}{l}
+\ident{counit}_{\cclrd{ \emptyset }}~\ident{Impl} (v, \langle \ldots \rangle)   \rightarrow v
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{cobind})} & \begin{array}{l}
+\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
+\\[-0.25em]
+\quad (\ident{Df}\langle f (\ident{Df}\langle v_0, \ldots, v_m \rangle), \ldots, f (\ident{Df}\langle v_{n}, \ldots, v_{m+n} \rangle) \rangle)
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{merge})} & \begin{array}{l}
+\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle v'_0, \ldots, v'_n\rangle)) \rightarrow
+\\[-0.25em]
+\quad (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{min}(m,n)}, v'_{\mathit{min}(m,n)}) \rangle)
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{split})} & \begin{array}{l}
+\ident{split}_{\cclrd{m}, \cclrd{n}} (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle)  \rightarrow
+\\[-0.25em]
+\quad \ident{Df}\langle v_0, \ldots, v_m \rangle, (\ident{Df}\langle b_0, \ldots, b_n\rangle
+\end{array}
+\\[0.5em]
+{\footnotesize(\emph{prev})} & \begin{array}{l}
+\ident{prev}_{\cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_n, v_{n+1} \rangle) \rightarrow
+\\[-0.25em]
+\quad \ident{Df}\langle v_0, \ldots, v_n \rangle
+\end{array}
+\end{array}
+\end{equation*}
+
+\caption{Additional constructs embedding implicit parameters into the language}
+\label{fig:semantics-ext-df}
+\end{figure}
 \newpage

### 2016-02-12 text/Chapter04.tex
Normalize data-flow and dataflow

@@
 properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live}) 
-and whole-context data-flow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
+and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
 structure. First, their lambda abstraction duplicates the requirements. Given a body with context 
@@
 Recall the lambda abstraction rules for the implicit parameters system (annotating the context
-with sets of required parameters) and the data-flow system (annotating the context with the
+with sets of required parameters) and the dataflow system (annotating the context with the
 number of past required values):
@@
 In order to capture both systems using a single calculus, we need a way of unifying the two
-systems. For the data-flow system, this can be achieved by over-approximating the number of 
+systems. For the dataflow system, this can be achieved by over-approximating the number of 
 required past elements:
@@
 In two of the three systems, some of the operators of the flat coeffect algebra coincide, but
-in the data-flow system all three are distinct. Similarly, the two special elements  
+in the dataflow system all three are distinct. Similarly, the two special elements  
 coincide in some, but not all systems. The required laws are motivated by the aim to capture
@@
 It is worth noting that, in some of the systems, the operators $\cpar$ and $\czip$ are the least 
-upper bound and the greatest lower bounds of a lattice. For example, in data-flow computations, they 
+upper bound and the greatest lower bounds of a lattice. For example, in dataflow computations, they 
 are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
@@
 use just by providing a flat coeffect algebra. The following summary defines the systems for implicit 
-parameters, liveness and data-flow. For the latter two, the general calculus has a lambda abstraction
+parameters, liveness and dataflow. For the latter two, the general calculus has a lambda abstraction
 that is compatible with those discussed in Chapter~\ref{ch:flat}, but includes implicit sub-coeffecting.
@@
 
-\begin{example}[Data-flow]
-In data-flow, context is annotated with natural numbers and the flat coeffect algebra is formed 
+\begin{example}[Dataflow]
+In dataflow, context is annotated with natural numbers and the flat coeffect algebra is formed 
 by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$.
@@
 As discussed earlier, sequential composition $\cseq$ is represented by $+$ and pointwise 
-composition $\cpar$ uses $\emph{max}$. For data-flow, we need a third separate operator for
+composition $\cpar$ uses $\emph{max}$. For dataflow, we need a third separate operator for
 lambda abstraction. Annotating the body with $\emph{min}(\cclrd{r}, \cclrd{s})$ ensures that
@@
 As required by the laws, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
-and $(\mathbb{N}, \mathit{min})$ forms a band. Note that data-flow is our first example where 
+and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where 
 $\cseq$ is not idempotent. The distributivity laws require the following to be the case:
@@
 As a further example that was not covered earlier, it is also possible to combine liveness analysis
-and data-flow. In the above calculus, $0$ denotes that we require the current value, but no previous
+and dataflow. In the above calculus, $0$ denotes that we require the current value, but no previous
 values. However, for constants, we do not even need the current value.
 
-\begin{example}[Optimized data-flow]
-In optimized data-flow, context is annotated with natural numbers extended with the $\bot$ element,
+\begin{example}[Optimized dataflow]
+In optimized dataflow, context is annotated with natural numbers extended with the $\bot$ element,
 that is $\mathbb{N}_{\bot} = \{\bot, 0, 1, 2, 3, \ldots \}$ such that $\forall n \in \mathbb{N}. \bot \leq n$.
@@
 which can be based on syntactic substitution. As we discuss later, the notion of context for
-data-flow is more complex.
+dataflow is more complex.
 
@@
 The property holds for the three systems we consider -- for implicit parameters, this is 
-an equality; for liveness and data-flow (which both use a lattice), the greatest lower bound 
+an equality; for liveness and dataflow (which both use a lattice), the greatest lower bound 
 is smaller than the least upper bound.
@@
 
-\paragraph{Data-flow.} 
-The type preservation property does not hold for data-flow. This case is similar to the example
+\paragraph{Dataflow.} 
+The type preservation property does not hold for dataflow. This case is similar to the example
 shown earlier with effectful computations. As a minimal example, consider the substitution of 
@@
 
-\paragraph{Implicit parameters.} In data-flow, there is no typing for the resulting expression that
+\paragraph{Implicit parameters.} In dataflow, there is no typing for the resulting expression that
 preserves the type of the function. However, this is not the case for all systems. Consider substituting
@@
 Liveness is an example of top-pointed coeffects as variables are annotated with 
-$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and data-flow are examples
+$\ident{L}$ and $\ident{D} \leq \ident{L}$, while implicit parameters and dataflow are examples
 of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
@@
 it holds for some flat coeffect systems, but not for all. Out of the examples we discuss, it 
-holds for implicit parameters, but does not hold for liveness and data-flow.
+holds for implicit parameters, but does not hold for liveness and dataflow.
 
@@
 $\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
-marking the context and function argument as live. In case of data-flow, the immediate coeffects 
+marking the context and function argument as live. In case of dataflow, the immediate coeffects 
 are made larger by the lambda abstraction -- the context requirements of the function value are 
@@
 \hspace{-1em}{\footnotesize Liveness} & $\cclrd{s} \sqcap (\cclrd{s} \sqcup \cclrd{r})$ & $\cclrd{s}$ \\ \hline
-\hspace{-1em}{\footnotesize Data-flow} & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
+\hspace{-1em}{\footnotesize Dataflow} & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
 \end{tabular}
@@
 \mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r}) &\geq& \mathit{min}(\cclrd{s_1}, \cclrd{s_2}) + \cclrd{r} 
-  &\quad& \textnormal{(data-flow)} \\
+  &\quad& \textnormal{(dataflow)} \\
 \end{array}
@@
 lambda abstraction -- for implicit parameters, the lambda function needs to split context 
-requirements using $\cclrd{r} \cup \cclrd{s}$, but for data-flow and liveness it suffices to 
+requirements using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to 
 duplicate the requirement $\cclrd{r}$ of the body. We introduced the $\czip$ operation as 
@@
 The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
-oriented coeffect calculi (such as our liveness and data-flow computations), we can use
+oriented coeffect calculi (such as our liveness and dataflow computations), we can use
 the (\emph{idabs}) rule and get an equivalent type system. This alternative formulation
@@
 the structure of the information tracked by the type system. We instantiated the system to 
-capture three specific systems, namely liveness, data-flow and implicit parameters. However,
+capture three specific systems, namely liveness, dataflow and implicit parameters. However,
 the system is more general and an capture numerous other applications outlined in 

### 2016-02-14 text/Chapter05.tex
Add implicit parameters and generalization (Chapter 5)

@@
 user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
-operations are the comonadically-inspired ones defined earlier.
-
-The Figure~\ref{fig:semantics-ext-df} shows extensions to the target language for modelling
-dataflow computations. We introduce a new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
-and a matching kind of expressions. We add a language primitive modelling the semantics of the
-\kvd{prev} keyword together with its typing rule (\emph{prev}) and we also add
-a typing rule (\emph{df}) that checks the types of the elements of the stream and also guarantees 
+operations are the comonadically-inspired ones defined earlier, together with an additional 
+\emph{domain-specific} operation that models the \kvd{prev} construct.
+
+The Figure~\ref{fig:semantics-ext-df} extensionds the target language with syntax, typing rules,
+additional translation rule and reductions for modelling dataflow computations. We introduce a 
+new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
+and a matching kind of expressions. We specify how the \kvd{prev} keyword is translated into a
+$\ident{prev}_\cclrd{r}$ operation of the target language nd we also add a typing rule
+(\emph{df}) that checks the types of the elements of the stream and also guarantees 
 that the number of elements in the stream matches the number in the coeffect.
-
 The additional reduction rules mirror the semantics that we discussed earlier when talking about
@@
 e &=& \ldots \sep \ident{Df}\langle e_0, \ldots, e_n\rangle \sep \ident{prev}_{\cclrd{n}}~e\\
-C &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_
+C &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_ \sep \ident{Df}\langle v_0, \ldots, v_{i-1}, \_, e_{i+1} \ldots, e_n\rangle
 \end{array}  
@@
 \begin{theorem}[Well-typedness of the \kvd{prev} translation]
-\label{thm:semantics-welltyped}
+\label{thm:semantics-df-welltyped}
 Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$,
@@
 \vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{df}): $e = \ident{Df}\langle e_0, \ldots, e_n\rangle$. If $e_i$ is not a value
+  then reduce using (\emph{ctx}) with context $\ident{Df}\langle v_0, \ldots, v_{i-1}, \_, e_{i+1} \ldots, e_n\rangle$.
+  Otherwise, $e_0, \ldots, e_n$ are values and so $\ident{Df}\langle e_0, \ldots, e_n\rangle$
+  is also a value.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
 Case (\emph{split}, \emph{prev}): Similar. Either sub-expression is not a value, or the type 
@@
 
-::::TODO::::
-
-The types of the comonadically-inspired operations are the same for each concrete coeffect DSL, but
-each DSL introduces its own \emph{values} of type $\ctyp{\cclrd{r}}\tau$ and also its own reduction
-rules that define how comonadically-inspired operations evaluate.
-
-We first consider dataflow computations. As discussed earlier in the
-semantics of dataflow, the indexed comonad for a context with $n$ past values carries $n+1$ 
-values. When reducing translated programs, the comonadic values will not be directly manipulated by the
-user code. In a programming language, it could be seen as an \emph{abstract data type} whose only
-operations are the comonadically-inspired ones defined earlier.
-
-The Figure~\ref{fig:semantics-ext-df} shows extensions to the target language for modelling
-dataflow computations. We introduce a new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
-and a matching kind of expressions. We add a language primitive modelling the semantics of the
-\kvd{prev} keyword together with its typing rule (\emph{prev}) and we also add
-a typing rule (\emph{df}) that checks the types of the elements of the stream and also guarantees 
-that the number of elements in the stream matches the number in the coeffect.
-
-The additional reduction rules mirror the semantics that we discussed earlier when talking about
-indexed dataflow comonad.
-
-\paragraph{Properties.}
+As discussed earlier (Example~\ref{thm:semantics-indexed-prod}), implicit parameters can be modelled 
+by an indexed product comonad, which annotates a value with additional context -- in our case, a 
+mapping from implicit parameter names to their values. In this section, we embed this model into
+the target language.
+
+As with dataflow computations, we take the core functional subset (Figure~\ref{fig:semantics-target}) 
+with comonadically-inspired extensions (Figure~\ref{fig:semantics-ext}) and we specify a new kind
+of values of type $\ctyp{\cclrd{r}}\tau$ and domain-specific reduction rules that specify how
+the operations propagate and access the context containing implicit parameter bindings.
+Again, the $\ctyp{\cclrd{r}}\tau$ values can be seen as an \emph{abstract data type}, which are 
+never manipulated directly.
+
+\paragraph{Domain-specific extensions.}
+
+The Figure~\ref{fig:semantics-ext-impl} shows extensions to the target language for modelling
+implicit parameters. A comonadic value with a coeffect $\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}$
+is modelled by a new kind of value written as 
+$\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots \ident{?p}_n \mapsto v_n \})$ which contains
+a value $v$ together with implicit parameter assignments for all the parameters specified in the
+coeffect. We add a corresponding kind of expression with its typing rule (\emph{impl}).
+
+There are also two domain-specific operations for working with implicit parameters. The
+$\ident{lookup}_{\cclrd{\ident{?p}}}$ operation reads a value of an implicit parameter and
+the $\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}$ operation adds a mapping assigning a value
+to an implicit parameter \ident{?p}. The typing rule (\emph{lookup}) specifies that the accessed
+parameter need to be a part of the context and the rule (\emph{letimpl}) specifies that the
+\ident{letimpl} operation extends the context with a new implicit parameter binding.
+
+The new translation rules specify how implicit parameter access, written as \ident{?p}, and
+implicit parameter binding, written as $\kvd{let}~\ident{?p}~=e_1~\kvd{in}~e_2$ are translated to the
+target language. The first one is straightforward. The binding is similar to the translation for
+function application -- we split the context, evaluate $e_1$ using the first part of the context
+$\ctx_1$ and then add the new binding to the remaining context $\ctx_2$.
+
+Finally, Figure~\ref{fig:semantics-ext-impl} also defines the reduction rules. The (\emph{lookup})
+rule accesses an implicit parameter and (\emph{letimpl}) adds a new binding. The reduction rules 
+for comonadically-inspired primitives closely model the product comonad discussed in 
+Example~\ref{thm:semantics-indexed-prod}. Reductions for (\emph{cobind}) and (\emph{split}) 
+restrict the set of available implicit parameters according to the annotations and 
+(\emph{merge}) combines them, preferring the values from the call-site.
+
+For the semantics of implicit parameter programs that we consider, the preference of call-site
+bindings over declaration-site bindings in (\emph{merge}) does not matter. The unique typing 
+derivations for implicit parameter coeffects obtained in Section~\ref{sec:flat-unique} always
+split implicit parameters into \emph{disjoint sets}.
 
@@
 \begin{array}{rcl}
-v &=& \ldots \sep \ident{Impl}(v, \langle \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n\rangle) \\
-e &=& \ldots \sep \ident{Impl}(e, \langle\ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n\rangle) \\
+v &=& \ldots \sep \ident{Impl}(v, \{ \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n\}) \\
+e &=& \ldots \sep \ident{Impl}(e, \{\ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n\}) \\
   && \hspace{1.4em} \sep \ident{lookup}_{\cclrd{\ident{?p}}}~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~e_1~e_2\\
-C &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_
-\end{array}  
-\end{equation*}
-
-\vspace{1.25em}
+C &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_ \\
+  && \hspace{1.4em} \sep \ident{Impl}(\_, \{ \ident{?p}_1\mapsto e_1, .., \ident{?p}_n\mapsto e_n\})\\
+  && \hspace{1.4em} \sep \ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, .., \ident{?p}_{i-1} \mapsto v_{i-1}, \ident{?p}_i \mapsto \_, \ident{?p}_{i+1} \mapsto v_{i+1}, .. \ident{?p}_n \mapsto e_n\})
+\end{array}  
+\end{equation*}
+
+\vspace{1em}
 \paragraph{Typing rules}
@@
 \begin{equation*}
+  \hspace{-2em}
 \tyrule{impl}
   { \Gamma \vdash e:\tau & \forall i\in\{ 1 \ldots n \}. ~ \Gamma \vdash e_i : \ident{num}}
-  { \Gamma \vdash \ident{Impl}(e, \langle \ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n \rangle) : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \}}}{\tau} }
-\end{equation*}
-\begin{equation*}
+  { \Gamma \vdash \ident{Impl}(e, \{ \ident{?p}_1\mapsto e_1, \ldots, \ident{?p}_n\mapsto e_n \}) : \ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \}}}{\tau} }
+\end{equation*}
+\begin{equation*}
+  \hspace{-2em}
 \tyrule{lookup}
-  { \Gamma \vdash e : \ctyp{\cclrd{ \{ \ldots, \ident{?p}, \ldots \} }}{\tau} }
+  { \Gamma \vdash e : \ctyp{\cclrd{ \{\ident{?p}\} }}{\tau} }
   { \Gamma \vdash \ident{lookup}_{\cclrd{\ident{?p}}}~e : \ident{num} }
@@
 \begin{equation*}
+  \hspace{-2em}
 \tyrule{letimpl}
@@
 
-\vspace{1.25em}
+\vspace{1em}
 \paragraph{Translation}
@@
 \begin{equation*}
+\begin{array}{rl}  
+(\footnotesize{\emph{lookup}}) &
 \semdef
-  { \coctx{\Gamma}{\cclrd{ \{ \ident{?p} \} }} \vdash \ident{?p} : \ident{num} }
+  { \coctx{\Gamma}{\cclrd{ \{\ident{?p}\} }} \vdash \ident{?p} : \ident{num} }
   {\lambda ctx.\ident{lookup}_{\cclrd{\ident{?p}}}~\ctx }
-\end{equation*}
-
-\begin{equation*}
+\\[2em]
+\begin{array}{l}(\footnotesize{\emph{letimpl}})\\[1.25em]~\end{array}&
 \semdefff
@@
         \quad \vdash \kvd{let}~\ident{?p}=e_1\\[-0.3em]
-        \quyad\qquad \kvd{in}~e_2 : \tau_2 \\[-0.3em]
+        \quad\qquad \kvd{in}~e_2 : \tau_2 \\[-0.3em]
         ~ \\[-0.3em]
@@
   \lambda\ctx.~~\\[-0.3em]
-    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cclrd{\cup}\; \cclrd{(s\setminus{\ident{?p}})}}~\ident{dup}~\ctx\\[-0.3em]
-    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{(s\setminus{\ident{?p}})}}~\ctx_0\\[-0.3em]
-    \quad g~(\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{(s\setminus{\ident{?p}})}}~(f~\ctx_1)~\ctx_2)
+    \quad \kvd{let}~\ctx_0 = \ident{map}_{\cclrd{r} \;\cclrd{\cup}\; \cclrd{(s\setminus\{\ident{?p}\})}}~\ident{dup}~\ctx\\[-0.3em]
+    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\cclrd{r}, \cclrd{(s\setminus\{\ident{?p}\})}}~\ctx_0\\[-0.3em]
+    \quad g~(\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{(s\setminus\{\ident{?p}\})}}~(f~\ctx_1)~\ctx_2)
   \end{array} }   
-\end{equation*}
-
-\vspace{1.25em}
+\end{array}
+\end{equation*}
+
+\vspace{1em}
 \paragraph{Reduction rules}
-
-::: TODO ::: Everything except for counit
 
@@
 {\footnotesize(\emph{counit})} & \begin{array}{l}
-\ident{counit}_{\cclrd{ \emptyset }}~\ident{Impl} (v, \langle \ldots \rangle)   \rightarrow v
+\ident{counit}_{\cclrd{ \emptyset }}~(\ident{Impl} (v, \ldots))   \rightarrow v
 \end{array}
@@
 {\footnotesize(\emph{cobind})} & \begin{array}{l}
-\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
+\ident{cobind}_{\cclrd{r}, \cclrd{s}}~f~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})) \rightarrow
 \\[-0.25em]
-\quad (\ident{Df}\langle f (\ident{Df}\langle v_0, \ldots, v_m \rangle), \ldots, f (\ident{Df}\langle v_{n}, \ldots, v_{m+n} \rangle) \rangle)
-\end{array}
+\quad \ident{Impl}(f~(\ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} )), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{s} \} )
+\end{array}
+\\[1em]
+{\footnotesize(\emph{merge})} & \begin{array}{l}
+\ident{merge}_{\cclrd{r}, \cclrd{s}} 
+              (~\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})  \\[-0.25em]
+\hspace{4.4em}    \ident{Impl}(v', \{ \ident{?p}'_1 \mapsto v'_1, \ldots, \ident{?p}'_{n}\mapsto v'_n \})~) \rightarrow \\[-0.25em]
+\quad \ident{Impl}((v, v'), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}\in\cclrd{r \setminus s}  \} \cup \{ \ident{?p}'_i \mapsto v'_i \;|\; \ident{?p}'\in\cclrd{s}  \})
+\end{array}
+\\[1.75em]
+{\footnotesize(\emph{split})} & \begin{array}{l}
+\ident{split}_{\cclrd{r}, \cclrd{s}} (\ident{Impl}((v,v'), \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightarrow
+\\[-0.25em]
+\quad \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} ), \ident{Impl}(v', \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{s} \} )
+\end{array}
+\\[1em]
+{\footnotesize(\emph{letimpl})} & \begin{array}{l}
+\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v'~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightarrow
+\\[-0.25em]
+\quad \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i\;|\; \ident{?p}_i \in \cclrd{r}, \ident{?p}_i \neq \ident{?p} \} \cup \{ \ident{?p} \mapsto v' \} )
+\end{array}
+\\[1em]
+{\footnotesize(\emph{lookup})} & \begin{array}{l}
+\ident{lookup}_{\cclrd{\ident{?p}_i}} (\ident{Impl}(v, \{\ident{?p}_i\mapsto v_i\}) \rightarrow v_i
+\end{array}
+\end{array}
+\end{equation*}
+
+\vspace{-0.5em}
+\caption{Additional constructs embedding implicit parameters into the language}
+\label{fig:semantics-ext-impl}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\newpage
+~
+\newpage
+\paragraph{Properties.}
+
+The core functional language subset (Figure~\ref{fig:semantics-target}) with the comonadically-inspired 
+extensions (Figure~\ref{fig:semantics-ext}) and the domain-specific extensions for implicit 
+parameters defined in Figure~\ref{fig:semantics-ext-impl} is safe. Furthermore, the 
+translation from coeffect calculus produces well-typed programs. Together, these two properties 
+also guarantee safety of the source coeffect calculus, when used with implicit parameters.
+
+As for dataflow computations, we prove safety by extending the preservation
+(Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog})
+for the core functional subset of the language, but it is worth noting that the key parts of
+the proofs are centered around the new reduction rules for comonadically-inspired primitives
+and newly defined \ident{Impl} values. These do not interact with the rest of the language in 
+any unexpected ways.
+
+\begin{theorem}[Well-typedness of the implicit parameters translation]
+\label{thm:semantics-impl-welltyped}
+Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$,
+the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
+and Figure~\ref{fig:semantics-ext-impl} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
+\end{theorem}
+\begin{proof}
+By rule induction over the derivation of the translation.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{var, num, abs, app}): As before.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{lookup}): The type of $\ctx$ has a coeffect  $\cclrd{ \{\ident{?p}\} }$
+  which includes the parameter \ident{?p} as required in order to use the (\emph{lookup}) typing rule.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm 
+Case (\emph{letimpl}): The type of $\ctx$ matches with the input type of
+  $\ident{map}_{  \cclrd{r \cup (s\setminus\{\ident{?p}\})} }$. After duplication and splitting
+  the context, $\ctx_1$ and $\ctx_2$ have types $\ctyp{\cclrd{r}}(\ldots)$ and
+  $\ctyp{\cclrd{s\setminus\{\ident{?p}\}}}(\ldots)$, respectively. This matches with the expected
+  types of $\ident{f}$ and $\ident{letimpl}$. The context returned by $\ident{letimpl}$ then 
+  matches the one required by $\ident{g}$.
+\end{proof}
+
+\begin{lemma}[Canonical forms]
+\label{thm:semantics-impl-canon}
+For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
+\begin{enumerate}
+  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
+  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
+  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
+  \item If $\tau=\ctyp{\cclrd{ \{ \ident{?p}_1, \ldots, \ident{?p}_n \} }}\tau_1$ then $e = \ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_n \mapsto v_n \} )$
+\end{enumerate}  
+\end{lemma}
+\begin{proof}
+  (1,2,3) as before; for (4) the last typing rule must have been (\emph{impl}).
+\end{proof}
+
+\begin{lemma}[Preservation under substitution]
+\label{thm:semantics-impl-pres-subst}
+For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$ 
+then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $\Gamma, x:\tau \vdash e : \tau'$ as before, with new
+  cases for $\ident{Impl}(e, \{ \ldots \})$, $\ident{lookup}_{\cclrd{\ident{?p}}}$ and 
+  $\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}$.
+\end{proof}
+
+\begin{theorem}[Type preservation]
+\label{thm:semantics-impl-pres}
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
+\end{theorem}
+\begin{proof}
+  Rule induction over $\rightarrow$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{fn}, \emph{prj}, \emph{ctx}): As before, using Lemma~\ref{thm:semantics-impl-pres-subst} for (\emph{fn}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{counit}): $e = \ident{counit}_{\cclrd{0}}(\ident{Impl}(v, \{\;\}))$. The last rule in the type
+  derivation of $e$ must have been (\emph{counit}) with $\Gamma \vdash \ident{Impl}(v, \{ \; \}) : \ctyp{\cclrd{\emptyset}}\tau$
+  which, in turn, required that $\Gamma \vdash v : \tau$.
+
+\newpage
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f~(\ident{Impl}(v, \{ \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n  \}))$.  
+  The last rule in the type derivation of $e$ must have been (\emph{cobind}) with a type $\tau = \ctyp{\cclrd{s}}\tau_2$ and assumptions
+  $\Gamma \vdash \ident{Impl}(v, \{ \ident{?p}_1\mapsto v_1, \ldots, \ident{?p}_n\mapsto v_n  \}) : \ctyp{\cclrd{r}}{\tau}$ and
+  $\Gamma \vdash f : \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$.
+  The reduced expression has a type $\ctyp{\cclrd{s}}\tau_2$:
+\begin{equation*}
+\inference
+  { \inference 
+      { \Gamma \vdash f : \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2 & 
+        \Gamma \vdash \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} ) :  \ctyp{\cclrd{r}}{\tau_1}}  
+      { \Gamma \vdash f~(\ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} )) : \tau_2} }
+  { \Gamma \vdash \ident{Impl}(f~(\ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{r} \} )), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{p}_i \in \cclrd{s} \}) : \ctyp{\cclrd{s}}\tau_2 }
+\end{equation*}
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{lookup}): $e = \ident{lookup}_{\cclrd{\ident{?p}_i}}(\ident{Impl}(v, \{\; \ldots, \ident{?p}_i \mapsto v_i \ldots \}))$. The last rule in the type
+  derivation must have been (\emph{lookup}) with $\tau = \ident{num}$ and an assumption $\Gamma\vdash\ident{Impl}(v, \{\; \ldots, \ident{?p}_i\mapsto v_i \ldots \}) : \ctyp{\cclrd{ \{\ldots,\ident{?p}_i,\ldots\} }}{\tau}$,
+  which requires $\Gamma \vdash v_i : \ident{num}$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{merge}, \emph{split}, \emph{letimpl}): Similar. In all three cases, the last typing rule in the derivation of $e$
+  guarantees that all values of all implicit parameters that are required for the reduction are available.
+\end{proof}
+
+\begin{theorem}[Progress]
+\label{thm:semantics-impl-prog}  
+  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+\end{theorem}
+\begin{proof}
+  By rule induction over $\vdash$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{num,abs,var,app,proj,tup}): As before, using the adapted canonical forms lemma
+  (Lemma~\ref{thm:semantics-impl-canon}) for (\emph{app}) and (\emph{proj}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{counit}): $e = {\ident{counit}_\cunit}~e_1$. If $e_1$ is not a value, it can be reduced 
+  using (\emph{ctx}) with context ${\ident{counit}_\cunit}~\_$, otherwise it is a value. From Lemma~\ref{thm:semantics-df-canon},
+  $e_1 = \ident{Impl}(v, \{\;\})$ and so we can apply (\emph{counit}) reduction rule.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2$. If $e_1$ is not a value,
+  reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{r},\cclrd{s}}}~\_~e$. If $e_2$ is
+  not a value reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{r},\cclrd{s}}}~v~\_$.
+  If both are values, then from Lemma~\ref{thm:semantics-impl-canon}, we have 
+  $e_2 = \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}_i \in \cclrd{r}\cup\cclrd{s} \})$ 
+  and we apply the (\emph{cobind}) reduction.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{merge}): $e = \ident{merge}_{\cclrd{r}, \cclrd{s}} e_1$. If $e_1$ is not a value, 
+  reduce using (\emph{ctx}) with context $e = \ident{merge}_{\cclrd{r}, \cclrd{s}}~\_$. If $e_1$ is
+  a value, it must be a pair of values $(   
+    \ident{Impl}(v, \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}_i \in\cclrd{r} \}),
+    \ident{Impl}(v', \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}_i \in\cclrd{s} \}))$
+  using Lemma~\ref{thm:semantics-df-canon} and it can reduce using (\emph{merge}) reduction.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{impl}): $e = \ident{Impl}(e', \{ \ident{?p}_1 \mapsto e_1, \ldots, \ident{?p}_n \mapsto e_n\})$. 
+  If $e$ is not a value, reduce using (\emph{ctx}) with context
+  $\ident{Impl}(\_, \{ \ident{?p}_1 \mapsto e_1, \ldots, \ident{?p}_n \mapsto e_n\})$. 
+  If $e_i$ is not a value, reduce using (\emph{ctx}) with context 
+  $\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{i-1} \mapsto v_{i-1}, \ident{?p}_i \mapsto \_, \ident{?p}_{i+1} \mapsto v_{i+1}, \ldots \ident{?p}_n \mapsto e_n\})$.
+  Otherwise, $e, e_0, \ldots, e_n$ are values and so 
+  $\ident{Impl}(e', \{ \ident{?p}_1 \mapsto e_1, \ldots, \ident{?p}_n \mapsto e_n\})$
+  is also a value.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm 
+Case (\emph{split}, \emph{letimpl}): Similar. Either sub-expression is not a value, 
+  or the type guarantees that it is a comonadic value with implicit parameter bindings
+  that enable the (\emph{split}) or (\emph{letimpl}) reduction, respectively.
+\end{proof}
+
+\begin{theorem}[Safety]
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow^* e'$ then either $e'$ is a value or there
+  exists $e''$ such that $e' \rightarrow e''$
+\end{theorem}
+\begin{proof}
+  Rule induction over $\rightarrow^*$ using Theorem~\ref{thm:semantics-impl-pres} and Theorem~\ref{thm:semantics-impl-prog}.
+\end{proof}
+
+\newpage
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Generalised safety of comonadic embedding}
+\label{sec:semantics-generalising}
+
+In Section~\ref{sec:semantics-proofs-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
+safety property of two concrete context-aware programming langauges based on the coeffect 
+language framework. The proofs for the two systems were very similar and they relied on the
+same key principle.
+
+The principle is that the coeffect annotation $\cclrd{r}$ on the type modelling 
+the indexed comonad structure $\ctyp{\cclrd{r}}\tau$ in the target language guarantees
+that the comonadic value will provide the necessary context. As a result the reductions for
+operations accessing the context do not get stuck. In case of dataflow, $\ident{prev}_\cclrd{n}$ 
+can always access the tail of the stream and and $\ident{counit}_\cunit$ can always access the 
+head (because the stream has a sufficient number of elements). In case of implicit parameters, 
+the context passed to $\ident{lookup}_{\cclrd{\ident{?p}}}$ will always contain a binding for
+$\ident{?p}$.
+
+Our core functional target language is not expressive enough to capture the relationship between
+the coeffect annotation and the structure of the \ident{Df} or \ident{Impl} value and so we
+resorted to adding those as ad-hoc extensions. However, given a target language with a sufficiently 
+expressive type system, the properties proved in Section~\ref{sec:semantics-proofs} would be
+guaranteed directly by the target language. This includes dependently-typed languages such as
+Idris or Agda \cite{other-idris,other-agda}, but type-level numerals and sets can also be encoded
+in the Haskell type system \cite{effects-embedding}.
+
+In other words, the flat coeffect type system, together with the translation for introduced in this 
+chapter, can be embedded into a Haskell-like languages and it can provide a succinct and safe way 
+of implementing context-aware domain specific languages.
+
+\paragraph{Liveness.}
+For example, if we wanted to follow the development in the previous section for liveness, we would
+extend the target language with two kinds of expressions, \ident{Dead} representing a dead context 
+with no value and \ident{Live} representing a context with a value:
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &=& \ldots \sep \ident{Dead} \sep \ident{Live}~e \\
+\end{array}  
+\end{equation*}
+%
+The typing rules promote the information about whether a value is available into the type-level 
+and so a context carrying a live value is marked as $\ctyp{\cclrd{\ident{L}}}{\tau}$ while a
+dead context has a type $\ctyp{\cclrd{\ident{D}}}{\tau}$.
+%
+\begin{equation*}
+\tyrule{live}
+  { \Gamma \vdash e:\tau }
+  { \Gamma \vdash \ident{Live}~e : \ctyp{\cclrd{\ident{L}}}{\tau} }
+\qquad
+\tyrule{dead}
+  { ~ }
+  { \Gamma \vdash \ident{Dead} : \ctyp{\cclrd{\ident{D}}}{\tau} }
+\end{equation*}
+%
+The above example can be encoded using generalized algebraic data types (GADTs) in Haskell. We
+define types \ident{D} and \ident{L} to capture the type annotations and we define a type
+$\ident{C}~r~a$ where the type parameter $r$ represents a coeffect annotation:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{data}~\ident{L} \\[-0.2em]
+\kvd{data}~\ident{D} \\[0.5em]
+\kvd{data}~\ident{C}~r~a~\kvd{where} \\[-0.2em]
+\quad \ident{Live} \hspace{0.45em}:: a \rightarrow \ident{C}~\ident{L}~a\\[-0.2em]
+\quad \ident{Dead} :: \ident{C}~\ident{D}~a
+\end{array}
+\end{equation*}
+%
+The operations of the flat coeffect algebra could then be encoded as type-level operations on
+the types $\ident{L}$ and $\ident{D}$ so that, for example $\ident{Seq}~a~b$ is $\ident{D}$ only
+if both $a$ and $b$ are $\ident{D}$. The types of the \ident{counit} and \ident{cobind} operations
+would encode the coeffect annotations as:
+
+\begin{equation*}
+\begin{array}{rcl}
+\ident{counit} &::& \ident{C}~\ident{L}~a \rightarrow a \\[-0.2em]
+\ident{cobind} &::& (\ident{C}~\ident{r}~a \rightarrow b) \rightarrow \ident{C}~(\ident{Seq}~r~s)~a \rightarrow \ident{C}~\ident{s}~b
+\end{array}
+\end{equation*}
+%
+Here, the additional type annotation ensures that \ident{counit} can only be called on a context
+that contains a value and so calling the operation is not going to fail. Similarly, the type of the
+\ident{cobind} operation now guarantees that if a function (used as the first argument) or the result
+require a live context, it will be called with a value $\ident{C}~\ident{L}~a$ that is guaranteed
+to contain a value.
+
+In our formalization, this is what guarantees that the reduction rules do 
+not get stuck. For liveness, those would look as:
+%
+\begin{equation*}
+\begin{array}{rl}
+{\footnotesize(\emph{counit})} &
+  \ident{counit}_{\cclrd{ \ident{L} }}~(\ident{Just}~v)   \rightarrow v
 \\[0.5em]
-{\footnotesize(\emph{merge})} & \begin{array}{l}
-\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle v'_0, \ldots, v'_n\rangle)) \rightarrow
-\\[-0.25em]
-\quad (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{min}(m,n)}, v'_{\mathit{min}(m,n)}) \rangle)
-\end{array}
-\\[0.5em]
-{\footnotesize(\emph{split})} & \begin{array}{l}
-\ident{split}_{\cclrd{m}, \cclrd{n}} (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle)  \rightarrow
-\\[-0.25em]
-\quad \ident{Df}\langle v_0, \ldots, v_m \rangle, (\ident{Df}\langle b_0, \ldots, b_n\rangle
-\end{array}
-\\[0.5em]
-{\footnotesize(\emph{prev})} & \begin{array}{l}
-\ident{prev}_{\cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_n, v_{n+1} \rangle) \rightarrow
-\\[-0.25em]
-\quad \ident{Df}\langle v_0, \ldots, v_n \rangle
-\end{array}
-\end{array}
-\end{equation*}
-
-\caption{Additional constructs embedding implicit parameters into the language}
-\label{fig:semantics-ext-df}
-\end{figure}
-\newpage
-
-~
-
-\newpage
-
-\section{Generalising}
-\label{sec:semantics-generalising}
-... works for any good comonad...
-
-\newpage
-\newpage
+{\footnotesize(\emph{cobind-1})} &
+  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~v)
+\\
+{\footnotesize(\emph{cobind-2})} &
+  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~\ident{Dead})
+\\
+{\footnotesize(\emph{cobind-3})} &
+  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
+\\
+{\footnotesize(\emph{cobind-4})} &
+  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
+\end{array}
+\end{equation*}
+%
+If we encoded the rules as functions in a dependently-typed language or in Haskell, the type
+preservation property would be guaranteed by the type system of the target language. The 
+equivalent of the progress property would then be guaranteed by the fact that the the operations
+are always defined.
+
+It is worth noting that this is where coeffects need a target language with a more expressive
+type system than monads. For monadic computations, it is sufficient to use a type $\ident{M}~a$
+which represents that \emph{some} effects may happen. The the type does not specify which effects
+and, indeed, this means that \emph{all possible effects} may happen.
+
+With coeffects, we need indexed comonads $\ident{C}~r~a$ where the annotation $r$ specifies what
+context may be required. Without the annotation, a type $\ident{C}~a$ would represent a comonadic
+context that has \emph{all possible context} available, which is rarely useful in practice.
 

### 2016-02-14 main.tex
Finish edits (Chapter 5)

@@
   and \emph{structural} coeffect calculi that capture a wide range of context-aware programming
-  languages. We discuss their type systems and semantics together with safety properties.}
+  languages discussed in the previous section. We present parameterized coeffect type system, 
+  categorical model based on indexed comonads. We use the model to give semantics in terms of 
+  translation to a target functional language and we present syntactic safety proof for 
+  context-aware languages with implicit parameters and dataflow.}
 \setcounter{part}{1}

### 2016-02-14 text/Chapter05.tex
Finish edits (Chapter 5)

@@
 
-\paragraph{Liveness.}
-For example, if we wanted to follow the development in the previous section for liveness, we would
-extend the target language with two kinds of expressions, \ident{Dead} representing a dead context 
-with no value and \ident{Live} representing a context with a value:
+\paragraph{Coeffects for liveness.}
+As an example, we consider the third instance of coeffect calculus that was discussed in 
+Chapter~\ref{ch:flat}. If we wanted to follow the development in the previous section for liveness, 
+we would extend the target language with two kinds of expressions, \ident{Dead} representing a dead
+context with no value and \ident{Live} representing a context with a value:
 %
@@
 %
-The above example can be encoded using generalized algebraic data types (GADTs) in Haskell. We
-define types \ident{D} and \ident{L} to capture the type annotations and we define a type
-$\ident{C}~r~a$ where the type parameter $r$ represents a coeffect annotation:
-%
-\begin{equation*}
-\begin{array}{l}
-\kvd{data}~\ident{L} \\[-0.2em]
-\kvd{data}~\ident{D} \\[0.5em]
-\kvd{data}~\ident{C}~r~a~\kvd{where} \\[-0.2em]
-\quad \ident{Live} \hspace{0.45em}:: a \rightarrow \ident{C}~\ident{L}~a\\[-0.2em]
-\quad \ident{Dead} :: \ident{C}~\ident{D}~a
-\end{array}
-\end{equation*}
-%
-The operations of the flat coeffect algebra could then be encoded as type-level operations on
-the types $\ident{L}$ and $\ident{D}$ so that, for example $\ident{Seq}~a~b$ is $\ident{D}$ only
-if both $a$ and $b$ are $\ident{D}$. The types of the \ident{counit} and \ident{cobind} operations
-would encode the coeffect annotations as:
-
-\begin{equation*}
-\begin{array}{rcl}
-\ident{counit} &::& \ident{C}~\ident{L}~a \rightarrow a \\[-0.2em]
-\ident{cobind} &::& (\ident{C}~\ident{r}~a \rightarrow b) \rightarrow \ident{C}~(\ident{Seq}~r~s)~a \rightarrow \ident{C}~\ident{s}~b
-\end{array}
-\end{equation*}
-%
-Here, the additional type annotation ensures that \ident{counit} can only be called on a context
-that contains a value and so calling the operation is not going to fail. Similarly, the type of the
-\ident{cobind} operation now guarantees that if a function (used as the first argument) or the result
-require a live context, it will be called with a value $\ident{C}~\ident{L}~a$ that is guaranteed
-to contain a value.
-
-In our formalization, this is what guarantees that the reduction rules do 
-not get stuck. For liveness, those would look as:
+Finally, we need to add reduction rules that define the meaning of the comonadically-inspired
+operations for liveness. Those follow the definitions given in Example~\ref{thm:semantics-indexed-opt}
+when discussing the categorical semantics:
 %
@@
 %
-If we encoded the rules as functions in a dependently-typed language or in Haskell, the type
-preservation property would be guaranteed by the type system of the target language. The 
-equivalent of the progress property would then be guaranteed by the fact that the the operations
-are always defined.
+This language extension is safe because the reductions respect the typing of the
+comonadically-inspired operations. The $\ident{counit}_\cunit$ reduction does not get stuck
+for well-typed terms because $\cunit = \cclrd{\ident{L}}$ in the coeffect algebra and thus its 
+argument is of type $\ctyp{\cclrd{\ident{L}}}{\tau}$ and will always be a value $\ident{Live}~v$.
+
+Similarly, the when reducing $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$, the typing ensures that
+the value passed as the second argument is of type $\ctyp{\cclrd{r}\cseq\cclrd{s}}{\tau_1}$.
+In case of liveness, $\cclrd{r}\,\cseq\,\cclrd{s}=\cclrd{\ident{L}}$ if either
+$\cclrd{r}=\cclrd{\ident{L}}$ or $\cclrd{s}=\cclrd{\ident{L}}$. This means that reductions
+(\emph{cobind-1}) and (\emph{cobind-2}) will not get stuck because the value will be
+$\ident{Live}~v$ and not \ident{Dead}. The reduction rules also preserve typing - the resulting
+value is of type $\ctyp{\cclrd{s}}\tau_2$, that is $\ident{Live}~v$ for (\emph{cobind-1}),
+(\emph{cobind-2}) and \ident{Dead} for (\emph{cobind-3}) and (\emph{cobind-4}).
+
+\paragraph{Encoding liveness in Haskell.}
+The liveness example can be encoded in Haskell using type-level features such as generalized
+algebraic data types (GADTs) and type families \cite{haskell-faking,haskell-promotion,haskell-families}.
+We do not aim to give a complete implementation, but to show that such encoding is possible and
+would provide the necessary safety guarantees.
+
+We first define types \ident{D} and \ident{L} to capture the coeffect annotations. 
+Then we define a comonadically-inspired type $\ident{C}~r~a$ as a GADT with cases for
+\ident{Live} and \ident{Dead} contexts. The type parameter $r$ represents a coeffect 
+annotation:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{data}~\ident{L} \\[-0.2em]
+\kvd{data}~\ident{D} \\[0.5em]
+\kvd{data}~\ident{C}~r~a~\kvd{where} \\[-0.2em]
+\quad \ident{Live} \hspace{0.45em}:: a \rightarrow \ident{C}~\ident{L}~a\\[-0.2em]
+\quad \ident{Dead} :: \ident{C}~\ident{D}~a
+\end{array}
+\end{equation*}
+%
+The definition matches with the typing rules (\emph{live}) and (\emph{dead}). The coeffect 
+annotation for a live value is \ident{L} and the annotation for a dead value is \ident{D}.
+To give the type of \ident{cobind}, we need a type-level function that encodes the operations
+of the flat coeffect algebra. We model $\cseq$ as $\ident{Seq}~a~b$. The operation is defined
+on types $\ident{L}$ and $\ident{D}$ and returns a type $\ident{D}$ if and only if both its
+arguments are $\ident{D}$:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{type}~\kvd{family}~\ident{Seq}~r~s :: *\\[-0.2em]
+\quad \kvd{type}~\kvd{instance}~\ident{Seq}~\ident{D}~\ident{D} = \ident{D}\\[-0.2em]
+\quad \kvd{type}~\kvd{instance}~\ident{Seq}~\ident{L}~s = \ident{L}\\[-0.2em]
+\quad \kvd{type}~\kvd{instance}~\ident{Seq}~r~\ident{L} = \ident{L}
+\end{array}
+\end{equation*}
+%
+The \ident{counit} and \ident{cobind} operations can then be defined as Haskell functions
+that have types corresponding to the typing rules (\emph{counit}) and (\emph{cobind}) given
+in Figure~\ref{fig:semantics-ext}:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\ident{counit} &::& \ident{C}~\ident{L}~a \rightarrow a \\[-0.2em]
+\ident{cobind} &::& (\ident{C}~\ident{r}~a \rightarrow b) \rightarrow \ident{C}~(\ident{Seq}~r~s)~a \rightarrow \ident{C}~\ident{s}~b
+\end{array}
+\end{equation*}
+%
+Here, the additional type annotation ensures that \ident{counit} can only be called on a context
+that contains a value and so calling the operation is not going to fail. Similarly, the type of the
+\ident{cobind} operation now guarantees that if a function (used as the first argument) or the result
+require a live context, it will be called with a value $\ident{C}~\ident{L}~a$ that is guaranteed
+to contain a value.
+
+\paragraph{Coeffects in dependently-typed languages.}
+If we use the above encoding, type preservation is guaranteed by the type 
+system of the target language. The equivalent of the progress property is guaranteed by the 
+fact that the the implementation of the operations is well-defined.
 
@@
 
-\section{Related and futrue work}
+\section{Related categorical structures}
 \label{sec:semantics-related}
 
-Most of the related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
-and we covered work related to individual concepts throughout the chapter. In this section, we
-do not repeat the discussion present elsewhere. Instead, we discuss one specific question that often 
-arises when discussing coeffects and that is \emph{when is a coeffect (not) an effect?}
-
-We start with a quick overview of the ways in which effects and coeffects differ and then
-we briefly look at one (but illustrative) example where the two concepts overlap. We focus 
-mainly on the equivalence between the \emph{categorical semantics}, which reveals the nature
-of the computations -- rather than considering just the syntactic aspects of the type system.
-
-\subsection{Properties and related notions}
-
-The flat indexed comonad structure is all we need to give the semantics of the flat coeffect 
-calculus. Before doing so in Section~\ref{sec:flat-semantics-calculus}, we briefly consider 
-additional properties and other categorical structures that have been proposed mainly in the 
-context of monads and effects and we look how they relate to indexed comonads.
-
-\paragraph{Shape preservation.}
+Related work leading to coeffects has already been discussed in Chapter~\ref{ch:pathways} 
+and we covered work related to individual concepts throughout the thesis. However, there is a
+number of related categorical structures that are related to our \emph{indexed comonads}
+(Section~\ref{sec:semantics-flat-idx}) that deserve additional discussion.
+
+In Section~\ref{sec:semantics-related-related}, we discuss related approaches to adding 
+indices to categorical structures (mostly monads). In Section~\ref{sec:semantics-related-monad},
+we discuss a question that often arises when discussing coeffects and that is 
+\emph{when is a coeffect (not) an effect?}
+
+\subsection{Indexed categorical structures}
+\label{sec:semantics-related-related}
+
 Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
@@
 
-~
-
 Indexed comonads are not restricted by this property of comonads. For example, given the indexed
@@
 just implicit parameters $\cclrd{s}$. 
+Thus \emph{indexed comonads} are a generalization of \emph{comonads} that captures structures that
+do not form a (non-indexed) comonad. However, indexing has been also discussed in the case of
+\emph{monads.}
 
@@
 \subsection{When is coeffect not a monad}
+\label{sec:semantics-related-monad}
+
 Coeffect systems differ from effect systems in three important ways:
@@
 rule, we can restrict the first parameter, which corresponds to the declaration site context:
-%
+
 \begin{equation*}
@@
 
+% ==================================================================================================
+%
+%     #####
+%    #     # #    # #    # #    #   ##   #####  #   #
+%    #       #    # ##  ## ##  ##  #  #  #    #  # #
+%     #####  #    # # ## # # ## # #    # #    #   #
+%          # #    # #    # #    # ###### #####    #
+%    #     # #    # #    # #    # #    # #   #    #
+%     #####   ####  #    # #    # #    # #    #   #
+%
+% ==================================================================================================
+
 \section{SUmmary}
 
-Next, we introduced the notion of \emph{flat indexed comonad}, which generalizes of comonads
+In the previous chapter, we defined \emph{type system} for flat coeffect calculi that uniformly
+captures the shared structure of context-aware computations. In this chapter, we completed
+the unification by providing \emph{semantics} for flat coeffect calculi and proving the
+\emph{safety} of coeffect languages for dataflow and implicit parameters. The semantics shown
+here also guides the implementation that is discussed later in Chapter~\ref{ch:impl}.
+
+The development presented in this chapter follows the well-known example of effects and monads.
+We introduced the notion of \emph{indexed comonad}, which generalizes comonads
 and adds additional operations needed to provide categorical semantics of the flat coeffect 
-calculus. The indices of the flat indexed comonad operations correspond to the
-coeffect annotations in the type system and provide a foundation for the design of the 
-calculus.
-
-In the upcoming chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
+calculus and we demonstrated how implicit parameters, liveness and dataflow computations form
+indexed comonads.
+
+We then used the comonadic semantics to define a \emph{comonadically-inspired translation}
+that turns programs written in a domain-specific coeffect language into a functional target 
+language. This is akin to the Haskell do notation for monads. Finally, we extended the target
+language with concrete implementation of comonadic operations for dataflow and implicit parameters
+and we presented syntactic safety proof. In summary, the proof states that well-typed 
+context-aware programs written in a coeffect language \emph{do not go wrong} (when translated
+to a simple functional language and evaluated).
+
+The proof relies on the fact that coeffect annotations (provided by the coeffect type system)
+guarantee that the required context is available in the comonadic value that represents the 
+context and we also discussed how this would guarantee safety in languages with sufficiently
+expressive type system such as Haskell.
+
+In the following chapter, we move from \emph{flat} coeffect calculi, tracking whole-context
 properties to \emph{structural} coeffect calculi, tracking per-variable information, thus

### 2016-03-06 main.tex
Move stuff in chapter 4, improve introductions

--- 2016-02-03 main.tex
@@
 
-\ctparttext{This part presents the key novel contributions of the thesis. We develop \emph{flat}
-  and \emph{structural} coeffect calculi that capture a wide range of context-aware programming
-  languages. We discuss their type systems and semantics together with safety properties.}
+% ==================================================================================================
+
+\ctparttext{In this part, we capture the similarities between the concrete context-aware langauges
+  presented in the previous chapter. We also develop the key novel technical contributions of
+  the thesis. We define a \emph{flat coeffect type system} (Chapter~\ref{ch:flat}) that is 
+  parameterized by a \emph{coeffect algebra} and a mechanism for choosing unique typing derivation. 
+  We instantiate a coeffect type system with a concrete coeffect algebra and procedure for choosing 
+  unique typing derivation for three languages to capture dataflow, implicit parameters and liveness.
+
+  The type system is complemented with a translational semantics for coeffect-based context-aware
+  programming languages (Chapter~\ref{ch:semantics}). The semantics is inspired by a categorical
+  model based on \emph{indexed comonads} and it translates source context-aware program into a 
+  target program in a simple functional language with comonadically-inspired primitives. We give 
+  concrete definition of the primitives for dataflow, implicit parameters and liveness and 
+  present a syntactic safety proof for these three languages.
+
+  The following page provides a detailed overview of the content of Chapters~\ref{ch:flat} and
+  Chapters~\ref{ch:semantics}, highlighting the split between general definitions and properties
+  (about the coeffect calculus) and concrete definitions and properties (about concrete
+  context-awre language). The Chapter~\ref{ch:structural} mirrors the same development for 
+  \emph{structural coeffect systems}.
+}
 \setcounter{part}{1}
 \part{Coeffect calculi}
+
+\begin{center}
+\begin{tabular}{ | l | l | l |}
+\hline
+\multicolumn{3}{ | c | }{\spacedlowsmallcaps{Chapter 4}} \\ \hline \hline
+\hspace{7.1em} & \textsc{Coeffect calculus} \hspace{2.5em} & \textsc{Language-specific} \hspace{6.9em} \\ \hline
+\textsc{Syntax} 
+  & \hspace{-0.5em}\begin{tabular}{l} Coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:flat-calculus-syntax}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Extensions such as \\[-0.3em] \ident{?param} and \kvd{prev} \\[-0.3em] (Section~\ref{sec:flat-calculus-examples})  \end{tabular} \\ \hline
+\textsc{Type system} 
+  & \hspace{-0.5em}\begin{tabular}{l} Abstract coeffect \\[-0.3em] algebra (Section~\ref{sec:flat-calculus-algebra})  \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Concrete instances \\[-0.3em] of the coeffect algebra \\[-0.3em] (Section~\ref{sec:flat-calculus-examples})  \end{tabular} \\ 
+\hline
+  & \hspace{-0.5em}\begin{tabular}{l} Coeffect type system \\[-0.3em] parameterized by \\[-0.3em] the coeffect algebra \\[-0.3em] (Section~\ref{sec:flat-calculus-types})  \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Typing for language- \\[-0.3em] specific extensions \\[-0.3em] (Section~\ref{sec:flat-calculus-examples})  \end{tabular} \\ 
+\hline
+  & 
+  & \hspace{-0.5em}\begin{tabular}{l} Procedure for determining \\[-0.3em] a unique typing derivation \\[-0.3em]  (Section~\ref{sec:flat-unique})  \end{tabular} \\ \hline
+\textsc{Properties} 
+  & \hspace{-0.5em}\begin{tabular}{l} Syntactic properties \\[-0.3em] of coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:flat-syntax})  \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Uniqueness of the above \\[-0.3em]  (Section~\ref{sec:flat-unique})  \end{tabular} \\ \hline
+\end{tabular}
+\end{center}
+
+~
+
+\begin{center}
+\begin{tabular}{ | l | l | l |}
+\hline
+\multicolumn{3}{ | c | }{\spacedlowsmallcaps{Chapter 5}} \\ \hline \hline
+& \textsc{Coeffect calculus} & \textsc{Language-specific} \\ \hline
+\textsc{Categorical} 
+  & \hspace{-0.5em}\begin{tabular}{l} Indexed comonads \\[-0.3em] (Sectiuon~\ref{sec:semantics-flat-idx}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Examples including indexed \\[-0.3em] product, list and maybe \\[-0.3em] comonads (Section~\ref{sec:semantics-flat-monoidal}) \end{tabular} \\ 
+\hline
+  & \hspace{-0.5em}\begin{tabular}{l} Categorical semantics \\[-0.4em] of coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:semantics-flat-calculus}) \end{tabular}
+  & \\ \hline
+\textsc{Translational} 
+  & \hspace{-0.5em}\begin{tabular}{l} Functional target \\[-0.3em] language (Section~\ref{sec:semantics-translation-target}) \end{tabular}
+  &  \\ 
+\hline
+  & \hspace{-0.5em}\begin{tabular}{l} Translation from coeffect \\[-0.3em] $\lambda$-calculus to target \\[-0.3em]  language (Section~\ref{sec:semantics-translation-transl}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Translation for language-\\[-0.3em] specific extensions (\kvd{prev}, \ident{?p}) \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\ \hline
+\textsc{Operational} 
+  & \hspace{-0.5em}\begin{tabular}{l} Abstract comonadically- \\[-0.3em] inspired primitives \\[-0.3em] (Section~\ref{sec:semantics-translation-transl}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Concrete reduction rules for \\[-0.3em] comonadically-inspired primitives \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\
+\hline
+  & 
+  & \hspace{-0.5em}\begin{tabular}{l} Reduction rules for language-\\[-0.3em] specific extensions (\kvd{prev}, \ident{?p}) \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\
+\hline
+  & \hspace{-0.5em}\begin{tabular}{l} Sketch of generalized \\[-0.3em] syntactic soundness \\[-0.3em] (Section~\ref{sec:semantics-generalising}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Syntactic soundness \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\ \hline
+\end{tabular}
+\end{center}
+
+% ==================================================================================================
 

### 2016-03-06 text/Chapter04.tex
Move stuff in chapter 4, improve introductions

@@
 \label{ch:flat}
-
-Successful programming language abstractions need to generalize a wide range of recurring
-problems while capturing the key commonalities. These two aims are typically in opposition -- 
-more general abstractions are less powerful, while less general abstractions cannot be
-used as often.
 
@@
 useful properties that we can see when we consider the systems separately. For this reason, we 
-discuss \emph{flat coeffects} and \emph{structural coeffects} separately.
-
-In this and the next chapter, we discuss the type system and the semantics of flat coeffect 
-systems, respectively. In this chapter, we develop a parameterized type system for flat coeffect
-systems and we study its general syntactic properties. We also consider variations of the type system
-that resolve the ambiguity of coeffectful lambda abstraction for concrete instances of the system.
-In the next chapter, we give operational meaning of concrete coeffect languages using the unified
-system and we discuss their safety.
+discuss \emph{flat coeffects} (Chapter~\ref{ch:flat} and Chapter~\ref{ch:semantics}) and 
+\emph{structural coeffects} (Chapter~\ref{ch:structural}) separately.
+
+In this chapter, we develop a \emph{flat coeffect calculus} that provides a type system for 
+tracking per-context properties of context-aware programming languages. The \emph{coeffect 
+calculus} captures the shared properties of such languages. It is parameterized by a \emph{flat
+coeffect algebra} and can be instantiated to track implicit parameters, liveness and number of
+required past values in dataflow languages.
+
+To capture contextual properties in full generality, the flat coeffect calculus permits multiple 
+valid typing derivations for a given term. To resolve the ambiguity arising from such generality, 
+each concrete context-aware language is also equipped with an algorithm for choosing a unique 
+typing derivation. In the next chapter, we give operational meaning for concrete coeffect 
+languages based on the flat coeffect calculus and we discuss their safety.
 
@@
 
-\item The type system permits multiple typing derivation for certain programs due to the ambiguity
-  inherent in conextual lambda abstraction rule. In Section~\ref{sec:flat-unique}, we discuss
-  variations of the type system that resolve the ambiguity and give a unique typing derivation
-  for the three coeffect systems covered in this chapter.
+\item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent 
+  in the conextual lambda abstraction rule. Each concrete context-aware language based on the 
+  coeffect calculus may resolve this ambiguity differently. We show the procedure for choosing
+  unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).
 
@@
   and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
-  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections motivate the laws
+  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections justify the laws
   of the flat coeffect algebra.
@@
 
-In the previous chapter, we looked at three important examples of systems that track whole-context 
+In the previous chapter, we looked at three examples of systems that track whole-context 
 properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live}) 
-and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a very similar 
-structure. First, their lambda abstraction duplicates the requirements. Given a body with context 
-requirements $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
-annotated with $\cclrd{r}$. Second, their application arises from the combination of \emph{sequential} 
-and \emph{pointwise} composition.
-
-The system for tracking of implicit parameters and similar (Section~\ref{sec:applications-flat-impl})
-differ in two ways. In lambda abstraction, they split the context requirements between the 
+and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a similar 
+structure in two ways. First, their lambda abstraction duplicates the demands. Given a body with 
+context demands $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
+annotated with $\cclrd{r}$. Second, the context demands in the type systems are combined using 
+two different operators (representing sequential and pointwise operations).
+
+The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl})
+differs. In lambda abstraction, it splits the context demands between the 
 declaration site and the call site and they use only a single operator on the indices, typically $\cup$.
 
-Despite the differences, the systems fit the same unified framework. This becomes apparent when
+Despite the differences, the systems fit the same framework. This becomes apparent when
 we consider the categorical structure (Section~\ref{sec:flat-semantics}). However, rather than 
@@
 a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
-in the structure, as explained in the previous chapter, most importantly in lambda abstraction
-(the relationship with monads is further discussed in Section~\ref{sec:flat-related}).
+in the structure of lambda abstraction as discussed in Section~\ref{sec:applications-structure-lam}
+(the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Reconciling lambda abstraction}
+\label{sec:flat-calculus-lambda}
+
+Recall the lambda abstraction rules for the implicit parameters system (annotating the context
+with sets of required parameters) and the dataflow system (annotating the context with the
+number of past required values):
+%
+\begin{equation*}
+\tyrule{param}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\;
+\tyrule{df1}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{n}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 }
+\end{equation*}
+%
+In order to capture both systems using a single calculus, we need a way of unifying the two
+systems. For the dataflow system, this can be achieved by over-approximating the number of 
+required past elements:
+%
+\begin{equation*}
+\tyrule{df2}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{\textnormal{min}(n, m)}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 }
+\end{equation*}
+%
+The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule. 
+Furthermore, if we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on 
+annotations of contexts), then the reverse is also true -- because 
+$\cclrd{\textit{min}(n, m)} \leq \cclrd{m}$ and $\cclrd{\textit{min}(n, m)} \leq \cclrd{n}$.
+In other words (\emph{df1}) is more precise, but (\emph{df2}) gives a sound over-approximation
+with a structure that can be unified with (\emph{param}).
+
+Using a rule such as (\emph{df2}) allows us to give a unified formulation of the flat coeffect 
+calculus in Section~\ref{sec:flat-calculus-types}, however the coeffect-specific handling of
+lambda abstraction remains important in practical implementation in order to obtain a unique
+typing derivation for each coeffect program as discussed in Section~\ref{sec:flat-unique}
+(and in the implementation in Chatper~\ref{ch:impl}).
 
@@
 
-The flat coeffect calculus is defined in terms of \emph{flat coeffect algebra}, which defines
-the structure of context annotations, such as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
-sets of implicit parameters, versions represented as integers or other values. The expressions of 
-the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type 
-\ident{num} as an example of a concrete base type with numerical constants written as $n$:
-%
-\begin{equation*}
-\begin{array}{rcl}
-e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
-\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
-\end{array}
-\end{equation*}
-%
-Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
-annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a 
-unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
-we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations} 
-rather than solving the problem of type reconstruction.
-
-We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
-represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
-It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
-context $\cclrd{r}$.
-
-In the categorically-inspired translation in the next chapter, the function 
-$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function 
-$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
-does not itself exist as a syntactical value in the coeffect calculus. This is because we use 
-comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in 
-the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
-The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Reconciling lambda abstraction}
-\label{sec:flat-calculus-lambda}
-
-Recall the lambda abstraction rules for the implicit parameters system (annotating the context
-with sets of required parameters) and the dataflow system (annotating the context with the
-number of past required values):
-%
-\begin{equation*}
-\tyrule{param}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\;
-\tyrule{df1}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{n}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 }
-\end{equation*}
-%
-In order to capture both systems using a single calculus, we need a way of unifying the two
-systems. For the dataflow system, this can be achieved by over-approximating the number of 
-required past elements:
-%
-\begin{equation*}
-\tyrule{df2}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{\textnormal{min}(n, m)}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 }
-\end{equation*}
-%
-The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule. 
-Furthermore, if we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on 
-annotations of contexts), then the reverse is also true -- because 
-$\cclrd{\textit{min}(n, m)} \leq \cclrd{m}$ and $\cclrd{\textit{min}(n, m)} \leq \cclrd{n}$.
-In other words (\emph{df1}) is more precise, but (\emph{df2}) gives a sound over-approximation
-with a structure that can be unified with (\emph{param}).
-
-Using a rule such as (\emph{df2}) allows us to give a unified formulation of the flat coeffect 
-calculus in Section~\ref{sec:flat-calculus-types}, however the coeffect-specific handling of
-lambda abstraction remains important in practical implementation in order to obtain a unique
-typing derivation for each coeffect program as discussed in Section~\ref{sec:flat-unique}
-(and in the implementation in Chatper~\ref{ch:impl}).
+The \emph{flat coeffect calculus} is parameterized by a \emph{flat coeffect algebra}, which defines
+the structure of context annotations, written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
+sets of implicit parameters, versions represented as integers or other values.
+
+Typing judgements the attach coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$ to the context
+and are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$.
 
@@
 
-Before looking at the type system in Figure~\ref{fig:flat-types}, let us clarify how the rules
+Before looking at the typing rules, let us clarify how the typing judgements
 should be understood. The coeffect calculus provides both analysis of context dependence (type 
@@
 \item \textsc{Analysis of context dependence.}
-Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context requirements}. This means
+Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context demands}. This means
 we can over-approximate them and require more than is actually needed at runtime.
 
-Syntactically, the typing rules should be read top-down. In function application, the context 
-requirements of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in 
-lambda abstraction, the requirements of a single expression (the body) are split between
+Syntactically, the typing rules are best read top-down. In function application, the context 
+demands of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in 
+lambda abstraction, the demands of a single expression (the body) are split between
 the declaration site and the call site.
@@
 The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
-position} in the semantic model (Section~\ref{sec:flat-semantics}). It means that we need to be
+position} in the semantic model (Section~\ref{sec:semantics-theory}). It means that we need to be
 careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
-exactly the opposite things. To disambiguate, we always use the term \emph{context requirements} 
+exactly the opposite things. To disambiguate, we always use the term \emph{context demands} 
 when using the syntactic view, especially in the rest of Chapter~\ref{chf:flat}, and 
@@
 which are mainly used in function application. The third operator, $\czip$ is used in lambda 
-abstraction and represents \emph{splitting} of context requirements (or, semantically, \emph{merging} 
+abstraction and represents \emph{splitting} of context demands (or, semantically, \emph{merging} 
 of available context capabilities).
@@
 \item For the $\czip$ operation, we require associativity and idempotence. The idempotence
-requirement makes it possible to duplicate the coeffects and place the same requirement on both
+demand makes it possible to duplicate the coeffects and place the same demand on both
 call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
@@
 Using the syntactic reading, the two operators represent \emph{merging} and \emph{splitting} of context 
-requirements -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context 
-requirements of the body are split between two positions in the conclusions; in the (\emph{app}) rule, 
-$\cpar$ appears in the conclusion and combines two context requirements from the assumptions.
+demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context 
+demands of the body are split between two positions in the conclusions; in the (\emph{app}) rule, 
+$\cpar$ appears in the conclusion and combines two context demands from the assumptions.
 
@@
 
+\subsection{Language syntax}
+\label{sec:flat-calculus-syntax}
+
+The expressions of 
+the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type 
+\ident{num} as an example of a concrete base type with numerical constants written as $n$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\end{array}
+\end{equation*}
+%
+Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
+annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a 
+unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
+we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations} 
+rather than solving the problem of type reconstruction.
+
+We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
+represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
+It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a 
+context $\cclrd{r}$.
+
+In the categorically-inspired translation in the next chapter, the function 
+$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function 
+$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
+does not itself exist as a syntactical value in the coeffect calculus. This is because we use 
+comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in 
+the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
+The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.
+
+%---------------------------------------------------------------------------------------------------
+
 \begin{figure}[t]
@@
 
-\subsection{Flat coeffect types}
+\subsection{Type system}
 \label{sec:flat-calculus-types}
@@
 The (\emph{abs}) rule is defined as discussed in Section~\ref{sec:flat-calculus-lambda}. The
-body is annotated with context requirements $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
-between the context-requirements on the declaration site $\cclrd{r}$ and context-requirements on
+body is annotated with context demands $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
+between the context-demands on the declaration site $\cclrd{r}$ and context-demands on
 the call site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.
 
-In function application (\emph{app}), context requirements of both expressions and the 
+In function application (\emph{app}), context demands of both expressions and the 
 function are combined. As discussed in Chapter~\ref{ch:applications}, the pointwise composition
-$\cpar$ is used to combine the context requirements of the expression representing a function 
-$\cclrd{r}$ and the context requirements of the argument, sequentially composed with the 
-context-requirements of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.
+$\cpar$ is used to combine the context demands of the expression representing a function 
+$\cclrd{r}$ and the context demands of the argument, sequentially composed with the 
+context-demands of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.
 
@@
 As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
-differs from the rule for effect systems in that it does not delay all context requirements. 
-In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the requirements
+differs from the rule for effect systems in that it does not delay all context demands. 
+In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
 can be satisfied either by the call-site or by the declaration-site. In case of dataflow and
@@
 where function abstraction captures all parameters that are available at the declaration-site and 
-places all other requirements on the call-site. For the example in the introduction, this means
+places all other demands on the call-site. For the example in the introduction, this means
 that the body of $f$ captures the value of \ident{?p} available from the declaration-site
@@
 Given a typing derivation for the body that produced constraints $C$, we generate an additional
-constraint that restricts $\cclrd{r}$ (declaration-site requirements) to those available in the
+constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
 current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds 
@@
 The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
-A function $(\lambda x.e)$ does not defer all context requirements of the body $e$ and may have
-immediate context requirements. Thus we say that $e$ is a value if it is a value in the usual
-sense \emph{and} has not immediate context requirements. We define this formally in 
+A function $(\lambda x.e)$ does not defer all context demands of the body $e$ and may have
+immediate context demands. Thus we say that $e$ is a value if it is a value in the usual
+sense \emph{and} has not immediate context demands. We define this formally in 
 Section~\ref{sec:flat-syntax-cbv}.
@@
 semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
-to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context requirements.
+to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context demands.
 
@@
 coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
-not automatically defer all context requirements.
+not automatically defer all context demands.
 
@@
 \noindent
-The call-by-value substitution substitutes a value, with context requirements $\cunit$, for a 
+The call-by-value substitution substitutes a value, with context demands $\cunit$, for a 
 variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
-context requirements of the term:
+context demands of the term:
 
@@
 $\coctx{\Gamma,  x\!:\!\sigma}{ \cclrd{r}  } \vdash e : \tau$, then substituting $v$ for $x$ does not
-change the type and context requirements, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
+change the type and context demands, that is $\coctx{\Gamma}{ \cclrd{r} } \vdash \subst{e}{x}{v} : \tau$.
 \end{lemma}
@@
 %
-Intuitively, this specifies that the $\czip$ operation (splitting of context requirements) 
+Intuitively, this specifies that the $\czip$ operation (splitting of context demands) 
 under-approximates the actual context capabilities while the $\cpar$ operation (combining of
-context requirements) over-approximates the actual context requirements.
+context demands) over-approximates the actual context demands.
 
@@
 
-Semantically, the coeffect over-approximates the actual requirements -- at run-time, the code does not 
+Semantically, the coeffect over-approximates the actual demands -- at run-time, the code does not 
 actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect 
@@
 of bottom-pointed coeffects. For top-pointed flat coeffects, the substitution lemma holds without 
-additional requirements:
+additional demands:
 
@@
 In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
-requirements. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
-then the context requirements arising from the substitution can be associated with the context
-$\Gamma$, leaving the context requirements of a function value unchanged. As a result, substitution 
-does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for 
+demands. However, if the system satisfies the strong condition that $\czip = \cseq = \cpar$ 
+then the context demands arising from the substitution can be associated with the context
+$\Gamma$, leaving the context demands of a function value unchanged. As a result, substitution 
+does not break soundness as in effect systems. The demand $\czip = \cseq = \cpar$ holds for 
 our implicit parameters example (all three operators are a set union) and for other set-like 
@@
 The flat system discussed here is \emph{flexible enough} to let us always re-associate new context 
-requirements (arising from the substitution) with the free-vari\-able context. In contrast, the 
+demands (arising from the substitution) with the free-vari\-able context. In contrast, the 
 structural system discussed in Chapter~\ref{ch:structural} is \emph{precise enough} to keep the 
@@
 marking the context and function argument as live. In case of dataflow, the immediate coeffects 
-are made larger by the lambda abstraction -- the context requirements of the function value are 
+are made larger by the lambda abstraction -- the context demands of the function value are 
 imposed on the declaration site of the new lambda abstraction. We remedy this limitation in 
@@
 The derivation starts with the expression $e$ and derives the type for $\lambda x.e~x$. The
-application yields context requirements $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
+application yields context demands $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
 original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
@@
 coeffects and can be used to define sub-coeffecting. Syntactically, an expression with context
-requirements $\cclrd{r}$ can be treated as an expression with greater context requirements:
+demands $\cclrd{r}$ can be treated as an expression with greater context demands:
 %
@@
 In Figure~\ref{fig:flat-types-sub}, we add sub-typing on function types, making it possible to treat
-a function with smaller context requirements as a function with greater context requirements:
+a function with smaller context demands as a function with greater context demands:
 %
@@
 Recall the (\emph{let}) rule in Figure~\ref{fig:flat-types}. It annotates the expression 
-$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context requirements $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
+$\kvd{let}~x=e_1~\kvd{in}~e_2$ with context demands $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
 This is a special case of typing an expression $(\lambda x.e_2)~e_1$, using the idempotence
@@
 lambda abstraction -- for implicit parameters, the lambda function needs to split context 
-requirements using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to 
-duplicate the requirement $\cclrd{r}$ of the body. We introduced the $\czip$ operation as 
+demands using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to 
+duplicate the demand $\cclrd{r}$ of the body. We introduced the $\czip$ operation as 
 a way of providing the additional abstraction.
@@
 Recall that $(\C, \czip)$ is a band, that is, $\czip$  is idempotent and associative. The 
-idempotence means that the context requirements of the body can be required from both the 
+idempotence means that the context demands of the body can be required from both the 
 declaration site and the call site. Thus, the (\emph{idabs}) typing is valid. For reference,
@@
 removes the non-determinism of type checking that arises from the splitting of context 
-requirements in the original (\emph{abs}) rule.
+demands in the original (\emph{abs}) rule.
 
@@
 \noindent
-Intuitively, this means that the context requirements of a function with multiple arguments can be 
+Intuitively, this means that the context demands of a function with multiple arguments can be 
 split arbitrarily between the declaration site and (multiple) call sites. In other words, it does
-not matter how the context requirements are satisfied.
+not matter how the context demands are satisfied.
 
@@
 properties (here, coeffects) of the expressions. The monoid laws then correspond to 
-the requirement that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
-requirement that $((), e) \simeq e \simeq (e, ())$ (unit).
+the demand that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
+demand that $((), e) \simeq e \simeq (e, ())$ (unit).
 
@@
 \noindent
-The Remark~\ref{thm:flat-tup-eq} motivates the requirement of the monoid structure 
+The Remark~\ref{thm:flat-tup-eq} motivates the demand of the monoid structure 
 $(\C, \cpar, \czero)$ of the flat coeffect algebra. We require only unit and associativity

### 2016-03-06 text/Chapter05.tex
Move stuff in chapter 4, improve introductions

@@
 In summary, coeffect type system, captures certain runtime 
-requirements of context-aware programs and (as we show in this chapter), eliminates common errors 
+demands of context-aware programs and (as we show in this chapter), eliminates common errors 
 related to working with context.
@@
 In the next section, we introduce \emph{indexed comonads}, which provide a more precise analysis 
-and let us model computations with context requirements $\cclrd{r}$ as functions 
+and let us model computations with context demands $\cclrd{r}$ as functions 
 $\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.
@@
 by $\ident{counit}$) together with the (associative) composition operation that composes computations 
-with contextual requirements as follows:
+with contextual demands as follows:
 %
@@
 The composition $\hat{\circ}$ best expresses the intention of indexed comonads. Given two functions
-with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
-requires $\cclrd{r}\,\cseq\,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
+with contextual demands $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
+requires $\cclrd{r}\,\cseq\,\cclrd{s}$. The contextual demands propagate \emph{backwards} and
 are attached to the input of the composed function.
@@
 \subsection{Functional target langauge}
+\label{sec:semantics-translation-target}
 
@@
 \subsection{Comonadically-inspired translation}
+\label{sec:semantics-translation-transl}
 
@@
 
-In Section~\ref{sec:semantics-proofs-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
+In Section~\ref{sec:semantics-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
 safety property of two concrete context-aware programming langauges based on the coeffect 
@@
 \item Syntactically, the second difference is in the lambda abstraction (\emph{abs}). In 
-  coeffect systems, the context requirements of the body can be split between (or duplicated
+  coeffect systems, the context demands of the body can be split between (or duplicated
   at) declaration site and call site, while monadic effect systems always defer all effects.
@@
 we use the $\cup$ operation for all three operations of the coeffect algebra. However, to enable
-splitting of implicit parameter requirements using the reader monad, we need to extend the 
+splitting of implicit parameter demands using the reader monad, we need to extend the 
 monad structure and change the translation of monadic lambda abstraction.
@@
 Lambda abstraction in effect systems does not provide a way of splitting the context
-requirements between declaration site and call site (or, semantically, combining the implicit 
+demands between declaration site and call site (or, semantically, combining the implicit 
 parameters available in the scope where the function is defined and those specified by the caller).
@@
 In the comonadic (\emph{cabs}) rule, the implicit parameters of the body are split. However,
-the monadic rule (\emph{mabs}) places all requirements on the call site. This follows from the
+the monadic rule (\emph{mabs}) places all demands on the call site. This follows from the
 fact that monadic semantics uses the \ident{unit} operation in the interpretation of lambda abstraction:
@@
 Another important aspect of the signature is that the function needs to be indexed by the coeffect
-annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context requirements
+annotations $\cclrd{r}, \cclrd{s}$. The indices determine how the input context demands
 $\cclrd{r \cup s}$ are split -- and thus guarantee determinism of the function at run-time.
@@
 As just demonstrated, we can extend monads so that the reader monad is capable of capturing
-the semantics of implicit parameters, including the splitting of implicit parameter requirements
+the semantics of implicit parameters, including the splitting of implicit parameter demands
 in lambda abstraction. Can we also go the other way round and \emph{restrict} the comonadic
-semantics so that all requirements are delayed as in the (\emph{mabs}) rule, thus modelling 
+semantics so that all demands are delayed as in the (\emph{mabs}) rule, thus modelling 
 fully dynamically scoped parameters?

### 2016-04-09 text/Chapter04.tex
Atom whitespace, working on implementation chapter

@@
 
-\paragraph{Implementation.}
-From the presentation in this section, it might appear that resolving the ambiguity related to
-lambda abstraction for implicit parameters requires a type system that is quite different from
-the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
-case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are 
-simpler.
-
-Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets 
-of implicit parameters to satisfy the constraints. We still need to track implicit parameters in 
-scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
-to the one from Figure~\ref{fig:flat-types}:
-%
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
-  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
-    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}} 
-\end{equation*}
-%
-Given a typing derivation for the body that produced constraints $C$, we generate an additional
-constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
-current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds 
-the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.
-
 %---------------------------------------------------------------------------------------------------
@@
 
-\paragraph{Implementation.}
-As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
-require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
-by (\emph{idabs}) can be easily obtained by generating additional constraints:
-%
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
-  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
-    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}} 
-\end{equation*}
-%
-Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
-coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by 
-(\emph{idabs}).
-
 % ==================================================================================================

### 2016-04-09 text/Chapter08.tex
Atom whitespace, working on implementation chapter

@@
-\chapter{Conclusions} 
-\label{ch:conclusions}
+% ==================================================================================================
 
-%---------------------------------------------------------------------------------------------------
+\chapter{Implementation}
+\label{ch:impl}
 
-How to make this practical?
+In the previous three chapters, we presented the theory of coeffects consisting of type system
+and comonadically inspired semantics of two parameterized coeffect calculi. The theory provides
+a framework that simplifies the implementation of safe context-aware programming languages. To
+support this claim, this chapter presents a prototype implementation of three coeffect languages --
+language with implicit parameters and both flat and structural versions of a data-flow language.
 
-codo notation / computation expressions
+The implementation directly follows the thoery presented in the previous three chapters. It
+consists of a common framework that provides type checking and translation to a simple functional
+target language with comonadically-inspired primitives. Each concrete context-aware language then
+adds a domain-specific rule for choosing unique typing derivation (as discussed in
+Section~\ref{sec:flat-unique}) together with a domain-specific definition of the
+comonadically-inspired primitives that define the runtime semantics (see Section~\ref{sec:semantics-proofs}).
 
-embedding algebraic things
+The main goal of the implementation is to show that the theory is practically useful and to present it
+in a more practical way. However, we do not intend to build a complete real-world programming language.
+For this reason, the implementation is available primarilly as an interactive web-based essay, though
+it can be also downloaded and run locally.
+
+\paragraph{Chapter structure and contributions}
+
+\begin{itemize}
+\item We discuss how the implementation follows from the theory presented earlier (Section~\ref{sec:impl-theory}).
+  This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
+  to a simple target langauge that is then iterpereted. We also discuss how the common framework makes it
+  easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).
+
+\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate ointeresting
+  aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
+  difference between flat and structural systems (Section~\ref{sec:impl-case-typing}) and the
+  comonadically-inspired translation (Section~\ref{sec:impl-case-transl}).
+
+\item We briefly outline some of the interesting technical details of the implementation
+  (Section~\ref{sec:impl-tech}). To make the implementation easily accessible, we make it
+  available via web and we also implement a limited support for type inference.
+
+\item The implementation is available not just as downloadable code, but also in the format of interactive essay
+  (Section~\ref{sec:impl-essay}), which aims to make coeffects accessible to a broader audience.
+  We conclude this chapter by discussing the most interesting aspects of the presentation.
+\end{itemize}
+
+% ==================================================================================================
+%
+%    #######
+%       #    #    # ######  ####  #####  #   #
+%       #    #    # #      #    # #    #  # #
+%       #    ###### #####  #    # #    #   #
+%       #    #    # #      #    # #####    #
+%       #    #    # #      #    # #   #    #
+%       #    #    # ######  ####  #    #   #
+%
+% ==================================================================================================
+
+\section{From theory to implementation}
+\label{sec:impl-theory}
+
+The theory discussed so far provides the two key components of the implementation. In
+Chapter~\ref{ch:flat}, we discussed the type checking of context-aware programs and
+Chapter~\ref{ch:semantics} models the execution of context-aware programs (in terms of
+translation and operational semantics). For structural coeffects, the same components are discussed
+in Chapter~\ref{ch:structural}. In this section, we discuss how those provide foundation for the
+implementation.
+
+\subsection{Type checking}
+\label{sec:impl-theory-typing}
+
+(Section~\ref{sec:flat-unique})
+
+\paragraph{Implementation.}
+From the presentation in this section, it might appear that resolving the ambiguity related to
+lambda abstraction for implicit parameters requires a type system that is quite different from
+the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
+case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are
+simpler.
+
+Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets
+of implicit parameters to satisfy the constraints. We still need to track implicit parameters in
+scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
+to the one from Figure~\ref{fig:flat-types}:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}}
+\end{equation*}
+%
+Given a typing derivation for the body that produced constraints $C$, we generate an additional
+constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
+current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds
+the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.
+
+\paragraph{Implementation.}
+As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
+require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
+by (\emph{idabs}) can be easily obtained by generating additional constraints:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}}
+\end{equation*}
+%
+Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
+coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by
+(\emph{idabs}).
+
+Type checking
+ * DSL specific constraint resolution
+ * easier for structural
+
+\subsection{Translation}
+\label{sec:impl-theory-transl}
+
+Translation ~ Semantics
+
+\subsection{Extensions}
+\label{sec:impl-theory-ext}
+
+It is really easy to add another DSL
+
+\section{Case studies}
+\label{sec:impl-case}
+
+\subsection{Typing}
+\label{sec:impl-case-typing}
+
+let both =
+  let ?fst = 100 in
+  fun trd -> ?fst + ?snd + trd in
+let ?fst = 200 in
+both 1
+
+num -{ ?snd:num }-> num
+
+let struct x y = x + prev y in
+struct
+
+num -{ 1 }-> num -{ 1 }-> num
+
+let flat x y = x + prev y in
+flat
+
+num -{ 0 }-> num -{ 1 }-> num
+
+\subsection{Translation}
+\label{sec:impl-case-transl}
+
+let ?param = 10 in
+fun x -> ?param + ?other
+
+let (ctx2, ctx3) = split (duplicate finput) in
+let ctx1 = letimpl ?param (ctx2, 10) in
+fun x ->
+  let ctx4 = merge (x, ctx1) in
+  let (ctx5, ctx6) = split (duplicate ctx4) in
+  lookup ?param ctx5 + lookup ?other ctx6
+
+fun x -> (fun v -> prev v) (prev x)
+
+fun x ->
+  let ctx1 = merge (x, sinput) in
+  (fun v ->
+    let ctx3 = merge (v, choose_0 ctx1) in
+    counit (prev ctx3)) (cobind (fun ctx2 ->
+      counit (prev ctx2)) (choose_1 ctx1))
+
+\section{Technical details}
+\label{sec:impl-tech}
+
+1. F# + JavaScript
+2. Single AST and parsing
+3. Type checking and constraint generation
+4. Constraint resolution
+
+\section{Interactive essay}
+\label{sec:impl-essay}

### 2016-04-10 text/Chapter08.tex
Implementation chapter

@@
 
-\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate ointeresting
+\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate interesting
   aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
@@
 
-\subsection{Type checking}
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Type checking and inference}
 \label{sec:impl-theory-typing}
 
-(Section~\ref{sec:flat-unique})
+To simplify the writing of context-aware programs, the implementation provides a limited form of
+type inference (Section~\ref{sec:impl-tech}). This is available just for convenience and so we do
+not claim any completeness or complexity result about the algorithm and we do not present full
+formalization. However, it is worth noting how the domain-specific procedures for choosing a unique
+type derivation (Section~\ref{sec:flat-unique}) are adapted.
 
-\paragraph{Implementation.}
-From the presentation in this section, it might appear that resolving the ambiguity related to
-lambda abstraction for implicit parameters requires a type system that is quite different from
-the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
-case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are
-simpler.
+The type inference works in the standard way \cite{types-mlessence,types-inference} by generating
+type constraints and solving them. Solving of type constraints is done in the standard way, but we
+additionally collect and solve \emph{coeffect constraints}. In order to obtain unique type
+derivation, we generate additional coeffect constraints in lambda abstraction of each flat
+coeffect language.
 
-Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets
-of implicit parameters to satisfy the constraints. We still need to track implicit parameters in
-scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
-to the one from Figure~\ref{fig:flat-types}:
+\paragraph{Flat implicit parameters.}
+
+As discussed in Section~\ref{sec:flat-unique}, when choosing unique typing derivation for
+implicit parameters, we keep track of the implicit parameters available in lexical scope
+(written as $\cclrd{\Delta}$). In lambda abstraction rule, the implicit parameters required by
+the body (tracked by $\ccclrd{r}$) are split so that all parameters available in lexical scope
+are captured and only the remaining parameters ($\cclrd{r}\setminus\cclrd{\Delta}$) are required
+from the caller of the function.
+
+From the presentation in Section~\ref{sec:flat-unique}, it might appear that resolving the ambiguity
+related to lambda abstraction for implicit parameters requires a type system different from
+the core flat coeffect type system shown earlier in Section ~\ref{sec:flat-calculus-types}.
+This is not the case.
+
+We track implicit parameters in scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from
+the implementation only generates an additional coeffect constraint. Writing
+$\coctx{\Gamma}{\cclrd{t}} \vdash e : \tau ~|~ C}$ for a judgement that generates coeffect
+constraints $C$, the (\emph{abs}) rule used for implicit parameters looks as follows:
 %
@@
 constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
-current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds
+current static scope $\cclrd{\Delta}$. It is not necessary to generate a constraint for the
+coeffect $\cclrd{s}$, because our constraint satisfaction algorithm finds
 the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.
 
-\paragraph{Implementation.}
-As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
-require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
-by (\emph{idabs}) can be easily obtained by generating additional constraints:
+\paragraph{Flat data-flow.}
+In context-aware language for data-flow (and in language with liveness tracking), the inherent
+ambiguity of the (\emph{abs}) rule is resolved by duplicating the context requirements of the body.
+In Section~\ref{sec:flat-unique}, this was defined by replacing the standard coeffect (\emph{abs})
+rule with a rule (\emph{idabs}) that uses an annotation $\cclrd{r}$ for the body of the function,
+declaration-site coeffect and call-site coeffect.
+
+As with implicit parameters, the implementation does not require changing the core (\emph{abs})
+typing rule of the flat coeffect system. Instead, the unique resulution is obtained by generating
+additional coeffect constraints:
 %
@@
 
-Type checking
- * DSL specific constraint resolution
- * easier for structural
+% --------------------------------------------------------------------------------------------------
 
-\subsection{Translation}
+\subsection{Execution of context-aware programs}
 \label{sec:impl-theory-transl}
 
-Translation ~ Semantics
+Context-aware programs are executed by translating the source program into a simple functional
+target language. For simplicity, programs in the simple target language are then interpreted, but
+they could equally be compiled using standard techniques for compiling functional code. The
+translation follows the rules defined in Section~\ref{sec:semantics-translation} (for flat coeffect
+languages) and Section~X (for structural coeffect languages). The result of the translation is a
+program that consists of the following:
 
-\subsection{Extensions}
+\begin{itemize}
+\item {\sc Functional constructs.} Those include binary operations, tuples, let binding,
+  constants, variables, function abstraction and application. The interpreter keeps a map of
+  assignments for variables in scope and recursively evaluates the expression.
+
+\item {\sc Comonadic operations.} Those are the comonadic primitives provided by indexed comonads --
+  \ident{cobind}, \ident{counit} together with \ident{merge} and \ident{split} for flat coeffects
+  or \ident{merge} and \ident{choose} for structural coeffects. The translation that produces these
+  is shared by all context-aware languages, but their definition in the interpreter is
+  domain-specific.
+
+\item {\sc Domain-specific operations.} Each context-aware language may additionally include
+  operations that model domain-specific operations. For data-flow, this is \ident{prev} (accessing
+  past values) and for implicit parameters, this is \ident{letimpl} and \ident{lookup} for implicit
+  parameter binding and access, respectively.
+\end{itemize}
+
+\noindent
+The fact that the prototype implementation is based on the theoretical framework provided by
+coeffect calculi means that it has the desirable properties proved in Section~\ref{sec:semantics-proofs}
+and Section~X. In particular, evaluating a well-typed context-aware program in a context that
+provides sufficient contextual capabilities will not cause an error.
+
+In the interactive essay (Section~\ref{sec:impl-essay}), we further use the coeffects to
+automatically generate a user interface that requires the user to provide the required contextual
+capabilities (past values for individual variables, or values for implicit parameters).
+
+Another benefit of using the common framework is that the implementation can be easily extended
+to support additional context-aware languages.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Supporting more context-aware languages}
 \label{sec:impl-theory-ext}
 
-It is really easy to add another DSL
+The prototype implementation supports two of the context-aware languages discussed in this thesis:
+implicit parameters and data-flow. The remaining examples are calculus for tracking variable
+liveness (of ordinary variables) and the structural language based on bounded reuse (counting
+number of times values are used). However, the prototype implementation is based on the
+common coeffect framework and makes it easy to add support for these and also for other context
+aware languages based on coeffects.
+
+In order to extend the implementation with support for liveness or bounded reuse tracking (or
+other context-aware language), the following 4 additions are required:
+
+\begin{enumerate}
+\item A domain-specific function abstraction rule that resolves the ambiguity in the general
+  (\emph{abs}) rule of the flat coeffect calculus. For liveness, the handling would be the same
+  as for data-flow, but for other flat coeffect systems, another resolution mechanism might be used
+  instead.
+
+\item A domain-specific instance of coeffect algebra needs to be provided. In order to support
+  the type inference in the prototype implementation, the constraint solver needs to be extended
+  to solve constraint using the coeffect algebra. For liveness, this would be solving simple
+  two-point lattice constraints.
+
+\item For evaluation, a new kind of comonadic values needs to be added. For liveness, this would be
+  an option value that may or may not contain a value. The semantics of comonadic operations on the
+  values needs to be defined.
+
+\item For context-aware languages that have additional primitives (such as \kvd{prev} or \ident{?param}),
+  the parser and AST needs to be extended, custom type-checking and translation rules added and
+  domain-specific primitive operations (with their semantics) provided. Liveness and bounded reuse
+  do not have additional custom syntax and so supporting these would not require this step.
+\end{enumerate}
+
+\noindent
+The list mirrors a list of steps that need to be done when supporting a new effectful computation
+in a language that supports monadic do-notation. The step (3) corresponds to implementing a new
+monad and (4) corresponds to adding monad-specific effectful operations. The step (2) applies when
+using indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
+have effectful/monadic counterpart is (1).
+
+When adding a new context-aware programming language support, much of the existing infrastructure
+can be reused. This includes the implementation of the core coeffect and type checking rules and
+also the translation for standard language constructs as well as the interpreter for the target
+language.
+
+% ==================================================================================================
+%
+%     #####
+%    #     #   ##    ####  ######     ####  ##### #    # #####  # ######  ####
+%    #        #  #  #      #         #        #   #    # #    # # #      #
+%    #       #    #  ####  #####      ####    #   #    # #    # # #####   ####
+%    #       ######      # #              #   #   #    # #    # # #           #
+%    #     # #    # #    # #         #    #   #   #    # #    # # #      #    #
+%     #####  #    #  ####  ######     ####    #    ####  #####  # ######  ####
+%
+% ==================================================================================================
 
@@
 
-\subsection{Typing}
+The prototype implementation illustrates a number of interesting aspects of coeffect systems.
+Those appear as examples in the interactive essay (discussed in Section~\ref{sec:impl-essay}), but
+we briefly review them in this section.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Typing context-aware programs}
 \label{sec:impl-case-typing}
 
-let both =
-  let ?fst = 100 in
-  fun trd -> ?fst + ?snd + trd in
-let ?fst = 200 in
-both 1
+We first consider two case studies of how coeffect type checking works. The first one exposes the
+resolution of the ambiguity in typing for implicit parameters and the second one exposes the
+difference between flat and structural system for data-flow.
 
-num -{ ?snd:num }-> num
+\paragraph{Abstraction for implicit parameters.}
+As discussed in Section~\ref{sec:impl-theory-typing}, the implementation of the language with
+implicit parameters resolves the ambiguity in the lambda abstraction by generating a coeffect
+constraint that restricts the set of parameters required from the declaration-site to those that
+are lexically available. Remaining parameters are required from the call-site. This is illustrated
+by the following example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{both}~=\\[-0.25em]
+\quad \kvd{let}~\ident{?fst}~=~100~\kvd{in}\\[-0.25em]
+\quad \kvd{fun}~\ident{trd}~\rightarrow~\ident{?fst}~+~\ident{?snd}~+~\ident{trd}~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{?fst}~=~200~\kvd{in}\\[-0.25em]
+\ident{both}~1
+\end{array}
+\end{equation*}
+%
+In this expression, the lambda function on line 3 requires implicit parameters \ident{?fst} and
+\ident{?snd}. Since \ident{?fst} is available in scope, the type of \ident{both} is a function that
+requires only \ident{?snd}. In the text-based notation used in the prototype, the type of the
+function \ident{both} is: {\tt num -\{?snd:num\}-> num}.
 
-let struct x y = x + prev y in
-struct
+\paragraph{Flat and structural data-flow.}
+In flat data-flow, the context requirements of the body is required from both the declaration-site
+and from the call-site. In structural data-flow, the context requirements are tracked separately
+for each variable, which provides a more precise type. Consider the following two examples (the
+\kvd{let} keyword is used to define a curried function of two arguments):
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{oldy}~\ident{x}~\ident{y}~=~\ident{x}~+~\kvd{prev}~\ident{y}~\kvd{in}\\[-0.25em]
+\ident{oldy}
+\end{array}
+\end{equation*}
+%
+When type checking the expression using the flat system, the type of \ident{oldy} is inferred as {\tt num -\{1\}-> num -\{1\}-> num},
+but when using the structural system, the type becomes {\tt num -\{0\}-> num -\{1\}-> num}.
 
-num -{ 1 }-> num -{ 1 }-> num
+This illustrates the difference between the two - the flat system keeps only one annotation
+for the whole body (which requires 1 past value). In lambda abstraction (or function declaration
+written using \kvd{let}), this requirement is duplicated. The structural system keeps information
+per-variable and so the resulting type reflects the fact that only the variable \idnet{y} appears
+inside \kvd{prev}.
 
-let flat x y = x + prev y in
-flat
-
-num -{ 0 }-> num -{ 1 }-> num
+% --------------------------------------------------------------------------------------------------
 
@@
 
-let ?param = 10 in
+\begin{equation*}
+\begin{array}{l}
+let ?param = 10 in\\[-0.25em]
 fun x -> ?param + ?other
+\end{array}
+\end{equation*}
 
-let (ctx2, ctx3) = split (duplicate finput) in
-let ctx1 = letimpl ?param (ctx2, 10) in
-fun x ->
-  let ctx4 = merge (x, ctx1) in
-  let (ctx5, ctx6) = split (duplicate ctx4) in
+\begin{equation*}
+\begin{array}{l}
+let (ctx2, ctx3) = split (duplicate finput) in\\[-0.25em]
+let ctx1 = letimpl ?param (ctx2, 10) in\\[-0.25em]
+fun x ->\\[-0.25em]
+  let ctx4 = merge (x, ctx1) in\\[-0.25em]
+  let (ctx5, ctx6) = split (duplicate ctx4) in\\[-0.25em]
   lookup ?param ctx5 + lookup ?other ctx6
+\end{array}
+\end{equation*}
 
+\begin{equation*}
+\begin{array}{l}
 fun x -> (fun v -> prev v) (prev x)
+\end{array}
+\end{equation*}
 
-fun x ->
-  let ctx1 = merge (x, sinput) in
-  (fun v ->
-    let ctx3 = merge (v, choose_0 ctx1) in
-    counit (prev ctx3)) (cobind (fun ctx2 ->
+\begin{equation*}
+\begin{array}{l}
+fun x ->\\[-0.25em]
+  let ctx1 = merge (x, sinput) in\\[-0.25em]
+  (fun v ->\\[-0.25em]
+    let ctx3 = merge (v, choose_0 ctx1) in\\[-0.25em]
+    counit (prev ctx3)) (cobind (fun ctx2 ->\\[-0.25em]
       counit (prev ctx2)) (choose_1 ctx1))
+\end{array}
+\end{equation*}
 

### 2016-04-21 text/Chapter08.tex
Implementation chapter

@@
 
-\item We briefly outline some of the interesting technical details of the implementation
-  (Section~\ref{sec:impl-tech}). To make the implementation easily accessible, we make it
-  available via web and we also implement a limited support for type inference.
-
-\item The implementation is available not just as downloadable code, but also in the format of interactive essay
-  (Section~\ref{sec:impl-essay}), which aims to make coeffects accessible to a broader audience.
-  We conclude this chapter by discussing the most interesting aspects of the presentation.
+\item The implementation is available not just as downloadable code, but also in the format of
+  interactive web-based essay (Section~\ref{sec:impl-essay}), which aims to make coeffects
+  accessible to a broader audience. We discuss the most interesting aspects of the essay format and
+  briefly discuss some of the interesting implementation detail (Section~\ref{sec:impl-essay-tech}).
 \end{itemize}
@@
 
-\subsection{Supporting more context-aware languages}
+\subsection{Supporting additional context-aware languages}
 \label{sec:impl-theory-ext}
@@
 
-\subsection{Translation}
+\subsection{Comonadically-inspired translation}
 \label{sec:impl-case-transl}
 
+In addition to running coeffect programs, the implementation can also print the result of the
+translation to the simple functional target language with comonadically-inspired primitives.
+The following two case studies illustrate important aspects of the translation for flat coeffect
+systems (Section~\ref{sec:semantics-translation}) and structural coeffect systems (Section~?).
+
+\paragraph{Merging implicit parameter contexts.}
+
+The following example illustrates the lambda abstraction for implicit parameters.
+It defines a parameter \ident{?param} and then returns a function value that captures it, but
+also requires an implicit parameter \ident{?other}:
+%
 \begin{equation*}
 \begin{array}{l}
-let ?param = 10 in\\[-0.25em]
-fun x -> ?param + ?other
+\kvd{let}~\ident{?param} ~=~ \num{10}~\kvd{in}\\[-0.25em]
+\kvd{fun}~\ident{x}~\rightarrow~\ident{?param} ~+~ \ident{?other}
 \end{array}
 \end{equation*}
-
+%
+Translating the code to the target language produces the code below. The reader is encouraged to
+view the translation in the interactive essay (Section~\ref{sec:impl-essay}), which displays the
+types and coeffect annotations of the individual values and primitives. As in the theory, the
+comonadically-inspired primitives are families of operations indexed by the coeffects (we
+omit the annotations here):
+%
 \begin{equation*}
 \begin{array}{l}
-let (ctx2, ctx3) = split (duplicate finput) in\\[-0.25em]
-let ctx1 = letimpl ?param (ctx2, 10) in\\[-0.25em]
-fun x ->\\[-0.25em]
-  let ctx4 = merge (x, ctx1) in\\[-0.25em]
-  let (ctx5, ctx6) = split (duplicate ctx4) in\\[-0.25em]
-  lookup ?param ctx5 + lookup ?other ctx6
+\kvd{let}~(\ident{ctx2}, \ident{ctx3}) ~=~ \ident{split}~(\ident{duplicate}~\kvd{finput})~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{ctx1} ~=~ \ident{letimpl}_{\ident{?param}}~(\ident{ctx2}, \num{10})~\kvd{in}\\[-0.25em]
+\kvd{fun}~\ident{x}~\rightarrow\\[-0.25em]
+\quad \kvd{let}~\ident{ctx4} ~=~ \ident{merge}~(\ident{x}, \ident{ctx1})~\kvd{in}\\[-0.25em]
+\quad \kvd{let}~(\ident{ctx5}, \ident{ctx6}) ~=~ \ident{split}~(\ident{duplicate}~\ident{ctx4})~\kvd{in}\\[-0.25em]
+\quad \ident{lookup}_{\ident{?param}}~\ident{ctx5} ~+~ \ident{lookup}_{\ident{?other}}~\ident{ctx6}
 \end{array}
 \end{equation*}
+%
+The \kvd{finput} value on the first line represents an empty context in which the expression is
+evaluated and is of type $\ctyp{\czero}{\ident{unit}}$. The context is duplicated; \ident{ctx3}
+is not needed, because $\num{10}$ is a constant; \ident{ctx2} is passed to \ident{letimpl}, which
+assigns an implicit parameter value in the newly returned context \ident{ctx1} of type
+$\ctyp{\cclrd{\{\ident{?param}\}}}{\ident{unit}}$ (this now carries the implicit parameter value, but
+it does not represent any ordinary variables, thus the \ident{unit} type representing an empty
+tuple).
 
+In the body of the function, the context \ident{ctx1} is merged with the context provided by the
+variable \ident{x}. The type of \ident{ctx4} is $\ctyp{\ccrld{\{\ident{?param},\ident{?other}\}}}{(\ident{num}\times\ident{unit})}$.
+This is then split into two parts that contain just one of the implicit parameters and those are
+then accessed using \ident{lookup}.
+
+\paragraph{Composition in structural data-flow.}
+
+In structural coeffect systems, the translation works differently in that the context passed to
+a sub-expression contains only assignments for the variables used in the sub-expression (in the
+flat version, we always duplicated the variable context before using \ident{split}). To illustrate
+this, consider the following simple function:
+%
 \begin{equation*}
 \begin{array}{l}
-fun x -> (fun v -> prev v) (prev x)
+\kvd{fun}~\ident{x} \rightarrow \kvd{fun}~\ident{y} \rightarrow \kvd{prev}~\ident{x}
 \end{array}
 \end{equation*}
-
+%
+In structural coeffect systems, the comonadic value is annotated with a vector of coeffect
+annotations that correspond to individual variables. The initial structural input \kvd{sinput}
+is a value of type $\ctyp{\cclrd{[]}}{()}$ containing no variables (we write $()$ rather than
+\ident{unit} to make that more explicit). The translated code then looks as follows:
+%
 \begin{equation*}
 \begin{array}{l}
-fun x ->\\[-0.25em]
-  let ctx1 = merge (x, sinput) in\\[-0.25em]
-  (fun v ->\\[-0.25em]
-    let ctx3 = merge (v, choose_0 ctx1) in\\[-0.25em]
-    counit (prev ctx3)) (cobind (fun ctx2 ->\\[-0.25em]
-      counit (prev ctx2)) (choose_1 ctx1))
+\kvd{fun}~\ident{x} \rightarrow\\[-0.25em]
+\quad \kvd{let}~\ident{ctx1} ~=~ \ident{merge}~(\ident{x}, \kvd{sinput}) \kvd{in}\\[-0.25em]
+\quad (\kvd{fun}~\ident{y} \rightarrow\\[-0.25em]
+\quad \quad \kvd{let}~\ident{ctx2} ~=~ \ident{merge}~(\ident{y}, \ident{ctx1})~\kvd{in}\\[-0.25em]
+\quad \quad \ident{counit}~(\ident{prev}~(\ident{choose}_{\langle 0,1\rangle}~\ident{ctx2}))
 \end{array}
 \end{equation*}
+%
+The two variables are merged with the initial context, obtaining a value \ident{ctx2} of type
+$\ctyp{\cclrd{[0, 1]}}{\ident{num}\times\ident{num}}$ that contains two values with 0 and 1 past
+values, respectively.
 
-\section{Technical details}
-\label{sec:impl-tech}
+For simplicity, the implementation does not use the \ident{split}/\ident{merge} pair of operations
+of the structural coeffects to obtain the correct subset of variables. This can be done, but it
+would make the translated code longer and more cumbersome. Instead, we use a higher-level operation
+\ident{choose} (which can be expressed in terms of \ident{split}/\ident{merge}) that projects the
+variable subset as specified by the index. Here, $\langle 0, 1\rangle$ means that the first variable
+should be dropped an the second one should be kept.
 
-1. F# + JavaScript
-2. Single AST and parsing
-3. Type checking and constraint generation
-4. Constraint resolution
+The resulting single-variable context is then passed to \ident{prev} (to shift the stream by one)
+and then to \ident{counit} to obtain the current value.
+
+% ==================================================================================================
+%
+%    #######
+%    #        ####   ####    ##   #   #
+%    #       #      #       #  #   # #
+%    #####    ####   ####  #    #   #
+%    #            #      # ######   #
+%    #       #    # #    # #    #   #
+%    #######  ####   ####  #    #   #
+%
+% ==================================================================================================
 
@@
 \label{sec:impl-essay}
+
+As explained in the introduction of this chapter, the purpose of the implementation presented in
+this thesis is not to provide a real-world programming language, but to support the theory
+discussed in the rest of the thesis. The goal is to explain the theory and inspire authors of
+real-world programming langauges to include support for context-aware programming, ideally using
+coeffects as a sound foundation. For this reason, the implementation needs to be:
+
+\begin{itemize}
+\item {\sc Accessible.}  Anyone interested should be able to use the implemented langauges without
+  downloading the source code and compiling it and without installing specialized software.
+
+\item {\sc Explorable.}  It should be possible to explore the inner workings -- how is the typing
+  derived, how is the source code translated to the target language and how is it evaluated.
+\end{itemize}
+
+\noindent
+To make the work \emph{accessible}, we implement sample context-aware languages in a way that makes it
+possible to use them in any standard web browser with JavaScript support (Section~\ref{sec:impl-essay-tech})
+without requiring any server-side component. Following the idea that ``the medium is the message'' \cite{essay-medium},
+we choose medium that encourages \emph{exploration} and make the implementation available not just
+as source code that can be compiled and run locally, but also in the format of
+interactive essay (Section~\ref{sec:impl-essay-features}). The live version of the essay can be
+found at: \url{http://tomasp.net/coeffects}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=16em]{images/eval1.png}}
+\boxed{\includegraphics[width=16em]{images/eval2.png}}
+\caption{Interactive evaluation of implicit parameters (left) and data-flow (right)}
+\label{fig:essay-eval}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Explorable language implementation}
+\label{sec:impl-essay-features}
+
+The interactive essay format used of the implementation is inspired by Bret Victor's work on
+\emph{explorable explanations} \cite{essay-explorable}:
+%
+\begin{quote}
+\emph{Do our reading environments encourage active reading? Or do they utterly oppose it? A
+  typical reading tool, such as a book or website, displays the author's argument, and nothing
+  else. The reader's line of thought remains internal and invisible, vague and speculative. We
+  form questions, but can't answer them. We consider alternatives, but can't explore them. We
+  question assumptions, but can't verify them. And so, in the end, we blindly trust, or blindly
+  don't, and we miss the deep understanding that comes from dialogue and exploration.}
+\end{quote}
+%
+The interactive essay we present encourages active reading in the sense summarized in Victor's quote.
+We show the reader an example (program, typing derivation or translation), but the reader is
+encouraged to modify it and see how the explanation in the essay adapts.
+
+The idea of active reading is older and has been encouraged in the context of art Josef Albers'
+classic 1963 work on color \cite{essay-albers} (which has been turned into an interactive essay
+60 years later \cite{essay-albers-app}). More recently, similar formats have been used to explain
+topics in areas such as signal processing \cite{essay-seeing} (explaining Fourrier transformations)
+and sociology \cite{essay-polygons} (visualizing and explaining game theoretical model of segregation
+in the society \cite{essay-segregation}). To our best knowledge, no such work exists in the area
+of programming language theory and so we briefly outline some of the interesting features that
+our essay provides in order to encourage active reading.
+
+\paragraph{Interactive program excution.}
+
+After providing the practical motivation for coeffects (based on Chapter~\ref{ch:intro}), the
+interactive essay shows the reader the two sample context-aware programming languages. Readers can
+write code in a panel that type checks the input, generates user interface for entering the
+required context and runs the sample code.
+
+The panels for implicit parameters and for data-flow computations are shown in Figure~\ref{fig:essay-eval}.
+The sample code on the left adds two implicit parameters and the generated UI lets the user
+enter implicit parameter values as required by the context in the typing judgement. The sample on
+the right calculates the average of the current and past value in a data-flow; the UI lets the user
+enter past values required by the type of the function.
+
+The essay guides the readers through a number of interesting programs (shown in Section~\ref{sec:impl-case-typing})
+and encourages them to run and try modifying them. For implicit parameters, this includes the case
+where an implicit parameter is available at both declaration-site and call-site (showing that the
+declaration-site value is captured). For data-flow, the examples include the comparison between the
+inferred type when using flat and structural type systems.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=16em]{images/df1.png}}
+\boxed{\includegraphics[width=16em]{images/df2.png}}
+\caption{Function smoothing the X coordinate (left) with a sample run (right).}
+\label{fig:essay-df}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Reactive data-flow.}
+
+The interactive program execution lets the reader run sample programs, but not in a realistic
+context. To show a more real-world scenario, the essay includes a widget shown in Figure~\ref{fig:essay-df}.
+This lets the user write a function taking a stream of X and Y coordinates and calculate value
+based on the current and past values of the mouse pointer. The X and Y values, together with the
+result are plotted using a live chart.
+
+In the example run shown above, the sample program calculates the average of the last 12 values
+of the X coordinate (green line in Figure~\ref{fig:essay-df}). The example also illustrates one
+practical use of the coeffect type system -- when running, the widget keeps the coordinates in a
+pre-allocated fixed-size array, because the coeffect type system guarantees that at most 12 past
+values will be accessed.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=32em]{images/types.png}}
+\caption{Source code with type information and explorable typing derivation.}
+\label{fig:essay-typing}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Explorable typing derivations.}
+
+Perhaps the most important aspect of the implemented context aware programming languages is their
+coeffect and type system. In a conventional implementation, the functioning of the type system
+would remain mostly hidden -- we would get an ``OK'' message for a well-typed program and a type
+error for invalid program (likely not very informative, given that reporting good error messages
+for type errors is a notoriously hard problem even for mature language implementations).
+
+The presented interactive essay provides two features to help the reader understand and explore
+typing derivations. First, when reading code samples in the text, tooltips show the typing of
+individual identifiers, most interestingly functions (Figure~\ref{fig:essay-typing}, above).
+
+Second, a later part of the essay provides a type checker that lets the user enter a source code
+in a context aware programming language and produces an explorable typing derivation for the
+program. The output (shown in Figure~\ref{fig:essay-typing}, below) displays a typing judgement
+with assumptions and conclusions and lets the reader navigate through the typing derivation by
+clicking on the assumptions or conclusions. This way, the reader can see how is the final typing
+derivation obtained, exploring interesting aspects, such as the abstraction rule (shown in
+Figure~\ref{fig:essay-typing}).
+
+\paragraph{Comonadically-inspired translation.}
+
+The last interactive element of the essay lets the reader explore the translation of source
+context aware language to a target simple functional language (Section~\ref{sec:impl-theory-transl}).
+Compared with the monadic do-notation \cite{other-haskell98}, the comonadic translation is more
+complex for two reasons. First, it merges all variables into a single (comonadic) value representing
+the context. Second, there is a flat and structural variation of the system. For these two reasons,
+understanding the translation based just on the rules is harder than for monads. The essay lets
+readers see translation for carefuly chosen case studies that illustrate the key aspects of the
+translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
+programs encouraging \emph{active reading}. Sample input and output are shown in
+Figure~\ref{fig:essay-transl}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=16em]{images/transl1.png}}
+\boxed{\includegraphics[width=16em]{images/transl2.png}}
+\caption{TBD}
+\label{fig:essay-transl}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Implementation overview}
+\label{sec:impl-essay-tech}
+
+The core part of our implementation mostly follows standard techniques for implementing type
+checkers and interpreters for statically-typed functional programming langauges. Two interesting
+aspects that are worth discussing is the JavaScript targetting (for running the langauge
+implementation in a web browser) and integration with client-side (JavaScript) libraries for
+building the user interface. The full source code can be obtained from
+\url{https://github.com/coeffects/coeffects-playground} and is structured as follows:
+%
+\begin{itemize}
+\item Parsing is implemented using smple parser combinator library; \strf{ast.fs} defines the
+  abstract syntax tree for the languages; \strf{parsec.fs} implements a small parser combinator
+  library and \strf{lexer.fs} with \strf{parser.fs} parse the source code by first tokenizing
+  the input and then parsing the stream of tokens.
+
+\item The type checking is implemented by \strf{typechecker.fs}, which annotates an untyped AST
+  with types and generates set of type and coeffect constraints. The constraints are later solved
+  (using domain-specific algorithms for each of the languages) in \strf{solver.fs}.
+
+\item Type-checked programs in context-aware languages are translated to simple target
+  functional subset of the language in \strf{translation.fs}; \strf{evaluation.fs} then interprets
+  programs in the target language. The interpretation does not handle the source
+  language and so programs containing context-aware contstructs cannot be interpreted directly.
+
+\item The user interface of the interactive essay discussed in Section~\ref{sec:impl-essay-features}
+  is implemented partly in F\# and partly in JavaScript. The two most important components include
+  a pretty-printer \strf{pretty.fs} which formats source code (with type tooltips) and typing
+  derivations and \strf{gui.fs} that implements user interaction (e.g.~navigation in explorable
+  typing derivations).
+\end{itemize}
+
+\noindent
+As discussed in Section~\ref{sec:impl-theory-ext}, the implementation can be easily extended to
+support additional context-aware programming languages. This is due to the fact that it is based
+on the unified theory of coeffects. In practice, adding support for liveness tracking would
+require adding a domain-specific constraint solver in \strf{solver.fs} and extending the
+interpreter in \strf{evaluation.fs} with a new kind of comonadically-inspired data type
+(indexed maybe comonad) with its associated operations.
+
+\section{Related work}
+\label{sec:impl-related}
+
+To our best knowledge, combining Bret Victor's idea of explorable explanations with programming
+language theory (as discussed in Section~\ref{sec:impl-essay}) is a novel contribution of our work.
+On the technical side, we build on a number of standard methods.
+
+\paragraph{Parsing and type checking.} The implementation of the parser, type checker and
+interpreter follows standard techniques for implementing functional programming languages.
+In order to be able to compile the implementation to JavaScript (see below), we built a
+small parser combinator library \cite{monad-parsing} rather than using one of the already
+available libraries \cite{other-fparsec}.
+
+\paragraph{Targetting JavaScript.} In order to make the implementation accessible to a broad
+audience, it can be executed in a web browser. This is achieved by automatically translating the
+implementation from F\# to JavaScript. We use an F\# library called FunScript \cite{other-funscript}
+(which is a more recent incarnation of the idea developed by the author \cite{app-fsharp-webtools}).
+We choose F\#, but similar tools exist for other functional languages such as OCaml \cite{other-js-ocaml}.
+It is worth noting that FunScript is implemented as a \emph{library} rather than as a compiler
+extension. This is done using the meta-programming capabilities of F\# \cite{app-fsharp-metaprog}.
+
+\paragraph{Client-side library integration.} An interesting aspect of the interactive essay
+user interface is the integration with third-party JavaScript libraries. We use a number of
+libraries including JQuery (for web browser DOM manipulation) and MathJax \cite{essay-mathjax}
+(for rendering of typing derivation). In order to call those from F\# code, we use a number
+of mapping methods described in \cite{app-age-of-web}. For example, the following F\# declaration
+is used to invoke the \ident{Queue} function in MathJax:
+%
+\begin{equation*}
+\begin{array}{l}
+\lbrack\langle\ident{JSEmitInline}(\str{MathJax.Hub.Queue(\{0\});})\rangle\rbrack\\[-0.25em]
+\kvd{let}~\ident{queueAction}~(\ident{f}:\ident{unit}~\rightarrow~\ident{unit}):\ident{unit}~=~\ident{failwith}~\str{JS only!}
+\end{array}
+\end{equation*}
+%
+The \ident{JSEmitInline} syntax on the first line is an attribute that instructs FunScript
+to compile all invocations of the \ident{queueAction} function into the JavaScript literal
+specified in the attribute (with \strf{\{0\}} replaced by the argument).
+
+% ==================================================================================================
+%
+%     #####
+%    #     # #    # #    # #    #   ##   #####  #   #
+%    #       #    # ##  ## ##  ##  #  #  #    #  # #
+%     #####  #    # # ## # # ## # #    # #    #   #
+%          # #    # #    # #    # ###### #####    #
+%    #     # #    # #    # #    # #    # #   #    #
+%     #####   ####  #    # #    # #    # #    #   #
+%
+% ==================================================================================================
+
+\section{Summary}
+
+This chapter supplements the theory of coeffects presented in the previous three chapters with
+a prototype implementation. We implemented three simple context-aware programming langauges that
+track implicit parameters and past values in dataflow computations (in flat and structural way).
+
+The implementation discussed in this chapter provides an evidence for our claim that the
+theory of coeffects can be used as a basis for a wide range of sound context-aware programming
+languages. Our implementation consists of a shared coeffect framework (handling type checking and
+translation). Each context-aware language then adds a domain-specific rule for choosing a unique
+typing derivation, an interpretation of comonadically-inspired primitives and (optionally)
+domain-specific primitives such as the \kvd{prev} construct for dataflow.
+
+We make the implementation available in the usual form (as source code that can be downloaded,
+compiled and executed), but we also present it in the form of interactive essay. This encourages
+\emph{active reading} and lets the reader explore a number of aspects of the implementation
+including the type checking (through explorable typing derivations) and the translation. The key
+contribution of this thesis is that it provides a unified way for \emph{thinking} about context in
+programming langauges and the interactive presentation of the implementation is aligned with this
+goal. Programming languages of the future will need a mechanism akin to coeffects and we aim to
+provide a convincing argument supported by a prototype implementation.

### 2016-04-25 text/Chapter04.tex
Edits (Chapter 4)

@@
 \emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing
-per-variable properties. As we show in Section~X, the systems can be further unified using a
-single abstraction, but such abstraction is \emph{less powerful} -- \ie~its generality hides
-useful properties that we can see when we consider the systems separately. For this reason, we
-discuss \emph{flat coeffects} (Chapter~\ref{ch:flat} and Chapter~\ref{ch:semantics}) and
-\emph{structural coeffects} (Chapter~\ref{ch:structural}) separately.
+per-variable properties. As we show in Section~\ref{sec:unified-unified}, the systems can be
+further unified using a single abstraction, but such abstraction is \emph{less powerful} -- \ie~its
+generality hides useful properties that we can see when we consider the systems separately. For
+this reason, we discuss \emph{flat coeffects} (Chapter~\ref{ch:flat} and Chapter~\ref{ch:semantics})
+and \emph{structural coeffects} (Chapter~\ref{ch:structural}) separately.
 
@@
 required past values in dataflow languages.
-
 To capture contextual properties in full generality, the flat coeffect calculus permits multiple
@@
 each concrete context-aware language is also equipped with an algorithm for choosing a unique
-typing derivation. In the next chapter, we give operational meaning for concrete coeffect
+typing derivation. This allows us to explore the language design landscape, while still follow
+the usual scoping rules for languages with established approaches (e.g.~implicit parameters in
+Haskell).
+
+In the next chapter, we give operational meaning for concrete coeffect
 languages based on the flat coeffect calculus and we discuss their safety.
@@
 \item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent
-  in the conextual lambda abstraction rule. Each concrete context-aware language based on the
-  coeffect calculus may resolve this ambiguity differently. We show the procedure for choosing
-  unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).
-
-\item We discuss syntactic properties of the calculus, covering type-pre\-ser\-vation for call-by-name
+  in contextual lambda abstraction. Each concrete context-aware language based on the
+  coeffect calculus must specify how such ambiguities are to be resolved. We give the procedure for
+  choosing unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).
+
+\item We discuss equational properties of the calculus, covering type-pre\-ser\-vation for call-by-name
   and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
-  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections justify the laws
-  of the flat coeffect algebra.
+  with subtyping and pairs (Section~\ref{sec:flat-exts}).
 \end{itemize}
@@
 and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a similar
-structure in two ways. First, their lambda abstraction duplicates the demands. Given a body with
-context demands $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
-annotated with $\cclrd{r}$. Second, the context demands in the type systems are combined using
-two different operators (representing sequential and pointwise operations).
-
-The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl})
-differs. In lambda abstraction, it splits the context demands between the
-declaration site and the call site and they use only a single operator on the indices, typically $\cup$.
+structure in two ways. First, lambda abstraction duplicates their \emph{context demands}. Given a
+body with context demands $\cclrd{r}$, the declaration site context \emph{as well as} the function
+arrow are annotated with $\cclrd{r}$. Second, the context demands in the type systems are combined
+using two different operators (representing sequential and pointwise operations).
+
+The system for tracking implicit parameters (Section~\ref{sec:applications-flat-impl})
+differs. In lambda abstraction, it partitions the context demands between the
+declaration site and the call site. Furthermore, the operator that combines context demands
+is $\cup$ for both sequential and pointwise composition.
 
 Despite the differences, the systems fit the same framework. This becomes apparent when
-we consider the categorical structure (Section~\ref{sec:flat-semantics}). However, rather than
-starting from the semantics, we first explain how the systems can be unified syntactically
-(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as a justification.
+we consider the categorical structure (Section~\ref{sec:flat-semantics}). Rather than starting
+from the categorical semantics, we first explain how the systems can be unified syntactically
+(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as an additional
+justification.
 
 The development in this chapter can be seen as a counterpart to the well-known development of
-\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:semantics} then links \emph{coeffects}
-with \emph{comonads} in the same way in which effect systems can be linked with monads
+\emph{effect systems} \cite{effects-gifford}. Chapter~\ref{ch:semantics} then links \emph{coeffects}
+with \emph{comonads} in the same way in which effect systems have been linked with monads
 \cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows
 a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
-in the structure of lambda abstraction as discussed in Section~\ref{sec:applications-structure-lam}
+in the structure of lambda abstraction as discussed briefly here and in Section~\ref{sec:applications-structure-lam}
 (the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).
@@
 
-\subsection{Reconciling lambda abstraction}
+\subsection{A unified treatment of lambda abstraction}
 \label{sec:flat-calculus-lambda}
 
-Recall the lambda abstraction rules for the implicit parameters system (annotating the context
-with sets of required parameters) and the dataflow system (annotating the context with the
+Recall the lambda abstraction rules for the implicit parameters coeffect system (annotating
+contexts with sets of required parameters) and the dataflow system (annotating contexts with the
 number of past required values):
@@
 %
-In order to capture both systems using a single calculus, we need a way of unifying the two
-systems. For the dataflow system, this can be achieved by over-approximating the number of
-required past elements:
+In order to capture both systems using a single calculus, we need a way of rewriting the
+(\emph{df1}) rule such that the annotation in the assumption is in the form $\cclrd{n \disc m}$
+for some operation $\cclrd{\disc}$. For the dataflow system, this can be achieved by using
+the $\cclrd{\emph{min}}$ function:
 %
@@
 %
-The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule.
-Furthermore, if we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on
+The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule. That is,
+a typing derivation using (\emph{df1}) is also valid when using (\emph{df2}). Furthermore,
+if we include sub-typing rule (on annotations of functions) and subcoeffecting rule (on
 annotations of contexts), then the reverse is also true -- because
 $\cclrd{\textit{min}(n, m)} \leq \cclrd{m}$ and $\cclrd{\textit{min}(n, m)} \leq \cclrd{n}$.
-In other words (\emph{df1}) is more precise, but (\emph{df2}) gives a sound over-approximation
-with a structure that can be unified with (\emph{param}).
-
-Using a rule such as (\emph{df2}) allows us to give a unified formulation of the flat coeffect
-calculus in Section~\ref{sec:flat-calculus-types}, however the coeffect-specific handling of
-lambda abstraction remains important in practical implementation in order to obtain a unique
-typing derivation for each coeffect program as discussed in Section~\ref{sec:flat-unique}
-(and in the implementation in Chatper~\ref{ch:impl}).
+In other words (\emph{df2}) permits an implicit subcoeffecting (and sub-typing) that is not
+possible when using the (\emph{df1}) rule, but it has a structure that can be unified
+with (\emph{param}).
 
@@
 
-The \emph{flat coeffect calculus} is parameterized by a \emph{flat coeffect algebra}, which defines
-the structure of context annotations, written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
-sets of implicit parameters, versions represented as integers or other values.
-
-Typing judgements the attach coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$ to the context
-and are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$.
-
-%---------------------------------------------------------------------------------------------------
-
-\subsection{Understanding flat coeffects}
-\label{sec:flat-calculus-undestanding}
-
-Before looking at the typing rules, let us clarify how the typing judgements
-should be understood. The coeffect calculus provides both analysis of context dependence (type
-system) and semantics for context (how it is propagated). These two aspects provide different
-ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
-used to define it.
-
-\begin{itemize}
-\item \textsc{Analysis of context dependence.}
-Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context demands}. This means
-we can over-approximate them and require more than is actually needed at runtime.
-
-Syntactically, the typing rules are best read top-down. In function application, the context
-demands of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in
-lambda abstraction, the demands of a single expression (the body) are split between
-the declaration site and the call site.
-
-\item \textsc{Semantics of context passing.}
-Semantically, coeffect annotations $\cclrd{r}$ mo\-del \emph{contextual capabilities}. This means
-that we can throw away capabilities, if a sub-expression requires fewer than we
-currently have.
-
-Semantically, the typing rules should be read bottom-up. In application, the capabilities
-provided to the term $e_1~e_2$ are \emph{split} between the two sub-expressions; in abstraction,
-the capabilities provided by the call site and declaration site are \emph{merged} and passed
-to the body.
-\end{itemize}
-
-The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
-position} in the semantic model (Section~\ref{sec:semantics-theory}). It means that we need to be
-careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
-exactly the opposite things. To disambiguate, we always use the term \emph{context demands}
-when using the syntactic view, especially in the rest of Chapter~\ref{chf:flat}, and
-\emph{context capabilities} or just \emph{available context} when using the semantic view,
-especially in Chapter~\ref{ch:semantics}.
+This section describes the \emph{flat coeffect calculus}. A small programming language based on the
+$\lambda$-calculus with a type system that statically tracks context demands. The calculus can
+capture different notions of context. The structure of context demands is provided by a
+\emph{flat coeffect algebra} (defined in the next section) which is an abstract algebraic structure
+that can be instantiated to model concrete context demands (sets of implicit parameters, number
+of past values as integers or other information). Annotations that specify context demands are
+written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$.
+
+We enrich types and typing judgements with coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$;
+typing judgements are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$. The expressions of
+the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type
+\ident{num} as an example of a concrete base type with numerical constants written as $n$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+\end{array}
+\end{equation*}
+%
+Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
+annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
+unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
+we can directly focus on the more interesting problem of finding unique \emph{coeffect annotations}
+rather than solving the problem of type reconstruction.
+
+We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
+represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
+It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a
+context $\cclrd{r}$.
+
+In the categorically-inspired translation in the next chapter, the function
+$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function
+$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
+does not itself exist as a syntactic value in the coeffect calculus. This is because we use
+comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in
+the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
+The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.
 
@@
 To make the flat coeffect system general enough, the algebra consists of three operations.
-Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{pointwise} composition,
+Two of them, $\cseq$ and $\cpar$, represent \emph{sequential} and \emph{pointwise} composition,
 which are mainly used in function application. The third operator, $\czip$ is used in lambda
-abstraction and represents \emph{splitting} of context demands (or, semantically, \emph{merging}
-of available context capabilities).
-
-In addition to the three operations, we also require two special values used to annotate
+abstraction and represents \emph{splitting} of context demands.
+
+In addition to the three operations, the algebra also requires two special values used to annotate
 variable access and constant access and a relation that defines the ordering.
@@
 A \emph{\cclrd{flat coeffect algebra}} $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is a set
-$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations
+$\C$ together with elements $\cunit, \czero \in \C$, binary relation $\cleq$ and binary operations
 $\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
-idempotent monoid, $(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup).
+idempotent monoid, $(\C, \czip)$ is a band (idempotent semigroup) and $(\C, \cleq)$ is a pre-order.
 That is, for all $r,s,t\in \C$:
@@
 \item For the $\czip$ operation, we require associativity and idempotence. The idempotence
-demand makes it possible to duplicate the coeffects and place the same demand on both
+demand makes it possible to duplicate the given coeffects and place the same demand on both
 call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
@@
 are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
-(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.
-
-Using the syntactic reading, the two operators represent \emph{merging} and \emph{splitting} of context
-demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context
-demands of the body are split between two positions in the conclusions; in the (\emph{app}) rule,
-$\cpar$ appears in the conclusion and combines two context demands from the assumptions.
+(we discuss the lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.
 
@@
 
-The flat coeffect algebra requires a pre-order relation $\cleq$. This will be used to introduce
-sub-coeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
+The flat coeffect algebra includes a pre-order relation $\cleq$. This will be used to introduce
+subcoeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
 coeffect algebra, as it will be useful for characterization of different kinds of coeffect calculi.
-When the idempotent monoid $(\C, \cpar, \czero)$ also has the commutative property (\ie~forms a
+When the idempotent monoid $(\C, \cpar, \czero)$ is also commutative (\ie~forms a
 semi-lattice), the $\cleq$ relation can be defined as the ordering of the semi-lattice:
@@
 This definition is consistent with all three examples that motivate flat coeffect calculus, but
-it cannot be used with the structural coeffects (where it fails for the bounded reuse
-calculus) and so we choose not to use it.
+it cannot be used with the structural coeffects (Chapter~\ref{ch:structural}), where it fails for
+the bounded reuse calculus. For this reason, we choose not to use it for flat coeffect calculus
+either.
 
@@
 
-\subsection{Language syntax}
-\label{sec:flat-calculus-syntax}
-
-The expressions of
-the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type
-\ident{num} as an example of a concrete base type with numerical constants written as $n$:
-%
-\begin{equation*}
-\begin{array}{rcl}
-e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
-\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
-\end{array}
-\end{equation*}
-%
-Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
-annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
-unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
-we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations}
-rather than solving the problem of type reconstruction.
-
-We discuss subtyping and pairs in Section~\ref{sec:flat-exts}. The type $\tau_1 \xrightarrow{\cclrd{r}} \tau_2$
-represents a function from $\tau_1$ to $\tau_2$ that requires additional context $\cclrd{r}$.
-It can be viewed as a pure function that takes $\tau_1$ \emph{with} or \emph{wrapped in} a
-context $\cclrd{r}$.
-
-In the categorically-inspired translation in the next chapter, the function
-$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function
-$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
-does not itself exist as a syntactical value in the coeffect calculus. This is because we use
-comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in
-the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
-The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.
-
-%---------------------------------------------------------------------------------------------------
-
 \begin{figure}[t]
@@
 \tyrule{var}
-  {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\cunit} \vdash x : \tau }
+  {~}
+  {\coctx{\Gamma}{\cunit} \vdash x : \tau }\quad(x : \tau \in \Gamma)
 \end{equation*}
@@
 \end{equation*}
+%
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad(\cclrd{r'} \cleq \cclrd{r})
+\end{equation*}
+%
 
@@
 between the context-demands on the declaration site $\cclrd{r}$ and context-demands on
-the call site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.
+the call site $\cclrd{s}$.
 
 In function application (\emph{app}), context demands of both expressions and the
-function are combined. As discussed in Chapter~\ref{ch:applications}, the pointwise composition
-$\cpar$ is used to combine the context demands of the expression representing a function
-$\cclrd{r}$ and the context demands of the argument, sequentially composed with the
-context-demands of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.
+function are combined. As discussed in Chapter~\ref{ch:applications}, sequential composition
+is used to combine the context-demands of the argument $\cclrd{s}$ with the context-demands
+of the function $\cclrd{t}$. The result $\cclrd{s}\,\cseq\,\cclrd{t}$ is then composed using
+pointwise composition with the context demands of the expression that represents the function
+$\cclrd{r}$, giving the coeffect $\cclrd{r}\,\cpar\,(\cclrd{s}\cseq\cclrd{t})$.
 
 The type system also includes a rule for let-binding. The rule is \emph{not} equivalent to the
-typing derivation for $(\lambda x.e_2)~e_1$, but it corresponds to \emph{one} possible typing
+derived rule for $(\lambda x.e_2)~e_1$, but it corresponds to \emph{one} possible typing
 derivation. As we show in \ref{sec:flat-exts-let}, the typing used in (\emph{let}) is more
-precise than the general rule that can be derived from $(\lambda x.e_2)~e_1$. Additional
-constructs such as pairs, sub-coeffecting and sub-typing are covered in Section~\ref{sec:flat-exts}.
+precise than the general rule that can be derived from $(\lambda x.e_2)~e_1$.
+
+To guide understanding of the system, we also show non-syntax-directed (\emph{sub}) rule for
+subcoeffecting. The rule states that an expression with context demands $\cclrd{r'}$ can be
+treated as an expression with greater context demands $\cclrd{r}$. We return to subcoeffecting,
+subtyping and additional constructs such as pairs in Section~\ref{sec:flat-exts}. When
+discussing procedure for choocing unique typing in Section~\ref{sec:flat-unique}, we consider
+only the syntax-directed part of the system.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Understanding flat coeffects}
+\label{sec:flat-calculus-undestanding}
+
+Before proceeding, let us clarify how the typing judgements should be understood. The coeffect
+calculus can be understood in two ways discussed in this and the next chapter. As a type system
+(Chapter~\ref{ch:flat}), it provides analysis of context dependence. As a semantics
+(Chapter~\ref{ch:semantics}), it specifies how context is propagated. These two readings provide
+different ways of interpreting the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the
+typing rules used to define it.
+
+\begin{itemize}
+\item \textsc{Analysis of context dependence.}
+Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context demands}. This means
+we can over-approximate them and require more in the type system than is needed at runtime.
+
+Syntactically, the typing rules are best read top-down (from assumptions to the consequent).
+In function application, the context demands of multiple assumptions (arising from two
+sub-expressions) are \emph{merged}; in lambda abstraction, the demands of a single expression
+(the body) are split between the declaration site and the call site.
+
+\item \textsc{Semantics of context passing.}
+Semantically, coeffect annotations $\cclrd{r}$ mo\-del \emph{contextual capabilities}. This means
+that we can throw away capabilities, if a sub-expression requires fewer than we
+currently have.
+
+Semantically, the typing rules should be read bottom-up (from the consequent to assumptions).
+In application, the capabilities provided to the term $e_1~e_2$ are \emph{split} between the
+two sub-expressions; in abstraction, the capabilities provided by the call site and declaration
+site are \emph{merged} and passed to the body.
+\end{itemize}
+
+\noindent
+For example, using the syntactic reading, the operators $\czip$ and $\cpar$ represent \emph{merging}
+and \emph{splitting} of context demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption
+and the combined context demands of the body are split between two positions in the conclusions;
+in the (\emph{app}) rule, $\cpar$ appears in the conclusion and combines two context demands
+from the assumptions.
+
+The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
+position} in the semantic model (Section~\ref{sec:semantics-theory}). It means that we need to be
+careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
+exactly the opposite things. To disambiguate, we always use the term \emph{context demands}
+when using the syntactic view, especially in the rest of Chapter~\ref{ch:flat}, and
+\emph{context capabilities} or just \emph{available context} when using the semantic view,
+especially in Chapter~\ref{ch:semantics}.
 
@@
 
-The flat coeffect calculus generalizes the flat systems discussed in
+The flat coeffect calculus generalizes the three flat systems discussed in
 Section~\ref{sec:applications-flat} of the previous chapter. We can instantiate it to a specific
-use just by providing a flat coeffect algebra. The following summary defines the systems for implicit
-parameters, liveness and dataflow. For the latter two, the general calculus has a lambda abstraction
-that is compatible with those discussed in Chapter~\ref{ch:flat}, but includes implicit sub-coeffecting.
+use just by providing a flat coeffect algebra.
 
 \begin{example}[Implicit parameters]
-Assuming \ident{Id} is a set of implicit parameter names, the flat coeffect algebra
-is formed by $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset, \emptyset, \subseteq)$.
+\label{ex:flat-impl}
+Assuming \ident{Id} is a set of implicit parameter names written $\ident{?p}$, the flat coeffect
+algebra is formed by $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset, \emptyset, \subseteq)$.
 \end{example}
@@
 \noindent
-For simplicity, we assume that all parameters have the same type $\ident{num}$ and so the annotations only
+For simplicity, assume that all parameters have the same type $\ident{num}$ and so the annotations only
 track sets of names. The definition uses a set union for all three operations. Both variables and
@@
 %
-The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of required
-parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all parameters, but
-it is possible to define a coeffect calculus that uses mappings from names to types (care is needed to
-avoid assigning multiple types to a parameter of the same type).
+The (\emph{param}) rule specifies that the accessed parameter $\ident{?p}$ needs to be in the set of
+required parameters $\cclrd{r}$. As discussed earlier, we use the same type $\ident{num}$ for all
+parameters, but it is also possible to define a coeffect calculus that uses mappings from names to
+types (care is needed to avoid assigning multiple types to a parameter of the same type).
 
@@
 Let $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ be a two-point lattice such that $\ident{D} \sqsubseteq \ident{L}$
-with a join $\sqcup$ and meet $\sqcap$. The flat coeffect algebra for liveness is then formed by
+with join $\sqcup$ and meet $\sqcap$. The flat coeffect algebra for liveness is then formed by
 $(\mathcal{L}, \sqcap, \sqcup, \sqcap, \ident{L}, \ident{D}, \sqsubseteq)$.
@@
 marked as live $\ident{L}$. When the body is dead $\ident{D}$, the declaration site and call site
-can be marked as dead $\ident{D}$, or as live $\ident{L}$. The latter is less precise, but
-permissible over-approximation, which could otherwise be obtained via sub-typing.
+can be marked as dead $\ident{D}$, or as live $\ident{L}$. The latter is less precise, but it
+is a valid derivation that could also be obtained via sub-typing.
 
@@
 both call site and declaration site annotations are equal or greater than the annotation
-of the body. As with liveness, this allows over-approximation.
+of the body.
 
@@
 As a further example that was not covered earlier, it is also possible to combine liveness analysis
-and dataflow. In the above calculus, $0$ denotes that we require the current value, but no previous
-values. However, for constants, we do not even need the current value.
+and dataflow. In the above dataflow calculus, $0$ denotes that we require the current value of some
+variable, but no previous values. However, for constants, we do not even need the current value.
 
@@
 
-\section{Choosing unique typing}
+\section{Choosing a unique typing}
 \label{sec:flat-unique}
@@
 Furthermore, the semantics of context-aware languages in Chapter~\ref{sec:applications} and
-also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivation} and so the same program
-could have a different meaning, depending on the typing derivation chosen. In this section, we
-specify how to choose \emph{unique} typing derivation in each of the coeffect systems we consider.
+also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivation} and so the meaning
+of a program depends on the typing derivation chosen. In this section, we specify how to choose
+the desired \emph{unique} typing derivation in each of the coeffect systems we consider.
 
@@
 \begin{array}{l}
-\kvd{let}~f~= (\kvd{let}~\ident{?x}=1~\kvd{in}~(\lambda y.~\ident{?x}))~\kvd{in} \\[-0.2em]
-\kvd{let}~\ident{?x}=2~\kvd{in}~f~0 \\[-0.2em]
+\kvd{let}~f~= (\kvd{let}~\ident{?x}=42~\kvd{in}~(\lambda y.~\ident{?x}))~\kvd{in} \\[-0.2em]
+\kvd{let}~\ident{?x}=666~\kvd{in}~f~0 \\[-0.2em]
 \end{array}
@@
 The coeffect calculus intentionally allows both of the options, acknowledging the fact that the
-choice needs to be made for each individual concrete context-aware programming language. In this
-section, we discuss the choices for implicit parameters, dataflow and liveness.
-
-In this section, we use the fact that the coeffect calculus uses Church-style syntax for lambda
-abstraction and has a type annotation for the type of the variable. This does not affect the
+choice needs to be made for each individual concrete context-aware programming language. In the
+above case, one typing derivation represents dynamic binding and the other static binding, but
+more subtleties arise when the nested expression uses multiple implicit parameters.
+
+In this section, we discuss the specific choices of typing derivation for implicit parameters,
+dataflow and liveness. We use the fact that the coeffect calculus uses Church-style syntax for
+lambda abstraction giving a type annotation for the bound variable. This does not affect the
 handling of coeffects (those are not defined by the type annotation), but it lets us prove
-uniqueness typing property of the specialized coeffect type systems. This shows that we define a
-\emph{unique} way of assigning coeffects to otherwise well-typed programs.
+\emph{uniqueness of typing}; a theorem showing that we define a \emph{unique} way of assigning
+coeffects to otherwise well-typed programs.
 
@@
 
-For implicit parameters we follow the behaviour implemented by Haskell \cite{app-implicit-parameters}
-where function abstraction captures all parameters that are available at the declaration-site and
-places all other demands on the call-site. For the example in the introduction, this means
+For implicit parameters we choose to follow the behaviour implemented by Haskell~\cite{app-implicit-parameters}
+where function abstraction captures all parameters that are statically available at the declaration-site
+and places all other demands on the call-site. For the example above, this means
 that the body of $f$ captures the value of \ident{?p} available from the declaration-site
 and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The
-program thus evaluates to $1$.
+program thus evaluates to a numerical value $42$.
 
 To express this behaviour formally, we extend the coeffect type system to additionally track
-implicit parameters that are currently in scope. The typing judgement becomes:
+implicit parameters that are currently in static scope. The typing judgement becomes:
 %
@@
 The coeffect on the declaration site becomes $\cclrd{\Delta}$ (capture all available parameters)
-and the latent coeffect attached to the function becomes $\cclrd{r\setminus\Delta}$ (require all
+and the latent coeffect attached to the function becomes $\cclrd{r\setminus\Delta}$ (require any
 remaining parameters from the call-site). Finally, in the (\emph{letpar}) rule, we add the newly
@@
 \paragraph{Properties.}
-If a program written in a coeffect language with implicit parameters is well-typed according to the
-type system presented in Figure~\ref{fig:flat-resolve-impl}, then the type system gives a unique
-derivation. We use this unique typing derivation to give the semantics of coeffect language
-with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm as
-discussed in Chatper~\ref{ch:impl}.
+If a program written in a coeffect language with implicit parameters is well-typed in a type
+system presented in Figure~\ref{fig:flat-resolve-impl} then this identifies the unique prefered
+derivation for the program. We use this unique typing derivation to give the semantics of coeffect
+language with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm
+as discussed in Chatper~\ref{ch:impl}.
 
@@
 
-In order to prove the uniqueness of typing theorem (Theorem~\ref{thm:flat-impl-unique}), we first
-need the inversion lemma (Lemma~\ref{thm:flat-impl-invert}).
+In order to prove the uniqueness of typing theorem (Theorem~\ref{thm:flat-impl-unique}), we follow
+the standard approach \cite{syntactic-tapl} and first give the inversion lemma (Lemma~\ref{thm:flat-impl-invert}).
 
@@
 
+\noindent
+Finally, we note that unique typing derivations obtained using the type system given in
+Figure~\ref{fig:flat-resolve-impl} are valid typing derivation under the original flat coeffect
+type system in Figure~\ref{fig:flat-types}.
+
+\begin{theorem}[Admisibility of unique typing for implicit parameters]
+\label{thm:flat-impl-admis}
+  If $\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash e : \tau$ (using the rules in
+  Figure~\ref{fig:flat-resolve-impl}) then also $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
+  (using the rules in Figure~\ref{fig:flat-types} and Example~\ref{ex:flat-impl}).
+\end{theorem}
+\begin{proof}
+  Each typing rule in the unique type derivation is a special case of the corresponding
+  typing rule in the flat coeffect calculus (ignoring the additional context $\cclrd{\Delta}$); the
+  splitting of coeffects in (\emph{abs}) in Figure~\ref{fig:flat-resolve-impl} is a special case
+  of splitting two sets using $\cclrd{\cup}$.
+\end{proof}
+
 %---------------------------------------------------------------------------------------------------
@@
 
-For livenes and dataflow, the (\emph{idabs}) rule provides the most precise coeffect.
-Assume we have lambda abstraction with body that has coeffects $\cclrd{r}$. The
+For liveness and dataflow, the (\emph{idabs}) rule provides the most precise coeffect.
+Assume we have a lambda abstraction with a body that has coeffects $\cclrd{r}$. The
 ordinary (\emph{abs}) rule requires us to find $\cclrd{s}, \cclrd{t}$ such that
@@
     $\cclrd{s}, \cclrd{t}$ such that the equality holds are $\cclrd{s}=\cclrd{t}=\cclrd{r}$.
- \item[--] For livenes, this is $\cclrd{r}=\cclrd{s} \sqcup \cclrd{t}$. When $\cclrd{r}=\ident{L}$,
+ \item[--] For liveness, this is $\cclrd{r}=\cclrd{s} \sqcup \cclrd{t}$. When $\cclrd{r}=\ident{L}$,
    the only solution is $\cclrd{s}=\cclrd{t}=\ident{L}$; when $\cclrd{r}=\ident{D}$, the most
@@
 
-The notion of ``more precise'' solution can be defined in terms of sub-coeffecting and subtyping.
+The notion of ``more precise'' solution can be defined in terms of subcoeffecting and subtyping.
 We return to this topic in Section~\ref{sec:flat-exts-lambda} and we also precisely characterise
@@
 implementation (Chatper~\ref{ch:impl}).
+
+We note that the unique typing derivation is admissible in the original coeffect type system.
+For dataflow and liveness, this follows directly from the fact that (\emph{idabs}) is a special
+case of the (\emph{abs}) rule and so we do not state this explicitly as in
+Theorem~\ref{thm:flat-impl-admis} for implicit parameters.
 
@@
 Before discussing the syntactic properties of general coeffect calculus formally, it should be
-clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
+clarified what is meant by providing a ``pathway to operational semantics'' in this section. We do
 that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
-Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
+Writing $\subst{e_1}{x}{e_2}$ for a standard capture-avoiding syntactic substitution, the following
 equations define four syntactic reductions on the terms:
@@
 (type preservation), then operational semantics can be defined as a repeated application of the
-rules, until a specified normal form (\ie~a value) is reached.
+rules, together with additional domain-specific rules for each context-aware langauge, until
+a specified normal form (\ie~a value) is reached.
 
@@
 
-The focus of this chapter is on the general coeffect system and so we do not discuss the operational
-semantics of the specific notions of context. However, some work in that area has been done
-by Brunel et al. \cite{coeffects-quantitative}. We discuss concrete semantics of implicit
-parameters and dataflow in Chapter~\ref{ch:semantics}.
+The focus of this chapter is on the general coeffect system and so we do not discuss the
+domain-specific reduction rules for individual context-aware language. Some work on operational
+semantics of general coeffect systems has been done by Brunel et al. \cite{coeffects-quantitative}.
+We give formal semantics of implicit parameters and dataflow in Chapter~\ref{ch:semantics} by
+translation to a simple functional programming language instead.)
 
@@
 The call-by-value evaluation strategy is often used by effectful languages. Here, an argument is
-first reduced to a \emph{value} before performing the substitution. In effectful languages,
-value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
+first reduced to a \emph{value} before performing the substitution. In effectful languages, the notion
+of value is defined syntactically. For example, in the \emph{Effect} language \cite{monads-effects-marriage},
 values are identifiers $x$ or functions $(\lambda x.e)$.
@@
 immediate context demands. Thus we say that $e$ is a value if it is a value in the usual
-sense \emph{and} has not immediate context demands. We define this formally in
+sense \emph{and} has no immediate context demands. We define this formally in
 Section~\ref{sec:flat-syntax-cbv}.
@@
 The call-by-value evaluation strategy preserves typing for a wide range of flat coeffect calculi,
-including all our three examples. However, it is rather weak -- in order to use it, the concrete
-semantics needs to provide a way for reducing context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
+including all our three examples. However, it is rather weak -- in order to use it, the domain-specific
+semantics needs to provide a way for reducing a context-dependent term $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
 to a value, \ie~a term $\coctx{\Gamma}{\cunit} \vdash e' : \tau$ with no context demands.
-
-\paragraph{Local soundness and completeness.}
-Two desirable properties of calculi, coined by Pfenning and Davies \cite{logic-modal-reconstruction},
-are \emph{local soundness} and \emph{local completeness}. They guarantee that the rules which
-introduce a function arrow (lambda abstraction) and eliminate it (application) are sufficiently
-strong, but not too strong.
-
-The local soundness property is witnessed by (call-by-name) $\beta$-reduction, which we discussed
-already. The local completeness is witnessed by the $\eta$-expansion rule. We discuss the flat
-coeffect algebra conditions under which the reduction holds in Section~\ref{sec:flat-syntax-cbn}.
 
@@
 As discussed in the previous section, call-by-value reduction can be used for most flat coeffect
-calculi, but it provides a very weak general model \ie~the hard work of reducing context-dependent
+calculi, but it provides a very weak general model. The hard work of reducing a context-dependent
 term to a \emph{value} has to be provided for each system. Syntactic values are defined
@@
 The syntactic form \emph{SynVal} captures syntactic values, but a context-dependency-free value in
-coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ does
-not automatically defer all context demands.
+coeffect calculus cannot be defined purely syntactically, because a function $(\lambda x.e)$ may
+still have context demands -- for example a function $(\lambda x.\kvd{prev}~x)$ has an
+immediate context demand $1$ (requiring 1 past value of all variables in the context).
 
@@
 \noindent
-The call-by-value substitution substitutes a value, with context demands $\cunit$, for a
+Call-by-value substitution substitutes a value, with context demands $\cunit$, for a
 variable, whose access is also annotated with $\cunit$. Thus, it does not affect the type and
@@
 By induction over the type derivation, using the fact that $x$ and $v$ are annotated
-with $\cunit$ and that $\Gamma$ is treated as a set in the flat calculus.
+with $\cunit$ and that variables are never removed from the set $\Gamma$ in the flat coeffect
+calculus.
 \end{proof}
@@
 flat coeffect algebra, which relates the $\czip$ and $\cpar$ operations. This is captured
-by the (\emph{approximation}) property:
+by the \emph{approximation} property:
 %
@@
 \label{thm:cbv-reduction}
-In a flat coeffect system with the (\emph{approxi\-mation}) property,
+In a flat coeffect system satisfying the approximation property, that is
+$\cclrd{r}\,\czip\,\cclrd{t}~\;\cleq~\;\cclrd{r}\,\cpar\,\cclrd{t}$,
 if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
@@
 From Lemma~\ref{thm:flat-subst-cbv}, $\subst{e}{x}{v}$ has the same coeffect annotation as $e$.
-As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply sub-coeffecting:
+As $\cclrd{r}\,\czip\,\cclrd{t}\;\cleq\;\cclrd{r}\,\cpar\,\cclrd{t}$, we can apply subcoeffecting:
 %
@@
 
-The typing is preserved for different reasons in two of our systems, so we briefly review the
+Typing is preserved for different reasons in two of our systems, so we briefly review the
 concrete examples. Then, we prove the substitution lemma for two special cases of flat coeffects
 (Lemma~\ref{thm:cbn-substitution-top} and Lemma~\ref{thm:cbn-substitution-bot}) and finally, we
-state the conditions under which typing preservation hold for flat coeffect calculi
+state the conditions under which typing preservation holds for flat coeffect calculi
 (Theorem~\ref{thm:cbn-flat}).
@@
 \paragraph{Dataflow.}
-The type preservation property does not hold for dataflow. This case is similar to the example
-shown earlier with effectful computations. As a minimal example, consider the substitution of
+Reducing an expression $(\lambda x.e_1)~e_2$ to  $\subst{e_1}{x}{e_2}$ does not always
+preserve the type of the expression in dataflow languages. This case is similar to the example shown
+earlier with effectful computations. As a minimal example, consider the substitution of
 a context-dependent expression $\kvd{prev}~z$ for a variable $y$ in a function $\lambda x.y$:
@@
 Semantically, the coeffect over-approximates the actual demands -- at run-time, the code does not
-actually access a previous value of the argument $x$. This cannot be captured by a flat coeffect
-system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
-
-\paragraph{Implicit parameters.} In dataflow, there is no typing for the resulting expression that
-preserves the type of the function. However, this is not the case for all systems. Consider substituting
+actually access a previous value of the argument $x$. This fact cannot be captured by a flat coeffect
+type system, but it can be captured using the structural system discussed in Chapter~\ref{ch:structural}.
+
+\paragraph{Implicit parameters.} In dataflow, substituting $\kvd{prev}~x$ for a variable $y$ in
+an expression $\lambda z.y$ changes the context demands attached to the type of the function.
+This is the case not just for the preferred unique typeing derivation, but for all possible typings
+that can be obtained using the (\emph{abs}) rule. However, this is not the case for all
+systems. Consider a substitution $\subst{\lambda x.y}{y}{\ident{?p}}$ that substitutes
 an implicit parameter access $\ident{?p}$ for a free variable $y$ under a lambda:
@@
 %
-The above shows one possible typing of the body -- one that does not change the coeffects of the
-function type and attaches all additional coeffects (implicit parameters) to the context. In case
-of implicit parameters (and, more generally, systems with set-like annotations) this is always
-possible.
+The (\emph{after}) judgement shows one possible typing of the body -- one that does not change the
+coeffects of the function type and attaches all additional coeffects (implicit parameters) to the
+context. In case of implicit parameters (and, more generally, systems with set-like annotations)
+this is always possible.
 
 \paragraph{Liveness.} In liveness, the type preservation also holds, but for a different reason. Consider
-substituting an arbitrary expression $e$ of type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
+a substitution $\subst{\lambda x.y}{y}{e}$ that substitutes an arbitrary expression $e$ of
+type $\tau_1$ with coeffects $\cclrd{r}$ for a variable $y$:
 %
@@
 
-\paragraph{Reduction theorem.}
+\paragraph{A grand CBN reduction theorem.}
 The above examples (implicit parameters and liveness) demonstrate two particular kinds of coeffect
-algebra for which typing preservation holds. Proving the type preservation separately provides
-more insight into how the systems work. We consider the two cases separately, but find a more general
-formulation for both of them.
+algebra for which call-by-name reduction preserves typing. Proving the type preservation separately
+provides more insight into how the systems work. We consider the two cases separately, but find a
+more general formulation for both of them.
 
@@
 In a top-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
-substituting an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$ does
-not change the coeffects of $e_r$:
+when we substitute  an expression $e_s$ with arbitrary coeffects $\cclrd{s}$ for a variable $x$ in $e_r$,
+the resulting expression is still typeable in a context with the original coeffect of $e_r$:
 %
@@
 \begin{proof}
-Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
+Using subcoeffecting ($\cclrd{s} \,\cleq\, \cunit$) and a variation of Lemma~\ref{thm:flat-subst-cbv}.
 \end{proof}
@@
 As variables are annotated with the top element $\cunit$, we can substitute the term $e_s$
-for any variable and use sub-coeffecting to get the original typing (because
+for any variable and use subcoeffecting to get the original typing (because
 $\cclrd{s} \,\cleq\, \cunit$).
@@
 $\Gamma$, leaving the context demands of a function value unchanged. As a result, substitution
-does not break soundness as in effect systems. The demand $\czip = \cseq = \cpar$ holds for
-our implicit parameters example (all three operators are a set union) and for other set-like
-coeffects. It allows the following substitution lemma:
+does not break soundness as in effect systems. The requirement $\czip = \cseq = \cpar$ holds for
+our implicit parameters example (all three operators are a set union) and for other coeffect
+systems that track sets of context demands discussed in Section~\ref{sec:applications-flat-distr}.
+It allows the following substitution lemma:
 
@@
 However, $\eta$-expansion preserves the type for implicit parameters and, more generally,
-for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$ is the
-standard $\eta$-reduction:
+for any flat coeffect algebra where $\cpar \,=\, \czip$. Assuming $\rightarrow_\eta$
+performs an expansion that turns a function-typed term $e$ to a
+syntactic function $\lambda x.e~x$, the following theorem holds:
 
@@
 original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
-is showing just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
-this is sufficient for showing type preservation.
+shows just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
+this is sufficient for type preservation.
 \end{proof}
@@
 there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
-these include liveness and implicit parameters. Moreover, for implicit parameters (and more
-generally, any set-like flat coeffect algebra), the $\eta$-expansion holds as well, giving us
-both local soundness and local completeness as coined by Pfenning and Davies \cite{logic-modal-reconstruction}.
+these include liveness and implicit parameters. Moreover, for implicit parameters the
+$\eta$-expansion holds as well, giving us both local soundness and local completeness as coined
+by Pfenning and Davies \cite{logic-modal-reconstruction}.
 
@@
 The laws are required for three reasons -- to be able to define the categorical structure
-in Section~\ref{sec:flat-semantics}, to prove equational properties in Section~\ref{sec:flat-syntax}
+in Section~\ref{sec:semantics-theory}, to prove equational properties in Section~\ref{sec:flat-syntax}
 and finally, to guarantee intuitive syntactic properties for constructs such as
@@
 
-In this section, we look at the last point. We consider sub-coeffecting and subtyping
+In this section, we turn to the last point. We consider subcoeffecting and subtyping
 (Section~\ref{sec:flat-exts-sub}), discuss what syntactic equivalences are permitted
@@
 The \emph{flat coeffect algebra} includes the $\cleq$ relation which captures the ordering of
-coeffects and can be used to define sub-coeffecting. Syntactically, an expression with context
-demands $\cclrd{r}$ can be treated as an expression with greater context demands:
+coeffects and can be used to define subcoeffecting. Syntactically, an expression with context
+demands $\cclrd{r'}$ can be treated as an expression with a greater context. This is captured
+by the (\emph{sub}) rule shown in Figure~\ref{fig:flat-types} (recall that for implicit
+parameters $\cleq = \cclrd{\subseteq}$):
 %
@@
 %
-Semantically, this means that we can \emph{drop} some of the provided context. For example,
-if an expression requires implicit parameters $\{ \ident{?p} \}$ it can be treated as requiring
-$\{ \ident{?p}, \ident{?q} \}$. The semantic function will then be provided with a dictionary
-containing both assignments and it can drop the value for the unused parameter \ident{?q}.
-
-Sub-coeffecting only affects the immediate coeffects attached to the free-variable context.
+Semantically, when read from the consequent to the assumption, this means that we can \emph{drop}
+some of the provided context. For example, if an expression requires implicit parameters
+$\{ \ident{?p} \}$ it can be treated as requiring $\{ \ident{?p}, \ident{?q} \}$. The semantic
+function will then be provided with a dictionary containing both assignments and it can ignore
+(or even actively drop) the value for the unused parameter \ident{?q}.
+
+Subcoeffecting only affects the immediate coeffects attached to the free-variable context.
 In Figure~\ref{fig:flat-types-sub}, we add sub-typing on function types, making it possible to treat
@@
 \begin{equation*}
-\tyrule{typ}
+\tyrule{sub-typ}
   { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau & \tau <: \tau' }
@@
 shows, the function type is contra-variant in the input and co-variant in the output. The
-(\emph{typ}) rule allows using sub-typing on expressions in the coeffect calculus.
+(\emph{sub-typ}) rule allows using sub-typing on expressions in the coeffect calculus.
 
@@
 $\kvd{let}~x=e_1~\kvd{in}~e_2$ with context demands $\cclrd{s}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{r})$.
-This is a special case of typing an expression $(\lambda x.e_2)~e_1$, using the idempotence
-of $\czip$ as follows:
+This rule can be derived from the typing derivation for an expression $(\lambda x.e_2)~e_1$ as
+a special case, by using the idempotence of $\czip$ as follows:
 %
@@
 %
-This design decision is similar to ML value restriction, but it works the other way round. Our
-\emph{let} binding is more restrictive than the typing of abstraction-application, rather than
-being more general. The choice is motivated by the fact that the typing obtained using the special
-rule for let-binding is more precise for all the examples consider
-in this chapter. Table~\ref{tab:flat-simplelet} shows how the coeffect annotations are simplified
-for our examples.
+This design decision is motivated by the fact that the typing obtained using the special rule for
+let-binding is more precise for all the examples consider in this chapter. To see this, assume
+an arbitrary splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Table~\ref{tab:flat-simplelet}
+shows the coeffect annotation derived from $(\lambda x.e_2)~e_1$, the coeffect annotation obtained
+by the (\emph{let}) rule and the simplified coeffect annotation using the particular flat coeffect
+algebras.
 
@@
 \begin{center}
-\begin{tabular}{ | l | c | c |}
+\begin{tabular}{ | l | c | c | c |}
 \hline
-& \textbf{\footnotesize Definition\hspace{1em}} & \textbf{\footnotesize Simplified\hspace{1em}} \\ \hline
-\hspace{-1em}{\footnotesize Implicit parameters} & $\cclrd{s} \cup (\cclrd{s} \cup \cclrd{r})$ & $\cclrd{s} \cup \cclrd{r}$ \\ \hline
-\hspace{-1em}{\footnotesize Liveness} & $\cclrd{s} \sqcap (\cclrd{s} \sqcup \cclrd{r})$ & $\cclrd{s}$ \\ \hline
-\hspace{-1em}{\footnotesize Dataflow} & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
+& \textbf{\footnotesize Derived \hspace{1em}} & \textbf{\footnotesize Definition\hspace{1em}} & \textbf{\footnotesize Simplified\hspace{1em}} \\ \hline
+\hspace{-1em}{\footnotesize Implicit parameters} & $\cclrd{s_1} \cup (\cclrd{s_2} \cup \cclrd{r})$ & $\cclrd{s} \cup (\cclrd{s} \cup \cclrd{r})$ & $\cclrd{s} \cup \cclrd{r}$ \\ \hline
+\hspace{-1em}{\footnotesize Liveness} & $\cclrd{s_1} \sqcap (\cclrd{s_2} \sqcup \cclrd{r})$ & $\cclrd{s} \sqcap (\cclrd{s} \sqcup \cclrd{r})$ & $\cclrd{s}$ \\ \hline
+\hspace{-1em}{\footnotesize Dataflow} & $\mathit{max}(\cclrd{s_1}, \cclrd{s_2} + \cclrd{r})$ & $\mathit{max}(\cclrd{s}, \cclrd{s} + \cclrd{r})$ & $\cclrd{s} + \cclrd{r}$ \\ \hline
 \end{tabular}
@@
 \vspace{-0.5em}
-\figcaption{Simplified annotation for let binding in sample flat calculi instances}
+\figcaption{Simplified coeffect annotation for let binding in three flat calculi instances}
 \label{tab:flat-simplelet}
@@
 
-The simplified annotations directly follow from the definitions of particular flat coeffect
-algebras. It is perhaps somewhat unexpected that the annotation can be simplified in different
-ways for different examples.
-
-To see that the simplified annotations are more precise, assume that we used arbitrary
-splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$ rather than idempotence. The
-``Definition'' column would use $\cclrd{s_1}$ and $\cclrd{s_2}$ for the first and second
-$\cclrd{s}$, respectively. The corresponding simplified annotation would have
-$\cclrd{s_1}\,\czip\,\cclrd{s_2}$ instead of $\cclrd{s}$. For all our systems, the
-simplified annotation (on the right) is more precise than the original (on the left):
+It is perhaps somewhat unexpected that the annotation can be simplified in different
+ways for different examples. However, for all our systems, the simplfied annotation
+(on the right) is more precise than the original (on the left). Recall that
+$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following three inequalities hold:
 %
@@
 %
-This property cannot be proved from other properties of the flat coeffect algebra. To make
+This property does not follow from the axioms of the flat coeffect algebra. To make
 the flat coeffect system as general as possible, we do not \emph{in general} require it as
-an additional axiom, although the above examples provide reasonable basis for requiring
-that the specialized annotation for let binding is the least possible annotation for the
-expression $(\lambda x.e_2)~e_1$.
+an additional axiom, although the above examples provide a reasonable basis for using
+the specialized (\emph{let}) rule in the flat coeffect system.
 
@@
 \label{thm:flat-alt-abs}
-For a flat coeffect calculus with a strictly oriented algebra, equipped with sub-coeffecting and
+For a flat coeffect calculus with a strictly oriented algebra, equipped with subcoeffecting and
 subtyping, the standard (abs) rule can be derived from the (idabs) rule.
@@
 \begin{proof}
-The following derives the conclusion of (\emph{abs}) using (\emph{idabs}), sub-coeffecting,
+The following derives the conclusion of (\emph{abs}) using (\emph{idabs}), subcoeffecting,
 sub-typing and the fact that the algebra is \emph{strictly oriented}:

### 2016-04-25 text/Chapter04.tex
Finish edits (Chapter 4)

@@
 in the dataflow system all three are distinct. Similarly, the two special elements
-coincide in some, but not all systems. The required laws are motivated by the aim to capture
+coincide in some, but not all systems. The required axioms are motivated by the aim to capture
 common properties of the three examples, without unnecessarily restricting the system:
@@
 \item The monoid $(\C, \cseq, \cunit)$ represents \emph{sequential} composition of (semantic)
-functions. The laws of a monoid are required in order to form a category structure in the
+functions. The monoid axioms are required in order to form a category structure in the
 semantics (Section~\ref{sec:flat-semantics}).
@@
 \ie~the case when the same context is passed to multiple (independent) computations. The monoid
-laws guarantee that usual syntactic transformations on tuples and the unit value
+axioms guarantee that usual syntactic transformations on tuples and the unit value
 (Section~\ref{sec:flat-exts}) preserve the coeffect. Idempotence holds for all our examples
@@
 constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
-definition satisfies the flat coeffect algebra laws because $(S, \cup, \emptyset)$ is an idempotent,
+definition satisfies the flat coeffect algebra axioms because $(S, \cup, \emptyset)$ is an idempotent,
 commutative monoid. The language has additional syntax for defining an implicit parameter and for
@@
 
-As required by the laws, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
+As required by the axioms, $(\mathbb{N}, +, 0)$ and $(\mathbb{N}, \mathit{max}, 0)$ form monoids
 and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where
-$\cseq$ is not idempotent. The distributivity laws require the following to be the case:
+$\cseq$ is not idempotent. The distributivity axioms require the following to be the case:
 $\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see.
@@
 $\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band (the extended
-\emph{min} is still idempotent and associative) and the distributivity law also holds
+\emph{min} is still idempotent and associative) and the distributivity axioms also hold
 for $\mathbb{N}_{\bot}$.
@@
 
-The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of laws.
-The laws are required for three reasons -- to be able to define the categorical structure
+The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of axioms.
+The axioms are required for three reasons -- to be able to define the categorical structure
 in Section~\ref{sec:semantics-theory}, to prove equational properties in Section~\ref{sec:flat-syntax}
@@
 This rule can be derived from the typing derivation for an expression $(\lambda x.e_2)~e_1$ as
-a special case, by using the idempotence of $\czip$ as follows:
+a special case. We use the idempotence of $\czip$ as follows:
 %
@@
 %
-This design decision is motivated by the fact that the typing obtained using the special rule for
-let-binding is more precise for all the examples consider in this chapter. To see this, assume
-an arbitrary splitting $\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Table~\ref{tab:flat-simplelet}
+This is one possible derivation, but other derivations may be valid for concrete coeffect system.
+The design decision of using this particular derivation for the typing of \kvd{let} is motivated by
+the fact that the typing obtained using the special rule is more precise for all the examples
+consider in this chapter. To see this, assume an arbitrary splitting
+$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Table~\ref{tab:flat-simplelet}
 shows the coeffect annotation derived from $(\lambda x.e_2)~e_1$, the coeffect annotation obtained
@@
 
-It is perhaps somewhat unexpected that the annotation can be simplified in different
-ways for different examples. However, for all our systems, the simplfied annotation
-(on the right) is more precise than the original (on the left). Recall that
-$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following three inequalities hold:
+It is perhaps somewhat unexpected that the annotation can be simplified in different ways for
+different examples. However, for all our systems, the simplfied annotation (right column in
+Table~\ref{tab:flat-simplelet}) is more precise than the original (left column). Recall that
+$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following inequalities hold:
 %
@@
 In other words, the inequality states that using idempotence, we get a more precise typing.
-Using the $\cgeq$ operator this property can be expressed using the abstract operators of the flat coeffect algebra as:
+Using the $\cgeq$ operator of flat coeffect algebra, this property can be expressed in general as:
 %
@@
 demands using $\cclrd{r} \cup \cclrd{s}$, but for dataflow and liveness it suffices to
-duplicate the demand $\cclrd{r}$ of the body. We introduced the $\czip$ operation as
-a way of providing the additional abstraction.
-
-In this section, we first identify coeffect calculi for which the simpler (\emph{idabs}) rule
-introduced in Section~\ref{ec:flat-unique-ldf} is sufficient. Then we look at syntactic
-transformations corresponding to other common properties of $\czip$.
+duplicate the demand $\cclrd{r}$ of the body. We consider coeffect calculi for which the
+simpler duplication of coeffects is sufficient.
 
@@
 idempotence means that the context demands of the body can be required from both the
-declaration site and the call site. Thus, the (\emph{idabs}) typing is valid. For reference,
-we repeat the earlier definitions here:
+declaration site and the call site. In Section~\ref{sec:flat-unique-ldf}, we introduced the
+(\emph{idabs}) rule (repeated below for reference), which uses the idempotence and duplicates
+coeffect annotations:
 
@@
 The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
-oriented coeffect calculi (such as our liveness and dataflow computations), we can use
-the (\emph{idabs}) rule and get an equivalent type system. This alternative formulation
-removes the non-determinism of type checking that arises from the splitting of context
-demands in the original (\emph{abs}) rule.
+oriented coeffect calculi (such as our liveness and dataflow computations), the (\emph{idabs})
+rule not only determines a unique typing derivation (as discussed in Section~\ref{sec:flat-unique-ldf}),
+but it gives (together with subtyping and subcoeffecting) an equivalent type system.
 
@@
 Intuitively, this means that the context demands of a function with multiple arguments can be
-split arbitrarily between the declaration site and (multiple) call sites. In other words, it does
-not matter how the context demands are satisfied.
+split arbitrarily between the declaration site and (multiple) call sites.
 
@@
 
-To show the key aspects of flat coeffect systems, the calculus introduced in Section~\ref{sec:flat-calculus}
-consists only of variables, abstraction, application and let binding. Here, we extend it
-with pairs and the unit value to sketch how it can be turned into a more complete programming
-language and to motivate the laws required about $\cpar$. The syntax of the language is extended as follows:
+To focus on the key aspects of flat coeffect systems, the calculus introduced in Section~\ref{sec:flat-calculus}
+consists only of variables, abstraction, application and let binding. Here, we extend it with pairs
+and the unit value to sketch how it can be turned into a more complete programming language and to
+further motivate the axioms for $\cpar$. The syntax of the language is extended as follows:
 
@@
 sub-expressions. Finally, the (\emph{proj}) rule is uninteresting, because $\pi_i$ can be
-viewed as a pure function.
+viewed as a pure unary function.
 
@@
 isomorphism that performs appropriate transformation on values, without affecting other
-properties (here, coeffects) of the expressions. The monoid laws then correspond to
-the demand that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
+properties (here, coeffects) of the expressions. The monoid axioms then correspond to
+the requirement that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
 demand that $((), e) \simeq e \simeq (e, ())$ (unit).
 
-Thanks to the properties of $\cpar$, the flat coeffect calculus obeys the monoid laws for pairs.
+Thanks to the properties of $\cpar$, the flat coeffect calculus obeys the monoid axioms for pairs.
 In the following, we assume that \ident{assoc} is a pure function transforming a pair $(x_1, (x_2, x_3))$
@@
 $(\C, \cpar, \czero)$ of the flat coeffect algebra. We require only unit and associativity
-laws. In our three examples, the $\cpar$ operator is also symmetric, which additionally
-gives us the property that $(e_1, e_2) \simeq (e_2, e_1)$.
+axioms. In our three examples, the $\cpar$ operator is also symmetric, which additionally
+guarantees that $(e_1, e_2) \simeq (e_2, e_1)$, which is a property that is expected to hold
+for $\lambda$-calculus.
 
@@
 capture three specific systems, namely liveness, dataflow and implicit parameters. However,
-the system is more general and an capture numerous other applications outlined in
+the system is more general and an capture various other applications outlined in
 Section~\ref{sec:applications-flat}.
@@
 done differently for each concrete coeffect system, depending on its specific notion of context.
-We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec}.
+We discussed this for implicit parameters, dataflow and liveness in Section~\ref{sec:flat-unique}
+and noted that the result for dataflow and liveness generalizes for any coeffect calculus
+with strictly oriented coeffect algebra (Remark~\ref{thm:flat-alt-abs}).
 
@@
 syntactic properties that hold for all flat coeffect systems satisfying certain additional
-conditions. In particular, two \emph{typing preservation} theorems prove that the operational
+conditions. In particular, two \emph{type preservation} theorems prove that the operational
 semantics for two classes of flat coeffect calculi (including liveness and implicit parameters)
-can be based on the standard call-by-name reduction.
+can be based on standard call-by-name reduction.
 
@@
 concrete discussion. We explain its category-theoretical motivation, we use it to define
-translational semantics (akin to Haskell's do notation) and we prove that well-typed programs
-in flat coeffect calcului for implicit parameters and dataflow do not get stuck.
+translational semantics (akin to Haskell's do notation) and we prove a soundness result that
+well-typed programs in flat coeffect calcului for implicit parameters and dataflow do not get
+stuck in the translated version.

### 2016-04-26 text/Chapter05.tex
Some more edits (Chapter 5)

@@
 The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
-of context-aware systems discussed in Chapter~\ref{ch:applications}. The coeffect calculus 
+of context-aware systems outlined in Chapter~\ref{ch:applications}. The coeffect calculus
 can be seen as a \emph{language framework} that simplifies the construction of concrete
@@
 
-We use comonads in a syntactic way, following the example of Haskell's use of monads. The 
-translation is the same for all coeffect languages, but the safety depends on the concrete
-coeffect-specific comonadically-inspired primitives. We prove the soundness of two concrete
-coeffect calculi (dataflow and implicit parameters). We note that the proof crucially relies on a 
-relationship between coeffect annotations (provided by the type system) and the comonadically-inspired
-primitives (defining the semantics), which makes it easy to extend it to other concrete 
-context-aware langauges.
+We use comonads in a syntactic way, following the example of Wadler and Thiemann \cite{monads-effects-marriage}
+and Haskell's use of monads. The translation is the same for all coeffect languages, but the safety
+depends on the concrete coeffect-specific comonadically-inspired primitives. We prove the soundness
+of two concrete coeffect calculi (dataflow and implicit parameters). We note that the proof
+crucially relies on a relationship between coeffect annotations (provided by the type system) and
+the comonadically-inspired primitives (defining the semantics), which makes it easy to extend it to
+other concrete context-aware langauges.
 
@@
   dual of monads (Section~\ref{sec:semantics-theory}) and we discuss how they provide semantics
-  for coeffect calculus. This provides an insight into how (and why) the calculus works and shows 
-  an intriguing link with effects and monads.
+  for coeffect calculus. This provides an insight into how (and why) the coeffect calculus works and
+  shows an intriguing link with effects and monads.
 
 \item We use indexed comonads to guide our \emph{translational semantics} of coeffect calculus
-  (Section~\ref{sec:semantics-translation}). We define a simple functional programming language
-  (with type system and operational semantics) and prove that well-typed programs in the language
-  do not get stuck. We extend it with uninterpreted comonadically-inspired primitives and
-  define a translation that turns well-typed context-aware coeffect programs into programs of 
-  our functional language.
-
-\item For two of the coeffect calculi discussed earlier (dataflow and implicit parameters),
+  (Section~\ref{sec:semantics-translation}). We define a simple sound functional programming
+  language (with type system and operational semantics). We extend it with uninterpreted
+  comonadically-inspired primitives and define a translation that turns well-typed context-aware
+  coeffect programs into programs of our functional language.
+
+\item For two sample coeffect calculi discussed earlier (dataflow and implicit parameters),
   we give reduction rules for the comonadically-inspired primitives and we extend the progress
@@
 
-\item We leave the details to future work, but we note that the proof for concrete coeffect language 
-  (dataflow and implicit parameters) can be generalized -- rather than reconsidering progress and 
-  preservation of the whole target language, we rely just on the correctness of the coeffect-specific 
-  comonadically-inspired primitives and abstraction mechanism provided by languages 
-  such as ML and Haskell (Section~\ref{sec:semantics-related}).
+\item We note that the proof for concrete coeffect language (dataflow and implicit parameters) can
+  be generalized -- rather than reconsidering progress and preservation of the whole target
+  language, we rely just on the correctness of the coeffect-specific comonadically-inspired
+  primitives and abstraction mechanism provided by languages such as ML and Haskell
+  (Section~\ref{sec:semantics-related}).
 \end{itemize}
@@
 
-This chapter links together all the different technical developments presented in this thesis.
+This chapter links together a number of different technical developments presented in this thesis.
 We take the flat coeffect calculus introduced in Chatper~\ref{ch:flat}, define its \emph{abstract
@@
 The key claim of this thesis is that writing context-aware programs using coeffects is easier and
-safer. In this chapter, we substantiate the claim by showing that programs written in the coeffect
-calculus and evaluated using the translation provided here do not ``go wrong''.
+less error-prone. In this chapter, we substantiate the claim by showing that programs written in the
+coeffect calculus and evaluated using the translation provided here do not ``go wrong''.
 
@@
  \quad\quad \ident{lookup}~\str{two}~\ident{params}~\kvd{in}\\[-0.2em]
-\ident{add}~0~(\ident{add}~\str{one}~10~\ident{params})
+\ident{add}~0~(\ident{cons}~\str{one}~10~\ident{params})
 \end{array}
@@
 provides a value for the implicit parameter $\cclrd{\ident{?one}}$, the program reduces to
-a value of an correct type (or never terminates).
+a value of the correct type (or never terminates).
 
@@
 a function of type $\ident{num}~\ident{list}\rightarrow\ident{num}$. The function fails for input
-lists containing only zero or one elements and this is not reflected in the type and it is not 
+lists containing only zero or one elements and this is not reflected in the type and is not
 enforced by the type checker.
@@
 
-In summary, coeffect type system, captures certain runtime 
+In summary, a coeffect type system, captures certain runtime
 demands of context-aware programs and (as we show in this chapter), eliminates common errors
@@
 
-The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
+The type system of the flat coeffect calculus arises syntactically, as a generalization of the examples
 discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical
@@
 semantics of contextual computations (mainly for dataflow) in terms of comonadic functions
-$\ctyp{}\tau_1 \rightarrow \tau_2$. Our \emph{indexed comonads} annotate the structure with 
-information about the required context, \ie~$\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
+$\ctyp{}\tau_1 \rightarrow \tau_2$. We introduce \emph{indexed comonads} that annotate the structure
+with information about the required context, \ie~$\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
 This is similar to the recent development on monads and effects by Katsumata \cite{monads-parametric}
-who parameterizes monads in a similar way to our parameterization of comonads.
+who parameterizes monads in a similar way to our indexed comonads.
 
@@
 
-As discussed in Section~\ref{sec:path-sem}, categorical semantics interprets terms as morphisms
-in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets typing 
-judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
-$\sem{\tau_1 \times \ldots \times \tau_n} \rightarrow \sem{\tau}$.
+As discussed in Section~\ref{sec:path-sem}, a categorical semantics interprets terms as morphisms
+in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets a term
+with a typing derivation leading to a judgement $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$
+as a morphism $\sem{\tau_1 \times \ldots \times \tau_n} \rightarrow \sem{\tau}$.
+
+For a well-defined semantics, we need to ensure that a well-typed term is assigned exactly one
+meaning. This can be achieved in a number of ways. First, we can prove the \emph{coherence}
+\cite{semantics-bltres} and show the morhphisms assigned to multiple typing derivations are equivalent.
+Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
+using the uniqye typing derivation specified in Section~\ref{sec:flat-unique}.
 
@@
 computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that
-approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
+approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$,
 for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models
-partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
+failures (the Maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list
 monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$
@@
 
-Following similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that 
+Following a similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that
 (\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent
@@
 
-The monadic and comonadic model outlined here represents at most a binary analysis of effects or 
+The monadic and comonadic model outlined aboe represents at most a binary analysis of effects or
 context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no
@@
 two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers
-require defining complex system of lifting to be composable. Consequently, they are usually used
+require the user to define complex systems of lifting to be composable. Consequently, they are usually used
 for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
@@
 The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
-sequential composition, where the annotation $\cunit$ represents a variable access. An indexed 
-comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the annotation 
-$\cclrd{r}$ determines what context is required.
+sequential composition, where the element $\cunit$ represents a variable access. An indexed
+comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the $\cclrd{r}$
+(also called \emph{annotation}) is a member of the set $\C$ and determines what context is required.
 
@@
 Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings
-$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
+$\ctyp{\cclrd{r}}{}$ indexed by elements of the monoid structure $\C$. Similarly, the $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$
 operation is now formed by a \emph{family} of mappings for different pairs of indices
 $\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations
-and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
+and we should include objects $\alpha, \beta$ (modeling types) as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
 For the purpose of this thesis, it is sufficient to omit the superscripts and treat $\ident{cobind}$ just
@@
 The $\ident{counit}$ operation is not defined for all $\cclrd{r} \in \C$, but only for
-the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
+the unit $\cunit$. Nevertheless we continue to write $\ident{counit}_{\cunit}$,
 but this is merely for symmetry and as a useful reminder to the reader. Crucially, this means that
@@
 
-If we look at the indices in the laws, we can see that the left and right identity 
+If we look at the indices in the comonad axioms, we can see that the left and right identity
 require $\cunit$ to be the unit of $\cseq$. Similarly, the associativity law implies the
@@
 
-\begin{example}[Indexed option]
-The \emph{indexed option comonad} is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
+\begin{example}[Indexed Maybe]
+The \emph{indexed Maybe comonad} is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$
 where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
@@
 \noindent
-The \emph{indexed option comonad} models the semantics of the liveness coeffect system discussed in 
+The \emph{indexed Maybe comonad} models the semantics of the liveness coeffect system discussed in
 Section~\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context
@@
 $(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
-As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
-= \alpha \times (\cclrd{r} \rightarrow \rho)$ represents a value $\alpha$ together with a function that
-associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
+We assume that, all implicit parameters have the type $\ident{num}$. The data type $\ctyp{\cclrd{r}}{\alpha}
+= \alpha \times (\cclrd{r} \rightarrow \ident{num})$ represents a value $\alpha$ together with a function that
+associates a parameter value $\ident{num}$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
 The cobind and counit operations are defined as:
@@
 \noindent
-In the definition, we use the notation $(a, g)$ for a pair containing the value of type $\alpha$
-together with $g$, which is a function $\cclrd{r} \rightarrow \rho$. The \ident{counit} operation
+In the definition, we use the notation $(a, g)$ for a pair containing a value of type $\alpha$
+together with $g$, which is a function of type $\cclrd{r} \rightarrow \ident{num}$. The \ident{counit} operation
 takes a value and a function (with empty set as a domain), ignores the function and extracts the
@@
 (\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed
-to \ident{cobind} is initially defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is 
+to \ident{cobind} is defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is
 valid in both cases.
@@
 Because of the asymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
-the duality between monads and comonads can no longer help us with defining the additional structure
+the duality between monads and comonads does not lead us towards the additional structure
 required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
@@
 This allows lifting of free variables into an effectful computation. In Haskell, strength can be expressed
-in the host language and so it is implicit.
+in the host language and so is implicit.
 
@@
 %
-The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
-contexts and is used to merge the context of the declaration site (containing free variables)
+The \ident{m} operation is needed in the semantics of lambda abstraction. Semantically, it represents
+merging of contextual capabilities attached to the variable contexts of the declaration site (containing free variables)
 and the call site (containing bound variable). For example, for implicit parameters, this combines
@@
 
-The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
+The semantics of flat coeffect calculus requires not only operations for \emph{merging}, but also for
 \emph{splitting} of contexts.
@@
 bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
-as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
-the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.
+as an informal analogy.}. We look at examples of this operation in the next section.
 
@@
 
-\begin{example}[Indexed option]
+\begin{example}[Indexed Maybe]
 \label{thm:semantics-indexed-opt}
@@
 \noindent
-Without the indexing, the \ident{merge} operations implements \emph{zip} on option values,
-returning an option only when both values are present. The behaviour of the \ident{split} 
+Without the indexing, the \ident{merge} operations implements \emph{zip} on Maybe values,
+returning a value only when both values are present. The behaviour of the \ident{split}
 operation is partly determined by the indices. When the input is \emph{dead}, both values have
@@
 $\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
-is $\alpha \times (\cclrd{r} \rightarrow \rho)$ where $\rho$ is the type of implicit parameter values.
+is $\alpha \times (\cclrd{r} \rightarrow \ident{num})$ where $\ident{num}$ is the type of implicit parameter values.
 The additional operations are defined as:
@@
 A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data
-types needs to form a comonad separately. For example, our indexed option does not form a family of 
+types needs to form a comonad separately. For example, our indexed Maybe does not form a family of
 comonads (again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given

### 2016-04-26 main.tex
Move conclusions
Implementation chapter
Edits (Chapter 5)

@@
 \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
+\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\ttfamily #1}}}
 \newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
@@
 % For definining semantics of things over a typing derivation
+\newcommand{\semdef}[2]{
+\begin{array}{rcl}
+  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
+  \sem{#1} & = & #2 \\
+\end{array}
+}
 \newcommand{\semdeff}[4]{
 \begin{array}{rcl}
-  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
+  \sem{#1} & = & #3 \\ \cline{1-1}\cline{3-3}
+  \sem{#2} & = & #4 \\
+\end{array}
+}
+\newcommand{\emdeff}[4]{
+\begin{array}{rcl}
+  \sem{#1} & = & #3 \\ \cline{1-1}\cline{3-3}
   #2 & = & #4 \\
@@
 \begin{array}{rcl}
-  #1 & = & #4 \\
-  #2 & = & #5 \\ \cline{1-1}\cline{3-3}
-  #3 & = & #6 \\
+  \sem{#1} & = & #4 \\
+  \sem{#2} & = & #5 \\ \cline{1-1}\cline{3-3}
+  \sem{#3} & = & #6 \\
 \end{array}
 }
-\newcommand{\semdef}[2]{
+\newcommand{\emdefff}[6]{
 \begin{array}{rcl}
-  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
-  #1 & = & #2 \\
+  \sem{#1} & = & #4 \\
+  \sem{#2} & = & #5 \\ \cline{1-1}\cline{3-3}
+  #3 & = & #6 \\
 \end{array}
@@
 \part{Coeffect calculi}
+\label{part:coeffect-calculi}
 
@@
 \textsc{Syntax}
-  & \hspace{-0.5em}\begin{tabular}{l} Coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:flat-calculus-syntax}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:flat-calculus}) \end{tabular}
   & \hspace{-0.5em}\begin{tabular}{l} Extensions such as \\[-0.3em] \ident{?param} and \kvd{prev} \\[-0.3em] (Section~\ref{sec:flat-calculus-examples})  \end{tabular} \\ \hline
@@
 %\include{text/Chapter07}
-%\include{text/Chapter08}
+\include{text/Chapter08}
+%\include{text/Chapter09}
 

### 2016-04-26 text/Chapter05.tex
Edits (Chapter 5)

@@
 \emph{domain-specific} coeffect lnguages. In the previous chapter, we discussed how it provides a
-type system that tracks the required context. In this chapter, we show that the langauge framework
+type system that tracks the required context. In this chapter, we show that the language framework
 also provides a way for defining the semantics of concrete domain-specific coeffect languages,
@@
 the comonadically-inspired primitives (defining the semantics), which makes it easy to extend it to
-other concrete context-aware langauges.
+other concrete context-aware languages.
 
@@
 \begin{array}{l}
-\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
+\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}`\rightarrow`\\[-0.2em]
  \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
@@
 flat coeffect calculus for liveness, it appears natural to model computations as functions
-$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalise comonads
+$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalize comonads
 to \emph{indexed comonads}.
@@
 single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
-\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
-bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
-as an informal analogy.}. We look at examples of this operation in the next section.
+\emph{greatest lower bound}. We look at examples of this operation in the next section.
 
@@
 \emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often
-the \emph{least upper-bound}\footnotemark[1].
+the \emph{least upper-bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
+bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
+as an informal analogy.}.
 
@@
 %
-Although we do not demand this as a general law, in all our systems, it is the case that
-$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$.
-This special case allows a simpler definition of \emph{indexed flat comonad} by expressing the
-\ident{split} operation in terms of lifting (sub-coeffecting) as follows:
+The axioms of flat coeffect algebra do not, in general, require that
+$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
+but the property holds for the three sample coeffect systems we consier. For systems with
+the above property, the \ident{split} operation can be expressed in terms of lifting
+(sub-coeffecting) as follows:
 %
@@
 \ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
-  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),
+  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),~
     \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
@@
 
-This alternative is valid for our examples, but we do not use it for three reasons. First, it
-requires making sub-coeffecting a part of the core definition. Second, this would be the only
-place where our semantics uses a variable \emph{twice} (in this case $c$). Thus using an explicit
-\ident{split} means that the structure required by our semantics does not need to provide variable
-duplication and our model could be embedded in linear or affine category. Finally, explicit
-\ident{split} is similar to the definition that is needed for structural coeffects in
-Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.
+This alternative definition is valid for our examples, but we do not use it for three reasons. First,
+it requires making sub-coeffecting a part of the core definition. Second, this would be the only
+place where our semantics uses a variable \emph{twice} (in this case $c$). Note therefore that
+our use of an explicit \ident{split} means that the structure required by our semantics does not
+need to provide variable duplication and our model could be embedded in linear or affine category.
+Finally, explicit \ident{split} is similar to the definition that is needed for structural coeffects
+in Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.
 
@@
 All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:semantics-flat-idx} can
-be extended into \emph{flat indexed comonads}. Note that this is not true \emph{in general},
+be extended into \emph{flat indexed comonads}. Note however that this cannot be done mechanically,
 because each example requires us to define additional operations, specific for the example.
@@
 \begin{example}[Monoidal comonads]
-Just like indexed comonads generalise co\-monads, the additional structure of
-flat indexed comonads generalises symmetric semimonoidal comonads of Uustalu
+Just like indexed comonads generalize co\-monads, the additional structure of
+flat indexed comonads generalizes the symmetric semimonoidal comonads of Uustalu
 and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
@@
 
-\begin{example}[Indexed Maybe]
+\begin{example}[Indexed Maybe comonad]
 \label{thm:semantics-indexed-opt}
 
-Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively
+The flat coeffect algebra for liveness defines $\cpar$ and $\czip$, respectively as $\sqcup$ and $\sqcap$
 and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined
@@
 For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
-$\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
+$\cleq$ is formed by the subset relation $\subseteq$. Recall that the comonadic data type $\ctyp{\cclrd{r}}{\alpha}$
 is $\alpha \times (\cclrd{r} \rightarrow \ident{num})$ where $\ident{num}$ is the type of implicit parameter values.
@@
 The \ident{split} operation splits the tuple and restricts the function (representing available
-implicit parameters) to the required sub-sets. The \ident{merge} operation is more
+implicit parameters) to the required subsets. The \ident{merge} operation is more
 interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
@@
 argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths.
-Finally, the lifting operation just drops some number of elements from a list.
 
@@
 \\[2em]
-\semdefff
+\emdefff
   {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
   {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\begin{array}{l}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
   {f}
@@
 comonad} structure, we can now define a single uniform semantics that is capable of capturing all
-our examples, as well as other computations that can be modelled by the structure.
+our examples, as well as various other computations.
 
@@
 may have different meaning (\eg~when working with implicit parameters) and so the semantics is
-defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a
-semantics to a term, we choose its unique coeffect-specific typing derivation as defined in
-Section~\ref{sec:flat-unique}.
+defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a semantics to
+a term, we need to choose a particular typing derivation. The algorithm for choosing a unique
+typing derivation for our three systems has been defined in Section~\ref{sec:flat-unique}.
 
 \paragraph{Contexts and types.}
-The modelling of contexts and functions generalizes the earlier concrete examples. We use the
-family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed) data-type that wraps the product of
-free variables of the context and the arguments of functions:
+The modelling of contexts and functions generalizes the concrete examples discussed in
+Chapter~\ref{ch:applications}. We use the family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed)
+data-type that wraps the product of free variables of the context and the arguments of functions:
 %
@@
 Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses
-$\ident{counit}_{\cunit}$ to extract a product of free variables and projection
+$\ident{counit}_{\cunit}$ to extract a product of free variables, followed by projection
 $\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) is interpreted as a curried
@@
 \paragraph{Properties.}
-We the categorical semantics  in Section~\ref{sec:semantics-translation} to define a translation
-that embeds context-dependent computations in a functional programming langauge, similarly to how
+The categorical semantics  in Section~\ref{sec:semantics-translation} defines a translation
+that embeds context-dependent computations in a functional programming language, similarly to how
 monads and the do notation provide a way of embedding effectful computations in Haskell.
@@
 is well-typed. We revisit the property in Lemma~\ref{thm:semantics-welltyped} once we define
-the target langauge and its typing.
+the target language and its typing.
 
@@
 \begin{proof}
-By analysis of the semantic rules in Figure~\ref{fig:semantics-flat}.
+By analysis of the semantic rules in Figure~\ref{fig:semantics-flat}. We need to check that the
+domains and codomains of the morphisms in the semantics (right-hand side) match.
 \end{proof}
@@
   \tau &  = & \ident{num} \sep \tau_1 \times \ldots \times \tau_n \sep \tau_1 \rightarrow \tau_2 \\
-  C  & = & (v_1, \ldots, v_{i-1}, \_, e_{i+1}, \ldots e_n) \sep v~\_ \sep \_~e \sep \pi_i~\_
+  K  & = & (v_1, \ldots, v_{i-1}, \_, e_{i+1}, \ldots e_n) \sep v~\_ \sep \_~e \sep \pi_i~\_
 \end{array}
@@
 \begin{array}{rll}
-  \footnotesize{(\emph{fn})}  & (\lambda x.e)~v \rightarrow e[x\leftarrow v] \\[0.25em]
-  \footnotesize{(\emph{prj})} & \pi_i (v_1, \ldots, v_n) \rightarrow v_i\\[0.25em]
-  \footnotesize{(\emph{ctx})}  & C[e] \rightarrow C[e']  & (\textnormal{when}~e\rightarrow e')\\[0.25em]
+  \footnotesize{(\emph{fn})}  & (\lambda x.e)~v \rightsquigarrow e[x\leftarrow v] \\[0.25em]
+  \footnotesize{(\emph{prj})} & \pi_i (v_1, \ldots, v_n) \rightsquigarrow v_i\\[0.25em]
+  \footnotesize{(\emph{ctx})}  & K[e] \rightsquigarrow K[e']  & (\textnormal{when}~e\rightsquigarrow e')\\[0.25em]
 \end{array}
@@
 
-\caption{Common syntax and reduction rules of the target langauge}
+\caption{Common syntax and reduction rules of the target language}
 \label{fig:semantics-target}
@@
 
-\section{Embedding coeffect languages}
+\section{Translational semantics}
 \label{sec:semantics-translation}
@@
 domain-specific language extensions (the \kvd{prev} keyword, or the \ident{?impl} syntax)
-is translated to a \emph{target} langauge that is not context-aware. The target language is
+is translated to a \emph{target} language that is not context-aware. The target language is
 a small functional language consisting of:
@@
 %
-The syntx, typing and reduction rules of the first part (simple functional language) are used by
-all concrete coeffect domain-specific langauges. The syntax and typing rules of the second part
+The syntax, typing and reduction rules of the first part (simple functional language) are common to
+all concrete coeffect domain-specific languages. The syntax and typing rules of the second part
 (comonadically-inspired) primitives are also shared by all coeffect DSLs, however the \emph{reduction
-rules} for the comonadically-inspired primitives differ -- they capture the concrete notion of
-context. Finally, the third part (domain-specific primitives) will differ for each coeffect DSL.
+rules} for the comonadically-inspired primitives differ -- they capture the concrete notions of
+context. Finally, the third part (domain-specific primitives) will differ for each coeffect
+domain-specific language.
 
@@
 
-\subsection{Functional target langauge}
+\subsection{Functional target language}
 \label{sec:semantics-translation-target}
 
-The target langauge for the translation is a simply typed lambda calculus with integers and tuples.
+The target language for the translation is a simply typed lambda calculus with integers and tuples.
 We include integers as an example of a concrete type. Tuples are needed by the translation, which
@@
 
-The syntax of the target programming langauge is shown in Figure~\ref{fig:semantics-target}.
+The syntax of the target programming language is shown in Figure~\ref{fig:semantics-target}.
 The values include numbers $n$, tuples and function values. The expressions include variables $x$,
@@
 %
-Finally, $C[e]$ defines the context in which sub-expressions are evaluated. Together with the
-evaluation rules shown in  Figure~\ref{fig:semantics-target}, this captures the standard
-call-by-name semantics of the common parts of the target language. The (standard) typing rules
-for the common expressions of the target langauge are also shown in Figure~\ref{fig:semantics-target}.
+Finally, $K[e]$ defines the syntactic evaluation context in which sub-expressions are evaluated.
+Together with the evaluation rules shown in  Figure~\ref{fig:semantics-target}, this captures the
+standard call-by-name semantics of the common parts of the target language. The (standard) typing rules
+for the common expressions of the target language are also shown in Figure~\ref{fig:semantics-target}.
 
@@
 
-The functional subset of the langauge described so far models a simple ML-like langauge. We
+The functional subset of the language described so far models a simple ML-like language. We
 choose call-by-value over call-by-name for no particular reason and Haskell-like language would
@@
 \label{thm:semantics-fp-pres-subst}
-For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
+For all $\Gamma, e, e', \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
 then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
@@
 \begin{proof}
-  Rule induction over $\rightarrow$.
+  Rule induction over $\rightsquigarrow$.
 
@@
 \label{thm:semantics-fp-prog}
-  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
 \end{theorem}
@@
 
-\begin{theorem}[Safety]
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow^* e'$ then either $e'$ is a value or there
-  exists $e''$ such that $e' \rightarrow e''$
+\begin{theorem}[Safety of functional target language]
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^* e'$ then either $e'$ is a value of type $\tau$ or
+  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
 \end{theorem}
 \begin{proof}
-  Rule induction over $\rightarrow^*$ using Theorem~\ref{thm:semantics-fp-pres} and Theorem~\ref{thm:semantics-fp-prog}.
+  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:semantics-fp-pres} and Theorem~\ref{thm:semantics-fp-prog}.
 \end{proof}
@@
 \paragraph{Language syntax}
-Given $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$, extend the syntax:
+Given $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$, extend the programming language
+syntax with the following constructs:
 
@@
   \tau &  = &\ldots \sep \ctyp{\cclrd{r}}{\tau} \\
-  C  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~\_~e \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
+  K  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~\_~e \sep \ident{cobind}_{\cclrd{s},\cclrd{r}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
     && \hspace{1.4em} \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~\_
@@
 
-\caption{Comonadically-inspired extensions for the target langauge}
+\caption{Comonadically-inspired extensions for the target language}
 \label{fig:semantics-ext}
@@
 comonadically-inspired operations. This is done in the same way for all concrete coeffect
-domain-specific languages and so we give the additional syntax, evaluation context and
-typing rules in Figure~\ref{fig:semantics-ext} for all languages we consider later in
+domain-specific languages and so we give the common additional syntax, evaluation context and
+typing rules once in Figure~\ref{fig:semantics-ext}. We consider examples later in
 Section~\ref{sec:semantics-proofs}.
@@
 correspond to the operations of indexed comonads. Note that the we embed the coeffect annotations
-into the target language -- these are known when translationg a typed term from a source language
-and they will be useful when proving that sufficient context (as specified by the coeffect
-annotations) is available.
-
-The figure defines the syntax and the typing rules, but it does not define the reduction rules.
-Those -- together with the values for a concrete notion of context -- will be defined separately
-for each individual coeffect language.
+into the target language -- these are known when translating a term with a chosen typing derivation
+from a source language and they will be useful when proving that sufficient context (as specified
+by the coeffect annotations) is available.
+
+Figure~\ref{fig:semantics-ext} defines the syntax and the typing rules, but it does not define the
+reduction rules. These -- together with the values for a concrete notion of context -- will be
+defined separately for each individual coeffect language.
 
@@
 \\[1.5em]
-\hspace{1.5em}\semdeff
+\hspace{1.5em}\emdeff
   {\coctx{\Gamma, x_i:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
   {\hspace{-1em}\begin{array}{l}
-    \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x_i.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2
+    \sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x_i.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
     \\[-0.3em]~\\[-0.3em]~ \\[-0.3em]~
@@
     \lambda \ctx.\lambda v.\\[-0.3em]
-    \quad \kvd{let}~\ident{assoc} = \lambda x.\\[-0.3em]
+    \quad \kvd{let}~\ident{reassoc} = \lambda x.\\[-0.3em]
     \quad\quad (\pi_1(\pi_1~x), \ldots, \pi_{i-1}(\pi_1~x), \pi_2~x)\\[-0.3em]
-    \quad f~(\ident{map}_{\cclrd{r}\czip\cclrd{s}}~\ident{assoc}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v)))
+    \quad f~(\ident{map}_{\cclrd{r}\czip\cclrd{s}}~\ident{reassoc}~(\ident{merge}_{\cclrd{r}, \cclrd{s}}~(\ctx, v)))
    \end{array} }
@@
 \\[4em]
-\semdefff
+\emdefff
   {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
   {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\begin{array}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\\[-0.3em]~\\[-0.3em]~\end{array}}
+  {\begin{array}{r}\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\\[-0.3em]~\\[-0.3em]~\end{array}}
   {f}
@@
 language are translated into the target language. The rules are very similar to those shown earlier
-in Figure~\ref{fig:semantics-flat}. The result is now written as source code in the
+in Figure~\ref{fig:semantics-flat}. The consequent is now written as source code in the
 target programming language rather than as composition of morphisms in a category. However,
-thanks to the equivalence between $\lambda$-calculus and category theory, both interpretations
-are equivalent.
+thanks to the relationship between $\lambda$-calculus and cartesian closed categories, both
+interpretations are equivalent.
 
@@
 that contains variable assignments. Previously, we assumed that the tuple is appropriately
-reassociated. Now, we perform the reassociation explicitly. We keep a flat tuple of variables, so
+reassociated. For programming language translation and the implementation (discussed in
+Chapter~\ref{ch:impl}), we perform the reassociation explicitly. We keep a flat tuple of variables, so
 given $\Gamma=x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n$, the tuple has a type $\tau_1\times\ldots\times\tau_n$.
@@
 
-In Section~\ref{sec:semantics-proofs}, we will prove safety properties of well-typed programs in the
-target language. Thanks to the fact that the translation produces a well-typed program (as complex
-as it might be) means that we are also proving safety of well-typed programs in the source
-context aware languages.
+In Section~\ref{sec:semantics-proofs}, we will prove safety properties of well-typed programs in
+the target language. Thanks to the fact that the translation produces a well-typed program
+means that we are also proving safety of well-typed programs in the source context-aware languages.
 
@@
 Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$
-written in a context-aware programming langugae that is translated to the target langauge as:
+written in a context-aware programming langugae that is translated to the target language as
+(we write $\ldots$ for the omitted part of the translation tree):
 %
@@
   \semdeff
-    {(\ldots)\qquad~}
-    {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
+    {~~(\ldots)~~}
+    {\coctx{}{\cclrd{r}} \vdash e : \tau }
     {(\ldots)}
@@
 %
-Then $f$ is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
+Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
 \end{theorem}
@@
 
-The previous two languages provide a general structrue that we now use to prove the safety of
-concrete context-aware programming languages based on the coeffect language framework. We
-consider a langauge for dataflow computations (Section~\ref{sec:semantics-proofs-df}) and
-for implicit parameters (Section~\ref{sec:semantics-proofs-impl}). In both cases, we extend
-the progress and preservation theorems of the functional subset of the target language, but the
-approach can be generalised as discussed in Section~\ref{sec:semantics-generalising}.
+The language defined in Figure~\ref{fig:semantics-target} and Figure~\ref{fig:semantics-ext}
+provide a general structure that we now use to prove the safety of various context-aware
+programming languages based on the coeffect language framework. As examples, we consider a language
+for dataflow computations (Section~\ref{sec:semantics-proofs-df}) and for implicit parameters
+(Section~\ref{sec:semantics-proofs-impl}). In both cases, we extend the progress and preservation
+theorems of the functional subset of the target language, but the approach can be generalized as
+discussed in Section~\ref{sec:semantics-generalising}.
+
+As outlined in the table at the beginning of Part~\ref{part:coeffect-calculi}, we now covered the
+parts of the semantics that are shared by all context-aware languages. This includes the functional
+target language with comonadically-inspired uninterpreted type $\ctyp{\cclrd{r}}{\tau}$ and the
+syntax for comonadically-inspired uninterpreted primitives such as
+$\ident{cobind}_{\cclrd{s},\cclrd{r}}$ and $\ident{counit}_{\cunit}$, together with their typing.
+
+Using dataflow and implicit parameters as two examples, we now add the domain-specific extensions
+needed for a concrete context-aware programming language. This includes syntax for values and
+expressions of the comonad-inspired type $\ctyp{\cclrd{r}}{\tau}$ and reduction rules for the
+comonadically-inspired operations ($\ident{cobind}_{\cclrd{s},\cclrd{r}}$, $\ident{counit}_{\cunit}$, etc.).
 
@@
 
-\subsection{Coeffect langauge for dataflow}
+\subsection{Coeffect language for dataflow}
 \label{sec:semantics-proofs-df}
@@
 that the number of elements in the stream matches the number in the coeffect.
-The additional reduction rules mirror the semantics that we discussed earlier when talking about
-indexed dataflow comonad.
+The additional reduction rules mirror the semantics that we discussed in
+Example~\ref{thm:semantics-indexed-list} when discussing the indexed list comonad.
 
@@
 e &=& \ldots \sep \ident{Df}\langle e_0, \ldots, e_n\rangle \sep \ident{prev}_{\cclrd{n}}~e\\
-C &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_ \sep \ident{Df}\langle v_0, \ldots, v_{i-1}, \_, e_{i+1} \ldots, e_n\rangle
+K &=& \ldots \sep \ident{prev}_{\cclrd{n}}~\_ \sep \ident{Df}\langle v_0, \ldots, v_{i-1}, \_, e_{i+1} \ldots, e_n\rangle
 \end{array}
@@
 {\footnotesize(\emph{counit})} & \begin{array}{l}
-\ident{counit}_{\cclrd{0}}(\ident{Df} \langle v_0 \rangle)   \rightarrow v_0
+\ident{counit}_{\cclrd{0}}(\ident{Df} \langle v_0 \rangle)   \rightsquigarrow v_0
 \end{array}
@@
 {\footnotesize(\emph{cobind})} & \begin{array}{l}
-\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightarrow
+\ident{cobind}_{\cclrd{m}, \cclrd{n}}~f~(\ident{Df}\langle v_0, \ldots v_{m+n} \rangle) \rightsquigarrow
 \\[-0.25em]
@@
 {\footnotesize(\emph{merge})} & \begin{array}{l}
-\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle v'_0, \ldots, v'_n\rangle)) \rightarrow
+\ident{merge}_{\cclrd{m}, \cclrd{n}} ((\ident{Df}\langle v_0, \ldots, v_m \rangle), (\ident{Df}\langle v'_0, \ldots, v'_n\rangle)) \rightsquigarrow
 \\[-0.25em]
@@
 {\footnotesize(\emph{split})} & \begin{array}{l}
-\ident{split}_{\cclrd{m}, \cclrd{n}} (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle)  \rightarrow
+\ident{split}_{\cclrd{m}, \cclrd{n}} (\ident{Df}\langle (v_0, b_0), \ldots, (v_{\mathit{max}(m,n)}, b_{\mathit{max}(m,n)}) \rangle)  \rightsquigarrow
 \\[-0.25em]
@@
 {\footnotesize(\emph{prev})} & \begin{array}{l}
-\ident{prev}_{\cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_n, v_{n+1} \rangle) \rightarrow
+\ident{prev}_{\cclrd{n}} (\ident{Df}\langle v_0, \ldots, v_n, v_{n+1} \rangle) \rightsquigarrow
 \\[-0.25em]
@@
 
-\caption{Additional constructs for modelling dataflow in the target langauge}
+\caption{Additional constructs for modelling dataflow in the target language}
 \label{fig:semantics-ext-df}
@@
 \paragraph{Properties.}
-Now consider a target langauge consisting of the core (ML-subset) defined by the syntax,
+Now consider a target language consisting of the core (ML-subset) defined by the syntax,
 reduction rules and typing rules given in Figure~\ref{fig:semantics-target} and comonadically-inspired
@@
 
-In order to prove type safety, we first extend the canonical forms lemma
-(Lemma~\ref{thm:semantics-fp-canon}) and the preservation under substitution lemma
+In order to prove type safety, we first extend the \emph{canonical forms lemma}
+(Lemma~\ref{thm:semantics-fp-canon}) and the \emph{preservation under substitution lemma}
 (Lemma~\ref{thm:semantics-fp-pres-subst}). Those need to consider the new (\emph{df})
@@
 rule for \kvd{prev} produces well-typed expressions. Finally, we extend the type preservation
-(Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog}).
+(Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog}) theorems.
 
@@
 the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
-and Figure~\ref{fig:semantics-ext-df} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
+and Figure~\ref{fig:semantics-ext-df} is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
 \end{theorem}
@@
 \label{thm:semantics-df-pres-subst}
-For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
+For all $\Gamma, e, e', \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
 then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
@@
 \label{thm:semantics-df-pres}
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow e'$ then $\Gamma \vdash e' : \tau$
 \end{theorem}
 \begin{proof}
-  Rule induction over $\rightarrow$.
+  Rule induction over $\rightsquigarrow$.
 
@@
 \label{thm:semantics-df-prog}
-  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
 \end{theorem}
@@
 
-\begin{theorem}[Safety]
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow^* e'$ then either $e'$ is a value or there
-  exists $e''$ such that $e' \rightarrow e''$
+\begin{theorem}[Safety of context-aware dataflow language]
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^{*} e'$ then either $e'$ is a value of type $\tau$ or
+  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
 \end{theorem}
 \begin{proof}
-  Rule induction over $\rightarrow^*$ using Theorem~\ref{thm:semantics-df-pres} and Theorem~\ref{thm:semantics-df-prog}.
+  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:semantics-df-pres} and Theorem~\ref{thm:semantics-df-prog}.
 \end{proof}
@@
 
-\subsection{Coeffect langauge for implicit parameters}
+\subsection{Coeffect language for implicit parameters}
 \label{sec:semantics-proofs-impl}
 
-As discussed earlier (Example~\ref{thm:semantics-indexed-prod}), implicit parameters can be modelled
-by an indexed product comonad, which annotates a value with additional context -- in our case, a
-mapping from implicit parameter names to their values. In this section, we embed this model into
-the target language.
+We now turn to our second example. As discussed earlier (Example~\ref{thm:semantics-indexed-prod}),
+implicit parameters can be modelled by an indexed product comonad, which annotates a value with
+additional context -- in our case, a mapping from implicit parameter names to their values. In this
+section, we embed this model into the target language.
 
@@
 Again, the $\ctyp{\cclrd{r}}\tau$ values can be seen as an \emph{abstract data type}, which are
-never manipulated directly.
+never manipulated directly, except by the comonadically-inspired operations
+($\ident{cobind}_{\cclrd{s},\cclrd{r}}$, $\ident{counit}_{\cunit}$, etc.).
 
@@
 rule accesses an implicit parameter and (\emph{letimpl}) adds a new binding. The reduction rules
-for comonadically-inspired primitives closely model the product comonad discussed in
-Example~\ref{thm:semantics-indexed-prod}. Reductions for (\emph{cobind}) and (\emph{split})
-restrict the set of available implicit parameters according to the annotations and
-(\emph{merge}) combines them, preferring the values from the call-site.
+closely model the product comonad discussed in Example~\ref{thm:semantics-indexed-prod}. Reductions
+for (\emph{cobind}) and (\emph{split}) restrict the set of available implicit parameters according
+to the annotations and (\emph{merge}) combines them, preferring the values from the call-site.
 
@@
 derivations for implicit parameter coeffects obtained in Section~\ref{sec:flat-unique} always
-split implicit parameters into \emph{disjoint sets}.
+split implicit parameters into \emph{disjoint sets}, so preferences do not come into play.
 
@@
   && \hspace{1.4em} \sep \ident{lookup}_{\cclrd{\ident{?p}}}~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~e_1~e_2\\
-C &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_ \\
+K &=& \ldots \sep \ident{lookup}_{\cclrd{\ident{?p}}}~\_ \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~\_~e \sep \ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v~\_ \\
   && \hspace{1.4em} \sep \ident{Impl}(\_, \{ \ident{?p}_1\mapsto e_1, .., \ident{?p}_n\mapsto e_n\})\\
@@
 {\footnotesize(\emph{counit})} & \begin{array}{l}
-\ident{counit}_{\cclrd{ \emptyset }}~(\ident{Impl} (v, \ldots))   \rightarrow v
+\ident{counit}_{\cclrd{ \emptyset }}~(\ident{Impl} (v, \ldots))   \rightsquigarrow v
 \end{array}
@@
 {\footnotesize(\emph{cobind})} & \begin{array}{l}
-\ident{cobind}_{\cclrd{r}, \cclrd{s}}~f~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})) \rightarrow
+\ident{cobind}_{\cclrd{r}, \cclrd{s}}~f~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})) \rightsquigarrow
 \\[-0.25em]
@@
               (~\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \})  \\[-0.25em]
-\hspace{4.4em}    \ident{Impl}(v', \{ \ident{?p}'_1 \mapsto v'_1, \ldots, \ident{?p}'_{n}\mapsto v'_n \})~) \rightarrow \\[-0.25em]
+\hspace{4.4em}    \ident{Impl}(v', \{ \ident{?p}'_1 \mapsto v'_1, \ldots, \ident{?p}'_{n}\mapsto v'_n \})~) \rightsquigarrow \\[-0.25em]
 \quad \ident{Impl}((v, v'), \{ \ident{?p}_i \mapsto v_i \;|\; \ident{?p}\in\cclrd{r \setminus s}  \} \cup \{ \ident{?p}'_i \mapsto v'_i \;|\; \ident{?p}'\in\cclrd{s}  \})
@@
 {\footnotesize(\emph{split})} & \begin{array}{l}
-\ident{split}_{\cclrd{r}, \cclrd{s}} (\ident{Impl}((v,v'), \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightarrow
+\ident{split}_{\cclrd{r}, \cclrd{s}} (\ident{Impl}((v,v'), \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightsquigarrow
 \\[-0.25em]
@@
 {\footnotesize(\emph{letimpl})} & \begin{array}{l}
-\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v'~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightarrow
+\ident{letimpl}_{\cclrd{\ident{?p}}, \cclrd{r}}~v'~(\ident{Impl}(v, \{ \ident{?p}_1 \mapsto v_1, \ldots, \ident{?p}_{n}\mapsto v_n \}))  \rightsquigarrow
 \\[-0.25em]
@@
 {\footnotesize(\emph{lookup})} & \begin{array}{l}
-\ident{lookup}_{\cclrd{\ident{?p}_i}} (\ident{Impl}(v, \{\ident{?p}_i\mapsto v_i\}) \rightarrow v_i
+\ident{lookup}_{\cclrd{\ident{?p}_i}} (\ident{Impl}(v, \{\ident{?p}_i\mapsto v_i\}) \rightsquigarrow v_i
 \end{array}
@@
 \paragraph{Properties.}
-
-The core functional language subset (Figure~\ref{fig:semantics-target}) with the comonadically-inspired
-extensions (Figure~\ref{fig:semantics-ext}) and the domain-specific extensions for implicit
-parameters defined in Figure~\ref{fig:semantics-ext-impl} is safe. Furthermore, the
-translation from coeffect calculus produces well-typed programs. Together, these two properties
-also guarantee safety of the source coeffect calculus, when used with implicit parameters.
-
-As for dataflow computations, we prove safety by extending the preservation
+We now prove the type safety of of a context-aware programming language with implicit parameters.
+To do this, we prove safety of the target functional language with specific extensions for implicit
+parameters and we show that the translation from context-aware programming language with implicit
+parameters produces well-typed programs in the target language.
+
+The target language consists of the core functional language subset (Figure~\ref{fig:semantics-target})
+with the comonadically-inspired extensions (Figure~\ref{fig:semantics-ext}) and the domain-specific
+extensions for implicit parameters defined in Figure~\ref{fig:semantics-ext-impl}. The well-typedness
+of the translation has been discussed earlier (Theorem~\ref{thm:semantics-welltyped}) and we extend
+it to cover operations specific for implicit parameters below (Theorem~\ref{thm:semantics-impl-welltyped}).
+
+As for dataflow computations, we prove the type safety by extending the preservation
 (Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog})
@@
 the translated program $f$ obtained using the rules in Figure~\ref{fig:semantics-translation}
-and Figure~\ref{fig:semantics-ext-impl} is well-typed, \ie~in the target langauge: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
+and Figure~\ref{fig:semantics-ext-impl} is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
 \end{theorem}
@@
 \label{thm:semantics-impl-pres-subst}
-For all $\Gamma, e, e' \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
+For all $\Gamma, e, e', \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
 then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
@@
 \label{thm:semantics-impl-pres}
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow e'$ then $\Gamma \vdash e' : \tau$
 \end{theorem}
 \begin{proof}
-  Rule induction over $\rightarrow$.
+  Rule induction over $\rightsquigarrow$.
 
@@
 \label{thm:semantics-impl-prog}
-  If $\vdash e : \tau$ then either $e$ is a value or there exits $e'$ such that $e \rightarrow e'$
+  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
 \end{theorem}
@@
 
-\begin{theorem}[Safety]
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow^* e'$ then either $e'$ is a value or there
-  exists $e''$ such that $e' \rightarrow e''$
+\begin{theorem}[Safety of context-aware langauge with implicit parameters]
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^* e'$ then either $e'$ is a value of type $\tau$ or
+  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
 \end{theorem}
 \begin{proof}
-  Rule induction over $\rightarrow^*$ using Theorem~\ref{thm:semantics-impl-pres} and Theorem~\ref{thm:semantics-impl-prog}.
+  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:semantics-impl-pres} and Theorem~\ref{thm:semantics-impl-prog}.
 \end{proof}
@@
 
-\section{Generalised safety of comonadic embedding}
+\section{Generalized safety of comonadic embedding}
 \label{sec:semantics-generalising}
@@
 In Section~\ref{sec:semantics-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
-safety property of two concrete context-aware programming langauges based on the coeffect
-language framework. The proofs for the two systems were very similar and they relied on the
+safety property of two concrete context-aware programming languages based on the coeffect
+language framework. The proofs for the two systems were very similar and relied on the
 same key principle.
@@
 {\footnotesize(\emph{counit})} &
-  \ident{counit}_{\cclrd{ \ident{L} }}~(\ident{Just}~v)   \rightarrow v
+  \ident{counit}_{\cclrd{ \ident{L} }}~(\ident{Just}~v)   \rightsquigarrow v
 \\[0.5em]
 {\footnotesize(\emph{cobind-1})} &
-  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~v)
+  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightsquigarrow \ident{Live}~(f~v)
 \\
 {\footnotesize(\emph{cobind-2})} &
-  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightarrow \ident{Live}~(f~\ident{Dead})
+  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{L}}}~f~(\ident{Live}~v) \rightsquigarrow \ident{Live}~(f~\ident{Dead})
 \\
 {\footnotesize(\emph{cobind-3})} &
-  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
+  \ident{cobind}_{\cclrd{\ident{L}}, \cclrd{\ident{D}}}~f~v \rightsquigarrow \ident{Dead}
 \\
 {\footnotesize(\emph{cobind-4})} &
-  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{D}}}~f~v \rightarrow \ident{Dead}
+  \ident{cobind}_{\cclrd{\ident{D}}, \cclrd{\ident{D}}}~f~v \rightsquigarrow \ident{Dead}
 \end{array}
@@
 (\emph{cobind-1}) and (\emph{cobind-2}) will not get stuck because the value will be
-$\ident{Live}~v$ and not \ident{Dead}. The reduction rules also preserve typing - the resulting
+$\ident{Live}~v$ and not \ident{Dead}. The reduction rules also preserve typing -- the resulting
 value is of type $\ctyp{\cclrd{s}}\tau_2$, that is $\ident{Live}~v$ for (\emph{cobind-1}),
@@
 The liveness example can be encoded in Haskell using type-level features such as generalized
-algebraic data types (GADTs) and type families \cite{haskell-faking,haskell-promotion,haskell-families}.
+algebraic data types (GADTs) and type families \cite{haskell-faking,haskell-promotion,haskell-families},
+which encode some of the features known from dependently-typed languages such as Agda \cite{other-agda}.
 We do not aim to give a complete implementation, but to show that such encoding is possible and
@@
 %
-Here, the additional type annotation ensures that \ident{counit} can only be called on a context
-that contains a value and so calling the operation is not going to fail. Similarly, the type of the
-\ident{cobind} operation now guarantees that if a function (used as the first argument) or the result
-require a live context, it will be called with a value $\ident{C}~\ident{L}~a$ that is guaranteed
-to contain a value.
+Here, the additional type parameter is used as a phantom type \cite{types-phantom} and ensures that
+\ident{counit} can only be called on a context that contains a value and so calling the operation
+is not going to fail. Similarly, the type of the \ident{cobind} operation now guarantees that if a
+function (used as the first argument) or the result require a live context, it will be called with
+a value $\ident{C}~\ident{L}~a$ that is guaranteed to contain a value.
 
@@
 type system than monads. For monadic computations, it is sufficient to use a type $\ident{M}~a$
-which represents that \emph{some} effects may happen. The the type does not specify which effects
+which represents that \emph{some} effect may happen. The the type does not specify which effects
 and, indeed, this means that \emph{all possible effects} may happen.
 
-With coeffects, we need indexed comonads $\ident{C}~r~a$ where the annotation $r$ specifies what
-context may be required. Without the annotation, a type $\ident{C}~a$ would represent a comonadic
-context that has \emph{all possible context} available, which is rarely useful in practice.
+With coeffects, we need to use indexed comonads $\ident{C}~r~a$ where the annotation $r$ specifies
+what context may be required. Without the annotation, a type $\ident{C}~a$ would represent a
+comonadic context that has \emph{all possible context} available, which is rarely useful in practice.
 

### 2016-04-27 text/Chapter05.tex
Finish edits in Chapters 4 and 5

@@
 
-\newpage
-~
-\newpage
 \paragraph{Properties.}
@@
 
-\newpage
-
 \vspace{0.25em}\noindent\hangindent=0.6cm
@@
 \end{proof}
-
-\newpage
 
@@
 the context changes from providing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to providing
-just implicit parameters $\cclrd{s}$.
-Thus \emph{indexed comonads} are a generalization of \emph{comonads} that captures structures that
-do not form a (non-indexed) comonad. However, indexing has been also discussed in the case of
-\emph{monads.}
+just implicit parameters $\cclrd{s}$. Thus \emph{indexed comonads} are a generalization of
+\emph{comonads} that captures structures that fail to form a comonad without indexing. In the
+rest of the section, we look at work that discusses indexing in the context of \emph{monads.}
 
@@
 \noindent
-A \emph{family of comonads} is more restrictive than an \emph{indexed comonad}, because each of the data
+A \emph{family of comonads} is not as expressive as an \emph{indexed comonads}. Many indexed
+comonads cannot be captured by a family of comonads. This is because each of the data
 types needs to form a comonad separately. For example, our indexed Maybe does not form a family of
@@
 indexed comonads) are closely related to our definition.  Although presented in a more general
-categorical framework (and using monads), the model defines the \ident{unit} operation only on the
-unit of a monoid and the \ident{bind} operation composes effect annotations using the provided monoidal
-structure.
+categorical framework (and using monads), the model (i) defines the \ident{unit} operation only on the
+unit of a monoid and (ii) the \ident{bind} operation composes effect annotations using the provided
+monoidal structure.
 
@@
   Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
-  computation, many of which cannot be captured by a monad (\eg~liveness or data-flow).
+  computation, many of which cannot be captured by a monad (\eg~liveness or dataflow).
+  In terms of program analysis \cite{other-optcomp}, monas capture forward dataflow analyses and
+  comonads correspond to backward dataflow analyses.
 
@@
   sequential composition and context merging ($\cpar, \cseq$, and $\czip$) while most effect systems
-  only use a single operation for sequential composition (used by monadic bind).
+  only use a single operation for sequential composition (used by monadic bind). Effect systems
+  may use a richer algebraic structure to support additional language constructs such as conditionals
+  \cite{effects-nielson,effects-revisited}, but not for abstraction and application.
 
@@
   coeffect systems, the context demands of the body can be split between (or duplicated
-  at) declaration site and call site, while monadic effect systems always defer all effects.
+  at) declaration site and call site, while lambda abstraction in monadic effect systems always
+  defer all effects -- creating a function value has no effect.
 \end{itemize}
@@
 \noindent
-Despite the differences, our implicit parameters example can be also represented by a monad.
-Semantically, the \emph{reader} monad is equivalent to the \emph{product} comonad. Syntactically,
-we use the $\cup$ operation for all three operations of the coeffect algebra. However, to enable
-splitting of implicit parameter demands using the reader monad, we need to extend the
-monad structure and change the translation of monadic lambda abstraction.
+Despite the differences, our implicit paremeters resemble, in many ways, the \emph{reader} monad.
+As discussed in Section~\ref{sec:semantics-related-monads}, the \emph{reader} monad is semantically
+equivalent to the \emph{product} comonad when we consider just sequential composition. For a
+language with lambda abstraction, we need a slight extension to the usual treatment of monads in
+order to model implicit parameters using a monad.
 
@@
 Before looking at the necessary extensions, consider the two ways of modelling implicit
-parameters. We assume that the function $\cclrd{r} \rightarrow \sigma$ is a lookup function
+parameters. We assume that the function $\cclrd{r} \rightarrow \ident{num}$ is a lookup function
 for reading implicit parameter values that is defined on a set $\cclrd{r}$. The two definitions
@@
 \noindent
+This equivalence shows an intriguing relationship between the \emph{product comonad} and
+\emph{reader monad}, but it cannot be extended beyond that. In particular, comonads that model
+dataflow computations or liveness do not have a corresponding monadic structure.
 This equivalence holds for monads and comonads (as well as \emph{indexed} monads
 and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
-the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging.
+the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging. This can be
+supported in monadic computations by adding an aditional operation discussed next.
 
@@
 \end{equation*}
+%
+This suggests that the \emph{reader monad} is a special case among monads. Our work suggests that
+passing read-only information to a computation is better captured by a product comonad, which also
+matches the intuition -- read-only information is a \emph{contextual capability}.
 
@@
 
-In the previous chapter, we defined \emph{type system} for flat coeffect calculi that uniformly
+In the previous chapter, we defined a \emph{type system} for flat coeffect calculi that uniformly
 captures the shared structure of context-aware computations. In this chapter, we completed
@@
 that turns programs written in a domain-specific coeffect language into a functional target
-language. This is akin to the Haskell do notation for monads. Finally, we extended the target
-language with concrete implementation of comonadic operations for dataflow and implicit parameters
-and we presented syntactic safety proof. In summary, the proof states that well-typed
+language. This is akin to the Haskell do-notation for monads. Finally, we extended the target
+language with concrete implementations of comonadic operations for dataflow and implicit parameters
+and we presented a syntactic safety proof. In summary, the proof states that well-typed
 context-aware programs written in a coeffect language \emph{do not go wrong} (when translated

### 2016-04-30 text/Chapter04.tex
Some more edits (Chapter 5)
Unique structural derivation and edits (Chapter 6)

@@
 
-\setcounter{chapter}{3}
 \chapter{Types for flat coeffects}
@@
 \vspace{0.5em}\noindent\hangindent=0.7cm
-Cases (\emph{var}), (\emph{const}), (\emph{app}), (\emph{let}), (\emph{param}) and (\emph{letpar})
+Cases (\emph{app}), (\emph{let}) and (\emph{letpar})
   similarly to (\emph{abs}).
@@
 \label{thm:flat-ldf-unique}
-In the type system for implicit parameters defined in Figure~\ref{fig:flat-types} with the (abs)
+In the type system for liveness and dataflow defined in Figure~\ref{fig:flat-types} with the (abs)
 rule replaced by (idabs), when $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and
@@
 \vspace{0.5em}\noindent\hangindent=0.6cm
-Case (\emph{abs}): $e = \lambda x\!:\!\tau_1.e_1$. $\tau = \tau_1 \xrightarrow{\cclrd{c}} \tau_2$
+Case (\emph{abs}): $e = \lambda x\!:\!\tau_1.e_1$. Then $\tau = \tau_1 \xrightarrow{\cclrd{c}} \tau_2$
   for some $\tau_2$ and $\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{c}} \vdash e : \tau_2}$.
@@
   By the induction hypothesis $\tau_2 = \tau_2'$ and $\cclrd{c}=\cclrd{c'}$ and therefore
-  also so $\tau = \tau'$.
+  also $\tau = \tau'$.
 
 \vspace{0.5em}\noindent\hangindent=0.7cm
-Cases (\emph{var}), (\emph{const}) are direct consequence of Lemma~\ref{thm:flat-impl-invert}.
+Cases (\emph{var}), (\emph{const}) are direct consequence of Lemma~\ref{thm:flat-ldf-invert}.
 
 \vspace{0.5em}\noindent\hangindent=0.7cm
-Cases (\emph{var}), (\emph{const}), (\emph{app}), (\emph{let}), (\emph{param}) and (\emph{letpar})
-  similarly to (\emph{abs}).
+Cases (\emph{app}) and (\emph{let}) similarly to (\emph{abs}).
 \end{proof}

### 2016-04-30 text/Chapter06.tex
Unique structural derivation and edits (Chapter 6)

@@
 
-In this chapter, we take the next step towards a unified theory of context-aware programming
-languages. As already discussed, we attempt to find abstractions that are \emph{general enough} 
-to capture a wide range of useful programming language features, but \emph{specific enough} to 
-let us identify and capture interesting properties of the context.
-
 In Chapter~\ref{ch:applications}, we discussed two notions of context. Context-aware programming
-languages that that capture whole-context properties were generalized by the \emph{flat coeffect 
-calculus} in Chapter~\ref{ch:flat}. Here, we consider per-variable contextual properties
-and we introduce the \emph{structural coeffect calculus}.
+languages that capture whole-context properties were generalized by the \emph{flat coeffect
+calculus} in Chapters~\ref{ch:flat} and \ref{ch:semantics}. Here, we consider per-variable
+contextual properties and we introduce the \emph{structural coeffect calculus}.
 
 The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness
-and data-flow), but provides relatively imprecise approximation and weak syntactical properties. 
+and dataflow), but provides relatively imprecise approximation and weak syntactical properties.
 Some of those examples can be also seen as per-variable properties. For those, structural coeffect
@@
 
-As we will see, the categorical semantics of the structural coeffect calculus is based on the
-same underlying structure as the semantics of the flat calculus. However, the syntactical properties
-of the calculus are quite different and so it is worth considering it in a separate chapter.
+We mirror the development for flat coeffect calculus and develop a small calculus with a type system
+that captures per-variable contextual properties. We outline its categorical semantics and use it as
+a basis for a translation that turns well-typed programs in context-aware languages into well-typed
+programs in a simple target functional language. We prove syntactic safety for a sample target
+language, showing that ``well-typed context-aware programs do not get stuck''.
 
@@
   by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the
-  system captures fine-grained liveness and data-flow information, as well as a calculus for bounded 
+  system captures fine-grained liveness and dataflow information, as well as a calculus for bounded
   reuse (checking how many times is a variable accessed).
 
-\item We extend indexed comonads introduced in the previous chapter to 
-  \emph{structural indexed comonads} and use them to provide the semantics of structural 
-  coeffect calculus (Section~\ref{sec:struct-semantics}). As with the flat version, the
-  semantics provides a motivation for the design of the calculus.
+\item We present a syntax-directed version of the calculus that is used to obtain unique typing
+  derivation for programs in structural coeffect calculus (Section~\ref{sec:struct-unique}).
+  Unlike in flat systems, the procedure for choosing a unique typing derivation is common to all
+  structural systems.
 
@@
   for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}).
+
+\item We extend indexed comonads introduced in the previous chapter to \emph{structural indexed
+  comonads} and use them to provide the semantics of structural coeffect calculus
+  (Section~\ref{sec:struct-semantics}). As with the flat version, the theory serves as a motivation
+  for syntactic translational semantics.
+
+\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
+  programs from the structural coeffect calculus into a simple functional language with
+  uninterpreted comonadically-inspired primitivies. We give concrete operational semantics for the
+  target language for one of our sample langauges and show that well-typed programs, produced by
+  translation from the coeffect calculus do not get stuck.
 \end{itemize}
@@
 
-Compared to the previous chapter, structural coeffect calculi we consider are more 
-homogeneous and so finding the common pattern is in some ways easier. However, the systems
-are more complicated as they need to keep annotations attached to individual variables
-and thus require explicit structural rules. Before looking at the system, we briefly consider
-the most important related work.
+Compared to Chapter~\ref{ch:flat}, structural coeffect calculi we consider are more homogeneous and
+so finding the common pattern is easier. However, the systems are more complicated as they need to
+keep annotations attached to individual variables and thus require explicit structural rules. Before
+looking at the system, we briefly consider the most important related work.
 
@@
 In the previous chapter, we discussed the correspondence between coeffects and effects (and
-between comonads and monads). As noted earlier, the $\lambda$-calculus is asymmetric in that
-an expression has multiple inputs (variables in the context), but just a single result 
-(the resulting value) and so monads and effects have no notion directly corresponding to 
-structural coeffect systems.
+between comonads and monads). As noted in Section~\ref{sec:applications-strucutre-vec}, the
+$\lambda$-calculus is asymmetric in that an expression has multiple inputs (variables in the
+context), but just a single result (the resulting value) and so monads and effects have no notion
+directly corresponding to structural coeffect systems.
 
 The work in this chapter is more closely related to sub-structural type systems \cite{substruct-attpl-intro}.
-While sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
-\emph{exchange} rules, our systems keep them, but use them to manipulate both the context
-and its annotations.
+Sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
+\emph{exchange} rules. In contrast, our systems keep all three structural rules, but use them to
+manipulate the coeffect annotations in a way that matches the variable manipulations.
 
@@
 %
-%       ###           ##                   ##                 
-%      #   #           #                    #                 
-%      #       ###     #     ###   #   #    #    #   #   ###  
-%      #          #    #    #   #  #   #    #    #   #  #     
-%      #       ####    #    #      #   #    #    #   #   ###  
-%      #   #  #   #    #    #   #  #  ##    #    #  ##      # 
-%       ###    ####   ###    ###    ## #   ###    ## #  ####  
+%     #####
+%    #     #   ##   #       ####  #    # #      #    #  ####
+%    #        #  #  #      #    # #    # #      #    # #
+%    #       #    # #      #      #    # #      #    #  ####
+%    #       ###### #      #      #    # #      #    #      #
+%    #     # #    # #      #    # #    # #      #    # #    #
+%     #####  #    # ######  ####   ####  ######  ####   ####
 %
@@
 \begin{array}{rcl}
-e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
-\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
+e &::=& x n \sep \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
 \end{array}
@@
 
-Free variable contexts are annotated with vectors of structural coeffect scalars. In what follows,
-we write the vectors of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging 
-over vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face 
+Free variable contexts are annotated with vectors of structural coeffect scalars. We write vectors
+of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging over vectors are
+written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face
 and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
 is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
-of coeffects is provided by the definition of \emph{structural coeffect algebra} below.
+of coeffects is provided by the definition of \emph{structural coeffect algebra} discussed next.
 
@@
 A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
-$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
-$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
-$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
+$\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
+$(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and a binary relation $\cleq$ such
+that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
 %
@@
 \noindent
-In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
+The structural coeffect scalar structure resembles flat coeffect algebra, but it differs in a
+number of important ways:
+%
+\begin{itemize}
+\item The $\cpar$ operation of structural coeffect scalar is not required to be idempotent. In structural
+  systems, we can track individual variable accesses and not requiring idempotence allows interesting
+  systems such as the one for bounded reuse (Section~\ref{sec:struct-semantics-examples}).
+
+\item In the flat coeffect calculus, we used the $\czip$ operation to merge the annotations of contexts
   available from the declaration site and the call site or, in the syntactic reading, to split the
-context requirements. The other difference is that the $\cpar$ operation of structural coeffect
-scalar is not required to be idempotent. The bounded reuse example uses $\cpar$ which is not 
-idempotent.
-
-Finally, we now also require two properties that relate the ordering $\cleq$ with coeffects composed
+  context requirements. In structural systems, this is handled directly by vectors and so $\czip$
+  is no longer needed.
+
+\item We now require two properties that relate the ordering $\cleq$ with coeffects composed
   via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma
@@
   (Lemma~\ref{thm:structural-substitution}).
-
-In the structural coeffect calculus, we use a vector instead -- combining and splitting of 
-coeffects becomes just vector concatenation or splitting, respectively; these are provided
-by the tensor product. The operations on vectors are indexed by integers representing the lengths
-of the vectors. The additional structure required by the type system for structural coeffect 
-calculi is given by the following definition.
+\end{itemize}
+
+\noindent
+In the structural coeffect calculus, the scalar coeffect structure is supplemented by a vector
+structure. Combining and splitting of coeffects becomes just vector concatenation or splitting,
+respectively; these are provided by the tensor product. The operations on vectors are indexed by
+integers representing the lengths of the vectors. The additional structure required by the type
+system for structural coeffect calculi is given by the following definition.
 
@@
 can be determined from the lengths of the matching free variable context vectors.
-
 More generally, we could see the the coeffect annotations as \emph{containers}
@@
 
-% -------------------------------------------------------------------------------------------------
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Syntax-driven typing rules:}
+\begin{equation*}
+\tyrule{var}
+  { }
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  { }
+  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x:\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Structural rules for context manipulation:}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+\begin{array}{l}
+ \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{2.5em}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~
+\begin{array}{l}
+ \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\vspace{1em}
+{\small c.) Non-syntax-directed subcoeffecting rule:}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+(\cclrd{s'} \cleq \cclrd{s})
+\end{equation*}
+
+\figcaption{Type system for the structural coeffect calculus}
+\label{fig:struct-types}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
 
@@
 \paragraph{Structural rules.}
-The remaining rules, shown in Figure~\ref{fig:struct-types} (b), are not syntax-directed.
-They allow different transformation of the free variable context. We include (\emph{sub}) 
-as one of the rules, allowing sub-coeffecting on coeffect scalars belonging to individual 
-variables. The remaining rules capture \emph{weakening}, \emph{exchange} and \emph{contraction} 
-known from sub-structural systems.
+The rules shown in Figure~\ref{fig:struct-types} (b) are not syntax-directed and
+allow different transformation of the free variable context. They correspond to the
+transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
+from sub-structural systems.
 
@@
 
-%---------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-{\small a.) Syntax-driven typing rules:}
-\begin{equation*}
-\tyrule{var}
-  { }
-  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{const}
-  {c \!:\! \tau \in \Delta}
-  {\coctx{()}{\alift{}} \vdash c : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
-   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-
-\vspace{1em}
-{\small b.) Structural rules for context manipulation:}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-(\cclrd{s'} \cleq \cclrd{s})
-\end{equation*}
-\begin{equation*}
-\tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
-  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau} 
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-\begin{array}{l}
- \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
- \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{2.5em} 
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-~
-\begin{array}{l}
- \xlen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
- \xlen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-
-\figcaption{Type system for the structural coeffect calculus}
-\label{fig:struct-types}
-\end{figure}
+\paragraph{Subcoeffecting.}
+Finbally, Figure~\ref{fig:struct-types} (c) shows the subcoeffecting rule (\emph{sub}).
+As with the flat coeffect calculus, we include it mainly to guide the understanding of the
+system, but we treat it as an additional rule and do not discuss it in the subsequent theory.
+The rule would be needed in languages with conditionals and other language features. It is
+worth noting that allowing subcoeffecting on coeffect scalars belonging to individual variables
+is sufficient for structural systems.
 
@@
 The other aspect that makes structural systems simpler is that they remove the need for
-separate let binding. As discussed in Section~\ref{sec:flat-calculus-let}, flat calculi include
+separate let binding. As discussed in Section~\ref{sec:flat-exts-let}, flat calculi include
 let binding that gives a \emph{more precise} typing than combination of abstraction and
@@
 The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
-presented in Section~\ref{sec:applications-structural}. Two of them -- structural data-flow and
+presented in Section~\ref{sec:applications-structural}. Two of them -- structural dataflow and
 structural liveness provide a more precise tracking of properties that can be tracked using
@@
 
-\begin{example}[Structural data-flow]
-In data-flow, context is annotated with natural numbers and the structural coeffect scalar is formed 
+\begin{example}[Structural dataflow]
+In dataflow, context is annotated with natural numbers and the structural coeffect scalar is formed
 by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
@@
 and so the above definitions are both valid. Similar construction can be used for the
-\emph{optimized data-flow} example from Section~\ref{sec:flat-calculus-examples}.
+\emph{optimized dataflow} example from Section~\ref{sec:flat-calculus-examples}.
 
 It is important to note that this gives us a systems with \emph{different} properties. The
-information are now tracked per-variable rather than for entire contexts. For data-flow,
+information are now tracked per-variable rather than for entire contexts. For dataflow,
 we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$
@@
 \noindent
-Similarly to the structural calculus for data-flow, the calculus for bounded variable reuse
+Similarly to the structural calculus for dataflow, the calculus for bounded variable reuse
 annotates each variable with an integer. However, the integer denotes how many times is the variable
@@
 %
-%      #####  #                                 
-%        #    #                                 
-%        #    # ##    ###    ###   # ##   #   # 
-%        #    ##  #  #   #  #   #  ##  #  #   # 
-%        #    #   #  #####  #   #  #      #  ## 
-%        #    #   #  #      #   #  #       ## # 
-%        #    #   #   ###    ###   #          # 
-%                                         #   # 
-%                                          ###                                                                         
+%       #
+%      # #   #    # #####   ####  # #    # # ##### #   #
+%     #   #  ##  ## #    # #    # # #    # #   #    # #
+%    #     # # ## # #####  #      # #    # #   #     #
+%    ####### #    # #    # #  ### # #    # #   #     #
+%    #     # #    # #    # #    # # #    # #   #     #
+%    #     # #    # #####   ####  #  ####  #   #     #
+%
 % ==================================================================================================
 
-\section{Categorical motivation}
-\label{sec:struct-semantics}
-
-When we introduced structural coeffect systems in Section~\ref{sec:applications-structural}, we 
-included a concrete semantics of structural liveness and bounded variable reuse. In this section,
-we generalize the examples using the notion of \emph{structural indexed comonad}, which is an 
-extension of the \emph{indexed comonad} structure. Similarly to \emph{flat indexed comonad}, the
-structural variant adds operations that are needed to embed full $\lambda$-calculus, this time 
-with per-variable contexts.
-
-As in the previous chapter, the main aim of this section is to motivate and explain the design of 
-the structural coeffect calculus shown in Section~\ref{sec:struct-calculus}. However, the semantics 
-also highlights the similarities and differences between the flat and structural system.
-Most of the differences arise from the fact that contexts in structural coeffect systems are treated 
-as \emph{vectors} rather than sets.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of vectors}
-
-Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing 
-judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a 
-morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model, 
-we can freely transform the value contained in the context modelled using an indexed comonad
-$\ctyp{\cclrd{r}}$.
-
-Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
-which transforms the value inside the context without affecting the coeffect annotation. Thus,
-we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
-$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value 
-$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
-
-The ability to freely transform the variable structure is not desirable in the model of 
-structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
-the coeffect annotations matches the structure of variables. To achieve this, we model vectors
-using a structure distinct from ordinary products which we denote $-\xtimes-$. 
-For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
-is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
-We assume that the operator is equipped with necessary associativity transformations allowing us
-to use it freely on more than two values.
-
-The operator is a bifunctor, but it is \emph{not} a product in the categorical sense. In particular, 
-there is no way to turn $\tau_1 \xtimes \ldots \xtimes \tau_n$ into $\tau_i$ (the structure does 
-not have projections) and so there is also no way of turning
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$ into
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{\tau_i}$, which would break the correspondence 
-between coeffect annotations and variable structure.
-
-The structure created using $-\xtimes-$ can be manipulated only using operations provided by
-the \emph{structural indexed comonad}, which operate over variable contexts contained in an 
-indexed comonad $\ctyp{\aclrd{\textbf{r}}}$ and are designed to preserve the correspondence
-between vectors and annotations.
-
-In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
-As mentioned, we assume that the use of the operator can be freely re-associated. For example, when
-calling an operation that requires input of the form 
-$(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
-we use an argument $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ and assume that the 
-appropriate transformation is inserted.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Indexed comonads, revisited}
-
-The semantics of structural coeffect calculus reuses the definition of \emph{indexed comonad} almost
-without change. The additional structure that is required for context manipulation (merging and 
-splitting) is different and is here provided by the \emph{structural indexed comonad} structure that 
-we introduce in this section.
-
-Recall the definition from Section~\ref{sec:flat-semantics-idx}, which defines an indexed como\-nad 
-over a monoid $(\C, \cseq, \cunit)$ as a triple $(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, 
-\ident{cobind}_{\cclrd{r}, \cclrd{s}})$. The triple consists of a family of object mappings $\ctyp{\cclrd{r}{}}$,
-and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
-
-In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$
-representing function values (appearing in the language), but also of the form
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$, 
-modelling expressions in a context. To capture this, we need to revisit the definition and use 
-\emph{coeffect vectors} in some of the operations.
-
-\begin{definition}
-Given a monoid $(\C, \cseq, \cunit)$ with a pointwise extension of the $\cseq$ operator to a vector 
-(written as $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$) and an operation lifting scalars to vectors 
-$\alift{-}$, an \emph{indexed comonad} over a category $\catc$ is a triple 
-$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}})$:
-
-\begin{compactitem}
-\item $\ctyp{\aclrd{\textbf{r}}}{}$ for all $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$ is a family of object mappings 
-\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\alift{\cunit}}{\alpha} \rightarrow \alpha$ 
-\item $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$ is a mapping $(\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta) 
-  \rightarrow (\ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{s}}}{\beta})$
-\end{compactitem}
-\end{definition}
-
-\noindent
-The object mapping $\ctyp{\aclrd{\textbf{r}}}$ is now indexed by a vector rather than by a scalar 
-$\ctyp{\cclrd{r}}$ as in the previous chapter. This new definition supersedes the old one, because a 
-flat coeffect annotation can be seen as singleton vectors.
-
-The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
-return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
-operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
-is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
-takes a vector of variables, but always produces just a single value. Thus the resulting function 
-also takes a vector of variables, but always returns a context with a vector containing just one value.
-In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
-no way of constructing values containing $\xtimes$.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Structural indexed comonads}
-
-The flat indexed comonad structure extends indexed comonads with operations 
-$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ that combine or 
-split the additional (flat) context and are annotated with the flat coeffect operations $\czip$ 
-and $\cpar$, respectively.
-
-In the structural version, the corresponding operations operate convert between a (wrapped) vector 
-of values represented using $\xtimes$ and ordinary pairs of contexts containing parts of the vector. 
-The vectors of coeffect annotations are split or merged using $\atimes$ of the structural coeffect 
-algebra, in a way that mirrors the wrapped vectors (variable structure).
-
-The following definition includes $\ident{lift}_{\cclrd{r'},\cclrd{r}}$, which is similar as
-before and models sub-coeffecting and also $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ which models
-duplication of a variable in a context needed for the semantics of contraction:
-
-\begin{definition}
-Given a structural coeffect algebra formed by $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$
-with operations $\alift{-}$ and $\aseq$, a \emph{structural indexed comonad} is an indexed comonad over 
-the monoid $(\C, \cseq, \cunit)$ equipped with families of operations $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$, 
-$\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$,
-$\ident{dup}_{\cclrd{r}, \cclrd{s}}$ and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
-%
-\begin{compactitem}
-\item $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
-  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}$
-\item $\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
-  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
-\item $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ is a family of mappings
-  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \xtimes \alpha)}$
-\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
-  $\ctyp{\alift{\cclrd{r'}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
-\end{compactitem}
-%
-Here, the following equalities must hold:
-\begin{equation*}
-\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~\equiv~ \ident{id} \qquad\qquad
-\ident{id} ~\equiv~ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} 
-\end{equation*}
-
-\end{definition}
-
-\noindent
-These operations differ from those of the flat indexed comonad in that the \ident{merge} and 
-\ident{split} operations are required to be inverse functions and to preserve the additional 
-information about the context. This was not required for the flat system where the operations
-could under-approximate or over-approximate. Note that the operations use $\xtimes$ to combine 
-or split the contained values. This means that they operate on free-variable vectors rather than 
-on ordinary products.
-
-The \ident{dup} mapping is a new operation that was not required for a flat calculus. It 
-takes a variable context with a single variable annotated with $\cclrd{r}\,\cpar\,\cclrd{s}$,
-duplicates the value of the variable $\alpha$ and splits the additional context between the two
-new variables. In a flat calculus, this operation was expressed using ordinary tuple 
-construction, which is not possible here -- the returned context needs to contain a 
-two-element vector $\alpha \xtimes \alpha$.
-
-Finally, the \ident{lift} mapping is almost the same as in the flat version. It
-operates on a singleton vector, which is equivalent to operating on a scalar as before.
-The operation could be extended to a vector in a pointwise way, but we keep it simple
-and perform sub-coeffecting on individual variables.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}[!t]
-\begin{equation*}
-\begin{array}{ll}
- \sem{\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }~\ctx =
-   \ident{counit}_{\cunit}~\ctx & (\emph{var})
-\\[0.4em]
-\sem{\coctx{\Gamma}{\czero} \vdash c_i : \tau }~\ctx =
-  \delta~(c_i) & (\emph{const})
-\\[0.4em]
-\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\atimes\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}~\ctx = & (\emph{app})\\[-0.25em]
-  \hspace{1.5em}
+\begin{figure}[t]
+\begin{equation*}
+\tyrule{var}
+  { }
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  { }
+  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \\
+   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash e_1~e_2 : \tau_2}
+\quad \coctx{\Gamma}{\aclrd{\textbf{c}}} = \ident{mergevars}(\cclrd{t}, \coctx{\Gamma_1}{\aclrd{\textbf{r}}}, \coctx{\Gamma_2}{\aclrd{\textbf{s}}})
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{t}}} \vdash e : \tau_2}
+  {\coctx{\Gamma_2}{\aclrd{\textbf{r}}} \vdash \lambda x:\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\quad (\coctx{\Gamma_2}{\aclrd{\textbf{r}}}), \tau_1, \cclrd{s} = \ident{findvar}_x(\coctx{\Gamma_1}{\aclrd{\textbf{t}}})
+\end{equation*}
+
+\vspace{1em}
+\begin{equation*}
 \begin{array}{l}
-  \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}~\ctx\\[-0.25em]
-  \kvd{in}~\sem{\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}~\ctx_1~
-      (\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~\sem{\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\ctx_2)
-  \end{array}
-\\[1.4em]
-\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }~\ctx = \lambda v.& (\emph{abs})\\[-0.25em]
-\hspace{1.75em}\sem{\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
-  ~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v))
-\\[0.4em]
-\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}~\ctx = & (\emph{weak}) \\
-  \hspace{1.75em}
-  \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx~~\kvd{in}~~
-  \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}~\ctx_1
-\\[0.4em]
-\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{sub})\\
-  \hspace{1.75em}
-  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~
-  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}, \alift{\cclrd{s'}}, \aclrd{\textbf{q}}}~\ident{lift}_{\cclrd{s}, \cclrd{s'}}~\ctx)
-\\[0.4em]
-\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
-        \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}~\ctx = & (\emph{exch})\\[-0.25em]
-  \hspace{1.75em}  \sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}
-     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
-  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}}, 
-     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx)
-\\[0.4em]
-\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} 
-        \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}~\ctx = & (\emph{contr})\\[-0.25em]
-  \hspace{1.75em}  \sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}
-      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]
-  \hspace{4em}  (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}}, 
-      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)
-\end{array}
-\end{equation*}
-\\[0.0em]
-{\small The definition uses the following auxiliary definitions:}
-\begin{equation*}
-\hspace{-2em}
-\begin{array}{l}
- \ident{swap}_{\cclrd{t},\cclrd{s}} ~:~ 
-  \ctyp{\alift{\cclrd{t}, \cclrd{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\alift{\cclrd{s}, \cclrd{t}}}{(\beta \xtimes \alpha)} \\[-0.25em]
-\ident{swap}_{\cclrd{t},\cclrd{s}}~\ctx = \\
-  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}}~\ctx \\[-0.25em]
-  \hspace{1.75em} \kvd{in}~\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)
- \\[0.4em]
-\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} ~:~
-  (\ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{s'}}}{\beta'}) \rightarrow
-    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta \xtimes \gamma)} \rightarrow
-    \ctyp{\aclrd{\textbf{r}} \atimes \aclrd{\textbf{s'}} \atimes \aclrd{\textbf{t}}}{(\alpha \xtimes \beta' \xtimes \gamma)}\\[-0.25em]
-\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f~\ctx =\\
-  \hspace{1.75em} \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx \\[-0.25em]
-  \hspace{1.75em} \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}~\ctx' \\[-0.25em] 
-  \hspace{1.75em} \kvd{in}~\ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~
-        (\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
-\end{array}
-\end{equation*}
-\figcaption{Categorical semantics of the structural coeffect calculus}
-\label{fig:struct-semantics}
-\vspace{-0.5em}
-\end{figure*}
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of structural calculus}
-
-The concrete semantics for liveness and bounded variable use shown in 
-Sections~\ref{sec:applications-struct-live} and \ref{sec:applications-struct-bll}
-suggests that semantics of structural coeffect calculi tend to be more complex 
-than semantics of flat coeffect calculi. The complexity comes from the fact
-that we need a more expressive representation of the variable context -- \eg~a vector
-of optional values (\ie~wrapped in the \ident{Maybe} data type). Additionally,
-the structural system needs to pass separate variable contexts to the sub-expressions.
-
-The latter aspect is fully captured by the semantics shown in this section. The
-earlier point is left to the concrete notion of structural coeffect. Our model still gives
-us the flexibility of defining the concrete representation of variable vectors. We explore
-a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
-at a unified categorical semantics defined in terms of \emph{structural indexed comonads}.
-
-\paragraph{Contexts and functions.}
-In the structural coeffect calculus, expressions in context are interpreted as functions taking
-a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
-such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
-structure is annotated with a scalar, such as $\ctyp{\cclrd{r}}$, which we treat as being equivalent
-to a singleton vector annotation $\ctyp{\alift{\cclrd{r}}}$:
-%
-\begin{equation*}
-\begin{array}{rcl}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
-  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
-\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
-\end{array}
-\end{equation*}
-%
-Note that the instances of flat indexed comonad ignored the fact that the variable 
-context wrapped in the data structure is a product. This is not generally the case for the 
-structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
-are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
-rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need to examine the structure
-of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
-
-\paragraph{Expressions.}
-A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. As in
-the previous chapter, the semantics is written in a programming language style using constructs
-such as let-binding rather than using a categorical (point-free) notation. As before, the semantics
-can equally be written using primitives such as currying, uncurrying, function pairing etc.
-
-The following summarizes how the standard syntax-driven rules work, highlighting the differences
-from the flat version:
-
-\begin{itemize}
-\item When accessing a variable (\emph{var}), the context now contains \emph{only} the accessed
-  variable and so the semantics is just $\ident{counit}_{\cunit}$ without a projection.
-  Constants (\emph{const}) are interpreted using a global dictionary $\delta$ as earlier.
-
-\item The semantics of flat function application first duplicated the context so that the same 
-  variables can be passed to both sub-expressions. This is no longer needed -- the 
-  (\emph{app}) rule splits the variables \emph{including} the additional context into two parts.
-  Passing the first context to the semantics of $e_1$ gives us a function
-  $\ctyp{\alift{\cclrd{t}}}{\tau_1} \rightarrow \tau_2$.
-
-  A value $\ctyp{\alift{\cclrd{t}}}{\tau_1}$ required in order to call the function is obtained by 
-  applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ to the semantics of $e_2$. The result 
-  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
-  is then called with the latter part of the split input context.
-
-\item The semantics of function abstraction (\emph{abs}) is syntactically the same as in the 
-  flat version -- the only difference is that we now merge a free-variable context with a 
-  singleton vector, both at the level of variable assignments and at the level of coeffect 
-  annotations.
-\end{itemize}
-
-\noindent
-The semantics for the non-syntax-driven rules (weakening, exchange, contraction) performs 
-transformations on the free-variable context. Weakening (\emph{weak}) splits the context and 
-ignores the part corresponding to the removed variable. If we were modelling the semantics in 
-a language with a linear type system, this would require an additional operation for ignoring an
-unused context annotated with $\aclrd{\czero}$.
-
-The remaining rules perform a transformation anywhere inside the free-variable vector. 
-To simplify writing the semantics, we define a helper 
-$\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
-that splits the variable vector into three parts, transforms the middle part and then 
-merges them, using the newly transformed middle part.
-
-The transformations on the middle part are quite simple. The (\emph{sub}) rule 
-uses $\ident{lift}_{\cclrd{s}, \cclrd{s'}}$ to discard some of the available additional context;
-the (\emph{exch}) rule swaps two single-variable contexts and the (\emph{contr}) rule
-uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$ operation to duplicate a variable while 
-splitting its additional context.
-
-\paragraph{Properties.}
-
-As in the flat calculus, the main reason for defining the categorical semantics in this chapter
-is to provide validation for the design of the calculus. The following correspondence theorem
-states that the annotations in the typing rules of the structural coeffect calculus correspond 
-to the indices of the semantics. Thus, the calculus captures a context-dependent property if
-it can be modelled by a \emph{structural indexed comonad}. As we show in the next section, 
-this is the case for all three discussed examples (liveness, data-flow, bounded variable reuse).
-
-\begin{theorem}[Correspondence]
-In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$ 
-and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types 
-$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$ 
-and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined 
-by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
+\ident{exch}_x(\coctx{\Gamma}{\aclrd{\textbf{t}}})~=~(\coctx{\Gamma_1,\Gamma_2}{\aclrd{\textbf{t}_1} \atimes \aclrd{\textbf{t}_2}}), \tau, \cclrd{s} \quad \textnormal{where} \\
+\qquad \xlen{\Gamma_1} = \alen{\aclrd{\textbf{t}_1}}~\textnormal{and}~\xlen{\Gamma_2} = \alen{\aclrd{\textbf{t}_2}} \\[-0.25em]
+\qquad x\!:\!\tau \in \Gamma~\textnormal{and}~
+  \coctx{\Gamma}{\aclrd{\textbf{t}}} = \coctx{\Gamma_1, x\!:\!\tau, \Gamma_2}{\aclrd{\textbf{t}_1} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{t}_2}}
+\\[0.5em]
+\ident{findvar}_x(\coctx{\Gamma}{\aclrd{\textbf{t}}})~=~ (\coctx{\Gamma}{\aclrd{\textbf{t}}}), \tau, \czero \qquad (\textnormal{otherwise})
+\\[1em]
+\ident{mergevars}(\cclrd{t}, \coctx{\Gamma_1}{\aclrd{\textbf{r}}}, \coctx{\Gamma_2}{\aclrd{\textbf{s}}})~=~
+  \coctx{\Gamma'_1, \Gamma'_2, \Gamma}{\aclrd{\textbf{r}'}\atimes(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}'}) \atimes \aclrd{\textbf{c}} } \quad \textnormal{where}\\
+\qquad \coctx{\Gamma_1}{\aclrd{\textbf{r}}} =
+    \coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}} \\
+\qquad \coctx{\Gamma_2}{\aclrd{\textbf{s}}} =
+    \coctx{y_1\!:\!\tau'_1, \ldots, y_m\!:\!\tau'_n}{\alift{\cclrd{s_1}, \ldots, \cclrd{s_m}}} \\
+\qquad \coctx{\Gamma}{\aclrd{\textbf{c}}} =
+    \coctx{z_1\!:\!\tau''_1, \ldots, z_k\!:\!\tau''_k}{\alift{\cclrd{c_1}, \ldots, \cclrd{c_k}}} \\[-0.25em]
+\qquad\quad \textnormal{such that}~\forall l\in \{1 \ldots k\}~\exists i,j.~(z_l\!:\!\tau''_l = x_i\!:\!\tau_i = y_j\!:\!\tau'_j) \\[-0.25em]
+\qquad\qquad \textnormal{and}~\cclrd{c_l} = \cclrd{r_i} \;\cpar\; (\cclrd{t} \,\cseq\, \cclrd{s_j}) \\
+\qquad \coctx{\Gamma_1}{\aclrd{\textbf{r}}} =
+    \coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}
+    ~\textnormal{such that}~x_i\!:\!\tau_i \notin \Gamma \\
+\qquad \coctx{\Gamma_2}{\aclrd{\textbf{s}}} =
+    \coctx{y_1\!:\!\tau'_1, \ldots, y_m\!:\!\tau'_n}{\alift{\cclrd{s_1}, \ldots, \cclrd{s_m}}}
+    ~\textnormal{such that}~y_i\!:\!\tau'_i \notin \Gamma \\
+\end{array}
+\end{equation*}
+
+\figcaption{Syntax-directed type system for the structural coeffect calculus}
+\label{fig:struct-syntax-types}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
+
+\section{Choosing a unique typing}
+\label{sec:struct-unique}
+
+In the structural coeffect calculus, the lambda abstraction rule does not introduce ambiguity
+in the typing. This is in contrast with flat coeffect systems (most importantly, the one for
+implicit parameters), where lambda abstraction allowed arbitrary splitting of context requirements.
+In structural coeffect systems, the context requirements placed on the call-site (attached to
+the function type) are those of the bound variable.
+
+However, the type system for structural coeffect calculus in Figure~\ref{fig:struct-types}
+introduces another kind of ambiguity due to the fact that non syntax-directed structural rules
+can be applied repeatedly and in arbitrary order. As with the semantics for flat coeffect calculus
+in Chapter~\ref{ch:semantics}, we define the semantics of the structural coeffect calculus over
+a \emph{typing derivation} and so the meaning of a program depends on the typing derivation chosen.
+In this section, we specify how to choose the desired \emph{unique} typing derivation.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Syntax-directed type system}
+
+In order to choose a unique typing derivation, we follow the example of sub-structural type systems
+\cite{substruct-attpl-intro} and introduce a syntax-directed version of the type system (also called
+\emph{algorithmic}). This replaces the non-syntax-directed rules for weakening, exchange and
+contraction with additional logic in places where contexts are combined. Given a typing derivation
+in the syntax-directed type system, we can then produce a matching typing derivation in the original
+type system.
+
+The syntax-directed version of the type system is shown in Figure~\ref{fig:struct-syntax-types}.
+The typing rules for variables (\emph{var}) and constants (\emph{const}) are the same as before.
+The two interesting rules are lambda abstraction and application.
+
+\paragraph{Lambda abstraction.}
+In lambda abstraction (\emph{abs}) in Figure~\ref{fig:struct-types}, we assume that the bound
+varible is the last variable of the context. In the syntax-directed system, we do not make the
+same assumption. Instead, we use an auxiliary relation $\ident{findvar}_x$ that takes a typing
+context $\coctx{\Gamma}{\aclrd{\textbf{t}}}$ and returns a context with the variable $x$ removed
+together with its type and coeffect.
+
+The $\ident{findvar}_x$ relation is defined by two disjoint cases. If the variable $x$ is not
+present in the context, the context is returned as is and the returned coeffect is $\czero$.
+This case corresponds to the (\emph{weak}) structural rule. If the variable is not found in the
+context, the relation does not place any restrictions on the type of the variable and so $\tau$
+can be any type (this is the only reason why $\ident{findvar}$ is a relation rather than a function).
+If the variable $x$ appears in the context then $\ident{findvar}_x$ removes it together with its
+corresponding coeffect annotation and returns it as the result.
+
+\paragraph{Function application.} In the (\emph{app}) rule in Figure~\ref{fig:struct-types}, we
+assume that the variable contexts of the two sub-expressions can be merged. This requires that they
+contain disjoint variables, which can be always obtained by exchange and contraction. In the
+syntax-driven systenm, we merge coeffects of shared variables explicitly. This is done in the
+\ident{mergevars} function.
+
+The \ident{mergevars} function returns context consisting of three parts. Parts $\Gamma_1$
+and $\Gamma_2$ represent variables that appear only in the first or the second context; part
+$\Gamma$ contains common variables. The coeffect annotations corresponding to $\Gamma_1$ are
+the original annotations from $\aclrd{\textbf{r}}$; the coeffects corresponding to $\Gamma_2$
+are composed with the coeffect of the function value $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}'}$
+as in the original (\emph{app}) rule. Finally, for shared variables, the coeffect is
+obtained by point-wise composition (as in contraction) of the coeffect for the two contexts
+$\cclrd{r_i} \;\cpar\; (\cclrd{t} \,\cseq\, \cclrd{s_j})$. The first coeffect corresponds to the
+context requirements in the sub-expression $e_1$ and the second coeffect corresponds to the
+function argument $e_2$ sequentially composed with the coeffect $\cclrd{t}$ of the function
+(as in ordinary application rule).
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Properties}
+
+The syntax-directed type checking presented in the previous section gives a unique typing derivation
+that can be automatically turned into a valid typing derivation of the original type system
+presented in Figure~\ref{fig:struct-types}. This gives us a unique typing derivation for the
+structural coeffect calculus. As with the unique typing derivation for the flat coeffect system,
+this is used to give semantics of terms of the structural coeffect calculus. We also note that
+a well-typed program in the original type system has a typing derivation in the syntax-driven
+version.
+
+As when discussing uniqueness of typing for flat coeffect systems (Section~\ref{sec:flat-unique}),
+we first give the inversion lemma (Lemma~\ref{thm:struct-invert}) and then prove the
+uniqueness of typing theorem (Theorem~\ref{thm:struct-unique}).
+
+\begin{lemma}[Inversion lemma for syntax-directed structural coeffects]
+\label{thm:struct-invert}
+For the type system defined in Figure~\ref{fig:struct-syntax-types}:
+%
+\begin{enumerate}
+\raggedright
+\item If $\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash x : \tau$ then $\Gamma = x\!:\!\tau$ and $\aclrd{\textbf{c}} = \alift{\cunit}$.
+\item If $\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash n : \tau$ then $\Gamma = ()$ and $\tau = \ident{num}$ and $\aclrd{\textbf{c}} = \alift{}$.
+\item If $\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash e_1~e_2 : \tau_2$
+  then there is some $\Gamma_1, \Gamma_2, \tau_1$ and some $\cclrd{t}, \aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ \\ such that
+  $\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2$ and
+  $\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1$ and also
+  $\coctx{\Gamma}{\aclrd{\textbf{c}}} = \ident{mergevars}(\cclrd{t}, \coctx{\Gamma_1}{\aclrd{\textbf{r}}}, \coctx{\Gamma_2}{\aclrd{\textbf{s}}})$.
+\item If $\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash \lambda x\!:\!\tau_1.e : \tau$
+  then there is some $\Gamma', \tau_2$ and some $\cclrd{s}, \aclrd{\textbf{t}}$ such that
+  $\coctx{\Gamma'}{\aclrd{\textbf{t}}} \vdash e : \tau_2$ and
+  $\tau = \tau_1 \xrightarrow{\cclrd{s}} \tau_2$ and also
+  $(\coctx{\Gamma}{\aclrd{\textbf{c}}}), \tau_1, \cclrd{s} = \ident{findvar}_x(\coctx{\Gamma_1}{\aclrd{\textbf{t}}})$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+Follows from the individual rules given in Figure~\ref{fig:struct-syntax-types}.
+\end{proof}
+
+\begin{theorem}[Uniqueness of syntax-directed structural coeffects]
+\label{thm:struct-unique}
+In the syntax-directed type system for structural coeffects defined in Figure~\ref{fig:struct-syntax-types},
+when $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and
+$\coctx{\Gamma}{\aclrd{\textbf{r}'}} \vdash e : \tau'$ then $\tau = \tau'$ and $\aclrd{\textbf{r}} = \aclrd{\textbf{r}'}$.
 \end{theorem}
 \begin{proof}
-By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
+Suppose that (A) $\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash e : \tau$ and
+(B) $\coctx{\Gamma}{\aclrd{\textbf{c}'}} \vdash e : \tau'$. We show by induction over the typing
+derivation of $\coctx{\Gamma}{\aclrd{\textbf{c}}} \vdash e : \tau$ that $\tau = \tau'$ and $\aclrd{\textbf{c}}=\aclrd{\textbf{c}'}$.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{abs}): $e = \lambda x\!:\!\tau_1.e_1$. Then $\tau = \tau_1 \xrightarrow{\cclrd{c}} \tau_2$
+  for some $\tau_2$ and $\coctx{\Gamma'}{\aclrd{\textbf{t}}} \vdash e : \tau_2$ for some $\Gamma', \aclrd{\textbf{t}}$
+  and also $(\coctx{\Gamma}{\aclrd{\textbf{c}}}), \tau_1, \cclrd{s} = \ident{findvar}_x(\coctx{\Gamma'}{\aclrd{\textbf{t}}})$.
+  By case (4) of Lemma~\ref{thm:struct-invert}, the final rule of the derivation (B) must
+  have also been (\emph{abs}) and this derivation has a sub-derivation with a conclusion
+  ${\coctx{\Gamma}{\aclrd{\textbf{c}'}} \vdash e : \tau_2'}$.
+  By the induction hypothesis $\tau_2 = \tau_2'$ and $\aclrd{\textbf{c}}=\aclrd{\textbf{c}'}$ and therefore
+  also so $\tau = \tau'$. Although $\ident{findvar}_x$ is a relation, it allows only one possible result
+  (because the type of the bound variable matches the type annotation).
+
+\vspace{0.5em}\noindent\hangindent=0.7cm
+Cases (\emph{var}), (\emph{const}) are direct consequence of Lemma~\ref{thm:struct-invert}.
+
+\vspace{0.5em}\noindent\hangindent=0.7cm
+Casea (\emph{app}) similarly to (\emph{abs}).
 \end{proof}
@@
 \noindent
-As in the flat calculus, the primitive operations of the structural indexed comonad are each
-annotated with a unique operations provided by the coeffect annotations. This means that the
-semantics fully determines the structure of the typing rules of the structural coeffect
-calculus.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Examples of structural indexed comonads}
-\label{sec:struct-semantics-examples}
-
-The categorical semantics for structural coeffect calculus is easily instantiated to give
-semantics for a concrete calculus. In this section, we revisit the three examples discussed
-throughout this chapter -- structural liveness, data-flow and bounded variable reuse. 
-Some aspects of the first two examples will be similar to flat versions discussed in
-Section~\ref{sec:flat-semantics} -- they are based on the same data structures (option and a 
-list, respectively), but the data structures are composed differently. Generally speaking,
-rather than having a data structure over a product of variables, we now have 
-a vector of variables over a specific data structure.
-
-The abstract semantics does not specify how vectors of variables should be represented, so this can
-vary in concrete instantiations. In all our examples, we represent a vector of variables as a 
-product written using $\times$. To distinguish between products representing vectors and ordinary
-products (\eg~a product of contexts returned by \ident{split}), we write vectors using 
-$\langle a, \ldots, b \rangle$ rather than the parentheses, used for ordinary tuples.
-
-\paragraph{Data-flow.}
-It is interesting to note that the semantics of data-flow and bounded variable reuse (discussed
-next) both keep a product of multiple values for each variable, so they are both built around an 
-\emph{indexed list} data structure. However, their \ident{cobind} and \ident{dup} operations work 
-differently. We start by looking at the structure modelling data-flow computations. For readability,
-variables in bold face (such as $\mathbf{a_1}$) range over vectors while ordinary notation (such
-as $a_1$) is used for individual values.
-
-\begin{example}[Indexed list for data-flow]
-\label{ex:struct-dfl}
-The indexed list model of data-flow computations is defined over a structural coeffect
-algebra $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
-is indexed by required number of past variables for each individual variable. It is defined
-over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
-containing a current value followed by $n_i$ past values:
-%
-\begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
-\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
-  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{(n_k+1)-\textnormal{times}}
-\end{equation*}
-%
-The mappings that define the structural indexed comonad include the \ident{split} and \ident{merge}
-operations that are shared by the other two examples (discussed below):
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{merge}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
-  (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle) = \\[-0.25em]
-\quad \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle
-\\[0.45em]
-\ident{split}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle} 
-   \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle = \\[-0.25em]
-\quad (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle)
-\end{array}
-\end{equation*}
-%
-The remaining mappings that are required by structural indexed comonad and capture the
-essence of data-flow computations are defined as:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{counit}_{0}~\langle \langle a_0 \rangle \rangle = a_0
-\\[0.45em]
-\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
-  \langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle = \\[-0.25em]
-\quad \langle\langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle, ~\ldots~, \\[-0.25em]
-\quad ~~~\; f \langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle~\; \rangle\rangle
-\\[0.45em]
-\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{\mathit{max}(m,n)} \rangle \rangle =
-   \langle \langle a_1, \ldots, a_m \rangle, \langle a_1, \ldots, a_n \rangle \rangle
-\\[0.45em]
-\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
-  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The definition of the indexed list data structure relies on the fact that the number of annotations
-corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
-containing $n_i+1$ values for the $i$-\textit{th} variable (the annotation represents the number of 
-required \emph{past} values so one more value is required).
-
-The \ident{split} and \ident{merge} operations are defined separately, because they are not specific
-to the example. They operate on the top-level vectors of variables (without looking at the 
-representation of the variable). This means that we can re-use the same definitions for the following
-two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will there represent options
-rather than lists).
-
-The mappings that explain how data-flow computations work are \ident{cobind} (representing sequential
-composition) and \ident{dup} (representing context sharing or parallel composition). In \ident{cobind},
-we get $k$ vectors corresponding to $k$ variables, each with $m+n_i$ values. The operation calls 
-$f$ $m$-times to obtain $m$ past values required as the result of type $\ctyp{\langle m \rangle} \beta$.
-
-The $\ident{dup}_{m,n}$ operation needs to produce a two-variable context containing $m$ and $n$ values,
-respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
-is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable
-and \ident{lift} drops additional past values that are not required.
-
-~
-
-\paragraph{Bounded reuse.}
-As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list 
-structure. Rather than representing possibly different past values that can be shared (see the 
-definition of \ident{dup}), the list now represents multiple copies of the same value as each value
-can only be accessed once. This semantics follows that of Girard \cite{logic-bounded}.
-
-\begin{example}[Indexed list for bounded reuse]
-The indexed list model of bounded variable reuse is defined over a structural coeffect
-algebra $(\mathbb{N}, \ast, +, 1, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$ 
-is a vector containing $n_i$ values of $i$-\textit{th} variable:
-%
-\begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
-\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
-  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{n_k-\textnormal{times}}
-\end{equation*}
-%
-The \ident{merge} and \ident{split} operations are defined as in Example~\ref{ex:struct-dfl}.
-The operations that capture the behaviour of bounded reuse are:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{counit}_{1}~\langle \langle a_0 \rangle \rangle = a_0
-\\[0.45em]
-\ident{lift}_{k', k} \langle\langle a_0, \ldots, a_{k'} \rangle\rangle = 
-  \langle\langle a_0, \ldots, a_k \rangle\rangle \qquad\qquad(\hspace{-0.95em}\textnormal{\footnotesize when}~k\leq k')
-\\[0.45em]
-\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{m+n} \rangle \rangle = 
-\quad \langle \langle a_1, \ldots, a_m \rangle, \langle a_{m+1}, \ldots, a_{m+n} \rangle \rangle
-\\[0.45em]
-\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
-  \langle \langle a_{1,0}, \ldots a_{1,m\ast{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m\ast{n_k}} \rangle \rangle = \\[-0.25em]
-\quad \langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1 - 1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k - 1}} \rangle \rangle, ~\ldots~, \\[-0.25em]
-\quad ~~ f \langle \langle a_{1,(m-1)\ast n_1}, \ldots a_{1,(m-1) \ast n_1} \rangle, \ldots, 
-                   \langle a_{k,m\ast n_k - 1}, \ldots a_{k,m\ast n_k - 1} \rangle \rangle~\; \rangle
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-The \ident{counit} and \ident{lift} operations are defined as previously -- variable access 
-extracts the only value of the only variable and sub-coeffecting allows discarding multiple
-copies of a value that are not needed.
-
-In the bounded variable reuse system, variable sharing is annotated with the $+$ operator (in contrast 
-with \textit{max} used in data-flow). The $\ident{dup}_{m,n}$ operation thus splits the $m+n$ available
-values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The 
-\ident{cobind} operation works similarly -- it splits $m\ast n_i$ available values of each
-variable into $m$ vectors containing $n_i$ copies and then calls the $f$ function $m$-times to 
-obtain $m$ resulting values without sharing any input value.
-
-\paragraph{Liveness.}
-In both data-flow and bounded reuse, the data type is defined as a vector of values obtained
-by applying the indexed list type constructor to types of individual variables. We can
-generalize this pattern. Given a parameterized (indexed) type constructor $D^l \alpha$, we define 
-$\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of a vector of $D^{l_i}$ types.
-For liveness, the definition lets us reuse some of the mappings used when defining the semantics of
-flat liveness. However, we cannot fully define the semantics of the structural version in terms of 
-the flat version -- the \ident{cobind} operation is different and we need an appropriate 
-\ident{dup} operation.
-
-\begin{example}[Structural indexed option]
-Given a structural coeffect algebra formed by $(\{ \ident{L}, \ident{D} \}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$
-and the indexed option data type $D^l$, such that $D^{\ident{D}}\alpha = 1$ and $D^{\ident{L}}\alpha=\alpha$,
-the data type for structural indexed option comonad is:
-%
-\begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} = 
-  D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
-\end{equation*}
-%
-The \ident{merge} and \ident{split} operations are defined as earlier. The remaining operations
-model variable liveness as follows:
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{cobind}_{\ident{L}, \langle l_1, \ldots, l_n \rangle}~f~\langle a_1, \ldots, a_n \rangle = \langle f~\langle a_1, \ldots, a_n \rangle \rangle \\[-0.25em]
-\ident{cobind}_{\ident{D}, \langle \ident{D}, \ldots, \ident{D} \rangle}~f~\langle (), \ldots, () \rangle = \langle \ident{D} \rangle
-\\[0.5em]
-\hspace{-0.5em}
-\begin{array}{lll}
-~\\[-0.85em]
-\ident{dup}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle (), () \rangle \\[-0.25em]
-\ident{dup}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle a, () \rangle \\[-0.25em]
-\ident{dup}_{\ident{D}, \ident{L}} \langle a \rangle &\narrow{=}& \langle (), a \rangle \\[-0.25em]
-\ident{dup}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a, a \rangle 
-\end{array}
-\qquad\quad
-\begin{array}{lll}
-\ident{counit}_{\ident{L}}~\langle a \rangle &\narrow{=}& a
-\\[0.45em]
-\ident{lift}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a \rangle \\[-0.25em]
-\ident{lift}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
-\ident{lift}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle () \rangle \\[-0.25em]
-\end{array}
-\end{array}
-\end{equation*}
-\end{example}
-
-\noindent
-When the expected result of the \ident{cobind} operation is dead (second case), the operation can
-ignore all inputs and directly return the unit value $()$. Otherwise, it passes the vector of 
-input variables to $f$ as-is -- no matter whether the individual values are live or dead. The
-\ident{L} annotation is a unit with respect to $\cap$ and so the annotations expected by $f$ are
-the same as those required by the result of \ident{cobind}.
-
-The \ident{dup} operation resembles with the flat version of \ident{split} -- this is expected as
-duplication in the flat calculus is performed by first duplicating the variable context (using 
-\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair. Depending on the
-required coeffect annotations, this may copy (duplicate) the value, or it may produce an empty context.
-
-Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
-$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models 
-sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
-it behaves as identity.
+As noted earlier, unique typing derivations obtained using the type system given in
+Figure~\ref{fig:struct-syntax-types} can be automatically turned into typing derivations
+of the original (non-syntax-directed) structural coeffect type system in Figure~\ref{fig:struct-types}.
+The following remark provides the details.
+
+\begin{remark}[Admisibility of unique typing for implicit parameters]
+\label{thm:flat-impl-admis}
+  If $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ (using the rules in
+  Figure~\ref{fig:struct-syntax-types}) then there is a uniqye typing derivation
+  using the typing rules from Figure~\ref{fig:struct-types} with a conclusion
+  $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ obtained by induction over
+  the original typing derivation as folows:
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{var}), (\emph{const}): The resulting typing derivation uses the corresponding
+  rule of the non-syntax-directed type system.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{abs}): Take the typing derivation for the sub-expression $e$. If the variable $x$
+  does not appear in $\Gamma_1$, apply (\emph{weak}) followed by (\emph{abs}). Otherwise
+  assume $\Gamma_1 = x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n$ and $x=x_i$. Apply (\emph{exch})
+  repeatedly on variables $x_i, x_{i+1}$ then $x_i, x_{i+2}$ and so on until it is applied
+  on $x_i, x_n$. At this point, $x_i$ is the last variable of the vector and we can apply
+  (\emph{abs}). This produces the same consequent as the one in the original typing derivation.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{app}): Take the typing derivations for the sub-expressions $e_1$ and $e_2$ in
+  free-variable contexts $\Gamma_1$ and $\Gamma_2$. For each variable $x$ that appears in both
+  $\Gamma_1$ and $\Gamma_2$, rename the variable to a fresh name $x'$ in $e_1$ and to another
+  fresh name $x''$ in $e_2$ and their typing derivations. Now we have disjoint contexts and
+  we can apply (\emph{app}) on the target derivations.
+
+\vspace{0.5em}\noindent\hspace{0.6cm}\hangindent=0.6cm
+  Next, apply (\emph{exch}) until $x'$ and $x''$ are last two variables in the vector and apply
+  (\emph{contr}), renaming both $x'$ and $x''$ to the original name $x$. Repeat this step for
+  all variables that were renamed. The resulting variable context is $\Gamma$ and the resulting
+  coeffect annotation is the same as in the original typing derivation.
+\end{remark}
 
@@
 %
-%       ###                  #                    #       #          
-%      #   #                 #                    #                  
-%      #      #   #  # ##   ####    ###    ###   ####    ##     ###  
-%       ###   #   #  ##  #   #         #  #   #   #       #    #   # 
-%          #  #  ##  #   #   #      ####  #       #       #    #     
-%      #   #   ## #  #   #   #  #  #   #  #   #   #  #    #    #   # 
-%       ###       #  #   #    ##    ####   ###     ##    ###    ###  
-%             #   #                                                  
-%              ###                                                   
+%     #####
+%    #     # #   # #    # #####   ##    ####  ##### #  ####
+%    #        # #  ##   #   #    #  #  #    #   #   # #    #
+%     #####    #   # #  #   #   #    # #        #   # #
+%          #   #   #  # #   #   ###### #        #   # #
+%    #     #   #   #   ##   #   #    # #    #   #   # #    #
+%     #####    #   #    #   #   #    #  ####    #   #  ####
 %
@@
 
-Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
-different notion of context. As before, the properties of the flat coeffect algebra guarantee that 
-certain equational properties hold in all instances of the structural coeffect calculus.
-
-In this section, we look at these common properties that we get ``for free'' from the structural 
-coeffect calculus. We start the discussion by briefly considering the key aspects
-that make the equational theory of flat and structural coeffects different.
+The properties of the structural coeffect algebra guarantee that certain equational properties
+hold in all instances of the structural coeffect calculus. In this section, we look at these common
+properties that we get ``for free'' from the structural coeffect calculus. We start the discussion
+by briefly considering the key aspects that make the equational theory of flat and structural
+coeffects different.
 
@@
 When discussing syntactic reductions for the flat calculus in Section~\ref{sec:flat-syntax},
-we noted that no single technique of specifying syntactic reduction works universally for all 
-flat coeffect calculi. We considered multiple alternatives (call-by-name, call-by-value, 
-internalized substitution) that can be used as the basis for operational semantics for
-different calculi that satisfy certain additional properties.
-
-The structural coeffect calculus has more preferable equational properties. In particular, 
-we can prove both $\beta$-reduction and $\eta$-expansion using just the properties of structural 
-coeffect algebra. For this reason, we focus on these two reductions in this section.
-Using the terminology of Pfenning and Davies \cite{logic-modal-reconstruction},
-the structural coeffect calculus satisfies both the \emph{local soundness} and 
-the \emph{local completeness} properties, witnessed by $\beta$-reduction and $\eta$-expansion,
-respectively.
+we noted that call-by-name reduction does not, in general, preserve typing  for all
+flat coeffect calculi. In the structural coeffect calculus, $\beta$-reduction and also $\eta$-expansion
+preserve typing for all instances of the calculus. Using the terminology of Pfenning and Davies
+\cite{logic-modal-reconstruction}, the structural coeffect calculus satisfies both the \emph{local
+soundness} and the \emph{local completeness} properties.
 
@@
 algebra used here are $\cup$, we can place the additional requirement on the outer context.
-Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.
+Note that this is not the \emph{only} possible typing, but it is a \emph{permissible} typing.
 
@@
 The following example shows a situation similar to the previous one. Here, we use
-structural data-flow calculus (writing $\kvd{prev}~e$ to obtain previous value of the 
+structural dataflow calculus (writing $\kvd{prev}~e$ to obtain previous value of the
 expression $e$) and we substitute $w+z$ for $y$:
@@
 
-~
-
-~
-
-~
-
 \begin{lemma}[Multi-nary substitution]
@@
 %
-%     ###                         ##                    #                 
-%    #   #                         #                                      
-%    #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##  
+%    #######
+%       #    #    # ######  ####  #####  #   #
+%       #    #    # #      #    # #    #  # #
+%       #    ###### #####  #    # #    #   #
+%       #    #    # #      #    # #####    #
+%       #    #    # #      #    # #   #    #
+%       #    #    # ######  ####  #    #   #
+%
+% ==================================================================================================
+
+\section{Categorical motivation}
+\label{sec:struct-semantics}
+
+To define the semantics of structural coeffect calculus, we follow the same approach as for flat
+coeffect calculus in Chapter~\ref{ch:semantics}. In this section, we define categorical semantics
+for the calculus in terms of \emph{structural indexed comonad}, which is an extension of the
+\emph{indexed comonad} structure. Similarly to \emph{flat indexed comonad}, the structural variant
+adds operations that are needed to embed full $\lambda$-calculus, this time with per-variable contexts.
+
+We use the semantics to guide the \emph{categorically-inspired translation} discussed in
+Section~\ref{sec:struct-transl}, which translates context-aware programs from the structural coeffect
+calculus to a simple target functional language with uninterpreted comonadically-inspired primitives
+(that correspond to operations of the structural indexed comonads). We then give operational
+semantics for a concrete context-aware language by giving the domain-specific reduction rules for
+the comonadically-inspired primitives. As an example, we use this to prove syntactic type safety
+of structural dataflow language in Section~\ref{sec:struct-transl-safety}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of vectors}
+
+Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing
+judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a
+morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model,
+we can freely transform the value contained in the context modelled using an indexed comonad
+$\ctyp{\cclrd{r}}$.
+
+Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
+which transforms the value inside the context without affecting the coeffect annotation. Thus,
+we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
+$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value
+$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
+
+The ability to freely transform the variable structure is not desirable in the model of
+structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
+the coeffect annotations matches the structure of variables. To achieve this, we model vectors
+using a structure distinct from ordinary products which we denote $-\xtimes-$.
+For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
+is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
+We assume that the operator is equipped with necessary associativity transformations allowing us
+to use it freely on more than two values.
+
+The operator is a bifunctor, but it is \emph{not} a product in the categorical sense. In particular,
+there is no way to turn $\tau_1 \xtimes \ldots \xtimes \tau_n$ into $\tau_i$ (the structure does
+not have projections) and so there is also no way of turning
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$ into
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{\tau_i}$, which would break the correspondence
+between coeffect annotations and variable structure.
+
+The structure created using $-\xtimes-$ can be manipulated only using operations provided by
+the \emph{structural indexed comonad}, which operate over variable contexts contained in an
+indexed comonad $\ctyp{\aclrd{\textbf{r}}}$ and are designed to preserve the correspondence
+between vectors and annotations.
+
+In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
+As mentioned, we assume that the use of the operator can be freely re-associated. For example, when
+calling an operation that requires input of the form
+$(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
+we use an argument $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ and assume that the
+appropriate transformation is inserted.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Indexed comonads, revisited}
+
+The semantics of structural coeffect calculus reuses the definition of \emph{indexed comonad} with
+a minimal change. The additional structure that is required for context manipulation (merging and
+splitting) is different and is here provided by the \emph{structural indexed comonad} structure that
+we introduce in this section.
+
+Recall the definition from Section~\ref{sec:semantics-flat-idx}, which defines an indexed como\-nad
+over a monoid $(\C, \cseq, \cunit)$ as a triple $(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit},
+\ident{cobind}_{\cclrd{r}, \cclrd{s}})$. The triple consists of a family of object mappings $\ctyp{\cclrd{r}{}}$,
+and two mappings that involve context-dependent morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
+
+In the structural coeffect calculus, we work with morphisms of the form $\ctyp{\cclrd{r}} \tau_1 \rightarrow \tau_2$
+representing function values (appearing in the language), but also of the form
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$,
+modelling expressions in a context. To capture this, we need to revisit the definition and use
+\emph{coeffect vectors} in some of the operations.
+
+\begin{definition}
+Given a monoid $(\C, \cseq, \cunit)$ with a pointwise extension of the $\cseq$ operator to a vector
+(written as $\cclrd{t}\,\aseq\,\aclrd{\textbf{s}}$) and an operation lifting scalars to vectors
+$\alift{-}$, an \emph{indexed comonad} over a category $\catc$ is a triple
+$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}})$:
+
+\begin{compactitem}
+\item $\ctyp{\aclrd{\textbf{r}}}{}$ for all $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$ is a family of object mappings
+\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\alift{\cunit}}{\alpha} \rightarrow \alpha$
+\item $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$ is a mapping $(\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta)
+  \rightarrow (\ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{s}}}{\beta})$
+\end{compactitem}
+\end{definition}
+
+\noindent
+The object mapping $\ctyp{\aclrd{\textbf{r}}}$ is now indexed by a vector rather than by a scalar
+$\ctyp{\cclrd{r}}$ as in the previous chapter. This new definition supersedes the old one, because a
+flat coeffect annotation can be seen as singleton vectors.
+
+The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
+return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
+operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
+is explained by the fact that the input function ($\ctyp{\aclrd{\textbf{r}}}{\alpha} \rightarrow \beta$)
+takes a vector of variables, but always produces just a single value. Thus the resulting function
+also takes a vector of variables, but always returns a context with a vector containing just one value.
+In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
+no way of constructing values containing $\xtimes$.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural indexed comonads}
+
+The flat indexed comonad structure extends indexed comonads with operations
+$\ident{merge}_{\cclrd{r},\cclrd{s}}$ and $\ident{split}_{\cclrd{r},\cclrd{s}}$ that combine or
+split the additional (flat) context and are annotated with the flat coeffect operations $\czip$
+and $\cpar$, respectively.
+
+In the structural version, the corresponding operations operate convert between a (wrapped) vector
+of values represented using $\xtimes$ and ordinary pairs of contexts containing parts of the vector.
+The vectors of coeffect annotations are split or merged using $\atimes$ of the structural coeffect
+algebra, in a way that mirrors the wrapped vectors (variable structure).
+
+The following definition includes $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ which models
+duplication of a variable in a context needed for the semantics of contraction:
+
+\begin{definition}
+Given a structural coeffect algebra formed by $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$
+with operations $\alift{-}$ and $\aseq$, a \emph{structural indexed comonad} is an indexed comonad over
+the monoid $(\C, \cseq, \cunit)$ equipped with families of operations $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$,
+$\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ and $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ where:
+%
+\begin{compactitem}
+\item $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
+  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}$
+\item $\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
+  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
+\item $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ is a family of mappings
+  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \xtimes \alpha)}$
+\end{compactitem}
+%
+Here, the following equalities must hold:
+\begin{equation*}
+\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~\equiv~ \ident{id} \qquad\qquad
+\ident{id} ~\equiv~ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}
+\end{equation*}
+
+\end{definition}
+
+\noindent
+These operations differ from those of the flat indexed comonad in that the \ident{merge} and
+\ident{split} operations are required to be inverse functions and to preserve the additional
+information about the context. This was not required for the flat system where the operations
+could under-approximate or over-approximate. Note that the operations use $\xtimes$ to combine
+or split the contained values. This means that they operate on free-variable vectors rather than
+on ordinary products.
+
+The \ident{dup} mapping is a new operation that was not required for a flat calculus. It
+takes a variable context with a single variable annotated with $\cclrd{r}\,\cpar\,\cclrd{s}$,
+duplicates the value of the variable $\alpha$ and splits the additional context between the two
+new variables. In a flat calculus, this operation was expressed using ordinary tuple
+construction, which is not possible here -- the returned context needs to contain a
+two-element vector $\alpha \xtimes \alpha$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[!t]
+The semantics is defined over a typing derivation:
+%
+\begin{equation*}
+\hspace{-3em}
+\begin{array}{ll}
+\hspace{4em}\semdef
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau}
+  {\ident{counit}_{\cunit}}
+& (\emph{var})
+\\[1em]
+\hspace{4.6em}\semdef
+  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
+  {\ident{const}~n}
+& (\emph{num})
+\\[1.5em]
+\hspace{1.40em}\semdeff
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
+  {f}
+  {f \circ \ident{curry}~\ident{merge}_{\aclrd{\textbf{r}}, \cclrd{s}} }
+& (\emph{abs})
+\\[2em]
+\hspace{-1.85em}\emdefff
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
+  {f}
+  {g}
+  {\hspace{-0.5em}\begin{array}{l}
+  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g)\\[-0.3em]
+  ~~\circ~~\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}\\[-0.3em]
+  \end{array}\hspace{-0.5em}}
+& (\emph{app})
+\\[1.5em]
+\hspace{1.55em}\semdeff
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}
+  {f}
+  {f\circ\ident{snd}\circ \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}}
+& (\emph{weak})
+\\[1.5em]
+\hspace{0.5em}\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2\\[-0.25em]\qquad}
+     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {f}
+  {\begin{array}{l}\hspace{-0.5em}f\,\circ\,\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}},
+     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}\\[-0.25em]
+     ~~(\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}\circ\ident{swap}\circ\ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}})\hspace{-5em}~ \end{array} }
+& (\emph{exch})
+\\[1.5em]
+\hspace{0em}\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
+  {f}
+  {\begin{array}{l}\hspace{-0.5em}f\,\circ\,\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}},
+      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}\hspace{-0.5em} \\[-0.25em]~\end{array}}
+& (\emph{contr})
+\end{array}
+\end{equation*}
+
+\vspace{2em}
+Assuming the following auxiliary operations:
+%
+\begin{equation*}
+\begin{array}{rcl}
+  \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f &=&
+    \ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}} \;\circ\;
+      \ident{id} \times(\ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} \,\circ\, f \times \ident{id} \,\circ\, \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}) \;\circ\;
+      \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}} \\[-0.3em]
+  \ident{map}_\cclrd{r}}~f &=& \ident{cobind}_{\cunit, \cclrd{r}}~(f\circ\ident{counit}_\cunit) \\[-0.3em]
+\end{array}
+\end{equation*}
+\begin{equation*}
+\begin{array}{rcl}
+  \ident{id}~x &=& x \\[-0.3em]
+  \ident{const}~v &=& \lambda x.v \\[-0.3em]
+  \ident{curry}~f~x~y &=& \lambda f.\lambda x.\lambda y.f~(x, y)\\[-0.3em]
+  \ident{snd}~(x,y) &=& y\\[-0.3em]
+  \ident{swap}~(x,y) &=& (y,x)\\[-0.3em]
+  f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.3em]
+  \ident{app}~(f, x) &=& f~x
+\end{array}
+\end{equation*}
+
+\figcaption{Categorical semantics of the structural coeffect calculus}
+\label{fig:struct-semantics}
+\end{figure*}
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of structural calculus}
+
+The concrete semantics for liveness and bounded variable use shown in
+Sections~\ref{sec:applications-struct-live} and \ref{sec:applications-struct-bll}
+suggests that semantics of structural coeffect calculi tend to be more complex
+than semantics of flat coeffect calculi. The complexity comes from the fact
+that we need a more expressive representation of the variable context -- \eg~a vector
+of optional values. Additionally, the structural system needs to pass separate variable
+contexts to the sub-expressions.
+
+The latter aspect is fully captured by the semantics shown in this section. The
+earlier point is left to the concrete notion of structural coeffect. Our model still gives
+us the flexibility of defining the concrete representation of variable vectors. We explore
+a number of examples in Section~\ref{sec:struct-semantics-examples} and start by looking
+at a unified categorical semantics defined in terms of \emph{structural indexed comonads}.
+
+\paragraph{Contexts and functions.}
+In the structural coeffect calculus, expressions in context are interpreted as functions taking
+a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
+such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
+structure is annotated with a scalar, such as $\ctyp{\cclrd{r}}$, which we treat as being equivalent
+to a singleton vector annotation $\ctyp{\alift{\cclrd{r}}}$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau}
+  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+Note that the instances of flat indexed comonad ignored the fact that the variable
+context wrapped in the data structure is a product. This is not generally the case for the
+structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
+are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
+rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need to examine the structure
+of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
+
+\paragraph{Expressions.}
+A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. The
+semantics is written as composition of morphisms using a number of auxiliary definitions. Due to
+the eqivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
+as specifying translation to a target functional langauge in Section~\ref{sec:struct-transl}.
+
+The following summarizes how the standard syntax-driven rules work, highlighting the differences
+from the flat version:
+
+\begin{itemize}
+\item When accessing a variable (\emph{var}), the context now contains \emph{only} the accessed
+  variable and so the semantics is just $\ident{counit}_{\cunit}$ without a projection.
+  Constants (\emph{const}) are interpreted by a constant function.
+
+\item The semantics of flat function application first duplicated the context so that the same
+  variables can be passed to both sub-expressions. This is no longer needed -- the
+  (\emph{app}) rule splits the variables \emph{including} the additional context into two parts.
+  Passing the first context to the semantics of $e_1$ gives us a function
+  $\ctyp{\alift{\cclrd{t}}}{\tau_1} \rightarrow \tau_2$.
+
+  A value $\ctyp{\alift{\cclrd{t}}}{\tau_1}$ required in order to call the function is obtained by
+  applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ to the semantics of $e_2$. The result
+  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
+  is then called with the latter part of the split input context.
+
+\item The semantics of function abstraction (\emph{abs}) is syntactically the same as in the
+  flat version -- the only difference is that we now merge a free-variable context with a
+  singleton vector, both at the level of variable assignments and at the level of coeffect
+  annotations.
+\end{itemize}
+
+\noindent
+The semantics for the non-syntax-driven rules (weakening, exchange, contraction) performs
+transformations on the free-variable context. Weakening (\emph{weak}) splits the context and
+ignores the part corresponding to the removed variable. If we were modelling the semantics in
+a language with a linear type system, this would require an additional operation for ignoring an
+unused context annotated with $\aclrd{\czero}$.
+
+The remaining rules perform a transformation anywhere inside the free-variable vector.
+To simplify writing the semantics, we define a helper
+$\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
+that splits the variable vector into three parts, transforms the middle part and then
+merges them, using the newly transformed middle part.
+
+The transformations on the middle part are quite simple. The (\emph{exch}) rule swaps two
+single-variable contexts and the (\emph{contr}) rule uses the $\ident{dup}_{\cclrd{s}, \cclrd{t}}$
+operation to duplicate a variable while splitting its additional context.
+
+\paragraph{Properties.}
+
+As in the flat calculus, the main reason for defining the categorical semantics in this chapter
+is to provide validation for the design of the calculus. The following correspondence theorem
+states that the annotations in the typing rules of the structural coeffect calculus correspond
+to the indices of the semantics. Thus, the calculus captures a context-dependent property if
+it can be modelled by a \emph{structural indexed comonad}. As we show in the next section,
+this is the case for all three discussed examples (liveness, dataflow, bounded variable reuse).
+
+\begin{theorem}[Correspondence]
+In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$
+and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types
+$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$
+and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined
+by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
+\end{theorem}
+\begin{proof}
+By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Examples of structural indexed comonads}
+\label{sec:struct-semantics-examples}
+
+The categorical semantics for structural coeffect calculus is easily instantiated to give
+semantics for a concrete calculus. In this section, we revisit the three examples discussed
+throughout this chapter -- structural liveness, dataflow and bounded variable reuse.
+Some aspects of the first two examples will be similar to flat versions discussed in
+Section~\ref{sec:semantics-flat-monoidal} -- they are based on the same data structures (option and a
+list, respectively), but the data structures are composed differently. Generally speaking,
+rather than having a data structure over a product of variables, we now have
+a vector of variables over a specific data structure.
+
+The abstract semantics does not specify how vectors of variables should be represented, so this can
+vary in concrete instantiations. In all our examples, we represent a vector of variables as a
+product written using $\times$. To distinguish between products representing vectors and ordinary
+products (\eg~a product of contexts returned by \ident{split}), we write vectors using
+$\langle a, \ldots, b \rangle$ rather than the parentheses, used for ordinary tuples.
+
+\paragraph{Dataflow.}
+It is interesting to note that the semantics of dataflow and bounded variable reuse (discussed
+next) both keep a product of multiple values for each variable, so they are both built around an
+\emph{indexed list} data structure. However, their \ident{cobind} and \ident{dup} operations work
+differently. We start by looking at the structure modelling dataflow computations. For readability,
+variables in bold face (such as $\mathbf{a_1}$) range over vectors while ordinary notation (such
+as $a_1$) is used for individual values.
+
+\begin{example}[Indexed list for dataflow]
+\label{thm:struct-dfl}
+The indexed list model of dataflow computations is defined over a structural coeffect
+algebra $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
+is indexed by required number of past variables for each individual variable. It is defined
+over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
+containing a current value followed by $n_i$ past values:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} =
+\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
+  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{(n_k+1)-\textnormal{times}}
+\end{equation*}
+%
+The mappings that define the structural indexed comonad include the \ident{split} and \ident{merge}
+operations that are shared by the other two examples (discussed below):
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{merge}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle}
+  (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle) = \\[-0.25em]
+\quad \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle
+\\[0.45em]
+\ident{split}_{\langle m_1, \ldots, m_k \rangle, \langle n_1, \ldots n_l \rangle}
+   \langle \mathbf{a_1}, \ldots, \mathbf{a_k}, \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle = \\[-0.25em]
+\quad (\langle \mathbf{a_1}, \ldots, \mathbf{a_k} \rangle, \langle \mathbf{b_1}, \ldots, \mathbf{b_l}\rangle)
+\end{array}
+\end{equation*}
+%
+The remaining mappings that are required by structural indexed comonad and capture the
+essence of dataflow computations are defined as:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{0}~\langle \langle a_0 \rangle \rangle = a_0
+\\[0.45em]
+\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
+  \langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle = \\[-0.25em]
+\quad \langle\langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle, ~\ldots~, \\[-0.25em]
+\quad ~~~\; f \langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle~\; \rangle\rangle
+\\[0.45em]
+\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{\mathit{max}(m,n)} \rangle \rangle =
+   \langle \langle a_1, \ldots, a_m \rangle, \langle a_1, \ldots, a_n \rangle \rangle
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The definition of the indexed list data structure relies on the fact that the number of annotations
+corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
+containing $n_i+1$ values for the $i$-\textit{th} variable (the annotation represents the number of
+required \emph{past} values so one more value is required).
+
+The \ident{split} and \ident{merge} operations are defined separately, because they are not specific
+to the example. They operate on the top-level vectors of variables (without looking at the
+representation of the variable). This means that we can re-use the same definitions for the following
+two examples (with the only difference that $\mathbf{a_i}, \mathbf{b_i}$ will there represent options
+rather than lists).
+
+The mappings that explain how dataflow computations work are \ident{cobind} (representing sequential
+composition) and \ident{dup} (representing context sharing or parallel composition). In \ident{cobind},
+we get $k$ vectors corresponding to $k$ variables, each with $m+n_i$ values. The operation calls
+$f$ $m$-times to obtain $m$ past values required as the result of type $\ctyp{\langle m \rangle} \beta$.
+
+The $\ident{dup}_{m,n}$ operation needs to produce a two-variable context containing $m$ and $n$ values,
+respectively, of the input variable. The input provides $\mathit{max}(m, n)$ values, so the definition
+is simply a matter of restriction. Finally, \ident{counit} extracts the value of its single variable.
+
+\paragraph{Bounded reuse.}
+As mentioned earlier, the semantics of calculus for bounded reuse is also based on the indexed list
+structure. Rather than representing possibly different past values that can be shared (see the
+definition of \ident{dup}), the list now represents multiple copies of the same value as each value
+can only be accessed once. This semantics follows that of Girard \cite{logic-bounded}.
+
+\begin{example}[Indexed list for bounded reuse]
+\label{thm:struct-bll}
+The indexed list model of bounded variable reuse is defined over a structural coeffect
+algebra $(\mathbb{N}, \ast, +, 1, 0, \leq)$. The data type $\ctyp{\langle n_1, \ldots, n_k\rangle}$
+is a vector containing $n_i$ values of $i$-\textit{th} variable:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} =
+\underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
+  \underbrace{(\alpha_k \times \ldots \times \alpha_k)}_{n_k-\textnormal{times}}
+\end{equation*}
+%
+The \ident{merge} and \ident{split} operations are defined as in Example~\ref{ex:struct-dfl}.
+The operations that capture the behaviour of bounded reuse are:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{counit}_{1}~\langle \langle a_0 \rangle \rangle = a_0
+\\[0.45em]
+\ident{dup}_{m, n} \langle \langle a_1, \ldots, a_{m+n} \rangle \rangle =
+\quad \langle \langle a_1, \ldots, a_m \rangle, \langle a_{m+1}, \ldots, a_{m+n} \rangle \rangle
+\\[0.45em]
+\ident{cobind}_{m, \langle n_1, \ldots, n_k \rangle}~f~
+  \langle \langle a_{1,0}, \ldots a_{1,m\ast{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m\ast{n_k}} \rangle \rangle = \\[-0.25em]
+\quad \langle\; f \langle \langle a_{1,0}, \ldots a_{1,{n_1 - 1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k - 1}} \rangle \rangle, ~\ldots~, \\[-0.25em]
+\quad ~~ f \langle \langle a_{1,(m-1)\ast n_1}, \ldots a_{1,(m-1) \ast n_1} \rangle, \ldots,
+                   \langle a_{k,m\ast n_k - 1}, \ldots a_{k,m\ast n_k - 1} \rangle \rangle~\; \rangle
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+The \ident{counit} operation is defined as previously -- it extracts the only value of the only
+variable In the bounded variable reuse system, variable sharing is annotated with the $+$ operator
+(in contrast with \textit{max} used in dataflow). The $\ident{dup}_{m,n}$ operation thus splits the $m+n$
+available values between two vectors of length $m$ and $n$, without \emph{sharing} a value. The
+\ident{cobind} operation works similarly -- it splits $m\ast n_i$ available values of each
+variable into $m$ vectors containing $n_i$ copies and then calls the $f$ function $m$-times to
+obtain $m$ resulting values without sharing any input value.
+
+\paragraph{Liveness.}
+In both dataflow and bounded reuse, the data type is defined as a vector of values obtained
+by applying the indexed list type constructor to types of individual variables. We can
+generalize this pattern. Given a parameterized (indexed) type constructor $D^l \alpha$, we define
+$\ctyp{\langle l_1, \ldots, l_n \rangle}$ in terms of a vector of $D^{l_i}$ types.
+For liveness, the definition lets us reuse some of the mappings used when defining the semantics of
+flat liveness. However, we cannot fully define the semantics of the structural version in terms of
+the flat version -- the \ident{cobind} operation is different and we need an appropriate
+\ident{dup} operation.
+
+\begin{example}[Structural indexed option]
+\label{thm:struct-live}
+Given a structural coeffect algebra formed by $(\{ \ident{L}, \ident{D} \}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$
+and the indexed option data type $D^l$, such that $D^{\ident{D}}\alpha = 1$ and $D^{\ident{L}}\alpha=\alpha$,
+the data type for structural indexed option comonad is:
+%
+\begin{equation*}
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} =
+  D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
+\end{equation*}
+%
+The \ident{merge} and \ident{split} operations are defined as earlier. The remaining operations
+model variable liveness as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{cobind}_{\ident{L}, \langle l_1, \ldots, l_n \rangle}~f~\langle a_1, \ldots, a_n \rangle = \langle f~\langle a_1, \ldots, a_n \rangle \rangle \\[-0.25em]
+\ident{cobind}_{\ident{D}, \langle \ident{D}, \ldots, \ident{D} \rangle}~f~\langle (), \ldots, () \rangle = \langle \ident{D} \rangle
+\\[0.5em]
+\hspace{-0.5em}
+\begin{array}{lll}
+~\\[-0.85em]
+\ident{dup}_{\ident{D}, \ident{D}} \langle () \rangle &\narrow{=}& \langle (), () \rangle \\[-0.25em]
+\ident{dup}_{\ident{L}, \ident{D}} \langle a \rangle &\narrow{=}& \langle a, () \rangle \\[-0.25em]
+\ident{dup}_{\ident{D}, \ident{L}} \langle a \rangle &\narrow{=}& \langle (), a \rangle \\[-0.25em]
+\ident{dup}_{\ident{L}, \ident{L}} \langle a \rangle &\narrow{=}& \langle a, a \rangle
+\end{array}
+\qquad\quad
+\begin{array}{lll}
+\ident{counit}_{\ident{L}}~\langle a \rangle &\narrow{=}& a
+\end{array}
+\end{array}
+\end{equation*}
+\end{example}
+
+\noindent
+When the expected result of the \ident{cobind} operation is dead (second case), the operation can
+ignore all inputs and directly return the unit value $()$. Otherwise, it passes the vector of
+input variables to $f$ as-is -- no matter whether the individual values are live or dead. The
+\ident{L} annotation is a unit with respect to $\cap$ and so the annotations expected by $f$ are
+the same as those required by the result of \ident{cobind}.
+
+The \ident{dup} operation resembles with the flat version of \ident{split} -- this is expected as
+duplication in the flat calculus is performed by first duplicating the variable context (using
+\ident{map}) and then applying \ident{split}. Here, the duplication returns a pair. Depending on the
+required coeffect annotations, this may copy (duplicate) the value, or it may produce an empty context.
+
+Finally, \ident{counit} extracts a value which is always present as guaranteed by the type
+$\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models
+sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
+it behaves as identity.
+
+\paragraph{Properties.}
+The concrete categortical semantics presented in this section is a generalization of the concrete
+semantics given when introducing context-aware programming languages in Chapter~\ref{ch:applications}.
+
+\begin{theorem}[Generalization]
+\label{thm:struct-generalization}
+
+Consider a typing derivation obtained according to the rules for finding unique typing derivations
+as specified in Section~\ref{sec:struct-unique} for a coeffect language with liveness, dataflow or
+bounded variable use.
+
+The semantics obtained by instantiating the rules in Figure~\ref{fig:struct-semantics} with the
+concrete operations defined in Example~\ref{thm:struct-live},
+Example~\ref{thm:struct-dfl} or Example~\ref{thm:struct-bll} is the
+same as the one defined in Figure~\ref{fig:applications-struct-livesem},
+Section~\ref{sec:applications-structural-dataflow} and Section~\ref{sec:applications-struct-bll},
+respectively.
+\end{theorem}
+\begin{proof}
+Simple expansion of the definitions for the unique typing derivation.
+\end{proof}
+
+% ==================================================================================================
+%
+%    #######
+%       #    #####    ##   #    #  ####  #        ##   ##### #  ####  #    #
+%       #    #    #  #  #  ##   # #      #       #  #    #   # #    # ##   #
+%       #    #    # #    # # #  #  ####  #      #    #   #   # #    # # #  #
+%       #    #####  ###### #  # #      # #      ######   #   # #    # #  # #
+%       #    #   #  #    # #   ## #    # #      #    #   #   # #    # #   ##
+%       #    #    # #    # #    #  ####  ###### #    #   #   #  ####  #    #
+%
+% ==================================================================================================
+
+\section{Translation}
+\label{sec:struct-transl}
+
+\subsection{Safety of concrete thing}
+\label{sec:struct-transl-safety}
+
+% ==================================================================================================
+%
+%     #####
+%    #     #  ####  #    #  ####  #      #    #  ####  #  ####  #    #
 %    #       #    # ##   # #    # #      #    # #      # #    # ##   #
-%    #      #   #  #   #  #        #    #   #   ###     #    #   #  #   # 
-%    #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   # 
-%     ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   # 
+%    #       #    # # #  # #      #      #    #  ####  # #    # # #  #
+%    #       #    # #  # # #      #      #    #      # # #    # #  # #
+%    #     # #    # #   ## #    # #      #    # #    # # #    # #   ##
+%     #####   ####  #    #  ####  ######  ####   ####  #  ####  #    #
 %
@@
 This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
-contextual properties. The system provides a precise analysis of liveness and data-flow and allows 
+contextual properties. The system provides a precise analysis of liveness and dataflow and allows
 other interesting uses such as tracking of variable accesses based on bounded linear logic.
@@
 calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
-it unifies our three examples (bounded reuse, data-flow and liveness) and it demonstrates
+it unifies our three examples (bounded reuse, dataflow and liveness) and it demonstrates
 suitability of the type system. The second technical result is equational theory for the structural

### 2016-05-02 text/Chapter05.tex
Unique structural derivation and edits (Chapter 6)
Translational semantics for structural coeffects (Chapter 6)

@@
 %    #######
-%      #       #    # #####  ###### #####  #####  # #    #  ####
-%      #       ##  ## #    # #      #    # #    # # ##   # #    #
-%      #####   # ## # #####  #####  #    # #    # # # #  # #
-%      #       #    # #    # #      #    # #    # # #  # # #  ###
-%      #       #    # #    # #      #    # #    # # #   ## #    #
-%      ####### #    # #####  ###### #####  #####  # #    #  ####
+%       #    #####    ##   #    #  ####  #        ##   ##### #  ####  #    #
+%       #    #    #  #  #  ##   # #      #       #  #    #   # #    # ##   #
+%       #    #    # #    # # #  #  ####  #      #    #   #   # #    # # #  #
+%       #    #####  ###### #  # #      # #      ######   #   # #    # #  # #
+%       #    #   #  #    # #   ## #    # #      #    #   #   # #    # #   ##
+%       #    #    # #    # #    #  ####  ###### #    #   #   #  ####  #    #
 %
@@
 \begin{figure}[t]
-\paragraph{Language syntax}
+\paragraph{Language syntax.}
 Given $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$, extend the programming language
@@
   e &  =  &\ldots \sep\ident{cobind}_{\cclrd{s},\cclrd{r}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~e
-         \sep \ident{split}_{\cclrd{r},\cclrd{s}}~e \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~e \\
+         \sep \ident{split}_{\cclrd{r},\cclrd{s}}~e \\
   \tau &  = &\ldots \sep \ctyp{\cclrd{r}}{\tau} \\
@@
     && \hspace{1.4em} \sep \ident{merge}_{\cclrd{r},\cclrd{s}}~\_
-           \sep \ident{split}_{\cclrd{r},\cclrd{s}}~\_ \sep \ident{lift}_{\cclrd{r},\cclrd{r'}}~\_ \\
+           \sep \ident{split}_{\cclrd{r},\cclrd{s}}~\_  \\
 \end{array}
@@
 
-\paragraph{Typing rules}
+\paragraph{Typing rules.}
 Given $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$, add the typing rules:
@@
 %     #####
-%    #     # #    # #    # #    #   ##   #####  #   #
-%    #       #    # ##  ## ##  ##  #  #  #    #  # #
-%     #####  #    # # ## # # ## # #    # #    #   #
-%          # #    # #    # #    # ###### #####    #
-%    #     # #    # #    # #    # #    # #   #    #
-%     #####   ####  #    # #    # #    # #    #   #
+%    #     #  ####  #    #  ####  #      #    #  ####  #  ####  #    #  ####
+%    #       #    # ##   # #    # #      #    # #      # #    # ##   # #
+%    #       #    # # #  # #      #      #    #  ####  # #    # # #  #  ####
+%    #       #    # #  # # #      #      #    #      # # #    # #  # #      #
+%    #     # #    # #   ## #    # #      #    # #    # # #    # #   ## #    #
+%     #####   ####  #    #  ####  ######  ####   ####  #  ####  #    #  ####
 %
@@
 
-\section{SUmmary}
+\section{Conclusions}
 

### 2016-05-02 text/Chapter06.tex
Translational semantics for structural coeffects (Chapter 6)

@@
 \subsection{Structural indexed comonads}
+\label{sec:struct-semantics-idxc}
 
@@
   {f}
-  {f \circ \ident{curry}~\ident{merge}_{\aclrd{\textbf{r}}, \cclrd{s}} }
+  {f \circ \ident{curry}~\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}} }
 & (\emph{abs})
@@
   {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.25em]~\end{array}}
   {f}
@@
   {\hspace{-0.5em}\begin{array}{l}
-  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g)\\[-0.3em]
-  ~~\circ~~\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}\\[-0.3em]
+  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g)\\[-0.25em]
+  ~~\circ~~\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}\\[-0.25em]
   \end{array}\hspace{-0.5em}}
@@
       \ident{id} \times(\ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} \,\circ\, f \times \ident{id} \,\circ\, \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}) \;\circ\;
-      \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}} \\[-0.3em]
-  \ident{map}_\cclrd{r}}~f &=& \ident{cobind}_{\cunit, \cclrd{r}}~(f\circ\ident{counit}_\cunit) \\[-0.3em]
+      \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}
 \end{array}
@@
 \begin{array}{rcl}
-  \ident{id}~x &=& x \\[-0.3em]
-  \ident{const}~v &=& \lambda x.v \\[-0.3em]
-  \ident{curry}~f~x~y &=& \lambda f.\lambda x.\lambda y.f~(x, y)\\[-0.3em]
-  \ident{snd}~(x,y) &=& y\\[-0.3em]
-  \ident{swap}~(x,y) &=& (y,x)\\[-0.3em]
-  f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.3em]
+  \ident{id}~x &=& x \\[-0.25em]
+  \ident{const}~v &=& \lambda x.v \\[-0.25em]
+  \ident{curry}~f~x~y &=& \lambda f.\lambda x.\lambda y.f~(x, y)\\[-0.25em]
+  \ident{fst}~(x,y) &=& x\\[-0.25em]
+  \ident{swap}~(x,y) &=& (y,x)\\[-0.25em]
+  f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.25em]
   \ident{app}~(f, x) &=& f~x
@@
 \end{figure*}
+
 % --------------------------------------------------------------------------------------------------
@@
 \begin{theorem}[Correspondence]
+\label{thm:struct-correspondence}
 In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$
@@
 
-\section{Translation}
+\section{Translational semantics}
 \label{sec:struct-transl}
+
+In the previous section, we used category theory to give a unified model capable of capturing
+the semantics of our three context-aware language based on the structural coeffect calculus.
+Although the categorical model is interesting on its own, we use it in the same way as in
+Chapter~\ref{ch:semantics} -- to define a translation from source context-aware languages to a
+simple target functional language. As for flat coeffect calculus, we show that the translation
+produces well-typed programs in the target language. For a sample context-aware programming
+language, we then show that well-typed programs (produced by the translation) do not get stuck.
+
+This section mirrors the development presented in Chapter~\ref{ch:semantics} for flat coeffect
+calculus. We extend the simple target language with additional constructs inspired by structural
+indexed comonads (Section~\ref{sec:struct-transl-ext}), define the translation to the target
+langauge (Section~\ref{sec:struct-transl-transl}) and prove safety of one sample language
+(Section~\ref{sec:struct-transl-safety}) -- we choose structural dataflow to allow easy
+comparison with the flat system.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\paragraph{Language syntax.}
+Given a structural coeffect algebra, extend the programming language syntax with the following
+constructs:
+
+\begin{equation*}
+\begin{array}{rcl}
+  e &  =  &\ldots \sep\ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e
+         \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~e \\
+  \tau &  = &\ldots \sep \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \\
+  K  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~\_~e \sep \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
+    && \hspace{1.4em} \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~\_
+           \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~\_ \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~\_ \\
+\end{array}
+\end{equation*}
+
+~
+
+~
+
+\paragraph{Typing rules.}
+Given a structural coeffect algebra, add the typing rules:
+
+\begin{equation*}
+\begin{array}{l}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\alift{\cunit}}{\tau}}
+  {\Gamma \vdash \ident{counit}_{\cunit}~e : \tau} \\
+\\
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
+  {\Gamma \vdash \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
+  \\
+\tyrule{merge}
+  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)} }
+  {\Gamma \vdash  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }\\
+\\
+\tyrule{split}
+  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
+  {\Gamma \vdash  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)}} \\
+\\
+\tyrule{dup}
+  {\Gamma \vdash  e : \ctyp{\alift{\cclrd{r}\cpar\cclrd{s}}}{\tau} }
+  {\Gamma \vdash  \ident{dup}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\alift{\cclrd{r},\cclrd{s}}}{(\tau \xtimes \tau)} } \\
+\end{array}
+\end{equation*}
+
+\caption{Comonadically-inspired extensions for structural coeffects}
+\label{fig:struct-semantics-ext}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Comonadically-inspired language extensions}
+\label{sec:struct-transl-ext}
+
+In Section~\ref{sec:semantics-translation-target}, we defined the syntax, typing rules and
+operational semantics for a simple functional programming language. We then extended it with
+uninterpreted constructs inspired by the \emph{flat indexed comonad} structure and used it as the
+translation target for the flat coeffect calculus. In this section, we take the same core language
+and extend it with constructs inspired by the \emph{structural indexed comonad}.
+
+Given a coeffect language with a structural coeffect algebra formed by
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq, \atimes$, we
+extend the core functional language with operations shown in Figure~\ref{fig:struct-semantics-ext}.
+The syntax extensions add comonadically-inspired operations that mirror those defined in
+Section~\ref{sec:struct-semantics-idxc}. The typing for the operations corresponds to their
+categorical counterparts.
+
+We also include an uninterpreted type $\ctyp{\aclrd{\textbf{r}}}{\tau_1\xtimes\ldots\xtimes\tau_k}$,
+which models a contextual (comonadic) value indexed by a vector of annotations. As in the
+categorical model for structural coeffects, context consisting of multiple variables is not modelled
+as ordinary tuple -- it can only be manipulated by the comonadically-inspired operations.
+In the target language, this is done by defining the $\ctyp{\aclrd{\textbf{r}}}{}$ type over
+zero or more underlying types. Here, our syntactic treatment differs slightly from the categorical
+model where objects created by $\xtimes$ were first-class values. In Figure~\ref{fig:struct-semantics-ext},
+we need to explicitly specify the types of individual components in typing rules for
+(\emph{merge}), (\emph{split}) and (\emph{cobind}).
+
+As with flat coeffects, the extensions described here are common for all concrete instances of
+structural context-aware languages. For each concrete language, we need to provide values of type
+$\ctyp{\aclrd{\textbf{r}}}{\tau}$ and reduction rules for comonadically-inspired operations.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[!t]
+The translation is defined over a typing derivation:
+%
+\begin{equation*}
+\begin{array}{ll}
+\hspace{3.25em}\semdef
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau}
+  {\lambda\ctx.\ident{counit}_{\cunit}~\ctx}
+& (\emph{var})
+\\[1em]
+\hspace{4em}\semdef
+  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
+  {\lambda\ctx.n}
+& (\emph{num})
+\\[1.5em]
+\hspace{0.8em}\semdeff
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
+  {f}
+  {\lambda\ctx.\lambda v.f~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v)) }
+& (\emph{abs})
+\\[2em]
+\hspace{-2.5em}\emdefff
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.25em]~\\[-0.25em]~\end{array}}
+  {f}
+  {g}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda\ctx.~~\\[-0.25em]
+    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}~\ctx\\[-0.25em]
+    \quad f~\ctx_1~(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g~\ctx_2)
+  \end{array} }
+& (\emph{app})
+\\[1.5em]
+\hspace{0em}\emdeff
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\begin{array}{l}\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}\\[-0.25em]~\\[-0.25em]~\\\end{array}}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda\ctx.~~\\[-0.25em]
+    \quad \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx\\[-0.25em]
+    \quad f~\ctx_1
+  \end{array} }
+& (\emph{weak})
+\\[1.5em]
+\hspace{0em}\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2\\[-0.25em]\qquad}
+     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]~\end{array}}
+  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
+  {\begin{array}{l}\hspace{-0.5em}
+    \lambda\ctx.~f (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}},\alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~(\lambda\ctx'.\\[-0.25em]
+     \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}} \\[-0.25em]
+     \quad \ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)~))\end{array} }
+& (\emph{exch})
+\\[1.5em]
+\hspace{-0.5em}\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
+  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
+  {\begin{array}{l}\hspace{-0.5em}\lambda\ctx.f~(\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}},
+      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}\\[-0.25em]
+      \qquad \ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)\hspace{-0.5em}\end{array}}
+& (\emph{contr})
+\end{array}
+\end{equation*}
+
+\vspace{2em}
+Assuming the following auxiliary definition:
+%
+\begin{equation*}
+\begin{array}{l}
+  \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} = \lambda f.\lambda \ctx.\\[-0.25em]
+  \qquad \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx\\[-0.25em]
+  \qquad \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}},\aclrd{\textbf{t}}}~\ctx'\\[-0.25em]
+  \qquad \ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~(\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
+\end{array}
+\end{equation*}
+
+\figcaption{Trnaslation from a structural coeffect calculus}
+\label{fig:struct-translation}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Comonadically-inspired translation}
+\label{sec:struct-transl-transl}
+
+When translating context-aware programs to the functional language, variable contexts become
+values of comonadically-inspired data types containing a value for each variable in the context.
+Function inputs become comonadically-inspired values containing exactly one variable. More formally:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}} &=&
+  \ctyp{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}{(\sem{\tau_1} \xtimes \ldots \xtimes \sem{\tau_n})} \\[0.5em]
+\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} & = & \ctyp{\alift{\cclrd{r}}}{\sem{\tau_1}} \rightarrow \sem{\tau_2} \\[-0.2em]
+\sem{\ident{num}} & = & \ident{num} \\
+\end{array}
+\end{equation*}
+
+\noindent
+The definition differs from the one for flat coeffects (Section~\ref{sec:semantics-translation-transl})
+in that the comonadically-inspired data type takes multiple type parameters (separated using $\xtimes$),
+rather than wrapping a regular tuple in the target language.
+
+The translation rules are defined in Figure~\ref{fig:struct-translation}. As in the case of the flat
+coeffect calculus, the definition directly follows the categorical semantics shown in
+Figure~\ref{fig:struct-translation}. We expand the definitions so that the result is a valid
+program in the target language rather than a composition of morphisms.
+
+As with flat coeffects, the correspondence property of the semantics (Theorem~\ref{thm:struct-correspondence})
+can now be adapted into well-typedness of the translation. Given a well-typed program in the
+structural coeffect calculus, the translation produces a well-typed program in the target
+language. This is true for all context-aware languages based on the structural coeffect calculus
+and it provides us with the first part of type safety theorem. The second part is type safety
+of the target language with concrete domain-specific extensions as discussed in
+Section~\ref{sec:struct-transl-safety}.
+
+\begin{theorem}[Well-typedness of the translation]
+\label{thm:semantics-welltyped}
+Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
+written in a structural context-aware programming langugae that is translated to the target language as
+(we write $\ldots$ for the omitted part of the translation tree):
+%
+\begin{equation*}
+  \semdeff
+    {~~(\ldots)~~}
+    {\coctx{}{\alift{}} \vdash e : \tau }
+    {(\ldots)}
+    { ~~f }
+\end{equation*}
+%
+Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
+\end{theorem}
+\begin{proof}
+By rule induction over the derivation of the translation. Given a judgement
+$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\aclrd{\textbf{c}}} \vdash e : \tau$ where
+$\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
+constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{var}): $\aclrd{\textbf{c}}=\alift{\cunit}$ and so ${\ident{counit}_\cunit}~\ctx$ is well-typed.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{num}): $\tau = \ident{num}$ and so the body $n$ is well-typed.
+
+%
+%
+%
+%
+TODO
+%
+%
+%
+%
+%
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\cclrd{r}}(\ldots)$ and the type of $v$ is $\ctyp{\cclrd{s}}\tau_1$,
+  calling $\ident{merge}_{\cclrd{r},\cclrd{s}}$ and reassociating produces
+  $\ctyp{\cclrd{r}\czip\cclrd{s}}(\ldots)$ as expected by $f$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{app}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of
+  $\ctx_1, \ctx_2$ are $\ctyp{\cclrd{r}}(\ldots)$ and $\ctyp{\cclrd{s}\cseq\cclrd{t}}(\ldots)$, respectively.
+  $g$ requires $\ctyp{\cclrd{s}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{s},\cclrd{t}}$ is
+  $\ctyp{\cclrd{t}}\tau_1$ as required by $f$.
+\end{proof}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 \label{sec:struct-transl-safety}
+
+\newpage
 

### 2016-05-04 text/Chapter05.tex
Add structural coeffect proofs (Chapter 6)

@@
 and a matching kind of expressions. We specify how the \kvd{prev} keyword is translated into a
-$\ident{prev}_\cclrd{r}$ operation of the target language nd we also add a typing rule
+$\ident{prev}_\cclrd{r}$ operation of the target language and we also add a typing rule
 (\emph{df}) that checks the types of the elements of the stream and also guarantees
@@
 \semdeff
-  { \coctx{\Gamma}{\cclrd{n}} \vdash e : \tau }
-  { \coctx{\Gamma}{\cclrd{n+1}} \vdash \kvd{prev}~e : \tau }
+  { \coctx{\Gamma}{\cclrd{n+1}} \vdash e : \tau }
+  { \coctx{\Gamma}{\cclrd{n}} \vdash \kvd{prev}~e : \tau }
   {f}

### 2016-05-04 text/Chapter06.tex
Add structural coeffect proofs (Chapter 6)

@@
   {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
-  {\Gamma \vdash \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
+  {\Gamma \vdash \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
   \\
@@
 \begin{theorem}[Well-typedness of the translation]
-\label{thm:semantics-welltyped}
+\label{thm:struct-welltyped}
 Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
@@
 
-%
-%
-%
-%
-TODO
-%
-%
-%
-%
-%
-
 \vspace{0.25em}\noindent\hangindent=0.6cm
-Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\cclrd{r}}(\ldots)$ and the type of $v$ is $\ctyp{\cclrd{s}}\tau_1$,
-  calling $\ident{merge}_{\cclrd{r},\cclrd{s}}$ and reassociating produces
-  $\ctyp{\cclrd{r}\czip\cclrd{s}}(\ldots)$ as expected by $f$.
+Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\aclrd{\textbf{r}}}(\ldots)$ and the type of $v$ is $\ctyp{\alift{\cclrd{s}}}\tau_1$,
+  calling $\ident{merge}_{\aclrd{\textbf{r}},\alift{\cclrd{s}}}$ produces a context of type
+  $\ctyp{\aclrd{\textbf{r}}\atimes\alift{\cclrd{s}}}(\ldots\xtimes\tau_1)$ as expected by $f$.
 
 \vspace{0.25em}\noindent\hangindent=0.6cm
-Case (\emph{app}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of
-  $\ctx_1, \ctx_2$ are $\ctyp{\cclrd{r}}(\ldots)$ and $\ctyp{\cclrd{s}\cseq\cclrd{t}}(\ldots)$, respectively.
-  $g$ requires $\ctyp{\cclrd{s}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{s},\cclrd{t}}$ is
-  $\ctyp{\cclrd{t}}\tau_1$ as required by $f$.
+Case (\emph{app}): After applying $\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t}\aseq\aclrd{\textbf{s}}}$ the types of
+  $\ctx_1, \ctx_2$ are $\ctyp{\aclrd{\textbf{r}}}(\ldots)$ and $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}(\ldots)$, respectively.
+  $g$ requires $\ctyp{\aclrd{\textbf{s}}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{t},\aclrd{\textbf{s}}}$ is
+  $\ctyp{\alift{\cclrd{t}}}\tau_1$ as required by $f$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{weak}): After applying $\ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}$ the type of
+  $\ctx_1$ is $\ctyp{\aclrd{\textbf{r}}}(\ldots)$ as required.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{exch}), (\emph{contr}): The auxiliary definition $\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
+  keeps parts of the context corresponding to coeffect annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{q}}$ unchanged and transforms
+  the nested part. In (\emph{exch}), the provided lambda function is of type $\ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau_1\xtimes\tau_2)}
+  \rightarrow \ctyp{\alift{\cclrd{t},\cclrd{s}}}{(\tau_2\xtimes\tau_1)}$. In (\emph{contr}), the type of
+  $\ident{dup}_{\cclrd{s},\cclrd{t}}$ is $\ctyp{\alift{\cclrd{s}\cpar\cclrd{t}}}{\tau} \rightarrow \ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau\xtimes\tau)}$
+  (assuming \ident{nest} is expanded, rather than treated as a value within the language).
 \end{proof}
@@
 
-\subsection{Safety of concrete thing}
+\subsection{Structural coeffect language for dataflow}
 \label{sec:struct-transl-safety}
 
-\newpage
+The target language with comonadically-inspired primitives provides a framework that can be used
+to model a variety of structural context-aware langauges and prove their safety. As outlined in
+Section~\ref{sec:semantics-generalising}, the key principle that guarantees the safety of the
+target language is a correspondence between coeffect annotations and the values they represent.
+In a more expressive target language (such as Haskell or Agda), it would be sufficient to
+provide an implementation of the comonadically-inspired primitives for the concrete domain-specific
+language.
+
+Our simple target language is not expressive enough to capture the correspondence in the type
+system and so we instead follow the same methodology as when discussing safety for flat coeffect
+languages in section Section~\ref{sec:semantics-proofs} and show safety for a sample concrete
+context-aware language. We consider the structural coeffect language for dataflow. The definitions
+can be compared with the flat version discussed in Section~\ref{sec:semantics-proofs-df}, which
+highlights the similarities and differences between the flat and structural notion of context.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-0.5em}
+\paragraph{Language syntax}
+\begin{equation*}
+\begin{array}{rcl}
+v &=& \ldots \sep \ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle
+    \hspace{9.5em} \textbf{v} ~=~ \langle v_0, \ldots, v_k \rangle \\
+e &=& \ldots \sep \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle \sep \ident{prev}_{\cclrd{n_0}, \ldots, \cclrd{n_k}}~e
+    \quad\qquad \textbf{e} ~=~ \langle e_0, \ldots, e_k \rangle \\
+K &=& \ldots \sep \ident{Df}\langle \textbf{v}_0, \ldots, \langle v_{j,0}, \ldots, v_{j, i-1}, \_, e_{j, i+1} \ldots, e_{j, n}\rangle, \ldots, \textbf{e}_k\rangle \\
+   && \ldots \sep \ident{prev}_{\cclrd{n_0}, \ldots, \cclrd{n_k}}~\_
+\end{array}
+\end{equation*}
+
+\vspace{0.75em}
+\paragraph{Typing rules}
+
+\begin{equation*}
+\tyrule{vec}
+  { \forall i\in\{0\ldots n\}. ~ \Gamma \vdash e_i : \tau}
+  { \Gamma \vdash_{\ident{vec}} \langle e_0, \ldots, e_n \rangle : \tau, \cclrd{n} }
+\end{equation*}
+\begin{equation*}
+\tyrule{df}
+  { \forall i\in\{0\ldots k\}. ~ \Gamma \vdash_{\ident{vec}} \textbf{e}_i : \tau_i, \cclrd{n_i}}
+  { \Gamma \vdash \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle : \ctyp{\alift{\cclrd{n_0}, \ldots, \cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
+\end{equation*}
+\begin{equation*}
+\tyrule{prev}
+  { \Gamma \vdash e : \ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
+  { \Gamma \vdash \ident{prev}_{\cclrd{n_0},\ldots,\cclrd{n_k}}~e : \ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
+\end{equation*}
+
+\vspace{1em}
+\paragraph{Translation}
+
+\begin{equation*}
+\semdeff
+  { \coctx{\Gamma}{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}} \vdash e : \tau }
+  { \coctx{\Gamma}{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}} \vdash \kvd{prev}~e : \tau }
+  {f}
+  {\lambda ctx.\ident{prev}_{\cclrd{n_0},\ldots,\cclrd{n_k}}~\ctx }
+\end{equation*}
+
+\vspace{1em}
+\paragraph{Reduction rules}
+
+\begin{equation*}
+\begin{array}{rl}
+{\footnotesize(\emph{counit})} & \begin{array}{l}
+\ident{counit}_{\cclrd{0}}(\ident{Df} \langle\langle v_0 \rangle\rangle)   \rightsquigarrow v_0
+\end{array}
+\\[0.25em]
+{\footnotesize(\emph{cobind})} & \begin{array}{l}
+\ident{cobind}_{\cclrd{m}, \alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}~f~\ident{Df}\langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle \rightsquigarrow \\[-0.25em]
+\quad \ident{Df}\langle\langle\; f (\ident{Df}\langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle), ~\ldots~, \\[-0.25em]
+\qquad ~~~\, f (\ident{Df}\langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle)~\; \rangle\rangle
+\end{array}
+\\[0.25em]
+{\footnotesize(\emph{merge})} & \begin{array}{l}
+\ident{merge}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}} ((\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle), (\ident{Df}\langle \textbf{v}'_0, \ldots, \textbf{v}'_l\rangle)) \rightsquigarrow
+\\[-0.25em]
+\quad \ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k, \textbf{v}'_0, \ldots, \textbf{v}'_l \rangle
+\end{array}
+\\[0.25em]
+{\footnotesize(\emph{split})} & \begin{array}{l}
+\ident{split}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}} (\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k, \textbf{v}'_0, \ldots, \textbf{v}'_l \rangle)  \rightsquigarrow
+\\[-0.25em]
+\quad (\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle, ~\ident{Df}\langle \textbf{v}'_0, \ldots, \textbf{v}'_l\rangle)
+\end{array}
+\\[0.25em]
+{\footnotesize(\emph{prev})} & \begin{array}{l}
+\ident{prev}_{\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}} (\ident{Df}\langle\langle v_{1,0}, \ldots, v_{1, n_1}, v_{1,{n_1}+1} \rangle,\ldots,\langle v_{k,0}, \ldots, v_{k, n_k}, v_{k,{n_k}+1} \rangle\rangle) \rightsquigarrow
+\\[-0.25em]
+\quad \ident{Df}\langle\langle v_{1,0}, \ldots, v_{1, n_1} \rangle,\ldots,\langle v_{k,0}, \ldots, v_{k, n_k} \rangle\rangle
+\end{array}
+\\[0.25em]
+{\footnotesize(\emph{dup})} & \begin{array}{l}
+\ident{dup}_{\cclrd{m},\cclrd{n}} (\ident{Df}\langle\langle v_0, \ldots, v_{\mathit{max}(m,n)} \rangle\rangle) \rightsquigarrow
+  \ident{Df}\langle\langle v_0, \ldots, v_m \rangle, \langle v_0, \ldots, v_n \rangle\rangle
+\end{array}
+\end{array}
+\end{equation*}
+
+\caption{Additional constructs for modelling structural dataflow}
+\label{fig:structural-ext-df}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Domain-specific extensions.}
+
+The Figure~\ref{fig:structural-ext-df} extends the target functional langauge with constructs,
+typing rules, translation rules and reduction rules needed for modelling structural dataflow.
+In the structural model, the type $\ctyp{\alift{\cclrd{r_0}, \ldots, \cclrd{r_n}}}{(\tau_0\xtimes\ldots\xtimes\tau_n)}$
+represents a structure that provides values and additional contexts for variables of types
+$\tau_0,\ldots,\tau_n$ with contextual capabilities as specified by corresponding coeffect annotations
+$\cclrd{r_0}, \ldots, \cclrd{r_n}$.
+
+In case of dataflow, the comonadically-inspired structure keeps vectors of past values for each of
+the variables in the context. In the syntax, we write $\textbf{v}$ and $\textbf{e}$ for vectors of
+values and expressions, respectively. The expression $\ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle$
+is then formed by a vector of variable assignments where each variable assignment is a vector
+of current and past values. When reducing expressions to values, we reduce values from left to
+right and from current-most value to the last past value. This is specified by the context $K$.
+
+When type-checking expressions that create the \ident{Df} values, we use an auxiliary judgement
+$\Gamma \vdash_{\ident{vec}} e : \tau, \cclrd{n}$. The judgement checks that a vector of expressions
+$\textbf{e}$ contains exactly $\cclrd{n}$ expressions of type $\tau$. This is caputred by the
+(\emph{vec}) rule, which is then used to check individual elements of the context in the
+(\emph{df}) rule.
+
+\paragraph{Properties.}
+
+Now consider a target language consisting of the core (ML-subset) defined by the syntax,
+reduction rules and typing rules given in Figure~\ref{fig:semantics-target} (Chapter~\ref{ch:semantics})
+with primitives inspired by structural indexed comonads defined in Figure~\ref{fig:struct-semantics-ext}
+and also concrete notion of comonadically-inspired value and reduction rules for dataflow as
+defined in Figure~\ref{fig:structural-ext-df}.
+
+As with the examples discussed in Chapter~\ref{ch:semantics}, the resulting language is type safe.
+Together with the well-typedness of the translation (Theorem~\ref{thm:struct-welltyped}), this
+guarantees type safety of the structural coeffect calculus for dataflow.
+In order to prove type safety, we first extend the \emph{canonical forms lemma}
+(Lemma~\ref{thm:semantics-fp-canon}) and the \emph{preservation under substitution lemma}
+(Lemma~\ref{thm:semantics-fp-pres-subst}). Those need to consider the new (\emph{df})
+and (\emph{prev}) typing rules and substitution under the newly introduced expression forms
+$\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\cclrd{n}}$. We show that the translation
+rule for \kvd{prev} produces well-typed expressions. Finally, we extend the type preservation
+(Theorem~\ref{thm:structural-fp-pres}) and progress (Theorem~\ref{thm:structural-fp-prog}) theorems.
+
+\begin{theorem}[Well-typedness of the \kvd{prev} translation]
+\label{thm:structural-df-welltyped}
+Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$,
+the translated program $f$ obtained using the rules in Figure~\ref{fig:struct-translation}
+and Figure~\ref{fig:structural-ext-df} is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
+\end{theorem}
+\begin{proof}
+By rule induction over the derivation of the translation.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{var, num, abs, app}): As before.
+
+\vspace{0.5em}\noindent\hangindent=0.6cm
+Case (\emph{prev}): Type of $\ctx$ is $\ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)}$
+  and so we can apply the (\emph{prev}) rule to obtain $\ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)}$
+  as required by $f$.
+\end{proof}
+
+\begin{lemma}[Canonical forms]
+\label{thm:structural-df-canon}
+For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
+\begin{enumerate}
+  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
+  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
+  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
+  \item If $\tau=\ctyp{\alift{\cclrd{n_1}, \ldots, \cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)}$
+    then $e = \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_n\rangle$ for some $\textbf{v}_i$
+    such that $\textbf{v}_i = \langle v_0, \ldots, v_{n_i}\rangle$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+  (1,2,3) as before; for (4) the last typing rule must have been (\emph{df}).
+\end{proof}
+
+\begin{lemma}[Preservation under substitution]
+\label{thm:structural-df-pres-subst}
+For all $\Gamma, e, e', \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
+then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
+\end{lemma}
+\begin{proof}
+  By induction over the derivation of $\Gamma, x:\tau \vdash e : \tau'$ as before, with new
+  cases for $\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\aclrd{\textbf{n}}}$.
+\end{proof}
+
+\begin{theorem}[Type preservation]
+\label{thm:structural-df-pres}
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow e'$ then $\Gamma \vdash e' : \tau$
+\end{theorem}
+\begin{proof}
+  Rule induction over $\rightsquigarrow$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{fn}, \emph{prj}, \emph{ctx}): As before, using Lemma~\ref{thm:structural-df-pres-subst} for (\emph{fn}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{counit}): $e = \ident{counit}_{\cclrd{0}}(\ident{Df} \langle\langle v_0 \rangle\rangle)$. The last rule in the type
+  derivation of $e$ must have been (\emph{counit}) with $\Gamma \vdash \ident{Df} \langle\lange v_0 \rangle\rangle : \ctyp{\alift{\cclrd{0}}}\tau$
+  and therefore $\Gamma \vdash v_0 : \tau$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{m}, \alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}~f~(\ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_{k} \rangle)$
+  such that $\forall i\in\{1\ldots k\}$ $\textbf{v}_i = \langle v_0, \ldots, v_{n_i}\rangle$. The last rule in the type derivation of $e$ must have been
+  (\emph{cobind}) with a type $\tau = \ctyp{\alift{\cclrd{m}}}\tau'$ and assumptions:
+\begin{itemize}[noitemsep,leftmargin=1.5cm,topsep=0pt]
+  \item[--] $\Gamma \vdash f : \ctyp{\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)} \rightarrow \tau_2$ and
+  \item[--] $\Gamma \vdash \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_{k} \rangle : \ctyp{\alift{\cclrd{m}+\cclrd{n_1},\ldots,\cclrd{m}+\cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)}$
+\end{itemize}
+%
+\hspace{0.5cm} Using the (\emph{df}) rule, the reduced expression has a type $\ctyp{\alift{\cclrd{m}}}\tau'$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{merge}, \emph{split}, \emph{next}): Similar. In all three cases, the last typing rule in the derivation of $e$
+  guarantees that the context contains correct number of vectors and each vector contains a
+  sufficient number of values of a correct type.
+\end{proof}
+
+\begin{theorem}[Progress]
+\label{thm:structural-df-prog}
+  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
+\end{theorem}
+\begin{proof}
+  By rule induction over $\vdash$.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{num,abs,var,app,proj,tup}): As before, using the adapted canonical forms lemma
+  (Lemma~\ref{thm:structural-df-canon}) for (\emph{app}) and (\emph{proj}).
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{counit}): $e = {\ident{counit}_\cunit}~e_1$. If $e_1$ is not a value, it can be reduced
+  using (\emph{ctx}) with context ${\ident{counit}_\cunit}~\_$, otherwise it is a value. From Lemma~\ref{thm:structural-df-canon},
+  $e_1 = \ident{Df}\langle\langle v \rangle\rangle$ and so we can apply (\emph{counit}) reduction rule.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{m}, \alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}~e_1~e_2$. If $e_1$ is not a value,
+  reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{m},\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}}~\_~e$. If $e_2$ is
+  not a value reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{m},\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}}~v~\_$.
+  If both are values then we have
+  $e_2 = \ident{Df}\langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle$
+  from Lemma~\ref{thm:structural-df-canon} and so we can apply the (\emph{cobind}) reduction.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{merge}): $e = \ident{merge}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}} e_1$. If $e_1$ is not a value,
+  reduce using (\emph{ctx}) with context $e = \ident{merge}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}}~\_$. If $e_1$ is
+  a value, it must be a pair of vectors $(\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle, \ident{Df}\langle \textbf{v}'_0, \ldots, \textbf{v}'_l\rangle)$
+  using Lemma~\ref{thm:structural-df-canon} and it can reduce using (\emph{merge}) reduction.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{df}): $e = \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_n\rangle$. If $\textbf{e}_i$ is not a value
+  then reduce using (\emph{ctx}) with the context $\ident{Df}\langle \ldots \rangle$.
+  Otherwise, $\textbf{e}_0, \ldots, \textbf{e}_n$ are values and so $\ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_n\rangle$
+  is also a value.
+
+\vspace{0.25em}\noindent\hangindent=0.6cm
+Case (\emph{split}, \emph{prev}): Similar. Either sub-expression is not a value, or the type
+  guarantees that it is a stream with correct number of elements to enable the (\emph{split})
+  or (\emph{prev}) reduction, respectively.
+\end{proof}
+
+\begin{theorem}[Safety of context-aware dataflow language]
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^{*} e'$ then either $e'$ is a value of type $\tau$ or
+  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
+\end{theorem}
+\begin{proof}
+  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:structural-df-pres} and Theorem~\ref{thm:structural-df-prog}.
+\end{proof}
 
@@
 framework, consisting of two calculi capturing properties of context-aware computations introduced
-in Chapter~\ref{ch:applications}. In the previous chapter, we focused on the whole-context
-properties of computations and we developed \emph{flat coeffect calculus} to capture them.
+in Chapter~\ref{ch:applications}. In Chapters~\ref{ch:flat} and \ref{ch:semantics}, we focused on
+whole-context properties of computations and we developed \emph{flat coeffect calculus} to capture them.
 This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
@@
 
-Following the structure of the previous chapter, the structural coeffect calculus is parameterized
+Following the structure of the previous two chapters, the structural coeffect calculus is parameterized
 by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations
@@
 instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
-as $\czip$.
-
-In order to keep track of separate annotations for each variable, we use a system with
+as $\czip$. In order to keep track of separate annotations for each variable, we use a system with
 explicit structural rules (contraction, weakening and exchange) that manipulate the structure
@@
 
-The chapter presents two technical results. The first is the semantics of structural coeffect
-calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
-it unifies our three examples (bounded reuse, dataflow and liveness) and it demonstrates
-suitability of the type system. The second technical result is equational theory for the structural
-coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
-typing preservation property for any instance of the structural coeffect calculus. These two
-strong properties are desirable for programming languages, but are often not satisfied
+The structural coeffect calculus has desirable equational properties that are satisfied only by
+certain flat coeffect calculi. In particular, we show that $\beta$-reduction and $\eta$-expansion
+have the typing preservation property for any instance of the structural coeffect calculus. These
+two strong properties are desirable for programming languages, but are often not satisfied
 (\eg~by languages with effects).
 
-Two questions remain open for the upcoming chapter. Firstly, is there a way to present
-flat and structural coeffect calculi in a uniform way as a single system? Secondly, what
-is the best pathway towards a practical implementation of programming languages based on
-coeffect systems?
+Finally, we discuss the semantics of the structural coeffect calculus in in terms of \emph{structural
+indexed comonads}. As in Chapter~\ref{ch:semantics}, we first use the categorical semantics
+to unify the different notions of context (bounded reuse, dataflow and liveness) and then use
+it as a basis for translation that turns well-typed programs written in the structural coeffect
+calculus into well-typed programs of a simple functional language. We give concrete reduction
+rules for the target language, modelling \emph{structural dataflow} and prove type safety
+of the language.

### 2016-05-11 main.tex
Unique structural derivation and edits (Chapter 6)
Add structural coeffect proofs (Chapter 6)
Switch order of chapters 7 and 8
Add binding to chapter 2, whitespace, tweaks

@@
 \usepackage{amsthm}
+\usepackage{enumitem}
 %\usepackage{catchfilebetweentags}
@@
 \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
-\definecolor{strclr}{rgb}{0.5,0.1,0.0}
-\definecolor{prepclr}{rgb}{0.0,0.0,0.0}
+\definecolor{numclr}{rgb}{0.0,0.4,0.0}
+\definecolor{strclr}{rgb}{0.4,0.3,0.0}
+\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
 
@@
 \newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
-\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
-\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\ttfamily #1}}}
-\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
+\newcommand{\num}[1]{\textcolor{numclr}{#1}}
+\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
+\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
+\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
 \newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
@@
 
+\urlstyle{sf}
+
 \newcommand{\figcaption}[1]{
@@
   \sem{#2} & = & #4 \\
+\end{array}
+}
+\newcommand{\mdeff}[4]{
+\begin{array}{rcl}
+  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
+  #2 & = & #4 \\
 \end{array}
@@
 
-%----------------------------------------------------------------------------------------
-%	THESIS CONTENT - CHAPTERS
-%----------------------------------------------------------------------------------------
+% ==================================================================================================
 
-% \ctparttext{You can put some informational part preamble text here.
-Illo principalmente su nos.
-Non message \emph{occidental} angloromanic da.
-Debitas effortio simplificate sia se, auxiliar summarios da que, se avantiate publicationes via.
-Pan in terra summarios, capital interlingua se que.
-Al via multo esser specimen, campo responder que da.
-Le usate medical addresses pro, europa origine sanctificate nos se.} % Text on the Part 1 page describing  the content in Part 1
+\ctparttext{  The computing ecosystem is becoming increasingly heterogeneous and rich. Modern
+  programs need to run on a variety of devices that are all different, but provide unique rich
+  capabilities. For example, application running on a phone can access the GPS sensor, while
+  application running in the cloud can access GPU computing resources. Both diversity and richness
+  can only be expected to increase with trends such as the internet of things.
 
-% \part{Some Kind of Manual} % First part of the thesis
+  In this thesis, we argue that the creating programming languages that allow the programmer to
+  better work with the environment or \emph{context} in which applications execute is the next
+  big challenge for programming language designers.
 
-% \include{Chapters/Chapter01} % Chapter 1
+  We start with a detailed discussion of the motivation for the thesis and an overview of our
+  methodology (Chapter~\ref{ch:intro}). Next, we discuss previous programming langauge research that
+  leads to the work presented in this thesis (Chapter~\ref{ch:pathways}) and we examine a number
+  of practical context-aware systems in detail (Chapter~\ref{ch:applications}), identifying two
+  kinds of context that we later capture by \emph{flat} and \emph{structural coeffects}.
+}
+\part{Context-aware programming}
+\label{part:context-aware}
 
-% \cleardoublepage % Empty page before the start of the next part
-
-%------------------------------------------------
-
-% \ctparttext{You can put some informational part preamble text here.
-Illo principalmente su nos.
-Non message \emph{occidental} angloromanic da.
-Debitas effortio simplificate sia se, auxiliar summarios da que, se avantiate publicationes via.
-Pan in terra summarios, capital interlingua se que.
-Al via multo esser specimen, campo responder que da.
-Le usate medical addresses pro, europa origine sanctificate nos se.} % Text on the Part 2 page describing the content in Part 2
-
-% \part{The Showcase} % Second part of the thesis
-
-%\include{text/Chapter01}
-%\include{text/Chapter02}
-%\include{text/Chapter03}
+\include{text/Chapter01}
+\include{text/Chapter02}
+\include{text/Chapter03}
 
@@
 }
-\setcounter{part}{1}
 \part{Coeffect calculi}
 \label{part:coeffect-calculi}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-\include{text/Chapter04}
-\include{text/Chapter05}
+%\include{text/Chapter04}
+%\include{text/Chapter05}
 %\include{text/Chapter06}
+
+\ctparttext{ TBD }
+\part{Towards practical coeffects}
+\label{part:context-aware}
+
 %\include{text/Chapter07}
-\include{text/Chapter08}
+%\include{text/Chapter08}
 %\include{text/Chapter09}
 
-%----------------------------------------------------------------------------------------
-%	THESIS CONTENT - APPENDICES
-%----------------------------------------------------------------------------------------
+% ==================================================================================================
 
-% \part{Appendix} % New part of the thesis for the appendix
+\cleardoublepage
+\include{Bibliography}
 
-% \include{Chapters/Chapter0B} % Appendix B - empty template
-
-%----------------------------------------------------------------------------------------
-%	POST-CONTENT THESIS PAGES
-%----------------------------------------------------------------------------------------
-
-\cleardoublepage\include{Bibliography} % Bibliography
-% \cleardoublepage\include{FrontBackMatter/Colophon} % Colophon
-% \cleardoublepage\include{FrontBackMatter/Declaration} % Declaration
-
-%----------------------------------------------------------------------------------------
+% --------------------------------------------------------------------------------------------------
 

### 2016-05-11 text/Chapter01.tex
Further work and conclusions (Chapter 7)
Add binding to chapter 2, whitespace, tweaks

@@
-%!TEX root = ../main.tex
-
-% ==================================================================================================
-
 \chapter{Why context-aware programming matters}
@@
 
-Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
-the goal of programming language research is generally to design languages that provide
-more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and
-more \emph{unified}. These are exactly the aims that we follow in this thesis. In this chapter, 
-we explain what the common problems in context-dependent programming are. In 
-Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we develop two simple calculi to 
-understand and capture the structure of those problems and, finally, Chapter~\ref{ch:unified} 
-unifies the two abstractions.
-
 % ==================================================================================================
@@
 %                                                                                                                                                                       % ==================================================================================================
+% ==================================================================================================
 
 \section{Why context-aware programming matters}
+\label{sec:intro-whymatters}
 
@@
    network data), we may have permission to access only some of the data and we may want to track
-   \emph{provenance} information. However, this is not checked -- if a program attempts to access
-   unavailable data, the access will be refused at run-time.
+   \emph{provenance} information. However, this is not checked statically -- if a program attempts
+   to access unavailable data, the access will be refused at run-time.
 
@@
 \begin{quote}
-Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
-not supported for type \texttt{System.String}.
+\emph{Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
+not supported for type \texttt{System.String}.}
 \end{quote}
@@
 The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
-\cite{app-sql-injection}. An attacker could enter \str{'; DROP TABLE Products -{}-} as their 
-\ident{name} and delete the database table ``Products''! For this reason, most libraries discourage 
+\cite{app-sql-injection}. An attacker could enter \str{'; DROP TABLE Products -{}-} as the
+\ident{name} and delete the database table ``Products''. For this reason, most libraries discourage
 building SQL commands by string concatenation, but there are still many systems that do so.
@@
 Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
-meta-data about variables that are in some ways special. Such meta-data can determine how the variables
+metadata about variables that are in some ways special. Such metadata can determine how the variables
 can be used. Here, \ident{name} comes from the user input. This information about the value should
@@
 In the security context, such marking of values (but at run-time) is called \emph{tainting}
-\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
-tracking \cite{app-provenance-future}. This can be useful when working with data in other contexts. 
-For example, data jounralsts might want to propagate meta-data about the quality and the 
-information source -- is the source trustworthy? Is the data up-to-date? Such meta-data could 
+\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance
+tracking} \cite{app-provenance-future}. This can be useful when working with data in other contexts.
+For example, data journalists might want to propagate metadata about the quality and the
+information source -- is the source trustworthy? Is the data up-to-date? Such metadata could
 propagate to the result and tell us important information about the calculated results.
@@
 
-The information about access patterns can be used to efficiently compile the computation
+Access patterns can be used to efficiently compile the computation
 by preallocating the necessary space (as we know which sub-range of the array might be accessed).
@@
 The context includes notions such as execution environment, capabilities provided by the
-environment or input and meta-data about the input and variables through which it is accessed.
+environment or input and metadata about the input and variables through which it is accessed.
 
@@
 \begin{itemize}
-\item \textsc{Flat coeffects} -- represent additional data, resources and meta-data that are
+\item \textsc{Flat coeffects} represent additional data, resources and metadata that are
    available in the execution environment (regardless of how they are accessed in a program).
@@
 
-\item \textsc{Structural coeffects} -- capture additional meta-data related to inputs. This can include
+\item \textsc{Structural coeffects} capture additional metadata related to inputs. This can include
    provenance (source of the input value), usage information (how often is the value
@@
 \begin{array}{l}
-\ident{\textbf{fetchLocalNews}}\\
+\ident{fetchLocalNews}\\
 \hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
@@
 
+\noindent
 The details of how the cross-compilation would work are out of the scope of this thesis. However,
@@
 expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
-\ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
+\ident{float} or \ident{string}, depending on the types of the variables. For example,
 the following is a valid typing judgement in F\# \cite{app-fsharp}:
@@
 in the context -- they also capture detailed information about the execution environment.
-In the above example, the system tracks meta-data about the variables and annotates \ident{password} 
+In the above example, the system tracks metadata about the variables and annotates \ident{password}
 as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
@@
   directions developed recently in the theory of programming languages (including type and effect
-  systems, categorical semantics and sub-structural logics).
+  systems, categorical semantics and substructural logics).
 
@@
 
-\item In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we present the key
-  contributions of this thesis. We develop the \emph{flat} and \emph{structural} calculi, show
-  how they capture important contextual properties and develop their categorical semantics using
-  a notion based on comonads. Chapter~\ref{ch:unified}  links the two systems using a single 
-  formalism that is capable of capturing both flat and structural properties and also discusses
-  an alternative presentation that is more suitable for automatic type inference of coeffects.
+\item In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we capture important contextual
+  properties using a simple theoretical models. We develop the \emph{flat coeffect calculus} that
+  captures per-context properties and \emph{structural coeffect calculus} that captures
+  per-variable properties. We give a type system for the calculi and study their equational
+  properties.
 
-\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis. 
-  One important direction deserves further exploration, and so Chapter~\ref{ch:unified}
-  also includes a brief discussion of a different approach to tracking contextual information that 
-  arises from modal logics. This is followed by an outline of the future work that is needed
-  in order to implement the presented theory in main-stream programming languages.
+\item In Chapter~\ref{ch:semantics} we give categorical semantics of the flat coeffect calculus.
+  This provides a unified way of defining the semantics of context-aware languages. We use the
+  categorical semantics as a basis for \emph{categorically-inspired translation} that turns
+  context-aware programs into programs in a simple functional language. We prove that well-typed
+  context-aware programs are translated to programs that ``do not go wrong''. The development
+  is repeated for structural coeffects in Chapter~\ref{ch:structural}.
+
+\item In Chapter~\ref{ch:impl}, we use the translation as a basis for a prototype
+  implementation of three simple context-aware programming languages. Our implementation serves as
+  an empirical evidence showing how coeffects simplify programming with context and provide
+  additional safety guarantees. We present the implementation in the format of an \emph{interactive
+  essay} (\url{http://tomasp.net/coeffects}) that encourages active reading and exploration of
+  various aspects of the theory.
+
+\item Related work is presented in Chapter~\ref{ch:pathways} and, together with further work,
+  throughout the thesis. Two important directions deserve furhter exploration. In Chapter~\ref{ch:further},
+  we outline a unified coeffect system that is capable of capturing both flat and structural
+  properties. We also include a brief discussion of a different approach to tracking contextual
+  information that arises from modal logics.
 \end{itemize}

### 2016-05-11 text/Chapter02.tex
Add binding to chapter 2, whitespace, tweaks

@@
-%!TEX root = ../main.tex
-
-% ==================================================================================================
-
 \chapter{Pathways to coeffects}
 \label{ch:pathways}
-
-%---------------------------------------------------------------------------------------------------
 
@@
 
-We start by looking at coeffects as the dual of effect systems (Section~\ref{sec:path-eff}). Next,
-we extend the duality to category theory, looking at the categorical dual of monads known as 
-\emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically inspired type systems 
+We start (Section~\ref{sec:path-binding}) by discussing how coeffects arise from the most common
+notion of context-dependence -- variable binding. Next, we look at coeffects as the dual of effect
+systems (Section~\ref{sec:path-eff}) and we extend the duality to category theory, looking at
+\emph{comonads} (Section~\ref{sec:path-sem}). Finally we consider logically inspired type systems
 that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).
@@
 
+%===================================================================================================
+%
+%    ######
+%    #     # # #    # #####  # #    #  ####
+%    #     # # ##   # #    # # ##   # #    #
+%    ######  # # #  # #    # # # #  # #
+%    #     # # #  # # #    # # #  # # #  ###
+%    #     # # #   ## #    # # #   ## #    #
+%    ######  # #    # #####  # #    #  ####
+%
+%===================================================================================================
+
+\section{Through static and dynamic binding}
+\label{sec:path-binding}
+
+Accessing a variable is arguably the simplest form of context-dependence, to the extent that we
+do not normally think of variables as a notion of context. However, variables fit well with our
+earlier description of context in programming: a block of code that accesses a variable can only
+be executed in an environment where the variable is available.
+
+In this section, we look at variable binding through the perspective of context-requirements.
+We discuss ordinary variable binding and Haskell's implicit parameters \cite{app-implicit-parameters},
+which provides an interesting point in the design space. We then discuss ambiguity that is inherent
+in variable binding and, more generally, context-aware programming languages, and a way of
+resolving it through \emph{type-directed semantics}.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Variable binding}
+\label{sec:path-binding-var}
+
+Variable access represents a form of context-dependence. For example, an expression $x+y$ can be
+only evaluated if the environment provides values for variables $x$ and $y$. A variable
+\emph{requirements} can be satisfied in two standard ways that are captured by \emph{dynamic} and
+\emph{static} (or lexical) variable binding. Consider the following simple program:
+
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{f}=\\[-0.25em]
+\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
+\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{x}=5~\kvd{in}\\
+\ident{f}~0
+\end{array}
+\end{equation*}
+%
+The program can be evaluated in two ways, depending on the variable binding mechanism:
+%
+\begin{itemize}
+\item \textsc{Static (lexical) binding.} In a langauge with static binding (such as ML or Java),
+  the variable \ident{x} inside the body of the lambda function is statically bound to the
+  declaration in the lexical scope -- that is, the variable on the second line -- and the
+  expression evaluates to 10.
+
+\item \textsc{Dynamic binding.} In a language with dynamic binding (some variants of LISP), the
+  variable value is dynamically bound to the last value that has been added to the environment
+  during program execution. Thus, the \ident{x} variable inside the lambda function refers to
+  \ident{x} defined on line 5 of the sample and the expression evaluates to 5.
+\end{itemize}
+
+\noindent
+When we view variable access as a context requirement, we can see that the body of the function
+($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
+\ident{y}. In static binding, the context requirements of the body can be placed on the scope in
+which the function is defined (declaration-site). In dynamic binding, the requirements are
+\emph{delayed} and are placed on the scope in which the function is called (call-site).
+
+In static and dynamic scoping, all variable requirements are always placed on one site. However,
+those are not the only two options. It is conceivable that a language would use a mechanism that
+splits variable requirements differently and combines aspects of dynamic and static binding. For
+example, the language could use statis binding by default, but resort to dynamic binding if a
+variable is not available in the lexical scope. One such system is discussed in the next section.
+
+Languages with static scoping generally check that all variable requirements are satisfied.
+Attempt to access a variable that is not available in scope will give a compile error. In contrast,
+languages with dynamic variable binding typically do not perform compile-time checks -- if a
+program attempts to access a variable that is not available in the environment, a runtime error
+occurs. This is possibly artifact of their implementation -- implementing static binding
+\emph{without} checking would be cumbersome and implementing checking for dynamic binding is
+requires a more sophisticated type system (Section~\ref{sec:path-effects-coeff}).
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Implicit parameter binding}
+
+Haskell uses static binding for ordinary variables, but it additionally provides a feature named
+\emph{implicit parameters} \cite{app-implicit-parameters} that adds a special kind of variables,
+written as \ident{?param}, which use a particular combination of static and dynamic binding.
+
+The following two examples are variations on the one discussed in the Section~\ref{sec:path-binding-var}.
+We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
+implicit parameter is defined both at declaration-site and at the call-site. On the right, the
+implicit parameter is available only at the call-site:
+
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{f}=\\[-0.25em]
+\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
+\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{?x}=5~\kvd{in}\\
+\ident{f}~0
+\end{array}\hspace{5em}\begin{array}{l}
+\kvd{let}~\ident{f}=\\[-0.25em]
+\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{?x}=5~\kvd{in}\\
+\ident{f}~0
+\end{array}
+\end{equation*}
+%
+The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
+dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
+is statically bound and the context requirement is satisfied using the declaration-site context.
+Otherwise, the context requirement is delayed and has to be satisfied at the call-site.
+In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
+delayed context requirements and the expression evalautes to $10$. In the example on the right,
+the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
+the function. The expression evaluates to $5$.
+
+In Haskell, the type system checks that bindings for all required implicit parameters are
+available. The type of the function \ident{f} on the left is $\kvd{int}\rightarrow\kvd{int}$, while
+the type of the \ident{f} function on the right is $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
+The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
+in the environment when calling the function. It is worth noting that the syntax is similar to the
+one used by type-class constraints. Those can be viewed as context requirements too
+(Section~\ref{sec:applications-flat-impl}).
+
+\paragraph{Thesis perspective.}
+The three different binding mechanisms discussed so far can be seen as different ways of splitting
+context requirements of a particular kind. Dynamic and static binding represent the opposite ends
+of the design space and Haskell's implicit parameters are an interesting point inside the wider space.
+
+In this thesis, we consider different notions of context, using implicit parameters as just one
+of several motivating examples. However, implicit parameters are a valuable example, because they
+clearly illustrate the ambiguity inherent in context-aware programs -- the context requirements of
+a function can be satisfied using the context available at declaration-site or using the context
+available at the call-site.
+
+We also aim to find a description of context-aware languages that does not make ad-hoc decisions
+about how context requirements are split between the declaration-site and the call-site. While
+Haskell's solution for implicit parameters might be the most reasonable one, different notions
+of context might require different domain-specific choices and the general framework of
+context-aware programming should make that possible.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Resolving ambiguity}
+\label{sec:path-binding-amb}
+
+In many practical programming languages, the value and semantics of an expression depends on the
+type derivation. Typically, the choice is hidden behind a mechanism that selects one preferred type
+derivation.
+
+This mechanism serves as an inspiration for our approach to resolving ambiguity inherent in
+context-aware programs. This section shows a brief example using the F\# language \cite{app-fsharp},
+before revisiting the implicit parameters example.
+
+~
+
+Consider an F\# lambda expression $(\lambda\ident{x}\rightarrow\ident{x.Length})$, which takes
+an object $\ident{x}$ and returns the value of its \ident{Length} property. F\# is a nominally-typed
+language meaning that, in isolation, the function has an ambiguous meaning\footnote{In contrast,
+in a structurally-typed language, the function would have a unique typing in isolation. In OCaml,
+the type would be $\langle \ident{Length}:\ident{'a}\rangle \rightarrow \ident{'a}$.}. It can be
+a function taking an array, it can be a function taking a string, or it can be a function taking
+one of the other .NET types that are equipped with the \ident{Length} property.
+
+The semantics of the function depends on the typing derivation. For example, for arrays, it is
+compiled using the \ident{ldlen} intermediate language (IL) instruction, while or strings, it is
+compiled using \ident{call} instruction (calling the property getter). In F\#, the compiler chooses
+an appropriate typing derivation. For example:
+%
+\begin{equation*}
+\begin{array}{l}
+  \lbrack \str{hello};\str{world} \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
+  \lbrack \ident{Array.empty};~\ident{Array.create}~\num{100}~\num{0}\; \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
+\end{array}
+\end{equation*}
+%
+The $|\hspace{-0.25em}>$ operator passes the value on the left to the function on the right. In
+the first case, the compiler infers that the type of the input is a list of strings and so the
+type of the lambda function becomes $\ident{string}\rightarrow\ident{int}$. In the second case,
+the list contains two arrays (empty array and array containing one hundred 0 values) and so
+the type of the lambda function is $\ident{int}\lbrack\,\rbrack\rightarrow\ident{int}$.
+The important points about the example are:
+%
+\begin{itemize}
+\item The semantics of the function $(\lambda\ident{x}\rightarrow\ident{x.Length})$
+  depends on its type. For arrays, it is compiled using a special IL instruction, while for
+  strings, it calls a property getter.
+
+\item The compiler chooses an appropriate typing derivation. In the above case, this is done
+  based on the context in which the expression appears, but other options are possible
+  (in some cases, there is a \emph{default} resolution; in some cases, the compiler requires
+  explicit typing annotation).
+\end{itemize}
+%
+The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
+possible typing derivations and its semantics varies depending on the type. If the lexical
+scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
+and it captures the value from the lexical scope. Otherwise, the type of the function is
+$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
+value from a hidden dictionary that is passed together with the input from the call-site.
+
+\paragraph{Thesis perspective.}
+Just like the F\# function in the above example, certain expressions in context-aware languages
+developed in this thesis have multiple valid typing derivations and their semantics depends on the
+type. In F\#, the compiler determines a unique typing derivation based on other parts of the program
+(or fails, if type is not uniquely determined). In our languages, we also determine a unique typing
+derivation. However, rather than relying on type information from other parts of the program, we
+explicitly define algorithm that chooses the preferred unique derivation.
+
+This approach decouples two important aspects of context-aware programming and lets us study them
+independently -- the semantics of context-aware programs and the domain-specific way of resolving
+how context requirements are satisfied. In our treatment of implicit parameters, we consider
+multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
+but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
+behaviour of Haskell).
+
+%===================================================================================================
+%
+%    #######
+%    #       ###### ###### ######  ####  #####     ####  #   #  ####  ##### ###### #    #  ####
+%    #       #      #      #      #    #   #      #       # #  #        #   #      ##  ## #
+%    #####   #####  #####  #####  #        #       ####    #    ####    #   #####  # ## #  ####
+%    #       #      #      #      #        #           #   #        #   #   #      #    #      #
+%    #       #      #      #      #    #   #      #    #   #   #    #   #   #      #    # #    #
+%    ####### #      #      ######  ####    #       ####    #    ####    #   ###### #    #  ####
+%
 %===================================================================================================
@@
 \label{sec:path-effects-coeff}
+
 When writing the judgements of coeffect systems, we want to emphasize the fact that coeffect
@@
 sub-expressions and latent context-requirements of the function.
-
-However, the (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
+The (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
 resource requirements of the body $\cclrd{r} \cup \cclrd{s}$ are split between the \emph{immediate
@@
 
-As will be discussed in more detail in Section~\ref{sec:applications-flat-impl}, this means that a 
-resource can be captured when a function is declared (e.g.~when it is constructed on the 
-server-side where database access is available), or when a function is called (\eg~when a function 
-created on server-side requires access to current time-zone, it can use the resource available on
- the client-side). In other words, resources in this example support both static (lexical) and
- dynamic scoping.
+This is where context-aware languages permit multiple valid typing derivations as discussed in
+Section~\ref{sec:path-binding-amb}. In the example here, a resource can be captured
+when a function is declared (e.g.~when it is constructed on the server-side where database access
+is available), or when a function is called (\eg~when a function created on server-side requires
+access to current time-zone, it can use the resource available on the client-side). In other words,
+resources in this example support both static (lexical) and dynamic scoping. Out of the multiple
+valid typing derivation, we would choose one -- for example, capturing only those server-side
+resources that are not available on the client-side\footnote{This can be characterized as
+``\emph{dynamic binding when possible, static binding when needed}'' and it is, quite curiously, the
+opposite choice than the one used by Haskell's implicit parameters.}. We discuss this
+system in detail in Section~\ref{sec:applications-flat-impl}.
 
@@
 
+%===================================================================================================
+%
+%     #####
+%    #     # ###### #    #   ##   #    # ##### #  ####   ####
+%    #       #      ##  ##  #  #  ##   #   #   # #    # #
+%     #####  #####  # ## # #    # # #  #   #   # #       ####
+%          # #      #    # ###### #  # #   #   # #           #
+%    #     # #      #    # #    # #   ##   #   # #    # #    #
+%     #####  ###### #    # #    # #    #   #   #  ####   ####
+%
 %===================================================================================================
@@
 monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in
-Haskell, list monad ($\mtyp{}{\tau} = \mu \gamma.1 + (\tau \times \gamma)$) and other.
+Haskell, list monad ($\mtyp{}{\tau} = 1 + (\tau \times \mtyp{}{\tau})$) and other.
 In programming language semantics, monads can be used in two distinct ways.
@@
 programming language without needing to add additional constructs corresponding to the primitives
-of our semantics (the alternative is discussed in Section~\ref{sec:unified-meta}). To clarify the 
+of our semantics (the alternative is discussed in Section~\ref{sec:further-meta}). To clarify the
 difference, the following two sections show a minimal example of both formal systems. We follow
@@
 access and the application of a primitive function $f$ is interpreted as follows:
-%
+
 \begin{equation*}
@@
 extend the lambda calculus with additional type(s) and expressions that correspond to monadic
-primitives. Assuming $\alpha$ is a primitive type:
+primitives:
 %
 \begin{align*}
-\tau &:= \alpha \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
+\tau &:= \ident{num} \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
    e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
@@
 Moggi regards the meta-language system as more fundamental, because \emph{``its models are more
-general''}. Indeed, this is a valid and reasonable perspective. Yet, we follow the first style,
-precisely because it is \emph{less general}. As discussed in Chapter~\ref{ch:intro}, our aim is 
-to develop concrete context-aware programming languages (together with their type theory and 
-semantics) rather than to build a general framework for reasoning about languages with 
-context-dependent properties.
+general''}. This is a valid and reasonable perspective. Yet, we follow the first style,
+precisely because it is \emph{less general}. Our aim is to develop concrete context-aware
+programming languages (together with their type systems and semantics) rather than to build a
+general framework for reasoning about languages with contextual properties.
 
@@
 \end{compactitem}
+%
 such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
+
 \begin{align}
@@
 To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
-$\ctyp{}{\tau} = \mu \gamma.\tau + (\tau \times \gamma)$. A value of the type is either
-the last element $\tau$ or an element followed by another non-empty list $\tau \times \gamma$
-(consisting of the head $\tau$ and the tail $\gamma$). Note that the list must be non-empty, 
+$\ctyp{}{\tau} = \tau + (\tau \times \ctyp{}{\tau})$. A value of the type is either
+the last element $\tau$ or an element followed by another non-empty list $\tau \times \ctyp{}{\tau}$
+(consisting of the head $\tau$ and the tail $\ctyp{}{\tau}$). Note that the list must be non-empty,
 otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In
@@
 The semantics follows that of effectful computations using monads. A variable access is interpreted
-using $\ident{counit}_\ctyp{}{}$ (obtain the value and ignore additional available context); composition
+using $\ident{counit}_\ctyp{}{}$ (extract the variable value); composition
 uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
 is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).
-
 For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an
@@
 We return to this topic when discussing lambda abstraction in Section~\ref{sec:applications-structure-lam}
-and semantics of flat coeffect systems in Section~\ref{sec:flat-semantics}.
+and semantics of flat coeffect systems in Section~\ref{sec:semantics-theory}.
 
@@
 
-\paragraph{Meta-language interpretation.} ~To briefly demonstrate the app\-roach that employs comonads
-as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
-Nanevski. We do not attempt to provide a precise overview of their work. The main purpose 
-of the following discussion is to provide a different intuition behind comonads, and to present an 
-example of a language that includes comonad as a type constructor, together with language primitives
-corresponding to comonadic operations\footnote{In fact, Pfenning, Davies and Nanevski 
-\cite{logic-modal-reconstruction,logic-cmtt} never mention comonads explicitly. This is done in later 
-work by Gabbay et al. \cite{logic-cmtt-semantics},  but the connection between the language and comonads
+\paragraph{Meta-language interpretation.} To demonstrate the approach that employs comonads
+as part of a meta-language, we look at an example inspired by the work of Pfenning et al.
+\cite{logic-modal-reconstruction,logic-cmtt}. We do not attempt to provide a precise overview of
+their work. The main purpose of the following discussion is to provide a different intuition behind
+comonads, and to present an example of a language that includes comonad as a type constructor,
+together with language primitives corresponding to comonadic operations\footnote{In fact,
+Pfenning et al. never mention comonads explicitly. This is done in later work by Gabbay et al.
+\cite{logic-cmtt-semantics},  but the connection between the language and comonads
 is not as direct as in case of monadic or comonadic semantics covered in the previous section.}.
@@
 This is another key aspect that distinguishes coeffects from effect systems. We return back to
-the meta-language approach of embedding comonads in Section~\ref{sec:unified-meta}.
+the meta-language approach of embedding comonads in Section~\ref{sec:further-meta}.
 
@@
 
-The \emph{meta-language} approach of Pfenning, Davies and Nanevski is closely related to
+The \emph{meta-language} approach of Pfenning et al. is closely related to
 our work. Most importantly, Contextual Modal Type Theory (CMTT) uses indexed $\square$ modality
@@
 %===================================================================================================
+%
+%     #####
+%    #     # #    # #####   ####  ##### #####  #    #  ####  ##### #    # #####    ##   #
+%    #       #    # #    # #        #   #    # #    # #    #   #   #    # #    #  #  #  #
+%     #####  #    # #####   ####    #   #    # #    # #        #   #    # #    # #    # #
+%          # #    # #    #      #   #   #####  #    # #        #   #    # #####  ###### #
+%    #     # #    # #    # #    #   #   #   #  #    # #    #   #   #    # #   #  #    # #
+%     #####   ####  #####   ####    #   #    #  ####   ####    #    ####  #    # #    # ######
+%
+%===================================================================================================
 
@@
 judgement: $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$. In other words, our work focuses on
-``what is happening on the left hand side of $\vdash$''.
+what is happening on the left hand side of $\vdash$.
 
@@
 specifically, about the sub-tree of the context. Although this is not the exact definition used in
-Chapter~\ref{ch:unified}, we could define contexts as follows:
+Chapter~\ref{ch:structural}, we could define contexts as follows:
 %
@@
 We can also annotate individual components. For example, a context containing variables $x,y,z$
-where only $x$ is used could be written as $\coctx{(x\!:\!\tau_1}{\ident{\cclrd{used}}}, \coctx{(y\!:\!\tau_2, z\!:\!\tau_3)}{\ident{\cclrd{unused}}}$.
+where only $x$ is used could be written as $\coctx{(x\!:\!\tau_1)}{\ident{\cclrd{used}}}, \coctx{(y\!:\!\tau_2, z\!:\!\tau_3)}{\ident{\cclrd{unused}}}$.
 
@@
 
+%===================================================================================================
+%
+%     #####  ####### ######
+%    #     # #     # #     #
+%    #       #     # #     #
+%    #       #     # ######
+%    #       #     # #
+%    #     # #     # #
+%     #####  ####### #
+%
 %===================================================================================================
@@
 %===================================================================================================
+%
+%     #####
+%    #     # #    # #    # #    #   ##   #####  #   #
+%    #       #    # ##  ## ##  ##  #  #  #    #  # #
+%     #####  #    # # ## # # ## # #    # #    #   #
+%          # #    # #    # #    # ###### #####    #
+%    #     # #    # #    # #    # #    # #   #    #
+%     #####   ####  #    # #    # #    # #    #   #
+%
+%===================================================================================================
 
@@
 
-This chapter presented three different pathways leading to the idea of coeffects. We also 
+This chapter presented four different pathways leading to the idea of coeffects. We also
 introduced the most important related work, although presenting related work was not the
@@
 
-The first pathway follows as a dualisation of well-known effect systems. However, this is not simply 
+The first pathway follows as a generalization of static and dynamic variable binding. Variable
+binding can be seen as the most primitive form of context-dependence and coeffects provide a
+generalization that can capture different binding mechanisms in a unified way. In the second
+pathway, we looked at the dual of well-known work on effect systems. However, this is not simply
 a syntactic transformation. As we further discuss in the next chapter, coeffect systems treat lambda
-abstraction differently. The second pathway follows by extending comonadic semantics of context-dependent 
+abstraction differently. The third pathway follows by extending comonadic semantics of context-dependent
 computations with indexing and building a type system analogous to effect system from the ``marriage of
-effects and monads''. Finally, the third pathway starts with substructural type systems. Coeffect
+effects and monads''. Finally, the fourth pathway starts with substructural type systems. Coeffect
 systems naturally arise by annotating bunches in bunched logics with additional information. In this

### 2016-05-11 text/Chapter03.tex
Add binding to chapter 2, whitespace, tweaks

@@
-%!TEX root = ../main.tex
-
 % ==================================================================================================
@@
   can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$.
-  We need to define the type system in a way that is similar to sub-structural systems 
+  We need to define the type system in a way that is similar to substructural systems
   (discussed in Section~\ref{sec:path-logic}) and add explicit rules for manipulating
@@
 The structural typing rules are shown in Figure~\ref{fig:applications-struct-live} (b). They mirror
-the rules know from sub-structural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
+the rules know from substructural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
 extends the context with a single unused variable $x$ and adds the $\ident{D}$ annotation to the

### 2016-05-11 text/Chapter06.tex
Add binding to chapter 2, whitespace, tweaks

@@
 systems give a more precise information about the context. However, we also look at other applications
-that arise from the work on sub-structural logics, a pathway to coeffects discussed in
+that arise from the work on substructural logics, a pathway to coeffects discussed in
 Section~\ref{sec:path-logic}.
@@
 
-The work in this chapter is more closely related to sub-structural type systems \cite{substruct-attpl-intro}.
-Sub-structural systems remove some or all of \emph{weakening}, \emph{contraction} and
+The work in this chapter is more closely related to substructural type systems \cite{substruct-attpl-intro}.
+Substructural systems remove some or all of \emph{weakening}, \emph{contraction} and
 \emph{exchange} rules. In contrast, our systems keep all three structural rules, but use them to
@@
 
-The type system for the structural coeffect calculus is similar to sub-structural type systems
+The type system for the structural coeffect calculus is similar to substructural type systems
 in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
 weakening, exchange or contraction -- this is done by checking the types of sub-expressions
-in disjoint parts of the free variable context. Unlike in sub-structural logics, our system
+in disjoint parts of the free variable context. Unlike in substructural logics, our system
 does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
@@
 The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
-using $\cunit$. Note that, as in sub-structural systems, the free variable context contains
+using $\cunit$. Note that, as in substructural systems, the free variable context contains
 \emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
@@
 transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
-from sub-structural systems.
+from substructural systems.
 
@@
 
-In order to choose a unique typing derivation, we follow the example of sub-structural type systems
+In order to choose a unique typing derivation, we follow the example of substructural type systems
 \cite{substruct-attpl-intro} and introduce a syntax-directed version of the type system (also called

### 2016-05-11 text/Chapter08.tex
Add binding to chapter 2, whitespace, tweaks

--- 2016-04-09 text/Chapter08.tex
@@
+\chapter{Related and future work}
+\label{ch:further}
+
+%===================================================================================================
+
+The main goal of this thesis is to provide a \emph{unified} calculus for tracking context
+dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified
+two kinds of contextual properties that we further covered separately. Flat coeffects in
+Chapter~\ref{ch:flat} track whole-context properties, while structural coeffects, covered in
+Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify the two
+notions.
+
+Although the results presented in this thesis are mainly of a theoretical nature, we indeed
+believe that coeffects should be integrated in main-stream programming languages. This is addressed
+in the second part of the chapter, where we outline possible approach for a practical implementation
+of coeffect systems. Finally, we conclude the chapter with a brief discussion of an alternative
+approach to coeffect system, based on the meta-language style.
+
+\paragraph{Chapter structure and contributions}
+\begin{itemize}
+\item We start by unifying the two kinds of coeffects discussed so far. In
+  Section~\ref{sec:further-unified}, we introduce the \emph{unified coeffect calculus}, which
+  generalizes flat and structural systems and can be instantiated to track both flat and structural
+  properties.
+
+\item We outline one possible approach for practical implementations of coeffects
+  (Section~\ref{sec:further-impl}). The technique is based on embedding comonadic computations
+  via a lightweight syntactic extension and using rich type systems for capturing the structure of
+  coeffect algebra.
+
+\item Finally, we discuss an alternative approach to defining coeffect systems based on the
+  meta-language style, which highlights the relationship between our work and related work arising
+  from modal logics such as CMTT \cite{logic-cmtt} (Section~\ref{sec:further-meta}).
+\end{itemize}
+
+\noindent
+The unified coeffect calculus presented in the first part of the chapter is a novel result that
+completes the theory developed in this thesis. The next two sections present important future and
+related work, respectively.
+
 % ==================================================================================================
 
-\chapter{Implementation}
-\label{ch:impl}
+\section{The unified coeffect calculus}
+\label{sec:further-unified}
 
-In the previous three chapters, we presented the theory of coeffects consisting of type system
-and comonadically inspired semantics of two parameterized coeffect calculi. The theory provides
-a framework that simplifies the implementation of safe context-aware programming languages. To
-support this claim, this chapter presents a prototype implementation of three coeffect languages --
-language with implicit parameters and both flat and structural versions of a data-flow language.
-
-The implementation directly follows the thoery presented in the previous three chapters. It
-consists of a common framework that provides type checking and translation to a simple functional
-target language with comonadically-inspired primitives. Each concrete context-aware language then
-adds a domain-specific rule for choosing unique typing derivation (as discussed in
-Section~\ref{sec:flat-unique}) together with a domain-specific definition of the
-comonadically-inspired primitives that define the runtime semantics (see Section~\ref{sec:semantics-proofs}).
-
-The main goal of the implementation is to show that the theory is practically useful and to present it
-in a more practical way. However, we do not intend to build a complete real-world programming language.
-For this reason, the implementation is available primarilly as an interactive web-based essay, though
-it can be also downloaded and run locally.
-
-\paragraph{Chapter structure and contributions}
+The type systems of the flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural
+coeffect calculus (Figure~\ref{fig:struct-types}) differ in a number of ways. Understanding the
+differences is the key to reconciling the two systems:
 
 \begin{itemize}
-\item We discuss how the implementation follows from the theory presented earlier (Section~\ref{sec:impl-theory}).
-  This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
-  to a simple target langauge that is then iterpereted. We also discuss how the common framework makes it
-  easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).
+\item The structural coeffect calculus contains explicit rules for context manipulation
+  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
+  explicitly, but are admissible.
 
-\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate ointeresting
-  aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
-  difference between flat and structural systems (Section~\ref{sec:impl-case-typing}) and the
-  comonadically-inspired translation (Section~\ref{sec:impl-case-transl}).
+\item In the structural coeffect calculus, the variable context is treated as a vector
+  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
+  the variable context is a finite partial function (from names to types) and is annotated with
+  a single (scalar) coeffect.
 
-\item We briefly outline some of the interesting technical details of the implementation
-  (Section~\ref{sec:impl-tech}). To make the implementation easily accessible, we make it
-  available via web and we also implement a limited support for type inference.
+\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
+  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
+  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
+  and so the structural coeffect algebra uses the same tensor product $\atimes$.
+\end{itemize}
 
-\item The implementation is available not just as downloadable code, but also in the format of interactive essay
-  (Section~\ref{sec:impl-essay}), which aims to make coeffects accessible to a broader audience.
-  We conclude this chapter by discussing the most interesting aspects of the presentation.
+\noindent
+In the unified calculus presented in this section, we address the three differences as follows.
+We include explicit rules for context manipulation in the calculus; in systems that arise from flat
+calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
+the structure of coeffect annotations using the notion of a ``container'' which can be specialized
+to obtain a single annotation or a vector of annotations. This could potentially be used to capture
+other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
+between splitting and merging of context requirements (using the notation $\apar$ and $\azip$,
+respectively). For structural coeffect calculi, the two operators coincide, but for flat
+coeffect calculi, they provide the needed flexibility.
+
+\subsection{Shapes and containers}
+Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of
+Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
+to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
+types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
+lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise
+as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
+For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
+specified length). More formally:
+
+\begin{definition}
+A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
+Here, $S$ is a set of shapes and $P$ is a shape-indexed family $P : S \rightarrow \ident{Set}$
+of positions.
+\end{definition}
+
+\noindent
+Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams,
+trees and the singleton data type (which contains exactly one element). Containers relevant
+to our work are lists and singleton data types:
+
+\begin{itemize}
+\item The container representing lists is given by $S \triangleleft P$ where
+  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a
+  given length $n$ is the set of indices $P(n)= \{ 1 \ldots n \}$.
+
+\item The container representing the singleton data type is given by $S \triangleleft P$ where
+  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
+  shape $\ast$ contains exactly one position. To follow the intuition based on offset or
+  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
 \end{itemize}
+
+\noindent
+In the unified coeffect calculus, the structure of coeffect annotations is defined by a
+container with additional operations (discussed later) that links it with the free-variable
+context $\Gamma$.
+
+\subsection{Structure of coeffects}
+In the structural coeffect calculus, the coeffect annotation was formed by a vector of coeffect
+scalars. The annotations in the unified coeffect calculus are similar, but a \emph{vector} is
+replaced with a more general \emph{container}. The primitive coeffect annotations in the unified
+calculus are formed of \emph{coeffect scalars}. The coeffect scalars are equipped with the same
+structure as in structural coeffect calculus (Definition~\ref{def:structural-scalar}). In this
+section, we refer to it as \emph{\cclrd{unified coeffect scalar}} (and we repeat the definition
+below). Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
+scalar values are attached to the free-variable context. Finally, we define the
+\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.
+
+As in the structural coeffect calculus, the contexts in the unified calculus are annotated with
+shape-indexed coeffects, written as $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$;
+functions take just a single input parameter and so are annotated with scalar coeffect values
+$\sigma \xrightarrow{\cclrd{r}} \tau$.
+
+\paragraph{Coeffect scalars.}
+The following definition of the coeffect scalar structure repeats the
+Definition~\ref{def:structural-scalar} from the previous chapter.
+
+\begin{definition}
+A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
+$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations
+$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
+$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
+%
+\begin{equation*}
+\begin{array}{ccr}
+r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
+\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
+\textnormal{(monoid)}
+\\
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
+\czero \;\cpar\; r = r = r \;\cpar\; \czero &
+\textnormal{(monoid)}
+\\
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
+t\; \cleq\; t &
+\textnormal{(pre-order)}
+\\
+\end{array}
+\end{equation*}
+%
+The following distributivity axioms hold:
+\begin{align*}
+\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+%
+In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
+\begin{equation*}
+\begin{array}{l}
+ \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
+ \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
+\end{array}
+\end{equation*}
+\end{definition}
+
+\noindent
+As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
+an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable,
+respectively.
+
+The $\cpar$ operation models combining of context requirements arising from multiple parts of a
+program. The meaning depends on the coeffect container. The operation can either combine requirements
+of individual variables (structural coeffects) or requirements attached to the whole context of
+multiple sub-exp\-ressions (flat coeffects).
+
+\paragraph{Coeffect containers.}
+The coeffect container is a container that determines how are scalar coeffects attached to
+free-variable contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed positions,
+the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.
+
+The mapping between the shape of the variable context and the shape of the coeffect annotation
+is not necessarily bijective. For example, coeffect annotations in flat systems have just a single
+shape $\SH = \{ \ast \}$.
+
+In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
+$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
+to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation
+$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in
+  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
+We usually write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
+The coeffect container is also equipped with an operation that appends shapes (when we concatenate
+variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
+context.
+
+\begin{definition}
+A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$
+comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a
+mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
+\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.
+
+The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable
+context, respectively. The $\stimes$ operation corresponds to concatenation of free-variable contexts.
+Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$,
+we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$.
+\end{definition}
+
+\noindent
+As discussed earlier, we use two kinds of coeffect containers that describe the structure of vectors
+(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):
+
+\begin{example}
+\label{ex:further-struct-shape}
+Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
+where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the
+number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$,
+respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} =
+{|}\Gamma_1{|} + {|}\Gamma_2{|}$.
+
+Therefore, a coeffect annotation is a \emph{vector}
+$\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$
+for each position (corresponding to a variable $x_i$ in the context).
+\end{example}
+
+\begin{example}
+\label{ex:further-flat-shape}
+Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
+The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
+with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.
+
+That is, there is a single shape
+$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
+Therefore, a coeffect annotation is drawn from $\C^{\{\sflat\}}$ which is isomorphic to $\C$
+and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
+\end{example}
+
+\begin{example}
+Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
+\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
+${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.
+This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
+$\SP(\sflat)=\emptyset$ and so coeffect annotations would be from the empty set $\C^\emptyset$.
+\end{example}
+
+\paragraph{Unified coeffect algebra.}
+The unified coeffect calculus annotates typing judgments with shape-indexed coeffect annotations.
+The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and a coeffect container
+to define shape-indexed coeffects and operations for manipulating these.
+
+The definition here reconciles the third point discussed in Section~\ref{sec:further-unified} --
+the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
+while structural coeffects use the tensor $\atimes$. In the unified calculus, we use two operators
+that can, however, coincide.
+
+\begin{definition}
+Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
+\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified
+coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
+is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
+combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
+%
+\begin{equation*}
+\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
+\end{equation*}
+\end{definition}
+
+\noindent
+A coeffect algebra induces the following three additional operations:
+%
+\begin{equation*}
+\begin{array}{rclcrcl}
+ \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP(\sunit)}  \\
+ \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
+\\[-0.5em]
+ \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
+ \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
+\\[-0.5em]
+ \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
+ \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
+\end{array}
+\end{equation*}
+%
+The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by
+the shape of a singleton context. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a
+vector by a scalar. As we always use bold face for vectors and ordinary face for scalars (as well
+as a distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
+$\sclrd{m}$ and write~$\aseq$.
+
+Finally, we define $\alen{-}$ as an operation that returns the shape of a given shape-indexed
+coeffect. The only purpose is to simplify notation, as we tend to avoid subscripts, but often need
+to specify that shapes of variable context and coeffect match,
+\eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.
+
+\paragraph{Splitting and merging coeffects.}
+The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts.
+For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in
+\C^{\SP(\sclrd{m})}$ and $\aclrd{\textbf{s}} \in \C^{\SP(\sclrd{n})}$. In the structural system, the
+context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The
+combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined
+coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}}
+\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.
+
+For structural coeffect systems such as bounded reuse, both $\apar$ and $\azip$ are
+just the tensor product $\times$ of vectors. For flat coeffect systems, the operations
+can be defined independently, letting $\azip=\czip$ and $\apar=\cpar$.
+
+The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:further-semantics}),
+where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
+a morphism that merges the capabilities provided by two contexts (in the syntactic
+reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
+is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
+context into two parts (in the syntactic reading, merges their context requirements).
+Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
+$\apar$ in \emph{conclusions}.
+
+% --------------------------------------------------------------------------------------------------
+
+\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
+\newcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
+\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
+
+\begin{figure}[t]
+
+{\small a.) Syntax-driven typing rules:}
+\begin{equation*}
+\tyrule{var}
+  { }
+  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{const}
+  {c \!:\! \tau \in \Delta}
+  {\coctx{()}{\azero} \vdash c : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\apar\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\azip\,\alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\azip\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\apar\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+
+\vspace{1.5em}
+{\small b.) Structural rules for context manipulation:}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+(\cclrd{s'} \cleq \cclrd{s})
+\end{equation*}
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
+  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+\begin{array}{l}
+ \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\hspace{2.5em}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~
+\begin{array}{l}
+ \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
+ \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+\figcaption{Type system for the unified coeffect calculus}
+\label{fig:unif-types}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Unified coeffect type system}
+\label{sec:further-types}
+The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
+shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
+how the unified system differs from the structural system.
+
+The type system for the unified coeffect calculus is parameterized by a coeffect scalar
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$
+and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
+As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
+use of variables -- the associativity is built-in. The order of variables matters, but can be changed
+using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
+are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
+annotated with coeffects scalars.
+
+\paragraph{Syntax-driven rules.}
+The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
+differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
+coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$.
+The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.
+
+In a structural system, the two annotations correspond to a singleton and empty vector, respectively.
+However, for a singleton shape with one position, the annotations are equivalent to annotating
+variables with a scalar $\cunit$ and constants with a scalar $\czero$.
+
+In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
+system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging
+of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
+in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.
+
+Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
+discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read
+the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
+information and conclusions split available contextual information.
+
+\paragraph{Structural rules.}
+The merging/splitting operations in the structural rules are changed in the same way as in the
+rules above. It is also worth noting that structural coeffect system used vectors of multiple
+elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch})
+rule denoted two-element vectors. In the unified system, this is replaced with merging/splitting
+of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.
+
+In structural systems, the two notations mean the same thing -- we are simply concatenating or splitting two
+singleton vectors. However, this generalization allows us to capture flat coeffect systems as well.
+The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
+$\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:further-flat},
+thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not
+affect the coeffect annotation are admissible for all flat systems embedded in the unified
+coeffect calculus.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffects}
+\label{sec:further-structural}
+
+The unified coeffect system uses a general notion of context shape, but it has been designed with
+structural and flat systems in mind. In this and the next section, we show how it captures
+the two coeffect systems from Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}.
+
+The unified calculus closely resembles the structural system and so using it to model structural
+systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector}
+of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect algebra formed by
+a vector (free monoid) of scalars.
+
+\begin{definition}
+\label{def:further-struct}
+Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system}
+is defined by:
+
+\begin{itemize}
+\item A coeffect container $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$ where $\SH=\mathbb{N}$ and
+  $\SP(\sclrd{n})=\{ 1 \ldots n \}$ and  ${|}x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n{|}=n$.
+
+\item A coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
+  shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
+  That is $\epsilon : \C^{\SP(\sclrd{0})}$ is an empty vector and
+  $\times : \C^{\SP(\sclrd{n})} \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{n + m})}$
+  appends vectors.
+\end{itemize}
+\end{definition}
+
+\noindent
+The definition is valid since the shape operations form a monoid
+$\sclrd{(\mathbb{N}, +, 0)}$ and $\slen{\textnormal{--}}$ (calculating the length of
+a list) is a monoid homomorphism from the free monoid to the monoid of
+shapes.
+
+\paragraph{Properties.}
+An important property of the unified system is that, when used in a structural way as discussed
+above, it gives calculi with the same properties as the structural system described in
+Chapter~\ref{ch:structural}. This can be easily seen by comparing the Figure~\ref{fig:unif-types}
+with the Figure~\ref{fig:struct-types} and using the free monoid interpretation of the unified
+coeffect algebra.
+
+\begin{remark}
+The system described in Definition~\ref{def:further-struct} is equivalent to the structural coeffect
+system described in Figure~\ref{fig:struct-types}. That is, a typing derivation using a structural
+coeffect embedded in the unified system is valid if and only if the corresponding derivation is
+valid in the structural system.
+\end{remark}
+
+\noindent
+Using the above definition, our unified coeffect system can capture per-variable coeffect properties
+discussed in Section~\ref{sec:applications-structural}. This includes the system for bounded reuse
+(which is only used in the structural form) and precise tracking of per-variable dataflow and liveness.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Flat coeffects}
+\label{sec:further-flat}
+
+The same unified coeffect system can be used to capture systems that track whole-context (flat)
+coeffects such as implicit parameters. This is achieved using a singleton-shaped container for
+coeffect annotations. The resulting system has explicit structural rules (and is syntactically
+different from the standard flat coeffect system), but we show that they are equivalent.
+
+Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:further-flat-shape}).
+In this setting, the context annotations $\C^{ \SP(\sflat) } = \{0\} \rightarrow \C$ are equivalent
+to coeffect scalars $\C$. In addition to the coeffect scalar structure, we need to define $\apar$
+and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$
+(merging of shaped coeffect annotations). The $\azip$ operation (corresponding to $\czip$
+in flat coeffect calculus) needs to be provided explicitly.
+
+\begin{definition}
+\label{def:further-flat}
+Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary operation
+$\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$,
+the unified coeffect algebra modelling a flat coeffect systems consists of:
+%
+\begin{itemize}{}
+\item A flat coeffect container $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ as defined in
+  Example~\ref{ex:further-flat-shape}.
+
+\item A flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$
+  and $\azero = \czero$ with an\\ additional binary operation $\azip = \czip$.
+\end{itemize}
+\end{definition}
+
+\noindent
+Intuitively, the requirement $(\cclrd{r}\,\czip\,\cclrd{s}) \;\cleq\; (\cclrd{r}\,\cpar\,\cclrd{s})$,
+which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
+denotes that splitting context requirements and then re-combining them preserves all the requirements
+from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
+it cannot lose requirements. This is fundamental for showing that exchange and contraction are admissible
+in the unified system.
+
+\paragraph{Properties.}
+To show that the typing of flat properties in the unified system is equivalent to the typing in the
+flat system, we show that a valid typing judgement in the first system is a valid typing judgement in
+the second system and vice versa.
+
+In one direction, we show that the unified system (capturing flat properties) permits weakening,
+contraction and exchange rules that do not change the coeffect annotations. This guarantees that
+a valid judgement in flat system is also valid in the unified system.
+
+\begin{lemma}
+\label{thm:further-weak}
+A unified coeffect calculus capturing flat properties admits weakening that does not change the
+coeffect annotation.
+\end{lemma}
+\begin{proof}
+The rule is admissible using the following derivation:
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma}{ \cclrd{r} } \vdash e : \tau}
+      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \;\apar\; \alift{ \czero }} \vdash e : \tau} }
+  { \inference
+      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r} \,\cpar\, \czero} \vdash e : \tau}
+      {\coctx{\Gamma,x \!:\! \tau_1}{\cclrd{r}} \vdash e : \tau} }
+\end{equation*}
+
+\noindent
+We write $\cclrd{r}$ rather than $\aclrd{\textbf{r}}$, because we are tracking flat properties.
+The first step is an application of the (\emph{weak}) rule. Next, we use the fact that
+$\apar=\cpar$ and $\alift{\czero}=\czero$, which is the unit of the monoid $(\C, \cpar, \czero)$.
+\end{proof}
+
+\begin{lemma}
+A unified coeffect calculus capturing flat properties admits exchange that does not change the
+coeffect annotation.
+\end{lemma}
+\begin{proof}
+We use the idempotence of $\czip$ and $\cpar$ together with the (\emph{exch}) rule:
+\begin{equation*}
+\inference
+  { \inference
+      { \inference
+          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r}} \vdash e : \tau}
+          {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r} \,\czip\, \cclrd{r}} \vdash e : \tau} }
+      { \coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau } }
+  { \inference
+      {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r}} \apar \alift{\cclrd{r}} \apar \cclrd{r}} \vdash e : \tau}
+      {\inference
+         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}\,\cpar\,\cclrd{r}} \vdash e : \tau}
+         {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r}} \vdash e : \tau} } }
+\end{equation*}
+
+\noindent
+Using idempotence, we first duplicate the annotation $\cclrd{r}$ to get a coeffect in the form
+$\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}$ as required by the
+assumption of the (\emph{exch}) rule. Note that $\alift{-}$ can be added freely as
+$\alift{\cclrd{r}}$ is equivalent to $\cclrd{r}$. After applying (\emph{exch}), we use
+idempotence of $\apar$.
+\end{proof}
+
+\begin{lemma}
+A unified coeffect calculus capturing flat properties admits contraction that does not change the
+coeffect annotation.
+\end{lemma}
+\begin{proof}
+Similarly to exchange, the proof uses idempotence of $\czip$ and $\cpar$:
+\begin{equation*}
+\inference
+  { \inference
+      {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \czip\, \cclrd{r} \czip\, \cclrd{r} \czip\, \cclrd{r}} \vdash e : \tau}
+      {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \azip \alift{\cclrd{r}} \azip \alift{\cclrd{r}} \azip \cclrd{r}} \vdash e : \tau} }
+  { \inference
+      {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \apar \alift{\cclrd{r} \,\cpar\, \cclrd{r}} \apar \cclrd{r}} \vdash \subst{e}{z,y}{x} : \tau}
+      {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{r} \,\cpar\, \cclrd{r} \,\cpar\, \cclrd{r}} \vdash \subst{e}{z,y}{x} : \tau} }
+\begin{array}{r} \\[2.5em] \qedhere \end{array}
+\end{equation*}
+\end{proof}
+
+\noindent
+In the last two cases, we need to turn the coeffect into a form that is required by the exchange
+and contraction rules. Aside from idempotence, we could use the unit property and obtain
+\eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this approach does not
+work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
+the existence of a unit for $\czip$).
+
+In the other direction, we need to show that any valid judgement in the unified system (tracking
+flat properties) is also valid in the flat system. The syntax-directed rules are the same in both
+systems, but we need to show that any use of (explicit) weakening, contraction and exchange can be
+derived in the flat system.
+
+\begin{lemma}
+Weakening, as defined in a unified coeffect calculus capturing flat properties, is admissible in
+the flat coeffect calculus.
+\end{lemma}
+\begin{proof}
+Similar to the proof in Lemma~\ref{thm:further-weak}. The property follows from the fact that
+$\bot=\alift{\czero}$ is the unit of $\cpar$.
+\end{proof}
+
+\begin{lemma}
+\label{thm:further-rev-contr}
+Contraction, as defined in a unified coeffect calculus capturing flat properties, is admissible in
+the flat coeffect calculus.
+\end{lemma}
+\begin{proof}
+The application of (\emph{contr}) rule has the following form:
+\begin{equation*}
+\inference
+ { \inference
+    {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}} \vdash e : \tau}
+    {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\azip\, \alift{\cclrd{s}} \,\azip\, \alift{\cclrd{t}} \,\azip\, \cclrd{q}} \vdash e : \tau} }
+ { \inference
+    {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\apar\, \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \,\apar\, \cclrd{q}} \vdash \subst{e}{z,y}{x} : \tau}
+    {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{q}} \vdash \subst{e}{z,y}{x} : \tau} }
+\end{equation*}
+%
+From Definition~\ref{def:further-flat}, we know that $(\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q})
+\;\cleq\; (\cclrd{r} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{q})$. Thus, the judgement
+ can be derived using the (\emph{sub}) rule of flat coeffect calculus.
+\end{proof}
+
+\begin{lemma}
+\label{thm:further-rev-exch}
+Exchange, as defined in a unified coeffect calculus capturing flat properties, is admissible in
+the flat coeffect calculus.
+\end{lemma}
+\begin{proof}
+The application of (\emph{exch}) rule has the following form:
+\begin{equation*}
+\inference
+ {\inference
+    {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q}} \vdash e : \tau}
+    {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\cclrd{r} \,\azip\, \alift{\cclrd{s}} \,\azip\, \alift{\cclrd{t}} \,\azip\, \cclrd{q}} \vdash e : \tau} }
+ {\inference
+    {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\apar\, \alift{\cclrd{t}} \,\apar\, \alift{\cclrd{s}} \,\apar\, \cclrd{q}} \vdash e : \tau}
+    {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\cclrd{r} \,\cpar\, \cclrd{t} \,\cpar\, \cclrd{s} \,\cpar\, \cclrd{q}} \vdash e : \tau} }
+\end{equation*}
+%
+From idempotence of $\czip$, we get that $\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t} \,\czip\, \cclrd{q} =
+\cclrd{r} \,\czip\, \cclrd{t} \,\czip\, \cclrd{s} \,\czip\, \cclrd{q}$. Thus the judgement can be derived using
+(\emph{sub}) as in contraction.
+\end{proof}
+
+\noindent
+A consequence of the equivalence discussed above is that the unified coeffect system can capture
+all properties that can be captured by the flat coeffect system -- including implicit parameters,
+rebindable resources, Haskell type classes (discussed by Orchard~\cite{comonads-dom-thesis}),
+data-flow and variable liveness.
+
+%--------------------------------------------------------------------------------------------------
+
+\subsection{Semantics of unified calculus}
+\label{sec:further-semantics}
+
+The semantics of the unified calculus can be obtained by generalizing the semantics of the structural
+calculus in the same way as we generalized the type system. However, it is worth exploring the type
+signatures of the operations of the underlying (comonad-based) structure.
+
+The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
+structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The
+index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of
+the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape
+$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in
+Section~\ref{sec:further-types}, we need to modify the domains and codomains of the \ident{merge},
+\ident{split} and \ident{dup} operations as follows.
+%
+\begin{equation*}
+\begin{array}{lcl}
+ \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
+  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
+ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
+  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
+ \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
+  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
+\end{array}
+\end{equation*}
+%
+The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
+\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
+(in \ident{merge}, but also in \ident{dup} which models contraction).
+This means that the splitting and merging the context can change the coeffect annotation:
+%
+\begin{equation*}
+\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
+  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
+  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
+\end{equation*}
+%
+For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
+splitting and then merging a context may lose some of the available contextual capabilities
+(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:further-rev-contr}
+and Lemma~\ref{thm:further-rev-exch} relied on sub-coeffecting in the flat system.
+
+\paragraph{Related and future work.}
+The semantics presented here serves main\-ly to inform the design of the type system. For this
+reason, we keep the semantics simple and close to the concrete notions used in the type system
+such as flat, structural and unified coeffect algebras. In particular, the indexed comonad
+structure consists of object mapping $\ctyp{r}{}$ that is indexed by a corresponding algebra:
+
+\begin{compactitem}
+\item[--] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
+\item[--] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is
+  indexed by vectors of scalars $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$
+\item[--] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
+  the mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is indexed by shape-indexed scalars
+  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP(\sclrd{m})}$
+\end{compactitem}
+
+\noindent
+In this treatment, the indices are formed by objects that are modelled outside of category theory.
+A fully categorical semantics can be found in a joint paper with Orchard \cite{coeffects-icfp14}.
+There, the index (coeffect scalars and shape-indexed scalars) are modelled as categories too.
+The definition is written using the following notation:
+
+\begin{compactitem}
+\item[--] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
+\item[--] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
+\end{compactitem}
+
+\noindent
+In the categorical semantics of coeffect calculi, coeffect scalars are modelled as a
+category of scalars $\cclrd{\mathbb{I}}$ such that $\textit{obj}(\cclrd{\mathbb{I}}) = \C$.
+The structure of the context (indexed comonad) over the input then becomes a functor
+indexed by the free-variable context shape $\slen{\Gamma}$ and coeffect annotations
+$\aclrd{\textbf{r}}$, \ie~$\ctyp{\slen{\Gamma}}_{\aclrd{\textbf{r}}}$:
+%
+\begin{equation*}
+\sem{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \ctyp{\slen{\Gamma}}_{\aclrd{R}}
+\sem{\Gamma} \rightarrow \sem{\tau}
+\end{equation*}
+%
+The structure $\ctyp{}$ can be thought of as a dependent product of functors $\ctyp{\sclrd{n}}$
+over possible shapes $\sclrd{n} \in \SH$ where:
+%
+\begin{equation*}
+\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]
+\end{equation*}
+%
+For a fixed context shape $\sclrd{n}$ the functor
+$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]$
+maps a coeffect indexed by positions $\SP(\sclrd{n})$ to a functor from a context
+$\mathbb{C}^{\sclrd{n}}$ to an object in $\mathbb{C}$. That is, given a coeffect annotation
+(matching the shape of the context), we get a functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$.
+From a programming perspective, this functor defines a data structure that models the additional
+context provided to the program. The shape of this data structure depends on the coeffect
+annotation $\cclrd{\mathbb{I}}^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
+the number of values needed for each variable and the functor will be
+formed by lists of length matching the required number.
+
+In the model above, shapes and positions are still treated as ordinary sets. Indeed, in the
+joint paper \cite{coeffects-icfp14}, we treat shapes as just sets of sets. In this chapter, we
+instead define \emph{coeffect container} as an extension of containers, which have a well-defined
+categorical structure \cite{semantics-containers}. A future work is adapting the semantics of
+the unified coeffect calculus so that it is based on the semantics of containers. However, note
+that context is \emph{not} a (simple) container, because containers are used as indices (coeffect
+annotations) of the object that wrap the free-variable context.
+
+% =================================================================================================
+%
+%      ####                         #       #                   ##
+%      #   #                        #                            #
+%      #   #  # ##    ###    ###   ####    ##     ###    ###     #
+%      ####   ##  #      #  #   #   #       #    #   #      #    #
+%      #      #       ####  #       #       #    #       ####    #
+%      #      #      #   #  #   #   #  #    #    #   #  #   #    #
+%      #      #       ####   ###     ##    ###    ###    ####   ###
+%
+% =================================================================================================
+
+\section{Towards practical coeffects}
+\label{sec:further-impl}
+
+As discussed earlier, the main focus of this thesis is the development of the much needed
+\emph{theory of context-aware programming languages} and so discussing the details of a practical
+implementation of coeffect tracking is beyond the scope of the thesis. However, this section
+briefly outlines one possible pathway towards this goal.
+
+Many of the examples of contextual computation that we discussed earlier have been implemented as
+a single-purpose programming language feature (\eg~implicit parameters \cite{app-implicit-parameters}
+or distributed computations \cite{app-distributed-ml5,app-distributed-links}). However,
+the main contribution of this thesis is that it captures
+\emph{multiple} different notions of context-aware computations using just a \emph{single}
+common abstraction. For this reason, we advocate that future practical implementations of
+coeffects should not be single-purpose language features, but rather reusable abstractions that
+can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
+
+In order to do this, programming languages need to provide two features; one that allows
+embedding of context-aware computations themselves in programs akin to the ``do'' notation
+in Haskell (Section~\ref{sec:further-impl-embed}) and one that allows tracking of the
+contextual information in the type system (Section~\ref{sec:further-impl-types}). To make
+constraint resolution in type inference easier, we also discuss how the coeffect algebra
+can be simplified (Section~\ref{unified-impl-semilattice})
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding contextual computations}
+\label{sec:further-impl-embed}
+
+The embedding of \emph{contextual} computations into programming languages can follow the successful
+model of \emph{effectful} computations. In purely functional programming languages such as Haskell,
+effectful computations are embedded by \emph{implementing} their model within the language
+and inserting the necessary (monadic) plumbing. This is made easier by the ``do'' notation or
+monad comprehensions \cite{other-haskell98,monads-compre}, which insert the monadic operations
+automatically.
+
+The recently proposed ``codo'' notation \cite{comonads-codo} provides similar automation for
+context-aware computations based on comonads. The notation follows the semantics of our
+flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based
+on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting future work --
+this would require explicitly manipulating individual context variables and application of structural
+rules, which is not needed in flat coeffects.
+
+In ML-like languages, effects (and many coeffects) are built-in into the language semantics,
+but they can still benefit from a special notation for explicitly marking effectful (coeffectful)
+blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
+that differ from the ``do'' notation in two ways. First, they support wider range of
+constructs, making it possible to wrap existing F\# code within a block without other changes.
+Second, they support other abstractions including monads, applicative functors and monad
+transformers. It would be interesting to see if computation expressions can be extended to
+support programming with computations based on flat/structural indexed comonads.
+
+More lightweight syntax for effectful computation can be obtained using
+techniques that automatically insert the necessary monadic plumbing (without
+special syntax). This has been done for effectful computations \cite{monads-lightweight-ml}
+and a similar approach would be worth exploring for coeffects.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect annotations as types}
+\label{sec:further-impl-types}
+
+The other aspect of practical implementation of coeffects is that of tracking the context requirements
+(coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose
+language feature) the type system needs to be able to capture various kinds of coeffect algebras.
+The structures required in this thesis include sets (with union or intersection), natural numbers (with
+addition, maximum, minimum and multiplication), two-point lattice (for liveness) and free monoids
+(vectors of annotations).
+
+The work \cite{effects-embedding} on embedding effect systems in Haskell demonstrated that the
+recent additions to the Haskell type system provide enough power to implement structures such
+as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful
+future direction for applied coeffects.
+
+In dependently-typed programming languages such as Agda or Idris \cite{other-agda,other-idris},
+the embedding of coeffects can be implemented more directly (as terms implementing sets or lattices
+can be lifted to the type level). However, we believe that coeffect tracking does not require full
+dependent types and can be made accessible in more mainstream languages. Dependent ML
+\cite{types-dependent-ml} provides an interesting example of a language with some dependent typing
+(\eg~computations with integers) which is still close to its non-dependently-typed predecessor ML.
+
+Another approach for embedding computations into the type system has been pioneered by F\#
+\emph{type providers} \cite{app-inforich}. Technically, type providers are compiler extensions
+that generate types based on external data sources or other information in order to provide easy
+access to data or services. A similar approach could be used for embedding \emph{algberas} such
+as coeffect algebras into the type system. An \emph{algebra provider} would then be a library that
+specifies the objects of the algebra, its equational laws and generalization rules for type inference.
+This could provide an easy-to-use way of embedding coeffect tracking in pragmatic languages
+such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
+\cite{types-units-of-measure}; these could be alternatively provided via one such \emph{algebra
+provider}.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Alternative formulation using coeffect lattice}
+\label{sec:further-impl-semilattice}
+
+Compared to typical effect systems based on sets, coeffect systems have richer structure consisting
+of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting
+properties, but it makes working with coeffect annotations difficult -- it complicates tasks
+such as type generalization and constraint solving in type inference. Here, we look at one possible
+simplification of the structure.
+
+\begin{itemize}
+\item As mentioned in Section~\ref{sec:flat-calculus-algebra}, when the monoid $(\C,\cpar,\czero)$ is a
+semi-lattice, the $\cleq$ relation can often be expressed in terms of the $\cpar$ operation
+($\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}$).
+This does not give a definition equivalent to our, but it is a simplification that is consistent
+with all examples in this thesis.
+
+\item Another approach that works for most (but not all) systems we discussed is to require the structure
+$(\C,\cpar,\czip)$ to form a lattice. This is consistent with all our structural coeffect systems as
+well as flat systems for liveness and data-flow, but not with the system for tracking of implicit
+parameters (or sets of rebindable resources).
+\end{itemize}
+
+\noindent
+The system for tracking implicit parameters does not fit the lattice-based model, because it uses the
+same operation $\cup$ for both $\cpar$ and $\czip$.
+
+This also means that the system for implicit parameters cannot be used with a system where lambda
+abstraction duplicates the context requirements (Section~\ref{sec:flat-exts-lambda}).
+
+\paragraph{Lattice-based coeffects.} Putting implicit parameters aside, there are many other coeffect
+systems for which we could use the following definition of coeffect scalars (for simplicity, we only
+consider flat systems):
+
+\begin{definition}
+A \emph{\cclrd{flat coeffect lattice}} $(\C, \cseq, \cleq, \cunit, \czero)$ is a set
+$\C$ together with elements $\cunit, \czero \in \C$ and an ordering $\cleq$ such
+that $(\C, \cseq, \cunit)$ is a monoid and $(\C, \cleq)$ is a lattice with the least
+element $\czero$.
+
+Assuming $\cpar$ and $\czip$ are the \emph{least upper bound} and \emph{greatest lower bound}
+of the lattice, we require the following distributivity axioms:
+%
+\begin{align*}
+\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
+\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
+\end{align*}
+\end{definition}
+
+\noindent
+The definition provides all the operations of a flat coeffect algebra, but it defines a more rigid
+structure. In particular, $\czip$ and $\cpar$ are both defined in terms of $\cleq$. This simplifies
+the system as the operations of the underlying indexed comonad structure can be expressed using just
+$\cseq$ and $\cleq$:
+%
+\begin{equation*}
+\begin{array}{lcll}
+ \ident{counit}_{\cunit} &\narrow{:}&
+    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
+ \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
+    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
+\ident{merge}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}&
+    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{t}}{(\alpha \times \beta)} &
+    (\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s}) \\
+\ident{split}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}&
+    \ctyp{\cclrd{t}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} &
+    (\cclrd{r}\,\cleq\,\cclrd{t}, \cclrd{s}\,\cleq\,\cclrd{t}) \\
+\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}&
+    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} &
+    (\cclrd{s}\,\cleq\,\cclrd{r})
+\end{array}
+\end{equation*}
+%
+Note that we still need \emph{two distinct} operators in the system. Analysis of our examples shows
+that the operator used for sequential composition $\cseq$ is not necessarily related in any way to
+the ordering on the lattice provided by $\cleq$.
+
+However, the operations $\ident{merge}$, $\ident{split}$ and $\ident{lift}$ can now all
+use ordering. For example, the \ident{merge} operation previously returned
+a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
+element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit sub-coeffecting
+to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and
+$\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
+This means that we can compose the original \ident{merge} operation which returns greatest lower
+bound with $\ident{lift}$ to get the (less precise) operation required here.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+{\small a.) Lattice-based flat systems (selected rules):}
+
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x:\tau_1}{\cclrd{t}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
+~~(\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s})
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
+   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
+  {\coctx{\Gamma}{\cclrd{u}} \vdash e_1~e_2 : \tau_2}
+~~(\cclrd{r}\,\cleq\,\cclrd{u}, (\cclrd{s} \,\cseq\, \cclrd{t})\,\cleq\,\cclrd{u})
+\end{equation*}
+
+\vspace{1.5em}
+{\small b.) Lattice-based structural systems (selected rules):}
+
+\begin{equation*}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{u}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
+~~(\cclrd{s}\,\cleq\,\cclrd{u}, \cclrd{t}\,\cleq\,\cclrd{u})
+\end{equation*}
+
+\figcaption{Flat and structural coeffects using lattice-based formulation}
+\label{fig:further-lattice-types}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Lattice-based type system.}
+\hspace{-1em}
+Using lattice-based coeffect structure leads to the typing rules shown in Figure~\ref{fig:further-lattice-types}.
+In the flat coeffect system, this affects both application and abstraction -- in abstraction,
+the context-requirements of the body should be smaller than the context available on the
+call site and declaration site. In application, we require more than what is required by
+the sub-expressions.
+
+Structural coeffect systems only require least upper bound $\cpar$ in the (\emph{contr}) rule.
+This can be rewritten similarly to the flat (\emph{app}) rule --  when contracting two variables, the
+context requirements of the resulting variable are greater than the requirements associated with
+each of the original variables.
+
+In summary, the lattice-based coeffect system has some advantages over the design used in this
+thesis. In particular, it uses better-known structures and is simpler. However, we do not use
+it as it rules out one of the important motivating examples (implicit parameters) and, even with
+the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.
+
+% =================================================================================================
+%
+%    #   #          #                    ##
+%    #   #          #                     #
+%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###
+%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   #
+%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    #####
+%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #
+%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###
+%                                                            #   #                #   #
+%                                                             ###                  ###
+% =================================================================================================
+
+\section{Coeffect meta-language}
+\label{sec:further-meta}
+
+In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
+language semantics as introduced by Moggi \cite{monad-notions}. The first approach is to use
+monads in the \emph{semantics} of an effectful language. The second approach is to extend the
+language with (additional) monadic constructs that can then be used for writing effectful
+monads explicitly.
+
+In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
+the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
+indexed comonads to give the semantics for the language and to derive type system for it.
+
+In this section, we briefly discuss the other technique. That is, we embed indexed
+comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
+context (semantically, this corresponds to an indexed comonad) and we add language constructs
+that correspond to the operations of indexed comonads.
+
+This section provides a brief sketch of the coeffect meta-language in order to highlight the relationship
+between coeffects and closely related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
+Developing the system further is an interesting future research direction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffects and contextual modal type theory}
+
+As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
+to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
+meta-languages that include $\square$ as a type constructor
+\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
+Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
+From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
+coeffect calculus.
+
+\paragraph{Context in CMTT and coeffects.}
+Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
+comonads (the $\square$ modality) into the meta-language, there are two important differences.
+
+Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
+requirements are formed by an abstract coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, etc.
+
+Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
+the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+constructor refers to a value of type $\tau$ \emph{together} with additional context specified
+by the coeffect annotation $\cclrd{r}$ (\eg~a list of past values or additional resources).
+
+In contrast in CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the
+original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the
+context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the
+two operations of a comonad:
+\begin{equation*}
+\begin{array}{rcl}
+ \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
+ \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow
+    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
+\end{array}
+\end{equation*}
+Two readings of the signatures are possible, but they give quite different meanings to the operations:
+
+\begin{itemize}
+\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value
+  and does not require any additional context; the \ident{cobind} operation requires
+  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to
+  evaluate the function, ending with a value $\beta$ together with remaining context
+  $\cclrd{s}$.
+\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation
+  that requires no additional context to obtain a $\alpha$ value; given a function that
+  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
+  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$
+  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
+  context requirements is eliminated by the function).
+\end{itemize}
+
+\noindent
+Although the different reading does not affect formal properties of the systems, it is important
+to understand the difference when discussing the two systems as they provide a different
+intuition.
+
+The sketch of a coeffect meta-language in the following section attempts to bridge the gap
+between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
+it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
+sets of variables. Future work on the coeffect meta-language would thus be an interesting
+development for both coeffect systems and CMTT.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+The coeffect meta-language could be designed using both flat and structural indexed comonads.
+For simplicity, this section only discusses the flat variant. The syntax of types and terms
+of the language includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
+language constructs:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
+ e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
+  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
+  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
+  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2
+\end{array}
+\end{equation*}
+%
+The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
+operation of the comonad. To define meta-language for flat indexed comonads, we also include
+\kvd{split} and \kvd{merge} that embed the corresponding operations.
+
+\paragraph{Types for coeffect meta-language.}
+The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
+shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing
+rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
+
+The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
+a value from a comonadic context and corresponds to variable access in coeffect calculi. The
+\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
+and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
+The result is a computation that produces a $\tau_2$ value with the remaining context specified
+by $\cclrd{s}$. Note that the expressions $e_2$ and $e_1$ correspond to the first and second
+arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following
+CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
+although it differs because of our generalization of comonads where \ident{bind} composes
+coeffect annotations rather than requiring the same annotation everywhere.}.
+
+The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some
+transformation on one or two values in a context and then add the new value as a fresh variable
+to the variable context. We omit sub-coeffecting, but it could be easily added following
+the method used elsewhere.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Typing rules for the simply typed $\lambda$-calculus}
+
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\Gamma \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
+   \Gamma \vdash e_2 : \tau_2 }
+  {\Gamma \vdash e_1~e_2 : \tau_1 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x:\tau_1 \vdash e : \tau_2}
+  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
+
+\begin{equation*}
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
+\end{equation*}
+\begin{equation*}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash !e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{split}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
+   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{merge}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
+   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
+   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
+  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
+\end{equation*}
+
+\figcaption{Type system for the (flat) coeffect meta-language}
+\label{fig:conclusions-cml}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding flat coeffect calculus}
+
+The \emph{meta-language} approach of embedding comonads in a language is more general than the
+\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
+a type system for context-aware programming languages.
+
+However, it is worth demonstrating that the (flat) coeffect calculus can be embedded in the
+meta-language described in the previous section. This may be desirable, \eg~when using the
+meta-language for reasoning about context-aware computations. We briefly consider the embedding
+as it illuminates the relationship between coeffects and CMTT (unfortunately it is not possible
+to embed coeffect calculi in CMTT because of the more general annotations structure).
+
+Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
+meta-language. Note that the translation is indexed by $v$, which is a name of variable used
+to represent the entire variable context of the source language. The translation is defined in
+Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in
+Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+the operations of an indexed comonad.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{l}
+ \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
+   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
+   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
+      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
+   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
+\end{array}
+\end{equation*}
+\figcaption{Embedding flat coeffect calculus in coeffect meta-language }
+\label{fig:conclusions-embed}
+\end{figure}
 
@@
 %
-%    #######
-%       #    #    # ######  ####  #####  #   #
-%       #    #    # #      #    # #    #  # #
-%       #    ###### #####  #    # #    #   #
-%       #    #    # #      #    # #####    #
-%       #    #    # #      #    # #   #    #
-%       #    #    # ######  ####  #    #   #
+%        ###                         ##                    #
+%       #   #                         #
+%       #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###
+%       #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #
+%       #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###
+%       #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      #
+%        ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####
 %
@@
 
-\section{From theory to implementation}
-\label{sec:impl-theory}
+\section{Conclusions}
 
-The theory discussed so far provides the two key components of the implementation. In
-Chapter~\ref{ch:flat}, we discussed the type checking of context-aware programs and
-Chapter~\ref{ch:semantics} models the execution of context-aware programs (in terms of
-translation and operational semantics). For structural coeffects, the same components are discussed
-in Chapter~\ref{ch:structural}. In this section, we discuss how those provide foundation for the
-implementation.
+In this chapter, we completed our quest of presenting a \emph{unified} system for tracking contextual
+properties and then we discussed two important topics of future and related work that require
+in-depth discussion.
 
-\subsection{Type checking}
-\label{sec:impl-theory-typing}
+The key technical contribution of this chapter is the \emph{unified coeffect system}
+(Section~\ref{sec:further-unified}), which unifies the flat and structural system presented
+in the previous two chapters. To achieve this, we introduced \emph{coeffect container}, which
+determines how are coeffect annotations attached to variable contexts. We then discussed two
+instances of the structure that capture flat and structural properties.
 
-(Section~\ref{sec:flat-unique})
-
-\paragraph{Implementation.}
-From the presentation in this section, it might appear that resolving the ambiguity related to
-lambda abstraction for implicit parameters requires a type system that is quite different from
-the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
-case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are
-simpler.
-
-Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets
-of implicit parameters to satisfy the constraints. We still need to track implicit parameters in
-scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
-to the one from Figure~\ref{fig:flat-types}:
-%
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
-  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
-    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}}
-\end{equation*}
-%
-Given a typing derivation for the body that produced constraints $C$, we generate an additional
-constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
-current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds
-the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.
-
-\paragraph{Implementation.}
-As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
-require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
-by (\emph{idabs}) can be easily obtained by generating additional constraints:
-%
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
-  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
-    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}}
-\end{equation*}
-%
-Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
-coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by
-(\emph{idabs}).
-
-Type checking
- * DSL specific constraint resolution
- * easier for structural
-
-\subsection{Translation}
-\label{sec:impl-theory-transl}
-
-Translation ~ Semantics
-
-\subsection{Extensions}
-\label{sec:impl-theory-ext}
-
-It is really easy to add another DSL
-
-\section{Case studies}
-\label{sec:impl-case}
-
-\subsection{Typing}
-\label{sec:impl-case-typing}
-
-let both =
-  let ?fst = 100 in
-  fun trd -> ?fst + ?snd + trd in
-let ?fst = 200 in
-both 1
-
-num -{ ?snd:num }-> num
-
-let struct x y = x + prev y in
-struct
-
-num -{ 1 }-> num -{ 1 }-> num
-
-let flat x y = x + prev y in
-flat
-
-num -{ 0 }-> num -{ 1 }-> num
-
-\subsection{Translation}
-\label{sec:impl-case-transl}
-
-let ?param = 10 in
-fun x -> ?param + ?other
-
-let (ctx2, ctx3) = split (duplicate finput) in
-let ctx1 = letimpl ?param (ctx2, 10) in
-fun x ->
-  let ctx4 = merge (x, ctx1) in
-  let (ctx5, ctx6) = split (duplicate ctx4) in
-  lookup ?param ctx5 + lookup ?other ctx6
-
-fun x -> (fun v -> prev v) (prev x)
-
-fun x ->
-  let ctx1 = merge (x, sinput) in
-  (fun v ->
-    let ctx3 = merge (v, choose_0 ctx1) in
-    counit (prev ctx3)) (cobind (fun ctx2 ->
-      counit (prev ctx2)) (choose_1 ctx1))
-
-\section{Technical details}
-\label{sec:impl-tech}
-
-1. F# + JavaScript
-2. Single AST and parsing
-3. Type checking and constraint generation
-4. Constraint resolution
-
-\section{Interactive essay}
-\label{sec:impl-essay}
+In the rest of the chapter, we discussed two topics of future and related work. First
+(Section~\ref{sec:further-impl}), we considered pathways to practical implementations of
+coeffect systems, including a discussion of a novel lattice-based coeffect system, which is simpler,
+but cannot capture all our motivating examples. Finally, we discussed how our work relates to
+meta-language based on comonads (Section~\ref{sec:further-meta}). We present a \emph{coeffect
+meta-language} that follows similar style to CMTT, but is based on indexed comonads. Next, we
+discussed how to embed flat coeffect systems in this coeffect meta-language, which elucidates the
+relationship between our work and CMTT.

### 2016-05-11 text/Chapter09.tex
Add binding to chapter 2, whitespace, tweaks

--- 2014-12-16 text/Chapter07.tex
@@
-%!TEX root = ../main.tex
-
 \chapter{Conclusions and further work} 
@@
 ~ The second category of context-aware systems discussed in Section~\ref{sec:applications-structural}
-captures \emph{per-variable} contextual properties. The systems discussed here resemble sub-structural
+captures \emph{per-variable} contextual properties. The systems discussed here resemble substructural
 logics, but rather than \emph{restricting} variable use, they \emph{track} how the variables are used.
@@
 coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
-to coeffect systems, categorical semantics of coeffects, and connections with sub-structural type 
+to coeffect systems, categorical semantics of coeffects, and connections with substructural type
 systems and bunched logics.

### 2016-05-12 text/AppendixA.tex
Context-aware systems edits (Chapter 3)

@@
 see, the fact that both branches require the same number of past values means that we need to use
-the sub-coeffecting rule.
+the subcoeffecting rule.
 
@@
 we derive the types for the sub-expressions of the application and the conditional. Note that 
-to obtain compatible function types, we use sub-coeffecting \emph{before} abstraction in the
+to obtain compatible function types, we use subcoeffecting \emph{before} abstraction in the
 typing of $(\lambda x.x)$ -- the type $\ident{int} \xrightarrow{\cclrd{1}} \ident{int}$ is  a

### 2016-05-12 text/AppendixB.tex
Context-aware systems edits (Chapter 3)

@@
 \item If $y\neq x$ then $\subst{y}{x}{e_s} = y$. Using the fact that $\cunit$ is the bottom element
-and sub-coeffecting:
+and subcoeffecting:
 \[
@@
 \paragraph{(sub)} As in the (\emph{contr}) case, in the (\emph{sub}) case we distinguish two situations.
-If the sub-coeffecting is applied to variable that is \emph{not} being substituted for, then the case is
-easy (sub-coeffecting does not interact with substitution), so we only consider the case when $x$ is one
+If the subcoeffecting is applied to variable that is \emph{not} being substituted for, then the case is
+easy (subcoeffecting does not interact with substitution), so we only consider the case when $x$ is one
 of the variables being substituted for:
@@
 \]
-The derivation first applies the standard (\emph{weak}) rule and then uses sub-coeffecting
+The derivation first applies the standard (\emph{weak}) rule and then uses subcoeffecting
 rule and the property $\czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})$ to obtain conclusion of the required form.

### 2016-05-12 text/Chapter03.tex
Context-aware systems edits (Chapter 3)

@@
-% ==================================================================================================
-
 \chapter{Context-aware systems}
@@
 
+This chapter reviews some of the existing context-aware programming abstractions and presents them
+in a uniform way. We start with disconnected examples, but at the end, we will see that they
+share a common pattern\footnote{The different properties captured by monads may appear similarly
+disconnected at first!}.
+
 \paragraph{Chapter structure and contributions}
 \begin{itemize}
-\item We start with a characterization of contextual properties. The Section~\ref{sec:applications-structure}
+\item We characterize contextual properties -- the Section~\ref{sec:applications-structure}
   explains what is a \emph{coeffect} and contrasts it with a better known notion of
@@
   between \emph{flat coeffect} systems (Section~\ref{sec:applications-flat}) and \emph{structural coeffect}
-  systems (Section~\ref{sec:applications-structural}). The fact that we find a common structure in
-  all systems presented here lets us develop unified coeffect calculi in the upcoming three chapters.
+  systems (Section~\ref{sec:applications-structural}). This common structure is precisely
+  captured by the two \emph{coeffect calculi} in the upcoming chapters.
 
@@
 \end{itemize}
-
-\noindent
-As mentioned, this chapter may appear as a collection of disconnected
-examples\footnote{The different properties captured by monads may appear similarly
-disconnected at first!}. But at the end, we will see that they share a common pattern.
 
@@
 
-In standard $\lambda$-calculus, variable access can be seen as a primitive operation that
-accesses the context. The variable access expression introduces a context requirement -- the
-expression $x$ is typeable only in a context that contains $x:\tau$ for some type $\tau$.
-
-The standard lambda abstraction (\emph{pure}), shown in Figure~\ref{fig:applications-abs},
-splits the free-variable context of an expression into two parts. At runtime, the value of the parameter
-has to be provided by the \emph{call site} (dynamic scope) and the remaining values are provided
-by the \emph{declaration site} (lexical scope). In the type checking, the splitting is determined
-syntactically -- the notation $\lambda x.e$ names the variable whose value comes from the call site.
+As discussed in Section~\ref{sec:path-binding}, variable access can be seen as a basic form
+of context requirement in $\lambda$-calculus. The expression $x$ is typeable only in a context
+that contains $x:\tau$ for some type $\tau$.
+
+In lexically scoped languages, lambda abstraction (\emph{pure}), as shown in
+Figure~\ref{fig:applications-abs}, splits the free-variable context of an expression into two parts.
+At runtime, the value of the parameter has to be provided by the \emph{call site} (dynamic scope)
+and the remaining values are provided by the \emph{declaration site} (lexical scope). In the type
+checking, the splitting is determined syntactically. The notation $\lambda x.e$ \emph{names} the
+variable whose value comes from the call site.
 
@@
 \emph{split} between the call site and declaration site. Furthermore, there is no syntactic
-structure that determines how the requirements are split. In the case of distributed
-programming, the resources can be freely associated with either of the two sites.
+structure that determines how the requirements are split. As mentioned in Section~\ref{sec:path-binding-amb},
+we decouple the definition of semantics from the domain-specific choice that determines how
+context requirements are satisfied. We capture the choice in the type and give semantics over a
+\emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
+example, by preferring resources available on the client over resources available on the server.
 
@@
 $\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
-The unification is discussed in Chapter~\ref{ch:unified}.
+The unification is outlined in Section~\ref{ch:further-unified}.
 
@@
 a value for $y$, which needs to be accessible at least within $(-1, 1)$. In this way, structural
-coeffects remove the non-determinism arising from the splitting of requirements in flat coeffect systems.
+coeffects remove the ambiguity arising from the splitting of requirements in flat coeffect systems.
 
@@
 $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$ with a product on the left, but a single value
-on the right. In both effect systems and coeffect systems, we \emph{write} the annotation as part of
+on the right. In both effect systems and coeffect systems, we write the annotation as part of
 the function arrow. However, in the underlying categorical model, effects are attached to the result
@@
 
-Structural coeffects have one coeffect annotation per each variable. Thus, the annotation consists
+Structural coeffects have one annotation per each variable. Thus, the annotation consists
 of multiple values -- one belonging to each variable. To distinguish between the overall annotation
 and individual (per-variable) annotations, we call the overall coeffect a \emph{vector} consisting of
-\emph{scalar} coeffects. This asymmetry also explains why coeffect systems are not trivially dual to
-effect systems.
+\emph{scalar} coeffects. This asymmetry also explains why coeffects are not trivially dual to
+effects.
 
@@
 over $\C$ is a tuple $\alift{ \cclrd{r_1}, \ldots, \cclrd{r_n} }$ of scalars.
-We use bold face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for vectors and
-normal face $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars\footnote{For better readability, the thesis
-also distinguishes different structures using colours. However ignoring the colour does not introduce
-any ambiguity.}. We also say that a \emph{shape} of a vector $\slen{\aclrd{\textbf{r}}}$ (or more generally any container)
+We use bold face letters like $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ for
+vectors and normal face $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars. We also say that a
+\emph{shape} of a vector $\slen{\aclrd{\textbf{r}}}$ (or more generally any container)
 determines the set of \emph{positions} in a vector. So, a vector of a shape (length) $n$ has positions
-$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Chapter~\ref{ch:unified} and also
+$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Section~\ref{ch:further-unified} and also
 discuss how our use relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.
@@
 $\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
-Finally, we often concatenate vectors -- for example, when joining two variable contexts.
+Finally, we often concatenate vectors, for example, when joining two variable contexts.
 Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
@@
 
-For three of the examples, we present a type system and a simple semantics. Although the examples
+For three of the examples, we present a type system and a simple semantics (given inductively over
+the typing derivation). We informally discuss how preferred typing derivation is chosen (to resolve
+the inherent ambiguity), but leave details to later chapters. Although the examples
 are not new, our novel presentation of the systems (and the fact that they appear side-by-side)
@@
 \paragraph{Implicit parameters.}
-Implicit parameters are a special kind of variables that support dynamic scoping.
-They make it possible to parameterise a computation (involving a long chain of function calls)
-without passing parameters explicitly as additional arguments of all involved functions.
+As discussed in Section~\ref{sec:path-binding-impl}, implicit parameters are a special kind of
+variables that support dynamic scoping. They make it possible to parameterise a computation
+(involving a long chain of function calls) without passing parameters explicitly as additional
+arguments of all involved functions.
 
@@
 
-A simple language with implicit parameters has an expression $\ident{?param}$ to read a parameter
-and an expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a
+A simple language with support for implicit parameters has an expression $\ident{?param}$ to read a
+parameter and an expression\footnote{Haskell uses $\kvd{let}~\ident{?p} = e_1~\kvd{in}~e_2$, but we use a
 different keyword to avoid confusion.} $\kvd{letdyn}~\ident{?param} = e_1~\kvd{in}~e_2$ that sets a
@@
 $\ident{?size}$:
-%
+
 \begin{equation*}
@@
 The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context
-$\{ \ident{?width} : \ident{int} \}$. The nested function (returned as a result) uses the parameter
+$\{ \ident{?width} \}$. The nested function (returned as a result) uses the parameter
 $\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters used by the
@@
 write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
-In a purely dynamically scoped system, they would have to be defined when the user invokes the nested function.
+We also assume that are implicit parameters are of type $\ident{num}$, so the annotation can be a
+simple set of names (rather than mapping from names to types). In a purely dynamically scoped system,
+implicit parameters would have to be defined when the user invokes the nested function.
 However, implicit parameters behave as a combination of lexical and dynamic scoping. This means
@@
 \tag{\emph{dynamic}}
-\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
-  (\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int}, \ident{?size} : \ident{int} \} } \ident{string})
+\ident{string} \xrightarrow{ \{ \ident{?width} \} }
+  (\ident{string} \xrightarrow{ \{ \ident{?width}, \ident{?size} \} } \ident{string})
 \end{equation}
@@
 \tag{\emph{mixed}}
-\ident{string} \xrightarrow{ \{ \ident{?width} : \ident{int} \} }
-  (\ident{string} \xrightarrow{ \{ \ident{?size} : \ident{int} \} } \ident{string})
+\ident{string} \xrightarrow{ \{ \ident{?width} \} }
+  (\ident{string} \xrightarrow{ \{ \ident{?size} \} } \ident{string})
 \end{equation}
@@
 This is not a complete list of possible typings, but it demonstrates the options. The (\emph{dynamic})
-case requires the parameter \ident{?width} twice (this may be confusing when the caller provides
-two different values). In the (\emph{mixed}) case, the nested function captures the \ident{?width}
-parameter available from the declaration site. Using the latter typing, the function can be called as follows:\
+case requires the parameter \ident{?width} twice (the caller may provide different value, in which
+case, the semantics needs to specify which value is preferred). In the (\emph{mixed}) case, the
+nested function captures the \ident{?width} parameter available from the declaration site. Using
+the latter typing, the function can be called as follows:
 %
@@
 For different typings of \ident{format}, different ways of calling it are valid. This illustrates
-the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect systems may
-introduce certain non-determinism in the typing. The following section shows how this looks in the
-type system for implicit parameters.
+the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect programs have
+multiple typing derivations and the semantics depends on the domain-specific choice of preferred
+typing. The following section shows how this looks in the type system for implicit parameters.
 
@@
 finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
-\ident{Names} \mapsto \ident{Types}$. The expressions include \ident{?param} to read implicit
+\ident{Names}$. The expressions include \ident{?param} to read implicit
 parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
@@
 introduces primitive context requirements is (\emph{param}). Accessing a parameter \ident{?param}
-of type $\tau$ requires it to be available in the context. The context may provide more (unused)
-implicit parameters thanks to the (\emph{sub}) rule.
+requires it to be available in the context. The context may provide more (unused)
+implicit parameters thanks to the subcoeffecting rule (\emph{sub}).
 
@@
 the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
-and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$). Finally, (\emph{letdyn}) defines an implicit
-parameter and removes it from the set of requirements.
+and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$) and thus an expression may have multiple
+valid typing derivations. Finally, (\emph{letdyn}) removes the bound parameter from the set of requirements.
 
@@
 
-\begin{figure}
-
-{\small The semantics is defined inductively over the typing derivation:}
-\begin{equation}
-\tag{\emph{var}}
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash x_i : \tau_i} = \lambda ((x_1, \ldots, x_n), \_) \rightarrow x_i
-\end{equation}
-\vspace{-1.5em}
-\begin{equation}
-\tag{\emph{param}}
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash ?p : \sigma} = \lambda (\_, f) \rightarrow f~?p
-\end{equation}
-\vspace{-1.5em}
-\begin{equation}
-\tag{\emph{sub}}
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau} = \lambda (x, f) \rightarrow
-  \sem{\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}~(x, \restr{f}{\cclrd{r'}})
-\end{equation}
-\vspace{-1.5em}
-\begin{equation}
-\tag{\emph{abs}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } = \lambda ((x_1, \ldots, x_n), f) \rightarrow\\[-0.25em]
-  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{r}\cup\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), f \uplus g)
-\end{array}
-\end{equation}
-\vspace{-1.0em}
-\begin{equation*}
-\tag{\emph{app}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{r}\cup\cclrd{s}\cup\cclrd{t}} \vdash e_1~e_2 : \tau_2 } = \lambda (x, f) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~(x, \restr{f}{\cclrd{r}})\\[-0.25em]
-  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~(x, \restr{f}{\cclrd{s}} ), \restr{f}{\cclrd{t}})
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation}
-\tag{\emph{letdyn}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{r \cup (s \setminus \{ \ident{?p}:\tau_1 \})}} \vdash \kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2 } =
-     \lambda (x, f) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~v=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 }~(x, \restr{f}{\cclrd{r}})\\[-0.25em]
-  \qquad\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2 }~
-     (x, \restr{f}{\cclrd{s} \setminus \cclrd{\{ \ident{?p}:\tau_1 \}}} \uplus \{ \ident{?p} \mapsto v \} )
-\end{array}
-\end{equation}
+\begin{figure}[t]
+\begin{equation*}
+\hspace{0em}
+\begin{array}{ll}
+\hspace{4.5em}\semdef
+  {\coctx{\Gamma}{\cclrd{r}} \vdash x_i : \tau_i}
+  {\lambda ((x_1, \ldots, x_n), \_) . x_i}
+& (\emph{var})
+\\[0.0em]
+\hspace{3.5em}\semdef
+  {\coctx{\Gamma}{\cclrd{r}} \vdash \ident{?p} : \ident{num}}
+  {\lambda (\_, f) . f~\ident{?p}}
+& (\emph{param})
+\\[1.5em]
+\hspace{5em}\semdeff
+  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
+  {f}
+  {\lambda (x, g) . f~(x, \restr{g}{\cclrd{r'}})}
+& (\emph{sub})
+\\[1.5em]
+\hspace{0em}\emdeff
+  {\coctx{\Gamma,y:\tau_1}{\cclrd{r}\cup\cclrd{s}} \vdash e : \tau_2}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}\\[-0.25em]~\end{array}}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda ((x_1, \ldots, x_n), g_1) . \lambda (y, g_2) .\\[-0.25em]
+  \quad f~((x_1, \ldots, x_n, y), g_1 \uplus g_2)
+  \end{array}\hspace{-1em}~}
+& (\emph{abs})
+\\[1.5em]
+\hspace{-0.8em}\emdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r}\cup\cclrd{s}\cup\cclrd{t}} \vdash e_1~e_2 : \tau_2}\\[-0.25em]~\end{array}}
+  {f_1}
+  {f_2}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda (x, g) . \\[-0.25em]
+  \quad(f_1~(x, \restr{g}{\cclrd{r}}))~(f_2~(x, \restr{g}{\cclrd{s}} ), \restr{g}{\cclrd{t}})
+  \end{array}\hspace{-2em}~}
+& (\emph{app})
+\\[1.5em]
+\hspace{-3.3em}\emdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \ident{num}}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r \cup (s \setminus \{ \ident{?p} \})}} \vdash\\[-0.25em]
+    \qquad\kvd{letdyn}~\ident{?p}=e_1~\kvd{in}~e_2 : \tau_2}\end{array}}
+  {f_1}
+  {f_2}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda (x, g) . f_2~(x, \restr{g}{\cclrd{s} \setminus \cclrd{\{ \ident{?p} \}}} \,\uplus \\[-0.25em]
+    \qquad \{ \ident{?p} \mapsto (f_1~(x, \restr{g}{\cclrd{r}})) \} )
+  \end{array}\hspace{-1em}~}
+& (\emph{letdyn})
+\end{array}
+\end{equation*}
 
 \vspace{1em}
-{\small Here $\uplus$ and $\restr{f}{\cclrd{r}}$ are auxiliary definitions:}
+Assuming the following auxiliary definitions:
 \begin{equation*}
@@
 \label{fig:applications-flat-implsem}
+\vspace{-0.5em}
 \end{figure}
@@
 but simply to provide a better explanation. A detailed semantics in terms of indexed comonads is
-shown in Chapter~\ref{ch:flat}.
-
-For simplicity, we assume that all implicit parameters have a type $\sigma$. In that setting,
-coeffect annotations $\cclrd{r}$ are just sets of names, \ie~$\cclrd{r}, \cclrd{s}, \cclrd{t} \subseteq \ident{Names}$.
+shown in Chapter~\ref{ch:semantics}.
+
 Given an expression $e$ of type $\tau$ that requires free variables $\Gamma$ and implicit parameters
@@
 with a dictionary of implicit parameters and returns $\tau$:
-
+%
 \begin{equation*}
 \sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau}
-  ~:~ (\tau_1 \times \ldots \times \tau_n) \times (\cclrd{r} \rightarrow \sigma) \rightarrow \tau
-\end{equation*}
-%
-The dictionary is represented as a function from $\cclrd{r}$ to $\sigma$. This means that it
-provides a $\sigma$ value for all implicit parameters that are required according to the typing.
+  ~:~ (\tau_1 \times \ldots \times \tau_n) \times (\cclrd{r} \rightarrow \ident{num}) \rightarrow \tau
+\end{equation*}
+%
+The dictionary is represented as a function from $\cclrd{r}$ to $\ident{num}$. This means that it
+provides a $\ident{num}$ value for all implicit parameters that are required according to the typing.
 Note that the domain of the function is not the set of all possible implicit parameter names, but
@@
 \begin{array}{l}
- \sem{\tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \tau_1 \times (\cclrd{s} \rightarrow \sigma) \rightarrow \tau_2
-\end{array}
-\end{equation*}
-%
-The definition of the semantics is shown in Figure~\ref{fig:applications-flat-implsem}. The
-(\emph{var}) and (\emph{param}) rules are simple -- they project the appropriate variable and
-implicit parameter, respectively.
-
-When an expression requires implicit parameters $\cclrd{r}$, the semantics always provides a
-dictionary defined \emph{exactly} on $\cclrd{r}$. To achieve this, the (\emph{sub}) rule restricts
-the function to $\cclrd{r'}$ (which is valid because $\cclrd{r'} \subseteq \cclrd{r}$).
+ \sem{\tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \tau_1 \times (\cclrd{s} \rightarrow \ident{num}) \rightarrow \tau_2
+\end{array}
+\end{equation*}
+%
+The definition of the semantics is shown in Figure~\ref{fig:applications-flat-implsem}. We use a
+notation that emphasizes the fact that the semantics is given over a typing derivation. On the
+left-hand side of $=$, we show the applied typing rule. The right-hand side of $=$ then shows the
+semantic functions assigned to the individual assumptions and the resulting semantics for the
+consequent.
+
+The (\emph{var}) and (\emph{param}) rules are simple -- they project the appropriate variable and
+implicit parameter, respectively. When an expression requires implicit parameters $\cclrd{r}$, the
+semantics always provides a dictionary defined \emph{exactly} on $\cclrd{r}$. To achieve this, the
+(\emph{sub}) rule restricts the function to $\cclrd{r'}$ (which is valid because $\cclrd{r'} \subseteq \cclrd{r}$).
 
 The most interesting rules are (\emph{abs}) and (\emph{app}). In abstraction, we get two dictionaries
-$f$ and $g$ (from the declaration site and call site, respectively), which are combined and passed
+$g_1$ and $g_2$ (from the declaration site and call site, respectively), which are combined and passed
 to the body of the function. The semantics prefers values from the call site, which is captured by
@@
 examining the semantic rules (and noting that the restriction and union always provide the
-expected set of parameters).
+expected set of parameters). This idea is captured more formally by the soundness proof for the
+operational semantics given in Chapter~\ref{ch:semantics}.
 
 \paragraph{Monadic semantics.}
-Implicit parameters are related to the \emph{reader mo\-nad}. The type
-$\tau_1 \times (\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2$ is equivalent to
-$\tau_1 \rightarrow ((\cclrd{r} \rightarrow \sigma) \rightarrow \tau_2)$ through currying. Thus, we can
-express the function as $\tau_1 \rightarrow M\tau_2$ for $M\tau = (\cclrd{r} \rightarrow \sigma) \rightarrow \tau$.
+Implicit parameters are related to the \emph{reader monad}. The type
+$\tau_1 \times (\cclrd{r} \rightarrow \ident{num}) \rightarrow \tau_2$ is equivalent to
+$\tau_1 \rightarrow ((\cclrd{r} \rightarrow \ident{num}) \rightarrow \tau_2)$ through currying. Thus, we can
+express the function as $\tau_1 \rightarrow M\tau_2$ for $M\tau = (\cclrd{r} \rightarrow \ident{num}) \rightarrow \tau$.
 Indeed, the reader monad can be used to model dynamic scoping. However, there is an important distinction
@@
 \begin{equation*}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } =
-      \lambda ((x_1, \ldots, x_n), \_) \rightarrow\\[-0.25em]
-  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{s}} \vdash e : \tau_2 }~((x_1, \ldots, x_n, y), g)
-\end{array}
+\emdeff
+  {\coctx{\Gamma,y:\tau_1}{\cclrd{r}} \vdash e : \tau_2}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{\emptyset}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}\\[-0.25em]~\end{array}}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda ((x_1, \ldots, x_n), \_) . \lambda (y, g) .\\[-0.25em]
+  \quad f~((x_1, \ldots, x_n, y), g)
+  \end{array}\hspace{-1em}~}
 \end{equation*}
@@
 
-As we discuss later in Section~\ref{sec:flat-related-monads}, the reader monad can be extended
+As we discuss later in Section~\ref{sec:semantics-related-monads}, the reader monad can be extended
 to model rebinding. However, later examples in this chapter, such as liveness in
@@
 \begin{equation*}
-\coctx{x\!:\!\alpha}{ \{ \ident{Num}_\alpha \} } \vdash x + x : \alpha
+\coctx{x\!:\!\alpha}{ \cclrd{\{ \ident{Num}_\alpha \}} } \vdash x + x : \alpha
 \end{equation*}
@@
 Implicit parameters are the simplest example of a system where function abstraction does not
-``delay'' all impurities of the body. Here, the term ``delay'' refers to the fact that some
+delay all impurities of the body. Here, the term ``delay'' refers to the fact that some
 implicit parameters may be captured (from the declaration site) at the time when the function is
@@
 is marshalled and sent to another machine, rebindable resources can be handled in two ways.
-First, if the resource is available on the target machine, the parameter may be \emph{rebound} to
-the resource on the new machine. This is captured by the dynamic scoping rule. Second, if the
+If the resource is available on the target machine, the parameter may be \emph{rebound} to
+the resource on the new machine. This is captured by the dynamic scoping rule. If the
 resource is not available on the target machine, the resource is either marshalled or a \emph{remote
@@
 Otherwise, the date and time needs to be obtained from the server too.
+
+~
 
@@
 
-{\small a.) Set-based type system for cross-compilation, inspired by Links \cite{app-distributed-links}}
+a.) Set-based type system for cross-compilation, inspired by Links \cite{app-distributed-links}
 
@@
 
-{\small b.) Version-based type system, inspired by Android API level \cite{app-android-multitarget}}
+b.) Version-based type system, inspired by Android API level \cite{app-android-multitarget}
 
@@
 
-Cross-compilation may seem similar to resource tracking (and thus to the tracking of implicit
-parameters), but it actually demonstrates a couple of new ideas that are important for flat coeffect
-systems. Unlike with implicit parameters, we will not present a specific existing system in this
-section -- instead we briefly look at two examples that let us explore the range of possibilities.
+Cross-compilation is similar to resource tracking (and thus to the tracking of implicit
+parameters), but it demonstrates a couple of new ideas that are important for flat coeffect
+systems. Unlike with implicit parameters, we will not give a full type system in this
+section, but we briefly look at two examples that explore the range of possibilities.
 
@@
 are sets of execution environments, \ie~$\cclrd{r}, \cclrd{s}, \cclrd{t} \subseteq \{ \ident{client},
-\ident{server}, \ident{database} \}$. Sub-coeffecting (\emph{sub}) lets us ignore some of the supported
+\ident{server}, \ident{database} \}$. Subcoeffecting (\emph{sub}) lets us ignore some of the supported
 execution environments; application (\emph{app}) can be only executed in the \emph{intersection} of the
@@
 
-Sub-coeffecting and application are trivially dual to the rules for implicit parameters. We just track
+Subcoeffecting and application are trivially dual to the rules for implicit parameters. We just track
 supported environments using intersection as opposed to tracking required parameters using union.
@@
 the highest level of the levels required by the sub-expressions and the function. The system uses
-yet another variant of lambda abstraction (\emph{abs}) -- the requirements of the body are duplicated
+yet another variant of lambda abstraction (\emph{abs}). The requirements of the body are duplicated
 and placed on \emph{both} the declaration site and the call site.
@@
 rather than on the \emph{language-semantics} style (see~Section~\ref{sec:path-sem-langs}). We
-return to this topic in Section~\ref{sec:unified-meta}.
+return to this topic in Section~\ref{sec:further-meta}.
 
@@
 
+~
+
 \emph{Live variable analysis} (LVA) \cite{app-modern-compiler} is a standard technique in compiler
@@
 variable was used and so we have to mark both the call site and the declaration site as live (this will
-be refined in a structural version of the system).
+be refined in a structural version).
 
@@
 \begin{array}{l}
-\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\
+\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\[-0.25em]
 \quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
@@
 
-\begin{figure}
-{\small a.) The operations of a two-point lattice $\mathcal{L} = \{\ident{L}, \ident{D}\}$
-where $\ident{D} \sqsubseteq \ident{L}$ are defined as:}
-%
-\begin{equation*}
-\begin{array}{rcl}
-\ident{L} \sqcup \ident{L} &=& \ident{L}\\
-\ident{D} \sqcup \ident{L} &=& \ident{D}\\
-\end{array}
-\qquad
-\begin{array}{rcl}
-\ident{L} \sqcup \ident{D} &=& \ident{D}\\
-\ident{D} \sqcup \ident{D} &=& \ident{D}
-\end{array}
-\qquad
-\begin{array}{rcl}
-\ident{L} \sqcap \ident{L} &=& \ident{L}\\
-\ident{D} \sqcap \ident{L} &=& \ident{L}\\
-\end{array}
-\qquad
-\begin{array}{rcl}
-\ident{L} \sqcap \ident{D} &=& \ident{L}\\
-\ident{D} \sqcap \ident{D} &=& \ident{D}
-\end{array}
-\end{equation*}
-
-{\small b.) Sequential composition of (semantic) functions composes annotations using $\sqcup$:}
-\begin{equation*}
-\begin{array}{llll}
-f : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 \qquad\qquad&
-g : \tau_2 \xrightarrow{\cclrd{s}} \tau_3 \qquad\qquad&
-g \circ f : \tau_1 \xrightarrow{\cclrd{r} \sqcup \cclrd{s}} \tau_3 \qquad\qquad& ~
-\\[0.75em]
-f : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 &
-g : \tau_2 \xrightarrow{\cclrd{\ident{L}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_3 & (1)
-\\[-0.10em]
-f : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 &
-g : \tau_2 \xrightarrow{\cclrd{\ident{L}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 & (2)
-\\[-0.10em]
-f : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 &
-g : \tau_2 \xrightarrow{\cclrd{\ident{D}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 & (3)
-\\[-0.10em]
-f : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 &
-g : \tau_2 \xrightarrow{\cclrd{\ident{D}}} \tau_3 &
-g \circ f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 & (4)
-\end{array}
-\end{equation*}
-
-{\small c.) Pointwise composition of (semantic) functions composes annotations using $\sqcap$:}
-\begin{equation*}
-\begin{array}{llll}
-f : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 \qquad\qquad&
-h : \tau_1 \xrightarrow{\cclrd{s}} \tau_3 \qquad\qquad&
-\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{r} \sqcap \cclrd{s}} \tau_2 \times \tau_3 \qquad&~
-\\[0.75em]
-f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 &
-h : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 \times \tau_3 & (1)
-\\[-0.10em]
-f : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 &
-h : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 \times \tau_3 & (2)
-\\[-0.10em]
-f : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 &
-h : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 \times \tau_3 & (3)
-\\[-0.10em]
-f : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 &
-h : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_3 &
-\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 \times \tau_3 & (4)
-\end{array}
-\end{equation*}
-
-\figcaption{Liveness annotations with sequential and pointwise composition}
-\label{fig:applications-flat-livealg}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \begin{figure}[t]
@@
   {x : \tau \in \Gamma}
-  {\coctx{\Gamma}{\cclrd{\ident{L} }} \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{const}
-  {c : \tau \in \Delta}
-  {\coctx{\Gamma}{ \cclrd{\ident{D}} } \vdash c : \tau }
+  {\coctx{\Gamma}{\cclrd{\cclrd{\ident{L}} }} \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{num}
+  {~}
+  {\coctx{\Gamma}{ \cclrd{\cclrd{\ident{D}}} } \vdash n : \ident{num} }
 \end{equation*}
@@
   {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \sqsubseteq \cclrd{r})
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad\quad(\cclrd{r'} \,\cclrd{\sqsubseteq}\, \cclrd{r})
 \end{equation*}
@@
    \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{t})} \vdash e_1~e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{r} \,\cclrd{\sqcup}\, (\cclrd{s} \,\cclrd{\sqcap}\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 A type system that captures whole-context liveness annotates the context with value of a
-two-point lattice $\mathcal{L} = \{ \ident{L}, \ident{D} \}$. The annotation \ident{L} marks
-the context as \emph{live} and \ident{D} stands for a \emph{dead} context.
-Figure~\ref{fig:applications-flat-livealg} (a) defines the ordering $\sqsubseteq$, meet $\sqcup$ and join
-operations $\sqcap$ of the lattice.
+two-point lattice $\mathcal{L} = \{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$. The annotation \cclrd{\ident{L}} marks
+the context as \emph{live} and \cclrd{\ident{D}} stands for a \emph{dead} context.
+Figure~\ref{fig:applications-flat-livealg} (a) defines the ordering $\cclrd{\sqsubseteq}$, meet $\cclrd{\sqcup}$ and join
+operations $\cclrd{\sqcap}$ of the lattice.
 
 The typing rules for tracking whole-context liveness are shown in Figure~\ref{fig:applications-flat-liveness}.
-The language now includes constants $c:\tau \in \Delta$. Accessing a constant (\emph{const}) annotates
-the context as dead using \ident{D}. This contrasts with variable access (\emph{var}), which marks the
-context as live using \ident{L}. A dead context (definitely not needed) can be treated as live context
+The language now includes numerical constants $n$. Accessing a constant (\emph{num}) annotates
+the context as dead using \cclrd{\ident{D}}. This contrasts with variable access (\emph{var}), which marks the
+context as live using \cclrd{\ident{L}}. A dead context (definitely not needed) can be treated as live context
 using the (\emph{sub}) rule. This captures the \emph{may} nature of the analysis.
@@
 as the result of $e_1$. However, we need more than just sequential composition. The same input
-context is passed to the expression $e_1$ (in order to get the function value) and to the
-sequentially composed function (to evaluate $e_2$ followed by the function call). This is captured
-by \emph{pointwise composition}.
+context is passed to the expression $e_1$ (in order to get the function value) and to a function
+obtained by sequential composition (first evaluate the argument $e_2$ and pass the result to the
+function value). This is captured by \emph{pointwise composition}.
 
 Consider first \emph{sequential composition} of (semantic) functions $f, g$ annotated with
-$\cclrd{r}, \cclrd{s}$. The composed function $g \circ f$ is annotated with $\cclrd{r} \sqcup \cclrd{s}$
+$\cclrd{r}, \cclrd{s}$. The composed function $g \circ f$ is annotated with $\cclrd{r} \cclrd{\sqcup} \cclrd{s}$
 as shown in Figure~\ref{fig:applications-flat-livealg} (b).
@@
 is live if either the parameter of $f$ or $h$ is live. The pointwise composition is written as
-$\langle f, h \rangle$ and it combines annotations using $\sqcap$ as shown in Figure~\ref{fig:applications-flat-livealg} (c).
+$\langle f, h \rangle$ and it combines annotations using $\cclrd{\sqcap}$ as shown in Figure~\ref{fig:applications-flat-livealg} (c).
 Here, the argument is not needed only when both $f$ and $h$ do not need it ($1$). In all other cases,
@@
 from $\Gamma$; if it is dead, then it does not require $\Gamma$ or $x$.
-As further discussed later in Section~?, the (\emph{let}) rule is again just a syntactic sugar for
+The (\emph{let}) rule is again just a syntactic sugar for
 $(\lambda x.e_2)~e_1$. This follows from the simple observation that
-$\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{r}) = \cclrd{r}$.
+$\cclrd{r} \;\cclrd{\sqcup}\; (\cclrd{s} \;\cclrd{\sqcap}\; \cclrd{r}) = \cclrd{r}$.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+a.) The operations of a two-point lattice $\mathcal{L} = \{\cclrd{\ident{L}}, \cclrd{\ident{D}}\}$
+where $\cclrd{\ident{D}} \,\cclrd{\sqsubseteq}\, \cclrd{\ident{L}}$ are:
+%
+\begin{equation*}
+\begin{array}{rcl}
+\cclrd{\ident{L}} \,\cclrd{\sqcup}\, \cclrd{\ident{L}} &=& \cclrd{\ident{L}}\\
+\cclrd{\ident{D}} \,\cclrd{\sqcup}\, \cclrd{\ident{L}} &=& \cclrd{\ident{D}}\\
+\end{array}
+\qquad
+\begin{array}{rcl}
+\cclrd{\ident{L}} \,\cclrd{\sqcup}\, \cclrd{\ident{D}} &=& \cclrd{\ident{D}}\\
+\cclrd{\ident{D}} \,\cclrd{\sqcup}\, \cclrd{\ident{D}} &=& \cclrd{\ident{D}}
+\end{array}
+\qquad
+\begin{array}{rcl}
+\cclrd{\ident{L}} \,\cclrd{\sqcap}\, \cclrd{\ident{L}} &=& \cclrd{\ident{L}}\\
+\cclrd{\ident{D}} \,\cclrd{\sqcap}\, \cclrd{\ident{L}} &=& \cclrd{\ident{L}}\\
+\end{array}
+\qquad
+\begin{array}{rcl}
+\cclrd{\ident{L}} \,\cclrd{\sqcap}\, \cclrd{\ident{D}} &=& \cclrd{\ident{L}}\\
+\cclrd{\ident{D}} \,\cclrd{\sqcap}\, \cclrd{\ident{D}} &=& \cclrd{\ident{D}}
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+b.) Sequential composition composes annotations using $\cclrd{\sqcup}$:
+\begin{equation*}
+\begin{array}{llll}
+f : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 \qquad\qquad&
+g : \tau_2 \xrightarrow{\cclrd{s}} \tau_3 \qquad\qquad&
+g \circ f : \tau_1 \xrightarrow{\cclrd{r} \cclrd{\sqcup} \cclrd{s}} \tau_3 \qquad\qquad& ~
+\\[0.75em]
+f : \tau_1 \xrightarrow{\cclrd{\cclrd{\ident{L}}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\cclrd{\ident{L}}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_3 & (1)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\cclrd{\ident{L}}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_3 & (2)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\cclrd{\cclrd{\ident{L}}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_3 & (3)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau_2 &
+g : \tau_2 \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau_3 &
+g \circ f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_3 & (4)
+\end{array}
+\end{equation*}
+
+\vspace{0.5em}
+c.) Pointwise composition composes annotations using $\cclrd{\sqcap}$:
+\begin{equation*}
+\begin{array}{llll}
+f : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 \qquad\qquad&
+h : \tau_1 \xrightarrow{\cclrd{s}} \tau_3 \qquad\qquad&
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{r} \cclrd{\sqcap} \cclrd{s}} \tau_2 \times \tau_3 \qquad&~
+\\[0.75em]
+f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_2 \times \tau_3 & (1)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_2 \times \tau_3 & (2)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{D}} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_2 \times \tau_3 & (3)
+\\[-0.10em]
+f : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_2 &
+h : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_3 &
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{ \cclrd{\ident{L}} }} \tau_2 \times \tau_3 & (4)
+\end{array}
+\end{equation*}
+
+\figcaption{Liveness annotations with sequential and pointwise composition}
+\label{fig:applications-flat-livealg}
+\end{figure}
 
@@
 %
-In the first case, the context is dead. In ($1$), the function's parameter is dead and so the
+In the first case ($1$), the context is dead. The function's parameter is dead and so the
 overall context is dead, even though the argument uses a variable $y$ -- the semantics evaluates
@@
 representation depends on the annotation -- in other words, the type is \emph{indexed} by
-the coeffect annotation. The indexing is discussed in details in Section~\ref{sec:flat-semantics-idx}.
+the coeffect annotation. The indexing is discussed in details in Section~\ref{sec:semantics-flat-idx}.
 For now, it suffices to define the semantics using two separate rules:
@@
 \begin{array}{rlrcl}
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{\ident{L}} } \vdash e : \tau}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{\cclrd{\ident{L}}} } \vdash e : \tau}
   &:& (\tau_1 \times \ldots \times \tau_n) &\narrow{\rightarrow}& \tau\\
-\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{\ident{D}} } \vdash e : \tau}
+\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{\cclrd{\ident{D}}} } \vdash e : \tau}
   &:& 1 &\narrow{\rightarrow}& \tau
@@
 \begin{array}{l}
-\sem{\tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2\\
-\sem{\tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
+\sem{\tau_1 \xrightarrow{\cclrd{\cclrd{\ident{L}}}} \tau_2} = \tau_1 \rightarrow \tau_2\\
+\sem{\tau_1 \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau_2} = 1 \rightarrow \tau_2
 \end{array}
@@
 input and returns a constant using a global dictionary $\delta$. In (\emph{var}), the context is live
-and so the semantics performs a projection. Sub-coeffecting is captured by two rules. A dead context
+and so the semantics performs a projection. Subcoeffecting is captured by two rules. A dead context
 can be treated as live using (\emph{abs-1}); in other cases, the annotation is not changed (\emph{abs-2}).
@@
 where the last rule handles all cases not covered by the first two. In (\emph{app-1}), we handle the
-case when the function $g$ does not require its argument -- $e_2$ is not used and instead, the function
+case when the function $f_2$ does not require its argument -- $x$ is not used and instead, the function
 is called with $()$ as the argument. The case (\emph{app-2}) covers the case when the expression
@@
 
-\begin{figure}
-\begin{equation}
-\tag{\emph{var}}
-\sem{\coctx{\Gamma}{\cclrd{ \ident{L} }} \vdash x_i : \tau_i} = \lambda (x_1, \ldots, x_n) \rightarrow x_i
-\end{equation}
-\vspace{-1.75em}
-\begin{equation}
-\tag{\emph{const}}
-\sem{\coctx{\Gamma}{\cclrd{ \ident{D} }} \vdash c_i : \tau_i} = \lambda () \rightarrow \delta(c_i)
-\end{equation}
-%
-%
-\vspace{-0.5em}
-%
-%
-\begin{equation}
-\tag{\emph{sub-1}}
-\sem{\coctx{\Gamma}{\cclrd{\ident{L} }} \vdash e : \tau} = \lambda x \rightarrow
-  \sem{\coctx{\Gamma}{\cclrd{\ident{D} }} \vdash e : \tau}~()
-\end{equation}
-\vspace{-1.75em}
-\begin{equation}
-\tag{\emph{sub-2}}
-\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau} = \lambda x \rightarrow
-  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}~x
-\end{equation}
-%
-%
-\vspace{-0.5em}
-%
-%
-\begin{equation}
-\tag{\emph{abs-1}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 } = \lambda (x_1, \ldots, x_n) \rightarrow\\[-0.25em]
-  \qquad\lambda y \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{\ident{L}}} \vdash e : \tau_2 }~(x_1, \ldots, x_n, y)
-\end{array}
-\end{equation}
-\vspace{-1.0em}
-\begin{equation}
-\tag{\emph{abs-2}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{\ident{D}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 } = \lambda () \rightarrow\\[-0.25em]
-  \qquad\lambda () \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{\ident{D}}} \vdash e : \tau_2 }~()
-\end{array}
-\end{equation}
-%
-%
-\vspace{0.0em}
-%
-%
-%
-\begin{equation*}
-\tag{\emph{app-1}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{ \ident{D} }} \tau_2 }~x
-    ~\kvd{in}~g~()
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\tag{\emph{app-2}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{ \ident{L} }} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{ \ident{L} }} \tau_2 }~x
-    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{ \ident{D} }} \vdash e_2 : \tau_1 }~())
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\tag{\emph{app-3}}
-\hspace{-0.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{r} \sqcup (\cclrd{s} \sqcap \cclrd{t}) } \vdash e_1~e_2 : \tau_2 } = \lambda x \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~x
-    ~\kvd{in}~g~(\sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }~x)
+\begin{figure}[t]
+\begin{equation*}
+\hspace{0em}
+\begin{array}{ll}
+\hspace{3.5em}\semdef
+  {\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash x_i : \tau_i}
+  {\lambda (x_1, \ldots, x_n) . x_i}
+& (\emph{var})
+\\[-0.75em]
+\hspace{2.75em}\semdef
+  {\coctx{\Gamma}{\cclrd{\ident{D}}} \vdash n : \ident{num}}
+  {\lambda () . n}
+& (\emph{num})
+\\[2em]
+\hspace{4.25em}\semdeff
+  {\coctx{\Gamma}{\cclrd{\ident{D}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash e : \tau}
+  {f}
+  {\lambda x . f~()}
+& (\emph{sub-1})
+\\[1em]
+\hspace{4.5em}\semdeff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
+  {f}
+  {\lambda x . f~x}
+& (\emph{sub-2})
+\\[2.5em]
+\hspace{-0.55em}\emdeff
+  {\coctx{\Gamma,y:\tau_1}{\cclrd{\ident{L}}} \vdash e : \tau_2}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2}\\[-0.25em]~\end{array}\hspace{-0.5em}~}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}\lambda (x_1, \ldots, x_n) . \lambda y .\\[-0.25em]
+    \quad f~(x_1, \ldots, x_n, y)\end{array}~~}
+& (\emph{abs-1})
+\\[1em]
+\hspace{-0.5em}\semdeff
+  {\coctx{\Gamma,y:\tau_1}{\cclrd{\ident{D}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\cclrd{\ident{D}}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2}
+  {f}
+  {\lambda () . \lambda () . f~()}
+& (\emph{abs-2})
+\\[2.5em]
+\hspace{0.75em}\semdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1~e_2 : \tau_2}
+  {f}
+  {\_}
+  {\lambda x . (f~x)~()}
+& (\emph{app-1})
+\\[2em]
+\hspace{0.75em}\semdefff
+  {\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{\ident{D}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma}{\cclrd{\ident{L}}} \vdash e_1~e_2 : \tau_2}
+  {f_1}
+  {f_2}
+  {\lambda x . (f_1~x)~(f_2~())}
+& (\emph{app-2})
+\\[2em]
+\hspace{-1.6em}\semdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma}{\cclrd{r} \;\cclrd{\sqcup}\; (\cclrd{s} \,\cclrd{\sqcap}\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
+  {f_1}
+  {f_2}
+  {\lambda x . (f_1~x)~(f_2~x)}
+& (\emph{app-3})
 \end{array}
@@
 \label{fig:applications-flat-livsem}
+\vspace{0.5em}
 \end{figure}
@@
   \begin{array}{ll}
-    nil     & \; \text{if $n=0$}\\
+    \ident{nil}     & \; \text{if $n=0$}\\
     x_{n-1} & \; \text{if $n>0$}
@@
    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m}, \cclrd{n} + \cclrd{p})} \vdash e_1~e_2 : \tau_2}
+  {\coctx{\Gamma}{\cclrd{\textnormal{max}}(\cclrd{m}, \cclrd{n+p})} \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
     \coctx{\Gamma, x:\tau_1}{\cclrd{n}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma}{\cclrd{n}+\cclrd{m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+  {\coctx{\Gamma}{\cclrd{n+m}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
 \end{equation*}
@@
 
-Variable access (\emph{var}) annotates the context with $0$; sub-coeffecting (\emph{sub}) allows
+Variable access (\emph{var}) annotates the context with $0$; subcoeffecting (\emph{sub}) allows
 us to require more values than is actually needed. Primitive context-requirements are introduced
@@
 g : \tau_2 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
-g \circ f : \tau_1 \xrightarrow{\cclrd{m} + \cclrd{n}} \tau_3 \qquad\qquad
+g \circ f : \tau_1 \xrightarrow{\cclrd{m+n}} \tau_3 \qquad\qquad
 \\
@@
 h : \tau_1 \xrightarrow{\cclrd{n}} \tau_3 \qquad\qquad&
-\langle f, h \rangle : \tau_1 \xrightarrow{\textnormal{max}(\cclrd{m}, \cclrd{s})} \tau_2 \times \tau_3 \qquad
+\langle f, h \rangle : \tau_1 \xrightarrow{\cclrd{\textnormal{max}}(\cclrd{m}, \cclrd{s})} (\tau_2 \times \tau_3) \qquad
 \end{array}
@@
 and so on. To produce $\ident{n}$ past values that are required as the input of $g$, we need
-$\cclrd{m}+\cclrd{n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
+$\cclrd{m+n}$ past values of type $\tau_1$. The pointwise composition is simpler. It uses
 the same stream to evaluate functions requiring $\cclrd{m}$ and $\cclrd{n}$ past values, and so it
@@
 \,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
-\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
+\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\quad(\kvd{prev}~\ident{counter})
 \end{array}
@@
 
-%---------------------------------------------------------------------------------------------------
-
-\begin{figure}
-\begin{equation}
-\tag{\emph{var}}
-\hspace{-1em}
-\sem{\coctx{\Gamma}{\cclrd{0}} \vdash x_i : \tau_i} = \lambda \langle(x_0, \ldots, x_n)\rangle \rightarrow x_i
-\end{equation}
-\vspace{-1.0em}
-\begin{equation}
-\tag{\emph{prev}}
-\hspace{-1.5em}
-\begin{array}{l}
-\sem{\coctx{\Gamma}{\cclrd{n}+1} \vdash \kvd{prev}~e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}+1} \rangle \rightarrow\\[-0.1em]
-  \qquad\sem{ \coctx{\Gamma}{ \cclrd{n} } \vdash e : \tau }~\langle \mathbf{v}_1, \ldots, \mathbf{v}_{ \cclrd{n}+1 }\rangle
-\end{array}
-\end{equation}
-\vspace{-0.5em}
-\begin{equation}
-\tag{\emph{sub}}
-\hspace{-1.5em}
-\begin{array}{l}
-\sem{\coctx{\Gamma}{\cclrd{n}} \vdash e : \tau} = \lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle \rightarrow\\[-0.1em]
-  \qquad\sem{ \coctx{\Gamma}{ \cclrd{n'} } \vdash e : \tau }~\langle \mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{n'} }\rangle
-\end{array}
-\end{equation}
-\vspace{-0.5em}
-\begin{equation}
-\tag{\emph{abs}}
-\hspace{-1.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\cclrd{n}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 } =
-    \lambda \langle \mathbf{v}_0, \ldots \mathbf{v}_{\cclrd{n}} \rangle \rightarrow\\[-0.1em]
-  \qquad\lambda (y, g) \rightarrow \sem{ \coctx{\Gamma,y:\tau_1}{\cclrd{n}} \vdash e : \tau_2
-    }~\langle (\mathbf{v}_0, y_0), \ldots, (\mathbf{v}_{ \cclrd{n}  }, y_{ \cclrd{n} } ) \rangle
-\end{array}
-\end{equation}
-\vspace{-0.5em}
-\begin{equation*}
-\tag{\emph{app}}
-\hspace{-1.5em}
-\begin{array}{l}
-  \sem{\coctx{\Gamma}{\textnormal{max}(\cclrd{m}, \cclrd{n}+\cclrd{p})} \vdash e_1~e_2 : \tau_2 } =
-    \lambda (\mathbf{v}_0, \ldots, \mathbf{v}_{ \textnormal{max}(\cclrd{m}, \cclrd{n}+\cclrd{p}) } ) \rightarrow \\[-0.1em]
-  \qquad\kvd{let}~g=\sem{ \coctx{\Gamma}{\cclrd{m}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{p}} \tau_2
-    }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{m} })\\[-0.1em]
-  \qquad\kvd{in}~g~(~\sem{ \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{n} }),\ldots, \\[-0.1em]
-  \qquad\qquad\;~\sem{ \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau_1 }~(\mathbf{v}_{ \cclrd{p} }, \ldots, \mathbf{v}_{ \cclrd{n}+\cclrd{p} })~)
-\end{array}
-\end{equation*}
-
-\figcaption{Semantics showing how past values are accessed in a data-flow language}
-\label{fig:applications-flat-dfsem}
-\end{figure}
-
-%---------------------------------------------------------------------------------------------------
-
 \paragraph{Semantics.}
-The sample language discussed in this section is a \emph{causal} data-flow language. This means that
+The language discussed in this section is a \emph{causal} data-flow language. This means that
 a computation can access \emph{past} values of the stream (but not future values). In the semantics,
@@
 a list of fixed length. As with liveness, this is a data structure \emph{indexed} by the coeffect
-annotation. We write $\tau^{\cclrd{n}}$ for a list containing $\cclrd{n}$ elements (which can be
-also viewed as an $\cclrd{n}$-element product $\tau \times \ldots \times \tau$).
+annotation. We write $\tau^{\cclrd{n}}$ for a list containing $\cclrd{n}$ elements, which can be
+also viewed as an $\cclrd{n}$-element product $\tau \times \ldots \times \tau$.
 
@@
 zero as in (\emph{var}).
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+\begin{equation*}
+\begin{array}{ll}
+\hspace{3em}\semdef
+  {\coctx{\Gamma}{\cclrd{0}} \vdash x_i : \tau_i}
+  {\lambda \langle(x_1, \ldots, x_n)\rangle . x_i}
+& (\emph{var})
+\\[1.5em]
+\hspace{-0.75em}\emdeff
+  {\coctx{\Gamma}{\cclrd{n}} \vdash e : \tau}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{n}+1} \vdash \kvd{prev}~e : \tau}\\[-0.25em]~\end{array}\hspace{-0.5em}~}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}\lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}+1} \rangle .\\[-0.25em]
+    \qquad f~\langle \mathbf{v}_1, \ldots, \mathbf{v}_{ \cclrd{n}+1 }\rangle\end{array}\hspace{-1em}~}
+& (\emph{prev})
+\\[1.5em]
+\hspace{3.25em}\emdeff
+  {\coctx{\Gamma}{\cclrd{n'}} \vdash e : \tau}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{n}} \vdash e : \tau}\\[-0.25em]~\end{array}\hspace{-0.5em}~}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}\lambda \langle \mathbf{v}_0, \ldots, \mathbf{v}_{\cclrd{n}} \rangle .\\[-0.25em]
+    \qquad f~\langle \mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{n'} }\rangle\end{array}\hspace{-1em}~}
+& (\emph{sub})
+\\[1.5em]
+\hspace{-1.5em}\emdeff
+  {\coctx{\Gamma,y:\tau_1}{\cclrd{n}} \vdash e : \tau_2}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{n}} \vdash \lambda y. e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2}\\[-0.25em]~\end{array}\hspace{-0.5em}~}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda \langle \mathbf{v}_0, \ldots \mathbf{v}_{\cclrd{n}} \rangle . \lambda \langle y_0, \ldots, y_{\cclrd{n}}\rangle .\\[-0.25em]
+  \quad f~\langle (\mathbf{v}_0, y_0), \ldots, (\mathbf{v}_{ \cclrd{n}  }, y_{ \cclrd{n} } ) \rangle
+  \end{array}\hspace{-1em}~}
+& (\emph{abs})
+\\[1.5em]
+\hspace{0.25em}\emdefff
+  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1}
+  {\begin{array}{l}\coctx{\Gamma}{\cclrd{\textnormal{max}}(\cclrd{m}, \cclrd{n+p})} \\[-0.25em]\qquad \vdash e_1~e_2 : \tau_2\\[-0.25em]~\\[-0.25em]~\end{array}}
+  {f_1}
+  {f_2}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda \langle\mathbf{v}_0, \ldots, \mathbf{v}_{ \textnormal{max}(\cclrd{m}, \cclrd{n}+\cclrd{p})} \rangle . \\[-0.1em]
+  \quad (f_1~\langle\mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{m} }\rangle)\\[-0.1em]
+  ~~\qquad\langle\; f_2~\langle\mathbf{v}_0, \ldots, \mathbf{v}_{ \cclrd{n} }\rangle,\ldots,\\[-0.25em]
+  ~~\qquad~\;\, f_2~\langle\mathbf{v}_{ \cclrd{p} }, \ldots, \mathbf{v}_{ \cclrd{n}+\cclrd{p} }\rangle~\rangle
+  \end{array}\hspace{-0.5em}~}
+& (\emph{app})
+\\[1.5em]
+\end{array}
+\end{equation*}
+
+\figcaption{Semantics showing how past values are accessed in a data-flow language}
+\label{fig:applications-flat-dfsem}
+\end{figure}
+
+%---------------------------------------------------------------------------------------------------
 
@@
 \emph{capabilities} to perform some operation. We can invert the intuition behind liveness and
-use it as a trivial example -- when the context is live, it contains a \emph{permission} to access
+use it as a trivial example. When the context is live, it contains a \emph{permission} to access
 variables. In this section, we briefly consider a system for safe locking of Flanagan and
@@
 
+\noindent
+\vspace{-0.5em}
 \begin{equation*}
@@
 \end{equation*}
-%
+\vspace{-0.5em}
+
+\noindent
 The rule requires that $e_1$ yields a value of a singleton type $m$. The type is added as an
@@
 the variable context $\Gamma$ is a vector $\langle y\!:\!\tau, x\!:\!\tau \rangle$. Only the variable $y$
-is used and so the annotated context is: $\coctx{y\!:\!\tau, x\!:\!\tau}{ \alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}} }$.
+is used and so the annotated context is: $\coctx{y\!:\!\tau, x\!:\!\tau}{ \alift{\cclrd{\cclrd{\ident{L}}}, \cclrd{\cclrd{\ident{D}}}} }$.
 When writing the contexts, we omit angle brackets around variables, but it should still be viewed
@@
 \item The fact that variables are now a vector means that we cannot freely re-order them. This
-  guarantees that $\coctx{x\!:\!\tau, y\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$
-  can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$.
+  guarantees that $\coctx{x\!:\!\tau, y\!:\!\tau}{\alift{\cclrd{\cclrd{\ident{L}}}, \cclrd{\cclrd{\ident{D}}}}}$
+  can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\cclrd{\ident{L}}}, \cclrd{\cclrd{\ident{D}}}}}$.
   We need to define the type system in a way that is similar to substructural systems
@@
   individual annotations to individual variables. This lets us unify and combine flat and
-  structural systems as discussed in Section~\ref{sec:unified-unified}, but the alternative is briefly
-  explored in Section~\ref{sec:unified-meta}.
+  structural systems as discussed in Section~\ref{sec:further-unified}, but the alternative is briefly
+  explored in Section~\ref{sec:further-meta}.
 \end{itemize}
@@
 The structural system for liveness uses the same two-point lattice of annotations
-$\mathcal{L}=\{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$ that was used by the flat system. We also use the
-$\sqcup, \sqcap$ and $\sqsubseteq$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.
+$\mathcal{L}=\{ \cclrd{\cclrd{\ident{L}}}, \cclrd{\cclrd{\ident{D}}} \}$ that was used by the flat system. We also use the
+$\cclrd{\sqcup}, \cclrd{\sqcap}$ and $\cclrd{\sqsubseteq}$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.
 
 The rules of the system are split into two groups. Figure~\ref{fig:applications-struct-live} (a) shows
-the standard syntax-driven rules plus sub-coeffecting. In (\emph{var}), the context contains just the
+the standard syntax-driven rules plus subcoeffecting. In (\emph{var}), the context contains just the
 single accessed variable, which is annotated as live. Unused variables can be introduced using weakening.
 A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
-sub-coeffecting rule (\emph{sub}) uses a pointwise extension of the $\sqsubseteq$ relation over two
+subcoeffecting rule (\emph{sub}) uses a pointwise extension of the $\cclrd{\sqsubseteq}$ relation over two
 vectors as defined in Section~\ref{sec:applications-strucutre-vec}.
 
-In the (\emph{abs}) rule, the variable context of the body $\Gamma, x\!:\!\tau_1$ is annotated with
-a vector $\mathbf{\aclrd{r}}\cons\alift{\cclrd{s}}$, where the vector $\mathbf{\aclrd{r}}$ corresponds
-to $\Gamma$ and the singleton annotation $\cclrd{s}$ corresponds to the variable $x$. Thus, the
-function is annotated with $\cclrd{s}$. Note that the free-variable context is annotated with vectors,
-but functions take only a single input and so are annotated with primitive annotations.
-
-The (\emph{app}) rule is similar to function applications in flat systems, but there is an important
-difference. In structural systems, the two sub-expressions have separate variable contexts
-$\Gamma_1$ and $\Gamma_2$. Therefore, the composed expression just concatenates the variables
-and their corresponding annotations. (We can still use the same variable in both sub-expressions
-thanks to the structural contraction rule.)
-
-The context $\Gamma_1$ is used to evaluate $e_1$ and is thus annotated with $\aclrd{\mathbf{r}}$.
-The annotation for $\Gamma_2$ is more interesting. It is a result of sequential composition of two
-semantic functions -- the first one takes the (multi-variable) context $\Gamma_2$ and evaluates
-$e_2$; the second takes the result of type $\tau_1$ and passes it to the function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$.
-The composition is defined as follows:
-%
-\begin{equation*}
-g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\aclrd{\mathbf{s}}} \sigma
-\qquad
-f : \sigma \xrightarrow{\cclrd{t}} \tau
-\qquad
-f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \,\sqcup\, \aclrd{\mathbf{s}}} \tau
-\end{equation*}
-%
-This definition is only for illustration and is revised in Chapter~\ref{ch:structural}. The function
-$g$ takes a product of multiple variables (and is annotated with a vector). The function $f$ takes
-just a single value and is annotated with the scalar. As in the flat system, sequential composition
-is modelled using $\sqcup$ -- but here, we use a scalar-vector extension of the operation. Finally,
-the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$).
-
 % --------------------------------------------------------------------------------------------------
@@
 \begin{figure}[t]
-{\small a.) Ordinary, syntax-driven rules along with sub-coeffecting}
+a.) Ordinary, syntax-driven rules along with subcoeffecting
+
 \begin{equation*}
@@
     \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \;\cclrd{\sqcup}\; \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
 \end{equation*}
@@
     \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \;\cclrd{\sqcup}\; \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
 \end{equation*}
@@
   {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
-\end{equation*}
-\vspace{0.5em}
-
-{\small b.) Structural rules for context manipulation}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \;\cclrd{\sqsubseteq}\; \aclrd{\textbf{r'}}
+\end{equation*}
+\vspace{1em}
+
+b.) Structural rules for context manipulation
+
 \begin{equation*}
@@
   {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
-  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\cclrd{\ident{D}}} }} \vdash e : \sigma}
 \end{equation*}
@@
   {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}\sqcap\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}\cclrd{\sqcap}\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
 ~
@@
 
+In the (\emph{abs}) rule, the variable context of the body $\Gamma, x\!:\!\tau_1$ is annotated with
+a vector $\mathbf{\aclrd{r}}\cons\alift{\cclrd{s}}$, where the vector $\mathbf{\aclrd{r}}$ corresponds
+to $\Gamma$ and the singleton annotation $\cclrd{s}$ corresponds to the variable $x$. Thus, the
+function is annotated with $\cclrd{s}$. Note that the free-variable context is annotated with vectors,
+but functions take only a single input and so are annotated with primitive annotations.
+
+The (\emph{app}) rule is similar to function applications in flat systems, but there is an important
+difference. In structural systems, the two sub-expressions have separate variable contexts
+$\Gamma_1$ and $\Gamma_2$. Therefore, the composed expression just concatenates the variables
+and their corresponding annotations. (We can still use the same variable in both sub-expressions
+thanks to the structural contraction rule.)
+
+The context $\Gamma_1$ is used to evaluate $e_1$ and is thus annotated with $\aclrd{\mathbf{r}}$.
+The annotation for $\Gamma_2$ is more interesting. It is a result of sequential composition of two
+semantic functions -- the first one takes the (multi-variable) context $\Gamma_2$ and evaluates
+$e_2$; the second takes the result of type $\tau_1$ and passes it to the function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$.
+The composition is defined as follows:
+%
+\begin{equation*}
+g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\aclrd{\mathbf{s}}} \sigma
+\qquad
+f : \sigma \xrightarrow{\cclrd{t}} \tau
+\qquad
+f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \,\cclrd{\sqcup}\, \aclrd{\mathbf{s}}} \tau
+\end{equation*}
+%
+This definition is only for illustration and is revised in Chapter~\ref{ch:structural}. The function
+$g$ takes a product of multiple variables (and is annotated with a vector). The function $f$ takes
+just a single value and is annotated with the scalar. As in the flat system, sequential composition
+is modelled using $\cclrd{\sqcup}$, but here we use a scalar-vector extension of the operation. Finally,
+the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$).
+
 \paragraph{Structural typing rules.}
@@
 the rules know from substructural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
-extends the context with a single unused variable $x$ and adds the $\ident{D}$ annotation to the
+extends the context with a single unused variable $x$ and adds the $\cclrd{\ident{D}}$ annotation to the
 vector of coeffects.
@@
 where both $z$ and $y$ are replaced by a single variable $x$. Their annotations $\cclrd{s}, \cclrd{t}$
-are combined into $\cclrd{s} \sqcap \cclrd{t}$, which means that $x$ is live if either $z$ or $y$
+are combined into $\cclrd{s} \;\cclrd{\sqcap}\; \cclrd{t}$, which means that $x$ is live if either $z$ or $y$
 were live in the original expression.
@@
 \paragraph{Example.} To demonstrate how the system works, we consider the expression
-$(\lambda x \rightarrow v)~y$. This is similar to an example where flat liveness mistakenly marks
+$(\lambda x . v)~y$. This is similar to an example where flat liveness mistakenly marks
 the entire context as live. Despite the fact that the variable $y$ is accessed (syntactically), it
@@
     {}
-    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash v : \tau } }
+    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}} }} \vdash v : \tau } }
 { \tyruler{abs}
-    { \coctx{v\!:\!\tau, x\!:\!\tau}{\alift{ \cclrd{\ident{L}},\cclrd{\ident{D}} }} \vdash v : \tau }
-    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash (\lambda x \rightarrow v) : \tau \xrightarrow{\cclrd{\ident{D}}} \tau }}
+    { \coctx{v\!:\!\tau, x\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}},\cclrd{\cclrd{\ident{D}}} }} \vdash v : \tau }
+    { \coctx{v\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}} }} \vdash (\lambda x . v) : \tau \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau }}
 \end{equation*}
@@
     \vspace{-1.2em}
-    \coctx{v\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash (\lambda x \rightarrow v) : \tau \xrightarrow{\cclrd{\ident{D}}} \tau
+    \coctx{v\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}} }} \vdash (\lambda x . v) : \tau \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau
     \end{array} &
-    \tyruler{var}{}{\coctx{y\!:\!\tau}{\alift{ \cclrd{\ident{L}} }} \vdash y : \tau}
+    \tyruler{var}{}{\coctx{y\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}} }} \vdash y : \tau}
   }
@@
   \inference
-  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\ident{L}} } \cons\, ({ \cclrd{\ident{D}} \sqcup \alift{\cclrd{\ident{L}}} })}
-  	     \vdash (\lambda x \rightarrow v)~y : \tau }
-  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\ident{L}}, \cclrd{\ident{D}} }} \vdash (\lambda x \rightarrow v)~y : \tau }
+  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}} } \cons\, ({ \cclrd{\cclrd{\ident{D}}} \,\cclrd{\sqcup}\, \alift{\cclrd{\cclrd{\ident{L}}}} })}
+  	     \vdash (\lambda x . v)~y : \tau }
+  	{ \coctx{v\!:\!\tau, y\!:\!\tau}{\alift{ \cclrd{\cclrd{\ident{L}}}, \cclrd{\cclrd{\ident{D}}} }} \vdash (\lambda x . v)~y : \tau }
   }
@@
 and the second one simplifies the coeffect annotation. The key part is the use of the scalar-vector
-operator $\cclrd{\ident{D}} \sqcup \alift{\cclrd{\ident{L}}}$. Using the definition of the scalar-vector
-extension, this equals $\alift{\cclrd{\ident{D}} \sqcup \cclrd{\ident{L}}}$ which is
-$\alift{\cclrd{\ident{D}}}$.
+operator $\cclrd{\cclrd{\ident{D}}} \;\cclrd{\sqcup}\; \alift{\cclrd{\cclrd{\ident{L}}}}$. Using the definition of the scalar-vector
+extension, this equals $\alift{\cclrd{\cclrd{\ident{D}}} \;\cclrd{\sqcup}\; \cclrd{\cclrd{\ident{L}}}}$ which is
+$\alift{\cclrd{\cclrd{\ident{D}}}}$.
 
@@
 the type is indexed by the corresponding annotation:
-%
+
 \begin{equation*}
 \begin{array}{r}
+\\[-0.5em]
 \sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau}
@@
 \textnormal{where}~\tau'_i = \begin{cases}
-\tau_i & (\cclrd{r_i} = \cclrd{\ident{L}})\\
-1      & (\cclrd{r_i} = \cclrd{\ident{D}})
+\tau_i & (\cclrd{r_i} = \cclrd{\cclrd{\ident{L}}})\\
+1      & (\cclrd{r_i} = \cclrd{\cclrd{\ident{D}}})
 \end{cases}
@@
 \end{equation*}
-%
+\vspace{0.5em}
+
+\noindent
 Note that the product of the free variables is not an ordinary tuple of our language, but a special
@@
 take just a single input and so they are interpreted in the same way as in flat calculus:
-%
-\begin{equation*}
-\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2 \qquad\qquad
-\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
-\end{equation*}
-%
+
+\vspace{-0.5em}
+\begin{equation*}
+\sem{\tau_1 \xrightarrow{\aclrd{\cclrd{\ident{L}}}} \tau_2} = \tau_1 \rightarrow \tau_2 \qquad\qquad
+\sem{\tau_1 \xrightarrow{\aclrd{\cclrd{\ident{D}}}} \tau_2} = 1 \rightarrow \tau_2
+\end{equation*}
+\vspace{-0.5em}
+
+\noindent
 The rules that define the semantics are shown in Figure~\ref{fig:applications-struct-livesem}.
@@
 works the same way regardless of whether the variables are live or dead. For simplicity, we omit
-sub-coeffecting, which just turns some of the available values $v_i$ to unit values $()$.
+subcoeffecting, which just turns some of the available values $v_i$ to unit values $()$.
 
 As dictated by the semantics, the application again needs to ``implement'' dead code elimination
-(otherwise the type system would be unsound). When the input parameter of the function $g$ is live
-(\emph{app-1}), we first evaluate $e_2$ and then pass the result to $g$. When the parameter is
+(otherwise the type system would be unsound). When the input parameter of the function $f_1$ is live
+(\emph{app-1}), we first evaluate $e_2$ and then pass the result to $f_1$. When the parameter is
 dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values in $\mathbf{v_2}$ can be
@@
 It drops the $()$ value and evaluates the expression in a context $\mathbf{v}$. Exchange (\emph{exch})
-simply swaps two variables. In contraction, we either duplicate a dead value (\emph{contr-1}),
-or a live value (\emph{contr-2}). In the latter, one of the duplicates may be dead and
-so we need to consider three separate cases.
+simply swaps two variables. In contraction, we duplicate the value (no matter whether it is dead
+or live) and we use an auxiliary definition $\restr{x}{\cclrd{r}}$ to replace a live value with
+$()$ when only one of the contracted variables is live.
 
@@
 \begin{figure}[t]
-{\small a.) Semantics of ordinary expressions}
-\begin{equation*}
-\tag{\emph{var}}
-\sem{\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} = \lambda(x) \rightarrow x
-\end{equation*}
-\vspace{-1.5em}
-\begin{equation*}
-\tag{\emph{const}}
-\sem{\coctx{()}{\alift{}} \vdash c : \tau} = \lambda () \rightarrow \delta(c)
-\end{equation*}
-\vspace{-1.5em}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{abs}}
-\begin{array}{l}
-\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda y . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} = \lambda \textbf{v} \rightarrow\\[-0.25em]
-\qquad\lambda y \rightarrow \sem{\coctx{\Gamma, y \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}~(\textbf{v}, y)
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{app-1}}
-\begin{array}{l}
-\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{L}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} =
-  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{L}}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
-  \qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }~\mathbf{v_2})
-\end{array}
-\end{equation*}
-\vspace{-1.0em}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{app-2}}
-\begin{array}{l}
-\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\ident{D}} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} =
-  \lambda (\mathbf{v_1}, \mathbf{v_2}) \rightarrow \\[-0.25em]
-  \qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\ident{D}}} \tau_2 }~\mathbf{v_1}~
-  \kvd{in}~g~()
-\end{array}
-\end{equation*}
-\vspace{0.5em}
-
-{\small b.) Semantics of structural context manipulation}
-\begin{equation*}
-\tag{\emph{weak}}
-\sem{\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\ident{D}} }} \vdash e : \sigma} =
-  \lambda(\textbf{v}, ()) \rightarrow
-    \sem{\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}~\textbf{v}
-\end{equation*}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{exch}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} =
-  \lambda(\mathbf{v_1}, y, x, \mathbf{v_2}) \rightarrow\\[-0.25em]
-\qquad
-  \sem{\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~
-  (\mathbf{v_1}, x, y, \mathbf{v_2})
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{contr-1}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, (), \mathbf{v_2}) \rightarrow \\[-0.25em]
-\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma} ~(\mathbf{v_1}, (), (), \mathbf{v_2})
-\end{array}
-\end{equation*}
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{contr-2}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} = \lambda(\mathbf{v_1}, x, \mathbf{v_2}) \rightarrow \\[-0.25em]
-\quad\begin{cases}
-~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, x, \mathbf{v_2})\\[-0.25em]
-~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{D}},\cclrd{\ident{L}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, (), x, \mathbf{v_2})\\[-0.25em]
-~~\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\ident{L}},\cclrd{\ident{D}}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}~(\mathbf{v_1}, x, (), \mathbf{v_2})
-\end{cases}
-\end{array}
-\end{equation*}
+a.) Semantics of ordinary expressions
+\begin{equation*}
+\hspace{-1em}
+\begin{array}{ll}
+\\[-2.5em]
+\hspace{4.75em}\semdef
+  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau}
+  {\lambda (x). x}
+& (\emph{var})
+\\[-0.5em]
+\hspace{4.5em}\semdef
+  {\coctx{()}{\alift{}} \vdash n : \ident{num}}
+  {\lambda (). n}
+& (\emph{num})
+\\[1.5em]
+\hspace{1.75em}\semdeff
+  {\coctx{\Gamma, y \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda y . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
+  {f}
+  {\lambda \textbf{v} . \lambda y . f~(\textbf{v}, y)}
+& (\emph{abs})
+\\[1.5em]
+\hspace{-0.75em}\semdefff
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\cclrd{\ident{L}}}} \tau_2}
+  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\cclrd{\ident{L}}} \,\cclrd{\sqcup}\, \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
+  {f_1}
+  {f_2}
+  {\lambda (\mathbf{v_1}, \mathbf{v_2}) . (f_1\;\mathbf{v_1})~(f_2\;\mathbf{v_2})}
+& (\emph{app-1})
+\\[2.5em]
+\hspace{-0.75em}\semdefff
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{\cclrd{\ident{D}}}} \tau_2}
+  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{\cclrd{\ident{D}}} \,\cclrd{\sqcup}\, \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
+  {f_1}
+  {\_}
+  {\lambda (\mathbf{v_1}, \mathbf{v_2}) . (f_1~\mathbf{v_1})~()}
+& (\emph{app-2})
+\end{array}
+\end{equation*}
+\vspace{1em}
+
+b.) Semantics of structural context manipulation\\
+Using the auxiliary definition $\restr{x}{\cclrd{\ident{L}}} = x$ and $\restr{x}{\cclrd{\ident{D}}} = ()$:
+
+\begin{equation*}
+\begin{array}{ll}
+\hspace{2em}\semdeff
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{\cclrd{\ident{D}}} }} \vdash e : \sigma}
+  {f}
+  {\lambda(\textbf{v}, ()) . f~\textbf{v} }
+& (\emph{weak})
+\\[1.5em]
+\hspace{0em}\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2\\[-0.25em]\qquad}
+     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda(\mathbf{v_1}, y, x, \mathbf{v_2}) .\\[-0.25em]
+    \quad f~(\mathbf{v_1}, x, y, \mathbf{v_2})\end{array} }
+& (\emph{exch})
+\\[1.5em]
+\hspace{-0.5em}\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
+  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
+  {\hspace{-0.5em}\begin{array}{l}\lambda(\mathbf{v_1}, x, \mathbf{v_2}) .\\[-0.25em]
+  \quad f~(\mathbf{v_1}, \restr{x}{\cclrd{s}}, \restr{x}{\cclrd{t}}, \mathbf{v_2})\end{array}\hspace{-0.5em}~}
+& (\emph{contr})
+\end{array}
+\end{equation*}
+
 \vspace{-0.5em}
@@
 transformation from the context in the (typing) conclusion to the (typing) assumption. How
-should coeffects be understood, in general, is discussed further in Section~\ref{sec:flat-calculus-undestanding}.
+should coeffects be understood, in general, is discussed further in Section~\ref{sec:flat-calculus-understanding}.
 
 The structural systems discussed in the upcoming sections are remarkably similar to the one
-shown here. We discuss two more examples to explore the design space, but we shall
-omit details that are the shared with the system in this section.
+shown here. We discuss two more examples to explore the design space, but omit details shared with
+the system in this section.
 
@@
 relate it with the original formulation.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+a.) Ordinary, syntax-driven rules along with subcoeffecting
+\begin{equation*}
+\tyrule{var}{}
+  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad
+    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+\end{equation*}
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
+\end{equation*}
+\vspace{0.5em}
+
+b.) Structural rules for context manipulation
+\begin{equation*}
+\tyrule{weak}
+  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
+  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma}
+\end{equation*}
+\begin{equation*}
+\tyrule{exch}
+  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+\quad
+\begin{array}{l}
+\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
+\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+\begin{equation*}
+\tyrule{contr}
+  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+~
+\begin{array}{l}
+\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
+\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
+\end{array}
+\end{equation*}
+
+\figcaption{Structural coeffect bounded reuse analysis}
+\label{fig:applications-struct-bll}
+\vspace{-1em}
+\end{figure}
 
@@
 
-Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates the number of accesses using
+Finally, the subcoeffecting rule (\emph{sub}) safely overapproximates the number of accesses using
 the pointwise $\leq$ relation. We can view any variable as being used a greater
 number of times than it actually is.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-{\small a.) Ordinary, syntax-driven rules along with sub-coeffecting}
-\begin{equation*}
-\tyrule{var}{}
-  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad
-    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad
-    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
-\end{equation*}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
-  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
-\end{equation*}
-\vspace{0.5em}
-
-{\small b.) Structural rules for context manipulation}
-\begin{equation*}
-\tyrule{weak}
-  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \sigma}
-  {\coctx{\Gamma,x \!:\! \tau}{\aclrd{\textbf{r}} \cons \alift{ \cclrd{0} }} \vdash e : \sigma}
-\end{equation*}
-\begin{equation*}
-\tyrule{exch}
-  {\coctx{\Gamma_1,x\!:\!\tau',y\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-  {\coctx{\Gamma_1,y\!:\!\tau,x\!:\!\tau',\Gamma_2}{\aclrd{\textbf{r}} \cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-\quad
-\begin{array}{l}
-\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
-\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-\begin{equation*}
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
-~
-\begin{array}{l}
-\slen{\Gamma_1} = \slen{\aclrd{\textbf{r}}}\\[-0.25em]
-\slen{\Gamma_2} = \slen{\aclrd{\textbf{s}}}
-\end{array}
-\end{equation*}
-
-\figcaption{Structural coeffect bounded reuse analysis}
-\label{fig:applications-struct-bll}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
 
@@
 
-The rules that define the semantics of bounded calculus are mostly the same as (or easy to adapt
-from) the semantic rules of liveness in Figure~\ref{fig:applications-struct-livesem}. The ones that
-differ are those that use sequential composition (application and let binding) and the contraction
-rule, which represents pointwise composition.
+The rules that define the semantics of bounded calculus are easy to adapt from the semantic rules
+of liveness in Figure~\ref{fig:applications-struct-livesem}. The ones that differ are those that use
+sequential composition (application and let binding) and the contraction rule, which represents
+pointwise composition.
 
@@
 of instances of a variable) as these are clear from the coeffect annotations. We assume that $\Gamma_2$
-contains $n$ variables and that $\aclrd{s}=\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}$:
-%
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{contr}}
+contains $n$ variables and that $\aclrd{s}=\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}$. First, consider
+the (\emph{contr}) rule:
+%
+\begin{equation*}
+\hspace{-1em}
 \begin{array}{l}
-\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em]
-\qquad  \lambda(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}+\cclrd{t}}), \mathbf{v_2}) \rightarrow \\[-0.25em]
-\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
-\qquad\qquad\qquad (\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}}), (x_{\cclrd{s}+1}, \ldots, x_{\cclrd{s}+\cclrd{t}} ), \mathbf{v_2})
-\end{array}
-\end{equation*}
-%
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{app}}
+\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{s} + \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
+  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
+  {\hspace{-0.5em}\begin{array}{l}\lambda(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}+\cclrd{t}}), \mathbf{v_2}) .\\[-0.25em]
+  \quad f~(\mathbf{v_1}, (x_1, \ldots, x_{\cclrd{s}}), (x_{\cclrd{s}+1}, \ldots, x_{\cclrd{s}+\cclrd{t}} ), \mathbf{v_2})\end{array}\hspace{-0.5em}~}
+\end{array}
+\end{equation*}
+%
+The semantic function is called with $\cclrd{s}+\cclrd{t}$ copies of
+a value for the $x$ variable. The values are split between $\cclrd{s}$ and $\cclrd{t}$ separate
+copies of variables $y$ and $z$, respectively. The (\emph{app}) rule is similar in that it needs to
+split the input variable context. However, it needs to split values of multiple variabless:
+%
+\begin{equation*}
+\hspace{-1em}
 \begin{array}{l}
-\sem{ \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} = \\[-0.25em]
-  \qquad\lambda (\mathbf{v_1}, ((x_{1,1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{n,\cclrd{t}\ast\cclrd{s}_n}) ) \rightarrow \\[-0.25em]
-  \qquad\qquad\kvd{let}~g = \sem{ \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }~\mathbf{v_1}\\[-0.25em]
-  \qquad\qquad\kvd{let}~\mathbf{y}_1 = ((x_{1,1}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{1,\cclrd{s}_n})  )\\[-0.25em]
-  \qquad\qquad\kvd{let}~\ldots \\[-0.25em]
-  \qquad\qquad\kvd{let}~\mathbf{y}_{ \cclrd{t} } = ((x_{1,(\cclrd{t}-1)\ast\cclrd{s}_1 + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, \\[-0.25em]
-  \hspace{8.2em}                                         (x_{n,(\cclrd{t}-1)\ast\cclrd{s}_n + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_n})  )\\[-0.25em]
-  \qquad\qquad\kvd{in}~g~(\sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_1, \ldots,
-    \sem{ \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}~\mathbf{y}_{\cclrd{t}})
-\\[1em]
-\end{array}
-\end{equation*}
-%
-In the (\emph{contr}) rule, the semantic function is called with $\cclrd{s}+\cclrd{t}$ copies of
-a value for the $x$ variable. The values are split between $\cclrd{s}$ and $\cclrd{t}$ separate
-copies of variables $y$ and $z$, respectively.
-
-The (\emph{app}) rule is similar in that it needs to split the input variable context. However,
-it needs to split values of multiple variables -- in $x_{i,j}$, the index $i$ stands for an
-index of the variable while $j$ is an index of one of multiple copies of the value. In the
-semantic function, the second
+\emdefff
+  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
+  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})}\\[-0.25em]
+    \qquad\vdash e_1 \, e_2 : \tau_2}\\[-0.25em]~\\[-0.25em]~\\[-0.25em]~\\[-0.25em]~\end{array}\hspace{-0.5em}~}
+  {f_1}
+  {f_2}
+  {\hspace{-0.5em}\begin{array}{l}
+  \lambda (\mathbf{v_1}, ((x_{1,1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), \ldots, (x_{n,1}, \ldots, x_{n,\cclrd{t}\ast\cclrd{s}_n}) ) .\\[-0.25em]
+  \quad(f_1~\textbf{v}_1)\\[-0.25em]
+  \qquad (~f_2~((x_{1,1}, \ldots, x_{1,\cclrd{s}_1}), \ldots, \\[-0.25em]
+  \hspace{4.0em}                                         ~(x_{n,1}, \ldots, x_{1,\cclrd{s}_n}) ~),\ldots,\\[-0.25em]
+  \qquad \;~f_2~((x_{1,(\cclrd{t}-1)\ast\cclrd{s}_1 + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_1}), ~\ldots~, \\[-0.25em]
+  \hspace{4.0em}                                         ~(x_{n,(\cclrd{t}-1)\ast\cclrd{s}_n + 1}, \ldots, x_{1,\cclrd{t}\ast\cclrd{s}_n})  ~)~)\\[-0.25em]
+  \end{array}}}
+\end{array}
+\end{equation*}
+\vspace{0.5em}
+
+\noindent
+In $x_{i,j}$, the index $i$ stands for an index of the variable while $j$ is an index of one of
+multiple copies of the value. In the semantic function, the second
 part of the context consists of $n$ variables where the multiplicity of each value is specified
@@
 (and omitting the terms), the weakening and contraction rules are written as follows:
-%
+
 \[
@@
 treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
-semantics} and \emph{meta-language} in Section~\ref{sec:unified-meta}.
+semantics} and \emph{meta-language} in Section~\ref{sec:further-meta}.
 
@@
 and unused variables (\cclrd{L}, \cclrd{D} and $1$, $0$, respectively) and in the operators used
-for sequential composition and contraction ($\sqcup$, $\sqcap$ and $\ast$, $+$, respectively).
+for sequential composition and contraction ($\cclrd{\sqcup}$, $\cclrd{\sqcap}$ and $\ast$, $+$, respectively).
 The algebraic structure capturing these operators is developed in Chapter~\ref{ch:structural}.
@@
 related to individual variables -- we could attach annotations to the variables themselves
-rather than attaching a \emph{vector} of annotations to the entire context. The main benefit
-of our approach is that it lets us unify flat and structural systems (Chapter~\ref{ch:unified}).
+rather than attaching a \emph{vector} of annotations to the entire context. One benefit
+of our approach is that it lets us unify flat and structural systems (Chapter~\ref{ch:further-unified}).
 
@@
 following example:
-
+%
 \begin{equation*}
@@
     \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} + \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
+  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \,\cclrd{+}\, \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
 \end{equation*}
@@
   {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}
-  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\textnormal{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
+  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{\textnormal{max}}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma}
 \end{equation*}
@@
 $0$, meaning that no past values are required. This is the same as in flat data-flow, but different
-from bounded reuse and liveness (where difference between using and not using a variable matters).
-Primitive requirements are introduced by the (\emph{prev}) rule, which increments the annotation of
-all variables in the context.
+from bounded reuse and liveness (where unused variables have a different coeffect). Primitive
+requirements are introduced by the (\emph{prev}) rule, which increments the annotations of
+all variables.
 
@@
 operations that were used in the (flat) application. In the structural system, these are used in
-(\emph{app}) and (\emph{contr}), respectively. Thus application combines coeffect annotations using
-$+$ and contraction using \emph{max}. This contrasts with bounded reuse, which uses $\ast$ and $+$,
+(\emph{app}) and (\emph{contr}). Thus application combines coeffect annotations using $+$ and
+contraction using \emph{max}. This contrasts with bounded reuse, which uses $\ast$ and $+$,
 respectively.
@@
 used just once), the pair provides past values (that can be reused and freely accessed).
-To illustrate the behaviour we consider the semantics of the \kvd{prev} construct and of the
-structural contraction rule:
-%
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{prev}}
+To illustrate the behaviour consider first the semantics of the \kvd{prev} expression:
+%
+\begin{equation*}
+\hspace{-1em}
 \begin{array}{l}
-\sem{\coctx{\Gamma}{\alift{(\cclrd{s}_1 + 1), \ldots, (\cclrd{s}_n + 1)}} \vdash \kvd{prev}~e : \tau} = \\[-0.25em]
-  \qquad\lambda ((x_{1,0}, \ldots, x_{1,{\cclrd{s}_1 + 1}}), \ldots, (x_{n,0}, \ldots, x_{n,{\cclrd{s}_n + 1}})) \rightarrow \\[-0.25em]
-\qquad\qquad\sem{\coctx{\Gamma}{\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}} \vdash e : \tau}\\[-0.25em]
-\qquad\qquad\qquad ((x_{1,0}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,0}, \ldots, x_{n,\cclrd{s}_n}))
-\end{array}
-\end{equation*}
-%
-\begin{equation*}
-\hspace{-0.5em}
-\tag{\emph{contr}}
-\begin{array}{l}
-\sem{\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\textnormal{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{\textbf{q}}} \vdash \subst{\subst{e}{z}{x}}{y}{x} : \sigma} =\\[-0.25em]
-\qquad  \lambda(\mathbf{v_1}, (x_0, x_1, \ldots, x_{ \textnormal{max}(\cclrd{s},\cclrd{t}) }), \mathbf{v_2}) \rightarrow \\[-0.25em]
-\qquad\qquad\sem{\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{\textbf{q}}} \vdash e : \sigma}\\[-0.25em]
-\qquad\qquad\qquad (\mathbf{v_1}, (x_0, \ldots, x_{\cclrd{s}}), (x_0, \ldots, x_{\cclrd{t}} ), \mathbf{v_2})
-\end{array}
-\end{equation*}
-%
-In (\emph{prev}), the semantic function is called with an argument that stores values of $n$
+\emdeff
+  {\coctx{\Gamma}{\alift{\cclrd{s}_1, \ldots, \cclrd{s}_n}} \vdash e : \tau}
+  {\begin{array}{l}\sem{\coctx{\Gamma}{\alift{(\cclrd{s}_1 \hspace{-0.2em}+\hspace{-0.2em} 1), \ldots, (\cclrd{s}_n \hspace{-0.2em}+\hspace{-0.2em} 1)}}\\[-0.25em]
+    \qquad\vdash \kvd{prev}~e : \tau}\end{array}\hspace{-0.5em}~}
+  {f}
+  {\hspace{-0.5em}\begin{array}{l}\lambda((x_{1,0}, \ldots, x_{1,{\cclrd{s}_1 + 1}}), \ldots, (x_{n,0}, \ldots, x_{n,{\cclrd{s}_n + 1}})).\\[-0.25em]
+  \quad f~((x_{1,0}, \ldots, x_{1,\cclrd{s}_1}), \ldots, (x_{n,0}, \ldots, x_{n,\cclrd{s}_n}))\end{array}\hspace{-0.5em}~}
+\end{array}
+\end{equation*}
+%
+Here, the semantic function is called with an argument that stores values of $n$
 variables, such that a variable $x_i$ has values ranging from $x_{i,0}$ to $x_{i,\cclrd{s}_i + 1}$.
 Thus, there is one current value, followed by $\cclrd{s_i} + 1$ past values. The expression $e$
-nested under $\kvd{prev}$ requires only $\cclrd{s_i}$ past values and so the semantics simply
-drops the last value.
-
-In the (\emph{contr}) rule, the semantic function receives $\emph{max}(\cclrd{s}, \cclrd{t})$ values
+nested under $\kvd{prev}$ requires only $\cclrd{s_i}$ past values and so the semantics
+drops the last value. The following shows the semantics of contraction:
+%
+\begin{equation*}
+\hspace{-1em}
+\begin{array}{l}
+\mdeff
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
+      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
+  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
+          \atimes \alift{\cclrd{\textnormal{max}}(\cclrd{s}, \cclrd{t})} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
+  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
+  {\hspace{-0.5em}\begin{array}{l}\lambda(\mathbf{v_1}, (x_0, x_1, \ldots, x_{ \textnormal{max}(\cclrd{s},\cclrd{t}) }), \mathbf{v_2}) .\\[-0.25em]
+  \quad f~((\mathbf{v_1}, (x_0, \ldots, x_{\cclrd{s}}), (x_0, \ldots, x_{\cclrd{t}} ), \mathbf{v_2}))\end{array}\hspace{-0.5em}~}
+\end{array}
+\end{equation*}
+%
+The semantic function receives $\emph{max}(\cclrd{s}, \cclrd{t})$ values
 of a specific variable $x$. It needs to produce values for two separate variables, $y$ and $z$ that
@@
 \cite{app-tainting-xss} and SQL injection attacks \cite{app-tainting-sql,app-tainting-wasp}.
-In the latter case, we want to check that SQL commands cannot be directly constructed from,
-potentially dangerous, inputs provided by the user. Consider the type checking of the following
+In the latter case, we want to check that SQL commands cannot be directly constructed from
+unchecked inputs provided by the user. Consider the type checking of the following
 expression in a context containing variables \ident{id} and \ident{msg}:
@@
 Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
-where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\ident{L}, \ident{H}\}$
-represents low and high secrecy, respectively with $\ident{L} \leq \ident{H}$ meaning that low security
+where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\cclrd{\ident{L}}, \ident{H}\}$
+represents low and high secrecy, respectively with $\cclrd{\ident{L}} \leq \ident{H}$ meaning that low security
 values can be treated as high security (but not the other way round).
@@
 \begin{equation*}
+\hspace{-2em}
 \tyrule{app}
@@
 using memory in that region and then deallocates the memory of the region:
-%
+
 \begin{equation*}
@@
 % --------------------------------------------------------------------------------------------------
+
+~
 
@@
 Chapter~\ref{ch:structural}, respectively). We can, however, unify the two variants into a single
-system discussed in Chapter~\ref{ch:unified}.
+system discussed in Section~\ref{ch:further-unified}.

### 2016-05-12 text/Chapter05.tex
Context-aware systems edits (Chapter 3)

@@
 
-\paragraph{Semantics of sub-coeffecting.}
-Although we do not include sub-coeffecting in the core flat coeffect calculus, it is an interesting
-extension to consider. Semantically, sub-coeffecting drops some of the available contextual
+\paragraph{Semantics of subcoeffecting.}
+Although we do not include subcoeffecting in the core flat coeffect calculus, it is an interesting
+extension to consider. Semantically, subcoeffecting drops some of the available contextual
 capabilities (drops some of the implicit parameters or some of the past values). This can be
@@
 the above property, the \ident{split} operation can be expressed in terms of lifting
-(sub-coeffecting) as follows:
+(subcoeffecting) as follows:
 %
@@
 This alternative definition is valid for our examples, but we do not use it for three reasons. First,
-it requires making sub-coeffecting a part of the core definition. Second, this would be the only
+it requires making subcoeffecting a part of the core definition. Second, this would be the only
 place where our semantics uses a variable \emph{twice} (in this case $c$). Note therefore that
@@
 to be dead (this is also the only solution of $\cclrd{\ident{D}} =\cclrd{r}\sqcap\cclrd{s}$), but when
-the input is \emph{live}, the operation can perform implicit sub-coeffecting and drop one of
+the input is \emph{live}, the operation can perform implicit subcoeffecting and drop one of
 the values.
@@
 
-\section{Conclusions}
+\section{Summary}
 

### 2016-05-12 text/Chapter06.tex
Context-aware systems edits (Chapter 3)

@@
 modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
-pointwise composition, as well as a relation $\cleq$ that defines sub-coeffecting.
+pointwise composition, as well as a relation $\cleq$ that defines subcoeffecting.
 
@@
 $\ctyp{\langle \ident{L} \rangle}{\alpha} \rightarrow \alpha$. The lifting operation models
-sub-coeffecting which can turn a context with a value into a dead context (second case); otherwise
+subcoeffecting which can turn a context with a value into a dead context (second case); otherwise
 it behaves as identity.

### 2016-05-12 text/Chapter08.tex
Context-aware systems edits (Chapter 3)

@@
 Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
-discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read
+discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-understanding}, we can also read
 the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
@@
 (and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:further-rev-contr}
-and Lemma~\ref{thm:further-rev-exch} relied on sub-coeffecting in the flat system.
+and Lemma~\ref{thm:further-rev-exch} relied on subcoeffecting in the flat system.
 
@@
 a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
-element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit sub-coeffecting
+element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit subcoeffecting
 to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and
@@
 transformation on one or two values in a context and then add the new value as a fresh variable
-to the variable context. We omit sub-coeffecting, but it could be easily added following
+to the variable context. We omit subcoeffecting, but it could be easily added following
 the method used elsewhere.

### 2016-05-12 text/Chapter03.tex
Fix references

@@
 $\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
-The unification is outlined in Section~\ref{ch:further-unified}.
+The unification is outlined in Section~\ref{sec:further-unified}.
 
@@
 determines the set of \emph{positions} in a vector. So, a vector of a shape (length) $n$ has positions
-$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Section~\ref{ch:further-unified} and also
+$\{ 1, 2, \ldots, n \}$. We discuss containers and shapes further in Section~\ref{sec:further-unified} and also
 discuss how our use relates to containers of Abbott, Altenkirch and Ghani \cite{types-containers}.
@@
 Chapter~\ref{ch:structural}, respectively). We can, however, unify the two variants into a single
-system discussed in Section~\ref{ch:further-unified}.
+system discussed in Section~\ref{sec:further-unified}.

### 2016-05-12 text/Chapter04.tex
Context-aware systems edits (Chapter 3)
Fix references

@@
 \emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing
-per-variable properties. As we show in Section~\ref{sec:unified-unified}, the systems can be
+per-variable properties. As we show in Section~\ref{sec:further-unified}, the systems can be
 further unified using a single abstraction, but such abstraction is \emph{less powerful} -- \ie~its
@@
 (Section~\ref{sec:flat-exts}) preserve the coeffect. Idempotence holds for all our examples
-and allows us to unify the flat and structural systems in Chapter~\ref{ch:unified}.
+and allows us to unify the flat and structural systems in Section~\ref{sec:further-unified}.
 
@@
 %---------------------------------------------------------------------------------------------------
-
+s
 \subsection{Understanding flat coeffects}
-\label{sec:flat-calculus-undestanding}
+\label{sec:flat-calculus-understanding}
 
@@
 % ==================================================================================================
-\section{Conclusions}
+\section{Summary}
 

### 2016-05-12 text/Chapter06.tex
Fix references

@@
 A benefit of this approach is that the typing judgements have the same structure as those of the
-flat coeffect calculus. As discussed in Chapter~\ref{ch:unified}, this makes it possible to unify
+flat coeffect calculus. As discussed in Section~\ref{sec:further-unified}, this makes it possible to unify
 the two systems.
@@
 More generally, we could see the the coeffect annotations as \emph{containers}
-\cite{types-containers}. This approach is used in Chapter~\ref{ch:unified} to unify flat
+\cite{types-containers}. This approach is used in Section~\ref{sec:further-unified} to unify flat
 and structural systems.

### 2016-05-12 responses/response.tex
Edits in implementation, consistency (Chapter 7)

@@
 is 10. Correspondingly, the type system only permits one typing -- with the parameter \ident{?x}
-required on the declaration-side (allowing no rebinding at the call-site). However, if the function
-value was defined in a context not containing \ident{?x}, the value from the call-site would be used.
+required on the declaration-side (allowing no rebinding at the call site). However, if the function
+value was defined in a context not containing \ident{?x}, the value from the call site would be used.
 Thus, Haskell uses a \emph{defaulting mechanism} for implicit parameters, similar to that of F\#
@@
 version) for the concrete systems, \eg~an upper bound on the number of required
-past values in causal data-flow.
+past values in causal dataflow.
 
@@
 context, but do not explicitly specify where the context comes from. In case of implicit parameters,
-the values can be provided either by the call-site or by the declaration-site.
+the values can be provided either by the call site or by the declaration-site.
 

### 2016-05-12 main.tex
Context-aware systems edits (Chapter 3)
Edits in implementation, consistency (Chapter 7)

@@
   We start with a detailed discussion of the motivation for the thesis and an overview of our
-  methodology (Chapter~\ref{ch:intro}). Next, we discuss previous programming langauge research that
+  methodology (Chapter~\ref{ch:intro}). Next, we discuss previous programming language research that
   leads to the work presented in this thesis (Chapter~\ref{ch:pathways}) and we examine a number
@@
 
-\include{text/Chapter01}
-\include{text/Chapter02}
-\include{text/Chapter03}
+%\include{text/Chapter01}
+%\include{text/Chapter02}
+%\include{text/Chapter03}
 
@@
 
-\ctparttext{In this part, we capture the similarities between the concrete context-aware langauges
+\ctparttext{In this part, we capture the similarities between the concrete context-aware languages
   presented in the previous chapter. We also develop the key novel technical contributions of
@@
 
-% ==================================================================================================
+% --------------------------------------------------------------------------------------------------
 
@@
 
-\ctparttext{ TBD }
+% ==================================================================================================
+
+\ctparttext{ In the first part of the thesis, we argued for the importance of \emph{context} in
+  programming languages. As programs execute in increasingly diverse and rich environments,
+  languages need to understand and check how programs use such context. In the second part, we
+  developed theoretical foundations (type system and semantics) for context-aware programming
+  languages. What remains to be done if context-aware programming languages are to become
+  ``the next big thing''?
+
+  In this part, we explore practical aspects of implementing context-aware programming languages
+  based on coeffects and related future work. We discuss a prototype implementation (Chapter~\ref{ch:impl}),
+  which links together all parts of the theory discussed in the previous part. Building a
+  production-ready programming language is outside the scope of the thesis, so we instead focus
+  on conveying the concept of coeffects to broader audience and make the implementation available
+  as a web-based interactive essay (Section~\ref{sec:impl-essay}) at: \url{http://tomasp.net/coeffects}.
+
+  In further work (Chapter~\ref{ch:further}), we outline unification of flat and structural
+  coeffects (Section~\ref{sec:further-unified}), which may be more suited for embedding in practical
+  languages and we discuss alternative approaches for using coeffects in programming languages
+  (\ref{sec:further-meta}).
+}
+
 \part{Towards practical coeffects}
@@
 
-%\include{text/Chapter07}
-%\include{text/Chapter08}
-%\include{text/Chapter09}
+\include{text/Chapter07}
+\include{text/Chapter08}
+\include{text/Chapter09}
 

### 2016-05-12 text/Abstract.tex
Edits in implementation, consistency (Chapter 7)

@@
 variable usage (\eg~in liveness analaysis and linear logics) or past values in 
-stream-based data-flow programming.
+stream-based dataflow programming.
 

### 2016-05-12 text/AppendixA.tex
Edits in implementation, consistency (Chapter 7)

@@
 This appendix gives full typing derivations for the examples presented in Chapter~\ref{ch:applications},
-which demonstrate simple coeffect calculi for implicit parameters, liveness and data-flow.
+which demonstrate simple coeffect calculi for implicit parameters, liveness and dataflow.
 
@@
 
-\section{Coeffect typing for data-flow}
+\section{Coeffect typing for dataflow}
 \label{sec:appendixa-dataflow}
@@
 The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
-maximal number of past values required by a data-flow computation. The discussion includes the 
+maximal number of past values required by a dataflow computation. The discussion includes the 
 following example:
@@
 
-\figcaption{Coeffect typing for data-flow}
+\figcaption{Coeffect typing for dataflow}
 \label{fig:appendixa-dfl}

### 2016-05-12 text/Chapter01.tex
Edits in implementation, consistency (Chapter 7)

@@
 The final example leaves the topic of cross-platform and distributed computing. We focus on
-checking how arrays are accessed. This is a simpler version of the data-flow programming
+checking how arrays are accessed. This is a simpler version of the dataflow programming
 examples used later in the thesis.
@@
   to context. The chapter presents type systems and semantics for a number of systems and analyses
-  (including data-flow, liveness analysis, distributed programming and Haskell's type classes).
+  (including dataflow, liveness analysis, distributed programming and Haskell's type classes).
   Our novel presentation reveals their similarity.

### 2016-05-12 text/Chapter02.tex
Context-aware systems edits (Chapter 3)
Edits in implementation, consistency (Chapter 7)

@@
 \begin{itemize}
-\item \textsc{Static (lexical) binding.} In a langauge with static binding (such as ML or Java),
+\item \textsc{Static (lexical) binding.} In a language with static binding (such as ML or Java),
   the variable \ident{x} inside the body of the lambda function is statically bound to the
@@
 which the function is defined (declaration-site). In dynamic binding, the requirements are
-\emph{delayed} and are placed on the scope in which the function is called (call-site).
+\emph{delayed} and are placed on the scope in which the function is called (call site).
 
@@
 \subsection{Implicit parameter binding}
+\label{sec:path-binding-impl}
 
@@
 We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
-implicit parameter is defined both at declaration-site and at the call-site. On the right, the
-implicit parameter is available only at the call-site:
+implicit parameter is defined both at declaration-site and at the call site. On the right, the
+implicit parameter is available only at the call site:
 
@@
 is statically bound and the context requirement is satisfied using the declaration-site context.
-Otherwise, the context requirement is delayed and has to be satisfied at the call-site.
+Otherwise, the context requirement is delayed and has to be satisfied at the call site.
 In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
@@
 a function can be satisfied using the context available at declaration-site or using the context
-available at the call-site.
+available at the call site.
 
 We also aim to find a description of context-aware languages that does not make ad-hoc decisions
-about how context requirements are split between the declaration-site and the call-site. While
+about how context requirements are split between the declaration-site and the call site. While
 Haskell's solution for implicit parameters might be the most reasonable one, different notions
@@
 $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
-value from a hidden dictionary that is passed together with the input from the call-site.
+value from a hidden dictionary that is passed together with the input from the call site.
 
@@
 The theoretical parts of this thesis extend the work of Uustalu and Vene who use comonads
-to give the semantics of data-flow computations \cite{comonads-dataflow} and more generally,
+to give the semantics of dataflow computations \cite{comonads-dataflow} and more generally,
 notions of \emph{context-dependent computations} \cite{comonads-notions}. The computations discussed
@@
 
-In causal data-flow, we can interpret the list as a list consisting of past values, with the
+In causal dataflow, we can interpret the list as a list consisting of past values, with the
 current value in the head. Then, the $\ident{cobind}$ operation calculates the current value
@@
 input $(l_n)$. In addition to the operations of comonad, the model also uses some operations that
-are specific to causal data-flow:
+are specific to causal dataflow:
 %
@@
 %
-The operation drops the first element from the list. In the data-flow interpretation, this means
+The operation drops the first element from the list. In the dataflow interpretation, this means
 that it returns the previous state of a value.
 
-Now, consider a simple data-flow language with single-variable contexts, variables,
+Now, consider a simple dataflow language with single-variable contexts, variables,
 primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
@@
 has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
-the language models data-flow and values are lists (of past values) is a matter of semantics,
+the language models dataflow and values are lists (of past values) is a matter of semantics,
 which is defined as follows:

### 2016-05-12 text/Chapter03.tex
Edits in implementation, consistency (Chapter 7)

@@
   The systems are adapted from diverse sources (type systems, static analyses, logics) and apply to
-  various domains (cross-compilation, liveness, distributed computing, data-flow, security), but
+  various domains (cross-compilation, liveness, distributed computing, dataflow, security), but
   share a common structure.
@@
 \item In addition, the coeffect systems for tracking the number of accessed past values in
-  data-flow languages (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
-  presents novel results and can be used to optimize data-flow programs.
+  dataflow languages (Sections~\ref{sec:applications-flat-dataflow} and \ref{sec:applications-structural-dataflow})
+  presents novel results and can be used to optimize dataflow programs.
 \end{itemize}
@@
 cross-compilation (Section~\ref{sec:applications-flat-distr}), liveness analysis
-(Section~\ref{sec:applications-flat-live}) and data-flow (Section~\ref{sec:applications-flat-dataflow}).
+(Section~\ref{sec:applications-flat-live}) and dataflow (Section~\ref{sec:applications-flat-dataflow}).
 
@@
 
-\subsection{Data-flow languages}
+\subsection{Dataflow languages}
 \label{sec:applications-flat-dataflow}
@@
 location in the array specified by a cursor. In this section, we make the example more concrete,
-using a simpler and better studied programming model, data-flow languages.
-
-Lucid \cite{app-lucid} is a declarative data-flow language designed by Wadge and Ashcroft. In Lucid,
+using a simpler and better studied programming model, dataflow languages.
+
+Lucid \cite{app-lucid} is a declarative dataflow language designed by Wadge and Ashcroft. In Lucid,
 variables represent streams and programs are written as transformations over streams. A function
@@
 
-The data-flow approach has been successfully used in domains such as development of real-time embedded
-application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the data-flow
+The dataflow approach has been successfully used in domains such as development of real-time embedded
+application where many \emph{synchronous languages} \cite{app-synchronous-lang} build on the dataflow
 paradigm. The following example is inspired by the Lustre \cite{app-synchronous-lustre} language
@@
 %
-When reading data-flow programs, we do not need to think about variables in terms of streams --
+When reading dataflow programs, we do not need to think about variables in terms of streams --
 we can see them as simple values. Most of the operations perform calculation just on the
@@
 and Vene later used comonads \cite{app-dataflow-essence}. In Chapter~\ref{ch:flat}, we extend
-the latter approach. The present chapter presents a sketch of a concrete data-flow semantics
+the latter approach. The present chapter presents a sketch of a concrete dataflow semantics
 defined directly on streams.
@@
 \kvd{prev} operation and track the maximal number of \emph{past values} needed. This is an
-important information for efficient implementation of data-flow languages. When we can guarantee
+important information for efficient implementation of dataflow languages. When we can guarantee
 that at most $x$ past values are accessed, the values can be stored in a pre-allocated buffer
@@
 
-\figcaption{Coeffect rules for tracking context-usage in data-flow language}
+\figcaption{Coeffect rules for tracking context-usage in dataflow language}
 \label{fig:applications-flat-dataflow}
@@
 The (\emph{app}) rule follows the same intuition as for liveness. It combines \emph{sequential}
-and \emph{pointwise} composition of semantic functions. In case of data-flow, the operations
+and \emph{pointwise} composition of semantic functions. In case of dataflow, the operations
 combine annotations using $+$ and \emph{max} operations:
@@
 \paragraph{Semantics.}
-The language discussed in this section is a \emph{causal} data-flow language. This means that
+The language discussed in this section is a \emph{causal} dataflow language. This means that
 a computation can access \emph{past} values of the stream (but not future values). In the semantics,
@@
 
-Uustalu and Vene \cite{comonads-notions} model causal data-flow computations using a non-empty list
+Uustalu and Vene \cite{comonads-notions} model causal dataflow computations using a non-empty list
 $\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$ over the input. A function $\tau_1 \rightarrow \tau_2$
@@
 
-\figcaption{Semantics showing how past values are accessed in a data-flow language}
+\figcaption{Semantics showing how past values are accessed in a dataflow language}
 \label{fig:applications-flat-dfsem}
@@
 \paragraph{Summary.}
-Type systems have been used in the context of data-flow languages, for example to check
+Type systems have been used in the context of dataflow languages, for example to check
 initialization properties \cite{app-dataflow-init}, but to our knowledge, not for checking
@@
 
-The most interesting point about the data-flow system is that it is remarkably similar to our
+The most interesting point about the dataflow system is that it is remarkably similar to our
 earlier liveness example. In the type system, abstraction (\emph{abs}) duplicates the context
@@
 \subsection{Permissions and safe locking}
-In the implicit parameters and data-flow examples, the context provides additional resources or
+In the implicit parameters and dataflow examples, the context provides additional resources or
 values that may be accessed at runtime. However, coeffects can also track \emph{permissions} or
@@
 useful structural version of the system; we consider an example inspired by linear logic; finally,
-we revisit data-flow to get a more precise analysis. Although quite different, the common pattern
+we revisit dataflow to get a more precise analysis. Although quite different, the common pattern
 among these three examples is somewhat easier to see, because they all track information about
@@
 
-\subsection{Data-flow languages revisited}
+\subsection{Dataflow languages revisited}
 \label{sec:applications-structural-dataflow}
 
-When discussing data-flow languages in an earlier section, we said that the context provides
+When discussing dataflow languages in an earlier section, we said that the context provides
 past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as
 a \emph{flat} property, which gives us a system that keeps the same number of past values for
-all variables. However, data-flow can also be adapted to a structural system which keeps the number
+all variables. However, dataflow can also be adapted to a structural system which keeps the number
 of required past values individually for each variable. Consider the
@@
 
-Turning a flat data-flow system to a structural data-flow system is a change similar to the one
+Turning a flat dataflow system to a structural dataflow system is a change similar to the one
 between flat ans structural liveness. In case of liveness analysis, we included the flat system
-only as an illustration (it is not practically useful). For data-flow, the flat system is less
+only as an illustration (it is not practically useful). For dataflow, the flat system is less
 precise, but still practically useful (simplicity may outweigh precision).
@@
 
-In data-flow, we annotate both used variables (\emph{var}) and unused variables (\emph{weak}) with
-$0$, meaning that no past values are required. This is the same as in flat data-flow, but different
+In dataflow, we annotate both used variables (\emph{var}) and unused variables (\emph{weak}) with
+$0$, meaning that no past values are required. This is the same as in flat dataflow, but different
 from bounded reuse and liveness (where unused variables have a different coeffect). Primitive
@@
 
-In flat data-flow, we identified sequential composition and pointwise composition as two primitive
+In flat dataflow, we identified sequential composition and pointwise composition as two primitive
 operations that were used in the (flat) application. In the structural system, these are used in
@@
 \paragraph{Semantics.}
-To define the semantics of our structural data-flow language, we can use the same approach as when
+To define the semantics of our structural dataflow language, we can use the same approach as when
 adapting flat liveness to structural liveness. Rather than wrapping the whole context in a type
@@
 \paragraph{Summary.}
-Two of the structural examples shown so far (liveness and data-flow) extend an earlier flat
+Two of the structural examples shown so far (liveness and dataflow) extend an earlier flat
 version of a similar system. We discuss this relation in general later. However, a flat system
@@
 
-The data-flow example demonstrates that the a flat system can also be turned into structural
+The dataflow example demonstrates that the a flat system can also be turned into structural
 system. In general, this only works for systems where lambda abstraction duplicates context
@@
 \emph{change} the context -- not just by evaluating certain code block in a locally modified context
-(\eg by wrapping it in $\ident{prev}$ in data-flow), but also by calling a function that, for example,
+(\eg by wrapping it in $\ident{prev}$ in dataflow), but also by calling a function that, for example,
 acquires new capabilities and returns those to the caller. Such actions appear to be closer to
@@
 as well as approaches based on monads. We looked at a number of applications including Haskell's implicit
-parameters and type classes, data-flow languages such as Lucid, liveness analysis and also a number of
+parameters and type classes, dataflow languages such as Lucid, liveness analysis and also a number of
 security properties.

### 2016-05-12 text/Chapter04.tex
Edits in implementation, consistency (Chapter 7)

@@
 The liveness example is interesting because it does not require any additional syntactic extensions
-to the langauge. It annotates constants and variables with $\ident{D}$ and $\ident{L}$, respectively
+to the language. It annotates constants and variables with $\ident{D}$ and $\ident{L}$, respectively
 and it captures how those annotation propagate through the remaining language constructs.
@@
 
-A simple dataflow langauge includes an additional construct $\kvd{prev}$ for accessing the
+A simple dataflow language includes an additional construct $\kvd{prev}$ for accessing the
 previous value in a stream with an additional typing rule that look as follows:
@@
 In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
-can be satisfied either by the call-site or by the declaration-site. In case of dataflow and
+can be satisfied either by the call site or by the declaration-site. In case of dataflow and
 liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
@@
   \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
-    parameter \ident{?x} is required from the call-site and the program produces $2$.
+    parameter \ident{?x} is required from the call site and the program produces $2$.
 \end{itemize}
@@
 where function abstraction captures all parameters that are statically available at the declaration-site
-and places all other demands on the call-site. For the example above, this means
+and places all other demands on the call site. For the example above, this means
 that the body of $f$ captures the value of \ident{?p} available from the declaration-site
@@
 and the latent coeffect attached to the function becomes $\cclrd{r\setminus\Delta}$ (require any
-remaining parameters from the call-site). Finally, in the (\emph{letpar}) rule, we add the newly
+remaining parameters from the call site). Finally, in the (\emph{letpar}) rule, we add the newly
 bound implicit parameter \ident{?p} to the static scope in the sub-expression $e_2$.
@@
 (type preservation), then operational semantics can be defined as a repeated application of the
-rules, together with additional domain-specific rules for each context-aware langauge, until
+rules, together with additional domain-specific rules for each context-aware language, until
 a specified normal form (\ie~a value) is reached.
@@
 abstraction. The body of a function requires certain context, but the context can be often
-provided by either the declaration-site or the call-site. Resolving this ambiguity has to be
+provided by either the declaration-site or the call site. Resolving this ambiguity has to be
 done differently for each concrete coeffect system, depending on its specific notion of context.

### 2016-05-12 text/Chapter05.tex
Edits in implementation, consistency (Chapter 7)

@@
 (\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent
-computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
+computations~\cite{comonads-notions}. For example, dataflow computations over non-empty lists
 are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
@@
 
-As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
+As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model dataflow
 computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.
@@
 type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
-%
+
 \begin{equation*}
@@
 the \emph{least upper-bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
-bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
+bounds in the liveness and dataflow examples, but not for implicit parameters. However, they remain useful
 as an informal analogy.}.
@@
 
-Our last example provides the semantics of data-flow computations. The flat coeffect algebra
+Our last example provides the semantics of dataflow computations. The flat coeffect algebra
 is formed by $(\mathbb{N}, +, \mathit{max}, \mathit{min}, 0, 0, \leq)$. In a
@@
 The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
-of the $f$ function, which itself requires $m$ past values. When combining two lists,
-$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter
-argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths.
+of $f$, which itself requires $m$ past values. When combining two lists, $\ident{merge}_{m,n}$
+behaves as \emph{zip} and produces a list of length of the shorter argument. When splitting a list,
+$\ident{split}_{m, n}$ needs the maximum of the lengths.
 
@@
 \begin{equation*}
-\hspace{-3em}
+\hspace{-2em}
 \begin{array}{ll}
@@
   ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
-  \end{array}\hspace{-0.5em}}
+  \end{array}\hspace{-2em}~}
 & (\emph{app})
@@
 In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
-computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed
+computations including implicit parameters, liveness and dataflow. Using the \emph{flat indexed
 comonad} structure, we can now define a single uniform semantics that is capable of capturing all
@@
 \begin{proof}
-Simple expansion of the definitions for the unique typing derivation.
+Expansion of the definitions, using the unique typing derivation
+for dataflow and liveness and any typing derivation for implicit parameters.
 \end{proof}
@@
 
-\caption{Common syntax and reduction rules of the target language}
+\figcaption{Common syntax and reduction rules of the target language}
 \label{fig:semantics-target}
@@
 
-\caption{Comonadically-inspired extensions for the target language}
+\figcaption{Comonadically-inspired extensions for the target language}
 \label{fig:semantics-ext}
@@
 
-\caption{Additional constructs for modelling dataflow in the target language}
+\figcaption{Additional constructs for modelling dataflow in the target language}
 \label{fig:semantics-ext-df}
@@
 for (\emph{cobind}) and (\emph{split}) restrict the set of available implicit parameters according
-to the annotations and (\emph{merge}) combines them, preferring the values from the call-site.
-
-For the semantics of implicit parameter programs that we consider, the preference of call-site
+to the annotations and (\emph{merge}) combines them, preferring the values from the call site.
+
+For the semantics of implicit parameter programs that we consider, the preference of call site
 bindings over declaration-site bindings in (\emph{merge}) does not matter. The unique typing
@@
 \begin{figure}[h!]
+\vspace{2em}
 \paragraph{Language syntax}
@@
 
-\vspace{-0.5em}
-\caption{Additional constructs embedding implicit parameters into the language}
+\figcaption{Additional constructs embedding implicit parameters into the language}
 \label{fig:semantics-ext-impl}
+\vspace{1em}
 \end{figure}
@@
 
-\begin{theorem}[Safety of context-aware langauge with implicit parameters]
+\begin{theorem}[Safety of context-aware language with implicit parameters]
   If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^* e'$ then either $e'$ is a value of type $\tau$ or
@@
 
-In Section~\ref{sec:semantics-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
+In Section~\ref{sec:semantics-proofs-df} and Section~\ref{sec:semantics-proofs-impl}, we proved the
 safety property of two concrete context-aware programming languages based on the coeffect

### 2016-05-12 text/Chapter06.tex
Edits in implementation, consistency (Chapter 7)

@@
   uninterpreted comonadically-inspired primitivies. We give concrete operational semantics for the
-  target language for one of our sample langauges and show that well-typed programs, produced by
+  target language for one of our sample languages and show that well-typed programs, produced by
   translation from the coeffect calculus do not get stuck.
@@
 implicit parameters), where lambda abstraction allowed arbitrary splitting of context requirements.
-In structural coeffect systems, the context requirements placed on the call-site (attached to
+In structural coeffect systems, the context requirements placed on the call site (attached to
 the function type) are those of the bound variable.
@@
 the eqivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
-as specifying translation to a target functional langauge in Section~\ref{sec:struct-transl}.
+as specifying translation to a target functional language in Section~\ref{sec:struct-transl}.
 
@@
 \begin{proof}
-Simple expansion of the definitions for the unique typing derivation.
+Expansion of the definitions for the unique typing derivation.
 \end{proof}
@@
 indexed comonads (Section~\ref{sec:struct-transl-ext}), define the translation to the target
-langauge (Section~\ref{sec:struct-transl-transl}) and prove safety of one sample language
+language (Section~\ref{sec:struct-transl-transl}) and prove safety of one sample language
 (Section~\ref{sec:struct-transl-safety}) -- we choose structural dataflow to allow easy
@@
 
-\caption{Comonadically-inspired extensions for structural coeffects}
+\figcaption{Comonadically-inspired extensions for structural coeffects}
 \label{fig:struct-semantics-ext}
@@
 The target language with comonadically-inspired primitives provides a framework that can be used
-to model a variety of structural context-aware langauges and prove their safety. As outlined in
+to model a variety of structural context-aware languages and prove their safety. As outlined in
 Section~\ref{sec:semantics-generalising}, the key principle that guarantees the safety of the
@@
 
-\caption{Additional constructs for modelling structural dataflow}
+\figcaption{Additional constructs for modelling structural dataflow}
 \label{fig:structural-ext-df}
@@
 
-The Figure~\ref{fig:structural-ext-df} extends the target functional langauge with constructs,
+The Figure~\ref{fig:structural-ext-df} extends the target functional language with constructs,
 typing rules, translation rules and reduction rules needed for modelling structural dataflow.

### 2016-05-12 text/Chapter07.tex
Edits in implementation, consistency (Chapter 7)

--- 2014-11-17 text/Chapter07.tex
@@
-%!TEX root = ../main.tex
+% ==================================================================================================
 
-\chapter{Conclusions} 
-\label{ch:conclusions}
+\chapter{Implementation}
+\label{ch:impl}
+
+In the previous three chapters, we presented two coeffect calculi that capture two kinds of
+contextual properties. The calculi are parameterized and can be instantiate to capture concrete
+notions of context. They consist of type systems (parameterized by coeffect algebra) and semantics
+(parameterized by small number of comonadically-inspired primitives). The theory can be seen as
+a framework that simplifies the implementation of safe context-aware programming languages. To
+support this claim, this chapter presents a prototype implementation of the coeffect framework and
+uses it to build three concrete languages -- language with implicit parameters and both flat and
+structural versions of a dataflow language.
+
+The implementation directly follows the thoery presented in Chapters~\ref{ch:flat}, \ref{ch:semantics}
+and \ref{ch:structural}. It consists of a common framework that provides type checking and
+translation to a simple functional target language with comonadically-inspired primitives. Each
+concrete context-aware language then adds a domain-specific rule for choosing a unique typing
+derivation (as discussed in Section~\ref{sec:flat-unique}) together with a domain-specific
+definition of the comonadically-inspired primitives that define the runtime semantics
+(see Section~\ref{sec:semantics-proofs}).
+
+The main goal of the implementation is to show that the theory is practically useful and to present it
+in a more practical way. However, we do not intend to build a complete real-world programming language.
+For this reason, the implementation is available primarilly as an interactive essay\footnote{The
+\emph{interactive essay} format is based on Bret Victor's work on \emph{explorable explanations} \cite{essay-explorable}
+and is further explained in Section~\ref{sec:impl-essay-features}} online at \url{http://tomasp.net/coeffects}.
+
+\paragraph{Chapter structure and contributions}
+
+\begin{itemize}
+\item We discuss how the implementation follows the theory (Section~\ref{sec:impl-theory}) presented earlier.
+  This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
+  to a simple target language that is then iterpereted. We also discuss how the common framework makes it
+  easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).
+
+\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate interesting
+  aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
+  difference between flat and structural systems (Section~\ref{sec:impl-case-typing}) and the
+  comonadically-inspired translation (Section~\ref{sec:impl-case-transl}).
+
+\item The implementation is available not just as downloadable code, but also in the format of
+  interactive essay (Section~\ref{sec:impl-essay}), which aims to make coeffects
+  accessible to a broader audience. We discuss the most interesting aspects of the web-based
+  presentation and briefly discuss some of the interesting implementation detail
+  (Section~\ref{sec:impl-essay-tech}).
+\end{itemize}
 
@@
 %
-%       ###                                            
-%      #   #                                           
-%      #      #   #  ## #   ## #    ###   # ##   #   # 
-%       ###   #   #  # # #  # # #      #  ##  #  #   # 
-%          #  #   #  # # #  # # #   ####  #      #  ## 
-%      #   #  #  ##  # # #  # # #  #   #  #       ## # 
-%       ###    ## #  #   #  #   #   ####  #          # 
-%                                                #   # 
-%                                                 ###  
-% =================================================================================================
+%    #######
+%       #    #    # ######  ####  #####  #   #
+%       #    #    # #      #    # #    #  # #
+%       #    ###### #####  #    # #    #   #
+%       #    #    # #      #    # #####    #
+%       #    #    # #      #    # #   #    #
+%       #    #    # ######  ####  #    #   #
+%
+% ==================================================================================================
 
-Some of the most fundamental academic work is not the one solving hard research problems, but the 
-one that changes how we understand the world. Some philosophers argue that \emph{language} is the 
-key for understanding how we think, while in science the dominant thinking is determined by 
-\emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
-In a way, programming languages play a similar role for computer science and software development.
+\section{From theory to implementation}
+\label{sec:impl-theory}
 
-This thesis aims to change how developers and programming language designers think about 
-\emph{context} or \emph{execution environments} of programs. Such context or execution environment
-has numerous forms -- mobile applications access network, GPS location or user's personal data; 
-journalists obtain information published on the web or through open government data initiatives. 
-This thesis aims to change our understanding of \emph{context} so that the above examples are 
-viewed uniformly through a single programming language abstraction, which we call \emph{coeffects}, 
-rather than as disjoint cases. This is done through the following three contributions.
+The theory discussed so far provides the two key components of the implementation. In
+Chapter~\ref{ch:flat}, we discussed the type checking of context-aware programs and
+Chapter~\ref{ch:semantics} models the execution of context-aware programs (in terms of
+translation and operational semantics). For structural coeffects, the same components are discussed
+in Chapter~\ref{ch:structural}. In this section, we discuss how teose provide foundation for the
+implementation.
 
@@
 
-\section{Unified presentation.}
+\subsection{Type checking and inference}
+\label{sec:impl-theory-typing}
 
-In $\lambda$-calculus, the term \emph{context} is usually used for the free-variable context.
-However, there are many other programming language features that are related to context or program's
-execution environment. In Chapter~\ref{ch:applications}, we revisit many of such features --
-including resource tracking in distributed computing, cross-platform development, data-flow 
-programming and liveness analysis, but also Haskell's type classes and implicit parameters.
+To simplify the writing of context-aware programs, the implementation provides a limited form of
+type inference (in contrast, previous chapters described just type \emph{checking}). This is
+available for convenience. We do not claim any completeness result about the algorithm and we
+do not present full formalization. However, it is worth noting how the domain-specific procedures
+for choosing a unique type derivation (Section~\ref{sec:flat-unique}) are adapted.
 
-The main contribution of the chapter is that it presents the disjoint language features in a 
-unified way. We show type systems and semantics for many of the languages, illuminating
-the fact that they are closely related.
+The type inference works in the standard way \cite{types-mlessence,types-inference} by generating
+type constraints and solving them. Solving of type constraints is done in the standard way, but we
+additionally collect and solve \emph{coeffect constraints}. In this section, we focus only on
+coeffect constraints (as generation and solving of type constraints is standard). The following
+(\emph{abs}) rule demonstrates the notation used in this section:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s} \}}
+\end{equation*}
+%
+The judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau ~|~ C}$ denotes that an expression $e$
+in a context $\coctx{\Gamma}{\cclrd{r}}$ has a type $\tau$ and produces coeffect constraints $C$.
+In the (\emph{abs}) rule, we annotate the function body, function type and declaration site with
+new coeffect variables $\cclrd{r}, \cclrd{t}$ and $\cclrd{s}$, respectively and we generate a
+coeffect constraint $\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}$ that captures the (\emph{abs}) rule
+from flat coeffect calculus (Figure~\ref{fig:flat-types}).
 
-Considering applications is one way of approaching the theory of coeffects introduced in this thesis.
-Other pathways to coeffects are discussed in Chapter~\ref{ch:pathways}, which looks at theoretical 
-developments leading to coeffects, including the work on effect systems, comonadic semantics and 
-linear logics.
+In order to obtain unique type derivation for each term (using the algorithms discussed in
+Section~\ref{sec:flat-unique}), we generate additional coeffect constraints for lambda abstraction
+of each flat coeffect language, as this is where flat coeffects permit multiple possible typings.
+
+\begin{example}[Flat implicit parameters.]
+As discussed in Section~\ref{sec:flat-unique}, when choosing unique typing derivation for
+implicit parameters, we keep track of the implicit parameters available in the lexical scope
+(written as $\cclrd{\Delta}$). In the lambda abstraction rule, the implicit parameters required by
+the body (tracked by $\ccclrd{r}$) are split so that all parameters available in lexical scope
+are captured and only the remaining parameters ($\cclrd{r}\setminus\cclrd{\Delta}$) are required
+from the caller of the function.
+
+From the presentation in Section~\ref{sec:flat-unique}, it might appear that resolving the ambiguity
+related to lambda abstraction for implicit parameters requires a type system different from
+the type system shown earlier in Section~\ref{sec:flat-calculus-types}. This is not the case.
+We track implicit parameters in scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from
+the implementation only generates an additional coeffect constraint. The adapted (\emph{abs}) rule
+for implicit parameters looks as follows:
+%
+\begin{equation*}
+\hspace{-2em}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta}, \cclrd{s}=\cclrd{t}\cclrd{\setminus \Delta} \}}
+\end{equation*}
+%
+Given a typing derivation for the body, we generate an additional constraint that restricts
+$\cclrd{r}$ (declaration site demands) to those available in the current static scope
+$\cclrd{\Delta}$ and a constraint that restricts the delayed (call site) demands
+$\cclrd{s}$ to $\cclrd{t}\cclrd{\setminus \Delta}$.
+\end{example}
+
+\begin{example}[Flat dataflow]
+In a context-aware language for dataflow (and in language with liveness tracking), the inherent
+ambiguity of the (\emph{abs}) rule is resolved by placing the context requirements of the body on
+both the declaration site and the call site. In Section~\ref{sec:flat-unique}, this was defined by
+replacing the standard coeffect (\emph{abs}) rule with a rule (\emph{idabs}) that uses an
+annotation $\cclrd{r}$ for the body of the function, declaration site coeffect and call site coeffect.
+
+As with implicit parameters, the implementation does not require changing the core (\emph{abs})
+typing rule of the flat coeffect system. Instead, the unique resolution is obtained by generating
+additional coeffect constraints:
+%
+\begin{equation*}
+\tyrule{abs}
+  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
+  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
+    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}}
+\end{equation*}
+%
+Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
+coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by
+(\emph{idabs}).
+\end{example}
 
@@
 
-\section{Flat coeffect calculus.}
+\subsection{Execution of context-aware programs}
+\label{sec:impl-theory-transl}
 
-The applications discussed in Chapter~\ref{ch:applications} fall into two categories. In the first
-category (Section~\ref{sec:applications-flat}), the additional contextual information are 
-\emph{whole-context} properties. They either capture properties of the execution environment or
-affect the whole free-variable context.
+Context-aware programs are executed by translating the source program into a simple functional
+target language with domain-specific primitives. For simplicity, programs in the simple target
+language are then interpreted, but they could equally be compiled using standard techniques for
+compiling functional code. The translation follows the rules defined in Section~\ref{sec:semantics-translation}
+(for flat coeffect languages) and Section~\ref{sec:struct-transl} (for structural coeffect
+languages). The result of the translation is a program that consists of the following:
 
-In Chapter~\ref{ch:flat}, we develop a \emph{flat coeffect calculus} which gives us an unified way
-of tracking \emph{whole-context} properties. The calculus is parameterized by a \emph{flat coeffect
-algebra} that captures the algebraic properties that contextual information satisfy. Concrete 
-instances of flat coeffects include Haskell's implicit parameters, whole-context liveness and
-whole-context data-flow.
+\begin{itemize}
+\item {\sc Functional constructs.} Those include binary operations, tuples, let binding,
+  constants, variables, function abstraction and application. The interpreter keeps a map of
+  assignments for variables in scope and recursively evaluates the expression.
 
-Our focus is on the syntactic properties of the calculus -- we give a type system 
-(Section~\ref{sec:flat-calculus}) and discuss equational theory of the calculus (Section~\ref{sec:flat-syntax}).
-In the flat coeffect calculus, $\beta$-reduction and $\eta$-expansion do not generally preserve
-type of expressions, but we identify two conditions when this is the case -- this gives us
-a basis for operational semantics of our calculi for liveness and implicit parameters. 
-The design of the type system is further validated by categorical semantics
-(Section~\ref{sec:flat-semantics}) which models flat coeffect calculus in terms of 
-\emph{flat indexed comonad} -- a structure based on categorical dual of monads.
+\item {\sc Comonadic operations.} Those are the comonadic primitives provided by indexed comonads --
+  \ident{cobind}, \ident{counit} together with \ident{merge} and \ident{split} for flat coeffects
+  or \ident{merge} and \ident{choose} for structural coeffects. The translation that produces these
+  is shared by all context-aware languages, but their definition in the interpreter is
+  domain-specific.
+
+\item {\sc Domain-specific operations.} Each context-aware language may additionally include
+  operations that model domain-specific operations. For dataflow, this is \ident{prev} (accessing
+  past values) and for implicit parameters, this is \ident{letimpl} and \ident{lookup} for implicit
+  parameter binding and access, respectively.
+\end{itemize}
+
+\noindent
+The fact that the prototype implementation is based on the theoretical framework provided by
+coeffect calculi means that it has the desirable properties proved in Section~\ref{sec:semantics-proofs}
+and Section~\ref{sec:struct-transl}. In particular, evaluating a well-typed context-aware program in
+a context that provides sufficient contextual capabilities will not cause an error.
+
+In the interactive essay (Section~\ref{sec:impl-essay}), we further use the coeffects to
+automatically generate a user interface that requires the user to provide the required contextual
+capabilities (past values for individual variables, or values for implicit parameters).
+
+Another benefit of using the common framework is that the implementation can be easily extended
+to support further languages with additional contextual properties.
 
@@
 
-\section{Structural coeffect calculus.}
+\subsection{Supporting additional context-aware languages}
+\label{sec:impl-theory-ext}
 
-In the second category of context-aware systems discussed in Section~\ref{sec:applications-structural},
-contextual properties are \emph{per-variable}. The systems discussed here resemble sub-structural
-logics, but rather than \emph{restricting} variable use, they \emph{track} additional information
-about how variables are used.
+The prototype implementation supports two of the context-aware languages discussed in this thesis:
+implicit parameters and dataflow. Those are sufficient to demonstrate all important aspects of the
+system, but the implementation could be extended to support the remaining languages discussed in
+the thesis. Thanks to the fact that the implementation is based on the common coeffect framework,
+this requires minimal amount of work. In order to extend the prototype implementation with support
+for liveness, bounded reuse tracking or other context-aware language, the following four additions
+are required:
 
-We unify systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
-which describes our \emph{structural coeffect calculus} (Section~\ref{sec:struct-calculus}). 
-Similarly to the flat variant, the calculus is parameterized by a \emph{structural coeffect algebra}. 
-Concrete instances of the calculus track bounded variable use (\ie~how many times is variable 
-accessed), data-flow properties (how many past values are needed) and liveness (\ie~can variable 
-be accessed).
+\begin{enumerate}
+\item A domain-specific function abstraction rule that resolves the ambiguity in the general
+  (\emph{abs}) rule of the flat coeffect calculus. For liveness, the handling would be the same
+  as for atafloww, but for other flat coeffect systems, another resolution mechanism might be used
+  instead.
 
-The structural coeffect calculus has desirable equational properties (Section~\ref{sec:struct-syntax}).
-In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all structural
-coeffect calculi. This follows from the fact that structural coeffect associates contextual requirements
-with individual variables and preserves the connection by including explicit structural rules
-(weakening, exchange and contraction).
+\item A domain-specific instance of the coeffect algebra needs to be provided. This consists of
+  defining the set of coeffect annotations and associated operations. In order to support
+  the type inference in the prototype implementation, the constraint solver needs to be extended
+  to solve constraint using the coeffect algebra. For liveness, this would be solving simple
+  two-point lattice constraints.
 
-The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
-thesis. In Chapter~\ref{ch:unified}, we develop the two calculi further by introducing a unified
-presentation of the two (by parameterizing the calculus by the \emph{shape} of context) and we present
-an alternative presentation that makes type inference easier.
+\item For evaluation, a new data type of comonadic values needs to be added. For liveness, this would
+  be an option value that may or may not contain a value. The semantics of comonadic operations on the
+  values needs to be defined.
+
+\item For context-aware languages that have additional primitives (such as $\kvd{prev}~e$ or \ident{?param}),
+  the parser and AST needs to be extended, custom type-checking and translation rules added and
+  domain-specific primitive operations (with their semantics) provided. Liveness and bounded reuse
+  do not have additional custom syntax and so supporting these does not require this step.
+\end{enumerate}
+
+\noindent
+The list mirrors a list of steps that need to be done when supporting a new effectful computation
+in a language that supports monadic do-notation. The step (3) corresponds to implementing a new
+monad and (4) corresponds to adding monad-specific effectful operations. The step (2) applies when
+using indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
+have a counterpart in effectful/monadic languages is (1).
+
+When adding a new context-aware programming language, much of the existing infrastructure
+can be reused. This includes the implementation of the core coeffect and type checking rules and
+also the translation for standard language constructs as well as the interpreter for the target
+language.
+
+% ==================================================================================================
+%
+%     #####
+%    #     #   ##    ####  ######     ####  ##### #    # #####  # ######  ####
+%    #        #  #  #      #         #        #   #    # #    # # #      #
+%    #       #    #  ####  #####      ####    #   #    # #    # # #####   ####
+%    #       ######      # #              #   #   #    # #    # # #           #
+%    #     # #    # #    # #         #    #   #   #    # #    # # #      #    #
+%     #####  #    #  ####  ######     ####    #    ####  #####  # ######  ####
+%
+% ==================================================================================================
+
+\section{Case studies}
+\label{sec:impl-case}
+
+The prototype implementation illustrates a number of interesting aspects of coeffect systems.
+Those appear as examples in the interactive essay (discussed in Section~\ref{sec:impl-essay}),
+but we briefly review them in this section.
 
@@
 
+\subsection{Typing context-aware programs}
+\label{sec:impl-case-typing}
+
+We first consider two case studies of how coeffect type checking works. The first one exposes the
+ambiguity resolution algorithm for the typing of implicit parameters and the second one exposes the
+difference between flat and structural system forataflowow.
+
+\vspace{-0.5em}
+\paragraph{Abstraction for implicit parameters.}
+As discussed in Section~\ref{sec:impl-theory-typing}, the implementation of the language with
+implicit parameters resolves the ambiguity in the lambda abstraction by generating a coeffect
+constraint that restricts the set of parameters required from the declaration site to those that
+are lexically available. Remaining parameters are required from the call site. This is illustrated
+by the following example:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{both}~=\\[-0.25em]
+\quad \kvd{let}~\ident{?fst}~=~100~\kvd{in}\\[-0.25em]
+\quad \kvd{fun}~\ident{trd}~\rightarrow~\ident{?fst}~+~\ident{?snd}~+~\ident{trd}~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{?fst}~=~200~\kvd{in}\\[-0.25em]
+\ident{both}~1
+\end{array}
+\end{equation*}
+%
+In this expression, the lambda function on the third line requires implicit parameters \ident{?fst} and
+\ident{?snd}. Since \ident{?fst} is available in scope, the type of \ident{both} is a function that
+requires only \ident{?snd}. In the text-based notation used in the prototype, the type of the
+function \ident{both} is: {\tt num -\{?snd:num\}-> num}.
+
+\vspace{-0.5em}
+\paragraph{Flat and structural dataflow.}
+In flat dataflow, the context requirements of the body is required from both the declaration site
+and from the call site. In structural dataflow, the context requirements are tracked separately
+for each variable, which provides a more precise type. Consider the following two examples (the
+\kvd{let} keyword is used to define a curried function of two arguments):
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{oldy}~\ident{x}~\ident{y}~=~\ident{x}~+~\kvd{prev}~\ident{y}~\kvd{in}\\[-0.25em]
+\ident{oldy}
+\end{array}
+\end{equation*}
+%
+When type checking the expression using the flat system, the type of \ident{oldy} is inferred as {\tt num -\{1\}-> num -\{1\}-> num},
+but when using the structural system, the type becomes {\tt num -\{0\}-> num -\{1\}-> num}.
+
+This illustrates the difference between the two -- the flat system keeps only one annotation
+for the whole body (which requires 1 past value). In lambda abstraction (or function declaration
+written using \kvd{let}), this requirement is duplicated. The structural system keeps information
+per-variable and so the resulting type reflects the fact that only the variable \idnet{y} appears
+inside \kvd{prev}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Comonadically-inspired translation}
+\label{sec:impl-case-transl}
+
+In addition to running coeffect programs, the implementation can also display the result of the
+translation to the simple functional target language with comonadically-inspired primitives.
+The following two case studies illustrate important aspects of the translation for flat coeffect
+systems (Section~\ref{sec:semantics-translation}) and structural coeffect systems (Section~\ref{sec:struct-transl}).
+
+\paragraph{Merging implicit parameter contexts.}
+
+The following example illustrates the lambda abstraction for implicit parameters.
+It defines a parameter \ident{?param} and then returns a function value that captures it, but
+also requires an implicit parameter \ident{?other}:
+
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~\ident{?param} ~=~ \num{10}~\kvd{in}\\[-0.25em]
+\kvd{fun}~\ident{x}~\rightarrow~\ident{?param} ~+~ \ident{?other}
+\end{array}
+\end{equation*}
+%
+Translating the code to the target language produces the code below. The reader is encouraged to
+view the translation in the interactive essay (Section~\ref{sec:impl-essay}), which displays the
+types and coeffect annotations of the individual values and primitives. As in the theory, the
+comonadically-inspired primitives are families of operations indexed by the coeffects (we
+omit the annotations here):
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{let}~(\ident{ctx2}, \ident{ctx3}) ~=~ \ident{split}~(\ident{duplicate}~\kvd{finput})~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{ctx1} ~=~ \ident{letimpl}_{\ident{?param}}~(\ident{ctx2}, \num{10})~\kvd{in}\\[-0.25em]
+\kvd{fun}~\ident{x}~\rightarrow\\[-0.25em]
+\quad \kvd{let}~\ident{ctx4} ~=~ \ident{merge}~(\ident{x}, \ident{ctx1})~\kvd{in}\\[-0.25em]
+\quad \kvd{let}~(\ident{ctx5}, \ident{ctx6}) ~=~ \ident{split}~(\ident{duplicate}~\ident{ctx4})~\kvd{in}\\[-0.25em]
+\quad \ident{lookup}_{\ident{?param}}~\ident{ctx5} ~+~ \ident{lookup}_{\ident{?other}}~\ident{ctx6}
+\end{array}
+\end{equation*}
+%
+The \kvd{finput} value on the first line models an empty context in which the expression is
+evaluated and is of type $\ctyp{\czero}{\ident{unit}}$. The $\czero$ annotation captures the
+fact that there are no implicit parameters in the context and the type $\ident{unit}$ specifies
+that the context carries no variables.
+
+The context is duplicated and the second part \ident{ctx3} is not needed, because $\num{10}$ is a
+constant. The first part \ident{ctx2} is passed to \ident{letimpl}, which assigns an implicit
+parameter value in the newly returned context \ident{ctx1} of type
+$\ctyp{\cclrd{\{\ident{?param}\}}}{\ident{unit}}$ (the hidden dictionary now contains a value
+for the implicit parameter \ident{?param}, but the context does not contain value bindings for
+any ordinary variables (the \ident{unit} type can be seen as an empty tuple).
+
+In the body of the function, the context \ident{ctx1} is merged with the context provided by the
+variable \ident{x}. The type of \ident{ctx4} is $\ctyp{\ccrld{\{\ident{?param},\ident{?other}\}}}{(\ident{num}\times\ident{unit})}$.
+This is then split into two parts that contain just one of the implicit parameters and those are
+then accessed using \ident{lookup}.
+
+\paragraph{Composition in structural dataflow.}
+
+In structural coeffect systems, the translation works differently in that the context passed to
+a sub-expression contains only assignments for the variables used in the sub-expression (in the
+flat version, we always duplicated the variable context before using \ident{split}). To illustrate
+this, consider the following simple function:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{fun}~\ident{x} \rightarrow \kvd{fun}~\ident{y} \rightarrow \kvd{prev}~\ident{x}
+\end{array}
+\end{equation*}
+%
+In structural coeffect systems, the comonadic value is annotated with a vector of coeffect
+annotations that correspond to individual variables. The initial structural input \kvd{sinput}
+is a value of type $\ctyp{\cclrd{[]}}{()}$ containing no variables (for structural coeffect systems,
+we write $()$ rather than \ident{unit} to make that more explicit). The translated code then looks
+as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{fun}~\ident{x} \rightarrow\\[-0.25em]
+\quad \kvd{let}~\ident{ctx1} ~=~ \ident{merge}~(\ident{x}, \kvd{sinput}) \kvd{in}\\[-0.25em]
+\quad (\kvd{fun}~\ident{y} \rightarrow\\[-0.25em]
+\quad \quad \kvd{let}~\ident{ctx2} ~=~ \ident{merge}~(\ident{y}, \ident{ctx1})~\kvd{in}\\[-0.25em]
+\quad \quad \ident{counit}~(\ident{prev}~(\ident{choose}_{\langle 0,1\rangle}~\ident{ctx2}))
+\end{array}
+\end{equation*}
+%
+The two variables are merged with the initial context, obtaining a value \ident{ctx2} of type
+$\ctyp{\cclrd{[0, 1]}}{\ident{num}\times\ident{num}}$ that contains two dataflow values with 0 and 1 past
+values, respectively.
+
+For simplicity, the implementation does not use the \ident{split}/\ident{merge} pair of operations
+of the structural coeffects to obtain the correct subset of variables. This can be done, but it
+would make the translated code longer and more cumbersome. Instead, we use a higher-level operation
+\ident{choose} (which can be expressed in terms of \ident{split}/\ident{merge}) that projects the
+variable subset as specified by the index. Here, $\langle 0, 1\rangle$ means that the first variable
+should be dropped an the second one should be kept.
+
+The resulting single-variable context is then passed to \ident{prev} (to shift the stream by one)
+and then to \ident{counit} to obtain the current value.
+
+% ==================================================================================================
+%
+%    #######
+%    #        ####   ####    ##   #   #
+%    #       #      #       #  #   # #
+%    #####    ####   ####  #    #   #
+%    #            #      # ######   #
+%    #       #    # #    # #    #   #
+%    #######  ####   ####  #    #   #
+%
+% ==================================================================================================
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=16em]{images/eval1.png}}
+\boxed{\includegraphics[width=16em]{images/eval2.png}}
+\caption{Interactive evaluation of implicit parameters (left) and dataflow (right)}
+\label{fig:essay-eval}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\section{Interactive essay}
+\label{sec:impl-essay}
+
+As explained in the introduction of this chapter, the purpose of the implementation presented in
+this thesis is not to provide a real-world programming language, but to support the theory
+discussed in the rest of the thesis. The goal is to explain the theory and inspire authors of
+real-world programming languages to include support for context-aware programming, ideally using
+coeffects as a sound foundation. For this reason, the implementation needs to be:
+
+\begin{itemize}
+\item {\sc Accessible.}  Anyone interested should be able to experiment with the implemented languages
+  without downloading the source code and compiling it and without installing specialized software.
+
+\item {\sc Explorable.}  It should be possible to explore the inner workings -- how is the typing
+  derived, how is the source code translated to the target language and how is it evaluated.
+\end{itemize}
+
+\noindent
+To make the work \emph{accessible}, we implement sample context-aware languages in a way that makes it
+possible to use them in any standard web browser with JavaScript support (Section~\ref{sec:impl-essay-tech})
+without requiring any server-side component. Following the idea that ``the medium is the message'' \cite{essay-medium},
+we choose a medium that encourages \emph{exploration} and make the implementation available not just
+as source code that can be compiled and run locally, but also in the format of
+an interactive essay (Section~\ref{sec:impl-essay-features}). The live version of the essay can be
+found at: \url{http://tomasp.net/coeffects}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=16em]{images/df1.png}}
+\boxed{\includegraphics[width=16em]{images/df2.png}}
+\caption{Function smoothing the X coordinate (left) with a sample run (right).}
+\label{fig:essay-df}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Explorable language implementation}
+\label{sec:impl-essay-features}
+
+The interactive essay format used of the implementation is based on Bret Victor's work on
+\emph{explorable explanations} \cite{essay-explorable}. Bret Victor describes what the interactivity
+enables as follows:
+%
+\begin{quote}
+\emph{Do our reading environments encourage active reading? Or do they utterly oppose it? A
+  typical reading tool, such as a book or website, displays the author's argument, and nothing
+  else. The reader's line of thought remains internal and invisible, vague and speculative. We
+  form questions, but can't answer them. We consider alternatives, but can't explore them. We
+  question assumptions, but can't verify them. And so, in the end, we blindly trust, or blindly
+  don't, and we miss the deep understanding that comes from dialogue and exploration.}
+\end{quote}
+%
+The interactive essay we present encourages active reading in the sense summarized in Victor's quote.
+We show the reader an example (program, typing derivation or translation), but the reader is
+encouraged to modify it and see how the explanation in the essay adapts.
+
+The idea of active reading is older and has been encouraged in the context of art Josef Albers'
+classic 1963 work on color \cite{essay-albers} (which was turned into an interactive essay
+60 years later \cite{essay-albers-app}). More recently, similar formats have been used to explain
+topics in areas such as signal processing \cite{essay-seeing} (explaining Fourier transformations)
+and sociology \cite{essay-polygons} (visualizing and explaining game theoretical model of segregation
+in the society \cite{essay-segregation}). To our best knowledge, the format has not previously
+been used in the area of programming language theory and so we briefly outline some of the
+interesting features that our essay provides in order to encourage active reading.
+
+\paragraph{Interactive program excution.}
+
+After providing the practical motivation for coeffects (based on Chapter~\ref{ch:intro}), the
+interactive essay shows the reader the two sample context-aware programming languages. Readers can
+write code in a panel that type checks the input, generates user interface for entering the
+required context and runs the sample code.
+
+The panels for implicit parameters and for dataflow computations are shown in Figure~\ref{fig:essay-eval}.
+The sample code on the left adds two implicit parameters and the generated UI lets the user
+enter implicit parameter values as required by the context in the typing judgement. The sample on
+the right calculates the average of the current and past value in a dataflow; the UI lets the user
+enter past values required by the type of the function.
+
+The essay guides the readers through a number of programs (shown in Section~\ref{sec:impl-case-typing})
+and encourages them to run and try modifying them. For implicit parameters, this includes the case
+where an implicit parameter is available at both declaration site and call site (showing that the
+declaration site value is captured). For dataflow, the examples include the comparison between the
+inferred type when using flat and structural type systems.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=32em]{images/types.png}}
+\caption{Source code with type information and explorable typing derivation.}
+\label{fig:essay-typing}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Reactive dataflow.}
+
+The interactive program execution lets the reader run sample programs, but not in a particuarly realistic
+context. To show a more real-world scenario, the essay includes a widget shown in Figure~\ref{fig:essay-df}.
+This lets the user write a function taking a stream of X and Y coordinates and calculate value
+based on the current and past values of the mouse pointer. The X and Y values, together with the
+result are plotted using a live chart.
+
+In the example run shown above, the sample program calculates the average of the last 12 values
+of the X coordinate (green line in Figure~\ref{fig:essay-df}). The example also illustrates one
+practical use of the coeffect type system -- when running, the widget keeps the coordinates in a
+pre-allocated fixed-size array, because the coeffect type system guarantees that at most 12 past
+values will be accessed.
+
+\paragraph{Explorable typing derivations.}
+
+Perhaps the most important aspect of the implemented context aware programming languages is their
+coeffect and type system. In a conventional implementation, the functioning of the type system
+would remain mostly hidden -- we would get an ``OK'' message for a well-typed program and a type
+error for invalid program (likely not very informative, given that reporting good error messages
+for type errors is a notoriously hard problem even for mature language implementations).
+
+The presented interactive essay provides two features to help the reader understand and explore
+typing derivations. First, when reading code samples in the text, tooltips show the typing of
+individual identifiers, most interestingly functions (Figure~\ref{fig:essay-typing}, above).
+
+Second, a later part of the essay provides a type checker that lets the user enter a source code
+in a context aware programming language and produces an explorable typing derivation for the
+program. The output (shown in Figure~\ref{fig:essay-typing}, below) displays a typing judgement
+with assumptions and conclusions and lets the reader navigate through the typing derivation by
+clicking on the assumptions or conclusions. This way, the reader can see how is the final typing
+derivation obtained, exploring interesting aspects, such as the abstraction rule (shown in
+Figure~\ref{fig:essay-typing}).
+
+\paragraph{Comonadically-inspired translation.}
+
+The last interactive element of the essay lets the reader explore the translation of source
+context aware language to a target simple functional language (Section~\ref{sec:impl-theory-transl}).
+Compared with the monadic do-notation \cite{other-haskell98}, the comonadic translation is more
+complex for two reasons.
+
+First, it merges all variables into a single (comonadic) value representing
+the context. Second, there is a flat and structural variation of the system. For these two reasons,
+understanding the translation based just on the rules is harder than for monads. The essay lets
+readers see translation for carefuly chosen case studies that illustrate the key aspects of the
+translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
+programs encouraging \emph{active reading}. Sample input and output are shown in
+Figure~\ref{fig:essay-transl}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\boxed{\includegraphics[width=16em]{images/transl1.png}}
+\boxed{\includegraphics[width=16em]{images/transl2.png}}
+\caption{Source code with comonadically-inspired translation}
+\label{fig:essay-transl}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Implementation overview}
+\label{sec:impl-essay-tech}
+
+The core part of our implementation mostly follows standard techniques for implementing type
+checkers and interpreters for statically-typed functional programming languages. Two interesting
+aspects that are worth discussing is the JavaScript targetting (for running the language
+implementation in a web browser) and integration with client-side (JavaScript) libraries for
+building the user interface. The full source code can be obtained from
+\url{https://github.com/coeffects/coeffects-playground}. It is structured as follows:
+%
+\begin{itemize}
+\item Parsing is implemented using a simple parser combinator library; \strf{ast.fs} defines the
+  abstract syntax tree for the languages; \strf{parsec.fs} implements the parser combinators
+  and \strf{lexer.fs} with \strf{parser.fs} parse the source code by first tokenizing
+  the input and then parsing the stream of tokens.
+
+\item The type checking is implemented by \strf{typechecker.fs}, which annotates an untyped AST
+  with types and generates set of type and coeffect constraints. The constraints are later solved
+  (using domain-specific algorithms for each of the languages) in \strf{solver.fs}.
+
+\item Type-checked programs in context-aware languages are translated to simple target
+  functional subset of the language in \strf{translation.fs}; \strf{evaluation.fs} then interprets
+  programs in the target language. The interpretation does not handle the source
+  language and so programs containing context-aware contstructs cannot be interpreted directly.
+
+\item The user interface of the interactive essay (Section~\ref{sec:impl-essay-features})
+  is implemented partly in F\# and partly in JavaScript. The two most important components include
+  a pretty-printer \strf{pretty.fs} which formats source code (with type tooltips) and typing
+  derivations and \strf{gui.fs} that implements user interaction (e.g.~navigation in explorable
+  typing derivations).
+\end{itemize}
+
+\noindent
+As discussed in Section~\ref{sec:impl-theory-ext}, the implementation can be easily extended to
+support additional context-aware programming languages. This is due to the fact that it is based
+on the unified theory of coeffects. In practice, adding support for liveness tracking would
+require adding a domain-specific constraint solver in \strf{solver.fs} and extending the
+interpreter in \strf{evaluation.fs} with a new kind of comonadically-inspired data type
+(indexed maybe comonad) with its associated operations.
+
+\section{Related work}
+\label{sec:impl-related}
+
+To our best knowledge, combining Bret Victor's idea of explorable explanations with programming
+language theory (as discussed in Section~\ref{sec:impl-essay}) is a novel contribution of our work.
+On the technical side, we build on a number of standard methods.
+
+\paragraph{Parsing and type checking.} The implementation of the parser, type checker and
+interpreter follows standard techniques for implementing functional programming languages.
+In order to be able to compile the implementation to JavaScript (see below), we built a
+small parser combinator library \cite{monad-parsing} rather than using one of the already
+available libraries \cite{other-fparsec}.
+
+\paragraph{Targetting JavaScript.} In order to make the implementation accessible to a broad
+audience, it can be executed in a web browser. This is achieved by automatically translating the
+implementation from F\# to JavaScript. We use an F\# library called FunScript \cite{other-funscript}
+(which is a more recent incarnation of the idea developed by the author \cite{app-fsharp-webtools}).
+We choose F\#, but similar tools exist for other functional languages such as OCaml \cite{other-js-ocaml}.
+It is worth noting that FunScript is implemented as a \emph{library} rather than as a compiler
+extension. This is done using the meta-programming capabilities of F\# \cite{app-fsharp-metaprog}.
+
+\paragraph{Client-side library integration.} An interesting aspect of the interactive essay
+user interface is the integration with third-party JavaScript libraries. We use a number of
+libraries including JQuery (for web browser DOM manipulation) and MathJax \cite{essay-mathjax}
+(for rendering of typing derivation). In order to call those from F\# code, we use a number
+of mapping methods described in \cite{app-age-of-web}. For example, the following F\# declaration
+is used to invoke the \ident{Queue} function in MathJax:
+%
+\begin{equation*}
+\begin{array}{l}
+\lbrack\langle\ident{JSEmitInline}(\str{MathJax.Hub.Queue(\{0\});})\rangle\rbrack\\[-0.25em]
+\kvd{let}~\ident{queueAction}~(\ident{f}:\ident{unit}~\rightarrow~\ident{unit}):\ident{unit}~=~\ident{failwith}~\str{JS only!}
+\end{array}
+\end{equation*}
+%
+The \ident{JSEmitInline} syntax on the first line is an attribute that instructs FunScript
+to compile all invocations of the \ident{queueAction} function into the JavaScript literal
+specified in the attribute (with \strf{\{0\}} replaced by the argument).
+
+% ==================================================================================================
+%
+%     #####
+%    #     # #    # #    # #    #   ##   #####  #   #
+%    #       #    # ##  ## ##  ##  #  #  #    #  # #
+%     #####  #    # # ## # # ## # #    # #    #   #
+%          # #    # #    # #    # ###### #####    #
+%    #     # #    # #    # #    # #    # #   #    #
+%     #####   ####  #    # #    # #    # #    #   #
+%
+% ==================================================================================================
+
+~
+
 \section{Summary}
+
+This chapter supplements the theory of coeffects presented in the previous three chapters with
+a prototype implementation. We implemented three simple context-aware programming languages that
+track implicit parameters and past values in dataflow computations, the latter both in flat
+(per-context) and structural (per-variable) way.
+
+The implementation discussed in this chapter provides an evidence for our claim that the
+theory of coeffects can be used as a basis for a wide range of sound context-aware programming
+languages. Our implementation consists of a shared coeffect framework (handling type checking and
+translation). Each context-aware language then adds a domain-specific rule for choosing a unique
+typing derivation, an interpretation of comonadically-inspired primitives and (optionally)
+domain-specific primitives such as the \kvd{prev} construct for dataflow.
+
+We make the implementation available in the usual form (as source code that can be downloaded,
+compiled and executed), but we also present it in the form of interactive essay. This encourages
+\emph{active reading} and lets the reader explore a number of aspects of the implementation
+including the type checking (through explorable typing derivations) and the translation. The key
+contribution of this thesis is that it provides a unified way for \emph{thinking} about context in
+programming languages and the interactive presentation of the implementation is aligned with this
+goal. Programming languages of the future will need a mechanism akin to coeffects and we aim to
+provide a convincing argument supported by a prototype implementation.

### 2016-05-12 text/Chapter08.tex
Edits in implementation, consistency (Chapter 7)

@@
 rebindable resources, Haskell type classes (discussed by Orchard~\cite{comonads-dom-thesis}),
-data-flow and variable liveness.
+dataflow and variable liveness.
 
@@
 $(\C,\cpar,\czip)$ to form a lattice. This is consistent with all our structural coeffect systems as
-well as flat systems for liveness and data-flow, but not with the system for tracking of implicit
+well as flat systems for liveness and dataflow, but not with the system for tracking of implicit
 parameters (or sets of rebindable resources).

### 2016-05-12 text/Chapter09.tex
Fix references
Edits in implementation, consistency (Chapter 7)

@@
 journalists obtain information published on the web or through open government data initiatives;
-in data-flow programming, we have access to past values of expressions.
+in dataflow programming, we have access to past values of expressions.
 This thesis aims to change our understanding of \emph{context} so that the above examples are
@@
 execution environment. In Chapter~\ref{ch:applications}, we revisit many of such features --
-including resource tracking in distributed computing, cross-platform development, data-flow
+including resource tracking in distributed computing, cross-platform development, dataflow
 programming and liveness analysis, but also Haskell's implicit parameters.
@@
 instances of flat coeffects include Haskell's implicit parameters, whole-context liveness and
-whole-context data-flow.
+whole-context dataflow.
 
@@
 Concrete instances of the calculus track bounded variable use (\ie~how many times is a variable
-accessed), data-flow properties (how many past values are needed) and liveness (\ie~can variable
+accessed), dataflow properties (how many past values are needed) and liveness (\ie~can variable
 be accessed).
@@
 The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
-thesis. In Chapter~\ref{ch:unified}, we discuss the similarities and differences between the two.
+thesis. In Chapter~\ref{ch:further}, we discuss the similarities and differences between the two.
 Although the calculi are worth considering separately (as they have different equational properties),
-they can be presented in a uniform way. In Section~\ref{sec:unified-unified}, we present the
+they can be presented in a uniform way. In Section~\ref{sec:further-unified}, we present the
 \emph{unified coeffect calculus}, which can be instantiate to track both per-variable and
@@
 This thesis aims to provide a similar unifying theory and tools for capturing context-dependence
-in programming languages. We showed that programming language features (liveness, data-flow,
+in programming languages. We showed that programming language features (liveness, dataflow,
 implicit parameters, etc.) that were previously treated separately can be captured by a common

### 2016-05-12 text/backup/Chapter01.tex
Edits in implementation, consistency (Chapter 7)

@@
 \todo{We present three calculi that model common notions of context-dependence and can be used
-as basis for developing context-aware programming langauges with static type systems.}
+as basis for developing context-aware programming languages with static type systems.}

### 2016-05-13 text/Abstract.tex
Add interactive essay to abstract

@@
-%!TEX root = ../main.tex
 \pdfbookmark[1]{Abstract}{Abstract} % Bookmark name visible in a PDF viewer
@@
 The development of programming languages needs to reflect important changes in the way
-programs execute. In recent years, this has included \eg~the development of parallel programming 
+programs execute. In recent years, this has included the development of parallel programming
 models (in reaction to the multi-core revolution) or improvements in data access technologies.
@@
 variable usage (\eg~in liveness analaysis and linear logics) or past values in
-stream-based dataflow programming.
+stream-based dataflow programming. Our first contribution is the discovery of the connection
+between the above examples and their novel presentation in the form of calculi (\emph{coeffect systems}).
+The presented type systems and formal semantics highlight the relationship between different
+notions of context.
 
-Our first contribution is the discovery of the connection between the above examples and
-their novel presentation in the form of calculi (\emph{coeffect systems}). The presented type 
-systems and formal semantics highlight the relationship between different notions of context. 
-Our second and third contributions are two unified coeffect calculi that capture commonalities 
-in the presented examples. In particular, our \emph{flat coeffect calculus} models languages 
+Our second contribution is the definition of two unified coeffect calculi that capture the common
+structure of the examples. In particular, our \emph{flat coeffect calculus} models languages
 with contextual properties of the execution environment and our \emph{structural coeffect
 calculus} models languages where the contextual properties are attached to the variable usage.
-We close the thesis with a \emph{unified coeffect calculus} that captures the two notions
-using a single parameterized system.
+We define the semantics of the calculi in terms of category theoretical structure of an
+\emph{indexed comonad} (based on dualisation of the well-known monad structure), use it
+to define operational semantics and prove safety result for the calculi.
 
-Although the focus of this thesis is on the syntactic properties of the presented 
-systems, we also discuss their category-theoretical motivation. We introduce the notion of
-an \emph{indexed} comonad (based on dualisation of the well-known monad structure) and show 
-how they provide semantics of the two coeffect calculi.
+Our third contribution is a novel presentation of our work in the form of web-based
+\emph{interactive essay}. This provides a simple implementation of three context-aware programming
+languages and lets the reader write and run simple context-aware programs, but also explore the
+theory behind the implementation including the typing derivation and semantics.
 

### 2016-05-13 text/Chapter08.tex
Edit thesis conclusions (Chapter 9)

@@
-\chapter{Related and future work}
+\chapter{Further work}
 \label{ch:further}
@@
 
+% =================================================================================================
+%
+% 	 #####                 #     #                                                #
+% 	 #                     #     #                                                #
+% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   #
+% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #
+% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###
+% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #
+% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   #
+%
+% =================================================================================================
+
+\section{Further work}
+\label{sec:conc-further}
+
+Most of the further work has been discussed throughout the thesis, so this section
+serves mainly as an overview. We follow the same structure as when discussing pathways to
+coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
+to coeffect systems, categorical semantics of coeffects, and connections with substructural type
+systems and bunched logics.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Syntactic coeffect systems.}
+The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
+variables, lambda abstraction, application and let-binding. This lets us focus on fundamental
+properties of coeffect systems (and explain how coeffect system differ from better-known effect
+systems), but it is hardly sufficient for a realistic programming language.
+
+Further work is to extend the coeffect systems presented here to a full programming language. A
+useful reference is the work of Nielson and Nielson~\cite{effects-nielson} who consider a
+similar development for effect systems, adding conditionals, recursion and polymorphism.
+
+Coeffect annotations in context-aware programming languages can contain significant amount
+of information. Thus, the programming language also needs a form of type inference that can propagate
+such information. This has been done for effect systems \cite{effects-polymorphic}. For
+\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
+for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
+to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
+be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where
+\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Language semantics.}
+
+As discussed in Section~\ref{sec:path-sem}, comonads can be used to define the semantics of a
+programming language in two ways. The first, ``language semantics'', approach is to use a
+single comonad to describe the semantics of a specific context-aware property. The second,
+``meta-language'' approach is to extend the language with explicit constructs representing the
+operations of the underlying comonad.
+
+The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
+guide for the development of the type systems. Further work includes more precise treatment of
+the categorical structure of indexed comonads. This has partly been done for flat coeffects by
+Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
+first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
+work is to develop a similar categorical model for the unified coeffect calculus, possibly using
+the categorical notion of containers \cite{types-containers}.
+
+We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:unified-meta}.
+Developing this technique further could unify coeffect systems with the work on Contextual Modal
+Type-Theory (CMTT) \cite{logic-cmtt} and allow other interesting applications of coeffects such
+as meta-programming \cite{logic-cmtt} and distributed computing with explicit modalities \cite{app-distributed-ml5}.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Substructural and bunched logics.}
+
+In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
+context} and to \emph{individual variables}, respectively. We later unified the two systems
+in Section~\ref{sec:unified-unified}.
+
+An intriguing question is whether coeffects can generalize \emph{bunched typing} \cite{substruct-bunched},
+which uses a tree-like structure of variable contexts (also discussed in Section~\ref{sec:path-logic}).
+The definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
+typing requires tree-like variable context, while we use a vector. Furthermore, bunched typing
+annotates internal nodes of the tree (sub-contexts are combined using ``;'' or using ``,'') rather
+than just variables. Finding a simple coeffect system that is capable of capturing bunched typing
+is thus an interesting further work. Indeed, this could lead to numerous uses of coeffects as
+bunched typing is the basis for widely-used separation logic \cite{substruct-separation-logic}.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Applications.}
+The last direction for further development is developing practical programming languages or
+libraries based on the theory presented in this thesis. We believe that the most useful approach
+is to extend programming languages with a reusable coeffect system and allow library developers
+define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
+For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
+``do'' notation.
+
+Another option is to use coeffects just as the underlying theory of a single-purpose programming
+language feature. When discussing why context-aware programming matters (Section~\ref{sec:intro-whymatters}),
+we covered a number of concrete problems that developers are facing today. Each of those would
+presents an important further work. The most important practical challenges are cross-compilation
+(often solved using complex $\prepk{\#if}$ pre-processor rules) and tracking of provenance
+and security properties.
+
 % ==================================================================================================

### 2016-05-13 text/Chapter09.tex
Edit thesis conclusions (Chapter 9)

@@
-\chapter{Conclusions and further work}
+\chapter{Conclusions}
 \label{ch:conclusions}
@@
 
-In this chapter, we give a brief summary of the technical development completed in this thesis
+In this chapter, we give a brief overview of the technical contributions of this thesis
 (Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
-common patterns and captures those using three \emph{coeffect calculi}. Next, we give an overview
-of further work (Section~\ref{sec:conc-further}), often referring to longer discussion in earlier
-chapters. Finally, Section~\ref{sec:conc-conclusions} concludes the thesis.
+common patterns and captures those using two \emph{coeffect calculi}. It gives formal semantics
+using categorical foundations and uses it as a basis for prototype implementation. Finally,
+Section~\ref{sec:conc-conclusions} concludes the thesis.
 
@@
 %
-%       ###
-%      #   #
-%      #      #   #  ## #   ## #    ###   # ##   #   #
-%       ###   #   #  # # #  # # #      #  ##  #  #   #
-%          #  #   #  # # #  # # #   ####  #      #  ##
-%      #   #  #  ##  # # #  # # #  #   #  #       ## #
-%       ###    ## #  #   #  #   #   ####  #          #
-%                                                #   #
-%                                                 ###
+%    #######
+%    #     # #    # ###### #####  #    # # ###### #    #
+%    #     # #    # #      #    # #    # # #      #    #
+%    #     # #    # #####  #    # #    # # #####  #    #
+%    #     # #    # #      #####  #    # # #      # ## #
+%    #     #  #  #  #      #   #   #  #  # #      ##  ##
+%    #######   ##   ###### #    #   ##   # ###### #    #
+%
 % =================================================================================================
 
-\section{Summary}
+\section{Overview of contributions}
 \label{sec:conc-summary}
 
-Modern computer programs run in rich and diverse environments or \emph{contexts}. The richness means
-that environments provides additional resources and capabilities that may be accessed by the program.
-The diversity means that programs often need to run in multiple different environments, such as mobile
-phones, servers, web browsers or even on the GPU. In this thesis, we present the foundations for
-building programming languages that simplify writing software for such rich and diverse environments.
+As observed in Chapter~\ref{ch:intro}, modern computer programs run in rich and diverse environments
+or \emph{contexts}. The richness means that environments provides additional resources and
+capabilities that may be accessed by the program. The diversity means that programs often need to
+run in multiple different environments, such as mobile phones, servers, web browsers or even on the
+GPU. In this thesis, we present the foundations for building programming languages that simplify
+writing software for such rich and diverse environments.
 
@@
 % --------------------------------------------------------------------------------------------------
+~
 
@@
 
-Our focus is on the syntactic properties of the calculus -- we give a type system
-(Section~\ref{sec:flat-calculus}) and discuss equational theory of the calculus (Section~\ref{sec:flat-syntax}).
-In the flat coeffect calculus, $\beta$-reduction and $\eta$-expansion do not generally preserve
-the type of an expression, but we identify two conditions when this is the case -- this gives us
-a basis for operational semantics of our calculi for liveness and implicit parameters.
-
-The design of the type system is further validated by categorical semantics
-(Section~\ref{sec:flat-semantics}) which models flat coeffect calculus in terms of
-\emph{flat indexed comonad} -- a structure based on a categorical dual of monads.
+We define a type system for the calculus (Section~\ref{sec:flat-calculus}),
+discuss how to resolve the ambiguity inherent in context-aware programming for sample
+languages we consider (Section~\ref{sec:flat-unique}) and study equational properties of the
+calculus (Section~\ref{sec:flat-syntax}). In the flat coeffect calculus, $\beta$-reduction and
+$\eta$-expansion do not generally preserve the type of an expression, but we identify two
+conditions when this is the case.
 
@@
 
-\paragraph{Unified calculus.}
+\paragraph{Semantics and safety.}
 
-The Chapters~\ref{ch:flat} and \ref{ch:structural} present the main novel technical contributions of this
-thesis. In Chapter~\ref{ch:further}, we discuss the similarities and differences between the two.
-Although the calculi are worth considering separately (as they have different equational properties),
-they can be presented in a uniform way. In Section~\ref{sec:further-unified}, we present the
-\emph{unified coeffect calculus}, which can be instantiate to track both per-variable and
-whole-context properties. The key trick is to parameterize the calculus by the \emph{shape} of
-a context, which can be a vector (corresponding to the vector of free variables) or a singleton
-container (attaching one annotation to the whole context).
+Both of the coeffect calculi are parameterized and can be instantiated to obtain a language with
+a type system that tracks concrete notion of context dependence. The coeffect calculus can be seen
+as a framework for building concrete domain-specific context-aware programming languages. In
+addition to the type system, the framework also provides a way for defining the semantics of
+concrete domain-specific context-aware languages, guides their implementation and simplifies
+safety proofs.
+
+This is done using a \emph{comonadically-inspired} semantics. We generalize the category-theoretical
+dual of monad to \emph{indexed comonad} and use it to define the semantics of context-aware
+programs for both the flat coeffect calculus (Chapter~\ref{ch:semantics}) and the structural coeffect
+calculus (Section~\ref{sec:struct-semantics}). Due to the ambiguity inherent in contextual properties,
+we define the semantics over a typing derivation, but we also give a domain-specific algorithm for
+choosing a unique typing derivation for each of our sample languages (Section~\ref{sec:flat-unique}
+and Section~\ref{sec:struct-unique}).
+
+We use comonads in a syntactic way, following the example of Haskell's use of monads and treat
+it as a \emph{translation} from source context-aware programming language to a simple target
+functional language. The target language includes uninterpreted comonadically-inspired primitives
+whose concrete reduction rules need to be provided for each concrete context-aware deomain-specific
+language. We give concrete definitions and prove type safety for three sample context-aware
+languages (Section~\ref{sec:semantics-proofs} and Section~\ref{sec:struct-transl}). Together with
+well-typedness of our translation, these guarantee that ``well-typed context-aware programs do
+not go wrong''.
+
+% --------------------------------------------------------------------------------------------------
+
+\paragraph{Interactive essay.}
+
+A part of the thesis is an interactive web-based essay (\url{http://tomasp.net/coeffects}),
+which implements three simple context-aware languages. In addition to more traditional prototype
+language implementations, the essay provides a number of novel features. As usual, it lets users
+write and evaluate sample programs, but it also lets them explore the details of the coeffect
+theory -- this includes an interactive way of exploring typing derivations and program semantics.
+The essay also provides a number of case studies that highlight interesting aspects of the theory.
+
+The implementation (Chapter~\ref{ch:impl}) serves two purposes. First, it links together all parts
+of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the two coeffect calculi
+to build a language framework that is then instantiated to implement three concrete context-aware
+languages. The  framework provides the type system and translation (based on the
+comonadically-inspired semantics) that is used for program execution. This shows the practical
+benefits of using the theory of coeffects as a basis of real-world programming languages.
+
+Second, the implementation uses novel kind of medium (Section~\ref{sec:impl-essay}) to make our work
+\emph{accessible} and \emph{explorable}. Building a production-ready programming language is outside
+the scope of the  thesis, so instead, our goal is to explain the theory and inspire authors of
+real-world programming languages to include support for context-aware programming, ideally using
+coeffects as a sound foundation. For this reason, we make the implementation accessible over web
+without requiring installation of specialized software and we make it explorable to encourage
+active reading.
 
@@
 %
-% 	 #####                 #     #                                                #
-% 	 #                     #     #                                                #
-% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   #
-% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #
-% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###
-% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #
-% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   #
-%
-% =================================================================================================
-
-\section{Further work}
-\label{sec:conc-further}
-
-Most of the further work has been discussed throughout the thesis, so this section
-serves mainly as an overview. We follow the same structure as when discussing pathways to
-coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
-to coeffect systems, categorical semantics of coeffects, and connections with substructural type
-systems and bunched logics.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Syntactic coeffect systems.}
-The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
-variables, lambda abstraction, application and let-binding. This lets us focus on fundamental
-properties of coeffect systems (and explain how coeffect system differ from better-known effect
-systems), but it is hardly sufficient for a realistic programming language.
-
-Further work is to extend the coeffect systems presented here to a full programming language. A
-useful reference is the work of Nielson and Nielson~\cite{effects-nielson} who consider a
-similar development for effect systems, adding conditionals, recursion and polymorphism.
-
-Coeffect annotations in context-aware programming languages can contain significant amount
-of information. Thus, the programming language also needs a form of type inference that can propagate
-such information. This has been done for effect systems \cite{effects-polymorphic}. For
-\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
-for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
-to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
-be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where
-\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Language semantics.}
-
-As discussed in Section~\ref{sec:path-sem}, comonads can be used to define the semantics of a
-programming language in two ways. The first, ``language semantics'', approach is to use a
-single comonad to describe the semantics of a specific context-aware property. The second,
-``meta-language'' approach is to extend the language with explicit constructs representing the
-operations of the underlying comonad.
-
-The categorical semantics in this thesis used the ``language semantics'' style, but mainly as a
-guide for the development of the type systems. Further work includes more precise treatment of
-the categorical structure of indexed comonads. This has partly been done for flat coeffects by
-Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
-first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
-work is to develop a similar categorical model for the unified coeffect calculus, possibly using
-the categorical notion of containers \cite{types-containers}.
-
-We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:unified-meta}.
-Developing this technique further could unify coeffect systems with the work on Contextual Modal
-Type-Theory (CMTT) \cite{logic-cmtt} and allow other interesting applications of coeffects such
-as meta-programming \cite{logic-cmtt} and distributed computing with explicit modalities \cite{app-distributed-ml5}.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Substructural and bunched logics.}
-
-In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
-context} and to \emph{individual variables}, respectively. We later unified the two systems
-in Section~\ref{sec:unified-unified}.
-
-An intriguing question is whether coeffects can generalize \emph{bunched typing} \cite{substruct-bunched},
-which uses a tree-like structure of variable contexts (also discussed in Section~\ref{sec:path-logic}).
-The definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
-typing requires tree-like variable context, while we use a vector. Furthermore, bunched typing
-annotates internal nodes of the tree (sub-contexts are combined using ``;'' or using ``,'') rather
-than just variables. Finding a simple coeffect system that is capable of capturing bunched typing
-is thus an interesting further work. Indeed, this could lead to numerous uses of coeffects as
-bunched typing is the basis for widely-used separation logic \cite{substruct-separation-logic}.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Applications.}
-The last direction for further development is developing practical programming languages or
-libraries based on the theory presented in this thesis. We believe that the most useful approach
-is to extend programming languages with a reusable coeffect system and allow library developers
-define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
-For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
-``do'' notation.
-
-Another option is to use coeffects just as the underlying theory of a single-purpose programming
-language feature. When discussing why context-aware programming matters (Section~\ref{sec:intro-whymatters}),
-we covered a number of concrete problems that developers are facing today. Each of those would
-presents an important further work. The most important practical challenges are cross-compilation
-(often solved using complex $\prepk{\#if}$ pre-processor rules) and tracking of provenance
-and security properties.
-
-% =================================================================================================
-%
-% 	  ###                         ##                    #
-% 	 #   #                         #
-% 	 #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##    ###
-% 	 #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  #  #
-% 	 #      #   #  #   #  #        #    #   #   ###     #    #   #  #   #   ###
-% 	 #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   #      #
-% 	  ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   #  ####
+%     #####
+%    #     # #    # #    # #    #   ##   #####  #   #
+%    #       #    # ##  ## ##  ##  #  #  #    #  # #
+%     #####  #    # # ## # # ## # #    # #    #   #
+%          # #    # #    # #    # ###### #####    #
+%    #     # #    # #    # #    # #    # #   #    #
+%     #####   ####  #    # #    # #    # #    #   #
 %

### 2016-05-16 text/Chapter04.tex
Edits in further work and the rest (Chapter 8)

@@
 concrete discussion. We explain its category-theoretical motivation, we use it to define
-translational semantics (akin to Haskell's do notation) and we prove a soundness result that
+translational semantics (akin to Haskell's ``do'' notation) and we prove a soundness result that
 well-typed programs in flat coeffect calcului for implicit parameters and dataflow do not get

### 2016-05-16 text/Chapter05.tex
Edits in further work and the rest (Chapter 8)

@@
 Effect systems provide a type system for tracking effects and monadic translation can be used as
-a basis for implementing effectful domain-specific languages (e.g.~through the do-notation in
+a basis for implementing effectful domain-specific languages (e.g.~through the ``do'' notation in
 Haskell).
@@
 that embeds context-dependent computations in a functional programming language, similarly to how
-monads and the do notation provide a way of embedding effectful computations in Haskell.
+monads and the ``do'' notation provide a way of embedding effectful computations in Haskell.
 
@@
 In this section, we follow the example of effects and monads and we use the semantics to define a
-translation akin to the do notation in Haskell.
+translation akin to the ``do'' notation in Haskell.
 
@@
 that turns programs written in a domain-specific coeffect language into a functional target
-language. This is akin to the Haskell do-notation for monads. Finally, we extended the target
+language. This is akin to the Haskell ``do'' notation for monads. Finally, we extended the target
 language with concrete implementations of comonadic operations for dataflow and implicit parameters

### 2016-05-16 text/Chapter07.tex
Edits in further work and the rest (Chapter 8)

@@
 The list mirrors a list of steps that need to be done when supporting a new effectful computation
-in a language that supports monadic do-notation. The step (3) corresponds to implementing a new
+in a language that supports monadic ``do'' notation. The step (3) corresponds to implementing a new
 monad and (4) corresponds to adding monad-specific effectful operations. The step (2) applies when
@@
 context aware language to a target simple functional language (Section~\ref{sec:impl-theory-transl}).
-Compared with the monadic do-notation \cite{other-haskell98}, the comonadic translation is more
+Compared with the monadic ``do'' notation \cite{other-haskell98}, the comonadic translation is more
 complex for two reasons.

### 2016-05-16 text/Chapter08.tex
Edits in further work and the rest (Chapter 8)

@@
 
-%===================================================================================================
-
-The main goal of this thesis is to provide a \emph{unified} calculus for tracking context
-dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified
-two kinds of contextual properties that we further covered separately. Flat coeffects in
-Chapter~\ref{ch:flat} track whole-context properties, while structural coeffects, covered in
-Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify the two
-notions.
-
-Although the results presented in this thesis are mainly of a theoretical nature, we indeed
-believe that coeffects should be integrated in main-stream programming languages. This is addressed
-in the second part of the chapter, where we outline possible approach for a practical implementation
-of coeffect systems. Finally, we conclude the chapter with a brief discussion of an alternative
-approach to coeffect system, based on the meta-language style.
+The prototype implementation discussed in the previous chapter directly follows the two
+coeffect calculi presented in this thesis. This links together all aspects of the coeffect
+theory (type system, semantics, safety proofs) and it provides practical support for the
+presented theoretical work. However, the theory of coeffects could be embedded in practical
+programming languages in a various ways and the theory could be further developed in a number
+of different directions.
+
+The most important alternative directions revisit three design decisions made in this thesis.
+First, rather than discussing flat and structural coeffects, we can treat them in a single
+unified calculus tracking both kinds of context dependence. Second, rather than using ``language
+semantics'' style (Section~\ref{sec:path-sem-langs}), we could use a ``meta-language'' style
+and extend the host language with explicit constructs for working with coeffects. Third,
+rather than implementing a prototype coeffect language, we could embed coeffects into an
+existing language using techniques inspired by Haskell's ``do'' notation.
 
@@
 \begin{itemize}
-\item We start by unifying the two kinds of coeffects discussed so far. In
-  Section~\ref{sec:further-unified}, we introduce the \emph{unified coeffect calculus}, which
-  generalizes flat and structural systems and can be instantiated to track both flat and structural
-  properties.
-
-\item We outline one possible approach for practical implementations of coeffects
-  (Section~\ref{sec:further-impl}). The technique is based on embedding comonadic computations
-  via a lightweight syntactic extension and using rich type systems for capturing the structure of
-  coeffect algebra.
-
-\item Finally, we discuss an alternative approach to defining coeffect systems based on the
-  meta-language style, which highlights the relationship between our work and related work arising
-  from modal logics such as CMTT \cite{logic-cmtt} (Section~\ref{sec:further-meta}).
+\item We start by unifying the two kinds of coeffects discussed so far. We introduce the
+  \emph{unified coeffect calculus} (Section~\ref{sec:further-unified}), which generalizes flat
+  and structural systems and can be instantiated to track both flat and structural properties.
+
+\item We discuss an alternative approach to defining coeffect systems based on the meta-language
+  style (Section~\ref{sec:further-meta}), which highlights the relationship between our work and
+  related work arising from modal logics such as CMTT \cite{logic-cmtt}.
+
+\item We consider a way of implementing coeffects based on embedding comonadic computations
+  via a lightweight syntactic extension and a rich type systems akin to Haskell's ``do'' notation
+  (Section~\ref{sec:further-impl}).
+
+\item Finally, we discuss extensions needed for real-world source languages with constructs
+  such as conditionals and further applications of coeffects arising from substructural and bunched
+  logics (Section~\ref{sec:further-theory})
 \end{itemize}
@@
   (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
-  explicitly, but are admissible.
+  explicitly, but they can be derived thanks to the fact that variables in the context can be
+  freely reordered.
 
@@
   coeffect calculus, the operations (which model appending and splitting vectors) are invertible
-  and so the structural coeffect algebra uses the same tensor product $\atimes$.
+  and so the structural coeffect algebra uses the same tensor product $\atimes$. However, the
+  $\cpar$ operation is still needed in the contraction rule.
 \end{itemize}
@@
 respectively). For structural coeffect calculi, the two operators coincide, but for flat
-coeffect calculi, they provide the needed flexibility.
+coeffect calculi, they differ and provide the needed flexibility.
 
@@
 Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of
-Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
-to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
+Abbott et al. \cite{types-containers}. Containers have later been linked to comonads by Ahman et al.
+\cite{comonads-containers}, but here we use them as part of the coeffect algebra, rather than to
+provide semantics of context-aware languages.
+
+Intuitively a container describes data
 types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
@@
 \begin{definition}
-A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
-$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations
-$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
-$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
+A \emph{\cclrd{unified coeffect scalar}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
+$\C$ together with elements $\cunit, \,\czero \in \C$, binary operations $\cseq, \,\cpar$ such that
+$(\C, \,\cseq, \,\cunit)$ and $(\C, \,\cpar, \,\czero)$ are monoids and a binary relation $\cleq$ such
+that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
 %
@@
   \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
-We usually write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
-The coeffect container is also equipped with an operation that appends shapes (when we concatenate
+We write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
+The coeffect container is equipped with an operation that appends shapes (when concatenating
 variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
@@
 
-Therefore, a coeffect annotation is a \emph{vector}
+Therefore, a coeffect annotation is a vector
 $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$
@@
 \begin{example}
-Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
+Similarly, we can also define a coeffect container with \emph{no} positions,
 \ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
@@
 Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
-\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified
-coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
+\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \aclrd{unified
+coeffect algebra} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
 is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
 combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
-%
+
 \begin{equation*}
@@
  \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
-\\[-0.5em]
+\end{array}
+\begin{array}{rclcrcl}
  \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
  \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
-\\[-0.5em]
+\end{array}
+\begin{array}{rclcrcl}
  \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
@@
 and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
-As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
-use of variables -- the associativity is built-in. The order of variables matters, but can be changed
+As in the structural system, free-variable contexts $\Gamma$ are treated as vectors (without duplicate
+variables) with associativity built-in. The order of variables matters, but can be changed
 using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
@@
 Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
-discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-understanding}, we can also read
+discussed in Section~\ref{sec:flat-calculus-understanding}, we can also read
 the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
@@
 
-%--------------------------------------------------------------------------------------------------
-
-\subsection{Semantics of unified calculus}
-\label{sec:further-semantics}
-
-The semantics of the unified calculus can be obtained by generalizing the semantics of the structural
-calculus in the same way as we generalized the type system. However, it is worth exploring the type
-signatures of the operations of the underlying (comonad-based) structure.
-
-The semantics of the unified coeffect calculus follows the same pattern as the semantics of the
-structural calculus (Section~\ref{sec:struct-semantics}) with a number of differences. The
-index $\aclrd{\textbf{r}}$ of the object mapping $\ctyp{\aclrd{\textbf{r}}}$ is an element of
-the structure $\C^{\SP(\sclrd{n})}$ for a specified container $\SH \triangleleft \SP$ and a shape
-$\sclrd{n}$ determined as $\sclrd{n} = \slen{\Gamma}$. To match the type system discussed in
-Section~\ref{sec:further-types}, we need to modify the domains and codomains of the \ident{merge},
-\ident{split} and \ident{dup} operations as follows.
-%
-\begin{equation*}
-\begin{array}{lcl}
- \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
-  \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \\
- \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} &\narrow{:}&
-  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \\
- \ident{dup}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
-  \ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}}\,\azip\,\alift{\cclrd{s}}}{(\alpha \xtimes \alpha)} \\
+% =================================================================================================
+%
+%    #   #          #                    ##
+%    #   #          #                     #
+%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###
+%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   #
+%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    #####
+%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #
+%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###
+%                                                            #   #                #   #
+%                                                             ###                  ###
+% =================================================================================================
+
+\section{Coeffect meta-language}
+\label{sec:further-meta}
+
+In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
+language semantics as introduced by Moggi \cite{monad-notions}. The first approach is to use
+monads in the \emph{semantics} of an effectful language. The second approach is to extend the
+language with (additional) monadic constructs that can then be used for writing effectful
+monads explicitly.
+
+In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
+the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
+indexed comonads to give the semantics for the language and to derive type system for it.
+
+In this section, we briefly discuss the other technique. That is, we embed indexed
+comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
+context (semantically, this corresponds to an indexed comonad) and we add language constructs
+that correspond to the operations of indexed comonads. The coeffect meta-language highlights the
+relationship between coeffects and closely related work on contextual modal type theory (CMTT)
+\cite{logic-cmtt}. Developing the system further is also an interesting future research direction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffects and contextual modal type theory}
+
+As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
+to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
+meta-languages that include $\square$ as a type constructor
+\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
+Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
+From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
+coeffect calculus.
+
+\paragraph{Context in CMTT and coeffects.}
+Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
+comonads (the $\square$ modality) into the meta-language, there are two important differences.
+
+Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
+makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
+requirements are formed by an abstract coeffect algebra, which is more general and can capture
+variable contexts, but also integers, two-point lattices, etc.
+
+Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
+the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+constructor refers to a value of type $\tau$ \emph{together} with additional context specified
+by the coeffect annotation $\cclrd{r}$ (\eg~a list of past values or additional resources).
+
+In contrast in CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the
+original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the
+context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the
+two operations of a comonad:
+\begin{equation*}
+\begin{array}{rcl}
+ \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
+ \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow
+    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
 \end{array}
 \end{equation*}
-%
-The key difference is that the vector concatenation $\atimes$ is replaced with $\apar$ in the
-\emph{domain} of the operations (in \ident{split}) and with $\azip$ in the codomain
-(in \ident{merge}, but also in \ident{dup} which models contraction).
-This means that the splitting and merging the context can change the coeffect annotation:
-%
-\begin{equation*}
-\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} \circ \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}} ~:~
-  \ctyp{\aclrd{\textbf{r}}\apar\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
-  \rightarrow \ctyp{\aclrd{\textbf{r}}\azip\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}
-\end{equation*}
-%
-For flat coeffect calculi where $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
-splitting and then merging a context may lose some of the available contextual capabilities
-(and increase context requirements). This illuminates why our proofs of Lemma~\ref{thm:further-rev-contr}
-and Lemma~\ref{thm:further-rev-exch} relied on subcoeffecting in the flat system.
-
-\paragraph{Related and future work.}
-The semantics presented here serves main\-ly to inform the design of the type system. For this
-reason, we keep the semantics simple and close to the concrete notions used in the type system
-such as flat, structural and unified coeffect algebras. In particular, the indexed comonad
-structure consists of object mapping $\ctyp{r}{}$ that is indexed by a corresponding algebra:
-
-\begin{compactitem}
-\item[--] In the flat coeffect calculus, $\ctyp{\cclrd{r}}{}$ is indexed by scalar coeffects $\cclrd{r} \in \C$
-\item[--] In the structural coeffect calculus, the object mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is
-  indexed by vectors of scalars $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{m}\in \sclrd{\mathbb{N}}} \C^{\sclrd{m}}$
-\item[--] In unified coeffect calculus with a coeffect container $\SH\,\sclrd{\triangleleft}\,\SP$,
-  the mapping $\ctyp{\aclrd{\textbf{r}}}{}$ is indexed by shape-indexed scalars
-  $\aclrd{\textbf{r}} \in \bigcup_{\sclrd{s}\in \SH} \C^{\SP(\sclrd{m})}$
-\end{compactitem}
-
-\noindent
-In this treatment, the indices are formed by objects that are modelled outside of category theory.
-A fully categorical semantics can be found in a joint paper with Orchard \cite{coeffects-icfp14}.
-There, the index (coeffect scalars and shape-indexed scalars) are modelled as categories too.
-The definition is written using the following notation:
-
-\begin{compactitem}
-\item[--] $[\mathbb{C}, \mathbb{D}]$ is a category of functors between categories $\mathbb{C}$ and $\mathbb{D}$
-\item[--] $B^A$ is an exponential object $A \Rightarrow B$ in a Cartesian-closed category
-\end{compactitem}
-
-\noindent
-In the categorical semantics of coeffect calculi, coeffect scalars are modelled as a
-category of scalars $\cclrd{\mathbb{I}}$ such that $\textit{obj}(\cclrd{\mathbb{I}}) = \C$.
-The structure of the context (indexed comonad) over the input then becomes a functor
-indexed by the free-variable context shape $\slen{\Gamma}$ and coeffect annotations
-$\aclrd{\textbf{r}}$, \ie~$\ctyp{\slen{\Gamma}}_{\aclrd{\textbf{r}}}$:
-%
-\begin{equation*}
-\sem{\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau} : \ctyp{\slen{\Gamma}}_{\aclrd{R}}
-\sem{\Gamma} \rightarrow \sem{\tau}
-\end{equation*}
-%
-The structure $\ctyp{}$ can be thought of as a dependent product of functors $\ctyp{\sclrd{n}}$
-over possible shapes $\sclrd{n} \in \SH$ where:
-%
-\begin{equation*}
-\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]
-\end{equation*}
-%
-For a fixed context shape $\sclrd{n}$ the functor
-$\ctyp{\sclrd{n}} : \cclrd{\mathbb{I}}^{\SP(\sclrd{n})} \rightarrow [\mathbb{C}^{\SP(\sclrd{n})}, \mathbb{C}]$
-maps a coeffect indexed by positions $\SP(\sclrd{n})$ to a functor from a context
-$\mathbb{C}^{\sclrd{n}}$ to an object in $\mathbb{C}$. That is, given a coeffect annotation
-(matching the shape of the context), we get a functor $\in [\mathbb{C}^{\sclrd{n}},\mathbb{C}]$.
-From a programming perspective, this functor defines a data structure that models the additional
-context provided to the program. The shape of this data structure depends on the coeffect
-annotation $\cclrd{\mathbb{I}}^{\sclrd{n}}$. For example, in bounded reuse, the annotation defines
-the number of values needed for each variable and the functor will be
-formed by lists of length matching the required number.
-
-In the model above, shapes and positions are still treated as ordinary sets. Indeed, in the
-joint paper \cite{coeffects-icfp14}, we treat shapes as just sets of sets. In this chapter, we
-instead define \emph{coeffect container} as an extension of containers, which have a well-defined
-categorical structure \cite{semantics-containers}. A future work is adapting the semantics of
-the unified coeffect calculus so that it is based on the semantics of containers. However, note
-that context is \emph{not} a (simple) container, because containers are used as indices (coeffect
-annotations) of the object that wrap the free-variable context.
+Two readings of the signatures are possible, but they give quite different meanings to the operations:
+
+\begin{itemize}
+\item \textsc{Coeffect interpretation.} The \ident{counit} operation extracts a value
+  and does not require any additional context; the \ident{cobind} operation requires
+  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to
+  evaluate the function, ending with a value $\beta$ together with remaining context
+  $\cclrd{s}$.
+\item \textsc{CMTT interpretation.} The \ident{counit} operation evaluates a computation
+  that requires no additional context to obtain a $\alpha$ value; given a function that
+  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
+  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$
+  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
+  context requirements is eliminated by the function).
+\end{itemize}
+
+\noindent
+Although the different reading does not affect formal properties of the systems, it is important
+to understand the difference when discussing the two systems as they provide a different
+intuition.
+
+The sketch of a coeffect meta-language in the following section attempts to bridge the gap
+between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
+it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
+sets of variables. Future work on the coeffect meta-language would thus be an interesting
+development for both coeffect systems and CMTT.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Coeffect meta-language}
+
+The coeffect meta-language could be designed using both flat and structural indexed comonads.
+For simplicity, this section only discusses the flat variant. The syntax of types and terms
+of the language includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
+language constructs:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
+ e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
+  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
+  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
+  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2
+\end{array}
+\end{equation*}
+%
+The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
+operation of the comonad. To define meta-language for flat indexed comonads, we also include
+\kvd{split} and \kvd{merge} that embed the corresponding operations.
+
+\paragraph{Types for coeffect meta-language.}
+The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
+shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing
+rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
+
+The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
+a value from a comonadic context and corresponds to variable access in coeffect calculi. The
+\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
+and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
+The result is a computation that produces a $\tau_2$ value with the remaining context specified
+by $\cclrd{s}$. Note that the expressions $e_2$ and $e_1$ correspond to the first and second
+arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following
+CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
+although it differs because of our generalization of comonads where \ident{bind} composes
+coeffect annotations rather than requiring the same annotation everywhere.}.
+
+The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some
+transformation on one or two values in a context and then add the new value as a fresh variable
+to the variable context. We omit subcoeffecting, but it could be easily added following
+the method used elsewhere.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+
+{\small a.) Typing rules for the simply typed $\lambda$-calculus}
+
+\begin{equation*}
+\tyrule{var}
+  {x : \tau \in \Gamma}
+  {\Gamma \vdash x : \tau }
+\end{equation*}
+\begin{equation*}
+\tyrule{app}
+  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
+   \Gamma \vdash e_2 : \tau_2 }
+  {\Gamma \vdash e_1~e_2 : \tau_1 }
+\end{equation*}
+\begin{equation*}
+\tyrule{abs}
+  {\Gamma, x:\tau_1 \vdash e : \tau_2}
+  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
+\end{equation*}
+
+\vspace{1em}
+{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
+
+\begin{equation*}
+\tyrule{cobind}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
+\end{equation*}
+\begin{equation*}
+\tyrule{counit}
+  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
+  {\Gamma \vdash !e : \tau}
+\end{equation*}
+\begin{equation*}
+\tyrule{split}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
+   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
+  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+\begin{equation*}
+\tyrule{merge}
+  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
+   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
+   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
+  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
+\end{equation*}
+
+\figcaption{Type system for the (flat) coeffect meta-language}
+\label{fig:conclusions-cml}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Embedding flat coeffect calculus}
+
+The \emph{meta-language} approach of embedding comonads in a language is more general than the
+\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
+a type system for context-aware programming languages.
+
+However, it is worth demonstrating that the (flat) coeffect calculus can be embedded in the
+meta-language described in the previous section. This may be desirable, \eg~when using the
+meta-language for reasoning about context-aware computations. We briefly consider the embedding
+as it illuminates the relationship between coeffects and CMTT (unfortunately it is not possible
+to embed coeffect calculi in CMTT because of the more general annotations structure).
+
+Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
+we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
+meta-language. Note that the translation is indexed by $v$, which is a name of variable used
+to represent the entire variable context of the source language. The translation is defined in
+Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in
+Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+the operations of an indexed comonad.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\begin{equation*}
+\begin{array}{l}
+ \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
+   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
+   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
+      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
+ \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
+   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
+\end{array}
+\end{equation*}
+\figcaption{Embedding flat coeffect calculus in coeffect meta-language }
+\label{fig:conclusions-embed}
+\end{figure}
 
@@
 %
-%      ####                         #       #                   ##
+% 	 #####                 #     #                                                #
 % 	 #                     #     #                                                #
-%      #   #  # ##    ###    ###   ####    ##     ###    ###     #
-%      ####   ##  #      #  #   #   #       #    #   #      #    #
-%      #      #       ####  #       #       #    #       ####    #
-%      #      #      #   #  #   #   #  #    #    #   #  #   #    #
-%      #      #       ####   ###     ##    ###    ###    ####   ###
+% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   #
+% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #
+% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###
+% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #
+% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   #
 %
@@
 
-\section{Towards practical coeffects}
+\section{Related and future work}
+\label{sec:further-further}
+
+Most of the related and future work has been discussed throughout the thesis. This section summarizes
+two interesting directions for future work and the related work in the area. First, we look at a way of
+embedding coeffects in existing languages such as Haskell that is inspired by the Haskell's ``do'' notation
+(Section~\ref{sec:further-impl}). Second, we discuss extensions to the theory of coeffects that
+are needed for real-world programming languages and we consider how the system could be extended
+to generalize languages based on substructural and bunched logics (Section~\ref{sec:further-theory}).
+
+\subsection{Embedded context-aware DSLs}
 \label{sec:further-impl}
-
-As discussed earlier, the main focus of this thesis is the development of the much needed
-\emph{theory of context-aware programming languages} and so discussing the details of a practical
-implementation of coeffect tracking is beyond the scope of the thesis. However, this section
-briefly outlines one possible pathway towards this goal.
 
@@
 \emph{multiple} different notions of context-aware computations using just a \emph{single}
-common abstraction. For this reason, we advocate that future practical implementations of
-coeffects should not be single-purpose language features, but rather reusable abstractions that
+common abstraction. For this reason, future practical implementations of coeffects should not be
+single-purpose language features, but rather reusable abstractions that
 can be instantiated with a concrete \emph{coeffect algebra} specified by the user.
-
-In order to do this, programming languages need to provide two features; one that allows
+One approach towards this goal is to build a specialized context-aware language such as the
+one implemented in Chapter~\ref{ch:impl}. An alternative method is to embed context-aware
+DSLs into a richer target functional language.
+
+In order to do this, the target programming language needs to provide two features; one that allows
 embedding of context-aware computations themselves in programs akin to the ``do'' notation
-in Haskell (Section~\ref{sec:further-impl-embed}) and one that allows tracking of the
-contextual information in the type system (Section~\ref{sec:further-impl-types}). To make
-constraint resolution in type inference easier, we also discuss how the coeffect algebra
-can be simplified (Section~\ref{unified-impl-semilattice})
+in Haskell and one that allows tracking of the contextual information in the type system.
 
@@
 
-\subsection{Embedding contextual computations}
-\label{sec:further-impl-embed}
+\paragraph{Embedding contextual computations}
 
@@
 
-\subsection{Coeffect annotations as types}
-\label{sec:further-impl-types}
+\paragraph{Coeffect annotations as types}
 
@@
 
-The work \cite{effects-embedding} on embedding effect systems in Haskell demonstrated that the
-recent additions to the Haskell type system provide enough power to implement structures such
+In a joint work on embedding effect systems in Haskell \cite{effects-embedding}, we demonstrated that
+the recent additions to the Haskell type system provide enough power to implement structures such
 as sets at the type level. Using these to embed coeffect systems in Haskell is one fruitful
-future direction for applied coeffects.
-
-In dependently-typed programming languages such as Agda or Idris \cite{other-agda,other-idris},
+future direction for implementing applied context-aware languages.
+
+In dependently-typed languages such as Agda or Idris \cite{other-agda,other-idris},
 the embedding of coeffects can be implemented more directly (as terms implementing sets or lattices
@@
 
-\subsection{Alternative formulation using coeffect lattice}
-\label{sec:further-impl-semilattice}
-
-Compared to typical effect systems based on sets, coeffect systems have richer structure consisting
-of four operations ($\cseq,\czip,\cpar$ and a relation $\cleq$). This allows capturing interesting
-properties, but it makes working with coeffect annotations difficult -- it complicates tasks
-such as type generalization and constraint solving in type inference. Here, we look at one possible
-simplification of the structure.
-
-\begin{itemize}
-\item As mentioned in Section~\ref{sec:flat-calculus-algebra}, when the monoid $(\C,\cpar,\czero)$ is a
-semi-lattice, the $\cleq$ relation can often be expressed in terms of the $\cpar$ operation
-($\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}$).
-This does not give a definition equivalent to our, but it is a simplification that is consistent
-with all examples in this thesis.
-
-\item Another approach that works for most (but not all) systems we discussed is to require the structure
-$(\C,\cpar,\czip)$ to form a lattice. This is consistent with all our structural coeffect systems as
-well as flat systems for liveness and dataflow, but not with the system for tracking of implicit
-parameters (or sets of rebindable resources).
-\end{itemize}
-
-\noindent
-The system for tracking implicit parameters does not fit the lattice-based model, because it uses the
-same operation $\cup$ for both $\cpar$ and $\czip$.
-
-This also means that the system for implicit parameters cannot be used with a system where lambda
-abstraction duplicates the context requirements (Section~\ref{sec:flat-exts-lambda}).
-
-\paragraph{Lattice-based coeffects.} Putting implicit parameters aside, there are many other coeffect
-systems for which we could use the following definition of coeffect scalars (for simplicity, we only
-consider flat systems):
-
-\begin{definition}
-A \emph{\cclrd{flat coeffect lattice}} $(\C, \cseq, \cleq, \cunit, \czero)$ is a set
-$\C$ together with elements $\cunit, \czero \in \C$ and an ordering $\cleq$ such
-that $(\C, \cseq, \cunit)$ is a monoid and $(\C, \cleq)$ is a lattice with the least
-element $\czero$.
-
-Assuming $\cpar$ and $\czip$ are the \emph{least upper bound} and \emph{greatest lower bound}
-of the lattice, we require the following distributivity axioms:
-%
-\begin{align*}
-\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
-\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
-\end{align*}
-\end{definition}
-
-\noindent
-The definition provides all the operations of a flat coeffect algebra, but it defines a more rigid
-structure. In particular, $\czip$ and $\cpar$ are both defined in terms of $\cleq$. This simplifies
-the system as the operations of the underlying indexed comonad structure can be expressed using just
-$\cseq$ and $\cleq$:
-%
-\begin{equation*}
-\begin{array}{lcll}
- \ident{counit}_{\cunit} &\narrow{:}&
-    \ctyp{\cunit}{\alpha} \rightarrow \alpha \\
- \ident{cobind}_{\cclrd{r}, \cclrd{s}} &\narrow{:}&
-    (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) \\
-\ident{merge}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}&
-    \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{t}}{(\alpha \times \beta)} &
-    (\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s}) \\
-\ident{split}_{\cclrd{r},\cclrd{s},\cclrd{t}} &\narrow{:}&
-    \ctyp{\cclrd{t}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} &
-    (\cclrd{r}\,\cleq\,\cclrd{t}, \cclrd{s}\,\cleq\,\cclrd{t}) \\
-\ident{lift}_{\cclrd{r},\cclrd{s}} &\narrow{:}&
-    \ctyp{\cclrd{r}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\alpha} &
-    (\cclrd{s}\,\cleq\,\cclrd{r})
-\end{array}
-\end{equation*}
-%
-Note that we still need \emph{two distinct} operators in the system. Analysis of our examples shows
-that the operator used for sequential composition $\cseq$ is not necessarily related in any way to
-the ordering on the lattice provided by $\cleq$.
-
-However, the operations $\ident{merge}$, $\ident{split}$ and $\ident{lift}$ can now all
-use ordering. For example, the \ident{merge} operation previously returned
-a result with greatest lower bound $\cclrd{r}\;\czip\;\cclrd{s}$. Now, we require any
-element that is smaller than $\cclrd{r}$ and $\cclrd{s}$. This adds an implicit subcoeffecting
-to the operation -- for any $\cclrd{t}$ such that $\cclrd{t}\;\cleq\;\cclrd{r}$ and
-$\cclrd{t}\;\cleq\;\cclrd{s}$, it is also the case that $\cclrd{t}\;\cleq\;(\cclrd{r}\,\czip\,\cclrd{s})$.
-This means that we can compose the original \ident{merge} operation which returns greatest lower
-bound with $\ident{lift}$ to get the (less precise) operation required here.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-{\small a.) Lattice-based flat systems (selected rules):}
-
-\begin{equation*}
-\tyrule{abs}
-  {\coctx{\Gamma, x:\tau_1}{\cclrd{t}} \vdash e : \tau_2}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-~~(\cclrd{t}\,\cleq\,\cclrd{r}, \cclrd{t}\,\cleq\,\cclrd{s})
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
-   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
-  {\coctx{\Gamma}{\cclrd{u}} \vdash e_1~e_2 : \tau_2}
-~~(\cclrd{r}\,\cleq\,\cclrd{u}, (\cclrd{s} \,\cseq\, \cclrd{t})\,\cleq\,\cclrd{u})
-\end{equation*}
-
-\vspace{1.5em}
-{\small b.) Lattice-based structural systems (selected rules):}
-
-\begin{equation*}
-\tyrule{contr}
-  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{u}} \atimes \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
-~~(\cclrd{s}\,\cleq\,\cclrd{u}, \cclrd{t}\,\cleq\,\cclrd{u})
-\end{equation*}
-
-\figcaption{Flat and structural coeffects using lattice-based formulation}
-\label{fig:further-lattice-types}
-\end{figure}
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Lattice-based type system.}
-\hspace{-1em}
-Using lattice-based coeffect structure leads to the typing rules shown in Figure~\ref{fig:further-lattice-types}.
-In the flat coeffect system, this affects both application and abstraction -- in abstraction,
-the context-requirements of the body should be smaller than the context available on the
-call site and declaration site. In application, we require more than what is required by
-the sub-expressions.
-
-Structural coeffect systems only require least upper bound $\cpar$ in the (\emph{contr}) rule.
-This can be rewritten similarly to the flat (\emph{app}) rule --  when contracting two variables, the
-context requirements of the resulting variable are greater than the requirements associated with
-each of the original variables.
-
-In summary, the lattice-based coeffect system has some advantages over the design used in this
-thesis. In particular, it uses better-known structures and is simpler. However, we do not use
-it as it rules out one of the important motivating examples (implicit parameters) and, even with
-the simplification, solving constraints with $\cseq$ and $\cleq$ is still an open question.
-
-% =================================================================================================
-%
-%    #   #          #                    ##
-%    #   #          #                     #
-%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###
-%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   #
-%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    #####
-%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #
-%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###
-%                                                            #   #                #   #
-%                                                             ###                  ###
-% =================================================================================================
-
-\section{Coeffect meta-language}
-\label{sec:further-meta}
-
-In Section~\ref{sec:path-sem-langs}, we discussed two ways of using monads in programming
-language semantics as introduced by Moggi \cite{monad-notions}. The first approach is to use
-monads in the \emph{semantics} of an effectful language. The second approach is to extend the
-language with (additional) monadic constructs that can then be used for writing effectful
-monads explicitly.
-
-In this thesis, we focused on the first approach. In both flat and structural coeffect calculi,
-the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
-indexed comonads to give the semantics for the language and to derive type system for it.
-
-In this section, we briefly discuss the other technique. That is, we embed indexed
-comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
-type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
-context (semantically, this corresponds to an indexed comonad) and we add language constructs
-that correspond to the operations of indexed comonads.
-
-This section provides a brief sketch of the coeffect meta-language in order to highlight the relationship
-between coeffects and closely related work on contextual modal type theory (CMTT) \cite{logic-cmtt}.
-Developing the system further is an interesting future research direction.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffects and contextual modal type theory}
-
-As discussed in Section~\ref{sec:path-sem-contextdep}, context-aware computations are related
-to modal logics -- comonads have been used to model the $\square$ modality and as a basis for
-meta-languages that include $\square$ as a type constructor
-\cite{logic-intuitionistic-modal,logic-modal-reconstruction,logic-intuitionistic-modal,logic-cmtt}.
-Nanevski et al. \cite{logic-cmtt} extend an S4 term language to a contextual modal type theory (CMTT).
-From the perspective of this thesis, CMTT can be viewed as a \emph{meta-language} version of our
-coeffect calculus.
-
-\paragraph{Context in CMTT and coeffects.}
-Aside from the fact that coeffect calculi use comonads for \emph{semantics} and CMTT embeds
-comonads (the $\square$ modality) into the meta-language, there are two important differences.
-
-Firstly, the \emph{context} in CMTT is a set of variables required by a computation, which
-makes CMTT useful for meta-programming and staged computations. In coeffect calculi, the context
-requirements are formed by an abstract coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, etc.
-
-Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
-the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
-constructor refers to a value of type $\tau$ \emph{together} with additional context specified
-by the coeffect annotation $\cclrd{r}$ (\eg~a list of past values or additional resources).
-
-In contrast in CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the
-original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the
-context $\Psi$ in order to produce value $\tau$. This also changes the interpretation of the
-two operations of a comonad:
-\begin{equation*}
-\begin{array}{rcl}
- \ident{counit} &\narrow{:}& \ctyp{\cunit}{\alpha} \rightarrow \alpha\\
- \ident{cobind} &\narrow{:}& (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) \rightarrow
-    \ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}
-\end{array}
-\end{equation*}
-Two readings of the signatures are possible, but they give quite different meanings to the operations:
-
-\begin{itemize}
-\item \emph{Coeffect interpretation.} The \ident{counit} operation extracts a value
-  and does not require any additional context; the \ident{cobind} operation requires
-  context $\cclrd{r}\,\cseq\,\cclrd{s}$, uses the part specified by $\cclrd{r}$ to
-  evaluate the function, ending with a value $\beta$ together with remaining context
-  $\cclrd{s}$.
-\item \emph{CMTT interpretation.} The \ident{counit} operation evaluates a computation
-  that requires no additional context to obtain a $\alpha$ value; given a function that
-  turns a computation requiring context $\cclrd{r}$ into a value $\beta$, the \ident{cobind}
-  operation can turn a computation that requires context $\cclrd{r}\,\cseq\,\cclrd{s}$
-  into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
-  context requirements is eliminated by the function).
-\end{itemize}
-
-\noindent
-Although the different reading does not affect formal properties of the systems, it is important
-to understand the difference when discussing the two systems as they provide a different
-intuition.
-
-The sketch of a coeffect meta-language in the following section attempts to bridge the gap
-between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
-it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
-sets of variables. Future work on the coeffect meta-language would thus be an interesting
-development for both coeffect systems and CMTT.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Coeffect meta-language}
-
-The coeffect meta-language could be designed using both flat and structural indexed comonads.
-For simplicity, this section only discusses the flat variant. The syntax of types and terms
-of the language includes the type constructor $\ctyp{\cclrd{r}}{\tau}$ and four additional
-language constructs:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tau &\narrow{::=}& \alpha \sep \tau_1 \rightarrow \tau_2 \sep \ctyp{\cclrd{r}}{\tau}\\
- e &\narrow{::=}& v \sep \lambda x.e \sep e_1~e_2 \sep !e    \\[-0.25em]
-  &        \sep& \kvd{let\;box}~x=e_1~\kvd{in}~e_2           \\[-0.25em]
-  &        \sep& \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 \\[-0.25em]
-  &        \sep& \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2
-\end{array}
-\end{equation*}
-%
-The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
-operation of the comonad. To define meta-language for flat indexed comonads, we also include
-\kvd{split} and \kvd{merge} that embed the corresponding operations.
-
-\paragraph{Types for coeffect meta-language.}
-The type system for the language is shown in Figure~\ref{fig:conclusions-cml}. The first part
-shows the usual typing rules for simply-typed $\lambda$-calculus. For simplicity, we omit typing
-rules for pairs, but those need to be present as the \kvd{merge} operation works on tuples.
-
-The second part of the typing rules is more interesting. The (\emph{counit}) operation extracts
-a value from a comonadic context and corresponds to variable access in coeffect calculi. The
-\kvd{let\;box} construct (\emph{cobind}) takes an input $e_1$ with context $\cclrd{r}\,\cseq\,\cclrd{s}$
-and a computation that turns a variable $x$ with a context $\cclrd{r}$ into a value $\tau_2$.
-The result is a computation that produces a $\tau_2$ value with the remaining context specified
-by $\cclrd{s}$. Note that the expressions $e_2$ and $e_1$ correspond to the first and second
-arguments of the \ident{cobind} operation. The keyword \kvd{let\;box} is chosen following
-CMTT\footnote{The rule is similar to the \ident{letbox} rule for ICML \cite[p. 14]{logic-cmtt},
-although it differs because of our generalization of comonads where \ident{bind} composes
-coeffect annotations rather than requiring the same annotation everywhere.}.
-
-The \kvd{split} and \kvd{merge} constructs follow a similar pattern. They both apply some
-transformation on one or two values in a context and then add the new value as a fresh variable
-to the variable context. We omit subcoeffecting, but it could be easily added following
-the method used elsewhere.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-
-{\small a.) Typing rules for the simply typed $\lambda$-calculus}
-
-\begin{equation*}
-\tyrule{var}
-  {x : \tau \in \Gamma}
-  {\Gamma \vdash x : \tau }
-\end{equation*}
-\begin{equation*}
-\tyrule{app}
-  {\Gamma \vdash e_1 : \tau_2 \rightarrow \tau_1 &
-   \Gamma \vdash e_2 : \tau_2 }
-  {\Gamma \vdash e_1~e_2 : \tau_1 }
-\end{equation*}
-\begin{equation*}
-\tyrule{abs}
-  {\Gamma, x:\tau_1 \vdash e : \tau_2}
-  {\Gamma \vdash \lambda x.e : \tau_1 \rightarrow \tau_2 }
-\end{equation*}
-
-\vspace{1em}
-{\small b.) Additional typing rules arising from \emph{flat indexed comonads}}
-
-\begin{equation*}
-\tyrule{cobind}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cseq\,\cclrd{s}}{\tau_1} & \Gamma, x : \ctyp{\cclrd{r}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{let\;box}~x=e_1~\kvd{in}~e_2 : \ctyp{\cclrd{s}}{\tau_2}}
-\end{equation*}
-\begin{equation*}
-\tyrule{counit}
-  {\Gamma \vdash e : \ctyp{\cunit}{\tau}}
-  {\Gamma \vdash !e : \tau}
-\end{equation*}
-\begin{equation*}
-\tyrule{split}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}\,\cpar\,\cclrd{s}}{\tau_1} &
-   \Gamma, x:\ctyp{\cclrd{r}}{\tau_1}, y:\ctyp{\cclrd{s}}{\tau_1} \vdash e_2 : \tau_2 }
-  {\Gamma \vdash \kvd{split}~e_1~\kvd{into}~x,y~\kvd{in}~e_2 : \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{merge}
-  {\Gamma \vdash e_1 : \ctyp{\cclrd{r}}{\tau_1} &
-   \Gamma \vdash e_2 : \ctyp{\cclrd{s}}{\tau_2} \\
-   \Gamma, x:\ctyp{\cclrd{r}\,\czip\,\cclrd{s}}{(\tau_1 \times \tau_2)} \vdash e_2 : \tau_3 }
-  {\Gamma \vdash \kvd{merge}~e_1, e_2~\kvd{into}~x~\kvd{in}~e_2 : \tau_3 }
-\end{equation*}
-
-\figcaption{Type system for the (flat) coeffect meta-language}
-\label{fig:conclusions-cml}
-\end{figure}
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Embedding flat coeffect calculus}
-
-The \emph{meta-language} approach of embedding comonads in a language is more general than the
-\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
-a type system for context-aware programming languages.
-
-However, it is worth demonstrating that the (flat) coeffect calculus can be embedded in the
-meta-language described in the previous section. This may be desirable, \eg~when using the
-meta-language for reasoning about context-aware computations. We briefly consider the embedding
-as it illuminates the relationship between coeffects and CMTT (unfortunately it is not possible
-to embed coeffect calculi in CMTT because of the more general annotations structure).
-
-Given a typing judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ in the flat coeffect calculus,
-we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
-meta-language. Note that the translation is indexed by $v$, which is a name of variable used
-to represent the entire variable context of the source language. The translation is defined in
-Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in
-Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
-the operations of an indexed comonad.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\begin{equation*}
-\begin{array}{l}
- \sem{ \coctx{\Gamma}{\cunit}\vdash x_i : \tau_i }_v = \pi_i (!v) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}\;\cpar\;(\cclrd{s}\,\cseq\,\cclrd{t})} \vdash e_1~e_2 : \tau_2 }_v =\\[-0.25em]
-   \qquad \kvd{split}~v~\kvd{into}~v_s, v_{rt}~\kvd{in} \\[-0.25em]
-   \qquad \sem{ \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 }_{v_s}
-      ~(\kvd{let\;box}~v_r = v_{rt}~\kvd{in}~\sem{ \coctx{\Gamma}{\cclrd{r}} \vdash e_2 : \tau_1}_{v_r}) \\[0.25em]
- \sem{ \coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e:\tau_1 \xrightarrow{\cclrd{s}} \tau_2 }_v = \lambda x. \\[-0.25em]
-   \qquad \kvd{merge}~v,x~\kvd{into}~v_{rs}~\kvd{in}~\sem{ \coctx{(\Gamma,x:\tau_1)}{\cclrd{r}\,\czip\,\cclrd{s}} \vdash e : \tau_2 }_{v_{rs}}
-\end{array}
-\end{equation*}
-\figcaption{Embedding flat coeffect calculus in coeffect meta-language }
-\label{fig:conclusions-embed}
-\end{figure}
-
-% =================================================================================================
-%
-% 	 #####                 #     #                                                #
-% 	 #                     #     #                                                #
-% 	 #      #   #  # ##   ####   # ##    ###   # ##          #   #   ###   # ##   #   #
-% 	 ####   #   #  ##  #   #     ##  #  #   #  ##  #         #   #  #   #  ##  #  #  #
-% 	 #      #   #  #       #     #   #  #####  #             # # #  #   #  #      ###
-% 	 #      #  ##  #       #  #  #   #  #      #             # # #  #   #  #      #  #
-% 	 #       ## #  #        ##   #   #   ###   #              # #    ###   #      #   #
-%
-% =================================================================================================
-
-\section{Further work}
-\label{sec:conc-further}
-
-Most of the further work has been discussed throughout the thesis, so this section
-serves mainly as an overview. We follow the same structure as when discussing pathways to
-coeffects (Chapter~\ref{ch:pathways}). We consider further work related to syntactic approach
-to coeffect systems, categorical semantics of coeffects, and connections with substructural type
-systems and bunched logics.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Syntactic coeffect systems.}
+\subsection{Extending the theory of coeffects }
+\label{sec:further-theory}
+
+\paragraph{Richer coeffect languages.}
 The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
@@
 of information. Thus, the programming language also needs a form of type inference that can propagate
-such information. This has been done for effect systems \cite{effects-polymorphic}. For
-\emph{flat coeffects}, type inference is complicated by the non-determinism in the typing rule
-for lambda abstraction, but the problem is simpler for \emph{structural coeffects}. In addition
-to the \emph{algebra providers} (discussed in Section~\ref{sec:unified-impl-types}), it would also
-be worth developing a bidirectional type system \cite{types-bidirectional} for coeffects, where
-\emph{some} annotations are required, but most types and coeffects are reconstructed automatically.
+such information. This has been done for effect systems \cite{effects-polymorphic}. For coeffects,
+the prototype implementation discussed in Chapter~\ref{ch:impl} provides basic support for
+type inference, however formal treatment of the algorithm is left to future work.
 
@@
 
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Applications.}
-The last direction for further development is developing practical programming languages or
-libraries based on the theory presented in this thesis. We believe that the most useful approach
-is to extend programming languages with a reusable coeffect system and allow library developers
-define their own coeffect algebras. This option has been discussed in Section~\ref{sec:unified-impl}.
-For languages like Haskell, this can be done through a light-weight syntactic sugar akin to the
-``do'' notation.
-
-Another option is to use coeffects just as the underlying theory of a single-purpose programming
-language feature. When discussing why context-aware programming matters (Section~\ref{sec:intro-whymatters}),
-we covered a number of concrete problems that developers are facing today. Each of those would
-presents an important further work. The most important practical challenges are cross-compilation
-(often solved using complex $\prepk{\#if}$ pre-processor rules) and tracking of provenance
-and security properties.
-
 % ==================================================================================================
@@
 
-In this chapter, we completed our quest of presenting a \emph{unified} system for tracking contextual
-properties and then we discussed two important topics of future and related work that require
-in-depth discussion.
-
+In this chapter, we looked at three alternative directions for further development of coeffects.
 The key technical contribution of this chapter is the \emph{unified coeffect system}
 (Section~\ref{sec:further-unified}), which unifies the flat and structural system presented
-in the previous two chapters. To achieve this, we introduced \emph{coeffect container}, which
-determines how are coeffect annotations attached to variable contexts. We then discussed two
+in the previous two chapters. This provides an alternative to the separate treatment of
+per-context and per-variable contextual dependencies at the cost of making the system more
+complex. To unify the two, we introduced \emph{coeffect container}, which determines how are
+coeffect annotations attached to variable contexts. We then discussed two
 instances of the structure that capture flat and structural properties.
 
-In the rest of the chapter, we discussed two topics of future and related work. First
-(Section~\ref{sec:further-impl}), we considered pathways to practical implementations of
-coeffect systems, including a discussion of a novel lattice-based coeffect system, which is simpler,
-but cannot capture all our motivating examples. Finally, we discussed how our work relates to
-meta-language based on comonads (Section~\ref{sec:further-meta}). We present a \emph{coeffect
-meta-language} that follows similar style to CMTT, but is based on indexed comonads. Next, we
-discussed how to embed flat coeffect systems in this coeffect meta-language, which elucidates the
+Next, we discussed how our work relates to meta-language based on comonads (Section~\ref{sec:further-meta}).
+We present a \emph{coeffect meta-language} that extends a simple functional programming language
+with additional language constructs corresponding to contextual operations. This is similar to the
+target languages used in the translational semantics in Chapter~\ref{ch:semantics}, but it follows
+a notation similar to the one used by contextual modal type theory (CMTT). This elucidates the
 relationship between our work and CMTT.
+
+Finally, we considered an alternative approach to practical implementation of coeffect systems.
+Rather than building a specialized programming language (as we did in Chapter~\ref{ch:impl}), we
+discuss the technical requirements for embedding coeffects as a domain-specific language (DSL) into
+a statically-typed functional language with rich type system such as Haskell. This can be done by
+following the succesfull example of Haskell's ``do'' notation.

### 2016-05-16 text/AppendixB.tex
Spell checking

@@
 
-This appendinx provides additional details for some of the proofs for equational theory
+This appendix provides additional details for some of the proofs for equational theory
 of flat coeffect calculus from Chapter~\ref{ch:flat} and structural coeffect calculus

### 2016-05-16 text/Chapter02.tex
Spell checking

@@
 authors consider unimportant or present the work differently than originally intended. The reason
-is to fulfil the second goal of the chapter. When we do so, this is explicitly said in the text.
+is to fulfill the second goal of the chapter. When we do so, this is explicitly said in the text.
 
@@
 splits variable requirements differently and combines aspects of dynamic and static binding. For
-example, the language could use statis binding by default, but resort to dynamic binding if a
+example, the language could use static binding by default, but resort to dynamic binding if a
 variable is not available in the lexical scope. One such system is discussed in the next section.
@@
 In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
-delayed context requirements and the expression evalautes to $10$. In the example on the right,
+delayed context requirements and the expression evaluates to $10$. In the example on the right,
 the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling

### 2016-05-16 text/Chapter03.tex
Spell checking

@@
 capture some \emph{output impurity}. In contrast, coeffect systems track what a program requires from
-the execution envionment, or \emph{input impurity}.
+the execution environment, or \emph{input impurity}.
 
@@
 In (\emph{var}), the context contains just a single variable and so we do not even need to apply
-projection; (\emph{cosnt}) receives no variables and uses global constant lookup function $\delta$.
+projection; (\emph{const}) receives no variables and uses global constant lookup function $\delta$.
 In (\emph{abs}), we obtain two parts of the context and combine them into $(\mathbf{v}, x)$. This
@@
 copies of variables $y$ and $z$, respectively. The (\emph{app}) rule is similar in that it needs to
-split the input variable context. However, it needs to split values of multiple variabless:
+split the input variable context. However, it needs to split values of multiple variables:
 %

### 2016-05-16 text/Chapter04.tex
Spell checking

--- 2016-05-12 text/Chapter04.tex
@@
 %
-Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
+Note that the lambda abstraction in the syntax is written in the Church-style and requires a type
 annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
@@
 subtyping and additional constructs such as pairs in Section~\ref{sec:flat-exts}. When
-discussing procedure for choocing unique typing in Section~\ref{sec:flat-unique}, we consider
+discussing procedure for choosing unique typing in Section~\ref{sec:flat-unique}, we consider
 only the syntax-directed part of the system.
@@
 If a program written in a coeffect language with implicit parameters is well-typed in a type
-system presented in Figure~\ref{fig:flat-resolve-impl} then this identifies the unique prefered
+system presented in Figure~\ref{fig:flat-resolve-impl} then this identifies the unique preferred
 derivation for the program. We use this unique typing derivation to give the semantics of coeffect
 language with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm
-as discussed in Chatper~\ref{ch:impl}.
+as discussed in Chapter~\ref{ch:impl}.
 
@@
 defines the semantics of coeffect program (Chapter~\ref{ch:semantics}) and it is used in the
-implementation (Chatper~\ref{ch:impl}).
+implementation (Chapter~\ref{ch:impl}).
 
@@
 an expression $\lambda z.y$ changes the context demands attached to the type of the function.
-This is the case not just for the preferred unique typeing derivation, but for all possible typings
+This is the case not just for the preferred unique typing derivation, but for all possible typings
 that can be obtained using the (\emph{abs}) rule. However, this is not the case for all
@@
 It is perhaps somewhat unexpected that the annotation can be simplified in different ways for
-different examples. However, for all our systems, the simplfied annotation (right column in
+different examples. However, for all our systems, the simplified annotation (right column in
 Table~\ref{tab:flat-simplelet}) is more precise than the original (left column). Recall that
@@
 concrete discussion. We explain its category-theoretical motivation, we use it to define
-translational semantics (akin to Haskell's do notation) and we prove a soundness result that
-well-typed programs in flat coeffect calcului for implicit parameters and dataflow do not get
+translational semantics (akin to Haskell's ``do'' notation) and we prove a soundness result that
+well-typed programs in flat coeffect calculi for implicit parameters and dataflow do not get
 stuck in the translated version.

### 2016-05-16 text/Chapter05.tex
Spell checking

@@
 can be seen as a \emph{language framework} that simplifies the construction of concrete
-\emph{domain-specific} coeffect lnguages. In the previous chapter, we discussed how it provides a
+\emph{domain-specific} coeffect languages. In the previous chapter, we discussed how it provides a
 type system that tracks the required context. In this chapter, we show that the language framework
@@
 This chapter links together a number of different technical developments presented in this thesis.
-We take the flat coeffect calculus introduced in Chatper~\ref{ch:flat}, define its \emph{abstract
+We take the flat coeffect calculus introduced in Chapter~\ref{ch:flat}, define its \emph{abstract
 comonadic semantics} and use it to define a translation that gives a \emph{concrete operational
@@
 Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
-using the uniqye typing derivation specified in Section~\ref{sec:flat-unique}.
+using the unique typing derivation specified in Section~\ref{sec:flat-unique}.
 
@@
 
-The monadic and comonadic model outlined aboe represents at most a binary analysis of effects or
+The monadic and comonadic model outlined above represents at most a binary analysis of effects or
 context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no
@@
 $\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
-but the property holds for the three sample coeffect systems we consier. For systems with
+but the property holds for the three sample coeffect systems we consider. For systems with
 the above property, the \ident{split} operation can be expressed in terms of lifting
@@
   (\emph{tup}) and $\Gamma \vdash (v_1, \ldots, v_n) : \tau_1 \times\ldots\times \tau_n$ and
-  $\tau = \tau_i$. After applyng (\emph{prj}) reduction, $e' = v_i$ and so $\Gamma \vdash e' : \tau_i$.
+  $\tau = \tau_i$. After applying (\emph{prj}) reduction, $e' = v_i$ and so $\Gamma \vdash e' : \tau_i$.
 
@@
 target programming language rather than as composition of morphisms in a category. However,
-thanks to the relationship between $\lambda$-calculus and cartesian closed categories, both
+thanks to the relationship between $\lambda$-calculus and Cartesian closed categories, both
 interpretations are equivalent.
@@
 Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$
-written in a context-aware programming langugae that is translated to the target language as
+written in a context-aware programming language that is translated to the target language as
 (we write $\ldots$ for the omitted part of the translation tree):
@@
 
-The Figure~\ref{fig:semantics-ext-df} extensionds the target language with syntax, typing rules,
+The Figure~\ref{fig:semantics-ext-df} extends the target language with syntax, typing rules,
 additional translation rule and reductions for modelling dataflow computations. We introduce a
@@
 reduction rules and typing rules given in Figure~\ref{fig:semantics-target} and comonadically-inspired
-primtives defined in Figure~\ref{fig:semantics-ext} and also concrete notion of comonadically-inspired
+primitives defined in Figure~\ref{fig:semantics-ext} and also concrete notion of comonadically-inspired
 value and reduction rules for dataflow as defined in Figure~\ref{fig:semantics-ext-df}.
@@
 
-\paragraph{Parameteric effect monads.}
+\paragraph{Parametric effect monads.}
 Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our
@@
   computation, many of which cannot be captured by a monad (\eg~liveness or dataflow).
-  In terms of program analysis \cite{other-optcomp}, monas capture forward dataflow analyses and
+  In terms of program analysis \cite{other-optcomp}, monads capture forward dataflow analyses and
   comonads correspond to backward dataflow analyses.
@@
 \noindent
-Despite the differences, our implicit paremeters resemble, in many ways, the \emph{reader} monad.
+Despite the differences, our implicit parameters resemble, in many ways, the \emph{reader} monad.
 As discussed in Section~\ref{sec:semantics-related-monads}, the \emph{reader} monad is semantically
@@
 the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging. This can be
-supported in monadic computations by adding an aditional operation discussed next.
+supported in monadic computations by adding an additional operation discussed next.
 

### 2016-05-16 text/Chapter06.tex
Spell checking

@@
   programs from the structural coeffect calculus into a simple functional language with
-  uninterpreted comonadically-inspired primitivies. We give concrete operational semantics for the
+  uninterpreted comonadically-inspired primitives. We give concrete operational semantics for the
   target language for one of our sample languages and show that well-typed programs, produced by
@@
 \paragraph{Subcoeffecting.}
-Finbally, Figure~\ref{fig:struct-types} (c) shows the subcoeffecting rule (\emph{sub}).
+Finally, Figure~\ref{fig:struct-types} (c) shows the subcoeffecting rule (\emph{sub}).
 As with the flat coeffect calculus, we include it mainly to guide the understanding of the
@@
 In lambda abstraction (\emph{abs}) in Figure~\ref{fig:struct-types}, we assume that the bound
-varible is the last variable of the context. In the syntax-directed system, we do not make the
+variable is the last variable of the context. In the syntax-directed system, we do not make the
 same assumption. Instead, we use an auxiliary relation $\ident{findvar}_x$ that takes a typing
@@
 contain disjoint variables, which can be always obtained by exchange and contraction. In the
-syntax-driven systenm, we merge coeffects of shared variables explicitly. This is done in the
+syntax-driven system, we merge coeffects of shared variables explicitly. This is done in the
 \ident{mergevars} function.
@@
 \vspace{0.5em}\noindent\hangindent=0.7cm
-Casea (\emph{app}) similarly to (\emph{abs}).
+Case (\emph{app}) similarly to (\emph{abs}).
 \end{proof}
@@
   If $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ (using the rules in
-  Figure~\ref{fig:struct-syntax-types}) then there is a uniqye typing derivation
+  Figure~\ref{fig:struct-syntax-types}) then there is a unique typing derivation
   using the typing rules from Figure~\ref{fig:struct-types} with a conclusion
   $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ obtained by induction over
-  the original typing derivation as folows:
+  the original typing derivation as follows:
 
@@
 semantics is written as composition of morphisms using a number of auxiliary definitions. Due to
-the eqivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
+the equivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
 as specifying translation to a target functional language in Section~\ref{sec:struct-transl}.
@@
 \paragraph{Properties.}
-The concrete categortical semantics presented in this section is a generalization of the concrete
+The concrete categorical semantics presented in this section is a generalization of the concrete
 semantics given when introducing context-aware programming languages in Chapter~\ref{ch:applications}.
@@
 
-\figcaption{Trnaslation from a structural coeffect calculus}
+\figcaption{Translation from a structural coeffect calculus}
 \label{fig:struct-translation}
@@
 Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
-written in a structural context-aware programming langugae that is translated to the target language as
+written in a structural context-aware programming language that is translated to the target language as
 (we write $\ldots$ for the omitted part of the translation tree):
@@
 $\Gamma \vdash_{\ident{vec}} e : \tau, \cclrd{n}$. The judgement checks that a vector of expressions
-$\textbf{e}$ contains exactly $\cclrd{n}$ expressions of type $\tau$. This is caputred by the
+$\textbf{e}$ contains exactly $\cclrd{n}$ expressions of type $\tau$. This is captured by the
 (\emph{vec}) rule, which is then used to check individual elements of the context in the

### 2016-05-16 text/Chapter07.tex
Spell checking

@@
 
-The implementation directly follows the thoery presented in Chapters~\ref{ch:flat}, \ref{ch:semantics}
+The implementation directly follows the theory presented in Chapters~\ref{ch:flat}, \ref{ch:semantics}
 and \ref{ch:structural}. It consists of a common framework that provides type checking and
@@
 in a more practical way. However, we do not intend to build a complete real-world programming language.
-For this reason, the implementation is available primarilly as an interactive essay\footnote{The
+For this reason, the implementation is available primarily as an interactive essay\footnote{The
 \emph{interactive essay} format is based on Bret Victor's work on \emph{explorable explanations} \cite{essay-explorable}
@@
   This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
-  to a simple target language that is then iterpereted. We also discuss how the common framework makes it
+  to a simple target language that is then interpreted. We also discuss how the common framework makes it
   easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).
@@
 translation and operational semantics). For structural coeffects, the same components are discussed
-in Chapter~\ref{ch:structural}. In this section, we discuss how teose provide foundation for the
+in Chapter~\ref{ch:structural}. In this section, we discuss how these provide foundation for the
 implementation.
@@
   (\emph{abs}) rule of the flat coeffect calculus. For liveness, the handling would be the same
-  as for atafloww, but for other flat coeffect systems, another resolution mechanism might be used
+  as for dataflow, but for other flat coeffect systems, another resolution mechanism might be used
   instead.
@@
 ambiguity resolution algorithm for the typing of implicit parameters and the second one exposes the
-difference between flat and structural system forataflowow.
+difference between flat and structural system for dataflow.
 
@@
 
-\paragraph{Interactive program excution.}
+\paragraph{Interactive program execution.}
 
@@
 
-The interactive program execution lets the reader run sample programs, but not in a particuarly realistic
+The interactive program execution lets the reader run sample programs, but not in a particularly realistic
 context. To show a more real-world scenario, the essay includes a widget shown in Figure~\ref{fig:essay-df}.
@@
 understanding the translation based just on the rules is harder than for monads. The essay lets
-readers see translation for carefuly chosen case studies that illustrate the key aspects of the
+readers see translation for carefully chosen case studies that illustrate the key aspects of the
 translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
@@
   programs in the target language. The interpretation does not handle the source
-  language and so programs containing context-aware contstructs cannot be interpreted directly.
+  language and so programs containing context-aware constructs cannot be interpreted directly.
 

### 2016-05-16 text/Chapter08.tex
Spell checking

@@
 that generate types based on external data sources or other information in order to provide easy
-access to data or services. A similar approach could be used for embedding \emph{algberas} such
+access to data or services. A similar approach could be used for embedding \emph{algebras} such
 as coeffect algebras into the type system. An \emph{algebra provider} would then be a library that
@@
 a statically-typed functional language with rich type system such as Haskell. This can be done by
-following the succesfull example of Haskell's ``do'' notation.
+following the successful example of Haskell's ``do'' notation.

### 2016-05-16 text/Chapter09.tex
Edits in further work and the rest (Chapter 8)
Spell checking

@@
 
-\section{Overview of contributions}
+\section{Contributions}
 \label{sec:conc-summary}
@@
 functional language. The target language includes uninterpreted comonadically-inspired primitives
-whose concrete reduction rules need to be provided for each concrete context-aware deomain-specific
+whose concrete reduction rules need to be provided for each concrete context-aware domain-specific
 language. We give concrete definitions and prove type safety for three sample context-aware

### 2016-05-29 text/Chapter01.tex
Spell checking
Addressing feedback, fix demands (Chapter 6)

@@
 To support cross-compilation, the language needs to be context-aware. Each of the function has a
-number of context requirements. The \ident{fetchNews} function needs to have access to a database;
+number of context demands. The \ident{fetchNews} function needs to have access to a database;
 \ident{fetchLocalNews} needs access to a GPS sensor and to a network (to perform the remote call).
@@
 
-The approach advocated in this thesis is to track information about context requirements using
+The approach advocated in this thesis is to track information about context demands using
 the type system. To make this practical, the system should also provide at least a partial support for
-automatic type inference, as the information about context requirements makes the types more complex.
+automatic type inference, as the information about context demands makes the types more complex.
 An inspiring example might be the F\# support for units of measure \cite{app-units} -- the user has
@@
 a function that takes an input of type $\tau_1$, produces a result of type $\tau_2$ and has additional
-context requirements specified by $c$. In the above example, the annotation $c$ is simply a set of
+context demands specified by $c$. In the above example, the annotation $c$ is simply a set of
 required resources or capabilities. However, a more complex structure could be used as well, for example,
@@
 
-\paragraph{Tracking context requirements.}
+\paragraph{Tracking context demands.}
 
-The systems for tracking of context requirements developed in this thesis are inspired by the idea
+The systems for tracking of context demands developed in this thesis are inspired by the idea
 of effect systems. To demonstrate our approach, consider the following call from the sample program
@@
 \item Related work is presented in Chapter~\ref{ch:pathways} and, together with further work,
-  throughout the thesis. Two important directions deserve furhter exploration. In Chapter~\ref{ch:further},
+  throughout the thesis. Two important directions deserve further exploration. In Chapter~\ref{ch:further},
   we outline a unified coeffect system that is capable of capturing both flat and structural

### 2016-05-29 text/Chapter02.tex
Addressing feedback, fix demands (Chapter 6)

@@
 ($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
-\ident{y}. In static binding, the context requirements of the body can be placed on the scope in
+\ident{y}. In static binding, the context demands of the body can be placed on the scope in
 which the function is defined (declaration-site). In dynamic binding, the requirements are
@@
 In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
-delayed context requirements and the expression evaluates to $10$. In the example on the right,
+delayed context demands and the expression evaluates to $10$. In the example on the right,
 the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
@@
 in the environment when calling the function. It is worth noting that the syntax is similar to the
-one used by type-class constraints. Those can be viewed as context requirements too
+one used by type-class constraints. Those can be viewed as context demands too
 (Section~\ref{sec:applications-flat-impl}).
@@
 The three different binding mechanisms discussed so far can be seen as different ways of splitting
-context requirements of a particular kind. Dynamic and static binding represent the opposite ends
+context demands of a particular kind. Dynamic and static binding represent the opposite ends
 of the design space and Haskell's implicit parameters are an interesting point inside the wider space.
@@
 of several motivating examples. However, implicit parameters are a valuable example, because they
-clearly illustrate the ambiguity inherent in context-aware programs -- the context requirements of
+clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
 a function can be satisfied using the context available at declaration-site or using the context
@@
 We also aim to find a description of context-aware languages that does not make ad-hoc decisions
-about how context requirements are split between the declaration-site and the call site. While
+about how context demands are split between the declaration-site and the call site. While
 Haskell's solution for implicit parameters might be the most reasonable one, different notions
@@
 independently -- the semantics of context-aware programs and the domain-specific way of resolving
-how context requirements are satisfied. In our treatment of implicit parameters, we consider
+how context demands are satisfied. In our treatment of implicit parameters, we consider
 multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
@@
 Most of the typing rules correspond to those of effect systems. Variable access (\emph{var})
-has no context requirements, application (\emph{app}) combines context requirements of the two
+has no context demands, application (\emph{app}) combines context demands of the two
 sub-expressions and latent context-requirements of the function.
@@
 from the marriage of monads and effects. We develop coeffect calculus with an type system that
-tracks the context requirements more precisely (by annotating the types) and we add indexing
+tracks the context demands more precisely (by annotating the types) and we add indexing
 to comonads and link the two by giving a formal semantics. The indexing allows us to capture

### 2016-05-29 text/Chapter03.tex
Addressing feedback, fix demands (Chapter 6)

@@
 systems using judgements of the form $\coctx{\Gamma}{\cclrd{\sigma}} \vdash e : \tau$, associating
-the context requirements with $\Gamma$. Thus, we extend the traditional notion of free-variable
+the context demands with $\Gamma$. Thus, we extend the traditional notion of free-variable
 context $\Gamma$ with richer notions of context. This notation emphasizes the right intuition,
@@
 example of flat coeffects. These could be, for example, a database, GPS sensor or access to the current
-time. We also outlined that such context requirements can be tracked as part of the typing assumption,
+time. We also outlined that such context demands can be tracked as part of the typing assumption,
 for example, say we have an expression $e$ that requires GPS coordinates and the current time.
@@
 we decouple the definition of semantics from the domain-specific choice that determines how
-context requirements are satisfied. We capture the choice in the type and give semantics over a
+context demands are satisfied. We capture the choice in the type and give semantics over a
 \emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
@@
 Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
-$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as
-$\aclrd{\textbf{r}}\times\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.
+$\{ 1, \ldots, m \}$, the associative operation for concatenation $\vtimes$ is defined as
+$\aclrd{\textbf{r}}\vtimes\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.
 
@@
 but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should
-be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.
+be seen as $\Gamma_1 \vtimes \langle x\!:\!\tau\rangle \vtimes \Gamma_2$.
 
@@
 Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that
-introduces primitive context requirements is (\emph{param}). Accessing a parameter \ident{?param}
+introduces primitive context demands is (\emph{param}). Accessing a parameter \ident{?param}
 requires it to be available in the context. The context may provide more (unused)
@@
 When we read the rules from the top to the bottom, application (\emph{app}) and let binding
-(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
+(\emph{let}) simply union the context demands of the sub-expressions. However, lambda abstraction
 (\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
@@
 Type classes and implicit parameters show two important points about flat coeffect systems.
-First, the context requirements are associated with some \emph{scope}, such as the body
+First, the context demands are associated with some \emph{scope}, such as the body
 of a function. Second, they are associated with the input. To call a function that takes an
@@
 Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce
-non-determinism. It simply duplicates the context requirements. However, this still matches the
+non-determinism. It simply duplicates the context demands. However, this still matches the
 property of coeffects that impurities cannot be delayed or thunked and attached just to the
@@
 \tyrule{let}
-  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad
+  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad
     \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
-  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \atimes (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
 \end{equation*}
@@
 reveals which parts of the systems differ and which parts are shared. In particular, both systems
-use the same vector operations ($\times$, $\alift{\cclrd{\textnormal{--}}}$) and also share the
+use the same vector operations ($\atimes$, $\alift{\cclrd{\textnormal{--}}}$) and also share the
 lambda abstraction rule (\emph{abs}). They differ in the primitive values used to annotate used

### 2016-05-29 text/Chapter04.tex
Addressing feedback, fix demands (Chapter 6)

@@
 \subsection{Implicit parameters}
+\label{sec:flat-unique-impl}
 
@@
 The focus of this chapter is on the general coeffect system and so we do not discuss the
-domain-specific reduction rules for individual context-aware language. Some work on operational
-semantics of general coeffect systems has been done by Brunel et al. \cite{coeffects-quantitative}.
+domain-specific reduction rules for individual context-aware language. Some work on both operational
+and denotational semantics of general coeffect systems has been done by Brunel et al.
+\cite{coeffects-quantitative} and Breuvart and Pagani \cite{coeffects-modelling}.
 We give formal semantics of implicit parameters and dataflow in Chapter~\ref{ch:semantics} by
@@
 In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$
-where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
+where $\czip = \cseq = \cpar$ is an idempotent and commutative operation and
 $\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:

### 2016-05-29 text/Chapter05.tex
Addressing feedback, fix demands (Chapter 6)

@@
   and preservation proofs, showing that well-typed programs produced by translation from two
-  coeffect languages do not get stuck (Section~\ref{sec:semantics-proofs})
+  coeffect languages do not go wrong (Section~\ref{sec:semantics-proofs})
 

### 2016-05-29 text/Chapter06.tex
Addressing feedback, fix demands (Chapter 6)

@@
 
-The flat coeffect system captures a number of interesting use-cases (implicit parameters, liveness
-and dataflow), but provides relatively imprecise approximation and weak syntactical properties.
-Some of those examples can be also seen as per-variable properties. For those, structural coeffect
-systems give a more precise information about the context. However, we also look at other applications
-that arise from the work on substructural logics, a pathway to coeffects discussed in
-Section~\ref{sec:path-logic}.
+The flat coeffect system captures a number of interesting use-cases. For some of those (liveness
+and dataflow), flat coeffects provide only imprecise approximation (for example, marking the whole
+context as live rather than marking individual variables). Dataflow and liveness (but not implicit
+parameters) can be also seen as per-variable properties. For those, structural coeffect systems
+capture more precise information about the context. However, we also look at other applications
+that arise from the work on substructural logics discussed in Section~\ref{sec:path-logic}.
 
@@
 programs in a simple target functional language. We prove syntactic safety for a sample target
-language, showing that ``well-typed context-aware programs do not get stuck''.
+language, showing that ``well-typed context-aware programs do not go wrong''.
 
@@
 \item We present a \emph{structural coeffect calculus} as a type system that is parameterized
-  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}. We show how the
-  system captures fine-grained liveness and dataflow information, as well as a calculus for bounded
+  by a \emph{structural coeffect algebra} (Section~\ref{sec:struct-calculus}). We show how the
+  system captures pre-variable liveness and dataflow information, as well as a calculus for bounded
   reuse (checking how many times is a variable accessed).
@@
 
-\item We discuss the equational theory of the calculus. For structural coeffects, the substitution
-  lemma holds for all instances of the calculus. Thus, we prove the type-preservation property
-  for all structural calculi for both call-by-name and call-by-value (Section~\ref{sec:struct-syntax}).
-
-\item We extend indexed comonads introduced in the previous chapter to \emph{structural indexed
+\item We discuss the equational theory of the calculus. We show that type-preservation
+  holds for all examples of the structural calculus we consider, for both call-by-name and call-by-value
+  reductions (Section~\ref{sec:struct-syntax}) and we
+  explore a number of extensions to the minimal calculus based on $\lambda$-calculus,
+  including subcoeffecting and let binding (Section~\ref{sec:structural-exts}).
+
+\item We extend the indexed comonads introduced in the previous chapter to \emph{structural indexed
   comonads} and use them to provide the semantics of structural coeffect calculus
@@
   target language for one of our sample languages and show that well-typed programs, produced by
-  translation from the coeffect calculus do not get stuck.
+  translation from the coeffect calculus do not go wrong.
 \end{itemize}
@@
 
-Compared to Chapter~\ref{ch:flat}, structural coeffect calculi we consider are more homogeneous and
-so finding the common pattern is easier. However, the systems are more complicated as they need to
-keep annotations attached to individual variables and thus require explicit structural rules. Before
-looking at the system, we briefly consider the most important related work.
+Compared to Chapter~\ref{ch:flat}, the structural coeffect calculi we consider are more homogeneous
+and so finding the common pattern is easier. However, the systems are more complicated as they need
+to keep annotations attached to individual variables and thus require explicit structural rules.
+Before looking at the system, we briefly consider the most important related work.
 
@@
 $\lambda$-calculus is asymmetric in that an expression has multiple inputs (variables in the
-context), but just a single result (the resulting value) and so monads and effects have no notion
-directly corresponding to structural coeffect systems.
-
-The work in this chapter is more closely related to substructural type systems \cite{substruct-attpl-intro}.
+context), but just a single result (the resulting value). So, while there is only one notion of
+monadic effect system, there are two separate notions of coeffect system - one that keeps coeffect
+annotations per-environment and one that keeps coeffect annotations per-variable.
+
+The work in this chapter is related to substructural type systems \cite{substruct-attpl-intro}.
 Substructural systems remove some or all of \emph{weakening}, \emph{contraction} and
@@
 
-Our work follows the ``language semantics'' style in that we provide a structural semantics to
-the terms of ordinary $\lambda$-calculus. By contrast the most closely related work has been done in
-the meta-language style, which extends the terms and types of a language with constructs for explicitly
-manipulating the context. This includes Contextual Modal Type Theory (CMTT) \cite{logic-cmtt}, where variables
-may be of a type $A[\Psi]$ denoting a value of type $A$ that requires context $\Psi$. In CMTT,
+Our work follows the ``language semantics'' style in that we provide semantics to the terms of
+ordinary $\lambda$-calculus. By contrast the closely related work on Contextual Modal
+Type Theory (CMTT) \cite{logic-cmtt} follows the meta-language style. CMTT extends the terms and
+types of a language with constructs for explicitly manipulating the context. Variables
+of type $A[\Psi]$ denote a value of type $A$ that requires context $\Psi$. In CMTT,
 $A[\Psi]$ is a first-class type, while structural coeffect systems do not expose coeffect
@@
 
-In the structural coeffect calculus, a vector of variables forming the free-variable context is
-annotated with a vector of primitive (scalar) coeffect annotations. These annotations differ
-for different coeffect calculi and their properties are captured by the definition of \emph{structural
-coeffect scalar} below. The scalar annotations can be \eg~integers (how many past values we
-need) or values of a two-point lattice specifying whether a variable is live or not.
-
-Scalar annotations are written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$ (following the style used in
-the previous chapter). Functions always have exactly one input variable and so they are annotated
-with a single coeffect scalar. Thus the expressions and types of structural coeffect calculi are the
-same as in the previous chapter. The only difference is that $\cclrd{r}, \cclrd{s}, \cclrd{t}$ now
-range over values of a \emph{structural coeffect scalar} rather than over values of a \emph{flat
-coeffect algebra} (the definitions are related, but not the same).
+In the structural coeffect calculus, functions are annotated with a primitive (scalar) coeffect
+annotations. A vector of variables forming the free-variable context is annotated with a vector of
+coeffect annotations. These annotations differ for various coeffect calculi; their properties are
+captured by the definition of \emph{structural coeffect scalar} below. The scalar annotations can
+be \eg~integers (how many past values we need) or values of a two-point lattice specifying whether
+a variable is live or not. The expressions and types of the structural coeffect calculus are defined
+as follows:
 %
@@
 \begin{array}{rcl}
-e &::=& x n \sep \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
+e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
 \tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
@@
 %
-In the previous chapter, the free variable context $\Gamma$ was treated as a set. In the
-structural coeffect calculus, the order of variables matters. Thus we treat a free variable
-context as a vector with a uniqueness condition. We also write $\xlen{-}$ for the length
-of the vector:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \Gamma &\narrow{=}& \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
- \xlen{\Gamma} &\narrow{=}& n
-\end{array}
-\end{equation*}
-%
-For readability, we use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
-typing judgements, but the free variable context should be understood as a vector.
-Furthermore, the notation $\Gamma_1, \Gamma_2$ is here seen as the tensor product.
-Given $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and
-$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
-$\Gamma_1, \Gamma_2 = \Gamma_1 \times \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
-
-Free variable contexts are annotated with vectors of structural coeffect scalars. We write vectors
-of coeffects as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$. Meta-variables ranging over vectors are
-written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ (using bold face
-and colour to distinguish them from scalar meta-variables) and the length of a coeffect vector
-is written as $\alen{\aclrd{\textbf{r}}}$. The structure for working with vectors
-of coeffects is provided by the definition of \emph{structural coeffect algebra} discussed next.
+The expressions and types of structural coeffect calculus are similar to those of the flat coeffect
+calculus with two differences. First, we omit the \kvd{let} construct in the core language. In
+structural coeffects, let binding can be defined as a derived rule using abstraction and application
+(Section~\ref{sec:structural-exts-let}). Second, the coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$
+on function type now range over values of a \emph{structural coeffect scalar}.
 
@@
 
-The \emph{structural coeffect scalar} structure is similar to \emph{flat coeffect algebra} with the
-exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
+The \emph{structural coeffect scalar} structure is similar to that of \emph{flat coeffect algebra}
+with the exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
 modelling sequential composition of computations and a monoid $(\C, \cpar, \czero)$ representing
-pointwise composition, as well as a relation $\cleq$ that defines subcoeffecting.
+pointwise composition, as well as the $\cleq$ relation.
 
@@
 %
-The following distributivity axioms hold:
+In addition, the following distributivity axioms hold:
 \begin{align*}
@@
 \end{align*}
-%
-In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
-\begin{equation*}
-\begin{array}{l}
- \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
- \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
-\end{array}
-\end{equation*}
 \end{definition}
@@
 \noindent
-The structural coeffect scalar structure resembles flat coeffect algebra, but it differs in a
-number of important ways:
+The structural coeffect scalar structure resembles flat coeffect algebra, but it differs in two important ways:
 %
@@
   systems, we can track individual variable accesses and not requiring idempotence allows interesting
-  systems such as the one for bounded reuse (Section~\ref{sec:struct-semantics-examples}).
+  systems such as that for bounded reuse (Section~\ref{sec:struct-semantics-examples}).
 
@@
   available from the declaration site and the call site or, in the syntactic reading, to split the
-  context requirements. In structural systems, this is handled directly by vectors and so $\czip$
-  is no longer needed.
-
-\item We now require two properties that relate the ordering $\cleq$ with coeffects composed
-  via $\cpar$ and with $\czero$. These were previously required for bottom-pointed substitution lemma
-  (Lemma~\ref{thm:cbn-substitution-bot}). Here, we require them for all structural coeffects as it holds
-  for all our examples; it is needed for the structural version of the substitution lemma
-  (Lemma~\ref{thm:structural-substitution}).
+  context demands. Structural systems \emph{append vectors} of annotations instead of
+  \emph{merging annotations} and so $\czip$ is no longer needed.
 \end{itemize}
@@
 In the structural coeffect calculus, the scalar coeffect structure is supplemented by a vector
-structure. Combining and splitting of coeffects becomes just vector concatenation or splitting,
-respectively; these are provided by the tensor product. The operations on vectors are indexed by
-integers representing the lengths of the vectors. The additional structure required by the type
-system for structural coeffect calculi is given by the following definition.
+structure. The vector structure is used to manipulate vectors of coeffect scalars attached to a
+variable context. The required structure is captured by the following definition.
 
@@
 \begin{itemize}
+\vspace{-0.3em}
 \item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
@@
 
-\item An operation that constructs a vector from scalars indexed by the vector length
+\vspace{-0.3em}
+\item A family of operations (indexed by the vector length) that construct a vector from scalars
   $\alift{-}_{\sclrd{n}} : \C \times \ldots \times \C \rightarrow \C^{\sclrd{n}}$
@@
 
+\vspace{-0.3em}
 \item A pointwise extension of the $\cseq$ operator written as $\cclrd{t} \,\aseq\, \aclrd{\textbf{s}}$
@@
 
+\vspace{-0.3em}
 \item An indexed tensor product $\atimes_{\sclrd{n},\sclrd{m}} : \C^{\sclrd{n}} \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{n}+\sclrd{m}}$
-  that is used in both directions -- for vector concatenation and for splitting --
+  that is bijective and is used in both directions -- for vector concatenation and for splitting --
   which is defined as $\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}
@@
 The fact that the tensor product $\atimes_{\sclrd{n},\sclrd{m}}$ is indexed by the lengths
-of the two vectors means that we can use it unambiguously for both concatenation of vectors
-and for splitting of vectors, provided that the lengths of the resulting vectors are known.
+of the two vectors means that we can use it unambiguously for concatenating two vectors and
+splitting of a vector, provided that the lengths of the resulting vectors are known.
 In the following text, we usually omit the indices and write just
@@
 can be determined from the lengths of the matching free variable context vectors.
-More generally, we could see the the coeffect annotations as \emph{containers}
+More generally, we could see the coeffect annotations as \emph{containers}
 \cite{types-containers}. This approach is used in Section~\ref{sec:further-unified} to unify flat
 and structural systems.
+
+%---------------------------------------------------------------------------------------------------
+
+\subsection{Structural coeffect types}
+\label{sec:struct-calculus-types}
+
+The type system for the structural coeffect calculus is shown in Figure~\ref{fig:struct-types}.
+It is similar to substructural type systems
+\cite{substruct-attpl-intro} in how it handles free variable contexts. In the type system for
+flat coeffects (Section~\ref{sec:flat-calculus-types}), the (\emph{var}) rule implicitly allows
+\emph{weakening} and \emph{exchange} by ignoring other variables in the context and
+(\emph{app}) implicitly allows \emph{contraction} by passing the same context to both sub-expressions.
+
+In the structural system, this is made explicit by adding \emph{structural rules}.
+While substructural type systems usually remove some of the rules, we keep all three and
+use them to track how variables are used. This is done by manipulating the coeffect annotations
+in parallel with manipulating the variable contexts. As in substructural type systems, (\emph{app})
+checks the types of sub-expressions in disjoint parts of the free variable contexts and
+(\emph{var}) requires the context to contain exactly one variable. The typing rule for let binding in
+structural coeffects is a derived rule and we discuss it later in Section~\ref{sec:structural-exts-let}.
+
+~
 
@@
   {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x:\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\end{equation*}
-\begin{equation*}
-\tyrule{let}
-  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
-    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
-  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
 \end{equation*}
@@
 
-\vspace{1em}
-{\small c.) Non-syntax-directed subcoeffecting rule:}
-\begin{equation*}
-\tyrule{sub}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
-\quad
-(\cclrd{s'} \cleq \cclrd{s})
-\end{equation*}
-
 \figcaption{Type system for the structural coeffect calculus}
@@
 
-\subsection{Structural coeffect types}
-\label{sec:struct-calculus-types}
-
-The type system for the structural coeffect calculus is similar to substructural type systems
-in how it handles free variable contexts. The \emph{syntax-driven} rules do not implicitly allow
-weakening, exchange or contraction -- this is done by checking the types of sub-expressions
-in disjoint parts of the free variable context. Unlike in substructural logics, our system
-does allow weakening, exchange and contraction, but via explicit \emph{structural} rules that
-perform corresponding transformation on the coeffect annotation.
+\paragraph{Variable contexts.}
+In Chapter~\ref{ch:flat}, the free variable context $\Gamma$ was treated as a set. In the type
+system for the structural coeffect calculus, the variable context is treated as a vector, with
+an additional condition that a cannot appear multiple times. We also write $\xlen{-}$ for the
+length of the vector:
+%
+\begin{equation*}
+\begin{array}{l}
+ \Gamma = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle \qquad \textnormal{such that}~~\forall i, j ~.~ i \neq j \implies x_i \neq x_j\\
+ \xlen{\langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle} = n
+\end{array}
+\end{equation*}
+%
+We use the usual notation $x_1\!:\!\tau_1, \ldots, x_1\!:\!\tau_1 \vdash e : \tau$ for
+typing judgements, but the free variable context should be understood as a vector.
+The notation $\Gamma_1, \Gamma_2$ is used for concatenation of vectors of variables. That is,
+given a context $\Gamma_1 = \langle x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n \rangle $ and a context
+$\Gamma_2 = \langle x_{n+1}\!:\!\tau_{n+1}, \ldots, x_m\!:\!\tau_m \rangle $ then
+$\Gamma_1, \Gamma_2 = \langle x_1\!:\!\tau_1, \ldots, x_m\!:\!\tau_m \rangle$.
+
+In the typing rules, free variable contexts are annotated with vectors of structural coeffect scalars,
+written as $\coctx{x_1\!:\tau_1,\ldtos,x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}$.
+Meta-variables ranging over coeffect vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
+(using bold face and colour to distinguish them from scalar meta-variables) and the length of a
+coeffect vector is written as $\alen{\aclrd{\textbf{r}}}$.
 
@@
 The variable access rule (\emph{var}) annotates the corresponding variable as being accessed
-using $\cunit$. Note that, as in substructural systems, the free variable context contains
+using $\cunit$. As in substructural systems, the free variable context contains
 \emph{only} the accessed variable. Other variables can be introduced using explicit weakening.
@@
 variable. The corresponding splitting is performed on the coeffect vector, uniquely associating
-the annotation $\cclrd{s}$ with the bound variable $x$. This means that the typing rule removes
-non-determinism of the type inference present in flat coeffect systems.
+the annotation $\cclrd{s}$ with the bound variable $x$. This form of typing rule obviates the
+ambiguity in splitting of context demands present in the flat coeffect systems.
 
@@
 \paragraph{Structural rules.}
-The rules shown in Figure~\ref{fig:struct-types} (b) are not syntax-directed and
-allow different transformation of the free variable context. They correspond to the
-transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
+These are shown in Figure~\ref{fig:struct-types} (b). The three structural rules are not
+syntax-directed and allow different transformation of the free variable context. They correspond
+to the transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
 from substructural systems.
@@
 then unify them using substitution. The resulting variable is annotated with $\cpar$ and it
-is the only place in the structural coeffect system where context requirements are combined
+is the only place in the structural coeffect system where context demands are combined
 (semantically, this is where the available context is shared).
-
-\paragraph{Subcoeffecting.}
-Finally, Figure~\ref{fig:struct-types} (c) shows the subcoeffecting rule (\emph{sub}).
-As with the flat coeffect calculus, we include it mainly to guide the understanding of the
-system, but we treat it as an additional rule and do not discuss it in the subsequent theory.
-The rule would be needed in languages with conditionals and other language features. It is
-worth noting that allowing subcoeffecting on coeffect scalars belonging to individual variables
-is sufficient for structural systems.
 
@@
 of $\czip$ in lambda abstraction and, as discussed in Section~\ref{sec:struct-syntax}, has
-a cleaner equational theory.
-
-\paragraph{Flat and structural context.} In flat systems, lambda abstraction splits context
-requirements using $\czip$ and application combines them using $\cpar$. In the structural
-version, both of these are replaced with $\atimes$. The $\czip$ operation is not needed, but
-$\cpar$ is still used in the (\emph{contr}) rule.
+a more desirable equational theory. By contrast, the flat system allows certain interesting
+use cases that \emph{rely} on the flexibility of $\czip$ in lambda abstraction (such as implicit
+parameters), that cannot be expressed in the structural system.
+
+In flat systems, lambda abstraction splits context demands using $\czip$ and application
+combines them using $\cpar$. In the structural version, both of these are replaced with $\atimes$.
+The $\czip$ operation is not needed, but note the use of $\cpar$ in the (\emph{contr}) rule.
 
@@
 Secondly, the operator $\cpar$ is used when the semantics passes a given context to multiple
-sub-expressions. In flat systems, the context is shared in (\emph{app}) and (\emph{pair}), because
-the sub-expressions may share variables. In structural systems, the sharing is isolated into an
-explicit contraction rule.
-
-\paragraph{Let binding.}
-The other aspect that makes structural systems simpler is that they remove the need for
-separate let binding. As discussed in Section~\ref{sec:flat-exts-let}, flat calculi include
-let binding that gives a \emph{more precise} typing than combination of abstraction and
-application. Structural coeffect systems avoid this issue.
-
-\begin{remark}[Let binding]
-\label{thm:structural-letbinding}
-In a structural coeffect calculus, the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$ can be
-seen as a derived rule, \ie~its typing is equivalent to the typing of the expression
-$(\lambda x.e_2)~e_1$.
-
-\end{remark}
-\begin{proof}
-Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that
-in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$.
-%
-\begin{equation*}
-\inference
-  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
-    \inference
-      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
-      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
-  { \inference
-      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
-      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
-\end{equation*}
-
-\noindent
-The assumptions and conclusions match those of the (\emph{let}) rule.
-\end{proof}
+sub-expressions. In flat systems, the context is shared in (\emph{app}) and the additional
+(\emph{pair}) rule for constructing tuple values, because the sub-expressions may share variables.
+In structural systems, the sharing is isolated into an explicit contraction rule.
 
@@
 flat systems. Formally, any flat coeffect algebra can be turned into a structural coeffect scalar
-(by dropping the $\czip$ operator), but this does not always give us a meaningful system -- for
-example, implicit parameters could be attached to individual variables, but, to our knowledge, this
-has not been used in any practical system.
+(by dropping the $\czip$ operator). This is useful for liveness and dataflow, but it does not yield
+a practically useful system for the flat algebra coeffect for implicit parameters.
 
@@
 because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
-scalar. This is the case, for example, for the bounded variable use.
+scalar. This is the case, for example, for the system tracking bounded variable use
+(Example~\ref{ex:structural-ex-bll}).
 
@@
 \noindent
-For the two examples that have both flat and structural version, obtaining the structural coeffect
-algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The laws
-required by a structural coeffect algebra are the same as those required by the flat version
+These two examples have both flat and structural versions. For them, obtaining the structural
+coeffect algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The
+laws required by a structural coeffect algebra are the same as those required by the flat version
 and so the above definitions are both valid. Similar construction can be used for the
@@
 
-It is important to note that this gives us a systems with \emph{different} properties. The
-information are now tracked per-variable rather than for entire contexts. For dataflow,
+It is important to note that this gives us a systems with \emph{different} properties.
+Information is now tracked per-variable rather than for entire contexts. For dataflow,
 we also need to adapt the typing rule for the \kvd{prev} construct. Here, we write $\aclrd{+}$
@@
 structural nature of the type system, it only increments the required number of values for
-variables that are used in the expression $e$. Annotations of other variables can be left
-unchanged.
-
-Before looking at the semantics and equational properties of structural coeffect systems,
-we consider bounded variable use, which is an example of structural system that does not
-have a flat counterpart.
+variables that are actually used in the expression $e$, whereas in the flat coeffect system the rule
+incremented the annotation for the whole context. Thanks to the structural nature of the system,
+annotations of variables that do not appear in the expression $e$ can be left unchanged.
+
+Before looking at properties of structural coeffect systems,
+we consider a system for tracking bounded variable use, which is an example of structural system
+that does not have a flat counterpart.
 
 \begin{example}[Bounded variable reuse]
+\label{ex:structural-ex-bll}
 The structural coeffect algebra for tracking bounded variable use is given by
@@
 Similarly to the structural calculus for dataflow, the calculus for bounded variable reuse
-annotates each variable with an integer. However, the integer denotes how many times is the variable
+annotates each variable with an integer. However, the integer now denotes how many times is the variable
 \emph{accessed} rather than how many \emph{past values} are needed. The resulting type system is
@@
   {\coctx{\Gamma_2}{\aclrd{\textbf{r}}} \vdash \lambda x:\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
-\quad (\coctx{\Gamma_2}{\aclrd{\textbf{r}}}), \tau_1, \cclrd{s} = \ident{findvar}_x(\coctx{\Gamma_1}{\aclrd{\textbf{t}}})
+\quad (\coctx{\Gamma_2}{\aclrd{\textbf{r}}}), \cclrd{s} = \ident{findvar}_{x,\tau_1}(\coctx{\Gamma_1}{\aclrd{\textbf{t}}})
 \end{equation*}
@@
 \begin{array}{l}
-\ident{exch}_x(\coctx{\Gamma}{\aclrd{\textbf{t}}})~=~(\coctx{\Gamma_1,\Gamma_2}{\aclrd{\textbf{t}_1} \atimes \aclrd{\textbf{t}_2}}), \tau, \cclrd{s} \quad \textnormal{where} \\
+\ident{findvar}_{x,\tau}(\coctx{\Gamma}{\aclrd{\textbf{t}}})~=~(\coctx{\Gamma_1,\Gamma_2}{\aclrd{\textbf{t}_1} \atimes \aclrd{\textbf{t}_2}}), \cclrd{s} \quad \textnormal{where} \\
 \qquad \xlen{\Gamma_1} = \alen{\aclrd{\textbf{t}_1}}~\textnormal{and}~\xlen{\Gamma_2} = \alen{\aclrd{\textbf{t}_2}} \\[-0.25em]
@@
 \\[0.5em]
-\ident{findvar}_x(\coctx{\Gamma}{\aclrd{\textbf{t}}})~=~ (\coctx{\Gamma}{\aclrd{\textbf{t}}}), \tau, \czero \qquad (\textnormal{otherwise})
+\ident{findvar}_{x,\tau}(\coctx{\Gamma}{\aclrd{\textbf{t}}})~=~ (\coctx{\Gamma}{\aclrd{\textbf{t}}}), \czero \qquad (\textnormal{otherwise})
 \\[1em]
@@
 in the typing. This is in contrast with flat coeffect systems (most importantly, the one for
-implicit parameters), where lambda abstraction allowed arbitrary splitting of context requirements.
-In structural coeffect systems, the context requirements placed on the call site (attached to
+implicit parameters), where lambda abstraction allowed arbitrary splitting of context demands.
+In structural coeffect systems, the context demands placed on the call site (attached to
 the function type) are those of the bound variable.
@@
 However, the type system for structural coeffect calculus in Figure~\ref{fig:struct-types}
-introduces another kind of ambiguity due to the fact that non syntax-directed structural rules
+introduces another kind of ambiguity due to the fact that non-syntax-directed structural rules
 can be applied repeatedly and in arbitrary order. As with the semantics for flat coeffect calculus
-in Chapter~\ref{ch:semantics}, we define the semantics of the structural coeffect calculus over
+in Chapter~\ref{ch:semantics}, we define the semantics of the structural coeffect calculus relative to
 a \emph{typing derivation} and so the meaning of a program depends on the typing derivation chosen.
-In this section, we specify how to choose the desired \emph{unique} typing derivation.
+In this section, we specify how to choose the desired \emph{unique} typing derivation, following
+the example of flat coeffect calculi as discussed in Section~\ref{sec:flat-unique}.
 
@@
 In order to choose a unique typing derivation, we follow the example of substructural type systems
-\cite{substruct-attpl-intro} and introduce a syntax-directed version of the type system (also called
-\emph{algorithmic}). This replaces the non-syntax-directed rules for weakening, exchange and
-contraction with additional logic in places where contexts are combined. Given a typing derivation
-in the syntax-directed type system, we can then produce a matching typing derivation in the original
-type system.
+\cite{substruct-attpl-intro} and introduce a syntax-directed version of the type system.
+This replaces the non-syntax-directed rules for weakening, exchange and contraction with
+more complexity in the rules where contexts are combined  (\emph{app}) and variables removed (\emph{abs}).
+
+Given a typing derivation in the deterministic syntax-directed type system, we then choose a
+typing derivation in the original type system that uniquely specifies how to apply weakening,
+contraction and exchange. The algorithm given in Proposition~\ref{thm:struct-unique-transl}
+inserts only the necessary structural rules to rearrange variable context into the shape
+required by the assumptions.
 
@@
 \paragraph{Lambda abstraction.}
-In lambda abstraction (\emph{abs}) in Figure~\ref{fig:struct-types}, we assume that the bound
+In the lambda abstraction (\emph{abs}) rule in Figure~\ref{fig:struct-types}, we assume that the bound
 variable is the last variable of the context. In the syntax-directed system, we do not make the
-same assumption. Instead, we use an auxiliary relation $\ident{findvar}_x$ that takes a typing
+same assumption. Instead, we use an auxiliary function $\ident{findvar}_{x,\tau}$ that takes a typing
 context $\coctx{\Gamma}{\aclrd{\textbf{t}}}$ and returns a context with the variable $x$ removed
-together with its type and coeffect.
-
-The $\ident{findvar}_x$ relation is defined by two disjoint cases. If the variable $x$ is not
-present in the context, the context is returned as is and the returned coeffect is $\czero$.
-This case corresponds to the (\emph{weak}) structural rule. If the variable is not found in the
-context, the relation does not place any restrictions on the type of the variable and so $\tau$
-can be any type (this is the only reason why $\ident{findvar}$ is a relation rather than a function).
-If the variable $x$ appears in the context then $\ident{findvar}_x$ removes it together with its
-corresponding coeffect annotation and returns it as the result.
-
-\paragraph{Function application.} In the (\emph{app}) rule in Figure~\ref{fig:struct-types}, we
-assume that the variable contexts of the two sub-expressions can be merged. This requires that they
+together with the coeffect originally attached to the variable. The $\ident{findvar}_x$ function is
+defined by two disjoint cases. The case when the variable $x$ is not present in the context
+corresponds to the (\emph{weak}) structural rule.
+
+\paragraph{Function application.} The (\emph{app}) rule in Figure~\ref{fig:struct-types}
+assumes that the variable contexts of the two sub-expressions can be merged. This requires that they
 contain disjoint variables, which can be always obtained by exchange and contraction. In the
@@
 
-The \ident{mergevars} function returns context consisting of three parts. Parts $\Gamma_1$
+As with $\ident{findvar}_{x,\tau}$, the \ident{mergevars} function is fully deterministic.
+It returns context consisting of three parts. Parts $\Gamma_1$
 and $\Gamma_2$ represent variables that appear only in the first or the second context; part
@@
 $\cclrd{r_i} \;\cpar\; (\cclrd{t} \,\cseq\, \cclrd{s_j})$. The first coeffect corresponds to the
-context requirements in the sub-expression $e_1$ and the second coeffect corresponds to the
+context demands in the sub-expression $e_1$ and the second coeffect corresponds to the
 function argument $e_2$ sequentially composed with the coeffect $\cclrd{t}$ of the function
@@
 The syntax-directed type checking presented in the previous section gives a unique typing derivation
-that can be automatically turned into a valid typing derivation of the original type system
+that can be automatically turned into one of the valid typing derivations of the original type system
 presented in Figure~\ref{fig:struct-types}. This gives us a unique typing derivation for the
 structural coeffect calculus. As with the unique typing derivation for the flat coeffect system,
-this is used to give semantics of terms of the structural coeffect calculus. We also note that
-a well-typed program in the original type system has a typing derivation in the syntax-driven
-version.
+the chosen typing derivation is used to give semantics of terms of the structural coeffect calculus.
+We also note that a well-typed program in the original type system has a typing derivation in the
+syntax-driven version.
 
 As when discussing uniqueness of typing for flat coeffect systems (Section~\ref{sec:flat-unique}),
-we first give the inversion lemma (Lemma~\ref{thm:struct-invert}) and then prove the
-uniqueness of typing theorem (Theorem~\ref{thm:struct-unique}).
+we first give an inversion lemma (Lemma~\ref{thm:struct-invert}) and then prove
+uniqueness of typing (Theorem~\ref{thm:struct-unique}).
 
@@
 \noindent
-As noted earlier, unique typing derivations obtained using the type system given in
-Figure~\ref{fig:struct-syntax-types} can be automatically turned into typing derivations
+As noted earlier, unique typing derivations obtained using the syntax-directed type system
+given in Figure~\ref{fig:struct-syntax-types} can be automatically turned into typing derivations
 of the original (non-syntax-directed) structural coeffect type system in Figure~\ref{fig:struct-types}.
-The following remark provides the details.
-
-\begin{remark}[Admisibility of unique typing for implicit parameters]
-\label{thm:flat-impl-admis}
+Unlike in the flat coeffect system, this does not determine how context demands are split
+(as this is done deterministically to match the variable bindings), but it specifies how are
+the structural rules (weakening, exchange and contraction) applied.
+The following proposition provides the details.
+
+\begin{proposition}[Choosing a unique typing derivation]
+\label{thm:struct-unique-transl}
   If $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ (using the rules in
@@
   coeffect annotation is the same as in the original typing derivation.
-\end{remark}
+\end{proposition}
+
+% ==================================================================================================
+%
+%    #######
+%    #       #    # ##### ###### #    #  ####  #  ####  #    #  ####
+%    #        #  #    #   #      ##   # #      # #    # ##   # #
+%    #####     ##     #   #####  # #  #  ####  # #    # # #  #  ####
+%    #         ##     #   #      #  # #      # # #    # #  # #      #
+%    #        #  #    #   #      #   ## #    # # #    # #   ## #    #
+%    ####### #    #   #   ###### #    #  ####  #  ####  #    #  ####
+%
+% ==================================================================================================
+
+\section{Syntactic properties and extensions}
+\label{sec:structural-exts}
+
+When discussing the structural coeffect calculus in Section~\ref{sec:struct-calculus}, we
+considered a language with variables, constants, application and abstraction. This lets us focus
+on the key properties of the coeffects, but it neglects a number of practical concerns.
+In this section, we extend the language with let binding and subcoeffecting. This is useful
+in practice, but it also shows other interesting aspects of the theory. Additional extensions that
+make the coeffect language practically useful are given by the implementation in Chapter~\ref{ch:impl}.
+
+\subsection{Let binding}
+\label{sec:structural-exts-let}
+
+In the flat coeffect calculus, we included a special typing rule for let binding. As discussed in
+Section~\ref{sec:flat-exts-let}, this provides a more precise typing than the rule derived from
+abstraction and application, because it removes the ambiguity introduced by abstraction. For the
+structural coeffect system, the typing rule for let binding can be treated as a derived rule.
+The following shows the structural typing for \kvd{let}:
+
+\begin{equation*}
+\tyrule{let}
+  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
+    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
+  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
+\end{equation*}
+%
+Thanks to the structural nature of the calculus, the coeffect $\cclrd{t}$ that is associated with
+the variable $x$ is uniquely determined (as in function abstraction). It is then sequentially
+composed with the coeffects attached to the variables that actually appear in the sub-expression
+$e_1$.
+
+\begin{proposition}[Let binding]
+\label{thm:structural-letbinding}
+In a structural coeffect calculus, the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$ can be
+seen as a derived rule, \ie~its typing is equivalent to the typing of the expression
+$(\lambda x.e_2)~e_1$.
+\end{proposition}
+\begin{proof}
+Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that
+in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$.
+%
+\begin{equation*}
+\inference
+  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
+    \inference
+      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
+      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
+  { \inference
+      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
+      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
+\end{equation*}
+
+\noindent
+The assumptions and conclusions match those of the (\emph{let}) rule.
+\end{proof}
+
+\subsection{Subcoeffecting}
+\label{sec:structural-exts-sub}
+
+When discussing the flat coeffect calculus in Section~\ref{sec:flat-calculus-algebra}, we noted
+that the $\cleq$ operation for \emph{flat coeffect algebra} can be defined in terms of $\cpar$
+as follows:
+%
+\begin{equation*}
+\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
+\end{equation*}
+%
+This is not the case for the \emph{structural coeffect scalar} structure. For example, in the
+calculus for tracking bounded reuse, the $\cpar$ operator is defined as $+$ (on integers) and
+$\cleq$ is just $\leq$ and so the above equivalence does not hold. For this reason, we included
+$\cleq$ as an explicit part of both of the structures.
+
+The subcoeffecting rule is not syntax directed and we did not include it in the core calculus
+in order to keep the discussion about choosing unique derivation in Section~\ref{sec:struct-unique}
+focused on the key problem -- structural rules. The subcoeffecting rule for structural coeffect
+calculus looks as follows:
+%
+\begin{equation*}
+\tyrule{sub}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s'}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+  {\coctx{\Gamma}{\aclrd{\textbf{r}} \atimes \alift{\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}
+\quad
+(\cclrd{s'} \cleq \cclrd{s})
+\end{equation*}
+%
+The sub-coeffecting is applied on individual variables rather than on the whole context, but it
+could be easily extended to a relation on vectors $\aclrd{\leq}$. Subtyping on functions can be
+defined in exactly the same way as for the flat coeffect calculus (Section~\ref{sec:flat-exts-sub}),
+because functions are annotated with a (single) coeffect scalar. It is worth noting that
+subcoeffecting is needed in Lemma~\ref{thm:structural-substitution} (discussed in the next
+section) when performing a substitution for a variable in an expression that does not contain the
+substituted variable.
 
@@
 
-The properties of the structural coeffect algebra guarantee that certain equational properties
-hold in all instances of the structural coeffect calculus. In this section, we look at these common
-properties that we get ``for free'' from the structural coeffect calculus. We start the discussion
-by briefly considering the key aspects that make the equational theory of flat and structural
-coeffects different.
+The properties of the structural coeffect algebra, together with two additional weak conditions,
+guarantee that certain equational properties on terms hold in all instances of the structural
+coeffect calculus that we consider in this thesis. In this section, we look at these common
+properties. In Section~\ref{sec:struct-syntax-intro}, we first briefly compare equational theory
+for flat coeffects (Section~\ref{sec:flat-syntax}) and structural coeffects (Section~\ref{sec:struct-syntactic-thms}).
 
@@
 
-When discussing syntactic reductions for the flat calculus in Section~\ref{sec:flat-syntax},
+When discussing syntactic reductions for the flat calculus,
 we noted that call-by-name reduction does not, in general, preserve typing  for all
@@
 
-\paragraph{Substitution for flat coeffects.}
+\paragraph{Substitution for flat coeffects (recap).}
 The less obvious (\emph{top-pointed}) variant of the substitution lemma for flat coeffects
@@
 to coincide. This enables substitution to preserve the type of expressions, because all
-additional requirements arising as the result of the substitution can be associated with the
-declaration context. For example, consider the following example where implicit parameter
+additional demands arising as the result of the substitution can be associated with the
+declaration context. For example, consider the following example where Haskell-style implicit parameter
 $\ident{?offset}$ is substituted for the variable $y$:
@@
 \paragraph{Substitution for structural coeffects.}
-In contrast, the substitution lemma (Lemma~\ref{thm:structural-substitution}, shown later on
+By contrast, the substitution lemma (Lemma~\ref{thm:structural-substitution},
 page~\pageref{thm:structural-substitution}) for structural coeffects can be proven because
@@
   &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{before}) \\
-\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}}
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{before}) \\
+\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\aclrd{\ast}\alift{\cclrd{1},\cclrd{1}}}
   &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{after}) \\
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{after}) \\
 \coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}}
   &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
-   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textit{final})
+   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & (\textit{equivalently})
 \end{array}
@@
 expression $w+z$ with free variables $w,z$ annotated with $\alift{\cclrd{1}, \cclrd{1}}$.
-The substitution applies the operation $\aseq$ (modelling sequential composition) to
-the annotation of the new context -- in the above example
-$\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}=\alift{\cclrd{2}, \cclrd{2}}$.
+The substitution applies the operation $\aclrd{\ast}$ (which stands for the sequential
+composition $\aseq$ from the structural coeffect algebra) to the annotation of the new context --
+in the above example the coeffect $\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}$ (\emph{after}) is
+equivalent to the coeffect $\alift{\cclrd{2}, \cclrd{2}}$ (\emph{equivalently}).
 
@@
 such as the bunched typing system \cite{substruct-bunched}. To express the substitution
-lemma later in this section, we first define the notion of a \emph{context with holes}:
+lemma later in this section, we follow the example of bunched type system and define the
+notion of a \emph{context with holes}. A context with holes is a context such as
+$\coctx{x_1\!:\!\tau_1, \ldots, x_k\!:\!\tau_k}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_k}}}$,
+where some of the variable typings $x_i\!:\!\tau_i$ and corresponding coeffects $\cclrd{r_i}$
+are replaced by \emph{holes} written as $\coctx{-}{-}$.
 
 \begin{definition}[Context with holes]
-A \emph{context with holes} is a context such as $\coctx{x_1\!:\!\tau_1, \ldots, x_k\!:\!\tau_k}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_k}}}$,
-where some of the variable typings $x_i\!:\!\tau_i$ and corresponding coeffects $\cclrd{r_i}$
-are replaced by \emph{holes} written as $-$.
-
-\begin{equation*}
-~\Delta[\,\coctx{-}{-}\,]_n = \Delta[\,\underbrace{\coctx{-}{-} \,|\, \ldots \,|\, \coctx{-}{-}}_{ n-\textit{times} } \,]
-\end{equation*}
+We write $\Delta[\,\coctx{-}{-}\,]_n$ for a \emph{context with $n$ holes} (in addition to some
+number of variables). A context with holes is defined inductively over the number of holes:
+%
 \begin{equation*}
@@
 $\coctx{\Delta}{\textbf{s}}[\,\coctx{\Gamma_1}{\aclrd{\mathbf{r_1}}} \,|\, \ldots \,|\, \coctx{\Gamma_n}{\aclrd{\mathbf{r_n}}} \,]$,
-which replaces the holes by the specified variables and corresponding coeffect annotations, is defined as:
+replaces holes by the specified variables and corresponding coeffect annotations and is defined as:
 %
@@
 
-Substitution can be seen as a form of sequencing, \ie~the term $\subst{e_1}{x}{e_2}$ should have
-the same type as the term $\kvd{let}~x=e_2~\kvd{in}~e_1$. In other words, we first need to obtain the
-value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with
-requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
-are annotated with $\cclrd{s} \,\aseq\, \aclrd{\textbf{t}}$, using the (scalar-vector extension)
-of the sequential composition operator $\cseq$.
-
-An operational reading of the resulting coeffect annotation is that, to evaluate each of the variables
-annotated with coeffects in the vector $\aclrd{\textbf{t}}$, we first need to evaluate the substituted
-expression with coeffects $\cclrd{s}$, followed by the rest of the expression with coeffects specified
-by $\aclrd{\textbf{t}}$.
-
 \begin{lemma}[Multi-nary substitution]
 \label{thm:structural-substitution}
-Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
+In a structural coeffect calculus with a \emph{structural coeffect scalar} such that
+$\cclrd{r}\,\cleq\,\cclrd{r'} \, \Rightarrow \, \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s})$
+and also $\czero \;\cleq\; (\czero\,\cseq\,\cclrd{r})$,
+given an expression with multiple holes that are filled by variables $x_1\!:\!\tau_1,\ldots,x_n\!:\!\tau_n$ with coeffects $\cclrd{s_1},\ldots,\cclrd{s_n}$:
 %
@@
 %
-then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
+substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
 where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
@@
 
+The Lemma~\ref{thm:structural-substitution} has two additional requirements on the structural coeffect
+scalar that are similar to those of the substitution lemma for bottom-pointed flat coeffect systems
+(Lemma~\ref{thm:cbn-substitution-bot}). Those two conditions are satisfied for all our examples
+(it would be reasonable to require them for \emph{all} structural coeffect scalars, but we prefer
+to keep the original definition more general). The two requirements are needed in the proof
+for substitution for subcoeffecting rule (discussed in Section~\ref{sec:structural-exts-sub}) and
+for weakening, respectively.
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Reduction and expansion}
+\label{sec:struct-syntactic-thms}
 
@@
 $\eta$-expansion preserves the types. Both of the properties hold for a system with any
-structural coeffect algebra.
+structural coeffect algebra that satisfies the additional weak requirements given in Lemma~\ref{thm:structural-substitution}.
 
@@
 \begin{theorem}[Type preservation]
-In a structural coeffect system with a structural coeffect algebra formed by
-$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
-if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$
+In a structural coeffect system with subcoeffecting (Section~\ref{sec:structural-exts-sub})
+and a structural coeffect algebra formed by
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$ that
+satisfies the requirements of Lemma~\ref{thm:structural-substitution},
+it holds that if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$
 and $e \rightarrow_{\beta} e'$ using the full $\beta$-reduction then
@@
 \begin{proof}
-Consider the typing derivation for the redex $(\lambda x.e_r)\,e_s$ before the reduction
-(note that this is similar to the typing of let binding discussed in Remark~\ref{thm:structural-letbinding},
-but we do not swap the two parts of the free-variable context):
+Consider the typing derivation for the redex $(\lambda x.e_r)\,e_s$:
 %
@@
 \noindent
-Because of the vector (free monoid) structure of coeffect annotations $\aclrd{\textbf{r}}$,
+Because of the vector structure of coeffect annotations $\aclrd{\textbf{r}}$,
 $\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$, these are uniquely associated with $\Gamma_r$,
 $\Gamma_s$, and $x$ respectively. Therefore, substituting $e_s$ (which has coeffects
-$\aclrd{\textbf{s}}$) for $x$ introduces the context requirements specified by $\aclrd{\textbf{s}}$
-which are composed with the requirements $\cclrd{t}$ associated with $x$, \ie~the variable being
+$\aclrd{\textbf{s}}$) for $x$ introduces the context demands specified by $\aclrd{\textbf{s}}$
+which are composed with the demands $\cclrd{t}$ associated with $x$, \ie~the variable being
 substituted.
@@
 effects of $e$ are hidden by the purity of $\lambda$-abstraction. In the coeffect calculus,
-the (\emph{abs}) rule allows immediate contextual requirements of $e$ to ``float outside''
+the (\emph{abs}) rule allows immediate contextual demands of $e$ to ``float outside''
 of the enclosing $\lambda$. Furthermore, the free monoid nature of $\atimes$ in structural
-coeffect systems allows the exact immediate requirements of $\lambda x . e x$ to match
+coeffect systems allows the exact immediate demands of $\lambda x . e x$ to match
 those of $e$.
@@
 
+~
+
 \section{Summary}

### 2016-05-29 text/Chapter07.tex
Addressing feedback, fix demands (Chapter 6)

@@
 In a context-aware language for dataflow (and in language with liveness tracking), the inherent
-ambiguity of the (\emph{abs}) rule is resolved by placing the context requirements of the body on
+ambiguity of the (\emph{abs}) rule is resolved by placing the context demands of the body on
 both the declaration site and the call site. In Section~\ref{sec:flat-unique}, this was defined by
@@
 \paragraph{Flat and structural dataflow.}
-In flat dataflow, the context requirements of the body is required from both the declaration site
-and from the call site. In structural dataflow, the context requirements are tracked separately
+In flat dataflow, the context demands of the body is required from both the declaration site
+and from the call site. In structural dataflow, the context demands are tracked separately
 for each variable, which provides a more precise type. Consider the following two examples (the

### 2016-05-29 text/Chapter08.tex
Addressing feedback, fix demands (Chapter 6)

@@
 
-\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
-  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
+\item In the flat coeffect calculus, we distinguish between \emph{splitting} context demands
+  and \emph{merging} of context demands ($\cpar$ and $\czip$, respectively). In the structural
   coeffect calculus, the operations (which model appending and splitting vectors) are invertible
@@
 other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
-between splitting and merging of context requirements (using the notation $\apar$ and $\azip$,
+between splitting and merging of context demands (using the notation $\apar$ and $\azip$,
 respectively). For structural coeffect calculi, the two operators coincide, but for flat
@@
 
-The $\cpar$ operation models combining of context requirements arising from multiple parts of a
+The $\cpar$ operation models combining of context demands arising from multiple parts of a
 program. The meaning depends on the coeffect container. The operation can either combine requirements
@@
 a morphism that merges the capabilities provided by two contexts (in the syntactic
-reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
+reading, splits the context demands), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
 is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
-context into two parts (in the syntactic reading, merges their context requirements).
+context into two parts (in the syntactic reading, merges their context demands).
 Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
@@
 system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging
-of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
-in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.
+of the unified coeffect algebra. As already mentioned, we use splitting of context demands $\azip$
+in rule \emph{assumptions} and merging of context demands $\apar$ in rule \emph{conclusions}.
 
@@
 which could be also written as $(\aclrd{r}\,\azip\,\aclrd{s}) \;\aclrd{\leq}\; (\aclrd{r}\,\apar\,\aclrd{s})$,
-denotes that splitting context requirements and then re-combining them preserves all the requirements
+denotes that splitting context demands and then re-combining them preserves all the requirements
 from the assumptions. The system may be imprecise and conclusions can overapproximate assumptions, but
@@
   into a computation that requires just $\cclrd{s}$ and contains $\beta$ (a part of the
-  context requirements is eliminated by the function).
+  context demands is eliminated by the function).
 \end{itemize}
@@
 
-The other aspect of practical implementation of coeffects is that of tracking the context requirements
+The other aspect of practical implementation of coeffects is that of tracking the context demands
 (coeffect annotations) in the type system. To achieve this (without resorting to a single-purpose

### 2016-05-29 text/backup/Chapter01.tex
Addressing feedback, fix demands (Chapter 6)

@@
 context requirement of the function body is a set of resources $\{\ident{Clock}, \ident{News}\}$.
-The context requirements are split between the \emph{declaration site} and \emph{call site}.
+The context demands are split between the \emph{declaration site} and \emph{call site}.
 However, there are multiple possible splittings. The splitting $\{\ident{News}\} \cup \{\ident{Clock}\}$
@@
 version. As we associate information with individual variables, lambda abstraction creates a function
-that requires the context requirements associated with the variable that is being abstracted over.
+that requires the context demands associated with the variable that is being abstracted over.
 
 For example, if we wrapped the earlier example above in a function taking \ident{a} (and using 
-\ident{b} from the declaration site) then the function would have context requirements $5$ -- that
+\ident{b} from the declaration site) then the function would have context demands $5$ -- that
 is the number associated with the variable \ident{a}.

### 2016-05-30 text/Chapter02.tex
Edits and tweaks (Chapter 8 and 9 mostly)

@@
 systems (Section~\ref{sec:path-eff}) and we extend the duality to category theory, looking at
-\emph{comonads} (Section~\ref{sec:path-sem}). Finally we consider logically inspired type systems
-that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).
+\emph{comonads} (Section~\ref{sec:path-sem}). We also consider type systems inspired by linear
+and bunched logic, which are closely related to our structural coeffects (Section~\ref{sec:path-logic}).
+Finally, we also consider practical motivations for context-aware programming (Section~\ref{sec:path-cop}).
 
@@
 a way that naturally leads to the coeffect systems developed later in the thesis. For this reason,
-we are not always faithful to the referenced work -- sometimes we focus on aspects that the
-authors consider unimportant or present the work differently than originally intended. The reason
-is to fulfill the second goal of the chapter. When we do so, this is explicitly said in the text.
+we are not always faithful to the referenced work. We present the work through the coeffect view
+and so we sometimes focus on aspects that authors consider unimportant or we present the work
+differently than originally intended. When we do so, this is explicitly stated in the text.
 
@@
 
-\section{Through static and dynamic binding}
+\section{Coeffects via static and dynamic binding}
 \label{sec:path-binding}
@@
 We discuss ordinary variable binding and Haskell's implicit parameters \cite{app-implicit-parameters},
-which provides an interesting point in the design space. We then discuss ambiguity that is inherent
-in variable binding and, more generally, context-aware programming languages, and a way of
+which provide an interesting point in the design space. Implicit parameters give an example of
+an ambiguity that arises more generally in context-aware programming, as well as one way of
 resolving it through \emph{type-directed semantics}.
+
+For a more general context-aware programming example, consider a program
+running on a laptop that accesses an implicit parameter representing a printer. When printing, the
+text may appear on my home printer (corresponding to static binding), or it may appear on the
+nearest printer based on my physical location (corresponding to dynamic binding).
 
@@
 only evaluated if the environment provides values for variables $x$ and $y$. A variable
-\emph{requirements} can be satisfied in two standard ways that are captured by \emph{dynamic} and
-\emph{static} (or lexical) variable binding. Consider the following simple program:
-
+\emph{requirements} can be satisfied in two standard ways that are characterized as \emph{dynamic}
+and \emph{static} (or lexical) variable binding. Consider the following simple program:
+%
 \begin{equation*}
@@
 example, the language could use static binding by default, but resort to dynamic binding if a
-variable is not available in the lexical scope. One such system is discussed in the next section.
-
-Languages with static scoping generally check that all variable requirements are satisfied.
-Attempt to access a variable that is not available in scope will give a compile error. In contrast,
+variable is not available in the lexical scope. One such system is Haskell's implicit parameters
+discussed in the next section.
+
+Languages with static scoping resolve variable bindings at compile-time. In contrast,
 languages with dynamic variable binding typically do not perform compile-time checks -- if a
 program attempts to access a variable that is not available in the environment, a runtime error
-occurs. This is possibly artifact of their implementation -- implementing static binding
-\emph{without} checking would be cumbersome and implementing checking for dynamic binding is
-requires a more sophisticated type system (Section~\ref{sec:path-effects-coeff}).
+occurs. This is possibly an artifact of their implementation -- implementing static binding
+\emph{without} checking would be cumbersome and implementing static checking for dynamic binding is
+requires a more sophisticated type system than traditional (Section~\ref{sec:path-effects-coeff}).
 
@@
 
-The following two examples are variations on the one discussed in the Section~\ref{sec:path-binding-var}.
-We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
-implicit parameter is defined both at declaration-site and at the call site. On the right, the
-implicit parameter is available only at the call site:
+The following two examples are variations on the one discussed in Section~\ref{sec:path-binding-var},
+obtained by replacing a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
+implicit parameter is declared both in the static scope and in the dynamic scope. On the right, the
+implicit parameter is available only in the dynamic scope:
 
@@
 %
-The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
-dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
+The binding rules for Haskell's implicit parameters can be summarized as ``\emph{static binding when possible,
+dynamic binding when needed}''. If an implicit parameter is available in static scope, then the value
 is statically bound and the context requirement is satisfied using the declaration-site context.
 Otherwise, the context requirement is delayed and has to be satisfied at the call site.
-In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
-delayed context demands and the expression evaluates to $10$. In the example on the right,
-the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
+In the example on the left, \ident{?x} is bound to $10$ and so the function \ident{f} has no
+delayed context demands and thus the expression evaluates to $10$. On the right,
+the context demand for \ident{?x} is \emph{delayed} and is satisfied via dynamic binding when calling
 the function. The expression evaluates to $5$.
@@
 In Haskell, the type system checks that bindings for all required implicit parameters are
-available. The type of the function \ident{f} on the left is $\kvd{int}\rightarrow\kvd{int}$, while
-the type of the \ident{f} function on the right is $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
+available and so no runtime errors can occur. The type of the function \ident{f} on the left is
+$\kvd{int}\rightarrow\kvd{int}$, while the type of the \ident{f} function on the right is
+$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
 The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
@@
 The three different binding mechanisms discussed so far can be seen as different ways of splitting
-context demands of a particular kind. Dynamic and static binding represent the opposite ends
-of the design space and Haskell's implicit parameters are an interesting point inside the wider space.
-
-In this thesis, we consider different notions of context, using implicit parameters as just one
-of several motivating examples. However, implicit parameters are a valuable example, because they
+context demands of a particular kind into static and dynamic parts. Dynamic and static binding
+represent the opposite ends of the design spectrum and Haskell's implicit parameters are an
+interesting point inside the wider spectrum.
+
+In this thesis, we consider various notions of context, using implicit parameters as one
+of several motivating examples. Implicit parameters are a particularly valuable example, because they
 clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
 a function can be satisfied using the context available at declaration-site or using the context
-available at the call site.
-
-We also aim to find a description of context-aware languages that does not make ad-hoc decisions
+available at the call site. Recall our earlier printer example -- a language that
+provides access to resources in the context needs to provide enough flexibility in handling such
+ambiguities, be it implicit parameter values available in scope or printers available in the
+physical environment.
+
+We aim to find a description of context-aware languages that does not make ad-hoc decisions
 about how context demands are split between the declaration-site and the call site. While
-Haskell's solution for implicit parameters might be the most reasonable one, different notions
-of context might require different domain-specific choices and the general framework of
-context-aware programming should make that possible.
+Haskell's solution for implicit parameters might be the most reasonable one for that particular
+case, this thesis argues that other notions of context require different domain-specific choices
+and the general framework of context-aware programming should make that possible.
 
@@
 
-In many practical programming languages, the value and semantics of an expression depends on the
-type derivation. Typically, the choice is hidden behind a mechanism that selects one preferred type
-derivation.
+In many practical programming languages, the value and semantics of an expression depends on its
+type derivation. In order to assign unique semantics to an expression, the choice is typically
+hidden behind a mechanism that selects one preferred type derivation.
 
 This mechanism serves as an inspiration for our approach to resolving ambiguity inherent in
-context-aware programs. This section shows a brief example using the F\# language \cite{app-fsharp},
+context-aware programs. This section discusses a brief example using the F\# language \cite{app-fsharp},
 before revisiting the implicit parameters example.
-
-~
 
@@
 type of the lambda function becomes $\ident{string}\rightarrow\ident{int}$. In the second case,
-the list contains two arrays (empty array and array containing one hundred 0 values) and so
+the list contains two arrays (an empty array and an array containing one hundred zero values) and so
 the type of the lambda function is $\ident{int}\lbrack\,\rbrack\rightarrow\ident{int}$.
@@
 %
-\begin{itemize}
+\begin{enumerate}
 \item The semantics of the function $(\lambda\ident{x}\rightarrow\ident{x.Length})$
@@
   based on the context in which the expression appears, but other options are possible
-  (in some cases, there is a \emph{default} resolution; in some cases, the compiler requires
+  (in some cases, there is a \emph{default} resolution; often, the compiler requires an
   explicit typing annotation).
-\end{itemize}
+
+\item An expression without a type derivation does not have semantics. For example, given
+  $\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})$, the F\# compiler fails to
+  infer a type; the expression is not well-typed and does not have a semantics.\footnote{Alternatively,
+  the compiler could choose default typing among multiple options. The F\# compiler does this
+  for the $+$ operator, which can be used on \ident{float} and \ident{int} types, but the compiler
+  chooses \ident{int} as the default.}
+\end{enumerate}
 %
 The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
-possible typing derivations and its semantics varies depending on the type. If the lexical
+possible typing derivations and its semantics varies depending on its type. If the lexical
 scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
@@
 type. In F\#, the compiler determines a unique typing derivation based on other parts of the program
-(or fails, if type is not uniquely determined). In our languages, we also determine a unique typing
-derivation. However, rather than relying on type information from other parts of the program, we
-explicitly define algorithm that chooses the preferred unique derivation.
+(if type is not uniquely determined, it either chooses a default or fails). In our languages, we
+also determine a unique typing derivation. However, rather than relying on type information from
+other parts of the program, we explicitly define an algorithm that chooses the preferred unique
+derivation (Section~Section~\ref{sec:flat-unique} and Section~\ref{sec:struct-unique}).
 
@@
 independently -- the semantics of context-aware programs and the domain-specific way of resolving
-how context demands are satisfied. In our treatment of implicit parameters, we consider
+ambiguities in how context demands are satisfied. In our treatment of implicit parameters, we consider
 multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
@@
 
-\section{Through type and effect systems}
+\section{Coeffects via type and effect systems}
 \label{sec:path-eff}
@@
 \subsection{Simple effect system.}
-The structure of a simple effect system\footnote{Most work on effect systems uses $\sigma$ for
+The structure of a simple effect system\footnote{Most work on effect systems uses $\sigma$ or $F$ for
 effect annotations. We use letters $\cclrd{r}, \cclrd{s}, \cclrd{t}$ and also distinguish effect
@@
 
-\section{Through language semantics}
+\section{Coeffects via language semantics}
 \label{sec:path-sem}
@@
 computation} and provides a way of constructing and composing effectful computations:
+
+~
 
@@
 access and the application of a primitive function $f$ is interpreted as follows:
-
+%
 \begin{equation*}
@@
 The types consist of the primitive type, function type and a type constructor that
-represents monadic computations. This means that the expressions in the language can create both
+represents monadic computations. Thus the expressions in the language can create both
 effect-free values, such as $\tau$ and computations $\mtyp{}{\tau}$. The additional expression
-$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no actual effects) from a
-value and $\kvd{let}_\mtyp{}{}$ is used to sequence effectful computations. In the semantics,
+$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no effects) from a
+value and $\kvd{let}_\mtyp{}{}$ is sequences effectful computations. In the semantics,
 monads are not needed to interpret variable access and application, they are only used in the
@@
 Moggi). Before looking at the details, we use the following definition of comonad:
+
+~
 
@@
 which is defined as follows:
-%
+
 \begin{equation*}
@@
 
-\section{Through substructural and bunched logics}
+\section{Coeffects via substructural and bunched logics}
 \label{sec:path-logic}
@@
 but it is also interesting on its own.
+
+%---------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\tyrule{exchange}
+  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_2 \vdash e : \gamma}
+  {\Gamma, y\!:\!\tau_2, x\!:\!\tau_1 \vdash e : \gamma}
+\end{equation*}
+\begin{equation*}
+\tyrule{weakening}
+  {\Gamma, \Delta \vdash e : \gamma}
+  {\Gamma, x\!:\!\tau, \Delta \vdash e : \gamma}
+\end{equation*}
+\begin{equation*}
+\tyrule{contraction}
+  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_1, \Delta \vdash e : \tau_2}
+  {\Gamma, x\!:\!\tau_1, \Delta \vdash \subst{e}{y}{x} : \tau_2}
+\end{equation*}
+
+\figcaption{Exchange, weakening and contraction typing rules}
+\label{fig:substructural-rules}
+\end{figure}
 
@@
 
-\begin{figure}
-\begin{equation*}
-\tyrule{exchange}
-  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_2 \vdash e : \gamma}
-  {\Gamma, y\!:\!\tau_2, x\!:\!\tau_1 \vdash e : \gamma}
-\end{equation*}
-\begin{equation*}
-\tyrule{weakening}
-  {\Gamma, \Delta \vdash e : \gamma}
-  {\Gamma, x\!:\!\tau, \Delta \vdash e : \gamma}
-\end{equation*}
-\begin{equation*}
-\tyrule{contraction}
-  {\Gamma, x\!:\!\tau_1, y\!:\!\tau_1, \Delta \vdash e : \tau_2}
-  {\Gamma, x\!:\!\tau_1, \Delta \vdash \subst{e}{y}{x} : \tau_2}
-\end{equation*}
-
-\figcaption{Exchange, weakening and contraction typing rules}
-\label{fig:substructural-rules}
-\end{figure}
-
 %---------------------------------------------------------------------------------------------------
@@
 \section{Context oriented programming}
+\label{sec:path-cop}
 

### 2016-05-30 text/Chapter04.tex
Edits and tweaks (Chapter 8 and 9 mostly)

@@
 In order to capture both systems using a single calculus, we need a way of rewriting the
-(\emph{df1}) rule such that the annotation in the assumption is in the form $\cclrd{n \disc m}$
-for some operation $\cclrd{\disc}$. For the dataflow system, this can be achieved by using
+(\emph{df1}) rule such that the annotation in the assumption is in the form $\cclrd{n \circ m}$
+for some operation $\cclrd{\circ}$. For the dataflow system, this can be achieved by using
 the $\cclrd{\emph{min}}$ function:

### 2016-05-30 text/Chapter07.tex
Edits and tweaks (Chapter 8 and 9 mostly)

@@
 written using \kvd{let}), this requirement is duplicated. The structural system keeps information
-per-variable and so the resulting type reflects the fact that only the variable \idnet{y} appears
+per-variable and so the resulting type reflects the fact that only the variable \ident{y} appears
 inside \kvd{prev}.
@@
 (which is a more recent incarnation of the idea developed by the author \cite{app-fsharp-webtools}).
-We choose F\#, but similar tools exist for other functional languages such as OCaml \cite{other-js-ocaml}.
+We choose F\#, but similar tools exist for other functional languages such as OCaml \cite{app-ocaml-js}.
 It is worth noting that FunScript is implemented as a \emph{library} rather than as a compiler

### 2016-05-30 text/Chapter08.tex
Edits and tweaks (Chapter 8 and 9 mostly)

@@
 The prototype implementation discussed in the previous chapter directly follows the two
-coeffect calculi presented in this thesis. This links together all aspects of the coeffect
+coeffect calculi presented in this thesis. It links together all aspects of the coeffect
 theory (type system, semantics, safety proofs) and it provides practical support for the
-presented theoretical work. However, the theory of coeffects could be embedded in practical
-programming languages in a various ways and the theory could be further developed in a number
-of different directions.
+presented theoretical work. In this chapter, we consider alternative ways of embedding the
+theory of coeffects in practical programming languages and we also outline further directions
+in which the theory can be further developed.
 
@@
 First, rather than discussing flat and structural coeffects, we can treat them in a single
-unified calculus tracking both kinds of context dependence. Second, rather than using ``language
-semantics'' style (Section~\ref{sec:path-sem-langs}), we could use a ``meta-language'' style
-and extend the host language with explicit constructs for working with coeffects. Third,
-rather than implementing a prototype coeffect language, we could embed coeffects into an
-existing language using techniques inspired by Haskell's ``do'' notation.
+unified calculus tracking both kinds of context dependence (Section~\ref{sec:further-unified}).
+Second, rather than using ``language semantics'' style (Section~\ref{sec:path-sem-langs}), we
+could use a ``meta-language'' style and extend the host language with explicit constructs for
+working with coeffects (Section~\ref{sec:further-meta}). Third, rather than implementing a
+prototype coeffect language, we could embed coeffects into an existing language using techniques
+inspired by Haskell's ``do'' notation (Section~\ref{sec:further-further}).
 
@@
   \emph{unified coeffect calculus} (Section~\ref{sec:further-unified}), which generalizes flat
-  and structural systems and can be instantiated to track both flat and structural properties.
+  and structural systems and which can be instantiated to track both flat and structural properties.
 
@@
 \noindent
-The unified coeffect calculus presented in the first part of the chapter is a novel result that
-completes the theory developed in this thesis. The next two sections present important future and
-related work, respectively.
+The unified coeffect calculus presented in Section~\ref{sec:further-unified} is a novel result that
+completes the theory developed in this thesis and has been discussed in detail in a separate
+joint publication by the author \cite{coeffects-icfp14}. We present it as further work here,
+because we do not supplement the unified calculus with the same operational treatment as the flat
+and structural coeffect calculi. Finally, the Sections~\ref{sec:further-meta} presents
+important related work and Section~\ref{sec:further-further} presents important further work.
 
@@
 \item The structural coeffect calculus contains explicit rules for context manipulation
-  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
-  explicitly, but they can be derived thanks to the fact that variables in the context can be
-  freely reordered.
+  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not
+  explicit, but they can be derived thanks to the fact that variables in the context can be
+  freely reordered. The (\emph{var}) rule differs to permit implicit exchange and weakening
+  in the flat system (by ignoring other variables in the context) and (\emph{app}) implicitly
+  permits contraction (by duplicating the context passed to sub-expressions).
 
@@
 
-\item In the flat coeffect calculus, we distinguish between \emph{splitting} context demands
-  and \emph{merging} of context demands ($\cpar$ and $\czip$, respectively). In the structural
-  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
-  and so the structural coeffect algebra uses the same tensor product $\atimes$. However, the
-  $\cpar$ operation is still needed in the contraction rule.
+\item In the flat coeffect calculus (Figure~\ref{fig:flat-types}), we distinguish between
+  \emph{splitting} context demands and \emph{merging} of context demands ($\cpar$ and $\czip$,
+  respectively). In the structural coeffect calculus, the operations (which model appending and
+  splitting vectors) are invertible and so the structural coeffect algebra uses the same tensor
+  product $\atimes$. In essence $\cpar$ and $\czip$ represent under- and over-approximation of
+  $\atimes$. In the structural coeffect system, the $\cpar$ operation is still needed
+  in the contraction rule.
 \end{itemize}
@@
 the structure of coeffect annotations using the notion of a ``container'' which can be specialized
-to obtain a single annotation or a vector of annotations. This could potentially be used to capture
-other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
-between splitting and merging of context demands (using the notation $\apar$ and $\azip$,
-respectively). For structural coeffect calculi, the two operators coincide, but for flat
+to obtain a single annotation or a vector of annotations. This could potentially be alternatively
+specialized to capture other structures such as trees in bunched typing \cite{substruct-bunched}.
+Finally, we keep the $\cpar$ operation (needed in the contraction rule), but we distinguish
+between splitting and merging of context demands (using the notation $\apar$
+and $\azip$, respectively). For structural coeffect calculi, the two operators coincide, but for flat
 coeffect calculi, they differ and provide the needed flexibility.
@@
 types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
-lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise
-as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
+lists)\footnote{Shapes can be intuitively thought of as sizes, but this is not precise
+as some containers have multiple shapes of the same length, \eg~for perfectly ballanced, left- and right-leaning trees.}.
 For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
@@
 A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
-Here, $S$ is a set of shapes and $P$ is a shape-indexed family $P : S \rightarrow \ident{Set}$
-of positions.
+Here, $S$ is a set of shapes and $P$ is a shape-indexed family of sets such that
+for each shape $s\in S$, $P_s$ (also written as $P(s)$) is a set of positions for shape $s$.
 \end{definition}
@@
 \subsection{Structure of coeffects}
-In the structural coeffect calculus, the coeffect annotation was formed by a vector of coeffect
-scalars. The annotations in the unified coeffect calculus are similar, but a \emph{vector} is
-replaced with a more general \emph{container}. The primitive coeffect annotations in the unified
-calculus are formed of \emph{coeffect scalars}. The coeffect scalars are equipped with the same
-structure as in structural coeffect calculus (Definition~\ref{def:structural-scalar}). In this
-section, we refer to it as \emph{\cclrd{unified coeffect scalar}} (and we repeat the definition
-below). Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
+In the structural coeffect calculus, coeffect annotations are formed by vectors of coeffect
+scalars. The annotations in the unified coeffect calculus are similar, but the notion of \emph{vector}
+is replaced with a more general \emph{container}. The primitive coeffect annotations in the unified
+calculus are formed by coeffect scalars, which have the same structure as in the structural
+coeffect calculus (Definition~\ref{def:structural-scalar}). In this section, we refer to it as
+\emph{\cclrd{unified coeffect scalar}} (we repeat the definition below for clarity).
+Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
 scalar values are attached to the free-variable context. Finally, we define the
@@
 \paragraph{Coeffect scalars.}
-The following definition of the coeffect scalar structure repeats the
-Definition~\ref{def:structural-scalar} from the previous chapter.
+The following definition of the coeffect scalar structure repeats
+Definition~\ref{def:structural-scalar} from the Chapter~\ref{ch:structural}, with the only change
+that we now refer to it as \emph{\cclrd{unified coeffect scalar}}.
 
 \begin{definition}
+\label{def:unified-scal}
 A \emph{\cclrd{unified coeffect scalar}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
-$\C$ together with elements $\cunit, \,\czero \in \C$, binary operations $\cseq, \,\cpar$ such that
-$(\C, \,\cseq, \,\cunit)$ and $(\C, \,\cpar, \,\czero)$ are monoids and a binary relation $\cleq$ such
+$\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
+$(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and a binary relation $\cleq$ such
 that $(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
@@
 an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable,
-respectively.
-
-The $\cpar$ operation models combining of context demands arising from multiple parts of a
-program. The meaning depends on the coeffect container. The operation can either combine requirements
+respectively. The $\cpar$ operation models combining of context demands arising from multiple parts of a
+program. Its meaning depends on the coeffect container. The operation can either combine requirements
 of individual variables (structural coeffects) or requirements attached to the whole context of
@@
 \paragraph{Coeffect containers.}
-The coeffect container is a container that determines how are scalar coeffects attached to
-free-variable contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed positions,
+The coeffect container is a container that determines how are scalar coeffects attached to free-variable
+contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed sets of positions,
 the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.
@@
 $\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
-to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation
-$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in
-  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
-We write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
-The coeffect container is equipped with an operation that appends shapes (when concatenating
+to $\Gamma$. We write $\sclrd{l} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$ and
+$\SP_{\sclrd{l}}$ is the \emph{set} of positions for the given shape. A coeffect $\aclrd{\textbf{r}}$
+can now be seen as a function $\SP_{\sclrd{l}} \rightarrow \C$ that returns a scalar coefect
+$\cclrd{r}\in\C$ for each of the position defined by the shape $\sclrd{l}$.
+We write the set of positions for a shape $\sclrd{l}$ as $\SP(\sclrd{l})$ and a function that
+returns scalar coeffect as an exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{l})}$.
+
+Additionally, the coeffect container is equipped with an operation that appends shapes (when concatenating
 variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
@@
 \begin{definition}
+\label{def:unified-cont}
 A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$
@@
 \noindent
-As discussed earlier, we use two kinds of coeffect containers that describe the structure of vectors
-(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):
+As discussed earlier, we use two kinds of coeffect containers. Those that describe the structure of vectors
+(for structural coeffects) and those that describe the shape of trivial singleton container (for flat coeffects):
 
@@
 \label{ex:further-struct-shape}
-Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
+A structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
 where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the
@@
 \label{ex:further-flat-shape}
-Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
+A flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
 The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
@@
 \begin{definition}
+\label{def:unified-alg}
 Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
@@
 combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
-
-\begin{equation*}
-\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
+%
+\begin{equation*}
+\begin{array}{rcl}
+\apar_{\sclrd{m},\sclrd{n}} &:& \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})} \\
+\azip_{\sclrd{m},\sclrd{n}} &:& \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
+\end{array}
 \end{equation*}
@@
 
-The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:further-semantics}),
+The difference between $\azip$ and $\apar$ is clarified by the semantics \cite{coeffects-icfp14},
 where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
@@
 \label{sec:further-types}
-The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
-shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
-how the unified system differs from the structural system.
+We define the unified coeffect system in Figure~\ref{fig:unif-types}. It resembles the structural
+type system shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we
+focus on how the unified system differs from the structural system.
 
 The type system for the unified coeffect calculus is parameterized by a coeffect scalar
-$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$
+$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ (Definition~\ref{def:unified-scal}) together with a
+coeffect algebra $(\azip, \apar, \azero)$ (Definition~\ref{def:unified-alg})
 and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
-As in the structural system, free-variable contexts $\Gamma$ are treated as vectors (without duplicate
-variables) with associativity built-in. The order of variables matters, but can be changed
+As in the structural system, free-variable contexts $\Gamma$ are treated as vectors of distinct
+variables with associativity built-in. The order of variables matters, but can be changed
 using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
-are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
-annotated with coeffects scalars.
+are shape-indexed coeffects (rather than simple vectors as before). As before, function arrows
+are annotated with coeffect scalars.
 
@@
 The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
-differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
-coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$.
+differently. The $\alift{-}$ operation does not create a \emph{vector}, but a coeffect container
+of shape $\sunit$ that returns the coeffect scalars $\cunit$ for all positions in the singleton shape.
 The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.
@@
 
+~
+
 Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
@@
 \paragraph{Structural rules.}
-The merging/splitting operations in the structural rules are changed in the same way as in the
-rules above. It is also worth noting that structural coeffect system used vectors of multiple
-elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch})
+The merging/splitting operations in the structural rules are generalized in the same way as in the
+rules above. It is also worth noting that structural coeffect systems use vectors of elements while
+the unified system uses the container structure.
+For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch})
 rule denoted two-element vectors. In the unified system, this is replaced with merging/splitting
@@
 singleton vectors. However, this generalization allows us to capture flat coeffect systems as well.
-The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
+The lifting operation in flat systems simply returns the lifted scalar; operators $\azip$ and
 $\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:further-flat},
-thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not
-affect the coeffect annotation are admissible for all flat systems embedded in the unified
-coeffect calculus.
+thanks to the properties of coeffect scalars, the operations of contraction, weakening and exchange
+that do not affect the coeffect annotation are admissible for all flat systems embedded in the
+unified coeffect calculus.
 
@@
 The unified calculus closely resembles the structural system and so using it to model structural
-systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector}
-of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect algebra formed by
-a vector (free monoid) of scalars.
+systems is easy -- given a (structural) coeffect scalar, we use the coeffect container that describes
+a \emph{vector} of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect
+algebra formed by a vector (free monoid) of scalars.
 
@@
 Flat coeffect systems are characterised by a singleton set of shapes (Example~\ref{ex:further-flat-shape}).
-In this setting, the context annotations $\C^{ \SP(\sflat) } = \{0\} \rightarrow \C$ are equivalent
+In this setting, the context annotations $\C^{ \SP(\sflat) }$ are defined as function values
+$\{0\} \rightarrow \C$. The domain of the function is a singleton set and so the values are equivalent
 to coeffect scalars $\C$. In addition to the coeffect scalar structure, we need to define $\apar$
 and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$
-(merging of shaped coeffect annotations). The $\azip$ operation (corresponding to $\czip$
-in flat coeffect calculus) needs to be provided explicitly.
+(merging of coeffect annotations). The $\azip$ operation (corresponding to $\czip$
+in flat coeffect calculus) is provided explicitly.
 
@@
 \begin{proof}
-The rule is admissible using the following derivation:
+The rule is admissible as shown using the following derivation:
 \begin{equation*}
@@
 \begin{lemma}
-A unified coeffect calculus capturing flat properties admits exchange that does not change the
+A unified coeffect calculus capturing flat properties admits exchange rule that does not change the
 coeffect annotation.
@@
 \begin{lemma}
-A unified coeffect calculus capturing flat properties admits contraction that does not change the
+A unified coeffect calculus capturing flat properties admits contraction rule that does not change the
 coeffect annotation.
@@
 \noindent
-In the last two cases, we need to turn the coeffect into a form that is required by the exchange
+In the last two lemmas, we need to turn the coeffect into a form that is required by the exchange
 and contraction rules. Aside from idempotence, we could use the unit property and obtain
 \eg~$\czero\,\czip\,\cclrd{r}\,\czip\,\czero\,\czip\,\czero$. However, this approach does not
-work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not ever require
+work because $\cpar$ and $\czip$ may have different unit elements (in fact, we do not even require
 the existence of a unit for $\czip$).
@@
 
-In this section, we briefly discuss the other technique. That is, we embed indexed
-comonads into a $\lambda$-calculus as additional constructs. To do that, we introduce the
+In this section, we briefly discuss the other technique. That is, we embed indexed comonads into
+a $\lambda$-calculus as additional constructs of the meta-langauge. To do that, we introduce the
 type constructor $\ctyp{\cclrd{r}}{\tau}$ which represents a value $\tau$ wrapped in additional
 context (semantically, this corresponds to an indexed comonad) and we add language constructs
-that correspond to the operations of indexed comonads. The coeffect meta-language highlights the
-relationship between coeffects and closely related work on contextual modal type theory (CMTT)
+that correspond to the operations of indexed comonads. The coeffect meta-language so derived highlights
+the relationship between coeffects and closely related work on contextual modal type theory (CMTT)
 \cite{logic-cmtt}. Developing the system further is also an interesting future research direction.
@@
 requirements are formed by an abstract coeffect algebra, which is more general and can capture
-variable contexts, but also integers, two-point lattices, etc.
+variable contexts, but also integers, two-point lattices for liveness, etc.
 
 Secondly, CMTT uses different intuitive understanding of the comonad (type constructor) and
-the associated operations. In categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
+the associated operations. In the categorical semantics of coeffect calculi, the $\ctyp{\cclrd{r}}{\tau}$
 constructor refers to a value of type $\tau$ \emph{together} with additional context specified
@@
 
-In contrast in CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the
+By contrast, in CMTT\footnote{To avoid using different notations, we write $\ctyp{\Psi}{\tau}$ instead of the
 original $[\Psi]{\tau}$} the type $\ctyp{\Psi}{\tau}$ models a value that \emph{requires} the
@@
 \end{equation*}
-Two readings of the signatures are possible, but they give quite different meanings to the operations:
+As discussed in Section~\ref{sec:path-sem-contextdep}, these signatures can be read in two different
+ways. One is the coeffect reading used in this thesis and another is the CMTT reading. The
+annotations have a different meaning in the two readings:
 
@@
 \noindent
-Although the different reading does not affect formal properties of the systems, it is important
+Although the different readings do not affect formal properties of the systems, it is important
 to understand the difference when discussing the two systems as they provide a different
@@
 
-The sketch of a coeffect meta-language in the following section attempts to bridge the gap
-between coeffects and CMTT. Just like CMTT, it embeds comonads as language constructs, but
-it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which tracks only
-sets of variables. Future work on the coeffect meta-language would thus be an interesting
-development for both coeffect systems and CMTT.
+In the following section, we present a sketch of a comonadically-inspired meta-language, which
+attempts to bridge the gap between coeffects and CMTT. Just like CMTT, it embeds comonads as language
+constructs, but it annotates them with a (flat) coeffect algebra, thus it generalizes CMTT which
+tracks only sets of variables. Future work on the general coeffect meta-language would thus be an
+interesting development for both coeffect systems and CMTT.
 
@@
 \end{equation*}
-%
+
+~
+
+~
+
 The $!e$ and $\kvd{let\;box}$ constructs correspond to the \ident{counit} and \ident{cobind}
-operation of the comonad. To define meta-language for flat indexed comonads, we also include
-\kvd{split} and \kvd{merge} that embed the corresponding operations.
+operation of the comonad. To make our meta-language expressive enough for flat indexed comonads,
+we also include \kvd{split} and \kvd{merge} that embed the corresponding operations.
 
@@
 transformation on one or two values in a context and then add the new value as a fresh variable
-to the variable context. We omit subcoeffecting, but it could be easily added following
-the method used elsewhere.
+to the variable context. We do not discuss subcoeffecting, but it could be easily added following
+the method used in Section~\ref{sec:semantics-flat-monoidal}.
 
@@
 
-The \emph{meta-language} approach of embedding comonads in a language is more general than the
-\emph{semantics} approach. This thesis focuses on a narrower use that better guides the design of
-a type system for context-aware programming languages.
-
-However, it is worth demonstrating that the (flat) coeffect calculus can be embedded in the
+The \emph{meta-language} approach of embedding comonads within a language is more general than the
+\emph{semantics} approach. This thesis focuses on a syntacitcally more restricted use that better
+guides the design of a type system for context-aware programming languages.
+
+We first demonstrate that the flat coeffect calculus can be embedded in the
 meta-language described in the previous section. This may be desirable, \eg~when using the
-meta-language for reasoning about context-aware computations. We briefly consider the embedding
-as it illuminates the relationship between coeffects and CMTT (unfortunately it is not possible
-to embed coeffect calculi in CMTT because of the more general annotations structure).
+meta-language for reasoning about context-aware computations. We briefly consider this embedding
+as it illuminates the relationship between coeffects and CMTT. It is not possible to embed
+coeffect calculi in CMTT without generalisation such as the one presented here, because coeffect
+systems have a more general structure of annotations.
 
@@
 we define $\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}_v$ as its embedding in the coeffect
-meta-language. Note that the translation is indexed by $v$, which is a name of variable used
-to represent the entire variable context of the source language. The translation is defined in
-Figure~\ref{fig:conclusions-embed}. The embedding resembles the semantics discussed in
+meta-language. The translation (Figure~\ref{fig:conclusions-embed}) is indexed by $v$, which is a name of variable used
+to represent the entire variable context of the source language. The embedding resembles the semantics discussed in
 Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
@@
 \end{equation*}
-\figcaption{Embedding flat coeffect calculus in coeffect meta-language }
+\figcaption{Embedding the flat coeffect calculus in the coeffect meta-language }
 \label{fig:conclusions-embed}
@@
 
-Most of the related and future work has been discussed throughout the thesis. This section summarizes
-two interesting directions for future work and the related work in the area. First, we look at a way of
-embedding coeffects in existing languages such as Haskell that is inspired by the Haskell's ``do'' notation
+This section summarizes two interesting directions for future work and the related
+work in the area. First, we look at a way of embedding coeffects in existing languages
+such as Haskell that is inspired by the Haskell's ``do'' notation for monads
 (Section~\ref{sec:further-impl}). Second, we discuss extensions to the theory of coeffects that
@@
 In order to do this, the target programming language needs to provide two features; one that allows
-embedding of context-aware computations themselves in programs akin to the ``do'' notation
-in Haskell and one that allows tracking of the contextual information in the type system.
+easiler embedding of context-aware computations themselves in programs akin to the ``do'' notation
+for monadsin Haskell and one that allows tracking of the contextual information in the type system.
 
@@
 flat coeffect calculus (Chapter~\ref{ch:flat}). Extending the ``codo'' notation to support calculi based
-on the structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting future work --
-this would require explicitly manipulating individual context variables and application of structural
-rules, which is not needed in flat coeffects.
+on the structural coeffect calculus (Chapter~\ref{ch:structural}) is interesting future work --
+this would require tracking the use of variables and the application of corresponding operations
+on the coeffect annotations, according to the structural rules.
 
@@
 
-More lightweight syntax for effectful computation can be obtained using
-techniques that automatically insert the necessary monadic plumbing (without
-special syntax). This has been done for effectful computations \cite{monads-lightweight-ml}
-and a similar approach would be worth exploring for coeffects.
+More lightweight syntax for effectful computation can be obtained by
+automatically insertin the necessary monadic plumbing (without
+special syntax). This has been done for effectful computations in Lightewight ML
+\cite{monads-lightweight-ml} and a similar approach would be worth exploring for coeffects.
 
@@
 dependent types and can be made accessible in more mainstream languages. Dependent ML
-\cite{types-dependent-ml} provides an interesting example of a language with some dependent typing
-(\eg~computations with integers) which is still close to its non-dependently-typed predecessor ML.
+\cite{types-dependent-ml} provides an interesting example of a language with limited dependent typing
+(arithmetic) which is still close to its non-dependently-typed predecessor ML.
 
@@
 as coeffect algebras into the type system. An \emph{algebra provider} would then be a library that
-specifies the objects of the algebra, its equational laws and generalization rules for type inference.
+specifies the objects of the algebra, its equational laws and rules for type inference.
 This could provide an easy-to-use way of embedding coeffect tracking in pragmatic languages
 such as F\#. It is worth noting that the mechanism could also subsume F\# units of measure
-\cite{types-units-of-measure}; these could be alternatively provided via one such \emph{algebra
+\cite{types-units-of-measure}; these could then be seen as one such \emph{algebra
 provider}.
@@
 The coeffect calculi presented in this thesis are based on a simple $\lambda$-calculus, comprising
-variables, lambda abstraction, application and let-binding. This lets us focus on fundamental
-properties of coeffect systems (and explain how coeffect system differ from better-known effect
+variables, lambda abstraction, application and let-binding. This approach lets us focus on fundamental
+properties of coeffect systems (and explain how coeffect system differ from the better-known effect
 systems), but it is hardly sufficient for a realistic programming language.
@@
 Further work is to extend the coeffect systems presented here to a full programming language. A
-useful reference is the work of Nielson and Nielson~\cite{effects-nielson} who consider a
+useful starting point is the work of Nielson and Nielson~\cite{effects-nielson} who consider a
 similar development for effect systems, adding conditionals, recursion and polymorphism.
@@
 the prototype implementation discussed in Chapter~\ref{ch:impl} provides basic support for
-type inference, however formal treatment of the algorithm is left to future work.
+type inference. This work has concentrated on the definition of coeffect systems and type checking;
+detailed treatment of type inference is future work.
 
@@
 
-We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:unified-meta}.
+We briefly outlined a calculus based on the ``meta-language'' approach in Section~\ref{sec:further-meta}.
 Developing this technique further could unify coeffect systems with the work on Contextual Modal
@@
 In the \emph{flat} and \emph{structural} coeffect calculi, we attach annotations to the \emph{entire
-context} and to \emph{individual variables}, respectively. We later unified the two systems
-in Section~\ref{sec:unified-unified}.
+context} and to \emph{individual variables}, respectively. This chapter sketched unification of
+the two systems.
 
@@
 which uses a tree-like structure of variable contexts (also discussed in Section~\ref{sec:path-logic}).
-The definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
-typing requires tree-like variable context, while we use a vector. Furthermore, bunched typing
-annotates internal nodes of the tree (sub-contexts are combined using ``;'' or using ``,'') rather
-than just variables. Finding a simple coeffect system that is capable of capturing bunched typing
-is thus an interesting further work. Indeed, this could lead to numerous uses of coeffects as
-bunched typing is the basis for widely-used separation logic \cite{substruct-separation-logic}.
+The current definition of the \emph{unified coeffect calculus} is likely not expressive enough -- bunched
+typing requires tree-like variable context, while we use a vector. Finding a simple coeffect system
+that is capable of capturing bunched typing is thus an interesting further work. Indeed, this could
+lead to numerous uses of coeffects as bunched typing is the basis for widely-used
+separation logic \cite{substruct-separation-logic}.
 
@@
 discuss the technical requirements for embedding coeffects as a domain-specific language (DSL) into
-a statically-typed functional language with rich type system such as Haskell. This can be done by
+a statically-typed functional language with a rich type system such as Haskell. This can be done by
 following the successful example of Haskell's ``do'' notation.

### 2016-05-30 text/Chapter09.tex
Edits and tweaks (Chapter 8 and 9 mostly)

@@
 
-Some of the most fundamental academic work is not the one solving hard research problems, but the
-one that changes how we understand the world. Some philosophers argue that \emph{language} is the
+Some of the most fundamental academic work is not that which solves hard research problems, but
+that which changes how we understand the world. Some philosophers argue that \emph{language} is the
 key for understanding how we think, while in science the dominant thinking is determined by
 \emph{paradigms} \cite{philosophy-kuhn} or \emph{research programmes} \cite{philosophy-lakatos}.
-In a way, programming languages play a similar role for computer science and software development.
+Programming languages play a similar role for computer science and software development.
 
 This thesis aims to change how developers and programming language designers think about
-\emph{context} or \emph{execution environments} of programs. Such context or execution environment
-has numerous forms -- mobile applications access network, GPS location or user's personal data;
+\emph{contexts} or \emph{execution environments} of programs. Such contexts or execution environments
+have numerous forms -- mobile applications access network, GPS location or user's personal data;
 journalists obtain information published on the web or through open government data initiatives;
@@
 (Section~\ref{sec:conc-summary}). The thesis looks at two kinds of context-dependence, identifies
-common patterns and captures those using two \emph{coeffect calculi}. It gives formal semantics
-using categorical foundations and uses it as a basis for prototype implementation. Finally,
+common patterns and captures those using two \emph{coeffect calculi}. It gives a formal semantics
+using categorical insights and uses it as a basis for prototype implementation. Finally,
 Section~\ref{sec:conc-conclusions} concludes the thesis.
@@
 
-The main contribution of the chapter is that it presents the disjoint language features in a
-unified way. We show type systems and semantics for many of the languages, illuminating
-the fact that they are closely related.
+The main contribution of Chapter~\ref{ch:applications} is that it presents the disjoint language
+features in a unified way. We show type systems and semantics for many of the languages,
+illuminating the fact that they are closely related.
 
@@
 $\eta$-expansion do not generally preserve the type of an expression, but we identify two
-conditions when this is the case.
+conditions satisfied by two of our instances when this is the case.
 
@@
 
-We unify systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
+We characterise systems with \emph{per-variable} contextual properties in Chapter~\ref{ch:structural},
 which describes our \emph{structural coeffect calculus} (Section~\ref{sec:struct-calculus}).
@@
 In particular, type preservation for $\beta$-reduction and $\eta$-expansion holds for all instances of the
-structural coeffect calculus. This follows from the fact that structural coeffect
-associates contextual requirements with individual variables and preserves the connection by
+structural coeffect calculus. This follows from the fact that structural coeffects
+associate contextual requirements with individual variables and preserves the connection by
 including explicit structural rules (weakening, exchange and contraction).
@@
 
-Both of the coeffect calculi are parameterized and can be instantiated to obtain a language with
+Both forms of coeffect calculi are parameterized and can be instantiated to obtain a language with
 a type system that tracks concrete notion of context dependence. The coeffect calculus can be seen
@@
 
-A part of the thesis is an interactive web-based essay (\url{http://tomasp.net/coeffects}),
-which implements three simple context-aware languages. In addition to more traditional prototype
-language implementations, the essay provides a number of novel features. As usual, it lets users
+Associated with this essay is an interactive web-based essay, which implements three simple
+context-aware languages (Chapter~\ref{ch:impl}). In addition to traditional prototype language
+implementations, the essay provides a number of novel features. As usual, it lets users
 write and evaluate sample programs, but it also lets them explore the details of the coeffect
@@
 
-The implementation (Chapter~\ref{ch:impl}) serves two purposes. First, it links together all parts
-of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the two coeffect calculi
-to build a language framework that is then instantiated to implement three concrete context-aware
-languages. The  framework provides the type system and translation (based on the
+The implementation (Section~\ref{ch:impl-theory}) serves two purposes. First, it links together all parts
+of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the flat and structural
+coeffect calculi to build a language framework that is then instantiated to implement three concrete
+context-aware languages. The  framework provides the type system and translation (based on the
 comonadically-inspired semantics) that is used for program execution. This shows the practical
@@
 
-Second, the implementation uses novel kind of medium (Section~\ref{sec:impl-essay}) to make our work
-\emph{accessible} and \emph{explorable}. Building a production-ready programming language is outside
-the scope of the  thesis, so instead, our goal is to explain the theory and inspire authors of
-real-world programming languages to include support for context-aware programming, ideally using
-coeffects as a sound foundation. For this reason, we make the implementation accessible over web
-without requiring installation of specialized software and we make it explorable to encourage
-active reading.
+Second, web-based essay (\url{http://tomasp.net/coeffects}) discussed in Section~\ref{sec:impl-essay}
+is novel from the pedagogical perspective. It makes our work \emph{accessible} and \emph{explorable}.
+Building a production-ready programming language is outside the scope of the  thesis, so instead,
+our goal is to explain the theory and inspire authors of real-world programming languages to
+include support for context-aware programming, ideally using coeffects as a sound foundation.
+For this reason, we make the implementation accessible over web without requiring installation
+of specialized software and we make it explorable to encourage \emph{active reading}.
 

### 2017-01-16 main.tex
Add interactive essay to abstract
Edits in further work and the rest (Chapter 8)
Addressing feedback, fix demands (Chapter 6)
Edits and tweaks (Chapter 8 and 9 mostly)
Edits pages 0-30

@@
 
-\newcommand{\atimes}{ \textcolor{aclr}{\times} }
+\newcommand{\vtimes}{ \mathbin{+\mkern-10mu+} }
+
+\newcommand{\atimes}{ \textcolor{aclr}{\mathbin{+\mkern-10mu+}} }
 \newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
@@
 
-\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
-\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
+\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{180}{$\perp$}\mkern-10mu\rotatebox[origin=c]{180}{$\perp$}} } }
+\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\perp\mkern-10mu\perp} } }
 
@@
 
-%\include{text/Titlepage} % Main title page
+\include{text/Titlepage} % Main title page
 
-%\cleardoublepage\include{text/Declaration} % Publications from the thesis page
+\cleardoublepage\include{text/Declaration} % Publications from the thesis page
 
-%\cleardoublepage\include{text/Abstract} % Abstract page
+\cleardoublepage\include{text/Abstract} % Abstract page
 
@@
   programs need to run on a variety of devices that are all different, but provide unique rich
-  capabilities. For example, application running on a phone can access the GPS sensor, while
-  application running in the cloud can access GPU computing resources. Both diversity and richness
+  capabilities. For example, an application running on a phone can access the GPS sensor, while
+  a service running in the cloud can access GPU computing resources. Both diversity and richness
   can only be expected to increase with trends such as the internet of things.
 
-  In this thesis, we argue that the creating programming languages that allow the programmer to
+  In this thesis, we argue that creating programming languages that allow the programmer to
   better work with the environment or \emph{context} in which applications execute is the next
@@
 
-%\include{text/Chapter01}
-%\include{text/Chapter02}
-%\include{text/Chapter03}
+\include{text/Chapter01}
+\include{text/Chapter02}
+\include{text/Chapter03}
 
@@
 
-%\include{text/Chapter04}
-%\include{text/Chapter05}
-%\include{text/Chapter06}
+\include{text/Chapter04}
+\include{text/Chapter05}
+\include{text/Chapter06}
 
@@
 
-%\appendix
-%\include{text/AppendixA}
-%\include{text/AppendixB}
+\appendix
+\include{text/AppendixA}
+\include{text/AppendixB}
 

### 2017-01-16 text/Abstract.tex
Edits pages 0-30

@@
 sensor on a phone and database on the server), but also more traditional notions such as
-variable usage (\eg~in liveness analaysis and linear logics) or past values in
+variable usage (\eg~in liveness analysis and linear logics) or past values in
 stream-based dataflow programming. Our first contribution is the discovery of the connection
@@
 \emph{indexed comonad} (based on dualisation of the well-known monad structure), use it
-to define operational semantics and prove safety result for the calculi.
+to define operational semantics and prove type safety of the calculi.
 

### 2017-01-16 text/Chapter01.tex
Edits pages 0-30

@@
 concerns that have led to the development of new programming languages. This helps to explain why
-context-aware programming is important. The examples are by no means representative, but they
-illustrate various kinds of motivations well.
+context-aware programming is important. The examples are by no means exhaustive, but they
+are representative of the various kinds of motivations.
 
@@
 
-Accessing ``big data'' sources is an example of a more subtle challenge. Initiatives like open
-government data\footnote{In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
-certainly make more data available. However, to access the data, one has to parse CSV and Excel files,
-issue SQL or SPARQL queries (to query database and the semantic web, respectively).
+Accessing ``big data'' is an example of a more subtle challenge. Initiatives like open
+government data make more data available, but to access the data, one has to parse CSV and 
+Excel files, issue SQL or SPARQL queries (to query database and the semantic web, respectively).
 
@@
 
-~\\
-
 \paragraph{Context-aware programming.}
@@
 expose such flaws. We look at a number of basic programs that rely on contextual
-information, we explain why the currently dominant solutions are inappropriate and then
-briefly outline how this thesis solves the problems.
+information, we explain why the currently dominant solutions are inappropriate.
+
+\paragraph{The thesis.}
+The key claim of this thesis is that the theory of \emph{coeffects} that we develop provides
+a unified way of capturing how programs rely on the context, or execution environment, 
+in which they execute. Perhaps most importantly, coeffects provide a single conceptual framework 
+that can capture a wide range of previously disconnected notions of context-awareness.
+
+Thanks to coeffects, programming languages for rich and diverse execution environments will
+be able to offer a simple, safe and unified programming model for interacting with the 
+environment. In this thesis, we develop the theoretical foundations for such languages and
+we develop a prototype implementation that illustrates the benefits of coeffects, but also 
+serves as a rich guide for future langauge implementers.
 
@@
 %     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###
-%                                                                                                                                                                       % ==================================================================================================
+%                                                                                                                                                                       
 % ==================================================================================================
@@
 The following sub-sections explore 4 examples in more detail. The examples are chosen to
-demonstrate two distinct forms of contexts that are studied in this thesis -- first two are
+demonstrate two distinct forms of contexts that are studied in this thesis -- the first two are
 related to the program environment and the latter two are associated with individual variables of
@@
 SQL discussed in the introduction, but it is a more broadly applicable technique for programming
-in heterogeneous environments. It lets developers targets multiple runtimes that have limited
+in heterogeneous environments. It lets developers target multiple runtimes that have limited
 execution capabilities.
@@
 uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
-not a technical limitation of LINQ, but a fundamental problem of the approach.
+not a technical limitation of LINQ, but a fundamental problem of an approach where code written 
+using libraries for one execution environment is translated and run in another, more limited, 
+execution environment.
 
@@
 Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
-metadata about variables that are in some ways special. Such metadata can determine how the variables
-can be used. Here, \ident{name} comes from the user input. This information about the value should
-be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
-not directly contain user input (in an unchecked form).
+metadata about values that are in some ways special. Such metadata can determine how the values
+can be used. Here, the value stored in \ident{name} comes from the user input. This information about 
+the value should be propagated to \ident{query}. The \ident{SqlCommand} object should then require 
+arguments that can not directly contain user input.
 
@@
 
-Consider a simple programming language with arrays where $n^{\mathit{th}}$ element of an
+Consider a simple programming language with arrays where the $n^{\textit{th}}$ element of an
 array \ident{arr} is accessed using $\ident{arr}[n]$. We focus on writing stencil computations
@@
 As an example, consider a news reader app consisting of a server-side component (which stores the
-news in an SQL database) and a number of clients applications for popular platforms (Android, Windows
+news in an SQL database) and a number of clients applications for popular platforms (iPhone, Windows
 Phone, etc.). A simplified code excerpt that might appear somewhere in the implementation is shown in
@@
 We assume that the language supports cross-compilation and splits the single program into three
-components: one for the server-side and two for the client-side, for iPhone and Windows platforms,
+components: one for the server-side and two for the client-side, for iPhone and Windows Phone,
 respectively. The cross-compilation could be done in a way similar to Links \cite{app-distributed-links},
@@
 Furthermore, integrating contextual information into the type system can provide information for
-modern developer tools. For example, many editors for F\# display inferred types when placing mouse
-pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
+modern developer tools. For example, many editors for F\# display inferred types when placing the 
+mouse pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
 %
@@
 
-The concrete syntax used here is just for illustration. Furthermore,
+The concrete syntax is just for illustration and
 some information could even be mapped to other visual representations -- for example, differently
 coloured backgrounds for platform-specific functions. The key point is that the type provides a
-number of useful information:
+number of pieces of useful information:
 
@@
 
-\item The \ident{gpsLocation} function accesses the GPS sensor and since we call it in
+\item The \ident{gpsLocation} function accesses the GPS sensor and since we call it 
   from \texttt{fetchLocalNews}, this function also requires GPS (the requirement is propagated
@@
 
-\item We can compile the program for two client-side platforms - the entry points are \ident{iPhoneMain}
+\item We can compile the program for two client-side platforms -- the entry points are \ident{iPhoneMain}
   and \ident{windowsMain} and require iOS and Windows user-interface libraries, together with
@@
 %
-This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
-be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
-but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.
+This judgement assumes that the type of $x$ and $y$ is \ident{int} and so the result must also
+be \ident{int}. The expression might also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
+but not in a context where types of $x$ and $y$ do not match.
 
@@
 certain kind. They can be used for properties that describe how programs \emph{affect} the environment.
-For context-aware languages, we instead need to track what programs \emph{require} from the environment.
+For context-aware languages, we instead need to track what programs \emph{require} from the 
+environment. This intuitive distinction is made more precise in Section~\ref{sec:semantics-related}.
 
@@
 \ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
-  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
+  \vdash \ident{query}(\ident{cmd}, \ident{password}, \kvd{time}) : \ident{news list}
 \end{equation*}
 %
-The expression queries a database and gets back a list of news values as the result. Recall from the
-earlier discussion that there are two contextual information that are desirable to track for this
-expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
+The expression queries a database and gets back a list of news based on the spefieid \kvd{time}. 
+There are three pieces of contextual information that are desirable to track for this expression. 
+First, the call to the \ident{query} primitive requires \emph{database access}. Second, 
 the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
-an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this
-information in the following way (we slightly refine the notation later):
+an unsecure network connection. Third, the expression uses a special construct \kvd{time}, which
+requires access to \emph{system clock}. The \emph{coeffect systems} developed in this thesis 
+capture this information in the following way (we slightly refine the notation later):
 %
@@
  \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
-   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
- \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
+   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database}, \ident{clock} \}}}  \\[-0.1em]
+ \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}, \kvd{time}) : \ident{news list}
 \end{array}
@@
 %
-Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
-context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
-in the context -- they also capture detailed information about the execution environment.
+The typing judgement includes an additional annotation that now captures 
+contextual requirements of the expression. The annotation is attched to the variable context 
+$\Gamma$. As discussed later (Section~\ref{sec:flat-calculus-understanding}), this better reflects
+how contextual information are used in the type system, but it also matches the semantics 
+(Section~\ref{sec:semantics-theory}). Rather than attaching additional information to the
+\emph{resulting type}, we attach them to the variable context $\Gamma$.
+
+In other words, the context in coeffect systems consists of the available variables, but also
+tracks detailed information about the execution environment.
 In the above example, the system tracks metadata about the variables and annotates \ident{password}
-as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
-that the execution requires an access to database.
+as sensitive and it tracks requirements about the execution environment, for example,
+that the execution requires clock.
 
@@
 
-It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell
+\paragraph{Context demands and lambda abstraction.}
+The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
+Given an effectful expression such as $\ident{print}~\str{hi}$, the function $\lambda x.\ident{print}~\str{hi}$
+is an effect free value that delays all effects. The message is printed when we run the function,
+but not when the function is declared.
+
+Contextual properties do not follow this pattern. As discussed in 
+Section~\ref{sec:applications-structure-lam}, context demands cannot always be delayed. For 
+example, consider a context-aware expression $\kvd{time}$ that requires access to system clock.
+If we create this function on a server in a distributed system, send it to a client and then
+run it, the function body can access the clock in two ways. It can use the clock of the server (when the
+function was created), or it can use the clock of the client (when the function was executed).
+We return to this example in Section~\ref{sec:applications-flat-distr}.
+
+Another important difference between effects and coeffects becomes apparent when we consider their
+semantics. It is well-known fact that \emph{effects} correspond to \emph{monads} and languages such as Haskell
 use monads to provide a limited form of effect system. An interesting observation made in this thesis
@@
 
-The key claim of this thesis is that programming languages need to provide better ways of capturing
-how programs rely on the context, or execution environment, in which they execute. This chapter
-shows why this is an important problem. We looked at a number of properties related to context that
-are currently handled in ad-hoc and error-prone ways. Next, we considered the properties in a
+This chapter shows why capturing how programs rely on the context, or execution environment, in 
+which they execute is an important problem. We looked at a number of properties related to context 
+that are currently handled in ad-hoc and error-prone ways. Next, we considered the properties in a
 simplified, but realistic example of a client/server application for displaying local news.
@@
 \item In Chapter~\ref{ch:impl}, we use the translation as a basis for a prototype
-  implementation of three simple context-aware programming languages. Our implementation serves as
-  an empirical evidence showing how coeffects simplify programming with context and provide
-  additional safety guarantees. We present the implementation in the format of an \emph{interactive
-  essay} (\url{http://tomasp.net/coeffects}) that encourages active reading and exploration of
-  various aspects of the theory.
+  implementation of three simple context-aware programming languages. We use a format of web-based
+  \emph{interactive essay} (available at \url{http://tomasp.net/coeffects}), which shows how 
+  coeffects simplify programming with context by allowing the reader to write and run simple 
+  context-aware programs, but also explore the theory behind the implementation including the 
+  typing derivation and the translation.
 
@@
 there is a unified notion of \emph{context}, capturing many common scenarios in programming,
-and that programming language designers need to provide ways for working with this context (using
-\emph{coeffects} or not). This greatly reduces the number of distinct concepts that software
+and that programming language designers need to provide ways for working with this context.
+This greatly reduces the number of distinct concepts that software
 developers need to keep in mind of when building applications for the rich and diverse

### 2017-01-16 text/Chapter02.tex
Edits pages 0-30

@@
 only evaluated if the environment provides values for variables $x$ and $y$. A variable
-\emph{requirements} can be satisfied in two standard ways that are characterized as \emph{dynamic}
+\emph{requirement} can be satisfied in two standard ways that are characterized as \emph{dynamic}
 and \emph{static} (or lexical) variable binding. Consider the following simple program:
@@
 \begin{array}{l}
-\kvd{let}~\ident{f}=\\[-0.25em]
-\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
-\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
+\kvd{let}~\ident{f}=\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
 \kvd{let}~\ident{x}=5~\kvd{in}\\
@@
 \item \textsc{Dynamic binding.} In a language with dynamic binding (some variants of LISP), the
-  variable value is dynamically bound to the last value that has been added to the environment
+  variable value is dynamically bound to the topmost value for the available kept on the stack 
   during program execution. Thus, the \ident{x} variable inside the lambda function refers to
@@
 example, the language could use static binding by default, but resort to dynamic binding if a
-variable is not available in the lexical scope. One such system is Haskell's implicit parameters
+variable is not available in the lexical scope. One such system is implicit parameters
 discussed in the next section.
@@
 Languages with static scoping resolve variable bindings at compile-time. In contrast,
-languages with dynamic variable binding typically do not perform compile-time checks -- if a
-program attempts to access a variable that is not available in the environment, a runtime error
-occurs. This is possibly an artifact of their implementation -- implementing static binding
-\emph{without} checking would be cumbersome and implementing static checking for dynamic binding is
-requires a more sophisticated type system than traditional (Section~\ref{sec:path-effects-coeff}).
+languages with dynamic variable binding cannot resolve variables at compile-time. They 
+typically perform runtime checks -- if a program attempts to access a variable that is not 
+available in the environment, a runtime error occurs. Implementing static checking for dynamic
+binding is also possible, but it requires a more sophisticated type system 
+(Section~\ref{sec:path-effects-coeff}), while implementing static binding \emph{without} checking 
+would be cumbersome and so dynamically scoped languages are often dynamically typed.
 
@@
 
-Haskell uses static binding for ordinary variables, but it additionally provides a feature named
+Haskell uses static binding for ordinary variables, but GHC additionally provides a feature named
 \emph{implicit parameters} \cite{app-implicit-parameters} that adds a special kind of variables,
@@
 other parts of the program, we explicitly define an algorithm that chooses the preferred unique
-derivation (Section~Section~\ref{sec:flat-unique} and Section~\ref{sec:struct-unique}).
+derivation (Section~\ref{sec:flat-unique} and Section~\ref{sec:struct-unique}).
 
@@
 multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
-but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
-behaviour of Haskell).
+but we uniquely choose one preferred typing (mimicking the behaviour of GHC for implicit parameters).
 
@@
 language that already supports effectful operations as a way of increasing the safety -- the type
-and effect system provides stronger guarantees than a plain type system. Filinsky
+and effect system provides stronger guarantees than a plain type system. Filinski
 \cite{effects-comprehensive} refers to this approach as \emph{descriptive}\footnote{In contrast
@@
 computation} and provides a way of constructing and composing effectful computations:
-
-~
-
+%
 \begin{definition}
@@
 monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in
-Haskell, list monad ($\mtyp{}{\tau} = 1 + (\tau \times \mtyp{}{\tau})$) and other.
+Haskell and list monad ($\mtyp{}{\tau} = 1 + (\tau \times \mtyp{}{\tau})$).
 In programming language semantics, monads can be used in two distinct ways.
@@
 need to extend the language in any way -- we assume that the language already contains the
-effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
+effectful primitives (such as the assignment operator $x \leftarrow e$). A judgement
 of the form $x\!:\!\tau_1 \vdash e : \tau_2$ is interpreted as a morphism $\tau_1 \rightarrow \mtyp{}{\tau_2}$,
@@
 
-\paragraph{Meta-language interpretation.} ~When designing meta-language based on monads, we need to
+\paragraph{Meta-language interpretation.} When designing a meta-language based on monads, we need to
 extend the lambda calculus with additional type(s) and expressions that correspond to monadic
@@
 $\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no effects) from a
-value and $\kvd{let}_\mtyp{}{}$ is sequences effectful computations. In the semantics,
+value and $\kvd{let}_\mtyp{}{}$ sequences effectful computations. In the semantics,
 monads are not needed to interpret variable access and application, they are only used in the
@@
 Moggi). Before looking at the details, we use the following definition of comonad:
-
-~
-
+%
 \begin{definition}
@@
 \noindent
-The definition is similar to a monad with ``reversed arrows''. Intuitively, the $\ident{counit}$
+The definition is dual to a monad. Intuitively, the $\ident{counit}$
 operation extracts a value $\alpha$ from a value that carries additional context $\ctyp{}{\alpha}$.
@@
 In the outline presented here, we ignored lambda abstraction. Similarly to monadic semantics,
-where lambda abstraction requires \emph{strong} monad, the comonadic semantics also requires
+where lambda abstraction requires a \emph{strong} monad, the comonadic semantics also requires
 additional structure called \emph{symmetric (semi)monoidal} comonads. This structure is
@@
 In languages inspired by modal logics, types can have the form $\square \tau$. In the work of
-Pfenning and Davies, this is the type of a term that is provable with no assumptions. In distributed
-programming language ML5 by Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus}, the
+Pfenning and Davies, this is the type of a term that is provable with no assumptions. In the
+ML5 language by Murphy et al. \cite{app-distributed-ml5,logic-distributed-calculus}, the
 $\square \tau$ type means \emph{mobile code}, that is code that can be evaluated at any node of a
@@
 context-dependent computations in terms of comonads. We provide the rest of the story known
-from the marriage of monads and effects. We develop coeffect calculus with an type system that
+from the marriage of monads and effects. We develop coeffect calculus with a type system that
 tracks the context demands more precisely (by annotating the types) and we add indexing
@@
 
-In the case of resources, the additional information about the context are simply added to the
-variable context (as a products), but we will later look at contextual properties that affect
+In the case of resources, the additional information about the context is added to the
+variable context (as a product), but we will later look at contextual properties that affect
 how variables are represented. More importantly, \emph{structural coeffects} link additional
@@
 One important note about bunched typing is that it requires a different interpretation. The omission
-of weakening and contraction in linear logic means that variable can be used exactly once.
+of weakening and contraction in linear logic means that variable must be used exactly once.
 In bunched typing, variables may still be duplicated, but only using the ``;'' separator.

### 2017-01-17 responses/response.tex
Edits pages 31-94

@@
 \paragraph{Resolution in Haskell.}
-In Haskell, the implicit parameters from the \emph{declaration-site} are preferred and so the result
+In Haskell, the implicit parameters from the \emph{declaration site} are preferred and so the result
 is 10. Correspondingly, the type system only permits one typing -- with the parameter \ident{?x}
@@
 context, but do not explicitly specify where the context comes from. In case of implicit parameters,
-the values can be provided either by the call site or by the declaration-site.
+the values can be provided either by the call site or by the declaration site.
 

### 2017-01-17 main.tex
Edits pages 31-94

@@
   programming languages (Chapter~\ref{ch:semantics}). The semantics is inspired by a categorical
-  model based on \emph{indexed comonads} and it translates source context-aware program into a
-  target program in a simple functional language with comonadically-inspired primitives. We give
+  model based on \emph{indexed comonads} and it translates a source context-aware program into a
+  target program in a simple functional language with comonadically-inspired primitives. We give a
   concrete definition of the primitives for dataflow, implicit parameters and liveness and
@@
   (about the coeffect calculus) and concrete definitions and properties (about concrete
-  context-awre language). The Chapter~\ref{ch:structural} mirrors the same development for
+  context-awre language). Chapter~\ref{ch:structural} mirrors the same development for
   \emph{structural coeffect systems}.
@@
 \textsc{Categorical}
-  & \hspace{-0.5em}\begin{tabular}{l} Indexed comonads \\[-0.3em] (Sectiuon~\ref{sec:semantics-flat-idx}) \end{tabular}
+  & \hspace{-0.5em}\begin{tabular}{l} Indexed comonads \\[-0.3em] (Section~\ref{sec:semantics-flat-idx}) \end{tabular}
   & \hspace{-0.5em}\begin{tabular}{l} Examples including indexed \\[-0.3em] product, list and maybe \\[-0.3em] comonads (Section~\ref{sec:semantics-flat-monoidal}) \end{tabular} \\

### 2017-01-17 text/Chapter02.tex
Edits pages 31-94

@@
 \ident{y}. In static binding, the context demands of the body can be placed on the scope in
-which the function is defined (declaration-site). In dynamic binding, the requirements are
+which the function is defined (declaration site). In dynamic binding, the requirements are
 \emph{delayed} and are placed on the scope in which the function is called (call site).
@@
 dynamic binding when needed}''. If an implicit parameter is available in static scope, then the value
-is statically bound and the context requirement is satisfied using the declaration-site context.
+is statically bound and the context requirement is satisfied using the declaration site context.
 Otherwise, the context requirement is delayed and has to be satisfied at the call site.
@@
 clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
-a function can be satisfied using the context available at declaration-site or using the context
+a function can be satisfied using the context available at declaration site or using the context
 available at the call site. Recall our earlier printer example -- a language that
@@
 We aim to find a description of context-aware languages that does not make ad-hoc decisions
-about how context demands are split between the declaration-site and the call site. While
+about how context demands are split between the declaration site and the call site. While
 Haskell's solution for implicit parameters might be the most reasonable one for that particular

### 2017-01-17 text/Chapter03.tex
Edits pages 31-94

@@
 \begin{itemize}
-\item We characterize contextual properties -- the Section~\ref{sec:applications-structure}
+\item We characterize contextual properties -- Section~\ref{sec:applications-structure}
   explains what is a \emph{coeffect} and contrasts it with a better known notion of
@@
 \label{sec:applications-structure-lam}
-
-The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
-The typical lambda abstraction rule for effect systems looks as (\emph{effect}) in
-Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how
-the effect analysis works as follows:
+As outlined in Section~\ref{sec:intro-theory}, the difference between effects and coeffects becomes 
+apparent when we consider lambda abstraction. The typical lambda abstraction rule for effect systems 
+looks as (\emph{effect}) in Figure~\ref{fig:applications-abs}. Wadler and 
+Thiemann~\cite{monads-effects-marriage} explain how the effect analysis works as follows:
 %
@@
 machine (\eg~current time), we can transfer the function with a context requirement and \emph{rebind}
-the resource. However, if the resource is not available (\eg.~GPS on the server), we need to a
+the resource. However, if the resource is not available (\eg.~GPS on the server), we need to 
 capture \emph{remote reference}.
@@
 
-We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
+We look at resources in distributed programming in more detail in Section~\ref{sec:applications-flat-distr}.
 The important point here is that in flat coeffect systems, contextual requirements are
@@
 In Section~\ref{sec:intro-why-array}, we used an example of tracking array access patterns. For every
-variable, the additional coeffect annotation keeps a range of indices that may be accessed relatively
-to the current cursor. For example, consider an expression
+variable, the additional coeffect annotation keeps a range of indices, relative to the current cursor,
+that may be accessed. For example, consider an expression
 $x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$.
@@
 denoting that we access only the current value in $x$, but we need access to both left and right
-neighbours in the $y$ array. In order to unify the flat and structural notions, we attach this information
-as a \emph{vector} of annotations associated with a \emph{vector} of variable and write:
+neighbours in the $y$ array. The context annotation is written as:
 $\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
-The unification is outlined in Section~\ref{sec:further-unified}.
+
+Note that we attach the per-variable information as a \emph{vector} of annotations 
+associated with a \emph{vector} of variables, which makes it possible to treat flat and structural
+coeffects uniformly (Section~\ref{sec:struct-semantics} and \ref{sec:semantics-theory}) and unify
+them into a single system (Section~\ref{sec:further-unified}), which is capable of tracking both
+flat per-context coeffects and structural per-variable coeffects. Attaching structural coeffect
+annotations directly to individual variables would simplify some aspects of our formalism; we choose
+not to follow that approach as we envision that a fully unified theory of coeffects (which we 
+attempt to approach in this thesis) would be able to capture variables, flat and structural coeffects 
+using just a single mechanism.
 
@@
 
-Before looking at concrete flat and structural systems, we briefly overview
-some notation used in this thesis. Structural coeffects keep annotations as \emph{vectors} and
-use a number of operations related to scalars and vectors.
-
 % --------------------------------------------------------------------------------------------------
@@
 
-Structural coeffects have one annotation per each variable. Thus, the annotation consists
+Structural coeffects have one annotation per variable. Thus, the annotation consists
 of multiple values -- one belonging to each variable. To distinguish between the overall annotation
@@
 scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
- $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
+ $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$
+ and $\aclrd{\textbf{r}}\circ\cclrd{s} = \alift{ \cclrd{r_1}\circ\cclrd{s}, \ldots, \cclrd{r_n}\circ\cclrd{s}}$.
 Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
@@
 
-In flat coeffect systems, the additional contextual information are independent of lexically scoped
+In flat coeffect systems, the additional contextual information is independent of lexically scoped
 variables. As such, flat coeffects capture properties where the execution environment provides some
@@
 $\ident{?size}$:
-
+%
 \begin{equation*}
@@
 write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
-We also assume that are implicit parameters are of type $\ident{num}$, so the annotation can be a
+We also assume that implicit parameters are of type $\ident{num}$, so the annotation can be a
 simple set of names (rather than mapping from names to types). In a purely dynamically scoped system,
@@
 of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are
-finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
+sets of names, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
 \ident{Names}$. The expressions include \ident{?param} to read implicit
@@
 language. In the type system, the (\emph{abs}) rule places context-requirements on both the
-declaration site and the call site. For implicit parameters, this rule introduces non-determinism
-in the type-inference, because the parameters can be split arbitrarily. However, as we show in the
-next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two
+declaration site and the call site. For implicit parameters, this rule means there the system does
+not have the pincipal type property, because the parameters can be split arbitrarily. As we show in 
+the next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two
 parts of context (implicit parameter dictionaries) that are provided by the call site and
@@
 Chapter~\ref{ch:intro}. This section explores the use case. We look at rebindable resources and
-cross-compilation. The structure of both is very similar to implicit parameters and type
-class constraints, but they demonstrate that there is a broader use for coeffect systems.
+cross-compilation. Both of these could be seen as an instance of implicit parameters, but we
+present them separately to illustrate other forms that coeffect systems can have. The examples
+given later (\eg~in Section~\ref{sec:applications-flat-dataflow}) show coeffects that cannot be
+easily seen as implicit parameters.
 
@@
 \begin{array}{l}
- \kvd{let}~\ident{recentNews} = \lambda () \rightarrow\\[-0.25em]
+ \kvd{let}~\ident{localNews} = \lambda () \rightarrow\\[-0.25em]
  \quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\[-0.25em]
- \quad\ident{query}~\ident{db}~\str{SELECT * WHERE Date > \%1}~(\kvd{access}~\ident{Clock})
-\end{array}
-\end{equation*}
-%
-When \ident{recentNews} is created on the server and sent to the client, a remote reference to
+ \quad\ident{query}~\ident{db}~\str{SELECT * WHERE Location = \%1}~(\kvd{access}~\ident{GPS})
+\end{array}
+\end{equation*}
+%
+When \ident{localNews} is created on the server and sent to the client, a remote reference to
 the database (available only on the server) must be captured. If the client device supports a
-clock, then \ident{Clock} can be locally \emph{rebound}, \eg, to accommodate time-zone changes.
-Otherwise, the date and time needs to be obtained from the server too.
-
-~
+GPS, then \ident{GPS} can be locally \emph{rebound} to provide news for the reader's area.
+Otherwise, the default location needs to be obtained from the server.
 
@@
 Lambda abstraction splits the requirements between declaration site (capturing remote reference)
-and call site (representing rebinding). For this reason, we do not discuss the system in details
+and call site (representing rebinding). For this reason, we do not discuss the system in detail
 and instead look at other uses.
@@
 \paragraph{Cross-compilation.}
-A related issue with distributed programming is the need to target increasing number of diverse
+A related issue with distributed programming is the need to target an increasing number of diverse
 platforms. Modern applications often need to run on multiple platforms (iOS, Android, Windows or
@@
 A single program is then automatically split and compiled to multiple target runtimes. This
-posses additional challenges -- it is necessary to check where each part of the program can run
+poses additional challenges -- it is necessary to check where each part of the program can run
 and statically guarantee that it will be possible to compile code to the required target
@@
 
-Subcoeffecting and application are trivially dual to the rules for implicit parameters. We just track
+Subcoeffecting and application are the same as the rules for implicit parameters. We just track
 supported environments using intersection as opposed to tracking required parameters using union.
@@
 \begin{itemize}
-\item The function is represented as executable code for an call site environment
+\item The function is represented as executable code for a call site environment
   and is executed there, possibly after it is marshalled and transferred to another machine.
@@
 variables are used. Nevertheless, we still look at the left-hand side of $\vdash$ and the structure
-of the typing rules and semantics will be very similar.
-
-~
+of the typing rules and the semantics will be very similar.
 
@@
 propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
-guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variables
-to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look
+guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variable
+uses to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look
 at an example with conditionals:
@@
 A type system that captures whole-context liveness annotates the context with value of a
-two-point lattice $\mathcal{L} = \{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$. The annotation \cclrd{\ident{L}} marks
-the context as \emph{live} and \cclrd{\ident{D}} stands for a \emph{dead} context.
-Figure~\ref{fig:applications-flat-livealg} (a) defines the ordering $\cclrd{\sqsubseteq}$, meet $\cclrd{\sqcup}$ and join
-operations $\cclrd{\sqcap}$ of the lattice.
+two-point lattice $\mathcal{L} = \{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$ where 
+$\cclrd{\ident{L}} \cclrd{\sqsubseteq} \cclrd{\ident{D}}$ (Figure~\ref{fig:applications-flat-livealg} (a)). 
+The annotation \cclrd{\ident{L}} marks the context as \emph{live} and \cclrd{\ident{D}} stands 
+for a \emph{dead} context.
 
@@
 \paragraph{Summary.}
-Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce
-non-determinism. It simply duplicates the context demands. However, this still matches the
+Unlike with implicit parameters, lambda abstraction for liveness analysis has the principal types
+property. It simply duplicates the context demands. However, this still matches the
 property of coeffects that impurities cannot be delayed or thunked and attached just to the
@@
 
-The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an
-operation that accesses a context. In case of arrays, the context is neighbourhood of a current
+Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an
+operation that accesses a context. In case of arrays, the context is the neighbourhood of a current
 location in the array specified by a cursor. In this section, we make the example more concrete,
@@
 
+\paragraph{Algebra of annotations.}
+The coeffect annotations used in the liveness example in Section~\ref{sec:applications-flat-live}
+form a two-point lattice. The annotations for dataflow do not form a lattice (as the absorption
+laws do not hold for addition and maximum), but the structure resembles that of \emph{semiring}.
+As discussed later in Section~\ref{sec:flat-calculus-algebra}, the coeffect annotations for many 
+of our examples form a semiring, but this is not the case in general.
+
+The coeffect annotations for dataflow forms a tropical semiring with $\cclrd{+}$ as the multiplication,
+$\cclrd{\textnormal{max}}$ as the addition, $\cclrd{-\infty}$ as the zero element and $\cclrd{0}$ as
+unit. In this section, we do not use the $\cclrd{-\infty}$ element, but this can be used to annotate
+unused variables as discussed in Section~\ref{sec:flat-calculus-examples}.
+
+% --------------------------------------------------------------------------------------------------
+
 \paragraph{Example.}
@@
 The language discussed in this section is a \emph{causal} dataflow language. This means that
-a computation can access \emph{past} values of the stream (but not future values). In the semantics,
+a computation can access \emph{past} values of the stream but not future values. In the semantics,
 we again need richer structure over the input.
@@
 \paragraph{Summary.}
-Type systems have been used in the context of dataflow languages, for example to check
-initialization properties \cite{app-dataflow-init}, but to our knowledge, not for checking
-the maximal number of required past values. Thus this section serves not just as an example,
-but also shows how coeffects can lead to novel results.
+Type systems have been used in the context of dataflow languages to check
+initialization properties \cite{app-dataflow-init} and resource-aware systems are capable
+of tracking how required values are accessed in dataflow programming \cite{app-pixie}.
+This section formulates the problem in terms of coeffects.
 
@@
 variables. In this section, we briefly consider a system for safe locking of Flanagan and
-Abadi~\cite{app-safe-locking} as one, more advanced example. Calculus of capabilities of
+Abadi~\cite{app-safe-locking} as one, more advanced example. The calculus of capabilities of
 Cray et al.~\cite{app-capabilities} is discussed later in Section~\ref{sec:applications-active-ccc}.
@@
 interesting from a theoretical perspective, but it is not practically useful. Here, we
-revisit the problem and define a structural system that tracks liveness per-variable.
+revisit the problem and define a structural system that tracks liveness of individual variables.
 
@@
 The structural typing rules are shown in Figure~\ref{fig:applications-struct-live} (b). They mirror
-the rules know from substructural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
+the rules known from substructural type systems (Section~\ref{sec:path-logic}). Weakening (\emph{weak})
 extends the context with a single unused variable $x$ and adds the $\cclrd{\ident{D}}$ annotation to the
@@
 the type is indexed by the corresponding annotation:
-
+%
 \begin{equation*}
@@
 \noindent
-Note that the product of the free variables is not an ordinary tuple of our language, but a special
-construction (we return to this topic in Section~\ref{sec:struct-semantics}). This follows from the
+Note that the product of the free variables is not a tuple of our language, but a special
+construction used only in the semantics. This follows from the
 asymmetry of $\lambda$-calculus, as discussed in Section~\ref{sec:applications-strucutre-vec}. Functions
 take just a single input and so they are interpreted in the same way as in flat calculus:
-
-\vspace{-0.5em}
+%
 \begin{equation*}
@@
 \end{equation*}
-\vspace{-0.5em}
-
-\noindent
+%
 The rules that define the semantics are shown in Figure~\ref{fig:applications-struct-livesem}.
-To make the definition simpler, we are somewhat vague when working with products. We write
-variables of product type such as $\mathbf{v}$ in bold-face and individual values like $x$ in
-normal face. We freely re-associate products and so $(\mathbf{v}, x)$ should not be seen as a
+To make the definition simpler, we treat the product $\tau'_1 \times \ldots \times \tau'_n$ as a
+flat list. Variables of product type are written in bold-face as $\mathbf{v}$ and individual values 
+are written in normal face as $x$. An expression $(\mathbf{v}, x)$ should not be seen as a
 nested product, but simply as a product containing all variables from the product $\mathbf{v}$
@@
 annotations. In a number of ways, the system is simpler than the flat coeffect calculi. In
-lambda abstraction, we simply annotate function with the annotation of a matching variable
+lambda abstraction, we simply annotate a function with the annotation of a matching variable
 (this rule is the same for all upcoming systems). In application, the \emph{pointwise} composition
@@
 how to build the context required by the sub-expressions and so the semantics implements
-transformation from the context in the (typing) conclusion to the (typing) assumption. How
-should coeffects be understood, in general, is discussed further in Section~\ref{sec:flat-calculus-understanding}.
+transformation from the context in the (typing) conclusion to the (typing) assumption. The two
+ways of understanding coeffects are discussed further in Section~\ref{sec:flat-calculus-understanding}.
 
@@
 Liveness analysis checks whether a variable is used or unused. With structural coeffects, we can go
-further and track how many times is the variable accessed. Girard et al. \cite{logic-bounded} coined
-this idea as \emph{bounded linear logic} and use it to restrict well-typed programs to
+further and track how many times is the variable accessed. This has been captured in \emph{bounded 
+linear logic} by Girard et al. \cite{logic-bounded} who use it to restrict well-typed programs to
 polynomial-time algorithms. We first introduce the system in our, coeffect, style and then
@@
 As discussed in Chapter~\ref{ch:structural}, the system preserves type and coeffect annotations under
-the $\beta$-reduction. Reducing the expression in this case gives $x + (x+y) + (x+y)$. This has the
+$\beta$-reduction. Reducing the expression in this case gives $x + (x+y) + (x+y)$. This has the
 same bounds as the original expression -- $x$ is used three times and $y$ twice.
@@
 The type system in Figure~\ref{fig:applications-struct-bll} annotates contexts with vectors of integers.
-The rules have the same structure as those of the system for liveness analysis. The only difference is
-how annotations are combined. Here, we use integer multiplication ($\ast$) for sequential composition
-and addition ($+$) for point-wise composition.
+The rules have the same structure as those of the system for liveness analysis and the annotations
+form a semiring with integer multiplication ($\ast$) for sequential composition and addition 
+($+$) for point-wise composition.
 
@@
 
-As a result, BLL does not need to attach additional annotation to the variable context
+As a result, BLL does not need to attach additional annotations to the variable context
 as a whole. The requirements are attached to individual variables and so our context
@@
 rather than attaching a \emph{vector} of annotations to the entire context. One benefit
-of our approach is that it lets us unify flat and structural systems (Chapter~\ref{ch:further-unified}).
+of our approach is that it lets us unify flat and structural systems (Section~\ref{sec:further-unified}).
 
@@
 past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as
-a \emph{flat} property, which gives us a system that keeps the same number of past values for
+a \emph{flat} property, which gives us a system that keeps an upper bound on past values for
 all variables. However, dataflow can also be adapted to a structural system which keeps the number
@@
 To evaluate a current value of the stream, we need the current value of \ident{left} and one past
-value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
+value of \ident{right}. A flat system is not able to capture this level of detail and simply
 requires $1$ past values of both streams in the variable context.
@@
 Turning a flat dataflow system to a structural dataflow system is a change similar to the one
-between flat ans structural liveness. In case of liveness analysis, we included the flat system
+between flat and structural liveness. In case of liveness analysis, we included the flat system
 only as an illustration (it is not practically useful). For dataflow, the flat system is less
@@
 The type system in Figure~\ref{fig:applications-struct-df} annotates the variable context with a
-vector of integers. This is similar as in the bounded reuse system, but the integers \emph{mean} a
+vector of integers. This is similar to the bounded reuse system, but the integers \emph{mean} a
 different thing. Consequently, they are also calculated differently. We omit rules that are the
@@
 
-The result is similar as the structure used for bounded reuse. The only difference is that, given
+The result is similar to the structure used for bounded reuse. The only difference is that, given
 a variable annotated with $\cclrd{r}$, we need $1+\cclrd{r}$ values. That is, we need the current
@@
 
-Static typed systems based on tainting have been use to prevent cross-site scripting attacks
+Type systems based on tainting have been use to prevent cross-site scripting attacks
 \cite{app-tainting-xss} and SQL injection attacks \cite{app-tainting-sql,app-tainting-wasp}.
@@
 In this example, \ident{id} must not come directly from a user input, because \ident{query} requires
-untainted string. Otherwise, the attacker could specify values such as \str{1; DROP TABLE Users}.
+an untainted string. Otherwise, the attacker could specify values such as \str{1; DROP TABLE Users}.
 The variable \ident{msg} may or may not be tainted, because it is not used in protected context
@@
 flag (for tainting) or more complex data (for provenance). In static checking, the information
-need to be associated with the variables in the variable context.
+needs to be associated with the variables in the variable context.
 
@@
 in \emph{secure information flow}. There, the aim is to guarantee that sensitive information (such
-as credit card number) cannot be leaked to contexts with low secrecy (\eg~sent via an unsecured
+as a credit card number) cannot be leaked to contexts with low secrecy (\eg~sent via an unsecured
 network channel). Volpano et al. \cite{app-secure-flow} provide the first (provably) sound type
-system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} surveys
+system that guarantees non-inference and Sabelfeld et al. \cite{app-secure-information-flow} survey
 more recent work. Information flow checking has been also integrated (as a single-purpose
@@
 Systems for secure information flow typically define a lattice of security classes $(\mathcal{S}, \leq)$
-where $\mathcal{S}$ is a finite set of classes and an ordering. For example a set $\{\cclrd{\ident{L}}, \ident{H}\}$
-represents low and high secrecy, respectively with $\cclrd{\ident{L}} \leq \ident{H}$ meaning that low security
+where $\mathcal{S}$ is a partially ordered finite set of classes. For example a set $\{\cclrd{\ident{L}}, \cclrd{\ident{H}}\}$
+represents low and high secrecy, respectively with $\cclrd{\ident{L}} \leq \cclrd{\ident{H}}$ meaning that low security
 values can be treated as high security (but not the other way round).
@@
 
-There is a number of systems that also capture contextual properties, but that make it possible to
+A number of systems capture contextual properties, but make it possible to
 \emph{change} the context -- not just by evaluating certain code block in a locally modified context
-(\eg by wrapping it in $\ident{prev}$ in dataflow), but also by calling a function that, for example,
+(\eg~by wrapping it in $\ident{prev}$ in dataflow), but also by calling a function that
 acquires new capabilities and returns those to the caller. Such actions appear to be closer to
@@
 using memory in that region and then deallocates the memory of the region:
-
+%
 \begin{equation*}
@@
 This is where systems with active context differ from systems with passive context. To type check
-such programs, we do not only need to know what context is required to call \ident{calculate}
+such programs, we not only need to know what context is required to call \ident{calculate}
 (\ie~context on the left-hand side of $\vdash$). We also need to know what effects an expression
-has on the context when it evaluates and the current context meeds to be updated after a function
+has on the context when it evaluates and the current context needs to be updated after a function
 call. This is an effectful property that would appear on the right-hand side of $\vdash$.
@@
 \emph{input} of the function. Without giving any details, the semantics generally has the
-following structure (with comonad to model coeffects on the left):
+following structure (with a comonad to model coeffects on the left):
 %
@@
 %
-Systems with active contexts require two annotations -- one that specifies the context required
-before the call is performed and one that specifies how the context changes after the call (this
-could be either a \emph{new} context or \emph{update} to the original context). Thus the structure
-of the semantics would look as follows (with comonad to model coeffects on the left and monad to
+Systems with active contexts require two annotations -- one specifies the context required
+before the call and one specifies how the context changes after the call (this
+could be a \emph{new} context or an \emph{update} to the original context). Thus the structure
+of the semantics would look as follows (with a comonad to model coeffects on the left and a monad to
 model effects on the right):
-%
+
 \begin{equation*}
@@
 
-~
-
 \paragraph{Software updating.}
 Another example of a system that uses contextual information actively is dynamic software updating
-(DSU) \cite{app-dsu-programs,app-dsu}. The DSU systems have the ability to update programs at
+(DSU) \cite{app-dsu-programs,app-dsu}. DSU systems have the ability to update programs at
 runtime without stopping them. For example, Proteus developed by Stoyle et al. \cite{app-dsu-mutatis}
@@
 
-This chapter was divided into two major sections. First, we looked at \emph{flat} systems, which
-track whole-context properties. Next, we look a \emph{structural} systems, which track per-variable
-properties. Both of the variants are useful and important -- for example, implicit parameters can
-only be expressed as \emph{flat} system, but liveness analysis is only useful as \emph{structural}.
+This chapter was divided into two sections. First, we looked at \emph{flat} systems, which
+track whole-context properties. Next, we looked at \emph{structural} systems, which track per-variable
+properties. Both are useful and important -- for example, implicit parameters can
+only be expressed as a \emph{flat} system, but liveness analysis is only useful as \emph{structural}.
 For this reason, we explore both of these variants in this thesis (Chapter~\ref{ch:flat} and
-Chapter~\ref{ch:structural}, respectively). We can, however, unify the two variants into a single
+Chapter~\ref{ch:structural}). We can, however, unify the two variants into a single
 system discussed in Section~\ref{sec:further-unified}.

### 2017-01-17 text/Chapter04.tex
Fix error
Edits pages 31-94

@@
 each concrete context-aware language is also equipped with an algorithm for choosing a unique
-typing derivation. This allows us to explore the language design landscape, while still follow
+typing derivation. This allows us to explore the language design landscape, while still following
 the usual scoping rules for languages with established approaches (e.g.~implicit parameters in
-Haskell).
+GHC).
 
@@
   in contextual lambda abstraction. Each concrete context-aware language based on the
-  coeffect calculus must specify how such ambiguities are to be resolved. We give the procedure for
-  choosing unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).
+  coeffect calculus must specify how such ambiguities are to be resolved. We give the procedures for
+  choosing unique typing derivations for our three examples (Section~\ref{sec:flat-unique}).
 
@@
 Despite the differences, the systems fit the same framework. This becomes apparent when
-we consider the categorical structure (Section~\ref{sec:flat-semantics}). Rather than starting
+we consider the categorical structure (Chapter~\ref{ch:semantics}). Rather than starting
 from the categorical semantics, we first explain how the systems can be unified syntactically
@@
 \cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows
-a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
+a similar style to that of effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
 in the structure of lambda abstraction as discussed briefly here and in Section~\ref{sec:applications-structure-lam}
@@
 %
-Note that the lambda abstraction in the syntax is written in the Church-style and requires a type
+Note that the lambda abstraction in the syntax is written in the Church style and requires a type
 annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
-unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
+unique typing derivation for context-aware computations. Using Church style lambda abstraction,
 we can directly focus on the more interesting problem of finding unique \emph{coeffect annotations}
@@
 \\
-r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
-r\; \cpar\; r = r~~~~&
+r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &r\; \cpar\; r = r~~~~&
 \czero \;\cpar\; r = r = r \;\cpar\; \czero
 \\
-r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
-r\; \czip\; r = r \qquad
+r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &r\; \czip\; r = r~~~~&~
 \\
-\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
+\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &
 t\; \cleq\; t \\[0.5em]
@@
 functions. The monoid axioms are required in order to form a category structure in the
-semantics (Section~\ref{sec:flat-semantics}).
+semantics (Section~\ref{sec:semantics-theory}).
 
@@
 upper bound and the greatest lower bounds of a lattice. For example, in dataflow computations, they
-are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
-(we discuss the lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.
+are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters.
+
+\paragraph{Semiring.}
+The $\cseq$ and $\cpar$ operations of the flat coeffect algebra together with the $\cunit$ and
+$\czero$ values resemble semiring. As required by semiring laws, both of the operations are 
+associative; $\cunit$ and $\czero$ are the units of $\cseq$ and $\cpar$, respectively, and the
+distributivity laws also hold. The two additional laws of semiring that are not required by a 
+flat coeffect algebra are symmetry of $\cseq$ and annihilation:
+%
+\begin{align*}
+\czero \, \cseq \, \cclrd{r} = \czero = \cclrd{r} \, \cseq \, \czero \qquad \cclrd{r}\,\cpar\,\cclrd{s} = \cclrd{s}\,\cpar\,\cclrd{r}
+\end{align*}
+%
+The symmetry of $\cseq$ holds for all our examples, but we do not include it so that we do not
+unnecessarily restrict the system. The annihilation holds for some of the examples (liveness
+and optimized dataflow in Section~\ref{sec:flat-calculus-examples}), but not for all of them.
+In particular, it does not hold for implicit parameters $\cclrd{\emptyset}\,\cclrd{\cup}\,\cclrd{r}\neq\cclrd{\emptyset}$
+and so requiring it would lead to a more restrictive structure than is desirable.
 
 \paragraph{Ordering.}
-
 The flat coeffect algebra includes a pre-order relation $\cleq$. This will be used to introduce
@@
 
-To guide understanding of the system, we also show non-syntax-directed (\emph{sub}) rule for
+To guide understanding of the system, we also show a non-syntax-directed (\emph{sub}) rule for
 subcoeffecting. The rule states that an expression with context demands $\cclrd{r'}$ can be
@@
 subtyping and additional constructs such as pairs in Section~\ref{sec:flat-exts}. When
-discussing procedure for choosing unique typing in Section~\ref{sec:flat-unique}, we consider
+discussing the procedure for choosing unique typing in Section~\ref{sec:flat-unique}, we consider
 only the syntax-directed part of the system.
@@
 %---------------------------------------------------------------------------------------------------
-s
+
 \subsection{Understanding flat coeffects}
@@
 (Chapter~\ref{ch:semantics}), it specifies how context is propagated. These two readings provide
-different ways of interpreting the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the
+different ways of interpreting the judgement $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the
 typing rules used to define it.
@@
 and the combined context demands of the body are split between two positions in the conclusions;
-in the (\emph{app}) rule, $\cpar$ appears in the conclusion and combines two context demands
+in the (\emph{app}) rule, $\cpar$ appears in the conclusion and merges two context demands
 from the assumptions.
@@
 track sets of names. The definition uses a set union for all three operations. Both variables and
-constants are are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
+constants are annotated with $\emptyset$ and the ordering is defined by $\subseteq$. The
 definition satisfies the flat coeffect algebra axioms because $(S, \cup, \emptyset)$ is an idempotent,
@@
 parameters, but it is also possible to define a coeffect calculus that uses mappings from names to
-types (care is needed to avoid assigning multiple types to a parameter of the same type).
+types.
 
@@
 the meet operation $\sqcap$ and pointwise composition $\cpar$ is modelled by join $\sqcup$.
-The two-point lattice is a commutative, idempotent monoid. Distributivity
+The two-point lattice induces a commutative, idempotent monoid. Distributivity
 $(r \sqcup s) \sqcap t = (r \sqcap t) \sqcup (s \sqcap t)$ does not hold for \emph{every}
@@
 and $(\mathbb{N}, \mathit{min})$ forms a band. Note that dataflow is our first example where
-$\cseq$ is not idempotent. The distributivity axioms require the following to be the case:
-$\mathit{max}(r,s) + t = \mathit{max}(r+t, s+t)$, which is easy to see.
+$\cseq$ is not idempotent. The distributivity axioms require addition to distribute over maximum,
+which is easy to see.
 
 A simple dataflow language includes an additional construct $\kvd{prev}$ for accessing the
-previous value in a stream with an additional typing rule that look as follows:
+previous value in a stream with an additional typing rule that looks as follows:
 %
@@
 \begin{example}[Optimized dataflow]
-In optimized dataflow, context is annotated with natural numbers extended with the $\bot$ element,
-that is $\mathbb{N}_{\bot} = \{\bot, 0, 1, 2, 3, \ldots \}$ such that $\forall n \in \mathbb{N}. \bot \leq n$.
-The flat coeffect algebra is $(\mathbb{N}_{\bot}, +, \mathit{max}, \mathit{min}, 0, \bot, \leq)$
-where $m + n$ is $\bot$ whenever $m=\bot$ or $n=\bot$ and \emph{min}, \emph{max} treat $\bot$ as the
+In optimized dataflow, context is annotated with natural numbers extended with the $-\infty$ element.
+The flat coeffect algebra is $(\mathbb{N} \cup \{-\infty\}, +, \mathit{max}, \mathit{min}, 0, \bot, \leq)$
+where $m + n$ is $-\infty$ whenever $m=-\infty$ or $n=-\infty$ and \emph{min}, \emph{max} treat $-\infty$ as the
 least element.
@@
 \noindent
-Note that $(\mathbb{N}_{\bot}, +, 0)$ is a monoid for the extended definition of $+$; for the
+Note that $(\mathbb{N}\cup\{-\infty\}, +, 0)$ is a monoid for the extended definition of $+$; for the
 bottom element $0 + \bot = \bot$ and for natural numbers $0 + n = n$. The structure
 $(\mathbb{N}, \emph{max}, \bot)$ is also a monoid, because $\bot$ is the least element and so
-$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band (the extended
-\emph{min} is still idempotent and associative) and the distributivity axioms also hold
-for $\mathbb{N}_{\bot}$.
+$\mathit{max}(n, \bot) = n$. Finally,  $(\mathbb{N}, \emph{min})$ is a band and the distributivity 
+axioms also hold for $\mathbb{N}\cup\{-\infty\}$.
 
@@
 
-As discussed in Chapter~\ref{sec:applications}, the lambda abstraction rule for coeffect systems
+As discussed in Chapter~\ref{ch:applications}, the lambda abstraction rule for coeffect systems
 differs from the rule for effect systems in that it does not delay all context demands.
 In case of implicit parameters (Section~\ref{sec:applications-flat-impl}), the demands
-can be satisfied either by the call site or by the declaration-site. In case of dataflow and
+can be satisfied either by the call site or by the declaration site. In case of dataflow and
 liveness, the rule discussed in Section~\ref{sec:flat-calculus} reintroduces similar ambiguity
@@
 
-Furthermore, the semantics of context-aware languages in Chapter~\ref{sec:applications} and
-also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivation} and so the meaning
+Furthermore, the semantics of context-aware languages in Chapter~\ref{ch:applications} and
+also in Chapter~\ref{ch:semantics} is defined over \emph{typing derivations} and so the meaning
 of a program depends on the typing derivation chosen. In this section, we specify how to choose
@@
 
-The most interesting case is that of implicit parameters. For example, consider the following
-program written using the coeffect calculus with implicit parameter extensions:
+The most interesting case is that of implicit parameters. Consider the following
+program written using the extended coeffect calculus:
 %
@@
 \begin{array}{l}
-\kvd{let}~f~= (\kvd{let}~\ident{?x}=42~\kvd{in}~(\lambda y.~\ident{?x}))~\kvd{in} \\[-0.2em]
-\kvd{let}~\ident{?x}=666~\kvd{in}~f~0 \\[-0.2em]
+\kvd{let}~\ident{?x}=\num{1}\\[-0.2em]
+\kvd{let}~f~= \lambda y.~\ident{?x}~\kvd{in} \\[-0.2em]
+\kvd{let}~\ident{?x}=2~\kvd{in}~f~0 \\[-0.2em]
 \end{array}
@@
   \item $f : \ident{num} \xrightarrow{\cclrd{\emptyset}} \ident{num}$ -- in this case, the value
-    of \ident{?x} is captured from the declaration-site and the program produces $1$.
+    of \ident{?x} is captured from the declaration site and the program produces $1$.
   \item $f : \ident{num} \xrightarrow{\cclrd{\{ \ident{?x} \}}} \ident{num}$ -- in this case, the
@@
 %
-The coeffect calculus intentionally allows both of the options, acknowledging the fact that the
+The coeffect calculus intentionally allows both options, acknowledging the fact that the
 choice needs to be made for each individual concrete context-aware programming language. In the
@@
 In this section, we discuss the specific choices of typing derivation for implicit parameters,
-dataflow and liveness. We use the fact that the coeffect calculus uses Church-style syntax for
+dataflow and liveness. We use the fact that the coeffect calculus uses Church style syntax for
 lambda abstraction giving a type annotation for the bound variable. This does not affect the
@@
 
-For implicit parameters we choose to follow the behaviour implemented by Haskell~\cite{app-implicit-parameters}
-where function abstraction captures all parameters that are statically available at the declaration-site
+For implicit parameters we choose to follow the behaviour implemented by GHC~\cite{app-implicit-parameters}
+where function abstraction captures all parameters that are statically available at the declaration site
 and places all other demands on the call site. For the example above, this means
-that the body of $f$ captures the value of \ident{?p} available from the declaration-site
+that the body of $f$ captures the value of \ident{?p} available from the declaration site
 and $f$ will be typed as a function requiring no parameters (coeffect $\cclrd{\emptyset}$). The
-program thus evaluates to a numerical value $42$.
+program thus evaluates to a numerical value $1$.
 
@@
 %
-Here, $\cclrd{\Delta}$ is a set of implicit parameters that are in scope at the declaration-site.
+Here, $\cclrd{\Delta}$ is a set of implicit parameters that are in scope at the declaration site.
 The modified typing rules are shown in Figure~\ref{fig:flat-resolve-impl}. The rules (\emph{var}),
@@
 \paragraph{Properties.}
-If a program written in a coeffect language with implicit parameters is well-typed in a type
+If a program written in a coeffect language with implicit parameters is well-typed in the type
 system presented in Figure~\ref{fig:flat-resolve-impl} then this identifies the unique preferred
-derivation for the program. We use this unique typing derivation to give the semantics of coeffect
+derivation for the program. We use this unique typing derivation to give the semantics of a coeffect
 language with implicit parameters in Chapter~\ref{ch:semantics} and we also implement this algorithm
@@
 
-The type system is more restrictive than the fully general one and it reject certain programs that
+The type system is more restrictive than the fully general one and it rejects certain programs that
 could be typed using the more general system. This is expected -- we are restricting the fully
@@
 Finally, we note that unique typing derivations obtained using the type system given in
-Figure~\ref{fig:flat-resolve-impl} are valid typing derivation under the original flat coeffect
+Figure~\ref{fig:flat-resolve-impl} are valid typing derivations under the original flat coeffect
 type system in Figure~\ref{fig:flat-types}.
@@
 
-Before discussing the syntactic properties of general coeffect calculus formally, it should be
+Before discussing the syntactic properties of the general coeffect calculus formally, it should be
 clarified what is meant by providing a ``pathway to operational semantics'' in this section. We do
@@
 Writing $\subst{e_1}{x}{e_2}$ for a standard capture-avoiding syntactic substitution, the following
-equations define four syntactic reductions on the terms:
+equations define three syntactic reductions on the terms:
 %
@@
 knowledge of the specific notion of context. If the reductions preserve the type of the expression
-(type preservation), then operational semantics can be defined as a repeated application of the
+(type preservation), then an operational semantics can be defined as a repeated application of the
 rules, together with additional domain-specific rules for each context-aware language, until
@@
 rules and then we focus on call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name
-(Section~\ref{sec:flat-syntax-cbn}) in more details.
+(Section~\ref{sec:flat-syntax-cbn}) in more detail.
 
@@
 parameters system. This means that certain coeffect systems support call-by-name evaluation strategy
-and could be embedded in purely functional language (such as Haskell).
+and could be embedded in a purely functional language (such as Haskell).
 
@@
 As discussed in Section~\ref{sec:flat-syntax-props}, the call-by-name strategy does not \emph{in
-general} preserve the type of a terms in coeffect calculi, but it does preserve the typing in two
+general} preserve the type of a term in coeffect calculi, but it does preserve the typing in two
 interesting cases.
@@
 For top-pointed coeffect algebra (using Lemma~\ref{thm:cbn-substitution-top}), the proof is similar
-to the one in Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
+to that of Theorem~\ref{thm:cbv-reduction}, using the facts that $\cclrd{s} \,\cleq\, \cunit$
 and $r \,\czip\, t \;=\; r \,\cpar\, t$.
@@
 application yields context demands $\cclrd{r}\,\cpar\,\cclrd{s}$. In order to recover the
-original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. Note that the derivation
+original typing, this must be equal to $\cclrd{r}\,\czip\,\cclrd{s}$. The derivation
 shows just one possible typing -- the expression $\lambda x.e~x$ has other types -- but
-this is sufficient for type preservation.
+this suffices.
 \end{proof}
@@
 In summary, flat coeffect calculi do not \emph{in general} permit call-by-name evaluation, but
-there are several cases where call-by-name evaluation can be used. Among the examples we discuss,
-these include liveness and implicit parameters. Moreover, for implicit parameters the
+there are several cases where call-by-name evaluation can be used. 
+These include liveness and implicit parameters. Moreover, for implicit parameters the
 $\eta$-expansion holds as well, giving us both local soundness and local completeness as coined
@@
 %
-This is one possible derivation, but other derivations may be valid for concrete coeffect system.
+This is one possible derivation, but other derivations may be valid for concrete coeffect systems.
 The design decision of using this particular derivation for the typing of \kvd{let} is motivated by
 the fact that the typing obtained using the special rule is more precise for all the examples
-consider in this chapter. To see this, assume an arbitrary splitting
-$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Table~\ref{tab:flat-simplelet}
+considered in this chapter. To see this, assume an arbitrary splitting
+$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. Figure~\ref{fig:flat-simplelet}
 shows the coeffect annotation derived from $(\lambda x.e_2)~e_1$, the coeffect annotation obtained
@@
 
-\begin{table}
+\begin{figure}
 \begin{center}
@@
 \figcaption{Simplified coeffect annotation for let binding in three flat calculi instances}
-\label{tab:flat-simplelet}
-\end{table}
+\label{fig:flat-simplelet}
+\end{figure}
 
@@
 different examples. However, for all our systems, the simplified annotation (right column in
-Table~\ref{tab:flat-simplelet}) is more precise than the original (left column). Recall that
-$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following inequalities hold:
+Table~\ref{fig:flat-simplelet}) is more precise than the original (left column). Recall that
+$\cclrd{s} = \cclrd{s_1}\,\czip\,\cclrd{s_2}$. The following holds:
 %
@@
 \noindent
-The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
+The practical consequence of Remark~\ref{thm:flat-alt-abs} is that, for strictly
 oriented coeffect calculi (such as our liveness and dataflow computations), the (\emph{idabs})
@@
 sub-expressions. Finally, the (\emph{proj}) rule is uninteresting, because $\pi_i$ can be
-viewed as a pure unary function.
+viewed as a unary function.
 
 \paragraph{Properties.}
-Pairs and the unit value in a lambda calculus typically form a monoid. Assuming $\simeq$ is an
-isomorphism that performs appropriate transformation on values, without affecting other
-properties (here, coeffects) of the expressions. The monoid axioms then correspond to
+Pairs and the unit value in a lambda calculus typically form a monoid up to isomorphism. Assuming 
+$\simeq$ is an isomorphism that performs appropriate transformation on values, without affecting other
+properties (here, coeffects) of the expressions, the monoid axioms then correspond to
 the requirement that $(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)$ (associativity) and the
@@
 capture three specific systems, namely liveness, dataflow and implicit parameters. However,
-the system is more general and an capture various other applications outlined in
+the system is more general and can capture various other applications outlined in
 Section~\ref{sec:applications-flat}.
@@
 abstraction. The body of a function requires certain context, but the context can be often
-provided by either the declaration-site or the call site. Resolving this ambiguity has to be
+provided by either the declaration site or the call site. Resolving this ambiguity has to be
 done differently for each concrete coeffect system, depending on its specific notion of context.
@@
 Finally, we introduced the equational theory for flat coeffect calculus. Although each
-concrete instance of flat coeffect calculus models different notion of context, there are
+concrete instance of flat coeffect calculus models a different notion of context, there are
 syntactic properties that hold for all flat coeffect systems satisfying certain additional

### 2017-01-17 text/Chapter05.tex
Edits pages 31-94

@@
 $\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) is interpreted as a curried
-function that takes the declaration-site context and a function argument, merges them using
+function that takes the declaration site context and a function argument, merges them using
 $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ and passes the result to the semantics of the body $f$.
@@
 For the semantics of implicit parameter programs that we consider, the preference of call site
-bindings over declaration-site bindings in (\emph{merge}) does not matter. The unique typing
+bindings over declaration site bindings in (\emph{merge}) does not matter. The unique typing
 derivations for implicit parameter coeffects obtained in Section~\ref{sec:flat-unique} always

### 2017-01-17 text/Chapter01.tex
Edits pages 95-124

@@
 \paragraph{The thesis.}
-The key claim of this thesis is that the theory of \emph{coeffects} that we develop provides
-a unified way of capturing how programs rely on the context, or execution environment, 
-in which they execute. Perhaps most importantly, coeffects provide a single conceptual framework 
-that can capture a wide range of previously disconnected notions of context-awareness.
+Our key claim is that the theory of \emph{coeffects} that we develop in this thesis provides
+a uniform and less error-prone way of writing programs that rely on the context, or
+execution environment, in which they run. Perhaps most importantly, coeffects provide a single 
+conceptual framework that can capture a wide range of previously disconnected notions of 
+context-awareness and thus make understanding of context easier.
 
@@
   context-aware programs into programs in a simple functional language. We prove that well-typed
-  context-aware programs are translated to programs that ``do not go wrong''. The development
+  context-aware programs are translated to programs that ``do not get stuck''. The development
   is repeated for structural coeffects in Chapter~\ref{ch:structural}.

### 2017-01-17 text/Chapter05.tex
Edits pages 95-124

@@
   and preservation proofs, showing that well-typed programs produced by translation from two
-  coeffect languages do not go wrong (Section~\ref{sec:semantics-proofs})
+  coeffect languages do not ``get stuck'' (Section~\ref{sec:semantics-proofs}).
 
@@
 
-The key claim of this thesis is that writing context-aware programs using coeffects is easier and
-less error-prone. In this chapter, we substantiate the claim by showing that programs written in the
-coeffect calculus and evaluated using the translation provided here do not ``go wrong''.
+In Chapter~\ref{ch:intro}, we claim that coeffects make writing context-aware programs
+less error-prone. In this chapter, we substantiate this claim by showing that programs written 
+in the coeffect calculus and evaluated using the translation provided here do not ``get stuck''.
 
@@
 \begin{array}{l}
-\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}`\rightarrow`\\[-0.2em]
+\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
  \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
@@
 assignment to the map). In that case, the type of \ident{add} is just $\ident{int}\rightarrow\ident{int}$
-and so the user does not know which implicit parameters it will need.
+and so there is no static account of which implicit parameters it will need.
 
@@
 
-In summary, a coeffect type system, captures certain runtime
+In summary, a coeffect type system captures certain runtime
 demands of context-aware programs and (as we show in this chapter), eliminates common errors
@@
 
-The correspondence between effect system and monads has been pointed out by Wadler and Thiemann
+The correspondence between effect systems and monads has been pointed out by Wadler and Thiemann
 \cite{monads-effects-marriage} and further explored by Atkey \cite{monads-parameterised-notions}
-and Vazou and Leijen \cite{monads-effects-remarriage}). This line of work relates effectful
+and Vazou and Leijen \cite{monads-effects-remarriage}. This line of work relates effectful
 functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations
@@
 meaning. This can be achieved in a number of ways. First, we can prove the \emph{coherence}
-\cite{semantics-bltres} and show the morhphisms assigned to multiple typing derivations are equivalent.
+\cite{semantics-bltres} and show the morphhisms assigned to multiple typing derivations are equivalent.
 Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
@@
 for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models
-failures (the Maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list
+failures (the Maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (set
 monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$
@@
 monadic computations -- sequential composition and strength for lifting free variables into the
-body of computation under a lambda abstraction.
+body of a computation under a lambda abstraction.
 
@@
 require the user to define complex systems of lifting to be composable. Consequently, they are usually used
-for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
-capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
+for capturing different classes of effects (exceptions, non-determinism, state), but not for
+capturing individual effects (\eg~a set of memory regions that may be accessed by a
 stateful computation).}.
@@
 \begin{compactitem}
-\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
+\item $C$ is a functor (mapping types to types) $C : \catc \rightarrow \catc$
 \item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$
@@
 From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
-\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries
-additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function
+\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value $\ctyp{}{\alpha}$ 
+that carries additional context. The $\ident{cobind}$ operation turns a context-dependent function
 $\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
@@
 \begin{example}[Non-empty list]
-A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
+Non-empty lists are recursive data types $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
 We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The
@@
 $(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
-We assume that, all implicit parameters have the type $\ident{num}$. The data type $\ctyp{\cclrd{r}}{\alpha}
+We assume that all implicit parameters have the type $\ident{num}$. The data type $\ctyp{\cclrd{r}}{\alpha}
 = \alpha \times (\cclrd{r} \rightarrow \ident{num})$ represents a value $\alpha$ together with a function that
@@
 value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to
-restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in oder to get
+restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in order to get
 implicit parameters required by the argument of $f$ and by the resulting computation, respectively
@@
 To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions}
-require \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
+require a \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
 %
@@
 \begin{example}[Monoidal comonads]
-Just like indexed comonads generalize co\-monads, the additional structure of
+Just as indexed comonads generalize co\-monads, the additional structure of
 flat indexed comonads generalizes the symmetric semimonoidal comonads of Uustalu
@@
 of $f$, which itself requires $m$ past values. When combining two lists, $\ident{merge}_{m,n}$
-behaves as \emph{zip} and produces a list of length of the shorter argument. When splitting a list,
+behaves as \emph{zip} and produces a list as long as the shorter argument. When splitting a list,
 $\ident{split}_{m, n}$ needs the maximum of the lengths.
@@
 may have different meaning (\eg~when working with implicit parameters) and so the semantics is
-defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a semantics to
+defined over a \emph{typing derivation} rather than over a \emph{term}. To assign a semantics to
 a term, we need to choose a particular typing derivation. The algorithm for choosing a unique
@@
 Chapter~\ref{ch:applications}. We use the family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed)
-data-type that wraps the product of free variables of the context and the arguments of functions:
+data type that wraps the product of free variables of the context and the arguments of functions:
 %
@@
 and wrap it with context required by the function $e_1$ by applying $\ident{cobind}_{\cclrd{s}, \cclrd{t}}$.
-The \ident{app} operation than applies the function (first element) on the argument (second element).
+The \ident{app} operation then applies the function (first element) on the argument (second element).
 Finally, numbers (\emph{num}) become constant functions that ignore the context.
@@
 Although the notion of indexed comonads presented in the previous section is novel and interesting
-in its own, the main reason for introducing it is that we can view it as a translation that provides
+on its own, the main reason for introducing it is that we can view it as a translation that provides
 embedding of context-aware domain-specific languages in a simple target functional language.
@@
 
-A context-aware \emph{source} program written using a concrete context-aware domain-specific
+A \emph{source} program written using a concrete context-aware domain-specific
 language (capturing dataflow, implicit parameters or other kinds of context awareness) with
@@
 a small functional language consisting of:
-%
+
 \begin{itemize}
@@
   \item Additional primitives that model contextual operations of each concrete coeffect language
-    (\emph{prev} for the \kvd{prev} keyword, \emph{lookup} for the \ident{?p} syntax and
-    \emph{letimpl} for the $\kvd{let}~\ident{?p}=\ldots$ notation).
+    (the \kvd{prev} keyword, lookup function for the \ident{?p} syntax and operation that models
+     implicit parameter binding for the $\kvd{let}~\ident{?p}=\ldots$ notation).
 \end{itemize}
@@
 all concrete coeffect domain-specific languages. The syntax and typing rules of the second part
-(comonadically-inspired) primitives are also shared by all coeffect DSLs, however the \emph{reduction
+(comonadically-inspired primitives) are also shared by all coeffect DSLs, however the \emph{reduction
 rules} for the comonadically-inspired primitives differ -- they capture the concrete notions of
@@
 Together with the evaluation rules shown in  Figure~\ref{fig:semantics-target}, this captures the
-standard call-by-name semantics of the common parts of the target language. The (standard) typing rules
+standard call-by-value semantics of the common parts of the target language. The (standard) typing rules
 for the common expressions of the target language are also shown in Figure~\ref{fig:semantics-target}.
@@
   For (1), the last typing rule must have been (\emph{num}); for (2), it must have been
-  (\emph{abs}) and for (3), the last typing rule must have been (\emph{tup})
+  (\emph{abs}) and for (3), the last typing rule must have been (\emph{tup}).
 \end{proof}
@@
 \label{thm:semantics-fp-pres}
-  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$
+  If $\Gamma \vdash e : \tau$ and $e \rightarrow e'$ then $\Gamma \vdash e' : \tau$.
 \end{theorem}
@@
 \label{thm:semantics-fp-prog}
-  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
+  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$.
 \end{theorem}
@@
 
-In this chapter, we use the same structure in a different way. Rather than treating the rules as
+In this section, we use the same structure in a different way. Rather than treating the rules as
 \emph{denotation} in categorical sense, we treat them as \emph{translation} from a source
@@
 The additional expressions such as $\ident{counit}_\cunit$ and $\ident{cobind}_{\cclrd{r,s}}$
-correspond to the operations of indexed comonads. Note that the we embed the coeffect annotations
+correspond to the operations of indexed comonads. Note that we embed the coeffect annotations
 into the target language -- these are known when translating a term with a chosen typing derivation
@@
 In Section~\ref{sec:semantics-proofs}, we will prove safety properties of well-typed programs in
-the target language. Thanks to the fact that the translation produces a well-typed program
+the target language. The fact that the translation produces a well-typed program
 means that we are also proving safety of well-typed programs in the source context-aware languages.
@@
 The language defined in Figure~\ref{fig:semantics-target} and Figure~\ref{fig:semantics-ext}
-provide a general structure that we now use to prove the safety of various context-aware
+provides a general structure that we now use to prove the safety of various context-aware
 programming languages based on the coeffect language framework. As examples, we consider a language
@@
 \begin{theorem}[Safety of context-aware dataflow language]
+\label{thm:semantics-sound-df}  
   If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^{*} e'$ then either $e'$ is a value of type $\tau$ or
@@
 \paragraph{Properties.}
-We now prove the type safety of of a context-aware programming language with implicit parameters.
+We now prove the type safety of a context-aware programming language with implicit parameters.
 To do this, we prove safety of the target functional language with specific extensions for implicit
@@
 \begin{theorem}[Safety of context-aware language with implicit parameters]
+\label{thm:semantics-sound-impl}
   If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^* e'$ then either $e'$ is a value of type $\tau$ or
@@
 operations accessing the context do not get stuck. In case of dataflow, $\ident{prev}_\cclrd{n}$
-can always access the tail of the stream and and $\ident{counit}_\cunit$ can always access the
+can always access the tail of the stream and $\ident{counit}_\cunit$ can always access the
 head (because the stream has a sufficient number of elements). In case of implicit parameters,
@@
 To give the type of \ident{cobind}, we need a type-level function that encodes the operations
-of the flat coeffect algebra. We model $\cseq$ as $\ident{Seq}~a~b$. The operation is defined
+of the flat coeffect algebra. We model $\cseq$ as $\ident{Seq}$. The operation is defined
 on types $\ident{L}$ and $\ident{D}$ and returns a type $\ident{D}$ if and only if both its
@@
 system of the target language. The equivalent of the progress property is guaranteed by the
-fact that the the implementation of the operations is well-defined.
+fact that the implementation of the operations is well-defined.
 
@@
 
-\subsection{When is coeffect not a monad}
+\subsection{When is a coeffect not a monad}
 \label{sec:semantics-related-monad}
@@
 \end{equation*}
+\qed
 \end{proof}
@@
 \noindent
-This equivalence shows an intriguing relationship between the \emph{product comonad} and
-\emph{reader monad}, but it cannot be extended beyond that. In particular, comonads that model
-dataflow computations or liveness do not have a corresponding monadic structure.
-This equivalence holds for monads and comonads (as well as \emph{indexed} monads
+Orchard \cite{comonads-dom-thesis} calls this intriguing relationship between the 
+\emph{product comonad} and \emph{reader monad} an \emph{equipotence} and notes that it 
+arises from adjunction between the monad and the comonad.
+
+The equivalence holds for monads and comonads (as well as \emph{indexed} monads
 and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
@@
 that turns programs written in a domain-specific coeffect language into a functional target
-language. This is akin to the Haskell ``do'' notation for monads. Finally, we extended the target
+language. Finally, we extended the target
 language with concrete implementations of comonadic operations for dataflow and implicit parameters
 and we presented a syntactic safety proof. In summary, the proof states that well-typed
-context-aware programs written in a coeffect language \emph{do not go wrong} (when translated
+context-aware programs written in a coeffect language \emph{do not get stuck} (when translated
 to a simple functional language and evaluated).

### 2017-01-17 text/Chapter06.tex
Edits and tweaks (Chapter 8 and 9 mostly)
Edits pages 95-124

@@
 programs in a simple target functional language. We prove syntactic safety for a sample target
-language, showing that ``well-typed context-aware programs do not go wrong''.
+language, showing that ``well-typed context-aware programs do not get stuck''.
 
@@
   target language for one of our sample languages and show that well-typed programs, produced by
-  translation from the coeffect calculus do not go wrong.
+  translation from the coeffect calculus do not get stuck.
 \end{itemize}
@@
 rule for \kvd{prev} produces well-typed expressions. Finally, we extend the type preservation
-(Theorem~\ref{thm:structural-fp-pres}) and progress (Theorem~\ref{thm:structural-fp-prog}) theorems.
+(Theorem~\ref{thm:semantics-fp-pres}) and progress (Theorem~\ref{thm:semantics-fp-prog}) theorems.
 

### 2017-01-22 main.tex
Edits pages 125-187

@@
 \usepackage{enumitem}
-%\usepackage{catchfilebetweentags}
 
@@
 \newcommand{\SP}{ {\textcolor{sclr}{P}} }
-\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
+\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleright P}} }
 \newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
@@
 
-  In further work (Chapter~\ref{ch:further}), we outline unification of flat and structural
+  In Chapter~\ref{ch:further}, we outline unification of flat and structural
   coeffects (Section~\ref{sec:further-unified}), which may be more suited for embedding in practical
   languages and we discuss alternative approaches for using coeffects in programming languages
-  (\ref{sec:further-meta}).
+  (Section~\ref{sec:further-meta}).
 }

### 2017-01-22 text/Chapter06.tex
Edits pages 125-187

@@
 that captures per-variable contextual properties. We outline its categorical semantics and use it as
-a basis for a translation that turns well-typed programs in context-aware languages into well-typed
+the basis for a translation that turns well-typed programs in context-aware languages into well-typed
 programs in a simple target functional language. We prove syntactic safety for a sample target
@@
   target language for one of our sample languages and show that well-typed programs, produced by
-  translation from the coeffect calculus do not get stuck.
+  translation from the coeffect calculus, do not get stuck.
 \end{itemize}
@@
 context), but just a single result (the resulting value). So, while there is only one notion of
-monadic effect system, there are two separate notions of coeffect system - one that keeps coeffect
+monadic effect system, there are two separate notions of coeffect system -- one that keeps coeffect
 annotations per-environment and one that keeps coeffect annotations per-variable.
@@
 In the structural coeffect calculus, functions are annotated with a primitive (scalar) coeffect
-annotations. A vector of variables forming the free-variable context is annotated with a vector of
+annotation. A vector of variables forming the free-variable context is annotated with a vector of
 coeffect annotations. These annotations differ for various coeffect calculi; their properties are
-captured by the definition of \emph{structural coeffect scalar} below. The scalar annotations can
+captured by the definition of \emph{structural coeffect algebra} below. The scalar annotations can
 be \eg~integers (how many past values we need) or values of a two-point lattice specifying whether
@@
 (Section~\ref{sec:structural-exts-let}). Second, the coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$
-on function type now range over values of a \emph{structural coeffect scalar}.
+on function type now range over values of a \emph{structural coeffect algebra}.
 
@@
 
-The \emph{structural coeffect scalar} structure is similar to that of \emph{flat coeffect algebra}
+The \emph{structural coeffect algebra} structure is similar to that of \emph{flat coeffect algebra}
 with the exception that it drops the $\czip$ operation. It only provides a monoid $(\C, \cseq, \cunit)$
@@
 \label{def:structural-scalar}
-A \emph{\cclrd{structural coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
+A \emph{\cclrd{structural coeffect algebra}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
 $\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
@@
 \noindent
-The structural coeffect scalar structure resembles flat coeffect algebra, but it differs in two important ways:
+The structural coeffect algebra follows the flat coeffect algebra and does not require symmetry of 
+$\cpar$ and the annihilation law ($\czero\,\cseq\,\cclrd{r} = \czero$), although both of the properties
+hold for the examples in Section~\ref{sec:struct-calculus-examples}. The structural coeffect algebra 
+differs from flat coeffect algebra in two important ways:
 %
 \begin{itemize}
-\item The $\cpar$ operation of structural coeffect scalar is not required to be idempotent. In structural
+\item The $\cpar$ operation of structural coeffect algebra is not required to be idempotent. In structural
   systems, we can track individual variable accesses and not requiring idempotence allows interesting
@@
 \noindent
-In the structural coeffect calculus, the scalar coeffect structure is supplemented by a vector
+In the structural coeffect calculus, the scalar coeffect algebra is supplemented by a vector
 structure. The vector structure is used to manipulate vectors of coeffect scalars attached to a
@@
 \begin{definition}
-A \emph{\aclrd{structural coeffect algebra}} is formed by a structural coeffect scalar
+A \emph{\aclrd{structural coeffect vector}} is formed by a structural coeffect algebra
 $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ equipped with the following additional structures:
@@
 \item Coeffect vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$, ranging over
-  structural coeffect scalars indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.
+  elements of the structural coeffect algebra indexed by vector lengths $\sclrd{m}, \sclrd{n} \in \sclrd{\mathbb{N}}$.
 
@@
 checks the types of sub-expressions in disjoint parts of the free variable contexts and
-(\emph{var}) requires the context to contain exactly one variable. The typing rule for let binding in
-structural coeffects is a derived rule and we discuss it later in Section~\ref{sec:structural-exts-let}.
-
-~
+(\emph{var}) requires the context to contain exactly one variable. The typing rule for let binding
+is a derived rule and we discuss it later in Section~\ref{sec:structural-exts-let}.
 
@@
 system for the structural coeffect calculus, the variable context is treated as a vector, with
-an additional condition that a cannot appear multiple times. We also write $\xlen{-}$ for the
+an additional condition that a variable cannot appear multiple times. We also write $\xlen{-}$ for the
 length of the vector:
@@
 
-In the typing rules, free variable contexts are annotated with vectors of structural coeffect scalars,
-written as $\coctx{x_1\!:\tau_1,\ldtos,x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}$.
+In the typing rules, free variable contexts are annotated with structural coeffect vectors,
+written as $\coctx{x_1\!:\tau_1,\ldots,x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}$.
 Meta-variables ranging over coeffect vectors are written as $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
@@
 These are shown in Figure~\ref{fig:struct-types} (b). The three structural rules are not
-syntax-directed and allow different transformation of the free variable context. They correspond
+syntax-directed and allow different transformations of the free variable context. They correspond
 to the transformations known as \emph{weakening}, \emph{exchange} and \emph{contraction}
@@
 This suggests that $\czip$ and $\cpar$ serve two roles in flat coeffects. First, they are used
-as over-approximations and under-approximations of $\atimes$. This is demonstrated by the
+as under-approximations and over-approximations of $\atimes$. This is demonstrated by the
 (\emph{approximation}) requirement introduced in Section~\ref{sec:flat-syntax-cbv}, which requires that
@@
 \subsection{Examples of structural coeffects}
-
-The structural coeffect calculus above can be instantiated to obtain the 3 structural coeffect calculi
+\label{sec:struct-calculus-examples}
+
+The structural coeffect calculus above can be instantiated to obtain the three structural coeffect calculi
 presented in Section~\ref{sec:applications-structural}. Two of them -- structural dataflow and
@@
 because there is no appropriate $\czip$ operator that could be added to form the flat coeffect
-scalar. This is the case, for example, for the system tracking bounded variable use
+algebra. This is the case, for example, for the system tracking bounded variable use
 (Example~\ref{ex:structural-ex-bll}).
@@
 \begin{example}[Structural liveness]
-The structural coeffect scalar for liveness is formed by
+The structural coeffect algebra for liveness is formed by
 $(\mathcal{L}, \sqcap, \sqcup, \ident{L}, \ident{D}, \sqsubseteq)$, where $\mathcal{L}=\{ \ident{L}, \ident{D} \}$ is
@@
 \begin{example}[Structural dataflow]
-In dataflow, context is annotated with natural numbers and the structural coeffect scalar is formed
+In dataflow, context is annotated with natural numbers and the structural coeffect algebra is formed
 by $(\mathbb{N}, +, \mathit{max}, 0, 0, \leq)$.
@@
 These two examples have both flat and structural versions. For them, obtaining the structural
-coeffect algebra is easy. As shown by the examples above, we simply omit the $\czip$ operation. The
-laws required by a structural coeffect algebra are the same as those required by the flat version
+coeffect vector is easy. As shown by the examples above, we simply omit the $\czip$ operation. The
+laws required by a structural coeffect vector are the same as those required by the flat version
 and so the above definitions are both valid. Similar construction can be used for the
@@
 
-It is important to note that this gives us a systems with \emph{different} properties.
+It is important to note that this gives us a system with \emph{different} properties.
 Information is now tracked per-variable rather than for entire contexts. For dataflow,
@@
 Before looking at properties of structural coeffect systems,
-we consider a system for tracking bounded variable use, which is an example of structural system
+we consider a system for tracking bounded variable use, which is an example of a structural system
 that does not have a flat counterpart.
@@
 
-However, the type system for structural coeffect calculus in Figure~\ref{fig:struct-types}
+However, the type system for the structural coeffect calculus in Figure~\ref{fig:struct-types}
 introduces another kind of ambiguity due to the fact that non-syntax-directed structural rules
-can be applied repeatedly and in arbitrary order. As with the semantics for flat coeffect calculus
+can be applied repeatedly and in arbitrary order. As with the semantics for the flat coeffect calculus
 in Chapter~\ref{ch:semantics}, we define the semantics of the structural coeffect calculus relative to
@@
 As with $\ident{findvar}_{x,\tau}$, the \ident{mergevars} function is fully deterministic.
-It returns context consisting of three parts. Parts $\Gamma_1$
+It returns a context consisting of three parts. Parts $\Gamma_1$
 and $\Gamma_2$ represent variables that appear only in the first or the second context; part
@@
 %
-This is not the case for the \emph{structural coeffect scalar} structure. For example, in the
+This is not the case for the \emph{structural coeffect algebra} structure. For example, in the
 calculus for tracking bounded reuse, the $\cpar$ operator is defined as $+$ (on integers) and
@@
 coeffect calculus that we consider in this thesis. In this section, we look at these common
-properties. In Section~\ref{sec:struct-syntax-intro}, we first briefly compare equational theory
+properties. In Section~\ref{sec:struct-syntax-intro}, we first briefly compare the equational theories
 for flat coeffects (Section~\ref{sec:flat-syntax}) and structural coeffects (Section~\ref{sec:struct-syntactic-thms}).
@@
 The following example shows a situation similar to the previous one. Here, we use
-structural dataflow calculus (writing $\kvd{prev}~e$ to obtain previous value of the
+structural dataflow calculus (writing $\kvd{prev}~e$ to obtain the previous value of the
 expression $e$) and we substitute $w+z$ for $y$:
@@
 The substitution applies the operation $\aclrd{\ast}$ (which stands for the sequential
-composition $\aseq$ from the structural coeffect algebra) to the annotation of the new context --
+composition $\aseq$ from the structural coeffect vector) to the annotation of the new context --
 in the above example the coeffect $\cclrd{2}\cclrd{\ast}\alift{\cclrd{1}, \cclrd{1}}$ (\emph{after}) is
@@
 $(\lambda x.x+x)~(y+z)$ we need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
-This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
+This is similar to substitution lemmas in other structural variants of $\lambda$-calculus,
 such as the bunched typing system \cite{substruct-bunched}. To express the substitution
@@
 \label{thm:structural-substitution}
-In a structural coeffect calculus with a \emph{structural coeffect scalar} such that
+In a structural coeffect calculus with a \emph{structural coeffect algebra} such that
 $\cclrd{r}\,\cleq\,\cclrd{r'} \, \Rightarrow \, \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s})$
@@
 
-The Lemma~\ref{thm:structural-substitution} has two additional requirements on the structural coeffect
+Lemma~\ref{thm:structural-substitution} has two additional requirements on the structural coeffect
 scalar that are similar to those of the substitution lemma for bottom-pointed flat coeffect systems
 (Lemma~\ref{thm:cbn-substitution-bot}). Those two conditions are satisfied for all our examples
-(it would be reasonable to require them for \emph{all} structural coeffect scalars, but we prefer
+(it would be reasonable to require them for \emph{all} structural coeffect algebras, but we prefer
 to keep the original definition more general). The two requirements are needed in the proof
@@
 
-In the Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
+In Chapter~\ref{ch:flat}, we discussed call-by-value separately from call-by-name, because the
 proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
@@
 $\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$
-arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
+arising from the monoid $(\C, \cseq, \cunit)$ of the coeffect algebra.
 \end{proof}
@@
 %
-The \ident{merge} and \ident{split} operations are defined as in Example~\ref{ex:struct-dfl}.
+The \ident{merge} and \ident{split} operations are defined as in Example~\ref{thm:struct-dfl}.
 The operations that capture the behaviour of bounded reuse are:
@@
 The \ident{counit} operation is defined as previously -- it extracts the only value of the only
-variable In the bounded variable reuse system, variable sharing is annotated with the $+$ operator
+variable. In the bounded variable reuse system, variable sharing is annotated with the $+$ operator
 (in contrast with \textit{max} used in dataflow). The $\ident{dup}_{m,n}$ operation thus splits the $m+n$
@@
 \label{thm:structural-df-pres}
-  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow e'$ then $\Gamma \vdash e' : \tau$
+  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow e'$ then $\Gamma \vdash e' : \tau$.
 \end{theorem}
@@
 \label{thm:structural-df-prog}
-  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
+  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$.
 \end{theorem}

### 2017-01-22 text/Chapter07.tex
Edits pages 125-187

@@
 In order to obtain unique type derivation for each term (using the algorithms discussed in
-Section~\ref{sec:flat-unique}), we generate additional coeffect constraints for lambda abstraction
+Section~\ref{sec:flat-unique}), we generate additional coeffect constraints for each lambda abstraction
 of each flat coeffect language, as this is where flat coeffects permit multiple possible typings.
@@
 \noindent
-The list mirrors a list of steps that need to be done when supporting a new effectful computation
-in a language that supports monadic ``do'' notation. The step (3) corresponds to implementing a new
-monad and (4) corresponds to adding monad-specific effectful operations. The step (2) applies when
-using indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
+The list is akin to the steps that are needed when supporting a new effectful computation in a 
+language that supports monadic ``do'' notation, with the exception that in step (4), we can also
+modify the parser and AST, as opposed to just defining primitive effectful functions.
+This could be addressed by making the host language more extensible and flexible. 
+The step (3) corresponds to implementing a new monad. The step (2) applies when using 
+indexing to track effects more precisely \cite{effects-embedding}. The only step that does not
 have a counterpart in effectful/monadic languages is (1).
@@
 
-The idea of active reading is older and has been encouraged in the context of art Josef Albers'
+The idea of active reading is older and has been encouraged in the context of art in Josef Albers'
 classic 1963 work on color \cite{essay-albers} (which was turned into an interactive essay
@@
 and sociology \cite{essay-polygons} (visualizing and explaining game theoretical model of segregation
-in the society \cite{essay-segregation}). To our best knowledge, the format has not previously
+in society \cite{essay-segregation}). To our best knowledge, the format has not previously
 been used in the area of programming language theory and so we briefly outline some of the
@@
 interactive essay shows the reader the two sample context-aware programming languages. Readers can
-write code in a panel that type checks the input, generates user interface for entering the
+write code in a panel that type checks the input, generates a user interface for entering the
 required context and runs the sample code.
@@
 understanding the translation based just on the rules is harder than for monads. The essay lets
-readers see translation for carefully chosen case studies that illustrate the key aspects of the
+readers see translations for carefully chosen case studies that illustrate the key aspects of the
 translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
@@
 checkers and interpreters for statically-typed functional programming languages. Two interesting
-aspects that are worth discussing is the JavaScript targetting (for running the language
+aspects that are worth discussing are the JavaScript targetting (for running the language
 implementation in a web browser) and integration with client-side (JavaScript) libraries for
@@
 
-The implementation discussed in this chapter provides an evidence for our claim that the
+The implementation discussed in this chapter provides evidence for our claim that the
 theory of coeffects can be used as a basis for a wide range of sound context-aware programming

### 2017-01-22 text/Chapter08.tex
Edits pages 125-187

@@
-\chapter{Further work}
+\chapter{Unified coeffect systems}
 \label{ch:further}
@@
 \item We consider a way of implementing coeffects based on embedding comonadic computations
-  via a lightweight syntactic extension and a rich type systems akin to Haskell's ``do'' notation
+  via a lightweight syntactic extension and a rich type system akin to Haskell's ``do'' notation
   (Section~\ref{sec:further-impl}).
@@
 lists)\footnote{Shapes can be intuitively thought of as sizes, but this is not precise
-as some containers have multiple shapes of the same length, \eg~for perfectly ballanced, left- and right-leaning trees.}.
+as some containers have multiple shapes of the same length, \eg~for perfectly balanced, left- and right-leaning trees.}.
 For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
@@
 \begin{definition}
-A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
+A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleright P$.
 Here, $S$ is a set of shapes and $P$ is a shape-indexed family of sets such that
-for each shape $s\in S$, $P_s$ (also written as $P(s)$) is a set of positions for shape $s$.
+for each shape $s\in S$, $P_s$, also written as $P(s)$, is a set of positions for shape $s$.
 \end{definition}
@@
 \begin{itemize}
-\item The container representing lists is given by $S \triangleleft P$ where
+\item The container representing lists is given by $S \triangleright P$ where
   shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a
@@
 
-\item The container representing the singleton data type is given by $S \triangleleft P$ where
+\item The container representing the singleton data type is given by $S \triangleright P$ where
   shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
@@
 \paragraph{Coeffect containers.}
-The coeffect container is a container that determines how are scalar coeffects attached to free-variable
+The coeffect container is a container that determines how scalar coeffects are attached to free-variable
 contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed sets of positions,
@@
 $\SP_{\sclrd{l}}$ is the \emph{set} of positions for the given shape. A coeffect $\aclrd{\textbf{r}}$
-can now be seen as a function $\SP_{\sclrd{l}} \rightarrow \C$ that returns a scalar coefect
+can now be seen as a function $\SP_{\sclrd{l}} \rightarrow \C$ that returns a scalar coeffect
 $\cclrd{r}\in\C$ for each of the position defined by the shape $\sclrd{l}$.
@@
 the term language is that of a simply-typed $\lambda$-calculus, and we used (flat or structural)
-indexed comonads to give the semantics for the language and to derive type system for it.
+indexed comonads to give the semantics for the language and to derive a type system for it.
 
@@
 to represent the entire variable context of the source language. The embedding resembles the semantics discussed in
-Section~\ref{sec:flat-semantics}. This is not surprising as the meta-language directly mirrors
+Section~\ref{sec:semantics-translation}. This is not surprising as the meta-language directly mirrors
 the operations of an indexed comonad.
@@
 work in the area. First, we look at a way of embedding coeffects in existing languages
-such as Haskell that is inspired by the Haskell's ``do'' notation for monads
+such as Haskell that is inspired by Haskell's ``do'' notation for monads
 (Section~\ref{sec:further-impl}). Second, we discuss extensions to the theory of coeffects that
@@
 In order to do this, the target programming language needs to provide two features; one that allows
-easiler embedding of context-aware computations themselves in programs akin to the ``do'' notation
-for monadsin Haskell and one that allows tracking of the contextual information in the type system.
+easier embedding of context-aware computations themselves in programs akin to the ``do'' notation
+for monads in Haskell and one that allows tracking of the contextual information in the type system.
 
@@
 blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
-that differ from the ``do'' notation in two ways. First, they support wider range of
+that differ from the ``do'' notation in two ways. First, they support a wider range of
 constructs, making it possible to wrap existing F\# code within a block without other changes.
@@
 
-Coeffect annotations in context-aware programming languages can contain significant amount
+Coeffect annotations in context-aware programming languages can contain a significant amount
 of information. Thus, the programming language also needs a form of type inference that can propagate
@@
 Orchard~\cite{comonads-dom-thesis}. A joint work with Orchard~\cite{coeffects-icfp14} shows the
-first steps for structural systems. As discussed in Section~\ref{sec:unified-semantics}, further
+first steps for structural systems. As discussed in Section~\ref{sec:further-unified}, further
 work is to develop a similar categorical model for the unified coeffect calculus, possibly using
@@
 In this chapter, we looked at three alternative directions for further development of coeffects.
-The key technical contribution of this chapter is the \emph{unified coeffect system}
-(Section~\ref{sec:further-unified}), which unifies the flat and structural system presented
+The \emph{unified coeffect system} (Section~\ref{sec:further-unified}) is a novel technical 
+contribution that unifies the flat and structural system presented
 in the previous two chapters. This provides an alternative to the separate treatment of
 per-context and per-variable contextual dependencies at the cost of making the system more
-complex. To unify the two, we introduced \emph{coeffect container}, which determines how are
-coeffect annotations attached to variable contexts. We then discussed two
+complex. To unify the two, we introduced \emph{coeffect container}, which determines how 
+coeffect annotations are attached to variable contexts. We then discussed two
 instances of the structure that capture flat and structural properties.

### 2017-01-22 text/Chapter09.tex
Edits pages 95-124
Edits pages 125-187

@@
 journalists obtain information published on the web or through open government data initiatives;
-in dataflow programming, we have access to past values of expressions.
-This thesis aims to change our understanding of \emph{context} so that the above examples are
-viewed uniformly through a single programming language abstraction, which we call \emph{coeffects},
-rather than as disjoint cases.
+in dataflow programming, we have access to past values of expressions. The notion of \emph{coeffect}
+provides a single programming language abstractions that allows us to view the above examples as
+instances of the same notion of context-aware properties, rather than as disjoint cases.
 
@@
 well-typedness of our translation, these guarantee that ``well-typed context-aware programs do
-not go wrong''.
+not get stuck''.
 
@@
 
-The implementation (Section~\ref{ch:impl-theory}) serves two purposes. First, it links together all parts
+The implementation serves two purposes. First, it links together all parts
 of the theory developed in this thesis (Section~\ref{sec:impl-case}). It uses the flat and structural
@@
 
-Second, web-based essay (\url{http://tomasp.net/coeffects}) discussed in Section~\ref{sec:impl-essay}
+Second, the web-based essay (\url{http://tomasp.net/coeffects}) discussed in Section~\ref{sec:impl-essay}
 is novel from the pedagogical perspective. It makes our work \emph{accessible} and \emph{explorable}.

### 2017-01-22 text/Chapter02.tex
Edit pages 188-208

@@
 The theoretical parts of this thesis extend the work of Uustalu and Vene who use comonads
-to give the semantics of dataflow computations \cite{comonads-dataflow} and more generally,
+to give the semantics of dataflow computations \cite{app-dataflow-essence} and more generally,
 notions of \emph{context-dependent computations} \cite{comonads-notions}. The computations discussed

### 2017-01-23 text/Chapter06.tex
Structural semantics edits

@@
-\chapter{The structural coeffect calculus}
+\chapter{Structural coeffect calculus}
 \label{ch:structural}
@@
 
-\subsection{Semantics of vectors}
-
-Recall that in the flat coeffect calculus, the context is interpreted as a product and so a typing
-judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\cclrd{r}} \vdash e : \tau$ is interpreted as a
-morphism $\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$. In this model,
-we can freely transform the value contained in the context modelled using an indexed comonad
-$\ctyp{\cclrd{r}}$.
-
-Previously, we defined the $\ident{map}$ function (in terms of $\ident{cobind}$ and $\ident{counit}$),
-which transforms the value inside the context without affecting the coeffect annotation. Thus,
-we can use $\ident{map}_{\cclrd{r}}~\pi_i$ to transform a context containing product of variables
-$\ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)}$ into a context containing a single value
-$\ctyp{\cclrd{r}}{\tau_i}$. This changes the carried value without affecting the coeffect $\cclrd{r}$.
-
-The ability to freely transform the variable structure is not desirable in the model of
-structural coeffect systems. Our aim is to guarantee (by construction) that the structure of
-the coeffect annotations matches the structure of variables. To achieve this, we model vectors
-using a structure distinct from ordinary products which we denote $-\xtimes-$.
-For example, the judgement $\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}} \vdash e : \tau$
-is modelled as a morphism $\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$.
-We assume that the operator is equipped with necessary associativity transformations allowing us
-to use it freely on more than two values.
-
-The operator is a bifunctor, but it is \emph{not} a product in the categorical sense. In particular,
-there is no way to turn $\tau_1 \xtimes \ldots \xtimes \tau_n$ into $\tau_i$ (the structure does
-not have projections) and so there is also no way of turning
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$ into
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{\tau_i}$, which would break the correspondence
-between coeffect annotations and variable structure.
-
-The structure created using $-\xtimes-$ can be manipulated only using operations provided by
-the \emph{structural indexed comonad}, which operate over variable contexts contained in an
-indexed comonad $\ctyp{\aclrd{\textbf{r}}}$ and are designed to preserve the correspondence
-between vectors and annotations.
-
-In what follows, we model (finite) vectors of length $n$ as $\tau_1 \xtimes \ldots \xtimes \tau_n$.
-As mentioned, we assume that the use of the operator can be freely re-associated. For example, when
-calling an operation that requires input of the form
-$(\tau_1 \xtimes \ldots \xtimes \tau_i) \xtimes (\tau_{i+1} \xtimes \ldots \xtimes \tau_n)$,
-we use an argument $(\tau_1 \xtimes \ldots \xtimes \tau_n)$ and assume that the
-appropriate transformation is inserted.
-
-% --------------------------------------------------------------------------------------------------
-
 \subsection{Indexed comonads, revisited}
@@
 representing function values (appearing in the language), but also of the form
-$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau$,
+$\ctyp{\alift{\cclrd{r_1},\ldots,\cclrd{r_n}}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau$,
 modelling expressions in a context. To capture this, we need to revisit the definition and use
@@
 \end{compactitem}
+such that, for all $f:\ctyp{\aclrd{\textbf{t}}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\alift{\cclrd{r}}}{\beta} \rightarrow \gamma$:
+\begin{align}
+\tag{\emph{left identity}}
+  \ident{cobind}_{\alift{\cclrd{s},\cunit}}~\ident{counit}_{\cunit} &= \idf{}
+  \\
+\tag{\emph{right identity}}
+  \ident{counit}_{\cunit} \circ \ident{cobind}_{\cunit, \aclrd{\textbf{t}}}~f &= f
+  \\
+\tag{\emph{associativity}}
+\hspace{-10em}
+  \ident{cobind}_{\cclrd{s},\cclrd{r}\aseq\aclrd{\textbf{t}}}~(g \circ \ident{cobind}_{\cclrd{r}, \aclrd{\textbf{t}}}~f) &=
+    (\ident{cobind}_{\cclrd{s}, \alift{\cclrd{r}}}~g) \circ (\ident{cobind}_{\cclrd{s}\cseq\cclrd{r},\aclrd{\textbf{t}}}~f)
+\end{align}
 \end{definition}
@@
 $\ctyp{\cclrd{r}}$ as in the previous chapter. This new definition supersedes the old one, because a
-flat coeffect annotation can be seen as singleton vectors.
+flat coeffect annotation can be seen as singleton vectors. The comonad laws are adjusted accordingly.
 
 The operation $\ident{counit}_{\cunit}$ operates on a vector of length one. This means that it will always
-return a single value rather than a vector created using $-\xtimes-$. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
+return a single value rather than a vector. The $\ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}$
 operation is, perhaps surprisingly, indexed by a coeffect vector and a coeffect scalar. This asymmetry
@@
 also takes a vector of variables, but always returns a context with a vector containing just one value.
-In other words, $\alpha$ may contain $\xtimes$, but $\beta$ may not, because the coeffect calculus has
-no way of constructing values containing $\xtimes$.
+
+In the definition of \emph{indexed comonad} given in this section, we use a vectors of scalar 
+coeffect annotations to match with the type system given in Section~\ref{sec:struct-calculus}. 
+A more general definition could be given in terms of modules over coeffect algebra, which
+provides a more general definition using standard structures of category theory. However,
+for the purpose of defining the translation in Section~\ref{sec:struct-transl}, the concrete
+definition using vectors of coeffect annotations suffices.
 
@@
 
-In the structural version, the corresponding operations operate convert between a (wrapped) vector
-of values represented using $\xtimes$ and ordinary pairs of contexts containing parts of the vector.
+In the structural version, the corresponding operations convert between a vector
+of values and pairs of contexts containing parts of a vector.
 The vectors of coeffect annotations are split or merged using $\atimes$ of the structural coeffect
-algebra, in a way that mirrors the wrapped vectors (variable structure).
+algebra, in a way that mirrors the variable structure.
 
@@
 \item $\ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
-  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)}$
+  $\ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta} \rightarrow \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \times \beta)}$
 \item $\ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}$ is a family of mappings
-  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \xtimes \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
+  $\ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\alpha \times \beta)} \rightarrow \ctyp{\aclrd{\textbf{r}}}{\alpha} \times \ctyp{\aclrd{\textbf{s}}}{\beta}$
 \item $\ident{dup}_{\cclrd{r}, \cclrd{s}}$ is a family of mappings
-  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \xtimes \alpha)}$
+  $\ctyp{\alift{\cclrd{r} \cpar \cclrd{s}}}{\alpha} \rightarrow \ctyp{\alift{\cclrd{r}, \cclrd{s}}}{(\alpha \times \alpha)}$
 \end{compactitem}
@@
 information about the context. This was not required for the flat system where the operations
-could under-approximate or over-approximate. Note that the operations use $\xtimes$ to combine
-or split the contained values. This means that they operate on free-variable vectors rather than
-on ordinary products.
+could under-approximate or over-approximate.
 
@@
 duplicates the value of the variable $\alpha$ and splits the additional context between the two
-new variables. In a flat calculus, this operation was expressed using ordinary tuple
-construction, which is not possible here -- the returned context needs to contain a
-two-element vector $\alpha \xtimes \alpha$.
+new variables.
 
@@
 In the structural coeffect calculus, expressions in context are interpreted as functions taking
-a vector (represented using $-\xtimes-$) wrapped in a structure indexed with a vector of annotations
+a vector wrapped in a structure indexed with a vector of annotations
 such as $\ctyp{\aclrd{\textbf{r}}}$. Functions take only a single variable as an input and so the
@@
 \sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau}
-  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)} \rightarrow \tau\\
+  &:& \ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
 \sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\alift{\cclrd{r}}}{\tau_1} \rightarrow \tau_2
@@
 structural indexed comonads -- the definitions shown in Section~\ref{sec:struct-semantics-examples}
-are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \xtimes \ldots \xtimes \tau_n)}$
-rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$. The need to examine the structure
-of the variable context is another reason for using $-\xtimes-$ when interpreting expressions in contexts.
+are given specifically for $\ctyp{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} }{(\tau_1 \times \ldots \times \tau_n)}$
+rather than more generally for $\ctyp{\aclrd{\textbf{r}}}{\alpha}$.
 
@@
   applying $\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}$ to the semantics of $e_2$. The result
-  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \xtimes \ldots \xtimes \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
+  $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}{(\ldots \times \ldots \times \ldots)} \rightarrow \ctyp{\alift{\cclrd{t}}}{\tau_1}$
   is then called with the latter part of the split input context.
@@
 is indexed by required number of past variables for each individual variable. It is defined
-over a vector of variables $\alpha_1 \xtimes \ldots \xtimes \alpha_k$ and it keeps a product
+over a vector of variables $\alpha_1 \times \ldots \times \alpha_k$ and it keeps a product
 containing a current value followed by $n_i$ past values:
@@
 \begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} =
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \times \ldots \times \alpha_k)} =
 \underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{(n_1+1)-\textnormal{times}} \times \ldots \times
@@
 The definition of the indexed list data structure relies on the fact that the number of annotations
-corresponds to the number of variables combined using $-\xtimes-$. It then creates a vector of lists
+corresponds to the number of variables combined using $-\times-$. It then creates a vector of lists
 containing $n_i+1$ values for the $i$-\textit{th} variable (the annotation represents the number of
@@
 \begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} =
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \times \ldots \times \alpha_k)} =
 \underbrace{(\alpha_1 \times \ldots \times \alpha_1)}_{n_1-\textnormal{times}} \times \ldots \times
@@
 \begin{equation*}
-\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \xtimes \ldots \xtimes \alpha_k)} =
+\ctyp{\langle n_1, \ldots, n_k\rangle}{(\alpha_1 \times \ldots \times \alpha_k)} =
   D^{n_1} \alpha_1 \times \ldots \times D^{n_k} \alpha_k
@@
          \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~e \\
-  \tau &  = &\ldots \sep \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \\
+  \tau &  = &\ldots \sep \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \times \ldots \times \tau_k)} \\
   K  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~\_~e \sep \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
@@
 \tyrule{cobind}
-  {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
+  {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \times \ldots \times \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \times \ldots \times \tau_k)} }
   {\Gamma \vdash \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
@@
 \tyrule{merge}
-  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)} }
-  {\Gamma \vdash  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }\\
+  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \times \ldots \times \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \times \ldots \times \tau_k)} }
+  {\Gamma \vdash  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \times \ldots \times \tau_k)} }\\
 \\
 \tyrule{split}
-  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
-  {\Gamma \vdash  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)}} \\
+  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \times \ldots \times \tau_k)} }
+  {\Gamma \vdash  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \times \ldots \times \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \times \ldots \times \tau_k)}} \\
 \\
@@
   {\Gamma \vdash  e : \ctyp{\alift{\cclrd{r}\cpar\cclrd{s}}}{\tau} }
-  {\Gamma \vdash  \ident{dup}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\alift{\cclrd{r},\cclrd{s}}}{(\tau \xtimes \tau)} } \\
+  {\Gamma \vdash  \ident{dup}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\alift{\cclrd{r},\cclrd{s}}}{(\tau \times \tau)} } \\
 \end{array}
@@
 
-We also include an uninterpreted type $\ctyp{\aclrd{\textbf{r}}}{\tau_1\xtimes\ldots\xtimes\tau_k}$,
+We also include an uninterpreted type $\ctyp{\aclrd{\textbf{r}}}{\tau_1\times\ldots\times\tau_k}$,
 which models a contextual (comonadic) value indexed by a vector of annotations. As in the
@@
 In the target language, this is done by defining the $\ctyp{\aclrd{\textbf{r}}}{}$ type over
-zero or more underlying types. Here, our syntactic treatment differs slightly from the categorical
-model where objects created by $\xtimes$ were first-class values. In Figure~\ref{fig:struct-semantics-ext},
+zero or more underlying types. In Figure~\ref{fig:struct-semantics-ext},
 we need to explicitly specify the types of individual components in typing rules for
@@
 \sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}} &=&
-  \ctyp{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}{(\sem{\tau_1} \xtimes \ldots \xtimes \sem{\tau_n})} \\[0.5em]
+  \ctyp{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}{(\sem{\tau_1} \times \ldots \times \sem{\tau_n})} \\[0.5em]
 \sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} & = & \ctyp{\alift{\cclrd{r}}}{\sem{\tau_1}} \rightarrow \sem{\tau_2} \\[-0.2em]
@@
 \noindent
-The definition differs from the one for flat coeffects (Section~\ref{sec:semantics-translation-transl})
-in that the comonadically-inspired data type takes multiple type parameters (separated using $\xtimes$),
-rather than wrapping a regular tuple in the target language.
-
 The translation rules are defined in Figure~\ref{fig:struct-translation}. As in the case of the flat
@@
 $\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
-constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.
+constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\times\ldots\times\sem{\tau_n}}) \rightarrow \sem{\tau}$.
 
@@
   calling $\ident{merge}_{\aclrd{\textbf{r}},\alift{\cclrd{s}}}$ produces a context of type
-  $\ctyp{\aclrd{\textbf{r}}\atimes\alift{\cclrd{s}}}(\ldots\xtimes\tau_1)$ as expected by $f$.
+  $\ctyp{\aclrd{\textbf{r}}\atimes\alift{\cclrd{s}}}(\ldots\times\tau_1)$ as expected by $f$.
 
@@
   keeps parts of the context corresponding to coeffect annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{q}}$ unchanged and transforms
-  the nested part. In (\emph{exch}), the provided lambda function is of type $\ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau_1\xtimes\tau_2)}
-  \rightarrow \ctyp{\alift{\cclrd{t},\cclrd{s}}}{(\tau_2\xtimes\tau_1)}$. In (\emph{contr}), the type of
-  $\ident{dup}_{\cclrd{s},\cclrd{t}}$ is $\ctyp{\alift{\cclrd{s}\cpar\cclrd{t}}}{\tau} \rightarrow \ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau\xtimes\tau)}$
+  the nested part. In (\emph{exch}), the provided lambda function is of type $\ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau_1\times\tau_2)}
+  \rightarrow \ctyp{\alift{\cclrd{t},\cclrd{s}}}{(\tau_2\times\tau_1)}$. In (\emph{contr}), the type of
+  $\ident{dup}_{\cclrd{s},\cclrd{t}}$ is $\ctyp{\alift{\cclrd{s}\cpar\cclrd{t}}}{\tau} \rightarrow \ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau\times\tau)}$
   (assuming \ident{nest} is expanded, rather than treated as a value within the language).
@@
   { \forall i\in\{0\ldots k\}. ~ \Gamma \vdash_{\ident{vec}} \textbf{e}_i : \tau_i, \cclrd{n_i}}
-  { \Gamma \vdash \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle : \ctyp{\alift{\cclrd{n_0}, \ldots, \cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
+  { \Gamma \vdash \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle : \ctyp{\alift{\cclrd{n_0}, \ldots, \cclrd{n_k}}}{(\tau_0\times\ldots\times\tau_k)} }
 \end{equation*}
@@
 \tyrule{prev}
-  { \Gamma \vdash e : \ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
-  { \Gamma \vdash \ident{prev}_{\cclrd{n_0},\ldots,\cclrd{n_k}}~e : \ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
+  { \Gamma \vdash e : \ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\times\ldots\times\tau_k)} }
+  { \Gamma \vdash \ident{prev}_{\cclrd{n_0},\ldots,\cclrd{n_k}}~e : \ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\times\ldots\times\tau_k)} }
 \end{equation*}
@@
 typing rules, translation rules and reduction rules needed for modelling structural dataflow.
-In the structural model, the type $\ctyp{\alift{\cclrd{r_0}, \ldots, \cclrd{r_n}}}{(\tau_0\xtimes\ldots\xtimes\tau_n)}$
+In the structural model, the type $\ctyp{\alift{\cclrd{r_0}, \ldots, \cclrd{r_n}}}{(\tau_0\times\ldots\times\tau_n)}$
 represents a structure that provides values and additional contexts for variables of types
@@
 \vspace{0.5em}\noindent\hangindent=0.6cm
-Case (\emph{prev}): Type of $\ctx$ is $\ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)}$
-  and so we can apply the (\emph{prev}) rule to obtain $\ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)}$
+Case (\emph{prev}): Type of $\ctx$ is $\ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\times\ldots\times\tau_k)}$
+  and so we can apply the (\emph{prev}) rule to obtain $\ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\times\ldots\times\tau_k)}$
   as required by $f$.
@@
   \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
-  \item If $\tau=\ctyp{\alift{\cclrd{n_1}, \ldots, \cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)}$
+  \item If $\tau=\ctyp{\alift{\cclrd{n_1}, \ldots, \cclrd{n_k}}}{(\tau_1\times\ldots\times\tau_k)}$
     then $e = \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_n\rangle$ for some $\textbf{v}_i$
@@
 \begin{itemize}[noitemsep,leftmargin=1.5cm,topsep=0pt]
-  \item[--] $\Gamma \vdash f : \ctyp{\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)} \rightarrow \tau_2$ and
-  \item[--] $\Gamma \vdash \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_{k} \rangle : \ctyp{\alift{\cclrd{m}+\cclrd{n_1},\ldots,\cclrd{m}+\cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)}$
+  \item[--] $\Gamma \vdash f : \ctyp{\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}{(\tau_1\times\ldots\times\tau_k)} \rightarrow \tau_2$ and
+  \item[--] $\Gamma \vdash \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_{k} \rangle : \ctyp{\alift{\cclrd{m}+\cclrd{n_1},\ldots,\cclrd{m}+\cclrd{n_k}}}{(\tau_1\times\ldots\times\tau_k)}$
 \end{itemize}
@@
 % ==================================================================================================
-
-~
 

### 2017-01-24 text/Chapter08.tex
Scalars vs algebras

@@
 coeffect calculus (Definition~\ref{def:structural-scalar}). In this section, we refer to it as
-\emph{\cclrd{unified coeffect scalar}} (we repeat the definition below for clarity).
+\emph{\cclrd{unified coeffect algebra}} (we repeat the definition below for clarity).
 Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
 scalar values are attached to the free-variable context. Finally, we define the
-\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.
+\emph{\aclrd{unified coeffect label}} which consists of shape-indexed coeffect algebra values.
 
@@
 
-\paragraph{Coeffect scalars.}
-The following definition of the coeffect scalar structure repeats
+\paragraph{Coeffect algebra.}
+The following definition of the coeffect algebra structure repeats
 Definition~\ref{def:structural-scalar} from the Chapter~\ref{ch:structural}, with the only change
-that we now refer to it as \emph{\cclrd{unified coeffect scalar}}.
+that we now refer to it as \emph{\cclrd{unified coeffect algebra}}.
 
@@
 \label{def:unified-scal}
-A \emph{\cclrd{unified coeffect scalar}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
+A \emph{\cclrd{unified coeffect algebra}} $(\C, \,\cseq, \,\cpar, \,\cunit, \,\czero, \,\cleq)$ is a set
 $\C$ together with elements $\cunit, \czero \in \C$, binary operations $\cseq, \cpar$ such that
@@
 
-\paragraph{Unified coeffect algebra.}
+\paragraph{Unified coeffect label.}
 The unified coeffect calculus annotates typing judgments with shape-indexed coeffect annotations.
-The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and a coeffect container
+The \emph{\aclrd{unified coeffect label}} combines a coeffect algebra and a coeffect container
 to define shape-indexed coeffects and operations for manipulating these.
@@
 \label{def:unified-alg}
-Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
+Given a \cclrd{unified coeffect algebra} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
 \sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \aclrd{unified
-coeffect algebra} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
+coeffect label} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
 is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
@@
 \noindent
-A coeffect algebra induces the following three additional operations:
+A coeffect label induces the following three additional operations:
 %
@@
 
-The type system for the unified coeffect calculus is parameterized by a coeffect scalar
+The type system for the unified coeffect calculus is parameterized by a coeffect algebra
 $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ (Definition~\ref{def:unified-scal}) together with a
-coeffect algebra $(\azip, \apar, \azero)$ (Definition~\ref{def:unified-alg})
+coeffect label $(\azip, \apar, \azero)$ (Definition~\ref{def:unified-alg})
 and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
@@
 system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging
-of the unified coeffect algebra. As already mentioned, we use splitting of context demands $\azip$
+of the unified coeffect label. As already mentioned, we use splitting of context demands $\azip$
 in rule \emph{assumptions} and merging of context demands $\apar$ in rule \emph{conclusions}.
-
-~
 
@@
 $\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:further-flat},
-thanks to the properties of coeffect scalars, the operations of contraction, weakening and exchange
+thanks to the properties of the coeffect algebra, the operations of contraction, weakening and exchange
 that do not affect the coeffect annotation are admissible for all flat systems embedded in the
@@
 The unified calculus closely resembles the structural system and so using it to model structural
-systems is easy -- given a (structural) coeffect scalar, we use the coeffect container that describes
+systems is easy -- given a (structural) coeffect algebra, we use the coeffect container that describes
 a \emph{vector} of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect
-algebra formed by a vector (free monoid) of scalars.
+label formed by a vector (free monoid) of scalars.
 
@@
 \label{def:further-struct}
-Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system}
+Given a coeffect algebra $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system}
 is defined by:
@@
 
-\item A coeffect algebra $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
+\item A coeffect label $\aclrd{(\times, \times, \epsilon)}$ where $\times$ and $\epsilon$ are
   shape-indexed versions of the binary operation and the unit of a free monoid over $\C$.
@@
 with the Figure~\ref{fig:struct-types} and using the free monoid interpretation of the unified
-coeffect algebra.
+coeffect label.
 
@@
 $\{0\} \rightarrow \C$. The domain of the function is a singleton set and so the values are equivalent
-to coeffect scalars $\C$. In addition to the coeffect scalar structure, we need to define $\apar$
+to coeffect scalars $\C$. In addition to the coeffect algebra structure, we need to define $\apar$
 and $\azip$. Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$
@@
 \label{def:further-flat}
-Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary operation
+Given a coeffect algebra $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a binary operation
 $\cclrd{\wedge} : \C \times \C \rightarrow \C$ such that $(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$,
-the unified coeffect algebra modelling a flat coeffect systems consists of:
+the unified coeffect label modelling a flat coeffect systems consists of:
 %
@@
 
-\item A flat coeffect algebra $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$
+\item A flat coeffect label $(\czip, \cpar, \czero)$, \ie~$\apar = \cpar$
   and $\azero = \czero$ with an\\ additional binary operation $\azip = \czip$.

### 2017-04-10 main.tex
:mortar_board: :beers: :sparkles: :wine_glass: :boom:
Eliminate type 3 fonts from techreport

@@
 \newcommand{\num}[1]{\textcolor{numclr}{#1}}
-\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
-\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
-\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
+\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
+\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\ttfamily #1}}}
+\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
 \newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
-\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
-\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
+\newcommand{\ident}[1]{\textnormal{\ttfamily #1}}
+\newcommand{\cmt}[1]{\textit{\ttfamily\textcolor{cmtclr}{#1}}}
 
@@
 
-\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
-\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
+\newcommand{\czero}{ \textcolor{cclr}{ \mathtt{ign} } }
+\newcommand{\cunit}{ \textcolor{cclr}{ \mathtt{use} } }
 \newcommand{\cseq}{ \textcolor{cclr}{ \circledast }}
@@
 \newcommand{\catc}{\mathcal{C}}
-\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }
+\newcommand{\idf}[1]{ {\textnormal{\ttfamily id}}_{#1} }
 
-\urlstyle{sf}
+\urlstyle{tt}
 
@@
 \begin{document}
+\setcounter{page}{3}
 
@@
 
-\pagenumbering{roman} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
+%\pagenumbering{roman} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
 
@@
 
-\include{text/Titlepage} % Main title page
+%\include{text/Titlepage} % Main title page
 
-\cleardoublepage\include{text/Declaration} % Publications from the thesis page
+%\cleardoublepage\include{text/Declaration} % Publications from the thesis page
 
 \cleardoublepage\include{text/Abstract} % Abstract page
+
+\cleardoublepage\include{text/Acknowledgements} % Publications from the thesis page
 
@@
 
-\pagenumbering{arabic} % Arabic page numbering for thesis content (1, 2, 3, etc)
+%\pagenumbering{arabic} % Arabic page numbering for thesis content (1, 2, 3, etc)
 %\setcounter{page}{90} % Uncomment to manually start the page counter at an arbitrary value (for example if you wish to count the pre-content pages in the page count)
